### 2016-08-30 pkg/glmertree/upre/paper/prediction_rule_ensembles.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@953 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{bigfoot}
\usepackage{ctable}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{placeins}
\usepackage{listings}
\linespread{1.2}
\usepackage{tabto}

%% for \usepackage{Sweave}
\SweaveOpts{echo=FALSE, results=hide}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries>>=
library("partykit")
@

\setkeys{Gin}{width=0.8\textwidth}

\title{Prediction rule ensembles in R}
\shorttitle{Prediction rule ensembles in R}
\leftheader{Prediction rule ensembles in R}
\rightheader{Prediction rule ensembles in R}
\author{M. Fokkema$^1$}
\affiliation{$^1$Universiteit Leiden}

%\acknowledgements{}

\begin{document}

\maketitle
\pagewiselinenumbers

\section{Abstract}
Prediction rule ensembles are a highly interpretable method for regression and classification. Prediction rule ensembles consist of a small set of simple decision rules, which are easy to evaluate and apply in practice. The current paper discusses rule ensemble methodology, and introduces the unbiased rule prediction ensemble (upre) algorithm. The upre algorithm and its software implementation are largely based on the prediction rule ensemble methodology of \citeA{FrieyPope08}, with some added features and improvement. To illustrate the algorithm and the \verb|R| package \verb|upre|, upre will be applied to real datasets and the results will presented. The software is well documented and freely available. It is hoped that readers, after reproducing the examples, will wave their hands in the air like they just don't care and exclaim 'Upre? I think it's super!'.

\section{Introduction}

Prediction rule ensembles provide accurate and interpretable methods for regression and classification. Prediction rules are logical statements of the form if [conditions] then [prediction], which are highly usable in clinical decision making. The rules in prediction rule ensembles can be seen as very simple decision trees \cite{FokkySmit15a, KatsyPach08}. Decision trees can be derived by, for example, the classification and regression trees (CART; \citeNP{BreiyFrie84}) algorithm, and such trees are well known for their interpretability and applicability in practical in decision making. However, single trees are also known for their instability: small changes in the training data may cause large changes in the resulting tree \cite<e.g., >{Dann00}.

A powerful solution to this problem of instability is combining the predictions of many single trees, with techniques like bagging, boosting and random forests. Ensembling trees has been shown to substantially improve predictive accuracy \cite<e.g., >{Brei96, Diet00, StroyMall09}. However, the tree ensembles generally consist of a large number of trees, and are therefore difficult to interpret, let alone apply without the help of statistical software. A trade-off between accuracy and interpretability seems to apply: single trees provide high interpretability but lower accuracy, whereas ensembles provide high accuracy but lower interpretability.

Rule-based ensembles aim to balance accuracy and interpretability: accuracy is improved by ensembling predictions and interpretability is improved by using prediction rules instead of full trees. Several algorithms for deriving rule-based ensembles have been developed. Most are aimed exclusively at classification, like ROCCER, \cite{PratyFlac05}, SLIPPER \cite{CoheySing99}, Lightweight Rule Induction \cite{WeisyIndu00} and C4.5rules \cite{Quin14}. In contrast, the Rulefit \cite{FrieyPope08}, ENDER \cite{DembyKotl10} and Node Harvest \cite{Mein10} algorithms can be applied to both classification and regression problems. Among these three, Rulefit is most explicitly aimed at balancing accuracy and interpretability, as it employs regularized regression to select only a small set of prediction rules for the final ensemble. In contrast, ENDER and Node Harvest produce ensembles with a large number of rules, limiting interpretability of the ensemble.

The aim of the current paper is to introduce the unbiased prediction rule ensemble (upre) algorithm and its software implementation. The upre algorithm is largely based on Rulefit, as described by \cite{FrieyPope08}, and derives a prediction rule ensembles in two stages. In the first stage, a large initial ensemble of rules is generated by taking the nodes from trees that are grown on a large number of samples drawn from the training dataset. This will be further described in the subsection Rule generation. In the second stage, a small final ensemble of prediction rules is selected by means of penalized regression, which will be further described in the subsection Rule selection. In the subsection Comparison with Rulefit, the differences and similarities between upre and Rulefit will be discussed. In the subsection Illustration, the upre algorithm will be applied to several real datasets, to illustrate the algorithm and its \verb|R| implementation. In the Simulation section, the predictive accuracy of upre will be assessed and compared with that of Rulefit and Node Harvest, using simulated datasets. In the Discussion, results will be summarized, and further research and future developments will be discussed.

\subsection{Rule generation}

The upre algorithm derives prediction rules from an ensemble of trees, which are sequentially grown on random samples of the training data. The trees grown by upre are conditional inference trees or ctrees \cite{HothyHorn06, ZeilHoth08}. Although tree-based ensemble algorithms generally use CART for growing trees, CART uses biased variable selection criteria: given two equally informative input variables, the variable with a larger number of unique and/or missing values has a higher probability of being selected as a splitting variable \cite<e.g., >{KimyLoh01, StroyBoul08}. The ctree algorithm provides unbiased variable selection, by using statistical tests for parameter instability to select splitting variables. For further details on the parameter instability tests and split selection, readers are referred to \citeNP{HothyZeil06}.

An example ctree is presented in Figure \ref{fig:regression_tree}. The first node of a tree does not represent an informative rule, as it applies to all observations in the sample. Therefore, although the tree in Figure \ref{fig:regression_tree} consists of nine nodes, the following eight prediction rules are derived from it:

\begin{multline*}
	\\
	r_1(\mathbf{x}) = I (x_4 \leq 82)\\
	r_2(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{a,c\})\\
	r_3(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\})\\
	r_4(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\} \cdot x_4 \leq 77)\\
	r_5(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\} \cdot x_4 > 77)\\
	r_6(\mathbf{x}) = I (x_4 > 82)\\
	r_7(\mathbf{x}) = I (x_4 > 82 \cdot x_5 \leq \textrm{seldom})\\
	r_8(\mathbf{x}) = I (x_4 > 82 \cdot x_5 > \textrm{seldom})\\
\end{multline*}

Where $\mathbf{x}$ is random vector of $n$ input variables, $I$ is an indicator of the truth of its argument, and $r_m(\mathbf{x})$ denotes prediction rule $m$, taking a value of 0 when its conditions do not apply, and a value of 1 when its conditions apply. Note that the tree in Figure \ref{fig:regression_tree} shows that input variables used for growing trees may be continuous (like $x_4$), unordered categorical (like $x_3$) or ordered categorical (like $x_5$). Also, that the same input variable may appear multiple times in the same rule (like $x_4$ in $r_4(\mathbf{x}) or r_5(\mathbf{x})$).

\begin{figure}[t!]
\setkeys{Gin}{width=\textwidth}
<<example-tree, fig=TRUE, width= 5, height= 5>>=
ex_tree <- party(
# node 0:
partynode(0L,
split = partysplit(2L, breaks = 82),
kids = list(
# node 1:
partynode(1L,
split = partysplit(1L, breaks = 2),
kids = list(
# node 2:
partynode(2L, info = ""), 
# node 3:
partynode(3L, 
split = partysplit(2L, breaks = 77),
kids = list(
# node 4:
partynode(4L, info = ""),
# node 5:
partynode(5L, info = ""))))),
# node 6:
partynode(6L,
split = partysplit(3L, breaks = 1),
kids = list(
# node 7:
partynode(7L, info = " "),
# node 8:
partynode(8L, info = " "))))), 
# dataset:
data.frame(x3 = factor(character(0), levels = c("a", "c", "b", "d", "e"), ordered = FALSE), 
x4 = numeric(0), 
x5 = factor(character(0), levels = c("seldom", "somtimes", "regularly"), ordered = TRUE))
)
plot(ex_tree, inner_panel = node_inner(ex_tree, id = FALSE), terminal_panel = node_terminal(ex_tree, id = FALSE, fill = "white"))
@
\label{fig:regression_tree}
\caption{Example conditional inference tree. Note that a conditional inference tree would normally depict the distribution of the outcome variable $y$ in each of the terminal nodes. For generating the initial ensemble of prediction rules, only the tree structure, not the distribution of the outcome variable is used, which is therefore omitted here.}
\end{figure}

To derive the ensemble of trees, approaches similar to bagging \cite{Brei96bagging}, boosting \cite{FreuyShap95}, random forests \cite{Brei01RandFor}, or a combination can be used. In either appraoch, a user-specified number of random samples (500, by default) of size $\eta$ are drawn from the training data with or without replacement. Setting $\eta = 1$ will result in sampling with replacement, or bootstrap sampling. Setting $\eta < 1$ will result in sampling without replacement, or subsampling. %In some circumstances, subsampling may lead to better results: Strobl psymeth paper and some other reference. 
Trees are grown on the random samples in a sequential fashion, with learning rate $\nu = 0, \dots, 1$. The learning rate $\nu$ corresponds to the shrinkage parameter in boosting, and controls the influence of earlier trees in the induction of new trees. Let $m = 1, \dots, M$ denote the random sample of observations drawn from the training data, and $f_m(\mathbf{x})$ the predictive function from the tree grown on subsample $m$. Then every tree $m > 1$ is grown using outcome variable $u_m$, instead of the original outcome variable $y$:

\begin{equation}
	u_m  = y - \sum_{k=1}^{m-1}{\nu \cdot f_k(\mathbf{x})} 
	% some kind of notation for residualized y should be used
	% first, the mean of y should be substracted, which is $f_0(\mathbf{x})$
\end{equation}

Note that setting $\nu = 0$ removes the influence of earlier trees, with each tree being grown using the original values of the outcome variable $y$. Setting $\nu = 1$ maximizes the influence of earlier trees. \citeA{FrieyPope03} have found small, nonzero values of $\nu$ (e.g., .01) to perform best for ensembles of small decision trees. In addition, they found their results to be fairly insensitive to the value selected for $\eta$, provided it was small (i.e., $\eta < N/2$ for $N \leq 500$ and $\eta < 10 \cdot \sqrt{N}$ for $N > 500$). The prediction rules in a upre ensemble can be based on a bagged ensemble by setting $\nu = 0$ and $\eta = 1$. The rules can be based on a random forest by setting $\nu = 0$, $\eta = 1$ and the number of input variables to be selected as candidates for each split to values $< n$. Note that $\nu$, $\eta$ and the number of input variables to be selected as candidates for each split can be varied independently from each other, to create a hybrid of bagging, boosting, and random forests.

Note that the original bagging and random forest algorithms make use of unpruned CART trees, which can be mimicked in the upre software implementation by setting the maximum tree depth to infinity. However, in situations with a large number of observations and few variables, \citeA{LinyJeon06} have found predictive accuracy of random forests to improve when tree size is limited. In addition, smaller trees provide shorter prediction rules, which are easier to interpret. Therefore, by default upre restricts maximal tree depth to 3, thereby limiting the maximum number of conditions of rules to 3. In effect, this limits the interactions that can be captured by the rule ensemble to second-order and three-variable interactions. Users may provide any other integer value for maximum tree depth, as well.

After the ensemble of trees is generated, every node from every tree is included in the initial rule ensemble, after removing perfectly collinear rules (i.e., rules with conditions or support identical to that of earlier rules).

In addition, users may choose the original input variables to be added as linear terms to the initial ensemble. Although non-linear functions of input variables are relatively easy for rules to approximate, linear functions are not. Therefore, addition of the input variables may improve sparsity and predictive accuracy of the final ensemble. To reduce the effect of possible outliers, input variables may be winsorized before entering the predictive model. The linear terms are given by:

\begin{equation}
	l_j(x_j) = \min(\delta_j^+, max(\delta_j^-, x_j))
	\label{eq:linear_terms}	
\end{equation}

where $j$ is an indicator for the input variable, and $\delta_j$ and $\delta_j$ are the $\beta$ and $(1-\beta)$ quantiles of the distribution of $x_j$ in the training data. The value chosen for $\beta$ reflect user's prior suspicion of the fraction of outliers. By default $\beta$ is set to .025. Setting $\beta = 0$ will result in the input variables not being winsorized.

\subsection{Selection of the final ensemble}

The initial ensemble consists of a large number of rules and/or linear functions, only a few of which may contribute to predictive accuracy on future observations. Therefore, coefficients for the rules and/or linear terms for the final predictive model are derived using penalized regression. When both rules and linear terms are included, the predictive model becomes:

\begin{equation}
	F(\mathbf{x}) = \hat{a}_0 + \sum_{k=1}^K{\hat{a}_k r_k(\mathbf{x}) + \sum_{j=1}^n{\hat{b}_j l_j(x_j)}} 
	\label{eq:predictive_model}
\end{equation}

Where $1 \dots j \dots n$ is an index for the input variable and $1 \dots k \dots K$ is an index for the rule in the initial ensemble. The coefficient values are estimated by means of penalized regression. When the final model is selected by lasso regression \cite{Tibs96}, the estimates satisfy the following minimization criterion:

\begin{equation}
	( \{ \hat{a}_k \}^K_0 , \{ \hat{b}_j \}^n_1  ) = 
	\arg\min_{ \{ \hat{a}_k \}^K_0 , \{ \hat{b}_j \}^n_1 } \sum_{i=1}^N{L \left( 
		y_i, \hat{a}_0 + \sum_{k=1}^K{\hat{a}_k r_k(\mathbf{x}_i) + \sum_{j=1}^n{\hat{b}_j l_j(x_ij)}}  \right)} +
	\lambda \left( \sum_{k=1}^K{| \hat{a}_k |} + \sum_{j=1}^n{| \hat{b}_j |} \right) 
	\label{eq:lasso_regression}
\end{equation}

The first term is the prediction risk over the $N$ training observations, and the second term is a penalty for non-zero coefficients of rules and linear terms. The influence of the penalty term is determined by the value of $\lambda \geq 0$. The value of $\lambda$ is taken to be that which minimizes the prediction risk on a separate validation dataset, or by multi-fold cross validation on the training data. The penalty term in Equation \ref{eq:lasso_regression} forces many coefficients to be set to zero, providing a final ensemble consisting of a small set of rules and/or linear terms. Alternatively, ridge or elastic net penalties \cite{ZouyHast05} may be applied in Equation \ref{eq:lasso_regression}, instead of the lasso penalty, but this will generally yield less sparse models.

The penalty term in \ref{eq:lasso_regression} is sensitive to linear transformations of the input variables: generally, terms with smaller variance are more heavily penalized. Users can give a linear terms the same weight in the penalty term as a typical rule, by selecting its normalized values to be used in minimizing Equation \ref{eq:lasso_regression}:

\begin{equation} 
	l_j(x_j) \leftarrow 0.4*l_j(x_j) / \mathit{std}(l_j(x_j))
\end{equation}

Alternatively, all rules and linear functions may be standardized prior to estimating Equation \ref{eq:lasso_regression}, giving each rule and linear term equal a-priori influence. However, rules with support close to zero or one are defined by a correspondingly small number of training observations, and reducing penalization of these terms may be unwanted.

% In the 2010 version of RuleFit, Huber loss was an option for estimating the final model. In the 2012 version, Huber loss is no longer used (it seems).

\subsection{Interpreting the final ensemble: variable importances and interactions}

\subsubsection{Importances}

The estimated coefficients for rules and/or linear terms in the final ensemble provide an unstandardized global measure of their contribution to the predictive model. Importances are provide standardized measures of their contribution to the predictive model, and can be calculated for rules, linear terms, as well as input variables (as input variables may appear in the final model as part of one or more rules, as well as a linear term). Variable importances provide a measure of the extent to which individual input variables influence the predictions of the final rule ensemble. Importances can be calculated globally (averaged over the whole training dataset) or locally (over a selected subregion of the input variable space). The global importance of a rule $r_k$ is given by

\begin{equation}
	I_k = |\hat{a}_k| \cdot \sqrt{s_k(1-s_k)} % equation 28 in F\&P
	\label{eq:global_rule_imp}
\end{equation}

where $s_k$ is the support of rule $r_k$ on the training data, given by

\begin{equation}
	s_k = \frac{1}{N} \sum^{N}_{i=1}{r_k(x_i)}
	\label{eq:global_support}
\end{equation}

The global importance of linear term $l_j(x_j)$ is given by

\begin{equation}
	I_j = |\hat{b}_j| \cdot \mathit{std}(l_j(x_j)) % equation 29 in F\&P
	\label{eq:global_linear_imp}
\end{equation}

where $std(l_j(x_j))$ is the standard deviation of the linear term $l_j(x_j)$ (Equation \ref{eq:linear_terms}) in the training data. The importance of an input variable $J_j$ is given by the sum of the importance of its linear term $I_j$, and a weighted sum of the importances of the rules $I_k$ in which the variable appears:

\begin{equation}
	J_j = I_j + \sum_{x_j \in r_k}{I_k / m_k}
	\label{eq:glob_inputvar_imp}
\end{equation}

where $m_k$ is the number of variables that define rule $k$. The second term in Equation \ref{eq:glob_inputvar_imp} shows that the importance of a rule is distributed equally over the input variables appearing in the rule\footnote{Note that when $x_j$ appears multiple times in the conditions of $r_k$, Equation \ref{eq:glob_inputvar_imp} is not completely accurate. In those instances, the importance of $r_k$ that is distributed to the importance $J_j$ equals $\frac{\textrm{the number of times } x_j \textrm{appears in the conditions of} r_k \times I_k}{\textrm{the total number of conditions in} r_k}$}.

Local importances for a selected subregion $S$ of the input variable space can be calculated by replacing the global support $\s_k$ in Equation \ref{eq:global_rule_imp} by the support in the selected subregion, which is given by

\begin{equation}
	s_k(S) = \frac{1}{|S|} \sum_{x_i \in S}{r_k(x_i)}
	\label{eq:local_support}
\end{equation}

where $|S|$ is the cardinality (number of training set observations) in subregion $S$. Similarly, the local importance of a linear term over a selected subregion can be calculated by replacing the global standard deviation $\mathit{std}(l_j(x_j))$ in Equation \ref{eq:global_linear_imp} by its standard deviation in the selected subregion:

\begin{equation}
	\mathit{std}(l_j(x_j), S) = \sqrt{ \frac{1}{|S| - 1} \sum_{x_i \in S}{(l_j(x_ij) - \bar{l}_j(x_j, S))^2} }
	\label{eq:local_sd}
\end{equation}

where $\bar{l}_j(x_j, S)$ is the mean of $l(x_j)$ in subregion $S$.

Local importances can also be calculated for a single point $\mathbf{x}$. The local importance of rule $r_k$ at a single point $\mathbf{x}$ is given by

\begin{equation}
	I_k(\mathbf{x}) = |\hat{a_k}| \cdot |r_k(\mathbf{x}) - s_k |)% equation 30 in F\&P
	\label{eq:local_rule_imp}
\end{equation}

The local importance of linear term $l_j(x_j)$ at a single value $x_j$ is given by

\begin{equation}
	I_j(x_j) = |\hat{b_j}| \cdot |l_j(x_j) - \bar{l}_j |)% equation 31 in F\&P
	\label{local_linear_imp}
\end{equation}

where $\bar{l}_j$ is the mean of $l_j(x_j)$ over the training data. The quantities in Equations \ref{eq:global_rule_imp}, \ref{eq:global_linear_imp}, \ref{eq:local_rule_imp} and \ref{eq:local_linear_imp} represent the absolute change in the (average) prediction $F(\mathbf{x})$ when the corresponding rule or linear term is dropped from the predictive model (Equation \ref{eq:predictive_model}), when the intercept would be adjusted accordingly (i.e., $\hat{a}_0 \leftarrow \hat{a}_0 + \hat{a}_k s_k$ for rules and $\hat{a}_0 \leftarrow \hat{a}_0 + \hat{b}_j \bar{l}_j$ for linear terms). The local importance of input variable $x_j$ at each point $\mathbf{x}$ is given by

\begin{equation}
	J_j(\mathbf{x}) = I_j(\mathbf{x}) + \sum{x_j \in r_k}{I_k(\mathbf{x}) / m_k} 
\end{equation}

\subsection{Partial dependence plots}

The influence of individual or pairs of predictor variables on the predictions of the final rule ensemble can be also assessed visually by plotting the partial dependence of $F(\mathbf{x})$ on individual or pairs of input variables. The partial dependence of $F(\mathbf{x})$ on a subset of input variable $s \subset \{1, ..., n\}$ is the expected value of $F(\mathbf{x})$ over the marginal joint distribution of variables $\mathbf{x}_{\backslash s}$ not in $\mathbf{x}_s$. It is defined as

\begin{equation}
	F_s(\mathbf{x}_s) = E_{\mathbf{x}_{\backslash s}} [F(\mathbf{x}_s, \mathbf{x}_{\backslash s})] % equation 39 in F/&P
\end{equation}

and can be estimated from the data by

\begin{equation}
	\hat{F}_s(\mathbf{x}_s) = \frac{1}{N} \sum^N_{i=1}{F(\mathbf{x}_s, \mathbf{x}_{i \backslash s})} % equation 40 in F/&P
\end{equation}

where $\{ \mathbf{x}_{i \backslash s} \}^N_1$ are the observations in the training dataset. The subset of input variables $s$ must consist of single or a pair of input variables to allow for plotting the partial dependence of $F(\mathbf{x})$ on $\mathbf{x}_s$.

\subsubsection{Interaction effects}

Prediction rules are well suited to capturing interaction effects of input variables. However, the presence of prediction rules involving multiple variables in the final ensemble are a necessary but not sufficient condition for the presence of interaction effects. Different rules involving the same sets of variables may eliminate the interaction effect of these variable in the final predictive model $F(\mathbf{x})$. Therefore, the presence of interaction effects should be assessed by analyzing the full predictive model $F(\mathbf{x})$, for example by means of partial dependence plots and computation of interaction statistics, which will be discusses in this section.

If an input variable $x_j$ does not interact with any of the other input variables, then the effect of $x_j$ and $x_{\backslash j}$ on $F(\mathbf{x})$ are additive, and therefore can be expressed as

\begin{equation}
	F(\mathbf{x}) = F_j(x_j) + F(\mathbf{x}_{\backslash j}) % equation 42
\end{equation}

where $F_j(x_j)$ is the partial dependence of $F(\mathbf{x})$ on $x_j$ and $F(\mathbf{x}_{\backslash j})$ is the partial dependence of $F(\mathbf{x})$ on $\mathbf{x}_{\backslash j}$. A statistics for testing whether $x_j$ interacts with any other input variable is given by

\begin{equation}
	H^2_j = \frac{\sum^N_{i=1}{ [F(\mathbf{x}_i) - \hat{F}_j(x_{ij}) - \hat{F}_{\backslash j}(x_{i \backslash j})]^2 }}{{\sum^N_{i=1}{ F^2(\mathbf{x}_i)}
	\label{eq:interaction_statistic}
\end{equation}

$H^2_j$ will differ from zero to the extent that $x_j$ is involved in interactions with other input variables. To assess whether $H^2_j$ differs significantly from zero, a null distribution for the statistic has to be derived. \cite{FrieyPope08} suggest the use of a variant of the parametric bootstrap \cite{EfroyTibs94} to derive a null distribution of statistic $ H_j^2$. The statistic $H^2_j$ is repeatedly computed on a series of artificial datasets from which interactions involving $x_j$ are absent. The distribution of test statistics thus obtained can be used as a reference distribution, to assess significance of the value of  $H^2_j$ based on the training data.

%To disencourage spurious interactions entering the model, F\&P (section 8.2) use a strategy to put an incentive on variables entering a tree, that do not appear earlier in the tree. I currently do not know of a way to do that for ctrees.

\subsection{Comparison with Rulefit}

- ctree instead of CART trees
- Rulefit uses an approach that can be best described as a combination of bagging and boosting: it takes subsamples\footnote{With subsampling, observations are drawn without replacement from the full dataset, so strictly speaking, Rulefit does not use a bagging approach, as this would involve bootstrap samples of the dataset.} of size $\eta$ of the training data, and grows trees in a sequential fashion with learning rate $\nu = 0, \dots, 1$.
- Rulefit allows for inducing trees in a manner that can be best described as a hybrid of bagging \cite{Brei96bagging} and boosting \cite{Frie01}. In some situations, bagged ensembles of trees may be outperformed by random forests in terms of predictive accuracy \cite{Diet00bbrf}. Also, Node Harvest derives prediction rules from the nodes of random forest trees. Therefore, upre allows for inducing trees in a bagging, boosting or random forest type manner, or a hybrid of the three.    
- Rulefit also provides forward stagewise regression
- Current implementation of rulefit does not provide user-friendly interface for rules inspection, does not provide value of lambda or intercept, and does not allow user to select a different values for lambda (e.g., lamba min or lambda 1se).
- In the pairplot function, rulefit does not seem to cross all possible values of x1 and x2, it may just take all observed combinations. Then, correlated x1 and x2 may result in a plot that looks like an interaction, but it is not. Check if my plotting function gives a different results, although it may be slower (but then, it is immediately clear why))

\subsection{Miscellaneous}

RuleFit returns the cross-validated model selection criterion value with standard error (when test.reps > 1), and number of terms in the resulting model, in the command line. 
The penalized regression is performed on the whole set of training observations. However, the value of the penalty parameter(s) is chosen to be the value that minimizes future prediction risk, estimated in a separate sample not used in training, or by full multi-fold cross validation.

"3.2. Missing values. An interesting property of NH is its natural ability to
cope with missing values. Once a fit is obtained, predictions for new data can be
obtained without use of imputation techniques or surrogate splits. To fit the node
harvest estimator with missing data, we replace missing values in the matrix X
by the imputation technique described in ...."

Users are adviced to not use the lambda path determined by \verb|glmnet|, include a way to determine own path?

Future implementations may support multivariate outcomes. ctree() allows for multivariate responses. If glmnet also allows for that, multivariate rule ensembles can be created.

\subsection{Extensions to random-effect models}

Schelldorder, Meier \& Buhlmann (2014) propose a two-step approach for high-dimensional GLMMs: In the first step, variable screening is performed, selecting a set of candidate active variables. In the second step, refitting by maximum likelihood (ML) estimation is performed to get accurate parameter estimates.
Groll \& Tutz (2010) propose a method that works by combining gradient ascent optimization with the Fisher scoring algorithm and is based on the approach of Goeman (2010).

\subsection{Alternative rule ensemble methods}

Meinshausen's Node harvest algorithm derives an initial ensemble Q which is very large (thousands of rules). The rules in the initial ensemble can be generated at random from the training data, without using a response variable. However, Meinshausen (2010) suggests to derive the rules from the trees in a random forest (Breiman, 2001). However, the trees are fitted on subsamples of the data (of size n/10), rather than on bootstrap samples. Next, the nodes that satisfy maximal interaction order and minimal node size constraints are added to the initial ensemble, provided that they are not already present in the set. If two or more nodes in the initial ensemble contain the exact same set of training observations, only one is randomly selected and included in the ensemble.

Node Harvest uses only the nodes (rules) in which a new observation `falls' for prediction, requiring only a small number of rules for a single observation. However, future observations may `fall' into any of the nodes of the initial ensemble, so all rules from the initial (and large) ensemble have to be retained to allow for prediction of future observations.

ROCCER \cite{PratyFlac05}, SLIPPER \cite{CoheySing99}, LRI \cite{WeisyIndu00}, and C4.5rules \cite{Quin14} are for classification problems only. \cite{DembyKotl10} can be used for both classification and regression. None of these methods are implemented in statistical software like R of SPSS.

Lightweight Rule Induction \cite{WeisyIndu00}: "A light weight rule induction method is described that generates compact Disjunctive Normal Form (DNF) rules. Each class has an equal number of unweighted rules. A new example is classified by applying all rules and assigning the example to the class with the most satisfied rules. The induction method attempts to minimize the training error with no pruning. An overall design is specified by setting limits on the size and number of rules. During training, cases are adaptively weighted using a simple cumulative error method. The induction method is nearly linear in time relative to an increase in the number of induced rules or the number of cases. Experimental results on large bench-mark datasets demonstrate that predictive performance can rival the best reported results in the literature."

SLIPPER \cite{CoheySing99}: "We describe SLIPPER - a new rule learner that generates rulesets by repeatedly boosting a simple, greedy rule builder. Like the rulesets built by other rule learners, the ensemble of rules created by SLIPPER is compact and comprehensible. This is made possible by imposing appropriate constraints on the rule-builder, and by use of a recently proposed generalization of Ad-aboost called confidence-rated boosting. In spite of its relative simplicity, SLIPPER is highly scalable, and an effective learner. Experimentally, SLIPPER scales no worse than O(n log n), where n is the number of examples, and on a set of 32 benchmark problems, SLIPPER achieves lower error rates than RIPPER 20 times, and lower error rates than C4.5rules 22 times."

\cite{Quin14} C4.5rules is for classification only

SLIPPER and ENDER take a boosting approach to rule induction.

ENDER \cite{DembyKotl10}: "We believe that ENDER can still be used for interpretation purposes. One way is to follow the approach given in Friedman and Popescu (2008) that relies on a postprocessing phase in which the rules are refitted by using the lasso regularization."

\bibliography{bib}

\end{document}

### 2016-08-30 pkg/glmertree/upre/paper/prediction_rule_ensembles.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@953 255f4249-2263-4417-b1ea-8498779e27a7

--- 2016-08-30 pkg/glmertree/upre/paper/prediction_rule_ensembles.Rnw
@@
 %% for \usepackage{Sweave}
-\SweaveOpts{echo=FALSE, results=hide}
+
 \setkeys{Gin}{width=0.8\textwidth}
-
-<<preliminaries>>=
-library("partykit")
-@
 
@@
 \begin{multline*}
-	\\
 	r_1(\mathbf{x}) = I (x_4 \leq 82)\\
@@
 \setkeys{Gin}{width=\textwidth}
-<<example-tree, fig=TRUE, width= 5, height= 5>>=
-ex_tree <- party(
-# node 0:
-partynode(0L,
-split = partysplit(2L, breaks = 82),
-kids = list(
-# node 1:
-partynode(1L,
-split = partysplit(1L, breaks = 2),
-kids = list(
-# node 2:
-partynode(2L, info = ""), 
-# node 3:
-partynode(3L, 
-split = partysplit(2L, breaks = 77),
-kids = list(
-# node 4:
-partynode(4L, info = ""),
-# node 5:
-partynode(5L, info = ""))))),
-# node 6:
-partynode(6L,
-split = partysplit(3L, breaks = 1),
-kids = list(
-# node 7:
-partynode(7L, info = " "),
-# node 8:
-partynode(8L, info = " "))))), 
-# dataset:
-data.frame(x3 = factor(character(0), levels = c("a", "c", "b", "d", "e"), ordered = FALSE), 
-x4 = numeric(0), 
-x5 = factor(character(0), levels = c("seldom", "somtimes", "regularly"), ordered = TRUE))
-)
-plot(ex_tree, inner_panel = node_inner(ex_tree, id = FALSE), terminal_panel = node_terminal(ex_tree, id = FALSE, fill = "white"))
-@
+\includegraphics{prediction_rule_ensembles-example-tree}
 \label{fig:regression_tree}
@@
 
+% !!! rules that are each others negative should also be removed!!!!
+
 In addition, users may choose the original input variables to be added as linear terms to the initial ensemble.
@@
 The coefficient values are estimated by means of penalized regression.
-When the final model is selected by lasso regression \cite{Tibs96}, the estimates satisfy the following minimization criterion:
+When the final model is selected by lasso regression \cite{Tibs96}, which is the default in the upre algorithm, the estimates satisfy the following minimization criterion:
 
@@
 The value of $\lambda$ is taken to be that which minimizes the prediction risk on a separate validation dataset, or by multi-fold cross validation on the training data.
-The penalty term in Equation \ref{eq:lasso_regression} forces many coefficients to be set to zero, providing a final ensemble consisting of a small set of rules and/or linear terms.
-Alternatively, ridge or elastic net penalties \cite{ZouyHast05} may be applied in Equation \ref{eq:lasso_regression}, instead of the lasso penalty, but this will generally yield less sparse models.
+The penalty term in \ref{eq:lasso_regression} forces many coefficients to be set to zero, providing a final ensemble consisting of a small set of rules and/or linear terms.
+Users can also select ridge or elastic net regression \cite{ZouyHast05} to derive the final ensemble.
+The penalty term in Equation \ref{eq:lasso_regression} would then replaced by a ridge or elastic net counterpart, respectively penalizing squared values of $\hat{a}_k$ and $\hat{b}_j$ instead, or as well as, absolute values.
 
-The penalty term in \ref{eq:lasso_regression} is sensitive to linear transformations of the input variables: generally, terms with smaller variance are more heavily penalized.
-Users can give a linear terms the same weight in the penalty term as a typical rule, by selecting its normalized values to be used in minimizing Equation \ref{eq:lasso_regression}:
+The penalty term in \ref{eq:lasso_regression} is affected by linear transformations of the input variables: generally, terms with smaller variance are more heavily penalized.
+Users can give linear terms the same influence on the penalty term as a typical rule, by selecting their normalized versions to be used in the estimation (Equation \ref{eq:lasso_regression}):
 
 \begin{equation} 
-	l_j(x_j) \leftarrow 0.4*l_j(x_j) / \mathit{std}(l_j(x_j))
+	l_j(x_j) <- 0.4*l_j(x_j) / \textit{std}(l_j(x_j))
 \end{equation}
 
-Alternatively, all rules and linear functions may be standardized prior to estimating Equation \ref{eq:lasso_regression}, giving each rule and linear term equal a-priori influence.
-However, rules with support close to zero or one are defined by a correspondingly small number of training observations, and reducing penalization of these terms may be unwanted.
+% Rules are not normalized.
+Rules with support close to zero or one are defined by a correspondingly small number of training observations, and therefore it is OK that lasso regression puts increased penalty on those.
 
 % In the 2010 version of RuleFit, Huber loss was an option for estimating the final model. In the 2012 version, Huber loss is no longer used (it seems).
+
+% if linear terms are normalized, the intercept should be corrected in the coef and/or predict function, as well.
 
@@
 \begin{equation}
-	I_j = |\hat{b}_j| \cdot \mathit{std}(l_j(x_j)) % equation 29 in F\&P
+	I_j = |\hat{b}_j| \cdot \textit{std}(l_j(x_j)) % equation 29 in F\&P
 	\label{eq:global_linear_imp}
@@
 where $|S|$ is the cardinality (number of training set observations) in subregion $S$.
-Similarly, the local importance of a linear term over a selected subregion can be calculated by replacing the global standard deviation $\mathit{std}(l_j(x_j))$ in Equation \ref{eq:global_linear_imp} by its standard deviation in the selected subregion:
+Similarly, the local importance of a linear term over a selected subregion can be calculated by replacing the global standard deviation $\textit{std}(l_j(x_j))$ in Equation \ref{eq:global_linear_imp} by its standard deviation in the selected subregion:
 
 \begin{equation}
-	\mathit{std}(l_j(x_j), S) = \sqrt{ \frac{1}{|S| - 1} \sum_{x_i \in S}{(l_j(x_ij) - \bar{l}_j(x_j, S))^2} }
+	\textit{std}(l_j(x_j), S) = \sqrt{ \frac{1}{|S| - 1} \sum_{x_i \in S}{(l_j(x_ij) - \bar{l}_j(x_j, S))^2} }
 	\label{eq:local_sd}
@@
 where $\bar{l}_j$ is the mean of $l_j(x_j)$ over the training data.
-The quantities in Equations \ref{eq:global_rule_imp}, \ref{eq:global_linear_imp}, \ref{eq:local_rule_imp} and \ref{eq:local_linear_imp} represent the absolute change in the (average) prediction $F(\mathbf{x})$ when the corresponding rule or linear term is dropped from the predictive model (Equation \ref{eq:predictive_model}), when the intercept would be adjusted accordingly (i.e., $\hat{a}_0 \leftarrow \hat{a}_0 + \hat{a}_k s_k$ for rules and $\hat{a}_0 \leftarrow \hat{a}_0 + \hat{b}_j \bar{l}_j$ for linear terms).
+The quantities in Equations \ref{eq:global_rule_imp}, \ref{eq:global_linear_imp}, \ref{eq:local_rule_imp} and \ref{eq:local_linear_imp} represent the absolute change in the (average) prediction $F(\mathbf{x})$ when the corresponding rule or linear term is dropped from the predictive model (Equation \ref{eq:predictive_model}), when the intercept would be adjusted accordingly (i.e., $\hat{a}_0 <- \hat{a}_0 + \hat{a}_k s_k$ for rules and $\hat{a}_0 <- \hat{a}_0 + \hat{b}_j \bar{l}_j$ for linear terms).
 The local importance of input variable $x_j$ at each point $\mathbf{x}$ is given by
@@
 
-The influence of individual or pairs of predictor variables on the predictions of the final rule ensemble can be also assessed visually by plotting the partial dependence of $F(\mathbf{x})$ on individual or pairs of input variables.
-The partial dependence of $F(\mathbf{x})$ on a subset of input variable $s \subset \{1, ..., n\}$ is the expected value of $F(\mathbf{x})$ over the marginal joint distribution of variables $\mathbf{x}_{\backslash s}$ not in $\mathbf{x}_s$.
-It is defined as
-
-\begin{equation}
-	F_s(\mathbf{x}_s) = E_{\mathbf{x}_{\backslash s}} [F(\mathbf{x}_s, \mathbf{x}_{\backslash s})] % equation 39 in F/&P
-\end{equation}
-
-and can be estimated from the data by
-
-\begin{equation}
-	\hat{F}_s(\mathbf{x}_s) = \frac{1}{N} \sum^N_{i=1}{F(\mathbf{x}_s, \mathbf{x}_{i \backslash s})} % equation 40 in F/&P
-\end{equation}
-
-where $\{ \mathbf{x}_{i \backslash s} \}^N_1$ are the observations in the training dataset.
-The subset of input variables $s$ must consist of single or a pair of input variables to allow for plotting the partial dependence of $F(\mathbf{x})$ on $\mathbf{x}_s$.
+The influence of individual predictor variables on the predictions of the final rule ensemble can also be represented visually in partial dependence plots
 
@@
 
-Prediction rules are well suited to capturing interaction effects of input variables.
-However, the presence of prediction rules involving multiple variables in the final ensemble are a necessary but not sufficient condition for the presence of interaction effects.
-Different rules involving the same sets of variables may eliminate the interaction effect of these variable in the final predictive model $F(\mathbf{x})$.
-Therefore, the presence of interaction effects should be assessed by analyzing the full predictive model $F(\mathbf{x})$, for example by means of partial dependence plots and computation of interaction statistics, which will be discusses in this section.
+One of the main advantages of tree-based methods is the automatic detection of non-linear and/or interaction effects.
+However, the presence of interactions or non-linear effects of single variables cannot simply be inferred from inspection of the rules in the final ensemble.
+On the one hand, variables involved in interaction effects do not necessarily appear in the same rules, and on the other hand, variables appearing in the same rules do not necessarily produce an interaction effect on the outcome variable.
+To assess non-linear effects and multi-way interactions of predictor variables, partial dependence plots and statistical tests are provided.
 
-If an input variable $x_j$ does not interact with any of the other input variables, then the effect of $x_j$ and $x_{\backslash j}$ on $F(\mathbf{x})$ are additive, and therefore can be expressed as
+In addition, the presence of interactions of (user-specified subsets of) predictor variables can be statistically tested.
+For these tests, a null model involving main effects of predictor variables only, is compared with a model involving interactions.
 
-\begin{equation}
-	F(\mathbf{x}) = F_j(x_j) + F(\mathbf{x}_{\backslash j}) % equation 42
-\end{equation}
-
-where $F_j(x_j)$ is the partial dependence of $F(\mathbf{x})$ on $x_j$ and $F(\mathbf{x}_{\backslash j})$ is the partial dependence of $F(\mathbf{x})$ on $\mathbf{x}_{\backslash j}$.
-A statistics for testing whether $x_j$ interacts with any other input variable is given by
-
-\begin{equation}
-	H^2_j = \frac{\sum^N_{i=1}{ [F(\mathbf{x}_i) - \hat{F}_j(x_{ij}) - \hat{F}_{\backslash j}(x_{i \backslash j})]^2 }}{{\sum^N_{i=1}{ F^2(\mathbf{x}_i)}
-	\label{eq:interaction_statistic}
-\end{equation}
-
-$H^2_j$ will differ from zero to the extent that $x_j$ is involved in interactions with other input variables.
-To assess whether $H^2_j$ differs significantly from zero, a null distribution for the statistic has to be derived.
-\cite{FrieyPope08} suggest the use of a variant of the parametric bootstrap \cite{EfroyTibs94} to derive a null distribution of statistic $ H_j^2$.
-The statistic $H^2_j$ is repeatedly computed on a series of artificial datasets from which interactions involving $x_j$ are absent.
-The distribution of test statistics thus obtained can be used as a reference distribution, to assess significance of the value of  $H^2_j$ based on the training data.
-
-%To disencourage spurious interactions entering the model, F\&P (section 8.2) use a strategy to put an incentive on variables entering a tree, that do not appear earlier in the tree.
-I currently do not know of a way to do that for ctrees.
+To disencourage spurious interactions entering the model, F\&P (section 8.2) use a strategy to put an incentive on variables entering a tree, that do not appear earlier in the tree.
+Meinshausen uses a random forest methodology for deriving the trees for the initial ensemble, which may have the same results: by selecting only a subset of the predictor variables for growing each tree, interaction effects will be disencouraged.
 
@@
 - Current implementation of rulefit does not provide user-friendly interface for rules inspection, does not provide value of lambda or intercept, and does not allow user to select a different values for lambda (e.g., lamba min or lambda 1se).
-- In the pairplot function, rulefit does not seem to cross all possible values of x1 and x2, it may just take all observed combinations.
-Then, correlated x1 and x2 may result in a plot that looks like an interaction, but it is not.
-Check if my plotting function gives a different results, although it may be slower (but then, it is immediately clear why))
 
@@
 
-Users are adviced to not use the lambda path determined by \verb|glmnet|, include a way to determine own path?
-
-Future implementations may support multivariate outcomes. ctree() allows for multivariate responses. If glmnet also allows for that, multivariate rule ensembles can be created.
+Users are advices to not use the lambda path determined by \verb|glmnet|, include a way to determine own path?
 
 \subsection{Extensions to random-effect models}
-
 Schelldorder, Meier \& Buhlmann (2014) propose a two-step approach for high-dimensional GLMMs: In the first step, variable screening is performed, selecting a set of candidate active variables.

### 2016-08-30 pkg/glmertree/building and checking.R

@@ -1,0 +1,47 @@
+setwd("C:/Users/tobii/Desktop/swReg")
+library(devtools)
+library(roxygen2)
+
+
+# process documentation:
+setwd("./swReg")
+document()
+
+setwd("..")
+check("swReg")
+uninstall("swReg")
+build("swReg")
+install("swReg")
+
+
+
+
+
+## create package:
+#create("upre")
+
+# process documentation:
+setwd("./upre")
+document()
+
+showPDFmanual <- function(package, lib.loc=NULL)
+{
+  path <- find.package(package, lib.loc)
+  system(paste(shQuote(file.path(R.home("bin"), "R")),
+               "CMD", "Rd2pdf",
+               shQuote(path)))
+} 
+showPDFmanual("upre")
+
+# check and build
+setwd("..")
+check("upre")
+uninstall("upre")
+build("upre")
+install("upre")
+
+
+
+# It is typically a design mistake to use ::: in your code since the corresponding object has probably 
+# been kept internal for a good reason. Consider contacting the package maintainer if you feel the need 
+# to access the object for anything but mere inspection. 

### 2016-08-30 pkg/glmertree/old initial and final rule ensemble functions (12-07-2016).R

@@ -1,0 +1,221 @@
+
+
+
+
+
+#' Derive initial rule ensemble
+#'
+#' \code{initial} derives a large initial ensemble of prediction rules.
+#' 
+#' @param formula regression formula; a symbolic description of the model to be fit.
+#' @param data a data frame containing the variables in the model.
+#' @param type character. type of base learners to be included in ensemble. Defaults to "both" (intial 
+#' ensemble included both rules and linear functions). Other option may be "rules" (for prediction rules 
+#' only) or "linear" (for linear functions only).
+#' @param weights an optional vercotr of weights to be used in deriving the trees.
+#' @param sampfrac numeric. fraction of randomly selected training observations to produce each tree. 
+#' Defaults to .5.
+#' @param ntrees numeric. total number of trees to be generated. Defaults to 100.
+#' @param seed numeric. random seed to be used in deriving the final ensemble (for reproducability). 
+#' Defaults to 42.
+#' @param maxdepth numeric. maximal depth of trees to be grown. Defaults to 3, resulting in trees with max 
+#' 15 nodes (8 terminal and 7 inner nodes), and therefore 15 max rules.
+#' @param memory.par numeric. learning rate for sequentially induced trees. Defaults to .01.
+#' @param remove.duplicates logical. Remove rules from the ensemble which have the exact same support in 
+#' training data?
+#' @param max.rules numeric. Approximate maximum number of rules to be generated. The number of rules in 
+#' the final ensemble will be smaller, due to the omission of rules with identical conditions or support.
+#' @return a list with \code{data} (the original dataset), \code{rulevars} dataset with 0-1 coded rule 
+#' memberships for each rule and observation, \code{rules} dataframe with rule descriptions
+#' @examples airq.init.ens <- initial(Ozone ~ ., data=airquality[complete.cases(airquality),])
+#' @export
+#' @import partykit Formula
+initial <- function(formula, data, type = "both", weights = rep(1, times = nrow(data)), sampfrac = .5, 
+                    ntrees = 100, seed = 42, maxdepth = 3, memory.par = 0.01, remove.duplicates = TRUE,
+                    max.rules = 2500)
+{
+  ###################
+  ## Preliminaries ##
+  ###################
+  
[…]

### 2016-08-30 pkg/glmertree/ruleensfunctions (06-07-2016).R

@@ -1,0 +1,205 @@
+#' Derive initial rule ensemble
+#'
+#' \code{initial} derives a large initial ensemble of prediction rules.
+#' 
+#' @param formula regression formula; a symbolic description of the model to be fit.
+#' @param data a data frame containing the variables in the model.
+#' @param type character. type of base learners to be included in ensemble. Defaults to "both" (intial ensemble included both rules and linear functions). Other option may be "rules" (for prediction rules only) or "linear" (for linear functions only).
+#' @param weights an optional vercotr of weights to be used in deriving the trees.
+#' @param sampfrac numeric. fraction of randomly selected training observations to produce each tree. Defaults to .5.
+#' @param ntrees numeric. total number of trees to be generated. Defaults to 100.
+#' @param seed numeric. random seed to be used in deriving the final ensemble (for reproducability). Defaults to 42.
+#' @param maxdepth numeric. maximal depth of trees to be grown. Defaults to 3, resulting in trees with max 8 terminal nodes.
+#' @param memory.par numeric. learning rate for sequentially induced trees. Defaults to .01.
+#' @return a list with \code{data} (the original dataset), \code{rulevars} dataset with 0-1 coded rule memberships for each rule and observation, \code{rules} dataframe with rule descriptions
+#' @examples airq.init.ens <- initial(Ozone ~ ., data=airquality[complete.cases(airquality),])
+initial <- function(formula, data, type = "both", weights = rep(1, times = nrow(data)), sampfrac = .5, 
+                    ntrees = 100, seed = 42, maxdepth = 3, memory.par = 0.01)
+{
+  ###################
+  ## Preliminaries ##
+  ###################
+  
+  if(missing(formula) || missing(data)) {
+    stop("Arguments formula and dataset are required")
+  }
+  if(!is.formula(formula) & !Formula::is.Formula(formula)) {
+    stop("Argument formula should be of class formula")
+  }
+  if(length(sampfrac) != 1 || sampfrac < 0.01 || sampfrac > 0.99) {
+    stop("Bad value for 'sampfrac'")
+  }
+  if(length(type) != 1 || (type != "rules" & type != "both" & type != "linear")) {
[…]

### 2016-08-30 pkg/glmertree/upre/R/upre.R

@@ -1,0 +1,1007 @@
+#' Derive an unbiased prediction rule ensemble
+#'
+#' \code{upre} derives a sparse ensemble of rules and/or linear functions for 
+#' prediction
+#' 
+#' @param formula a symbolic description of the model to be fit of the form 
+#' \code{y ~ x1 + x2 + ...+ xn}. Node that input variables should not have 'rule' 
+#' in their name, and the formula may NOT exclude the intercept (e.g., by using 
+#' \code{+ 0} or \code{- 1}).
+#' @param data matrix or data.frame containing the variables in the model. When a
+#' matrix is specified, it must be of class \code{"numeric"} (the input and output 
+#' variable must be continuous; the input variables may be 0-1 coded variables). 
+#' When a data.frame is specified, the output variable must be of 
+#' class \code{"numeric"} and must be a continuous variable; the input variables 
+#' must be of class \code{"numeric"} (for continuous input variables), 
+#' \code{"logical"} (for binary variables), \code{"factor"} (for nominal input 
+#' variables with 2 or more levels), or \code{"ordered" "factor"} (for 
+#' ordered input variables).
+#' @param type character. Type of base learners to be included in ensemble. 
+#' Defaults to "both" (intial ensemble included both rules and linear functions). 
+#' Other option may be "rules" (for prediction rules only) or "linear" (for 
+#' linear functions only).
+#' @param weights an optional vector of observation weights to be used for 
+#' deriving the ensemble.
+#' @param sampfrac numeric value greater than 0, and smaller than or equal to 1. 
+#' Fraction of randomly selected training observations used to produce each tree. 
+#' Setting this to values < 1 will result in subsamples being drawn without 
+#' replacement (i.e., subsampling). Setting this equal to 1 will result in 
+#' bootstrap sampling.
+#' @param seed numeric. Random seed to be used in deriving the final ensemble 
+#' (for reproducability).
+#' @param maxdepth numeric. Maximal depth of trees to be grown. Defaults to 3,
+#' resulting in trees with max 15 nodes (8 terminal and 7 inner nodes), and 
+#' therefore max 15 rules.
+#' @param learnrate numeric. Learning rate for sequentially induced trees.
+#' @param removeduplicates logical. Remove rules from the ensemble which have 
+#' the exact same support in training data?
+#' @param maxrules numeric. Approximate maximum number of rules to be generated. 
[…]

### 2016-08-30 pkg/glmertree/upre/paper/Sweave code.R

@@ -1,0 +1,3 @@
+setwd("C:/Users/tobii/Desktop/swReg/upre/paper")
+Sweave("prediction_rule_ensembles.Rnw")
+tools::texi2dvi("prediction_rule_ensembles.tex", pdf = TRUE, clean = TRUE)

### 2016-08-30 pkg/glmertree/upre/paper/simulated data.R

@@ -1,0 +1,34 @@
+###################
+## Simulate data ##
+###################
+
+set.seed(42)
+simdata <- data.frame(
+  x1 = runif(500, -10, 30),
+  x2 = runif(500, 35, 50),
+  x3 = runif(500, 10, 12),
+  x4 = rnorm(500, 5, 5),
+  x5 = rnorm(500, 0, 10),
+  x6 = rnorm(500, -10, 15),
+  x7 = rnorm(500, sd = 7.5),
+  x8 = rnorm(500, sd = 7.5),
+  x9 = runif(500, 8, 14),
+  x10 = rnorm(500, -5, 7)
+)
+rules <- list(
+  rule1 = "x1 > 5 & x2 < 40",
+  rule2 = "x3 > 11.25 & x4 < 6",
+  rule3 = "x5 > 2 & x6 < -12.5"
+)
+error <- rnorm(500, 0, 5)
+for(i in 1:length(rules)) {
+  simdata[[paste("rule",i,sep="")]] <- as.numeric(with(simdata, eval(parse(text = rules[[i]]))))
+}
+
+simdata$y <- with(simdata, x8*.15 + x9*.8 + rule1*15 + rule2*15 + rule3*15) + error
+cor(simdata$y, with(simdata, x8*.15 + x9*.8 + rule1*3 + rule2*3 + rule3*3))
+cor(simdata$y, error)
+apply(simdata, 2, sd)
+cor(simdata[,c(paste("rule",1:3,sep=""),paste("x",7:10,sep=""),"y")])
+simdata2 <- simdata[,c(paste("x", 1:10, sep=""), "y")]
+simdata2 <- round(simdata2, digits = 1)

### 2016-08-30 pkg/glmertree/upre/paper/unbiased rule ensembles.R

@@ -1,0 +1,193 @@
+library(partykit)
+library(glmnet)
+library(MatrixModels)
+source("C:/Users/tobii/Desktop/swReg/upre/R/upre.R")
+
+## test on air quality data:
+airq <- airquality[complete.cases(airquality),]
+dim(airq)
+# some tests to see what ctree does with factors:
+plot(ctree(Ozone ~ ., data=airq))
+airq.f <- airq
+airq.f$Ozone <- factor(airq$Ozone > median(airq$Ozone))
+plot(ctree(Ozone ~ ., data=airq.f))
+airq.f2 <- airq
+airq.f2$Ozone <- cut(airq$Ozone, breaks = 4)
+plot(ctree(Ozone ~ ., data=airq.f2))
+
+plot(ctree(Ozone ~ offset(Temp) + Temp, data=airq.f))
+plot(ctree(Ozone ~ Temp, data=airq.f))
+tmp <- airq$Temp
+# ctree does not require a family. Automatically uses continous or (multi)nominal methods. 
+# if learning rate > 0, ctree cannot be used for factors, as it does not take offset argument, 
+# but glmtree does. Note: this gives a different tree than ctree!
+plot(glmtree(Ozone ~ 1 | Temp, data=airq.f, family = "binomial", offset = tmp))
+
+predict(glmtree(Ozone ~ 1 | Temp, data=airq.f, family = "binomial"), newdata = airq.f)
+
+
+
+
+
+airq.ens <- upre(Ozone ~ ., data=airq)
+airq.ens.f <- upre(Ozone ~ ., data=airq.f, learnrate = 0)
+airq.ens.f2 <- upre(Ozone ~ ., data=airq.f, learnrate = .000001)
+predict(airq.ens.f)
+predict(airq.ens.f, type = "response")
+predict(airq.ens.f, type = "class")
+predict(airq.ens.f2)
+predict(airq.ens.f2, type = "response")
+predict(airq.ens.f2, type = "class")
+
+airq.coefs <- coef(airq.ens)
+airq.coefs[airq.coefs$coefficient>0,]
+importance(airq.ens)
+plot(airq.ens$glmnet.fit)
+
+airq.f.coefs <- coef(airq.ens.f)
+airq.f.coefs[airq.f.coefs$coefficient!=0,]
+importance(airq.ens.f)
+plot(airq.ens.f$glmnet.fit)
+
+singleplot(airq.ens, "Temp")
[…]

### 2016-09-02 pkg/glmertree/paper/treatment_interactions.Rnw
Paper and simulation update: - Linear to piecewise interactions simulation added - Various changes in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@961 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{bigfoot}
\usepackage{ctable}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{placeins}
\linespread{1.2}
\usepackage{tabto}

%% for \usepackage{Sweave}
\SweaveOpts{echo=FALSE, results=hide}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries>>=
library("foreign")
library("glmertree")
library("lattice")
@

\title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
\shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
\leftheader{Treatment-Subgroup Interactions with GLMM Trees}
\rightheader{Treatment-Subgroup Interactions with GLMM Trees}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

\acknowledgements{The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application. The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014. The visit was supported by the Institute.}

\begin{document}

\maketitle
\pagewiselinenumbers

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but datasets suited for their detection may often have a clustered or multilevel structure. For example, observations in a dataset may be nested within research centers or trials. Therefore, we propose generalized linear mixed-effects model trees (GLMM tree), allowing for the detection of treatment-subgroup interactions, as well as taking into account the nested structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with thattrees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of datasets, and trees without random effects in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4 and 33\%, respectively). Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal accuracy in predicting treatment-effect differences, but better accuracy in detecting interactions. We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

\section{Introduction}

In research on the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified (or personalized) medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. Often, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized. This may hamper identification of which treatment works best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate instead of test such hypotheses.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several tree-based algorithms to the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, which allows users to select from a range of statistical methods, including trees. As such, the \citeA{ZhanyTsia12Stat} method can also be referred to as a tree-based method.

Often, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection, also take into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. GLMM tree allows for taking into account the clustered structure of datasets \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree. In the Empirical Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects. In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. Finally, in the Discussion we summarize the results and point out some directions for future research.

\subsection{Artificial motivating dataset}

<<artificial-data>>=
set.seed(123)
treatment <- rbinom(n=150,size=1,prob=1/2)
duration <- round(matrix(rnorm(150,mean=5,sd=3),ncol=1))
duration[duration<0] <- 1
anxiety <- round(rnorm(150,mean=10,sd=3))
age <- round(rnorm(150,mean=30,sd=10))
age[age<18] <- 18
age[age>75] <- 75
cutscore1 <- 5+.44*3
cutscore2 <- 10
node1t1 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==0, 1, 0)
node1t2 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==1, 1, 0)
node2t1 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==0, 1, 0)
node2t2 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==1, 1, 0)
node3t1 <- ifelse(duration>cutscore1 & treatment==0, 1, 0)
node3t2 <- ifelse(duration>cutscore1 & treatment==1, 1, 0)
beta1 <- -2
beta2 <- 0
beta3 <- 2.5
error <- rnorm(150,0,2)
cluster <- error+rnorm(150, 0, 6)
int_values <- rnorm(10, 0, 1)
tmp <- rep(int_values,each=15)
tmp <- tmp[order(tmp)]
rand_int <- rep(NA, times=150)
rand_int[order(cluster)] <- tmp
cluster[order(cluster)] <- rep(1:10, each=15)
cor(error,rand_int)
true_error <- error - rand_int
true_eff <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
     + .4*treatment)
outcome <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
                      + .4*treatment + true_error + rand_int)
outcome[outcome<0] <- 1
example <- data.frame(treatment, outcome, duration, anxiety, cluster, age)
example$treatment <- factor(example$treatment, labels = c("Treatment 1", "Treatment 2"))
@

We created generated a dataset representing observations on 150 participants in a randomized clinical trial. Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).

The simulated dataset has 3 subgroups with differential treatment effectiveness. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, Treatment~1 is more effective than Treatment~2: the mean of the response variable is 7 for Treatment~1, and 11 for Treatment~2. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment~1, and 9 for Treatment~2. The third subgroup consists of observations with duration $>6$. In this subgroup, Treatment~2 is more effective than Treatment~1: the mean value of the response variable is 12 for Treatment~1, and 7 for Treatment~2.

Participants were part of one of ten clusters, each with a different value for the intercept. Data was generated such that covariates and cluster-specific intercepts were uncorrelated. Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.

<<example-trees>>=
exampleglm <- lmtree(outcome ~ treatment | age, data = example)
exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = example)
exampleglmmtree <- lmertree(outcome ~ treatment | cluster | age + duration + anxiety, data = example)
@

\subsection{Model-based recursive partitioning}

The rationale behind MOB is that a global parametric model may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parameters of more complex models across terminal nodes.

For example, let us take a global GLM to estimate the overall treatment effect in the artificial motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment group. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper; that is, two treatment conditions and no additional covariates are included in the GLM. Furthermore, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

\begin{figure}[t!]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a global GLM for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

The result of fitting a global GLM to the artificial motivating dataset is depicted in Figure~\ref{fig:fixedeffects}; the boxplots show the distribution of the posttreatment depression scores in both treatment groups. The global model does not describe the data well: there is substantial residual variance and the slope of the regression line is nearly zero. This does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

The MOB algorithm detects such moderation, by testing for parameter stability over a set of auxiliary covariates or partitioning variables. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. By definition, the empirical scores of all observations in a dataset sum to zero. When the model is correctly specified, the expected value of the score for each observation is also zero. Therefore, under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}. To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). The binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on a GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node. As a result, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all coefficients interact with the subgroup indicator. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

\begin{figure}[t!]
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
\label{fig:example_mobtree}
\end{figure}

Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the artificial motivating dataset. By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above. However, the two subgroups in the middle do not represent true subgroups, and this may be due to the clustered structure of the dataset not being taken into account.

\subsection{Generalized linear mixed-effects model trees}

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

For simplicity, we assume that only cluster-specific intercepts are included in the model in the current paper. However, multiple random-effects coefficients can easily be included.

Note that, if the random-effects coefficients were known, the model could be estimated by a simple GLM as in Equation \ref{eq:fixedeffects} where $z_{i}^{\top}b$ would only be added as an offset (i.e., a variable with a fixed coefficient of 1) to the linear predictor.

Although GLMMs are adequately suited for taking into account the nested structure of a datasets, the global fixed-effects model in Equation \ref{eq:mixedeffects} may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} and random effects are incorporated as well: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random effects.

For estimating GLMM trees, we take this approach a step further: we take a GLM tree, instead of CART tree with constant fits to estimate the fixed-effects part of the GLMM. This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:

\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

  \item[Step 1:] Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$), with $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.

  \item[Step 2:] Estimate random effects in the mixed-effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.

  \item[Step 3:] Repeat Steps~1 and~2 until convergence.
\end{description}

The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown. In every iteration, the GLM tree and coefficients $\beta_{j(r)}$ and $b_{(r)}$ are re-estimated. The GLM tree is estimated, given the estimated random effects from the last iteration, and the random effects are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}.

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the artificial motivating dataset is presented. As can be seen, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.

\begin{figure}[t!]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

\section{Empirical Evaluation: Method}

To assess and compare the performance of GLMM tree, we performed three simulation studies. In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions. In Study II, we compared the accuracy of GLMM tree with that of GLM tree in datasets without treatment-subgroup interactions, allowing for assessment of the Type-I error rate. In Study III, we compared the accuracy of GLMM tree with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

\subsection{Software}

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge). This package provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.

In all simulations, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing. The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight.

\subsection{Treatment-subgroup interaction design} 
For generating datasets with treatment-subgroup interactions, we used one of the treatment-subgroup interaction designs from \citeA{DussyMech14}, which is depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by values on the (true) partitioning variables $U_2$, and $U_1$ or $U_5$. The other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

\begin{figure}[t!]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(3L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0), U5 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

\subsubsection{Data-generating parameters} 
In generating datasets, we varied seven parameters of the data-generating process:

\begin{enumerate} 
\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}. 
\item Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes. In datasets with treatment-subgroup interactions, the treatment effect difference varied only in nodes 4 and 7, with levels $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

\subsection{Variable distributions}
Following the approach of \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between $U$ variables vary according to the third facet of the data-generating design described above.

To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design. For each cluster, a single value $b_m$ was drawn from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design. When $b$ was correlated with one of the potential partitioning variables, this variable was randomly selected.

To generate node-specific fixed effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3. In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$. For every observation, we generated a binomial variable (with probability .5) as an indicator for treatment type. Random errors $\epsilon$ were drawn from a normal distribution with $\mu_{\epsilon} = 0$ and $\sigma_{\epsilon} = 5$.

Finally, the response variable was calculated as the sum of the (node-specific) fixed effects, the random intercept and the error term: $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.

\subsection{Assessment of performance}

Firstly, the total number of nodes in estimated GLM and GLMM tree were calculated to assess the extent to which the algorithms may detect spurious subgroups. Also, the number of nodes allow for assessing Type-I (the extent to which datasets are erroneously partitioned) and Type-II (the extent to which datasets are erroneously not partitioned) error.

Secondly, in datasets with treatment-subgroup interactions, we assessed the accuracy of estimated models. For GLM and GLMM trees, an accurately recovered tree was defined as a tree with (1) the true tree size (i.e., tree size $= 7$ in datasets with treatment-subgroup interactions), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$. Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).

Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) in each dataset. Consequently, predictive accuracy was only be assessed in datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions. To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets. Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from 'new' clusters.

\section{Empirical Evaluation: Results}

\subsection{Study I: Piecewise interactions}

For each cell of the data-generating design described above, 50 datasets were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets with piecewise treatment-subgroup interactions. In these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).

<<simulation-data>>=
load("treespecs_long.dat")
treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)
treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered=T)
treespecs.long$treatdiff <- factor(as.numeric(as.character(treespecs.long$treatdiff)), ordered=T)
treespecs.long$truetree.values <- as.numeric(treespecs.long$truetree.values)-1
levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="bi and splitting U correlated"] <- "b correlated with splitting U"
levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="bi and non-splitting U correlated"] <- "b correlated with non-splitting U"
levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="uncorrelated"] <- "b and U uncorrelated"
treespecs.long <- treespecs.long[treespecs.long$correlation.ind!="GLMM",]
treespecs.long$correlation.ind <- factor(treespecs.long$correlation.ind)
treespecs.long$treesize.ind <- factor(treespecs.long$treesize.ind)
@

\begin{figure}[t!]
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent levels of sample size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

\subsubsection{Tree size}

GLMM tree yielded trees with an average size of 7.15 nodes ($\mathrm{SD}=0.61$), whereas GLM tree yielded an average tree size of 8.15 nodes ($\mathrm{SD}=2.05$), indicating that GLM trees involved more spurious splits than GLMM trees, on average. The estimated probability that a dataset was erroneously not partitioned (the Type-II error) was 0 for both algorithms.

To find the most important predictors of tree size, we performed an ANOVA with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors of tree size were algorithm type, sample size, variance of the random intercept, and the correlation between partitioning variables. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treesize_interact}).

Figure~\ref{fig:xyplot_treesize_interact} indicates that average tree size increases slightly with sample size, which may be due to increased power. In the absence of random effects (i.e., $\sigma_{b}=0$), both GLM and GLMM trees have about 7 nodes on average. However, clear differences in tree size were observed when $\sigma_b > 0$. When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average. When random intercepts are not correlated with a partitioning covariate and sample size is large (i.e., 500 or 1000), GLM and GLMM trees are about equally sized and have approximately the true tree size, on average. When random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10). This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.

\begin{figure}[t!]
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.acc <- aggregate(formula=truetree.values ~ correlation.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=correlation.ind, type="b", 
ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
)
@
\caption{Tree accuracy of GLM and GLMM tree. Tree accuracy is defined as the proportion of datasets in which the true tree was accurately recovered. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$, columns represent levels of sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

\subsubsection{Accuracy of recovered trees}

To assess the accuracy of trees recovered by both algorithms, we inspected variables and values that were selected for partitioning. For the first split, GLMM tree selected the true partitioning variable ($U_2$) in all datasets, and GLM tree in all but one datasets. The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}). GLM tree did show somewhat higher variability in recovering the splitting value for the first split than GLMM tree ($\mathrm{SD}=0.154$ and $\mathrm{SD}=0.127$, respectively).

GLMM tree accurately recovered the partition in 90.40\% of datasets, whereas GLM tree accurately recovered the partition in only 61.44\% of datasets. To find the most important predictors of tree accuracy, we fitted a GLM with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and the level of dependence between the partitioning variables and the random intercept. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treeaccuracy}).

Figure~\ref{fig:xyplot_treeaccuracy} indicates that in the absence of random effects, the trees recovered by GLM and GLMM tree were about equally accurate. In the presence of random effects, GLM trees were much less accurate than GLMM trees. This was found for all sample sizes, when random effects were correlated with a partitioning covariate. This effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable. When random intercepts were not correlated with one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., $N = 200$).

\begin{figure}[t!]
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T),
scales=list(y=list(limits = c(0,1))))
)
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Rows represent the levels of the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences, columns represent levels of sample size.}
\label{fig:xyplot_correlations}
\end{figure}

\subsubsection{Predictive accuracy of trees}
To assess predictive accuracy, correlations between the true and predicted treatment-effect differences of both algorithms were calculated in test datasets. The predicted treatment-effect differences of GLMM tree were closer to the true differences than those of GLM tree, with a mean correlation of .93 ($\mathrm{SD}=0.12$) for GLMM tree and .88 ($\mathrm{SD}=0.19$) for GLM tree. An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the treatment difference effect size. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_correlations}).

Figure~\ref{fig:xyplot_correlations} shows clear main effects of sample size and treatment-effect differences for both algorithms. In the absence of random effects (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were about equally accurate. However, it should be noted that in the absence of random effects, GLMM tree tends to outperform GLM tree in smaller sample sizes (i.e., 200), whereas GLM tree tends to outperform GLMM tree in larger sample sizes (i.e., 500 and 1000). In the presence of random effects, GLM tree predictions were much less accurate than those of GLMM tree when treatment-effect differences were smaller (i.e., Cohen's $d = .5$). This effect was also observed, but much less pronounced, when treatment-effect differences were larger (i.e., Cohen's $d = 1.0$). In addition, with increasing sample size and treatment-effect differences, the difference in predictive accuracy between GLM and GLMM tree becomes smaller.

\subsection{Study II: Type-I error}

\subsubsection{Design}
For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there is only a main effect of treatment in the population. Put differently, the number of subgroups or terminal nodes in these datasets was $J=1$, and there was only a single value of $\beta_j = \beta$ in every dataset. For the datasets without treatment-subgroup interactions, we varied the same parameters of the data-generating process as in Study I. However, the sixth facet of the data-generating process had only two levels for these datasets, as there were no `true' partitioning variables in these datasets. Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.

\subsubsection{Tree size and Type-I error}
In datasets without treatment-subgroup interactions, the average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree. The overall Type-I error rate was only .04 for GLMM tree, and .33 for GLM tree. Again, main predictors of tree size were found to be sample size, variance of the random intercepts, and dependence between the random intercept and one of the partitioning variables. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treesize_nointeract}). Figure~\ref{fig:xyplot_treesize_nointeract} indicates that in the absence of random effects (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tend to create trees of size 1. In the presence of random effects, GLMM tree still tends to create trees of size 1, but GLM tree creates much larger trees, when the random effects are correlated with one of the partitioning covariates. This effect increases with sample size.

\begin{figure}[t!]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("noint_treesizes_long.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

\subsection{Study III: Linear and piecewise interactions}

\subsubsection{Design}
To compare the performance of GLMM tree with that of GLMMs with pre-specified interaction effects, we created datasets in which interactions were varied from purely linear to purely piecewise. The number of cells in the simulation design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters). The fifth facet of the data-generating design was limited to two levels ($\sigma_b = 2.5$ and $\sigma_b = 7.5$). As in Study II, we limited the sixth facet to two levels (the random intercept was either correlated to one of the partitioning variables, or not). To counter unidentifiability due to potentially large numbers of pre-specified interaction terms in the GLMMs, we limited the number of covariates to $k=5$ and $k=10$ (instead of $k=5$ and $k=15$). Finally, to allow for a smooth transition from linear to piecewise interactions, the seventh facet of the interaction design (effect size of treatment differences) was replaced by a factor with three levels, controlling the type of interactions: purely linear interactions, purely piecewise interactions, and a combination of both. As a result, $50 \times 3 \times 2 \times 2 \times 1 \times 2 \times 2 \times 3$ = 7,200 training datasets were generated for this study.

All input variables were generated as described above. To generate purely linear interactions, outcome variable $y$ was generated as the sum of:

\begin{enumerate}
\item a main effect of $U_2$ (coefficient = 1.5) 
\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = -.25), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = .25)
\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = -.25), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = .25)
\end{enumerate}

Note that the direction (positive or negative) and type (main or interaction) of these effects correspond to those in Figure~\ref{fig:modelC}. All interactions were created by using centered $U_k$ variables, calculated by subtracting the variable means. To generate purely piecewise interactions, we used the partition depicted in Figure~\ref{fig:modelC}. We estimated $\bar{y}$ for each of the treatment-subgroups by calculating subgroup means in a large dataset (N = $10^6$) with purely linear interactions. For each observation in datasets with purely piecewise interactions, the treatment-subgroup specific value for $\bar{y}$ was imputed. To generate combined linear and piecewise interactions, $y$ values according to the purely piecewise and linear models were summed and divided by 2.

GLMMs were estimated by specifying main effects for all covariates $U_k$ and the treatment indicator, first-order (or two-way) interactions between all pairs of covariates $U_k$, and second-order (or three-way) interactions between all pairs of covariates $U_k$ and treatment. Accurately recovered GLMMs were defined to be models in which significant (i.e., absolute value of the $t$-test statistic $\geq 1.96$) effects were found for $U_2$ and (treatment) interactions of $U_1$ and $U_2$ and $U_2$ and $U_5$. Accurately recovered GLMM trees were defined to be trees in which the first split involved $U_2$, the second split to the right involved $U_1$ and the third split to the right involved $U_5$. Note that this criterion of accuracy is somewhat more lenient than in Study I.

<<linear-simulation-data>>=
load("pw_treespecs_long.dat")
treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)
treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered=T)
treespecs.long$type <- factor(treespecs.long$type, ordered = TRUE, levels = 
c("linear", "both", "piecewise"))
@

\subsubsection{Accuracy of recovered models}

Overall, GLMM tree accurately recovered main and interaction effects in 53.58\% of datasets, wheareas GLMMs accurately recovered main and interaction effects in 34.81\% of datasets. To find the most important predictors of model accuracy, we performed an ANOVA, in which the algorithm type, sample size, the number of covariates and type of interaction (linear, piecewise or both) were found to be most important. These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_accuracy_pw}.

Figure~\ref{fig:xyplot_accuracy_pw} indicates a clear effect of sample size: when the number of covariates is large (i.e., 10) GLMMs do not recover all true effects in the dataset, whereas they always seem to recover the true effects when the number of covariates is small (i.e., 5), and the effects are partly linear. However, when the effects are purely piecewise, GLMMs never recover the true effects in the dataset, unless sample size is very large (i.e., $N = 1000$). At the same time, Figure~\ref{fig:xyplot_accuracy_pw} indicates that the accuracy of GLMM trees is unaffected by the number of covariates. Furthermore, GLMM tree performs best in recovering purely piecewise interactions, and poorest in recovering purely linear interactions. Finally, the performance of GLMM tree is affected by sample size: generally, the true effects are more often recovered in datasets with larger sample sizes.

\begin{figure}[t!]
<<accuracy-pw-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.acc <- aggregate(formula=truevars ~ truevars.ind + N + type + np, FUN=mean, 
data=treespecs.long)
xyplot(truevars ~ type | N + np, data = aggdata.acc, groups=truevars.ind, type="b",
ylab="proportion of accurately recovered models", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
@
\caption{Model accuracy for GLMMs and GLMM trees. In an accurate model, the true main and interaction effects were accurately recovered. Columns represent sample size, rows represent the number of covariates in the model.}
\label{fig:xyplot_accuracy_pw}
\end{figure}

\subsubsection{Predictive accuracy}

Both algorithms showed equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees. In terms of bias, GLMM treatment-effect difference predictions deviated more from the true values than the predictions of GLMM tree. The mean true treatment-effect difference averaged over all datasets was .04 (SD = 1.22), whereas the mean predicted treatment-effect differences averaged over all datasets for GLMMs was -.26 (SD = 100.33) and for GLMM trees .03 (SD = 3.88). Furthermore, it should be noted that the full models were used for prediction with GLMMs and GLMM trees. That is, predictions for GLMMs in datasets with 10 covariates may involve 11 (significant or non-significant) main effects, 100 first-order interactions and 100 second-order interactions. As the maximum tree depth of GLMM trees was set to 4, GLMM trees used a maximum of only 8 terminal nodes for predictions.

The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (linear, piecewise or both). These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}. As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely linear interactions. The predictive accuracy of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs decreases when the number of covariates increases. Finally, for both algorithms, predictive accuracy improves somewhat when sample size increases.

\begin{figure}[t!]
<<correlation-pw-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + np + type + N, FUN=mean, 
data=treespecs.long)
xyplot(correlation.values ~ type | N + np, data = aggdata.cor, groups=correlation.ind, type="b",
ylab="correlation", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
@
\caption{Average predictive accuracy of GLMMs and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Columns represent sample size, rows represent the number of covariates.}
\label{fig:xyplot_correlation_pw}
\end{figure}

\section{Application: Individual patient-level meta-analysis dataset on treatments for depression}

<<depression-data>>=
metadata <- read.dta("Database IPDMA CBT PHA Version 11.dta")
metadata[metadata == 999] <- NA
metadata[metadata == 888] <- NA
vars <- c("studyid", "Tx_group", "Age", "Gender", "education", "ComorbidAnxietyDisorder", "HRSDt0", "HRSDt1")
factors <- c("studyid", "Tx_group", "Gender", "education", "ComorbidAnxietyDisorder")
metadata$education <- factor(metadata$education, ordered = TRUE)
for (i in 1:length(factors)) {metadata[, factors[i]] <- factor(metadata[,factors[i]])}
metadata <- metadata[vars] # select only relevant variables
metadata <- metadata[complete.cases(metadata[, vars]), ] # select only complete data
metadata <- metadata[!metadata$Tx_group == "placebo", ] # remove placebo observations
metadata$Tx_group <- factor(metadata$Tx_group)
@

\subsection{Method}

To further illustrate and compare the application of GLM and GLMM tree, we applied both algorithms to a dataset from a individual-patient data meta-analysis of \citeA{CuijyWeit14}. This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression. The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression. Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}). \citeA{CuijyWeit14} found no indication that gender predicted or moderated treatment outcomes. Further details on the dataset can be found in \citeA{CuijyWeit14}.

In our analyses, posttreatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score. The predictor variable in the linear model was treatment type (0 = CBT and 1 = PHA). An indicator for study was used as the cluster indicator.

In RCTs, ANCOVAs are often employed, to linearly control posttreatment values on the outcome measure for pretreatment values. Therefore, we estimated the GLM and GLMM trees using posttreatment HAM-D scores, controlled for the linear effects of pretreatment HAM-D scores. The trees were grown using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score. As a result, data from 694 patients from 7 studies were included in the analyses. Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.

To provide a standardized estimate of the treatment effect differences in the final nodes of the trees, we calculated node-specific Cohen's $d$ values. Cohen's $d$ was calculated by dividing the node-specific predicted treatment outcome difference by the node-specific pooled standard deviation. Confidence intervals for Cohen's $d$ could not be calculated, as these can not take into account the exploratory nature of our analyses (i.e., variable and split point selection). The predictive accuracy of GLM and GLMM trees was assessed by calculating average correlations between observed and predicted HAM-D posttreatment scores, based on 50-fold cross validation.

The results of recursive partitioning techniques are known to be potentially unstable, in the sense that samll changes in the dataset may substantially alter the variables or values selected for partitioning. Therefore, we used the R-package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values. Using the \verb|stabletree| function, we calculated the number of times a variable and value were selected for partitioning, over 500 subsamples of size $.9 \times N$ of the dataset.

<<depression-trees>>=
## offset
metadata$HRSDfit <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))

## GLM tree
lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, offset = HRSDfit)

## GLMM tree
lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(HRSDfit) | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, ranefstart = metadata$HRSDfit)

## GLM tree with effect sizes: 
GLMtree_app_eff <- party(
	partynode(1L,
		split = partysplit(1L, breaks = 2), kids = list(
			partynode(2L, info = c(
				expression(''),
				expression(widehat(mu)[CBT] == '13.25'),
				expression(''),
				expression(widehat(mu)[PHA] == '10.29'),
				expression(''),
				expression(widehat(sigma)[pooled] == '7.52'),
				expression(''),
				expression(widehat(d) == '0.39'),
				expression(''))),
			partynode(3L, split = partysplit(2L, breaks = 1), kids = list(
				partynode(4L, info = c(
				expression(''),
				expression(widehat(mu)[CBT] == '7.82'),
				expression(''),
				expression(widehat(mu)[PHA] == '7.55'),
				expression(''),
				expression(widehat(sigma)[pooled] == '5.51'),
				expression(''),
				expression(widehat(d) == '0.05'),
				expression(''))),
			partynode(5L, info = c(
				expression(''),
				expression(widehat(mu)[CBT] == '10.35'),
				expression(''),
				expression(widehat(mu)[PHA] == '7.69'),
				expression(''),
				expression(widehat(sigma)[pooled] == '6.64'),
				expression(''),
				expression(widehat(d) == '0.40'),
				expression(''))))))),
	data.frame(education = metadata$education, ComorbidAnxietyDisorder = metadata$ComorbidAnxietyDisorder)
)

## GLMM tree with effect sizes:
GLMMtree_app_eff <- party(
	partynode(1L,
		split = partysplit(2L, breaks = 1), kids = list(
			partynode(2L, info = c(
				expression(''),
				expression(widehat(mu)[CBT] == '8.27'),
				expression(''),
				expression(widehat(mu)[PHA] == '7.99'),
				expression(''),
				expression(widehat(sigma)[pooled] == '5.86'),
				expression(''),
				expression(widehat(d) == '0.05'),
				expression(''))),
			partynode(3L, info = c(
				expression(''),
				expression(widehat(mu)[CBT] == '10.84'),
				expression(''),
				expression(widehat(mu)[PHA] == '8.21'),
				expression(''),
				expression(widehat(sigma)[pooled] == '6.80'),
				expression(''),
				expression(widehat(d) == '0.39'),
				expression(''))))),
	data.frame(education = metadata$education, ComorbidAnxietyDisorder = metadata$ComorbidAnxietyDisorder)
)
@

\begin{figure}[t!]
\setkeys{Gin}{width=0.75\textwidth}
<<depression-glmtree, fig=TRUE, height=6, width=8>>=
plot(lmtree_app)
plot(GLMtree_app_eff, tp_args = list(FUN = identity, height = 10, width = 14), tnex = 1.5)
@
\caption{Upper panel: GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA). Lower panel: Subgroup-specific descriptive statistics.}
\label{fig:lmtree_C&W}
\end{figure}

\subsection{Results}

The tree and effects sizes resulting from application of GLM tree are presented in Figures~\ref{fig:lmtree_C&W} and \ref{fig:lmtree_C&W_eff} and those resulting from application of GLMM tree are presented in Figures~\ref{fig:lmertree_C&W} and \ref{fig:lmertree_C&W_eff}.

The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education. By taking into account study-specific intercepts, the GLMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split made by GLM tree is a spurious one. The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable. The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Figure~\ref{fig:lmertree_C&W_eff}). For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Figure~\ref{fig:lmertree_C&W_eff}). The estimated intraclass correlation coefficient for the GLMM tree was .05.

Assessment of predictive accuracy by means of 50-fold cross validation indicated better predictive accuracy for GLMM tree than GLM tree. The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathit{var}=.067$) for GLMM tree, and .19 ($\mathit{var}=.084$) for GLM tree. This indicates that GLMM tree provided higher predictive accuracy, and also somewhat lower variability of predictive accuracy than GLM tree.

Table \ref{tab:stability} presents statistics on the variables selected for partitioning in subsamples of the dataset. Note that the selection frequencies do not add up to 1, as trees may involve multiple, or no splits. Table \ref{tab:stability} indicates that the presence of a comorbid anxiety disorder was selected for partitioning in the majority ofGLMM trees grown on subsamples of the dataset, and all other variables were selected in less than 3\% of the subsamples. As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.

\begin{figure}[t!]
\setkeys{Gin}{width=0.5\textwidth}
<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
plot(lmertree_app)
@
%\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).}
\caption{Upper panel: GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA). Lower panel: Subgroup-specific descriptive statistics.}
\label{fig:lmertree_C&W}
\end{figure}

\begin{figure}[t!]
\setkeys{Gin}{width=0.4\textwidth}
<<depression-glmmtree-eff, fig=TRUE, height=3, width=4>>=
plot(GLMMtree_app_eff, tp_args = list(FUN = identity, height = 10, width = 14), tnex = 1.5)
@
\caption{Statistics and effect sizes for GLM tree in Figure \ref{fig:lmertree_C&W}.}
\label{fig:lmertree_C&W_eff}
\end{figure}

\begin{table}
	\small
	\caption{Variable selection statistics}
	\begin{tabular}{lcc}
		\thickline
		&	\multicolumn{2}{c}{Selection frequency} \\
		Variable				&	GLM tree&	GLMM tree	\\
		\hline
		Education               &	0.956	&	0.014		\\
		ComorbidAnxietyDisorder	&	0.398	&	0.528		\\
		HRSDt0                  &	0.034	&	0.002		\\
		Age                     &	0.000	&	0.022		\\
		Gender                  &	0.002	&	0.004		\\
		\hline
		\multicolumn{3}{l}{\textit{Note.} Frequencies are calculated over 500 random}\\
		\multicolumn{3}{l}{subsamples of size $.9 \times N$ of the complete dataset.}\\
		\multicolumn{3}{l}{Frequencies do not add up to 1, as trees may involve}\\
		\multicolumn{3}{l}{multiple, or no splits.}\\
	\end{tabular}
	\label{tab:stability}
\end{table}

\section{Discussion}

In the current paper, we presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters. As such, we hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies.

GLMM tree performed very well in recovering treatment-subgroup interactions, accurately recovering interactions in 90\% of datasets with treatment-subgroup interactions. In contrast, GLM tree accurately recovered interactions in only 61\% of these datasets. In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets. Put differently, the Type-I error rate of GLMM tree very closely approximated the $\alpha$ level used for testing parameter stability, whereas the Type-I error rate of GLM tree clearly exceeded this value.

The better performance of GLMM tree was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables. In these instances, random effects gave rise to spurious subgroup detection (spurious splits) by GLM tree, both in datasets with and without treatment-subgroup interactions.

GLMM tree also provided better predictive accuracy than GLM tree, with an average correlation between true and predicted treatment differences of .94 for GLMM tree, and .88 for GLM tree. GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, treatment-effect differences were relatively small (i.e., Cohen's $d=.5$), and/or sample size was small (i.e., $ N < 1,000$). Such treatment-effect differences and sample sizes may be quite common, even in multi-center clinical trials, and GLMM tree may provide a helpful tool for subgroup detection in those instances.

As expected, when random effects were absent from the simulated datasets, GLM and GLMM tree yielded very similar predictive accuracy. This indicates that GLMM tree can be applied whenever cluster-specific random effects are expected: In the absence of random effects, GLM tree and GLMM tree are expected to perform about equally well and in the presence of random effects GLMM tree is expected to outperform GLM tree.

Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided similar predictive accuracy, on average. When interactions were purely linear, GLMMs outperformed GLMM trees, and when interactions were purely piecewise, GLMM trees outperformed GLMMs. However, GLMM tree may have a clear advantage when there are a large number of potential moderator variables (i.e., $> 5$). With 10 potential moderator variables, $t$-tests of GLMMs were unable to identify the true moderator variables, whereas the number of potential moderator variables did not influence performance of GLMM tree. Furthermore, the number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex predictive models. The trees in our simulations were limited to a maximum number of 8 terminal nodes, and may therefore be much easier to use for prediction in practical decision-making contexts.

In the Application, we found GLMM tree to provide results that are easily interpretable, and also more accurate than a GLM tree without random effects. In addition, node-specific means and variances can be used to calculate effect sizes, to judge clinical relevance of the findings, as would often be done in RCTs or meta-analysis. Although we have limited ourselves to calculating Cohen's $d$ in the current paper, equivalent values of the success rate difference or the number needed to treat can be calculated, but this would involve additional distributional assumptions \cite{KraeyKupf06}. Node-specific effect sizes can also be used to prune trees, when a researcher prefers to have a final tree which is based on statistical as well as clinical significance. A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significant in the tree-growing process.

These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures. However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified model. That is, if there are subgroups with respect to the partitioning variables specified, so that there are different parameters of the GLM in each of these subgroups, then GLMM tree will accurately recover those subgroups. However, as with any data-analytic method, misspecification of the model will negatively affect performance. One source of misspecification would be the omission of relevant variables in the GLM, or as potential partitioning variables. Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups. However, the performance of GLMM tree, in contrast to that of GLMMs with pre-specified interactions, was not negatively affected by the number of potential moderator variables.

As discussed in the Introduction, several tree-based methods for treatment-subgroup interaction detection are available. These methods have different objectives, and there is not yet an agreed-upon single best method. In a simulation study, \citeA{SiesyVanM16} found the method of \citeA{ZhanyTsia12Stat} to perform best, followed by model-based recursive partitioning. However, the method of Zhang et al. performed worst under some conditions of the simulation study in terms of the Type I error rate. Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations were limited to GLM and GLMM-based MOB.

Furthermore, it should be stressed that tree-based methods are exploratory tools. They can be used to detect predictors, interactions and non-linear effects in a data-driven way, but users should take the exploratory nature of such analyses into account. The resulting trees are potentially unstable, and stability of the results should be assessed, preferably in a dataset not used for training, or by multi-fold cross-validation or resampling techniques.

In conclusion, GLMM tree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatment effect differences, both in the presence and absence of cluster-specific random effects. Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials or individual-level patient data meta-analyses.

\nolinenumbers
\bibliographystyle{apacite}
\bibliography{bib}

\end{document}

### 2016-09-02 pkg/glmertree/paper/Application with CV.R

@@ -73,7 +73,7 @@ lm_es
 aggregate(metadata$HRSDt1, by = list(metadata$Tx_group, predict(lm_o, type = "node")), FUN = length)
 
 
-## 50 fold CV
+## 50-fold CV:
 metadata <- metadata[,vars]
 library(peperr)
 set.seed(32)
@@ -109,7 +109,7 @@ for (i in 1:50) {
 }  
 
 
-## Assess 50 fold CV results
+## Assess 50-fold CV results
 lmtreecors <- list()
 lmertreecors <- list()
 for (i in 1:50){
@@ -120,3 +120,28 @@ mean(unlist(lmtreecors))
 mean(unlist(lmertreecors))
 var(unlist(lmtreecors))
 var(unlist(lmertreecors))
+
+
+
+
+## Assess tree stability:
+library(stablelearner)
+# create trees without offset (otherwise, stabletree gives errors):
+lm_tree <- lmtree(HRSDt1 - HRSDfit ~ Tx_group | Age + Gender +
+                    education + ComorbidAnxietyDisorder + HRSDt0, data = metadata)
+lmer_tree <- lmertree(HRSDt1 - HRSDfit ~ Tx_group | (1 | studyid) |
+                        Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+                      data = metadata)
+set.seed(15)
+subsample <- function (S = 500) {
+  sampfun <- function(n) replicate(S, sample(1L:n, round(.9*n), replace = FALSE))
+  list(method = "Subsampling", sampler = sampfun)
+}
+
+s_lm_tree <- stabletree(lm_tree, sampler = subsample)
+s_lmer_tree <- stabletree(lmer_tree, sampler = subsample)
+summary(s_lm_tree)
+plot(s_lm_tree)
+summary(s_lmer_tree)
+plot(s_lmer_tree)
+

### 2016-09-02 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/Evaluation of results.R

@@ -1,16 +1,18 @@
-library(lme4)
-library(partykit)
+# install.packages("glmertree", repos="http://R-Forge.R-project.org")
+library(glmertree)
 
-# Evaluate tree sizes
-REEMobtreesize <- vector()
-Mobtreesize <- vector()
+
+## Evaluate tree size:
+
+GLMMtreesize <- vector()
+GLMtreesize <- vector()
 for (c in 1:50){
-  load(paste("treesize.REEMobs", c, sep=""))
-  REEMobtreesize <- c(REEMobtreesize, unlist(treesize.REEMobs))
-  load(paste("treesize.Mobs", c, sep=""))
-  Mobtreesize <- c(Mobtreesize, unlist(treesize.Mobs))  
+  load(paste("treesize.GLMM", c, sep=""))
+  GLMMtreesize <- c(GLMMtreesize, unlist(treesize.GLMM))
+  load(paste("treesize.GLM", c, sep=""))
+  GLMtreesize <- c(GLMtreesize, unlist(treesize.GLM))  
 }
-treespecs <- data.frame(REEMobtreesize, Mobtreesize)
+treespecs <- data.frame(GLMMtreesize, GLMtreesize)
 
 load("descriptions1")
 for (i in 1:length(descriptions)) {
@@ -31,221 +33,280 @@ treespecs$corUbi <- rep(factor(treespecs$corUbi[1:length(descriptions)]), times=
 treespecs$numbclus <- rep(factor(treespecs$numbclus[1:length(descriptions)]), times=50)
 treespecs$sigmabi <- rep(factor(treespecs$sigmabi[1:length(descriptions)]), times=50)
 
-table(treespecs$REEMobtreesize)
-prop.table(table(treespecs$REEMobtreesize))
-mean(treespecs$REEMobtreesize)
-sd(treespecs$REEMobtreesize)
+table(treespecs$GLMMtreesize)
+prop.table(table(treespecs$GLMMtreesize))
+mean(treespecs$GLMMtreesize)
+sd(treespecs$GLMMtreesize)
 
-table(treespecs$Mobtreesize)
-prop.table(table(treespecs$Mobtreesize))
-mean(treespecs$Mobtreesize)
-sd(treespecs$Mobtreesize)
+table(treespecs$GLMtreesize)
+prop.table(table(treespecs$GLMtreesize))
[…]

### 2016-09-02 pkg/glmertree/paper/Simulation replication/Datasets without treat-sub interactions/generate training data.R

@@ -1,7 +1,7 @@
 N <- list() # different levels for N
 N[[1]] <- 200; N[[2]] <- 500; N[[3]] <- 1000 
 
-bi <- list() # different numbers and values for the study intercepts
+bi <- list() # different number of clusters and variances of random intercepts
 bi$numbclus <- c(5,10,25)
 bi$sigmas <- c(0, 5, 10)
 
@@ -26,7 +26,7 @@ descriptions <- list()
 
 set.seed(24061983)
 
-for (c in 1:50) {  
+for (c in 1:1) {  
 counter <- 0
 for (d in 1:length(bi$sigmas)) {# d is counter for sigma of b_i's
   for (e in 1:length(bi$numbclus)) {# e is counter for number of clusters
@@ -72,7 +72,7 @@ for (d in 1:length(bi$sigmas)) {# d is counter for sigma of b_i's
               r$Y[r$T==1] <- r$Y[r$T==1]-diff/2              
               r$Y[r$T==2] <- r$Y[r$T==2]+diff/2
               
-              # add study intercept variable to datasets, which is randomly correlated with X1 thru X5
+              # add cluster-specific intercept to datasets, which is randomly correlated with X1 thru X5
               sigma_bi <- bi$sigmas[d]
               if(sigma_bi==0) {int_vals <- rep(rnorm(numbclus,0,300),each=nobs/numbclus)}
               if(sigma_bi!=0) {int_vals <- rep(rnorm(numbclus,0,bi$sigmas[d]),each=nobs/numbclus)} # generate values to be use as random intercepts
@@ -90,7 +90,9 @@ for (d in 1:length(bi$sigmas)) {# d is counter for sigma of b_i's
               r$bi[order(r$bi)] <- int_vals
               if(sigma_bi==0){r$Y <- r$Y}
               if(sigma_bi!=0){r$Y <- r$Y + r$bi}
-          
+              r$cluster <- factor(r$bi)
+              levels(r$cluster) <- 1:length(unique(r$bi))
+              
               # generate and add error to Y
               r$errorY <- rnorm(N[[i]], sd=5)
               r$Y <- r$Y + r$errorY
@@ -98,7 +100,7 @@ for (d in 1:length(bi$sigmas)) {# d is counter for sigma of b_i's
               # write dataset to list and add description
               counter <- counter+1
               r$T <- factor(r$T)
[…]

### 2016-09-02 pkg/glmertree/paper/Simulation replication/Linear to piecewise interactions/Analysis of datasets.R

@@ -1,0 +1,158 @@
+#install.packages("glmertree", repos="http://R-Forge.R-project.org")
+#library(glmertree)
+library(partykit)
+library(lme4)
+
+# source most recent glmertree code:
+source("C:/Users/tobii/Desktop/swReg/glmertree/R/glmertree.r")
+
+## Fit GLMM trees on training data:
+for (c in 1:50) {
+  print(c)
+  load(paste("datasets", c, sep=""))
+  load(paste("descriptions", c, sep=""))
+  GLMMtrees <- list()
+  for (i in 1:length(datasets)) {
+    print(i)
+    if (descriptions[[i]][[3]] == "number of covariates = 5") {
+      GLMMtreeformula <- Y ~ T | cluster | X1 + X2 + X3 + X4 + X5
+    }
+    if (descriptions[[i]][[3]] == "number of covariates = 10") {
+      GLMMtreeformula <- Y ~ T | cluster | X1 + X2 + X3 + X4 + X5 + X6 + X7 + 
+        X8 + X9 + X10
+    }
+    GLMMtrees[[i]] <- lmertree(GLMMtreeformula, data=datasets[[i]], maxdepth=4)
+  }
+  save(GLMMtrees, file=paste("GLMMtrees", c, sep=""))
+  rm(list=ls())
+  source("C:/Users/tobii/Desktop/swReg/glmertree/R/glmertree.r")
+}
+
+
+## Fit GLMMs on training data:
+for (c in 1:50) {
+  print(c)
+  load(paste("datasets", c, sep=""))
+  load(paste("descriptions", c, sep=""))
+  GLMMs <- list()
+  for (i in 1:length(datasets)) {
+    if (descriptions[[i]][[3]] == "number of covariates = 5") {
+    GLMMformula <- Y ~ (1 | cluster) + T + X1 + X2 + X3 + X4 + X5 + 
+      T:(X1 + X2 + X3 + X4 + X5) + (X1 + X2 + X3 + X4 + X5)^2 + 
+      T:(X1 + X2 + X3 + X4 + X5)^2
+    }
+    if (descriptions[[i]][[3]] == "number of covariates = 10") {
+      if(descriptions[[i]][[1]] == "N = 200") {
+        GLMMformula <- Y ~ (1 | cluster) + X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + T +
[…]

### 2016-09-02 pkg/glmertree/paper/Simulation replication/Linear to piecewise interactions/Evaluation of results.R

@@ -1,0 +1,220 @@
+# install.packages("glmertree", repos="http://R-Forge.R-project.org")
+library(glmertree)
+
+
+## Evaluate tree size:
+
+GLMMtreesize <- vector()
+for (c in 1:50){
+  load(paste("treesize.GLMM", c, sep=""))
+  GLMMtreesize <- c(GLMMtreesize, unlist(treesize.GLMM))
+}
+treespecs <- data.frame(GLMMtreesize)
+
+load("descriptions1")
+for (i in 1:length(descriptions)) {
+  treespecs[i,"N"] <- substr(descriptions[[i]][[1]], 5, 8)
+  treespecs[i,"rho"] <- substr(descriptions[[i]][[2]], 7, 9)
+  treespecs[i,"np"] <- substr(descriptions[[i]][[3]], 24, 25)
+  treespecs[i,"type"] <- substr(descriptions[[i]][[4]], 8, 18)
+  treespecs[i,"corUbi"] <- substr(descriptions[[i]][[5]], 32, 64)
+  treespecs[i,"numbclus"] <- substr(descriptions[[i]][[6]], 34, 35)
+  treespecs[i,"sigmabi"] <- substr(descriptions[[i]][[7]], 12, 14)  
+}
+
+treespecs$N <- rep(factor(treespecs$N[1:length(descriptions)]), times=50)
+treespecs$rho <- rep(factor(treespecs$rho[1:length(descriptions)]), times=50)
+treespecs$np <- rep(factor(treespecs$np[1:length(descriptions)]), times=50)
+treespecs$type <- rep(factor(treespecs$type[1:length(descriptions)]), times=50)
+treespecs$corUbi <- rep(factor(treespecs$corUbi[1:length(descriptions)]), times=50)
+treespecs$numbclus <- rep(factor(treespecs$numbclus[1:length(descriptions)]), times=50)
+treespecs$sigmabi <- rep(factor(treespecs$sigmabi[1:length(descriptions)]), times=50)
+
+treespecs.long <- data.frame(treesize=treespecs["GLMMtreesize"], N=treespecs$N, rho=treespecs$rho, 
+                             np=treespecs$np, type=treespecs$type, corUb=treespecs$corUbi, 
+                             numbclus=treespecs$numbclus, sigmab=treespecs$sigmabi, datasetID=
+                               factor(1:nrow(treespecs)))
[…]

### 2016-09-02 pkg/glmertree/paper/Simulation replication/Linear to piecewise interactions/generate test data.R

@@ -1,0 +1,165 @@
+N <- list() # different levels for N
+N[[1]] <- 200
+N[[2]] <- 500
+N[[3]]<- 1000
+
+bi <- list() # different numbers and values for the study intercepts
+bi$numbclus <- c(25)
+bi$sigmas <- c(2.5, 7.5)
+
+rho <- list() # levels of intercorrelations between the covariates
+rho[[1]] <- matrix(0, nrow=20, ncol=20)
+rho[[2]] <- matrix(.3, nrow=20, ncol=20)
+diag(rho[[1]]) <- 1
+diag(rho[[2]]) <- 1
+
+corUbi <- list()
+corUbi[[1]] <- "uncorrelated"
+corUbi[[2]] <- "bi and U correlated"
+
+pXj <- list() # levels of number of covariates
+pXj[[1]] <- 5; pXj[[2]] <- 10 
+
+types <- list("linear","both","piecewise") # different levels for the difference in treatment outcome
+
+
+testdata <- list()
+testdescriptions <- list()
+
+set.seed(38916032)
+
+for (c in 1:50) {  
+  counter <- 0
+  for (d in 1:length(bi$sigmas)) {# d is counter for sigma_b value
+    for (e in 1:length(bi$numbclus)) {# e is counter for number of clusters
+      numbclus <- bi$numbclus[[e]] # numbclus is number of different random intercept values
+      for (f in 1:length(corUbi)) { # f is counter for correlation between U and b
+        for (g in 1:length(types)) { # g is counter for type
+          type <- types[[g]]
+          for (h in 1:length(pXj)) { # h is counter for number of covariates
+            p <- pXj[[h]]
+            for (i in 1:length(N)) { # i is counter for sample size N
+              nobs <- N[[i]]
+              for (j in 1:length(rho)) { # j is counter for intercorrelations between covariates
+                L <- chol(rho[[j]]) # Cholesky decomposition of correlation matrices
+                nvars <- dim(rho[[j]])[1]
+                
+                # Random variables that follow an M correlation matrix
+                r <- t(L) %*% matrix(rnorm(nvars*nobs, sd=10), nrow=nvars, ncol=nobs)
+                r <- t(r)
+                
+                # create centered covariate datasets
+                r <- as.data.frame(r)
+                names(r) <- paste("X", 1:15, sep="")
+                
+                # add population means to covariates
[…]

### 2016-09-02 pkg/glmertree/paper/Simulation replication/Linear to piecewise interactions/generate training data.R

@@ -1,40 +1,78 @@
 N <- list() # different levels for N
-N[[1]] <- 200; N[[2]] <- 500; N[[3]] <- 1000 
+N[[1]] <- 200
+N[[2]] <- 500
+N[[3]]<- 1000
 
 bi <- list() # different numbers and values for the study intercepts
-bi$numbclus <- c(5,10,25)
-bi$sigmas <- c(0, 5, 10)
+bi$numbclus <- c(25)
+bi$sigmas <- c(2.5, 7.5)
 
 rho <- list() # levels of intercorrelations between the covariates
 rho[[1]] <- matrix(0, nrow=20, ncol=20)
-diag(rho[[1]]) <- 1
 rho[[2]] <- matrix(.3, nrow=20, ncol=20)
+diag(rho[[1]]) <- 1
 diag(rho[[2]]) <- 1
 
 corUbi <- list()
 corUbi[[1]] <- "uncorrelated"
-corUbi[[2]] <- "bi and splitting U correlated"
-corUbi[[3]] <- "bi and non-splitting U correlated" 
+corUbi[[2]] <- "bi and U correlated"
 
 pXj <- list() # levels of number of covariates
-pXj[[1]] <- 5; pXj[[2]] <- 15 
+pXj[[1]] <- 5; pXj[[2]] <- 10 
 
-muYdiff <- list() # different levels for the difference in treatment outcome
-muYdiff[[1]] <- 2.5; muYdiff[[2]] <- 5
+types <- list("linear","both","piecewise") # different levels for the difference in treatment outcome
+
+
+### Get the piecewise nodemeans:
+set.seed(1)
+X1 <- rnorm(10e6, mean = 10, sd = 10)
+X2 <- rnorm(10e6, mean = 30, sd = 10)
+X5 <- rnorm(10e6, mean = 70, sd = 10)
+Treat <- rbinom(10e6, 1, .5)
+node3 = X2<=30 & X1<=17
+node4 = X2<=30 & X1>17
+node6 = X2>30 & X5<=63
+node7 = X2>30 & X5>63
+fact1 <- rep(NA, times = 10e6)
+fact1[node3 & Treat] <- "3T2"
+fact1[node4 & Treat] <- "4T2"
+fact1[node6 & Treat] <- "6T2"
+fact1[node7 & Treat] <- "7T2"
+fact1[node3 & !Treat] <- "3T1"
[…]

### 2016-09-05 pkg/glmertree/paper/treatment_interactions.Rnw
Small touch-ups in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@963 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine.
-Tree-based algorithms are helpful tools for the  detection of such interactions, but datasets suited for their detection may often have a clustered or multilevel structure.
-For example, observations in a dataset may be nested within research centers or trials.
-Therefore, we propose generalized linear mixed-effects model trees (GLMM tree), allowing for the detection of treatment-subgroup interactions, as well as taking into account the nested structure of a dataset.
+Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure.
+Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, as well accounting for the clustered structure of a dataset.
 The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters.
-In a simulation study, we evaluate the performance of GLMM tree and compare it with thattrees without random effects, and GLMMs with pre-specified interactions.
+In a simulation study, we evaluate the performance of GLMM tree and compare it with that of trees without random effects, and GLMMs with pre-specified interactions.
 GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of datasets, and trees without random effects in 61\% of datasets.
 GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4 and 33\%, respectively).
-Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal accuracy in predicting treatment-effect differences, but better accuracy in detecting interactions.
+Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on avergae), but better accuracy in detecting interactions.
 We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression.
@@
 Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments.
-Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
-Several tree-based algorithms to the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview).
-Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, which allows users to select from a range of statistical methods, including trees.
-As such, the \citeA{ZhanyTsia12Stat} method can also be referred to as a tree-based method.
+Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
+As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions.
+Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
+Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.
 
-Often, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model.
+In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model.
 For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}.
@@
 In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}.
-In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}.
-However, none of the purely tree-based methods for treatment-subgroup interaction detection, also take into account the clustered structure of a dataset.
+For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}.
+However, none of the purely tree-based methods for treatment-subgroup interaction detection take into account the clustered structure of a dataset.
 Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.
@@
 
-In what follows, we introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree.
+In what follows, we first introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree.
 In the Empirical Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects.
@@
 where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function.
-$\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment group.
-Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper; that is, two treatment conditions and no additional covariates are included in the GLM.
-Furthermore, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.
+$\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.
+Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper.
+Also, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.
 However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.
@@
 
-The MOB algorithm detects such moderation, by testing for parameter stability over a set of auxiliary covariates or partitioning variables.
+The MOB algorithm can be used to detect such moderation, by testing for parameter stability over a set of auxiliary covariates or partitioning variables.
 When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor.
@@
 More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function.
-By definition, the empirical scores of all observations in a dataset sum to zero.
-When the model is correctly specified, the expected value of the score for each observation is also zero.
-Therefore, under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}.
+When the model is correctly specified, the expected value of the score for each observation equals zero.
+Therefore, under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}.
 To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.
@@
 Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree.
-If the partitioning is based on a GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node.
-As a result, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
+If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
 %
@@
 By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
-Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above.
-However, the two subgroups in the middle do not represent true subgroups, and this may be due to the clustered structure of the dataset not being taken into account.
+Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above.
+However, the two subgroups in the middle do not represent true subgroups, which may be due to the clustered structure of the dataset not being taken into account.
 
@@
 Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
-Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$.
-Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
+Further, $b$ is a random vector of length $M$, each $m$th element representing the random intercept for cluster $m$; again, for simplicity we assume that only cluster-specific intercepts are included in the model, but multiple random-effects coefficients can easily be included.
+Furthermore, within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
 The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).
-
-For simplicity, we assume that only cluster-specific intercepts are included in the model in the current paper. However, multiple random-effects coefficients can easily be included.
 
@@
 
-Although GLMMs are adequately suited for taking into account the nested structure of a datasets, the global fixed-effects model in Equation \ref{eq:mixedeffects} may not describe the data well.
+Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part may not describe the data well.
 Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} and random effects are incorporated as well:
@@
 For other response distributions, the \verb|glmtree| function would be available.
-For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
+For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
 
@@
 
-\subsection{Treatment-subgroup interaction design} 
+\subsection{Simulation design}
+
+\subsubsection{Treatment-subgroup interactions} 
 For generating datasets with treatment-subgroup interactions, we used one of the treatment-subgroup interaction designs from \citeA{DussyMech14}, which is depicted in Figure~\ref{fig:modelC}.
@@
 \item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
-\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
+\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
 \item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition.
@@
 
-\subsection{Variable distributions}
+\subsubsection{Variable distributions}
 Following the approach of \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$.
@@
 
-\subsection{Assessment of performance}
+\subsubsection{Assessment of performance}
 
@@
 
-Figure~\ref{fig:xyplot_treesize_interact} indicates that average tree size increases slightly with sample size, which may be due to increased power.
-In the absence of random effects (i.e., $\sigma_{b}=0$), both GLM and GLMM trees have about 7 nodes on average.
+Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions.
+On average, tree size increases slightly with sample size for GLMM tree, and dramatically for GLM tree.
+In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size.
 However, clear differences in tree size were observed when $\sigma_b > 0$.
 When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average.
-When random intercepts are not correlated with a partitioning covariate and sample size is large (i.e., 500 or 1000), GLM and GLMM trees are about equally sized and have approximately the true tree size, on average.
-When random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10).
+When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized.
+However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially wit larger sample sizes (i.e., 500 or 1000) and when $\sigma_b$ values (i.e., 10).
 This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.
@@
 The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).
-GLM tree did show somewhat higher variability in recovering the splitting value for the first split than GLMM tree ($\mathrm{SD}=0.154$ and $\mathrm{SD}=0.127$, respectively).
 
-GLMM tree accurately recovered the partition in 90.40\% of datasets, whereas GLM tree accurately recovered the partition in only 61.44\% of datasets.
+However, further splits were more accurately recovered by GLMM tree, which accurately recovered the partition in 90.40\% of datasets, and GLM tree in only 61.44\% of datasets.
 To find the most important predictors of tree accuracy, we fitted a GLM with algorithm type and the parameters of the data-generating process as independent variables.
@@
 This effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable.
-When random intercepts were not correlated with one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., $N = 200$).
+However, when random intercepts were not correlated with one of the $U_k$ variables, GLMM tree outperformed GLM tree only when sample size was small (i.e., $N = 200$).
 
@@
 \subsubsection{Predictive accuracy of trees}
-To assess predictive accuracy, correlations between the true and predicted treatment-effect differences of both algorithms were calculated in test datasets.
 The predicted treatment-effect differences of GLMM tree were closer to the true differences than those of GLM tree, with a mean correlation of .93 ($\mathrm{SD}=0.12$) for GLMM tree and .88 ($\mathrm{SD}=0.19$) for GLM tree.
-An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the treatment difference effect size.
+An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the treatment-difference effect size.
 These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_correlations}).
@@
 In the absence of random effects (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were about equally accurate.
-However, it should be noted that in the absence of random effects, GLMM tree tends to outperform GLM tree in smaller sample sizes (i.e., 200), whereas GLM tree tends to outperform GLMM tree in larger sample sizes (i.e., 500 and 1000).
-In the presence of random effects, GLM tree predictions were much less accurate than those of GLMM tree when treatment-effect differences were smaller (i.e., Cohen's $d = .5$).
+In the presence of random effects, GLM tree predictions were clearly less accurate than those of GLMM tree when treatment-effect differences were smaller (i.e., Cohen's $d = .5$).
 This effect was also observed, but much less pronounced, when treatment-effect differences were larger (i.e., Cohen's $d = 1.0$).
-In addition, with increasing sample size and treatment-effect differences, the difference in predictive accuracy between GLM and GLMM tree becomes smaller.
+In other words, GLMM tree outperforms GLM tree in the presence of random effects, especially when sample size and treatment-effect differences are smaller.
 
@@
 For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there is only a main effect of treatment in the population.
-Put differently, the number of subgroups or terminal nodes in these datasets was $J=1$, and there was only a single value of $\beta_j = \beta$ in every dataset.
-For the datasets without treatment-subgroup interactions, we varied the same parameters of the data-generating process as in Study I.
+Put differently, there was only a single global value of $\beta_j = \beta$ in every dataset.
+For these datasets, we varied the same parameters of the data-generating process as in Study I.
 However, the sixth facet of the data-generating process had only two levels for these datasets, as there were no `true' partitioning variables in these datasets.
@@
 \subsubsection{Tree size and Type-I error}
-In datasets without treatment-subgroup interactions, the average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree.
+In datasets without treatment-subgroup interactions, average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree.
 The overall Type-I error rate was only .04 for GLMM tree, and .33 for GLM tree.
@@
 To generate purely piecewise interactions, we used the partition depicted in Figure~\ref{fig:modelC}.
-We estimated $\bar{y}$ for each of the treatment-subgroups by calculating subgroup means in a large dataset (N = $10^6$) with purely linear interactions.
-For each observation in datasets with purely piecewise interactions, the treatment-subgroup specific value for $\bar{y}$ was imputed.
+We calculated treatment-subgroup specific means $\hat{\mu}_{y}$ in a large dataset (N = $10^6$) with purely linear interactions.
+For each observation in datasets with purely piecewise interactions, this treatment-subgroup specific $\hat{\mu}_{y}$ value was imputed.
 To generate combined linear and piecewise interactions, $y$ values according to the purely piecewise and linear models were summed and divided by 2.
@@
 
-Both algorithms showed equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees.
-In terms of bias, GLMM treatment-effect difference predictions deviated more from the true values than the predictions of GLMM tree.
-The mean true treatment-effect difference averaged over all datasets was .04 (SD = 1.22), whereas the mean predicted treatment-effect differences averaged over all datasets for GLMMs was -.26 (SD = 100.33) and for GLMM trees .03 (SD = 3.88).
-Furthermore, it should be noted that the full models were used for prediction with GLMMs and GLMM trees.
-That is, predictions for GLMMs in datasets with 10 covariates may involve 11 (significant or non-significant) main effects, 100 first-order interactions and 100 second-order interactions.
-As the maximum tree depth of GLMM trees was set to 4, GLMM trees used a maximum of only 8 terminal nodes for predictions.
-
-The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (linear, piecewise or both).
-These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.
-As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely linear interactions.
-The predictive accuracy of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs decreases when the number of covariates increases.
-Finally, for both algorithms, predictive accuracy improves somewhat when sample size increases.
-
 \begin{figure}[t!]
@@
 \end{figure}
+
+Both algorithms showed equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees.
+In terms of bias, GLMM treatment-effect difference predictions deviated more from the true values than the predictions of GLMM tree.
+The mean true treatment-effect difference averaged over all datasets was .04 (SD = 1.22), whereas the mean predicted treatment-effect differences averaged over all datasets for GLMMs were -.26 (SD = 100.33) and for GLMM trees .03 (SD = 3.88).
+Furthermore, it should be noted that the models used for prediction with GLMM trees were more complex than the GLMMs used for prediction.
+Predictions for GLMMs in datasets with 10 covariates involved 11 (significant or non-significant) main effects, 100 first-order interactions and 100 second-order interactions.
+As the maximum tree depth of GLMM trees was set to 4, GLMM trees used a maximum of only 8 terminal nodes for predictions.
+
+The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (linear, piecewise or both).
+These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.
+As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely linear interactions.
+Furthermore, performance of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases.
+This indicates that GLMM tree should be preferred in more exploratory situations, where the number of potential moderator variables is large, as GLMMs may do very poorly in terms of predictive accuracy, especially when interactions are piecewise.
+On the other hand, GLMM tree may perform poorly when in small sample sizes (i.e., $N=200$), when interactions are purely linear.
+
+\FloatBarrier
 
@@
 
-To further illustrate and compare the application of GLM and GLMM tree, we applied both algorithms to a dataset from a individual-patient data meta-analysis of \citeA{CuijyWeit14}.
+To further illustrate and the use of GLM and GLMM tree, we applied both algorithms to a dataset from a individual-patient data meta-analysis of \citeA{CuijyWeit14}.
 This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
@@
 plot(lmtree_app)
+@
+<<depression-glmtree-effs, fig=TRUE, height=6, width=8>>=
 plot(GLMtree_app_eff, tp_args = list(FUN = identity, height = 10, width = 14), tnex = 1.5)
@@
 
-The tree and effects sizes resulting from application of GLM tree are presented in Figures~\ref{fig:lmtree_C&W} and \ref{fig:lmtree_C&W_eff} and those resulting from application of GLMM tree are presented in Figures~\ref{fig:lmertree_C&W} and \ref{fig:lmertree_C&W_eff}.
+The tree and effects sizes resulting from application of GLM tree are presented in Figure~\ref{fig:lmtree_C&W}.
+Those resulting from application of GLMM tree are presented in Figure~\ref{fig:lmertree_C&W}.
 
@@
 The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
-The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Figure~\ref{fig:lmertree_C&W_eff}).
-For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Figure~\ref{fig:lmertree_C&W_eff}).
+The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Figure~\ref{fig:lmertree_C&W}).
+For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Figure~\ref{fig:lmertree_C&W}).
 The estimated intraclass correlation coefficient for the GLMM tree was .05.
@@
 
-\begin{figure}[t!]
+\begin{figure}
 \setkeys{Gin}{width=0.5\textwidth}
@@
 @
-%\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).}
-\caption{Upper panel: GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).
-Lower panel: Subgroup-specific descriptive statistics.}
-\label{fig:lmertree_C&W}
-\end{figure}
-
-\begin{figure}[t!]
-\setkeys{Gin}{width=0.4\textwidth}
-<<depression-glmmtree-eff, fig=TRUE, height=3, width=4>>=
+<<depression-glmmtree-effs, fig=TRUE, height=4, width=4.7>>=
 plot(GLMMtree_app_eff, tp_args = list(FUN = identity, height = 10, width = 14), tnex = 1.5)
 @
-\caption{Statistics and effect sizes for GLM tree in Figure \ref{fig:lmertree_C&W}.}
-\label{fig:lmertree_C&W_eff}
+\caption{Left panel: GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
+The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).
+Right panel: Subgroup-specific descriptive statistics.}
+\label{fig:lmertree_C&W}
 \end{figure}
@@
 		\hline
-		Education               &	0.956	&	0.014		\\
-		ComorbidAnxietyDisorder	&	0.398	&	0.528		\\
-		HRSDt0                  &	0.034	&	0.002		\\
-		Age                     &	0.000	&	0.022		\\
-		Gender                  &	0.002	&	0.004		\\
+		Education               &	.956	&	.014		\\
+		ComorbidAnxietyDisorder	&	.398	&	.528		\\
+		HRSDt0                  &	.034	&	.002		\\
+		Age                     &	.000	&	.022		\\
+		Gender                  &	.002	&	.004		\\
 		\hline
 		\multicolumn{3}{l}{\textit{Note.} Frequencies are calculated over 500 random}\\
-		\multicolumn{3}{l}{subsamples of size $.9 \times N$ of the complete dataset.}\\
-		\multicolumn{3}{l}{Frequencies do not add up to 1, as trees may involve}\\
-		\multicolumn{3}{l}{multiple, or no splits.}\\
+		\multicolumn{3}{l}{subsamples of the complete dataset. Frequencies}\\
+		\multicolumn{3}{l}{do not add up to 1, as trees may involve multiple}\\
+		\multicolumn{3}{l}{or no splits.}\\
 	\end{tabular}
@@
 \end{table}
+
+\FloatBarrier
 
@@
 With 10 potential moderator variables, $t$-tests of GLMMs were unable to identify the true moderator variables, whereas the number of potential moderator variables did not influence performance of GLMM tree.
+Therefore, our results indicate that GLMM trees are better suited than GLMMs for exploratory analyses, in which moderator variables need to be selected from a larger number of covariates.
 Furthermore, the number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex predictive models.
@@
 The resulting trees are potentially unstable, and stability of the results should be assessed, preferably in a dataset not used for training, or by multi-fold cross-validation or resampling techniques.
+In the Application, we have shown how the stability of splitting variable selection can be assessed using resampling techniques.
 

### 2016-09-05 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@964 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Furthermore, performance of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases.
-This indicates that GLMM tree should be preferred in more exploratory situations, where the number of potential moderator variables is large, as GLMMs may do very poorly in terms of predictive accuracy, especially when interactions are piecewise.
-On the other hand, GLMM tree may perform poorly when in small sample sizes (i.e., $N=200$), when interactions are purely linear.
+This indicates that GLMM tree may be especially useful for exploratory purposes.
 

### 2016-09-15 pkg/glmertree/paper/treatment_interactions.Rnw
Updated paper based on Niels comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@980 255f4249-2263-4417-b1ea-8498779e27a7

@@
 As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions.
-Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
+Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
 Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.
@@
 For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}.
-However, none of the purely tree-based methods for treatment-subgroup interaction detection take into account the clustered structure of a dataset.
+However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset.
 Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.
@@
 For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}.
-GLMM tree allows for taking into account the clustered structure of datasets \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.
+In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14), GLMM tree allows for taking into account the clustered structure of datasets.
+In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.
 
 In what follows, we first introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree.
-In the Empirical Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects.
+In the Simulation-based Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects.
+The comparison with GLM-based MOB allows for assessing the extent to which including random-effects estimation improves the performance of MOB trees.
+The comparison with GLMMs with pre-specified interactions will allow for assessing the extent to which using a tree-based method will improve detection of treatment-subgroup interactions.
 In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression.
@@
 
-Participants were part of one of ten clusters, each with a different value for the intercept.
-Data was generated such that covariates and cluster-specific intercepts were uncorrelated.
-Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.
+Participants were part of one of ten clusters, each with a different value for the random intercept.
+Data were generated such that covariates and cluster-specific intercepts were uncorrelated.
+43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.
 
@@
 This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different.
-However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parameters of more complex models across terminal nodes.
+However, CART trees detect differences in the mean (or median) value across terminal nodes, whereas MOB trees detect differences in parameters of more complex models across terminal nodes.
 
-For example, let us take a global GLM to estimate the overall treatment effect in the artificial motivating dataset.
+For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset.
 The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
@@
 
-The result of fitting a global GLM to the artificial motivating dataset is depicted in Figure~\ref{fig:fixedeffects}; the boxplots show the distribution of the posttreatment depression scores in both treatment groups.
+The result of fitting a global GLM to the motivating dataset is depicted in Figure~\ref{fig:fixedeffects}; the boxplots show the distribution of the posttreatment depression scores in both treatment groups.
 The global model does not describe the data well: there is substantial residual variance and the slope of the regression line is nearly zero.
@@
 
-The MOB algorithm can be used to detect such moderation, by testing for parameter stability over a set of auxiliary covariates or partitioning variables.
+The MOB algorithm can be used to detect such moderation, by testing for parameter stability over a set of auxiliary covariates, also known as \textit{partitioning variables}.
 When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor.
@@
 If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3).
-The binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2).
+This partition is created using $U_{k*}$, the variable with the minimal p-value in step (2).
 The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}.
@@
 %
-Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all coefficients interact with the subgroup indicator.
+Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all $x$-variables interact with the subgroup indicator.
 Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
-$\beta^{*}$ would have length $2J$, and contain the subgroup-specific fixed-effects coefficients.
+$\beta^{*}$ would have length $2J$ as well, containing subgroup-specific fixed-effects coefficients.
 
@@
 
-Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the artificial motivating dataset.
+Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the motivating dataset.
 By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
@@
 
-For estimating GLMM trees, we take this approach a step further: we take a GLM tree, instead of CART tree with constant fits to estimate the fixed-effects part of the GLMM.
+For estimating GLMM trees, we take this approach a step further: instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree.
 This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes.
@@
 
-In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the artificial motivating dataset is presented.
+In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented.
 As can be seen, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.
@@
 
-\section{Empirical Evaluation: Method}
+\section{Simulation-based Evaluation: Method}
 
-To assess and compare the performance of GLMM tree, we performed three simulation studies.
+To assess the performance of GLMM tree, we performed three simulation studies.
 In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions.
@@
 
-R \cite{R14} was used for generation and analysis of all datasets.
+R \cite{R14} was used for the generation and analysis of all datasets.
 The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions.
@@
 
+\subsubsection{Continuous interactions}
+
+The subgroups depicted in Figure~\ref{fig:modelC} are piecewise functions of the partitioning variables, therefore these interactions can be referred to as piecewise interactions.
+However, interactions may also consist of continuous functions of the partitioning variables.
+For comparing the accuracy of GLMM tree with that of GLMMs with pre-specified interaction terms, we generated datasets in which interactions were varied to be purely piecewise, purely continuous, or a combination of both.
+To generate purely continuous interactions, the outcome variable $y$ was generated as the sum of:
+
+\begin{enumerate}
+	\item a main effect of $U_2$ (coefficient = 1.5) 
+	\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = -.25), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = .25)
+	\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = -.25), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = .25)
+\end{enumerate}
+
+Note that the direction (positive or negative) and type (main or interaction) of these effects correspond to those in Figure~\ref{fig:modelC}.
+All interactions were created by using centered $U_k$ variables, calculated by subtracting the variable means.
+
 \subsubsection{Data-generating parameters} 
@@
 Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) in each dataset.
-Consequently, predictive accuracy was only be assessed in datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.
+Consequently, predictive accuracy was only assessed in datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.
 To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets.
-Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from 'new' clusters.
+Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from `new' clusters.
 
-\section{Empirical Evaluation: Results}
+\section{Simulation-based Evaluation: Results}
 
-\subsection{Study I: Piecewise interactions}
+\subsection{Study I: Detection of Treatment-Subgroup Interactions}
 
-For each cell of the data-generating design described above, 50 datasets were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets with piecewise treatment-subgroup interactions.
+For each cell of the data-generating design described above, 50 datasets were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets with treatment-subgroup interactions.
 In these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).
@@
 When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized.
-However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially wit larger sample sizes (i.e., 500 or 1000) and when $\sigma_b$ values (i.e., 10).
+However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially with larger sample sizes (i.e., 500 or 1000) and larger $\sigma_b$ values (i.e., 10).
 This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.
@@
 \subsubsection{Design}
-For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there is only a main effect of treatment in the population.
+In addition to the accuracy of treatment-subgroup interaction detection, we assessed the Type-I error: the extent to which datasets without treatment-subgroup interactions may erroneously be partitioned.
+For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there was only a main effect of treatment in the population.
 Put differently, there was only a single global value of $\beta_j = \beta$ in every dataset.
@@
 
-\subsection{Study III: Linear and piecewise interactions}
+\subsection{Study III: Piecewise vs. Continuous Interactions}
 
 \subsubsection{Design}
-To compare the performance of GLMM tree with that of GLMMs with pre-specified interaction effects, we created datasets in which interactions were varied from purely linear to purely piecewise.
-The number of cells in the simulation design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters).
+
+To compare the performance of GLMM tree with that of GLMMs with pre-specified interaction effects, we created datasets in which interactions were varied from purely continuous to purely piecewise.
+For these datasets, the seventh facet of the interaction design (effect size of treatment differences) was replaced by a factor with three levels, controlling the type of interactions: purely continuous interactions, purely piecewise interactions, and a combination of both.
+Purely continuous interactions were generated as described above, in the Method section.
+To allow for a smooth transition from purely continuous to purely continuous interactions, we aimed to keep the proportion of variance explained by each of the covariates equal across piecewise, continuous and combined interactions.
+Therefore, for piecewise interactions, we used the partition depicted in Figure 4, but not the coefficients.
+The coefficients used for data-generation were replaced by the subgroup-specific means, given purely continuous interactions.
+To generate the combined conitnuous and piecewise interactions, $y$ values according to the purely piecewise and continuous interaction models were summed and divided by 2.
+
+Furthermore, in this simulation study, the number of cells in the simulation design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters), as the earlier simulations indicated no effects of the number of clusters.
 The fifth facet of the data-generating design was limited to two levels ($\sigma_b = 2.5$ and $\sigma_b = 7.5$).
-As in Study II, we limited the sixth facet to two levels (the random intercept was either correlated to one of the partitioning variables, or not).
-To counter unidentifiability due to potentially large numbers of pre-specified interaction terms in the GLMMs, we limited the number of covariates to $k=5$ and $k=10$ (instead of $k=5$ and $k=15$).
-Finally, to allow for a smooth transition from linear to piecewise interactions, the seventh facet of the interaction design (effect size of treatment differences) was replaced by a factor with three levels, controlling the type of interactions: purely linear interactions, purely piecewise interactions, and a combination of both.
+As in Study II, we limited the sixth facet to two levels (the random intercept was either correlated to one of the covariates, or not).
+To counter unidentifiability due to potentially large numbers of pre-specified interaction terms in the GLMMs, the second level of the second facet (number of potential partitioning covariates) was set to $k=10$, instead of the value of $k=15$ used in the earlier simulations.
 As a result, $50 \times 3 \times 2 \times 2 \times 1 \times 2 \times 2 \times 3$ = 7,200 training datasets were generated for this study.
-
-All input variables were generated as described above. To generate purely linear interactions, outcome variable $y$ was generated as the sum of:
-
-\begin{enumerate}
-\item a main effect of $U_2$ (coefficient = 1.5) 
-\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = -.25), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = .25)
-\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = -.25), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = .25)
-\end{enumerate}
-
-Note that the direction (positive or negative) and type (main or interaction) of these effects correspond to those in Figure~\ref{fig:modelC}.
-All interactions were created by using centered $U_k$ variables, calculated by subtracting the variable means.
-To generate purely piecewise interactions, we used the partition depicted in Figure~\ref{fig:modelC}.
-We calculated treatment-subgroup specific means $\hat{\mu}_{y}$ in a large dataset (N = $10^6$) with purely linear interactions.
-For each observation in datasets with purely piecewise interactions, this treatment-subgroup specific $\hat{\mu}_{y}$ value was imputed.
-To generate combined linear and piecewise interactions, $y$ values according to the purely piecewise and linear models were summed and divided by 2.
 
@@
 treespecs.long$type <- factor(treespecs.long$type, ordered = TRUE, levels = 
-c("linear", "both", "piecewise"))
+c("continuous", "both", "piecewise"))
 @
@@
 Overall, GLMM tree accurately recovered main and interaction effects in 53.58\% of datasets, wheareas GLMMs accurately recovered main and interaction effects in 34.81\% of datasets.
-To find the most important predictors of model accuracy, we performed an ANOVA, in which the algorithm type, sample size, the number of covariates and type of interaction (linear, piecewise or both) were found to be most important.
+To find the most important predictors of model accuracy, we performed an ANOVA, in which the algorithm type, sample size, the number of covariates and type of interaction (continuous, piecewise or both) were found to be most important.
 These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_accuracy_pw}.
 
-Figure~\ref{fig:xyplot_accuracy_pw} indicates a clear effect of sample size: when the number of covariates is large (i.e., 10) GLMMs do not recover all true effects in the dataset, whereas they always seem to recover the true effects when the number of covariates is small (i.e., 5), and the effects are partly linear.
-However, when the effects are purely piecewise, GLMMs never recover the true effects in the dataset, unless sample size is very large (i.e., $N = 1000$).
-At the same time, Figure~\ref{fig:xyplot_accuracy_pw} indicates that the accuracy of GLMM trees is unaffected by the number of covariates.
-Furthermore, GLMM tree performs best in recovering purely piecewise interactions, and poorest in recovering purely linear interactions.
-Finally, the performance of GLMM tree is affected by sample size: generally, the true effects are more often recovered in datasets with larger sample sizes.
+Figure~\ref{fig:xyplot_accuracy_pw} shows a clear effect of interaction type: GLMM tree performs best in datasets with purely piecewise interactions, and poorest in datasets with purely continuous interactions.
+GLMMs perform best in datasets with purely continuous interactions, and poorest in datasets with purely piecewise interactions.
+Figure~\ref{fig:xyplot_accuracy_pw} also shows a clear effect of the number of covariates: with a large number of covariates, GLMMs never recover all true effects in the dataset, whereas they always recover the true effects when the number of covariates is small and interaction effects are at least partly continuous.
+At the same time, the accuracy of GLMM tree is unaffected by the number of covariates.
+Finally, the performance of both GLMM tree and GLMMs is affected by sample size: the effects are more often recovered in datasets with larger sample sizes.
 
@@
 
-Both algorithms showed equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees.
+Both algorithms showed about equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees.
 In terms of bias, GLMM treatment-effect difference predictions deviated more from the true values than the predictions of GLMM tree.
 The mean true treatment-effect difference averaged over all datasets was .04 (SD = 1.22), whereas the mean predicted treatment-effect differences averaged over all datasets for GLMMs were -.26 (SD = 100.33) and for GLMM trees .03 (SD = 3.88).
-Furthermore, it should be noted that the models used for prediction with GLMM trees were more complex than the GLMMs used for prediction.
+Furthermore, it should be noted that the predictive models for GLMM trees were simpler than those of the GLMMs.
 Predictions for GLMMs in datasets with 10 covariates involved 11 (significant or non-significant) main effects, 100 first-order interactions and 100 second-order interactions.
-As the maximum tree depth of GLMM trees was set to 4, GLMM trees used a maximum of only 8 terminal nodes for predictions.
+Predictions for GLMM tree involved a maximum  of only 8 terminal nodes, as their maximum depth was restricted.
 
-The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (linear, piecewise or both).
+The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (continuous, piecewise or both).
 These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.
-As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely linear interactions.
+As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely continuous interactions.
 Furthermore, performance of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases.
@@
 
-To further illustrate and the use of GLM and GLMM tree, we applied both algorithms to a dataset from a individual-patient data meta-analysis of \citeA{CuijyWeit14}.
+To further illustrate the use of GLM and GLMM tree, we applied both algorithms to a dataset from a individual-patient data meta-analysis of \citeA{CuijyWeit14}.
 This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
@@
 
-The results of recursive partitioning techniques are known to be potentially unstable, in the sense that samll changes in the dataset may substantially alter the variables or values selected for partitioning.
+The results of recursive partitioning techniques are known to be potentially unstable, in the sense that small changes in the dataset may substantially alter the variables or values selected for partitioning.
 Therefore, we used the R-package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values.
@@
 The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
-By taking into account study-specific intercepts, the GLMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split made by GLM tree is a spurious one.
+By taking into account study-specific intercepts, the GLMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split made by GLM tree may be a spurious one.
 The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
@@
 As such, we hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies.
+We performed simulation studies to assess the accuracy of GLMM tree in detecting treatment-subgroup interactions and illustrated the application of GLMM tree on an existing dataset.
 
@@
 GLMM tree also provided better predictive accuracy than GLM tree, with an average correlation between true and predicted treatment differences of .94 for GLMM tree, and .88 for GLM tree.
-GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, treatment-effect differences were relatively small (i.e., Cohen's $d=.5$), and/or sample size was small (i.e., $ N < 1,000$).
+GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, differences in treatment effects across terminal nodes were relatively small (i.e., Cohen's $d=.5$), and/or sample size was small (i.e., $ N < 1,000$).
 Such treatment-effect differences and sample sizes may be quite common, even in multi-center clinical trials, and GLMM tree may provide a helpful tool for subgroup detection in those instances.
@@
 Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided similar predictive accuracy, on average.
-When interactions were purely linear, GLMMs outperformed GLMM trees, and when interactions were purely piecewise, GLMM trees outperformed GLMMs.
+When interactions were purely continuous, GLMMs outperformed GLMM trees, and when interactions were purely piecewise, GLMM trees outperformed GLMMs.
 However, GLMM tree may have a clear advantage when there are a large number of potential moderator variables (i.e., $> 5$).
@@
 
-In the Application, we found GLMM tree to provide results that are easily interpretable, and also more accurate than a GLM tree without random effects.
-In addition, node-specific means and variances can be used to calculate effect sizes, to judge clinical relevance of the findings, as would often be done in RCTs or meta-analysis.
+These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures.
+However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified.
+That is, if there are subgroups with respect to the variables specified as potential partitioning variables and these subgroups have different values for the parameters of the GLM, then GLMM tree will accurately recover those subgroups.
+However, misspecification of the model will negatively affect performance.
+The most important source of misspecification would be the omission of relevant variables, either in the GLM or as partitioning variables.
+When relevant variables are omitted, GLMM tree can only approximate the true subgroups using the specified variables.
+Another source of misspecification would be the inclusion of irrelevant variables.
+Our simulations indicate that the performance of GLMM tree was not negatively affected by the number of potential moderator variables specified, but the power to detect subgroups may still be reduced with larger numbers of potential partitioning variables.
+Including irrelevant variables in the GLM may also negatively affect performance, although we have not assessed this in our simulations.
+
+In the Application, we found GLMM tree to provide results that were more easily interpretable, and also more accurate than a GLM tree without random effects.
+In addition, to judge clinical relevance of the findings, we calculated node-specific effect sizes, as would often be done in RCTs or meta-analysis.
 Although we have limited ourselves to calculating Cohen's $d$ in the current paper, equivalent values of the success rate difference or the number needed to treat can be calculated, but this would involve additional distributional assumptions \cite{KraeyKupf06}.
@@
 A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significant in the tree-growing process.
-
-These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures.
-However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified model.
-That is, if there are subgroups with respect to the partitioning variables specified, so that there are different parameters of the GLM in each of these subgroups, then GLMM tree will accurately recover those subgroups.
-However, as with any data-analytic method, misspecification of the model will negatively affect performance.
-One source of misspecification would be the omission of relevant variables in the GLM, or as potential partitioning variables.
-Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups.
-However, the performance of GLMM tree, in contrast to that of GLMMs with pre-specified interactions, was not negatively affected by the number of potential moderator variables.
 

### 2016-10-06 pkg/glmertree/paper/treatment_interactions.Rnw
some small improvements in pages 1-6
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@996 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure.
-Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, as well accounting for the clustered structure of a dataset.
+Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset.
 The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters.
-In a simulation study, we evaluate the performance of GLMM tree and compare it with that of trees without random effects, and GLMMs with pre-specified interactions.
-GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of datasets, and trees without random effects in 61\% of datasets.
-GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4 and 33\%, respectively).
-Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on avergae), but better accuracy in detecting interactions.
-We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression.
+In a simulation study, we evaluate the performance of GLMM trees and compare it with that of trees without random effects, and GLMMs with pre-specified interactions.
+GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets.
+GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4\% and 33\%, respectively).
+Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on average), but better accuracy in detecting interactions.
+We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression.
 We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
@@
 As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions.
-Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
+Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview).
 Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.
@@
 However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset.
-Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.
+Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM trees.
 
@@
 For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}.
-In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14), GLMM tree allows for taking into account the clustered structure of datasets.
+In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}), GLMM tree allows for taking into account the clustered structure of datasets.
 In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.
@@
 
-We created generated a dataset representing observations on 150 participants in a randomized clinical trial.
+We created a dataset representing observations on 150 participants in a randomized clinical trial.
 Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory.
@@
 
+See Figure~\ref{fig:example_glimmertree} for the GLMM tree fitted to the simulated data set which recovers the true underlying structure.
+
 <<example-trees>>=
@@
 @
+
+\begin{figure}[t!]
+<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
+plot(exampleglmmtree)
+@
+\caption{GLMM tree of the motivating example dataset.
+Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
+\label{fig:example_glimmertree}
+\end{figure}
 
@@
 Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper.
-Also, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.
+Also, for the continuous response variable in the motivating data set, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.
 However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.
@@
 In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented.
-As can be seen, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.
-
-\begin{figure}[t!]
-<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
-plot(exampleglmmtree)
-@
-\caption{GLMM tree of the motivating example dataset.
-Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
-\label{fig:example_glimmertree}
-\end{figure}
+As already mentioned above, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.
 

### 2016-10-09 pkg/glmertree/paper/treatment_interactions.Rnw
various improvements -> still some open READMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@997 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \setkeys{Gin}{width=0.8\textwidth}
+
+%% for comments
+\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
 
@@
 
-Participants were part of one of ten clusters, each with a different value for the random intercept.
-Data were generated such that covariates and cluster-specific intercepts were uncorrelated.
-43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.
-
-See Figure~\ref{fig:example_glimmertree} for the GLMM tree fitted to the simulated data set which recovers the true underlying structure.
-
 <<example-trees>>=
@@
 \begin{figure}[t!]
+\centering
 <<exampleglmmtree, fig=TRUE, width=9, height=7>>=
@@
 \end{figure}
+
+Participants were part of one of ten clusters, each with a different value for the random intercept.
+Data were generated such that covariates and cluster-specific intercepts were uncorrelated.
+43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.
+
+See Figure~\ref{fig:example_glimmertree} for the GLMM tree fitted to the simulated data set which recovers the true underlying structure.
 
@@
 \begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.5\textwidth}
@@
 
-If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3).
-This partition is created using $U_{k*}$, the variable with the minimal p-value in step (2).
-The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}.
-In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.
+If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a $p$-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3).
+This partition is created using $U_{{k^*}}$, the variable with the minimal $p$-value in step (2).
+The split point for $U_{{k^*}}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}.
+In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).
+
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=\textwidth}
+<<exampleglmtree, fig=TRUE, width=12, height=9>>=
+plot(exampleglmtree)
+@
+\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset.
+Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
+\label{fig:example_mobtree}
+\end{figure}
 
@@
 Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
-$\beta^{*}$ would have length $2J$ as well, containing subgroup-specific fixed-effects coefficients.
-
-\begin{figure}[t!]
-\setkeys{Gin}{width=\textwidth}
-<<exampleglmtree, fig=TRUE, width=12, height=9>>=
-plot(exampleglmtree)
-@
-\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset.
-Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
-\label{fig:example_mobtree}
-\end{figure}
+The corresponding $\beta^{*}$ would have length $2J$ as well, containing subgroup-specific fixed-effects coefficients.
 
@@
 Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above.
-However, the two subgroups in the middle do not represent true subgroups, which may be due to the clustered structure of the dataset not being taken into account.
+However, the two subgroups in the middle do not represent true subgroups, which is due to the clustered structure of the dataset not being taken into account.
 
@@
 %
-Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
-Further, $b$ is a random vector of length $M$, each $m$th element representing the random intercept for cluster $m$; again, for simplicity we assume that only cluster-specific intercepts are included in the model, but multiple random-effects coefficients can easily be included.
+For a simple random intercept $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
+Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$.
+For simplicity, we just employ such a cluster-specific intercept here but further random effects can be easily included in $z_i$.
 Furthermore, within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
@@
 Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part may not describe the data well.
-Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} and random effects are incorporated as well:
+Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} while still adjusting for random effects:
 %
@@
 
-For estimating GLMM trees, we take this approach a step further: instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree.
-This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes.
-In addition, GLMM trees can be estimated for continuous, as well as binary and count variables.
+For estimating GLMM trees, we take this approach two steps further: (1)~Instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree.
+This allows not only for detection of differences in intercepts/levels across terminal nodes, but also for detection of differences in slopes/regression coefficients such as treatment effects.
+(2)~By using generalized linear (mixed) models, the response may also be a binary or count variable instead of a continuous variable.
 The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
-
 \begin{description}
@@
 Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}.
+Typically, this converges if the tree does not change from one iteration to the next.
 
@@
 
-\section{Simulation-based Evaluation: Method}
+\section{Simulation-based evaluation: Method}
 
-To assess the performance of GLMM tree, we performed three simulation studies.
+To assess the performance of GLMM tree, we carried out three simulation studies.
 In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions.
@@
 
-R \cite{R14} was used for the generation and analysis of all datasets.
-The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions.
-For other response distributions, the \verb|glmtree| function would be available.
+R \cite{R} was used for the generation and analysis of all datasets.
+The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| is available for other response distributions.
 For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
-
+%
 For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
@@
 
+\readme{Z: My impression is that the description of the data-generating process is too far ahead of the simulation itself.
+I would recommend to first explain what is the same in all three simulation studies.
+And then what is specific could maybe be included at the beginning of each of the individual studies so that readers have it in mind when looking at the results.
+So in this section here I would just briefly explain the idea and defer all details until later.}
+
 \subsubsection{Treatment-subgroup interactions} 
@@
 \begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.8\textwidth}
@@
 	\item a main effect of $U_2$ (coefficient = 1.5) 
-	\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = -.25), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = .25)
-	\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = -.25), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = .25)
+	\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = $-.25$), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = $.25$)
+	\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = $-.25$), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = $.25$)
 \end{enumerate}
+
+\readme{Z: Also include the formula for the linear predictor in this setup so that this can be easily matched with the GLMM estimated.}
 
@@
 
-\section{Simulation-based Evaluation: Results}
+\section{Simulation-based evaluation: Results}
 
@@
 \begin{figure}[t!]
+\centering
 <<treesize-xyplot, fig=TRUE, width=9, height=9>>=
@@
 \begin{figure}[t!]
+\centering
 <<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
@@
 \begin{figure}[t!]
+\centering
 <<correlation-xyplot, fig=TRUE, width=9, height=7>>=
@@
 
+\readme{Z: Wouldn't it be more natural to just look at P(tree size greater than 1), i.e., the probability of splitting at all.
+Then it could be worked out more clearly that GLMM trees approximately keep the significance level while GLM trees generally have an inflated irror.}
+
 \subsubsection{Design}
@@
 \begin{figure}[t!]
+\centering
 <<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
@@
 \begin{figure}[t!]
+\centering
 <<accuracy-pw-xyplot, fig=TRUE, width=9, height=7>>=
@@
 \begin{figure}[t!]
+\centering
 <<correlation-pw-xyplot, fig=TRUE, width=9, height=7>>=
@@
 
-To further illustrate the use of GLM and GLMM tree, we applied both algorithms to a dataset from a individual-patient data meta-analysis of \citeA{CuijyWeit14}.
+To further illustrate the use of GLM and GLMM tree, we applied both algorithms to a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}.
 This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
@@
 The results of recursive partitioning techniques are known to be potentially unstable, in the sense that small changes in the dataset may substantially alter the variables or values selected for partitioning.
-Therefore, we used the R-package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values.
+Therefore, we used the R~package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values.
 Using the \verb|stabletree| function, we calculated the number of times a variable and value were selected for partitioning, over 500 subsamples of size $.9 \times N$ of the dataset.
@@
 \begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.75\textwidth}
-<<depression-glmtree, fig=TRUE, height=6, width=8>>=
-plot(lmtree_app)
+<<depression-glmtree, fig=TRUE, height=9, width=8>>=
+combine_panels <- function(panel1, panel2, party2 = NULL) {
+  function(node) {
+    nid <- id_node(node)
+    pushViewport(viewport(
+      layout = grid.layout(nrow = 2, ncol = 1, heights = c(1.1, 0.7)),
+      name = paste("node_mob_mypanel", nid, sep = "")))
+    grid.rect(gp = gpar(fill = "white", col = 0))
+    pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+    panel1(node)
+    popViewport()
+    pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+    node2 <- if(is.null(party2)) node else node_party(party2[nid])
+    panel2(node2)
+    popViewport(2)
+  }
+}
+
+plot(lmtree_app, tnex = 3.5, terminal_panel = combine_panels(
+  panel1 = node_bivplot(lmtree_app),
+  panel2 = node_terminal(GLMtree_app_eff, FUN = identity, height = 10, width = 14, id = FALSE),
+  party2 = GLMtree_app_eff
+))
 @
-<<depression-glmtree-effs, fig=TRUE, height=6, width=8>>=
-plot(GLMtree_app_eff, tp_args = list(FUN = identity, height = 10, width = 14), tnex = 1.5)
-@
-\caption{Upper panel: GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).
+\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
+Upper panel: The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
 Lower panel: Subgroup-specific descriptive statistics.}
@@
 Note that the selection frequencies do not add up to 1, as trees may involve multiple, or no splits.
-Table \ref{tab:stability} indicates that the presence of a comorbid anxiety disorder was selected for partitioning in the majority ofGLMM trees grown on subsamples of the dataset, and all other variables were selected in less than 3\% of the subsamples.
+Table \ref{tab:stability} indicates that the presence of a comorbid anxiety disorder was selected for partitioning in the majority of GLMM trees grown on subsamples of the dataset, and all other variables were selected in less than 3\% of the subsamples.
 As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.
@@
 \begin{figure}
+\centering
 \setkeys{Gin}{width=0.5\textwidth}
-<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
-plot(lmertree_app)
+<<depression-glmmtree, fig=TRUE, height=7.5, width=5.5>>=
+plot(lmertree_app$tree, tnex = 3.5, terminal_panel = combine_panels(
+  panel1 = node_bivplot(lmertree_app$tree),
+  panel2 = node_terminal(GLMMtree_app_eff, FUN = identity, height = 10, width = 14, id = FALSE),
+  party2 = GLMMtree_app_eff
+))
 @
-<<depression-glmmtree-effs, fig=TRUE, height=4, width=4.7>>=
-plot(GLMMtree_app_eff, tp_args = list(FUN = identity, height = 10, width = 14), tnex = 1.5)
-@
-\caption{Left panel: GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).
+\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
+Left panel: The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
 Right panel: Subgroup-specific descriptive statistics.}

### 2016-11-07 pkg/glmertree/paper/treatment_interactions.Rnw
Several changes in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1013 255f4249-2263-4417-b1ea-8498779e27a7

@@
 GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets.
-GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4\% and 33\%, respectively).
-Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on average), but better accuracy in detecting interactions.
+GLMM tree also outperformed trees without random effects in terms of predictive accuracy and Type-I error rates.
+Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed better predictive accuracy.
 We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression.
-We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
+We conclude that GLMM tree is a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
 \\
@@
 However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset.
-Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM trees.
+Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model tree, or GLMM tree.
 
@@
 We created a dataset representing observations on 150 participants in a randomized clinical trial.
-Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory.
+Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the pos-ttreatment total score on a depression inventory.
 For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).
@@
 Data were generated such that covariates and cluster-specific intercepts were uncorrelated.
-43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.
+43\% of variance in post-treatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.
 
@@
 For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset.
-The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
+The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function:
 %
@@
 Also, for the continuous response variable in the motivating data set, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.
-However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.
+However, the model can easily accommodate additional treatment conditions and covariates, and binary or count outcome variables.
 
@@
 Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$.
-For simplicity, we just employ such a cluster-specific intercept here but further random effects can be easily included in $z_i$.
-Furthermore, within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
+For simplicity we only employ a cluster-specific intercept here, but further random effects can be easily included in $z_i$.
+Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
 The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).
@@
 For estimating GLMM trees, we take this approach two steps further: (1)~Instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree.
-This allows not only for detection of differences in intercepts/levels across terminal nodes, but also for detection of differences in slopes/regression coefficients such as treatment effects.
+This allows not only for detection of differences in intercepts across terminal nodes, but also for detection of differences in slopes such as treatment effects.
 (2)~By using generalized linear (mixed) models, the response may also be a binary or count variable instead of a continuous variable.
@@
 In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented.
-As already mentioned above, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.
+By taking into account the clustering of observations, the spurious split involving the anxiety variable no longer appears in the tree and the true treatment subgroups have been recovered.
 
-\section{Simulation-based evaluation: Method}
+\section{Simulation-based evaluation}
 
-To assess the performance of GLMM tree, we carried out three simulation studies.
-In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions.
-In Study II, we compared the accuracy of GLMM tree with that of GLM tree in datasets without treatment-subgroup interactions, allowing for assessment of the Type-I error rate.
-In Study III, we compared the accuracy of GLMM tree with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.
+To assess the performance of GLMM tree, we carried out three simulation studies: In Studies I and II, we compared the performance of GLMM tree with that of GLM tree.
+In Study I, we assessed the accuracy of both tree-based algorithms in datasets with treatment-subgroup interactions.
+In Study II, we assessed the Type-I error of both tree-based algorithms in datasets without treatment-subgroup interactions.
+In Study III, we compared the performance of GLMM tree in detecting treatment interactions with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.
+
+\subsection{General simulation design}
+
+In all simulation studies, we varied the following data-generating parameters:
+
+\begin{enumerate} 
+\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
+\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
+\item Two levels for the intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
+\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
+\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
+\item Two levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates.  
+\end{enumerate}
+
+In addition, each simulation study required a different design for main and interaction effects, which will be described in their respective sections below.
+
+Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$.
+Means for the other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$.
+All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$.
+Correlations between $U$ variables vary according to the third facet of the data-generating design described above.
+
+To generate the cluster-specific intercepts $b_m$, we partitioned the sample into $M$ equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design.
+For each cluster, a single value $b_m$ was drawn from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design.
+When $b$ was correlated with one of the potential partitioning variables, this variable was randomly selected.
+
+For every observation, we generated a binomial variable (with probability .5) as an indicator for treatment type.
+Random errors $\epsilon$ were drawn from a normal distribution with $\mu_{\epsilon} = 0$ and $\sigma_{\epsilon} = 5$.
+The value of the outcome variable $y_i$ was calculated as the sum of the random intercept, (node-specific) fixed effects and the random error term.
 
@@
 R \cite{R} was used for the generation and analysis of all datasets.
-The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| is available for other response distributions.
-For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
-%
+The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| for other response distributions.
+For estimating GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
+
 For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
@@
 
-In all simulations, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing.
+The significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing, in all simulations.
 The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight.
 
-\subsection{Simulation design}
+In Study III, the \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in GLMMs.
+The \verb|lmerTest| package calculates effective degrees of freedom and $p$-values based on Satterthwaite approximations.
 
-\readme{Z: My impression is that the description of the data-generating process is too far ahead of the simulation itself.
-I would recommend to first explain what is the same in all three simulation studies.
-And then what is specific could maybe be included at the beginning of each of the individual studies so that readers have it in mind when looking at the results.
-So in this section here I would just briefly explain the idea and defer all details until later.}
-
-\subsubsection{Treatment-subgroup interactions} 
-For generating datasets with treatment-subgroup interactions, we used one of the treatment-subgroup interaction designs from \citeA{DussyMech14}, which is depicted in Figure~\ref{fig:modelC}.
-Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes.
-The four subgroups are characterized by values on the (true) partitioning variables $U_2$, and $U_1$ or $U_5$.
-The other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.
+\subsection{Study I: Method}
 
@@
 
-\subsubsection{Continuous interactions}
+\subsubsection{Treatment-subgroup interaction design}
 
-The subgroups depicted in Figure~\ref{fig:modelC} are piecewise functions of the partitioning variables, therefore these interactions can be referred to as piecewise interactions.
-However, interactions may also consist of continuous functions of the partitioning variables.
-For comparing the accuracy of GLMM tree with that of GLMMs with pre-specified interaction terms, we generated datasets in which interactions were varied to be purely piecewise, purely continuous, or a combination of both.
-To generate purely continuous interactions, the outcome variable $y$ was generated as the sum of:
+For generating datasets with treatment-subgroup (or piecewise) interactions, we used a design from \citeA{DussyMech14} which is depicted in Figure~\ref{fig:modelC}.
+Figure~\ref{fig:modelC} shows four subgroups, characterized by values of the (true) partitioning variables $U_2$, and $U_1$ or $U_5$.
+Two of the subgroups have mean differences in treatment outcome, indicated by a non-zero value of $\beta_{j1}$, and two subgroups do not have mean differences in treatment outcome, indicated by a $\beta_{j1}$ value of 0.
+
+In this simulation study, some of the potential partitioning covariates were true partitioning covariates ($U_1$, $U_2$ and $U_5$), whereas others were not (e.g., $U_3$ and $U_4$).
+Therefore, an extra level was added to the sixth facet of the \textit{General simulation design} described above.
+Also, to assess the effect of the size of the treatment-effect differences in nodes 3 and 7, an additional facet was added:
 
 \begin{enumerate}
-	\item a main effect of $U_2$ (coefficient = 1.5) 
-	\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = $-.25$), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = $.25$)
-	\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = $-.25$), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = $.25$)
+\item[6.] Three levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the true partitioning covariates ($U_1$, $U_2$ or $U_5$), $b$ correlated with one of the noise variables ($U_3$ or $U_4$.
+\item[7.] Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes.
+The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
 \end{enumerate}
 
-\readme{Z: Also include the formula for the linear predictor in this setup so that this can be easily matched with the GLMM estimated.}
-
-Note that the direction (positive or negative) and type (main or interaction) of these effects correspond to those in Figure~\ref{fig:modelC}.
-All interactions were created by using centered $U_k$ variables, calculated by subtracting the variable means.
-
-\subsubsection{Data-generating parameters} 
-In generating datasets, we varied seven parameters of the data-generating process:
-
-\begin{enumerate} 
-\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
-\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
-\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
-\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
-\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
-\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition.
-However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}.
-\item Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes.
-In datasets with treatment-subgroup interactions, the treatment effect difference varied only in nodes 4 and 7, with levels $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
-\end{enumerate}
-
-\subsubsection{Variable distributions}
-Following the approach of \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$.
-The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$.
-All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$.
-Correlations between $U$ variables vary according to the third facet of the data-generating design described above.
-
-To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design.
-For each cluster, a single value $b_m$ was drawn from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design.
-When $b$ was correlated with one of the potential partitioning variables, this variable was randomly selected.
-
-To generate node-specific fixed effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3.
-In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$.
-For every observation, we generated a binomial variable (with probability .5) as an indicator for treatment type.
-Random errors $\epsilon$ were drawn from a normal distribution with $\mu_{\epsilon} = 0$ and $\sigma_{\epsilon} = 5$.
-
-Finally, the response variable was calculated as the sum of the (node-specific) fixed effects, the random intercept and the error term: $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.
+Again, 50 datasets were generated for each cell of the design.
+As a result, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2 = 32,400$ datasets were generated.
+In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.
 
@@
 
-Firstly, the total number of nodes in estimated GLM and GLMM tree were calculated to assess the extent to which the algorithms may detect spurious subgroups.
-Also, the number of nodes allow for assessing Type-I (the extent to which datasets are erroneously partitioned) and Type-II (the extent to which datasets are erroneously not partitioned) error.
+Performance of the algorithms was assessed by means of the total number of nodes in estimated GLM and GLMM trees, the accuracy of the resulting trees and predictive accuracy.
+An accurately recovered tree was defined as a tree with (1) a total of seven nodes (2) the first split involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$.
+The allowance of $\pm 5$ equals plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).
 
-Secondly, in datasets with treatment-subgroup interactions, we assessed the accuracy of estimated models.
-For GLM and GLMM trees, an accurately recovered tree was defined as a tree with (1) the true tree size (i.e., tree size $= 7$ in datasets with treatment-subgroup interactions), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$.
-Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).
-
-Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) in each dataset.
-Consequently, predictive accuracy was only assessed in datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.
+Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences in every dataset.
 To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets.
@@
 
-\section{Simulation-based evaluation: Results}
+To identify the most important predictors of tree size, tree accuracy and predictive accuracy, we used ANOVAs and GLMs with algorithm type and the parameters of the data-generating process as independent variables.
+First-order interactions between algorithm type and each of the data-generating parameters were also included.
+The most important predictors were further assessed with graphical displays.
 
-\subsection{Study I: Detection of Treatment-Subgroup Interactions}
-
-For each cell of the data-generating design described above, 50 datasets were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets with treatment-subgroup interactions.
-In these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).
+\subsection{Study I: Results}
 
@@
 
+\subsubsection{Tree size}
+
 \begin{figure}[t!]
@@
 type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
-abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
-)
+abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
 @
@@
 
-\subsubsection{Tree size}
-
-GLMM tree yielded trees with an average size of 7.15 nodes ($\mathrm{SD}=0.61$), whereas GLM tree yielded an average tree size of 8.15 nodes ($\mathrm{SD}=2.05$), indicating that GLM trees involved more spurious splits than GLMM trees, on average.
+GLMM tree yielded trees with an average size of 7.15 nodes ($\mathrm{SD}=0.61$), whereas GLM tree yielded an trees with an average size of 8.15 nodes ($\mathrm{SD}=2.05$), indicating that GLM trees involved more spurious splits than GLMM trees on average.
 The estimated probability that a dataset was erroneously not partitioned (the Type-II error) was 0 for both algorithms.
 
-To find the most important predictors of tree size, we performed an ANOVA with algorithm type and the parameters of the data-generating process as independent variables.
-First-order interactions between algorithm type and each of the data-generating parameters were also included.
-The most important predictors of tree size were algorithm type, sample size, variance of the random intercept, and the correlation between partitioning variables.
-These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treesize_interact}).
-
-Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions.
-On average, tree size increases slightly with sample size for GLMM tree, and dramatically for GLM tree.
+An ANOVA indicated that the most important predictors of tree size were algorithm type, sample size, variance of the random intercept, and the correlation between partitioning variables.
+The graphical display in Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions.
 In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size.
 However, clear differences in tree size were observed when $\sigma_b > 0$.
-When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average.
+When sample size is small and the random intercept is not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average.
 When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized.
-However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially with larger sample sizes (i.e., 500 or 1000) and larger $\sigma_b$ values (i.e., 10).
-This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.
+However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially with larger sample sizes and larger $\sigma_b$ values.
+
+\subsubsection{Accuracy of recovered trees}
 
@@
 
-\subsubsection{Accuracy of recovered trees}
-
 To assess the accuracy of trees recovered by both algorithms, we inspected variables and values that were selected for partitioning.
@@
 
-However, further splits were more accurately recovered by GLMM tree, which accurately recovered the partition in 90.40\% of datasets, and GLM tree in only 61.44\% of datasets.
-To find the most important predictors of tree accuracy, we fitted a GLM with algorithm type and the parameters of the data-generating process as independent variables.
-First-order interactions between algorithm type and each of the data-generating parameters were also included.
-The most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and the level of dependence between the partitioning variables and the random intercept.
-These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treeaccuracy}).
+Further splits were more accurately recovered by GLMM tree, which accurately recovered the partition in 90.40\% of datasets, and GLM tree in only 61.44\% of datasets.
+A GLM with tree accuracy as the outcome variable indicates that the most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and level of dependence between the partitioning variables and the random intercept.
+The graphical display in Figure~\ref{fig:xyplot_treeaccuracy} indicates that in the absence of random effects, the trees recovered by GLM and GLMM tree were about equally accurate.
+In the presence of random effects, GLM trees were much less accurate than GLMM trees when random effects were correlated with a partitioning covariate.
+When random intercepts were not correlated with one of the $U_k$ variables, GLMM tree outperformed GLM tree when sample size was small (i.e., $N = 200$).
 
-Figure~\ref{fig:xyplot_treeaccuracy} indicates that in the absence of random effects, the trees recovered by GLM and GLMM tree were about equally accurate.
-In the presence of random effects, GLM trees were much less accurate than GLMM trees.
-This was found for all sample sizes, when random effects were correlated with a partitioning covariate.
-This effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable.
-However, when random intercepts were not correlated with one of the $U_k$ variables, GLMM tree outperformed GLM tree only when sample size was small (i.e., $N = 200$).
+\subsubsection{Predictive accuracy}
 
@@
 auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T),
-scales=list(y=list(limits = c(0,1))))
+scales=list(y=list(limits = c(-.1,1.1))))
 )
@@
 \caption{Average predictive accuracy of GLM and GLMM trees.
-Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2.
+Predictive accuracy of trees is defined as the correlation between the true and predicted outcome differences between Treatment~1 and~2.
 Rows represent the levels of the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences, columns represent levels of sample size.}
@@
 
-\subsubsection{Predictive accuracy of trees}
 The predicted treatment-effect differences of GLMM tree were closer to the true differences than those of GLM tree, with a mean correlation of .93 ($\mathrm{SD}=0.12$) for GLMM tree and .88 ($\mathrm{SD}=0.19$) for GLM tree.
-An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the treatment-difference effect size.
-These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_correlations}).
-
-Figure~\ref{fig:xyplot_correlations} shows clear main effects of sample size and treatment-effect differences for both algorithms.
+An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the effect size of the treatment difference.
+The graphical display in Figure~\ref{fig:xyplot_correlations} shows clear main effects of sample size and treatment-effect differences for both algorithms.
 In the absence of random effects (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were about equally accurate.
@@
 
-\subsection{Study II: Type-I error}
-
-\readme{Z: Wouldn't it be more natural to just look at P(tree size greater than 1), i.e., the probability of splitting at all.
-Then it could be worked out more clearly that GLMM trees approximately keep the significance level while GLM trees generally have an inflated irror.}
+\subsection{Study II: Method}
 
 \subsubsection{Design}
-In addition to the accuracy of treatment-subgroup interaction detection, we assessed the Type-I error: the extent to which datasets without treatment-subgroup interactions may erroneously be partitioned.
-For assessing Type-I error of GLM and GLMM tree, we generated datasets without treatment-subgroup interactions, in which there was only a main effect of treatment in the population.
+In the second simulation study we assessed the Type-I error of GLM and GLMM tree.
+Type-I error was defined as the proportion of datasets without treatment-subgroup interactions which were erroneously partitioned by the algorithm.
+In the simulated datasets in this study there was only a main effect of treatment in the population.
 Put differently, there was only a single global value of $\beta_j = \beta$ in every dataset.
-For these datasets, we varied the same parameters of the data-generating process as in Study I.
-However, the sixth facet of the data-generating process had only two levels for these datasets, as there were no `true' partitioning variables in these datasets.
-Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.
 
-\subsubsection{Tree size and Type-I error}
+To assess the effect of the treatment-effect difference $\beta$, an additional facet was added to the \textit{General simulation design} described above: 
+\begin{enumerate}
+\item[7.] Two levels for $\beta$, the unstandardized mean difference in treatment outcomes.
+The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
+\end{enumerate}
+
+For each cell of the \textit{General simulation design} described above, 50 datasets were generated.
+As a result, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2 = 21,600$ datasets without treatment-subgroup interactions were generated.
+In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta + z_{i}^{\top} b_{m} + \epsilon_{i}$.
+
+\subsubsection{Assessment of performance}
+
+The total number of nodes in estimated GLM and GLMM tree were calculated.
+A tree of size $>1$ was classified as a Type-I error.
+To identify the most important predictors of Type-I error, we used a GLMs with algorithm type and the parameters of the data-generating process as independent variables.
+First-order interactions between algorithm type and each of the data-generating parameters were also included.
+The most important predictors were further assessed with a graphical display.
+
+\subsection{Study II: Results}
+
 In datasets without treatment-subgroup interactions, average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree.
-The overall Type-I error rate was only .04 for GLMM tree, and .33 for GLM tree.
-Again, main predictors of tree size were found to be sample size, variance of the random intercepts, and dependence between the random intercept and one of the partitioning variables.
-These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treesize_nointeract}).
-Figure~\ref{fig:xyplot_treesize_nointeract} indicates that in the absence of random effects (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tend to create trees of size 1.
-In the presence of random effects, GLMM tree still tends to create trees of size 1, but GLM tree creates much larger trees, when the random effects are correlated with one of the partitioning covariates.
-This effect increases with sample size.
+The average Type-I error rate was only .04 for GLMM tree, and .33 for GLM tree.
+Main predictors of type-I error were found to be sample size, variance of the random intercept, and dependence between the random intercept and one of the partitioning variables.
+
+The graphical display in Figure~\ref{fig:xyplot_treesize_nointeract} indicates that GLMM tree has a Type-I error rate somewhat below the pre-specified $\alpha$ level under all circumstances.
+The same goes for GLM tree, whenever the random intercept has 0 variance, or the random intercept is not correlated to one of the partitioning covariates.
+However, when the random intercept is correlated to one of the potential partitioning covariates, the type-I error rapidly increases for GLM tree.
+With large sample sizes and large variance of the random intercept, GLM tree will almost certainly create spurious splits.
 
 \begin{figure}[t!]
-\centering
 <<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
@@
 treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
-aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
-print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
-type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
-abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
-lines=TRUE, points=TRUE)))
+treesizes.long$treesize.valuesD <- as.numeric(treesizes.long$treesize.values>1)
+treesize.glm <- glm(treesize.valuesD ~ treesize.ind + N + rho + np + treatdiff + numbclus + 
+sigmabm + corUbm + treesize.ind*(N + rho + np + treatdiff + numbclus + sigmabm + corUbm), 
+data=treesizes.long, family="binomial")
+summary(treesize.glm)
+#N, corUb & sigmab have strongest main and/or interaction effects
+aggdata.size <- aggregate(formula=treesize.valuesD ~ treesize.ind + N + sigmabm + corUbm, 
+FUN=mean, data=treesizes.long)
+levels(aggdata.size$corUbm)[levels(aggdata.size$corUbm)=="b and random U correlated"] <- 
+"b and U correlated"
+levels(aggdata.size$corUbm)[levels(aggdata.size$corUbm)=="uncorrelated"] <- 
+"b and U uncorrelated"
+xyplot(treesize.valuesD ~ sigmabm | N + corUbm, data = aggdata.size, groups=treesize.ind, 
+type="b", ylab="Type-I error", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
+auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
 @
-\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions.
-Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
+\caption{Type-I error of GLM and GLMM trees. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
 \label{fig:xyplot_treesize_nointeract}
@@
 
-\subsection{Study III: Piecewise vs. Continuous Interactions}
+\subsection{Study III: Method}
 
-\subsubsection{Design}
+\subsubsection{Interaction design}
 
-To compare the performance of GLMM tree with that of GLMMs with pre-specified interaction effects, we created datasets in which interactions were varied from purely continuous to purely piecewise.
-For these datasets, the seventh facet of the interaction design (effect size of treatment differences) was replaced by a factor with three levels, controlling the type of interactions: purely continuous interactions, purely piecewise interactions, and a combination of both.
-Purely continuous interactions were generated as described above, in the Method section.
-To allow for a smooth transition from purely continuous to purely continuous interactions, we aimed to keep the proportion of variance explained by each of the covariates equal across piecewise, continuous and combined interactions.
-Therefore, for piecewise interactions, we used the partition depicted in Figure 4, but not the coefficients.
-The coefficients used for data-generation were replaced by the subgroup-specific means, given purely continuous interactions.
-To generate the combined conitnuous and piecewise interactions, $y$ values according to the purely piecewise and continuous interaction models were summed and divided by 2.
+The treatment-subgroup interactions in Study I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables $U_1$, $U_2$ and $U_5$.
+Trees are pre-eminently suited for recovering such piecewise interactions, but may have difficulty when the true interactions are continuous functions of moderator variables (for example, $U_1 \cdot U_2$).
+At the same time, linear regression models with pre-specified interaction terms may perform well in recovering continuous interactions, but may have difficulty in recovering piecewise interactions.
+Therefore, in the third simulation study, we added a seventh facet to the \textit{General simulation design} described above:
 
-Furthermore, in this simulation study, the number of cells in the simulation design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters), as the earlier simulations indicated no effects of the number of clusters.
+\begin{enumerate}
+	\item[7.] Three levels for interaction type: continuous, piecewise and a combination of both. 
+\end{enumerate}
+
+For datasets with purely piecewise interactions, the same partition as in Study II (depicted in Figure \ref{fig:modelC}) was used.
+In other words, the outcome variable in this design was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
+
+For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure \ref{fig:modelC} was used.
+However, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes comprised continuous main and (treatment) interaction effects of the partitioning variables.
+These node-specific effects are described in Table \ref{tab:continuous_terms}.
+$\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure \ref{fig:modelC}.
+Continuous interactions were created using centered $U_k$ variables, calculated by subtracting the variable mean.
+
+For datasets with purely continuous interactions, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$.
+Here, $\beta$ has a global value and no subscript, and comprised purely continuous main and (treatment) interaction effects of the moderator variables $U$, as described in Table \ref{tab:continuous_terms}.
+
+\begin{table}
+\caption{Fixed-effects terms in simulations with continuous and combined continuous and piecewise interaction designs.}
+\begin{tabular}{lccccc}
+\thickline
+term							&	$\beta_3$&	$\beta_4$&	$\beta_6$&	$\beta_7$&	$\beta$		\\
+\hline
+intercept						&	27		&	27		&	27		&	27		&	27			\\
+$U_2$							&	.1		&	.1		&	.1		&	.1		&	.1			\\
+$U_2 \cdot U_1$ 				&	-.357	&	0		&	0		&	0		&	-.357		\\
+$U_2 \cdot U_5$	 				&	0		&	0		&	0		&	.357	&	.357		\\
+$U_2 \cdot U_1 \cdot treatment$	&	-.151	&	0		&	0		&	0		&	-.151		\\
+$U_2 \cdot U_5 \cdot treatment$	&	0		&	0		&	0		&	.151	&	.151		\\
+\hline
+\multicolumn{6}{l}{\small{\textit{Note. }Subscripted $\beta$ values refer to the combined}}\\	\multicolumn{6}{l}{\small piecewise and continuous interaction design, with the}\\
+\multicolumn{6}{l}{\small values of the subscripts denoting the terminal nodes in}\\
+\multicolumn{6}{l}{\small Figure \ref{fig:modelC}. $\beta$ without a subscript refers to the global}\\
+\multicolumn{6}{l}{\small coefficients in the purely continuous interaction design.}\\
+\end{tabular}	
+\label{tab:continuous_terms}
+\end{table}
+
+Furthermore, in this simulation study, the number of cells in the design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters), as Study I and II indicated no effects of the number of clusters.
 The fifth facet of the data-generating design was limited to two levels ($\sigma_b = 2.5$ and $\sigma_b = 7.5$).
-As in Study II, we limited the sixth facet to two levels (the random intercept was either correlated to one of the covariates, or not).
-To counter unidentifiability due to potentially large numbers of pre-specified interaction terms in the GLMMs, the second level of the second facet (number of potential partitioning covariates) was set to $k=10$, instead of the value of $k=15$ used in the earlier simulations.
 As a result, $50 \times 3 \times 2 \times 2 \times 1 \times 2 \times 2 \times 3$ = 7,200 training datasets were generated for this study.
 
-GLMMs were estimated by specifying main effects for all covariates $U_k$ and the treatment indicator, first-order (or two-way) interactions between all pairs of covariates $U_k$, and second-order (or three-way) interactions between all pairs of covariates $U_k$ and treatment.
-Accurately recovered GLMMs were defined to be models in which significant (i.e., absolute value of the $t$-test statistic $\geq 1.96$) effects were found for $U_2$ and (treatment) interactions of $U_1$ and $U_2$ and $U_2$ and $U_5$.
-Accurately recovered GLMM trees were defined to be trees in which the first split involved $U_2$, the second split to the right involved $U_1$ and the third split to the right involved $U_5$.
-Note that this criterion of accuracy is somewhat more lenient than in Study I.
+\subsubsection{GLMMs with pre-specified interactions}
+GLMMs were estimated by specifying main effects for all covariates $U_k$ and the treatment indicator, first-order interactions between all pairs of covariates $U_k$, and second-order interactions between all pairs of covariates $U_k$ and treatment.
+Continuous predictor variables were centered by subtracting the mean value, before calculating and including the interaction term in the GLMM.
+Predictions were obtained by refitting a GLMM on the training data, using only the predictors with p-values $< .05$ in the original GLMM.
+Predictions for test observations were calculated using the refitted GLMMs.
+
+\subsubsection{Assessment of performance} 
+Predictive accuracy was assessed in terms of the correlation between the true and predicted treatment-effect differences in test datasets.
+To identify the most important predictors of predictive accuracy, we used an ANOVA with algorithm type and the parameters of the data-generating process as independent variables.
+First-order interactions between algorithm type and each of the data-generating parameters were also included.
+The most important predictors were further assessed with a graphical display.
 
@@
 treespecs.long$type <- factor(treespecs.long$type, ordered = TRUE, levels = 
-c("continuous", "both", "piecewise"))
+c("linear", "both", "piecewise"))
+levels(treespecs.long$type) <- gsub("linear","continuous", levels(treespecs.long$type))
 @
 
-\subsubsection{Accuracy of recovered models}
-
-Overall, GLMM tree accurately recovered main and interaction effects in 53.58\% of datasets, wheareas GLMMs accurately recovered main and interaction effects in 34.81\% of datasets.
-To find the most important predictors of model accuracy, we performed an ANOVA, in which the algorithm type, sample size, the number of covariates and type of interaction (continuous, piecewise or both) were found to be most important.
-These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_accuracy_pw}.
-
-Figure~\ref{fig:xyplot_accuracy_pw} shows a clear effect of interaction type: GLMM tree performs best in datasets with purely piecewise interactions, and poorest in datasets with purely continuous interactions.
-GLMMs perform best in datasets with purely continuous interactions, and poorest in datasets with purely piecewise interactions.
-Figure~\ref{fig:xyplot_accuracy_pw} also shows a clear effect of the number of covariates: with a large number of covariates, GLMMs never recover all true effects in the dataset, whereas they always recover the true effects when the number of covariates is small and interaction effects are at least partly continuous.
-At the same time, the accuracy of GLMM tree is unaffected by the number of covariates.
-Finally, the performance of both GLMM tree and GLMMs is affected by sample size: the effects are more often recovered in datasets with larger sample sizes.
-
-\begin{figure}[t!]
-\centering
-<<accuracy-pw-xyplot, fig=TRUE, width=9, height=7>>=
-aggdata.acc <- aggregate(formula=truevars ~ truevars.ind + N + type + np, FUN=mean, 
-data=treespecs.long)
-xyplot(truevars ~ type | N + np, data = aggdata.acc, groups=truevars.ind, type="b",
-ylab="proportion of accurately recovered models", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
-auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
-@
-\caption{Model accuracy for GLMMs and GLMM trees.
-In an accurate model, the true main and interaction effects were accurately recovered.
-Columns represent sample size, rows represent the number of covariates in the model.}
-\label{fig:xyplot_accuracy_pw}
-\end{figure}
-
-\subsubsection{Predictive accuracy}
+\subsection{Study III: Results}
 
@@
 ylab="correlation", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
-auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
+auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T),
+scales=list(y=list(limits = c(-.1,1.1))))
 @
@@
 
-Both algorithms showed about equal average correlations between the true and predicted treatment-effect differences: average accuracy was .60 (SD = .36) for GLMMs and .60 (SD = .32) for GLMM trees.
-In terms of bias, GLMM treatment-effect difference predictions deviated more from the true values than the predictions of GLMM tree.
-The mean true treatment-effect difference averaged over all datasets was .04 (SD = 1.22), whereas the mean predicted treatment-effect differences averaged over all datasets for GLMMs were -.26 (SD = 100.33) and for GLMM trees .03 (SD = 3.88).
-Furthermore, it should be noted that the predictive models for GLMM trees were simpler than those of the GLMMs.
-Predictions for GLMMs in datasets with 10 covariates involved 11 (significant or non-significant) main effects, 100 first-order interactions and 100 second-order interactions.
-Predictions for GLMM tree involved a maximum  of only 8 terminal nodes, as their maximum depth was restricted.
-
+GLMM tree showed somewhat higher average correlations between the true and predicted treatment-effect differences: average accuracy was .54 (SD = .40) for GLMM trees and .54 (SD = .43) for GLMMs.
 The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (continuous, piecewise or both).
 These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.
+
 As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely continuous interactions.
-Furthermore, performance of GLMM tree is hardly affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases.
-This indicates that GLMM tree may be especially useful for exploratory purposes.
-
-\FloatBarrier
+Performance of both methods improves with increasing sample size.
+Furthermore, performance of GLMM tree is not affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous.
+This indicates that GLMM tree may be especially useful for exploratory purposes, where there are many potential moderator variables.
+Also, GLMM trees may provide simpler models, as the average number of fixed-effects predictor variables (main effects and interactions) evaluated for prediction with GLMMs was 12.30, whereas the average number of inner nodes that need to be evaluated for prediction with GLMM tree was only 3.38.
 
@@
 Confidence intervals for Cohen's $d$ could not be calculated, as these can not take into account the exploratory nature of our analyses (i.e., variable and split point selection).
-The predictive accuracy of GLM and GLMM trees was assessed by calculating average correlations between observed and predicted HAM-D posttreatment scores, based on 50-fold cross validation.
+
+To compare the two tree-based approaches with a more classical approach of interaction detection, we also fitted a GLMM with pre-specified interactions on the dataset.
+In the GLMM, the posttreatment HAM-D scores were controlled for the linear effects of pretreatment HAM-D scores, and then regressed on a random intercept, main effects of treatment and the potential moderators (partitioning variables), and interactions between treatment and the potential moderators.
+As it is not known in advance how to interact the potential moderators, higher-order interactions were not included.
+
+The predictive accuracy of the GLM and GLMM trees and the GLMM was assessed by calculating average correlations between observed and predicted HAM-D posttreatment scores, based on 50-fold cross validation.
 
@@
 Therefore, we used the R~package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values.
-Using the \verb|stabletree| function, we calculated the number of times a variable and value were selected for partitioning, over 500 subsamples of size $.9 \times N$ of the dataset.
+Using the \verb|stabletree| function, we calculated the number of times a variable and value were selected for partitioning over 500 subsamples of size $.9 \times N$ of the dataset.
 
@@
 
-Assessment of predictive accuracy by means of 50-fold cross validation indicated better predictive accuracy for GLMM tree than GLM tree.
-The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathit{var}=.067$) for GLMM tree, and .19 ($\mathit{var}=.084$) for GLM tree.
-This indicates that GLMM tree provided higher predictive accuracy, and also somewhat lower variability of predictive accuracy than GLM tree.
+The GLMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: b = 2.29, p = .002; interaction with treatment: b = -2.10, p = .028).
+Also, the GLMM indicated an interaction between treatment and age (b = .10, p = .018).
+
+Assessment of predictive accuracy by means of 50-fold cross validation indicated better predictive accuracy for GLMM tree than for GLM tree and the GLMM with pre-specified interactions.
+The correlation between true and predicted posttreatment HAM-D total scores averaged over 50 folds was .272 ($\mathit{var}=.067$) for GLMM tree, .233 ($\mathit{var}=.064$) for the GLMM with prespecified interactions and .190 ($\mathit{var}=.084$) for GLM tree.
 
@@
 
-Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided similar predictive accuracy, on average.
-When interactions were purely continuous, GLMMs outperformed GLMM trees, and when interactions were purely piecewise, GLMM trees outperformed GLMMs.
-However, GLMM tree may have a clear advantage when there are a large number of potential moderator variables (i.e., $> 5$).
-With 10 potential moderator variables, $t$-tests of GLMMs were unable to identify the true moderator variables, whereas the number of potential moderator variables did not influence performance of GLMM tree.
-Therefore, our results indicate that GLMM trees are better suited than GLMMs for exploratory analyses, in which moderator variables need to be selected from a larger number of covariates.
-Furthermore, the number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex predictive models.
-The trees in our simulations were limited to a maximum number of 8 terminal nodes, and may therefore be much easier to use for prediction in practical decision-making contexts.
+Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided somewhat better accuracy, on average.
+When interactions were purely continuous, GLMMs outperformed GLMM trees, and when interactions were partly or purely piecewise, GLMM trees outperformed GLMMs.
+In addition, GLMM tree may have a clear advantage when there are a large number of potential moderator variables (i.e., $> 5$).
+Therefore, GLMM trees may be better suited than GLMMs for exploratory analyses, in which moderator variables need to be selected from a larger number of covariates.
+Furthermore, the number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models.
+The trees in our simulations were limited to a maximum number of 7 inner nodes, and may therefore be much easier to use for prediction in practical decision-making contexts.
 

### 2016-11-07 pkg/glmertree/paper/Simulation replication/Datasets without treat-sub interactions/Analysis of datasets + evaluation.R

@@ -1,9 +1,8 @@
 # install.packages("glmertree", repos="http://R-Forge.R-project.org") 
 library(glmertree)
-names(datasets[[1]])
 
 # Fit trees on training data
-for (c in 1:1) {
+for (c in 1:50) {
   load(paste("datasets", c, sep=""))
   load(paste("descriptions", c, sep=""))
   GLMMtrees <- list()
@@ -65,7 +64,6 @@ for (i in 1:length(descriptions)) {
   treesizes[i,"numbclus"] <- substr(descriptions[[i]][[6]], 34, 35) # is number of random intercept values
   treesizes[i,"sigmabi"] <- substr(descriptions[[i]][[7]], 12, 14) # max and -min random intercept value    
 }
-save(treesizes,file="treesizes")
 
 treesizes$N <- factor(treesizes$N)
 treesizes$rho <-  factor(treesizes$rho)            
@@ -83,7 +81,6 @@ treesizes$corUbi <- rep(treesizes$corUbi[1:length(descriptions)], times=1)
 treesizes$numbclus <- rep(treesizes$numbclus[1:length(descriptions)], times=1)
 treesizes$sigmabi <- rep(treesizes$sigmabi[1:length(descriptions)], times=1)
 
-treesizes
 table(treesizes[,1]) # GLMM trees
 prop.table(table(treesizes[,1])) # GLMM trees
 table(treesizes[,2]) # GLM trees
@@ -99,12 +96,16 @@ tmp <- as.character(treesizes.long$treesize.ind)
 tmp[treesizes.long$treesize.ind=="GLMMtreesize"] <- "GLMM tree"
 tmp[treesizes.long$treesize.ind=="GLMtreesize"] <- "GLM tree"
 treesizes.long$treesize.ind <- factor(tmp)
-
 mean(treesizes[,1]);sd(treesizes[,1]) # GLMM tree
 mean(treesizes[,2]);sd(treesizes[,2]) # GLM tree
 
+save(treesizes.long, file="noint_treesizes_long.dat")
+load("noint_treesizes_long.dat")
+
 # create lattice xyplots
 library(lattice)
+
+## ANOVA for treesize
 treesize.anova <- aov(treesize.values ~ treesize.ind + N + rho + np + treatdiff + numbclus + sigmabm + corUbm +
                         treesize.ind*(N + rho + np + treatdiff + numbclus + sigmabm + corUbm), 
                       data=treesizes.long)
[…]

### 2016-12-17 pkg/glmertree/paper/treatment_interactions.Rnw
Minor paper updates and updated simulation files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1054 255f4249-2263-4417-b1ea-8498779e27a7

@@
 GLMM tree also outperformed trees without random effects in terms of predictive accuracy and Type-I error rates.
-Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed better predictive accuracy.
-We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression.
+Compared to GLMMs with pre-specified interaction effects, GLMM tree showed better predictive accuracy, on average.
+We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression.
 We conclude that GLMM tree is a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
@@
 As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions.
-Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview).
+Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov16, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview).
 Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.
@@
 For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}.
-In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}), GLMM tree allows for taking into account the clustered structure of datasets.
+In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}), GLMM tree allows for taking into account the clustered structure of datasets.
 In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.
 
-In what follows, we first introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree.
+In what follows, we first introduce GLM-based MOB and the GLMM tree algorithm.
 In the Simulation-based Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects.
@@
 Finally, in the Discussion we summarize the results and point out some directions for future research.
+But first, we will introduce an artificial motivating dataset, which will be used to illustrate treatment-subgroup interaction detection with GLM-based MOB and GLMM tree.
 
@@
 We created a dataset representing observations on 150 participants in a randomized clinical trial.
-Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the pos-ttreatment total score on a depression inventory.
+Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the post-treatment total score on a depression inventory.
 For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).
@@
 The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$.
-In this subgroup, Treatment~1 is more effective than Treatment~2: the mean of the response variable is 7 for Treatment~1, and 11 for Treatment~2.
+In this subgroup, Treatment~1 is more effective than Treatment~2.
 The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$.
-In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment~1, and 9 for Treatment~2.
+In this subgroup, both therapies are equally effective.
 The third subgroup consists of observations with duration $>6$.
-In this subgroup, Treatment~2 is more effective than Treatment~1: the mean value of the response variable is 12 for Treatment~1, and 7 for Treatment~2.
+In this subgroup, Treatment~2 is more effective than Treatment~1.
 
@@
 @
-
-\begin{figure}[t!]
-\centering
-<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
-plot(exampleglmmtree)
-@
-\caption{GLMM tree of the motivating example dataset.
-Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
-\label{fig:example_glimmertree}
-\end{figure}
 
@@
 %
-For a simple random intercept $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
+For a simple random intercept, $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
 Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$.
@@
 
+\begin{figure}[t!]
+\centering
+<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
+plot(exampleglmmtree)
+@
+\caption{GLMM tree of the motivating example dataset.
+Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
+\label{fig:example_glimmertree}
+\end{figure}
+
 \section{Simulation-based evaluation}
@@
 To assess the performance of GLMM tree, we carried out three simulation studies: In Studies I and II, we compared the performance of GLMM tree with that of GLM tree.
-In Study I, we assessed the accuracy of both tree-based algorithms in datasets with treatment-subgroup interactions.
-In Study II, we assessed the Type-I error of both tree-based algorithms in datasets without treatment-subgroup interactions.
+In Study I, we assessed the accuracy of the two tree algorithms in datasets with treatment-subgroup interactions.
+In Study II, we assessed the Type-I error of the two tree algorithms in datasets without treatment-subgroup interactions.
 In Study III, we compared the performance of GLMM tree in detecting treatment interactions with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.
@@
 
-In all simulation studies, we varied the following data-generating parameters:
+In all simulation studies, the following data-generating parameters were varied:
 
@@
 \item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
-\item Two levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates.  
+\item Two levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
 \end{enumerate}
-
-In addition, each simulation study required a different design for main and interaction effects, which will be described in their respective sections below.
 
@@
 All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$.
-Correlations between $U$ variables vary according to the third facet of the data-generating design described above.
 
@@
 In this simulation study, some of the potential partitioning covariates were true partitioning covariates ($U_1$, $U_2$ and $U_5$), whereas others were not (e.g., $U_3$ and $U_4$).
-Therefore, an extra level was added to the sixth facet of the \textit{General simulation design} described above.
-Also, to assess the effect of the size of the treatment-effect differences in nodes 3 and 7, an additional facet was added:
+Therefore, an extra level was added to the sixth facet of the \textit{General simulation design}:
 
@@
 \item[6.] Three levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the true partitioning covariates ($U_1$, $U_2$ or $U_5$), $b$ correlated with one of the noise variables ($U_3$ or $U_4$.
+\end{enumerate}
+
+Also, to assess the effect of treatment-effect differences in nodes 3 and 7, an additional facet was added:
+
+\begin{enumerate}
 \item[7.] Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes.
@@
 
-Again, 50 datasets were generated for each cell of the design.
-As a result, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2 = 32,400$ datasets were generated.
+50 datasets were generated for each cell of the design, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2 = 32,400$ datasets were generated.
 In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.
@@
 
-To assess the accuracy of trees recovered by both algorithms, we inspected variables and values that were selected for partitioning.
 For the first split, GLMM tree selected the true partitioning variable ($U_2$) in all datasets, and GLM tree in all but one datasets.
@@
 
-The total number of nodes in estimated GLM and GLMM tree were calculated.
-A tree of size $>1$ was classified as a Type-I error.
-To identify the most important predictors of Type-I error, we used a GLMs with algorithm type and the parameters of the data-generating process as independent variables.
+The total number of nodes in estimated GLM and GLMM trees were calculated and a tree of size $>1$ was classified as a Type-I error.
+To identify the most important predictors of Type-I error, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables.
 First-order interactions between algorithm type and each of the data-generating parameters were also included.
@@
 The same goes for GLM tree, whenever the random intercept has 0 variance, or the random intercept is not correlated to one of the partitioning covariates.
-However, when the random intercept is correlated to one of the potential partitioning covariates, the type-I error rapidly increases for GLM tree.
+However, when the random intercept is correlated with one of the potential partitioning covariates, the type-I error rapidly increases for GLM tree.
 With large sample sizes and large variance of the random intercept, GLM tree will almost certainly create spurious splits.
@@
 For datasets with purely continuous interactions, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$.
-Here, $\beta$ has a global value and no subscript, and comprised purely continuous main and (treatment) interaction effects of the moderator variables $U$, as described in Table \ref{tab:continuous_terms}.
+Here, $\beta$ has a global value and no subscript, comprising purely continuous main and (treatment) interaction effects of the moderator variables $U$, as described in Table \ref{tab:continuous_terms}.
 
@@
 Continuous predictor variables were centered by subtracting the mean value, before calculating and including the interaction term in the GLMM.
-Predictions were obtained by refitting a GLMM on the training data, using only the predictors with p-values $< .05$ in the original GLMM.
-Predictions for test observations were calculated using the refitted GLMMs.
 
@@
 Predictive accuracy was assessed in terms of the correlation between the true and predicted treatment-effect differences in test datasets.
+As the full GLMM models were likely to overfit, predictions for GLMMs with pre-specified interactions were obtained by refitting a GLMM on the training data, using only the predictors with p-values $< .05$ in the original GLMM.
+Predictions for test observations were obtained from these refitted GLMMs.
+
 To identify the most important predictors of predictive accuracy, we used an ANOVA with algorithm type and the parameters of the data-generating process as independent variables.
@@
 
-GLMM tree showed somewhat higher average correlations between the true and predicted treatment-effect differences: average accuracy was .54 (SD = .40) for GLMM trees and .54 (SD = .43) for GLMMs.
-The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (continuous, piecewise or both).
-These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.
+GLMM tree showed somewhat higher average correlations between the true and predicted treatment-effect differences: average accuracy was .54 (SD = .40) for GLMM trees and .51 (SD = .43) for GLMMs.
+The most important predictors of predictive accuracy were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.
 
 As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely continuous interactions.
+GLMMs perform poorly when interactions are not purely piecewise, and GLMM tree performace poorly when interactions are purely linear.
+
 Performance of both methods improves with increasing sample size.
@@
 This indicates that GLMM tree may be especially useful for exploratory purposes, where there are many potential moderator variables.
-Also, GLMM trees may provide simpler models, as the average number of fixed-effects predictor variables (main effects and interactions) evaluated for prediction with GLMMs was 12.30, whereas the average number of inner nodes that need to be evaluated for prediction with GLMM tree was only 3.38.
+
+GLMM trees may also provide simpler models, as the GLMMs included 12.30 significant predictor variables (main effects and interactions), on average.
+GLMM trees had 3.38 inner nodes on average, requiring less variables to be evaluated for making predictions.
 
@@
 As such, we hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies.
-We performed simulation studies to assess the accuracy of GLMM tree in detecting treatment-subgroup interactions and illustrated the application of GLMM tree on an existing dataset.
+Also, we illustrated the application of GLMM tree on an existing dataset.
 
@@
 GLMM tree also provided better predictive accuracy than GLM tree, with an average correlation between true and predicted treatment differences of .94 for GLMM tree, and .88 for GLM tree.
-GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, differences in treatment effects across terminal nodes were relatively small (i.e., Cohen's $d=.5$), and/or sample size was small (i.e., $ N < 1,000$).
+GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, differences in treatment effects across terminal nodes were relatively small (Cohen's $d=.5$), and/or sample size was small ($< 1,000$).
 Such treatment-effect differences and sample sizes may be quite common, even in multi-center clinical trials, and GLMM tree may provide a helpful tool for subgroup detection in those instances.
@@
 Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided somewhat better accuracy, on average.
-When interactions were purely continuous, GLMMs outperformed GLMM trees, and when interactions were partly or purely piecewise, GLMM trees outperformed GLMMs.
-In addition, GLMM tree may have a clear advantage when there are a large number of potential moderator variables (i.e., $> 5$).
+GLMM tree performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at leas partially piecewise.
+Our findings also indicated a clear advantage for GLMM tree when there are a large number of potential moderator variables (i.e., $> 5$).
 Therefore, GLMM trees may be better suited than GLMMs for exploratory analyses, in which moderator variables need to be selected from a larger number of covariates.
@@
 Node-specific effect sizes can also be used to prune trees, when a researcher prefers to have a final tree which is based on statistical as well as clinical significance.
-A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significant in the tree-growing process.
+A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significance in the tree-growing process.
 

### 2016-12-17 pkg/glmertree/paper/Application with CV.R

@@ -1,5 +1,7 @@
 library(foreign)
 library(glmertree)
+library(lmerTest)
+source("C:\\Users\\tobii\\Desktop\\swReg\\glmertree\\R\\glmertree.R")
 
 # DATA PREPARATION
 metadata <- read.dta("Database IPDMA CBT PHA Version 11.dta")
@@ -32,6 +34,15 @@ lmer_f <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(HRSDfit) |
 plot(lm_f); print(lm_f)
 plot(lm_o); print(lm_o)
 plot(lmer_f); print(lmer_f)
+
+## Fit GLMM wit h pre-specified interactions:
+metadata2 <- metadata
+metadata2$education <- as.numeric(metadata2$education)
+GLMM <- lmer(HRSDt1 - HRSDfit ~ (1 | studyid) +  Tx_group*(Age + Gender + education + 
+           ComorbidAnxietyDisorder + HRSDt0), data = metadata2)
+summary(GLMM)
+
+
 
 ## calculate predictions
 metadata$lmtreepred <- predict(lm_f, newdata = metadata)
@@ -82,6 +93,10 @@ testdata <- list()
 traindata <- list()
 lmertrees <- list()
 lmtrees <- list()
+GLMMs <- list()
+GLMMs2 <- list()
+traindata2 <- list()
+testdata2 <- list()
 for (i in 1:50) {
   print(i)
   trainids <- samplething$sample.index[[i]]
@@ -92,35 +107,71 @@ for (i in 1:50) {
   traindata[[i]]$HRSDfit <- fitted(lm(HRSDt1 ~ HRSDt0, data = traindata[[i]]))
   testdata[[i]]$HRSDfit <- predict(lm(HRSDt1 ~ HRSDt0, data = traindata[[i]]), newdata = testdata[[i]])
 
+  # fit GLMM:
+  traindata2[[i]] <- traindata[[i]]
+  traindata2[[i]]$education <- as.numeric(traindata2[[i]]$education)
+  testdata2[[i]] <- testdata[[i]]
+  testdata2[[i]]$education <- as.numeric(testdata2[[i]]$education)
+  
+  traindata2[[i]] <- data.frame(studyid = traindata2[[i]]$studyid, 
+                                HRSDt1 = traindata2[[i]]$HRSDt1,
+                                HRSDfit = traindata2[[i]]$HRSDfit,
+                                model.matrix(~ Tx_group*(Age + Gender + education + 
[…]

### 2017-04-10 pkg/devel/partykit/R/modelparty.R
minsize default as in old mob
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1234 255f4249-2263-4417-b1ea-8498779e27a7

@@ -120,7 +120,7 @@ mob_control <- function(
   minbucket = minsize,
   minprob = 0.01,
   minsplit = minsize,
-  minsize = 20L,
+  minsize = NULL,
   stump = FALSE,
   caseweights = TRUE,
   dfsplit = TRUE,
@@ -186,6 +186,12 @@ mob_control <- function(
       stop("mincriterion = 1 - alpha must be between 0 and 1.")
   }
   
+  ### make sure right criterion is used for exhaustive search
+  if(testflavour == "exhaustive"){
+    criterion <- "statistic"
+    logmincriterion <- -Inf
+  }
+  
   c(.urp_control(criterion = criterion,
                  logmincriterion = log(mincriterion), minsplit = minsplit, 
                  minbucket = minbucket, minprob = minprob, nmax = Inf, 
@@ -220,10 +226,18 @@ mob <- function
   ...
 ) {
 
-  ### make sure right criterion is used for exhaustive search
-  if(control$testflavour == "exhaustive"){
-    control$criterion <- "statistic"
-    control$logmincriterion <- -Inf
+
+  
+  ### if minsize is NULL, set to 10 * number of parameters
+  if (is.null(control$minbucket) | is.null(control$minsplit)) {
+    Fmla <- as.Formula(formula)
+    n_coef <- length(attr(terms(Fmla, lhs = 0, rhs = 1), "term.labels"))
+    if(n_coef == 0) n_coef <- 1
+    n_y <- length(attr(terms(Fmla, lhs = 1, rhs = 0), "term.labels"))
+    if(n_y == 0) n_y <- 1
+    minsize <- as.integer(ceiling(10L * n_coef/n_y))
+    if (is.null(control$minbucket)) control$minbucket <- minsize
+    if (is.null(control$minsplit)) control$minsplit <- minsize
   }
   
   ### get the call and the calling environment for .urp_tree

### 2017-04-24 pkg/glmertree/paper/treatment_interactions.Rnw
Paper update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1262 255f4249-2263-4417-b1ea-8498779e27a7

@@
-\documentclass[nobf,doc]{apa}
+\documentclass[doc]{apa}
 \usepackage{amssymb}
@@
 \usepackage{tabto}
+\usepackage{Sweave}
+\usepackage{subfig}
 
@@
 
-\acknowledgements{The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application.
+\acknowledgements{The authors would like 
+to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application.
 The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014.
@@
 
+\headinglevels{four}
+
 \begin{document}
@@
 Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine.
-Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure.
+Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research.
 Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset.
-The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters.
-In a simulation study, we evaluate the performance of GLMM trees and compare it with that of trees without random effects, and GLMMs with pre-specified interactions.
-GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets.
-GLMM tree also outperformed trees without random effects in terms of predictive accuracy and Type-I error rates.
-Compared to GLMMs with pre-specified interaction effects, GLMM tree showed better predictive accuracy, on average.
+The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters.
+In a short tutorial, we show how GLMM trees can be fit with the R package \verb|glmertree|.
+In a simulation study, GLMM tree showed higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees.
+Also, GLMM tree showed somewhat higher predictive accuracy than linear mixed-effects model with pre-specified interaction effect, on average.
 We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression.
@@
 
-In research on the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified (or personalized) medicine is becoming increasingly important.
+In research on the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized or stratified medicine is becoming increasingly important.
 Stratified medicine presents the challenge of finding which patients respond best to which treatments.
@@
 
-Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments.
-Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
-As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions.
+Tree-based methods are such hypothesis-generating methods.
+Tree-based methods, also known as recursive partitioning methods, recursively split the space spanned by the predictor variables into rectangular regions, containing observations that are increasingly similar with respect to the outcome.
+Several tree-based methods take the mean or majority class of a single dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}.
+Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM approach of \citeA{Ciam91} is the earliest example.
+
+Due to the recursive nature of the splitting, the rectangular regions of the partition can be graphically depicted as nodes in a decision tree, as shown in the example in Figure \ref{fig:exampletree}.
+The partition in Figure \ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety.
+The resulting tree has a depth of two, as the longest path travels along two splits.
+Each of the splits in the tree is defined by a splitting variable and value.
+The first split in the tree separates the observations into two subgroups, based on the splitting variable duration and a splitting value of 8, yielding two rectangular regions, represented by node 2 and node 3.
+As the observations in node 2 are not further split, node 2 is a terminal node.
+Node 3 is an inner node, as the observations in this node are further split into nodes 4 and 5, based on the anxiety variable.
+
+If the partition in Figure \ref{fig:exampletree} would be used for prediction of a new observation, the new observation would be assigned to one of the terminal nodes according to its values on the splitting variables.
+The prediction is then determined by the distribution of the training observations within the terminal node.
+For example, the prediction may be the node-specific mean of a single continuous variable.
+In the current paper, we focus on trees where the terminal nodes consist of a linear (LM) or generalized linear model (GLM), in which case the predicted value for a new observation is determined by the node-specific parameter estimates of the (G)LM.
+
+\begin{figure}[tb!]
+\setkeys{Gin}{width=0.5\textwidth}
+\centering
+\subfloat[Rectangular areas]{
+<<example_decision_tree2, fig = TRUE>>=
+options(prompt = " ", continue = " ")
+plot(c(0, 18), c(0, 18), type = "n", xlab = "duration", ylab = "anxiety")
+abline(v = 8, h = NA)
+lines(x = c(-1, 8), y = c(10,10))
+text(label = "node 3", x = 3.5, y = 4.5)
+text(label = "node 4", x = 3.5, y = 14)
+text(label = "node 5", x = 13.5, y = 9)
+@
+}
+\subfloat[Decision tree]{
+<<example-decision-tree, fig=TRUE>>=
+fig1 <- party(
+  partynode(1L,
+    split = partysplit(1L, breaks = 8),
+      kids = list(
+        partynode(3L, 
+          split = partysplit(2L, breaks = 10),
+          kids = list(
+            partynode(4L, info = ""),
+            partynode(5L, info = ""))),
+  partynode(2L, info = ""))),
+  data.frame(duration = numeric(0), anxiety = numeric(0))
+)
+plot(fig1)
+@
+}
+\caption{\small Example of a recursive partition. In the left panel, the partition is depicted as a set of rectangular areas. In the right panel, the same partition is depicted as a decision tree.}
+\label{fig:exampletree}
+\end{figure}
+
+Tree-based methods are particularly useful for exploratory purposes, because they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
+As such, they are preeminently suited to the detection of treatment-subgroup interactions.
 Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov16, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview).
-Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.
-
-In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model.
-For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}.
+Also, \citeA{ZhanyTsia12Bio} and \citeA{ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.
+
+In many instances, researchers may want to detect treatment-subgroup interactions in clustered or nested datasets.
+For example, in individual-level patient data meta-analyses, where datasets of multiple clinical trials on the same treatments are pooled.
 In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}.
-In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}.
+In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92}.
 For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}.
 However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset.
-Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model tree, or GLMM tree.
+Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of interactions and non-linearities in GLMM type models: generalized linear mixed-effects model tree, or GLMM tree.
 
@@
 For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}.
-In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}), GLMM tree allows for taking into account the clustered structure of datasets.
+In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}, GLMM tree allows for taking into account the clustered structure of datasets.
 In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.
 
-In what follows, we first introduce GLM-based MOB and the GLMM tree algorithm.
-In the Simulation-based Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects.
-The comparison with GLM-based MOB allows for assessing the extent to which including random-effects estimation improves the performance of MOB trees.
-The comparison with GLMMs with pre-specified interactions will allow for assessing the extent to which using a tree-based method will improve detection of treatment-subgroup interactions.
-In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression.
-Finally, in the Discussion we summarize the results and point out some directions for future research.
-But first, we will introduce an artificial motivating dataset, which will be used to illustrate treatment-subgroup interaction detection with GLM-based MOB and GLMM tree.
-
-\subsection{Artificial motivating dataset}
+The remainder of this paper is structured into four sections: In the first section, we introduce the GLMM tree algorithm using an artificial motivating dataset with treatment-subgroup interactions.
+The second section is a short tutorial on how to fit GLMM trees using the R package \verb|glmertree|.
+In the third section, we compare the performance of GLMM tree with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions.
+In the fourth section, we apply GLMM tree to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression.
+In the fith and last section we summarize the results and discuss limitations and directions for future research.
+
+\section{GLMM tree algorithm}
+
+\subsubsection{Artificial motivating dataset}
+
+We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. The R code used to generate the example dataset is provided in the Appendix.
+
+The dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2).
+The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology.
+The potential moderator variables are duration, age and anxiety.
+Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment.
+Finally, each patient was part of one of ten clusters, each having a different value for the random intercept, uncorrelated to the partitioning variables.
+
+The outcome variable was generated such that there were three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure \ref{fig:exampletree}: The first subgroup of patients corresponds to node~3 and shows lower post-treatment depression in Treatment~1 than in Treatment~2.
+The second subgroup of patients corresponds to node~4 and shows about equal post-treatment depression in both treatment conditions.
+The third subgroup of patients corresponds to node~5 and shows lower post-treatment depression in Treatment~2 than in Treatment~1.
+Thus, duration and anxiety are true partitioning or moderator variables, whereas age is not.
+Anticipating the final results of our analyses, the treatment-subgroup interactions are depicted in Figure~\ref{fig:example_glimmertree}, which shows the GLMM tree that accurately recovered the treatment-subgroup interactions.
 
@@
 set.seed(123)
-treatment <- rbinom(n=150,size=1,prob=1/2)
-duration <- round(matrix(rnorm(150,mean=5,sd=3),ncol=1))
-duration[duration<0] <- 1
-anxiety <- round(rnorm(150,mean=10,sd=3))
-age <- round(rnorm(150,mean=30,sd=10))
-age[age<18] <- 18
-age[age>75] <- 75
-cutscore1 <- 5+.44*3
-cutscore2 <- 10
-node1t1 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==0, 1, 0)
-node1t2 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==1, 1, 0)
-node2t1 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==0, 1, 0)
-node2t2 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==1, 1, 0)
-node3t1 <- ifelse(duration>cutscore1 & treatment==0, 1, 0)
-node3t2 <- ifelse(duration>cutscore1 & treatment==1, 1, 0)
-beta1 <- -2
-beta2 <- 0
-beta3 <- 2.5
-error <- rnorm(150,0,2)
-cluster <- error+rnorm(150, 0, 6)
-int_values <- rnorm(10, 0, 1)
-tmp <- rep(int_values,each=15)
-tmp <- tmp[order(tmp)]
-rand_int <- rep(NA, times=150)
-rand_int[order(cluster)] <- tmp
-cluster[order(cluster)] <- rep(1:10, each=15)
-cor(error,rand_int)
-true_error <- error - rand_int
-true_eff <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
-     + .4*treatment)
-outcome <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
-                      + .4*treatment + true_error + rand_int)
-outcome[outcome<0] <- 1
-example <- data.frame(treatment, outcome, duration, anxiety, cluster, age)
-example$treatment <- factor(example$treatment, labels = c("Treatment 1", "Treatment 2"))
-@
-
-We created a dataset representing observations on 150 participants in a randomized clinical trial.
-Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the post-treatment total score on a depression inventory.
-For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).
-
-The simulated dataset has 3 subgroups with differential treatment effectiveness.
-The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$.
-In this subgroup, Treatment~1 is more effective than Treatment~2.
-The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$.
-In this subgroup, both therapies are equally effective.
-The third subgroup consists of observations with duration $>6$.
-In this subgroup, Treatment~2 is more effective than Treatment~1.
-
-<<example-trees>>=
-exampleglm <- lmtree(outcome ~ treatment | age, data = example)
-exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = example)
-exampleglmmtree <- lmertree(outcome ~ treatment | cluster | age + duration + anxiety, data = example)
-@
-
-Participants were part of one of ten clusters, each with a different value for the random intercept.
-Data were generated such that covariates and cluster-specific intercepts were uncorrelated.
-43\% of variance in post-treatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.
-
-See Figure~\ref{fig:example_glimmertree} for the GLMM tree fitted to the simulated data set which recovers the true underlying structure.
-
-\subsection{Model-based recursive partitioning}
-
-The rationale behind MOB is that a global parametric model may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
-This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different.
-However, CART trees detect differences in the mean (or median) value across terminal nodes, whereas MOB trees detect differences in parameters of more complex models across terminal nodes.
-
-For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset.
-The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function:
+## Generate treatment indicator and covariates:
+treatment <- rbinom(n = 150, size = 1, prob = .5)
+duration <- round(rnorm(150, mean = 7, sd = 3))
+anxiety <- round(rnorm(150, mean = 10, sd = 3))
+age <- round(rnorm(150, mean = 45, sd = 10))
+## Generate tree structure:
+node1t1 <- ifelse(duration <= 8 & anxiety <= 10 & treatment == 0, -2, 0)
+node1t2 <- ifelse(duration <= 8 & anxiety <= 10 & treatment == 1, 2, 0)
+node3t1 <- ifelse(duration > 8 & treatment == 0, 2.5, 0)
+node3t2 <- ifelse(duration > 8 & treatment == 1, -2.5, 0)
+## Generate error:
+error <- rnorm(150, 0, 2)
+## Generate cluster indicators and random intercepts:
+cluster <- error + rnorm(150, 0, 6)
+rand_int <- sort(rep(rnorm(10, 0, 1), each = 15))
+rand_int[order(cluster)] <- rand_int 
+cluster[order(cluster)] <- rep(1:10, each = 15)
+error <- error - rand_int
+## Generate outcome variable:
+outcome <- round(9 + node1t1 + node1t2 + node3t1 + node3t2 + .4*treatment + error + rand_int)
+## Combine into dataset:
+treatment <- factor(treatment, labels = c("Treatment 1", "Treatment 2"))
+exampledata <- data.frame(outcome, duration, anxiety, cluster, age, treatment)
+## Grow trees:
+exampleglm <- lmtree(outcome ~ treatment | age, data = exampledata)
+exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = exampledata)
+exampleglmmtree <- lmertree(outcome ~ treatment | cluster | age + duration + anxiety, data = exampledata)
+examplelmm <- lmer(outcome ~ treatment + (1|cluster) + age + duration + anxiety + treatment*(age + duration + anxiety), data = exampledata)
+@
+
+\subsubsection{Model-based recursive partitioning}
+
+The rationale behind MOB is that a global parametric model may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find better-fitting models in each cell of the partition.
+For example, to assess the effect of treatment, we may fit a global GLM using only the treatment indicator as a predictor of the outcome.
+The expectation $\mu_i$ of outcome $y_i$ given the treatment regressor $x_i$ is then modeled through a linear predictor and suitable link function:
 %
@@
 where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function.
-$\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.
-Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper.
-Also, for the continuous response variable in the motivating data set, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.
+$\beta$ is a vector of fixed-effects regression coefficients.
+For simplicity, in the current paper we assume only two treatment groups and no further covariates in the GLM, so $x_{i}$ and $\beta$ each have length 2.
+The first element of $\beta$ then corresponds to the mean of the linear predictor in the first treatment group and the second element corresponds to the mean difference in the linear predictor between the first and second treatment groups.
+For the continuous response variable in the motivating data set, we employ the identity link function and assume a normal distribution for the error (denoted by $\epsilon_i = y_i - \mu_i$) with mean zero and variance $\sigma_{\epsilon}^2$.
 However, the model can easily accommodate additional treatment conditions and covariates, and binary or count outcome variables.
 
-\begin{figure}[t!]
+\begin{figure}[tb!]
 \centering
@@
 @
-\caption{Example of a global GLM for treatment outcomes, based on the artificial motivating dataset ($N=150$).
-The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
+\caption{\small Example of a global GLM for treatment outcomes, based on the artificial motivating dataset ($N=150$).
+The y-axis represent treatment outcome (post-treatment depression).
+The dot for Treatment~1 represents the first and the slope of the regression line represents the second element of $\beta$.}
 \label{fig:fixedeffects}
@@
 
-The result of fitting a global GLM to the motivating dataset is depicted in Figure~\ref{fig:fixedeffects}; the boxplots show the distribution of the posttreatment depression scores in both treatment groups.
-The global model does not describe the data well: there is substantial residual variance and the slope of the regression line is nearly zero.
-This does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.
-
-The MOB algorithm can be used to detect such moderation, by testing for parameter stability over a set of auxiliary covariates, also known as \textit{partitioning variables}.
+Obviously, such a simple, global GLM will not fit the data well, especially in the presence of moderators.
+For illustrational purposes, however, we take it as a starting point to illustrate MOB.
+The global GLM fitted to the motivating example dataset is depicted in Figure~\ref{fig:fixedeffects}.
+As the boxplots show, there is little difference between the global effects of the two treatments and there is considerable residual variance.
+
+The MOB algorithm can be used to partition the dataset using additional covariates and find better fitting local models.
+To this end, the MOB algorithm tests for parameter stability over a set of auxiliary covariates, also known as \textit{partitioning variables}.
 When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor.
-To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.
-
-More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function.
-When the model is correctly specified, the expected value of the score for each observation equals zero.
-Therefore, under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}.
-To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.
-
-If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a $p$-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3).
-This partition is created using $U_{{k^*}}$, the variable with the minimal $p$-value in step (2).
-The split point for $U_{{k^*}}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}.
+To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) statistically test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.
+
+In step (2) a test statistic quantifying parameter instability is calculated for every potential partitioning variable.
+As the distribution of these test statistics under the null hypothesis of parameter stability is known, a $p$-value for every partitioning variable can be calculated.
+Note that a more in-depth discussion of the parameter stability tests is outside the scope of this paper, but can be found in \citeA{ZeilyHorn07} and \citeA{ZeilyHoth08}.
+
+If at least one of the partitioning variables yields a $p$-value below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3).
+This partition is created using $U_{{k^*}}$, the partitioning variable with the minimal $p$-value in step (2).
+The split point for $U_{k^*}$ is selected by taking the value that minimizes the sum of the values of the objective function in both partitions.
+In other words, for every possible split point, the objective function is minimized separately in the two resulting subsets of the data; the split point yielding the minimum sum of the objective functions is selected.
 In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).
 
-\begin{figure}[t!]
+\begin{figure}[tb!]
 \centering
@@
 @
-\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset.
-Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
+\caption{\small GLM tree grown on the artificial motivating dataset.
+Three additional covariates (pre-treatment anxiety, duration and age) were used as potential splitting variables.
+The y-axes in the terminal nodes represent the treatment outcome (post-treatment depression severity).}
 \label{fig:example_mobtree}
@@
 
-Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree.
+The partition resulting from application of MOB can be depicted as a decision tree.
 If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
@@
 %
-Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all $x$-variables interact with the subgroup indicator.
-Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
-The corresponding $\beta^{*}$ would have length $2J$ as well, containing subgroup-specific fixed-effects coefficients.
-
-Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the motivating dataset.
-By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
-Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above.
-However, the two subgroups in the middle do not represent true subgroups, which is due to the clustered structure of the dataset not being taken into account.
-
-\subsection{Generalized linear mixed-effects model trees}
-
-For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate.
-The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
+
+To illustrate, we fitted a GLM tree on the artificial motivating dataset.
+In addition to the treatment indicator and treatment outcome used to fit the earlier GLM, we specified the anxiety, duration and age variables as potential partitioning variables.
+Figure~\ref{fig:example_mobtree} shows the resulting GLM tree.
+MOB partitioned the observations into four subgroups, each with a different estimate $\beta_j$.
+Age was correctly not identified as a partitioning variable and the left- and rightmost nodes are in accordance with the true treatment-subgroup interactions described above.
+However, the two nodes in the middle do not represent true subgroups, possibly due to the dependence of observations within clusters not being taken into account.
+
+\subsubsection{Including random effects}
+
+For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a mixed-effects model would be more appropriate.
+The GLM in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
 %
@@
 %
-For a simple random intercept, $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
-Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$.
-For simplicity we only employ a cluster-specific intercept here, but further random effects can be easily included in $z_i$.
+For a random-intercept only model, $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
+Further, $b$ is a random vector of length $M$, each $m$-th element corresponding to the random intercept for cluster $m$.
+For simplicity, we employ a cluster-specific intercept only, but further random effects can easily be included in $z_i$.
 Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
+Also, the errors $\epsilon$ are assumed to follow a normal distribution, with constant variance across clusters.
 The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).
 
-Note that, if the random-effects coefficients were known, the model could be estimated by a simple GLM as in Equation \ref{eq:fixedeffects} where $z_{i}^{\top}b$ would only be added as an offset (i.e., a variable with a fixed coefficient of 1) to the linear predictor.
-
-Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part may not describe the data well.
+Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part $x_{i}^{\top}\beta$ may not describe the data well.
 Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} while still adjusting for random effects:
@@
 To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}.
-In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects part is estimated as usual.
-To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random effects.
+In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects parameters are estimated as usual.
+To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random-effects parameters.
+
+\begin{figure}[tb!]
+\centering
+<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
+plot(exampleglmmtree)
+@
+\caption{\small GLMM tree of the motivating example dataset.
+Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
+\label{fig:example_glimmertree}
+\end{figure}
 
@@
 The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
+
 \begin{description}
@@
 
-  \item[Step 1:] Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$), with $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.
-
-  \item[Step 2:] Estimate random effects in the mixed-effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.
+  \item[Step 1:] Set $r = r+1$. Estimate a GLM tree using $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.
+
+  \item[Step 2:] Fit the mixed-effects model $g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b$ with subgroups $j(r)$ from the GLM tree estimated in Step~1.
+Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
 
@@
 The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown.
-In every iteration, the GLM tree and coefficients $\beta_{j(r)}$ and $b_{(r)}$ are re-estimated.
-The GLM tree is estimated, given the estimated random effects from the last iteration, and the random effects are estimated, given the estimated GLM tree from the current iteration.
-Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}.
+In every iteration, the GLM tree is re-estimated in step (1) and the fixed- and random-effects parameters are re-estimated in step (2).
+Note that the random-effects part of the model is not partitioned, but estimated globally.
+Only the fixed effects are estimated locally, within the cells of the partition.
+Convergence of the algorithm is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}.
 Typically, this converges if the tree does not change from one iteration to the next.
@@
 In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented.
-By taking into account the clustering of observations, the spurious split involving the anxiety variable no longer appears in the tree and the true treatment subgroups have been recovered.
-
-\begin{figure}[t!]
-\centering
-<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
+In addition to the treatment indicator, treatment outcome and the potential partitioning variables, the GLMM tree has also included a random intercept term with respect to the cluster indicator.
+As a result, the dependence between observations is taken into accoutn, the true treatment subgroups have been recovered and the spurious split involving the anxiety variable no longer appears in the tree.
+
+\FloatBarrier
+
+\section{Tutorial}
+
+We implemented the GLMM tree algorithm in the R package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
+The package makes use of the \verb|partykit| package \cite{HothyZeil15} and \verb|lme4| package \cite{BateyMeac12} to estimate trees and mixed-effects models, respectively.
+The latest version of the \verb|glmertree| package can be downloaded, installed and loaded as follows:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+install.packages("glmertree", repos="http://R-Forge.R-project.org")
+library(glmertree)
+@
+\end{small}
+
+Package documentation and examples can be accessed as follows:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+?glmertree
+@
+\end{small}
+
+The main functions in the \verb|glmertree| package are \verb|lmtree()|, for continuous outcome variables, and \verb|glmertree()|, for binary or count outcome variables.
+Both functions require the user to specify at least two arguments: \verb|formula| and \verb|data|.
+The data we will be using is the motivating example dataset described in the previous section.
+This dataset is included in the \verb|partykit| package and can be loaded as follows:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+data(a_m_data)
+summary(a_m_data)
+@
+\end{small}
+
+The model formula to be specified consists of a left- and right hand side.
+The left hand side of the model formula (preceding the tilde symbol) specifies the outcome variable.
+The right hand side consists of three parts, separated by vertical bars: The first part specifies the predictor variable(s) of the (generalized) linear model, the second part specifies the random effects and the third part specifies the potential partitioning variables:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+exampleglmmtree <- lmertree(depression ~ treatment | cluster | age + duration + 
+                              anxiety, data = a_m_data)
+@
+\end{small}
+
+Note that in the example above, the partitioning variables are continuous, but (ordered) categorical partitioning variables may also be specified.
+Also, we specified only a single variable in the random-effects part, resulting in estimation of a random intercept with respect to \verb|cluster|.
+More complex random effects can also be specified: for example, specifying the random-effects part as \verb=| (1 + age | cluster) |= would yield a model with a random intercept as well as a random slope for age with respect to cluster.
+The brackets are necessary to protect the vertical bars in the random effects formulation.
+
+Now, we can use the \verb|plot| method to recreate the tree depicted in Figure \ref{fig:example_glimmertree}:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
 plot(exampleglmmtree)
 @
-\caption{GLMM tree of the motivating example dataset.
-Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
-\label{fig:example_glimmertree}
-\end{figure}
+\end{small}
+
+In every inner node of the tree, the splitting variable and corresponding p-value from the parameter stability test is reported.
+To control for multiple testing, the p-values are Bonferroni corrected, by default.
+This can be turned off by adding \verb|bonferroni = FALSE| to the function call, yielding a less conservative criterion for the parameter stability tests, but note that this will increase the likelihood of overfitting.
+The significance level $\alpha$ equals .05 by default, but a different value, say for example .01, can be specified by including \verb|alpha = .01| in the function call.
+
+The predictions of the random effects can be plotted by adding \verb|plotranef = TRUE| to the function call:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+plot(exampleglmmtree, plotranef = TRUE)
+@
+\end{small}
+
+To obtain numerical results, \verb|print|, \verb|coef| and \verb|ranef| methods can be used:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+print(exampleglmmtree)
+coef(exampleglmmtree)
+ranef(exampleglmmtree)
+@
+\end{small}
+
+To obtain predicted values, the \verb|predict| method can be used:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+predict(exampleglmmtree, newdata = a_m_data[1:10,])
+@
+\end{small}
+
+When \verb|newdata| is not specified, predictions for the training observations are returned, by default.
+Random effects can be excluded from the predictions by adding \verb|re.form = NA|.
+This is useful, for example, when \verb|newdata| is specified, but the new observations do not have a cluster indicator or are from new clusters:
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+predict(exampleglmmtree, newdata = a_m_data[1:10, -4], re.form = NA)
+@
+\end{small}
+
+Residuals of the fitted GLMM tree can be obtained with the \verb|residuals| method. This can be useful for assessing potential misspecification of the model (e.g., heteroscedasticity):
+
+\begin{small}
+<<eval = FALSE, echo = TRUE>>=
+resids <- residuals(exampleglmmtree)
+preds <- predict(exampleglmmtree)
+plot(a_m_data$cluster, resids)
+scatter.smooth(preds, resids)
+@
+\end{small}
+
+The first plot did not indicate substantial variation in error variances across levels of the random effects.
+The second plot of fitted values against residuals also did not reveal a pattern indicating model misspecification.
 
@@
 
-To assess the performance of GLMM tree, we carried out three simulation studies: In Studies I and II, we compared the performance of GLMM tree with that of GLM tree.
-In Study I, we assessed the accuracy of the two tree algorithms in datasets with treatment-subgroup interactions.
-In Study II, we assessed the Type-I error of the two tree algorithms in datasets without treatment-subgroup interactions.
-In Study III, we compared the performance of GLMM tree in detecting treatment interactions with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.
-
-\subsection{General simulation design}
+To assess the performance of GLMM tree, we carried out three simulation studies: In Study I we assessed and compared the accuracy of GLMM tree, linear-model based MOB and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions.
+In Study II, we assessed and compared the Type-I error of GLMM tree and linear-model based MOB in datasets without treatment-subgroup interactions.
+In Study III, we assessed and compared the performance of GLMM tree and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions.
+As the outcome variable was continuous in all simulated dataset, the GLMM tree algorithm and the trees resulting from its application will be referred to as LMM tree(s).
+
+\subsubsection{General simulation design}
 
@@
 \begin{enumerate} 
-\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
-\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
-\item Two levels for the intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
-\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
-\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
-\item Two levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
+\item Sample size: $N=200$, $N=500$, $N=1000$.
+\item Number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
+\item Intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
+\item Number of clusters: $M=5$, $M=10$, $M=25$.
+\item Population SD of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
+\item Intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
 \end{enumerate}
@@
 Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$.
-Means for the other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$.
+Means for other potential partitioning covariates were drawn from a discrete uniform distribution on the interval $[-70,70]$.
 All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$.
 
-To generate the cluster-specific intercepts $b_m$, we partitioned the sample into $M$ equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design.
+To generate the cluster-specific intercepts, we partitioned the sample into $M$ equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design.
 For each cluster, a single value $b_m$ was drawn from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design.
-When $b$ was correlated with one of the potential partitioning variables, this variable was randomly selected.
+If $b$ was correlated with one of the potential partitioning variables, the correlated variable was randomly selected.
 
@@
 
-\subsection{Software}
-
-R \cite{R} was used for the generation and analysis of all datasets.
-The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| for other response distributions.
-For estimating GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
-
-For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
-This package provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.
-
-The significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing, in all simulations.
-The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight.
-
-In Study III, the \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in GLMMs.
+Due to the large number of cells in the simulation design, the most important predictors of accuracy were determined by means of ANOVAs and/or GLMs.
+The most important predictors of accuracy where then assessed through graphical displays.
+The ANOVAs and GLMs included main effects of algorithm type and the parameters of the data-generating process, as well as first-order interactions between algorithm type and each of the data-generating parameters.
+
+\subsubsection{Software}
+
+R \cite{R} was used for data generation and analyses.
+The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating LM trees, using the \verb|lmtree| function.
+For estimation of LMM trees the \verb|lmertree| function of the \verb|glmertree| package (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge) was used.
+The significance level $\alpha$ for the parameter instability tests was set to .05 for all trees, with a Bonferroni correction applied for multiple testing.
+The minimum number of observations per node in trees was set to 20 and maximum tree depth was set to three, thus limiting the number of terminal nodes to eight in every tree.
+
+The \verb|REEMtree| package \cite{SelaySimo11} was employed for estimating MERTs, using default settings.
+For estimating LMMs the \verb|lmer| function function from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
+The \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in LMMs in Study III.
 The \verb|lmerTest| package calculates effective degrees of freedom and $p$-values based on Satterthwaite approximations.
 
-\subsection{Study I: Method}
-
-\begin{figure}[t!]
+\subsection{Study I: Performance of LMM tree, LM tree and MERT in datasets with treatment-subgroup interactions}
+
+\subsubsection{Method}
+
+\begin{figure}[tb!]
 \centering
@@
 @
-\caption{Data-generating model for treatment-subgroup interactions.
-Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
+\caption{\small Data-generating model for treatment-subgroup interactions.
+Parameter $d_j$ denotes the node-specific standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
 \label{fig:modelC}
@@
 
-\subsubsection{Treatment-subgroup interaction design}
-
-For generating datasets with treatment-subgroup (or piecewise) interactions, we used a design from \citeA{DussyMech14} which is depicted in Figure~\ref{fig:modelC}.
-Figure~\ref{fig:modelC} shows four subgroups, characterized by values of the (true) partitioning variables $U_2$, and $U_1$ or $U_5$.
+\paragraph{Treatment-subgroup interaction design}
+
+For generating datasets with treatment-subgroup interactions, we used a design from \citeA{DussyMech14} which is depicted in Figure~\ref{fig:modelC}.
+Figure~\ref{fig:modelC} shows four subgroups, characterized by values of the partitioning variables $U_2$, and $U_1$ or $U_5$.
 Two of the subgroups have mean differences in treatment outcome, indicated by a non-zero value of $\beta_{j1}$, and two subgroups do not have mean differences in treatment outcome, indicated by a $\beta_{j1}$ value of 0.
 
-In this simulation study, some of the potential partitioning covariates were true partitioning covariates ($U_1$, $U_2$ and $U_5$), whereas others were not (e.g., $U_3$ and $U_4$).
+In this simulation design, some of the potential partitioning covariates are true partitioning covariates, the others are noise variable.
 Therefore, an extra level was added to the sixth facet of the \textit{General simulation design}:
@@
 \begin{enumerate}
-\item[6.] Three levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the true partitioning covariates ($U_1$, $U_2$ or $U_5$), $b$ correlated with one of the noise variables ($U_3$ or $U_4$.
+\item[6.] Intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the true partitioning covariates ($U_1$, $U_2$ or $U_5$), $b$ correlated with one of the noise variables ($U_3$ or $U_4$).
 \end{enumerate}
 
-Also, to assess the effect of treatment-effect differences in nodes 3 and 7, an additional facet was added:
+An additional facet was added to assess the effect of the magnitude of treatment-effect differences:
 
 \begin{enumerate}
-\item[7.] Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes.
-The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
+\item[7.] Two levels for the mean difference in treatment outcomes: The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
 \end{enumerate}
 
-50 datasets were generated for each cell of the design, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2 = 32,400$ datasets were generated.
-In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.
-
-\subsubsection{Assessment of performance}
-
-Performance of the algorithms was assessed by means of the total number of nodes in estimated GLM and GLMM trees, the accuracy of the resulting trees and predictive accuracy.
-An accurately recovered tree was defined as a tree with (1) a total of seven nodes (2) the first split involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$.
-The allowance of $\pm 5$ equals plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).
-
-Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences in every dataset.
+For each cell of the design, 50 datasets were generated. In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.
+
+\paragraph{Assessment of performance}
+
+Performance of the algorithms was assessed by means of tree size, tree accuracy and predictive accuracy.
+An accurately recovered tree was defined as a tree with (1) seven nodes in total, (2) the first split involving variable $U_2$ with a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ with a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ with a value of $63 \pm 5$.
+The allowance of $\pm 5$ equals plus or minus half the population SD of the partitioning variable ($\sigma_{U_k}$).
+
+For MERT, the number of nodes and tree accuracy was not assessed, as the treatment-subgroup interaction design in Figure \ref{fig:modelC} corresponds to a large number of regression tree structures, that would all be different but also correct.
+Therefore, performance of MERT was only assessed in terms of predictive accuracy.
+
+Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences.
 To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets.
@@
 
-To identify the most important predictors of tree size, tree accuracy and predictive accuracy, we used ANOVAs and GLMs with algorithm type and the parameters of the data-generating process as independent variables.
-First-order interactions between algorithm type and each of the data-generating parameters were also included.
-The most important predictors were further assessed with graphical displays.
-
-\subsection{Study I: Results}
+For MERT, predicted treatment-effect differences were obtained by fitting two MERTS on the training data: one using observations in the first treatment condition and one using observations in the second treatment condition.
+Predictions of treatment-effect differences for test observations were obtained by dropping test observations down both trees and taking the difference between the two predicted values.
+This approach was taken as it yielded higher predictive accuracy than the alternative of fitting a single MERT using treatment as a potential partitioning variable.
+
+\subsubsection{Results}
 
 <<simulation-data>>=
-load("treespecs_long.dat")
-treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)
-treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered=T)
-treespecs.long$treatdiff <- factor(as.numeric(as.character(treespecs.long$treatdiff)), ordered=T)
+load("treespecs_study1.dat")
+treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered = TRUE)
+treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered = TRUE)
+treespecs.long$treatdiff <- factor(as.numeric(as.character(treespecs.long$treatdiff)), ordered = TRUE)
 treespecs.long$truetree.values <- as.numeric(treespecs.long$truetree.values)-1
@@
 levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="uncorrelated"] <- "b and U uncorrelated"
-treespecs.long <- treespecs.long[treespecs.long$correlation.ind!="GLMM",]
-treespecs.long$correlation.ind <- factor(treespecs.long$correlation.ind)
-treespecs.long$treesize.ind <- factor(treespecs.long$treesize.ind)
-@
-
-\subsubsection{Tree size}
-
-\begin{figure}[t!]
+treespecs.long2 <- treespecs.long[!is.na(treespecs.long$treesize.values),]
+treespecs.long2$treesize.ind <- factor(treespecs.long2$treesize.ind)
+treespecs.long2$truetree.ind <- factor(treespecs.long2$truetree.ind)
+treespecs.long2$truetree.values <- as.numeric(treespecs.long2$truetree.values)-1
+@
+
+\paragraph{Tree size}
+
+\begin{figure}[tb!]
 \centering
 <<treesize-xyplot, fig=TRUE, width=9, height=9>>=
-aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
-print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
-type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
-abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
-@
-\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions.
-Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent levels of sample size.}
-\label{fig:xyplot_treesize_interact}
+treespecs.long2$N <- factor(as.numeric(as.character(treespecs.long2$N)), ordered = TRUE)
+treespecs.long2$sigmab <- factor(as.numeric(as.character(treespecs.long2$sigmab)), ordered = TRUE)
+levels(treespecs.long2$treesize.ind)[levels(treespecs.long2$treesize.ind)=="GLM tree"] <- "LM tree"
+levels(treespecs.long2$treesize.ind)[levels(treespecs.long2$treesize.ind)=="GLMM tree"] <- "LMM tree"
+aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN = mean, 
+                     data = treespecs.long2)
+levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "bi and splitting U correlated"] <- "b correlated with splitting U"
+levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "bi and non-splitting U correlated"] <- "b correlated with non-splitting U"
+levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "uncorrelated"] <- "b and U uncorrelated"
+xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups = treesize.ind, type = "b", 
+       ylab = "tree size", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), abline = c(7,0),
+       auto.key = list(space = "top", columns = 2, title = " ", cex.title = 1, lines = TRUE, points = TRUE))
+@
+\caption{\small Average tree size of LM and LMM trees.
+Tree size is defined as the total number of nodes in a tree.
+Rows represent the correlation between the random intercepts and one of the partitioning variables, columns represent sample size.
+The horizontal line in each graph indicates the 'true' tree size (the total number of nodes in the tree used for generating the interactions).}
+\label{fig:xyplot_treesize}
 \end{figure}
 
-GLMM tree yielded trees with an average size of 7.15 nodes ($\mathrm{SD}=0.61$), whereas GLM tree yielded an trees with an average size of 8.15 nodes ($\mathrm{SD}=2.05$), indicating that GLM trees involved more spurious splits than GLMM trees on average.
-The estimated probability that a dataset was erroneously not partitioned (the Type-II error) was 0 for both algorithms.
-
-An ANOVA indicated that the most important predictors of tree size were algorithm type, sample size, variance of the random intercept, and the correlation between partitioning variables.
-The graphical display in Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions.
-In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size.
-However, clear differences in tree size were observed when $\sigma_b > 0$.
-When sample size is small and the random intercept is not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average.
-When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized.
-However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially with larger sample sizes and larger $\sigma_b$ values.
-
-\subsubsection{Accuracy of recovered trees}
-
-\begin{figure}[t!]
+The average size of LMM trees was 7.15 nodes ($\mathrm{SD}=0.61$), whereas the average size of LM trees was 8.15 nodes ($\mathrm{SD}=2.05$), indicating that LM trees tend to involve more spurious splits than LMM trees.
+The effects of the most important predictors of tree size are depicted in Figure~\ref{fig:xyplot_treesize}.
+The average size of LMM trees was close to the true tree size in all conditions.
+In the absence of random effects, this was also the case for LM trees.
+In the presence of random effects that are correlated to a (potential) partitioning variable, LM tree starts to create spurious splits, especially with larger $\sigma_b$ values.
+In the presence of random effects that are uncorrelated to the other variables in the model, LM trees lack power to detect treatment-subgroup interactions if sample size is small (i.e., $N=200$).
+With larger sample sizes, LM trees showed about the true tree size, on average.
+
+\paragraph{Accuracy of recovered trees}
+
+\begin{figure}[tb!]
 \centering
 <<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
-aggdata.acc <- aggregate(formula=truetree.values ~ correlation.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
-print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=correlation.ind, type="b", 
-ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
-auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
-)
-@
-\caption{Tree accuracy of GLM and GLMM tree.
+levels(treespecs.long2$truetree.ind)[levels(treespecs.long2$truetree.ind)=="GLM tree"] <- "LM tree"
+levels(treespecs.long2$truetree.ind)[levels(treespecs.long2$truetree.ind)=="GLMM tree"] <- "LMM tree"
+aggdata.acc <- aggregate(formula = truetree.values ~ truetree.ind + N + sigmab + corUb, FUN = mean, data = treespecs.long2)
+print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups = truetree.ind, type = "b", 
+ylab = "tree accuracy", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), 
+auto.key = list(space = "top", columns = 2, title = " ", cex.title = 1, lines = TRUE, points = TRUE)))
+@
+\caption{\small Tree accuracy of LM and LMM tree.
 Tree accuracy is defined as the proportion of datasets in which the true tree was accurately recovered.
-Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$, columns represent levels of sample size.}
+Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$, columns represent sample size.}
 \label{fig:xyplot_treeaccuracy}
@@
 
-For the first split, GLMM tree selected the true partitioning variable ($U_2$) in all datasets, and GLM tree in all but one datasets.
-The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).
-
-Further splits were more accurately recovered by GLMM tree, which accurately recovered the partition in 90.40\% of datasets, and GLM tree in only 61.44\% of datasets.
-A GLM with tree accuracy as the outcome variable indicates that the most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and level of dependence between the partitioning variables and the random intercept.
-The graphical display in Figure~\ref{fig:xyplot_treeaccuracy} indicates that in the absence of random effects, the trees recovered by GLM and GLMM tree were about equally accurate.
-In the presence of random effects, GLM trees were much less accurate than GLMM trees when random effects were correlated with a partitioning covariate.
-When random intercepts were not correlated with one of the $U_k$ variables, GLMM tree outperformed GLM tree when sample size was small (i.e., $N = 200$).
-
-\subsubsection{Predictive accuracy}
-
-\begin{figure}[t!]
+The estimated probability that a dataset was erroneously not partitioned (Type-II error) was 0 for both algorithms.
+For the first split, LMM tree selected the true partitioning variable ($U_2$) in all datasets, and LM tree in all but one datasets.
+The mean splitting value of the first split was 29.94 for LM as well as LMM trees, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).
+
+Further splits were more accurately recovered by LMM tree, which accurately recovered the partition in 90.40\% of datasets, whereas LM tree accurately recovered the partition in 61.44\% of datasets.
+The effects of the four most important predictors of tree accuracy are depicted in Figure~\ref{fig:xyplot_treeaccuracy}.
+In the absence of random effects, LM and LMM tree were about equally accurate.
+In the presence of random effects, LM trees were much less accurate than LMM trees when random effects were correlated with a partitioning covariate.
+When random intercepts were not correlated with one of the $U_k$ variables, LMM trees outperformed LM trees only when sample size was small (i.e., $N = 200$).
+
+\paragraph{Predictive accuracy}
+
+\begin{figure}[tb!]
 \centering
 <<correlation-xyplot, fig=TRUE, width=9, height=7>>=
-aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
-print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
-type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
-auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T),
-scales=list(y=list(limits = c(-.1,1.1))))
-)
-@
-\caption{Average predictive accuracy of GLM and GLMM trees.
-Predictive accuracy of trees is defined as the correlation between the true and predicted outcome differences between Treatment~1 and~2.
-Rows represent the levels of the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences, columns represent levels of sample size.}
+levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLM tree"] <- "LM tree"
+levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLMM tree"] <- "LMM tree"
+aggdata.cor <- aggregate(formula = correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN = mean, data = treespecs.long)
+print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups = correlation.ind, 
+type = "b", ylab = "correlation", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), 
+auto.key = list(space = "top", columns = 3, title = " ", cex.title = 1, lines = TRUE, points = TRUE),
+scales = list(y = list(limits = c(-.1,1.1)))))
+@
+\caption{\small Average predictive accuracy of LM and LMM trees.
+Predictive accuracy is defined as the correlation between the true and predicted treatment-effect differences.
+Rows represent absolute treatment-effect differences in subgroups with treatment-effect differences, columns represent sample size.}
 \label{fig:xyplot_correlations}
@@
 
-The predicted treatment-effect differences of GLMM tree were closer to the true differences than those of GLM tree, with a mean correlation of .93 ($\mathrm{SD}=0.12$) for GLMM tree and .88 ($\mathrm{SD}=0.19$) for GLM tree.
-An ANOVA indicated that the most important predictors of predictive accuracy were algorithm type, sample size, variance of the random intercept, and the effect size of the treatment difference.
-The graphical display in Figure~\ref{fig:xyplot_correlations} shows clear main effects of sample size and treatment-effect differences for both algorithms.
-In the absence of random effects (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were about equally accurate.
-In the presence of random effects, GLM tree predictions were clearly less accurate than those of GLMM tree when treatment-effect differences were smaller (i.e., Cohen's $d = .5$).
-This effect was also observed, but much less pronounced, when treatment-effect differences were larger (i.e., Cohen's $d = 1.0$).
-In other words, GLMM tree outperforms GLM tree in the presence of random effects, especially when sample size and treatment-effect differences are smaller.
-
-\subsection{Study II: Method}
-
-\subsubsection{Design}
-In the second simulation study we assessed the Type-I error of GLM and GLMM tree.
-Type-I error was defined as the proportion of datasets without treatment-subgroup interactions which were erroneously partitioned by the algorithm.
-In the simulated datasets in this study there was only a main effect of treatment in the population.
+The predicted treatment-effect differences of LMM tree showed an average correlation of .93 (SD = .13) with the true differences.
+LM tree and MERT showed lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively.
+The most important predictors of predictive accuracy are depicted in Figure \ref{fig:xyplot_correlations}.
+Performance of all three algorithms improves with increasing sample size and treatment-effect differences.
+Furthermore, LMM tree and MERT are not much affected by the presence and magnitude of random effects in the data.
+LMM tree performs most accurate in most conditions and is never outperformed by the other methods.
+MERT performs least accurate in most conditions and never outperforms the other methods.
+The differences in accuracy become less pronounced with larger sample and effect sizes.
+
+\FloatBarrier
+
+\subsection{Study II: Type-I error of LM and LMM trees}
+
+\subsubsection{Method}
+
+\paragraph{Design}
+
+In the second simulation study we assessed the Type-I error rate of LM and LMM tree.
+In the datasets in this study there was only a main effect of treatment in the population.
 Put differently, there was only a single global value of $\beta_j = \beta$ in every dataset.
-
-To assess the effect of the treatment-effect difference $\beta$, an additional facet was added to the \textit{General simulation design} described above: 
+A Type-I error was defined as the proportion of datasets without treatment-subgroup interactions which were erroneously partitioned by the algorithm.
+
+To assess the effect of the treatment-effect difference $\beta$, an additional facet wasadded to the \textit{General simulation design}: 
 \begin{enumerate}
-\item[7.] Two levels for $\beta$, the unstandardized mean difference in treatment outcomes.
-The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
+\item[7.] Two levels for $\beta$, the global mean difference in treatment outcomes: $\beta = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$) and $\beta = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
 \end{enumerate}
 
-For each cell of the \textit{General simulation design} described above, 50 datasets were generated.
-As a result, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2 = 21,600$ datasets without treatment-subgroup interactions were generated.
-In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta + z_{i}^{\top} b_{m} + \epsilon_{i}$.
-
-\subsubsection{Assessment of performance}
-
-The total number of nodes in estimated GLM and GLMM trees were calculated and a tree of size $>1$ was classified as a Type-I error.
-To identify the most important predictors of Type-I error, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables.
-First-order interactions between algorithm type and each of the data-generating parameters were also included.
-The most important predictors were further assessed with a graphical display.
-
-\subsection{Study II: Results}
-
-In datasets without treatment-subgroup interactions, average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree.
-The average Type-I error rate was only .04 for GLMM tree, and .33 for GLM tree.
-Main predictors of type-I error were found to be sample size, variance of the random intercept, and dependence between the random intercept and one of the partitioning variables.
-
-The graphical display in Figure~\ref{fig:xyplot_treesize_nointeract} indicates that GLMM tree has a Type-I error rate somewhat below the pre-specified $\alpha$ level under all circumstances.
-The same goes for GLM tree, whenever the random intercept has 0 variance, or the random intercept is not correlated to one of the partitioning covariates.
-However, when the random intercept is correlated with one of the potential partitioning covariates, the type-I error rapidly increases for GLM tree.
-With large sample sizes and large variance of the random intercept, GLM tree will almost certainly create spurious splits.
-
-\begin{figure}[t!]
+For each cell in the simulation design, 50 datasets were generated. In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta + z_{i}^{\top} b_{m} + \epsilon_{i}$.
+
+\paragraph{Assessment of performance}
+
+To assess the Type-I error rates of LM and LMM trees, tree sizes were calculated and trees of size $>1$ were classified as Type-I errors.
+The nominal Type-I error rate for both LM and LMM tree equals .05, corresponding to the pre-specified significance level $\alpha$ for the parameter instability tests.
+
+\subsubsection{Results}
+
+\begin{figure}[tb!]
+\centering
 <<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
@@
 treesizes.long$treesize.valuesD <- as.numeric(treesizes.long$treesize.values>1)
-treesize.glm <- glm(treesize.valuesD ~ treesize.ind + N + rho + np + treatdiff + numbclus + 
-sigmabm + corUbm + treesize.ind*(N + rho + np + treatdiff + numbclus + sigmabm + corUbm), 
-data=treesizes.long, family="binomial")
-summary(treesize.glm)
+levels(treesizes.long$treesize.ind)[levels(treesizes.long$treesize.ind)=="GLM tree"] <- "LM tree"
+levels(treesizes.long$treesize.ind)[levels(treesizes.long$treesize.ind)=="GLMM tree"] <- "LMM tree"
 #N, corUb & sigmab have strongest main and/or interaction effects
@@
 @
-\caption{Type-I error of GLM and GLMM trees. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
+\caption{\small Type-I error rate of LM and LMM trees. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
 \label{fig:xyplot_treesize_nointeract}
@@
 
-\subsection{Study III: Method}
-
-\subsubsection{Interaction design}
-
-The treatment-subgroup interactions in Study I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables $U_1$, $U_2$ and $U_5$.
+In datasets without treatment-subgroup interactions, average tree size was 1.09 ($\mathrm{SD}=0.44$) for LMM tree, and 2.02 ($\mathrm{SD}=1.68$) for LM tree.
+The average Type-I error rate was only .04 for LMM tree, and .33 for LM tree.
+Main predictors of type-I error are depicted in Figure~\ref{fig:xyplot_treesize_nointeract}, which shows that LMM trees have a Type-I error rate somewhat below the pre-specified $\alpha$ level in all conditions.
+The same goes for LM tree, when random effects are absent, or uncorrelated to one of the partitioning covariates.
+When the random intercept is correlated with one of the potential partitioning covariates, the type-I error rapidly increases for LM tree.
+With increasing sample size or random-effects variance, LM tree will yield a larger number of spurious splits.
+
+\FloatBarrier
+
+\subsection{Study III: Recovery of piecewise and continuous interactions by LMM trees and LMMs with pre-specified interactions}
+
+\subsubsection{Method}
+
+\paragraph{Interaction design}
+
+The treatment-subgroup interactions in Study I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables.
 Trees are pre-eminently suited for recovering such piecewise interactions, but may have difficulty when the true interactions are continuous functions of moderator variables (for example, $U_1 \cdot U_2$).
@@
 \begin{enumerate}
-	\item[7.] Three levels for interaction type: continuous, piecewise and a combination of both. 
+	\item[7.] Three levels for interaction type: continuous, piecewise and combined piecewise-continuous interactions. 
 \end{enumerate}
 
-For datasets with purely piecewise interactions, the same partition as in Study II (depicted in Figure \ref{fig:modelC}) was used.
+For datasets with purely piecewise interactions, the same partition as in Study I (Figure \ref{fig:modelC}) was used.
 In other words, the outcome variable in this design was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
 
-For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure \ref{fig:modelC} was used.
-However, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes comprised continuous main and (treatment) interaction effects of the partitioning variables.
-These node-specific effects are described in Table \ref{tab:continuous_terms}.
-$\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure \ref{fig:modelC}.
-Continuous interactions were created using centered $U_k$ variables, calculated by subtracting the variable mean.
-
-For datasets with purely continuous interactions, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$.
-Here, $\beta$ has a global value and no subscript, comprising purely continuous main and (treatment) interaction effects of the moderator variables $U$, as described in Table \ref{tab:continuous_terms}.
-
-\begin{table}
-\caption{Fixed-effects terms in simulations with continuous and combined continuous and piecewise interaction designs.}
-\begin{tabular}{lccccc}
+For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure \ref{fig:modelC} was also used.
+However, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes now comprised continuous main and (treatment) interaction effects of the partitioning variables.
+The corresponding node-specific parameters are presented in Table \ref{tab:continuous_terms}.
+The $\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure \ref{fig:modelC}.
+The interaction terms were created using centered $U_k$ variables, calculated by subtracting their variable means.
+
+In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by the single $\beta$ column in Table \ref{tab:continuous_terms}.
+
+\begin{table}[tb!]
+\small
+\caption{\small Fixed-effects terms in simulations with continuous and combined continuous and piecewise interaction designs.}
+\begin{tabular}{lcccccc}
 \thickline
-term							&	$\beta_3$&	$\beta_4$&	$\beta_6$&	$\beta_7$&	$\beta$		\\
+Term							&	$\beta_3$&	$\beta_4$&	$\beta_6$&	$\beta_7$&&	$\beta$		\\
 \hline
-intercept						&	27		&	27		&	27		&	27		&	27			\\
-$U_2$							&	.1		&	.1		&	.1		&	.1		&	.1			\\
-$U_2 \cdot U_1$ 				&	-.357	&	0		&	0		&	0		&	-.357		\\
-$U_2 \cdot U_5$	 				&	0		&	0		&	0		&	.357	&	.357		\\
-$U_2 \cdot U_1 \cdot treatment$	&	-.151	&	0		&	0		&	0		&	-.151		\\
-$U_2 \cdot U_5 \cdot treatment$	&	0		&	0		&	0		&	.151	&	.151		\\
+intercept						&	27		&	27		&	27		&	27		&&	27			\\
+$U_2$							&	.1		&	.1		&	.1		&	.1		&&	.1			\\
+$U_2 \cdot U_1$ 				&	-.357	&	0		&	0		&	0		&&	-.357		\\
+$U_2 \cdot U_5$	 				&	0		&	0		&	0		&	.357	&&	.357		\\
+$U_2 \cdot U_1 \cdot treatment$	&	-.151	&	0		&	0		&	0		&&	-.151		\\
+$U_2 \cdot U_5 \cdot treatment$	&	0		&	0		&	0		&	.151	&&	.151		\\
 \hline
-\multicolumn{6}{l}{\small{\textit{Note. }Subscripted $\beta$ values refer to the combined}}\\	\multicolumn{6}{l}{\small piecewise and continuous interaction design, with the}\\
-\multicolumn{6}{l}{\small values of the subscripts denoting the terminal nodes in}\\
-\multicolumn{6}{l}{\small Figure \ref{fig:modelC}. $\beta$ without a subscript refers to the global}\\
-\multicolumn{6}{l}{\small coefficients in the purely continuous interaction design.}\\
+\multicolumn{7}{l}{\footnotesize{\textit{Note. }Subscripted $\beta$ values refer to the terminal nodes}}\\	
+\multicolumn{7}{l}{\footnotesize in Figure \ref{fig:modelC} for the combined piecewise and continuous}\\
+\multicolumn{7}{l}{\footnotesize interaction design; $\beta$ without subscript refers to the global}\\
+\multicolumn{7}{l}{\footnotesize coefficients in the continuous interaction design.}\\
 \end{tabular}
@@
 The fifth facet of the data-generating design was limited to two levels ($\sigma_b = 2.5$ and $\sigma_b = 7.5$).
-As a result, $50 \times 3 \times 2 \times 2 \times 1 \times 2 \times 2 \times 3$ = 7,200 training datasets were generated for this study.
-
-\subsubsection{GLMMs with pre-specified interactions}
-GLMMs were estimated by specifying main effects for all covariates $U_k$ and the treatment indicator, first-order interactions between all pairs of covariates $U_k$, and second-order interactions between all pairs of covariates $U_k$ and treatment.
-Continuous predictor variables were centered by subtracting the mean value, before calculating and including the interaction term in the GLMM.
-
-\subsubsection{Assessment of performance} 
+For every cell of the design, 50 datasets were generated.
+
+\paragraph{LMMs with pre-specified interactions}
+LMMs were estimated by specifying main effects for all covariates $U_k$ and the treatment indicator, first-order interactions between all pairs of covariates $U_k$, and second-order interactions between all pairs of covariates $U_k$ and treatment.
+Continuous predictor variables were centered by subtracting the mean value, before calculating and including the interaction term in the LMM.
+
+\paragraph{Assessment of performance} 
 Predictive accuracy was assessed in terms of the correlation between the true and predicted treatment-effect differences in test datasets.
-As the full GLMM models were likely to overfit, predictions for GLMMs with pre-specified interactions were obtained by refitting a GLMM on the training data, using only the predictors with p-values $< .05$ in the original GLMM.
-Predictions for test observations were obtained from these refitted GLMMs.
-
-To identify the most important predictors of predictive accuracy, we used an ANOVA with algorithm type and the parameters of the data-generating process as independent variables.
-First-order interactions between algorithm type and each of the data-generating parameters were also included.
-The most important predictors were further assessed with a graphical display.
-
-<<linear-simulation-data>>=
+As the full LMM models were likely to overfit, LMMs were refitted on the training data, using only the predictors with $p$-values $<.05$ in the original LMM.
+Predictions for test observations were obtained using the refitted LMMs.
+
+\subsubsection{Results}
+
+\begin{figure}[tb!]
+\centering
+<<correlation-pw-xyplot, fig=TRUE, width=9, height=7>>=
 load("pw_treespecs_long.dat")
@@
 levels(treespecs.long$type) <- gsub("linear","continuous", levels(treespecs.long$type))
-@
-
-\subsection{Study III: Results}
-
-\begin{figure}[t!]
-\centering
-<<correlation-pw-xyplot, fig=TRUE, width=9, height=7>>=
+levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLMM"] <- "LMM"
+levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLMM tree"] <- "LMM tree"
 aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + np + type + N, FUN=mean, 
@@
 @
-\caption{Average predictive accuracy of GLMMs and GLMM trees.
+\caption{\small Average predictive accuracy of LMMs and LMM trees.
 Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2.
@@
 
-GLMM tree showed somewhat higher average correlations between the true and predicted treatment-effect differences: average accuracy was .54 (SD = .40) for GLMM trees and .51 (SD = .43) for GLMMs.
-The most important predictors of predictive accuracy were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.
-
-As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely continuous interactions.
-GLMMs perform poorly when interactions are not purely piecewise, and GLMM tree performace poorly when interactions are purely linear.
+On average, LMM tree showed somewhat higher accuracy: the average correlation between true and predicted treatment-effect differences was .54 (SD = .40) for LMM trees and .51 (SD = .43) for LMMs.
+The effects of the most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlation_pw}.
+As Figure~\ref{fig:xyplot_correlation_pw} indicates, LMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas LMMs show highest predictive accuracy in datasets with purely continuous interactions.
+LMMs perform poorly when interactions are not purely piecewise, and LMM trees perform poorly only when interactions are purely linear.
 
 Performance of both methods improves with increasing sample size.
-Furthermore, performance of GLMM tree is not affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous.
-This indicates that GLMM tree may be especially useful for exploratory purposes, where there are many potential moderator variables.
-
-GLMM trees may also provide simpler models, as the GLMMs included 12.30 significant predictor variables (main effects and interactions), on average.
-GLMM trees had 3.38 inner nodes on average, requiring less variables to be evaluated for making predictions.
-
-\section{Application: Individual patient-level meta-analysis dataset on treatments for depression}
+Furthermore, performance of LMM tree is not affected by the number of covariates, whereas the predictive accuracy of LMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous.
+This indicates that LMM tree is especially useful for exploratory purposes, where there are many potential moderator variables.
+
+In terms of interpretability, LMM trees will often provide simpler models: The LMMs included 12.30 significant terms on average, whereas LMM trees had 3.38 inner nodes on average, requiring only about 3-4 variables to be evaluated for making predictions.
+
+\FloatBarrier
+
+\section{Application: Individual patient-level meta-analysis on treatments for depression}
 
@@
 
-\subsection{Method}
-
-To further illustrate the use of GLM and GLMM tree, we applied both algorithms to a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}.
+\subsubsection{Method}
+
+\paragraph{Dataset}
+To illustrate the application of GLMM tree on real data, we use a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}.
 This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
@@
 Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}).
-\citeA{CuijyWeit14} found no indication that gender predicted or moderated treatment outcomes.
-Further details on the dataset can be found in \citeA{CuijyWeit14}.
-
-In our analyses, posttreatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score.
+\citeA{CuijyWeit14} found no indication that gender predicted or moderated treatment outcome.
+
+In our analyses, post-treatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pre-treatment HAM-D score.
 The predictor variable in the linear model was treatment type (0 = CBT and 1 = PHA).
@@
 
-In RCTs, ANCOVAs are often employed, to linearly control posttreatment values on the outcome measure for pretreatment values.
-Therefore, we estimated the GLM and GLMM trees using posttreatment HAM-D scores, controlled for the linear effects of pretreatment HAM-D scores.
-The trees were grown using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score.
-As a result, data from 694 patients from 7 studies were included in the analyses.
-Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.
+In RCTs, ANCOVAs are often employed, to linearly control post-treatment values on the outcome measure for pre-treatment values.
+Therefore, post-treatment HAM-D scores, controlled for the linear effects of pre-treatment HAM-D scores were taken as the outcome variable.
+All models were fitted using data of the 694 patients from 7 studies, for which complete data was available.
+Results of our analysis may therefore not be fully representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.
+
+\paragraph{Models and comparisons}
+
+As the outcome variable is continuous, we employed an identity link and Gaussian error distribution.
+The tree resulting from GLMM tree will therefore be refered to as an LMM tree.
+To compare the accuracy of the LMM tree, we also fitted an LM tree and an LMM with pre-specified interactions to the data.
+In the LMM, the outcome variable was regressed on a random intercept, main effects of treatment and the potential moderators (partitioning variables) and interactions between treatment and the potential moderators.
+As it is not known in advance how to interact the potential moderators, higher-order interactions were not included.
+
+\paragraph{Effect size}
 
@@
 Cohen's $d$ was calculated by dividing the node-specific predicted treatment outcome difference by the node-specific pooled standard deviation.
-Confidence intervals for Cohen's $d$ could not be calculated, as these can not take into account the exploratory nature of our analyses (i.e., variable and split point selection).
-
-To compare the two tree-based approaches with a more classical approach of interaction detection, we also fitted a GLMM with pre-specified interactions on the dataset.
-In the GLMM, the posttreatment HAM-D scores were controlled for the linear effects of pretreatment HAM-D scores, and then regressed on a random intercept, main effects of treatment and the potential moderators (partitioning variables), and interactions between treatment and the potential moderators.
-As it is not known in advance how to interact the potential moderators, higher-order interactions were not included.
-
-The predictive accuracy of the GLM and GLMM trees and the GLMM was assessed by calculating average correlations between observed and predicted HAM-D posttreatment scores, based on 50-fold cross validation.
+
+\paragraph{Predictive accuracy and stability}
+
+Predictive accuracy of each method was assessed by calculating average correlation between observed and predicted HAM-D post-treatment scores, based on 50-fold cross validation.
 
@@
 )
-@
-
-\begin{figure}[t!]
-\centering
-\setkeys{Gin}{width=0.75\textwidth}
-<<depression-glmtree, fig=TRUE, height=9, width=8>>=
+
+## Panel-combining function:
 combine_panels <- function(panel1, panel2, party2 = NULL) {
@@
 }
-
-plot(lmtree_app, tnex = 3.5, terminal_panel = combine_panels(
-  panel1 = node_bivplot(lmtree_app),
-  panel2 = node_terminal(GLMtree_app_eff, FUN = identity, height = 10, width = 14, id = FALSE),
-  party2 = GLMtree_app_eff
-))
-@
-\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-Upper panel: The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
-Lower panel: Subgroup-specific descriptive statistics.}
-\label{fig:lmtree_C&W}
-\end{figure}
-
-\subsection{Results}
-
-The tree and effects sizes resulting from application of GLM tree are presented in Figure~\ref{fig:lmtree_C&W}.
-Those resulting from application of GLMM tree are presented in Figure~\ref{fig:lmertree_C&W}.
-
-The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
-By taking into account study-specific intercepts, the GLMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split made by GLM tree may be a spurious one.
-The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
-The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Figure~\ref{fig:lmertree_C&W}).
-For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Figure~\ref{fig:lmertree_C&W}).
+@
+
+\subsubsection{Results}
+
+The trees and effect sizes resulting from application of LMM and LM tree are presented in Figure~\ref{fig:lmertree_C&W} and Figure~\ref{fig:lmtree_C&W}, respectively.
+The LM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
+By taking into account study-specific intercepts, the LMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split in the LM tree may be spurious.
+The LMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
+The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$).
+For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$).
 The estimated intraclass correlation coefficient for the GLMM tree was .05.
 
-The GLMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: b = 2.29, p = .002; interaction with treatment: b = -2.10, p = .028).
-Also, the GLMM indicated an interaction between treatment and age (b = .10, p = .018).
-
-Assessment of predictive accuracy by means of 50-fold cross validation indicated better predictive accuracy for GLMM tree than for GLM tree and the GLMM with pre-specified interactions.
-The correlation between true and predicted posttreatment HAM-D total scores averaged over 50 folds was .272 ($\mathit{var}=.067$) for GLMM tree, .233 ($\mathit{var}=.064$) for the GLMM with prespecified interactions and .190 ($\mathit{var}=.084$) for GLM tree.
-
-Table \ref{tab:stability} presents statistics on the variables selected for partitioning in subsamples of the dataset.
-Note that the selection frequencies do not add up to 1, as trees may involve multiple, or no splits.
-Table \ref{tab:stability} indicates that the presence of a comorbid anxiety disorder was selected for partitioning in the majority of GLMM trees grown on subsamples of the dataset, and all other variables were selected in less than 3\% of the subsamples.
-As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.
-
-\begin{figure}
+\begin{figure}[tb!]
 \centering
@@
 @
-\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-Left panel: The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
-Right panel: Subgroup-specific descriptive statistics.}
+\caption{\small LMM tree for prediction of treatment outcomes.
+Upper terminal nodes: y-axes represent post-treatment HAM-D scores, x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
+Lower terminal nodes: Subgroup-specific descriptive statistics.}
 \label{fig:lmertree_C&W}
@@
 
-\begin{table}
+\begin{figure}[tb!]
+\centering
+\setkeys{Gin}{width=0.75\textwidth}
+<<depression-glmtree, fig=TRUE, height=9, width=8>>=
+plot(lmtree_app, tnex = 3.5, terminal_panel = combine_panels(
+  panel1 = node_bivplot(lmtree_app),
+  panel2 = node_terminal(GLMtree_app_eff, FUN = identity, height = 10, width = 14, id = FALSE),
+  party2 = GLMtree_app_eff
+))
+@
+\caption{\small LM tree for prediction of treatment outcomes.
+Upper terminal nodes: y-axes represent post-treatment HAM-D scores, x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
+Lower terminal nodes: Subgroup-specific descriptive statistics.}
+\label{fig:lmtree_C&W}
+\end{figure}
+
+The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: b = 2.29, p = .002; interaction with treatment: b = -2.10, p = .028).
+Also, the GLMM indicated an interaction between treatment and age (b = .10, p = .018).
+
+Assessment of predictive accuracy by means of 50-fold cross validation indicated better predictive accuracy for the LMM tree than for the LM tree and the LMM.
+The correlation between true and predicted post-treatment HAM-D total scores averaged over 50 folds was .272 ($\mathit{var}=.067$) for LMM tree, .233 ($\mathit{var}=.064$) for the LMM with pre-specified interactions and .190 ($\mathit{var}=.084$) for the LM tree.
+
+Table \ref{tab:stability} presents statistics on the variables selected for partitioning in subsamples of the dataset.
+Presence of a comorbid anxiety disorder was selected for partitioning in the majority of LMM trees grown on subsamples of the dataset, while the other variables were selected in at most 4\% of the subsamples.
+As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.
+
+\begin{table}[tb!]
 	\small
-	\caption{Variable selection statistics}
+	\caption{\small Variable selection statistics}
 	\begin{tabular}{lcc}
@@
 		&	\multicolumn{2}{c}{Selection frequency} \\
-		Variable				&	GLM tree&	GLMM tree	\\
+		Variable				&	LM tree&	LMM tree	\\
 		\hline
@@
 		\hline
-		\multicolumn{3}{l}{\textit{Note.} Frequencies are calculated over 500 random}\\
-		\multicolumn{3}{l}{subsamples of the complete dataset. Frequencies}\\
-		\multicolumn{3}{l}{do not add up to 1, as trees may involve multiple}\\
-		\multicolumn{3}{l}{or no splits.}\\
+		\multicolumn{3}{l}{\footnotesize \textit{Note.} Frequencies are calculated over 500 random}\\
+		\multicolumn{3}{l}{\footnotesize subsamples of the complete dataset. Frequencies do not}\\
+		\multicolumn{3}{l}{\footnotesize add up to 1, as trees may involve multiple or no splits.}\\
 	\end{tabular}
@@
 
-In the current paper, we presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters.
-As such, we hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies.
-Also, we illustrated the application of GLMM tree on an existing dataset.
-
-GLMM tree performed very well in recovering treatment-subgroup interactions, accurately recovering interactions in 90\% of datasets with treatment-subgroup interactions.
-In contrast, GLM tree accurately recovered interactions in only 61\% of these datasets.
-In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.
-Put differently, the Type-I error rate of GLMM tree very closely approximated the $\alpha$ level used for testing parameter stability, whereas the Type-I error rate of GLM tree clearly exceeded this value.
+\subsubsection{Summary}
+
+We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters.
+We hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies.
+
+GLMM tree accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions.
+In contrast, GLM tree accurately recovered treatment-subgroup interactions in only 61\% of these datasets.
+In terms of predictive accuracy, GLMM tree outperformed GLM tree as well as MERT, with an average correlation between true and predicted treatment-effect differences of .94 for GLMM tree, .88 for GLM tree and .75 for MERT.
+The Type-I error rate of GLMM tree very closely approximated the $\alpha$ level used for testing parameter stability: GLMM tree erroneously detected subgroups in 4\% of datasets without treatment-subgroup interactions, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.
 
 The better performance of GLMM tree was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables.
-In these instances, random effects gave rise to spurious subgroup detection (spurious splits) by GLM tree, both in datasets with and without treatment-subgroup interactions.
-
-GLMM tree also provided better predictive accuracy than GLM tree, with an average correlation between true and predicted treatment differences of .94 for GLMM tree, and .88 for GLM tree.
-GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, differences in treatment effects across terminal nodes were relatively small (Cohen's $d=.5$), and/or sample size was small ($< 1,000$).
-Such treatment-effect differences and sample sizes may be quite common, even in multi-center clinical trials, and GLMM tree may provide a helpful tool for subgroup detection in those instances.
-
-As expected, when random effects were absent from the simulated datasets, GLM and GLMM tree yielded very similar predictive accuracy.
-This indicates that GLMM tree can be applied whenever cluster-specific random effects are expected: In the absence of random effects, GLM tree and GLMM tree are expected to perform about equally well and in the presence of random effects GLMM tree is expected to outperform GLM tree.
-
-Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided somewhat better accuracy, on average.
-GLMM tree performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at leas partially piecewise.
-Our findings also indicated a clear advantage for GLMM tree when there are a large number of potential moderator variables (i.e., $> 5$).
-Therefore, GLMM trees may be better suited than GLMMs for exploratory analyses, in which moderator variables need to be selected from a larger number of covariates.
-Furthermore, the number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models.
-The trees in our simulations were limited to a maximum number of 7 inner nodes, and may therefore be much easier to use for prediction in practical decision-making contexts.
-
-These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures.
-However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified.
-That is, if there are subgroups with respect to the variables specified as potential partitioning variables and these subgroups have different values for the parameters of the GLM, then GLMM tree will accurately recover those subgroups.
-However, misspecification of the model will negatively affect performance.
-The most important source of misspecification would be the omission of relevant variables, either in the GLM or as partitioning variables.
-When relevant variables are omitted, GLMM tree can only approximate the true subgroups using the specified variables.
-Another source of misspecification would be the inclusion of irrelevant variables.
-Our simulations indicate that the performance of GLMM tree was not negatively affected by the number of potential moderator variables specified, but the power to detect subgroups may still be reduced with larger numbers of potential partitioning variables.
-Including irrelevant variables in the GLM may also negatively affect performance, although we have not assessed this in our simulations.
-
-In the Application, we found GLMM tree to provide results that were more easily interpretable, and also more accurate than a GLM tree without random effects.
-In addition, to judge clinical relevance of the findings, we calculated node-specific effect sizes, as would often be done in RCTs or meta-analysis.
-Although we have limited ourselves to calculating Cohen's $d$ in the current paper, equivalent values of the success rate difference or the number needed to treat can be calculated, but this would involve additional distributional assumptions \cite{KraeyKupf06}.
-Node-specific effect sizes can also be used to prune trees, when a researcher prefers to have a final tree which is based on statistical as well as clinical significance.
-A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significance in the tree-growing process.
-
-As discussed in the Introduction, several tree-based methods for treatment-subgroup interaction detection are available.
-These methods have different objectives, and there is not yet an agreed-upon single best method.
-In a simulation study, \citeA{SiesyVanM16} found the method of \citeA{ZhanyTsia12Stat} to perform best, followed by model-based recursive partitioning.
+In those datasets, GLM tree is likely to detect spurious splits and subgroups.
+At the same time, GLM tree showed less power to detect the true subgroups in the presence of random effects.
+As expecte, the accuracy of MERT was not affected much by the presence of random effects, but only approached the accuracy of GLMM tree in datasets with the largest sample and effect sizes.
+GLMM tree especially outperformed the other methods when effect size was small (i.e., Cohen's $d=.5$) and sample size was small (i.e., 200).
+Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM tree may therefore provide a helpful tool for subgroup detection in those instances.
+
+When random effects were absent from the simulated datasets, GLM and GLMM tree yielded very similar predictive accuracy.
+This finding is of practical importance, as it indicates that application of GLMM tree will not `hurt': GLM tree and GLMM tree are expected to perform equally well in the absence of random effects, while GLMM tree will likely outperform GLM tree in the presence of random effects.
+
+Compared to linear mixed-effects models with pre-specified interactions, GLMM tree provided somewhat better accuracy, on average.
+As expected, GLMM tree performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise.
+We found a clear advantage of GLMM tree in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited than GLMMs for exploratory analyses.
+Also, GLMM trees may be easier to interpret: The number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models.
+The trees in our simulations were limited to a maximum depth of three, requiring evaluation of at most 3 variables to make a prediction or decision in practice.
+
+In the Application, we obtained similar findings: GLMM tree yielded higher predictive accuracy, while using a smaller number of variables for prediction than GLM tree and a GLMM with pre-specified interactions.
+In addition, the GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.
+
+\subsubsection{Limitations and future research}
+
+Recursive partitioning methods were originally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}.
+However, GLMM tree is obviously a parametric tool, as it fits fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors.
+Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM tree.
+
+Furthermore, misspecification of partitioning and fixed- and random-effects variables will also reduce accuracy of the resulting GLMM tree.
+If relevant variables are omitted, or incorrectly specified, GLMM tree can only approximate the true subgroups using the specified variables.
+Our simulations indicate that LM tree detected spurious subgroups as a result of misspecifying (that is, not including) the random effects.
+Reduced accuracy and spurious splits can also be expected to occur when relevant random effects are not included when specifying the GLMM tree model.
+Furthermore, as the random effects are estimated globally, misspecification of the random effects can have a strong impact on the resulting GLMM tree model.
+
+Another source of misspecification is the inclusion of irrelevant variables.
+Although our simulations indicate that the performance of GLMM tree was not negatively affected by increasing the number of noise variables specified for partitioning from 2 to 12, the power to detect subgroups may be reduced with much larger numbers of noise variables.
+Including irrelevant variables in the random or fixed effects may also negatively affect accuracy of GLMM tree, but we have not assessed this in our study.
+
+Users can reduce the risk of misspecification when fitting a GLMM tree in two ways.
+Firstly, by carefully specifying the predictors of the GLM, the partitioning variables and the random effects.
+Secondly, by inspecting residuals and assessing stability of the resulting model.
+In the Tutorial we have shown how residuals can be plotted to assess potential misspecification.
+In the Application we have shown how the \verb|stablelearner| package can be used to assess tree stability.
+However, more research on the effects that various types of misspecification will have on the performance of GLMM tree is required.
+
+As GLMM tree fits more complex models than non-parametric tree-based methods, like CART for example, larger sample sizes are needed to fit the model.
+How much larger likely depends on the complexity of the specified model: Our simulations show that with fixed and random-effects specifications with only a single predictor, a sample size of 200 is sufficient to detect subgroups with moderate differences in treatment effect.
+More complex fixed- and random-effects specifications will require larger sample sizes; how these affect the performance of GLMM tree requires further research.
+
+In the Introduction we mentioned several existing tree-based methods for treatment-subgroup interaction detection.
+These methods have different objectives and there is not yet an agreed-upon single best method.
+In a simulation study, \citeA{SiesyVanM16} found the method of \citeA{ZhanyTsia12Stat} to perform best, followed by MOB.
 However, the method of Zhang et al.
 performed worst under some conditions of the simulation study in terms of the Type I error rate.
-Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations were limited to GLM and GLMM-based MOB.
-
-Furthermore, it should be stressed that tree-based methods are exploratory tools.
-They can be used to detect predictors, interactions and non-linear effects in a data-driven way, but users should take the exploratory nature of such analyses into account.
-The resulting trees are potentially unstable, and stability of the results should be assessed, preferably in a dataset not used for training, or by multi-fold cross-validation or resampling techniques.
-In the Application, we have shown how the stability of splitting variable selection can be assessed using resampling techniques.
-
-In conclusion, GLMM tree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatment effect differences, both in the presence and absence of cluster-specific random effects.
-Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials or individual-level patient data meta-analyses.
+Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations and comparisons only focused on GLMM tree and GLM-based MOB.
+
+\subsubsection{Conclusion}
+
+Our results indicate that GLMM tree provides accurate recovery of treatment-subgroup interactions and prediction of treatment effects, both in the presence and absence of random effects and interactions.
+Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, for example in multi-center trials or individual-level patient data meta-analyses.
 
@@
 
+\begin{appendix}
+
+\section{R code for generating artificial motivating dataset}
+
+\begin{small}
+<<appendixA_1, echo = TRUE>>=
+set.seed(123)
+treatment <- rbinom(n = 150, size = 1, prob = .5)
+duration <- round(rnorm(150, mean = 7, sd = 3))
+anxiety <- round(rnorm(150, mean = 10, sd = 3))
+age <- round(rnorm(150, mean = 45, sd = 10))
+error <- rnorm(150, 0, 2)
+cluster <- error + rnorm(150, 0, 6)
+rand_int <- sort(rep(rnorm(10, 0, 1), each = 15))
+rand_int[order(cluster)] <- rand_int 
+error <- error - rand_int
+cluster[order(cluster)] <- rep(1:10, each = 15)
+node3t1 <- ifelse(duration <= 8 & anxiety <= 10 & treatment == 0, -2, 0)
+node3t2 <- ifelse(duration <= 8 & anxiety <= 10 & treatment == 1, 2, 0)
+node5t1 <- ifelse(duration > 8 & treatment == 0, 2.5, 0)
+node5t2 <- ifelse(duration > 8 & treatment == 1, -2.5, 0)
+depression <- round(9 + node3t1 + node3t2 + node5t1 + node5t2 + .4*treatment + 
+                   error + rand_int)
+treatment <- factor(treatment, labels = c("Treatment 1", "Treatment 2"))
+a_m_data <- data.frame(depression, treatment, cluster, age, anxiety, duration)
+@
+\end{small}
+
+\end{appendix}
+
 \end{document}

### 2017-04-24 pkg/partykit/R/ctree.R
mv partykit 1.1-2 to partykitR1 for later reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1264 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,464 +1,0 @@
-
-### calculate quad p-values
-.MPinv <- function (X, tol = sqrt(.Machine$double.eps)) 
-{
-    if (length(dim(X)) > 2L || !(is.numeric(X) || is.complex(X))) 
-        stop("'X' must be a numeric or complex matrix")
-    if (!is.matrix(X)) 
-        X <- as.matrix(X)
-    Xsvd <- svd(X)
-    if (is.complex(X)) 
-        Xsvd$u <- Conj(Xsvd$u)
-    Positive <- Xsvd$d > max(tol * Xsvd$d[1L], 0)
-    Xplus <- if (all(Positive)) 
-        Xsvd$v %*% (1/Xsvd$d * t(Xsvd$u))
-    else if (!any(Positive)) 
-        array(0, dim(X)[2L:1L])
-    else Xsvd$v[, Positive, drop = FALSE] %*% ((1/Xsvd$d[Positive]) * 
-        t(Xsvd$u[, Positive, drop = FALSE]))
-    list(Xplus = Xplus, rank = sum(Positive))
-}
-
-.pX2 <- function(lin, exp, cov, pval = TRUE) {
-    if (length(lin) == 1) {
-        if (cov < .Machine$double.eps) return(c(-Inf, -Inf))
-        X2 <- ((lin - exp)^2) / cov
-        df <- 1
-    } else {
-        v <- diag(V <- matrix(cov, ncol = length(lin)))
-        ### remove elements with zero variance first
-        lin <- as.vector(lin)[v > 0]
-        exp <- as.vector(exp)[v > 0]
-        V <- V[v > 0, v > 0, drop = FALSE]
-        v <- v[v > 0]
-        if (length(v) == 0) return(c(-Inf, -Inf))
-        tmp <- matrix(lin - exp, ncol = 1)
-        Xplus <- .MPinv(V)
-        X2 <- crossprod(tmp, Xplus$Xplus) %*% tmp
-        df <- Xplus$rank
-    }
-    
-    X2 <- pmax(0, X2) ## Z: X2 may be slightly negative...
-    
-    if (pval)
-        return(c(log(X2), pchisq(X2, df = df, lower.tail = TRUE, 
-                                 log.p = TRUE)))
-    return(c(log(X2), NA))
-}
-
-### calculate max-T p-value
-.pmaxT <- function(lin, exp, cov, pval = TRUE) {
-
-    if (length(lin) == 1) {
[…]

### 2017-04-24 pkg/partykit/R/glmtree.R

@@ -1,144 +1,0 @@
-## simple wrapper function to specify fitter and return class
-glmtree <- function(formula, data, subset, na.action, weights, offset, cluster,
-  family = gaussian, epsilon = 1e-8, maxit = 25, ...)
-{
-  ## use dots for setting up mob_control
-  control <- mob_control(...)
-
-  ## keep call
-  cl <- match.call(expand.dots = TRUE)
-
-  ## extend formula if necessary
-  f <- Formula::Formula(formula)
-  if(length(f)[2L] == 1L) {
-    attr(f, "rhs") <- c(list(1), attr(f, "rhs"))
-    formula[[3L]] <- formula(f)[[3L]]
-  } else {
-    f <- NULL
-  }
-
-  ## process family
-  if(inherits(family, "family")) {
-    fam <- TRUE
-  } else {
-    fam <- FALSE
-    if(is.character(family)) family <- get(family)
-    if(is.function(family)) family <- family()
-  }
-
-  ## call mob
-  m <- match.call(expand.dots = FALSE)
-  if(!is.null(f)) m$formula <- formula
-  m$fit <- glmfit
-  m$control <- control
-  m$epsilon <- epsilon
-  m$maxit <- maxit
-  if("..." %in% names(m)) m[["..."]] <- NULL
-  if(!fam) m$family <- family
-  m[[1L]] <- as.call(quote(partykit::mob))
-  rval <- eval(m, parent.frame())
-  
-  ## extend class and keep original call
-  rval$info$call <- cl
-  rval$info$family <- family$family
-  class(rval) <- c("glmtree", class(rval))
-  return(rval)
-}
-
-## actual fitting function for mob()
-glmfit <- function(y, x, start = NULL, weights = NULL, offset = NULL, cluster = NULL, ...,
-  estfun = FALSE, object = FALSE)
-{
-  ## catch control arguments
-  args <- list(...)
-  ctrl <- list()
-  for(n in c("epsilon", "maxit")) {
-    if(n %in% names(args)) {
-      ctrl[[n]] <- args[[n]]
-      args[[n]] <- NULL
-    }
-  }
-  args$control <- do.call("glm.control", ctrl)
-  
-  ## add intercept-only regressor matrix (if missing)
-  ## NOTE: does not have terms/formula
-  if(is.null(x)) x <- matrix(1, nrow = NROW(y), ncol = 1L,
-    dimnames = list(NULL, "(Intercept)"))
-  
[…]

### 2017-04-24 pkg/partykit/R/palmtree.R

@@ -1,191 +1,0 @@
-utils::globalVariables(c(".tree", ".lm", ".weights"))
-
-palmtree <- function(formula, data, weights = NULL, family = NULL,
-  lmstart = NULL, abstol = 0.001, maxit = 100, 
-  dfsplit = TRUE, verbose = FALSE, plot = FALSE, ...)
-{
-  ## remember call
-  cl <- match.call()
-
-  ## process family
-  if(!is.null(family)) {
-    if(!inherits(family, "family")) {
-      if(is.character(family)) family <- get(family)
-      if(is.function(family)) family <- family()
-    }
-  }
-  
-  ## formula processing (full, tree, regression)
-  ff <- Formula::as.Formula(formula)
-  tf <- formula(ff, lhs = 1L, rhs = c(1L, 3L))
-  rf <- formula(ff, lhs = 1L, rhs = 1L)
-  ## intercept in lmtree?
-  intercept <- attr(terms(rf), "intercept") > 0L
-  if(!intercept) rf <- update(rf, . ~ . + 1)
-  ## without tree
-  rf0 <- formula(Formula::as.Formula(rf, formula(ff, lhs = 0L, rhs = 2L)),
-    lhs = 1L, rhs = c(1L, 2L), collapse = TRUE)
-  ## with tree
-  rf <- if(intercept) update(rf, . ~ .tree / .) else update(rf, . ~ .tree:.)
-  rf <- formula(Formula::as.Formula(rf, formula(ff, lhs = 0L, rhs = 2L)),
-    lhs = 1L, rhs = c(1L, 2L), collapse = TRUE)
-
-  ## weights
-  data$.weights <- if(is.null(weights)) rep(1, nrow(data)) else weights
-
-  ## initialization
-  iteration <- 0L
-  if (is.null(lmstart)) {
-    lmstart <- if(is.null(family)) {
-      lm(rf0, data = data, weights = .weights)
-    } else {
-      glm(rf0, data = data, weights = .weights, family = family)
-    }
-    lmstart <- predict(lmstart)
-  }
-  data$.lm <- lmstart  
-
-  continue <- TRUE
-  oldloglik <- -Inf
-
-  ## iterate between palm and lmtree estimation
-  while (continue) {
-    iteration <- iteration + 1L
-
-    ## lmtree
-    tree <- if(is.null(family)) {
-      lmtree(tf, data = data, offset = .lm, weights = .weights, dfsplit = FALSE, ...)
-    } else {
[…]

### 2017-04-24 pkg/partykit/tests/bugfixes.R

@@ -1,766 +1,0 @@
-
-datLB <-
-structure(list(Site = c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
-1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
-3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 
-4L, 4L, 4L, 4L, 5L, 5L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 6L, 6L, 6L, 
-6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
-7L, 7L, 7L, 7L, 7L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 
-8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 
-9L, 9L, 9L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 
-2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 
-3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
-4L, 5L, 5L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 
-6L, 6L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
-7L, 7L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 
-8L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 
-1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 
[…]

### 2017-04-24 pkg/partykit/tests/ctree-ctree.R

@@ -1,163 +1,0 @@
-
-### compare partykit::ctree and party::ctree wrt predictions
-### and computing time. But not during R CMD check
-
-library("partykit")
-library("party")
-
-set.seed(29)
-
-n <- 200
-dgp <- function(n, response = c("numeric", "ordered", "binary", "factor"), na = FALSE) {
-
-    response <- match.arg(response)
-    x <- runif(n * 10)
-    x <- matrix(x, ncol = 10)
-    x <- cbind(x, x[,1:2] + cbind(rnorm(n, sd = 0.1), rnorm(n, sd = 0.1)))
-    x1 <- sample(gl(4, n / 4))
-    x2 <- sample(ordered(gl(4, n / 4)))
-    y <- as.integer(interaction(x[,1] > 0.5, x[,2] > 0.25))
-    y <- rnorm(n, mean = y)
-    if (na) x[sample(1:(n * 10), n)] <- NA
-    if (response == "numeric")
-        return(data.frame(x, x1, x2, y = y))
-    if (response == "binary") {
-        y <- factor(y < median(y))
-        return(data.frame(x, x1, x2, y = y))
-    } else {
-        y <- cut(y, 
-            breaks = c(-Inf, quantile(y, c(0.25, 0.5, 0.75)), Inf))
-    }
-    if (response == "ordered")
-        return(data.frame(x, x1, x2, y = ordered(y)))
-    if (response == "factor")
-        return(data.frame(x, x1, x2, y = y))
-}
-
-args <- list(response = c("numeric", "ordered", "binary", "factor"),
-             na = c(FALSE, TRUE),
-             teststat = c("quad", "max"),
-             testtype = c("Bonferroni", "Univariate", "Teststatistic"),
-             mincriterion = c(0.8, 0.9, 0.95),
-             maxsurrogate = 0:2, maxdepth = 3)
-
-gr <- do.call("expand.grid", args)
-tgr <- gr
-gr[["response"]] <- as.character(gr[["response"]])
-gr[["teststat"]] <- as.character(gr[["teststat"]])
-gr[["testtype"]] <- as.character(gr[["testtype"]])
-
-### random splits are done differently in party::ctree
-gr <- subset(gr, !na | (na & maxsurrogate > 0))
-
-
-fun <- function(args) {
-
-    learn <- dgp(n, response = args$response, na = args$na)
[…]

### 2017-04-24 pkg/partykitR1/R/meanvartree.R

@@ -1,0 +1,165 @@
+meanvarfit <- function(y, x = NULL, start = NULL, weights = NULL, offset = NULL,
+  ..., estfun = FALSE, object = FALSE)
+{
+  m <- mean(y * weights)/mean(weights)
+  res <- y - m
+  s2 <- mean(res^2 * weights)/mean(weights)
+
+  rval <- list(
+    coefficients = structure(c(m, log(s2)), .Names = c("mean", "log(variance)")),
+    objfun = -sum(weights * dnorm(y, mean = m, sd = sqrt(s2), log = TRUE)),
+    estfun = if(estfun) cbind(res, (res^2 - s2)/2) * (1/s2) * weights else NULL,
+    object = NULL
+  )
+}
+
+meanvartree <- function(formula, data, na.action, weights, subset, minsplit = 10L, ...)
+{
+  ## keep call
+  cl <- match.call(expand.dots = TRUE)
+
+  ## use dots for setting up mob_control
+  control <- mob_control(minsplit = minsplit, ...)
+
+  ## call mob
+  m <- match.call(expand.dots = FALSE)
+  stopifnot(require("Formula"))
+  m$formula <- formula(as.Formula(formula, ~ 1), rhs = 2:1)
+  m$fit <- meanvarfit
+  m$control <- control
+  m$minsplit <- NULL
+  if("..." %in% names(m)) m[["..."]] <- NULL
+  m[[1L]] <- as.name("mob")
+  rval <- eval(m, parent.frame())
+  
+  ## extend class and keep original call
+  rval$info$call <- cl
+  rval <- as.constparty(rval)
+  class(rval) <- c("constparty", "party", "modelparty")
+  return(rval)
+}
+
+mixforest <- function(formula, data, na.action, weights, subset,
+  ntree = 500L, mtry = 3L, minsplit = 10L, replace = TRUE, fraction = 0.632, ...)
+{
+  ## call and formula
+  cl <- match.call()
+  if(missing(data)) data <- environment(formula)
+  mf <- match.call(expand.dots = FALSE)
+  m <- match(c("formula", "data", "subset", "na.action", "weights"), names(mf), 0)
+  mf <- mf[c(1L, m)]
[…]

### 2017-04-24 pkg/partykitR1/R/mtree.R

@@ -1,0 +1,293 @@
+mtree_ols <- list(
+  fit = function(y, x = NULL, start = NULL, weights = NULL, offset = NULL,
+    ..., estfun = FALSE, object = FALSE)
+  {
+    if(!is.null(offset)) y <- y - offset
+    m <- mean(y * weights)/mean(weights)
+    res <- y - m
+    rss <- sum(res^2 * weights)
+    
+    list(
+      coefficients = c("mean" = m),
+      objfun = rss,
+      estfun = if(estfun) res * weights else NULL,
+      object = if(object) list(nuisance = c("log(variance)" = log(rss/sum(weights)))) else NULL
+    )
+  },
+  
+  ddist = function(x, par, log = FALSE) {
+    par <- matrix(par, ncol = 2L)
+    dnorm(x, mean = par[, 1L], sd = sqrt(exp(par[, 2L])), log = log)
+  },
+
+  pdist = function(q, par, lower.tail = TRUE, log.p = FALSE) {
+    par <- matrix(par, ncol = 2L)
+    pnorm(q, mean = par[, 1L], sd = sqrt(exp(par[, 2L])), lower.tail = lower.tail, log = log)
+  },
+
+  qdist = function(p, par, lower.tail = TRUE, log.p = FALSE) {
+    par <- matrix(par, ncol = 2L)
+    qnorm(p, mean = par[, 1L], sd = sqrt(exp(par[, 2L])), lower.tail = lower.tail, log = log)
+  }
+)
+
+mtree_norm <- list(
+  fit = function(y, x = NULL, start = NULL, weights = NULL, offset = NULL,
+    ..., estfun = FALSE, object = FALSE)
+  {
+    if(!is.null(offset)) y <- y - offset
+    m <- mean(y * weights)/mean(weights)
+    res <- y - m
+    s2 <- mean(res^2 * weights)/mean(weights)
+
+    list(
+      coefficients = structure(c(m, log(s2)), .Names = c("mean", "log(variance)")),
+      objfun = -sum(weights * dnorm(y, mean = m, sd = sqrt(s2), log = TRUE)),
+      estfun = if(estfun) cbind(res, (res^2 - s2)/2) * (1/s2) * weights else NULL,
+      object = NULL
+    )
+  },
+  
+  ddist = function(x, par, log = FALSE) {
+    par <- matrix(par, ncol = 2L)
[…]

### 2017-04-24 pkg/partykitR1/tests/bugfixes.R

@@ -1,0 +1,766 @@
+
+datLB <-
+structure(list(Site = c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
+1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
+3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 
+4L, 4L, 4L, 4L, 5L, 5L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 6L, 6L, 6L, 
+6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
+7L, 7L, 7L, 7L, 7L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 
+8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 
+9L, 9L, 9L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 
+2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 
+3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
+4L, 5L, 5L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 
+6L, 6L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
+7L, 7L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 
+8L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 
+1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 
[…]

### 2017-04-24 pkg/partykitR1/tests/ctree-ctree.R

@@ -1,0 +1,163 @@
+
+### compare partykit::ctree and party::ctree wrt predictions
+### and computing time. But not during R CMD check
+
+library("partykitR1")
+library("party")
+
+set.seed(29)
+
+n <- 200
+dgp <- function(n, response = c("numeric", "ordered", "binary", "factor"), na = FALSE) {
+
+    response <- match.arg(response)
+    x <- runif(n * 10)
+    x <- matrix(x, ncol = 10)
+    x <- cbind(x, x[,1:2] + cbind(rnorm(n, sd = 0.1), rnorm(n, sd = 0.1)))
+    x1 <- sample(gl(4, n / 4))
+    x2 <- sample(ordered(gl(4, n / 4)))
+    y <- as.integer(interaction(x[,1] > 0.5, x[,2] > 0.25))
+    y <- rnorm(n, mean = y)
+    if (na) x[sample(1:(n * 10), n)] <- NA
+    if (response == "numeric")
+        return(data.frame(x, x1, x2, y = y))
+    if (response == "binary") {
+        y <- factor(y < median(y))
+        return(data.frame(x, x1, x2, y = y))
+    } else {
+        y <- cut(y, 
+            breaks = c(-Inf, quantile(y, c(0.25, 0.5, 0.75)), Inf))
+    }
+    if (response == "ordered")
+        return(data.frame(x, x1, x2, y = ordered(y)))
+    if (response == "factor")
+        return(data.frame(x, x1, x2, y = y))
+}
+
+args <- list(response = c("numeric", "ordered", "binary", "factor"),
+             na = c(FALSE, TRUE),
+             teststat = c("quad", "max"),
+             testtype = c("Bonferroni", "Univariate", "Teststatistic"),
+             mincriterion = c(0.8, 0.9, 0.95),
+             maxsurrogate = 0:2, maxdepth = 3)
+
+gr <- do.call("expand.grid", args)
+tgr <- gr
+gr[["response"]] <- as.character(gr[["response"]])
+gr[["teststat"]] <- as.character(gr[["teststat"]])
+gr[["testtype"]] <- as.character(gr[["testtype"]])
+
+### random splits are done differently in party::ctree
+gr <- subset(gr, !na | (na & maxsurrogate > 0))
+
+
+fun <- function(args) {
+
+    learn <- dgp(n, response = args$response, na = args$na)
[…]

### 2017-04-24 pkg/partykitR1/tests/regtest-split.R

@@ -1,0 +1,72 @@
+library("partykitR1")
+set.seed(1)
+
+dat <- data.frame(v1 = as.double(1:100))
+
+sv1 <- partysplit(as.integer(1), breaks = as.double(50))
+character_split(sv1, dat)
+stopifnot(all(kidids_split(sv1, dat) == ((dat$v1 > 50) + 1)))
+
+sv1 <- partysplit(as.integer(1), breaks = as.double(50), 
+                index = as.integer(c(2, 1)))
+character_split(sv1, dat)
+stopifnot(all(kidids_split(sv1, dat) == ((dat$v1 <= 50) + 1)))
+
+sv1 <- partysplit(as.integer(1), breaks = as.double(50), right = FALSE)
+character_split(sv1, dat)
+stopifnot(all(kidids_split(sv1, dat) == ((dat$v1 >= 50) + 1)))
+
+sv1 <- partysplit(as.integer(1), breaks = as.double(50), 
+                index = as.integer(c(2, 1)), right = FALSE)
+character_split(sv1, dat)
+stopifnot(all(kidids_split(sv1, dat) == ((dat$v1 < 50) + 1)))
+
+sv1 <- partysplit(as.integer(1), breaks = as.double(c(25, 75)))
+character_split(sv1, dat)
+stopifnot(all(kidids_split(sv1, dat) == 
+              as.integer(cut(dat$v1, c(-Inf, 25, 75, Inf)))))
+
+sv1 <- partysplit(as.integer(1), breaks = as.double(c(25, 75)), right = FALSE)
+character_split(sv1, dat)
+stopifnot(all(kidids_split(sv1, dat) == 
+              as.integer(cut(dat$v1, c(-Inf, c(25, 75), Inf), right = FALSE))))
+
+sv1 <- partysplit(as.integer(1), breaks = as.double(c(25, 75)), 
+                index = as.integer(3:1), right = FALSE)
+character_split(sv1, dat)
+stopifnot(all(kidids_split(sv1, dat) == 
+              (3:1)[as.integer(cut(dat$v1, c(-Inf, c(25, 75), Inf), right = FALSE))]))
+
+
+dat$v2 <- gl(4, 25)
+
+sv2 <- partysplit(as.integer(2), index = as.integer(c(1, 2, 1, 2)))
+character_split(sv2, dat)
+kidids_split(sv2, dat)
+
[…]

### 2017-04-24 pkg/partykit/R/modelparty_R1.R
mv devel version to pkg/partykit; see README.txt
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1269 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,0 +1,1126 @@
+mob <- function(formula, data, subset, na.action, weights, offset, cluster,
+  fit, control = mob_control(), ...)
+{
+  ## check fitting function
+  fitargs <- names(formals(fit))
+  if(!all(c("y", "x", "start", "weights", "offset") %in% fitargs)) {
+    stop("no suitable fitting function specified")
+  }
+
+  ## augment fitting function (if necessary)
+  if(!all(c("estfun", "object") %in% fitargs)) {
+    afit <- function(y,
+      x = NULL, start = NULL, weights = NULL, offset = NULL, cluster = NULL, ...,
+      estfun = FALSE, object = FALSE)
+    {
+      obj <- if("cluster" %in% fitargs) {
+        fit(y = y, x = x, start = start, weights = weights, offset = offset, cluster = cluster, ...)
+      } else {
+        fit(y = y, x = x, start = start, weights = weights, offset = offset, ...)
+      }
+      list(
+        coefficients = coef(obj),
+        objfun = -as.numeric(logLik(obj)),
+        estfun = if(estfun) sandwich::estfun(obj) else NULL,
+        object = if(object) obj else NULL
+      )
+    }
+  } else {
+    if("cluster" %in% fitargs) {
+      afit <- fit
+    } else {
+      afit <- function(y, x = NULL, start = NULL, weights = NULL, offset = NULL, cluster = NULL, ..., estfun = FALSE, object = FALSE) {
+        fit(y = y, x = x, start = start, weights = weights, offset = offset, ..., estfun = estfun, object = object)
+      }
+    }
+  }
+
+  ## call
+  cl <- match.call()
+  if(missing(data)) data <- environment(formula)
+  mf <- match.call(expand.dots = FALSE)
+  m <- match(c("formula", "data", "subset", "na.action", "weights", "offset", "cluster"), names(mf), 0L)
+  mf <- mf[c(1L, m)]
+  mf$drop.unused.levels <- TRUE
+
+  ## formula FIXME: y ~ . or y ~ x | .
+  oformula <- as.formula(formula)
+  formula <- Formula::as.Formula(formula)
+  if(length(formula)[2L] < 2L) {
+    formula <- Formula::Formula(formula(Formula::as.Formula(formula(formula), ~ 0), rhs = 2L:1L))
[…]

### 2017-04-24 pkg/partykit/R/party_R1.R

@@ -1,663 +1,0 @@
-## FIXME: data in party
-##   - currently assumed to be a data.frame
-##   - potentially empty
-##   - the following are all assumed to work:
-##     dim(data), names(data)
-##     sapply(data, class), lapply(data, levels)
-##   - potentially these need to be modified if data/terms
-##     should be able to deal with data bases
-
-party <- function(node, data, fitted = NULL, terms = NULL, names = NULL, info = NULL) {
-
-    stopifnot(inherits(node, "partynode"))
-    stopifnot(inherits(data, "data.frame"))
-    ### make sure all split variables are there 
-    ids <- nodeids(node)[!nodeids(node) %in% nodeids(node, terminal = TRUE)]
-    varids <- unique(unlist(nodeapply(node, ids = ids, FUN = function(x) 
-        varid_split(split_node(x)))))
-    stopifnot(varids %in% 1:ncol(data))
-    
-    if(!is.null(fitted)) {
-        stopifnot(inherits(fitted, "data.frame"))
-        stopifnot(nrow(data) == 0L | nrow(data) == nrow(fitted))
-        
-	# try to provide default variable "(fitted)"
-	if(nrow(data) > 0L) {
-          if(!("(fitted)" %in% names(fitted))) 
-            fitted[["(fitted)"]] <- fitted_node(node, data = data)
-	} else {
-	  stopifnot("(fitted)" == names(fitted)[1L])
-	}
-
-        nt <- nodeids(node, terminal = TRUE)
-        stopifnot(all(fitted[["(fitted)"]] %in% nt))
-
-        node <- as.partynode(node, from = 1L)
-        nt2 <- nodeids(node, terminal = TRUE)
-        fitted[["(fitted)"]] <- nt2[match(fitted[["(fitted)"]], nt)]
-    } else {
-        node <- as.partynode(node, from = 1L)
-	# default "(fitted)"
-	if(nrow(data) > 0L & missing(fitted)) 
-          fitted <- data.frame("(fitted)" = fitted_node(node, 
-            data = data), check.names = FALSE)
-    }
-    
-    party <- list(node = node, data = data, fitted = fitted, 
-                  terms = NULL, names = NULL, info = info)
-    class(party) <- "party"
-
-    if(!is.null(terms)) {
-        stopifnot(inherits(terms, "terms"))
-	party$terms <- terms
-    }
-
-    if (!is.null(names)) {
-        n <- length(nodeids(party, terminal = FALSE))
[…]

### 2017-04-24 pkg/glmertree/paper/treatment_interactions.Rnw
Paper update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1272 255f4249-2263-4417-b1ea-8498779e27a7

@@
 <<simulation-data>>=
-load("treespecs_study1.dat")
+load("treespecs_studyI.dat")
 treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered = TRUE)
@@
 <<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
-load("noint_treesizes_long.dat")
+load("treespecs_studyII.dat")
 treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
@@
 <<correlation-pw-xyplot, fig=TRUE, width=9, height=7>>=
-load("pw_treespecs_long.dat")
+load("treespecs_studyIII.dat")
 treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)

### 2017-04-26 pkg/partykit/R/mob-methods_R2.R
model.frame.modelparty always returns _model.frame_ now; use <obj> if get_all_vars is what you want
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1280 255f4249-2263-4417-b1ea-8498779e27a7

@@ -9,8 +9,11 @@ getCall.modelparty <- function(x, ...) x$info$call
 
 model.frame.modelparty <- function(formula, ...)
 {
+  ### formula$data is get_all_vars, not model.frame
   mf <- formula$data
-  if(nrow(mf) > 0L) return(mf)
+  ### always compute model.frame
+  if(nrow(mf) > 0L) 
+      return(model.frame(formula$info$Formula, data = mf))
 
   dots <- list(...)
   nargs <- dots[match(c("data", "na.action", "subset"), names(dots), 0L)]

### 2017-04-26 pkg/partykit/R/mob-plot_R2.R

@@ -7,7 +7,7 @@ node_bivplot <- function(mobobj, which = NULL, id = TRUE, pop = TRUE,
                          mainlab = NULL, ...)
 {
   ## obtain dependent variable
-  mf <- model.frame(mobobj$info$Formula, data = model.frame(mobobj))
+  mf <- model.frame(mobobj)
   y <- Formula::model.part(mobobj$info$Formula, mf, lhs = 1L, rhs = 0L)
   if(isTRUE(ylab)) ylab <- names(y)
   if(identical(ylab, FALSE)) ylab <- ""

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
small touch-ups in abstract and intro -> tutorial already mentioned to be in vignette only
try to use "GLMM trees" (plural) or "GLMM tree algorithm" (singular) more
consistently

a few further style touch ups: Figure~\ref{} / node~X / minus in math mode



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1294 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research.
-Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset.
+Therefore, we propose the generalized linear mixed-effects model tree (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset.
 The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters.
-In a short tutorial, we show how GLMM trees can be fit with the R package \verb|glmertree|.
-In a simulation study, GLMM tree showed higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees.
-Also, GLMM tree showed somewhat higher predictive accuracy than linear mixed-effects model with pre-specified interaction effect, on average.
-We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression.
-We conclude that GLMM tree is a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
+In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees.
+Also, GLMM tree show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effect, on average.
+We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression.
+We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.
+An implementation is available in the R package \verb|glmertree| and a short tutorial demonstrating how to fit GLMM trees in practice is provided in the accompanying vignette.\\
 \\
@@
 In research on the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized or stratified medicine is becoming increasingly important.
-Stratified medicine presents the challenge of finding which patients respond best to which treatments.
+Stratified medicine presents the challenge of discovering which patients respond best to which treatments.
 This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}.
@@
 
-Due to the recursive nature of the splitting, the rectangular regions of the partition can be graphically depicted as nodes in a decision tree, as shown in the example in Figure \ref{fig:exampletree}.
-The partition in Figure \ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety.
+Due to the recursive nature of the splitting, the rectangular regions of the partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}.
+The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety.
 The resulting tree has a depth of two, as the longest path travels along two splits.
 Each of the splits in the tree is defined by a splitting variable and value.
-The first split in the tree separates the observations into two subgroups, based on the splitting variable duration and a splitting value of 8, yielding two rectangular regions, represented by node 2 and node 3.
-As the observations in node 2 are not further split, node 2 is a terminal node.
-Node 3 is an inner node, as the observations in this node are further split into nodes 4 and 5, based on the anxiety variable.
-
-If the partition in Figure \ref{fig:exampletree} would be used for prediction of a new observation, the new observation would be assigned to one of the terminal nodes according to its values on the splitting variables.
+The first split in the tree separates the observations into two subgroups, based on the splitting variable duration and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3.
+As the observations in node~2 are not further split, node~2 is a terminal node.
+Node~3 is an inner node, as the observations in this node are further split into nodes~4 and~5, based on the anxiety variable.
+
+If the partition in Figure~\ref{fig:exampletree} would be used for prediction of a new observation, the new observation would be assigned to one of the terminal nodes according to its values on the splitting variables.
 The prediction is then determined by the distribution of the training observations within the terminal node.
 For example, the prediction may be the node-specific mean of a single continuous variable.
-In the current paper, we focus on trees where the terminal nodes consist of a linear (LM) or generalized linear model (GLM), in which case the predicted value for a new observation is determined by the node-specific parameter estimates of the (G)LM.
+In the current paper, we focus on trees where the terminal nodes consist of a linear (LM) or generalized linear model (GLM), in which case the predicted value for a new observation is determined by the node-specific parameter estimates of the (G)LM, while also adjusting for random effects.
 
@@
 However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset.
-Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of interactions and non-linearities in GLMM type models: generalized linear mixed-effects model tree, or GLMM tree.
+Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM trees.
 
@@
 In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}, GLMM tree allows for taking into account the clustered structure of datasets.
-In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.
+In contrast to previously suggested regression trees with random effects \cite<e.g., >{HajjyBell11,SelaySimo12}, GLMM trees allow for treatment effect estimation, with continuous as well as non-continuous response variables.
 
 The remainder of this paper is structured into four sections: In the first section, we introduce the GLMM tree algorithm using an artificial motivating dataset with treatment-subgroup interactions.
-The second section is a short tutorial on how to fit GLMM trees using the R package \verb|glmertree|.
-In the third section, we compare the performance of GLMM tree with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions.
-In the fourth section, we apply GLMM tree to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression.
+The second section is a short tutorial on how to fit GLMM trees using the R package \verb|glmertree| \readme{still needs to be adapted: vignette}.
+In the third section, we compare the performance of GLMM trees with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions.
+In the fourth section, we apply a GLMM tree to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression.
 In the fith and last section we summarize the results and discuss limitations and directions for future research.
@@
 
-The outcome variable was generated such that there were three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure \ref{fig:exampletree}: The first subgroup of patients corresponds to node~3 and shows lower post-treatment depression in Treatment~1 than in Treatment~2.
+The outcome variable was generated such that there were three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: The first subgroup of patients corresponds to node~3 and shows lower post-treatment depression in Treatment~1 than in Treatment~2.
 The second subgroup of patients corresponds to node~4 and shows about equal post-treatment depression in both treatment conditions.
@@
 
-Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part $x_{i}^{\top}\beta$ may not describe the data well.
-Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} while still adjusting for random effects:
+Although the random-effects part of the GLMM in Equation~\ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part $x_{i}^{\top}\beta$ may not describe the data well.
+Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation~\ref{eq:fixedeffects_MOB} while still adjusting for random effects:
 %
@@
 <<eval = FALSE, echo = TRUE>>=
-data(a_m_data)
+data("a_m_data", package = "glmertree")
 summary(a_m_data)
@@
 Also, we specified only a single variable in the random-effects part, resulting in estimation of a random intercept with respect to \verb|cluster|.
-More complex random effects can also be specified: for example, specifying the random-effects part as \verb=| (1 + age | cluster) |= would yield a model with a random intercept as well as a random slope for age with respect to cluster.
+More complex random effects can also be specified: for example, specifying the random-effects part as \verb=...
+| (1 + age | cluster) | ...= would yield a model with a random intercept as well as a random slope for age with respect to cluster.
 The brackets are necessary to protect the vertical bars in the random effects formulation.
 
-Now, we can use the \verb|plot| method to recreate the tree depicted in Figure \ref{fig:example_glimmertree}:
+Now, we can use the \verb|plot| method to recreate the tree depicted in Figure~\ref{fig:example_glimmertree}:
 
@@
 
-For MERT, the number of nodes and tree accuracy was not assessed, as the treatment-subgroup interaction design in Figure \ref{fig:modelC} corresponds to a large number of regression tree structures, that would all be different but also correct.
+For MERT, the number of nodes and tree accuracy was not assessed, as the treatment-subgroup interaction design in Figure~\ref{fig:modelC} corresponds to a large number of regression tree structures, that would all be different but also correct.
 Therefore, performance of MERT was only assessed in terms of predictive accuracy.
@@
 LM tree and MERT showed lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively.
-The most important predictors of predictive accuracy are depicted in Figure \ref{fig:xyplot_correlations}.
+The most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlations}.
 Performance of all three algorithms improves with increasing sample size and treatment-effect differences.
@@
 
-For datasets with purely piecewise interactions, the same partition as in Study I (Figure \ref{fig:modelC}) was used.
+For datasets with purely piecewise interactions, the same partition as in Study I (Figure~\ref{fig:modelC}) was used.
 In other words, the outcome variable in this design was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
 
-For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure \ref{fig:modelC} was also used.
+For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure~\ref{fig:modelC} was also used.
 However, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes now comprised continuous main and (treatment) interaction effects of the partitioning variables.
-The corresponding node-specific parameters are presented in Table \ref{tab:continuous_terms}.
-The $\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure \ref{fig:modelC}.
+The corresponding node-specific parameters are presented in Table~\ref{tab:continuous_terms}.
+The $\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure~\ref{fig:modelC}.
 The interaction terms were created using centered $U_k$ variables, calculated by subtracting their variable means.
 
-In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by the single $\beta$ column in Table \ref{tab:continuous_terms}.
+In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by the single $\beta$ column in Table~\ref{tab:continuous_terms}.
 
@@
 \thickline
-Term							&	$\beta_3$&	$\beta_4$&	$\beta_6$&	$\beta_7$&&	$\beta$		\\
-\hline
+Term					&	$\beta_3$&	$\beta_4$		&	$\beta_6$&	$\beta_7$		&&	$\beta$		\\ \hline
 intercept				&	27		&	27		&	27		&	27		&&	27		\\
 $U_2$					&	.1		&	.1		&	.1		&	.1		&&	.1		\\
-$U_2 \cdot U_1$ 				&	-.357	&	0		&	0		&	0		&&	-.357		\\
+$U_2 \cdot U_1$ 			&	$-$.357		&	0		&	0		&	0		&&	$-$.357		\\
 $U_2 \cdot U_5$	 			&	0		&	0		&	0		&	.357		&&	.357		\\
-$U_2 \cdot U_1 \cdot treatment$	&	-.151	&	0		&	0		&	0		&&	-.151		\\
-$U_2 \cdot U_5 \cdot treatment$	&	0		&	0		&	0		&	.151	&&	.151		\\
-\hline
-\multicolumn{7}{l}{\footnotesize{\textit{Note. }Subscripted $\beta$ values refer to the terminal nodes}}\\	
-\multicolumn{7}{l}{\footnotesize in Figure \ref{fig:modelC} for the combined piecewise and continuous}\\
+$U_2 \cdot U_1 \cdot \mbox{treatment}$	&	$-$.151		&	0		&	0		&	0		&&	$-$.151		\\
+$U_2 \cdot U_5 \cdot \mbox{treatment}$	&	0		&	0		&	0		&	.151		&&	.151		\\ \hline
+\multicolumn{7}{l}{\footnotesize{\emph{Note:} Subscripted $\beta$ values refer to the terminal nodes}}\\	
+\multicolumn{7}{l}{\footnotesize in Figure~\ref{fig:modelC} for the combined piecewise and continuous}\\
 \multicolumn{7}{l}{\footnotesize interaction design; $\beta$ without subscript refers to the global}\\
@@
 
-Table \ref{tab:stability} presents statistics on the variables selected for partitioning in subsamples of the dataset.
+Table~\ref{tab:stability} presents statistics on the variables selected for partitioning in subsamples of the dataset.
 Presence of a comorbid anxiety disorder was selected for partitioning in the majority of LMM trees grown on subsamples of the dataset, while the other variables were selected in at most 4\% of the subsamples.

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
improvements in methods section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1295 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. The R code used to generate the example dataset is provided in the Appendix.
+We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm.
+The R code used to generate the example dataset is provided in the Appendix \readme{I would put this into the appendix of the vignette instead.}.
 
@@
 Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment.
-Finally, each patient was part of one of ten clusters, each having a different value for the random intercept, uncorrelated to the partitioning variables.
-
-The outcome variable was generated such that there were three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: The first subgroup of patients corresponds to node~3 and shows lower post-treatment depression in Treatment~1 than in Treatment~2.
-The second subgroup of patients corresponds to node~4 and shows about equal post-treatment depression in both treatment conditions.
-The third subgroup of patients corresponds to node~5 and shows lower post-treatment depression in Treatment~2 than in Treatment~1.
+Finally, each patient was part of one of ten clusters, each having a different value for the random intercept, uncorrelated with the partitioning variables.
+
+The outcome variable was generated such that there are three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: For the first subgroup of patients (node~3) with short duration ($\le 8$) months of depressive symptoms prior to treatment and low anxiety scores ($\le 10$), Treatment~1 leads to lower post-treatment depression than in Treatment~2.
+For the second subgroup of patients (node~4) with short duration but high anxiety scores ($> 10$) post-treatment depression is about equal in both treatment conditions.
+For the third subgroup of patients (node~5) with long duration ($> 8$ months) Treatment~2 leads to lower post-treatment depression than Treatment~1.
 Thus, duration and anxiety are true partitioning or moderator variables, whereas age is not.
 Anticipating the final results of our analyses, the treatment-subgroup interactions are depicted in Figure~\ref{fig:example_glimmertree}, which shows the GLMM tree that accurately recovered the treatment-subgroup interactions.
+\readme{I've incorporated a characterization of the groups in terms of duration and anxiety here.
+But probably a better description of these variables is needed earlier.
+Possibly in a table with some summary statistics, e.g., mean/sd/range.
+Finally, the description in the manual Rd pages of the package should also be improved.}
 
@@
 
-The rationale behind MOB is that a global parametric model may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find better-fitting models in each cell of the partition.
-For example, to assess the effect of treatment, we may fit a global GLM using only the treatment indicator as a predictor of the outcome.
-The expectation $\mu_i$ of outcome $y_i$ given the treatment regressor $x_i$ is then modeled through a linear predictor and suitable link function:
+The rationale behind MOB is that a single global GLM (or other parametric model) may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find better-fitting models in each cell of the partition.
+For example, to assess the effect of treatment, we may first fit a global GLM where the treatment indicator has the same effect/coefficient on the outcome for all observations.
+Subsequently, the data may be partitioned recursively with respect to other covariates, leading to separate models with different treatment effects/coefficients in each subsample.
+\readme{Expanded on the treatment GLM MOB explanation to give more context and motivation.}
+
+More formally, in a single global GLM the expectation $\mu_i$ of outcome $y_i$ given the treatment regressor $x_i$ is modeled through a linear predictor and suitable link function:
 %
@@
 $\beta$ is a vector of fixed-effects regression coefficients.
-For simplicity, in the current paper we assume only two treatment groups and no further covariates in the GLM, so $x_{i}$ and $\beta$ each have length 2.
-The first element of $\beta$ then corresponds to the mean of the linear predictor in the first treatment group and the second element corresponds to the mean difference in the linear predictor between the first and second treatment groups.
+For simplicity, in the current paper we focus on two treatment groups and no further covariates in the GLM, so that in our illustrations $x_{i}$ and $\beta$ both have length 2.
 For the continuous response variable in the motivating data set, we employ the identity link function and assume a normal distribution for the error (denoted by $\epsilon_i = y_i - \mu_i$) with mean zero and variance $\sigma_{\epsilon}^2$.
-However, the model can easily accommodate additional treatment conditions and covariates, and binary or count outcome variables.
+Thus, the first element of $\beta$ then corresponds to the mean of the linear predictor in the first treatment group and the second element corresponds to the mean difference in the linear predictor between the first and second treatment groups.
+However, the model can easily accommodate additional treatment conditions and covariates, as well as binary or count/Poisson outcome variables.
 
@@
 Obviously, such a simple, global GLM will not fit the data well, especially in the presence of moderators.
-For illustrational purposes, however, we take it as a starting point to illustrate MOB.
+For expository purposes, however, we take it as a starting point to illustrate MOB.
 The global GLM fitted to the motivating example dataset is depicted in Figure~\ref{fig:fixedeffects}.
@@
 
-The MOB algorithm can be used to partition the dataset using additional covariates and find better fitting local models.
-To this end, the MOB algorithm tests for parameter stability over a set of auxiliary covariates, also known as \textit{partitioning variables}.
+The MOB algorithm can be used to partition the dataset using additional covariates and find better-fitting local models.
+To this end, the MOB algorithm tests for parameter stability over a set of auxiliary covariates, also called \textit{partitioning variables}.
 When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor.
@@
 As the distribution of these test statistics under the null hypothesis of parameter stability is known, a $p$-value for every partitioning variable can be calculated.
-Note that a more in-depth discussion of the parameter stability tests is outside the scope of this paper, but can be found in \citeA{ZeilyHorn07} and \citeA{ZeilyHoth08}.
+Note that a more in-depth discussion of the parameter stability tests is beyond the scope of this paper, but can be found in \citeA{ZeilyHorn07} and \citeA{ZeilyHoth08}.
 
 If at least one of the partitioning variables yields a $p$-value below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3).
-This partition is created using $U_{{k^*}}$, the partitioning variable with the minimal $p$-value in step (2).
+This partition is created using $U_{{k^*}}$ \readme{the U variables have not been introduced, no symbol has been used for the partitioning variables so far}, the partitioning variable with the minimal $p$-value in step (2).
 The split point for $U_{k^*}$ is selected by taking the value that minimizes the sum of the values of the objective function in both partitions.
@@
 Age was correctly not identified as a partitioning variable and the left- and rightmost nodes are in accordance with the true treatment-subgroup interactions described above.
-However, the two nodes in the middle do not represent true subgroups, possibly due to the dependence of observations within clusters not being taken into account.
+However, the two nodes in the middle have an unnecessary split and thus do not represent true subgroups, possibly due to the dependence of observations within clusters not being taken into account.
 
@@
 \caption{\small GLMM tree of the motivating example dataset.
-Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
+Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months) and age were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
 \label{fig:example_glimmertree}

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
additional remarks about the tutorial (vignette to-be)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1296 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+\readme{This tutorial should be turned into a vignette and then the R output, graphics, etc. should be shown as well.}
+
 We implemented the GLMM tree algorithm in the R package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
@@
 
-The main functions in the \verb|glmertree| package are \verb|lmtree()|, for continuous outcome variables, and \verb|glmertree()|, for binary or count outcome variables.
+The main functions in the \verb|glmertree| package are \verb|lmertree()|, for continuous outcome variables, and \verb|glmertree()|, for binary or count outcome variables.
 Both functions require the user to specify at least two arguments: \verb|formula| and \verb|data|.
@@
 The predictions of the random effects can be plotted by adding \verb|plotranef = TRUE| to the function call:
+\readme{As already indicated by e-mail, plotranef is an awkward name and should be improved, e.g., to which.}
 

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
small touch-ups in simulation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1297 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-To assess the performance of GLMM tree, we carried out three simulation studies: In Study I we assessed and compared the accuracy of GLMM tree, linear-model based MOB and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions.
-In Study II, we assessed and compared the Type-I error of GLMM tree and linear-model based MOB in datasets without treatment-subgroup interactions.
-In Study III, we assessed and compared the performance of GLMM tree and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions.
+To assess the performance of GLMM trees, we carried out three simulation studies: In Study~I we assessed and compared the accuracy of GLMM trees, linear-model based MOB and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions.
+In Study~II, we assessed and compared the Type-I error of GLMM trees and linear-model based MOB in datasets without treatment-subgroup interactions.
+In Study~III, we assessed and compared the performance of GLMM trees and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions.
 As the outcome variable was continuous in all simulated dataset, the GLMM tree algorithm and the trees resulting from its application will be referred to as LMM tree(s).
@@
 \item Number of clusters: $M=5$, $M=10$, $M=25$.
-\item Population SD of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
+\item Population standard deviation (SD) of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
 \item Intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
@@
 For estimating LMMs the \verb|lmer| function function from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
-The \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in LMMs in Study III.
+The \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in LMMs in Study~III.
 The \verb|lmerTest| package calculates effective degrees of freedom and $p$-values based on Satterthwaite approximations.
 
-\subsection{Study I: Performance of LMM tree, LM tree and MERT in datasets with treatment-subgroup interactions}
+\subsection{Study~I: Performance of LMM tree, LM tree and MERT in datasets with treatment-subgroup interactions}
 
@@
 
-For MERT, predicted treatment-effect differences were obtained by fitting two MERTS on the training data: one using observations in the first treatment condition and one using observations in the second treatment condition.
+For MERT, predicted treatment-effect differences were obtained by fitting two MERTs on the training data: one using observations in the first treatment condition and one using observations in the second treatment condition.
 Predictions of treatment-effect differences for test observations were obtained by dropping test observations down both trees and taking the difference between the two predicted values.
@@
 In the absence of random effects, this was also the case for LM trees.
-In the presence of random effects that are correlated to a (potential) partitioning variable, LM tree starts to create spurious splits, especially with larger $\sigma_b$ values.
+In the presence of random effects that are correlated to a (potential) partitioning variable, LM trees start to create spurious splits, especially with larger $\sigma_b$ values.
 In the presence of random effects that are uncorrelated to the other variables in the model, LM trees lack power to detect treatment-subgroup interactions if sample size is small (i.e., $N=200$).
@@
 The estimated probability that a dataset was erroneously not partitioned (Type-II error) was 0 for both algorithms.
-For the first split, LMM tree selected the true partitioning variable ($U_2$) in all datasets, and LM tree in all but one datasets.
+For the first split, LMM trees selected the true partitioning variable ($U_2$) in all datasets, and LM trees in all but one datasets.
 The mean splitting value of the first split was 29.94 for LM as well as LMM trees, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).
 
-Further splits were more accurately recovered by LMM tree, which accurately recovered the partition in 90.40\% of datasets, whereas LM tree accurately recovered the partition in 61.44\% of datasets.
+Further splits were more accurately recovered by LMM trees yielding 90.40\% accuracy for the full partition comparted to only 61.44\% for LM trees.
 The effects of the four most important predictors of tree accuracy are depicted in Figure~\ref{fig:xyplot_treeaccuracy}.
@@
 
-The predicted treatment-effect differences of LMM tree showed an average correlation of .93 (SD = .13) with the true differences.
-LM tree and MERT showed lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively.
-The most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlations}.
+The predicted treatment-effect differences of LMM tree show an average correlation of .93 (SD = .13) with the true differences.
+LM trees and MERT show lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively.
+The most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlations} \readme{Is it necessary to span the whole unit interval on the y-axis?
+Using 0.3 to 1.1 should be more than sufficient, I guess}.
 Performance of all three algorithms improves with increasing sample size and treatment-effect differences.
-Furthermore, LMM tree and MERT are not much affected by the presence and magnitude of random effects in the data.
-LMM tree performs most accurate in most conditions and is never outperformed by the other methods.
-MERT performs least accurate in most conditions and never outperforms the other methods.
+Furthermore, LMM trees and MERT are not much affected by the presence and magnitude of random effects in the data.
+LMM trees perform most accurately in most conditions and are never outperformed by the other methods.
+MERT performs least accurately in most conditions and never outperforms the other methods.
 The differences in accuracy become less pronounced with larger sample and effect sizes.
@@
 
-\subsection{Study II: Type-I error of LM and LMM trees}
+\subsection{Study~II: Type-I error of LM and LMM trees}
 
@@
 
-To assess the effect of the treatment-effect difference $\beta$, an additional facet wasadded to the \textit{General simulation design}: 
+To assess the effect of the treatment-effect difference $\beta$, an additional facet was added to the \textit{General simulation design}: 
 \begin{enumerate}
@@
 To assess the Type-I error rates of LM and LMM trees, tree sizes were calculated and trees of size $>1$ were classified as Type-I errors.
-The nominal Type-I error rate for both LM and LMM tree equals .05, corresponding to the pre-specified significance level $\alpha$ for the parameter instability tests.
+The nominal Type-I error rate for both LM and LMM trees equals .05, corresponding to the pre-specified significance level $\alpha$ for the parameter instability tests.
 
@@
 
-In datasets without treatment-subgroup interactions, average tree size was 1.09 ($\mathrm{SD}=0.44$) for LMM tree, and 2.02 ($\mathrm{SD}=1.68$) for LM tree.
-The average Type-I error rate was only .04 for LMM tree, and .33 for LM tree.
+In datasets without treatment-subgroup interactions, average tree size was 1.09 ($\mathrm{SD}=0.44$) for LMM trees, and 2.02 ($\mathrm{SD}=1.68$) for LM trees.
+The average Type-I error rate was only .04 for LMM trees, and .33 for LM trees.
 Main predictors of type-I error are depicted in Figure~\ref{fig:xyplot_treesize_nointeract}, which shows that LMM trees have a Type-I error rate somewhat below the pre-specified $\alpha$ level in all conditions.
-The same goes for LM tree, when random effects are absent, or uncorrelated to one of the partitioning covariates.
-When the random intercept is correlated with one of the potential partitioning covariates, the type-I error rapidly increases for LM tree.
-With increasing sample size or random-effects variance, LM tree will yield a larger number of spurious splits.
+The same goes for LM trees, when random effects are absent, or uncorrelated to one of the partitioning covariates.
+When the random intercept is correlated with one of the potential partitioning covariates, the type-I error rapidly increases for LM trees.
+With increasing sample size or random-effects variance, LM trees will yield a larger number of spurious splits.
 
@@
 
-\subsection{Study III: Recovery of piecewise and continuous interactions by LMM trees and LMMs with pre-specified interactions}
+\subsection{Study~III: Recovery of piecewise and continuous interactions by LMM trees and LMMs with pre-specified interactions}
 
@@
 
-The treatment-subgroup interactions in Study I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables.
+The treatment-subgroup interactions in Study~I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables.
 Trees are pre-eminently suited for recovering such piecewise interactions, but may have difficulty when the true interactions are continuous functions of moderator variables (for example, $U_1 \cdot U_2$).
@@
 
-For datasets with purely piecewise interactions, the same partition as in Study I (Figure~\ref{fig:modelC}) was used.
+For datasets with purely piecewise interactions, the same partition as in Study~I (Figure~\ref{fig:modelC}) was used.
 In other words, the outcome variable in this design was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
@@
 
-Furthermore, in this simulation study, the number of cells in the design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters), as Study I and II indicated no effects of the number of clusters.
+Furthermore, in this simulation study, the number of cells in the design was reduced by limiting the fourth facet of the data-generating design to a single level ($M=25$ clusters), as Study~I and II indicated no effects of the number of clusters.
 The fifth facet of the data-generating design was limited to two levels ($\sigma_b = 2.5$ and $\sigma_b = 7.5$).
@@
 
-On average, LMM tree showed somewhat higher accuracy: the average correlation between true and predicted treatment-effect differences was .54 (SD = .40) for LMM trees and .51 (SD = .43) for LMMs.
+On average, LMM trees showed somewhat higher accuracy: the average correlation between true and predicted treatment-effect differences was .54 (SD = .40) for LMM trees and .51 (SD = .43) for LMMs.
 The effects of the most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlation_pw}.
@@
 Performance of both methods improves with increasing sample size.
-Furthermore, performance of LMM tree is not affected by the number of covariates, whereas the predictive accuracy of LMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous.
-This indicates that LMM tree is especially useful for exploratory purposes, where there are many potential moderator variables.
+Furthermore, performance of LMM trees is not affected by the number of covariates, whereas the predictive accuracy of LMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous.
+This indicates that LMM trees are especially useful for exploratory purposes, where there are many potential moderator variables.
 
@@
 \paragraph{Dataset}
-To illustrate the application of GLMM tree on real data, we use a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}.
+To illustrate the use of GLMM trees in real application, we employ a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}.
 This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
comments and small touch-ups in application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1298 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-As the outcome variable is continuous, we employed an identity link and Gaussian error distribution.
-The tree resulting from GLMM tree will therefore be refered to as an LMM tree.
+As the outcome variable is continuous, we employed an identity link and Gaussian response distribution.
+The resulting GLMM tree will therefore be refered to as an LMM tree.
 To compare the accuracy of the LMM tree, we also fitted an LM tree and an LMM with pre-specified interactions to the data.
@@
 The results of recursive partitioning techniques are known to be potentially unstable, in the sense that small changes in the dataset may substantially alter the variables or values selected for partitioning.
-Therefore, we used the R~package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values.
-Using the \verb|stabletree| function, we calculated the number of times a variable and value were selected for partitioning over 500 subsamples of size $.9 \times N$ of the dataset.
+Therefore, following \citeA{PhilyZeil16}, subsampling is used to assess the stability of the selected splitting variables and values.
+More precisely, variable selection frequencies of the trees are computed from 500 subsamples, each comprising 90\% of the full dataset.
 
@@
 
-The trees and effect sizes resulting from application of LMM and LM tree are presented in Figure~\ref{fig:lmertree_C&W} and Figure~\ref{fig:lmtree_C&W}, respectively.
+The trees and effect sizes resulting from application of LMM and LM trees are presented in Figure~\ref{fig:lmertree_C&W} and Figure~\ref{fig:lmtree_C&W}, respectively.
 The LM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
@@
 
-The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: b = 2.29, p = .002; interaction with treatment: b = -2.10, p = .028).
-Also, the GLMM indicated an interaction between treatment and age (b = .10, p = .018).
+The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: $b = 2.29$, $p = .002$; interaction with treatment: $b = -2.10$, $p = .028$).
+Also, the GLMM indicated an interaction between treatment and age ($b = .10$, $p = .018$).
 
@@
 The correlation between true and predicted post-treatment HAM-D total scores averaged over 50 folds was .272 ($\mathit{var}=.067$) for LMM tree, .233 ($\mathit{var}=.064$) for the LMM with pre-specified interactions and .190 ($\mathit{var}=.084$) for the LM tree.
+\readme{SD rather than var is used in the manuscript up to this point.}
 
@@
 As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.
+\readme{Practitioners might still worry that the CAD variable is not selected more often -- only in a bit more than 50\% of the cases.
+Maybe this would be better in bootstrap sampling rather than resampling because the sample size is not reduzed?}
 

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
discussion probably needs some more work...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1299 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+\readme{This summary is quite extensive. Personally, I would prefer a more condensed summary of `highlights'.}
+
 We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters.
-We hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies.
-
-GLMM tree accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions.
-In contrast, GLM tree accurately recovered treatment-subgroup interactions in only 61\% of these datasets.
-In terms of predictive accuracy, GLMM tree outperformed GLM tree as well as MERT, with an average correlation between true and predicted treatment-effect differences of .94 for GLMM tree, .88 for GLM tree and .75 for MERT.
-The Type-I error rate of GLMM tree very closely approximated the $\alpha$ level used for testing parameter stability: GLMM tree erroneously detected subgroups in 4\% of datasets without treatment-subgroup interactions, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.
-
-The better performance of GLMM tree was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables.
-In those datasets, GLM tree is likely to detect spurious splits and subgroups.
-At the same time, GLM tree showed less power to detect the true subgroups in the presence of random effects.
-As expecte, the accuracy of MERT was not affected much by the presence of random effects, but only approached the accuracy of GLMM tree in datasets with the largest sample and effect sizes.
-GLMM tree especially outperformed the other methods when effect size was small (i.e., Cohen's $d=.5$) and sample size was small (i.e., 200).
-Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM tree may therefore provide a helpful tool for subgroup detection in those instances.
-
-When random effects were absent from the simulated datasets, GLM and GLMM tree yielded very similar predictive accuracy.
-This finding is of practical importance, as it indicates that application of GLMM tree will not `hurt': GLM tree and GLMM tree are expected to perform equally well in the absence of random effects, while GLMM tree will likely outperform GLM tree in the presence of random effects.
-
-Compared to linear mixed-effects models with pre-specified interactions, GLMM tree provided somewhat better accuracy, on average.
-As expected, GLMM tree performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise.
-We found a clear advantage of GLMM tree in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited than GLMMs for exploratory analyses.
+We hypothesized GLMM trees to be well suited for the detection of treatment-subgroup interactions in clustered datasets and confirmed this by our simulation studies.
+
+GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions.
+In contrast, GLM trees accurately recovered treatment-subgroup interactions in only 61\% of these datasets.
+In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERT, with an average correlation between true and predicted treatment-effect differences of .94 for GLMM trees, .88 for GLM trees and .75 for MERT.
+The Type-I error rate of GLMM trees very closely approximated the $\alpha$ level used for testing parameter stability: GLMM trees erroneously detect subgroups in 4\% of datasets without treatment-subgroup interactions, whereas GLM trees erroneously detect subgroups in 33\% of those datasets.
+
+The better performance of GLMM trees was mostly observed when random effects in the datasets were sizeable, and random intercepts were correlated with potential partitioning variables.
+In those datasets, GLM trees are likely to detect spurious splits and subgroups.
+At the same time, GLM trees showed less power to detect the true subgroups in the presence of random effects.
+As expected, the accuracy of MERT was not affected much by the presence of random effects, but only approached the accuracy of GLMM trees in datasets with the largest sample and effect sizes.
+GLMM trees especially outperformed the other methods when effect size was small (i.e., Cohen's $d=.5$) and sample size was small (i.e., 200).
+Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM trees may therefore provide a helpful tool for subgroup detection in those instances.
+
+When random effects were absent from the simulated datasets, GLM and GLMM trees yielded very similar predictive accuracy.
+This finding is of practical importance, as it indicates that application of GLMM trees will not `hurt': GLM trees and GLMM trees are expected to perform equally well in the absence of random effects, while GLMM trees will likely outperform GLM trees in the presence of random effects.
+
+Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy, on average.
+As expected, GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise.
+We found a clear advantage of GLMM trees in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited than GLMMs for exploratory analyses.
 Also, GLMM trees may be easier to interpret: The number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models.
@@
 
-In the Application, we obtained similar findings: GLMM tree yielded higher predictive accuracy, while using a smaller number of variables for prediction than GLM tree and a GLMM with pre-specified interactions.
+In the Application, we obtained similar findings: the GLMM tree yielded higher predictive accuracy, while using a smaller number of variables for prediction than the GLM tree and a GLMM with pre-specified interactions.
 In addition, the GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.
@@
 
+\readme{I would emphasize early on that GLMM trees rely on the properties of their building blocks, namely GLMMs and MOBs.
+The effect of misspecifications are studied to some degree for both methods.
+For GLMM one could cite papers discussing this.
+For MOB, or trees more broadly, there are also some references.
+One of these could be the `To Split or to Mix?' paper by Frick, Strobl, and Zeileis.}
+
 Recursive partitioning methods were originally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}.
-However, GLMM tree is obviously a parametric tool, as it fits fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors.
+However, GLMM trees are obviously a parametric tool, as they fit fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors.
 Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM tree.
@@
 If relevant variables are omitted, or incorrectly specified, GLMM tree can only approximate the true subgroups using the specified variables.
-Our simulations indicate that LM tree detected spurious subgroups as a result of misspecifying (that is, not including) the random effects.
+Our simulations indicate that LM trees detect spurious subgroups as a result of misspecifying (that is, not including) the random effects.
 Reduced accuracy and spurious splits can also be expected to occur when relevant random effects are not included when specifying the GLMM tree model.

### 2017-05-12 pkg/glmertree/paper/treatment_interactions.Rnw
Some paper revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1327 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usepackage{lscape,dcolumn,hhline}
-\usepackage{graphicx,epic,eepic,rotating}
 \usepackage[normalem]{ulem}
@@
 \usepackage{Sweave}
-\usepackage{subfig}
+\usepackage[caption=false]{subfig}
 
@@
 \rightheader{Treatment-Subgroup Interactions with GLMM Trees}
-\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
-\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}
+\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^1$}
+\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich}
 
@@
 In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees.
-Also, GLMM tree show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effect, on average.
+Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effect, on average.
 We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression.
-We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.
-An implementation is available in the R package \verb|glmertree| and a short tutorial demonstrating how to fit GLMM trees in practice is provided in the accompanying vignette.\\
+We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
 \\
@@
 Tree-based methods are such hypothesis-generating methods.
-Tree-based methods, also known as recursive partitioning methods, recursively split the space spanned by the predictor variables into rectangular regions, containing observations that are increasingly similar with respect to the outcome.
-Several tree-based methods take the mean or majority class of a single dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}.
+Tree-based methods, also known as recursive partitioning methods, split observations repeatedly into groups so that they become increasingly similar with respect to the outcome within each group.
+Several tree-based methods take the mean of a continuous dependent variable or the majority class of a categorical dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}.
 Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM approach of \citeA{Ciam91} is the earliest example.
@@
 Each of the splits in the tree is defined by a splitting variable and value.
-The first split in the tree separates the observations into two subgroups, based on the splitting variable duration and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3.
+The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3.
 As the observations in node~2 are not further split, node~2 is a terminal node.
@@
 If the partition in Figure~\ref{fig:exampletree} would be used for prediction of a new observation, the new observation would be assigned to one of the terminal nodes according to its values on the splitting variables.
-The prediction is then determined by the distribution of the training observations within the terminal node.
+The prediction is then based on the estimated distribution of the outcome variable within that terminal node.
 For example, the prediction may be the node-specific mean of a single continuous variable.
@@
 As such, they are preeminently suited to the detection of treatment-subgroup interactions.
-Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov16, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview).
+Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov16, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15, AtheyImbe16}).
 Also, \citeA{ZhanyTsia12Bio} and \citeA{ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.
@@
 For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}.
-In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}, GLMM tree allows for taking into account the clustered structure of datasets.
+In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}, GLMM trees allow for taking into account the clustered structure of datasets.
 In contrast to previously suggested regression trees with random effects \cite<e.g., >{HajjyBell11,SelaySimo12}, GLMM trees allow for treatment effect estimation, with continuous as well as non-continuous response variables.
@@
 The remainder of this paper is structured into four sections: In the first section, we introduce the GLMM tree algorithm using an artificial motivating dataset with treatment-subgroup interactions.
-The second section is a short tutorial on how to fit GLMM trees using the R package \verb|glmertree| \readme{still needs to be adapted: vignette}.
-In the third section, we compare the performance of GLMM trees with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions.
-In the fourth section, we apply a GLMM tree to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression.
-In the fith and last section we summarize the results and discuss limitations and directions for future research.
+In the second section, we compare the performance of GLMM trees with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions.
+In the third section, we apply the GLMM tree algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression.
+In the fourth and last section we summarize the results and discuss limitations and directions for future research.
+A tutorial on how to fit GLMM trees using the R package \verb|glmertree| is included as supplementary material.
+In the tutorial, the artificial motivating dataset is used, allowing users to recreate the trees and models to be fitted in the next section.
 
@@
 We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm.
-The R code used to generate the example dataset is provided in the Appendix \readme{I would put this into the appendix of the vignette instead.}.
-
-The dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2).
+This dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2).
 The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology.
@@
 Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment.
-Finally, each patient was part of one of ten clusters, each having a different value for the random intercept, uncorrelated with the partitioning variables.
-
-The outcome variable was generated such that there are three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: For the first subgroup of patients (node~3) with short duration ($\le 8$) months of depressive symptoms prior to treatment and low anxiety scores ($\le 10$), Treatment~1 leads to lower post-treatment depression than in Treatment~2.
-For the second subgroup of patients (node~4) with short duration but high anxiety scores ($> 10$) post-treatment depression is about equal in both treatment conditions.
-For the third subgroup of patients (node~5) with long duration ($> 8$ months) Treatment~2 leads to lower post-treatment depression than Treatment~1.
+Summary statistics of these variables are provided in Table~\ref{tab:sum_stats}.
+Each patient was part of one of ten clusters, each having a different value for the random intercept, which was generated from a standard normal distribution and uncorrelated with the partitioning variables.
+
+\begin{table}[h!]
+\small
+\caption{\small Summary statistics for partitioning and outcome variables in the artificial motivating dataset}
+\begin{tabular}{lcccc}
+\thickline
+ & min & max & mean & sd\\
+\hline
+depression & 3 & 16 & 9.12 & 2.66 \\
+age & 18 & 69 & 45 & 9.56 \\
+anxeity & 3 & 18 &  10.26 & 3.05 \\
+duration & 1 & 17 & 6.97 & 2.90 \\
+\hline
+\end{tabular}
+\label{tab:sum_stats}
+\end{table}
+
+The outcome variable was generated such that there are three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: For the first subgroup of patients (node~3) with short duration ($\le 8$) months of depressive symptoms prior to treatment and low anxiety scores ($\le 10$), Treatment~1 leads to lower post-treatment depression than in Treatment~2 (true mean difference = 2).
+For the second subgroup of patients (node~4) with short duration but high anxiety scores ($> 10$) post-treatment depression is about equal in both treatment conditions (true mean difference = 0).
+For the third subgroup of patients (node~5) with long duration ($> 8$ months) Treatment~2 leads to lower post-treatment depression than Treatment~1 (true mean difference = 2.5).
 Thus, duration and anxiety are true partitioning or moderator variables, whereas age is not.
 Anticipating the final results of our analyses, the treatment-subgroup interactions are depicted in Figure~\ref{fig:example_glimmertree}, which shows the GLMM tree that accurately recovered the treatment-subgroup interactions.
-\readme{I've incorporated a characterization of the groups in terms of duration and anxiety here.
-But probably a better description of these variables is needed earlier.
-Possibly in a table with some summary statistics, e.g., mean/sd/range.
-Finally, the description in the manual Rd pages of the package should also be improved.}
 
@@
 Subsequently, the data may be partitioned recursively with respect to other covariates, leading to separate models with different treatment effects/coefficients in each subsample.
-\readme{Expanded on the treatment GLM MOB explanation to give more context and motivation.}
 
@@
 The MOB algorithm can be used to partition the dataset using additional covariates and find better-fitting local models.
-To this end, the MOB algorithm tests for parameter stability over a set of auxiliary covariates, also called \textit{partitioning variables}.
+To this end, the MOB algorithm tests for parameter stability with respect to each of a set of auxiliary covariates, also called \textit{partitioning variables}, which we will denote by $U$.
 When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor.
-To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) statistically test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.
+To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) statistically test for parameter instability with respect to each of a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.
 
@@
 If at least one of the partitioning variables yields a $p$-value below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3).
-This partition is created using $U_{{k^*}}$ \readme{the U variables have not been introduced, no symbol has been used for the partitioning variables so far}, the partitioning variable with the minimal $p$-value in step (2).
-The split point for $U_{k^*}$ is selected by taking the value that minimizes the sum of the values of the objective function in both partitions.
-In other words, for every possible split point, the objective function is minimized separately in the two resulting subsets of the data; the split point yielding the minimum sum of the objective functions is selected.
+This partition is created using $U_{k^*}$, the partitioning variable with the minimal $p$-value in step (2).
+The split point for $U_{k^*}$ is selected by taking the value that minimizes the instability as measured by the sum of the values of two loss functions, one for each of the resulting subgroups.
+In other words, the loss function is minimized separately in the two subgroups resulting from every possible split point and the split point yielding the minimum sum of the loss functions is selected.
 In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).
@@
 For simplicity, we employ a cluster-specific intercept only, but further random effects can easily be included in $z_i$.
-Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
-Also, the errors $\epsilon$ are assumed to follow a normal distribution, with constant variance across clusters.
+Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$ and that the errors $\epsilon$ have constant variance across clusters.
 The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).
@@
 %
+In the GLMM tree model, the fixed effects $\beta_j$ are local parameters, their value depending on subgroup $j$, but the random effects $b$ are global.
 To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}.
@@
 In every iteration, the GLM tree is re-estimated in step (1) and the fixed- and random-effects parameters are re-estimated in step (2).
-Note that the random-effects part of the model is not partitioned, but estimated globally.
+Note that the random effects are not partitioned, but estimated globally.
 Only the fixed effects are estimated locally, within the cells of the partition.
@@
 In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented.
-In addition to the treatment indicator, treatment outcome and the potential partitioning variables, the GLMM tree has also included a random intercept term with respect to the cluster indicator.
-As a result, the dependence between observations is taken into accoutn, the true treatment subgroups have been recovered and the spurious split involving the anxiety variable no longer appears in the tree.
+In addition to the treatment indicator, treatment outcome and the potential partitioning variables, the GLMM tree algorithm has also taken a random intercept with respect to the cluster indicator into account.
+As a result, the dependence between observations is taken into account, the true treatment subgroups have been recovered and the spurious split involving the anxiety variable no longer appears in the tree.
 
 \FloatBarrier
-
-\section{Tutorial}
-
-\readme{This tutorial should be turned into a vignette and then the R output, graphics, etc. should be shown as well.}
-
-We implemented the GLMM tree algorithm in the R package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
-The package makes use of the \verb|partykit| package \cite{HothyZeil15} and \verb|lme4| package \cite{BateyMeac12} to estimate trees and mixed-effects models, respectively.
-The latest version of the \verb|glmertree| package can be downloaded, installed and loaded as follows:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-install.packages("glmertree", repos="http://R-Forge.R-project.org")
-library(glmertree)
-@
-\end{small}
-
-Package documentation and examples can be accessed as follows:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-?glmertree
-@
-\end{small}
-
-The main functions in the \verb|glmertree| package are \verb|lmertree()|, for continuous outcome variables, and \verb|glmertree()|, for binary or count outcome variables.
-Both functions require the user to specify at least two arguments: \verb|formula| and \verb|data|.
-The data we will be using is the motivating example dataset described in the previous section.
-This dataset is included in the \verb|partykit| package and can be loaded as follows:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-data("a_m_data", package = "glmertree")
-summary(a_m_data)
-@
-\end{small}
-
-The model formula to be specified consists of a left- and right hand side.
-The left hand side of the model formula (preceding the tilde symbol) specifies the outcome variable.
-The right hand side consists of three parts, separated by vertical bars: The first part specifies the predictor variable(s) of the (generalized) linear model, the second part specifies the random effects and the third part specifies the potential partitioning variables:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-exampleglmmtree <- lmertree(depression ~ treatment | cluster | age + duration + 
-                            anxiety, data = a_m_data)
-@
-\end{small}
-
-Note that in the example above, the partitioning variables are continuous, but (ordered) categorical partitioning variables may also be specified.
-Also, we specified only a single variable in the random-effects part, resulting in estimation of a random intercept with respect to \verb|cluster|.
-More complex random effects can also be specified: for example, specifying the random-effects part as \verb=...
-| (1 + age | cluster) | ...= would yield a model with a random intercept as well as a random slope for age with respect to cluster.
-The brackets are necessary to protect the vertical bars in the random effects formulation.
-
-Now, we can use the \verb|plot| method to recreate the tree depicted in Figure~\ref{fig:example_glimmertree}:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-plot(exampleglmmtree)
-@
-\end{small}
-
-In every inner node of the tree, the splitting variable and corresponding p-value from the parameter stability test is reported.
-To control for multiple testing, the p-values are Bonferroni corrected, by default.
-This can be turned off by adding \verb|bonferroni = FALSE| to the function call, yielding a less conservative criterion for the parameter stability tests, but note that this will increase the likelihood of overfitting.
-The significance level $\alpha$ equals .05 by default, but a different value, say for example .01, can be specified by including \verb|alpha = .01| in the function call.
-
-The predictions of the random effects can be plotted by adding \verb|plotranef = TRUE| to the function call:
-\readme{As already indicated by e-mail, plotranef is an awkward name and should be improved, e.g., to which.}
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-plot(exampleglmmtree, plotranef = TRUE)
-@
-\end{small}
-
-To obtain numerical results, \verb|print|, \verb|coef| and \verb|ranef| methods can be used:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-print(exampleglmmtree)
-coef(exampleglmmtree)
-ranef(exampleglmmtree)
-@
-\end{small}
-
-To obtain predicted values, the \verb|predict| method can be used:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-predict(exampleglmmtree, newdata = a_m_data[1:10,])
-@
-\end{small}
-
-When \verb|newdata| is not specified, predictions for the training observations are returned, by default.
-Random effects can be excluded from the predictions by adding \verb|re.form = NA|.
-This is useful, for example, when \verb|newdata| is specified, but the new observations do not have a cluster indicator or are from new clusters:
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-predict(exampleglmmtree, newdata = a_m_data[1:10, -4], re.form = NA)
-@
-\end{small}
-
-Residuals of the fitted GLMM tree can be obtained with the \verb|residuals| method. This can be useful for assessing potential misspecification of the model (e.g., heteroscedasticity):
-
-\begin{small}
-<<eval = FALSE, echo = TRUE>>=
-resids <- residuals(exampleglmmtree)
-preds <- predict(exampleglmmtree)
-plot(a_m_data$cluster, resids)
-scatter.smooth(preds, resids)
-@
-\end{small}
-
-The first plot did not indicate substantial variation in error variances across levels of the random effects.
-The second plot of fitted values against residuals also did not reveal a pattern indicating model misspecification.
 
@@
 In Study~III, we assessed and compared the performance of GLMM trees and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions.
-As the outcome variable was continuous in all simulated dataset, the GLMM tree algorithm and the trees resulting from its application will be referred to as LMM tree(s).
+As the outcome variable was continuous in all simulated datasets, the GLMM tree algorithm and trees resulting from its application will be referred to as LMM tree(s).
 
@@
 
-\subsection{Study~I: Performance of LMM tree, LM tree and MERT in datasets with treatment-subgroup interactions}
+\subsection{Study~I: Performance of LMM trees, LM trees and MERTs in datasets with treatment-subgroup interactions}
 
@@
 For MERT, the number of nodes and tree accuracy was not assessed, as the treatment-subgroup interaction design in Figure~\ref{fig:modelC} corresponds to a large number of regression tree structures, that would all be different but also correct.
-Therefore, performance of MERT was only assessed in terms of predictive accuracy.
+Therefore, performance of MERTs was only assessed in terms of predictive accuracy.
 
@@
 treespecs.long2$treesize.ind <- factor(treespecs.long2$treesize.ind)
-treespecs.long2$truetree.ind <- factor(treespecs.long2$truetree.ind)
-treespecs.long2$truetree.values <- as.numeric(treespecs.long2$truetree.values)-1
 @
@@
 The effects of the four most important predictors of tree accuracy are depicted in Figure~\ref{fig:xyplot_treeaccuracy}.
-In the absence of random effects, LM and LMM tree were about equally accurate.
+In the absence of random effects, LM and LMM trees were about equally accurate.
 In the presence of random effects, LM trees were much less accurate than LMM trees when random effects were correlated with a partitioning covariate.
@@
 type = "b", ylab = "correlation", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), 
-auto.key = list(space = "top", columns = 3, title = " ", cex.title = 1, lines = TRUE, points = TRUE),
-scales = list(y = list(limits = c(-.1,1.1)))))
+auto.key = list(space = "top", columns = 3, title = " ", cex.title = 1, lines = TRUE, points = TRUE), ylim = c(0.3, 1.1)))
 @
@@
 
-The predicted treatment-effect differences of LMM tree show an average correlation of .93 (SD = .13) with the true differences.
-LM trees and MERT show lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively.
-The most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlations} \readme{Is it necessary to span the whole unit interval on the y-axis?
-Using 0.3 to 1.1 should be more than sufficient, I guess}.
+The predicted treatment-effect differences of LMM trees show an average correlation of .93 (SD = .13) with the true differences.
+LM trees and MERTs show lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively.
+The most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlations}.
 Performance of all three algorithms improves with increasing sample size and treatment-effect differences.
-Furthermore, LMM trees and MERT are not much affected by the presence and magnitude of random effects in the data.
+Furthermore, LMM trees and MERTs are not much affected by the presence and magnitude of random effects in the data.
 LMM trees perform most accurately in most conditions and are never outperformed by the other methods.
-MERT performs least accurately in most conditions and never outperforms the other methods.
-The differences in accuracy become less pronounced with larger sample and effect sizes.
+MERTs perform the least accurate in most conditions and never outperforms the other methods, but the differences in accuracy become less pronounced with larger sample and effect sizes.
 
@@
 
-The treatment-subgroup interactions in Study~I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables.
-Trees are pre-eminently suited for recovering such piecewise interactions, but may have difficulty when the true interactions are continuous functions of moderator variables (for example, $U_1 \cdot U_2$).
+The interactions in Study~I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables.
+Trees are preeminently suited for recovering such piecewise or subgroup interactions, but may have difficulty when the true interactions are continuous functions of moderator variables (for example, $U_1 \cdot U_2$).
 At the same time, linear regression models with pre-specified interaction terms may perform well in recovering continuous interactions, but may have difficulty in recovering piecewise interactions.
@@
 
-For datasets with purely piecewise interactions, the same partition as in Study~I (Figure~\ref{fig:modelC}) was used.
+To generate datasets with purely piecewise interactions, the same partition as in Study~I (Figure~\ref{fig:modelC}) was used.
 In other words, the outcome variable in this design was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
@@
 For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure~\ref{fig:modelC} was also used.
-However, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes now comprised continuous main and (treatment) interaction effects of the partitioning variables.
-The corresponding node-specific parameters are presented in Table~\ref{tab:continuous_terms}.
+In addition, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes now comprised continuous main and (treatment) interaction effects of the partitioning variables.
+In other words, the partitioning variables $U_2$, $U_1$ and $U_5$ appear are also part of the linear predictor $x_i$, as part of the terms presented in Table~\ref{tab:continuous_terms}.
+The corresponding node-specific $\beta_j$ parameters are also presented in Table~\ref{tab:continuous_terms}.
 The $\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure~\ref{fig:modelC}.
 The interaction terms were created using centered $U_k$ variables, calculated by subtracting their variable means.
+Again, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
 
 In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by the single $\beta$ column in Table~\ref{tab:continuous_terms}.
+The outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$.
 
@@
 Predictive accuracy was assessed in terms of the correlation between the true and predicted treatment-effect differences in test datasets.
-As the full LMM models were likely to overfit, LMMs were refitted on the training data, using only the predictors with $p$-values $<.05$ in the original LMM.
+As full LMMs may be likely to overfit, LMMs were refitted on the training data, using only the predictors with $p$-values $<.05$ in the original LMM.
 Predictions for test observations were obtained using the refitted LMMs.
@@
 As Figure~\ref{fig:xyplot_correlation_pw} indicates, LMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas LMMs show highest predictive accuracy in datasets with purely continuous interactions.
-LMMs perform poorly when interactions are not purely piecewise, and LMM trees perform poorly only when interactions are purely linear.
-
-Performance of both methods improves with increasing sample size.
+LMM trees perform poorly only when interactions are purely linear, whereas LMMs perform poorly when interactions are not purely continuous.
+Strikingly, Figure~\ref{fig:xyplot_correlations_pw} suggests that LMMs perform somewhat more accurately in the presence of purely piecewise interactions than in the presence of partly continuous interactions, but only with larger sample sizes and a smaller number of potential moderator variables.
+
+Performance of both LMM trees and LMMS improves with increasing sample size.
 Furthermore, performance of LMM trees is not affected by the number of covariates, whereas the predictive accuracy of LMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous.
@@
 
-In terms of interpretability, LMM trees will often provide simpler models: The LMMs included 12.30 significant terms on average, whereas LMM trees had 3.38 inner nodes on average, requiring only about 3-4 variables to be evaluated for making predictions.
+Finally, it should be noted that in terms of interpretability, LMM trees often provide simpler models: The LMMs included 12.30 significant terms on average, whereas LMM trees had 3.38 inner nodes on average, requiring only about 3-4 variables to be evaluated for making predictions.
 
@@
 \paragraph{Dataset}
-To illustrate the use of GLMM trees in real application, we employ a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}.
+To illustrate the use of GLMM trees in real data applications, we employ a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}.
 This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
@@
 
-As the outcome variable is continuous, we employed an identity link and Gaussian response distribution.
-The resulting GLMM tree will therefore be refered to as an LMM tree.
-To compare the accuracy of the LMM tree, we also fitted an LM tree and an LMM with pre-specified interactions to the data.
-In the LMM, the outcome variable was regressed on a random intercept, main effects of treatment and the potential moderators (partitioning variables) and interactions between treatment and the potential moderators.
+As the outcome variable is continuousTo this end, the MOB algorithm tests for parameter stability with respect to
+each of a set of auxiliary covariates, we employed an identity link and Gaussian response distribution.
+The resulting GLMM trees will therefore be refered to as LMM trees.
+To compare the accuracy of LMM trees, we also fitted LM trees and LMMs with pre-specified interactions to the data.
+In the LMMs, the outcome variable was regressed on a random intercept, main effects of treatment and the potential moderators (partitioning variables) and interactions between treatment and the potential moderators.
 As it is not known in advance how to interact the potential moderators, higher-order interactions were not included.
@@
 
-\paragraph{Predictive accuracy and stability}
+\paragraph{Predictive accuracy}
 
 Predictive accuracy of each method was assessed by calculating average correlation between observed and predicted HAM-D post-treatment scores, based on 50-fold cross validation.
+
+\paragraph{Stability}
 
@@
 
-The trees and effect sizes resulting from application of LMM and LM trees are presented in Figure~\ref{fig:lmertree_C&W} and Figure~\ref{fig:lmtree_C&W}, respectively.
+\begin{figure}[tb!]
+\centering
+\setkeys{Gin}{width=0.75\textwidth}
+<<depression-glmtree, fig=TRUE, height=9, width=8>>=
+plot(lmtree_app, tnex = 3.5, terminal_panel = combine_panels(
+  panel1 = node_bivplot(lmtree_app),
+  panel2 = node_terminal(GLMtree_app_eff, FUN = identity, height = 10, width = 14, id = FALSE),
+  party2 = GLMtree_app_eff
+))
+@
+\caption{\small LM tree for prediction of treatment outcomes.
+Upper terminal nodes: y-axes represent post-treatment HAM-D scores, x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
+Lower terminal nodes: Subgroup-specific descriptive statistics.}
+\label{fig:lmtree_C&W}
+\end{figure}
+
+Figure~\ref{fig:lmtree_C&W} and Figure~\ref{fig:lmertree_C&W} present the LM and LMM trees fitted on the IPDMA data.
 The LM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
@@
 
-\begin{figure}[tb!]
-\centering
-\setkeys{Gin}{width=0.75\textwidth}
-<<depression-glmtree, fig=TRUE, height=9, width=8>>=
-plot(lmtree_app, tnex = 3.5, terminal_panel = combine_panels(
-  panel1 = node_bivplot(lmtree_app),
-  panel2 = node_terminal(GLMtree_app_eff, FUN = identity, height = 10, width = 14, id = FALSE),
-  party2 = GLMtree_app_eff
-))
-@
-\caption{\small LM tree for prediction of treatment outcomes.
-Upper terminal nodes: y-axes represent post-treatment HAM-D scores, x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
-Lower terminal nodes: Subgroup-specific descriptive statistics.}
-\label{fig:lmtree_C&W}
-\end{figure}
-
 The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: $b = 2.29$, $p = .002$; interaction with treatment: $b = -2.10$, $p = .028$).
@@
 Assessment of predictive accuracy by means of 50-fold cross validation indicated better predictive accuracy for the LMM tree than for the LM tree and the LMM.
-The correlation between true and predicted post-treatment HAM-D total scores averaged over 50 folds was .272 ($\mathit{var}=.067$) for LMM tree, .233 ($\mathit{var}=.064$) for the LMM with pre-specified interactions and .190 ($\mathit{var}=.084$) for the LM tree.
-\readme{SD rather than var is used in the manuscript up to this point.}
+The correlation between true and predicted post-treatment HAM-D total scores averaged over 50 folds was .272 ($\mathit{SD}=.260$) for LMM tree, .233 ($\mathit{SD}=.252$) for the LMM with pre-specified interactions and .190 ($\mathit{SD}=.290$) for the LM tree.
 
@@
 As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.
-\readme{Practitioners might still worry that the CAD variable is not selected more often -- only in a bit more than 50\% of the cases.
-Maybe this would be better in bootstrap sampling rather than resampling because the sample size is not reduzed?}
 
@@
 
-\readme{This summary is quite extensive. Personally, I would prefer a more condensed summary of `highlights'.}
-
 We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters.
@@
 
-GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions.
-In contrast, GLM trees accurately recovered treatment-subgroup interactions in only 61\% of these datasets.
-In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERT, with an average correlation between true and predicted treatment-effect differences of .94 for GLMM trees, .88 for GLM trees and .75 for MERT.
-The Type-I error rate of GLMM trees very closely approximated the $\alpha$ level used for testing parameter stability: GLMM trees erroneously detect subgroups in 4\% of datasets without treatment-subgroup interactions, whereas GLM trees erroneously detect subgroups in 33\% of those datasets.
-
-The better performance of GLMM trees was mostly observed when random effects in the datasets were sizeable, and random intercepts were correlated with potential partitioning variables.
-In those datasets, GLM trees are likely to detect spurious splits and subgroups.
-At the same time, GLM trees showed less power to detect the true subgroups in the presence of random effects.
-As expected, the accuracy of MERT was not affected much by the presence of random effects, but only approached the accuracy of GLMM trees in datasets with the largest sample and effect sizes.
-GLMM trees especially outperformed the other methods when effect size was small (i.e., Cohen's $d=.5$) and sample size was small (i.e., 200).
-Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM trees may therefore provide a helpful tool for subgroup detection in those instances.
-
-When random effects were absent from the simulated datasets, GLM and GLMM trees yielded very similar predictive accuracy.
-This finding is of practical importance, as it indicates that application of GLMM trees will not `hurt': GLM trees and GLMM trees are expected to perform equally well in the absence of random effects, while GLMM trees will likely outperform GLM trees in the presence of random effects.
+GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions, outperforming GLM trees without random effects.
+In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average.
+In datasets without treatment-subgroup effects, GLMM trees showed a Type-I error rate 4\$, compared with a Type-I error rate of 33\% for GLM trees.
+
+Better performance of GLMM trees was mostly observed when random effects were sizeable and correlated with potential partitioning variables.
+In those circumstances, GLM trees are likely to detect spurious splits and subgroups.
+At the same time, GLM trees showed less power to detect subgroups in the presence of random effects.
+As expected, accuracy of MERTs was not affected by the presence of random effects, but only approached GLMM trees' accuracy in datasets with the largest sample and effect sizes.
+Especially with smaller effect and sample sizes (i.e., Cohen's $d=.5$ and/or $N = 200$), GLMM tree outperformed the other tree methods.
+Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM trees may provide a useful tool for subgroup detection in those instances.
+
+In the absence of random effects, GLM and GLMM trees yielded very similar predictive accuracy.
+This finding is of practical importance, indicating that application of GLMM trees will not `hurt' predictive accuray when in fact, random effects are absent.
 
 Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy, on average.
-As expected, GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise.
-We found a clear advantage of GLMM trees in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited than GLMMs for exploratory analyses.
-Also, GLMM trees may be easier to interpret: The number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models.
-The trees in our simulations were limited to a maximum depth of three, requiring evaluation of at most 3 variables to make a prediction or decision in practice.
-
-In the Application, we obtained similar findings: the GLMM tree yielded higher predictive accuracy, while using a smaller number of variables for prediction than the GLM tree and a GLMM with pre-specified interactions.
-In addition, the GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.
-
-\subsubsection{Limitations and future research}
+GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise.
+GLMM trees performed better in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited for exploration that GLMMs.
+Also, GLMM trees may be easier to interpret and apply: The number of terms in a GLMM increases quadratically with the number of predictors, yielding complex models.
+GLMM trees in our simulations had a maximum tree depth of three, requiring evaluation of at most three variables to make a prediction for new observations.
+
+In the Application, the GLMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the other two methods (GLM tree and a GLMM with pre-specified interactions).
+The GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.
+
+\subsubsection{Limitations}
 
@@
 
-Recursive partitioning methods were originally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}.
-However, GLMM trees are obviously a parametric tool, as they fit fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors.
-Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM tree.
-
-Furthermore, misspecification of partitioning and fixed- and random-effects variables will also reduce accuracy of the resulting GLMM tree.
-If relevant variables are omitted, or incorrectly specified, GLMM tree can only approximate the true subgroups using the specified variables.
+Recursive partitioning methods were origfinally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}.
+GLMM trees take MOB and GLMMs as building blocks, thereby relying on the properties of these building blocks.
+As a result, GLMM trees may inherit their sensitivities to model misspecification as well as their robustness against it.
+We will address potential types of misspecification in fitting GLMM trees:
+
+\readme{MF: GLMM tree at least tackles one type of random-effects misspecification: ignoring them.
+\cite{BarryLevy13} and \cite{Barr13} advice to keep it maximal is at repeated measures designs.
+\cite{McCuyNeuh11}: "A key conclusion is the large degree of robustness of maximum likelihood for a wide variety of commonly encountered situation".
+....
+fit fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors.
+Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM trees.}
+
+Furthermore, misspecification of partitioning and fixed- and random-effects variables will also reduce accuracy of the resulting GLMM trees.
+If relevant variables are omitted or incorrectly specified, GLMM trees can only approximate the true subgroups using the specified variables.
 Our simulations indicate that LM trees detect spurious subgroups as a result of misspecifying (that is, not including) the random effects.
 Reduced accuracy and spurious splits can also be expected to occur when relevant random effects are not included when specifying the GLMM tree model.
-Furthermore, as the random effects are estimated globally, misspecification of the random effects can have a strong impact on the resulting GLMM tree model.
+Furthermore, as the random effects are estimated globally, misspecification of the random effects may have a strong impact on the resulting GLMM tree model.
 
 Another source of misspecification is the inclusion of irrelevant variables.
-Although our simulations indicate that the performance of GLMM tree was not negatively affected by increasing the number of noise variables specified for partitioning from 2 to 12, the power to detect subgroups may be reduced with much larger numbers of noise variables.
-Including irrelevant variables in the random or fixed effects may also negatively affect accuracy of GLMM tree, but we have not assessed this in our study.
-
-Users can reduce the risk of misspecification when fitting a GLMM tree in two ways.
-Firstly, by carefully specifying the predictors of the GLM, the partitioning variables and the random effects.
-Secondly, by inspecting residuals and assessing stability of the resulting model.
-In the Tutorial we have shown how residuals can be plotted to assess potential misspecification.
-In the Application we have shown how the \verb|stablelearner| package can be used to assess tree stability.
-However, more research on the effects that various types of misspecification will have on the performance of GLMM tree is required.
-
-As GLMM tree fits more complex models than non-parametric tree-based methods, like CART for example, larger sample sizes are needed to fit the model.
+Although our simulations indicate that the performance of GLMM trees was not negatively affected by increasing the number of noise variables specified for partitioning from 2 to 12, the power to detect subgroups may be reduced with much larger numbers of noise variables.
+Including irrelevant variables in the random or fixed effects may also negatively affect accuracy of GLMM trees, but we have not assessed this in our study.
+
+Users can evaluate potential misspecification by inspecting residuals and predictions, and assessing tree stability.
+The vigette included as Supplementary Material shows how residuals andrandom-effects predictions can be plotted to assess potential misspecification.
+In the Application we have shown how tree stability can be assessed using the methods suggested by \cite{PhilyZeil16}.
+Future studies may focus on the effects of possible misspecification on the performance of GLMM trees.
+
+As GLMM trees fit more complex models than non-parametric tree-based methods, like CART for example, larger sample sizes are needed to fit the model.
 How much larger likely depends on the complexity of the specified model: Our simulations show that with fixed and random-effects specifications with only a single predictor, a sample size of 200 is sufficient to detect subgroups with moderate differences in treatment effect.
-More complex fixed- and random-effects specifications will require larger sample sizes; how these affect the performance of GLMM tree requires further research.
+More complex fixed- and random-effects specifications will require larger sample sizes; how these affect the performance of GLMM trees requires further research.
+
+\subsubsection{Future directions}
+
+\readme{MF: Add something about implementing non-normal non-parametric or mixture ranef}
 
@@
 performed worst under some conditions of the simulation study in terms of the Type I error rate.
-Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations and comparisons only focused on GLMM tree and GLM-based MOB.
+Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations and comparisons only focused on GLMM trees and GLM trees.
 
@@
 
-Our results indicate that GLMM tree provides accurate recovery of treatment-subgroup interactions and prediction of treatment effects, both in the presence and absence of random effects and interactions.
-Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, for example in multi-center trials or individual-level patient data meta-analyses.
+Our results indicate that GLMM trees provide accurate recovery of treatment-subgroup interactions and prediction of treatment effects, both in the presence and absence of random effects and interactions.
+Therefore, GLMM trees offer a promising method for detecting treatment-subgroup interactions in clustered datasets, for example in multi-center trials or individual-level patient data meta-analyses.
 
@@
 
-\begin{appendix}
-
-\section{R code for generating artificial motivating dataset}
-
-\begin{small}
-<<appendixA_1, echo = TRUE>>=
-set.seed(123)
-treatment <- rbinom(n = 150, size = 1, prob = .5)
-duration <- round(rnorm(150, mean = 7, sd = 3))
-anxiety <- round(rnorm(150, mean = 10, sd = 3))
-age <- round(rnorm(150, mean = 45, sd = 10))
-error <- rnorm(150, 0, 2)
-cluster <- error + rnorm(150, 0, 6)
-rand_int <- sort(rep(rnorm(10, 0, 1), each = 15))
-rand_int[order(cluster)] <- rand_int 
-error <- error - rand_int
-cluster[order(cluster)] <- rep(1:10, each = 15)
-node3t1 <- ifelse(duration <= 8 & anxiety <= 10 & treatment == 0, -2, 0)
-node3t2 <- ifelse(duration <= 8 & anxiety <= 10 & treatment == 1, 2, 0)
-node5t1 <- ifelse(duration > 8 & treatment == 0, 2.5, 0)
-node5t2 <- ifelse(duration > 8 & treatment == 1, -2.5, 0)
-depression <- round(9 + node3t1 + node3t2 + node5t1 + node5t2 + .4*treatment + 
-                   error + rand_int)
-treatment <- factor(treatment, labels = c("Treatment 1", "Treatment 2"))
-a_m_data <- data.frame(depression, treatment, cluster, age, anxiety, duration)
-@
-\end{small}
-
-\end{appendix}
-
 \end{document}

### 2017-05-12 pkg/glmertree/paper/treatment_interactions.Rnw
Minor paper revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1328 255f4249-2263-4417-b1ea-8498779e27a7

@@
 LMM trees perform poorly only when interactions are purely linear, whereas LMMs perform poorly when interactions are not purely continuous.
-Strikingly, Figure~\ref{fig:xyplot_correlations_pw} suggests that LMMs perform somewhat more accurately in the presence of purely piecewise interactions than in the presence of partly continuous interactions, but only with larger sample sizes and a smaller number of potential moderator variables.
+Strikingly, Figure~\ref{fig:xyplot_correlation_pw} suggests that LMMs perform somewhat more accurately in the presence of purely piecewise interactions than in the presence of partly continuous interactions, but only with larger sample sizes and a smaller number of potential moderator variables.
 
@@
 This indicates that LMM trees are especially useful for exploratory purposes, where there are many potential moderator variables.
-
-Finally, it should be noted that in terms of interpretability, LMM trees often provide simpler models: The LMMs included 12.30 significant terms on average, whereas LMM trees had 3.38 inner nodes on average, requiring only about 3-4 variables to be evaluated for making predictions.
+In addition, LMM trees may often provide simpler models: Whereas the LMMs included 12.30 significant terms on average, LMM trees had 3.38 inner nodes on average, requiring only about 3-4 variables to be evaluated for making predictions.
 
@@
 In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average.
-In datasets without treatment-subgroup effects, GLMM trees showed a Type-I error rate 4\$, compared with a Type-I error rate of 33\% for GLM trees.
+In datasets without treatment-subgroup effects, GLMM trees showed a Type-I error rate of 4\%, compared with a Type-I error rate of 33\% for GLM trees.
 
@@
 
-Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy, on average.
+Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy on average.
 GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise.
-GLMM trees performed better in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited for exploration that GLMMs.
-Also, GLMM trees may be easier to interpret and apply: The number of terms in a GLMM increases quadratically with the number of predictors, yielding complex models.
-GLMM trees in our simulations had a maximum tree depth of three, requiring evaluation of at most three variables to make a prediction for new observations.
-
-In the Application, the GLMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the other two methods (GLM tree and a GLMM with pre-specified interactions).
+GLMM trees performed better in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited for exploration than GLMMs.
+
+In the Application, we obtained similar results: the GLMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the other two methods (GLM tree and a GLMM with pre-specified interactions).
 The GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.

### 2017-05-18 pkg/partykit/inst/urp/urp.Rnw
mv vignette skeleton for the time being
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1345 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nojss]{jss}

%\VignetteIndexEntry{urp: Internals of Unbiased Recursive Partitioning}
%\VignetteKeywords{conditional inference, model-based recursive partitioning}
%\VignettePackage{partykit}

%% packages
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{rotating}
%% need no \usepackage{Sweave}

%% commands
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

\hyphenation{Qua-dra-tic}

\title{URP: Internals of Unbiased Recursive Partitioning}
\Plaintitle{URP: urp: Internals of Unbiased Recursive Partitioning}

\author{Torsten Hothorn\\Universit\"at Z\"urich
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

\Abstract{
tbd
}
\Keywords{conditional inference, model-based recursive partitioning}

\Address{
  Torsten Hothorn\\
  Institut f\"ur Sozial- und Pr\"aventivmedizin, Abteilung Biostatistik \\
  Universit\"at Z\"urich \\
  Hirschengraben 84\\
  CH-8001 Z\"urich, Switzerland \\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://user.math.uzh.ch/hothorn/}\\

  Achim Zeileis \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Achim.Zeileis@R-project.org} \\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

\begin{document}

<<setup, echo = FALSE, results = hide>>=
options(width = 70, SweaveHooks = list(leftpar = 
    function() par(mai = par("mai") * c(1, 1.1, 1, 1))))
require("partykit")
set.seed(290875)
@

\setkeys{Gin}{width=\textwidth}

\section{Overview}

\section{Exported: URP tree}

Main workhorse

<<.urptree, eval = FALSE>>=
.urp_tree <- function
(
    call,               ### match.call of user-visible function
    frame,              ### parent.frame of user-visible function
    data = NULL, 
    data_asis = FALSE, 
    control,            ### .urp_control() or more
    trafofun,           ### function for transformations
    doFit = TRUE        ### grow tree or set-up only
) {}
@

\begin{description}
\item[\code{call}] a call with \code{formula}, 
                   \code{data}, \code{subset}, 
                   \code{weights}, \code{offset}, 
                   \code{cluster}, \code{strata} arguments.
\code{formula} is of the form \code{y ~ x | z} where \code{y} is the
response, \code{x} are covariates in the model to be partitioned and
\code{z} are partitioning variables. \code{y ~ z} is interpreted as
\code{y ~ 1 | z}.
\item[\code{frame}] a frame to evaluate the call in.
\item[\code{data}] optional \code{data.frame}.
\item[\code{data\_asis}] a logical indicating to use \code{data} and
  not reevaluating the call.
\item[\code{control}] control arguments
\item[\code{trafofun}] a function for the computation of scores / data
transformations, see below.
\item[\code{doFit}] a logical indicating to actually fit the tree or to
evaluate the call only.
\end{description}

The control arguments allow to change the behaviour of the tree growing
algorithm, for example to use permutation or $M$-fluctuation tests for
variable selection. The most important choice is are the scores to be used,
these are defined by \code{trafofun}.

\subsection{Argument trafofun}

This is a function with arguments

<<trafofun, eval = FALSE>>=
trafofun(formula, data, ctrl)
@

Here, \code{formula} is the model part \code{y ~ x} of the full formula
\code{y ~ x | z}. This function returns a function with arguments
<<trafo, eval = FALSE>>=
function(subset, estfun, object, info = NULL) {}
@

\begin{description}
\item[\code{subset}] an integer vector specifying the subset of
  observations in the current node (\proglang{R} indexing).
\item[\code{estfun}] a logical indicating to return the scores.
\item[\code{object}] a logical indicating to return a fitted model.
\item[\code{info}] additional information, typically a list with a
\code{coefficients} slot to be used as starting values for model fitting.
\end{description}

This function returns a list with slots
\begin{description}
\item[\code{estfun}] the matrix of score functions
\item[\code{coefficients}] optional model coefficients
\item[\code{objfun}] optional objective function to be maximised
\item[\code{object}] optional model object
\item[\code{converged}] a logical indicating model convergance
\end{description}

\section{Not Exported: URP fit}

<<urp_fit, eval = FALSE>>=
.urp_fit <- function
(
    data, 
    partyvars, 
    cluster, 
    ctrl
) {}
@

returns a function

<<urp_fit-ret, eval = FALSE>>=
function(trafo, subset, weights) {}
@

with returns a function

<<selectfun, eval = FALSE>>=
function(y = NULL, trafo, subset = integer(0), 
         weights = integer(0), whichvar, info = NULL)  {}
@

which returns a list with arguments

\begin{description}
\item[\code{criteria}] a $2 \times J$ matrix with rows \code{statistic} and
                       \code{p.value} on the LOG scale. 
\item[\code{splitfun}] a function \code{function(whichvar, minbucket)}
returning a \code{partysplit} object or \code{NULL}.
\end{description}

\section{No Exported: URP node}

<<urp-node, eval = FALSE>>=
.urp_node <- function
(
    id = 1L,                    ### id of this node
    data,                       ### full data, readonly
    selectfun,                  ### variable selection
                                ### and split function
    svselectfun = selectfun,    ### same for surrogate splits
    partyvars,                  ### partytioning variables
                                ### a subset of 1:ncol(data)
    weights = integer(0L),      ### optional case weights
    subset,                     ### subset of 1:nrow(data)
                                ### for identifying obs for this node
    ctrl,                       ### .urp_control()
    info = NULL,
    cenv = NULL                 ### environment for depth and maxid
) {}
@

%% \bibliography{party}

\end{document}

### 2017-05-18 pkg/model4you/inst/JORS/manuscript_model4you.Rnw
Adding forst infos to manuscript
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1348 255f4249-2263-4417-b1ea-8498779e27a7

%% Journal of Open Research Software Latex template -- Created By Stephen Bonner and John Brennan, Durham Universtiy, UK.

\documentclass{jors}

%% Set the header information
\pagestyle{fancy}
\definecolor{mygray}{gray}{0.6}
\renewcommand\headrule{}
\rhead{\footnotesize 3}
\rhead{\textcolor{gray}{UP JORS software Latex paper template version 0.1}}

\begin{document}

{\bf Software paper for submission to the Journal of Open Research Software} \\

To complete this template, please replace the blue text with your own. The paper has three main sections: (1) Overview; (2) Availability; (3) Reuse potential. \\

Please submit the completed paper to: editor.jors@ubiquitypress.com

\rule{\textwidth}{1pt}

\section*{(1) Overview}

\vspace{0.5cm}

\section*{Title}
model4you: An R package for personalised treatment effect estimation

\section*{Paper Authors}
1. Seibold, Heidi; 2. Zeileis, Achim; 3. Hothorn, Torsten

\section*{Paper Author Roles and Affiliations}
1. PhD student; Biostatistics Department, Epidemiology, Biostatistics \& Prevention
Institute, University of Zurich

2. Professor; Department of Statistics, Faculty of Economics and Statistics, University of
Innsbruck

3. Professor; Biostatistics Department, Epidemiology, Biostatistics \& Prevention
Institute, University of Zurich

\section*{Abstract}

\textcolor{blue}{A short (ca. 100 word) summary of the software being described: what problem the software addresses, how it was implemented and architected, where it is stored, and its reuse potential.}

\section*{Keywords}

personalised medicine; subgroup analysis; model-based recursive partitioning; unbiased trees;
treatment effect; random forest

\section*{Introduction}

\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles. A short comparison with software which implements similar functionality should be included in this section. }

\section*{Implementation and architecture}

Based on partykit, merely convenience interface.

\textcolor{blue}{How the software was implemented, with details of the architecture where relevant. Use of relevant diagrams is appropriate. Please also describe any variants and associated implementation differences.}

\section*{Quality control}

Tests, code in this manual, examples in documentation

\textcolor{blue}{Detail the level of testing that has been carried out on the code (e.g. unit, functional, load etc.), and in which environments. If not already included in the software documentation, provide details of how a user could quickly understand if the software is working (e.g. providing examples of running the software with sample input and output data). }

\section*{(2) Availability}
\vspace{0.5cm}
\section*{Operating system}

Should work on all operating systems that run R.
\textcolor{blue}{Please include minimum version compatibility.}

\section*{Programming language}
R (version 3.1.0 or higher)

\section*{Additional system requirements}
None.
\textcolor{blue}{E.g. memory, disk space, processor, input devices, output devices.}

\section*{Dependencies}
R, partykit package (version 1.2 or higher)

\section*{List of contributors}
Same as the authors:
Heidi Seibold, Achim Zeileis and Torsten Hothorn

\section*{Software location:}

{\bf Archive} \textcolor{blue}{(e.g. institutional repository, general repository) (required ??? please see instructions on journal website for depositing archive copy of software in a suitable repository)}

\begin{description}[noitemsep,topsep=0pt]
	\item[Name:] CRAN
	\item[Persistent identifier:] \url{https://cran.r-project.org/web/packages/model4you/index.html} %TODO upload to CRAN
	\item[Licence:] GPL-2
	\item[Publisher:] Heidi Seibold 
	\item[Version published:] \textcolor{blue}{The version number of the software archived.}
	\item[Date published:] \textcolor{blue}{dd/mm/yy}
\end{description}

{\bf Code repository} \textcolor{blue}{(e.g. SourceForge, GitHub etc.) (required)}

\begin{description}[noitemsep,topsep=0pt]
	\item[Name:] R-forge
	\item[Persistent identifier:] \url{https://r-forge.r-project.org/projects/partykit/}
	\item[Licence:] GPL-2
	\item[Date published:] 
\end{description}

\section*{Language}

English

\section*{(3) Reuse potential}

The software is intentionally written to make usage as simple as possible. We hope that researchers use it for analysis of:
\begin{itemize}
  \item Clinical trials
  \item A-B testing
  \item Other situations where the comparison of two decisions is of interest
\end{itemize}
We encourage users to use the partykit forum on R-forge
(\url{https://r-forge.r-project.org/forum/forum.php?forum_id=852}) or the party
tag on Stackoverflow (\url{http://stackoverflow.com/questions/tagged/party}) in
case of questions or problems.

\textcolor{blue}{Please describe in as much detail as possible the ways in which the software could be reused by other researchers both within and outside of your field. This should include the use cases for the software, and also details of how the software might be modified or extended (including how contributors should contact you) if appropriate. Also you must include details of what support mechanisms are in place for this software (even if there is no support).}

\section*{Acknowledgements}

\textcolor{blue}{Please add any relevant acknowledgements to anyone else who supported the project in which the software was created, but did not work directly on the software itself.}

\section*{Funding statement}
Heidi Seibold and Torsten Hothorn were financially supported by the Swiss
National Science Foundation (grant 205321\_163456).

\textcolor{blue}{If the software resulted from funded research please give the funder and grant number.}

\section*{Competing interests}

The authors declare that they have no competing interests.

\section*{References}

\textcolor{blue}{Please enter references in the Harvard style and include a DOI where available, citing them in the text with a number in square brackets, e.g. \\ }

\textcolor{blue}{[1] Piwowar, H A 2011 Who Shares? Who Doesn't? Factors Associated with Openly Archiving Raw Research Data. PLoS ONE 6(7): e18657. DOI: \\ http://dx.doi.org/10.1371/journal.pone.0018657.}

\vspace{2cm}

\rule{\textwidth}{1pt}

{ \bf Copyright Notice} \\
Authors who publish with this journal agree to the following terms: \\

Authors retain copyright and grant the journal right of first publication with the work simultaneously licensed under a  \href{http://creativecommons.org/licenses/by/3.0/}{Creative Commons Attribution License} that allows others to share the work with an acknowledgement of the work's authorship and initial publication in this journal. \\

Authors are able to enter into separate, additional contractual arrangements for the non-exclusive distribution of the journal's published version of the work (e.g., post it to an institutional repository or publish it in a book), with an acknowledgement of its initial publication in this journal. \\

By submitting this paper you agree to the terms of this Copyright Notice, which will apply to this submission if and when it is published by this journal.

\end{document}

### 2017-05-22 pkg/model4you/inst/JORS/manuscript_model4you.Rnw
first example in manuscript
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1362 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \documentclass{jors}
+
+%% packages
+\usepackage{amsmath}
+\usepackage{tikz}
+\usetikzlibrary{shapes,arrows,positioning}
 
@@
 \begin{document}
+
+<<setup, echo=FALSE>>=
+knitr::opts_chunk$set(cache = TRUE)
+@
 
@@
 }
+
+\begin{figure}
+\centering
+
+\tikzstyle{block} = [rectangle, draw,
+    text width=10em, text centered, rounded corners, minimum height=4em]
+\tikzstyle{exp} = [rectangle,
+    text width=10em, rounded corners, minimum height=4em]
+\tikzstyle{line} = [draw, ->, line width=0.2em]
+
+\begin{tikzpicture}[node distance = 7em, auto]
+  \node [block] (clin) {Clinical trial};
+  \node [block, below of=clin] (pdat) {Patient and treatment information, trial outcome};
+  \node [block, below of=pdat] (frst) {Model-based forest};
+  \node [block, below of=frst] (pmod) {Personalised models};
+  \node [block, below of=pmod] (plot) {Dependence plot};
+
+  \node [exp, right=1em of clin] {Real world};
+  \node [exp, right=1em of pdat] {Data};
+  \node [exp, right=1em of frst] {Black box model};
+  \node [exp, right=1em of pmod] {Interpretable model};
+  \node [exp, right=1em of plot] {Visual explanation};
+
+  \path [line] (clin) -- (pdat) node [pos=0.5, left] {data collection};
+  \path [line] (pdat) -- (frst) node [pos=0.5, left] {computation of forest};
+  \path [line] (frst) -- (pmod) node [pos=0.5, left] {similarity weights};
+  \path [line] (pmod) -- (plot) node [pos=0.5, left] {model coefficients};
+\end{tikzpicture}
+
+\caption{Personalised models workflow: from real world to interpretable
+visualisation.}
+\end{figure}
+
+<<GBSG2_prep>>=
+library("model4you")
+library("survival")
+library("ggplot2")
+library("gridExtra")
+data("GBSG2", package = "TH.data")
+
+set.seed(2017)
+@
+
+<<GBSG2_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for GBSG2 data.">>=
+bmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, model = TRUE)
+survplot(bmod)
+
+## extract coefficients and confidence intervals
+coefs <- coef(bmod)
+ci <- confint(bmod)
+coefs <- cbind(coefs, ci)
+
+## prepare for plotting
+digits <- 2
+cf <- format(round(coefs, digits), nsmall = digits)
+colnams <- colnames(cf)
+colnams[1] <- "theta"
+grid.table(cf, cols = colnams,
+           theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
+           vp = viewport(x = 0, y = 0, 
+                         width = 0.7, height = 0.4,
+                         just = c("left", "bottom")))
+@
+
+<<GBSG2_tree, fig.width = 9, fig.cap = "Personalised model tree for GBSG2 data.">>=
+tr <- pmtree(bmod)
+plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survplot, 
+                                          confint = TRUE))
+@
+
+<<GBSG2_pmod>>=
+frst <- pmforest(object = bmod, ntree = 100)
+
+coeffun <- function(model) {
+  ## model coefficients
+  coefs <- c(coef(model), scale = model$scale)
+
+  ## difference in median survival 
+  p = 0.5
+  coefs["median_s0"] <- qweibull(p = p, shape = 1/coefs["scale"], 
+                                 scale = exp(coefs["(Intercept)"]))
+  coefs["median_s1"] <- qweibull(p = p, shape = 1/coefs["scale"], 
+                                 scale = exp(coefs["(Intercept)"] + coefs[2]))
+  coefs["median_sdiff"] <- coefs["median_s1"] - coefs["median_s0"]
+
+  return(coefs)
+}
+coefs <- pmodel(x = frst, fun = coeffun, OOB = TRUE)
+
+## add to data and plot
+dpdat <- cbind(coefs, GBSG2)
+znam <- names(GBSG2)[!(names(GBSG2) %in% c("horTh", "time", "cens"))]
+
+dplot <- function(z) {
+  p <- ggplot(dpdat, aes_string(x = z, y = "median_sdiff")) 
+  if(is.factor(dpdat[, z])) {
+    p + geom_boxplot()
+  } else {
+    p + geom_point(alpha = 0.2) + 
+      geom_smooth(fill = NA, method = "loess")
+  }
+}
+dps <- lapply(znam, dplot)
+do.call(grid.arrange, dps)
+
+do.call(grid.arrange, lapply(dps, function(p) {
+  dmin <- min(dpdat$median_sdiff)
+  dq <- quantile(dpdat$median_sdiff, 0.99)
+  p + coord_cartesian(ylim = c(dmin, dq))
+  }))
+@
 

### 2017-05-23 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1370 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \section{Abstract}
-Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine.
+Identification of subgroups of patients for whom treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine.
 Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research.
@@
 The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters.
-In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees.
-Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effect, on average.
+In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than linear-model based recursive partitioning and mixed-effects regression trees.
+Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effects, on average.
 We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression.
@@
 
-Due to the recursive nature of the splitting, the rectangular regions of the partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}.
+Due to the recursive nature of the splitting, the rectangular regions of a recursive partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}.
 The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety.
@@
 Summary statistics of these variables are provided in Table~\ref{tab:sum_stats}.
-Each patient was part of one of ten clusters, each having a different value for the random intercept, which was generated from a standard normal distribution and uncorrelated with the partitioning variables.
+Each patient was part of one of ten clusters, each having a different value for the random intercept, which were generated from a standard normal distribution and uncorrelated with the partitioning variables.
 
@@
 
+\begin{minipage}{\linewidth}
 \begin{description}
@@
   \item[Step 3:] Repeat Steps~1 and~2 until convergence.
+
 \end{description}
+\end{minipage}
+\\
+\\
 
@@
 
-To assess the performance of GLMM trees, we carried out three simulation studies: In Study~I we assessed and compared the accuracy of GLMM trees, linear-model based MOB and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions.
+To assess the performance of GLMM trees, we carried out three simulation studies: In Study~I we assessed and compared the accuracy of GLMM trees, linear-model based MOB (LM trees) and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions.
 In Study~II, we assessed and compared the Type-I error of GLMM trees and linear-model based MOB in datasets without treatment-subgroup interactions.
@@
 R \cite{R} was used for data generation and analyses.
-The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating LM trees, using the \verb|lmtree| function.
+The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15, pkg:partykit}) was employed for estimating LM trees, using the \verb|lmtree| function.
 For estimation of LMM trees the \verb|lmertree| function of the \verb|glmertree| package (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge) was used.
@@
 
-The \verb|REEMtree| package \cite{SelaySimo11} was employed for estimating MERTs, using default settings.
-For estimating LMMs the \verb|lmer| function function from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
+The \verb|REEMtree| package (version 0.9.3; \citeNP{SelaySimo11}) was employed for estimating MERTs, using default settings.
+For estimating LMMs the \verb|lmer| function function from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac15, pkg:lme4}) was employed, using restricted maximum likelihood (REML) estimation.
 The \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in LMMs in Study~III.
@@
 In this simulation design, some of the potential partitioning covariates are true partitioning covariates, the others are noise variable.
-Therefore, an extra level was added to the sixth facet of the \textit{General simulation design}:
+Therefore, in addition to the \textit{General simulation design}, the following facet was added in the this study:
 
@@
 
-An additional facet was added to assess the effect of the magnitude of treatment-effect differences:
+To assess the effect of the magnitude of treatment-effect differences, the following facet was added in this study:
 
@@
 Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences.
-To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets.
+To prevent overly optimistic estimates of predictive accuracy, predictive accuracy was assessed using test datasets.
 Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from `new' clusters.
 
-For MERT, predicted treatment-effect differences were obtained by fitting two MERTs on the training data: one using observations in the first treatment condition and one using observations in the second treatment condition.
-Predictions of treatment-effect differences for test observations were obtained by dropping test observations down both trees and taking the difference between the two predicted values.
-This approach was taken as it yielded higher predictive accuracy than the alternative of fitting a single MERT using treatment as a potential partitioning variable.
+The best approach for including treatment effects in MERTs is not completely obvious.
+Firstly, a single MERT may be fitted, where treatment is included as one of the potential partitioning variables.
+Predictions of treatment-effect differences can then be obtained by dropping test observations down the tree twice, once for every level of the treatment indicator.
+Secondly,  two MERTs may be fitted: one using observations in the first treatment condition and one using observations in the second treatment condition.
+Predictions of treatment-effect differences can then be obtained by dropping a test observation down each of the two trees.
+We tried both approaches: the second approach yielded higher predictive accuracy, as the first approach often did not pick up the treatment indicator as a predictor.
+Therefore, we have taken the second approach of fitting two MERTs to each dataset in our simulations.
 
@@
 With larger sample sizes, LM trees showed about the true tree size, on average.
+Tree size of MERTs was not assessed, as a single true tree size for MERTs could not be derived from the design in Figure~\ref{fig:modelC}.
 
@@
 When random intercepts were not correlated with one of the $U_k$ variables, LMM trees outperformed LM trees only when sample size was small (i.e., $N = 200$).
+Tree accuracy of MERTs was not assessed, as a single accurate tree structure for MERTs could not be derived from the design in Figure~\ref{fig:modelC}.
 
@@
 
-The predicted treatment-effect differences of LMM trees show an average correlation of .93 (SD = .13) with the true differences.
+The predicted treatment-effect differences of LMM trees show an average correlation of .93 ($\mathrm{SD = .13}$) with the true differences.
 LM trees and MERTs show lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively.
@@
 
-To assess the effect of the treatment-effect difference $\beta$, an additional facet was added to the \textit{General simulation design}: 
+To assess the effect of the treatment-effect difference $\beta$, in addition to the \textit{General simulation design}, the following facet was added in this study: 
 \begin{enumerate}
@@
 At the same time, linear regression models with pre-specified interaction terms may perform well in recovering continuous interactions, but may have difficulty in recovering piecewise interactions.
-Therefore, in the third simulation study, we added a seventh facet to the \textit{General simulation design} described above:
+Therefore, in the third simulation study, in addition to the \textit{General simulation design} described above, the following facet was added:
 
@@
 
-For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure~\ref{fig:modelC} was also used.
+For generating datasets with both piecewise and continuous interactions, the partition as depicted in Figure~\ref{fig:modelC} was also used.
 In addition, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes now comprised continuous main and (treatment) interaction effects of the partitioning variables.
-In other words, the partitioning variables $U_2$, $U_1$ and $U_5$ appear are also part of the linear predictor $x_i$, as part of the terms presented in Table~\ref{tab:continuous_terms}.
+In other words, the partitioning variables $U_2$, $U_1$ and $U_5$ also appear in the linear predictor $x_i$, as part of the terms presented in Table~\ref{tab:continuous_terms}.
 The corresponding node-specific $\beta_j$ parameters are also presented in Table~\ref{tab:continuous_terms}.
@@
 The interaction terms were created using centered $U_k$ variables, calculated by subtracting their variable means.
-Again, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
-
-In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by the single $\beta$ column in Table~\ref{tab:continuous_terms}.
+Again, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$.
+
+In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by terms and the single column for $\beta$ in Table~\ref{tab:continuous_terms}.
 The outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$.
@@
 
-As the outcome variable is continuousTo this end, the MOB algorithm tests for parameter stability with respect to
-each of a set of auxiliary covariates, we employed an identity link and Gaussian response distribution.
+As the outcome variable is continuous, we employed an identity link and Gaussian response distribution.
 The resulting GLMM trees will therefore be refered to as LMM trees.
@@
 For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$).
-The estimated intraclass correlation coefficient for the GLMM tree was .05.
+The estimated intraclass correlation coefficient for the LMM tree was .05.
 
@@
 
-The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: $b = 2.29$, $p = .002$; interaction with treatment: $b = -2.10$, $p = .028$).
-Also, the GLMM indicated an interaction between treatment and age ($b = .10$, $p = .018$).
+The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the LMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: $b = 2.29$, $p = .002$; interaction with treatment: $b = -2.10$, $p = .028$).
+Also, the LMM indicated an interaction between treatment and age ($b = .10$, $p = .018$).
 
@@
 
-We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters.
-We hypothesized GLMM trees to be well suited for the detection of treatment-subgroup interactions in clustered datasets and confirmed this by our simulation studies.
-
-GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions, outperforming GLM trees without random effects.
-In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average.
-In datasets without treatment-subgroup effects, GLMM trees showed a Type-I error rate of 4\%, compared with a Type-I error rate of 33\% for GLM trees.
-
-Better performance of GLMM trees was mostly observed when random effects were sizeable and correlated with potential partitioning variables.
-In those circumstances, GLM trees are likely to detect spurious splits and subgroups.
-At the same time, GLM trees showed less power to detect subgroups in the presence of random effects.
-As expected, accuracy of MERTs was not affected by the presence of random effects, but only approached GLMM trees' accuracy in datasets with the largest sample and effect sizes.
-Especially with smaller effect and sample sizes (i.e., Cohen's $d=.5$ and/or $N = 200$), GLMM tree outperformed the other tree methods.
-Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM trees may provide a useful tool for subgroup detection in those instances.
-
-In the absence of random effects, GLM and GLMM trees yielded very similar predictive accuracy.
-This finding is of practical importance, indicating that application of GLMM trees will not `hurt' predictive accuray when in fact, random effects are absent.
-
-Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy on average.
-GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise.
-GLMM trees performed better in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited for exploration than GLMMs.
-
-In the Application, we obtained similar results: the GLMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the other two methods (GLM tree and a GLMM with pre-specified interactions).
-The GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.
-
-\subsubsection{Limitations}
-
-\readme{I would emphasize early on that GLMM trees rely on the properties of their building blocks, namely GLMMs and MOBs.
-The effect of misspecifications are studied to some degree for both methods.
-For GLMM one could cite papers discussing this.
-For MOB, or trees more broadly, there are also some references.
-One of these could be the `To Split or to Mix?' paper by Frick, Strobl, and Zeileis.}
-
-Recursive partitioning methods were origfinally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}.
-GLMM trees take MOB and GLMMs as building blocks, thereby relying on the properties of these building blocks.
-As a result, GLMM trees may inherit their sensitivities to model misspecification as well as their robustness against it.
-We will address potential types of misspecification in fitting GLMM trees:
-
-\readme{MF: GLMM tree at least tackles one type of random-effects misspecification: ignoring them.
-\cite{BarryLevy13} and \cite{Barr13} advice to keep it maximal is at repeated measures designs.
-\cite{McCuyNeuh11}: "A key conclusion is the large degree of robustness of maximum likelihood for a wide variety of commonly encountered situation".
-....
-fit fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors.
-Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM trees.}
-
-Furthermore, misspecification of partitioning and fixed- and random-effects variables will also reduce accuracy of the resulting GLMM trees.
-If relevant variables are omitted or incorrectly specified, GLMM trees can only approximate the true subgroups using the specified variables.
-Our simulations indicate that LM trees detect spurious subgroups as a result of misspecifying (that is, not including) the random effects.
-Reduced accuracy and spurious splits can also be expected to occur when relevant random effects are not included when specifying the GLMM tree model.
-Furthermore, as the random effects are estimated globally, misspecification of the random effects may have a strong impact on the resulting GLMM tree model.
-
-Another source of misspecification is the inclusion of irrelevant variables.
-Although our simulations indicate that the performance of GLMM trees was not negatively affected by increasing the number of noise variables specified for partitioning from 2 to 12, the power to detect subgroups may be reduced with much larger numbers of noise variables.
-Including irrelevant variables in the random or fixed effects may also negatively affect accuracy of GLMM trees, but we have not assessed this in our study.
-
-Users can evaluate potential misspecification by inspecting residuals and predictions, and assessing tree stability.
-The vigette included as Supplementary Material shows how residuals andrandom-effects predictions can be plotted to assess potential misspecification.
-In the Application we have shown how tree stability can be assessed using the methods suggested by \cite{PhilyZeil16}.
-Future studies may focus on the effects of possible misspecification on the performance of GLMM trees.
-
-As GLMM trees fit more complex models than non-parametric tree-based methods, like CART for example, larger sample sizes are needed to fit the model.
-How much larger likely depends on the complexity of the specified model: Our simulations show that with fixed and random-effects specifications with only a single predictor, a sample size of 200 is sufficient to detect subgroups with moderate differences in treatment effect.
-More complex fixed- and random-effects specifications will require larger sample sizes; how these affect the performance of GLMM trees requires further research.
-
-\subsubsection{Future directions}
-
-\readme{MF: Add something about implementing non-normal non-parametric or mixture ranef}
-
-In the Introduction we mentioned several existing tree-based methods for treatment-subgroup interaction detection.
+We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of global random-effects parameters.
+We hypothesized GLMM trees to be well suited for the detection of treatment-subgroup interactions in clustered datasets.
+We confirmed this through our simulation studies and by applying the algorithm to an existing dataset on the effects of depression treatments.
+
+Our simulations focused on the performance of the GLMM tree algorithm in datasets with continuous response variables.
+The resulting LMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions, outperforming LM trees without random effects.
+In terms of predictive accuracy, LMM trees outperformed LM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average.
+In datasets without treatment-subgroup effects, LMM trees showed a rather low Type-I error rate of 4\%, compared with a Type-I error rate of 33\% for LM trees.
+
+Better performance of LMM trees was mostly observed when random effects were sizeable and correlated with potential partitioning variables.
+In those circumstances, LM trees are likely to detect spurious splits and subgroups.
+Especially with smaller effect and sample sizes (i.e., Cohen's $d=.5$ and/or $N = 200$), LMM trees outperformed the other tree methods.
+As such effect and sample sizes are quite common in multi-center clinical trials, the GLMM tree algorithm may provide a useful tool for subgroup detection in those instances.
+
+In the absence of random effects, LM and LMM trees yielded very similar predictive accuracy.
+This finding is of practical importance, indicating that application of the GLMM tree algorithm will not reduce accuracy when in fact, random effects are absent from the data.
+
+Compared to LMMs with pre-specified interactions, LMM trees provided somewhat better accuracy, on average.
+LMM trees performed much better than LMMs when interactions were at least partly piecewise.
+However, the performance ofLMM trees deteriorated when the interactions were purely continuous functions of the predictor variables, in which case LMMs with pre-specified interactions performed very well.
+The performance of LMMs deteriorated with a larger number of predictor variables, whereas the performance of LMM trees was not affected by this, confirming our expectation that LMM trees are better suited for exploration than LMMs.
+
+In the Application, we found the LMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the LM tree and the LMM with pre-specified interactions.
+The LMM trees obtained over repeated subsamples of the training data proved to be relatively stable.
+
+\subsubsection{Limitations and future directions}
+
+Recursive partitioning methods were originally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}.
+The GLMM tree algorithm takes MOB trees and GLMMs as building blocks and thereby inherits some of their sensitivity to model misspecifications.
+
+As with GLMMs, misspecification of the random effects may negatively affect the accuracy of GLMM trees.
+Previous research on GLMMs has shown that misspecifying the shape of the random-effects distribution reduces the accuracy of random-effects predictions, but has little effects on the fixed-effects parameter estimates \cite<e.g., >{NeuhyMcCu12}.
+This indicates that for GLMM trees, misspecification of the shape of the random-effects distribution will affect random-effects predictions, but will have little effect on the estimated tree.
+The incorrect specification of a predictor as having a fixed instead of a random effect may be more likely to result in spurious splits or poorer parameter estimates.
+Therefore, to check for model misspecification, users are advised to visually inspect residuals and predicted random effects.
+The tutorial included in the supplementary material shows how residuals and random effects predictions can be obtained and plotted.
+
+As with MOB trees, GLMM trees will perform best when the true data structure is tree-shaped.
+Trees may need many splits to approximate structures with a different shape, which may be more unstable under small perturbations.
+Therefore, in the Application we have shown how tree stability can be assessed using the methods developed by \citeA{PhilyZeil16}.
+Furthermore, when relevant partitioning variables are omitted, GLMM tree can only approximate the subgroups using the specified variables.
+Although our simulations did not involve scenarios where relevant partitioning variables were omitted, the results of \citeA{FricyStro14} indicate that MOB trees can detect parameter instability even when covariates are only loosely connected to the group structure.
+
+GLMM trees rely more on user specification of predictor variables than methods like CART and MERT.
+On the one hand, this increases the danger of model misspecification.
+On the other hand, it increases power and accuracy if the fixed-effects predictor variable(s) are correctly specified.
+For example, the best approach for assessing treatment-effect differences with MERTs is not completely obvious.
+We tried two approaches in our simulations: firstly, fitting a single MERT with the treatment indicator as one of the potential partitioning variables and secondly, fitting separate MERTs for each level of the treatment indicator.
+With the first approach, MERTs often did not detect the treatment indicator as a predictor variable in datasets with treatment-subgroup interactions.
+The second approach yielded better accuracy, indicating that MERTs profit from user specification of the relevant predictor variable(s).
+The GLMM tree algorithm provides a straightforward approach for including relevant fixed-effects predictor variables in the model.
+Our results suggest this yields higher predictive accuracy if the model is correctly specified.
+However, our simulations have not assessed the effect of misspecification of the fixed-effects predictor variable(s) and such misspecification may reduce predictive accuracy.
+
+For recursive partitioning methods fitting local parametric models only, the danger of model misspecification may be limited \cite<e.g., >{Ciam91, SiciyMass08, SuyWang04}.
+GLMM trees, however, fit a global random-effect model in addition to local fixed-effects regression models and may therefore be less robust against model misspecification.
+Further research on the effects of model misspecifications on the performance of GLMM trees is therefore needed.
+
+In the Introduction, we mentioned several existing tree-based methods for treatment-subgroup interaction detection.
 These methods have different objectives and there is not yet an agreed-upon single best method.
@@
 performed worst under some conditions of the simulation study in terms of the Type I error rate.
-Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations and comparisons only focused on GLMM trees and GLM trees.
+Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations and comparisons mostly focused on LMM trees and LM trees.
 

### 2017-05-25 pkg/model4you/inst/JORS/man_model4you.Rnw
renaming manuscript (weird latex error -> black background)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1379 255f4249-2263-4417-b1ea-8498779e27a7
flow charts and model for one observation (does not work yet)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1381 255f4249-2263-4417-b1ea-8498779e27a7

--- 2017-05-22 pkg/model4you/inst/JORS/manuscript_model4you.Rnw
@@
 <<setup, echo=FALSE>>=
-knitr::opts_chunk$set(cache = TRUE)
+knitr::opts_chunk$set(cache = TRUE, message = FALSE)
 @
@@
 \centering
+\include{treefamily}
+\caption{The family of model-based trees and forests.}
+\end{figure}
 
-\tikzstyle{block} = [rectangle, draw,
-    text width=10em, text centered, rounded corners, minimum height=4em]
-\tikzstyle{exp} = [rectangle,
-    text width=10em, rounded corners, minimum height=4em]
-\tikzstyle{line} = [draw, ->, line width=0.2em]
-
-\begin{tikzpicture}[node distance = 7em, auto]
-  \node [block] (clin) {Clinical trial};
-  \node [block, below of=clin] (pdat) {Patient and treatment information, trial outcome};
-  \node [block, below of=pdat] (frst) {Model-based forest};
-  \node [block, below of=frst] (pmod) {Personalised models};
-  \node [block, below of=pmod] (plot) {Dependence plot};
-
-  \node [exp, right=1em of clin] {Real world};
-  \node [exp, right=1em of pdat] {Data};
-  \node [exp, right=1em of frst] {Black box model};
-  \node [exp, right=1em of pmod] {Interpretable model};
-  \node [exp, right=1em of plot] {Visual explanation};
-
-  \path [line] (clin) -- (pdat) node [pos=0.5, left] {data collection};
-  \path [line] (pdat) -- (frst) node [pos=0.5, left] {computation of forest};
-  \path [line] (frst) -- (pmod) node [pos=0.5, left] {similarity weights};
-  \path [line] (pmod) -- (plot) node [pos=0.5, left] {model coefficients};
-\end{tikzpicture}
-
+\begin{figure}
+\centering
+\include{pmworkflow}
 \caption{Personalised models workflow: from real world to interpretable
@@
 
+<<GBSG2_forest>>=
+frst <- pmforest(object = bmod, ntree = 100)
+@
+
+<<GBSG_pmod1>>=
+newpatient_i <- data.frame(age = 60, menostat = "Post", tsize = 20, tgrade = "II",
+                         pnodes = 3, progrec = 50, estrec = 60)
+mod_i <- pmodel(x = frst, fun = identity, newdata = newpatient_i)
+@
+
 <<GBSG2_pmod>>=
-frst <- pmforest(object = bmod, ntree = 100)
-
 coeffun <- function(model) {

### 2017-05-25 pkg/model4you/inst/JORS/pmworkflow.tex
flow charts and model for one observation (does not work yet)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1381 255f4249-2263-4417-b1ea-8498779e27a7

--- 2017-05-18 pkg/model4you/inst/JORS/manuscript_model4you.Rnw
@@
-%% Journal of Open Research Software Latex template -- Created By Stephen Bonner and John Brennan, Durham Universtiy, UK.
+\tikzstyle{block} = [rectangle, draw,
+    text width=13em, text centered, rounded corners, minimum height=3em]
+\tikzstyle{exp} = [rectangle,
+    text width=10em, rounded corners, minimum height=4em]
+\tikzstyle{line} = [draw, ->, line width=0.2em]
 
-\documentclass{jors}
+\begin{tikzpicture}[node distance = 7em, auto]
+  \node [block] (clin) {Clinical trial};
+  \node [block, below of=clin] (pdat) {Patient and treatment information, trial outcome};
+  \node [block, below of=pdat] (frst) {Model-based forest};
+  \node [block, below of=frst] (pmod) {Personalised models};
+  \node [block, below of=pmod] (plot) {Dependence plot};
 
-%% Set the header information
-\pagestyle{fancy}
-\definecolor{mygray}{gray}{0.6}
-\renewcommand\headrule{}
-\rhead{\footnotesize 3}
-\rhead{\textcolor{gray}{UP JORS software Latex paper template version 0.1}}
+  \node [exp, right=1em of clin] {Real world};
+  \node [exp, right=1em of pdat] {Data};
+  \node [exp, right=1em of frst] {Black box model};
+  \node [exp, right=1em of pmod] {Interpretable model};
+  \node [exp, right=1em of plot] {Visual explanation};
 
-\begin{document}
-
-{\bf Software paper for submission to the Journal of Open Research Software} \\
-
-To complete this template, please replace the blue text with your own. The paper has three main sections: (1) Overview; (2) Availability; (3) Reuse potential. \\
-
-Please submit the completed paper to: editor.jors@ubiquitypress.com
-
-\rule{\textwidth}{1pt}
-
-\section*{(1) Overview}
-
-\vspace{0.5cm}
-
-\section*{Title}
-model4you: An R package for personalised treatment effect estimation
-
-\section*{Paper Authors}
-1. Seibold, Heidi; 2. Zeileis, Achim; 3. Hothorn, Torsten
-
-\section*{Paper Author Roles and Affiliations}
-1. PhD student; Biostatistics Department, Epidemiology, Biostatistics \& Prevention
-Institute, University of Zurich
-
-2. Professor; Department of Statistics, Faculty of Economics and Statistics, University of
-Innsbruck
-
-3. Professor; Biostatistics Department, Epidemiology, Biostatistics \& Prevention
-Institute, University of Zurich
-
-\section*{Abstract}
-
-\textcolor{blue}{A short (ca.
-100 word) summary of the software being described: what problem the software addresses, how it was implemented and architected, where it is stored, and its reuse potential.}
-
-\section*{Keywords}
-
-personalised medicine; subgroup analysis; model-based recursive partitioning; unbiased trees;
-treatment effect; random forest
-
-\section*{Introduction}
-
-\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles.
-A short comparison with software which implements similar functionality should be included in this section.
-}
-
-\section*{Implementation and architecture}
-
-Based on partykit, merely convenience interface.
-
-\textcolor{blue}{How the software was implemented, with details of the architecture where relevant.
-Use of relevant diagrams is appropriate.
-Please also describe any variants and associated implementation differences.}
-
-\section*{Quality control}
-
-Tests, code in this manual, examples in documentation
-
-\textcolor{blue}{Detail the level of testing that has been carried out on the code (e.g.
-unit, functional, load etc.), and in which environments.
-If not already included in the software documentation, provide details of how a user could quickly understand if the software is working (e.g.
-providing examples of running the software with sample input and output data).
-}
-
-\section*{(2) Availability}
-\vspace{0.5cm}
-\section*{Operating system}
-
-Should work on all operating systems that run R.
-\textcolor{blue}{Please include minimum version compatibility.}
-
-\section*{Programming language}
-R (version 3.1.0 or higher)
-
-\section*{Additional system requirements}
-None.
-\textcolor{blue}{E.g. memory, disk space, processor, input devices, output devices.}
-
-\section*{Dependencies}
-R, partykit package (version 1.2 or higher)
-
-\section*{List of contributors}
-Same as the authors:
-Heidi Seibold, Achim Zeileis and Torsten Hothorn
-
-\section*{Software location:}
-
-{\bf Archive} \textcolor{blue}{(e.g.
-institutional repository, general repository) (required ???
-please see instructions on journal website for depositing archive copy of software in a suitable repository)}
-
-\begin{description}[noitemsep,topsep=0pt]
-	\item[Name:] CRAN
-	\item[Persistent identifier:] \url{https://cran.r-project.org/web/packages/model4you/index.html} %TODO upload to CRAN
-	\item[Licence:] GPL-2
-	\item[Publisher:] Heidi Seibold 
-	\item[Version published:] \textcolor{blue}{The version number of the software archived.}
-	\item[Date published:] \textcolor{blue}{dd/mm/yy}
-\end{description}
-
-{\bf Code repository} \textcolor{blue}{(e.g. SourceForge, GitHub etc.) (required)}
-
-\begin{description}[noitemsep,topsep=0pt]
-	\item[Name:] R-forge
-	\item[Persistent identifier:] \url{https://r-forge.r-project.org/projects/partykit/}
-	\item[Licence:] GPL-2
-	\item[Date published:] 
-\end{description}
-
-\section*{Language}
-
-English
-
-\section*{(3) Reuse potential}
-
-The software is intentionally written to make usage as simple as possible. We hope that researchers use it for analysis of:
-\begin{itemize}
-  \item Clinical trials
-  \item A-B testing
-  \item Other situations where the comparison of two decisions is of interest
-\end{itemize}
-We encourage users to use the partykit forum on R-forge
-(\url{https://r-forge.r-project.org/forum/forum.php?forum_id=852}) or the party
-tag on Stackoverflow (\url{http://stackoverflow.com/questions/tagged/party}) in
-case of questions or problems.
-
-\textcolor{blue}{Please describe in as much detail as possible the ways in which the software could be reused by other researchers both within and outside of your field.
-This should include the use cases for the software, and also details of how the software might be modified or extended (including how contributors should contact you) if appropriate.
-Also you must include details of what support mechanisms are in place for this software (even if there is no support).}
-
-\section*{Acknowledgements}
-
-\textcolor{blue}{Please add any relevant acknowledgements to anyone else who supported the project in which the software was created, but did not work directly on the software itself.}
-
-\section*{Funding statement}
-Heidi Seibold and Torsten Hothorn were financially supported by the Swiss
-National Science Foundation (grant 205321\_163456).
-
-\textcolor{blue}{If the software resulted from funded research please give the funder and grant number.}
-
-\section*{Competing interests}
-
-The authors declare that they have no competing interests.
-
-\section*{References}
-
-\textcolor{blue}{Please enter references in the Harvard style and include a DOI where available, citing them in the text with a number in square brackets, e.g. \\ }
-
-\textcolor{blue}{[1] Piwowar, H A 2011 Who Shares?
-Who Doesn't?
-Factors Associated with Openly Archiving Raw Research Data.
-PLoS ONE 6(7): e18657.
-DOI: \\ http://dx.doi.org/10.1371/journal.pone.0018657.}
-
-\vspace{2cm}
-
-\rule{\textwidth}{1pt}
-
-{ \bf Copyright Notice} \\
-Authors who publish with this journal agree to the following terms: \\
-
-Authors retain copyright and grant the journal right of first publication with the work simultaneously licensed under a  \href{http://creativecommons.org/licenses/by/3.0/}{Creative Commons Attribution License} that allows others to share the work with an acknowledgement of the work's authorship and initial publication in this journal.
-\\
-
-Authors are able to enter into separate, additional contractual arrangements for the non-exclusive distribution of the journal's published version of the work (e.g., post it to an institutional repository or publish it in a book), with an acknowledgement of its initial publication in this journal.
-\\
-
-By submitting this paper you agree to the terms of this Copyright Notice, which will apply to this submission if and when it is published by this journal.
-
-\end{document}
+  \path [line] (clin) -- (pdat) node [pos=0.5, left] {data collection};
+  \path [line] (pdat) -- (frst) node [pos=0.5, left] {computation of forest};
+  \path [line] (frst) -- (pmod) node [pos=0.5, left] {similarity weights};
+  \path [line] (pmod) -- (plot) node [pos=0.5, left] {model coefficients};
+\end{tikzpicture}

### 2017-05-25 pkg/model4you/inst/JORS/treefamily.tex
flow charts and model for one observation (does not work yet)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1381 255f4249-2263-4417-b1ea-8498779e27a7

% \documentclass{article}
% \usepackage{tikz}
% \usetikzlibrary{shapes,arrows,positioning}
% 
% \begin{document}

\tikzstyle{every node}=[font=\scriptsize\ttfamily]
% Nodes
\tikzstyle{fun} = [rectangle, draw, align=center, fill=white,
    minimum width=8em, text centered, rounded corners, minimum height=1.5em]
% Define the layers to draw the diagram
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}

\begin{tikzpicture}[node distance = 7em, auto]
%% Draw nodes
  \node (ctree) [fun] {partykit::ctree};
  \node (mob) [fun, right=1em of ctree] {partykit::mob};

  \node (f1) [below=2em of mob] {};
  \node [fun, right=1em of f1] (lmtree) {partykit::lmtree};
  \node [fun, below=0.5em of lmtree] (glmtree) {partykit::glmtree};

  \node (f2) [below=1.3em of glmtree] {};
  \node [fun, right=2em of lmtree] (lmertree) {glmertree::lmertree};
  \node [fun, below=0.5em of lmertree] (glmertree) {glmertree::glmertree};
  \node [fun, above=1em of lmertree] (lagsarlmtree) {lagsarlmtree::lagsarlmtree}; 
  \node [fun, right=1em of f2] (palmtree) {palmtree::palmtree};

  \node [fun, below=4em of glmtree] (raschtree) {psychotree::raschtree};
  \node [fun, below=0.5em of raschtree] (rstree) {psychotree::rstree};
  \node [fun, below=0.5em of rstree] (pctree) {psychotree::pctree};
  \node [fun, below=0.5em of pctree] (bttree) {psychotree::bttree};

  \node [fun, below=1.2em of bttree] (betatree) {betareg::betatree};

  \node (ytrafo) [ellipse, draw, fill=white, below=9em of ctree] {ytrafo};
  \node (f2) [below=22em of mob] {};
  \node [fun, left=0em of f2] (disttree) {disttree::disttree};
  \node [fun, below=1.2em of disttree] (trafotree) {trtf::trafotree};
  \node [fun, below=1.2em of trafotree] (pmtree) {model4you::pmtree};
  \node [fun, right=1em of pmtree] (pmforest) {model4you::pmforest};
  \node [fun, above=1.2em of pmforest] (traforest) {trtf::trafoforest};
  \node [fun, above=1.2em of traforest] (distforest) {disttree::distforest};

  \node (cforest) [fun, below=30em of ctree] {partykit::cforest};
  \node (ytrafo2) [ellipse, draw, fill=white, right=13em of cforest] {ytrafo};

%% Draw arrows
\begin{pgfonlayer}{background}
  \draw[->] (mob) |- (lmtree);
  \draw[->] (mob) |- (glmtree);
  \draw[->, dashed] (lmtree) |- (lmertree);
  \draw[->, dashed] (glmtree) |- (glmertree);
  \draw[->] (lmtree) -- ++(1,0) |- (lagsarlmtree);
  \draw[->, dashed] (lmtree) |- (palmtree);
  % \draw[->, dashed] (glmtree) |- (palmtree);
  \draw[->] (mob) |- (raschtree);
  \draw[->] (mob) |- (rstree);
  \draw[->] (mob) |- (pctree);
  \draw[->] (mob) |- (bttree);
  \draw[->] (mob) |- (betatree);
  \draw[->, dotted] (ytrafo) -- (mob);
  \draw[->] (ctree) |- (pmtree);
  \draw[->] (ctree) |- (disttree);
  \draw[->] (mob) -- (disttree);
  \draw[->] (ctree) |- (trafotree);
  \draw[->, thick] (pmtree) -- (pmforest);
  \draw[->, thick] (disttree) -- (distforest);
  \draw[->, thick] (trafotree) -- (traforest);
  \draw[->, thick] (ctree) -- ++(-2,0) |- (cforest);
  \draw[->] (cforest) -- ++(19em,0) |- (pmforest);
  \draw[->] (cforest) -- ++(19em,0) |- (distforest);
  \draw[->] (cforest) -- ++(19em,0) |- (traforest);

\end{pgfonlayer}
\end{tikzpicture}

% \end{document}

### 2017-05-30 pkg/model4you/inst/JORS/man_model4you.Rnw
abstract and start intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1397 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usetikzlibrary{shapes,arrows,positioning}
+
+%% commands
 
@@
 <<setup, echo=FALSE>>=
+library("knitr")
 knitr::opts_chunk$set(cache = TRUE, message = FALSE)
+
+knit_hooks$set(inline = function(x) { 
+  if (is.numeric(x)) return(knitr:::format_sci(x, 'latex')) 
+  highr:::hi_latex(x) 
+}) 
 @
@@
 \section*{Abstract}
+Typical models estimating treatment effecs assume that the treatment effect is the same for all 
+individuals observed and 
+for possible future individuals. Model-based recursive partitioning allows to relax this assumption.
+With model-based trees one can compute treatment effects for different strata of individuals. The
+strata are found in a data driven fashion and depend on characteristics of the individuals. 
+Model-based random forests allow for a similarity estimation between individuals in terms of 
+model estimates and through this we can estimate personalised models.
+The R package \emph{model4you} (available on CRAN) implements these stratified and personalised models with a
+focus on ease of use and interpretability so that clinicians and other users can take
+the model they usually use for the estimation of the average treatment 
+effect and with a few lines of code get a visualisation that is easy to understand and interpret.
 
-\textcolor{blue}{A short (ca.
+%\textcolor{blue}{A short (ca.
 100 word) summary of the software being described: what problem the software addresses, how it was implemented and architected, where it is stored, and its reuse potential.}
@@
 
-\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles.
+%\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles.
 A short comparison with software which implements similar functionality should be included in this section.
 }
+Studies in various fields randomly assign individuals one of two groups with
+different exposure and then measure a response. For example, in clinical trials
+patients are assigned to one of two treatment groups where usually one treatment
+group receives a new treatment or drug and the other treatment group receives
+the standard of care or a placebo.
+
+Treatment effect estimation is often done using simple models with the binary
+treatment indicator as only covariate. In the example of a clinical trial the
+treatment indicator would be $1$ if the patient receives the new treatment and
+$0$ if the patient receives standard of care.
+In R such a simple model can be estimated as follows:
+<<eval=FALSE>>=
+base_model <- model(response ~ treatment, data)
+@
+with \Sexpr{'response'} being the response measured, \Sexpr{'treatment'} being the
+treatment indicator and \Sexpr{'data'} being the data set containing these variables.
+The function \Sexpr{'model()'} can be replaced for example by \Sexpr{'lm()'} to estimate
+a linear model, \Sexpr{'glm()'} to estimate a generalised linear model or 
+\Sexpr{'survreg()'} to estimate a parametric survival model. 
+These models estimate intercept and treatment effect for
+all individuals in the \Sexpr{'data'} and allow for predicting the response 
+of other individuals given they do or don't receive the treatment of interest.
+
+If you do not believe that all individuals have the same intercept and treatment
+effect the R package \emph{model4you} gives you two options to loosen up this 
+strict model: 
+\paragraph{(1) Model-based trees} identify subgroups where within the subgroups the model 
+parameters are similar and between groups the model parameters are different. 
+<<eval=FALSE>>=
+strat_models <- pmtree(base_model)
+@
+
+\paragraph{(2) Personalised models} use model-based random forests to estimate similarity
+of individuals in terms of model parameters. For each individual a personalised model can
+be estimated based on a weighted set of the original data, where the similarity measure 
+corresponds to the weight.
+<<eval=FALSE>>=
+pers_models <- pmodel(base_model)
+@
 
@@
 library("survival")
-library("ggplot2")
-library("gridExtra")
 data("GBSG2", package = "TH.data")

### 2017-06-01 pkg/model4you/inst/JORS/man_model4you.Rnw
second example, references, size in tikz, ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1425 255f4249-2263-4417-b1ea-8498779e27a7

@@
 %% packages
+\usepackage{natbib}
+
+\usepackage{subfig}
 \usepackage{amsmath}
@@
 \usetikzlibrary{shapes,arrows,positioning}
-
-%% commands
 
@@
 library("knitr")
-knitr::opts_chunk$set(cache = TRUE, message = FALSE)
+knitr::opts_chunk$set(cache = TRUE, message = FALSE, fig.align = "center")
 
@@
 \paragraph{(1) Model-based trees} identify subgroups where within the subgroups the model 
-parameters are similar and between groups the model parameters are different. 
+parameters are similar and between groups the model parameters are different. This is 
+achieved by finding instabilities in the model parameters with respect to a 
+variable (characteristicum) and recursively partitioning the data into groups. If, for
+example the algorithm finds that men and women have differing treatment effects, the
+data is partitioned into two subgroups. Details on model-based trees in general can 
+be found in \citep{zeileis_model-based_2008} and for the special use case in clinical 
+trial in \citep{seibold_model-based_2016}.
+
+Just a single line of code lets the user compute a model-based tree in R:
 <<eval=FALSE>>=
@@
 be estimated based on a weighted set of the original data, where the similarity measure 
-corresponds to the weight.
+corresponds to the weight. Details on the personalised models can be found in 
+\citep{seibold_individual_2017}.
+
+In R, to get from the initial base model to personalised models for all observations
+in the training data, type:
 <<eval=FALSE>>=
-pers_models <- pmodel(base_model)
+pm_forest <- pmforest(base_model)
+pers_models <- pmodel(pm_forest)
 @
-
-\begin{figure}
-\centering
-\include{treefamily}
-\caption{The family of model-based trees and forests.}
-\end{figure}
 
@@
 
+In the following we will present two example applications for model-based trees
+and personalised models. For this we need to load the package and -- to ensure
+reproducibility -- set a random seed. Also for plots we will need the \emph{ggplot2}
+package and we use a black and white theme.
+<<prep>>=
+library("model4you")
+set.seed(2017)
+
+library("ggplot2")
+theme_set(theme_classic())
+@
+
+\paragraph{Mathematics exam analysis}
+<<math_data>>=
+data("MathExam14W", package = "psychotools")
+
+## scale points achieved to [0, 100] percent
+MathExam14W$tests <- 100 * MathExam14W$tests/26
+MathExam14W$pcorrect <- 100 * MathExam14W$nsolved/13
+
+## select variables to be used
+MathExam <- MathExam14W[ , c("pcorrect", "group", "tests", "study",
+                             "attempt", "semester", "gender")]
+@
+
+<<math_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
+## compute base model
+bmod_math <- lm(pcorrect ~ group, data = MathExam)
+lmplot(bmod_math, densest = TRUE)
+@
+
+<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for GBSG2 data.">>=  
+## compute tree
+(tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2)))
+plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lmplot, 
+                                               confint = TRUE))
+@
+
+<<math_methods>>=  
+## predict
+newdat <- MathExam[1:5, ]
+
+# terminal nodes
+(nodes <- predict(tr_math, type = "node", newdata = newdat))
+
+# response
+(pr <- predict(tr_math, type = "pass", newdata = newdat))
+
+# response including confidence intervals, see ?predict.lm
+(pr1 <- predict(tr_math, type = "pass", newdata = newdat,
+                predict_args = list(interval = "confidence")))
+@
+
+<<math_pmodl>>=
+forest_math <- pmforest(bmod_math)
+pmod_math <- pmodel(forest_math)
+head(pmod_math)
+@
+
+<<math_dp1, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
+dpdat_math <- cbind(pmod_math, MathExam)
+
+ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
+  geom_point(alpha = 0.2, size = 1) + 
+  geom_smooth(fill = NA, method = "loess") +
+  theme(legend.position = "none")
+
+library("ggbeeswarm")
+ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
+  geom_quasirandom(alpha = 0.5, size = 1) +
+  theme(legend.position = "none")
+@
+
+<<math_dp2, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=
+ggplot(dpdat_math, aes(x = attempt, y = group2, color = attempt)) +
+  geom_quasirandom(alpha = 0.5) +
+  theme(legend.position = "none")
+
+ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
+  geom_quasirandom(alpha = 0.5) +
+  theme(legend.position = "none")
+@
+
+\paragraph{Analysis of the German Breast Cancer Study Group 2}
+
 <<GBSG2_prep>>=
-library("model4you")
 library("survival")
 data("GBSG2", package = "TH.data")
-
-set.seed(2017)
 @
@@
 <<GBSG2_forest>>=
-frst <- pmforest(object = bmod, ntree = 100)
+frst <- pmforest(bmod)
 @
@@
 Based on partykit, merely convenience interface.
+
+\begin{figure}
+\centering
+\include{treefamily}
+\caption{The family of model-based trees and forests.}
+\end{figure}
 
@@
 
-\section*{References}
+% \section*{References}
+\bibliographystyle{plain}
+\bibliography{ref_model4you}
 
-\textcolor{blue}{Please enter references in the Harvard style and include a DOI where available, citing them in the text with a number in square brackets, e.g. \\ }
-
-\textcolor{blue}{[1] Piwowar, H A 2011 Who Shares?
+% \textcolor{blue}{Please enter references in the Harvard style and include a DOI where available, citing them in the text with a number in square brackets, e.g. \\ }
+% 
+% \textcolor{blue}{[1] Piwowar, H A 2011 Who Shares?
 Who Doesn't?

### 2017-06-05 pkg/model4you/inst/JORS/man_model4you.Rnw
Just pretty linebreaks
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1454 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \section*{Paper Author Roles and Affiliations}
-1. PhD student; Biostatistics Department, Epidemiology, Biostatistics \& Prevention
-Institute, University of Zurich
+1. PhD student; Biostatistics Department, Epidemiology, Biostatistics \&
+Prevention Institute, University of Zurich
 
-2. Professor; Department of Statistics, Faculty of Economics and Statistics, University of
-Innsbruck
+2. Professor; Department of Statistics, Faculty of Economics and Statistics,
+University of Innsbruck
 
-3. Professor; Biostatistics Department, Epidemiology, Biostatistics \& Prevention
-Institute, University of Zurich
+3. Professor; Biostatistics Department, Epidemiology, Biostatistics \&
+Prevention Institute, University of Zurich
 
 \section*{Abstract}
-Typical models estimating treatment effecs assume that the treatment effect is the same for all 
-individuals observed and 
-for possible future individuals. Model-based recursive partitioning allows to relax this assumption.
-With model-based trees one can compute treatment effects for different strata of individuals. The
-strata are found in a data driven fashion and depend on characteristics of the individuals. 
-Model-based random forests allow for a similarity estimation between individuals in terms of 
-model estimates and through this we can estimate personalised models.
-The R package \emph{model4you} (available on CRAN) implements these stratified and personalised models with a
-focus on ease of use and interpretability so that clinicians and other users can take
-the model they usually use for the estimation of the average treatment 
-effect and with a few lines of code get a visualisation that is easy to understand and interpret.
+Typical models estimating treatment effecs assume that the treatment effect is
+the same for all individuals observed and for possible future individuals.
+Model-based recursive partitioning allows to relax this assumption.  With
+model-based trees one can compute treatment effects for different strata of
+individuals. The strata are found in a data driven fashion and depend on
+characteristics of the individuals.  Model-based random forests allow for a
+similarity estimation between individuals in terms of model estimates and
+through this we can estimate personalised models.  The R package
+\emph{model4you} (available on CRAN) implements these stratified and
+personalised models with a focus on ease of use and interpretability so that
+clinicians and other users can take the model they usually use for the
+estimation of the average treatment effect and with a few lines of code get a
+visualisation that is easy to understand and interpret.
 
@@
 
-personalised medicine; subgroup analysis; model-based recursive partitioning; unbiased trees;
-treatment effect; random forest
+personalised medicine; subgroup analysis; model-based recursive partitioning;
+unbiased trees; treatment effect; random forest
 
@@
 different exposure and then measure a response. For example, in clinical trials
-patients are assigned to one of two treatment groups where usually one treatment
-group receives a new treatment or drug and the other treatment group receives
-the standard of care or a placebo.
+patients are assigned to one of two treatment groups where usually one
+treatment group receives a new treatment or drug and the other treatment group
+receives the standard of care or a placebo.
 
@@
 treatment indicator would be $1$ if the patient receives the new treatment and
-$0$ if the patient receives standard of care.
-In R such a simple model can be estimated as follows:
+$0$ if the patient receives standard of care.  In R such a simple model can be
+estimated as follows:
 <<eval=FALSE>>=
@@
 @
-with \Sexpr{'response'} being the response measured, \Sexpr{'treatment'} being the
-treatment indicator and \Sexpr{'data'} being the data set containing these variables.
-The function \Sexpr{'model()'} can be replaced for example by \Sexpr{'lm()'} to estimate
-a linear model, \Sexpr{'glm()'} to estimate a generalised linear model or 
-\Sexpr{'survreg()'} to estimate a parametric survival model. 
-These models estimate intercept and treatment effect for
-all individuals in the \Sexpr{'data'} and allow for predicting the response 
-of other individuals given they do or don't receive the treatment of interest.
+with \Sexpr{'response'} being the response measured, \Sexpr{'treatment'} being
+the treatment indicator and \Sexpr{'data'} being the data set containing these
+variables.  The function \Sexpr{'model()'} can be replaced for example by
+\Sexpr{'lm()'} to estimate a linear model, \Sexpr{'glm()'} to estimate a
+generalised linear model or \Sexpr{'survreg()'} to estimate a parametric
+survival model.  These models estimate intercept and treatment effect for all
+individuals in the \Sexpr{'data'} and allow for predicting the response of
+other individuals given they do or don't receive the treatment of interest.
 
-If you do not believe that all individuals have the same intercept and treatment
-effect the R package \emph{model4you} gives you two options to loosen up this 
-strict model: 
-\paragraph{(1) Model-based trees} identify subgroups where within the subgroups the model 
-parameters are similar and between groups the model parameters are different. This is 
-achieved by finding instabilities in the model parameters with respect to a 
-variable (characteristicum) and recursively partitioning the data into groups. If, for
-example the algorithm finds that men and women have differing treatment effects, the
-data is partitioned into two subgroups. Details on model-based trees in general can 
-be found in \citep{zeileis_model-based_2008} and for the special use case in clinical 
-trial in \citep{seibold_model-based_2016}.
+If you do not believe that all individuals have the same intercept and
+treatment effect the R package \emph{model4you} gives you two options to loosen
+up this strict model: \paragraph{(1) Model-based trees} identify subgroups
+where within the subgroups the model parameters are similar and between groups
+the model parameters are different. This is achieved by finding instabilities
+in the model parameters with respect to a variable (characteristicum) and
+recursively partitioning the data into groups. If, for example the algorithm
+finds that men and women have differing treatment effects, the data is
+partitioned into two subgroups. Details on model-based trees in general can be
+found in \citep{zeileis_model-based_2008} and for the special use case in
+clinical trial in \citep{seibold_model-based_2016}.
 
@@
 
-\paragraph{(2) Personalised models} use model-based random forests to estimate similarity
-of individuals in terms of model parameters. For each individual a personalised model can
-be estimated based on a weighted set of the original data, where the similarity measure 
-corresponds to the weight. Details on the personalised models can be found in 
-\citep{seibold_individual_2017}.
+\paragraph{(2) Personalised models} use model-based random forests to estimate
+similarity of individuals in terms of model parameters. For each individual a
+personalised model can be estimated based on a weighted set of the original
+data, where the similarity measure corresponds to the weight. Details on the
+personalised models can be found in \citep{seibold_individual_2017}.
 
-In R, to get from the initial base model to personalised models for all observations
-in the training data, type:
+In R, to get from the initial base model to personalised models for all
+observations in the training data, type:
 <<eval=FALSE>>=
@@
 and personalised models. For this we need to load the package and -- to ensure
-reproducibility -- set a random seed. Also for plots we will need the \emph{ggplot2}
-package and we use a black and white theme.
+reproducibility -- set a random seed. Also for plots we will need the
+\emph{ggplot2} package and we use a black and white theme.
 <<prep>>=
@@
 (tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2)))
-plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lmplot, 
+plot(tr_math, terminal_panel = node_pmterminal(tr_math, 
+                                               plotfun = lmplot, 
                                                confint = TRUE))
@@
 
-The software is intentionally written to make usage as simple as possible. We hope that researchers use it for analysis of:
+The software is intentionally written to make usage as simple as possible. We
+hope that researchers use it for analysis of:
 \begin{itemize}

### 2017-06-21 pkg/disttree/inst/useR2017/slides.Rnw
added slides for presentation at the useR2017
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1468 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[11pt,compress,t]{beamer}

\usepackage{Sweave} 
\usepackage{graphicx}
\setbeamertemplate{caption}{\insertcaption} 
\usetheme{Z}
%% includes a replacement for \usepackage{Sweave}
\usepackage{amsmath,tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
\graphicspath{{plots/}}

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}

%% need no \usepackage{Sweave}
\definecolor{InputColor}{rgb}{0,0,0.3}
\definecolor{OutputColor}{rgb}{0.2,0.2,0.2}

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("ggplot2")
theme_set(theme_bw(base_size = 18))
library("disttree")
#library("partykitR1")
library("gamlss")
library("lattice")
library("crch")
library("latex2exp")
library("parallel")
library("gamlss.cens")
gen.cens(NO, type = "left")

## define distribution list:
# dist_list_normal
{

  dist_list_normal <- list()

  parnames <- c("mu", "sigma")
  etanames <- c("mu", "log(sigma)")

  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {

    val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
    if(!log) val <- exp(val)

    # par <- c(eta[1], exp(eta[2]))
    # val <- dnorm(y, mean = par[1], sd = par[2], log = log)

    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y))
      val <- sum(weights * val, na.rm = TRUE)
    }
    return(val)
  }

  sdist <- function(y, eta, weights = NULL, sum = FALSE) {

    score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
                   -1 + exp(-2*eta[2] + log((y-eta[1])^2)))

    # par <- c(eta[1], exp(eta[2])) 
    # score <- cbind(1/par[2]^2 * (y-par[1]), 
    #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))

    score <- as.matrix(score)
    colnames(score) <- etanames
    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y))
      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
      score[score==Inf] = 1.7e308
      score <- colSums(weights * score, na.rm = TRUE)
    }
    return(score)
  }

  hdist <- function(y, eta, weights = NULL) {    
    ny <- length(y)
    if(is.null(weights)) weights <- rep.int(1, ny)

    d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
    d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-eta[1]) * exp(-2*eta[2]), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
    d2ld.etasigma2 <- sum(weights * (-2)*exp(log((y-eta[1])^2) - 2*eta[2]), na.rm = TRUE)

    # par <- c(eta[1], exp(eta[2]))                           
    # d2ld.etamu2 <- sum(weights * rep.int(-1/par[2]^2, ny))
    # d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-par[1])/par[2]^2), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
    # d2ld.etasigma2 <- sum(weights * (-2)*(y-par[1])^2/par[2]^2, na.rm = TRUE)

    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etamu.d.etasigma, d2ld.etasigma2), nrow = 2)
    colnames(hess) <- rownames(hess) <-  etanames

    return(hess)
  }

  ## additional functions pdist, qdist, rdist
  pdist <- pnorm
  qdist <- qnorm
  rdist <- rnorm

  link <- c("identity", "log")

  linkfun <- function(par) {
    eta <- c(par[1], log(par[2]))
    names(eta) <- etanames
    return(eta)
  }

  linkinv <- function(eta) {
    par <- c(eta[1], exp(eta[2]))
    names(par) <- parnames
    return(par)
  }

  linkinvdr <- function(eta) {
    dpardeta <- c(1, exp(eta[2]))
    names(dpardeta) <- parnames
    return(dpardeta)
  }

  startfun <- function(y, weights = NULL){
    if(is.null(weights)) {
      mu <- mean(y)
      sigma <- sqrt(1/length(y) * sum((y - mu)^2))
    } else {
      mu <- weighted.mean(y, weights)
      sigma <- sqrt(1/sum(weights) * sum(weights * (y - mu)^2))
    }
    starteta <- c(mu, log(sigma))
    names(starteta) <- etanames
    return(starteta)
  }

  mle <- TRUE

  dist_list_normal <- list(family.name = "Normal Distribution",
                           ddist = ddist, 
                           sdist = sdist, 
                           hdist = hdist,
                           pdist = pdist,
                           qdist = qdist,
                           rdist = rdist,
                           link = link, 
                           linkfun = linkfun, 
                           linkinv = linkinv, 
                           linkinvdr = linkinvdr,
                           startfun = startfun,
                           mle = mle
  )
}

# dist_list_cens_normal
{

  dist_list_cens_normal <- list()

  parnames <- c("mu", "sigma")
  etanames <- c("mu", "log(sigma)")

  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
    par <- c(eta[1], exp(eta[2]))
    val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
    if(sum) {
      if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
      val <- sum(weights * val, na.rm = TRUE)
    }
    return(val)
  }

  sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
    par <- c(eta[1], exp(eta[2]))
    # y[y==0] <- 1e-323

    score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
    score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
    score <- cbind(score_m, score_s)
    score <- as.matrix(score)
    colnames(score) <- etanames
    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y)[1])
      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
      score[score==Inf] = 1.7e308
      score <- colSums(weights * score, na.rm = TRUE)
      #if(any(is.nan(score))) print(c(eta, "y", y))
    }
    return(score)
  }

  hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
    ny <- length(y)
    if(is.null(weights)) weights <- rep.int(1, ny)

    par <- c(eta[1], exp(eta[2]))                           
    # y[y==0] <- 1e-323

    d2mu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
    d2sigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
    dmudsigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu.sigma", left = left, right = right) # FIX: order?
    dsigmadmu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma.mu", left = left, right = right) # FIX: order?
    dsigma <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)

    d2ld.etamu2 <- sum(weights * d2mu, na.rm = TRUE)
    d2ld.etamu.d.etasigma <- sum(weights * dmudsigma * par[2], na.rm = TRUE)
    d2ld.etasigma.d.etamu <- sum(weights * dsigmadmu * par[2], na.rm = TRUE)
    d2ld.etasigma2 <- sum(weights * (d2sigma * exp(2*eta[2]) + dsigma * par[2]), na.rm = TRUE)

    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etasigma.d.etamu, d2ld.etasigma2), nrow = 2)
    colnames(hess) <- rownames(hess) <-  etanames

    return(hess)
  }

  ## additional functions pdist, qdist, rdist
  pdist <- function(q, eta, lower.tail = TRUE, log.p = FALSE) crch:::pcnorm(q, mean = eta[1], sd = eta[2], 
                                                                            lower.tail = lower.tail, log.p = log.p, 
                                                                            left = left, right = right)
  qdist <- function(p, eta, lower.tail = TRUE, log.p = FALSE) crch:::qcnorm(p, mean = eta[1], sd = eta[2], 
                                                                            lower.tail = lower.tail, log.p = log.p, 
                                                                            left = left, right = right)
  rdist <- function(n, eta) crch:::rcnorm(n, mean = eta[1], sd = eta[2], left = left, right = right)

  link <- c("identity", "log")

  linkfun <- function(par) {
    eta <- c(par[1], log(par[2]))
    names(eta) <- etanames
    return(eta)
  }

  linkinv <- function(eta) {
    par <- c(eta[1], exp(eta[2]))
    names(par) <- parnames
    return(par)
  }

  linkinvdr <- function(eta) {
    dpardeta <- c(1, exp(eta[2]))
    names(dpardeta) <- parnames
    return(dpardeta)
  }

  startfun <- function(y, weights = NULL){
    yc <- pmax(0,y)  # optional ?
    if(is.null(weights)) {
      mu <- mean(yc)
      sigma <- sqrt(1/length(yc) * sum((yc - mu)^2))
    } else {
      mu <- weighted.mean(yc, weights)
      sigma <- sqrt(1/sum(weights) * sum(weights * (yc - mu)^2))
    }
    starteta <- c(mu, log(sigma))
    names(starteta) <- etanames
    return(starteta)
  }

  mle <- FALSE

  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
                                ddist = ddist, 
                                sdist = sdist, 
                                hdist = hdist,
                                pdist = pdist,
                                qdist = qdist,
                                rdist = rdist,
                                link = link, 
                                linkfun = linkfun, 
                                linkinv = linkinv, 
                                linkinvdr = linkinvdr,
                                startfun = startfun,
                                mle = mle
  )
}

######## data generating process based on a predefined tree or a parameter function
# if tree: 2 splits => 3 terminal nodes
# 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
#
# input: n            ..... nr of observations
#        family       ..... distribution of the generated observations
#        split.matrix ..... indices of the variables used for the splits together with the corresponding split points
#        parm.matrix  ..... set of distribution parameters for each subgroup
#        fun          ..... parameter function
#
# output: data.frame with generated observations y (generated seperatly in each subgroups with given distributions parameters),
#                         the given split variables x1, ..., x10 for each observation and
#                         the index of the subgroup for each observation or
#                         the distribution parameter for each observation

# FIX: until now only complete lists can be handed over (dist_list...)
dgp <- function(n, family = dist_list_normal, 
                fun = NULL,
                split.matrix = matrix(nrow = 2, ncol = 2), 
                par.matrix = matrix(nrow = 3, ncol = 2), 
                round.sp = 3)
{

  # generating the possible split variables
  x1 <- runif(n,-0.4,1)
  x2 <- runif(n,-10,10)
  x3 <- runif(n,0,100)
  x4 <- x1 + rnorm(n, sd = 0.1)
  x5 <- x1 + rnorm(n, sd = 0.3)
  x6 <- rbinom(n,1,0.5)
  x7 <- rbinom(n,1,0.5)
  x8 <- sample(1:4, n, replace = TRUE) 
  x9 <- sample(1:5, n, replace = TRUE) 
  x10 <- sample(1:7, n, replace = TRUE)
  #x11 <- runif(n,-0.5,1)
  x <- cbind(x1,x2,x3,x4,x5,x6,x7,x8,x9,x10)
  # reduce nr of possible split points by rounding values of split variables
  x <- round(x, digits = round.sp)

  y <- vector(mode = "numeric", length = n)

  # getting the random function
  if(is.function(family)) family <- family()
  if(inherits(family, "gamlss.family")) {
    rfun <- get(paste0("r",family$family[[1]]))
  } else {
    if(family$family.name == "Normal Distribution") rfun <- rNO
    if(family$family.name == "censored Normal Distribution") rfun <- rNO
    if(family$family.name == "Poisson Distribution") rfun <- rPO
    if(family$family.name == "Weibull Distribution") rfun <- survival:::rsurvreg 
  }

  if(is.null(fun)) {
    index <- vector(mode = "numeric", length = n)
    sv = split.matrix[,1]    # index of split variables
    sp = split.matrix[,2]    # split points
    par = par.matrix

    # splitting based on the given split variables and split points and
    # generating the observations seperatly for each subgroup (with the corresponding given distribution parameters)
    for(k in 1:n){
      if(x[k, sv[1]] <= sp[1]){
        y[k] <- do.call(rfun, as.list(c(1,par[1,])))
        index[k] <- 2L
      } else {
        if(x[k, sv[2]] <= sp[2]){
          y[k] <- do.call(rfun, as.list(c(1,par[2,]))) 
          index[k] <- 4L
        } else {
          y[k] <- do.call(rfun, as.list(c(1,par[3,])))
          index[k] <- 5L
        }
      }
    }
    d <- as.data.frame(cbind(y, x, index))
    colnames(d) <- c("y", paste0("x", c(1:10)), "index")

  } else {

    dpar <- fun(x)

    if(!(family$family.name == "Poisson Distribution")){
      y <- rfun(n, dpar[,1], dpar[,2])
      d <- as.data.frame(cbind(y, x, dpar))
      colnames(d) <- c("y", paste0("x", c(1:10)), "mu", "sigma")
    } else {
      y <- rfun(n, dpar)
      d <- as.data.frame(cbind(y, x, dpar))
      colnames(d) <- c("y", paste0("x", c(1:10)), "lambda")
    }
  }

  if(family$family.name == "censored Normal Distribution") {
    #d$ystar <- d$y
    d$y <- pmax(d$y, 0)
  }

  return(d)
}

############################################ 
# parameter functions

### plotting function for one data set
plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE) 
{
  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
                    df$fitted.par,
                    dt$fitted.par,
                    #coef(dt)[paste(dt$fitted[,1]),],
                    g$mu.fv, g$sigma.fv)

  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
                          "fitted.mu.df","fitted.sigma.df",
                          "fitted.mu.dt","fitted.sigma.dt",
                          "fitted.mu.g","fitted.sigma.g")
  sp <- plotdata[order(plotdata["x1"]),]

  if(!(any(dt.par, df.par, g.par, all.par))){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "true parameters", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 2, col = 'black')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 2, col = 'black')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
           col = c('black','grey'), lty = 1, bty = 'n')
  }

  # disttree
  if(dt.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = 'green4')
    lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
           col = c('black','grey','green4', 'greenyellow'), lty = 1, bty = 'n')
  }

  # distforest
  if(df.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = 'red')
    lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
    lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
           col = c('black','grey','red', 'orange'), lty = 1, bty = 'n')
  }

  # gamlss
  if(g.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = 'blue')
    lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
    lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
           col = c('black','grey','blue', 'lightblue'), lty = 1, bty = 'n')
  }

  # disttree vs distforest vs gamlss
  if(all.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = 'green4')
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = 'red')
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = 'blue')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
                        TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
           col = c('black','green4','red','blue'), lty = 1, bty = 'n')
  }
}

### plotting function for one data set and sigma only
plot1sigma <- function(learndata, dt=NULL, df=NULL, g=NULL) 
{
  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
                    df$fitted.par,
                    dt$fitted.par,
                    #coef(dt)[paste(dt$fitted[,1]),],
                    g$mu.fv, g$sigma.fv)

  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
                          "fitted.mu.df","fitted.sigma.df",
                          "fitted.mu.dt","fitted.sigma.dt",
                          "fitted.mu.g","fitted.sigma.g")
  sp <- plotdata[order(plotdata["x1"]),]

  plot(y = sp$y-sp$true.mu, x = sp$x, type = "p", col="grey", xlab = "x", ylab = "y",  main = "disttree", xaxt='n', yaxt='n', xlab = "", ylab = "")
  lines(x = sp$x, y = sp$true.sigma, type = "l", col = 'black')
  lines(x = sp$x, y = -sp$true.sigma, type = "l", col = 'black')
  lines(x = sp$x, y = sp$fitted.sigma.dt, type = "l", col = 'green4')
  lines(x = sp$x, y = -sp$fitted.sigma.dt, type = "l", col = 'green4')
  lines(x = sp$x, y = sp$fitted.sigma.df, type = "l", col = 'red')
  lines(x = sp$x, y = -sp$fitted.sigma.df, type = "l", col = 'red')
  lines(x = sp$x, y = sp$fitted.sigma.g, type = "l", col = 'blue')
  lines(x = sp$x, y = -sp$fitted.sigma.g, type = "l", col = 'blue')
  legend('topleft', c(TeX('$\\sigma$'), TeX('$\\hat{\\sigma}$ disttree'),
                      TeX('$\\hat{\\sigma}$ distforest'), TeX('$\\hat{\\sigma}$ gamlss')), 
         col = c('black','green4','red','blue'), lty = 1)
}

@

%\bibliography{ref.bib}

\begin{document}
\SweaveOpts{concordance=TRUE}

\title{Distributional Trees and Forests}
\author{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
\URL{https://R-Forge.R-project.org/projects/partykit/}

\section{Distributional Trees and Forests}

\subsection{Motivation}

\begin{frame}[fragile]
\frametitle{Motivation}
\vspace{0.5cm}
\begin{figure}[!htb]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<1->{
<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
lines(x = x, y = ytrue, col = "red", lwd = 1.5, main = "")
@
}
\end{center}
\endminipage
\visible<2->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<2->{
<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
lines(x = x, y = ytrue, col = "red", lwd = 1.5, main = "")
@
}
\end{center}
\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\visible<3->{
<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
lines(x = x, y = ytrue, col = "red", lwd = 1.5)
lines(x = x, y = ytrue+0.1+var, col = "red", lty = 2)
lines(x = x, y = ytrue-0.1-var, col = "red", lty = 2)
@
}
\end{center}
\endminipage

\vspace{0.5cm}
\minipage{0.25\textwidth}
\begin{center}
\visible<1->{
LM, GLM\\
\vspace{0.5cm}
\code{lm}\\
\code{glm}\\
\vspace{1.5cm}}
\end{center}
\endminipage
\hspace{1.1cm}
\minipage{0.25\textwidth}
\begin{center}
\visible<2->{
GAM\\
\vspace{0.5cm}
\code{mgcv}\\
\code{VGAM}\\
$\ldots$\\
\vspace{1cm}}
\end{center}
\endminipage
\hspace{1.1cm}
\minipage{0.25\textwidth}
\begin{center}
\visible<3->{
GAMLSS\\
\vspace{0.5cm}
\code{gamlss}\\
\code{mgcv}\\
\code{VGAM}\\
\code{gamboostLSS}\\
$\ldots$}
\end{center}
\endminipage
\end{figure}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
\vspace{0.1cm}
\begin{figure}[!htb]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = ytree, col = "red", lwd = 1.5)
@
\visible<1->{
<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = ytree, col = "red", lwd = 1.5)
@
}
\end{center}

\endminipage
\visible<2->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
 \begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = yforest, col = "red", lwd = 1.5, main = "")
@
\visible<2->{
<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = ytree, col = "red", lwd = 1.5)
@
}
\end{center}

\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
  \visible<3->{
  \begin{center}
  \begin{tikzpicture}
  \draw[line width=0.01mm, black] (0,0) rectangle (2.6,2.24);
  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
  \end{tikzpicture}
  \end{center}}
\endminipage

\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
\visible<1->{
Regression Tree\\
\vspace{0.5cm}
\resizebox{0.3\textwidth}{!}{
\begin{tikzpicture}
  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (1, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0.5, 1) {};
  \draw[-, line width=1pt] (n0) -- (n1);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (1.5, 1) {};
  \draw[-, line width=1pt] (n0) -- (n2);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (1, 0) {};
  \draw[-, line width=1pt] (n2) -- (n3);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (2, 0) {};
  \draw[-, line width=1pt] (n2) -- (n4);
\end{tikzpicture}}
\vspace{0.3cm}\\
\code{rpart}\\
\code{party(kit)}\\
\vspace{1cm}}
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
\visible<2->{
Random Forest\\
\vspace{0.5cm}
\resizebox{0.9\textwidth}{!}{
\begin{tikzpicture}
  \node[ellipse, fill=HighlightBlue!70, align=center] (n00) at (1, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n01) at (0.5, 1) {};
  \draw[-, line width=1pt] (n00) -- (n01);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n02) at (1.5, 1) {};
  \draw[-, line width=1pt] (n00) -- (n02);

  \node[ellipse, fill=HighlightBlue!70, align=center] (n10) at (3, 2) {};
  \node[ellipse, fill=HighlightBlue!70, align=center] (n11) at (2.5, 1) {};
  \draw[-, line width=1pt] (n10) -- (n11);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n12) at (3.5, 1) {};
  \draw[-, line width=1pt] (n10) -- (n12);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n13) at (2, 0) {};
  \draw[-, line width=1pt] (n11) -- (n13);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n14) at (2.8, 0) {};
  \draw[-, line width=1pt] (n11) -- (n14);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n15) at (3.2, 0) {};
  \draw[-, line width=1pt] (n12) -- (n15);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n16) at (4, 0) {};
  \draw[-, line width=1pt] (n12) -- (n16);

  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n21);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n22);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
  \draw[-, line width=1pt] (n22) -- (n23);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
  \draw[-, line width=1pt] (n22) -- (n24);
\end{tikzpicture}
}
\vspace{0.5cm}\\
\code{randomForest}\\
\code{ranger}\\
\code{party(kit)}\\
$\ldots$}
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.68cm}
\visible<3->{
Distributional Trees and Forests\\
\vspace{1.7cm}
\code{disttree}\\
based on \code{partykit}\\
\vspace{1cm}}
\end{center}
\endminipage
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Goals}
\vspace{1cm}
{\bf Tree:}
\begin{itemize}
  \item Specify the complete distribution in each subgroup.\\
  (location, scale and shape)
  \item Automatic detection of steps and abrupt changes. %(data driven)
  \item Capture non-linear and non-additive effects and interactions.
\end{itemize}
\vspace{0.5cm}
{\bf Forest:}
\begin{itemize}
  \item Smoother effects.
  \item Stabelization of the model.
\end{itemize}
\end{frame}

\subsection{Building Distributional Trees and Forests}
\begin{frame}
\frametitle{Building Distributional Trees and Forests}
{\bf Tree:}
\begin{enumerate}
\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
\item Estimate $\hat{\theta}$ via maximum likelihood.
\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
%\item Assess whether the \emph{model scores} are associated with
%    (or change along) any of the available covariates -- e.g.,
%    using parameter instability tests (\emph{strucchange}) or
%    conditional inference (\emph{coin}).
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose breakpoint with highest improvement in log-likelihood.
\item Repeat steps 2--4 recursively in the subgroups
    until some stopping criterion is met.
    %-- e.g., for significance or sample size.
%\item Choose the variable with the strongest association
%\item Choose the split point which leads to the highest improvement of the model
%\item Split and repeat 2-6 in each node until a stopping criterion is met
\end{enumerate}

\vspace{0.4cm}
{\bf Forest:} Ensemble of trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Prediction}
\vspace{1cm}
{\bf Tree:}\\
\vspace{0.1cm}
\hspace{0.7cm} Estimate $\hat{\theta}$ on the subsample of the learning data which \\
\hspace{0.7cm} ends up in the same terminal node as the new observation.
%\item Evaluate final node for new observation
%\item Return corresponding distribution (parameters)

\vspace{0.5cm}

{\bf Forest:}\\
\vspace{0.1cm}
\hspace{0.7cm} Estimate $\hat{\theta}$ on the whole learning data but weighted by \\
\hspace{0.7cm} the number of trees in which a learning observation ends up\\
\hspace{0.7cm} in the same terminal node as the new observation.
% but weighted by similarities of the learning data and the new observation over all trees
%\item Evaluate final node for new observation in each of the trees and investigate similarities with each of the learning observations
%\item Fit a distributional model for the new observation with different influences of the learning observations defined by weights
\end{frame}

\begin{frame}
\frametitle{Fitting a Tree}
\vspace*{-0.4cm}
\begin{center} 
DGP:  $Y \sim \mathcal{N}(\mu(X), \sigma(X))$
<<dgp_tree, echo=FALSE, results=hide>>=
data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
names(data) <- c("","","")
fig <- party(
  partynode(1L,
            split = partysplit(2L, breaks = 0.4),
            kids = list(
              partynode(2L, info = c(
                "n = 200",
                "True parameters:",
                expression(mu == '4'),
                expression(sigma == '1')
              )),
              partynode(3L,
                        split = partysplit(3L, breaks = 0.8),
                        kids = list(
                          partynode(4L, info = c(
                            "n = 200",
                            "True parameters:",
                            expression(mu == '12'),
                            expression(sigma == '3')
                          )),
                          partynode(5L, info = c(
                            "n = 100",
                            "True parameters:",
                            expression(mu == '4'),
                            expression(sigma == '3')
                          )))))),
  data
)
@

\setkeys{Gin}{width=0.75\linewidth}
<<plottree_dgp, fig=TRUE, echo=FALSE>>=
pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
plot(fig,
     tp_args = list(FUN = identity, width = 18.7, fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)], pval = FALSE),
     drop_terminal = TRUE, tnex = 1.5)
@
\end{center}
\end{frame}

<<tree, echo=FALSE, results=hide>>=
set.seed(7)
nobs <- 500
x <- runif(nobs, 0, 1)
mu <- sigma <- ytrue <- numeric(length = nobs)
for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
y <- rnorm(nobs, mean = mu, sd = sigma)
#y <- rcnorm(nobs, mean = mu, sd = sigma, left = 0)
ytrue <- mu
data <- data.frame(cbind(y,x, ytrue))
tree <- disttree(y~x, data = data, family = NO(), type.tree = "mob")
#tree <- disttree(y~x, data = data, family = dist_list_cens_normal)

#FUN <- function(x) {
#                cf <- x$coefficients
#                cf <- matrix(cf, ncol = 1, dimames = list(c(expression(mu), expression(sigma)), ""))
#                c(sprintf("n = %s", x$nobs), "Estimated parameters:", 
#                  strwrap(capture.output(print(cf, digits = 4L))[-1L]))
#}

#FUN <- function(x) {
#                cf <- x$coefficients
#                cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), 
#                  ""))
#                c(sprintf("n = %s", x$nobs), "Estimated parameters:", 
#                  strwrap(capture.output(print(cf, digits = 4L))[-1L]))
#            }
#FUN <- function (x) 
#{
#  cf <- x$coefficients
#  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
#  c(sprintf("n = %s", x$nobs), "Estimated parameters:", c(c(expression(mu), round(cf[1],3)), 
#                                                          c(expression(sigma), round(cf[2],3))))
#}
@

\begin{frame}[fragile]
\frametitle{Fitting a Tree}
\begin{center}
\vspace*{-0.4cm}
Model: \code{disttree(y~x)}\\
\setkeys{Gin}{width=0.75\linewidth}
<<plottree_estpar, fig=TRUE, echo=FALSE>>=
pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
plot(tree, drop = TRUE, tnex = 1.5, #FUN = FUN,
     tp_args = list(fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)]))
@
\end{center}
\end{frame}

%\begin{frame}
%\frametitle{Fitting a Tree}
%\center
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{tree_estpar}
%\end{frame}

\begin{frame}[fragile]
\frametitle{Fitting a Tree}
\begin{center}
\vspace*{-0.4cm}
Model: \code{disttree(y~x)}\\
\setkeys{Gin}{width=0.75\linewidth}
<<plottree_box, fig=TRUE, echo=FALSE>>=
plot(as.constparty(tree), tnex = 1.5,
     tp_args = list(fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)]))
@
\end{center}
\end{frame}

%\begin{frame}
%\frametitle{Fitting a Tree}
%\center
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{tree_box}
%\end{frame}

<<prelimminaries2, echo=FALSE, results=hide>>=
set.seed(54)
nobs <- 500
x <- runif(nobs, 0, 1)
mu <- sigma <- ytrue <- numeric(length = nobs)
for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
y <- rnorm(nobs, mean = mu, sd = sigma)
ytrue <- mu
data <- data.frame(cbind(y,x, ytrue))

alldata <- cbind(data, mu, sigma)
odata <- alldata[order(alldata["x"]),]
@

\begin{frame}[fragile]
\frametitle{Fitting a Tree}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<plottree_xyplot, fig=TRUE, echo=FALSE>>=
par(mar=c(5.1,4.1,4.1,3.1))
plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "darkgrey")
#plot(y=odata$y, x=odata$x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
lines(x = odata$x, y = odata$mu, col = "red", lwd = 1.5, main = "")
lines(x = odata$x, y = odata$mu + odata$sigma, col = "red", lwd = 1.5, lty = 2, main = "")
lines(x = odata$x, y = odata$mu - odata$sigma, col = "red", lwd = 1.5, lty = 2, main = "")
@
\end{center}
\end{frame}

%\begin{frame}
%\frametitle{Fitting a Tree}
%\center
%\vspace*{-0.7cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{tree_xyplot}
%\end{frame}

%\SweaveOpts{eval=FALSE}
<<echo = FALSE, results=hide>>=
set.seed(723)
fun <- function(x){cbind(10*exp(-((4*x[,1]-2)^4)), 0.5 + 2*abs(x[,1]))}
learndata <- dgp(300, family = dist_list_normal, round.sp = 4, fun = fun)
control <-  ctree_control(teststat = "quad", testtype = "Bonferroni", mincriterion = 0.7, minbucket = 20L)
dt <- disttree(y~x1, data=learndata, family=dist_list_normal, type.tree = "ctree", control = control)
control <-  ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 7L)
df <- distforest(y~x1, data=learndata, family=dist_list_normal, ntree = 500L, type.tree = "ctree", control = control)
g <- gamlss(y ~ pb(x1), sigma.formula = ~pb(x1), data = learndata, trace = FALSE)
@

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<oneforest_exp1, fig=TRUE, echo=FALSE>>=
plot1(learndata, dt=dt, df=df, g=g)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<oneforest_exp1_dt, fig=TRUE, echo=FALSE>>=
plot1(learndata, dt=dt, df=df, g=g, dt.par = TRUE)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<oneforest_exp1_g, fig=TRUE, echo=FALSE>>=
plot1(learndata, dt=dt, df=df, g=g, g.par = TRUE)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<oneforest_exp1_df, fig=TRUE, echo=FALSE>>=
plot1(learndata, dt=dt, df=df, g=g, df.par = TRUE)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<oneforest_exp1_all, fig=TRUE, echo=FALSE>>=
plot1(learndata, dt=dt, df=df, g=g, all.par = TRUE)
@
\end{center}
\end{frame}

%\begin{frame}[fragile]
%\frametitle{Simulation}
%\begin{center}
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%<<oneforest_exp1_sigma, fig=TRUE, echo=FALSE>>=
%plot1sigma(learndata,dt, df, g)
%@
%\end{center}
%\end{frame}

<<preliminaries3, echo=FALSE, results=hide>>=
load("~/svn/partykit/pkg/disttree/presentation/simlist20170621.rda")

# plots for various data sets
# results given in simlist
# plot RMSE
plot_rmse <- function(simlist, type = c("exp", "par")){

  if(type == "exp"){
    #dev.off() 
    rmse <- cbind(simlist$av.rmse.exp.true.dt, simlist$av.rmse.exp.true.df, simlist$av.rmse.exp.true.g,
                  simlist$av.rmse.exp.obs.dt, simlist$av.rmse.exp.obs.df, simlist$av.rmse.exp.obs.g)
    colnames(rmse) <- c("dt.true", "df.true", "g.true", "dt.obs", "df.obs", "g.obs")
    ylim <- c(min(rmse), max(rmse))
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = "forestgreen", ylim = ylim,
         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
    lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = 'forestgreen')
    lines(x = simlist$x.axis, y = rmse[,"df.true"], type = "l", col = 'red')
    lines(x = simlist$x.axis, y = rmse[,"df.obs"], type = "l", lty = 2, col = 'red')
    lines(x = simlist$x.axis, y = rmse[,"g.true"], type = "l", col = 'blue')
    lines(x = simlist$x.axis, y = rmse[,"g.obs"], type = "l", lty = 2, col = 'blue')
    #legend('left', c("disttree", "distforest", "gamlss"), col = c('forestgreen', 'red','blue'), lty = 1)
    mtext(text= "smooth", side = 1, line = 2, adj = 0)
    mtext(text= "steep", side = 1, line = 2, adj = 1)
    mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
    mtext(text= "RMSE", side = 2, line = 1)
    mtext(text= "disttree", side = 2, col = "darkgreen", las = 1, line = 0.2, padj = 6.5)
    mtext(text= "distforest", side = 2, col = "red", las = 1, line = 0.2, padj = 13)
    mtext(text= "gamlss", side = 2, col = "blue", las = 1, line = 0.3, padj = 16)

    # add arrows
    parsave <- par(new = TRUE,  mar = c(0,0,0,0))
    plot(0,0,xlim=c(0,1),ylim=c(0,1),type='n',xlab='', ylab='', col='white', axes = FALSE)

    #add arrow on the right side (downwards, for RMSE)
    arrows(0.99,0.47, 0.99,0.3,lwd=1) ## add arrow
    text(0.99,0.5, 'better')
    segments(0.99,0.53,0.99,0.7,lwd=1)

    #add arrows below
    arrows(0.45,0.05,0.3,0.05,lwd=1) ## add arrow
    arrows(0.6,0.05,0.75,0.05,lwd=1) ## add arrow

    # plot parameters  reset to prior values
    par(parsave)
  }

  if(type == "par"){
    rmse <- cbind(simlist$av.rmse.mu.dt, simlist$av.rmse.mu.df, simlist$av.rmse.mu.g, 
                  simlist$av.rmse.sigma.dt, simlist$av.rmse.sigma.df, simlist$av.rmse.sigma.g)
    colnames(rmse) <- c("mu.dt", "mu.df", "mu.g", "sigma.dt", "sigma.df", "sigma.g")
    ylim <- c(min(rmse), max(rmse))
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = "forestgreen", ylim = ylim,
         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
    lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = 'forestgreen')
    lines(x = simlist$x.axis, y = rmse[,"mu.df"], type = "l", col = 'red')
    lines(x = simlist$x.axis, y = rmse[,"sigma.df"], type = "l", lty = 2, col = 'red')
    lines(x = simlist$x.axis, y = rmse[,"mu.g"], type = "l", col = 'blue')
    lines(x = simlist$x.axis, y = rmse[,"sigma.g"], type = "l", lty = 2, col = 'blue')
    #legend('topleft', c("disttree", "distforest", "gamlss"), col = c('forestgreen', 'red','blue'), lty = 1)
    mtext(text= "smooth", side = 1, line = 2, adj = 0)
    mtext(text= "steep", side = 1, line = 2, adj = 1)
    mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
    mtext(text= "RMSE", side = 2, line = 1)
    mtext(text= "disttree", side = 2, col = "darkgreen", las = 1, line = 0.2, padj = -4)
    mtext(text= "distforest", side = 2, col = "red", las = 1, line = 0.2, padj = 8)
    mtext(text= "gamlss", side = 2, col = "blue", las = 1, line = 0.3, padj = 14)

    # add arrows
    parsave <- par(new = TRUE,  mar = c(0,0,0,0))
    plot(0,0,xlim=c(0,1),ylim=c(0,1),type='n',xlab='',ylab='',col='white', axes = FALSE)

    #add arrow on the right side (downwards, for RMSE)
    arrows(0.99,0.47,0.99,0.3,lwd=1) ## add arrow
    text(0.99,0.5, 'better')
    segments(0.99,0.53,0.99,0.7,lwd=1)

    #add arrows below
    arrows(0.45,0.05,0.3,0.05,lwd=1) ## add arrow
    arrows(0.6,0.05,0.75,0.05,lwd=1) ## add arrow

    # plot parameters  reset to prior values
    par(parsave)
  }
}

# plot loglikelihood
plot_ll <- function(simlist){
  ll <- cbind(simlist$av.loglik.dt, simlist$av.loglik.df, simlist$av.loglik.g)
  colnames(ll) <- c("dt", "df","g")
  ylim <- c(min(ll), max(ll))
  par(mar=c(5.1,4.1,4.1,3.1))
  plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = "forestgreen", ylim = ylim,
       xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
  lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = 'red')
  lines(x = simlist$x.axis, y = ll[,"g"], type = "l", col = 'blue')
  #legend('topright', c("disttree", "distforest", "gamlss"), col = c('forestgreen', 'red','blue'), lty = 1)
  mtext(text= "smooth", side = 1, line = 2, adj = 0)
  mtext(text= "steep", side = 1, line = 2, adj = 1)
  mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
  mtext(text= "Log-Likelihood", side = 2, line = 1)
  mtext(text= "disttree", side = 2, col = "darkgreen", las = 1, line = 0.2, padj = 12)
  mtext(text= "distforest", side = 2, col = "red", las = 1, line = 0.2, padj = -8.5)
  mtext(text= "gamlss", side = 2, col = "blue", las = 1, line = 0.3, padj = -16)

  # add arrows
  parsave <- par(new = TRUE,  mar = c(0,0,0,0))
  plot(0,0,xlim=c(0,1),ylim=c(0,1),type='n',xlab='',ylab='',col='white', axes = FALSE)

  #add arrow on the right side (upwards, for LL)
  arrows(0.99,0.53,0.99,0.7,lwd=1) ## add arrow
  text(0.99,0.5, 'better')
  segments(0.99,0.3,0.99,0.47,lwd=1)

  #add arrows below
  arrows(0.45,0.05,0.25,0.05,lwd=1) ## add arrow
  arrows(0.6,0.05,0.8,0.05,lwd=1) ## add arrow

  # plot parameters  reset to prior values
  par(parsave)
}
@

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<plotsim_ll, fig=TRUE, echo=FALSE>>=
plot_ll(simlist)
@
\end{center}
\end{frame}

%\begin{frame}[fragile]
%\frametitle{Simulation}
%\begin{center}
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{sim_ll}
%\end{center}
%\end{frame}

%\begin{frame}[fragile]
%\frametitle{Simulation}
%\begin{center}
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{sim_RMSE_par}
%\end{center}
%\end{frame}

%\begin{frame}[fragile]
%\frametitle{Simulation}
%\begin{center}
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{sim_RMSE_exp}
%\end{center}
%\end{frame}

\begin{frame}
\frametitle{Software}
\vspace{0.4cm}
R-package \pkg{disttree} available on R-Forge:\\
\vspace{0.2cm}
\url{https://R-Forge.R-project.org/projects/partykit/}
\vspace{1cm}
Main functions:
\vspace{0.2cm}
\begin{itemize}
\item \code{distfit()}
\item \code{disttree()}
\item \code{distforest()}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{References}
\nocite{*}
%\nocite{hothorn2006unbiased, zeileis2008model, hothorn2015partykit, stasinopoulos2007generalized, seibold2016model}
{\footnotesize
\bibliography{ref.bib}}
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
plot_rmse(simlist, type = "par")
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=
plot_rmse(simlist, type = "exp")
@
\end{center}
\end{frame}

\end{document}

### 2017-06-21 pkg/disttree/inst/useR2017/slides.Rnw
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1470 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\usepackage{Sweave} 
-\usepackage{graphicx}
 \setbeamertemplate{caption}{\insertcaption} 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
 
@@
 \frametitle{Goals}
-\vspace{1cm}
+
+\textbf{Distributional:} Specify the complete probability distribution
+(including location, scale and shape).
+
+\medskip
+
 {\bf Tree:}
 \begin{itemize}
-  \item Specify the complete distribution in each subgroup.\\
-  (location, scale and shape)
   \item Automatic detection of steps and abrupt changes. %(data driven)
@@
 \end{itemize}
-\vspace{0.5cm}
+
+\medskip
+
 {\bf Forest:}
@@
   \item Smoother effects.
-  \item Stabelization of the model.
+  \item Stabilization and regularization of the model.
 \end{itemize}
@@
 data <- data.frame(cbind(y,x, ytrue))
-tree <- disttree(y~x, data = data, family = NO(), type.tree = "mob")
-#tree <- disttree(y~x, data = data, family = dist_list_cens_normal)
+tree <- disttree(y ~ x, data = data, family = NO(), type.tree = "mob")
+#tree <- disttree(y ~ x, data = data, family = dist_list_cens_normal)
 
@@
 \vspace*{-0.4cm}
-Model: \code{disttree(y~x)}\\
+Model: \code{disttree(y ~ x)}\\
 \setkeys{Gin}{width=0.75\linewidth}
@@
 <<preliminaries3, echo=FALSE, results=hide>>=
-load("~/svn/partykit/pkg/disttree/presentation/simlist20170621.rda")
+load("simlist20170621.rda")
 

### 2017-06-21 pkg/disttree/inst/useR2017/slides.Rnw
further touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1471 255f4249-2263-4417-b1ea-8498779e27a7

@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "true parameters", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", xaxt='n', yaxt='n', xlab = "", ylab = "")
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')

### 2017-06-22 pkg/disttree/inst/useR2017/slides.Rnw
restructuring of tree plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1474 255f4249-2263-4417-b1ea-8498779e27a7

@@
 gen.cens(NO, type = "left")
+
+coltree <- hcl(c(128,100,30))
+colforest <- hcl(c(10,100,30))
+colgamlss <- hcl(c(260,100,30))
 
@@
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = 'green4')
+    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = coltree)
     lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
@@
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey','green4', 'greenyellow'), lty = 1, bty = 'n')
+           col = c('black','grey',coltree, 'greenyellow'), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = 'red')
+    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = colforest)
     lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
@@
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey','red', 'orange'), lty = 1, bty = 'n')
+           col = c('black','grey',colforest, 'orange'), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = 'blue')
+    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = colgamlss)
     lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
@@
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
-           col = c('black','grey','blue', 'lightblue'), lty = 1, bty = 'n')
+           col = c('black','grey',colgamlss, 'lightblue'), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = 'green4')
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = 'red')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = 'blue')
+    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = coltree)
+    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = colforest)
+    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = colgamlss)
     legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
                         TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
-           col = c('black','green4','red','blue'), lty = 1, bty = 'n')
-  }
-}
-
-### plotting function for one data set and sigma only
-plot1sigma <- function(learndata, dt=NULL, df=NULL, g=NULL) 
-{
-  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
-                    df$fitted.par,
-                    dt$fitted.par,
-                    #coef(dt)[paste(dt$fitted[,1]),],
-                    g$mu.fv, g$sigma.fv)
-
-  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
-                          "fitted.mu.df","fitted.sigma.df",
-                          "fitted.mu.dt","fitted.sigma.dt",
-                          "fitted.mu.g","fitted.sigma.g")
-  sp <- plotdata[order(plotdata["x1"]),]
-
-  plot(y = sp$y-sp$true.mu, x = sp$x, type = "p", col="grey", xlab = "x", ylab = "y",  main = "disttree", xaxt='n', yaxt='n', xlab = "", ylab = "")
-  lines(x = sp$x, y = sp$true.sigma, type = "l", col = 'black')
-  lines(x = sp$x, y = -sp$true.sigma, type = "l", col = 'black')
-  lines(x = sp$x, y = sp$fitted.sigma.dt, type = "l", col = 'green4')
-  lines(x = sp$x, y = -sp$fitted.sigma.dt, type = "l", col = 'green4')
-  lines(x = sp$x, y = sp$fitted.sigma.df, type = "l", col = 'red')
-  lines(x = sp$x, y = -sp$fitted.sigma.df, type = "l", col = 'red')
-  lines(x = sp$x, y = sp$fitted.sigma.g, type = "l", col = 'blue')
-  lines(x = sp$x, y = -sp$fitted.sigma.g, type = "l", col = 'blue')
-  legend('topleft', c(TeX('$\\sigma$'), TeX('$\\hat{\\sigma}$ disttree'),
-                      TeX('$\\hat{\\sigma}$ distforest'), TeX('$\\hat{\\sigma}$ gamlss')), 
-         col = c('black','green4','red','blue'), lty = 1)
+           col = c('black',coltree,colforest,colgamlss), lty = 1, bty = 'n')
+  }
 }
@@
 \frametitle{Motivation}
-\vspace{0.5cm}
+\vspace{-0.41cm}
 \begin{figure}[!htb]
@@
 plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
-lines(x = x, y = ytrue, col = "red", lwd = 1.5, main = "")
+box(lwd=5)
+lines(x = x, y = ytrue, col = "red", lwd = 5, main = "")
 @
@@
 plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
-lines(x = x, y = ytrue, col = "red", lwd = 1.5)
+box(lwd=5)
+lines(x = x, y = ytrue, col = "red", lwd=5)
 lines(x = x, y = ytrue+0.1+var, col = "red", lty = 2)
@@
 \frametitle{Motivation}
-\vspace{0.1cm}
+\vspace{-0.2cm}
 \begin{figure}[!htb]
@@
 plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
+box(lwd=5)
 #lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
-lines(x = x, y = ytree, col = "red", lwd = 1.5)
+lines(x = x, y = ytree, col = "red", lwd=5)
 @
@@
 plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
-#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
-lines(x = x, y = ytree, col = "red", lwd = 1.5)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+lines(x = x, y = ytree, col = "red", lwd=5)
 @
@@
 for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
-#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
-lines(x = x, y = yforest, col = "red", lwd = 1.5, main = "")
 @
@@
 plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
-#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
-lines(x = x, y = ytree, col = "red", lwd = 1.5)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+lines(x = x, y = yforest, col = "red", lwd=5, main = "")
 @
@@
   \begin{tikzpicture}
-  \draw[line width=0.01mm, black] (0,0) rectangle (2.6,2.24);
+  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
   \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
@@
 \visible<3->{
-Distributional Trees and Forests\\
+Distributional trees and forests\\
 \vspace{1.7cm}
@@
 
-\textbf{Distributional:} Specify the complete probability distribution
-(including location, scale and shape).
+\textbf{Distributional:} \\
+\hspace{0.3cm} Specify the complete probability distribution\\
+\hspace{0.3cm} (including location, scale and shape).
 
@@
 \frametitle{Prediction}
-\vspace{1cm}
+%\vspace{1cm}
 {\bf Tree:}\\
-\vspace{0.1cm}
-\hspace{0.7cm} Estimate $\hat{\theta}$ on the subsample of the learning data which \\
-\hspace{0.7cm} ends up in the same terminal node as the new observation.
-%\item Evaluate final node for new observation
-%\item Return corresponding distribution (parameters)
-
+\begin{enumerate}
+\item Select the subsample of the learning data which ends up in the same terminal node as the new observation.
+\item Estimate $\hat{\theta}$ on this subsample.
+\end{enumerate}
 \vspace{0.5cm}
@@
 {\bf Forest:}\\
-\vspace{0.1cm}
-\hspace{0.7cm} Estimate $\hat{\theta}$ on the whole learning data but weighted by \\
-\hspace{0.7cm} the number of trees in which a learning observation ends up\\
-\hspace{0.7cm} in the same terminal node as the new observation.
+\begin{enumerate}
+\item For each learning observation calculate a weight by counting the number of trees in which it ends up in the same terminal node as the new observation.
+\item Estimate $\hat{\theta}$ on the wheighted learning data.
+\end{enumerate}
 % but weighted by similarities of the learning data and the new observation over all trees
@@
 \frametitle{Fitting a Tree}
-\vspace*{-0.4cm}
+\vspace*{-0.5cm}
 \begin{center} 
 DGP:  $Y \sim \mathcal{N}(\mu(X), \sigma(X))$
+\vspace*{-0.21cm}
 <<dgp_tree, echo=FALSE, results=hide>>=
 data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
-names(data) <- c("","","")
+names(data) <- c("x","x","x")
 fig <- party(
@@
                 "n = 200",
-                "True parameters:",
+                "   True parameters:   ",
                 expression(mu == '4'),
@@
                             "n = 200",
-                            "True parameters:",
+                            "   True parameters:   ",
                             expression(mu == '12'),
@@
                             "n = 100",
-                            "True parameters:",
+                            "   True parameters:   ",
                             expression(mu == '4'),
@@
 )
-@
-
-\setkeys{Gin}{width=0.75\linewidth}
+
+node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
+    gp = gpar()) 
+{
+    meta <- obj$data
+    nam <- names(obj)
+    extract_label <- function(node) {
+        if (is.terminal(node)) 
+            return(rep.int("", 2L))
+        varlab <- character_split(split_node(node), meta)$name
+        if (abbreviate > 0L) 
+            varlab <- abbreviate(varlab, as.integer(abbreviate))
+        if (pval) {
+            nullna <- function(x) is.null(x) || is.na(x)
+            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
+                silent = TRUE))
+            pval <- if (inherits(pval, "try-error")) 
+                FALSE
+            else pval
+        }
+        if (pval) {
+            pvalue <- node$info$p.value
+            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
+                paste("p =", round(pvalue, digits = 3L)))
+        }
+        else {
+            plab <- ""
+        }
+        return(c(varlab, plab))
+    }
+    maxstr <- function(node) {
+        lab <- extract_label(node)
+        klab <- if (is.terminal(node)) 
+            ""
+        else unlist(lapply(kids_node(node), maxstr))
+        lab <- c(lab, klab)
+        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
+        lab <- lab[which.max(nchar(lab))]
+        if (length(lab) < 1L) 
+            lab <- ""
+        return(lab)
+    }
+    nstr <- maxstr(node_party(obj))
+    if (nchar(nstr) < 6) 
+        nstr <- "aAAAAa"
+    rval <- function(node) {
+        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
+            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
+            height = unit(3, "lines"), name = paste("node_inner", 
+                id_node(node), sep = ""), gp = gp)
+        pushViewport(node_vp)
+        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
+            seq(0.8, 1, by = 0.01))
+        yell <- sqrt(xell * (1 - xell))
+        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
+        lab <- extract_label(node)
+        fill <- rep(fill, length.out = 2L)
+        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
+            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
+        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
+            "lines"))
+        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
+        #    "lines"))
+        grid.text(lab[2L], y = unit(1, "lines"))
+        if (id) {
+            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
+                "npc"), width = max(unit(1, "lines"), unit(1.3, 
+                "strwidth", nam[id_node(node)])), height = max(unit(1, 
+                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
+            pushViewport(nodeIDvp)
+            grid.rect(gp = gpar(fill = fill[2]))
+            grid.text(nam[id_node(node)])
+            popViewport()
+        }
+        upViewport()
+    }
+    return(rval)
+}
+
+class(node_inner_ext) <- "grapcon_generator"
+
+@
+
+\setkeys{Gin}{width=0.74\linewidth}
 <<plottree_dgp, fig=TRUE, echo=FALSE>>=
 pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-plot(fig,
-     tp_args = list(FUN = identity, width = 18.7, fill = pal[c(1, 3)]), 
-     ip_args = list(fill = pal[c(2, 3)], pval = FALSE),
-     drop_terminal = TRUE, tnex = 1.5)
+plot(fig, inner_panel = node_inner_ext,
+     tp_args = list(FUN = identity, width = 18, fill = pal[c(1, 3)]), 
+     ip_args = list(fill = pal[c(2, 3)]),
+     drop_terminal = TRUE, tnex = 1.4)
 @
@@
 \begin{center}
-\vspace*{-0.4cm}
+\vspace*{-0.5cm}
 Model: \code{disttree(y ~ x)}\\
-\setkeys{Gin}{width=0.75\linewidth}
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.74\linewidth}
 <<plottree_estpar, fig=TRUE, echo=FALSE>>=
 pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-plot(tree, drop = TRUE, tnex = 1.5, #FUN = FUN,
+plot(tree, drop = TRUE, tnex = 1.4, #FUN = FUN,
      tp_args = list(fill = pal[c(1, 3)]), 
      ip_args = list(fill = pal[c(2, 3)]))
+
 @
@@
 \begin{center}
-\vspace*{-0.4cm}
-Model: \code{disttree(y~x)}\\
-\setkeys{Gin}{width=0.75\linewidth}
+\vspace*{-0.5cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.74\linewidth}
 <<plottree_box, fig=TRUE, echo=FALSE>>=
-plot(as.constparty(tree), tnex = 1.5,
+plot(as.constparty(tree), tnex = 1.4,
      tp_args = list(fill = pal[c(1, 3)]), 
@@
 %\SweaveOpts{eval=FALSE}
-<<echo = FALSE, results=hide>>=
+<<sim_1dataset, echo=FALSE, results=hide>>=
 set.seed(723)
@@
 dt <- disttree(y~x1, data=learndata, family=dist_list_normal, type.tree = "ctree", control = control)
+
+if(file.exists("sim_forest.rda")){
+  load("sim_forest.rda")
+} else {
   control <-  ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 7L)
   df <- distforest(y~x1, data=learndata, family=dist_list_normal, ntree = 500L, type.tree = "ctree", control = control)
+  df$info <- df$fitted <- df$weights <- df$nodes <- df$data <- df$terms <- df$loglik <- df$trafo <- NULL
+  save(df, file = "sim_forest.rda")
+}
 g <- gamlss(y ~ pb(x1), sigma.formula = ~pb(x1), data = learndata, trace = FALSE)
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = "forestgreen", ylim = ylim,
+    plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = coltree, ylim = ylim,
          xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
-    lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = 'forestgreen')
-    lines(x = simlist$x.axis, y = rmse[,"df.true"], type = "l", col = 'red')
-    lines(x = simlist$x.axis, y = rmse[,"df.obs"], type = "l", lty = 2, col = 'red')
-    lines(x = simlist$x.axis, y = rmse[,"g.true"], type = "l", col = 'blue')
-    lines(x = simlist$x.axis, y = rmse[,"g.obs"], type = "l", lty = 2, col = 'blue')
-    #legend('left', c("disttree", "distforest", "gamlss"), col = c('forestgreen', 'red','blue'), lty = 1)
-    mtext(text= "smooth", side = 1, line = 2, adj = 0)
-    mtext(text= "steep", side = 1, line = 2, adj = 1)
+    lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = coltree)
+    lines(x = simlist$x.axis, y = rmse[,"df.true"], type = "l", col = colforest)
+    lines(x = simlist$x.axis, y = rmse[,"df.obs"], type = "l", lty = 2, col = colforest)
+    lines(x = simlist$x.axis, y = rmse[,"g.true"], type = "l", col = colgamlss)
+    lines(x = simlist$x.axis, y = rmse[,"g.obs"], type = "l", lty = 2, col = colgamlss)
+    #legend('left', c("disttree", "distforest", "gamlss"), col = c(coltree, colforest, colgamlss), lty = 1)
+    mtext(text= "smooth", side = 1, line = 1, adj = 0)
+    mtext(text= "steep", side = 1, line = 1, adj = 1)
     mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
     mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = "darkgreen", las = 1, line = 0.2, padj = 6.5)
-    mtext(text= "distforest", side = 2, col = "red", las = 1, line = 0.2, padj = 13)
-    mtext(text= "gamlss", side = 2, col = "blue", las = 1, line = 0.3, padj = 16)
+    mtext(text= "disttree", side = 2, col = coltree, las = 1, line = 0.2, padj = 6.5)
+    mtext(text= "distforest", side = 2, col = colforest, las = 1, line = 0.2, padj = 13)
+    mtext(text= "gamlss", side = 2, col = colgamlss, las = 1, line = 0.3, padj = 16)
 
@@
     #add arrows below
-    arrows(0.45,0.05,0.3,0.05,lwd=1) ## add arrow
-    arrows(0.6,0.05,0.75,0.05,lwd=1) ## add arrow
+    arrows(0.45,0.09,0.3,0.09,lwd=1) ## add arrow
+    arrows(0.6,0.09,0.75,0.09,lwd=1) ## add arrow
 
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = "forestgreen", ylim = ylim,
+    plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = coltree, ylim = ylim,
          xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
-    lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = 'forestgreen')
-    lines(x = simlist$x.axis, y = rmse[,"mu.df"], type = "l", col = 'red')
-    lines(x = simlist$x.axis, y = rmse[,"sigma.df"], type = "l", lty = 2, col = 'red')
-    lines(x = simlist$x.axis, y = rmse[,"mu.g"], type = "l", col = 'blue')
-    lines(x = simlist$x.axis, y = rmse[,"sigma.g"], type = "l", lty = 2, col = 'blue')
-    #legend('topleft', c("disttree", "distforest", "gamlss"), col = c('forestgreen', 'red','blue'), lty = 1)
-    mtext(text= "smooth", side = 1, line = 2, adj = 0)
-    mtext(text= "steep", side = 1, line = 2, adj = 1)
+    lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = coltree)
+    lines(x = simlist$x.axis, y = rmse[,"mu.df"], type = "l", col = colforest)
+    lines(x = simlist$x.axis, y = rmse[,"sigma.df"], type = "l", lty = 2, col = colforest)
+    lines(x = simlist$x.axis, y = rmse[,"mu.g"], type = "l", col = colgamlss)
+    lines(x = simlist$x.axis, y = rmse[,"sigma.g"], type = "l", lty = 2, col = colgamlss)
+    #legend('topleft', c("disttree", "distforest", "gamlss"), col = c(coltree, colforest,colgamlss), lty = 1)
+    mtext(text= "smooth", side = 1, line = 1, adj = 0)
+    mtext(text= "steep", side = 1, line = 1, adj = 1)
     mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
     mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = "darkgreen", las = 1, line = 0.2, padj = -4)
-    mtext(text= "distforest", side = 2, col = "red", las = 1, line = 0.2, padj = 8)
-    mtext(text= "gamlss", side = 2, col = "blue", las = 1, line = 0.3, padj = 14)
+    mtext(text= "disttree", side = 2, col = coltree, las = 1, line = 0.2, padj = -4)
+    mtext(text= "distforest", side = 2, col = colforest, las = 1, line = 0.2, padj = 8)
+    mtext(text= "gamlss", side = 2, col = colgamlss, las = 1, line = 0.3, padj = 14)
 
@@
   par(mar=c(5.1,4.1,4.1,3.1))
-  plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = "forestgreen", ylim = ylim,
+  plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = coltree, ylim = ylim,
        xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
-  lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = 'red')
-  lines(x = simlist$x.axis, y = ll[,"g"], type = "l", col = 'blue')
-  #legend('topright', c("disttree", "distforest", "gamlss"), col = c('forestgreen', 'red','blue'), lty = 1)
-  mtext(text= "smooth", side = 1, line = 2, adj = 0)
-  mtext(text= "steep", side = 1, line = 2, adj = 1)
+  lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = colforest)
+  lines(x = simlist$x.axis, y = ll[,"g"], type = "l", col = colgamlss)
+  #legend('topright', c("disttree", "distforest", "gamlss"), col = c(coltree, colforest,colgamlss), lty = 1)
+  mtext(text= "smooth", side = 1, line = 1, adj = 0)
+  mtext(text= "steep", side = 1, line = 1, adj = 1)
   mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
   mtext(text= "Log-Likelihood", side = 2, line = 1)
-  mtext(text= "disttree", side = 2, col = "darkgreen", las = 1, line = 0.2, padj = 12)
-  mtext(text= "distforest", side = 2, col = "red", las = 1, line = 0.2, padj = -8.5)
-  mtext(text= "gamlss", side = 2, col = "blue", las = 1, line = 0.3, padj = -16)
+  mtext(text= "disttree", side = 2, col = coltree, las = 1, line = 0.2, padj = 12)
+  mtext(text= "distforest", side = 2, col = colforest, las = 1, line = 0.2, padj = -8.5)
+  mtext(text= "gamlss", side = 2, col = colgamlss, las = 1, line = 0.3, padj = -16)
 
@@
   #add arrows below
-  arrows(0.45,0.05,0.25,0.05,lwd=1) ## add arrow
-  arrows(0.6,0.05,0.8,0.05,lwd=1) ## add arrow
+  arrows(0.45,0.09,0.3,0.09,lwd=1) ## add arrow
+  arrows(0.6,0.09,0.75,0.09,lwd=1) ## add arrow
 
@@
 \vspace{0.2cm}
-\url{https://R-Forge.R-project.org/projects/partykit/}
+\url{https://R-Forge.R-project.org/projects/partykit/}\\
 \vspace{1cm}

### 2017-06-22 pkg/disttree/inst/useR2017/slides.Rnw
modified plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1476 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-coltree <- hcl(c(128,100,30))
-colforest <- hcl(c(10,100,30))
-colgamlss <- hcl(c(260,100,30))
+pal <- hcl(c(10, 128, 260), 100, 50)
+names(pal) <- c("forest", "tree", "gamlss")
 
@@
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = coltree)
+    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
     lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
@@
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey',coltree, 'greenyellow'), lty = 1, bty = 'n')
+           col = c('black','grey',pal["tree"], 'greenyellow'), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = colforest)
+    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
     lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
@@
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey',colforest, 'orange'), lty = 1, bty = 'n')
+           col = c('black','grey',pal["forest"], 'orange'), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = colgamlss)
+    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 1.5)
     lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
@@
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
-           col = c('black','grey',colgamlss, 'lightblue'), lty = 1, bty = 'n')
+           col = c('black','grey',pal["gamlss"], 'lightblue'), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = coltree)
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = colforest)
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = colgamlss)
+    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
+    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
+    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
     legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
                         TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
-           col = c('black',coltree,colforest,colgamlss), lty = 1, bty = 'n')
+           col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = 'n')
   }
@@
 <<plottree_dgp, fig=TRUE, echo=FALSE>>=
-pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+pal2 <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
 plot(fig, inner_panel = node_inner_ext,
-     tp_args = list(FUN = identity, width = 18, fill = pal[c(1, 3)]), 
-     ip_args = list(fill = pal[c(2, 3)]),
+     tp_args = list(FUN = identity, width = 18, fill = pal2[c(1, 3)]), 
+     ip_args = list(fill = pal2[c(2, 3)]),
      drop_terminal = TRUE, tnex = 1.4)
@@
 #tree <- disttree(y ~ x, data = data, family = dist_list_cens_normal)
-
-#FUN <- function(x) {
-#                cf <- x$coefficients
-#                cf <- matrix(cf, ncol = 1, dimames = list(c(expression(mu), expression(sigma)), ""))
-#                c(sprintf("n = %s", x$nobs), "Estimated parameters:", 
-#                  strwrap(capture.output(print(cf, digits = 4L))[-1L]))
-#}
-
-#FUN <- function(x) {
-#                cf <- x$coefficients
-#                cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), 
-#                  ""))
-#                c(sprintf("n = %s", x$nobs), "Estimated parameters:", 
-#                  strwrap(capture.output(print(cf, digits = 4L))[-1L]))
-#            }
-#FUN <- function (x) 
-#{
-#  cf <- x$coefficients
-#  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
-#  c(sprintf("n = %s", x$nobs), "Estimated parameters:", c(c(expression(mu), round(cf[1],3)), 
-#                                                          c(expression(sigma), round(cf[2],3))))
-#}
 @
@@
 <<plottree_estpar, fig=TRUE, echo=FALSE>>=
-pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-plot(tree, drop = TRUE, tnex = 1.4, #FUN = FUN,
-     tp_args = list(fill = pal[c(1, 3)]), 
-     ip_args = list(fill = pal[c(2, 3)]))
-
+# function for output in terminal panels
+FUN <- function (x) 
+{
+  cf <- x$coefficients
+  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
+  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste("mu == ", round(cf[1],3))), 
+                                                        parse(text = paste("sigma == ", round(cf[2],3))))
+}
+
+# alternative to using FUN:
+# change info slot in each terminal node and use FUN = identity
+if(FALSE){
+info2 <- tree$node$kids[[1]]$info 
+info4 <- tree$node$kids[[2]]$kids[[1]]$info 
+info5 <- tree$node$kids[[2]]$kids[[2]]$info
+
+cf <- info2$coefficients
+info2 <- c(sprintf("n = %s", info2$nobs), "Estimated parameters:", parse(text = paste("mu == ", round(cf[1],3))), 
+                                                          parse(text = paste("sigma == ", round(cf[2],3))))
+
+cf <- info4$coefficients
+info4 <- c(sprintf("n = %s", info4$nobs), "Estimated parameters:", parse(text = paste("mu == ", round(cf[1],3))), 
+                                                          parse(text = paste("sigma == ", round(cf[2],3))))
+
+cf <- info5$coefficients
+info5 <- c(sprintf("n = %s", info5$nobs), "Estimated parameters:", parse(text = paste("sigma == ", round(cf[2],3))), 
+                                                          parse(text = paste("mu == ", round(cf[1],3))))
+
+tree$node$kids[[1]]$info <- info2
+tree$node$kids[[2]]$kids[[1]]$info <- info4 
+tree$node$kids[[2]]$kids[[2]]$info <- info5
+
+class(tree) <- "party"
+}
+
+pal2 <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+
+## plot version using FUN and tree of class 'disttree'
+plot(tree, drop = TRUE, tnex = 1.4, FUN = FUN,
+     tp_args = list(fill = pal2[c(1, 3)], width = 18), 
+     ip_args = list(fill = pal2[c(2, 3)]))
+
+## plot version using replaced info slots and tree of class 'party'
+if(FALSE){
+  plot(tree, drop = TRUE, tnex = 1.4,
+       tp_args = list(fill = pal2[c(1, 3)], FUN = identity, width = 18), 
+       ip_args = list(fill = pal2[c(2, 3)]))
+}
 @
@@
 plot(as.constparty(tree), tnex = 1.4,
-     tp_args = list(fill = pal[c(1, 3)]), 
-     ip_args = list(fill = pal[c(2, 3)]))
+     tp_args = list(fill = pal2[c(1, 3)]), 
+     ip_args = list(fill = pal2[c(2, 3)]))
 @
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = coltree, ylim = ylim,
-         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
-    lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = coltree)
-    lines(x = simlist$x.axis, y = rmse[,"df.true"], type = "l", col = colforest)
-    lines(x = simlist$x.axis, y = rmse[,"df.obs"], type = "l", lty = 2, col = colforest)
-    lines(x = simlist$x.axis, y = rmse[,"g.true"], type = "l", col = colgamlss)
-    lines(x = simlist$x.axis, y = rmse[,"g.obs"], type = "l", lty = 2, col = colgamlss)
-    #legend('left', c("disttree", "distforest", "gamlss"), col = c(coltree, colforest, colgamlss), lty = 1)
+    plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = pal["tree"], ylim = ylim,
+         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"df.true"], type = "l", col = pal["forest"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"df.obs"], type = "l", lty = 2, col = pal["forest"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"g.true"], type = "l", col = pal["gamlss"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"g.obs"], type = "l", lty = 2, col = pal["gamlss"], lwd = 2)
+    #legend('left', c("disttree", "distforest", "gamlss"), col = c(pal["tree"], pal["forest"], pal["gamlss"]), lty = 1)
     mtext(text= "smooth", side = 1, line = 1, adj = 0)
@@
     mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = coltree, las = 1, line = 0.2, padj = 6.5)
-    mtext(text= "distforest", side = 2, col = colforest, las = 1, line = 0.2, padj = 13)
-    mtext(text= "gamlss", side = 2, col = colgamlss, las = 1, line = 0.3, padj = 16)
+    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 6.5)
+    mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = 13)
+    mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = 16)
 
@@
     #add arrows below
-    arrows(0.45,0.09,0.3,0.09,lwd=1) ## add arrow
-    arrows(0.6,0.09,0.75,0.09,lwd=1) ## add arrow
+    arrows(0.45,0.085,0.3,0.085,lwd=1) ## add arrow
+    arrows(0.6,0.085,0.75,0.085,lwd=1) ## add arrow
 
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = coltree, ylim = ylim,
-         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
-    lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = coltree)
-    lines(x = simlist$x.axis, y = rmse[,"mu.df"], type = "l", col = colforest)
-    lines(x = simlist$x.axis, y = rmse[,"sigma.df"], type = "l", lty = 2, col = colforest)
-    lines(x = simlist$x.axis, y = rmse[,"mu.g"], type = "l", col = colgamlss)
-    lines(x = simlist$x.axis, y = rmse[,"sigma.g"], type = "l", lty = 2, col = colgamlss)
-    #legend('topleft', c("disttree", "distforest", "gamlss"), col = c(coltree, colforest,colgamlss), lty = 1)
+    plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = pal["tree"], ylim = ylim,
+         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"mu.df"], type = "l", col = pal["forest"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"sigma.df"], type = "l", lty = 2, col = pal["forest"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"mu.g"], type = "l", col = pal["gamlss"], lwd = 2)
+    lines(x = simlist$x.axis, y = rmse[,"sigma.g"], type = "l", lty = 2, col = pal["gamlss"], lwd = 2)
+    #legend('topleft', c("disttree", "distforest", "gamlss"), col = c(pal["tree"], pal["forest"],pal["gamlss"]), lty = 1)
     mtext(text= "smooth", side = 1, line = 1, adj = 0)
@@
     mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = coltree, las = 1, line = 0.2, padj = -4)
-    mtext(text= "distforest", side = 2, col = colforest, las = 1, line = 0.2, padj = 8)
-    mtext(text= "gamlss", side = 2, col = colgamlss, las = 1, line = 0.3, padj = 14)
+    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = -4)
+    mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = 8)
+    mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = 14)
 
@@
   par(mar=c(5.1,4.1,4.1,3.1))
-  plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = coltree, ylim = ylim,
-       xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n')
-  lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = colforest)
-  lines(x = simlist$x.axis, y = ll[,"g"], type = "l", col = colgamlss)
-  #legend('topright', c("disttree", "distforest", "gamlss"), col = c(coltree, colforest,colgamlss), lty = 1)
+  plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = pal["tree"], ylim = ylim,
+       xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', lwd = 2)
+  lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = pal["forest"], lwd = 2)
+  lines(x = simlist$x.axis, y = ll[,"g"], type = "l", col = pal["gamlss"], lwd = 2)
+  #legend('topright', c("disttree", "distforest", "gamlss"), col = c(pal["tree"], pal["forest"],pal["gamlss"]), lty = 1)
   mtext(text= "smooth", side = 1, line = 1, adj = 0)
@@
   mtext(text= "Log-Likelihood", side = 2, line = 1)
-  mtext(text= "disttree", side = 2, col = coltree, las = 1, line = 0.2, padj = 12)
-  mtext(text= "distforest", side = 2, col = colforest, las = 1, line = 0.2, padj = -8.5)
-  mtext(text= "gamlss", side = 2, col = colgamlss, las = 1, line = 0.3, padj = -16)
+  mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 12)
+  mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = -8.5)
+  mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = -16)
 
@@
   #add arrows below
-  arrows(0.45,0.09,0.3,0.09,lwd=1) ## add arrow
-  arrows(0.6,0.09,0.75,0.09,lwd=1) ## add arrow
+  arrows(0.45,0.085,0.3,0.085,lwd=1) ## add arrow
+  arrows(0.6,0.085,0.75,0.085,lwd=1) ## add arrow
 

### 2017-06-22 pkg/disttree/inst/useR2017/slides.Rnw
changed loaded simlist and deleted other rda-files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1477 255f4249-2263-4417-b1ea-8498779e27a7

@@
     legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
-           col = c('black','grey'), lty = 1, bty = 'n')
+           col = c('black','black'), lty = c(1,2,2), bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 1.5)
+    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
     lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
@@
 <<preliminaries3, echo=FALSE, results=hide>>=
-load("simlist20170621.rda")
+load("simlist20170621b.rda")
 
@@
     mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 6.5)
+    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 5)
     mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = 13)
-    mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = 16)
+    mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = 16.5)
 
@@
     mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = -4)
+    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = -7)
     mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = 8)
@@
   mtext(text= "Log-Likelihood", side = 2, line = 1)
-  mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 12)
-  mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = -8.5)
+  mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 16)
+  mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = -8)
   mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = -16)

### 2017-06-22 pkg/disttree/inst/useR2017/slides.Rnw
changed citation and reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1478 255f4249-2263-4417-b1ea-8498779e27a7
added function 'gensimlist.R' and rda-files with data needed for plots; in 'slides.Rnw' it is first checked whether these rda-files already exist, otherwise they are generated; for 'simlist.rda' an external function 'gensimlist.R' is sourced and generates simlist
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1479 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
+%\usepackage[style=authoryear, backend=bibtex]{biblatex}
 \graphicspath{{plots/}}
@@
 
-<<prelimminaries2, echo=FALSE, results=hide>>=
+<<treedata, echo=FALSE, results=hide>>=
 set.seed(54)
@@
 }
+
 g <- gamlss(y ~ pb(x1), sigma.formula = ~pb(x1), data = learndata, trace = FALSE)
@@
 
-<<preliminaries3, echo=FALSE, results=hide>>=
-load("simlist20170621b.rda")
+<<get_simlist, echo=FALSE, results=hide>>=
+if(file.exists("simlist.rda")){
+  load("simlist.rda")
+} else {
+  source("gensimlist.R")
+  simlist <- gensimlist(seed = 7, nrep = 100, ntree = 100)
+  save(simlist, file = "simlist.rda")
+}
 

### 2017-06-22 pkg/disttree/inst/useR2017/gensimlist.R

@@ -1,0 +1,551 @@
+### generates simlist which will then be plotted in slides.Rnw
+gensimlist <- function(seed = 7, nrep = 100, ntree = 100)
+{
+  library("disttree")
+  library("gamlss")
+  library("parallel")
+  
+  ## define distribution list:
+  # dist_list_normal
+  {
+    
+    dist_list_normal <- list()
+    
+    parnames <- c("mu", "sigma")
+    etanames <- c("mu", "log(sigma)")
+    
+    
+    ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {     
+      
+      val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
+      if(!log) val <- exp(val)
+      
+      # par <- c(eta[1], exp(eta[2]))
+      # val <- dnorm(y, mean = par[1], sd = par[2], log = log)
+      
+      if(sum) {
+        if(is.null(weights)) weights <- rep.int(1, length(y))
+        val <- sum(weights * val, na.rm = TRUE)
+      }
+      return(val)
+    }
+    
+    
+    sdist <- function(y, eta, weights = NULL, sum = FALSE) {   
+      
+      score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
+                     -1 + exp(-2*eta[2] + log((y-eta[1])^2)))
+      
+      # par <- c(eta[1], exp(eta[2])) 
+      # score <- cbind(1/par[2]^2 * (y-par[1]), 
+      #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))
+      
+      score <- as.matrix(score)
+      colnames(score) <- etanames
+      if(sum) {
+        if(is.null(weights)) weights <- rep.int(1, length(y))
+        # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
+        score[score==Inf] = 1.7e308
+        score <- colSums(weights * score, na.rm = TRUE)
+      }
+      return(score)
+    }
+    
+    
+    hdist <- function(y, eta, weights = NULL) {    
+      ny <- length(y)
+      if(is.null(weights)) weights <- rep.int(1, ny)
+      
+      d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
[…]

### 2017-06-23 pkg/disttree/inst/useR2017/slides.Rnw
added new data for 'simlist'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1482 255f4249-2263-4417-b1ea-8498779e27a7

@@
   source("gensimlist.R")
-  simlist <- gensimlist(seed = 7, nrep = 100, ntree = 100)
+  simlist <- gensimlist(seed = 9, nrep = 150, ntree = 100)
   save(simlist, file = "simlist.rda")

### 2017-06-23 pkg/disttree/inst/useR2017/slides.Rnw
touch-ups for colors/graphics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1483 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+## HCL palette
 pal <- hcl(c(10, 128, 260), 100, 50)
@@
 
-  # disttree vs distforest vs gamlss
+  # disttree vs. distforest vs. gamlss
   if(all.par){
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
@@
 par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
+plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
 box(lwd=5)
-lines(x = x, y = ytrue, col = "red", lwd = 5, main = "")
+lines(x = x, y = ytrue, col = pal["forest"], lwd = 5, main = "")
 @
@@
 par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
+plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
 box(lwd=5)
-lines(x = x, y = ytrue, col = "red", lwd=5)
-lines(x = x, y = ytrue+0.1+var, col = "red", lty = 2)
-lines(x = x, y = ytrue-0.1-var, col = "red", lty = 2)
+lines(x = x, y = ytrue, col = pal["forest"], lwd=5)
+lines(x = x, y = ytrue+0.1+var, col = pal["forest"], lty = 1, lwd = 2)
+lines(x = x, y = ytrue-0.1-var, col = pal["forest"], lty = 1, lwd = 2)
 @
@@
 for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
-plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
+plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
 box(lwd=5)
 #lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
-lines(x = x, y = ytree, col = "red", lwd=5)
+lines(x = x, y = ytree, col = pal["forest"], lwd=5)
 @
@@
 par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
+plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
 box(lwd=5)
 #lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = ytree, col = "red", lwd=5)
+lines(x = x, y = ytree, col = pal["forest"], lwd=5)
 @
@@
 par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
+plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
 box(lwd=5)
 #lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = yforest, col = "red", lwd=5, main = "")
+lines(x = x, y = yforest, col = pal["forest"], lwd=5, main = "")
 @
@@
   cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
-  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste("mu == ", round(cf[1],3))), 
-                                                        parse(text = paste("sigma == ", round(cf[2],3))))
-}
-
-# alternative to using FUN:
-# change info slot in each terminal node and use FUN = identity
-if(FALSE){
-info2 <- tree$node$kids[[1]]$info 
-info4 <- tree$node$kids[[2]]$kids[[1]]$info 
-info5 <- tree$node$kids[[2]]$kids[[2]]$info
-
-cf <- info2$coefficients
-info2 <- c(sprintf("n = %s", info2$nobs), "Estimated parameters:", parse(text = paste("mu == ", round(cf[1],3))), 
-                                                          parse(text = paste("sigma == ", round(cf[2],3))))
-
-cf <- info4$coefficients
-info4 <- c(sprintf("n = %s", info4$nobs), "Estimated parameters:", parse(text = paste("mu == ", round(cf[1],3))), 
-                                                          parse(text = paste("sigma == ", round(cf[2],3))))
-
-cf <- info5$coefficients
-info5 <- c(sprintf("n = %s", info5$nobs), "Estimated parameters:", parse(text = paste("sigma == ", round(cf[2],3))), 
-                                                          parse(text = paste("mu == ", round(cf[1],3))))
-
-tree$node$kids[[1]]$info <- info2
-tree$node$kids[[2]]$kids[[1]]$info <- info4 
-tree$node$kids[[2]]$kids[[2]]$info <- info5
-
-class(tree) <- "party"
+  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
+                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
 }
@@
      ip_args = list(fill = pal2[c(2, 3)]))
-
-## plot version using replaced info slots and tree of class 'party'
-if(FALSE){
-  plot(tree, drop = TRUE, tnex = 1.4,
-       tp_args = list(fill = pal2[c(1, 3)], FUN = identity, width = 18), 
-       ip_args = list(fill = pal2[c(2, 3)]))
-}
 @
@@
 par(mar=c(5.1,4.1,4.1,3.1))
-plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "darkgrey")
-#plot(y=odata$y, x=odata$x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
-lines(x = odata$x, y = odata$mu, col = "red", lwd = 1.5, main = "")
-lines(x = odata$x, y = odata$mu + odata$sigma, col = "red", lwd = 1.5, lty = 2, main = "")
-lines(x = odata$x, y = odata$mu - odata$sigma, col = "red", lwd = 1.5, lty = 2, main = "")
+plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "slategray")
+#plot(y=odata$y, x=odata$x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray")
+lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
+lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
+lines(x = odata$x, y = odata$mu - odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
 @
@@
     plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = pal["tree"], ylim = ylim,
-         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', lwd = 2)
+         xlab = "", ylab = "", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', lwd = 2)
     lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
@@
     plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = pal["tree"], ylim = ylim,
-         xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', lwd = 2)
+         xlab = "", ylab = "", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', lwd = 2)
     lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
@@
   plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = pal["tree"], ylim = ylim,
-       xlab = "", ylab = "", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', lwd = 2)
+       xlab = "", ylab = "", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', lwd = 2)
   lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = pal["forest"], lwd = 2)

### 2017-06-23 pkg/disttree/inst/useR2017/slides.Rnw
use transparency rather than line type for scale
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1484 255f4249-2263-4417-b1ea-8498779e27a7

@@
 par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal["forest"], lwd=5)
-lines(x = x, y = ytrue+0.1+var, col = pal["forest"], lty = 1, lwd = 2)
-lines(x = x, y = ytrue-0.1-var, col = pal["forest"], lty = 1, lwd = 2)
+plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
+  col = hcl(10, 100, 50, alpha = 0.2), border = "transparent")
+lines(x, ytrue, col = pal["forest"], lwd=5)
+points(x, y, col = "slategray", pch = 19)
+box(lwd = 5)
 @

### 2017-06-23 pkg/disttree/inst/useR2017/slides.Rnw
different representation of variance in the plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1485 255f4249-2263-4417-b1ea-8498779e27a7

--- 2017-06-23 pkg/disttree/inst/useR2017/slides.Rnw
@@
 names(pal) <- c("forest", "tree", "gamlss")
+
+pallight <- hcl(c(10, 128, 260), 100, 50, alpha = 0.25)
+names(pallight) <- c("forest", "tree", "gamlss")
+
+transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)
 
@@
 ### plotting function for one data set
-plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE) 
+plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE, all.var = FALSE) 
 {
@@
 
-  if(!(any(dt.par, df.par, g.par, all.par))){
+  if(!(any(dt.par, df.par, g.par, all.par, all.var))){
     par(mar=c(5.1,4.1,4.1,3.1))
@@
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 2, col = 'black')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 2, col = 'black')
+    polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
+  col = transpgrey, border = "transparent")
+    #lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 1, col = 'grey')
+    #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
     legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
-           col = c('black','black'), lty = c(1,2,2), bty = 'n')
+           col = c('black',transpgrey, transpgrey), lty = c(1,1,1), bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
+    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
+    #col = transpgrey, border = "transparent")
     lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
+    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2.5)
+    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
+  col = pallight["tree"], border = "transparent")
+    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
+    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
     legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey',pal["tree"], 'greenyellow'), lty = 1, bty = 'n')
+           col = c('black','grey',pal["tree"], pallight["tree"]), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
+    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
+    #col = transpgrey, border = "transparent")
     lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
-    lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
+    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2.5)
+    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
+  col = pallight["forest"], border = "transparent")
+    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
+    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
     legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey',pal["forest"], 'orange'), lty = 1, bty = 'n')
+           col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = 'n')
   }
@@
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
+    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
+    #col = transpgrey, border = "transparent")
     lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
     lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
+    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2.5)
+    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
+  col = pallight["gamlss"], border = "transparent")
+    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
+    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
     legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                         TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
-           col = c('black','grey',pal["gamlss"], 'lightblue'), lty = 1, bty = 'n')
+           col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = 'n')
   }
@@
   }
+
+  # variance: disttree vs. distforest vs. gamlss
+  if(all.var){
+    par(mar=c(5.1,4.1,4.1,3.1))
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
+    #col = transpgrey, border = "transparent")
+    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'black')
+    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'black')
+    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
+  col = pallight["tree"], border = "transparent")
+    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
+    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
+    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
+  col = pallight["forest"], border = "transparent")
+    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
+    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
+    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
+  col = pallight["gamlss"], border = "transparent")
+    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
+    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
+    legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
+                        TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
+           col = c('black',pallight["tree"],pallight["forest"],pallight["gamlss"]), lty = 1, bty = 'n')
+  }
 }
@@
 box(lwd=5)
-lines(x = x, y = ytrue, col = pal["forest"], lwd = 5, main = "")
+lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
 @
@@
 par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal["forest"], lwd=5)
-lines(x = x, y = ytrue+0.1+var, col = pal["forest"], lty = 1, lwd = 2)
-lines(x = x, y = ytrue-0.1-var, col = pal["forest"], lty = 1, lwd = 2)
-@
+plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
+  col = pallight["forest"], border = "transparent")
+lines(x, ytrue, col = pal["forest"], lwd=7)
+points(x, y, col = "slategray", pch = 19)
+box(lwd = 5)
+@
+
 }
@@
 for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
-plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
-lines(x = x, y = ytree, col = pal["forest"], lwd=5)
 @
@@
 #lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = ytree, col = pal["forest"], lwd=5)
+lines(x = x, y = ytree, col = pal["forest"], lwd=7)
 @
@@
 #lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = yforest, col = pal["forest"], lwd=5, main = "")
+lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
 @
@@
 par(mar=c(5.1,4.1,4.1,3.1))
-plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "slategray")
+plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "grey")
 #plot(y=odata$y, x=odata$x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray")
 lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
-lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
-lines(x = odata$x, y = odata$mu - odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
+polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
+  col = pallight["forest"], border = "transparent")
+
+#lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
+#lines(x = odata$x, y = odata$mu - odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
 @
@@
 plot1(learndata, dt=dt, df=df, g=g, all.par = TRUE)
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Simulation}
+\begin{center}
+\vspace*{-1cm}
+\setkeys{Gin}{width=0.8\linewidth}
+<<oneforest_exp1_allvar, fig=TRUE, echo=FALSE>>=
+plot1(learndata, dt=dt, df=df, g=g, all.var = TRUE)
 @
@@
 \begin{itemize}
-\item \code{distfit()}
-\item \code{disttree()}
-\item \code{distforest()}
+\item \code{distfit()}: distributional fit, no covariates, gamlss.family, ML
+\item \code{disttree()}: tree algorithm \code{ctree} or \code{mob}, \\
+\hspace{2.18cm} covariates as splitting variables
+\item \code{distforest()}: ensemble of trees,\\
+\hspace{2.58cm} covariates as splitting variables
 \end{itemize}
 \end{frame}
+
+\subsection{References}
 
@@
 \frametitle{References}
-\nocite{*}
-%\nocite{hothorn2006unbiased, zeileis2008model, hothorn2015partykit, stasinopoulos2007generalized, seibold2016model}
-{\footnotesize
-\bibliography{ref.bib}}
+
+\footnotesize
+
+Hothorn T, Hornik K, Zeileis A (2006).\\
+ \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}\\
+ \textit{Journal of Computational and Graphical Statistics},
+ \textbf{15}(3), 651--674.\\
+ %URL~\url{http://    }
+
+\bigskip
+
+Zeileis A, Hothorn T, Hornik K (2008).\\
+ \dquote{Model-Based Recursive Partitioning.}\\
+  \textit{Journal of Computational and Graphical Statistics},
+  \textbf{17}(2), 492--514.\\
+  %\doi{10.1198/106186008X319331}
+
+\bigskip
+
+Hothorn T, Zeileis A (2015).\\
+ \dquote{Partykit: A Modular Toolkit for Recursive Partytioning in R.}\\
+ \textit{Journal of Machine Learning Research},
+ \textbf{16}, 3905--3909.\\
+
+\bigskip
+
+Stasinopoulos DM, Rigby RA, et al. (2007).\\
+  \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in R.}\\
+  \textit{Journal of Statistical Software}, 
+  \textbf{23}(7), 1--46.\\
+
+\bigskip
+
+Seibold H, Zeileis A, Hothorn T (2016).\\
+  \dquote{Model-Based Recursive Partitioning for Subgroup Analyses.}\\
+  \textit{The international journal of biostatistics}, 
+  \textbf{12}(1), 45--63.\\
+
 \end{frame}

### 2017-06-23 pkg/disttree/inst/useR2017/slides.Rnw
added comments on the 3 main functions 'distfit', 'disttree' and 'distforest'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1486 255f4249-2263-4417-b1ea-8498779e27a7
capitalized first word and added point in slide 'Software'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1488 255f4249-2263-4417-b1ea-8498779e27a7
edits in software and references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1489 255f4249-2263-4417-b1ea-8498779e27a7

@@
     #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
-           col = c('black',transpgrey, transpgrey), lty = c(1,1,1), bty = 'n')
+    legend('topleft', expression(mu %+-% sigma), bty = 'n')
   }
@@
 \vspace{0.4cm}
-R-package \pkg{disttree} available on R-Forge:\\
-\vspace{0.2cm}
+\textbf{Package:} \emph{disttree} available on R-Forge at\\
+
+\medskip
+
 \url{https://R-Forge.R-project.org/projects/partykit/}\\
-\vspace{1cm}
-Main functions:
-\vspace{0.2cm}
-\begin{itemize}
-\item \code{distfit()}: distributional fit, no covariates, gamlss.family, ML
-\item \code{disttree()}: tree algorithm \code{ctree} or \code{mob}, \\
-\hspace{2.18cm} covariates as splitting variables
-\item \code{distforest()}: ensemble of trees,\\
-\hspace{2.58cm} covariates as splitting variables
-\end{itemize}
+
+\bigskip
+\bigskip
+
+\textbf{Main functions:}
+
+\medskip
+
+\begin{tabular}{ll}
+\code{distfit}    & Distributional fits (ML, \code{gamlss.family} or custom \code{list}).\\
+                  & No covariates. \\
+\code{disttree}   & Distributional trees (\code{ctree}/\code{mob} + \code{distfit}).\\
+                  & Covariates as partitioning variables. \\
+\code{distforest} & Distributional forests (ensemble of \code{disttree}s).\\
+                  & Covariates as partitioning variables.
+\end{tabular}
+
 \end{frame}
@@
 
-Hothorn T, Hornik K, Zeileis A (2006).\\
- \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}\\
+Hothorn T, Hornik K, Zeileis A (2006).
+ \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
  \textit{Journal of Computational and Graphical Statistics},
- \textbf{15}(3), 651--674.\\
- %URL~\url{http://    }
-
-\bigskip
-
-Zeileis A, Hothorn T, Hornik K (2008).\\
- \dquote{Model-Based Recursive Partitioning.}\\
+ \textbf{15}(3), 651--674.
+ \doi{10.1198/106186006X133933}
+
+\medskip
+
+Zeileis A, Hothorn T, Hornik K (2008).
+ \dquote{Model-Based Recursive Partitioning.}
   \textit{Journal of Computational and Graphical Statistics},
-  \textbf{17}(2), 492--514.\\
-  %\doi{10.1198/106186008X319331}
-
-\bigskip
-
-Hothorn T, Zeileis A (2015).\\
- \dquote{Partykit: A Modular Toolkit for Recursive Partytioning in R.}\\
+  \textbf{17}(2), 492--514.
+  \doi{10.1198/106186008X319331}
+
+\medskip
+
+Hothorn T, Zeileis A (2015).
+ \dquote{partykit: A Modular Toolkit for Recursive Partytioning in R.}
  \textit{Journal of Machine Learning Research},
- \textbf{16}, 3905--3909.\\
-
-\bigskip
-
-Stasinopoulos DM, Rigby RA, et al. (2007).\\
-  \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in R.}\\
+ \textbf{16}, 3905--3909.
+ \url{http://www.jmlr.org/papers/v16/hothorn15a.html}
+
+\medskip
+
+Stasinopoulos DM, Rigby RA (2007).
+  \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in R.}
   \textit{Journal of Statistical Software}, 
-  \textbf{23}(7), 1--46.\\
-
-\bigskip
-
-Seibold H, Zeileis A, Hothorn T (2016).\\
-  \dquote{Model-Based Recursive Partitioning for Subgroup Analyses.}\\
-  \textit{The international journal of biostatistics}, 
-  \textbf{12}(1), 45--63.\\
+  \textbf{23}(7), 1--46.
+  \doi{10.18637/jss.v023.i07}
+
+\medskip
+
+Seibold H, Zeileis A, Hothorn T (2017).
+  \dquote{Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.}
+  \textit{Statistical Methods in Medical Research}, 
+  \textbf{12}(1), 45--63.
+  \doi{10.1177/0962280217693034}
+
+\medskip
+
+Hothorn T, Zeileis A (2017).
+  \dquote{Transformation Forests.}
+  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
+  \url{http://arxiv.org/abs/1701.02110}
 

### 2017-06-24 pkg/disttree/inst/useR2017/slides.Rnw
plots: now colored title instead of colored lines in legend
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1490 255f4249-2263-4417-b1ea-8498779e27a7

@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
@@
     #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    legend('topleft', expression(mu %+-% sigma), bty = 'n')
+    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
+    #       col = c('black', transpgrey), lty = 1, bty = "n", lwd = 2.5)
+    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
   }
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", col.main = pal["tree"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
@@
     #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey',pal["tree"], pallight["tree"]), lty = 1, bty = 'n')
+    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
+    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
+    #       col = c('black','grey', pal["tree"], pallight["tree"]), lty = 1, bty = "n", lwd = 2.5)
+    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
   }
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", col.main = pal["forest"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
@@
     #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
-    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-           col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = 'n')
+    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
+    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
+    #       col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = "n", lwd = 2.5)
+    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
   }
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
@@
     #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
-           col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = 'n')
+    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
+    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
+    #       col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
+    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
   }
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
+    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
+    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
+    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
+
     lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
@@
     lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
-    legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
-                        TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
-           col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = 'n')
+    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
+    #                    TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
+    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
+    legend(x = -0.5, y = 13.95, expression(mu), bty = "n")
   }
@@
     par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
+    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
+    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
+    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
+    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
+
     #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
@@
     #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-    legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
-                        TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
-           col = c('black',pallight["tree"],pallight["forest"],pallight["gamlss"]), lty = 1, bty = 'n')
-  }
-}
-
-@
-
-%\bibliography{ref.bib}
+    #legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
+    #                    TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
+    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
+  legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
+  }
+
+}
+
+@
 
@@
 par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
+plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
 box(lwd=5)
@@
 par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "slategray", pch = 19)
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
 box(lwd=5)
@@
 plot(fig, inner_panel = node_inner_ext,
-     tp_args = list(FUN = identity, width = 18, fill = pal2[c(1, 3)]), 
+     tp_args = list(FUN = identity, width = 18, height = 5.002, fill = pal2[c(1, 3)]), 
      ip_args = list(fill = pal2[c(2, 3)]),
@@
 plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "grey")
-#plot(y=odata$y, x=odata$x, xaxt='n', yaxt='n', ann=FALSE, col = "slategray")
+#plot(y=odata$y, x=odata$x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray")
 lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
@@
   col = pallight["forest"], border = "transparent")
-
+legend(x = -0.071, y = 21.6, expression(mu  %+-%  sigma), bty = "n")
 #lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
@@
     plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = pal["tree"], ylim = ylim,
-         xlab = "", ylab = "", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', lwd = 2)
+         xlab = "", ylab = "", xaxt="n", yaxt="n", lwd = 2)
+    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
+    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
+    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
     lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
@@
     parsave <- par(new = TRUE,  mar = c(0,0,0,0))
-    plot(0,0,xlim=c(0,1),ylim=c(0,1),type='n',xlab='', ylab='', col='white', axes = FALSE)
+    plot(0,0,xlim=c(0,1),ylim=c(0,1),type="n",xlab='', ylab='', col='white', axes = FALSE)
 
@@
     plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = pal["tree"], ylim = ylim,
-         xlab = "", ylab = "", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', lwd = 2)
+         xlab = "", ylab = "", xaxt="n", yaxt="n", lwd = 2)
+    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
+    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
+    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
+    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
     lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
@@
     parsave <- par(new = TRUE,  mar = c(0,0,0,0))
-    plot(0,0,xlim=c(0,1),ylim=c(0,1),type='n',xlab='',ylab='',col='white', axes = FALSE)
+    plot(0,0,xlim=c(0,1),ylim=c(0,1),type="n",xlab='',ylab='',col='white', axes = FALSE)
 
@@
   plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = pal["tree"], ylim = ylim,
-       xlab = "", ylab = "", main = "disttree vs. distforest vs. gamlss", xaxt='n', yaxt='n', lwd = 2)
+       xlab = "", ylab = "", xaxt="n", yaxt="n", lwd = 2)
+  title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
+  title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
+  title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
+  title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
+  title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
+
   lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = pal["forest"], lwd = 2)
@@
   #legend('topright', c("disttree", "distforest", "gamlss"), col = c(pal["tree"], pal["forest"],pal["gamlss"]), lty = 1)
+  #mtext(text= "disttree", side = 3, col = pal["tree"], line = 1, adj = 0.3)
+  #mtext(text= "distforest", side = 3, col = pal["forest"], line = 1, adj = 0.5)
+  #mtext(text= "gamlss", side = 3, col = pal["gamlss"], line = 1, adj = 0.7)
+  #mtext(text= "vs.", side = 3, col = "black", line = 1, adj = 0.4)
+  #mtext(text= "vs.", side = 3, col = "black", line = 1, adj = 0.6)
   mtext(text= "smooth", side = 1, line = 1, adj = 0)
@@
   mtext(text= "Log-Likelihood", side = 2, line = 1)
-  mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 16)
+  mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 16.7)
   mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = -8)
@@
   parsave <- par(new = TRUE,  mar = c(0,0,0,0))
-  plot(0,0,xlim=c(0,1),ylim=c(0,1),type='n',xlab='',ylab='',col='white', axes = FALSE)
+  plot(0,0,xlim=c(0,1),ylim=c(0,1),type="n",xlab='', ylab='', col='white', axes = FALSE)
 

### 2017-06-26 pkg/disttree/inst/useR2017/slides.Rnw
distribution used in DGP now written as conditional distribution; renaming of 3rd color-vector 'pal2'->'paltrees'; 1st plot in 'motivation' starts one step later
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1491 255f4249-2263-4417-b1ea-8498779e27a7

@@
 @
-\visible<1->{
+\visible<2->{
 <<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
@@
 \endminipage
-\visible<2->{{\LARGE$\rightarrow$}}
+\visible<3->{{\LARGE$\rightarrow$}}
 \minipage{0.29\textwidth}
@@
 @
-\visible<2->{
+\visible<3->{
 <<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
@@
 \endminipage
-\visible<3->{{\LARGE$\rightarrow$}}
+\visible<4->{{\LARGE$\rightarrow$}}
 \minipage{0.29\textwidth}
@@
 @
-\visible<3->{
+\visible<4->{
 <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
@@
 \begin{center}
-\visible<1->{
+\visible<2->{
 LM, GLM\\
@@
 \begin{center}
-\visible<2->{
+\visible<3->{
 GAM\\
@@
 \begin{center}
-\visible<3->{
+\visible<4->{
 GAMLSS\\
@@
 \item For each learning observation calculate a weight by counting the number of trees in which it ends up in the same terminal node as the new observation.
-\item Estimate $\hat{\theta}$ on the wheighted learning data.
+\item Estimate $\hat{\theta}$ on the weighted learning data.
 \end{enumerate}
@@
 \begin{center} 
-DGP:  $Y \sim \mathcal{N}(\mu(X), \sigma(X))$
+DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
 \vspace*{-0.21cm}
@@
 <<plottree_dgp, fig=TRUE, echo=FALSE>>=
-pal2 <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
 plot(fig, inner_panel = node_inner_ext,
-     tp_args = list(FUN = identity, width = 18, height = 5.002, fill = pal2[c(1, 3)]), 
-     ip_args = list(fill = pal2[c(2, 3)]),
+     tp_args = list(FUN = identity, width = 18, fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]),
      drop_terminal = TRUE, tnex = 1.4)
@@
 
-pal2 <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
 
@@
 plot(tree, drop = TRUE, tnex = 1.4, FUN = FUN,
-     tp_args = list(fill = pal2[c(1, 3)], width = 18), 
-     ip_args = list(fill = pal2[c(2, 3)]))
+     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
 @
@@
 plot(as.constparty(tree), tnex = 1.4,
-     tp_args = list(fill = pal2[c(1, 3)]), 
-     ip_args = list(fill = pal2[c(2, 3)]))
+     tp_args = list(fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
 @
@@
   col = pallight["forest"], border = "transparent")
-legend(x = -0.071, y = 21.6, expression(mu  %+-%  sigma), bty = "n")
+legend(x = -0.0715, y = 21.65, expression(mu  %+-%  sigma), bty = "n")
 #lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)

### 2017-06-26 pkg/disttree/inst/useR2017/slides.Rnw
added density plot for example disttree
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1492 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-%\begin{frame}
-%\frametitle{Fitting a Tree}
-%\center
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{tree_estpar}
-%\end{frame}
-
 \begin{frame}[fragile]
@@
 <<plottree_box, fig=TRUE, echo=FALSE>>=
-plot(as.constparty(tree), tnex = 1.4,
+plot(as.constparty(tree), tnex = 1.4, drop = TRUE,
      tp_args = list(fill = paltrees[c(1, 3)]), 
@@
 
+\begin{frame}[fragile]
+\frametitle{Fitting a Tree}
+\begin{center}
+\vspace*{-0.5cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.74\linewidth}
+<<plottree_dens_h, fig=TRUE, echo=FALSE>>=
+node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
+                          main = "", xlab = "", ylab = "Density", id = TRUE,
+                          fill = "lightgray", col = "darkred", lwd = 1.5, ...) 
+{
+  ylines <- 3
+  if (is.null(xscale)) xscale <- c(-5.1,22.5)
+  if (is.null(yscale)) yscale <- c(-0.05,0.5)
+  xr <- xscale
+  yr <- yscale
+
+  if (horizontal) {
+    yyy <- xscale
+    xscale <- yscale
+    yscale <- yyy
+  }
+
+  rval <- function(node) {
+    yrange <- seq(from = -20, to = 90)/4
+    ydens <- node$info$object$ddist(yrange)
+
+    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3, 
+                                            widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")), 
+                                            heights = unit(c(1, 1), c("lines", "null"))), 
+                       width = unit(1, "npc"), 
+                       height = unit(1, "npc") - unit(2, "lines"), 
+                       name = paste("node_density",node$id, sep = ""))
+    pushViewport(top_vp)
+    grid.rect(gp = gpar(fill = "white", col = 0))
+    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
+    pushViewport(top)
+    mainlab <- paste(ifelse(id, paste("Node", node$id, "(n = "), "n = "), node$info$nobs, ifelse(id, ")", ""), sep = "")
+
+    grid.text(mainlab)
+    popViewport()
+    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2, 
+                     xscale = xscale, yscale = yscale, 
+                     name = paste("node_density",  node$id, "plot", sep = ""))
+    pushViewport(plot)
+    yd <- ydens
+    xd <- yrange
+    if (horizontal) {
+      yyy <- xd
+      xd <- yd
+      yd <- yyy
+      yyy <- xr
+      xr <- yr
+      yr <- yyy
+    }
+
+    grid.lines(x = xd, y = yd, default.units = "native", 
+               gp = gpar(col = col))
+    grid.xaxis()
+    grid.yaxis()
+    grid.rect(gp = gpar(fill = "transparent"))
+    upViewport(2)
+  }
+  return(rval)
+}
+
+class(node_density) <- "grapcon_generator"
+
+plot(tree, tnex = 1.4, drop = TRUE,
+     terminal_panel = node_density(horizontal = TRUE, col = pal["forest"], lwd = 5),
+     tp_args = list(fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Fitting a Tree}
+\begin{center}
+\vspace*{-0.5cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.74\linewidth}
+<<plottree_dens_v, fig=TRUE, echo=FALSE>>=
+plot(tree, tnex = 1.4, drop = TRUE,
+     terminal_panel = node_density(horizontal = FALSE),
+     tp_args = list(fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
 %\begin{frame}

### 2017-06-26 pkg/disttree/inst/useR2017/slides.Rnw
changed 'col' and 'lwd' in density plot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1493 255f4249-2263-4417-b1ea-8498779e27a7

@@
     grid.lines(x = xd, y = yd, default.units = "native", 
-               gp = gpar(col = col))
+               gp = gpar(col = col, lwd = lwd))
     grid.xaxis()
@@
 plot(tree, tnex = 1.4, drop = TRUE,
-     terminal_panel = node_density(horizontal = TRUE, col = pal["forest"], lwd = 5),
+     terminal_panel = node_density(horizontal = TRUE, col = pal["forest"], lwd = 2),
      tp_args = list(fill = paltrees[c(1, 3)]), 
@@
 plot(tree, tnex = 1.4, drop = TRUE,
-     terminal_panel = node_density(horizontal = FALSE),
+     terminal_panel = node_density(horizontal = FALSE, col = pal["forest"], lwd = 2),
      tp_args = list(fill = paltrees[c(1, 3)]), 

### 2017-06-26 pkg/disttree/inst/useR2017/slides.Rnw
density shading / new UIBK corporate design
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1494 255f4249-2263-4417-b1ea-8498779e27a7

@@
-\documentclass[11pt,compress,t]{beamer}
+\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
+\usetheme[nototalframenumber]{uibk}
+\setfaculty{volkswirtschaftundstatistik}{en}
+
+\title{Distributional Trees and Forests}
+\author{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
+\URL{http://R-Forge.R-project.org/projects/partykit}
+\Conference{useR! 2017}
+\date{2017-07-05}
 
 \setbeamertemplate{caption}{\insertcaption} 
-\usetheme{Z}
 %% includes a replacement for \usepackage{Sweave}
@@
 \begin{document}
-
-\title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
-\URL{https://R-Forge.R-project.org/projects/partykit/}
 
@@
       yr <- yyy
-    }
-
+      rxd <- rep(0, length(xd))
+      ryd <- rev(yd)
+    } else {
+      rxd <- rev(xd)
+      ryd <- rep(0, length(yd))
+    }
+
+    grid.polygon(x = c(xd, rxd), y = c(yd, ryd), default.units = "native",
+              gp = gpar(col = "black", fill = fill))
     grid.lines(x = xd, y = yd, default.units = "native", 
@@
 \begin{tabular}{ll}
-\code{distfit}    & Distributional fits (ML, \code{gamlss.family} or custom \code{list}).\\
+\code{distfit}    & Distributional fits (ML, \code{gamlss.family}/custom \code{list}).\\
                   & No covariates. \\

### 2017-06-27 pkg/disttree/inst/useR2017/slides.Rnw
changed size of plots and formated density plot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1495 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber]{uibk}
-\setfaculty{volkswirtschaftundstatistik}{en}
+\usetheme[nototalframenumber, noslidenumber]{uibk}
+%\setfaculty{volkswirtschaftundstatistik}{en}
 
@@
 \textbf{Distributional:} \\
-\hspace{0.3cm} Specify the complete probability distribution\\
-\hspace{0.3cm} (including location, scale and shape).
+\vspace{0.1 cm}
+\hspace{0.5cm} Specify the complete probability distribution\\
+\hspace{0.5cm} (including location, scale and shape).
 
@@
 \frametitle{Fitting a Tree}
-\vspace*{-0.5cm}
+\vspace*{-0.1cm}
 \begin{center} 
@@
 
-\setkeys{Gin}{width=0.74\linewidth}
+\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
 <<plottree_dgp, fig=TRUE, echo=FALSE>>=
@@
      ip_args = list(fill = paltrees[c(2, 3)]),
-     drop_terminal = TRUE, tnex = 1.4)
+     drop_terminal = TRUE, tnex = 1.7)
 @
@@
 \begin{center}
-\vspace*{-0.5cm}
+\vspace*{-0.1cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.74\linewidth}
+\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
 <<plottree_estpar, fig=TRUE, echo=FALSE>>=
@@
 ## plot version using FUN and tree of class 'disttree'
-plot(tree, drop = TRUE, tnex = 1.4, FUN = FUN,
+plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
      tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
@@
 \begin{center}
-\vspace*{-0.5cm}
+\vspace*{-0.1cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.74\linewidth}
+\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
 <<plottree_box, fig=TRUE, echo=FALSE>>=
-plot(as.constparty(tree), tnex = 1.4, drop = TRUE,
-     tp_args = list(fill = paltrees[c(1, 3)]), 
+plot(as.constparty(tree), tnex = 1.7, drop = TRUE,
+     tp_args = list(fill = paltrees[c(1, 3)], ylines = 1.5), 
      ip_args = list(fill = paltrees[c(2, 3)]))
@@
 \begin{center}
-\vspace*{-0.5cm}
+\vspace*{-0.1cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.74\linewidth}
-<<plottree_dens_h, fig=TRUE, echo=FALSE>>=
+\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
+<<plottree_dens, fig=TRUE, echo=FALSE>>=
 node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
-                          main = "", xlab = "", ylab = "Density", id = TRUE,
-                          fill = "lightgray", col = "darkred", lwd = 1.5, ...) 
+                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
+                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
 {
-  ylines <- 3
+  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
+  ylines <- 1.5
   if (is.null(xscale)) xscale <- c(-5.1,22.5)
@@
 
+    if (rug) {
+      nodeobs <- node$info$object$y
+      if (horizontal) {
+        grid.rect(x = xscale[1], y = nodeobs , height = 0, width = xscale[1], 
+                  default.units = "native", just = c("right", "bottom"))
+      } else {
+        grid.rect(x = nodeobs, y = yscale[1], 
+                  width = 0, height = abs(yscale[1]), default.units = "native", 
+                  just = c("center", "bottom"))
+        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
+        #           default.units = "native")
+        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
+        #           default.units = "native")
+      }
+    }
+
     grid.polygon(x = c(xd, rxd), y = c(yd, ryd), default.units = "native",
-              gp = gpar(col = "black", fill = fill))
-    grid.lines(x = xd, y = yd, default.units = "native", 
-               gp = gpar(col = col, lwd = lwd))
+              gp = gpar(col = "black", fill = fill, lwd = lwd))
+    #grid.lines(x = xd, y = yd, default.units = "native", 
+    #           gp = gpar(col = col, lwd = lwd))
     grid.xaxis()
@@
 
-plot(tree, tnex = 1.4, drop = TRUE,
-     terminal_panel = node_density(horizontal = TRUE, col = pal["forest"], lwd = 2),
-     tp_args = list(fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]))
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Fitting a Tree}
-\begin{center}
-\vspace*{-0.5cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.74\linewidth}
-<<plottree_dens_v, fig=TRUE, echo=FALSE>>=
-plot(tree, tnex = 1.4, drop = TRUE,
-     terminal_panel = node_density(horizontal = FALSE, col = pal["forest"], lwd = 2),
+plot(tree, tnex = 1.7, drop = TRUE,
+     terminal_panel = node_density,
      tp_args = list(fill = paltrees[c(1, 3)]), 
@@
 \begin{center}
-\vspace*{-1cm}
+\vspace*{-1.03cm}
 \setkeys{Gin}{width=0.8\linewidth}
@@
   col = pallight["forest"], border = "transparent")
-legend(x = -0.0715, y = 21.65, expression(mu  %+-%  sigma), bty = "n")
+legend(x = -0.0715, y = 21.62, expression(mu  %+-%  sigma), bty = "n")
 #lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)

### 2017-06-28 pkg/disttree/inst/useR2017/slides.Rnw
changed plot size of example tree
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1496 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \frametitle{Fitting a Tree}
-\vspace*{-0.1cm}
+\vspace*{-0.12cm}
 \begin{center} 
@@
 
-\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
 <<plottree_dgp, fig=TRUE, echo=FALSE>>=
@@
 \begin{center}
-\vspace*{-0.1cm}
+\vspace*{-0.12cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
 <<plottree_estpar, fig=TRUE, echo=FALSE>>=
@@
 \begin{center}
-\vspace*{-0.1cm}
+\vspace*{-0.12cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
 <<plottree_box, fig=TRUE, echo=FALSE>>=
@@
 \begin{center}
-\vspace*{-0.1cm}
+\vspace*{-0.12cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.65\linewidth, height=0.65\linewidth}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
 <<plottree_dens, fig=TRUE, echo=FALSE>>=

### 2017-06-28 pkg/disttree/inst/useR2017/slides.Rnw
semi-transparent rug
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1497 255f4249-2263-4417-b1ea-8498779e27a7

@@
         grid.rect(x = xscale[1], y = nodeobs , height = 0, width = xscale[1], 
-                  default.units = "native", just = c("right", "bottom"))
+                  default.units = "native", just = c("right", "bottom"),
+		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
       } else {
@@
                   width = 0, height = abs(yscale[1]), default.units = "native", 
-                  just = c("center", "bottom"))
+                  just = c("center", "bottom"),
+		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
         #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 

### 2017-06-28 pkg/disttree/inst/useR2017/slides.Rnw
changed yscale of the density plot and added Heidi as coauthor on the title slide
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1498 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
+\author{Lisa Schlosser, Torsten Hothorn, \\
+Heidi Seibold, Achim Zeileis}
 \URL{http://R-Forge.R-project.org/projects/partykit}
@@
   if (is.null(xscale)) xscale <- c(-5.1,22.5)
-  if (is.null(yscale)) yscale <- c(-0.05,0.5)
+  if (is.null(yscale)) yscale <- c(-0.05,0.45)
   xr <- xscale

### 2017-07-12 pkg/model4you/inst/JORS/man_model4you.Rnw
more text
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1500 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\documentclass{jors}
+\documentclass{josr}
 
@@
 \section*{Abstract}
-Typical models estimating treatment effecs assume that the treatment effect is
-the same for all individuals observed and for possible future individuals.
-Model-based recursive partitioning allows to relax this assumption.  With
+Typical models estimating treatment effects assume that the treatment effect is
+the same for all individuals.
+Model-based recursive partitioning allows to relax this assumption and allows for
+the estimation of stratified treatment effects (model-based trees) or even
+personalised treatment effects (model-based forests).  With
 model-based trees one can compute treatment effects for different strata of
@@
 characteristics of the individuals.  Model-based random forests allow for a
-similarity estimation between individuals in terms of model estimates and
-through this we can estimate personalised models.  The R package
-\emph{model4you} (available on CRAN) implements these stratified and
+similarity estimation between individuals in terms of model estimates. The similarity
+measure can then be used to estimate personalised models. 
+The R package \emph{model4you} implements these stratified and
 personalised models with a focus on ease of use and interpretability so that
@@
 }
-Studies in various fields randomly assign individuals one of two groups with
+Studies in various fields randomly assign individuals to one of two groups with
 different exposure and then measure a response. For example, in clinical trials
@@
 
-If you do not believe that all individuals have the same intercept and
-treatment effect the R package \emph{model4you} gives you two options to loosen
-up this strict model: \paragraph{(1) Model-based trees} identify subgroups
+For cases where the assumption that all individuals have the same intercept
+and treatment effect is too strict the R package \emph{model4you} offers two 
+options:
+
+\paragraph{(1) Model-based trees} identify subgroups
 where within the subgroups the model parameters are similar and between groups
@@
 clinical trial in \citep{seibold_model-based_2016}.
-
 Just a single line of code lets the user compute a model-based tree in R:
@@
 personalised models can be found in \citep{seibold_individual_2017}.
-
-In R, to get from the initial base model to personalised models for all
-observations in the training data, type:
+To compute personalised models for all observations in the training data is 
+simple:
 <<eval=FALSE>>=
@@
 
-\begin{figure}
-\centering
-\include{pmworkflow}
-\caption{Personalised models workflow: from real world to interpretable
-visualisation.}
-\end{figure}
+% \begin{figure}
+% \centering
+% \include{pmworkflow}
+% \caption{Personalised models workflow: from real world to interpretable
+% visualisation.}
+% \end{figure}
 
@@
 and personalised models. For this we need to load the package and -- to ensure
-reproducibility -- set a random seed. Also for plots we will need the
-\emph{ggplot2} package and we use a black and white theme.
+reproducibility -- set a random seed. Also for plots we will need packages
+\emph{ggplot2} and \emph{gridExtra}.
 <<prep>>=
@@
 theme_set(theme_classic())
+library("gridExtra")
 @
@@
 \paragraph{Mathematics exam analysis}
+In 2014 first-year business and economics students at the University of Innsbruck
+were divided into two examination groups. Group 1 wrote the exam in the morning and
+group 2 started after the first group finished.
+
 <<math_data>>=
@@
 lmplot(bmod_math, densest = TRUE)
+
+## extract coefficients and confidence intervals
+coefs <- coef(bmod_math)
+ci <- confint(bmod_math)
+coefs <- cbind(coefs, ci)
+
+## prepare for plotting
+digits <- 2
+cf <- format(round(coefs, digits), nsmall = digits)
+colnams <- colnames(cf)
+colnams[1] <- "theta"
+gridExtra::grid.table(
+  cf, cols = colnams,
+  theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
+  vp = viewport(x = 0.5, y = 0, width = 0.7, height = 0.4,
+                just = c("center", "bottom"))
+)
 @
@@
 
-\paragraph{Analysis of the German Breast Cancer Study Group 2}
-
-<<GBSG2_prep>>=
-library("survival")
-data("GBSG2", package = "TH.data")
-@
-
-<<GBSG2_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for GBSG2 data.">>=
-bmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, model = TRUE)
-survplot(bmod)
-
-## extract coefficients and confidence intervals
-coefs <- coef(bmod)
-ci <- confint(bmod)
-coefs <- cbind(coefs, ci)
-
-## prepare for plotting
-digits <- 2
-cf <- format(round(coefs, digits), nsmall = digits)
-colnams <- colnames(cf)
-colnams[1] <- "theta"
-grid.table(cf, cols = colnams,
-           theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
-           vp = viewport(x = 0, y = 0, 
-                         width = 0.7, height = 0.4,
-                         just = c("left", "bottom")))
-@
-
-<<GBSG2_tree, fig.width = 9, fig.cap = "Personalised model tree for GBSG2 data.">>=
-tr <- pmtree(bmod)
-plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survplot, 
-                                          confint = TRUE))
-@
-
-<<GBSG2_forest>>=
-frst <- pmforest(bmod)
-@
-
-<<GBSG_pmod1>>=
-newpatient_i <- data.frame(age = 60, menostat = "Post", tsize = 20, tgrade = "II",
-                           pnodes = 3, progrec = 50, estrec = 60)
-mod_i <- pmodel(x = frst, fun = identity, newdata = newpatient_i)
-@
-
-<<GBSG2_pmod>>=
-coeffun <- function(model) {
-  ## model coefficients
-  coefs <- c(coef(model), scale = model$scale)
-
-  ## difference in median survival 
-  p = 0.5
-  coefs["median_s0"] <- qweibull(p = p, shape = 1/coefs["scale"], 
-                                 scale = exp(coefs["(Intercept)"]))
-  coefs["median_s1"] <- qweibull(p = p, shape = 1/coefs["scale"], 
-                                 scale = exp(coefs["(Intercept)"] + coefs[2]))
-  coefs["median_sdiff"] <- coefs["median_s1"] - coefs["median_s0"]
-
-  return(coefs)
-}
-coefs <- pmodel(x = frst, fun = coeffun, OOB = TRUE)
-
-## add to data and plot
-dpdat <- cbind(coefs, GBSG2)
-znam <- names(GBSG2)[!(names(GBSG2) %in% c("horTh", "time", "cens"))]
-
-dplot <- function(z) {
-  p <- ggplot(dpdat, aes_string(x = z, y = "median_sdiff")) 
-  if(is.factor(dpdat[, z])) {
-    p + geom_boxplot()
-  } else {
-    p + geom_point(alpha = 0.2) + 
-      geom_smooth(fill = NA, method = "loess")
-  }
-}
-dps <- lapply(znam, dplot)
-do.call(grid.arrange, dps)
-
-do.call(grid.arrange, lapply(dps, function(p) {
-  dmin <- min(dpdat$median_sdiff)
-  dq <- quantile(dpdat$median_sdiff, 0.99)
-  p + coord_cartesian(ylim = c(dmin, dq))
-}))
-@
+% \paragraph{Analysis of the German Breast Cancer Study Group 2}
+% 
+% <<GBSG2_prep>>=
+% library("survival")
+% data("GBSG2", package = "TH.data")
+% @
+% 
+% <<GBSG2_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for GBSG2 data.">>=
+% bmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, model = TRUE)
+% survplot(bmod)
+% 
+% ## extract coefficients and confidence intervals
+% coefs <- coef(bmod)
+% ci <- confint(bmod)
+% coefs <- cbind(coefs, ci)
+% 
+% ## prepare for plotting
+% digits <- 2
+% cf <- format(round(coefs, digits), nsmall = digits)
+% colnams <- colnames(cf)
+% colnams[1] <- "theta"
+% gridExtra::grid.table(
+%   cf, cols = colnams,
+%   theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
+%   vp = viewport(x = 0, y = 0, width = 0.7, height = 0.4,
+%                 just = c("left", "bottom"))
+% )
+% @
+% 
+% <<GBSG2_tree, fig.width = 9, fig.cap = "Personalised model tree for GBSG2 data.">>=
+% tr <- pmtree(bmod)
+% plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survplot, 
+%                                           confint = TRUE))
+% @
+% 
+% <<GBSG2_forest>>=
+% frst <- pmforest(bmod)
+% @
+% 
+% <<GBSG_pmod1>>=
+% newpatient_i <- data.frame(age = 60, menostat = "Post", tsize = 20, tgrade = "II",
+%                            pnodes = 3, progrec = 50, estrec = 60)
+% mod_i <- pmodel(x = frst, fun = identity, newdata = newpatient_i)
+% @
+% 
+% <<GBSG2_pmod>>=
+% coeffun <- function(model) {
+%   ## model coefficients
+%   coefs <- c(coef(model), scale = model$scale)
+%   
+%   ## difference in median survival 
+%   p = 0.5
+%   coefs["median_s0"] <- qweibull(p = p, shape = 1/coefs["scale"], 
+%                                  scale = exp(coefs["(Intercept)"]))
+%   coefs["median_s1"] <- qweibull(p = p, shape = 1/coefs["scale"], 
+%                                  scale = exp(coefs["(Intercept)"] + coefs[2]))
+%   coefs["median_sdiff"] <- coefs["median_s1"] - coefs["median_s0"]
+%   
+%   return(coefs)
+% }
+% coefs <- pmodel(x = frst, fun = coeffun, OOB = TRUE)
+% 
+% ## add to data and plot
+% dpdat <- cbind(coefs, GBSG2)
+% znam <- names(GBSG2)[!(names(GBSG2) %in% c("horTh", "time", "cens"))]
+% 
+% dplot <- function(z) {
+%   p <- ggplot(dpdat, aes_string(x = z, y = "median_sdiff")) 
+%   if(is.factor(dpdat[, z])) {
+%     p + geom_boxplot()
+%   } else {
+%     p + geom_point(alpha = 0.2) + 
+%       geom_smooth(fill = NA, method = "loess")
+%   }
+% }
+% dps <- lapply(znam, dplot)
+% do.call(grid.arrange, dps)
+% 
+% do.call(grid.arrange, lapply(dps, function(p) {
+%   dmin <- min(dpdat$median_sdiff)
+%   dq <- quantile(dpdat$median_sdiff, 0.99)
+%   p + coord_cartesian(ylim = c(dmin, dq))
+% }))
+% @
 
 \section*{Implementation and architecture}
-
-Based on partykit, merely convenience interface.
-
 \begin{figure}
@@
 \include{treefamily}
-\caption{The family of model-based trees and forests.}
+\caption{The family of model-based trees and forests. Square nodes represent
+functions, round nodes represent function arguments. Solid lines represent 
+direct usage, dashed lines indirect usage (iteration between tree and model 
+estimation), and dotted lines represent close similarity but no dependency.}
+\label{fig:family}
 \end{figure}
+
+The R package \emph{model4you} is based on the \emph{partykit} package. In fact, 
+everything that can be done with \emph{model4you} can already be done with 
+\emph{partykit} (version $\geq$ 1.2-0). \emph{model4you} is merely a convenience 
+interface and focusses on ease of use and interpretability. Various other packages
+depend on \emph{partykit}. Figure~\ref{fig:family} shows all packages depending
+on \emph{partykit} that deal with model-based trees and forests. Functions 
+\Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::mob()'} provide the basis for all
+others. Note that \Sexpr{'partykit::ctree()'} computes conditional inference trees
+but by specifying argument \Sexpr{'ytrafo'} appropriately, it can also compute
+model-based trees. This functionality is used for packages \emph{disttree} 
+(partitioning of Generalized Additive Models for Location, Scale and Shape),
+\emph{trtf} (partitioning of transformation models) and \emph{model4you}. 
+Many of the packages shown in Figure~\ref{fig:family} are available on CRAN. 
+The remaining packages are under development and available on 
+\url{https://r-forge.r-project.org/projects/partykit/}.
 

### 2017-07-13 pkg/model4you/inst/JORS/man_model4you.Rnw
small stuff
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1504 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Typical models estimating treatment effects assume that the treatment effect is
-the same for all individuals.
-Model-based recursive partitioning allows to relax this assumption and allows for
-the estimation of stratified treatment effects (model-based trees) or even
-personalised treatment effects (model-based forests).  With
-model-based trees one can compute treatment effects for different strata of
-individuals. The strata are found in a data driven fashion and depend on
-characteristics of the individuals.  Model-based random forests allow for a
-similarity estimation between individuals in terms of model estimates. The similarity
-measure can then be used to estimate personalised models. 
-The R package \emph{model4you} implements these stratified and
+the same for all individuals.  Model-based recursive partitioning allows to
+relax this assumption and allows for the estimation of stratified treatment
+effects (model-based trees) or even personalised treatment effects (model-based
+forests).  With model-based trees one can compute treatment effects for
+different strata of individuals. The strata are found in a data driven fashion
+and depend on characteristics of the individuals.  Model-based random forests
+allow for a similarity estimation between individuals in terms of model
+estimates. The similarity measure can then be used to estimate personalised
+models.  The R package \emph{model4you} implements these stratified and
 personalised models with a focus on ease of use and interpretability so that
@@
 
-For cases where the assumption that all individuals have the same intercept
-and treatment effect is too strict the R package \emph{model4you} offers two 
+For cases where the assumption that all individuals have the same intercept and
+treatment effect is too strict the R package \emph{model4you} offers two
 options:
 
-\paragraph{(1) Model-based trees} identify subgroups
-where within the subgroups the model parameters are similar and between groups
-the model parameters are different. This is achieved by finding instabilities
-in the model parameters with respect to a variable (characteristicum) and
-recursively partitioning the data into groups. If, for example the algorithm
-finds that men and women have differing treatment effects, the data is
-partitioned into two subgroups. Details on model-based trees in general can be
-found in \citep{zeileis_model-based_2008} and for the special use case in
-clinical trial in \citep{seibold_model-based_2016}.
-Just a single line of code lets the user compute a model-based tree in R:
+\paragraph{(1) Model-based trees} identify subgroups where within the subgroups
+the model parameters are similar and between groups the model parameters are
+different. This is achieved by finding instabilities in the model parameters
+with respect to a variable (characteristicum) and recursively partitioning the
+data into groups. If, for example the algorithm finds that men and women have
+differing treatment effects, the data is partitioned into two subgroups.
+Details on model-based trees in general can be found in
+\citep{zeileis_model-based_2008} and for the special use case in clinical trial
+in \citep{seibold_model-based_2016}.  Just a single line of code lets the user
+compute a model-based tree in R:
 <<eval=FALSE>>=
@@
 data, where the similarity measure corresponds to the weight. Details on the
-personalised models can be found in \citep{seibold_individual_2017}.
-To compute personalised models for all observations in the training data is 
+personalised models can be found in \citep{seibold_individual_2017}.  To
+compute personalised models for all observations in the training data is
 simple:
@@
 
-The R package \emph{model4you} is based on the \emph{partykit} package. In fact, 
-everything that can be done with \emph{model4you} can already be done with 
-\emph{partykit} (version $\geq$ 1.2-0). \emph{model4you} is merely a convenience 
-interface and focusses on ease of use and interpretability. Various other packages
-depend on \emph{partykit}. Figure~\ref{fig:family} shows all packages depending
-on \emph{partykit} that deal with model-based trees and forests. Functions 
-\Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::mob()'} provide the basis for all
-others. Note that \Sexpr{'partykit::ctree()'} computes conditional inference trees
-but by specifying argument \Sexpr{'ytrafo'} appropriately, it can also compute
-model-based trees. This functionality is used for packages \emph{disttree} 
-(partitioning of Generalized Additive Models for Location, Scale and Shape),
-\emph{trtf} (partitioning of transformation models) and \emph{model4you}. 
-Many of the packages shown in Figure~\ref{fig:family} are available on CRAN. 
-The remaining packages are under development and available on 
+The R package \emph{model4you} is based on the \emph{partykit} package. In
+fact, everything that can be done with \emph{model4you} can already be done
+with \emph{partykit} (version $\geq$ 1.2-0). \emph{model4you} is merely a
+convenience interface and focusses on ease of use and interpretability. Various
+other packages depend on \emph{partykit}. Figure~\ref{fig:family} shows all
+packages depending on \emph{partykit} that deal with model-based trees and
+forests. Functions \Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::mob()'}
+provide the basis for all others. Note that \Sexpr{'partykit::ctree()'}
+computes conditional inference trees but by specifying argument
+\Sexpr{'ytrafo'} appropriately, it can also compute model-based trees. This
+functionality is used for packages \emph{disttree} (partitioning of Generalized
+Additive Models for Location, Scale and Shape), \emph{trtf} (partitioning of
+transformation models) and \emph{model4you}.  Many of the packages shown in
+Figure~\ref{fig:family} are available on CRAN.  The remaining packages are
+under development and available on
 \url{https://r-forge.r-project.org/projects/partykit/}.

### 2017-07-13 pkg/model4you/inst/JORS/man_model4you.Rnw
reuse potential
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1505 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \section*{Quality control}
-
-Tests, code in this manual, examples in documentation
+All packages on CRAN undergo standard checks for compatibility with the R
+package ecosystem.  The R package contains examples and tests. These were run
+and checked on Linux 86\_64 and Windows.
 
@@
 
-The software is intentionally written to make usage as simple as possible. We
-hope that researchers use it for analysis of:
-\begin{itemize}
-\item Clinical trials
-\item A-B testing
-\item Other situations where the comparison of two decisions is of interest
-\end{itemize}
-We encourage users to use the partykit forum on R-forge
-(\url{https://r-forge.r-project.org/forum/forum.php?forum_id=852}) or the party
-tag on Stackoverflow (\url{http://stackoverflow.com/questions/tagged/party}) in
-case of questions or problems.
+The software is intentionally written to make usage as simple as possible. The
+most prominent use case are clinical trials where the assumption of an average
+treatment effect for all patients is too strict and the efficacy of the
+treatment depends on patient characteristics (e.g. gender, biomarkers, etc.).
+For subgroup analyses (stratified treatment effects) model-based trees
+(\Sexpr{'pmtree()'}) can be used; For personalised treatment effects
+model-based forests (\Sexpr{'pmforest()'}) provide a way of estimating
+similarity between patients and using this similarity measure to estimate
+personalised models (\Sexpr{'pmodel()'}).  The target audience are people who
+deal with heterogeneous treatment effects, such as medical researchers,
+pharmaceutical companies or analysts in marketing (A-B testing).  In general
+the software is useful to researchers dealing with scenarios where two
+exposures are compared and responses of subjects possibly depend on other
+variables.
+
+We encourage users to use the party tag on Stackoverflow
+(\url{http://stackoverflow.com/questions/tagged/party}) in case of questions or
+problems.
 

### 2017-07-14 pkg/model4you/inst/JORS/man_model4you.Rnw
start vignette and delete GBSG2 from manuscript
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1506 255f4249-2263-4417-b1ea-8498779e27a7

@@
 library("knitr")
-knitr::opts_chunk$set(cache = TRUE, message = FALSE, fig.align = "center")
+knitr::opts_chunk$set(cache = FALSE, message = FALSE, fig.align = "center")
 
@@
 
-<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for GBSG2 data.">>=  
+<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for math exam.">>=  
 ## compute tree
@@
 @
-
-% \paragraph{Analysis of the German Breast Cancer Study Group 2}
-% 
-% <<GBSG2_prep>>=
-% library("survival")
-% data("GBSG2", package = "TH.data")
-% @
-% 
-% <<GBSG2_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for GBSG2 data.">>=
-% bmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, model = TRUE)
-% survplot(bmod)
-% 
-% ## extract coefficients and confidence intervals
-% coefs <- coef(bmod)
-% ci <- confint(bmod)
-% coefs <- cbind(coefs, ci)
-% 
-% ## prepare for plotting
-% digits <- 2
-% cf <- format(round(coefs, digits), nsmall = digits)
-% colnams <- colnames(cf)
-% colnams[1] <- "theta"
-% gridExtra::grid.table(
-%   cf, cols = colnams,
-%   theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
-%   vp = viewport(x = 0, y = 0, width = 0.7, height = 0.4,
-%                 just = c("left", "bottom"))
-% )
-% @
-% 
-% <<GBSG2_tree, fig.width = 9, fig.cap = "Personalised model tree for GBSG2 data.">>=
-% tr <- pmtree(bmod)
-% plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survplot, 
-%                                           confint = TRUE))
-% @
-% 
-% <<GBSG2_forest>>=
-% frst <- pmforest(bmod)
-% @
-% 
-% <<GBSG_pmod1>>=
-% newpatient_i <- data.frame(age = 60, menostat = "Post", tsize = 20, tgrade = "II",
-%                            pnodes = 3, progrec = 50, estrec = 60)
-% mod_i <- pmodel(x = frst, fun = identity, newdata = newpatient_i)
-% @
-% 
-% <<GBSG2_pmod>>=
-% coeffun <- function(model) {
-%   ## model coefficients
-%   coefs <- c(coef(model), scale = model$scale)
-%   
-%   ## difference in median survival 
-%   p = 0.5
-%   coefs["median_s0"] <- qweibull(p = p, shape = 1/coefs["scale"], 
-%                                  scale = exp(coefs["(Intercept)"]))
-%   coefs["median_s1"] <- qweibull(p = p, shape = 1/coefs["scale"], 
-%                                  scale = exp(coefs["(Intercept)"] + coefs[2]))
-%   coefs["median_sdiff"] <- coefs["median_s1"] - coefs["median_s0"]
-%   
-%   return(coefs)
-% }
-% coefs <- pmodel(x = frst, fun = coeffun, OOB = TRUE)
-% 
-% ## add to data and plot
-% dpdat <- cbind(coefs, GBSG2)
-% znam <- names(GBSG2)[!(names(GBSG2) %in% c("horTh", "time", "cens"))]
-% 
-% dplot <- function(z) {
-%   p <- ggplot(dpdat, aes_string(x = z, y = "median_sdiff")) 
-%   if(is.factor(dpdat[, z])) {
-%     p + geom_boxplot()
-%   } else {
-%     p + geom_point(alpha = 0.2) + 
-%       geom_smooth(fill = NA, method = "loess")
-%   }
-% }
-% dps <- lapply(znam, dplot)
-% do.call(grid.arrange, dps)
-% 
-% do.call(grid.arrange, lapply(dps, function(p) {
-%   dmin <- min(dpdat$median_sdiff)
-%   dq <- quantile(dpdat$median_sdiff, 0.99)
-%   p + coord_cartesian(ylim = c(dmin, dq))
-% }))
-% @
 

### 2017-07-19 pkg/model4you/inst/JORS/man_model4you.Rnw
first draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1507 255f4249-2263-4417-b1ea-8498779e27a7

@@
 library("knitr")
-knitr::opts_chunk$set(cache = FALSE, message = FALSE, fig.align = "center")
+knitr::opts_chunk$set(cache = TRUE, message = FALSE, fig.align = "center")
 
@@
 the same for all individuals.  Model-based recursive partitioning allows to
-relax this assumption and allows for the estimation of stratified treatment
-effects (model-based trees) or even personalised treatment effects (model-based
-forests).  With model-based trees one can compute treatment effects for
-different strata of individuals. The strata are found in a data driven fashion
-and depend on characteristics of the individuals.  Model-based random forests
-allow for a similarity estimation between individuals in terms of model
-estimates. The similarity measure can then be used to estimate personalised
-models.  The R package \emph{model4you} implements these stratified and
-personalised models with a focus on ease of use and interpretability so that
-clinicians and other users can take the model they usually use for the
-estimation of the average treatment effect and with a few lines of code get a
-visualisation that is easy to understand and interpret.
+relax this assumption and to estimate stratified treatment effects (model-based
+trees) or even personalised treatment effects (model-based forests).  With
+model-based trees one can compute treatment effects for different strata of
+individuals. The strata are found in a data driven fashion and depend on
+characteristics of the individuals.  Model-based random forests allow for a
+similarity estimation between individuals in terms of model parameters (e.g.\
+intercept and treatment effect). The similarity measure can then be used to
+estimate personalised models.  The R package \emph{model4you} implements these
+stratified and personalised models with a focus on ease of use and
+interpretability so that clinicians and other users can take the model they
+usually use for the estimation of the average treatment effect and with a few
+lines of code get a visualisation that is easy to understand and interpret.
 
@@
 Details on model-based trees in general can be found in
-\citep{zeileis_model-based_2008} and for the special use case in clinical trial
-in \citep{seibold_model-based_2016}.  Just a single line of code lets the user
-compute a model-based tree in R:
+\citep{zeileis_model-based_2008} and for the special use case for stratified
+treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
+line of code lets the user compute a model-based tree in R:
 <<eval=FALSE>>=
@@
 data, where the similarity measure corresponds to the weight. Details on the
-personalised models can be found in \citep{seibold_individual_2017}.  To
-compute personalised models for all observations in the training data is
+personalised models can be found in \citep{seibold_individual_2017}.  
+Computing personalised models for all observations in the training data is
 simple:
@@
 
-In the following we will present two example applications for model-based trees
+In the following we will present an example application for model-based trees
 and personalised models. For this we need to load the package and -- to ensure
-reproducibility -- set a random seed. Also for plots we will need packages
-\emph{ggplot2} and \emph{gridExtra}.
+reproducibility -- set a random seed. Also for visualisations we need packages
+\emph{ggplot2}, \emph{ggbeeswarm} and \emph{gridExtra}.
 <<prep>>=
@@
 theme_set(theme_classic())
+library("ggbeeswarm")
 library("gridExtra")
@@
 
-\paragraph{Mathematics exam analysis}
+\paragraph{Mathematics exam analysis:}
 In 2014 first-year business and economics students at the University of Innsbruck
 were divided into two examination groups. Group 1 wrote the exam in the morning and
-group 2 started after the first group finished.
+group 2 started after the first group finished. The exams for the two groups were 
+slightly different. Translated into clinical trial language, the exam groups are
+the treatment groups and the two different exams are the treatments.
+The data can be accessed and prepared as follows:
 
@@
 
-<<math_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
-## compute base model
+To investigate whether the exam was fair, we asses whether the two groups
+differ in terms of the percentage of correctly answered exam questions. This
+can be done using a simple linear model regressing the percentage points of
+correct anwers on the exam group. 
+<<math_bmod>>=
 bmod_math <- lm(pcorrect ~ group, data = MathExam)
-lmplot(bmod_math, densest = TRUE)
-
-## extract coefficients and confidence intervals
-coefs <- coef(bmod_math)
-ci <- confint(bmod_math)
-coefs <- cbind(coefs, ci)
-
-## prepare for plotting
-digits <- 2
-cf <- format(round(coefs, digits), nsmall = digits)
-colnams <- colnames(cf)
-colnams[1] <- "theta"
-gridExtra::grid.table(
-  cf, cols = colnams,
-  theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
-  vp = viewport(x = 0.5, y = 0, width = 0.7, height = 0.4,
-                just = c("center", "bottom"))
-)
 @
 
-<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for math exam.">>=  
-## compute tree
-(tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2)))
-plot(tr_math, terminal_panel = node_pmterminal(tr_math, 
-                                               plotfun = lmplot, 
-                                               confint = TRUE))
+The model can be visualised by plotting the estimated densities
+(see Figure~\ref{fig:math_bmod_vis}):
+<<math_bmod_vis, fig.height = 3, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
+lmplot(bmod_math)
 @
 
-<<math_methods>>=  
-## predict
-newdat <- MathExam[1:5, ]
-
-# terminal nodes
-(nodes <- predict(tr_math, type = "node", newdata = newdat))
-
-# response
-(pr <- predict(tr_math, type = "pass", newdata = newdat))
-
-# response including confidence intervals, see ?predict.lm
-(pr1 <- predict(tr_math, type = "pass", newdata = newdat,
-                predict_args = list(interval = "confidence")))
+A tree based on this model can be computed and visualised in only two lines of
+code:
+<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for math exam.">>=  
+tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2))
+plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lmplot))
 @
 
+The tree (see Figure~\ref{fig:math_tree}) divides students based on the
+percentage of successful online \Sexpr{'tests'}. These online tests were
+conducted biweekly thrughout the semester.  The largest difference between the
+two exam groups is in the students who did very well in the online tests (more
+than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
+tree thus gives us much more information on the fairness of the exam than the
+simple linear model, which is that it does not seem to be fair for students who
+did very well thrughout the semester (at this point we should state that the
+students self selected into the two exam groups which might also be the reason
+for differences in exam performance).
+
+Estimating personalised models is almost as simple as the stratified models:
 <<math_pmodl>>=
 forest_math <- pmforest(bmod_math)
-pmod_math <- pmodel(forest_math)
-head(pmod_math)
+pmods_math <- pmodel(forest_math)
+
+## model parameters of first 6 students
+head(pmods_math)
 @
 
-<<math_dp1, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
-dpdat_math <- cbind(pmod_math, MathExam)
+Dependence plots with the group effect (treatment effect) on the y-achsis and the
+student characteristics on the x-axis are a good way of visualising the personalised 
+models and for getting knowledge about the interactions between student characteristics
+and the treatment. Since the percentage of successfull online tests is measured on a
+grid, a bee plot possibly shows the relationships even better than the scatter plot
+(both shown in Figure~\ref{fig:math_dp1}).
+<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
+dpdat_math <- cbind(pmods_math, MathExam)
 
-ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
+ggplot(dpdat_math, aes(x = tests, y = group2)) +
   geom_point(alpha = 0.2, size = 1) + 
@@
 
-library("ggbeeswarm")
 ggplot(dpdat_math, aes(x = tests, y = group2, color = tests)) +
@@
 
-<<math_dp2, fig.width=4, fig.height=4, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=
+For the number of previous attempts to pass the exam and the gender box plots, 
+bee plots or a combination thereof can be used (Figure~\ref{fig:math_dp2}).
+<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=
+
 ggplot(dpdat_math, aes(x = attempt, y = group2, color = attempt)) +
@@
 ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
+  geom_boxplot() +
   geom_quasirandom(alpha = 0.5) +
@@
 @
+
+With the tools provided by the \emph{model4you} package it is
+very simple to create understandable stratified and personalised models and 
+compelling visualisations that can be used to communicate theses models.
 
@@
 
-\textcolor{blue}{How the software was implemented, with details of the architecture where relevant.
+%\textcolor{blue}{How the software was implemented, with details of the architecture where relevant.
 Use of relevant diagrams is appropriate.
@@
 
-\textcolor{blue}{Detail the level of testing that has been carried out on the code (e.g.
+%\textcolor{blue}{Detail the level of testing that has been carried out on the code (e.g.
 unit, functional, load etc.), and in which environments.
@@
 Should work on all operating systems that run R.
-\textcolor{blue}{Please include minimum version compatibility.}
+%\textcolor{blue}{Please include minimum version compatibility.}
 
@@
 None.
-\textcolor{blue}{E.g. memory, disk space, processor, input devices, output devices.}
+%\textcolor{blue}{E.g. memory, disk space, processor, input devices, output devices.}
 
@@
 
-{\bf Archive} \textcolor{blue}{(e.g.
+{\bf Archive} %\textcolor{blue}{(e.g.
 institutional repository, general repository) (required ???
@@
 
-{\bf Code repository} \textcolor{blue}{(e.g. SourceForge, GitHub etc.) (required)}
+{\bf Code repository} %\textcolor{blue}{(e.g. SourceForge, GitHub etc.) (required)}
 
@@
 \item[Licence:] GPL-2
-\item[Date published:] 
+\item[Date published:] \textcolor{blue}{dd/mm/yy} %TODO
 \end{description}
@@
 
-\textcolor{blue}{Please describe in as much detail as possible the ways in which the software could be reused by other researchers both within and outside of your field.
+%\textcolor{blue}{Please describe in as much detail as possible the ways in which the software could be reused by other researchers both within and outside of your field.
 This should include the use cases for the software, and also details of how the software might be modified or extended (including how contributors should contact you) if appropriate.
@@
 \section*{Acknowledgements}
+%TODO
 
-\textcolor{blue}{Please add any relevant acknowledgements to anyone else who supported the project in which the software was created, but did not work directly on the software itself.}
+%\textcolor{blue}{Please add any relevant acknowledgements to anyone else who supported the project in which the software was created, but did not work directly on the software itself.}
 
@@
 
-\textcolor{blue}{If the software resulted from funded research please give the funder and grant number.}
+%\textcolor{blue}{If the software resulted from funded research please give the funder and grant number.}
 

### 2017-08-14 pkg/model4you/R/plot-pmtree.R
vignette with knitr
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1528 255f4249-2263-4417-b1ea-8498779e27a7

@@ -23,7 +23,8 @@
 #' 
 #' @importFrom ggplot2 ggplot geom_line theme_classic aes_string xlim xlab scale_linetype_discrete
 #' @export
-lmplot <- function(mod, data = NULL, densest = FALSE, theme = theme_classic()) {
+lmplot <- function(mod, data = NULL, densest = FALSE, theme = theme_classic(),
+                   yrange = NULL) {
   
   ## get formula and data
   modcall <- getCall(mod)
@@ -34,7 +35,7 @@ lmplot <- function(mod, data = NULL, densest = FALSE, theme = theme_classic()) {
   xdat <- unique(get_all_vars(xformula, data = data))
   ydat <- get_all_vars(yformula, data = data)
   ynam <- names(ydat) # as.character(yformula[[2]])
-  yrange <- range(ydat)
+  if(is.null(yrange)) yrange <- range(ydat)
   
   ## get density functions for each treatment group
   k <- 50
@@ -84,7 +85,8 @@ lmplot <- function(mod, data = NULL, densest = FALSE, theme = theme_classic()) {
 #' @importFrom survival Surv survreg
 #' @importFrom Formula as.Formula 
 #' @export
-survplot <- function(mod, data = NULL, theme = theme_classic()) {
+survplot <- function(mod, data = NULL, theme = theme_classic(),
+                     yrange = NULL) {
   
   ## get formula and data
   modcall <- getCall(mod)
@@ -93,7 +95,10 @@ survplot <- function(mod, data = NULL, theme = theme_classic()) {
   yformula <- formula(modformula, lhs = 1, rhs = 0)
   if(is.null(data)) data <- eval(modcall$data)
   xdat <- unique(get_all_vars(xformula, data = data))
-  ymax <- max(model.frame(yformula, data = data))
+  if(is.null(yrange)) {
+    ymax <- max(model.frame(yformula, data = data))
+    yrange <- c(0, ymax)
+  }
   
   ## get survivor functions for each treatment group
   p <- seq(.01, .99, by=.02)
@@ -109,7 +114,7 @@ survplot <- function(mod, data = NULL, theme = theme_classic()) {
   xnam <- attr(terms(xformula), "term.labels")
   ggplot(data = pr, aes_string(x = "pr", y = "probability", group = xnam, 
                                color = xnam)) + 
-    geom_line() + coord_cartesian(xlim = c(0, ymax)) + 
[…]

### 2017-08-14 pkg/model4you/R/pmtree.R

@@ -1,4 +1,4 @@
-#' Compute model-based forest from model.
+#' Compute model-based tree from model.
 #'
 #' Input a parametric model and get a forest.
 #'

### 2017-08-18 pkg/model4you/inst/JORS/man_model4you.Rnw
update plot fcts and add plot for logistic regression
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1534 255f4249-2263-4417-b1ea-8498779e27a7

@@
 The model can be visualised by plotting the estimated densities
-(see Figure~\ref{fig:math_bmod_vis}):
-<<math_bmod_vis, fig.height = 3, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
-lmplot(bmod_math)
+(see Figure~\ref{fig:math_bmod_vis} (a)):
+<<eval=FALSE>>=  
+lm_plot(bmod_math)
+@
+
+As the plot is a \emph{ggplot2} graphic it can be maniulated with the usual
+\emph{ggplot2} layers, e.g.\ with
+<<eval=FALSE>>=
+lm_plot(bmod_math) + labs(y = "Estimated Density", x = "% Correct")
+@
+or equivalently with 
+<<eval=FALSE>>=
+myplot <- function(model, ...) {
+  lm_plot(model, ...) + 
+    labs(y = "Estimated Density", x = "% Correct")
+}
+myplot(bmod_math)
+@
+The result is shown in Figure~\ref{fig:math_bmod_vis} (b).
+<<math_bmod_vis, echo=FALSE, fig.cap = "Base model for Mathematics Exam data.", fig.subcap=c("Figure made with \\texttt{lm\\_plot()}.", "Figure made with manipulated \\texttt{lm\\_plot()}."), out.width="0.45\\textwidth", fig.height=3, fig.width=5>>=
+lm_plot(bmod_math)
+myplot <- function(model, ...) {
+  lm_plot(model, ...) + 
+    labs(y = "Estimated Density", x = "% Correct")
+}
+myplot(bmod_math)
 @
@@
 tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2))
-plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lmplot))
+plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = myplot))
 @

### 2017-08-18 pkg/model4you/tests/test-pmtree.R

@@ -7,11 +7,11 @@ data(GBSG2, package = "TH.data")
 
 ## base model
 bmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, model = TRUE)
-survplot(bmod)
+survreg_plot(bmod)
 
 ## partitioned model
 tr <- pmtree(bmod)
-plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survplot, 
+plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survreg_plot, 
                                           confint = TRUE))
 summary(tr)
 summary(tr, node = 1:2)

### 2017-08-30 pkg/disttree/inst/draft/draft.Rnw
first draft with simulation in a separated R file 'gensim.R' and one example of the results in 'simres.rda'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1546 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nojss]{jss}

%\VignetteIndexEntry{ctree: Conditional Inference Trees}
%\VignetteDepends{coin, TH.data, survival, strucchange}
%\VignetteKeywords{conditional inference, non-parametric models, recursive partitioning}
%\VignettePackage{partykit}

%% packages
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{rotating}
\usepackage{caption}
\captionsetup{format=hang}
\usepackage{subcaption}
\usepackage{Sweave}
\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
%\usepackage[pagebackref=true]{hyperref}
%\usepackage{doi}

%% commands
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

\hyphenation{Qua-dra-tic}

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("ggplot2")
theme_set(theme_bw(base_size = 18))
library("disttree")
library("Formula")  # FIX ME: should be imported in disttree
library("gamlss")
library("randomForest")
library("lattice")
library("crch")
library("latex2exp")
library("parallel")
library("gamlss.cens")
gen.cens(NO, type = "left")
library("mgcv")
library("MASS")

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 30, 90, 180), 100, 50)
names(pal) <- c("forest", "tree", "gamlss", "randomForest", "bamlss", "gamboostLSS", "cforest")

pallight <- hcl(c(10, 128, 260, 290, 30, 90, 180), 100, 50, alpha = 0.25)
names(pallight) <- c("forest", "tree", "gamlss", "randomForest", "bamlss", "gamboostLSS", "cforest")

transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)

## define distribution list:
# dist_list_normal
{

  dist_list_normal <- list()

  parnames <- c("mu", "sigma")
  etanames <- c("mu", "log(sigma)")

  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {

    val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
    if(!log) val <- exp(val)

    # par <- c(eta[1], exp(eta[2]))
    # val <- dnorm(y, mean = par[1], sd = par[2], log = log)

    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y))
      val <- sum(weights * val, na.rm = TRUE)
    }
    return(val)
  }

  sdist <- function(y, eta, weights = NULL, sum = FALSE) {

    score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
                   -1 + exp(-2*eta[2] + log((y-eta[1])^2)))

    # par <- c(eta[1], exp(eta[2])) 
    # score <- cbind(1/par[2]^2 * (y-par[1]), 
    #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))

    score <- as.matrix(score)
    colnames(score) <- etanames
    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y))
      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
      score[score==Inf] = 1.7e308
      score <- colSums(weights * score, na.rm = TRUE)
    }
    return(score)
  }

  hdist <- function(y, eta, weights = NULL) {    
    ny <- length(y)
    if(is.null(weights)) weights <- rep.int(1, ny)

    d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
    d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-eta[1]) * exp(-2*eta[2]), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
    d2ld.etasigma2 <- sum(weights * (-2)*exp(log((y-eta[1])^2) - 2*eta[2]), na.rm = TRUE)

    # par <- c(eta[1], exp(eta[2]))                           
    # d2ld.etamu2 <- sum(weights * rep.int(-1/par[2]^2, ny))
    # d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-par[1])/par[2]^2), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
    # d2ld.etasigma2 <- sum(weights * (-2)*(y-par[1])^2/par[2]^2, na.rm = TRUE)

    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etamu.d.etasigma, d2ld.etasigma2), nrow = 2)
    colnames(hess) <- rownames(hess) <-  etanames

    return(hess)
  }

  ## additional functions pdist, qdist, rdist
  pdist <- pnorm
  qdist <- qnorm
  rdist <- rnorm

  link <- c("identity", "log")

  linkfun <- function(par) {
    eta <- c(par[1], log(par[2]))
    names(eta) <- etanames
    return(eta)
  }

  linkinv <- function(eta) {
    par <- c(eta[1], exp(eta[2]))
    names(par) <- parnames
    return(par)
  }

  linkinvdr <- function(eta) {
    dpardeta <- c(1, exp(eta[2]))
    names(dpardeta) <- parnames
    return(dpardeta)
  }

  startfun <- function(y, weights = NULL){
    if(is.null(weights)) {
      mu <- mean(y)
      sigma <- sqrt(1/length(y) * sum((y - mu)^2))
    } else {
      mu <- weighted.mean(y, weights)
      sigma <- sqrt(1/sum(weights) * sum(weights * (y - mu)^2))
    }
    starteta <- c(mu, log(sigma))
    names(starteta) <- etanames
    return(starteta)
  }

  mle <- TRUE

  dist_list_normal <- list(family.name = "Normal Distribution",
                           ddist = ddist, 
                           sdist = sdist, 
                           hdist = hdist,
                           pdist = pdist,
                           qdist = qdist,
                           rdist = rdist,
                           link = link, 
                           linkfun = linkfun, 
                           linkinv = linkinv, 
                           linkinvdr = linkinvdr,
                           startfun = startfun,
                           mle = mle,
                           gamlssobj = FALSE,
                           censored = FALSE
  )
}

# dist_list_cens_normal
{

  dist_list_cens_normal <- list()

  parnames <- c("mu", "sigma")
  etanames <- c("mu", "log(sigma)")

  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
    par <- c(eta[1], exp(eta[2]))
    val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
    if(sum) {
      if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
      val <- sum(weights * val, na.rm = TRUE)
    }
    return(val)
  }

  sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
    par <- c(eta[1], exp(eta[2]))
    # y[y==0] <- 1e-323

    score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
    score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
    score <- cbind(score_m, score_s)
    score <- as.matrix(score)
    colnames(score) <- etanames
    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y)[1])
      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
      score[score==Inf] = 1.7e308
      score <- colSums(weights * score, na.rm = TRUE)
      #if(any(is.nan(score))) print(c(eta, "y", y))
    }
    return(score)
  }

  hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
    ny <- length(y)
    if(is.null(weights)) weights <- rep.int(1, ny)

    par <- c(eta[1], exp(eta[2]))                           
    # y[y==0] <- 1e-323

    d2mu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
    d2sigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
    dmudsigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu.sigma", left = left, right = right) # FIX: order?
    dsigmadmu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma.mu", left = left, right = right) # FIX: order?
    dsigma <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)

    d2ld.etamu2 <- sum(weights * d2mu, na.rm = TRUE)
    d2ld.etamu.d.etasigma <- sum(weights * dmudsigma * par[2], na.rm = TRUE)
    d2ld.etasigma.d.etamu <- sum(weights * dsigmadmu * par[2], na.rm = TRUE)
    d2ld.etasigma2 <- sum(weights * (d2sigma * exp(2*eta[2]) + dsigma * par[2]), na.rm = TRUE)

    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etasigma.d.etamu, d2ld.etasigma2), nrow = 2)
    colnames(hess) <- rownames(hess) <-  etanames

    return(hess)
  }

  ## additional functions pdist, qdist, rdist
  pdist <- function(q, eta, lower.tail = TRUE, log.p = FALSE) crch:::pcnorm(q, mean = eta[1], sd = eta[2], 
                                                                            lower.tail = lower.tail, log.p = log.p, 
                                                                            left = left, right = right)
  qdist <- function(p, eta, lower.tail = TRUE, log.p = FALSE) crch:::qcnorm(p, mean = eta[1], sd = eta[2], 
                                                                            lower.tail = lower.tail, log.p = log.p, 
                                                                            left = left, right = right)
  rdist <- function(n, eta) crch:::rcnorm(n, mean = eta[1], sd = eta[2], left = left, right = right)

  link <- c("identity", "log")

  linkfun <- function(par) {
    eta <- c(par[1], log(par[2]))
    names(eta) <- etanames
    return(eta)
  }

  linkinv <- function(eta) {
    par <- c(eta[1], exp(eta[2]))
    names(par) <- parnames
    return(par)
  }

  linkinvdr <- function(eta) {
    dpardeta <- c(1, exp(eta[2]))
    names(dpardeta) <- parnames
    return(dpardeta)
  }

  startfun <- function(y, weights = NULL){
    yc <- pmax(0,y)  # optional ?
    if(is.null(weights)) {
      mu <- mean(yc)
      sigma <- sqrt(1/length(yc) * sum((yc - mu)^2))
    } else {
      mu <- weighted.mean(yc, weights)
      sigma <- sqrt(1/sum(weights) * sum(weights * (yc - mu)^2))
    }
    starteta <- c(mu, log(sigma))
    names(starteta) <- etanames
    return(starteta)
  }

  mle <- FALSE

  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
                                ddist = ddist, 
                                sdist = sdist, 
                                hdist = hdist,
                                pdist = pdist,
                                qdist = qdist,
                                rdist = rdist,
                                link = link, 
                                linkfun = linkfun, 
                                linkinv = linkinv, 
                                linkinvdr = linkinvdr,
                                startfun = startfun,
                                mle = mle,
                                gamlssobj = FALSE,
                                censored = TRUE
  )
}

######## data generating process based on a predefined tree or a parameter function
# if tree: 2 splits => 3 terminal nodes
# 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
#
# input: n            ..... nr of observations
#        family       ..... distribution of the generated observations
#        split.matrix ..... indices of the variables used for the splits together with the corresponding split points
#        parm.matrix  ..... set of distribution parameters for each subgroup
#        fun          ..... parameter function
#
# output: data.frame with generated observations y (generated seperatly in each subgroups with given distributions parameters),
#                         the given split variables x1, ..., x10 for each observation and
#                         the index of the subgroup for each observation or
#                         the distribution parameter for each observation

# FIX: until now only complete lists can be handed over (dist_list...)
dgp <- function(n, family = dist_list_normal, 
                fun = NULL,
                split.matrix = matrix(nrow = 2, ncol = 2), 
                par.matrix = matrix(nrow = 3, ncol = 2), 
                round.sp = 3)
{

  # generating the possible split variables
  x1 <- runif(n,-0.4,1)
  x2 <- runif(n,-10,10)
  x3 <- runif(n,0,100)
  x4 <- x1 + rnorm(n, sd = 0.1)
  x5 <- x1 + rnorm(n, sd = 0.3)
  x6 <- rbinom(n,1,0.5)
  x7 <- rbinom(n,1,0.5)
  x8 <- sample(1:4, n, replace = TRUE) 
  x9 <- sample(1:5, n, replace = TRUE) 
  x10 <- sample(1:7, n, replace = TRUE)
  #x11 <- runif(n,-0.5,1)
  x <- cbind(x1,x2,x3,x4,x5,x6,x7,x8,x9,x10)
  # reduce nr of possible split points by rounding values of split variables
  x <- round(x, digits = round.sp)

  y <- vector(mode = "numeric", length = n)

  # getting the random function
  if(is.function(family)) family <- family()
  if(inherits(family, "gamlss.family")) {
    rfun <- get(paste0("r",family$family[[1]]))
  } else {
    if(family$family.name == "Normal Distribution") rfun <- rNO
    if(family$family.name == "censored Normal Distribution") rfun <- rNO
    if(family$family.name == "Poisson Distribution") rfun <- rPO
    if(family$family.name == "Weibull Distribution") rfun <- survival:::rsurvreg 
  }

  if(is.null(fun)) {
    index <- vector(mode = "numeric", length = n)
    sv = split.matrix[,1]    # index of split variables
    sp = split.matrix[,2]    # split points
    par = par.matrix

    # splitting based on the given split variables and split points and
    # generating the observations seperatly for each subgroup (with the corresponding given distribution parameters)
    for(k in 1:n){
      if(x[k, sv[1]] <= sp[1]){
        y[k] <- do.call(rfun, as.list(c(1,par[1,])))
        index[k] <- 2L
      } else {
        if(x[k, sv[2]] <= sp[2]){
          y[k] <- do.call(rfun, as.list(c(1,par[2,]))) 
          index[k] <- 4L
        } else {
          y[k] <- do.call(rfun, as.list(c(1,par[3,])))
          index[k] <- 5L
        }
      }
    }
    d <- as.data.frame(cbind(y, x, index))
    colnames(d) <- c("y", paste0("x", c(1:10)), "index")

  } else {

    dpar <- fun(x)

    if(!(family$family.name == "Poisson Distribution")){
      y <- rfun(n, dpar[,1], dpar[,2])
      d <- as.data.frame(cbind(y, x, dpar))
      colnames(d) <- c("y", paste0("x", c(1:10)), "mu", "sigma")
    } else {
      y <- rfun(n, dpar)
      d <- as.data.frame(cbind(y, x, dpar))
      colnames(d) <- c("y", paste0("x", c(1:10)), "lambda")
    }
  }

  if(family$family.name == "censored Normal Distribution") {
    #d$ystar <- d$y
    d$y <- pmax(d$y, 0)
  }

  return(d)
}

### plotting function for one data set
plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE, all.var = FALSE) 
{
  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
                    df$fitted.par,
                    dt$fitted.par,
                    #coef(dt)[paste(dt$fitted[,1]),],
                    g$mu.fv, g$sigma.fv)

  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
                          "fitted.mu.df","fitted.sigma.df",
                          "fitted.mu.dt","fitted.sigma.dt",
                          "fitted.mu.g","fitted.sigma.g")
  sp <- plotdata[order(plotdata["x1"]),]

  if(!(any(dt.par, df.par, g.par, all.par, all.var))){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
  col = transpgrey, border = "transparent")
    #lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 1, col = 'grey')
    #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
    #       col = c('black', transpgrey), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # disttree
  if(dt.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", col.main = pal["tree"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2.5)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
  col = pallight["tree"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
    #       col = c('black','grey', pal["tree"], pallight["tree"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # distforest
  if(df.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", col.main = pal["forest"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2.5)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
  col = pallight["forest"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
    #       col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # gamlss
  if(g.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2.5)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
  col = pallight["gamlss"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
    #       col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # disttree vs. distforest vs. gamlss
  if(all.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)

    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
    #                    TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 13.95, expression(mu), bty = "n")
  }

  # variance: disttree vs. distforest vs. gamlss
  if(all.var){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)

    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'black')
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
  col = pallight["tree"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
  col = pallight["forest"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
  col = pallight["gamlss"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
    #legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
    #                    TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
  legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

}

## function to estimate standard deviation of randomForest for a new observation
# (in randomForest the argument 'keep.inbag' must be set to TRUE)
rf_getsd <- function(object, newobs, rfdata){

  # get predictions for the new observations from all trees
  pred.newobs <- predict(object, predict.all = TRUE, newdata = newobs)

  # vector where the standard deviations from all trees are stored
  sd_trees <- numeric(length = object$ntree)

  # loop over all trees of the forest
  for(i in 1:object$ntree){

    # get data used to build this tree
    obsid <- rep.int(c(1:NROW(rfdata)), as.vector(object$inbag[,i]))
    obs_tree <- rfdata[obsid,]
    rownames(obs_tree) <- c(1:NROW(obs_tree))
    # get predictions for this data from this tree
    pred.obs_tree <- predict(object, newdata = obs_tree, predict.all = TRUE)$individual[,i]

    # get prediction for the new observation from this tree
    pred.newobs_tree <- pred.newobs$individual[,i]

    # get part of the data that ends up in the same terminal node (has the same prediction)
    obs_node <- obs_tree[pred.obs_tree == pred.newobs_tree,]

    sd_trees[i] <- sd(obs_node$accel)
  }

  # average of sd over all trees
  sd_newobs <- mean(sd_trees, na.rm = TRUE)

  return(sd_newobs)   
}

## function to estimate standard deviation of cforest for a new observation
cf_getsd <- function(cf, newdata = NULL){

  # get IDs of predicted nodes for the learning data cfdata (does not have to be handed over as cf was learned on cfdata)
  pred.node.learn <- predict(cf, type = "node")

  # get IDs of predicted nodes for the new observations
  if(is.null(newdata)) {
    pred.node.new <- pred.node.learn
    newdata <- cf$data
  } else {
    pred.node.new <- predict(cf, newdata = newdata, type = "node")
  }

  sdnew <- numeric(length = NROW(newdata))

  for(i in 1:NROW(newdata)){
    for(t in 1:cf$info$call$ntree){
      nodedata <- cf$data[(pred.node.learn[[t]] == pred.node.new[[t]][i]),]
      sdnew[i] <- sd(nodedata[,paste(cf$terms[[2]])])
    }
  }

  return(sdnew)   
}

## simulation plot functions
# plot RMSE
plot_rmse <- function(simres, type = c("exp", "par")){

  if(type == "exp"){
    rmse <- NULL
    colnames <- NULL
    col <- NULL
    legendnames <- NULL
    if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.dt"], simres$resmat[,"av.rmse.exp.obs.dt"])
      colnames <- c(colnames, "dt.true", "dt.obs")
      col <- c(col, pal["tree"])
      legendnames <- c(legendnames, "disttree")
    }
    if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.df"], simres$resmat[,"av.rmse.exp.obs.df"])
      colnames <- c(colnames, "df.true", "df.obs")
      col <- c(col, pal["forest"])
      legendnames <- c(legendnames, "distforest")
    }
    if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.g"], simres$resmat[,"av.rmse.exp.obs.g"])
      colnames <- c(colnames, "g.true", "g.obs")
      col <- c(col, pal["gamlss"])
      legendnames <- c(legendnames, "gamlss")
    }
    if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.b"], simres$resmat[,"av.rmse.exp.obs.b"])
      colnames <- c(colnames, "b.true", "b.obs")
      col <- c(col, pal["bamlss"])
      legendnames <- c(legendnames, "bamlss")
    }
    if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.gb"], simres$resmat[,"av.rmse.exp.obs.gb"])
      colnames <- c(colnames, "gb.true", "gb.obs")
      col <- c(col, pal["gamboostLSS"])
      legendnames <- c(legendnames, "gamboostLSS")
    }
    if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.rf"], simres$resmat[,"av.rmse.exp.obs.rf"])
      colnames <- c(colnames, "rf.true", "rf.obs")
      col <- c(col, pal["randomForest"])
      legendnames <- c(legendnames, "randomForest")
    }
    if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.cf"], simres$resmat[,"av.rmse.exp.obs.cf"])
      colnames <- c(colnames, "cf.true","cf.obs")
      col <- c(col, pal["cforest"])
      legendnames <- c(legendnames, "cforest")
    }

    colnames(rmse) <- colnames
    ylim <- c(min(rmse), max(rmse))

    plot(x = simres$x.axis, y = rmse[,1], type = "l", col = col[1], ylim = ylim,
         xlab = "kappa", ylab = "RMSE")
    lines(x = simres$x.axis, y = rmse[,2], type = "l", lty = 2, col = col[1])

    for(i in 1:(length(col)-1)){
      lines(x = simres$x.axis, y = rmse[,(2*i+1)], type = "l", col = col[i+1])
      lines(x = simres$x.axis, y = rmse[,(2*i+2)], type = "l", lty = 2, col = col[i+1])
    }
    legend('topleft', legendnames, 
           col = col, lty = 1, cex = 0.7)
  }

  if(type == "par"){

    rmse <- NULL
    colnames <- NULL
    col <- NULL
    legendnames <- NULL
    if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.dt"], simres$resmat[,"av.rmse.sigma.dt"])
      colnames <- c(colnames, "dt.true", "dt.obs")
      col <- c(col, pal["tree"])
      legendnames <- c(legendnames, "disttree")
    }
    if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.df"], simres$resmat[,"av.rmse.sigma.df"])
      colnames <- c(colnames, "df.true", "df.obs")
      col <- c(col, pal["forest"])
      legendnames <- c(legendnames, "distforest")
    }
    if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.g"], simres$resmat[,"av.rmse.sigma.g"])
      colnames <- c(colnames, "g.true", "g.obs")
      col <- c(col, pal["gamlss"])
      legendnames <- c(legendnames, "gamlss")
    }
    if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.b"], simres$resmat[,"av.rmse.sigma.b"])
      colnames <- c(colnames, "b.true", "b.obs")
      col <- c(col, pal["bamlss"])
      legendnames <- c(legendnames, "bamlss")
    }
    if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.gb"], simres$resmat[,"av.rmse.sigma.gb"])
      colnames <- c(colnames, "gb.true", "gb.obs")
      col <- c(col, pal["gamboostLSS"])
      legendnames <- c(legendnames, "gamboostLSS")
    }
    if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.rf"], simres$resmat[,"av.rmse.sigma.rf"])
      colnames <- c(colnames, "rf.true", "rf.obs")
      col <- c(col, pal["randomForest"])
      legendnames <- c(legendnames, "randomForest")
    }
    if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.cf"], simres$resmat[,"av.rmse.sigma.cf"])
      colnames <- c(colnames, "cf.true","cf.obs")
      col <- c(col, pal["cforest"])
      legendnames <- c(legendnames, "cforest")
    }

    colnames(rmse) <- colnames
    ylim <- c(min(rmse), max(rmse))

    plot(x = simres$x.axis, y = rmse[,1], type = "l", col = col[1], ylim = ylim,
         xlab = "kappa", ylab = "RMSE")
    lines(x = simres$x.axis, y = rmse[,2], type = "l", lty = 2, col = col[1])

    for(i in 1:(length(col)-1)){
      lines(x = simres$x.axis, y = rmse[,(2*i+1)], type = "l", col = col[i+1])
      lines(x = simres$x.axis, y = rmse[,(2*i+2)], type = "l", lty = 2, col = col[i+1])
    }
    legend('topleft', legendnames, 
           col = col, lty = 1, cex = 0.7)
  }
}

# plot loglikelihood
plot_ll <- function(simres){

    ll <- NULL
    colnames <- NULL
    col <- NULL
    legendnames <- NULL
    if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
      ll <- cbind(ll, simres$resmat[,"av.loglik.dt"])
      colnames <- c(colnames, "dt.ll")
      col <- c(col, pal["tree"])
      legendnames <- c(legendnames, "disttree")
    }
    if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
      ll <- cbind(ll, simres$resmat[,"av.loglik.df"])
      colnames <- c(colnames, "df.ll")
      col <- c(col, pal["forest"])
      legendnames <- c(legendnames, "distforest")
    }
    if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
      ll <- cbind(ll, simres$resmat[,"av.loglik.g"])
      colnames <- c(colnames, "g.true")
      col <- c(col, pal["gamlss"])
      legendnames <- c(legendnames, "gamlss")
    }
    if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
      ll <- cbind(ll, simres$resmat[,"av.loglik.b"])
      colnames <- c(colnames, "b.true")
      col <- c(col, pal["bamlss"])
      legendnames <- c(legendnames, "bamlss")
    }
    if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
      ll <- cbind(ll, simres$resmat[,"av.loglik.gb"])
      colnames <- c(colnames, "gb.true")
      col <- c(col, pal["gamboostLSS"])
      legendnames <- c(legendnames, "gamboostLSS")
    }
    if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
      ll <- cbind(ll, simres$resmat[,"av.loglik.rf"])
      colnames <- c(colnames, "rf.true")
      col <- c(col, pal["randomForest"])
      legendnames <- c(legendnames, "randomForest")
    }
    if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
      ll <- cbind(ll, simres$resmat[,"av.loglik.cf"])
      colnames <- c(colnames, "cf.true")
      col <- c(col, pal["cforest"])
      legendnames <- c(legendnames, "cforest")
    }

    colnames(ll) <- colnames
    ylim <- c(min(ll), max(ll))

    plot(x = simres$x.axis, y = ll[,1], type = "l", col = col[1], ylim = ylim,
         xlab = "kappa", ylab = "log-likelihood")

    for(i in 2:length(col)){
      lines(x = simres$x.axis, y = ll[,i], type = "l", col = col[i])
    }
    legend('topleft', legendnames, 
           col = col, lty = 1, cex = 0.7)
}
@

\title{\texttt{disttree}: Distributional Trees and Forests}
\Plaintitle{disttree: Distributional Trees and Forests}

\author{Lisa Schlosser\\Universit\"at Innsbruck
   \And Torsten Hothorn\\Universit\"at Z\"urich
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}

\Abstract{
In regression analysis one is interested in the relationship between a dependent variable and one or more explanatory variables. Various methods to fit statistical models to the data set have been developed, starting from linear models considering only the mean of the response variable and ranging to probabilistic models where all parameters of a distribution are fit to the given data set.\\
If there is a strong variation within the data it might be advantageous to split the data first into more homogeneous subgroups based on given covariates and then fit a local model in each subgroup rather than fitting one global model to the whole data set. This can be done by applying regression trees and forests.\\
Both of these two concepts, parametric modeling and algorithmic trees, have been investigated and developed further, however, mostly separated from each other. Therefore, our goal is to embed the progress made in the field of probabilistic modeling in the idea of algorithmic tree and forest models. In particular, more flexible models such as GAMLSS (\cite{stasinopoulos2005}) should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc. while non-additive effects of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree. In distributional forests an ensemble of distributional trees is built and used to calculate weights which are then included in the fitting process of a distributional model.
}
\Keywords{parametric models, trees and forests, recursive partitioning, maximum likelihood}

\Address{
  Lisa Schlosser \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Lisa.Schlosser@uibk.ac.at} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/lisa-schlosser.html}\\

  Torsten Hothorn\\
  Institut f\"ur Sozial- und Pr\"aventivmedizin, Abteilung Biostatistik \\
  Universit\"at Z\"urich \\
  Hirschengraben 84\\
  CH-8001 Z\"urich, Switzerland \\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://user.math.uzh.ch/hothorn/}\\

  Achim Zeileis \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Achim.Zeileis@R-project.org} \\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

\begin{document}
\SweaveOpts{concordance=TRUE}

\SweaveOpts{eps=FALSE, keep.source=TRUE, eval = TRUE}

\section{Introduction}

In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables. This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family. (In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination. (In the fitting function \code{gam} types of smoothing terms can be selected.)\\
However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
(GAMLSS) (\cite{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM. In that way a whole distribution can be specified. The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
This development in the field of parametric modeling is illustrated in figure \ref{Devel_parmod}.\\

\begin{figure}[t!]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage
{\LARGE$\rightarrow$}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage
{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage
\vspace{0.0cm}
\minipage{0.25\textwidth}
\begin{center}
\hspace{0.4cm}
LM, GLM 
\end{center}
\endminipage
\hspace{1.4cm}
\minipage{0.25\textwidth}
\begin{center}
GAM
\end{center}
\endminipage
\hspace{1.4cm}
\minipage{0.25\textwidth}
\begin{center}
GAMLSS
\end{center}
\endminipage
\caption{\label{Devel_parmod}Development in the field of parametric modeling}
\end{figure}

\begin{figure}[t!]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@

<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
\end{center}
\endminipage
{\LARGE$\rightarrow$}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@

<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
\end{center}
\endminipage
{\LARGE$\rightarrow$}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_distforest, echo=FALSE, results=hide>>=
## distforest
set.seed(4723)
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
xt <- 2*(x-0.5)
var <- exp(-(2*xt)^4)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
x <- c(1:nobs)/nobs
@

<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
points(x, y, col = "slategray", pch = 19)
box(lwd=5)
@
\end{center}
\endminipage

\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Regression Tree 
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Random Forest 
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
Distributional Forest 
\end{center}
\endminipage\\
\minipage{0.29\textwidth}
\vspace{0.2cm}
\center {\LARGE$\downarrow$}\\
\vspace{-0.4cm}
%\hspace{0.0cm}
\begin{center}
<<motivation_disttree, echo=FALSE, results=hide>>=
## disttree
set.seed(74)
x <- c(1:nobs)/nobs
var <- ytree/3
y <- ytrue + rnorm(nobs,0,var)
@

<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
  col = pallight["forest"], border = "transparent")
box(lwd=5)
@
\vspace{-0.2cm}\\
Distributional Tree
\vspace{0.2cm}
\end{center}
\endminipage
\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests}
\end{figure}

All these parametric models work very well as long as only additive effects appear in the considered data. A different type of model that can capture non-additive effects as well are regression trees. Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node. One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of figure \ref{Devel_treeforest}. One way how to tackle this problem is to not just consider one tree but an ensemble of trees. This is the idea of random forests. A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set. Then the results of all trees are combined to get a final model which unlike a tree model can also deal with smooth effects in a reasonable way.\\ 
In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
\\
In the past the combination of parametric data models and algorithmic tree models has gained more and more interest since it allows for a more flexible and clear structure. \\
Algorithms that only fit constants to each final node and therefore only model the mean of the response variable (such as CART, Breiman 1984) might lead to very large trees which complicates analyzing and interpreting. Algorithms such as GUIDE (Loh 2002), CRUISE (Kim and Loh 2001) and LOTUS (Chan and Loh 2004) were among the first to provide trees with parametric models in each terminal node. The MOB algorithm (Zeileis, Hothorn and Hornik 2008), an algorithm for model-based recursive partitioning, extended this idea by fitting parametric models to each node and using the gained information of the fitted model to create further splits in inner nodes. Using this algorithm as a framework to build trees LMs and GLMs can be fitted in the nodes of the trees. However, up to now trees which only model the mean of the response variable in its nodes are still more common and widely used.\\
Random forests (\cite{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\cite{hothorn2015partykit}, \cite{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
\\
But as mentioned before, until now the above listed development in the field of probabilistic modeling has not yet been fully integrated in the idea of algorithmic tree and forest models. Probabilistic models are rarely used in combination with tree algorithms and more flexible models such as GAMLSSs have not yet been applied as models in the nodes of a tree.\\
\\
Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests. In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc. while non-additive effects or interactions of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree. In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
These (future) steps are illustrated in figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
\\
\\
Now we take a look at a classical text book example, the mcycle data set from the package \pkg{MASS}. It consists of a series of measurements of head acceleration in a simulated motorcycle accident. The response variable \code{accel} represents the recorded head acceleration meassured in g and the explanatory variable \code{times} is meassured in milliseconds after the impact.\\
We compare the performance of the following four functions on this data set: \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest}\\
This situation is particularly well suited for a GAMLSS because there is only one regressor and physically one would expect a smooth function. However, this example shows, that the distributional forest can also capture the changes quite well albeit not as smoothly as the GAMLSS. Moreover, it illustrates that the distributional tree detects the changes as well but models them with rough steps.\\
\\
To make it possible to compare the models also on the estimated scale parameter, here the standard deviation of the random forest is estimated in each terminal node of each tree and then averaged, in the same way as the mean parameter is estimated. This makes it possible to specify a distribution, however, the splits of a random forest are only made in the location parameter while a distributional tree/forest considers splits in all distribution parameters.\\
\\
In figure \ref{plot_mcycle} the results of a distributional forest are plotted on the top left and then compared to the results of a distributional tree, a GAMLSS and a random forest, regarding the location parameter on the top right and the scale parameter below.\\
While the smooth function of \code{gamlss} seems to model especially the middle part (for \code{times} between 17 and 40) better than the wiggly functions resulting from the trees it can also be observed that at \code{times} $=14$ there is a rapid drop which is captuerd more precisely by the tree-based models while \code{gamlss} starts descending earlier. Also for the highest values of \code{times} the smooth function increases again while the forests stay almost at the same level and the tree stays at a fixed value which seems more appropriate in this situation.\\
Comparing the two forest models, the distributional forest seems to be a bit more stable while the random forest includes more rapid changes of a wider range.\\
Looking at the plot of the scale parameter, in general the distributional forest estimates higher parameters than the random forest.\\

<<echo=FALSE, results=hide>>=

## FIX ME: distforest uses subsets, randomForest uses bootstraps

## randomForest: https://stackoverflow.com/questions/28417826/nodesize-parameter-ignored-in-randomforest-package
## nodesize = the minimum number of observations that must exist in a node in order for a split to be attempted.
## -> nodesize = minsplit

if(file.exists("mcycle_plotdata.rda")){
  load("mcycle_plotdata.rda")
} else {
  set.seed(7)
  dt <- disttree(accel ~ times, data = mcycle, family = NO(), type.tree = "mob", control = mob_control(minsize = 7L))
  df <- distforest(accel ~ times, data = mcycle, family = NO(), ntree = 300L, 
                   control = ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 5L, minsplit = 10L, intersplit = TRUE))
  rf <- randomForest(accel ~ times, data = mcycle, ntree = 300L, nodesize = 10L, 
                     keep.inbag = TRUE, replace = FALSE)
  #ga <- gam(list(accel ~ s(times), ~ s(times)), data = mcycle, family=gaulss())
  ga <- gam(list(accel ~ s(times,k=20,bs="ad"), ~ s(times)), data = mcycle, family=gaulss())
  #b <- bamlss(list(accel ~ s(times), accel ~ s(times)), data = mcycle)save(df, file = "mcycle_plotdata.rda")

  ## get 'fitted.sigma.rf'
  rf_sd <- numeric(length(NROW(mcycle)))

  for(i in 1:NROW(mcycle)){
    rf_sd[i] <- rf_getsd(rf, newobs = mcycle[i,], rfdata = mcycle)
  }

  plotdata <- cbind(mcycle,
                    df$fitted.par,
                    dt$fitted.par,
                    #g$mu.fv, g$sigma.fv,
                    ga$fitted.values[,1], 1/ga$fitted.values[,2],
                    rf$predicted, rf_sd)

  colnames(plotdata) <- c("times", "accel",
                          "fitted.mu.df","fitted.sigma.df",
                          "fitted.mu.dt","fitted.sigma.dt",
                          "fitted.mu.g","fitted.sigma.g",
                          "fitted.mu.rf","fitted.sigma.rf")

  sp <- plotdata[order(plotdata["times"]),]

  save(sp, file = "mcycle_plotdata.rda")
}
@

\begin{figure}[t!]
\begin{subfigure}{0.65\textwidth}
<<plot_df, fig=TRUE, echo=FALSE>>=
# distforest
plot(mcycle)
lines(x = sp$times, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
polygon(c(sp$times, rev(sp$times)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
        col = pallight["forest"], border = "transparent")
@
%\caption{\label{plot_mcycle_df}Distributional Forest, \code{distforest}}
\end{subfigure}
\hspace{-2cm}
\begin{subfigure}{0.65\textwidth}
<<plot_mu_vs, fig=TRUE, echo=FALSE>>=
# disttree vs. distforest vs. gamlss vs. randomForest
plot(x = sp$times, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 1.5,
     ylim = c(-134,75), ylab = "accel", xlab = "times")
lines(x = sp$times, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 1.5)
lines(x = sp$times, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 1.5)
lines(x = sp$times, y = sp$fitted.mu.rf, type = "l", col = pal["randomForest"], lwd = 1.5)
legend('topleft', c("disttree", "distforest", "gamlss", "randomForest"), 
       col = c(pal["tree"], pal["forest"], pal["gamlss"], pal["randomForest"]), lty = 1, cex = 1)
@
%\caption{\label{plot_mcycle_mu_vs} Location parameter: \code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
\end{subfigure}
\vspace{-1cm}

\hspace{4cm}
\begin{subfigure}{0.65\textwidth}
<<plot_sigma_vs_lines, fig=TRUE, echo=FALSE>>=
# variance: disttree vs. distforest vs. gamlss (lines)
plot(x = sp$times, y = sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2,
     ylim = c(0,50), ylab = "variance of accel", xlab = "times")
#lines(x = sp$times, y = -sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
lines(x = sp$times, y = sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
#lines(x = sp$times, y = -sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
lines(x = sp$times, y = sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
#lines(x = sp$times, y = -sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
lines(x = sp$times, y = sp$fitted.sigma.rf, type = "l", col = pal["randomForest"], lwd = 2)
#lines(x = sp$times, y = -sp$fitted.sigma.rf, type = "l", col = pal["randomForest"], lwd = 2)
legend('topleft', c("disttree", "distforest", "gamlss", "randomForest"), 
       col = c(pal["tree"], pal["forest"], pal["gamlss"], pal["randomForest"]), lty = 1, cex = 1)
@
\end{subfigure}
\caption{\label{plot_mcycle}top left: location and scale parameter of \code{distforest}\\
top right: location parameter of all four models\\
bottom: scale parameter of all four models}
\end{figure}

\newpage
\section{Methodology}
\label{Methodology}
\subsection{Distributional Fit}
Fitting a distributional model to a set of observations without considering any covariates is a well known procedure which can for example be done by applying the maximum likelihood method. This is also the chosen approach in the methodology described in this section.\\
The goal is to fit a distributional model $D(Y, \theta)$ to a response variable $Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters $\theta \in \Theta$ in case of a \textit{k}-parametric distribution. The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided. Therefore, the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$. This is done by the following maximization.
$$
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$
where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response variable $Y$.\\
In that way, a whole distribution is specified with all its features, including location, scale and shape. However, fitting one global model to the whole data set might be too much of a generalization and therefore not represent the data and its features in a reasonable way. Moreover, if covariates are available it is desirable to include them into the model as they might provide important information. In particular, they can be used to separate the data into more homogeneous subgroups. Doing so after fitting a global model to the whole data set and then fitting a local model to each subgroup should improve the model remarkably. This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.

\subsection{Distributional Tree}
After having fit a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is now to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ to decide if the data set should be split into subgroups. In particular, the information provided by the fitted model are the scores 
$$
s(\hat{\theta}, y_i) = \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i).
$$ 
These values express how well the model fits the data regarding each parameter. Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and each observation.\\
Statistical tests are applied to find out whether any significant connection/dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$ is given. If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest connection/dependency.\\
In a next step a local distributional model is fit in each of the subgroups. Repeating this procedure of fitting a model, applying statistical tests to evaluate the goodness of fit and splitting the data set depending on the test results leads to a tree structure. In that way the learning data set is separated into \textit{B} disjoint segments such that
$$
\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b
$$
where each segment is represented by a terminal node of the tree.\\

The exact way of determining how and where to split varies over different tree algorithms. For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm or the ctree algorithm. Summed up, the steps of building a distributional tree using one of these algorithms are
\begin{enumerate}
\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
\item Estimate $\bold{\hat{\theta}}$ via maximum likelihood.
\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
%\item Assess whether the \emph{model scores} are associated with
%    (or change along) any of the available covariates -- e.g.,
%    using parameter instability tests (\emph{strucchange}) or
%    conditional inference (\emph{coin}).
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose the breakpoint with the highest improvement in log-likelihood.
\item Repeat steps 2--4 recursively in the subgroups
    until some stopping criterion is met.
    %-- e.g., for significance or sample size.
%\item Choose the variable with the strongest association
%\item Choose the split point which leads to the highest improvement of the model
%\item Split and repeat 2-6 in each node until a stopping criterion is met
\end{enumerate}
The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.\\
Executing these steps results in a stepwise parametric model where a complete distribution is fit in each node of the tree.\\
\\
For distributional trees the strategy of how to make a prediction for a new observation $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ is very simple and straight forward. All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new observation are selected and a distributional model is fit on the subset. To select this subset a weight vector $\bold{\omega^{tree}} = (\omega^{tree}_1, \ldots, \omega^{tree}_n)$ is included in the fitting process.
$$
\omega^{tree}_i(\bold{z}^{\ast}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z}^{\ast} \in \mathcal{B}_b))
$$
The predicted distribution of the new observation is then fully specified by the estimated parameter $\hat{\theta}(\bold{z}^{\ast})$ where
$$
\hat{\theta}(\bold{z}^{\ast}) = \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{tree}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i).
$$

While the clear structure of such a model is advantageous in analyzing it, the abrupt changes are often too rough and impose steps on the model even in case of smooth effects. In these situations bagging can solve the problem by combining an ensemble of trees such that wrongly detected abrupt changes of the model are turned into almost smooth transitions. In that way, all kinds of effects can be modeled, which is is also one the of the main advantages of forest models (taking bagging one step further as the variance is reduced by sampling the covariates considered for the splitting process in each node).

\subsection{Distributional Forest}
A distributional forest consists of an ensemble of distributional trees $\{t_s\}_{s=1,\ldots,T}$. Each of these trees is built on a different data set which can either be a bootstrap or a subset of the original data set. Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables. In that way the correlation among the resulting trees is reduced and therefores also the variance of the model.\\
\\
Forest models differ mainly in the way how the trees are combined to make predictions. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.\\
Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new observation. This difference is represented in the weight vector, as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
$\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
For a new observation $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ this set of weights is obtained by the following procedure: For each observation $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which it ends up in the same terminal node as the new observation $\bold{z}^{\ast}$. 
$$
\omega^{forest}_i(\bold{z}^{\ast}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z}^{\ast} \in \mathcal{B}^s_b))
$$
where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new observation $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way, leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
$$
\hat{\theta}(\bold{z}^{\ast}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{forest}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i)
$$
While in a distributional tree the same distribution is specified for all observations being in the same node a distributional forest predicts and fits a different distribution for each different observation which is due to the individual weights calculated separately for each observation.

\newpage
\section{Application}
\subsection{Continuation of the mcycle example: Cross Validation}
In the introduction the four functions \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest} have been applied on the mcycle data set and their results haven been visualized and discussed. To get a more profound impression of their performances we now want to compare them in a cross validation framework on this text book example data set. The 133 observations are randomly split into 10 subsets roughly of the same size. After learning the model on 9 of these subgroups the RMSE of the estimated location parameter and the log-likelihood function are evaluated on the 10th part. Each subset is chosen once as a test data set and the mean of the resulting values is stored. This procedure is repeated 9 times. This comparison is illustrated in figure \ref{plot_cross_mean}, based on the RMSE on the left and based on the log-likelihood on the right. (Extrapolation of the GAMLSS would lead to very small values of the log-likelihood. To avoid this, the first and the last observation are always included in the learn data set.) \\
As we have already seen in the introduction the distributional tree obviously struggles with the smooth effects while the two forest models can capture the changes quite well. As far as the RMSE of the estimated location parameter is concerned the GAMLSS plays to its strength as a smooth parametric model and leads to the best results followed by the distributional forest and the random forest which might both be a bit to wiggly.\\
However, regarding the log-likelihood the forests perform better than the GAMLSS. This is mainly due to the fact that the GAMLSS models the scale parameter close to zero for observations with \code{times} $<10$. This leads to very low values of the log-likelihood function if one or more of these observations are included in the test data set. As the estimated scale parameter of the tree based models are higher for these specific observations this problem does not occure here.\\
Overall it can be seen that the distributional forest which fits the whole distribution and considers splits in both parameters performs better than the random forest which only models the mean and therefore also only inlcudes splits in the location parameter.

<<echo=FALSE, results=hide>>=
#### cross validation
if((file.exists("mcycle_cross_mean.rda")) & (file.exists("mcycle_cross_all.rda"))){
  load("mcycle_cross_mean.rda")
  load("mcycle_cross_all.rda")
  nrep_cross <- NROW(cross_all)
} else {

  set.seed(704)
  #set.seed(7054)
  nrep_cross <- 10

  rmse <- matrix(nrow = nrep_cross * 10, ncol = 4)
  ll <- matrix(nrow = nrep_cross * 10, ncol = 4)
  mean_rmse <- matrix(nrow = nrep_cross, ncol = 4)
  mean_ll <- matrix(nrow = nrep_cross, ncol = 4)
  errors <- numeric(length = nrep_cross * 10)
  colnames(rmse) <- colnames(mean_rmse) <- c("rmse_disttree", "rmse_distforest", "rmse_gamlss", "rmse_randomForest")
  colnames(ll) <- colnames(mean_ll) <- c("ll_disttree", "ll_distforest", "ll_gamlss", "ll_randomForest")

  tdata <- list()
  ldata <- list()

  for(k in 1:nrep_cross) {

    # separate data set (without the first and the last observation) into 10 parts (9 of size 13 and 1 of size 14):
    # (first and last observation are not in testid -> always stay in learndata)
    id <- c(2:(NROW(mcycle)-1))
    testid <- list()
    for(i in 1:9){
      testid[[i]] <- sample(id, 13, replace = FALSE)
      id <- id[!(id %in% testid[[i]])]
    }
    testid[[10]] <- id

    rmse_dt_mu <- rmse_df_mu <- rmse_g_mu <- rmse_rf_mu <- 
      ll_dt <- ll_df <- ll_g <- ll_rf <- numeric(length = 10)

    for(i in 1:10){
      tdata[[(k-1)*10+i]] <- testdata <- mcycle[testid[[i]], ]
      ldata[[(k-1)*10+i]] <- learndata <- mcycle[-testid[[i]], ]

      #testdata <- mcycle[testid[[i]], ]
      #learndata <- mcycle[-testid[[i]], ]

      dt <- disttree(accel ~ times, data = learndata, family = NO(), type.tree = "mob", control = mob_control(minsize = 7L))
      df <- distforest(accel ~ times, data = learndata, family = NO(), ntree = 300L, 
                       control = ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 5L, minsplit = 10L, intersplit = TRUE))
      rf <- randomForest(accel ~ times, data = learndata, ntree = 300L, nodesize = 10L, keep.inbag = TRUE, replace = FALSE)
      ga <- gam(list(accel ~ s(times,k=20,bs="ad"), ~ s(times)), data = learndata, family=gaulss())
      #ga <- gam(list(accel ~ s(times), ~ s(times)), data = learndata, family=gaulss())
      errors[(k-1)*10+i] <- (ga$outer.info$conv != "full convergence")

      # get predicted values
      dt.pred.par <- predict(dt, newdata = testdata, type = "parameter")
      df.pred.par <- predict(df, newdata = testdata, type = "parameter")
      rf.pred.resp <- predict(rf, newdata = testdata)
      rf.sd <- numeric(length(NROW(testdata)))
      for(l in 1:NROW(testdata)){
        rf.sd[l] <- rf_getsd(rf, newobs = testdata[l,], rfdata = learndata)
      }
      g.pred.par <- predict(ga, newdata = testdata, type = "response")
      g.pred.mu <- g.pred.par[,1]
      g.pred.sigma <- 1/g.pred.par[,2]

      # RMSE location parameter
      rmse_dt_mu[i] <- sqrt(mean((dt.pred.par$mu - testdata$accel)^2))
      rmse_df_mu[i] <- sqrt(mean((df.pred.par$mu - testdata$accel)^2))
      rmse_g_mu[i] <- sqrt(mean((g.pred.mu - testdata$accel)^2))
      rmse_rf_mu[i] <- sqrt(mean((rf.pred.resp - testdata$accel)^2))

      # loglikelihood
      dtll <- dfll <- gll <- rfll <- 0
      for(j in 1:(nrow(testdata))){

        dtll_j <- dnorm(testdata[j,"accel"], mean = dt.pred.par$mu[j], sd = dt.pred.par$sigma[j], log=TRUE)
        dfll_j <- dnorm(testdata[j,"accel"], mean = df.pred.par$mu[j], sd = df.pred.par$sigma[j], log=TRUE)
        gll_j <- dnorm(testdata[j,"accel"], mean = g.pred.mu[j], sd = g.pred.sigma[j], log=TRUE)
        rfll_j <- dnorm(testdata[j,"accel"], mean = rf.pred.resp[j], sd = rf.sd[j], log=TRUE)

        dtll <- if(is.na(dtll_j)) {
          print(dt.pred.par[j,], testdata[j,"accel"]) 
          dtll + (-5)
        } else {dtll + dtll_j}
        dfll <- if(is.na(dfll_j)) {
          print(df.pred.par[j,], testdata[j,"accel"]) 
          dfll + (-5)
        } else {dfll + dfll_j}
        gll <- if(is.na(gll_j)) {
          print(g.pred.mu[j], g.pred.sigma[j], testdata[j,"accel"]) 
          gll + (-5)
        } else {gll + gll_j}
        rfll <- if(is.na(rfll_j)) {
          print(c(rf.pred.resp[j], rf.sd[j]), testdata[j,"accel"]) 
          rfll + (-5)
        } else {rfll + rfll_j}
      }

      ll_dt[i] <- dtll
      ll_df[i] <- dfll
      ll_g[i] <- gll
      ll_rf[i] <- rfll

      rmse[i+(k-1)*10,] <- c(rmse_dt_mu[i], rmse_df_mu[i], rmse_g_mu[i], rmse_rf_mu[i])
      ll[i+(k-1)*10,] <- c(ll_dt[i], ll_df[i], ll_g[i], ll_rf[i])
    }

    # store data
    mean_rmse[k,] <- c(mean(rmse_dt_mu), mean(rmse_df_mu), mean(rmse_g_mu), mean(rmse_rf_mu))
    mean_ll[k,] <- c(mean(ll_dt), mean(ll_df), mean(ll_g), mean(ll_rf))

  }

  cross_all <- cbind(rmse, ll)
  cross_mean <- cbind(mean_rmse, mean_ll)
  save(cross_all, file = "mcycle_cross_all.rda")
  save(cross_mean, file = "mcycle_cross_mean.rda")

}
@

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<mcycle_cross_rmse_mean, fig=TRUE, echo=FALSE>>=
# parallel boxplots RMSE
box_rmse <- cross_mean[, c("rmse_disttree", "rmse_randomForest", "rmse_distforest", "rmse_gamlss")]
colnames(box_rmse) <- c("disttree", "randomForest", "distforest", "gamlss")
boxplot.matrix(box_rmse, ylab = "RMSE")
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<mcycle_cross_ll_mean, fig=TRUE, echo=FALSE>>=
# parallel boxplots log-likelihood
box_ll <- cross_mean[, c("ll_disttree", "ll_randomForest", "ll_distforest", "ll_gamlss")]
#box_ll <- box_ll[box_ll[,1]>-400,]
#box_ll <- box_ll[box_ll[,2]>-400,]
box_ll <- box_ll[box_ll[,"ll_gamlss"]>-70,]
#box_ll <- box_ll[box_ll[,4]>-400,]
colnames(box_ll) <- c("disttree", "randomForest", "distforest", "gamlss")
boxplot.matrix(box_ll, ylab = "log-likelihood")
@
\end{subfigure}
%\center
\caption{\label{plot_cross_mean}10-times-10 cross validation on the mcycle data, averages (>-70):\\ \code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
\end{figure}

%\begin{figure}[t!]
%\begin{subfigure}{0.6\textwidth}
%<<mcycle_cross_rmse_all, fig=TRUE, echo=FALSE>>=
%# parallel boxplots RMSE
%box_rmse <- cross_all[, c("rmse_disttree", "rmse_randomForest", "rmse_distforest", %"rmse_gamlss")]
%colnames(box_rmse) <- c("disttree", "randomForest", "distforest", "gamlss")
%boxplot.matrix(box_rmse, ylab = "RMSE")
%@
%%\caption{\label{plot_boot_rmse}Bootstraps: RMSE of the location parameter, \code{dist%tree%} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
%\end{subfigure}
%\hspace{-1cm}
%\begin{subfigure}{0.6\textwidth}
%<<mcycle_cross_ll_all, fig=TRUE, echo=FALSE>>=
%# parallel boxplots log-likelihood
%box_ll <- cross_all[, c("ll_disttree", "ll_randomForest", "ll_distforest", "ll_gamlss"%)]
%#box_ll <- box_ll[box_ll[,1]>-400,]
%#box_ll <- box_ll[box_ll[,2]>-400,]
%box_ll <- box_ll[box_ll[,"ll_gamlss"]>-100,]
%#box_ll <- box_ll[box_ll[,4]>-400,]
%colnames(box_ll) <- c("disttree", "randomForest", "distforest", "gamlss")
%boxplot.matrix(box_ll, ylab = "log-likelihood")
%@
%\end{subfigure}
%%\center
%\caption{\label{plot_cross_ll_all}mcycle 10-times-10 Cross validation (>-100): %\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
%\end{figure}

\newpage
\section{Simulation Study}
Within this simulation study the performances of distributional trees and forests are tested and compared to other models, showing situations where one or the other method is more advantageous than others. By doing so the following hypotheses are investigated:
\begin{itemize}%[leftmargin=0.5cm]
\item When applying a tree algorithm the use of a fitting function which returns scores is advantageous to find correct splits. In particular it is crucial to find splits in parameters other than the location parameter. Hence a forest consisting of trees which were built using scores for all distribution parameters leads to better results than a simple forest considering only the mean.
\item In the case of smooth parameter functions without any abrupt changes a distributional forest performs quite well, even though other models such as GAMLSS might be more successful in these situations. However, once the smooth parameter functions include very steep sections the distributional forest outperforms other models (and in extreme cases a distributional tree does as well).
\item Once non-additive effects or interactions appear within the parameter functions distributional trees and forests clearly show their advantages over other models.
\end{itemize}

The following methods are compared to distributional trees and forests:
\begin{itemize}
\item gamlss \cite{stasinopoulos2005} fits a separate generalized additive model to each of the distribution models. In that way it specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.
\item bamlss \cite{bamlss2017} also provides an infrastructure for estimating probabilistic distributional regression but within a Bayesian Framework. Again each parameter is modelled by (possibly) complex additive terms.
\item gamboostLSS \cite{gamboostLSS2017} offers boosting models for fitting  generalized additive models for location, shape and scale
\item randomForest \cite{randomForest2002} combines an ensemble of classification and regression trees applying bagging and reducing the correlation among trees through a randomized selection of possible split variables within each node.
\item cforest \cite{hothorn2015partykit} builds a forest in a similar way but consisting of ctrees applying the ctree algorithm mentioned in section \ref{Methodology}. In this simulation study no transformation function is included.
\end{itemize}

The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and two covariates $X_1$ and $X_2$.\\
The covariates are generated based on the following distribution assumptions:
\begin{center}
\begin{tabular}{ l  c  l }
$X_1$ & $\sim$ & \textit{Unif}([-0.4,1]) \\ 
$X_2$ & $\sim$ & \textit{Unif}([-10,10])
\end{tabular}
\end{center}
The distribution of the response variable is a normal distribution where the corresponding distribution parameters are defined by parameter functions depending on the covariates. 
\vspace{-1cm}
\begin{center}
$$ Y \sim \mathcal{N} (f(x_1, x_2, \kappa))$$
with
$$f(x_1, x_2, \kappa) = \left(\mu(x_1,x_2,\kappa) , \sigma(x_1,x_2,\kappa)\right)$$ 
\end{center}

An additional factor $\kappa$ defines the steepness of certain sections of the parameter function. For low values of $\kappa$ the resulting parameters change rather slowly over varying values of the covariates while high values of $\kappa$ lead to abrupt changes. \\
Including non-additive effects creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated. Here a non-additive effects is created by separating the generated data into two subgroups depending on the value of $X_2$. In particular, within the function $\mu$ each observation is assigned with a probability of being in subgroup 1. This probability is specified by a smooth function and used to generate a binary variable indicating in which subgroup the observation is. Depending on the subgroup the actual smooth parameter function is defined.
\vspace{-1cm}
\begin{center}
$$\mu(x_1,x_2,\kappa)  = 10 \cdot \Big(1-s(x_1,x_2,\kappa)+ \big(2\cdot s(x_1,x_2,\kappa) - 1\big) \cdot \text{\code{rbinom}}\left(n,1,P_1\right)\Big)
$$
with the distribution function of the logistic distribution as a smooth function
$$
s(x_2,\kappa) = \text{\code{plogis}}\left(\kappa^2 \cdot \frac{x_2 - 3}{10}\right)
$$
and the probability of being in subgroup 1
$$
P_1(x_1,\kappa) = e^{-(10\cdot x_1 - 5)^{2 \cdot \kappa}}.
$$
\end{center}
In this scenerio the scale parameter $\sigma$ depends on the location parameter $\mu$ in the following way:
\vspace{-1cm}
\begin{center}
$$\sigma(x_1,x_2,\kappa)  = 2 + \frac{2 \cdot \mu(x_1,x_2,\kappa)}{5}$$
\end{center}
The  methods are compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by the out-of-sample log-likelihood of the fitted models.\\
\\
Generally, a smooth parametric model such as GAMLSS should be the most successful method if the parameter function is smooth with no rapid changes. However, if there are any jumps or non-additive effects or interactions, the trees should be more effective in detecting these steps and therefore also lead to a better fitting forest. But this situation might also appear for smooth functions which contain sections with very steep slopes. Therefore, it is now of great interest, whether the concept of GAMLSS or of trees and forests covers a broader range of functions where it is superior to the other one.\\
\\
Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups. This is the point where the tree structure provides a better solution. However, a single tree can not keep up with the performance of \code{gamlss} when it comes to modeling the smooth functions within the subgroups. Offering a good compromise the forest takes advantages of the tree structure to find subgroups but can also model smooth effects if enough trees are included.\\
However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
\\
Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation. For $\kappa=1$ the performance of all three methods is similar. However, as $\kappa$ increases the RMSE of gamlss increases while the RMSE of trees and forests stays at almost the same level or even decreases sligthly as soon as $\kappa$ is greater or equal 2. The same conclusion can be drawn from looking at the log-likelihood where the resulting values of gamlss decrease rapidly while the values of trees and forests only decrease a little at the beginning but then increase.\\
This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become. Especially regarding the scale parameter the trees are clearly advantageous over the forests.

(As it has already been demonstrated in the first examples on single data sets the performances of \code{distforest} and \code{disttree} are not as successful as the one of \code{gamlss} for low values of $\kappa$ but remarkably improve as this values increases and are superior to \code{gamlss} for high values of $\kappa$. This is illustrated by an almost constant RMSE of \code{distforest} and \code{disttree} while the RMSE of \code{gamlss} increases. Comparing the tree and the forest model it can be observed that the forest seems to perform better for lower $\kappa$ while very steep sections lead to a better performance of trees.

In this section the results of the three methods are compared over an increasing value of $\kappa$. For each step they are applied on various data sets and the resulting values of the RMSE and the log-likelihood are averaged within each step. In that way it can be investigated how the results change for a changing value of $\kappa$.\\
Again, a normally distributed response variable and the same smooth parameter function are considered.\\

<<echo=FALSE, results=hide>>=
#### cross validation
if(file.exists("simres.rda")){
  load("simres.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/paper/gensim.R")
  source("gensim.R")
  simres <- gensim(seedconst = 7, nrep = 20, ntree = 100L,
                   cov.sep = FALSE,
                   fix.mu = FALSE,
                   fix.sigma = FALSE,
                   mu.sigma.interaction = FALSE,
                   gamboost_cvr = FALSE,
                   eval_disttree = TRUE,
                   eval_distforest = TRUE,
                   eval_gamlss = TRUE,
                   eval_bamlss = TRUE,
                   eval_gamboostLSS = TRUE,
                   eval_randomForest = TRUE,
                   eval_cforest = TRUE)
  save(simres, file = "simres.rda")
}
@

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<sim_rmse_exp, fig=TRUE, echo=FALSE>>=
plot_rmse(simres, type = "exp")
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<sim_rmse_par, fig=TRUE, echo=FALSE>>=
plot_rmse(simres, type = "par")
@
\end{subfigure}
\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE)\\
left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
right: continuous lines represent the location parameter, dotted lines the scale parameter}
\end{figure}

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<sim_ll, fig=TRUE, echo=FALSE>>=
plot_ll(simres)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<sim_levelplot, fig=TRUE, echo=FALSE>>=
levelfun <- function(x) simres$fun(x,5)

ncuts <- 100
colR <- colorRampPalette(c("blue","green"))                                       
cols <- colR(ncuts+1)
griddata <- expand.grid(list(x1 = seq(-0.4, 1, 0.01), x2 = seq(-10, 10, 0.2)))
griddata$mu <- levelfun(griddata)
levelplot(mu ~ x1 * x2, data = griddata, xlab = "x1", ylab = "x2", 
          region = TRUE, cuts = ncuts, col.regions = cols)

@
\end{subfigure}
\caption{\label{plot_sim_ll}left: results of the simulation study: log-likelihood\\ %\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest} vs. %\code{bamlss} vs. \code{gamboostLSS}\\
right: levelplot of the parameter function for $\kappa=5$}
\end{figure}

%\newpage
%\section{References}

%Hothorn T, Hornik K, Zeileis A (2006). 
%``Unbiased Recursive Partitioning: A Conditional Inference Framework.''
%\textit{Journal of Computational and Graphical Statistics}, 
%\textbf{15}(3), 651--674. 
%\doi{10.1198/106186006X133933}\\
%\\
%Zeileis A, Hothorn T, Hornik K (2008). 
%``Model-Based Recursive Partitioning.''
%\textit{Journal of Computational and Graphical Statistics}, 
%\textbf{17}(2), 492--514. 
%\doi{10.1198/106186008X319331}\\
%\\
%Hothorn T, Zeileis A (2015). 
%``partykit: A Modular Toolkit for Recursive Partytioning in R.''
%\textit{Journal of Machine Learning Research}, 
%\textbf{16}, 3905--3909. 
%\url{http://www.jmlr.org/papers/v16/hothorn15a.html}\\
%\\
%Stasinopoulos DM, Rigby RA (2007).
%``Generalized Additive Models for Location Scale and Shape (GAMLSS) in R.''
%\textit{Journal of Statistical Software}, 
%\textbf{23}(7), 1--46.
%\doi{10.18637/jss.v023.i07}\\
%\\
%Stasinopoulos DM, Rigby RA (2005).
%``Generalized Additive Models for Location Scale and Shape (with discussion)''
%\textit{Applied Statistics},
%\textbf{54}(3), 507--554.
%\doi{10.1111/j.1467-9876.2005.00510.x}\\
%\\
%Seibold H, Zeileis A, Hothorn T (2017).
%``Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.''
%\textit{Statistical Methods in Medical Research}, 
%\textbf{12}(1), 45--63.
%\doi{10.1177/0962280217693034}\\
%\\
%Hothorn T, Zeileis A (2017).
%``Transformation Forests.''
%\emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
%\url{http://arxiv.org/abs/1701.02110}\\
%\\
%Messner JW, Mayr GJ, Zeileis A (2016).
%``Heteroscedastic Censored and Truncated Regression with {crch}''
%\textit{The R Journal},
%\textbf{8}(1), 173--181.
%\url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}\\
%\\
%Zeileis A, Wiel MA, Hornik K, Hothorn T (2008).
%``Implementing a {Class} of {Permutation} {Tests}: {The} {Coin} {Package}''
%\textit{Journal of Statistical Software},
%\textbf{28}(8), 1--23.
%\doi{10.18637/jss.v028.i08}\\
%%publisher={American Statistical Association}
%\\
%Zeileis A, Hothorn T, Hornik K (2010).
%``Party with the {MOB}: {Model-Based} {Recursive} {Partitioning} in {R}''
%\textit{R package version 0.9-9999}.
%\url{https://cran.r-project.org/web/packages/party/vignettes/MOB.pdf}\\
%\\
%Breiman L (2001).
%``Random {Forests}''
%\textit{Machine Learning},
%\textbf{45}(1), 5--32
%%\doi{10.1023/A:1010933404324}
%%publisher={Springer}\\
%\\
%Loh WY (2011).
%``Classification and {Regression} {Trees}''
%\textit{Wiley Interdisciplinary Reviews: Data Mining and Knowledge Discovery},
%\textbf{1}(1), 14--23.
%\doi{10.1002/widm.8}
%%publisher={Wiley Online Library}\\
%\\
%Hothorn T (Maintainer), Hornik K, Strobl C, Zeileis A (2015).
%``Package 'party' ''
%\textit{Package Reference Manual for Party Version 0.9--0.998},
%\textbf{16}, 37.\\
%%  \doi{}
%\\
%Yee TW (2010).
%``The {VGAM} {Package} for {Categorical} {Data} {Analysis}''
%\textit{Journal of Statistical Software},
%\textbf{32}(10), 1--34.
%\url{http://www.jstatsoft.org/v32/i10/}\\
%\\
%Umlauf N, Klein N, Zeileis A (2017).
%``{BAMLSS}: {B}ayesian Additive Models for Location, Scale
%  and Shape (and Beyond)''
%\textit{Working Papers in Economics and Statistics, Research
%  Platform Empirical and Experimental Economics, Universit\"at
%  Innsbruck},
%\textbf{2017-05},
%\url{http://EconPapers.RePEc.org/RePEc:inn:wpaper:2017-05}\\
%%type = {Working Paper},
%%number = {2017-05},
%%month = {February},
%\\
%Benjamin Hofner and Andreas Mayr and Nora Fenske and
%  Matthias Schmid (2017). 
%``{gamboostLSS}: Boosting Methods for {GAMLSS} Models''
%\textit{{R} package version 2.0-0}.
%\url{https://CRAN.R-project.org/package=gamboostLSS}\\
%\\
%Liaw A, Wiener M (2002).
%``Classification and Regression by randomForest''
%\textit{R News},
%\textbf{2}(3), 18--22.
%\url{http://CRAN.R-project.org/doc/Rnews/}

%\newpage
\nocite{hothorn2006unbiased, zeileis2008model, hothorn2015partykit, Messner2016, loh2011classification, stasinopoulos2007generalized, Seibold2016, Seibold2017, breiman2001random, stasinopoulos2005, yee2010vgam}
% zeileis2008implementing, 
% zeileis2010party, 
% hothorn2015package,

\newpage
\bibliography{ref.bib}

\end{document}

### 2017-08-31 pkg/disttree/inst/draft/draft.Rnw
edited abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1550 255f4249-2263-4417-b1ea-8498779e27a7

@@
 In distributional forests an ensemble of distributional trees is built and used to calculate weights which are then included in the fitting process of a distributional model.
+These forest models can detect smooth effects as well as abrupt changes and interactions in a reasonable way without needing any kind of variable selection or information about the expected effects advance and therefore offers a good compromise particularly in complex settings.
 }
@@
   simres <- gensim(seedconst = 7, nrep = 20, ntree = 100L,
-                   cov.sep = FALSE,
+                   cov.sep = TRUE,
                    fix.mu = FALSE,

### 2017-09-09 pkg/disttree/inst/draft/draft.Rnw
added optional argument 'noise_sd' to set the standard deviation of the noise variable within the data generating process 'dgp()'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1567 255f4249-2263-4417-b1ea-8498779e27a7

@@
   #        fun          ..... parameter function
+  #        noise_sd     ..... optional parameter to set the standard deviation of the noise variable
   #
@@
                   par.matrix = matrix(nrow = 3, ncol = 2), 
-                round.sp = 3)
+                  round.sp = 3,
+                  noise_sd = 0)
   {
@@
 
+    if(noise_sd > 0) d$y <- d$y + rnorm(n, mean = 0, sd = noise_sd)
+
     return(d)
@@
 ### plotting function for one data set
-plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE, all.var = FALSE) 
+plot1 <- function(learndata, 
+                  dt=NULL, df=NULL, g=NULL, gb=NULL, b=NULL, rf=NULL, cf=NULL, 
+                  mu=TRUE, sigma=TRUE) 
 {
@@
 The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
-This development in the field of parametric modeling is illustrated in figure \ref{Devel_parmod}.\\
+This development in the field of parametric modeling is illustrated in Figure \ref{Devel_parmod}.\\
 
@@
 Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node.
-One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of figure \ref{Devel_treeforest}.
+One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure \ref{Devel_treeforest}.
 One way how to tackle this problem is to not just consider one tree but an ensemble of trees.
@@
 In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
-These (future) steps are illustrated in figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
+These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
 \\
@@
 \\
-In figure \ref{plot_mcycle} the results of a distributional forest are plotted on the top left and then compared to the results of a distributional tree, a GAMLSS and a random forest, regarding the location parameter on the top right and the scale parameter below.\\
+In Figure \ref{plot_mcycle} the results of a distributional forest are plotted on the top left and then compared to the results of a distributional tree, a GAMLSS and a random forest, regarding the location parameter on the top right and the scale parameter below.\\
 While the smooth function of \code{gamlss} seems to model especially the middle part (for \code{times} between 17 and 40) better than the wiggly functions resulting from the trees it can also be observed that at \code{times} $=14$ there is a rapid drop which is captuerd more precisely by the tree-based models while \code{gamlss} starts descending earlier.
@@
 \section{Application}
-\subsection{Continuation of the mcycle example: Cross Validation}
+\subsection{Continuation of the mcycle example: cross validation}
 In the introduction the four functions \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest} have been applied on the mcycle data set and their results haven been visualized and discussed.
@@
 This procedure is repeated 9 times.
-This comparison is illustrated in figure \ref{plot_cross_mean}, based on the RMSE on the left and based on the log-likelihood on the right.
+This comparison is illustrated in Figure \ref{plot_cross_mean}, based on the RMSE on the left and based on the log-likelihood on the right.
 (Extrapolation of the GAMLSS would lead to very small values of the log-likelihood.
@@
   source("gensim.R")
-  simres <- gensim(seedconst = 7, nrep = 20, ntree = 100L,
+  simres <- gensim(seedconst = 7423, nrep = 20, ntree = 100L,
                    cov.sep = TRUE,

### 2017-09-10 pkg/disttree/inst/draft/draft.Rnw
small bugfixes and additional parameter 'censpar'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1574 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-### plotting function for one data set
-plot1 <- function(learndata, 
-                  dt=NULL, df=NULL, g=NULL, gb=NULL, b=NULL, rf=NULL, cf=NULL, 
-                  mu=TRUE, sigma=TRUE) 
-{
-  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
-                    df$fitted.par,
-                    dt$fitted.par,
-                    #coef(dt)[paste(dt$fitted[,1]),],
-                    g$mu.fv, g$sigma.fv)
-
-  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
-                          "fitted.mu.df","fitted.sigma.df",
-                          "fitted.mu.dt","fitted.sigma.dt",
-                          "fitted.mu.g","fitted.sigma.g")
-  sp <- plotdata[order(plotdata["x1"]),]
-
-  if(!(any(dt.par, df.par, g.par, all.par, all.var))){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-  col = transpgrey, border = "transparent")
-    #lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
-    #       col = c('black', transpgrey), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # disttree
-  if(dt.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", col.main = pal["tree"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
-  col = pallight["tree"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-    #       col = c('black','grey', pal["tree"], pallight["tree"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # distforest
-  if(df.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", col.main = pal["forest"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
-  col = pallight["forest"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
-    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-    #       col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # gamlss
-  if(g.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
-  col = pallight["gamlss"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
-    #       col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # disttree vs. distforest vs. gamlss
-  if(all.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
-    #                    TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
-    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 13.95, expression(mu), bty = "n")
-  }
-
-  # variance: disttree vs. distforest vs. gamlss
-  if(all.var){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'black')
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
-  col = pallight["tree"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
-  col = pallight["forest"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
-  col = pallight["gamlss"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
-    #                    TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
-    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-  legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-}
-
 ## function to estimate standard deviation of randomForest for a new observation
@@
 
-The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and two covariates $X_1$ and $X_2$.\\
+The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and one or both of the two covariates $X_1$ and $X_2$.\\
 The covariates are generated based on the following distribution assumptions:
@@
 \end{center}
-The distribution of the response variable is a normal distribution where the corresponding distribution parameters are defined by parameter functions depending on the covariates. 
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% 1 covariate
+In a first example the distribution of the response variable is a left censored normal distribution with censoring point 0 where the corresponding distribution parameters are defined by a parameter function $f$ depending on one covariate only.
+
+\begin{center}
+$$ Y \sim \text{c}\mathcal{N} (f(x_1, \kappa))$$
+with
+$$f(x_1, \kappa) = \left(\mu(x_1,\kappa) , \sigma(x_2,\kappa)\right)$$ 
+\end{center}
+
+An additional factor $\kappa$ defines the steepness of certain sections of the parameter function.
+For low values of $\kappa$ the resulting parameters change rather slowly over varying values of the covariates while high values of $\kappa$ lead to abrupt changes.
+\\
+
 \vspace{-1cm}
+\begin{center}
+$$
+\mu(x_1,\kappa) = 
+\begin{cases}
+    4 + 8 \cdot \exp{(-(4 \cdot x_1-0.7)^{(2\cdot \kappa)})},& \text{if } x_1 \geq 0\\
+    0,              & \text{if } x_1 < 0
+\end{cases}
+$$
+
+$$
+\sigma(x_1,\kappa) = 
+\begin{cases}
+    1+3 \cdot \|x_1\|,& \text{if } x_1 \geq 0\\
+    0.00001,              & \text{if } x_1 < 0
+\end{cases}
+$$
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% 2 covariates
+In a second example we consider a normaly distributed response variable with the following parameter function depending on two covariates.
+%\vspace{-1cm}
 \begin{center}
@@
 
-An additional factor $\kappa$ defines the steepness of certain sections of the parameter function.
-For low values of $\kappa$ the resulting parameters change rather slowly over varying values of the covariates while high values of $\kappa$ lead to abrupt changes.
-\\
-Including non-additive effects creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated.
+Including non-additive effects and interactions creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated.
 Here a non-additive effects is created by separating the generated data into two subgroups depending on the value of $X_2$.

### 2017-09-10 pkg/disttree/inst/draft/gensim.R

@@ -23,6 +23,7 @@ gensim <- function(seedconst = 7, nrep = 100, ntree = 100,
                    noise_sd = 0,
                    onecov = TRUE,
                    censNO = TRUE,
+                   censpar = 0.7,
                    cov.sep = FALSE,
                    fix.mu = FALSE,
                    fix.sigma = FALSE,
@@ -694,7 +695,7 @@ gensim <- function(seedconst = 7, nrep = 100, ntree = 100,
     } else {
       if(fix.sigma){
         fun <- function(x, kappa){
-          mu <- 4 + 8 * (exp(-(4*x[,1]-0.7)^(2*kappa)))
+          mu <- 4 + 8 * (exp(-(4*x[,1]-censpar)^(2*kappa)))
           #mu <- 2 + 10 * (exp(-(4*x[,1]-2)^(2*kappa)))
           if(censNO) mu[x[,1]<0] <- 0
           sigma <- 3
@@ -705,7 +706,7 @@ gensim <- function(seedconst = 7, nrep = 100, ntree = 100,
       } else {
         if(mu.sigma.interaction){
           fun <- function(x, kappa){
-            mu <- 4 + 8 * (exp(-(4*x[,1]-0.7)^(2*kappa)))
+            mu <- 4 + 8 * (exp(-(4*x[,1]-censpar)^(2*kappa)))
             #mu <- 2 + 10 * (exp(-(4*x[,1]-2)^(2*kappa)))
             if(censNO) mu[x[,1]<0] <- 0
             sigma <- 2 + mu/4
@@ -715,7 +716,7 @@ gensim <- function(seedconst = 7, nrep = 100, ntree = 100,
           }
         } else {
           fun <- function(x, kappa){
-            mu <- 4 + 8 * (exp(-(4*x[,1]-0.7)^(2*kappa)))
+            mu <- 4 + 8 * (exp(-(4*x[,1]-censpar)^(2*kappa)))
             #mu <- 2 + 10 * (exp(-(4*x[,1]-2)^(2*kappa)))
             if(censNO) mu[x[,1]<0] <- 0
             sigma <- 1+3*abs(x[,1])

### 2017-09-13 pkg/disttree/inst/draft/draft.Rnw
additional arguments 'legend' and 'ylim' in plotting functions 'plot_ll' and 'plot_rmse'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1576 255f4249-2263-4417-b1ea-8498779e27a7

@@
   # (in randomForest the argument 'keep.inbag' must be set to TRUE)
-rf_getsd <- function(object, newobs, rfdata){
-
+  rf_getsd <- function(rf, newdata, rfdata){
+
+    rf_sd <- numeric(length = NROW(newdata))
+
+    for(k in 1:NROW(newdata)){
+      newobs <- newdata[k,]
       # get predictions for the new observations from all trees
-  pred.newobs <- predict(object, predict.all = TRUE, newdata = newobs)
+      pred.newobs <- predict(rf, predict.all = TRUE, newdata = newobs)
 
       # vector where the standard deviations from all trees are stored
-  sd_trees <- numeric(length = object$ntree)
+      sd_trees <- numeric(length = rf$ntree)
 
       # loop over all trees of the forest
-  for(i in 1:object$ntree){
+      for(i in 1:rf$ntree){
 
         # get data used to build this tree
-    obsid <- rep.int(c(1:NROW(rfdata)), as.vector(object$inbag[,i]))
+        obsid <- rep.int(c(1:NROW(rfdata)), as.vector(rf$inbag[,i]))
         obs_tree <- rfdata[obsid,]
@@
         # get predictions for this data from this tree
-    pred.obs_tree <- predict(object, newdata = obs_tree, predict.all = TRUE)$individual[,i]
+        pred.obs_tree <- predict(rf, newdata = obs_tree, predict.all = TRUE)$individual[,i]
 
@@
 
-    sd_trees[i] <- sd(obs_node$accel)
+        sd_trees[i] <- sd(obs_node$y)
       }
@@
       sd_newobs <- mean(sd_trees, na.rm = TRUE)
-
-  return(sd_newobs)   
+      rf_sd[k] <- sd_newobs
+    }
+    return(rf_sd)   
   }
@@
 # plot RMSE
-plot_rmse <- function(simres, type = c("exp", "par")){
+plot_rmse <- function(simres, type = c("exp", "par"), ylim = NULL, legend = TRUE){
 
@@
     colnames(rmse) <- colnames
-    ylim <- c(min(rmse), max(rmse))
+    if(is.null(ylim)) ylim <- c(min(na.omit(rmse)), max(na.omit(rmse)))
 
@@
     }
-    legend('topleft', legendnames, 
+    if(legend) legend('topleft', legendnames, 
            col = col, lty = 1, cex = 0.7)
@@
 # plot loglikelihood
-plot_ll <- function(simres){
+plot_ll <- function(simres, ylim = NULL, legend = TRUE){
 
@@
     colnames(ll) <- colnames
-    ylim <- c(min(ll), max(ll))
+    if(is.null(ylim)) ylim <- c(min(na.omit(ll)), max(na.omit(ll)))
 
@@
   ## get 'fitted.sigma.rf'
-  rf_sd <- numeric(length(NROW(mcycle)))
-
-  for(i in 1:NROW(mcycle)){
-    rf_sd[i] <- rf_getsd(rf, newobs = mcycle[i,], rfdata = mcycle)
-  }
+  rf_sd <- rf_getsd(rf, newdata = mcycle, rfdata = mcycle)
 
@@
       rf.pred.resp <- predict(rf, newdata = testdata)
-      rf.sd <- numeric(length(NROW(testdata)))
-      for(l in 1:NROW(testdata)){
-        rf.sd[l] <- rf_getsd(rf, newobs = testdata[l,], rfdata = learndata)
-      }
+
+      rf.sd <- rf_getsd(rf, newdata = testdata, rfdata = learndata)
+
       g.pred.par <- predict(ga, newdata = testdata, type = "response")

### 2017-09-19 pkg/disttree/inst/draft/draft.Rnw
adapted formula and text to the new simulation function (results not updated yet)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1581 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-######## data generating process based on a predefined tree or a parameter function
-  # if tree: 2 splits => 3 terminal nodes
-  # 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
-  #
-  # input: n            ..... nr of observations
-  #        family       ..... distribution of the generated observations
-  #        split.matrix ..... indices of the variables used for the splits together with the corresponding split points
-  #        parm.matrix  ..... set of distribution parameters for each subgroup
-  #        fun          ..... parameter function
-  #        noise_sd     ..... optional parameter to set the standard deviation of the noise variable
-  #
-  # output: data.frame with generated observations y (generated seperatly in each subgroups with given distributions parameters),
-  #                         the given split variables x1, ..., x10 for each observation and
-  #                         the index of the subgroup for each observation or
-  #                         the distribution parameter for each observation
-
-  # FIX: until now only complete lists can be handed over (dist_list...)
-  dgp <- function(n, family = dist_list_normal, 
-                  fun = NULL,
-                  split.matrix = matrix(nrow = 2, ncol = 2), 
-                  par.matrix = matrix(nrow = 3, ncol = 2), 
-                  round.sp = 3,
-                  noise_sd = 0)
-  {
-
-    # generating the possible split variables
-    x1 <- runif(n,-0.4,1)
-    x2 <- runif(n,-10,10)
-    x3 <- runif(n,0,100)
-    x4 <- x1 + rnorm(n, sd = 0.1)
-    x5 <- x1 + rnorm(n, sd = 0.3)
-    x6 <- rbinom(n,1,0.5)
-    x7 <- rbinom(n,1,0.5)
-    x8 <- sample(1:4, n, replace = TRUE) 
-    x9 <- sample(1:5, n, replace = TRUE) 
-    x10 <- sample(1:7, n, replace = TRUE)
-    #x11 <- runif(n,-0.5,1)
-    x <- cbind(x1,x2,x3,x4,x5,x6,x7,x8,x9,x10)
-    # reduce nr of possible split points by rounding values of split variables
-    x <- round(x, digits = round.sp)
-
-    y <- vector(mode = "numeric", length = n)
-
-    # getting the random function
-    if(is.function(family)) family <- family()
-    if(inherits(family, "gamlss.family")) {
-      rfun <- get(paste0("r",family$family[[1]]))
-    } else {
-      if(family$family.name == "Normal Distribution") rfun <- rNO
-      if(family$family.name == "censored Normal Distribution") rfun <- rNO
-      if(family$family.name == "Poisson Distribution") rfun <- rPO
-      if(family$family.name == "Weibull Distribution") rfun <- survival:::rsurvreg 
-    }
-
-    if(is.null(fun)) {
-      index <- vector(mode = "numeric", length = n)
-      sv = split.matrix[,1]    # index of split variables
-      sp = split.matrix[,2]    # split points
-      par = par.matrix
-
-      # splitting based on the given split variables and split points and
-      # generating the observations seperatly for each subgroup (with the corresponding given distribution parameters)
-      for(k in 1:n){
-        if(x[k, sv[1]] <= sp[1]){
-          y[k] <- do.call(rfun, as.list(c(1,par[1,])))
-          index[k] <- 2L
-        } else {
-          if(x[k, sv[2]] <= sp[2]){
-            y[k] <- do.call(rfun, as.list(c(1,par[2,]))) 
-            index[k] <- 4L
-          } else {
-            y[k] <- do.call(rfun, as.list(c(1,par[3,])))
-            index[k] <- 5L
-          }
-        }
-      }
-      d <- as.data.frame(cbind(y, x, index))
-      colnames(d) <- c("y", paste0("x", c(1:10)), "index")
-
-    } else {
-
-      dpar <- fun(x)
-
-      if(!(family$family.name == "Poisson Distribution")){
-        y <- rfun(n, dpar[,1], dpar[,2])
-        d <- as.data.frame(cbind(y, x, dpar))
-        colnames(d) <- c("y", paste0("x", c(1:10)), "mu", "sigma")
-      } else {
-        y <- rfun(n, dpar)
-        d <- as.data.frame(cbind(y, x, dpar))
-        colnames(d) <- c("y", paste0("x", c(1:10)), "lambda")
-      }
-    }
-
-    if(family$family.name == "censored Normal Distribution") {
-      #d$ystar <- d$y
-      d$y <- pmax(d$y, 0)
-    }
-
-    if(noise_sd > 0) d$y <- d$y + rnorm(n, mean = 0, sd = noise_sd)
-
-    return(d)
-  }
-
 ## function to estimate standard deviation of randomForest for a new observation
@@
 
-\title{\texttt{disttree}: Distributional Trees and Forests}
+\title{Distributional Trees and Forests}
 \Plaintitle{disttree: Distributional Trees and Forests}
@@
 \begin{itemize}%[leftmargin=0.5cm]
-\item When applying a tree algorithm the use of a fitting function which returns scores is advantageous to find correct splits.
+\item When applying a tree algorithm the use of a fitting function which returns scores for each distribution parameter is advantageous to find correct splits.
 In particular it is crucial to find splits in parameters other than the location parameter.
-Hence a forest consisting of trees which were built using scores for all distribution parameters leads to better results than a simple forest considering only the mean.
-\item In the case of smooth parameter functions without any abrupt changes a distributional forest performs quite well, even though other models such as GAMLSS might be more successful in these situations.
-However, once the smooth parameter functions include very steep sections the distributional forest outperforms other models (and in extreme cases a distributional tree does as well).
-\item Once non-additive effects or interactions appear within the parameter functions distributional trees and forests clearly show their advantages over other models.
+Hence a distributional forest consisting of distributional trees leads to better results than a simple forest considering only the mean.
+\item In the case of smooth parameter functions without any abrupt changes / non-additive effects or interactions a smooth parameteric model such as GAMLSS is clearly the first choice while a stepfunction can be modelled best by applying a tree-structured model.
+Even though for both of these extreme situations a specificly advandageous method is provided, a forest model can be a good compromise and performs well in a wide range of settings including the two mentioned before.
+Therefore, a forest is a recommendable choice particularly when no information about the type of parameter function is provided.
+\item While specifying interactions and selecting covariates is necessary to achieve reasonable results in parameteric models such as GAMLSS, this is done automatically within trees and forests.
+For that reason, especially in case of a high number of covariates and no information about their effects on the response distributional, trees and forests provide a framework where no additional effort is required for the model specification.
 \end{itemize}
@@
 
-The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and one or both of the two covariates $X_1$ and $X_2$.\\
-The covariates are generated based on the following distribution assumptions:
+The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and ten covariates $X_1, \ldots, X_{10}$ which are all independent und identically distributed following a uniform distribution.
+$$
+X_i \sim \textit{Unif}([-1,1]) \qquad \forall i \in \{1,\ldots,10\}
+$$
+
+The distribution of the response variable is a left censored normal distribution with censoring point 0 where the corresponding distribution parameters are defined by a parameter function $f$ depending each on one covariate.
+The location parameter $\mu$ depends on $X_1$ and the scale parameter $\sigma$ depends on $X_2$.
+The other covariates $X_3, \ldots, X_{10}$ are noise variables.\\
+The latent variable $Y^{\ast}$ is generated based on the following setting
 \begin{center}
-\begin{tabular}{ l  c  l }
-$X_1$ & $\sim$ & \textit{Unif}([-0.4,1]) \\ 
-$X_2$ & $\sim$ & \textit{Unif}([-10,10])
-\end{tabular}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 1 covariate
-In a first example the distribution of the response variable is a left censored normal distribution with censoring point 0 where the corresponding distribution parameters are defined by a parameter function $f$ depending on one covariate only.
-
-\begin{center}
-$$ Y \sim \text{c}\mathcal{N} (f(x_1, \kappa))$$
+$$ Y^{\ast} \sim \mathcal{N} (f(X_1, X_2, \kappa))$$
 with
-$$f(x_1, \kappa) = \left(\mu(x_1,\kappa) , \sigma(x_2,\kappa)\right)$$ 
+$$f(X_1, X_2, \kappa) = \left(\mu(X_1,\kappa) , \sigma(X_2,\kappa)\right)$$ 
 \end{center}
@@
 $$
-\mu(x_1,\kappa) = 
-\begin{cases}
-    4 + 8 \cdot \exp{(-(4 \cdot x_1-0.7)^{(2\cdot \kappa)})},& \text{if } x_1 \geq 0\\
-    0,              & \text{if } x_1 < 0
+\mu(X_1,\kappa) = \mu_{\text{base}} + 8 \cdot \exp{(-(3 \cdot X_1-1)^{(2\cdot \kappa)})}
+$$
+
+$$
+\sigma(X_2,\kappa) = 2 + 2 \cdot (1-\Lambda(\kappa^{1.8} \cdot 5 \cdot (X_2+0.5)))
+$$
+where $\Lambda$ is the cumulative distribution function of the logistic distribution
+$$
+\Lambda(\omega) = \frac{1}{1 + \exp(-\omega)}
+$$
+The response variable $Y$ is defined as 
+$$
+Y = \begin{cases}
+    Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
+    0,       & \text{if } else
 \end{cases}
 $$
-
-$$
-\sigma(x_1,\kappa) = 
-\begin{cases}
-    1+3 \cdot \|x_1\|,& \text{if } x_1 \geq 0\\
-    0.00001,              & \text{if } x_1 < 0
-\end{cases}
-$$
 \end{center}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 2 covariates
-In a second example we consider a normaly distributed response variable with the following parameter function depending on two covariates.
-%\vspace{-1cm}
-\begin{center}
-$$ Y \sim \mathcal{N} (f(x_1, x_2, \kappa))$$
-with
-$$f(x_1, x_2, \kappa) = \left(\mu(x_1,x_2,\kappa) , \sigma(x_1,x_2,\kappa)\right)$$ 
-\end{center}
-
-Including non-additive effects and interactions creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated.
+%Including non-additive effects and interactions creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated.
 Here a non-additive effects is created by separating the generated data into two subgroups depending on the value of $X_2$.
@@
 Depending on the subgroup the actual smooth parameter function is defined.
-\vspace{-1cm}
-\begin{center}
-$$\mu(x_1,x_2,\kappa)  = 10 \cdot \Big(1-s(x_1,x_2,\kappa)+ \big(2\cdot s(x_1,x_2,\kappa) - 1\big) \cdot \text{\code{rbinom}}\left(n,1,P_1\right)\Big)
-$$
-with the distribution function of the logistic distribution as a smooth function
-$$
-s(x_2,\kappa) = \text{\code{plogis}}\left(\kappa^2 \cdot \frac{x_2 - 3}{10}\right)
-$$
-and the probability of being in subgroup 1
-$$
-P_1(x_1,\kappa) = e^{-(10\cdot x_1 - 5)^{2 \cdot \kappa}}.
-$$
-\end{center}
-In this scenerio the scale parameter $\sigma$ depends on the location parameter $\mu$ in the following way:
-\vspace{-1cm}
-\begin{center}
-$$\sigma(x_1,x_2,\kappa)  = 2 + \frac{2 \cdot \mu(x_1,x_2,\kappa)}{5}$$
-\end{center}
+%\vspace{-1cm}
+%\begin{center}
+%$$\mu(x_1,x_2,\kappa)  = 10 \cdot \Big(1-s(x_1,x_2,\kappa)+ \big(2\cdot s(x_1,x_2,\kappa) - 1\big) \cdot \text{\code{rbinom}}\left(n,1%,P_1\right)\Big)
+%$$
+%with the distribution function of the logistic distribution as a smooth function
+%$$
+%s(x_2,\kappa) = \text{\code{plogis}}\left(\kappa^2 \cdot \frac{x_2 - 3}{10}\right)
+%$$
+%and the probability of being in subgroup 1
+%$$
+%P_1(x_1,\kappa) = e^{-(10\cdot x_1 - 5)^{2 \cdot \kappa}}.
+%$$
+%\end{center}
+%In this scenerio the scale parameter $\sigma$ depends on the location parameter $\mu$ in the following way:
+%\vspace{-1cm}
+%\begin{center}
+%$$\sigma(x_1,x_2,\kappa)  = 2 + \frac{2 \cdot \mu(x_1,x_2,\kappa)}{5}$$
+%\end{center}
+
 The  methods are compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by the out-of-sample log-likelihood of the fitted models.\\
@@
 \\
-Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups.
+%Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups.
 This is the point where the tree structure provides a better solution.
@@
 Offering a good compromise the forest takes advantages of the tree structure to find subgroups but can also model smooth effects if enough trees are included.\\
-However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
-\\
-Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation.
+%However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
+%\\
+%Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation.
 For $\kappa=1$ the performance of all three methods is similar.
@@
 The same conclusion can be drawn from looking at the log-likelihood where the resulting values of gamlss decrease rapidly while the values of trees and forests only decrease a little at the beginning but then increase.\\
-This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
-For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become.
+%This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
+%For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become.
 Especially regarding the scale parameter the trees are clearly advantageous over the forests.
 
-(As it has already been demonstrated in the first examples on single data sets the performances of \code{distforest} and \code{disttree} are not as successful as the one of \code{gamlss} for low values of $\kappa$ but remarkably improve as this values increases and are superior to \code{gamlss} for high values of $\kappa$.
+%(As it has already been demonstrated in the first examples on single data sets the performances of \code{distforest} and \code{disttree} are not as successful as the one of \code{gamlss} for low values of $\kappa$ but remarkably improve as this values increases and are superior to \code{gamlss} for high values of $\kappa$.
 This is illustrated by an almost constant RMSE of \code{distforest} and \code{disttree} while the RMSE of \code{gamlss} increases.

### 2017-09-20 pkg/disttree/inst/draft/draft.Rnw
adapted further parts of the simulation section to the new simulation setup and added section 'rain'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1585 255f4249-2263-4417-b1ea-8498779e27a7

@@
 In distributional forests an ensemble of distributional trees is built and used to calculate weights which are then included in the fitting process of a distributional model.
-These forest models can detect smooth effects as well as abrupt changes and interactions in a reasonable way without needing any kind of variable selection or information about the expected effects advance and therefore offers a good compromise particularly in complex settings.
+These forest models can detect smooth effects as well as abrupt changes and interactions in a reasonable way without needing any kind of variable selection or information a!
+ bout the expected effects advance and therefore offers a good compromise particularly in complex settings. 
 }
@@
 A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set.
-Then the results of all trees are combined to get a final model which unlike a tree model can also deal with smooth effects in a reasonable way.\\
+Then the results of all trees are combined to get a final model which unlike a tree model c!
+ an also deal with smooth effects in a reasonable way.\\ 
 In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
@@
   ga <- gam(list(accel ~ s(times,k=20,bs="ad"), ~ s(times)), data = mcycle, family=gaulss())
-  #b <- bamlss(list(accel ~ s(times), accel ~ s(times)), data = mcycle)save(df, file = "mcycle_plotdata.rda")
+  #b <- bamlss(list(accel ~ s(times), accel ~ s(times)), data = mcycle)
+  #save(df, file = "mcycle_plotdata.rda")
 
@@
 (Extrapolation of the GAMLSS would lead to very small values of the log-likelihood.
-To avoid this, the first and the last observation are always included in the learn data set.) \\
+To avoid this, the first and the l!
+ ast observation are always included in the learn data set.) \\
 As we have already seen in the introduction the distributional tree obviously struggles with the smooth effects while the two forest models can capture the changes quite well.
@@
 \section{Simulation Study}
-Within this simulation study the performances of distributional trees and forests are tested and compared to other models, showing situations where one or the other method is more advantageous than others.
+Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.
 By doing so the following hypotheses are investigated:
@@
 Hence a distributional forest consisting of distributional trees leads to better results than a simple forest considering only the mean.
-\item In the case of smooth parameter functions without any abrupt changes / non-additive effects or interactions a smooth parameteric model such as GAMLSS is clearly the first choice while a stepfunction can be modelled best by applying a tree-structured model.
+\item In the case of smooth parameter functions without any abrupt changes / non-additive effects a smooth parameteric model such as GAMLSS is clearly the first choice while a stepfunction can be modelled best by applying a tree-structured model.
 Even though for both of these extreme situations a specificly advandageous method is provided, a forest model can be a good compromise and performs well in a wide range of settings including the two mentioned before.
-Therefore, a forest is a recommendable choice particularly when no information about the type of parameter function is provided.
+(Therefore, a forest is a recommendable choice particularly when no information about the type of parameter function is provided.)
 \item While specifying interactions and selecting covariates is necessary to achieve reasonable results in parameteric models such as GAMLSS, this is done automatically within trees and forests.
-For that reason, especially in case of a high number of covariates and no information about their effects on the response distributional, trees and forests provide a framework where no additional effort is required for the model specification.
+For that reason, distributional trees and forests provide a framework where no additional effort is required for the model specification.
+This feature plays to its strength especially in case of a high number of covariates and no information about their effects on the response.
 \end{itemize}
 
-The following methods are compared to distributional trees and forests:
+%Generally, a smooth parametric model such as GAMLSS should be the most successful method if the parameter function is smooth with no rapid changes.
+However, if there are any jumps or non-additive effects or interactions, the trees should be more effective in detecting these steps and therefore also lead to a better fitting forest.
+But this situation might also appear for smooth functions which contain sections with very steep slopes.
+Therefore, it is now of great interest, whether the concept of smooth parametric models or of trees and forests covers a broader range of functions where it is superior to the other.\\
+
+The following methods are applied and compared to distributional trees and forests:
 \begin{itemize}
@@
 
-The distribution of the response variable is a left censored normal distribution with censoring point 0 where the corresponding distribution parameters are defined by a parameter function $f$ depending each on one covariate.
-The location parameter $\mu$ depends on $X_1$ and the scale parameter $\sigma$ depends on $X_2$.
-The other covariates $X_3, \ldots, X_{10}$ are noise variables.\\
-The latent variable $Y^{\ast}$ is generated based on the following setting
-\begin{center}
-$$ Y^{\ast} \sim \mathcal{N} (f(X_1, X_2, \kappa))$$
-with
-$$f(X_1, X_2, \kappa) = \left(\mu(X_1,\kappa) , \sigma(X_2,\kappa)\right)$$ 
-\end{center}
-
-An additional factor $\kappa$ defines the steepness of certain sections of the parameter function.
+The distribution of the response variable is a left censored normal distribution with censoring point 0.
+The corresponding distribution parameters are defined by a parameter function $f$ depending on the covariates and an additional factor $\kappa$ which regulates the steepness of certain sections of the parameter function.
 For low values of $\kappa$ the resulting parameters change rather slowly over varying values of the covariates while high values of $\kappa$ lead to abrupt changes.
 \\
-
-\vspace{-1cm}
+The location parameter $\mu$ depends on $X_1$ and $\kappa$ and the scale parameter $\sigma$ depends on $X_2$ and $\kappa$. The other covariates $X_3, \ldots, X_{10}$ are used as noise variables.\\
+The latent variable $Y^{\ast}$ is generated based on the following setting.
 \begin{center}
 $$ 
+Y^{\ast} \sim \mathcal{N} (f(X_1, X_2, \kappa))
+$$
+with
+$$
+f(X_1, X_2, \kappa) = \left(\mu(X_1,\kappa) , \sigma(X_2,\kappa)\right)
+$$ 
+and
+$$
 \mu(X_1,\kappa) = \mu_{\text{base}} + 8 \cdot \exp{(-(3 \cdot X_1-1)^{(2\cdot \kappa)})}
@@
 $$
-Y = \begin{cases}
+Y = 
+\begin{cases}
     Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
@@
 
-%Including non-additive effects and interactions creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated.
-Here a non-additive effects is created by separating the generated data into two subgroups depending on the value of $X_2$.
-In particular, within the function $\mu$ each observation is assigned with a probability of being in subgroup 1.
-This probability is specified by a smooth function and used to generate a binary variable indicating in which subgroup the observation is.
-Depending on the subgroup the actual smooth parameter function is defined.
-%\vspace{-1cm}
-%\begin{center}
-%$$\mu(x_1,x_2,\kappa)  = 10 \cdot \Big(1-s(x_1,x_2,\kappa)+ \big(2\cdot s(x_1,x_2,\kappa) - 1\big) \cdot \text{\code{rbinom}}\left(n,1%,P_1\right)\Big)
-%$$
-%with the distribution function of the logistic distribution as a smooth function
-%$$
-%s(x_2,\kappa) = \text{\code{plogis}}\left(\kappa^2 \cdot \frac{x_2 - 3}{10}\right)
-%$$
-%and the probability of being in subgroup 1
-%$$
-%P_1(x_1,\kappa) = e^{-(10\cdot x_1 - 5)^{2 \cdot \kappa}}.
-%$$
-%\end{center}
-%In this scenerio the scale parameter $\sigma$ depends on the location parameter $\mu$ in the following way:
-%\vspace{-1cm}
-%\begin{center}
-%$$\sigma(x_1,x_2,\kappa)  = 2 + \frac{2 \cdot \mu(x_1,x_2,\kappa)}{5}$$
-%\end{center}
-
-The  methods are compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by the out-of-sample log-likelihood of the fitted models.\\
+To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied to one generated data set only where $\kappa$ is set to 1 (top plots) and then to a second data set where $\kappa$ is set to 10 (bottom plots).\\
+In the left plots the location parameter function $\mu(X_1, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $(x_2, x_3, \ldots, x_{10}) = (0,0, \ldots, 0)$.
 \\
-Generally, a smooth parametric model such as GAMLSS should be the most successful method if the parameter function is smooth with no rapid changes.
-However, if there are any jumps or non-additive effects or interactions, the trees should be more effective in detecting these steps and therefore also lead to a better fitting forest.
-But this situation might also appear for smooth functions which contain sections with very steep slopes.
-Therefore, it is now of great interest, whether the concept of GAMLSS or of trees and forests covers a broader range of functions where it is superior to the other one.\\
+In the right plots the scale parameter function $\sigma(X_2, \kappa)$ is plotted along the covariate $X_2$ together with the estimated values of the different models for fixed covariates $(x_1, x_3, \ldots, x_{10}) = (0.4, 0, \ldots, 0)$.
+\\
+
+<<echo=FALSE, results=hide>>=
+source("onecov.R")
+if(file.exists("onecov1.rda")){
+  load("onecov1.rda")
+} else {
+  onecov1 <- sim_onecov(kappa = 1, nobs = 700,
+                         seedconst = 7, ntree = 100,
+                         formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
+                         tree_minsplit = 40, tree_mincrit = 0.95,
+                         forest_minsplit = 20, forest_mincrit = 0, 
+                         type.tree = "ctree",
+                         mubase = 0,
+                         censNO = TRUE,
+                         fix.mu = FALSE,
+                         fix.sigma = FALSE,
+                         mu.sigma.interaction = FALSE,
+                         gamboost_cvr = FALSE,
+                         eval_disttree = TRUE,
+                         eval_distforest = TRUE,
+                         eval_gamlss = TRUE,
+                         eval_bamlss = TRUE,
+                         eval_gamboostLSS = TRUE,
+                         eval_randomForest = TRUE,
+                         eval_cforest = TRUE)
+  save(onecov1, file = "onecov1.rda")
+}
+
+if(file.exists("onecov10.rda")){
+  load("onecov10.rda")
+} else {
+  onecov10 <- sim_onecov(kappa = 10, nobs = 700,
+                         seedconst = 7, ntree = 100,
+                         formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
+                         tree_minsplit = 40, tree_mincrit = 0.95,
+                         forest_minsplit = 20, forest_mincrit = 0, 
+                         type.tree = "ctree",
+                         mubase = 0,
+                         censNO = TRUE,
+                         fix.mu = FALSE,
+                         fix.sigma = FALSE,
+                         mu.sigma.interaction = FALSE,
+                         gamboost_cvr = FALSE,
+                         eval_disttree = TRUE,
+                         eval_distforest = TRUE,
+                         eval_gamlss = TRUE,
+                         eval_bamlss = TRUE,
+                         eval_gamboostLSS = TRUE,
+                         eval_randomForest = TRUE,
+                         eval_cforest = TRUE)
+  save(onecov10, file = "onecov10.rda")
+}
+@
+
+\begin{figure}[t!]
+\begin{subfigure}{0.6\textwidth}
+<<sim_onecov1_mu, fig=TRUE, echo=FALSE>>=
+plot_onecov(learndata = onecov1$learndata, parfun = onecov1$fun, 
+            dt = onecov1$dt, df = onecov1$df, g = onecov1$g, b = onecov1$b, 
+            gb = onecov1$gb, rf = onecov1$rf, cf = onecov1$cf, 
+            compare_mu = TRUE, fix_x2 = 0,
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.6\textwidth}
+<<sim_onecov1_sigma, fig=TRUE, echo=FALSE>>=
+plot_onecov(learndata = onecov1$learndata, parfun = onecov1$fun, 
+            dt = onecov1$dt, df = onecov1$df, g = onecov1$g, b = onecov1$b, 
+            gb = onecov1$gb, rf = onecov1$rf, cf = onecov1$cf, 
+            compare_sigma = TRUE, fix_x1 = 0.4,
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+@
+\end{subfigure}
+\caption{\label{plot_sim_onecov1} ... }
+\end{figure}
+
+\begin{figure}[t!]
+\begin{subfigure}{0.6\textwidth}
+<<sim_onecov10_mu, fig=TRUE, echo=FALSE>>=
+plot_onecov(learndata = onecov10$learndata, parfun = onecov10$fun, 
+            dt = onecov10$dt, df = onecov10$df, g = onecov10$g, b = onecov10$b, 
+            gb = onecov10$gb, rf = onecov10$rf, cf = onecov10$cf, 
+            compare_mu = TRUE, fix_x2 = 0,
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.6\textwidth}
+<<sim_onecov10_sigma, fig=TRUE, echo=FALSE>>=
+plot_onecov(learndata = onecov10$learndata, parfun = onecov10$fun, 
+            dt = onecov10$dt, df = onecov10$df, g = onecov10$g, b = onecov10$b, 
+            gb = onecov10$gb, rf = onecov10$rf, cf = onecov10$cf, 
+            compare_sigma = TRUE, fix_x1 = 0.4,
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+@
+\end{subfigure}
+\caption{\label{plot_sim_onecov10} ... }
+\end{figure}
+
+Now, to get a more profound impression the methods are evaluated on various data sets.
+The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step ...
+data sets are generated.
+The methods are then compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by the out-of-sample log-likelihood of the fitted models, averaged within each step.
+(In that way it can be investigated how the results change for a changing value of $\kappa$.)
+
+<<echo=FALSE, results=hide>>=
+if(file.exists("simres.rda")){
+  load("simres.rda")
+} else {
+  #source("~/svn/partykit/pkg/disttree/inst/draft/.R")
+  source("gensim.R")
+  simres <- gensim(seedconst = 7, nrep = 40, ntree = 100,
+                   nsteps = 9, stepsize = 1, kappa.start = 1,
+                   formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
+                   nobs = 400, testnobs = 200L,
+                   tree_minsplit = 14, tree_mincrit = 0.95,
+                   forest_minsplit = 7, forest_mincrit = 0, 
+                   type.tree = "ctree",
+                   censNO = TRUE,
+                   fix.mu = FALSE,
+                   fix.sigma = FALSE,
+                   mu.sigma.interaction = FALSE,
+                   gamboost_cvr = FALSE,
+                   eval_disttree = TRUE,
+                   eval_distforest = TRUE,
+                   eval_gamlss = TRUE,
+                   eval_bamlss = TRUE,
+                   eval_gamboostLSS = TRUE,
+                   eval_randomForest = TRUE,
+                   eval_cforest = TRUE,
+                   mubase = 0)
+  save(simres, file = "simres.rda")
+}
+@
+
+\begin{figure}[t!]
+\begin{subfigure}{0.6\textwidth}
+<<sim_rmse_exp, fig=TRUE, echo=FALSE>>=
+plot_rmse(simres, type = "exp")
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.6\textwidth}
+<<sim_rmse_par, fig=TRUE, echo=FALSE>>=
+plot_rmse(simres, type = "par")
+@
+\end{subfigure}
+\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE)\\
+left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
+right: continuous lines represent the location parameter, dotted lines the scale parameter}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{subfigure}{0.6\textwidth}
+<<sim_ll, fig=TRUE, echo=FALSE>>=
+plot_ll(simres)
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.6\textwidth}
+<<sim_levelplot, fig=TRUE, echo=FALSE>>=
+levelfun <- function(x) simres$fun(x,5)
+
+ncuts <- 100
+colR <- colorRampPalette(c("blue","green"))                                       
+cols <- colR(ncuts+1)
+griddata <- expand.grid(list(x1 = seq(-0.4, 1, 0.01), x2 = seq(-10, 10, 0.2)))
+griddata$mu <- levelfun(griddata)
+levelplot(mu ~ x1 * x2, data = griddata, xlab = "x1", ylab = "x2", 
+          region = TRUE, cuts = ncuts, col.regions = cols)
+
+@
+\end{subfigure}
+\caption{\label{plot_sim_ll}left: results of the simulation study: log-likelihood\\ %\code{disttree} vs.
+\code{distforest} vs.
+\code{gamlss} vs.
+\code{randomForest} vs.
+%\code{bamlss} vs.
+\code{gamboostLSS}\\
+right: levelplot of the parameter function for $\kappa=5$}
+\end{figure}
+
 \\
@@
 
-In this section the results of the three methods are compared over an increasing value of $\kappa$.
-For each step they are applied on various data sets and the resulting values of the RMSE and the log-likelihood are averaged within each step.
-In that way it can be investigated how the results change for a changing value of $\kappa$.\\
-Again, a normally distributed response variable and the same smooth parameter function are considered.\\
+\section{Rain}
+In this section precipation data of .. years is analyzed. The response variable is the meassured total precipiation within 24 hours of each day in July at the station ... .\\
+The following methods are applied on this data set:
+\begin{itemize}
+\item distributional tree
+\item distributional forest
+\item gamlss
+\item gamboostLSS
+\item bamlss
+\item EMOS
+\end{itemize}
+
+These methods are compared in a 10x10 cross validation framework based on the resulting RMSE, log-likelihood and CRPS.\\
+These values are illustrated in the following boxplots.
 
 <<echo=FALSE, results=hide>>=
-#### cross validation
-if(file.exists("simres.rda")){
-  load("simres.rda")
+#### cross validation rain
+if(file.exists("rainres.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rainres.rda")
+  load("rainres.rda")
 } else {
-  #source("~/svn/partykit/pkg/disttree/paper/gensim.R")
-  source("gensim.R")
-  simres <- gensim(seedconst = 7423, nrep = 20, ntree = 100L,
-                   cov.sep = TRUE,
-                   fix.mu = FALSE,
-                   fix.sigma = FALSE,
-                   mu.sigma.interaction = FALSE,
-                   gamboost_cvr = FALSE,
-                   eval_disttree = TRUE,
-                   eval_distforest = TRUE,
-                   eval_gamlss = TRUE,
-                   eval_bamlss = TRUE,
-                   eval_gamboostLSS = TRUE,
-                   eval_randomForest = TRUE,
-                   eval_cforest = TRUE)
-  save(simres, file = "simres.rda")
-}
+  source("~/svn/partykit/pkg/disttree/inst/draft/rain.R")
+  #source("gensim.R")
+  rainres <- rain(seedconst = 7423)
+  save(rainres, file = "rainres.rda")
+}
+
+rain_rmse <- matrix(0, ncol = ncol(rainres[[1]]$rmse), length(rainres))
+rain_ll <- matrix(0, ncol = ncol(rainres[[1]]$ll), length(rainres))
+rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres))
+
+for(i in 1:length(rainres)){
+  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse)
+  rain_ll[i,] <- colMeans(rainres[[i]]$ll)
+  rain_crps[i,] <- colMeans(rainres[[i]]$crps)
+}
+
+colnames(rain_rmse) <- colnames(rain_ll) <- colnames(rain_crps) <-
+  colnames(rainres[[1]]$rmse)
 @
@@
 \begin{figure}[t!]
-\begin{subfigure}{0.6\textwidth}
-<<sim_rmse_exp, fig=TRUE, echo=FALSE>>=
-plot_rmse(simres, type = "exp")
+\begin{subfigure}{0.3\textwidth}
+<<rain_rmse_box, fig=TRUE, echo=FALSE>>=
+boxplot(rain_rmse)
 @
@@
 \hspace{-1cm}
-\begin{subfigure}{0.6\textwidth}
-<<sim_rmse_par, fig=TRUE, echo=FALSE>>=
-plot_rmse(simres, type = "par")
+\begin{subfigure}{0.3\textwidth}
+<<rain_ll_box, fig=TRUE, echo=FALSE>>=
+boxplot(rain_ll)
 @
 \end{subfigure}
-\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE)\\
+\hspace{-1cm}
+\begin{subfigure}{0.3\textwidth}
+<<rain_crps_box, fig=TRUE, echo=FALSE>>=
+boxplot(rain_crps)
+@
+\end{subfigure}
+\caption{\label{boxplots_rain}Results 10x10 cross validation on the rain data set (RMSE, log-likelihood and CRPS)\\
 left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\

### 2017-09-26 pkg/disttree/inst/draft/draft.Rnw
added handling of NAs in averaging over the results within the cross-validation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1591 255f4249-2263-4417-b1ea-8498779e27a7

@@
   source("gensim.R")
-  simres <- gensim(seedconst = 7, nrep = 40, ntree = 100,
-                   nsteps = 9, stepsize = 1, kappa.start = 1,
+simres <- gensim(seedconst = 7423, nrep = 10, ntree = 100,
+                 nsteps = 4, stepsize = 3, kappa.start = 1,
                  formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
-                   nobs = 400, testnobs = 200L,
-                   tree_minsplit = 14, tree_mincrit = 0.95,
-                   forest_minsplit = 7, forest_mincrit = 0, 
+                 nobs = 700, testnobs = 200L,
+                 tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95,
+                 forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
                  type.tree = "ctree",
@@
                  eval_gamlss = TRUE,
-                   eval_bamlss = TRUE,
+                 eval_bamlss = FALSE,
                  eval_gamboostLSS = TRUE,
@@
                  mubase = 0)
+
   save(simres, file = "simres.rda")
@@
 for(i in 1:length(rainres)){
-  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse)
-  rain_ll[i,] <- colMeans(rainres[[i]]$ll)
-  rain_crps[i,] <- colMeans(rainres[[i]]$crps)
+  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse, na.rm = TRUE)
+  rain_ll[i,] <- colMeans(rainres[[i]]$ll, na.rm = TRUE)
+  rain_crps[i,] <- colMeans(rainres[[i]]$crps, na.rm = TRUE)
 }

### 2017-09-29 pkg/model4you/inst/JORS/man_model4you.Rnw
comments Torsten
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1596 255f4249-2263-4417-b1ea-8498779e27a7

@@
 treatment group receives a new treatment or drug and the other treatment group
-receives the standard of care or a placebo.
+receives the standard of care or a placebo. Other examples are in A-B testing
+in marketing studies or any other two group comparisons such as the mathematics
+exam discussed below, where students were divided into different exam groups
+and received slightly different exam tasks. In the following we will refer to
+the two groups as \emph{treatment groups} and to the group indicator as
+\emph{treatment indicator}, which always takes values $0$
+(individual in first group) and $1$ (individual in second group).
 
@@
 @
+Note that \Sexpr{'pmtree()'} uses the data given in the call of the base 
+model. It automatically uses variables not used in the model formula 
+(in the example above \Sexpr{'response ~ treatment'}) as potential subgroup 
+defining variables. This can be edited using the \Sexpr{'zformula'} argument.
 
@@
 @ 
+Again here the potential effect-modifying variables are taken by default
+as all variable not given in the model formula and can be defined using 
+the \Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.
 
@@
 \paragraph{Mathematics exam analysis:}
-In 2014 first-year business and economics students at the University of Innsbruck
-were divided into two examination groups. Group 1 wrote the exam in the morning and
-group 2 started after the first group finished. The exams for the two groups were 
-slightly different. Translated into clinical trial language, the exam groups are
-the treatment groups and the two different exams are the treatments.
-The data can be accessed and prepared as follows:
+In 2014 first-year business and economics students at the University of
+Innsbruck were divided into two examination groups. Group 1 wrote the exam in
+the morning and group 2 started after the first group finished. The exams for
+the two groups were slightly different.  The data can be accessed and prepared
+as follows:
 
@@
 percentage of successful online \Sexpr{'tests'}. These online tests were
-conducted biweekly thrughout the semester.  The largest difference between the
+conducted biweekly throughout the semester.  The largest difference between the
 two exam groups is in the students who did very well in the online tests (more
@@
 
-Dependence plots with the group effect (treatment effect) on the y-achsis and the
-student characteristics on the x-axis are a good way of visualising the personalised 
-models and for getting knowledge about the interactions between student characteristics
-and the treatment. Since the percentage of successfull online tests is measured on a
-grid, a bee plot possibly shows the relationships even better than the scatter plot
-(both shown in Figure~\ref{fig:math_dp1}).
+Dependence plots with the group effect (treatment effect) on the y-achsis and
+the student characteristics on the x-axis are a good way of visualising the
+personalised models and for getting knowledge about the interactions between
+student characteristics and the treatment. Since the percentage of successfull
+online tests is measured on a grid, a bee plot possibly shows the relationships
+even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1}).
 <<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
@@
 \section*{Implementation and architecture}
-\begin{figure}
-\centering
-\include{treefamily}
-\caption{The family of model-based trees and forests. Square nodes represent
-functions, round nodes represent function arguments. Solid lines represent
-direct usage, dashed lines indirect usage (iteration between tree and model
-estimation), and dotted lines represent close similarity but no dependency.}
-\label{fig:family}
-\end{figure}
+The R package \emph{model4you} is focussed on ease of use and interpretability.
+Users can take a simple model that they know and understand as basis and simply
+plug it into \Sexpr{'pmtree()'} or \Sexpr{'pmforest()'} depending on whether
+they want subgroup wise or personalised models. The basis for these
+functionalities is provided by the \emph{partykit} package which is a widely
+used R package for trees and forests. The \emph{model4you} package provides
+wrappers for the well implemented and tested functions
+\Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::cforest()'} and extends the
+functionalities where needed to improve usability and interpretability.
 
-The R package \emph{model4you} is based on the \emph{partykit} package. In
-fact, everything that can be done with \emph{model4you} can already be done
-with \emph{partykit} (version $\geq$ 1.2-0). \emph{model4you} is merely a
-convenience interface and focusses on ease of use and interpretability. Various
-other packages depend on \emph{partykit}. Figure~\ref{fig:family} shows all
-packages depending on \emph{partykit} that deal with model-based trees and
-forests. Functions \Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::mob()'}
-provide the basis for all others. Note that \Sexpr{'partykit::ctree()'}
-computes conditional inference trees but by specifying argument
-\Sexpr{'ytrafo'} appropriately, it can also compute model-based trees. This
-functionality is used for packages \emph{disttree} (partitioning of Generalized
-Additive Models for Location, Scale and Shape), \emph{trtf} (partitioning of
-transformation models) and \emph{model4you}.  Many of the packages shown in
-Figure~\ref{fig:family} are available on CRAN.  The remaining packages are
-under development and available on
-\url{https://r-forge.r-project.org/projects/partykit/}.
+The \emph{partykit} package provides the basis for functionalities in other
+packages namely \emph{glmertree}, \emph{psychotree}, \emph{betareg} (all on
+CRAN), \emph{trtf}, \emph{disttree}, \emph{lagsarlmtree} and \emph{palmtree}
+(all available on R-Forge with a plan to move them to CRAN).
 

### 2017-10-05 pkg/disttree/inst/draft/draft.Rnw
adapted to the structure of the results from rain_cross and rain_pred
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1622 255f4249-2263-4417-b1ea-8498779e27a7

--- 2017-09-20 pkg/disttree/inst/draft/draft.Rnw
@@
 # (in randomForest the argument 'keep.inbag' must be set to TRUE)
-rf_getsd <- function(rf, newdata, rfdata){
+rf_getsd <- function(rf, newdata = NULL, rfdata){
+
+  if(is.null(newdata)) newdata <- rfdata
 
@@
    \And Torsten Hothorn\\Universit\"at Z\"urich
-   \And Achim Zeileis\\Universit\"at Innsbruck}
-\Plainauthor{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
+   \And Achim Zeileis\\Universit\"at Innsbruck
+   \And Reto Stauffer\\Universit\"at Innsbruck}
+\Plainauthor{Lisa Schlosser, Torsten Hothorn, Achim Zeileis, Reto Stauffer}
 
@@
   URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
+
+  Reto Stauffer \\
+  Department of Statistics \\
+  Faculty of Economics and Statistics \\
+  Universit\"at Innsbruck \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
+  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/stauffer/reto-stauffer.html}\\
 }
@@
 
+\SweaveOpts{eval = FALSE}
+
 <<echo=FALSE, results=hide>>=
-source("onecov.R")
-if(file.exists("onecov1.rda")){
-  load("onecov1.rda")
+source("oneset.R")
+if(file.exists("oneset1.rda")){
+  load("oneset1.rda")
 } else {
-  onecov1 <- sim_onecov(kappa = 1, nobs = 700,
+  oneset1 <- sim_oneset(kappa = 7, nobs = 400,
                        seedconst = 7, ntree = 100,
-                         formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
-                         tree_minsplit = 40, tree_mincrit = 0.95,
-                         forest_minsplit = 20, forest_mincrit = 0, 
+                       formula = y~x1+x2+x3+x4+x5+x6, 
+                       tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95, 
+                       forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
+                       forest_mtry = 3,
                        type.tree = "ctree",
-                         mubase = 0,
                        censNO = TRUE,
@@
                        mu.sigma.interaction = FALSE,
+                       pred_fix = TRUE,
+                       pred_fix_x1 = -0.5,
+                       pred_fix_x4 = 0,
                        gamboost_cvr = FALSE,
@@
                        eval_gamlss = TRUE,
-                         eval_bamlss = TRUE,
-                         eval_gamboostLSS = TRUE,
-                         eval_randomForest = TRUE,
-                         eval_cforest = TRUE)
-  save(onecov1, file = "onecov1.rda")
-}
-
-if(file.exists("onecov10.rda")){
-  load("onecov10.rda")
+                       eval_gamboostLSS = FALSE,
+                       eval_randomForest = FALSE,
+                       eval_cforest = FALSE,
+                       mubase = 3,
+                       sigmabase = 3)
+  save(oneset1, file = "oneset1.rda")
+}
+
+if(file.exists("oneset10.rda")){
+  load("oneset10.rda")
 } else {
-  onecov10 <- sim_onecov(kappa = 10, nobs = 700,
+  oneset10 <- sim_oneset(kappa = 10, nobs = 400,
                        seedconst = 7, ntree = 100,
-                         formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
-                         tree_minsplit = 40, tree_mincrit = 0.95,
-                         forest_minsplit = 20, forest_mincrit = 0, 
+                       formula = y~x1+x2+x3+x4+x5+x6, 
+                       tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95, 
+                       forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
+                       forest_mtry = 3,
                        type.tree = "ctree",
-                         mubase = 0,
                        censNO = TRUE,
@@
                        eval_gamlss = TRUE,
-                         eval_bamlss = TRUE,
-                         eval_gamboostLSS = TRUE,
-                         eval_randomForest = TRUE,
-                         eval_cforest = TRUE)
-  save(onecov10, file = "onecov10.rda")
+                       eval_gamboostLSS = FALSE,
+                       eval_randomForest = FALSE,
+                       eval_cforest = FALSE,
+                       mubase = 3,
+                       sigmabase = 3)
+  save(oneset10, file = "oneset10.rda")
 }
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_onecov1_mu, fig=TRUE, echo=FALSE>>=
-plot_onecov(learndata = onecov1$learndata, parfun = onecov1$fun, 
-            dt = onecov1$dt, df = onecov1$df, g = onecov1$g, b = onecov1$b, 
-            gb = onecov1$gb, rf = onecov1$rf, cf = onecov1$cf, 
-            compare_mu = TRUE, fix_x2 = 0,
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+<<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
+plot_oneset(oneset1, 
+            compare_mu = TRUE, 
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
+            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
+            )
 @
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_onecov1_sigma, fig=TRUE, echo=FALSE>>=
-plot_onecov(learndata = onecov1$learndata, parfun = onecov1$fun, 
-            dt = onecov1$dt, df = onecov1$df, g = onecov1$g, b = onecov1$b, 
-            gb = onecov1$gb, rf = onecov1$rf, cf = onecov1$cf, 
-            compare_sigma = TRUE, fix_x1 = 0.4,
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+<<sim_oneset1_sigma, fig=TRUE, echo=FALSE>>=
+plot_oneset(oneset1, 
+            compare_sigma_line = TRUE, 
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
+            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
+            )
 @
 \end{subfigure}
-\caption{\label{plot_sim_onecov1} ... }
+\caption{\label{plot_sim_oneset1} ... }
 \end{figure}
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_onecov10_mu, fig=TRUE, echo=FALSE>>=
-plot_onecov(learndata = onecov10$learndata, parfun = onecov10$fun, 
-            dt = onecov10$dt, df = onecov10$df, g = onecov10$g, b = onecov10$b, 
-            gb = onecov10$gb, rf = onecov10$rf, cf = onecov10$cf, 
-            compare_mu = TRUE, fix_x2 = 0,
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+<<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
+plot_oneset(oneset10,
+            compare_mu = TRUE,
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
+            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
+            )
 @
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_onecov10_sigma, fig=TRUE, echo=FALSE>>=
-plot_onecov(learndata = onecov10$learndata, parfun = onecov10$fun, 
-            dt = onecov10$dt, df = onecov10$df, g = onecov10$g, b = onecov10$b, 
-            gb = onecov10$gb, rf = onecov10$rf, cf = onecov10$cf, 
-            compare_sigma = TRUE, fix_x1 = 0.4,
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
+<<sim_oneset10_sigma, fig=TRUE, echo=FALSE>>=
+plot_oneset(oneset10, 
+            compare_sigma_line = TRUE, 
+            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
+            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
+            )
 @
 \end{subfigure}
-\caption{\label{plot_sim_onecov10} ... }
+\caption{\label{plot_sim_oneset10} ... }
 \end{figure}
@@
   source("gensim.R")
-  simres <- gensim(seedconst = 7, nrep = 40, ntree = 100,
-                   nsteps = 9, stepsize = 1, kappa.start = 1,
-                   formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
+  simres <- gensim(seedconst = 7, nrep = 20, ntree = 100,
+                   nsteps = 5, stepsize = 2,
+                   formula = y~x1+x2+x3+x4+x5+x6,
                    nobs = 400, testnobs = 200L,
-                   tree_minsplit = 14, tree_mincrit = 0.95,
-                   forest_minsplit = 7, forest_mincrit = 0, 
-                   type.tree = "ctree",
-                   censNO = TRUE,
+                   tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95,
+                   forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
+                   forest_mtry = 3,
                    fix.mu = FALSE,
@@
                    mu.sigma.interaction = FALSE,
+                   mubase = 1,
+                   sigmabase = 3,
+                   censNO = TRUE,
                    gamboost_cvr = FALSE,
@@
                    eval_gamlss = TRUE,
-                   eval_bamlss = TRUE,
-                   eval_gamboostLSS = TRUE,
-                   eval_randomForest = TRUE,
-                   eval_cforest = TRUE,
-                   mubase = 0)
+                   eval_bamlss = FALSE,
+                   eval_gamboostLSS = FALSE,
+                   eval_randomForest = FALSE,
+                   eval_cforest = FALSE)
+
   save(simres, file = "simres.rda")
@@
 cols <- colR(ncuts+1)
-griddata <- expand.grid(list(x1 = seq(-0.4, 1, 0.01), x2 = seq(-10, 10, 0.2)))
+griddata <- expand.grid(list(x1 = seq(-1, 1, 0.01), x4 = seq(0, 1, 0.2)))
 griddata$mu <- levelfun(griddata)
-levelplot(mu ~ x1 * x2, data = griddata, xlab = "x1", ylab = "x2", 
+levelplot(mu ~ x1 * x4, data = griddata, xlab = "x1", ylab = "x4", 
           region = TRUE, cuts = ncuts, col.regions = cols)
@@
 
-\\
 %Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups.
@@
 
+\SweaveOpts{eval = TRUE}
+
+\newpage
 \section{Rain}
-In this section precipation data of .. years is analyzed. The response variable is the meassured total precipiation within 24 hours of each day in July at the station ... .\\
+In this section precipation data of 30 years is analyzed. The response variable is the meassured total precipiation within 24 hours of each day in July at the station ... .\\
 The following methods are applied on this data set:
@@
 \item gamboostLSS
-\item bamlss
 \item EMOS
@@
 #### cross validation rain
-if(file.exists("rainres.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rainres.rda")
-  load("rainres.rda")
+if(file.exists("rain_Axams.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams.rda")
+  load("rain_Axams.rda")
 } else {
-  source("~/svn/partykit/pkg/disttree/inst/draft/rain.R")
-  #source("gensim.R")
-  rainres <- rain(seedconst = 7423)
-  save(rainres, file = "rainres.rda")
-}
-
-rain_rmse <- matrix(0, ncol = ncol(rainres[[1]]$rmse), length(rainres))
-rain_ll <- matrix(0, ncol = ncol(rainres[[1]]$ll), length(rainres))
-rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres))
-
-for(i in 1:length(rainres)){
-  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse)
-  rain_ll[i,] <- colMeans(rainres[[i]]$ll)
-  rain_crps[i,] <- colMeans(rainres[[i]]$crps)
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross.R")
+  source("rain_cross.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+  setwd("~/svn/partykit/pkg/disttree/inst/draft/")
+  rainres <- rain_cross(stationname = "Axams", 
+                        seedconst = 7, ntree = 100,
+                        forest_mtry = 27,
+                        gamboost_cvr = TRUE,
+                        frac = FALSE)
+}
+
+rain_rmse <- matrix(0, ncol = ncol(rainres[[1]]$rmse), length(rainres)-1)
+rain_ll <- matrix(0, ncol = ncol(rainres[[1]]$ll), length(rainres)-1)
+rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres)-1)
+
+for(i in 1:(length(rainres)-1)){
+  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse, na.rm = TRUE)
+  rain_ll[i,] <- colMeans(rainres[[i]]$ll, na.rm = TRUE)
+  rain_crps[i,] <- colMeans(rainres[[i]]$crps, na.rm = TRUE)
 }
@@
   colnames(rainres[[1]]$rmse)
+
 @
@@
 \end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.3\textwidth}
+<<rain_crps_box, fig=TRUE, echo=FALSE>>=
+boxplot(1 - rain_crps / rain_crps[,7]) 
+abline(h = 0, col = 2, lwd = 2)
+@
+\end{subfigure}
 \caption{\label{boxplots_rain}Results 10x10 cross validation on the rain data set (RMSE, log-likelihood and CRPS)\\
@@
 right: continuous lines represent the location parameter, dotted lines the scale parameter}
-\end{figure}
-
-\begin{figure}[t!]
-\begin{subfigure}{0.6\textwidth}
-<<sim_ll, fig=TRUE, echo=FALSE>>=
-plot_ll(simres)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.6\textwidth}
-<<sim_levelplot, fig=TRUE, echo=FALSE>>=
-levelfun <- function(x) simres$fun(x,5)
-
-ncuts <- 100
-colR <- colorRampPalette(c("blue","green"))                                       
-cols <- colR(ncuts+1)
-griddata <- expand.grid(list(x1 = seq(-0.4, 1, 0.01), x2 = seq(-10, 10, 0.2)))
-griddata$mu <- levelfun(griddata)
-levelplot(mu ~ x1 * x2, data = griddata, xlab = "x1", ylab = "x2", 
-          region = TRUE, cuts = ncuts, col.regions = cols)
-
-@
-\end{subfigure}
-\caption{\label{plot_sim_ll}left: results of the simulation study: log-likelihood\\ %\code{disttree} vs.
-\code{distforest} vs.
-\code{gamlss} vs.
-\code{randomForest} vs.
-%\code{bamlss} vs.
-\code{gamboostLSS}\\
-right: levelplot of the parameter function for $\kappa=5$}
 \end{figure}

### 2017-10-05 pkg/disttree/inst/draft/draft.Rnw
started description of rain data set
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1624 255f4249-2263-4417-b1ea-8498779e27a7

@@
 In this section precipation data of 30 years is analyzed. The response variable is the meassured total precipiation within 24 hours of each day in July at the station ... .\\
+
+\begin{tabular}{| l | c |}
+\hline
+basic covariates & variations
+\hline
+tppow: total precipitation power transformed & tppow_mean, tppow_spread, tppow_min, tppow_max \\
+& tppow_0612, tppow_1218, tppow_1824, tppow_2430 \\
+\hline
+capepow:    & capepow_mean, capepow_spread, capepow_min, capepow_max \\
+& capepow_0612, capepow_1218, capepow_1824, capepow_2430 \\
+\hline
+tcloc: total cloud coverage & \\
+\hline
+\hline
+\end{table}
+
 The following methods are applied on this data set:
-\begin{itemize}
-\item distributional tree
-\item distributional forest
-\item gamlss
-\item gamboostLSS
-\item EMOS
-\end{itemize}
+
+\begin{tabular}{| l | c  c  c  c |}
+\hline
+model & mu link & sigma link & mu covariates & sigma covariates \\
+\hline
+disttree & identity & log & all covariates & all covariates \\ 
+\hline
+distforest & identity & log & all covariates & all covariates \\
+\hline
+gamlss & identity & log & manually selected covariates: & manually selected covariates \\
+\hline
+gamboostLSS & identity & log & all covariates & all covariates \\
+\hline
+EMOS & identity & log & tppow_mean & tppow_spread \\
+\hline
+\end{table}
 

### 2017-10-06 pkg/disttree/inst/draft/draft.Rnw
added prediction on one station with 27 years as learning data and 3 years as testing data
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1629 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\begin{tabular}{| l | c |}
-\hline
-basic covariates & variations
-\hline
-tppow: total precipitation power transformed & tppow_mean, tppow_spread, tppow_min, tppow_max \\
-& tppow_0612, tppow_1218, tppow_1824, tppow_2430 \\
-\hline
-capepow:    & capepow_mean, capepow_spread, capepow_min, capepow_max \\
-& capepow_0612, capepow_1218, capepow_1824, capepow_2430 \\
-\hline
-tcloc: total cloud coverage & \\
-\hline
-\hline
-\end{table}
+\begin{tabular}{| l | c | l |}
+\hline
+basic covariates & nr & variations\\
+\hline
+\textbf{tp{\_}pow}: total precipitation & 12 & mean, std dev, min, max (all for 6--30 UTC)\\
+\hspace*{\fill} (power transformed by 1.6) & & mean: 6--12, 12--18, 12--24, 24--30 UTC\\
+& & std dev: 6--12, 12--18, 12--24, 24--30 UTC\\
+\hline
+\textbf{cape{\_}pow}: convective available & 12 & mean, std dev, min, max (all for 6--30 UTC)\\
+\hspace*{\fill} potential energy & & mean: 6--12, 12--18, 12--24, 24--30 UTC\\
+& & std dev: 6--12, 12--18, 12--24, 24--30 UTC\\
+\hline
+dswrf: downwards short & 4-6 & mean, std dev, min, max (all for 0--30 UTC)\\
+\hspace*{\fill}ware radiation flux (sunshine) & & mean of the spread, max of the spread, (min of the spread)\\
+\hline
+msl: mean sea level pressure & 6 &
+mean of the mean, max of the mean, min of the mean,
+mean of the spread, max of the spread, min of the spread \\
+\hline
+pwat: preciptable water & 6 &
+mean of the mean, max of the mean, min of the mean, 
+mean of the spread, max of the spread, min of the spread \\
+\hline
+tmax: 2m maximum temperature & 6 &
+mean of the mean, max of the mean, min of the mean, 
+mean of the spread, max of the spread, min of the spread \\
+\hline
+tcolc: Total column-integrated condensate & 6 &
+mean of the mean, max of the mean, min of the mean, 
+mean of the spread, max of the spread, min of the spread \\
+\hline
+t500: temperature on 500 hPa & 6 &
+mean of the mean, max of the mean, min of the mean, 
+mean of the spread, max of the spread, min of the spread \\
+\hline
+t700: temperature on 700 hPa & 6 &
+mean of the mean, max of the mean, min of the mean, 
+mean of the spread, max of the spread, min of the spread \\
+\hline
+t850: temperature on 850 hPa & 6 &
+mean of the mean, max of the mean, min of the mean, 
+mean of the spread, max of the spread, min of the spread \\
+\hline
+tdiff500850: temperature difference & 3 &
+mean, min, max \\
+\hline
+tdiff500700: temperature difference & 3 &
+mean, min, max \\
+\hline
+tdiff700850: temperature difference & 3 &
+mean, min, max \\
+\hline
+msl{\_}diff: temperature difference & 1 & 
+- \\
+\hline
+\end{tabular}
 
@@
 \hline
-EMOS & identity & log & tppow_mean & tppow_spread \\
-\hline
-\end{table}
-
-These methods are compared in a 10x10 cross validation framework based on the resulting RMSE, log-likelihood and CRPS.\\
+EMOS & identity & log & tppowmean & tppowspread \\
+\hline
+\end{tabular}
+
+\subsection{Station Axams}
+<<echo=FALSE, results=hide>>=
+#### Axams prediction 27 - 3
+if(file.exists("rain_Axams_pred.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
+  load("rain_Axams_pred.rda")
+} else {
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
+  source("rain_axams_pred.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+  res <- rain_axams_pred(seedconst = 7, ntree = 100,
+                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
+                         gamboost_cvr = FALSE,
+                         frac = FALSE)
+
+  save(res, file = "rain_Axams_pred.rda")
+}
+
+# prepare data for plot of estimated density functions
+pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
+pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+cbind(df_exp, res$testdata[pdays,"robs"])
+
+# plot predicted distributions together with observations
+#set.seed(res$call$seedconst)
+set.seed(7)
+x <- sort(runif(500,-2,12))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
+
+@
+
+\begin{figure}[t!]
+<<rain_Axams_pred, fig=TRUE, echo=FALSE>>=
+plot(x = x, y = y1, type = "l", col = "red", 
+         main = paste0("July ", pday, dayending), ylab = "density", ylim = c(0,max(y1, y2, y3) + 0.01))
+    lines(x = x, y = y2, type = "l", col = "blue")
+    lines(x = x, y = y3, type = "l", col = "darkgreen")
+    legend('topright', c("2010", "2011", "2012"), col = c("red", "blue", "darkgreen"), lty = 1, cex = 1)
+
+    p1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+    p2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+    p3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+
+    points(x = p1[1], y = p1[2], col = "red")
+    points(x = p2[1], y = p2[2], col = "blue")
+    points(x = p3[1], y = p3[2], col = "darkgreen")
+
+    lines(x = c(p1[1], p1[1]), y = c(p1[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(p2[1], p2[1]), y = c(p2[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(p3[1], p3[1]), y = c(p3[2], 0), col = "darkgrey", type = "l", lty = 2)
+@
+
+\caption{\label{}}
+\end{figure}
+
+The methods are compared in a 10x10 cross validation framework based on the resulting RMSE, log-likelihood and CRPS.\\
 These values are illustrated in the following boxplots.
@@
   gen.cens("NO", type = "left")
-  setwd("~/svn/partykit/pkg/disttree/inst/draft/")
   rainres <- rain_cross(stationname = "Axams", 
@@
                         frac = FALSE)
+
+  save(rainres, file = "rain_Axams.rda")
 }

### 2017-10-08 pkg/disttree/inst/draft/draft.Rnw
added plots to the rain section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1634 255f4249-2263-4417-b1ea-8498779e27a7

--- 2017-10-05 pkg/disttree/inst/draft/draft.Rnw
@@
 \usepackage{Sweave}
+\usepackage{enumitem}
 \usepackage{tikz}
@@
 
+\SweaveOpts{eval = TRUE}
 \section{Introduction}
@@
 These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
-\\
-\\
-Now we take a look at a classical text book example, the mcycle data set from the package \pkg{MASS}.
-It consists of a series of measurements of head acceleration in a simulated motorcycle accident.
-The response variable \code{accel} represents the recorded head acceleration meassured in g and the explanatory variable \code{times} is meassured in milliseconds after the impact.\\
-We compare the performance of the following four functions on this data set: \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest}\\
-This situation is particularly well suited for a GAMLSS because there is only one regressor and physically one would expect a smooth function.
-However, this example shows, that the distributional forest can also capture the changes quite well albeit not as smoothly as the GAMLSS.
-Moreover, it illustrates that the distributional tree detects the changes as well but models them with rough steps.\\
-\\
-To make it possible to compare the models also on the estimated scale parameter, here the standard deviation of the random forest is estimated in each terminal node of each tree and then averaged, in the same way as the mean parameter is estimated.
-This makes it possible to specify a distribution, however, the splits of a random forest are only made in the location parameter while a distributional tree/forest considers splits in all distribution parameters.\\
-\\
-In Figure \ref{plot_mcycle} the results of a distributional forest are plotted on the top left and then compared to the results of a distributional tree, a GAMLSS and a random forest, regarding the location parameter on the top right and the scale parameter below.\\
-While the smooth function of \code{gamlss} seems to model especially the middle part (for \code{times} between 17 and 40) better than the wiggly functions resulting from the trees it can also be observed that at \code{times} $=14$ there is a rapid drop which is captuerd more precisely by the tree-based models while \code{gamlss} starts descending earlier.
-Also for the highest values of \code{times} the smooth function increases again while the forests stay almost at the same level and the tree stays at a fixed value which seems more appropriate in this situation.\\
-Comparing the two forest models, the distributional forest seems to be a bit more stable while the random forest includes more rapid changes of a wider range.\\
-Looking at the plot of the scale parameter, in general the distributional forest estimates higher parameters than the random forest.\\
-
-<<echo=FALSE, results=hide>>=
-
-## FIX ME: distforest uses subsets, randomForest uses bootstraps
-
-## randomForest: https://stackoverflow.com/questions/28417826/nodesize-parameter-ignored-in-randomforest-package
-## nodesize = the minimum number of observations that must exist in a node in order for a split to be attempted.
-## -> nodesize = minsplit
-
-if(file.exists("mcycle_plotdata.rda")){
-  load("mcycle_plotdata.rda")
-} else {
-  set.seed(7)
-  dt <- disttree(accel ~ times, data = mcycle, family = NO(), type.tree = "mob", control = mob_control(minsize = 7L))
-  df <- distforest(accel ~ times, data = mcycle, family = NO(), ntree = 300L, 
-                   control = ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 5L, minsplit = 10L, intersplit = TRUE))
-  rf <- randomForest(accel ~ times, data = mcycle, ntree = 300L, nodesize = 10L, 
-                     keep.inbag = TRUE, replace = FALSE)
-  #ga <- gam(list(accel ~ s(times), ~ s(times)), data = mcycle, family=gaulss())
-  ga <- gam(list(accel ~ s(times,k=20,bs="ad"), ~ s(times)), data = mcycle, family=gaulss())
-  #b <- bamlss(list(accel ~ s(times), accel ~ s(times)), data = mcycle)
-  #save(df, file = "mcycle_plotdata.rda")
-
-  ## get 'fitted.sigma.rf'
-  rf_sd <- rf_getsd(rf, newdata = mcycle, rfdata = mcycle)
-
-  plotdata <- cbind(mcycle,
-                    df$fitted.par,
-                    dt$fitted.par,
-                    #g$mu.fv, g$sigma.fv,
-                    ga$fitted.values[,1], 1/ga$fitted.values[,2],
-                    rf$predicted, rf_sd)
-
-  colnames(plotdata) <- c("times", "accel",
-                          "fitted.mu.df","fitted.sigma.df",
-                          "fitted.mu.dt","fitted.sigma.dt",
-                          "fitted.mu.g","fitted.sigma.g",
-                          "fitted.mu.rf","fitted.sigma.rf")
-
-  sp <- plotdata[order(plotdata["times"]),]
-
-  save(sp, file = "mcycle_plotdata.rda")
-}
-@
-
-\begin{figure}[t!]
-\begin{subfigure}{0.65\textwidth}
-<<plot_df, fig=TRUE, echo=FALSE>>=
-# distforest
-plot(mcycle)
-lines(x = sp$times, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
-polygon(c(sp$times, rev(sp$times)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
-        col = pallight["forest"], border = "transparent")
-@
-%\caption{\label{plot_mcycle_df}Distributional Forest, \code{distforest}}
-\end{subfigure}
-\hspace{-2cm}
-\begin{subfigure}{0.65\textwidth}
-<<plot_mu_vs, fig=TRUE, echo=FALSE>>=
-# disttree vs. distforest vs. gamlss vs. randomForest
-plot(x = sp$times, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 1.5,
-     ylim = c(-134,75), ylab = "accel", xlab = "times")
-lines(x = sp$times, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 1.5)
-lines(x = sp$times, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 1.5)
-lines(x = sp$times, y = sp$fitted.mu.rf, type = "l", col = pal["randomForest"], lwd = 1.5)
-legend('topleft', c("disttree", "distforest", "gamlss", "randomForest"), 
-       col = c(pal["tree"], pal["forest"], pal["gamlss"], pal["randomForest"]), lty = 1, cex = 1)
-@
-%\caption{\label{plot_mcycle_mu_vs} Location parameter: \code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
-\end{subfigure}
-\vspace{-1cm}
-
-\hspace{4cm}
-\begin{subfigure}{0.65\textwidth}
-<<plot_sigma_vs_lines, fig=TRUE, echo=FALSE>>=
-# variance: disttree vs. distforest vs. gamlss (lines)
-plot(x = sp$times, y = sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2,
-     ylim = c(0,50), ylab = "variance of accel", xlab = "times")
-#lines(x = sp$times, y = -sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
-lines(x = sp$times, y = sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-#lines(x = sp$times, y = -sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-lines(x = sp$times, y = sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-#lines(x = sp$times, y = -sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-lines(x = sp$times, y = sp$fitted.sigma.rf, type = "l", col = pal["randomForest"], lwd = 2)
-#lines(x = sp$times, y = -sp$fitted.sigma.rf, type = "l", col = pal["randomForest"], lwd = 2)
-legend('topleft', c("disttree", "distforest", "gamlss", "randomForest"), 
-       col = c(pal["tree"], pal["forest"], pal["gamlss"], pal["randomForest"]), lty = 1, cex = 1)
-@
-\end{subfigure}
-\caption{\label{plot_mcycle}top left: location and scale parameter of \code{distforest}\\
-top right: location parameter of all four models\\
-bottom: scale parameter of all four models}
-\end{figure}
 
@@
 While in a distributional tree the same distribution is specified for all observations being in the same node a distributional forest predicts and fits a different distribution for each different observation which is due to the individual weights calculated separately for each observation.
-
-\newpage
-\section{Application}
-\subsection{Continuation of the mcycle example: cross validation}
-In the introduction the four functions \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest} have been applied on the mcycle data set and their results haven been visualized and discussed.
-To get a more profound impression of their performances we now want to compare them in a cross validation framework on this text book example data set.
-The 133 observations are randomly split into 10 subsets roughly of the same size.
-After learning the model on 9 of these subgroups the RMSE of the estimated location parameter and the log-likelihood function are evaluated on the 10th part.
-Each subset is chosen once as a test data set and the mean of the resulting values is stored.
-This procedure is repeated 9 times.
-This comparison is illustrated in Figure \ref{plot_cross_mean}, based on the RMSE on the left and based on the log-likelihood on the right.
-(Extrapolation of the GAMLSS would lead to very small values of the log-likelihood.
-To avoid this, the first and the l!
- ast observation are always included in the learn data set.) \\
-As we have already seen in the introduction the distributional tree obviously struggles with the smooth effects while the two forest models can capture the changes quite well.
-As far as the RMSE of the estimated location parameter is concerned the GAMLSS plays to its strength as a smooth parametric model and leads to the best results followed by the distributional forest and the random forest which might both be a bit to wiggly.\\
-However, regarding the log-likelihood the forests perform better than the GAMLSS.
-This is mainly due to the fact that the GAMLSS models the scale parameter close to zero for observations with \code{times} $<10$.
-This leads to very low values of the log-likelihood function if one or more of these observations are included in the test data set.
-As the estimated scale parameter of the tree based models are higher for these specific observations this problem does not occure here.\\
-Overall it can be seen that the distributional forest which fits the whole distribution and considers splits in both parameters performs better than the random forest which only models the mean and therefore also only inlcudes splits in the location parameter.
-
-<<echo=FALSE, results=hide>>=
-#### cross validation
-if((file.exists("mcycle_cross_mean.rda")) & (file.exists("mcycle_cross_all.rda"))){
-  load("mcycle_cross_mean.rda")
-  load("mcycle_cross_all.rda")
-  nrep_cross <- NROW(cross_all)
-} else {
-
-  set.seed(704)
-  #set.seed(7054)
-  nrep_cross <- 10
-
-  rmse <- matrix(nrow = nrep_cross * 10, ncol = 4)
-  ll <- matrix(nrow = nrep_cross * 10, ncol = 4)
-  mean_rmse <- matrix(nrow = nrep_cross, ncol = 4)
-  mean_ll <- matrix(nrow = nrep_cross, ncol = 4)
-  errors <- numeric(length = nrep_cross * 10)
-  colnames(rmse) <- colnames(mean_rmse) <- c("rmse_disttree", "rmse_distforest", "rmse_gamlss", "rmse_randomForest")
-  colnames(ll) <- colnames(mean_ll) <- c("ll_disttree", "ll_distforest", "ll_gamlss", "ll_randomForest")
-
-  tdata <- list()
-  ldata <- list()
-
-  for(k in 1:nrep_cross) {
-
-    # separate data set (without the first and the last observation) into 10 parts (9 of size 13 and 1 of size 14):
-    # (first and last observation are not in testid -> always stay in learndata)
-    id <- c(2:(NROW(mcycle)-1))
-    testid <- list()
-    for(i in 1:9){
-      testid[[i]] <- sample(id, 13, replace = FALSE)
-      id <- id[!(id %in% testid[[i]])]
-    }
-    testid[[10]] <- id
-
-    rmse_dt_mu <- rmse_df_mu <- rmse_g_mu <- rmse_rf_mu <- 
-      ll_dt <- ll_df <- ll_g <- ll_rf <- numeric(length = 10)
-
-    for(i in 1:10){
-      tdata[[(k-1)*10+i]] <- testdata <- mcycle[testid[[i]], ]
-      ldata[[(k-1)*10+i]] <- learndata <- mcycle[-testid[[i]], ]
-
-      #testdata <- mcycle[testid[[i]], ]
-      #learndata <- mcycle[-testid[[i]], ]
-
-      dt <- disttree(accel ~ times, data = learndata, family = NO(), type.tree = "mob", control = mob_control(minsize = 7L))
-      df <- distforest(accel ~ times, data = learndata, family = NO(), ntree = 300L, 
-                       control = ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 5L, minsplit = 10L, intersplit = TRUE))
-      rf <- randomForest(accel ~ times, data = learndata, ntree = 300L, nodesize = 10L, keep.inbag = TRUE, replace = FALSE)
-      ga <- gam(list(accel ~ s(times,k=20,bs="ad"), ~ s(times)), data = learndata, family=gaulss())
-      #ga <- gam(list(accel ~ s(times), ~ s(times)), data = learndata, family=gaulss())
-      errors[(k-1)*10+i] <- (ga$outer.info$conv != "full convergence")
-
-      # get predicted values
-      dt.pred.par <- predict(dt, newdata = testdata, type = "parameter")
-      df.pred.par <- predict(df, newdata = testdata, type = "parameter")
-      rf.pred.resp <- predict(rf, newdata = testdata)
-
-      rf.sd <- rf_getsd(rf, newdata = testdata, rfdata = learndata)
-
-      g.pred.par <- predict(ga, newdata = testdata, type = "response")
-      g.pred.mu <- g.pred.par[,1]
-      g.pred.sigma <- 1/g.pred.par[,2]
-
-      # RMSE location parameter
-      rmse_dt_mu[i] <- sqrt(mean((dt.pred.par$mu - testdata$accel)^2))
-      rmse_df_mu[i] <- sqrt(mean((df.pred.par$mu - testdata$accel)^2))
-      rmse_g_mu[i] <- sqrt(mean((g.pred.mu - testdata$accel)^2))
-      rmse_rf_mu[i] <- sqrt(mean((rf.pred.resp - testdata$accel)^2))
-
-      # loglikelihood
-      dtll <- dfll <- gll <- rfll <- 0
-      for(j in 1:(nrow(testdata))){
-
-        dtll_j <- dnorm(testdata[j,"accel"], mean = dt.pred.par$mu[j], sd = dt.pred.par$sigma[j], log=TRUE)
-        dfll_j <- dnorm(testdata[j,"accel"], mean = df.pred.par$mu[j], sd = df.pred.par$sigma[j], log=TRUE)
-        gll_j <- dnorm(testdata[j,"accel"], mean = g.pred.mu[j], sd = g.pred.sigma[j], log=TRUE)
-        rfll_j <- dnorm(testdata[j,"accel"], mean = rf.pred.resp[j], sd = rf.sd[j], log=TRUE)
-
-        dtll <- if(is.na(dtll_j)) {
-          print(dt.pred.par[j,], testdata[j,"accel"]) 
-          dtll + (-5)
-        } else {dtll + dtll_j}
-        dfll <- if(is.na(dfll_j)) {
-          print(df.pred.par[j,], testdata[j,"accel"]) 
-          dfll + (-5)
-        } else {dfll + dfll_j}
-        gll <- if(is.na(gll_j)) {
-          print(g.pred.mu[j], g.pred.sigma[j], testdata[j,"accel"]) 
-          gll + (-5)
-        } else {gll + gll_j}
-        rfll <- if(is.na(rfll_j)) {
-          print(c(rf.pred.resp[j], rf.sd[j]), testdata[j,"accel"]) 
-          rfll + (-5)
-        } else {rfll + rfll_j}
-      }
-
-      ll_dt[i] <- dtll
-      ll_df[i] <- dfll
-      ll_g[i] <- gll
-      ll_rf[i] <- rfll
-
-      rmse[i+(k-1)*10,] <- c(rmse_dt_mu[i], rmse_df_mu[i], rmse_g_mu[i], rmse_rf_mu[i])
-      ll[i+(k-1)*10,] <- c(ll_dt[i], ll_df[i], ll_g[i], ll_rf[i])
-    }
-
-    # store data
-    mean_rmse[k,] <- c(mean(rmse_dt_mu), mean(rmse_df_mu), mean(rmse_g_mu), mean(rmse_rf_mu))
-    mean_ll[k,] <- c(mean(ll_dt), mean(ll_df), mean(ll_g), mean(ll_rf))
-
-  }
-
-  cross_all <- cbind(rmse, ll)
-  cross_mean <- cbind(mean_rmse, mean_ll)
-  save(cross_all, file = "mcycle_cross_all.rda")
-  save(cross_mean, file = "mcycle_cross_mean.rda")
-
-}
-@
-
-\begin{figure}[t!]
-\begin{subfigure}{0.6\textwidth}
-<<mcycle_cross_rmse_mean, fig=TRUE, echo=FALSE>>=
-# parallel boxplots RMSE
-box_rmse <- cross_mean[, c("rmse_disttree", "rmse_randomForest", "rmse_distforest", "rmse_gamlss")]
-colnames(box_rmse) <- c("disttree", "randomForest", "distforest", "gamlss")
-boxplot.matrix(box_rmse, ylab = "RMSE")
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.6\textwidth}
-<<mcycle_cross_ll_mean, fig=TRUE, echo=FALSE>>=
-# parallel boxplots log-likelihood
-box_ll <- cross_mean[, c("ll_disttree", "ll_randomForest", "ll_distforest", "ll_gamlss")]
-#box_ll <- box_ll[box_ll[,1]>-400,]
-#box_ll <- box_ll[box_ll[,2]>-400,]
-box_ll <- box_ll[box_ll[,"ll_gamlss"]>-70,]
-#box_ll <- box_ll[box_ll[,4]>-400,]
-colnames(box_ll) <- c("disttree", "randomForest", "distforest", "gamlss")
-boxplot.matrix(box_ll, ylab = "log-likelihood")
-@
-\end{subfigure}
-%\center
-\caption{\label{plot_cross_mean}10-times-10 cross validation on the mcycle data, averages (>-70):\\ \code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
-\end{figure}
-
-%\begin{figure}[t!]
-%\begin{subfigure}{0.6\textwidth}
-%<<mcycle_cross_rmse_all, fig=TRUE, echo=FALSE>>=
-%# parallel boxplots RMSE
-%box_rmse <- cross_all[, c("rmse_disttree", "rmse_randomForest", "rmse_distforest", %"rmse_gamlss")]
-%colnames(box_rmse) <- c("disttree", "randomForest", "distforest", "gamlss")
-%boxplot.matrix(box_rmse, ylab = "RMSE")
-%@
-%%\caption{\label{plot_boot_rmse}Bootstraps: RMSE of the location parameter, \code{dist%tree%} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
-%\end{subfigure}
-%\hspace{-1cm}
-%\begin{subfigure}{0.6\textwidth}
-%<<mcycle_cross_ll_all, fig=TRUE, echo=FALSE>>=
-%# parallel boxplots log-likelihood
-%box_ll <- cross_all[, c("ll_disttree", "ll_randomForest", "ll_distforest", "ll_gamlss"%)]
-%#box_ll <- box_ll[box_ll[,1]>-400,]
-%#box_ll <- box_ll[box_ll[,2]>-400,]
-%box_ll <- box_ll[box_ll[,"ll_gamlss"]>-100,]
-%#box_ll <- box_ll[box_ll[,4]>-400,]
-%colnames(box_ll) <- c("disttree", "randomForest", "distforest", "gamlss")
-%boxplot.matrix(box_ll, ylab = "log-likelihood")
-%@
-%\end{subfigure}
-%%\center
-%\caption{\label{plot_cross_ll_all}mcycle 10-times-10 Cross validation (>-100): %\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
-%\end{figure}
 
@@
 \SweaveOpts{eval = TRUE}
-
+\newpage
 \newpage
 \section{Rain}
-In this section precipation data of 30 years is analyzed. The response variable is the meassured total precipiation within 24 hours of each day in July at the station ... .\\
+In this application we analyze precipitation data meassured over 28 years at 95 (?) stations all over Tyrol and Vorarlberg.
+The considered data is provided by .....
+and consists of the meassured total precipiation within 24 hours of each day in July from 1985 until 2012, both included.
+Additionally, numerical forecasts from the ECMWF are used as covariates.
+These forecasts are ensemble predictions consisting of ...
+individual forecasts based on perturbed initiall conditions.\\
+In particular, 14 basic forecast variables are considered with up to 12 variations of these variables being included as covariates.
+Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standarad deviation, the minimum and the maximum over all these individual predictions.
+Another variation refers to the time periods which can be over 24 hours or over time windows of 6 hours.
+Another possible option is to consider the mean over 24 hours for each individual prediction and then take the mean over the whole ensemble.
+In the following table listing all covariates this last option is referred to as the ensemble mean of mean.\\
+The response variable total precicipitation is power transformed by a factor of 1.6.
+This is done due to experience and knowledge about possible distributions of precipitation meassurements.
+The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
+
+\newpage
+The ensemble forecast for the total precipitation is represented by 12 covariates: 
+\begin{itemize}[noitemsep]
+\item ensemble mean of the sum of the total precipitation from 6--30 UTC
+\item ensemble standard deviation of the sum of the total precipitation from 6--30 UTC
+\item ensemble minimum of the sum of the total precipitation from 6--30 UTC
+\item ensemble maximum of the sum of the total precipitation from 6--30 UTC
+\item ensemble mean of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC and from 24--30 UTC
+\item ensemble standard deviation of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC and from 24--30 UTC
+\end{itemize}
+
+The preciptable water is represented by 6 covariates:
+\begin{itemize}[noitemsep]
+\item the ensemble mean of the mean of preciptable water over the time period 6--30 UTC 
+\item the ensemble mean of the minimum of preciptable water over the time period 6--30 UTC 
+\item the ensemble mean of the maximum of preciptable water over the time period 6--30 UTC 
+\item the ensemble standard deviation of the mean of preciptable water over the time period 6--30 UTC 
+\item the ensemble standard deviation of the minimum of preciptable water over the time period 6--30 UTC 
+\item the ensemble standard deviation of the maximum of preciptable water over the time period 6--30 UTC 
+\end{itemize}
+
+\newpage
+\begin{tabular}{| l | c | l |}
+\hline
+basic covariates & nr & variations\\
+\hline
+\textbf{tp{\_}pow}: total precipitation    & 12 & ensemble mean, ensemble std dev, \\
+\hspace*{\fill} power transformed (by 1.6) &    & ensemble min, ensemble max\\
+\textbf{cape{\_}pow}: convective           &    & (all for 6--30 UTC)\\
+\hspace*{0.4cm} available potential energy &    & ensemble mean: 6--12, 12--18,\\
+\hspace*{\fill}  power transformed (by 1.6)&    & \hspace{2.5cm} 18--24, 24--30 UTC\\
+\hspace*{\fill}                            &    & ensemble std dev: 6--12, 12--18,\\
+\hspace*{\fill}                            &    & \hspace{2.85cm}  18--24, 24--30 UTC\\
+\hline
+\textbf{dswrf}: downwards short ware & 6 & ensemble mean of mean, \\
+\hspace*{\fill} radiation flux (sunshine) &     & ensemble mean of min,\\
+\textbf{msl}: mean sea level pressure  &     &  ensemble mean of max,\\
+\textbf{pwat}: preciptable water &   & ensemble std dev of mean,\\
+\textbf{tmax}: 2m maximum        &   & ensemble std dev of min,\\
+\hspace*{1.15cm} temperature  &   & ensemble std dev of max,\\
+\textbf{tcolc}: total column-integrated &   & (all for 6--30 UTC)\\
+\hspace*{1.05cm} condensate     &   & \\
+\textbf{t500}: temperature     &  & \\
+\hspace*{1cm} on 500 hPA     &   & \\
+\textbf{t700}: temperature     &  & \\
+\hspace*{1cm} on 700 hPA     &   & \\
+\textbf{t850}: temperature     &   & \\
+\hspace{1cm} on 850 hPA     &   & \\
+\hline
+\textbf{tdiff500850}: temperature         & 3 & ensemble mean,\\
+\hspace*{\fill} difference 500 to 850 hPA &   & ensemble min,\\
+\textbf{tdiff500700}: temperature         &  & ensemble max\\
+\hspace*{\fill} difference 500 to 700 hPA &   & (all for 6--30 UTC)\\
+\textbf{tdiff700850}: temperature          &  & \\
+\hspace*{\fill} difference 700 to 850 hPA  &   & \\
+\hline
+\textbf{msl{\_}diff}: mean sea level  & 1 & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
+\hspace{1.65cm} difference            &   & (for 6--30 UTC)\\
+\hline
+\end{tabular}
+
+\newpage
+
 The following methods are applied on this data set:
-\begin{itemize}
-\item distributional tree
-\item distributional forest
-\item gamlss
-\item gamboostLSS
-\item EMOS
-\end{itemize}
-
-These methods are compared in a 10x10 cross validation framework based on the resulting RMSE, log-likelihood and CRPS.\\
-These values are illustrated in the following boxplots.
+
+\begin{tabular}{| l | c  | c  | c  | c |}
+\hline
+ & \textbf{mu link} & \textbf{sigma link} & \textbf{mu covariates} & \textbf{sigma covariates} \\
+\hline
+\textbf{disttree} & identity & log & all covariates & all covariates \\ 
+\hline
+\textbf{distforest} & identity & log & all covariates & all covariates \\
+\hline
+\textbf{gamlss} & identity & log & tppow{\_}mean, & tppow{\_}sprd,\\
+& & & tppow{\_}max, & dswrf{\_}sprd{\_}mean, \\
+& & & tppow{\_}mean1218, & tppow{\_}sprd1218,\\ 
+& & & capepow{\_}mean1218, & capepow{\_}mean1218,\\
+& & & dswrf{\_}mean{\_}mean, & tcolc{\_}sprd{\_}mean,\\
+& & & tcolc{\_}mean{\_}mean, & tdiff500850{\_}mean\\
+& & & pwat{\_}mean{\_}mean, & \\
+& & & tdiff500850{\_}mean, & \\
+& & & msl{\_}diff & \\
+\hline
+\textbf{gamboostLSS} & identity & log & all covariates & all covariates \\
+\hline
+\textbf{EMOS} & identity & log & tppow{\_}mean & tppow{\_}spread \\
+\hline
+\end{tabular}
+\\
+manuall selected covariates in gamlss\\
+mention interactions:\\
+tppow{\_}mean1218 $\cdot$ capepow{\_}mean1218,\\
+tppow{\_}sprd1218 $\cdot$ capepow{\_}mean1218\\
+\\
+(advantage: no sepcification needed in disttree/distforest)
+
+\SweaveOpts{eval = TRUE}
+\newpage
+\subsection{Station Axams}
+one station \\
+Axams, closest to Ibk\\
+ca 800 observations\\
+
+\subsubsection{Predictions}
+Fit one model on the first 25 years.\\
+predictions for the 26th, 27th and 28th year\\
+plot predicted distributions together with obersvations for one day in July of 26th, 27th and 28th year  
+<<echo=FALSE, results=hide>>=
+#### Axams prediction 25 - 3
+if(file.exists("rain_Axams_pred.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
+  load("rain_Axams_pred.rda")
+} else {
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
+  source("rain_axams_pred.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+  res <- rain_axams_pred(seedconst = 7, ntree = 100,
+                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
+                         gamboost_cvr = FALSE,
+                         frac = FALSE)
+
+  save(res, file = "rain_Axams_pred.rda")
+}
+
+# prepare data for plot of estimated density functions
+pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
+pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+cbind(df_exp, res$testdata[pdays,"robs"])
+
+# plot predicted distributions together with observations
+#set.seed(res$call$seedconst)
+set.seed(7)
+x <- c(0.01, sort(runif(500,0.01,8)))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
+
+# point mass (slightly shifted)
+pm1 <- c(0.04, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
+pm2 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
+pm3 <- c(-0.1, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+
+# predictions
+pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+
+#legendheight
+lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
+lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
+lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+@
+
+\begin{figure}[t!]
+<<rain_Axams_pred, fig=TRUE, echo=FALSE>>=
+plot(x = x, y = y1, type = "l", col = "red", 
+     main = paste0("July ", pday, dayending), ylab = "density", 
+     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
+     xlim = c(-1.5,8))
+
+lines(x = x, y = y2, type = "l", col = "blue")
+    lines(x = x, y = y3, type = "l", col = "darkgreen")
+    #legend('topright', c("2010", "2011", "2012"), col = c("red", "blue", "darkgreen"), lty = 1, cex = 1)
+
+    # plot point mass
+    lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
+    lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
+    lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
+    points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
+    points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
+    points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
+
+    # plot predictions
+    points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
+    points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
+    points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
+
+    lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
+
+    # add labels
+    text(x = -0.8, y = lh1, labels = "2010", col = "red", cex = 0.8)
+    text(x = -0.8, y = lh2, labels = "2011", col = "blue", cex = 0.8)
+    text(x = -0.8, y = lh3, labels = "2012", col = "darkgreen", cex = 0.8)
+@
+\caption{\label{}}
+\end{figure}
+
+Variable importance\\
+<<variabble_importance, echo=FALSE, eval=TRUE>>=
+if(file.exists("vimp100_crps.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
+  load("vimp100_crps.rda")
+} else {
+
+  # locally redefine logLik.distforest to use crps in varimp()
+  logLik.distforest <- function(object, newdata = NULL, ...) {
+    if(is.null(newdata)) {
+      newdata <- object$data
+    }
+
+    # predict parameter
+    pdf <- predict(object, newdata = newdata, type = "parameter")
+    df_mu <- pdf$mu
+    df_sigma <- pdf$sigma
+
+    # calculate CRPS
+    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
+
+    return(structure(crps, df = NA, class = "logLik"))
+  }
+
+  #vimp_crps <- varimp(res$df, nperm = 1L)
+  #vimp10_crps <- varimp(res$df, nperm = 10L)
+  vimp100_crps <- varimp(res$df, nperm = 100L)
+  #vimp1000_crps <- varimp(res$df, nperm = 1000L)
+
+  #save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
+  #save(vimp10_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp10_crps.rda")
+  save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
+  #save(vimp1000_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp1000_crps.rda")
+
+  rm(logLik.distforest) 
+}
+
+most_impvar <- sort(vimp100_crps, decreasing = TRUE)[c(1:20)]
+most_impvar <- most_impvar[most_impvar>2.7]
+names_mi <- names(most_impvar)
+covnames <- colnames(res$learndata)
+id_mi <- match(names_mi, covnames)
+
+datafix <- res$learndata
+datafix1 <- datafix2 <- datafix3 <- datafix4 <- 
+  datafix5 <- datafix5 <- datafix6 <- datafix7 <- datafix8 <- datafix
+
+# fix all variables except for the most important ones to their mean
+for(i in 3:NCOL(datafix))  if(i!=id_mi[1]) datafix1[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+for(i in 3:NCOL(datafix))  if(i!=id_mi[2]) datafix2[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+for(i in 3:NCOL(datafix))  if(i!=id_mi[3]) datafix3[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+for(i in 3:NCOL(datafix))  if(i!=id_mi[4]) datafix4[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+for(i in 3:NCOL(datafix))  if(i!=id_mi[5]) datafix5[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+for(i in 3:NCOL(datafix))  if(i!=id_mi[6]) datafix6[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+for(i in 3:NCOL(datafix))  if(i!=id_mi[7]) datafix7[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+for(i in 3:NCOL(datafix))  if(i!=id_mi[8]) datafix8[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
+
+pred1 <- predict(res$df, newdata  = datafix1, type = "parameter")
+plotdata1 <- cbind(datafix1, pred1)
+sp1 <- plotdata1[order(datafix1[,id_mi[1]]),]
+
+# plot effect of most important variables on parameter
+if(FALSE){
+
+  pred1 <- predict(res$df, newdata  = datafix1, type = "parameter")
+  plotdata1 <- cbind(datafix1, pred1)
+  sp1 <- plotdata1[order(datafix1[,id_mi[1]]),]
+  plot(x = sp1[,id_mi[1]], y = sp1[, "mu"], type = "l", xlab = names_mi[1], ylab = "mu")
+  plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = "sigma")
+
+  pred2 <- predict(res$df, newdata  = datafix2, type = "parameter")
+  plotdata2 <- cbind(datafix2, pred2)
+  sp2 <- plotdata2[order(datafix2[,id_mi[2]]),]
+  plot(x = sp2[,id_mi[2]], y = sp2[, "mu"], type = "l", xlab = names_mi[2], ylab = "mu")
+  plot(x = sp2[,id_mi[2]], y = sp2[, "sigma"], type = "l", xlab = names_mi[2], ylab = "sigma")
+
+  pred3 <- predict(res$df, newdata  = datafix3, type = "parameter")
+  plotdata3 <- cbind(datafix3, pred3)
+  sp3 <- plotdata3[order(datafix3[,id_mi[3]]),]
+  plot(x = sp3[,id_mi[3]], y = sp3[, "mu"], type = "l", xlab = names_mi[3], ylab = "mu")
+  plot(x = sp3[,id_mi[3]], y = sp3[, "sigma"], type = "l", xlab = names_mi[3], ylab = "sigma")
+
+  pred4 <- predict(res$df, newdata  = datafix4, type = "parameter")
+  plotdata4 <- cbind(datafix4, pred4)
+  sp4 <- plotdata4[order(datafix4[,id_mi[4]]),]
+  plot(x = sp4[,id_mi[4]], y = sp4[, "mu"], type = "l", xlab = names_mi[4], ylab = "mu")
+  plot(x = sp4[,id_mi[4]], y = sp4[, "sigma"], type = "l", xlab = names_mi[4], ylab = "sigma")
+
+  pred5 <- predict(res$df, newdata  = datafix5, type = "parameter")
+  plotdata5 <- cbind(datafix5, pred5)
+  sp5 <- plotdata5[order(datafix5[,id_mi[5]]),]
+  plot(x = sp5[,id_mi[5]], y = sp5[, "mu"], type = "l", xlab = names_mi[5], ylab = "mu")
+  plot(x = sp5[,id_mi[5]], y = sp5[, "sigma"], type = "l", xlab = names_mi[5], ylab = "sigma")
+
+  pred6 <- predict(res$df, newdata  = datafix6, type = "parameter")
+  plotdata6 <- cbind(datafix6, pred6)
+  sp6 <- plotdata6[order(datafix6[,id_mi[6]]),]
+  plot(x = sp6[,id_mi[6]], y = sp6[, "mu"], type = "l", xlab = names_mi[6], ylab = "mu")
+  plot(x = sp6[,id_mi[6]], y = sp6[, "sigma"], type = "l", xlab = names_mi[6], ylab = "sigma")
+
+  pred7 <- predict(res$df, newdata  = datafix7, type = "parameter")
+  plotdata7 <- cbind(datafix7, pred7)
+  sp7 <- plotdata7[order(datafix7[,id_mi[7]]),]
+  plot(x = sp7[,id_mi[7]], y = sp7[, "mu"], type = "l", xlab = names_mi[7], ylab = "mu")
+  plot(x = sp7[,id_mi[7]], y = sp7[, "sigma"], type = "l", xlab = names_mi[7], ylab = "sigma")
+
+  pred8 <- predict(res$df, newdata  = datafix8, type = "parameter")
+  plotdata8 <- cbind(datafix8, pred8)
+  sp8 <- plotdata8[order(datafix8[,id_mi[8]]),]
+  plot(x = sp8[,id_mi[8]], y = sp8[, "mu"], type = "l", xlab = names_mi[8], ylab = "mu")
+  plot(x = sp8[,id_mi[8]], y = sp8[, "sigma"], type = "l", xlab = names_mi[8], ylab = "sigma")
+
+}
+
+@
+
+\begin{figure}[t!]
+<<fig = TRUE, echo = FALSE>>=
+par(mfrow = c(1,2))
+plot(x = sp1[,id_mi[1]], y = sp1[, "mu"], type = "l", xlab = names_mi[1], ylab = "mu")
+plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = "sigma")
+@
+\end{figure}
+marginal effect of covariate with highest variable importance in distforest\\
+(other covariates? no effects for 4th variable, probably because only splits after several other splits)
+
+\SweaveOpts{eval = TRUE}
+
+\subsubsection{Cross-Validation}
+randomly split into 10 (roughly) equally sized data sets, learn on 9, evaluate on the 10th\\
+average over 10 variations\\
+repeat 10 times\\
+compare rmse, ll, crps score skills with reference model EMOS log\\
 
@@
 \begin{subfigure}{0.3\textwidth}
-<<rain_rmse_box, fig=TRUE, echo=FALSE>>=
+<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
 boxplot(rain_rmse)
@@
 \begin{subfigure}{0.3\textwidth}
-<<rain_ll_box, fig=TRUE, echo=FALSE>>=
-boxplot(rain_ll)
+<<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
+boxplot(rain_ll, ylim = c(-2.8,-1.3))
+#boxplot(rain_ll)
 @
@@
 \begin{subfigure}{0.3\textwidth}
-<<rain_crps_box, fig=TRUE, echo=FALSE>>=
+<<rain_cross_axams_crps, fig=TRUE, echo=FALSE>>=
 boxplot(rain_crps)
@@
 \begin{subfigure}{0.3\textwidth}
-<<rain_crps_box, fig=TRUE, echo=FALSE>>=
-boxplot(1 - rain_crps / rain_crps[,7]) 
+<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE>>=
+boxplot(1 - rain_crps / rain_crps[,6]) 
 abline(h = 0, col = 2, lwd = 2)
@@
 \end{subfigure}
-\caption{\label{boxplots_rain}Results 10x10 cross validation on the rain data set (RMSE, log-likelihood and CRPS)\\
+\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams (RMSE, log-likelihood and CRPS)\\
+left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
+right: continuous lines represent the location parameter, dotted lines the scale parameter}
+\end{figure}
+
+\subsection{Predictions over all stations}
+
+NAs in gamlss in 13 stations!!!\\
+% sum(is.na(crps[,3]))
+<<eval = TRUE, echo = FALSE>>=
+#### prediction over all stations 25 - 3
+if(file.exists("rain_pred.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred.rda")
+  load("rain_pred.rda")
+} else {
+
+  source("rain_pred.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+
+  res <- rain_pred(seedconst = 7,
+                   gamboost_cvr = TRUE,
+                   frac = FALSE)
+
+  save(res, file = "rain_pred.rda")
+}
+
+ll <- res[[1]]$results["ll",]
+for(i in 2:(length(res)-1)) ll <- rbind(ll, res[[i]]$results["ll",])
+
+rmse <- res[[1]]$results["rmse",]
+for(i in 2:(length(res)-1)) rmse <- rbind(rmse, res[[i]]$results["rmse",])
+
+crps <- res[[1]]$results["crps",]
+for(i in 2:(length(res)-1))  crps <- rbind(crps, res[[i]]$results["crps",])
+
+colnames(ll) <- colnames(rmse) <- colnames(crps) <- colnames(res[[1]]$results)
+
+@
+
+\begin{figure}[t!]
+\begin{subfigure}{0.3\textwidth}
+<<rain_pred_rmse, fig=TRUE, echo=FALSE>>=
+boxplot(rmse)
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.3\textwidth}
+<<rain_pred_ll, fig=TRUE, echo=FALSE>>=
+boxplot(ll, ylim = c(-2.8,-1.3))
+#boxplot(ll)
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.3\textwidth}
+<<rain_pred_crps, fig=TRUE, echo=FALSE>>=
+boxplot(crps)
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.3\textwidth}
+<<rain_pred_skills_score, fig=TRUE, echo=FALSE>>=
+# skills score
+  boxplot(1 - crps/crps[,6])
+  abline(h = 0, col = "red")
+@
+\end{subfigure}
+\caption{\label{boxplots_rain}Results of predictions over all stations (RMSE, log-likelihood and CRPS)\\
 left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\

### 2017-10-08 pkg/disttree/inst/draft/draft.Rnw
restructured plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1635 255f4249-2263-4417-b1ea-8498779e27a7

@@
 @
-\caption{\label{}}
+\caption{\label{}Predictions at station Axams for one day in July of 2010, 2011 and 2012}
 \end{figure}
@@
 \begin{figure}[t!]
-<<fig = TRUE, echo = FALSE>>=
-par(mfrow = c(1,2))
+\begin{subfigure}{0.5\textwidth}
+<<rain_axams_varim_margeffects1_mu, fig=TRUE, echo=FALSE>>=
 plot(x = sp1[,id_mi[1]], y = sp1[, "mu"], type = "l", xlab = names_mi[1], ylab = "mu")
+@
+\end{subfigure}
+\hspace{-1cm}
+\begin{subfigure}{0.5\textwidth}
+<<rain_axams_varim_margeffects1_sigma, fig=TRUE, echo=FALSE>>=
 plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = "sigma")
 @
+\end{subfigure}
+\caption{\label{varimp}marginal effect of covariate with highest variable importance in distforest}
 \end{figure}
-marginal effect of covariate with highest variable importance in distforest\\
 (other covariates? no effects for 4th variable, probably because only splits after several other splits)
@@
 
+\newpage
 \subsubsection{Cross-Validation}
@@
 \begin{figure}[t!]
-\begin{subfigure}{0.3\textwidth}
+\begin{subfigure}{0.5\textwidth}
 <<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
@@
 \hspace{-1cm}
-\begin{subfigure}{0.3\textwidth}
+\begin{subfigure}{0.5\textwidth}
 <<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
@@
 \end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.3\textwidth}
-<<rain_cross_axams_crps, fig=TRUE, echo=FALSE>>=
-boxplot(rain_crps)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.3\textwidth}
+\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
+left: RMSE, right: log-likelihood}
+\end{figure}
+
+\begin{figure}[t!]
 <<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE>>=
@@
 @
-\end{subfigure}
-\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams (RMSE, log-likelihood and CRPS)\\
-left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
-right: continuous lines represent the location parameter, dotted lines the scale parameter}
+\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams \\
+CRPS skills score with reference model EMOS log}
 \end{figure}
 
+\newpage
 \subsection{Predictions over all stations}
@@
 NAs in gamlss in 13 stations!!!\\
+for example in 13th station in complete{\_}stations\\
+\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  : 
+  NA's in the working vector or weights for parameter mu}
+\\
 % sum(is.na(crps[,3]))
@@
 \begin{figure}[t!]
-\begin{subfigure}{0.3\textwidth}
-<<rain_pred_rmse, fig=TRUE, echo=FALSE>>=
-boxplot(rmse)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.3\textwidth}
-<<rain_pred_ll, fig=TRUE, echo=FALSE>>=
-boxplot(ll, ylim = c(-2.8,-1.3))
-#boxplot(ll)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.3\textwidth}
-<<rain_pred_crps, fig=TRUE, echo=FALSE>>=
-boxplot(crps)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.3\textwidth}
 <<rain_pred_skills_score, fig=TRUE, echo=FALSE>>=
@@
 @
-\end{subfigure}
-\caption{\label{boxplots_rain}Results of predictions over all stations (RMSE, log-likelihood and CRPS)\\
-left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
-right: continuous lines represent the location parameter, dotted lines the scale parameter}
+\caption{\label{boxplots_rain}Results of predictions over all stations: CRPS skills score with reference model EMOS log}
 \end{figure}

### 2017-10-09 pkg/disttree/inst/draft/draft.Rnw
restructured tables
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1643 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usepackage{enumitem}
+\usepackage{array, makecell}
 \usepackage{tikz}
@@
 In particular, 14 basic forecast variables are considered with up to 12 variations of these variables being included as covariates.
-Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standarad deviation, the minimum and the maximum over all these individual predictions.
-Another variation refers to the time periods which can be over 24 hours or over time windows of 6 hours.
-Another possible option is to consider the mean over 24 hours for each individual prediction and then take the mean over the whole ensemble.
-In the following table listing all covariates this last option is referred to as the ensemble mean of mean.\\
+Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standarad deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble min and ensemble max.
+All of these values can be calculated on the whole day (6--30 UTC) or a 6h time window of the day.\\
+Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standarad deviation, noted as ensemble mean of mean/min/max or ensemble std dev of mean/min/max respectively.\\
+The following table shows all variables together with the number of variations and the provided variations listed in the last column.\\
+\\
 The response variable total precicipitation is power transformed by a factor of 1.6.
@@
 \newpage
-\begin{tabular}{| l | c | l |}
+\begin{tabular}{| l  c  l |}
 \hline
-basic covariates & nr & variations\\
+\thead{basic covariates} & \thead{nr} & \thead{variations}\\
 \hline
-\textbf{tp{\_}pow}: total precipitation    & 12 & ensemble mean, ensemble std dev, \\
-\hspace*{\fill} power transformed (by 1.6) &    & ensemble min, ensemble max\\
-\textbf{cape{\_}pow}: convective           &    & (all for 6--30 UTC)\\
-\hspace*{0.4cm} available potential energy &    & ensemble mean: 6--12, 12--18,\\
-\hspace*{\fill}  power transformed (by 1.6)&    & \hspace{2.5cm} 18--24, 24--30 UTC\\
-\hspace*{\fill}                            &    & ensemble std dev: 6--12, 12--18,\\
-\hspace*{\fill}                            &    & \hspace{2.85cm}  18--24, 24--30 UTC\\
+\textbf{tp{\_}pow}: total precipitation    & 12 variations:  & ensemble mean of sum, \\
+\hspace*{\fill} power transformed (by 1.6) &    & ensemble std dev of sum, \\ 
+\textbf{cape{\_}pow}: convective           &    & ensemble min of sum, \\
+\hspace*{0.4cm} available potential energy &    & ensemble max of sum\\
+\hspace*{\fill}  power transformed (by 1.6)&    & \qquad all over 6--30 UTC\\
+\hspace*{\fill}                            &    & ensemble mean of sum over\\
+\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
+\hspace*{\fill}                            &    & ensemble std dev of sum over\\
+\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
 \hline
-\textbf{dswrf}: downwards short ware & 6 & ensemble mean of mean, \\
+\textbf{dswrf}: downwards short ware      & 6 variations: & ensemble mean of mean, \\
 \hspace*{\fill} radiation flux (sunshine) &   & ensemble mean of min,\\
@@
 \hspace*{1.15cm} temperature              &   & ensemble std dev of max,\\
-\textbf{tcolc}: total column-integrated &   & (all for 6--30 UTC)\\
+\textbf{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
 \hspace*{1.05cm} condensate  &   & \\
@@
 \hline
-\textbf{tdiff500850}: temperature         & 3 & ensemble mean,\\
-\hspace*{\fill} difference 500 to 850 hPA &   & ensemble min,\\
-\textbf{tdiff500700}: temperature         &  & ensemble max\\
-\hspace*{\fill} difference 500 to 700 hPA &   & (all for 6--30 UTC)\\
+\textbf{tdiff500850}: temperature         & 3 variations: & ensemble mean of diff in mean,\\
+\hspace*{\fill} difference 500 to 850 hPA &   & ensemble min of diff in mean,\\
+\textbf{tdiff500700}: temperature         &   & ensemble max of diff in mean\\
+\hspace*{\fill} difference 500 to 700 hPA &   & \qquad all over 6--30 UTC\\
 \textbf{tdiff700850}: temperature         &   & \\
@@
 \hline
-\textbf{msl{\_}diff}: mean sea level  & 1 & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
-\hspace{1.65cm} difference            &   & (for 6--30 UTC)\\
+\textbf{msl{\_}diff}: mean sea level  & 1 variation: & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
+\hspace{1.65cm} difference            &   & \qquad over 6--30 UTC\\
 \hline
@@
 
-\begin{tabular}{| l | c  | c  | c  | c |}
+\begin{tabular}{| l | c   c   c   c |}
 \hline
@@
 \\
+
 manuall selected covariates in gamlss\\
@@
 Axams, closest to Ibk\\
-ca 800 observations\\
+865 observations\\
 
 \subsubsection{Predictions}
-Fit one model on the first 25 years.\\
+Fit one model on the first 25 years (1985--2009, both included)\\
 predictions for the 26th, 27th and 28th year\\
@@
                          forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
-                         gamboost_cvr = FALSE,
+                         gamboost_cvr = TRUE,
                          frac = FALSE)
@@
 for example in 13th station in complete{\_}stations\\
-\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  : 
-  NA's in the working vector or weights for parameter mu}
+\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
+\text{NA's in the working vector or weights for parameter mu}
 \\

### 2017-10-09 pkg/disttree/inst/draft/draft.Rnw
further descriptions in the rain section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1644 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \\
-
-manuall selected covariates in gamlss\\
-mention interactions:\\
-tppow{\_}mean1218 $\cdot$ capepow{\_}mean1218,\\
-tppow{\_}sprd1218 $\cdot$ capepow{\_}mean1218\\
-\\
-(advantage: no sepcification needed in disttree/distforest)
+For the gamlss model the covariates were selected manually based on expert knowledge.
+Additionally, interactions are specified in advance, once between tppow{\_}mean1218 and capepow{\_}mean1218 for the location parameter and once between tppow{\_}sprd1218 and capepow{\_}mean1218 for the l ocatioin parameter.
+By putting extra effort in the specification of the formulas the gamlss model is provided with additional information and is therefore expected to profit from this advantage.\\
 
@@
 \subsection{Station Axams}
-one station \\
-Axams, closest to Ibk\\
-865 observations\\
+First of all we focus on one observation station, Axams in Tyrol. This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol.\\
+As for all other stations, daily observations of the month of July of 28 years are available, starting in 1985 and ending in 2012.\\
 
 \subsubsection{Predictions}
-Fit one model on the first 25 years (1985--2009, both included)\\
-predictions for the 26th, 27th and 28th year\\
-plot predicted distributions together with obersvations for one day in July of 26th, 27th and 28th year  
+In a first step one model is fit on the first 25 years (1985--2009, both included) and then predictions for the 26th, 27th and 28th year (2010, 2011 and 2012) are made.\\
+To illustrate the results, one day of July is chosen and the estimated distributions for this day of all three prediction years are plotted together with the corresponding obersvations.  
 <<echo=FALSE, results=hide>>=
@@
 
-Variable importance\\
+Analyzing variable importance of the distforest model leads to the following list of the covariates which show the highest values in variable importance.\\
+In the following plot the marginal effect of these variables on the distribution parameters are plotted. (All other covariates are kept at their mean.)
 <<variabble_importance, echo=FALSE, eval=TRUE>>=

### 2017-10-10 pkg/disttree/inst/draft/draft.Rnw
changed tables and skills score plot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1648 255f4249-2263-4417-b1ea-8498779e27a7

@@
    \And Torsten Hothorn\\Universit\"at Z\"urich
-   \And Achim Zeileis\\Universit\"at Innsbruck
-   \And Reto Stauffer\\Universit\"at Innsbruck}
-\Plainauthor{Lisa Schlosser, Torsten Hothorn, Achim Zeileis, Reto Stauffer}
+   \And Reto Stauffer\\Universit\"at Innsbruck
+   \And Achim Zeileis\\Universit\"at Innsbruck}
+\Plainauthor{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
 
@@
 \newpage
+\begin{table}[h]
+%\resizebox{1.4\textwidth}{!}{
+\begin{minipage}{\textwidth}
 \begin{tabular}{| l  c  l |}
 \hline
-\thead{basic covariates} & \thead{nr} & \thead{variations}\\
+Basic covariates & {\#} & Variations\\
 \hline
-\textbf{tp{\_}pow}: total precipitation    & 12 variations:  & ensemble mean of sum, \\
+\textbf{tp}: total precipitation            & 12 variations:  & ensemble mean of sum, \\
 \hspace*{\fill} power transformed (by 1.6) &    & ensemble std dev of sum, \\ 
-\textbf{cape{\_}pow}: convective           &    & ensemble min of sum, \\
-\hspace*{0.4cm} available potential energy &    & ensemble max of sum\\
+\textbf{cape}: convective                  &    & ensemble minimum of sum, \\
+\hspace*{0.4cm} available potential energy &    & ensemble maximum of sum\\
 \hspace*{\fill}  power transformed (by 1.6)&    & \qquad all over 6--30 UTC\\
@@
 \hline
-\textbf{dswrf}: downwards short ware      & 6 variations: & ensemble mean of mean, \\
-\hspace*{\fill} radiation flux (sunshine) &   & ensemble mean of min,\\
-\textbf{msl}: mean sea level pressure     &   &  ensemble mean of max,\\
-\textbf{pwat}: preciptable water          &   & ensemble std dev of mean,\\
-\textbf{tmax}: 2m maximum                 &   & ensemble std dev of min,\\
-\hspace*{1.15cm} temperature              &   & ensemble std dev of max,\\
+\textbf{dswrf}: downwards short ware      & 6 variations: & ensemble mean of mean values, \\
+\hspace*{\fill} radiation flux (sunshine) &   & ensemble mean of minimal values,\\
+\textbf{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
+\textbf{pwat}: preciptable water          &   & ensemble std dev of mean values,\\
+\textbf{tmax}: 2m maximum                 &   & ensemble std dev of minimal values,\\
+\hspace*{1.15cm} temperature              &   & ensemble std dev of maximal values,\\
 \textbf{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
@@
 \hline
-\textbf{tdiff500850}: temperature         & 3 variations: & ensemble mean of diff in mean,\\
-\hspace*{\fill} difference 500 to 850 hPA &   & ensemble min of diff in mean,\\
-\textbf{tdiff500700}: temperature         &   & ensemble max of diff in mean\\
+\textbf{tdiff500850}: temperature         & 3 variations: & ensemble mean of difference in mean,\\
+\hspace*{\fill} difference 500 to 850 hPA &   & ensemble minimum of difference in mean,\\
+\textbf{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
 \hspace*{\fill} difference 500 to 700 hPA &   & \qquad all over 6--30 UTC\\
@@
 \end{tabular}
+\caption[Table caption text]{Basic covariates together with the number and the type of variations}
+\label{covariates}
+\end{minipage} 
+%}
+\end{table}
 
@@
 
-\begin{tabular}{| l | c   c   c   c |}
+\begin{table}[h]
+%\resizebox{1.4\textwidth}{!}{
+\begin{minipage}{\textwidth}
+\begin{tabular}{| l | l   l   l |}
 \hline
- & \textbf{mu link} & \textbf{sigma link} & \textbf{mu covariates} & \textbf{sigma covariates} \\
+ & \textbf{Type} & \textbf{Location ($\mu$)} & \textbf{Scale ($\sigma$)} \\
 \hline
-\textbf{disttree} & identity & log & all covariates & all covariates \\ 
+\textbf{Forest} & recursive    & all & all \\ 
+                & partitioning &     &     \\ 
 \hline
-\textbf{distforest} & identity & log & all covariates & all covariates \\
+\textbf{GAM (prespecified)} & spline  & tp{\_}mean, & tppow{\_}sprd,\\
+                            & in each & tp{\_}max,  & dswrf{\_}sprd{\_}mean,\\
+ & & tp{\_}mean1218 $\cdot$ & tp{\_}sprd1218 $\cdot$\\ 
+ & & \quad cape{\_}mean1218, & \quad  cape{\_}mean1218,\\
+ & & dswrf{\_}mean{\_}mean, & tcolc{\_}sprd{\_}mean,\\
+ & & tcolc{\_}mean{\_}mean, & tdiff500850{\_}mean\\
+ & & pwat{\_}mean{\_}mean, & \\
+ & & tdiff500850{\_}mean, & \\
+ & & msl{\_}diff & \\
 \hline
-\textbf{gamlss} & identity & log & tppow{\_}mean, & tppow{\_}sprd,\\
-& & & tppow{\_}max, & dswrf{\_}sprd{\_}mean, \\
-& & & tppow{\_}mean1218, & tppow{\_}sprd1218,\\ 
-& & & capepow{\_}mean1218, & capepow{\_}mean1218,\\
-& & & dswrf{\_}mean{\_}mean, & tcolc{\_}sprd{\_}mean,\\
-& & & tcolc{\_}mean{\_}mean, & tdiff500850{\_}mean\\
-& & & pwat{\_}mean{\_}mean, & \\
-& & & tdiff500850{\_}mean, & \\
-& & & msl{\_}diff & \\
+\textbf{GAM (boosted)} & spline  & all & all \\
+                       & in each &     &     \\
 \hline
-\textbf{gamboostLSS} & identity & log & all covariates & all covariates \\
-\hline
-\textbf{EMOS} & identity & log & tppow{\_}mean & tppow{\_}spread \\
+\textbf{EMOS} & linear & tp{\_}mean & tp{\_}spread \\
 \hline
 \end{tabular}
-\\
+\caption[Table caption text]{Table of applied methods together with their type (way of including the covariates) and the included covariates for each distribution parameter}
+\label{model_specification}
+\end{minipage} 
+%}
+\end{table}
+
 For the gamlss model the covariates were selected manually based on expert knowledge.
@@
 
+# skills score
+s <- 1 - crps[, 2:4]/crps[,6]
 @
@@
 <<rain_pred_skills_score, fig=TRUE, echo=FALSE>>=
-# skills score
-  boxplot(1 - crps/crps[,6])
-  abline(h = 0, col = "red")
+  matplot(t(s), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, xlab = "", ylab = "", xlim = c(0.5, 3.5))
+  boxplot(s, add = TRUE)
+  abline(h = 0, col = "gold", lwd = 2)
 @

### 2017-10-10 pkg/disttree/inst/draft/draft.Rnw
variable importance: barplot of top 10 instead of plot of marginal effects on distribution parameter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1649 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+  Reto Stauffer \\
+  Department of Statistics \\
+  Faculty of Economics and Statistics \\
+  Universit\"at Innsbruck \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
+  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/stauffer/reto-stauffer.html}\\
+
   Achim Zeileis \\
@@
 
-  Reto Stauffer \\
-  Department of Statistics \\
-  Faculty of Economics and Statistics \\
-  Universit\"at Innsbruck \\
-  Universit\"atsstr.~15 \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/stauffer/reto-stauffer.html}\\
 }
@@
 <<variabble_importance, echo=FALSE, eval=TRUE>>=
-if(file.exists("vimp100_crps.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
-  load("vimp100_crps.rda")
+if(file.exists("vimp_crps.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
+  load("vimp_crps.rda")
 } else {
@@
 
-  #vimp_crps <- varimp(res$df, nperm = 1L)
+  vimp_crps <- varimp(res$df, nperm = 1L)
   #vimp10_crps <- varimp(res$df, nperm = 10L)
-  vimp100_crps <- varimp(res$df, nperm = 100L)
+  #vimp100_crps <- varimp(res$df, nperm = 100L)
   #vimp1000_crps <- varimp(res$df, nperm = 1000L)
 
-  #save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
+  save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
   #save(vimp10_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp10_crps.rda")
-  save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
+  #save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
   #save(vimp1000_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp1000_crps.rda")
@@
 
-most_impvar <- sort(vimp100_crps, decreasing = TRUE)[c(1:20)]
-most_impvar <- most_impvar[most_impvar>2.7]
-names_mi <- names(most_impvar)
-covnames <- colnames(res$learndata)
-id_mi <- match(names_mi, covnames)
-
-datafix <- res$learndata
-datafix1 <- datafix2 <- datafix3 <- datafix4 <- 
-  datafix5 <- datafix5 <- datafix6 <- datafix7 <- datafix8 <- datafix
-
-# fix all variables except for the most important ones to their mean
-for(i in 3:NCOL(datafix))  if(i!=id_mi[1]) datafix1[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-for(i in 3:NCOL(datafix))  if(i!=id_mi[2]) datafix2[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-for(i in 3:NCOL(datafix))  if(i!=id_mi[3]) datafix3[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-for(i in 3:NCOL(datafix))  if(i!=id_mi[4]) datafix4[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-for(i in 3:NCOL(datafix))  if(i!=id_mi[5]) datafix5[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-for(i in 3:NCOL(datafix))  if(i!=id_mi[6]) datafix6[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-for(i in 3:NCOL(datafix))  if(i!=id_mi[7]) datafix7[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-for(i in 3:NCOL(datafix))  if(i!=id_mi[8]) datafix8[,i] <- rep(mean(datafix[,i], na.rm = TRUE), NROW(datafix))
-
-pred1 <- predict(res$df, newdata  = datafix1, type = "parameter")
-plotdata1 <- cbind(datafix1, pred1)
-sp1 <- plotdata1[order(datafix1[,id_mi[1]]),]
-
-# plot effect of most important variables on parameter
-if(FALSE){
-
-  pred1 <- predict(res$df, newdata  = datafix1, type = "parameter")
-  plotdata1 <- cbind(datafix1, pred1)
-  sp1 <- plotdata1[order(datafix1[,id_mi[1]]),]
-  plot(x = sp1[,id_mi[1]], y = sp1[, "mu"], type = "l", xlab = names_mi[1], ylab = "mu")
-  plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = "sigma")
-
-  pred2 <- predict(res$df, newdata  = datafix2, type = "parameter")
-  plotdata2 <- cbind(datafix2, pred2)
-  sp2 <- plotdata2[order(datafix2[,id_mi[2]]),]
-  plot(x = sp2[,id_mi[2]], y = sp2[, "mu"], type = "l", xlab = names_mi[2], ylab = "mu")
-  plot(x = sp2[,id_mi[2]], y = sp2[, "sigma"], type = "l", xlab = names_mi[2], ylab = "sigma")
-
-  pred3 <- predict(res$df, newdata  = datafix3, type = "parameter")
-  plotdata3 <- cbind(datafix3, pred3)
-  sp3 <- plotdata3[order(datafix3[,id_mi[3]]),]
-  plot(x = sp3[,id_mi[3]], y = sp3[, "mu"], type = "l", xlab = names_mi[3], ylab = "mu")
-  plot(x = sp3[,id_mi[3]], y = sp3[, "sigma"], type = "l", xlab = names_mi[3], ylab = "sigma")
-
-  pred4 <- predict(res$df, newdata  = datafix4, type = "parameter")
-  plotdata4 <- cbind(datafix4, pred4)
-  sp4 <- plotdata4[order(datafix4[,id_mi[4]]),]
-  plot(x = sp4[,id_mi[4]], y = sp4[, "mu"], type = "l", xlab = names_mi[4], ylab = "mu")
-  plot(x = sp4[,id_mi[4]], y = sp4[, "sigma"], type = "l", xlab = names_mi[4], ylab = "sigma")
-
-  pred5 <- predict(res$df, newdata  = datafix5, type = "parameter")
-  plotdata5 <- cbind(datafix5, pred5)
-  sp5 <- plotdata5[order(datafix5[,id_mi[5]]),]
-  plot(x = sp5[,id_mi[5]], y = sp5[, "mu"], type = "l", xlab = names_mi[5], ylab = "mu")
-  plot(x = sp5[,id_mi[5]], y = sp5[, "sigma"], type = "l", xlab = names_mi[5], ylab = "sigma")
-
-  pred6 <- predict(res$df, newdata  = datafix6, type = "parameter")
-  plotdata6 <- cbind(datafix6, pred6)
-  sp6 <- plotdata6[order(datafix6[,id_mi[6]]),]
-  plot(x = sp6[,id_mi[6]], y = sp6[, "mu"], type = "l", xlab = names_mi[6], ylab = "mu")
-  plot(x = sp6[,id_mi[6]], y = sp6[, "sigma"], type = "l", xlab = names_mi[6], ylab = "sigma")
-
-  pred7 <- predict(res$df, newdata  = datafix7, type = "parameter")
-  plotdata7 <- cbind(datafix7, pred7)
-  sp7 <- plotdata7[order(datafix7[,id_mi[7]]),]
-  plot(x = sp7[,id_mi[7]], y = sp7[, "mu"], type = "l", xlab = names_mi[7], ylab = "mu")
-  plot(x = sp7[,id_mi[7]], y = sp7[, "sigma"], type = "l", xlab = names_mi[7], ylab = "sigma")
-
-  pred8 <- predict(res$df, newdata  = datafix8, type = "parameter")
-  plotdata8 <- cbind(datafix8, pred8)
-  sp8 <- plotdata8[order(datafix8[,id_mi[8]]),]
-  plot(x = sp8[,id_mi[8]], y = sp8[, "mu"], type = "l", xlab = names_mi[8], ylab = "mu")
-  plot(x = sp8[,id_mi[8]], y = sp8[, "sigma"], type = "l", xlab = names_mi[8], ylab = "sigma")
-
-}
-
 @
@@
 \begin{figure}[t!]
-\begin{subfigure}{0.5\textwidth}
+%\begin{subfigure}{0.5\textwidth}
 <<rain_axams_varim_margeffects1_mu, fig=TRUE, echo=FALSE>>=
-plot(x = sp1[,id_mi[1]], y = sp1[, "mu"], type = "l", xlab = names_mi[1], ylab = "mu")
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.5\textwidth}
-<<rain_axams_varim_margeffects1_sigma, fig=TRUE, echo=FALSE>>=
-plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = "sigma")
-@
-\end{subfigure}
-\caption{\label{varimp}marginal effect of covariate with highest variable importance in distforest}
+par(mar = c(2,10,2,2))
+barplot(sort(vimp_crps, decreasing = TRUE)[1:10], horiz = TRUE, las = 1, axes = FALSE)
+@
+%\end{subfigure}
+%\hspace{-1cm}
+%\begin{subfigure}{0.5\textwidth}
+%<<rain_axams_varim_margeffects1_sigma, fig=TRUE, echo=FALSE>>=
+%plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = %"sigma")
+%@
+%\end{subfigure}
+\caption{\label{varimp} Variable imortance: top 10 list for the forest model}
 \end{figure}
-(other covariates? no effects for 4th variable, probably because only splits after several other splits)
 

### 2017-10-11 pkg/disttree/inst/draft/draft.Rnw
added PIT-histograms and QQR-plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1651 255f4249-2263-4417-b1ea-8498779e27a7

@@
 library("MASS")
+library("countreg")
 
@@
 par(mar = c(2,10,2,2))
-barplot(sort(vimp_crps, decreasing = TRUE)[1:10], horiz = TRUE, las = 1, axes = FALSE)
+barplot(sort(vimp_crps, decreasing = TRUE)[1:10], horiz = TRUE, las = 1, cex.names = 0.8, axes = FALSE)
 @
@@
 \caption{\label{varimp} Variable imortance: top 10 list for the forest model}
+\end{figure}
+
+<<pit_qqr, echo=FALSE, eval=TRUE>>=
+
+## pit histograms and qqr-plots
+
+## get predicted parameter
+# for learndata and testdata
+{
+
+  ## disttree
+  # in sample
+  pdt <- predict(res$dt, type = "parameter")
+  dt_mu_l <- pdt$mu 
+  dt_sigma_l <- pdt$sigma 
+  pit_dt_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = dt_mu_l, sd = dt_sigma_l))
+  pit_dt_l[res$learndata[,"robs"]>0, 1] <- pit_dt_l[res$learndata[,"robs"]>0, 2]
+
+  # out of sample
+  pdt <- predict(res$dt, type = "parameter", newdata = res$testdata)
+  dt_mu_t <- pdt$mu 
+  dt_sigma_t <- pdt$sigma 
+  pit_dt_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = dt_mu_t, sd = dt_sigma_t))
+  pit_dt_t[res$testdata[,"robs"]>0, 1] <- pit_dt_t[res$testdata[,"robs"]>0, 2]
+
+  ## distforest
+  # in sample
+  pdf <- predict(res$df, type = "parameter")
+  df_mu_l <- pdf$mu
+  df_sigma_l <- pdf$sigma
+  pit_df_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = df_mu_l, sd = df_sigma_l))
+  pit_df_l[res$learndata[,"robs"]>0, 1] <- pit_df_l[res$learndata[,"robs"]>0, 2]
+
+  # out of sample
+  pdf <- predict(res$df, type = "parameter", newdata = res$testdata)
+  df_mu_t <- pdf$mu
+  df_sigma_t <- pdf$sigma
+  pit_df_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = df_mu_t, sd = df_sigma_t))
+  pit_df_t[res$testdata[,"robs"]>0, 1] <- pit_df_t[res$testdata[,"robs"]>0, 2]
+
+  ## gamlss
+  # in sample
+  g_learndata <- res$learndata
+  g_learndata$robs <- Surv(g_learndata$robs, g_learndata$robs>0, type="left")
+  g_mu_l <- predict(res$g, what = "mu", type = "response")
+  g_sigma_l <- predict(res$g, what = "sigma", type = "response")
+  pit_g_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = g_mu_l, sd = g_sigma_l))
+  pit_g_l[res$learndata[,"robs"]>0, 1] <- pit_g_l[res$learndata[,"robs"]>0, 2]
+
+  # out of sample
+  g_mu_t <- predict(res$g, what = "mu", type = "response", newdata = res$testdata)
+  g_sigma_t <- predict(res$g, what = "sigma", type = "response", newdata = res$testdata)
+  pit_g_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = g_mu_t, sd = g_sigma_t))
+  pit_g_t[res$testdata[,"robs"]>0, 1] <- pit_g_t[res$testdata[,"robs"]>0, 2]
+
+  #gamboostLSS
+  pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response")
+  gb_mu_l <- pgb$mu
+  gb_sigma_l <- pgb$sigma
+  pit_gb_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = gb_mu_l, sd = gb_sigma_l))
+  pit_gb_l[res$learndata[,"robs"]>0, 1] <- pit_gb_l[res$learndata[,"robs"]>0, 2]
+
+  # out of sample
+  pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response", newdata = res$testdata)
+  gb_mu_t <- pgb$mu
+  gb_sigma_t <- pgb$sigma
+  pit_gb_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = gb_mu_t, sd = gb_sigma_t))
+  pit_gb_t[res$testdata[,"robs"]>0, 1] <- pit_gb_t[res$testdata[,"robs"]>0, 2]
+
+  #EMOS
+  ml_mu_l <- predict(res$ml, type = "location")     # returns parameter on response scale
+  ml_sigma_l <- predict(res$ml, type = "scale")
+  pit_ml_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = ml_mu_l, sd = ml_sigma_l))
+  pit_ml_l[res$learndata[,"robs"]>0, 1] <- pit_ml_l[res$learndata[,"robs"]>0, 2]
+
+  # out of sample
+  ml_mu_t <- predict(res$ml, type = "location", newdata = res$testdata)     # returns parameter on response scale
+  ml_sigma_t <- predict(res$ml, type = "scale", newdata = res$testdata)
+  pit_ml_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = ml_mu_t, sd = ml_sigma_t))
+  pit_ml_t[res$testdata[,"robs"]>0, 1] <- pit_ml_t[res$testdata[,"robs"]>0, 2]
+
+}
+
+@
+
+\begin{figure}[t!]
+%\begin{subfigure}{0.5\textwidth}
+<<rain_axams_pit_insample, fig=TRUE, echo=FALSE>>=
+
+# in sample
+set.seed(7)
+par(mfrow = c(2,2))
+pithist(pit_df_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+pithist(pit_g_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+pithist(pit_gb_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+pithist(pit_ml_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+
+@
+\caption{\label{pit_insample} PIT-Histograms for station Axams (in sample)}
+\end{figure}
+
+\begin{figure}[t!]
+%\begin{subfigure}{0.5\textwidth}
+<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE>>=
+
+# out of sample
+set.seed(7)
+
+par(mfrow = c(2,2))
+pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+
+@
+\caption{\label{pit_oosample} PIT-Histograms for station Axams (out of sample)}
+\end{figure}
+
+\begin{figure}[t!]
+%\begin{subfigure}{0.5\textwidth}
+<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
+# in sample
+set.seed(7)
+par(mfrow = c(2,2))
+set.seed(1)
+qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
+qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
+qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
+qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
+
+@
+\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
+\end{figure}
+
+\begin{figure}[t!]
+%\begin{subfigure}{0.5\textwidth}
+<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE>>=
+
+# out of sample
+set.seed(7)
+par(mfrow = c(2,2))
+set.seed(1)
+qqrplot(pit_df_t, nsim = 1, main = "distforest (out of bag)")
+qqrplot(pit_g_t, nsim = 1, main = "gamlss (out of bag)")
+qqrplot(pit_gb_t, nsim = 1, main = "gamboostLSS (out of bag)")
+qqrplot(pit_ml_t, nsim = 1, main = "EMOS (out of bag)")
+
+@
+\caption{\label{qqr_oosample} QQR-Plot for station Axams (out of sample)}
 \end{figure}

### 2017-10-11 pkg/disttree/inst/draft/draft.Rnw
started restructuring introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1655 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 00d108acb8c525c3510b6d65f5cc0cc5bd9ddb99

@@
 \Abstract{
+In the field of probabilistic forecasting classical linear models such as ensemle model output statistics (EMOS) are still commonly and widely used.
+This is due to their simplicity and their interpretability.
+However, the restriction to modelling only linear effects required further extensions.
+Generalized additive models (GAM) which can also capture non-linear effects allow for more flexibilty.\\
+only location\\
+GAMLSS\\
+trees and forests not yet applied\\
+less effort\\
+variable specification\\
+complex terrain\\
+
 In regression analysis one is interested in the relationship between a dependent variable and one or more explanatory variables.

### 2017-10-12 pkg/disttree/inst/draft/draft.Rnw
rewritten introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1656 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8b67b094fd0aa44fd6065c721cb382b831b61842

--- 2017-10-11 pkg/disttree/inst/draft/draft.Rnw
@@
 pal <- hcl(c(10, 128, 260, 290, 30, 90, 180), 100, 50)
-names(pal) <- c("forest", "tree", "gamlss", "randomForest", "bamlss", "gamboostLSS", "cforest")
+names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
 
 pallight <- hcl(c(10, 128, 260, 290, 30, 90, 180), 100, 50, alpha = 0.25)
-names(pallight) <- c("forest", "tree", "gamlss", "randomForest", "bamlss", "gamboostLSS", "cforest")
+names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
 
@@
 }
+
+# plot crpse
+plot_crps <- function(simres, ylim = NULL, legend = TRUE){
+
+  crps <- NULL
+  colnames <- NULL
+  col <- NULL
+  legendnames <- NULL
+  if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
+    crps <- cbind(crps, simres$resmat[,"av.crps.dt"])
+    colnames <- c(colnames, "dt.crps")
+    col <- c(col, pal["tree"])
+    legendnames <- c(legendnames, "disttree")
+  }
+  if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
+    crps <- cbind(crps, simres$resmat[,"av.crps.df"])
+    colnames <- c(colnames, "df.crps")
+    col <- c(col, pal["forest"])
+    legendnames <- c(legendnames, "distforest")
+  }
+  if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
+    crps <- cbind(crps, simres$resmat[,"av.crps.g"])
+    colnames <- c(colnames, "g.true")
+    col <- c(col, pal["gamlss"])
+    legendnames <- c(legendnames, "gamlss")
+  }
+  if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
+    crps <- cbind(crps, simres$resmat[,"av.crps.b"])
+    colnames <- c(colnames, "b.true")
+    col <- c(col, pal["bamlss"])
+    legendnames <- c(legendnames, "bamlss")
+  }
+  if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
+    crps <- cbind(crps, simres$resmat[,"av.crps.gb"])
+    colnames <- c(colnames, "gb.true")
+    col <- c(col, pal["gamboostLSS"])
+    legendnames <- c(legendnames, "gamboostLSS")
+  }
+  if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
+    crps <- cbind(crps, simres$resmat[,"av.crps.rf"])
+    colnames <- c(colnames, "rf.true")
+    col <- c(col, pal["randomForest"])
+    legendnames <- c(legendnames, "randomForest")
+  }
+  if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
+    crps <- cbind(crps, simres$resmat[,"av.crps.cf"])
+    colnames <- c(colnames, "cf.true")
+    col <- c(col, pal["cforest"])
+    legendnames <- c(legendnames, "cforest")
+  }
+
+  colnames(crps) <- colnames
+  if(is.null(ylim)) ylim <- c(min(na.omit(crps)), max(na.omit(crps)))
+
+  plot(x = simres$x.axis, y = crps[,1], type = "l", col = col[1], ylim = ylim,
+       xlab = "kappa", ylab = "CRPS")
+
+  for(i in 2:length(col)){
+    lines(x = simres$x.axis, y = crps[,i], type = "l", col = col[i])
+  }
+  if(legend) legend('topleft', legendnames, 
+                    col = col, lty = 1, cex = 0.7)
+}
 @
@@
 \Abstract{
-In regression analysis one is interested in the relationship between a dependent variable and one or more explanatory variables.
+In the field of probabilistic forecasting classical linear models such as ensemle model output statistics (EMOS) are still common and widely used.
+This is due to their simplicity and their interpretability.
+However, the restriction to modelling only linear effects requires further extensions which allow for more flexibility.
+Generalized additive models for location, scale, and shape (GAMLSS) can also capture non-linear effects as each distribution parameter can be modelled separately by its own GAM.
+A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models.
+However, up to now, this class of models has not yet been integrated in probabilistic forecasting as no general distributional framework is available for these methods.
+Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
+Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge/experience from the meteorology or based on a computationally more demanding boosting approach.
+
+%In regression analysis one is interested in the relationship between a dependent variable and one or more explanatory variables.
 Various methods to fit statistical models to the data set have been developed, starting from linear models considering only the mean of the response variable and ranging to probabilistic models where all parameters of a distribution are fit to the given data set.\\
-If there is a strong variation within the data it might be advantageous to split the data first into more homogeneous subgroups based on given covariates and then fit a local model in each subgroup rather than fitting one global model to the whole data set.
+%If there is a strong variation within the data it might be advantageous to split the data first into more homogeneous subgroups based on given covariates and then fit a local model in each subgroup rather than fitting one global model to the whole data set.
 This can be done by applying regression trees and forests.\\
-Both of these two concepts, parametric modeling and algorithmic trees, have been investigated and developed further, however, mostly separated from each other.
+%Both of these two concepts, parametric modeling and algorithmic trees, have been investigated and developed further, however, mostly separated from each other.
 Therefore, our goal is to embed the progress made in the field of probabilistic modeling in the idea of algorithmic tree and forest models.
@@
 In distributional forests an ensemble of distributional trees is built and used to calculate weights which are then included in the fitting process of a distributional model.
-These forest models can detect smooth effects as well as abrupt changes and interactions in a reasonable way without needing any kind of variable selection or information a!
- bout the expected effects advance and therefore offers a good compromise particularly in complex settings. 
-}
-\Keywords{parametric models, trees and forests, recursive partitioning, maximum likelihood}
+These forest models can detect smooth effects as well as abrupt changes and interactions in a reasonable way without needing any kind of variable selection or information about the expected effects advance and therefore offers a good compromise particularly in complex settings.
+}
+\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}
 
@@
 \section{Introduction}
-
-In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
-Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables.
+A typical application in probabilistic forecasting in meteorology is the analysis of precipitation data.
+The main two main interests lie in estimating the probability of rain as well as predicting the expected amount of precipitation.
+In this research area various methods have been developed and tested in the last decades.
+Basic linear models such as ensemle model output statistics (EMOS) were introduced in ...
+and are still common and widely used.
+Despite their simplicity and interpretability, the limitation to modelling only linear effects has risen the interest in applying more flexible models.
+This feature is provided by generalized additive models (GAM) which can also model non-linear effects.
+To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS) offer a framework in which each distribution parameter can be modelled separately by its own GAM.
+In Figure \ref{Devel_parmod} the development of parametric models starting from simple limear models (LM) and ranging to GAMLSS is illustrated.\\
+These flexible parameteric methods provide well fitting models when applied on precipitation data in a rather smooth/homogeneous setting, e.g.
+areas where no abrupt changes in meteorological/climatological events appear.
+However, for example in complex terrain/mountaineous regions various factors can change rapidly and their effects are difficult to be modelled/represented by a smooth parameteric model.
+In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models.
+Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process.
+Therefore, no additional effort or knowledge for the specification of the model in advance is required.
+However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well.
+In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothened in case of smooth effects while abrupt changes are stil captured by the tree structure of the model.\\
+For these reasons, regression trees and particularly random forests provide valueable features which can contribute to improving models for probabilistic forecasting.
+But, up to now, no general distributional framework is available for these methods.
+Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
+(The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure \ref{Devel_treeforest}.)\\
+
+%In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
+%Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables.
 This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family.
 (In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
-Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination.
+%Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination.
 (In the fitting function \code{gam} types of smoothing terms can be selected.)\\
-However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
-(GAMLSS) (\cite{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
+%However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
+%(GAMLSS) (\cite{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
 In that way a whole distribution can be specified.
 The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
-This development in the field of parametric modeling is illustrated in Figure \ref{Devel_parmod}.\\
+%This development in the field of parametric modeling is illustrated in Figure \ref{Devel_parmod}.\\
 

### 2017-10-12 pkg/disttree/inst/draft/draft.Rnw
small corrections in introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1657 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d17cfcbb10c4c02e8f93c96abc010a4969462b5a

@@
 
-All these parametric models work very well as long as only additive effects appear in the considered data.
+%All these parametric models work very well as long as only additive effects appear in the considered data.
 A different type of model that can capture non-additive effects as well are regression trees.
@@
 A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set.
-Then the results of all trees are combined to get a final model which unlike a tree model c!
- an also deal with smooth effects in a reasonable way.\\ 
-In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
-\\
-In the past the combination of parametric data models and algorithmic tree models has gained more and more interest since it allows for a more flexible and clear structure. \\
-Algorithms that only fit constants to each final node and therefore only model the mean of the response variable (such as CART, Breiman 1984) might lead to very large trees which complicates analyzing and interpreting.
+Then the results of all trees are combined to get a final model which unlike a tree model can also deal with smooth effects in a reasonable way.\\
+%In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
+%\\
+%In the past the combination of parametric data models and algorithmic tree models has gained more and more interest since it allows for a more flexible and clear structure. \\
+%Algorithms that only fit constants to each final node and therefore only model the mean of the response variable (such as CART, Breiman 1984) might lead to very large trees which complicates analyzing and interpreting.
 Algorithms such as GUIDE (Loh 2002), CRUISE (Kim and Loh 2001) and LOTUS (Chan and Loh 2004) were among the first to provide trees with parametric models in each terminal node.
@@
 However, up to now trees which only model the mean of the response variable in its nodes are still more common and widely used.\\
-Random forests (\cite{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\cite{hothorn2015partykit}, \cite{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
-\\
-But as mentioned before, until now the above listed development in the field of probabilistic modeling has not yet been fully integrated in the idea of algorithmic tree and forest models.
+%Random forests (\cite{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\cite{hothorn2015partykit}, \cite{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
+%\\
+%But as mentioned before, until now the above listed development in the field of probabilistic modeling has not yet been fully integrated in the idea of algorithmic tree and forest models.
 Probabilistic models are rarely used in combination with tree algorithms and more flexible models such as GAMLSSs have not yet been applied as models in the nodes of a tree.\\
-\\
-Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests.
+%\\
+%Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests.
 In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc.
@@
 In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
-These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
+%These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
 

### 2017-10-12 pkg/disttree/inst/draft/draft.Rnw
included log-likelihood function for precipitation application in methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1659 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 74dc4faf87f883faecf43b19207c340cef92a159

@@
 The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided.
-Therefore, the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$.
-This is done by the following maximization.
+In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distribution with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
+Therefore, the corresponding log-likelihood function $\ell_{tp}$ with parameter vector $\theta = c(\mu, \sigma)$ is
+$$
+\ell_{tp}(\mu, \sigma; Y) = 
+\begin{cases}
+    \log\left(\frac{1}{\sigma} \cdot \phi(\frac{Y - \mu}{\sigma}) \right), & \text{if } Y > 0\\
+    \vspace{0.1cm}
+    \log\left(\Phi(\frac{-\mu}{\sigma})\right), & \text{if } Y = 0
+\end{cases}
+$$
+where $\phi$ and $\Phi$ are the probability density function and the distribution function of the standard normal distribution $\mathcal{N}(0,1)$.\\
+With this specification the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$. This is done by the following maximization.
 $$
@@
 \newpage
-\newpage
 \section{Rain}
@@
 
-\newpage
-The ensemble forecast for the total precipitation is represented by 12 covariates: 
-\begin{itemize}[noitemsep]
-\item ensemble mean of the sum of the total precipitation from 6--30 UTC
-\item ensemble standard deviation of the sum of the total precipitation from 6--30 UTC
-\item ensemble minimum of the sum of the total precipitation from 6--30 UTC
-\item ensemble maximum of the sum of the total precipitation from 6--30 UTC
-\item ensemble mean of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC and from 24--30 UTC
-\item ensemble standard deviation of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC and from 24--30 UTC
-\end{itemize}
-
-The preciptable water is represented by 6 covariates:
-\begin{itemize}[noitemsep]
-\item the ensemble mean of the mean of preciptable water over the time period 6--30 UTC 
-\item the ensemble mean of the minimum of preciptable water over the time period 6--30 UTC 
-\item the ensemble mean of the maximum of preciptable water over the time period 6--30 UTC 
-\item the ensemble standard deviation of the mean of preciptable water over the time period 6--30 UTC 
-\item the ensemble standard deviation of the minimum of preciptable water over the time period 6--30 UTC 
-\item the ensemble standard deviation of the maximum of preciptable water over the time period 6--30 UTC 
-\end{itemize}
+%\newpage
+%The ensemble forecast for the total precipitation is represented by 12 covariates: 
+%\begin{itemize}[noitemsep]
+%\item ensemble mean of the sum of the total precipitation from 6--30 UTC
+%\item ensemble standard deviation of the sum of the total precipitation from 6--30 UTC
+%\item ensemble minimum of the sum of the total precipitation from 6--30 UTC
+%\item ensemble maximum of the sum of the total precipitation from 6--30 UTC
+%\item ensemble mean of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC %and from 24--30 UTC
+%\item ensemble standard deviation of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, %from 18--24 UTC and from 24--30 UTC
+%\end{itemize}
+
+%The preciptable water is represented by 6 covariates:
+%\begin{itemize}[noitemsep]
+%\item the ensemble mean of the mean of preciptable water over the time period 6--30 UTC 
+%\item the ensemble mean of the minimum of preciptable water over the time period 6--30 UTC 
+%\item the ensemble mean of the maximum of preciptable water over the time period 6--30 UTC 
+%\item the ensemble standard deviation of the mean of preciptable water over the time period 6--30 UTC 
+%\item the ensemble standard deviation of the minimum of preciptable water over the time period 6--30 UTC 
+%\item the ensemble standard deviation of the maximum of preciptable water over the time period 6--30 UTC 
+%\end{itemize}
 

### 2017-10-13 pkg/disttree/inst/draft/draft.Rnw
further changes and new text in section probabilistic precipitation forecasting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1660 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c74d4c49c115f5b487d790e5b1b5703ba5dfc235

@@
 However, the restriction to modelling only linear effects requires further extensions which allow for more flexibility.
-Generalized additive models for location, scale, and shape (GAMLSS) can also capture non-linear effects as each distribution parameter can be modelled separately by its own GAM.
+Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modelled separately by its own GAM.
 A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models.
@@
 \endminipage
-\hspace{1.4cm}
+\hspace{1.7cm}
 \minipage{0.25\textwidth}
@@
 The exact way of determining how and where to split varies over different tree algorithms.
-For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm or the ctree algorithm.
-Summed up, the steps of building a distributional tree using one of these algorithms are
+(For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm or the ctree algorithm.) Summed up, the steps of building a distributional tree using one of these algorithms are
 \begin{enumerate}
@@
 \newpage
-\section{Rain}
-In this application we analyze precipitation data meassured over 28 years at 95 (?) stations all over Tyrol and Vorarlberg.
+\section{Probabilistic Forecasting of Precipitation}
+As explained in the introduction distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valueable in probabilistic forecasting but have not yet been applied in this kind of settings.
+To show their strengths in practice, these new methods are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
+In particular, the data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both included.
+The observations were meassured at 95 (?) stations all over Tyrol (and regions close to the frontier).
 The considered data is provided by .....
-and consists of the meassured total precipiation within 24 hours of each day in July from 1985 until 2012, both included.
-Additionally, numerical forecasts from the ECMWF are used as covariates.
+Next to the variable of interest, the meassured total precipiation within 24 hours of each day, numerical forecasts from the ECMWF are used as covariates.
 These forecasts are ensemble predictions consisting of ...
@@
 Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standarad deviation, noted as ensemble mean of mean/min/max or ensemble std dev of mean/min/max respectively.\\
-The following table shows all variables together with the number of variations and the provided variations listed in the last column.\\
-\\
+Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.\\
 The response variable total precicipitation is power transformed by a factor of 1.6.
@@
 
-\newpage
-\begin{table}[h]
+%\newpage
+\begin{table}[t!]
 %\resizebox{1.4\textwidth}{!}{
@@
 
-\newpage
-
-The following methods are applied on this data set:
-
-\begin{table}[h]
+Next to distributional forests other methods / the same methods as in the simulation section are applied to this data set and their results are compared based on RMSE, log-likelihood values and CRPS values.
+The chosen methods and their specifications are listed in Table \ref{model_specification}.
+\begin{table}[t!]
 %\resizebox{1.4\textwidth}{!}{
@@
 
-For the gamlss model the covariates were selected manually based on expert knowledge.
-Additionally, interactions are specified in advance, once between tppow{\_}mean1218 and capepow{\_}mean1218 for the location parameter and once between tppow{\_}sprd1218 and capepow{\_}mean1218 for the l ocatioin parameter.
+For the gamlss model the covariates were selected manually based on meteorological knowledge and experience.
+Additionally, interactions are specified in advance, once between tp{\_}mean1218 and cape{\_}mean1218 for the location parameter and once between tp{\_}sprd1218 and cape{\_}mean1218 for the locatioin parameter.
 By putting extra effort in the specification of the formulas the gamlss model is provided with additional information and is therefore expected to profit from this advantage.\\

### 2017-10-18 pkg/model4you/inst/JORS/man_model4you.Rnw
small changes of text
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1661 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 414146413ce2ec20a2a697e59276c2b39360e41b

@@
 \Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::cforest()'} and extends the
-functionalities where needed to improve usability and interpretability.
+functionalities to allow for the computation of personalised models and to 
+improve usability and interpretability.
 
@@
 CRAN), \emph{trtf}, \emph{disttree}, \emph{lagsarlmtree} and \emph{palmtree}
-(all available on R-Forge with a plan to move them to CRAN).
+(all available on R-Forge, publishing on CRAN planned).
 

### 2017-10-18 pkg/disttree/inst/draft/draft.Rnw
added description of RMSE, CRPS and log-likelihood; changed level colors of mapplot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1663 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d7b8e4498074b174aa7e8f052f99094e1c1ba0a0

@@
 
-%\VignetteIndexEntry{ctree: Conditional Inference Trees}
-%\VignetteDepends{coin, TH.data, survival, strucchange}
-%\VignetteKeywords{conditional inference, non-parametric models, recursive partitioning}
-%\VignettePackage{partykit}
+%\VignetteIndexEntry{disttree: Distributional Trees and Forests}
+%\VignetteDepends{partykit, gamlss.dist}
+%\VignetteKeywords{regression trees, random forests, distributional regression, recursive partitioning}
+%\VignettePackage{disttree}
 
@@
 ## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 30, 90, 180), 100, 50)
+pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
 names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
 
-pallight <- hcl(c(10, 128, 260, 290, 30, 90, 180), 100, 50, alpha = 0.25)
+pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
 names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
@@
 \Abstract{
-In the field of probabilistic forecasting classical linear models such as ensemle model output statistics (EMOS) are still common and widely used.
+In the field of probabilistic forecasting classical linear models such as ensemble model output statistics (EMOS) are still common and widely used.
 This is due to their simplicity and their interpretability.
 However, the restriction to modelling only linear effects requires further extensions which allow for more flexibility.
-Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modelled separately by its own GAM.
+Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own GAM.
 A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models.
@@
 In this research area various methods have been developed and tested in the last decades.
-Basic linear models such as ensemle model output statistics (EMOS) were introduced in ...
+Basic linear models such as ensemble model output statistics (EMOS) were introduced in ...
 and are still common and widely used.
@@
 This feature is provided by generalized additive models (GAM) which can also model non-linear effects.
-To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS) offer a framework in which each distribution parameter can be modelled separately by its own GAM.
-In Figure \ref{Devel_parmod} the development of parametric models starting from simple limear models (LM) and ranging to GAMLSS is illustrated.\\
-These flexible parameteric methods provide well fitting models when applied on precipitation data in a rather smooth/homogeneous setting, e.g.
+To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
+In Figure \ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.\\
+These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth/homogeneous setting, e.g.
 areas where no abrupt changes in meteorological/climatological events appear.
-However, for example in complex terrain/mountaineous regions various factors can change rapidly and their effects are difficult to be modelled/represented by a smooth parameteric model.
+However, for example in complex terrain/mountainous regions various factors can change rapidly and their effects are difficult to be modeled/represented by a smooth parametric model.
 In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models.
@@
 However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well.
-In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothened in case of smooth effects while abrupt changes are stil captured by the tree structure of the model.\\
-For these reasons, regression trees and particularly random forests provide valueable features which can contribute to improving models for probabilistic forecasting.
+In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothened in case of smooth effects while abrupt changes are still captured by the tree structure of the model.\\
+For these reasons, regression trees and particularly random forests provide valuable features which can contribute to improving models for probabilistic forecasting.
 But, up to now, no general distributional framework is available for these methods.
@@
 Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables.
-In that way the correlation among the resulting trees is reduced and therefores also the variance of the model.\\
+In that way the correlation among the resulting trees is reduced and therefore also the variance of the model.\\
 \\
@@
 Hence a distributional forest consisting of distributional trees leads to better results than a simple forest considering only the mean.
-\item In the case of smooth parameter functions without any abrupt changes / non-additive effects a smooth parameteric model such as GAMLSS is clearly the first choice while a stepfunction can be modelled best by applying a tree-structured model.
-Even though for both of these extreme situations a specificly advandageous method is provided, a forest model can be a good compromise and performs well in a wide range of settings including the two mentioned before.
+\item In the case of smooth parameter functions without any abrupt changes / non-additive effects a smooth parametric model such as GAMLSS is clearly the first choice while a stepfunction can be modeled best by applying a tree-structured model.
+Even though for both of these extreme situations a specifically advantageous method is provided, a forest model can be a good compromise and performs well in a wide range of settings including the two mentioned before.
 (Therefore, a forest is a recommendable choice particularly when no information about the type of parameter function is provided.)
-\item While specifying interactions and selecting covariates is necessary to achieve reasonable results in parameteric models such as GAMLSS, this is done automatically within trees and forests.
+\item While specifying interactions and selecting covariates is necessary to achieve reasonable results in parametric models such as GAMLSS, this is done automatically within trees and forests.
 For that reason, distributional trees and forests provide a framework where no additional effort is required for the model specification.
@@
 \item bamlss \cite{bamlss2017} also provides an infrastructure for estimating probabilistic distributional regression but within a Bayesian Framework.
-Again each parameter is modelled by (possibly) complex additive terms.
+Again each parameter is modeled by (possibly) complex additive terms.
 \item gamboostLSS \cite{gamboostLSS2017} offers boosting models for fitting  generalized additive models for location, shape and scale
@@
 \section{Probabilistic Forecasting of Precipitation}
-As explained in the introduction distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valueable in probabilistic forecasting but have not yet been applied in this kind of settings.
+As explained in the introduction distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
 To show their strengths in practice, these new methods are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
 In particular, the data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both included.
-The observations were meassured at 95 (?) stations all over Tyrol (and regions close to the frontier).
+The observations were measured at 95 (?) stations all over Tyrol (and regions close to the frontier).
 The considered data is provided by .....
-Next to the variable of interest, the meassured total precipiation within 24 hours of each day, numerical forecasts from the ECMWF are used as covariates.
+Next to the variable of interest, the measured total precipitation within 24 hours of each day (measured from 6 UTC to 6 UTC of the following day), numerical forecasts from the ECMWF are used as covariates.
 These forecasts are ensemble predictions consisting of ...
-individual forecasts based on perturbed initiall conditions.\\
+individual forecasts based on perturbed initial conditions.\\
 In particular, 14 basic forecast variables are considered with up to 12 variations of these variables being included as covariates.
-Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standarad deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble min and ensemble max.
+Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standard deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble min and ensemble max.
 All of these values can be calculated on the whole day (6--30 UTC) or a 6h time window of the day.\\
-Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standarad deviation, noted as ensemble mean of mean/min/max or ensemble std dev of mean/min/max respectively.\\
+Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/min/max or ensemble std dev of mean/min/max respectively.\\
 Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.\\
-The response variable total precicipitation is power transformed by a factor of 1.6.
-This is done due to experience and knowledge about possible distributions of precipitation meassurements.
+The response variable total precipitation is power transformed by a factor of 1.6.
+This is done due to experience and knowledge about possible distributions of precipitation measurements.
 The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
@@
 
-For the gamlss model the covariates were selected manually based on meteorological knowledge and experience.
-Additionally, interactions are specified in advance, once between tp{\_}mean1218 and cape{\_}mean1218 for the location parameter and once between tp{\_}sprd1218 and cape{\_}mean1218 for the locatioin parameter.
-By putting extra effort in the specification of the formulas the gamlss model is provided with additional information and is therefore expected to profit from this advantage.\\
+For the prespecified GAM model the covariates were selected manually based on meteorological knowledge and experience.
+Additionally, interactions are specified in advance, once between tp{\_}mean1218 and cape{\_}mean1218 for the location parameter and once between tp{\_}sprd1218 and cape{\_}mean1218 for the scale parameter.
+By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.\\
+
+\subsubsection{Means of Comparison}
+In order to compare the performance of the chosen methods three different values are calculated for the resulting models: the root of the mean squared error (RMSE), the log-likelihood value and the continuous ranked probability score (CRPS).
+All of them are evaluated out of sample, e.g.
+on a part of the data set that has not been used to build the model.
+Therefore, overfitting would lead to worse results.\\
+\\
+The RMSE represents how close the estimated expected values are to the true observations.
+For a response variable $Y$ and a set of observations $\textbf{y} = \{y_i\}_{i = 1,\ldots,n}$ with the corresponding estimated parameters $\hat{\boldsymbol{\theta}} = \{\hat{\theta}_i\}_{i = 1,\ldots,n}$ the RMSE is defined as follows.
+
+$$
+\text{RMSE}(Y, \boldsymbol{y}, \boldsymbol{\hat{\theta}}) = \sqrt{\frac{1}{n} \sum_{i=1}^n (\mathbb{E}_{\hat{\theta}_i}(Y) - y_i)^2}
+$$
+
+In most cases this value does not measure the distributional fit as only the expected value is considered.
+However, in the case of a censored normal distribution the expected value of the observed variable is a transformation of the expected value of the latent variable which includes the location and the scale parameter.\\
+\\
+A typical way to evaluate the distributional fit is provided by the log-likelihood value.
+For each observation $y$ and the corresponding predicted parameter $\hat{\theta}$ the value of the log-likelihood function $\ell(\hat{theta}; y)$ is calculated and then averaged over all observations in the testing data set.\\
+\\
+The CRPS is another value that measures the accurarcy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+$$
+\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \le z))^2 dz
+$$
+where \(\mathbf{1}\) is the indicator function.
+Again, the value is averaged over all observations in the testing data set.
+As the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
+One big advantage in terms of interpretability is that the resulsting value is on the scale of the observation.
+\\
+\\
+As we are interested in comparing the performance of the distributional forest to those of other models, we choose one model as a reference and calculated the skills score with respect to the CRPS.
+In that way we obtain a measure of how much a method improves the model fit in terms of CRPS values compared to the reference method.
+$$
+\text{SkillsScore(method)} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
+$$
 
@@
 \subsection{Station Axams}
-First of all we focus on one observation station, Axams in Tyrol. This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol.\\
-As for all other stations, daily observations of the month of July of 28 years are available, starting in 1985 and ending in 2012.\\
-
-\subsubsection{Predictions}
-In a first step one model is fit on the first 25 years (1985--2009, both included) and then predictions for the 26th, 27th and 28th year (2010, 2011 and 2012) are made.\\
-To illustrate the results, one day of July is chosen and the estimated distributions for this day of all three prediction years are plotted together with the corresponding obersvations.  
+First of all we focus on one observation station, Axams in Tyrol.
+This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol.
+As for all other stations, daily observations of the month of July of 28 years are available, starting in 1985 and ending in 2012, both years included.\\
+%\subsubsection{Predictions}
+In a first step one model is learned on the first 25 years (1985--2009, both included) and then predictions for the 26th, 27th and 28th year (2010, 2011 and 2012) are made.
+To illustrate the results, one day of July is chosen and the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations in Figure \ref{axams_onestation}.\\
+The estimated distribution parameters for 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting prediction of $3.185$ close to the observed value of $3.001$.
+Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
+\\
 <<echo=FALSE, results=hide>>=
@@
 lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+
 @
@@
      main = paste0("July ", pday, dayending), ylab = "density", 
+     xlab = "total precipitation (power transformed)",
      ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
@@
 @
-\caption{\label{}Predictions at station Axams for one day in July of 2010, 2011 and 2012}
+\caption{\label{axams_oneday}Predictions at station Axams for one day in July of 2010, 2011 and 2012}
 \end{figure}
 
-Analyzing variable importance of the distforest model leads to the following list of the covariates which show the highest values in variable importance.\\
-In the following plot the marginal effect of these variables on the distribution parameters are plotted. (All other covariates are kept at their mean.)
+Analyzing the distforest model in this setting it can be seen that the covariates listed in the following plot show the highest values in variable importance.\\
+As expected, the numerical forecasts variable for total precipitation is the most frequently selected. In particular, three variations of this variable lead the top ten of this list.\\  
 <<variabble_importance, echo=FALSE, eval=TRUE>>=
@@
 %\end{subfigure}
-\caption{\label{varimp} Variable imortance: top 10 list for the forest model}
+\caption{\label{varimp} Variable importance: top 10 list for the forest model}
 \end{figure}
+
+To get another impression of the quality of the predictions made by the tested methods two PIT-histograms are plotted for each of them, one for the in-sample predictions and one for the out-of-sample predictions.\\
 
@@
 
-\begin{figure}[t!]
+%\begin{figure}[t!]
 %\begin{subfigure}{0.5\textwidth}
-<<rain_axams_pit_insample, fig=TRUE, echo=FALSE>>=
-
-# in sample
-set.seed(7)
-par(mfrow = c(2,2))
-pithist(pit_df_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-pithist(pit_g_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-pithist(pit_gb_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-pithist(pit_ml_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-
-@
-\caption{\label{pit_insample} PIT-Histograms for station Axams (in sample)}
-\end{figure}
+%<<rain_axams_pit_insample, fig=TRUE, echo=FALSE>>=
+%
+%# in sample
+%set.seed(7)
+%par(mfrow = c(2,2))
+%pithist(pit_df_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%pithist(pit_g_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%pithist(pit_gb_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%pithist(pit_ml_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%
+%@
+%\caption{\label{pit_insample} PIT-Histograms for station Axams (in sample)}
+%\end{figure}
 
@@
 
-\begin{figure}[t!]
+%\begin{figure}[t!]
 %\begin{subfigure}{0.5\textwidth}
-<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
-# in sample
-set.seed(7)
-par(mfrow = c(2,2))
-set.seed(1)
-qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
-qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
-qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
-qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
-
-@
-\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
-\end{figure}
-
-\begin{figure}[t!]
-%\begin{subfigure}{0.5\textwidth}
-<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE>>=
-
-# out of sample
-set.seed(7)
-par(mfrow = c(2,2))
-set.seed(1)
-qqrplot(pit_df_t, nsim = 1, main = "distforest (out of bag)")
-qqrplot(pit_g_t, nsim = 1, main = "gamlss (out of bag)")
-qqrplot(pit_gb_t, nsim = 1, main = "gamboostLSS (out of bag)")
-qqrplot(pit_ml_t, nsim = 1, main = "EMOS (out of bag)")
-
-@
-\caption{\label{qqr_oosample} QQR-Plot for station Axams (out of sample)}
-\end{figure}
+%<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
+%# in sample
+%set.seed(7)
+%par(mfrow = c(2,2))
+%set.seed(1)
+%qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
+%qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
+%qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
+%qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
+
+%@
+%\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
+%\end{figure}
+
+%\begin{figure}[t!]
+%%\begin{subfigure}{0.5\textwidth}
+%<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE>>=
+%    
+%# out of sample
+%set.seed(7)
+%par(mfrow = c(2,2))
+%set.seed(1)
+%qqrplot(pit_df_t, nsim = 1, main = "distforest (out of bag)")
+%qqrplot(pit_g_t, nsim = 1, main = "gamlss (out of bag)")
+%qqrplot(pit_gb_t, nsim = 1, main = "gamboostLSS (out of bag)")
+%qqrplot(pit_ml_t, nsim = 1, main = "EMOS (out of bag)")
+
+%@
+%\caption{\label{qqr_oosample} QQR-Plot for station Axams (out of sample)}
+%\end{figure}
 
@@
 \subsubsection{Cross-Validation}
-randomly split into 10 (roughly) equally sized data sets, learn on 9, evaluate on the 10th\\
-average over 10 variations\\
-repeat 10 times\\
-compare rmse, ll, crps score skills with reference model EMOS log\\
-
+After analyzing the results of the models on one data set only where visualization is easier we now want to get a more profound impression of the performance of the different methods.
+For that reason a cross validation framework is considered for the station Axams.
+The data set over all 28 years is split randomly into 10 roughly equally sized data sets.
+While 9 of them are used as learning data the 10th is the testing data for which predictions are made and evaluated.
+This is done by calculating the RMSE, the log-likelihood value and the CRPS value for each of the observations in the testing data and then averaging them.
+Each of the 10 parts is selected as testing data once which leads to 10 different settings over which, again, the average values are stored.
+This whole procedure is repeated 10 times and the results are represented by the boxplots in Figure \ref{boxplots_crps_axams}.
+For the CRPS the skills score is the chosen measurement with the EMOS model as the reference model represented by the horizontal line at height 0.\\
 <<echo=FALSE, results=hide>>=
@@
 
-\begin{figure}[t!]
-\begin{subfigure}{0.5\textwidth}
-<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
-boxplot(rain_rmse)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.5\textwidth}
-<<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
-boxplot(rain_ll, ylim = c(-2.8,-1.3))
-#boxplot(rain_ll)
-@
-\end{subfigure}
-\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
-left: RMSE, right: log-likelihood}
-\end{figure}
+%\begin{figure}[t!]
+%\begin{subfigure}{0.5\textwidth}
+%<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
+%boxplot(rain_rmse)
+%@
+%\end{subfigure}
+%\hspace{-1cm}
+%\begin{subfigure}{0.5\textwidth}
+%<<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
+%boxplot(rain_ll, ylim = c(-2.8,-1.3))
+%#boxplot(rain_ll)
+%@
+%\end{subfigure}
+%\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
+%left: RMSE, right: log-likelihood}
+%\end{figure}
 
@@
 @
-\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams \\
+\caption{\label{boxplots_crps_axams}Results 10x10 cross validation at the station Axams \\
 CRPS skills score with reference model EMOS log}
@@
 
+This evaluation shows that in this setting both GAM models and the distforest perform better than the EMOS model.
+While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distforest even reaches an improvement rate of close to 6 percent.\\
+From this result we can conclude that the distributional forest provides a framework where only little or almost no effort is required to build a model which performs even better than other methods that are commonly used in this kind of application.\\
+For the specified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
+To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decision.
+Moreover, by making them is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not match / correspond to the truth.\\
+This risk is avoided in the boosted GAM as all covariates are included and the selection of the influential covariates is done within a boosting framework.
+However, the trade off here is a remarkably higher computational effort.\\
+To build a distributional forest the user does not have to put any additional effort in it and therefore is not required to provide any expert knowledge or additional information about the setting.
+Variable selection is done automatically and interactions are detected by the tree structure of the model.\\
+In this particular setting, a very high number of covariates is available which intensifies / increases the impact of the above mentioned strength of the forest model.\\
+Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in performance.
+
 \newpage
 \subsection{Predictions over all stations}
-
-NAs in gamlss in 13 stations!!!\\
-for example in 13th station in complete{\_}stations\\
-\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
-\text{NA's in the working vector or weights for parameter mu}
-\\
+After having only focused on one observation station we are now going to apply the methods on all stations.
+At each of them the first 25 years are used as learning data set and predictions are made for the last 3 years of available observations.
+For these out-of-sample predictions the skills score with respect to the CRPS and with the EMOS model as reference is calculated for each method at each station.
+In Figure \ref{boxplots_rain} the boxplots of the skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.\\
+%NAs in gamlss in 13 stations!!!\\
+%for example in 13th station in complete{\_}stations\\
+%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
+%\text{NA's in the working vector or weights for parameter mu}
+%\\
 % sum(is.na(crps[,3]))
@@
 s <- 1 - crps[, 2:4]/crps[,6]
+
+## prepare date for map which shows where distforest performed better than gamlss or gamboostLSS
+# based on the crps skills score (reference modell EMOS log)
+
+library("raster") # dem (digital elevation model)
+library("sp")     # gadm www.gadm.org/country
+
+load("plot_map_rain/demo/tirol.gadm.rda")
+load("plot_map_rain/demo/tirol.dem.rda")
+load("plot_map_rain/demo/ehyd.statlist.rda")
+
+# Create SpatialPointsDataFrame from station list
+sp <- SpatialPointsDataFrame(subset(ehyd.statlist[res$complete_stations,],
+                                    select=c(lon,lat)),
+                             data = subset(ehyd.statlist[res$complete_stations,],
+                                           select = -c(lon,lat)),
+                             proj4string = crs(tirol.dem))
+
+library("colorspace")
+crps_ssc_df <- 1 - crps[,2]/crps[,6]
+crps_ssc_gb <- 1 - crps[,4]/crps[,6]
+crps_ssc_g <- 1 - crps[,3]/crps[,6]
+crps_ssc_ml <- 1 - crps[,6]/crps[,6]    # <- rep(0, length(crps_ssc_g))
+crps_ssc <-  1- crps/crps[,6]
+# abs(crps_ssc_df - crps_ssc_g)
+# boxplot(abs(crps_ssc_df - crps_ssc_g))
+# hist(abs(crps_ssc_df - crps_ssc_g))
+
+# set the colour according to whether distforest performed better than gamlss and
+# set the size of the points on the map according to the difference to the forest model
+pal0 <- hcl(c(10, 260, 290, 50), 100, 80)
+pal2 <- hcl(c(10, 260, 290, 50), 100, 65)
+pal5 <- hcl(c(10, 260, 290, 50), 100, 50)
+#pie(rep(1, 3), col = c(pal0[1], pal2[1], pal5[1]))
+
+col_crps_ssc <- numeric(length = length(crps_ssc_df))
+crps_ssc_best <- numeric(length = length(crps_ssc_df))
+diff_crps_ssc <- numeric(length = length(crps_ssc_df))
+transp_col_crps_ssc <- numeric(length = length(crps_ssc_df))
+level_col_crps_ssc <- numeric(length = length(crps_ssc_df))
+
+for(i in 1: length(crps_ssc_df)){
+  col_crps_ssc[i] <- c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"])[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
+  crps_ssc_best[i] <- which.max(crps_ssc[i,])
+  diff_crps_ssc[i] <- crps_ssc[i,crps_ssc_best[i]] - crps_ssc[i,2]
+  if(crps_ssc_best[i] == 2){
+    reference_crps_ssc <- max(c(crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]), na.rm = TRUE)
+    diff_crps_df <- crps_ssc_df[i] - reference_crps_ssc
+    level_col_crps_ssc[i] <- pal0[1]
+    if(diff_crps_df > 0.02) level_col_crps_ssc[i] <- pal2[1]
+    if(diff_crps_df > 0.05) level_col_crps_ssc[i] <- pal5[1]
+  } else {
+    level_col_crps_ssc[i] <- pal0[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
+    if(diff_crps_ssc[i] > 0.02) level_col_crps_ssc[i] <- pal2[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
+    if(diff_crps_ssc[i] > 0.05) level_col_crps_ssc[i] <- pal5[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
+  }
+
+}
+
+# ehyd.statlist[res$complete_stations,][col_crps_ssc == pal["gamlss"],]
+# table(crps_ssc_best)
+# Axams (Station 77) is the 70th complete station
+# boxplot(diff_crps_ssc[diff_crps_ssc>0])
+# abline(h = diff_crps_ssc[70], col = "red")
+
+# define transperancy of points according to the difference to the forest model (for those points where the forest is not the best one) 
+# alpha parameter between 0 and 1
+alpha_crps_ssc <- (diff_crps_ssc*10) + 0.15
+#alpha_crps_ssc <- (diff_crps_ssc*10/0.85) * 0.5 + 0.5
+alpha_crps_ssc[diff_crps_ssc == 0] <- 1
+add.alpha <- function(col, alpha=1){
+  if(missing(col))
+    stop("Please provide a vector of colours.")
+  apply(sapply(col, col2rgb)/255, 2, 
+                     function(x) 
+                       rgb(x[1], x[2], x[3], alpha=alpha))  
+}
+for(i in 1:length(col_crps_ssc)) transp_col_crps_ssc[i] <- add.alpha(col_crps_ssc[i], alpha = alpha_crps_ssc[i])
+
 @
@@
 \caption{\label{boxplots_rain}Results of predictions over all stations: CRPS skills score with reference model EMOS log}
+\end{figure}
+
+On the map in Figure \ref{map} each station is represented by a pointed colored according to the method that performed best in terms of CRPS at this station.
+For those points where the distributional forest is no the best performing method the intensity of the color represents how much better (in terms of CRPS) the method is compared to the distributional forest.
+\begin{figure}[t!]
+<<map, fig=TRUE, echo=FALSE>>=
+
+  raster::image(tirol.dem, col = rev(gray.colors(100)),
+                main="Stations in Tyrol")
+  plot(tirol.gadm, add = TRUE, ylab = "latitude", xlab = "longitude")
+  points(sp, pch = 21, bg = level_col_crps_ssc, col = 1)
+  #points(sp, pch = 21, bg = col_crps_ssc, col = 1)
+  #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
+  legend("left", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("df", "g", "gb", "ml"), cex = 0.8)
+
+@
+\caption{\label{map}Results of predictions over all stations: CRPS skills score (Forest vs. specified GAM vs, boosted GAM)}
 \end{figure}

### 2017-10-25 pkg/model4you/inst/JORS/man_model4you.Rnw
comments Achim
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1665 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 41aa96587e88cdda1f89cc2bd0161efd6d33a9d3

@@
 
-\paragraph{(1) Model-based trees} identify subgroups where within the subgroups
+\paragraph{1. Model-based trees} identify subgroups where within the subgroups
 the model parameters are similar and between groups the model parameters are
@@
 
-\paragraph{(2) Personalised models} use model-based random forests to estimate
+\paragraph{2. Personalised models} use model-based random forests to estimate
 similarity of individuals in terms of model parameters. For each individual a
@@
 the \Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.
-
-% \begin{figure}
-% \centering
-% \include{pmworkflow}
-% \caption{Personalised models workflow: from real world to interpretable
-% visualisation.}
-% \end{figure}
 
@@
 \item[Name:] CRAN
-\item[Persistent identifier:] \url{https://cran.r-project.org/web/packages/model4you/index.html} %TODO upload to CRAN
-\item[Licence:] GPL-2
+\item[Persistent identifier:] \url{https://cran.r-project.org/package=model4you} %TODO upload to CRAN
+\item[Licence:] GPL-2 | GPL-3
 \item[Publisher:] Heidi Seibold 
@@
 \item[Persistent identifier:] \url{https://r-forge.r-project.org/projects/partykit/}
-\item[Licence:] GPL-2
+\item[Licence:] GPL-2 | GPL-3
 \item[Date published:] \textcolor{blue}{dd/mm/yy} %TODO
@@
 \section*{Competing interests}
-
 The authors declare that they have no competing interests.

### 2017-10-27 pkg/model4you/inst/JORS/man_model4you.Rnw
typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1666 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 271e0965ae354d08d23e9ecb527d8f0be3fa30dd

@@
 simple linear model, which is that it does not seem to be fair for students who
-did very well thrughout the semester (at this point we should state that the
+did very well throughout the semester (at this point we should state that the
 students self selected into the two exam groups which might also be the reason
@@
 
-Dependence plots with the group effect (treatment effect) on the y-achsis and
+Dependence plots with the group effect (treatment effect) on the y-axsis and
 the student characteristics on the x-axis are a good way of visualising the
@@
 \section*{Implementation and architecture}
-The R package \emph{model4you} is focussed on ease of use and interpretability.
+The R package \emph{model4you} is focused on ease of use and interpretability.
 Users can take a simple model that they know and understand as basis and simply

### 2017-10-29 pkg/disttree/inst/draft/draft.Rnw
added text and edited plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1668 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 382285f56f532d0e6fa71d57578f3e26a9ac055e

@@
 \Abstract{
-In the field of probabilistic forecasting classical linear models such as ensemble model output statistics (EMOS) are still common and widely used.
+In the field of probabilistic forecasting classical linear models such as ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) provide common and widely used tools.
 This is due to their simplicity and their interpretability.
-However, the restriction to modelling only linear effects requires further extensions which allow for more flexibility.
-Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own GAM.
+However, the restriction to modeling only linear effects requires further extensions which allow for more flexibility.
+Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
 A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models.
@@
 A typical application in probabilistic forecasting in meteorology is the analysis of precipitation data.
-The main two main interests lie in estimating the probability of rain as well as predicting the expected amount of precipitation.
+The two main interests lie in estimating the probability of rain as well as predicting the expected amount of precipitation.
 In this research area various methods have been developed and tested in the last decades.
-Basic linear models such as ensemble model output statistics (EMOS) were introduced in ...
-and are still common and widely used.
-Despite their simplicity and interpretability, the limitation to modelling only linear effects has risen the interest in applying more flexible models.
+Basic linear models such as ensemble model output statistics (EMOS) were introduced more than a decade ago (\cite{gneiting2005calibrated}) and are still common and widely used.
+Despite their simplicity and interpretability, the limitation to modeling only linear effects has risen the interest in applying more flexible models.
 This feature is provided by generalized additive models (GAM) which can also model non-linear effects.
+However, often only the expectation of the underlying distribution is estimated.
 To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
@@
 Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-(The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure \ref{Devel_treeforest}.)\\
+The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure \ref{Devel_treeforest}.\\
 
@@
 In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distribution with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
-Therefore, the corresponding log-likelihood function $\ell_{tp}$ with parameter vector $\theta = c(\mu, \sigma)$ is
+Therefore, the corresponding log-likelihood function $\ell_{tp}$ with parameter vector $\theta = (\mu, \sigma)$ is
 $$
@@
 Moreover, if covariates are available it is desirable to include them into the model as they might provide important information.
-In particular, they can be used to separate the data into more homogeneous subgroups.
+In our application on precipitation data a wide range of covariates are available from numerical esemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions.
+In particular, these covariates can be used to separate the data into more homogeneous subgroups.
 Doing so after fitting a global model to the whole data set and then fitting a local model to each subgroup should improve the model remarkably.
@@
 \newpage
+\newpage
 \section{Probabilistic Forecasting of Precipitation}
@@
 individual forecasts based on perturbed initial conditions.\\
-In particular, 14 basic forecast variables are considered with up to 12 variations of these variables being included as covariates.
-Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standard deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble min and ensemble max.
+In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates.
+Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standard deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble minimum and ensemble maximum.
 All of these values can be calculated on the whole day (6--30 UTC) or a 6h time window of the day.\\
-Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/min/max or ensemble std dev of mean/min/max respectively.\\
-Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.\\
-The response variable total precipitation is power transformed by a factor of 1.6.
+Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std dev of mean/minimal/maximal values respectively.\\
+Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
+The response variable, total precipitation (tp), is power transformed by a factor of 1.6.
 This is done due to experience and knowledge about possible distributions of precipitation measurements.
@@
 %\resizebox{1.4\textwidth}{!}{
-\begin{minipage}{\textwidth}
+%\begin{minipage}{\textwidth}
+\begin{center}
 \begin{tabular}{| l | l   l   l |}
 \hline
- & \textbf{Type} & \textbf{Location ($\mu$)} & \textbf{Scale ($\sigma$)} \\
+ & \textbf{Type} & \textbf{Location ($\boldsymbol{\mu}$)} & \textbf{Scale ($\boldsymbol{\log(\sigma)}$)} \\
 \hline
@@
 \label{model_specification}
-\end{minipage} 
 %}
+\end{center}
 \end{table}
@@
 By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.\\
+In the EMOS model the logarithmic function is chosen as the link function for the scale parameter.
+This choice has been made after testing and comparing three link functions (identity, quadratic and logarithmic function) by selecting the one leading to the best performance of the EMOS model in this application.
 
@@
 All of them are evaluated out of sample, e.g.
-on a part of the data set that has not been used to build the model.
-Therefore, overfitting would lead to worse results.\\
+on a part of the data set that has not been used to build the model.\\
+%Therefore, overfitting would lead to worse results.\\
 \\
@@
 A typical way to evaluate the distributional fit is provided by the log-likelihood value.
-For each observation $y$ and the corresponding predicted parameter $\hat{\theta}$ the value of the log-likelihood function $\ell(\hat{theta}; y)$ is calculated and then averaged over all observations in the testing data set.\\
+For each observation $y$ and the corresponding predicted parameter $\hat{\theta}$ the value of the log-likelihood function $\ell(\hat{\theta}; y)$ is calculated and then averaged over all observations in the testing data set.
+The higher the value the better the predictions of the model are.\\
 \\
-The CRPS is another value that measures the accurarcy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+The CRPS is another value that measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
 $$
@@
 As the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
-One big advantage in terms of interpretability is that the resulsting value is on the scale of the observation.
+One big advantage in terms of interpretability is that the resulting value is on the scale of the observation.
 \\
@@
 \SweaveOpts{eval = TRUE}
-\newpage
 \subsection{Station Axams}
@@
 In a first step one model is learned on the first 25 years (1985--2009, both included) and then predictions for the 26th, 27th and 28th year (2010, 2011 and 2012) are made.
-To illustrate the results, one day of July is chosen and the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations in Figure \ref{axams_onestation}.\\
-The estimated distribution parameters for 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting prediction of $3.185$ close to the observed value of $3.001$.
+To illustrate the results, one day of July is chosen and the density functions of the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations in Figure \ref{axams_onestation}.\\
+The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
 Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
-\\
+
 <<echo=FALSE, results=hide>>=
@@
 lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-
 @
@@
 \begin{figure}[t!]
+\begin{center}
 <<rain_Axams_pred, fig=TRUE, echo=FALSE>>=
@@
 @
-\caption{\label{axams_oneday}Predictions at station Axams for one day in July of 2010, 2011 and 2012}
+\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012}
+\end{center}
 \end{figure}
 
-Analyzing the distforest model in this setting it can be seen that the covariates listed in the following plot show the highest values in variable importance.\\
-As expected, the numerical forecasts variable for total precipitation is the most frequently selected. In particular, three variations of this variable lead the top ten of this list.\\  
+As seen in Table \ref{model_specification} the forest model includes all 82 covariates.
+In a single tree model it can easily be observed which covariates are used as split variables and hence have an impact on the response variable.
+In a forest model though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree.
+To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been chosen.
+The 10 leading covariates of the above described forest model for the station Axams learned on the first 25 year are listed in Figure \ref{varimp}.
+As expected, the numerical forecast for total precipitation is the most frequently selected.
+In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.
 <<variabble_importance, echo=FALSE, eval=TRUE>>=
@@
 \begin{figure}[t!]
-%\begin{subfigure}{0.5\textwidth}
-<<rain_axams_varim_margeffects1_mu, fig=TRUE, echo=FALSE>>=
+\begin{center}
+\begin{subfigure}{0.7\textwidth}
+<<rain_axams_varim, fig=TRUE, echo=FALSE>>=
 par(mar = c(2,10,2,2))
-barplot(sort(vimp_crps, decreasing = TRUE)[1:10], horiz = TRUE, las = 1, cex.names = 0.8, axes = FALSE)
-@
-%\end{subfigure}
+barplot(sort(vimp_crps, decreasing = TRUE)[1:10], horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE, names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = TRUE)[1:10])))
+@
+\end{subfigure}
 %\hspace{-1cm}
@@
 \caption{\label{varimp} Variable importance: top 10 list for the forest model}
+\end{center}
 \end{figure}
 
-To get another impression of the quality of the predictions made by the tested methods two PIT-histograms are plotted for each of them, one for the in-sample predictions and one for the out-of-sample predictions.\\
-
-<<pit_qqr, echo=FALSE, eval=TRUE>>=
+To get another impression of the quality of the predictions made by the tested methods a probability integral transform (PIT) histograms is plotted in Figure \ref{pit_oosample} for out-of-sample predictions of each of the models.
+A perfectly fitting model would lead to a histogram with all bars being of height 1.
+Overall, it can be stated that all four model provide well fitting models while the forest seems to be the one being closest to the ideal case.
+
+<<pit_qqr, echo=FALSE, eval=TRUE, results=hide>>=
 
@@
   ## disttree
+  if(FALSE){
     # in sample
@@
     pit_dt_l[res$learndata[,"robs"]>0, 1] <- pit_dt_l[res$learndata[,"robs"]>0, 2]
-
+  }  
   # out of sample
@@
   ## distforest
+  if(FALSE){
     # in sample
@@
     pit_df_l[res$learndata[,"robs"]>0, 1] <- pit_df_l[res$learndata[,"robs"]>0, 2]
-
+  }
   # out of sample
@@
   ## gamlss
-  # in sample
   g_learndata <- res$learndata
   g_learndata$robs <- Surv(g_learndata$robs, g_learndata$robs>0, type="left")
+  if(FALSE){
+    # in sample
     g_mu_l <- predict(res$g, what = "mu", type = "response")
@@
     pit_g_l[res$learndata[,"robs"]>0, 1] <- pit_g_l[res$learndata[,"robs"]>0, 2]
-
+  }  
   # out of sample
@@
   #gamboostLSS
+  if(FALSE){
+    #in sample
     pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response")
@@
     pit_gb_l[res$learndata[,"robs"]>0, 1] <- pit_gb_l[res$learndata[,"robs"]>0, 2]
-
+  }  
   # out of sample
@@
   #EMOS
+  if(FALSE){
+    # in sample
     ml_mu_l <- predict(res$ml, type = "location")     # returns parameter on response scale
@@
     pit_ml_l[res$learndata[,"robs"]>0, 1] <- pit_ml_l[res$learndata[,"robs"]>0, 2]
-
+  }  
   # out of sample
@@
 \begin{figure}[t!]
-%\begin{subfigure}{0.5\textwidth}
+\begin{center}
+\begin{subfigure}{0.8\textwidth}
 <<rain_axams_pit_oosample, fig=TRUE, echo=FALSE>>=
@@
 par(mfrow = c(2,2))
-pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-
-@
+pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Forest")
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "GAM (prespecified)")
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "GAM (boosted)")
+pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS")
+
+@
+\end{subfigure}
 \caption{\label{pit_oosample} PIT-Histograms for station Axams (out of sample)}
+\end{center}
 \end{figure}
@@
 \SweaveOpts{eval = TRUE}
-
-\newpage
 \subsubsection{Cross-Validation}
-After analyzing the results of the models on one data set only where visualization is easier we now want to get a more profound impression of the performance of the different methods.
+After analyzing the results of the models on one data set only where visualization is easier we now want to get a more profound impression of the performances of the different methods.
 For that reason a cross validation framework is considered for the station Axams.
@@
 Each of the 10 parts is selected as testing data once which leads to 10 different settings over which, again, the average values are stored.
-This whole procedure is repeated 10 times and the results are represented by the boxplots in Figure \ref{boxplots_crps_axams}.
-For the CRPS the skills score is the chosen measurement with the EMOS model as the reference model represented by the horizontal line at height 0.\\
+This whole procedure is repeated 10 times.
+The results regarding the CRPS values are illustrated by the boxplots in Figure \ref{boxplots_crps_axams} where the skills score of the CRPS is the chosen measurement scale with the EMOS model as the reference model represented by the horizontal line at height 0.\\
+The boxplots of the RMSE and the log-likelihood values can be found in the appendix.
 <<echo=FALSE, results=hide>>=
@@
 
-%\begin{figure}[t!]
-%\begin{subfigure}{0.5\textwidth}
-%<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
-%boxplot(rain_rmse)
-%@
-%\end{subfigure}
-%\hspace{-1cm}
-%\begin{subfigure}{0.5\textwidth}
-%<<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
-%boxplot(rain_ll, ylim = c(-2.8,-1.3))
-%#boxplot(rain_ll)
-%@
-%\end{subfigure}
-%\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
-%left: RMSE, right: log-likelihood}
-%\end{figure}
-
 \begin{figure}[t!]
+\begin{center}
+\center
 <<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE>>=
-boxplot(1 - rain_crps / rain_crps[,6]) 
-abline(h = 0, col = 2, lwd = 2)
-@
-\caption{\label{boxplots_crps_axams}Results 10x10 cross validation at the station Axams \\
-CRPS skills score with reference model EMOS log}
+boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)"),
+        ylab = "CRPS skills score") 
+abline(h = 0, col = pal["EMOS"], lwd = 2)
+@
+\center
+\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 10x10 cross validation at the station Axams}
+\end{center}
 \end{figure}
 
-This evaluation shows that in this setting both GAM models and the distforest perform better than the EMOS model.
+The CRPS skills score boxplot shows that in this setting both GAM models and the forest perform better than the EMOS model.
 While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distforest even reaches an improvement rate of close to 6 percent.\\
@@
 For the specified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
-To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decision.
-Moreover, by making them is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not match / correspond to the truth.\\
+To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decisions.
+Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not match / correspond to the truth.\\
 This risk is avoided in the boosted GAM as all covariates are included and the selection of the influential covariates is done within a boosting framework.
@@
 
-\newpage
+%\newpage
 \subsection{Predictions over all stations}
-After having only focused on one observation station we are now going to apply the methods on all stations.
-At each of them the first 25 years are used as learning data set and predictions are made for the last 3 years of available observations.
-For these out-of-sample predictions the skills score with respect to the CRPS and with the EMOS model as reference is calculated for each method at each station.
-In Figure \ref{boxplots_rain} the boxplots of the skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.\\
+Until now the focus has been on one observation station only.
+To show that the presented results do not depend on the choice of this station we are now going to apply the methods on all stations.
+At each of them, the first 25 years are used as learning data set and predictions are made for the last 3 years of available observations.
+For these out-of-sample predictions the CRPS skills score with respect to the EMOS model as reference is calculated for each method at each station.
+In Figure \ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.\\
+Again, the corresponding boxplots of the RMSE and the log-likelihood values can be found in the appendix.
 %NAs in gamlss in 13 stations!!!\\
@@
 s <- 1 - crps[, 2:4]/crps[,6]
+colnames(s) <- c("Forest", "GAM (prespecified)", "GAM (boosted)")
 
 ## prepare date for map which shows where distforest performed better than gamlss or gamboostLSS
-# based on the crps skills score (reference modell EMOS log)
+# based on the crps skills score (reference model EMOS)
 
@@
 for(i in 1:length(col_crps_ssc)) transp_col_crps_ssc[i] <- add.alpha(col_crps_ssc[i], alpha = alpha_crps_ssc[i])
-
-@
-
-\begin{figure}[t!]
-<<rain_pred_skills_score, fig=TRUE, echo=FALSE>>=
-  matplot(t(s), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, xlab = "", ylab = "", xlim = c(0.5, 3.5))
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<rain_all_crps_skills_score, fig=TRUE, echo=FALSE>>=
+  matplot(t(s), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
+          lty = 1, axes = FALSE, 
+          xlab = "", ylab = "CRPS skills score", xlim = c(0.5, 3.5))
   boxplot(s, add = TRUE)
-  abline(h = 0, col = "gold", lwd = 2)
-@
-\caption{\label{boxplots_rain}Results of predictions over all stations: CRPS skills score with reference model EMOS log}
+  abline(h = 0, col = pal["EMOS"], lwd = 2)
+@
+\caption{\label{boxplot_crps_all}CRPS skills score with reference model EMOS for predictions over all stations}
+\end{center}
 \end{figure}
 
-On the map in Figure \ref{map} each station is represented by a pointed colored according to the method that performed best in terms of CRPS at this station.
-For those points where the distributional forest is no the best performing method the intensity of the color represents how much better (in terms of CRPS) the method is compared to the distributional forest.
-\begin{figure}[t!]
+On the map in Figure \ref{map} each observation station is represented by a point colored according to the method that performed best in terms of CRPS at this station.
+For those points where the distributional forest is not the best performing method the intensity of the color represents how much better (in terms of CRPS) the method is compared to the distributional forest.
+The more intense the color is the bigger the difference.
+In case of the forest being the best model, the second best method is chosen as reference.
+
+\begin{figure}[h!]
+\begin{center}
 <<map, fig=TRUE, echo=FALSE>>=
@@
   raster::image(tirol.dem, col = rev(gray.colors(100)),
-                main="Stations in Tyrol")
-  plot(tirol.gadm, add = TRUE, ylab = "latitude", xlab = "longitude")
+                main="Stations in Tyrol", ylab = "latitude", xlab = "longitude")
+  plot(tirol.gadm, add = TRUE)
   points(sp, pch = 21, bg = level_col_crps_ssc, col = 1)
@@
   #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
-  legend("left", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("df", "g", "gb", "ml"), cex = 0.8)
-
-@
-\caption{\label{map}Results of predictions over all stations: CRPS skills score (Forest vs. specified GAM vs, boosted GAM)}
+  legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS"), cex = 1)
+
+@
+\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions (Forest vs. GAM (prespecified) vs. GAM (boosted))}
+\end{center}
 \end{figure}
@@
 %\url{http://CRAN.R-project.org/doc/Rnews/}
+%\\
+%Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
+%``Calibrated probabilistic forecasting using ensemble model output statistics and minimum CRPS estimation''
+%\textit{Monthly Weather Review},
+%\textbf{133}(5), 1098--1118.
+%\doi{https://doi.org/10.1175/MWR2904.1}
 
@@
 \newpage
+\newpage
 \bibliography{ref.bib}
 
+\newpage
+\section*{Appendix}
+
+\begin{figure}[h!]
+\begin{center}
+%\begin{subfigure}{0.5\textwidth}
+<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
+boxplot(rain_rmse[,c(2,3,4)], ylab = "RMSE",
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
+@
+%\end{subfigure}
+\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at the station Axams}
+\end{center}
+\end{figure}
+%\hspace{-1cm}
+\begin{figure}[h!]
+\begin{center}
+%\begin{subfigure}{0.5\textwidth}
+<<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
+boxplot(rain_ll[,c(2,3,4)], ylim = c(-2.8,-1.3), ylab = "log-likelihood",
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
+#boxplot(rain_ll)
+@
+%\end{subfigure}
+%\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
+%left: RMSE, right: log-likelihood}
+\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams}
+\end{center}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<rain_all_rmse, fig=TRUE, echo=FALSE>>=
+  matplot(t(rmse[,2:4]), type = "l", lwd = 1, 
+          col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, 
+          xlab = "", ylab = "RMSE", xlim = c(0.5, 3.5))
+  boxplot(rmse[,2:4], add = TRUE)
+@
+\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations}
+\end{center}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<rain_all_ll, fig=TRUE, echo=FALSE>>=
+  matplot(t(ll[,2:4]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
+          lty = 1, axes = FALSE, 
+          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 3.5), ylim = c(-4,-1))
+  boxplot(ll[,2:4], add = TRUE)
+@
+\caption{\label{boxplot_ll_all}Log-likelihood vallues for predictions over all stations (extracted interval $\[-4,1\]$)}
+\end{center}
+\end{figure}
+
 \end{document}

### 2017-10-29 pkg/disttree/inst/draft/draft.Rnw
added text and edited plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1669 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: aa95e88ffd121635152b7f9876eba6b2ebb1188b

@@
 <<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
-boxplot(rain_ll[,c(2,3,4)], ylim = c(-2.8,-1.3), ylab = "log-likelihood",
+boxplot(rain_ll[,c(2,3,4)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
         names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
@@
 %left: RMSE, right: log-likelihood}
-\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams}
+\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams (extracted interval [-2.8,-1.3])}
 \end{center}
@@
           xlab = "", ylab = "RMSE", xlim = c(0.5, 3.5))
-  boxplot(rmse[,2:4], add = TRUE)
+  boxplot(rmse[,2:4], add = TRUE,
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
 @
@@
           xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 3.5), ylim = c(-4,-1))
-  boxplot(ll[,2:4], add = TRUE)
-@
-\caption{\label{boxplot_ll_all}Log-likelihood vallues for predictions over all stations (extracted interval $\[-4,1\]$)}
+  boxplot(ll[,2:4], add = TRUE,
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
+@
+\caption{\label{boxplot_ll_all}Log-likelihood vallues for predictions over all stations (extracted interval [-4,-1])}
 \end{center}

### 2017-10-30 pkg/disttree/inst/draft/draft.Rnw
added text and edited plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1671 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6e63b6f5b31dba9a5850a5784e714a0a48b629f6

@@
 Moreover, if covariates are available it is desirable to include them into the model as they might provide important information.
-In our application on precipitation data a wide range of covariates are available from numerical esemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions.
+In our application on precipitation data a wide range of covariates are available from numerical ensemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions.
 In particular, these covariates can be used to separate the data into more homogeneous subgroups.
@@
 plot(x = x, y = y1, type = "l", col = "red", 
-     main = paste0("July ", pday, dayending), ylab = "density", 
-     xlab = "total precipitation (power transformed)",
+     main = paste0("July ", pday, dayending), ylab = "Density", 
+     xlab = "Total precipitation (power transformed)",
      ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
@@
 To get another impression of the quality of the predictions made by the tested methods a probability integral transform (PIT) histograms is plotted in Figure \ref{pit_oosample} for out-of-sample predictions of each of the models.
-A perfectly fitting model would lead to a histogram with all bars being of height 1.
+A perfectly fitting model would lead to a histogram with all bars being of height 1 as indicated by the red horizontal line.
 Overall, it can be stated that all four model provide well fitting models while the forest seems to be the one being closest to the ideal case.
@@
   raster::image(tirol.dem, col = rev(gray.colors(100)),
-                main="Stations in Tyrol", ylab = "latitude", xlab = "longitude")
+                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude")
   plot(tirol.gadm, add = TRUE)
@@
 @
-\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions (Forest vs. GAM (prespecified) vs. GAM (boosted))}
+\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions}
 \end{center}
@@
 %\begin{subfigure}{0.5\textwidth}
-<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
-boxplot(rain_rmse[,c(2,3,4)], ylab = "RMSE",
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
+<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE, width=10, height=7>>=
+boxplot(rain_rmse[,c(2,3,4,6)], ylab = "RMSE",
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
 @
@@
 %\begin{subfigure}{0.5\textwidth}
-<<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
-boxplot(rain_ll[,c(2,3,4)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
+<<rain_cross_axams_ll, fig=TRUE, echo=FALSE, width=10, height=7>>=
+boxplot(rain_ll[,c(2,3,4,6)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
 #boxplot(rain_ll)
@@
 \begin{center}
-<<rain_all_rmse, fig=TRUE, echo=FALSE>>=
-  matplot(t(rmse[,2:4]), type = "l", lwd = 1, 
+<<rain_all_rmse, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
+  matplot(t(rmse[,c(2,3,4,6)]), type = "l", lwd = 1, 
           col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, 
-          xlab = "", ylab = "RMSE", xlim = c(0.5, 3.5))
-  boxplot(rmse[,2:4], add = TRUE,
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
+          xlab = "", ylab = "RMSE", xlim = c(0.5, 4.5))
+  boxplot(rmse[,c(2,3,4,6)], add = TRUE,
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
 @
@@
 \begin{center}
-<<rain_all_ll, fig=TRUE, echo=FALSE>>=
-  matplot(t(ll[,2:4]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
+<<rain_all_ll, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
+  matplot(t(ll[,c(2,3,4,6)]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
           lty = 1, axes = FALSE, 
-          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 3.5), ylim = c(-4,-1))
-  boxplot(ll[,2:4], add = TRUE,
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
-@
-\caption{\label{boxplot_ll_all}Log-likelihood vallues for predictions over all stations (extracted interval [-4,-1])}
+          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 4.5), ylim = c(-4,-1))
+  boxplot(ll[,c(2,3,4,6)], add = TRUE,
+        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
+@
+\caption{\label{boxplot_ll_all}Log-likelihood values for predictions over all stations (extracted interval [-4,-1])}
 \end{center}

### 2017-11-02 pkg/model4you/inst/JORS/man_model4you.Rnw
license
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1673 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 47ec1755f698d4658449179abcecd7394f07a3c2
comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1674 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9417b46a63a073560603668aa7189ec475fd79fd

@@
 model. It automatically uses variables not used in the model formula 
-(in the example above \Sexpr{'response ~ treatment'}) as potential subgroup 
+(in the example above 
+%% <TH> bei mir fehlt hier die tilde </TH>
+\Sexpr{'response ~ treatment'}) as potential subgroup 
 defining variables. This can be edited using the \Sexpr{'zformula'} argument.
@@
 @
+%% <TH> hier koennte man ggf. die Parameter einmal herzeigen und
+%% interpretieren </TH>
 
+%% <TH> warum brauchen wir die Abb 2x? Bitte nur einmal richtig ohne die
+%% gg-Freuerei. /TH>
 The model can be visualised by plotting the estimated densities
@@
 even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1}).
+%% <TH> ich wuerde die Achsenbeschriftung ein bisschen besser machen, also
+%% keine Variablennamen sondern lesbare Beschreibungen </TH>
 <<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
@@
 they want subgroup wise or personalised models. The basis for these
-functionalities is provided by the \emph{partykit} package which is a widely
+functionalities is provided by the \emph{partykit} 
+%% <TH> hier fehlt das Zitat (CRAN + MLOSS) </TH>
+package which is a widely
 used R package for trees and forests. The \emph{model4you} package provides
@@
 \item[Persistent identifier:] \url{https://cran.r-project.org/package=model4you} %TODO upload to CRAN
-\item[Licence:] GPL-2 | GPL-3
+\item[Licence:] GPL-2 $\vert$ GPL-3
 \item[Publisher:] Heidi Seibold 
@@
 \item[Persistent identifier:] \url{https://r-forge.r-project.org/projects/partykit/}
-\item[Licence:] GPL-2 | GPL-3
+\item[Licence:] GPL-2 $\vert$ GPL-3
 \item[Date published:] \textcolor{blue}{dd/mm/yy} %TODO
@@
 Heidi Seibold and Torsten Hothorn were financially supported by the Swiss
-National Science Foundation (grant 205321\_163456).
+National Science Foundation (grants 205321\_163456 and IZSEZ0\_177091).
 
@@
 The authors declare that they have no competing interests.
+
+%% <TH> Gib noch einen Link zum SMMR paper rein, das findet man sonst nicht
+%% </TH>
 

### 2017-11-04 pkg/model4you/inst/JORS/man_model4you.Rnw
working on comments from Torsten
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1675 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 273888732638b6079c9c85f2c71efc5062aef546

--- 2017-10-27 pkg/model4you/inst/JORS/man_model4you.Rnw
@@
 
+\usepackage{booktabs}
+
+%% Nice inline code
+\makeatletter
+\newcommand*{\textalltt}{}
+\DeclareRobustCommand*{\textalltt}{%
+  \begingroup
+    \let\do\@makeother
+    \dospecials
+    \catcode`\\=\z@
+    \catcode`\{=\@ne
+    \catcode`\}=\tw@
+    \verbatim@font\@noligs
+    \@vobeyspaces
+    \frenchspacing
+    \@textalltt
+}
+\newcommand*{\@textalltt}[1]{%
+    #1%
+  \endgroup
+}
+\makeatother
+
 %% Set the header information
@@
 
-<<setup, echo=FALSE>>=
+<<setup, echo=FALSE, message=FALSE>>=
 library("knitr")
@@
 
+# knit_hooks$set(inline = function(x) {
+#   if (is.numeric(x)) return(knitr:::format_sci(x, 'latex'))
+#   highr:::hi_latex(x)
+# })
+
+library("dplyr")
+library("stringr")
+
 knit_hooks$set(inline = function(x) { 
   if (is.numeric(x)) return(knitr:::format_sci(x, 'latex')) 
-  highr:::hi_latex(x) 
+  highr:::hi_latex(x) %>%  str_replace("hlopt", "textalltt")
   }) 
@@
 @
-
+The estimtates and confidence intervals of this model can be computed via
+<<>>=
+cbind(estimate = coef(bmod_math), confint(bmod_math))
+@
+% The estimtates and confidence intervals of this model are shown in 
+% Table~\ref{tab:bmod}.
+% <<math_bmod_coef, results="asis", echo=FALSE>>=
+% library("Hmisc")
+% cf <- round(cbind(estimate = coef(bmod_math), confint(bmod_math)), 2)
+% colnams <- dimnames(cf)[[2]]
+% colnams <- gsub("\\%", "\\\\%", colnams)
+% colnams[1] <- "$\\theta$"
+% 
+% latex(cf,
+%       booktabs = TRUE,
+%       rowlabel = "", file = "", 
+%       colheads = colnams,
+%       label = "tab:bmod", caption = "Estimate and confidence intervals for the base model."
+%       )
+% @
+%% <TH> warum brauchen wir die Abb 2x? Bitte nur einmal richtig ohne die
+%% gg-Freuerei. /TH>
 The model can be visualised by plotting the estimated densities
-(see Figure~\ref{fig:math_bmod_vis} (a)):
-<<eval=FALSE>>=  
+(see Figure~\ref{fig:math_bmod_vis}):
+<<math_bmod_vis, fig.height=2.5, fig.width=5, out.width="0.7\\textwidth", fig.cap = "Density estimates of base model for the Mathematics Exam data.">>=  
 lm_plot(bmod_math)
 @
-
-As the plot is a \emph{ggplot2} graphic it can be maniulated with the usual
-\emph{ggplot2} layers, e.g.\ with
-<<eval=FALSE>>=
-lm_plot(bmod_math) + labs(y = "Estimated Density", x = "% Correct")
-@
-or equivalently with 
-<<eval=FALSE>>=
-myplot <- function(model, ...) {
-  lm_plot(model, ...) + 
-    labs(y = "Estimated Density", x = "% Correct")
-}
-myplot(bmod_math)
-@
-The result is shown in Figure~\ref{fig:math_bmod_vis} (b).
-<<math_bmod_vis, echo=FALSE, fig.cap = "Base model for Mathematics Exam data.", fig.subcap=c("Figure made with \\texttt{lm\\_plot()}.", "Figure made with manipulated \\texttt{lm\\_plot()}."), out.width="0.45\\textwidth", fig.height=3, fig.width=5>>=
-lm_plot(bmod_math)
-myplot <- function(model, ...) {
-  lm_plot(model, ...) + 
-    labs(y = "Estimated Density", x = "% Correct")
-}
-myplot(bmod_math)
-@
+Both the estimates and confidence intervals and the density curves suggest
+that there is almost no difference between the two groups. But does this
+really hold for all types of students?
+% As the plot is a \emph{ggplot2} graphic it can be maniulated with the usual
+% \emph{ggplot2} layers, e.g.\ with
+% <<eval=FALSE>>=
+% lm_plot(bmod_math) + labs(y = "Estimated Density", x = "% Correct")
+% @
+% or equivalently with 
+% <<eval=FALSE>>=
+% myplot <- function(model, ...) {
+%   lm_plot(model, ...) + 
+%     labs(y = "Estimated Density", x = "% Correct")
+% }
+% myplot(bmod_math)
+% @
+% The result is shown in Figure~\ref{fig:math_bmod_vis} (b).
+% <<math_bmod_vis, echo=FALSE, fig.cap = "Base model for Mathematics Exam data.", fig.subcap=c("Figure made with \\texttt{lm\\_plot()}.", "Figure made with manipulated \\texttt{lm\\_plot()}."), out.width="0.45\\textwidth", fig.height=3, fig.width=5>>=
+% lm_plot(bmod_math)
+% myplot <- function(model, ...) {
+%   lm_plot(model, ...) + 
+%     labs(y = "Estimated Density", x = "% Correct")
+% }
+% myplot(bmod_math)
+% @
 
@@
 code:
-<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for math exam.">>=  
+<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for the Mathematics Exam datam.">>=  
 tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2))
-plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = myplot))
+plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lm_plot))
 @
@@
   geom_smooth(fill = NA, method = "loess") +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  ylab("individual estmated group effect")
 
@@
   geom_quasirandom(alpha = 0.5, size = 1) +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  ylab("individual estmated group effect")
 @
@@
   geom_quasirandom(alpha = 0.5) +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  ylab("individual estmated group effect")
 
@@
   geom_quasirandom(alpha = 0.5) +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  ylab("individual estmated group effect")
 @
@@
 they want subgroup wise or personalised models. The basis for these
-functionalities is provided by the \emph{partykit} package which is a widely
-used R package for trees and forests. The \emph{model4you} package provides
+functionalities is provided by the \emph{partykit} 
+package which is a widely
+used R package for trees and forests \citep{hothorn_partykit_2015, hothorn_partykit_2017}. 
+The \emph{model4you} package provides
 wrappers for the well implemented and tested functions
@@
 % \section*{References}
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 \bibliography{ref_model4you}

### 2017-11-04 pkg/model4you/inst/JORS/man_model4you.Rnw
ylab and clean up Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1676 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 99142485cdebba65dca043f41940f6e2ed11a4d4

@@
 @
-% The estimtates and confidence intervals of this model are shown in 
-% Table~\ref{tab:bmod}.
-% <<math_bmod_coef, results="asis", echo=FALSE>>=
-% library("Hmisc")
-% cf <- round(cbind(estimate = coef(bmod_math), confint(bmod_math)), 2)
-% colnams <- dimnames(cf)[[2]]
-% colnams <- gsub("\\%", "\\\\%", colnams)
-% colnams[1] <- "$\\theta$"
-% 
-% latex(cf,
-%       booktabs = TRUE,
-%       rowlabel = "", file = "", 
-%       colheads = colnams,
-%       label = "tab:bmod", caption = "Estimate and confidence intervals for the base model."
-%       )
-% @
-%% <TH> warum brauchen wir die Abb 2x? Bitte nur einmal richtig ohne die
-%% gg-Freuerei. /TH>
 The model can be visualised by plotting the estimated densities
@@
 really hold for all types of students?
-% As the plot is a \emph{ggplot2} graphic it can be maniulated with the usual
-% \emph{ggplot2} layers, e.g.\ with
-% <<eval=FALSE>>=
-% lm_plot(bmod_math) + labs(y = "Estimated Density", x = "% Correct")
-% @
-% or equivalently with 
-% <<eval=FALSE>>=
-% myplot <- function(model, ...) {
-%   lm_plot(model, ...) + 
-%     labs(y = "Estimated Density", x = "% Correct")
-% }
-% myplot(bmod_math)
-% @
-% The result is shown in Figure~\ref{fig:math_bmod_vis} (b).
-% <<math_bmod_vis, echo=FALSE, fig.cap = "Base model for Mathematics Exam data.", fig.subcap=c("Figure made with \\texttt{lm\\_plot()}.", "Figure made with manipulated \\texttt{lm\\_plot()}."), out.width="0.45\\textwidth", fig.height=3, fig.width=5>>=
-% lm_plot(bmod_math)
-% myplot <- function(model, ...) {
-%   lm_plot(model, ...) + 
-%     labs(y = "Estimated Density", x = "% Correct")
-% }
-% myplot(bmod_math)
-% @
 
@@
 tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2))
-plot(tr_math, terminal_panel = node_pmterminal(tr_math, plotfun = lm_plot))
+plot(tr_math, terminal_panel = node_pmterminal(tr_math, 
+                                               plotfun = lm_plot))
 @
@@
   theme(legend.position = "none") +
-  ylab("individual estmated group effect")
+  ylab("estmated individual\nexam group effect")
 
@@
   theme(legend.position = "none") +
-  ylab("individual estmated group effect")
+  ylab("estmated individual\nexam group effect")
 @

### 2017-11-04 pkg/disttree/inst/draft/draft.Rnw
edited introduction and methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1677 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8a97252ee7386f3b2504e2505c66c08b6554b3ed

@@
 \Abstract{
-In the field of probabilistic forecasting classical linear models such as ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) provide common and widely used tools.
+In the field of probabilistic forecasting linear models such as ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) provide common and widely used tools.
 This is due to their simplicity and their interpretability.
@@
 Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
-A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models.
+A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models such as regression trees (\cite{loh2011classification}) or random forests (\cite{breiman2001random}).
 However, up to now, this class of models has not yet been integrated in probabilistic forecasting as no general distributional framework is available for these methods.
@@
 In this research area various methods have been developed and tested in the last decades.
-Basic linear models such as ensemble model output statistics (EMOS) were introduced more than a decade ago (\cite{gneiting2005calibrated}) and are still common and widely used.
+(Multiple) Linear models such as ensemble model output statistics (EMOS) were introduced more than a decade ago (\cite{gneiting2005calibrated}) and are still common and widely used.
 Despite their simplicity and interpretability, the limitation to modeling only linear effects has risen the interest in applying more flexible models.
@@
 However, often only the expectation of the underlying distribution is estimated.
-To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
+To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
 In Figure \ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.\\
@@
 However, for example in complex terrain/mountainous regions various factors can change rapidly and their effects are difficult to be modeled/represented by a smooth parametric model.
-In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models.
+In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models such as regression trees (\cite{loh2011classification}) or random forests (\cite{breiman2001random}).
 Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process.
@@
 However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well.
-In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothened in case of smooth effects while abrupt changes are still captured by the tree structure of the model.\\
+In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothed out in case of smooth effects while abrupt changes are still captured by the tree structure of the model.\\
 For these reasons, regression trees and particularly random forests provide valuable features which can contribute to improving models for probabilistic forecasting.
@@
 The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure \ref{Devel_treeforest}.\\
+After explaining the novel methodology in section \ref{Methodology} of this paper and presenting results from a simulation study in section \ref{Simulation} the focus will then be put on an application in the field of probabilistic forecasting in section \ref{ProbabilisticForecasting}.
+The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
+In a first setting only one observation station is taken into consideration and based on the data/information from 25 years probabilistic predictions are made for the 3 successive years.
+A brief outlook on the results of this setting is given by Figure \ref{axams_onestation}.
+For the 24th of July of 2010, 2011 and 2012 the density functions of the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations.
+The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
+Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
+Next to the expected values the predictions provided by these distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
+
+<<echo=FALSE, results=hide>>=
+#### Axams prediction 25 - 3
+if(file.exists("rain_Axams_pred.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
+  load("rain_Axams_pred.rda")
+} else {
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
+  source("rain_axams_pred.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+  res <- rain_axams_pred(seedconst = 7, ntree = 100,
+                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
+                         gamboost_cvr = TRUE,
+                         frac = FALSE)
+
+  save(res, file = "rain_Axams_pred.rda")
+}
+
+# prepare data for plot of estimated density functions
+pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
+pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+cbind(df_exp, res$testdata[pdays,"robs"])
+
+# plot predicted distributions together with observations
+#set.seed(res$call$seedconst)
+set.seed(7)
+x <- c(0.01, sort(runif(500,0.01,8)))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
+
+# point mass (slightly shifted)
+pm1 <- c(0.04, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
+pm2 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
+pm3 <- c(-0.1, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+
+# predictions
+pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+
+#legendheight
+lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
+lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
+lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+@
+
+\begin{figure}[t!]
+\begin{center}
+<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
+plot(x = x, y = y1, type = "l", col = "red", 
+     main = paste0("July ", pday, dayending), ylab = "Density", 
+     xlab = "Total precipitation (power transformed)",
+     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
+     xlim = c(-1.5,8))
+
+lines(x = x, y = y2, type = "l", col = "blue")
+    lines(x = x, y = y3, type = "l", col = "darkgreen")
+    #legend('topright', c("2010", "2011", "2012"), col = c("red", "blue", "darkgreen"), lty = 1, cex = 1)
+
+    # plot point mass
+    lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
+    lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
+    lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
+    points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
+    points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
+    points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
+
+    # plot predictions
+    points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
+    points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
+    points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
+
+    lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
+
+    # add labels
+    text(x = -0.8, y = lh1, labels = "2010", col = "red", cex = 0.8)
+    text(x = -0.8, y = lh2, labels = "2011", col = "blue", cex = 0.8)
+    text(x = -0.8, y = lh3, labels = "2012", col = "darkgreen", cex = 0.8)
+@
+\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.}
+\end{center}
+\end{figure}
 
@@
 \endminipage
-\caption{\label{Devel_parmod}Development in the field of parametric modeling}
+\caption{\label{Devel_parmod}Development in the field of parametric modeling.}
 \end{figure}
@@
 \label{Methodology}
+
+\subsection{Motivation (Review/Problems)}
+As illustrated in Figure \ref{Devel_parmod} a lot of developement has been made in the field of parametric modeling and a wide range of tools can be applied in research areas such as probabilistic forecasting.
+But as mentioned before, further features such as the ability to capture non-additive effects or interactions are desirable in many applications which demands for more flexible models.
+Since tree structured models provide these required features a combination of them with parametric models can offer a framework that is able to deal with a wide range of different modeling situations.
+In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree structered models step by step starting with a simple distributional fit in subsection \ref{distfit}, then building a distributional tree in subsection \ref{disttree} and finally a distributional forest in subsection \ref{distforest}.
+
 \subsection{Distributional Fit}
+\label{distfit}
 Fitting a distributional model to a set of observations without considering any covariates is a well known procedure which can for example be done by applying the maximum likelihood method.
@@
 The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided.
-In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distribution with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
-Therefore, the corresponding log-likelihood function $\ell_{tp}$ with parameter vector $\theta = (\mu, \sigma)$ is
-$$
-\ell_{tp}(\mu, \sigma; Y) = 
+Using the GAMLSS framework (\cite{stasinopoulos2005}) a wide range of distributions can be selected where each distribution parameter can be specified separatly.
+This enables the model to deal with various features that can usually be very challenging in model fitting such as censoring, truncation or heavy tails.\\
+In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
+Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
+$$
+\ell(\mu, \sigma; Y) = 
 \begin{cases}
@@
 $$
-where $\phi$ and $\Phi$ are the probability density function and the distribution function of the standard normal distribution $\mathcal{N}(0,1)$.\\
-With this specification the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$. This is done by the following maximization.
+where $\phi$ and $\Phi$ are the probability density function and the distribution function of the standard normal distribution $\mathcal{N}(0,1)$.
+Next to the censored normal distribution the censored logistic distribution is also often chosen for the modeling of precipitation amounts.
+In \cite{Gebetsberger2017} this choice has been made in order to be able to deal with heavy tails which often appear in precipitation data.
+Another frequent application of probabilistic forecasting in meteorology is the modeling of temperature.
+In this case a normal distribution is often selected.\\
+With the specification of the distribution family and the log-likelihood function the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$.
+This is done by the following maximization.
 $$
@@
 In our application on precipitation data a wide range of covariates are available from numerical ensemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions.
-In particular, these covariates can be used to separate the data into more homogeneous subgroups.
-Doing so after fitting a global model to the whole data set and then fitting a local model to each subgroup should improve the model remarkably.
-This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.
+\\
+A common way of including these covariates is to consider them as regressors in a distributional model such that each parameter $\theta$ depends on covariates $\bold{X} = (X_1,\ldots,X_m)$.
+This could for example be done by a linear function $g$.
+In that way the fitted model would be specified as $D(Y, \theta = g(\bold{X}))$.
+In order to avoid the restriction to linear effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ to specify a more flexible model $D(Y, \theta = f_1(\bold{X}) + \ldots + f_p(\bold{X}))$ applying a GAM structure.
+Another approach to distributional regression is proposed by \cite{Klein2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
+These include non-linear effects, spatial effects, random coefficients and interaction surfaces which extends the flexibility compared to parametric models such as GAMLSS modeling only smooth effects.\\
+However, the restriction to additive models remains and therefore a different approach of how to include covariates that also allows for non-additive effects and interactions is now considered.
+In particular, the covariates can be used to separate the data into more homogeneous subgroups.
+Doing so after fitting a global model to the whole data set (without considering any covariates) and then fitting a local model to each subgroup should improve the model remarkably.
+This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.\\
 
 \subsection{Distributional Tree}
-After having fit a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is now to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ to decide if the data set should be split into subgroups.
+\label{disttree}
+After having fit a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is now to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors) to decide if the data set should be split into subgroups.
 In particular, the information provided by the fitted model are the scores
@@
 \subsection{Distributional Forest}
+\label{distforest}
 A distributional forest consists of an ensemble of distributional trees $\{t_s\}_{s=1,\ldots,T}$.
@@
 Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables.
-In that way the correlation among the resulting trees is reduced and therefore also the variance of the model.\\
+In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\cite{breiman2001random}).\\
 \\
@@
 \section{Simulation Study}
+\label{Simulation}
 Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.
@@
 \end{subfigure}
-\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE)\\
+\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE).\\
 left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
@@
 \section{Probabilistic Forecasting of Precipitation}
+\label{ProbabilisticForecasting}
 As explained in the introduction distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
@@
 \begin{minipage}{\textwidth}
-\begin{tabular}{| l  c  l |}
+\begin{tabular}{ l  c  l }
 \hline
@@
 \end{tabular}
-\caption[Table caption text]{Basic covariates together with the number and the type of variations}
+\caption[Table caption text]{Basic covariates together with the number and the type of variations.}
 \label{covariates}
@@
 \begin{center}
-\begin{tabular}{| l | l   l   l |}
+\begin{tabular}{ l  l   l   l }
 \hline
@@
 \end{tabular}
-\caption[Table caption text]{Table of applied methods together with their type (way of including the covariates) and the included covariates for each distribution parameter}
+\caption[Table caption text]{Table of applied methods together with their type (way of including the covariates) and the included covariates for each distribution parameter.}
 \label{model_specification}
@@
 As for all other stations, daily observations of the month of July of 28 years are available, starting in 1985 and ending in 2012, both years included.\\
-%\subsubsection{Predictions}
-In a first step one model is learned on the first 25 years (1985--2009, both included) and then predictions for the 26th, 27th and 28th year (2010, 2011 and 2012) are made.
+
+%%\subsubsection{Predictions}
+%In a first step one model is learned on the first 25 years (1985--2009, both included) and then predictions for the 26th, 27th and 28th year (2010, 2011 and 2012) are made.
 To illustrate the results, one day of July is chosen and the density functions of the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations in Figure \ref{axams_onestation}.\\
-The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
+%The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
 Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
 
+\SweaveOpts{eval = FALSE}
 <<echo=FALSE, results=hide>>=
@@
 \begin{center}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE>>=
+<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
 plot(x = x, y = y1, type = "l", col = "red", 
@@
 @
-\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012}
+\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.}
 \end{center}
 \end{figure}
+
+\SweaveOpts{eval = TRUE}
 
@@
 \begin{subfigure}{0.7\textwidth}
-<<rain_axams_varim, fig=TRUE, echo=FALSE>>=
+<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
 par(mar = c(2,10,2,2))
-barplot(sort(vimp_crps, decreasing = TRUE)[1:10], horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE, names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = TRUE)[1:10])))
+barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
+        horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE, 
+        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
 @
@@
 %\end{subfigure}
-\caption{\label{varimp} Variable importance: top 10 list for the forest model}
+\caption{\label{varimp} Variable importance: top 10 list for the forest model.}
 \end{center}
@@
 \begin{subfigure}{0.8\textwidth}
-<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE>>=
+<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
 
@@
 \end{subfigure}
-\caption{\label{pit_oosample} PIT-Histograms for station Axams (out of sample)}
+\caption{\label{pit_oosample} PIT-Histograms for station Axams (out of sample).}
 \end{center}
@@
 \center
-<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE>>=
+<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5>>=
 boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
         names = c("Forest", "GAM (prespecified)", "GAM (boosted)"),
-        ylab = "CRPS skills score") 
+        ylab = "CRPS skills score", col = "lightgray") 
 abline(h = 0, col = pal["EMOS"], lwd = 2)
@@
 \center
-\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 10x10 cross validation at the station Axams}
+\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 10x10 cross validation at the station Axams.}
 \end{center}
@@
 <<rain_all_crps_skills_score, fig=TRUE, echo=FALSE>>=
-  matplot(t(s), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
+  matplot(t(s[,]), type = "l", lwd = 2, 
+          col = c(rep(gray(0.5, alpha = 0.3),69),
+                  #gray(0.1, alpha = 0.8), 
+                  "brown",
+                  rep(gray(0.5, alpha = 0.3),15)), 
           lty = 1, axes = FALSE, 
           xlab = "", ylab = "CRPS skills score", xlim = c(0.5, 3.5))
-  boxplot(s, add = TRUE)
+  # Station Axams is the 77th station which is the 70th in the list of complete stations
+  boxplot(s, add = TRUE, col = "transparent")
   abline(h = 0, col = pal["EMOS"], lwd = 2)
-@
-\caption{\label{boxplot_crps_all}CRPS skills score with reference model EMOS for predictions over all stations}
+
+@
+\caption{\label{boxplot_crps_all}CRPS skills score with reference model EMOS for predictions over all stations.}
 \end{center}
@@
 @
-\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions}
+\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions.}
 \end{center}
@@
 %\end{subfigure}
-\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at the station Axams}
+\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at the station Axams.}
 \end{center}
@@
 %left: RMSE, right: log-likelihood}
-\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams (extracted interval [-2.8,-1.3])}
+\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams (extracted interval [-2.8,-1.3]).}
 \end{center}
@@
 @
-\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations}
+\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations.}
 \end{center}
@@
 @
-\caption{\label{boxplot_ll_all}Log-likelihood values for predictions over all stations (extracted interval [-4,-1])}
+\caption{\label{boxplot_ll_all}Log-likelihood values for predictions over all stations (extracted interval [-4,-1]).}
 \end{center}

### 2017-11-08 pkg/disttree/inst/draft/draft.Rnw
edited plots and text in section 'simulation' and restructured section 'probabilistic forcasting'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1686 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bda48e60a0ced43e0b039206cc583a72ab3beca8

@@
 The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure \ref{Devel_treeforest}.\\
-After explaining the novel methodology in section \ref{Methodology} of this paper and presenting results from a simulation study in section \ref{Simulation} the focus will then be put on an application in the field of probabilistic forecasting in section \ref{ProbabilisticForecasting}.
-The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
-In a first setting only one observation station is taken into consideration and based on the data/information from 25 years probabilistic predictions are made for the 3 successive years.
-A brief outlook on the results of this setting is given by Figure \ref{axams_onestation}.
-For the 24th of July of 2010, 2011 and 2012 the density functions of the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations.
-The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
-Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
-Next to the expected values the predictions provided by these distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
-
-<<echo=FALSE, results=hide>>=
-#### Axams prediction 25 - 3
-if(file.exists("rain_Axams_pred.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
-  load("rain_Axams_pred.rda")
-} else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
-  source("rain_axams_pred.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-  res <- rain_axams_pred(seedconst = 7, ntree = 100,
-                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
-                         gamboost_cvr = TRUE,
-                         frac = FALSE)
-
-  save(res, file = "rain_Axams_pred.rda")
-}
-
-# prepare data for plot of estimated density functions
-pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
-
-pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
-pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, res$testdata[pdays,"robs"])
-
-# plot predicted distributions together with observations
-#set.seed(res$call$seedconst)
-set.seed(7)
-x <- c(0.01, sort(runif(500,0.01,8)))
-y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
-y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
-y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
-dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
-
-# point mass (slightly shifted)
-pm1 <- c(0.04, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
-pm2 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
-pm3 <- c(-0.1, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
-
-# predictions
-pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-
-#legendheight
-lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
-lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
-lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-@
-
-\begin{figure}[t!]
-\begin{center}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
-plot(x = x, y = y1, type = "l", col = "red", 
-     main = paste0("July ", pday, dayending), ylab = "Density", 
-     xlab = "Total precipitation (power transformed)",
-     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
-     xlim = c(-1.5,8))
-
-lines(x = x, y = y2, type = "l", col = "blue")
-    lines(x = x, y = y3, type = "l", col = "darkgreen")
-    #legend('topright', c("2010", "2011", "2012"), col = c("red", "blue", "darkgreen"), lty = 1, cex = 1)
-
-    # plot point mass
-    lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
-    lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
-    lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
-    points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
-    points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
-    points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
-
-    # plot predictions
-    points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
-    points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
-    points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
-
-    lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
-    lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
-    lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
-
-    # add labels
-    text(x = -0.8, y = lh1, labels = "2010", col = "red", cex = 0.8)
-    text(x = -0.8, y = lh2, labels = "2011", col = "blue", cex = 0.8)
-    text(x = -0.8, y = lh3, labels = "2012", col = "darkgreen", cex = 0.8)
-@
-\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.}
-\end{center}
-\end{figure}
 
@@
 
+After explaining the novel methodology in Section \ref{Methodology} of this paper and presenting results from a simulation study in Section \ref{Simulation} the focus will then be put on an application in the field of probabilistic forecasting in Section \ref{ProbabilisticForecasting}.
+The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
+In a first setting only one observation station is taken into consideration and based on the data/information from 25 years probabilistic predictions are made for the 3 successive years.
+A brief outlook on the results of this setting is given by Figure \ref{axams_onestation}.
+For the 24th of July of 2010, 2011 and 2012 the density functions of the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations.
+The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
+Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
+Next to the expected values the predictions provided by these distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
+
+<<echo=FALSE, results=hide>>=
+#### Axams prediction 25 - 3
+if(file.exists("rain_Axams_pred.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
+  load("rain_Axams_pred.rda")
+} else {
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
+  source("rain_axams_pred.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+  res <- rain_axams_pred(seedconst = 7, ntree = 100,
+                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
+                         gamboost_cvr = TRUE,
+                         frac = FALSE)
+
+  save(res, file = "rain_Axams_pred.rda")
+}
+
+# prepare data for plot of estimated density functions
+pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
+pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+cbind(df_exp, res$testdata[pdays,"robs"])
+
+# plot predicted distributions together with observations
+#set.seed(res$call$seedconst)
+set.seed(7)
+x <- c(0.01, sort(runif(500,0.01,8)))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
+
+# point mass (slightly shifted)
+pm1 <- c(0.04, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
+pm2 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
+pm3 <- c(-0.1, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+
+# predictions
+pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+
+#legendheight
+lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
+lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
+lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+@
+
+\begin{figure}[t!]
+\begin{center}
+<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
+plot(x = x, y = y1, type = "l", col = "red", 
+     main = paste0("July ", pday, dayending), ylab = "Density", 
+     xlab = "Total precipitation (power transformed)",
+     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
+     xlim = c(-1.5,8))
+
+lines(x = x, y = y2, type = "l", col = "blue")
+    lines(x = x, y = y3, type = "l", col = "darkgreen")
+    #legend('topright', c("2010", "2011", "2012"), col = c("red", "blue", "darkgreen"), lty = 1, cex = 1)
+
+    # plot point mass
+    lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
+    lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
+    lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
+    points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
+    points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
+    points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
+
+    # plot predictions
+    points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
+    points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
+    points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
+
+    lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
+    lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
+
+    # add labels
+    text(x = -0.8, y = lh1, labels = "2010", col = "red", cex = 0.8)
+    text(x = -0.8, y = lh2, labels = "2011", col = "blue", cex = 0.8)
+    text(x = -0.8, y = lh3, labels = "2012", col = "darkgreen", cex = 0.8)
+@
+\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.}
+\end{center}
+\end{figure}
+
 \newpage
@@
 \subsection{Motivation (Review/Problems)}
-As illustrated in Figure \ref{Devel_parmod} a lot of developement has been made in the field of parametric modeling and a wide range of tools can be applied in research areas such as probabilistic forecasting.
+As illustrated in Figure \ref{Devel_parmod} a lot of development has been made in the field of parametric modeling and a wide range of tools can be applied in research areas such as probabilistic forecasting.
 But as mentioned before, further features such as the ability to capture non-additive effects or interactions are desirable in many applications which demands for more flexible models.
 Since tree structured models provide these required features a combination of them with parametric models can offer a framework that is able to deal with a wide range of different modeling situations.
-In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree structered models step by step starting with a simple distributional fit in subsection \ref{distfit}, then building a distributional tree in subsection \ref{disttree} and finally a distributional forest in subsection \ref{distforest}.
+In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree structured models step by step starting with a simple distributional fit in Subsection \ref{distfit}, then building a distributional tree in Subsection \ref{disttree} and finally a distributional forest in Subsection \ref{distforest}.
 
@@
 The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided.
-Using the GAMLSS framework (\cite{stasinopoulos2005}) a wide range of distributions can be selected where each distribution parameter can be specified separatly.
+Using the GAMLSS framework (\cite{stasinopoulos2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
 This enables the model to deal with various features that can usually be very challenging in model fitting such as censoring, truncation or heavy tails.\\
@@
 
-The following methods are applied and compared to distributional trees and forests:
-\begin{itemize}
-\item gamlss \cite{stasinopoulos2005} fits a separate generalized additive model to each of the distribution models.
-In that way it specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.
-\item bamlss \cite{bamlss2017} also provides an infrastructure for estimating probabilistic distributional regression but within a Bayesian Framework.
-Again each parameter is modeled by (possibly) complex additive terms.
-\item gamboostLSS \cite{gamboostLSS2017} offers boosting models for fitting  generalized additive models for location, shape and scale
-\item randomForest \cite{randomForest2002} combines an ensemble of classification and regression trees applying bagging and reducing the correlation among trees through a randomized selection of possible split variables within each node.
-\item cforest \cite{hothorn2015partykit} builds a forest in a similar way but consisting of ctrees applying the ctree algorithm mentioned in section \ref{Methodology}.
-In this simulation study no transformation function is included.
-\end{itemize}
-
-The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and ten covariates $X_1, \ldots, X_{10}$ which are all independent und identically distributed following a uniform distribution.
-$$
-X_i \sim \textit{Unif}([-1,1]) \qquad \forall i \in \{1,\ldots,10\}
+To prove these statements distributional trees and forests are evaluated on generated data sets.
+Their results are then compared to those of a GAMLSS (\cite{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution models and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
+The generated data sets consist of the values of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
+$$
+X_1, X_2, X_3 \sim \textit{Unif}([-1,1])
+$$
+$$
+X_4, X_5, X_6 \sim \textit{Binom}(1, 0.5)
 $$
@@
 \\
-The location parameter $\mu$ depends on $X_1$ and $\kappa$ and the scale parameter $\sigma$ depends on $X_2$ and $\kappa$. The other covariates $X_3, \ldots, X_{10}$ are used as noise variables.\\
+The location parameter $\mu$ depends on $X_1, X_4$ and $\kappa$ and the scale parameter $\sigma$ depends on $X_1, X_4$ and $\kappa$.
+The other covariates $X_2, X_3, X_5$ and $X_6$ are used as noise variables.\\
 The latent variable $Y^{\ast}$ is generated based on the following setting.
@@
 $$ 
-Y^{\ast} \sim \mathcal{N} (f(X_1, X_2, \kappa))
+Y^{\ast} \sim \mathcal{N} (f(X_1, X_4, \kappa))
 $$
@@
 $$
-\mu(X_1,\kappa) = \mu_{\text{base}} + 8 \cdot \exp{(-(3 \cdot X_1-1)^{(2\cdot \kappa)})}
-$$
-
-$$
-\sigma(X_2,\kappa) = 2 + 2 \cdot (1-\Lambda(\kappa^{1.8} \cdot 5 \cdot (X_2+0.5)))
-$$
-where $\Lambda$ is the cumulative distribution function of the logistic distribution
-$$
-\Lambda(\omega) = \frac{1}{1 + \exp(-\omega)}
-$$
+\mu(X_1, X_4, \kappa) = 0.2 + 7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \mathbf{1}(X_1 < 0.4) + 
+                              7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2} \cdot \mathbf{1}(X_1 \geq 0.4) + 
+                              3 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot X_4
+$$
+
+$$
+\sigma(X_2,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.7)
+$$
+%where $\Lambda$ is the cumulative distribution function of the logistic distribution
+%$$
+%\Lambda(\omega) = \frac{1}{1 + \exp(-\omega)}
+%$$
 The response variable $Y$ is defined as 
@@
     Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
-    0,       & \text{if } else
+    0,       & else
 \end{cases}
@@
 
-To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied to one generated data set only where $\kappa$ is set to 1 (top plots) and then to a second data set where $\kappa$ is set to 10 (bottom plots).\\
+To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to 1 (top plots in Figure \ref{}) and then to a second data set where $\kappa$ is set to 10 (bottom plots in Figure \ref{}).\\
 In the left plots the location parameter function $\mu(X_1, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $(x_2, x_3, \ldots, x_{10}) = (0,0, \ldots, 0)$.
@@
 } else {
-  oneset1 <- sim_oneset(kappa = 7, nobs = 400,
-                       seedconst = 7, ntree = 100,
+
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+
+  kappa <- 1
+  fun <- function(x){
+    mu <- 0.2 + 
+      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
+      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 10/2) * as.numeric(x[,1]>= 0.4) #+
+    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) +
+    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
+    sigma <- 0.5 + 
+      #mu/5 + 
+      2 * (x[,1] < -0.4) + 
+      3 * x[,4] #+
+    #1*(x[,1] < -0.7) 
+    #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
+    #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
+    #5 * abs(x[,1]) * x[,4]
+    par <- cbind(mu, sigma)
+    return(par)
+  }
+
+  oneset1 <- sim_oneset(parfun = fun, nobs = 500, seedconst = 74, ntree = 100,
                         formula = y~x1+x2+x3+x4+x5+x6, 
-                       tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95, 
-                       forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
-                       forest_mtry = 3,
-                       type.tree = "ctree",
+                        #formula = y~x1+x2+x4+x5,
+                        #formula = y~x1+x2+x3,
+                        tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
+                        forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
+                        forest_mtry = 3, type.tree = "ctree", 
                         censNO = TRUE, 
-                       fix.mu = FALSE,
-                       fix.sigma = FALSE,
-                       mu.sigma.interaction = FALSE,
                         pred_fix = TRUE,
-                       pred_fix_x1 = -0.5,
-                       pred_fix_x4 = 0,
+                        pred_fix_x1 = 0.9,
+                        pred_fix_x2 = 0.9,
+                        pred_fix_x4 = 1,
                         gamboost_cvr = FALSE,
@@
                         eval_gamlss = TRUE,
-                       eval_gamboostLSS = FALSE,
-                       eval_randomForest = FALSE,
-                       eval_cforest = FALSE,
-                       mubase = 3,
-                       sigmabase = 3)
+                        nrunif = 3,
+                        nrcovform = 6)
+
   save(oneset1, file = "oneset1.rda")
@@
 
-if(file.exists("oneset10.rda")){
+if(file.exists("oneset10.rda")){Hans-Brenner-Platz
   load("oneset10.rda")
 } else {
-  oneset10 <- sim_oneset(kappa = 10, nobs = 400,
-                       seedconst = 7, ntree = 100,
+
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+
+  kappa <- 10
+  fun <- function(x){
+    mu <- 0.2 + 
+      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
+      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 10/2) * as.numeric(x[,1]>= 0.4) #+
+    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) +
+    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
+    sigma <- 0.5 + 
+      #mu/5 + 
+      2 * (x[,1] < -0.4) + 
+      3 * x[,4] #+
+    #1*(x[,1] < -0.7) 
+    #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
+    #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
+    #5 * abs(x[,1]) * x[,4]
+    par <- cbind(mu, sigma)
+    return(par)
+  }
+
+  oneset10 <- sim_oneset(parfun = fun, nobs = 500, seedconst = 74, ntree = 100,
                         formula = y~x1+x2+x3+x4+x5+x6, 
-                       tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95, 
-                       forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
-                       forest_mtry = 3,
-                       type.tree = "ctree",
+                        #formula = y~x1+x2+x4+x5,
+                        #formula = y~x1+x2+x3,
+                        tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
+                        forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
+                        forest_mtry = 3, type.tree = "ctree", 
                         censNO = TRUE, 
-                       fix.mu = FALSE,
-                       fix.sigma = FALSE,
-                       mu.sigma.interaction = FALSE,
                         pred_fix = TRUE,
-                       pred_fix_x1 = -0.5,
-                       pred_fix_x4 = 0,
+                        pred_fix_x1 = 0.9,
+                        pred_fix_x2 = 0.9,
+                        pred_fix_x4 = 1,
                         gamboost_cvr = FALSE,
@@
                         eval_gamlss = TRUE,
-                       eval_gamboostLSS = FALSE,
-                       eval_randomForest = FALSE,
-                       eval_cforest = FALSE,
-                       mubase = 3,
-                       sigmabase = 3)
+                        nrunif = 3,
+                        nrcovform = 6)
+
   save(oneset10, file = "oneset10.rda")
@@
 <<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset1, 
+plot_oneset(oneset1, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
               compare_mu = TRUE, 
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
-            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
-            )
+              add_dt = TRUE,
+              add_df = TRUE,
+              add_g = TRUE,
+              nomodel = FALSE)
 @
@@
 <<sim_oneset1_sigma, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset1, 
+  plot_oneset(oneset1, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
+              compare_mu = FALSE,
               compare_sigma_line = TRUE, 
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
-            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
-            )
+              add_dt = TRUE,
+              add_df = TRUE,
+              add_g = TRUE,
+              nomodel = TRUE)
 @
@@
 <<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset10,
+plot_oneset(oneset10, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
               compare_mu = TRUE, 
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
-            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
-            )
+              add_dt = TRUE,
+              add_df = TRUE,
+              add_g = TRUE,
+              nomodel = FALSE)
 @
@@
 <<sim_oneset10_sigma, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset10, 
+  plot_oneset(oneset10, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
+              compare_mu = FALSE,
               compare_sigma_line = TRUE, 
-            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
-            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
-            )
+              add_dt = TRUE,
+              add_df = TRUE,
+              add_g = TRUE,
+              nomodel = TRUE)
 @
@@
 
-Now, to get a more profound impression the methods are evaluated on various data sets.
+In the case of smooth effects of the coefficients on the parameter function the GAMLSS is closest to the true parameter while the steps of the tree are cleary too rough.
+By combining various trees the forest can smooth out these steps, however, not as well as the GAMLSS.\\
+Looking at the other extreme situation in which the parameter functions include abrupt changes the tree captures them very well wich also leads to a good fit of the forest model.
+In contrast to the two tree structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter function.\\
+\\
+
+Now, to get a more profound impression, the methods are evaluated on various data sets.
 The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step ...
-data sets are generated.
-The methods are then compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by the out-of-sample log-likelihood of the fitted models, averaged within each step.
-(In that way it can be investigated how the results change for a changing value of $\kappa$.)
+data sets are generated.\\
+\\
+%\subsubsection{Means of Comparison}
+In order to compare the performance of the chosen methods over various data sets, the continuous ranked probability score (CRPS) is calculated for each model on each data set.
+
+%three different values are calculated for the resulting models: the root of the mean squared error (RMSE), the log-likelihood value and the continuous ranked probability score (CRPS).
+All of them are evaluated out of sample, e.g.
+on a part of the data set that has not been used to build the model.\\
+%Therefore, overfitting would lead to worse results.\\
+%\\
+%The RMSE represents how close the estimated expected values are to the true observations.
+For a response variable $Y$ and a set of observations $\textbf{y} = \{y_i\}_{i = 1,\ldots,n}$ with the corresponding estimated parameters $\hat{\boldsymbol{\theta}} = \{\hat{\theta}_i\}_{i = 1,\ldots,n}$ the RMSE is defined as follows.
+
+%$$
+%\text{RMSE}(Y, \boldsymbol{y}, \boldsymbol{\hat{\theta}}) = \sqrt{\frac{1}{n} \sum_{i=1}^n (\mathbb{E}_{\hat{\theta}_i}(Y) - y_i)^2}
+%$$
+
+%In most cases this value does not measure the distributional fit as only the expected value is considered.
+However, in the case of a censored normal distribution the expected value of the observed variable is a transformation of the expected value of the latent variable which includes the location and the scale parameter.\\
+%\\
+%A typical way to evaluate the distributional fit is provided by the log-likelihood value.
+For each observation $y$ and the corresponding predicted parameter $\hat{\theta}$ the value of the log-likelihood function $\ell(\hat{\theta}; y)$ is calculated and then averaged over all observations in the testing data set.
+The higher the value the better the predictions of the model are.\\
+%\\
+%The CRPS is another value that measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+$$
+\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
+$$
+where \(\mathbf{1}\) is the indicator function. \\
+In this simulation study the CRPS is evaluated out of sample, e.g.
+on a data set that has not been used to build the model but has been generated by the same process and with the same settings as the original data set.
+In the following this second data set on which the CRPS is evaluated is called testing data set while the original data set on which the models are built on is called learning data set.
+The CRPS value is first calculated for each observation in the testing data set and then the average of these values is taken.
+As the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
+In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observation.
+\\
+
+%The methods are then compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by %the out-of-sample log-likelihood of the fitted models, averaged within each step.
+%(In that way it can be investigated how the results change for a changing value of $\kappa$.)
+
+\SweaveOpts{eval = TRUE}
 
@@
 
+%\begin{figure}[t!]
+%\begin{subfigure}{0.6\textwidth}
+%<<sim_rmse_exp, fig=TRUE, echo=FALSE>>=
+%plot_rmse(simres, type = "exp")
+%@
+%\end{subfigure}
+%\hspace{-1cm}
+%\begin{subfigure}{0.6\textwidth}
+%<<sim_rmse_par, fig=TRUE, echo=FALSE>>=
+%plot_rmse(simres, type = "par")
+%@
+%\end{subfigure}
+%\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE).\\
+%left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference %between the observations and the estimated expected value\\
+%right: continuous lines represent the location parameter, dotted lines the scale parameter}
+%\end{figure}
+
 \begin{figure}[t!]
-\begin{subfigure}{0.6\textwidth}
-<<sim_rmse_exp, fig=TRUE, echo=FALSE>>=
-plot_rmse(simres, type = "exp")
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.6\textwidth}
-<<sim_rmse_par, fig=TRUE, echo=FALSE>>=
-plot_rmse(simres, type = "par")
-@
-\end{subfigure}
-\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE).\\
-left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
-right: continuous lines represent the location parameter, dotted lines the scale parameter}
+\begin{center}
+%\begin{subfigure}{0.6\textwidth}
+<<sim_crps, fig=TRUE, echo=FALSE>>=
+plot_crps(simres)
+@
+\caption{\label{plot_sim_crps}Results of the simulation study: CRPS} %\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest} vs. %\code{bamlss} vs. \code{gamboostLSS}\\
+%right: levelplot of the parameter function for $\kappa=5$}
+\end{center}
 \end{figure}
 
-\begin{figure}[t!]
-\begin{subfigure}{0.6\textwidth}
-<<sim_ll, fig=TRUE, echo=FALSE>>=
-plot_ll(simres)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.6\textwidth}
-<<sim_levelplot, fig=TRUE, echo=FALSE>>=
-levelfun <- function(x) simres$fun(x,5)
-
-ncuts <- 100
-colR <- colorRampPalette(c("blue","green"))                                       
-cols <- colR(ncuts+1)
-griddata <- expand.grid(list(x1 = seq(-1, 1, 0.01), x4 = seq(0, 1, 0.2)))
-griddata$mu <- levelfun(griddata)
-levelplot(mu ~ x1 * x4, data = griddata, xlab = "x1", ylab = "x4", 
-          region = TRUE, cuts = ncuts, col.regions = cols)
-
-@
-\end{subfigure}
-\caption{\label{plot_sim_ll}left: results of the simulation study: log-likelihood\\ %\code{disttree} vs.
+%\end{subfigure}
+%\hspace{-1cm}
+%\begin{subfigure}{0.6\textwidth}
+%<<sim_levelplot, fig=TRUE, echo=FALSE>>=
+%levelfun <- function(x) simres$fun(x,5)
+%
+%ncuts <- 100
+%colR <- colorRampPalette(c("blue","green"))                                       
+%cols <- colR(ncuts+1)
+%griddata <- expand.grid(list(x1 = seq(-1, 1, 0.01), x4 = seq(0, 1, 0.2)))
+%griddata$mu <- levelfun(griddata)
+%levelplot(mu ~ x1 * x4, data = griddata, xlab = "x1", ylab = "x4", 
+%          region = TRUE, cuts = ncuts, col.regions = cols)
+
+%@
+%\end{subfigure}
+%\caption{\label{plot_sim_ll}left: results of the simulation study: log-likelihood\\ %\code{disttree} vs.
 \code{distforest} vs.
@@
 \code{gamboostLSS}\\
-right: levelplot of the parameter function for $\kappa=5$}
-\end{figure}
+%right: levelplot of the parameter function for $\kappa=5$}
+%\end{figure}
+
+Figure \ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study.
+The parametric GAMLSS can easily deal with the smooth effects that appear for low values of $\kappa$ and outperforms the tree and forest models in this situation.
+However, it clearly starts to struggle as $\kappa$ increases and the effects turn into abrupt changes.
+This behaviour has already been detected and discussed above when looking at only single data sets and is now proven by an almost monotone increase of the CRPS.\\
+On the contrary, the imposed/enforced steps of the tree model lead to a high CRPS value in the situation of smooth effects while this value decreases quickly as the slopes in the parameter functions get steeper.
+For high values of $\kappa$ resulting almost in step functions the tree model perfomes the best as stated by the low CRPS value.\\
+Therefore, both extreme situations show to have a preferable method to choose which clearly outperforms the others.
+But overall the forest model stays more or less on the same level of goodness of fit regardless of the type of effects in the parameter functions.
+As this level lies inbetween the ones of the other two methods the forest model provides a good compromise.
+Of course, if the type of effects is known in advance one should choose the corresponding adequate method, but as this is usually not the case applying a distributional forest offers a solution which can deal with a wide variety of situations and in that way secures a good fit of the resulting model.
 
@@
 \label{ProbabilisticForecasting}
+
 As explained in the introduction distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
 To show their strengths in practice, these new methods are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
-In particular, the data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both included.
+
+\subsection{Setting}
+The data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both years included.
 The observations were measured at 95 (?) stations all over Tyrol (and regions close to the frontier).
 The considered data is provided by .....
-Next to the variable of interest, the measured total precipitation within 24 hours of each day (measured from 6 UTC to 6 UTC of the following day), numerical forecasts from the ECMWF are used as covariates.
+Next to the variable of interest which is the measured total precipitation within 24 hours of each day (measured from 6 UTC to 6 UTC of the following day), numerical forecasts (from ...
+) are used as covariates.
 These forecasts are ensemble predictions consisting of ...
-individual forecasts based on perturbed initial conditions.\\
+individual forecasts based on varying/perturbed initial conditions.\\
 In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates.
-Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standard deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble minimum and ensemble maximum.
-All of these values can be calculated on the whole day (6--30 UTC) or a 6h time window of the day.\\
+Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble minimum and ensemble maximum.
+All of these values can be calculated on the sums over a whole day (6--30 UTC) or a 6h time window of the day.\\
 Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std dev of mean/minimal/maximal values respectively.\\
 Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
-The response variable, total precipitation (tp), is power transformed by a factor of 1.6.
+The response variable \texttt{tp} (total precipitation) is power transformed by a factor of 1.6.
 This is done due to experience and knowledge about possible distributions of precipitation measurements.
 The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
-
-%\newpage
-%The ensemble forecast for the total precipitation is represented by 12 covariates: 
-%\begin{itemize}[noitemsep]
-%\item ensemble mean of the sum of the total precipitation from 6--30 UTC
-%\item ensemble standard deviation of the sum of the total precipitation from 6--30 UTC
-%\item ensemble minimum of the sum of the total precipitation from 6--30 UTC
-%\item ensemble maximum of the sum of the total precipitation from 6--30 UTC
-%\item ensemble mean of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC %and from 24--30 UTC
-%\item ensemble standard deviation of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, %from 18--24 UTC and from 24--30 UTC
-%\end{itemize}
-
-%The preciptable water is represented by 6 covariates:
-%\begin{itemize}[noitemsep]
-%\item the ensemble mean of the mean of preciptable water over the time period 6--30 UTC 
-%\item the ensemble mean of the minimum of preciptable water over the time period 6--30 UTC 
-%\item the ensemble mean of the maximum of preciptable water over the time period 6--30 UTC 
-%\item the ensemble standard deviation of the mean of preciptable water over the time period 6--30 UTC 
-%\item the ensemble standard deviation of the minimum of preciptable water over the time period 6--30 UTC 
-%\item the ensemble standard deviation of the maximum of preciptable water over the time period 6--30 UTC 
-%\end{itemize}
+The same transformation has been chosen for the variable \texttt{cape} (convective available potential energy).
 
@@
 \hline
-\textbf{tp}: total precipitation            & 12 variations:  & ensemble mean of sum, \\
-\hspace*{\fill} power transformed (by 1.6) &    & ensemble std dev of sum, \\ 
-\textbf{cape}: convective                  &    & ensemble minimum of sum, \\
-\hspace*{0.4cm} available potential energy &    & ensemble maximum of sum\\
-\hspace*{\fill}  power transformed (by 1.6)&    & \qquad all over 6--30 UTC\\
-\hspace*{\fill}                            &    & ensemble mean of sum over\\
-\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
-\hspace*{\fill}                            &    & ensemble std dev of sum over\\
-\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
+\textbf{tp}: total precipitation            & 12 variations:  & ensemble mean of sums over 24h, \\
+\hspace*{\fill} power transformed (by 1.6) &    & ensemble std dev of sums over 24h, \\ 
+\textbf{cape}: convective                  &    & ensemble minimum of sums over 24h, \\
+\hspace*{0.4cm} available potential energy &    & ensemble maximum of sums over 24h\\
+\hspace*{\fill}  power transformed (by 1.6)&    & \qquad all for 6--30 UTC\\
+\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
+\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
+\hspace*{\fill}                            &    & ensemble std dev of sums over 6h\\
+\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
 \hline
@@
 
-Next to distributional forests other methods / the same methods as in the simulation section are applied to this data set and their results are compared based on RMSE, log-likelihood values and CRPS values.
+\subsection{Models}
+Next to distributional forests other methods are applied to this data set and their results are compared based on RMSE, log-likelihood values and CRPS values.
 The chosen methods and their specifications are listed in Table \ref{model_specification}.
@@
 For the prespecified GAM model the covariates were selected manually based on meteorological knowledge and experience.
-Additionally, interactions are specified in advance, once between tp{\_}mean1218 and cape{\_}mean1218 for the location parameter and once between tp{\_}sprd1218 and cape{\_}mean1218 for the scale parameter.
+Additionally, interactions are specified in advance, once between \texttt{tp{\_}mean1218} and \texttt{cape{\_}mean1218} for the location parameter and once between \texttt{tp{\_}sprd1218} and \texttt{cape{\_}mean1218} for the scale parameter.
 By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.\\
 In the EMOS model the logarithmic function is chosen as the link function for the scale parameter.
-This choice has been made after testing and comparing three link functions (identity, quadratic and logarithmic function) by selecting the one leading to the best performance of the EMOS model in this application.
-
-\subsubsection{Means of Comparison}
-In order to compare the performance of the chosen methods three different values are calculated for the resulting models: the root of the mean squared error (RMSE), the log-likelihood value and the continuous ranked probability score (CRPS).
-All of them are evaluated out of sample, e.g.
-on a part of the data set that has not been used to build the model.\\
-%Therefore, overfitting would lead to worse results.\\
-\\
-The RMSE represents how close the estimated expected values are to the true observations.
-For a response variable $Y$ and a set of observations $\textbf{y} = \{y_i\}_{i = 1,\ldots,n}$ with the corresponding estimated parameters $\hat{\boldsymbol{\theta}} = \{\hat{\theta}_i\}_{i = 1,\ldots,n}$ the RMSE is defined as follows.
-
-$$
-\text{RMSE}(Y, \boldsymbol{y}, \boldsymbol{\hat{\theta}}) = \sqrt{\frac{1}{n} \sum_{i=1}^n (\mathbb{E}_{\hat{\theta}_i}(Y) - y_i)^2}
-$$
-
-In most cases this value does not measure the distributional fit as only the expected value is considered.
-However, in the case of a censored normal distribution the expected value of the observed variable is a transformation of the expected value of the latent variable which includes the location and the scale parameter.\\
-\\
-A typical way to evaluate the distributional fit is provided by the log-likelihood value.
-For each observation $y$ and the corresponding predicted parameter $\hat{\theta}$ the value of the log-likelihood function $\ell(\hat{\theta}; y)$ is calculated and then averaged over all observations in the testing data set.
-The higher the value the better the predictions of the model are.\\
-\\
-The CRPS is another value that measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
-$$
-\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \le z))^2 dz
-$$
-where \(\mathbf{1}\) is the indicator function.
-Again, the value is averaged over all observations in the testing data set.
-As the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
-One big advantage in terms of interpretability is that the resulting value is on the scale of the observation.
-\\
-\\
-As we are interested in comparing the performance of the distributional forest to those of other models, we choose one model as a reference and calculated the skills score with respect to the CRPS.
-In that way we obtain a measure of how much a method improves the model fit in terms of CRPS values compared to the reference method.
-$$
-\text{SkillsScore(method)} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
-$$
+This choice has been made based on the results of \cite{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared.
+Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.
 
 \SweaveOpts{eval = TRUE}
-\subsection{Station Axams}
+\subsection{Application on one station}
 First of all we focus on one observation station, Axams in Tyrol.
@@
 As for all other stations, daily observations of the month of July of 28 years are available, starting in 1985 and ending in 2012, both years included.\\
-
-%%\subsubsection{Predictions}
-%In a first step one model is learned on the first 25 years (1985--2009, both included) and then predictions for the 26th, 27th and 28th year (2010, 2011 and 2012) are made.
-To illustrate the results, one day of July is chosen and the density functions of the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations in Figure \ref{axams_onestation}.\\
-%The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
-Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
-
-\SweaveOpts{eval = FALSE}
-<<echo=FALSE, results=hide>>=
-#### Axams prediction 25 - 3
-if(file.exists("rain_Axams_pred.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
-  load("rain_Axams_pred.rda")
-} else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
-  source("rain_axams_pred.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-  res <- rain_axams_pred(seedconst = 7, ntree = 100,
-                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
-                         gamboost_cvr = TRUE,
-                         frac = FALSE)
-
-  save(res, file = "rain_Axams_pred.rda")
-}
-
-# prepare data for plot of estimated density functions
-pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
-
-pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
-pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, res$testdata[pdays,"robs"])
-
-# plot predicted distributions together with observations
-#set.seed(res$call$seedconst)
-set.seed(7)
-x <- c(0.01, sort(runif(500,0.01,8)))
-y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
-y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
-y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
-dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
-
-# point mass (slightly shifted)
-pm1 <- c(0.04, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
-pm2 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
-pm3 <- c(-0.1, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
-
-# predictions
-pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-
-#legendheight
-lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
-lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
-lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-@
-
-\begin{figure}[t!]
-\begin{center}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
-plot(x = x, y = y1, type = "l", col = "red", 
-     main = paste0("July ", pday, dayending), ylab = "Density", 
-     xlab = "Total precipitation (power transformed)",
-     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
-     xlim = c(-1.5,8))
-
-lines(x = x, y = y2, type = "l", col = "blue")
-    lines(x = x, y = y3, type = "l", col = "darkgreen")
-    #legend('topright', c("2010", "2011", "2012"), col = c("red", "blue", "darkgreen"), lty = 1, cex = 1)
-
-    # plot point mass
-    lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
-    lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
-    lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
-    points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
-    points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
-    points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
-
-    # plot predictions
-    points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
-    points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
-    points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
-
-    lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
-    lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
-    lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
-
-    # add labels
-    text(x = -0.8, y = lh1, labels = "2010", col = "red", cex = 0.8)
-    text(x = -0.8, y = lh2, labels = "2011", col = "blue", cex = 0.8)
-    text(x = -0.8, y = lh3, labels = "2012", col = "darkgreen", cex = 0.8)
-@
-\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.}
-\end{center}
-\end{figure}
-
-\SweaveOpts{eval = TRUE}
-
+The results of fitting one distributional forest to the first 25 years of the data set and predicting the total amount of precipitation for one day of each of the three successive years have been shown in the introduction in Figure \ref{axams_onestation} as a motivational example for a probabilistic forecast.
+Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on PIT-histograms.\\
+\\
 As seen in Table \ref{model_specification} the forest model includes all 82 covariates.
@@
 \SweaveOpts{eval = TRUE}
-\subsubsection{Cross-Validation}
-After analyzing the results of the models on one data set only where visualization is easier we now want to get a more profound impression of the performances of the different methods.
+%\subsubsection{Cross-Validation}
+After analyzing the results of the models on one data set only where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
 For that reason a cross validation framework is considered for the station Axams.
@@
 While 9 of them are used as learning data the 10th is the testing data for which predictions are made and evaluated.
-This is done by calculating the RMSE, the log-likelihood value and the CRPS value for each of the observations in the testing data and then averaging them.
+This is done by calculating (the RMSE, the log-likelihood value and) the CRPS value for each of the observations in the testing data and then averaging them.
 Each of the 10 parts is selected as testing data once which leads to 10 different settings over which, again, the average values are stored.
 This whole procedure is repeated 10 times.
+\\
+As we are interested in comparing the performance of the distributional forest to those of other models, we choose one model as a reference and calculated the skills score with respect to the CRPS.
+In that way we obtain a measure of how much a method improves the model fit in terms of CRPS values compared to the reference method.
+$$
+\text{SkillsScore(method)} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
+$$
 The results regarding the CRPS values are illustrated by the boxplots in Figure \ref{boxplots_crps_axams} where the skills score of the CRPS is the chosen measurement scale with the EMOS model as the reference model represented by the horizontal line at height 0.\\
-The boxplots of the RMSE and the log-likelihood values can be found in the appendix.
+(The boxplots of the RMSE and the log-likelihood values can be found in the appendix.)
 <<echo=FALSE, results=hide>>=
@@
 %\newpage
-\subsection{Predictions over all stations}
+\subsection{Application on all stations}
 Until now the focus has been on one observation station only.
@@
 \begin{center}
-<<map, fig=TRUE, echo=FALSE>>=
+<<map, fig=TRUE, echo=FALSE, height=6, width=9>>=
 
   raster::image(tirol.dem, col = rev(gray.colors(100)),
-                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude")
+                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
+                xlim = c(9.8,13.2), 
+                ylim = c(46.6, 47.8))
   plot(tirol.gadm, add = TRUE)
-  points(sp, pch = 21, bg = level_col_crps_ssc, col = 1)
+  points(sp, pch = 21, bg = level_col_crps_ssc, col = 1, cex = 1.5)
   #points(sp, pch = 21, bg = col_crps_ssc, col = 1)
   #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
-  legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS"), cex = 1)
+  legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS"), cex = 1, bty = "n")
 

### 2017-11-09 pkg/disttree/inst/draft/draft.Rnw
edited plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1689 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c648d42d2eccbc7bc58393b5abcb0904ab23579b

@@
 \usepackage{enumitem}
+\usepackage{graphicx}
 \usepackage{array, makecell}
@@
 lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+
+## pit histograms and qqr-plots
+
+## get predicted parameter
+# for learndata and testdata
+{
+  ## disttree
+  if(FALSE){
+    # in sample
+    pdt <- predict(res$dt, type = "parameter")
+    dt_mu_l <- pdt$mu 
+    dt_sigma_l <- pdt$sigma 
+    pit_dt_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = dt_mu_l, sd = dt_sigma_l))
+    pit_dt_l[res$learndata[,"robs"]>0, 1] <- pit_dt_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  pdt <- predict(res$dt, type = "parameter", newdata = res$testdata)
+  dt_mu_t <- pdt$mu 
+  dt_sigma_t <- pdt$sigma 
+  pit_dt_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = dt_mu_t, sd = dt_sigma_t))
+  pit_dt_t[res$testdata[,"robs"]>0, 1] <- pit_dt_t[res$testdata[,"robs"]>0, 2]
+
+  ## distforest
+  if(FALSE){
+    # in sample
+    pdf <- predict(res$df, type = "parameter")
+    df_mu_l <- pdf$mu
+    df_sigma_l <- pdf$sigma
+    pit_df_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = df_mu_l, sd = df_sigma_l))
+    pit_df_l[res$learndata[,"robs"]>0, 1] <- pit_df_l[res$learndata[,"robs"]>0, 2]
+  }
+  # out of sample
+  pdf <- predict(res$df, type = "parameter", newdata = res$testdata)
+  df_mu_t <- pdf$mu
+  df_sigma_t <- pdf$sigma
+  pit_df_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = df_mu_t, sd = df_sigma_t))
+  pit_df_t[res$testdata[,"robs"]>0, 1] <- pit_df_t[res$testdata[,"robs"]>0, 2]
+
+  ## gamlss
+  g_learndata <- res$learndata
+  g_learndata$robs <- Surv(g_learndata$robs, g_learndata$robs>0, type="left")
+  if(FALSE){
+    # in sample
+    g_mu_l <- predict(res$g, what = "mu", type = "response")
+    g_sigma_l <- predict(res$g, what = "sigma", type = "response")
+    pit_g_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = g_mu_l, sd = g_sigma_l))
+    pit_g_l[res$learndata[,"robs"]>0, 1] <- pit_g_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  g_mu_t <- predict(res$g, what = "mu", type = "response", newdata = res$testdata)
+  g_sigma_t <- predict(res$g, what = "sigma", type = "response", newdata = res$testdata)
+  pit_g_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = g_mu_t, sd = g_sigma_t))
+  pit_g_t[res$testdata[,"robs"]>0, 1] <- pit_g_t[res$testdata[,"robs"]>0, 2]
+
+  #gamboostLSS
+  if(FALSE){
+    #in sample
+    pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response")
+    gb_mu_l <- pgb$mu
+    gb_sigma_l <- pgb$sigma
+    pit_gb_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = gb_mu_l, sd = gb_sigma_l))
+    pit_gb_l[res$learndata[,"robs"]>0, 1] <- pit_gb_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response", newdata = res$testdata)
+  gb_mu_t <- pgb$mu
+  gb_sigma_t <- pgb$sigma
+  pit_gb_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = gb_mu_t, sd = gb_sigma_t))
+  pit_gb_t[res$testdata[,"robs"]>0, 1] <- pit_gb_t[res$testdata[,"robs"]>0, 2]
+
+  #EMOS
+  if(FALSE){
+    # in sample
+    ml_mu_l <- predict(res$ml, type = "location")     # returns parameter on response scale
+    ml_sigma_l <- predict(res$ml, type = "scale")
+    pit_ml_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = ml_mu_l, sd = ml_sigma_l))
+    pit_ml_l[res$learndata[,"robs"]>0, 1] <- pit_ml_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  ml_mu_t <- predict(res$ml, type = "location", newdata = res$testdata)     # returns parameter on response scale
+  ml_sigma_t <- predict(res$ml, type = "scale", newdata = res$testdata)
+  pit_ml_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = ml_mu_t, sd = ml_sigma_t))
+  pit_ml_t[res$testdata[,"robs"]>0, 1] <- pit_ml_t[res$testdata[,"robs"]>0, 2]
+
+}
 @
@@
 
-\SweaveOpts{eval = FALSE}
+\SweaveOpts{eval = TRUE}
 
 <<echo=FALSE, results=hide>>=
-source("oneset.R")
-if(file.exists("oneset1.rda")){
-  load("oneset1.rda")
-} else {
+
+if(!(file.exists("plot_oneset1_mu.jpeg") & file.exists("plot_oneset1_sigma.jpeg") &
+   file.exists("plot_oneset10_mu.jpeg") & file.exists("plot_oneset10_sigma.jpeg"))){
 
@@
   gen.cens("NO", type = "left")
-
-  kappa <- 1
-  fun <- function(x){
-    mu <- 0.2 + 
-      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
-      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 10/2) * as.numeric(x[,1]>= 0.4) #+
-    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) +
-    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
-    sigma <- 0.5 + 
-      #mu/5 + 
-      2 * (x[,1] < -0.4) + 
-      3 * x[,4] #+
-    #1*(x[,1] < -0.7) 
-    #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
-    #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
-    #5 * abs(x[,1]) * x[,4]
-    par <- cbind(mu, sigma)
-    return(par)
-  }
-
-  oneset1 <- sim_oneset(parfun = fun, nobs = 500, seedconst = 74, ntree = 100,
-                        formula = y~x1+x2+x3+x4+x5+x6, 
-                        #formula = y~x1+x2+x4+x5,
-                        #formula = y~x1+x2+x3,
-                        tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
-                        forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
-                        forest_mtry = 3, type.tree = "ctree", 
-                        censNO = TRUE, 
-                        pred_fix = TRUE,
-                        pred_fix_x1 = 0.9,
-                        pred_fix_x2 = 0.9,
-                        pred_fix_x4 = 1,
-                        gamboost_cvr = FALSE,
-                        eval_disttree = TRUE,
-                        eval_distforest = TRUE,
-                        eval_gamlss = TRUE,
-                        nrunif = 3,
-                        nrcovform = 6)
-
-  save(oneset1, file = "oneset1.rda")
-}
-
-if(file.exists("oneset10.rda")){Hans-Brenner-Platz
-  load("oneset10.rda")
-} else {
-
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-
-  kappa <- 10
-  fun <- function(x){
-    mu <- 0.2 + 
-      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
-      (10 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 10/2) * as.numeric(x[,1]>= 0.4) #+
-    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) +
-    #(3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
-    sigma <- 0.5 + 
-      #mu/5 + 
-      2 * (x[,1] < -0.4) + 
-      3 * x[,4] #+
-    #1*(x[,1] < -0.7) 
-    #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
-    #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
-    #5 * abs(x[,1]) * x[,4]
-    par <- cbind(mu, sigma)
-    return(par)
-  }
-
-  oneset10 <- sim_oneset(parfun = fun, nobs = 500, seedconst = 74, ntree = 100,
-                        formula = y~x1+x2+x3+x4+x5+x6, 
-                        #formula = y~x1+x2+x4+x5,
-                        #formula = y~x1+x2+x3,
-                        tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
-                        forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
-                        forest_mtry = 3, type.tree = "ctree", 
-                        censNO = TRUE, 
-                        pred_fix = TRUE,
-                        pred_fix_x1 = 0.9,
-                        pred_fix_x2 = 0.9,
-                        pred_fix_x4 = 1,
-                        gamboost_cvr = FALSE,
-                        eval_disttree = TRUE,
-                        eval_distforest = TRUE,
-                        eval_gamlss = TRUE,
-                        nrunif = 3,
-                        nrcovform = 6)
-
-  save(oneset10, file = "oneset10.rda")
-}
+  source("oneset.R")
+
+}
+
 @
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset1, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
-              compare_mu = TRUE, 
-              add_dt = TRUE,
-              add_df = TRUE,
-              add_g = TRUE,
-              nomodel = FALSE)
-@
+\includegraphics{plot_oneset1_mu.jpeg}
 \end{subfigure}
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_oneset1_sigma, fig=TRUE, echo=FALSE>>=
-  plot_oneset(oneset1, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
-              compare_mu = FALSE,
-              compare_sigma_line = TRUE, 
-              add_dt = TRUE,
-              add_df = TRUE,
-              add_g = TRUE,
-              nomodel = TRUE)
-@
+\includegraphics{plot_oneset1_sigma.jpeg}
 \end{subfigure}
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset10, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
-              compare_mu = TRUE, 
-              add_dt = TRUE,
-              add_df = TRUE,
-              add_g = TRUE,
-              nomodel = FALSE)
-@
+\includegraphics{plot_oneset10_mu.jpeg}
 \end{subfigure}
@@
 \begin{subfigure}{0.6\textwidth}
-<<sim_oneset10_sigma, fig=TRUE, echo=FALSE>>=
-  plot_oneset(oneset10, only_dt = FALSE, only_df = FALSE, only_g = FALSE,
-              compare_mu = FALSE,
-              compare_sigma_line = TRUE, 
-              add_dt = TRUE,
-              add_df = TRUE,
-              add_g = TRUE,
-              nomodel = TRUE)
-@
+\includegraphics{plot_oneset10_sigma.jpeg}
 \end{subfigure}
@@
 
-<<pit_qqr, echo=FALSE, eval=TRUE, results=hide>>=
-
-## pit histograms and qqr-plots
-
-## get predicted parameter
-# for learndata and testdata
-{
-
-  ## disttree
-  if(FALSE){
-    # in sample
-    pdt <- predict(res$dt, type = "parameter")
-    dt_mu_l <- pdt$mu 
-    dt_sigma_l <- pdt$sigma 
-    pit_dt_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = dt_mu_l, sd = dt_sigma_l))
-    pit_dt_l[res$learndata[,"robs"]>0, 1] <- pit_dt_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  pdt <- predict(res$dt, type = "parameter", newdata = res$testdata)
-  dt_mu_t <- pdt$mu 
-  dt_sigma_t <- pdt$sigma 
-  pit_dt_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = dt_mu_t, sd = dt_sigma_t))
-  pit_dt_t[res$testdata[,"robs"]>0, 1] <- pit_dt_t[res$testdata[,"robs"]>0, 2]
-
-  ## distforest
-  if(FALSE){
-    # in sample
-    pdf <- predict(res$df, type = "parameter")
-    df_mu_l <- pdf$mu
-    df_sigma_l <- pdf$sigma
-    pit_df_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = df_mu_l, sd = df_sigma_l))
-    pit_df_l[res$learndata[,"robs"]>0, 1] <- pit_df_l[res$learndata[,"robs"]>0, 2]
-  }
-  # out of sample
-  pdf <- predict(res$df, type = "parameter", newdata = res$testdata)
-  df_mu_t <- pdf$mu
-  df_sigma_t <- pdf$sigma
-  pit_df_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = df_mu_t, sd = df_sigma_t))
-  pit_df_t[res$testdata[,"robs"]>0, 1] <- pit_df_t[res$testdata[,"robs"]>0, 2]
-
-  ## gamlss
-  g_learndata <- res$learndata
-  g_learndata$robs <- Surv(g_learndata$robs, g_learndata$robs>0, type="left")
-  if(FALSE){
-    # in sample
-    g_mu_l <- predict(res$g, what = "mu", type = "response")
-    g_sigma_l <- predict(res$g, what = "sigma", type = "response")
-    pit_g_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = g_mu_l, sd = g_sigma_l))
-    pit_g_l[res$learndata[,"robs"]>0, 1] <- pit_g_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  g_mu_t <- predict(res$g, what = "mu", type = "response", newdata = res$testdata)
-  g_sigma_t <- predict(res$g, what = "sigma", type = "response", newdata = res$testdata)
-  pit_g_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = g_mu_t, sd = g_sigma_t))
-  pit_g_t[res$testdata[,"robs"]>0, 1] <- pit_g_t[res$testdata[,"robs"]>0, 2]
-
-  #gamboostLSS
-  if(FALSE){
-    #in sample
-    pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response")
-    gb_mu_l <- pgb$mu
-    gb_sigma_l <- pgb$sigma
-    pit_gb_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = gb_mu_l, sd = gb_sigma_l))
-    pit_gb_l[res$learndata[,"robs"]>0, 1] <- pit_gb_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response", newdata = res$testdata)
-  gb_mu_t <- pgb$mu
-  gb_sigma_t <- pgb$sigma
-  pit_gb_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = gb_mu_t, sd = gb_sigma_t))
-  pit_gb_t[res$testdata[,"robs"]>0, 1] <- pit_gb_t[res$testdata[,"robs"]>0, 2]
-
-  #EMOS
-  if(FALSE){
-    # in sample
-    ml_mu_l <- predict(res$ml, type = "location")     # returns parameter on response scale
-    ml_sigma_l <- predict(res$ml, type = "scale")
-    pit_ml_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = ml_mu_l, sd = ml_sigma_l))
-    pit_ml_l[res$learndata[,"robs"]>0, 1] <- pit_ml_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  ml_mu_t <- predict(res$ml, type = "location", newdata = res$testdata)     # returns parameter on response scale
-  ml_sigma_t <- predict(res$ml, type = "scale", newdata = res$testdata)
-  pit_ml_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = ml_mu_t, sd = ml_sigma_t))
-  pit_ml_t[res$testdata[,"robs"]>0, 1] <- pit_ml_t[res$testdata[,"robs"]>0, 2]
-
-}
-
-@
-
 %\begin{figure}[t!]

### 2017-11-13 pkg/disttree/inst/draft/draft.Rnw
edited function in simulation study and further small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1699 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 13c4e6f6caf810ca0b677773bf2f8eb4c3ffa451

@@
 
-## pit histograms and qqr-plots
+## PIT histograms and qqr-plots
 
@@
 Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.
-By doing so the following hypotheses are investigated:
-\begin{itemize}%[leftmargin=0.5cm]
-\item When applying a tree algorithm the use of a fitting function which returns scores for each distribution parameter is advantageous to find correct splits.
-In particular it is crucial to find splits in parameters other than the location parameter.
-Hence a distributional forest consisting of distributional trees leads to better results than a simple forest considering only the mean.
-\item In the case of smooth parameter functions without any abrupt changes / non-additive effects a smooth parametric model such as GAMLSS is clearly the first choice while a stepfunction can be modeled best by applying a tree-structured model.
+
+%In the case of smooth parameter functions without any abrupt changes / non-additive effects a smooth parametric model such as GAMLSS is clearly the first choice while a stepfunction can be modeled best by applying a tree-structured model.
 Even though for both of these extreme situations a specifically advantageous method is provided, a forest model can be a good compromise and performs well in a wide range of settings including the two mentioned before.
 (Therefore, a forest is a recommendable choice particularly when no information about the type of parameter function is provided.)
-\item While specifying interactions and selecting covariates is necessary to achieve reasonable results in parametric models such as GAMLSS, this is done automatically within trees and forests.
-For that reason, distributional trees and forests provide a framework where no additional effort is required for the model specification.
-This feature plays to its strength especially in case of a high number of covariates and no information about their effects on the response.
-\end{itemize}
 
@@
 
-To prove these statements distributional trees and forests are evaluated on generated data sets.
-Their results are then compared to those of a GAMLSS (\cite{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution models and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
-The generated data sets consist of the values of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
+The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\cite{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
+The generated data sets include 500 observations which each includes the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
 $$
@@
 \\
-The location parameter $\mu$ depends on $X_1, X_4$ and $\kappa$ and the scale parameter $\sigma$ depends on $X_1, X_4$ and $\kappa$.
+The location parameter $\mu$ depends on $X_1, X_4$ and $\kappa$ and the scale parameter $\sigma$ depends on $\mu, X_1$ and $\kappa$.
 The other covariates $X_2, X_3, X_5$ and $X_6$ are used as noise variables.\\
@@
 $$
-f(X_1, X_2, \kappa) = \left(\mu(X_1,\kappa) , \sigma(X_2,\kappa)\right)
+f(X_1, X_4, \kappa) = \left(\mu(X_1, X_4, \kappa) , \sigma(\mu,X_1,\kappa)\right)
 $$ 
 and
-$$
-\mu(X_1, X_4, \kappa) = 0.2 + 7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \mathbf{1}(X_1 < 0.4) + 
-                              7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2} \cdot \mathbf{1}(X_1 \geq 0.4) + 
-                              3 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot X_4
-$$
-
-$$
-\sigma(X_2,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.7)
+\begin{align*}
+\mu(X_1, X_4, \kappa) = 0.2 
++ \begin{cases}
+    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
+    & \text{if } X_1 < 0.4 \text{ and } X_4 = 0\\
+    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
+    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 0\\
+    10 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
+    & \text{if } X_1 < 0.4 \text{ and } X_4 = 1\\
+    10 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{10}{2},       
+    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 1
+\end{cases}
+\end{align*}
+
+$$
+\sigma(\mu, X_1,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.7)
 $$
@@
     Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
-    0,       & else
+    0,       & \text{else}
 \end{cases}
@@
 
-To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to 1 (top plots in Figure \ref{}) and then to a second data set where $\kappa$ is set to 10 (bottom plots in Figure \ref{}).\\
-In the left plots the location parameter function $\mu(X_1, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $(x_2, x_3, \ldots, x_{10}) = (0,0, \ldots, 0)$.
+To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to 1 (Figure \ref{plot_sim_oneset1}) and then to a second data set where $\kappa$ is set to 10 (bottom plots in Figure \ref{plot_sim_oneset10}).\\
+In these plot the location parameter function $\mu(X_1, X_4, \kappa)$ (left window) and the the scale parameter function $\sigma(\mu, X_1, \kappa)$ (right window) are plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $x_4 = 0$ and $(x_2, x_3, x_5, x_6) = (0,0, \ldots, 0)$.
 \\
-In the right plots the scale parameter function $\sigma(X_2, \kappa)$ is plotted along the covariate $X_2$ together with the estimated values of the different models for fixed covariates $(x_1, x_3, \ldots, x_{10}) = (0.4, 0, \ldots, 0)$.
-\\
 
 \SweaveOpts{eval = TRUE}
-
 <<echo=FALSE, results=hide>>=
@@
 \end{subfigure}
-\caption{\label{plot_sim_oneset1} ... }
+\caption{\label{plot_sim_oneset1} Smooth parameter functions ($\kappa = 1$) together with the predicted parameter of the tree model, the forest model and the GAMLSS.}
 \end{figure}
@@
 \end{subfigure}
-\caption{\label{plot_sim_oneset10} ... }
+\caption{\label{plot_sim_oneset10} Parameter functions with abrupt changes ($\kappa = 10$) together with the predicted parameter of the tree model, the forest model and the GAMLSS.}
 \end{figure}
 
-In the case of smooth effects of the coefficients on the parameter function the GAMLSS is closest to the true parameter while the steps of the tree are cleary too rough.
+In the case of smooth effects of the coefficients on the parameter function ($\kappa = 1$, Figure \ref{plot_sim_oneset1}) the fitted parameters of the GAMLSS are closest to the true parameters while the steps of the tree are cleary too rough.
 By combining various trees the forest can smooth out these steps, however, not as well as the GAMLSS.\\
-Looking at the other extreme situation in which the parameter functions include abrupt changes the tree captures them very well wich also leads to a good fit of the forest model.
-In contrast to the two tree structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter function.\\
-\\
+Looking at the other extreme situation in which the parameter functions include abrupt changes ($\kappa = 10$, Figure \ref{plot_sim_oneset10}) the tree captures them very well wich also leads to a good fit of the forest model.
+In contrast to the two tree structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter functions.\\
 
 Now, to get a more profound impression, the methods are evaluated on various data sets.
-The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step ...
-data sets are generated.\\
-\\
+The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step 100 data sets are generated.\\
 %\subsubsection{Means of Comparison}
@@
 
-Figure \ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study.
+Figure \ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study and supports the assumptions made above based on the results for the two specific situations in Figures \ref{plot_sim_oneset10} and \ref{plot_sim_oneset10}.
+Moreover, it can be observed how the stepwise transition from the one extreme situation to the other influences the performance of the applied methods.\\
 The parametric GAMLSS can easily deal with the smooth effects that appear for low values of $\kappa$ and outperforms the tree and forest models in this situation.
 However, it clearly starts to struggle as $\kappa$ increases and the effects turn into abrupt changes.
-This behaviour has already been detected and discussed above when looking at only single data sets and is now proven by an almost monotone increase of the CRPS.\\
+%This behaviour has already been detected and discussed above when looking at only single data sets and is now proven by an almost monotone increase of the CRPS.\\
 On the contrary, the imposed/enforced steps of the tree model lead to a high CRPS value in the situation of smooth effects while this value decreases quickly as the slopes in the parameter functions get steeper.
@@
 As this level lies inbetween the ones of the other two methods the forest model provides a good compromise.
-Of course, if the type of effects is known in advance one should choose the corresponding adequate method, but as this is usually not the case applying a distributional forest offers a solution which can deal with a wide variety of situations and in that way secures a good fit of the resulting model.
+Of course, if the type of effects is known in advance one should choose the corresponding adequate method, but as this is usually not the case, applying a distributional forest offers a solution which can deal with a wide variety of situations and in that way secures a good fit of the resulting model.
 
@@
 
-As explained in the introduction distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
+As explained in the introduction, distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
 To show their strengths in practice, these new methods are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
@@
 \subsection{Setting}
-The data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both years included.
+The data set used for training and validation consists of observed daily
+precipitation sums provided by the national hydrographical service
+(\citealt{ehyd}) and numerical weather forecasts from the {U.S.} national
+oceanic and atmospheric administration (NOAA).
+Both, observations and weather forecasts are available from 1985 trough the
+year 2012, focussing on July only in this article.\\
+Observations are available on a daily basis, each observation representing the
+total precipitation sum fallen within 24~hours observed at 6~UTC at \textbf{95}
+stations all over Tyrol and it's close surrounding. The observations are
+undergoing a rigorous quality check before made available to the public.  As
+covariates the second generation reforecast data set of the global ensemble
+forecast system (GEFS) is used (\citealt{hamill2013}).  This data set consists
+of an 11-member ensemble based on a fixed version of the numerical model and a
+horizontal grid-spacing of about $50~\times~50 km^2$ initialized at 0~UTC from
+Dezember 1984 to precent. Each of the 11 ensemble members uses slightly
+different perturbed initial conditions trying to predict the situation specific
+uncertenty of the atmospheric state.
+
+%The data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both years included.
 The observations were measured at 95 (?) stations all over Tyrol (and regions close to the frontier).
 The considered data is provided by .....
-Next to the variable of interest which is the measured total precipitation within 24 hours of each day (measured from 6 UTC to 6 UTC of the following day), numerical forecasts (from ...
+%Next to the variable of interest which is the measured total precipitation within 24 hours of each day (measured from 6 UTC to 6 UTC of the following day), numerical forecasts (from ...
 ) are used as covariates.
@@
 individual forecasts based on varying/perturbed initial conditions.\\
+
 In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates.
@@
 Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
+To remove the skewness of ...
+(the distribution of variables representing precipitation amounts), a power transformation has been used frequently (Box and Cox 1964).
+In literature, cubic (\citealt{stidd1973}) or square root (\citealt{hutchinson1998b}) transformations have often been suggested but may vary for different climatic zones or temporal aggregation periods.
 The response variable \texttt{tp} (total precipitation) is power transformed by a factor of 1.6.
-This is done due to experience and knowledge about possible distributions of precipitation measurements.
+This factor is chosen due to experience and knowledge about possible distributions of precipitation measurements.
 The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
@@
 \subsection{Models}
-Next to distributional forests other methods are applied to this data set and their results are compared based on RMSE, log-likelihood values and CRPS values.
+Next to distributional forests three other methods are applied to this data set and their results are compared based on (RMSE, log-likelihood values and) CRPS values.
+One of the other models is an EMOS models as this is a common choice in this type of setting / research field and therefore provides a valueable reference.
+The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameters (denoted as GAM (prespecified)) and once applying a boosting algorithm (GAM (boosted)).
 The chosen methods and their specifications are listed in Table \ref{model_specification}.
@@
 The results of fitting one distributional forest to the first 25 years of the data set and predicting the total amount of precipitation for one day of each of the three successive years have been shown in the introduction in Figure \ref{axams_onestation} as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on PIT-histograms.\\
+Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on PIT histograms.\\
 \\
@@
 %\end{subfigure}
-\caption{\label{varimp} Variable importance: top 10 list for the forest model.}
+\caption{\label{varimp} Variable importance: Top 10 list for the forest model.}
 \end{center}
@@
 %@
-%\caption{\label{pit_insample} PIT-Histograms for station Axams (in sample)}
+%\caption{\label{pit_insample} PIT histograms for station Axams (in sample)}
 %\end{figure}
@@
 \end{subfigure}
-\caption{\label{pit_oosample} PIT-Histograms for station Axams (out of sample).}
+\caption{\label{pit_oosample} PIT histograms for station Axams (out of sample).}
 \end{center}
@@
 For these out-of-sample predictions the CRPS skills score with respect to the EMOS model as reference is calculated for each method at each station.
-In Figure \ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.\\
-Again, the corresponding boxplots of the RMSE and the log-likelihood values can be found in the appendix.
+In Figure \ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
+The red line represents the results for the station Axams.\\
+(Again, the corresponding boxplots of the RMSE and the log-likelihood values can be found in the appendix.)
 %NAs in gamlss in 13 stations!!!\\
@@
 
+\section{Discussion}
+
 %\newpage

### 2017-11-14 pkg/disttree/inst/draft/draft.Rnw
edited plots, small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1708 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fc2feb7958c8ee282b9501c9b3b83be030b69569

@@
 
-## function to estimate standard deviation of randomForest for a new observation
-# (in randomForest the argument 'keep.inbag' must be set to TRUE)
-rf_getsd <- function(rf, newdata = NULL, rfdata){
-
-  if(is.null(newdata)) newdata <- rfdata
-
-  rf_sd <- numeric(length = NROW(newdata))
-
-  for(k in 1:NROW(newdata)){
-    newobs <- newdata[k,]
-    # get predictions for the new observations from all trees
-    pred.newobs <- predict(rf, predict.all = TRUE, newdata = newobs)
-
-    # vector where the standard deviations from all trees are stored
-    sd_trees <- numeric(length = rf$ntree)
-
-    # loop over all trees of the forest
-    for(i in 1:rf$ntree){
-
-      # get data used to build this tree
-      obsid <- rep.int(c(1:NROW(rfdata)), as.vector(rf$inbag[,i]))
-      obs_tree <- rfdata[obsid,]
-      rownames(obs_tree) <- c(1:NROW(obs_tree))
-      # get predictions for this data from this tree
-      pred.obs_tree <- predict(rf, newdata = obs_tree, predict.all = TRUE)$individual[,i]
-
-      # get prediction for the new observation from this tree
-      pred.newobs_tree <- pred.newobs$individual[,i]
-
-      # get part of the data that ends up in the same terminal node (has the same prediction)
-      obs_node <- obs_tree[pred.obs_tree == pred.newobs_tree,]
-
-      sd_trees[i] <- sd(obs_node$y)
-    }
-
-    # average of sd over all trees
-    sd_newobs <- mean(sd_trees, na.rm = TRUE)
-    rf_sd[k] <- sd_newobs
-  }
-  return(rf_sd)   
-}
-
-## function to estimate standard deviation of cforest for a new observation
-cf_getsd <- function(cf, newdata = NULL){
-
-  # get IDs of predicted nodes for the learning data cfdata (does not have to be handed over as cf was learned on cfdata)
-  pred.node.learn <- predict(cf, type = "node")
-
-  # get IDs of predicted nodes for the new observations
-  if(is.null(newdata)) {
-    pred.node.new <- pred.node.learn
-    newdata <- cf$data
-  } else {
-    pred.node.new <- predict(cf, newdata = newdata, type = "node")
-  }
-
-  sdnew <- numeric(length = NROW(newdata))
-
-  for(i in 1:NROW(newdata)){
-    for(t in 1:cf$info$call$ntree){
-      nodedata <- cf$data[(pred.node.learn[[t]] == pred.node.new[[t]][i]),]
-      sdnew[i] <- sd(nodedata[,paste(cf$terms[[2]])])
-    }
-  }
-
-  return(sdnew)   
-}
-
-## simulation plot functions
 # plot RMSE
@@
       col <- c(col, pal["tree"])
-      legendnames <- c(legendnames, "disttree")
+      legendnames <- c(legendnames, "tree")
     }
@@
       col <- c(col, pal["forest"])
-      legendnames <- c(legendnames, "distforest")
+      legendnames <- c(legendnames, "forest")
     }
@@
     plot(x = simres$x.axis, y = rmse[,1], type = "l", col = col[1], ylim = ylim,
-         xlab = "kappa", ylab = "RMSE")
+         xlab = expression(kappa), ylab = "RMSE")
     lines(x = simres$x.axis, y = rmse[,2], type = "l", lty = 2, col = col[1])
@@
     col <- c(col, pal["tree"])
-      legendnames <- c(legendnames, "disttree")
+    legendnames <- c(legendnames, "tree")
   }
@@
     col <- c(col, pal["forest"])
-      legendnames <- c(legendnames, "distforest")
+    legendnames <- c(legendnames, "forest")
   }
@@
   plot(x = simres$x.axis, y = ll[,1], type = "l", col = col[1], ylim = ylim,
-         xlab = "kappa", ylab = "log-likelihood")
+       xlab = expression(kappa), ylab = "log-likelihood")
 
@@
     col <- c(col, pal["tree"])
-    legendnames <- c(legendnames, "disttree")
+    legendnames <- c(legendnames, "tree")
   }
@@
     col <- c(col, pal["forest"])
-    legendnames <- c(legendnames, "distforest")
+    legendnames <- c(legendnames, "forest")
   }
@@
   plot(x = simres$x.axis, y = crps[,1], type = "l", col = col[1], ylim = ylim,
-       xlab = "kappa", ylab = "CRPS")
+       xlab = expression(kappa), ylab = "CRPS")
 
@@
   }
-  if(legend) legend('topleft', legendnames, 
-                    col = col, lty = 1, cex = 0.7)
+  if(legend) legend('topright', legendnames, 
+                    col = col, lty = 1, cex = 1, bty = "n")
 }
@@
 \Abstract{
-In the field of probabilistic forecasting linear models such as ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) provide common and widely used tools.
+In the field of probabilistic forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools.
 This is due to their simplicity and their interpretability.
-However, the restriction to modeling only linear effects requires further extensions which allow for more flexibility.
-Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
-A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models such as regression trees (\cite{loh2011classification}) or random forests (\cite{breiman2001random}).
+However, the restriction to model only linear effects requires further extensions which allow for more flexibility.
+Generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
+A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
 However, up to now, this class of models has not yet been integrated in probabilistic forecasting as no general distributional framework is available for these methods.
 Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge/experience from the meteorology or based on a computationally more demanding boosting approach.
+Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge/experience from meteorology or based on a computationally more demanding boosting approach.
 
@@
 Therefore, our goal is to embed the progress made in the field of probabilistic modeling in the idea of algorithmic tree and forest models.
-In particular, more flexible models such as GAMLSS (\cite{stasinopoulos2005}) should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc.
+In particular, more flexible models such as GAMLSS (\citealt{stasinopoulos2005}) should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc.
 while non-additive effects of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree.
@@
 A typical application in probabilistic forecasting in meteorology is the analysis of precipitation data.
-The two main interests lie in estimating the probability of rain as well as predicting the expected amount of precipitation.
+The two main interests lie in estimating the probability of precipitation as well as predicting the expected amount of precipitation.
 In this research area various methods have been developed and tested in the last decades.
-(Multiple) Linear models such as ensemble model output statistics (EMOS) were introduced more than a decade ago (\cite{gneiting2005calibrated}) and are still common and widely used.
+(Multiple) Linear models such as ensemble model output statistics (EMOS) were introduced more than a decade ago (\citealt{gneiting2005calibrated}) and are still common and widely used.
 Despite their simplicity and interpretability, the limitation to modeling only linear effects has risen the interest in applying more flexible models.
@@
 However, often only the expectation of the underlying distribution is estimated.
-To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
-In Figure \ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.\\
+To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
+In Figure~\ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.\\
 These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth/homogeneous setting, e.g.
@@
 However, for example in complex terrain/mountainous regions various factors can change rapidly and their effects are difficult to be modeled/represented by a smooth parametric model.
-In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models such as regression trees (\cite{loh2011classification}) or random forests (\cite{breiman2001random}).
+In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
 Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process.
-Therefore, no additional effort or knowledge for the specification of the model in advance is required.
+Therefore, no additional effort or knowledge for the specification of the model is required in advance.
 However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well.
@@
 Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure \ref{Devel_treeforest}.\\
+The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure~\ref{Devel_treeforest}.\\
 
@@
 %However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
-%(GAMLSS) (\cite{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
+%(GAMLSS) (\citealt{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
 In that way a whole distribution can be specified.
 The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
-%This development in the field of parametric modeling is illustrated in Figure \ref{Devel_parmod}.\\
+%This development in the field of parametric modeling is illustrated in Figure~\ref{Devel_parmod}.\\
 
@@
 Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node.
-One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure \ref{Devel_treeforest}.
+One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure~\ref{Devel_treeforest}.
 One way how to tackle this problem is to not just consider one tree but an ensemble of trees.
@@
 However, up to now trees which only model the mean of the response variable in its nodes are still more common and widely used.\\
-%Random forests (\cite{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\cite{hothorn2015partykit}, \cite{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
+%Random forests (\citealt{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\citealt{hothorn2015partykit}, \citealt{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
 %\\
@@
 In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
-%These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
-
-After explaining the novel methodology in Section \ref{Methodology} of this paper and presenting results from a simulation study in Section \ref{Simulation} the focus will then be put on an application in the field of probabilistic forecasting in Section \ref{ProbabilisticForecasting}.
+%These (future) steps are illustrated in Figure~\ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
+
+After explaining the novel methodology in Section~\ref{Methodology} of this paper and presenting results from a simulation study in Section~\ref{Simulation} the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}.
 The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
-In a first setting only one observation station is taken into consideration and based on the data/information from 25 years probabilistic predictions are made for the 3 successive years.
-A brief outlook on the results of this setting is given by Figure \ref{axams_onestation}.
-For the 24th of July of 2010, 2011 and 2012 the density functions of the estimated distributions for this day of all three prediction years are plotted together with the corresponding observations.
-The estimated distribution parameters for the 24th of July of 2010 are comparably high with location $3.146$ and scale $1.902$ which leads to a well fitting predicted expectation of $3.185$ close to the observed value of $3.001$.
-Also for 2011 and 2012 the estimated expected values of 1.085 and 0.483 provide reasonable predictions as the observed values are 1.121 and 0 respectively.\\
-Next to the expected values the predictions provided by these distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
+In a first setting only one observation station is taken into consideration.
+The models are trained on 25 years and prbabilistic predictions of the consecutive 3 years are validated.
+A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}.
+For the 24th of July of 2010, 2011 and 2012 the density functions of the distributions estimated by a forest model are plotted together with the corresponding observations.
+This illustrates that well fitting predicted expected values are provided by this model as they are close to the observed values.
+Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
 
 <<echo=FALSE, results=hide>>=
-#### Axams prediction 25 - 3
-if(file.exists("rain_Axams_pred.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
-  load("rain_Axams_pred.rda")
+#### Axams prediction 24 - 4
+if(file.exists("rain_Axams_pred_24to4.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred_24to4.rda")
+  load("rain_Axams_pred_24to4.rda")
 } else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
-  source("rain_axams_pred.R")
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred_24to4.R")
+  source("rain_axams_pred_24to4.R")
   library("gamlss.cens")
@@
 
-# prepare data for plot of estimated density functions
+#### prepare data for plot of estimated density functions
+# predictions for one day (in each of the four years) 
+# (19th of July 2011 is missing)
 pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
 
-pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
+pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
 pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
@@
 y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
 dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
@@
 # point mass (slightly shifted)
-pm1 <- c(0.04, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
-pm2 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
-pm3 <- c(-0.1, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
+pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
+pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
 
@@
 pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(res$testdata[pdays,"robs"][4], crch::dcnorm(res$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
 
@@
 lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
 
@@
      main = paste0("July ", pday, dayending), ylab = "Density", 
-     xlab = "Total precipitation (power transformed)",
-     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
+     #xlab = expression(Total~precipitation~group("[", list(mm^(1/1.6), plain(per 24h)),"]")),
+     xlab = expression(Total~precipitation~group("[", mm^(1/1.6)~per~day,"]")),
+     #{\frac{1}{1.6}}$,
+     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
      xlim = c(-1.5,8))
@@
 lines(x = x, y = y3, type = "l", col = "darkgreen")
-    #legend('topright', c("2010", "2011", "2012"), col = c("red", "blue", "darkgreen"), lty = 1, cex = 1)
+lines(x = x, y = y4, type = "l", col = "purple")
+legend('topright', c("predicted distribution", "point mass at censoring point", "observation"),
+       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 1)
 
@@
 lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
+lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)
+
 points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
@@
 points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
+points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)
 
@@
 points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
+points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
 
@@
 lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)
 
 # add labels
-    text(x = -0.8, y = lh1, labels = "2010", col = "red", cex = 0.8)
-    text(x = -0.8, y = lh2, labels = "2011", col = "blue", cex = 0.8)
-    text(x = -0.8, y = lh3, labels = "2012", col = "darkgreen", cex = 0.8)
-@
-\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.}
+text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
+text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
+text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
+text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
+@
+\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.
+Observations and predictions are left censored at 0.
+The corresponding point mass is shown at the censoring point (0).}
 \end{center}
@@
 \subsection{Motivation (Review/Problems)}
-As illustrated in Figure \ref{Devel_parmod} a lot of development has been made in the field of parametric modeling and a wide range of tools can be applied in research areas such as probabilistic forecasting.
+As illustrated in Figure~\ref{Devel_parmod} a lot of development has been made in the field of parametric modeling and a wide range of tools can be applied in research areas such as probabilistic forecasting.
 But as mentioned before, further features such as the ability to capture non-additive effects or interactions are desirable in many applications which demands for more flexible models.
 Since tree structured models provide these required features a combination of them with parametric models can offer a framework that is able to deal with a wide range of different modeling situations.
-In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree structured models step by step starting with a simple distributional fit in Subsection \ref{distfit}, then building a distributional tree in Subsection \ref{disttree} and finally a distributional forest in Subsection \ref{distforest}.
+In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree structured models step by step.
+We start with a simple distributional fit in Subsection~\ref{distfit}, then build a distributional tree in Subsection~\ref{disttree} and finally a distributional forest in Subsection~\ref{distforest}.
 
@@
 The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided.
-Using the GAMLSS framework (\cite{stasinopoulos2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
+Using the GAMLSS framework (\citealt{stasinopoulos2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
 This enables the model to deal with various features that can usually be very challenging in model fitting such as censoring, truncation or heavy tails.\\
@@
 Next to the censored normal distribution the censored logistic distribution is also often chosen for the modeling of precipitation amounts.
-In \cite{Gebetsberger2017} this choice has been made in order to be able to deal with heavy tails which often appear in precipitation data.
+In \citealt{Gebetsberger2017} this choice has been made in order to be able to deal with heavy tails which often appear in precipitation data.
 Another frequent application of probabilistic forecasting in meteorology is the modeling of temperature.
@@
 However, fitting one global model to the whole data set might be too much of a generalization and therefore not represent the data and its features in a reasonable way.
-Moreover, if covariates are available it is desirable to include them into the model as they might provide important information.
+Moreover, if covariates are available it is desirable to include them into the model as they might provide important additional information.
 In our application on precipitation data a wide range of covariates are available from numerical ensemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions.
@@
 In order to avoid the restriction to linear effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ to specify a more flexible model $D(Y, \theta = f_1(\bold{X}) + \ldots + f_p(\bold{X}))$ applying a GAM structure.
-Another approach to distributional regression is proposed by \cite{Klein2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
+Another approach to distributional regression is proposed by \citealt{Klein2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
 These include non-linear effects, spatial effects, random coefficients and interaction surfaces which extends the flexibility compared to parametric models such as GAMLSS modeling only smooth effects.\\
-However, the restriction to additive models remains and therefore a different approach of how to include covariates that also allows for non-additive effects and interactions is now considered.
+However, the restriction to additive models remains.
+Therefore, a different approach of how to include covariates that also allows for non-additive effects and interactions is considered in the next subsection of this article.
 In particular, the covariates can be used to separate the data into more homogeneous subgroups.
@@
 \label{disttree}
-After having fit a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is now to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors) to decide if the data set should be split into subgroups.
+Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors) to decide whether the data set should be split into subgroups or not.
 In particular, the information provided by the fitted model are the scores
@@
 These values express how well the model fits the data regarding each parameter.
-Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and each observation.\\
-Statistical tests are applied to find out whether any significant connection/dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$ is given.
+Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.\\
+Statistical tests are applied to check whether there is a significant connection/dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$.
 If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest connection/dependency.\\
@@
 
-The exact way of determining how and where to split varies over different tree algorithms.
+The exact way of determining how and where to split varies across different tree algorithms.
 (For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm or the ctree algorithm.) Summed up, the steps of building a distributional tree using one of these algorithms are
@@
 \item Repeat steps 2--4 recursively in the subgroups
-    until some stopping criterion is met.
+    until some stopping criterion is reached.
     %-- e.g., for significance or sample size.
@@
 \\
-For distributional trees the strategy of how to make a prediction for a new observation $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ is very simple and straight forward.
-All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new observation are selected and a distributional model is fit on the subset.
-To select this subset a weight vector $\bold{\omega^{tree}} = (\omega^{tree}_1, \ldots, \omega^{tree}_n)$ is included in the fitting process.
+For distributional trees the strategy of how to make a prediction for a new set of covariates $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ is very simple and straight forward.
+All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new set of covariates are selected and a distributional model is fit on the subset.
+To select this subset a weight vector $\bold{\omega}^{tree} = (\omega^{tree}_1, \ldots, \omega^{tree}_n)$ is included in the fitting process.
 $$
@@
 $$
-The predicted distribution of the new observation is then fully specified by the estimated parameter $\hat{\theta}(\bold{z}^{\ast})$ where
+The predicted distribution of the new set of covariates is then fully specified by the estimated parameter $\hat{\theta}(\bold{z}^{\ast})$ where
 $$
@@
 
-While the clear structure of such a model is advantageous in analyzing it, the abrupt changes are often too rough and impose steps on the model even in case of smooth effects.
+While the clear structure of such a tree model is advantageous in interpreting and analyzing it, the abrupt changes are often too rough and impose steps on the model even if the true underlying effect is smooth.
 In these situations bagging can solve the problem by combining an ensemble of trees such that wrongly detected abrupt changes of the model are turned into almost smooth transitions.
@@
 Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables.
-In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\cite{breiman2001random}).\\
+In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).\\
 \\
@@
 In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.\\
-Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new observation.
-This difference is represented in the weight vector, as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
+Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates.
+This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
 $\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
-For a new observation $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ this set of weights is obtained by the following procedure: For each observation $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which it ends up in the same terminal node as the new observation $\bold{z}^{\ast}$.
+For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each observation $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$.
 $$
@@
 where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
-This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new observation $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way, leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
+This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way, leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
 $$
@@
 $$
-While in a distributional tree the same distribution is specified for all observations being in the same node a distributional forest predicts and fits a different distribution for each different observation which is due to the individual weights calculated separately for each observation.
+While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates which is due to the individual weights calculated separately for each set of covariates.
 
@@
 
-The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\cite{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
-The generated data sets include 500 observations which each includes the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
+The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\citealt{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
+Each generated data set includes 500 samples consisiting of the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
 $$
@@
 The distribution of the response variable is a left censored normal distribution with censoring point 0.
-The corresponding distribution parameters are defined by a parameter function $f$ depending on the covariates and an additional factor $\kappa$ which regulates the steepness of certain sections of the parameter function.
+The corresponding distribution parameters are defined by a parameter function $f$ depending on the covariates and an additional factor $\kappa$, also known as the smoothness constraint, which regulates the steepness of certain sections of the parameter function.
 For low values of $\kappa$ the resulting parameters change rather slowly over varying values of the covariates while high values of $\kappa$ lead to abrupt changes.
@@
 
-To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to 1 (Figure \ref{plot_sim_oneset1}) and then to a second data set where $\kappa$ is set to 10 (bottom plots in Figure \ref{plot_sim_oneset10}).\\
-In these plot the location parameter function $\mu(X_1, X_4, \kappa)$ (left window) and the the scale parameter function $\sigma(\mu, X_1, \kappa)$ (right window) are plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $x_4 = 0$ and $(x_2, x_3, x_5, x_6) = (0,0, \ldots, 0)$.
+To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to 1 and then to a second data set where $\kappa$ is set to 10.
+In Figure~\ref{plot_sim_oneset_mu} the location parameter function $\mu(X_1, X_4, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $x_4 = 1$ and $(x_2, x_3, x_5, x_6) = (0,0,0,0)$.
 \\
@@
 
-if(!(file.exists("plot_oneset1_mu.jpeg") & file.exists("plot_oneset1_sigma.jpeg") &
-   file.exists("plot_oneset10_mu.jpeg") & file.exists("plot_oneset10_sigma.jpeg"))){
+source("oneset.R")
+
+#if(file.exists("oneset1.rda")){
+#  load("oneset1.rda")
+#} else {
+
+  #define parameter function
+  kappa <- 1
+  fun1 <- function(x){
+    mu <- 0.2 + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
+      (3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
+    sigma <- 0.5 + 
+      mu/5 + 
+      #(x[,1]>0.4)  #*x[,4]#+
+      2*(x[,1] < -0.4) 
+      #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
+      #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
+      #5 * abs(x[,1]) * x[,4]
+    par <- cbind(mu, sigma)
+    return(par)
+  }
 
@@
   gen.cens("NO", type = "left")
-  source("oneset.R")
-
-}
-
+
+  oneset1 <- sim_oneset_vary_x1(parfun = fun1, nobs = 500, seedconst = 74, ntree = 100,
+                                 formula = y~x1+x2+x3+x4+x5+x6, 
+                                 tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
+                                 forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
+                                 forest_mtry = 3, type.tree = "ctree", 
+                                 censNO = TRUE, 
+                                 pred_fix_x2 = 0.9,
+                                 pred_fix_x4 = 1,
+                                 nrunif = 3,
+                                 nrcovform = 6)
+
+#  save(oneset1, file = "oneset1.rda")
+#}
+
+#if(file.exists("oneset10.rda")){
+#  load("oneset10.rda")
+#} else {
+
+  #define parameter function
+  kappa <- 10
+  fun10 <- function(x){
+    mu <- 0.2 + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
+      (3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
+    sigma <- 0.5 + 
+      mu/5 + 
+      #(x[,1]>0.4)  #*x[,4]#+
+      2*(x[,1] < -0.4) 
+      #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
+      #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
+      #5 * abs(x[,1]) * x[,4]
+    par <- cbind(mu, sigma)
+    return(par)
+  }
+
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+
+  oneset10 <- sim_oneset_vary_x1(parfun = fun10, nobs = 500, seedconst = 74, ntree = 100,
+                                 formula = y~x1+x2+x3+x4+x5+x6, 
+                                 tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
+                                 forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
+                                 forest_mtry = 3, type.tree = "ctree", 
+                                 censNO = TRUE, 
+                                 pred_fix_x2 = 0.9,
+                                 pred_fix_x4 = 1,
+                                 nrunif = 3,
+                                 nrcovform = 6)
+
+#  save(oneset10, file = "oneset10.rda")
+#}
 @
@@
 \begin{subfigure}{0.6\textwidth}
-\includegraphics{plot_oneset1_mu.jpeg}
+<<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
+plot_oneset(oneset1, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
+@
 \end{subfigure}
@@
 \begin{subfigure}{0.6\textwidth}
-\includegraphics{plot_oneset1_sigma.jpeg}
+<<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
+plot_oneset(oneset10, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
+@
 \end{subfigure}
-\caption{\label{plot_sim_oneset1} Smooth parameter functions ($\kappa = 1$) together with the predicted parameter of the tree model, the forest model and the GAMLSS.}
+\caption{\label{plot_sim_oneset_mu} Smooth parameter function (left window, $\kappa = 1$) and parameter function with abrupt changes (right window, $\kappa = 10$) together with the predicted location parameter $\mu$ of the tree model, the forest model and the GAMLSS,  plotted along $x_1$ while all the other covariates are kept at fixed values.}
 \end{figure}
 
-\begin{figure}[t!]
-\begin{subfigure}{0.6\textwidth}
-\includegraphics{plot_oneset10_mu.jpeg}
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.6\textwidth}
-\includegraphics{plot_oneset10_sigma.jpeg}
-\end{subfigure}
-\caption{\label{plot_sim_oneset10} Parameter functions with abrupt changes ($\kappa = 10$) together with the predicted parameter of the tree model, the forest model and the GAMLSS.}
-\end{figure}
-
-In the case of smooth effects of the coefficients on the parameter function ($\kappa = 1$, Figure \ref{plot_sim_oneset1}) the fitted parameters of the GAMLSS are closest to the true parameters while the steps of the tree are cleary too rough.
+In the case of smooth effects of the coefficients on the parameter function ($\kappa = 1$, left window in Figure~\ref{plot_sim_oneset_mu}) the fitted parameters of the GAMLSS are closest to the true parameters while the steps of the tree are cleary too rough.
 By combining various trees the forest can smooth out these steps, however, not as well as the GAMLSS.\\
-Looking at the other extreme situation in which the parameter functions include abrupt changes ($\kappa = 10$, Figure \ref{plot_sim_oneset10}) the tree captures them very well wich also leads to a good fit of the forest model.
+Looking at the other extreme situation in which the parameter functions include abrupt changes ($\kappa = 10$, right window in Figure~\ref{plot_sim_oneset_mu}) the tree captures them very well wich also leads to a good fit of the forest model.
 In contrast to the two tree structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter functions.\\
 
-Now, to get a more profound impression, the methods are evaluated on various data sets.
-The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step 100 data sets are generated.\\
+To get a more profound impression, the methods are evaluated on various data sets. The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step 100 data sets are generated.\\
 %\subsubsection{Means of Comparison}
@@
 where \(\mathbf{1}\) is the indicator function. \\
-In this simulation study the CRPS is evaluated out of sample, e.g.
-on a data set that has not been used to build the model but has been generated by the same process and with the same settings as the original data set.
-In the following this second data set on which the CRPS is evaluated is called testing data set while the original data set on which the models are built on is called learning data set.
-The CRPS value is first calculated for each observation in the testing data set and then the average of these values is taken.
+In this simulation study the CRPS is evaluated out of sample on a data set that has not been used to build the model but has been generated by the same process and with the same settings as the original data set.
+In the following this second data set is called testing data set while the original data set on which the models are built on is called learning data set.
+The mean CRPS over the whole testing data set is used for validation.
 As the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
@@
   source("gensim.R")
-  simres <- gensim(seedconst = 7, nrep = 20, ntree = 100,
-                   nsteps = 5, stepsize = 2,
+
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+
+  simfun <- function(x, kappa){
+    mu <- 0.2 + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
+      (3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
+    sigma <- 0.5 + 
+      mu/5 + 
+      #(x[,1]>0.4)  #*x[,4]#+
+      2*(x[,1] < -0.4) 
+    #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
+    #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
+    #5 * abs(x[,1]) * x[,4]
+    par <- cbind(mu, sigma)
+    return(par)
+  }
+
+  simres <- gensim(parfun = simfun,
+                   seedconst = 7, nrep = 100, ntree = 100,
+                   nsteps = 10, stepsize = 1,
                    formula = y~x1+x2+x3+x4+x5+x6,
-                   nobs = 400, testnobs = 200L,
-                   tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95,
-                   forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
+                   nobs = 500, testnobs = 500L,
+                   tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95,
+                   forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
                    forest_mtry = 3,
-                   fix.mu = FALSE,
-                   fix.sigma = FALSE,
-                   mu.sigma.interaction = FALSE,
-                   mubase = 1,
-                   sigmabase = 3,
                    censNO = TRUE,
@@
                    eval_randomForest = FALSE,
-                   eval_cforest = FALSE)
+                   eval_cforest = FALSE,
+                   nrunif = 3,
+                   nrcovform = 6)
 
@@
 
-Figure \ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study and supports the assumptions made above based on the results for the two specific situations in Figures \ref{plot_sim_oneset10} and \ref{plot_sim_oneset10}.
+Figure~\ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study and supports the assumptions made above based on the results for the two specific situations in Figure~\ref{plot_sim_oneset_mu}.
 Moreover, it can be observed how the stepwise transition from the one extreme situation to the other influences the performance of the applied methods.\\
@@
 %This behaviour has already been detected and discussed above when looking at only single data sets and is now proven by an almost monotone increase of the CRPS.\\
-On the contrary, the imposed/enforced steps of the tree model lead to a high CRPS value in the situation of smooth effects while this value decreases quickly as the slopes in the parameter functions get steeper.
+On the contrary, the imposed/enforced steps of the tree model lead to a high CRPS in the situation of smooth effects while decreasing quickly as the slopes in the parameter functions get steeper.
 For high values of $\kappa$ resulting almost in step functions the tree model perfomes the best as stated by the low CRPS value.\\
@@
 As this level lies inbetween the ones of the other two methods the forest model provides a good compromise.
-Of course, if the type of effects is known in advance one should choose the corresponding adequate method, but as this is usually not the case, applying a distributional forest offers a solution which can deal with a wide variety of situations and in that way secures a good fit of the resulting model.
+If the type of effects is known in advance one should choose the corresponding adequate method.
+But as this is usually not the case, applying a distributional forest offers a good and flexible solution which can deal with a wide variety of situations and in that way ensures a reasonable well fit for a wide rango of situations where no prior information on the form of the effects is available.
 
@@
 Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std dev of mean/minimal/maximal values respectively.\\
-Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
+Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
 To remove the skewness of ...
@@
 \textbf{t500}: temperature   &   & \\
-\hspace*{1cm} on 500 hPA     &   & \\
+\hspace*{1cm} on 500 hPa     &   & \\
 \textbf{t700}: temperature   &   & \\
-\hspace*{1cm} on 700 hPA     &   & \\
+\hspace*{1cm} on 700 hPa     &   & \\
 \textbf{t850}: temperature   &   & \\
-\hspace{1cm} on 850 hPA      &   & \\
+\hspace{1cm} on 850 hPa      &   & \\
 \hline
 \textbf{tdiff500850}: temperature         & 3 variations: & ensemble mean of difference in mean,\\
-\hspace*{\fill} difference 500 to 850 hPA &   & ensemble minimum of difference in mean,\\
+\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
 \textbf{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
-\hspace*{\fill} difference 500 to 700 hPA &   & \qquad all over 6--30 UTC\\
+\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
 \textbf{tdiff700850}: temperature         &   & \\
-\hspace*{\fill} difference 700 to 850 hPA &   & \\
+\hspace*{\fill} difference 700 to 850 hPa &   & \\
 \hline
@@
 The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameters (denoted as GAM (prespecified)) and once applying a boosting algorithm (GAM (boosted)).
-The chosen methods and their specifications are listed in Table \ref{model_specification}.
+The chosen methods and their specifications are listed in Table~\ref{model_specification}.
 \begin{table}[t!]
@@
 In the EMOS model the logarithmic function is chosen as the link function for the scale parameter.
-This choice has been made based on the results of \cite{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared.
+This choice has been made based on the results of \citealt{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared.
 Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.
@@
 As for all other stations, daily observations of the month of July of 28 years are available, starting in 1985 and ending in 2012, both years included.\\
-The results of fitting one distributional forest to the first 25 years of the data set and predicting the total amount of precipitation for one day of each of the three successive years have been shown in the introduction in Figure \ref{axams_onestation} as a motivational example for a probabilistic forecast.
+The results of fitting one distributional forest to the first 25 years of the data set and predicting the total amount of precipitation for one day of each of the three successive years have been shown in the introduction in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast.
 Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on PIT histograms.\\
 \\
-As seen in Table \ref{model_specification} the forest model includes all 82 covariates.
+As seen in Table~\ref{model_specification} the forest model includes all 82 covariates.
 In a single tree model it can easily be observed which covariates are used as split variables and hence have an impact on the response variable.
@@
 To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been chosen.
-The 10 leading covariates of the above described forest model for the station Axams learned on the first 25 year are listed in Figure \ref{varimp}.
+The 10 leading covariates of the above described forest model for the station Axams learned on the first 25 year are listed in Figure~\ref{varimp}.
 As expected, the numerical forecast for total precipitation is the most frequently selected.
@@
 
-To get another impression of the quality of the predictions made by the tested methods a probability integral transform (PIT) histograms is plotted in Figure \ref{pit_oosample} for out-of-sample predictions of each of the models.
+To get another impression of the quality of the predictions made by the tested methods a probability integral transform (PIT) histograms is plotted in Figure~\ref{pit_oosample} for out-of-sample predictions of each of the models.
 A perfectly fitting model would lead to a histogram with all bars being of height 1 as indicated by the red horizontal line.
@@
 For that reason a cross validation framework is considered for the station Axams.
-The data set over all 28 years is split randomly into 10 roughly equally sized data sets.
-While 9 of them are used as learning data the 10th is the testing data for which predictions are made and evaluated.
+The data set over all 28 years is split into 7 subsets each consisting of 4 randomly selected years.
+While 6 of these subsets are used as learning data the 7th is the testing data for which predictions are made and evaluated.
 This is done by calculating (the RMSE, the log-likelihood value and) the CRPS value for each of the observations in the testing data and then averaging them.
-Each of the 10 parts is selected as testing data once which leads to 10 different settings over which, again, the average values are stored.
+Each of the 7 parts is selected as testing data once which leads to 7 different settings over which, again, the average values are stored.
 This whole procedure is repeated 10 times.
@@
 $$
-The results regarding the CRPS values are illustrated by the boxplots in Figure \ref{boxplots_crps_axams} where the skills score of the CRPS is the chosen measurement scale with the EMOS model as the reference model represented by the horizontal line at height 0.\\
+The results regarding the CRPS values are illustrated by the boxplots in Figure~\ref{boxplots_crps_axams} where the skills score of the CRPS is the chosen measurement scale with the EMOS model as the reference model represented by the horizontal line at height 0.\\
 (The boxplots of the RMSE and the log-likelihood values can be found in the appendix.)
@@
 \center
-\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 10x10 cross validation at the station Axams.}
+\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 7x10 cross validation at the station Axams.}
 \end{center}
@@
 For these out-of-sample predictions the CRPS skills score with respect to the EMOS model as reference is calculated for each method at each station.
-In Figure \ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
+In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
 The red line represents the results for the station Axams.\\
@@
 
-On the map in Figure \ref{map} each observation station is represented by a point colored according to the method that performed best in terms of CRPS at this station.
+On the map in Figure~\ref{map} each observation station is represented by a point colored according to the method that performed best in terms of CRPS at this station.
 For those points where the distributional forest is not the best performing method the intensity of the color represents how much better (in terms of CRPS) the method is compared to the distributional forest.

### 2017-11-15 pkg/devel/partykit/R/cforest.R
experimental
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1713 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9f640b8e6a9a7f4d1402d4ba695ab2cbbea039a2

@@ -1,0 +1,271 @@
+
+### constructor for forest objects
+constparties <- function(nodes, data, weights, fitted = NULL, terms = NULL, info = NULL) {
+
+    stopifnot(all(sapply(nodes, function(x) inherits(x, "partynode"))))
+    stopifnot(inherits(data, "data.frame"))
+    stopifnot(inherits(weights, "list"))
+
+    if(!is.null(fitted)) {
+        stopifnot(inherits(fitted, "data.frame"))
+        stopifnot(nrow(data) == 0L | nrow(data) == nrow(fitted))
+        if (nrow(data) == 0L)
+            stopifnot("(response)" %in% names(fitted))
+    } else {
+        stopifnot(nrow(data) > 0L)
+        stopifnot(!is.null(terms))
+        fitted <- data.frame("(response)" = model.response(model.frame(terms, data = data, 
+                                                                       na.action = na.pass)),
+                             check.names = FALSE)
+    }
+
+    ret <- list(nodes = nodes, data = data, weights = weights, fitted = fitted)
+    class(ret) <- c("constparties", "parties")
+
+    if(!is.null(terms)) {
+        stopifnot(inherits(terms, "terms"))
+        ret$terms <- terms
+    }
+
+    if (!is.null(info))
+        ret$info <- info
+
+    ret
+}
+
+.perturb <- function(replace = FALSE, fraction = .632) {
+    ret <- function(prob) {
+        if (replace) {
+            rw <- rmultinom(1, size = length(prob), prob = prob)
+        } else {
+            rw <- integer(length(prob))
+            i <- sample(1:length(prob), ceiling(fraction * length(prob)), prob = prob)
+            rw[i] <- 1L
+        }
+        as.integer(rw)
+    }
+    ret
+}
+
+cforest <- function
+(
+    formula,
+    data,   
+    weights,
+    subset, 
+    offset, 
+    cluster,
+    strata,
+    na.action = na.pass,
+    control = ctree_control(
+        teststat = "quad", testtype = "Univ", mincriterion = 0,
+        saveinfo = FALSE, ...),
+    ytrafo = NULL, 
+    scores = NULL, 
+    ntree = 500L, 
+    perturb = list(replace = FALSE, fraction = 0.632),
+    mtry = ceiling(sqrt(nvar)), 
+    applyfun = NULL,
+    cores = NULL, 
+    trace = FALSE,
+    ...
+) {
+   
+    ### get the call and the calling environment for .urp_tree
[…]

### 2017-11-15 pkg/devel/partykit/R/ctree.R

@@ -1,0 +1,502 @@
+
+.ctree_test <- function(model, trafo, data, subset, weights, j, SPLITONLY = FALSE, ctrl) {
+
+    ix <- data[[j, type = "index"]]
+    iy <- data[["yx", type = "index"]]
+    Y <- model$estfun
+
+    if (!is.null(iy)) {
+        stopifnot(NROW(levels(iy)) == (NROW(Y) - 1))
+        return(.ctree_test_2d(data = data, j = j, Y = Y, iy = iy, 
+                              subset = subset, weights = weights, 
+                              SPLITONLY = SPLITONLY, ctrl = ctrl))
+    }
+
+    stopifnot(NROW(Y) == length(ix))
+
+    NAyx <- data[["yx", type = "missing"]]
+    NAz <- data[[j, type = "missing"]]
+    if (ctrl$MIA && (ctrl$splittest || SPLITONLY)) {
+        subsetNArm <- subset[!(subset %in% NAyx)]
+    } else {
+        subsetNArm <- subset[!(subset %in% c(NAyx, NAz))]
+    }
+
+    return(.ctree_test_1d(data = data, j = j, Y = Y, subset = subsetNArm, 
+                          weights = weights, SPLITONLY = SPLITONLY, ctrl = ctrl))
+}
+
+.ctree_split <- function(model, trafo, data, subset, weights, j, SPLITONLY = TRUE, ctrl)
+    .ctree_test(model, trafo, data, subset, weights, j, SPLITONLY, ctrl)
+
+.partysplit <- function(varid, breaks = NULL, index = NULL, right = TRUE, 
+                        prob = NULL, info = NULL) {
+    ret <- list(varid = varid, breaks = breaks, index = index, right = right, 
+                prob = prob, info = info)
+    class(ret) <- "partysplit"
+    ret
+}
+
+.ctree_test_1d <- function(data, j, Y, subset, weights, SPLITONLY = FALSE, ctrl) {
+
+    x <- data[[j]]
+    MIA <- FALSE
+    if (ctrl$MIA) MIA <- (length(NAs <- data[[j, type = "missing"]]) > 0)
+
+    ### X for (ordered) factors is always dummy matrix
+    if (is.factor(x) || is.ordered(x))
+        X <- data[[j, type = "index"]]
+
+    scores <- data[[j, type = "scores"]]
+    ORDERED <- is.ordered(x) || is.numeric(x)
+
[…]

### 2017-11-15 pkg/devel/partykit/R/extree.R

@@ -1,0 +1,894 @@
+
+### unbiased recursive partitioning: set up new node
+.extree_node <- function
+(
+    id = 1L, 			### id of this node
+    data, 			### full data, readonly
+    trafo,
+    selectfun, 			### variable selection
+    splitfun,                   ### split selection
+    svselectfun,                ### same for surrogate splits
+    svsplitfun,                 ### same for surrogate splits
+    partyvars, 			### partytioning variables
+                                ### a subset of 1:ncol(model.frame(data))
+    weights = integer(0L),	### optional case weights
+    subset, 			### subset of 1:nrow(data)
+                                ### for identifying obs for this node
+    ctrl, 			### extree_control()
+    info = NULL,
+    cenv = NULL			### environment for depth and maxid
+) {
+
+    ### depth keeps track of the depth of the tree
+    ### which has to be < than maxdepth
+    ### maxit is the largest id in the left subtree
+    if (is.null(cenv)) {
+        cenv <- new.env()
+        assign("depth", 0L, envir = cenv)
+    }
+    depth <- get("depth", envir = cenv)
+    assign("maxid", id, envir = cenv)
+    if (depth >= ctrl$maxdepth)
+        return(partynode(as.integer(id)))
+
+    ### check for stumps
+    if (id > 1L && ctrl$stump) 
+        return(partynode(as.integer(id)))
+
+    ### sw is basically the number of observations
+    ### which has to be > minsplit in order to consider
+    ### the node for splitting
+    if (length(weights) > 0L) {
+        if (ctrl$caseweights) {
+            sw <- sum(weights[subset]) 
+        } else {
+            sw <- sum(weights[subset] > 0L)
+        }
+    } else {
+        sw <- length(subset)
+    }
+    if (sw < ctrl$minsplit) 
+        return(partynode(as.integer(id)))
+
+    svars <- which(partyvars > 0)
+    if (ctrl$mtry < Inf) {
+        mtry <- min(sum(partyvars > 0), ctrl$mtry)
+        svars <- .resample(partyvars, mtry, prob = partyvars)
+    } 
+
+    thismodel <- trafo(subset = subset, weights = weights, info = info, 
+                       estfun = TRUE, object = TRUE)
+    if (is.null(thismodel))
+        return(partynode(as.integer(id)))
+
+    ### update sample size constraints on possible splits
[…]

### 2017-11-15 pkg/devel/partykit/R/mob-methods_R2.R

@@ -1,0 +1,446 @@
+
+## methods concerning call/formula/terms/etc.
+## (default methods work for terms and update)
+
+formula.modelparty <- function(x, extended = FALSE, ...)
+  if(extended) x$info$Formula else x$info$formula
+
+getCall.modelparty <- function(x, ...) x$info$call
+
+### temporarily moved to modelparty_R2.R
+if (FALSE) {
+model.frame.modelparty <- function(formula, ...)
+{
+  ### formula$data is get_all_vars, not model.frame
+  mf <- formula$data
+  ### always compute model.frame
+  if(nrow(mf) > 0L) 
+      return(model.frame(formula$info$Formula, data = mf))
+
+  dots <- list(...)
+  nargs <- dots[match(c("data", "na.action", "subset"), names(dots), 0L)]
+  mf <- formula$info$call
+  mf <- mf[c(1L, match(c("formula", "data", "subset", "na.action"), names(mf), 0L))]
+  mf$drop.unused.levels <- TRUE
+  mf[[1L]] <- quote(stats::model.frame)
+  mf[names(nargs)] <- nargs
+  if(is.null(env <- environment(formula$info$terms))) env <- parent.frame()
+  mf$formula <- Formula::Formula(as.formula(mf$formula))
+  eval(mf, env)
+}
+}
+
+weights.modelparty <- function(object, ...) {
+  fit <- object$fitted
+  ww <- if(!is.null(w <- fit[["(weights)"]])) w else rep.int(1L, NROW(fit))
+  structure(ww, .Names = rownames(fit))
+}
+
+## methods concerning model/parameters/loglik/etc.
+
+coef.modelparty <- function(object, node = NULL, drop = TRUE, ...) {
+  if(is.null(node)) node <- nodeids(object, terminal = TRUE)
+  cf <- do.call("rbind", nodeapply(object, ids = node, FUN = function(n) info_node(n)$coefficients))
+  if(drop) drop(cf) else cf
+}
+
+refit.modelparty <- function(object, node = NULL, drop = TRUE, ...)
+{
+  ## by default use all ids
+  if(is.null(node)) node <- nodeids(object)
+  
+  ## estimated coefficients
+  cf <- nodeapply(object, ids = node, FUN = function(n) info_node(n)$coefficients)
+  
+  ## model.frame
+  mf <- model.frame(object)
+  weights <- weights(object)
+  offset <- model.offset(mf)
[…]

### 2017-11-15 pkg/devel/partykit/R/modelparty_R2.R

@@ -1,0 +1,445 @@
+## variable selection: given model scores, conduct
+## all M-fluctuation tests for orderins in z
+.mfluc_test <- function(model, trafo, data, subset, weights, j, SPLITONLY, ctrl)
+{
+
+  stopifnot(!SPLITONLY)
+  stopifnot(is.null(data[["yx", type = "index"]]))
+  z <- data[[j]][subset]
+  estfun <- model$estfun[subset,,drop = FALSE]
+  cluster <- data[["(cluster)"]][subset]
+  if(length(weights) == 0) {
+      weights <- rep(1, NROW(estfun))
+  } else {
+      weights <- weights[subset]
+  }
+  obj <- model$obj
+
+  if(length(unique(z)) < 2L) return(list(statistic = NA, p.value = NA))
+  
+  ## set up return values
+  m <- NCOL(z)
+  pval <- rep.int(NA_real_, m)
+  stat <- rep.int(0, m)
+  ifac <- rep.int(FALSE, m)
+  
+  ## estimating functions (dropping zero weight observations)
+  process <- as.matrix(estfun)
+  ww0 <- (weights > 0)
+  process <- process[ww0, , drop = FALSE]
+  if(!is.null(cluster)) cluster <- droplevels(cluster[ww0]) 
+  weights <- weights[ww0]
+  z <- z[ww0]
+  k <- NCOL(process)
+  n <- NROW(process)
+  nobs <- if(ctrl$caseweights && any(weights != 1L)) sum(weights) else n
+
+  
+  ## stop if all collumn values in process are the same
+  if(all(apply(process, 2, function(x) length(unique(x))) == 1)) 
+    return(list(statistic = NA, p.value = NA))
+  
+  ## scale process
+  process <- process/sqrt(nobs)
+  vcov <- ctrl$vcov
+  if(is.null(obj)) vcov <- "opg"
+  if(vcov != "opg") {
+    bread <- vcov(obj) * nobs
+  }
+  if(vcov != "info") {
+    meat <- if(is.null(cluster)) {
+      crossprod(process/sqrt(weights))
+    } else {
+      crossprod(as.matrix(apply(process/sqrt(weights), 2L, tapply, cluster, sum)))
+    }
+  }
+  ## from strucchange
+  root.matrix <- function(X) {
+    if((ncol(X) == 1L)&&(nrow(X) == 1L)) return(sqrt(X)) else {
+      X.eigen <- eigen(X, symmetric = TRUE)
[…]

### 2017-11-15 pkg/devel/partykit/R/party.R

@@ -1,0 +1,682 @@
+## FIXME: data in party
+##   - currently assumed to be a data.frame
+##   - potentially empty
+##   - the following are all assumed to work:
+##     dim(data), names(data)
+##     sapply(data, class), lapply(data, levels)
+##   - potentially these need to be modified if data/terms
+##     should be able to deal with data bases
+
+party <- function(node, data, fitted = NULL, terms = NULL, names = NULL, info = NULL) {
+
+    stopifnot(inherits(node, "partynode"))
+    stopifnot(inherits(data, "data.frame"))
+    ### make sure all split variables are there 
+    ids <- nodeids(node)[!nodeids(node) %in% nodeids(node, terminal = TRUE)]
+    varids <- unique(unlist(nodeapply(node, ids = ids, FUN = function(x) 
+        varid_split(split_node(x)))))
+    stopifnot(varids %in% 1:ncol(data))
+    
+    if(!is.null(fitted)) {
+        stopifnot(inherits(fitted, "data.frame"))
+        stopifnot(nrow(data) == 0L | nrow(data) == nrow(fitted))
+        
+	# try to provide default variable "(fitted)"
+	if(nrow(data) > 0L) {
+          if(!("(fitted)" %in% names(fitted))) 
+            fitted[["(fitted)"]] <- fitted_node(node, data = data)
+	} else {
+	  stopifnot("(fitted)" == names(fitted)[1L])
+	}
+
+        nt <- nodeids(node, terminal = TRUE)
+        stopifnot(all(fitted[["(fitted)"]] %in% nt))
+
+        node <- as.partynode(node, from = 1L)
+        nt2 <- nodeids(node, terminal = TRUE)
+        fitted[["(fitted)"]] <- nt2[match(fitted[["(fitted)"]], nt)]
+    } else {
+        node <- as.partynode(node, from = 1L)
+	# default "(fitted)"
+	if(nrow(data) > 0L & missing(fitted)) 
+          fitted <- data.frame("(fitted)" = fitted_node(node, 
+            data = data), check.names = FALSE)
+    }
+    
+    party <- list(node = node, data = data, fitted = fitted, 
+                  terms = NULL, names = NULL, info = info)
+    class(party) <- "party"
+
+    if(!is.null(terms)) {
+        stopifnot(inherits(terms, "terms"))
+	party$terms <- terms
+    }
+
+    if (!is.null(names)) {
+        n <- length(nodeids(party, terminal = FALSE))
+        if (length(names) != n)
[…]

### 2017-11-15 pkg/devel/partykit/R/plot.R

@@ -1,0 +1,1006 @@
+.nobs_party <- function(party, id = 1L) {
+  dat <- data_party(party, id = id)
+  if("(weights)" %in% names(dat)) sum(dat[["(weights)"]]) else NROW(dat)
+}
+
+node_inner <- function(obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", gp = gpar())
+{
+  meta <- obj$data
+  nam <- names(obj)
+
+  extract_label <- function(node) {
+    if(is.terminal(node)) return(rep.int("", 2L))
+
+    varlab <- character_split(split_node(node), meta)$name
+    if(abbreviate > 0L) varlab <- abbreviate(varlab, as.integer(abbreviate))
+
+    ## FIXME: make more flexible rather than special-casing p-value
+    if(pval) {
+      nullna <- function(x) is.null(x) || is.na(x)
+      pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), silent = TRUE))
+      pval <- if(inherits(pval, "try-error")) FALSE else pval
+    }
+    if(pval) {
+      pvalue <- node$info$p.value
+      plab <- ifelse(pvalue < 10^(-3L),
+        paste("p <", 10^(-3L)),
+    	paste("p =", round(pvalue, digits = 3L)))
+    } else {
+      plab <- ""
+    }
+    return(c(varlab, plab))
+  }
+
+  maxstr <- function(node) {
+      lab <- extract_label(node)
+      klab <- if(is.terminal(node)) "" else unlist(lapply(kids_node(node), maxstr))
+      lab <- c(lab, klab)
+      lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
+      lab <- lab[which.max(nchar(lab))]
+      if(length(lab) < 1L) lab <- ""
+      return(lab)
+  }
+
+  nstr <- maxstr(node_party(obj))
+  if(nchar(nstr) < 6) nstr <- "aAAAAa"
+
+  ### panel function for the inner nodes
+  rval <- function(node) {  
+    node_vp <- viewport(
+      x = unit(0.5, "npc"),
+      y = unit(0.5, "npc"),
+      width = unit(1, "strwidth", nstr) * 1.3, 
+      height = unit(3, "lines"),
+      name = paste("node_inner", id_node(node), sep = ""),
+      gp = gp
+    )
+    pushViewport(node_vp)
+
[…]

### 2017-11-15 pkg/devel/partykit/inst/extree/extree.R

@@ -1,0 +1,888 @@
+
+### unbiased recursive partitioning: set up new node
+.extree_node <- function
+(
+    id = 1L, 			### id of this node
+    data, 			### full data, readonly
+    trafo,
+    selectfun, 			### variable selection
+    splitfun,                   ### split selection
+    svselectfun,                ### same for surrogate splits
+    svsplitfun,                 ### same for surrogate splits
+    partyvars, 			### partytioning variables
+                                ### a subset of 1:ncol(model.frame(data))
+    weights = integer(0L),	### optional case weights
+    subset, 			### subset of 1:nrow(data)
+                                ### for identifying obs for this node
+    ctrl, 			### extree_control()
+    info = NULL,
+    cenv = NULL			### environment for depth and maxid
+) {
+
+    ### depth keeps track of the depth of the tree
+    ### which has to be < than maxdepth
+    ### maxit is the largest id in the left subtree
+    if (is.null(cenv)) {
+        cenv <- new.env()
+        assign("depth", 0L, envir = cenv)
+    }
+    depth <- get("depth", envir = cenv)
+    assign("maxid", id, envir = cenv)
+    if (depth >= ctrl$maxdepth)
+        return(partynode(as.integer(id)))
+
+    ### check for stumps
+    if (id > 1L && ctrl$stump) 
+        return(partynode(as.integer(id)))
+
+    ### sw is basically the number of observations
+    ### which has to be > minsplit in order to consider
+    ### the node for splitting
+    if (length(weights) > 0L) {
+        if (ctrl$caseweights) {
+            sw <- sum(weights[subset]) 
+        } else {
+            sw <- sum(weights[subset] > 0L)
+        }
+    } else {
+        sw <- length(subset)
+    }
+    if (sw < ctrl$minsplit) 
+        return(partynode(as.integer(id)))
+
+    svars <- which(partyvars > 0)
+    if (ctrl$mtry < Inf) {
+        mtry <- min(sum(partyvars > 0), ctrl$mtry)
+        svars <- .resample(partyvars, mtry, prob = partyvars)
+    } 
+
+    thismodel <- trafo(subset = subset, weights = weights, info = info, 
+                       estfun = TRUE, object = TRUE)
+    if (is.null(thismodel))
+        return(partynode(as.integer(id)))
+
+    ### update sample size constraints on possible splits
[…]

### 2017-11-15 pkg/devel/partykit/inst/extree/modelparty.R

@@ -1,0 +1,423 @@
+## variable selection: given model scores, conduct
+## all M-fluctuation tests for orderins in z
+.mfluc_test <- function(model, trafo, data, subset, weights, j, SPLITONLY, ctrl)
+{
+
+  stopifnot(!SPLITONLY)
+  stopifnot(is.null(data[["yx", type = "index"]]))
+  z <- data[[j]][subset]
+  estfun <- model$estfun[subset,,drop = FALSE]
+  cluster <- data[["(cluster)"]][subset]
+  if(length(weights) == 0) {
+      weights <- rep(1, NROW(estfun))
+  } else {
+      weights <- weights[subset]
+  }
+  obj <- model$obj
+
+  if(length(unique(z)) < 2L) return(list(statistic = NA, p.value = NA))
+  
+  ## set up return values
+  m <- NCOL(z)
+  pval <- rep.int(NA_real_, m)
+  stat <- rep.int(0, m)
+  ifac <- rep.int(FALSE, m)
+  
+  ## estimating functions (dropping zero weight observations)
+  process <- as.matrix(estfun)
+  ww0 <- (weights > 0)
+  process <- process[ww0, , drop = FALSE]
+  if(!is.null(cluster)) cluster <- droplevels(cluster[ww0]) 
+  weights <- weights[ww0]
+  z <- z[ww0]
+  k <- NCOL(process)
+  n <- NROW(process)
+  nobs <- if(ctrl$caseweights && any(weights != 1L)) sum(weights) else n
+
+  
+  ## stop if all collumn values in process are the same
+  if(all(apply(process, 2, function(x) length(unique(x))) == 1)) 
+    return(list(statistic = NA, p.value = NA))
+  
+  ## scale process
+  process <- process/sqrt(nobs)
+  vcov <- ctrl$vcov
+  if(is.null(obj)) vcov <- "opg"
+  if(vcov != "opg") {
+    bread <- vcov(obj) * nobs
+  }
+  if(vcov != "info") {
+    meat <- if(is.null(cluster)) {
+      crossprod(process/sqrt(weights))
+    } else {
+      crossprod(as.matrix(apply(process/sqrt(weights), 2L, tapply, cluster, sum)))
+    }
+  }
+  ## from strucchange
+  root.matrix <- function(X) {
+    if((ncol(X) == 1L)&&(nrow(X) == 1L)) return(sqrt(X)) else {
+      X.eigen <- eigen(X, symmetric = TRUE)
[…]

### 2017-11-15 pkg/devel/partykit/inst/extree/tmp.R

@@ -1,0 +1,12 @@
+
+library("profmem")
+
+d <- iris
+profmem(class(d) <- "list")
+
+profmem(x <- d[["Sepal.Width"]])
+
+profmem(y <- as.double(x))
+
+
+

### 2017-11-15 pkg/devel/partykit/src/ctreesplit.c

@@ -1,0 +1,203 @@
+
+#include "ctree.h"
+
+void C_split_numeric (SEXP x, SEXP y, int *iweights, int minbucket, 
+                      double *expinf, double *covinf, double *breaks) {
+
+     int *orderx, i, j, k, n, q, sw = 0, count;
+     double dsw = 0.0, *dx, lastx, thisx, xmax;
+     double cstat = 0.0, laststat = 0.0;
+     double *linstat, *explinstat, *covlinstat, *dy;
+
+     n = NROW(x);
+     if (NCOL(x) != 1 || !isReal(x))
+         error("");
+
+     /* orderx = order(x) */
+     orderx = Calloc(n, int);
+     for (i = 0; i < n; i++) orderx[i] = i;
+     dx = Calloc(n, double);
+     for (i = 0; i < n; i++) dx[i] = REAL(x)[i];
+     rsort_with_index(dx, orderx, n);
+     Free(dx);
+     dx = REAL(x);
+     xmax = dx[orderx[n-1]];
+
+     q = NCOL(y);
+     dy = REAL(y); 
+     
+     linstat = Calloc(q, double);
+     explinstat = Calloc(q, double);
+     covlinstat = Calloc(q * q, double);
+     
+     for (j = 0; j < q; j++) {
+         linstat[j] = 0.0;
+         explinstat[j] = 0.0;
+         for (k = 0; k < q; k++) 
+             covlinstat[j * q + k] = 0.0;
+     }
+
+     dsw = 0;
+     lastx = NA_REAL;
+     breaks[0] = NA_REAL;
+     
+     for (i = 0; i < n; i++) {
+         if (iweights[i] == 0 || ISNA(dx[i])) continue;
+         sw += iweights[i];
+     }
+
+     C_ExpInf(dy, q, iweights, sw, n, expinf);
+     C_CovInf(dy, q, iweights, sw, n, expinf, covinf);
+     
+      /* check sample size contraint */
+     if (sw >= minbucket) {
+     for (i = 0; i < (n - 1); i++) {
+
+         /* value of the orderx[i]th largest observation */
+         thisx = dx[orderx[i]];
+         
+         /* nothing do to for zero weights or NAs */
+         if (iweights[orderx[i]] == 0 || ISNA(thisx)) continue;
+
+         /* sum of the weights of all x <= thisx */
+         dsw += (double) iweights[orderx[i]];
+
+         /* linear statistic for y ~ (x <= thisx) */
[…]

### 2017-11-15 pkg/devel/partykit/tests/ctree-ctree.R

@@ -1,0 +1,95 @@
+
+### compare partykit::ctree and party::ctree wrt predictions
+### and computing time. But not during R CMD check
+
+library("partykit")
+library("party")
+
+set.seed(29)
+
+n <- 200
+dgp <- function(n, response = c("numeric", "ordered", "binary", "factor"), na = FALSE) {
+
+    response <- match.arg(response)
+    x <- runif(n * 10)
+    x <- matrix(x, ncol = 10)
+    x <- cbind(x, x[,1:2] + cbind(rnorm(n, sd = 0.1), rnorm(n, sd = 0.1)))
+    x1 <- sample(gl(4, n / 4))
+    x2 <- sample(ordered(gl(4, n / 4)))
+    y <- as.integer(interaction(x[,1] > 0.5, x[,2] > 0.25))
+    y <- rnorm(n, mean = y)
+    if (na) x[sample(1:(n * 10), n)] <- NA
+    if (response == "numeric")
+        return(data.frame(x, x1, x2, y = y))
+    if (response == "binary") {
+        y <- factor(y < median(y))
+        return(data.frame(x, x1, x2, y = y))
+    } else {
+        y <- cut(y, 
+            breaks = c(-Inf, quantile(y, c(0.25, 0.5, 0.75)), Inf))
+    }
+    if (response == "ordered")
+        return(data.frame(x, x1, x2, y = ordered(y)))
+    if (response == "factor")
+        return(data.frame(x, x1, x2, y = y))
+}
+
+args <- list(response = c("numeric", "ordered", "binary", "factor"),
+             na = c(FALSE, TRUE),
+             teststat = c("quad", "max"),
+             testtype = c("Bonferroni", "Univariate", "Teststatistic"),
+             mincriterion = c(0.8, 0.9, 0.95),
+             maxsurrogate = 0:2, maxdepth = 3)
+
+gr <- do.call("expand.grid", args)
+tgr <- gr
+gr[["response"]] <- as.character(gr[["response"]])
+gr[["teststat"]] <- as.character(gr[["teststat"]])
+gr[["testtype"]] <- as.character(gr[["testtype"]])
+
+### random splits are done differently in party::ctree
+gr <- subset(gr, !na | (na & maxsurrogate > 0))
+
+
+fun <- function(args) {
+
+    learn <- dgp(n, response = args$response, na = args$na)
[…]

### 2017-11-15 pkg/devel/partykit/tests/regtest-glmtree_R2.R

@@ -1,0 +1,197 @@
+
+library("partykit")
+library("sandwich")
+
+set.seed(29)
+n <- 1000
+x <- runif(n)
+z <- runif(n)
+y <- rnorm(n, mean = x * c(-1, 1)[(z > 0.7) + 1], sd = 3)
+z_noise <- factor(sample(1:3, size = n, replace = TRUE))
+d <- data.frame(y = y, x = x, z = z, z_noise = z_noise)
+
+
+fmla <- as.formula("y ~ x | z + z_noise")
+fmly <- gaussian()
+fit <- partykit:::glmfit
+
+# versions of the data
+d1 <- d
+d1$z <- signif(d1$z, digits = 1)
+
+k <- 20
+zs_noise <- matrix(rnorm(n*k), nrow = n)
+colnames(zs_noise) <- paste0("z_noise_", 1:k)
+d2 <- cbind(d, zs_noise)
+fmla2 <- as.formula(paste("y ~ x | z + z_noise +",
+                          paste0("z_noise_", 1:k, collapse = " + ")))
+
+
+d3 <- d2
+d3$z <- factor(sample(1:3, size = n, replace = TRUE, prob = c(0.1, 0.5, 0.4)))
+d3$y <- rnorm(n, mean = x * c(-1, 1)[(d3$z == 2) + 1], sd = 3)
+
+## check weights
+w <- rep(1, n)
+w[1:10] <- 2
+(mw1 <- glmtree(formula = fmla, data = d, weights = w))
+(mw2 <- glmtree(formula = fmla, data = d, weights = w, caseweights = FALSE))
+
+
+
+## check dfsplit
+(mmfluc2 <- mob(formula = fmla, data = d, fit = partykit:::glmfit))
+(mmfluc3 <- glmtree(formula = fmla, data = d))
+(mmfluc3_dfsplit <- glmtree(formula = fmla, data = d, dfsplit = 10))
+
+
+## check tests
+library("strucchange")
+sctest(mmfluc3, node = 1) # does not yet work
+
+x <- mmfluc3
+(tst3 <- nodeapply(x, ids = nodeids(x), function(n) n$info$criterion))
+
+
+
+
+## check logLik and AIC
+logLik(mmfluc2)
+logLik(mmfluc3)
+logLik(mmfluc3_dfsplit)
+logLik(glm(y ~ x, data = d))
+
+AIC(mmfluc3)
+AIC(mmfluc3_dfsplit)
+
+
+
+## check inner and terminal
+options <- list(NULL,
[…]

### 2017-11-15 pkg/devel/partykit/tests/regtest-lmtree_R2.R

@@ -1,0 +1,56 @@
+library("partykit")
+
+set.seed(29)
+n <- 1000
+x <- runif(n)
+z <- runif(n)
+y <- rnorm(n, mean = x * c(-1, 1)[(z > 0.7) + 1], sd = 3)
+z_noise <- factor(sample(1:3, size = n, replace = TRUE))
+d <- data.frame(y = y, x = x, z = z, z_noise = z_noise)
+
+
+fmla <- as.formula("y ~ x | z + z_noise")
+
+(m_mob <- mob(formula = fmla, data = d, fit = partykit:::lmfit))
+(m_lm2 <- lmtree(formula = fmla, data = d))
+
+
+mods <- nodeapply(m_lm2, ids = nodeids(m_lm2, terminal = TRUE), function(x) x$info$object)
+sum(sapply(mods, function(x) sum(x$residuals^2)))
+
+## glm versus lm / logLik versus sum of squared residuals
+lm_big <- lmtree(formula = fmla, data = d, testflavour = "exhaustive")
+glm_big <- glmtree(formula = fmla, data = d, testflavour = "exhaustive")
+
+width(lm_big)
+width(glm_big)
+
+AIC(lm_big)
+AIC(glm_big)
+# AIC of lmtree must be worse (greater) or equal than of glmtree because AIC assesses the logLik,
+# which is optimized in glm but noth in lm (there it is the sum of squared residuals).
+# In splitting the different objective functions may lead to differences, because the logLik
+# contains the term log(sum of squared residuals), i.e. we optimize
+# log(sum of squared residuals in left node) + log(sum of squared residuals in right node);
+# In lmtree we optimize
+# sum of squared residuals in left node + sum of squared residuals in right node
+
+
+## pruning
+lm_aic <- prune.modelparty(lm_big, type = "AIC")
+lm_bic <- prune.modelparty(lm_big, type = "BIC")
+
+width(lm_big)
+width(lm_aic)
+width(lm_bic)
+
+glm_aic <- prune.modelparty(glm_big, type = "AIC")
+glm_bic <- prune.modelparty(glm_big, type = "BIC")
+
+width(glm_big)
+width(glm_aic)
+width(glm_bic)
+
+
+
+

### 2017-11-15 pkg/devel/partykit/vignettes/mob.R

@@ -1,0 +1,683 @@
+### R code from vignette source 'mob.Rnw'
+
+###################################################
+### code chunk number 1: setup
+###################################################
+library("partykit")
+options(prompt = "R> ", continue = "+  ", digits = 4, useFancyQuotes = FALSE)
+
+
+###################################################
+### code chunk number 2: PimaIndiansDiabetes
+###################################################
+data("PimaIndiansDiabetes", package = "mlbench")
+
+
+###################################################
+### code chunk number 3: PimIndiansDiabetes-formula
+###################################################
+pid_formula <- diabetes ~ glucose | pregnant + pressure + triceps +
+  insulin + mass + pedigree + age
+
+
+###################################################
+### code chunk number 4: logit
+###################################################
+logit <- function(y, x, start = NULL, weights = NULL, offset = NULL, ...) {
+  glm(y ~ 0 + x, family = binomial, start = start, ...)
+}
+
+
+###################################################
+### code chunk number 5: PimaIndiansDiabetes-mob
+###################################################
+pid_tree <- mob(pid_formula, data = PimaIndiansDiabetes, fit = logit)
+
+
+###################################################
+### code chunk number 6: PimaIndiansDiabetes-print
+###################################################
+pid_tree
+
+
+###################################################
+### code chunk number 7: PimaIndiansDiabetes-glmtree
+###################################################
+pid_tree2 <- glmtree(diabetes ~ glucose | pregnant +
+  pressure + triceps + insulin + mass + pedigree + age,
+  data = PimaIndiansDiabetes, family = binomial)
+
+
+###################################################
+### code chunk number 8: PimaIndiansDiabetes-plot
+###################################################
+plot(pid_tree)
+
+
+###################################################
+### code chunk number 9: PimaIndiansDiabetes-plot2
+###################################################
+plot(pid_tree2, tp_args = list(ylines = 1, margins = c(1.5, 1.5, 1.5, 2.5)))
+
+
+###################################################
+### code chunk number 10: PimaIndiansDiabetes-sctest1
+###################################################
+library("strucchange")
+sctest(pid_tree, node = 1)
+
+
+###################################################
+### code chunk number 11: PimaIndiansDiabetes-sctest2
+###################################################
+sctest(pid_tree, node = 2)
+
+
+###################################################
+### code chunk number 12: PimaIndiansDiabetes-sctest3
[…]

### 2017-11-15 pkg/disttree/inst/draft/draft.Rnw
edited plots, small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1709 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 46dd1abe8808febd4c056e4e8fe58ed62a22c2b4
changed setup for predictions, now learn data set consists of 24 years and predictions are made for 4 years
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1714 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a45939a53927c461a631b2b5fabc3a24da2ef2c5

@@
 In a first setting only one observation station is taken into consideration.
-The models are trained on 25 years and prbabilistic predictions of the consecutive 3 years are validated.
+The models are trained on 24 years and prbabilistic predictions of the consecutive 4 years are validated.
 A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}.
-For the 24th of July of 2010, 2011 and 2012 the density functions of the distributions estimated by a forest model are plotted together with the corresponding observations.
+For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a forest model are plotted together with the corresponding observations.
 This illustrates that well fitting predicted expected values are provided by this model as they are close to the observed values.
@@
 
-  save(res, file = "rain_Axams_pred.rda")
+  save(res, file = "rain_Axams_pred_24to4.rda")
 }
@@
 }
+
+## Variable importance
+if(file.exists("vimp_crps.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
+  load("vimp_crps.rda")
+} else {
+
+  # locally redefine logLik.distforest to use crps in varimp()
+  logLik.distforest <- function(object, newdata = NULL, ...) {
+    if(is.null(newdata)) {
+      newdata <- object$data
+    }
+
+    # predict parameter
+    pdf <- predict(object, newdata = newdata, type = "parameter")
+    df_mu <- pdf$mu
+    df_sigma <- pdf$sigma
+
+    # calculate CRPS
+    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
+
+    return(structure(crps, df = NA, class = "logLik"))
+  }
+
+  vimp_crps <- varimp(res$df, nperm = 1L)
+  #vimp10_crps <- varimp(res$df, nperm = 10L)
+  #vimp100_crps <- varimp(res$df, nperm = 100L)
+  #vimp1000_crps <- varimp(res$df, nperm = 1000L)
+
+  save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
+  #save(vimp10_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp10_crps.rda")
+  #save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
+  #save(vimp1000_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp1000_crps.rda")
+
+  rm(logLik.distforest) 
+}
+
 @
@@
      main = paste0("July ", pday, dayending), ylab = "Density", 
-     #xlab = expression(Total~precipitation~group("[", list(mm^(1/1.6), plain(per 24h)),"]")),
-     xlab = expression(Total~precipitation~group("[", mm^(1/1.6)~per~day,"]")),
+     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
+     #xlab = expression(Total~precipitation~group("[", mm^(1/1.6)~per~day,"]")),
      #{\frac{1}{1.6}}$,
@@
 legend('topright', c("predicted distribution", "point mass at censoring point", "observation"),
-       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 1)
+       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
 
@@
 @
-\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2010, 2011 and 2012.
+\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2009, 2010, 2011 and 2012.
 Observations and predictions are left censored at 0.
@@
 
-\subsection{Motivation (Review/Problems)}
+\subsection{Motivation}
 As illustrated in Figure~\ref{Devel_parmod} a lot of development has been made in the field of parametric modeling and a wide range of tools can be applied in research areas such as probabilistic forecasting.
@@
     & \text{if } X_1 < 0.4 \text{ and } X_4 = 1\\
-    10 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{10}{2},       
+    13 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
     & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 1
@@
 $$
-\sigma(\mu, X_1,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.7)
+\sigma(\mu, X_1,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.4)
 $$
@@
 
-To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to 1 and then to a second data set where $\kappa$ is set to 10.
-In Figure~\ref{plot_sim_oneset_mu} the location parameter function $\mu(X_1, X_4, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $x_4 = 1$ and $(x_2, x_3, x_5, x_6) = (0,0,0,0)$.
+To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to~1 and then to a second data set where $\kappa$ is set to 10.
+In Figure~\ref{plot_sim_oneset_mu} the location parameter function $\mu(X_1, X_4, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $x_4 = 0$ and $(x_2, x_3, x_5, x_6) = (0,0,0,0)$.
 \\
@@
   #define parameter function
-  kappa <- 1
+  #kappa <- 1
   fun1 <- function(x){
     mu <- 0.2 + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
-      (3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
+      (7 * (exp(-(3*x[,1]-1.2)^(2*1)))) * as.numeric(x[,1] < 0.4) + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*1))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
+      (3 * (exp(-(3*x[,1]-1.2)^(2*1)))) * x[,4]
     sigma <- 0.5 + 
@@
       2*(x[,1] < -0.4) 
-      #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
-      #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
+      #plogis((1^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
+      #4*(1-plogis((1^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
       #5 * abs(x[,1]) * x[,4]
@@
                                  pred_fix_x2 = 0.9,
-                                 pred_fix_x4 = 1,
+                                 pred_fix_x4 = 0,
                                  nrunif = 3,
@@
   #define parameter function
-  kappa <- 10
+  #kappa <- 10
   fun10 <- function(x){
     mu <- 0.2 + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
-      (3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
+      (7 * (exp(-(3*x[,1]-1.2)^(2*10)))) * as.numeric(x[,1] < 0.4) + 
+      (7 * (exp(-(3*x[,1]-1.2)^(2*10))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
+      (3 * (exp(-(3*x[,1]-1.2)^(2*10)))) * x[,4]
     sigma <- 0.5 + 
@@
       2*(x[,1] < -0.4) 
-      #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
-      #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
+      #plogis((10^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
+      #4*(1-plogis((10^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
       #5 * abs(x[,1]) * x[,4]
@@
   simres <- gensim(parfun = simfun,
-                   seedconst = 7, nrep = 100, ntree = 100,
+                   seedconst = 74, nrep = 100, ntree = 100,
                    nsteps = 10, stepsize = 1,
@@
 @
-\caption{\label{plot_sim_crps}Results of the simulation study: CRPS} %\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest} vs. %\code{bamlss} vs. \code{gamboostLSS}\\
+\caption{\label{plot_sim_crps}CRPS of the three models evaluated on various data sets over increasing $\kappa$} %\code{disttree} vs.
+\code{distforest} vs.
+\code{gamlss} vs.
+\code{randomForest} vs.
+%\code{bamlss} vs.
+\code{gamboostLSS}\\
 %right: levelplot of the parameter function for $\kappa=5$}
@@
 \subsection{Setting}
-The data set used for training and validation consists of observed daily
-precipitation sums provided by the national hydrographical service
-(\citealt{ehyd}) and numerical weather forecasts from the {U.S.} national
-oceanic and atmospheric administration (NOAA).
-Both, observations and weather forecasts are available from 1985 trough the
-year 2012, focussing on July only in this article.\\
-Observations are available on a daily basis, each observation representing the
-total precipitation sum fallen within 24~hours observed at 6~UTC at \textbf{95}
-stations all over Tyrol and it's close surrounding. The observations are
-undergoing a rigorous quality check before made available to the public.  As
-covariates the second generation reforecast data set of the global ensemble
-forecast system (GEFS) is used (\citealt{hamill2013}).  This data set consists
-of an 11-member ensemble based on a fixed version of the numerical model and a
-horizontal grid-spacing of about $50~\times~50 km^2$ initialized at 0~UTC from
-Dezember 1984 to precent. Each of the 11 ensemble members uses slightly
-different perturbed initial conditions trying to predict the situation specific
-uncertenty of the atmospheric state.
+The data set used for training and validation consists of observed daily precipitation sums provided by the National Hydrographical Service (\citealt{ehyd}) and numerical weather forecasts from the {U.S.} National Oceanic and Atmospheric Administration (NOAA).
+Both, observations and weather forecasts are available from 1985 trough the year 2012, focussing on July only in this article.\\ 
+Observations are available on a daily basis, each observation representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and it's close surrounding.
+The observations are undergoing a rigorous quality check before made available to the public.
+As covariates the second generation reforecast data set of the global ensemble forecast system (GEFS) is used (\citealt{hamill2013}).
+This data set consists of an 11-member ensemble based on a fixed version of the numerical model and a horizontal grid-spacing of about $50~\times~50 \text{km}^2$ initialized at 0~UTC from Dezember 1984 to precent.
+Each of the 11 ensemble members uses slightly different perturbed initial conditions trying to predict the situation specific uncertenty of the atmospheric state.
 
@@
 In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates.
-Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble minimum and ensemble maximum.
-All of these values can be calculated on the sums over a whole day (6--30 UTC) or a 6h time window of the day.\\
+Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std dev, ensemble minimum and ensemble maximum.
+All of these values can be calculated on the sums over 24 hours (6--30 UTC) or a 6 hour time window.\\
 Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std dev of mean/minimal/maximal values respectively.\\
 Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
-To remove the skewness of ...
-(the distribution of variables representing precipitation amounts), a power transformation has been used frequently (Box and Cox 1964).
+To remove large parts of the skewness of precipitation data, a power transformation has been used frequently (Box and Cox 1964).
 In literature, cubic (\citealt{stidd1973}) or square root (\citealt{hutchinson1998b}) transformations have often been suggested but may vary for different climatic zones or temporal aggregation periods.
-The response variable \texttt{tp} (total precipitation) is power transformed by a factor of 1.6.
-This factor is chosen due to experience and knowledge about possible distributions of precipitation measurements.
+In \citealt{stauffer2017a} optimization methods have been applied to find the best transformation factor which resulted in a value of $\frac{1}{1.6}$.
+Since the data used for this optimization covers almost the same geographical region as considered in this article the same factor is chosen here for the power transformation of the response variable \texttt{tp} (total precipitation).
 The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
-The same transformation has been chosen for the variable \texttt{cape} (convective available potential energy).
+For simplicity the same transformation has been chosen for the variable \texttt{cape} (convective available potential energy).
 
@@
 \textbf{tp}: total precipitation            & 12 variations:  & ensemble mean of sums over 24h, \\
-\hspace*{\fill} power transformed (by 1.6) &    & ensemble std dev of sums over 24h, \\ 
+\hspace*{\fill} power transformed (by $\frac{1}{1.6}$) &    & ensemble std dev of sums over 24h, \\ 
 \textbf{cape}: convective                  &    & ensemble minimum of sums over 24h, \\
 \hspace*{0.4cm} available potential energy &    & ensemble maximum of sums over 24h\\
-\hspace*{\fill}  power transformed (by 1.6)&    & \qquad all for 6--30 UTC\\
+\hspace*{\fill}  power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
 \hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
@@
 \textbf{dswrf}: downwards short ware      & 6 variations: & ensemble mean of mean values, \\
-\hspace*{\fill} radiation flux (sunshine) &   & ensemble mean of minimal values,\\
+\hspace*{\fill} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
 \textbf{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
@@
 \subsection{Models}
-Next to distributional forests three other methods are applied to this data set and their results are compared based on (RMSE, log-likelihood values and) CRPS values.
-One of the other models is an EMOS models as this is a common choice in this type of setting / research field and therefore provides a valueable reference.
-The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameters (denoted as GAM (prespecified)) and once applying a boosting algorithm (GAM (boosted)).
+Next to distributional forests three other methods are applied to this data set and their results are compared based on CRPS values.
+One of the reference models is an EMOS models as this is a common choice in this type of setting / research field.
+The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameters and once applying a boosting algorithm).
 The chosen methods and their specifications are listed in Table~\ref{model_specification}.
@@
 \hline
-\textbf{GAM (prespecified)} & spline  & tp{\_}mean, & tppow{\_}sprd,\\
+\textbf{prespecified GAM} & spline  & tp{\_}mean, & tppow{\_}sprd,\\
                             & in each & tp{\_}max,  & dswrf{\_}sprd{\_}mean,\\
- & & tp{\_}mean1218 $\cdot$ & tp{\_}sprd1218 $\cdot$\\ 
+ & & tp{\_}mean1218 $\ast$ & tp{\_}sprd1218 $\ast$\\ 
  & & \quad cape{\_}mean1218, & \quad  cape{\_}mean1218,\\
@@
 \hline
-\textbf{GAM (boosted)} & spline  & all & all \\
+\textbf{boosted GAM} & spline  & all & all \\
                        & in each &     &     \\
@@
 Additionally, interactions are specified in advance, once between \texttt{tp{\_}mean1218} and \texttt{cape{\_}mean1218} for the location parameter and once between \texttt{tp{\_}sprd1218} and \texttt{cape{\_}mean1218} for the scale parameter.
+These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation.
 By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.\\
@@
 This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol.
-As for all other stations, daily observations of the month of July of 28 years are available, starting in 1985 and ending in 2012, both years included.\\
-The results of fitting one distributional forest to the first 25 years of the data set and predicting the total amount of precipitation for one day of each of the three successive years have been shown in the introduction in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on PIT histograms.\\
+As for all other stations, daily observations of the month of July of 28 years are available from 1985 through the year 2012.\\
+The results of fitting one distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast.
+Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on probability integral transform (PIT) histograms.\\
 \\
-As seen in Table~\ref{model_specification} the forest model includes all 82 covariates.
-In a single tree model it can easily be observed which covariates are used as split variables and hence have an impact on the response variable.
+As shown in Table~\ref{model_specification} the forest model includes all 82 covariates.
+A single tree model allows to easily analize which covariates are used as split variables in the nodes and hence have an impact on the response variable.
 In a forest model though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree.
-To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been chosen.
-The 10 leading covariates of the above described forest model for the station Axams learned on the first 25 year are listed in Figure~\ref{varimp}.
+To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied.
+The 10 leading covariates of the above described forest model for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}.
 As expected, the numerical forecast for total precipitation is the most frequently selected.
 In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.
-<<variabble_importance, echo=FALSE, eval=TRUE>>=
-if(file.exists("vimp_crps.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
-  load("vimp_crps.rda")
-} else {
-
-  # locally redefine logLik.distforest to use crps in varimp()
-  logLik.distforest <- function(object, newdata = NULL, ...) {
-    if(is.null(newdata)) {
-      newdata <- object$data
-    }
-
-    # predict parameter
-    pdf <- predict(object, newdata = newdata, type = "parameter")
-    df_mu <- pdf$mu
-    df_sigma <- pdf$sigma
-
-    # calculate CRPS
-    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
-
-    return(structure(crps, df = NA, class = "logLik"))
-  }
-
-  vimp_crps <- varimp(res$df, nperm = 1L)
-  #vimp10_crps <- varimp(res$df, nperm = 10L)
-  #vimp100_crps <- varimp(res$df, nperm = 100L)
-  #vimp1000_crps <- varimp(res$df, nperm = 1000L)
-
-  save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
-  #save(vimp10_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp10_crps.rda")
-  #save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
-  #save(vimp1000_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp1000_crps.rda")
-
-  rm(logLik.distforest) 
-}
-
-@
 
@@
 
-To get another impression of the quality of the predictions made by the tested methods a probability integral transform (PIT) histograms is plotted in Figure~\ref{pit_oosample} for out-of-sample predictions of each of the models.
-A perfectly fitting model would lead to a histogram with all bars being of height 1 as indicated by the red horizontal line.
-Overall, it can be stated that all four model provide well fitting models while the forest seems to be the one being closest to the ideal case.
+To get another impression of the quality of the predictions made by the tested methods a PIT histogram is plotted in Figure~\ref{pit_oosample} for out-of-sample predictions of each of the models. 
+A perfectly calibrated model would lead to a PIT histogram with a uniform distribution (a density of 1 for each bin as indicated by the red horizontal line).
+Overall, it can be stated that all four model provide well calibrated forecasts.
 
@@
 pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Forest")
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "GAM (prespecified)")
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "GAM (boosted)")
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM")
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM")
 pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS")
@@
 The data set over all 28 years is split into 7 subsets each consisting of 4 randomly selected years.
-While 6 of these subsets are used as learning data the 7th is the testing data for which predictions are made and evaluated.
-This is done by calculating (the RMSE, the log-likelihood value and) the CRPS value for each of the observations in the testing data and then averaging them.
+While 6 of these subsets are used as learning data the 7th is the testing data set for which predictions are made and evaluated.
+This is done by calculating the CRPS value for each of the observations in the testing data set and then averaging them.
 Each of the 7 parts is selected as testing data once which leads to 7 different settings over which, again, the average values are stored.
@@
 As we are interested in comparing the performance of the distributional forest to those of other models, we choose one model as a reference and calculated the skills score with respect to the CRPS.
-In that way we obtain a measure of how much a method improves the model fit in terms of CRPS values compared to the reference method.
+In that way we obtain a measure of how much a method improves the model performance in terms of CRPS values compared to the reference method.
 $$
@@
 $$
-The results regarding the CRPS values are illustrated by the boxplots in Figure~\ref{boxplots_crps_axams} where the skills score of the CRPS is the chosen measurement scale with the EMOS model as the reference model represented by the horizontal line at height 0.\\
-(The boxplots of the RMSE and the log-likelihood values can be found in the appendix.)
+The resulting skills scores are illustrated in Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model represented by the horizontal line at height 0.
 <<echo=FALSE, results=hide>>=
 #### cross validation rain
-if(file.exists("rain_Axams.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams.rda")
-  load("rain_Axams.rda")
+if(file.exists("rain_Axams_7x10.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_7x10.rda")
+  load("rain_Axams_7x10.rda")
 } else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross.R")
-  source("rain_cross.R")
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross_7x10.R")
+  source("rain_cross_7x10.R")
   library("gamlss.cens")
@@
 
-  save(rainres, file = "rain_Axams.rda")
+  save(rainres, file = "rain_Axams_7x10.rda")
 }
@@
 boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)"),
+        names = c("Forest", "prespecified GAM", "boosted GAM"),
         ylab = "CRPS skills score", col = "lightgray") 
@@
 \center
-\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 7x10 cross validation at the station Axams.}
+\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 7x10 cross validation at station Axams.}
 \end{center}
@@
 
-The CRPS skills score boxplot shows that in this setting both GAM models and the forest perform better than the EMOS model.
-While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distforest even reaches an improvement rate of close to 6 percent.\\
-From this result we can conclude that the distributional forest provides a framework where only little or almost no effort is required to build a model which performs even better than other methods that are commonly used in this kind of application.\\
+The CRPS skills score boxplot shows that, in this setting, both GAM models and the forest perform better than the EMOS model.
+While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the forest even reaches an improvement rate of close to 6 percent.\\
+From this result we can conclude that the distributional forest provides an alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.\\
 For the specified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
@@
 Until now the focus has been on one observation station only.
-To show that the presented results do not depend on the choice of this station we are now going to apply the methods on all stations.
-At each of them, the first 25 years are used as learning data set and predictions are made for the last 3 years of available observations.
+To show that the presented results do not depend on the choice of this particular station we are now going to apply the methods on all 95 stations.
+At each of them, the first 24 years are used as learning data set and predictions are made for the last 4 years of available observations.
 For these out-of-sample predictions the CRPS skills score with respect to the EMOS model as reference is calculated for each method at each station.
 In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
-The red line represents the results for the station Axams.\\
-(Again, the corresponding boxplots of the RMSE and the log-likelihood values can be found in the appendix.)
+The blue line represents the results for the station Axams.
 %NAs in gamlss in 13 stations!!!\\
@@
 <<eval = TRUE, echo = FALSE>>=
-#### prediction over all stations 25 - 3
-if(file.exists("rain_pred.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred.rda")
-  load("rain_pred.rda")
+#### prediction over all stations 24 - 4
+if(file.exists("rain_pred_24to4.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
+  load("rain_pred_24to4.rda")
 } else {
 
-  source("rain_pred.R")
+  source("rain_pred_24to4.R")
   library("gamlss.cens")
@@
 
-  save(res, file = "rain_pred.rda")
+  save(res, file = "rain_pred_24to4.rda")
 }
@@
 s <- 1 - crps[, 2:4]/crps[,6]
-colnames(s) <- c("Forest", "GAM (prespecified)", "GAM (boosted)")
+colnames(s) <- c("Forest", "prespecified GAM", "boosted GAM")
 
@@
 #pie(rep(1, 3), col = c(pal0[1], pal2[1], pal5[1]))
+palone <- hcl(250, 100, c(90, 60, 30))
+#pie(rep(1, 3), col = palone)
 
@@
 diff_crps_ssc <- numeric(length = length(crps_ssc_df))
+diff_crps_df <- numeric(length = length(crps_ssc_df))
 transp_col_crps_ssc <- numeric(length = length(crps_ssc_df))
 level_col_crps_ssc <- numeric(length = length(crps_ssc_df))
-
+pch_crps_ssc <- numeric(length = length(crps_ssc_df))
+pch_intensity_crps_ssc <- numeric(length = length(crps_ssc_df))
+reference_crps_ssc <- numeric(length = length(crps_ssc_df))
+
+# get best model (loop over all stations)
 for(i in 1: length(crps_ssc_df)){
   col_crps_ssc[i] <- c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"])[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
+  pch_crps_ssc[i] <- c(21, 24, 25, 22)[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
+
   crps_ssc_best[i] <- which.max(crps_ssc[i,])
   diff_crps_ssc[i] <- crps_ssc[i,crps_ssc_best[i]] - crps_ssc[i,2]
+
+  pch_intensity_crps_ssc[i] <- 1
+  if(diff_crps_ssc[i] > 0.02) pch_intensity_crps_ssc[i] <- 2
+  if(diff_crps_ssc[i] > 0.05) pch_intensity_crps_ssc[i] <- 3
+
+  # falls forest am besten
   if(crps_ssc_best[i] == 2){
-    reference_crps_ssc <- max(c(crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]), na.rm = TRUE)
-    diff_crps_df <- crps_ssc_df[i] - reference_crps_ssc
+    reference_crps_ssc[i] <- max(c(crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]), na.rm = TRUE)
+    diff_crps_df[i] <- crps_ssc_df[i] - reference_crps_ssc[i]
     level_col_crps_ssc[i] <- pal0[1]
-    if(diff_crps_df > 0.02) level_col_crps_ssc[i] <- pal2[1]
-    if(diff_crps_df > 0.05) level_col_crps_ssc[i] <- pal5[1]
+    pch_intensity_crps_ssc[i] <- 1
+    if(diff_crps_df[i] > 0.02) {
+      level_col_crps_ssc[i] <- pal2[1]
+      pch_intensity_crps_ssc[i] <- 2
+    }
+    if(diff_crps_df[i] > 0.05) {
+      level_col_crps_ssc[i] <- pal5[1]
+      pch_intensity_crps_ssc[i] <- 3
+    }
   } else {
@@
                   #gray(0.1, alpha = 0.8), 
-                  "brown",
+                  "blue",
                   rep(gray(0.5, alpha = 0.3),15)), 
@@
 
-On the map in Figure~\ref{map} each observation station is represented by a point colored according to the method that performed best in terms of CRPS at this station.
+Figure~\ref{map} shows the location of all stations in the data set.
+Each observation station is represented by a point colored according to the method that performed best in terms of CRPS at this station.
 For those points where the distributional forest is not the best performing method the intensity of the color represents how much better (in terms of CRPS) the method is compared to the distributional forest.
-The more intense the color is the bigger the difference.
+The more intense the color, the bigger the difference.
 In case of the forest being the best model, the second best method is chosen as reference.
@@
 
+  if(FALSE){
     raster::image(tirol.dem, col = rev(gray.colors(100)),
@@
     #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
-  legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS"), cex = 1, bty = "n")
+    legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
+
+  }
+
+    raster::image(tirol.dem, col = rev(gray.colors(100)),
+                  main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
+                  xlim = c(9.8,13.2), 
+                  ylim = c(46.6, 47.8))
+    plot(tirol.gadm, add = TRUE)
+    points(sp, pch = pch_crps_ssc, bg = palone[pch_intensity_crps_ssc], col = 1, cex = 1.5)
+    #points(sp, pch = 21, bg = col_crps_ssc, col = 1)
+    #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
+    legend("topleft", pch=c(21,24,25,22), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
 
@@
 
-\newpage
-\section*{Appendix}
-
-\begin{figure}[h!]
-\begin{center}
-%\begin{subfigure}{0.5\textwidth}
-<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE, width=10, height=7>>=
-boxplot(rain_rmse[,c(2,3,4,6)], ylab = "RMSE",
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
-@
-%\end{subfigure}
-\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at the station Axams.}
-\end{center}
-\end{figure}
-%\hspace{-1cm}
-\begin{figure}[h!]
-\begin{center}
-%\begin{subfigure}{0.5\textwidth}
-<<rain_cross_axams_ll, fig=TRUE, echo=FALSE, width=10, height=7>>=
-boxplot(rain_ll[,c(2,3,4,6)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
-#boxplot(rain_ll)
-@
-%\end{subfigure}
-%\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
-%left: RMSE, right: log-likelihood}
-\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams (extracted interval [-2.8,-1.3]).}
-\end{center}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<rain_all_rmse, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
-  matplot(t(rmse[,c(2,3,4,6)]), type = "l", lwd = 1, 
-          col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, 
-          xlab = "", ylab = "RMSE", xlim = c(0.5, 4.5))
-  boxplot(rmse[,c(2,3,4,6)], add = TRUE,
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
-@
-\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations.}
-\end{center}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<rain_all_ll, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
-  matplot(t(ll[,c(2,3,4,6)]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
-          lty = 1, axes = FALSE, 
-          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 4.5), ylim = c(-4,-1))
-  boxplot(ll[,c(2,3,4,6)], add = TRUE,
-        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
-@
-\caption{\label{boxplot_ll_all}Log-likelihood values for predictions over all stations (extracted interval [-4,-1]).}
-\end{center}
-\end{figure}
+%\newpage
+%\section*{Appendix}
+
+%\begin{figure}[h!]
+%\begin{center}
+%%\begin{subfigure}{0.5\textwidth}
+%<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE, width=10, height=7>>=
+%boxplot(rain_rmse[,c(2,3,4,6)], ylab = "RMSE",
+%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
+%@
+%%\end{subfigure}
+%\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at station Axams.}
+%\end{center}
+%\end{figure}
+%%\hspace{-1cm}
+%\begin{figure}[h!]
+%\begin{center}
+%%\begin{subfigure}{0.5\textwidth}
+%<<rain_cross_axams_ll, fig=TRUE, echo=FALSE, width=10, height=7>>=
+%boxplot(rain_ll[,c(2,3,4,6)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
+%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
+%#boxplot(rain_ll)
+%@
+%%\end{subfigure}
+%%\caption{\label{boxplots_rain}Results 10x10 cross validation at station Axams\\
+%%left: RMSE, right: log-likelihood}
+%\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at station Axams (extracted interval [-2.8,-1.3]).}
+%\end{center}
+%\end{figure}
+
+%\begin{figure}[h!]
+%\begin{center}
+%<<rain_all_rmse, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
+%  matplot(t(rmse[,c(2,3,4,6)]), type = "l", lwd = 1, 
+%          col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, 
+%          xlab = "", ylab = "RMSE", xlim = c(0.5, 4.5))
+%  boxplot(rmse[,c(2,3,4,6)], add = TRUE,
+%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
+%@
+%\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations.}
+%\end{center}
+%\end{figure}
+
+%\begin{figure}[h!]
+%\begin{center}
+%<<rain_all_ll, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
+%  matplot(t(ll[,c(2,3,4,6)]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
+%          lty = 1, axes = FALSE, 
+%          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 4.5), ylim = c(-4,-1))
+%  boxplot(ll[,c(2,3,4,6)], add = TRUE,
+%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
+%@
+%\caption{\label{boxplot_ll_all}Log-likelihood values for predictions over all stations (extracted interval [-4,-1]).}
+%\end{center}
+%\end{figure}
 

### 2017-11-16 pkg/devel/partykit/R/cforest.R
mob/cforest
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1722 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: da835fa31c6660c3f581fc89a234cce3662f6503

@@ -73,62 +73,37 @@ cforest <- function
    
     ### get the call and the calling environment for .urp_tree
     call <- match.call(expand.dots = FALSE)
-    call$na.action <- na.action
-    frame <- parent.frame()
-    if (missing(data)) {   
-        data <- NULL
-        data_asis <- FALSE
-    } else {
-        data_asis <- missing(weights) && missing(subset) &&
-                     missing(cluster) && missing(offset)   
-    }
+    m <- match(c("formula", "data", "subset", "na.action", "weights", "offset", "cluster", 
+                 "scores", "ytrafo", "control"), names(call), 0L)
+    ctreecall <- call[c(1L, m)]
+    ctreecall$doFit <- FALSE
+    ctreecall[[1L]] <- quote(partykit:::ctree)
+    tree <- eval(ctreecall, parent.frame())
 
-    ### <FIXME> should be xtrafo
-    if (!is.null(scores)) {
-        if (missing(data)) 
-            stop("can deal with scores with data being missing")
-        for (n in names(scores)) {
-            sc <- scores[[n]]
-            if (is.ordered(data[[n]]) &&
-                nlevels(data[[n]]) == length(sc)) {
-                attr(data[[n]], "scores") <- as.numeric(sc)
-            } else {
-                warning("scores for variable ", sQuote(n), " ignored")
-            }
-        }
-    }
-    #### </FIXME>
+    d <- tree$d
+    updatefun <- tree$update
 
-    call$weights <- NULL ### NOTE: trees are unweighted, weights enter sampling!
-    trafofun <- function(...) .ctreetrafo(..., ytrafo = ytrafo)
-    tree <- .urp_tree(call, frame, data = data, data_asis = data_asis, control = control,
-                      trafofun = trafofun, doFit = FALSE)
-
-    nvar <- length(tree$partyvars)
+    nvar <- sum(d$variables$z > 0)
     control$mtry <- mtry
     control$applyfun <- lapply
  
-    ### <FIXME> we need tree$partyvars here, avoid calling .urp_tree twice
-    tree <- .urp_tree(call, frame, data = data, data_asis = data_asis, control = control,
-                      trafofun = trafofun, doFit = FALSE)
-    ### </FIXME>
-
-    strata <- tree$mf[["(strata)"]]
+    strata <- d[["(strata)"]]
     if (!is.null(strata)) {
[…]

### 2017-11-16 pkg/devel/partykit/R/ctree.R

@@ -377,6 +377,7 @@ ctree <- function(formula, data, subset, na.action = na.pass, weights, offset, c
         nf <- names(formals(ytrafo))
         if (all(c("data", "weights", "control") %in% nf))
             ytrafo <- ytrafo(data = d, weights = weights, control = control)
+        nf <- names(formals(ytrafo))
         stopifnot(all(c("subset", "weights", "info", "estfun", "object") %in% nf) ||
                   all(c("y", "x", "weights", "offset", "start") %in% nf))
     } else {

### 2017-11-16 pkg/devel/partykit/R/extree.R

@@ -53,7 +53,7 @@
     svars <- which(partyvars > 0)
     if (ctrl$mtry < Inf) {
         mtry <- min(sum(partyvars > 0), ctrl$mtry)
-        svars <- .resample(partyvars, mtry, prob = partyvars)
+        svars <- .resample(svars, mtry, prob = partyvars[partyvars > 0])
     } 
 
     thismodel <- trafo(subset = subset, weights = weights, info = info, 
@@ -259,22 +259,17 @@ extree_fit <- function(data, trafo, converged, selectfun = NULL,
             iy <- data[["yx", type = "index"]]
             if (is.null(iy)) {
                 NAyx <- data[["yx", type = "missing"]]
-                if (length(NAyx) > 0) {
-                    y <- matrix(0, nrow = nrow(model.frame(data)), ncol = ncol(yx$y))
-                    y[-NAyx, ,drop = FALSE] <- yx$y
-                    x <- matrix(0, nrow = nrow(model.frame(data)), ncol = ncol(yx$x))
-                    x[-NAyx, ,drop = FALSE] <- yx$x
-                    offset <- numeric(nrow(model.frame(data)))
-                    offset[-NAyx] <- attr(yx$x, "offset")
-                } else {
-                    y <- yx$y
-                    x <- yx$x
-                    offset <- attr(yx$x, "offset")
-                }
+                y <- yx$y
+                x <- yx$x
+                offset <- attr(yx$x, "offset")
                 subset <- subset[!(subset %in% NAyx)]
-                y <- y[subset,,drop = FALSE]
-                if (!is.null(yx$x)) {
-                    ax <- attributes(yx$x)
+                if (NCOL(y) > 1) {
+                    y <- y[subset,,drop = FALSE]
+                } else {
+                    y <- y[subset]
+                }
+                if (!is.null(x)) {
+                    ax <- attributes(x)
                     ax$dim <- NULL
                     ax$dimnames <- NULL
                     x <- x[subset,,drop = FALSE]
@@ -288,7 +283,7 @@ extree_fit <- function(data, trafo, converged, selectfun = NULL,
                 } else {
                     obj <- trafo(y = y, x = x, offset = offset, weights = w, start = info$coef, ...)
[…]

### 2017-11-16 pkg/devel/partykit/R/modelparty_R2.R

@@ -220,16 +220,15 @@ mob_control <- function(
   terminal = "object",
   mtry = Inf, 
   nmax = Inf,
+  ytype = c("vector", "data.frame", "matrix"),
   # deprecated
-  verbose, xtype, ytype, prune
+  verbose, xtype, prune
 ) {
   
   if (!missing("verbose"))
     warning("argument verbose deprecated")
   if (!missing("xtype"))
     warning("argument xtype deprecated")
-  if (!missing("ytype"))
-    warning("argument ytype deprecated")
   if (!missing("prune"))
     warning("argument prune deprecated")
     
@@ -260,7 +259,7 @@ mob_control <- function(
     list(breakties = breakties, 
          intersplit = intersplit, parm = parm, dfsplit = dfsplit, 
          restart = restart, model = model, vcov = match.arg(vcov), 
-         ordinal = match.arg(ordinal), 
+         ordinal = match.arg(ordinal), ytype = match.arg(ytype),
          nrep = nrep, terminal = terminal, inner = inner, trim = trim))
 }
 
@@ -299,7 +298,7 @@ mob <- function
       }
       list(
         coefficients = coef(obj),
-        objfun = as.numeric(logLik(obj)),
+        objfun = -as.numeric(logLik(obj)),
         estfun = if(estfun) sandwich::estfun(obj) else NULL,
         object = if(object) obj else NULL
       )
@@ -321,6 +320,7 @@ mob <- function
     mf <- mf[c(1L, m)]
     mf$yx <- "matrix"
     mf$nmax <- control$nmax
+    mf$ytype <- control$ytype
     ## evaluate model.frame
     mf[[1L]] <- quote(partykit:::extree_data)
 
@@ -349,7 +349,7 @@ mob <- function
         } else {
             n_coef <- length(cf)
         }
-        minsize <- as.integer(ceiling(10L * n_coef/ncol(d$yx$y)))
+        minsize <- as.integer(ceiling(10L * n_coef))
         if (is.null(control$minbucket)) control$minbucket <- minsize
         if (is.null(control$minsplit)) control$minsplit <- minsize
     }

### 2017-11-22 pkg/disttree/inst/draft/draft.Rnw
draft without simulation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1753 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6237ac5a31c1022e265155a69a9b7f8f546b3c9a

@@
 \usepackage{tikz}
+\usetikzlibrary{arrows.meta}
 \usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
@@
 \begin{figure}[t!]
+\begin{tikzpicture}
+\node (a) at (0,0){
 \minipage{0.29\textwidth}
@@
 \end{center}
-\endminipage
-{\LARGE$\rightarrow$}
+\endminipage};
+\node (b) at (5,0) 
+{
 \minipage{0.29\textwidth}
@@
 \end{center}
-\endminipage
-{{\LARGE$\rightarrow$}}
+\endminipage};
+\node (c) at (10,0) 
+{
 \minipage{0.29\textwidth}
@@
 \end{center}
-\endminipage
+\endminipage};
+\node (d) at (0,-2) 
+{
+\minipage{0.29\textwidth}
+\begin{center}
 \vspace{0.0cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\hspace{0.4cm}
 LM, GLM  
 \end{center}
-\endminipage
-\hspace{1.7cm}
-\minipage{0.25\textwidth}
+\endminipage};
+\node (e) at (5,-2) 
+{
+\minipage{0.29\textwidth}
 \begin{center}
+\vspace{0.0cm}
 GAM 
 \end{center}
-\endminipage
-\hspace{1.7cm}
-\minipage{0.25\textwidth}
+\endminipage};
+\node at (10,-2)
+{
+\minipage{0.28\textwidth}
 \begin{center}
+\vspace{0.0cm}
 GAMLSS 
 \end{center}
-\endminipage
+\endminipage};
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
+\end{tikzpicture}
 \caption{\label{Devel_parmod}Development in the field of parametric modeling.}
@@
 \end{center}
-\endminipage
-{\LARGE$\rightarrow$}
+\endminipage};
+\node (c) at (10,0) 
+{
 \minipage{0.29\textwidth}
@@
 \end{center}
-\endminipage
-
+\endminipage};
+\node (d) at (0,-2) 
+{
 \minipage{0.29\textwidth}
@@
 \end{center}
-\endminipage
-\hspace{0.65cm}
+\endminipage};
+\node (e) at (5,-2) 
+{
 \minipage{0.29\textwidth}
@@
 \end{center}
-\endminipage
-\hspace{0.65cm}
+\endminipage};
+\node at (10,-2)
+{
 \minipage{0.28\textwidth}
@@
 \end{center}
-\endminipage\\
+\endminipage};
+\node (g) at (0,-5.4) 
+{
 \minipage{0.29\textwidth}
 \vspace{0.2cm}
-\center {\LARGE$\downarrow$}\\
-\vspace{-0.4cm}
 %\hspace{0.0cm}
@@
 @
-\vspace{-0.2cm}\\
+\vspace{-0.2cm}%\\
 Distributional Tree
@@
 \end{center}
-\endminipage
+\endminipage};
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](0,-2.4)--(0,-3.2);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.2,-5.4)--(7.9,-2.5);
+\end{tikzpicture}
 \caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests}
 \end{figure}
+
+%\SweaveOpts{eval = FALSE}
 
@@
 
-After explaining the novel methodology in Section~\ref{Methodology} of this paper and presenting results from a simulation study in Section~\ref{Simulation} the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}.
+After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}.
 The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
 In a first setting only one observation station is taken into consideration.
-The models are trained on 24 years and prbabilistic predictions of the consecutive 4 years are validated.
+The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years are validated.
 A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}.
@@
 In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
-Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
+(The power transformation is used to remove large parts of the skewness of the data as further explained in \ref{setting}.) Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
 $$
@@
 These values express how well the model fits the data regarding each parameter.
-Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.\\
+Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.
+This enables the model to be sensitive to changes in each of the parameters which is one of the main advantages of distributional trees over other tree models considering only the mean of the distribution.\\
 Statistical tests are applied to check whether there is a significant connection/dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$.
@@
 In that way, all kinds of effects can be modeled, which is is also one the of the main advantages of forest models (taking bagging one step further as the variance is reduced by sampling the covariates considered for the splitting process in each node).
+\\
+\\
+Common forest models such as random forests consist of trees only considering the mean of the distribution of the response variable.
+For that reason, changes in other parameters can only be detected if they are connected to changes in the location parameter, e.g.
+if they depend on the location parameter.
+Therefore, building a forest of distributional trees which can also capture separate changes in each of the distribution parameters leads to an improvement of the model.
+This valuable feature of detecting separate changes is also provided by GAMLSS as each parameter is modeled separately.
+However, in this case this separation can also be a disadvantage as it hinders the capturing of interactions unless they are specified explicitly by the user.
+These two examples/situations provide once again a good reason to combine the two frameworks of flexible parametric modeling and tree structured models in order to profit from their advantages and compensate possible disadvantages.
+This is the pursued goal in introducing distributional forests.
 
@@
 $$
-While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates which is due to the individual weights calculated separately for each set of covariates.
-
-\newpage
-\section{Simulation Study}
-\label{Simulation}
-Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.
+While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates which is due to the individual weights calculated separately for each set of covariates.\\
+\\
+As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree structured modeling it provides a model that can deal with a wide range of situations.
+Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably by combining various tree and smoothing the steps imposed by a single tree.
+As scores are considered for each distribution parameter (separate) effects in different parameters can be captured.
+At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.\\
+Even though many modeling situations demand for the use of a specific model providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance what kind of situation this is.
+In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.
+
+\SweaveOpts{eval = FALSE}
+%\newpage
+%\section{Simulation Study}
+%\label{Simulation}
+%Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.
 
@@
 
-The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\citealt{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
-Each generated data set includes 500 samples consisiting of the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
-$$
-X_1, X_2, X_3 \sim \textit{Unif}([-1,1])
-$$
-$$
-X_4, X_5, X_6 \sim \textit{Binom}(1, 0.5)
-$$
-
-The distribution of the response variable is a left censored normal distribution with censoring point 0.
+%The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\citealt{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
+%Each generated data set includes 500 samples consisiting of the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
+%$$
+%X_1, X_2, X_3 \sim \textit{Unif}([-1,1])
+%$$
+%$$
+%X_4, X_5, X_6 \sim \textit{Binom}(1, 0.5)
+%$$
+
+%The distribution of the response variable is a left censored normal distribution with censoring point 0.
 The corresponding distribution parameters are defined by a parameter function $f$ depending on the covariates and an additional factor $\kappa$, also known as the smoothness constraint, which regulates the steepness of certain sections of the parameter function.
@@
 \\
-The location parameter $\mu$ depends on $X_1, X_4$ and $\kappa$ and the scale parameter $\sigma$ depends on $\mu, X_1$ and $\kappa$.
+%The location parameter $\mu$ depends on $X_1, X_4$ and $\kappa$ and the scale parameter $\sigma$ depends on $\mu, X_1$ and $\kappa$.
 The other covariates $X_2, X_3, X_5$ and $X_6$ are used as noise variables.\\
-The latent variable $Y^{\ast}$ is generated based on the following setting.
-\begin{center}
-$$ 
-Y^{\ast} \sim \mathcal{N} (f(X_1, X_4, \kappa))
-$$
-with
-$$
-f(X_1, X_4, \kappa) = \left(\mu(X_1, X_4, \kappa) , \sigma(\mu,X_1,\kappa)\right)
-$$ 
-and
-\begin{align*}
-\mu(X_1, X_4, \kappa) = 0.2 
-+ \begin{cases}
-    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
-    & \text{if } X_1 < 0.4 \text{ and } X_4 = 0\\
-    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
-    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 0\\
-    10 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
-    & \text{if } X_1 < 0.4 \text{ and } X_4 = 1\\
-    13 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
-    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 1
-\end{cases}
-\end{align*}
-
-$$
-\sigma(\mu, X_1,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.4)
-$$
-%where $\Lambda$ is the cumulative distribution function of the logistic distribution
+%The latent variable $Y^{\ast}$ is generated based on the following setting.
+%\begin{center}
 %$$ 
-%\Lambda(\omega) = \frac{1}{1 + \exp(-\omega)}
+%Y^{\ast} \sim \mathcal{N} (f(X_1, X_4, \kappa))
 %$$
-The response variable $Y$ is defined as 
-$$
-Y = 
-\begin{cases}
-    Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
-    0,       & \text{else}
-\end{cases}
-$$
-\end{center}
-
-To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to~1 and then to a second data set where $\kappa$ is set to 10.
+%with
+%$$
+%f(X_1, X_4, \kappa) = \left(\mu(X_1, X_4, \kappa) , \sigma(\mu,X_1,\kappa)\right)
+%$$ 
+%and
+%\begin{align*}
+%\mu(X_1, X_4, \kappa) = 0.2 
+%+ \begin{cases}
+%    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
+%    & \text{if } X_1 < 0.4 \text{ and } X_4 = 0\\
+%    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
+%    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 0\\
+%    10 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
+%    & \text{if } X_1 < 0.4 \text{ and } X_4 = 1\\
+%    13 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
+%    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 1
+%\end{cases}
+%\end{align*}
+
+%$$
+%\sigma(\mu, X_1,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.4)
+%$$
+%%where $\Lambda$ is the cumulative distribution function of the logistic distribution
+%%$$
+%%\Lambda(\omega) = \frac{1}{1 + \exp(-\omega)}
+%%$$
+%The response variable $Y$ is defined as 
+%$$
+%Y = 
+%\begin{cases}
+%    Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
+%    0,       & \text{else}
+%\end{cases}
+%$$
+%\end{center}
+
+%To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to~1 and then to a second data set where $\kappa$ is set to 10.
 In Figure~\ref{plot_sim_oneset_mu} the location parameter function $\mu(X_1, X_4, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $x_4 = 0$ and $(x_2, x_3, x_5, x_6) = (0,0,0,0)$.
@@
 
-\SweaveOpts{eval = TRUE}
-<<echo=FALSE, results=hide>>=
+%\SweaveOpts{eval = TRUE}
+<<eval = FALSE, echo=FALSE, results=hide>>=
 
@@
 
-\begin{figure}[t!]
-\begin{subfigure}{0.6\textwidth}
-<<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset1, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
-@
-\end{subfigure}
-\hspace{-1cm}
-\begin{subfigure}{0.6\textwidth}
-<<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
-plot_oneset(oneset10, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
-@
-\end{subfigure}
-\caption{\label{plot_sim_oneset_mu} Smooth parameter function (left window, $\kappa = 1$) and parameter function with abrupt changes (right window, $\kappa = 10$) together with the predicted location parameter $\mu$ of the tree model, the forest model and the GAMLSS,  plotted along $x_1$ while all the other covariates are kept at fixed values.}
-\end{figure}
-
-In the case of smooth effects of the coefficients on the parameter function ($\kappa = 1$, left window in Figure~\ref{plot_sim_oneset_mu}) the fitted parameters of the GAMLSS are closest to the true parameters while the steps of the tree are cleary too rough.
+%\begin{figure}[t!]
+%\begin{subfigure}{0.6\textwidth}
+%<<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
+%plot_oneset(oneset1, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
+%@
+%\end{subfigure}
+%\hspace{-1cm}
+%\begin{subfigure}{0.6\textwidth}
+%<<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
+%plot_oneset(oneset10, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
+%@
+%\end{subfigure}
+%\caption{\label{plot_sim_oneset_mu} Smooth parameter function (left window, $\kappa = 1$) and parameter function with abrupt changes (right %window, $\kappa = 10$) together with the predicted location parameter $\mu$ of the tree model, the forest model and the GAMLSS,  plotted along %$x_1$ while all the other covariates are kept at fixed values.}
+%\end{figure}
+
+%In the case of smooth effects of the coefficients on the parameter function ($\kappa = 1$, left window in Figure~\ref{plot_sim_oneset_mu}) the fitted parameters of the GAMLSS are closest to the true parameters while the steps of the tree are cleary too rough.
 By combining various trees the forest can smooth out these steps, however, not as well as the GAMLSS.\\
-Looking at the other extreme situation in which the parameter functions include abrupt changes ($\kappa = 10$, right window in Figure~\ref{plot_sim_oneset_mu}) the tree captures them very well wich also leads to a good fit of the forest model.
+%Looking at the other extreme situation in which the parameter functions include abrupt changes ($\kappa = 10$, right window in Figure~\ref{plot_sim_oneset_mu}) the tree captures them very well wich also leads to a good fit of the forest model.
 In contrast to the two tree structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter functions.\\
 
-To get a more profound impression, the methods are evaluated on various data sets. The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step 100 data sets are generated.\\
-%\subsubsection{Means of Comparison}
-In order to compare the performance of the chosen methods over various data sets, the continuous ranked probability score (CRPS) is calculated for each model on each data set.
+%To get a more profound impression, the methods are evaluated on various data sets. The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step 100 data sets are generated.\\
+%%\subsubsection{Means of Comparison}
+%In order to compare the performance of the chosen methods over various data sets, the continuous ranked probability score (CRPS) is calculated for each model on each data set.
 
@@
 %The CRPS is another value that measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
-The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
-$$
-\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
-$$
-where \(\mathbf{1}\) is the indicator function. \\
-In this simulation study the CRPS is evaluated out of sample on a data set that has not been used to build the model but has been generated by the same process and with the same settings as the original data set.
+%The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+%$$
+%\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
+%$$
+%where \(\mathbf{1}\) is the indicator function. \\
+%In this simulation study the CRPS is evaluated out of sample on a data set that has not been used to build the model but has been generated by the same process and with the same settings as the original data set.
 In the following this second data set is called testing data set while the original data set on which the models are built on is called learning data set.
-The mean CRPS over the whole testing data set is used for validation.
+%The mean CRPS over the whole testing data set is used for validation.
 As the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
@@
 
-\SweaveOpts{eval = TRUE}
-
-<<echo=FALSE, results=hide>>=
+%\SweaveOpts{eval = FALSE}
+
+<<eval = FALSE, echo=FALSE, results=hide>>=
 if(file.exists("simres.rda")){
@@
 
-\begin{figure}[t!]
-\begin{center}
-%\begin{subfigure}{0.6\textwidth}
-<<sim_crps, fig=TRUE, echo=FALSE>>=
-plot_crps(simres)
-@
-\caption{\label{plot_sim_crps}CRPS of the three models evaluated on various data sets over increasing $\kappa$} %\code{disttree} vs.
-\code{distforest} vs.
-\code{gamlss} vs.
-\code{randomForest} vs.
-%\code{bamlss} vs.
-\code{gamboostLSS}\\
-%right: levelplot of the parameter function for $\kappa=5$}
-\end{center}
-\end{figure}
+%\begin{figure}[t!]
+%\begin{center}
+%%\begin{subfigure}{0.6\textwidth}
+%<<sim_crps, fig=TRUE, echo=FALSE>>=
+%plot_crps(simres)
+%@
+%\caption{\label{plot_sim_crps}CRPS of the three models evaluated on various data sets over increasing $\kappa$} 
+%%\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest} vs. %\code{bamlss} vs. \code{gamboostLSS}\\
+%%right: levelplot of the parameter function for $\kappa=5$}
+%\end{center}
+%\end{figure}
 
@@
 
-Figure~\ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study and supports the assumptions made above based on the results for the two specific situations in Figure~\ref{plot_sim_oneset_mu}.
+%Figure~\ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study and supports the assumptions made above based on the results for the two specific situations in Figure~\ref{plot_sim_oneset_mu}.
 Moreover, it can be observed how the stepwise transition from the one extreme situation to the other influences the performance of the applied methods.\\
-The parametric GAMLSS can easily deal with the smooth effects that appear for low values of $\kappa$ and outperforms the tree and forest models in this situation.
+%The parametric GAMLSS can easily deal with the smooth effects that appear for low values of $\kappa$ and outperforms the tree and forest models in this situation.
 However, it clearly starts to struggle as $\kappa$ increases and the effects turn into abrupt changes.
 %This behaviour has already been detected and discussed above when looking at only single data sets and is now proven by an almost monotone increase of the CRPS.\\
-On the contrary, the imposed/enforced steps of the tree model lead to a high CRPS in the situation of smooth effects while decreasing quickly as the slopes in the parameter functions get steeper.
+%On the contrary, the imposed/enforced steps of the tree model lead to a high CRPS in the situation of smooth effects while decreasing quickly as the slopes in the parameter functions get steeper.
 For high values of $\kappa$ resulting almost in step functions the tree model perfomes the best as stated by the low CRPS value.\\
-Therefore, both extreme situations show to have a preferable method to choose which clearly outperforms the others.
+%Therefore, both extreme situations show to have a preferable method to choose which clearly outperforms the others.
 But overall the forest model stays more or less on the same level of goodness of fit regardless of the type of effects in the parameter functions.
@@
 \subsection{Setting}
+\label{setting}
 The data set used for training and validation consists of observed daily precipitation sums provided by the National Hydrographical Service (\citealt{ehyd}) and numerical weather forecasts from the {U.S.} National Oceanic and Atmospheric Administration (NOAA).
-Both, observations and weather forecasts are available from 1985 trough the year 2012, focussing on July only in this article.\\ 
+Both, observations and weather forecasts are available from 1985 trough the year 2012, focusing on July only in this article.\\ 
 Observations are available on a daily basis, each observation representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and it's close surrounding.
@@
 As covariates the second generation reforecast data set of the global ensemble forecast system (GEFS) is used (\citealt{hamill2013}).
-This data set consists of an 11-member ensemble based on a fixed version of the numerical model and a horizontal grid-spacing of about $50~\times~50 \text{km}^2$ initialized at 0~UTC from Dezember 1984 to precent.
-Each of the 11 ensemble members uses slightly different perturbed initial conditions trying to predict the situation specific uncertenty of the atmospheric state.
+This data set consists of an 11-member ensemble based on a fixed version of the numerical model and a horizontal grid-spacing of about $50~\times~50 \text{km}^2$ initialized at 0~UTC from December 1984 to present.
+Each of the 11 ensemble members uses slightly different perturbed initial conditions trying to predict the situation specific uncertainty of the atmospheric state.
 
@@
 
-\subsection{Models}
+\subsection{Models and Evaluation}
 Next to distributional forests three other methods are applied to this data set and their results are compared based on CRPS values.
@@
 This choice has been made based on the results of \citealt{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared.
-Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.
-
-\SweaveOpts{eval = TRUE}
+Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.\\
+\\
+In order to compare the performance of the chosen methods, the continuous ranked probability score (CRPS) is calculated for each model.
+The CRPS measures the accuracy of a probabilistic model.
+For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+$$
+\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
+$$
+where \(\mathbf{1}\) is the indicator function. \\
+In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model and the mean CRPS over this data set is used for validation.
+Since the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
+In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observation.
+\\
+
+%\SweaveOpts{eval = TRUE}
 \subsection{Application on one station}
@@
 As shown in Table~\ref{model_specification} the forest model includes all 82 covariates.
-A single tree model allows to easily analize which covariates are used as split variables in the nodes and hence have an impact on the response variable.
+A single tree model allows to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable.
 In a forest model though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree.
@@
 
-\SweaveOpts{eval = TRUE}
+%\SweaveOpts{eval = TRUE}
 %\subsubsection{Cross-Validation}
@@
 
+\SweaveOpts{eval = FALSE}
 %\newpage
@@
 % sum(is.na(crps[,3]))
-<<eval = TRUE, echo = FALSE>>=
+%<<eval = TRUE, echo = FALSE>>=
+<<echo = FALSE>>=
 #### prediction over all stations 24 - 4
@@
 
-ll <- res[[1]]$results["ll",]
-for(i in 2:(length(res)-1)) ll <- rbind(ll, res[[i]]$results["ll",])
-
-rmse <- res[[1]]$results["rmse",]
-for(i in 2:(length(res)-1)) rmse <- rbind(rmse, res[[i]]$results["rmse",])
-
-crps <- res[[1]]$results["crps",]
-for(i in 2:(length(res)-1))  crps <- rbind(crps, res[[i]]$results["crps",])
-
-colnames(ll) <- colnames(rmse) <- colnames(crps) <- colnames(res[[1]]$results)
+ll_all <- res[[1]]$results["ll",]
+for(i in 2:(length(res)-1)) ll_all <- rbind(ll_all, res[[i]]$results["ll",])
+
+rmse_all <- res[[1]]$results["rmse",]
+for(i in 2:(length(res)-1)) rmse_all <- rbind(rmse_all, res[[i]]$results["rmse",])
+
+crps_all <- res[[1]]$results["crps",]
+for(i in 2:(length(res)-1))  crps_all <- rbind(crps_all, res[[i]]$results["crps",])
+
+colnames(ll_all) <- colnames(rmse_all) <- colnames(crps_all) <- colnames(res[[1]]$results)
 
 # skills score
-s <- 1 - crps[, 2:4]/crps[,6]
+s <- 1 - crps_all[, 2:4]/crps_all[,6]
 colnames(s) <- c("Forest", "prespecified GAM", "boosted GAM")
 
-## prepare date for map which shows where distforest performed better than gamlss or gamboostLSS
-# based on the crps skills score (reference model EMOS)
+## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
+
+crps <- crps_all[,c("distforest", "gamlss", "gamboostLSS", "EMOS log")]
+
+## best method
+bst <- apply(crps, 1, which.min)
+
+## distance of forest to best other method
+dst <- crps[,1] - crps[cbind(1:nrow(crps), apply(crps[, -1], 1, which.min) + 1)]
+
+## breaks/groups
+brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
+grp <- cut(dst, breaks = brk)
+
+## HCL colors (relatively flashy, essentially CARTO Tropic)
+clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
 
@@
 
-library("colorspace")
-crps_ssc_df <- 1 - crps[,2]/crps[,6]
-crps_ssc_gb <- 1 - crps[,4]/crps[,6]
-crps_ssc_g <- 1 - crps[,3]/crps[,6]
-crps_ssc_ml <- 1 - crps[,6]/crps[,6]    # <- rep(0, length(crps_ssc_g))
-crps_ssc <-  1- crps/crps[,6]
-# abs(crps_ssc_df - crps_ssc_g)
-# boxplot(abs(crps_ssc_df - crps_ssc_g))
-# hist(abs(crps_ssc_df - crps_ssc_g))
-
-# set the colour according to whether distforest performed better than gamlss and
-# set the size of the points on the map according to the difference to the forest model
-pal0 <- hcl(c(10, 260, 290, 50), 100, 80)
-pal2 <- hcl(c(10, 260, 290, 50), 100, 65)
-pal5 <- hcl(c(10, 260, 290, 50), 100, 50)
-#pie(rep(1, 3), col = c(pal0[1], pal2[1], pal5[1]))
-palone <- hcl(250, 100, c(90, 60, 30))
-#pie(rep(1, 3), col = palone)
-
-col_crps_ssc <- numeric(length = length(crps_ssc_df))
-crps_ssc_best <- numeric(length = length(crps_ssc_df))
-diff_crps_ssc <- numeric(length = length(crps_ssc_df))
-diff_crps_df <- numeric(length = length(crps_ssc_df))
-transp_col_crps_ssc <- numeric(length = length(crps_ssc_df))
-level_col_crps_ssc <- numeric(length = length(crps_ssc_df))
-pch_crps_ssc <- numeric(length = length(crps_ssc_df))
-pch_intensity_crps_ssc <- numeric(length = length(crps_ssc_df))
-reference_crps_ssc <- numeric(length = length(crps_ssc_df))
-
-# get best model (loop over all stations)
-for(i in 1: length(crps_ssc_df)){
-  col_crps_ssc[i] <- c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"])[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
-  pch_crps_ssc[i] <- c(21, 24, 25, 22)[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
-
-  crps_ssc_best[i] <- which.max(crps_ssc[i,])
-  diff_crps_ssc[i] <- crps_ssc[i,crps_ssc_best[i]] - crps_ssc[i,2]
-
-  pch_intensity_crps_ssc[i] <- 1
-  if(diff_crps_ssc[i] > 0.02) pch_intensity_crps_ssc[i] <- 2
-  if(diff_crps_ssc[i] > 0.05) pch_intensity_crps_ssc[i] <- 3
-
-  # falls forest am besten
-  if(crps_ssc_best[i] == 2){
-    reference_crps_ssc[i] <- max(c(crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]), na.rm = TRUE)
-    diff_crps_df[i] <- crps_ssc_df[i] - reference_crps_ssc[i]
-    level_col_crps_ssc[i] <- pal0[1]
-    pch_intensity_crps_ssc[i] <- 1
-    if(diff_crps_df[i] > 0.02) {
-      level_col_crps_ssc[i] <- pal2[1]
-      pch_intensity_crps_ssc[i] <- 2
-    }
-    if(diff_crps_df[i] > 0.05) {
-      level_col_crps_ssc[i] <- pal5[1]
-      pch_intensity_crps_ssc[i] <- 3
-    }
-  } else {
-    level_col_crps_ssc[i] <- pal0[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
-    if(diff_crps_ssc[i] > 0.02) level_col_crps_ssc[i] <- pal2[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
-    if(diff_crps_ssc[i] > 0.05) level_col_crps_ssc[i] <- pal5[which.max(c(crps_ssc_df[i], crps_ssc_g[i], crps_ssc_gb[i], crps_ssc_ml[i]))]
-  }
-
-}
-
-# ehyd.statlist[res$complete_stations,][col_crps_ssc == pal["gamlss"],]
-# table(crps_ssc_best)
-# Axams (Station 77) is the 70th complete station
-# boxplot(diff_crps_ssc[diff_crps_ssc>0])
-# abline(h = diff_crps_ssc[70], col = "red")
-
-# define transperancy of points according to the difference to the forest model (for those points where the forest is not the best one) 
-# alpha parameter between 0 and 1
-alpha_crps_ssc <- (diff_crps_ssc*10) + 0.15
-#alpha_crps_ssc <- (diff_crps_ssc*10/0.85) * 0.5 + 0.5
-alpha_crps_ssc[diff_crps_ssc == 0] <- 1
-add.alpha <- function(col, alpha=1){
-  if(missing(col))
-    stop("Please provide a vector of colours.")
-  apply(sapply(col, col2rgb)/255, 2, 
-                     function(x) 
-                       rgb(x[1], x[2], x[3], alpha=alpha))  
-}
-for(i in 1:length(col_crps_ssc)) transp_col_crps_ssc[i] <- add.alpha(col_crps_ssc[i], alpha = alpha_crps_ssc[i])
+#library("colorspace")
 @
@@
 
-  if(FALSE){
+  layout(cbind(1, 2), width = c(9, 1))
+  par(mar = c(5,4,4,0.1))
   raster::image(tirol.dem, col = rev(gray.colors(100)),
@@
   plot(tirol.gadm, add = TRUE)
-    points(sp, pch = 21, bg = level_col_crps_ssc, col = 1, cex = 1.5)
-    #points(sp, pch = 21, bg = col_crps_ssc, col = 1)
-    #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
-    legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
-
-  }
-
-    raster::image(tirol.dem, col = rev(gray.colors(100)),
-                  main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
-                  xlim = c(9.8,13.2), 
-                  ylim = c(46.6, 47.8))
-    plot(tirol.gadm, add = TRUE)
-    points(sp, pch = pch_crps_ssc, bg = palone[pch_intensity_crps_ssc], col = 1, cex = 1.5)
-    #points(sp, pch = 21, bg = col_crps_ssc, col = 1)
-    #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
+  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
   legend("topleft", pch=c(21,24,25,22), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
 
-@
-\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions.}
+  par(mar = c(5,0.2,5,2.3))
+  ## legend
+  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
+       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
+  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
+  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))
+
+@
+\caption{\label{map}Map of Tyrol with all observation stations.
+The type of symbol indicates which model performed best with respect to the CRPS.
+The color indicates the difference of the CRPS of the forest model and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).}
 \end{center}
@@
 \section{Discussion}
+
+\section*{Computational Details}
+The corresponding implementation to the novel methods presented in this article is available on R-Forge in an \textsf{R}-package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}(.
+It uses the functions \code{mob} (\citealt{zeileis2010party}() and \code{ctree} (\citealt{hothorn2006unbiased}), both being provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}), as a framework for building trees and forests.
 

### 2017-11-23 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1754 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 937236edecd689c8513921b0194a3773b8f4c0cc

@@
 \usepackage{amsmath}
-\usepackage{thumbpdf}
-\usepackage{rotating}
-\usepackage{caption}
-\captionsetup{format=hang}
-\usepackage{subcaption}
+\usepackage{thumbpdf,lmodern}
+%\usepackage{rotating}
+%\usepackage{caption}
+%\captionsetup{format=hang}
+%\usepackage{subcaption}
 \usepackage{Sweave}
-\usepackage{enumitem}
-\usepackage{graphicx}
+%\usepackage{enumitem}
+%\usepackage{graphicx}
 \usepackage{array, makecell}
@@
 \Address{
-  Lisa Schlosser \\
+  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
   Department of Statistics \\
@@
   6020 Innsbruck, Austria \\
-  E-mail: \email{Lisa.Schlosser@uibk.ac.at} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/lisa-schlosser.html}\\
+  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/lisa-schlosser.html}, \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\
 
@@
   URL: \url{http://user.math.uzh.ch/hothorn/}\\
-
-  Reto Stauffer \\
-  Department of Statistics \\
-  Faculty of Economics and Statistics \\
-  Universit\"at Innsbruck \\
-  Universit\"atsstr.~15 \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/stauffer/reto-stauffer.html}\\
-
-  Achim Zeileis \\
-  Department of Statistics \\
-  Faculty of Economics and Statistics \\
-  Universit\"at Innsbruck \\
-  Universit\"atsstr.~15 \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Achim.Zeileis@R-project.org} \\
-  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
 
@@
 Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
+\\
 
@@
 {
+  set.seed(7)
   ## disttree
@@
 plot(x = x, y = y1, type = "l", col = "red", 
-     main = paste0("July ", pday, dayending), ylab = "Density", 
+     main = paste0("July ", pday), ylab = "Density", 
      xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
@@
 @
-\caption{\label{axams_onestation}Predictions at station Axams for one day in July of 2009, 2010, 2011 and 2012.
+\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest model learned on data from 1985--2008.
 Observations and predictions are left censored at 0.
@@
 
-\newpage
+%\newpage
 \section{Methodology}
@@
 In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
-(The power transformation is used to remove large parts of the skewness of the data as further explained in \ref{setting}.) Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
+(The power transformation is used to remove large parts of the skewness of the data as further explained in \ref{data}.) Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
 $$
@@
 The exact way of determining how and where to split varies across different tree algorithms.
-(For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm or the ctree algorithm.) Summed up, the steps of building a distributional tree using one of these algorithms are
+For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm (\citealt{zeileis2008model}) or the ctree algorithm (\\citealt{hothorn2006unbiased}).
+Summed up, the steps of building a distributional tree using one of these algorithms are
 \begin{enumerate}
@@
 \SweaveOpts{eval = TRUE}
-\newpage
-\newpage
-\section{Probabilistic Forecasting of Precipitation}
+%\newpage
+%\newpage
+\section{Probabilistic Forecasting of Precipitation in Complex Terrain}
 \label{ProbabilisticForecasting}
@@
 
-\subsection{Setting}
-\label{setting}
+\subsection{Data}
+\label{data}
 The data set used for training and validation consists of observed daily precipitation sums provided by the National Hydrographical Service (\citealt{ehyd}) and numerical weather forecasts from the {U.S.} National Oceanic and Atmospheric Administration (NOAA).
@@
 Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std dev, ensemble minimum and ensemble maximum.
-All of these values can be calculated on the sums over 24 hours (6--30 UTC) or a 6 hour time window.\\
+All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
 Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std dev of mean/minimal/maximal values respectively.\\
@@
 \hline
-\textbf{tp}: total precipitation            & 12 variations:  & ensemble mean of sums over 24h, \\
+\textbf{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
 \hspace*{\fill} power transformed (by $\frac{1}{1.6}$) &    & ensemble std dev of sums over 24h, \\ 
@@
 \hline
-\textbf{dswrf}: downwards short ware      & 6 variations: & ensemble mean of mean values, \\
+\textbf{dswrf}: downwards short ware      & 6 & ensemble mean of mean values, \\
 \hspace*{\fill} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
@@
 \textbf{pwat}: preciptable water          &   & ensemble std dev of mean values,\\
-\textbf{tmax}: 2m maximum                 &   & ensemble std dev of minimal values,\\
-\hspace*{1.15cm} temperature              &   & ensemble std dev of maximal values,\\
+\textbf{tmax}: 2m maximum temperature     &   & ensemble std dev of minimal values,\\
+\hspace*{1.15cm}                          &   & ensemble std dev of maximal values,\\
 \textbf{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
 \hspace*{1.05cm} condensate               &   & \\
-\textbf{t500}: temperature   &   & \\
-\hspace*{1cm} on 500 hPa     &   & \\
-\textbf{t700}: temperature   &   & \\
-\hspace*{1cm} on 700 hPa     &   & \\
-\textbf{t850}: temperature   &   & \\
-\hspace{1cm} on 850 hPa      &   & \\
+\textbf{t500}: temperature on 500 hPa     &   & \\
+\hspace*{1cm}                             &   & \\
+\textbf{t700}: temperature on 700 hPa     &   & \\
+\hspace*{1cm}                             &   & \\
+\textbf{t850}: temperature on 850 hPa     &   & \\
+\hspace{1cm}                              &   & \\
 \hline
-\textbf{tdiff500850}: temperature         & 3 variations: & ensemble mean of difference in mean,\\
+\textbf{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
 \hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
@@
 \hline
-\textbf{msl{\_}diff}: mean sea level  & 1 variation: & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
+\textbf{msl{\_}diff}: mean sea level  & 1 & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
 \hspace{1.65cm} difference            &   & \qquad over 6--30 UTC\\
@@
 \end{tabular}
-\caption[Table caption text]{Basic covariates together with the number and the type of variations.}
+\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
 \label{covariates}
@@
 \hline
- & \textbf{Type} & \textbf{Location ($\boldsymbol{\mu}$)} & \textbf{Scale ($\boldsymbol{\log(\sigma)}$)} \\
+Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$) \\
 \hline
@@
 \\
+For the purpose of comparing the performance of a distributional forest to those of other models the so-called skills score can be calculated with respect to the CRPS and one model being chosen as a reference.
+In that way we obtain a measure of how much a method improves the model performance in terms of CRPS values compared to the reference method.
+$$
+\text{SkillsScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
+$$
 
 %\SweaveOpts{eval = TRUE}
-\subsection{Application on one station}
+\subsection{Application for one station}
 First of all we focus on one observation station, Axams in Tyrol.
@@
 To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied.
-The 10 leading covariates of the above described forest model for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}.
-As expected, the numerical forecast for total precipitation is the most frequently selected.
+Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described forest model for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}.
+As expected, the numerical forecast for total precipitation is the most influential variable.
 In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.
@@
 \begin{center}
-\begin{subfigure}{0.7\textwidth}
+%\begin{subfigure}{0.7\textwidth}
 <<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
@@
 barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
-        horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE, 
+        horiz = TRUE, las = 1, cex.names = 1.2, axes = TRUE, 
         names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
-@
-\end{subfigure}
+#axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(-1,0.5,0))
+@
+%\end{subfigure}
 %\hspace{-1cm}
@@
 %\end{subfigure}
-\caption{\label{varimp} Variable importance: Top 10 list for the forest model.}
+\caption{\label{varimp} Variable importance based on mean decrease in accuracy regarding the CRPS: Top 10 list for the forest model.}
 \end{center}
@@
 \begin{center}
-\begin{subfigure}{0.8\textwidth}
+%\begin{subfigure}{0.8\textwidth}
 <<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
@@
 @
-\end{subfigure}
-\caption{\label{pit_oosample} PIT histograms for station Axams (out of sample).}
+%\end{subfigure}
+\caption{\label{pit_oosample}Out-of-sample PIT histograms for station Axams assessed in 2009--2012 based on models learned on data from 1985--2008.}
 \end{center}
@@
 %\subsubsection{Cross-Validation}
-After analyzing the results of the models on one data set only where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
+\indent After analyzing the results of the models on one data set only where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
 For that reason a cross validation framework is considered for the station Axams.
@@
 This whole procedure is repeated 10 times.
-\\
-As we are interested in comparing the performance of the distributional forest to those of other models, we choose one model as a reference and calculated the skills score with respect to the CRPS.
-In that way we obtain a measure of how much a method improves the model performance in terms of CRPS values compared to the reference method.
-$$
-\text{SkillsScore(method)} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
-$$
 The resulting skills scores are illustrated in Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model represented by the horizontal line at height 0.
@@
 \center
-\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 7x10 cross validation at station Axams.}
+\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 10 times 7-fold cross validation at station Axams.}
 \end{center}
@@
 In this particular setting, a very high number of covariates is available which intensifies / increases the impact of the above mentioned strength of the forest model.\\
-Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in performance.
-
-\SweaveOpts{eval = FALSE}
+Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
+
+\SweaveOpts{eval = TRUE}
 %\newpage
-\subsection{Application on all stations}
+\subsection{Application for all stations}
 Until now the focus has been on one observation station only.
@@
 In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
-The blue line represents the results for the station Axams.
+The green line and points represent the results for the station Axams.
 %NAs in gamlss in 13 stations!!!\\
@@
 
-ll_all <- res[[1]]$results["ll",]
-for(i in 2:(length(res)-1)) ll_all <- rbind(ll_all, res[[i]]$results["ll",])
-
-rmse_all <- res[[1]]$results["rmse",]
-for(i in 2:(length(res)-1)) rmse_all <- rbind(rmse_all, res[[i]]$results["rmse",])
+#ll_all <- res[[1]]$results["ll",]
+#for(i in 2:(length(res)-1)) ll_all <- rbind(ll_all, res[[i]]$results["ll",])
+
+#rmse_all <- res[[1]]$results["rmse",]
+#for(i in 2:(length(res)-1)) rmse_all <- rbind(rmse_all, res[[i]]$results["rmse",])
 
@@
 
-colnames(ll_all) <- colnames(rmse_all) <- colnames(crps_all) <- colnames(res[[1]]$results)
+#colnames(ll_all) <- colnames(rmse_all) <- 
+  colnames(crps_all) <- colnames(res[[1]]$results)
 
@@
 brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
+#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
 grp <- cut(dst, breaks = brk)
@@
   matplot(t(s[,]), type = "l", lwd = 2, 
-          col = c(rep(gray(0.5, alpha = 0.3),69),
-                  #gray(0.1, alpha = 0.8), 
-                  "blue",
-                  rep(gray(0.5, alpha = 0.3),15)), 
+          col = gray(0.5, alpha = 0.2),
           lty = 1, axes = FALSE, 
           xlab = "", ylab = "CRPS skills score", xlim = c(0.5, 3.5))
+lines(s[70,], col = "limegreen", type = "o", pch = 19, lwd = 2)  
 # Station Axams is the 77th station which is the 70th in the list of complete stations
@@
 
-  par(mar = c(5,0.2,5,2.3))
+  par(mar = c(0.5,0.2,0.5,2.3))
   ## legend
@@
 \section*{Computational Details}
-The corresponding implementation to the novel methods presented in this article is available on R-Forge in an \textsf{R}-package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}(.
-It uses the functions \code{mob} (\citealt{zeileis2010party}() and \code{ctree} (\citealt{hothorn2006unbiased}), both being provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}), as a framework for building trees and forests.
+The corresponding implementation to the novel methods presented in this article is available on R-Forge in an \textsf{R} package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}).
+The function \code{disttree} uses either the \code{mob} or \code{ctree} function as a framework to build a tree.
+Similarly the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a forest.
+All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}).
 
@@
 
-\newpage
-\newpage
+%\newpage
+%\newpage
 \bibliography{ref.bib}

### 2017-11-24 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1757 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 80fa692cfa1a87e6b2d6eb4ca6acabd3ffcd6f78

@@
 In the field of probabilistic forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools.
-This is due to their simplicity and their interpretability.
+This is due to their simplicity and interpretability.
 However, the restriction to model only linear effects requires further extensions which allow for more flexibility.
@@
 Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge/experience from meteorology or based on a computationally more demanding boosting approach.
+Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
 
@@
 A typical application in probabilistic forecasting in meteorology is the analysis of precipitation data.
-The two main interests lie in estimating the probability of precipitation as well as predicting the expected amount of precipitation.
+The main interest lies in estimating the probability of precipitation as well as predicting the expected amount of precipitation.
 In this research area various methods have been developed and tested in the last decades.
@@
 In Figure~\ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.\\
-These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth/homogeneous setting, e.g.
+These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g.
 areas where no abrupt changes in meteorological/climatological events appear.
-However, for example in complex terrain/mountainous regions various factors can change rapidly and their effects are difficult to be modeled/represented by a smooth parametric model.
+However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model.
 In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
@@
 \end{tikzpicture}
-\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests}
+\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests.}
 \end{figure}
@@
 
+\newpage
 \subsection{Distributional Fit}
@@
 Moreover, if covariates are available it is desirable to include them into the model as they might provide important additional information.
-In our application on precipitation data a wide range of covariates are available from numerical ensemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions.
+In our application on precipitation data a wide range of covariates are available from numerical ensemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest and have an influence on the predictions.
 \\
@@
 This enables the model to be sensitive to changes in each of the parameters which is one of the main advantages of distributional trees over other tree models considering only the mean of the distribution.\\
-Statistical tests are applied to check whether there is a significant connection/dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$.
-If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest connection/dependency.\\
+Statistical tests are applied to check whether there is a significant dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$.
+If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest dependency.\\
 In a next step a local distributional model is fit in each of the subgroups.
@@
 The exact way of determining how and where to split varies across different tree algorithms.
-For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm (\citealt{zeileis2008model}) or the ctree algorithm (\\citealt{hothorn2006unbiased}).
+For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm (\citealt{zeileis2008model}) or the ctree algorithm (\citealt{hothorn2006unbiased}).
 Summed up, the steps of building a distributional tree using one of these algorithms are
@@
 However, in this case this separation can also be a disadvantage as it hinders the capturing of interactions unless they are specified explicitly by the user.
-These two examples/situations provide once again a good reason to combine the two frameworks of flexible parametric modeling and tree structured models in order to profit from their advantages and compensate possible disadvantages.
+These two cases provide once again a good reason to combine the two frameworks of flexible parametric modeling and tree structured models in order to profit from their advantages and compensate possible disadvantages.
 This is the pursued goal in introducing distributional forests.
@@
 Next to distributional forests three other methods are applied to this data set and their results are compared based on CRPS values.
-One of the reference models is an EMOS models as this is a common choice in this type of setting / research field.
+One of the reference models is an EMOS models as this is a common choice in this type of setting and research field.
 The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameters and once applying a boosting algorithm).
@@
 par(mfrow = c(2,2))
-pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Forest")
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM")
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM")
-pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS")
+pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Forest", ylim = c(0,1.4))
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM", ylim = c(0,1.4))
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM", ylim = c(0,1.4))
+pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.4))
 
@@
 To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decisions.
-Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not match / correspond to the truth.\\
+Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not correspond to the truth.\\
 This risk is avoided in the boosted GAM as all covariates are included and the selection of the influential covariates is done within a boosting framework.
@@
 Variable selection is done automatically and interactions are detected by the tree structure of the model.\\
-In this particular setting, a very high number of covariates is available which intensifies / increases the impact of the above mentioned strength of the forest model.\\
+In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the forest model.\\
 Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
@@
 Figure~\ref{map} shows the location of all stations in the data set.
-Each observation station is represented by a point colored according to the method that performed best in terms of CRPS at this station.
-For those points where the distributional forest is not the best performing method the intensity of the color represents how much better (in terms of CRPS) the method is compared to the distributional forest.
-The more intense the color, the bigger the difference.
-In case of the forest being the best model, the second best method is chosen as reference.
+Each observation station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station.
+The color of the symbol indicates the difference of the CRPS of the forest model and the CRPS of the best perfoming among the other three models (prespecified GAM, boosted GAM and EMOS).
 
@@
 The function \code{disttree} uses either the \code{mob} or \code{ctree} function as a framework to build a tree.
-Similarly the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a forest.
+Similarly, the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a forest.
 All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}).

### 2017-11-27 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1758 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 81caa5ae1744d544d82350e6d40ef169a35eb40a

@@
 \section{Introduction}
-A typical application in probabilistic forecasting in meteorology is the analysis of precipitation data.
+A typical application in meteorology is to improve probabilistic precipitation forecasts.
 The main interest lies in estimating the probability of precipitation as well as predicting the expected amount of precipitation.
@@
 Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
-\\
+In section \ref{ProbabilisticForecasting} this example will be further discussed and it will be shown that the predictions made by the forest model are reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, the scale parameter, and therefore the whole distribution.
 
@@
 In \citealt{Gebetsberger2017} this choice has been made in order to be able to deal with heavy tails which often appear in precipitation data.
-Another frequent application of probabilistic forecasting in meteorology is the modeling of temperature.
+Another frequent application of probabilistic forecasting in meteorology is the modeling of near surface air temperature.
 In this case a normal distribution is often selected.\\
@@
 Moreover, if covariates are available it is desirable to include them into the model as they might provide important additional information.
-In our application on precipitation data a wide range of covariates are available from numerical ensemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest and have an influence on the predictions.
+In our application on precipitation data a wide range of covariates are available from a numerical ensemble prediction system considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest and have an influence on the predictions.
 \\
@@
 However, the restriction to additive models remains.
-Therefore, a different approach of how to include covariates that also allows for non-additive effects and interactions is considered in the next subsection of this article.
+Therefore, a different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the next subsection of this article.
 In particular, the covariates can be used to separate the data into more homogeneous subgroups.
@@
 if they depend on the location parameter.
-Therefore, building a forest of distributional trees which can also capture separate changes in each of the distribution parameters leads to an improvement of the model.
-This valuable feature of detecting separate changes is also provided by GAMLSS as each parameter is modeled separately.
+Therefore, building a forest of distributional trees which can also capture separate effects in each of the distribution parameters leads to an improvement of the model.
+This valuable feature of detecting separate effects is also provided by GAMLSS as each parameter is modeled separately.
 However, in this case this separation can also be a disadvantage as it hinders the capturing of interactions unless they are specified explicitly by the user.
@@
 $\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
-For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each observation $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$.
+For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$.
 $$
@@
 where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
-This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way, leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
+This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
 $$
@@
 $$
-While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates which is due to the individual weights calculated separately for each set of covariates.\\
+While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates due to the individual weights calculated separately for each set of covariates.\\
 \\
 As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree structured modeling it provides a model that can deal with a wide range of situations.
-Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably by combining various tree and smoothing the steps imposed by a single tree.
+Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably well by combining various tree and smoothing the steps imposed by a single tree.
 As scores are considered for each distribution parameter (separate) effects in different parameters can be captured.
 At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.\\
-Even though many modeling situations demand for the use of a specific model providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance what kind of situation this is.
+Even though many situations demand for the use of specificaly taylored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
 In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.
@@
 As covariates the second generation reforecast data set of the global ensemble forecast system (GEFS) is used (\citealt{hamill2013}).
-This data set consists of an 11-member ensemble based on a fixed version of the numerical model and a horizontal grid-spacing of about $50~\times~50 \text{km}^2$ initialized at 0~UTC from December 1984 to present.
+This data set consists of an 11-member ensemble based on a fixed version of the numerical model and a horizontal grid-spacing of about $50~\times~50 \text{km}^2$ initialized daily at 0~UTC from December 1984 to present providing forecasts on 26 hourly temporal resolution.
 Each of the 11 ensemble members uses slightly different perturbed initial conditions trying to predict the situation specific uncertainty of the atmospheric state.
@@
 In \citealt{stauffer2017a} optimization methods have been applied to find the best transformation factor which resulted in a value of $\frac{1}{1.6}$.
-Since the data used for this optimization covers almost the same geographical region as considered in this article the same factor is chosen here for the power transformation of the response variable \texttt{tp} (total precipitation).
+Since the observation data set used for this optimization contains the observations considered in this article the same factor is chosen here for the power transformation of the response variable \texttt{tp} (total precipitation).
 The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
@@
 \hline
-\textbf{dswrf}: downwards short ware      & 6 & ensemble mean of mean values, \\
+\textbf{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
 \hspace*{\fill} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
@@
 \hline
-\textbf{msl{\_}diff}: mean sea level  & 1 & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
-\hspace{1.65cm} difference            &   & \qquad over 6--30 UTC\\
+\textbf{msl{\_}diff}: mean sea level pressure & 1 & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
+\hspace{1.45cm} difference            &   & \qquad over 6--30 UTC\\
 \hline
@@
 \subsection{Models and Evaluation}
-Next to distributional forests three other methods are applied to this data set and their results are compared based on CRPS values.
-One of the reference models is an EMOS models as this is a common choice in this type of setting and research field.
-The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameters and once applying a boosting algorithm).
+Next to distributional forests three other methods are applied to this data set and their results are compared based on the continuous ranked probability score (CRPS).
+One of the reference models is an EMOS model as this is a common choice in this type of setting and research field.
+The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameter and once applying a boosting algorithm.
 The chosen methods and their specifications are listed in Table~\ref{model_specification}.
@@
 where \(\mathbf{1}\) is the indicator function. \\
-In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model and the mean CRPS over this data set is used for validation.
+In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model.
+The mean CRPS over thi data set is used for validation.
 Since the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
@@
 For the purpose of comparing the performance of a distributional forest to those of other models the so-called skills score can be calculated with respect to the CRPS and one model being chosen as a reference.
-In that way we obtain a measure of how much a method improves the model performance in terms of CRPS values compared to the reference method.
+In that way we obtain a measure of how much a method improves the model performance in terms of CRPS compared to the reference method.
 $$
@@
 %\end{subfigure}
-\caption{\label{varimp} Variable importance based on mean decrease in accuracy regarding the CRPS: Top 10 list for the forest model.}
+\caption{\label{varimp} Variable importance of the forest model for station Axams learned on data from 1985--2008 and assessed in 2009--2012 based on mean decrease in accuracy regarding the CRPS (top 10 list).}
 \end{center}
@@
 The CRPS skills score boxplot shows that, in this setting, both GAM models and the forest perform better than the EMOS model.
-While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the forest even reaches an improvement rate of close to 6 percent.\\
-From this result we can conclude that the distributional forest provides an alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.\\
-For the specified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
+While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the forest even reaches an improvement rate close to 6 percent.\\
+From this result we can conclude that the distributional forest provides a powerful alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.\\
+For the prespecified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
 To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decisions.
@@
 In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
-The green line and points represent the results for the station Axams.
+The green line represents the results for the station Axams.
 %NAs in gamlss in 13 stations!!!\\
@@
 @
-\caption{\label{boxplot_crps_all}CRPS skills score with reference model EMOS for predictions over all stations.}
+\caption{\label{boxplot_crps_all}CRPS skills score over all stations with reference model EMOS. The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
 \end{center}

### 2017-11-27 pkg/disttree/inst/CMStatistics2017/slides.Rnw
slides for CMStatistics 2017 in London
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1759 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2da9a09935c92c5cfb76ed4efb7b17250e26aeb1

--- 2017-06-28 pkg/disttree/inst/useR2017/slides.Rnw
@@
 \usetheme[nototalframenumber, noslidenumber]{uibk}
-%\setfaculty{volkswirtschaftundstatistik}{en}
 
 \title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn, \\
-Heidi Seibold, Achim Zeileis}
-\URL{http://R-Forge.R-project.org/projects/partykit}
-\Conference{useR! 2017}
-\date{2017-07-05}
+\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Achim Zeileis}
+\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
+\URL{https://R-Forge.R-project.org/projects/partykit/}
+\headerimage{3}
 
@@
 \usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-%\usepackage[style=authoryear, backend=bibtex]{biblatex}
 \graphicspath{{plots/}}
@@
 \definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
-
-%% need no \usepackage{Sweave}
-\definecolor{InputColor}{rgb}{0,0,0.3}
-\definecolor{OutputColor}{rgb}{0.2,0.2,0.2}
 
@@
                            startfun = startfun,
-                           mle = mle
+                           mle = mle,
+                           gamlssobj = FALSE,
+                           censored = FALSE
   )
@@
                                 startfun = startfun,
-                                mle = mle
+                                mle = mle,
+                                gamlssobj = FALSE,
+                                censored = TRUE
   )

### 2017-11-27 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
slides for CMStatistics 2017 in London
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1759 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2da9a09935c92c5cfb76ed4efb7b17250e26aeb1

--- 2017-11-27 pkg/disttree/inst/CMStatistics2017/slides.Rnw
@@
 ## HCL palette
-pal <- hcl(c(10, 128, 260), 100, 50)
-names(pal) <- c("forest", "tree", "gamlss")
-
-pallight <- hcl(c(10, 128, 260), 100, 50, alpha = 0.25)
-names(pallight) <- c("forest", "tree", "gamlss")
+pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
+names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
+
+pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
+names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
 
@@
 
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace*{0.2cm}
+Data:
+\begin{itemize}
+\item Daily observations from 95 stations: \\
+24h precipitation sums from July of 28 years.
+\item Numerical ensemble predictions of meteorological quantities\\
+\vspace{0.1cm}
+$\rightarrow$ Covariates:\\
+\begin{itemize}
+\item Total precipitation
+\item Convective available potential energy
+\item Downwards short wave radiation flux (``sunshine'')
+\item Mean sea level pressure 
+\item Mean sea level pressure difference
+\item Preciptable water
+\item 2m maximum temperature
+\item Total column-integrated condensate
+\item Temperature 
+\item Temperature differences in altitude
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace*{0.2cm}
+Application for one station:
+\begin{itemize}
+\item Learn forest model on data from 24 years (1985--2008).
+\item Evaluate on 4 years (2009--2012).
+\end{itemize}
+\end{frame}
+
+<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
+#setwd("~/svn/partykit/pkg/disttree/inst/CMStatistics2017")
+load("data_Axams.rda")
+
+# tree and forest formula
+{
+  dt.formula <- df.formula <- 
+      robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
+      tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
+      tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
+      capepow_mean + capepow_sprd + capepow_min + capepow_max + 
+      capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
+      capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
+      dswrf_mean_mean + dswrf_mean_max +  
+      dswrf_sprd_mean + dswrf_sprd_max + 
+      msl_mean_mean + msl_mean_min + msl_mean_max + 
+      msl_sprd_mean + msl_sprd_min + msl_sprd_max +
+      pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
+      pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
+      tmax_mean_mean + tmax_mean_min + tmax_mean_max +
+      tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
+      tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
+      tcolc_sprd_mean + tcolc_sprd_min + tcolc_sprd_max +
+      t500_mean_mean + t500_mean_min + t500_mean_max +
+      t700_mean_mean + t700_mean_min + t700_mean_max +
+      t850_mean_mean + t850_mean_min + t850_mean_max +
+      t500_sprd_mean + t500_sprd_min + t500_sprd_max +
+      t700_sprd_mean + t700_sprd_min + t700_sprd_max +
+      t850_sprd_mean + t850_sprd_min + t850_sprd_max +
+      tdiff500850_mean + tdiff500850_min + tdiff500850_max +
+      tdiff700850_mean + tdiff700850_min + tdiff700850_max +
+      tdiff500700_mean + tdiff500700_min + tdiff500700_max +
+      msl_diff
+
+  }
+
+learndata <- data_axams[data_axams$year < 109,]
+testdata <- data_axams[data_axams$year %in% c(109, 110, 111, 112),]
+
+##############################################################
+# fitting the models
+set.seed(7)
+
+df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
+                 type.tree = "ctree", ntree = 100, censtype = "left", censpoint = 0,
+                 control = ctree_control(teststat = "quad", testtype = "Univ",
+                                         intersplit = TRUE,
+                                         mincriterion = 0, minsplit = 50,
+                                         minbucket = 20), mtry = 27)
+
+#### prepare data for plot of estimated density functions
+# predictions for one day (in each of the four years) 
+# (19th of July 2011 is missing)
+pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
+pdf <- predict(df, newdata = testdata[pdays,], type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+cbind(df_exp, testdata[pdays,"robs"])
+
+# plot predicted distributions together with observations
+#set.seed(res$call$seedconst)
+set.seed(7)
+x <- c(0.01, sort(runif(500,0.01,8)))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
+dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
+
+# point mass (slightly shifted)
+pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
+pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
+pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
+
+# predictions
+pred1 <- c(testdata[pdays,"robs"][1], crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(testdata[pdays,"robs"][2], crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(testdata[pdays,"robs"][3], crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(testdata[pdays,"robs"][4], crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+
+#legendheight
+lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
+lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
+lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
+@
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\begin{center}
+\vspace*{-0.2cm}
+<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=5.2>>=
+plot(x = x, y = y1, type = "l", col = "red", 
+     main = paste0("Axams, July ", pday), ylab = "Density", 
+     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
+     #xlab = expression(Total~precipitation~group("[", mm^(1/1.6)~per~day,"]")),
+     #{\frac{1}{1.6}}$,
+     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
+     xlim = c(-1.5,8))
+
+lines(x = x, y = y2, type = "l", col = "blue")
+lines(x = x, y = y3, type = "l", col = "darkgreen")
+lines(x = x, y = y4, type = "l", col = "purple")
+legend('topright', c("predicted distribution", "point mass at censoring point", "observation"),
+       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
+
+# plot point mass
+lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
+lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
+lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
+lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)
+
+points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
+points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
+points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
+points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)
+
+# plot predictions
+points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
+points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
+points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
+points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
+
+lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)
+
+# add labels
+text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
+text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
+text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
+text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace*{0.2cm}
+Application for one station:
+\begin{itemize}
+\item Learn forest model on data from 24 years.
+\item Evaluate on 4 years.
+\item 10 times 7-fold cross validation.
+\item Compare to commonly used methods:\\
+\begin{itemize}
+\item Prespecified GAM
+\item Boosted GAM
+\item Ensemble Model Output Statistics (EMOS) 
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+<<echo=FALSE, results=hide>>=
+#### cross validation rain
+if(file.exists("rain_Axams_7x10.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_7x10.rda")
+  load("rain_Axams_7x10.rda")
+} else {
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross_7x10.R")
+  source("rain_cross_7x10.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+  rainres <- rain_cross(stationname = "Axams", 
+                        seedconst = 7, ntree = 100,
+                        forest_mtry = 27,
+                        gamboost_cvr = TRUE,
+                        frac = FALSE)
+
+  save(rainres, file = "rain_Axams_7x10.rda")
+}
+
+rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres)-1)
+
+for(i in 1:(length(rainres)-1)){
+  rain_crps[i,] <- colMeans(rainres[[i]]$crps, na.rm = TRUE)
+}
+
+colnames(rain_crps) <- colnames(rainres[[1]]$rmse)
+
+@
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\begin{center}
+\vspace*{-0.2cm}
+<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.2>>=
+boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
+        main = "Cross Validation (with reference model EMOS)",
+        names = c("Forest", "prespecified GAM", "boosted GAM"),
+        ylab = "CRPS skills score", col = "lightgray") 
+abline(h = 0, col = pal["EMOS"], lwd = 2)
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace*{0.2cm}
+Application for all stations:
+\begin{itemize}
+\item Learn forest model on data from 24 years (1985--2008).
+\item Evaluate on 4 years (2009--2012).
+\item Compare to commonly used methods:\\
+\begin{itemize}
+\item Prespecified GAM
+\item Boosted GAM
+\item Ensemble Model Output Statistics (EMOS) 
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+<<echo = FALSE>>=
+#### prediction over all stations 24 - 4
+if(file.exists("rain_pred_24to4.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
+  load("rain_pred_24to4.rda")
+} else {
+
+  source("rain_pred_24to4.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+
+  res <- rain_pred(seedconst = 7,
+                   gamboost_cvr = TRUE,
+                   frac = FALSE)
+
+  save(res, file = "rain_pred_24to4.rda")
+}
+
+crps_all <- res[[1]]$results["crps",]
+for(i in 2:(length(res)-1))  crps_all <- rbind(crps_all, res[[i]]$results["crps",])
+
+colnames(crps_all) <- colnames(res[[1]]$results)
+
+# skills score
+s <- 1 - crps_all[, 2:4]/crps_all[,6]
+colnames(s) <- c("Forest", "prespecified GAM", "boosted GAM")
+
+## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
+
+crps <- crps_all[,c("distforest", "gamlss", "gamboostLSS", "EMOS log")]
+
+## best method
+bst <- apply(crps, 1, which.min)
+
+## distance of forest to best other method
+dst <- crps[,1] - crps[cbind(1:nrow(crps), apply(crps[, -1], 1, which.min) + 1)]
+
+## breaks/groups
+brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
+#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
+grp <- cut(dst, breaks = brk)
+
+## HCL colors (relatively flashy, essentially CARTO Tropic)
+clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
+
+library("raster") # dem (digital elevation model)
+library("sp")     # gadm www.gadm.org/country
+
+load("../draft/plot_map_rain/demo/tirol.gadm.rda")
+load("../draft/plot_map_rain/demo/tirol.dem.rda")
+load("../draft/plot_map_rain/demo/ehyd.statlist.rda")
+
+# Create SpatialPointsDataFrame from station list
+sp <- SpatialPointsDataFrame(subset(ehyd.statlist[res$complete_stations,],
+                                    select=c(lon,lat)),
+                             data = subset(ehyd.statlist[res$complete_stations,],
+                                           select = -c(lon,lat)),
+                             proj4string = crs(tirol.dem))
+
+@
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\begin{center}
+\vspace*{0cm}
+
+<<map, fig=TRUE, echo=FALSE, width=10, height = 6.5>>=
+
+layout(cbind(1, 2), width = c(9, 1))
+par(mar = c(5,4,4,0.1))
+raster::image(tirol.dem, col = rev(gray.colors(100)),
+              main="Observation Stations", ylab = "Latitude", xlab = "Longitude", 
+              xlim = c(9.8,13.2), 
+              ylim = c(46.6, 47.8), 
+              cex.lab=1.5, cex.axis=1.2, cex.main=2)
+plot(tirol.gadm, add = TRUE)
+points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.9)
+legend("topleft", pch=c(21,24,25,22), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1.2, bty = "n")
+
+par(mar = c(0.5,0.2,0.5,2.3))
+## legend
+plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
+     xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
+rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
+axis(4, at = brk, las = 1, mgp=c(0,0,-1.2), cex = 1.4)
+
+@
+\end{center}
+\end{frame}
+
 %\begin{frame}[fragile]
@@
 
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
-plot_rmse(simlist, type = "par")
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=
-plot_rmse(simlist, type = "exp")
-@
-\end{center}
-\end{frame}
-
 \end{document}

### 2017-11-28 pkg/partykit/inst/extree/ctree.R
1.2-0 with extree
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1775 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8f197995840bfb2534c6f1b15e454b3a85dd1892

@@ -1,593 +1,0 @@
-
-.ctree_test <- function(model, trafo, data, subset, weights, j, SPLITONLY = FALSE, ctrl) {
-
-    ix <- data[[j, type = "index"]]
-    iy <- data[["yx", type = "index"]]
-    Y <- model$estfun
-
-    if (!is.null(iy)) {
-        stopifnot(NROW(levels(iy)) == (NROW(Y) - 1))
-        return(.ctree_test_2d(data = data, j = j, Y = Y, iy = iy, 
-                              subset = subset, weights = weights, 
-                              SPLITONLY = SPLITONLY, ctrl = ctrl))
-    }
-
-    stopifnot(NROW(Y) == length(ix))
-
-    NAyx <- data[["yx", type = "missing"]]
-    NAz <- data[[j, type = "missing"]]
-    if (ctrl$MIA) {
-        subsetNArm <- subset[!(subset %in% NAyx)]
-    } else {
-        subsetNArm <- subset[!(subset %in% c(NAyx, NAz))]
-    }
-
-    return(.ctree_test_1d(data = data, j = j, Y = Y, subset = subsetNArm, 
-                          weights = weights, SPLITONLY = SPLITONLY, ctrl = ctrl))
-}
-
-.ctree_split <- function(model, trafo, data, subset, weights, j, SPLITONLY = TRUE, ctrl)
-    .ctree_test(model, trafo, data, subset, weights, j, SPLITONLY, ctrl)
-
-.partysplit <- function(varid, breaks = NULL, index = NULL, right = TRUE, 
-                        prob = NULL, info = NULL) {
-    ret <- list(varid = varid, breaks = breaks, index = index, right = right, 
-                prob = prob, info = info)
-    class(ret) <- "partysplit"
-    ret
-}
-
-.ctree_test_1d <- function(data, j, Y, subset, weights, SPLITONLY = FALSE, ctrl) {
-
-    x <- data[[j]]
-    MIA <- FALSE
-    if (ctrl$MIA) MIA <- (length(NAs <- data[[j, type = "missing"]]) > 0)
-
-    ### X for (ordered) factors is always dummy matrix
-    if (is.factor(x) || is.ordered(x))
-        X <- data[[j, type = "index"]]
-
-    scores <- data[[j, type = "scores"]]
-    ORDERED <- is.ordered(x) || is.numeric(x)
-
-    ux <- Xleft <- Xright <- NULL
-    
[…]

### 2017-11-28 pkg/partykit/inst/extree/modelparty.R

@@ -1,423 +1,0 @@
-## variable selection: given model scores, conduct
-## all M-fluctuation tests for orderins in z
-.mfluc_test <- function(model, trafo, data, subset, weights, j, SPLITONLY, ctrl)
-{
-
-  stopifnot(!SPLITONLY)
-  stopifnot(is.null(data[["yx", type = "index"]]))
-  z <- data[[j]][subset]
-  estfun <- model$estfun[subset,,drop = FALSE]
-  cluster <- data[["(cluster)"]][subset]
-  if(length(weights) == 0) {
-      weights <- rep(1, NROW(estfun))
-  } else {
-      weights <- weights[subset]
-  }
-  obj <- model$obj
-
-  if(length(unique(z)) < 2L) return(list(statistic = NA, p.value = NA))
-  
-  ## set up return values
-  m <- NCOL(z)
-  pval <- rep.int(NA_real_, m)
-  stat <- rep.int(0, m)
-  ifac <- rep.int(FALSE, m)
-  
-  ## estimating functions (dropping zero weight observations)
-  process <- as.matrix(estfun)
-  ww0 <- (weights > 0)
-  process <- process[ww0, , drop = FALSE]
-  if(!is.null(cluster)) cluster <- droplevels(cluster[ww0]) 
-  weights <- weights[ww0]
-  z <- z[ww0]
-  k <- NCOL(process)
-  n <- NROW(process)
-  nobs <- if(ctrl$caseweights && any(weights != 1L)) sum(weights) else n
-
-  
-  ## stop if all collumn values in process are the same
-  if(all(apply(process, 2, function(x) length(unique(x))) == 1)) 
-    return(list(statistic = NA, p.value = NA))
-  
-  ## scale process
-  process <- process/sqrt(nobs)
-  vcov <- ctrl$vcov
-  if(is.null(obj)) vcov <- "opg"
-  if(vcov != "opg") {
-    bread <- vcov(obj) * nobs
-  }
-  if(vcov != "info") {
-    meat <- if(is.null(cluster)) {
-      crossprod(process/sqrt(weights))
-    } else {
-      crossprod(as.matrix(apply(process/sqrt(weights), 2L, tapply, cluster, sum)))
-    }
-  }
-  ## from strucchange
-  root.matrix <- function(X) {
-    if((ncol(X) == 1L)&&(nrow(X) == 1L)) return(sqrt(X)) else {
-      X.eigen <- eigen(X, symmetric = TRUE)
[…]

### 2017-11-28 pkg/partykit/tests/regtest-cforest.R

@@ -24,7 +24,7 @@ p_party <- predict(cf_party)
 stopifnot(max(abs(p_partykit - p_party)) < .Machine$double.eps)
 
 prettytree(cf_party@ensemble[[1]], inames = names(airq)[-1])
-party(cf_partykit$nodes[[1]], data = airq)
+party(cf_partykit$nodes[[1]], data = model.frame(cf_partykit))
 
 v_party <- do.call("rbind", lapply(1:5, function(i) party::varimp(cf_party)))
 
@@ -56,7 +56,7 @@ p_party <- do.call("rbind", treeresponse(cf_party))
 stopifnot(max(abs(unclass(p_partykit) - unclass(p_party))) < .Machine$double.eps)
 
 prettytree(cf_party@ensemble[[1]], inames = names(iris)[-5])
-party(cf_partykit$nodes[[1]], data = iris)
+party(cf_partykit$nodes[[1]], data = model.frame(cf_partykit))
 
 v_party <- do.call("rbind", lapply(1:5, function(i) party::varimp(cf_party)))
 

### 2017-11-28 pkg/partykit/tests/regtest-party.R

@@ -98,7 +98,7 @@ nt <- node_party(ctree(Species ~ ., data = iris))
 (ntp <- nodeprune(nt, 4)) ### partynode method
 
 ### check if both methods do the same
-p1 <- predict(party(ntp, data = iris), type = "node")
+p1 <- predict(party(ntp, data = model.frame(ct)), type = "node")
 p2 <- predict(ctp, type = "node")
 stopifnot(max(abs(p1 - p2)) == 0)
 

### 2017-11-30 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1783 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b2edc10747e23a82ab71433996ebfedfbd688f12

@@
 At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.\\
-Even though many situations demand for the use of specificaly taylored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
+Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
 In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.
@@
 In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model.
-The mean CRPS over thi data set is used for validation.
+The mean CRPS over this data set is used for validation.
 Since the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
@@
 Each observation station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station.
-The color of the symbol indicates the difference of the CRPS of the forest model and the CRPS of the best perfoming among the other three models (prespecified GAM, boosted GAM and EMOS).
+The color of the symbol indicates the difference of the CRPS of the forest model and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
 

### 2017-11-30 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
started rewrting application slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1784 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7b13f2926dc4e27ba18ac49528154a9ffdae2ac1

@@
 \title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Achim Zeileis}
+\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
@@
 library("disttree")
+library("Formula")
 #library("partykitR1")
@@
 \vspace*{0.2cm}
+Task:
+\begin{itemize}
+\item Goal: Predict total precipitation amount.
+\item Input: Numerical Weather Prediction (NWP) with gridsize ... .
+\item 
+\vspace{0.1cm}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace*{0.2cm}
 Data:
@@
 boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
-        main = "Cross Validation (with reference model EMOS)",
+        main = "Cross validation (with reference model EMOS)",
         names = c("Forest", "prespecified GAM", "boosted GAM"),
@@
 \vspace*{0.2cm}
-Application for all stations:
+Application for all 95 stations:
 \begin{itemize}

### 2017-11-30 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
different introduction to application
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1787 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7dee437966faf1aaed07c41f532ec96595d9037e

@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 
@@
 \vspace*{0.2cm}
-Task:
+%Task:
 \begin{itemize}
 \item Goal: Predict total precipitation amount.
-\item Input: Numerical Weather Prediction (NWP) with gridsize ... .
-\item 
-\vspace{0.1cm}
+\item Input: Numerical Weather Prediction (NWP).
+\item Idea: Statistical Model with numerical predictions as covariates.
+\item Data:
+\begin{itemize}
+\item 95 observation stations in Tyrol.
+\item Daily 24h precipitation sums from July of 28 years (1985--2012).
+\item Numerical ensemble predictions of meteorological quantities.
+\end{itemize}
+\item Distribution Assumption: $(\text{Total precipitation})^\frac{1}{1/6} \sim \textit{c}\matchcal{N}$
 \end{itemize}
@@
 \vspace*{0.2cm}
-Data:
+Variable Selection:
 \begin{itemize}
-\item Daily observations from 95 stations: \\
-24h precipitation sums from July of 28 years.
-\item Numerical ensemble predictions of meteorological quantities\\
-\vspace{0.1cm}
-$\rightarrow$ Covariates:\\
+\item Complex terrain \rightarrow unknown influences.
+\item 85 Covariates:\\
 \begin{itemize}
@@
 \item 10 times 7-fold cross validation.
-\item Compare to commonly used methods:\\
+\item Compare to commonly used heteroscedastic censored Gaussian models:\\
 \begin{itemize}
-\item Prespecified GAM
-\item Boosted GAM
-\item Ensemble Model Output Statistics (EMOS) 
+\item Prespecified GAM: variable selection based on meteorological expert knowledge.
+\item Boosted GAM: automatic variable selection.
+\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
 \end{itemize}

### 2017-12-02 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
added references and edited introduction to application (precipitation forecasts)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1797 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8105fc32dc51df48cc0ba2673eb4610eef9d1460

@@
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
+\setbeamerfont{footnote}{size=\tiny}
+\renewcommand\footnoterule{}
 \URL{https://R-Forge.R-project.org/projects/partykit/}
@@
 \frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
+\vspace*{-0.1cm}
 %Task:
@@
 \item Input: Numerical Weather Prediction (NWP).
-\item Idea: Statistical Model with numerical predictions as covariates.
+\item Idea: Statistical model with numerical predictions as \\
+\hspace{1cm}covariates.
 \item Data:
@@
 \item 95 observation stations in Tyrol.
-\item Daily 24h precipitation sums from July of 28 years (1985--2012).
-\item Numerical ensemble predictions of meteorological quantities.
+\item Daily 24h precipitation sums from July of 28 years (1985--2012)\footnote{provided by the National Hydrographical Service (EHYD)}.
+\item Numerical ensemble predictions of meteorological quantities\footnote{provided by the {U.S.} National Oceanic and Atmospheric Administration (NOAA)\\
+\hspace{0.32cm}using the Global Ensemble Forecast System (GEFS)}: 11 members each using slightly different perturbed initial conditions.
 \end{itemize}
-\item Distribution Assumption: $(\text{Total precipitation})^\frac{1}{1/6} \sim \textit{c}\matchcal{N}$
+\item Distribution Assumption:\\
+\vspace{0.1cm}
+\hspace{1cm}$(\text{Total precipitation})^\frac{1}{1.6} \sim \textit{cens}\mathcal{N}(\mu,\sigma^2)$
 \end{itemize}
@@
 \begin{itemize}
-\item Complex terrain \rightarrow unknown influences.
-\item 85 Covariates:\\
+\item Complex terrain $\rightarrow$ unknown influences.
+\item 82 Covariates deduced from 14 base variables \\
+(variations such as ensemble minimum/maximum/mean/standard deviation):\\
 \begin{itemize}
@@
 \item Evaluate on 4 years (2009--2012).
-\item Compare to commonly used methods:\\
+\item Compare to commonly used heteroscedastic censored Gaussian models:\\
 \begin{itemize}
-\item Prespecified GAM
-\item Boosted GAM
-\item Ensemble Model Output Statistics (EMOS) 
+\item Prespecified GAM: variable selection based on meteorological expert knowledge.
+\item Boosted GAM: automatic variable selection.
+\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{References}
+\frametitle{References (1)}
 
@@
 
+\begin{frame}
+\frametitle{References (2)}
+
+\footnotesize
+
+% Reference for the observation data set
+BMLFUW (2016).
+  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtschaft ({BMLFUW}), {A}bteilung {IV/4} - {W}asserhaushalt}
+  \text{Available at \url{http://ehyd.gv.at}}, 
+  \text{Accessed: {F}ebruary 29 2016}.
+
+\medskip
+
+% Reference for the GFS reforecast version 2 data set 
+Hamill T M, Bates G T, Whitaker J S, Murray D R, Fiorino M, Galarneau T J Jr., Zhu Y, Lapenta W (2013).
+  \dquote{{NOAA}'s Second-Generation Global Medium-Range Ensemble Reforecast Dataset}
+  \textit{Bulletin of the American Meteorological Society}, 
+  \textbf{94}(10), 1553--1565.
+  \doi = {10.1175/BAMS-D-12-00014.1}
+
+\medskip
+
+Messner J W,  Mayr G J, Zeileis A (2016).
+  \dquote{Heteroscedastic Censored and Truncated Regression with {crch}}
+  \textit{The R Journal}, 
+  \textbf{8}(1), 173--181.
+  \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
+
+\medskip
+
+Breiman L (2001).
+  \dquote{Random {Forests}}
+  \textit{Machine Learning}, 
+  \textbf{45}(1), 5--32.
+  \doi{10.1023/A:1010933404324}
+
+\medskip
+
+Hofner B, Mayr A, Fenske N, Schmid M (2017).
+  \dquote{{gamboostLSS}: Boosting Methods for {GAMLSS} Models}
+  \textit{{R} package version 2.0-0},
+  \url{https://CRAN.R-project.org/package=gamboostLSS}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{References (3)}
+
+\footnotesize
+Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
+  \dquote{Calibrated probabilistic forecasting using ensemble model output statistics and minimum CRPS estimation}
+  \textit{Monthly Weather Review}, 
+  \textbf{133}(5), 1098--1118.
+  \doi{https://doi.org/10.1175/MWR2904.1}
+
+\medskip
+
+Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
+  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: Unanimous Predictions, Heavy Tails, and Link Functions}
+  \textit{Monthly Weather Review}, 
+  \textbf{Forthcoming}.
+  \doi{10.1175/MWR-D-16-0388.1}
+
+\medskip
+
+{CGIAR-CSI} (2016).
+  \dquote{{SRTM} 90m Digital Elevation Database {v4.1}}
+  \text{Available at \url{http://srtm.csi.cgiar.org}}, 
+  \text{Accessed: {J}uly 10 2016}.
+
+\medskip
+
+Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
+  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored Additive Regression Model}
+  \textit{International Journal of Climatology}, 
+  \textbf{37}(7), 3264--3275.
+  \doi{10.1002/joc.4913}
+
+\end{frame}
+
 \end{document}

### 2017-12-03 pkg/disttree/inst/draft/draft.Rnw
changed ylim of PIT-histograms
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1800 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7f4575af896e18b07dce31f56c0928d8c12bfe90

@@
 par(mfrow = c(2,2))
-pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Forest", ylim = c(0,1.4))
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM", ylim = c(0,1.4))
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM", ylim = c(0,1.4))
-pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.4))
+pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Forest", ylim = c(0,1.5))
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM", ylim = c(0,1.5))
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM", ylim = c(0,1.5))
+pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
 

### 2017-12-03 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
small changes in application section; extra tree section dropped
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1801 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 148499199a69ac05d6e9457cc8e931fdffa9304d

@@
 \item Estimate $\hat{\theta}$ via maximum likelihood.
-\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
+\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $z_i$.
 %\item Assess whether the \emph{model scores} are associated with
@@
 \end{frame}
-
-\begin{frame}
-\frametitle{Fitting a Tree}
-\vspace*{-0.12cm}
-\begin{center} 
-DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
-\vspace*{-0.21cm}
-<<dgp_tree, echo=FALSE, results=hide>>=
-data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
-names(data) <- c("x","x","x")
-fig <- party(
-  partynode(1L,
-            split = partysplit(2L, breaks = 0.4),
-            kids = list(
-              partynode(2L, info = c(
-                "n = 200",
-                "   True parameters:   ",
-                expression(mu == '4'),
-                expression(sigma == '1')
-              )),
-              partynode(3L,
-                        split = partysplit(3L, breaks = 0.8),
-                        kids = list(
-                          partynode(4L, info = c(
-                            "n = 200",
-                            "   True parameters:   ",
-                            expression(mu == '12'),
-                            expression(sigma == '3')
-                          )),
-                          partynode(5L, info = c(
-                            "n = 100",
-                            "   True parameters:   ",
-                            expression(mu == '4'),
-                            expression(sigma == '3')
-                          )))))),
-  data
-)
-
-node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
-    gp = gpar()) 
-{
-    meta <- obj$data
-    nam <- names(obj)
-    extract_label <- function(node) {
-        if (is.terminal(node)) 
-            return(rep.int("", 2L))
-        varlab <- character_split(split_node(node), meta)$name
-        if (abbreviate > 0L) 
-            varlab <- abbreviate(varlab, as.integer(abbreviate))
-        if (pval) {
-            nullna <- function(x) is.null(x) || is.na(x)
-            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
-                silent = TRUE))
-            pval <- if (inherits(pval, "try-error")) 
-                FALSE
-            else pval
-        }
-        if (pval) {
-            pvalue <- node$info$p.value
-            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
-                paste("p =", round(pvalue, digits = 3L)))
-        }
-        else {
-            plab <- ""
-        }
-        return(c(varlab, plab))
-    }
-    maxstr <- function(node) {
-        lab <- extract_label(node)
-        klab <- if (is.terminal(node)) 
-            ""
-        else unlist(lapply(kids_node(node), maxstr))
-        lab <- c(lab, klab)
-        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
-        lab <- lab[which.max(nchar(lab))]
-        if (length(lab) < 1L) 
-            lab <- ""
-        return(lab)
-    }
-    nstr <- maxstr(node_party(obj))
-    if (nchar(nstr) < 6) 
-        nstr <- "aAAAAa"
-    rval <- function(node) {
-        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
-            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
-            height = unit(3, "lines"), name = paste("node_inner", 
-                id_node(node), sep = ""), gp = gp)
-        pushViewport(node_vp)
-        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
-            seq(0.8, 1, by = 0.01))
-        yell <- sqrt(xell * (1 - xell))
-        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
-        lab <- extract_label(node)
-        fill <- rep(fill, length.out = 2L)
-        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
-            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
-        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
-            "lines"))
-        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
-        #    "lines"))
-        grid.text(lab[2L], y = unit(1, "lines"))
-        if (id) {
-            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
-                "npc"), width = max(unit(1, "lines"), unit(1.3, 
-                "strwidth", nam[id_node(node)])), height = max(unit(1, 
-                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
-            pushViewport(nodeIDvp)
-            grid.rect(gp = gpar(fill = fill[2]))
-            grid.text(nam[id_node(node)])
-            popViewport()
-        }
-        upViewport()
-    }
-    return(rval)
-}
-
-class(node_inner_ext) <- "grapcon_generator"
-
-@
-
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_dgp, fig=TRUE, echo=FALSE>>=
-paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-plot(fig, inner_panel = node_inner_ext,
-     tp_args = list(FUN = identity, width = 18, fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]),
-     drop_terminal = TRUE, tnex = 1.7)
-@
-\end{center}
-\end{frame}
-
-<<tree, echo=FALSE, results=hide>>=
-set.seed(7)
-nobs <- 500
-x <- runif(nobs, 0, 1)
-mu <- sigma <- ytrue <- numeric(length = nobs)
-for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
-for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
-y <- rnorm(nobs, mean = mu, sd = sigma)
-#y <- rcnorm(nobs, mean = mu, sd = sigma, left = 0)
-ytrue <- mu
-data <- data.frame(cbind(y,x, ytrue))
-tree <- disttree(y ~ x, data = data, family = NO(), type.tree = "mob")
-#tree <- disttree(y ~ x, data = data, family = dist_list_cens_normal)
-@
-
-\begin{frame}[fragile]
-\frametitle{Fitting a Tree}
-\begin{center}
-\vspace*{-0.12cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_estpar, fig=TRUE, echo=FALSE>>=
-# function for output in terminal panels
-FUN <- function (x) 
-{
-  cf <- x$coefficients
-  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
-  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
-                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
-}
-
-paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-
-## plot version using FUN and tree of class 'disttree'
-plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
-     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
-     ip_args = list(fill = paltrees[c(2, 3)]))
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Fitting a Tree}
-\begin{center}
-\vspace*{-0.12cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_box, fig=TRUE, echo=FALSE>>=
-plot(as.constparty(tree), tnex = 1.7, drop = TRUE,
-     tp_args = list(fill = paltrees[c(1, 3)], ylines = 1.5), 
-     ip_args = list(fill = paltrees[c(2, 3)]))
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Fitting a Tree}
-\begin{center}
-\vspace*{-0.12cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_dens, fig=TRUE, echo=FALSE>>=
-node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
-                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
-                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
-{
-  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
-  ylines <- 1.5
-  if (is.null(xscale)) xscale <- c(-5.1,22.5)
-  if (is.null(yscale)) yscale <- c(-0.05,0.45)
-  xr <- xscale
-  yr <- yscale
-
-  if (horizontal) {
-    yyy <- xscale
-    xscale <- yscale
-    yscale <- yyy
-  }
-
-  rval <- function(node) {
-    yrange <- seq(from = -20, to = 90)/4
-    ydens <- node$info$object$ddist(yrange)
-
-    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3, 
-                                            widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")), 
-                                            heights = unit(c(1, 1), c("lines", "null"))), 
-                       width = unit(1, "npc"), 
-                       height = unit(1, "npc") - unit(2, "lines"), 
-                       name = paste("node_density",node$id, sep = ""))
-    pushViewport(top_vp)
-    grid.rect(gp = gpar(fill = "white", col = 0))
-    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
-    pushViewport(top)
-    mainlab <- paste(ifelse(id, paste("Node", node$id, "(n = "), "n = "), node$info$nobs, ifelse(id, ")", ""), sep = "")
-
-    grid.text(mainlab)
-    popViewport()
-    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2, 
-                     xscale = xscale, yscale = yscale, 
-                     name = paste("node_density",  node$id, "plot", sep = ""))
-    pushViewport(plot)
-    yd <- ydens
-    xd <- yrange
-    if (horizontal) {
-      yyy <- xd
-      xd <- yd
-      yd <- yyy
-      yyy <- xr
-      xr <- yr
-      yr <- yyy
-      rxd <- rep(0, length(xd))
-      ryd <- rev(yd)
-    } else {
-      rxd <- rev(xd)
-      ryd <- rep(0, length(yd))
-    }
-
-    if (rug) {
-      nodeobs <- node$info$object$y
-      if (horizontal) {
-        grid.rect(x = xscale[1], y = nodeobs , height = 0, width = xscale[1], 
-                  default.units = "native", just = c("right", "bottom"),
-		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
-      } else {
-        grid.rect(x = nodeobs, y = yscale[1], 
-                  width = 0, height = abs(yscale[1]), default.units = "native", 
-                  just = c("center", "bottom"),
-		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
-        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
-        #           default.units = "native")
-        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
-        #           default.units = "native")
-      }
-    }
-
-    grid.polygon(x = c(xd, rxd), y = c(yd, ryd), default.units = "native",
-              gp = gpar(col = "black", fill = fill, lwd = lwd))
-    #grid.lines(x = xd, y = yd, default.units = "native", 
-    #           gp = gpar(col = col, lwd = lwd))
-    grid.xaxis()
-    grid.yaxis()
-    grid.rect(gp = gpar(fill = "transparent"))
-    upViewport(2)
-  }
-  return(rval)
-}
-
-class(node_density) <- "grapcon_generator"
-
-plot(tree, tnex = 1.7, drop = TRUE,
-     terminal_panel = node_density,
-     tp_args = list(fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]))
-@
-\end{center}
-\end{frame}
-%\begin{frame}
-%\frametitle{Fitting a Tree}
-%\center
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{tree_box}
-%\end{frame}
-
-<<treedata, echo=FALSE, results=hide>>=
-set.seed(54)
-nobs <- 500
-x <- runif(nobs, 0, 1)
-mu <- sigma <- ytrue <- numeric(length = nobs)
-for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
-for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
-y <- rnorm(nobs, mean = mu, sd = sigma)
-ytrue <- mu
-data <- data.frame(cbind(y,x, ytrue))
-
-alldata <- cbind(data, mu, sigma)
-odata <- alldata[order(alldata["x"]),]
-@
-
-\begin{frame}[fragile]
-\frametitle{Fitting a Tree}
-\begin{center}
-\vspace*{-1.03cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plottree_xyplot, fig=TRUE, echo=FALSE>>=
-par(mar=c(5.1,4.1,4.1,3.1))
-plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "grey")
-#plot(y=odata$y, x=odata$x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray")
-lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
-polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
-  col = pallight["forest"], border = "transparent")
-legend(x = -0.0715, y = 21.62, expression(mu  %+-%  sigma), bty = "n")
-#lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
-#lines(x = odata$x, y = odata$mu - odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
-@
-\end{center}
-\end{frame}
-
-%\begin{frame}
-%\frametitle{Fitting a Tree}
-%\center
-%\vspace*{-0.7cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{tree_xyplot}
-%\end{frame}
 
@@
 \frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{-0.1cm}
+\vspace*{0.5cm}
 %Task:
@@
 \item Input: Numerical Weather Prediction (NWP).
+\item Problem: Uncertain initial conditions and \\
+\hspace{1.7cm}unknown processes.
 \item Idea: Statistical model with numerical predictions as \\
 \hspace{1cm}covariates.
-\item Data:
+\item Distribution Assumption:\\
+\vspace{0.2cm}
+\hspace{1cm}$(\text{Total precipitation})^\frac{1}{1.6} \sim \textit{cens}\mathcal{N}(\mu,\sigma^2)$
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace*{0.2cm}
+Data:
 \begin{itemize}
@@
 \item Numerical ensemble predictions of meteorological quantities\footnote{provided by the {U.S.} National Oceanic and Atmospheric Administration (NOAA)\\
-\hspace{0.32cm}using the Global Ensemble Forecast System (GEFS)}: 11 members each using slightly different perturbed initial conditions.
+\hspace{0.32cm}using the Global Ensemble Forecast System (GEFS)}: \\
+\hspace{0.5cm}11 members each using slightly different \\
+\hspace{0.5cm}perturbed initial conditions.
 \end{itemize}
-\item Distribution Assumption:\\
-\vspace{0.1cm}
-\hspace{1cm}$(\text{Total precipitation})^\frac{1}{1.6} \sim \textit{cens}\mathcal{N}(\mu,\sigma^2)$
-\end{itemize}
+
 \end{frame}
@@
 \item Complex terrain $\rightarrow$ unknown influences.
-\item 82 Covariates deduced from 14 base variables \\
-(variations such as ensemble minimum/maximum/mean/standard deviation):\\
+\item 82 Covariates deduced from 14 base variables:
 \begin{itemize}
@@
 \end{itemize}
+\item Variations such as ensemble minimum/maximum/mean/standard deviation.
 \end{itemize}

### 2017-12-05 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1811 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 27b29f2adc84e0d571404a7ca6042f9a154f148d

@@
 Both, observations and weather forecasts are available from 1985 trough the year 2012, focusing on July only in this article.\\ 
-Observations are available on a daily basis, each observation representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and it's close surrounding.
+Observations are available on a daily basis, each observation representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and its close surrounding.
 The observations are undergoing a rigorous quality check before made available to the public.
@@
 \hline
-\textbf{prespecified GAM} & spline  & tp{\_}mean, & tppow{\_}sprd,\\
+\textbf{prespecified GAM} & spline  & tp{\_}mean, & tp{\_}sprd,\\
                             & in each & tp{\_}max,  & dswrf{\_}sprd{\_}mean,\\
@@
 \hline
-\textbf{EMOS} & linear & tp{\_}mean & tp{\_}spread \\
+\textbf{EMOS} & linear & tp{\_}mean & tp{\_}sprd \\
 \hline

### 2017-12-05 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
edited application section and references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1813 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 252f6545fa6410ba6db6141f20a5ec1c441237cd

@@
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
-\setbeamerfont{footnote}{size=\tiny}
-\renewcommand\footnoterule{}
+%\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
 \URL{https://R-Forge.R-project.org/projects/partykit/}
@@
 \frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.5cm}
-%Task:
+\vspace{0.4cm}
+\textbf{Goal:} Predict total precipitation amount in alpine regions.\\
+\medskip
+\textbf{Input:} Numerical Weather Prediction (NWP):\\ 
+Global physical model numerically simulating atmospheric processes on $50\times 50 \text{km}^2$ grid.\\
+\medskip
+\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
+\vspace{1.3cm}
+%\bigskip
+\textbf{Distribution Assumption:}
+%\vspace{0.1cm}
+%\hspace{1cm}
+$(\text{Total precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2)$\\
+\bigskip
+\textbf{Idea:} Statistical model with numerical predictions as covariates.
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{0.4cm}
+\textbf{Observation data} (National Hydrographical Service):\\
 \begin{itemize}
-\item Goal: Predict total precipitation amount.
-\item Input: Numerical Weather Prediction (NWP).
-\item Problem: Uncertain initial conditions and \\
-\hspace{1.7cm}unknown processes.
-\item Idea: Statistical model with numerical predictions as \\
-\hspace{1cm}covariates.
-\item Distribution Assumption:\\
-\vspace{0.2cm}
-\hspace{1cm}$(\text{Total precipitation})^\frac{1}{1.6} \sim \textit{cens}\mathcal{N}(\mu,\sigma^2)$
+\item 95 observation stations in Tyrol, Austria.\\
+\item Daily 24h precipitation sums from July of 28 years (1985--2012).
+\end{itemize}
+\vspace{1cm}
+%\bigskip
+\textbf{NWP model output} (Global Ensemble Forecast System):\\
+\begin{itemize}
+\item Numerical ensemble predictions of various meteorological quantities.\\
+\item 11 members each using slightly different perturbed initial conditions.
 \end{itemize}
@@
 \vspace*{0.2cm}
-Data:
-\begin{itemize}
-\item 95 observation stations in Tyrol.
-\item Daily 24h precipitation sums from July of 28 years (1985--2012)\footnote{provided by the National Hydrographical Service (EHYD)}.
-\item Numerical ensemble predictions of meteorological quantities\footnote{provided by the {U.S.} National Oceanic and Atmospheric Administration (NOAA)\\
-\hspace{0.32cm}using the Global Ensemble Forecast System (GEFS)}: \\
-\hspace{0.5cm}11 members each using slightly different \\
-\hspace{0.5cm}perturbed initial conditions.
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
-Variable Selection:
-\begin{itemize}
-\item Complex terrain $\rightarrow$ unknown influences.
-\item 82 Covariates deduced from 14 base variables:
+\textbf{Base Variables:}
 \begin{itemize}
@@
 \item Mean sea level pressure
-\item Mean sea level pressure difference
 \item Preciptable water
@@
 \end{itemize}
-\item Variations such as ensemble minimum/maximum/mean/standard deviation.
-\end{itemize}
+\medskip
+\textbf{Variations:} Ensemble min/max/mean/standard deviation\\
+\vspace{0.1cm}
+$\rightarrow$ 82 covariates.
+
 \end{frame}
@@
 \frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
+\vspace{0.2cm}
 Application for one station:
@@
 \item 10 times 7-fold cross validation.
-\item Compare to commonly used heteroscedastic censored Gaussian models:\\
-\begin{itemize}
-\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-\item Boosted GAM: automatic variable selection.
-\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
+\item Compare to commonly used heteroscedastic censored Gaussian models:
 \end{itemize}
-\end{itemize}
+%\begin{itemize}
+%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
+%\item Boosted GAM: automatic variable selection.
+%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
+%\end{itemize}
+\vspace{0.1cm}
+\begin{table}[t!]
+\footnotesize
+\nodindent
+\hskip-0.4cm\begin{tabular}{ l l l }
+\hline
+Model & Type & Variable Selection \\
+\hline
+\vspace{-0.15cm}
+\textbf{Forest} & recursive    & automatic \\ 
+                & partitioning &           \\ 
+\hline
+\vspace{-0.15cm}
+\textbf{prespecified GAM} & spline  & based on expert \\
+                          & in each & knowledge      \\
+\hline
+\vspace{-0.15cm}
+\textbf{boosted GAM} & spline  & automatic  \\
+                     & in each &            \\
+\hline
+\vspace{-0.15cm}
+\textbf{EMOS} & linear & total precipitation mean \\
+              &        & and standard deviation   \\
+\hline
+\end{tabular}
+\end{table}
+
 \end{frame}
@@
 \frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
+\vspace{0.2cm}
 Application for all 95 stations:
@@
 \item Compare to commonly used heteroscedastic censored Gaussian models:\\
-\begin{itemize}
-\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-\item Boosted GAM: automatic variable selection.
-\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
 \end{itemize}
-\end{itemize}
+%\begin{itemize}
+%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
+%\item Boosted GAM: automatic variable selection.
+%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
+%\end{itemize}
+\vspace{0.1cm}
+\begin{table}[t!]
+\footnotesize
+\nodindent
+\hskip-0.4cm\begin{tabular}{ l l l }
+\hline
+Model & Type & Variable Selection \\
+\hline
+\vspace{-0.15cm}
+\textbf{Forest} & recursive    & automatic \\ 
+                & partitioning &           \\ 
+\hline
+\vspace{-0.15cm}
+\textbf{prespecified GAM} & spline  & based on expert \\
+                          & in each & knowledge      \\
+\hline
+\vspace{-0.15cm}
+\textbf{boosted GAM} & spline  & automatic  \\
+                     & in each &            \\
+\hline
+\vspace{-0.15cm}
+\textbf{EMOS} & linear & total precipitation mean \\
+              &        & and standard deviation   \\
+\hline
+\end{tabular}
+\end{table}
 \end{frame}
@@
 \begin{frame}
-\frametitle{References (1)}
+\frametitle{References}
+
+\vspace{-0.2cm}
 
 \footnotesize
-
 Hothorn T, Hornik K, Zeileis A (2006).
@@
 
+%\medskip
+
+%Seibold H, Zeileis A, Hothorn T (2017).
+%  \dquote{Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.}
+%  \textit{Statistical Methods in Medical Research}, 
+%  \textbf{12}(1), 45--63.
+%  \doi{10.1177/0962280217693034}
+
+%\medskip
+
+%Hothorn T, Zeileis A (2017).
+%  \dquote{Transformation Forests.}
+%  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
+%  \url{http://arxiv.org/abs/1701.02110}
+
 \medskip
 
-Seibold H, Zeileis A, Hothorn T (2017).
-  \dquote{Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.}
-  \textit{Statistical Methods in Medical Research}, 
-  \textbf{12}(1), 45--63.
-  \doi{10.1177/0962280217693034}
+Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
+  \dquote{Calibrated probabilistic forecasting using ensemble model output statistics and minimum CRPS estimation}
+  \textit{Monthly Weather Review}, 
+  \textbf{133}(5), 1098--1118.
+  \doi{10.1175/MWR2904.1}
 
@@
 
-Hothorn T, Zeileis A (2017).
-  \dquote{Transformation Forests.}
-  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
-  \url{http://arxiv.org/abs/1701.02110}
+Hofner B, Mayr A, Fenske N, Schmid M (2017).
+  \dquote{{gamboostLSS}: Boosting Methods for {GAMLSS} Models}
+  \textit{{R} package version 2.0-0},
+  \url{https://CRAN.R-project.org/package=gamboostLSS}
 
@@
 \frametitle{References (2)}
+
+\vspace{-0.2cm}
 
@@
   \textbf{94}(10), 1553--1565.
-  \doi = {10.1175/BAMS-D-12-00014.1}
+  \doi{10.1175/BAMS-D-12-00014.1}
 
 \medskip
+
+%Breiman L (2001).
+%  \dquote{Random {Forests}}
+%  \textit{Machine Learning}, 
+%  \textbf{45}(1), 5--32.
+%  \doi{10.1023/A:1010933404324}
 
@@
   \textit{The R Journal}, 
-  \textbf{8}(1), 173--181.
+  \textbf{8}(1), 173--181.\\
+  \scriptsize
   \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
+  \footnotesize
 
@@
 
-Breiman L (2001).
-  \dquote{Random {Forests}}
-  \textit{Machine Learning}, 
-  \textbf{45}(1), 5--32.
-  \doi{10.1023/A:1010933404324}
-
-\medskip
-
-Hofner B, Mayr A, Fenske N, Schmid M (2017).
-  \dquote{{gamboostLSS}: Boosting Methods for {GAMLSS} Models}
-  \textit{{R} package version 2.0-0},
-  \url{https://CRAN.R-project.org/package=gamboostLSS}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{References (3)}
-
-\footnotesize
-Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-  \dquote{Calibrated probabilistic forecasting using ensemble model output statistics and minimum CRPS estimation}
-  \textit{Monthly Weather Review}, 
-  \textbf{133}(5), 1098--1118.
-  \doi{https://doi.org/10.1175/MWR2904.1}
+Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
+  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored Additive Regression Model}
+  \textit{International Journal of Climatology}, 
+  \textbf{37}(7), 3264--3275.
+  \doi{10.1002/joc.4913}
 
@@
 
-\medskip
-
-{CGIAR-CSI} (2016).
-  \dquote{{SRTM} 90m Digital Elevation Database {v4.1}}
-  \text{Available at \url{http://srtm.csi.cgiar.org}}, 
-  \text{Accessed: {J}uly 10 2016}.
-
-\medskip
-
-Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
-  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored Additive Regression Model}
-  \textit{International Journal of Climatology}, 
-  \textbf{37}(7), 3264--3275.
-  \doi{10.1002/joc.4913}
+%\medskip
+
+%{CGIAR-CSI} (2016).
+%  \dquote{{SRTM} 90m Digital Elevation Database {v4.1}}
+%  \text{Available at \url{http://srtm.csi.cgiar.org}}, 
+%  \text{Accessed: {J}uly 10 2016}.
 

### 2017-12-06 pkg/disttree/inst/draft/draft.Rnw
adapted to changes in ref.bib
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1826 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e46f6205a2fc840e4d01b0a2634347faa119cebd

@@
 %\newpage
-\nocite{hothorn2006unbiased, zeileis2008model, hothorn2015partykit, Messner2016, loh2011classification, stasinopoulos2007generalized, Seibold2016, Seibold2017, breiman2001random, stasinopoulos2005, yee2010vgam}
+\nocite{hothorn2006unbiased, zeileis2008model, hothorn2015partykit, Messner2016, loh2011classification, stasinopoulos2007generalized, Seibold2016, Seibold2017, breiman2001random, stasinopoulos2005, yee2010vgam, gamboostLSS2016}
 % zeileis2008implementing, 

### 2017-12-06 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
small changes in formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1827 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 31f5403ff6dc4a176afb5023f168f7c3208fa5f6

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber, noslidenumber]{uibk}
+\usetheme[nototalframenumber, noslidenumber, url]{uibk}
 
@@
 \vspace{0.4cm}
-\textbf{Goal:} Predict total precipitation amount in alpine regions.\\
+\textbf{Goal:} Predict daily precipitation amount in alpine regions.\\
 \medskip
-\textbf{Input:} Numerical Weather Prediction (NWP):\\ 
-Global physical model numerically simulating atmospheric processes on $50\times 50 \text{km}^2$ grid.\\
+\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
 \medskip
@@
 %\bigskip
-\textbf{Distribution Assumption:}
-%\vspace{0.1cm}
-%\hspace{1cm}
-$(\text{Total precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2)$\\
+
+\textbf{Idea:} Statistical model with numerical predictions as covariates.\\
 \bigskip
-\textbf{Idea:} Statistical model with numerical predictions as covariates.
+\textbf{Distribution assumption:}\\
+\vspace{-0.3cm}
+%\hspace{1.5cm}
+%\center
+$$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
+
 \end{frame}
@@
 \item 95 observation stations in Tyrol, Austria.\\
-\item Daily 24h precipitation sums from July of 28 years (1985--2012).
+\item Daily 24h precipitation sums from July over 28 years (1985--2012).
 \end{itemize}
@@
 \begin{itemize}
-\item Numerical ensemble predictions of various meteorological quantities.\\
+\item Ensemble predictions of various meteorological quantities related to precipitation, temperature,
+air pressure, \dots.\\
 \item 11 members each using slightly different perturbed initial conditions.
@@
 \vspace*{0.2cm}
-\textbf{Base Variables:}
+\textbf{Base variables:}
 \begin{itemize}
@@
 \vspace*{0.2cm}
-Application for one station:
+\textbf{Application for one station:}
 \begin{itemize}
@@
 \vspace{0.2cm}
-Application for one station:
+\textbf{Application for one station:}
 \begin{itemize}
@@
 \vspace{0.2cm}
-Application for all 95 stations:
+\textbf{Application for all 95 stations:}
 \begin{itemize}
@@
 \item Evaluate on 4 years (2009--2012).
-\item Compare to commonly used heteroscedastic censored Gaussian models:\\
+\item Compare to commonly used heteroscedastic censored Gaussian models. %:\\
 \end{itemize}
@@
 %\end{itemize}
-\vspace{0.1cm}
-\begin{table}[t!]
-\footnotesize
-\nodindent
-\hskip-0.4cm\begin{tabular}{ l l l }
-\hline
-Model & Type & Variable Selection \\
-\hline
-\vspace{-0.15cm}
-\textbf{Forest} & recursive    & automatic \\ 
-                & partitioning &           \\ 
-\hline
-\vspace{-0.15cm}
-\textbf{prespecified GAM} & spline  & based on expert \\
-                          & in each & knowledge      \\
-\hline
-\vspace{-0.15cm}
-\textbf{boosted GAM} & spline  & automatic  \\
-                     & in each &            \\
-\hline
-\vspace{-0.15cm}
-\textbf{EMOS} & linear & total precipitation mean \\
-              &        & and standard deviation   \\
-\hline
-\end{tabular}
-\end{table}
+
+%\vspace{0.1cm}
+%\begin{table}[t!]
+%\footnotesize
+%\nodindent
+%\hskip-0.4cm\begin{tabular}{ l l l }
+%\hline
+%Model & Type & Variable Selection \\
+%\hline
+%\vspace{-0.15cm}
+%\textbf{Forest} & recursive    & automatic \\ 
+%                & partitioning &           \\ 
+%\hline
+%\vspace{-0.15cm}
+%\textbf{prespecified GAM} & spline  & based on expert \\
+%                          & in each & knowledge      \\
+%\hline
+%\vspace{-0.15cm}
+%\textbf{boosted GAM} & spline  & automatic  \\
+%                     & in each &            \\
+%\hline
+%\vspace{-0.15cm}
+%\textbf{EMOS} & linear & total precipitation mean \\
+%              &        & and standard deviation   \\
+%\hline
+%\end{tabular}
+%\end{table}
 \end{frame}
@@
 raster::image(tirol.dem, col = rev(gray.colors(100)),
-              main="Observation Stations", ylab = "Latitude", xlab = "Longitude", 
+              main="Observation stations", ylab = "Latitude", xlab = "Longitude", 
               xlim = c(9.8,13.2), 
@@
 Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-  \dquote{Calibrated probabilistic forecasting using ensemble model output statistics and minimum CRPS estimation}
+  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and Minimum CRPS Estimation}
   \textit{Monthly Weather Review}, 
@@
 
-Hofner B, Mayr A, Fenske N, Schmid M (2017).
-  \dquote{{gamboostLSS}: Boosting Methods for {GAMLSS} Models}
-  \textit{{R} package version 2.0-0},
-  \url{https://CRAN.R-project.org/package=gamboostLSS}
+Hofner B, Mayr A, Schmid M (2016).
+  \dquote{{gamboostLSS}: An R Package for Model Building and Variable Selection in the {GAMLSS} Framework}
+  \textit{Journal of Statistical Software},
+  \textbf{74}(1), 1--31.
+  \doi{10.18637/jss.v074.i01}
 
@@
   \textit{Monthly Weather Review}, 
-  \textbf{Forthcoming}.
+  \text{Forthcoming}.
   \doi{10.1175/MWR-D-16-0388.1}

### 2017-12-08 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
resized map and small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1828 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ea82a0595ee89a9a4db0f8fa815d3dfc0f249111

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber, noslidenumber, url]{uibk}
+\usetheme[nototalframenumber, noslidenumber]{uibk}
 
@@
 \textbf{Distribution assumption:}\\
-\vspace{-0.3cm}
+\vspace{-0.25cm}
 %\hspace{1.5cm}
@@
 \footnotesize
-\nodindent
+
 \hskip-0.4cm\begin{tabular}{ l l l }
@@
 \begin{center}
-\vspace*{0cm}
-
-<<map, fig=TRUE, echo=FALSE, width=10, height = 6.5>>=
+%\vspace*{-2cm}
+
+\setkeys{Gin}{width=0.95\textwidth}
+<<map, fig=TRUE, echo=FALSE, width=10, height=6.5>>=
 
 layout(cbind(1, 2), width = c(9, 1))
-par(mar = c(5,4,4,0.1))
+par(mar = c(5,4,4,0.01))
 raster::image(tirol.dem, col = rev(gray.colors(100)),
@@
               xlim = c(9.8,13.2), 
-              ylim = c(46.6, 47.8), 
-              cex.lab=1.5, cex.axis=1.2, cex.main=2)
+              ylim = c(46.6, 47.85), 
+              cex.lab=1.3, cex.axis=1.2, cex.main=1.7)
 plot(tirol.gadm, add = TRUE)
-points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.9)
+points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.7)
 legend("topleft", pch=c(21,24,25,22), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1.2, bty = "n")
 
-par(mar = c(0.5,0.2,0.5,2.3))
+par(mar = c(0.4,0.2,0.4,2.33))
 ## legend
@@
 rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
-axis(4, at = brk, las = 1, mgp=c(0,0,-1.2), cex = 1.4)
+axis(4, at = brk, las = 1, mgp=c(0,0,-1.2), cex = 1.5)
 

### 2017-12-12 pkg/disttree/inst/draft/draft.Rnw
edited abstract + small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1838 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8d9ec958e3a353474a2967fb843084fd92958def

@@
 \Abstract{
-In the field of probabilistic forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools.
-This is due to their simplicity and interpretability.
+While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to also model all other distribution parameters.
+In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods.
+For example in the field of probabilistic weather forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools allowing for clear interpretations.
+Usually acting on the assumption of a normally distributed response variable both parameters, the mean and the variance, depend linearly on covariates in this type of model.
 However, the restriction to model only linear effects requires further extensions which allow for more flexibility.
-Generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
-A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
-However, up to now, this class of models has not yet been integrated in probabilistic forecasting as no general distributional framework is available for these methods.
-Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
+Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
+A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
+This class of models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
+We are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
+Applying these novel approaches to precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
 
@@
 \section{Introduction}
+In recent years there has been a paradigm shift from prediction/regression models for the mean to distributional regression approaches for probabilistic forecasting.
+
 A typical application in meteorology is to improve probabilistic precipitation forecasts.
@@
 To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
-In Figure~\ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.\\
+In Figure~\ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.
+
 These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g.
@@
 However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model.
-In this situation a different approach which can deal with these abrupt changes would be to apply tree structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
+In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
 Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process.
@@
 However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well.
-In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothed out in case of smooth effects while abrupt changes are still captured by the tree structure of the model.\\
+In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothed out in case of smooth effects while abrupt changes are still captured by the tree structure of the model.
+
 For these reasons, regression trees and particularly random forests provide valuable features which can contribute to improving models for probabilistic forecasting.
@@
 Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure~\ref{Devel_treeforest}.\\
+The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure~\ref{Devel_treeforest}.
 
@@
 \end{tikzpicture}
-\caption{\label{Devel_parmod}Development in the field of parametric modeling.}
+\caption{\label{Devel_parmod}Development in the field of parametric modeling: (Generalized) linear models (LM, GLM; left), generalized additive models (GAM; middle), generalized additive models for location, scale, and shape (GAMLSS; right).}
 \end{figure}
@@
 points(x, y, col = "slategray", pch = 19)
-box(lwd=5)
+box(lwd=10)
 @
@@
   col = pallight["forest"], border = "transparent")
-box(lwd=5)
+box(lwd=10)
 @
@@
 But as mentioned before, further features such as the ability to capture non-additive effects or interactions are desirable in many applications which demands for more flexible models.
-Since tree structured models provide these required features a combination of them with parametric models can offer a framework that is able to deal with a wide range of different modeling situations.
-In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree structured models step by step.
+Since tree-structured models provide these required features a combination of them with parametric models can offer a framework that is able to deal with a wide range of different modeling situations.
+In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree-structured models step by step.
 We start with a simple distributional fit in Subsection~\ref{distfit}, then build a distributional tree in Subsection~\ref{disttree} and finally a distributional forest in Subsection~\ref{distforest}.
@@
 \newpage
-\subsection{Distributional Fit}
+\subsection{Distributional fit}
 \label{distfit}
 Fitting a distributional model to a set of observations without considering any covariates is a well known procedure which can for example be done by applying the maximum likelihood method.
-This is also the chosen approach in the methodology described in this section.\\
+This is also the chosen approach in the methodology described in this section.
+
 The goal is to fit a distributional model $D(Y, \theta)$ to a response variable $Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters $\theta \in \Theta$ in case of a \textit{k}-parametric distribution.
@@
 Using the GAMLSS framework (\citealt{stasinopoulos2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
-This enables the model to deal with various features that can usually be very challenging in model fitting such as censoring, truncation or heavy tails.\\
+This enables the model to deal with various features that can usually be very challenging in model fitting such as censoring, truncation or heavy tails.
+
 In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
@@
 Another frequent application of probabilistic forecasting in meteorology is the modeling of near surface air temperature.
-In this case a normal distribution is often selected.\\
+In this case a normal distribution is often selected.
+
 With the specification of the distribution family and the log-likelihood function the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$.
@@
 $$
-where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response variable $Y$.\\
+where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response variable $Y$.
+
 In that way, a whole distribution is specified with all its features, including location, scale and shape.
@@
 In our application on precipitation data a wide range of covariates are available from a numerical ensemble prediction system considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest and have an influence on the predictions.
-\\
+
 A common way of including these covariates is to consider them as regressors in a distributional model such that each parameter $\theta$ depends on covariates $\bold{X} = (X_1,\ldots,X_m)$.
@@
 Another approach to distributional regression is proposed by \citealt{Klein2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
-These include non-linear effects, spatial effects, random coefficients and interaction surfaces which extends the flexibility compared to parametric models such as GAMLSS modeling only smooth effects.\\
+These include non-linear effects, spatial effects, random coefficients and interaction surfaces which extends the flexibility compared to parametric models such as GAMLSS modeling only smooth effects.
+
 However, the restriction to additive models remains.
@@
 Doing so after fitting a global model to the whole data set (without considering any covariates) and then fitting a local model to each subgroup should improve the model remarkably.
-This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.\\
-
-\subsection{Distributional Tree}
+This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.
+
+\subsection{Distributional tree}
 \label{disttree}
@@
 Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.
-This enables the model to be sensitive to changes in each of the parameters which is one of the main advantages of distributional trees over other tree models considering only the mean of the distribution.\\
+This enables the model to be sensitive to changes in each of the parameters which is one of the main advantages of distributional trees over other tree models considering only the mean of the distribution.
+
 Statistical tests are applied to check whether there is a significant dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$.
-If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest dependency.\\
+If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest dependency.
+
 In a next step a local distributional model is fit in each of the subgroups.
@@
 $$
-where each segment is represented by a terminal node of the tree.\\
+where each segment is represented by a terminal node of the tree.
 
@@
 \end{enumerate}
-The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.\\
-Executing these steps results in a stepwise parametric model where a complete distribution is fit in each node of the tree.\\
-\\
+The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.
+
+Executing these steps results in a stepwise parametric model where a complete distribution is fit in each node of the tree.
+
 For distributional trees the strategy of how to make a prediction for a new set of covariates $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ is very simple and straight forward.
@@
 In that way, all kinds of effects can be modeled, which is is also one the of the main advantages of forest models (taking bagging one step further as the variance is reduced by sampling the covariates considered for the splitting process in each node).
-\\
-\\
+
 Common forest models such as random forests consist of trees only considering the mean of the distribution of the response variable.
@@
 However, in this case this separation can also be a disadvantage as it hinders the capturing of interactions unless they are specified explicitly by the user.
-These two cases provide once again a good reason to combine the two frameworks of flexible parametric modeling and tree structured models in order to profit from their advantages and compensate possible disadvantages.
+These two cases provide once again a good reason to combine the two frameworks of flexible parametric modeling and tree-structured models in order to profit from their advantages and compensate possible disadvantages.
 This is the pursued goal in introducing distributional forests.
 
-\subsection{Distributional Forest}
+\subsection{Distributional forest}
 \label{distforest}
@@
 Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables.
-In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).\\
-\\
+In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).
+
 Forest models differ mainly in the way how the trees are combined to make predictions.
-In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.\\
+In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.
+
 Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates.
-This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
-$\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
+This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights
+
+$\bold{\omega}^{forest}~=~(\omega^{forest}_1, \ldots, \omega^{forest}_n)$.
+
 For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$.
@@
 $$
-While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates due to the individual weights calculated separately for each set of covariates.\\
-\\
-As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree structured modeling it provides a model that can deal with a wide range of situations.
+While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates due to the individual weights calculated separately for each set of covariates.
+
+As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree-structured modeling it provides a model that can deal with a wide range of situations.
 Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably well by combining various tree and smoothing the steps imposed by a single tree.
 As scores are considered for each distribution parameter (separate) effects in different parameters can be captured.
-At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.\\
+At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.
+
 Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
@@
 %Looking at the other extreme situation in which the parameter functions include abrupt changes ($\kappa = 10$, right window in Figure~\ref{plot_sim_oneset_mu}) the tree captures them very well wich also leads to a good fit of the forest model.
-In contrast to the two tree structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter functions.\\
+In contrast to the two tree-structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter functions.\\
 
@@
 \SweaveOpts{eval = TRUE}
-%\newpage
-%\newpage
-\section{Probabilistic Forecasting of Precipitation in Complex Terrain}
+
+\section{Probabilistic precipitation forecasting in complex terrain}
 \label{ProbabilisticForecasting}
 
-As explained in the introduction, distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
+As explained in the introduction, distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree-structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
 To show their strengths in practice, these new methods are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
@@
 The data set used for training and validation consists of observed daily precipitation sums provided by the National Hydrographical Service (\citealt{ehyd}) and numerical weather forecasts from the {U.S.} National Oceanic and Atmospheric Administration (NOAA).
-Both, observations and weather forecasts are available from 1985 trough the year 2012, focusing on July only in this article.\\ 
+Both, observations and weather forecasts are available from 1985 trough the year 2012, focusing on July only in this article.
+
 Observations are available on a daily basis, each observation representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and its close surrounding.
@@
 In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates.
-Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std dev, ensemble minimum and ensemble maximum.
+Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std.
+dev., ensemble minimum and ensemble maximum.
 All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
-Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std dev of mean/minimal/maximal values respectively.\\
+Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std.
+dev.
+of mean/minimal/maximal values respectively.\\
 Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
@@
 In \citealt{stauffer2017a} optimization methods have been applied to find the best transformation factor which resulted in a value of $\frac{1}{1.6}$.
-Since the observation data set used for this optimization contains the observations considered in this article the same factor is chosen here for the power transformation of the response variable \texttt{tp} (total precipitation).
+Since the observation data set used for this optimization contains the observations considered in this article the same factor is chosen here for the power transformation of the response variable \emph{tp} (total precipitation).
 The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
-For simplicity the same transformation has been chosen for the variable \texttt{cape} (convective available potential energy).
+For simplicity the same transformation has been chosen for the variable \emph{cape} (convective available potential energy).
 
@@
 \hline
-\textbf{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
-\hspace*{\fill} power transformed (by $\frac{1}{1.6}$) &    & ensemble std dev of sums over 24h, \\ 
-\textbf{cape}: convective                  &    & ensemble minimum of sums over 24h, \\
-\hspace*{0.4cm} available potential energy &    & ensemble maximum of sums over 24h\\
-\hspace*{\fill}  power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
+\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
+\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. dev. of sums over 24h, \\ 
+\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
+\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
+\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
 \hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
 \hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
-\hspace*{\fill}                            &    & ensemble std dev of sums over 6h\\
+\hspace*{\fill}                            &    & ensemble std. dev. of sums over 6h\\
 \hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
 \hline
-\textbf{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
-\hspace*{\fill} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
-\textbf{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
-\textbf{pwat}: preciptable water          &   & ensemble std dev of mean values,\\
-\textbf{tmax}: 2m maximum temperature     &   & ensemble std dev of minimal values,\\
-\hspace*{1.15cm}                          &   & ensemble std dev of maximal values,\\
-\textbf{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
-\hspace*{1.05cm} condensate               &   & \\
-\textbf{t500}: temperature on 500 hPa     &   & \\
+\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
+\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
+\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
+\emph{pwat}: preciptable water          &   & ensemble std. dev. of mean values,\\
+\emph{tmax}: 2m maximum temperature     &   & ensemble std. dev. of minimal values,\\
+\hspace*{1.15cm}                          &   & ensemble std. dev. of maximal values,\\
+\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
+\hspace*{0.95cm} condensate               &   & \\
+\emph{t500}: temperature on 500 hPa     &   & \\
 \hspace*{1cm}                             &   & \\
-\textbf{t700}: temperature on 700 hPa     &   & \\
+\emph{t700}: temperature on 700 hPa     &   & \\
 \hspace*{1cm}                             &   & \\
-\textbf{t850}: temperature on 850 hPa     &   & \\
+\emph{t850}: temperature on 850 hPa     &   & \\
 \hspace{1cm}                              &   & \\
 \hline
-\textbf{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
+\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
 \hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
-\textbf{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
+\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
 \hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
-\textbf{tdiff700850}: temperature         &   & \\
+\emph{tdiff700850}: temperature         &   & \\
 \hspace*{\fill} difference 700 to 850 hPa &   & \\
 \hline
-\textbf{msl{\_}diff}: mean sea level pressure & 1 & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
-\hspace{1.45cm} difference            &   & \qquad over 6--30 UTC\\
+\emph{msl{\_}diff}: mean sea level pressure & 1 & msl{\_}mean{\_}max $-$ msl{\_}mean{\_}min\\
+\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
 \hline
@@
 
-\subsection{Models and Evaluation}
+\subsection{Models and evaluation}
 Next to distributional forests three other methods are applied to this data set and their results are compared based on the continuous ranked probability score (CRPS).
@@
 For the prespecified GAM model the covariates were selected manually based on meteorological knowledge and experience.
-Additionally, interactions are specified in advance, once between \texttt{tp{\_}mean1218} and \texttt{cape{\_}mean1218} for the location parameter and once between \texttt{tp{\_}sprd1218} and \texttt{cape{\_}mean1218} for the scale parameter.
+Additionally, interactions are specified in advance, once between \emph{tp{\_}mean1218} and \emph{cape{\_}mean1218} for the location parameter and once between \emph{tp{\_}sprd1218} and \emph{cape{\_}mean1218} for the scale parameter.
 These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation.
@@
 
-\section*{Computational Details}
+\section*{Computational details}
 The corresponding implementation to the novel methods presented in this article is available on R-Forge in an \textsf{R} package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}).

### 2017-12-13 pkg/disttree/inst/draft/draft.Rnw
edited abstract und introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1846 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f8f3b10ed1fb5f09dcec721eb8ca8403aea1f1af

@@
 While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to also model all other distribution parameters.
-In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods.
-For example in the field of probabilistic weather forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools allowing for clear interpretations.
+In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting.
+In generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects.
+A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
+This class of models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
+We are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
+Applying these novel approaches to precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
+
+%For example in the field of probabilistic weather forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools allowing for clear interpretations.
 Usually acting on the assumption of a normally distributed response variable both parameters, the mean and the variance, depend linearly on covariates in this type of model.
@@
 \section{Introduction}
-In recent years there has been a paradigm shift from prediction/regression models for the mean to distributional regression approaches for probabilistic forecasting.
-
-A typical application in meteorology is to improve probabilistic precipitation forecasts.
-The main interest lies in estimating the probability of precipitation as well as predicting the expected amount of precipitation.
-In this research area various methods have been developed and tested in the last decades.
-(Multiple) Linear models such as ensemble model output statistics (EMOS) were introduced more than a decade ago (\citealt{gneiting2005calibrated}) and are still common and widely used.
-Despite their simplicity and interpretability, the limitation to modeling only linear effects has risen the interest in applying more flexible models.
-This feature is provided by generalized additive models (GAM) which can also model non-linear effects.
-However, often only the expectation of the underlying distribution is estimated.
-To obtain probabilistic predictions, generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
-In Figure~\ref{Devel_parmod} the development of parametric models starting from simple linear models (LM) and ranging to GAMLSS is illustrated.
-
+In regression analysis various models have been developed to describe the relationship between a response variable and a (set of) covariate(s).
+Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects as illustrated in Figure~\ref{Devel_parmod}.
+However, since these models consider only the expectation of the underlying distribution further development was required.
+Therefore, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
+For example in the field of precipitation forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) usually acting on the assumption of a normally distributed response variable allow for both parameters, the mean and the variance, to depend linearly on covariates.
+Based on the resulting model the probability of precipitation can be estimated as well as the expected amount of precipitation.
+Therefore, these models provide common and widely used tools, also due to their intuitive model structure and their interpretability.
+However, the restriction to model only linear effects in the parameters has risen the interest in applying more flexible models.
+Generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
 These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g.
@@
 points(x, y, col = "slategray", pch = 19)
-box(lwd=10)
+box(lwd=12)
 @
@@
   col = pallight["forest"], border = "transparent")
-box(lwd=10)
+box(lwd=12)
 @
@@
 \draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](0,-2.4)--(0,-3.2);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.2,-5.4)--(7.9,-2.5);
+\draw[-{>[scale=4, length=2, width=3]},line width=1.4pt](7.1,0)--(7.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=1.4pt](0,-2.4)--(0,-3.2);
+\draw[-{>[scale=4, length=2, width=3]},line width=1.4pt](2.2,-5.4)--(7.9,-2.5);
 \end{tikzpicture}
-\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests.}
+\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests: regression tree (top left), distributional tree (bottom), random forest (top middle), distributional forest (top right).}
 \end{figure}
@@
 
-\newpage
 \subsection{Distributional fit}
@@
 dev.
-of mean/minimal/maximal values respectively.\\
-Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.\\
+of mean/minimal/maximal values respectively.
+Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.
+Altogether 82 covariates are available in this setting.
+
 To remove large parts of the skewness of precipitation data, a power transformation has been used frequently (Box and Cox 1964).
@@
 These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation.
-By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.\\
+By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.
+
 In the EMOS model the logarithmic function is chosen as the link function for the scale parameter.
 This choice has been made based on the results of \citealt{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared.
-Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.\\
-\\
+Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.
+
 In order to compare the performance of the chosen methods, the continuous ranked probability score (CRPS) is calculated for each model.
@@
 $$
-where \(\mathbf{1}\) is the indicator function. \\
+where \(\mathbf{1}\) is the indicator function.
+
 In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model.
@@
 In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observation.
-\\
+
 For the purpose of comparing the performance of a distributional forest to those of other models the so-called skills score can be calculated with respect to the CRPS and one model being chosen as a reference.
@@
 This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol.
-As for all other stations, daily observations of the month of July of 28 years are available from 1985 through the year 2012.\\
+As for all other stations, daily observations of the month of July of 28 years are available from 1985 through the year 2012.
+
 The results of fitting one distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on probability integral transform (PIT) histograms.\\
-\\
+Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on probability integral transform (PIT) histograms (\citealt{gneiting2007}).
+
 As shown in Table~\ref{model_specification} the forest model includes all 82 covariates.
@@
 The CRPS skills score boxplot shows that, in this setting, both GAM models and the forest perform better than the EMOS model.
-While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the forest even reaches an improvement rate close to 6 percent.\\
-From this result we can conclude that the distributional forest provides a powerful alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.\\
+While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the forest even reaches an improvement rate close to 6 percent.
+
+From this result we can conclude that the distributional forest provides a powerful alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.
+
 For the prespecified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
 To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decisions.
-Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not correspond to the truth.\\
+Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not correspond to the truth.
 This risk is avoided in the boosted GAM as all covariates are included and the selection of the influential covariates is done within a boosting framework.
-However, the trade off here is a remarkably higher computational effort.\\
+However, the trade off here is a remarkably higher computational effort.
 To build a distributional forest the user does not have to put any additional effort in it and therefore is not required to provide any expert knowledge or additional information about the setting.
-Variable selection is done automatically and interactions are detected by the tree structure of the model.\\
-In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the forest model.\\
+Variable selection is done automatically and interactions are detected by the tree structure of the model.
+In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the forest model.
 Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
@@
 %\newpage
-\nocite{hothorn2006unbiased, zeileis2008model, hothorn2015partykit, Messner2016, loh2011classification, stasinopoulos2007generalized, Seibold2016, Seibold2017, breiman2001random, stasinopoulos2005, yee2010vgam, gamboostLSS2016}
+\nocite{Messner2016, stasinopoulos2007generalized, Seibold2016, Seibold2017, yee2010vgam, gamboostLSS2016, srtm}
 % zeileis2008implementing, 

### 2017-12-13 pkg/disttree/inst/draft/oneset.R

@@ -1,0 +1,1567 @@
+################################
+### simulation function for one data set
+sim_oneset <- function(parfun,
+                       nobs = 500,
+                       seedconst = 7, ntree = 100,
+                       formula = y~x1+x2+x3+x4+x5+x6, 
+                       tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95, 
+                       forest_minsplit = 7, forest_minbucket = 4, forest_mincrit = 0, 
+                       forest_mtry = 3,
+                       type.tree = "ctree",
+                       censNO = TRUE,
+                       pred_fix = FALSE,
+                       pred_fix_x1 = -0.5,
+                       pred_fix_x2 = -0.5,
+                       pred_fix_x4 = 0,
+                       nrunif = 3,
+                       nrcovform = 6)
+{
+  ### preliminaries
+  library("disttree")
+  library("gamlss")
+  library("lattice")
+  library("crch")
+  library("latex2exp")
+  library("parallel")
+  library("gamlss.cens")
+  library("Formula")  # FIX ME: should be imported in disttree
+  library("survival")
+  gen.cens(NO, type = "left")
+  
+  cl <- match.call()
+  
+  
+  
+  ## define distribution list:
+  # dist_list_normal
+  {
+    
+    dist_list_normal <- list()
+    
+    parnames <- c("mu", "sigma")
+    etanames <- c("mu", "log(sigma)")
+    
+    
+    ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {     
+      
+      val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
+      if(!log) val <- exp(val)
+      
+      # par <- c(eta[1], exp(eta[2]))
+      # val <- dnorm(y, mean = par[1], sd = par[2], log = log)
+      
+      if(sum) {
+        if(is.null(weights)) weights <- rep.int(1, length(y))
+        val <- sum(weights * val, na.rm = TRUE)
+      }
+      return(val)
+    }
+    
+    
+    sdist <- function(y, eta, weights = NULL, sum = FALSE) {   
+      
+      score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
+                     -1 + exp(-2*eta[2] + log((y-eta[1])^2)))
+      
+      # par <- c(eta[1], exp(eta[2])) 
+      # score <- cbind(1/par[2]^2 * (y-par[1]), 
[…]

### 2017-12-13 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
small corrections in references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1847 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: de22cdaccf48054e3835abf38da4dba78bc71585

@@
 Hothorn T, Zeileis A (2015).
- \dquote{partykit: A Modular Toolkit for Recursive Partytioning in R.}
+ \dquote{\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
  \textit{Journal of Machine Learning Research},
@@
 Stasinopoulos DM, Rigby RA (2007).
-  \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in R.}
+  \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in \textsf{R}.}
   \textit{Journal of Statistical Software}, 
@@
 Hofner B, Mayr A, Schmid M (2016).
-  \dquote{{gamboostLSS}: An R Package for Model Building and Variable Selection in the {GAMLSS} Framework}
+  \dquote{\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and Variable Selection in the {GAMLSS} Framework}
   \textit{Journal of Statistical Software},
@@
 BMLFUW (2016).
-  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtschaft ({BMLFUW}), {A}bteilung {IV/4} - {W}asserhaushalt}
+  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtschaft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
   \text{Available at \url{http://ehyd.gv.at}}, 
-  \text{Accessed: {F}ebruary 29 2016}.
+  \text{Accessed: 2016--02--29}.
 
@@
 Messner J W,  Mayr G J, Zeileis A (2016).
-  \dquote{Heteroscedastic Censored and Truncated Regression with {crch}}
-  \textit{The R Journal}, 
+  \dquote{Heteroscedastic Censored and Truncated Regression with \textbf{crch}}
+  \textit{The \textsf{R} Journal}, 
   \textbf{8}(1), 173--181.\\
@@
 %  \text{Available at \url{http://srtm.csi.cgiar.org}}, 
-%  \text{Accessed: {J}uly 10 2016}.
+%  \text{Accessed: 2016--07--10}.
 

### 2017-12-13 pkg/disttree/inst/draft/draft.Rnw
playing around with bg and lwd in Fig 2; new title suggestion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1849 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0320dbc9d9bf15d4efd7c7b2e0a40188917fd6a5

@@
 
-%\VignetteIndexEntry{disttree: Distributional Trees and Forests}
+%\VignetteIndexEntry{disttree: Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
 %\VignetteDepends{partykit, gamlss.dist}
@@
 
-\title{Distributional Trees and Forests}
-\Plaintitle{disttree: Distributional Trees and Forests}
+\title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
 
@@
 <<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
+par(mar = c(2, 0, 2, 0))
 plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+usr <- par("usr")
+rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
 polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
@@
 points(x, y, col = "slategray", pch = 19)
-box(lwd=12)
+box(lwd=8)
 @
@@
 <<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+par(mar = c(2, 0, 2, 0))
+plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
+usr <- par("usr")
+rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
 lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
@@
   col = pallight["forest"], border = "transparent")
-box(lwd=12)
+points(x, y, col = "slategray", pch = 19)
+box(lwd=8)
 @
@@
 \draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=1.4pt](7.1,0)--(7.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=1.4pt](0,-2.4)--(0,-3.2);
-\draw[-{>[scale=4, length=2, width=3]},line width=1.4pt](2.2,-5.4)--(7.9,-2.5);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](7.1,0)--(7.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](0,-2.4)--(0,-3.2);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](2.2,-5.4)--(7.9,-2.5);
 \end{tikzpicture}

### 2017-12-14 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1877 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c7f3930e4c0f9ce6bfdf8827e518c3aadfc5c792

@@
 \Abstract{
-While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to also model all other distribution parameters.
+While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to model all distribution parameters.
 In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting.
@@
 This class of models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
-We are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-Applying these novel approaches to precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
+In particular, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
+Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
 
@@
 Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects as illustrated in Figure~\ref{Devel_parmod}.
-However, since these models consider only the expectation of the underlying distribution further development was required.
-Therefore, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
+These models consider only the expectation of the underlying distribution, however, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
 For example in the field of precipitation forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) usually acting on the assumption of a normally distributed response variable allow for both parameters, the mean and the variance, to depend linearly on covariates.
 Based on the resulting model the probability of precipitation can be estimated as well as the expected amount of precipitation.
-Therefore, these models provide common and widely used tools, also due to their intuitive model structure and their interpretability.
-However, the restriction to model only linear effects in the parameters has risen the interest in applying more flexible models.
-Generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
+Therefore, these models provide common and widely used tools, also due to their intuitive structure and interpretability.
+However, the restriction to model only linear effects has risen the interest in applying more flexible models.
+Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
 These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g.
@@
 For these reasons, regression trees and particularly random forests provide valuable features which can contribute to improving models for probabilistic forecasting.
-But, up to now, no general distributional framework is available for these methods.
-Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
+Therefore, we are going to introduce a novel general distributional framework for these methods by presenting distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
 The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure~\ref{Devel_treeforest}.
@@
 After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}.
-The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
+The new method is applied on precipitation data in a mountainous region with a very large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
 In a first setting only one observation station is taken into consideration.
@@
 
-With the specification of the distribution family and the log-likelihood function the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$.
+With the specification of the distribution family and the log-likelihood function the task of fitting a distributional model turns into the task of determining the distribution parameter~$\theta$.
 This is done by the following maximization.

### 2017-12-18 pkg/disttree/inst/draft/draft.Rnw
comments TH
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1891 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3ef54735073fbcd974b60f9b815351ab387000f6

@@
 A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
-This class of models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
+%%%<TH> no references in abstract </TH>
+This class of models 
+%%%<TH> "this" means trees and forests? </TH>
+will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
 In particular, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
+%%%<TH> break-up very long sentence </TH>
 Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
@@
   Torsten Hothorn\\
-  Institut f\"ur Sozial- und Pr\"aventivmedizin, Abteilung Biostatistik \\
+  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
   Universit\"at Z\"urich \\
@@
 Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects as illustrated in Figure~\ref{Devel_parmod}.
+%%%<TH> more explantion for Fig 1 needed </TH>
 These models consider only the expectation of the underlying distribution, however, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
@@
 However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model.
-In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
+In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees
+%%%<TH> citation Loh does not fit, maybe omit citations anyway </TH>
+(\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
 Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process.
@@
 The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure~\ref{Devel_treeforest}.
+%%% <TH> better description for Fig 2 </TH>
 
@@
 For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a forest model are plotted together with the corresponding observations.
-This illustrates that well fitting predicted expected values are provided by this model as they are close to the observed values.
+This illustrates that well fitting predicted expected values
+%%% well fitting predicted expected values -> appropriate probabilistic forecasts? </TH>
+are provided by this model as they are close to the observed values.
 Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
@@
 This is also the chosen approach in the methodology described in this section.
+%%% <TH> is is repeated on page 5, remove here </TH>
 
@@
 \label{disttree}
-Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors) to decide whether the data set should be split into subgroups or not.
-In particular, the information provided by the fitted model are the scores
+Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors)
+%%% <TH> X -> Z is confusing. Either we always write theta = g(X) where
+%%% g is now a tree or we explain that theta = g(X)(Z) is what we want.
+%%% Here, we don't partition models with X variables, so we should stick to X. </TH>
+to decide whether the data set should be split into subgroups or not. In particular, the information provided by the fitted model are the scores 
 $$
@@
 In a next step a local distributional model is fit in each of the subgroups.
-Repeating this procedure of fitting a model, applying statistical tests to evaluate the goodness of fit and splitting the data set depending on the test results leads to a tree structure.
-In that way the learning data set is separated into \textit{B} disjoint segments such that
+Repeating this procedure of fitting a model, applying statistical tests to evaluate the goodness of fit %%%<TH> avoid goodness of fit, this term is too ambiguous here </TH>
+and splitting the data set depending on the test results leads to a tree structure. In that way the learning data set is separated into \textit{B} disjoint segments such that
 $$
@@
 The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.
+%%% <TH> we need a little more detail here or a reference to a better
+%%% description of mob/ctree </TH>
 
@@
 
-For distributional trees the strategy of how to make a prediction for a new set of covariates $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ is very simple and straight forward.
+For distributional trees the strategy of how to make a prediction for a new set of covariates $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ 
+%%% <TH> \bold{z} is sufficient, remove ^\ast </TH>
+is very simple and straight forward. 
+%%% <TH> better avoid this judgement </TH>
 All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new set of covariates are selected and a distributional model is fit on the subset.
 To select this subset a weight vector $\bold{\omega}^{tree} = (\omega^{tree}_1, \ldots, \omega^{tree}_n)$ is included in the fitting process.
+%%% <TH> tree -> \text{tree}
 $$
@@
 
+%%% <TH> we need to be much more precise here: random forests is Breiman's and Cutler's
+%%% random forest. Maybe it is better to not mention this method at any time. </TH>
 Common forest models such as random forests consist of trees only considering the mean of the distribution of the response variable.
@@
 This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights
+%%% <TH> we need references for the nearest neighbor weights, copy from
+%%% "Transformation Forests" </TH>
+%%% <TH> tree -> \text{tree}, forest -> \text{forest} </TH>
 
@@
 
-For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$.
+For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$:
 $$
@@
 where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
+
+%%%<TH> we need to cite "Transformation Forests" for the interpretation of
+%%% random forests as adaptive nearest neighbors </TH>
+
 This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
@@
 
+%%% <TH> this should go into the discussion or the interpretation of the
+%%% empirical results </TH>
 Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
@@
 
+%%% <TH> hm, now we _are_ using a transformation model without estimating the
+%%% transformation. </TH>
 To remove large parts of the skewness of precipitation data, a power transformation has been used frequently (Box and Cox 1964).
@@
 where \(\mathbf{1}\) is the indicator function.
+%%% <TH> y > z??? </TH>
 
@@
 Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on probability integral transform (PIT) histograms (\citealt{gneiting2007}).
-
+%%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>
+
+%%%<TH> always us the term "distributional forest"
 As shown in Table~\ref{model_specification} the forest model includes all 82 covariates.

### 2017-12-22 pkg/model4you/inst/reproducing_papers/2016_subgroup_analyses_ijb.Rmd
reproduce papers with model4you
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1908 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 58bb0202c2b609fc149088d695b33158752f58d5

---
title: "model4you reproducing "Model-Based Recursive Partitioning for Subgroup Analyses""
author: "Heidi Seibold"
date: "December 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("model4you")
library("survival")
load("data/ALSFRSdata.rda")
load("data/ALSsurvdata.rda")
```

## Base models
```{r}
bm_alsfrs <- glm(ALSFRS.Start ~ Riluzole + offset(log(ALSFRS.halfYearAfter)), 
                 data = ALSFRSdata, family = gaussian(link = "log"))
summary(bm_alsfrs)

bm_alssurv_wb <- survreg(Surv(survival.time, cens) ~ Riluzole, 
                         data = ALSsurvdata, dist = "weibull")
summary(bm_alssurv_wb)

bm_alssurv_cox <- coxph(Surv(survival.time, cens) ~ Riluzole, 
                        data = ALSsurvdata)
summary(bm_alssurv_cox)
```

## Trees

```{r}
set.seed(123)

(tr_alsfrs <- pmtree(bm_alsfrs, control = ctree_control(maxdepth = 2)))
plot(tr_alsfrs, terminal_panel = node_pmterminal(tr_alsfrs, plotfun = lm_plot, 
                                                 confint = FALSE))

(tr_alssurv_wb <- pmtree(bm_alssurv_wb, control = ctree_control(maxdepth = 2)))
plot(tr_alssurv_wb, terminal_panel = node_pmterminal(tr_alssurv_wb, plotfun = survreg_plot, 
                                                     confint = TRUE))

(tr_alssurv_cox <- pmtree(bm_alssurv_cox, control = ctree_control(maxdepth = 2)))
```

### 2017-12-22 pkg/model4you/inst/reproducing_papers/2017_individual_treatment_effect_als_smmr.Rmd
reproduce papers with model4you
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1908 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 58bb0202c2b609fc149088d695b33158752f58d5

---
title: model4you reproducing "Individual treatment effect prediction for amyotrophic
  lateral sclerosis patients"
author: "Heidi Seibold"
date: "December 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r}
library("model4you")
library("survival")
library("ggplot2")
load("data/ALSFRSdata.rda")
load("data/ALSsurvdata.rda")
```

## Base models
```{r}
bm_alsfrs <- glm(ALSFRS.Start ~ Riluzole + offset(log(ALSFRS.halfYearAfter)), 
                 data = ALSFRSdata, family = gaussian(link = "log"))
summary(bm_alsfrs)

bm_alssurv <- survreg(Surv(survival.time, cens) ~ Riluzole, 
                      data = ALSsurvdata, dist = "weibull")
summary(bm_alssurv)
```
## Forests
```{r}
set.seed(123)

frst_alsfrs <- pmforest(bm_alsfrs, ntree = 100, 
                        control = ctree_control(teststat = "max", testtype = "Univ",
                                                mincriterion = 0.95, minsplit = 40, 
                                                minbucket = 30, lookahead = TRUE))

frst_alssurv <- pmforest(bm_alssurv, ntree = 100, 
                         control = ctree_control(teststat = "max", testtype = "Univ",
                                                 mincriterion = 0.95, minsplit = 40, 
                                                 minbucket = 30, lookahead = TRUE))
```

## Personalised models
```{r}
pm_alsfrs <- pmodel(frst_alsfrs)

coeffun <- function(model) {
  ## model coefficients
  coefs <- c(coef(model), scale = model$scale)

  ## difference in median survival 
  p = 0.5
  coefs["median_s0"] <- qweibull(p = p, shape = 1/coefs["scale"], 
                                 scale = exp(coefs["(Intercept)"]))
  coefs["median_s1"] <- qweibull(p = p, shape = 1/coefs["scale"], 
                                 scale = exp(coefs["(Intercept)"] + coefs["aA"]))
  coefs["median_sdiff"] <- coefs["median_s1"] - coefs["median_s0"]

  return(coefs)
}
pm_alssurv <- pmodel(frst_alssurv, fun = coeffun)
```

## Variable importance
```{r}
vi_alsfrs <- varimp(frst_alsfrs, 
                    risk = function(x, ...) -objfun(x, sum = TRUE, ...), 
                    nperm = 1)
var <- names(vi_alsfrs)
vi_alsfrs_d <- data.frame(variable = factor(var, levels = var[order(vi_alsfrs)]), 
                          VI = vi_alsfrs)

ggplot(vi_alsfrs_d, aes(x = variable, y = VI)) + 
  geom_bar(stat = "identity", width = .1) +
  coord_flip() +
  theme(panel.grid.major.y = element_blank())

vi_alssurv <- varimp(frst_alssurv, 
                     risk = function(x, ...) -objfun(x, sum = TRUE, ...), 
                     nperm = 1)
var <- names(vi_alssurv)
vi_alssurv_d <- data.frame(variable = factor(var, levels = var[order(vi_alssurv)]), 
                          VI = vi_alssurv)

ggplot(vi_alssurv_d, aes(x = variable, y = VI)) + 
  geom_bar(stat = "identity", width = .1) +
  coord_flip() +
  theme(panel.grid.major.y = element_blank())
```

## dependence plot
```{r}
dp_alsfrs <- cbind(pm_alsfrs, ALSFRSdata)

dp_plot_alsfrs <- function(var) {
  print(ggplot(dp_alsfrs, aes_string(y = "RiluzoleYes", x = var)) +
    geom_point() + geom_smooth(se = FALSE))
}
dp_plot_alsfrs("time_onset_treatment")
dp_plot_alsfrs("value_creatinine")
dp_plot_alsfrs("value_phosphorus")
dp_plot_alsfrs("subjectliters_fvc")
```

### 2018-01-04 pkg/disttree/inst/draft/draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1902 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 58d1788f010af26e5f2bfa24b247f6e9b58a4305
small corrections and additional references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1919 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b4519e640ab265fb8ea81a1915a085e5a036d9c7

@@
       col <- c(col, pal["forest"])
-      legendnames <- c(legendnames, "forest")
+      legendnames <- c(legendnames, "distributional forest")
     }
@@
     col <- c(col, pal["forest"])
-    legendnames <- c(legendnames, "forest")
+    legendnames <- c(legendnames, "distributional forest")
   }
@@
 In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting.
-In generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects.
-A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
-%%%<TH> no references in abstract </TH>
-This class of models 
-%%%<TH> "this" means trees and forests? </TH>
-will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
+In generalized additive models for location, scale and shape %(GAMLSS, \citealt{stasinopoulos2005})
+each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects.
+A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees %(\citealt{loh2011classification})
+or random forests. %(\citealt{breiman2001random}). 
+%%%<TH> "random forest" generisch; Breiman and Cutler's randomForest </TH>
+This class of tree and forest models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
 In particular, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
@@
   E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/lisa-schlosser.html}, \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/}, \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\
 
@@
 In regression analysis various models have been developed to describe the relationship between a response variable and a (set of) covariate(s).
-Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects as illustrated in Figure~\ref{Devel_parmod}.
-%%%<TH> more explantion for Fig 1 needed </TH>
+Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects.
+In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate.
+In the left picture the expected value of the response variable is modeled by a linear function in the covariate.
+In the middle a GAM is applied where a smooth function allows for more flexibility.
+
 These models consider only the expectation of the underlying distribution, however, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
@@
 However, the restriction to model only linear effects has risen the interest in applying more flexible models.
-Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM.
+Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM as illustrated in the right plot of Figure~\ref{Devel_parmod}.
 These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g.
@@
 However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model.
-In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees
+In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees (\citealt{breiman1984classification})
 %%%<TH> citation Loh does not fit, maybe omit citations anyway </TH>
-(\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
+%(\citealt{loh2011classification}) 
+or random forests such as Breiman and Cutler's randomForest (\citealt{breiman2001random}).
 Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process.
@@
 Therefore, we are going to introduce a novel general distributional framework for these methods by presenting distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-The developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated in Figure~\ref{Devel_treeforest}.
-%%% <TH> better description for Fig 2 </TH>
+In Figure~\ref{Devel_treeforest} the developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated for the case of one covariate.
+In the top left picture the step function modeling the mean of the response variable represents the division into three terminal subgroups by a regression tree.
+This is extended in the bottom left picture where a distributional tree models not only the mean but also the variance of the response variable by its own step function.
+The top middle picture shows that combining an ensemble of trees to a random forest can lead to smoother effects while the tree structure assures that abrupt changes are captured as well.
+Again, an extension of the random forest model considering only the mean is made in the top right picture.
+Here a distributional forest models the mean and the variance capturing abrupt changes but also smooth effects.
 
@@
 A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}.
-For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a forest model are plotted together with the corresponding observations.
+For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a distributional forest are plotted together with the corresponding observations.
 This illustrates that well fitting predicted expected values
@@
 Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
-In section \ref{ProbabilisticForecasting} this example will be further discussed and it will be shown that the predictions made by the forest model are reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, the scale parameter, and therefore the whole distribution.
+In section \ref{ProbabilisticForecasting} this example will be further discussed and it will be shown that the predictions made by the distributional forest are reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, the scale parameter, and therefore the whole distribution.
 
@@
 @
-\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest model learned on data from 1985--2008.
+\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
 Observations and predictions are left censored at 0.
@@
 \begin{cases}
-    \log\left(\frac{1}{\sigma} \cdot \phi(\frac{Y - \mu}{\sigma}) \right), & \text{if } Y > 0\\
+    \log\left(\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right), & \text{if } Y > 0\\
     \vspace{0.1cm}
-    \log\left(\Phi(\frac{-\mu}{\sigma})\right), & \text{if } Y = 0
+    \log\left(\Phi\left(\frac{-\mu}{\sigma}\right)\right), & \text{if } Y = 0
 \end{cases}
@@
 
-A common way of including these covariates is to consider them as regressors in a distributional model such that each parameter $\theta$ depends on covariates $\bold{X} = (X_1,\ldots,X_m)$.
+A common way of including these covariates is to consider them as regressors in a distributional model such that each parameter $\theta$ depends on covariates $\bold{Z} = (Z_1,\ldots,Z_m)$.
 This could for example be done by a linear function $g$.
-In that way the fitted model would be specified as $D(Y, \theta = g(\bold{X}))$.
-In order to avoid the restriction to linear effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ to specify a more flexible model $D(Y, \theta = f_1(\bold{X}) + \ldots + f_p(\bold{X}))$ applying a GAM structure.
+In that way the fitted model would be specified as $D(Y, \theta = g(\bold{Z}))$.
+In order to avoid the restriction to linear effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ to specify a more flexible model $D(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure.
 Another approach to distributional regression is proposed by \citealt{Klein2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
@@
 \label{disttree}
-Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors)
+Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ %(from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors)
 %%% <TH> X -> Z is confusing. Either we always write theta = g(X) where
@@
 In a next step a local distributional model is fit in each of the subgroups.
-Repeating this procedure of fitting a model, applying statistical tests to evaluate the goodness of fit %%%<TH> avoid goodness of fit, this term is too ambiguous here </TH>
-and splitting the data set depending on the test results leads to a tree structure. In that way the learning data set is separated into \textit{B} disjoint segments such that
+Repeating this procedure of fitting a model, applying statistical tests to evaluate parameter instability and splitting the data set depending on the test results leads to a tree structure.
+In that way the learning data set is separated into \textit{B} disjoint segments such that
 $$
@@
 %%% <TH> we need a little more detail here or a reference to a better
-%%% description of mob/ctree </TH>
+%%% description of mob/ctree OR appendix</TH>
 
@@
 
-For distributional trees the strategy of how to make a prediction for a new set of covariates $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ 
-%%% <TH> \bold{z} is sufficient, remove ^\ast </TH>
-is very simple and straight forward. 
-%%% <TH> better avoid this judgement </TH>
+For distributional trees a prediction for a new set of covariates $\bold{z} = (z_1, \ldots, z_m)$ is made with the following strategy: 
 All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new set of covariates are selected and a distributional model is fit on the subset.
-To select this subset a weight vector $\bold{\omega}^{tree} = (\omega^{tree}_1, \ldots, \omega^{tree}_n)$ is included in the fitting process.
-%%% <TH> tree -> \text{tree}
+To select this subset a weight vector $\bold{w}^{\text{tree}} = (w^{\text{tree}}_1, \ldots, w^{\text{tree}}_n)$ is included in the fitting process.
 $$
-\omega^{tree}_i(\bold{z}^{\ast}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z}^{\ast} \in \mathcal{B}_b))
+w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b))
 $$
-The predicted distribution of the new set of covariates is then fully specified by the estimated parameter $\hat{\theta}(\bold{z}^{\ast})$ where
+The predicted distribution of the new set of covariates is then fully specified by the estimated parameter $\hat{\theta}(\bold{z})$ where
 $$
-\hat{\theta}(\bold{z}^{\ast}) = \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{tree}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i).
+\hat{\theta}(\bold{z}) = \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\theta; y_i).
 $$
@@
 %%% random forest. Maybe it is better to not mention this method at any time. </TH>
-Common forest models such as random forests consist of trees only considering the mean of the distribution of the response variable.
+Common forest models such as Breiman and Cutler's randomForest consist of trees only considering the mean of the distribution of the response variable.
 For that reason, changes in other parameters can only be detected if they are connected to changes in the location parameter, e.g.
@@
 Forest models differ mainly in the way how the trees are combined to make predictions.
+Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{z}$ and the resulting subgroups and weighing schemes.
 In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.
-
-Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates.
-This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights
+In transformation forests by \cite{transformationforests2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations.
+This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``random survival forests'' by \cite{ishwaran2008}.
+The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
+%Similar observations have a high probability of ending up in the same terminal node whereas this probability is low for quite different observations. 
+%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (“bagging survival trees”, Hothorn et al. 2004). 
+In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 
+$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
+$$
+w^{\text{forest}}_i(\bold{z}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z} \in \mathcal{B}^s_b))
+$$
+where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.
+
+Therefore, contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates.
+Compared to the formula for predictions made by distributional trees the vector of tree weights $\bold{w}^{\text{tree}}$ containing only zeros and ones is now replaced by a vector of integer valued forest weights.
 %%% <TH> we need references for the nearest neighbor weights, copy from
 %%% "Transformation Forests" </TH>
-%%% <TH> tree -> \text{tree}, forest -> \text{forest} </TH>
-
-$\bold{\omega}^{forest}~=~(\omega^{forest}_1, \ldots, \omega^{forest}_n)$.
-
-For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$:
+%%%<TH> we need to cite "Transformation Forests" for the interpretation of
+%%% random forests as adaptive local ML </TH>
+This individual set of weights $\bold{w}^{\text{forest}}(\bold{z}) = (w^{\text{forest}}_1(\bold{z}), \ldots, w^{\text{forest}}_n(\bold{z}))$ for the new set of covariates $\bold{z}$ can now be included in the estimation process of the distribution parameters and in that way leads to its individual parameter vector $\hat{\theta}(\bold{z})$.
 $$
-\omega^{forest}_i(\bold{z}^{\ast}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z}^{\ast} \in \mathcal{B}^s_b))
-$$
-where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
-
-%%%<TH> we need to cite "Transformation Forests" for the interpretation of
-%%% random forests as adaptive nearest neighbors </TH>
-
-This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
-$$
-\hat{\theta}(\bold{z}^{\ast}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{forest}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i)
+\hat{\theta}(\bold{z}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\theta; y_i)
 $$
@@
 \hline
-\emph{msl{\_}diff}: mean sea level pressure & 1 & msl{\_}mean{\_}max $-$ msl{\_}mean{\_}min\\
+\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
 \hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
@@
 \hline
-\textbf{Forest} & recursive    & all & all \\ 
+\textbf{distributional forest} & recursive    & all & all \\ 
                 & partitioning &     &     \\ 
 \hline
-\textbf{prespecified GAM} & spline  & tp{\_}mean, & tp{\_}sprd,\\
-                            & in each & tp{\_}max,  & dswrf{\_}sprd{\_}mean,\\
- & & tp{\_}mean1218 $\ast$ & tp{\_}sprd1218 $\ast$\\ 
- & & \quad cape{\_}mean1218, & \quad  cape{\_}mean1218,\\
- & & dswrf{\_}mean{\_}mean, & tcolc{\_}sprd{\_}mean,\\
- & & tcolc{\_}mean{\_}mean, & tdiff500850{\_}mean\\
- & & pwat{\_}mean{\_}mean, & \\
- & & tdiff500850{\_}mean, & \\
- & & msl{\_}diff & \\
+\textbf{prespecified GAM} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
+                            & in each & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean},\\
+ & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
+ & & \quad \emph{cape{\_}mean1218}, & \quad \emph{cape{\_}mean1218},\\
+ & & \emph{dswrf{\_}mean{\_}mean}, & \emph{tcolc{\_}sprd{\_}mean},\\
+ & & \emph{tcolc{\_}mean{\_}mean}, & \emph{tdiff500850{\_}mean}\\
+ & & \emph{pwat{\_}mean{\_}mean}, & \\
+ & & \emph{tdiff500850{\_}mean}, & \\
+ & & \emph{msl{\_}diff} & \\
 \hline
@@
 \hline
-\textbf{EMOS} & linear & tp{\_}mean & tp{\_}sprd \\
+\textbf{EMOS} & linear & \emph{tp{\_}mean} & \emph{tp{\_}sprd} \\
 \hline
@@
 The results of fitting one distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of this forest model and compare its results to those of the GAM models and the EMOS model based on probability integral transform (PIT) histograms (\citealt{gneiting2007}).
+Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on probability integral transform (PIT) histograms (\citealt{gneiting2007}).
 %%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>
@@
 %%%<TH> always us the term "distributional forest"
-As shown in Table~\ref{model_specification} the forest model includes all 82 covariates.
+As shown in Table~\ref{model_specification} the distributional forest includes all 82 covariates.
 A single tree model allows to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable.
-In a forest model though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree.
+In a distributional forest though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree.
 To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied.
-Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described forest model for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}.
+Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}.
 As expected, the numerical forecast for total precipitation is the most influential variable.
@@
 barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
-        horiz = TRUE, las = 1, cex.names = 1.2, axes = TRUE, 
+        horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE,
+        font.axis = 3, #list(family="HersheySerif", face=3),
         names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
-#axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(-1,0.5,0))
+axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
 @
@@
 %\end{subfigure}
-\caption{\label{varimp} Variable importance of the forest model for station Axams learned on data from 1985--2008 and assessed in 2009--2012 based on mean decrease in accuracy regarding the CRPS (top 10 list).}
+\caption{\label{varimp} Variable importance of the distributional forest for station Axams learned on data from 1985--2008 and assessed in 2009--2012 based on mean decrease in accuracy regarding the CRPS (top 10 list).}
 \end{center}
@@
 par(mfrow = c(2,2))
-pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Forest", ylim = c(0,1.5))
+pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "distributional forest", ylim = c(0,1.5))
 pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM", ylim = c(0,1.5))
@@
 \center
-<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5>>=
+<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
 boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
-        names = c("Forest", "prespecified GAM", "boosted GAM"),
+        names = c("distributional forest", "prespecified GAM", "boosted GAM"),
         ylab = "CRPS skills score", col = "lightgray") 
@@
 
-The CRPS skills score boxplot shows that, in this setting, both GAM models and the forest perform better than the EMOS model.
-While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the forest even reaches an improvement rate close to 6 percent.
+The CRPS skills score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
+While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
 
@@
 Variable selection is done automatically and interactions are detected by the tree structure of the model.
-In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the forest model.
+In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the distributional forest model.
 Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
@@
 s <- 1 - crps_all[, 2:4]/crps_all[,6]
-colnames(s) <- c("Forest", "prespecified GAM", "boosted GAM")
+colnames(s) <- c(" distributional forest ", "prespecified GAM", "boosted GAM")
 
@@
 \begin{center}
-<<rain_all_crps_skills_score, fig=TRUE, echo=FALSE>>=
+<<rain_all_crps_skills_score, fig=TRUE, echo=FALSE, width = 6.5>>=
   matplot(t(s[,]), type = "l", lwd = 2, 
@@
 Each observation station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station.
-The color of the symbol indicates the difference of the CRPS of the forest model and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
+The color of the symbol indicates the difference of the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
 
@@
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend("topleft", pch=c(21,24,25,22), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
+  legend("topleft", pch=c(21,24,25,22), legend = c("distributional forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
 
@@
 The type of symbol indicates which model performed best with respect to the CRPS.
-The color indicates the difference of the CRPS of the forest model and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).}
+The color indicates the difference of the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).}
 \end{center}
@@
 The function \code{disttree} uses either the \code{mob} or \code{ctree} function as a framework to build a tree.
-Similarly, the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a forest.
+Similarly, the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a distributional forest.
 All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}).

### 2018-01-08 pkg/model4you/inst/reproducing_papers/2017_individual_treatment_effect_als_smmr.Rmd
reproducing papers
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1927 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0a30edddd44a1d218ade492fa1d92b42bb06195c

@@
 ---
-title: model4you reproducing "Individual treatment effect prediction for amyotrophic
-  lateral sclerosis patients"
+title: "model4you reproducing: Individual treatment effect prediction for amyotrophic lateral sclerosis patients"
 author: "Heidi Seibold"

### 2018-01-08 pkg/model4you/R/plot-pmtree.R

@@ -112,7 +112,7 @@ binomial_glm_plot <- function(mod, data = NULL, plot_data = FALSE, theme = theme
 #'
 #' Can be used on its own but is also useable as plotfun in 
 #' \code{\link{node_pmterminal}}.
-#'
+#' 
 #' @param mod A model of class lm.
 #' @param data optional data frame. If NULL the data stored in mod is used.
 #' @param densest should additional to the model density kernel density estimates
@@ -121,8 +121,11 @@ binomial_glm_plot <- function(mod, data = NULL, plot_data = FALSE, theme = theme
 #' @param yrange Range of the y variable to be used for plotting. 
 #'  If NULL the range in the data will be used.
 #'
+#' @details In case of an offset, the value of the offset variable will be set to
+#' the median of the values in the data.
+#'
 #' @examples 
-#' ## exmaple taken from ?lm
+#' ## example taken from ?lm
 #' ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
 #' trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
 #' group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
@@ -130,6 +133,13 @@ binomial_glm_plot <- function(mod, data = NULL, plot_data = FALSE, theme = theme
 #' data <- data.frame(weight, group)
 #' lm.D9 <- lm(weight ~ group, data = data)
 #' lm_plot(lm.D9)
+#' 
+#' ## example taken from ?glm (modified version)
+#' data(anorexia, package = "MASS")
+#' anorexia$treatment <- anorexia$Treat != "Cont"
+#' anorex.1 <- glm(Postwt ~ treatment + offset(Prewt),
+#'                 family = gaussian, data = anorexia)
+#' lm_plot(anorex.1)
 #' 
 #' 
 #' @importFrom ggplot2 ggplot geom_line theme_classic aes_string xlim xlab scale_linetype_discrete
@@ -144,16 +154,22 @@ lm_plot <- function(mod, data = NULL, densest = FALSE, theme = theme_classic(),
   modcall <- getCall(mod)
   modformula <- as.Formula(eval(modcall$formula))
   xformula <- formula(modformula, lhs = 0, rhs = 1)
+  offset_id <- attr(terms(xformula), "offset")
   yformula <- formula(modformula, lhs = 1, rhs = 0)
   if(is.null(data)) data <- eval(modcall$data)
[…]

### 2018-01-10 pkg/model4you/inst/reproducing_papers/2017_individual_treatment_effect_als_smmr.Rmd
trying to get varimp to work
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1933 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b62ce9ea3101aa7f0c075f5a02dbfe3b24494af4

@@
 ```{r}
+set.seed(123)
 vi_alsfrs <- varimp(frst_alsfrs, 

### 2018-01-10 pkg/model4you/R/pmtree-methods.R

@@ -126,16 +126,17 @@ objfun.pmtree <- function(x,
   
   ## if the order does not matter, this is faster
   if(sum) {
+    if(!is.null(weights) & is.null(newdata)) {
+      # which_node[weights == 0] <- NA
+      newdata <- x$data
+    } 
     
     ## get contributions of objfun in each node
     get_objfun_node_unordered <- function(nd) {
-      # if(is.null(newdata)) {
-      #   sum(objfun(mods[[as.character(nd)]], weights = weights[which_node == nd]))
-      # } else {
+      wn <- (which_node == nd) & !is.na(which_node)
       sum(objfun(mods[[as.character(nd)]],
-                 newdata = newdata[which_node == nd, ], 
-                 weights = weights[which_node == nd]))
-      # }
+                 newdata = newdata[wn, ], 
+                 weights = weights[wn]))
     }
     
     ## return the unordered contributions

### 2018-01-10 pkg/model4you/inst/reproducing_papers/2017_individual_treatment_effect_als_smmr.Rmd
finish reproducing papers
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1936 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 536dba33ea2da730ec3044df21a3de516be0d3c2

@@
 ---
-title: "model4you reproducing: Individual treatment effect prediction for amyotrophic lateral sclerosis patients"
+title: 'model4you reproducing: Individual treatment effect prediction for amyotrophic
+  lateral sclerosis patients'
 author: "Heidi Seibold"
 date: "December 22, 2017"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 ```{r}
+dp_plot <- function(var, data) {
+  p <- ggplot(data, aes_string(y = "RiluzoleYes", x = var))
+
+  if(is.factor(data[, var])) {
+    p <- p + geom_boxplot()
+  } else {
+    p <- p + geom_point() + geom_smooth(se = FALSE)
+  }
+
+  print(p)
+}
+
 dp_alsfrs <- cbind(pm_alsfrs, ALSFRSdata)
 
-dp_plot_alsfrs <- function(var) {
-  print(ggplot(dp_alsfrs, aes_string(y = "RiluzoleYes", x = var)) +
-    geom_point() + geom_smooth(se = FALSE))
-}
-dp_plot_alsfrs("time_onset_treatment")
-dp_plot_alsfrs("value_creatinine")
-dp_plot_alsfrs("value_phosphorus")
-dp_plot_alsfrs("subjectliters_fvc")
+dp_plot("time_onset_treatment", dp_alsfrs)
+dp_plot("value_creatinine", dp_alsfrs)
+dp_plot("value_phosphorus", dp_alsfrs)
+dp_plot("subjectliters_fvc", dp_alsfrs)
+
+dp_alssurv <- cbind(pm_alssurv, ALSsurvdata)
+
+dp_plot("age", dp_alssurv)
+dp_plot("time_onset_treatment", dp_alssurv)
+dp_plot("weakness", dp_alssurv)
+dp_plot("height", dp_alssurv)
+
 ```

### 2018-01-11 pkg/model4you/inst/reproducing_papers/2017_individual_treatment_effect_als_smmr.Rmd
simple varimp function
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1940 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f47c6c85720b0d11aa63be84672957b5ad03ddeb

@@
 ## Variable importance
-```{r}
+```{r varimp}
 set.seed(123)
-vi_alsfrs <- varimp(frst_alsfrs, 
-                    risk = function(x, ...) -objfun(x, sum = TRUE, ...), 
-                    nperm = 1)
+vi_alsfrs <- varimp(frst_alsfrs)
 var <- names(vi_alsfrs)
@@
 
-vi_alssurv <- varimp(frst_alssurv, 
-                     risk = function(x, ...) -objfun(x, sum = TRUE, ...), 
-                     nperm = 1)
+vi_alssurv <- varimp(frst_alssurv)
 var <- names(vi_alssurv)

### 2018-01-11 pkg/model4you/tests/test-pmodel.R

@@ -38,10 +38,10 @@ basemodel_lm <- lm(y ~ a, data = simdata)
 basemodel_wb <- survreg(Surv(y, cens) ~ a, data = simdata_s)
 
 ## forest
-frst_lm <- pmforest(basemodel_lm, ntree = 40, 
+frst_lm <- pmforest(basemodel_lm, ntree = 10, 
                     perturb = list(replace = FALSE, fraction = 0.632),
                     control = ctree_control(mincriterion = 0, lookahead = TRUE))
-frst_wb <- pmforest(basemodel_wb, ntree = 40, 
+frst_wb <- pmforest(basemodel_wb, ntree = 10, 
                     perturb = list(replace = FALSE, fraction = 0.632),
                     control = ctree_control(mincriterion = 0, lookahead = TRUE))
 
@@ -80,9 +80,5 @@ summary(coefs_wb)
 
 
 ## Variable importance
-of <- function(x, newdata = NULL, weights = NULL, 
-               perm = NULL, ...) {
-  - objfun(x, newdata = newdata, weights = weights, perm = perm, sum = TRUE, ...)
-}
-vi_lm <- varimp(frst_lm, risk = of, nperm = 1)
+vi_lm <- varimp(frst_lm)
 plot(data.frame(variable = names(vi_lm), VI = vi_lm))

### 2018-01-21 pkg/disttree/inst/draft/draft.Rnw
small corrections and qqr-plot now in appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1967 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 25666be79535b36b256c592f8f05a8a432ed2147

@@
 In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 
-$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
+$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_i,\bold{z}_i)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_i,\bold{z}_i)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
 $$
@@
 Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std.
-dev., ensemble minimum and ensemble maximum.
+deviation, ensemble minimum and ensemble maximum.
 All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
 Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std.
-dev.
-of mean/minimal/maximal values respectively.
+deviation of mean/minimal/maximal values respectively.
 Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.
@@
 \emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
-\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. dev. of sums over 24h, \\ 
+\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
 \emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
@@
 \hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
-\hspace*{\fill}                            &    & ensemble std. dev. of sums over 6h\\
+\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
 \hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
@@
 \emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
-\emph{pwat}: preciptable water          &   & ensemble std. dev. of mean values,\\
-\emph{tmax}: 2m maximum temperature     &   & ensemble std. dev. of minimal values,\\
-\hspace*{1.15cm}                          &   & ensemble std. dev. of maximal values,\\
+\emph{pwat}: preciptable water          &   & ensemble std. deviation of mean values,\\
+\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimal values,\\
+\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
 \emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
@@
 \end{figure}
-
-%\begin{figure}[t!]
-%\begin{subfigure}{0.5\textwidth}
-%<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
-%# in sample
-%set.seed(7)
-%par(mfrow = c(2,2))
-%set.seed(1)
-%qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
-%qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
-%qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
-%qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
-
-%@
-%\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
-%\end{figure}
-
-%\begin{figure}[t!]
-%%\begin{subfigure}{0.5\textwidth}
-%<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE>>=
-%    
-%# out of sample
-%set.seed(7)
-%par(mfrow = c(2,2))
-%set.seed(1)
-%qqrplot(pit_df_t, nsim = 1, main = "distforest (out of bag)")
-%qqrplot(pit_g_t, nsim = 1, main = "gamlss (out of bag)")
-%qqrplot(pit_gb_t, nsim = 1, main = "gamboostLSS (out of bag)")
-%qqrplot(pit_ml_t, nsim = 1, main = "EMOS (out of bag)")
-
-%@
-%\caption{\label{qqr_oosample} QQR-Plot for station Axams (out of sample)}
-%\end{figure}
 
@@
 %\newpage
-%\section*{Appendix}
+\section*{Appendix}
+
+%\begin{figure}[t!]
+%\begin{subfigure}{0.5\textwidth}
+%<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
+%# in sample
+%set.seed(7)
+%par(mfrow = c(2,2))
+%set.seed(1)
+%qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
+%qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
+%qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
+%qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
+
+%@
+%\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
+%\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+%\begin{subfigure}{0.8\textwidth}
+<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
+
+# out of sample
+set.seed(7)
+par(mfrow = c(2,2))
+set.seed(1)
+qqrplot(pit_df_t, nsim = 1, main = "distributional forest", ylim = c(-5,5))
+qqrplot(pit_g_t, nsim = 1, main = "prespecified GAM", ylim = c(-5,5))
+qqrplot(pit_gb_t, nsim = 1, main = "boosted GAM", ylim = c(-5,5))
+qqrplot(pit_ml_t, nsim = 1, main = "EMOS", ylim = c(-5,5))
+
+@
+\caption{\label{qqr_oosample} Out-of-sample QQR-plot for station Axams assessed in 2009--2012 based on models learned on data from 1985--2008.}
+\end{center}
+\end{figure}
 

### 2018-01-24 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1970 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 02f42108363b7fd61e0c06bede33de04319d1cc4

@@
 \title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
+\Shorttitle{Distributional Regression Forests for Probabilistic Precipitation Forecasting}
 
@@
 \section{Introduction}
-In regression analysis various models have been developed to describe the relationship between a response variable and a (set of) covariate(s).
+In regression analysis various models have been developed to describe the relationship between a response variable and a set of covariates.
 Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects.
@@
 \end{tikzpicture}
-\caption{\label{Devel_parmod}Development in the field of parametric modeling: (Generalized) linear models (LM, GLM; left), generalized additive models (GAM; middle), generalized additive models for location, scale, and shape (GAMLSS; right).}
+\caption{\label{Devel_parmod}Parametric modeling developments.
+(Generalized) linear models (left), generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
 \end{figure}
@@
 \end{tikzpicture}
-\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests: regression tree (top left), distributional tree (bottom), random forest (top middle), distributional forest (top right).}
+\caption{\label{Devel_treeforest}Tree and forest developments. Regression tree (top left), distributional tree (bottom), random forest (top middle), distributional forest (top right).}
 \end{figure}
@@
 \caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
-Observations and predictions are left censored at 0.
-The corresponding point mass is shown at the censoring point (0).}
+Observations are left censored at 0.
+The corresponding predicted point mass is shown at the censoring point (0).}
 \end{center}
@@
 \end{tabular}
-\caption[Table caption text]{Table of applied methods together with their type (way of including the covariates) and the included covariates for each distribution parameter.}
+\caption[Table caption text]{Overview of models with type of covariate dependency and included covariates for each distribution parameter.}
 \label{model_specification}
@@
 \begin{center}
-%\begin{subfigure}{0.7\textwidth}
+%\begin{subfigure}{0.65\textwidth}
 <<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
@@
         horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE,
+        xlab = "Variable importance: mean decrease in CRPS",
         font.axis = 3, #list(family="HersheySerif", face=3),
@@
 %\end{subfigure}
-\caption{\label{varimp} Variable importance of the distributional forest for station Axams learned on data from 1985--2008 and assessed in 2009--2012 based on mean decrease in accuracy regarding the CRPS (top 10 list).}
+\caption{\label{varimp} CRPS based variable importance for the top 10 covariates in the distributional forest. Based on data for station Axams, learning period 1985--2008 and assessed in 2009--2012.}
 \end{center}
@@
 %\end{subfigure}
-\caption{\label{pit_oosample}Out-of-sample PIT histograms for station Axams assessed in 2009--2012 based on models learned on data from 1985--2008.}
+\caption{\label{pit_oosample}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
 \end{center}
@@
 \center
-\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 10 times 7-fold cross validation at station Axams.}
+\caption{\label{boxplots_crps_axams}CRPS skills score from the 10 times 7-fold cross validation at station Axams (1985--2012). The horizontal orange line pertains to the reference model EMOS.}
 \end{center}
@@
 @
-\caption{\label{boxplot_crps_all}CRPS skills score over all stations with reference model EMOS. The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
+\caption{\label{boxplot_crps_all}CRPS skills score for each station (grey lines) and aggregated over all stations (boxplots).
+Station Axams is highlighted in greend and the horizontal orange line pertains to the reference model EMOS.
+The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
 \end{center}
@@
 @
-\caption{\label{map}Map of Tyrol with all observation stations.
-The type of symbol indicates which model performed best with respect to the CRPS.
-The color indicates the difference of the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).}
+\caption{\label{map}Map of Tyrol showing which model performed best at which station (type of symbol).
+The color codes whether the distributional forest had lower or higher CRPS (red or blue) compared to the best of the other three models (prespecified GAM, boosted GAM and EMOS).}
 \end{center}
@@
 par(mfrow = c(2,2))
-set.seed(1)
-qqrplot(pit_df_t, nsim = 1, main = "distributional forest", ylim = c(-5,5))
-qqrplot(pit_g_t, nsim = 1, main = "prespecified GAM", ylim = c(-5,5))
-qqrplot(pit_gb_t, nsim = 1, main = "boosted GAM", ylim = c(-5,5))
-qqrplot(pit_ml_t, nsim = 1, main = "EMOS", ylim = c(-5,5))
-
-@
-\caption{\label{qqr_oosample} Out-of-sample QQR-plot for station Axams assessed in 2009--2012 based on models learned on data from 1985--2008.}
+qqrplot(pit_df_t, nsim = 100, main = "distributional forest", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_g_t, nsim = 100, main = "prespecified GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_gb_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+
+qqrplot(pit_ml_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+@
+\caption{\label{qqr_oosample} Out-of-sample residual QQ-plot (2009--2012) for station Axams based on models learned on data from 1985--2008.}
 \end{center}

### 2018-01-29 pkg/disttree/inst/draft/draft.Rnw
added reference and edited appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1971 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d6a762d056d0747706e40a0df3fab209aad02f2f

@@
 The exact way of determining how and where to split varies across different tree algorithms.
-For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm (\citealt{zeileis2008model}) or the ctree algorithm (\citealt{hothorn2006unbiased}).
-Summed up, the steps of building a distributional tree using one of these algorithms are
+For distributional trees one possible choice is a permutation test based algorithm which is explained in detail in the appendix.
+Summed up, the steps of building a distributional tree are:
 \begin{enumerate}
 \item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\bold{\hat{\theta}}$ via maximum likelihood.
+\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(y_i, \theta) = 0$
 \item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
@@
 \label{distforest}
-A distributional forest consists of an ensemble of distributional trees $\{t_s\}_{s=1,\ldots,T}$.
+A distributional forest consists of an ensemble of $T$ distributional trees.
 Each of these trees is built on a different data set which can either be a bootstrap or a subset of the original data set.
@@
 Forest models differ mainly in the way how the trees are combined to make predictions.
-Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{z}$ and the resulting subgroups and weighing schemes.
+Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{Z}$ and the resulting subgroups and weighing schemes.
 In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.
 In transformation forests by \cite{transformationforests2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations.
-This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``random survival forests'' by \cite{ishwaran2008}.
+This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``bagging survival trees'' by \cite{hothorn2004}.
 The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
@@
 $$
-w^{\text{forest}}_i(\bold{z}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z} \in \mathcal{B}^s_b))
-$$
-where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.
+w^{\text{forest}}_i(\bold{z}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
+$$
+where $T$ is the number of trees of the forest, $B^t$ is the number of terminal nodes of the \textit{t}-th tree and $\mathcal{B}^t_b$ represents the \textit{b}-th terminal node of the \textit{t}-th tree.
 
@@
 At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.
-
-%%% <TH> this should go into the discussion or the interpretation of the
-%%% empirical results </TH>
-Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
-In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.
 
@@
 The results of fitting one distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on probability integral transform (PIT) histograms (\citealt{gneiting2007}).
+Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots.
+In the appendix the corresponding probability integral transform (PIT) histograms (\citealt{gneiting2007}) can be found.
 %%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>
@@
 <<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
-par(mar = c(2,10,2,2))
+par(mar = c(3,10,2,2))
 barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
@@
 
-To get another impression of the quality of the predictions made by the tested methods a PIT histogram is plotted in Figure~\ref{pit_oosample} for out-of-sample predictions of each of the models.
-A perfectly calibrated model would lead to a PIT histogram with a uniform distribution (a density of 1 for each bin as indicated by the red horizontal line).
-Overall, it can be stated that all four model provide well calibrated forecasts.
+In Figure~\ref{pit_oosample} another impression of the quality of the predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models.
+For these plots 100 randomized quantile residuals are simulated per observation.
+The closer the points are to the diagonal line the better the estimated distributional model fits the data.
+Overall, it can be stated that all four models provide well calibrated forecasts.
+
+\begin{figure}[t!]
+\begin{center}
+%\begin{subfigure}{0.8\textwidth}
+<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
+
+# out of sample
+set.seed(7)
+par(mfrow = c(2,2))
+qqrplot(pit_df_t, nsim = 100, main = "distributional forest", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_g_t, nsim = 100, main = "prespecified GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_gb_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+
+qqrplot(pit_ml_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+@
+\caption{\label{qqr_oosample} Out-of-sample residual QQ-plot (2009--2012) for station Axams based on models learned on data from 1985--2008.}
+\end{center}
+\end{figure}
 
@@
 %\begin{subfigure}{0.5\textwidth}
-%<<rain_axams_pit_insample, fig=TRUE, echo=FALSE>>=
-%
+%<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
 %# in sample
@@
 %par(mfrow = c(2,2))
-%pithist(pit_df_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%pithist(pit_g_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%pithist(pit_gb_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%pithist(pit_ml_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%
+%set.seed(1)
+%qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
+%qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
+%qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
+%qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
+
 %@
-%\caption{\label{pit_insample} PIT histograms for station Axams (in sample)}
+%\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
 %\end{figure}
-
-\begin{figure}[t!]
-\begin{center}
-%\begin{subfigure}{0.8\textwidth}
-<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
-
-# out of sample
-set.seed(7)
-
-par(mfrow = c(2,2))
-pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "distributional forest", ylim = c(0,1.5))
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM", ylim = c(0,1.5))
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM", ylim = c(0,1.5))
-pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
-
-@
-%\end{subfigure}
-\caption{\label{pit_oosample}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
-\end{center}
-\end{figure}
 
@@
 \begin{center}
-\center
 <<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
@@
 @
-\center
 \caption{\label{boxplots_crps_axams}CRPS skills score from the 10 times 7-fold cross validation at station Axams (1985--2012). The horizontal orange line pertains to the reference model EMOS.}
@@
 \begin{center}
-<<map, fig=TRUE, echo=FALSE, height=6, width=9>>=
+\setkeys{Gin}{width=0.99\textwidth}
+<<map, fig=TRUE, echo=FALSE, height=6.1, width=9>>=
 
@@
                 xlim = c(9.8,13.2), 
-                ylim = c(46.6, 47.8))
+                ylim = c(46.6, 47.85))
   plot(tirol.gadm, add = TRUE)
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend("topleft", pch=c(21,24,25,22), legend = c("distributional forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
-
+  legend(x = 9.8, y = 47.795, pch=c(21,24,25,22), legend = c("distributional forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
+  text(x = 10.3, y = 47.8, labels = "Models with lowest CRPS")
+  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.74), line = 0.3)
   par(mar = c(0.5,0.2,0.5,2.3))
@@
 
+%%% <TH> this should go into the discussion or the interpretation of the
+%%% empirical results </TH>
+Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
+In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.
+
 \section*{Computational details}
@@
 %\newpage
-%\section{References}
-
-%Hothorn T, Hornik K, Zeileis A (2006). 
-%``Unbiased Recursive Partitioning: A Conditional Inference Framework.''
-%\textit{Journal of Computational and Graphical Statistics}, 
-%\textbf{15}(3), 651--674. 
-%\doi{10.1198/106186006X133933}\\
-%\\
-%Zeileis A, Hothorn T, Hornik K (2008). 
-%``Model-Based Recursive Partitioning.''
-%\textit{Journal of Computational and Graphical Statistics}, 
-%\textbf{17}(2), 492--514. 
-%\doi{10.1198/106186008X319331}\\
-%\\
-%Hothorn T, Zeileis A (2015). 
-%``partykit: A Modular Toolkit for Recursive Partytioning in R.''
-%\textit{Journal of Machine Learning Research}, 
-%\textbf{16}, 3905--3909. 
-%\url{http://www.jmlr.org/papers/v16/hothorn15a.html}\\
-%\\
-%Stasinopoulos DM, Rigby RA (2007).
-%``Generalized Additive Models for Location Scale and Shape (GAMLSS) in R.''
-%\textit{Journal of Statistical Software}, 
-%\textbf{23}(7), 1--46.
-%\doi{10.18637/jss.v023.i07}\\
-%\\
-%Stasinopoulos DM, Rigby RA (2005).
-%``Generalized Additive Models for Location Scale and Shape (with discussion)''
-%\textit{Applied Statistics},
-%\textbf{54}(3), 507--554.
-%\doi{10.1111/j.1467-9876.2005.00510.x}\\
-%\\
-%Seibold H, Zeileis A, Hothorn T (2017).
-%``Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.''
-%\textit{Statistical Methods in Medical Research}, 
-%\textbf{12}(1), 45--63.
-%\doi{10.1177/0962280217693034}\\
-%\\
-%Hothorn T, Zeileis A (2017).
-%``Transformation Forests.''
-%\emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
-%\url{http://arxiv.org/abs/1701.02110}\\
-%\\
-%Messner JW, Mayr GJ, Zeileis A (2016).
-%``Heteroscedastic Censored and Truncated Regression with {crch}''
-%\textit{The R Journal},
-%\textbf{8}(1), 173--181.
-%\url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}\\
-%\\
-%Zeileis A, Wiel MA, Hornik K, Hothorn T (2008).
-%``Implementing a {Class} of {Permutation} {Tests}: {The} {Coin} {Package}''
-%\textit{Journal of Statistical Software},
-%\textbf{28}(8), 1--23.
-%\doi{10.18637/jss.v028.i08}\\
-%%publisher={American Statistical Association}
-%\\
-%Zeileis A, Hothorn T, Hornik K (2010).
-%``Party with the {MOB}: {Model-Based} {Recursive} {Partitioning} in {R}''
-%\textit{R package version 0.9-9999}.
-%\url{https://cran.r-project.org/web/packages/party/vignettes/MOB.pdf}\\
-%\\
-%Breiman L (2001).
-%``Random {Forests}''
-%\textit{Machine Learning},
-%\textbf{45}(1), 5--32
-%%\doi{10.1023/A:1010933404324}
-%%publisher={Springer}\\
-%\\
-%Loh WY (2011).
-%``Classification and {Regression} {Trees}''
-%\textit{Wiley Interdisciplinary Reviews: Data Mining and Knowledge Discovery},
-%\textbf{1}(1), 14--23.
-%\doi{10.1002/widm.8}
-%%publisher={Wiley Online Library}\\
-%\\
-%Hothorn T (Maintainer), Hornik K, Strobl C, Zeileis A (2015).
-%``Package 'party' ''
-%\textit{Package Reference Manual for Party Version 0.9--0.998},
-%\textbf{16}, 37.\\
-%%  \doi{}
-%\\
-%Yee TW (2010).
-%``The {VGAM} {Package} for {Categorical} {Data} {Analysis}''
-%\textit{Journal of Statistical Software},
-%\textbf{32}(10), 1--34.
-%\url{http://www.jstatsoft.org/v32/i10/}\\
-%\\
-%Umlauf N, Klein N, Zeileis A (2017).
-%``{BAMLSS}: {B}ayesian Additive Models for Location, Scale
-%  and Shape (and Beyond)''
-%\textit{Working Papers in Economics and Statistics, Research
-%  Platform Empirical and Experimental Economics, Universit\"at
-%  Innsbruck},
-%\textbf{2017-05},
-%\url{http://EconPapers.RePEc.org/RePEc:inn:wpaper:2017-05}\\
-%%type = {Working Paper},
-%%number = {2017-05},
-%%month = {February},
-%\\
-%Benjamin Hofner and Andreas Mayr and Nora Fenske and
-%  Matthias Schmid (2017). 
-%``{gamboostLSS}: Boosting Methods for {GAMLSS} Models''
-%\textit{{R} package version 2.0-0}.
-%\url{https://CRAN.R-project.org/package=gamboostLSS}\\
-%\\
-%Liaw A, Wiener M (2002).
-%``Classification and Regression by randomForest''
-%\textit{R News},
-%\textbf{2}(3), 18--22.
-%\url{http://CRAN.R-project.org/doc/Rnews/}
-%\\
-%Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-%``Calibrated probabilistic forecasting using ensemble model output statistics and minimum CRPS estimation''
-%\textit{Monthly Weather Review},
-%\textbf{133}(5), 1098--1118.
-%\doi{https://doi.org/10.1175/MWR2904.1}
-
-%\newpage
 \nocite{Messner2016, stasinopoulos2007generalized, Seibold2016, Seibold2017, yee2010vgam, gamboostLSS2016, srtm}
@@
 
+\subsection*{Tree algorithm}
+The tree algorithm used in the applications discussed in this paper is explained in the following.
+For notational simplicity, the testing and splitting procedure is described for the root node with observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$.
+In each child node the corresponding subset of observations depends on the assignment made by the foregoing split.\\
+
+\begin{itemize}
+\item A distributional model $D(Y, \theta)$ is fit to the set of observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution parameter $\hat{\theta}$.
+This parameter can be a multidimensional vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, depending on the type of distribution.
+The estimation is done by maximizing the log-likelihood function $\ell(\theta; Y)$, i.e.
+$$
+\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
+$$
+which is equal to solving
+$$
+\sum_{i=1}^n s(\theta, y_i) = \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\theta; y_i) = 0
+$$
+for $\theta$.
+
+\item Calculate score values 
+$$
+s(\hat{\theta}, y_i) = 
+\begin{pmatrix} 
+s(\hat{\theta}, y_1) \\
+s(\hat{\theta}, y_2) \\
+\vdots \\
+s(\hat{\theta}, y_n)
+\end{pmatrix} =
+\begin{pmatrix} 
+s(\hat{\theta}, y_1)_1 & s(\hat{\theta}, y_1)_2 & \ldots & s(\hat{\theta}, y_1)_K\\
+s(\hat{\theta}, y_2)_1 & s(\hat{\theta}, y_2)_2 & \ldots & s(\hat{\theta}, y_2)_K\\
+\vdots & \vdots & \ddots & \vdots \\
+s(\hat{\theta}, y_n)_1 & s(\hat{\theta}, y_n)_2 & \ldots & s(\hat{\theta}, y_n)_K
+\end{pmatrix}
+$$
+
+\item Test for independence between each column of scores $\left(s(\hat{\theta}, y_i)_k\right)_{i=1,\ldots,n}$ for $k\in\{1,\ldots,K\}$, and split variable $Z_j \in \{Z_1, \ldots, Z_m\}$.  
+\begin{align*}
+H_0^{1,j}:  \left(s(\hat{\theta}, y_i)_1\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j \\
+H_0^{2,j}:  \left(s(\hat{\theta}, y_i)_2\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j \\
+\vdots \hspace{3cm}\\
+H_0^{K,j}:  \left(s(\hat{\theta}, y_i)_K\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j 
+\end{align*}
+
+To test these hypotheses permutation testing with the multivariate linear statistic
+$$
+T_{j} = \sum_{i=1}^n g_j(z_{ji}) \cdot s(\hat{\theta}, y_i)
+$$
+is applied.
+The type of the transformation function $g_j$ depends on the type of the split variable $Z_j$.
+If $Z_j$ is numeric then $g_j$ is simply the identity function $g_j(z_{ji}) = z_{ji}$.
+If $Z_j$ is a categorical variable with $L$ categories then $g_j(z_{ji}) = e_L(z_{ji}) = (\I(z_{ji} = 1), \ldots, \I(z_{ji} = L))$ such that $g_j$ is a unit vector where the element corresponding to the value of $z_{ji}$ is $1$.
+Observations with missing values are excluded from the sums.
+
+With the conditional expectation $\mu_{j}$ and the covariance $\Sigma_{j}$ as derived by \cite{strasser1999} the test statistic can be standardized.
+For a numeric split variable this results in the Pearson correlation coefficient
+$$
+c(t_{j},\mu_{j},\Sigma_{j}) = \left|\frac{t_{j} - \mu_{j}}{\sqrt{\Sigma_{j}}}\right|.
+$$
+Otherwise the standardized form is
+$$
+c(t_{j},\mu_{jk},\Sigma_{j}) = \max_{l=1,\ldots,L}\left|\frac{(t_{j} - \mu_{j})_l}{\sqrt{(\Sigma_{j})_{ll}}}\right|.
+$$
+The smaller the p-value corresponding to the standardized test statistic $c(t_{j},\mu_{j},\Sigma_{j})$ is the stronger the discrepancy from the assumption of independence between the scores and the split variable $Z_j$.
+
+\item After Bonferroni adjusting the p-vlaues check whether any of the resulting p-values is beneath the selected significance level.
+If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest p-value to the scores corresponding to the distribution parameters relevant for the split.
+
+\item Choose the breakpoint that leads to the highest discrepancy between score functions in the two resulting subgroups. This difference can be measured by the linear statistic
+$$
+T_{j^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\hat{\theta}, y_i)
+$$
+where $\mathcal{B}_{1r}$ is the first of the two new subgroups that are defined by splitting in split point $r$ of variable $Z_{j^{\ast}}$. The split point is then chosen as follows:
+$$
+r^{\ast} = \argmin_{r} c(t_{j^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{j^{\ast}}^r).
+$$
+
+\item Repeat the testing and splitting procedure in each of the resulting subgroups until some stopping criterion is reached.
+This criterion can for example be a minimal number of observations in a node or a minimal p-value for the statistical tests.
+
+\end{itemize}
+
+This permutation test based tree algorithm is presented in \citealt{hothorn2006unbiased} as the ctree algorithm.
+A different framework to build a tree is provided by the MOB algorithm which is based on M-fluctuation tests (\citealt{zeileis2008model}).
+
 %\begin{figure}[t!]
 %\begin{subfigure}{0.5\textwidth}
-%<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
+%<<rain_axams_pit_insample, fig=TRUE, echo=FALSE>>=
+%
 %# in sample
@@
 %par(mfrow = c(2,2))
-%set.seed(1)
-%qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
-%qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
-%qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
-%qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
-
+%pithist(pit_df_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%pithist(pit_g_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%pithist(pit_gb_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%pithist(pit_ml_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
+%
 %@
-%\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
+%\caption{\label{pit_insample} PIT histograms for station Axams (in sample)}
 %\end{figure}
 
+\subsection*{Probability integral transform (PIT) histograms}
 \begin{figure}[t!]
@@
 %\begin{subfigure}{0.8\textwidth}
-<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
+<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
 
 # out of sample
-set.seed(7)
+set.seed(4)
+
 par(mfrow = c(2,2))
-qqrplot(pit_df_t, nsim = 100, main = "distributional forest", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_g_t, nsim = 100, main = "prespecified GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_gb_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-
-qqrplot(pit_ml_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-@
-\caption{\label{qqr_oosample} Out-of-sample residual QQ-plot (2009--2012) for station Axams based on models learned on data from 1985--2008.}
+pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "distributional forest", ylim = c(0,1.5))
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM", ylim = c(0,1.5))
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM", ylim = c(0,1.5))
+pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
+
+@
+%\end{subfigure}
+\caption{\label{pit_oosample}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
 \end{center}

### 2018-02-07 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1985 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 833d641068b635820b6d55dbdff06b99d2fa9671

@@
 In regression analysis various models have been developed to describe the relationship between a response variable and a set of covariates.
-Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects.
+Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has led to the development of generalized additive models (GAM) also capturing non-linear effects.
 In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate.
@@
 
-These models consider only the expectation of the underlying distribution, however, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
+These models consider only the expectation of the underlying distribution.
+However, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
 For example in the field of precipitation forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) usually acting on the assumption of a normally distributed response variable allow for both parameters, the mean and the variance, to depend linearly on covariates.
@@
 \subsection{Application for one station}
+\label{Application for one station}
 First of all we focus on one observation station, Axams in Tyrol.
@@
 Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots.
-In the appendix the corresponding probability integral transform (PIT) histograms (\citealt{gneiting2007}) can be found.
 %%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>
@@
 
-In Figure~\ref{pit_oosample} another impression of the quality of the predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models.
+In Figure~\ref{qqr_oosample} another impression of the quality of the predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models.
 For these plots 100 randomized quantile residuals are simulated per observation.
@@
 Overall, it can be stated that all four models provide well calibrated forecasts.
+In the appendix the corresponding probability integral transform (PIT) histograms (\citealt{gneiting2007}) can be found.
 
@@
 \subsection*{Probability integral transform (PIT) histograms}
+Probability integral transform (PIT) histograms \cite{gneiting2007} are a commonly used tool to assess the predictive power of distributional forecasting methods, particularly in the field of meteorological forecasting \\
+Let $x$ be an observation with true distribution function $G$ and predicted distribution function $F$.
+Then the corresponding probability integral transform (PIT) value is $p = F(x)$.
+In the ideal case of a perfect prediction and $F$ being continuous $p$ follows a uniform distribution.
+This uniformity can be analysed applying a histogram.
+The closer the bars of a PIT histogram are to the value 1, the closer the distribution of $p$ is to a uniform distribution and therefore the better the predicted distribution models the true distribution.
+
+The PIT histograms in Figure~\ref{pit_oosample} correspond to the forecasts presented in Section \ref{Application for one station}. 
+Daily observations of total precipitation from station Axams are considered and predictions are made for the 24th of July 2009, 2010, 2011 and 2012, based on data from the previous 24 years.
+
 \begin{figure}[t!]

### 2018-02-07 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1986 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 664c83d3f02ebb28f15d4c70a6c08764667c492c
revised large parts of intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1988 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e3bfcc5ad9c4a6e7c4106b2334bf896610df77c1

@@
 \section{Introduction}
-In regression analysis various models have been developed to describe the relationship between a response variable and a set of covariates.
-Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has led to the development of generalized additive models (GAM) also capturing non-linear effects.
-In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate.
-In the left picture the expected value of the response variable is modeled by a linear function in the covariate.
-In the middle a GAM is applied where a smooth function allows for more flexibility.
-
-These models consider only the expectation of the underlying distribution.
-However, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting.
-For example in the field of precipitation forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) usually acting on the assumption of a normally distributed response variable allow for both parameters, the mean and the variance, to depend linearly on covariates.
-Based on the resulting model the probability of precipitation can be estimated as well as the expected amount of precipitation.
-Therefore, these models provide common and widely used tools, also due to their intuitive structure and interpretability.
-However, the restriction to model only linear effects has risen the interest in applying more flexible models.
-Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM as illustrated in the right plot of Figure~\ref{Devel_parmod}.
-These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g.
-areas where no abrupt changes in meteorological/climatological events appear.
-However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model.
-In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees (\citealt{breiman1984classification})
-%%%<TH> citation Loh does not fit, maybe omit citations anyway </TH>
-%(\citealt{loh2011classification}) 
-or random forests such as Breiman and Cutler's randomForest (\citealt{breiman2001random}).
-Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process.
-Therefore, no additional effort or knowledge for the specification of the model is required in advance.
-However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well.
-In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothed out in case of smooth effects while abrupt changes are still captured by the tree structure of the model.
-
-For these reasons, regression trees and particularly random forests provide valuable features which can contribute to improving models for probabilistic forecasting.
-Therefore, we are going to introduce a novel general distributional framework for these methods by presenting distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-In Figure~\ref{Devel_treeforest} the developments in the field of tree-structured models together with the novel approaches of distributional trees and forests are illustrated for the case of one covariate.
-In the top left picture the step function modeling the mean of the response variable represents the division into three terminal subgroups by a regression tree.
-This is extended in the bottom left picture where a distributional tree models not only the mean but also the variance of the response variable by its own step function.
-The top middle picture shows that combining an ensemble of trees to a random forest can lead to smoother effects while the tree structure assures that abrupt changes are captured as well.
-Again, an extension of the random forest model considering only the mean is made in the top right picture.
-Here a distributional forest models the mean and the variance capturing abrupt changes but also smooth effects.
-
-%In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
-%Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables.
-This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family.
-(In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
-%Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination.
-(In the fitting function \code{gam} types of smoothing terms can be selected.)\\
-%However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
-%(GAMLSS) (\citealt{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
-In that way a whole distribution can be specified.
-The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
-%This development in the field of parametric modeling is illustrated in Figure~\ref{Devel_parmod}.\\
+
+In regression analysis various models have been developed to describe the
+relationship between a response variable and a set of covariates. The classical
+model is the linear model (LM) where the conditional mean of the response
+is modeled through a linear function of the covariates (see the left panel of
+Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
+this has been extended in various directions including:
+\begin{itemize}
+  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1986})
+    encompassing an additional nonlinear link function for the conditional mean.
+  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
+    allowing for smooth nonlinear effects in the covariates
+    (Figure~\ref{Devel_parmod}, middle).
+  \item Generalized additive models for location, scale, and shape
+    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
+    approach. In GAMLSS, each parameter of a statistical distribution can depend
+    on an additive predictor of the covariates comprising linear and/or
+    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
+\end{itemize}
+Thus, the above-mentioned models provide a broad toolbox for capturing different
+aspects of the response (mean only vs.\ full distribution) and different types
+of dependencies on the covariates (linear vs.\ nonlinear additive terms).
 
@@
 \end{figure}
+
+While in many applications conditional mean regression models had been receiving
+the most attention, there has been a paradigm shift over the last decade towards
+distributional regression models. An important reason for this is that in many
+fields forecasts of the mean are not the only (or not even the main) concern but
+that there is an increasing interest in probabilistic forecasts. Quantities of
+interest typically include exceedence probabilities for certain thresholds of the
+response or quantiles of the response distribution. As an example, consider
+weather forecasting where there is less interest in the mean amount of
+precipitation on the next day. Instead, the probability of rain vs.\ no rain
+is typically more relevant or, in some situations, a prediction interval of
+expected precipitation (say from the expected 10\% to 90\% quantile). Similar
+considerations apply for other meteorological quantities and hence attention
+in the weather forecasting literature has been shifting from classical linear
+models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
+non-homogenous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
+The NGR typically describes the mean of some meteorological response variable through
+the average of the corresponding quantity from an ensemble of physically-based
+numerical weather predictions (NWPs). Similarly, the variance of the response
+is captured through the variance of the ensemble of NWP outputs. Thus, the NGR
+considers both the mean as well as the uncertainty of the NWP model outputs
+to obtain probabilistic forecasts calibrated to a particular site.
+%% late: introduce MOS and EMOS? more details about precipitation forecasting
 
@@
 
-%\SweaveOpts{eval = FALSE}
+For these reasons, regression trees and particularly random forests provide
+valuable features which can contribute to improving models for probabilistic
+forecasting. Therefore, we are going to introduce a novel general distributional
+framework for these methods by presenting distributional trees and forests which
+embed the idea of flexible parametric modeling into the framework of regression
+trees and random forests. In Figure~\ref{Devel_treeforest} the developments in
+the field of tree-structured models together with the novel approaches of
+distributional trees and forests are illustrated for the case of one covariate.
+In the top left picture the step function modeling the mean of the response
+variable represents the division into three terminal subgroups by a regression
+tree. This is extended in the bottom left picture where a distributional tree
+models not only the mean but also the variance of the response variable by its
+own step function. The top middle picture shows that combining an ensemble of
+trees to a random forest can lead to smoother effects while the tree structure
+assures that abrupt changes are captured as well. Again, an extension of the
+random forest model considering only the mean is made in the top right picture.
+Here a distributional forest models the mean and the variance capturing abrupt
+changes but also smooth effects.
+
+However, for example in complex
+terrain in mountainous regions various factors can change rapidly and their
+effects are difficult to be represented by a smooth parametric model. In this
+situation a different approach which can deal with these abrupt changes would be
+to apply tree-structured models such as regression trees
+(\citealt{breiman1984classification}) or random forests such as Breiman and
+Cutler's randomForest (\citealt{breiman2001random}). Moreover, these models are
+particularly advantageous in case of many covariates with unknown effects or
+interactions as the variable selection is done automatically within the model
+fitting process. Therefore, no additional effort or knowledge for the
+specification of the model is required in advance. However, even though their
+main advantage of being able to model non-additive effects and interactions
+without any necessary prespecification can play to their strength, regression
+trees are not an adequate choice if (additional) smooth effects appear as well.
+In this situation a random forest provides a good compromise as it combines a
+set of trees such that the steps imposed by the trees are smoothed out in case
+of smooth effects while abrupt changes are still captured by the tree structure
+of the model.
+
+%In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
+%Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables.
+This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family.
+(In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
+%Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination.
+(In the fitting function \code{gam} types of smoothing terms can be selected.)\\
+%However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
+%(GAMLSS) (\citealt{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
+In that way a whole distribution can be specified.
+The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
+%This development in the field of parametric modeling is illustrated in Figure~\ref{Devel_parmod}.\\
 
@@
 \item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(y_i, \theta) = 0$
+\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(y_i, \hat{\theta}) = 0$.
 \item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
@@
 %Similar observations have a high probability of ending up in the same terminal node whereas this probability is low for quite different observations. 
-%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (“bagging survival trees”, Hothorn et al. 2004). 
+%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (bagging survival trees, Hothorn et al. 2004). 
 In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 

### 2018-02-08 pkg/disttree/inst/draft/draft.Rnw
revised 2nd part of intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1989 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4e1723563d4c1e5ffbb1f3395fabdef2a3826325

@@
 \begin{itemize}
-  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1986})
+  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1983})
     encompassing an additional nonlinear link function for the conditional mean.
@@
 
-For these reasons, regression trees and particularly random forests provide
-valuable features which can contribute to improving models for probabilistic
-forecasting. Therefore, we are going to introduce a novel general distributional
-framework for these methods by presenting distributional trees and forests which
-embed the idea of flexible parametric modeling into the framework of regression
-trees and random forests. In Figure~\ref{Devel_treeforest} the developments in
-the field of tree-structured models together with the novel approaches of
-distributional trees and forests are illustrated for the case of one covariate.
-In the top left picture the step function modeling the mean of the response
-variable represents the division into three terminal subgroups by a regression
-tree. This is extended in the bottom left picture where a distributional tree
-models not only the mean but also the variance of the response variable by its
-own step function. The top middle picture shows that combining an ensemble of
-trees to a random forest can lead to smoother effects while the tree structure
-assures that abrupt changes are captured as well. Again, an extension of the
-random forest model considering only the mean is made in the top right picture.
-Here a distributional forest models the mean and the variance capturing abrupt
-changes but also smooth effects.
-
-However, for example in complex
-terrain in mountainous regions various factors can change rapidly and their
-effects are difficult to be represented by a smooth parametric model. In this
-situation a different approach which can deal with these abrupt changes would be
-to apply tree-structured models such as regression trees
-(\citealt{breiman1984classification}) or random forests such as Breiman and
-Cutler's randomForest (\citealt{breiman2001random}). Moreover, these models are
+The before discussed models with the response depending on additive linear or smooth nonlinear terms
+of the covariates are powerful tools as long as only smooth effects are to be considered.
+A different approach that can also deal with non-additive effects
+and possibly unknown interactions of covariates is to apply tree-based methods. 
+A regression tree (\citealt{breiman1984classification})
+works on the idea of separating the data into more homogeneous subgroups based on
+covariates. However, the resulting step function as illustrated in the left panel of
+Figure \ref{Devel_treeforest} can often be too rough and imposes abrupt changes on the model/data even in
+case of smooth transitions. Random forest models such as Breiman and
+Cutler's randomForest (\citealt{breiman2001random}) encounter this problem/disadvantage by combining an
+ensemble of trees such that steps are smoothed out in case of smooth effects while abrupt changes
+are still captured by its tree-based structure (see the top middel panel of Figure \ref{Devel_treeforest}).
+While classic regression trees and random forests only model the mean of the response
+we are now going to combine the idea of these tree based-methods with the progress
+made in the field of (flexible) parametric modelling. In particular, the development from mean regression
+models to distributional models will be embeded in the framework or regression trees and
+random forests providing distributional trees and forests. This development is illustrated
+in the bottom panel of Figure~\ref{Devel_treeforest} for distributional trees and in the top right panel
+for distributional forests.
+
+In many modeling situations, especially when the type of effects are not known in advance,
+it can be challenging to specify a well fitting model. The novel methodology of distributional 
+trees and forests provides a framework that covers a wide range of effects including linear and nonlinear
+additive and non-additive effects as well as interactions of covariates. Moreover, these models are
 particularly advantageous in case of many covariates with unknown effects or
@@
 fitting process. Therefore, no additional effort or knowledge for the
-specification of the model is required in advance. However, even though their
-main advantage of being able to model non-additive effects and interactions
-without any necessary prespecification can play to their strength, regression
-trees are not an adequate choice if (additional) smooth effects appear as well.
-In this situation a random forest provides a good compromise as it combines a
-set of trees such that the steps imposed by the trees are smoothed out in case
-of smooth effects while abrupt changes are still captured by the tree structure
-of the model.
+specification of the model is required in advance.
+
+For these reasons, distributional trees and forests provide
+valuable features which can contribute to improving models for probabilistic
+forecasting. For example in precipitation/weather forecasting (for complex terrain) 
+in mountainous regions various influencing factors can change rapidly and their effects are 
+difficult to be represented by a smooth parametric model. (Additionally, a wide grid over
+this complex terrain can lead to abrupt changes within just one cell.)
+
+In the application discussed in this paper precipitation forecasts are made for a mountainous region
+in the Alps covering mainly Tyrol, Austria, and adjacent areas. Predictions are based on data from 
+85 observation stations and 28 years (1985--2012). The total amount of precipitation within 24 hours
+is modeled with the assumption of the underlying distribution being a censored normal distribution 
+with censoring point 0. As covariates a large number of meteorlogical quantities are provided by 
+ensembles of numerical weather prediction models. 
+Next to distributional other methods commonly used in this research area are applied and their results are compared. 
+In a first setting only one observation station is taken into consideration. The models are trained on data from 
+24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of 
+this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 
+the density functions of the distributions estimated by a distributional forest are plotted together with the 
+corresponding observations. This illustrates that appropriate probabilistic forecasts 
+are provided by this model as they are close to the observed values. Beside the expected values the predictions 
+provided by distributional models offer a wide range of information as they specify the whole distribution and 
+therefore enable further methods of inference and analysis. In section \ref{ProbabilisticForecasting} this example 
+will be further discussed and it will be shown that the predictions made by the distributional forest are 
+reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, 
+the scale parameter, and therefore the whole distribution.
 
@@
 
-After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}.
+%After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}.
 The new method is applied on precipitation data in a mountainous region with a very large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
@@
 %%% well fitting predicted expected values -> appropriate probabilistic forecasts? </TH>
-are provided by this model as they are close to the observed values.
+%are provided by this model as they are close to the observed values.
 Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.

### 2018-02-08 pkg/disttree/inst/draft/draft.Rnw
edited reference for GLM
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1992 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d9c1a42add31205b35ba4fbb6c3d6f922ea74101

@@
     encompassing an additional nonlinear link function for the conditional mean.
-  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
+  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1989})
     allowing for smooth nonlinear effects in the covariates

### 2018-02-08 pkg/disttree/inst/draft/draft.Rnw
edited reference for GLM
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1995 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e3035a23815155ee68617f65719cc6eebbf7ef6e
small corrections in the intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1997 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 51e9aa7d193d49f77f5be8a452a8756bd1c0cf84

--- 2018-02-08 pkg/disttree/inst/draft/draft.Rnw
@@
 \begin{itemize}
-  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1983})
+  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1989})
     encompassing an additional nonlinear link function for the conditional mean.
@@
 models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
-non-homogenous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
+non-homogeneous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
 The NGR typically describes the mean of some meteorological response variable through
@@
 covariates. However, the resulting step function as illustrated in the left panel of
-Figure \ref{Devel_treeforest} can often be too rough and imposes abrupt changes on the model/data even in
+Figure \ref{Devel_treeforest} can often be too rough and impose abrupt changes on the model even in
 case of smooth transitions. Random forest models such as Breiman and
-Cutler's randomForest (\citealt{breiman2001random}) encounter this problem/disadvantage by combining an
+Cutler's randomForest (\citealt{breiman2001random}) encounter this problem by combining an
 ensemble of trees such that steps are smoothed out in case of smooth effects while abrupt changes
-are still captured by its tree-based structure (see the top middel panel of Figure \ref{Devel_treeforest}).
+are still captured by its tree-based structure (see the top middle panel of Figure \ref{Devel_treeforest}).
+
 While classic regression trees and random forests only model the mean of the response
 we are now going to combine the idea of these tree based-methods with the progress
-made in the field of (flexible) parametric modelling. In particular, the development from mean regression
-models to distributional models will be embeded in the framework or regression trees and
+made in the field of flexible parametric modelling. In particular, the development from mean regression
+models to distributional models will be embedded in the framework of regression trees and
 random forests providing distributional trees and forests. This development is illustrated
@@
 
-In many modeling situations, especially when the type of effects are not known in advance,
+In many modeling situations, especially when the types of effects are not known in advance,
 it can be challenging to specify a well fitting model. The novel methodology of distributional 
@@
 valuable features which can contribute to improving models for probabilistic
-forecasting. For example in precipitation/weather forecasting (for complex terrain) 
-in mountainous regions various influencing factors can change rapidly and their effects are 
-difficult to be represented by a smooth parametric model. (Additionally, a wide grid over
-this complex terrain can lead to abrupt changes within just one cell.)
+forecasting. For example weather forecasting in mountainous regions various influencing factors 
+can change rapidly and their effects are  difficult to be represented by a smooth parametric model. 
+Additionally, large spacing of the grid applied in numerical weather prediction models over 
+complex terrain can lead to abrupt changes within just one cell.
 
 In the application discussed in this paper precipitation forecasts are made for a mountainous region
-in the Alps covering mainly Tyrol, Austria, and adjacent areas. Predictions are based on data from 
+in the Alps covering mainly Tyrol in Austria and adjacent areas. Predictions are based on data from 
 85 observation stations and 28 years (1985--2012). The total amount of precipitation within 24 hours
 is modeled with the assumption of the underlying distribution being a censored normal distribution 
-with censoring point 0. As covariates a large number of meteorlogical quantities are provided by 
+with censoring point 0. As covariates a large number of meteorological quantities are provided by 
 ensembles of numerical weather prediction models. 
-Next to distributional other methods commonly used in this research area are applied and their results are compared. 
-In a first setting only one observation station is taken into consideration. The models are trained on data from 
-24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of 
-this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 
-the density functions of the distributions estimated by a distributional forest are plotted together with the 
-corresponding observations. This illustrates that appropriate probabilistic forecasts 
-are provided by this model as they are close to the observed values. Beside the expected values the predictions 
-provided by distributional models offer a wide range of information as they specify the whole distribution and 
-therefore enable further methods of inference and analysis. In section \ref{ProbabilisticForecasting} this example 
-will be further discussed and it will be shown that the predictions made by the distributional forest are 
-reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, 
-the scale parameter, and therefore the whole distribution.
+Next to distributional forests other methods commonly used in this research area are applied and their 
+results are compared. In a first setting only one observation station is taken into consideration.
+The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years 
+are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. 
+For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated 
+by a distributional forest are plotted together with the corresponding observations. This illustrates 
+that appropriate probabilistic forecasts are provided by this model as they are close to the observed 
+values. Beside the expected values the predictions made by distributional models offer a wide range 
+of information as they specify the whole distribution and therefore enable further methods of inference 
+and analysis. In section \ref{ProbabilisticForecasting} this application will be further discussed and it 
+will be shown that the predictions provided by the distributional forest are reasonably well fitting, not 
+only for the mean parameter of the distribution but also for its second parameter, the scale parameter, 
+and therefore the whole distribution.
 
@@
 \caption{\label{boxplot_crps_all}CRPS skills score for each station (grey lines) and aggregated over all stations (boxplots).
-Station Axams is highlighted in greend and the horizontal orange line pertains to the reference model EMOS.
+Station Axams is highlighted in green and the horizontal orange line pertains to the reference model EMOS.
 The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
@@
 
-\item After Bonferroni adjusting the p-vlaues check whether any of the resulting p-values is beneath the selected significance level.
+\item After Bonferroni adjusting the p-values check whether any of the resulting p-values is beneath the selected significance level.
 If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest p-value to the scores corresponding to the distribution parameters relevant for the split.

### 2018-02-09 pkg/disttree/inst/Psychoco2018/slides.Rnw
slides for Psychoco2018
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1998 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1ffa37e05550d5c003b55bbc095743e245168dd3

--- 2017-11-27 pkg/disttree/inst/CMStatistics2017/slides.Rnw
@@
 
-\title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Achim Zeileis}
+\title{Distributional Regression Forests for Probabilistic Modeling and Forecasting}
+\author{Achim Zeileis, Torsten Hothorn,\\ Lisa Schlosser, Heidi Seibold}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
@@
 \hspace{0.5cm} Specify the complete probability distribution\\
-\hspace{0.5cm} (including location, scale and shape).
+\hspace{0.5cm} (including location, scale, and shape).
 
@@
 \end{frame}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Global likelihood estimation}
+\begin{itemize}
+\item Specified log-likelihood function $$\ell(\theta; y_i)$$ with distribution parameter $\theta$ and observation $y_i$ of a learning data set $\{y_i\}_{i = 1,\ldots,n}$.
+
+\bigskip
+
+\item Estimation of $\hat{\bold{\theta}}$ via maximum likelihood 
+$$
+\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
+$$
+equivalent to solving 
+$$
+\sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
+$$
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Adaptive local likelihood estimation}
+\medskip
+{\bf Tree:}\\
+\medskip
+For a set of covariates $\bold{x}$, a learning data set $\{(y_i, \bold{x}_i)\}_{i =1,\ldots,n}$ and subgroups $\mathcal{B}_1, \ldots, \mathcal{B}_B$ calculate weights
+$$
+w^{\text{tree}}_i(\bold{x}) = \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b))
+$$
+which are then included in the estimation of $\hat{\theta}$
+$$
+\hat{\theta}_{\text{tree}}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{x}) \cdot \ell(\theta; y_i).
+$$
+\end{frame}
+
+\begin{frame}
+\frametitle{Adaptive local likelihood estimation}
+\medskip
+{\bf Forest:}\\
+\medskip
+For a set of covariates $\bold{x}$, a learning data set $\{(y_i, \bold{x}_i)\}_{i =1,\ldots,n}$, $T$~trees and corresponding subgroups $\mathcal{B}_1^t, \ldots, \mathcal{B}_B^t$ calculate weights
+$$
+w^{\text{forest}}_i(\bold{x}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
+$$
+which are then included in the estimation of $\hat{\theta}$
+$$
+\hat{\theta}_{\text{forest}}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{x}) \cdot \ell(\theta; y_i).
+$$
+\end{frame}
+
+\begin{frame}
+\frametitle{Adaptive local likelihood estimation}
+{\bf Base model:}\\
+\vspace{-0.5cm}
+$$
+\hat{\theta}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
+$$
+
+{\bf Tree:}\\
+\vspace{-0.5cm}
+$$
+w^{\text{tree}}_i(\bold{x}) = \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b))
+$$
+$$
+\hat{\theta}_{\text{tree}}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{x}) \cdot \ell(\theta; y_i).
+$$
+
+{\bf Forest:}\\
+\vspace{-0.5cm}
+$$
+w^{\text{forest}}_i(\bold{x}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
+$$
+$$
+\hat{\theta}_{\text{forest}}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{x}) \cdot \ell(\theta; y_i)
+$$
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{frame}
@@
 \begin{frame}
+\frametitle{Choosing a distribution}
+
+\medskip
+{\bf Based on:}
+\begin{itemize}
+\item Recommendations by subject matter literature.
+\item Experiences from other modelling situations.
+\item Common simplifications/generalizations/assumptions.
+\end{itemize}
+
+\bigskip
+{\bf Examples:}
+\begin{itemize}
+\item Survival times: Weibull distribution, exponential distribution.
+\item Wages, precipitation: censored normal distribution.
+\item Temperature: normal distribution.
+\item Count of events/accidents (insurance): Poisson distribution.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Alternative: Transformation trees and forests}
+
+\medskip
+
+{\bf General framework / Conditional model:}
+$$
+\mathbb{P}(Y\leq y|X=x) = \mathbb{P}_{Y|X=x} = \mathbb{P}_{Y, \theta(x)} 
+$$
+
+\bigskip
+
+{\bf Transformation model:}
+$$
+\mathbb{P}(Y\leq y|X=x) = \Phi(\bold{a}_{Bs,d}(y)^T \theta(x)) 
+$$
+
+\begin{itemize} 
+\item $\bold{a}_{Bs,d}(y)^T \theta(x)$ is a smooth, monotone Bernstein polynomial of degree $d$.
+\item $d=1$ means $\mathbb{P}(Y\leq y|X=x) = \mathcal{N}(\mu(x),\sigma^2(x))$.
+\item $d=5$ is surprisingly flexible.
+\end{itemize}
+%All ``classical'' distributions:  Distributional forests
+\end{frame}
+
+\begin{frame}
 \frametitle{Software}

### 2018-02-09 pkg/disttree/inst/Psychoco2018/slides169.Rnw
slides in 16:9 format
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1999 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4b21fdf8bf79b5d028d605e0b989556cda932f5e

--- 2018-02-09 pkg/disttree/inst/Psychoco2018/slides.Rnw
@@
-\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
+\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
 \usetheme[nototalframenumber, noslidenumber]{uibk}
@@
 \title{Distributional Regression Forests for Probabilistic Modeling and Forecasting}
-\author{Achim Zeileis, Torsten Hothorn,\\ Lisa Schlosser, Heidi Seibold}
+\author{Achim Zeileis, Torsten Hothorn, Lisa Schlosser, Heidi Seibold}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
@@
 \begin{figure}[!htb]
-\minipage{0.29\textwidth}
+\minipage{0.2\textwidth}
 \begin{center}
@@
 \endminipage
+\hspace{0.1cm}
 \visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\hspace{0.1cm}
+\minipage{0.2\textwidth}
 \begin{center}
@@
 \endminipage
+\hspace{0.1cm}
 \visible<4->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\hspace{0.1cm}
+\minipage{0.2\textwidth}
 \begin{center}
@@
 \vspace{0.5cm}
-\minipage{0.25\textwidth}
+\minipage{0.2\textwidth}
 \begin{center}
@@
 \endminipage
-\hspace{1.1cm}
-\minipage{0.25\textwidth}
+\hspace{1.03cm}
+\minipage{0.2\textwidth}
 \begin{center}
@@
 \endminipage
+\hspace{0.1cm}
 \visible<2->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\hspace{0.1cm}
+\minipage{0.2\textwidth}
  \begin{center}
@@
 \endminipage
+\hspace{0.1cm}
 \visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\hspace{0.1cm}
+\minipage{0.2\textwidth}
   \visible<3->{
@@
   \begin{tikzpicture}
-  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
-  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
+  \draw[line width=0.2mm, black] (0,0) rectangle (2.25,2);
+  \node[rectangle, align=center] (sq) at (1.12, 1) {\Huge{?}};
   \end{tikzpicture}
@@
 
-\minipage{0.29\textwidth}
+\minipage{0.23\textwidth}
 \begin{center}
@@
 \endminipage
-\hspace{0.65cm}
-\minipage{0.29\textwidth}
+\hspace{0.58cm}
+\minipage{0.23\textwidth}
 \begin{center}
@@
 }
-\vspace{0.5cm}\\
+\vspace{0.44cm}\\
 \code{randomForest}\\
@@
 \endminipage
-\hspace{0.65cm}
-\minipage{0.28\textwidth}
-\begin{center}
-\vspace{0.68cm}
+\hspace{0.58cm}
+\minipage{0.23\textwidth}
+\begin{center}
+\vspace{0.5cm}
 \visible<3->{
 Distributional trees and forests\\
-\vspace{1.7cm}
+\vspace{1.73cm}
 \code{disttree}\\
-based on \code{partykit}\\
+based on \\
+\code{partykit}\\
 \vspace{1cm}}
@@
 \hspace{0.5cm} Specify the complete probability distribution\\
-\hspace{0.5cm} (including location, scale, and shape).
+\hspace{0.5cm} (including location, scale and shape).
 
@@
 \end{frame}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{frame}
 \frametitle{Fitting a Tree}
-\vspace*{-0.12cm}
+\vspace*{-0.8cm}
 \begin{center} 
@@
 
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+\setkeys{Gin}{width=0.55\linewidth, height=0.55\linewidth}
 <<plottree_dgp, fig=TRUE, echo=FALSE>>=
@@
 \begin{center}
-\vspace*{-0.12cm}
+\vspace*{-0.8cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+\setkeys{Gin}{width=0.55\linewidth, height=0.55\linewidth}
 <<plottree_estpar, fig=TRUE, echo=FALSE>>=
@@
 \begin{center}
-\vspace*{-0.12cm}
+\vspace*{-0.8cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+\setkeys{Gin}{width=0.55\linewidth, height=0.55\linewidth}
 <<plottree_box, fig=TRUE, echo=FALSE>>=
@@
 \begin{center}
-\vspace*{-0.12cm}
+\vspace*{-0.8cm}
 Model: \code{disttree(y ~ x)}\\
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+\setkeys{Gin}{width=0.55\linewidth, height=0.55\linewidth}
 <<plottree_dens, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1.03cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<plottree_xyplot, fig=TRUE, echo=FALSE>>=
@@
   col = pallight["forest"], border = "transparent")
-legend(x = -0.0715, y = 21.62, expression(mu  %+-%  sigma), bty = "n")
+legend(x = -0.0715, y = 21.65, expression(mu  %+-%  sigma), bty = "n")
 #lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<oneforest_exp1, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<oneforest_exp1_dt, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<oneforest_exp1_g, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<oneforest_exp1_df, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<oneforest_exp1_all, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<oneforest_exp1_allvar, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<plotsim_ll, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
+\setkeys{Gin}{width=0.6\linewidth}
 <<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=

### 2018-02-09 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2001 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 884e22091c178fa4bc9af79a241b6f4d07efda47

@@
 In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting.
-In generalized additive models for location, scale and shape %(GAMLSS, \citealt{stasinopoulos2005})
+In generalized additive models for location, scale, and shape %(GAMLSS, \citealt{stasinopoulos2005})
 each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects.
@@
 However, the restriction to model only linear effects requires further extensions which allow for more flexibility.
-Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
+Generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
 A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
@@
 Therefore, our goal is to embed the progress made in the field of probabilistic modeling in the idea of algorithmic tree and forest models.
-In particular, more flexible models such as GAMLSS (\citealt{stasinopoulos2005}) should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc.
+In particular, more flexible models such as GAMLSS (\citealt{stasinopoulos2005}) should be fitted in the nodes of a tree in order to capture location, scale, and shape as well as censoring, tail behavior etc.
 while non-additive effects of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree.
@@
 to obtain probabilistic forecasts calibrated to a particular site.
+
 %% late: introduce MOS and EMOS? more details about precipitation forecasting
+%In ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) this 
+%idea is taken one step further. Usually acting on the assumption of a normally distributed
+%response variable this model allows for both parameters, the mean and the variance, to depend 
+%linearly on covariates provided by NWPs. However, the restriction to model only linear effects 
+%has risen the interest in applying more flexible models such as offered by the GAMLSS framework.
 
@@
 
-The before discussed models with the response depending on additive linear or smooth nonlinear terms
-of the covariates are powerful tools as long as only smooth effects are to be considered.
-A different approach that can also deal with non-additive effects
-and possibly unknown interactions of covariates is to apply tree-based methods. 
-A regression tree (\citealt{breiman1984classification})
-works on the idea of separating the data into more homogeneous subgroups based on
-covariates. However, the resulting step function as illustrated in the left panel of
-Figure \ref{Devel_treeforest} can often be too rough and impose abrupt changes on the model even in
-case of smooth transitions. Random forest models such as Breiman and
-Cutler's randomForest (\citealt{breiman2001random}) encounter this problem by combining an
-ensemble of trees such that steps are smoothed out in case of smooth effects while abrupt changes
-are still captured by its tree-based structure (see the top middle panel of Figure \ref{Devel_treeforest}).
-
-While classic regression trees and random forests only model the mean of the response
+The so far discussed parametric models with the response depending on additive linear or 
+smooth nonlinear terms of the covariates are powerful tools as long as only smooth effects 
+are to be considered. A different approach that can also deal with non-additive effects and 
+possibly unknown interactions of covariates is to apply tree-based methods. 
+A regression tree (\citealt{breiman1984classification}) works on the idea of separating 
+the data into more homogeneous subgroups based on covariates. However, the resulting step 
+function, as illustrated in the left panel of Figure \ref{Devel_treeforest}, can often be too 
+rough and impose abrupt changes on the model even in case of smooth transitions. Random 
+forest models such as Breiman and Cutler's randomForest (\citealt{breiman2001random}) 
+solve this problem by combining an ensemble of trees such that steps are smoothed 
+out in case of smooth effects while abrupt changes are still captured by its tree-based 
+structure (see the top middle panel of Figure \ref{Devel_treeforest}).
+
+While classical regression trees and random forests only model the mean of the response
 we are now going to combine the idea of these tree based-methods with the progress
-made in the field of flexible parametric modelling. In particular, the development from mean regression
+made in the field of flexible parametric modeling. In particular, the development from mean regression
 models to distributional models will be embedded in the framework of regression trees and
@@
 specification of the model is required in advance.
-
 For these reasons, distributional trees and forests provide
 valuable features which can contribute to improving models for probabilistic
-forecasting. For example weather forecasting in mountainous regions various influencing factors 
+forecasting. For example in weather forecasting in mountainous regions various influencing factors 
 can change rapidly and their effects are  difficult to be represented by a smooth parametric model. 
 Additionally, large spacing of the grid applied in numerical weather prediction models over 
-complex terrain can lead to abrupt changes within just one cell.
+complex terrain can lead to abrupt changes within just one cell of the grid.
 
@@
 with censoring point 0. As covariates a large number of meteorological quantities are provided by 
-ensembles of numerical weather prediction models. 
+ensembles of NWPs. 
 Next to distributional forests other methods commonly used in this research area are applied and their 
@@
 (In the fitting function \code{gam} types of smoothing terms can be selected.)\\
-%However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
+%However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale,and shape
 %(GAMLSS) (\citealt{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
@@
 %Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests.
-In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc.
+In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale,and shape as well as censoring, tail behavior etc.
 while non-additive effects or interactions of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree.
@@
 
-In that way, a whole distribution is specified with all its features, including location, scale and shape.
+In that way, a whole distribution is specified with all its features, including location, scale, and shape.
 However, fitting one global model to the whole data set might be too much of a generalization and therefore not represent the data and its features in a reasonable way.

### 2018-02-09 pkg/disttree/inst/Psychoco2018/slides.Rnw
revisions: flow
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2002 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: aefa79f5043688378d3cd846b6a686e55bfe60d3

--- 2017-11-27 pkg/disttree/inst/CMStatistics2017/slides.Rnw
@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber, noslidenumber]{uibk}
-
-\title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Achim Zeileis}
+\usetheme[nototalframenumber, noslidenumber, url]{uibk}
+
+\title{Distributional Regression Forests for Probabilistic Modeling and Forecasting}
+\author{Lisa Schlosser, Torsten Hothorn, Heidi Seibold, Achim Zeileis}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
-\URL{https://R-Forge.R-project.org/projects/partykit/}
-\headerimage{3}
+\URL{https://eeecon.uibk.ac.at/~zeileis/}
+\headerimage{4}
 
@@
 \begin{figure}[!htb]
-\minipage{0.29\textwidth}
+\minipage{0.28\textwidth}
 \begin{center}
@@
 \visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.28\textwidth}
 \begin{center}
@@
 \visible<4->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.28\textwidth}
 \begin{center}
@@
 \visible<2->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.28\textwidth}
  \begin{center}
@@
 \visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.28\textwidth}
   \visible<3->{
@@
 
-\minipage{0.29\textwidth}
+\minipage{0.28\textwidth}
 \begin{center}
@@
 \hspace{0.65cm}
-\minipage{0.29\textwidth}
+\minipage{0.28\textwidth}
 \begin{center}
@@
 
-\subsection{Building Distributional Trees and Forests}
-\begin{frame}
-\frametitle{Building Distributional Trees and Forests}
-{\bf Tree:}
-\begin{enumerate}
-\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\hat{\theta}$ via maximum likelihood.
-\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
-%\item Assess whether the \emph{model scores} are associated with
-%    (or change along) any of the available covariates -- e.g.,
-%    using parameter instability tests (\emph{strucchange}) or
-%    conditional inference (\emph{coin}).
-\item Split the sample along the partitioning variable with the strongest association or instability.
-    Choose breakpoint with highest improvement in log-likelihood.
-\item Repeat steps 2--4 recursively in the subgroups
-    until some stopping criterion is met.
-    %-- e.g., for significance or sample size.
-%\item Choose the variable with the strongest association
-%\item Choose the split point which leads to the highest improvement of the model
-%\item Split and repeat 2-6 in each node until a stopping criterion is met
-\end{enumerate}
-
-\vspace{0.4cm}
-{\bf Forest:} Ensemble of trees.
-\begin{itemize}
-\item Bootstrap or subsamples.
-\item Random input variable sampling.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Prediction}
-%\vspace{1cm}
-{\bf Tree:}\\
-\begin{enumerate}
-\item Select the subsample of the learning data which ends up in the same terminal node as the new observation.
-\item Estimate $\hat{\theta}$ on this subsample.
-\end{enumerate}
-\vspace{0.5cm}
-
-{\bf Forest:}\\
-\begin{enumerate}
-\item For each learning observation calculate a weight by counting the number of trees in which it ends up in the same terminal node as the new observation.
-\item Estimate $\hat{\theta}$ on the weighted learning data.
-\end{enumerate}
-% but weighted by similarities of the learning data and the new observation over all trees
-%\item Evaluate final node for new observation in each of the trees and investigate similarities with each of the learning observations
-%\item Fit a distributional model for the new observation with different influences of the learning observations defined by weights
-\end{frame}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{frame}
@@
 
+\begin{frame}
+\frametitle{Global likelihood estimation}
+\begin{itemize}
+\item Specified log-likelihood function $$\ell(\theta; y_i)$$ with distribution parameter $\theta$ and observation $y_i$ of a learning data set $\{y_i\}_{i = 1,\ldots,n}$.
+
+\bigskip
+
+\item Estimation of $\hat{\bold{\theta}}$ via maximum likelihood 
+$$
+\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i).
+$$
+%% equivalent to solving 
+%% $$
+%% \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
+%% $$
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Adaptive local likelihood estimation}
+
+$$
+\hat{\theta}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i).
+$$
+
+Needed: Weights.
+
+Via tree: 0/1.
+
+Via forest: averaged across trees.
+
+\end{frame}
+
+\subsection{Building Distributional Trees and Forests}
+\begin{frame}
+\frametitle{Building Distributional Trees and Forests}
+{\bf Tree:}
+\begin{enumerate}
+\item Estimate $\hat{\theta}$ via maximum likelihood.
+\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
+\item Split the sample along the partitioning variable with the strongest association or instability.
+    Choose breakpoint with highest improvement in log-likelihood.
+\item Repeat steps 1--3 recursively until some stopping criterion is met, yielding
+    $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
+\end{enumerate}
+
+\bigskip
+
+{\bf Forest:} Ensemble of $T$ trees.
+\begin{itemize}
+\item Bootstrap or subsamples.
+\item Random input variable sampling.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Adaptive local likelihood estimation}
+
+{\bf Estimator:}
+$$
+\hat{\theta}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i)
+$$
+
+\bigskip
+
+\textbf{Weights:}
+
+\begin{eqnarray*}
+w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.3cm]
+w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.3cm]
+w^{\text{forest}}_i(\bold{x}) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
+\end{eqnarray*}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Simulation}
+
+Overview
+
+\end{frame}
+
 %\SweaveOpts{eval=FALSE}
@@
 \begin{frame}
+\frametitle{Choosing a distribution}
+
+\medskip
+{\bf Based on:}
+\begin{itemize}
+\item Recommendations by subject matter literature.
+\item Experiences from other modelling situations.
+\item Common simplifications/generalizations/assumptions.
+\end{itemize}
+
+\bigskip
+{\bf Examples:}
+\begin{itemize}
+\item Survival times: Weibull distribution, exponential distribution.
+\item Wages, precipitation: censored normal distribution.
+\item Temperature: normal distribution.
+\item Count of events/accidents (insurance): Poisson distribution.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Alternative: Transformation model}
+
+\medskip
+
+{\bf Distribution function:}
+$$
+F(y) = \Phi(\bold{a}_{Bs,d}(y)^\top \theta) 
+$$
+
+\begin{itemize} 
+\item $\bold{a}_{Bs,d}(y)^\top \theta$ is a smooth, monotone Bernstein polynomial of degree $d$.
+\item $d=1$ corresponds to $\mathcal{N}(\mu,\sigma^2)$.
+\item $d=5$ is surprisingly flexible.
+\end{itemize}
+
+\medskip
+
+\textbf{Example:} Body Mass Index explained by lifestyle factors (Switzerland).
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Alternative: Transformation model}
+
+\vspace*{0.4cm}
+
+\includegraphics[width=\textwidth]{cmpx-tree.pdf}
+
+\end{frame}
+
+\begin{frame}
 \frametitle{Software}
@@
 
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
-plot_rmse(simlist, type = "par")
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=
-plot_rmse(simlist, type = "exp")
-@
-\end{center}
-\end{frame}
+%% \begin{frame}[fragile]
+%% \frametitle{Simulation}
+%% \begin{center}
+%% \vspace*{-1cm}
+%% \setkeys{Gin}{width=0.8\linewidth}
+%% <<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
+%% plot_rmse(simlist, type = "par")
+%% @
+%% \end{center}
+%% \end{frame}
+
+%% \begin{frame}[fragile]
+%% \frametitle{Simulation}
+%% \begin{center}
+%% \vspace*{-1cm}
+%% \setkeys{Gin}{width=0.8\linewidth}
+%% <<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=
+%% plot_rmse(simlist, type = "exp")
+%% @
+%% \end{center}
+%% \end{frame}
 

### 2018-02-09 pkg/disttree/inst/draft/draft.Rnw
trigger rebuild
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2003 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e44297efc7abc249892430749c53aa42251febb0

@@
 
-In regression analysis various models have been developed to describe the
+In regression analysis a wide range of models has been developed to describe the
 relationship between a response variable and a set of covariates. The classical
@@
 \begin{itemize}
-  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1989})
+  \item \emph{Generalized linear models} (GLMs, \citealp{Nelder+Wedderburn:1972})
     encompassing an additional nonlinear link function for the conditional mean.
-  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
+  \item \emph{Generalized additive models} (GAMs, \citealp{Hastie+Tibshirani:1986})
     allowing for smooth nonlinear effects in the covariates
     (Figure~\ref{Devel_parmod}, middle).
-  \item Generalized additive models for location, scale, and shape
+  \item \emph{Generalized additive models for location, scale, and shape}
     (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
@@
 
-While in many applications conditional mean regression models had been receiving
+While in many applications conditional mean regression models have been receiving
 the most attention, there has been a paradigm shift over the last decade towards
@@
 fields forecasts of the mean are not the only (or not even the main) concern but
-that there is an increasing interest in probabilistic forecasts. Quantities of
+instead there is an increasing interest in probabilistic forecasts. Quantities of
 interest typically include exceedence probabilities for certain thresholds of the
@@
 
-The so far discussed parametric models with the response depending on additive linear or 
-smooth nonlinear terms of the covariates are powerful tools as long as only smooth effects 
-are to be considered. A different approach that can also deal with non-additive effects and 
-possibly unknown interactions of covariates is to apply tree-based methods. 
-A regression tree (\citealt{breiman1984classification}) works on the idea of separating 
-the data into more homogeneous subgroups based on covariates. However, the resulting step 
-function, as illustrated in the left panel of Figure \ref{Devel_treeforest}, can often be too 
-rough and impose abrupt changes on the model even in case of smooth transitions. Random 
-forest models such as Breiman and Cutler's randomForest (\citealt{breiman2001random}) 
-solve this problem by combining an ensemble of trees such that steps are smoothed 
-out in case of smooth effects while abrupt changes are still captured by its tree-based 
-structure (see the top middle panel of Figure \ref{Devel_treeforest}).
-
+In summary, the models discussed so far provide a broad and powerful toolset for
+parametric distributional fits depending on a specified set of additive linear or
+smooth nonlinear terms. A rather different approach to capturing the dependence on
+covariates are tree-based models.
+\begin{itemize}
+  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
+    split the data into more homogeneous subgroups and can thus capture abrupt shifts
+    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
+    Furthermore, trees automatically carry out a forward selection of covariates and
+    their interactions. 
+  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
+    an ensemble of trees fitted to resampled versions of the learning data. This
+    stabilizes the recursive partitions from individual trees and hence better
+    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
+\end{itemize}
 While classical regression trees and random forests only model the mean of the response
-we are now going to combine the idea of these tree based-methods with the progress
-made in the field of flexible parametric modeling. In particular, the development from mean regression
-models to distributional models will be embedded in the framework of regression trees and
-random forests providing distributional trees and forests. This development is illustrated
-in the bottom panel of Figure~\ref{Devel_treeforest} for distributional trees and in the top right panel
-for distributional forests.
+we propose to follow the ideas from (semi-)parametric regression modeling -- as outlined
+in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
+yielding two novel techniques:
+\begin{itemize}
+  \item \emph{Distributional trees} split the data into more homogenous groups with
+    respect to a parametric distribution, thus capturing changes in any distribution
+    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
+  \item \emph{Distributional forests} utilize an ensemble of distributional trees
+    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
+    top right).
+\end{itemize}
 

### 2018-02-10 pkg/disttree/inst/draft/draft.Rnw
finished revising intro (for now)
various clean-ups (removing commented LaTeX and R code)
graphics and label enhancements



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2005 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9888e50ba8f7c219354e60299b6488da4bd3de7c

@@
 
-## define distribution list:
-# dist_list_normal
-{
-
-  dist_list_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {
-
-    val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
-    if(!log) val <- exp(val)
-
-    # par <- c(eta[1], exp(eta[2]))
-    # val <- dnorm(y, mean = par[1], sd = par[2], log = log)
-
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE) {
-
-    score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
-                   -1 + exp(-2*eta[2] + log((y-eta[1])^2)))
-
-    # par <- c(eta[1], exp(eta[2])) 
-    # score <- cbind(1/par[2]^2 * (y-par[1]), 
-    #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))
-
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
-    d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-eta[1]) * exp(-2*eta[2]), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    d2ld.etasigma2 <- sum(weights * (-2)*exp(log((y-eta[1])^2) - 2*eta[2]), na.rm = TRUE)
-
-    # par <- c(eta[1], exp(eta[2]))                           
-    # d2ld.etamu2 <- sum(weights * rep.int(-1/par[2]^2, ny))
-    # d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-par[1])/par[2]^2), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    # d2ld.etasigma2 <- sum(weights * (-2)*(y-par[1])^2/par[2]^2, na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etamu.d.etasigma, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- pnorm
-  qdist <- qnorm
-  rdist <- rnorm
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    if(is.null(weights)) {
-      mu <- mean(y)
-      sigma <- sqrt(1/length(y) * sum((y - mu)^2))
-    } else {
-      mu <- weighted.mean(y, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (y - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- TRUE
-
-  dist_list_normal <- list(family.name = "Normal Distribution",
-                           ddist = ddist, 
-                           sdist = sdist, 
-                           hdist = hdist,
-                           pdist = pdist,
-                           qdist = qdist,
-                           rdist = rdist,
-                           link = link, 
-                           linkfun = linkfun, 
-                           linkinv = linkinv, 
-                           linkinvdr = linkinvdr,
-                           startfun = startfun,
-                           mle = mle,
-                           gamlssobj = FALSE,
-                           censored = FALSE
-  )
-}
-
 # dist_list_cens_normal
@@
   Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
+  Universit\"at Innsbruck \\
   Department of Statistics \\
   Faculty of Economics and Statistics \\
-  Universit\"at Innsbruck \\
   Universit\"atsstr.~15 \\
   6020 Innsbruck, Austria \\
-  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/}, \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\
+  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
+  \email{Achim.Zeileis@R-project.org} \\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
+  \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\
 
   Torsten Hothorn\\
+  Universit\"at Z\"urich \\
   Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
-  Universit\"at Z\"urich \\
   Hirschengraben 84\\
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 <<motivation_GLM, echo=FALSE, results=hide>>=
@@
 @
-\end{center}
 \endminipage};
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 <<motivation_GAM, echo=FALSE, results=hide>>=
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 <<motivation_GAMLSS, echo=FALSE, results=hide>>=
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 \vspace{0.0cm}
 LM, GLM  
-\end{center}
 \endminipage};
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 \vspace{0.0cm}
 GAM 
-\end{center}
 \endminipage};
@@
 \minipage{0.28\textwidth}
-\begin{center}
+\centering
 \vspace{0.0cm}
 GAMLSS 
-\end{center}
 \endminipage};
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 <<motivation_regtree, echo=FALSE, results=hide>>=
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 <<motivation_randforest, echo=FALSE, results=hide>>=
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 <<motivation_distforest, echo=FALSE, results=hide>>=
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 \vspace{0.0cm}
 Regression Tree 
-\end{center}
 \endminipage};
@@
 \minipage{0.29\textwidth}
-\begin{center}
+\centering
 \vspace{0.0cm}
 Random Forest 
-\end{center}
 \endminipage};
@@
 \minipage{0.28\textwidth}
-\begin{center}
+\centering
 \vspace{0.0cm}
 Distributional Forest 
-\end{center}
 \endminipage};
@@
 %\hspace{0.0cm}
-\begin{center}
+\centering
 <<motivation_disttree, echo=FALSE, results=hide>>=
@@
 \vspace{0.2cm}
-\end{center}
 \endminipage};
@@
 \end{tikzpicture}
-\caption{\label{Devel_treeforest}Tree and forest developments. Regression tree (top left), distributional tree (bottom), random forest (top middle), distributional forest (top right).}
+\caption{\label{Devel_treeforest}Tree and forest developments.
+Regression tree (top left), distributional tree (bottom left), random forest
+(top middle), and distributional forest (top right).}
 \end{figure}
@@
 While classical regression trees and random forests only model the mean of the response
-we propose to follow the ideas from (semi-)parametric regression modeling -- as outlined
+we propose to follow the ideas from GAMLSS modeling -- as outlined
 in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
@@
 \end{itemize}
-
-In many modeling situations, especially when the types of effects are not known in advance,
-it can be challenging to specify a well fitting model. The novel methodology of distributional 
-trees and forests provides a framework that covers a wide range of effects including linear and nonlinear
-additive and non-additive effects as well as interactions of covariates. Moreover, these models are
-particularly advantageous in case of many covariates with unknown effects or
-interactions as the variable selection is done automatically within the model
-fitting process. Therefore, no additional effort or knowledge for the
-specification of the model is required in advance.
-For these reasons, distributional trees and forests provide
-valuable features which can contribute to improving models for probabilistic
-forecasting. For example in weather forecasting in mountainous regions various influencing factors 
-can change rapidly and their effects are  difficult to be represented by a smooth parametric model. 
-Additionally, large spacing of the grid applied in numerical weather prediction models over 
-complex terrain can lead to abrupt changes within just one cell of the grid.
-
-In the application discussed in this paper precipitation forecasts are made for a mountainous region
-in the Alps covering mainly Tyrol in Austria and adjacent areas. Predictions are based on data from 
-85 observation stations and 28 years (1985--2012). The total amount of precipitation within 24 hours
-is modeled with the assumption of the underlying distribution being a censored normal distribution 
-with censoring point 0. As covariates a large number of meteorological quantities are provided by 
-ensembles of NWPs. 
-Next to distributional forests other methods commonly used in this research area are applied and their 
-results are compared. In a first setting only one observation station is taken into consideration.
-The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years 
-are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. 
-For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated 
-by a distributional forest are plotted together with the corresponding observations. This illustrates 
-that appropriate probabilistic forecasts are provided by this model as they are close to the observed 
-values. Beside the expected values the predictions made by distributional models offer a wide range 
-of information as they specify the whole distribution and therefore enable further methods of inference 
-and analysis. In section \ref{ProbabilisticForecasting} this application will be further discussed and it 
-will be shown that the predictions provided by the distributional forest are reasonably well fitting, not 
-only for the mean parameter of the distribution but also for its second parameter, the scale parameter, 
-and therefore the whole distribution.
-
-%In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
-%Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables.
-This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family.
-(In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
-%Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination.
-(In the fitting function \code{gam} types of smoothing terms can be selected.)\\
-%However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale,and shape
-%(GAMLSS) (\citealt{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM.
-In that way a whole distribution can be specified.
-The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
-%This development in the field of parametric modeling is illustrated in Figure~\ref{Devel_parmod}.\\
-
-%All these parametric models work very well as long as only additive effects appear in the considered data.
-A different type of model that can capture non-additive effects as well are regression trees.
-Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node.
-One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure~\ref{Devel_treeforest}.
-One way how to tackle this problem is to not just consider one tree but an ensemble of trees.
-This is the idea of random forests.
-A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set.
-Then the results of all trees are combined to get a final model which unlike a tree model can also deal with smooth effects in a reasonable way.\\
-%In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
-%\\
-%In the past the combination of parametric data models and algorithmic tree models has gained more and more interest since it allows for a more flexible and clear structure. \\
-%Algorithms that only fit constants to each final node and therefore only model the mean of the response variable (such as CART, Breiman 1984) might lead to very large trees which complicates analyzing and interpreting.
-Algorithms such as GUIDE (Loh 2002), CRUISE (Kim and Loh 2001) and LOTUS (Chan and Loh 2004) were among the first to provide trees with parametric models in each terminal node.
-The MOB algorithm (Zeileis, Hothorn and Hornik 2008), an algorithm for model-based recursive partitioning, extended this idea by fitting parametric models to each node and using the gained information of the fitted model to create further splits in inner nodes.
-Using this algorithm as a framework to build trees LMs and GLMs can be fitted in the nodes of the trees.
-However, up to now trees which only model the mean of the response variable in its nodes are still more common and widely used.\\
-%Random forests (\citealt{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\citealt{hothorn2015partykit}, \citealt{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
-%\\
-%But as mentioned before, until now the above listed development in the field of probabilistic modeling has not yet been fully integrated in the idea of algorithmic tree and forest models.
-Probabilistic models are rarely used in combination with tree algorithms and more flexible models such as GAMLSSs have not yet been applied as models in the nodes of a tree.\\
-%\\
-%Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests.
-In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale,and shape as well as censoring, tail behavior etc.
-while non-additive effects or interactions of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree.
-In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
-%These (future) steps are illustrated in Figure~\ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
-
-%After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}.
-The new method is applied on precipitation data in a mountainous region with a very large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area.
-In a first setting only one observation station is taken into consideration.
-The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years are validated.
-A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}.
-For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a distributional forest are plotted together with the corresponding observations.
-This illustrates that well fitting predicted expected values
-%%% well fitting predicted expected values -> appropriate probabilistic forecasts? </TH>
-%are provided by this model as they are close to the observed values.
-Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.
-In section \ref{ProbabilisticForecasting} this example will be further discussed and it will be shown that the predictions made by the distributional forest are reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, the scale parameter, and therefore the whole distribution.
+In the following, particular focus is given to distributional forests as a method for obtaining
+probabilistic forecasts by leveraging the strengths of random forests: the ability to
+capture both smooth and abruptly changing functions along with simultaneous selection of
+variables and possibly complex interactions. Thus, these properties make the method particularly
+appealing in case of many covariates with unknown effects where it would be challenging
+to specify a GAMLSS.
+
+In weather forecasting, these properties are appealing in mountainous regions and complex terrain
+where small-scale effects are not well-resolved in the NWP outputs. As these are computed at
+a coarser scale, there may be abrupt local effects in the meteorological observations within a single
+NWP grid cell. To illustrate this in practice, precipitation forecasts are obtained with distributional
+forests at 85 meteorological stations in a mountainous region in the Alps, covering mainly Tyrol,
+Austria, and adjacent areas. More specifically, a zero-censored Gaussian distribution is
+employed to model 24-hour total precipitation so that the zero-censored point mass describes the
+probability of no precipitation on a given day (see Figure~\ref{axams_onestation}). Forecasts
+for July are established based on data from the same month in the years 1985--2012 along with
+a large number of NWP ensemble outputs as covariates. As Figure~\ref{axams_onestation} shows,
+the station-wise models yield a full distributional forecast for each day -- here for the same
+day (July~24) at one station (Axams) over four years (2009--2012) -- based on the previous 24~years as training data.
+The corresponding observations conform reasonably well with the predictions.
 
@@
 \begin{figure}[t!]
-\begin{center}
+\centering
+\setkeys{Gin}{width=0.8\textwidth}
 <<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
@@
      xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
-     #xlab = expression(Total~precipitation~group("[", mm^(1/1.6)~per~day,"]")),
-     #{\frac{1}{1.6}}$,
      ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
@@
 lines(x = x, y = y4, type = "l", col = "purple")
-legend('topright', c("predicted distribution", "point mass at censoring point", "observation"),
+legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
        bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
@@
 \caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
-Observations are left censored at 0.
+Observations are left-censored at 0.
 The corresponding predicted point mass is shown at the censoring point (0).}
-\end{center}
 \end{figure}
 
-%\newpage
 \section{Methodology}
@@
 In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree-structured models step by step.
-We start with a simple distributional fit in Subsection~\ref{distfit}, then build a distributional tree in Subsection~\ref{disttree} and finally a distributional forest in Subsection~\ref{distforest}.
+We start with a simple distributional fit in Section~\ref{distfit}, then build a distributional tree in Section~\ref{disttree} and finally a distributional forest in Section~\ref{distforest}.
 
@@
 
-In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0.
-(The power transformation is used to remove large parts of the skewness of the data as further explained in \ref{data}.) Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
+In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left-censored with censoring point 0.
+(The power transformation is used to remove large parts of the skewness of the data as further explained in Section~\ref{data}.) Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
 $$
@@
 
-%The distribution of the response variable is a left censored normal distribution with censoring point 0.
+%The distribution of the response variable is a left-censored normal distribution with censoring point 0.
 The corresponding distribution parameters are defined by a parameter function $f$ depending on the covariates and an additional factor $\kappa$, also known as the smoothness constraint, which regulates the steepness of certain sections of the parameter function.
@@
 
-%\begin{figure}[t!]
-%\begin{subfigure}{0.6\textwidth}
-%<<sim_rmse_exp, fig=TRUE, echo=FALSE>>=
-%plot_rmse(simres, type = "exp")
-%@
-%\end{subfigure}
-%\hspace{-1cm}
-%\begin{subfigure}{0.6\textwidth}
-%<<sim_rmse_par, fig=TRUE, echo=FALSE>>=
-%plot_rmse(simres, type = "par")
-%@
-%\end{subfigure}
-%\caption{\label{plot_sim_rmse}Results of the simulation study (RMSE).\\
-%left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference %between the observations and the estimated expected value\\
-%right: continuous lines represent the location parameter, dotted lines the scale parameter}
-%\end{figure}
-
-%\begin{figure}[t!]
-%\begin{center}
-%%\begin{subfigure}{0.6\textwidth}
-%<<sim_crps, fig=TRUE, echo=FALSE>>=
-%plot_crps(simres)
-%@
-%\caption{\label{plot_sim_crps}CRPS of the three models evaluated on various data sets over increasing $\kappa$} 
-%%\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest} vs. %\code{bamlss} vs. \code{gamboostLSS}\\
-%%right: levelplot of the parameter function for $\kappa=5$}
-%\end{center}
-%\end{figure}
-
-%\end{subfigure}
-%\hspace{-1cm}
-%\begin{subfigure}{0.6\textwidth}
-%<<sim_levelplot, fig=TRUE, echo=FALSE>>=
-%levelfun <- function(x) simres$fun(x,5)
-%
-%ncuts <- 100
-%colR <- colorRampPalette(c("blue","green"))                                       
-%cols <- colR(ncuts+1)
-%griddata <- expand.grid(list(x1 = seq(-1, 1, 0.01), x4 = seq(0, 1, 0.2)))
-%griddata$mu <- levelfun(griddata)
-%levelplot(mu ~ x1 * x4, data = griddata, xlab = "x1", ylab = "x4", 
-%          region = TRUE, cuts = ncuts, col.regions = cols)
-
-%@
-%\end{subfigure}
-%\caption{\label{plot_sim_ll}left: results of the simulation study: log-likelihood\\ %\code{disttree} vs.
-\code{distforest} vs.
-\code{gamlss} vs.
-\code{randomForest} vs.
-%\code{bamlss} vs.
-\code{gamboostLSS}\\
-%right: levelplot of the parameter function for $\kappa=5$}
-%\end{figure}
-
-%Figure~\ref{plot_sim_crps} clearly illustrates the strengths of the different methods applied in this simulation study and supports the assumptions made above based on the results for the two specific situations in Figure~\ref{plot_sim_oneset_mu}.
-Moreover, it can be observed how the stepwise transition from the one extreme situation to the other influences the performance of the applied methods.\\
-%The parametric GAMLSS can easily deal with the smooth effects that appear for low values of $\kappa$ and outperforms the tree and forest models in this situation.
-However, it clearly starts to struggle as $\kappa$ increases and the effects turn into abrupt changes.
-%This behaviour has already been detected and discussed above when looking at only single data sets and is now proven by an almost monotone increase of the CRPS.\\
-%On the contrary, the imposed/enforced steps of the tree model lead to a high CRPS in the situation of smooth effects while decreasing quickly as the slopes in the parameter functions get steeper.
-For high values of $\kappa$ resulting almost in step functions the tree model perfomes the best as stated by the low CRPS value.\\
-%Therefore, both extreme situations show to have a preferable method to choose which clearly outperforms the others.
-But overall the forest model stays more or less on the same level of goodness of fit regardless of the type of effects in the parameter functions.
-As this level lies inbetween the ones of the other two methods the forest model provides a good compromise.
-If the type of effects is known in advance one should choose the corresponding adequate method.
-But as this is usually not the case, applying a distributional forest offers a good and flexible solution which can deal with a wide variety of situations and in that way ensures a reasonable well fit for a wide rango of situations where no prior information on the form of the effects is available.
-
-%Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups.
-This is the point where the tree structure provides a better solution.
-However, a single tree can not keep up with the performance of \code{gamlss} when it comes to modeling the smooth functions within the subgroups.
-Offering a good compromise the forest takes advantages of the tree structure to find subgroups but can also model smooth effects if enough trees are included.\\
-%However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
-%\\
-%Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation.
-For $\kappa=1$ the performance of all three methods is similar.
-However, as $\kappa$ increases the RMSE of gamlss increases while the RMSE of trees and forests stays at almost the same level or even decreases sligthly as soon as $\kappa$ is greater or equal 2.
-The same conclusion can be drawn from looking at the log-likelihood where the resulting values of gamlss decrease rapidly while the values of trees and forests only decrease a little at the beginning but then increase.\\
-%This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
-%For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become.
-Especially regarding the scale parameter the trees are clearly advantageous over the forests.
-
-%(As it has already been demonstrated in the first examples on single data sets the performances of \code{distforest} and \code{disttree} are not as successful as the one of \code{gamlss} for low values of $\kappa$ but remarkably improve as this values increases and are superior to \code{gamlss} for high values of $\kappa$.
-This is illustrated by an almost constant RMSE of \code{distforest} and \code{disttree} while the RMSE of \code{gamlss} increases.
-Comparing the tree and the forest model it can be observed that the forest seems to perform better for lower $\kappa$ while very steep sections lead to a better performance of trees.
-
 \SweaveOpts{eval = TRUE}
@@
 Each of the 11 ensemble members uses slightly different perturbed initial conditions trying to predict the situation specific uncertainty of the atmospheric state.
-
-%The data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both years included.
-The observations were measured at 95 (?) stations all over Tyrol (and regions close to the frontier).
-The considered data is provided by .....
-%Next to the variable of interest which is the measured total precipitation within 24 hours of each day (measured from 6 UTC to 6 UTC of the following day), numerical forecasts (from ...
-) are used as covariates.
-These forecasts are ensemble predictions consisting of ...
-individual forecasts based on varying/perturbed initial conditions.\\
 
@@
 Since the observation data set used for this optimization contains the observations considered in this article the same factor is chosen here for the power transformation of the response variable \emph{tp} (total precipitation).
-The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.
+The resulting transformed variable is expected to be normally distributed and left-censored at censoring point 0.
 For simplicity the same transformation has been chosen for the variable \emph{cape} (convective available potential energy).
@@
 %\begin{minipage}{\textwidth}
-\begin{center}
+\centering
 \begin{tabular}{ l l l l }
@@
 \hline
-\textbf{distributional forest} & recursive    & all & all \\ 
+\textbf{Distributional forest} & recursive    & all & all \\ 
                 & partitioning &     &     \\ 
 \hline
-\textbf{prespecified GAM} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
+\textbf{Prespecified GAM} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
                             & in each & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean},\\
@@
 \hline
-\textbf{boosted GAM} & spline  & all & all \\
+\textbf{Boosted GAM} & spline  & all & all \\
                        & in each &     &     \\
@@
 %}
-\end{center}
 \end{table}
@@
 
-%\SweaveOpts{eval = TRUE}
 \subsection{Application for one station}
@@
 \begin{figure}[t!]
-\begin{center}
-%\begin{subfigure}{0.65\textwidth}
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
 <<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
@@
 barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
-        horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE,
+        horiz = TRUE, las = 1, axes = FALSE,
         xlab = "Variable importance: mean decrease in CRPS",
@@
 @
-%\end{subfigure}
-%\hspace{-1cm}
-%\begin{subfigure}{0.5\textwidth}
-%<<rain_axams_varim_margeffects1_sigma, fig=TRUE, echo=FALSE>>=
-%plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = %"sigma")
-%@
-%\end{subfigure}
-\caption{\label{varimp} CRPS based variable importance for the top 10 covariates in the distributional forest. Based on data for station Axams, learning period 1985--2008 and assessed in 2009--2012.}
-\end{center}
+\caption{\label{varimp} CRPS-based variable importance for the top 10 covariates in the distributional forest. Based on data for station Axams, learning period 1985--2008 and assessed in 2009--2012.}
 \end{figure}
@@
 \begin{figure}[t!]
-\begin{center}
-%\begin{subfigure}{0.8\textwidth}
+\centering
+\setkeys{Gin}{width=0.99\textwidth}
 <<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
-
 # out of sample
@@
 par(mfrow = c(2,2))
-qqrplot(pit_df_t, nsim = 100, main = "distributional forest", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_g_t, nsim = 100, main = "prespecified GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_gb_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
 qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
 
-qqrplot(pit_ml_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+qqrplot(pit_ml_t, nsim = 100, main = "Boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
 @
 \caption{\label{qqr_oosample} Out-of-sample residual QQ-plot (2009--2012) for station Axams based on models learned on data from 1985--2008.}
-\end{center}
 \end{figure}
 
-%\begin{figure}[t!]
-%\begin{subfigure}{0.5\textwidth}
-%<<rain_axams_qqr_insample, fig=TRUE, echo=FALSE>>=
-%# in sample
-%set.seed(7)
-%par(mfrow = c(2,2))
-%set.seed(1)
-%qqrplot(pit_df_l, nsim = 1, main = "distforest (in bag)")
-%qqrplot(pit_g_l, nsim = 1, main = "gamlss (in bag)")
-%qqrplot(pit_gb_l, nsim = 1, main = "gamboostLSS (in bag)")
-%qqrplot(pit_ml_l, nsim = 1, main = "EMOS (in bag)")
-
-%@
-%\caption{\label{qqr_insample} QQR-Plot for station Axams (in sample)}
-%\end{figure}
-
-%\SweaveOpts{eval = TRUE}
-%\subsubsection{Cross-Validation}
-\indent After analyzing the results of the models on one data set only where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
+After analyzing the results of the models on one data set only where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
 For that reason a cross validation framework is considered for the station Axams.
@@
 if(file.exists("rain_Axams_7x10.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_7x10.rda")
   load("rain_Axams_7x10.rda")
 } else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross_7x10.R")
   source("rain_cross_7x10.R")
@@
 
-colnames(rain_rmse) <- colnames(rain_ll) <- colnames(rain_crps) <-
-  colnames(rainres[[1]]$rmse)
-
+colnames(rain_rmse) <- colnames(rain_ll) <- colnames(rain_crps) <- colnames(rainres[[1]]$rmse)
 @
@@
 \begin{figure}[t!]
-\begin{center}
+\centering
 <<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
 boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
-        names = c("distributional forest", "prespecified GAM", "boosted GAM"),
+        names = c("Distributional forest", "Prespecified GAM", "Boosted GAM"),
         ylab = "CRPS skills score", col = "lightgray") 
@@
 \caption{\label{boxplots_crps_axams}CRPS skills score from the 10 times 7-fold cross validation at station Axams (1985--2012). The horizontal orange line pertains to the reference model EMOS.}
-\end{center}
 \end{figure}
@@
 s <- 1 - crps_all[, 2:4]/crps_all[,6]
-colnames(s) <- c(" distributional forest ", "prespecified GAM", "boosted GAM")
+colnames(s) <- c("Distributional forest", "Prespecified GAM", "Boosted GAM")
 
@@
 
-\begin{figure}[h!]
-\begin{center}
+\begin{figure}[t!]
+\centering
 <<rain_all_crps_skills_score, fig=TRUE, echo=FALSE, width = 6.5>>=
@@
 The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
-\end{center}
 \end{figure}
@@
 
-\begin{figure}[h!]
-\begin{center}
+\begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.99\textwidth}
@@
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend(x = 9.8, y = 47.795, pch=c(21,24,25,22), legend = c("distributional forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1, bty = "n")
+  legend(x = 9.8, y = 47.795, pch=c(21,24,25,22), legend = c("Distributional forest", "Prespecified GAM", "Boosted GAM", "EMOS"), cex = 1, bty = "n")
   text(x = 10.3, y = 47.8, labels = "Models with lowest CRPS")
@@
 The color codes whether the distributional forest had lower or higher CRPS (red or blue) compared to the best of the other three models (prespecified GAM, boosted GAM and EMOS).}
-\end{center}
 \end{figure}
@@
 
-%\newpage
-\section*{Appendix}
-
-\subsection*{Tree algorithm}
+\newpage
+\begin{appendix}
+
+\section{Tree algorithm}
 The tree algorithm used in the applications discussed in this paper is explained in the following.
@@
 
-\subsection*{Probability integral transform (PIT) histograms}
+\section{Probability integral transform (PIT) histograms}
 Probability integral transform (PIT) histograms \cite{gneiting2007} are a commonly used tool to assess the predictive power of distributional forecasting methods, particularly in the field of meteorological forecasting \\
@@
 
-The PIT histograms in Figure~\ref{pit_oosample} correspond to the forecasts presented in Section \ref{Application for one station}. 
+The PIT histograms in Figure~\ref{pit_oosample} correspond to the forecasts presented in Section~\ref{Application for one station}. 
 Daily observations of total precipitation from station Axams are considered and predictions are made for the 24th of July 2009, 2010, 2011 and 2012, based on data from the previous 24 years.
@@
 \begin{figure}[t!]
-\begin{center}
+\centering
 %\begin{subfigure}{0.8\textwidth}
@@
 par(mfrow = c(2,2))
-pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "distributional forest", ylim = c(0,1.5))
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "prespecified GAM", ylim = c(0,1.5))
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAM", ylim = c(0,1.5))
+pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAM", ylim = c(0,1.5))
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAM", ylim = c(0,1.5))
 pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
@@
 \caption{\label{pit_oosample}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
-\end{center}
 \end{figure}
 
-%\begin{figure}[h!]
-%\begin{center}
-%%\begin{subfigure}{0.5\textwidth}
-%<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE, width=10, height=7>>=
-%boxplot(rain_rmse[,c(2,3,4,6)], ylab = "RMSE",
-%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
-%@
-%%\end{subfigure}
-%\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at station Axams.}
-%\end{center}
-%\end{figure}
-%%\hspace{-1cm}
-%\begin{figure}[h!]
-%\begin{center}
-%%\begin{subfigure}{0.5\textwidth}
-%<<rain_cross_axams_ll, fig=TRUE, echo=FALSE, width=10, height=7>>=
-%boxplot(rain_ll[,c(2,3,4,6)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
-%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
-%#boxplot(rain_ll)
-%@
-%%\end{subfigure}
-%%\caption{\label{boxplots_rain}Results 10x10 cross validation at station Axams\\
-%%left: RMSE, right: log-likelihood}
-%\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at station Axams (extracted interval [-2.8,-1.3]).}
-%\end{center}
-%\end{figure}
-
-%\begin{figure}[h!]
-%\begin{center}
-%<<rain_all_rmse, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
-%  matplot(t(rmse[,c(2,3,4,6)]), type = "l", lwd = 1, 
-%          col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, 
-%          xlab = "", ylab = "RMSE", xlim = c(0.5, 4.5))
-%  boxplot(rmse[,c(2,3,4,6)], add = TRUE,
-%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
-%@
-%\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations.}
-%\end{center}
-%\end{figure}
-
-%\begin{figure}[h!]
-%\begin{center}
-%<<rain_all_ll, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
-%  matplot(t(ll[,c(2,3,4,6)]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
-%          lty = 1, axes = FALSE, 
-%          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 4.5), ylim = c(-4,-1))
-%  boxplot(ll[,c(2,3,4,6)], add = TRUE,
-%        names = c("Forest", "prespecified GAM", "boosted GAM", "EMOS")) 
-%@
-%\caption{\label{boxplot_ll_all}Log-likelihood values for predictions over all stations (extracted interval [-4,-1]).}
-%\end{center}
-%\end{figure}
+\end{appendix}
 

### 2018-02-10 pkg/disttree/inst/draft/draft.Rnw
skills score -> skill score
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2006 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 514c18b54b5cc44f59a8aa3f79683a4d4514d93a

@@
 
-For the purpose of comparing the performance of a distributional forest to those of other models the so-called skills score can be calculated with respect to the CRPS and one model being chosen as a reference.
+For the purpose of comparing the performance of a distributional forest to those of other models the so-called skill score can be calculated with respect to the CRPS and one model being chosen as a reference.
 In that way we obtain a measure of how much a method improves the model performance in terms of CRPS compared to the reference method.
@@
 This whole procedure is repeated 10 times.
-The resulting skills scores are illustrated in Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model represented by the horizontal line at height 0.
+The resulting skill scores are illustrated in Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model represented by the horizontal line at height 0.
 <<echo=FALSE, results=hide>>=
@@
 \centering
-<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
+<<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
 boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
         names = c("Distributional forest", "Prespecified GAM", "Boosted GAM"),
-        ylab = "CRPS skills score", col = "lightgray") 
+        ylab = "CRPS skill score", col = "lightgray") 
 abline(h = 0, col = pal["EMOS"], lwd = 2)
 @
-\caption{\label{boxplots_crps_axams}CRPS skills score from the 10 times 7-fold cross validation at station Axams (1985--2012). The horizontal orange line pertains to the reference model EMOS.}
+\caption{\label{boxplots_crps_axams}CRPS skill score from the 10 times 7-fold cross validation at station Axams (1985--2012). The horizontal orange line pertains to the reference model EMOS.}
 \end{figure}
 
-The CRPS skills score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
+The CRPS skill score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
 While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
@@
 At each of them, the first 24 years are used as learning data set and predictions are made for the last 4 years of available observations.
-For these out-of-sample predictions the CRPS skills score with respect to the EMOS model as reference is calculated for each method at each station.
-In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skills score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
+For these out-of-sample predictions the CRPS skill score with respect to the EMOS model as reference is calculated for each method at each station.
+In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skill score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
 The green line represents the results for the station Axams.
@@
 
-# skills score
+# skill score
 s <- 1 - crps_all[, 2:4]/crps_all[,6]
@@
 \centering
-<<rain_all_crps_skills_score, fig=TRUE, echo=FALSE, width = 6.5>>=
+<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, width = 6.5>>=
   matplot(t(s[,]), type = "l", lwd = 2, 
@@
           lty = 1, axes = FALSE, 
-          xlab = "", ylab = "CRPS skills score", xlim = c(0.5, 3.5))
+          xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
 lines(s[70,], col = "limegreen", type = "o", pch = 19, lwd = 2)  
@@
 @
-\caption{\label{boxplot_crps_all}CRPS skills score for each station (grey lines) and aggregated over all stations (boxplots).
+\caption{\label{boxplot_crps_all}CRPS skill score for each station (grey lines) and aggregated over all stations (boxplots).
 Station Axams is highlighted in green and the horizontal orange line pertains to the reference model EMOS.

### 2018-02-10 pkg/disttree/inst/Psychoco2018/slides.Rnw
revised slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2007 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 625699e0cd3e940e2219554aee82644435002814

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber, noslidenumber, url]{uibk}
+%\usetheme[nototalframenumber, noslidenumber, url]{uibk}
+\usetheme[nototalframenumber, noslidenumber]{uibk}
 
@@
 \graphicspath{{plots/}}
+\usepackage{geometry}
 
@@
 \begin{frame}
-\frametitle{Fitting a Tree}
+\frametitle{Distributional trees}
 \vspace*{-0.12cm}
@@
 \begin{frame}[fragile]
-\frametitle{Fitting a Tree}
+\frametitle{Distributional trees}
 \begin{center}
@@
 %\begin{frame}
-%\frametitle{Fitting a Tree}
+%\frametitle{Distributional trees}
 %\center
@@
 \begin{itemize}
-\item Specified log-likelihood function $$\ell(\theta; y_i)$$ with distribution parameter $\theta$ and observation $y_i$ of a learning data set $\{y_i\}_{i = 1,\ldots,n}$.
+\item Specify distribution family with distribution function 
+$$F(y) = \mathbb{P}_{\theta}(Y\leq y)$$ 
+and log-likelihood function $$\ell(\theta; y).$$ 
+The full distribution is then defined by the parameter vector $\theta \in \Theta$ representing location, scale, and shape.
 
@@
 
-\item Estimation of $\hat{\bold{\theta}}$ via maximum likelihood 
+\item Estimate $\hat{\bold{\theta}}$ via maximum likelihood 
 $$
@@
 $$
+for a learning data set $\{y_i\}_{i = 1,\ldots,n}.$
 %% equivalent to solving 
@@
 
-Needed: Weights.
-
-Via tree: 0/1.
-
-Via forest: averaged across trees.
-
-\end{frame}
-
-\subsection{Building Distributional Trees and Forests}
+\bigskip
+
+How to calculate weights $\{w_i(\bold{x})\}_{i=1,\dots,n}$?\\
+
+\bigskip
+
+Two strategies:
+\begin{itemize}
+\item Build a tree and let $w_i(\bold{x}) \in \{0,1\}$ indicate whether $\bold{x}$ is in the same subgroup as the $i$-th learning observation.
+\item Build an ensemble of trees on resampled versions of the learning data and let $w_i(\bold{x}) \in  \mathbb{N}_0$ count the number of trees in which $\bold{x}$ is in the same subgroup as the $i$-th learning observation.
+%\item Via a tree model: $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ is in the same terminal node as the $i$-th learning observation.
+%\item Via a forest model: $w_i(\bold{x}) \in  \mathbb{N}_0$ counts the number of trees in which $\bold{x}$ is in the same terminal node as the $i$-th learning observation. 
+\end{itemize}
+
+\end{frame}
+
+\subsection{Building distributional trees and forests}
 \begin{frame}
-\frametitle{Building Distributional Trees and Forests}
+\frametitle{Building distributional trees and forests}
 {\bf Tree:}
@@
 \frametitle{Simulation}
-
-Overview
-
+{\bf Models}:
+
+\smallskip
+
+\hspace{0.55 em} disttree, distforest (100 trees), gamlss\\
+
+\bigskip
+
+{\bf DGP}:
+
+\smallskip
+
+\begin{tabular}{ll}
+Response:         & $y \sim \mathcal{N}(\mu(x), \sigma(x))$ with\\
+                  & \qquad $\mu(x) =  10 \cdot \exp\left(-(4 \cdot x - 2)^{4}\right)$\\
+                  & \qquad $\sigma(x) = 0.5 + 2 \cdot |x|$\\
+Covariate:        & $x \sim  \mathcal{U}([-0.4,1])$\\
+\# observations:  & 300 
+\end{tabular}
 \end{frame}
@@
 
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%<<oneforest_exp1_sigma, fig=TRUE, echo=FALSE>>=
-%plot1sigma(learndata,dt, df, g)
-%@
-%\end{center}
-%\end{frame}
+\begin{frame}
+\frametitle{Simulation}
+
+{\bf Models}:
+
+\footnotesize
+\hspace{0.55 em} disttree, distforest (100 trees), gamlss\\
+
+\bigskip
+
+\normalsize
+{\bf DGP}:
+
+\smallskip
+
+\footnotesize
+\begin{tabular}{ll}
+Response:         & $y \sim \mathcal{N}(\mu(x), \sigma(x))$ with\\
+                 & \qquad $\mu(x) =  10 \cdot \exp\left(-(4 \cdot x - 2)^{2 \cdot \kappa}\right)$\\
+                 & \qquad $\sigma(x) = 0.5 + 2 \cdot |x|$\\
+                 & (increasing $\kappa$ leads to steeper transitions)\\
+Covariate:        & $x \sim  \mathcal{U}([-0.4,1])$\\
+\# observations:  & 1000 
+\end{tabular}
+
+\bigskip
+
+\normalsize
+{\bf Evaluation} over increasing $\kappa$:
+
+\smallskip
+
+\footnotesize
+\begin{tabular}{ll}
+Interval for $\kappa$:     & $[1, 71]$ \\
+Stepsize for $\kappa$:     & 7 \\
+\# replications per step:  & 150 
+\end{tabular}
+
+\end{frame}
 
@@
        xlab = "", ylab = "", xaxt="n", yaxt="n", lwd = 2)
+  axis(1, at=c(1,71), lab=c(1, 71), mgp=c(3, .5, 0))
   title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
@@
   #mtext(text= "vs.", side = 3, col = "black", line = 1, adj = 0.6)
-  mtext(text= "smooth", side = 1, line = 1, adj = 0)
-  mtext(text= "steep", side = 1, line = 1, adj = 1)
-  mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
+  mtext(text= "smooth", side = 1, line = 2, adj = 0)
+  mtext(text= "steep", side = 1, line = 2, adj = 1)
+  mtext(text= TeX('$\\kappa$'), side = 1, line = 2)
   mtext(text= "Log-Likelihood", side = 2, line = 1)
@@
   #add arrows below
-  arrows(0.45,0.085,0.3,0.085,lwd=1) ## add arrow
-  arrows(0.6,0.085,0.75,0.085,lwd=1) ## add arrow
+  arrows(0.45,0.047,0.3,0.047,lwd=1) ## add arrow
+  arrows(0.6,0.047,0.75,0.047,lwd=1) ## add arrow
 
@@
 \begin{itemize}
-\item Recommendations by subject matter literature.
-\item Experiences from other modelling situations.
+\item Recommendations by specialist literature.
+\item Expert subject matter knowledge.
+\item Experiences from other modeling situations.
 \item Common simplifications/generalizations/assumptions.
@@
 \begin{frame}
-\frametitle{Alternative: Transformation model}
-
-\medskip
+\frametitle{Alternative: Transformation models}
+{\bf Advantages:}
+\begin{itemize}
+\item Does not require specification of distribution family.
+\item More flexible framework.
+\end{itemize}
+
+\bigskip
 
@@
 
+%\newgeometry{left=0.2cm, right=0.2cm, top = 0.5cm, bottom=0.5cm}
 \begin{frame}
@@
 \vspace*{0.4cm}
-
-\includegraphics[width=\textwidth]{cmpx-tree.pdf}
-
-\end{frame}
+\includegraphics[width=1.05\textwidth]{cmpx-tree.pdf}
+
+\end{frame}
+%\restoregeometry
 

### 2018-02-10 pkg/disttree/inst/Psychoco2018/slides.Rnw
tried to reduce complexity in estimation and simulation sections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2008 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3447011d4cf715bb6885af7875d01db529eecda0

@@
 %\usetheme[nototalframenumber, noslidenumber, url]{uibk}
-\usetheme[nototalframenumber, noslidenumber]{uibk}
+\usetheme[nototalframenumber, noslidenumber, url]{uibk}
 
@@
 
-\textbf{Distributional:} \\
-\vspace{0.1 cm}
-\hspace{0.5cm} Specify the complete probability distribution\\
-\hspace{0.5cm} (including location, scale, and shape).
+\textbf{Distributional:}
+\begin{itemize}
+  \item Specify the complete probability distribution (including location, scale, and shape).
+\end{itemize}
 
@@
 
-{\bf Tree:}
+\textbf{Tree:}
 \begin{itemize}
@@
 
-{\bf Forest:}
+\textbf{Forest:}
 \begin{itemize}
@@
 \begin{itemize}
-\item Specify distribution family with distribution function 
-$$F(y) = \mathbb{P}_{\theta}(Y\leq y)$$ 
-and log-likelihood function $$\ell(\theta; y).$$ 
-The full distribution is then defined by the parameter vector $\theta \in \Theta$ representing location, scale, and shape.
-
-\bigskip
+
+\item Specify a parametric distribution family $F(\cdot; \theta)$ with parameter vector
+  $\theta \in \Theta$ capturing location, scale, shape.
+
+\item Cumulative distribution function and log-likelihood:
+  \begin{eqnarray*}
+  F(y; \theta)    & = & \mathbb{P}_{\theta}(Y\leq y)\\
+  \ell(\theta; y) & = & \log(f(y; \theta))
+  \end{eqnarray*}
 
 \item Estimate $\hat{\bold{\theta}}$ via maximum likelihood
-$$
-\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i).
-$$
-for a learning data set $\{y_i\}_{i = 1,\ldots,n}.$
+  based on a learning sample $y_1,\dots, y_n$:
+\[
+\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
+\]
+
 %% equivalent to solving 
@@
 
-$$
+\textbf{Idea:} Covariates captured through adaptive weights.
+\[
 \hat{\theta}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i).
-$$
+\]
 
@@
 
-How to calculate weights $\{w_i(\bold{x})\}_{i=1,\dots,n}$?\\
+\textbf{Question:} How to choose weighting function $w_i(\bold{x})$?
 
@@
 
-Two strategies:
+\textbf{Possible answers:} Based on learning sample $y_1, \dots, y_n$ and (possibly new) observation $\bold{x}$.
 \begin{itemize}
-\item Build a tree and let $w_i(\bold{x}) \in \{0,1\}$ indicate whether $\bold{x}$ is in the same subgroup as the $i$-th learning observation.
-\item Build an ensemble of trees on resampled versions of the learning data and let $w_i(\bold{x}) \in  \mathbb{N}_0$ count the number of trees in which $\bold{x}$ is in the same subgroup as the $i$-th learning observation.
-%\item Via a tree model: $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ is in the same terminal node as the $i$-th learning observation.
-%\item Via a forest model: $w_i(\bold{x}) \in  \mathbb{N}_0$ counts the number of trees in which $\bold{x}$ is in the same terminal node as the $i$-th learning observation. 
+\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $y_i$ are classified into the same subgroup.
+\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $y_i$ across trees.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Building distributional trees and forests}
-{\bf Tree:}
+\frametitle{Distributional trees and forests}
+\textbf{Tree:}
 \begin{enumerate}
-\item Estimate $\hat{\theta}$ via maximum likelihood.
+\item Estimate $\hat{\theta}$ via maximum likelihood (without covariates).
 \item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
@@
 
-{\bf Forest:} Ensemble of $T$ trees.
+\textbf{Forest:} Ensemble of $T$ trees.
 \begin{itemize}
@@
 
-{\bf Estimator:}
-$$
+\textbf{Estimator:}
+\[
 \hat{\theta}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i)
-$$
+\]
 
@@
 \frametitle{Simulation}
-{\bf Models}:
-
-\smallskip
-
-\hspace{0.55 em} disttree, distforest (100 trees), gamlss\\
+
+\textbf{Models:} \code{disttree}, \code{distforest} (100 trees), \code{gamlss}.
 
@@
 
-{\bf DGP}:
-
-\smallskip
-
-\begin{tabular}{ll}
-Response:         & $y \sim \mathcal{N}(\mu(x), \sigma(x))$ with\\
-                  & \qquad $\mu(x) =  10 \cdot \exp\left(-(4 \cdot x - 2)^{4}\right)$\\
-                  & \qquad $\sigma(x) = 0.5 + 2 \cdot |x|$\\
-Covariate:        & $x \sim  \mathcal{U}([-0.4,1])$\\
-\# observations:  & 300 
-\end{tabular}
+\textbf{Data:}
+\begin{eqnarray*}
+  y         & \sim & \mathcal{N}(\mu(x), \sigma(x))\\
+  x         & \sim & \mathcal{U}(-0.4, 1) \\
+  \mu(x)    & =    & 10 \cdot \exp\left\{ -(4 \cdot x - 2)^{2 \cdot \kappa} \right\}\\
+  \sigma(x) & =    & 0.5 + 2 \cdot |x|
+\end{eqnarray*}
+
+\bigskip
+
+\textbf{Parameters:}
+\begin{itemize}
+  \item 1 replication: $n = 300$, $\kappa = 2$.
+  \item 150 replications: $n = 1000$, $\kappa = 1, 8, 15, \dots, 71$.
+\end{itemize}
 \end{frame}
@@
 \end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Simulation}
-
-{\bf Models}:
-
-\footnotesize
-\hspace{0.55 em} disttree, distforest (100 trees), gamlss\\
-
-\bigskip
-
-\normalsize
-{\bf DGP}:
-
-\smallskip
-
-\footnotesize
-\begin{tabular}{ll}
-Response:         & $y \sim \mathcal{N}(\mu(x), \sigma(x))$ with\\
-                 & \qquad $\mu(x) =  10 \cdot \exp\left(-(4 \cdot x - 2)^{2 \cdot \kappa}\right)$\\
-                 & \qquad $\sigma(x) = 0.5 + 2 \cdot |x|$\\
-                 & (increasing $\kappa$ leads to steeper transitions)\\
-Covariate:        & $x \sim  \mathcal{U}([-0.4,1])$\\
-\# observations:  & 1000 
-\end{tabular}
-
-\bigskip
-
-\normalsize
-{\bf Evaluation} over increasing $\kappa$:
-
-\smallskip
-
-\footnotesize
-\begin{tabular}{ll}
-Interval for $\kappa$:     & $[1, 71]$ \\
-Stepsize for $\kappa$:     & 7 \\
-\# replications per step:  & 150 
-\end{tabular}
-
 \end{frame}
@@
 \medskip
-{\bf Based on:}
+\textbf{Based on:}
 \begin{itemize}
@@
 \bigskip
-{\bf Examples:}
+\textbf{Examples:}
 \begin{itemize}
@@
 \begin{frame}
-\frametitle{Alternative: Transformation models}
-{\bf Advantages:}
+\frametitle{Transformation models}
+
+\textbf{Advantages:}
 \begin{itemize}
@@
 
-{\bf Distribution function:}
-$$
-F(y) = \Phi(\bold{a}_{Bs,d}(y)^\top \theta) 
-$$
-
+\textbf{Distribution function:}
+\[
+F(y; \theta) = \Phi(\bold{a}_{Bs,d}(y)^\top \theta) 
+\]
 \begin{itemize} 
@@
 
-%\newgeometry{left=0.2cm, right=0.2cm, top = 0.5cm, bottom=0.5cm}
 \begin{frame}
-\frametitle{Alternative: Transformation model}
+\frametitle{Transformation models}
 
@@
 \end{frame}
-%\restoregeometry
 
@@
 \begin{tabular}{ll}
-\code{distfit}    & Distributional fits (ML, \code{gamlss.family}/custom \code{list}).\\
+\code{distfit}    & Distributional fit (ML, \code{gamlss.family}/custom \code{list}).\\
                   & No covariates. \\
-\code{disttree}   & Distributional trees (\code{ctree}/\code{mob} + \code{distfit}).\\
+\code{disttree}   & Distributional tree (\code{ctree}/\code{mob} + \code{distfit}).\\
                   & Covariates as partitioning variables. \\
-\code{distforest} & Distributional forests (ensemble of \code{disttree}s).\\
+\code{distforest} & Distributional forest (\code{disttree} ensemble).\\
                   & Covariates as partitioning variables.
@@
 \footnotesize
+
+Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
+  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
+  \emph{Manuscript under preparation}.
+
+\medskip
+
+Hothorn T, Zeileis A (2017).
+  \dquote{Transformation Forests.}
+  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
+  \url{http://arxiv.org/abs/1701.02110}
+
+\medskip
+
+Hothorn T, Zeileis A (2015).
+ \dquote{partykit: A Modular Toolkit for Recursive Partytioning in R.}
+ \textit{Journal of Machine Learning Research},
+ \textbf{16}, 3905--3909.
+ \url{http://www.jmlr.org/papers/v16/hothorn15a.html}
+
+\medskip
 
@@
 
-Hothorn T, Zeileis A (2015).
- \dquote{partykit: A Modular Toolkit for Recursive Partytioning in R.}
- \textit{Journal of Machine Learning Research},
- \textbf{16}, 3905--3909.
- \url{http://www.jmlr.org/papers/v16/hothorn15a.html}
-
-\medskip
-
 Stasinopoulos DM, Rigby RA (2007).
@@
 
-\medskip
-
-Seibold H, Zeileis A, Hothorn T (2017).
-  \dquote{Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.}
-  \textit{Statistical Methods in Medical Research}, 
-  \textbf{12}(1), 45--63.
-  \doi{10.1177/0962280217693034}
-
-\medskip
-
-Hothorn T, Zeileis A (2017).
-  \dquote{Transformation Forests.}
-  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
-  \url{http://arxiv.org/abs/1701.02110}
-
-\end{frame}
-
-%% \begin{frame}[fragile]
-%% \frametitle{Simulation}
-%% \begin{center}
-%% \vspace*{-1cm}
-%% \setkeys{Gin}{width=0.8\linewidth}
-%% <<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
-%% plot_rmse(simlist, type = "par")
-%% @
-%% \end{center}
-%% \end{frame}
-
-%% \begin{frame}[fragile]
-%% \frametitle{Simulation}
-%% \begin{center}
-%% \vspace*{-1cm}
-%% \setkeys{Gin}{width=0.8\linewidth}
-%% <<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=
-%% plot_rmse(simlist, type = "exp")
-%% @
-%% \end{center}
-%% \end{frame}
+\end{frame}
 

### 2018-02-11 pkg/disttree/inst/draft/draft.Rnw
more work on intro, cleanup, and some comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2009 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e8c8925f6162dab52e04413de746c6671c303ade

@@
 options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-library("ggplot2")
-theme_set(theme_bw(base_size = 18))
 library("disttree")
@@
 library("gamlss")
-library("randomForest")
-library("lattice")
 library("crch")
-library("latex2exp")
-library("parallel")
 library("gamlss.cens")
@@
 library("mgcv")
-library("MASS")
 library("countreg")
@@
 Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
-
-%For example in the field of probabilistic weather forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools allowing for clear interpretations.
-Usually acting on the assumption of a normally distributed response variable both parameters, the mean and the variance, depend linearly on covariates in this type of model.
-However, the restriction to model only linear effects requires further extensions which allow for more flexibility.
-Generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM).
-A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}).
-This class of models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
-We are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-Applying these novel approaches to precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
-
-%In regression analysis one is interested in the relationship between a dependent variable and one or more explanatory variables.
-Various methods to fit statistical models to the data set have been developed, starting from linear models considering only the mean of the response variable and ranging to probabilistic models where all parameters of a distribution are fit to the given data set.\\
-%If there is a strong variation within the data it might be advantageous to split the data first into more homogeneous subgroups based on given covariates and then fit a local model in each subgroup rather than fitting one global model to the whole data set.
-This can be done by applying regression trees and forests.\\
-%Both of these two concepts, parametric modeling and algorithmic trees, have been investigated and developed further, however, mostly separated from each other.
-Therefore, our goal is to embed the progress made in the field of probabilistic modeling in the idea of algorithmic tree and forest models.
-In particular, more flexible models such as GAMLSS (\citealt{stasinopoulos2005}) should be fitted in the nodes of a tree in order to capture location, scale, and shape as well as censoring, tail behavior etc.
-while non-additive effects of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree.
-In distributional forests an ensemble of distributional trees is built and used to calculate weights which are then included in the fitting process of a distributional model.
-These forest models can detect smooth effects as well as abrupt changes and interactions in a reasonable way without needing any kind of variable selection or information about the expected effects advance and therefore offers a good compromise particularly in complex settings.
 }
@@
 appealing in case of many covariates with unknown effects where it would be challenging
-to specify a GAMLSS.
-
-In weather forecasting, these properties are appealing in mountainous regions and complex terrain
-where small-scale effects are not well-resolved in the NWP outputs. As these are computed at
-a coarser scale, there may be abrupt local effects in the meteorological observations within a single
-NWP grid cell. To illustrate this in practice, precipitation forecasts are obtained with distributional
-forests at 85 meteorological stations in a mountainous region in the Alps, covering mainly Tyrol,
-Austria, and adjacent areas. More specifically, a zero-censored Gaussian distribution is
-employed to model 24-hour total precipitation so that the zero-censored point mass describes the
-probability of no precipitation on a given day (see Figure~\ref{axams_onestation}). Forecasts
-for July are established based on data from the same month in the years 1985--2012 along with
-a large number of NWP ensemble outputs as covariates. As Figure~\ref{axams_onestation} shows,
-the station-wise models yield a full distributional forecast for each day -- here for the same
-day (July~24) at one station (Axams) over four years (2009--2012) -- based on the previous 24~years as training data.
-The corresponding observations conform reasonably well with the predictions.
+to specify a distributional regression model like GAMLSS.
+
+In weather forecasting, these properties are appealing in mountainous regions
+and complex terrain where small-scale effects are not well-resolved in the NWP
+outputs. As these are computed at a coarser scale, there may be abrupt local
+effects within a single NWP grid cell. To illustrate this in practice,
+precipitation forecasts are obtained with distributional forests at 95
+meteorological stations in a mountainous region in the Alps, covering mainly
+Tyrol, Austria, and adjacent areas. More specifically, a zero-censored Gaussian
+distribution is employed to model 24-hour total precipitation so that the
+zero-censored point mass describes the probability of no precipitation on a
+given day (see Figure~\ref{axams_onestation}). Forecasts for July are
+established based on data from the same month in the years 1985--2012 along with
+82~covariates derived from a wide range of NWP ensemble outputs. As
+Figure~\ref{axams_onestation} shows, the station-wise models yield a full
+distributional forecast for each day -- here for the same day (July~24) at one
+station (Axams) over four years (2009--2012) -- based on the previous 24~years
+as training data. The corresponding observations conform reasonably well with
+the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
+performance of distributional forests in this forecasting task in more detail.
+It is shown that they perform at least on par and sometimes clearly better than
+three alternative zero-censored Gaussian models: a standard ensemble model
+output statistics approach (EMOS, \citealp{gneiting2005calibrated}), a GAMLSS
+with regressors prespecified based on meteorological expertise following
+\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
+\citep{gamboostLSS2016} using non-homogenous boosting
+\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
+82~available regressors.
 
@@
 Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.
-Altogether 82 covariates are available in this setting.
+Altogether 82~covariates are available in this setting.
 
@@
 
+\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAM. Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}
+
+\fixme{A description of the boosting approach is currently missing completely.}
+
 In the EMOS model the logarithmic function is chosen as the link function for the scale parameter.
@@
 %%%<TH> always us the term "distributional forest"
-As shown in Table~\ref{model_specification} the distributional forest includes all 82 covariates.
+As shown in Table~\ref{model_specification} the distributional forest includes all 82~covariates.
 A single tree model allows to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable.
@@
 \section*{Computational details}
-The corresponding implementation to the novel methods presented in this article is available on R-Forge in an \textsf{R} package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}).
+The corresponding implementation to the novel methods presented in this article is available on \textsf{R}-Forge in an \textsf{R} package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}).
 The function \code{disttree} uses either the \code{mob} or \code{ctree} function as a framework to build a tree.
@@
 All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}).
+
+\fixme{Avoid nocite, make all relevant citations properly in the text.}
 

### 2018-02-13 pkg/disttree/inst/draft/draft.Rnw
disttree and distforest without family list stored in inner nodes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2014 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 38e2320ef5303575cdb1185228fee02dbd65ece3

@@
 
-# dist_list_cens_normal
-{
-
-  dist_list_cens_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
-    par <- c(eta[1], exp(eta[2]))
-    val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
-    if(sum) {
-      if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
-    par <- c(eta[1], exp(eta[2]))
-    # y[y==0] <- 1e-323
-
-    score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
-    score <- cbind(score_m, score_s)
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y)[1])
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-      #if(any(is.nan(score))) print(c(eta, "y", y))
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    par <- c(eta[1], exp(eta[2]))                           
-    # y[y==0] <- 1e-323
-
-    d2mu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    d2sigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-    dmudsigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu.sigma", left = left, right = right) # FIX: order?
-    dsigmadmu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma.mu", left = left, right = right) # FIX: order?
-    dsigma <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-
-    d2ld.etamu2 <- sum(weights * d2mu, na.rm = TRUE)
-    d2ld.etamu.d.etasigma <- sum(weights * dmudsigma * par[2], na.rm = TRUE)
-    d2ld.etasigma.d.etamu <- sum(weights * dsigmadmu * par[2], na.rm = TRUE)
-    d2ld.etasigma2 <- sum(weights * (d2sigma * exp(2*eta[2]) + dsigma * par[2]), na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etasigma.d.etamu, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- function(q, eta, lower.tail = TRUE, log.p = FALSE) crch:::pcnorm(q, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  qdist <- function(p, eta, lower.tail = TRUE, log.p = FALSE) crch:::qcnorm(p, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  rdist <- function(n, eta) crch:::rcnorm(n, mean = eta[1], sd = eta[2], left = left, right = right)
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    yc <- pmax(0,y)  # optional ?
-    if(is.null(weights)) {
-      mu <- mean(yc)
-      sigma <- sqrt(1/length(yc) * sum((yc - mu)^2))
-    } else {
-      mu <- weighted.mean(yc, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (yc - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- FALSE
-
-  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
-                                ddist = ddist, 
-                                sdist = sdist, 
-                                hdist = hdist,
-                                pdist = pdist,
-                                qdist = qdist,
-                                rdist = rdist,
-                                link = link, 
-                                linkfun = linkfun, 
-                                linkinv = linkinv, 
-                                linkinvdr = linkinvdr,
-                                startfun = startfun,
-                                mle = mle,
-                                gamlssobj = FALSE,
-                                censored = TRUE
-  )
-}
-
 # plot RMSE
@@
 %% late: introduce MOS and EMOS? more details about precipitation forecasting
-%In ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) this 
-%idea is taken one step further. Usually acting on the assumption of a normally distributed
-%response variable this model allows for both parameters, the mean and the variance, to depend 
-%linearly on covariates provided by NWPs. However, the restriction to model only linear effects 
-%has risen the interest in applying more flexible models such as offered by the GAMLSS framework.
+%In ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) this idea is taken one step further.
+Usually acting on the assumption of a normally distributed response variable this model allows for both parameters, the mean and the variance, to depend linearly on covariates provided by NWPs.
+However, the restriction to model only linear effects has risen the interest in applying more flexible models such as offered by the GAMLSS framework.
 

### 2018-02-13 pkg/disttree/inst/draft/rain_axams_pred_24to4.R
disttree and distforest without family list stored in inner nodes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2014 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 38e2320ef5303575cdb1185228fee02dbd65ece3

@@ -26,141 +26,7 @@ rain_axams_pred <- function(seedconst = 7, ntree = 100,
   
   #setwd("~/svn/partykit/pkg/disttree/inst/draft/")
   #setwd("~/disttree/inst/draft/")
-  
-  
-  # dist_list_cens_normal
-  {
-    
-    dist_list_cens_normal <- list()
-    
-    parnames <- c("mu", "sigma")
-    etanames <- c("mu", "log(sigma)")
-    
-    ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
-      par <- c(eta[1], exp(eta[2]))
-      val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
-      if(sum) {
-        if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
-        val <- sum(weights * val, na.rm = TRUE)
-      }
-      return(val)
-    }
-    
-    
-    sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
-      par <- c(eta[1], exp(eta[2]))
-      # y[y==0] <- 1e-323
-      
-      score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-      score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
-      score <- cbind(score_m, score_s)
-      score <- as.matrix(score)
-      colnames(score) <- etanames
-      if(sum) {
-        if(is.null(weights)) weights <- rep.int(1, length(y)[1])
-        # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
-        score[score==Inf] = 1.7e308
-        score <- colSums(weights * score, na.rm = TRUE)
-        #if(any(is.nan(score))) print(c(eta, "y", y))
-      }
-      return(score)
-    }
-    
-    
-    hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
-      ny <- length(y)
-      if(is.null(weights)) weights <- rep.int(1, ny)
-      
-      par <- c(eta[1], exp(eta[2]))                           
-      # y[y==0] <- 1e-323
-      
[…]

### 2018-02-15 pkg/disttree/inst/draft/draft.Rnw
revised parts of the application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2015 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b07a2cfd1928cbe372b854305feb99a1e2712a18

@@
 
-\SweaveOpts{eval = FALSE}
-%\newpage
-%\section{Simulation Study}
-%\label{Simulation}
-%Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.
-
-%In the case of smooth parameter functions without any abrupt changes / non-additive effects a smooth parametric model such as GAMLSS is clearly the first choice while a stepfunction can be modeled best by applying a tree-structured model.
-Even though for both of these extreme situations a specifically advantageous method is provided, a forest model can be a good compromise and performs well in a wide range of settings including the two mentioned before.
-(Therefore, a forest is a recommendable choice particularly when no information about the type of parameter function is provided.)
-
-%Generally, a smooth parametric model such as GAMLSS should be the most successful method if the parameter function is smooth with no rapid changes.
-However, if there are any jumps or non-additive effects or interactions, the trees should be more effective in detecting these steps and therefore also lead to a better fitting forest.
-But this situation might also appear for smooth functions which contain sections with very steep slopes.
-Therefore, it is now of great interest, whether the concept of smooth parametric models or of trees and forests covers a broader range of functions where it is superior to the other.\\
-
-%The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\citealt{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
-%Each generated data set includes 500 samples consisiting of the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
-%$$
-%X_1, X_2, X_3 \sim \textit{Unif}([-1,1])
-%$$
-%$$
-%X_4, X_5, X_6 \sim \textit{Binom}(1, 0.5)
-%$$
-
-%The distribution of the response variable is a left-censored normal distribution with censoring point 0.
-The corresponding distribution parameters are defined by a parameter function $f$ depending on the covariates and an additional factor $\kappa$, also known as the smoothness constraint, which regulates the steepness of certain sections of the parameter function.
-For low values of $\kappa$ the resulting parameters change rather slowly over varying values of the covariates while high values of $\kappa$ lead to abrupt changes.
-\\
-%The location parameter $\mu$ depends on $X_1, X_4$ and $\kappa$ and the scale parameter $\sigma$ depends on $\mu, X_1$ and $\kappa$.
-The other covariates $X_2, X_3, X_5$ and $X_6$ are used as noise variables.\\
-%The latent variable $Y^{\ast}$ is generated based on the following setting.
-%\begin{center}
-%$$ 
-%Y^{\ast} \sim \mathcal{N} (f(X_1, X_4, \kappa))
-%$$
-%with
-%$$
-%f(X_1, X_4, \kappa) = \left(\mu(X_1, X_4, \kappa) , \sigma(\mu,X_1,\kappa)\right)
-%$$ 
-%and
-%\begin{align*}
-%\mu(X_1, X_4, \kappa) = 0.2 
-%+ \begin{cases}
-%    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
-%    & \text{if } X_1 < 0.4 \text{ and } X_4 = 0\\
-%    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
-%    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 0\\
-%    10 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
-%    & \text{if } X_1 < 0.4 \text{ and } X_4 = 1\\
-%    13 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
-%    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 1
-%\end{cases}
-%\end{align*}
-
-%$$
-%\sigma(\mu, X_1,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.4)
-%$$
-%%where $\Lambda$ is the cumulative distribution function of the logistic distribution
-%%$$
-%%\Lambda(\omega) = \frac{1}{1 + \exp(-\omega)}
-%%$$
-%The response variable $Y$ is defined as 
-%$$
-%Y = 
-%\begin{cases}
-%    Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
-%    0,       & \text{else}
-%\end{cases}
-%$$
-%\end{center}
-
-%To illustrate the parameter functions and the impact of $\kappa$ together with the results of the considered methods they are first applied on one generated data set only where $\kappa$ is set to~1 and then to a second data set where $\kappa$ is set to 10.
-In Figure~\ref{plot_sim_oneset_mu} the location parameter function $\mu(X_1, X_4, \kappa)$ is plotted along the covariate $X_1$ together with the estimated values of the different models for fixed covariates $x_4 = 0$ and $(x_2, x_3, x_5, x_6) = (0,0,0,0)$.
-\\
-
-%\SweaveOpts{eval = TRUE}
-<<eval = FALSE, echo=FALSE, results=hide>>=
-
-source("oneset.R")
-
-#if(file.exists("oneset1.rda")){
-#  load("oneset1.rda")
-#} else {
-
-  #define parameter function
-  #kappa <- 1
-  fun1 <- function(x){
-    mu <- 0.2 + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*1)))) * as.numeric(x[,1] < 0.4) + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*1))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
-      (3 * (exp(-(3*x[,1]-1.2)^(2*1)))) * x[,4]
-    sigma <- 0.5 + 
-      mu/5 + 
-      #(x[,1]>0.4)  #*x[,4]#+
-      2*(x[,1] < -0.4) 
-      #plogis((1^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
-      #4*(1-plogis((1^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
-      #5 * abs(x[,1]) * x[,4]
-    par <- cbind(mu, sigma)
-    return(par)
-  }
-
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-
-  oneset1 <- sim_oneset_vary_x1(parfun = fun1, nobs = 500, seedconst = 74, ntree = 100,
-                                 formula = y~x1+x2+x3+x4+x5+x6, 
-                                 tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
-                                 forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
-                                 forest_mtry = 3, type.tree = "ctree", 
-                                 censNO = TRUE, 
-                                 pred_fix_x2 = 0.9,
-                                 pred_fix_x4 = 0,
-                                 nrunif = 3,
-                                 nrcovform = 6)
-
-#  save(oneset1, file = "oneset1.rda")
-#}
-
-#if(file.exists("oneset10.rda")){
-#  load("oneset10.rda")
-#} else {
-
-  #define parameter function
-  #kappa <- 10
-  fun10 <- function(x){
-    mu <- 0.2 + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*10)))) * as.numeric(x[,1] < 0.4) + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*10))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
-      (3 * (exp(-(3*x[,1]-1.2)^(2*10)))) * x[,4]
-    sigma <- 0.5 + 
-      mu/5 + 
-      #(x[,1]>0.4)  #*x[,4]#+
-      2*(x[,1] < -0.4) 
-      #plogis((10^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
-      #4*(1-plogis((10^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
-      #5 * abs(x[,1]) * x[,4]
-    par <- cbind(mu, sigma)
-    return(par)
-  }
-
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-
-  oneset10 <- sim_oneset_vary_x1(parfun = fun10, nobs = 500, seedconst = 74, ntree = 100,
-                                 formula = y~x1+x2+x3+x4+x5+x6, 
-                                 tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95, 
-                                 forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
-                                 forest_mtry = 3, type.tree = "ctree", 
-                                 censNO = TRUE, 
-                                 pred_fix_x2 = 0.9,
-                                 pred_fix_x4 = 0,
-                                 nrunif = 3,
-                                 nrcovform = 6)
-
-#  save(oneset10, file = "oneset10.rda")
-#}
-@
-
-%\begin{figure}[t!]
-%\begin{subfigure}{0.6\textwidth}
-%<<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
-%plot_oneset(oneset1, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
-%@
-%\end{subfigure}
-%\hspace{-1cm}
-%\begin{subfigure}{0.6\textwidth}
-%<<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
-%plot_oneset(oneset10, compare_mu =TRUE, add_dt = TRUE, add_df = TRUE, add_g = TRUE, plot_legend = TRUE, w_sigma = FALSE)
-%@
-%\end{subfigure}
-%\caption{\label{plot_sim_oneset_mu} Smooth parameter function (left window, $\kappa = 1$) and parameter function with abrupt changes (right %window, $\kappa = 10$) together with the predicted location parameter $\mu$ of the tree model, the forest model and the GAMLSS,  plotted along %$x_1$ while all the other covariates are kept at fixed values.}
-%\end{figure}
-
-%In the case of smooth effects of the coefficients on the parameter function ($\kappa = 1$, left window in Figure~\ref{plot_sim_oneset_mu}) the fitted parameters of the GAMLSS are closest to the true parameters while the steps of the tree are cleary too rough.
-By combining various trees the forest can smooth out these steps, however, not as well as the GAMLSS.\\
-%Looking at the other extreme situation in which the parameter functions include abrupt changes ($\kappa = 10$, right window in Figure~\ref{plot_sim_oneset_mu}) the tree captures them very well wich also leads to a good fit of the forest model.
-In contrast to the two tree-structured models the GAMLSS has its difficulties with these steps and therefore fluctuates around the true parameter functions.\\
-
-%To get a more profound impression, the methods are evaluated on various data sets. The value of $\kappa$ is increasing from 1 to 10 by a stepsize of 1 and for each step 100 data sets are generated.\\
-%%\subsubsection{Means of Comparison}
-%In order to compare the performance of the chosen methods over various data sets, the continuous ranked probability score (CRPS) is calculated for each model on each data set.
-
-%three different values are calculated for the resulting models: the root of the mean squared error (RMSE), the log-likelihood value and the continuous ranked probability score (CRPS).
-All of them are evaluated out of sample, e.g.
-on a part of the data set that has not been used to build the model.\\
-%Therefore, overfitting would lead to worse results.\\
-%\\
-%The RMSE represents how close the estimated expected values are to the true observations.
-For a response variable $Y$ and a set of observations $\textbf{y} = \{y_i\}_{i = 1,\ldots,n}$ with the corresponding estimated parameters $\hat{\boldsymbol{\theta}} = \{\hat{\theta}_i\}_{i = 1,\ldots,n}$ the RMSE is defined as follows.
-
-%$$
-%\text{RMSE}(Y, \boldsymbol{y}, \boldsymbol{\hat{\theta}}) = \sqrt{\frac{1}{n} \sum_{i=1}^n (\mathbb{E}_{\hat{\theta}_i}(Y) - y_i)^2}
-%$$
-
-%In most cases this value does not measure the distributional fit as only the expected value is considered.
-However, in the case of a censored normal distribution the expected value of the observed variable is a transformation of the expected value of the latent variable which includes the location and the scale parameter.\\
-%\\
-%A typical way to evaluate the distributional fit is provided by the log-likelihood value.
-For each observation $y$ and the corresponding predicted parameter $\hat{\theta}$ the value of the log-likelihood function $\ell(\hat{\theta}; y)$ is calculated and then averaged over all observations in the testing data set.
-The higher the value the better the predictions of the model are.\\
-%\\
-%The CRPS is another value that measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
-%The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
-%$$
-%\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
-%$$
-%where \(\mathbf{1}\) is the indicator function. \\
-%In this simulation study the CRPS is evaluated out of sample on a data set that has not been used to build the model but has been generated by the same process and with the same settings as the original data set.
-In the following this second data set is called testing data set while the original data set on which the models are built on is called learning data set.
-%The mean CRPS over the whole testing data set is used for validation.
-As the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
-In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observation.
-\\
-
-%The methods are then compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by %the out-of-sample log-likelihood of the fitted models, averaged within each step.
-%(In that way it can be investigated how the results change for a changing value of $\kappa$.)
-
-%\SweaveOpts{eval = FALSE}
-
-<<eval = FALSE, echo=FALSE, results=hide>>=
-if(file.exists("simres.rda")){
-  load("simres.rda")
-} else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/.R")
-  source("gensim.R")
-
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-
-  simfun <- function(x, kappa){
-    mu <- 0.2 + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * as.numeric(x[,1] < 0.4) + 
-      (7 * (exp(-(3*x[,1]-1.2)^(2*kappa))) * (1/2) + 7/2) * as.numeric(x[,1]>= 0.4) +
-      (3 * (exp(-(3*x[,1]-1.2)^(2*kappa)))) * x[,4]
-    sigma <- 0.5 + 
-      mu/5 + 
-      #(x[,1]>0.4)  #*x[,4]#+
-      2*(x[,1] < -0.4) 
-    #plogis((kappa^(1.8)) * 4 * (x[,2] + 0.7)) #* x[,4]
-    #4*(1-plogis((kappa^(1.8)) * 2 * (x[,1] + 0.5))) * x[,4]
-    #5 * abs(x[,1]) * x[,4]
-    par <- cbind(mu, sigma)
-    return(par)
-  }
-
-  simres <- gensim(parfun = simfun,
-                   seedconst = 74, nrep = 100, ntree = 100,
-                   nsteps = 10, stepsize = 1,
-                   formula = y~x1+x2+x3+x4+x5+x6,
-                   nobs = 500, testnobs = 500L,
-                   tree_minsplit = 20, tree_minbucket = 7, tree_mincrit = 0.95,
-                   forest_minsplit = 20, forest_minbucket = 7, forest_mincrit = 0, 
-                   forest_mtry = 3,
-                   censNO = TRUE,
-                   gamboost_cvr = FALSE,
-                   eval_disttree = TRUE,
-                   eval_distforest = TRUE,
-                   eval_gamlss = TRUE,
-                   eval_bamlss = FALSE,
-                   eval_gamboostLSS = FALSE,
-                   eval_randomForest = FALSE,
-                   eval_cforest = FALSE,
-                   nrunif = 3,
-                   nrcovform = 6)
-
-  save(simres, file = "simres.rda")
-}
-@
-
 \SweaveOpts{eval = TRUE}
@@
 
-As explained in the introduction, distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree-structured models which provide features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
-To show their strengths in practice, these new methods are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
+As explained in the introduction, distributional forests provide a new approach in probabilistic forecasting by embedding the ideas of flexible parametric modeling into the framework of tree-structured models.
+In that way this novel methodology offers additional features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
+To show their strengths in practice, distributional forests are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
+Their results are compared to those of other zero-censored Gaussian models including a standard ensemble model output approach and two GAMLSS models.
+For one of the GAMLSS models the regressors have been selected and interactions have been specified in advance based on subject-matter knowledge and experience from meteorology.
+The other one is based on a computationally more demanding boosting approach.
 
@@
 \label{data}
-The data set used for training and validation consists of observed daily precipitation sums provided by the National Hydrographical Service (\citealt{ehyd}) and numerical weather forecasts from the {U.S.} National Oceanic and Atmospheric Administration (NOAA).
+The data set used for training and validation consists of observed precipitation sums provided by the National Hydrographical Service (\citealt{ehyd}) and numerical weather forecasts from the {U.S.} National Oceanic and Atmospheric Administration (NOAA).
 Both, observations and weather forecasts are available from 1985 trough the year 2012, focusing on July only in this article.
 
-Observations are available on a daily basis, each observation representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and its close surrounding.
+Observations are available on a daily basis, each representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and its close surrounding.
 The observations are undergoing a rigorous quality check before made available to the public.
@@
 In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates.
-Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std.
+Since an ensemble of individual predictions is provided for each forecast variable, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std.
 deviation, ensemble minimum and ensemble maximum.
-All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
+All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.
 Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std.
@@
 \subsection{Models and evaluation}
-Next to distributional forests three other methods are applied to this data set and their results are compared based on the continuous ranked probability score (CRPS).
-One of the reference models is an EMOS model as this is a common choice in this type of setting and research field.
-The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameter and once applying a boosting algorithm.
-The chosen methods and their specifications are listed in Table~\ref{model_specification}.
+
+Next to distributional forests three other methods are applied to this data set:
+
+\begin{itemize}
+
+\item An ensemble model output statistics (EMOS, \citealp{gneiting2005calibrated}) model where the location parameter and the logarithm of the scale parameter depend linearly on the ensemble mean forecast of the total daily precipitation sum and the corresponding ensemble standard deviation respectively.
+The choice of the logarithmic function as link function for the scale parameter has been made based on the results of \citealt{Gebetsberger2017} where different link functions have been compared.
+Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function led to the best performance of the EMOS model in this application.
+
+\item A GAMLSS where for each distribution parameter the selection of covariates and the specification of interactions is done in advance, both based on meteorological expert knowledge following \cite{Stauffer+Umlauf+Messner:2017} (prespecified GAM).
+Among the available 82 regressors, 9 where selected for the location parameter and 6 for the scale parameter.
+Additionally, for each distribution parameter one interaction term is specified in which convective availbale energy is interacting with total precipitation (with the ensemble mean for the location parameter and with the ensemble standard deviation for the scale parameter).
+%In this terms an ensemble mean of 6 hour sums of convective available energy is interacting with an ensemble mean or standarad deviation of 6 hour sums of total precipitation.  
+These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation.
+By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information (and is therefore expected to profit from this advantage).
+\item A boosted GAMLSS \citep{gamboostLSS2016} using non-homogenous boosting
+\citep{Messner+Mayr+Zeileis:2017} for an automatic selection of covariates (boosted GAM).
+The idea of this (boosting) approach is to iteratively minimize a loss function where in each step only the coefficient leading to the highest improvement of the current model is updated.
+As the initial coefficients are all zero, only the most important covariates are included with nonzero coefficients while less relevant variables are ignored if the procedure is stopped before convergence.
+\end{itemize}
+
+%\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAM. Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}
+
+%\fixme{A description of the boosting approach is currently missing completely.}
+
+The selected methods and their specifications are listed in Table~\ref{model_specification}.
+
 \begin{table}[t!]
@@
 
-For the prespecified GAM model the covariates were selected manually based on meteorological knowledge and experience.
-Additionally, interactions are specified in advance, once between \emph{tp{\_}mean1218} and \emph{cape{\_}mean1218} for the location parameter and once between \emph{tp{\_}sprd1218} and \emph{cape{\_}mean1218} for the scale parameter.
-These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation.
-By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.
-
-\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAM. Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}
-
-\fixme{A description of the boosting approach is currently missing completely.}
-
-In the EMOS model the logarithmic function is chosen as the link function for the scale parameter.
-This choice has been made based on the results of \citealt{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared.
-Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.
-
 In order to compare the performance of the chosen methods, the continuous ranked probability score (CRPS) is calculated for each model. 
-The CRPS measures the accuracy of a probabilistic model.
-For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+\fixme{Adding reference for CPRS}
+
+The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
 $$
@@
 
-In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model.
-The mean CRPS over this data set is used for validation.
+In this application the CRPS is evaluated out of sample, e.g.
+on a part of the data set that has not been used to build the model.
+The mean CRPS over this subset of the data is used for validation.
 Since the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
-In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observation.
-
-For the purpose of comparing the performance of a distributional forest to those of other models the so-called skill score can be calculated with respect to the CRPS and one model being chosen as a reference.
-In that way we obtain a measure of how much a method improves the model performance in terms of CRPS compared to the reference method.
-$$
-\text{SkillsScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
+In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observations.
+
+For the purpose of comparing the models' performances with respect to a reference model the so-called skill score of the CRPS can be calculated.
+In that way a measure of how much a method improves the model fit in terms of CRPS compared to the reference method is obtained.
+$$
+\text{SkillScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
 $$
@@
 This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol.
-As for all other stations, daily observations of the month of July of 28 years are available from 1985 through the year 2012.
-
-The results of fitting one distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast.
+As for all other stations, daily observations and numerical weather predictions are available for the month of July of 28 years from 1985 through the year 2012.
+In Figure~\ref{axams_onestation} the results of fitting a distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown as a motivational example for a probabilistic forecast.
 Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots.
@@
 As shown in Table~\ref{model_specification} the distributional forest includes all 82~covariates.
-A single tree model allows to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable.
-In a distributional forest though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree.
+A single tree model would allow to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable.
+In a forest model though, for each node in each of the trees a subset of these covariates is chosen randomly as possible split variables to select one from.
+Moreover, each tree is built on a resampled version of the learning data.
+For these reasons the choices of split variables vary over the trees of a forest making it difficult to analyze the impact of covariates just by investigating the different trees.
 To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied.
-Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}.
+Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 yearz are listed in Figure~\ref{varimp}.
 As expected, the numerical forecast for total precipitation is the most influential variable.
@@
 This is done by calculating the CRPS value for each of the observations in the testing data set and then averaging them.
-Each of the 7 parts is selected as testing data once which leads to 7 different settings over which, again, the average values are stored.
+Each of the 7 parts is selected as testing data once which leads to 7 different settings of which again the average values are stored.
 This whole procedure is repeated 10 times.
@@
 \end{figure}
-
-The CRPS skill score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
+This CRPS skill score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
 While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
-
-From this result we can conclude that the distributional forest provides a powerful alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.
-
-For the prespecified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
-To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decisions.
-Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not correspond to the truth.
-This risk is avoided in the boosted GAM as all covariates are included and the selection of the influential covariates is done within a boosting framework.
-However, the trade off here is a remarkably higher computational effort.
-To build a distributional forest the user does not have to put any additional effort in it and therefore is not required to provide any expert knowledge or additional information about the setting.
-Variable selection is done automatically and interactions are detected by the tree structure of the model.
-In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the distributional forest model.
-Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
 
@@
 At each of them, the first 24 years are used as learning data set and predictions are made for the last 4 years of available observations.
-For these out-of-sample predictions the CRPS skill score with respect to the EMOS model as reference is calculated for each method at each station.
+For these out-of-sample predictions the CRPS skill score with the EMOS model as reference model is calculated for each method at each station.
 In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skill score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
@@
 
+From these result we can conclude that the distributional forest provides a powerful alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.
+
+For the prespecified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
+To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decisions.
+Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not correspond to the truth.
+This risk is avoided in the boosted GAM as all covariates are included and the selection of the influential covariates is done within a boosting framework.
+However, the trade off here is a remarkably higher computational effort.
+To build a distributional forest the user does not have to put any additional effort in it and therefore is not required to provide any expert knowledge or additional information about the setting.
+Variable selection is done automatically and interactions are detected by the tree structure of the model.
+In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the distributional forest model.
+Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
+
 \section{Discussion}

### 2018-02-18 pkg/disttree/inst/draft/draft.Rnw
revised parts of the application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2029 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 77d2a192a80e3385189f9dd53e5109389a89a824

@@
 \begin{itemize}
-  \item \emph{Distributional trees} split the data into more homogenous groups with
+  \item \emph{Distributional trees} split the data into more homogeneous groups with
     respect to a parametric distribution, thus capturing changes in any distribution
-    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
+    parameter like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
   \item \emph{Distributional forests} utilize an ensemble of distributional trees
@@
 \cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
-\citep{gamboostLSS2016} using non-homogenous boosting
+\citep{gamboostLSS2016} using non-homogeneous boosting
 \citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
@@
 However, the restriction to additive models remains.
-Therefore, a different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the next subsection of this article.
+Therefore, a different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the following.
 In particular, the covariates can be used to separate the data into more homogeneous subgroups.
@@
 
+%% TO DO: valueable features, why relevant in weather forecasting, what was missing here before? (complex terrain, many covariates, all kinds of effects could appear, interactions unknown)
+
+In weather forecasting, various factors can make it challenging to build a well fitting model.
+Complex physical processes in the atmosphere depend on possibly rapidly changing circumstances.
+Often it is hard to tell which of these circumstances have a relevant impact on the quantity of interest and what type of effect that can be or how they might interact.
+A profound basis of information can be obtained from numerical weather prediction (NWP) models as they provide predictions for a large number of meteorological quantities.
+However, while including too many variables in a model can lead to overfitting, excluding variables in advance can result in a loss of valuable information.
+Therefore, the task of selecting variables and specifying interactions forms the basis of building an adequate model.
+Moreover, this model has to be able to deal with a wide range of effects as abrupt changes can appear as well as smooth transitions.
+Especially if the considered region includes complex terrain, a large grid spacing of NWP models can lead to abrupt local effects within a single NWP grid cell.
+
 As explained in the introduction, distributional forests provide a new approach in probabilistic forecasting by embedding the ideas of flexible parametric modeling into the framework of tree-structured models.
-In that way this novel methodology offers additional features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings.
+The ability to deal with a wide variety of effects and an automatic variable selection and detection of interactions are features provided by distributional forests that could be particularly valuable in probabilistic (weather) forecasting in order to tackle the above mentioned challenges.
 To show their strengths in practice, distributional forests are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
@@
 \end{table}
+\fixme{dswrf\_mean/sprd\_min not included in formula but in the 82 covariates}
 
@@
 Among the available 82 regressors, 9 where selected for the location parameter and 6 for the scale parameter.
-Additionally, for each distribution parameter one interaction term is specified in which convective availbale energy is interacting with total precipitation (with the ensemble mean for the location parameter and with the ensemble standard deviation for the scale parameter).
+Additionally, for each distribution parameter one interaction term is specified in which convective available energy is interacting with total precipitation (with the ensemble mean for the location parameter and with the ensemble standard deviation for the scale parameter).
 %In this terms an ensemble mean of 6 hour sums of convective available energy is interacting with an ensemble mean or standarad deviation of 6 hour sums of total precipitation.  
@@
 By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information (and is therefore expected to profit from this advantage).
-\item A boosted GAMLSS \citep{gamboostLSS2016} using non-homogenous boosting
+\item A boosted GAMLSS \citep{gamboostLSS2016} using non-homogeneous boosting
 \citep{Messner+Mayr+Zeileis:2017} for an automatic selection of covariates (boosted GAM).
@@
 To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied.
-Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 yearz are listed in Figure~\ref{varimp}.
+Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 years are listed in Figure~\ref{varimp}.
 As expected, the numerical forecast for total precipitation is the most influential variable.

### 2018-02-19 pkg/disttree/inst/draft/draft.Rnw
changed format of references and revised parts of the application and methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2030 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fe018959bfd40c6c95a1abd42210653c4f28fc5d

@@
 In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting.
-In generalized additive models for location, scale, and shape %(GAMLSS, \citealt{stasinopoulos2005})
+In generalized additive models for location, scale, and shape %(GAMLSS, \citealt{Rigby+Stasinopoulos:2005})
 each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects.
-A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees %(\citealt{loh2011classification})
-or random forests. %(\citealt{breiman2001random}). 
+A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees %(\citealt{Loh:2011})
+or random forests. %(\citealt{Breiman:2001}). 
 %%%<TH> "random forest" generisch; Breiman and Cutler's randomForest </TH>
@@
   \item \emph{Generalized additive models for location, scale, and shape}
-    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
+    (GAMLSS, \citealt{Rigby+Stasinopoulos:2005}) adopting a probabilistic modeling
     approach. In GAMLSS, each parameter of a statistical distribution can depend
@@
 models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
-non-homogeneous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
+non-homogeneous Gaussian regression (NGR) of \cite{Gneiting+Raftery+Westveld:2005}.
 The NGR typically describes the mean of some meteorological response variable through
@@
 %% late: introduce MOS and EMOS? more details about precipitation forecasting
-%In ensemble model output statistics (EMOS, \cite{gneiting2005calibrated}) this idea is taken one step further.
+%In ensemble model output statistics (EMOS, \cite{Gneiting+Raftery+Westveld:2005}) this idea is taken one step further.
 Usually acting on the assumption of a normally distributed response variable this model allows for both parameters, the mean and the variance, to depend linearly on covariates provided by NWPs.
@@
 \begin{itemize}
-  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
+  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
     split the data into more homogeneous subgroups and can thus capture abrupt shifts
@@
     their interactions. 
-  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
+  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
     an ensemble of trees fitted to resampled versions of the learning data. This
@@
 three alternative zero-censored Gaussian models: a standard ensemble model
-output statistics approach (EMOS, \citealp{gneiting2005calibrated}), a GAMLSS
+output statistics approach (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}), a GAMLSS
 with regressors prespecified based on meteorological expertise following
 \cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
-\citep{gamboostLSS2016} using non-homogeneous boosting
+\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
 \citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
@@
 \subsection{Motivation}
-As illustrated in Figure~\ref{Devel_parmod} a lot of development has been made in the field of parametric modeling and a wide range of tools can be applied in research areas such as probabilistic forecasting.
-But as mentioned before, further features such as the ability to capture non-additive effects or interactions are desirable in many applications which demands for more flexible models.
-Since tree-structured models provide these required features a combination of them with parametric models can offer a framework that is able to deal with a wide range of different modeling situations.
-In this section we are going to introduce the novel methodology which embeds parametric modeling into the idea of tree-structured models step by step.
-We start with a simple distributional fit in Section~\ref{distfit}, then build a distributional tree in Section~\ref{disttree} and finally a distributional forest in Section~\ref{distforest}.
+As illustrated in Figure~\ref{Devel_parmod} a wide range of tools have been developed in the field of parametric modeling (which can be applied in research areas such as probabilistic forecasting).
+The shift from modeling only the expected value of the response to probabilistic modeling allows for much broader inference.
+In many applications further features such as the ability to capture non-additive effects or interactions are desirable which demands for more flexible models.
+Since tree-structured models provide these required features combining them with parametric models can offer a framework that is able to deal with a wide variety of different modeling situations.
+
+In this section the novel methodology which embeds parametric modeling into the idea of tree-structured models will be introduced step by step.
+Starting with a simple distributional fit in Section~\ref{distfit} will be followed by building a distributional tree in Section~\ref{disttree} and finally a distributional forest in Section~\ref{distforest}.
 
@@
 The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided.
-Using the GAMLSS framework (\citealt{stasinopoulos2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
+Using the GAMLSS framework (\citealt{Rigby+Stasinopoulos:2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
 This enables the model to deal with various features that can usually be very challenging in model fitting such as censoring, truncation or heavy tails.
@@
 Next to the censored normal distribution the censored logistic distribution is also often chosen for the modeling of precipitation amounts.
-In \citealt{Gebetsberger2017} this choice has been made in order to be able to deal with heavy tails which often appear in precipitation data.
+In \cite{Gebetsberger+Messner+Mayr:2017} this choice has been made in order to be able to deal with heavy tails which often appear in precipitation data.
 Another frequent application of probabilistic forecasting in meteorology is the modeling of near surface air temperature.
@@
 In order to avoid the restriction to linear effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ to specify a more flexible model $D(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure.
-Another approach to distributional regression is proposed by \citealt{Klein2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
+Another approach to distributional regression is proposed by \cite{Klein+Kneib+Lang:2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
 These include non-linear effects, spatial effects, random coefficients and interaction surfaces which extends the flexibility compared to parametric models such as GAMLSS modeling only smooth effects.
@@
 Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables.
-In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).
+In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{Breiman:2001}).
 
 Forest models differ mainly in the way how the trees are combined to make predictions.
-Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{Z}$ and the resulting subgroups and weighing schemes.
+Following the strategy explained and proposed by \cite{Hothorn+Zeileis:2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{Z}$ and the resulting subgroups and weighing schemes.
 In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.
-In transformation forests by \cite{transformationforests2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations.
-This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``bagging survival trees'' by \cite{hothorn2004}.
+In transformation forests by \cite{Hothorn+Zeileis:2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations.
+This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{Lin+Jeon:2006}, in ``quantile regression forests''  by  \cite{Meinshausen:2006} and in ``bagging survival trees'' by \cite{Hothorn+Lausen+Benner:2004}.
 The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
@@
 As explained in the introduction, distributional forests provide a new approach in probabilistic forecasting by embedding the ideas of flexible parametric modeling into the framework of tree-structured models.
-The ability to deal with a wide variety of effects and an automatic variable selection and detection of interactions are features provided by distributional forests that could be particularly valuable in probabilistic (weather) forecasting in order to tackle the above mentioned challenges.
+The ability to deal with a wide variety of effects and an automatic variable selection and detection of interactions are features provided by distributional forests that could be particularly valuable in probabilistic weather forecasting in order to tackle the above mentioned challenges.
 To show their strengths in practice, distributional forests are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
@@
 The observations are undergoing a rigorous quality check before made available to the public.
-As covariates the second generation reforecast data set of the global ensemble forecast system (GEFS) is used (\citealt{hamill2013}).
+As covariates the second generation reforecast data set of the global ensemble forecast system (GEFS) is used (\citealt{Hamill+Bates+Whitaker:2013}).
 This data set consists of an 11-member ensemble based on a fixed version of the numerical model and a horizontal grid-spacing of about $50~\times~50 \text{km}^2$ initialized daily at 0~UTC from December 1984 to present providing forecasts on 26 hourly temporal resolution.
@@
 To remove large parts of the skewness of precipitation data, a power transformation has been used frequently (Box and Cox 1964).
-In literature, cubic (\citealt{stidd1973}) or square root (\citealt{hutchinson1998b}) transformations have often been suggested but may vary for different climatic zones or temporal aggregation periods.
-In \citealt{stauffer2017a} optimization methods have been applied to find the best transformation factor which resulted in a value of $\frac{1}{1.6}$.
+In literature, cubic (\citealt{Stidd:1973}) or square root (\citealt{Hutchinson:1998}) transformations have often been suggested but may vary for different climatic zones or temporal aggregation periods.
+In \cite{Stauffer+Mayr+Messner:2017} optimization methods have been applied to find the best transformation factor which resulted in a value of $\frac{1}{1.6}$.
 Since the observation data set used for this optimization contains the observations considered in this article the same factor is chosen here for the power transformation of the response variable \emph{tp} (total precipitation).
@@
 
-\item An ensemble model output statistics (EMOS, \citealp{gneiting2005calibrated}) model where the location parameter and the logarithm of the scale parameter depend linearly on the ensemble mean forecast of the total daily precipitation sum and the corresponding ensemble standard deviation respectively.
-The choice of the logarithmic function as link function for the scale parameter has been made based on the results of \citealt{Gebetsberger2017} where different link functions have been compared.
+\item An ensemble model output statistics (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}) model where the location parameter and the logarithm of the scale parameter depend linearly on the ensemble mean forecast of the total daily precipitation sum and the corresponding ensemble standard deviation respectively.
+The choice of the logarithmic function as link function for the scale parameter has been made based on the results of \cite{Gebetsberger+Messner+Mayr:2017} where different link functions have been compared.
 Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function led to the best performance of the EMOS model in this application.
@@
 By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information (and is therefore expected to profit from this advantage).
-\item A boosted GAMLSS \citep{gamboostLSS2016} using non-homogeneous boosting
+\item A boosted GAMLSS \citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
 \citep{Messner+Mayr+Zeileis:2017} for an automatic selection of covariates (boosted GAM).
@@
 First of all we focus on one observation station, Axams in Tyrol.
-This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol.
+This station was selected due to its geographical closeness to Innsbruck, the capital of Tyrol.
 As for all other stations, daily observations and numerical weather predictions are available for the month of July of 28 years from 1985 through the year 2012.
 In Figure~\ref{axams_onestation} the results of fitting a distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots.
+Staying with this fitted model we are now going to analyze variable importance of the distributional forest model and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots.
 %%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>
@@
 As shown in Table~\ref{model_specification} the distributional forest includes all 82~covariates.
-A single tree model would allow to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable.
-In a forest model though, for each node in each of the trees a subset of these covariates is chosen randomly as possible split variables to select one from.
+A single tree model would allow to easily analyze which covariates are used as split variables in the nodes and hence have a significant impact on the response variable.
+In a forest model though, for each node in each of the trees a subset of the covariates is chosen randomly as possible split variables to select one from.
 Moreover, each tree is built on a resampled version of the learning data.
@@
 Overall, it can be stated that all four models provide well calibrated forecasts.
-In the appendix the corresponding probability integral transform (PIT) histograms (\citealt{gneiting2007}) can be found.
+This is also supported by the corresponding probability integral transform (PIT) histograms (\citealt{Gneiting+Balabdaoui+Raftery:2007}) in the appendix.
 
@@
 @
-\caption{\label{qqr_oosample} Out-of-sample residual QQ-plot (2009--2012) for station Axams based on models learned on data from 1985--2008.}
+\caption{\label{qqr_oosample} Out-of-sample residual QQ-plots (2009--2012) for station Axams based on models learned on data from 1985--2008.}
 \end{figure}
 
-After analyzing the results of the models on one data set only where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
-For that reason a cross validation framework is considered for the station Axams.
+After analyzing the results of the models on only one data set where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
+For that reason the four methods are evaluated in a cross validation framework for the station Axams.
 The data set over all 28 years is split into 7 subsets each consisting of 4 randomly selected years.
@@
 The resulting skill scores are illustrated in Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model represented by the horizontal line at height 0.
+This CRPS skill score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
+While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
 <<echo=FALSE, results=hide>>=
@@
 \end{figure}
-This CRPS skill score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
-While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
 
@@
 Until now the focus has been on one observation station only.
-To show that the presented results do not depend on the choice of this particular station we are now going to apply the methods on all 95 stations.
-At each of them, the first 24 years are used as learning data set and predictions are made for the last 4 years of available observations.
+To test and compare the four methods under different (geographical) circumstances we are now going to apply them on all 95 stations.
+At each of them, the first 24 years are used as learning data set and predictions are made for the ensuing 4 years of available observations.
 For these out-of-sample predictions the CRPS skill score with the EMOS model as reference model is calculated for each method at each station.
@@
 
-Figure~\ref{map} shows the location of all stations in the data set.
-Each observation station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station.
-The color of the symbol indicates the difference of the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
+Figure~\ref{map} shows a map of Tyrol with the location of all observation stations included in the data set.
+Each station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station.
+The color of the symbol indicates the difference between the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
+Even though for some stations, especially in East Tyrol, one of the two GAMs and for a few others the EMOS model leads to the best results, for the majority of stations the distributional forest performes best.
 
@@
 
-From these result we can conclude that the distributional forest provides a powerful alternative framework where only little or almost no effort is required to set up the model and which is competitive to existing methods that are commonly used for this kind of application or even outperforms them.
-
+From these results we can conclude that a distributional forest provides a powerful alternative framework requiring only little effort to set up a model which is competitive to commonly used methods or even outperforms them.
 For the prespecified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
@@
 In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the distributional forest model.
-Therefore, in comparing these methods one advantage of distributional forests is clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
+Therefore, in comparing these methods the strengths of distributional forests are clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
 
@@
 Similarly, the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a distributional forest.
-All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}).
+All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{Hothorn+Zeileis:2015}).
 
@@
 %\newpage
-\nocite{Messner2016, stasinopoulos2007generalized, Seibold2016, Seibold2017, yee2010vgam, gamboostLSS2016, srtm}
-% zeileis2008implementing, 
-% zeileis2010party, 
+\nocite{Messner+Mayr+Zeileis:2016, Stasinopoulos+Rigby:2007, Seibold+Zeileis+Hothorn:2016, Seibold+Zeileis+Hothorn:2017, Yee:2010, Hofner+Mayr+Schmid:2016, srtm}
+% Zeileis+Wiel+Hornik:2008, 
+% Zeileis+Hothorn+Hornik:2010, 
 % hothorn2015package,
@@
 
-With the conditional expectation $\mu_{j}$ and the covariance $\Sigma_{j}$ as derived by \cite{strasser1999} the test statistic can be standardized.
+With the conditional expectation $\mu_{j}$ and the covariance $\Sigma_{j}$ as derived by \cite{Strasser+Weber:1999} the test statistic can be standardized.
 For a numeric split variable this results in the Pearson correlation coefficient
@@
 
-This permutation test based tree algorithm is presented in \citealt{hothorn2006unbiased} as the ctree algorithm.
-A different framework to build a tree is provided by the MOB algorithm which is based on M-fluctuation tests (\citealt{zeileis2008model}).
+This permutation test based tree algorithm is presented in \cite{Hothorn+Hornik+Zeileis:2006} as the ctree algorithm.
+A different framework to build a tree is provided by the MOB algorithm which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).
 
@@
 \section{Probability integral transform (PIT) histograms}
-Probability integral transform (PIT) histograms \cite{gneiting2007} are a commonly used tool to assess the predictive power of distributional forecasting methods, particularly in the field of meteorological forecasting \\
+Probability integral transform (PIT) histograms \cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used tool to assess the predictive power of distributional forecasting methods, particularly in the field of meteorological forecasting \\
 Let $x$ be an observation with true distribution function $G$ and predicted distribution function $F$.

### 2018-02-19 pkg/disttree/inst/draft/rain_pred_24to4.R
changed format of references and revised parts of the application and methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2030 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fe018959bfd40c6c95a1abd42210653c4f28fc5d

@@ -28,139 +28,6 @@ rain_pred <- function(seedconst = 7, ntree = 100,
   #setwd("~/disttree/inst/draft/")
   
   
-  # dist_list_cens_normal
-  {
-    
-    dist_list_cens_normal <- list()
-    
-    parnames <- c("mu", "sigma")
-    etanames <- c("mu", "log(sigma)")
-    
-    ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
-      par <- c(eta[1], exp(eta[2]))
-      val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
-      if(sum) {
-        if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
-        val <- sum(weights * val, na.rm = TRUE)
-      }
-      return(val)
-    }
-    
-    
-    sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
-      par <- c(eta[1], exp(eta[2]))
-      # y[y==0] <- 1e-323
-      
-      score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-      score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
-      score <- cbind(score_m, score_s)
-      score <- as.matrix(score)
-      colnames(score) <- etanames
-      if(sum) {
-        if(is.null(weights)) weights <- rep.int(1, length(y)[1])
-        # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
-        score[score==Inf] = 1.7e308
-        score <- colSums(weights * score, na.rm = TRUE)
-        #if(any(is.nan(score))) print(c(eta, "y", y))
-      }
-      return(score)
-    }
-    
-    
-    hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
-      ny <- length(y)
-      if(is.null(weights)) weights <- rep.int(1, ny)
-      
-      par <- c(eta[1], exp(eta[2]))                           
-      # y[y==0] <- 1e-323
-      
[…]

### 2018-02-20 pkg/disttree/inst/draft/draft.Rnw
revised parts of the application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2031 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ca4ee8f26f125d9b558b9052119f254f506eea3c

@@
 %%%<TH> break-up very long sentence </TH>
-Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
+Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
 }
@@
 To show their strengths in practice, distributional forests are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
-Their results are compared to those of other zero-censored Gaussian models including a standard ensemble model output approach and two GAMLSS models.
-For one of the GAMLSS models the regressors have been selected and interactions have been specified in advance based on subject-matter knowledge and experience from meteorology.
+Their results are compared to those of other zero-censored Gaussian models including a standard ensemble model output approach and two GAMLSS.
+For one of the GAMLSS the regressors have been selected and interactions have been specified in advance based on subject-matter knowledge and experience from meteorology.
 The other one is based on a computationally more demanding boosting approach.
@@
 Altogether 82~covariates are available in this setting.
+\fixme{Mention that dswrf\_mean/sprd\_min is not included?}
 
@@
 \end{table}
-\fixme{dswrf\_mean/sprd\_min not included in formula but in the 82 covariates}
 
@@
 These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation.
-By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information (and is therefore expected to profit from this advantage).
+By putting extra effort in the specification of the formulas the prespecified GAM is provided with additional information and is therefore expected to profit from this advantage.
 \item A boosted GAMLSS \citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
 \citep{Messner+Mayr+Zeileis:2017} for an automatic selection of covariates (boosted GAM).
-The idea of this (boosting) approach is to iteratively minimize a loss function where in each step only the coefficient leading to the highest improvement of the current model is updated.
+The idea of this boosting approach is to iteratively minimize a loss function where in each step only the coefficient leading to the highest improvement of the current model is updated.
 As the initial coefficients are all zero, only the most important covariates are included with nonzero coefficients while less relevant variables are ignored if the procedure is stopped before convergence.
@@
 In Figure~\ref{axams_onestation} the results of fitting a distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of the distributional forest model and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots.
+Staying with this fitted model we are now going to analyze variable importance of the distributional forest model and compare its results to those of the GAMs and the EMOS model based on residual QQ-plots.
 %%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>
@@
 %%%<TH> always us the term "distributional forest"
-As shown in Table~\ref{model_specification} the distributional forest includes all 82~covariates.
+As shown in Table~\ref{model_specification} the distributional forest includes all of the available covariates.
 A single tree model would allow to easily analyze which covariates are used as split variables in the nodes and hence have a significant impact on the response variable.
@@
 
-In Figure~\ref{qqr_oosample} another impression of the quality of the predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models.
+In Figure~\ref{qqr_oosample} an impression of the quality of predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models.
 For these plots 100 randomized quantile residuals are simulated per observation.
@@
 This is done by calculating the CRPS value for each of the observations in the testing data set and then averaging them.
-Each of the 7 parts is selected as testing data once which leads to 7 different settings of which again the average values are stored.
+Each of the 7 parts is selected once as testing data which leads to 7 different settings of which again the average values are stored.
 This whole procedure is repeated 10 times.
 The resulting skill scores are illustrated in Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model represented by the horizontal line at height 0.
-This CRPS skill score boxplot shows that, in this setting, both GAM models and the distributional forest perform better than the EMOS model.
-While the two GAM models lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
+This CRPS skill score boxplot shows that, in this setting, both GAMs and the distributional forest perform better than the EMOS model.
+While the two GAMs lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
 <<echo=FALSE, results=hide>>=
@@
 Until now the focus has been on one observation station only.
-To test and compare the four methods under different (geographical) circumstances we are now going to apply them on all 95 stations.
+To test and compare the four methods under different geographical circumstances we are now going to apply them on all 95 stations.
 At each of them, the first 24 years are used as learning data set and predictions are made for the ensuing 4 years of available observations.
@@
 The color of the symbol indicates the difference between the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
-Even though for some stations, especially in East Tyrol, one of the two GAMs and for a few others the EMOS model leads to the best results, for the majority of stations the distributional forest performes best.
+Even though for some stations, especially in East Tyrol, one of the two GAMs and for a few others the EMOS model leads to the best results, for the majority of stations the distributional forest performs best.
 
@@
 From these results we can conclude that a distributional forest provides a powerful alternative framework requiring only little effort to set up a model which is competitive to commonly used methods or even outperforms them.
-For the prespecified GAM model the covariates needed to be selected and interactions needed to be defined in advance.
-To do so, meteorological experience and knowledge is required and a lot of thought has to be put in these decisions.
-Moreover, making such decisions is always equivalent to influencing the resulting model and comes along with the risk of having missed important facts or even imposing certain effects on the model which might not correspond to the truth.
-This risk is avoided in the boosted GAM as all covariates are included and the selection of the influential covariates is done within a boosting framework.
-However, the trade off here is a remarkably higher computational effort.
-To build a distributional forest the user does not have to put any additional effort in it and therefore is not required to provide any expert knowledge or additional information about the setting.
-Variable selection is done automatically and interactions are detected by the tree structure of the model.
-In this particular setting, a very high number of covariates is available which intensifies the impact of the above mentioned strength of the distributional forest model.
-Therefore, in comparing these methods the strengths of distributional forests are clearly outlined as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an improvement in the predictive performance.
+Even though a proper selection of covariates and specification of interactions in advance can lead to better fitting (and probably less complex) models, this presupposes profound knowledge of all influencing factors.
+Otherwise, important aspects might be missed or effects which do not correspond to the truth might be imposed on the model.
+This risk can be avoided by applying a boosting algorithm for an automatic variable selection.
+In this case the trade-off is often a remarkably higher computational effort.
+To build a distributional forest the user is not required to provide any expert knowledge or additional information about the setting as variable selection and detection of interactions are done automatically by the tree-structure of the model.
+This feature is particularly advantageous in case of a very high number of covariates.
+Therefore, the above discussed application illustrates the strengths of distributional forests as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an overall improvement in the predictive performance in this setting.
 

### 2018-02-21 pkg/disttree/inst/draft/draft.Rnw
semi-transparent qqrplot als PNG not PDF to save some space
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2032 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 94fb2867fd31d37aab65684837f91540d8ad028f

@@
 \setkeys{Gin}{width=0.99\textwidth}
-<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
+<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
 # out of sample
@@
 par(mfrow = c(2,2))
-qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
-
-qqrplot(pit_ml_t, nsim = 100, main = "Boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
+
+qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAM", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAM", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
 @

### 2018-02-21 pkg/disttree/inst/draft/draft.Rnw
revised parts of the application and methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2033 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a93165a03225f5a85c9c0c4fb0a7ff3dd3724a65

@@
 \subsection{Motivation}
-As illustrated in Figure~\ref{Devel_parmod} a wide range of tools have been developed in the field of parametric modeling (which can be applied in research areas such as probabilistic forecasting).
-The shift from modeling only the expected value of the response to probabilistic modeling allows for much broader inference.
-In many applications further features such as the ability to capture non-additive effects or interactions are desirable which demands for more flexible models.
-Since tree-structured models provide these required features combining them with parametric models can offer a framework that is able to deal with a wide variety of different modeling situations.
-
+As illustrated in Figure~\ref{Devel_parmod} a wide range of tools has been developed in the field of parametric modeling. 
+%The shift from conditional mean regression models to distributional regression models allows for much broader inference.
+In many applications further features such as the ability to capture non-additive effects or interactions are desirable which demands for more flexibility.
+Especially the shift from conditional mean regression models to distributional regression models has increased the application spectrum of these models as it allows for much broader inference.
+In some cases further features such as the ability to capture non-additive effects or interactions are desirable which demands for more flexibility.
+Since tree-structured models provide these required features, combining them with parametric models can offer a framework that is able to deal with a wide variety of different modeling situations.
 In this section the novel methodology which embeds parametric modeling into the idea of tree-structured models will be introduced step by step.
@@
 The goal is to fit a distributional model $D(Y, \theta)$ to a response variable $Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters $\theta \in \Theta$ in case of a \textit{k}-parametric distribution.
-The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided.
+The distribution family with density function $f$ has to be specified in advance such that a log-likelihood function $\ell(\theta; Y) = \log(f(Y; \theta)$ is provided.
 Using the GAMLSS framework (\citealt{Rigby+Stasinopoulos:2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
-This enables the model to deal with various features that can usually be very challenging in model fitting such as censoring, truncation or heavy tails.
+This enables the model to deal with various features that can make the task of fitting a model very challenging such as censoring, truncation or heavy tails.
 
@@
 
-However, the restriction to additive models remains.
-Therefore, a different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the following.
+A different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the following.
 In particular, the covariates can be used to separate the data into more homogeneous subgroups.
@@
 These values express how well the model fits the data regarding each parameter.
-Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.
+Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimation, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.
 This enables the model to be sensitive to changes in each of the parameters which is one of the main advantages of distributional trees over other tree models considering only the mean of the distribution.
@@
 If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest dependency.
-
-In a next step a local distributional model is fit in each of the subgroups.
+The exact way of determining how and where to split varies across different tree algorithms.
+For distributional trees one possible choice is a permutation test based algorithm which is explained in detail in the appendix.
+After splitting the data the next step is to fit a local distributional model in each of the subgroups.
 Repeating this procedure of fitting a model, applying statistical tests to evaluate parameter instability and splitting the data set depending on the test results leads to a tree structure.
@@
 
-The exact way of determining how and where to split varies across different tree algorithms.
-For distributional trees one possible choice is a permutation test based algorithm which is explained in detail in the appendix.
 Summed up, the steps of building a distributional tree are:
@@
 \item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(y_i, \hat{\theta}) = 0$.
-\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
+\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable~$Z_l$.
 %\item Assess whether the \emph{model scores} are associated with
@@
 
-From these results we can conclude that a distributional forest provides a powerful alternative framework requiring only little effort to set up a model which is competitive to commonly used methods or even outperforms them.
+\section{Discussion}
+
+From the results in Section~\ref{ProbabilisticForecasting} we can conclude that distributional forests provide a powerful alternative framework requiring only little effort to set up a model which is competitive to commonly used methods or even outperforms them.
 Even though a proper selection of covariates and specification of interactions in advance can lead to better fitting (and probably less complex) models, this presupposes profound knowledge of all influencing factors.
@@
 Therefore, the above discussed application illustrates the strengths of distributional forests as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an overall improvement in the predictive performance in this setting.
-
-\section{Discussion}
+Especially in situations where no or only little information is available about possible influences, effects or interactions this novel methodology can be seen as a good and recommendable compromise being able to deal with a wide range of modelling situations in a reasonable way.
 
@@
 %%% empirical results </TH>
-Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
+%Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
 In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.
@@
 \section{Tree algorithm}
+\label{treealgorithm}
 The tree algorithm used in the applications discussed in this paper is explained in the following.

### 2018-02-22 pkg/disttree/inst/draft/draft.Rnw
added references in the application and computational detail section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2034 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 37094590550942f9d36a0395f7dfd337ea39f181

@@
 
-Figure~\ref{map} shows a map of Tyrol with the location of all observation stations included in the data set.
+Figure~\ref{map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation Data, \cite{srtm}) with the location of all observation stations included in the data set.
 Each station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station.
@@
 All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{Hothorn+Zeileis:2015}).
+To specify a distribution family the user can hand over one of the following objects as input argument in \code{distfit}, \code{disttree} and \code{distforest}:
+\begin{itemize}
+\item a GAMLLS family object from the \textsf{R} package \textbf{gamlss.dist} (\citealt{Stasinopoulos+Rigby:2007}).
+\item a family list generating function from the package \textbf{disttree}.
+\item a complete family list from the package \textbf{disttree}.
+\item a custom list containing all required information about the distribution family.
+\end{itemize}
+
+In Section~\ref{ProbabilisticForecasting} the prespecified GAMs were built using the function \code{gamlss} from the \textsf{R} package \textbf{gamlss.dist} (\citealt{Stasinopoulos+Rigby:2007}).
+For the boosted GAMs the function \code{gamboostLSS} from the \textsf{R} package \textbf{gamboostLSS} (\citealt{Hofner+Mayr+Schmid:2016}) was applied.
+The EMOS models were built using the function \code{crch} from the \textsf{R} package \textbf{crch} (\citealt{Messner+Mayr+Zeileis:2016})
 
@@
 %\newpage
-\nocite{Messner+Mayr+Zeileis:2016, Stasinopoulos+Rigby:2007, Seibold+Zeileis+Hothorn:2016, Seibold+Zeileis+Hothorn:2017, Yee:2010, Hofner+Mayr+Schmid:2016, srtm}
+\nocite{Seibold+Zeileis+Hothorn:2016, Seibold+Zeileis+Hothorn:2017, Yee:2010}
 % Zeileis+Wiel+Hornik:2008, 
 % Zeileis+Hothorn+Hornik:2010,
-% hothorn2015package,
 

### 2018-02-22 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2035 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c352b3b2e7f55bd89de4deb979846abc4d4bbfe2

@@
 \subsection{Motivation}
-As illustrated in Figure~\ref{Devel_parmod} a wide range of tools has been developed in the field of parametric modeling. 
-%The shift from conditional mean regression models to distributional regression models allows for much broader inference.
-In many applications further features such as the ability to capture non-additive effects or interactions are desirable which demands for more flexibility.
-Especially the shift from conditional mean regression models to distributional regression models has increased the application spectrum of these models as it allows for much broader inference.
-In some cases further features such as the ability to capture non-additive effects or interactions are desirable which demands for more flexibility.
-Since tree-structured models provide these required features, combining them with parametric models can offer a framework that is able to deal with a wide variety of different modeling situations.
-In this section the novel methodology which embeds parametric modeling into the idea of tree-structured models will be introduced step by step.
-Starting with a simple distributional fit in Section~\ref{distfit} will be followed by building a distributional tree in Section~\ref{disttree} and finally a distributional forest in Section~\ref{distforest}.
+As illustrated in Figure~\ref{Devel_parmod} a wide range of tools has been developed 
+in the field of parametric modeling. Especially the shift from conditional mean 
+regression models to distributional regression models has increased the application 
+spectrum by allowing for much broader inference. In some cases further features 
+such as the ability to capture non-additive effects or interactions are desirable 
+which demands for more flexibility. Since tree-structured models provide these 
+required features, combining them with parametric models can offer a framework 
+that is able to deal with a wide variety of different modeling situations. 
+In this section the novel methodology which embeds parametric modeling into 
+the idea of tree-structured models will be introduced step by step. Starting 
+with a simple distributional fit in Section~\ref{distfit} will be followed by 
+building a distributional tree in Section~\ref{disttree} and finally a distributional 
+forest in Section~\ref{distforest}.
 
@@
 \label{distfit}
-Fitting a distributional model to a set of observations without considering any covariates is a well known procedure which can for example be done by applying the maximum likelihood method.
-This is also the chosen approach in the methodology described in this section.
+Fitting a distributional model to a set of observations without considering any 
+covariates is a well known procedure which can for example be done by applying 
+the maximum likelihood method. This is also the chosen approach in the methodology 
+described in this section. 
 %%% <TH> is is repeated on page 5, remove here </TH>
 
-The goal is to fit a distributional model $D(Y, \theta)$ to a response variable $Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters $\theta \in \Theta$ in case of a \textit{k}-parametric distribution.
-The distribution family with density function $f$ has to be specified in advance such that a log-likelihood function $\ell(\theta; Y) = \log(f(Y; \theta)$ is provided.
-Using the GAMLSS framework (\citealt{Rigby+Stasinopoulos:2005}) a wide range of distributions can be selected where each distribution parameter can be specified separately.
-This enables the model to deal with various features that can make the task of fitting a model very challenging such as censoring, truncation or heavy tails.
-
-In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left-censored with censoring point 0.
-(The power transformation is used to remove large parts of the skewness of the data as further explained in Section~\ref{data}.) Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is
+The goal is to fit a distributional model $D(Y, \theta)$ to a response variable 
+$Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters 
+$\theta \in \Theta$ in case of a \textit{k}-parametric distribution. The distribution 
+family with density function $f$ has to be specified in advance such that a 
+log-likelihood function $\ell(\theta; Y) = \log(f(Y; \theta)$ is provided. Using 
+the GAMLSS framework (\citealt{Rigby+Stasinopoulos:2005}) a wide range of distributions 
+can be selected where each distribution parameter can be specified separately. 
+This enables the model to deal with various features that can make the task of 
+fitting a model very challenging such as censoring, truncation or heavy tails.
+
+In our application of precipitation forecasting the response variable, which is 
+the power transformed total precipitation amount within 24 hours, is expected 
+to be normally distributed with mean/location parameter $\mu$ and standard 
+deviation/scale parameter $\sigma$ and left-censored with censoring point 0. 
+(The power transformation is used to remove large parts of the skewness of 
+the data as further explained in Section~\ref{data}.) Therefore, the 
+corresponding log-likelihood function with parameter vector 
+$\theta = (\mu, \sigma)$ is
 $$
@@
 $$
-where $\phi$ and $\Phi$ are the probability density function and the distribution function of the standard normal distribution $\mathcal{N}(0,1)$.
-Next to the censored normal distribution the censored logistic distribution is also often chosen for the modeling of precipitation amounts.
-In \cite{Gebetsberger+Messner+Mayr:2017} this choice has been made in order to be able to deal with heavy tails which often appear in precipitation data.
-Another frequent application of probabilistic forecasting in meteorology is the modeling of near surface air temperature.
-In this case a normal distribution is often selected.
-
-With the specification of the distribution family and the log-likelihood function the task of fitting a distributional model turns into the task of determining the distribution parameter~$\theta$.
-This is done by the following maximization.
+where $\phi$ and $\Phi$ are the probability density function and the 
+distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
+Next to the censored normal distribution also the censored logistic 
+distribution is often chosen for the modeling of precipitation amounts. In 
+\cite{Gebetsberger+Messner+Mayr:2017} this choice has been made in order 
+to be able to deal with heavy tails which often appear in precipitation 
+data. Another frequent application of probabilistic forecasting in meteorology 
+is the modeling of near surface air temperature. In this case a normal 
+distribution is often selected.
+
+With the specification of the distribution family and its log-likelihood 
+function the task of fitting a distributional model turns into the task 
+of determining the distribution parameter~$\theta$. This is done by the 
+following maximization.
 $$
@@
 $$
-where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response variable $Y$.
-
-In that way, a whole distribution is specified with all its features, including location, scale, and shape.
-However, fitting one global model to the whole data set might be too much of a generalization and therefore not represent the data and its features in a reasonable way.
-Moreover, if covariates are available it is desirable to include them into the model as they might provide important additional information.
-In our application on precipitation data a wide range of covariates are available from a numerical ensemble prediction system considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest and have an influence on the predictions.
-
-A common way of including these covariates is to consider them as regressors in a distributional model such that each parameter $\theta$ depends on covariates $\bold{Z} = (Z_1,\ldots,Z_m)$.
-This could for example be done by a linear function $g$.
-In that way the fitted model would be specified as $D(Y, \theta = g(\bold{Z}))$.
-In order to avoid the restriction to linear effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ to specify a more flexible model $D(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure.
-Another approach to distributional regression is proposed by \cite{Klein+Kneib+Lang:2015} where each parameter of a possibly complex distribution can be modeled by an additive term consisting of a variety of different functional effect types.
-These include non-linear effects, spatial effects, random coefficients and interaction surfaces which extends the flexibility compared to parametric models such as GAMLSS modeling only smooth effects.
-
-A different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the following.
-In particular, the covariates can be used to separate the data into more homogeneous subgroups.
-Doing so after fitting a global model to the whole data set (without considering any covariates) and then fitting a local model to each subgroup should improve the model remarkably.
-This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.
+where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response 
+variable $Y$.
+
+In that way, a whole distribution is specified with all its features, 
+including location, scale, and shape. However, fitting one global model 
+to the whole data set might be too much of a generalization and therefore 
+not represent the data and its features in a reasonable way. Moreover, 
+if covariates are available it is desirable to include them into the model 
+as they might provide important additional information. In our application 
+on precipitation data a wide range of covariates is available from a numerical 
+ensemble prediction system considering precipitation, temperature, sea level 
+pressure and other meteorological quantities which might be of interest and 
+have an influence on the predictions.
+
+A common way of including these covariates is to consider them as regressors 
+in a distributional model such that each parameter $\theta$ depends on covariates 
+$\bold{Z} = (Z_1,\ldots,Z_m)$. This could for example be done by a linear function 
+$g$. In that way the fitted model would be specified as 
+$D(Y, \theta = g(\bold{Z}))$. In order to avoid the restriction to linear 
+effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ 
+to specify a more flexible model 
+$D(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure. 
+Another approach to distributional regression is proposed by 
+\cite{Klein+Kneib+Lang:2015} where each parameter of a possibly complex 
+distribution can be modeled by an additive term consisting of a variety of 
+different functional effect types. These include non-linear effects, spatial 
+effects, random coefficients and interaction surfaces which extends the 
+flexibility compared to parametric models such as GAMLSS modeling only 
+smooth effects.
+
+A different approach on how to include covariates that also allows for 
+non-additive effects and interactions is considered in the following. 
+In particular, the covariates can be used to separate the data into more 
+homogeneous subgroups. Doing so after fitting a global model to the whole 
+data set (without considering any covariates) and then fitting a local 
+model to each subgroup should improve the model remarkably. This 
+procedure of splitting a data set into subgroups can be done by 
+applying a tree algorithm.
 
@@
 \label{disttree}
-Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ %(from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors)
-%%% <TH> X -> Z is confusing. Either we always write theta = g(X) where
-%%% g is now a tree or we explain that theta = g(X)(Z) is what we want.
-%%% Here, we don't partition models with X variables, so we should stick to X. </TH>
-to decide whether the data set should be split into subgroups or not. In particular, the information provided by the fitted model are the scores 
+Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ 
+the idea is to use the information provided by this model together with a 
+set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ 
+to decide whether the data set should be split into subgroups or not. 
+In particular, the information provided by the fitted model are the scores 
 $$
@@
 $$ 
-These values express how well the model fits the data regarding each parameter.
-Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimation, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.
-This enables the model to be sensitive to changes in each of the parameters which is one of the main advantages of distributional trees over other tree models considering only the mean of the distribution.
-
-Statistical tests are applied to check whether there is a significant dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$.
-If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest dependency.
-The exact way of determining how and where to split varies across different tree algorithms.
-For distributional trees one possible choice is a permutation test based algorithm which is explained in detail in the appendix.
-After splitting the data the next step is to fit a local distributional model in each of the subgroups.
-Repeating this procedure of fitting a model, applying statistical tests to evaluate parameter instability and splitting the data set depending on the test results leads to a tree structure.
-In that way the learning data set is separated into \textit{B} disjoint segments such that
+These values express how well the model fits the data regarding each 
+parameter. Ideally they should fluctuate randomly around zero, 
+similar to residuals in ordinary least squares estimation, but in 
+this case there is a score value for each pair of estimated parameter 
+$\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ 
+and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$. This enables the 
+model to be sensitive to changes in each of the parameters which is one 
+of the main advantages of distributional trees over other tree models 
+considering only the mean of the distribution.
+
+Statistical tests are applied to check whether there is a significant 
+dependency between the scores and each of the partitioning variables 
+$Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$. If so, the data set is split 
+into subgroups based on the values of the partitioning variable that 
+shows the highest dependency. The exact way of determining how and where 
+to split varies across different tree algorithms. For distributional trees 
+one possible choice is a permutation test based algorithm which is explained 
+in detail in the appendix.
+
+After splitting the data the next step is to 
+fit a local distributional model in each of the subgroups. Repeating this 
+procedure of fitting a model, applying statistical tests to evaluate parameter 
+instability and splitting the data set depending on the test results leads 
+to a tree structure. In that way the learning data set is separated into 
+\textit{B} disjoint segments such that
 $$
@@
 \begin{enumerate}
-\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(y_i, \hat{\theta}) = 0$.
-\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable~$Z_l$.
-%\item Assess whether the \emph{model scores} are associated with
-%    (or change along) any of the available covariates -- e.g.,
-%    using parameter instability tests (\emph{strucchange}) or
-%    conditional inference (\emph{coin}).
-\item Split the sample along the partitioning variable with the strongest association or instability.
-    Choose the breakpoint with the highest improvement in log-likelihood.
-\item Repeat steps 2--4 recursively in the subgroups
-    until some stopping criterion is reached.
-    %-- e.g., for significance or sample size.
-%\item Choose the variable with the strongest association
-%\item Choose the split point which leads to the highest improvement of the model
-%\item Split and repeat 2-6 in each node until a stopping criterion is met
+\item Specify a distribution family with log-likelihood function $\ell(\theta; y)$.
+\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(\hat{\theta}, y_i) = 0$.
+\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ 
+and each partitioning variable~$Z_l$.
+\item Split the sample along the partitioning variable with the strongest association 
+or instability. Choose the breakpoint with the highest improvement in log-likelihood.
+\item Repeat steps 2--4 recursively in the subgroups until some stopping criterion is reached.
+
 \end{enumerate}
-The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.
+The stopping criterion in step 5 can for example be a minimum number of 
+observations in a node or a p-value for the statistical tests applied in step 3.
 %%% <TH> we need a little more detail here or a reference to a better
@@
 
-Executing these steps results in a stepwise parametric model where a complete distribution is fit in each node of the tree.
-
-For distributional trees a prediction for a new set of covariates $\bold{z} = (z_1, \ldots, z_m)$ is made with the following strategy: 
-All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new set of covariates are selected and a distributional model is fit on the subset.
-To select this subset a weight vector $\bold{w}^{\text{tree}} = (w^{\text{tree}}_1, \ldots, w^{\text{tree}}_n)$ is included in the fitting process.
-$$
-w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b))
-$$
-The predicted distribution of the new set of covariates is then fully specified by the estimated parameter $\hat{\theta}(\bold{z})$ where
+Executing these steps results in a stepwise parametric model where a complete 
+distribution is fit in each node of the tree.
+
+For distributional trees a prediction for a (possibly new) set of covariates 
+$\bold{z} = (z_1, \ldots, z_m)$ is made with the following strategy: 
+All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ 
+which end up in the same terminal node of the tree as $\bold{z}$ are selected 
+and a distributional model is fit on this subset. Selecting these observations
+is done by including a weight vector 
+$\bold{w}^{\text{tree}} = (w^{\text{tree}}_1, \ldots, w^{\text{tree}}_n)$ 
+in the fitting process where
+$$
+w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)).
+$$
+The predicted distribution for $\bold{z}$ is then fully 
+specified by the estimated parameter $\hat{\theta}(\bold{z})$ where
 $$
@@
 
-While the clear structure of such a tree model is advantageous in interpreting and analyzing it, the abrupt changes are often too rough and impose steps on the model even if the true underlying effect is smooth.
-In these situations bagging can solve the problem by combining an ensemble of trees such that wrongly detected abrupt changes of the model are turned into almost smooth transitions.
-In that way, all kinds of effects can be modeled, which is is also one the of the main advantages of forest models (taking bagging one step further as the variance is reduced by sampling the covariates considered for the splitting process in each node).
+While the clear structure of such a tree model is advantageous in 
+interpreting and analyzing it, the abrupt changes are often too rough 
+and impose steps on the model even if the true underlying effect is 
+smooth. In these situations bagging can solve the problem by combining 
+an ensemble of trees such that wrongly detected abrupt changes of the 
+model are turned into almost smooth transitions. In that way, all kinds 
+of effects can be modeled, which is also one the of the main advantages 
+of forest models taking bagging one step further as the variance is 
+reduced by sampling the covariates considered for the splitting process 
+in each node.
 
@@
 %%% random forest. Maybe it is better to not mention this method at any time. </TH>
-Common forest models such as Breiman and Cutler's randomForest consist of trees only considering the mean of the distribution of the response variable.
-For that reason, changes in other parameters can only be detected if they are connected to changes in the location parameter, e.g.
-if they depend on the location parameter.
-Therefore, building a forest of distributional trees which can also capture separate effects in each of the distribution parameters leads to an improvement of the model.
-This valuable feature of detecting separate effects is also provided by GAMLSS as each parameter is modeled separately.
-However, in this case this separation can also be a disadvantage as it hinders the capturing of interactions unless they are specified explicitly by the user.
-These two cases provide once again a good reason to combine the two frameworks of flexible parametric modeling and tree-structured models in order to profit from their advantages and compensate possible disadvantages.
-This is the pursued goal in introducing distributional forests.
+Common forest models (such as Breiman and Cutler's randomForest) consist of 
+trees only considering the mean of the distribution of the response variable. 
+For that reason, changes in other parameters can only be detected if they 
+are connected to changes in the location parameter, e.g. if they depend on 
+the location parameter. As explained before, a distributional tree considers
+scores for all distribution parameters. Therefore, a forest consisting of 
+distributional trees can also capture separate effects in each of the 
+distribution parameters. For that reason distributional forests are
+now going to be introduced.
 
@@
 A distributional forest consists of an ensemble of $T$ distributional trees. 
-Each of these trees is built on a different data set which can either be a bootstrap or a subset of the original data set.
-Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables.
-In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{Breiman:2001}).
+Each of these trees is built on a different data set which can either be a bootstrap 
+or a subset of the original data set. Moreover, in each node a subset of the 
+partitioning variables is chosen randomly as candidates for splitting variables. 
+In that way the correlation among the resulting trees is reduced and therefore also 
+the variance of the model (\citealt{Breiman:2001}).
 
 Forest models differ mainly in the way how the trees are combined to make predictions. 
-Following the strategy explained and proposed by \cite{Hothorn+Zeileis:2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{Z}$ and the resulting subgroups and weighing schemes.
-In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.
-In transformation forests by \cite{Hothorn+Zeileis:2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations.
-This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{Lin+Jeon:2006}, in ``quantile regression forests''  by  \cite{Meinshausen:2006} and in ``bagging survival trees'' by \cite{Hothorn+Lausen+Benner:2004}.
-The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
-%Similar observations have a high probability of ending up in the same terminal node whereas this probability is low for quite different observations. 
-%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (bagging survival trees, Hothorn et al. 2004). 
-In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 
-$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_i,\bold{z}_i)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_i,\bold{z}_i)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
+Following the strategy explained and proposed by \cite{Hothorn+Zeileis:2017} we 
+interpret random forests as adaptive local likelihood estimators as distributional 
+models are fit locally in subgroups while the term ``adaptive'' refers to the 
+dependence on the splitting variables $\bold{Z}$ and the resulting subgroups 
+and weighing schemes. In this case the forest itself is only used to evaluate 
+neighborhood relationships among the observations which can then be included 
+in the fitting process. In transformation forests by \cite{Hothorn+Zeileis:2017} 
+so-called ``nearest neighbor weights'' are calculated as a measure of distance 
+between observations. This approach has also already been applied in several 
+random forest-type methods such as in ``adaptive nearest neighbors'' by 
+\cite{Lin+Jeon:2006}, in ``quantile regression forests''  by  \cite{Meinshausen:2006} 
+and in ``bagging survival trees'' by \cite{Hothorn+Lausen+Benner:2004}. 
+The idea of nearest neighbor weights is to measure similarity of a pair of 
+observations by counting the number of trees in which it is asigned to the 
+same terminal node.
+%Similar observations have a high probability of ending up in the same terminal 
+%node whereas this probability is low for quite different observations. 
+%although the first random forest-type algorithm for the estimation of 
+%conditional distribution functions was published more than a decade ago 
+%(bagging survival trees, Hothorn et al. 2004). 
+In particular, for a (possibly new) set of covariates 
+$\bold{z} = (z_1, \ldots, z_m) $  such a set of weights 
+$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ 
+is obtained by the following procedure: For each tuple $(y_i,\bold{z}_i)$ in 
+the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated 
+by counting the number of trees in which $(y_i,\bold{z}_i)$ ends up in the same 
+terminal node as $\bold{z}$:
 $$
@@
 $$
-where $T$ is the number of trees of the forest, $B^t$ is the number of terminal nodes of the \textit{t}-th tree and $\mathcal{B}^t_b$ represents the \textit{b}-th terminal node of the \textit{t}-th tree.
-
-Therefore, contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates.
-Compared to the formula for predictions made by distributional trees the vector of tree weights $\bold{w}^{\text{tree}}$ containing only zeros and ones is now replaced by a vector of integer valued forest weights.
+where $T$ is the number of trees of the forest, $B^t$ is the number of terminal 
+nodes of the \textit{t}-th tree and $\mathcal{B}^t_b$ represents the \textit{b}-th 
+terminal node of the \textit{t}-th tree.
+
+Therefore, contrary to distributional trees, in distributional forests the whole 
+learning data set can be used to fit a distributional model and specify the 
+predicted distribution for a set of covariates. Compared to the formula for
+predictions made by distributional trees the vector of tree weights 
+$\bold{w}^{\text{tree}}$ containing only zeros and ones is now replaced by a 
+vector of integer valued forest weights.
 %%% <TH> we need references for the nearest neighbor weights, copy from
@@
 %%% random forests as adaptive local ML </TH>
-This individual set of weights $\bold{w}^{\text{forest}}(\bold{z}) = (w^{\text{forest}}_1(\bold{z}), \ldots, w^{\text{forest}}_n(\bold{z}))$ for the new set of covariates $\bold{z}$ can now be included in the estimation process of the distribution parameters and in that way leads to its individual parameter vector $\hat{\theta}(\bold{z})$.
+This individual set of weights 
+$\bold{w}^{\text{forest}}(\bold{z}) = (w^{\text{forest}}_1(\bold{z}), \ldots, w^{\text{forest}}_n(\bold{z}))$ 
+for the set of covariates $\bold{z}$ can now be included in the 
+estimation process of the distribution parameters and in that way leads 
+to its individual parameter vector $\hat{\theta}(\bold{z})$ where
 $$
@@
 $$
-While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates due to the individual weights calculated separately for each set of covariates.
-
-As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree-structured modeling it provides a model that can deal with a wide range of situations.
-Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably well by combining various tree and smoothing the steps imposed by a single tree.
-As scores are considered for each distribution parameter (separate) effects in different parameters can be captured.
-At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.
+While in a distributional tree the same distribution is specified for all 
+sets of covariates being in the same node, a distributional forest predicts 
+and fits a different distribution for each different set of covariates due 
+to the individual weights calculated separately for each set of covariates.
+
+As this novel methodology combines the advantages of smooth and flexible 
+parametric modeling with those of tree-structured modeling it provides a 
+model that can deal with a wide range of situations. Abrupt changes and 
+non-additive effects can be captured by its tree structure while smooth 
+effects can be modeled reasonably well by combining various tree and 
+smoothing the steps imposed by a single tree. As scores are considered 
+for each distribution parameter (separate) effects in different parameters 
+can be captured. At the same time, also interactions are detected 
+automatically by splitting the data in a way that is sensitive to changes 
+in either each of the parameters or combined in various parameters.
 
@@
 
-%% TO DO: valueable features, why relevant in weather forecasting, what was missing here before? (complex terrain, many covariates, all kinds of effects could appear, interactions unknown)
-
-In weather forecasting, various factors can make it challenging to build a well fitting model.
-Complex physical processes in the atmosphere depend on possibly rapidly changing circumstances.
-Often it is hard to tell which of these circumstances have a relevant impact on the quantity of interest and what type of effect that can be or how they might interact.
-A profound basis of information can be obtained from numerical weather prediction (NWP) models as they provide predictions for a large number of meteorological quantities.
-However, while including too many variables in a model can lead to overfitting, excluding variables in advance can result in a loss of valuable information.
-Therefore, the task of selecting variables and specifying interactions forms the basis of building an adequate model.
-Moreover, this model has to be able to deal with a wide range of effects as abrupt changes can appear as well as smooth transitions.
-Especially if the considered region includes complex terrain, a large grid spacing of NWP models can lead to abrupt local effects within a single NWP grid cell.
-
-As explained in the introduction, distributional forests provide a new approach in probabilistic forecasting by embedding the ideas of flexible parametric modeling into the framework of tree-structured models.
-The ability to deal with a wide variety of effects and an automatic variable selection and detection of interactions are features provided by distributional forests that could be particularly valuable in probabilistic weather forecasting in order to tackle the above mentioned challenges.
-To show their strengths in practice, distributional forests are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities.
-Their results are compared to those of other zero-censored Gaussian models including a standard ensemble model output approach and two GAMLSS.
-For one of the GAMLSS the regressors have been selected and interactions have been specified in advance based on subject-matter knowledge and experience from meteorology.
-The other one is based on a computationally more demanding boosting approach.
+In weather forecasting, various factors can make it challenging to 
+build a well fitting model. Complex physical processes in the 
+atmosphere depend on possibly rapidly changing circumstances. 
+Often it is hard to tell which of these circumstances have a 
+relevant impact on the quantity of interest and what type of 
+effect that can be or how they might interact. A profound 
+basis of information can be obtained from numerical weather 
+prediction (NWP) models as they provide predictions for a 
+large number of meteorological quantities. However, while 
+including too many variables in a model can lead to overfitting, 
+excluding variables in advance can result in a loss of valuable 
+information. Therefore, the task of selecting variables and 
+specifying interactions forms the basis of building an adequate 
+model. Moreover, this model has to be able to deal with a wide 
+range of effects as abrupt changes can appear as well as smooth 
+transitions. Especially if the considered region includes complex 
+terrain, a large grid spacing of NWP models can lead to abrupt 
+local effects within a single NWP grid cell.
+
+As explained in the introduction, distributional forests provide 
+a new approach in probabilistic forecasting by embedding the ideas 
+of flexible parametric modeling into the framework of tree-structured 
+models. The ability to deal with a wide variety of effects and an 
+automatic variable selection and detection of interactions are 
+features provided by distributional forests that could be particularly 
+valuable in probabilistic weather forecasting in order to tackle 
+the above mentioned challenges.
+To show their strengths in practice, distributional forests are 
+now applied to probabilistic precipitation forecasts in a 
+mountainous region based on a large number of numerical weather 
+prediction quantities. Their results are compared to those of 
+other zero-censored Gaussian models including a standard ensemble 
+model output approach and two GAMLSS. For one of the GAMLSS the 
+regressors have been selected and interactions have been specified 
+in advance based on subject-matter knowledge and experience from 
+meteorology. The other one is based on a computationally more 
+demanding boosting approach.
 
@@
 \label{data}
-The data set used for training and validation consists of observed precipitation sums provided by the National Hydrographical Service (\citealt{ehyd}) and numerical weather forecasts from the {U.S.} National Oceanic and Atmospheric Administration (NOAA).
-Both, observations and weather forecasts are available from 1985 trough the year 2012, focusing on July only in this article.
-
-Observations are available on a daily basis, each representing the total precipitation sum fallen within 24~hours observed at 6~UTC at 95 stations all over Tyrol and its close surrounding.
-The observations are undergoing a rigorous quality check before made available to the public.
-As covariates the second generation reforecast data set of the global ensemble forecast system (GEFS) is used (\citealt{Hamill+Bates+Whitaker:2013}).
-This data set consists of an 11-member ensemble based on a fixed version of the numerical model and a horizontal grid-spacing of about $50~\times~50 \text{km}^2$ initialized daily at 0~UTC from December 1984 to present providing forecasts on 26 hourly temporal resolution.
-Each of the 11 ensemble members uses slightly different perturbed initial conditions trying to predict the situation specific uncertainty of the atmospheric state.
-
-In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates.
-Since an ensemble of individual predictions is provided for each forecast variable, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std.
-deviation, ensemble minimum and ensemble maximum.
-All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.
-Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std.
-deviation of mean/minimal/maximal values respectively.
-Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.
-Altogether 82~covariates are available in this setting.
+The data set used for training and validation consists of observed 
+precipitation sums provided by the National Hydrographical Service 
+(\citealt{ehyd}) and numerical weather forecasts from the {U.S.} 
+National Oceanic and Atmospheric Administration (NOAA).
+Both, observations and weather forecasts are available from 1985 
+trough the year 2012, focusing on July only in this article.
+
+Observations are available on a daily basis, each representing the 
+total precipitation sum fallen within 24~hours observed at 6~UTC 
+at 95 stations all over Tyrol and its close surrounding. The 
+observations are undergoing a rigorous quality check before made 
+available to the public.  As covariates the second generation 
+reforecast data set of the global ensemble forecast system (GEFS) 
+is used (\citealt{Hamill+Bates+Whitaker:2013}).  This data set 
+consists of an 11-member ensemble based on a fixed version of the 
+numerical model and a horizontal grid-spacing of about $50~\times~50 
+\text{km}^2$ initialized daily at 0~UTC from December 1984 to present 
+providing forecasts on 26 hourly temporal resolution. Each of the 
+11 ensemble members uses slightly different perturbed initial 
+conditions trying to predict the situation specific uncertainty 
+of the atmospheric state.
+
+In particular, 14 basic forecast variables are considered with up 
+to 12 variations of each of these variables being included as 
+covariates. Since an ensemble of individual predictions is provided 
+for each forecast variable, the possible variations include the mean, 
+the standard deviation, the minimum and the maximum over all ensemble 
+members, noted as ensemble mean, ensemble std. deviation, ensemble 
+minimum and ensemble maximum. All of these values can be calculated 
+on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour 
+time window. Another possible option is to consider the mean, minimum 
+or maximum over 24 hours for each individual prediction and then take 
+the ensemble mean or ensemble standard deviation, noted as ensemble 
+mean of mean/minimal/maximal values or ensemble std. deviation of 
+mean/minimal/maximal values respectively. Table~\ref{covariates} 
+shows all variables together with the number of variations and 
+the provided variations listed in the last column. Altogether 
+82~covariates are available in this setting. 
 \fixme{Mention that dswrf\_mean/sprd\_min is not included?}
@@
 %%% transformation. </TH>
-To remove large parts of the skewness of precipitation data, a power transformation has been used frequently (Box and Cox 1964).
-In literature, cubic (\citealt{Stidd:1973}) or square root (\citealt{Hutchinson:1998}) transformations have often been suggested but may vary for different climatic zones or temporal aggregation periods.
-In \cite{Stauffer+Mayr+Messner:2017} optimization methods have been applied to find the best transformation factor which resulted in a value of $\frac{1}{1.6}$.
-Since the observation data set used for this optimization contains the observations considered in this article the same factor is chosen here for the power transformation of the response variable \emph{tp} (total precipitation).
-The resulting transformed variable is expected to be normally distributed and left-censored at censoring point 0.
-For simplicity the same transformation has been chosen for the variable \emph{cape} (convective available potential energy).
+To remove large parts of the skewness of precipitation data, 
+a power transformation has been used frequently (Box and Cox 1964). 
+In literature, cubic (\citealt{Stidd:1973}) or square root 
+(\citealt{Hutchinson:1998}) transformations have often been 
+suggested but may vary for different climatic zones or temporal 
+aggregation periods. In \cite{Stauffer+Mayr+Messner:2017} 
+optimization methods have been applied to find the best transformation 
+factor which resulted in a value of $\frac{1}{1.6}$. Since the 
+observation data set used for this optimization contains the 
+observations considered in this article the same factor is 
+chosen here for the power transformation of the response variable 
+\emph{tp} (total precipitation). The resulting transformed 
+variable is expected to be normally distributed and left-censored 
+at censoring point 0. For simplicity the same transformation has 
+been chosen for the variable \emph{cape} (convective available 
+potential energy).
 
@@
 
-\item An ensemble model output statistics (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}) model where the location parameter and the logarithm of the scale parameter depend linearly on the ensemble mean forecast of the total daily precipitation sum and the corresponding ensemble standard deviation respectively.
-The choice of the logarithmic function as link function for the scale parameter has been made based on the results of \cite{Gebetsberger+Messner+Mayr:2017} where different link functions have been compared.
-Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function led to the best performance of the EMOS model in this application.
-
-\item A GAMLSS where for each distribution parameter the selection of covariates and the specification of interactions is done in advance, both based on meteorological expert knowledge following \cite{Stauffer+Umlauf+Messner:2017} (prespecified GAM).
-Among the available 82 regressors, 9 where selected for the location parameter and 6 for the scale parameter.
-Additionally, for each distribution parameter one interaction term is specified in which convective available energy is interacting with total precipitation (with the ensemble mean for the location parameter and with the ensemble standard deviation for the scale parameter).
-%In this terms an ensemble mean of 6 hour sums of convective available energy is interacting with an ensemble mean or standarad deviation of 6 hour sums of total precipitation.  
-These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation.
-By putting extra effort in the specification of the formulas the prespecified GAM is provided with additional information and is therefore expected to profit from this advantage.
-\item A boosted GAMLSS \citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
-\citep{Messner+Mayr+Zeileis:2017} for an automatic selection of covariates (boosted GAM).
-The idea of this boosting approach is to iteratively minimize a loss function where in each step only the coefficient leading to the highest improvement of the current model is updated.
-As the initial coefficients are all zero, only the most important covariates are included with nonzero coefficients while less relevant variables are ignored if the procedure is stopped before convergence.
+\item An ensemble model output statistics 
+(EMOS, \citealp{Gneiting+Raftery+Westveld:2005}) model where the 
+location parameter and the logarithm of the scale parameter depend 
+linearly on the ensemble mean forecast of the total daily precipitation 
+sum and the corresponding ensemble standard deviation respectively.
+The choice of the logarithmic function as link function for the scale 
+parameter has been made based on the results of 
+\cite{Gebetsberger+Messner+Mayr:2017} where different link functions 
+have been compared. Moreover, testing the identity, quadratic and 
+logarithmic function as link functions in this setting supported 
+this choice as the logarithmic function led to the best performance 
+of the EMOS model in this application.
+\item A GAMLSS where for each distribution parameter the selection 
+of covariates and the specification of interactions is done in 
+advance, both based on meteorological expert knowledge following 
+\cite{Stauffer+Umlauf+Messner:2017} (prespecified GAM). Among the 
+available 82 regressors, 9 where selected for the location parameter 
+and 6 for the scale parameter. Additionally, for each distribution 
+parameter one interaction term is specified in which convective 
+available energy is interacting with total precipitation (with the 
+ensemble mean for the location parameter and with the ensemble 
+standard deviation for the scale parameter). These interactions 
+might capture events with a high potential for moderate or severe 
+thunderstorms with large amounts of precipitation. By putting 
+extra effort in the specification of the formulas the prespecified 
+GAM is provided with additional information and is therefore 
+expected to profit from this advantage.
+\item A boosted GAMLSS \citep{Hofner+Mayr+Schmid:2016} using 
+non-homogeneous boosting \citep{Messner+Mayr+Zeileis:2017} 
+for an automatic selection of covariates (boosted GAM). 
+The idea of this boosting approach is to iteratively minimize 
+a loss function where in each step only the coefficient leading 
+to the highest improvement of the current model is updated. As 
+the initial coefficients are all zero, only the most important 
+covariates are included with nonzero coefficients while less 
+relevant variables are ignored if the procedure is stopped before 
+convergence.
 \end{itemize}
@@
 \begin{table}[t!]
-%\resizebox{1.4\textwidth}{!}{
-%\begin{minipage}{\textwidth}
 \centering
@@
 \end{tabular}
-\caption[Table caption text]{Overview of models with type of covariate dependency and included covariates for each distribution parameter.}
+\caption[Table caption text]{Overview of models with type of 
+covariate dependency and included covariates for each distribution 
+parameter.}
 \label{model_specification}
-%}
 \end{table}
 
-In order to compare the performance of the chosen methods, the continuous ranked probability score (CRPS) is calculated for each model. 
+In order to compare the performance of the chosen methods, the continuous 
+ranked probability score (CRPS) is calculated for each model. 
 \fixme{Adding reference for CPRS}
 
-The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
+The CRPS measures the accuracy of a probabilistic model. 
+For one observation $y$ with the predicted distribution 
+function $F$ the CRPS value is calculated by
 $$
@@
 
-In this application the CRPS is evaluated out of sample, e.g.
-on a part of the data set that has not been used to build the model.
-The mean CRPS over this subset of the data is used for validation.
-Since the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit.
-In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observations.
-
-For the purpose of comparing the models' performances with respect to a reference model the so-called skill score of the CRPS can be calculated.
-In that way a measure of how much a method improves the model fit in terms of CRPS compared to the reference method is obtained.
+In this application the CRPS is evaluated out of sample, 
+e.g. on a part of the data set that has not been used to 
+build the model. The mean CRPS over this subset of the data 
+is used for validation. Since the CRPS can be seen as a 
+generalization of the mean absolute error, a low value indicates 
+a good fit. In applications, one big advantage in terms of 
+interpretability is that the resulting value is on the scale of 
+the observations.
+
+For the purpose of comparing the models' performances with respect 
+to a reference model the so-called skill score of the CRPS can be 
+calculated. In that way a measure of how much a method improves the
+model fit in terms of CRPS compared to the reference method is obtained.
 $$
@@
 \label{Application for one station}
-First of all we focus on one observation station, Axams in Tyrol.
-This station was selected due to its geographical closeness to Innsbruck, the capital of Tyrol.
-As for all other stations, daily observations and numerical weather predictions are available for the month of July of 28 years from 1985 through the year 2012.
-In Figure~\ref{axams_onestation} the results of fitting a distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown as a motivational example for a probabilistic forecast.
-Staying with this fitted model we are now going to analyze variable importance of the distributional forest model and compare its results to those of the GAMs and the EMOS model based on residual QQ-plots.
-%%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>
+First of all we focus on one observation station, Axams in Tyrol. This 
+station was selected due to its geographical closeness to Innsbruck, 
+the capital of Tyrol. As for all other stations, daily observations and 
+numerical weather predictions are available for the month of July of 
+28~years from 1985 through the year 2012. In Figure~\ref{axams_onestation} 
+the results of fitting a distributional forest to the first 24 years of the 
+data set and predicting the total amount of precipitation for one specific 
+day of each of the four successive years have been shown as a motivational 
+example for a probabilistic forecast. Staying with this fitted model we are 
+now going to analyze variable importance of the distributional forest model 
+and compare its results to those of the GAMs and the EMOS model based on 
+residual QQ-plots.
 
 %%%<TH> always us the term "distributional forest"
-As shown in Table~\ref{model_specification} the distributional forest includes all of the available covariates.
-A single tree model would allow to easily analyze which covariates are used as split variables in the nodes and hence have a significant impact on the response variable.
-In a forest model though, for each node in each of the trees a subset of the covariates is chosen randomly as possible split variables to select one from.
-Moreover, each tree is built on a resampled version of the learning data.
-For these reasons the choices of split variables vary over the trees of a forest making it difficult to analyze the impact of covariates just by investigating the different trees.
-To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied.
-Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 years are listed in Figure~\ref{varimp}.
-As expected, the numerical forecast for total precipitation is the most influential variable.
-In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.
+As shown in Table~\ref{model_specification} the distributional forest 
+includes all of the available covariates. A single tree model would allow 
+to easily analyze which covariates are used as split variables in the nodes 
+and hence have a significant impact on the response variable. In a forest 
+model though, for each node in each of the trees a subset of the covariates 
+is chosen randomly as possible split variables to select one from. Moreover, 
+each tree is built on a resampled version of the learning data. For these 
+reasons the choices of split variables vary over the trees of a forest making 
+it difficult to analyze the impact of covariates just by investigating the 
+different trees. To get an overview of which covariates have the most influential 
+effect on the response variable a standard measurement of variable importance 
+for forest models has been applied. Based on the mean decrease in accuracy 
+regarding the CRPS the 10 leading covariates of the above described distributional 
+forest for station Axams learned on the first 24 years are listed in 
+Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation 
+is the most influential variable. In particular, five variations of this variable 
+are among the top ten of the list of covariates ordered by variable importance.
 
@@
 @
-\caption{\label{varimp} CRPS-based variable importance for the top 10 covariates in the distributional forest. Based on data for station Axams, learning period 1985--2008 and assessed in 2009--2012.}
+\caption{\label{varimp} CRPS-based variable importance for the top 10 covariates 
+in the distributional forest. Based on data for station Axams, learning period 
+1985--2008 and assessed in 2009--2012.}
 \end{figure}
 
-In Figure~\ref{qqr_oosample} an impression of the quality of predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models.
-For these plots 100 randomized quantile residuals are simulated per observation.
-The closer the points are to the diagonal line the better the estimated distributional model fits the data.
-Overall, it can be stated that all four models provide well calibrated forecasts.
-This is also supported by the corresponding probability integral transform (PIT) histograms (\citealt{Gneiting+Balabdaoui+Raftery:2007}) in the appendix.
+In Figure~\ref{qqr_oosample} an impression of the quality of predictions made 
+by the tested methods is given by residual QQ-plots for out-of-sample predictions 
+of each of the models. For these plots 100 randomized quantile residuals are 
+simulated per observation. The closer the points are to the diagonal line the 
+better the estimated distributional model fits the data. Overall, it can be 
+stated that all four models provide well calibrated forecasts. This is also 
+supported by the corresponding probability integral transform (PIT) histograms 
+(\citealt{Gneiting+Balabdaoui+Raftery:2007}) in the appendix.
 
@@
 @
-\caption{\label{qqr_oosample} Out-of-sample residual QQ-plots (2009--2012) for station Axams based on models learned on data from 1985--2008.}
+\caption{\label{qqr_oosample} Out-of-sample residual QQ-plots (2009--2012) 
+for station Axams based on models learned on data from 1985--2008.}
 \end{figure}
 
-After analyzing the results of the models on only one data set where visualization is easier, we now want to get a more profound impression of the performances of the different methods.
-For that reason the four methods are evaluated in a cross validation framework for the station Axams.
-The data set over all 28 years is split into 7 subsets each consisting of 4 randomly selected years.
-While 6 of these subsets are used as learning data the 7th is the testing data set for which predictions are made and evaluated.
-This is done by calculating the CRPS value for each of the observations in the testing data set and then averaging them.
-Each of the 7 parts is selected once as testing data which leads to 7 different settings of which again the average values are stored.
-This whole procedure is repeated 10 times.
-The resulting skill scores are illustrated in Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model represented by the horizontal line at height 0.
-This CRPS skill score boxplot shows that, in this setting, both GAMs and the distributional forest perform better than the EMOS model.
-While the two GAMs lead to an improvement of around 4 percent with the boosted version being slightly ahead, the distributional forest even reaches an improvement rate close to 6 percent.
+After analyzing the results of the models on only one data set where 
+visualization is easier, we now want to get a more profound impression 
+of the performances of the different methods. For that reason the four 
+methods are evaluated in a cross validation framework for the station Axams. 
+The data set over all 28 years is split into 7 subsets each consisting of 
+4 randomly selected years. While 6 of these subsets are used as learning 
+data the 7th is the testing data set for which predictions are made and 
+evaluated. This is done by calculating the CRPS value for each of the 
+observations in the testing data set and then averaging them. Each of the 
+7 parts is selected once as testing data which leads to 7 different settings 
+of which again the average values are stored. This whole procedure is repeated 
+10 times. The resulting skill scores are illustrated in 
+Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model 
+represented by the horizontal line at height 0. This CRPS skill score boxplot 
+shows that, in this setting, both GAMs and the distributional forest perform 
+better than the EMOS model. While the two GAMs lead to an improvement of 
+around 4 percent with the boosted version being slightly ahead, the 
+distributional forest even reaches an improvement rate close to 6 percent.
 <<echo=FALSE, results=hide>>=
@@
 @
-\caption{\label{boxplots_crps_axams}CRPS skill score from the 10 times 7-fold cross validation at station Axams (1985--2012). The horizontal orange line pertains to the reference model EMOS.}
+\caption{\label{boxplots_crps_axams}CRPS skill score from the 10 times 
+7-fold cross validation at station Axams (1985--2012). The horizontal orange 
+line pertains to the reference model EMOS.}
 \end{figure}
@@
 \subsection{Application for all stations}
-Until now the focus has been on one observation station only.
-To test and compare the four methods under different geographical circumstances we are now going to apply them on all 95 stations.
-At each of them, the first 24 years are used as learning data set and predictions are made for the ensuing 4 years of available observations.
-For these out-of-sample predictions the CRPS skill score with the EMOS model as reference model is calculated for each method at each station.
-In Figure~\ref{boxplot_crps_all} the boxplots of the CRPS skill score values are plotted and show that the distributional forest leads to the highest improvement averaged over all stations.
+Until now the focus has been on one observation station only. To test and 
+compare the four methods under different geographical circumstances we are 
+now going to apply them on all 95 stations. At each of them, the first 
+24 years are used as learning data set and predictions are made for the 
+ensuing 4 years of available observations. For these out-of-sample predictions 
+the CRPS skill score with the EMOS model as reference model is calculated 
+for each method at each station. In Figure~\ref{boxplot_crps_all} the boxplots 
+of the CRPS skill score values are plotted and show that the distributional 
+forest leads to the highest improvement averaged over all stations.
 The green line represents the results for the station Axams.
@@
 @
-\caption{\label{boxplot_crps_all}CRPS skill score for each station (grey lines) and aggregated over all stations (boxplots).
-Station Axams is highlighted in green and the horizontal orange line pertains to the reference model EMOS.
-The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
+\caption{\label{boxplot_crps_all}CRPS skill score for each station (grey lines) 
+and aggregated over all stations (boxplots). Station Axams is highlighted in green 
+and the horizontal orange line pertains to the reference model EMOS. The models are 
+learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
 \end{figure}
 
-Figure~\ref{map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation Data, \cite{srtm}) with the location of all observation stations included in the data set.
-Each station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station.
-The color of the symbol indicates the difference between the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
-Even though for some stations, especially in East Tyrol, one of the two GAMs and for a few others the EMOS model leads to the best results, for the majority of stations the distributional forest performs best.
+Figure~\ref{map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation 
+Data, \cite{srtm}) with the location of all observation stations included in the 
+data set. Each station is illustrated by a symbol with the type of symbol 
+representing the method that performed best in terms of CRPS at this station. 
+The color of the symbol indicates the difference between the CRPS of the 
+distributional forest and the CRPS of the best performing among the other 
+three models (prespecified GAM, boosted GAM and EMOS).
+Even though for some stations, especially in East Tyrol, one of the two GAMs 
+and for a few others the EMOS model leads to the best results, for the 
+majority of stations the distributional forest performs best.
 
@@
 @
-\caption{\label{map}Map of Tyrol showing which model performed best at which station (type of symbol).
-The color codes whether the distributional forest had lower or higher CRPS (red or blue) compared to the best of the other three models (prespecified GAM, boosted GAM and EMOS).}
+\caption{\label{map}Map of Tyrol showing which model performed best at 
+which station (type of symbol). The color codes whether the distributional 
+forest had lower or higher CRPS (red or blue) compared to the best of the 
+other three models (prespecified GAM, boosted GAM and EMOS).}
 \end{figure}
@@
 
-From the results in Section~\ref{ProbabilisticForecasting} we can conclude that distributional forests provide a powerful alternative framework requiring only little effort to set up a model which is competitive to commonly used methods or even outperforms them.
-Even though a proper selection of covariates and specification of interactions in advance can lead to better fitting (and probably less complex) models, this presupposes profound knowledge of all influencing factors.
-Otherwise, important aspects might be missed or effects which do not correspond to the truth might be imposed on the model.
-This risk can be avoided by applying a boosting algorithm for an automatic variable selection.
-In this case the trade-off is often a remarkably higher computational effort.
-To build a distributional forest the user is not required to provide any expert knowledge or additional information about the setting as variable selection and detection of interactions are done automatically by the tree-structure of the model.
-This feature is particularly advantageous in case of a very high number of covariates.
-Therefore, the above discussed application illustrates the strengths of distributional forests as they offer a very flexible framework which enables an easy use of the method and at the same time lead to an overall improvement in the predictive performance in this setting.
-Especially in situations where no or only little information is available about possible influences, effects or interactions this novel methodology can be seen as a good and recommendable compromise being able to deal with a wide range of modelling situations in a reasonable way.
-
-%%% <TH> this should go into the discussion or the interpretation of the
-%%% empirical results </TH>
-%Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required.
-In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.
+From the results in Section~\ref{ProbabilisticForecasting} we can conclude 
+that distributional forests provide a powerful alternative framework requiring 
+only little effort to set up a model which is competitive to commonly used 
+methods or even outperforms them. Even though a proper selection of covariates 
+and specification of interactions in advance can lead to better fitting 
+(and probably less complex) models, this presupposes profound knowledge of 
+all influencing factors. Otherwise, important aspects might be missed or 
+effects which do not correspond to the truth might be imposed on the model. 
+This risk can be avoided by applying a boosting algorithm for an automatic 
+variable selection. In this case the trade-off is often a remarkably higher 
+computational effort. To build a distributional forest the user is not 
+required to provide any expert knowledge or additional information about 
+the setting as variable selection and detection of interactions are done 
+automatically by the tree-structure of the model. This feature is 
+particularly advantageous in case of a very high number of covariates. 
+Therefore, the above discussed application illustrates the strengths of 
+distributional forests as they offer a very flexible framework which 
+enables an easy use of the method and at the same time lead to an overall 
+improvement in the predictive performance in this setting. Especially 
+in situations where no or only little information is available about 
+possible influences, effects or interactions this novel methodology 
+can be seen as a good and recommendable compromise being able to deal 
+with a wide range of modelling situations in a reasonable way.
+
+%Even though many situations demand for the use of specifically tailored 
+%models providing the needed features and therefore being clearly advantageous 
+%and the best choice, it is often hard to tell in advance which features 
+%will be required. In these cases a distributional forest is a 
+%recommendable compromise providing reasonable models and performing 
+%well without any prespecification or additional knowledge needed in advance.
 
 \section*{Computational details}
-The corresponding implementation to the novel methods presented in this article is available on \textsf{R}-Forge in an \textsf{R} package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}).
-The function \code{disttree} uses either the \code{mob} or \code{ctree} function as a framework to build a tree.
-Similarly, the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a distributional forest.
-All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{Hothorn+Zeileis:2015}).
-To specify a distribution family the user can hand over one of the following objects as input argument in \code{distfit}, \code{disttree} and \code{distforest}:
+The corresponding implementation to the novel methods presented 
+in this article is available on \textsf{R}-Forge in an \textsf{R} 
+package called \textbf{disttree} within the project \textbf{partykit} 
+(\url{https://R-Forge.R-project.org/projects/partykit/}). The function 
+\code{disttree} uses either the \code{mob} or \code{ctree} function 
+as a framework to build a tree. Similarly, the function 
+\code{distforest} is based on the \code{cforest} function which 
+provides a framework to build a distributional forest. All three 
+functions, \code{mob}, \code{ctree} and \code{cforest}, are provided 
+in the package \textbf{partykit} (\citealt{Hothorn+Zeileis:2015}). 
+To specify a distribution family the user can hand over one of the 
+following objects as input argument in \code{distfit}, \code{disttree} 
+and \code{distforest}: 
 \begin{itemize}
-\item a GAMLLS family object from the \textsf{R} package \textbf{gamlss.dist} (\citealt{Stasinopoulos+Rigby:2007}).
+\item a GAMLLS family object from the \textsf{R} package \textbf{gamlss} 
+(\citealt{Stasinopoulos+Rigby:2007}).
 \item a family list generating function from the package \textbf{disttree}.
 \item a complete family list from the package \textbf{disttree}.
-\item a custom list containing all required information about the distribution family.
+\item a custom list containing all required information about the 
+distribution family.
 \end{itemize}
 
-In Section~\ref{ProbabilisticForecasting} the prespecified GAMs were built using the function \code{gamlss} from the \textsf{R} package \textbf{gamlss.dist} (\citealt{Stasinopoulos+Rigby:2007}).
-For the boosted GAMs the function \code{gamboostLSS} from the \textsf{R} package \textbf{gamboostLSS} (\citealt{Hofner+Mayr+Schmid:2016}) was applied.
-The EMOS models were built using the function \code{crch} from the \textsf{R} package \textbf{crch} (\citealt{Messner+Mayr+Zeileis:2016})
+In Section~\ref{ProbabilisticForecasting} the prespecified GAMs were 
+built using the function \code{gamlss} from the \textsf{R} package 
+\textbf{gamlss.dist} (\citealt{Stasinopoulos+Rigby:2007}). For the 
+boosted GAMs the function \code{gamboostLSS} from the \textsf{R} 
+package \textbf{gamboostLSS} (\citealt{Hofner+Mayr+Schmid:2016}) 
+was applied. The EMOS models were built using the function 
+\code{crch} from the \textsf{R} package \textbf{crch} 
+(\citealt{Messner+Mayr+Zeileis:2016})
 
@@
 
-%\newpage
 \nocite{Seibold+Zeileis+Hothorn:2016, Seibold+Zeileis+Hothorn:2017, Yee:2010}
@@
 \label{treealgorithm}
-The tree algorithm used in the applications discussed in this paper is explained in the following.
-For notational simplicity, the testing and splitting procedure is described for the root node with observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$.
-In each child node the corresponding subset of observations depends on the assignment made by the foregoing split.\\
+The tree algorithm used in the applications discussed in this paper 
+is explained in the following. For notational simplicity, the testing 
+and splitting procedure is described for the root node with observations 
+$\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. In each child node the 
+corresponding subset of observations depends on the assignment made 
+by the foregoing split.
 
 \begin{itemize}
-\item A distributional model $D(Y, \theta)$ is fit to the set of observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution parameter $\hat{\theta}$.
-This parameter can be a multidimensional vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, depending on the type of distribution.
-The estimation is done by maximizing the log-likelihood function $\ell(\theta; Y)$, i.e.
+\item A distributional model $D(Y, \theta)$ is fit to the set of 
+observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution 
+parameter $\hat{\theta}$. This parameter can be a multidimensional 
+vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, 
+depending on the type of distribution. The estimation is done by 
+maximizing the log-likelihood function $\ell(\theta; Y)$, i.e.
 $$
@@
 $$
-is applied.
-The type of the transformation function $g_j$ depends on the type of the split variable $Z_j$.
-If $Z_j$ is numeric then $g_j$ is simply the identity function $g_j(z_{ji}) = z_{ji}$.
-If $Z_j$ is a categorical variable with $L$ categories then $g_j(z_{ji}) = e_L(z_{ji}) = (\I(z_{ji} = 1), \ldots, \I(z_{ji} = L))$ such that $g_j$ is a unit vector where the element corresponding to the value of $z_{ji}$ is $1$.
-Observations with missing values are excluded from the sums.
-
-With the conditional expectation $\mu_{j}$ and the covariance $\Sigma_{j}$ as derived by \cite{Strasser+Weber:1999} the test statistic can be standardized.
-For a numeric split variable this results in the Pearson correlation coefficient
+is applied. The type of the transformation function $g_j$ depends on 
+the type of the split variable $Z_j$. If $Z_j$ is numeric then $g_j$ 
+is simply the identity function $g_j(z_{ji}) = z_{ji}$. If $Z_j$ is 
+a categorical variable with $L$ categories then 
+$g_j(z_{ji}) = e_L(z_{ji}) = (\I(z_{ji} = 1), \ldots, \I(z_{ji} = L))$ 
+such that $g_j$ is a unit vector where the element corresponding to 
+the value of $z_{ji}$ is $1$. Observations with missing values are 
+excluded from the sums.
+
+With the conditional expectation $\mu_{j}$ and the covariance 
+$\Sigma_{j}$ as derived by \cite{Strasser+Weber:1999} the test 
+statistic can be standardized. For a numeric split variable this 
+results in the Pearson correlation coefficient
 $$
@@
 $$
-The smaller the p-value corresponding to the standardized test statistic $c(t_{j},\mu_{j},\Sigma_{j})$ is the stronger the discrepancy from the assumption of independence between the scores and the split variable $Z_j$.
-
-\item After Bonferroni adjusting the p-values check whether any of the resulting p-values is beneath the selected significance level.
-If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest p-value to the scores corresponding to the distribution parameters relevant for the split.
-
-\item Choose the breakpoint that leads to the highest discrepancy between score functions in the two resulting subgroups. This difference can be measured by the linear statistic
+The smaller the p-value corresponding to the standardized test 
+statistic $c(t_{j},\mu_{j},\Sigma_{j})$ is the stronger the discrepancy 
+from the assumption of independence between the scores and the split 
+variable $Z_j$.
+
+\item After Bonferroni adjusting the p-values check whether any of 
+the resulting p-values is beneath the selected significance level. 
+If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest 
+p-value to the scores corresponding to the distribution parameters 
+relevant for the split.
+
+\item Choose the breakpoint that leads to the highest discrepancy 
+between score functions in the two resulting subgroups. This 
+difference can be measured by the linear statistic
 $$
@@
 $$
-where $\mathcal{B}_{1r}$ is the first of the two new subgroups that are defined by splitting in split point $r$ of variable $Z_{j^{\ast}}$. The split point is then chosen as follows:
+where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
+are defined by splitting in split point $r$ of variable $Z_{j^{\ast}}$. 
+The split point is then chosen as follows:
 $$
@@
 
-\item Repeat the testing and splitting procedure in each of the resulting subgroups until some stopping criterion is reached.
-This criterion can for example be a minimal number of observations in a node or a minimal p-value for the statistical tests.
+\item Repeat the testing and splitting procedure in each of the resulting 
+subgroups until some stopping criterion is reached. This criterion 
+can for example be a minimal number of observations in a node or a 
+minimal p-value for the statistical tests.
 
@@
 
-This permutation test based tree algorithm is presented in \cite{Hothorn+Hornik+Zeileis:2006} as the ctree algorithm.
-A different framework to build a tree is provided by the MOB algorithm which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).
-
-%\begin{figure}[t!]
-%\begin{subfigure}{0.5\textwidth}
-%<<rain_axams_pit_insample, fig=TRUE, echo=FALSE>>=
-%
-%# in sample
-%set.seed(7)
-%par(mfrow = c(2,2))
-%pithist(pit_df_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%pithist(pit_g_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%pithist(pit_gb_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%pithist(pit_ml_l, nsim = 1000, breaks = seq(0, 1, length.out = 9))
-%
-%@
-%\caption{\label{pit_insample} PIT histograms for station Axams (in sample)}
-%\end{figure}
+This permutation test based tree algorithm is presented in 
+\cite{Hothorn+Hornik+Zeileis:2006} as the ctree algorithm. 
+A different framework to build a tree is provided by the MOB algorithm 
+which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).
 
 \section{Probability integral transform (PIT) histograms}
-Probability integral transform (PIT) histograms \cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used tool to assess the predictive power of distributional forecasting methods, particularly in the field of meteorological forecasting \\
-Let $x$ be an observation with true distribution function $G$ and predicted distribution function $F$.
-Then the corresponding probability integral transform (PIT) value is $p = F(x)$.
-In the ideal case of a perfect prediction and $F$ being continuous $p$ follows a uniform distribution.
-This uniformity can be analysed applying a histogram.
-The closer the bars of a PIT histogram are to the value 1, the closer the distribution of $p$ is to a uniform distribution and therefore the better the predicted distribution models the true distribution.
-
-The PIT histograms in Figure~\ref{pit_oosample} correspond to the forecasts presented in Section~\ref{Application for one station}. 
-Daily observations of total precipitation from station Axams are considered and predictions are made for the 24th of July 2009, 2010, 2011 and 2012, based on data from the previous 24 years.
+Probability integral transform (PIT) histograms 
+\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used tool to 
+assess the predictive power of distributional forecasting methods, 
+particularly in the field of meteorological forecasting.
+
+Let $x$ be an observation with true distribution function $G$ and 
+predicted distribution function $F$. Then the corresponding probability 
+integral transform (PIT) value is $p = F(x)$. In the ideal case of a 
+perfect prediction and $F$ being continuous $p$ follows a uniform distribution. 
+This uniformity can be analysed applying a histogram. The closer the bars 
+of a PIT histogram are to the value 1, the closer the distribution of $p$ 
+is to a uniform distribution and therefore the better the predicted 
+distribution models the true distribution.
+
+The PIT histograms in Figure~\ref{pit_oosample} correspond to the forecasts 
+presented in Section~\ref{Application for one station}. Daily observations 
+of total precipitation from station Axams are considered and predictions 
+are made for the 24th of July 2009, 2010, 2011 and 2012, based on data 
+from the previous 24 years.
 

### 2018-02-22 pkg/disttree/inst/draft/draft.Rnw
additional reference for Breiman and Cutler's random forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2036 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d696b83c5fe4e3bac0f3b0cc16a8f1cafb3db3b9

@@
 %%% random forest. Maybe it is better to not mention this method at any time. </TH>
-Common forest models (such as Breiman and Cutler's randomForest) consist of 
-trees only considering the mean of the distribution of the response variable. 
-For that reason, changes in other parameters can only be detected if they 
-are connected to changes in the location parameter, e.g. if they depend on 
-the location parameter. As explained before, a distributional tree considers
-scores for all distribution parameters. Therefore, a forest consisting of 
+Common forest models such as the original forest technique by 
+\cite{Breiman+Cutler:2004} consist of trees only considering the mean of the 
+distribution of the response variable. For that reason, changes in other 
+parameters can only be detected if they are connected to changes in the 
+location parameter, e.g. if they depend on the location parameter. 
+As explained before, a distributional tree considers scores for all 
+distribution parameters. Therefore, a forest consisting of 
 distributional trees can also capture separate effects in each of the 
@@
 
-\fixme{Avoid nocite, make all relevant citations properly in the text.}
-
-\nocite{Seibold+Zeileis+Hothorn:2016, Seibold+Zeileis+Hothorn:2017, Yee:2010}
-% Zeileis+Wiel+Hornik:2008, 
-% Zeileis+Hothorn+Hornik:2010,
-
-%\newpage
-%\newpage
 \bibliography{ref.bib}
@@
 The tree algorithm used in the applications discussed in this paper 
-is explained in the following. For notational simplicity, the testing 
+is going to be explained following the ideas presented in 
+\cite{Seibold+Zeileis+Hothorn:2017}. For notational simplicity, the testing 
 and splitting procedure is described for the root node with observations 

### 2018-02-23 pkg/disttree/inst/draft/draft.Rnw
small corrections and additional reference for Breiman and Cutler's random forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2038 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 81c7c29c962330b2090d54d5fec5e755ea6ef090

@@
 82~covariates are available in this setting.
+
 \fixme{Mention that dswrf\_mean/sprd\_min is not included?}
@@
 location parameter and the logarithm of the scale parameter depend 
-linearly on the ensemble mean forecast of the total daily precipitation 
-sum and the corresponding ensemble standard deviation respectively.
+linearly on the ensemble mean forecast of total daily precipitation 
+sums and the corresponding ensemble standard deviation respectively.
 The choice of the logarithmic function as link function for the scale 
@@
 The CRPS measures the accuracy of a probabilistic model. 
-For one observation $y$ with the predicted distribution 
+For one observation $y$ with predicted distribution 
 function $F$ the CRPS value is calculated by
@@
 In this application the CRPS is evaluated out of sample, 
-e.g. on a part of the data set that has not been used to 
+e.g. on a part of the data set that has not been considered to 
 build the model. The mean CRPS over this subset of the data 
@@
 After analyzing the results of the models on only one data set where 
-visualization is easier, we now want to get a more profound impression 
-of the performances of the different methods. For that reason the four 
-methods are evaluated in a cross validation framework for the station Axams. 
+visualization is easier, the methods are now going to be evaluated
+in a cross validation framework for the station Axams. 
 The data set over all 28 years is split into 7 subsets each consisting of 

### 2018-02-26 pkg/partykit/inst/guideliketest/results.Rnw
comparing results of different test statistics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2039 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 86884d7441449eb5bfc473319355ac3b9648023f

\documentclass{article}

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
source("guidelike.R")
load("sim/simres20180224_1step.rda")
@

\title{Comparison of different testing functions}

\begin{document}

\section{All: 1 step}
\setkeys{Gin}{width=\linewidth}
<<all_TTF1, fig=TRUE, echo=TRUE, height=10, width=8>>=
simres$test <- factor(simres$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = simres, 
       subset = binary_beta == TRUE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = FALSE, ylim = c(0,1))
@

\newpage  
<<all_TFF1, fig=TRUE, echo=TRUE, height=10, width=8>>=
simres$test <- factor(simres$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = simres, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = FALSE, ylim = c(0,1))
@

\newpage
\section{Influence of binarization of scores and/or categorization of split variables: 1 step}

\subsection{ctree}
<<ctree_cat_bin_TFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree", "ctree_cat", "ctree_bin", "ctree_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<ctree_cat_bin_FFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree", "ctree_cat", "ctree_bin", "ctree_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\subsection{mfluc}

<<mfluc_cat_bin_TFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<mfluc_cat_bin_FFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\section{GUIDE vs ctree and mfluc with binarized scores and categorized splitvariables: 1 step}
<<guide_vs_cat_bin_TFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc_cat_bin", "ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_vs_cat_bin_FFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc_cat_bin", "ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\section{GUIDE versions: 1 step}

<<guide_versions_TTF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_versions_TFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_versions_FTF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_versions_FFF1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\section{check alpha: 1 step}
\subsection{ctree}
<<check_alpha_ctree1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subdata <- subset(simres, test %in% c("ctree", "ctree_cat", "ctree_bin", "ctree_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,0.1))
@

\newpage  
\subsection{mfluc}
<<check_alpha_mfluc1, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,0.1))
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%load data set for 2 step function
<<>>=
load("sim/simres20180224_2step.rda")
@

\newpage
\section{All: 2 steps}
<<all_TTF2, fig=TRUE, echo=TRUE, height=10, width=8>>=
simres$test <- factor(simres$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = simres, 
       subset = binary_beta == TRUE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = FALSE, ylim = c(0,1))
@

\newpage  
<<all_TFF2, fig=TRUE, echo=TRUE, height=10, width=8>>=
simres$test <- factor(simres$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = simres, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = FALSE, ylim = c(0,1))
@

\newpage  
<<all_FTF2, fig=TRUE, echo=TRUE, height=10, width=8>>=
simres$test <- factor(simres$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = simres, 
       subset = binary_beta == FALSE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = FALSE, ylim = c(0,1))
@

\newpage  
<<all_FFF2, fig=TRUE, echo=TRUE, height=10, width=8>>=
simres$test <- factor(simres$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = simres, 
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = FALSE, ylim = c(0,1))
@

\newpage
\section{Influence of binarization of scores and/or categorization of split variables: 2 steps}

\subsection{ctree}
<<ctree_cat_bin_TFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree", "ctree_cat", "ctree_bin", "ctree_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<ctree_cat_bin_FFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree", "ctree_cat", "ctree_bin", "ctree_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\subsection{mfluc}

<<mfluc_cat_bin_TTF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<mfluc_cat_bin_TFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<mfluc_cat_bin_FTF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<mfluc_cat_bin_FFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\section{GUIDE vs ctree and mfluc with binarized scores and categorized splitvariables: 2 steps}
<<guide_vs_cat_bin_TFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc_cat_bin", "ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_vs_cat_bin_FFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("mfluc_cat_bin", "ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\section{GUIDE versions}
<<guide_versions_TTF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_versions_TFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_versions_FTF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == TRUE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage  
<<guide_versions_FFF2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subset(simres, test %in% c("ctree_cat_bin", "ctree_max_cat_bin", "guide_coin_12", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == FALSE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,1))
@

\newpage
\section{check alpha: 2 steps}
\subsection{ctree}
<<check_alpha_ctree2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subdata <- subset(simres, test %in% c("ctree", "ctree_cat", "ctree_bin", "ctree_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,0.1))
@

\newpage  
\subsection{mfluc}
<<check_alpha_mfluc2, fig=TRUE, echo=TRUE, height=9, width=8>>=
subdata <- subdata <- subset(simres, test %in% c("mfluc", "mfluc_cat", "mfluc_bin", "mfluc_cat_bin", "guide_coin_12"))
subdata$test <- factor(subdata$test)
xyplot(prop_T ~ delta | xi + vary_beta, groups = ~ test, 
       data = subdata,
       subset = binary_beta == TRUE & binary_regressor == FALSE & only_intercept == FALSE, 
       type = "b", auto.key = TRUE, ylim = c(0,0.1))
@
\end{document}

### 2018-02-26 pkg/partykit/inst/guideliketest/guidelike.R
comparing results of different test statistics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2039 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 86884d7441449eb5bfc473319355ac3b9648023f

@@ -146,7 +146,7 @@ if(FALSE){
   if(length(unique(x))<2) return(list(p.value = log(1-1), statistic = log(0)))
   
   
-  # categorize residuals
+  # categorize scores
   if(is.null(ctrl$ygroups)){
     # split Y into 2 parts based on whether residuals (here: scores) are positive or negative
     # separately for each parameter
@@ -648,12 +648,18 @@ if(FALSE){
   iy <- data$yxindex ### data[["yx", type = "index"]]
   Y <- model$estfun
   
-  # categorize residuals
+  # categorize scores
   if(is.null(ctrl$ygroups)){
     
     # split Y into 2 parts based on whether residuals (here: scores) are positive or negative
-    Y <-(Y>0)*1    # *1 to store numerical instead of logical values
-  
+    #Y <- (-1)^((Y>0)+1)    # FIX ME: Y should be a factor, but linear tests in ctree only work for numeric values
+    Y <- sign(Y)            # FIX ME: Y should be a factor, but linear tests in ctree only work for numeric values    
+    #if(NCOL(Y) == 1) Y <- data.frame(factor(Y>0))
+    #if(NCOL(Y) > 1){
+    #  Ybin <- data.frame(factor(Y[,1]))
+    #  for(i in 2:NCOL(Y)) Ybin <- cbind(Ybin, factor(Y[,i]>0))
+    #  Y <- Ybin
+    #}
   } else {
     
     if(length(ctrl$ygroups)>1) ybreaks <- ctrl$ygroups
@@ -710,12 +716,18 @@ if(FALSE){
   
   estfun <- model$estfun[subset,,drop = FALSE]
   
-  # categorize residuals
+  # categorize scores
   if(is.null(ctrl$ygroups)){
     
     # split estfun into 2 parts based on whether residuals (here: scores) are positive or negative
-    estfun <- (estfun>0)*1    # *1 to store numerical instead of logical values
-    
+    #estfun <- (-1)^((estfun>0)+1)    # FIX ME: Y should be a factor, but linear tests in ctree only work for numeric values
+    estfun <- sign(estfun)            # FIX ME: Y should be a factor, but linear tests in ctree only work for numeric values
+    #if(NCOL(estfun) == 1) estfun <- data.frame(factor(estfun>0))
+    #if(NCOL(estfun) > 1){
[…]

### 2018-03-09 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2047 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9784e6640d6cb1bd533d84ac01cdec942cf931d7

@@
 terrain, a large grid spacing of NWP models can lead to abrupt 
-local effects within a single NWP grid cell.
+local effects within a single grid cell.
 
@@
 of flexible parametric modeling into the framework of tree-structured 
-models. The ability to deal with a wide variety of effects and an 
-automatic variable selection and detection of interactions are 
-features provided by distributional forests that could be particularly 
+models. In that way, they offer features that could be particularly 
 valuable in probabilistic weather forecasting in order to tackle 
-the above mentioned challenges.
+the above mentioned challenges. Among these features are the ability 
+to deal with a wide variety of effects and an automatic variable 
+selection and detection of interactions. 
 To show their strengths in practice, distributional forests are 
@@
 
-\item An ensemble model output statistics 
-(EMOS, \citealp{Gneiting+Raftery+Westveld:2005}) model where the 
+\item Ensemble model output statistics 
+(EMOS, \citealp{Gneiting+Raftery+Westveld:2005}): The 
 location parameter and the logarithm of the scale parameter depend 
@@
 of the EMOS model in this application.
-\item A GAMLSS where for each distribution parameter the selection 
+\item Prespecified GAMLSS: For each distribution parameter the selection 
 of covariates and the specification of interactions is done in 
 advance, both based on meteorological expert knowledge following 
-\cite{Stauffer+Umlauf+Messner:2017} (prespecified GAM). Among the 
-available 82 regressors, 9 where selected for the location parameter 
+\cite{Stauffer+Umlauf+Messner:2017}. Among the 
+available 82 regressors, 9 are selected for the location parameter 
 and 6 for the scale parameter. Additionally, for each distribution 
@@
 expected to profit from this advantage.
-\item A boosted GAMLSS \citep{Hofner+Mayr+Schmid:2016} using 
-non-homogeneous boosting \citep{Messner+Mayr+Zeileis:2017} 
-for an automatic selection of covariates (boosted GAM). 
+\item Boosted GAMLSS: Non-homogeneous boosting 
+\citep{Messner+Mayr+Zeileis:2017} is used for an automatic 
+selection of covariates \citep{Hofner+Mayr+Schmid:2016}. 
 The idea of this boosting approach is to iteratively minimize 
@@
 
-%\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAM. Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}
+%\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAMLSS.
+Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}
 
@@
 \hline
-\textbf{Prespecified GAM} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
+\textbf{Prespecified GAMLSS} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
                             & in each & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean},\\
@@
 \hline
-\textbf{Boosted GAM} & spline  & all & all \\
+\textbf{Boosted GAMLSS} & spline  & all & all \\
                        & in each &     &     \\
@@
 First of all we focus on one observation station, Axams in Tyrol. This 
-station was selected due to its geographical closeness to Innsbruck, 
+station is selected due to its geographical closeness to Innsbruck, 
 the capital of Tyrol. As for all other stations, daily observations and 
@@
 now going to analyze variable importance of the distributional forest model 
-and compare its results to those of the GAMs and the EMOS model based on 
-residual QQ-plots.
+and compare its results to those of the prespecified and the boosted GAMLSS 
+and the EMOS model based on residual QQ-plots.
 
@@
 qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAM", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAM", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
 qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
@@
 represented by the horizontal line at height 0. This CRPS skill score boxplot 
-shows that, in this setting, both GAMs and the distributional forest perform 
-better than the EMOS model. While the two GAMs lead to an improvement of 
+shows that, in this setting, both GAMLSS and the distributional forest perform 
+better than the EMOS model. While the two GAMLSS lead to an improvement of 
 around 4 percent with the boosted version being slightly ahead, the 
@@
 \centering
-<<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
+<<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 7>>=
 boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
-        names = c("Distributional forest", "Prespecified GAM", "Boosted GAM"),
+        names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
         ylab = "CRPS skill score", col = "lightgray") 
@@
 s <- 1 - crps_all[, 2:4]/crps_all[,6]
-colnames(s) <- c("Distributional forest", "Prespecified GAM", "Boosted GAM")
+colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
 
@@
 \centering
-<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, width = 6.5>>=
+<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, width = 7>>=
   matplot(t(s[,]), type = "l", lwd = 2, 
@@
 distributional forest and the CRPS of the best performing among the other 
-three models (prespecified GAM, boosted GAM and EMOS).
-Even though for some stations, especially in East Tyrol, one of the two GAMs 
+three models (prespecified GAMLSS, boosted GAMLSS and EMOS).
+Even though for some stations, especially in East Tyrol, one of the two GAMLSS 
 and for a few others the EMOS model leads to the best results, for the 
@@
                 xlim = c(9.8,13.2), 
-                ylim = c(46.6, 47.85))
+                ylim = c(46.6, 47.87))
   plot(tirol.gadm, add = TRUE)
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend(x = 9.8, y = 47.795, pch=c(21,24,25,22), legend = c("Distributional forest", "Prespecified GAM", "Boosted GAM", "EMOS"), cex = 1, bty = "n")
-  text(x = 10.3, y = 47.8, labels = "Models with lowest CRPS")
-  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.74), line = 0.3)
+  legend(x = 9.8, y = 47.815, pch=c(21,24,25,22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
+  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
+  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
   par(mar = c(0.5,0.2,0.5,2.3))
@@
 forest had lower or higher CRPS (red or blue) compared to the best of the 
-other three models (prespecified GAM, boosted GAM and EMOS).}
+other three models (prespecified GAMLSS, boosted GAMLSS and EMOS).}
 \end{figure}
@@
 (\url{https://R-Forge.R-project.org/projects/partykit/}). The function
-\code{disttree} uses either the \code{mob} or \code{ctree} function 
-as a framework to build a tree. Similarly, the function 
-\code{distforest} is based on the \code{cforest} function which 
-provides a framework to build a distributional forest. All three 
+\code{distfit} fits a distributional model and is called within the
+tree-building function \code{disttree} and the forest-building
+function \code{distforest}. 
+Within \code{disttree} either the \code{mob} or \code{ctree} function 
+can be applied as a framework to build a tree. Similarly, the function 
+\code{distforest} is based on the function \code{cforest}. All three 
 functions, \code{mob}, \code{ctree} and \code{cforest}, are provided 
@@
 \begin{itemize}
-\item a GAMLLS family object from the \textsf{R} package \textbf{gamlss} 
+\item a GAMLLS family object from the \textsf{R} package \textbf{gamlss.dist} 
 (\citealt{Stasinopoulos+Rigby:2007}).
@@
 
-In Section~\ref{ProbabilisticForecasting} the prespecified GAMs were 
+In Section~\ref{ProbabilisticForecasting} the prespecified GAMLSS were 
 built using the function \code{gamlss} from the \textsf{R} package 
-\textbf{gamlss.dist} (\citealt{Stasinopoulos+Rigby:2007}). For the 
-boosted GAMs the function \code{gamboostLSS} from the \textsf{R} 
+\textbf{gamlss} (\citealt{Stasinopoulos+Rigby:2007}). For the 
+boosted GAMLSS the function \code{gamboostLSS} from the \textsf{R} 
 package \textbf{gamboostLSS} (\citealt{Hofner+Mayr+Schmid:2016}) 
@@
 pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAM", ylim = c(0,1.5))
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAM", ylim = c(0,1.5))
+pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
+pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
 pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))

### 2018-03-09 pkg/disttree/inst/draft/draft.Rnw
revised Sections 3.1 and 3.2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2048 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d2e5153443c5ec375fb5f6b5bed13334c7d77794

@@
   URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
-  \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\
+  \phantom{URL: }\url{https://retostauffer.org/},\\
+  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
 
@@
 meteorological stations in a mountainous region in the Alps, covering mainly
-Tyrol, Austria, and adjacent areas. More specifically, a zero-censored Gaussian
+Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{map}).
+More specifically, a zero-censored Gaussian
 distribution is employed to model 24-hour total precipitation so that the
@@
 
-In weather forecasting, various factors can make it challenging to 
-build a well fitting model. Complex physical processes in the 
-atmosphere depend on possibly rapidly changing circumstances. 
-Often it is hard to tell which of these circumstances have a 
-relevant impact on the quantity of interest and what type of 
-effect that can be or how they might interact. A profound 
-basis of information can be obtained from numerical weather 
-prediction (NWP) models as they provide predictions for a 
-large number of meteorological quantities. However, while 
-including too many variables in a model can lead to overfitting, 
-excluding variables in advance can result in a loss of valuable 
-information. Therefore, the task of selecting variables and 
-specifying interactions forms the basis of building an adequate 
-model. Moreover, this model has to be able to deal with a wide 
-range of effects as abrupt changes can appear as well as smooth 
-transitions. Especially if the considered region includes complex 
-terrain, a large grid spacing of NWP models can lead to abrupt 
-local effects within a single grid cell.
-
-As explained in the introduction, distributional forests provide 
-a new approach in probabilistic forecasting by embedding the ideas 
-of flexible parametric modeling into the framework of tree-structured 
-models. In that way, they offer features that could be particularly 
-valuable in probabilistic weather forecasting in order to tackle 
-the above mentioned challenges. Among these features are the ability 
-to deal with a wide variety of effects and an automatic variable 
-selection and detection of interactions. 
-To show their strengths in practice, distributional forests are 
-now applied to probabilistic precipitation forecasts in a 
-mountainous region based on a large number of numerical weather 
-prediction quantities. Their results are compared to those of 
-other zero-censored Gaussian models including a standard ensemble 
-model output approach and two GAMLSS. For one of the GAMLSS the 
-regressors have been selected and interactions have been specified 
-in advance based on subject-matter knowledge and experience from 
-meteorology. The other one is based on a computationally more 
-demanding boosting approach.
+Many weather forecasting models leverage the strengths of modern
+numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}
+and produce high-quality forecasts adapted to specific sites/locations
+through distributional regression models like EMOS
+(ensemble model output statistics, \citealp{Gneiting+Raftery+Westveld:2005}).
+In the case of precipiation forecasting, EMOS explains past precipitation
+sums by the NWP output ``total precipitation'' (\emph{tp}), using its
+ensemble mean as the predictor variable for the location parameter and
+the ensemble standard deviation for the scale parameter.
+
+While this approach alone is already highly effective in the plains,
+it typically does not perform as well in complex terrain due to unresolved
+effects in the NWP system. For example, in the Tyrolean Alps -- considered
+in the following application case study -- the NWP grid cells of $50 \times 50$
+km$^2$ are too coarse to capture mountains, valleys, etc. Therefore,
+it is often possible to substantially improve the EMOS by including
+further predictor variables, either from local observations or from
+other NWP outputs. Unfortunately, it is typically unclear which
+variables are relevant for improving the predictions. Simply including
+all available variables may be computationally burdensome and
+can lead to overfitting but, on the other hand, excluding too many variables
+may result in a loss of valuable information. Therefore, selecting
+the relevant variables and interactions is crucial for improving the
+forecasting model.
+
+In the following, it is illustrated how distributional regression forests
+can solve this problem of automatically selecting relevant variables,
+interactions, and potentially nonlinear effects. For fitting the forest
+only the response distribution and the list of potential predictor variables
+need to be specified (along with a few algorithmic details) and then
+the model fit is determined by the forest in a data-driven way. Here, we
+employ a zero-censored Gaussian distribution and 82~predictor variables
+computed from ensemble means and spreads of various NWP outputs.
+The predictive performance of the forest is compared to three other
+zero-censored Gaussian models: (a) a standard basic EMOS, (b) a GAMLSS
+with prespecified effects and interactions based on meteorological
+knowedge/experience, and (c) a boosted GAMLSS with automatic
+selection of smooth additive terms based on all 82~predictor variables.
 
@@
 \label{data}
-The data set used for training and validation consists of observed 
-precipitation sums provided by the National Hydrographical Service 
-(\citealt{ehyd}) and numerical weather forecasts from the {U.S.} 
-National Oceanic and Atmospheric Administration (NOAA).
-Both, observations and weather forecasts are available from 1985 
-trough the year 2012, focusing on July only in this article.
-
-Observations are available on a daily basis, each representing the 
-total precipitation sum fallen within 24~hours observed at 6~UTC 
-at 95 stations all over Tyrol and its close surrounding. The 
-observations are undergoing a rigorous quality check before made 
-available to the public.  As covariates the second generation 
-reforecast data set of the global ensemble forecast system (GEFS) 
-is used (\citealt{Hamill+Bates+Whitaker:2013}).  This data set 
-consists of an 11-member ensemble based on a fixed version of the 
-numerical model and a horizontal grid-spacing of about $50~\times~50 
-\text{km}^2$ initialized daily at 0~UTC from December 1984 to present 
-providing forecasts on 26 hourly temporal resolution. Each of the 
-11 ensemble members uses slightly different perturbed initial 
-conditions trying to predict the situation specific uncertainty 
-of the atmospheric state.
-
-In particular, 14 basic forecast variables are considered with up 
-to 12 variations of each of these variables being included as 
-covariates. Since an ensemble of individual predictions is provided 
-for each forecast variable, the possible variations include the mean, 
-the standard deviation, the minimum and the maximum over all ensemble 
-members, noted as ensemble mean, ensemble std. deviation, ensemble 
-minimum and ensemble maximum. All of these values can be calculated 
-on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour 
-time window. Another possible option is to consider the mean, minimum 
-or maximum over 24 hours for each individual prediction and then take 
-the ensemble mean or ensemble standard deviation, noted as ensemble 
-mean of mean/minimal/maximal values or ensemble std. deviation of 
-mean/minimal/maximal values respectively. Table~\ref{covariates} 
-shows all variables together with the number of variations and 
-the provided variations listed in the last column. Altogether 
-82~covariates are available in this setting.
-
-\fixme{Mention that dswrf\_mean/sprd\_min is not included?}
-
-%%% <TH> hm, now we _are_ using a transformation model without estimating the
-%%% transformation. </TH>
-To remove large parts of the skewness of precipitation data, 
-a power transformation has been used frequently (Box and Cox 1964). 
-In literature, cubic (\citealt{Stidd:1973}) or square root 
-(\citealt{Hutchinson:1998}) transformations have often been 
-suggested but may vary for different climatic zones or temporal 
-aggregation periods. In \cite{Stauffer+Mayr+Messner:2017} 
-optimization methods have been applied to find the best transformation 
-factor which resulted in a value of $\frac{1}{1.6}$. Since the 
-observation data set used for this optimization contains the 
-observations considered in this article the same factor is 
-chosen here for the power transformation of the response variable 
-\emph{tp} (total precipitation). The resulting transformed 
-variable is expected to be normally distributed and left-censored 
-at censoring point 0. For simplicity the same transformation has 
-been chosen for the variable \emph{cape} (convective available 
-potential energy).
-
-%\newpage
+
 \begin{table}[t!]
@@
 
+Training and validation data consist of observed 
+daily precipitation sums provided by the National Hydrographical Service 
+(\citealt{ehyd}) and numerical weather forecasts from the U.S.~National
+Oceanic and Atmospheric Administration (NOAA).
+Both, observations and forecasts are available for 1985--2012 and
+the analysis is exemplified using July as the month with the most precipitation
+in Tyrol.
+
+Observations are obtained for 95~stations all over Tyrol and surroundings,
+providing 24~hour precipitation sums measured at 6~UTC and rigorously
+quality-checked by the National Hydrographical Service. NWP outputs
+are obtained from the second generation reforecast data set of the
+global ensemble forecast system \citep[GEFS,][]{Hamill+Bates+Whitaker:2013}. 
+This data set consists of an 11-member ensemble based on a fixed version of the 
+numerical model and a horizontal grid spacing of about $50 \times 50$ 
+km$^2$ initialized daily at 0~UTC from December 1984 to present 
+providing forecasts on 26 hourly temporal resolution. Each of the 
+11 ensemble members uses slightly different perturbed initial 
+conditions trying to predict the situation specific uncertainty 
+of the atmospheric state.
+
+From the GEFS outputs 14 basic forecast variables are considered with up 
+to 12 variations of each of these variables such as mean/maximum/minimum
+or different aggregation time ranges, etc. A detailed overview is
+provided in Table~\ref{covariates}, yielding 82~predictor variables in total.
+\fixme{Mention that dswrf\_mean/sprd\_min is not included?}
+
+To remove large parts of the skewness of precipitation data, 
+a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
+cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
+However, the power parameter may vary for different climatic zones or temporal 
+aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
+in their choice of $\frac{1}{1.6}$ as a suitable power parameter for the
+region of Tyrol. The same power transformation is applied to both the
+observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
+``convective available potential energy'' (\emph{cape}).
+
 \subsection{Models and evaluation}
-
-Next to distributional forests three other methods are applied to this data set:
-
-\begin{itemize}
-
-\item Ensemble model output statistics 
-(EMOS, \citealp{Gneiting+Raftery+Westveld:2005}): The 
-location parameter and the logarithm of the scale parameter depend 
-linearly on the ensemble mean forecast of total daily precipitation 
-sums and the corresponding ensemble standard deviation respectively.
-The choice of the logarithmic function as link function for the scale 
-parameter has been made based on the results of 
-\cite{Gebetsberger+Messner+Mayr:2017} where different link functions 
-have been compared. Moreover, testing the identity, quadratic and 
-logarithmic function as link functions in this setting supported 
-this choice as the logarithmic function led to the best performance 
-of the EMOS model in this application.
-\item Prespecified GAMLSS: For each distribution parameter the selection 
-of covariates and the specification of interactions is done in 
-advance, both based on meteorological expert knowledge following 
-\cite{Stauffer+Umlauf+Messner:2017}. Among the 
-available 82 regressors, 9 are selected for the location parameter 
-and 6 for the scale parameter. Additionally, for each distribution 
-parameter one interaction term is specified in which convective 
-available energy is interacting with total precipitation (with the 
-ensemble mean for the location parameter and with the ensemble 
-standard deviation for the scale parameter). These interactions 
-might capture events with a high potential for moderate or severe 
-thunderstorms with large amounts of precipitation. By putting 
-extra effort in the specification of the formulas the prespecified 
-GAM is provided with additional information and is therefore 
-expected to profit from this advantage.
-\item Boosted GAMLSS: Non-homogeneous boosting 
-\citep{Messner+Mayr+Zeileis:2017} is used for an automatic 
-selection of covariates \citep{Hofner+Mayr+Schmid:2016}. 
-The idea of this boosting approach is to iteratively minimize 
-a loss function where in each step only the coefficient leading 
-to the highest improvement of the current model is updated. As 
-the initial coefficients are all zero, only the most important 
-covariates are included with nonzero coefficients while less 
-relevant variables are ignored if the procedure is stopped before 
-convergence.
-\end{itemize}
-
-%\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAMLSS.
-Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}
-
-%\fixme{A description of the boosting approach is currently missing completely.}
-
-The selected methods and their specifications are listed in Table~\ref{model_specification}.
 
@@
 \hline
-Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$) \\
-\hline
-\textbf{Distributional forest} & recursive    & all & all \\ 
-                & partitioning &     &     \\ 
-\hline
-\textbf{Prespecified GAMLSS} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
-                            & in each & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean},\\
+Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$)                                  \\ \hline
+Distributional forest & recursive    & all                & all                           \\ 
+                      & partitioning &                    &                               \\ \hline
+EMOS                  & linear       & \emph{tp{\_}mean}  & \emph{tp{\_}sprd}             \\ \hline
+Prespecified GAMLSS   & spline       & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},            \\
+                      & in each      & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean}, \\
  & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
@@
  & & \emph{tdiff500850{\_}mean}, & \\
- & & \emph{msl{\_}diff} & \\
-\hline
-\textbf{Boosted GAMLSS} & spline  & all & all \\
-                       & in each &     &     \\
-\hline
-\textbf{EMOS} & linear & \emph{tp{\_}mean} & \emph{tp{\_}sprd} \\
-\hline
+ & & \emph{msl{\_}diff} & \\ \hline
+Boosted GAMLSS        & spline  & all & all \\
+                      & in each &     &     \\ \hline
 \end{tabular}
@@
 
-In order to compare the performance of the chosen methods, the continuous 
-ranked probability score (CRPS) is calculated for each model. 
-\fixme{Adding reference for CPRS}
-
-The CRPS measures the accuracy of a probabilistic model. 
-For one observation $y$ with predicted distribution 
-function $F$ the CRPS value is calculated by
+The following zero-censored Gaussian regression models are employed
+in the empirical case study, see also Table~\ref{model_specification} for
+further details:
+%
+\begin{itemize}
+
+\item \emph{Distributional forest:} All 82~predictor variables are
+  considered for learning a forest of 100 trees. Bootstrap
+  sampling is employed for each tree using a third of the predictors
+  in each split of the tree (``mtry''). Parameters are estimated by
+  adaptive local likelihood based on the forest weights, as described
+  in Section~\ref{Methodology}.
+
+\item \emph{EMOS:} Standard ensemble model output statistics are used
+  with the total precipitation ensemble mean as regressor in the location
+  submodel and the ensemble standard deviation in the scale submodel.
+  The parameters are estimated by maximum likelihood, using an identity
+  link for the location and a log link for the scale
+  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.
+
+\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
+  the most relevant predictors, based on meteorological expert knowledge
+  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
+  on the 82~available variables, 8 terms are included in the location
+  submodel and 6 in the scale submodel. Both involve an interaction of
+  \emph{tp} and \emph{cape} in the afternoon (between 12 and 18 UTC)
+  to capture the potential for thunderstorms that frequently occur in
+  summer afternoons in the Alps. The model is estimated by maximum
+  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.
+
+\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
+  automatically from all 82~available variables, using non-cyclic boosting
+  for parameter estimation \citep{Hofner+Mayr+Schmid:2016,Messner+Mayr+Zeileis:2017}.
+  This updates the predictor terms for the location or scale submodels iteratively
+  by maximizing the log-likelihood only for the variable yielding the
+  highest improvement. The iteration stops early -- before fully maximizing
+  the in-sample likelihood -- based on a (computationally intensive)
+  out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
+  the number of boosting iterations (``mstop'') is: $50, 75, \dots, 975, 1000$.
+
+\end{itemize}
+
+The predictive performance in terms of full probabilistic forecasts is
+assessed using the continuous ranked probability score (CRPS, \fixme{reference}).
+For each of the models this assesses the discrepancy of the
+predicted distribution function $F$ from the observation $y$ by
 $$
@@
 $$
-where \(\mathbf{1}\) is the indicator function.
-%%% <TH> y > z??? </TH>
-
-In this application the CRPS is evaluated out of sample, 
-e.g. on a part of the data set that has not been considered to 
-build the model. The mean CRPS over this subset of the data 
-is used for validation. Since the CRPS can be seen as a 
-generalization of the mean absolute error, a low value indicates 
-a good fit. In applications, one big advantage in terms of 
-interpretability is that the resulting value is on the scale of 
-the observations.
-
-For the purpose of comparing the models' performances with respect 
-to a reference model the so-called skill score of the CRPS can be 
-calculated. In that way a measure of how much a method improves the
-model fit in terms of CRPS compared to the reference method is obtained.
-$$
-\text{SkillScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
+where \(\mathbf{1}\) is the indicator function. In the subsequent
+applications, the mean CRPS is always evaluated out of sample, 
+either using cross-validation or a hold-out data set (2009--2012)
+that was not used for learning (1985--2008). CRPS is a proper
+scoring rule \fixme{citation} and lower values indicate a better fit.
+
+To assess differences in the improvement of the forests and GAMLSS
+models over the basic EMOS, a CRPS-based skill score with EMOS as
+the reference method is computed as well:
+$$
+\text{SkillScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{EMOS}}}.
 $$
@@
 par(mfrow = c(2,2))
-
 qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
 qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
 qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
 @
@@
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend(x = 9.8, y = 47.815, pch=c(21,24,25,22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
+  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
   text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
@@
 \begin{itemize}
-\item a GAMLLS family object from the \textsf{R} package \textbf{gamlss.dist} 
+\item a GAMLSS family object from the \textsf{R} package \textbf{gamlss.dist} 
 (\citealt{Stasinopoulos+Rigby:2007}).
@@
 pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
+pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
 pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
 pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
-pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
 

### 2018-03-12 pkg/disttree/inst/draft/draft.Rnw
added citation 'Strictly Proper Scoring Rules, Prediction, and Estimation'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2049 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ca75cf4cb10efe684d275e254c35bc26ba2d388f
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2051 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9df3d18f4f2d11499fb3b49815c955313053d37a

@@
 established based on data from the same month in the years 1985--2012 along with
-82~covariates derived from a wide range of NWP ensemble outputs. As
+80~covariates derived from a wide range of NWP ensemble outputs. As
 Figure~\ref{axams_onestation} shows, the station-wise models yield a full
@@
 \citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
-82~available regressors.
+80~available regressors.
 
@@
 the model fit is determined by the forest in a data-driven way. Here, we
-employ a zero-censored Gaussian distribution and 82~predictor variables
+employ a zero-censored Gaussian distribution and 80~predictor variables
 computed from ensemble means and spreads of various NWP outputs.
@@
 knowedge/experience, and (c) a boosted GAMLSS with automatic
-selection of smooth additive terms based on all 82~predictor variables.
+selection of smooth additive terms based on all 80~predictor variables.
 
@@
 or different aggregation time ranges, etc. A detailed overview is
-provided in Table~\ref{covariates}, yielding 82~predictor variables in total.
-\fixme{Mention that dswrf\_mean/sprd\_min is not included?}
+provided in Table~\ref{covariates}, yielding 80~predictor variables in total.
+(Since the minimum of the ``variable downwards short wave radiation flux'' (\emph{dswrf}) over 24 hours is always 0, the ensemble mean and standard deviation of these minimal values are not considered.) \fixme{Mention that dswrf\_mean/sprd\_min is not included?}
 
@@
 
-\item \emph{Distributional forest:} All 82~predictor variables are
+\item \emph{Distributional forest:} All 80~predictor variables are
   considered for learning a forest of 100 trees. Bootstrap
@@
   following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
-  on the 82~available variables, 8 terms are included in the location
+  on the 80~available variables, 8 terms are included in the location
   submodel and 6 in the scale submodel. Both involve an interaction of
@@
 \item \emph{Boosted GAMLSS:} Smooth additive splines are selected
-  automatically from all 82~available variables, using non-cyclic boosting
+  automatically from all 80~available variables, using non-cyclic boosting
   for parameter estimation \citep{Hofner+Mayr+Schmid:2016,Messner+Mayr+Zeileis:2017}.
@@
 that was not used for learning (1985--2008). CRPS is a proper
-scoring rule \fixme{citation} and lower values indicate a better fit.
+scoring rule \citep{Gneiting+Raftery:2007} and lower values indicate a better fit.
 
@@
 day of each of the four successive years have been shown as a motivational 
-example for a probabilistic forecast. Staying with this fitted model we are 
-now going to analyze variable importance of the distributional forest model 
-and compare its results to those of the prespecified and the boosted GAMLSS 
+example for a probabilistic forecast. Staying with this fitted model,
+variable importance of the distributional forest model will be analyzed and its
+performance will be compared to those of the prespecified and the boosted GAMLSS 
 and the EMOS model based on residual QQ-plots.
@@
 forest for station Axams learned on the first 24 years are listed in 
-Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation 
-is the most influential variable. In particular, five variations of this variable 
-are among the top ten of the list of covariates ordered by variable importance.
+Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation (\emph{tp}) is the most influential variable. In particular, five variations of 
+this variable are among the top ten of the list of covariates ordered by 
+variable importance.
 
@@
 represented by the horizontal line at height 0. This CRPS skill score boxplot 
-shows that, in this setting, both GAMLSS and the distributional forest perform 
-better than the EMOS model. While the two GAMLSS lead to an improvement of 
-around 4 percent with the boosted version being slightly ahead, the 
+shows that, in this setting, both GAMLSS models and the distributional forest 
+perform better than the EMOS model. While the two GAMLSS lead to an improvement 
+of around 4 percent with the boosted version being slightly ahead, the 
 distributional forest even reaches an improvement rate close to 6 percent.
@@
 Until now the focus has been on one observation station only. To test and 
-compare the four methods under different geographical circumstances we are 
-now going to apply them on all 95 stations. At each of them, the first 
+compare the four methods under different geographical circumstances they
+are applied on all 95 stations. For each station, the first 
 24 years are used as learning data set and predictions are made for the 
@@
 for each method at each station. In Figure~\ref{boxplot_crps_all} the boxplots 
-of the CRPS skill score values are plotted and show that the distributional 
+of the CRPS skill score values are plotted illustrating that the distributional 
 forest leads to the highest improvement averaged over all stations.
@@
 Even though for some stations, especially in East Tyrol, one of the two GAMLSS
-and for a few others the EMOS model leads to the best results, for the 
+models and for a few others the EMOS model leads to the best results, for the 
 majority of stations the distributional forest performs best.
@@
 can be seen as a good and recommendable compromise being able to deal 
-with a wide range of modelling situations in a reasonable way.
+with a wide range of modeling situations in a reasonable way.
 

### 2018-03-12 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2052 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 231a95db8ddaca3a64bf17d59717eb05b4c68937

@@
 of determining the distribution parameter~$\theta$. This is done by the 
-following maximization.
+following maximization based on a learning sample $\{y_i\}_{i=1,\ldots,n}$ 
+of the response variable $Y$.
 $$
@@
 $$
-where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response 
-variable $Y$.
-
+which is equal to solving 
+$$
+\sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = 0.
+$$
 In that way, a whole distribution is specified with all its features, 
@@
 \label{disttree}
-Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ 
+Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$, 
 the idea is to use the information provided by this model together with a 

### 2018-03-17 pkg/disttree/inst/MIP-Seminar/slides.Rnw
first draft of slides for PhD-Seminar
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2053 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: be6eb4965a7015ec7f3014964e62ad0fce63684d

--- 2017-12-13 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
@@
 
-\title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
+\title{Flexible Probabilistic Forecasting with Distributional Trees and Forests}
+\author{Lisa Schlosser}
+%\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
-%\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
-\URL{https://R-Forge.R-project.org/projects/partykit/}
+%\URL{https://R-Forge.R-project.org/projects/partykit/}
 \headerimage{3}
@@
 %% includes a replacement for \usepackage{Sweave}
+\usepackage{changepage}
 \usepackage{amsmath,tikz}
-\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
+\usepackage{calc}
+\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
 \graphicspath{{plots/}}
@@
 library("disttree")
-library("Formula")
 #library("partykitR1")
@@
 ## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
-
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
+pal <- hcl(c(10, 128, 260), 100, 50)
+names(pal) <- c("forest", "tree", "gamlss")
+
+pallight <- hcl(c(10, 128, 260), 100, 50, alpha = 0.25)
+names(pallight) <- c("forest", "tree", "gamlss")
 
 transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-## define distribution list:
-# dist_list_normal
-{
-
-  dist_list_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {
-
-    val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
-    if(!log) val <- exp(val)
-
-    # par <- c(eta[1], exp(eta[2]))
-    # val <- dnorm(y, mean = par[1], sd = par[2], log = log)
-
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE) {
-
-    score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
-                   -1 + exp(-2*eta[2] + log((y-eta[1])^2)))
-
-    # par <- c(eta[1], exp(eta[2])) 
-    # score <- cbind(1/par[2]^2 * (y-par[1]), 
-    #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))
-
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
-    d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-eta[1]) * exp(-2*eta[2]), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    d2ld.etasigma2 <- sum(weights * (-2)*exp(log((y-eta[1])^2) - 2*eta[2]), na.rm = TRUE)
-
-    # par <- c(eta[1], exp(eta[2]))                           
-    # d2ld.etamu2 <- sum(weights * rep.int(-1/par[2]^2, ny))
-    # d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-par[1])/par[2]^2), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    # d2ld.etasigma2 <- sum(weights * (-2)*(y-par[1])^2/par[2]^2, na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etamu.d.etasigma, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- pnorm
-  qdist <- qnorm
-  rdist <- rnorm
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    if(is.null(weights)) {
-      mu <- mean(y)
-      sigma <- sqrt(1/length(y) * sum((y - mu)^2))
-    } else {
-      mu <- weighted.mean(y, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (y - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- TRUE
-
-  dist_list_normal <- list(family.name = "Normal Distribution",
-                           ddist = ddist, 
-                           sdist = sdist, 
-                           hdist = hdist,
-                           pdist = pdist,
-                           qdist = qdist,
-                           rdist = rdist,
-                           link = link, 
-                           linkfun = linkfun, 
-                           linkinv = linkinv, 
-                           linkinvdr = linkinvdr,
-                           startfun = startfun,
-                           mle = mle,
-                           gamlssobj = FALSE,
-                           censored = FALSE
-  )
-}
-
-# dist_list_cens_normal
-{
-
-  dist_list_cens_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
-    par <- c(eta[1], exp(eta[2]))
-    val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
-    if(sum) {
-      if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
-    par <- c(eta[1], exp(eta[2]))
-    # y[y==0] <- 1e-323
-
-    score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
-    score <- cbind(score_m, score_s)
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y)[1])
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-      #if(any(is.nan(score))) print(c(eta, "y", y))
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    par <- c(eta[1], exp(eta[2]))                           
-    # y[y==0] <- 1e-323
-
-    d2mu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    d2sigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-    dmudsigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu.sigma", left = left, right = right) # FIX: order?
-    dsigmadmu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma.mu", left = left, right = right) # FIX: order?
-    dsigma <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-
-    d2ld.etamu2 <- sum(weights * d2mu, na.rm = TRUE)
-    d2ld.etamu.d.etasigma <- sum(weights * dmudsigma * par[2], na.rm = TRUE)
-    d2ld.etasigma.d.etamu <- sum(weights * dsigmadmu * par[2], na.rm = TRUE)
-    d2ld.etasigma2 <- sum(weights * (d2sigma * exp(2*eta[2]) + dsigma * par[2]), na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etasigma.d.etamu, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- function(q, eta, lower.tail = TRUE, log.p = FALSE) crch:::pcnorm(q, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  qdist <- function(p, eta, lower.tail = TRUE, log.p = FALSE) crch:::qcnorm(p, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  rdist <- function(n, eta) crch:::rcnorm(n, mean = eta[1], sd = eta[2], left = left, right = right)
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    yc <- pmax(0,y)  # optional ?
-    if(is.null(weights)) {
-      mu <- mean(yc)
-      sigma <- sqrt(1/length(yc) * sum((yc - mu)^2))
-    } else {
-      mu <- weighted.mean(yc, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (yc - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- FALSE
-
-  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
-                                ddist = ddist, 
-                                sdist = sdist, 
-                                hdist = hdist,
-                                pdist = pdist,
-                                qdist = qdist,
-                                rdist = rdist,
-                                link = link, 
-                                linkfun = linkfun, 
-                                linkinv = linkinv, 
-                                linkinvdr = linkinvdr,
-                                startfun = startfun,
-                                mle = mle,
-                                gamlssobj = FALSE,
-                                censored = TRUE
-  )
-}
 
@@
 
-\subsection{Motivation}
-
 \begin{frame}[fragile]
-\frametitle{Motivation}
-\vspace{-0.41cm}
-\begin{figure}[!htb]
-\minipage{0.29\textwidth}
+\frametitle{Statistical modeling}
+\vspace{0.2cm}
+Approximation of reality based on a set of observations.\\
+\vspace{0.2cm}
+\visible<1, 4-> {Assumption:} 
+\vspace{-1.1cm}
 \begin{center}
+\begin{tikzpicture}
+\visible<1, 4->{
+\node[] at (0,0.4) {X};
+}
+\visible<2-3>{
+\node[inner sep=0pt] (today) at (-1.9,-0.2)
+    {\includegraphics[width=.38\textwidth]{today.png}};
+}
+\draw[->] (0.4,0.4)--(0.9,0.4);
+\draw[] (1,0) rectangle (2.5,0.8);
+\node[font=\scriptsize] at (1.75,0.4) {nature};
+\draw[->] (2.6,0.4)--(3.1,0.4);
+\visible<1, 4->{
+\node[] at (3.5,0.4) {Y};
+}
+\visible<2-3>{
+\node[inner sep=0pt] (future) at (5.4,-0.2)
+    {\includegraphics[width=.38\textwidth]{future.png}};
+}
+\end{tikzpicture}
+\end{center}
+\visible<3>{
+\vspace{0.5cm}
+Y = daily precipitation sums\\
+\vspace{0.1cm}
+X = observed precipitation of today,\\
+\qquad current temperature/humidity/\dots,\\
+\qquad physical model output
+}
+
+\vspace{-3.5cm}
+\visible<5->{
+\textbf{Application/Motivation:}
+\begin{itemize}
+\item Prediction.
+\item Analysis of dependencies and interactions.
+\end{itemize}
+}
+\vspace{0.5cm}
+\visible<6->{
+\textbf{Two main approaches / ``cultures'':}
+\begin{itemize}
+\item Data modeling.
+\item Algorithmic modeling.
+\end{itemize}
+}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Statistical modeling}
+\vspace{0.1cm}
+\textbf{Data modeling:}
+\vspace{0.3cm}
+\begin{center}
+\begin{tikzpicture}
+\node[] at (0,0.4) {X};
+\draw[->] (0.4,0.4)--(0.9,0.4);
+\draw[] (1,0) rectangle (2.5,0.8);
+\node[font=\scriptsize] at (1.75,0.4) {f(X; par)};
+\draw[->] (2.6,0.4)--(3.1,0.4);
+\node[] at (3.5,0.4) {Y};
+\end{tikzpicture}
+\end{center}
+
+\begin{itemize}
+\item Approximate data generating process up to a few unknown processes.
+\item Mathematical equations, model assumptions.
+\item Data: estimation of parameters.
+\end{itemize}
+
+\vspace{0.2cm}
+
+\visible<2-3>{
+\begin{minipage}{0.65\textwidth}
+\textbf{Example:} Linear model\\
+$$\visible<3>{\mathbb{E}(Y|X) = } f(X; \alpha) = \alpha_0 + \alpha_1 \cdot X$$  
+\end{minipage}
+\vspace{-0.6cm}
+\hfill
 <<motivation_GLM, echo=FALSE, results=hide>>=
@@
 @
-\visible<2->{
-<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
+\begin{minipage}{0.3\textwidth}
+<<plot_motivation_GLM_single, fig=TRUE, echo=FALSE>>=
 par(mar=c(2,0,2,0))
@@
 @
-}
-\end{center}
-\endminipage
-\visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
-\begin{center}
+\end{minipage}
+}
+
+\vspace{-2.35cm}
+\visible<4>{
+\begin{minipage}{0.65\textwidth}
+\textbf{Example:} Generalized additive model\\
+$$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
+\end{minipage}
+\hfill
 <<motivation_GAM, echo=FALSE, results=hide>>=
@@
 @
-\visible<3->{
+\begin{minipage}{0.3\textwidth}
 <<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
@@
 @
-}
-\end{center}
-\endminipage
-\visible<4->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
-\begin{center}
+\end{minipage}
+}
+
+\vspace{-3cm}
+\visible<5>{
+\begin{minipage}{0.65\textwidth}
+\vspace{0.8cm}
+\textbf{Example:} Generalized additive model\\
+for location, scale, and shape
+\begin{align*}
+\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
+\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
+\end{align*}
+\end{minipage}
+\hfill
 <<motivation_GAMLSS, echo=FALSE, results=hide>>=
@@
 @
-\visible<4->{
+\begin{minipage}{0.3\textwidth}
 <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
@@
 @
-
-}
-\end{center}
-\endminipage
-
-\vspace{0.5cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\visible<2->{
-LM, GLM\\
-\vspace{0.5cm}
-\code{lm}\\
-\code{glm}\\
-\vspace{1.5cm}}
-\end{center}
-\endminipage
-\hspace{1.1cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\visible<3->{
-GAM\\
-\vspace{0.5cm}
-\code{mgcv}\\
-\code{VGAM}\\
-$\ldots$\\
-\vspace{1cm}}
-\end{center}
-\endminipage
-\hspace{1.1cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\visible<4->{
-GAMLSS\\
-\vspace{0.5cm}
-\code{gamlss}\\
-\code{mgcv}\\
-\code{VGAM}\\
-\code{gamboostLSS}\\
-$\ldots$}
-\end{center}
-\endminipage
-\end{figure}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Motivation}
-\vspace{-0.2cm}
-\begin{figure}[!htb]
-\minipage{0.29\textwidth}
-\begin{center}
+\end{minipage}
+}
+\end{frame}
+
+<<raintree, echo=FALSE, results=hide>>=
+load("weather.rda")
+#d <- data.frame(rain24 = numeric(0), rain = logical(0), wind = numeric(0), 
+#                humidity = numeric(0), airpressure = numeric(0), temperature = numeric(0))
+
+sp_rain <- partysplit(2L, index = 1:2)
+sp_humidity <- partysplit(4L, breaks = 80)
+
+pn <- partynode(1L, split = sp_rain, kids = list(
+  partynode(2L, split = sp_humidity, kids = list(
+      partynode(3L, info = "5 mm/24h"),
+      partynode(4L, info = "8 mm/24h"))),
+  partynode(5L, info = "18 mm/24h")))
+
+p <- party(pn, weather)
+@
 <<motivation_regtree, echo=FALSE, results=hide>>=
@@
 @
-\visible<1->{
+<<motivation_randforest, echo=FALSE, results=hide>>=
+## Random Forest
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
+@
+
+\begin{frame}[fragile]
+\frametitle{Statistical modeling}
+\textbf{Algorithmic modeling:}
+
+\begin{center}
+\begin{tikzpicture}
+\node[] at (0,0.4) {X};
+\draw[->] (0.3,0.4)--(0.8,0.4);
+\draw[] (1,0) rectangle (2.5,0.8);
+\node[font=\scriptsize] at (1.75,0.4) {nature};
+\draw[->] (2.7,0.4)--(3.2,0.4);
+\node[] at (3.5,0.4) {Y};
+\draw[<->] (0.1,0.1) .. controls (0.2, -0.4) and (0.4, -0.5) .. (0.8,-0.5);
+\draw[<->] (2.7,-0.5) .. controls (3.1, -0.5) and (3.3, -0.4) .. (3.4,0.1);
+\node[font=\scriptsize] at (1.75,-0.5) {algorithm};
+\end{tikzpicture}
+\end{center}
+
+\begin{itemize}
+\item Flexible approximation with no/few assumptions.
+\item Data-driven model fitting.
+\end{itemize}
+
+\visible<2-3>{
+\begin{minipage}{0.65\textwidth}
+\vspace{0.5cm}
+\textbf{Example:} Tree model\\
+\vspace{-0.8cm}
+\begin{center}
+\resizebox{0.6\textwidth}{!}{
+<<plot_raintree, fig=TRUE, echo=FALSE, width=5, height=3>>=
+paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+plot(p,
+     tp_args = list(FUN = identity, width = 9, fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]),
+     drop_terminal = TRUE, tnex = 1)
+@
+}
+\end{center}
+\end{minipage}
+}
+\visible<3>{
+\hfill
+\begin{minipage}{0.3\textwidth}
 <<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
@@
 @
-}
-\end{center}
-
-\endminipage
-\visible<2->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\end{minipage}
+}
+
+\vspace{-3cm}
+\visible<4-5>{
+\begin{minipage}{0.65\textwidth}
+\textbf{Example:} Forest model\\
 \begin{center}
-<<motivation_randforest, echo=FALSE, results=hide>>=
-## Random Forest
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-@
-\visible<2->{
-<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
-@
-}
-\end{center}
-
-\endminipage
-\visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
-  \visible<3->{
-  \begin{center}
-  \begin{tikzpicture}
-  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
-  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
-  \end{tikzpicture}
-  \end{center}}
-\endminipage
-
-\minipage{0.29\textwidth}
-\begin{center}
-\vspace{0.0cm}
-\visible<1->{
-Regression Tree\\
-\vspace{0.5cm}
-\resizebox{0.3\textwidth}{!}{
-\begin{tikzpicture}
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (1, 2) {};
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0.5, 1) {};
-  \draw[-, line width=1pt] (n0) -- (n1);
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (1.5, 1) {};
-  \draw[-, line width=1pt] (n0) -- (n2);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (1, 0) {};
-  \draw[-, line width=1pt] (n2) -- (n3);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (2, 0) {};
-  \draw[-, line width=1pt] (n2) -- (n4);
-\end{tikzpicture}}
-\vspace{0.3cm}\\
-\code{rpart}\\
-\code{party(kit)}\\
-\vspace{1cm}}
-\end{center}
-\endminipage
-\hspace{0.65cm}
-\minipage{0.29\textwidth}
-\begin{center}
-\vspace{0.0cm}
-\visible<2->{
-Random Forest\\
-\vspace{0.5cm}
-\resizebox{0.9\textwidth}{!}{
+\vspace{-0.2cm}
+\resizebox{0.5\textwidth}{!}{
 \begin{tikzpicture}
@@
 }
-\vspace{0.5cm}\\
-\code{randomForest}\\
-\code{ranger}\\
-\code{party(kit)}\\
-$\ldots$}
 \end{center}
-\endminipage
-\hspace{0.65cm}
-\minipage{0.28\textwidth}
-\begin{center}
-\vspace{0.68cm}
-\visible<3->{
-Distributional trees and forests\\
-\vspace{1.7cm}
-\code{disttree}\\
-based on \code{partykit}\\
-\vspace{1cm}}
-\end{center}
-\endminipage
-\end{figure}
+\end{minipage}
+}
+\visible<5>{
+\hfill
+\begin{minipage}{0.3\textwidth}
+<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
+@
+\end{minipage}
+}
 \end{frame}
@@
 
-\textbf{Distributional:} \\
+\textbf{Synthesis:} \\
 \vspace{0.1 cm}
-\hspace{0.5cm} Specify the complete probability distribution\\
-\hspace{0.5cm} (including location, scale and shape).
-
-\medskip
-
-{\bf Tree:}
 \begin{itemize}
-  \item Automatic detection of steps and abrupt changes. %(data driven)
-  \item Capture non-linear and non-additive effects and interactions.
+  \item Probabilistic forecast\\
+  $\rightarrow$ Distributional approach
+  \item Automatic detection of steps and abrupt changes,\\
+  automatic selection of variables and interactions\\
+  $\rightarrow$ Tree approach
+  \item Smoothing and regularization\\
+  $\rightarrow$ Forest approach
 \end{itemize}
-
-\medskip
-
-{\bf Forest:}
-\begin{itemize}
-  \item Smoother effects.
-  \item Stabilization and regularization of the model.
-\end{itemize}
-\end{frame}
-
-\subsection{Building Distributional Trees and Forests}
+\end{frame}
+
+<<dgp_tree, echo=FALSE, results=hide>>=
+data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
+names(data) <- c("x","x","x")
+fig <- party(
+  partynode(1L,
+            split = partysplit(2L, breaks = 0.4),
+            kids = list(
+              partynode(2L, info = c(
+                "n = 200",
+                "   True parameters:   ",
+                expression(mu == '4'),
+                expression(sigma == '1')
+              )),
+              partynode(3L,
+                        split = partysplit(3L, breaks = 0.8),
+                        kids = list(
+                          partynode(4L, info = c(
+                            "n = 200",
+                            "   True parameters:   ",
+                            expression(mu == '12'),
+                            expression(sigma == '3')
+                          )),
+                          partynode(5L, info = c(
+                            "n = 100",
+                            "   True parameters:   ",
+                            expression(mu == '4'),
+                            expression(sigma == '3')
+                          )))))),
+  data
+)
+
+node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
+    gp = gpar()) 
+{
+    meta <- obj$data
+    nam <- names(obj)
+    extract_label <- function(node) {
+        if (is.terminal(node)) 
+            return(rep.int("", 2L))
+        varlab <- character_split(split_node(node), meta)$name
+        if (abbreviate > 0L) 
+            varlab <- abbreviate(varlab, as.integer(abbreviate))
+        if (pval) {
+            nullna <- function(x) is.null(x) || is.na(x)
+            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
+                silent = TRUE))
+            pval <- if (inherits(pval, "try-error")) 
+                FALSE
+            else pval
+        }
+        if (pval) {
+            pvalue <- node$info$p.value
+            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
+                paste("p =", round(pvalue, digits = 3L)))
+        }
+        else {
+            plab <- ""
+        }
+        return(c(varlab, plab))
+    }
+    maxstr <- function(node) {
+        lab <- extract_label(node)
+        klab <- if (is.terminal(node)) 
+            ""
+        else unlist(lapply(kids_node(node), maxstr))
+        lab <- c(lab, klab)
+        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
+        lab <- lab[which.max(nchar(lab))]
+        if (length(lab) < 1L) 
+            lab <- ""
+        return(lab)
+    }
+    nstr <- maxstr(node_party(obj))
+    if (nchar(nstr) < 6) 
+        nstr <- "aAAAAa"
+    rval <- function(node) {
+        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
+            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
+            height = unit(3, "lines"), name = paste("node_inner", 
+                id_node(node), sep = ""), gp = gp)
+        pushViewport(node_vp)
+        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
+            seq(0.8, 1, by = 0.01))
+        yell <- sqrt(xell * (1 - xell))
+        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
+        lab <- extract_label(node)
+        fill <- rep(fill, length.out = 2L)
+        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
+            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
+        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
+            "lines"))
+        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
+        #    "lines"))
+        grid.text(lab[2L], y = unit(1, "lines"))
+        if (id) {
+            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
+                "npc"), width = max(unit(1, "lines"), unit(1.3, 
+                "strwidth", nam[id_node(node)])), height = max(unit(1, 
+                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
+            pushViewport(nodeIDvp)
+            grid.rect(gp = gpar(fill = fill[2]))
+            grid.text(nam[id_node(node)])
+            popViewport()
+        }
+        upViewport()
+    }
+    return(rval)
+}
+
+class(node_inner_ext) <- "grapcon_generator"
+
+@
+
+<<tree, echo=FALSE, results=hide>>=
+set.seed(7)
+nobs <- 500
+x <- runif(nobs, 0, 1)
+mu <- sigma <- ytrue <- numeric(length = nobs)
+for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
+for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
+y <- rnorm(nobs, mean = mu, sd = sigma)
+#y <- rcnorm(nobs, mean = mu, sd = sigma, left = 0)
+ytrue <- mu
+data <- data.frame(cbind(y,x, ytrue))
+tree <- disttree(y ~ x, data = data, family = NO(), type.tree = "mob")
+#tree <- disttree(y ~ x, data = data, family = dist_list_cens_normal)
+@
+
+\begin{frame}[fragile]
+\frametitle{Fitting a Tree}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+<<plottree_estpar, fig=TRUE, echo=FALSE>>=
+# function for output in terminal panels
+FUN <- function (x) 
+{
+  cf <- x$coefficients
+  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
+  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
+                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
+}
+
+paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+
+## plot version using FUN and tree of class 'disttree'
+plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
+     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Fitting a Tree}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+<<plottree_dens, fig=TRUE, echo=FALSE>>=
+node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
+                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
+                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
+{
+  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
+  ylines <- 1.5
+  if (is.null(xscale)) xscale <- c(-5.1,22.5)
+  if (is.null(yscale)) yscale <- c(-0.05,0.45)
+  xr <- xscale
+  yr <- yscale
+
+  if (horizontal) {
+    yyy <- xscale
+    xscale <- yscale
+    yscale <- yyy
+  }
+
+  rval <- function(node) {
+    yrange <- seq(from = -20, to = 90)/4
+    ydens <- node$info$object$ddist(yrange)
+
+    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3, 
+                                            widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")), 
+                                            heights = unit(c(1, 1), c("lines", "null"))), 
+                       width = unit(1, "npc"), 
+                       height = unit(1, "npc") - unit(2, "lines"), 
+                       name = paste("node_density",node$id, sep = ""))
+    pushViewport(top_vp)
+    grid.rect(gp = gpar(fill = "white", col = 0))
+    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
+    pushViewport(top)
+    mainlab <- paste(ifelse(id, paste("Node", node$id, "(n = "), "n = "), node$info$nobs, ifelse(id, ")", ""), sep = "")
+
+    grid.text(mainlab)
+    popViewport()
+    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2, 
+                     xscale = xscale, yscale = yscale, 
+                     name = paste("node_density",  node$id, "plot", sep = ""))
+    pushViewport(plot)
+    yd <- ydens
+    xd <- yrange
+    if (horizontal) {
+      yyy <- xd
+      xd <- yd
+      yd <- yyy
+      yyy <- xr
+      xr <- yr
+      yr <- yyy
+      rxd <- rep(0, length(xd))
+      ryd <- rev(yd)
+    } else {
+      rxd <- rev(xd)
+      ryd <- rep(0, length(yd))
+    }
+
+    if (rug) {
+      nodeobs <- node$info$object$y
+      if (horizontal) {
+        grid.rect(x = xscale[1], y = nodeobs , height = 0, width = xscale[1], 
+                  default.units = "native", just = c("right", "bottom"),
+		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
+      } else {
+        grid.rect(x = nodeobs, y = yscale[1], 
+                  width = 0, height = abs(yscale[1]), default.units = "native", 
+                  just = c("center", "bottom"),
+		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
+        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
+        #           default.units = "native")
+        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
+        #           default.units = "native")
+      }
+    }
+
+    grid.polygon(x = c(xd, rxd), y = c(yd, ryd), default.units = "native",
+              gp = gpar(col = "black", fill = fill, lwd = lwd))
+    #grid.lines(x = xd, y = yd, default.units = "native", 
+    #           gp = gpar(col = col, lwd = lwd))
+    grid.xaxis()
+    grid.yaxis()
+    grid.rect(gp = gpar(fill = "transparent"))
+    upViewport(2)
+  }
+  return(rval)
+}
+
+class(node_density) <- "grapcon_generator"
+
+plot(tree, tnex = 1.7, drop = TRUE,
+     terminal_panel = node_density,
+     tp_args = list(fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
+
 \begin{frame}
 \frametitle{Building Distributional Trees and Forests}
+\begin{minipage}{0.75\textwidth}
 {\bf Tree:}
 \begin{enumerate}
-\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\hat{\theta}$ via maximum likelihood.
-\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $z_i$.
-%\item Assess whether the \emph{model scores} are associated with
-%    (or change along) any of the available covariates -- e.g.,
-%    using parameter instability tests (\emph{strucchange}) or
-%    conditional inference (\emph{coin}).
-\item Split the sample along the partitioning variable with the strongest association or instability.
-    Choose breakpoint with highest improvement in log-likelihood.
-\item Repeat steps 2--4 recursively in the subgroups
+\item<3-> Fit a global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
+Estimate model parameters $\hat{\theta}$.
+\item<5-> Evaluate how well the model fits the data (for each parameter and each observation).
+\item<6-> Choose the covariate $X$ with the strongest influence or association to
+goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
+\end{enumerate}
+\end{minipage}
+\begin{minipage}{0.2\textwidth}
+\vspace{-0.5cm}
+\begin{tikzpicture}
+\visible<2-3>{
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$Y$};
+}
+\visible<4->{
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$\mathcal{D}(Y;\hat{\theta}$)};
+}
+\visible<8>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
+\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<9->{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
+\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\end{tikzpicture}
+\end{minipage}
+\vspace{0.1cm}
+\begin{adjustwidth}{-0.0em}{-3em}
+\begin{enumerate}
+\setcounter{enumi}{3}
+\item<7-> Find the split point $p$ which leads to the highest improvement.
+\item<10-> Repeat steps 1--4 recursively in the subgroups
     until some stopping criterion is met.
-    %-- e.g., for significance or sample size.
-%\item Choose the variable with the strongest association
-%\item Choose the split point which leads to the highest improvement of the model
-%\item Split and repeat 2-6 in each node until a stopping criterion is met
 \end{enumerate}
-
+\end{adjustwidth}
 \vspace{0.4cm}
+\visible<11->{
 {\bf Forest:} Ensemble of trees.
 \begin{itemize}
-\item Bootstrap or subsamples.
-\item Random input variable sampling.
+\item Varying subsamples.
+\item Combine/average over trees.
 \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Prediction}
-%\vspace{1cm}
-{\bf Tree:}\\
-\begin{enumerate}
-\item Select the subsample of the learning data which ends up in the same terminal node as the new observation.
-\item Estimate $\hat{\theta}$ on this subsample.
-\end{enumerate}
-\vspace{0.5cm}
-
-{\bf Forest:}\\
-\begin{enumerate}
-\item For each learning observation calculate a weight by counting the number of trees in which it ends up in the same terminal node as the new observation.
-\item Estimate $\hat{\theta}$ on the weighted learning data.
-\end{enumerate}
-% but weighted by similarities of the learning data and the new observation over all trees
-%\item Evaluate final node for new observation in each of the trees and investigate similarities with each of the learning observations
-%\item Fit a distributional model for the new observation with different influences of the learning observations defined by weights
-\end{frame}
-
-%\SweaveOpts{eval=FALSE}
-<<sim_1dataset, echo=FALSE, results=hide>>=
-set.seed(723)
-fun <- function(x){cbind(10*exp(-((4*x[,1]-2)^4)), 0.5 + 2*abs(x[,1]))}
-learndata <- dgp(300, family = dist_list_normal, round.sp = 4, fun = fun)
-control <-  ctree_control(teststat = "quad", testtype = "Bonferroni", mincriterion = 0.7, minbucket = 20L)
-dt <- disttree(y~x1, data=learndata, family=dist_list_normal, type.tree = "ctree", control = control)
-
-if(file.exists("sim_forest.rda")){
-  load("sim_forest.rda")
-} else {
-  control <-  ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 7L)
-  df <- distforest(y~x1, data=learndata, family=dist_list_normal, ntree = 500L, type.tree = "ctree", control = control)
-  df$info <- df$fitted <- df$weights <- df$nodes <- df$data <- df$terms <- df$loglik <- df$trafo <- NULL
-  save(df, file = "sim_forest.rda")
-}
-
-g <- gamlss(y ~ pb(x1), sigma.formula = ~pb(x1), data = learndata, trace = FALSE)
-@
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<oneforest_exp1, fig=TRUE, echo=FALSE>>=
-plot1(learndata, dt=dt, df=df, g=g)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<oneforest_exp1_dt, fig=TRUE, echo=FALSE>>=
-plot1(learndata, dt=dt, df=df, g=g, dt.par = TRUE)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<oneforest_exp1_g, fig=TRUE, echo=FALSE>>=
-plot1(learndata, dt=dt, df=df, g=g, g.par = TRUE)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<oneforest_exp1_df, fig=TRUE, echo=FALSE>>=
-plot1(learndata, dt=dt, df=df, g=g, df.par = TRUE)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<oneforest_exp1_all, fig=TRUE, echo=FALSE>>=
-plot1(learndata, dt=dt, df=df, g=g, all.par = TRUE)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<oneforest_exp1_allvar, fig=TRUE, echo=FALSE>>=
-plot1(learndata, dt=dt, df=df, g=g, all.var = TRUE)
-@
-\end{center}
-\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%<<oneforest_exp1_sigma, fig=TRUE, echo=FALSE>>=
-%plot1sigma(learndata,dt, df, g)
-%@
-%\end{center}
+}
+\end{frame}
+
+%\begin{frame}
+%\frametitle{Building Distributional Trees and Forests}
+%{\bf Tree:}
+%\begin{enumerate}
+%\item Fit a global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
+%Estimate model parameters $\hat{\theta}$.
+%\item Evaluate how well the model fits the data (for each parameter).
+%\item Choose the covariate $X$ with the strongest influence or association to
+%goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
+%\item Find the split which leads to the highest improvement.
+%\item Repeat steps 1--4 recursively in the subgroups
+%    until some stopping criterion is met.
+%\end{enumerate}
+
+%\vspace{0.4cm}
+%{\bf Forest:} Ensemble of trees.
+%\begin{itemize}
+%\item Varying subsamples.
+%\item Average over trees.
+%\end{itemize}
 %\end{frame}
 
-<<get_simlist, echo=FALSE, results=hide>>=
-if(file.exists("simlist.rda")){
-  load("simlist.rda")
-} else {
-  source("gensimlist.R")
-  simlist <- gensimlist(seed = 9, nrep = 150, ntree = 100)
-  save(simlist, file = "simlist.rda")
-}
-
-# plots for various data sets
-# results given in simlist
-# plot RMSE
-plot_rmse <- function(simlist, type = c("exp", "par")){
-
-  if(type == "exp"){
-    #dev.off() 
-    rmse <- cbind(simlist$av.rmse.exp.true.dt, simlist$av.rmse.exp.true.df, simlist$av.rmse.exp.true.g,
-                  simlist$av.rmse.exp.obs.dt, simlist$av.rmse.exp.obs.df, simlist$av.rmse.exp.obs.g)
-    colnames(rmse) <- c("dt.true", "df.true", "g.true", "dt.obs", "df.obs", "g.obs")
-    ylim <- c(min(rmse), max(rmse))
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(x = simlist$x.axis, y = rmse[,"dt.true"], type = "l", col = pal["tree"], ylim = ylim,
-         xlab = "", ylab = "", xaxt="n", yaxt="n", lwd = 2)
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-    lines(x = simlist$x.axis, y = rmse[,"dt.obs"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"df.true"], type = "l", col = pal["forest"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"df.obs"], type = "l", lty = 2, col = pal["forest"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"g.true"], type = "l", col = pal["gamlss"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"g.obs"], type = "l", lty = 2, col = pal["gamlss"], lwd = 2)
-    #legend('left', c("disttree", "distforest", "gamlss"), col = c(pal["tree"], pal["forest"], pal["gamlss"]), lty = 1)
-    mtext(text= "smooth", side = 1, line = 1, adj = 0)
-    mtext(text= "steep", side = 1, line = 1, adj = 1)
-    mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
-    mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 5)
-    mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = 13)
-    mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = 16.5)
-
-    # add arrows
-    parsave <- par(new = TRUE,  mar = c(0,0,0,0))
-    plot(0,0,xlim=c(0,1),ylim=c(0,1),type="n",xlab='', ylab='', col='white', axes = FALSE)
-
-    #add arrow on the right side (downwards, for RMSE)
-    arrows(0.99,0.47, 0.99,0.3,lwd=1) ## add arrow
-    text(0.99,0.5, 'better')
-    segments(0.99,0.53,0.99,0.7,lwd=1)
-
-    #add arrows below
-    arrows(0.45,0.085,0.3,0.085,lwd=1) ## add arrow
-    arrows(0.6,0.085,0.75,0.085,lwd=1) ## add arrow
-
-    # plot parameters  reset to prior values
-    par(parsave)
-  }
-
-  if(type == "par"){
-    rmse <- cbind(simlist$av.rmse.mu.dt, simlist$av.rmse.mu.df, simlist$av.rmse.mu.g, 
-                  simlist$av.rmse.sigma.dt, simlist$av.rmse.sigma.df, simlist$av.rmse.sigma.g)
-    colnames(rmse) <- c("mu.dt", "mu.df", "mu.g", "sigma.dt", "sigma.df", "sigma.g")
-    ylim <- c(min(rmse), max(rmse))
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(x = simlist$x.axis, y = rmse[,"mu.dt"], type = "l", col = pal["tree"], ylim = ylim,
-         xlab = "", ylab = "", xaxt="n", yaxt="n", lwd = 2)
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-    lines(x = simlist$x.axis, y = rmse[,"sigma.dt"], type = "l", lty = 2, col = pal["tree"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"mu.df"], type = "l", col = pal["forest"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"sigma.df"], type = "l", lty = 2, col = pal["forest"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"mu.g"], type = "l", col = pal["gamlss"], lwd = 2)
-    lines(x = simlist$x.axis, y = rmse[,"sigma.g"], type = "l", lty = 2, col = pal["gamlss"], lwd = 2)
-    #legend('topleft', c("disttree", "distforest", "gamlss"), col = c(pal["tree"], pal["forest"],pal["gamlss"]), lty = 1)
-    mtext(text= "smooth", side = 1, line = 1, adj = 0)
-    mtext(text= "steep", side = 1, line = 1, adj = 1)
-    mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
-    mtext(text= "RMSE", side = 2, line = 1)
-    mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = -7)
-    mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = 8)
-    mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = 14)
-
-    # add arrows
-    parsave <- par(new = TRUE,  mar = c(0,0,0,0))
-    plot(0,0,xlim=c(0,1),ylim=c(0,1),type="n",xlab='',ylab='',col='white', axes = FALSE)
-
-    #add arrow on the right side (downwards, for RMSE)
-    arrows(0.99,0.47,0.99,0.3,lwd=1) ## add arrow
-    text(0.99,0.5, 'better')
-    segments(0.99,0.53,0.99,0.7,lwd=1)
-
-    #add arrows below
-    arrows(0.45,0.085,0.3,0.085,lwd=1) ## add arrow
-    arrows(0.6,0.085,0.75,0.085,lwd=1) ## add arrow
-
-    # plot parameters  reset to prior values
-    par(parsave)
-  }
-}
-
-# plot loglikelihood
-plot_ll <- function(simlist){
-  ll <- cbind(simlist$av.loglik.dt, simlist$av.loglik.df, simlist$av.loglik.g)
-  colnames(ll) <- c("dt", "df","g")
-  ylim <- c(min(ll), max(ll))
-  par(mar=c(5.1,4.1,4.1,3.1))
-  plot(x = simlist$x.axis, y = ll[,"dt"], type = "l", col = pal["tree"], ylim = ylim,
-       xlab = "", ylab = "", xaxt="n", yaxt="n", lwd = 2)
-  title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-  title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-  title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-  title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-  title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-
-  lines(x = simlist$x.axis, y = ll[,"df"], type = "l", col = pal["forest"], lwd = 2)
-  lines(x = simlist$x.axis, y = ll[,"g"], type = "l", col = pal["gamlss"], lwd = 2)
-  #legend('topright', c("disttree", "distforest", "gamlss"), col = c(pal["tree"], pal["forest"],pal["gamlss"]), lty = 1)
-  #mtext(text= "disttree", side = 3, col = pal["tree"], line = 1, adj = 0.3)
-  #mtext(text= "distforest", side = 3, col = pal["forest"], line = 1, adj = 0.5)
-  #mtext(text= "gamlss", side = 3, col = pal["gamlss"], line = 1, adj = 0.7)
-  #mtext(text= "vs.", side = 3, col = "black", line = 1, adj = 0.4)
-  #mtext(text= "vs.", side = 3, col = "black", line = 1, adj = 0.6)
-  mtext(text= "smooth", side = 1, line = 1, adj = 0)
-  mtext(text= "steep", side = 1, line = 1, adj = 1)
-  mtext(text= TeX('$\\kappa$'), side = 1, line = 1)
-  mtext(text= "Log-Likelihood", side = 2, line = 1)
-  mtext(text= "disttree", side = 2, col = pal["tree"], las = 1, line = 0.2, padj = 16.7)
-  mtext(text= "distforest", side = 2, col = pal["forest"], las = 1, line = 0.2, padj = -8)
-  mtext(text= "gamlss", side = 2, col = pal["gamlss"], las = 1, line = 0.3, padj = -16)
-
-  # add arrows
-  parsave <- par(new = TRUE,  mar = c(0,0,0,0))
-  plot(0,0,xlim=c(0,1),ylim=c(0,1),type="n",xlab='', ylab='', col='white', axes = FALSE)
-
-  #add arrow on the right side (upwards, for LL)
-  arrows(0.99,0.53,0.99,0.7,lwd=1) ## add arrow
-  text(0.99,0.5, 'better')
-  segments(0.99,0.3,0.99,0.47,lwd=1)
-
-  #add arrows below
-  arrows(0.45,0.085,0.3,0.085,lwd=1) ## add arrow
-  arrows(0.6,0.085,0.75,0.085,lwd=1) ## add arrow
-
-  # plot parameters  reset to prior values
-  par(parsave)
-}
-@
-
-\begin{frame}[fragile]
-\frametitle{Simulation}
-\begin{center}
-\vspace*{-1cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plotsim_ll, fig=TRUE, echo=FALSE>>=
-plot_ll(simlist)
-@
-\end{center}
-\end{frame}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\SweaveOpts{eval=TRUE}
 
@@
 \medskip
-\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
+\textbf{Physical knowledge:} Numerical Weather Prediction (NWP) system.\\
 \medskip
-\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
-\vspace{1.3cm}
-%\bigskip
-
+\textbf{Statistical post-processing} necessary due to uncertain initial conditions and unresolved processes.
+
+%Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
+\end{frame}
+
+<<echo = FALSE, results=hide>>=
+library("zoo")
+
+Sys.setenv("TZ"="UTC")
+
+# -------------------------------------------------------------------
+# GFS ensemble forecast
+# -------------------------------------------------------------------
+
+GFS   <- read.table("ensemble_plot/GENS_00_innsbruck-flughafen.txt", header = TRUE)
+init  <- min(as.POSIXct(GFS$timestamp-GFS$step*3600,origin="1970-01-01"))
+title <- sprintf("%s\n%s",
+                 "Global Forecast System (GFS) Ensemble Forecast for Innsbruck, Airport",
+                 sprintf("Forecast initialized %s",strftime(init,"%Y-%m-%d %H:%M UTC")))
+
+getZoo <- function( x, variable ) {
+  x <- subset( x, varname == variable )
+  if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
+  # Else create zoo
+  x <- zoo( subset(x,select=-c(varname,timestamp,step)),
+            as.POSIXct(x$timestamp,origin="1970-01-01") )
+  x
+}
+GFS_t2m  <- getZoo( GFS, "tmp2m" )
+# temperature bias: model simulates temperature for height 1675 (~1070 m higher than true height)
+# => add 10.7 degrees to account for bias (approx. 1 degree / 100 m)
+GFS_t2m <- GFS_t2m + 10.7
+GFS_rain <- getZoo( GFS, "apcpsfc" )
+
+# Two POSIXct vectors for the axis
+main  <- seq(min(as.POSIXct(as.Date(index(GFS_t2m)))),max(index(GFS_t2m)),by=86400)
+minor <- seq(min(index(GFS_t2m)),max(index(GFS_t2m)),by=3*3600)
+
+# GFS_t2m_mean <- rowMeans(GFS_t2m)
+GFS_t2m_mean <- GFS_t2m
+GFS_t2m_mean[,1] <- rowMeans(GFS_t2m)
+GFS_t2m_mean <- GFS_t2m_mean[,1]
+
+# GFS_rain_mean <- rowMeans(GFS_rain)
+GFS_rain_mean <- GFS_rain
+GFS_rain_mean[,1] <- rowMeans(GFS_rain)
+GFS_rain_mean <- GFS_rain_mean[,1]
+@
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{-0.3cm}
+\setkeys{Gin}{width=1.1\linewidth}
+<<ensemble_1_plot, fig=TRUE, echo=FALSE, width=9>>=
+par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
+layout( matrix(1:2,ncol=1) )
+plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+#plot( GFS_t2m_mean, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+abline( v = main, lty = 2 )
+mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
+plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+#plot( GFS_rain_mean, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+abline( v = main, lty = 2 )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
+
+axis( side = 1, at = minor, strftime(minor,"%H") )
+axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
+      line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{-0.3cm}
+\setkeys{Gin}{width=1.1\linewidth}
+<<ensemble_2_plot, fig=TRUE, echo=FALSE, width=9>>=
+par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
+layout( matrix(1:2,ncol=1) )
+plot( GFS_t2m[,c(1,2)], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+abline( v = main, lty = 2 )
+mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
+plot( GFS_rain[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+abline( v = main, lty = 2 )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
+
+axis( side = 1, at = minor, strftime(minor,"%H") )
+axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
+      line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{-0.3cm}
+\setkeys{Gin}{width=1.1\linewidth}
+<<ensemble_3_plot, fig=TRUE, echo=FALSE, width=9>>=
+par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
+layout( matrix(1:2,ncol=1) )
+plot( GFS_t2m[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+abline( v = main, lty = 2 )
+mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
+plot( GFS_rain[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+abline( v = main, lty = 2 )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
+
+axis( side = 1, at = minor, strftime(minor,"%H") )
+axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
+      line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{-0.3cm}
+\setkeys{Gin}{width=1.1\linewidth}
+<<ensemble_20_plot, fig=TRUE, echo=FALSE, width=9>>=
+par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
+layout( matrix(1:2,ncol=1) )
+plot( GFS_t2m, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+abline( v = main, lty = 2 )
+mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
+plot( GFS_rain, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+abline( v = main, lty = 2 )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
+
+axis( side = 1, at = minor, strftime(minor,"%H") )
+axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
+      line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{0.1cm}
 \textbf{Idea:} Statistical model with numerical predictions as covariates.\\
-\bigskip
+\medskip
 \textbf{Distribution assumption:}\\
 \vspace{-0.25cm}
-%\hspace{1.5cm}
-%\center
 $$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.4cm}
-\textbf{Observation data} (National Hydrographical Service):\\
+\textbf{Data:}
+\begin{itemize}
+\item \textbf{Y:} Observation data (National Hydrographical Service):\\
 \begin{itemize}
 \item 95 observation stations in Tyrol, Austria.\\
-\item Daily 24h precipitation sums from July over 28 years (1985--2012).
+\item Daily 24h precipitation sums (July, 1985--2012).
+%\item Daily 24h precipitation sums from July over 28 years (1985--2012).
 \end{itemize}
-\vspace{1cm}
-%\bigskip
-\textbf{NWP model output} (Global Ensemble Forecast System):\\
-\begin{itemize}
-\item Ensemble predictions of various meteorological quantities related to precipitation, temperature,
-air pressure, \dots.\\
-\item 11 members each using slightly different perturbed initial conditions.
+
+\item \textbf{X:} NWP model output (Global Ensemble Forecast System):\\
+Ensemble predictions of various meteorological quantities (related to precipitation, temperature, air pressure, sunshine, \dots $\rightarrow$ 82 covariates)
 \end{itemize}
@@
 \vspace*{0.2cm}
-\textbf{Base variables:}
-\begin{itemize}
-\item Total precipitation
-\item Convective available potential energy
-\item Downwards short wave radiation flux (``sunshine'')
-\item Mean sea level pressure
-\item Preciptable water
-\item 2m maximum temperature
-\item Total column-integrated condensate
-\item Temperature 
-\item Temperature differences in altitude
-\end{itemize}
-\medskip
-\textbf{Variations:} Ensemble min/max/mean/standard deviation\\
-\vspace{0.1cm}
-$\rightarrow$ 82 covariates.
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
-\textbf{Application for one station:}
+\textbf{Application for one stations:}
 \begin{itemize}
@@
 
-<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
-#setwd("~/svn/partykit/pkg/disttree/inst/CMStatistics2017")
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\SweaveOpts{eval=FALSE}
+
+<<pred_axams, echo=FALSE, results=hide>>=
 load("data_Axams.rda")
@@
 \vspace*{-0.2cm}
-<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=5.2>>=
+<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, height=5.2>>=
 plot(x = x, y = y1, type = "l", col = "red", 
@@
 \vspace{0.2cm}
-\textbf{Application for one station:}
+\textbf{Application for all 95 station:}
 \begin{itemize}
@@
 \item Evaluate on 4 years.
-\item 10 times 7-fold cross validation.
 \item Compare to commonly used heteroscedastic censored Gaussian models:
 \end{itemize}
-%\begin{itemize}
-%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-%\item Boosted GAM: automatic variable selection.
-%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
-%\end{itemize}
-\vspace{0.1cm}
-\begin{table}[t!]
-\footnotesize
-
-\hskip-0.4cm\begin{tabular}{ l l l }
-\hline
-Model & Type & Variable Selection \\
-\hline
-\vspace{-0.15cm}
-\textbf{Forest} & recursive    & automatic \\ 
-                & partitioning &           \\ 
-\hline
-\vspace{-0.15cm}
-\textbf{prespecified GAM} & spline  & based on expert \\
-                          & in each & knowledge      \\
-\hline
-\vspace{-0.15cm}
-\textbf{boosted GAM} & spline  & automatic  \\
-                     & in each &            \\
-\hline
-\vspace{-0.15cm}
-\textbf{EMOS} & linear & total precipitation mean \\
-              &        & and standard deviation   \\
-\hline
-\end{tabular}
-\end{table}
-
-\end{frame}
-
-<<echo=FALSE, results=hide>>=
-#### cross validation rain
-if(file.exists("rain_Axams_7x10.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_7x10.rda")
-  load("rain_Axams_7x10.rda")
-} else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross_7x10.R")
-  source("rain_cross_7x10.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-  rainres <- rain_cross(stationname = "Axams", 
-                        seedconst = 7, ntree = 100,
-                        forest_mtry = 27,
-                        gamboost_cvr = TRUE,
-                        frac = FALSE)
-
-  save(rainres, file = "rain_Axams_7x10.rda")
-}
-
-rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres)-1)
-
-for(i in 1:(length(rainres)-1)){
-  rain_crps[i,] <- colMeans(rainres[[i]]$crps, na.rm = TRUE)
-}
-
-colnames(rain_crps) <- colnames(rainres[[1]]$rmse)
-
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\begin{center}
-\vspace*{-0.2cm}
-<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.2>>=
-boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
-        main = "Cross validation (with reference model EMOS)",
-        names = c("Forest", "prespecified GAM", "boosted GAM"),
-        ylab = "CRPS skills score", col = "lightgray") 
-abline(h = 0, col = pal["EMOS"], lwd = 2)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.2cm}
-\textbf{Application for all 95 stations:}
-\begin{itemize}
-\item Learn forest model on data from 24 years (1985--2008).
-\item Evaluate on 4 years (2009--2012).
-\item Compare to commonly used heteroscedastic censored Gaussian models. %:\\
-\end{itemize}
-%\begin{itemize}
-%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-%\item Boosted GAM: automatic variable selection.
-%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
-%\end{itemize}
-
-%\vspace{0.1cm}
-%\begin{table}[t!]
-%\footnotesize
-%\nodindent
-%\hskip-0.4cm\begin{tabular}{ l l l }
-%\hline
-%Model & Type & Variable Selection \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{Forest} & recursive    & automatic \\ 
-%                & partitioning &           \\ 
-%\hline
-%\vspace{-0.15cm}
-%\textbf{prespecified GAM} & spline  & based on expert \\
-%                          & in each & knowledge      \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{boosted GAM} & spline  & automatic  \\
-%                     & in each &            \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{EMOS} & linear & total precipitation mean \\
-%              &        & and standard deviation   \\
-%\hline
-%\end{tabular}
-%\end{table}
 \end{frame}
@@
 if(file.exists("rain_pred_24to4.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
   load("rain_pred_24to4.rda")
@@
 
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_ll}
-%\end{center}
-%\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_RMSE_par}
-%\end{center}
-%\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_RMSE_exp}
-%\end{center}
-%\end{frame}
-
 \begin{frame}

### 2018-03-17 pkg/disttree/inst/Psychoco2018/slides.Rnw
last touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2016 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7d31a34b271f923b7723ce153d34e431da46f5f8
first draft of slides for PhD-Seminar
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2053 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: be6eb4965a7015ec7f3014964e62ad0fce63684d

@@
 %\usetheme[nototalframenumber, noslidenumber, url]{uibk}
-\usetheme[nototalframenumber, noslidenumber, url]{uibk}
+\usetheme[nototalframenumber, noslidenumber]{uibk}
 
@@
 transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-## define distribution list:
-# dist_list_normal
-{
-
-  dist_list_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {
-
-    val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
-    if(!log) val <- exp(val)
-
-    # par <- c(eta[1], exp(eta[2]))
-    # val <- dnorm(y, mean = par[1], sd = par[2], log = log)
-
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE) {
-
-    score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
-                   -1 + exp(-2*eta[2] + log((y-eta[1])^2)))
-
-    # par <- c(eta[1], exp(eta[2])) 
-    # score <- cbind(1/par[2]^2 * (y-par[1]), 
-    #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))
-
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
-    d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-eta[1]) * exp(-2*eta[2]), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    d2ld.etasigma2 <- sum(weights * (-2)*exp(log((y-eta[1])^2) - 2*eta[2]), na.rm = TRUE)
-
-    # par <- c(eta[1], exp(eta[2]))                           
-    # d2ld.etamu2 <- sum(weights * rep.int(-1/par[2]^2, ny))
-    # d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-par[1])/par[2]^2), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    # d2ld.etasigma2 <- sum(weights * (-2)*(y-par[1])^2/par[2]^2, na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etamu.d.etasigma, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- pnorm
-  qdist <- qnorm
-  rdist <- rnorm
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    if(is.null(weights)) {
-      mu <- mean(y)
-      sigma <- sqrt(1/length(y) * sum((y - mu)^2))
-    } else {
-      mu <- weighted.mean(y, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (y - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- TRUE
-
-  dist_list_normal <- list(family.name = "Normal Distribution",
-                           ddist = ddist, 
-                           sdist = sdist, 
-                           hdist = hdist,
-                           pdist = pdist,
-                           qdist = qdist,
-                           rdist = rdist,
-                           link = link, 
-                           linkfun = linkfun, 
-                           linkinv = linkinv, 
-                           linkinvdr = linkinvdr,
-                           startfun = startfun,
-                           mle = mle,
-                           gamlssobj = FALSE,
-                           censored = FALSE
-  )
-}
-
-# dist_list_cens_normal
-{
-
-  dist_list_cens_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
-    par <- c(eta[1], exp(eta[2]))
-    val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
-    if(sum) {
-      if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
-    par <- c(eta[1], exp(eta[2]))
-    # y[y==0] <- 1e-323
-
-    score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
-    score <- cbind(score_m, score_s)
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y)[1])
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-      #if(any(is.nan(score))) print(c(eta, "y", y))
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    par <- c(eta[1], exp(eta[2]))                           
-    # y[y==0] <- 1e-323
-
-    d2mu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    d2sigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-    dmudsigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu.sigma", left = left, right = right) # FIX: order?
-    dsigmadmu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma.mu", left = left, right = right) # FIX: order?
-    dsigma <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-
-    d2ld.etamu2 <- sum(weights * d2mu, na.rm = TRUE)
-    d2ld.etamu.d.etasigma <- sum(weights * dmudsigma * par[2], na.rm = TRUE)
-    d2ld.etasigma.d.etamu <- sum(weights * dsigmadmu * par[2], na.rm = TRUE)
-    d2ld.etasigma2 <- sum(weights * (d2sigma * exp(2*eta[2]) + dsigma * par[2]), na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etasigma.d.etamu, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- function(q, eta, lower.tail = TRUE, log.p = FALSE) crch:::pcnorm(q, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  qdist <- function(p, eta, lower.tail = TRUE, log.p = FALSE) crch:::qcnorm(p, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  rdist <- function(n, eta) crch:::rcnorm(n, mean = eta[1], sd = eta[2], left = left, right = right)
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    yc <- pmax(0,y)  # optional ?
-    if(is.null(weights)) {
-      mu <- mean(yc)
-      sigma <- sqrt(1/length(yc) * sum((yc - mu)^2))
-    } else {
-      mu <- weighted.mean(yc, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (yc - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- FALSE
-
-  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
-                                ddist = ddist, 
-                                sdist = sdist, 
-                                hdist = hdist,
-                                pdist = pdist,
-                                qdist = qdist,
-                                rdist = rdist,
-                                link = link, 
-                                linkfun = linkfun, 
-                                linkinv = linkinv, 
-                                linkinvdr = linkinvdr,
-                                startfun = startfun,
-                                mle = mle,
-                                gamlssobj = FALSE,
-                                censored = TRUE
-  )
-}
 
@@
 \begin{frame}
-\frametitle{Choosing a distribution}
-
-\medskip
-\textbf{Based on:}
+\frametitle{Model specification}
+
+\textbf{Covariates:} Automatically through adaptive forest weights.
+
+\bigskip
+
+\textbf{Response:} Distributional specification needed.
 \begin{itemize}
-\item Recommendations by specialist literature.
-\item Expert subject matter knowledge.
-\item Experiences from other modeling situations.
-\item Common simplifications/generalizations/assumptions.
+\item Continuous responses: Gaussian, \dots
+\item Limited responses: Censored Gaussian, \dots
+\item Survival times: Exponential, Weibull, \dots
+\item Count: Poisson, negative binomial, \dots
 \end{itemize}
@@
 \bigskip
-\textbf{Examples:}
-\begin{itemize}
-\item Survival times: Weibull distribution, exponential distribution.
-\item Wages, precipitation: censored normal distribution.
-\item Temperature: normal distribution.
-\item Count of events/accidents (insurance): Poisson distribution.
-\end{itemize}
+
+\textbf{Guidance:} Literature, theory, experience, \dots
+
+\bigskip
+
+\textbf{Alternative:} Transformation models.
+
 \end{frame}

### 2018-03-17 pkg/disttree/inst/MIP-Seminar/ensemble_plot/GENS.R

@@ -1,0 +1,64 @@
+# -------------------------------------------------------------------
+# - NAME:        GENS.R
+# - AUTHOR:      Reto Stauffer
+# - DATE:        2018-03-13
+# -------------------------------------------------------------------
+# - DESCRIPTION:
+# -------------------------------------------------------------------
+# - EDITORIAL:   2018-03-13, RS: Created file on thinkreto.
+# -------------------------------------------------------------------
+# - L@ST MODIFIED: 2018-03-13 17:43 on marvin
+# -------------------------------------------------------------------
+
+
+    library( "zoo" )
+
+    Sys.setenv("TZ"="UTC")
+
+# -------------------------------------------------------------------
+# GFS ensemble forecast
+# -------------------------------------------------------------------
+
+    GFS   <- read.table( "GENS_00_innsbruck-flughafen.txt", header = TRUE )
+    init  <- min(as.POSIXct(GFS$timestamp-GFS$step*3600,origin="1970-01-01"))
+    title <- sprintf("%s\n%s",
+                "Global Forecast System (GFS) Ensemble Forecast",
+                sprintf("Forecast initialized %s",strftime(init,"%Y-%m-%d %H:%M UTC")))
+
+    getZoo <- function( x, variable ) {
+        x <- subset( x, varname == variable )
+        if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
+        # Else create zoo
+        x <- zoo( subset(x,select=-c(varname,timestamp,step)),
+                  as.POSIXct(x$timestamp,origin="1970-01-01") )
+        x
+    }
+    GFS_t2m  <- getZoo( GFS, "tmp2m" )
+    GFS_rain <- getZoo( GFS, "apcpsfc" )
+
+    # Two POSIXct vectors for the axis
+    main  <- seq(min(as.POSIXct(as.Date(index(GFS_t2m)))),max(index(GFS_t2m)),by=86400)
+    minor <- seq(min(index(GFS_t2m)),max(index(GFS_t2m)),by=3*3600)
+
+    par(mar=c(0.2,0,0.2,0),oma=c(5,5,5,5))
+    layout( matrix(1:2,ncol=1) )
+    plot( GFS_t2m, screen=1, xaxs="i", xaxt="n" )
+    abline( v = main, lty = 2 )
+    mtext( side = 3, line = 2, cex = 1.2, font = 2, title )
[…]

### 2018-03-17 pkg/disttree/inst/MIP-Seminar/rain_cross_7x10.R

@@ -41,6 +41,140 @@ rain_cross <- function(stationname, seedconst = 7, ntree = 100,
   #setwd("~/disttree/inst/draft/")
   
   
+  # dist_list_cens_normal
+  {
+    
+    dist_list_cens_normal <- list()
+    
+    parnames <- c("mu", "sigma")
+    etanames <- c("mu", "log(sigma)")
+    
+    ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
+      par <- c(eta[1], exp(eta[2]))
+      val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
+      if(sum) {
+        if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
+        val <- sum(weights * val, na.rm = TRUE)
+      }
+      return(val)
+    }
+    
+    
+    sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
+      par <- c(eta[1], exp(eta[2]))
+      # y[y==0] <- 1e-323
+      
+      score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
+      score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
+      score <- cbind(score_m, score_s)
+      score <- as.matrix(score)
+      colnames(score) <- etanames
+      if(sum) {
+        if(is.null(weights)) weights <- rep.int(1, length(y)[1])
+        # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
+        score[score==Inf] = 1.7e308
+        score <- colSums(weights * score, na.rm = TRUE)
+        #if(any(is.nan(score))) print(c(eta, "y", y))
+      }
+      return(score)
+    }
+    
+    
+    hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
+      ny <- length(y)
+      if(is.null(weights)) weights <- rep.int(1, ny)
+      
+      par <- c(eta[1], exp(eta[2]))                           
+      # y[y==0] <- 1e-323
+      
[…]

### 2018-03-17 pkg/disttree/inst/MIP-Seminar/ensemble_plot/Spatial.R
first draft of slides for PhD-Seminar
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2053 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: be6eb4965a7015ec7f3014964e62ad0fce63684d

@@ -1,0 +1,102 @@
+# -------------------------------------------------------------------
+# - NAME:        Spatial.R
+# - AUTHOR:      Reto Stauffer
+# - DATE:        2018-03-13
+# -------------------------------------------------------------------
+# - DESCRIPTION:
+# -------------------------------------------------------------------
+# - EDITORIAL:   2018-03-13, RS: Created file on thinkreto.
+# -------------------------------------------------------------------
+# - L@ST MODIFIED: 2018-03-14 12:46 on marvin
+# -------------------------------------------------------------------
+
+
+    library( "raster" )
+    library( "ncdf4" )
+
+
+    step <- 36
+    step <- 240
+
+    ncfiles <- list.files("NetCDF",sprintf("^GFS_[0-9]{10}_%02d_[0-9]{2}.nc$",step))
+    if ( length(ncfiles)==0 ) stop("No netcdf files found")
+    ncfiles <- sprintf("NetCDF/%s",sort(ncfiles))
+
+
+    # Helper function (matrix + lon/lat vectors -> raster)
+    data2raster <- function( x, lons, lats, flip = TRUE ) {
+        delta <- min(diff(lons)) / 2
+        x <- t(x); if ( flip ) x <- x[nrow(x):1,]
+        r <- raster( x, xmn = min(lons)-delta, xmx = max(lons)+delta,
+                        ymn = min(lats)-delta, ymx = max(lats)+delta )
+        r
+    }
+
+    res <- list(t2m=list(),tp=list())
+    for ( file in ncfiles ) {
+        # Open NetCDF file
+        nc <- nc_open( file )
+        # Extract member from file name
+        mem <- as.integer(gsub("\\.nc","",regmatches(file,regexpr("[0-9]{2}.nc",file))))
+        # Loading longitude/latitude arrays
+        lats <- ncvar_get(nc,"latitude")
+        lons <- ncvar_get(nc,"longitude")
+        # Extracting data
+        tmp_t2m <- ncvar_get( nc, "t2m" ) - 273.15
+        tmp_tp  <- ncvar_get( nc, "tp" )
+        # Convert to raster and store
+        res$t2m[[sprintf("member_%d",mem)]] <- data2raster( tmp_t2m, lons, lats )
[…]

### 2018-03-18 pkg/disttree/inst/MIP-Seminar/slides.Rnw
edited format and scaling
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2058 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0086a424149496d3ed0e7ad9ad1e57220f66e62a

@@
 \frametitle{Statistical modeling}
-\vspace{0.2cm}
+\vspace{0.18cm}
 Approximation of reality based on a set of observations.\\
-\vspace{0.2cm}
-\visible<1, 4-> {Assumption:} 
-\vspace{-1.1cm}
+\vspace{0.19cm}
+\visible<2, 5-> {Assumption:} 
+\vspace{-0.7cm}
 \begin{center}
 \begin{tikzpicture}
-\visible<1, 4->{
+\visible<2, 5->{
 \node[] at (0,0.4) {X};
 }
-\visible<2-3>{
-\node[inner sep=0pt] (today) at (-1.9,-0.2)
-    {\includegraphics[width=.38\textwidth]{today.png}};
-}
+\visible<3-4>{
+\node[inner sep=0pt] (today) at (-1.8,-0.4)
+    {\includegraphics[width=.365\textwidth]{today.png}};
+\node[font=\scriptsize] at (-1.9,-2) {15.03.2018};
+}
+\visible<2->{
 \draw[->] (0.4,0.4)--(0.9,0.4);
@@
 \draw[->] (2.6,0.4)--(3.1,0.4);
-\visible<1, 4->{
+}
+\visible<2, 5->{
 \node[] at (3.5,0.4) {Y};
 }
-\visible<2-3>{
-\node[inner sep=0pt] (future) at (5.4,-0.2)
-    {\includegraphics[width=.38\textwidth]{future.png}};
+\visible<3-4>{
+\node[inner sep=0pt] (future) at (5.3,-0.4)
+    {\includegraphics[width=.365\textwidth]{future.png}};
+\node[font=\scriptsize] at (5.4,-2) {16.03.2018};
 }
@@
 \end{center}
-\visible<3>{
-\vspace{0.5cm}
+\visible<4>{
+\vspace{0.3cm}
 Y = daily precipitation sums\\
-\vspace{0.1cm}
+\vspace{0.15cm}
 X = observed precipitation of today,\\
@@
 
-\vspace{-3.5cm}
-\visible<5->{
+\vspace{-4cm}
+\visible<6->{
 \textbf{Application/Motivation:}
@@
 \vspace{0.5cm}
-\visible<6->{
+\visible<7->{
 \textbf{Two main approaches / ``cultures'':}
@@
 
+<<motivation_GLM, echo=FALSE, results=hide>>=
+nobs <- 200
+## GLM
+set.seed(7)
+x <- c(1:nobs)/nobs
+ytrue <- 1+x*1.5
+y <- ytrue + rnorm(nobs,0,0.3)
+@
+<<motivation_GAM, echo=FALSE, results=hide>>=
+## GAM
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3 
+y <- ytrue + rnorm(nobs,0,0.3)
+@
+<<motivation_GAMLSS, echo=FALSE, results=hide>>=
+## GAMLSS
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3
+var <- exp(-(2*x)^2)/2
+y <- ytrue + rnorm(nobs, 0, 0.1 + var)
+@
+
 \begin{frame}[fragile]
 \frametitle{Statistical modeling}
+\vspace{0.2cm}
+\textbf{Data modeling:}
 \vspace{0.1cm}
-\textbf{Data modeling:}
-\vspace{0.3cm}
 \begin{center}
@@
 \end{center}
-
+\vspace{-0.4cm}
 \begin{itemize}
@@
 \end{itemize}
-
-\vspace{0.2cm}
-
-\visible<2-3>{
+\only<1>{\vspace{1.5cm}}
+
+\vspace{0.1cm}
+
+\visible<2->{
 \begin{minipage}{0.65\textwidth}
-\textbf{Example:} Linear model\\
+\textbf{Example:} 
+\only<2-3>{Linear model}\only<4>{Generalized additive model}\only<5>{Generalized additive model\\ for location, scale, and shape}
+\\
+\only<2-3>{
 $$\visible<3>{\mathbb{E}(Y|X) = } f(X; \alpha) = \alpha_0 + \alpha_1 \cdot X$$  
+}
+\only<4>{
+$$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
+}
+\only<5>{
+\vspace{-0.8cm}
+\begin{align*}
+\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
+\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
+\end{align*}
+}
 \end{minipage}
-\vspace{-0.6cm}
-\hfill
-<<motivation_GLM, echo=FALSE, results=hide>>=
-nobs <- 200
-## GLM
-set.seed(7)
-x <- c(1:nobs)/nobs
-ytrue <- 1+x*1.5
-y <- ytrue + rnorm(nobs,0,0.3)
-@
+%\vspace{-0.6cm}
+%\hfill
 \begin{minipage}{0.3\textwidth}
+\only<2-3>{
 <<plot_motivation_GLM_single, fig=TRUE, echo=FALSE>>=
@@
 @
-\end{minipage}
-}
-
-\vspace{-2.35cm}
-\visible<4>{
-\begin{minipage}{0.65\textwidth}
-\textbf{Example:} Generalized additive model\\
-$$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
-\end{minipage}
-\hfill
-<<motivation_GAM, echo=FALSE, results=hide>>=
-## GAM
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3 
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-\begin{minipage}{0.3\textwidth}
+}
+\only<4>{
 <<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
@@
 @
-\end{minipage}
-}
-
-\vspace{-3cm}
-\visible<5>{
-\begin{minipage}{0.65\textwidth}
-\vspace{0.8cm}
-\textbf{Example:} Generalized additive model\\
-for location, scale, and shape
-\begin{align*}
-\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
-\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
-\end{align*}
-\end{minipage}
-\hfill
-<<motivation_GAMLSS, echo=FALSE, results=hide>>=
-## GAMLSS
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3
-var <- exp(-(2*x)^2)/2
-y <- ytrue + rnorm(nobs, 0, 0.1 + var)
-@
-\begin{minipage}{0.3\textwidth}
+}
+\only<5>{
 <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
@@
 @
+}
 \end{minipage}
@@
 \frametitle{Statistical modeling}
+\vspace{0.2cm}
 \textbf{Algorithmic modeling:}
-
+\vspace{0.1cm}
 \begin{center}
@@
 \node[] at (0,0.4) {X};
-\draw[->] (0.3,0.4)--(0.8,0.4);
+\draw[->] (0.4,0.4)--(0.9,0.4);
 \draw[] (1,0) rectangle (2.5,0.8);
 \node[font=\scriptsize] at (1.75,0.4) {nature};
-\draw[->] (2.7,0.4)--(3.2,0.4);
+\draw[->] (2.6,0.4)--(3.1,0.4);
 \node[] at (3.5,0.4) {Y};
@@
 \end{itemize}
-
-\visible<2-3>{
+\only<1>{\vspace{1cm}}
+\visible<2->{
 \begin{minipage}{0.65\textwidth}
-\vspace{0.5cm}
-\textbf{Example:} Tree model\\
-\vspace{-0.8cm}
+%\vspace{-0.5cm}
+\textbf{Example:} \only<2>{Tree model}\only<3>{Forest model}\\
+%\vspace{-0.8cm}
 \begin{center}
+\only<2>{
+\vspace{-0.7cm}
 \resizebox{0.6\textwidth}{!}{
-<<plot_raintree, fig=TRUE, echo=FALSE, width=5, height=3>>=
+<<plot_raintree, fig=TRUE, echo=FALSE, width=4.2, height=2.6>>=
+par(mar=c(0,0,0,0), oma=c(0,0,0.3,0))
 paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
@@
 }
-\end{center}
-\end{minipage}
-}
-\visible<3>{
-\hfill
-\begin{minipage}{0.3\textwidth}
-<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-@
-\end{minipage}
-}
-
-\vspace{-3cm}
-\visible<4-5>{
-\begin{minipage}{0.65\textwidth}
-\textbf{Example:} Forest model\\
-\begin{center}
-\vspace{-0.2cm}
-\resizebox{0.5\textwidth}{!}{
+}
+\only<3>{
+\resizebox{0.6\textwidth}{!}{
 \begin{tikzpicture}
@@
 }
+}
 \end{center}
 \end{minipage}
-}
-\visible<5>{
-\hfill
 \begin{minipage}{0.3\textwidth}
+\begin{center}
+\only<2>{
+\vspace{-0.29cm}
+<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+lines(x = x, y = ytree, col = pal["forest"], lwd=7)
+@
+}
+\only<3>{
 <<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
@@
 @
+}
+\end{center}
 \end{minipage}

### 2018-03-18 pkg/disttree/inst/MIP-Seminar/slides.Rnw
cumulated precipitation sums for ensemble plots (now 24h instead of 6h sums)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2059 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5fd9ce0a3a15edaeefd9c2af2883d8bb622ef9b3

@@
 \vspace{0.3cm}
-Y = daily precipitation sums\\
+X = observed precipitation of today,\\
+\qquad current temperature / humidity / \dots,\\
+\qquad physical model output\\
 \vspace{0.15cm}
-X = observed precipitation of today,\\
-\qquad current temperature/humidity/\dots,\\
-\qquad physical model output
+Y = daily precipitation sums
 }
@@
 \end{frame}
-
-<<motivation_GLM, echo=FALSE, results=hide>>=
-nobs <- 200
-## GLM
-set.seed(7)
-x <- c(1:nobs)/nobs
-ytrue <- 1+x*1.5
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-<<motivation_GAM, echo=FALSE, results=hide>>=
-## GAM
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3 
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-<<motivation_GAMLSS, echo=FALSE, results=hide>>=
-## GAMLSS
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3
-var <- exp(-(2*x)^2)/2
-y <- ytrue + rnorm(nobs, 0, 0.1 + var)
-@
 
@@
 \begin{minipage}{0.3\textwidth}
+<<motivation_GLM, echo=FALSE, results=hide>>=
+nobs <- 200
+## GLM
+set.seed(7)
+x <- c(1:nobs)/nobs
+ytrue <- 1+x*1.5
+y <- ytrue + rnorm(nobs,0,0.3)
+@
 \only<2-3>{
@@
 }
+<<motivation_GAM, echo=FALSE, results=hide>>=
+## GAM
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3 
+y <- ytrue + rnorm(nobs,0,0.3)
+@
 \only<4>{
@@
 }
+<<motivation_GAMLSS, echo=FALSE, results=hide>>=
+## GAMLSS
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3
+var <- exp(-(2*x)^2)/2
+y <- ytrue + rnorm(nobs, 0, 0.1 + var)
+@
 \only<5>{
@@
 \frametitle{Goals}
-
+\vspace{0.3 cm}
 \textbf{Synthesis:} \\
-\vspace{0.1 cm}
+\vspace{0.2 cm}
 \begin{itemize}
@@
   $\rightarrow$ Distributional approach
+  \medskip
   \item Automatic detection of steps and abrupt changes,\\
@@
   $\rightarrow$ Tree approach
+  \medskip
   \item Smoothing and regularization\\
@@
 \item<5-> Evaluate how well the model fits the data (for each parameter and each observation).
-\item<6-> Choose the covariate $X$ with the strongest influence or association to
+\item<7-> Choose the covariate $X$ with the strongest influence or association to
 goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
@@
 }
-\visible<8>{
+\visible<6-8>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
+\draw[-, line width=0.8pt] (n0) -- (n1);
+\draw[-, line width=0.8pt] (n0) -- (n2);
+}
+\visible<9>{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
@@
 }
-\visible<9->{
+\visible<10->{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
@@
 \setcounter{enumi}{3}
-\item<7-> Find the split point $p$ which leads to the highest improvement.
-\item<10-> Repeat steps 1--4 recursively in the subgroups
+\item<8-> Find the split point $p$ which leads to the highest improvement.
+\item<11-> Repeat steps 1--4 recursively in the subgroups
     until some stopping criterion is met.
@@
 \vspace{0.4cm}
-\visible<11->{
+\visible<12->{
 {\bf Forest:} Ensemble of trees.
@@
 \end{frame}
-
-%\begin{frame}
-%\frametitle{Building Distributional Trees and Forests}
-%{\bf Tree:}
-%\begin{enumerate}
-%\item Fit a global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
-%Estimate model parameters $\hat{\theta}$.
-%\item Evaluate how well the model fits the data (for each parameter).
-%\item Choose the covariate $X$ with the strongest influence or association to
-%goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
-%\item Find the split which leads to the highest improvement.
-%\item Repeat steps 1--4 recursively in the subgroups
-%    until some stopping criterion is met.
-%\end{enumerate}
-
-%\vspace{0.4cm}
-%{\bf Forest:} Ensemble of trees.
-%\begin{itemize}
-%\item Varying subsamples.
-%\item Average over trees.
-%\end{itemize}
-%\end{frame}
 
@@
 GFS_t2m <- GFS_t2m + 10.7
+
 GFS_rain <- getZoo( GFS, "apcpsfc" )
+# precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
+GFS_rain24 <- GFS_rain[c((1:10)*4),]
+for(i in 1:NROW(GFS_rain24)){
+  GFS_rain24[i,] <- colSums(GFS_rain[c((4*i-3):(4*i)),])
+}
+# first value of rain is 18 hours later than temperature 
+#(both start at 6:00 but rain is cumulated over 24 hours)
+# -> drop first 3 values of temperature
+GFS_t2m <- GFS_t2m[-c(1:3),]
 
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+#plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
 #plot( GFS_rain_mean, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+plot( GFS_rain24[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+#plot( GFS_rain[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
 abline( v = main, lty = 2 )
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+plot( GFS_rain24[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+#plot( GFS_rain[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
 abline( v = main, lty = 2 )
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+plot( GFS_rain24, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+#plot( GFS_rain, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
 abline( v = main, lty = 2 )
@@
 \textbf{Idea:} Statistical model with numerical predictions as covariates.\\
-\medskip
+\vspace{0.4cm}
 \textbf{Distribution assumption:}\\
@@
 $$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
+%\vspace{2cm}
+\visible<2->{
 \textbf{Data:}
+%\vspace{-1cm}
 \begin{itemize}
@@
 \end{itemize}
-
+\vspace{0.1cm}
 \item \textbf{X:} NWP model output (Global Ensemble Forecast System):\\
-Ensemble predictions of various meteorological quantities (related to precipitation, temperature, air pressure, sunshine, \dots $\rightarrow$ 82 covariates)
+\vspace{0.14cm}
+\quad \small{Ensemble predictions of various meteorological quantities\\
+\quad (precipitation, temperature, air pressure, sunshine, \dots)\\
+\quad $\rightarrow$ 82 covariates.}
 \end{itemize}
+}
 \end{frame}

### 2018-03-19 pkg/disttree/inst/MIP-Seminar/slides.Rnw
small corrections and shorter list of references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2061 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 532f82ea394628d36ebf14f42e9777ad485840a2

@@
 \item Prediction.
-\item Analysis of dependencies and interactions.
+\item Caption of dependencies and interactions.
 \end{itemize}
@@
 \textbf{Example:} 
-\only<2-3>{Linear model}\only<4>{Generalized additive model}\only<5>{Generalized additive model\\ for location, scale, and shape}
+\only<2-3>{Linear model}\only<4>{Generalized additive model}\only<5-6>{Generalized additive model\\ for location, scale, and shape}
 \\
@@
 }
-\only<5>{
+\only<5-6>{
 \vspace{-0.8cm}
 \begin{align*}
-\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
-\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
+\visible<6>{\mathbb{E}(Y|X) = }\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
+\visible<6>{\sqrt{Var(Y|X)} = }\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
 \end{align*}
@@
 @
-\only<5>{
+\only<5-6>{
 <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\SweaveOpts{eval=FALSE}
+\SweaveOpts{eval=TRUE}
 
@@
 
-\vspace{-0.2cm}
+\vspace{0.5cm}
+%\vspace{-0.2cm}
 
 \footnotesize
-Hothorn T, Hornik K, Zeileis A (2006).
- \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
- \textit{Journal of Computational and Graphical Statistics},
- \textbf{15}(3), 651--674.
- \doi{10.1198/106186006X133933}
-
-\medskip
-
-Zeileis A, Hothorn T, Hornik K (2008).
- \dquote{Model-Based Recursive Partitioning.}
-  \textit{Journal of Computational and Graphical Statistics},
-  \textbf{17}(2), 492--514.
-  \doi{10.1198/106186008X319331}
-
-\medskip
+
+%Hothorn T, Hornik K, Zeileis A (2006).
+% \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
+% \textit{Journal of Computational and Graphical Statistics},
+% \textbf{15}(3), 651--674.
+% \doi{10.1198/106186006X133933}
+
+%\medskip
+
+%Zeileis A, Hothorn T, Hornik K (2008).
+% \dquote{Model-Based Recursive Partitioning.}
+%  \textit{Journal of Computational and Graphical Statistics},
+%  \textbf{17}(2), 492--514.
+%  \doi{10.1198/106186008X319331}
+
+%\medskip
+
+Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
+  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
+  \emph{Manuscript under preparation}.
+
+\bigskip
 
@@
 
-\medskip
+\bigskip
 
@@
 
-\medskip
-
-Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and Minimum CRPS Estimation}
-  \textit{Monthly Weather Review}, 
-  \textbf{133}(5), 1098--1118.
-  \doi{10.1175/MWR2904.1}
-
-\medskip
-
-Hofner B, Mayr A, Schmid M (2016).
-  \dquote{\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and Variable Selection in the {GAMLSS} Framework}
-  \textit{Journal of Statistical Software},
-  \textbf{74}(1), 1--31.
-  \doi{10.18637/jss.v074.i01}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{References (2)}
-
-\vspace{-0.2cm}
-
-\footnotesize
-
-% Reference for the observation data set
-BMLFUW (2016).
-  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtschaft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
-  \text{Available at \url{http://ehyd.gv.at}}, 
-  \text{Accessed: 2016--02--29}.
-
-\medskip
+%\medskip
+
+%Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
+%  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and Minimum CRPS Estimation}
+%  \textit{Monthly Weather Review}, 
+%  \textbf{133}(5), 1098--1118.
+%  \doi{10.1175/MWR2904.1}
+
+%\medskip
+
+%Hofner B, Mayr A, Schmid M (2016).
+%  \dquote{\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and Variable Selection in the {GAMLSS} Framework}
+%  \textit{Journal of Statistical Software},
+%  \textbf{74}(1), 1--31.
+%  \doi{10.18637/jss.v074.i01}
+
+\end{frame}
+
+%\begin{frame}
+%\frametitle{References (2)}
+
+%\vspace{-0.2cm}
+
+%\footnotesize
+
+%% Reference for the observation data set
+%BMLFUW (2016).
+%  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtschaft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
+%  \text{Available at \url{http://ehyd.gv.at}}, 
+%  \text{Accessed: 2016--02--29}.
+
+%\medskip
 
 % Reference for the GFS reforecast version 2 data set 
-Hamill T M, Bates G T, Whitaker J S, Murray D R, Fiorino M, Galarneau T J Jr., Zhu Y, Lapenta W (2013).
-  \dquote{{NOAA}'s Second-Generation Global Medium-Range Ensemble Reforecast Dataset}
-  \textit{Bulletin of the American Meteorological Society}, 
-  \textbf{94}(10), 1553--1565.
-  \doi{10.1175/BAMS-D-12-00014.1}
-
-\medskip
+%Hamill T M, Bates G T, Whitaker J S, Murray D R, Fiorino M, Galarneau T J Jr., Zhu Y, Lapenta W (2013).
+%  \dquote{{NOAA}'s Second-Generation Global Medium-Range Ensemble Reforecast Dataset}
+%  \textit{Bulletin of the American Meteorological Society}, 
+%  \textbf{94}(10), 1553--1565.
+%  \doi{10.1175/BAMS-D-12-00014.1}
+
+%\medskip
 
@@
 
-Messner J W,  Mayr G J, Zeileis A (2016).
-  \dquote{Heteroscedastic Censored and Truncated Regression with \textbf{crch}}
-  \textit{The \textsf{R} Journal}, 
-  \textbf{8}(1), 173--181.\\
-  \scriptsize
-  \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
-  \footnotesize
-
-\medskip
-
-Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
-  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored Additive Regression Model}
-  \textit{International Journal of Climatology}, 
-  \textbf{37}(7), 3264--3275.
-  \doi{10.1002/joc.4913}
-
-\medskip
-
-Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
-  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: Unanimous Predictions, Heavy Tails, and Link Functions}
-  \textit{Monthly Weather Review}, 
-  \text{Forthcoming}.
-  \doi{10.1175/MWR-D-16-0388.1}
+%Messner J W,  Mayr G J, Zeileis A (2016).
+%  \dquote{Heteroscedastic Censored and Truncated Regression with \textbf{crch}}
+%  \textit{The \textsf{R} Journal}, 
+%  \textbf{8}(1), 173--181.\\
+%  \scriptsize
+%  \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
+%  \footnotesize
+
+%\medskip
+
+%Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
+%  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored Additive Regression Model}
+%  \textit{International Journal of Climatology}, 
+%  \textbf{37}(7), 3264--3275.
+%  \doi{10.1002/joc.4913}
+
+%\medskip
+
+%Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
+%  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: Unanimous Predictions, Heavy Tails, and Link %Functions}
+%  \textit{Monthly Weather Review}, 
+%  \text{Forthcoming}.
+%  \doi{10.1175/MWR-D-16-0388.1}
 
@@
 
-\end{frame}
+%\end{frame}
 

### 2018-03-20 pkg/disttree/inst/draft/draft.Rnw
merged conference version of abstract into manuscript draft, some smaller touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2063 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9996ea3546043b2e9c1118daa67c0fd67c538abe

@@
 \Abstract{
-While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to model all distribution parameters.
-In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting.
-In generalized additive models for location, scale, and shape %(GAMLSS, \citealt{Rigby+Stasinopoulos:2005})
-each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects.
-A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees %(\citealt{Loh:2011})
-or random forests. %(\citealt{Breiman:2001}). 
-%%%<TH> "random forest" generisch; Breiman and Cutler's randomForest </TH>
-This class of tree and forest models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods.
-In particular, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests.
-%%%<TH> break-up very long sentence </TH>
-Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
+To obtain a probabilistic model for a dependent variable based on some set of
+explanatory variables, a distributional approach is often adopted where the
+parameter(s) of the distribution are linked to regressors. In many classical
+models this often only captures the location/expectation of the distribution but
+over the last decade there has been increasing interest in distributional
+regression approaches modeling all parameters including location, scale, and
+shape. Notably, so-called non-homogenous Gaussian regression models both
+mean and variance of a Gaussian response variable and is particularly
+popular in weather forecasting and, more generally, the GAMLSS framework
+allows to establish generalized additive models for location, scale, and shape
+with smooth linear or nonlinear effects.
+%
+However, in situations where variable selection is required and/or there are
+non-smooth dependencies or interactions (especially unknown or of high-order),
+it is challenging to establish a good GAMLSS. A natural alternative in these
+situations would be the application of regression trees or random forests but,
+so far, no general distributional framework is available for these. Therefore,
+a framework for distributional trees and forests is proposed that blends
+regression trees and random forests with classical distributions from the GAMLSS
+framework as well as their censored or truncated counterparts.
+%
+To illustrate these novel approaches in practice, they are employed to obtain
+probabilistic precipitation forecasts at numerous locations in a mountainous
+region (Tyrol, Austria) based on a very large number of numerical weather
+prediction quantities. Comparisons with GAMLSS using either a specification
+based on prior meteorological knowledge or a computationally more demanding
+boosting approach show that distributional random forests automatically
+select variables and interactions, performing on par or often even better than
+the GAMLSS approaches.
 }
@@
 this has been extended in various directions including:
+
+\pagebreak
+
 \begin{itemize}
@@
 
+<<echo=FALSE, results=hide>>=
+#### Axams prediction 24 - 4
+if(file.exists("rain_Axams_pred_24to4.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred_24to4.rda")
+  load("rain_Axams_pred_24to4.rda")
+} else {
+  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred_24to4.R")
+  source("rain_axams_pred_24to4.R")
+  library("gamlss.cens")
+  gen.cens("NO", type = "left")
+  res <- rain_axams_pred(seedconst = 7, ntree = 100,
+                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
+                         gamboost_cvr = TRUE,
+                         frac = FALSE)
+
+  save(res, file = "rain_Axams_pred_24to4.rda")
+}
+
+#### prepare data for plot of estimated density functions
+# predictions for one day (in each of the four years) 
+# (19th of July 2011 is missing)
+pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
+pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+cbind(df_exp, res$testdata[pdays,"robs"])
+
+# plot predicted distributions together with observations
+#set.seed(res$call$seedconst)
+set.seed(7)
+x <- c(0.01, sort(runif(500,0.01,8)))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
+dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
+
+# point mass (slightly shifted)
+pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
+pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
+pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
+
+# predictions
+pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(res$testdata[pdays,"robs"][4], crch::dcnorm(res$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+
+#legendheight
+lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
+lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
+lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
+
+## PIT histograms and qqr-plots
+
+## get predicted parameter
+# for learndata and testdata
+{
+  set.seed(7)
+  ## disttree
+  if(FALSE){
+    # in sample
+    pdt <- predict(res$dt, type = "parameter")
+    dt_mu_l <- pdt$mu 
+    dt_sigma_l <- pdt$sigma 
+    pit_dt_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = dt_mu_l, sd = dt_sigma_l))
+    pit_dt_l[res$learndata[,"robs"]>0, 1] <- pit_dt_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  pdt <- predict(res$dt, type = "parameter", newdata = res$testdata)
+  dt_mu_t <- pdt$mu 
+  dt_sigma_t <- pdt$sigma 
+  pit_dt_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = dt_mu_t, sd = dt_sigma_t))
+  pit_dt_t[res$testdata[,"robs"]>0, 1] <- pit_dt_t[res$testdata[,"robs"]>0, 2]
+
+  ## distforest
+  if(FALSE){
+    # in sample
+    pdf <- predict(res$df, type = "parameter")
+    df_mu_l <- pdf$mu
+    df_sigma_l <- pdf$sigma
+    pit_df_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = df_mu_l, sd = df_sigma_l))
+    pit_df_l[res$learndata[,"robs"]>0, 1] <- pit_df_l[res$learndata[,"robs"]>0, 2]
+  }
+  # out of sample
+  pdf <- predict(res$df, type = "parameter", newdata = res$testdata)
+  df_mu_t <- pdf$mu
+  df_sigma_t <- pdf$sigma
+  pit_df_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = df_mu_t, sd = df_sigma_t))
+  pit_df_t[res$testdata[,"robs"]>0, 1] <- pit_df_t[res$testdata[,"robs"]>0, 2]
+
+  ## gamlss
+  g_learndata <- res$learndata
+  g_learndata$robs <- Surv(g_learndata$robs, g_learndata$robs>0, type="left")
+  if(FALSE){
+    # in sample
+    g_mu_l <- predict(res$g, what = "mu", type = "response")
+    g_sigma_l <- predict(res$g, what = "sigma", type = "response")
+    pit_g_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = g_mu_l, sd = g_sigma_l))
+    pit_g_l[res$learndata[,"robs"]>0, 1] <- pit_g_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  g_mu_t <- predict(res$g, what = "mu", type = "response", newdata = res$testdata)
+  g_sigma_t <- predict(res$g, what = "sigma", type = "response", newdata = res$testdata)
+  pit_g_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = g_mu_t, sd = g_sigma_t))
+  pit_g_t[res$testdata[,"robs"]>0, 1] <- pit_g_t[res$testdata[,"robs"]>0, 2]
+
+  #gamboostLSS
+  if(FALSE){
+    #in sample
+    pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response")
+    gb_mu_l <- pgb$mu
+    gb_sigma_l <- pgb$sigma
+    pit_gb_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = gb_mu_l, sd = gb_sigma_l))
+    pit_gb_l[res$learndata[,"robs"]>0, 1] <- pit_gb_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response", newdata = res$testdata)
+  gb_mu_t <- pgb$mu
+  gb_sigma_t <- pgb$sigma
+  pit_gb_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = gb_mu_t, sd = gb_sigma_t))
+  pit_gb_t[res$testdata[,"robs"]>0, 1] <- pit_gb_t[res$testdata[,"robs"]>0, 2]
+
+  #EMOS
+  if(FALSE){
+    # in sample
+    ml_mu_l <- predict(res$ml, type = "location")     # returns parameter on response scale
+    ml_sigma_l <- predict(res$ml, type = "scale")
+    pit_ml_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = ml_mu_l, sd = ml_sigma_l))
+    pit_ml_l[res$learndata[,"robs"]>0, 1] <- pit_ml_l[res$learndata[,"robs"]>0, 2]
+  }  
+  # out of sample
+  ml_mu_t <- predict(res$ml, type = "location", newdata = res$testdata)     # returns parameter on response scale
+  ml_sigma_t <- predict(res$ml, type = "scale", newdata = res$testdata)
+  pit_ml_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = ml_mu_t, sd = ml_sigma_t))
+  pit_ml_t[res$testdata[,"robs"]>0, 1] <- pit_ml_t[res$testdata[,"robs"]>0, 2]
+
+}
+
+## Variable importance
+if(file.exists("vimp_crps.rda")){
+  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
+  load("vimp_crps.rda")
+} else {
+
+  # locally redefine logLik.distforest to use crps in varimp()
+  logLik.distforest <- function(object, newdata = NULL, ...) {
+    if(is.null(newdata)) {
+      newdata <- object$data
+    }
+
+    # predict parameter
+    pdf <- predict(object, newdata = newdata, type = "parameter")
+    df_mu <- pdf$mu
+    df_sigma <- pdf$sigma
+
+    # calculate CRPS
+    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
+
+    return(structure(crps, df = NA, class = "logLik"))
+  }
+
+  vimp_crps <- varimp(res$df, nperm = 1L)
+  #vimp10_crps <- varimp(res$df, nperm = 10L)
+  #vimp100_crps <- varimp(res$df, nperm = 100L)
+  #vimp1000_crps <- varimp(res$df, nperm = 1000L)
+
+  save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
+  #save(vimp10_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp10_crps.rda")
+  #save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
+  #save(vimp1000_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp1000_crps.rda")
+
+  rm(logLik.distforest) 
+}
+
+@
+
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.8\textwidth}
+<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
+plot(x = x, y = y1, type = "l", col = "red", 
+     main = paste0("July ", pday), ylab = "Density", 
+     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
+     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
+     xlim = c(-1.5,8))
+
+lines(x = x, y = y2, type = "l", col = "blue")
+lines(x = x, y = y3, type = "l", col = "darkgreen")
+lines(x = x, y = y4, type = "l", col = "purple")
+legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
+       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
+
+# plot point mass
+lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
+lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
+lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
+lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)
+
+points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
+points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
+points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
+points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)
+
+# plot predictions
+points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
+points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
+points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
+points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
+
+lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)
+
+# add labels
+text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
+text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
+text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
+text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
+@
+\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
+Observations are left-censored at 0.
+The corresponding predicted point mass is shown at the censoring point (0).}
+\end{figure}
+
 In weather forecasting, these properties are appealing in mountainous regions
@@
 
-<<echo=FALSE, results=hide>>=
-#### Axams prediction 24 - 4
-if(file.exists("rain_Axams_pred_24to4.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred_24to4.rda")
-  load("rain_Axams_pred_24to4.rda")
-} else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred_24to4.R")
-  source("rain_axams_pred_24to4.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-  res <- rain_axams_pred(seedconst = 7, ntree = 100,
-                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
-                         gamboost_cvr = TRUE,
-                         frac = FALSE)
-
-  save(res, file = "rain_Axams_pred_24to4.rda")
-}
-
-#### prepare data for plot of estimated density functions
-# predictions for one day (in each of the four years) 
-# (19th of July 2011 is missing)
-pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
-
-pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
-pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, res$testdata[pdays,"robs"])
-
-# plot predicted distributions together with observations
-#set.seed(res$call$seedconst)
-set.seed(7)
-x <- c(0.01, sort(runif(500,0.01,8)))
-y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
-y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
-y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
-y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
-dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
-
-# point mass (slightly shifted)
-pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
-pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
-pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
-pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
-
-# predictions
-pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(res$testdata[pdays,"robs"][4], crch::dcnorm(res$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
-
-#legendheight
-lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
-lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
-lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
-
-## PIT histograms and qqr-plots
-
-## get predicted parameter
-# for learndata and testdata
-{
-  set.seed(7)
-  ## disttree
-  if(FALSE){
-    # in sample
-    pdt <- predict(res$dt, type = "parameter")
-    dt_mu_l <- pdt$mu 
-    dt_sigma_l <- pdt$sigma 
-    pit_dt_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = dt_mu_l, sd = dt_sigma_l))
-    pit_dt_l[res$learndata[,"robs"]>0, 1] <- pit_dt_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  pdt <- predict(res$dt, type = "parameter", newdata = res$testdata)
-  dt_mu_t <- pdt$mu 
-  dt_sigma_t <- pdt$sigma 
-  pit_dt_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = dt_mu_t, sd = dt_sigma_t))
-  pit_dt_t[res$testdata[,"robs"]>0, 1] <- pit_dt_t[res$testdata[,"robs"]>0, 2]
-
-  ## distforest
-  if(FALSE){
-    # in sample
-    pdf <- predict(res$df, type = "parameter")
-    df_mu_l <- pdf$mu
-    df_sigma_l <- pdf$sigma
-    pit_df_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = df_mu_l, sd = df_sigma_l))
-    pit_df_l[res$learndata[,"robs"]>0, 1] <- pit_df_l[res$learndata[,"robs"]>0, 2]
-  }
-  # out of sample
-  pdf <- predict(res$df, type = "parameter", newdata = res$testdata)
-  df_mu_t <- pdf$mu
-  df_sigma_t <- pdf$sigma
-  pit_df_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = df_mu_t, sd = df_sigma_t))
-  pit_df_t[res$testdata[,"robs"]>0, 1] <- pit_df_t[res$testdata[,"robs"]>0, 2]
-
-  ## gamlss
-  g_learndata <- res$learndata
-  g_learndata$robs <- Surv(g_learndata$robs, g_learndata$robs>0, type="left")
-  if(FALSE){
-    # in sample
-    g_mu_l <- predict(res$g, what = "mu", type = "response")
-    g_sigma_l <- predict(res$g, what = "sigma", type = "response")
-    pit_g_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = g_mu_l, sd = g_sigma_l))
-    pit_g_l[res$learndata[,"robs"]>0, 1] <- pit_g_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  g_mu_t <- predict(res$g, what = "mu", type = "response", newdata = res$testdata)
-  g_sigma_t <- predict(res$g, what = "sigma", type = "response", newdata = res$testdata)
-  pit_g_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = g_mu_t, sd = g_sigma_t))
-  pit_g_t[res$testdata[,"robs"]>0, 1] <- pit_g_t[res$testdata[,"robs"]>0, 2]
-
-  #gamboostLSS
-  if(FALSE){
-    #in sample
-    pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response")
-    gb_mu_l <- pgb$mu
-    gb_sigma_l <- pgb$sigma
-    pit_gb_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = gb_mu_l, sd = gb_sigma_l))
-    pit_gb_l[res$learndata[,"robs"]>0, 1] <- pit_gb_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response", newdata = res$testdata)
-  gb_mu_t <- pgb$mu
-  gb_sigma_t <- pgb$sigma
-  pit_gb_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = gb_mu_t, sd = gb_sigma_t))
-  pit_gb_t[res$testdata[,"robs"]>0, 1] <- pit_gb_t[res$testdata[,"robs"]>0, 2]
-
-  #EMOS
-  if(FALSE){
-    # in sample
-    ml_mu_l <- predict(res$ml, type = "location")     # returns parameter on response scale
-    ml_sigma_l <- predict(res$ml, type = "scale")
-    pit_ml_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = ml_mu_l, sd = ml_sigma_l))
-    pit_ml_l[res$learndata[,"robs"]>0, 1] <- pit_ml_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  ml_mu_t <- predict(res$ml, type = "location", newdata = res$testdata)     # returns parameter on response scale
-  ml_sigma_t <- predict(res$ml, type = "scale", newdata = res$testdata)
-  pit_ml_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = ml_mu_t, sd = ml_sigma_t))
-  pit_ml_t[res$testdata[,"robs"]>0, 1] <- pit_ml_t[res$testdata[,"robs"]>0, 2]
-
-}
-
-## Variable importance
-if(file.exists("vimp_crps.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
-  load("vimp_crps.rda")
-} else {
-
-  # locally redefine logLik.distforest to use crps in varimp()
-  logLik.distforest <- function(object, newdata = NULL, ...) {
-    if(is.null(newdata)) {
-      newdata <- object$data
-    }
-
-    # predict parameter
-    pdf <- predict(object, newdata = newdata, type = "parameter")
-    df_mu <- pdf$mu
-    df_sigma <- pdf$sigma
-
-    # calculate CRPS
-    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
-
-    return(structure(crps, df = NA, class = "logLik"))
-  }
-
-  vimp_crps <- varimp(res$df, nperm = 1L)
-  #vimp10_crps <- varimp(res$df, nperm = 10L)
-  #vimp100_crps <- varimp(res$df, nperm = 100L)
-  #vimp1000_crps <- varimp(res$df, nperm = 1000L)
-
-  save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
-  #save(vimp10_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp10_crps.rda")
-  #save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
-  #save(vimp1000_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp1000_crps.rda")
-
-  rm(logLik.distforest) 
-}
-
-@
-
-\begin{figure}[t!]
-\centering
-\setkeys{Gin}{width=0.8\textwidth}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
-plot(x = x, y = y1, type = "l", col = "red", 
-     main = paste0("July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
-     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-1.5,8))
-
-lines(x = x, y = y2, type = "l", col = "blue")
-lines(x = x, y = y3, type = "l", col = "darkgreen")
-lines(x = x, y = y4, type = "l", col = "purple")
-legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
-       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
-
-# plot point mass
-lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
-lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
-lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
-lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)
-
-points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
-points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
-points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
-points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)
-
-# plot predictions
-points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
-points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
-points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
-points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
-
-lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)
-
-# add labels
-text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
-text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
-text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
-text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
-@
-\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
-Observations are left-censored at 0.
-The corresponding predicted point mass is shown at the censoring point (0).}
-\end{figure}
-
 \section{Methodology}
@@
 
-\subsection{Motivation}
 As illustrated in Figure~\ref{Devel_parmod} a wide range of tools has been developed 
@@
 
-The goal is to fit a distributional model $D(Y, \theta)$ to a response variable 
+The goal is to fit a distributional model $\mathcal{D}(Y, \theta)$ to a response variable 
 $Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters 
@@
 $g$. In that way the fitted model would be specified as 
-$D(Y, \theta = g(\bold{Z}))$. In order to avoid the restriction to linear 
+$\mathcal{D}(Y, \theta = g(\bold{Z}))$. In order to avoid the restriction to linear 
 effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ 
 to specify a more flexible model 
-$D(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure. 
+$\mathcal{D}(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure. 
 Another approach to distributional regression is proposed by 
@@
 \label{disttree}
-Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$, 
+Given the fit of a global distributional model $\mathcal{D}(Y, \bold{\hat{\theta}})$, 
 the idea is to use the information provided by this model together with a 
@@
 \begin{itemize}
-\item A distributional model $D(Y, \theta)$ is fit to the set of 
+\item A distributional model $\mathcal{D}(Y, \theta)$ is fit to the set of 
 observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution 

### 2018-03-20 pkg/disttree/inst/draft/draft.Rnw
work on Section 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2064 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1b2913aa57cea1e52d9efe57cccf8e122c331578

@@
 %% packages
-\usepackage{amstext}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{thumbpdf,lmodern}
-%\usepackage{rotating}
-%\usepackage{caption}
-%\captionsetup{format=hang}
-%\usepackage{subcaption}
-\usepackage{Sweave}
-%\usepackage{enumitem}
-%\usepackage{graphicx}
-\usepackage{array, makecell}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta}
-\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-%\usepackage[pagebackref=true]{hyperref}
-%\usepackage{doi}
+\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}
+
+%% tikz
+\usepackage{array,makecell,tikz}
+\usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
+
+%% need no \usepackage{Sweave}
+\SweaveOpts{concordance=FALSE, eps=FALSE, keep.source=TRUE, eval = TRUE}
 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
-
-\SweaveOpts{eps=FALSE, keep.source=TRUE, eval = TRUE}
-
-\SweaveOpts{eval = TRUE}
+
 \section{Introduction}
@@
 
-As illustrated in Figure~\ref{Devel_parmod} a wide range of tools has been developed 
-in the field of parametric modeling. Especially the shift from conditional mean 
-regression models to distributional regression models has increased the application 
-spectrum by allowing for much broader inference. In some cases further features 
-such as the ability to capture non-additive effects or interactions are desirable 
-which demands for more flexibility. Since tree-structured models provide these 
-required features, combining them with parametric models can offer a framework 
-that is able to deal with a wide variety of different modeling situations. 
-In this section the novel methodology which embeds parametric modeling into 
-the idea of tree-structured models will be introduced step by step. Starting 
-with a simple distributional fit in Section~\ref{distfit} will be followed by 
-building a distributional tree in Section~\ref{disttree} and finally a distributional 
-forest in Section~\ref{distforest}.
+To embed the distributional approach from GAMLSS into regression trees and random
+forests, we proceed in three steps. (1)~To fix notation, we briefly review fitting
+distributions using standard maximum likelihood in in Section~\ref{distfit}.
+(2)~A recursive partitioning strategy based on the corresponding scores (or
+gradients) is introduced in Section~\ref{disttree}, leading to distributional
+trees. (3)~Ensembles of distributional trees fitted to randomized subsamples are
+employed to establish distributional forests in Section~\ref{distforest}.
+
+The general distributional notation is exemplified in all three steps for the
+zero-censored Gaussian distribution. The latter is employed in the empirical
+in the empirical case study in Section~\ref{ProbabilisticForecasting} to model
+power-transformed daily precipitation amounts.
 
@@
 \label{distfit}
-Fitting a distributional model to a set of observations without considering any 
-covariates is a well known procedure which can for example be done by applying 
-the maximum likelihood method. This is also the chosen approach in the methodology 
-described in this section. 
-%%% <TH> is is repeated on page 5, remove here </TH>
-
-The goal is to fit a distributional model $\mathcal{D}(Y, \theta)$ to a response variable 
-$Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters 
-$\theta \in \Theta$ in case of a \textit{k}-parametric distribution. The distribution 
-family with density function $f$ has to be specified in advance such that a 
-log-likelihood function $\ell(\theta; Y) = \log(f(Y; \theta)$ is provided. Using 
-the GAMLSS framework (\citealt{Rigby+Stasinopoulos:2005}) a wide range of distributions 
-can be selected where each distribution parameter can be specified separately. 
-This enables the model to deal with various features that can make the task of 
-fitting a model very challenging such as censoring, truncation or heavy tails.
-
-In our application of precipitation forecasting the response variable, which is 
-the power transformed total precipitation amount within 24 hours, is expected 
-to be normally distributed with mean/location parameter $\mu$ and standard 
-deviation/scale parameter $\sigma$ and left-censored with censoring point 0. 
-(The power transformation is used to remove large parts of the skewness of 
-the data as further explained in Section~\ref{data}.) Therefore, the 
-corresponding log-likelihood function with parameter vector 
-$\theta = (\mu, \sigma)$ is
+
+A distributional model $\mathcal{D}(Y, \bm{\theta})$ is considered for the response
+variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
+\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$. Based on the
+corresponding probability density function $f(Y; \bm{\theta})$ the log-likelihood
+function is defined $\ell(\bm{\theta}; Y) = \log\{f(Y; \bm{\theta})\}$. The GAMLSS
+framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
+distributional families with parameterizations corresponding to location, scale,
+and shape. Furthermore, censoring and/or truncation of these distributions
+can be incorporated in the usual straightforward way \fixme{reference needed}.
+
+To capture both location and scale of the probabilistic precipitation forecasts
+while accounting for a point mass at zero precipitation (i.e., dry days without
+rain), zero-censored Gaussian distribution with with mean/location parameter
+$\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
+the corresponding log-likelihood function with parameter vector 
+$\bm{\theta} = (\mu, \sigma)$ is
 $$
@@
 \begin{cases}
-    \log\left(\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right), & \text{if } Y > 0\\
-    \vspace{0.1cm}
-    \log\left(\Phi\left(\frac{-\mu}{\sigma}\right)\right), & \text{if } Y = 0
+    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
+    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
 \end{cases}
@@
 distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
-Next to the censored normal distribution also the censored logistic 
-distribution is often chosen for the modeling of precipitation amounts. In 
-\cite{Gebetsberger+Messner+Mayr:2017} this choice has been made in order 
-to be able to deal with heavy tails which often appear in precipitation 
-data. Another frequent application of probabilistic forecasting in meteorology 
-is the modeling of near surface air temperature. In this case a normal 
-distribution is often selected.
+Other distributions $\mathcal{D}$ and corresponding log-likelihoods
+$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
+censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
+zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.
 
@@
 function the task of fitting a distributional model turns into the task 
-of determining the distribution parameter~$\theta$. This is done by the 
-following maximization based on a learning sample $\{y_i\}_{i=1,\ldots,n}$ 
-of the response variable $Y$.
-$$
-\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
-$$
-which is equal to solving 
-$$
-\sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = 0.
-$$
-In that way, a whole distribution is specified with all its features, 
-including location, scale, and shape. However, fitting one global model 
-to the whole data set might be too much of a generalization and therefore 
-not represent the data and its features in a reasonable way. Moreover, 
-if covariates are available it is desirable to include them into the model 
-as they might provide important additional information. In our application 
-on precipitation data a wide range of covariates is available from a numerical 
-ensemble prediction system considering precipitation, temperature, sea level 
-pressure and other meteorological quantities which might be of interest and 
-have an influence on the predictions.
-
-A common way of including these covariates is to consider them as regressors 
-in a distributional model such that each parameter $\theta$ depends on covariates 
-$\bold{Z} = (Z_1,\ldots,Z_m)$. This could for example be done by a linear function 
-$g$. In that way the fitted model would be specified as 
-$\mathcal{D}(Y, \theta = g(\bold{Z}))$. In order to avoid the restriction to linear 
-effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ 
-to specify a more flexible model 
-$\mathcal{D}(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure. 
-Another approach to distributional regression is proposed by 
-\cite{Klein+Kneib+Lang:2015} where each parameter of a possibly complex 
-distribution can be modeled by an additive term consisting of a variety of 
-different functional effect types. These include non-linear effects, spatial 
-effects, random coefficients and interaction surfaces which extends the 
-flexibility compared to parametric models such as GAMLSS modeling only 
-smooth effects.
-
-A different approach on how to include covariates that also allows for 
-non-additive effects and interactions is considered in the following. 
-In particular, the covariates can be used to separate the data into more 
-homogeneous subgroups. Doing so after fitting a global model to the whole 
-data set (without considering any covariates) and then fitting a local 
-model to each subgroup should improve the model remarkably. This 
-procedure of splitting a data set into subgroups can be done by 
-applying a tree algorithm.
+of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
+maximum likelihood (ML) based on the learning sample
+$\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
+likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
+$$ \label{eq:mle}
+\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
+$$
+Equivalently, this can be defined based on the corresponding first-order
+conditions
+$$ \label{eq:foc}
+\sum_{i = 1}^n s(\bm{\hat \theta}, y_i) = 0,
+$$
+where $s(\bm{\theta}, y_i)$ is the associated score function
+$$ \label{eq:score}
+s(\bm{\theta}, y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
+$$ 
+The latter is subsequently employed as a general goodness-of-fit measure
+for individual observations $y_i$ and parameters $\bm{\theta}$.
 
@@
 \label{disttree}
-Given the fit of a global distributional model $\mathcal{D}(Y, \bold{\hat{\theta}})$, 
-the idea is to use the information provided by this model together with a 
-set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ 
-to decide whether the data set should be split into subgroups or not. 
-In particular, the information provided by the fitted model are the scores 
-$$
-s(\hat{\theta}, y_i) = \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i).
-$$ 
+
+Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
+for reasonably representing the data and covariates $\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
+employed to capture differences in the distribution parameters $\bm{\theta}$.
+In weather forecasting, these covariates typically include the output
+from numerical weather prediction systems and/or lagged weather observations.
+
+To incorporate the covariates into the distributional model, GAMLSS considers
+them as regressors so that an additive predictor
+$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
+with link function $g_j(\cdot)$ is used for every parameter $\theta_j$
+($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
+nonlinear effects, spatial effects, random coefficients, or interaction surfaces
+\citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
+additive terms and their functional forms in advance which is often difficult in
+practice, especially if the number of covariates $m$ is large.
+
+Regression trees generally take a different approach for automatically
+including covariates in a data-driven way that also allows for abrupt
+changes, non-additive effects, and interactions. In the context of
+distributional models the goal is to partition the covariate space
+$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
+distributional model for the response $Y$ can be found with segment-specific
+parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
+($b = 1, \dots, B$) partition the covariate space
+$$ \label{eq:partition}
+\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
+$$
+and a local distributional model $\mathcal{D}(Y, \bm{theta}^(b))$
+(i.e., with segment-specific parameters $\bm{theta}^(b)$) is fitted to the
+response $Y$ in that segment.
+
+To find the segments $\mathcal{B}_b$ that are (approximately) homogenous
+with respect to the distributional model with given parameters, the idea
+is to use a gradient-based recursive partitioning approach. In a given
+subsample of the learning data this fits the model by ML (see
+Equation~\ref{eq:mle}) and then assess the goodness of fit by looking at the
+scores $s(\bm{\hat \theta}; y_i)$ (see Equation~\ref{eq:score}).
+
+\fixme{still more to do}
+
 These values express how well the model fits the data regarding each 
@@
 $$
-\hat{\theta}(\bold{z}) = \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\theta; y_i).
+\hat{\theta}(\bold{z}) = \argmax_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\theta; y_i).
 $$
@@
 $$
-w^{\text{forest}}_i(\bold{z}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
+w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
 $$
@@
 $$
-\hat{\theta}(\bold{z}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\theta; y_i)
+\hat{\theta}(\bold{z}) =  \argmax_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\theta; y_i)
 $$
@@
 $$
-\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
+\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
 $$

### 2018-03-21 pkg/disttree/inst/draft/draft.Rnw
revised Section 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2065 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 07729d4692a600b75d13731454cad8d58097015e

@@
 parameter(s) of the distribution are linked to regressors. In many classical
-models this often only captures the location/expectation of the distribution but
+models this only captures the location/expectation of the distribution but
 over the last decade there has been increasing interest in distributional
 regression approaches modeling all parameters including location, scale, and
-shape. Notably, so-called non-homogenous Gaussian regression models both
-mean and variance of a Gaussian response variable and is particularly
+shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
+mean and variance of a Gaussian response and is particularly
 popular in weather forecasting and, more generally, the GAMLSS framework
@@
 %
-However, in situations where variable selection is required and/or there are
+However, when variable selection is required and/or there are
 non-smooth dependencies or interactions (especially unknown or of high-order),
@@
 region (Tyrol, Austria) based on a very large number of numerical weather
-prediction quantities. Comparisons with GAMLSS using either a specification
-based on prior meteorological knowledge or a computationally more demanding
-boosting approach show that distributional random forests automatically
-select variables and interactions, performing on par or often even better than
-the GAMLSS approaches.
+prediction quantities. It is shown that distributional random forests
+automatically select variables and interactions, performing on par or often even
+better than GAMLSS specified either through prior meteorological knowledge or a
+computationally more demanding boosting approach.
 }
@@
 \setkeys{Gin}{width=0.8\textwidth}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
+<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4>>=
+par(mar = c(4, 4, 3, 1.5))
 plot(x = x, y = y1, type = "l", col = "red", 
@@
 The latter is subsequently employed as a general goodness-of-fit measure
-for individual observations $y_i$ and parameters $\bm{\theta}$.
+to assess how well the distribution with parameters $\bm{\theta}$ fits
+one individual observation $y_i$.
 
@@
 Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
-for reasonably representing the data and covariates $\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
+for reasonably representing the response distribution and covariates
+$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
 employed to capture differences in the distribution parameters $\bm{\theta}$.
@@
 To incorporate the covariates into the distributional model, GAMLSS considers
-them as regressors so that an additive predictor
-$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
-with link function $g_j(\cdot)$ is used for every parameter $\theta_j$
+them as regressors in additive predictors
+$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$.
+Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
 ($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
@@
 \citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
-additive terms and their functional forms in advance which is often difficult in
+additive terms and their functional forms in advance which can be challenging in
 practice, especially if the number of covariates $m$ is large.
@@
 Regression trees generally take a different approach for automatically
-including covariates in a data-driven way that also allows for abrupt
+including covariates in a data-driven way and allowing for abrupt
 changes, non-additive effects, and interactions. In the context of
@@
 $$
-and a local distributional model $\mathcal{D}(Y, \bm{theta}^(b))$
-(i.e., with segment-specific parameters $\bm{theta}^(b)$) is fitted to the
+and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
+(i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
 response $Y$ in that segment.
@@
 with respect to the distributional model with given parameters, the idea
-is to use a gradient-based recursive partitioning approach. In a given
+is to use a gradient-based recursive-partitioning approach. In a given
 subsample of the learning data this fits the model by ML (see
-Equation~\ref{eq:mle}) and then assess the goodness of fit by looking at the
-scores $s(\bm{\hat \theta}; y_i)$ (see Equation~\ref{eq:score}).
-
-\fixme{still more to do}
-
-These values express how well the model fits the data regarding each 
-parameter. Ideally they should fluctuate randomly around zero, 
-similar to residuals in ordinary least squares estimation, but in 
-this case there is a score value for each pair of estimated parameter 
-$\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ 
-and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$. This enables the 
-model to be sensitive to changes in each of the parameters which is one 
-of the main advantages of distributional trees over other tree models 
-considering only the mean of the distribution.
-
-Statistical tests are applied to check whether there is a significant 
-dependency between the scores and each of the partitioning variables 
-$Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$. If so, the data set is split 
-into subgroups based on the values of the partitioning variable that 
-shows the highest dependency. The exact way of determining how and where 
-to split varies across different tree algorithms. For distributional trees 
-one possible choice is a permutation test based algorithm which is explained 
-in detail in the appendix.
-
-After splitting the data the next step is to 
-fit a local distributional model in each of the subgroups. Repeating this 
-procedure of fitting a model, applying statistical tests to evaluate parameter 
-instability and splitting the data set depending on the test results leads 
-to a tree structure. In that way the learning data set is separated into 
-\textit{B} disjoint segments such that
-$$
-\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b
-$$
-where each segment is represented by a terminal node of the tree.
-
-Summed up, the steps of building a distributional tree are:
+Equation~\ref{eq:mle}) and then assesses the goodness of fit by assessing the
+corresponding scores $s(\bm{\hat \theta}; y_i)$ (see Equation~\ref{eq:score}).
+
+In summary, distributional trees are fitted recursively via:
 \begin{enumerate}
-\item Specify a distribution family with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(\hat{\theta}, y_i) = 0$.
-\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ 
-and each partitioning variable~$Z_l$.
-\item Split the sample along the partitioning variable with the strongest association 
-or instability. Choose the breakpoint with the highest improvement in log-likelihood.
-\item Repeat steps 2--4 recursively in the subgroups until some stopping criterion is reached.
-
+\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
+  in the current subsample.
+\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
+  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
+\item Split the sample along the partitioning variable $Z_{l,i}^*$ with the
+  strongest association or instability. Choose the breakpoint with the highest
+  improvement in the log-likelihood or the highest discrepancy .
+\item Repeat steps 1--3 recursively in the subsamples until these become too
+  small or there is no significant association/instability (or some other
+  stopping criterion is reached).
 \end{enumerate}
-The stopping criterion in step 5 can for example be a minimum number of 
-observations in a node or a p-value for the statistical tests applied in step 3.
-%%% <TH> we need a little more detail here or a reference to a better
-%%% description of mob/ctree OR appendix</TH>
-
-Executing these steps results in a stepwise parametric model where a complete 
-distribution is fit in each node of the tree.
-
-For distributional trees a prediction for a (possibly new) set of covariates 
-$\bold{z} = (z_1, \ldots, z_m)$ is made with the following strategy: 
-All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ 
-which end up in the same terminal node of the tree as $\bold{z}$ are selected 
-and a distributional model is fit on this subset. Selecting these observations
-is done by including a weight vector 
-$\bold{w}^{\text{tree}} = (w^{\text{tree}}_1, \ldots, w^{\text{tree}}_n)$ 
-in the fitting process where
-$$
-w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)).
-$$
-The predicted distribution for $\bold{z}$ is then fully 
-specified by the estimated parameter $\hat{\theta}(\bold{z})$ where
-$$
-\hat{\theta}(\bold{z}) = \argmax_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\theta; y_i).
-$$
-
-While the clear structure of such a tree model is advantageous in 
-interpreting and analyzing it, the abrupt changes are often too rough 
-and impose steps on the model even if the true underlying effect is 
-smooth. In these situations bagging can solve the problem by combining 
-an ensemble of trees such that wrongly detected abrupt changes of the 
-model are turned into almost smooth transitions. In that way, all kinds 
-of effects can be modeled, which is also one the of the main advantages 
-of forest models taking bagging one step further as the variance is 
-reduced by sampling the covariates considered for the splitting process 
-in each node.
-
-%%% <TH> we need to be much more precise here: random forests is Breiman's and Cutler's
-%%% random forest. Maybe it is better to not mention this method at any time. </TH>
-Common forest models such as the original forest technique by 
-\cite{Breiman+Cutler:2004} consist of trees only considering the mean of the 
-distribution of the response variable. For that reason, changes in other 
-parameters can only be detected if they are connected to changes in the 
-location parameter, e.g. if they depend on the location parameter. 
-As explained before, a distributional tree considers scores for all 
-distribution parameters. Therefore, a forest consisting of 
-distributional trees can also capture separate effects in each of the 
-distribution parameters. For that reason distributional forests are
-now going to be introduced.
+%
+Different inference techniques can be used for assessing the association between
+scores and covariates in step~3. In the following we use the general class of
+permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
+also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
+Alternatively, one could use asymptotic M-fluctuation tests for parameter
+instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
+\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
+Appendix~\ref{app:tree}.
+
+For obtaining probabilistic predictions from the tree for a (possibly new) set
+of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
+be ``sent down'' the tree and the corresponding segment-specific MLE has to be
+obtained. This can also be understood as a weighted MLE where the weights
+select those observations from the learning sample that fall into the same
+segment:
+$$
+w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
+$$
+where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
+for a given $\bold{z}$ is then fully specified by the estimated parameter
+$\bm{\hat \theta}(\bold{z})$ where
+$$
+\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
+$$
 
@@
 \label{distforest}
-A distributional forest consists of an ensemble of $T$ distributional trees. 
-Each of these trees is built on a different data set which can either be a bootstrap 
-or a subset of the original data set. Moreover, in each node a subset of the 
-partitioning variables is chosen randomly as candidates for splitting variables. 
-In that way the correlation among the resulting trees is reduced and therefore also 
-the variance of the model (\citealt{Breiman:2001}).
-
-Forest models differ mainly in the way how the trees are combined to make predictions. 
-Following the strategy explained and proposed by \cite{Hothorn+Zeileis:2017} we 
-interpret random forests as adaptive local likelihood estimators as distributional 
-models are fit locally in subgroups while the term ``adaptive'' refers to the 
-dependence on the splitting variables $\bold{Z}$ and the resulting subgroups 
-and weighing schemes. In this case the forest itself is only used to evaluate 
-neighborhood relationships among the observations which can then be included 
-in the fitting process. In transformation forests by \cite{Hothorn+Zeileis:2017} 
-so-called ``nearest neighbor weights'' are calculated as a measure of distance 
-between observations. This approach has also already been applied in several 
-random forest-type methods such as in ``adaptive nearest neighbors'' by 
-\cite{Lin+Jeon:2006}, in ``quantile regression forests''  by  \cite{Meinshausen:2006} 
-and in ``bagging survival trees'' by \cite{Hothorn+Lausen+Benner:2004}. 
-The idea of nearest neighbor weights is to measure similarity of a pair of 
-observations by counting the number of trees in which it is asigned to the 
-same terminal node.
-%Similar observations have a high probability of ending up in the same terminal 
-%node whereas this probability is low for quite different observations. 
-%although the first random forest-type algorithm for the estimation of 
-%conditional distribution functions was published more than a decade ago 
-%(bagging survival trees, Hothorn et al. 2004). 
-In particular, for a (possibly new) set of covariates 
-$\bold{z} = (z_1, \ldots, z_m) $  such a set of weights 
-$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ 
-is obtained by the following procedure: For each tuple $(y_i,\bold{z}_i)$ in 
-the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated 
-by counting the number of trees in which $(y_i,\bold{z}_i)$ ends up in the same 
-terminal node as $\bold{z}$:
-$$
-w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
-$$
-where $T$ is the number of trees of the forest, $B^t$ is the number of terminal 
-nodes of the \textit{t}-th tree and $\mathcal{B}^t_b$ represents the \textit{b}-th 
-terminal node of the \textit{t}-th tree.
-
-Therefore, contrary to distributional trees, in distributional forests the whole 
-learning data set can be used to fit a distributional model and specify the 
-predicted distribution for a set of covariates. Compared to the formula for
-predictions made by distributional trees the vector of tree weights 
-$\bold{w}^{\text{tree}}$ containing only zeros and ones is now replaced by a 
-vector of integer valued forest weights.
-%%% <TH> we need references for the nearest neighbor weights, copy from
-%%% "Transformation Forests" </TH>
-%%%<TH> we need to cite "Transformation Forests" for the interpretation of
-%%% random forests as adaptive local ML </TH>
-This individual set of weights 
-$\bold{w}^{\text{forest}}(\bold{z}) = (w^{\text{forest}}_1(\bold{z}), \ldots, w^{\text{forest}}_n(\bold{z}))$ 
-for the set of covariates $\bold{z}$ can now be included in the 
-estimation process of the distribution parameters and in that way leads 
-to its individual parameter vector $\hat{\theta}(\bold{z})$ where
-$$
-\hat{\theta}(\bold{z}) =  \argmax_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\theta; y_i)
-$$
-While in a distributional tree the same distribution is specified for all 
-sets of covariates being in the same node, a distributional forest predicts 
-and fits a different distribution for each different set of covariates due 
-to the individual weights calculated separately for each set of covariates.
-
-As this novel methodology combines the advantages of smooth and flexible 
-parametric modeling with those of tree-structured modeling it provides a 
-model that can deal with a wide range of situations. Abrupt changes and 
-non-additive effects can be captured by its tree structure while smooth 
-effects can be modeled reasonably well by combining various tree and 
-smoothing the steps imposed by a single tree. As scores are considered 
-for each distribution parameter (separate) effects in different parameters 
-can be captured. At the same time, also interactions are detected 
-automatically by splitting the data in a way that is sensitive to changes 
-in either each of the parameters or combined in various parameters.
+
+While the simple recursive structure of a tree model is easy to visualize and
+interpret, the abrupt changes are often too rough, instable, and impose steps
+on the model even if the true underlying effect is smooth. Hence, ensemble
+methods such as bagging or random forests \citep{Breiman:2001} typically smooth
+the effects, stabilize the model, and improve predictive performance.
+
+While classical random forests \citep{Breiman:2001} grow ensembles of trees
+that pick up changes in the location of the response across the covariates,
+\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. These
+pick up changes in the ``direction'' of any distribution parameter by
+considering the full score vector for choosing splitting variables and split
+points. Each of the distributional trees is grown on a different data set
+obtained through bootstrap sampling (or subsampling) and in each node only a
+random subset of the covariates $\bold{Z}$ is considered. As usual in random
+forests, this reduces the correlation among the trees and stabilizes the
+variance of the model.
+
+To obtain probabilistic predictions from a distributional forest, it still
+needs to be specified how to compute the parameter estimates
+$\bm{\hat \theta}(\bold{z})$ for a (potentially new) set of covariates $\bold{z}$.
+Following \cite{Hothorn+Zeileis:2017} we interpret random forests as adaptive
+local likelihood estimators using the averaged ``nearest neighbor weights''
+\citep{Lin+Jeon:2006} from the $T$~trees in the forest
+$$
+w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} \mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
+$$
+Thus, these $w^{\text{forest}}_i(\bold{z}) \in [0, 1]$ whereas
+$w^{\text{tree}}_i(\bold{z}) \in \{0, 1\}$. Hence, weights cannot only be $0$
+or $1$ but change more smoothly, giving high weight to those observations $i$
+from the learning sample that co-occur in the same segment $\mathcal{B}_b^t$
+as the new observation $\bold{z}$ for many of the trees $t = 1, \dots, T$.
+Consequently, the parameter estimates may, in principle, change for every
+observation and can be obtained by
+$$
+\bm{\hat \theta}(\bold{z}) =  \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
+$$
+In summary, this yields a parametric distributional regression model 
+(through the score-based approach) that can capture both abrupt effects and
+high-order interactions (through the trees) and smooth effects (through
+the forest).
 
@@
 provided in Table~\ref{covariates}, yielding 80~predictor variables in total.
-(Since the minimum of the ``variable downwards short wave radiation flux'' (\emph{dswrf}) over 24 hours is always 0, the ensemble mean and standard deviation of these minimal values are not considered.) \fixme{Mention that dswrf\_mean/sprd\_min is not included?}
+(Since the minimum of the ``variable downwards short wave radiation flux'',
+\emph{dswrf}, over 24 hours is always 0, the ensemble mean and standard
+deviation of these minimal values are not considered.)
 
@@
 \label{Application for one station}
-First of all we focus on one observation station, Axams in Tyrol. This 
-station is selected due to its geographical closeness to Innsbruck, 
-the capital of Tyrol. As for all other stations, daily observations and 
-numerical weather predictions are available for the month of July of 
+
+In a first step, we show a detailed comparison of the competing models
+for one observation station, Axams in Tyrol (selected somewhat arbitrarily
+as the closest in the data set to Innsbruck, capital of Tyrol). As for all
+other stations, daily observations and  numerical weather predictions are available for the month of July of 
 28~years from 1985 through the year 2012. In Figure~\ref{axams_onestation} 
@@
 
-\section{Tree algorithm}
-\label{treealgorithm}
+\section{Tree algorithm} \label{app:tree}
 The tree algorithm used in the applications discussed in this paper 
@@
 \begin{itemize}
-\item A distributional model $\mathcal{D}(Y, \theta)$ is fit to the set of 
+\item A distributional model $\mathcal{D}(Y, \bm{\theta})$ is fit to the set of 
 observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution 
-parameter $\hat{\theta}$. This parameter can be a multidimensional 
+parameter $\bm{\hat \theta}$. This parameter can be a multidimensional 
 vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, 
 depending on the type of distribution. The estimation is done by 
-maximizing the log-likelihood function $\ell(\theta; Y)$, i.e.
+maximizing the log-likelihood function $\ell(\bm{\theta}; Y)$, i.e.
 $$

### 2018-03-21 pkg/disttree/inst/draft/draft.Rnw
work on Axams section and Appendix B, unified all label{}
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2067 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a78ecf18e2dcf18535dd81ad1c45b17dc05c3474

@@
 %\VignetteIndexEntry{disttree: Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
-%\VignetteDepends{partykit, gamlss.dist}
-%\VignetteKeywords{regression trees, random forests, distributional regression, recursive partitioning}
+%\VignetteDepends{disttree, crch, countreg, gamlss, gamlss.cens}
+%\VignetteKeywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}
 %\VignettePackage{disttree}
@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{concordance=FALSE, eps=FALSE, keep.source=TRUE, eval = TRUE}
+\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
 
@@
 
-\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
 <<preliminaries, echo=FALSE, results=hide>>=
@@
 library("disttree")
-library("Formula")  # FIX ME: should be imported in disttree
+library("crch")
+library("countreg")
 library("gamlss")
-library("crch")
 library("gamlss.cens")
 gen.cens(NO, type = "left")
-library("mgcv")
-library("countreg")
 
@@
 \section{Introduction}
+\label{sec:introduction}
 
@@
 is modeled through a linear function of the covariates (see the left panel of
-Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
+Figure~\ref{fig:gam} for a schematic illustration). Over the last decades
 this has been extended in various directions including:
@@
     allowing for smooth nonlinear effects in the covariates
-    (Figure~\ref{Devel_parmod}, middle).
+    (Figure~\ref{fig:gam}, middle).
   \item \emph{Generalized additive models for location, scale, and shape}
@@
     on an additive predictor of the covariates comprising linear and/or
-    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
+    smooth nonlinear terms (Figure~\ref{fig:gam}, right).
 \end{itemize}
@@
 \end{tikzpicture}
-\caption{\label{Devel_parmod}Parametric modeling developments.
-(Generalized) linear models (left), generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
+\caption{\label{fig:gam}Parametric modeling developments. (Generalized) linear models (left), generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
 \end{figure}
@@
 
-%% late: introduce MOS and EMOS? more details about precipitation forecasting
-%In ensemble model output statistics (EMOS, \cite{Gneiting+Raftery+Westveld:2005}) this idea is taken one step further.
-Usually acting on the assumption of a normally distributed response variable this model allows for both parameters, the mean and the variance, to depend linearly on covariates provided by NWPs.
-However, the restriction to model only linear effects has risen the interest in applying more flexible models such as offered by the GAMLSS framework.
-
 \begin{figure}[t!]
@@
 \end{tikzpicture}
-\caption{\label{Devel_treeforest}Tree and forest developments.
+\caption{\label{fig:tree}Tree and forest developments.
 Regression tree (top left), distributional tree (bottom left), random forest
@@
     split the data into more homogeneous subgroups and can thus capture abrupt shifts
-    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
+    (Figure~\ref{fig:tree}, top left) and approximate nonlinear functions.
     Furthermore, trees automatically carry out a forward selection of covariates and
@@
     stabilizes the recursive partitions from individual trees and hence better
-    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
+    approximates smooth functions (Figure~\ref{fig:tree}, top middle)
 \end{itemize}
@@
 we propose to follow the ideas from GAMLSS modeling -- as outlined
-in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
+in Figure~\ref{fig:gam} -- and combine tree-based methods with parametric distributional models,
 yielding two novel techniques:
@@
     respect to a parametric distribution, thus capturing changes in any distribution
-    parameter like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
+    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left)
   \item \emph{Distributional forests} utilize an ensemble of distributional trees
-    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
+    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
     top right).
@@
 
-## PIT histograms and qqr-plots
+## PIT histograms and qqrplots
 
@@
 @
-\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
+\caption{\label{fig:axams-dist}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
 Observations are left-censored at 0.
@@
 meteorological stations in a mountainous region in the Alps, covering mainly
-Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{map}).
+Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
 More specifically, a zero-censored Gaussian
@@
 zero-censored point mass describes the probability of no precipitation on a
-given day (see Figure~\ref{axams_onestation}). Forecasts for July are
+given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
 established based on data from the same month in the years 1985--2012 along with
 80~covariates derived from a wide range of NWP ensemble outputs. As
-Figure~\ref{axams_onestation} shows, the station-wise models yield a full
+Figure~\ref{fig:axams-dist} shows, the station-wise models yield a full
 distributional forecast for each day -- here for the same day (July~24) at one
@@
 as training data. The corresponding observations conform reasonably well with
-the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
+the predictions. In Section~\ref{sec:precipitation} we investigate the
 performance of distributional forests in this forecasting task in more detail.
@@
 three alternative zero-censored Gaussian models: a standard ensemble model
-output statistics approach (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}), a GAMLSS
-with regressors prespecified based on meteorological expertise following
-\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
+output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
+on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
+\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
 \citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
@@
 \section{Methodology}
-\label{Methodology}
+\label{sec:methods}
 
@@
 forests, we proceed in three steps. (1)~To fix notation, we briefly review fitting
-distributions using standard maximum likelihood in in Section~\ref{distfit}.
+distributions using standard maximum likelihood in in Section~\ref{sec:distfit}.
 (2)~A recursive partitioning strategy based on the corresponding scores (or
-gradients) is introduced in Section~\ref{disttree}, leading to distributional
+gradients) is introduced in Section~\ref{sec:disttree}, leading to distributional
 trees. (3)~Ensembles of distributional trees fitted to randomized subsamples are
-employed to establish distributional forests in Section~\ref{distforest}.
+employed to establish distributional forests in Section~\ref{sec:distforest}.
 
@@
 zero-censored Gaussian distribution. The latter is employed in the empirical
-in the empirical case study in Section~\ref{ProbabilisticForecasting} to model
+in the empirical case study in Section~\ref{sec:precipitation} to model
 power-transformed daily precipitation amounts.
@@
 \subsection{Distributional fit}
-\label{distfit}
+\label{sec:distfit}
 
@@
 \subsection{Distributional tree}
-\label{disttree}
+\label{sec:disttree}
 
@@
 \subsection{Distributional forest}
-\label{distforest}
+\label{sec:distforest}
 
@@
 
-\SweaveOpts{eval = TRUE}
-
 \section{Probabilistic precipitation forecasting in complex terrain}
-\label{ProbabilisticForecasting}
+\label{sec:precipitation}
 
@@
 \subsection{Data}
-\label{data}
+\label{sec:data}
 
@@
 \caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
-\label{covariates}
+\label{tab:covariates}
 \end{minipage} 
@@
 or different aggregation time ranges, etc. A detailed overview is
-provided in Table~\ref{covariates}, yielding 80~predictor variables in total.
+provided in Table~\ref{tab:covariates}, yielding 80~predictor variables in total.
 (Since the minimum of the ``variable downwards short wave radiation flux'',
@@
 \subsection{Models and evaluation}
+\label{sec:evaluation}
 
@@
 parameter.}
-\label{model_specification}
+\label{tab:models}
 \end{table}
@@
 The following zero-censored Gaussian regression models are employed
-in the empirical case study, see also Table~\ref{model_specification} for
+in the empirical case study, see also Table~\ref{tab:models} for
 further details:
@@
   adaptive local likelihood based on the forest weights, as described
-  in Section~\ref{Methodology}.
+  in Section~\ref{sec:methods}.
 
@@
 $$
-\text{SkillScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{EMOS}}}.
+\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{EMOS}}}.
 $$
@@
 \subsection{Application for one station}
-\label{Application for one station}
+\label{sec:axams}
 
@@
 for one observation station, Axams in Tyrol (selected somewhat arbitrarily
-as the closest in the data set to Innsbruck, capital of Tyrol). As for all
-other stations, daily observations and  numerical weather predictions are available for the month of July of 
-28~years from 1985 through the year 2012. In Figure~\ref{axams_onestation} 
-the results of fitting a distributional forest to the first 24 years of the 
-data set and predicting the total amount of precipitation for one specific 
-day of each of the four successive years have been shown as a motivational 
-example for a probabilistic forecast. Staying with this fitted model,
-variable importance of the distributional forest model will be analyzed and its
-performance will be compared to those of the prespecified and the boosted GAMLSS 
-and the EMOS model based on residual QQ-plots.
-
-%%%<TH> always us the term "distributional forest"
-As shown in Table~\ref{model_specification} the distributional forest 
-includes all of the available covariates. A single tree model would allow 
-to easily analyze which covariates are used as split variables in the nodes 
-and hence have a significant impact on the response variable. In a forest 
-model though, for each node in each of the trees a subset of the covariates 
-is chosen randomly as possible split variables to select one from. Moreover, 
-each tree is built on a resampled version of the learning data. For these 
-reasons the choices of split variables vary over the trees of a forest making 
-it difficult to analyze the impact of covariates just by investigating the 
-different trees. To get an overview of which covariates have the most influential 
-effect on the response variable a standard measurement of variable importance 
-for forest models has been applied. Based on the mean decrease in accuracy 
-regarding the CRPS the 10 leading covariates of the above described distributional 
-forest for station Axams learned on the first 24 years are listed in 
-Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation (\emph{tp}) is the most influential variable. In particular, five variations of 
-this variable are among the top ten of the list of covariates ordered by 
-variable importance.
-
-\begin{figure}[t!]
-\centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
-par(mar = c(3,10,2,2))
-barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
-        horiz = TRUE, las = 1, axes = FALSE,
-        xlab = "Variable importance: mean decrease in CRPS",
-        font.axis = 3, #list(family="HersheySerif", face=3),
-        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
-axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
-@
-\caption{\label{varimp} CRPS-based variable importance for the top 10 covariates 
-in the distributional forest. Based on data for station Axams, learning period 
-1985--2008 and assessed in 2009--2012.}
-\end{figure}
-
-In Figure~\ref{qqr_oosample} an impression of the quality of predictions made 
-by the tested methods is given by residual QQ-plots for out-of-sample predictions 
-of each of the models. For these plots 100 randomized quantile residuals are 
-simulated per observation. The closer the points are to the diagonal line the 
-better the estimated distributional model fits the data. Overall, it can be 
-stated that all four models provide well calibrated forecasts. This is also 
-supported by the corresponding probability integral transform (PIT) histograms 
-(\citealt{Gneiting+Balabdaoui+Raftery:2007}) in the appendix.
+as the closest in the data set to Innsbruck, capital of Tyrol that three of the
+authors are based in). As for all other stations, daily precipitation observations and
+numerical weather predictions are available for the month of July from 1985
+to 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
+forecasts from the distributional forest, trained on 1985--2008, for July 24
+in 2009--2012 have already been shown as a motivational example. The figure
+depicts the point masses at zero (i.e., the probability of a dry day) and
+the forecasted probability density function for the total amount of precipitation,
+showing that these differ considerably in location, scale, and the amount of
+censoring while conforming quite well with the actual precipitation observations
+from those days. But, of course, this is just one selected day and we are interested
+in assessing how well the distributional fits are calibrated and what the
+relative predictive performance is.
 
@@
 set.seed(7)
-par(mfrow = c(2,2))
-qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
-@
-\caption{\label{qqr_oosample} Out-of-sample residual QQ-plots (2009--2012) 
+par(mfrow = c(2, 2))
+qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_ml_t, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_g_t,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+@
+\caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
 for station Axams based on models learned on data from 1985--2008.}
@@
 
-After analyzing the results of the models on only one data set where 
-visualization is easier, the methods are now going to be evaluated
-in a cross validation framework for the station Axams. 
-The data set over all 28 years is split into 7 subsets each consisting of 
-4 randomly selected years. While 6 of these subsets are used as learning 
-data the 7th is the testing data set for which predictions are made and 
-evaluated. This is done by calculating the CRPS value for each of the 
-observations in the testing data set and then averaging them. Each of the 
-7 parts is selected once as testing data which leads to 7 different settings 
-of which again the average values are stored. This whole procedure is repeated 
-10 times. The resulting skill scores are illustrated in 
-Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model 
-represented by the horizontal line at height 0. This CRPS skill score boxplot 
-shows that, in this setting, both GAMLSS models and the distributional forest 
-perform better than the EMOS model. While the two GAMLSS lead to an improvement 
-of around 4 percent with the boosted version being slightly ahead, the 
-distributional forest even reaches an improvement rate close to 6 percent.
+To assess calibration Figure~\ref{fig:axams-qq} shows residual QQ~plots for
+out-of-sample predictions (2009--2012) from the different models trained on
+1985--2008. Due to the point masses at zero 100 draws from the randomized
+quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
+gray. Overall, the randomized quantile residuals conform quite well with the
+theoretical standard normal quantile (i.e., form a straight line), indicating
+that all four models are sufficiently well calibrated. This is also 
+supported by the corresponding PIT (probability integral transform) histograms 
+in Appendix~\ref{app:pit}.
+
 <<echo=FALSE, results=hide>>=
@@
 @
-\caption{\label{boxplots_crps_axams}CRPS skill score from the 10 times 
+\caption{\label{fig:axams-crps}CRPS skill score from the 10 times 
 7-fold cross validation at station Axams (1985--2012). The horizontal orange 
@@
 
-\SweaveOpts{eval = TRUE}
-%\newpage
+To assess predictive performance, a full cross-validation is carried out
+rather than relying on just the one test set for the years 2009--2012.
+To do so, 10 times 7-fold cross-validation is carried out where each
+splits the available 28~years into 7~subsets of 4~randomly selected years.
+The models learned on 6~folds (= 24~years) and evaluated on the 7-th fold
+(= 4~years) using the average CRPS across all observations.
+The resulting 10~CRPS skill scores are displayed by boxplots in
+Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
+line at a CRPSS of 0). Clearly, both GAMLSS models and the distributional forest 
+perform better than the EMOS model. While the two GAMLSS lead to an improvement 
+of around 4~percent, the distributional forest has a slightly higher improvement
+of around 5.5~percent.
+
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
+par(mar = c(3,10,2,2))
+barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
+        horiz = TRUE, las = 1, axes = FALSE,
+        xlab = "Variable importance: mean decrease in CRPS",
+        font.axis = 3, #list(family="HersheySerif", face=3),
+        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
+axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
+@
+\caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
+in the distributional forest. Based on data for station Axams, learning period 
+1985--2008 and assessed in 2009--2012.}
+\end{figure}
+
+Finally, it is of interest how this improvement in predictive performance
+by the distributional forest is accomplished, i.e., which of the 80~covariates
+are selected in the trees of the forest. As the 100~trees of the forest
+
+\fixme{still not finished}
+
+As shown in Table~\ref{tab:models} the distributional forest 
+includes all of the available covariates. A single tree model would allow 
+to easily analyze which covariates are used as split variables in the nodes 
+and hence have a significant impact on the response variable. In a forest 
+model though, for each node in each of the trees a subset of the covariates 
+is chosen randomly as possible split variables to select one from. Moreover, 
+each tree is built on a resampled version of the learning data. For these 
+reasons the choices of split variables vary over the trees of a forest making 
+it difficult to analyze the impact of covariates just by investigating the 
+different trees.
+
+To get an overview of which covariates have the most influential 
+effect on the response variable a standard measurement of variable importance 
+for forest models has been applied. Based on the mean decrease in accuracy 
+regarding the CRPS the 10 leading covariates of the above described distributional 
+forest for station Axams learned on the first 24 years are listed in 
+Figure~\ref{fig:axams-varimp}. As expected, the numerical forecast for total precipitation
+(\emph{tp}) is the most influential variable. In particular, five variations of 
+this variable are among the top ten of the list of covariates ordered by 
+variable importance.
+
 \subsection{Application for all stations}
+\label{sec:all}
+
 Until now the focus has been on one observation station only. To test and 
@@
 the CRPS skill score with the EMOS model as reference model is calculated 
-for each method at each station. In Figure~\ref{boxplot_crps_all} the boxplots 
+for each method at each station. In Figure~\ref{fig:all-crps} the boxplots 
 of the CRPS skill score values are plotted illustrating that the distributional 
@@
                              proj4string = crs(tirol.dem))
-
-#library("colorspace")
 @
@@
 @
-\caption{\label{boxplot_crps_all}CRPS skill score for each station (grey lines) 
+\caption{\label{fig:all-crps}CRPS skill score for each station (grey lines) 
 and aggregated over all stations (boxplots). Station Axams is highlighted in green 
@@
 
-Figure~\ref{map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation 
+Figure~\ref{fig:all-map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation 
 Data, \cite{srtm}) with the location of all observation stations included in the 
@@
 @
-\caption{\label{map}Map of Tyrol showing which model performed best at 
+\caption{\label{fig:all-map}Map of Tyrol showing which model performed best at 
 which station (type of symbol). The color codes whether the distributional 
@@
 \section{Discussion}
-
-From the results in Section~\ref{ProbabilisticForecasting} we can conclude 
+\label{sec:discussion}
+
+From the results in Section~\ref{sec:precipitation} we can conclude 
 that distributional forests provide a powerful alternative framework requiring 
@@
 
-In Section~\ref{ProbabilisticForecasting} the prespecified GAMLSS were 
+In Section~\ref{sec:precipitation} the prespecified GAMLSS were 
 built using the function \code{gamlss} from the \textsf{R} package 
@@
 
-\section{Tree algorithm} \label{app:tree}
+\section{Tree algorithm}
+\label{app:tree}
+
+\fixme{Shorten/omit parts that are in the main paper already. Focus in
+the description on what is actually used in the application, i.e.,
+reduce the emphasis of the modularity of the framework.}
+
+\fixme{Adapt notation.}
+
 The tree algorithm used in the applications discussed in this paper 
@@
 
-\section{Probability integral transform (PIT) histograms}
-Probability integral transform (PIT) histograms 
-\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used tool to 
-assess the predictive power of distributional forecasting methods, 
-particularly in the field of meteorological forecasting.
-
-Let $x$ be an observation with true distribution function $G$ and 
-predicted distribution function $F$. Then the corresponding probability 
-integral transform (PIT) value is $p = F(x)$. In the ideal case of a 
-perfect prediction and $F$ being continuous $p$ follows a uniform distribution. 
-This uniformity can be analysed applying a histogram. The closer the bars 
-of a PIT histogram are to the value 1, the closer the distribution of $p$ 
-is to a uniform distribution and therefore the better the predicted 
-distribution models the true distribution.
-
-The PIT histograms in Figure~\ref{pit_oosample} correspond to the forecasts 
-presented in Section~\ref{Application for one station}. Daily observations 
-of total precipitation from station Axams are considered and predictions 
-are made for the 24th of July 2009, 2010, 2011 and 2012, based on data 
-from the previous 24 years.
+\section{PIT histograms}
+\label{app:pit}
+
+As an alternative to QQ~plots based on randomized quantile residuals,
+PIT (probability integral transform) histograms 
+\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used to 
+assess how well-calibrated distributional fits are.
+Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
+probabilistic forecasts as Figure~\ref{fig:qq}, confirming that
+despite some deviations the overall calibration of all models is acceptable.
 
@@
 %\end{subfigure}
-\caption{\label{pit_oosample}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
+\caption{\label{fig:pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
 \end{figure}

### 2018-03-21 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2068 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 23275f833564eea728d6714a822b1b4fc2d89988

@@
 forests, we proceed in three steps. (1)~To fix notation, we briefly review fitting
-distributions using standard maximum likelihood in in Section~\ref{sec:distfit}.
+distributions using standard maximum likelihood in Section~\ref{sec:distfit}.
 (2)~A recursive partitioning strategy based on the corresponding scores (or
@@
 zero-censored Gaussian distribution. The latter is employed in the empirical 
-in the empirical case study in Section~\ref{sec:precipitation} to model
-power-transformed daily precipitation amounts.
+case study in Section~\ref{sec:precipitation} to model power-transformed daily 
+precipitation amounts.
 
@@
 while accounting for a point mass at zero precipitation (i.e., dry days without
-rain), zero-censored Gaussian distribution with with mean/location parameter
+rain), zero-censored Gaussian distribution with mean/location parameter
 $\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
@@
   and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
-\item Split the sample along the partitioning variable $Z_{l,i}^*$ with the
+\item Split the sample along the partitioning variable $Z_l^*$ with the
   strongest association or instability. Choose the breakpoint with the highest
@@
 (ensemble model output statistics, \citealp{Gneiting+Raftery+Westveld:2005}).
-In the case of precipiation forecasting, EMOS explains past precipitation
+In the case of precipitation forecasting, EMOS explains past precipitation
 sums by the NWP output ``total precipitation'' (\emph{tp}), using its
@@
 with prespecified effects and interactions based on meteorological
-knowedge/experience, and (c) a boosted GAMLSS with automatic
+knowledge/experience, and (c) a boosted GAMLSS with automatic
 selection of smooth additive terms based on all 80~predictor variables.
@@
 splits the available 28~years into 7~subsets of 4~randomly selected years.
-The models learned on 6~folds (= 24~years) and evaluated on the 7-th fold
+The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
 (= 4~years) using the average CRPS across all observations.
@@
 PIT (probability integral transform) histograms 
-\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used to 
+\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used tool to 
 assess how well-calibrated distributional fits are.
 Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
-probabilistic forecasts as Figure~\ref{fig:qq}, confirming that
+probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that
 despite some deviations the overall calibration of all models is acceptable.

### 2018-03-22 pkg/disttree/inst/draft/draft.Rnw
further improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2069 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: aa76bab9089ba006676822a08f8e104a7a08ab11

@@
 
-transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)
+transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
 
@@
 box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
 lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@@
 box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
 lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@@
 
-lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)
+lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2)
+lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2)
+lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2)
+lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2)
 
@@
 and shape. Furthermore, censoring and/or truncation of these distributions
-can be incorporated in the usual straightforward way \fixme{reference needed}.
+can be incorporated in the usual straightforward way
+\citep[see e.g.,][Chapter~7.2]{Long:1997}.
 
@@
 for one observation station, Axams in Tyrol (selected somewhat arbitrarily
-as the closest in the data set to Innsbruck, capital of Tyrol that three of the
-authors are based in). As for all other stations, daily precipitation observations and
+as the closest in the data set to Innsbruck, capital of Tyrol and base of three of the
+authors). As for all other stations, daily precipitation observations and
 numerical weather predictions are available for the month of July from 1985
@@
 are selected in the trees of the forest. As the 100~trees of the forest
-
-\fixme{still not finished}
-
-As shown in Table~\ref{tab:models} the distributional forest 
-includes all of the available covariates. A single tree model would allow 
-to easily analyze which covariates are used as split variables in the nodes 
-and hence have a significant impact on the response variable. In a forest 
-model though, for each node in each of the trees a subset of the covariates 
-is chosen randomly as possible split variables to select one from. Moreover, 
-each tree is built on a resampled version of the learning data. For these 
-reasons the choices of split variables vary over the trees of a forest making 
-it difficult to analyze the impact of covariates just by investigating the 
-different trees.
-
-To get an overview of which covariates have the most influential 
-effect on the response variable a standard measurement of variable importance 
-for forest models has been applied. Based on the mean decrease in accuracy 
-regarding the CRPS the 10 leading covariates of the above described distributional 
-forest for station Axams learned on the first 24 years are listed in 
-Figure~\ref{fig:axams-varimp}. As expected, the numerical forecast for total precipitation
-(\emph{tp}) is the most influential variable. In particular, five variations of 
-this variable are among the top ten of the list of covariates ordered by 
-variable importance.
+do not allow to simply assess the variables' role graphically, a common
+solution for random forests in general is to consider variable importance
+measures. Here, this is defined as the amount of change in CRPS when
+the association between one covariate and the response variable is
+artificially broken through permutation (and thus also breaking the
+association to the remaining covariates).
+
+Figure~\ref{fig:axams-varimp} shows 10~covariates with the highest permutation
+importance (i.e., change in CRPS) for station Axams. As expected the NWP outputs
+for total precipitation (\emph{tp}) are particularly important along with total
+column-integrated condensate (\emph{tcolc}). Also, both variables occur
+in various transformations such as means (either of the full day or certain parts
+of the afternoon), spreads, or minima/maxima. Thus, while the covariates
+themselves are not surprising, a specification with all the transformations
+would typically not have been considered in a GAMLSS.
+
+\pagebreak
 
@@
 
-Until now the focus has been on one observation station only. To test and 
-compare the four methods under different geographical circumstances they
-are applied on all 95 stations. For each station, the first 
-24 years are used as learning data set and predictions are made for the 
-ensuing 4 years of available observations. For these out-of-sample predictions 
-the CRPS skill score with the EMOS model as reference model is calculated 
-for each method at each station. In Figure~\ref{fig:all-crps} the boxplots 
-of the CRPS skill score values are plotted illustrating that the distributional 
-forest leads to the highest improvement averaged over all stations.
-The green line represents the results for the station Axams.
-%NAs in gamlss in 13 stations!!!\\
-%for example in 13th station in complete{\_}stations\\
-%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
-%\text{NA's in the working vector or weights for parameter mu}
-%\\
-% sum(is.na(crps[,3]))
-%<<eval = TRUE, echo = FALSE>>=
 <<echo = FALSE>>=
@@
 
-\begin{figure}[t!]
-\centering
-<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, width = 7>>=
+\begin{figure}[p!]
+\centering
+<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, height = 4.5, width = 7>>=
+par(mar = c(3.1, 4.1, 1.1, 2.1))
 matplot(t(s[,]), type = "l", lwd = 2, 
@@
 @
-\caption{\label{fig:all-crps}CRPS skill score for each station (grey lines) 
-and aggregated over all stations (boxplots). Station Axams is highlighted in green 
+\caption{\label{fig:all-crps}CRPS skill score for each station (gray lines
+with boxplots superimposed). Station Axams is highlighted in green 
 and the horizontal orange line pertains to the reference model EMOS. The models are 
-learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
-\end{figure}
-
-Figure~\ref{fig:all-map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation 
-Data, \cite{srtm}) with the location of all observation stations included in the 
-data set. Each station is illustrated by a symbol with the type of symbol 
-representing the method that performed best in terms of CRPS at this station. 
-The color of the symbol indicates the difference between the CRPS of the 
-distributional forest and the CRPS of the best performing among the other 
-three models (prespecified GAMLSS, boosted GAMLSS and EMOS).
-Even though for some stations, especially in East Tyrol, one of the two GAMLSS
-models and for a few others the EMOS model leads to the best results, for the 
-majority of stations the distributional forest performs best.
-
-\begin{figure}[t!]
-\centering
+learned on 1985--2008 and validated for 2009--2012.}
+
+\vspace*{0.4cm}
+
 \setkeys{Gin}{width=0.99\textwidth}
@@
 @
-\caption{\label{fig:all-map}Map of Tyrol showing which model performed best at 
-which station (type of symbol). The color codes whether the distributional 
-forest had lower or higher CRPS (red or blue) compared to the best of the 
-other three models (prespecified GAMLSS, boosted GAMLSS and EMOS).}
+\caption{\label{fig:all-map}Map of Tyrol coding the best-performing model
+for each station (type of symbol). The color codes whether the distributional 
+forest had higher (green) or lower (red) CRPS compared to the best of the 
+other three models.}
 \end{figure}
+
+After considering only one observation station up to now, a second step
+evaluates and compares the competing methods on all 95~stations. As in the previous
+section, all models are learned on the first 24~years and evaluated by the
+average CRPS on the last 4~years. More specifically, the CRPS skill score
+against the EMOS model is computed for the out-of-sample predictions at each
+station and visualized by parallel coordinates plots with boxplots superimposed
+in Figure~\ref{fig:all-crps}. Overall distributional forests have a slightly
+higher improvement in CRPSS compared to the two GAMLSS which is best seen
+by looking at the boxplots and also in the green line representating the results
+for station Axams. The underlying parallel coordinates additionally
+bring out that the prespecified GAMLSS sometimes performs rather differently
+(sometimes better, sometimes worse) compared to the two data-driven models.
+Finally, because some CRPSS become negative, it is also clear that the simple
+EMOS performs better than the other models for a few of the stations.
+
+%NAs in gamlss in 13 stations!!!\\
+%for example in 13th station in complete{\_}stations\\
+%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
+%\text{NA's in the working vector or weights for parameter mu}
+%\\
+% sum(is.na(crps[,3]))
+%<<eval = TRUE, echo = FALSE>>=
+
+To assess whether these differences in predictive performance are due to differences
+in the topology, Figure~\ref{fig:all-map} shows a brief summary for each station
+on a map of Tyrol (based on the SRTM 90m Digital Elevation 
+Data, \cite{srtm}). Each station is illustrated by a symbol that conveys which
+model performed best in terms of CRPS on the last 4~years of the data.
+Additionally, the color of the symbol indicates the CRPS difference between 
+distributional forest and the best-performing other model. Green signals that
+the distributional forest performs better than the other models whereas red
+signals that another model performs better. Overall the distributional forest
+performs on par (gray) or better (green) for the majority of stations. Only
+for a few stations in the north-east EMOS performs best, and in East Tyrol
+the prespecified EMOS performs particularly well.
 
@@
 
-From the results in Section~\ref{sec:precipitation} we can conclude 
-that distributional forests provide a powerful alternative framework requiring 
-only little effort to set up a model which is competitive to commonly used 
-methods or even outperforms them. Even though a proper selection of covariates 
-and specification of interactions in advance can lead to better fitting 
-(and probably less complex) models, this presupposes profound knowledge of 
-all influencing factors. Otherwise, important aspects might be missed or 
-effects which do not correspond to the truth might be imposed on the model. 
-This risk can be avoided by applying a boosting algorithm for an automatic 
-variable selection. In this case the trade-off is often a remarkably higher 
-computational effort. To build a distributional forest the user is not 
-required to provide any expert knowledge or additional information about 
-the setting as variable selection and detection of interactions are done 
-automatically by the tree-structure of the model. This feature is 
-particularly advantageous in case of a very high number of covariates. 
-Therefore, the above discussed application illustrates the strengths of 
-distributional forests as they offer a very flexible framework which 
-enables an easy use of the method and at the same time lead to an overall 
-improvement in the predictive performance in this setting. Especially 
-in situations where no or only little information is available about 
-possible influences, effects or interactions this novel methodology 
-can be seen as a good and recommendable compromise being able to deal 
-with a wide range of modeling situations in a reasonable way.
-
-%Even though many situations demand for the use of specifically tailored 
-%models providing the needed features and therefore being clearly advantageous 
-%and the best choice, it is often hard to tell in advance which features 
-%will be required. In these cases a distributional forest is a 
-%recommendable compromise providing reasonable models and performing 
-%well without any prespecification or additional knowledge needed in advance.
+Distributional regression modeling is combined with tree-based modeling
+to obtain a novel and flexible method for probabilistic forecasting.
+The resulting distributional trees and forests can capture abrupt and
+nonlinear effects and interactions in a data-driven way. By basing the
+split point and split variable selection on a full likelihood and
+corresponding score function, the trees and forests can not only pick
+up changes in the location but also the scale or shape of a distributional
+family.
+
+Distributional forests are an attractive alternative when prespecifying or boosting
+all effects and interactions in a GAMLSS model is challenging. They are
+rather straightforward to specify requiring only little prior subject
+matter knowledge and also work well in the presence of many potential
+covariates. The application to precipitation forecasting in complex terrain
+illustrates that distributional forests often perform on par or even
+better than their GAMLSS counterparts. Hence, they form a useful addition
+to the toolbox of probabilistic forecasts, especially in meteorology.
 
 \section*{Computational details}
-The corresponding implementation to the novel methods presented 
-in this article is available on \textsf{R}-Forge in an \textsf{R} 
-package called \textbf{disttree} within the project \textbf{partykit} 
+
+The proposed methods are in the \textsf{R} package \textbf{disttree} on
+\textsf{R}-Forge within the \textbf{partykit} project
 (\url{https://R-Forge.R-project.org/projects/partykit/}). The function
-\code{distfit} fits a distributional model and is called within the
-tree-building function \code{disttree} and the forest-building
-function \code{distforest}. 
-Within \code{disttree} either the \code{mob} or \code{ctree} function 
-can be applied as a framework to build a tree. Similarly, the function 
-\code{distforest} is based on the function \code{cforest}. All three 
-functions, \code{mob}, \code{ctree} and \code{cforest}, are provided 
-in the package \textbf{partykit} (\citealt{Hothorn+Zeileis:2015}). 
-To specify a distribution family the user can hand over one of the 
-following objects as input argument in \code{distfit}, \code{disttree} 
-and \code{distforest}: 
-\begin{itemize}
-\item a GAMLSS family object from the \textsf{R} package \textbf{gamlss.dist} 
-(\citealt{Stasinopoulos+Rigby:2007}).
-\item a family list generating function from the package \textbf{disttree}.
-\item a complete family list from the package \textbf{disttree}.
-\item a custom list containing all required information about the 
-distribution family.
-\end{itemize}
-
-In Section~\ref{sec:precipitation} the prespecified GAMLSS were 
-built using the function \code{gamlss} from the \textsf{R} package 
-\textbf{gamlss} (\citealt{Stasinopoulos+Rigby:2007}). For the 
-boosted GAMLSS the function \code{gamboostLSS} from the \textsf{R} 
-package \textbf{gamboostLSS} (\citealt{Hofner+Mayr+Schmid:2016}) 
-was applied. The EMOS models were built using the function 
-\code{crch} from the \textsf{R} package \textbf{crch} 
-(\citealt{Messner+Mayr+Zeileis:2016})
+\code{distfit} fits distributional models by maximum likelihood, which is
+used as the basis for the tree-building function \code{disttree}, upon which
+the \code{distforest} is built. All functions can either be used with
+GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
+(\citealt{Stasinopoulos+Rigby:2007}) or with custom lists containing
+all required information about the distribution family.
+
+In addition to \textbf{disttree}, Section~\ref{sec:precipitation} employs
+the functions \code{crch} from the \textsf{R} package \textbf{crch} 
+\citep{Messner+Mayr+Zeileis:2016} for the EMOS models,
+\code{gamlss} from the \textsf{R} package \textbf{gamlss}
+\citep{Stasinopoulos+Rigby:2007} for the prespecified GAMLSS, and
+\code{gamboostLSS} from the \textsf{R} package \textbf{gamboostLSS}
+\citep{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
+
+\fixme{replication code?}
 

### 2018-03-25 pkg/disttree/inst/draft/draft.Rnw
edited appendix A
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2078 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b33ee8556344dfc4124d758d78310a39b872f15a

@@
 
-# plot RMSE
-plot_rmse <- function(simres, type = c("exp", "par"), ylim = NULL, legend = TRUE){
-
-  if(type == "exp"){
-    rmse <- NULL
-    colnames <- NULL
-    col <- NULL
-    legendnames <- NULL
-    if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.dt"], simres$resmat[,"av.rmse.exp.obs.dt"])
-      colnames <- c(colnames, "dt.true", "dt.obs")
-      col <- c(col, pal["tree"])
-      legendnames <- c(legendnames, "tree")
-    }
-    if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.df"], simres$resmat[,"av.rmse.exp.obs.df"])
-      colnames <- c(colnames, "df.true", "df.obs")
-      col <- c(col, pal["forest"])
-      legendnames <- c(legendnames, "distributional forest")
-    }
-    if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.g"], simres$resmat[,"av.rmse.exp.obs.g"])
-      colnames <- c(colnames, "g.true", "g.obs")
-      col <- c(col, pal["gamlss"])
-      legendnames <- c(legendnames, "gamlss")
-    }
-    if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.b"], simres$resmat[,"av.rmse.exp.obs.b"])
-      colnames <- c(colnames, "b.true", "b.obs")
-      col <- c(col, pal["bamlss"])
-      legendnames <- c(legendnames, "bamlss")
-    }
-    if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.gb"], simres$resmat[,"av.rmse.exp.obs.gb"])
-      colnames <- c(colnames, "gb.true", "gb.obs")
-      col <- c(col, pal["gamboostLSS"])
-      legendnames <- c(legendnames, "gamboostLSS")
-    }
-    if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.rf"], simres$resmat[,"av.rmse.exp.obs.rf"])
-      colnames <- c(colnames, "rf.true", "rf.obs")
-      col <- c(col, pal["randomForest"])
-      legendnames <- c(legendnames, "randomForest")
-    }
-    if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.cf"], simres$resmat[,"av.rmse.exp.obs.cf"])
-      colnames <- c(colnames, "cf.true","cf.obs")
-      col <- c(col, pal["cforest"])
-      legendnames <- c(legendnames, "cforest")
-    }
-
-    colnames(rmse) <- colnames
-    if(is.null(ylim)) ylim <- c(min(na.omit(rmse)), max(na.omit(rmse)))
-
-    plot(x = simres$x.axis, y = rmse[,1], type = "l", col = col[1], ylim = ylim,
-         xlab = expression(kappa), ylab = "RMSE")
-    lines(x = simres$x.axis, y = rmse[,2], type = "l", lty = 2, col = col[1])
-
-    for(i in 1:(length(col)-1)){
-      lines(x = simres$x.axis, y = rmse[,(2*i+1)], type = "l", col = col[i+1])
-      lines(x = simres$x.axis, y = rmse[,(2*i+2)], type = "l", lty = 2, col = col[i+1])
-    }
-    if(legend) legend('topleft', legendnames, 
-                      col = col, lty = 1, cex = 0.7)
-  }
-
-  if(type == "par"){
-
-    rmse <- NULL
-    colnames <- NULL
-    col <- NULL
-    legendnames <- NULL
-    if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.dt"], simres$resmat[,"av.rmse.sigma.dt"])
-      colnames <- c(colnames, "dt.true", "dt.obs")
-      col <- c(col, pal["tree"])
-      legendnames <- c(legendnames, "tree")
-    }
-    if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.df"], simres$resmat[,"av.rmse.sigma.df"])
-      colnames <- c(colnames, "df.true", "df.obs")
-      col <- c(col, pal["forest"])
-      legendnames <- c(legendnames, "distributional forest")
-    }
-    if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.g"], simres$resmat[,"av.rmse.sigma.g"])
-      colnames <- c(colnames, "g.true", "g.obs")
-      col <- c(col, pal["gamlss"])
-      legendnames <- c(legendnames, "gamlss")
-    }
-    if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.b"], simres$resmat[,"av.rmse.sigma.b"])
-      colnames <- c(colnames, "b.true", "b.obs")
-      col <- c(col, pal["bamlss"])
-      legendnames <- c(legendnames, "bamlss")
-    }
-    if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.gb"], simres$resmat[,"av.rmse.sigma.gb"])
-      colnames <- c(colnames, "gb.true", "gb.obs")
-      col <- c(col, pal["gamboostLSS"])
-      legendnames <- c(legendnames, "gamboostLSS")
-    }
-    if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.rf"], simres$resmat[,"av.rmse.sigma.rf"])
-      colnames <- c(colnames, "rf.true", "rf.obs")
-      col <- c(col, pal["randomForest"])
-      legendnames <- c(legendnames, "randomForest")
-    }
-    if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
-      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.cf"], simres$resmat[,"av.rmse.sigma.cf"])
-      colnames <- c(colnames, "cf.true","cf.obs")
-      col <- c(col, pal["cforest"])
-      legendnames <- c(legendnames, "cforest")
-    }
-
-    colnames(rmse) <- colnames
-    if(is.null(ylim)) ylim <- c(min(na.omit(rmse)), max(na.omit(rmse)))
-
-    plot(x = simres$x.axis, y = rmse[,1], type = "l", col = col[1], ylim = ylim,
-         xlab = expression(kappa), ylab = "RMSE")
-    lines(x = simres$x.axis, y = rmse[,2], type = "l", lty = 2, col = col[1])
-
-    for(i in 1:(length(col)-1)){
-      lines(x = simres$x.axis, y = rmse[,(2*i+1)], type = "l", col = col[i+1])
-      lines(x = simres$x.axis, y = rmse[,(2*i+2)], type = "l", lty = 2, col = col[i+1])
-    }
-    if(legend) legend('topleft', legendnames, 
-                      col = col, lty = 1, cex = 0.7)
-  }
-}
-
-# plot loglikelihood
-plot_ll <- function(simres, ylim = NULL, legend = TRUE){
-
-  ll <- NULL
-  colnames <- NULL
-  col <- NULL
-  legendnames <- NULL
-  if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
-    ll <- cbind(ll, simres$resmat[,"av.loglik.dt"])
-    colnames <- c(colnames, "dt.ll")
-    col <- c(col, pal["tree"])
-    legendnames <- c(legendnames, "tree")
-  }
-  if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
-    ll <- cbind(ll, simres$resmat[,"av.loglik.df"])
-    colnames <- c(colnames, "df.ll")
-    col <- c(col, pal["forest"])
-    legendnames <- c(legendnames, "distributional forest")
-  }
-  if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
-    ll <- cbind(ll, simres$resmat[,"av.loglik.g"])
-    colnames <- c(colnames, "g.true")
-    col <- c(col, pal["gamlss"])
-    legendnames <- c(legendnames, "gamlss")
-  }
-  if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
-    ll <- cbind(ll, simres$resmat[,"av.loglik.b"])
-    colnames <- c(colnames, "b.true")
-    col <- c(col, pal["bamlss"])
-    legendnames <- c(legendnames, "bamlss")
-  }
-  if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
-    ll <- cbind(ll, simres$resmat[,"av.loglik.gb"])
-    colnames <- c(colnames, "gb.true")
-    col <- c(col, pal["gamboostLSS"])
-    legendnames <- c(legendnames, "gamboostLSS")
-  }
-  if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
-    ll <- cbind(ll, simres$resmat[,"av.loglik.rf"])
-    colnames <- c(colnames, "rf.true")
-    col <- c(col, pal["randomForest"])
-    legendnames <- c(legendnames, "randomForest")
-  }
-  if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
-    ll <- cbind(ll, simres$resmat[,"av.loglik.cf"])
-    colnames <- c(colnames, "cf.true")
-    col <- c(col, pal["cforest"])
-    legendnames <- c(legendnames, "cforest")
-  }
-
-  colnames(ll) <- colnames
-  if(is.null(ylim)) ylim <- c(min(na.omit(ll)), max(na.omit(ll)))
-
-  plot(x = simres$x.axis, y = ll[,1], type = "l", col = col[1], ylim = ylim,
-       xlab = expression(kappa), ylab = "log-likelihood")
-
-  for(i in 2:length(col)){
-    lines(x = simres$x.axis, y = ll[,i], type = "l", col = col[i])
-  }
-  if(legend) legend('topleft', legendnames, 
-                    col = col, lty = 1, cex = 0.7)
-}
-
-# plot crpse
-plot_crps <- function(simres, ylim = NULL, legend = TRUE){
-
-  crps <- NULL
-  colnames <- NULL
-  col <- NULL
-  legendnames <- NULL
-  if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
-    crps <- cbind(crps, simres$resmat[,"av.crps.dt"])
-    colnames <- c(colnames, "dt.crps")
-    col <- c(col, pal["tree"])
-    legendnames <- c(legendnames, "tree")
-  }
-  if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
-    crps <- cbind(crps, simres$resmat[,"av.crps.df"])
-    colnames <- c(colnames, "df.crps")
-    col <- c(col, pal["forest"])
-    legendnames <- c(legendnames, "distributional forest")
-  }
-  if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
-    crps <- cbind(crps, simres$resmat[,"av.crps.g"])
-    colnames <- c(colnames, "g.true")
-    col <- c(col, pal["gamlss"])
-    legendnames <- c(legendnames, "gamlss")
-  }
-  if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
-    crps <- cbind(crps, simres$resmat[,"av.crps.b"])
-    colnames <- c(colnames, "b.true")
-    col <- c(col, pal["bamlss"])
-    legendnames <- c(legendnames, "bamlss")
-  }
-  if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
-    crps <- cbind(crps, simres$resmat[,"av.crps.gb"])
-    colnames <- c(colnames, "gb.true")
-    col <- c(col, pal["gamboostLSS"])
-    legendnames <- c(legendnames, "gamboostLSS")
-  }
-  if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
-    crps <- cbind(crps, simres$resmat[,"av.crps.rf"])
-    colnames <- c(colnames, "rf.true")
-    col <- c(col, pal["randomForest"])
-    legendnames <- c(legendnames, "randomForest")
-  }
-  if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
-    crps <- cbind(crps, simres$resmat[,"av.crps.cf"])
-    colnames <- c(colnames, "cf.true")
-    col <- c(col, pal["cforest"])
-    legendnames <- c(legendnames, "cforest")
-  }
-
-  colnames(crps) <- colnames
-  if(is.null(ylim)) ylim <- c(min(na.omit(crps)), max(na.omit(crps)))
-
-  plot(x = simres$x.axis, y = crps[,1], type = "l", col = col[1], ylim = ylim,
-       xlab = expression(kappa), ylab = "CRPS")
-
-  for(i in 2:length(col)){
-    lines(x = simres$x.axis, y = crps[,i], type = "l", col = col[i])
-  }
-  if(legend) legend('topright', legendnames, 
-                    col = col, lty = 1, cex = 1, bty = "n")
-}
 @
@@
 higher improvement in CRPSS compared to the two GAMLSS which is best seen
-by looking at the boxplots and also in the green line representating the results
+by looking at the boxplots and also in the green line representing the results
 for station Axams. The underlying parallel coordinates additionally
@@
 
-The tree algorithm used in the applications discussed in this paper 
-is going to be explained following the ideas presented in 
-\cite{Seibold+Zeileis+Hothorn:2017}. For notational simplicity, the testing 
-and splitting procedure is described for the root node with observations 
-$\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. In each child node the 
-corresponding subset of observations depends on the assignment made 
-by the foregoing split.
-
-\begin{itemize}
-\item A distributional model $\mathcal{D}(Y, \bm{\theta})$ is fit to the set of 
-observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution 
-parameter $\bm{\hat \theta}$. This parameter can be a multidimensional 
-vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, 
-depending on the type of distribution. The estimation is done by 
-maximizing the log-likelihood function $\ell(\bm{\theta}; Y)$, i.e.
-$$
-\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
-$$
-which is equal to solving
-$$
-\sum_{i=1}^n s(\theta, y_i) = \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\theta; y_i) = 0
-$$
-for $\theta$.
-
-\item Calculate score values 
-$$
-s(\hat{\theta}, y_i) = 
+In the following, the tree algorithm applied in the empirical case study 
+discussed in this paper is going to be explained. For notational simplicity, 
+the testing and splitting procedure is described for the root node with 
+observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
+In each child node the corresponding subset of observations depends on 
+the assignment made by the foregoing split.
+
+After fitting a distributional model $\mathcal{D}(Y, \bm{\theta})$ to 
+the set of observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
+Section~\ref{sec:distfit} the resulting estimated parameter 
+$\bm{\hat{\theta}} = 
+(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
+can be plugged in the score function $s(\bm{\theta}, Y)$.
+In that way a meassurement of goodness-of-fit is obtained for each
+parameter $\theta_j$ and each observation $y_i$.\\
+To use this information, statistical tests are employed to detect
+dependencies between the score values
+$$
+s(\bm{\hat{\theta}}, y) = 
+%\begin{pmatrix} 
+%s(\bm{\hat{\theta}}, y_1) \\
+%s(\bm{\hat{\theta}}, y_2) \\
+%\vdots \\
+%s(\bm{\hat{\theta}}, y_n)
+%\end{pmatrix} =
 \begin{pmatrix} 
-s(\hat{\theta}, y_1) \\
-s(\hat{\theta}, y_2) \\
-\vdots \\
-s(\hat{\theta}, y_n)
-\end{pmatrix} =
-\begin{pmatrix} 
-s(\hat{\theta}, y_1)_1 & s(\hat{\theta}, y_1)_2 & \ldots & s(\hat{\theta}, y_1)_K\\
-s(\hat{\theta}, y_2)_1 & s(\hat{\theta}, y_2)_2 & \ldots & s(\hat{\theta}, y_2)_K\\
+s(\bm{\hat{\theta}}, y_1)_1 & s(\bm{\hat{\theta}}, y_1)_2 & \ldots & s(\bm{\hat{\theta}}, y_1)_k\\
+s(\bm{\hat{\theta}}, y_2)_1 & s(\bm{\hat{\theta}}, y_2)_2 & \ldots & s(\bm{\hat{\theta}}, y_2)_k\\
 \vdots & \vdots & \ddots & \vdots \\
-s(\hat{\theta}, y_n)_1 & s(\hat{\theta}, y_n)_2 & \ldots & s(\hat{\theta}, y_n)_K
+s(\bm{\hat{\theta}}, y_n)_1 & s(\bm{\hat{\theta}}, y_n)_2 & \ldots & s(\bm{\hat{\theta}}, y_n)_k
 \end{pmatrix}
 $$
-
-\item Test for independence between each column of scores $\left(s(\hat{\theta}, y_i)_k\right)_{i=1,\ldots,n}$ for $k\in\{1,\ldots,K\}$, and split variable $Z_j \in \{Z_1, \ldots, Z_m\}$.  
+and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
 \begin{align*}
-H_0^{1,j}:  \left(s(\hat{\theta}, y_i)_1\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j \\
-H_0^{2,j}:  \left(s(\hat{\theta}, y_i)_2\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j \\
-\vdots \hspace{3cm}\\
-H_0^{K,j}:  \left(s(\hat{\theta}, y_i)_K\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j 
+H_0^l:  s(\bm{\hat{\theta}}, y) \qquad \bot \qquad Z_l 
 \end{align*}
 
-To test these hypotheses permutation testing with the multivariate linear statistic
-$$
-T_{j} = \sum_{i=1}^n g_j(z_{ji}) \cdot s(\hat{\theta}, y_i)
-$$
-is applied. The type of the transformation function $g_j$ depends on 
-the type of the split variable $Z_j$. If $Z_j$ is numeric then $g_j$ 
-is simply the identity function $g_j(z_{ji}) = z_{ji}$. If $Z_j$ is 
-a categorical variable with $L$ categories then 
-$g_j(z_{ji}) = e_L(z_{ji}) = (\I(z_{ji} = 1), \ldots, \I(z_{ji} = L))$ 
-such that $g_j$ is a unit vector where the element corresponding to 
-the value of $z_{ji}$ is $1$. Observations with missing values are 
+These hypotheses are assessed by applying permutation testing with 
+the multivariate linear statistic
+$$
+T_{l} = \sum_{i=1}^n v_l(z_{li}) \cdot s(\bm{\hat{\theta}}, y_i).
+$$
+The type of the transformation function $v_l$ depends on 
+the type of the split variable $Z_l$. If $Z_l$ is numeric then $v_l$ 
+is simply the identity function $v_l(z_{li}) = z_{li}$. If $Z_l$ is 
+a categorical variable with $H$ categories then 
+$v_l(z_{li}) = e_H(z_{li}) = (\I(z_{ji} = 1), \ldots, \I(z_{li} = H))$ 
+such that $v_l$ is a unit vector where the element corresponding to 
+the value of $z_{li}$ is $1$. Observations with missing values are 
 excluded from the sums.
 
-With the conditional expectation $\mu_{j}$ and the covariance 
-$\Sigma_{j}$ as derived by \cite{Strasser+Weber:1999} the test 
-statistic can be standardized. For a numeric split variable this 
-results in the Pearson correlation coefficient
-$$
-c(t_{j},\mu_{j},\Sigma_{j}) = \left|\frac{t_{j} - \mu_{j}}{\sqrt{\Sigma_{j}}}\right|.
-$$
-Otherwise the standardized form is
-$$
-c(t_{j},\mu_{jk},\Sigma_{j}) = \max_{l=1,\ldots,L}\left|\frac{(t_{j} - \mu_{j})_l}{\sqrt{(\Sigma_{j})_{ll}}}\right|.
-$$
+With the conditional expectation $\mu_{l}$ and the covariance 
+$\Sigma_{l}$ as derived by \cite{Strasser+Weber:1999} the test 
+statistic can be standardized. The observed multivariate linear 
+statistic $t$ is mapped into the real line by a  univariate test 
+statistic $c$. In the application of this paper a quadratic form 
+is chosen, such that
+$$
+c_{quad}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
+$$
+where $\Sigma^+$ is the Moore-Penrose inverse of $\Sigma$. Alternatively,
+the maximum of the absolute values of the standardized linear statistic
+can be considered.
+
 The smaller the p-value corresponding to the standardized test 
-statistic $c(t_{j},\mu_{j},\Sigma_{j})$ is the stronger the discrepancy 
+statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
 from the assumption of independence between the scores and the split 
-variable $Z_j$.
-
-\item After Bonferroni adjusting the p-values check whether any of 
-the resulting p-values is beneath the selected significance level. 
-If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest 
-p-value to the scores corresponding to the distribution parameters 
-relevant for the split.
-
-\item Choose the breakpoint that leads to the highest discrepancy 
-between score functions in the two resulting subgroups. This 
-difference can be measured by the linear statistic
-$$
-T_{j^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\hat{\theta}, y_i)
+variable $Z_l$.
+
+After Bonferroni adjusting the p-values it has to be assessed whether
+any of the resulting p-values is beneath the selected significance level. 
+If so, the partitioning variable $Z_{l^\ast}$ with the lowest p-value 
+is chosen as splitting variable.
+
+As split point the breakpoint that leads to the highest discrepancy 
+between score functions in the two resulting subgroups is selected. 
+This difference can be measured by the linear statistic
+$$
+T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
 $$
 where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
-are defined by splitting in split point $r$ of variable $Z_{j^{\ast}}$. 
+are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
 The split point is then chosen as follows:
 $$
-r^{\ast} = \argmin_{r} c(t_{j^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{j^{\ast}}^r).
-$$
-
-\item Repeat the testing and splitting procedure in each of the resulting 
+r^{\ast} = \argmin_{r} c(t_{l^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{l^{\ast}}^r).
+$$
+
+Repeat the testing and splitting procedure in each of the resulting 
 subgroups until some stopping criterion is reached. This criterion 
 can for example be a minimal number of observations in a node or a 
-minimal p-value for the statistical tests.
-
-\end{itemize}
+minimal p-value for the statistical tests. In that way pre-pruning
+is applied in order to find right-sized trees and hence 
+avoid overfitting.
 
 This permutation test based tree algorithm is presented in 
-\cite{Hothorn+Hornik+Zeileis:2006} as the ctree algorithm. 
+\cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
 A different framework to build a tree is provided by the MOB algorithm 

### 2018-03-26 pkg/disttree/inst/MIP-Seminar/slides.Rnw
small changes in format of map
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2093 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b6f41f638775db4196792b5e77c09db65abfb7a7

@@
 \usepackage{calc}
+\usepackage{graphicx}
 \usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
@@
 \quad (precipitation, temperature, air pressure, sunshine, \dots)\\
-\quad $\rightarrow$ 82 covariates.}
+\quad $\rightarrow$ 80 covariates.}
 \end{itemize}
@@
   layout(cbind(1, 2), width = c(9, 1))
-par(mar = c(5,4,4,0.01))
+  par(mar = c(5,4,4,0.1))
   raster::image(tirol.dem, col = rev(gray.colors(100)),
-              main="Observation stations", ylab = "Latitude", xlab = "Longitude", 
+                main="Stations in Tyrol", cex.main=1.7,
+                ylab = "Latitude", xlab = "Longitude", 
                 xlim = c(9.8,13.2), 
-              ylim = c(46.6, 47.85), 
-              cex.lab=1.3, cex.axis=1.2, cex.main=1.7)
+                ylim = c(46.6, 47.87))
   plot(tirol.gadm, add = TRUE)
-points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.7)
-legend("topleft", pch=c(21,24,25,22), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1.2, bty = "n")
-
-par(mar = c(0.4,0.2,0.4,2.33))
+  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
+  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
+  text(x = 10.25, y = 47.82, labels = "Models with lowest CRPS")
+  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
+  par(mar = c(0.5,0.2,0.5,2.3))
   ## legend
@@
   rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
-axis(4, at = brk, las = 1, mgp=c(0,0,-1.2), cex = 1.5)
+  axis(4, at = brk, las = 1, mgp=c(0,-0.4,-1.2))
 

### 2018-03-27 pkg/disttree/inst/draft/draft.Rnw
added reference for CRPS, revised Appendix A and computational details, and added some further touch ups (especially in the application section)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2096 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4d1f1b869d374a943bb328ea03d5b3b4f55dfa64

@@
 in the weather forecasting literature has been shifting from classical linear 
-models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
+deterministic models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
 non-homogeneous Gaussian regression (NGR) of \cite{Gneiting+Raftery+Westveld:2005}.
@@
 numerical weather predictions (NWPs). Similarly, the variance of the response
-is captured through the variance of the ensemble of NWP outputs. Thus, the NGR
-considers both the mean as well as the uncertainty of the NWP model outputs
+is captured through the variance of the ensemble of NWPs. Thus, the NGR
+considers both the mean as well as the uncertainty of the ensemble predictions
 to obtain probabilistic forecasts calibrated to a particular site.
@@
 \vspace{0.0cm}
-Regression Tree 
+Regression tree 
 \endminipage};
@@
 \vspace{0.0cm}
-Random Forest 
+Random forest 
 \endminipage};
@@
 \vspace{0.0cm}
-Distributional Forest 
+Distributional forest 
 \endminipage};
@@
 \vspace{-0.2cm}%\\
-Distributional Tree
+Distributional tree
 \vspace{0.2cm}
@@
 variables and possibly complex interactions. Thus, these properties make the method particularly
-appealing in case of many covariates with unknown effects where it would be challenging
-to specify a distributional regression model like GAMLSS.
+appealing in case of many covariates with unknown effects and interactions where it would be 
+challenging to specify a distributional regression model like GAMLSS.
 
@@
 \centering
-\setkeys{Gin}{width=0.8\textwidth}
+\setkeys{Gin}{width=0.75\textwidth}
 <<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4>>=
@@
 
-In weather forecasting, these properties are appealing in mountainous regions
-and complex terrain where small-scale effects are not well-resolved in the NWP
-outputs. As these are computed at a coarser scale, there may be abrupt local
-effects within a single NWP grid cell. To illustrate this in practice,
+In weather forecasting, these properties are especially appealing in 
+mountainous regions and complex terrain where a wide range of local-scale
+effects are not yet resolved by the NWP models. 
+Thus, effects with abrupt changes and possibly non-linear interactions 
+might be required to account for site-specific unresolved features.
+To illustrate this in practice,
 precipitation forecasts are obtained with distributional forests at 95
@@
 Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
-More specifically, a zero-censored Gaussian
+More specifically, a zero left-censored Gaussian
 distribution is employed to model 24-hour total precipitation so that the
-zero-censored point mass describes the probability of no precipitation on a
-given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
-established based on data from the same month in the years 1985--2012 along with
-80~covariates derived from a wide range of NWP ensemble outputs. As
+zero-censored point mass describes the probability of observing no precipitation 
+on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
+established based on data from the same month over the years 1985--2012 including
+80~covariates derived from a wide range of different NWP quantities. As
 Figure~\ref{fig:axams-dist} shows, the station-wise models yield a full
-distributional forecast for each day -- here for the same day (July~24) at one
+distributional forecast for each day -- here for one specific day (July~24) at one
 station (Axams) over four years (2009--2012) -- based on the previous 24~years
@@
 \citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
-\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
+\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection among all
 80~available regressors.
@@
 To capture both location and scale of the probabilistic precipitation forecasts
-while accounting for a point mass at zero precipitation (i.e., dry days without
+while accounting for a point mass at zero (i.e., dry days without
 rain), zero-censored Gaussian distribution with mean/location parameter
@@
 In weather forecasting, these covariates typically include the output
-from numerical weather prediction systems and/or lagged weather observations.
+from numerical weather prediction systems and/or lagged observations.
 
@@
 (i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
-response $Y$ in that segment.
+response $Y$ in each segment.
 
@@
 
-In summary, distributional trees are fitted recursively via:
+To sum up, distributional trees are fitted recursively via:
 \begin{enumerate}
@@
 
-Many weather forecasting models leverage the strengths of modern
-numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}
-and produce high-quality forecasts adapted to specific sites/locations
-through distributional regression models like EMOS
-(ensemble model output statistics, \citealp{Gneiting+Raftery+Westveld:2005}).
-In the case of precipitation forecasting, EMOS explains past precipitation
-sums by the NWP output ``total precipitation'' (\emph{tp}), using its
-ensemble mean as the predictor variable for the location parameter and
-the ensemble standard deviation for the scale parameter.
+Many statistical weather forecasting models leverage the strengths of modern
+numerical weather prediction (NWP) systems
+\citep[see][]{Bauer+Thorpe+Brunet:2015}.  
+One frequently used method based on
+distributional regression models is the ensemble model output statistics 
+(EMOS) approach first proposed by
+\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
+specific quantities and sites/locations.  
+In case of precipitation forecasting EMOS typically uses the ensemble mean of 
+``total precipitation'' (\emph{tp}) forecasts as predictor for the location parameter and
+the corresponding ensemble standard deviation for the scale part of the statistical model
+to correct for possible errors of the ensemble in both, the expectation but also
+the uncertainty of a specific forecast.
 
@@
 in the following application case study -- the NWP grid cells of $50 \times 50$
-km$^2$ are too coarse to capture mountains, valleys, etc. Therefore,
-it is often possible to substantially improve the EMOS by including
-further predictor variables, either from local observations or from
-other NWP outputs. Unfortunately, it is typically unclear which
-variables are relevant for improving the predictions. Simply including
+km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
+Therefore, it is often possible to substantially improve the predictive 
+performance of the EMOS by including additional predictor variables, either 
+from local observations or an NWP model. Unfortunately, it is typically unknown 
+which variables are relevant for improving the predictions. Simply including
 all available variables may be computationally burdensome and
@@
 may result in a loss of valuable information. Therefore, selecting
-the relevant variables and interactions is crucial for improving the
-forecasting model.
+the relevant variables and interactions among all possible covariates is 
+crucial for improving the statistical forecasting model.
 
@@
 \hline
-\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
+\emph{tp}: total precipitation,           & 12  & ensemble mean of sums over 24h, \\
 \hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
 \emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
-\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
-\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
+\hspace*{0.9cm} potential energy, &    & ensemble maximum of sums over 24h\\
+\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 \\
 \hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
-\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
+\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
 \hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
-\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
+\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
 \hline
 \emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
-\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
+\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimum values$^*$,\\
 \emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
-\emph{pwat}: preciptable water          &   & ensemble std. deviation of mean values,\\
-\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimal values,\\
+\emph{pwat}: precipitable water         &   & ensemble std. deviation of mean values,\\
+\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimum values$^*$,\\
 \hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
-\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
+\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 \\
 \hspace*{0.95cm} condensate               &   & \\
@@
 \emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
-\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
+\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 \\
 \emph{tdiff700850}: temperature         &   & \\
@@
 \emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
-\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
+\hspace{1.6cm} difference            &   & \qquad over 6--30 \\
 \hline
 \end{tabular}
-\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
+\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
+Time periods indicate aggregation time periods in hours after NWP model
+initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
+0600\,UTC to 0600\,UTC of the next day). $^*$Minimum values of \emph{dswrf}
+over 24\,h are always zero and thus neglected.}
 \label{tab:covariates}
@@
 Both, observations and forecasts are available for 1985--2012 and
-the analysis is exemplified using July as the month with the most precipitation
+the analysis is exemplified using July, the month with the most precipitation
 in Tyrol.
@@
 Observations are obtained for 95~stations all over Tyrol and surroundings,
-providing 24~hour precipitation sums measured at 6~UTC and rigorously
+providing 24~hour precipitation sums measured at 0600\,UTC and rigorously
 quality-checked by the National Hydrographical Service. NWP outputs
@@
 This data set consists of an 11-member ensemble based on a fixed version of the 
-numerical model and a horizontal grid spacing of about $50 \times 50$ 
-km$^2$ initialized daily at 0~UTC from December 1984 to present 
-providing forecasts on 26 hourly temporal resolution. Each of the 
+numerical model and a horizontal grid spacing of about {$50 \times 50$ km$^2$}
+initialized daily at 0000~UTC from December 1984 to present 
+providing forecasts on a 6\ hourly temporal resolution. Each of the 
 11 ensemble members uses slightly different perturbed initial 
-conditions trying to predict the situation specific uncertainty 
+conditions to predict the situation specific uncertainty 
 of the atmospheric state.
 
-From the GEFS outputs 14 basic forecast variables are considered with up 
-to 12 variations of each of these variables such as mean/maximum/minimum
-or different aggregation time ranges, etc. A detailed overview is
+From the GEFS, 14 basic forecast variables are considered with up 
+to 12 variations each such as mean/maximum/minimum
+over different aggregation time periods. A detailed overview is
 provided in Table~\ref{tab:covariates}, yielding 80~predictor variables in total.
-(Since the minimum of the ``variable downwards short wave radiation flux'',
-\emph{dswrf}, over 24 hours is always 0, the ensemble mean and standard
-deviation of these minimal values are not considered.)
 
@@
 covariate dependency and included covariates for each distribution 
-parameter.}
+parameter. \emph{A}$\ast$\emph{B} indicates an interaction between
+covariate \emph{A} and \emph{B}.}
 \label{tab:models}
@@
 The following zero-censored Gaussian regression models are employed
-in the empirical case study, see also Table~\ref{tab:models} for
+in the empirical case study, see Table~\ref{tab:models} for
 further details:
@@
   in each split of the tree (``mtry''). Parameters are estimated by
-  adaptive local likelihood based on the forest weights, as described
+  adaptive local likelihood based on the forest weights as described
   in Section~\ref{sec:methods}.
 
-\item \emph{EMOS:} Standard ensemble model output statistics are used
-  with the total precipitation ensemble mean as regressor in the location
-  submodel and the ensemble standard deviation in the scale submodel.
+\item \emph{EMOS:} Standard ensemble model output statistics
+  models use the ensemble mean of total precipitation as regressor in the location
+  submodel and the corresponding ensemble standard deviation in the scale submodel.
   The parameters are estimated by maximum likelihood, using an identity
-  link for the location and a log link for the scale
+  link for the location part and a log link for the scale part
   \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.
@@
 \item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
-  the most relevant predictors, based on meteorological expert knowledge
+  the most relevant predictors based on meteorological expert knowledge
   following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
@@
   submodel and 6 in the scale submodel. Both involve an interaction of
-  \emph{tp} and \emph{cape} in the afternoon (between 12 and 18 UTC)
+  \emph{tp} and \emph{cape} in the afternoon (between 1200\,UTC and 1800\,UTC)
   to capture the potential for thunderstorms that frequently occur in
@@
   by maximizing the log-likelihood only for the variable yielding the
-  highest improvement. The iteration stops early -- before fully maximizing
+  biggest improvement. The iteration stops early -- before fully maximizing
   the in-sample likelihood -- based on a (computationally intensive)
@@
 The predictive performance in terms of full probabilistic forecasts is
-assessed using the continuous ranked probability score (CRPS, \fixme{reference}).
+assessed using the continuous ranked probability score (CRPS, \citealt{Hersbach:2000}).
 For each of the models this assesses the discrepancy of the
@@
 $$
-where \(\mathbf{1}\) is the indicator function. In the subsequent
+where \(\mathbf{1}(\cdot)\) is the indicator function. In the subsequent
 applications, the mean CRPS is always evaluated out of sample, 
 either using cross-validation or a hold-out data set (2009--2012)
-that was not used for learning (1985--2008). CRPS is a proper
-scoring rule \citep{Gneiting+Raftery:2007} and lower values indicate a better fit.
+that was not used for learning (1985--2008). CRPS is a proper scoring rule
+\citep{Gneiting+Raftery:2007} often used within the meteorological community.
+Lower values indicate better performance.
 
@@
 models over the basic EMOS, a CRPS-based skill score with EMOS as
-the reference method is computed as well:
+the reference method is computed:
 $$
@@
 In a first step, we show a detailed comparison of the competing models
-for one observation station, Axams in Tyrol (selected somewhat arbitrarily
-as the closest in the data set to Innsbruck, capital of Tyrol and base of three of the
+for one observation site, Axams in Tyrol (selected somewhat arbitrarily;
+the station in the data set closest to Innsbruck, capital of Tyrol and base of three of the
 authors). As for all other stations, daily precipitation observations and
 numerical weather predictions are available for the month of July from 1985
-to 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
+through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
 forecasts from the distributional forest, trained on 1985--2008, for July 24
 in 2009--2012 have already been shown as a motivational example. The figure
+shows that the model properly
 depicts the point masses at zero (i.e., the probability of a dry day) and
-the forecasted probability density function for the total amount of precipitation,
-showing that these differ considerably in location, scale, and the amount of
-censoring while conforming quite well with the actual precipitation observations
-from those days. But, of course, this is just one selected day and we are interested
-in assessing how well the distributional fits are calibrated and what the
-relative predictive performance is.
+the forecasted probability density function for the total amount of precipitation.
+The three sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
+censoring while conforming quite well with the actual observations
+from these days. While this is a nice illustrative example we are  
+interested in the overall predictive performance and calibration of the
+distributional fits.
 
@@
 gray. Overall, the randomized quantile residuals conform quite well with the
-theoretical standard normal quantile (i.e., form a straight line), indicating
-that all four models are sufficiently well calibrated. This is also 
-supported by the corresponding PIT (probability integral transform) histograms 
-in Appendix~\ref{app:pit}.
+theoretical standard normal quantile (i.e., form a straight line close to the 
+diagonal), indicating that all four models are sufficiently well calibrated. 
+This is also supported by the corresponding probability integral transform
+(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) histograms in
+Appendix~\ref{app:pit}.
 
@@
 
-To assess predictive performance, a full cross-validation is carried out
-rather than relying on just the one test set for the years 2009--2012.
-To do so, 10 times 7-fold cross-validation is carried out where each
+To assess the predictive performance, a full cross-validation is carried out
+rather than relying on just the one fixed test set for the years 2009--2012.
+To do so, a 10 times 7-fold cross-validation is carried out where each
 splits the available 28~years into 7~subsets of 4~randomly selected years.
@@
 Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
-line at a CRPSS of 0). Clearly, both GAMLSS models and the distributional forest 
-perform better than the EMOS model. While the two GAMLSS lead to an improvement 
+line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
+perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
 of around 4~percent, the distributional forest has a slightly higher improvement
-of around 5.5~percent.
+of around 5.5~percent in median.
 
@@
 
-Figure~\ref{fig:axams-varimp} shows 10~covariates with the highest permutation
+Figure~\ref{fig:axams-varimp} shows the 10~covariates with the highest permutation
 importance (i.e., change in CRPS) for station Axams. As expected the NWP outputs
@@
 would typically not have been considered in a GAMLSS.
-
-\pagebreak
+\fixme{Reto: Denke, dass man das zu leicht attakieren kann. Man sieht an der variable importance
+die nichtlinearitaet nicht bzw. moegliche Interaktionen. Die Variablen so in ein GAMLSS
+zu werfen haben wir aber auch schon getan. Hier muss noch ein besserer Abschluss hin denke ich.}
+
+%\pagebreak
 
@@
 forest had higher (green) or lower (red) CRPS compared to the best of the 
-other three models.}
+other three models. The background shows the local topography, the darker the
+higher \citep{srtm}.}
 \end{figure}
 
-After considering only one observation station up to now, a second step
-evaluates and compares the competing methods on all 95~stations. As in the previous
+After considering only one observational site up to now, a second step
+evaluates and compares the competing methods on all 95 available stations. As in the previous
 section, all models are learned on the first 24~years and evaluated by the
@@
 station and visualized by parallel coordinates plots with boxplots superimposed
-in Figure~\ref{fig:all-crps}. Overall distributional forests have a slightly
+in Figure~\ref{fig:all-crps}. Overall, distributional forests have a slightly
 higher improvement in CRPSS compared to the two GAMLSS which is best seen
-by looking at the boxplots and also in the green line representing the results
+by looking at the boxplots and the green line representing the results
 for station Axams. The underlying parallel coordinates additionally
@@
 (sometimes better, sometimes worse) compared to the two data-driven models.
-Finally, because some CRPSS become negative, it is also clear that the simple
-EMOS performs better than the other models for a few of the stations.
-
+Values below zero show that, for some stations, EMOS performs better
+than the more complex statistical methods.
 %NAs in gamlss in 13 stations!!!\\
@@
 To assess whether these differences in predictive performance are due to differences
-in the topology, Figure~\ref{fig:all-map} shows a brief summary for each station
-on a map of Tyrol (based on the SRTM 90m Digital Elevation 
-Data, \cite{srtm}). Each station is illustrated by a symbol that conveys which
+in the topography, Figure~\ref{fig:all-map} shows a brief spatial summary of all
+stations. Each station is illustrated by a symbol that conveys which
 model performed best in terms of CRPS on the last 4~years of the data.
@@
 for a few stations in the north-east EMOS performs best, and in East Tyrol
-the prespecified EMOS performs particularly well.
+the prespecified GAMLSS performs particularly well.
 
@@
 corresponding score function, the trees and forests can not only pick
-up changes in the location but also the scale or shape of a distributional
+up changes in the location but also the scale or shape of any distributional
 family.
 
-Distributional forests are an attractive alternative when prespecifying or boosting
-all effects and interactions in a GAMLSS model is challenging. They are
-rather straightforward to specify requiring only little prior subject
-matter knowledge and also work well in the presence of many potential
-covariates. The application to precipitation forecasting in complex terrain
-illustrates that distributional forests often perform on par or even
-better than their GAMLSS counterparts. Hence, they form a useful addition
-to the toolbox of probabilistic forecasts, especially in meteorology.
+Distributional forests are an attractive alternative when prespecifying or
+boosting all possible effects and interactions in a GAMLSS model is
+challenging.  Distributional forests are rather straightforward to specify
+requiring only little prior subject matter knowledge and also work well in the
+presence of many potential covariates. The application to precipitation
+forecasting in complex terrain illustrates that distributional forests often
+perform on par or even better than their GAMLSS counterparts. Hence, they form
+a useful addition to the already available toolbox of probabilistic forecasts
+for disciplines such as meteorology.
 
@@
 
-\fixme{replication code?}
+The fitted distributional forest for July 24 and
+observation station Axams illustrated in Figure~\ref{fig:axams-dist} 
+is reproducible using the \textsf{R} code provided as demo in the package
+\textbf{disttree}. This also includes fitting the other zero-censored
+Gaussian models considered in this paper and generating the corresponding 
+PIT histograms (Figure~\ref{fig:axams-pit})
+and QQR plots (Figure~\ref{fig:axams-qq}).
+Full replication of all results can be obtained using the companion 
+\textsf{R} package \textbf{RainTyrol} and the \textsf{R} code of
+\code{replication_code.R} provided in the package \textbf{disttree}.
+
+\fixme{Replication code is already available as described above,
+but still needs some small corrections (naming, parallelization,
+exports/imports/suggests in package RainTyrol, ...) which are in
+process}
 
@@
 
-\fixme{Shorten/omit parts that are in the main paper already. Focus in
-the description on what is actually used in the application, i.e.,
-reduce the emphasis of the modularity of the framework.}
-
-\fixme{Adapt notation.}
+\fixme{Has been revised, but maybe still needs some improvement. Too long?}
 
@@
 can be plugged in the score function $s(\bm{\theta}, Y)$.
-In that way a meassurement of goodness-of-fit is obtained for each
+In that way a measurement of goodness-of-fit is obtained for each
 parameter $\theta_j$ and each observation $y_i$.\\
@@
 
-After Bonferroni adjusting the p-values it has to be assessed whether
+After Bonferroni-adjusting the p-values it has to be assessed whether
 any of the resulting p-values is beneath the selected significance level. 
@@
 As an alternative to QQ~plots based on randomized quantile residuals,
-PIT (probability integral transform) histograms 
-\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used tool to 
-assess how well-calibrated distributional fits are.
-Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
-probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that
-despite some deviations the overall calibration of all models is acceptable.
+probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
+histograms are a commonly used tool to assess how well-calibrated distributional 
+fits are. Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
+probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
+some deviations the overall calibration of all models is acceptable.
 

### 2018-03-27 pkg/disttree/inst/MIP-Seminar/slides.Rnw
added observations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2097 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b88abb1ae358e46429b39966d98827ce03aa5df4

@@
 \item Varying subsamples.
-\item Combine/average over trees.
+\item Combine / average over trees.
 \end{itemize}
@@
 
+# observations
+load("STAGEobs_tawes_11121.rda")
+# "2018-03-14 01:00:00 CET" in line 321
+obs <- obs[-c(1:320),]
+
 getZoo <- function( x, variable ) {
@@
 GFS_t2m <- GFS_t2m + 10.7
+# first value of rain is 18 hours later than temperature 
+#(both start at 6:00 but rain is cumulated over 24 hours)
+# -> drop first 3 values of temperature
+GFS_t2m <- GFS_t2m[-c(1:3),]
+GFS_t2m$obs <- obs[,"obs$tl"]
+GFS_t2m <- GFS_t2m[c(1:40),]
 
 GFS_rain <- getZoo( GFS, "apcpsfc" )
+GFS_rain$obs <- obs[,"rr6"]
 # precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
@@
 }
-# first value of rain is 18 hours later than temperature 
-#(both start at 6:00 but rain is cumulated over 24 hours)
-# -> drop first 3 values of temperature
-GFS_t2m <- GFS_t2m[-c(1:3),]
+## FIX ME: add observations
 
@@
 plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+#plot( GFS_t2m_mean, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+abline( v = main, lty = 2 )
+mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
+plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+#plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+#plot( GFS_rain_mean, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+abline( v = main, lty = 2 )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
+
+axis( side = 1, at = minor, strftime(minor,"%H") )
+axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
+      line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{-0.3cm}
+\setkeys{Gin}{width=1.1\linewidth}
+<<ensemble_1_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
+par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
+layout( matrix(1:2,ncol=1) )
+plot( GFS_t2m[,c(1,22)], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
 #plot( GFS_t2m_mean, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)

### 2018-03-28 pkg/disttree/inst/MIP-Seminar/slides.Rnw
added observations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2105 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9e9d396370bd7d99f5cb77ac614b2a1fadee0262

--- 2018-03-26 pkg/disttree/inst/MIP-Seminar/slides.Rnw
@@
 
+# observations
+load("STAGEobs_tawes_11121.rda")
+# "2018-03-14 00:00:00" in line 321 (first entry for temperature)
+# "2018-03-13 06:00:00" in line 317 (first entry for rain -> aggregated to 24h sums)
+# "2018-03-23 00:00:00" in line 357
+obs <- obs[318:357,]
+
 getZoo <- function( x, variable ) {
@@
 GFS_t2m <- GFS_t2m + 10.7
+GFS_t2m$obs <- obs[,"obs$tl"]
+# first value of rain is 18 hours later than temperature 
+#(both start at 6:00 but rain is cumulated over 24 hours)
+# -> drop first 3 values of temperature
+GFS_t2m <- GFS_t2m[-c(1:3),]
 
 GFS_rain <- getZoo( GFS, "apcpsfc" )
+GFS_rain$obs <- obs[,"rr6"]
 # precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
@@
 }
-# first value of rain is 18 hours later than temperature 
-#(both start at 6:00 but rain is cumulated over 24 hours)
-# -> drop first 3 values of temperature
-GFS_t2m <- GFS_t2m[-c(1:3),]
 
@@
 GFS_t2m_mean <- GFS_t2m
-GFS_t2m_mean[,1] <- rowMeans(GFS_t2m)
+GFS_t2m_mean[,1] <- rowMeans(GFS_t2m[,-NCOL(GFS_t2m)])
 GFS_t2m_mean <- GFS_t2m_mean[,1]
@@
 # GFS_rain_mean <- rowMeans(GFS_rain)
-GFS_rain_mean <- GFS_rain
-GFS_rain_mean[,1] <- rowMeans(GFS_rain)
-GFS_rain_mean <- GFS_rain_mean[,1]
+GFS_rain24_mean <- GFS_rain24
+GFS_rain24_mean[,1] <- rowMeans(GFS_rain24[,-NCOL(GFS_rain24)])
+GFS_rain24_mean <- GFS_rain24_mean[,1]
 @
@@
 
-axis( side = 1, at = minor, strftime(minor,"%H") )
+#axis( side = 1, at = minor, strftime(minor,"%H") )
 axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
+#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
 
@@
 \setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_2_plot, fig=TRUE, echo=FALSE, width=9>>=
+<<ensemble_1_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
 par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
 layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,c(1,2)], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", 
+      ylim = range(GFS_t2m)*1.05 )
+lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
 abline( v = main, lty = 2 )
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
-#plot( GFS_rain[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", 
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
 abline( v = main, lty = 2 )
@@
 \setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_3_plot, fig=TRUE, echo=FALSE, width=9>>=
+<<ensemble_2_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
 par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
 layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+plot( GFS_t2m[,c(1,2)], screen=1, xaxs="i", xaxt="n", 
+      ylim = range(GFS_t2m)*1.05 )
+lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
 abline( v = main, lty = 2 )
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
-#plot( GFS_rain[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+plot( GFS_rain24[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
 abline( v = main, lty = 2 )
@@
 \setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_20_plot, fig=TRUE, echo=FALSE, width=9>>=
+<<ensemble_3_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
 par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
 layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+plot( GFS_t2m[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", 
+      ylim = range(GFS_t2m)*1.05 )
+lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
 abline( v = main, lty = 2 )
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
-#plot( GFS_rain, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
+plot( GFS_rain24[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
 abline( v = main, lty = 2 )
@@
 
-axis( side = 1, at = minor, strftime(minor,"%H") )
+#axis( side = 1, at = minor, strftime(minor,"%H") )
 axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
+#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
+
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{-0.3cm}
+\setkeys{Gin}{width=1.1\linewidth}
+<<ensemble_20_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
+par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
+layout( matrix(1:2,ncol=1) )
+plot( GFS_t2m[,-NCOL(GFS_t2m)], screen=1, xaxs="i", xaxt="n", 
+      ylim = range(GFS_t2m)*1.05 )
+lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
+abline( v = main, lty = 2 )
+mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
+plot( GFS_rain24[,-NCOL(GFS_rain24)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
+abline( v = main, lty = 2 )
+mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
+
+#axis( side = 1, at = minor, strftime(minor,"%H") )
+axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
+      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
+#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
 

### 2018-03-31 pkg/disttree/inst/MIP-Seminar/slides.Rnw
revised slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2108 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ee8043bb64140d2bca8e56dc8a484547bc43bd13

@@
 \frametitle{Statistical modeling}
-\vspace{0.18cm}
-Approximation of reality based on a set of observations.\\
-\vspace{0.19cm}
-\visible<2, 5-> {Assumption:} 
+\visible<2,3, 6-> {Assumption:}
 \vspace{-0.7cm}
@@
 \begin{tikzpicture}
-\visible<2, 5->{
-\node[] at (0,0.4) {X};
-}
-\visible<3-4>{
+\visible<2,3, 6-> {\node[] at (0,0.4) {X};}
+\visible<4-5>{
 \node[inner sep=0pt] (today) at (-1.8,-0.4)
@@
 }
-\visible<2, 5->{
+\visible<2,3, 6->{
 \node[] at (3.5,0.4) {Y};
 }
-\visible<3-4>{
+\visible<4-5>{
 \node[inner sep=0pt] (future) at (5.3,-0.4)
@@
 \end{center}
-\visible<4>{
-\vspace{0.3cm}
+\visible<3, 6->{
+\vspace{-2cm}
+Approximation of reality based on a set of observations.\\
+}
+\visible<5>{
+\vspace{2cm}
 X = observed precipitation of today,\\
@@
 \vspace{0.15cm}
-Y = daily precipitation sums
-}
-
-\vspace{-4cm}
-\visible<6->{
+Y = daily precipitation sums\\
+}
+
+\vspace{-3.5cm}
+\visible<7->{
 \textbf{Application/Motivation:}
@@
 \vspace{0.5cm}
-\visible<7->{
+\visible<8->{
 \textbf{Two main approaches / ``cultures'':}
@@
 \begin{frame}[fragile]
-\frametitle{Statistical modeling}
-\vspace{0.2cm}
-\textbf{Data modeling:}
-\vspace{0.1cm}
+\frametitle{Data modeling}
+\vspace{-0.1cm}
 \begin{center}
@@
 \vspace{-0.4cm}
+\only<1-6>{
 \begin{itemize}
@@
 \end{itemize}
-\only<1>{\vspace{1.5cm}}
+}
+\only<7-10>{
+\vspace{1cm}
+Summary:
+\begin{itemize}
+\item<8-10> Stochastic model to approximate data generating process.\\
+\only<9-10>{$\Rightarrow$ fixed $f$\\}
+\item<10> Disadvantage: Risk of incorrect assumptions.
+\end{itemize}
+}
+\only<1>{\vspace{1cm}}
 
@@
 
-\visible<2->{
+\visible<2-6>{
 \begin{minipage}{0.65\textwidth}
@@
 $$\visible<3>{\mathbb{E}(Y|X) = } f(X; \alpha) = \alpha_0 + \alpha_1 \cdot X$$  
+\vspace{0.15cm}
 }
@@
 $$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
+\vspace{0.15cm}
 }
@@
 sp_rain <- partysplit(2L, index = 1:2)
-sp_humidity <- partysplit(4L, breaks = 80)
+sp_humidity <- partysplit(4L, breaks = 70)
 
@@
   partynode(2L, split = sp_humidity, kids = list(
-      partynode(3L, info = "5 mm/24h"),
-      partynode(4L, info = "8 mm/24h"))),
-  partynode(5L, info = "18 mm/24h")))
+      partynode(3L, info = "4 mm/24h"),
+      partynode(4L, info = "9 mm/24h"))),
+  partynode(5L, info = "15 mm/24h")))
 
@@
 kappa <- 12
+nobs <- 120
 x <- c(1:nobs)/nobs
-ytrue <- ytree <- yforest <- numeric(length = length(x))
-for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
-y <- ytrue + rnorm(nobs,0,0.3)
-for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
+ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- numeric(length = length(x))
+for(i in 1:nobs) ytrue[i] <- if(x[i]<0.55) 4 else 4+5*plogis(kappa*(3*(x[i]-0.36)-1))
+y <- ytrue + rnorm(nobs,0,2)
+y_up <- rep.int(15, length(x)) + rnorm(nobs,0,1)
+for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
+for(i in 1:nobs) ytree2[i] <- if(x[i]<0.6) 4 else 9
+for(i in 1:nobs) ytree3[i] <- if(x[i]<0.75) 4 else 9
+for(i in 1:nobs) ytree4[i] <- if(x[i]<0.55) 4 else 9
+ytree_up <- rep.int(15, length(x))
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.6) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
 @
@@
 ## Random Forest
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
+#%for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
 @
@@
 \begin{frame}[fragile]
-\frametitle{Statistical modeling}
-\vspace{0.2cm}
-\textbf{Algorithmic modeling:}
-\vspace{0.1cm}
+\frametitle{Algorithmic modeling}
+\vspace{-0.1cm}
 \begin{center}
@@
 \vspace{-0.4cm}
+\visible<2->{
 \begin{itemize}
-\item Flexible approximation with no/few assumptions.
+\item Flexible approximation with no/few assumptions %\\
+%$\Rightarrow$ no fixed $f$.
 \item Data-driven model fitting.
 \end{itemize}
-\only<1>{\vspace{1cm}}
-\visible<2->{
-\begin{minipage}{0.65\textwidth}
+}
+\only<2>{\vspace{1cm}}
+\visible<3->{
+\begin{minipage}{0.6\textwidth}
 %\vspace{-0.5cm}
-\textbf{Example:} \only<2>{Tree model}\only<3>{Forest model}\\
+\textbf{Example:} \only<3-6>{Tree model\only<3>{\\}\only<4-6>{s}}\only<7->{Forest model}\\
 %\vspace{-0.8cm}
 \begin{center}
-\only<2>{
-\vspace{-0.7cm}
+\only<3>{
+\vspace{-1cm}
 \resizebox{0.6\textwidth}{!}{
-<<plot_raintree, fig=TRUE, echo=FALSE, width=4.2, height=2.6>>=
-par(mar=c(0,0,0,0), oma=c(0,0,0.3,0))
+<<plot_raintree, fig=TRUE, echo=FALSE, width=4, height=2.7>>=
+par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
 paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
@@
 }
-}
-\only<3>{
-\resizebox{0.6\textwidth}{!}{
+
+\vspace{-0.7cm}
+\begin{flushleft}
+\noindent {\small X = rain now, humdity\\
+Y = rain 24h}
+\end{flushleft}}
+\only<4->{
+\vspace{-0.5cm}
+\resizebox{0.55\textwidth}{!}{
 \begin{tikzpicture}
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n00) at (1, 2) {};
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n01) at (0.5, 1) {};
+%\draw[ellipse, draw=black, align=center, scale = 0.7, minimum width=170pt, minimum height = 35pt] (set) at (1, 3) {};
+\draw[gray] (1,3.5) ellipse (3 and 0.8);
+\begin{scope}
+\clip (1,3.5) ellipse (3 and 0.8);
+\pgfmathsetseed{7}
+\foreach \p in {1,...,200} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
+\end{scope}
+\visible<5->{
+\node[ellipse, draw=black!50!green, align=center, minimum width=80pt, minimum height = 40pt, line width = 3pt] (subset1) at (0.2, 3.5) {};
+}
+\visible<7->{
+\node[ellipse, draw = blue, align=center, minimum width=100pt, minimum height = 30pt, line width = 3pt] (subset2) at (1.2, 3.6) {};
+}
+\visible<9->{
+\node[ellipse, draw = orange, align=center, minimum width=120pt, minimum height = 20pt, line width = 3pt] (subset3) at (1.6, 3.2) {};
+}
+\visible<6->{
+  \node[ellipse, fill=black!50!green, align=center] (n00) at (-1, 2) {};
+  \node[ellipse, fill=black!50!green, align=center] (n01) at (-1.25, 1.25) {};
   \draw[-, line width=1pt] (n00) -- (n01);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n02) at (1.5, 1) {};
-  \draw[-, line width=1pt] (n00) -- (n02);
-
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n10) at (3, 2) {};
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n11) at (2.5, 1) {};
+  \node[rectangle, fill=black!50!green, align=center] (n02) at (-1.5, 0.5) {};
+  \draw[-, line width=1pt] (n01) -- (n02);
+  \node[rectangle, fill=black!50!green, align=center] (n03) at (-1, 0.5) {};
+  \draw[-, line width=1pt] (n01) -- (n03);
+  \node[rectangle, fill=black!50!green, align=center] (n04) at (-0.5, 0.5) {};
+  \draw[-, line width=1pt] (n00) -- (n04);
+}
+\visible<8->{
+  \node[ellipse, fill=blue, align=center] (n10) at (1, 2) {};
+  \node[ellipse, fill=blue, align=center] (n11) at (0.75, 1.25) {};
   \draw[-, line width=1pt] (n10) -- (n11);
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n12) at (3.5, 1) {};
-  \draw[-, line width=1pt] (n10) -- (n12);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n13) at (2, 0) {};
+  \node[rectangle, fill=blue, align=center] (n12) at (0.5, 0.5) {};
+  \draw[-, line width=1pt] (n11) -- (n12);
+  \node[rectangle, fill=blue, align=center] (n13) at (1, 0.5) {};
   \draw[-, line width=1pt] (n11) -- (n13);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n14) at (2.8, 0) {};
-  \draw[-, line width=1pt] (n11) -- (n14);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n15) at (3.2, 0) {};
-  \draw[-, line width=1pt] (n12) -- (n15);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n16) at (4, 0) {};
-  \draw[-, line width=1pt] (n12) -- (n16);
-
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
+  \node[rectangle, fill=blue, align=center] (n14) at (1.5, 0.5) {};
+  \draw[-, line width=1pt] (n10) -- (n14);
+}
+\visible<10->{
+  \node[ellipse, fill=orange, align=center] (n20) at (3, 2) {};
+  \node[ellipse, fill=orange, align=center] (n21) at (2.75, 1.25) {};
   \draw[-, line width=1pt] (n20) -- (n21);
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
-  \draw[-, line width=1pt] (n20) -- (n22);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
-  \draw[-, line width=1pt] (n22) -- (n23);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
-  \draw[-, line width=1pt] (n22) -- (n24);
+  \node[rectangle, fill=orange, align=center] (n22) at (2.5, 0.5) {};
+  \draw[-, line width=1pt] (n21) -- (n22);
+  \node[rectangle, fill=orange, align=center] (n23) at (3, 0.5) {};
+  \draw[-, line width=1pt] (n21) -- (n23);
+  \node[rectangle, fill=orange, align=center] (n24) at (3.5, 0.5) {};
+  \draw[-, line width=1pt] (n20) -- (n24);
+}
+\visible<11->{
+\draw[-, line width=1pt] (-1, -0.2) -- (0.2,-0.7);
+\draw[-, line width=1pt] (1, -0.2) -- (1,-0.7);
+\draw[-, line width=1pt] (3, -0.2) -- (1.8,-0.7);
+\node[rectangle, align=center] (combine) at (1, -1) {\large average};
+}
 \end{tikzpicture}
@@
 \end{minipage}
-\begin{minipage}{0.3\textwidth}
+\hspace{-1cm}
+\begin{minipage}{0.4\textwidth}
 \begin{center}
-\only<2>{
-\vspace{-0.29cm}
-<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+\only<3-5>{
+%\vspace{-0.29cm}
+<<plot_motivation_regtree, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(3,3,2,5))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+points(x=x, y = y_up, col = "slategray", pch = 19)
 box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
 lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-@
-}
-\only<3>{
-<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.4, cex = 3)
+@
+%\hspace{-1cm}
+}
+\only<6-7>{
+<<plot_motivation_randforest1, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(3,3,2,5))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+points(x=x, y = y_up, col = "slategray", pch = 19)
 box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
+lines(x = x, y = ytree, col = pal["forest"], lwd=7)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
+mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.4, cex = 3)
+@
+}
+\only<8-9>{
+<<plot_motivation_randforest2, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(3,3,2,5))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+points(x=x, y = y_up, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytree, col = pal["forest"], lwd=7)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
+lines(x = x, y = ytree3, col = "blue", lwd=7)
+mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.4, cex = 3)
+@
+}
+\only<10>{
+<<plot_motivation_randforest3, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(3,3,2,5))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+points(x=x, y = y_up, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytree, col = pal["forest"], lwd=7)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
+lines(x = x, y = ytree3, col = "blue", lwd=7)
+lines(x = x, y = ytree4, col = "orange", lwd=7)
+mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.4, cex = 3)
+@
+}
+\only<11->{
+<<plot_motivation_randforest4, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(3,3,2,5))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+points(x=x, y = y_up, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = yforest, col = pal["forest"], lwd=7)
+mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.4, cex = 3)
 @
@@
 \frametitle{Goals}
-\vspace{0.3 cm}
-\textbf{Synthesis:} \\
-\vspace{0.2 cm}
+\vspace{-0.7 cm}
+\begin{minipage}{0.7\textwidth}
+\textbf{Available:} \\
+\vspace{-0.3 cm}
 \begin{itemize}
-  \item Probabilistic forecast\\
+  \item<2-> Probabilistic forecast\\
   $\rightarrow$ Distributional approach
-  \medskip
-  \item Automatic detection of steps and abrupt changes,\\
+  \item<3-> Automatic detection of steps and abrupt changes,
   automatic selection of variables and interactions\\
   $\rightarrow$ Tree approach
-  \medskip
-  \item Smoothing and regularization\\
+  \item<4-> Smoothing and regularization\\
   $\rightarrow$ Forest approach
 \end{itemize}
+\end{minipage}
+<<motivation_GAMLSS_syn, echo=FALSE, results=hide>>=
+## GAMLSS
+set.seed(7)
+nobs <- 200
+x_g <- c(1:nobs)/nobs
+x_g <- 2*(x_g-0.5)
+ytrue_g <- x_g^3
+var <- exp(-(2*x_g)^2)/2
+y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var)
+@
+<<motivation_regtree_syn, echo=FALSE, results=hide>>=
+## Reg. Tree
+set.seed(7)
+nobs <- 120
+kappa <- 12
+x <- c(1:nobs)/nobs
+ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- numeric(length = length(x))
+for(i in 1:nobs) ytrue[i] <- if(x[i]<0.55) 4 else 4+5*plogis(kappa*(3*(x[i]-0.36)-1))
+y <- ytrue + rnorm(nobs,0,2)
+y_up <- rep.int(15, length(x)) + rnorm(nobs,0,1)
+for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
+for(i in 1:nobs) ytree2[i] <- if(x[i]<0.6) 4 else 9
+for(i in 1:nobs) ytree3[i] <- if(x[i]<0.75) 4 else 9
+for(i in 1:nobs) ytree4[i] <- if(x[i]<0.55) 4 else 9
+ytree_up <- rep.int(15, length(x))
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.6) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
+@
+\begin{minipage}{0.25\textwidth}
+\vspace{0.5cm}
+\visible<2->{
+<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(4,0,4,0))
+plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var, rev(ytrue_g - 0.1 - var)),
+  col = pallight["forest"], border = "transparent")
+lines(x_g, ytrue_g, col = pal["forest"], lwd=12)
+points(x_g, y_g, col = "slategray", pch = 19)
+box(lwd = 5)
+@
+}
+\visible<3->{
+<<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(4,0,4,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+points(x=x, y = y_up, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytree, col = pal["forest"], lwd=12)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=12, lty = 2)
+@
+}
+\visible<4->{
+<<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=7>>=
+par(mar=c(4,0,4,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+points(x=x, y = y_up, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=12, lty = 2)
+lines(x = x, y = yforest, col = pal["forest"], lwd=12)
+@
+}
+\end{minipage}
+
+%\vspace{0.1cm}
+
+\visible<5->{
+\textbf{Missing:} Synthesis\\
+}
+
+\vspace{0.5cm}
+
+\visible<6->{
+$\Rightarrow$ PhD-Thesis  
+}
+\visible<7->{
+(what I've been working on for the last 2 years)
+}
 \end{frame}
@@
 
+\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
+
 \begin{frame}[fragile]
 \frametitle{Fitting a Tree}
+\begin{minipage}{0.78\textwidth}
 \begin{center}
-\vspace*{-0.12cm}
+\vspace{0.2cm}
+\visible<2->{
 Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_estpar, fig=TRUE, echo=FALSE>>=
+}
+\vspace{-0.2cm}
+\only<2-3>{
+%\setkeys{Gin}{width=0.\linewidth, height=0.68\linewidth}
+<<plottree_estpar, fig=TRUE, echo=FALSE, width = 7>>=
 # function for output in terminal panels
@@
 @
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Fitting a Tree}
-\begin{center}
-\vspace*{-0.12cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_dens, fig=TRUE, echo=FALSE>>=
+}
+\only<4->{
+%\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+<<plottree_dens, fig=TRUE, echo=FALSE, width = 7>>=
 node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
@@
 @
+}
+\visible<3->{
+R-package \code{disttree}
+}
 \end{center}
+\end{minipage}
+\begin{minipage}{0.21\textwidth}
+
+\vspace{1.9cm}
+
+\center
+\visible<5->{
+Algorithmic \\
+Modelling\\
+}
+
+\vspace{1.3cm}
+
+\visible<6->{
+Data \\
+Modelling
+}
+
+\vspace{1.4cm}
+
+\visible<7->{
+Combined \visible<8->{\checkmark}
+}
+\end{minipage}
 \end{frame}
+
+%<<densities, eval=TRUE, echo=FALSE, results=hide>>=
+%x <- c(-300:300)/100
+%set.seed(7)
+%y1 <- dnorm(x, mean = 0, sd = 0.7)
+%y2 <- dnorm(x, mean = 0.8, sd = 0.5)
+%y3 <- dnorm(x, mean = -0.7, sd = 1.2)
+%@
+
+%<<density1, fig=TRUE, echo=FALSE>>=
+%par(mar=c(0,0,0,0))
+%plot(x, y1, type = "l", col = "red", ylim = c(0,0.8), lwd=2,
+%     xaxt='n', yaxt='n')
+%@
+
+%<<density2, fig=TRUE, echo=FALSE>>=
+%par(mar=c(0,0,0,0))
+%plot(x, y2, type = "l", col = "red", ylim = c(0,0.8), lwd=2,
+%     xaxt='n', yaxt='n')
+%@
+
+%<<density3, fig=TRUE, echo=FALSE>>=
+%par(mar=c(0,0,0,0))
+%plot(x, y3, type = "l", col = "red", ylim = c(0,0.8), lwd=2,
+%     xaxt='n', yaxt='n')
+%@
 
@@
 \frametitle{Building Distributional Trees and Forests}
-\begin{minipage}{0.75\textwidth}
+\begin{minipage}{0.7\textwidth}
 {\bf Tree:}
 \begin{enumerate}
-\item<3-> Fit a global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
+\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
 Estimate model parameters $\hat{\theta}$.
-\item<5-> Evaluate how well the model fits the data (for each parameter and each observation).
-\item<7-> Choose the covariate $X$ with the strongest influence or association to
-goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
+\item<6-> Evaluate goodnes of fit \\
+(for each parameter and each observation).
+\item<8-> Choose covariate $X$ with strongest influence on goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
 \end{enumerate}
 \end{minipage}
-\begin{minipage}{0.2\textwidth}
-\vspace{-0.5cm}
+\begin{minipage}{0.25\textwidth}
+\vspace{-1cm}
 \begin{tikzpicture}
@@
 }
-\visible<4->{
+\visible<4>{
 \node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$\mathcal{D}(Y;\hat{\theta}$)};
 }
-\visible<6-8>{
+\visible<5->{
+\node[inner sep=0pt] (density1) at (0.8,2)
+    {\includegraphics[width=0.6\textwidth]{density1.jpeg}};
+}
+\visible<7-8>{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
-\draw[-, line width=0.8pt] (n0) -- (n1);
-\draw[-, line width=0.8pt] (n0) -- (n2);
-}
-\visible<9>{
+\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1);
+\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2);
+}
+\visible<9-10>{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
-\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
-}
-\visible<10->{
+\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<11>{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
-\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<11->{
+\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<12->{
+\node[inner sep=0pt] (density2) at (-0.1,-0.25)
+    {\includegraphics[width=0.5\textwidth]{density2.jpeg}};
+}
+\visible<12->{
+\node[inner sep=0pt] (density3) at (1.7,-0.25)
+    {\includegraphics[width=0.5\textwidth]{density3.jpeg}};
 }
@@
 \item<8-> Find the split point $p$ which leads to the highest improvement.
-\item<11-> Repeat steps 1--4 recursively in the subgroups
+\item<10-> Repeat steps 1--4 recursively in the subgroups
     until some stopping criterion is met.
@@
 \vspace{0.4cm}
-\visible<12->{
+\visible<14->{
 {\bf Forest:} Ensemble of trees.
@@
 
-load("../draft/plot_map_rain/demo/tirol.gadm.rda")
-load("../draft/plot_map_rain/demo/tirol.dem.rda")
+load("../draft/plot_map_rain/data/tirol.rda")
+load("../draft/plot_map_rain/data/dem.rda")
 load("../draft/plot_map_rain/demo/ehyd.statlist.rda")
@@
                                            select = -c(lon,lat)),
-                             proj4string = crs(tirol.dem))
+                             proj4string = crs(dem))
 
@@
   par(mar = c(5,4,4,0.1))
-  raster::image(tirol.dem, col = rev(gray.colors(100)),
+  raster::image(dem, col = rev(gray.colors(100)),
                 main="Stations in Tyrol", cex.main=1.7,
@@
                 ylim = c(46.6, 47.87))
-  plot(tirol.gadm, add = TRUE)
+  plot(tirol, add = TRUE)
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)

### 2018-03-31 pkg/disttree/inst/MIP-Seminar/slides.Rnw
revised slides and added crps results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2109 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f5c423c6c38d0914ba4f9ba541fb0cefaf171c7a

@@
 library("parallel")
+library("RainTyrol")
 library("gamlss.cens")
@@
 \begin{minipage}{0.25\textwidth}
-\vspace{0.5cm}
+\vspace{0.7cm}
 \visible<2->{
 <<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(4,0,4,0))
+par(mar=c(3,0,3,0))
 plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
@@
 <<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(4,0,4,0))
+par(mar=c(3,0,3,0))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 <<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(4,0,4,0))
+par(mar=c(3,0,3,0))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 #### prediction over all stations 24 - 4
-if(file.exists("rain_pred_24to4.rda")){
-  load("rain_pred_24to4.rda")
+#### prediction over all stations 24 - 4
+if(file.exists("crps_24to4_all.rda")){
+  load("crps_24to4_all.rda")
 } else {
 
-  source("rain_pred_24to4.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-
-  res <- rain_pred(seedconst = 7,
-                   gamboost_cvr = TRUE,
-                   frac = FALSE)
-
-  save(res, file = "rain_pred_24to4.rda")
-}
-
-crps_all <- res[[1]]$results["crps",]
-for(i in 2:(length(res)-1))  crps_all <- rbind(crps_all, res[[i]]$results["crps",])
-
-colnames(crps_all) <- colnames(res[[1]]$results)
-
-# skills score
-s <- 1 - crps_all[, 2:4]/crps_all[,6]
-colnames(s) <- c("Forest", "prespecified GAM", "boosted GAM")
+  data("StationsTyrol")
+  stations <- StationsTyrol$name
+  test <- 2009:2012
+  train <- 1985:2008
+
+  res_24to4_all <- mclapply(1:length(stations),
+                            function(i){
+
+                              set.seed(7)
+
+                              res <- eval(station = stations[i],
+                                          train = train,
+                                          test = test,
+                                          gamboost_cvr = TRUE)
+
+                              return(res)
+                            },
+                            mc.cores = detectCores() - 1
+  )
+
+  # extract crps
+  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
+  # loop over all stations
+  for(i in 1:length(stations)){
+    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
+  }
+
+  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
+  rownames(crps_24to4_all) <- stations
+
+  save(crps_24to4_all, file = "crps_24to4_all.rda")
+  save(res_24to4_all, file = "res_24to4_all.rda")
+
+}
+
+# skill score
+s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
+colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
 
@@
 
-crps <- crps_all[,c("distforest", "gamlss", "gamboostLSS", "EMOS log")]
-
-## best method
-bst <- apply(crps, 1, which.min)
-
-## distance of forest to best other method
-dst <- crps[,1] - crps[cbind(1:nrow(crps), apply(crps[, -1], 1, which.min) + 1)]
-
-## breaks/groups
+crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]
+
+# best method
+bst <- apply(crps_map, 1, which.min)
+
+# distance of forest to best other method
+dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]
+
+# breaks/groups
 brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
@@
 
-## HCL colors (relatively flashy, essentially CARTO Tropic)
+# HCL colors (relatively flashy, essentially CARTO Tropic)
 clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
@@
 load("../draft/plot_map_rain/data/dem.rda")
-load("../draft/plot_map_rain/demo/ehyd.statlist.rda")
-
+
+data(StationsTyrol)
 # Create SpatialPointsDataFrame from station list
-sp <- SpatialPointsDataFrame(subset(ehyd.statlist[res$complete_stations,],
+sp <- SpatialPointsDataFrame(subset(StationsTyrol,
                                     select=c(lon,lat)),
-                             data = subset(ehyd.statlist[res$complete_stations,],
+                             data = subset(StationsTyrol,
                                            select = -c(lon,lat)),
                              proj4string = crs(dem))
-
 @
@@
 
+  ## plot map of Tyrol with all 95 observations
   layout(cbind(1, 2), width = c(9, 1))
@@
                 ylab = "Latitude", xlab = "Longitude",
+                cex.lab = 1.4,
                 xlim = c(9.8,13.2), 

### 2018-04-01 pkg/disttree/inst/draft/Axams_24to4.R
reran all calculations and saved all results, adapted replication code and calculations within draft.Rnw, replaced data for background of map by data from CC-BY, added corresponding reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2110 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2e17c4e8585e34e07908172e0958318f66e90739

@@ -1,0 +1,175 @@
+#######################################################
+### Probabilistic Forecasting on Precipitation Data ###
+#######################################################
+
+Axams_24to4 <- function(){
+  
+  library("disttree")
+  
+  #####
+  # load observations and covariates 
+  data("RainAxams")
+  
+  
+  #####
+  # formula
+  {
+    
+    # tree and forest formula
+    dt.formula <- df.formula <- 
+      robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
+      tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
+      tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
+      capepow_mean + capepow_sprd + capepow_min + capepow_max + 
+      capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
+      capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
+      dswrf_mean_mean + dswrf_mean_max + 
+      dswrf_sprd_mean + dswrf_sprd_max +
+      msl_mean_mean + msl_mean_min + msl_mean_max + 
+      msl_sprd_mean + msl_sprd_min + msl_sprd_max +
+      pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
+      pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
+      tmax_mean_mean + tmax_mean_min + tmax_mean_max +
+      tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
+      tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
+      tcolc_sprd_mean + tcolc_sprd_min + tcolc_sprd_max +
+      t500_mean_mean + t500_mean_min + t500_mean_max +
+      t700_mean_mean + t700_mean_min + t700_mean_max +
+      t850_mean_mean + t850_mean_min + t850_mean_max +
+      t500_sprd_mean + t500_sprd_min + t500_sprd_max +
+      t700_sprd_mean + t700_sprd_min + t700_sprd_max +
+      t850_sprd_mean + t850_sprd_min + t850_sprd_max +
+      tdiff500850_mean + tdiff500850_min + tdiff500850_max +
+      tdiff700850_mean + tdiff700850_min + tdiff700850_max +
[…]

### 2018-04-01 pkg/disttree/inst/draft/draft.Rnw
reran all calculations and saved all results, adapted replication code and calculations within draft.Rnw, replaced data for background of map by data from CC-BY, added corresponding reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2110 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2e17c4e8585e34e07908172e0958318f66e90739

@@
 gen.cens(NO, type = "left")
+library("RainTyrol")
 
@@
 #### Axams prediction 24 - 4
-if(file.exists("rain_Axams_pred_24to4.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred_24to4.rda")
-  load("rain_Axams_pred_24to4.rda")
+if(file.exists("Axams_24to4.rda")){
+  load("Axams_24to4.rda")
 } else {
-  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred_24to4.R")
-  source("rain_axams_pred_24to4.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-  res <- rain_axams_pred(seedconst = 7, ntree = 100,
-                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
-                         gamboost_cvr = TRUE,
-                         frac = FALSE)
-
-  save(res, file = "rain_Axams_pred_24to4.rda")
+  source("Axams_24to4.R")
+  Axams_24to4 <- Axams_24to4()
+  save(Axams_24to4, file = "Axams_24to4.rda")
 }
@@
 pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
-pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
+pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata[pdays,], type = "parameter")
 df_mu <- pdf$mu
@@
 df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, res$testdata[pdays,"robs"])
+cbind(df_exp, Axams_24to4$testdata[pdays,"robs"])
 
 # plot predicted distributions together with observations
-#set.seed(res$call$seedconst)
 set.seed(7)
@@
 # predictions
-pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(res$testdata[pdays,"robs"][4], crch::dcnorm(res$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+pred1 <- c(Axams_24to4$testdata[pdays,"robs"][1], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(Axams_24to4$testdata[pdays,"robs"][2], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(Axams_24to4$testdata[pdays,"robs"][3], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(Axams_24to4$testdata[pdays,"robs"][4], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
 
@@
 
+############
+# get predicted parameter of all models for testdata
+
+# distributional tree
+pdt <- predict(Axams_24to4$dt, newdata = Axams_24to4$testdata, type = "parameter")
+dt_mu <- pdt$mu
+dt_sigma <- pdt$sigma
+dt_exp <- pnorm(dt_mu/dt_sigma) * (dt_mu + dt_sigma * (dnorm(dt_mu/dt_sigma) / pnorm(dt_mu/dt_sigma)))
+
+# distributional forest
+pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata, type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+
+# prespecified GAM
+g_mu <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "mu", type = "response", data = Axams_24to4$g_learndata)
+g_sigma <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "sigma", type = "response", data = Axams_24to4$g_learndata)
+g_exp <- pnorm(g_mu/g_sigma) * (g_mu + g_sigma * (dnorm(g_mu/g_sigma) / pnorm(g_mu/g_sigma)))
+
+# boosted GAM
+pgb <- predict(Axams_24to4$gb, newdata = Axams_24to4$testdata, parameter = c("mu","sigma"), type = "response")
+gb_mu <- pgb$mu
+gb_sigma <- pgb$sigma
+gb_exp <- pnorm(gb_mu/gb_sigma) * (gb_mu + gb_sigma * (dnorm(gb_mu/gb_sigma) / pnorm(gb_mu/gb_sigma)))
+
+# EMOS
+ml_mu <- predict(Axams_24to4$ml, type = "location", newdata = Axams_24to4$testdata)
+ml_sigma <- predict(Axams_24to4$ml, type = "scale", newdata = Axams_24to4$testdata)
+ml_exp <- pnorm(ml_mu/ml_sigma) * (ml_mu + ml_sigma * (dnorm(ml_mu/ml_sigma) / pnorm(ml_mu/ml_sigma)))
+
+############
 ## PIT histograms and qqrplots
@@
   set.seed(7)
-  ## disttree
-  if(FALSE){
-    # in sample
-    pdt <- predict(res$dt, type = "parameter")
-    dt_mu_l <- pdt$mu 
-    dt_sigma_l <- pdt$sigma 
-    pit_dt_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = dt_mu_l, sd = dt_sigma_l))
-    pit_dt_l[res$learndata[,"robs"]>0, 1] <- pit_dt_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  pdt <- predict(res$dt, type = "parameter", newdata = res$testdata)
-  dt_mu_t <- pdt$mu 
-  dt_sigma_t <- pdt$sigma 
-  pit_dt_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = dt_mu_t, sd = dt_sigma_t))
-  pit_dt_t[res$testdata[,"robs"]>0, 1] <- pit_dt_t[res$testdata[,"robs"]>0, 2]
-
-  ## distforest
-  if(FALSE){
-    # in sample
-    pdf <- predict(res$df, type = "parameter")
-    df_mu_l <- pdf$mu
-    df_sigma_l <- pdf$sigma
-    pit_df_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = df_mu_l, sd = df_sigma_l))
-    pit_df_l[res$learndata[,"robs"]>0, 1] <- pit_df_l[res$learndata[,"robs"]>0, 2]
-  }
-  # out of sample
-  pdf <- predict(res$df, type = "parameter", newdata = res$testdata)
-  df_mu_t <- pdf$mu
-  df_sigma_t <- pdf$sigma
-  pit_df_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = df_mu_t, sd = df_sigma_t))
-  pit_df_t[res$testdata[,"robs"]>0, 1] <- pit_df_t[res$testdata[,"robs"]>0, 2]
-
-  ## gamlss
-  g_learndata <- res$learndata
-  g_learndata$robs <- Surv(g_learndata$robs, g_learndata$robs>0, type="left")
-  if(FALSE){
-    # in sample
-    g_mu_l <- predict(res$g, what = "mu", type = "response")
-    g_sigma_l <- predict(res$g, what = "sigma", type = "response")
-    pit_g_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = g_mu_l, sd = g_sigma_l))
-    pit_g_l[res$learndata[,"robs"]>0, 1] <- pit_g_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  g_mu_t <- predict(res$g, what = "mu", type = "response", newdata = res$testdata)
-  g_sigma_t <- predict(res$g, what = "sigma", type = "response", newdata = res$testdata)
-  pit_g_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = g_mu_t, sd = g_sigma_t))
-  pit_g_t[res$testdata[,"robs"]>0, 1] <- pit_g_t[res$testdata[,"robs"]>0, 2]
-
-  #gamboostLSS
-  if(FALSE){
-    #in sample
-    pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response")
-    gb_mu_l <- pgb$mu
-    gb_sigma_l <- pgb$sigma
-    pit_gb_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = gb_mu_l, sd = gb_sigma_l))
-    pit_gb_l[res$learndata[,"robs"]>0, 1] <- pit_gb_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  pgb <- predict(res$gb, parameter = c("mu","sigma"), type = "response", newdata = res$testdata)
-  gb_mu_t <- pgb$mu
-  gb_sigma_t <- pgb$sigma
-  pit_gb_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = gb_mu_t, sd = gb_sigma_t))
-  pit_gb_t[res$testdata[,"robs"]>0, 1] <- pit_gb_t[res$testdata[,"robs"]>0, 2]
-
-  #EMOS
-  if(FALSE){
-    # in sample
-    ml_mu_l <- predict(res$ml, type = "location")     # returns parameter on response scale
-    ml_sigma_l <- predict(res$ml, type = "scale")
-    pit_ml_l <- cbind(0, pnorm(res$learndata[,"robs"], mean = ml_mu_l, sd = ml_sigma_l))
-    pit_ml_l[res$learndata[,"robs"]>0, 1] <- pit_ml_l[res$learndata[,"robs"]>0, 2]
-  }  
-  # out of sample
-  ml_mu_t <- predict(res$ml, type = "location", newdata = res$testdata)     # returns parameter on response scale
-  ml_sigma_t <- predict(res$ml, type = "scale", newdata = res$testdata)
-  pit_ml_t <- cbind(0, pnorm(res$testdata[,"robs"], mean = ml_mu_t, sd = ml_sigma_t))
-  pit_ml_t[res$testdata[,"robs"]>0, 1] <- pit_ml_t[res$testdata[,"robs"]>0, 2]
+
+  # distributional tree
+  pit_dt <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = dt_mu, sd = dt_sigma))
+  pit_dt[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_dt[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # distributional forest
+  pit_df <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = df_mu, sd = df_sigma))
+  pit_df[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_df[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # prespecified GAM
+  pit_g <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = g_mu, sd = g_sigma))
+  pit_g[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_g[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # boosted GAM
+  pit_gb <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = gb_mu, sd = gb_sigma))
+  pit_gb[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_gb[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # EMOS
+  pit_ml <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = ml_mu, sd = ml_sigma))
+  pit_ml[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_ml[Axams_24to4$testdata[,"robs"]>0, 2]
 
@@
 if(file.exists("vimp_crps.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
   load("vimp_crps.rda")
@@
 
+  set.seed(7)
   # locally redefine logLik.distforest to use crps in varimp()
@@
 
-  vimp_crps <- varimp(res$df, nperm = 1L)
-  #vimp10_crps <- varimp(res$df, nperm = 10L)
-  #vimp100_crps <- varimp(res$df, nperm = 100L)
-  #vimp1000_crps <- varimp(res$df, nperm = 1000L)
-
-  save(vimp_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp_crps.rda")
-  #save(vimp10_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp10_crps.rda")
-  #save(vimp100_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp100_crps.rda")
-  #save(vimp1000_crps, file = "~/svn/partykit/pkg/disttree/inst/draft/vimp1000_crps.rda")
+  vimp_crps <- varimp(Axams_24to4$df, nperm = 1L)
+
+  save(vimp_crps, file = "vimp_crps.rda")
 
@@
 par(mfrow = c(2, 2))
-qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_ml_t, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_g_t,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_df, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_ml, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_g,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_gb, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
 @
@@
 #### cross validation rain
-if(file.exists("rain_Axams_7x10.rda")){
-  load("rain_Axams_7x10.rda")
+if(file.exists("crps_cross.rda")){
+  load("crps_cross.rda")
 } else {
-  source("rain_cross_7x10.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-  rainres <- rain_cross(stationname = "Axams", 
-                        seedconst = 7, ntree = 100,
-                        forest_mtry = 27,
-                        gamboost_cvr = TRUE,
-                        frac = FALSE)
-
-  save(rainres, file = "rain_Axams_7x10.rda")
+
+  nrep_cross <- 10
+  seed <- 7
+
+  res_cross <- mclapply(1:nrep_cross,
+                        function(i){
+
+                          set.seed(seed*i)
+
+                          # randomly split data in 7 parts each including 4 years
+                          years <- 1985:2012
+                          testyears <- list()
+                          for(j in 1:7){
+                            testyears[[j]] <- sample(years, 4, replace = FALSE)
+                            years <- years[!(years %in% testyears[[j]])]
                           }
 
-rain_rmse <- matrix(0, ncol = ncol(rainres[[1]]$rmse), length(rainres)-1)
-rain_ll <- matrix(0, ncol = ncol(rainres[[1]]$ll), length(rainres)-1)
-rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres)-1)
-
-for(i in 1:(length(rainres)-1)){
-  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse, na.rm = TRUE)
-  rain_ll[i,] <- colMeans(rainres[[i]]$ll, na.rm = TRUE)
-  rain_crps[i,] <- colMeans(rainres[[i]]$crps, na.rm = TRUE)
+                          #crps <- matrix(nrow = 7, ncol = 7)
+                          reslist <- list()
+                          for(k in 1:7){
+                            test <- testyears[[k]]
+                            train <- c(1985:2012)[!c(1985:2012) %in% test]
+
+                            res <- eval(station = "Axams",
+                                        train = train,
+                                        test = test,
+                                        gamboost_cvr = TRUE)
+
+                            #crps[k,] <- res$crps
+                            reslist[[k]] <- res
                           }
 
-colnames(rain_rmse) <- colnames(rain_ll) <- colnames(rain_crps) <- colnames(rainres[[1]]$rmse)
+                          #colnames(crps) <- names(res$crps)
+                          return(reslist)
+                        },
+                        mc.cores = detectCores() - 1
+  )
+
+  # extract CRPS
+  crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
+  # loop over all repetitions
+  for(i in 1:length(res_cross)){
+    #loop over all 7 folds (for 7 methods)
+    crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
+    for(j in 1:length(res_cross[[1]])){
+      crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
+    }
+    crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
+  }
+  colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps)
+
+  save(crps_cross, file = "crps_cross.rda")
+  save(res_cross, file = "res_cross.rda")
+
+}
 @
@@
 <<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 7>>=
-boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
+boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
         names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
@@
 #### prediction over all stations 24 - 4
-if(file.exists("rain_pred_24to4.rda")){
-  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
-  load("rain_pred_24to4.rda")
+if(file.exists("crps_24to4_all.rda")){
+  load("crps_24to4_all.rda")
 } else {
 
-  source("rain_pred_24to4.R")
-  library("gamlss.cens")
-  gen.cens("NO", type = "left")
-
-  res <- rain_pred(seedconst = 7,
-                   gamboost_cvr = TRUE,
-                   frac = FALSE)
-
-  save(res, file = "rain_pred_24to4.rda")
+  data("StationsTyrol")
+  stations <- StationsTyrol$name
+  test <- 2009:2012
+  train <- 1985:2008
+
+  res_24to4_all <- mclapply(1:length(stations),
+                            function(i){
+
+                              set.seed(7)
+
+                              res <- eval(station = stations[i],
+                                          train = train,
+                                          test = test,
+                                          gamboost_cvr = TRUE)
+
+                              return(res)
+                            },
+                            mc.cores = detectCores() - 1
+  )
+
+  # extract crps
+  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
+  # loop over all stations
+  for(i in 1:length(stations)){
+    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
   }
 
-#ll_all <- res[[1]]$results["ll",]
-#for(i in 2:(length(res)-1)) ll_all <- rbind(ll_all, res[[i]]$results["ll",])
-
-#rmse_all <- res[[1]]$results["rmse",]
-#for(i in 2:(length(res)-1)) rmse_all <- rbind(rmse_all, res[[i]]$results["rmse",])
-
-crps_all <- res[[1]]$results["crps",]
-for(i in 2:(length(res)-1))  crps_all <- rbind(crps_all, res[[i]]$results["crps",])
-
-#colnames(ll_all) <- colnames(rmse_all) <- 
-  colnames(crps_all) <- colnames(res[[1]]$results)
+  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
+  rownames(crps_24to4_all) <- stations
+
+  save(crps_24to4_all, file = "crps_24to4_all.rda")
+  save(res_24to4_all, file = "res_24to4_all.rda")
+
+}
 
 # skill score
-s <- 1 - crps_all[, 2:4]/crps_all[,6]
+s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
 colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
@@
 
-load("plot_map_rain/demo/tirol.gadm.rda")
-load("plot_map_rain/demo/tirol.dem.rda")
-load("plot_map_rain/demo/ehyd.statlist.rda")
-
+load("plot_map_rain/data/tirol.rda")
+load("plot_map_rain/data/dem.rda")
+
+data(StationsTyrol)
 # Create SpatialPointsDataFrame from station list
-sp <- SpatialPointsDataFrame(subset(ehyd.statlist[res$complete_stations,],
+sp <- SpatialPointsDataFrame(subset(StationsTyrol,
                                     select=c(lon,lat)),
-                             data = subset(ehyd.statlist[res$complete_stations,],
+                             data = subset(StationsTyrol,
                                            select = -c(lon,lat)),
-                             proj4string = crs(tirol.dem))
+                             proj4string = crs(dem))
 @
@@
 
+  ## plot map of Tyrol with all 95 observations
   layout(cbind(1, 2), width = c(9, 1))
   par(mar = c(5,4,4,0.1))
-  raster::image(tirol.dem, col = rev(gray.colors(100)),
+  raster::image(dem, col = rev(gray.colors(100)),
                 main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
@@
 forest had higher (green) or lower (red) CRPS compared to the best of the 
-other three models. The background shows the local topography, the darker the
-higher \citep{srtm}.}
+other three models. The background shows the local topography \citep{Robinson:2014}.}
 \end{figure}
@@
 par(mfrow = c(2,2))
-pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
-pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
-pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
-pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
+pithist(pit_df, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
+pithist(pit_ml, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
+pithist(pit_g, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
+pithist(pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
 

### 2018-04-03 pkg/disttree/inst/draft/draft.Rnw
now calling function evalmodels instead of eval
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2121 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 70b7de01b854ba21101f6cbf01857c84866fcb2c

@@
 
-                            res <- eval(station = "Axams",
+                            res <- evalmodels(station = "Axams",
                                               train = train,
@@
 
-                              res <- eval(station = stations[i],
+                              res <- evalmodels(station = stations[i],
                                                 train = train,

### 2018-04-04 pkg/disttree/inst/MIP-Seminar/slides.Rnw
revised slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2123 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5db251d5602213d07a40d20468176d8d0d8bfafb

@@
 options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
+
+set.seed(7)
+
 library("ggplot2")
@@
 \item<8-10> Stochastic model to approximate data generating process.\\
-\only<9-10>{$\Rightarrow$ fixed $f$\\}
+\only<9-10>{$\Rightarrow$ fixed $f$.\\}
 \item<10> Disadvantage: Risk of incorrect assumptions.
@@
 \only<5-6>{
-\vspace{-0.8cm}
+\vspace{-1.3cm}
+
 \begin{align*}
@@
 \begin{itemize}
-\item Flexible approximation with no/few assumptions %\\
+\item Flexible approximation with no/few assumptions. %\\
 %$\Rightarrow$ no fixed $f$.
@@
 \only<3>{
-\vspace{-1cm}
+\vspace{-0.96cm}
 \resizebox{0.6\textwidth}{!}{
@@
 \begin{flushleft}
-\noindent {\small X = rain now, humdity\\
-Y = rain 24h}
+\noindent {\small X = rain today, humidity\\
+Y = rain tomorrow}
 \end{flushleft}}
@@
 %\draw[ellipse, draw=black, align=center, scale = 0.7, minimum width=170pt, minimum height = 35pt] (set) at (1, 3) {};
-\draw[gray] (1,3.5) ellipse (3 and 0.8);
+\draw[gray,rounded corners=10pt] (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
+%\draw[gray] (1,3.5) ellipse (3 and 0.8);
 \begin{scope}
-\clip (1,3.5) ellipse (3 and 0.8);
+\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
+%\clip (1,3.5) ellipse (3 and 0.8);
 \pgfmathsetseed{7}
@@
 \visible<5->{
-\node[ellipse, draw=black!50!green, align=center, minimum width=80pt, minimum height = 40pt, line width = 3pt] (subset1) at (0.2, 3.5) {};
+\node[ellipse, draw=black!50!green, align=center, minimum width=80pt, minimum height = 40pt, line width = 3pt] (subset1) at (0.1, 3.5) {};
 }
@@
 \visible<9->{
-\node[ellipse, draw = orange, align=center, minimum width=120pt, minimum height = 20pt, line width = 3pt] (subset3) at (1.6, 3.2) {};
+\node[ellipse, draw = orange, align=center, minimum width=120pt, minimum height = 25pt, line width = 3pt] (subset3) at (1.6, 3.3) {};
 }
@@
 \begin{center}
+\vspace{0.1cm}
+
 \only<3-5>{
-%\vspace{-0.29cm}
+%\vspace{0.01cm}
 <<plot_motivation_regtree, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5))
+par(mar=c(3,3,2,5.5))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
-mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.4, cex = 3)
-@
-%\hspace{-1cm}
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
+mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.5, cex = 3)
+@
 }
 \only<6-7>{
+%\vspace{0.01cm}
+\hspace{-0.2cm}
 <<plot_motivation_randforest1, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5))
+par(mar=c(3,3,2,5.5))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
 lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.4, cex = 3)
+lines(x = x, y = ytree_up, col = pal["tree"], lwd=7)
+mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.5, cex = 3)
 @
@@
 \only<8-9>{
+%\vspace{0.01cm}
+\hspace{-0.3cm}
 <<plot_motivation_randforest2, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5))
+par(mar=c(3,3,2,5.5))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
 lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
 lines(x = x, y = ytree3, col = "blue", lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.4, cex = 3)
+lines(x = x, y = ytree_up, col = "blue", lwd=7)
+mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.5, cex = 3)
 @
@@
 \only<10>{
+%\vspace{0.01cm}
+\hspace{-0.3cm}
 <<plot_motivation_randforest3, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5))
+par(mar=c(3,3,2,5.5))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = ytree_up, col = "orange", lwd=7)
 lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
@@
 lines(x = x, y = ytree4, col = "orange", lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.4, cex = 3)
+mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.5, cex = 3)
 @
@@
 \only<11->{
+%\vspace{0.01cm}
+\hspace{-0.3cm}
 <<plot_motivation_randforest4, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5))
+par(mar=c(3,3,2,5.5))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 box(lwd=5)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7, lty = 2)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
 lines(x = x, y = yforest, col = pal["forest"], lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 15, line = 0.5, cex = 3, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 8, line = 0.5, cex = 3, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.4, cex = 3)
+mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("humidity", side = 1, line = 1.5, cex = 3)
 @
@@
 \begin{frame}[fragile]
-\frametitle{Fitting a Tree}
+\frametitle{Distributional Tree}
 \begin{minipage}{0.78\textwidth}
@@
 \visible<2->{
-Model: \code{disttree(y ~ x)}\\
+Model: \code{disttree(y $\sim$ x)}\\
 }
@@
 Algorithmic \\
-Modelling\\
+Modeling\\
 }
@@
 Data \\
-Modelling
+Modeling
 }
@@
 Estimate model parameters $\hat{\theta}$.
-\item<6-> Evaluate goodnes of fit \\
+\item<6-> Evaluate goodness of fit \\
 (for each parameter and each observation).
-\item<8-> Choose covariate $X$ with strongest influence on goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
+\item<8-> Choose covariate $X$ with strongest influence on goodness of fit of $\mathcal{D}(Y; \hat{\theta})$ as split variable.
 \end{enumerate}
@@
 \vspace{0.4cm}
-\visible<14->{
+\visible<13->{
 {\bf Forest:} Ensemble of trees.
@@
 \vspace*{0.2cm}
-\textbf{Application for one stations:}
+\textbf{Application for one station:}
 \begin{itemize}
@@
 \vspace{0.2cm}
-\textbf{Application for all 95 station:}
+\textbf{Application for all 95 stations:}
 \begin{itemize}
@@
 \item Evaluate on 4 years.
-\item Compare to commonly used heteroscedastic censored Gaussian models:
+\item Compare to commonly used heteroscedastic censored Gaussian models.
 \end{itemize}
@@
                 ylab = "Latitude", xlab = "Longitude",
-                cex.lab = 1.4,
+                cex.lab = 1.3,
                 xlim = c(9.8,13.2), 

### 2018-04-04 pkg/disttree/inst/MIP-Seminar/slides.Rnw
revised slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2125 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fbe982d0fbe31324ad2cb9e32e15b077fefcc8b5

@@
 # if tree: 2 splits => 3 terminal nodes
-# 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
+# 10 split variables (x1, ..., x10) are available (5 numeric, 2 binary, 3 categorical)
 #
@@
 \begin{frame}[fragile]
-\frametitle{Statistical modeling}
+\frametitle{Statistical Modeling}
 \visible<2,3, 6-> {Assumption:}
@@
 \begin{frame}[fragile]
-\frametitle{Data modeling}
+\frametitle{Data Modeling}
 \vspace{-0.1cm}
@@
 x <- c(1:nobs)/nobs
-ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- numeric(length = length(x))
+ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- 
+  ytree_up <- ytree_up2 <- ytree_up3 <- ytree_up4 <- 
+  yforest <- numeric(length = length(x))
 for(i in 1:nobs) ytrue[i] <- if(x[i]<0.55) 4 else 4+5*plogis(kappa*(3*(x[i]-0.36)-1))
@@
 for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
-for(i in 1:nobs) ytree2[i] <- if(x[i]<0.6) 4 else 9
-for(i in 1:nobs) ytree3[i] <- if(x[i]<0.75) 4 else 9
-for(i in 1:nobs) ytree4[i] <- if(x[i]<0.55) 4 else 9
+for(i in 1:nobs) ytree2[i] <- if(x[i]<0.63) 4.3 else 9.4
+for(i in 1:nobs) ytree3[i] <- if(x[i]<0.75) 3.7 else 9.2
+for(i in 1:nobs) ytree4[i] <- if(x[i]<0.55) 4.4 else 8.7
 ytree_up <- rep.int(15, length(x))
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.6) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
+ytree_up2 <- rep.int(14.5, length(x))
+ytree_up3 <- rep.int(15.3, length(x))
+ytree_up4 <- rep.int(15.5, length(x))
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.63) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
 @
@@
 \begin{frame}[fragile]
-\frametitle{Algorithmic modeling}
+\frametitle{Algorithmic Modeling}
 \vspace{-0.1cm}
@@
 %\vspace{-0.5cm}
-\textbf{Example:} \only<3-6>{Tree model\only<3>{\\}\only<4-6>{s}}\only<7->{Forest model}\\
+\textbf{Example:} \only<3>{Tree model}\only<4-10>{Subsampling}\only<3>{\\}\only<11->{Forest model}\\
 %\vspace{-0.8cm}
@@
 \begin{flushleft}
-\noindent {\small X = rain today, humidity\\
-Y = rain tomorrow}
+\noindent {\small X = rain now, humidity\\
+Y = rain 24h}
 \end{flushleft}}
@@
 mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
 mtext("humidity", side = 1, line = 1.5, cex = 3)
@@
 lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["tree"], lwd=7)
+lines(x = x, y = ytree_up2, col = pal["tree"], lwd=7)
 mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
 mtext("humidity", side = 1, line = 1.5, cex = 3)
@@
 lines(x = x, y = ytree3, col = "blue", lwd=7)
-lines(x = x, y = ytree_up, col = "blue", lwd=7)
+lines(x = x, y = ytree_up2, col = pal["tree"], lwd=7)
+lines(x = x, y = ytree_up3, col = "blue", lwd=7)
 mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
 mtext("humidity", side = 1, line = 1.5, cex = 3)
@@
 lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = "orange", lwd=7)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
 lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
 lines(x = x, y = ytree3, col = "blue", lwd=7)
+lines(x = x, y = ytree_up2, col = pal["tree"], lwd=7)
+lines(x = x, y = ytree_up3, col = "blue", lwd=7)
 lines(x = x, y = ytree4, col = "orange", lwd=7)
+lines(x = x, y = ytree_up4, col = "orange", lwd=7)
 mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
 mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("today", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain tomorrow", side = 2, las = FALSE, line = 1, cex = 3)
+mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
+mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
 mtext("humidity", side = 1, line = 1.5, cex = 3)
@@
 \frametitle{Goals}
-\vspace{-0.7 cm}
+\vspace{-0.4cm}
 \begin{minipage}{0.7\textwidth}
 \textbf{Available:} \\
-\vspace{-0.3 cm}
+\vspace{-0.5 cm}
 \begin{itemize}
-  \item<2-> Probabilistic forecast\\
-  $\rightarrow$ Distributional approach
-  \item<3-> Automatic detection of steps and abrupt changes,
-  automatic selection of variables and interactions\\
-  $\rightarrow$ Tree approach
-  \item<4-> Smoothing and regularization\\
-  $\rightarrow$ Forest approach
+  \item Distributional regression:\\
+  Probabilistic forecast.
+  \item Regression Trees:\\
+  Automatic detection of abrupt changes,
+  and selection of variables/interactions.
+  \item Random Forests:\\
+  Smoothing and regularization.
 \end{itemize}
+
+\vspace{1cm}
+
+\visible<2->{
+\textbf{Missing:} Synthesis
+
+\vspace{0.2cm}
+
+$\Rightarrow$ PhD-Thesis
+
+\vspace{0.1cm}
+
+%{\small 
+(what I've been working on the last 2 years)
+}
 \end{minipage}
@@
 ytrue_g <- x_g^3
-var <- exp(-(2*x_g)^2)/2
-y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var)
+var_g <- exp(-(2*x_g)^2)/2
+y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var_g)
 @
@@
 ## Reg. Tree
-set.seed(7)
-nobs <- 120
+set.seed(74)
+nobs <- 170
 kappa <- 12
 x <- c(1:nobs)/nobs
-ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- numeric(length = length(x))
-for(i in 1:nobs) ytrue[i] <- if(x[i]<0.55) 4 else 4+5*plogis(kappa*(3*(x[i]-0.36)-1))
-y <- ytrue + rnorm(nobs,0,2)
-y_up <- rep.int(15, length(x)) + rnorm(nobs,0,1)
+ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- var <- var_up <- numeric(length = length(x))
+for(i in 1:nobs) ytrue[i] <- if(x[i]<0.2) 4 else 4+5*plogis(kappa*(3*(x[i]-0.33)-1))
+ytrue_up <- rep.int(15, length(x))
+for(i in 1:nobs) var_up[i] <- if(x[i]<0.15) 3 else (if(x[i]<0.35) 2 else (if(x[i]<0.5) 2 else (if(x[i]<0.7) 1 else (if(x[i]<0.8) 1.5 else (if(x[i]<0.9) 2 else (2.5))))))
+for(i in 1:nobs) var[i] <- if(x[i]<0.15) 4 else (if(x[i]<0.2) 3 else (if(x[i]<0.25) 2 else (if(x[i]<0.4) 1 else (if(x[i]<0.8) 2 else (if(x[i]<0.9) 2.5 else (3))))))
+y <- ytrue + rnorm(nobs,0, sqrt(var))
+y_up <- ytrue_up + rnorm(nobs,0, sqrt(var_up))
 for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
-for(i in 1:nobs) ytree2[i] <- if(x[i]<0.6) 4 else 9
-for(i in 1:nobs) ytree3[i] <- if(x[i]<0.75) 4 else 9
-for(i in 1:nobs) ytree4[i] <- if(x[i]<0.55) 4 else 9
 ytree_up <- rep.int(15, length(x))
@@
 \begin{minipage}{0.25\textwidth}
-\vspace{0.7cm}
-\visible<2->{
-<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,0,3,0))
+\vspace{0.4cm}
+<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=8.5>>=
+par(mar=c(1,0,1,0))
 plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
-polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var, rev(ytrue_g - 0.1 - var)),
+polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g)),
   col = pallight["forest"], border = "transparent")
-lines(x_g, ytrue_g, col = pal["forest"], lwd=12)
+lines(x_g, ytrue_g, col = pal["forest"], lwd=10)
 points(x_g, y_g, col = "slategray", pch = 19)
@@
 @
-}
-\visible<3->{
-<<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,0,3,0))
+
+<<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=8.5>>=
+par(mar=c(1,0,1,0))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 box(lwd=5)
-lines(x = x, y = ytree, col = pal["forest"], lwd=12)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=12, lty = 2)
-@
-}
-\visible<4->{
-<<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,0,3,0))
+lines(x = x, y = ytree, col = pal["forest"], lwd=10)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=10)
+@
+
+<<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=8.5>>=
+par(mar=c(1,0,1,0))
 plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
@@
 box(lwd=5)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=12, lty = 2)
-lines(x = x, y = yforest, col = pal["forest"], lwd=12)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=10)
+lines(x = x, y = yforest, col = pal["forest"], lwd=10)
+@
+
+\vspace{0.2cm}
+
+\visible<2->{
+<<plot_motivation_distforest, fig=TRUE, echo=FALSE, width=6>>=
+par(mar=c(0,0,3,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
+polygon(c(x, rev(x)), c(yforest + var, rev(yforest - var)),
+  col = pallight["forest"], border = "transparent")
+polygon(c(x, rev(x)), c(ytrue_up + var_up, rev(ytrue_up - var_up)),
+  col = pallight["forest"], border = "transparent")
+points(x=x, y = y_up, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytree_up, col = pal["forest"], lwd=10)
+lines(x = x, y = yforest, col = pal["forest"], lwd=10)
 @
@@
 \end{minipage}
-
-%\vspace{0.1cm}
-
-\visible<5->{
-\textbf{Missing:} Synthesis\\
-}
-
-\vspace{0.5cm}
-
-\visible<6->{
-$\Rightarrow$ PhD-Thesis  
-}
-\visible<7->{
-(what I've been working on for the last 2 years)
-}
 \end{frame}
@@
 layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
+plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05, col = "slategray" )
 #plot( GFS_t2m_mean, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
@@
 mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
 #plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
@@
 plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05 )
+      ylim = range(GFS_t2m)*1.05, col = "slategray" )
 lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
@@
 plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
 lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
@@
 plot( GFS_t2m[,c(1,2)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05 )
+      ylim = range(GFS_t2m)*1.05, col = "slategray" )
 lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
@@
 plot( GFS_rain24[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
 lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
@@
 plot( GFS_t2m[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05 )
+      ylim = range(GFS_t2m)*1.05, col = "slategray" )
 lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
@@
 plot( GFS_rain24[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
 lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
@@
 plot( GFS_t2m[,-NCOL(GFS_t2m)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05 )
+      ylim = range(GFS_t2m)*1.05, col = "slategray" )
 lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
@@
 plot( GFS_rain24[,-NCOL(GFS_rain24)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
+      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
 lines( GFS_rain24[,"obs"], col = "red", lwd = 2)

### 2018-04-05 pkg/disttree/inst/draft/draft.Rnw
small corrections and additional reference for Generalized Random Forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2126 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 14070a331da16bc471dee29f4ce6ae11c3a72ec3

@@
 \begin{cases}
-    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
-    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
+    %\log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
+    %\log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
+    \log\left\{\sigma^{-1} \cdot \phi\left((Y - \mu) \cdot \sigma^{-1}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
+    \log\left\{\Phi\left((-\mu) \cdto \sigma^{-1}\right)\right\}, & \text{if } Y = 0
 \end{cases}
 $$
+
 where $\phi$ and $\Phi$ are the probability density function and the 
@@
 $$ \label{eq:foc}
-\sum_{i = 1}^n s(\bm{\hat \theta}, y_i) = 0,
-$$
-where $s(\bm{\theta}, y_i)$ is the associated score function
+\sum_{i = 1}^n s(\bm{\hat{\theta}}, y_i) = 0,
+$$
+where $s(\bm{\hat \theta}, y_i)$ is the associated score function
 $$ \label{eq:score}
-s(\bm{\theta}, y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
+s(\bm{\hat{\theta}}, y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i) \mid_{\bm{\theta} = \bm{\hat{\theta}}}.
 $$ 
@@
 the forest).
+
+Distributional random forests share some concepts and algorithmic aspects 
+with other generalizations of Breiman and Cutler's random forests. Nearest 
+neighbor weights are employed for aggregation in survival forests 
+\citep{Hothorn+Lausen+Benner:2004},
+quantile regression forests \citep{Meinshausen:2006},
+transformation forests \citep{Hothorn+Zeileis:2017},
+and generalized random forests for causal inferences 
+\citep{Athey+Tibshirani+Wager:2017}.
+These procedures aggregate over trees fitted to specific score functions 
+(e.g., log rank scores in survival trees, model residuals in transformation 
+or generalized forests). Distributional random forests, in contrast to these 
+nonparametric approaches, provide a compromise between model flexibility 
+and interpretability: The parameters of a problem-specific distribution 
+(zero left-censored normal for precipitation) have a clear meaning but 
+may depend on external variables in a quite general way.
 
@@
 aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
-in their choice of $\frac{1}{1.6}$ as a suitable power parameter for the
+in their choice of $1.6^{-1}}$ as a suitable power parameter for the
 region of Tyrol. The same power transformation is applied to both the
@@
   following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
-  on the 80~available variables, 8 terms are included in the location
-  submodel and 6 in the scale submodel. Both involve an interaction of
+  on the 80~available variables, eight terms are included in the location
+  submodel and six in the scale submodel. Both involve an interaction of
   \emph{tp} and \emph{cape} in the afternoon (between 1200\,UTC and 1800\,UTC)
@@
 splits the available 28~years into 7~subsets of 4~randomly selected years.
-The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
+The models are learned on six~folds (= 24~years) and evaluated on the 7-th fold
 (= 4~years) using the average CRPS across all observations.
-The resulting 10~CRPS skill scores are displayed by boxplots in
+The resulting ten~CRPS skill scores are displayed by boxplots in
 Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
@@
 \textsf{R} package \textbf{RainTyrol} and the \textsf{R} code of
-\code{replication_code.R} provided in the package \textbf{disttree}.
+\code{replication.R} provided in the package \textbf{disttree}.
 
@@
 (\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
-can be plugged in the score function $s(\bm{\theta}, Y)$.
+can be plugged in the score function $s(\bm{\hat{\theta}}, Y)$.
 In that way a measurement of goodness-of-fit is obtained for each
@@
 \begin{align*}
-H_0^l:  s(\bm{\hat{\theta}}, y) \qquad \bot \qquad Z_l 
+H_0^l:  s(\bm{\hat{\theta}}, Y) \qquad \bot \qquad Z_l 
 \end{align*}
@@
 statistic can be standardized. The observed multivariate linear 
-statistic $t$ is mapped into the real line by a  univariate test 
+statistic $t$ is mapped onto the real line by a  univariate test 
 statistic $c$. In the application of this paper a quadratic form 
@@
 $$
-c_{quad}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
+c_{\text{quad}}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
 $$
@@
 
-The smaller the p-value corresponding to the standardized test 
+The smaller the $p$-value corresponding to the standardized test 
 statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
@@
 
-After Bonferroni-adjusting the p-values it has to be assessed whether
-any of the resulting p-values is beneath the selected significance level. 
-If so, the partitioning variable $Z_{l^\ast}$ with the lowest p-value 
+After Bonferroni-adjusting the $p$-values it has to be assessed whether
+any of the resulting $p$-values is beneath the selected significance level. 
+If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
 is chosen as splitting variable.
 
-As split point the breakpoint that leads to the highest discrepancy 
-between score functions in the two resulting subgroups is selected. 
+The breakpoint that leads to the highest discrepancy between score functions 
+in the two resulting subgroups is selected as split point. 
 This difference can be measured by the linear statistic
@@
 
-Repeat the testing and splitting procedure in each of the resulting 
+One repeats the testing and splitting procedure in each of the resulting 
 subgroups until some stopping criterion is reached. This criterion 
 can for example be a minimal number of observations in a node or a 
-minimal p-value for the statistical tests. In that way pre-pruning
+minimal $p$-value for the statistical tests. In that way pre-pruning
 is applied in order to find right-sized trees and hence 
@@
 histograms are a commonly used tool to assess how well-calibrated distributional 
-fits are. Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
+fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
 probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
@@
 %\end{subfigure}
-\caption{\label{fig:pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
+\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
 \end{figure}

### 2018-04-05 pkg/disttree/inst/draft/draft.Rnw
final touches incorporating Torsten's and Reto's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2127 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 44a48df3d37767b050b25054e7f90ee80823d816

@@
 \documentclass[nojss]{jss}
+\shortcites{Hamill+Bates+Whitaker:2013}
 
@@
 %\VignetteDepends{disttree, crch, countreg, gamlss, gamlss.cens}
-%\VignetteKeywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}
+%\VignetteKeywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
 %\VignettePackage{disttree}
@@
 
-\hyphenation{Qua-dra-tic}
-
-%% colors
-\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
-\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
-
 <<preliminaries, echo=FALSE, results=hide>>=
@@
 transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
 @
@@
 explanatory variables, a distributional approach is often adopted where the
-parameter(s) of the distribution are linked to regressors. In many classical
-models this only captures the location/expectation of the distribution but
+parameters of the distribution are linked to regressors. In many classical
+models this only captures the location of the distribution but
 over the last decade there has been increasing interest in distributional
@@
 mean and variance of a Gaussian response and is particularly
-popular in weather forecasting and, more generally, the GAMLSS framework
+popular in weather forecasting. More generally, the GAMLSS framework
 allows to establish generalized additive models for location, scale, and shape
@@
 so far, no general distributional framework is available for these. Therefore,
-a framework for distributional trees and forests is proposed that blends
+a framework for distributional regression trees and forests is proposed that blends
 regression trees and random forests with classical distributions from the GAMLSS
@@
 To illustrate these novel approaches in practice, they are employed to obtain
-probabilistic precipitation forecasts at numerous locations in a mountainous
-region (Tyrol, Austria) based on a very large number of numerical weather
-prediction quantities. It is shown that distributional random forests
+probabilistic precipitation forecasts at numerous sites in a mountainous
+region (Tyrol, Austria) based on a large number of numerical weather
+prediction quantities. It is shown that the novel distributional regression forests
 automatically select variables and interactions, performing on par or often even
@@
 }
-\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}
+\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
 
@@
 \end{tikzpicture}
-\caption{\label{fig:gam}Parametric modeling developments. (Generalized) linear models (left), generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
+\caption{\label{fig:gam}Parametric modeling developments. (Generalized) linear models (left),
+generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
 \end{figure}
@@
 interest typically include exceedence probabilities for certain thresholds of the
-response or quantiles of the response distribution. As an example, consider
+response or quantiles of the response distribution. Specifically, consider
 weather forecasting where there is less interest in the mean amount of
@@
 \begin{itemize}
-  \item \emph{Distributional trees} split the data into more homogeneous groups with
+  \item \emph{Distributional regression trees} (for short: \emph{distributional trees}) split the data into more homogeneous groups with
     respect to a parametric distribution, thus capturing changes in any distribution
     parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left)
-  \item \emph{Distributional forests} utilize an ensemble of distributional trees
+  \item \emph{Distributional regression forests} (for short: \emph{distributional forests}) utilize an ensemble of distributional trees
     for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
@@
 \setkeys{Gin}{width=0.75\textwidth}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4>>=
-par(mar = c(4, 4, 3, 1.5))
-plot(x = x, y = y1, type = "l", col = "red", 
+<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=3.8>>=
+par(mar = c(3.8, 4, 2.3, 1.5))
+plot(x = x, y = y1, type = "l", col = pal[1], lwd = 1.3, 
      main = paste0("July ", pday), ylab = "Density", 
@@
 
-lines(x = x, y = y2, type = "l", col = "blue")
-lines(x = x, y = y3, type = "l", col = "darkgreen")
-lines(x = x, y = y4, type = "l", col = "purple")
+lines(x = x, y = y2, type = "l", col = pal[3], lwd = 1.3)
+lines(x = x, y = y3, type = "l", col = pal[2], lwd = 1.3)
+lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
 legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
-       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
+       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)
 
 # plot point mass
-lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
-lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
-lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
-lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)
-
-points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
-points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
-points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
-points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)
+lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = pal[1], type = "l", lwd = 1.3)
+lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = pal[3], type = "l", lwd = 1.3)
+lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = pal[2], type = "l", lwd = 1.3)
+lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = pal[4], type = "l", lwd = 1.3)
+
+points(x = pm1[1], y = pm1[2], col = pal[1], pch = 19)
+points(x = pm2[1], y = pm2[2], col = pal[3], pch = 19)
+points(x = pm3[1], y = pm3[2], col = pal[2], pch = 19)
+points(x = pm4[1], y = pm4[2], col = pal[4], pch = 19)
 
 # plot predictions
-points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
-points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
-points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
-points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
-
-lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2)
-lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2)
-lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2)
-lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2)
+points(x = pred1[1], y = pred1[2], col = pal[1], pch = 4)
+points(x = pred2[1], y = pred2[2], col = pal[3], pch = 4)
+points(x = pred3[1], y = pred3[2], col = pal[2], pch = 4)
+points(x = pred4[1], y = pred4[2], col = pal[4], pch = 4)
+
+lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
 
 # add labels
-text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
-text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
-text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
-text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
-@
-\caption{\label{fig:axams-dist}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008.
-Observations are left-censored at 0.
-The corresponding predicted point mass is shown at the censoring point (0).}
+text(x = -0.8, y = lh1, labels = "2009", col = pal[1], cex = 0.8)
+text(x = -0.8, y = lh2, labels = "2010", col = pal[3], cex = 0.8)
+text(x = -0.8, y = lh3, labels = "2011", col = pal[2], cex = 0.8)
+text(x = -0.8, y = lh4, labels = "2012", col = pal[4], cex = 0.8)
+@
+\caption{\label{fig:axams-dist}Total precipitation predictions by a distributional forest
+at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
+Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
 \end{figure}
@@
 effects are not yet resolved by the NWP models. 
-Thus, effects with abrupt changes and possibly non-linear interactions 
+Thus, effects with abrupt changes and possibly nonlinear interactions 
 might be required to account for site-specific unresolved features.
@@
 Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
-More specifically, a zero left-censored Gaussian
-distribution is employed to model 24-hour total precipitation so that the
+More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
 zero-censored point mass describes the probability of observing no precipitation 
@@
 80~covariates derived from a wide range of different NWP quantities. As
-Figure~\ref{fig:axams-dist} shows, the station-wise models yield a full
+Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
 distributional forecast for each day -- here for one specific day (July~24) at one
 station (Axams) over four years (2009--2012) -- based on the previous 24~years
-as training data. The corresponding observations conform reasonably well with
+as learning data. The corresponding observations conform reasonably well with
 the predictions. In Section~\ref{sec:precipitation} we investigate the
 performance of distributional forests in this forecasting task in more detail.
-It is shown that they perform at least on par and sometimes clearly better than
-three alternative zero-censored Gaussian models: a standard ensemble model
+Compared to three alternative zero-censored Gaussian models distributional forests
+perform at least on par and sometimes clearly better while requiring no expert
+knowledge for the model specification. The three alternatives are: a standard ensemble model
 output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
@@
 \citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
-\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection among all
+\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
 80~available regressors.
@@
 The general distributional notation is exemplified in all three steps for the
-zero-censored Gaussian distribution. The latter is employed in the empirical 
+Gaussian distribution, left-censored at zero. The latter is employed in the empirical 
 case study in Section~\ref{sec:precipitation} to model power-transformed daily 
@@
 variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
-\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$. Based on the
-corresponding probability density function $f(Y; \bm{\theta})$ the log-likelihood
-function is defined $\ell(\bm{\theta}; Y) = \log\{f(Y; \bm{\theta})\}$. The GAMLSS
+\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$ and
+corresponding log-likelihood function $\ell(\bm{\theta}; Y)$. The GAMLSS
 framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
@@
 while accounting for a point mass at zero (i.e., dry days without
-rain), zero-censored Gaussian distribution with mean/location parameter
-$\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
+rain), a Gaussian distribution left-censored at zero (for short: zero-censored Gaussian)
+with location parameter $\mu$ and scale parameter $\sigma$ is employed. Therefore,
 the corresponding log-likelihood function with parameter vector 
@@
 \begin{cases}
-    %\log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
-    %\log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
-    \log\left\{\sigma^{-1} \cdot \phi\left((Y - \mu) \cdot \sigma^{-1}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
-    \log\left\{\Phi\left((-\mu) \cdto \sigma^{-1}\right)\right\}, & \text{if } Y = 0
+    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
+    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
 \end{cases}
@@
 of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
-maximum likelihood (ML) based on the learning sample
+maximum likelihood (ML) based on the learning sample with observations
 $\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
@@
 $$
-where $s(\bm{\hat \theta}, y_i)$ is the associated score function
+where $s(\bm{\theta}; y_i)$ is the associated score function
 $$ \label{eq:score}
-s(\bm{\hat{\theta}}, y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i) \mid_{\bm{\theta} = \bm{\hat{\theta}}}.
+s(\bm{\theta}; y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
 $$ 
@@
 In weather forecasting, these covariates typically include the output
-from numerical weather prediction systems and/or lagged observations.
-
-To incorporate the covariates into the distributional model, GAMLSS considers
-them as regressors in additive predictors
-$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$.
+from numerical weather prediction systems and/or lagged meteorological observations.
+
+To incorporate the covariates into the distributional model, 
+they are considered as regressors in additive predictors
+$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
+in GAMLSS.
 Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
@@
 additive terms and their functional forms in advance which can be challenging in
-practice, especially if the number of covariates $m$ is large.
+practice and potentially require domain knowledge, especially if the number of covariates $m$ is large.
 
@@
 including covariates in a data-driven way and allowing for abrupt
-changes, non-additive effects, and interactions. In the context of
+changes, nonlinear and non-additive effects, and interactions. In the context of
 distributional models the goal is to partition the covariate space
@@
 on the model even if the true underlying effect is smooth. Hence, ensemble
-methods such as bagging or random forests \citep{Breiman:2001} typically smooth
-the effects, stabilize the model, and improve predictive performance.
-
-While classical random forests \citep{Breiman:2001} grow ensembles of trees
+methods such as bagging or random forests \citep{Breiman:2001} are typically
+applied to smooth the effects, stabilize the model, and improve predictive performance.
+
+While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
 that pick up changes in the location of the response across the covariates,
@@
 $$
-w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} \mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
-$$
+w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
+\frac{\mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
+$$
+where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
+segment of the $t$-th tree.
 Thus, these $w^{\text{forest}}_i(\bold{z}) \in [0, 1]$ whereas
@@
 
-Distributional random forests share some concepts and algorithmic aspects 
+Distributional forests share some concepts and algorithmic aspects 
 with other generalizations of Breiman and Cutler's random forests. Nearest 
@@
 (e.g., log rank scores in survival trees, model residuals in transformation 
-or generalized forests). Distributional random forests, in contrast to these 
+or generalized forests). Distributional forests, in contrast to these 
 nonparametric approaches, provide a compromise between model flexibility 
 and interpretability: The parameters of a problem-specific distribution 
-(zero left-censored normal for precipitation) have a clear meaning but 
+(zero-censored Gaussian for precipitation) have a clear meaning but 
 may depend on external variables in a quite general way.
@@
 \cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
-specific quantities and sites/locations.  
+specific quantities and sites.  
 In case of precipitation forecasting EMOS typically uses the ensemble mean of 
@@
 effects in the NWP system. For example, in the Tyrolean Alps -- considered
-in the following application case study -- the NWP grid cells of $50 \times 50$
+in the following case study -- the NWP grid cells of $50 \times 50$
 km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
@@
 performance of the EMOS by including additional predictor variables, either 
-from local observations or an NWP model. Unfortunately, it is typically unknown 
+from local meteorological observations or an NWP model. Unfortunately, it is typically unknown 
 which variables are relevant for improving the predictions. Simply including
@@
 
-In the following, it is illustrated how distributional regression forests
-can solve this problem of automatically selecting relevant variables,
-interactions, and potentially nonlinear effects. For fitting the forest
-only the response distribution and the list of potential predictor variables
-need to be specified (along with a few algorithmic details) and then
-the model fit is determined by the forest in a data-driven way. Here, we
-employ a zero-censored Gaussian distribution and 80~predictor variables
-computed from ensemble means and spreads of various NWP outputs.
-The predictive performance of the forest is compared to three other
-zero-censored Gaussian models: (a) a standard basic EMOS, (b) a GAMLSS
-with prespecified effects and interactions based on meteorological
-knowledge/experience, and (c) a boosted GAMLSS with automatic
+In the following, it is illustrated how distributional forests can
+solve this problem without requiring prior expert knowledge.  For fitting the
+forest only the response distribution and the list of potential predictor
+variables need to be specified (along with a few algorithmic details) and then
+the relevant variables, interactions, and potentially nonlinear effects are
+determined automatically in a data-driven way. Here, we employ a zero-censored
+Gaussian distribution and 80~predictor variables computed from ensemble means
+and spreads of various NWP outputs. The predictive performance of the forest is
+compared to three other zero-censored Gaussian models: (a)~a standard basic
+EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
+meteorological knowledge/experience, and (c)~a boosted GAMLSS with automatic
 selection of smooth additive terms based on all 80~predictor variables.
@@
 \emph{tp}: total precipitation,           & 12  & ensemble mean of sums over 24h, \\
-\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
+\hspace*{0.5cm} power transformed (by $1.6^{-1}$) &    & ensemble std. deviation of sums over 24h, \\ 
 \emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
 \hspace*{0.9cm} potential energy, &    & ensemble maximum of sums over 24h\\
-\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 \\
+\hspace*{0.9cm} power transformed (by $1.6^{-1}$)&    & \qquad all for 6--30 \\
 \hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
@@
 
-Training and validation data consist of observed 
+Learning and validation data consist of observed 
 daily precipitation sums provided by the National Hydrographical Service 
@@
 Observations are obtained for 95~stations all over Tyrol and surroundings,
-providing 24~hour precipitation sums measured at 0600\,UTC and rigorously
+providing 24-hour precipitation sums measured at 0600\,UTC and rigorously
 quality-checked by the National Hydrographical Service. NWP outputs
@@
 initialized daily at 0000~UTC from December 1984 to present 
-providing forecasts on a 6\ hourly temporal resolution. Each of the 
+providing forecasts on a 6-hourly temporal resolution. Each of the 
 11 ensemble members uses slightly different perturbed initial 
-conditions to predict the situation specific uncertainty 
+conditions to predict the situation-specific uncertainty 
 of the atmospheric state.
@@
 aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
-in their choice of $1.6^{-1}}$ as a suitable power parameter for the
+in their choice of $1.6^{-1}$ as a suitable power parameter for the
 region of Tyrol. The same power transformation is applied to both the
@@
 covariate dependency and included covariates for each distribution 
-parameter. \emph{A}$\ast$\emph{B} indicates an interaction between
+parameter. \emph{A} $\ast$ \emph{B} indicates an interaction between
 covariate \emph{A} and \emph{B}.}
@@
   on the 80~available variables, eight terms are included in the location
-  submodel and six in the scale submodel. Both involve an interaction of
+  submodel and five in the scale submodel. Both involve an interaction of
   \emph{tp} and \emph{cape} in the afternoon (between 1200\,UTC and 1800\,UTC)
@@
 
-In a first step, we show a detailed comparison of the competing models
-for one observation site, Axams in Tyrol (selected somewhat arbitrarily;
-the station in the data set closest to Innsbruck, capital of Tyrol and base of three of the
-authors). As for all other stations, daily precipitation observations and
-numerical weather predictions are available for the month of July from 1985
-through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
-forecasts from the distributional forest, trained on 1985--2008, for July 24
-in 2009--2012 have already been shown as a motivational example. The figure
-shows that the model properly
-depicts the point masses at zero (i.e., the probability of a dry day) and
-the forecasted probability density function for the total amount of precipitation.
-The three sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
-censoring while conforming quite well with the actual observations
-from these days. While this is a nice illustrative example we are  
-interested in the overall predictive performance and calibration of the
-distributional fits.
-
 \begin{figure}[t!]
@@
 \end{figure}
+
+In a first step, we show a detailed comparison of the competing models
+for one observation site, Axams in Tyrol (in the center of the study area,
+see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation observations and
+numerical weather predictions are available for the month of July from 1985
+through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
+forecasts from the distributional forest, trained on 1985--2008, for July 24
+in 2009--2012 have already been shown as a motivational example. The figure
+shows that the model properly
+depicts the point masses at zero (i.e., the probability of a dry day) and
+the forecasted probability density function for the total amount of precipitation.
+The four sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
+censoring while conforming quite well with the actual observations
+from these days. While this is a nice illustrative example we are  
+interested in the overall predictive performance and calibration of the
+distributional fits.
 
@@
 splits the available 28~years into 7~subsets of 4~randomly selected years.
-The models are learned on six~folds (= 24~years) and evaluated on the 7-th fold
+The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
 (= 4~years) using the average CRPS across all observations.
-The resulting ten~CRPS skill scores are displayed by boxplots in
+The resulting 10~CRPS skill scores are displayed by boxplots in
 Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
@@
 \setkeys{Gin}{width=0.7\textwidth}
-<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
-par(mar = c(3,10,2,2))
+<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
+par(mar = c(2.5,10,1,2))
 barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
@@
 of the afternoon), spreads, or minima/maxima. Thus, while the covariates
-themselves are not surprising, a specification with all the transformations
-would typically not have been considered in a GAMLSS.
-\fixme{Reto: Denke, dass man das zu leicht attakieren kann. Man sieht an der variable importance
-die nichtlinearitaet nicht bzw. moegliche Interaktionen. Die Variablen so in ein GAMLSS
-zu werfen haben wir aber auch schon getan. Hier muss noch ein besserer Abschluss hin denke ich.}
+themselves are not surprising, selecting a GAMLSS with a particular combination
+of all the transformations would be much more challenging.
 
@@
           xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
-lines(s[70,], col = "limegreen", type = "o", pch = 19, lwd = 2)  
+lines(s[70,], col = pal[2], type = "o", pch = 19, lwd = 2)  
 # Station Axams is the 77th station which is the 70th in the list of complete stations
@@
   plot(tirol, add = TRUE)
+  points(sp[70,], pch = 19, col = "black", cex = 1.85)
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
@@
 forest had higher (green) or lower (red) CRPS compared to the best of the 
-other three models. The background shows the local topography \citep{Robinson:2014}.}
+other three models. The background shows the local topography \citep{Robinson:2014}.
+Station Axams is highlighted in bold.}
 \end{figure}
@@
 
-The proposed methods are in the \textsf{R} package \textbf{disttree} on
-\textsf{R}-Forge within the \textbf{partykit} project
+The proposed methods are in the \textsf{R} package \textbf{disttree}
+(version~\Sexpr{packageVersion("disttree")}) based on the \textbf{partykit}
+package (version~\Sexpr{packageVersion("partykit")}), both available 
+on \textsf{R}-Forge at
 (\url{https://R-Forge.R-project.org/projects/partykit/}). The function
@@
 GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
-(\citealt{Stasinopoulos+Rigby:2007}) or with custom lists containing
-all required information about the distribution family.
+(\citealt{Stasinopoulos+Rigby:2007}, version~\Sexpr{packageVersion("gamlss.dist")})
+or with custom lists containing all required information about the distribution family.
 
 In addition to \textbf{disttree}, Section~\ref{sec:precipitation} employs
-the functions \code{crch} from the \textsf{R} package \textbf{crch} 
-\citep{Messner+Mayr+Zeileis:2016} for the EMOS models,
-\code{gamlss} from the \textsf{R} package \textbf{gamlss}
-\citep{Stasinopoulos+Rigby:2007} for the prespecified GAMLSS, and
-\code{gamboostLSS} from the \textsf{R} package \textbf{gamboostLSS}
-\citep{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
+\textsf{R} package \textbf{crch} 
+\citep[version~\Sexpr{packageVersion("crch")}]{Messner+Mayr+Zeileis:2016} for the EMOS models,
+\textbf{gamlss}
+\citep[version~\Sexpr{packageVersion("gamlss")}]{Stasinopoulos+Rigby:2007} for the prespecified GAMLSS, and
+\textbf{gamboostLSS}
+\citep[version~\Sexpr{packageVersion("gamboostLSS")}]{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
 
 The fitted distributional forest for July 24 and
-observation station Axams illustrated in Figure~\ref{fig:axams-dist} 
-is reproducible using the \textsf{R} code provided as demo in the package
-\textbf{disttree}. This also includes fitting the other zero-censored
+observation station Axams (including Figure~\ref{fig:axams-dist})
+is reproducible using 
+\code{demo("RainAxams", package = "disttree")}.
+This also includes fitting the other zero-censored
 Gaussian models considered in this paper and generating the corresponding 
-PIT histograms (Figure~\ref{fig:axams-pit})
-and QQR plots (Figure~\ref{fig:axams-qq}).
-Full replication of all results can be obtained using the companion 
-\textsf{R} package \textbf{RainTyrol} and the \textsf{R} code of
-\code{replication.R} provided in the package \textbf{disttree}.
-
-\fixme{Replication code is already available as described above,
-but still needs some small corrections (naming, parallelization,
-exports/imports/suggests in package RainTyrol, ...) which are in
-process}
+QQ~plots (Figure~\ref{fig:axams-qq}) and
+PIT histograms (Figure~\ref{fig:axams-pit}).
+Full replication of all results can be obtained with
+\code{demo("RainTyrol", package = "disttree")}
+requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
+also available within the \textsf{R}-Forge~project.
 
@@
 
-\fixme{Has been revised, but maybe still needs some improvement. Too long?}
-
 In the following, the tree algorithm applied in the empirical case study 
-discussed in this paper is going to be explained. For notational simplicity, 
-the testing and splitting procedure is described for the root node with 
-observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
-In each child node the corresponding subset of observations depends on 
-the assignment made by the foregoing split.
+discussed in this paper is explained. For notational simplicity, 
+the testing and splitting procedure is described for the root node, i.e.,
+the entire learning sample with observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
+In each child node the corresponding subsample depends on 
+the foregoing split(s).
 
 After fitting a distributional model $\mathcal{D}(Y, \bm{\theta})$ to 
-the set of observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
+the learning sample with observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
 Section~\ref{sec:distfit} the resulting estimated parameter 
@@
 (\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
-can be plugged in the score function $s(\bm{\hat{\theta}}, Y)$.
-In that way a measurement of goodness-of-fit is obtained for each
-parameter $\theta_j$ and each observation $y_i$.\\
+can be plugged in the score function $s(\bm{\theta}, Y)$.
+In that way a goodness-of-fit measurement is obtained for each
+parameter $\theta_j$ and each observation $y_i$.
 To use this information, statistical tests are employed to detect
@@
 s(\bm{\hat{\theta}}, y_1)_1 & s(\bm{\hat{\theta}}, y_1)_2 & \ldots & s(\bm{\hat{\theta}}, y_1)_k\\
-s(\bm{\hat{\theta}}, y_2)_1 & s(\bm{\hat{\theta}}, y_2)_2 & \ldots & s(\bm{\hat{\theta}}, y_2)_k\\
 \vdots & \vdots & \ddots & \vdots \\
@@
 
-These hypotheses are assessed by applying permutation testing with 
+These hypotheses are assessed with permutation tests based on
 the multivariate linear statistic
@@
 variable $Z_l$.
-
+%
 After Bonferroni-adjusting the $p$-values it has to be assessed whether
@@
 in the two resulting subgroups is selected as split point. 
-This difference can be measured by the linear statistic
+This is measured by the linear statistic
 $$
@@
 $$
-
 One repeats the testing and splitting procedure in each of the resulting 
@@
 
+\begin{figure}[b!]
+\centering
+%\begin{subfigure}{0.8\textwidth}
+<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7>>=
+
+# out of sample
+set.seed(4)
+
+par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
+pithist(pit_df, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
+pithist(pit_ml, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
+pithist(pit_g, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
+pithist(pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
+
+@
+%\end{subfigure}
+\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
+\end{figure}
+
 As an alternative to QQ~plots based on randomized quantile residuals,
@@
 
-\begin{figure}[t!]
-\centering
-%\begin{subfigure}{0.8\textwidth}
-<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
-
-# out of sample
-set.seed(4)
-
-par(mfrow = c(2,2))
-pithist(pit_df, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
-pithist(pit_ml, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
-pithist(pit_g, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
-pithist(pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
-
-@
-%\end{subfigure}
-\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
-\end{figure}
-
 \end{appendix}

### 2018-04-07 pkg/disttree/inst/draft/draft.Rnw
deleted not used color names
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2134 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3641611a5da40f14767c78cddfcb88fbf2e7b8eb

@@
 pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
-
 pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
-
 transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
@@
 box(lwd=5)
-lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
+lines(x = x, y = ytrue, col = pal[1], lwd = 7, main = "")
 @
@@
 polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
-  col = pallight["forest"], border = "transparent")
-lines(x, ytrue, col = pal["forest"], lwd=7)
+  col = pallight[1], border = "transparent")
+lines(x, ytrue, col = pal[1], lwd=7)
 points(x, y, col = "slategray", pch = 19)
@@
 #lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
-lines(x = x, y = ytree, col = pal["forest"], lwd=7)
+lines(x = x, y = ytree, col = pal[1], lwd=7)
 @
@@
 #lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
-lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
+lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
 @
@@
 polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
-  col = pallight["forest"], border = "transparent")
-lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
+  col = pallight[1], border = "transparent")
+lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
 points(x, y, col = "slategray", pch = 19)
@@
 rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
-lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
+lines(x = x, y = ytree, col = pal[1], lwd=7, main = "")
 polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
-  col = pallight["forest"], border = "transparent")
+  col = pallight[1], border = "transparent")
 points(x, y, col = "slategray", pch = 19)
@@
 <<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 7>>=
+par(mar = c(2.5,2,1,2))
 boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
@@
         ylab = "CRPS skill score", col = "lightgray") 
-abline(h = 0, col = pal["EMOS"], lwd = 2)
+abline(h = 0, col = pal[5], lwd = 2)
 @
@@
 
-load("plot_map_rain/data/tirol.rda")
-load("plot_map_rain/data/dem.rda")
-
-data(StationsTyrol)
+#load("plot_map_rain/data/tirol.rda")
+#load("plot_map_rain/data/dem.rda")
+
+data(StationsTyrol, package = "RainTyrol")
+data(MapTyrol, package = "RainTyrol")
+data(MapTyrol_border, package = "RainTyrol")
 # Create SpatialPointsDataFrame from station list
@@
   boxplot(s, add = TRUE, col = "transparent")
-  abline(h = 0, col = pal["EMOS"], lwd = 2)
+  abline(h = 0, col = pal[5], lwd = 2)
 

### 2018-04-09 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2139 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3dbb05a0185a04b39327a106aa37a18860b620d5

@@
 \draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](7.1,0)--(7.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](0,-2.4)--(0,-3.2);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](2.2,-5.4)--(7.9,-2.5);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](0,-2.4)--(0,-3.2);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.2,-5.4)--(7.9,-2.5);
 \end{tikzpicture}

### 2018-04-12 pkg/disttree/inst/draft/draft.Rnw
added missing dots at sentence endings
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2148 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a16fac85a966d6a4caf60b2b758df112d1c8a1d8

@@
     stabilizes the recursive partitions from individual trees and hence better
-    approximates smooth functions (Figure~\ref{fig:tree}, top middle)
+    approximates smooth functions (Figure~\ref{fig:tree}, top middle).
 \end{itemize}
@@
     respect to a parametric distribution, thus capturing changes in any distribution
-    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left)
+    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left).
   \item \emph{Distributional regression forests} (for short: \emph{distributional forests}) utilize an ensemble of distributional trees
@@
   strongest association or instability. Choose the breakpoint with the highest
-  improvement in the log-likelihood or the highest discrepancy .
+  improvement in the log-likelihood or the highest discrepancy.
 \item Repeat steps 1--3 recursively in the subsamples until these become too

### 2018-04-12 pkg/disttree/inst/MIP-Seminar/slides.Rnw
last touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2149 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 900e9ecccfd891c2b1cf9bcba64ae4ac93104716

@@
 \frametitle{Statistical Modeling}
-\visible<2,3, 6-> {Assumption:}
+\visible<2,3, 5-> {Assumption:}
 \vspace{-0.7cm}
@@
 \begin{tikzpicture}
-\visible<2,3, 6-> {\node[] at (0,0.4) {X};}
-\visible<4-5>{
+\visible<2,3, 5-> {\node[] at (0,0.4) {X};}
+\visible<4>{
 \node[inner sep=0pt] (today) at (-1.8,-0.4)
     {\includegraphics[width=.365\textwidth]{today.png}};
-\node[font=\scriptsize] at (-1.9,-2) {15.03.2018};
+\node[font=\scriptsize] at (-1.9,-2) {2018-03-15};
 }
@@
 }
-\visible<2,3, 6->{
+\visible<2,3, 5->{
 \node[] at (3.5,0.4) {Y};
 }
-\visible<4-5>{
+\visible<4>{
 \node[inner sep=0pt] (future) at (5.3,-0.4)
     {\includegraphics[width=.365\textwidth]{future.png}};
-\node[font=\scriptsize] at (5.4,-2) {16.03.2018};
+\node[font=\scriptsize] at (5.4,-2) {2018-03-16};
 }
@@
 \end{center}
-\visible<3, 6->{
+\visible<3, 5->{
 \vspace{-2cm}
@@
 }
-\visible<5>{
+\visible<4>{
 \vspace{2cm}
@@
 \vspace{-3.5cm}
-\visible<7->{
+\visible<6->{
 \textbf{Application/Motivation:}
@@
 \vspace{0.5cm}
-\visible<8->{
+\visible<7->{
 \textbf{Two main approaches / ``cultures'':}
@@
 \vspace{-0.4cm}
-\only<1-6>{
+\only<1-5>{
 \begin{itemize}
@@
 }
-\only<7-10>{
+\only<6->{
 \vspace{1cm}
@@
 \begin{itemize}
-\item<8-10> Stochastic model to approximate data generating process.\\
-\only<9-10>{$\Rightarrow$ fixed $f$.\\}
-\item<10> Disadvantage: Risk of incorrect assumptions.
+\item Stochastic model to approximate data generating process.\\
+\only{$\Rightarrow$ fixed $f$.\\}
+\item Disadvantage: Risk of incorrect assumptions.
 \end{itemize}
@@
 
-\visible<2-6>{
+\visible<2-5>{
 \begin{minipage}{0.65\textwidth}
@@
 }
-\only<5-6>{
-\vspace{-1.3cm}
+\only<5>{
+\vspace{-1.29cm}
 
 \begin{align*}
-\visible<6>{\mathbb{E}(Y|X) = }\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
-\visible<6>{\sqrt{Var(Y|X)} = }\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
+\visible<5>{\mathbb{E}(Y|X) = }\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
+\visible<5>{\sqrt{Var(Y|X)} = }\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
 \end{align*}
@@
 @
-\only<5-6>{
+\only<5>{
 <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
@@
 \vspace{0.2cm}
-\visible<2->{
+\visible<1->{
 Model: \code{disttree(y $\sim$ x)}\\
@@
 \vspace{-0.2cm}
-\only<2-3>{
+\only<1>{
 %\setkeys{Gin}{width=0.\linewidth, height=0.68\linewidth}
@@
 }
-\only<4->{
+\only<2->{
 %\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
@@
 }
-\visible<3->{
+\visible<1->{
 R-package \code{disttree}
@@
 \center
-\visible<5->{
+\visible<3->{
 Algorithmic \\
@@
 
-\visible<6->{
+\visible<4->{
 Data \\
@@
 
-\visible<7->{
-Combined \visible<8->{\checkmark}
+\visible<5->{
+Combined \visible<6->{\checkmark}
 }
@@
   \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
-  \emph{Manuscript under preparation}.
+  \emph{arXiv 1804.02921}, arXiv.org E-Print Archive.
+  \url{http://arxiv.org/abs/1804.02921}
 

### 2018-06-13 pkg/disttree/inst/eeecon_workshop/slides.Rnw
presentation at the eeecon workshop (2018-06-27)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2190 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dff0d574fef171aacc7d4d1fffb6f1d98f87e649

--- 2018-04-12 pkg/disttree/inst/MIP-Seminar/slides.Rnw
@@
 \title{Flexible Probabilistic Forecasting with Distributional Trees and Forests}
-\author{Lisa Schlosser}
+\author{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
 %\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
-%\URL{https://R-Forge.R-project.org/projects/partykit/}
+\URL{https://R-Forge.R-project.org/projects/partykit/}
 \headerimage{3}
@@
 \graphicspath{{plots/}}
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
 
@@
 
+# setwd("~/svn/partykit/pkg/disttree/inst/eeecon_workshop")
 library("ggplot2")
@@
 \item Prediction.
-\item Caption of dependencies and interactions.
+\item Capturing dependencies and interactions.
 \end{itemize}
@@
   Probabilistic forecast.
-  \item Regression Trees:\\
+  \item Regression trees:\\
   Automatic detection of abrupt changes,
   and selection of variables/interactions.
-  \item Random Forests:\\
+  \item Random forests:\\
   Smoothing and regularization.
@@
 
-$\Rightarrow$ PhD-Thesis
+$\Rightarrow$ Distributional trees and forests
 
@@
 
-%{\small 
-(what I've been working on the last 2 years)
 }
@@
 \visible<1->{
-R-package \code{disttree}
+R~package~\code{disttree}
 }
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\begin{frame}
+%\frametitle{Global Likelihood Estimation}
+%\begin{itemize}
+
+%\item Specify a parametric distribution family $F(\cdot;\theta)$ with parameter vector
+%  $\theta \in \Theta$ capturing location, scale, shape.
+
+%\item Log-likelihood function:
+%  \begin{eqnarray*}
+%  \ell(\theta; y) & = & \log(f(y; \theta))
+%  \end{eqnarray*}
+
+%\item Estimate $\hat{\bold{\theta}}$ via maximum likelihood
+%  based on a learning sample $y_1,\dots, y_n$:
+%\[
+%\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
+%\]
+
+%% equivalent to solving 
+%% $$
+%% \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
+%% $$
+%\end{itemize}
+
+%\end{frame}
+
+%\begin{frame}
+%\frametitle{Adaptive Local Likelihood Estimation}
+
+%\textbf{Idea:} Capture covariates through adaptive weights.
+%\[
+%\hat{\theta}(\bold{x}) = \argmax_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell%(\theta; y_i).
+%\]
+
+%\bigskip
+
+%\textbf{Question:} How to choose weighting function $w_i(\bold{x})$?
+
+%\bigskip
+
+%\textbf{Possible answers:} Based on learning sample $\{(y_i, \bold{x}_i)\}_{i=1, \dots, n}$ and (possibly new) observation $\bold{x}$.
+%\begin{itemize}
+%\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $\bold{x}_i$ are classified into the same subgroup.
+%\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $\bold{x}_i$ across trees.
+%\end{itemize}
+
+%\textbf{Possible answers:} Based on learning sample $y_1, \dots, y_n$ and (possibly new) observation $\bold{x}$.
+%\begin{itemize}
+%\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $y_i$ are classified into the same subgroup.
+%\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $y_i$ across trees.
+%\end{itemize}
+
+%\end{frame}
+
 \begin{frame}
@@
 \item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
-Estimate model parameters $\hat{\theta}$.
-\item<6-> Evaluate goodness of fit \\
-(for each parameter and each observation).
-\item<8-> Choose covariate $X$ with strongest influence on goodness of fit of $\mathcal{D}(Y; \hat{\theta})$ as split variable.
+Estimate $\hat{\theta}$ via maximum likelihood\\
+$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
+\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
 \end{enumerate}
 \end{minipage}
-\begin{minipage}{0.25\textwidth}
-\vspace{-1cm}
+\begin{minipage}{0.23\textwidth}
+%\vspace{-0.1cm}
 \begin{tikzpicture}
 \visible<2-3>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$Y$};
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
 }
 \visible<4>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$\mathcal{D}(Y;\hat{\theta}$)};
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{D}(Y;\hat{\theta}$)};
 }
 \visible<5->{
-\node[inner sep=0pt] (density1) at (0.8,2)
+\node[inner sep=0pt] (density1) at (0.8, 1.7)
     {\includegraphics[width=0.6\textwidth]{density1.jpeg}};
 }
-\visible<7-8>{
+\visible<6-8>{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
-\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1);
-\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2);
+\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1);
+\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2);
 }
@@
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
-\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2) node [midway, right] {\scriptsize $X > p$};
+\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
 }
@@
 \visible<11->{
-\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2) node [midway, right] {\scriptsize $X > p$};
+\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
 }
 \visible<12->{
-\node[inner sep=0pt] (density2) at (-0.1,-0.25)
+\node[inner sep=0pt] (density2) at (-0.1,-0.1)
     {\includegraphics[width=0.5\textwidth]{density2.jpeg}};
@@
 \visible<12->{
-\node[inner sep=0pt] (density3) at (1.7,-0.25)
+\node[inner sep=0pt] (density3) at (1.7,-0.1)
     {\includegraphics[width=0.5\textwidth]{density3.jpeg}};
@@
 \vspace{0.1cm}
-\begin{adjustwidth}{-0.0em}{-3em}
+%\begin{adjustwidth}{-0.0em}{-1em}
 \begin{enumerate}
-\setcounter{enumi}{3}
-\item<8-> Find the split point $p$ which leads to the highest improvement.
-\item<10-> Repeat steps 1--4 recursively in the subgroups
-    until some stopping criterion is met.
+\setcounter{enumi}{2}
+\item<8-> Split along the covariate $X$ with strongest association or instability and at breakpoint $p$ with highest improvement \\
+in log-likelihood.
+\item<10-> Repeat steps 1--3 recursively until some stopping criterion \\
+is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
 \end{enumerate}
-\end{adjustwidth}
+%\end{adjustwidth}
 \vspace{0.4cm}
 \visible<13->{
-{\bf Forest:} Ensemble of trees.
+{\bf Forest:} Ensemble of $T$ trees.
 \begin{itemize}
-\item Varying subsamples.
-\item Combine / average over trees.
+\item Bootstrap or subsamples.
+\item Random input variable sampling.
 \end{itemize}
 }
+\end{frame}
+
+\begin{frame}
+\frametitle{Adaptive Local Likelihood Estimation}
+
+Parameter estimator for a \only<1-3>{global}\only<4->{local} model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$}:
+
+\[
+\hat{\theta}\visible<2->{(\bold{x})} =  \argmax_{\theta \in \Theta} \sum_{i=1}^n \visible<2->{w_i(\bold{x}) \cdot} \ell(\theta; y_i)
+\]
+
+\bigskip
+
+\visible<3->{
+\textbf{Weights:}
+\begin{eqnarray*}
+w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.25cm]
+\visible<4->{
+w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.2cm]
+\visible<5->{
+w^{\text{forest}}_i(\bold{x}) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
+\end{eqnarray*}
+}}}
 \end{frame}
@@
   \url{http://arxiv.org/abs/1804.02921}
+and 
+  \emph{Working Papers, Faculty of Economics and Statistics, University of Innsbruck},
+  \url{https://EconPapers.repec.org/RePEc:inn:wpaper:2018-08}
 

### 2018-06-13 pkg/disttree/inst/eeecon_workshop/rain_pred_24to4.R
presentation at the eeecon workshop (2018-06-27)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2190 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dff0d574fef171aacc7d4d1fffb6f1d98f87e649

@@ -1,0 +1,1049 @@
+## IDEA: learn models on the first 24 years and evaluate it on the 25th, 26th, 27th and 28th year
+# average the resulting CRPS and loglikelihood over the 123 days (2011-07-19 missing) within one station
+# and repeat this procedure over all stations (with a sufficient number of years of observations)
+
+rain_pred <- function(seedconst = 7, ntree = 100,
+                      tree_minsplit = 50, tree_minbucket = 20, tree_mincrit = 0.95,
+                      forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0,
+                      forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
+                      type.tree = "ctree",
+                      gamboost_cvr = FALSE,
+                      frac = FALSE)
+{
+  
+  cl <- match.call()
+  
+  library("disttree")
+  library("gamlss")
+  library("gamlss.dist")
+  library("gamlss.cens")
+  gen.cens(NO, type = "left")
+  library("gamboostLSS")
+  library("crch")
+  library("scoringRules")
+  library("parallel")
+  
+  
+  #setwd("~/svn/partykit/pkg/disttree/inst/draft/")
+  #setwd("~/disttree/inst/draft/")
+  
+  
+  # dist_list_cens_normal
+  {
+    
+    dist_list_cens_normal <- list()
+    
+    parnames <- c("mu", "sigma")
+    etanames <- c("mu", "log(sigma)")
+    
+    ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
+      par <- c(eta[1], exp(eta[2]))
+      val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
+      if(sum) {
+        if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
+        val <- sum(weights * val, na.rm = TRUE)
+      }
+      return(val)
+    }
+    
+    
+    sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
+      par <- c(eta[1], exp(eta[2]))
+      # y[y==0] <- 1e-323
+      
+      score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
[…]

### 2018-06-13 pkg/disttree/inst/eeecon_workshop/slides.Rnw
added xyplot of example tree
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2192 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 12ee18406508c04643f82b17c5a16cd868b54f5d

@@
 
+<<treedata, echo=FALSE, results=hide>>=
+set.seed(54)
+nobs <- 500
+x <- runif(nobs, 0, 1)
+mu <- sigma <- ytrue <- numeric(length = nobs)
+for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
+for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
+y <- rnorm(nobs, mean = mu, sd = sigma)
+ytrue <- mu
+data <- data.frame(cbind(y,x, ytrue))
+
+alldata <- cbind(data, mu, sigma)
+odata <- alldata[order(alldata["x"]),]
+@
+
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\begin{center}
+\vspace*{-1.03cm}
+\setkeys{Gin}{width=0.8\linewidth}
+<<plottree_xyplot, fig=TRUE, echo=FALSE>>=
+par(mar=c(5.1,4.1,4.1,3.1))
+plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "grey")
+#plot(y=odata$y, x=odata$x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray")
+lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
+polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
+  col = pallight["forest"], border = "transparent")
+legend(x = -0.0715, y = 21.62, expression(mu  %+-%  sigma), bty = "n")
+#lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
+#lines(x = odata$x, y = odata$mu - odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
+@
+\end{center}
+\end{frame}
+
 %<<densities, eval=TRUE, echo=FALSE, results=hide>>=
@@
 
-Parameter estimator for a \only<1-3>{global}\only<4->{local} model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$}:
+\textbf{Parameter estimator} for \only<1-3>{\textbf{a global}}\only<4->{\textbf{an adaptive local}}\\ model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$} :
 
@@
 \begin{eqnarray*}
-w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.25cm]
+w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.2cm]
 \visible<4->{
-w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.2cm]
+w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.1cm]
 \visible<5->{

### 2018-06-13 pkg/disttree/inst/draft/draft.Rnw
corrected path for plot data
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2193 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 130a874c4c456ef1f02c7b5853cee9f7129642e2

@@
 data(MapTyrol, package = "RainTyrol")
-data(MapTyrol_border, package = "RainTyrol")
+# data(MapTyrol_border, package = "RainTyrol")
 # Create SpatialPointsDataFrame from station list
@@
                                            select = -c(lon,lat)),
-                             proj4string = crs(dem))
+                             proj4string = crs(MapTyrol$RasterLayer))
 @
@@
   par(mar = c(5,4,4,0.1))
-  raster::image(dem, col = rev(gray.colors(100)),
+  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
                 main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
@@
                 ylim = c(46.6, 47.87))
-  plot(tirol, add = TRUE)
+  plot(MapTyrol$SpatialPolygons, add = TRUE)
   points(sp[70,], pch = 19, col = "black", cex = 1.85)

### 2018-07-03 pkg/disttree/inst/draft/draft.Rnw
small changes based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2203 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a912077ffdb33d4856387ef53078b62d4e5c6db6

@@
 
+\numberwithin{equation}{subsection}
+
 <<preliminaries, echo=FALSE, results=hide>>=
@@
 instead there is an increasing interest in probabilistic forecasts. Quantities of
-interest typically include exceedence probabilities for certain thresholds of the
+interest typically include exceedance probabilities for certain thresholds of the
 response or quantiles of the response distribution. Specifically, consider
@@
 is typically more relevant or, in some situations, a prediction interval of
-expected precipitation (say from the expected 10\% to 90\% quantile). Similar
+expected precipitation (say from the expected \textit{10\%} to \textit{90\%} quantile). Similar
 considerations apply for other meteorological quantities and hence attention
@@
 
+# switch x-axis back to untransformed scale
+x <- x^(1.6)
+
 # point mass (slightly shifted)
-pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
-pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
-pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
-pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
+pm1 <- c(0.15, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.05
+pm2 <- c(0.05, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.01
+pm3 <- c(-0.05, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.03
+pm4 <- c(-0.15, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.07
 
 # predictions
-pred1 <- c(Axams_24to4$testdata[pdays,"robs"][1], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(Axams_24to4$testdata[pdays,"robs"][2], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(Axams_24to4$testdata[pdays,"robs"][3], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(Axams_24to4$testdata[pdays,"robs"][4], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+pred1 <- c(Axams_24to4$testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(Axams_24to4$testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(Axams_24to4$testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(Axams_24to4$testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
 
@@
      main = paste0("July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
+     xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
+     # xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
      ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-1.5,8))
+     xlim = c(-2,28))     #  xlim = c(-1.5,8))
 
@@
 # add labels
-text(x = -0.8, y = lh1, labels = "2009", col = pal[1], cex = 0.8)
-text(x = -0.8, y = lh2, labels = "2010", col = pal[3], cex = 0.8)
-text(x = -0.8, y = lh3, labels = "2011", col = pal[2], cex = 0.8)
-text(x = -0.8, y = lh4, labels = "2012", col = pal[4], cex = 0.8)
+text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
+text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
+text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
+text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
 @
@@
 likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
-$$ \label{eq:mle}
+\begin{equation}
+\label{eq:mle}
 \bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
-$$
+\end{equation}
 Equivalently, this can be defined based on the corresponding first-order
 conditions
-$$ \label{eq:foc}
+\begin{equation}
+\label{eq:foc}
 \sum_{i = 1}^n s(\bm{\hat{\theta}}, y_i) = 0,
-$$
+\end{equation}
 where $s(\bm{\theta}; y_i)$ is the associated score function
-$$ \label{eq:score}
+\begin{equation}
+\label{eq:score}
 s(\bm{\theta}; y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
-$$ 
+\end{equation}
 The latter is subsequently employed as a general goodness-of-fit measure
@@
 ($b = 1, \dots, B$) partition the covariate space
-$$ \label{eq:partition}
+\begin{equation}
+\label{eq:partition}
 \mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
-$$
+\end{equation}
 and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
@@
 \begin{minipage}{\textwidth}
+\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
+Time periods indicate aggregation time periods in hours after NWP model
+initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
+0600\,UTC to 0600\,UTC of the next day). $^*$Minimum values of \emph{dswrf}
+over 24\,h are always zero and thus neglected.}
+\label{tab:covariates}
 \begin{tabular}{ l  c  l }
@@
 \end{tabular}
-\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
-Time periods indicate aggregation time periods in hours after NWP model
-initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
-0600\,UTC to 0600\,UTC of the next day). $^*$Minimum values of \emph{dswrf}
-over 24\,h are always zero and thus neglected.}
-\label{tab:covariates}
 \end{minipage} 
@@
 \centering
+\caption[Table caption text]{Overview of models with type of 
+covariate dependency and included covariates for each distribution 
+parameter. \emph{A} $\ast$ \emph{B} indicates an interaction between
+covariate \emph{A} and \emph{B}.}
+\label{tab:models}
 \begin{tabular}{ l l l l }
@@
 \end{tabular}
-\caption[Table caption text]{Overview of models with type of 
-covariate dependency and included covariates for each distribution 
-parameter. \emph{A} $\ast$ \emph{B} indicates an interaction between
-covariate \emph{A} and \emph{B}.}
-\label{tab:models}
 \end{table}
@@
   in Section~\ref{sec:methods}.
+  % stopping criteria:
+  % tree_minsplit = 50, tree_minbucket = 20, tree_mincrit = 0.95,
+  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0
 

### 2018-07-05 pkg/disttree/inst/draft/draft.Rnw
corrections of the notation in the appendix based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2206 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 620244dc3a000fb987c43694ba665c7163b6e5f2

@@
 \documentclass[nojss]{jss}
+\shortcites{Hamill+Bates+Whitaker:2013}
+
+%\VignetteIndexEntry{disttree: Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
+%\VignetteDepends{disttree, crch, countreg, gamlss, gamlss.cens}
+%\VignetteKeywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
+%\VignettePackage{disttree}
+
+%% packages
+\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}
+
+%% tikz
+\usepackage{array,makecell,tikz}
+\usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
+
+%% need no \usepackage{Sweave}
+\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
+
+%% commands
+\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
+\renewcommand{\Prob}{\mathbb{P} }
+\renewcommand{\E}{\mathbb{E}}
+\newcommand{\V}{\mathbb{V}}
+\newcommand{\Var}{\mathbb{V}}
+\newcommand{\R}{\mathbb{R} }
+\newcommand{\N}{\mathbb{N} }
+\newcommand{\C}{\mathbb{C} }
+\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+\newcommand{\LS}{\mathcal{L}_n}
+\newcommand{\TS}{\mathcal{T}_n}
+\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
+\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
+\newcommand{\F}{\mathcal{F}}
+\newcommand{\A}{\mathcal{A}}
+\newcommand{\yn}{y_{\text{new}}}
+\newcommand{\z}{\mathbf{z}}
+\newcommand{\X}{\mathbf{X}}
+\newcommand{\Y}{\mathbf{Y}}
+\newcommand{\sX}{\mathcal{X}}
+\newcommand{\sY}{\mathcal{Y}}
+\newcommand{\T}{\mathbf{T}}
+\newcommand{\x}{\mathbf{x}}
+\renewcommand{\a}{\mathbf{a}}
+\newcommand{\xn}{\mathbf{x}_{\text{new}}}
+\newcommand{\y}{\mathbf{y}}
+\newcommand{\w}{\mathbf{w}}
+\newcommand{\ws}{\mathbf{w}_\cdot}
+\renewcommand{\t}{\mathbf{t}}
+\newcommand{\M}{\mathbf{M}}
+\renewcommand{\vec}{\text{vec}}
+\newcommand{\B}{\mathbf{B}}
+\newcommand{\K}{\mathbf{K}}
+\newcommand{\W}{\mathbf{W}}
+\newcommand{\D}{\mathbf{D}}
+\newcommand{\I}{\mathbf{I}}\documentclass[nojss]{jss}
 \shortcites{Hamill+Bates+Whitaker:2013}
@@
 \newcommand{\I}{\mathbf{I}}
+\newcommand{\bS}{\mathbf{S}}
+\newcommand{\cellx}{\pi_n[\x]}
+\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
+\newcommand{\err}{\text{Err}}
+\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
+\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
+\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
+\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
+\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
+\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
+\newcommand{\bft}{\mathbf{t}}
+
+<<preliminaries, echo=FALSE, results=hide>>=
+options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
+library("disttree")
+library("crch")
+library("countreg")
+library("gamlss")
+library("gamlss.cens")
+gen.cens(NO, type = "left")
+library("RainTyrol")
+
+## HCL palette
+pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
+pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
+transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
+@
+
+\title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
+\Shorttitle{Distributional Regression Forests for Probabilistic Precipitation Forecasting}
+
+\author{Lisa Schlosser\\Universit\"at Innsbruck
+   \And Torsten Hothorn\\Universit\"at Z\"urich
+   \And Reto Stauffer\\Universit\"at Innsbruck
+   \And Achim Zeileis\\Universit\"at Innsbruck}
+\Plainauthor{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
+
+\Abstract{
+To obtain a probabilistic model for a dependent variable based on some set of
+explanatory variables, a distributional approach is often adopted where the
+parameters of the distribution are linked to regressors. In many classical
+models this only captures the location of the distribution but
+over the last decade there has been increasing interest in distributional
+regression approaches modeling all parameters including location, scale, and
+shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
+mean and variance of a Gaussian response and is particularly
+popular in weather forecasting. More generally, the GAMLSS framework
+allows to establish generalized additive models for location, scale, and shape
+with smooth linear or nonlinear effects.
+%
+However, when variable selection is required and/or there are
+non-smooth dependencies or interactions (especially unknown or of high-order),
+it is challenging to establish a good GAMLSS. A natural alternative in these
+situations would be the application of regression trees or random forests but,
+so far, no general distributional framework is available for these. Therefore,
+a framework for distributional regression trees and forests is proposed that blends
+regression trees and random forests with classical distributions from the GAMLSS
+framework as well as their censored or truncated counterparts.
+%
+To illustrate these novel approaches in practice, they are employed to obtain
+probabilistic precipitation forecasts at numerous sites in a mountainous
+region (Tyrol, Austria) based on a large number of numerical weather
+prediction quantities. It is shown that the novel distributional regression forests
+automatically select variables and interactions, performing on par or often even
+better than GAMLSS specified either through prior meteorological knowledge or a
+computationally more demanding boosting approach.
+}
+\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
+
+\Address{
+  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
+  Universit\"at Innsbruck \\
+  Department of Statistics \\
+  Faculty of Economics and Statistics \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
+  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
+  \email{Achim.Zeileis@R-project.org} \\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
+  \phantom{URL: }\url{https://retostauffer.org/},\\
+  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
+
+  Torsten Hothorn\\
+  Universit\"at Z\"urich \\
+  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
+  Hirschengraben 84\\
+  CH-8001 Z\"urich, Switzerland \\
+  E-mail: \email{Torsten.Hothorn@R-project.org}\\
+  URL: \url{http://user.math.uzh.ch/hothorn/}\\
+
+}
+
+\begin{document}
+
+\section{Introduction}
+\label{sec:introduction}
+
+In regression analysis a wide range of models has been developed to describe the
+relationship between a response variable and a set of covariates. The classical
+model is the linear model (LM) where the conditional mean of the response
+is modeled through a linear function of the covariates (see the left panel of
+Figure~\ref{fig:gam} for a schematic illustration). Over the last decades
+this has been extended in various directions including:
+
+\pagebreak
+
+\begin{itemize}
+  \item \emph{Generalized linear models} (GLMs, \citealp{Nelder+Wedderburn:1972})
+    encompassing an additional nonlinear link function for the conditional mean.
+  \item \emph{Generalized additive models} (GAMs, \citealp{Hastie+Tibshirani:1986})
+    allowing for smooth nonlinear effects in the covariates
+    (Figure~\ref{fig:gam}, middle).
+  \item \emph{Generalized additive models for location, scale, and shape}
+    (GAMLSS, \citealt{Rigby+Stasinopoulos:2005}) adopting a probabilistic modeling
+    approach. In GAMLSS, each parameter of a statistical distribution can depend
+    on an additive predictor of the covariates comprising linear and/or
+    smooth nonlinear terms (Figure~\ref{fig:gam}, right).
+\end{itemize}
+Thus, the above-mentioned models provide a broad toolbox for capturing different
+aspects of the response (mean only vs.\ full distribution) and different types
+of dependencies on the covariates (linear vs.\ nonlinear additive terms).
+
+\begin{figure}[t!]
+\begin{tikzpicture}
+\node (a) at (0,0){
+\minipage{0.29\textwidth}
+\centering
+<<motivation_GLM, echo=FALSE, results=hide>>=
+nobs <- 200
+## GLM
+set.seed(7)
+x <- c(1:nobs)/nobs
+ytrue <- 1+x*1.5
+y <- ytrue + rnorm(nobs,0,0.3)
+@
+
+<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytrue, col = pal[1], lwd = 7, main = "")
+@
+\endminipage};
+\node (b) at (5,0) 
+{
+\minipage{0.29\textwidth}
+\centering
+<<motivation_GAM, echo=FALSE, results=hide>>=
+## GAM
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3 
+y <- ytrue + rnorm(nobs,0,0.3)
+@
+
+<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytrue, col = pal[1], lwd = 7, main = "")
+@
+\endminipage};
+\node (c) at (10,0) 
+{
+\minipage{0.29\textwidth}
+\centering
+<<motivation_GAMLSS, echo=FALSE, results=hide>>=
+## GAMLSS
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3
+var <- exp(-(2*x)^2)/2
+y <- ytrue + rnorm(nobs, 0, 0.1 + var)
+@
+
+<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
+  col = pallight[1], border = "transparent")
+lines(x, ytrue, col = pal[1], lwd=7)
+points(x, y, col = "slategray", pch = 19)
+box(lwd = 5)
+@
+\endminipage};
+\node (d) at (0,-2) 
+{
+\minipage{0.29\textwidth}
+\centering
+\vspace{0.0cm}
+LM, GLM  
+\endminipage};
+\node (e) at (5,-2) 
+{
+\minipage{0.29\textwidth}
+\centering
+\vspace{0.0cm}
+GAM 
+\endminipage};
+\node at (10,-2)
+{
+\minipage{0.28\textwidth}
+\centering
+\vspace{0.0cm}
+GAMLSS 
+\endminipage};
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
+\end{tikzpicture}
+\caption{\label{fig:gam}Parametric modeling developments. (Generalized) linear models (left),
+generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
+\end{figure}
+
+While in many applications conditional mean regression models have been receiving
+the most attention, there has been a paradigm shift over the last decade towards
+distributional regression models. An important reason for this is that in many
+fields forecasts of the mean are not the only (or not even the main) concern but
+instead there is an increasing interest in probabilistic forecasts. Quantities of
+interest typically include exceedence probabilities for certain thresholds of the
+response or quantiles of the response distribution. Specifically, consider
+weather forecasting where there is less interest in the mean amount of
+precipitation on the next day. Instead, the probability of rain vs.\ no rain
+is typically more relevant or, in some situations, a prediction interval of
+expected precipitation (say from the expected 10\% to 90\% quantile). Similar
+considerations apply for other meteorological quantities and hence attention
+in the weather forecasting literature has been shifting from classical linear 
+deterministic models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
+non-homogeneous Gaussian regression (NGR) of \cite{Gneiting+Raftery+Westveld:2005}.
+The NGR typically describes the mean of some meteorological response variable through
+the average of the corresponding quantity from an ensemble of physically-based
+numerical weather predictions (NWPs). Similarly, the variance of the response
+is captured through the variance of the ensemble of NWPs. Thus, the NGR
+considers both the mean as well as the uncertainty of the ensemble predictions
+to obtain probabilistic forecasts calibrated to a particular site.
+
+\begin{figure}[t!]
+\begin{tikzpicture}
+\node (a) at (0,0){
+\minipage{0.29\textwidth}
+\centering
+<<motivation_regtree, echo=FALSE, results=hide>>=
+## Reg. Tree
+set.seed(7)
+kappa <- 12
+x <- c(1:nobs)/nobs
+ytrue <- ytree <- yforest <- numeric(length = length(x))
+for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
+y <- ytrue + rnorm(nobs,0,0.3)
+for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
+@
+
+<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
+lines(x = x, y = ytree, col = pal[1], lwd=7)
+@
+\endminipage};
+\node (b) at (5,0) 
+{
+\minipage{0.29\textwidth}
+\centering
+<<motivation_randforest, echo=FALSE, results=hide>>=
+## Random Forest
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
+@
+
+<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
+lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
+@
+\endminipage};
+\node (c) at (10,0) 
+{
+\minipage{0.29\textwidth}
+\centering
+<<motivation_distforest, echo=FALSE, results=hide>>=
+## distforest
+set.seed(4723)
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
+xt <- 2*(x-0.5)
+var <- exp(-(2*xt)^4)/2
+y <- ytrue + rnorm(nobs, 0, 0.1 + var)
+x <- c(1:nobs)/nobs
+@
+
+<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
+par(mar = c(2, 0, 2, 0))
+plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+usr <- par("usr")
+rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
+polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
+  col = pallight[1], border = "transparent")
+lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
+points(x, y, col = "slategray", pch = 19)
+box(lwd=8)
+@
+\endminipage};
+\node (d) at (0,-2) 
+{
+\minipage{0.29\textwidth}
+\centering
+\vspace{0.0cm}
+Regression tree 
+\endminipage};
+\node (e) at (5,-2) 
+{
+\minipage{0.29\textwidth}
+\centering
+\vspace{0.0cm}
+Random forest 
+\endminipage};
+\node at (10,-2)
+{
+\minipage{0.28\textwidth}
+\centering
+\vspace{0.0cm}
+Distributional forest 
+\endminipage};
+\node (g) at (0,-5.4) 
+{
+\minipage{0.29\textwidth}
+\vspace{0.2cm}
+%\hspace{0.0cm}
+\centering
+<<motivation_disttree, echo=FALSE, results=hide>>=
+## disttree
+set.seed(74)
+x <- c(1:nobs)/nobs
+var <- ytree/3
+y <- ytrue + rnorm(nobs,0,var)
+@
+
+<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
+par(mar = c(2, 0, 2, 0))
+plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
+usr <- par("usr")
+rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
+lines(x = x, y = ytree, col = pal[1], lwd=7, main = "")
+polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
+  col = pallight[1], border = "transparent")
+points(x, y, col = "slategray", pch = 19)
+box(lwd=8)
+@
+\vspace{-0.2cm}%\\
+Distributional tree
+\vspace{0.2cm}
+\endminipage};
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](0,-2.4)--(0,-3.2);
+\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.2,-5.4)--(7.9,-2.5);
+\end{tikzpicture}
+\caption{\label{fig:tree}Tree and forest developments.
+Regression tree (top left), distributional tree (bottom left), random forest
+(top middle), and distributional forest (top right).}
+\end{figure}
+
+In summary, the models discussed so far provide a broad and powerful toolset for
+parametric distributional fits depending on a specified set of additive linear or
+smooth nonlinear terms. A rather different approach to capturing the dependence on
+covariates are tree-based models.
+\begin{itemize}
+  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
+    split the data into more homogeneous subgroups and can thus capture abrupt shifts
+    (Figure~\ref{fig:tree}, top left) and approximate nonlinear functions.
+    Furthermore, trees automatically carry out a forward selection of covariates and
+    their interactions. 
+  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
+    an ensemble of trees fitted to resampled versions of the learning data. This
+    stabilizes the recursive partitions from individual trees and hence better
+    approximates smooth functions (Figure~\ref{fig:tree}, top middle).
+\end{itemize}
+While classical regression trees and random forests only model the mean of the response
+we propose to follow the ideas from GAMLSS modeling -- as outlined
+in Figure~\ref{fig:gam} -- and combine tree-based methods with parametric distributional models,
+yielding two novel techniques:
+\begin{itemize}
+  \item \emph{Distributional regression trees} (for short: \emph{distributional trees}) split the data into more homogeneous groups with
+    respect to a parametric distribution, thus capturing changes in any distribution
+    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left).
+  \item \emph{Distributional regression forests} (for short: \emph{distributional forests}) utilize an ensemble of distributional trees
+    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
+    top right).
+\end{itemize}
+In the following, particular focus is given to distributional forests as a method for obtaining
+probabilistic forecasts by leveraging the strengths of random forests: the ability to
+capture both smooth and abruptly changing functions along with simultaneous selection of
+variables and possibly complex interactions. Thus, these properties make the method particularly
+appealing in case of many covariates with unknown effects and interactions where it would be 
+challenging to specify a distributional regression model like GAMLSS.
+
+<<echo=FALSE, results=hide>>=
+#### Axams prediction 24 - 4
+if(file.exists("Axams_24to4.rda")){
+  load("Axams_24to4.rda")
+} else {
+  source("Axams_24to4.R")
+  Axams_24to4 <- Axams_24to4()
+  save(Axams_24to4, file = "Axams_24to4.rda")
+}
+
+#### prepare data for plot of estimated density functions
+# predictions for one day (in each of the four years) 
+# (19th of July 2011 is missing)
+pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
+pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata[pdays,], type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+cbind(df_exp, Axams_24to4$testdata[pdays,"robs"])
+
+# plot predicted distributions together with observations
+set.seed(7)
+x <- c(0.01, sort(runif(500,0.01,8)))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
+dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
+
+# point mass (slightly shifted)
+pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
+pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
+pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
+pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
+
+# predictions
+pred1 <- c(Axams_24to4$testdata[pdays,"robs"][1], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(Axams_24to4$testdata[pdays,"robs"][2], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(Axams_24to4$testdata[pdays,"robs"][3], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(Axams_24to4$testdata[pdays,"robs"][4], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+
+#legendheight
+lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
+lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
+lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
+
+############
+# get predicted parameter of all models for testdata
+
+# distributional tree
+pdt <- predict(Axams_24to4$dt, newdata = Axams_24to4$testdata, type = "parameter")
+dt_mu <- pdt$mu
+dt_sigma <- pdt$sigma
+dt_exp <- pnorm(dt_mu/dt_sigma) * (dt_mu + dt_sigma * (dnorm(dt_mu/dt_sigma) / pnorm(dt_mu/dt_sigma)))
+
+# distributional forest
+pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata, type = "parameter")
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
+
+# prespecified GAM
+g_mu <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "mu", type = "response", data = Axams_24to4$g_learndata)
+g_sigma <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "sigma", type = "response", data = Axams_24to4$g_learndata)
+g_exp <- pnorm(g_mu/g_sigma) * (g_mu + g_sigma * (dnorm(g_mu/g_sigma) / pnorm(g_mu/g_sigma)))
+
+# boosted GAM
+pgb <- predict(Axams_24to4$gb, newdata = Axams_24to4$testdata, parameter = c("mu","sigma"), type = "response")
+gb_mu <- pgb$mu
+gb_sigma <- pgb$sigma
+gb_exp <- pnorm(gb_mu/gb_sigma) * (gb_mu + gb_sigma * (dnorm(gb_mu/gb_sigma) / pnorm(gb_mu/gb_sigma)))
+
+# EMOS
+ml_mu <- predict(Axams_24to4$ml, type = "location", newdata = Axams_24to4$testdata)
+ml_sigma <- predict(Axams_24to4$ml, type = "scale", newdata = Axams_24to4$testdata)
+ml_exp <- pnorm(ml_mu/ml_sigma) * (ml_mu + ml_sigma * (dnorm(ml_mu/ml_sigma) / pnorm(ml_mu/ml_sigma)))
+
+############
+## PIT histograms and qqrplots
+
+## get predicted parameter
+# for learndata and testdata
+{
+  set.seed(7)
+
+  # distributional tree
+  pit_dt <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = dt_mu, sd = dt_sigma))
+  pit_dt[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_dt[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # distributional forest
+  pit_df <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = df_mu, sd = df_sigma))
+  pit_df[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_df[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # prespecified GAM
+  pit_g <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = g_mu, sd = g_sigma))
+  pit_g[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_g[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # boosted GAM
+  pit_gb <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = gb_mu, sd = gb_sigma))
+  pit_gb[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_gb[Axams_24to4$testdata[,"robs"]>0, 2]
+
+  # EMOS
+  pit_ml <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = ml_mu, sd = ml_sigma))
+  pit_ml[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_ml[Axams_24to4$testdata[,"robs"]>0, 2]
+
+}
+
+## Variable importance
+if(file.exists("vimp_crps.rda")){
+  load("vimp_crps.rda")
+} else {
+
+  set.seed(7)
+  # locally redefine logLik.distforest to use crps in varimp()
+  logLik.distforest <- function(object, newdata = NULL, ...) {
+    if(is.null(newdata)) {
+      newdata <- object$data
+    }
+
+    # predict parameter
+    pdf <- predict(object, newdata = newdata, type = "parameter")
+    df_mu <- pdf$mu
+    df_sigma <- pdf$sigma
+
+    # calculate CRPS
+    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
+
+    return(structure(crps, df = NA, class = "logLik"))
+  }
+
+  vimp_crps <- varimp(Axams_24to4$df, nperm = 1L)
+
+  save(vimp_crps, file = "vimp_crps.rda")
+
+  rm(logLik.distforest) 
+}
+
+@
+
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.75\textwidth}
+<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=3.8>>=
+par(mar = c(3.8, 4, 2.3, 1.5))
+plot(x = x, y = y1, type = "l", col = pal[1], lwd = 1.3, 
+     main = paste0("July ", pday), ylab = "Density", 
+     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
+     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
+     xlim = c(-1.5,8))
+
+lines(x = x, y = y2, type = "l", col = pal[3], lwd = 1.3)
+lines(x = x, y = y3, type = "l", col = pal[2], lwd = 1.3)
+lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
+legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
+       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)
+
+# plot point mass
+lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = pal[1], type = "l", lwd = 1.3)
+lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = pal[3], type = "l", lwd = 1.3)
+lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = pal[2], type = "l", lwd = 1.3)
+lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = pal[4], type = "l", lwd = 1.3)
+
+points(x = pm1[1], y = pm1[2], col = pal[1], pch = 19)
+points(x = pm2[1], y = pm2[2], col = pal[3], pch = 19)
+points(x = pm3[1], y = pm3[2], col = pal[2], pch = 19)
+points(x = pm4[1], y = pm4[2], col = pal[4], pch = 19)
+
+# plot predictions
+points(x = pred1[1], y = pred1[2], col = pal[1], pch = 4)
+points(x = pred2[1], y = pred2[2], col = pal[3], pch = 4)
+points(x = pred3[1], y = pred3[2], col = pal[2], pch = 4)
+points(x = pred4[1], y = pred4[2], col = pal[4], pch = 4)
+
+lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+
+# add labels
+text(x = -0.8, y = lh1, labels = "2009", col = pal[1], cex = 0.8)
+text(x = -0.8, y = lh2, labels = "2010", col = pal[3], cex = 0.8)
+text(x = -0.8, y = lh3, labels = "2011", col = pal[2], cex = 0.8)
+text(x = -0.8, y = lh4, labels = "2012", col = pal[4], cex = 0.8)
+@
+\caption{\label{fig:axams-dist}Total precipitation predictions by a distributional forest
+at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
+Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
+\end{figure}
+
+In weather forecasting, these properties are especially appealing in 
+mountainous regions and complex terrain where a wide range of local-scale
+effects are not yet resolved by the NWP models. 
+Thus, effects with abrupt changes and possibly nonlinear interactions 
+might be required to account for site-specific unresolved features.
+To illustrate this in practice,
+precipitation forecasts are obtained with distributional forests at 95
+meteorological stations in a mountainous region in the Alps, covering mainly
+Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
+More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
+zero-censored point mass describes the probability of observing no precipitation 
+on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
+established based on data from the same month over the years 1985--2012 including
+80~covariates derived from a wide range of different NWP quantities. As
+Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
+distributional forecast for each day -- here for one specific day (July~24) at one
+station (Axams) over four years (2009--2012) -- based on the previous 24~years
+as learning data. The corresponding observations conform reasonably well with
+the predictions. In Section~\ref{sec:precipitation} we investigate the
+performance of distributional forests in this forecasting task in more detail.
+Compared to three alternative zero-censored Gaussian models distributional forests
+perform at least on par and sometimes clearly better while requiring no expert
+knowledge for the model specification. The three alternatives are: a standard ensemble model
+output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
+on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
+\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
+\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
+\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
+80~available regressors.
+
+\section{Methodology}
+\label{sec:methods}
+
+To embed the distributional approach from GAMLSS into regression trees and random
+forests, we proceed in three steps. (1)~To fix notation, we briefly review fitting
+distributions using standard maximum likelihood in Section~\ref{sec:distfit}.
+(2)~A recursive partitioning strategy based on the corresponding scores (or
+gradients) is introduced in Section~\ref{sec:disttree}, leading to distributional
+trees. (3)~Ensembles of distributional trees fitted to randomized subsamples are
+employed to establish distributional forests in Section~\ref{sec:distforest}.
+
+The general distributional notation is exemplified in all three steps for the
+Gaussian distribution, left-censored at zero. The latter is employed in the empirical 
+case study in Section~\ref{sec:precipitation} to model power-transformed daily 
+precipitation amounts.
+
+\subsection{Distributional fit}
+\label{sec:distfit}
+
+A distributional model $\mathcal{D}(Y, \bm{\theta})$ is considered for the response
+variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
+\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$ and
+corresponding log-likelihood function $\ell(\bm{\theta}; Y)$. The GAMLSS
+framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
+distributional families with parameterizations corresponding to location, scale,
+and shape. Furthermore, censoring and/or truncation of these distributions
+can be incorporated in the usual straightforward way
+\citep[see e.g.,][Chapter~7.2]{Long:1997}.
+
+To capture both location and scale of the probabilistic precipitation forecasts
+while accounting for a point mass at zero (i.e., dry days without
+rain), a Gaussian distribution left-censored at zero (for short: zero-censored Gaussian)
+with location parameter $\mu$ and scale parameter $\sigma$ is employed. Therefore,
+the corresponding log-likelihood function with parameter vector 
+$\bm{\theta} = (\mu, \sigma)$ is
+$$
+\ell(\mu, \sigma; Y) = 
+\begin{cases}
+    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
+    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
+\end{cases}
+$$
+
+where $\phi$ and $\Phi$ are the probability density function and the 
+distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
+Other distributions $\mathcal{D}$ and corresponding log-likelihoods
+$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
+censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
+zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.
+
+With the specification of the distribution family and its log-likelihood 
+function the task of fitting a distributional model turns into the task 
+of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
+maximum likelihood (ML) based on the learning sample with observations
+$\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
+likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
+$$ \label{eq:mle}
+\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
+$$
+Equivalently, this can be defined based on the corresponding first-order
+conditions
+$$ \label{eq:foc}
+\sum_{i = 1}^n s(\bm{\hat{\theta}}, y_i) = 0,
+$$
+where $s(\bm{\theta}; y_i)$ is the associated score function
+$$ \label{eq:score}
+s(\bm{\theta}; y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
+$$ 
+The latter is subsequently employed as a general goodness-of-fit measure
+to assess how well the distribution with parameters $\bm{\theta}$ fits
+one individual observation $y_i$.
+
+\subsection{Distributional tree}
+\label{sec:disttree}
+
+Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
+for reasonably representing the response distribution and covariates
+$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
+employed to capture differences in the distribution parameters $\bm{\theta}$.
+In weather forecasting, these covariates typically include the output
+from numerical weather prediction systems and/or lagged meteorological observations.
+
+To incorporate the covariates into the distributional model, 
+they are considered as regressors in additive predictors
+$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
+in GAMLSS.
+Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
+($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
+nonlinear effects, spatial effects, random coefficients, or interaction surfaces
+\citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
+additive terms and their functional forms in advance which can be challenging in
+practice and potentially require domain knowledge, especially if the number of covariates $m$ is large.
+
+Regression trees generally take a different approach for automatically
+including covariates in a data-driven way and allowing for abrupt
+changes, nonlinear and non-additive effects, and interactions. In the context of
+distributional models the goal is to partition the covariate space
+$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
+distributional model for the response $Y$ can be found with segment-specific
+parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
+($b = 1, \dots, B$) partition the covariate space
+$$ \label{eq:partition}
+\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
+$$
+and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
+(i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
+response $Y$ in each segment.
+
+To find the segments $\mathcal{B}_b$ that are (approximately) homogenous
+with respect to the distributional model with given parameters, the idea
+is to use a gradient-based recursive-partitioning approach. In a given
+subsample of the learning data this fits the model by ML (see
+Equation~\ref{eq:mle}) and then assesses the goodness of fit by assessing the
+corresponding scores $s(\bm{\hat \theta}; y_i)$ (see Equation~\ref{eq:score}).
+
+To sum up, distributional trees are fitted recursively via:
+\begin{enumerate}
+\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
+  in the current subsample.
+\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
+  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
+\item Split the sample along the partitioning variable $Z_l^*$ with the
+  strongest association or instability. Choose the breakpoint with the highest
+  improvement in the log-likelihood or the highest discrepancy.
+\item Repeat steps 1--3 recursively in the subsamples until these become too
+  small or there is no significant association/instability (or some other
+  stopping criterion is reached).
+\end{enumerate}
+%
+Different inference techniques can be used for assessing the association between
+scores and covariates in step~3. In the following we use the general class of
+permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
+also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
+Alternatively, one could use asymptotic M-fluctuation tests for parameter
+instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
+\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
+Appendix~\ref{app:tree}.
+
+For obtaining probabilistic predictions from the tree for a (possibly new) set
+of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
+be ``sent down'' the tree and the corresponding segment-specific MLE has to be
+obtained. This can also be understood as a weighted MLE where the weights
+select those observations from the learning sample that fall into the same
+segment:
+$$
+w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
+$$
+where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
+for a given $\bold{z}$ is then fully specified by the estimated parameter
+$\bm{\hat \theta}(\bold{z})$ where
+$$
+\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
+$$
+
+\subsection{Distributional forest}
+\label{sec:distforest}
+
+While the simple recursive structure of a tree model is easy to visualize and
+interpret, the abrupt changes are often too rough, instable, and impose steps
+on the model even if the true underlying effect is smooth. Hence, ensemble
+methods such as bagging or random forests \citep{Breiman:2001} are typically
+applied to smooth the effects, stabilize the model, and improve predictive performance.
+
+While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
+that pick up changes in the location of the response across the covariates,
+\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. These
+pick up changes in the ``direction'' of any distribution parameter by
+considering the full score vector for choosing splitting variables and split
+points. Each of the distributional trees is grown on a different data set
+obtained through bootstrap sampling (or subsampling) and in each node only a
+random subset of the covariates $\bold{Z}$ is considered. As usual in random
+forests, this reduces the correlation among the trees and stabilizes the
+variance of the model.
+
+To obtain probabilistic predictions from a distributional forest, it still
+needs to be specified how to compute the parameter estimates
+$\bm{\hat \theta}(\bold{z})$ for a (potentially new) set of covariates $\bold{z}$.
+Following \cite{Hothorn+Zeileis:2017} we interpret random forests as adaptive
+local likelihood estimators using the averaged ``nearest neighbor weights''
+\citep{Lin+Jeon:2006} from the $T$~trees in the forest
+$$
+w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
+\frac{\mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
+$$
+where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
+segment of the $t$-th tree.
+Thus, these $w^{\text{forest}}_i(\bold{z}) \in [0, 1]$ whereas
+$w^{\text{tree}}_i(\bold{z}) \in \{0, 1\}$. Hence, weights cannot only be $0$
+or $1$ but change more smoothly, giving high weight to those observations $i$
+from the learning sample that co-occur in the same segment $\mathcal{B}_b^t$
+as the new observation $\bold{z}$ for many of the trees $t = 1, \dots, T$.
+Consequently, the parameter estimates may, in principle, change for every
+observation and can be obtained by
+$$
+\bm{\hat \theta}(\bold{z}) =  \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
+$$
+In summary, this yields a parametric distributional regression model 
+(through the score-based approach) that can capture both abrupt effects and
+high-order interactions (through the trees) and smooth effects (through
+the forest).
+
+Distributional forests share some concepts and algorithmic aspects 
+with other generalizations of Breiman and Cutler's random forests. Nearest 
+neighbor weights are employed for aggregation in survival forests 
+\citep{Hothorn+Lausen+Benner:2004},
+quantile regression forests \citep{Meinshausen:2006},
+transformation forests \citep{Hothorn+Zeileis:2017},
+and generalized random forests for causal inferences 
+\citep{Athey+Tibshirani+Wager:2017}.
+These procedures aggregate over trees fitted to specific score functions 
+(e.g., log rank scores in survival trees, model residuals in transformation 
+or generalized forests). Distributional forests, in contrast to these 
+nonparametric approaches, provide a compromise between model flexibility 
+and interpretability: The parameters of a problem-specific distribution 
+(zero-censored Gaussian for precipitation) have a clear meaning but 
+may depend on external variables in a quite general way.
+
+\section{Probabilistic precipitation forecasting in complex terrain}
+\label{sec:precipitation}
+
+Many statistical weather forecasting models leverage the strengths of modern
+numerical weather prediction (NWP) systems
+\citep[see][]{Bauer+Thorpe+Brunet:2015}.  
+One frequently used method based on
+distributional regression models is the ensemble model output statistics 
+(EMOS) approach first proposed by
+\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
+specific quantities and sites.  
+In case of precipitation forecasting EMOS typically uses the ensemble mean of 
+``total precipitation'' (\emph{tp}) forecasts as predictor for the location parameter and
+the corresponding ensemble standard deviation for the scale part of the statistical model
+to correct for possible errors of the ensemble in both, the expectation but also
+the uncertainty of a specific forecast.
+
+While this approach alone is already highly effective in the plains,
+it typically does not perform as well in complex terrain due to unresolved
+effects in the NWP system. For example, in the Tyrolean Alps -- considered
+in the following case study -- the NWP grid cells of $50 \times 50$
+km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
+Therefore, it is often possible to substantially improve the predictive 
+performance of the EMOS by including additional predictor variables, either 
+from local meteorological observations or an NWP model. Unfortunately, it is typically unknown 
+which variables are relevant for improving the predictions. Simply including
+all available variables may be computationally burdensome and
+can lead to overfitting but, on the other hand, excluding too many variables
+may result in a loss of valuable information. Therefore, selecting
+the relevant variables and interactions among all possible covariates is 
+crucial for improving the statistical forecasting model.
+
+In the following, it is illustrated how distributional forests can
+solve this problem without requiring prior expert knowledge.  For fitting the
+forest only the response distribution and the list of potential predictor
+variables need to be specified (along with a few algorithmic details) and then
+the relevant variables, interactions, and potentially nonlinear effects are
+determined automatically in a data-driven way. Here, we employ a zero-censored
+Gaussian distribution and 80~predictor variables computed from ensemble means
+and spreads of various NWP outputs. The predictive performance of the forest is
+compared to three other zero-censored Gaussian models: (a)~a standard basic
+EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
+meteorological knowledge/experience, and (c)~a boosted GAMLSS with automatic
+selection of smooth additive terms based on all 80~predictor variables.
+
+\subsection{Data}
+\label{sec:data}
+
+\begin{table}[t!]
+%\resizebox{1.4\textwidth}{!}{
+\begin{minipage}{\textwidth}
+\begin{tabular}{ l  c  l }
+\hline
+Basic covariates & {\#} & Variations\\
+\hline
+\emph{tp}: total precipitation,           & 12  & ensemble mean of sums over 24h, \\
+\hspace*{0.5cm} power transformed (by $1.6^{-1}$) &    & ensemble std. deviation of sums over 24h, \\ 
+\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
+\hspace*{0.9cm} potential energy, &    & ensemble maximum of sums over 24h\\
+\hspace*{0.9cm} power transformed (by $1.6^{-1}$)&    & \qquad all for 6--30 \\
+\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
+\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
+\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
+\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
+\hline
+\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
+\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimum values$^*$,\\
+\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
+\emph{pwat}: precipitable water         &   & ensemble std. deviation of mean values,\\
+\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimum values$^*$,\\
+\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
+\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 \\
+\hspace*{0.95cm} condensate               &   & \\
+\emph{t500}: temperature on 500 hPa     &   & \\
+\hspace*{1cm}                             &   & \\
+\emph{t700}: temperature on 700 hPa     &   & \\
+\hspace*{1cm}                             &   & \\
+\emph{t850}: temperature on 850 hPa     &   & \\
+\hspace{1cm}                              &   & \\
+\hline
+\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
+\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
+\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
+\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 \\
+\emph{tdiff700850}: temperature         &   & \\
+\hspace*{\fill} difference 700 to 850 hPa &   & \\
+\hline
+\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
+\hspace{1.6cm} difference            &   & \qquad over 6--30 \\
+\hline
+\end{tabular}
+\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
+Time periods indicate aggregation time periods in hours after NWP model
+initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
+0600\,UTC to 0600\,UTC of the next day). $^*$Minimum values of \emph{dswrf}
+over 24\,h are always zero and thus neglected.}
+\label{tab:covariates}
+\end{minipage} 
+%}
+\end{table}
+
+Learning and validation data consist of observed 
+daily precipitation sums provided by the National Hydrographical Service 
+(\citealt{ehyd}) and numerical weather forecasts from the U.S.~National
+Oceanic and Atmospheric Administration (NOAA).
+Both, observations and forecasts are available for 1985--2012 and
+the analysis is exemplified using July, the month with the most precipitation
+in Tyrol.
+
+Observations are obtained for 95~stations all over Tyrol and surroundings,
+providing 24-hour precipitation sums measured at 0600\,UTC and rigorously
+quality-checked by the National Hydrographical Service. NWP outputs
+are obtained from the second generation reforecast data set of the
+global ensemble forecast system \citep[GEFS,][]{Hamill+Bates+Whitaker:2013}. 
+This data set consists of an 11-member ensemble based on a fixed version of the 
+numerical model and a horizontal grid spacing of about {$50 \times 50$ km$^2$}
+initialized daily at 0000~UTC from December 1984 to present 
+providing forecasts on a 6-hourly temporal resolution. Each of the 
+11 ensemble members uses slightly different perturbed initial 
+conditions to predict the situation-specific uncertainty 
+of the atmospheric state.
+
+From the GEFS, 14 basic forecast variables are considered with up 
+to 12 variations each such as mean/maximum/minimum
+over different aggregation time periods. A detailed overview is
+provided in Table~\ref{tab:covariates}, yielding 80~predictor variables in total.
+
+To remove large parts of the skewness of precipitation data, 
+a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
+cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
+However, the power parameter may vary for different climatic zones or temporal 
+aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
+in their choice of $1.6^{-1}$ as a suitable power parameter for the
+region of Tyrol. The same power transformation is applied to both the
+observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
+``convective available potential energy'' (\emph{cape}).
+
+\subsection{Models and evaluation}
+\label{sec:evaluation}
+
+\begin{table}[t!]
+\centering
+\begin{tabular}{ l l l l }
+\hline
+Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$)                                  \\ \hline
+Distributional forest & recursive    & all                & all                           \\ 
+                      & partitioning &                    &                               \\ \hline
+EMOS                  & linear       & \emph{tp{\_}mean}  & \emph{tp{\_}sprd}             \\ \hline
+Prespecified GAMLSS   & spline       & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},            \\
+                      & in each      & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean}, \\
+ & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
+ & & \quad \emph{cape{\_}mean1218}, & \quad \emph{cape{\_}mean1218},\\
+ & & \emph{dswrf{\_}mean{\_}mean}, & \emph{tcolc{\_}sprd{\_}mean},\\
+ & & \emph{tcolc{\_}mean{\_}mean}, & \emph{tdiff500850{\_}mean}\\
+ & & \emph{pwat{\_}mean{\_}mean}, & \\
+ & & \emph{tdiff500850{\_}mean}, & \\
+ & & \emph{msl{\_}diff} & \\ \hline
+Boosted GAMLSS        & spline  & all & all \\
+                      & in each &     &     \\ \hline
+\end{tabular}
+\caption[Table caption text]{Overview of models with type of 
+covariate dependency and included covariates for each distribution 
+parameter. \emph{A} $\ast$ \emph{B} indicates an interaction between
+covariate \emph{A} and \emph{B}.}
+\label{tab:models}
+\end{table}
+
+The following zero-censored Gaussian regression models are employed
+in the empirical case study, see Table~\ref{tab:models} for
+further details:
+%
+\begin{itemize}
+
+\item \emph{Distributional forest:} All 80~predictor variables are
+  considered for learning a forest of 100 trees. Bootstrap
+  sampling is employed for each tree using a third of the predictors
+  in each split of the tree (``mtry''). Parameters are estimated by
+  adaptive local likelihood based on the forest weights as described
+  in Section~\ref{sec:methods}.
+
+\item \emph{EMOS:} Standard ensemble model output statistics
+  models use the ensemble mean of total precipitation as regressor in the location
+  submodel and the corresponding ensemble standard deviation in the scale submodel.
+  The parameters are estimated by maximum likelihood, using an identity
+  link for the location part and a log link for the scale part
+  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.
+
+\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
+  the most relevant predictors based on meteorological expert knowledge
+  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
+  on the 80~available variables, eight terms are included in the location
+  submodel and five in the scale submodel. Both involve an interaction of
+  \emph{tp} and \emph{cape} in the afternoon (between 1200\,UTC and 1800\,UTC)
+  to capture the potential for thunderstorms that frequently occur in
+  summer afternoons in the Alps. The model is estimated by maximum
+  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.
+
+\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
+  automatically from all 80~available variables, using non-cyclic boosting
+  for parameter estimation \citep{Hofner+Mayr+Schmid:2016,Messner+Mayr+Zeileis:2017}.
+  This updates the predictor terms for the location or scale submodels iteratively
+  by maximizing the log-likelihood only for the variable yielding the
+  biggest improvement. The iteration stops early -- before fully maximizing
+  the in-sample likelihood -- based on a (computationally intensive)
+  out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
+  the number of boosting iterations (``mstop'') is: $50, 75, \dots, 975, 1000$.
+
+\end{itemize}
+
+The predictive performance in terms of full probabilistic forecasts is
+assessed using the continuous ranked probability score (CRPS, \citealt{Hersbach:2000}).
+For each of the models this assesses the discrepancy of the
+predicted distribution function $F$ from the observation $y$ by
+$$
+\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
+$$
+where \(\mathbf{1}(\cdot)\) is the indicator function. In the subsequent
+applications, the mean CRPS is always evaluated out of sample, 
+either using cross-validation or a hold-out data set (2009--2012)
+that was not used for learning (1985--2008). CRPS is a proper scoring rule
+\citep{Gneiting+Raftery:2007} often used within the meteorological community.
+Lower values indicate better performance.
+
+To assess differences in the improvement of the forests and GAMLSS
+models over the basic EMOS, a CRPS-based skill score with EMOS as
+the reference method is computed:
+$$
+\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{EMOS}}}.
+$$
+
+\subsection{Application for one station}
+\label{sec:axams}
+
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.99\textwidth}
+<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
+# out of sample
+set.seed(7)
+par(mfrow = c(2, 2))
+qqrplot(pit_df, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_ml, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_g,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+qqrplot(pit_gb, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
+@
+\caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
+for station Axams based on models learned on data from 1985--2008.}
+\end{figure}
+
+In a first step, we show a detailed comparison of the competing models
+for one observation site, Axams in Tyrol (in the center of the study area,
+see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation observations and
+numerical weather predictions are available for the month of July from 1985
+through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
+forecasts from the distributional forest, trained on 1985--2008, for July 24
+in 2009--2012 have already been shown as a motivational example. The figure
+shows that the model properly
+depicts the point masses at zero (i.e., the probability of a dry day) and
+the forecasted probability density function for the total amount of precipitation.
+The four sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
+censoring while conforming quite well with the actual observations
+from these days. While this is a nice illustrative example we are  
+interested in the overall predictive performance and calibration of the
+distributional fits.
+
+To assess calibration Figure~\ref{fig:axams-qq} shows residual QQ~plots for
+out-of-sample predictions (2009--2012) from the different models trained on
+1985--2008. Due to the point masses at zero 100 draws from the randomized
+quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
+gray. Overall, the randomized quantile residuals conform quite well with the
+theoretical standard normal quantile (i.e., form a straight line close to the 
+diagonal), indicating that all four models are sufficiently well calibrated. 
+This is also supported by the corresponding probability integral transform
+(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) histograms in
+Appendix~\ref{app:pit}.
+
+<<echo=FALSE, results=hide>>=
+#### cross validation rain
+if(file.exists("crps_cross.rda")){
+  load("crps_cross.rda")
+} else {
+
+  nrep_cross <- 10
+  seed <- 7
+
+  res_cross <- mclapply(1:nrep_cross,
+                        function(i){
+
+                          set.seed(seed*i)
+
+                          # randomly split data in 7 parts each including 4 years
+                          years <- 1985:2012
+                          testyears <- list()
+                          for(j in 1:7){
+                            testyears[[j]] <- sample(years, 4, replace = FALSE)
+                            years <- years[!(years %in% testyears[[j]])]
+                          }
+
+                          #crps <- matrix(nrow = 7, ncol = 7)
+                          reslist <- list()
+                          for(k in 1:7){
+                            test <- testyears[[k]]
+                            train <- c(1985:2012)[!c(1985:2012) %in% test]
+
+                            res <- evalmodels(station = "Axams",
+                                              train = train,
+                                              test = test,
+                                              gamboost_cvr = TRUE)
+
+                            #crps[k,] <- res$crps
+                            reslist[[k]] <- res
+                          }
+
+                          #colnames(crps) <- names(res$crps)
+                          return(reslist)
+                        },
+                        mc.cores = detectCores() - 1
+  )
+
+  # extract CRPS
+  crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
+  # loop over all repetitions
+  for(i in 1:length(res_cross)){
+    #loop over all 7 folds (for 7 methods)
+    crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
+    for(j in 1:length(res_cross[[1]])){
+      crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
+    }
+    crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
+  }
+  colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps)
+
+  save(crps_cross, file = "crps_cross.rda")
+  save(res_cross, file = "res_cross.rda")
+
+}
+@
+
+\begin{figure}[t!]
+\centering
+<<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 7>>=
+par(mar = c(2.5,2,1,2))
+boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
+        names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
+        ylab = "CRPS skill score", col = "lightgray") 
+abline(h = 0, col = pal[5], lwd = 2)
+@
+\caption{\label{fig:axams-crps}CRPS skill score from the 10 times 
+7-fold cross validation at station Axams (1985--2012). The horizontal orange 
+line pertains to the reference model EMOS.}
+\end{figure}
+
+To assess the predictive performance, a full cross-validation is carried out
+rather than relying on just the one fixed test set for the years 2009--2012.
+To do so, a 10 times 7-fold cross-validation is carried out where each
+splits the available 28~years into 7~subsets of 4~randomly selected years.
+The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
+(= 4~years) using the average CRPS across all observations.
+The resulting 10~CRPS skill scores are displayed by boxplots in
+Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
+line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
+perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
+of around 4~percent, the distributional forest has a slightly higher improvement
+of around 5.5~percent in median.
+
+\begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
+par(mar = c(2.5,10,1,2))
+barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
+        horiz = TRUE, las = 1, axes = FALSE,
+        xlab = "Variable importance: mean decrease in CRPS",
+        font.axis = 3, #list(family="HersheySerif", face=3),
+        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
+axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
+@
+\caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
+in the distributional forest. Based on data for station Axams, learning period 
+1985--2008 and assessed in 2009--2012.}
+\end{figure}
+
+Finally, it is of interest how this improvement in predictive performance
+by the distributional forest is accomplished, i.e., which of the 80~covariates
+are selected in the trees of the forest. As the 100~trees of the forest
+do not allow to simply assess the variables' role graphically, a common
+solution for random forests in general is to consider variable importance
+measures. Here, this is defined as the amount of change in CRPS when
+the association between one covariate and the response variable is
+artificially broken through permutation (and thus also breaking the
+association to the remaining covariates).
+
+Figure~\ref{fig:axams-varimp} shows the 10~covariates with the highest permutation
+importance (i.e., change in CRPS) for station Axams. As expected the NWP outputs
+for total precipitation (\emph{tp}) are particularly important along with total
+column-integrated condensate (\emph{tcolc}). Also, both variables occur
+in various transformations such as means (either of the full day or certain parts
+of the afternoon), spreads, or minima/maxima. Thus, while the covariates
+themselves are not surprising, selecting a GAMLSS with a particular combination
+of all the transformations would be much more challenging.
+
+%\pagebreak
+
+\subsection{Application for all stations}
+\label{sec:all}
+
+<<echo = FALSE>>=
+#### prediction over all stations 24 - 4
+if(file.exists("crps_24to4_all.rda")){
+  load("crps_24to4_all.rda")
+} else {
+
+  data("StationsTyrol")
+  stations <- StationsTyrol$name
+  test <- 2009:2012
+  train <- 1985:2008
+
+  res_24to4_all <- mclapply(1:length(stations),
+                            function(i){
+
+                              set.seed(7)
+
+                              res <- evalmodels(station = stations[i],
+                                                train = train,
+                                                test = test,
+                                                gamboost_cvr = TRUE)
+
+                              return(res)
+                            },
+                            mc.cores = detectCores() - 1
+  )
+
+  # extract crps
+  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
+  # loop over all stations
+  for(i in 1:length(stations)){
+    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
+  }
+
+  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
+  rownames(crps_24to4_all) <- stations
+
+  save(crps_24to4_all, file = "crps_24to4_all.rda")
+  save(res_24to4_all, file = "res_24to4_all.rda")
+
+}
+
+# skill score
+s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
+colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
+
+## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
+
+crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]
+
+# best method
+bst <- apply(crps_map, 1, which.min)
+
+# distance of forest to best other method
+dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]
+
+# breaks/groups
+brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
+#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
+grp <- cut(dst, breaks = brk)
+
+# HCL colors (relatively flashy, essentially CARTO Tropic)
+clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
+
+library("raster") # dem (digital elevation model)
+library("sp")     # gadm www.gadm.org/country
+
+#load("plot_map_rain/data/tirol.rda")
+#load("plot_map_rain/data/dem.rda")
+
+data(StationsTyrol, package = "RainTyrol")
+data(MapTyrol, package = "RainTyrol")
+# data(MapTyrol_border, package = "RainTyrol")
+# Create SpatialPointsDataFrame from station list
+sp <- SpatialPointsDataFrame(subset(StationsTyrol,
+                                    select=c(lon,lat)),
+                             data = subset(StationsTyrol,
+                                           select = -c(lon,lat)),
+                             proj4string = crs(MapTyrol$RasterLayer))
+@
+
+\begin{figure}[p!]
+\centering
+<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, height = 4.5, width = 7>>=
+par(mar = c(3.1, 4.1, 1.1, 2.1))
+matplot(t(s[,]), type = "l", lwd = 2, 
+          col = gray(0.5, alpha = 0.2),
+          lty = 1, axes = FALSE, 
+          xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
+lines(s[70,], col = pal[2], type = "o", pch = 19, lwd = 2)  
+# Station Axams is the 77th station which is the 70th in the list of complete stations
+  boxplot(s, add = TRUE, col = "transparent")
+  abline(h = 0, col = pal[5], lwd = 2)
+
+@
+\caption{\label{fig:all-crps}CRPS skill score for each station (gray lines
+with boxplots superimposed). Station Axams is highlighted in green 
+and the horizontal orange line pertains to the reference model EMOS. The models are 
+learned on 1985--2008 and validated for 2009--2012.}
+
+\vspace*{0.4cm}
+
+\setkeys{Gin}{width=0.99\textwidth}
+<<map, fig=TRUE, echo=FALSE, height=6.1, width=9>>=
+
+  ## plot map of Tyrol with all 95 observations
+  layout(cbind(1, 2), width = c(9, 1))
+  par(mar = c(5,4,4,0.1))
+  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
+                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
+                xlim = c(9.8,13.2), 
+                ylim = c(46.6, 47.87))
+  plot(MapTyrol$SpatialPolygons, add = TRUE)
+  points(sp[70,], pch = 19, col = "black", cex = 1.85)
+  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
+  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
+  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
+  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
+  par(mar = c(0.5,0.2,0.5,2.3))
+  ## legend
+  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
+       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
+  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
+  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))
+
+@
+\caption{\label{fig:all-map}Map of Tyrol coding the best-performing model
+for each station (type of symbol). The color codes whether the distributional 
+forest had higher (green) or lower (red) CRPS compared to the best of the 
+other three models. The background shows the local topography \citep{Robinson:2014}.
+Station Axams is highlighted in bold.}
+\end{figure}
+
+After considering only one observational site up to now, a second step
+evaluates and compares the competing methods on all 95 available stations. As in the previous
+section, all models are learned on the first 24~years and evaluated by the
+average CRPS on the last 4~years. More specifically, the CRPS skill score
+against the EMOS model is computed for the out-of-sample predictions at each
+station and visualized by parallel coordinates plots with boxplots superimposed
+in Figure~\ref{fig:all-crps}. Overall, distributional forests have a slightly
+higher improvement in CRPSS compared to the two GAMLSS which is best seen
+by looking at the boxplots and the green line representing the results
+for station Axams. The underlying parallel coordinates additionally
+bring out that the prespecified GAMLSS sometimes performs rather differently
+(sometimes better, sometimes worse) compared to the two data-driven models.
+Values below zero show that, for some stations, EMOS performs better
+than the more complex statistical methods.
+%NAs in gamlss in 13 stations!!!\\
+%for example in 13th station in complete{\_}stations\\
+%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
+%\text{NA's in the working vector or weights for parameter mu}
+%\\
+% sum(is.na(crps[,3]))
+%<<eval = TRUE, echo = FALSE>>=
+
+To assess whether these differences in predictive performance are due to differences
+in the topography, Figure~\ref{fig:all-map} shows a brief spatial summary of all
+stations. Each station is illustrated by a symbol that conveys which
+model performed best in terms of CRPS on the last 4~years of the data.
+Additionally, the color of the symbol indicates the CRPS difference between 
+distributional forest and the best-performing other model. Green signals that
+the distributional forest performs better than the other models whereas red
+signals that another model performs better. Overall the distributional forest
+performs on par (gray) or better (green) for the majority of stations. Only
+for a few stations in the north-east EMOS performs best, and in East Tyrol
+the prespecified GAMLSS performs particularly well.
+
+\section{Discussion}
+\label{sec:discussion}
+
+Distributional regression modeling is combined with tree-based modeling
+to obtain a novel and flexible method for probabilistic forecasting.
+The resulting distributional trees and forests can capture abrupt and
+nonlinear effects and interactions in a data-driven way. By basing the
+split point and split variable selection on a full likelihood and
+corresponding score function, the trees and forests can not only pick
+up changes in the location but also the scale or shape of any distributional
+family.
+
+Distributional forests are an attractive alternative when prespecifying or
+boosting all possible effects and interactions in a GAMLSS model is
+challenging.  Distributional forests are rather straightforward to specify
+requiring only little prior subject matter knowledge and also work well in the
+presence of many potential covariates. The application to precipitation
+forecasting in complex terrain illustrates that distributional forests often
+perform on par or even better than their GAMLSS counterparts. Hence, they form
+a useful addition to the already available toolbox of probabilistic forecasts
+for disciplines such as meteorology.
+
+\section*{Computational details}
+
+The proposed methods are in the \textsf{R} package \textbf{disttree}
+(version~\Sexpr{packageVersion("disttree")}) based on the \textbf{partykit}
+package (version~\Sexpr{packageVersion("partykit")}), both available 
+on \textsf{R}-Forge at
+(\url{https://R-Forge.R-project.org/projects/partykit/}). The function
+\code{distfit} fits distributional models by maximum likelihood, which is
+used as the basis for the tree-building function \code{disttree}, upon which
+the \code{distforest} is built. All functions can either be used with
+GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
+(\citealt{Stasinopoulos+Rigby:2007}, version~\Sexpr{packageVersion("gamlss.dist")})
+or with custom lists containing all required information about the distribution family.
+
+In addition to \textbf{disttree}, Section~\ref{sec:precipitation} employs
+\textsf{R} package \textbf{crch} 
+\citep[version~\Sexpr{packageVersion("crch")}]{Messner+Mayr+Zeileis:2016} for the EMOS models,
+\textbf{gamlss}
+\citep[version~\Sexpr{packageVersion("gamlss")}]{Stasinopoulos+Rigby:2007} for the prespecified GAMLSS, and
+\textbf{gamboostLSS}
+\citep[version~\Sexpr{packageVersion("gamboostLSS")}]{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
+
+The fitted distributional forest for July 24 and
+observation station Axams (including Figure~\ref{fig:axams-dist})
+is reproducible using 
+\code{demo("RainAxams", package = "disttree")}.
+This also includes fitting the other zero-censored
+Gaussian models considered in this paper and generating the corresponding 
+QQ~plots (Figure~\ref{fig:axams-qq}) and
+PIT histograms (Figure~\ref{fig:axams-pit}).
+Full replication of all results can be obtained with
+\code{demo("RainTyrol", package = "disttree")}
+requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
+also available within the \textsf{R}-Forge~project.
+
+\bibliography{ref.bib}
+
+\newpage
+\begin{appendix}
+
+\section{Tree algorithm}
+\label{app:tree}
+
+In the following, the tree algorithm applied in the empirical case study 
+discussed in this paper is explained. For notational simplicity, 
+the testing and splitting procedure is described for the root node, i.e.,
+the entire learning sample with observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
+In each child node the corresponding subsample depends on 
+the foregoing split(s).
+
+After fitting a distributional model $\mathcal{D}(Y, \bm{\theta})$ to 
+the learning sample with observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
+Section~\ref{sec:distfit} the resulting estimated parameter 
+$\bm{\hat{\theta}} = 
+(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
+can be plugged in the score function $s(\bm{\theta}, Y)$.
+In that way a goodness-of-fit measurement is obtained for each
+parameter $\theta_j$ and each observation $y_i$.
+To use this information, statistical tests are employed to detect
+dependencies between the score values
+$$
+s(\bm{\hat{\theta}}, y) = 
+%\begin{pmatrix} 
+%s(\bm{\hat{\theta}}, y_1) \\
+%s(\bm{\hat{\theta}}, y_2) \\
+%\vdots \\
+%s(\bm{\hat{\theta}}, y_n)
+%\end{pmatrix} =
+\begin{pmatrix} 
+s(\bm{\hat{\theta}}, y_1)_1 & s(\bm{\hat{\theta}}, y_1)_2 & \ldots & s(\bm{\hat{\theta}}, y_1)_k\\
+\vdots & \vdots & \ddots & \vdots \\
+s(\bm{\hat{\theta}}, y_n)_1 & s(\bm{\hat{\theta}}, y_n)_2 & \ldots & s(\bm{\hat{\theta}}, y_n)_k
+\end{pmatrix}
+$$
+and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
+\begin{align*}
+H_0^l:  s(\bm{\hat{\theta}}, Y) \qquad \bot \qquad Z_l 
+\end{align*}
+
+These hypotheses are assessed with permutation tests based on
+the multivariate linear statistic
+$$
+T_{l} = \sum_{i=1}^n v_l(z_{li}) \cdot s(\bm{\hat{\theta}}, y_i).
+$$
+The type of the transformation function $v_l$ depends on 
+the type of the split variable $Z_l$. If $Z_l$ is numeric then $v_l$ 
+is simply the identity function $v_l(z_{li}) = z_{li}$. If $Z_l$ is 
+a categorical variable with $H$ categories then 
+$v_l(z_{li}) = e_H(z_{li}) = (\I(z_{ji} = 1), \ldots, \I(z_{li} = H))$ 
+such that $v_l$ is a unit vector where the element corresponding to 
+the value of $z_{li}$ is $1$. Observations with missing values are 
+excluded from the sums.
+
+With the conditional expectation $\mu_{l}$ and the covariance 
+$\Sigma_{l}$ as derived by \cite{Strasser+Weber:1999} the test 
+statistic can be standardized. The observed multivariate linear 
+statistic $t$ is mapped onto the real line by a  univariate test 
+statistic $c$. In the application of this paper a quadratic form 
+is chosen, such that
+$$
+c_{\text{quad}}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
+$$
+where $\Sigma^+$ is the Moore-Penrose inverse of $\Sigma$. Alternatively,
+the maximum of the absolute values of the standardized linear statistic
+can be considered.
+
+The smaller the $p$-value corresponding to the standardized test 
+statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
+from the assumption of independence between the scores and the split 
+variable $Z_l$.
+%
+After Bonferroni-adjusting the $p$-values it has to be assessed whether
+any of the resulting $p$-values is beneath the selected significance level. 
+If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
+is chosen as splitting variable.
+
+The breakpoint that leads to the highest discrepancy between score functions 
+in the two resulting subgroups is selected as split point. 
+This is measured by the linear statistic
+$$
+T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
+$$
+where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
+are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
+The split point is then chosen as follows:
+$$
+r^{\ast} = \argmin_{r} c(t_{l^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{l^{\ast}}^r).
+$$
+One repeats the testing and splitting procedure in each of the resulting 
+subgroups until some stopping criterion is reached. This criterion 
+can for example be a minimal number of observations in a node or a 
+minimal $p$-value for the statistical tests. In that way pre-pruning
+is applied in order to find right-sized trees and hence 
+avoid overfitting.
+
+This permutation test based tree algorithm is presented in 
+\cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
+A different framework to build a tree is provided by the MOB algorithm 
+which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).
+
+\section{PIT histograms}
+\label{app:pit}
+
+\begin{figure}[b!]
+\centering
+%\begin{subfigure}{0.8\textwidth}
+<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7>>=
+
+# out of sample
+set.seed(4)
+
+par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
+pithist(pit_df, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
+pithist(pit_ml, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
+pithist(pit_g, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
+pithist(pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
+
+@
+%\end{subfigure}
+\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
+\end{figure}
+
+As an alternative to QQ~plots based on randomized quantile residuals,
+probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
+histograms are a commonly used tool to assess how well-calibrated distributional 
+fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
+probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
+some deviations the overall calibration of all models is acceptable.
+
+\end{appendix}
+
+\end{document}
+
 \newcommand{\bS}{\mathbf{S}}
@@
 $$
-T_{l} = \sum_{i=1}^n v_l(z_{li}) \cdot s(\bm{\hat{\theta}}, y_i).
-$$
-The type of the transformation function $v_l$ depends on 
-the type of the split variable $Z_l$. If $Z_l$ is numeric then $v_l$ 
-is simply the identity function $v_l(z_{li}) = z_{li}$. If $Z_l$ is 
-a categorical variable with $H$ categories then 
-$v_l(z_{li}) = e_H(z_{li}) = (\I(z_{ji} = 1), \ldots, \I(z_{li} = H))$ 
-such that $v_l$ is a unit vector where the element corresponding to 
-the value of $z_{li}$ is $1$. Observations with missing values are 
-excluded from the sums.
-
-With the conditional expectation $\mu_{l}$ and the covariance 
-$\Sigma_{l}$ as derived by \cite{Strasser+Weber:1999} the test 
-statistic can be standardized. The observed multivariate linear 
-statistic $t$ is mapped onto the real line by a  univariate test 
-statistic $c$. In the application of this paper a quadratic form 
-is chosen, such that
-$$
-c_{\text{quad}}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
-$$
-where $\Sigma^+$ is the Moore-Penrose inverse of $\Sigma$. Alternatively,
+T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\bm{\hat{\theta}}, Y_i)\right)
+$$
+where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
+type of the transformation function $v_l$ depends on the type of the split 
+variable $Z_l$. If $Z_l$ is numeric then $v_l$ is simply the identity 
+function $v_l(Z_{li}) = Z_{li}$ and therefore $T_l \in \mathbb{R}^k$ as 
+the ``vec'' operator converts the $1 \times k$ matrix into a $k$ column vector. 
+If $Z_l$ is a categorical variable with $H$ categories then 
+$v_l(Z_{li}) = e_H(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$ 
+such that $v_l$ is a $H$-dimensional unit vector where the element corresponding to 
+the value of $Z_{li}$ is $1$. In this case the statistic $T_l \in \mathbb{R}^{H \cdot k}$
+as the ``vec'' operator converts the $H \times k$ matrix into a $k$ column vector 
+by column-wise combination. Observations with missing values are excluded from the sums.
+
+With the conditional expectation $\mu_l$ and the covariance 
+$\Sigma_l$ of $T_l$ as derived by 
+\cite{Strasser+Weber:1999} the test statistic can be standardized. 
+The observed multivariate linear statistic $t_l$ which is either an
+$k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of Y,
+is mapped onto the real line by a univariate test statistic $c$. 
+In the application of this paper a quadratic form is chosen, such that
+$$
+c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
+$$
+where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$. Alternatively,
 the maximum of the absolute values of the standardized linear statistic
@@
 If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
-is chosen as splitting variable.
+is chosen as splitting variable. Otherwise no further split is made is in this node.
 

### 2018-07-05 pkg/disttree/inst/draft/draft.Rnw
corrections of the notation in the appendix based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2207 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c611cfca307dec01674a360cb58c46458a3f21f2

@@
 $$
-T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
-$$
-where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
+T_{l^{\ast}}^r = \min_{s=1,2}\sum_{i \in \mathcal{B}_{sr}} s(\bm{\hat{\theta}}, Y_i)
+$$
+where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the two new subgroups, without 
+any particular ordering, that 
 are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 

### 2018-07-06 pkg/disttree/inst/draft/draft.Rnw
corrections of the notation in the appendix based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2208 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0504b4ebb31ccd61d1b01de0268b1ddbe780f023

--- 2018-07-05 pkg/disttree/inst/draft/draft.Rnw
@@
 $$
-T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
-$$
-where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
-are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
+T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\bm{\hat{\theta}}, Y_i)
+$$
+for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
+two new subgroups, without any particular ordering, that are defined by 
+splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
 The split point is then chosen as follows:
 $$
-r^{\ast} = \argmin_{r} c(t_{l^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{l^{\ast}}^r).
+r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
 $$

### 2018-07-06 pkg/disttree/inst/draft/draft.Rnw
corrections of the notation in the appendix based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2209 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dfb295f6de34e9b4bfe8390b2700a37e61a391fd

@@
 \documentclass[nojss]{jss}
-\shortcites{Hamill+Bates+Whitaker:2013}
-
-%\VignetteIndexEntry{disttree: Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
-%\VignetteDepends{disttree, crch, countreg, gamlss, gamlss.cens}
-%\VignetteKeywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
-%\VignettePackage{disttree}
-
-%% packages
-\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}
-
-%% tikz
-\usepackage{array,makecell,tikz}
-\usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-
-%% need no \usepackage{Sweave}
-\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
-
-%% commands
-\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\renewcommand{\Prob}{\mathbb{P} }
-\renewcommand{\E}{\mathbb{E}}
-\newcommand{\V}{\mathbb{V}}
-\newcommand{\Var}{\mathbb{V}}
-\newcommand{\R}{\mathbb{R} }
-\newcommand{\N}{\mathbb{N} }
-\newcommand{\C}{\mathbb{C} }
-\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
-\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
-\newcommand{\LS}{\mathcal{L}_n}
-\newcommand{\TS}{\mathcal{T}_n}
-\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
-\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
-\newcommand{\F}{\mathcal{F}}
-\newcommand{\A}{\mathcal{A}}
-\newcommand{\yn}{y_{\text{new}}}
-\newcommand{\z}{\mathbf{z}}
-\newcommand{\X}{\mathbf{X}}
-\newcommand{\Y}{\mathbf{Y}}
-\newcommand{\sX}{\mathcal{X}}
-\newcommand{\sY}{\mathcal{Y}}
-\newcommand{\T}{\mathbf{T}}
-\newcommand{\x}{\mathbf{x}}
-\renewcommand{\a}{\mathbf{a}}
-\newcommand{\xn}{\mathbf{x}_{\text{new}}}
-\newcommand{\y}{\mathbf{y}}
-\newcommand{\w}{\mathbf{w}}
-\newcommand{\ws}{\mathbf{w}_\cdot}
-\renewcommand{\t}{\mathbf{t}}
-\newcommand{\M}{\mathbf{M}}
-\renewcommand{\vec}{\text{vec}}
-\newcommand{\B}{\mathbf{B}}
-\newcommand{\K}{\mathbf{K}}
-\newcommand{\W}{\mathbf{W}}
-\newcommand{\D}{\mathbf{D}}
-\newcommand{\I}{\mathbf{I}}\documentclass[nojss]{jss}
 \shortcites{Hamill+Bates+Whitaker:2013}
@@
 \newcommand{\I}{\mathbf{I}}
-\newcommand{\bS}{\mathbf{S}}
-\newcommand{\cellx}{\pi_n[\x]}
-\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
-\newcommand{\err}{\text{Err}}
-\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
-\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
-\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
-\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
-\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
-\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
-\newcommand{\bft}{\mathbf{t}}
-
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-library("disttree")
-library("crch")
-library("countreg")
-library("gamlss")
-library("gamlss.cens")
-gen.cens(NO, type = "left")
-library("RainTyrol")
-
-## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-@
-
-\title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
-\Shorttitle{Distributional Regression Forests for Probabilistic Precipitation Forecasting}
-
-\author{Lisa Schlosser\\Universit\"at Innsbruck
-   \And Torsten Hothorn\\Universit\"at Z\"urich
-   \And Reto Stauffer\\Universit\"at Innsbruck
-   \And Achim Zeileis\\Universit\"at Innsbruck}
-\Plainauthor{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
-
-\Abstract{
-To obtain a probabilistic model for a dependent variable based on some set of
-explanatory variables, a distributional approach is often adopted where the
-parameters of the distribution are linked to regressors. In many classical
-models this only captures the location of the distribution but
-over the last decade there has been increasing interest in distributional
-regression approaches modeling all parameters including location, scale, and
-shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
-mean and variance of a Gaussian response and is particularly
-popular in weather forecasting. More generally, the GAMLSS framework
-allows to establish generalized additive models for location, scale, and shape
-with smooth linear or nonlinear effects.
-%
-However, when variable selection is required and/or there are
-non-smooth dependencies or interactions (especially unknown or of high-order),
-it is challenging to establish a good GAMLSS. A natural alternative in these
-situations would be the application of regression trees or random forests but,
-so far, no general distributional framework is available for these. Therefore,
-a framework for distributional regression trees and forests is proposed that blends
-regression trees and random forests with classical distributions from the GAMLSS
-framework as well as their censored or truncated counterparts.
-%
-To illustrate these novel approaches in practice, they are employed to obtain
-probabilistic precipitation forecasts at numerous sites in a mountainous
-region (Tyrol, Austria) based on a large number of numerical weather
-prediction quantities. It is shown that the novel distributional regression forests
-automatically select variables and interactions, performing on par or often even
-better than GAMLSS specified either through prior meteorological knowledge or a
-computationally more demanding boosting approach.
-}
-\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
-
-\Address{
-  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
-  Universit\"at Innsbruck \\
-  Department of Statistics \\
-  Faculty of Economics and Statistics \\
-  Universit\"atsstr.~15 \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
-  \email{Achim.Zeileis@R-project.org} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
-  \phantom{URL: }\url{https://retostauffer.org/},\\
-  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
-
-  Torsten Hothorn\\
-  Universit\"at Z\"urich \\
-  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
-  Hirschengraben 84\\
-  CH-8001 Z\"urich, Switzerland \\
-  E-mail: \email{Torsten.Hothorn@R-project.org}\\
-  URL: \url{http://user.math.uzh.ch/hothorn/}\\
-
-}
-
-\begin{document}
-
-\section{Introduction}
-\label{sec:introduction}
-
-In regression analysis a wide range of models has been developed to describe the
-relationship between a response variable and a set of covariates. The classical
-model is the linear model (LM) where the conditional mean of the response
-is modeled through a linear function of the covariates (see the left panel of
-Figure~\ref{fig:gam} for a schematic illustration). Over the last decades
-this has been extended in various directions including:
-
-\pagebreak
-
-\begin{itemize}
-  \item \emph{Generalized linear models} (GLMs, \citealp{Nelder+Wedderburn:1972})
-    encompassing an additional nonlinear link function for the conditional mean.
-  \item \emph{Generalized additive models} (GAMs, \citealp{Hastie+Tibshirani:1986})
-    allowing for smooth nonlinear effects in the covariates
-    (Figure~\ref{fig:gam}, middle).
-  \item \emph{Generalized additive models for location, scale, and shape}
-    (GAMLSS, \citealt{Rigby+Stasinopoulos:2005}) adopting a probabilistic modeling
-    approach. In GAMLSS, each parameter of a statistical distribution can depend
-    on an additive predictor of the covariates comprising linear and/or
-    smooth nonlinear terms (Figure~\ref{fig:gam}, right).
-\end{itemize}
-Thus, the above-mentioned models provide a broad toolbox for capturing different
-aspects of the response (mean only vs.\ full distribution) and different types
-of dependencies on the covariates (linear vs.\ nonlinear additive terms).
-
-\begin{figure}[t!]
-\begin{tikzpicture}
-\node (a) at (0,0){
-\minipage{0.29\textwidth}
-\centering
-<<motivation_GLM, echo=FALSE, results=hide>>=
-nobs <- 200
-## GLM
-set.seed(7)
-x <- c(1:nobs)/nobs
-ytrue <- 1+x*1.5
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-
-<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal[1], lwd = 7, main = "")
-@
-\endminipage};
-\node (b) at (5,0) 
-{
-\minipage{0.29\textwidth}
-\centering
-<<motivation_GAM, echo=FALSE, results=hide>>=
-## GAM
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3 
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-
-<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal[1], lwd = 7, main = "")
-@
-\endminipage};
-\node (c) at (10,0) 
-{
-\minipage{0.29\textwidth}
-\centering
-<<motivation_GAMLSS, echo=FALSE, results=hide>>=
-## GAMLSS
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3
-var <- exp(-(2*x)^2)/2
-y <- ytrue + rnorm(nobs, 0, 0.1 + var)
-@
-
-<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
-polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
-  col = pallight[1], border = "transparent")
-lines(x, ytrue, col = pal[1], lwd=7)
-points(x, y, col = "slategray", pch = 19)
-box(lwd = 5)
-@
-\endminipage};
-\node (d) at (0,-2) 
-{
-\minipage{0.29\textwidth}
-\centering
-\vspace{0.0cm}
-LM, GLM  
-\endminipage};
-\node (e) at (5,-2) 
-{
-\minipage{0.29\textwidth}
-\centering
-\vspace{0.0cm}
-GAM 
-\endminipage};
-\node at (10,-2)
-{
-\minipage{0.28\textwidth}
-\centering
-\vspace{0.0cm}
-GAMLSS 
-\endminipage};
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
-\end{tikzpicture}
-\caption{\label{fig:gam}Parametric modeling developments. (Generalized) linear models (left),
-generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
-\end{figure}
-
-While in many applications conditional mean regression models have been receiving
-the most attention, there has been a paradigm shift over the last decade towards
-distributional regression models. An important reason for this is that in many
-fields forecasts of the mean are not the only (or not even the main) concern but
-instead there is an increasing interest in probabilistic forecasts. Quantities of
-interest typically include exceedence probabilities for certain thresholds of the
-response or quantiles of the response distribution. Specifically, consider
-weather forecasting where there is less interest in the mean amount of
-precipitation on the next day. Instead, the probability of rain vs.\ no rain
-is typically more relevant or, in some situations, a prediction interval of
-expected precipitation (say from the expected 10\% to 90\% quantile). Similar
-considerations apply for other meteorological quantities and hence attention
-in the weather forecasting literature has been shifting from classical linear 
-deterministic models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
-non-homogeneous Gaussian regression (NGR) of \cite{Gneiting+Raftery+Westveld:2005}.
-The NGR typically describes the mean of some meteorological response variable through
-the average of the corresponding quantity from an ensemble of physically-based
-numerical weather predictions (NWPs). Similarly, the variance of the response
-is captured through the variance of the ensemble of NWPs. Thus, the NGR
-considers both the mean as well as the uncertainty of the ensemble predictions
-to obtain probabilistic forecasts calibrated to a particular site.
-
-\begin{figure}[t!]
-\begin{tikzpicture}
-\node (a) at (0,0){
-\minipage{0.29\textwidth}
-\centering
-<<motivation_regtree, echo=FALSE, results=hide>>=
-## Reg. Tree
-set.seed(7)
-kappa <- 12
-x <- c(1:nobs)/nobs
-ytrue <- ytree <- yforest <- numeric(length = length(x))
-for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
-y <- ytrue + rnorm(nobs,0,0.3)
-for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
-@
-
-<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
-lines(x = x, y = ytree, col = pal[1], lwd=7)
-@
-\endminipage};
-\node (b) at (5,0) 
-{
-\minipage{0.29\textwidth}
-\centering
-<<motivation_randforest, echo=FALSE, results=hide>>=
-## Random Forest
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-@
-
-<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
-lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
-@
-\endminipage};
-\node (c) at (10,0) 
-{
-\minipage{0.29\textwidth}
-\centering
-<<motivation_distforest, echo=FALSE, results=hide>>=
-## distforest
-set.seed(4723)
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-xt <- 2*(x-0.5)
-var <- exp(-(2*xt)^4)/2
-y <- ytrue + rnorm(nobs, 0, 0.1 + var)
-x <- c(1:nobs)/nobs
-@
-
-<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
-par(mar = c(2, 0, 2, 0))
-plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
-usr <- par("usr")
-rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
-polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
-  col = pallight[1], border = "transparent")
-lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
-points(x, y, col = "slategray", pch = 19)
-box(lwd=8)
-@
-\endminipage};
-\node (d) at (0,-2) 
-{
-\minipage{0.29\textwidth}
-\centering
-\vspace{0.0cm}
-Regression tree 
-\endminipage};
-\node (e) at (5,-2) 
-{
-\minipage{0.29\textwidth}
-\centering
-\vspace{0.0cm}
-Random forest 
-\endminipage};
-\node at (10,-2)
-{
-\minipage{0.28\textwidth}
-\centering
-\vspace{0.0cm}
-Distributional forest 
-\endminipage};
-\node (g) at (0,-5.4) 
-{
-\minipage{0.29\textwidth}
-\vspace{0.2cm}
-%\hspace{0.0cm}
-\centering
-<<motivation_disttree, echo=FALSE, results=hide>>=
-## disttree
-set.seed(74)
-x <- c(1:nobs)/nobs
-var <- ytree/3
-y <- ytrue + rnorm(nobs,0,var)
-@
-
-<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
-par(mar = c(2, 0, 2, 0))
-plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
-usr <- par("usr")
-rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
-lines(x = x, y = ytree, col = pal[1], lwd=7, main = "")
-polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
-  col = pallight[1], border = "transparent")
-points(x, y, col = "slategray", pch = 19)
-box(lwd=8)
-@
-\vspace{-0.2cm}%\\
-Distributional tree
-\vspace{0.2cm}
-\endminipage};
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](0,-2.4)--(0,-3.2);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.2,-5.4)--(7.9,-2.5);
-\end{tikzpicture}
-\caption{\label{fig:tree}Tree and forest developments.
-Regression tree (top left), distributional tree (bottom left), random forest
-(top middle), and distributional forest (top right).}
-\end{figure}
-
-In summary, the models discussed so far provide a broad and powerful toolset for
-parametric distributional fits depending on a specified set of additive linear or
-smooth nonlinear terms. A rather different approach to capturing the dependence on
-covariates are tree-based models.
-\begin{itemize}
-  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
-    split the data into more homogeneous subgroups and can thus capture abrupt shifts
-    (Figure~\ref{fig:tree}, top left) and approximate nonlinear functions.
-    Furthermore, trees automatically carry out a forward selection of covariates and
-    their interactions. 
-  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
-    an ensemble of trees fitted to resampled versions of the learning data. This
-    stabilizes the recursive partitions from individual trees and hence better
-    approximates smooth functions (Figure~\ref{fig:tree}, top middle).
-\end{itemize}
-While classical regression trees and random forests only model the mean of the response
-we propose to follow the ideas from GAMLSS modeling -- as outlined
-in Figure~\ref{fig:gam} -- and combine tree-based methods with parametric distributional models,
-yielding two novel techniques:
-\begin{itemize}
-  \item \emph{Distributional regression trees} (for short: \emph{distributional trees}) split the data into more homogeneous groups with
-    respect to a parametric distribution, thus capturing changes in any distribution
-    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left).
-  \item \emph{Distributional regression forests} (for short: \emph{distributional forests}) utilize an ensemble of distributional trees
-    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
-    top right).
-\end{itemize}
-In the following, particular focus is given to distributional forests as a method for obtaining
-probabilistic forecasts by leveraging the strengths of random forests: the ability to
-capture both smooth and abruptly changing functions along with simultaneous selection of
-variables and possibly complex interactions. Thus, these properties make the method particularly
-appealing in case of many covariates with unknown effects and interactions where it would be 
-challenging to specify a distributional regression model like GAMLSS.
-
-<<echo=FALSE, results=hide>>=
-#### Axams prediction 24 - 4
-if(file.exists("Axams_24to4.rda")){
-  load("Axams_24to4.rda")
-} else {
-  source("Axams_24to4.R")
-  Axams_24to4 <- Axams_24to4()
-  save(Axams_24to4, file = "Axams_24to4.rda")
-}
-
-#### prepare data for plot of estimated density functions
-# predictions for one day (in each of the four years) 
-# (19th of July 2011 is missing)
-pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
-
-pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
-pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata[pdays,], type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, Axams_24to4$testdata[pdays,"robs"])
-
-# plot predicted distributions together with observations
-set.seed(7)
-x <- c(0.01, sort(runif(500,0.01,8)))
-y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
-y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
-y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
-y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
-dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
-
-# point mass (slightly shifted)
-pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
-pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
-pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
-pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
-
-# predictions
-pred1 <- c(Axams_24to4$testdata[pdays,"robs"][1], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(Axams_24to4$testdata[pdays,"robs"][2], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(Axams_24to4$testdata[pdays,"robs"][3], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(Axams_24to4$testdata[pdays,"robs"][4], crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
-
-#legendheight
-lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
-lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
-lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
-
-############
-# get predicted parameter of all models for testdata
-
-# distributional tree
-pdt <- predict(Axams_24to4$dt, newdata = Axams_24to4$testdata, type = "parameter")
-dt_mu <- pdt$mu
-dt_sigma <- pdt$sigma
-dt_exp <- pnorm(dt_mu/dt_sigma) * (dt_mu + dt_sigma * (dnorm(dt_mu/dt_sigma) / pnorm(dt_mu/dt_sigma)))
-
-# distributional forest
-pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata, type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-
-# prespecified GAM
-g_mu <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "mu", type = "response", data = Axams_24to4$g_learndata)
-g_sigma <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "sigma", type = "response", data = Axams_24to4$g_learndata)
-g_exp <- pnorm(g_mu/g_sigma) * (g_mu + g_sigma * (dnorm(g_mu/g_sigma) / pnorm(g_mu/g_sigma)))
-
-# boosted GAM
-pgb <- predict(Axams_24to4$gb, newdata = Axams_24to4$testdata, parameter = c("mu","sigma"), type = "response")
-gb_mu <- pgb$mu
-gb_sigma <- pgb$sigma
-gb_exp <- pnorm(gb_mu/gb_sigma) * (gb_mu + gb_sigma * (dnorm(gb_mu/gb_sigma) / pnorm(gb_mu/gb_sigma)))
-
-# EMOS
-ml_mu <- predict(Axams_24to4$ml, type = "location", newdata = Axams_24to4$testdata)
-ml_sigma <- predict(Axams_24to4$ml, type = "scale", newdata = Axams_24to4$testdata)
-ml_exp <- pnorm(ml_mu/ml_sigma) * (ml_mu + ml_sigma * (dnorm(ml_mu/ml_sigma) / pnorm(ml_mu/ml_sigma)))
-
-############
-## PIT histograms and qqrplots
-
-## get predicted parameter
-# for learndata and testdata
-{
-  set.seed(7)
-
-  # distributional tree
-  pit_dt <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = dt_mu, sd = dt_sigma))
-  pit_dt[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_dt[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # distributional forest
-  pit_df <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = df_mu, sd = df_sigma))
-  pit_df[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_df[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # prespecified GAM
-  pit_g <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = g_mu, sd = g_sigma))
-  pit_g[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_g[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # boosted GAM
-  pit_gb <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = gb_mu, sd = gb_sigma))
-  pit_gb[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_gb[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # EMOS
-  pit_ml <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = ml_mu, sd = ml_sigma))
-  pit_ml[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_ml[Axams_24to4$testdata[,"robs"]>0, 2]
-
-}
-
-## Variable importance
-if(file.exists("vimp_crps.rda")){
-  load("vimp_crps.rda")
-} else {
-
-  set.seed(7)
-  # locally redefine logLik.distforest to use crps in varimp()
-  logLik.distforest <- function(object, newdata = NULL, ...) {
-    if(is.null(newdata)) {
-      newdata <- object$data
-    }
-
-    # predict parameter
-    pdf <- predict(object, newdata = newdata, type = "parameter")
-    df_mu <- pdf$mu
-    df_sigma <- pdf$sigma
-
-    # calculate CRPS
-    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
-
-    return(structure(crps, df = NA, class = "logLik"))
-  }
-
-  vimp_crps <- varimp(Axams_24to4$df, nperm = 1L)
-
-  save(vimp_crps, file = "vimp_crps.rda")
-
-  rm(logLik.distforest) 
-}
-
-@
-
-\begin{figure}[t!]
-\centering
-\setkeys{Gin}{width=0.75\textwidth}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=3.8>>=
-par(mar = c(3.8, 4, 2.3, 1.5))
-plot(x = x, y = y1, type = "l", col = pal[1], lwd = 1.3, 
-     main = paste0("July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
-     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-1.5,8))
-
-lines(x = x, y = y2, type = "l", col = pal[3], lwd = 1.3)
-lines(x = x, y = y3, type = "l", col = pal[2], lwd = 1.3)
-lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
-legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
-       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)
-
-# plot point mass
-lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = pal[1], type = "l", lwd = 1.3)
-lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = pal[3], type = "l", lwd = 1.3)
-lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = pal[2], type = "l", lwd = 1.3)
-lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = pal[4], type = "l", lwd = 1.3)
-
-points(x = pm1[1], y = pm1[2], col = pal[1], pch = 19)
-points(x = pm2[1], y = pm2[2], col = pal[3], pch = 19)
-points(x = pm3[1], y = pm3[2], col = pal[2], pch = 19)
-points(x = pm4[1], y = pm4[2], col = pal[4], pch = 19)
-
-# plot predictions
-points(x = pred1[1], y = pred1[2], col = pal[1], pch = 4)
-points(x = pred2[1], y = pred2[2], col = pal[3], pch = 4)
-points(x = pred3[1], y = pred3[2], col = pal[2], pch = 4)
-points(x = pred4[1], y = pred4[2], col = pal[4], pch = 4)
-
-lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-
-# add labels
-text(x = -0.8, y = lh1, labels = "2009", col = pal[1], cex = 0.8)
-text(x = -0.8, y = lh2, labels = "2010", col = pal[3], cex = 0.8)
-text(x = -0.8, y = lh3, labels = "2011", col = pal[2], cex = 0.8)
-text(x = -0.8, y = lh4, labels = "2012", col = pal[4], cex = 0.8)
-@
-\caption{\label{fig:axams-dist}Total precipitation predictions by a distributional forest
-at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
-Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
-\end{figure}
-
-In weather forecasting, these properties are especially appealing in 
-mountainous regions and complex terrain where a wide range of local-scale
-effects are not yet resolved by the NWP models. 
-Thus, effects with abrupt changes and possibly nonlinear interactions 
-might be required to account for site-specific unresolved features.
-To illustrate this in practice,
-precipitation forecasts are obtained with distributional forests at 95
-meteorological stations in a mountainous region in the Alps, covering mainly
-Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
-More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
-zero-censored point mass describes the probability of observing no precipitation 
-on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
-established based on data from the same month over the years 1985--2012 including
-80~covariates derived from a wide range of different NWP quantities. As
-Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
-distributional forecast for each day -- here for one specific day (July~24) at one
-station (Axams) over four years (2009--2012) -- based on the previous 24~years
-as learning data. The corresponding observations conform reasonably well with
-the predictions. In Section~\ref{sec:precipitation} we investigate the
-performance of distributional forests in this forecasting task in more detail.
-Compared to three alternative zero-censored Gaussian models distributional forests
-perform at least on par and sometimes clearly better while requiring no expert
-knowledge for the model specification. The three alternatives are: a standard ensemble model
-output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
-on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
-\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
-\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
-\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
-80~available regressors.
-
-\section{Methodology}
-\label{sec:methods}
-
-To embed the distributional approach from GAMLSS into regression trees and random
-forests, we proceed in three steps. (1)~To fix notation, we briefly review fitting
-distributions using standard maximum likelihood in Section~\ref{sec:distfit}.
-(2)~A recursive partitioning strategy based on the corresponding scores (or
-gradients) is introduced in Section~\ref{sec:disttree}, leading to distributional
-trees. (3)~Ensembles of distributional trees fitted to randomized subsamples are
-employed to establish distributional forests in Section~\ref{sec:distforest}.
-
-The general distributional notation is exemplified in all three steps for the
-Gaussian distribution, left-censored at zero. The latter is employed in the empirical 
-case study in Section~\ref{sec:precipitation} to model power-transformed daily 
-precipitation amounts.
-
-\subsection{Distributional fit}
-\label{sec:distfit}
-
-A distributional model $\mathcal{D}(Y, \bm{\theta})$ is considered for the response
-variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
-\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$ and
-corresponding log-likelihood function $\ell(\bm{\theta}; Y)$. The GAMLSS
-framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
-distributional families with parameterizations corresponding to location, scale,
-and shape. Furthermore, censoring and/or truncation of these distributions
-can be incorporated in the usual straightforward way
-\citep[see e.g.,][Chapter~7.2]{Long:1997}.
-
-To capture both location and scale of the probabilistic precipitation forecasts
-while accounting for a point mass at zero (i.e., dry days without
-rain), a Gaussian distribution left-censored at zero (for short: zero-censored Gaussian)
-with location parameter $\mu$ and scale parameter $\sigma$ is employed. Therefore,
-the corresponding log-likelihood function with parameter vector 
-$\bm{\theta} = (\mu, \sigma)$ is
-$$
-\ell(\mu, \sigma; Y) = 
-\begin{cases}
-    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
-    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
-\end{cases}
-$$
-
-where $\phi$ and $\Phi$ are the probability density function and the 
-distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
-Other distributions $\mathcal{D}$ and corresponding log-likelihoods
-$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
-censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
-zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.
-
-With the specification of the distribution family and its log-likelihood 
-function the task of fitting a distributional model turns into the task 
-of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
-maximum likelihood (ML) based on the learning sample with observations
-$\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
-likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
-$$ \label{eq:mle}
-\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
-$$
-Equivalently, this can be defined based on the corresponding first-order
-conditions
-$$ \label{eq:foc}
-\sum_{i = 1}^n s(\bm{\hat{\theta}}, y_i) = 0,
-$$
-where $s(\bm{\theta}; y_i)$ is the associated score function
-$$ \label{eq:score}
-s(\bm{\theta}; y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
-$$ 
-The latter is subsequently employed as a general goodness-of-fit measure
-to assess how well the distribution with parameters $\bm{\theta}$ fits
-one individual observation $y_i$.
-
-\subsection{Distributional tree}
-\label{sec:disttree}
-
-Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
-for reasonably representing the response distribution and covariates
-$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
-employed to capture differences in the distribution parameters $\bm{\theta}$.
-In weather forecasting, these covariates typically include the output
-from numerical weather prediction systems and/or lagged meteorological observations.
-
-To incorporate the covariates into the distributional model, 
-they are considered as regressors in additive predictors
-$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
-in GAMLSS.
-Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
-($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
-nonlinear effects, spatial effects, random coefficients, or interaction surfaces
-\citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
-additive terms and their functional forms in advance which can be challenging in
-practice and potentially require domain knowledge, especially if the number of covariates $m$ is large.
-
-Regression trees generally take a different approach for automatically
-including covariates in a data-driven way and allowing for abrupt
-changes, nonlinear and non-additive effects, and interactions. In the context of
-distributional models the goal is to partition the covariate space
-$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
-distributional model for the response $Y$ can be found with segment-specific
-parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
-($b = 1, \dots, B$) partition the covariate space
-$$ \label{eq:partition}
-\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
-$$
-and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
-(i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
-response $Y$ in each segment.
-
-To find the segments $\mathcal{B}_b$ that are (approximately) homogenous
-with respect to the distributional model with given parameters, the idea
-is to use a gradient-based recursive-partitioning approach. In a given
-subsample of the learning data this fits the model by ML (see
-Equation~\ref{eq:mle}) and then assesses the goodness of fit by assessing the
-corresponding scores $s(\bm{\hat \theta}; y_i)$ (see Equation~\ref{eq:score}).
-
-To sum up, distributional trees are fitted recursively via:
-\begin{enumerate}
-\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
-  in the current subsample.
-\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
-  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
-\item Split the sample along the partitioning variable $Z_l^*$ with the
-  strongest association or instability. Choose the breakpoint with the highest
-  improvement in the log-likelihood or the highest discrepancy.
-\item Repeat steps 1--3 recursively in the subsamples until these become too
-  small or there is no significant association/instability (or some other
-  stopping criterion is reached).
-\end{enumerate}
-%
-Different inference techniques can be used for assessing the association between
-scores and covariates in step~3. In the following we use the general class of
-permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
-also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
-Alternatively, one could use asymptotic M-fluctuation tests for parameter
-instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
-\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
-Appendix~\ref{app:tree}.
-
-For obtaining probabilistic predictions from the tree for a (possibly new) set
-of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
-be ``sent down'' the tree and the corresponding segment-specific MLE has to be
-obtained. This can also be understood as a weighted MLE where the weights
-select those observations from the learning sample that fall into the same
-segment:
-$$
-w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
-$$
-where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
-for a given $\bold{z}$ is then fully specified by the estimated parameter
-$\bm{\hat \theta}(\bold{z})$ where
-$$
-\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
-$$
-
-\subsection{Distributional forest}
-\label{sec:distforest}
-
-While the simple recursive structure of a tree model is easy to visualize and
-interpret, the abrupt changes are often too rough, instable, and impose steps
-on the model even if the true underlying effect is smooth. Hence, ensemble
-methods such as bagging or random forests \citep{Breiman:2001} are typically
-applied to smooth the effects, stabilize the model, and improve predictive performance.
-
-While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
-that pick up changes in the location of the response across the covariates,
-\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. These
-pick up changes in the ``direction'' of any distribution parameter by
-considering the full score vector for choosing splitting variables and split
-points. Each of the distributional trees is grown on a different data set
-obtained through bootstrap sampling (or subsampling) and in each node only a
-random subset of the covariates $\bold{Z}$ is considered. As usual in random
-forests, this reduces the correlation among the trees and stabilizes the
-variance of the model.
-
-To obtain probabilistic predictions from a distributional forest, it still
-needs to be specified how to compute the parameter estimates
-$\bm{\hat \theta}(\bold{z})$ for a (potentially new) set of covariates $\bold{z}$.
-Following \cite{Hothorn+Zeileis:2017} we interpret random forests as adaptive
-local likelihood estimators using the averaged ``nearest neighbor weights''
-\citep{Lin+Jeon:2006} from the $T$~trees in the forest
-$$
-w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
-\frac{\mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
-$$
-where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
-segment of the $t$-th tree.
-Thus, these $w^{\text{forest}}_i(\bold{z}) \in [0, 1]$ whereas
-$w^{\text{tree}}_i(\bold{z}) \in \{0, 1\}$. Hence, weights cannot only be $0$
-or $1$ but change more smoothly, giving high weight to those observations $i$
-from the learning sample that co-occur in the same segment $\mathcal{B}_b^t$
-as the new observation $\bold{z}$ for many of the trees $t = 1, \dots, T$.
-Consequently, the parameter estimates may, in principle, change for every
-observation and can be obtained by
-$$
-\bm{\hat \theta}(\bold{z}) =  \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
-$$
-In summary, this yields a parametric distributional regression model 
-(through the score-based approach) that can capture both abrupt effects and
-high-order interactions (through the trees) and smooth effects (through
-the forest).
-
-Distributional forests share some concepts and algorithmic aspects 
-with other generalizations of Breiman and Cutler's random forests. Nearest 
-neighbor weights are employed for aggregation in survival forests 
-\citep{Hothorn+Lausen+Benner:2004},
-quantile regression forests \citep{Meinshausen:2006},
-transformation forests \citep{Hothorn+Zeileis:2017},
-and generalized random forests for causal inferences 
-\citep{Athey+Tibshirani+Wager:2017}.
-These procedures aggregate over trees fitted to specific score functions 
-(e.g., log rank scores in survival trees, model residuals in transformation 
-or generalized forests). Distributional forests, in contrast to these 
-nonparametric approaches, provide a compromise between model flexibility 
-and interpretability: The parameters of a problem-specific distribution 
-(zero-censored Gaussian for precipitation) have a clear meaning but 
-may depend on external variables in a quite general way.
-
-\section{Probabilistic precipitation forecasting in complex terrain}
-\label{sec:precipitation}
-
-Many statistical weather forecasting models leverage the strengths of modern
-numerical weather prediction (NWP) systems
-\citep[see][]{Bauer+Thorpe+Brunet:2015}.  
-One frequently used method based on
-distributional regression models is the ensemble model output statistics 
-(EMOS) approach first proposed by
-\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
-specific quantities and sites.  
-In case of precipitation forecasting EMOS typically uses the ensemble mean of 
-``total precipitation'' (\emph{tp}) forecasts as predictor for the location parameter and
-the corresponding ensemble standard deviation for the scale part of the statistical model
-to correct for possible errors of the ensemble in both, the expectation but also
-the uncertainty of a specific forecast.
-
-While this approach alone is already highly effective in the plains,
-it typically does not perform as well in complex terrain due to unresolved
-effects in the NWP system. For example, in the Tyrolean Alps -- considered
-in the following case study -- the NWP grid cells of $50 \times 50$
-km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
-Therefore, it is often possible to substantially improve the predictive 
-performance of the EMOS by including additional predictor variables, either 
-from local meteorological observations or an NWP model. Unfortunately, it is typically unknown 
-which variables are relevant for improving the predictions. Simply including
-all available variables may be computationally burdensome and
-can lead to overfitting but, on the other hand, excluding too many variables
-may result in a loss of valuable information. Therefore, selecting
-the relevant variables and interactions among all possible covariates is 
-crucial for improving the statistical forecasting model.
-
-In the following, it is illustrated how distributional forests can
-solve this problem without requiring prior expert knowledge.  For fitting the
-forest only the response distribution and the list of potential predictor
-variables need to be specified (along with a few algorithmic details) and then
-the relevant variables, interactions, and potentially nonlinear effects are
-determined automatically in a data-driven way. Here, we employ a zero-censored
-Gaussian distribution and 80~predictor variables computed from ensemble means
-and spreads of various NWP outputs. The predictive performance of the forest is
-compared to three other zero-censored Gaussian models: (a)~a standard basic
-EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
-meteorological knowledge/experience, and (c)~a boosted GAMLSS with automatic
-selection of smooth additive terms based on all 80~predictor variables.
-
-\subsection{Data}
-\label{sec:data}
-
-\begin{table}[t!]
-%\resizebox{1.4\textwidth}{!}{
-\begin{minipage}{\textwidth}
-\begin{tabular}{ l  c  l }
-\hline
-Basic covariates & {\#} & Variations\\
-\hline
-\emph{tp}: total precipitation,           & 12  & ensemble mean of sums over 24h, \\
-\hspace*{0.5cm} power transformed (by $1.6^{-1}$) &    & ensemble std. deviation of sums over 24h, \\ 
-\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
-\hspace*{0.9cm} potential energy, &    & ensemble maximum of sums over 24h\\
-\hspace*{0.9cm} power transformed (by $1.6^{-1}$)&    & \qquad all for 6--30 \\
-\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
-\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
-\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
-\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
-\hline
-\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
-\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimum values$^*$,\\
-\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
-\emph{pwat}: precipitable water         &   & ensemble std. deviation of mean values,\\
-\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimum values$^*$,\\
-\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
-\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 \\
-\hspace*{0.95cm} condensate               &   & \\
-\emph{t500}: temperature on 500 hPa     &   & \\
-\hspace*{1cm}                             &   & \\
-\emph{t700}: temperature on 700 hPa     &   & \\
-\hspace*{1cm}                             &   & \\
-\emph{t850}: temperature on 850 hPa     &   & \\
-\hspace{1cm}                              &   & \\
-\hline
-\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
-\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
-\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
-\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 \\
-\emph{tdiff700850}: temperature         &   & \\
-\hspace*{\fill} difference 700 to 850 hPa &   & \\
-\hline
-\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
-\hspace{1.6cm} difference            &   & \qquad over 6--30 \\
-\hline
-\end{tabular}
-\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
-Time periods indicate aggregation time periods in hours after NWP model
-initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
-0600\,UTC to 0600\,UTC of the next day). $^*$Minimum values of \emph{dswrf}
-over 24\,h are always zero and thus neglected.}
-\label{tab:covariates}
-\end{minipage} 
-%}
-\end{table}
-
-Learning and validation data consist of observed 
-daily precipitation sums provided by the National Hydrographical Service 
-(\citealt{ehyd}) and numerical weather forecasts from the U.S.~National
-Oceanic and Atmospheric Administration (NOAA).
-Both, observations and forecasts are available for 1985--2012 and
-the analysis is exemplified using July, the month with the most precipitation
-in Tyrol.
-
-Observations are obtained for 95~stations all over Tyrol and surroundings,
-providing 24-hour precipitation sums measured at 0600\,UTC and rigorously
-quality-checked by the National Hydrographical Service. NWP outputs
-are obtained from the second generation reforecast data set of the
-global ensemble forecast system \citep[GEFS,][]{Hamill+Bates+Whitaker:2013}. 
-This data set consists of an 11-member ensemble based on a fixed version of the 
-numerical model and a horizontal grid spacing of about {$50 \times 50$ km$^2$}
-initialized daily at 0000~UTC from December 1984 to present 
-providing forecasts on a 6-hourly temporal resolution. Each of the 
-11 ensemble members uses slightly different perturbed initial 
-conditions to predict the situation-specific uncertainty 
-of the atmospheric state.
-
-From the GEFS, 14 basic forecast variables are considered with up 
-to 12 variations each such as mean/maximum/minimum
-over different aggregation time periods. A detailed overview is
-provided in Table~\ref{tab:covariates}, yielding 80~predictor variables in total.
-
-To remove large parts of the skewness of precipitation data, 
-a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
-cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
-However, the power parameter may vary for different climatic zones or temporal 
-aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
-in their choice of $1.6^{-1}$ as a suitable power parameter for the
-region of Tyrol. The same power transformation is applied to both the
-observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
-``convective available potential energy'' (\emph{cape}).
-
-\subsection{Models and evaluation}
-\label{sec:evaluation}
-
-\begin{table}[t!]
-\centering
-\begin{tabular}{ l l l l }
-\hline
-Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$)                                  \\ \hline
-Distributional forest & recursive    & all                & all                           \\ 
-                      & partitioning &                    &                               \\ \hline
-EMOS                  & linear       & \emph{tp{\_}mean}  & \emph{tp{\_}sprd}             \\ \hline
-Prespecified GAMLSS   & spline       & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},            \\
-                      & in each      & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean}, \\
- & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
- & & \quad \emph{cape{\_}mean1218}, & \quad \emph{cape{\_}mean1218},\\
- & & \emph{dswrf{\_}mean{\_}mean}, & \emph{tcolc{\_}sprd{\_}mean},\\
- & & \emph{tcolc{\_}mean{\_}mean}, & \emph{tdiff500850{\_}mean}\\
- & & \emph{pwat{\_}mean{\_}mean}, & \\
- & & \emph{tdiff500850{\_}mean}, & \\
- & & \emph{msl{\_}diff} & \\ \hline
-Boosted GAMLSS        & spline  & all & all \\
-                      & in each &     &     \\ \hline
-\end{tabular}
-\caption[Table caption text]{Overview of models with type of 
-covariate dependency and included covariates for each distribution 
-parameter. \emph{A} $\ast$ \emph{B} indicates an interaction between
-covariate \emph{A} and \emph{B}.}
-\label{tab:models}
-\end{table}
-
-The following zero-censored Gaussian regression models are employed
-in the empirical case study, see Table~\ref{tab:models} for
-further details:
-%
-\begin{itemize}
-
-\item \emph{Distributional forest:} All 80~predictor variables are
-  considered for learning a forest of 100 trees. Bootstrap
-  sampling is employed for each tree using a third of the predictors
-  in each split of the tree (``mtry''). Parameters are estimated by
-  adaptive local likelihood based on the forest weights as described
-  in Section~\ref{sec:methods}.
-
-\item \emph{EMOS:} Standard ensemble model output statistics
-  models use the ensemble mean of total precipitation as regressor in the location
-  submodel and the corresponding ensemble standard deviation in the scale submodel.
-  The parameters are estimated by maximum likelihood, using an identity
-  link for the location part and a log link for the scale part
-  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.
-
-\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
-  the most relevant predictors based on meteorological expert knowledge
-  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
-  on the 80~available variables, eight terms are included in the location
-  submodel and five in the scale submodel. Both involve an interaction of
-  \emph{tp} and \emph{cape} in the afternoon (between 1200\,UTC and 1800\,UTC)
-  to capture the potential for thunderstorms that frequently occur in
-  summer afternoons in the Alps. The model is estimated by maximum
-  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.
-
-\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
-  automatically from all 80~available variables, using non-cyclic boosting
-  for parameter estimation \citep{Hofner+Mayr+Schmid:2016,Messner+Mayr+Zeileis:2017}.
-  This updates the predictor terms for the location or scale submodels iteratively
-  by maximizing the log-likelihood only for the variable yielding the
-  biggest improvement. The iteration stops early -- before fully maximizing
-  the in-sample likelihood -- based on a (computationally intensive)
-  out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
-  the number of boosting iterations (``mstop'') is: $50, 75, \dots, 975, 1000$.
-
-\end{itemize}
-
-The predictive performance in terms of full probabilistic forecasts is
-assessed using the continuous ranked probability score (CRPS, \citealt{Hersbach:2000}).
-For each of the models this assesses the discrepancy of the
-predicted distribution function $F$ from the observation $y$ by
-$$
-\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
-$$
-where \(\mathbf{1}(\cdot)\) is the indicator function. In the subsequent
-applications, the mean CRPS is always evaluated out of sample, 
-either using cross-validation or a hold-out data set (2009--2012)
-that was not used for learning (1985--2008). CRPS is a proper scoring rule
-\citep{Gneiting+Raftery:2007} often used within the meteorological community.
-Lower values indicate better performance.
-
-To assess differences in the improvement of the forests and GAMLSS
-models over the basic EMOS, a CRPS-based skill score with EMOS as
-the reference method is computed:
-$$
-\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{EMOS}}}.
-$$
-
-\subsection{Application for one station}
-\label{sec:axams}
-
-\begin{figure}[t!]
-\centering
-\setkeys{Gin}{width=0.99\textwidth}
-<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
-# out of sample
-set.seed(7)
-par(mfrow = c(2, 2))
-qqrplot(pit_df, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_ml, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_g,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_gb, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-@
-\caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
-for station Axams based on models learned on data from 1985--2008.}
-\end{figure}
-
-In a first step, we show a detailed comparison of the competing models
-for one observation site, Axams in Tyrol (in the center of the study area,
-see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation observations and
-numerical weather predictions are available for the month of July from 1985
-through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
-forecasts from the distributional forest, trained on 1985--2008, for July 24
-in 2009--2012 have already been shown as a motivational example. The figure
-shows that the model properly
-depicts the point masses at zero (i.e., the probability of a dry day) and
-the forecasted probability density function for the total amount of precipitation.
-The four sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
-censoring while conforming quite well with the actual observations
-from these days. While this is a nice illustrative example we are  
-interested in the overall predictive performance and calibration of the
-distributional fits.
-
-To assess calibration Figure~\ref{fig:axams-qq} shows residual QQ~plots for
-out-of-sample predictions (2009--2012) from the different models trained on
-1985--2008. Due to the point masses at zero 100 draws from the randomized
-quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
-gray. Overall, the randomized quantile residuals conform quite well with the
-theoretical standard normal quantile (i.e., form a straight line close to the 
-diagonal), indicating that all four models are sufficiently well calibrated. 
-This is also supported by the corresponding probability integral transform
-(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) histograms in
-Appendix~\ref{app:pit}.
-
-<<echo=FALSE, results=hide>>=
-#### cross validation rain
-if(file.exists("crps_cross.rda")){
-  load("crps_cross.rda")
-} else {
-
-  nrep_cross <- 10
-  seed <- 7
-
-  res_cross <- mclapply(1:nrep_cross,
-                        function(i){
-
-                          set.seed(seed*i)
-
-                          # randomly split data in 7 parts each including 4 years
-                          years <- 1985:2012
-                          testyears <- list()
-                          for(j in 1:7){
-                            testyears[[j]] <- sample(years, 4, replace = FALSE)
-                            years <- years[!(years %in% testyears[[j]])]
-                          }
-
-                          #crps <- matrix(nrow = 7, ncol = 7)
-                          reslist <- list()
-                          for(k in 1:7){
-                            test <- testyears[[k]]
-                            train <- c(1985:2012)[!c(1985:2012) %in% test]
-
-                            res <- evalmodels(station = "Axams",
-                                              train = train,
-                                              test = test,
-                                              gamboost_cvr = TRUE)
-
-                            #crps[k,] <- res$crps
-                            reslist[[k]] <- res
-                          }
-
-                          #colnames(crps) <- names(res$crps)
-                          return(reslist)
-                        },
-                        mc.cores = detectCores() - 1
-  )
-
-  # extract CRPS
-  crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
-  # loop over all repetitions
-  for(i in 1:length(res_cross)){
-    #loop over all 7 folds (for 7 methods)
-    crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
-    for(j in 1:length(res_cross[[1]])){
-      crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
-    }
-    crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
-  }
-  colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps)
-
-  save(crps_cross, file = "crps_cross.rda")
-  save(res_cross, file = "res_cross.rda")
-
-}
-@
-
-\begin{figure}[t!]
-\centering
-<<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 7>>=
-par(mar = c(2.5,2,1,2))
-boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
-        names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
-        ylab = "CRPS skill score", col = "lightgray") 
-abline(h = 0, col = pal[5], lwd = 2)
-@
-\caption{\label{fig:axams-crps}CRPS skill score from the 10 times 
-7-fold cross validation at station Axams (1985--2012). The horizontal orange 
-line pertains to the reference model EMOS.}
-\end{figure}
-
-To assess the predictive performance, a full cross-validation is carried out
-rather than relying on just the one fixed test set for the years 2009--2012.
-To do so, a 10 times 7-fold cross-validation is carried out where each
-splits the available 28~years into 7~subsets of 4~randomly selected years.
-The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
-(= 4~years) using the average CRPS across all observations.
-The resulting 10~CRPS skill scores are displayed by boxplots in
-Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
-line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
-perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
-of around 4~percent, the distributional forest has a slightly higher improvement
-of around 5.5~percent in median.
-
-\begin{figure}[t!]
-\centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
-par(mar = c(2.5,10,1,2))
-barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
-        horiz = TRUE, las = 1, axes = FALSE,
-        xlab = "Variable importance: mean decrease in CRPS",
-        font.axis = 3, #list(family="HersheySerif", face=3),
-        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
-axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
-@
-\caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
-in the distributional forest. Based on data for station Axams, learning period 
-1985--2008 and assessed in 2009--2012.}
-\end{figure}
-
-Finally, it is of interest how this improvement in predictive performance
-by the distributional forest is accomplished, i.e., which of the 80~covariates
-are selected in the trees of the forest. As the 100~trees of the forest
-do not allow to simply assess the variables' role graphically, a common
-solution for random forests in general is to consider variable importance
-measures. Here, this is defined as the amount of change in CRPS when
-the association between one covariate and the response variable is
-artificially broken through permutation (and thus also breaking the
-association to the remaining covariates).
-
-Figure~\ref{fig:axams-varimp} shows the 10~covariates with the highest permutation
-importance (i.e., change in CRPS) for station Axams. As expected the NWP outputs
-for total precipitation (\emph{tp}) are particularly important along with total
-column-integrated condensate (\emph{tcolc}). Also, both variables occur
-in various transformations such as means (either of the full day or certain parts
-of the afternoon), spreads, or minima/maxima. Thus, while the covariates
-themselves are not surprising, selecting a GAMLSS with a particular combination
-of all the transformations would be much more challenging.
-
-%\pagebreak
-
-\subsection{Application for all stations}
-\label{sec:all}
-
-<<echo = FALSE>>=
-#### prediction over all stations 24 - 4
-if(file.exists("crps_24to4_all.rda")){
-  load("crps_24to4_all.rda")
-} else {
-
-  data("StationsTyrol")
-  stations <- StationsTyrol$name
-  test <- 2009:2012
-  train <- 1985:2008
-
-  res_24to4_all <- mclapply(1:length(stations),
-                            function(i){
-
-                              set.seed(7)
-
-                              res <- evalmodels(station = stations[i],
-                                                train = train,
-                                                test = test,
-                                                gamboost_cvr = TRUE)
-
-                              return(res)
-                            },
-                            mc.cores = detectCores() - 1
-  )
-
-  # extract crps
-  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
-  # loop over all stations
-  for(i in 1:length(stations)){
-    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
-  }
-
-  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
-  rownames(crps_24to4_all) <- stations
-
-  save(crps_24to4_all, file = "crps_24to4_all.rda")
-  save(res_24to4_all, file = "res_24to4_all.rda")
-
-}
-
-# skill score
-s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
-colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
-
-## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
-
-crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]
-
-# best method
-bst <- apply(crps_map, 1, which.min)
-
-# distance of forest to best other method
-dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]
-
-# breaks/groups
-brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
-#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
-grp <- cut(dst, breaks = brk)
-
-# HCL colors (relatively flashy, essentially CARTO Tropic)
-clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
-
-library("raster") # dem (digital elevation model)
-library("sp")     # gadm www.gadm.org/country
-
-#load("plot_map_rain/data/tirol.rda")
-#load("plot_map_rain/data/dem.rda")
-
-data(StationsTyrol, package = "RainTyrol")
-data(MapTyrol, package = "RainTyrol")
-# data(MapTyrol_border, package = "RainTyrol")
-# Create SpatialPointsDataFrame from station list
-sp <- SpatialPointsDataFrame(subset(StationsTyrol,
-                                    select=c(lon,lat)),
-                             data = subset(StationsTyrol,
-                                           select = -c(lon,lat)),
-                             proj4string = crs(MapTyrol$RasterLayer))
-@
-
-\begin{figure}[p!]
-\centering
-<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, height = 4.5, width = 7>>=
-par(mar = c(3.1, 4.1, 1.1, 2.1))
-matplot(t(s[,]), type = "l", lwd = 2, 
-          col = gray(0.5, alpha = 0.2),
-          lty = 1, axes = FALSE, 
-          xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
-lines(s[70,], col = pal[2], type = "o", pch = 19, lwd = 2)  
-# Station Axams is the 77th station which is the 70th in the list of complete stations
-  boxplot(s, add = TRUE, col = "transparent")
-  abline(h = 0, col = pal[5], lwd = 2)
-
-@
-\caption{\label{fig:all-crps}CRPS skill score for each station (gray lines
-with boxplots superimposed). Station Axams is highlighted in green 
-and the horizontal orange line pertains to the reference model EMOS. The models are 
-learned on 1985--2008 and validated for 2009--2012.}
-
-\vspace*{0.4cm}
-
-\setkeys{Gin}{width=0.99\textwidth}
-<<map, fig=TRUE, echo=FALSE, height=6.1, width=9>>=
-
-  ## plot map of Tyrol with all 95 observations
-  layout(cbind(1, 2), width = c(9, 1))
-  par(mar = c(5,4,4,0.1))
-  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
-                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
-                xlim = c(9.8,13.2), 
-                ylim = c(46.6, 47.87))
-  plot(MapTyrol$SpatialPolygons, add = TRUE)
-  points(sp[70,], pch = 19, col = "black", cex = 1.85)
-  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
-  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
-  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
-  par(mar = c(0.5,0.2,0.5,2.3))
-  ## legend
-  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
-       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
-  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
-  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))
-
-@
-\caption{\label{fig:all-map}Map of Tyrol coding the best-performing model
-for each station (type of symbol). The color codes whether the distributional 
-forest had higher (green) or lower (red) CRPS compared to the best of the 
-other three models. The background shows the local topography \citep{Robinson:2014}.
-Station Axams is highlighted in bold.}
-\end{figure}
-
-After considering only one observational site up to now, a second step
-evaluates and compares the competing methods on all 95 available stations. As in the previous
-section, all models are learned on the first 24~years and evaluated by the
-average CRPS on the last 4~years. More specifically, the CRPS skill score
-against the EMOS model is computed for the out-of-sample predictions at each
-station and visualized by parallel coordinates plots with boxplots superimposed
-in Figure~\ref{fig:all-crps}. Overall, distributional forests have a slightly
-higher improvement in CRPSS compared to the two GAMLSS which is best seen
-by looking at the boxplots and the green line representing the results
-for station Axams. The underlying parallel coordinates additionally
-bring out that the prespecified GAMLSS sometimes performs rather differently
-(sometimes better, sometimes worse) compared to the two data-driven models.
-Values below zero show that, for some stations, EMOS performs better
-than the more complex statistical methods.
-%NAs in gamlss in 13 stations!!!\\
-%for example in 13th station in complete{\_}stations\\
-%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
-%\text{NA's in the working vector or weights for parameter mu}
-%\\
-% sum(is.na(crps[,3]))
-%<<eval = TRUE, echo = FALSE>>=
-
-To assess whether these differences in predictive performance are due to differences
-in the topography, Figure~\ref{fig:all-map} shows a brief spatial summary of all
-stations. Each station is illustrated by a symbol that conveys which
-model performed best in terms of CRPS on the last 4~years of the data.
-Additionally, the color of the symbol indicates the CRPS difference between 
-distributional forest and the best-performing other model. Green signals that
-the distributional forest performs better than the other models whereas red
-signals that another model performs better. Overall the distributional forest
-performs on par (gray) or better (green) for the majority of stations. Only
-for a few stations in the north-east EMOS performs best, and in East Tyrol
-the prespecified GAMLSS performs particularly well.
-
-\section{Discussion}
-\label{sec:discussion}
-
-Distributional regression modeling is combined with tree-based modeling
-to obtain a novel and flexible method for probabilistic forecasting.
-The resulting distributional trees and forests can capture abrupt and
-nonlinear effects and interactions in a data-driven way. By basing the
-split point and split variable selection on a full likelihood and
-corresponding score function, the trees and forests can not only pick
-up changes in the location but also the scale or shape of any distributional
-family.
-
-Distributional forests are an attractive alternative when prespecifying or
-boosting all possible effects and interactions in a GAMLSS model is
-challenging.  Distributional forests are rather straightforward to specify
-requiring only little prior subject matter knowledge and also work well in the
-presence of many potential covariates. The application to precipitation
-forecasting in complex terrain illustrates that distributional forests often
-perform on par or even better than their GAMLSS counterparts. Hence, they form
-a useful addition to the already available toolbox of probabilistic forecasts
-for disciplines such as meteorology.
-
-\section*{Computational details}
-
-The proposed methods are in the \textsf{R} package \textbf{disttree}
-(version~\Sexpr{packageVersion("disttree")}) based on the \textbf{partykit}
-package (version~\Sexpr{packageVersion("partykit")}), both available 
-on \textsf{R}-Forge at
-(\url{https://R-Forge.R-project.org/projects/partykit/}). The function
-\code{distfit} fits distributional models by maximum likelihood, which is
-used as the basis for the tree-building function \code{disttree}, upon which
-the \code{distforest} is built. All functions can either be used with
-GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
-(\citealt{Stasinopoulos+Rigby:2007}, version~\Sexpr{packageVersion("gamlss.dist")})
-or with custom lists containing all required information about the distribution family.
-
-In addition to \textbf{disttree}, Section~\ref{sec:precipitation} employs
-\textsf{R} package \textbf{crch} 
-\citep[version~\Sexpr{packageVersion("crch")}]{Messner+Mayr+Zeileis:2016} for the EMOS models,
-\textbf{gamlss}
-\citep[version~\Sexpr{packageVersion("gamlss")}]{Stasinopoulos+Rigby:2007} for the prespecified GAMLSS, and
-\textbf{gamboostLSS}
-\citep[version~\Sexpr{packageVersion("gamboostLSS")}]{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
-
-The fitted distributional forest for July 24 and
-observation station Axams (including Figure~\ref{fig:axams-dist})
-is reproducible using 
-\code{demo("RainAxams", package = "disttree")}.
-This also includes fitting the other zero-censored
-Gaussian models considered in this paper and generating the corresponding 
-QQ~plots (Figure~\ref{fig:axams-qq}) and
-PIT histograms (Figure~\ref{fig:axams-pit}).
-Full replication of all results can be obtained with
-\code{demo("RainTyrol", package = "disttree")}
-requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
-also available within the \textsf{R}-Forge~project.
-
-\bibliography{ref.bib}
-
-\newpage
-\begin{appendix}
-
-\section{Tree algorithm}
-\label{app:tree}
-
-In the following, the tree algorithm applied in the empirical case study 
-discussed in this paper is explained. For notational simplicity, 
-the testing and splitting procedure is described for the root node, i.e.,
-the entire learning sample with observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
-In each child node the corresponding subsample depends on 
-the foregoing split(s).
-
-After fitting a distributional model $\mathcal{D}(Y, \bm{\theta})$ to 
-the learning sample with observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
-Section~\ref{sec:distfit} the resulting estimated parameter 
-$\bm{\hat{\theta}} = 
-(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
-can be plugged in the score function $s(\bm{\theta}, Y)$.
-In that way a goodness-of-fit measurement is obtained for each
-parameter $\theta_j$ and each observation $y_i$.
-To use this information, statistical tests are employed to detect
-dependencies between the score values
-$$
-s(\bm{\hat{\theta}}, y) = 
-%\begin{pmatrix} 
-%s(\bm{\hat{\theta}}, y_1) \\
-%s(\bm{\hat{\theta}}, y_2) \\
-%\vdots \\
-%s(\bm{\hat{\theta}}, y_n)
-%\end{pmatrix} =
-\begin{pmatrix} 
-s(\bm{\hat{\theta}}, y_1)_1 & s(\bm{\hat{\theta}}, y_1)_2 & \ldots & s(\bm{\hat{\theta}}, y_1)_k\\
-\vdots & \vdots & \ddots & \vdots \\
-s(\bm{\hat{\theta}}, y_n)_1 & s(\bm{\hat{\theta}}, y_n)_2 & \ldots & s(\bm{\hat{\theta}}, y_n)_k
-\end{pmatrix}
-$$
-and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
-\begin{align*}
-H_0^l:  s(\bm{\hat{\theta}}, Y) \qquad \bot \qquad Z_l 
-\end{align*}
-
-These hypotheses are assessed with permutation tests based on
-the multivariate linear statistic
-$$
-T_{l} = \sum_{i=1}^n v_l(z_{li}) \cdot s(\bm{\hat{\theta}}, y_i).
-$$
-The type of the transformation function $v_l$ depends on 
-the type of the split variable $Z_l$. If $Z_l$ is numeric then $v_l$ 
-is simply the identity function $v_l(z_{li}) = z_{li}$. If $Z_l$ is 
-a categorical variable with $H$ categories then 
-$v_l(z_{li}) = e_H(z_{li}) = (\I(z_{ji} = 1), \ldots, \I(z_{li} = H))$ 
-such that $v_l$ is a unit vector where the element corresponding to 
-the value of $z_{li}$ is $1$. Observations with missing values are 
-excluded from the sums.
-
-With the conditional expectation $\mu_{l}$ and the covariance 
-$\Sigma_{l}$ as derived by \cite{Strasser+Weber:1999} the test 
-statistic can be standardized. The observed multivariate linear 
-statistic $t$ is mapped onto the real line by a  univariate test 
-statistic $c$. In the application of this paper a quadratic form 
-is chosen, such that
-$$
-c_{\text{quad}}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
-$$
-where $\Sigma^+$ is the Moore-Penrose inverse of $\Sigma$. Alternatively,
-the maximum of the absolute values of the standardized linear statistic
-can be considered.
-
-The smaller the $p$-value corresponding to the standardized test 
-statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
-from the assumption of independence between the scores and the split 
-variable $Z_l$.
-%
-After Bonferroni-adjusting the $p$-values it has to be assessed whether
-any of the resulting $p$-values is beneath the selected significance level. 
-If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
-is chosen as splitting variable.
-
-The breakpoint that leads to the highest discrepancy between score functions 
-in the two resulting subgroups is selected as split point. 
-This is measured by the linear statistic
-$$
-T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
-$$
-where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
-are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
-The split point is then chosen as follows:
-$$
-r^{\ast} = \argmin_{r} c(t_{l^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{l^{\ast}}^r).
-$$
-One repeats the testing and splitting procedure in each of the resulting 
-subgroups until some stopping criterion is reached. This criterion 
-can for example be a minimal number of observations in a node or a 
-minimal $p$-value for the statistical tests. In that way pre-pruning
-is applied in order to find right-sized trees and hence 
-avoid overfitting.
-
-This permutation test based tree algorithm is presented in 
-\cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
-A different framework to build a tree is provided by the MOB algorithm 
-which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).
-
-\section{PIT histograms}
-\label{app:pit}
-
-\begin{figure}[b!]
-\centering
-%\begin{subfigure}{0.8\textwidth}
-<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7>>=
-
-# out of sample
-set.seed(4)
-
-par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
-pithist(pit_df, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
-pithist(pit_ml, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
-pithist(pit_g, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
-pithist(pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
-
-@
-%\end{subfigure}
-\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
-\end{figure}
-
-As an alternative to QQ~plots based on randomized quantile residuals,
-probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
-histograms are a commonly used tool to assess how well-calibrated distributional 
-fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
-probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
-some deviations the overall calibration of all models is acceptable.
-
-\end{appendix}
-
-\end{document}
-
 \newcommand{\bS}{\mathbf{S}}

### 2018-07-09 pkg/disttree/inst/draft/draft.Rnw
added explanation to definition of parameter estimator for trees and setting of stopping criteria, based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2211 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ad6a509853f4aa584d93dbde69ad1d4218873347

@@
 $$
+Even though this parameter depends on the given $z$ it has to be noted that
+it is the same for all other observations that are in the same terminal node of 
+the tree as $z$ and therefore does not necessarily have to be recalculated for
+each of them.
 
@@
   adaptive local likelihood based on the forest weights as described
-  in Section~\ref{sec:methods}.
+  in Section~\ref{sec:methods}. As stopping criteria the minimal number
+  of observations to perform a split (\code{minsplit}) is set to 50, 
+  the minimal number of observations in a terminal node (\code{minsbucket})
+  is set to 20 and the value that must be exceeded in the statistical tests 
+  (e.g. $1-\text{p-vlaue$}) for variable selection is kept at the default value $0$.
   % stopping criteria:
-  % tree_minsplit = 50, tree_minbucket = 20, tree_mincrit = 0.95,
   % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0
+
+  % default in ctree_control within distforest:
+  % mincriterion = 0
+
+  % default values in ctree_control:
+  % alpha = 0.05, mincriterion = 1 - alpha,
+  % minsplit = 20L, minbucket = 7L,
+  % minprob = 0.01
 
@@
 \caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
-for station Axams based on models learned on data from 1985--2008.}
+for station Axams based on models learned on data from 1985--2008. 
+(Residuals of precipitation are on the powertransformed scale.)}
 \end{figure}
@@
 To do so, a 10 times 7-fold cross-validation is carried out where each
-splits the available 28~years into 7~subsets of 4~randomly selected years.
+splits the available 28~years into 7~subsets of 4~randomly selected and 
+therefore mostly not subsequent years.
 The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold

### 2018-07-15 pkg/disttree/inst/JSM2018/slides.Rnw
slides for JSM 2018
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2219 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2c56fe0e7f2977f91d1dc0720277c2e07b385288

--- 2018-06-13 pkg/disttree/inst/eeecon_workshop/slides.Rnw
@@
 
-\title{Flexible Probabilistic Forecasting with Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
-%\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
+\title{Distributional Trees and Forests}
+\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
+%\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
 \URL{https://R-Forge.R-project.org/projects/partykit/}
@@
 %% includes a replacement for \usepackage{Sweave}
-\usepackage{changepage}
 \usepackage{amsmath,tikz}
-\usepackage{calc}
-\usepackage{graphicx}
-\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
+\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
 \graphicspath{{plots/}}
+
 \newcommand{\argmax}{\operatorname{argmax}\displaylimits}
@@
 
-# setwd("~/svn/partykit/pkg/disttree/inst/eeecon_workshop")
+# setwd("~/svn/partykit/pkg/disttree/inst/JSM2018")
 library("ggplot2")
@@
 library("parallel")
-library("RainTyrol")
 library("gamlss.cens")
@@
 # if tree: 2 splits => 3 terminal nodes
-# 10 split variables (x1, ..., x10) are available (5 numeric, 2 binary, 3 categorical)
+# 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
 #
@@
 
-\begin{frame}[fragile]
-\frametitle{Statistical Modeling}
-\visible<2,3, 5-> {Assumption:}
-\vspace{-0.7cm}
-\begin{center}
-\begin{tikzpicture}
-\visible<2,3, 5-> {\node[] at (0,0.4) {X};}
-\visible<4>{
-\node[inner sep=0pt] (today) at (-1.8,-0.4)
-    {\includegraphics[width=.365\textwidth]{today.png}};
-\node[font=\scriptsize] at (-1.9,-2) {2018-03-15};
-}
-\visible<2->{
-\draw[->] (0.4,0.4)--(0.9,0.4);
-\draw[] (1,0) rectangle (2.5,0.8);
-\node[font=\scriptsize] at (1.75,0.4) {nature};
-\draw[->] (2.6,0.4)--(3.1,0.4);
-}
-\visible<2,3, 5->{
-\node[] at (3.5,0.4) {Y};
-}
-\visible<4>{
-\node[inner sep=0pt] (future) at (5.3,-0.4)
-    {\includegraphics[width=.365\textwidth]{future.png}};
-\node[font=\scriptsize] at (5.4,-2) {2018-03-16};
-}
-\end{tikzpicture}
-\end{center}
-\visible<3, 5->{
-\vspace{-2cm}
-Approximation of reality based on a set of observations.\\
-}
-\visible<4>{
-\vspace{2cm}
-X = observed precipitation of today,\\
-\qquad current temperature / humidity / \dots,\\
-\qquad physical model output\\
-\vspace{0.15cm}
-Y = daily precipitation sums\\
-}
-
-\vspace{-3.5cm}
-\visible<6->{
-\textbf{Application/Motivation:}
-\begin{itemize}
-\item Prediction.
-\item Capturing dependencies and interactions.
-\end{itemize}
-}
-\vspace{0.5cm}
-\visible<7->{
-\textbf{Two main approaches / ``cultures'':}
-\begin{itemize}
-\item Data modeling.
-\item Algorithmic modeling.
-\end{itemize}
-}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Data Modeling}
-\vspace{-0.1cm}
-\begin{center}
-\begin{tikzpicture}
-\node[] at (0,0.4) {X};
-\draw[->] (0.4,0.4)--(0.9,0.4);
-\draw[] (1,0) rectangle (2.5,0.8);
-\node[font=\scriptsize] at (1.75,0.4) {f(X; par)};
-\draw[->] (2.6,0.4)--(3.1,0.4);
-\node[] at (3.5,0.4) {Y};
-\end{tikzpicture}
-\end{center}
-\vspace{-0.4cm}
-\only<1-5>{
-\begin{itemize}
-\item Approximate data generating process up to a few unknown processes.
-\item Mathematical equations, model assumptions.
-\item Data: estimation of parameters.
-\end{itemize}
-}
-\only<6->{
-\vspace{1cm}
-Summary:
-\begin{itemize}
-\item Stochastic model to approximate data generating process.\\
-\only{$\Rightarrow$ fixed $f$.\\}
-\item Disadvantage: Risk of incorrect assumptions.
-\end{itemize}
-}
-\only<1>{\vspace{1cm}}
-
-\vspace{0.1cm}
-
-\visible<2-5>{
-\begin{minipage}{0.65\textwidth}
-\textbf{Example:} 
-\only<2-3>{Linear model}\only<4>{Generalized additive model}\only<5-6>{Generalized additive model\\ for location, scale, and shape}
-\\
-\only<2-3>{
-$$\visible<3>{\mathbb{E}(Y|X) = } f(X; \alpha) = \alpha_0 + \alpha_1 \cdot X$$  
-\vspace{0.15cm}
-}
-\only<4>{
-$$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
-\vspace{0.15cm}
-}
-\only<5>{
-\vspace{-1.29cm}
-
-\begin{align*}
-\visible<5>{\mathbb{E}(Y|X) = }\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
-\visible<5>{\sqrt{Var(Y|X)} = }\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
-\end{align*}
-}
-\end{minipage}
-%\vspace{-0.6cm}
-%\hfill
-\begin{minipage}{0.3\textwidth}
+\subsection{Motivation}
+
+\begin{frame}[fragile]
+\frametitle{Motivation}
+\vspace{-0.41cm}
+\begin{figure}[!htb]
+\minipage{0.29\textwidth}
+\begin{center}
 <<motivation_GLM, echo=FALSE, results=hide>>=
@@
 @
-\only<2-3>{
-<<plot_motivation_GLM_single, fig=TRUE, echo=FALSE>>=
+\visible<2->{
+<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
 par(mar=c(2,0,2,0))
@@
 }
+\end{center}
+\endminipage
+\visible<3->{{\LARGE$\rightarrow$}}
+\minipage{0.29\textwidth}
+\begin{center}
 <<motivation_GAM, echo=FALSE, results=hide>>=
@@
 @
-\only<4>{
+\visible<3->{
 <<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
@@
 }
+\end{center}
+\endminipage
+\visible<4->{{\LARGE$\rightarrow$}}
+\minipage{0.29\textwidth}
+\begin{center}
 <<motivation_GAMLSS, echo=FALSE, results=hide>>=
@@
 @
-\only<5>{
+\visible<4->{
 <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
@@
 @
-}
-\end{minipage}
-}
-\end{frame}
-
-<<raintree, echo=FALSE, results=hide>>=
-load("weather.rda")
-#d <- data.frame(rain24 = numeric(0), rain = logical(0), wind = numeric(0), 
-#                humidity = numeric(0), airpressure = numeric(0), temperature = numeric(0))
-
-sp_rain <- partysplit(2L, index = 1:2)
-sp_humidity <- partysplit(4L, breaks = 70)
-
-pn <- partynode(1L, split = sp_rain, kids = list(
-  partynode(2L, split = sp_humidity, kids = list(
-      partynode(3L, info = "4 mm/24h"),
-      partynode(4L, info = "9 mm/24h"))),
-  partynode(5L, info = "15 mm/24h")))
-
-p <- party(pn, weather)
-@
+
+}
+\end{center}
+\endminipage
+
+\vspace{0.5cm}
+\minipage{0.25\textwidth}
+\begin{center}
+\visible<2->{
+LM, GLM\\
+\vspace{0.5cm}
+\code{lm}\\
+\code{glm}\\
+\vspace{1.5cm}}
+\end{center}
+\endminipage
+\hspace{1.1cm}
+\minipage{0.25\textwidth}
+\begin{center}
+\visible<3->{
+GAM\\
+\vspace{0.5cm}
+\code{mgcv}\\
+\code{VGAM}\\
+$\ldots$\\
+\vspace{1cm}}
+\end{center}
+\endminipage
+\hspace{1.1cm}
+\minipage{0.25\textwidth}
+\begin{center}
+\visible<4->{
+GAMLSS\\
+\vspace{0.5cm}
+\code{gamlss}\\
+\code{mgcv}\\
+\code{VGAM}\\
+\code{gamboostLSS}\\
+$\ldots$}
+\end{center}
+\endminipage
+\end{figure}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Motivation}
+\vspace{-0.2cm}
+\begin{figure}[!htb]
+\minipage{0.29\textwidth}
+\begin{center}
 <<motivation_regtree, echo=FALSE, results=hide>>=
@@
 kappa <- 12
-nobs <- 120
 x <- c(1:nobs)/nobs
-ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- 
-  ytree_up <- ytree_up2 <- ytree_up3 <- ytree_up4 <- 
-  yforest <- numeric(length = length(x))
-for(i in 1:nobs) ytrue[i] <- if(x[i]<0.55) 4 else 4+5*plogis(kappa*(3*(x[i]-0.36)-1))
-y <- ytrue + rnorm(nobs,0,2)
-y_up <- rep.int(15, length(x)) + rnorm(nobs,0,1)
-for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
-for(i in 1:nobs) ytree2[i] <- if(x[i]<0.63) 4.3 else 9.4
-for(i in 1:nobs) ytree3[i] <- if(x[i]<0.75) 3.7 else 9.2
-for(i in 1:nobs) ytree4[i] <- if(x[i]<0.55) 4.4 else 8.7
-ytree_up <- rep.int(15, length(x))
-ytree_up2 <- rep.int(14.5, length(x))
-ytree_up3 <- rep.int(15.3, length(x))
-ytree_up4 <- rep.int(15.5, length(x))
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.63) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
-@
+ytrue <- ytree <- yforest <- numeric(length = length(x))
+for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
+y <- ytrue + rnorm(nobs,0,0.3)
+for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
+@
+\visible<1->{
+<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+lines(x = x, y = ytree, col = pal["forest"], lwd=7)
+@
+}
+\end{center}
+
+\endminipage
+\visible<2->{{\LARGE$\rightarrow$}}
+\minipage{0.29\textwidth}
+ \begin{center}
 <<motivation_randforest, echo=FALSE, results=hide>>=
 ## Random Forest
-#%for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-@
-
-\begin{frame}[fragile]
-\frametitle{Algorithmic Modeling}
-\vspace{-0.1cm}
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
+@
+\visible<2->{
+<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
+lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
+@
+}
+\end{center}
+
+\endminipage
+\visible<3->{{\LARGE$\rightarrow$}}
+\minipage{0.29\textwidth}
+  \visible<3->{
   \begin{center}
   \begin{tikzpicture}
-\node[] at (0,0.4) {X};
-\draw[->] (0.4,0.4)--(0.9,0.4);
-\draw[] (1,0) rectangle (2.5,0.8);
-\node[font=\scriptsize] at (1.75,0.4) {nature};
-\draw[->] (2.6,0.4)--(3.1,0.4);
-\node[] at (3.5,0.4) {Y};
-\draw[<->] (0.1,0.1) .. controls (0.2, -0.4) and (0.4, -0.5) .. (0.8,-0.5);
-\draw[<->] (2.7,-0.5) .. controls (3.1, -0.5) and (3.3, -0.4) .. (3.4,0.1);
-\node[font=\scriptsize] at (1.75,-0.5) {algorithm};
+  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
+  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
   \end{tikzpicture}
+  \end{center}}
+\endminipage
+
+\minipage{0.29\textwidth}
+\begin{center}
+\vspace{0.0cm}
+\visible<1->{
+Regression Tree\\
+\vspace{0.5cm}
+\resizebox{0.3\textwidth}{!}{
+\begin{tikzpicture}
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (1, 2) {};
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0.5, 1) {};
+  \draw[-, line width=1pt] (n0) -- (n1);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (1.5, 1) {};
+  \draw[-, line width=1pt] (n0) -- (n2);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (1, 0) {};
+  \draw[-, line width=1pt] (n2) -- (n3);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (2, 0) {};
+  \draw[-, line width=1pt] (n2) -- (n4);
+\end{tikzpicture}}
+\vspace{0.3cm}\\
+\code{rpart}\\
+\code{party(kit)}\\
+\vspace{1cm}}
 \end{center}
-\vspace{-0.4cm}
+\endminipage
+\hspace{0.65cm}
+\minipage{0.29\textwidth}
+\begin{center}
+\vspace{0.0cm}
 \visible<2->{
-\begin{itemize}
-\item Flexible approximation with no/few assumptions. %\\
-%$\Rightarrow$ no fixed $f$.
-\item Data-driven model fitting.
-\end{itemize}
-}
-\only<2>{\vspace{1cm}}
+Random Forest\\
+\vspace{0.5cm}
+\resizebox{0.9\textwidth}{!}{
+\begin{tikzpicture}
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n00) at (1, 2) {};
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n01) at (0.5, 1) {};
+  \draw[-, line width=1pt] (n00) -- (n01);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n02) at (1.5, 1) {};
+  \draw[-, line width=1pt] (n00) -- (n02);
+
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n10) at (3, 2) {};
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n11) at (2.5, 1) {};
+  \draw[-, line width=1pt] (n10) -- (n11);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n12) at (3.5, 1) {};
+  \draw[-, line width=1pt] (n10) -- (n12);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n13) at (2, 0) {};
+  \draw[-, line width=1pt] (n11) -- (n13);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n14) at (2.8, 0) {};
+  \draw[-, line width=1pt] (n11) -- (n14);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n15) at (3.2, 0) {};
+  \draw[-, line width=1pt] (n12) -- (n15);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n16) at (4, 0) {};
+  \draw[-, line width=1pt] (n12) -- (n16);
+
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
+  \draw[-, line width=1pt] (n20) -- (n21);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
+  \draw[-, line width=1pt] (n20) -- (n22);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
+  \draw[-, line width=1pt] (n22) -- (n23);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
+  \draw[-, line width=1pt] (n22) -- (n24);
+\end{tikzpicture}
+}
+\vspace{0.5cm}\\
+\code{randomForest}\\
+\code{ranger}\\
+\code{party(kit)}\\
+$\ldots$}
+\end{center}
+\endminipage
+\hspace{0.65cm}
+\minipage{0.28\textwidth}
+\begin{center}
+\vspace{0.68cm}
 \visible<3->{
-\begin{minipage}{0.6\textwidth}
-%\vspace{-0.5cm}
-\textbf{Example:} \only<3>{Tree model}\only<4-10>{Subsampling}\only<3>{\\}\only<11->{Forest model}\\
-%\vspace{-0.8cm}
-\begin{center}
-\only<3>{
-\vspace{-0.96cm}
-\resizebox{0.6\textwidth}{!}{
-<<plot_raintree, fig=TRUE, echo=FALSE, width=4, height=2.7>>=
-par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
-paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-plot(p,
-     tp_args = list(FUN = identity, width = 9, fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]),
-     drop_terminal = TRUE, tnex = 1)
-@
-}
-
-\vspace{-0.7cm}
-\begin{flushleft}
-\noindent {\small X = rain now, humidity\\
-Y = rain 24h}
-\end{flushleft}}
-\only<4->{
-\vspace{-0.5cm}
-\resizebox{0.55\textwidth}{!}{
-\begin{tikzpicture}
-%\draw[ellipse, draw=black, align=center, scale = 0.7, minimum width=170pt, minimum height = 35pt] (set) at (1, 3) {};
-\draw[gray,rounded corners=10pt] (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
-%\draw[gray] (1,3.5) ellipse (3 and 0.8);
-\begin{scope}
-\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
-%\clip (1,3.5) ellipse (3 and 0.8);
-\pgfmathsetseed{7}
-\foreach \p in {1,...,200} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
-\end{scope}
-\visible<5->{
-\node[ellipse, draw=black!50!green, align=center, minimum width=80pt, minimum height = 40pt, line width = 3pt] (subset1) at (0.1, 3.5) {};
-}
-\visible<7->{
-\node[ellipse, draw = blue, align=center, minimum width=100pt, minimum height = 30pt, line width = 3pt] (subset2) at (1.2, 3.6) {};
-}
-\visible<9->{
-\node[ellipse, draw = orange, align=center, minimum width=120pt, minimum height = 25pt, line width = 3pt] (subset3) at (1.6, 3.3) {};
-}
-\visible<6->{
-  \node[ellipse, fill=black!50!green, align=center] (n00) at (-1, 2) {};
-  \node[ellipse, fill=black!50!green, align=center] (n01) at (-1.25, 1.25) {};
-  \draw[-, line width=1pt] (n00) -- (n01);
-  \node[rectangle, fill=black!50!green, align=center] (n02) at (-1.5, 0.5) {};
-  \draw[-, line width=1pt] (n01) -- (n02);
-  \node[rectangle, fill=black!50!green, align=center] (n03) at (-1, 0.5) {};
-  \draw[-, line width=1pt] (n01) -- (n03);
-  \node[rectangle, fill=black!50!green, align=center] (n04) at (-0.5, 0.5) {};
-  \draw[-, line width=1pt] (n00) -- (n04);
-}
-\visible<8->{
-  \node[ellipse, fill=blue, align=center] (n10) at (1, 2) {};
-  \node[ellipse, fill=blue, align=center] (n11) at (0.75, 1.25) {};
-  \draw[-, line width=1pt] (n10) -- (n11);
-  \node[rectangle, fill=blue, align=center] (n12) at (0.5, 0.5) {};
-  \draw[-, line width=1pt] (n11) -- (n12);
-  \node[rectangle, fill=blue, align=center] (n13) at (1, 0.5) {};
-  \draw[-, line width=1pt] (n11) -- (n13);
-  \node[rectangle, fill=blue, align=center] (n14) at (1.5, 0.5) {};
-  \draw[-, line width=1pt] (n10) -- (n14);
-}
-\visible<10->{
-  \node[ellipse, fill=orange, align=center] (n20) at (3, 2) {};
-  \node[ellipse, fill=orange, align=center] (n21) at (2.75, 1.25) {};
-  \draw[-, line width=1pt] (n20) -- (n21);
-  \node[rectangle, fill=orange, align=center] (n22) at (2.5, 0.5) {};
-  \draw[-, line width=1pt] (n21) -- (n22);
-  \node[rectangle, fill=orange, align=center] (n23) at (3, 0.5) {};
-  \draw[-, line width=1pt] (n21) -- (n23);
-  \node[rectangle, fill=orange, align=center] (n24) at (3.5, 0.5) {};
-  \draw[-, line width=1pt] (n20) -- (n24);
-}
-\visible<11->{
-\draw[-, line width=1pt] (-1, -0.2) -- (0.2,-0.7);
-\draw[-, line width=1pt] (1, -0.2) -- (1,-0.7);
-\draw[-, line width=1pt] (3, -0.2) -- (1.8,-0.7);
-\node[rectangle, align=center] (combine) at (1, -1) {\large average};
-}
-\end{tikzpicture}
-}
-}
+Distributional trees and forests\\
+\vspace{1.7cm}
+\code{disttree}\\
+based on \code{partykit}\\
+\vspace{1cm}}
 \end{center}
-\end{minipage}
-\hspace{-1cm}
-\begin{minipage}{0.4\textwidth}
-\begin{center}
-\vspace{0.1cm}
-
-\only<3-5>{
-%\vspace{0.01cm}
-<<plot_motivation_regtree, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5.5))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.5, cex = 3)
-@
-}
-\only<6-7>{
-%\vspace{0.01cm}
-\hspace{-0.2cm}
-<<plot_motivation_randforest1, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5.5))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
-lines(x = x, y = ytree_up2, col = pal["tree"], lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.5, cex = 3)
-@
-}
-\only<8-9>{
-%\vspace{0.01cm}
-\hspace{-0.3cm}
-<<plot_motivation_randforest2, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5.5))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
-lines(x = x, y = ytree3, col = "blue", lwd=7)
-lines(x = x, y = ytree_up2, col = pal["tree"], lwd=7)
-lines(x = x, y = ytree_up3, col = "blue", lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.5, cex = 3)
-@
-}
-\only<10>{
-%\vspace{0.01cm}
-\hspace{-0.3cm}
-<<plot_motivation_randforest3, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5.5))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
-lines(x = x, y = ytree2, col = pal["tree"], lwd=7)
-lines(x = x, y = ytree3, col = "blue", lwd=7)
-lines(x = x, y = ytree_up2, col = pal["tree"], lwd=7)
-lines(x = x, y = ytree_up3, col = "blue", lwd=7)
-lines(x = x, y = ytree4, col = "orange", lwd=7)
-lines(x = x, y = ytree_up4, col = "orange", lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.5, cex = 3)
-@
-}
-\only<11->{
-%\vspace{0.01cm}
-\hspace{-0.3cm}
-<<plot_motivation_randforest4, fig=TRUE, echo=FALSE, width=7>>=
-par(mar=c(3,3,2,5.5))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=7)
-lines(x = x, y = yforest, col = pal["forest"], lwd=7)
-mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
-mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
-mtext("humidity", side = 1, line = 1.5, cex = 3)
-@
-}
-\end{center}
-\end{minipage}
-}
+\endminipage
+\end{figure}
 \end{frame}
@@
 \frametitle{Goals}
-\vspace{-0.4cm}
-\begin{minipage}{0.7\textwidth}
-\textbf{Available:} \\
-\vspace{-0.5 cm}
+
+\textbf{Distributional:} \\
 \begin{itemize}
-  \item Distributional regression:\\
-  Probabilistic forecast.
-  \item Regression trees:\\
-  Automatic detection of abrupt changes,
-  and selection of variables/interactions.
-  \item Random forests:\\
-  Smoothing and regularization.
+\item Specify the complete probability distribution (including location, scale and shape).
 \end{itemize}
 
-\vspace{1cm}
-
-\visible<2->{
-\textbf{Missing:} Synthesis
-
-\vspace{0.2cm}
-
-$\Rightarrow$ Distributional trees and forests
-
-\vspace{0.1cm}
-
-}
-\end{minipage}
-<<motivation_GAMLSS_syn, echo=FALSE, results=hide>>=
-## GAMLSS
-set.seed(7)
-nobs <- 200
-x_g <- c(1:nobs)/nobs
-x_g <- 2*(x_g-0.5)
-ytrue_g <- x_g^3
-var_g <- exp(-(2*x_g)^2)/2
-y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var_g)
-@
-<<motivation_regtree_syn, echo=FALSE, results=hide>>=
-## Reg. Tree
-set.seed(74)
-nobs <- 170
-kappa <- 12
-x <- c(1:nobs)/nobs
-ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- var <- var_up <- numeric(length = length(x))
-for(i in 1:nobs) ytrue[i] <- if(x[i]<0.2) 4 else 4+5*plogis(kappa*(3*(x[i]-0.33)-1))
-ytrue_up <- rep.int(15, length(x))
-for(i in 1:nobs) var_up[i] <- if(x[i]<0.15) 3 else (if(x[i]<0.35) 2 else (if(x[i]<0.5) 2 else (if(x[i]<0.7) 1 else (if(x[i]<0.8) 1.5 else (if(x[i]<0.9) 2 else (2.5))))))
-for(i in 1:nobs) var[i] <- if(x[i]<0.15) 4 else (if(x[i]<0.2) 3 else (if(x[i]<0.25) 2 else (if(x[i]<0.4) 1 else (if(x[i]<0.8) 2 else (if(x[i]<0.9) 2.5 else (3))))))
-y <- ytrue + rnorm(nobs,0, sqrt(var))
-y_up <- ytrue_up + rnorm(nobs,0, sqrt(var_up))
-for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
-ytree_up <- rep.int(15, length(x))
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.6) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
-@
-\begin{minipage}{0.25\textwidth}
-\vspace{0.4cm}
-<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=8.5>>=
-par(mar=c(1,0,1,0))
-plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
-polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g)),
-  col = pallight["forest"], border = "transparent")
-lines(x_g, ytrue_g, col = pal["forest"], lwd=10)
-points(x_g, y_g, col = "slategray", pch = 19)
-box(lwd = 5)
-@
-
-<<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=8.5>>=
-par(mar=c(1,0,1,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree, col = pal["forest"], lwd=10)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=10)
-@
-
-<<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=8.5>>=
-par(mar=c(1,0,1,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=10)
-lines(x = x, y = yforest, col = pal["forest"], lwd=10)
-@
-
-\vspace{0.2cm}
-
-\visible<2->{
-<<plot_motivation_distforest, fig=TRUE, echo=FALSE, width=6>>=
-par(mar=c(0,0,3,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
-     ylim = c(min(c(y,y_up)), max(c(y,y_up))))
-polygon(c(x, rev(x)), c(yforest + var, rev(yforest - var)),
-  col = pallight["forest"], border = "transparent")
-polygon(c(x, rev(x)), c(ytrue_up + var_up, rev(ytrue_up - var_up)),
-  col = pallight["forest"], border = "transparent")
-points(x=x, y = y_up, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytree_up, col = pal["forest"], lwd=10)
-lines(x = x, y = yforest, col = pal["forest"], lwd=10)
-@
-}
-\end{minipage}
-\end{frame}
-
+\medskip
+
+{\bf Tree:}
+\begin{itemize}
+  \item Automatic variable selection and detection of steps and abrupt changes. %(data driven)
+  \item Capture non-linear and non-additive effects and interactions.
+\end{itemize}
+
+\medskip
+
+{\bf Forest:}
+\begin{itemize}
+  \item Smoother effects.
+  \item Stabilization and regularization of the model.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Distributional trees}
+\vspace*{-0.12cm}
+\begin{center} 
+DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
+\vspace*{-0.21cm}
 <<dgp_tree, echo=FALSE, results=hide>>=
@@
 
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+<<plottree_dgp, fig=TRUE, echo=FALSE>>=
+paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+plot(fig, inner_panel = node_inner_ext,
+     tp_args = list(FUN = identity, width = 18, fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]),
+     drop_terminal = TRUE, tnex = 1.7)
+@
+\end{center}
+\end{frame}
+
 <<tree, echo=FALSE, results=hide>>=
@@
 
-\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
-
-\begin{frame}[fragile]
-\frametitle{Distributional Tree}
-\begin{minipage}{0.78\textwidth}
-\begin{center}
-\vspace{0.2cm}
-\visible<1->{
-Model: \code{disttree(y $\sim$ x)}\\
-}
-\vspace{-0.2cm}
-\only<1>{
-%\setkeys{Gin}{width=0.\linewidth, height=0.68\linewidth}
-<<plottree_estpar, fig=TRUE, echo=FALSE, width = 7>>=
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+<<plottree_estpar, fig=TRUE, echo=FALSE>>=
 # function for output in terminal panels
@@
 @
-}
-\only<2->{
-%\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_dens, fig=TRUE, echo=FALSE, width = 7>>=
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+<<plottree_box, fig=TRUE, echo=FALSE>>=
+plot(as.constparty(tree), tnex = 1.7, drop = TRUE,
+     tp_args = list(fill = paltrees[c(1, 3)], ylines = 1.5), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+<<plottree_dens, fig=TRUE, echo=FALSE>>=
 node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
@@
 @
-}
-\visible<1->{
-R~package~\code{disttree}
-}
 \end{center}
-\end{minipage}
-\begin{minipage}{0.21\textwidth}
-
-\vspace{1.9cm}
-
-\center
-\visible<3->{
-Algorithmic \\
-Modeling\\
-}
-
-\vspace{1.3cm}
-
-\visible<4->{
-Data \\
-Modeling
-}
-
-\vspace{1.4cm}
-
-\visible<5->{
-Combined \visible<6->{\checkmark}
-}
-\end{minipage}
-\end{frame}
+\end{frame}
+%\begin{frame}
+%\frametitle{Distributional trees}
+%\center
+%\vspace*{-1cm}
+%\setkeys{Gin}{width=0.8\linewidth}
+%\includegraphics{tree_box}
+%\end{frame}
 
@@
 
-%<<densities, eval=TRUE, echo=FALSE, results=hide>>=
-%x <- c(-300:300)/100
-%set.seed(7)
-%y1 <- dnorm(x, mean = 0, sd = 0.7)
-%y2 <- dnorm(x, mean = 0.8, sd = 0.5)
-%y3 <- dnorm(x, mean = -0.7, sd = 1.2)
-%@
-
-%<<density1, fig=TRUE, echo=FALSE>>=
-%par(mar=c(0,0,0,0))
-%plot(x, y1, type = "l", col = "red", ylim = c(0,0.8), lwd=2,
-%     xaxt='n', yaxt='n')
-%@
-
-%<<density2, fig=TRUE, echo=FALSE>>=
-%par(mar=c(0,0,0,0))
-%plot(x, y2, type = "l", col = "red", ylim = c(0,0.8), lwd=2,
-%     xaxt='n', yaxt='n')
-%@
-
-%<<density3, fig=TRUE, echo=FALSE>>=
-%par(mar=c(0,0,0,0))
-%plot(x, y3, type = "l", col = "red", ylim = c(0,0.8), lwd=2,
-%     xaxt='n', yaxt='n')
-%@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 %\begin{frame}
-%\frametitle{Global Likelihood Estimation}
-%\begin{itemize}
-
-%\item Specify a parametric distribution family $F(\cdot;\theta)$ with parameter vector
-%  $\theta \in \Theta$ capturing location, scale, shape.
-
-%\item Log-likelihood function:
-%  \begin{eqnarray*}
-%  \ell(\theta; y) & = & \log(f(y; \theta))
-%  \end{eqnarray*}
-
-%\item Estimate $\hat{\bold{\theta}}$ via maximum likelihood
-%  based on a learning sample $y_1,\dots, y_n$:
-%\[
-%\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
-%\]
-
-%% equivalent to solving 
-%% $$
-%% \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
-%% $$
-%\end{itemize}
-
+%\frametitle{Distributional trees}
+%\center
+%\vspace*{-0.7cm}
+%\setkeys{Gin}{width=0.8\linewidth}
+%\includegraphics{tree_xyplot}
 %\end{frame}
 
-%\begin{frame}
-%\frametitle{Adaptive Local Likelihood Estimation}
-
-%\textbf{Idea:} Capture covariates through adaptive weights.
-%\[
-%\hat{\theta}(\bold{x}) = \argmax_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell%(\theta; y_i).
-%\]
-
-%\bigskip
-
-%\textbf{Question:} How to choose weighting function $w_i(\bold{x})$?
-
-%\bigskip
-
-%\textbf{Possible answers:} Based on learning sample $\{(y_i, \bold{x}_i)\}_{i=1, \dots, n}$ and (possibly new) observation $\bold{x}$.
-%\begin{itemize}
-%\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $\bold{x}_i$ are classified into the same subgroup.
-%\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $\bold{x}_i$ across trees.
-%\end{itemize}
-
-%\textbf{Possible answers:} Based on learning sample $y_1, \dots, y_n$ and (possibly new) observation $\bold{x}$.
-%\begin{itemize}
-%\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $y_i$ are classified into the same subgroup.
-%\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $y_i$ across trees.
-%\end{itemize}
-
-%\end{frame}
-
+\subsection{Building Distributional Trees and Forests}
 \begin{frame}
 \frametitle{Building Distributional Trees and Forests}
-\begin{minipage}{0.7\textwidth}
 {\bf Tree:}
 \begin{enumerate}
-\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
-Estimate $\hat{\theta}$ via maximum likelihood\\
-$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
-\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
+\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
+\item Estimate $\hat{\theta}$ via maximum likelihood.
+\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable.
+%\item Assess whether the \emph{model scores} are associated with
+%    (or change along) any of the available covariates -- e.g.,
+%    using parameter instability tests (\emph{strucchange}) or
+%    conditional inference (\emph{coin}).
+\item Choose the partitioning variable with the strongest association or instability as split variable.
+\item Choose the breakpoint along this variable with highest improvement in log-likelihood and split the sample.
+\item Repeat steps 2--5 recursively in the subgroups
+    until some stopping criterion is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
+    %-- e.g., for significance or sample size.
+%\item Choose the variable with the strongest association
+%\item Choose the split point which leads to the highest improvement of the model
+%\item Split and repeat 2-6 in each node until a stopping criterion is met
 \end{enumerate}
-\end{minipage}
-\begin{minipage}{0.23\textwidth}
-%\vspace{-0.1cm}
-\begin{tikzpicture}
-\visible<2-3>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
-}
-\visible<4>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{D}(Y;\hat{\theta}$)};
-}
-\visible<5->{
-\node[inner sep=0pt] (density1) at (0.8, 1.7)
-    {\includegraphics[width=0.6\textwidth]{density1.jpeg}};
-}
-\visible<6-8>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
-\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1);
-\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2);
-}
-\visible<9-10>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
-\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
-}
-\visible<11>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
-}
-\visible<11->{
-\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
-}
-\visible<12->{
-\node[inner sep=0pt] (density2) at (-0.1,-0.1)
-    {\includegraphics[width=0.5\textwidth]{density2.jpeg}};
-}
-\visible<12->{
-\node[inner sep=0pt] (density3) at (1.7,-0.1)
-    {\includegraphics[width=0.5\textwidth]{density3.jpeg}};
-}
-\end{tikzpicture}
-\end{minipage}
-\vspace{0.1cm}
-%\begin{adjustwidth}{-0.0em}{-1em}
-\begin{enumerate}
-\setcounter{enumi}{2}
-\item<8-> Split along the covariate $X$ with strongest association or instability and at breakpoint $p$ with highest improvement \\
-in log-likelihood.
-\item<10-> Repeat steps 1--3 recursively until some stopping criterion \\
-is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
-\end{enumerate}
-%\end{adjustwidth}
-\vspace{0.4cm}
-\visible<13->{
+
+\vspace{0.2cm}
 {\bf Forest:} Ensemble of $T$ trees.
@@
 \end{itemize}
-}
 \end{frame}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\SweaveOpts{eval=TRUE}
-
 \begin{frame}[fragile]
@@
 \medskip
-\textbf{Physical knowledge:} Numerical Weather Prediction (NWP) system.\\
+\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
 \medskip
-\textbf{Statistical post-processing} necessary due to uncertain initial conditions and unresolved processes.
-
-%Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
+\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
+\vspace{1.3cm}
+%\bigskip
+
+%\textbf{Idea:} Statistical model with numerical predictions as covariates.\\
+%\bigskip
+%\textbf{Distribution assumption:}\\
+%\vspace{-0.25cm}
+%\hspace{1.5cm}
+%\center
+%$$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
+
 \end{frame}
@@
 \frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.1cm}
+\vspace{0.4cm}
+\textbf{Goal:} Predict daily precipitation amount in alpine regions.\\
+\medskip
+\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
+\medskip
+\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
+\vspace{1.3cm}
+%\bigskip
+
+\only<2->{
 \textbf{Idea:} Statistical model with numerical predictions as covariates.\\
-\vspace{0.4cm}
+\bigskip
 \textbf{Distribution assumption:}\\
 \vspace{-0.25cm}
+%\hspace{1.5cm}
+%\center
 $$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
-%\vspace{2cm}
-\visible<2->{
-\textbf{Data:}
-%\vspace{-1cm}
+}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{0.4cm}
+\textbf{Observation data} (National Hydrographical Service):\\
 \begin{itemize}
-\item \textbf{Y:} Observation data (National Hydrographical Service):\\
+\item Daily 24h precipitation sums from July over 28 years (1985--2012).\\
+\item 95 observation stations in Tyrol, Austria.
+\end{itemize}
+\vspace{1cm}
+%\bigskip
+\textbf{NWP model output} (Global Ensemble Forecast System):\\
 \begin{itemize}
-\item 95 observation stations in Tyrol, Austria.\\
-\item Daily 24h precipitation sums (July, 1985--2012).
-%\item Daily 24h precipitation sums from July over 28 years (1985--2012).
+\item Ensemble predictions of various meteorological quantities related to precipitation, temperature,
+air pressure, \dots.\\
+\item 11 members each using slightly different perturbed initial conditions.
 \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace*{0.2cm}
+\textbf{Base variables:}
+\begin{itemize}
+\item Total precipitation
+\item Convective available potential energy
+\item Downwards short wave radiation flux (``sunshine'')
+\item Mean sea level pressure
+\item Preciptable water
+\item 2m maximum temperature
+\item Total column-integrated condensate
+\item Temperature 
+\item Temperature differences in altitude
+\end{itemize}
+\medskip
+\textbf{Variations:} Ensemble min/max/mean/standard deviation\\
 \vspace{0.1cm}
-\item \textbf{X:} NWP model output (Global Ensemble Forecast System):\\
-\vspace{0.14cm}
-\quad \small{Ensemble predictions of various meteorological quantities\\
-\quad (precipitation, temperature, air pressure, sunshine, \dots)\\
-\quad $\rightarrow$ 80 covariates.}
-\end{itemize}
-}
+$\rightarrow$ 80 covariates.
+
 \end{frame}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\SweaveOpts{eval=TRUE}
-
-<<pred_axams, echo=FALSE, results=hide>>=
-load("data_Axams.rda")
+<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
+
+#####
+# load observations and covariates 
+data("RainAxams")
 
@@
 
-learndata <- data_axams[data_axams$year < 109,]
-testdata <- data_axams[data_axams$year %in% c(109, 110, 111, 112),]
+  # learning data: 24 years (1985 - 2008, both inlcuded)
+  # testing data: 4 successive years (2009, 2010, 2011, 2012)
+  learndata <- RainAxams[RainAxams$year < 2009,]
+  testdata <- RainAxams[RainAxams$year %in% c(2009, 2010, 2011, 2012),]
 
@@
 # plot predicted distributions together with observations
-#set.seed(res$call$seedconst)
 set.seed(7)
@@
 
+# switch x-axis back to untransformed scale
+x <- x^(1.6)
+
 # point mass (slightly shifted)
-pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
-pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
-pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
-pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))
+pm1 <- c(0.15, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.05
+pm2 <- c(0.05, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.01
+pm3 <- c(-0.05, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.03
+pm4 <- c(-0.15, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.07
 
 # predictions
-pred1 <- c(testdata[pdays,"robs"][1], crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(testdata[pdays,"robs"][2], crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(testdata[pdays,"robs"][3], crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(testdata[pdays,"robs"][4], crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+pred1 <- c(testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
 
@@
 \vspace*{-0.2cm}
-<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, height=5.2>>=
+<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=5.2>>=
 plot(x = x, y = y1, type = "l", col = "red", 
      main = paste0("Axams, July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
-     #xlab = expression(Total~precipitation~group("[", mm^(1/1.6)~per~day,"]")),
-     #{\frac{1}{1.6}}$,
+     xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
+     # xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
      ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-1.5,8))
+     xlim = c(-2,28))     #  xlim = c(-1.5,8))
 
@@
 lines(x = x, y = y4, type = "l", col = "purple")
-legend('topright', c("predicted distribution", "point mass at censoring point", "observation"),
+legend('topright', c("Predicted distribution", "Point mass at censoring point", "Observation"),
        bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
@@
 # add labels
-text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
-text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
-text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
-text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
+text(x = -1.7, y = lh1, labels = "2009", col = "red", cex = 0.8)         # -0.8
+text(x = -1.7, y = lh2, labels = "2010", col = "blue", cex = 0.8)        # -0.8
+text(x = -1.7, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)   # -0.8
+text(x = -1.7, y = lh4, labels = "2012", col = "purple", cex = 0.8)      # -0.8
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\vspace{0.2cm}
+\textbf{Application for one station:}
+\begin{itemize}
+\item Learn forest model on data from 24 years.
+\item Evaluate on 4 years.
+\item 10 times 7-fold cross validation.
+\item Compare to commonly used heteroscedastic censored Gaussian models:
+\end{itemize}
+%\begin{itemize}
+%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
+%\item Boosted GAM: automatic variable selection.
+%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
+%\end{itemize}
+\vspace{0.1cm}
+\begin{table}[t!]
+\footnotesize
+
+\hskip-0.4cm\begin{tabular}{ l l l }
+\hline
+Model & Type & Variable Selection \\
+\hline
+\vspace{-0.15cm}
+\textbf{Forest} & recursive    & automatic \\ 
+                & partitioning &           \\ 
+\hline
+\vspace{-0.15cm}
+\textbf{prespecified GAM} & spline  & based on expert \\
+                          & in each & knowledge      \\
+\hline
+\vspace{-0.15cm}
+\textbf{boosted GAM} & spline  & automatic  \\
+                     & in each &            \\
+\hline
+\vspace{-0.15cm}
+\textbf{EMOS} & linear & total precipitation mean \\
+              &        & and standard deviation   \\
+\hline
+\end{tabular}
+\end{table}
+
+\end{frame}
+
+<<echo=FALSE, results=hide>>=
+#### cross validation rain
+if(file.exists("crps_cross.rda")){
+  load("crps_cross.rda")
+} else {
+
+  nrep_cross <- 10
+  seed <- 7
+
+  res_cross <- mclapply(1:nrep_cross,
+                        function(i){
+
+                          set.seed(seed*i)
+
+                          # randomly split data in 7 parts each including 4 years
+                          years <- 1985:2012
+                          testyears <- list()
+                          for(j in 1:7){
+                            testyears[[j]] <- sample(years, 4, replace = FALSE)
+                            years <- years[!(years %in% testyears[[j]])]
+                          }
+
+                          #crps <- matrix(nrow = 7, ncol = 7)
+                          reslist <- list()
+                          for(k in 1:7){
+                            test <- testyears[[k]]
+                            train <- c(1985:2012)[!c(1985:2012) %in% test]
+
+                            res <- evalmodels(station = "Axams",
+                                              train = train,
+                                              test = test,
+                                              gamboost_cvr = TRUE)
+
+                            #crps[k,] <- res$crps
+                            reslist[[k]] <- res
+                          }
+
+                          #colnames(crps) <- names(res$crps)
+                          return(reslist)
+                        },
+                        mc.cores = detectCores() - 1
+  )
+
+  # extract CRPS
+  crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
+  # loop over all repetitions
+  for(i in 1:length(res_cross)){
+    #loop over all 7 folds (for 7 methods)
+    crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
+    for(j in 1:length(res_cross[[1]])){
+      crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
+    }
+    crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
+  }
+  colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps)
+
+  save(crps_cross, file = "crps_cross.rda")
+}
+@
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Precipitation}
+\begin{center}
+\vspace*{-0.2cm}
+<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.6, width=6.8>>=
+#par(mar = c(2.5,2,1,2))
+boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
+        names = c("Distributional Forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
+        main = "Cross validation (with reference model EMOS)",
+        cex.main=1.4,
+        cex.lab=1.2,
+        ylab = "CRPS skill score", col = "lightgray") 
+abline(h = 0, col = pal["EMOS"], lwd = 2)
 @
@@
 \begin{itemize}
-\item Learn forest model on data from 24 years.
-\item Evaluate on 4 years.
-\item Compare to commonly used heteroscedastic censored Gaussian models.
+\item Learn forest model on data from 24 years (1985--2008).
+\item Evaluate on 4 years (2009--2012).
+\item Compare to commonly used heteroscedastic censored Gaussian models. %:\\
 \end{itemize}
+%\begin{itemize}
+%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
+%\item Boosted GAM: automatic variable selection.
+%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
+%\end{itemize}
+
+%\vspace{0.1cm}
+%\begin{table}[t!]
+%\footnotesize
+%\nodindent
+%\hskip-0.4cm\begin{tabular}{ l l l }
+%\hline
+%Model & Type & Variable Selection \\
+%\hline
+%\vspace{-0.15cm}
+%\textbf{Forest} & recursive    & automatic \\ 
+%                & partitioning &           \\ 
+%\hline
+%\vspace{-0.15cm}
+%\textbf{prespecified GAM} & spline  & based on expert \\
+%                          & in each & knowledge      \\
+%\hline
+%\vspace{-0.15cm}
+%\textbf{boosted GAM} & spline  & automatic  \\
+%                     & in each &            \\
+%\hline
+%\vspace{-0.15cm}
+%\textbf{EMOS} & linear & total precipitation mean \\
+%              &        & and standard deviation   \\
+%\hline
+%\end{tabular}
+%\end{table}
 \end{frame}
@@
 <<echo = FALSE>>=
-#### prediction over all stations 24 - 4
 #### prediction over all stations 24 - 4
@@
   save(crps_24to4_all, file = "crps_24to4_all.rda")
-  save(res_24to4_all, file = "res_24to4_all.rda")
-
 }
@@
 
-load("../draft/plot_map_rain/data/tirol.rda")
-load("../draft/plot_map_rain/data/dem.rda")
-
-data(StationsTyrol)
+#load("plot_map_rain/data/tirol.rda")
+#load("plot_map_rain/data/dem.rda")
+
+data(StationsTyrol, package = "RainTyrol")
+data(MapTyrol, package = "RainTyrol")
+# data(MapTyrol_border, package = "RainTyrol")
 # Create SpatialPointsDataFrame from station list
@@
   par(mar = c(5,4,4,0.1))
-  raster::image(dem, col = rev(gray.colors(100)),
+  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
                 main="Stations in Tyrol", cex.main=1.7,
@@
                 ylim = c(46.6, 47.87))
-  plot(tirol, add = TRUE)
+  plot(MapTyrol$SpatialPolygons, add = TRUE)
+  #points(sp[70,], pch = 19, col = "black", cex = 1.85)
   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
@@
 
+%\begin{frame}[fragile]
+%\frametitle{Simulation}
+%\begin{center}
+%\vspace*{-1cm}
+%\setkeys{Gin}{width=0.8\linewidth}
+%\includegraphics{sim_ll}
+%\end{center}
+%\end{frame}
+
+%\begin{frame}[fragile]
+%\frametitle{Simulation}
+%\begin{center}
+%\vspace*{-1cm}
+%\setkeys{Gin}{width=0.8\linewidth}
+%\includegraphics{sim_RMSE_par}
+%\end{center}
+%\end{frame}
+
+%\begin{frame}[fragile]
+%\frametitle{Simulation}
+%\begin{center}
+%\vspace*{-1cm}
+%\setkeys{Gin}{width=0.8\linewidth}
+%\includegraphics{sim_RMSE_exp}
+%\end{center}
+%\end{frame}
+
 \begin{frame}
@@
 
-\vspace{0.5cm}
-%\vspace{-0.2cm}
+\vspace{-0.2cm}
 
 \footnotesize
-
-%Hothorn T, Hornik K, Zeileis A (2006).
-% \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
-% \textit{Journal of Computational and Graphical Statistics},
-% \textbf{15}(3), 651--674.
-% \doi{10.1198/106186006X133933}
-
-%\medskip
-
-%Zeileis A, Hothorn T, Hornik K (2008).
-% \dquote{Model-Based Recursive Partitioning.}
-%  \textit{Journal of Computational and Graphical Statistics},
-%  \textbf{17}(2), 492--514.
-%  \doi{10.1198/106186008X319331}
-
-%\medskip
-
 Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
@@
   \url{http://arxiv.org/abs/1804.02921}
-and 
-  \emph{Working Papers, Faculty of Economics and Statistics, University of Innsbruck},
-  \url{https://EconPapers.repec.org/RePEc:inn:wpaper:2018-08}
-
-\bigskip
+
+\medskip
+
+Hothorn T, Hornik K, Zeileis A (2006).
+ \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
+ \textit{Journal of Computational and Graphical Statistics},
+ \textbf{15}(3), 651--674.
+ \doi{10.1198/106186006X133933}
+
+\medskip
+
+Zeileis A, Hothorn T, Hornik K (2008).
+ \dquote{Model-Based Recursive Partitioning.}
+  \textit{Journal of Computational and Graphical Statistics},
+  \textbf{17}(2), 492--514.
+  \doi{10.1198/106186008X319331}
+
+\medskip
 
@@
 
-\bigskip
+\medskip
 
@@
 
-%\medskip
+\medskip
+
+Hofner B, Mayr A, Schmid M (2016).
+  \dquote{\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and Variable Selection in the {GAMLSS} Framework}
+  \textit{Journal of Statistical Software},
+  \textbf{74}(1), 1--31.
+  \doi{10.18637/jss.v074.i01}
+
+\end{frame}
+
+%\begin{frame}
+%\frametitle{References (2)}
+
+%\vspace{-0.2cm}
+
+%\footnotesize
 
 %Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-%  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and Minimum CRPS Estimation}
+%  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and %Minimum CRPS Estimation}
 %  \textit{Monthly Weather Review}, 
@@
 
-%Hofner B, Mayr A, Schmid M (2016).
-%  \dquote{\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and Variable Selection in the {GAMLSS} Framework}
-%  \textit{Journal of Statistical Software},
-%  \textbf{74}(1), 1--31.
-%  \doi{10.18637/jss.v074.i01}
-
-\end{frame}
-
-%\begin{frame}
-%\frametitle{References (2)}
-
-%\vspace{-0.2cm}
-
-%\footnotesize
-
-%% Reference for the observation data set
+% Reference for the observation data set
 %BMLFUW (2016).
-%  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtschaft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
+%  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtsch%aft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
 %  \text{Available at \url{http://ehyd.gv.at}}, 
@@
 % Reference for the GFS reforecast version 2 data set 
-%Hamill T M, Bates G T, Whitaker J S, Murray D R, Fiorino M, Galarneau T J Jr., Zhu Y, Lapenta W (2013).
+%Hamill T M, Bates G T, Whitaker J S, Murray D R, Fiorino M, Galarneau T J Jr., Zhu Y, Lapenta %W (2013).
 %  \dquote{{NOAA}'s Second-Generation Global Medium-Range Ensemble Reforecast Dataset}
@@
 %Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
-%  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored Additive Regression Model}
+%  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored %Additive Regression Model}
 %  \textit{International Journal of Climatology}, 
@@
 %Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
-%  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: Unanimous Predictions, Heavy Tails, and Link %Functions}
+%  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: %Unanimous Predictions, Heavy Tails, and Link Functions}
 %  \textit{Monthly Weather Review}, 

### 2018-07-15 pkg/partykit/inst/guideliketest/draft/draft.Rnw
first draft for comparison of guidelike tests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2220 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9fdcf7424d7bf6503d08c5a06b41a885fba48c7e

\documentclass[nojss]{jss}

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern}

%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}

%% commands
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

<<preliminaries, echo=FALSE, results=hide>>=
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
source("../guidelike.R")
source("../tests.R")
load("../sim/simres20180422_1step.rda")
sim1step <- simres
#load("../sim/simres20180422_2step.rda")
#sim2step <- simres
rm(simres)
@

\title{Comparing Testing Strategies of Unbiased Tree Algorithms}
%\Shorttitle{Comparing Different Testing Approaches in Tree Algorithms}

\author{Lisa Schlosser\\Universit\"at Innsbruck
   \And Torsten Hothorn\\Universit\"at Z\"urich
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}

\Abstract{Comparing Testing Strategies of Unbiased Tree Algorithms}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

\Address{
  Lisa Schlosser, Achim Zeileis \\
  Universit\"at Innsbruck \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\

  Torsten Hothorn\\
  Universit\"at Z\"urich \\
  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
  Hirschengraben 84\\
  CH-8001 Z\"urich, Switzerland \\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://user.math.uzh.ch/hothorn/}\\

}

\begin{document}

\section{Introduction}
In many situations fitting one global model to a data set can be very challenging, especially if this data contains lots of different features with high/extreme variations or complex interactions. Therefore, separating the data into more homogeneous subgroups based on (a set of) covariates first can simplify the task and fitting a local model to each of the resulting subgroups often leads to better results. This separation of the data can be done by applying a tree algorithm. Many different approaches and algorithms have been developed in this research field. While they all follow the idea of splitting the data in a way such that some objective function is minimized/maximized, they differ in their approach to the crucial decision that has to made before each split: In which variable and at which break point should a split be performed. (Early tree algorithms rely on exhaustive search procedures to find the best split point and split variable in one step. However, it has been shown that this is not only computatitionaly expensive but also biased towards split variables with many possible split points. Therefore, succeeding algorithms select a split variable in a first step and then search for the best split point within this variable only in a second step. 
In this paper the focus will be put on the task of selecting the (right/best) split variable. 
In particular the strategy of the GUIDE \citep{Loh:2002} algorithm is compared to the one of the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} and the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008}. Moreover, adapted versions of these algorithms are evaluated as well in order to compare effects of specific differences separately. 
After introducing the setting for the simulation study in Section~\ref{sec:settings} and a general scheme on how to build a tree in Section~\ref{sec:tree}, the three main differences will be discussed in Sections~\ref{sec:scores}, \ref{sec:categorization} and \ref{sec:binarization}.

\section{Simulation Settings}
\label{sec:settings}
\subsection{Data Generating Process}
The different approaches are evaluated on generated data sets. Each data set consists of a response variable, one regressor, one split variable and nine noise variables. In particular, the variables listed in Table~\ref{tab:variables} together with the corresponding distributions are included in the data generating process.

\vspace{0.2cm}

\begin{table}
\begin{center}
\begin{tabular}{l l l}
Name                    & Notation                   & Specification \\
\hline
response                & $Y$                        & $\mathcal{N}(\mu(X, Z_1), \sigma^2)$ \\
mean of $Y$             & $\mu$                      & linear function of $X$ and $Z_1$ \\
variance of $Y$         & $\sigma^2$                 & 1\\
regressor               & $X$                        & $\mathcal{U}([-1,1])$  or $(-1)^{\mathcal{B}(0.5)}$ \\
true split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
coefficients            & $\beta_0$, $\beta_1$       & functions of $Z_1$\\
true split point        & $\xi$                      & $\in \{0, 0.2, 0.5, 0.8\}$\\
effect size             & $\delta$                   & $\in \{0, 0.25, 0.5, 0.75, 1\}$
\end{tabular}
\caption[Table caption text]{Variables included in the data generating process.}
\label{tab:variables}
\end{center}
\end{table}

\vspace{0.4cm}

\noindent The location parameter $\mu$ of the normally distributed response variable $Y$ depends linearly on the regressor variable $X$ where the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation (can) depend on the true split variable $Z_1$ as explained below.
\begin{center}
$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$
\end{center}

\noindent For the coefficients $\beta_0$ and $\beta_1$ the following three different scenarios are considered:
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at 1).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed (at 0).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}

\noindent The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, can either be binary, e.g. a step function, or a continuous function. In the binary case the following definition is applied
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{i}  \quad \text{if } Z_1 < \xi \\
+\delta \cdot (-1)^{i}  \quad \text{if } Z_1 \geq \xi
\end{cases}
\end{align*}
while in the continuous case a simple linear function is considered in the form of
$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in oposite directions.

%\noindent Depending on whether the coefficients are binary or continuous, in the case of only one coefficient varying it is defined by 
%\begin{align*}
%\beta_{0/1}(Z_1) = \begin{cases}
%+\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%-\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%Z_1 \cdot \delta \qquad \text{if } \texttt{binary\_beta=FALSE}
%\end{cases}
%\end{align*}

%\noindent while in the case of both coefficients varying they are defined by:
%\begin{align*}
%\beta_{0}(Z_1) = \begin{cases}
%-\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%+\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%Z_1 \cdot \delta \qquad \text{if } \texttt{binary\_beta=FALSE}
%\end{cases}
%\end{align*}
%\begin{align*}
%\beta_{1}(Z_1) = \begin{cases}
%+\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%-\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%- Z_1 \cdot  \delta \hspace{0.5cm} \text{if } \texttt{binary\_beta=FALSE}
%\end{cases}
%\end{align*}

%\newpage
\begin{figure}%[t!]
\setkeys{Gin}{width=1.0\linewidth}
<<dgp_bin_beta_cont_reg, fig=TRUE, echo=FALSE>>=
set.seed(7)

nobs <- 300
sigma <- 0.5
xi <- 0
delta <- 1
beta0 <- 0
beta1 <- 1

par(mfrow = c(3,3))

d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
          vary_beta = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_0")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_0")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_0")

d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
          vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_1")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_1")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_1")

d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
          vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary all")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary all")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary all")
@
\caption{\label{fig:dgp}{Generated data of 300 observations for binary coefficients $\beta_0$ and $\beta_1$, continuous regressor $X$, variance parameter $\sigma = 0.5$, effect size $\delta = 1$ and split point $\xi = 0$. Left panels: response $y$ on regressor $X$; middle panels: intercept $\beta_0$ on split variable $Z_1$; middle panels: slope coefficients $\beta_1$ on split variable $Z_1$; each for varying $\beta_0$ (top), varying $\beta_1$ (middle) and both coefficients varying (bottom).}}
\end{figure}

%\newpage
\begin{figure}%[t!]
\setkeys{Gin}{width=1.0\linewidth}
<<dgp_cont_beta_bin_reg, fig=TRUE, echo=FALSE>>=
set.seed(7)

nobs <- 300
sigma <- 0.5
xi <- 0
delta <- 1
beta0 <- 0
beta1 <- 1

par(mfrow = c(3,3))

d <- dgp_stump(nobs = nobs, binary_beta = FALSE, binary_regressor = TRUE, only_intercept = FALSE,
          vary_beta = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_0")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_0")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_0")

d <- dgp_stump(nobs = nobs, binary_beta = FALSE, binary_regressor = TRUE, only_intercept = FALSE,
          vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_1")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_1")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_1")

d <- dgp_stump(nobs = nobs, binary_beta = FALSE, binary_regressor = TRUE, only_intercept = FALSE,
          vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary all")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary all")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary all")
@
\caption{\label{fig:dgp}{Generated data of 300 observations for continuous coefficients $\beta_0$ and $\beta_1$, binary regressor $X$, variance parameter $\sigma = 0.5$, effect size $\delta = 1$. Left panels: response $y$ on regressor $X$; middle panels: intercept $\beta_0$ on split variable $Z_1$; middle panels: slope coefficients $\beta_1$ on split variable $Z_1$; each for varying $\beta_0$ (top), varying $\beta_1$ (middle) and both coefficients varying (bottom).}}
\end{figure}

\newpage
\section{Building a Tree}
\label{sec:tree}
The aim of building a tree model is to partition the data into smaller and more homogeneous subgroups based on (a set of) covariates. 
Various different tree building algorithms have been developed, all following the same basic structure:
\begin{enumerate}
\item Starting with the whole data set, some form of discrepancy measure is calculated. This can for example be the sum of deviations from a typical or average value of the data set or it can be based on a fitted model such as the residual sum of squares (RSS).
\item Among all available covariates one is selected as split variable and a value within the range/measurable space of this variable is chosen as split point such that after splitting the data the discrepancy measure is minimal within the resulting subgroups.
% One of all available covariates together with a value within the range of this variable is chosen as split variable with the corresponding split point such that after splitting the data the discrepancy measure is minimal within the resulting subgroups.
\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criteria is reached or the subgroup consists of only one or all equal observations. (In the second case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
\end{enumerate}

% For notational simplicity the following generic tree algorithm will be explained for the root node containing observations ... 
% At first the whole data set is considered. In order to find the best way of splitting the data some kind of discrepancy measure is calculated, often to a mean % or average value of the data. Then the data is split into subgroups such that the decrease in the discrepancy measure is maximal. This procedure is then
% repeated in each of the subgroups until some stopping criterion is met.
% In a first step a global model is fit to the whole data set and its goodness of fit is evaluated. This can for example be done by calculating the residual sum of squares (RSS) or any other error measurement.

The crucial part in such an algorithm is the selection of a split variable and the corresponding split point. Algorithms such as CART make this decision in one step by performing an exhaustive search over all possible split points in all available split variables. However, this method is not only computationally demanding but also biased as it favors variables with a high number of possible splits. Therefore, other algorithms, such as GUIDE, QUEST, CTree and MOB tackle the splitting decision in two separate steps: First, a split variable is chosen and, second, a split point in the chosen split variable is selected. One way to search for the best split variable is to apply independence tests between the values of the discrepancy measurement for each observation and the corresponding values of each of the possible split variables and select the one showing the highest dependence. Once the split variable is chosen the search area for the optimal split point is reduced to the range of this split variable. A possible but computationally demanding approach to finding the best split point is an exhaustive search over all possible split points in the split variable. However, other strategies have been developed as well, for example based on two sample statistics such as applied in the CTree algorithm.

% After splitting the data into subgroups it has to be checked whether a predefined stopping criterion has already been fulfilled. Otherwise, the foregoing steps are repeated for each of the new subgroups, e.g. a model is fitted in each subgroup, goodness of fit is evaluated and split variables and split points are selected. This procedure is repeated until a stopping criterion is met or, if no criterion has been set in advance, until each subgroup consists of only one value or only equal values. In the second case the resulting tree is of maximal size and can be reduced by (post-)pruning.

%Summed up, the steps of building a model-based tree are:
%\begin{enumerate}
%\item Fit a global model to the (whole) data set.
%\item Evaluate the goodness of fit of this model.
%\item Select the split variable showing the highest dependence to the values of the goodness of fit-measurement and the split point leading to the highest improvement in goodness of fit and divide the data into subgroups.
%\item Repeat steps 1--3 for each resulting subgroup until a predefined stopping criterion is met.
%\end{enumerate}

\newpage
\section{Which scores should be used?}
\label{sec:scores}

In this simulation study the models that are fit to the data in each node of the tree are linear models. The fitting is done by estimating the coefficient vector $\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$ via minimizing an objective function $\Psi(y_i,x_i,\boldsymbol{\beta})$ summed up over all observations of the learning data set that are in the current node, for notational simplicity generally denoted by $\{(y_i, x_i)\}_{i=1\dots n}$. Here, this sum is the residual sum of squares. Information about the goodness of fit of the resulting model can be obtained by calculating so called scores. These are the values of the objective function derived by the coefficients with the given data and the estimated coefficients $\hat{\boldsymbol{\beta}}$ plugged in:

$$
s(y_i, x_i, \hat{\boldsymbol{\beta}}) = \frac{\partial \Psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
$$

which is in this case a vector of two dimensions as $\boldsymbol{\beta}$ is two-dimensional. In that way a score value indicating how well the model fits the data is given separately for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$ and $j \in \{1,2\}$.

In order to choose a split variable dependency tests are employed on the score values and each of the possible split variables. The variable showing the highest dependency is selected. Now the question arises which of the score values should be considered for these tests. Different approaches have been established. While the GUIDE-algorithm \citep{Loh:2002} only includes the scores of the intercept coefficient $\beta_0$, the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} and the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} consider all available score values. In Figure~\ref{fig:which_scores} the results of these two strategies applied on the before described simulation setting (for \texttt{binary\_beta = FALSE} and  \texttt{binary\_regressor = TRUE}) can be compared. Additionally, an adapted version of the GUIDE algorithm including all available score values has been employed as well.
For changes only in the intercept (left panel) the correct splitting variable ($Z_1$) is detected successfully by all methods with increasing effect size. While mfluc performs best, CTree follows second, slightly ahead of the original GUIDE version which performs better than the adapted version. In case of changes in the slope coefficient (middle panel) $\beta_1$ mfluc, CTree and the adapted GUIDE version again lead to similar results while the original GUIDE version fails in detecting the correct splitting variable as it was to be expected. In the third scenario where both coefficients vary over $Z_1$ (right panel) the adapted GUIDE version slightly outperforms the original version while CTree and mfluc are again ahead.
These results clearly illustrate that overall the decision to include all available scores when testing for dependencies in order to select split variables is advantageous over considering only the scores regarding the intercept. Only in situations where changes appear solely in the intercept the original GUIDE version can lead to slighter better results than the adapted version. However, since this particular type of scenario usually can not be assumed to be the case in advance, its strength in this particular situation is clearly outweighed by the obvious disadvantages in all other situations.

%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
\begin{figure}
\setkeys{Gin}{width=\linewidth}
<<which_scores_1step_z1_binreg, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim1step, test %in% c("ctree_max","mfluc",
                                        "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test)

xyplot(prop_z1 ~ delta | vary_beta, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == FALSE & binary_regressor == TRUE & 
                only_intercept == FALSE & xi == 0, 
       type = "b", auto.key = TRUE, layout= c(3,1),
       index.cond = list(c(2,3,1)))
@
\caption{\label{fig:which_scores}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for continuous coefficients, binary regressor, $\sigma = 1$, 250 observations and 100 repetitions per step. Each panel represents one scenario for the coefficients: varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and both coefficients varying (right panel). The performances of 4 algorithms are compared: mfluc, tree, adapted GUIDE, original GUIDE.}
\end{figure}

\newpage
\noindent Within the GUIDE algorithm two additional steps are made before testing for dependencies between scores and split variables. 
\begin{itemize}
\item The scores are binarized, i.e. depending on whether the actual values are positive or negative they are replaced by $+1$ or $-1$. 
\item The observed values of the possible split variables are categorized into 4 groups separated by the 1st, 2nd and 3rd quartile.
\end{itemize}
The effects of these two steps are going to be investigated in detail in the following two sections.

\section{Categorize the splitting variables?}
\label{sec:categorization}
To find out whether a categorization of the possible splitting variables leads to an improvement or a deterioration of performance the two algorithms CTree and MOB are applied, once in their original version without a categorization and once in an adapted version with a categorization. Moreover, they are compared to the adapted GUIDE version which includes scores of all available scores such as CTree and MOB do.

In Figure~\ref{fig:cat_ctree} the effect of the categorization on the performance of the CTree algorithm is illustrated. Only for the split point $\xi = 0.5$ it leads to an improvement which was to be expected since this is one of the quartiles used for the categorization. However, apart from this point categorizing the values of the split variable has a negative effect on the performance, even at $\xi = 0$ which is also one of the used quartiles.
Similar conclusions can be drawn looking at Figure~\ref{fig:cat_mfluc} where the effect on the MOB algorithm is shown. In this case, the categorization has almost no effect at the quartiles $\xi = 0$ and $\xi=0.5$, but again, leads to a decrease in the proportion of cases where the true splitting variable is detected.

Therefore, it can be summarized that a categorization of the values of the split variable does not lead to any advantages unless the true split point corresponds to one of the quartiles used for the categorization. In most situations it even causes the performance to get worse.

%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
\begin{figure}[h!]
\setkeys{Gin}{width=\linewidth}
<<cat_ctree_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim1step, test %in% c("ctree","ctree_cat", 
                                        "guide_sum_12"))
subdata$test <- factor(subdata$test)

xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & 
                only_intercept == FALSE & vary_beta == "beta1", 
       type = "b", auto.key = TRUE , layout= c(4,1))

@
\caption{\label{fig:cat_ctree}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step. Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of 3 algorithms are compared: CTree, categorized versions of CTree, adapted GUIDE.}
\end{figure}

\begin{figure}[h!]
\setkeys{Gin}{width=\linewidth}
<<cat_mfluc_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim1step, test %in% c("mfluc","mfluc_cat", 
                                        "guide_sum_12"))
subdata$test <- factor(subdata$test)

xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & 
                only_intercept == FALSE & vary_beta == "beta1", 
       type = "b", auto.key = TRUE , layout= c(4,1))

@
\caption{\label{fig:cat_mfluc}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step. Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of 3 algorithms are compared: mfluc, categorized version of mfluc, adapted GUIDE.}
\end{figure}

\newpage
\section{Binarize the response variable?}
\label{sec:binarization}
To find out whether a binarization of the scores leads to an improvement or a deterioration of performance the two algorithms CTree and MOB are applied, once in their original version without a binarization and once in an adapted version with a binarization. Moreover, they are compared to the adapted GUIDE version which includes scores of all available scores such as CTree and MOB do.

%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
\begin{figure}[h!]
\setkeys{Gin}{width=\linewidth}
<<bin_ctree_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim1step, test %in% c("ctree","ctree_bin", 
                                        "guide_sum_12"))
subdata$test <- factor(subdata$test)

xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & 
                only_intercept == FALSE & vary_beta == "beta1", 
       type = "b", auto.key = TRUE , layout= c(4,1))

@
\caption{\label{fig:bin_ctree}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step. Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of 3 algorithms are compared: CTree, categorized versions of CTree, adapted GUIDE.}
\end{figure}

\begin{figure}[h!]
\setkeys{Gin}{width=\linewidth}
<<bin_mfluc_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim1step, test %in% c("mfluc","mfluc_bin", 
                                        "guide_sum_12"))
subdata$test <- factor(subdata$test)

xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
       data = subdata, 
       subset = binary_beta == TRUE & binary_regressor == FALSE & 
                only_intercept == FALSE & vary_beta == "beta1", 
       type = "b", auto.key = TRUE , layout= c(4,1))

@
\caption{\label{fig:bin_mfluc}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step. Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of 3 algorithms are compared: mfluc, categorized version of mfluc, adapted GUIDE.}
\end{figure}

\newpage
\bibliography{ref.bib}

\end{document}

### 2018-07-24 pkg/disttree/inst/JSM2018/slides.Rnw
formating and small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2237 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 21389262e5b55eb5644a2d0d8c0d6d96efa6c07d

@@
 %\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
-\URL{https://R-Forge.R-project.org/projects/partykit/}
+\URL{https://R-Forge.R-project.org/projects/partykit/ \\
+https://partykit.org}
 \headerimage{3}
@@
 \visible<1->{
-Regression Tree\\
+Regression tree\\
 \vspace{0.5cm}
@@
 \visible<2->{
-Random Forest\\
+Random forest\\
 \vspace{0.5cm}
@@
 \begin{frame}
-\frametitle{Distributional trees}
+\frametitle{Distributional Trees}
 \vspace*{-0.12cm}
@@
 \begin{frame}[fragile]
-\frametitle{Distributional trees}
+\frametitle{Distributional Trees}
 \begin{center}
@@
 %\begin{frame}
-%\frametitle{Distributional trees}
+%\frametitle{Distributional Trees}
 %\center
@@
 \item Choose the partitioning variable with the strongest association or instability as split variable.
-\item Choose the breakpoint along this variable with highest improvement in log-likelihood and split the sample.
-\item Repeat steps 2--5 recursively in the subgroups
-    until some stopping criterion is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
+\item Choose the breakpoint along this variable leading to the highest improvement in log-likelihood and split the sample.
+\item Repeat steps 2--5 recursively until some stopping criterion is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
     %-- e.g., for significance or sample size.
@@
 
-\textbf{Parameter estimator} for \only<1-3>{\textbf{a global}}\only<4->{\textbf{an adaptive local}}\\ model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$} :
+\textbf{Parameter estimator} for \only<1-2>{\textbf{a global}}\only<3->{\textbf{an adaptive local}}\\ model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$} :
 
@@
 
-\visible<3->{
+\visible<2->{
 \textbf{Weights:}
 \begin{eqnarray*}
-w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.2cm]
+w^{\text{global}}_i(\bold{x})   & = & 1 \\[0.2cm]
+\visible<3->{
+w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.1cm]
 \visible<4->{
-w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.1cm]
-\visible<5->{
 w^{\text{forest}}_i(\bold{x}) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
@@
 # switch x-axis back to untransformed scale
-x <- x^(1.6)
+# x <- x^(1.6)
 
 # point mass (slightly shifted)
-pm1 <- c(0.15, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.05
-pm2 <- c(0.05, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.01
-pm3 <- c(-0.05, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.03
-pm4 <- c(-0.15, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.07
+pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.15
+pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.05
+pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.05
+pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.15
 
 # predictions
-pred1 <- c(testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+pred1 <- c(testdata[pdays,"robs"][1], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(testdata[pdays,"robs"][2], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(testdata[pdays,"robs"][3], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(testdata[pdays,"robs"][4], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
 
@@
      main = paste0("Axams, July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
-     # xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
+     # xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
+     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
      ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-2,28))     #  xlim = c(-1.5,8))
+     xlim = c(-1.5,8))     #  xlim = c(-2,28))
 
@@
 # plot predictions
-points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
-points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
-points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
-points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
+points(x = pred1[1], y = pred1[2], col = "red", pch = 4, cex = 1.4)
+points(x = pred2[1], y = pred2[2], col = "blue", pch = 4, cex = 1.4)
+points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4, cex = 1.4)
+points(x = pred4[1], y = pred4[2], col = "purple", pch = 4, cex = 1.4)
 
@@
 # add labels
-text(x = -1.7, y = lh1, labels = "2009", col = "red", cex = 0.8)         # -0.8
-text(x = -1.7, y = lh2, labels = "2010", col = "blue", cex = 0.8)        # -0.8
-text(x = -1.7, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)   # -0.8
-text(x = -1.7, y = lh4, labels = "2012", col = "purple", cex = 0.8)      # -0.8
+text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)         # -1.7 
+text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)        # -1.7
+text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)   # -1.7
+text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)      # -1.7
 @
@@
 \vspace{-0.15cm}
-\textbf{Forest} & recursive    & automatic \\ 
+\textbf{Distributional forest} & recursive    & automatic \\ 
                                & partitioning &           \\ 
@@
 \vspace{-0.15cm}
-\textbf{prespecified GAM} & spline  & based on expert \\
+\textbf{Prespecified GAM} & spline  & based on expert \\
                           & in each & knowledge      \\
@@
 \vspace{-0.15cm}
-\textbf{boosted GAM} & spline  & automatic  \\
+\textbf{Boosted GAM} & spline  & automatic  \\
                      & in each &            \\
@@
 boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
-        names = c("Distributional Forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
+        names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
         main = "Cross validation (with reference model EMOS)",
@@
 
-%\begin{frame}
-%\frametitle{References (2)}
-
-%\vspace{-0.2cm}
-
-%\footnotesize
-
-%Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-%  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and %Minimum CRPS Estimation}
-%  \textit{Monthly Weather Review}, 
-%  \textbf{133}(5), 1098--1118.
-%  \doi{10.1175/MWR2904.1}
-
-%\medskip
+\begin{frame}
+\frametitle{References (2)}
+
+%\vspace{-0.1cm}
+
+\footnotesize
+
+Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
+  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and Minimum CRPS Estimation}
+  \textit{Monthly Weather Review}, 
+  \textbf{133}(5), 1098--1118.
+  \doi{10.1175/MWR2904.1}
+
+\medskip
 
@@
 %BMLFUW (2016).
-%  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtsch%aft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
+%  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtschaft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
 %  \text{Available at \url{http://ehyd.gv.at}}, 
@@
 
-%Breiman L (2001).
-%  \dquote{Random {Forests}}
-%  \textit{Machine Learning}, 
-%  \textbf{45}(1), 5--32.
-%  \doi{10.1023/A:1010933404324}
-
-%Messner J W,  Mayr G J, Zeileis A (2016).
-%  \dquote{Heteroscedastic Censored and Truncated Regression with \textbf{crch}}
-%  \textit{The \textsf{R} Journal}, 
-%  \textbf{8}(1), 173--181.\\
-%  \scriptsize
-%  \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
-%  \footnotesize
-
-%\medskip
-
-%Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
-%  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored %Additive Regression Model}
-%  \textit{International Journal of Climatology}, 
-%  \textbf{37}(7), 3264--3275.
-%  \doi{10.1002/joc.4913}
-
-%\medskip
-
-%Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
-%  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: %Unanimous Predictions, Heavy Tails, and Link Functions}
-%  \textit{Monthly Weather Review}, 
-%  \text{Forthcoming}.
-%  \doi{10.1175/MWR-D-16-0388.1}
+Breiman L (2001).
+  \dquote{Random {Forests}}
+  \textit{Machine Learning}, 
+  \textbf{45}(1), 5--32.
+  \doi{10.1023/A:1010933404324}
+
+\medskip
+
+Messner J W,  Mayr G J, Zeileis A (2016).
+  \dquote{Heteroscedastic Censored and Truncated Regression with \textbf{crch}}
+  \textit{The \textsf{R} Journal}, 
+  \textbf{8}(1), 173--181.\\
+  \scriptsize
+  \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
+
+\medskip
+\footnotesize
+
+Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
+  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored Additive Regression Model}
+  \textit{International Journal of Climatology}, 
+  \textbf{37}(7), 3264--3275.
+  \doi{10.1002/joc.4913}
+
+\medskip
+
+Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
+  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: Unanimous Predictions, Heavy Tails, and Link Functions}
+  \textit{Monthly Weather Review}.
+  \doi{10.1175/MWR-D-16-0388.1}
 
@@
 
-%\end{frame}
+\end{frame}
 

### 2018-07-25 pkg/glmertree/R/glmertree.R
weights argument can now be passed with standard and non-standard evaluation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2238 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7358dadd3fb37b27ffb913bdd2ec9a58c96a6d26

@@ -2,7 +2,7 @@ utils::globalVariables(c(".tree", ".ranef", ".weights", ".cluster"))
 
 lmertree <- function(formula, data, weights = NULL, cluster = NULL,
                      ranefstart = NULL, offset = NULL, joint = TRUE,
-                     abstol = 0.001, maxit = 100, dfsplit = TRUE, 
+                     abstol = 0.001, maxit = 100L, dfsplit = TRUE, 
                      verbose = FALSE, plot = FALSE, 
                      lmer.control = lmerControl(), ...)
 {
@@ -16,14 +16,12 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
   
   ## process offset:
   q_offset <- substitute(offset)
-  offset <- try(offset, silent = TRUE)
-  if (inherits(offset, "try-error") || length(offset) != nrow(data)) {
+  if (!is.null(q_offset)) {
     offset <- eval(q_offset, data)
   }
 
   ## process cluster:
   q_cluster <- substitute(cluster)
-  cluster <- try(cluster, silent = TRUE)
   if (!is.null(q_cluster)) {
     data$.cluster <- eval(q_cluster, data)
     if (length(eval(q_cluster, data)) != nrow(data)) {
@@ -34,6 +32,17 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
       !inherits(data$.cluster, c("numeric", "character", "factor", "integer"))) {
     warning("Argument 'cluster' should specify an object of class numeric, factor or character, or NULL.", .immediate = TRUE)
   } 
+  
+  # process weights:
+  q_weights <- substitute(weights)
+  if (!is.null(q_weights)) {
+    data$.weights <- eval(q_weights, data)
+    weights <- eval(q_weights, data)
+    ## Note: Assigning weights and offset as variables outside data prevents lmer from yielding an error
+  } else {
+    data$.weights <- rep(1L, times = nrow(data))
+  }
+  
   
   ## formula processing (full, tree, random)
   ff <- Formula::as.Formula(formula)
@@ -51,9 +60,6 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
   } else {
     rf <- formula(ff, lhs = 1L, rhs = 2L)
   }
-  
-  ## weights
-  data$.weights <- if (is.null(weights)) rep(1, nrow(data)) else weights
   
   ## initialization
   iteration <- 0L
[…]

### 2018-08-29 pkg/disttree/inst/draft/draft.Rnw
highlighted changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2268 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 394c73009df503993cb104295fbe472d06896c85

--- 2018-07-06 pkg/disttree/inst/draft/draft.Rnw
@@
 %% tikz
-\usepackage{array,makecell,tikz}
+\usepackage{array,makecell,tikz,color,soul}
 \usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
@@
 
-# predictions
-pred1 <- c(Axams_24to4$testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(Axams_24to4$testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(Axams_24to4$testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(Axams_24to4$testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+# observations
+obs1 <- c(Axams_24to4$testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+obs2 <- c(Axams_24to4$testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+obs3 <- c(Axams_24to4$testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+obs4 <- c(Axams_24to4$testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
 
@@
 
-# plot predictions
-points(x = pred1[1], y = pred1[2], col = pal[1], pch = 4)
-points(x = pred2[1], y = pred2[2], col = pal[3], pch = 4)
-points(x = pred3[1], y = pred3[2], col = pal[2], pch = 4)
-points(x = pred4[1], y = pred4[2], col = pal[4], pch = 4)
-
-lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+# plot observations
+points(x = obs1[1], y = obs1[2], col = pal[1], pch = 4, cex = 1.4)
+points(x = obs2[1], y = obs2[2], col = pal[3], pch = 4, cex = 1.4)
+points(x = obs3[1], y = obs3[2], col = pal[2], pch = 4, cex = 1.4)
+points(x = obs4[1], y = obs4[2], col = pal[4], pch = 4, cex = 1.4)
+
+lines(x = c(obs1[1], obs1[1]), y = c(obs1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(obs2[1], obs2[1]), y = c(obs2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(obs3[1], obs3[1]), y = c(obs3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
+lines(x = c(obs4[1], obs4[1]), y = c(obs4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
 
@@
 @
-\caption{\label{fig:axams-dist}Total precipitation predictions by a distributional forest
+\caption{\label{fig:axams-dist}\hl{Total precipitation predictions by a distributional forest
 at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
-Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
+Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}}
 \end{figure}
@@
 $$
+\hl{Even though this parameter depends on the given $z$ it has to be noted that
+it is the same for all other observations that are in the same terminal node of 
+the tree as $z$ and therefore does not necessarily have to be recalculated for
+each of them.}
 
@@
   sampling is employed for each tree using a third of the predictors
-  in each split of the tree (``mtry''). Parameters are estimated by
+  in each split of the tree (\code{mtry}). Parameters are estimated by
   adaptive local likelihood based on the forest weights as described
   in Section~\ref{sec:methods}. 
+  \hl{As stopping criteria the minimal number
+  of observations to perform a split} (\code{minsplit}) \hl{is set to 50, 
+  the minimal number of observations in a terminal node} (\code{minbucket})
+  \hl{is set to 20 and the significance level for variable selection} (\code{alpha}) 
+  \hl{is kept at the default value $1$.}
   % stopping criteria:
-  % tree_minsplit = 50, tree_minbucket = 20, tree_mincrit = 0.95,
   % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0
+
+  % default in ctree_control within distforest:
+  % mincriterion = 0
+
+  % default values in ctree_control:
+  % alpha = 0.05, mincriterion = 1 - alpha,
+  % minsplit = 20L, minbucket = 7L,
+  % minprob = 0.01
 
@@
 rather than relying on just the one fixed test set for the years 2009--2012.
-To do so, a 10 times 7-fold cross-validation is carried out where each
-splits the available 28~years into 7~subsets of 4~randomly selected years.
+\hl{To do so, a 10 times 7-fold cross-validation is carried out where each
+splits the available 28~years into 7~subsets of 4~randomly selected and 
+therefore mostly not subsequent years.}
 The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
@@
 $$
-where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
+\hl{where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
 type of the transformation function $v_l$ depends on the type of the split 
@@
 as the ``vec'' operator converts the $H \times k$ matrix into a $k$ column vector 
-by column-wise combination. Observations with missing values are excluded from the sums.
+by column-wise combination.} Observations with missing values are excluded from the sums.
 
@@
 \cite{Strasser+Weber:1999} the test statistic can be standardized. 
-The observed multivariate linear statistic $t_l$ which is either an
+\hl{The observed multivariate linear statistic $t_l$ which is either an
 $k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of Y,
 is mapped onto the real line by a univariate test statistic $c$.
-In the application of this paper a quadratic form is chosen, such that
+In the application of this paper a quadratic form is chosen, such that}
 $$
@@
 $$
-where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$. Alternatively,
+\hl{where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.} Alternatively,
 the maximum of the absolute values of the standardized linear statistic
@@
 If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
-is chosen as splitting variable. Otherwise no further split is made is in this node.
+is chosen as splitting variable. \hl{Otherwise no further split is made is in this node.}
 
@@
 $$
-for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
+\hl{for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
 two new subgroups, without any particular ordering, that are defined by 
 splitting in split point $r$ of variable $Z_{l^{\ast}}$.
-The split point is then chosen as follows:
+The split point is then chosen as follows:}
 $$

### 2018-09-09 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
first draft of supplements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2277 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ca029d6448db189b0c0d78097537cf73806205c0

\documentclass[nojss]{jss}

\usepackage{graphicx, natbib}
\pagestyle{empty}
\usepackage{Sweave}
\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}

\title{Supplement 1: Testing different distributions}

\author{Lisa Schlosser\\Universit\"at Innsbruck
   \And Torsten Hothorn\\Universit\"at Z\"urich
   \And Reto Stauffer\\Universit\"at Innsbruck
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}

\begin{document}
\SweaveOpts{concordance=TRUE}

\section{Introduction}
This application study is supplementary material to the paper ''Distributional 
Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''. 
In Section~3 the presented distributional forest model is employed to obtain 
probabilistic precipitation forecasts in a mountainous region assuming that the 
response variable (the power-transformed daily precipitation sum) follows a 
left-censored Gaussian distribution with censoring point~$0$.

To assess how appropriate this distributional assumption is we test the robustness
of the results with respect to two questions: (1)~Are the tails of the distribution
captured appropriately by a Gaussian distribution or are \emph{heavier tails} required?
(2)~Is the point mass at zero (= no rain) driven by the same determinants as the
positive observations (= rain) or should this be captured by a \emph{separate parameter}?
For (1) we compare the censored Gaussian with a \emph{censored logistic} distribution and
for (2) a Gaussian \emph{hurdle model} (or two-part model) is employed instead of the
censored Gaussian.

The distribution can not only be exchanged for the distributional forest 
but for all the models evaluated in this application:

\begin{itemize}

\item \emph{Distributional forest:} All predictor variables are
  considered for learning a forest where bootstrap
  sampling is employed for each tree. Parameters are estimated by
  adaptive local likelihood based on forest weights.

\item \emph{EMOS:} Standard ensemble model output statistics
  models use the ensemble mean of total precipitation as regressor in the location
  submodel and the corresponding ensemble standard deviation in the scale submodel.
  The parameters are estimated by maximum likelihood, using an identity
  link for the location part and a log link for the scale part (following the advice of
  \cite{Gebetsberger+Messner+Mayr:2017}).

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
  the most relevant predictors based on meteorological expert knowledge
  following \cite{Stauffer+Umlauf+Messner:2017}. The model is estimated by maximum
  penalized likelihood using a backfitting algorithm (\cite{Stasinopoulos+Rigby:2007}).

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
  automatically from all available variables, using non-cyclic boosting
  for parameter estimation (\cite{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}).
  This updates the predictor terms for the location or scale submodels iteratively
  by maximizing the log-likelihood only for the variable yielding the
  biggest improvement. The iteration stops early -- before fully maximizing
  the in-sample likelihood -- based on a (computationally intensive)
  out-of-bag bootstrap estimate of the log-likelihood.

\end{itemize}

\section{Distributions applied / Distributional approaches}

The distribution applied in the paper is based on a Gaussian distribution 
for which the probability density function $f$ and the distribution 
function $F$ are defined by
\begin{align}
f_{\textit{Gaussian}}(Y; \mu, \sigma) &= 
     \frac{1}{\sigma} \cdot \phi\left(\frac{Y-\mu}{\sigma}\right) \\
F_{\textit{Gaussian}}(Y; \mu, \sigma) &=
    \Phi\left(\frac{Y-\mu}{\sigma}\right)
\end{align}
where $\phi$ and $\Phi$ are the probability density function and the 
distribution function of the standard normal distribution $\mathcal{N}(0,1)$.

Alternatively one could also choose a distribution with heavier tails 
such as a logistic distribution with corresponding probability density 
function and distribution function

\begin{align}
f_{\textit{logistic}}(Y; \mu, \sigma) &= 
     \frac{\exp{\left(-\frac{Y - \mu}{\sigma}\right)}}
     {\sigma \cdot \left(1 + \exp{\left(-\frac{Y - \mu}{\sigma}\right)}\right)^2} \\
F_{\textit{logistic}}(Y; \mu, \sigma) &=
     \frac{1}{1+ \exp{\left(\frac{\mu}{\sigma}\right)}}.
\end{align}

Regardless of the chosen distribution it has to be considered 
that precipitation data usually includes a point mass at zero.
In order to deal with this point mass one of two common strategies 
can be used.
\begin{itemize}
\item Censored distribution (\cite{Winkelmann:2006}): 
In general, censoring data usually leads 
to a point mass at the censoring point which has to be considered in 
the modeling process. Therefore, censored distributions are applied as 
they are able to capture this point mass. 
In precipitation data, observations can not be negative, thus, left-censoring
it at zero does not change the data. However, there is a naturally given
point mass at zero (days without rain) making a censored distribution an 
appropriate choice to model precipitation data.
For a chosen distribution with probability density function $f$ and
distribution function $F$ the log-likelihood function of the corresponding 
censored distribution is
\begin{equation}
\ell_{\textit{cens}}(\mu, \sigma; Y) = 
\begin{cases}
    \log\left\{f(Y; \mu, \sigma) \right\}, & \text{if } Y > 0 \\[0.2cm]
    \log\left\{F(0; \mu, \sigma) \right\}, & \text{if } Y = 0
\end{cases}
\end{equation}
as seen in Equation~2.1.1. of the paper for a Gaussian distribution.

\item Hurdle model / Two-part model (\cite{Cragg:1971}, \cite{Mullahy:1986}): 
An additional parameter $\nu$ describing the probability that there is 
any precipitation at all, 
i.e. $\nu = \mathbb{P}(Y>0)$, is estimated by a binary model. 
Then, a second model is fit to the subset of the data containing 
only positive observations applying a left-truncated at zero 
distribution to estimate the location parameter $\mu$ and scale 
parameter $\sigma$ of the underlying latent variable.
The three-parametric log-likelihood function for the full hurdle model 
with parameter vector $\bm{\theta} = (\mu, \sigma, \nu)$ is
\begin{equation} \label{eq:hurdle}
\ell_{\textit{hurdle}}(\mu, \sigma, \nu; Y) = 
\begin{cases}
    \log\left\{\nu \cdot \frac{f(Y; \mu, \sigma)}{1 - F(0; \mu, \sigma)} \right\}, & \text{if } Y > 0 \\[0.2cm]
    \log(1 - \nu), & \text{if } Y = 0.
\end{cases}
\end{equation}

\end{itemize}

In the paper a left-censored at zero Gaussian distribution is chosen.
Here, we additionally apply a left-censored at zero logistic distribution
and a hurdle model based on a Gaussian distribution.

In the log-likelihood function in \ref{eq:hurdle} properties of the
logarithmic function can be applied to split the term for $Y>0$ 
into two additive terms separating the parameters $\mu$ and $\sigma$
from the additional parameter $\nu$. Therefore, $\nu$ can be modeled 
by a separate model in the GAMLSS framework. This approach is also
chosen for the distributional forest model in this application study.

The models are compared based on the continuous ranked probability 
score (CRPS). For the censored models the closed form following 
\cite{Jordan:2018} is used while a numerical approximation is 
calculated for the hurdle models.

\section{Stations}
Out of the 95 observation stations 18 are considered for this 
benchmark study (see Figure~\ref{fig:map}). 
They have been selected based on the region
they are in, their altitude and the performance of the 
different models in the application in the paper such that
all of these features are represented equally.

\begin{figure}[h]
\vspace*{-3cm}
<<map, fig=TRUE, echo=FALSE>>=
##  map
data("StationsTyrol", package = "RainTyrol")
data("MapTyrol", package = "RainTyrol")
library("sp")
sp <- SpatialPointsDataFrame(subset(StationsTyrol, select = c(lon, lat)),
                             data = subset(StationsTyrol, select = -c(lon, lat)),
                             proj4string = raster::crs(MapTyrol$RasterLayer))

plot(MapTyrol$SpatialPolygons)
points(sp, pch = 21, col = "black", las = 1, cex = 1)
points(sp[c(5, 6, 16, 20, 23, 25, 32, 33, 46, 47, 48, 56, 57, 70, 71, 79, 82, 83),], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
@
\vspace*{-3cm}
\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and the 18 considered stations highlighted in pink.}
\end{figure}

\section{Results}

\begin{figure}[h]
\setkeys{Gin}{width=0.99\textwidth}
<<results_lattice, fig=TRUE, echo=FALSE, results=hide, height=10, width=8>>=
library("lattice")
##### 
# HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)

load("Rain_distributions.rda")
means <- results$means[results$means$method != "disttree",]
means$method <- factor(means$method, 
                       levels = c("distforest",
                                  "gamlss",
                                  "gamboostLSS",
                                  "EMOS"),
                       labels = c("Distributional forest",
                                  "Prespecified GAMLSS",
                                  "Boosted GAMLSS",
                                  "EMOS"))

strip.background.settings <- trellis.par.get("strip.background")
strip.background.settings$col <- "gray"
trellis.par.set("strip.background", strip.background.settings)

strip.border.settings <- trellis.par.get("strip.border")
strip.border.settings$col <- "black"
trellis.par.set("strip.border", strip.border.settings)

superpose.line.settings <- trellis.par.get("superpose.line")
superpose.line.settings$col <- hcl(c(128, 260, 290, 50), 100, 50)
trellis.par.set("superpose.line", superpose.line.settings)

superpose.symbol.settings <- trellis.par.get("superpose.symbol")
superpose.symbol.settings$pch <- c(16,24,25,15)
superpose.symbol.settings$col <- hcl(c(128, 260, 290, 50), 100, 50)
superpose.symbol.settings$fill <- hcl(c(128, 260, 290, 50), 100, 50)
trellis.par.set("superpose.symbol", superpose.symbol.settings)

xyplot(crps ~ distribution | station, groups = ~ method, 
       data = means, auto.key = TRUE, 
       type = "o", lwd = 2, lty = 1, layout = c(3,6))
@
\caption{\label{fig:lattice}CRPS values of all 4 methods, once applied with each of the three distributions for each of the 18 observation stations.}
\end{figure}

\begin{figure}[h]
\setkeys{Gin}{width=0.99\textwidth}
<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7>>=

  # crps
  par(mfrow = c(1,3))
  boxplot(crps~method, data = means, subset = distribution == "gaussian", main = "gaussian",
          ylim = c(0.65,1.15), axes = FALSE, ylab = "CRPS")
  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
  axis(1, 0:5, c("",
                 "Distr.
                 forest",
                 "Presp.
                 GAMLSS",
                 "Boosted
                 GAMLSS",
                 "EMOS", ""), las=2)
  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)

  boxplot(crps~method, data = means, subset = distribution == "hgaussian", main = "hgaussian",
          ylim = c(0.65,1.15), las = 2, axes = FALSE)
  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
  axis(1, 0:5, c("",
                 "Distr.
                 forest",
                 "Presp.
                 GAMLSS",
                 "Boosted
                 GAMLSS",
                 "EMOS", ""), las=2)
  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
  boxplot(crps~method, data = means, subset = distribution == "logistic", main = "logistic",
          ylim = c(0.65,1.15), las = 2, axes = FALSE)
  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
  axis(1, 0:5, c("",
                 "Distr.
                 forest",
                 "Presp.
                 GAMLSS",
                 "Boosted
                 GAMLSS",
                 "EMOS", ""), las=2)
  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)

@
\caption{\label{fig:box_crps}(Mean) CRPS values of each of the 18 observation stations, separated by method and distribution.}
\end{figure}

\begin{figure}[h]
\setkeys{Gin}{width=0.99\textwidth}
<<results_box_crpss_EMOS, fig=TRUE, echo=FALSE, height=3, width=7>>=
  # crps skill score by method (reference: EMOS)
  means$crps_ss_method <- means$crps
  means$crps_ss_method[means$method == "Distributional forest"] <- 
    1 - means$crps[means$method == "Distributional forest"] / means$crps[means$method == "EMOS"]
  means$crps_ss_method[means$method == "Prespecified GAMLSS"] <- 
    1 - means$crps[means$method == "Prespecified GAMLSS"] / means$crps[means$method == "EMOS"]
  means$crps_ss_method[means$method == "Boosted GAMLSS"] <- 
    1 - means$crps[means$method == "Boosted GAMLSS"] / means$crps[means$method == "EMOS"]
  means$crps_ss_method[means$method == "EMOS"] <- 
    1 - means$crps[means$method == "EMOS"] / means$crps[means$method == "EMOS"]

  means_sel <- means[means$method %in% c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),]
  means_sel$method <- factor(means_sel$method, levels(means_sel$method)[c(1:3)])

  par(mfrow = c(1,3))
  boxplot(crps_ss_method~method, data = means_sel, 
          subset = (distribution == "gaussian"), main = "gaussian",
          ylim = c(-0.12,0.23), axes = FALSE, ylab = "CRPS skill score")
  axis(2, seq(-0.15, 0.25, 0.05), c(seq(-0.15, 0, 0.05), seq(0.05, 0.25, 0.05)), las = 2)
  axis(1, 0:4, c("",
                 "Distr.
                 forest",
                 "Presp.
                 GAMLSS",
                 "Boosted
                 GAMLSS", ""), las=2)
  axis(3, 0:4, lwd.ticks = 0, labels = FALSE)
  axis(4, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
  abline(h = 0, col = pal[5], lwd = 2)
  boxplot(crps_ss_method~method, data = means_sel, 
          subset = (distribution == "logistic"), main = "logistic",
          ylim = c(-0.12,0.23), axes = FALSE)
  axis(2, seq(-0.15, 0.25, 0.05), c(seq(-0.15, 0, 0.05), seq(0.05, 0.25, 0.05)), las = 2)
  axis(1, 0:4, c("",
                 "Distr.
                 forest",
                 "Presp.
                 GAMLSS",
                 "Boosted
                 GAMLSS", ""), las=2)
  axis(3, 0:4, lwd.ticks = 0, labels = FALSE)
  axis(4, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
  abline(h = 0, col = pal[5], lwd = 2)
  boxplot(crps_ss_method~method, data = means_sel, 
          subset = (distribution == "hgaussian"), main = "hgaussian",
          ylim = c(-0.12,0.23), axes = FALSE)
  axis(2, seq(-0.15, 0.25, 0.05), c(seq(-0.15, 0, 0.05), seq(0.05, 0.25, 0.05)), las = 2)
  axis(1, 0:4, c("",
                 "Distr.
                 forest",
                 "Presp.
                 GAMLSS",
                 "Boosted
                 GAMLSS", ""), las=2)
  axis(3, 0:4, lwd.ticks = 0, labels = FALSE)
  axis(4, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
  abline(h = 0, col = pal[5], lwd = 2)
@
\caption{\label{fig:box_crpss_EMOS}(Mean) CRPS skill score values of each of the 18 observation stations, separated by method and distribution. EMOS is chosen as reference model.}
\end{figure}

\begin{figure}[h]
\setkeys{Gin}{width=0.99\textwidth}
<<results_box_crpss_gaussian, fig=TRUE, echo=FALSE, height=3, width=7>>=
  # crps skill score by distribution (reference: gaussian)
  means$crps_ss_dist <- means$crps
  means$crps_ss_dist[means$distribution == "gaussian"] <- 
    1 - means$crps[means$distribution == "gaussian"] / means$crps[means$distribution == "gaussian"]
  means$crps_ss_dist[means$distribution == "logistic"] <- 
    1 - means$crps[means$distribution == "logistic"] / means$crps[means$distribution == "gaussian"]
  means$crps_ss_dist[means$distribution == "hgaussian"] <- 
    1 - means$crps[means$distribution == "hgaussian"] / means$crps[means$distribution == "gaussian"]

  means_sel <- means[means$distribution %in% c("logistic", "hgaussian"),]
  means_sel$distribution <- factor(means_sel$distribution, levels(means_sel$distribution)[c(2:3)])

  par(mfrow = c(1,4))
  boxplot(crps_ss_dist~distribution, data = means_sel, 
          subset = (method == "Distributional forest"), main = "Distributional forest",
          ylim = c(-0.12,0.07), las = 2, ylab = "CRPS skill score")
  abline(h = 0, col = pal[5], lwd = 2)
  boxplot(crps_ss_dist~distribution, data = means_sel, 
          subset = (method == "Prespecified GAMLSS"), main = "Prespecified GAMLSS",
          ylim = c(-0.12,0.07), las = 2)
  abline(h = 0, col = pal[5], lwd = 2)
  boxplot(crps_ss_dist~distribution, data = means_sel, 
          subset = (method == "Boosted GAMLSS"), main = "Boosted GAMLSS",
          ylim = c(-0.12,0.07), las = 2)
  abline(h = 0, col = pal[5], lwd = 2)
  boxplot(crps_ss_dist~distribution, data = means_sel, 
          subset = (method == "EMOS"), main = "EMOS",
          ylim = c(-0.12,0.07), las = 2)
  abline(h = 0, col = pal[5], lwd = 2)
@
\caption{\label{fig:box_crpss_gaussian}(Mean) CRPS skill score values of each of the 18 observation stations, separated by method and distribution. Gaussian is chosen as reference distribution.}
\end{figure}

\section{Discussion}
As explained above the the parameter $\nu$ is modeled separately from $\mu$ and $\sigma$ 
in the GAMLSS framework which is why two separate models can be built. For the forest model
an alternative approach would be to build just one model employing the three-parametric
distribution family. However, testing both versions of the forest model in this application study shows that they both perform almost equally. Over the 18 considered stations the 
absolute difference in terms of CRPS between the two-forest model and the three-parametric forest is on average 0.0078 and lies between 0.0014 and 0.0267 for each station. Comparing
these values to the mean CRPS values in Figure~\ref{fig:box_crps} lying between 0.6 and 1.2
confirms the relatively small difference between the two versions of the forest model.

<<compare_forests, results=hide, echo=FALSE>>=
  load("crps_forests.rda")
  sum(crps_forests[,"forest_2p"] > crps_forests[,"forest_h"])
  summary(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"])
  summary(abs(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"]))
  colMeans(crps_forests)  
@

\bibliography{ref_supplement_distributions.bib}

\end{document}

### 2018-09-09 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
first draft of supplements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2277 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ca029d6448db189b0c0d78097537cf73806205c0

\documentclass[a4paper,11pt]{article}

\usepackage{Sweave}
\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}

\begin{document}
\SweaveOpts{concordance=TRUE}

\title{Supplement 2: Stationwise Evaluation}
\maketitle

\section{Introduction}
This application study is supplementary material to the paper ''Distributional 
Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''. 
In Section~3 the presented distributional forest model is employed to obtain 
probabilistic precipitation forecasts in a mountainous region.
Next to an evaluation over all 95 observation stations further investigations
including a cross-validation have been performed for one selected station, Axams.
Here, this more extensive evaluation is carried out for 18 observation stations.

\section{Models and evaluation}
As in the paper a distributional forest applying a left-censored at zero 
Gaussian distribution is evaluated together with three other zero-censored 
Gaussian models: (a)~a standard basic EMOS, (b)~a GAMLSS with prespecified 
effects and interactions based on meteorological knowledge/experience, and 
(c)~a boosted GAMLSS with automatic selection of smooth additive terms based 
on all 80~predictor variables.

First, they are learned on data from 24 years (1985--2008) and evaluated 
on the following 4 years (2009--2012). 
To compare the performance of the
models the continuous ranked probability score (CRPS) is calculated
for each observation in the testing data set and than averaged.
Moreover, for each model it is timed, how long the fitting process takes 
as well as how long it takes them to make one single prediction 
(on average over all testing observations).

For the forest model variable importance is analyzed and the top 10
variables are illustrated. All four models are also compared
visually by residual Q-Q-plots and PIT-histograms.

Second, all four models are evaluated in a 10 times 7-fold cross-validation.
Again, their performances are compared based on the average CRPS values 
and the fitting time is meassured for all models as well as the prediction
time. however, contrary to the first model, the prediction time now refers 
to the time needed on average to make predictions for the whole testing data, 
not just for one observation.
For the cross-validation the CRPS skill score is illustrated in a boxplot 
where the EMOS model is chosen as reference model, represented by the horizontal
line at 0.

\section{Stations}
Out of the 95 observation stations 18 are considered for this 
benchmark study. They have been selected based on the region
they are in, their altitude and the performance of the 
different models in the application in the paper such that
all of these features are represented equally.

\vspace*{-3cm}

\begin{figure}[h]
<<map_stationwise, fig=TRUE, echo=FALSE>>=
##  map
data("StationsTyrol", package = "RainTyrol")
data("MapTyrol", package = "RainTyrol")
library("sp")
sp <- SpatialPointsDataFrame(subset(StationsTyrol, select = c(lon, lat)),
                             data = subset(StationsTyrol, select = -c(lon, lat)),
                             proj4string = raster::crs(MapTyrol$RasterLayer))

plot(MapTyrol$SpatialPolygons)
points(sp, pch = 21, col = "black", las = 1, cex = 1)
points(sp[c(5, 6, 16, 20, 23, 25, 32, 33, 46, 47, 48, 56, 57, 70, 71, 79, 82, 83),], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
@
\vspace*{-3cm}
\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and the 18 considered stations highlighted in pink.}
\end{figure}

\section{Discussion}
Looking at prediction times for one single observation, the distributional forest is on average
the slowest model, however, the difference is not too big. This changes regarding predictions
for the a set of observation due to the fact that the forest makes predictions based
on local neighbour weights. Therefor, for each observation its own weight matrix has to be included in the prediction process making it a more time consuming procedure.

%\bibliography{ref_supplement_stationwise.bib}

\end{document}

### 2018-09-14 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
revised version of supplement 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2282 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 01c0f077e4047722b3019c3d9da67a06e3535524

@@
-\documentclass[nojss]{jss}
+%\documentclass[a4paper,11pt]{article}
+\documentclass[nojss, shortnames]{jss}
 
@@
 \Plainauthor{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
+
+\Abstract{In the application discussed in the paper ''Distributional 
+Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+the underlying distribution is assumed to a be left-censored at zero Gaussian distribution.
+In this suppplement the exact same evaluations as in the paper are performed employing
+other distribution assumption. In particular, a left-censored at zero logistic distribution 
+and a hurdle-model are applied and their results are compared to those presented in the paper.}
+
+\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
+
+\Address{
+  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
+  Universit\"at Innsbruck \\
+  Department of Statistics \\
+  Faculty of Economics and Statistics \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
+  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
+  \email{Achim.Zeileis@R-project.org} \\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
+  \phantom{URL: }\url{https://retostauffer.org/},\\
+  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
+
+  Torsten Hothorn\\
+  Universit\"at Z\"urich \\
+  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
+  Hirschengraben 84\\
+  CH-8001 Z\"urich, Switzerland \\
+  E-mail: \email{Torsten.Hothorn@R-project.org}\\
+  URL: \url{http://user.math.uzh.ch/hothorn/}\\
+
+}
 
@@
 
-The distribution can not only be exchanged for the distributional forest 
-but for all the models evaluated in this application:
+To compare the performance of the distributional forest to other commonly used methods the exact same models as in the paper are evaluated on the same data set and applying the same settings while only the distribution is exchanged.
+Further details on the models can be found in Section~3.2..
+In particular, Table~2 gives an overview over the models and their specifications.
+As a brief reminder the approaches are summarized below.
 
@@
 
-\item \emph{EMOS:} Standard ensemble model output statistics
-  models use the ensemble mean of total precipitation as regressor in the location
-  submodel and the corresponding ensemble standard deviation in the scale submodel.
-  The parameters are estimated by maximum likelihood, using an identity
-  link for the location part and a log link for the scale part (following the advice of
-  \cite{Gebetsberger+Messner+Mayr:2017}).
+\item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
+Standard ensemble model output statistics models use the ensemble 
+mean of total precipitation as regressor in the location submodel 
+and the corresponding ensemble standard deviation in the scale submodel.
+The parameters are estimated by maximum likelihood.
 
@@
 for parameter estimation (\cite{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}).
-  This updates the predictor terms for the location or scale submodels iteratively
-  by maximizing the log-likelihood only for the variable yielding the
-  biggest improvement. The iteration stops early -- before fully maximizing
-  the in-sample likelihood -- based on a (computationally intensive)
-  out-of-bag bootstrap estimate of the log-likelihood.
+The optimal stopping point for the iteration is evaluated based on 
+a (computationally intensive) out-of-bag bootstrap estimate of the log-likelihood.
 
@@
 F_{\textit{logistic}}(Y; \mu, \sigma) &=
-     \frac{1}{1+ \exp{\left(\frac{\mu}{\sigma}\right)}}.
+     \frac{1}{1+ \exp{\left(-\frac{Y - \mu}{\sigma}\right)}}.
 \end{align}
 
-Regardless of the chosen distribution it has to be considered 
-that precipitation data usually includes a point mass at zero.
-In order to deal with this point mass one of two common strategies 
-can be used.
+Both distributions have to be adapted in our application to capture that
+precipitation is non-negative with a point mass at zero (= no rain).
+Two common strategies for this are to employ either \emph{left-censoring at zero}
+or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
+a truncated (postive) part \citep[see][for an overview]{Winkelmann:2006}.
 \begin{itemize}
-\item Censored distribution (\cite{Winkelmann:2006}): 
-In general, censoring data usually leads 
-to a point mass at the censoring point which has to be considered in 
-the modeling process. Therefore, censored distributions are applied as 
-they are able to capture this point mass. 
-In precipitation data, observations can not be negative, thus, left-censoring
-it at zero does not change the data. However, there is a naturally given
-point mass at zero (days without rain) making a censored distribution an 
-appropriate choice to model precipitation data.
+\item Censored distribution \citep{Winkelmann:2006}: 
+%In general, censoring data usually leads 
+%to a point mass at the censoring point which has to be considered in 
+%the modeling process. Therefore, censored distributions are applied as 
+%they are able to capture this point mass. 
+%In precipitation data, observations can not be negative, thus, left-censoring
+%it at zero does not change the data. However, there is a naturally given
+%point mass at zero (days without rain) making a censored distribution an 
+%appropriate choice to model precipitation data.
 For a chosen distribution with probability density function $f$ and
@@
 as seen in Equation~2.1.1. of the paper for a Gaussian distribution.
+For positive values of $Y$ the censored version of the density function
+equals the original non-censored version. For $Y=0$ the censored density
+function takes the values of the probability mass lying left of the 
+censoring point. Therefore, the parameters $\mu$ and $\sigma$ control
+both parts, the point mass and the positive part of the distribution.
 
@@
 An additional parameter $\nu$ describing the probability that there is 
-any precipitation at all, 
-i.e. $\nu = \mathbb{P}(Y>0)$, is estimated by a binary model. 
-Then, a second model is fit to the subset of the data containing 
-only positive observations applying a left-truncated at zero 
-distribution to estimate the location parameter $\mu$ and scale 
-parameter $\sigma$ of the underlying latent variable.
+any precipitation at all, i.e. $\nu = \mathbb{P}(Y>0)$, is included in the model.
+%Then, a second model is fit to the subset of the data containing 
+%only positive observations applying a left-truncated at zero 
+%distribution to estimate the location parameter $\mu$ and scale 
+%parameter $\sigma$ of the underlying latent variable.
 The three-parametric log-likelihood function for the full hurdle model 
@@
 \end{equation}
-
+For $Y=0$ the probability that it won't rain is expressed by $1-\nu$.
+For $Y>0$ the probability that it will rain $\nu$ is multiplied with the left-truncated at zero density function.
+Contrary to the censored model, here it is only the parameter $\nu$ controlling the point mass while $\mu$ and $\sigma$ only influence the distribution of the positive observations.
 \end{itemize}
@@
 \section{Stations}
-Out of the 95 observation stations 18 are considered for this 
+Out of the 95 observation stations 15 are considered for this 
 benchmark study (see Figure~\ref{fig:map}). 
@@
 \vspace*{-3cm}
-<<map, fig=TRUE, echo=FALSE>>=
+\setkeys{Gin}{width=0.99\textwidth}
+<<map_stationwise, fig=TRUE, echo=FALSE>>=
 ##  map
@@
 plot(MapTyrol$SpatialPolygons)
-points(sp, pch = 21, col = "black", las = 1, cex = 1)
-points(sp[c(5, 6, 16, 20, 23, 25, 32, 33, 46, 47, 48, 56, 57, 70, 71, 79, 82, 83),], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
+points(sp, pch = 21, col = "darkgrey", las = 1, cex = 0.6)
+points(sp[c(5, 6, 20, 23, 25, 32, 33, 46, 47, 56, 57, 70, 79, 82, 83),], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
+# stationname beneath
+text(sp[c(25, 47, 57, 70),], 
+     labels = StationsTyrol$name[c(25, 47, 57, 70)],
+     pos=1, cex=0.75)
+# stationname left
+text(sp[c(20, 32, 46, 79, 83),], 
+     labels = StationsTyrol$name[c(20, 32, 46, 79, 83)],
+     pos=2, cex=0.75)
+# stationname above
+text(sp[c(6, 33, 82),], 
+     labels = StationsTyrol$name[c(6, 33, 82)],
+     pos=3, cex=0.75)
+# stationname right
+text(sp[c(5, 23, 56),], 
+     labels = StationsTyrol$name[c(5, 23, 56)],
+     pos=4, cex=0.75)
 @
-\vspace*{-3cm}
-\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and the 18 considered stations highlighted in pink.}
+\vspace*{-3.5cm}
+\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and the 15 considered stations highlighted in pink.}
 \end{figure}
@@
 \section{Results}
+In Figure~\ref{fig:lattice} each panel belongs to one of the 15 observation stations.
+The three distributions (left-censored at zero Gaussian distribution = \textit{gaussian}, hurdle model based on the Gaussian distribution = \textit{hgaussian}, left-censored at zero logistic distribution = \textit{logistic}) are compared based on the resulting CRPS values for each of the applied methods that are represented by different colours and symbols.
+Overall it can be stated that the forest model performs almost equally well with all three distributions.
+Therefore it seems to be very robust against missspecifications of the distribution.
+For the two GAMLSS models the censored Gaussian and the hurdle model perform similarly at most of the station while applying the censored logistic distribution leads to an increase in CRPS, i.e.
+to a decline in performance, especially for the prespecified GAMLSS.
+For the EMOS model exchanging the distribution does not have a strong effect on its performance, similar to the forest model.
+In particular, for the two censored distributions the results are almost equal, only for the hurdle model the CRPS values vary slightly, however in both directions, depending on the station.
+
+When applying a hurdle model for the stations Oetz and Innvervillgraten the prespecified GAMLSS computationally failed to build/fit a model.
+Therefore this method is not represented in Figure~\ref{fig:lattice} for these two cases.
+
+In stations like Ladis-Neuegg, Matrei in Osttirol or Oetz all methods seem to be very robust and stable over varying distributions.
+The strongest influence of an exchange of the distribution can be seen for example at stations Jungholz and St.Johann im Walde for the prespecified GAMLSS.
+
+In Figure\ref{fig:box_crpss_gaussian} the tested distributions are compared for each method based on CRPS skill score with the censored Gaussian distribution as reference represented by the horizontal line at zero.
+The hurdle model leads to a slight improvement for the forest model and the prespecified GAMLSS.
+Following the median this improvement is a bit stronger for the boosted GAMLSS, however, with a higher deviation as represented by the larger size of the box, similarly as for the EMOS, however with a negative influence indicated by the negative median.
+Choosing a logistic distribution instead of a Gaussian clearly reduces the strength of the two GAMLSS models, while it has almost no influence on the forest and the EMOS models.
+
+Alltogether it can be stated that the left-censored at zero Gaussian distribution seems to be an appropriate choice while the other two distributions are possible and mostly also reasonable alternatives.
+However, they do not lead to an overall improvement in performance for this application.
 
@@
 @
-\caption{\label{fig:lattice}CRPS values of all 4 methods, once applied with each of the three distributions for each of the 18 observation stations.}
+\caption{\label{fig:lattice}CRPS values of all 4 methods, once applied with each of the three distributions for each of the 15 observation stations.}
 \end{figure}
@@
 \setkeys{Gin}{width=0.99\textwidth}
-<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7>>=
+<<results_box_crpss_gaussian, fig=TRUE, echo=FALSE, height=3, width=7>>=
+  # crps skill score by distribution (reference: gaussian)
+  means$crps_ss_dist <- means$crps
+  means$crps_ss_dist[means$distribution == "gaussian"] <- 
+    1 - means$crps[means$distribution == "gaussian"] / means$crps[means$distribution == "gaussian"]
+  means$crps_ss_dist[means$distribution == "logistic"] <- 
+    1 - means$crps[means$distribution == "logistic"] / means$crps[means$distribution == "gaussian"]
+  means$crps_ss_dist[means$distribution == "hgaussian"] <- 
+    1 - means$crps[means$distribution == "hgaussian"] / means$crps[means$distribution == "gaussian"]
 
-  # crps
-  par(mfrow = c(1,3))
-  boxplot(crps~method, data = means, subset = distribution == "gaussian", main = "gaussian",
-          ylim = c(0.65,1.15), axes = FALSE, ylab = "CRPS")
-  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
-  axis(1, 0:5, c("",
-                 "Distr.
-                 forest",
-                 "Presp.
-                 GAMLSS",
-                 "Boosted
-                 GAMLSS",
-                 "EMOS", ""), las=2)
-  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
-  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
+  means_sel <- means[means$distribution %in% c("logistic", "hgaussian"),]
+  means_sel$distribution <- factor(means_sel$distribution, levels(means_sel$distribution)[c(2:3)])
 
-  boxplot(crps~method, data = means, subset = distribution == "hgaussian", main = "hgaussian",
-          ylim = c(0.65,1.15), las = 2, axes = FALSE)
-  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
-  axis(1, 0:5, c("",
-                 "Distr.
-                 forest",
-                 "Presp.
-                 GAMLSS",
-                 "Boosted
-                 GAMLSS",
-                 "EMOS", ""), las=2)
-  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
-  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
-  boxplot(crps~method, data = means, subset = distribution == "logistic", main = "logistic",
-          ylim = c(0.65,1.15), las = 2, axes = FALSE)
-  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
-  axis(1, 0:5, c("",
-                 "Distr.
-                 forest",
-                 "Presp.
-                 GAMLSS",
-                 "Boosted
-                 GAMLSS",
-                 "EMOS", ""), las=2)
-  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
-  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
-
+  par(mfrow = c(1,4))
+  boxplot(crps_ss_dist~distribution, data = means_sel, 
+          subset = (method == "Distributional forest"), main = "Distributional forest",
+          ylim = c(-0.12,0.07), las = 2, ylab = "CRPS skill score")
+  abline(h = 0, col = pal[5], lwd = 2)
+  boxplot(crps_ss_dist~distribution, data = means_sel, 
+          subset = (method == "Prespecified GAMLSS"), main = "Prespecified GAMLSS",
+          ylim = c(-0.12,0.07), las = 2)
+  abline(h = 0, col = pal[5], lwd = 2)
+  boxplot(crps_ss_dist~distribution, data = means_sel, 
+          subset = (method == "Boosted GAMLSS"), main = "Boosted GAMLSS",
+          ylim = c(-0.12,0.07), las = 2)
+  abline(h = 0, col = pal[5], lwd = 2)
+  boxplot(crps_ss_dist~distribution, data = means_sel, 
+          subset = (method == "EMOS"), main = "EMOS",
+          ylim = c(-0.12,0.07), las = 2)
+  abline(h = 0, col = pal[5], lwd = 2)
 @
-\caption{\label{fig:box_crps}(Mean) CRPS values of each of the 18 observation stations, separated by method and distribution.}
+\caption{\label{fig:box_crpss_gaussian}(Mean) CRPS skill score values of each of the 15 observation stations, separated by method and distribution. Gaussian is chosen as reference distribution.}
 \end{figure}
@@
 @
-\caption{\label{fig:box_crpss_EMOS}(Mean) CRPS skill score values of each of the 18 observation stations, separated by method and distribution. EMOS is chosen as reference model.}
+\caption{\label{fig:box_crpss_EMOS}(Mean) CRPS skill score values of each of the 15 observation stations, separated by method and distribution. EMOS is chosen as reference model.}
 \end{figure}
@@
 \setkeys{Gin}{width=0.99\textwidth}
-<<results_box_crpss_gaussian, fig=TRUE, echo=FALSE, height=3, width=7>>=
-  # crps skill score by distribution (reference: gaussian)
-  means$crps_ss_dist <- means$crps
-  means$crps_ss_dist[means$distribution == "gaussian"] <- 
-    1 - means$crps[means$distribution == "gaussian"] / means$crps[means$distribution == "gaussian"]
-  means$crps_ss_dist[means$distribution == "logistic"] <- 
-    1 - means$crps[means$distribution == "logistic"] / means$crps[means$distribution == "gaussian"]
-  means$crps_ss_dist[means$distribution == "hgaussian"] <- 
-    1 - means$crps[means$distribution == "hgaussian"] / means$crps[means$distribution == "gaussian"]
+<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7>>=
 
-  means_sel <- means[means$distribution %in% c("logistic", "hgaussian"),]
-  means_sel$distribution <- factor(means_sel$distribution, levels(means_sel$distribution)[c(2:3)])
+  # crps
+  par(mfrow = c(1,3))
+  boxplot(crps~method, data = means, subset = distribution == "gaussian", main = "gaussian",
+          ylim = c(0.65,1.15), axes = FALSE, ylab = "CRPS")
+  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
+  axis(1, 0:5, c("",
+                 "Distr.
+                 forest",
+                 "Presp.
+                 GAMLSS",
+                 "Boosted
+                 GAMLSS",
+                 "EMOS", ""), las=2)
+  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
+  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
 
-  par(mfrow = c(1,4))
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "Distributional forest"), main = "Distributional forest",
-          ylim = c(-0.12,0.07), las = 2, ylab = "CRPS skill score")
-  abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "Prespecified GAMLSS"), main = "Prespecified GAMLSS",
-          ylim = c(-0.12,0.07), las = 2)
-  abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "Boosted GAMLSS"), main = "Boosted GAMLSS",
-          ylim = c(-0.12,0.07), las = 2)
-  abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "EMOS"), main = "EMOS",
-          ylim = c(-0.12,0.07), las = 2)
-  abline(h = 0, col = pal[5], lwd = 2)
+  boxplot(crps~method, data = means, subset = distribution == "hgaussian", main = "hgaussian",
+          ylim = c(0.65,1.15), las = 2, axes = FALSE)
+  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
+  axis(1, 0:5, c("",
+                 "Distr.
+                 forest",
+                 "Presp.
+                 GAMLSS",
+                 "Boosted
+                 GAMLSS",
+                 "EMOS", ""), las=2)
+  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
+  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
+  boxplot(crps~method, data = means, subset = distribution == "logistic", main = "logistic",
+          ylim = c(0.65,1.15), las = 2, axes = FALSE)
+  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
+  axis(1, 0:5, c("",
+                 "Distr.
+                 forest",
+                 "Presp.
+                 GAMLSS",
+                 "Boosted
+                 GAMLSS",
+                 "EMOS", ""), las=2)
+  axis(3, 0:5, lwd.ticks = 0, labels = FALSE)
+  axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
+
 @
-\caption{\label{fig:box_crpss_gaussian}(Mean) CRPS skill score values of each of the 18 observation stations, separated by method and distribution. Gaussian is chosen as reference distribution.}
+\caption{\label{fig:box_crps}(Mean) CRPS values of each of the 15 observation stations, separated by method and distribution.}
 \end{figure}
@@
 \section{Discussion}
+As mentioned by \cite{Hofner+Mayr+Schmid:2016} the methods provided by GAMLSS ``can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters''.
+These problems have also occured in a few situations for the
+prespecified GAMLSS model within the cross-validation. In case of this method failing to 
+build a model it is represented by \code{NA}s in the evaluation.
+
 As explained above the the parameter $\nu$ is modeled separately from $\mu$ and $\sigma$ 
@@
 an alternative approach would be to build just one model employing the three-parametric
-distribution family. However, testing both versions of the forest model in this application study shows that they both perform almost equally. Over the 18 considered stations the 
-absolute difference in terms of CRPS between the two-forest model and the three-parametric forest is on average 0.0078 and lies between 0.0014 and 0.0267 for each station. Comparing
-these values to the mean CRPS values in Figure~\ref{fig:box_crps} lying between 0.6 and 1.2
-confirms the relatively small difference between the two versions of the forest model.
+distribution family. However, testing both versions of the forest model in this application study shows that they both perform almost equally. Figure~\ref{fig:box_crpss_forests} shows
+the CRPS skill score of the three-parametric forest model over the considered 15 observation
+stations with the two-part forest model as reference. Even though this alternative version
+of the forest model seems to perform better on average, according to the median this 
+improvement is not even one percent, hence neglectably small. Therefore, the two foreset version can be stated to perform almost equally.
 
-<<compare_forests, results=hide, echo=FALSE>>=
+\begin{figure}[h]
+\center
+\setkeys{Gin}{width=0.5\textwidth}
+<<compare_forests, fig=TRUE, results=hide, echo=FALSE, width=4, height=4>>=
   load("crps_forests.rda")
-  sum(crps_forests[,"forest_2p"] > crps_forests[,"forest_h"])
-  summary(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"])
-  summary(abs(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"]))
-  colMeans(crps_forests)  
+  crps_ss_forest_2p <- data.frame(1 - crps_forests[,"forest_2p"] / crps_forests[,"forest_h"])
+  names(crps_ss_forest_2p) <- "Three-parametric forest"
+  boxplot(crps_ss_forest_2p, ylab = "CRPS skill score", names = c("Three-parametric forest"))
+  abline(h = 0, col = pal[5], lwd = 2)
+  axis(1, 0:2, c("","Three-parametric forest", ""), las=1)
+  #sum(crps_forests[,"forest_2p"] > crps_forests[,"forest_h"])
+  #summary(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"])
+  #summary(abs(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"]))
+  #colMeans(crps_forests)  
 @
+\caption{\label{fig:box_crpss_forests}(Mean) CRPS skill score values of each of the 15 observation stations for the three-parametric forest model with the two part forest model as reference model.}
+\end{figure}
 
-\bibliography{ref_supplement_distributions.bib}
+%bibliographystyle{jss}
+\bibliography{ref_supplement.bib}
 

### 2018-09-15 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
revised versions of supplement 1 and 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2283 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 13a7f6f3c52fe3004fd87673ab56c156c7cb84e3

@@
-\documentclass[a4paper,11pt]{article}
+%\documentclass[a4paper,11pt]{article}
+\documentclass[nojss, shortnames]{jss}
 
+\usepackage{graphicx, natbib}
+\pagestyle{empty}
 \usepackage{Sweave}
-\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}
+\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,pdfpages,hyperref}
+
+\title{Supplement 2: Stationwise Evaluation}
+
+\author{Lisa Schlosser\\Universit\"at Innsbruck
+   \And Torsten Hothorn\\Universit\"at Z\"urich
+   \And Reto Stauffer\\Universit\"at Innsbruck
+   \And Achim Zeileis\\Universit\"at Innsbruck}
+\Plainauthor{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
+
+\Abstract{In the application discussed in the paper ''Distributional 
+Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+the presented method is evaluated on 95 observation stations. More extensive
+evaluations including a cross-validation are preformed for one station, Axams.
+In this supplement this exact same analysis as for station Axams is
+made for 14 further observation stations.}
+
+\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
+
+\Address{
+  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
+  Universit\"at Innsbruck \\
+  Department of Statistics \\
+  Faculty of Economics and Statistics \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
+  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
+  \email{Achim.Zeileis@R-project.org} \\
+  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
+  \phantom{URL: }\url{https://retostauffer.org/},\\
+  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
+
+  Torsten Hothorn\\
+  Universit\"at Z\"urich \\
+  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
+  Hirschengraben 84\\
+  CH-8001 Z\"urich, Switzerland \\
+  E-mail: \email{Torsten.Hothorn@R-project.org}\\
+  URL: \url{http://user.math.uzh.ch/hothorn/}\\
+
+}
 
@@
 \SweaveOpts{concordance=TRUE}
-
-\title{Supplement 2: Stationwise Evaluation}
-\maketitle
 
@@
 including a cross-validation have been performed for one selected station, Axams.
-Here, this more extensive evaluation is carried out for 18 observation stations.
+Here, this more extensive evaluation is carried out, applying the exact same settings 
+as in the paper, for Axams and 14 further observation stations. Additionally, computation
+times are measured.
 
@@
 Gaussian distribution is evaluated together with three other zero-censored 
-Gaussian models: (a)~a standard basic EMOS, (b)~a GAMLSS with prespecified 
-effects and interactions based on meteorological knowledge/experience, and 
-(c)~a boosted GAMLSS with automatic selection of smooth additive terms based 
-on all 80~predictor variables.
+Gaussian models. Further details on all four models can be found in Section~3.2.. 
+In particular, Table~2 gives an overview over the models and their specifications. 
+As a brief reminder the approaches are summarized below.
 
-First, they are learned on data from 24 years (1985--2008) and evaluated 
-on the following 4 years (2009--2012). 
-To compare the performance of the
-models the continuous ranked probability score (CRPS) is calculated
-for each observation in the testing data set and than averaged.
-Moreover, for each model it is timed, how long the fitting process takes 
-as well as how long it takes them to make one single prediction 
-(on average over all testing observations).
+\begin{itemize}
 
-For the forest model variable importance is analyzed and the top 10
-variables are illustrated. All four models are also compared
-visually by residual Q-Q-plots and PIT-histograms.
+\item \emph{Distributional forest:} All predictor variables are
+considered for learning a forest where bootstrap
+sampling is employed for each tree. Parameters are estimated by
+adaptive local likelihood based on forest weights.
 
-Second, all four models are evaluated in a 10 times 7-fold cross-validation.
-Again, their performances are compared based on the average CRPS values 
-and the fitting time is meassured for all models as well as the prediction
-time. however, contrary to the first model, the prediction time now refers 
-to the time needed on average to make predictions for the whole testing data, 
-not just for one observation.
-For the cross-validation the CRPS skill score is illustrated in a boxplot 
-where the EMOS model is chosen as reference model, represented by the horizontal
-line at 0.
+\item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
+Standard ensemble model output statistics models use the ensemble 
+mean of total precipitation as regressor in the location submodel 
+and the corresponding ensemble standard deviation in the scale submodel.
+The parameters are estimated by maximum likelihood.
+
+\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
+the most relevant predictors based on meteorological expert knowledge
+following \cite{Stauffer+Umlauf+Messner:2017}. The model is estimated by maximum
+penalized likelihood using a backfitting algorithm (\cite{Stasinopoulos+Rigby:2007}).
+
+\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
+automatically from all available variables, using non-cyclic boosting
+for parameter estimation (\cite{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}).
+The optimal stopping point for the iteration is evaluated based on 
+a (computationally intensive) out-of-bag bootstrap estimate of the log-likelihood.
+
+\end{itemize}
+
+%\subsection{Subset selection}
+The models are evaluated in two different scenarios.
+\begin{itemize}
+\item Ordered/Base model: The learning data consists of the first 24 years (1985--2008) and the testing data of the following 4 years (2009--2012). 
+\item Cross-validation: The 28 years are randomly split in 7 blocks of 4 years which are not necessarily successive.
+Each of the blocks is chosen once as testing data while the remaining 6 blocks are the learning data.
+This procedure is repeated 10 times resulting in a 10 times 7-fold cross-validation framework.
+
+\end{itemize}
+
+The performance of the models are compared based on the following evaluations.
+
+\begin{itemize}
+\item CRPS: The continuous ranked probability score (CRPS) is calculated
+for each observation in the testing data set and then averaged. For the cross-validation
+setup the mean of the resulting 70 CRPS average values is listed and all 70 average values
+are illustrated in a boxplot where the EMOS model is chosen as reference model, 
+represented by the horizontal line at 0.
+
+\item Fit time: It is measured how long it takes to fit the model to the learning data.
+
+\item Predict time: It is measured how long it takes the model to make predictions.
+For the ordered model the predict time refers to the time needed to make one single prediction.
+The mean predict time over all observations in the test data is considered.
+For the cross-validation the predict time refers to the time needed to make all predictions for the test data. The mean predict time over all 70 evaluations in the cross-validation is considered.
+
+\item Variable importance: For the forest model it is investigated which variables have the strongest influence on the model and its goodness of fit.
+For each variable this is measured based on mean decrease in CRPS after permutation.
+The 10 variables showing the highest values are listed in a barplot.
+
+\item Residual QQ plots: For each of the models an out-of-sample residual QQ plot for the ordered model provides a visualization of goodness of fit.
+Randomized residual quantiles are compared to standard normal quantiles.
+
+\item PIT histograms: As an additional visualization out-of-sample probability integral transform histograms illustrated how well-calibrated the distributional fits are for the ordered model.
+
+\end{itemize}
 
 \section{Stations}
-Out of the 95 observation stations 18 are considered for this 
+Out of the 95 observation stations 15 are considered for this 
 benchmark study. They have been selected based on the region
@@
 
-\vspace*{-3cm}
+Stations: \hyperlink{sumAxams.1}{Axams},
+\hyperlink{sumLech.1}{Lech},
+\hyperlink{sumZuers.1}{Zuers},
+\hyperlink{sumSeeimPaznaun.1}{See im Paznaun},
+\hyperlink{sumOetz.1}{Oetz},
+\hyperlink{sumJungholz.1}{Jungholz},
+\hyperlink{sumKoessen.1}{Koessen},
+\hyperlink{sumWalchsee.1}{Walchsee},
+\hyperlink{sumGinzling.1}{Ginzling},
+\hyperlink{sumRotholz.1}{Rotholz},
+\hyperlink{sumInnervillgraten.1}{Innervillgraten},
+\hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol},
+\hyperlink{sumStJohannimWalde.1}{St.Johann im Walde},
+\hyperlink{sumLadisNeuegg.1}{Ladis-Neuegg},
+\hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut},
 
 \begin{figure}[h]
+\setkeys{Gin}{width=0.99\textwidth}
+\vspace*{-3cm}
 <<map_stationwise, fig=TRUE, echo=FALSE>>=
@@
 plot(MapTyrol$SpatialPolygons)
-points(sp, pch = 21, col = "black", las = 1, cex = 1)
-points(sp[c(5, 6, 16, 20, 23, 25, 32, 33, 46, 47, 48, 56, 57, 70, 71, 79, 82, 83),], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
+points(sp, pch = 21, col = "darkgray", las = 1, cex = 0.6)
+points(sp[c(5, 6, 20, 23, 25, 32, 33, 46, 47, 56, 57, 70, 79, 82, 83),], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
+# stationname beneath
+text(sp[c(25, 47, 57, 70),], 
+     labels = StationsTyrol$name[c(25, 47, 57, 70)],
+     pos=1, cex=0.75)
+# stationname left
+text(sp[c(20, 32, 46, 79, 83),], 
+     labels = StationsTyrol$name[c(20, 32, 46, 79, 83)],
+     pos=2, cex=0.75)
+# stationname above
+text(sp[c(6, 33, 82),], 
+     labels = StationsTyrol$name[c(6, 33, 82)],
+     pos=3, cex=0.75)
+# stationname right
+text(sp[c(5, 23, 56),], 
+     labels = StationsTyrol$name[c(5, 23, 56)],
+     pos=4, cex=0.75)
 @
-\vspace*{-3cm}
-\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and the 18 considered stations highlighted in pink.}
+\vspace*{-3.5cm}
+\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and the 15 considered stations highlighted in pink.}
 \end{figure}
 
+\section{Results}
+Comparing CRPS skill score values from the cross-validation the distributional forest
+and the two GAMLSS versions show clear improvements compared to the reference model EMOS.
+At many stations the forest and either the prespecified GAMLSS (e.g.
+at station Innervillgraten) or the boosted GAMLSS (e.g.
+at station Lech or Zuers) are ahead, almost on the same level, while the second GAMLSS is slightly behind.
+At stations Axams, See im Paznaun and Ginzling the forest clearly leads to the highest improvement in terms of CRPS.
+Station St.Johann im Walde stands out as it is the only of the 15 stations where the prespecified GAMLSS is clearly ahead of the other models.
+But overall it can be stated that all three models perform well with only small variations over stations.
+Even if one of the evaluated models shows clear advantages in the ordered model (such as the EMOS at station Rotholz or the prespecified GAMLSS at station Innervillgraten) this is evened out within the cross-validation.
+
+Concerning fitting times either the prespecified GAMLSS (e.g.
+St.Johann im Walde, Innervillgraten) or the distributional forest (e.g.
+Zuers, Walchsee, Jungholz, Oetz, Koessen, Ginzling) are the fastest while the boosted GAMLSS usually takes more time, especially if the time required for the selection of the optimal value for \code{mstop} a multiple of the times of the other models is needed.
+
+As to prediction times the EMOS is by far the fastest due to its simple linear structure.
+The boosted GAMLSS is the second fastest followed by the prespecified GAMLSS and the distributional forest on the 4th place.
+This order holds for all stations and also for both subsample settings.
+However, while all four models are comparably fast making one single prediction the forest needs about ten times more time to make predictions for the whole learning data set.
+This is due to the maximum-likelihood estimation based on observation-specific weights matrices which have to be calculated separately for each observation.
+
+Concerning variable importance three covariates based on total precipitation (\code{tp\_min}, \code{tp\_max}, \code{tp\_mean}) are at the top of the list for almost all stations (except for the two stations Matrei in Osttirol and Ochsengarten-Obergut where a covariate describing total column-integrated condensate is on the third place).
+This was to be expected since the aim is to predict amounts of precipitation.
+The other members of the top ten variables vary slightly over stations, however, further covariates concerning total precipitation, total column-integrated condensate and temperature differences are often among them.
+
 \section{Discussion}
+Overall, the results presented in this supplement show that Axams is an observation station without any unusual features that would lead to outstanding performances of any of the methods.
+Therefore, it is a reasonable choice as a representative for all 95 station.
+Examples for stations where the forest model performs better or worse can be found easily, however, in general the methods perform similarly as for station Axams.
+Also at most observation stations in East Tyrol where the prespecified GAMLSS performed clearly better in the ordered subsample setting the cross-validation confirms that the forest model performs at almost equally well over varying subsample selections.
+
+As to fit times, all models achieve similar results, however, the cross-validation applied to find the optimal stopping value \code{mstop} for the boosted GAMLSS is extremely time consuming. 
 Looking at prediction times for one single observation, the distributional forest is on average
-the slowest model, however, the difference is not too big. This changes regarding predictions
-for the a set of observation due to the fact that the forest makes predictions based
-on local neighbour weights. Therefor, for each observation its own weight matrix has to be included in the prediction process making it a more time consuming procedure.
+the slowest model, however, the difference to the predict times of the other models is not too big.
+This changes regarding predictions for a whole set of observation due to the fact that the forest computes each single prediction separately as it is based on an observation-specific
+weight matrix.
+
+As mentioned by \cite{Hofner+Mayr+Schmid:2016} the methods provided by GAMLSS ``can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters''.
+These problems have also occured in a few situations for the
+prespecified GAMLSS model within the cross-validation. In case of this method failing to 
+build a model it is represented by \code{NA}s in the evaluation.
 
@@
 
+\includepdf[pages=-, link, linkname=sumAxams]{Axams.pdf}
+\includepdf[pages=-, link, linkname=sumLech]{Lech.pdf}
+\includepdf[pages=-, link, linkname=sumZuers]{Zuers.pdf}
+\includepdf[pages=-, link, linkname=sumSeeimPaznaun]{SeeimPaznaun.pdf}
+\includepdf[pages=-, link, linkname=sumOetz]{Oetz.pdf}
+\includepdf[pages=-, link, linkname=sumJungholz]{Jungholz.pdf}
+\includepdf[pages=-, link, linkname=sumKoessen]{Koessen.pdf}
+\includepdf[pages=-, link, linkname=sumWalchsee]{Walchsee.pdf}
+\includepdf[pages=-, link, linkname=sumGinzling]{Ginzling.pdf}
+\includepdf[pages=-, link, linkname=sumRotholz]{Rotholz.pdf}
+\includepdf[pages=-, link, linkname=sumInnervillgraten]{Innervillgraten.pdf}
+\includepdf[pages=-, link, linkname=sumMatreiinOsttirol]{MatreiinOsttirol.pdf}
+\includepdf[pages=-, link, linkname=sumStJohannimWalde]{StJohannimWalde.pdf}
+\includepdf[pages=-, link, linkname=sumLadisNeuegg]{LadisNeuegg.pdf}
+\includepdf[pages=-, link, linkname=sumOchsengartenObergut]{OchsengartenObergut.pdf}
+
+%bibliographystyle{jss}
+\bibliography{ref_supplement.bib}
+
 \end{document}
+
+%% https://cran.r-project.org/web/packages/gamboostLSS/vignettes/gamboostLSS_Tutorial.pdf
+%% As a consequence, gamboostLSS is a convenient alternative to the AIC-based variable selection methods implemented in gamlss.
+The latter methods can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters.

### 2018-09-15 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
revised versions of supplement 1 and 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2283 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 13a7f6f3c52fe3004fd87673ab56c156c7cb84e3

@@
 
-Alltogether it can be stated that the left-censored at zero Gaussian distribution seems to be an appropriate choice while the other two distributions are possible and mostly also reasonable alternatives.
-However, they do not lead to an overall improvement in performance for this application.
-
 \begin{figure}[h]
@@
 \section{Discussion}
-As mentioned by \cite{Hofner+Mayr+Schmid:2016} the methods provided by GAMLSS ``can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters''.
-These problems have also occured in a few situations for the
-prespecified GAMLSS model within the cross-validation. In case of this method failing to 
-build a model it is represented by \code{NA}s in the evaluation.
+Alltogether it can be stated that the left-censored at zero Gaussian distribution seems to be an appropriate choice while the other two distributions are possible and mostly also reasonable alternatives.
+However, they do not lead to an overall improvement in performance for this application.
 

### 2018-09-21 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
revised versions of the two supplements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2288 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4fd840560f1fff3def8ab5426de9972e78214d72

@@
-%\documentclass[a4paper,11pt]{article}
 \documentclass[nojss, shortnames]{jss}
@@
 Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
-the underlying distribution is assumed to a be left-censored at zero Gaussian distribution.
-In this suppplement the exact same evaluations as in the paper are performed employing
-other distribution assumption. In particular, a left-censored at zero logistic distribution 
-and a hurdle-model are applied and their results are compared to those presented in the paper.}
+the underlying distribution is assumed to be a left-censored at zero Gaussian distribution.
+In this supplement the exact same evaluations as in the paper are performed employing
+other distribution assumptions. In particular, a left-censored at zero logistic distribution 
+and a hurdle-model are applied and their results are compared to those of the left-censored 
+at zero Gaussian distribution presented in the paper.}
 
-\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
+\Keywords{random forests, GAMLSS, hurdle model, logistic, Gaussian}
 
@@
 URL: \url{http://user.math.uzh.ch/hothorn/}\\
-
 }
@@
 \section{Introduction}
-This application study is supplementary material to the paper ''Distributional 
-Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''. 
-In Section~3 the presented distributional forest model is employed to obtain 
+This application study is supplementary material to the application presented
+in Section~3 of the paper.
+There, the presented distributional forest model is employed to obtain 
 probabilistic precipitation forecasts in a mountainous region assuming that the 
@@
 
-To compare the performance of the distributional forest to other commonly used methods the exact same models as in the paper are evaluated on the same data set and applying the same settings while only the distribution is exchanged.
-Further details on the models can be found in Section~3.2..
-In particular, Table~2 gives an overview over the models and their specifications.
-As a brief reminder the approaches are summarized below.
+To compare the performance of the distributional forest to other commonly used
+methods the exact same models as in the paper are evaluated on the same data set 
+and applying the same settings while only the distribution is exchanged. Further 
+details on the models can be found in Section~3.2. of the paper. In particular, 
+Table~2 gives an overview over the models and their specifications. As a brief 
+reminder the approaches are summarized below.
 
@@
 
-\item \emph{Distributional forest:} All predictor variables are
-considered for learning a forest where bootstrap
-sampling is employed for each tree. Parameters are estimated by
-adaptive local likelihood based on forest weights.
+\item \emph{Distributional forest:} All predictor variables are considered for 
+learning a forest where subsampling is employed for each tree. Parameters are 
+estimated by adaptive local likelihood based on forest weights.
 
@@
 function $F$ are defined by
+%
 \begin{align}
-f_{\textit{Gaussian}}(Y; \mu, \sigma) &= 
+f_{\mathrm{Gaussian}}(Y; \mu, \sigma) &= 
 \frac{1}{\sigma} \cdot \phi\left(\frac{Y-\mu}{\sigma}\right) \\
-F_{\textit{Gaussian}}(Y; \mu, \sigma) &=
+F_{\mathrm{Gaussian}}(Y; \mu, \sigma) &=
 \Phi\left(\frac{Y-\mu}{\sigma}\right)
@@
 function and distribution function
-
+%
 \begin{align}
-f_{\textit{logistic}}(Y; \mu, \sigma) &= 
+f_{\mathrm{logistic}}(Y; \mu, \sigma) &= 
 \frac{\exp{\left(-\frac{Y - \mu}{\sigma}\right)}}
 {\sigma \cdot \left(1 + \exp{\left(-\frac{Y - \mu}{\sigma}\right)}\right)^2} \\
-F_{\textit{logistic}}(Y; \mu, \sigma) &=
+F_{\mathrm{logistic}}(Y; \mu, \sigma) &=
 \frac{1}{1+ \exp{\left(-\frac{Y - \mu}{\sigma}\right)}}.
 \end{align}
-
+%
 Both distributions have to be adapted in our application to capture that
@@
 or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
-a truncated (postive) part \citep[see][for an overview]{Winkelmann:2006}.
+a truncated (positive) part \citep[see][for an overview]{Winkelmann:2006}.
 \begin{itemize}
@@
 censored distribution is
+%
 \begin{equation}
-\ell_{\textit{cens}}(\mu, \sigma; Y) = 
+\ell_{\mathrm{cens}}(\mu, \sigma; Y) = 
 \begin{cases}
@@
 \end{equation}
+%
 as seen in Equation~2.1.1. of the paper for a Gaussian distribution.
@@
 any precipitation at all, i.e. $\nu = \mathbb{P}(Y>0)$, is included in the model.
-%Then, a second model is fit to the subset of the data containing 
-%only positive observations applying a left-truncated at zero 
-%distribution to estimate the location parameter $\mu$ and scale 
-%parameter $\sigma$ of the underlying latent variable.
 The three-parametric log-likelihood function for the full hurdle model 
@@
 \begin{equation} \label{eq:hurdle}
-\ell_{\textit{hurdle}}(\mu, \sigma, \nu; Y) = 
+\ell_{\mathrm{hurdle}}(\mu, \sigma, \nu; Y) = 
 \begin{cases}
@@
 For $Y=0$ the probability that it won't rain is expressed by $1-\nu$. 
-For $Y>0$ the probability that it will rain $\nu$ is multiplied with the left-truncated at zero density function.
-Contrary to the censored model, here it is only the parameter $\nu$ controlling the point mass while $\mu$ and $\sigma$ only influence the distribution of the positive observations.
+For $Y>0$ the probability that it will rain $\nu$ is multiplied with the 
+left-truncated at zero density function. Contrary to the censored model, 
+here it is only the parameter $\nu$ controlling the point mass while $\mu$ 
+and $\sigma$ only influence the distribution of the positive observations.
+
 \end{itemize}
@@
 
-In the log-likelihood function in \ref{eq:hurdle} properties of the
+In the log-likelihood function in Equation~\ref{eq:hurdle} properties of the
 logarithmic function can be applied to split the term for $Y>0$ 
@@
 from the additional parameter $\nu$. Therefore, $\nu$ can be modeled 
-by a separate model in the GAMLSS framework. This approach is also
+by a separate model in the GAMLSS framework. The same approach is
 chosen for the distributional forest model in this application study.
@@
 \section{Stations}
-Out of the 95 observation stations 15 are considered for this 
-benchmark study (see Figure~\ref{fig:map}). 
-They have been selected based on the region
-they are in, their altitude and the performance of the 
-different models in the application in the paper such that
-all of these features are represented equally.
+Out of the 95 observation stations 15 are considered for this benchmark 
+study (see Figure~\ref{fig:map}). 
+They have been selected based on the region they are in, their altitude 
+and the performance of the different models in the application in the 
+paper such that all of these features are represented equally.
+
+Selected stations: Axams, Lech, Zuers, See im Paznaun, Jungholz, Ladis-Neuegg, 
+Oetz, Ochsengarten-Obergut, Ginzling, Rotholz, Walchsee, Koessen, 
+Innervillgraten, Matrei in Osttirol, St.Johann im Walde.
 
@@
 points(sp, pch = 21, col = "darkgrey", las = 1, cex = 0.6)
-points(sp[c(5, 6, 20, 23, 25, 32, 33, 46, 47, 56, 57, 70, 79, 82, 83),], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
+points(sp[c(5, 6, 20, 23, 25, 32, 33, 46, 47, 56, 57, 70, 79, 82, 83),], 
+       pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
 # stationname beneath
@@
 \vspace*{-3.5cm}
-\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and the 15 considered stations highlighted in pink.}
+\caption{\label{fig:map}Map of Tyrol with all 95 observation stations and 
+the 15 considered stations highlighted in pink.}
 \end{figure}
@@
 In Figure~\ref{fig:lattice} each panel belongs to one of the 15 observation stations. 
-The three distributions (left-censored at zero Gaussian distribution = \textit{gaussian}, hurdle model based on the Gaussian distribution = \textit{hgaussian}, left-censored at zero logistic distribution = \textit{logistic}) are compared based on the resulting CRPS values for each of the applied methods that are represented by different colours and symbols.
-Overall it can be stated that the forest model performs almost equally well with all three distributions.
+The three distributions \textit{gaussian} (left-censored at zero Gaussian distribution), 
+\textit{hgaussian} (hurdle model based on the Gaussian distribution) and 
+\textit{logistic} (left-censored at zero logistic distribution) are compared based 
+on the resulting CRPS values for each of the four applied methods that are represented 
+by different colors and symbols. 
+Overall, the forest model performs almost equally well for all three distributions. 
 Therefore it seems to be very robust against missspecifications of the distribution.
-For the two GAMLSS models the censored Gaussian and the hurdle model perform similarly at most of the station while applying the censored logistic distribution leads to an increase in CRPS, i.e.
-to a decline in performance, especially for the prespecified GAMLSS.
-For the EMOS model exchanging the distribution does not have a strong effect on its performance, similar to the forest model.
-In particular, for the two censored distributions the results are almost equal, only for the hurdle model the CRPS values vary slightly, however in both directions, depending on the station.
+For the two GAMLSS models the censored Gaussian and the hurdle model perform similarly 
+at most of the stations while applying the censored logistic distribution leads to an 
+increase in CRPS, i.e. to a decline in performance, especially for the prespecified 
+GAMLSS (for example at stations Zuers and Jungholz).
+For the EMOS model exchanging the distribution does not have a strong effect on its 
+performance, similar to the forest model.
+All in all, exchanging the underlying distribution does not lead to clear 
+improvements of any of the methods, especially the censored Gaussian and the 
+hurdle model achieve almost the same results. However, the robustness against 
+missspecification of the distribution varies slightly over stations. For example 
+at station Matrei in Osttirol all methods seem to be very stable while stronger 
+influences can be seen at station Jungholz.
 
-When applying a hurdle model for the stations Oetz and Innvervillgraten the prespecified GAMLSS computationally failed to build/fit a model.
-Therefore this method is not represented in Figure~\ref{fig:lattice} for these two cases.
+\textit{Note:} When applying a hurdle model for the stations Oetz and 
+Innvervillgraten the prespecified GAMLSS computationally failed to build/fit 
+a model. Therefore this method is not represented in the corresponding panels 
+of Figure~\ref{fig:lattice} for these two cases.
 
-In stations like Ladis-Neuegg, Matrei in Osttirol or Oetz all methods seem to be very robust and stable over varying distributions.
-The strongest influence of an exchange of the distribution can be seen for example at stations Jungholz and St.Johann im Walde for the prespecified GAMLSS.
+In Figure~\ref{fig:box_crpss_gaussian} the tested distributions are compared 
+for each method based on CRPS skill score with the censored Gaussian distribution 
+as reference represented by the horizontal line at zero. Again, the forest model 
+and the EMOS show only slight variations (on average not even 1\%) over different 
+distributions while the two GAMLSS models perform even or slightly better applying 
+a hurdle model and loose strength when employing a logistic distribution compared 
+to a Gaussian distribution.
 
-In Figure\ref{fig:box_crpss_gaussian} the tested distributions are compared for each method based on CRPS skill score with the censored Gaussian distribution as reference represented by the horizontal line at zero.
-The hurdle model leads to a slight improvement for the forest model and the prespecified GAMLSS.
-Following the median this improvement is a bit stronger for the boosted GAMLSS, however, with a higher deviation as represented by the larger size of the box, similarly as for the EMOS, however with a negative influence indicated by the negative median.
-Choosing a logistic distribution instead of a Gaussian clearly reduces the strength of the two GAMLSS models, while it has almost no influence on the forest and the EMOS models.
-
-\begin{figure}[h]
+\begin{figure}
 \setkeys{Gin}{width=0.99\textwidth}
@@
 load("Rain_distributions.rda")
-means <- results$means[results$means$method != "disttree",]
+stations <- c("Axams", "Lech", "Zuers", "See im Paznaun", "Jungholz", 
+              "Ladis-Neuegg", "Oetz", "Ochsengarten-Obergut",
+              "Ginzling", "Rotholz", "Walchsee", "Koessen", 
+              "Innervillgraten", "Matrei in Osttirol", 
+              "St.Johann im Walde")
+
+means <- results$means[(results$means$method != "disttree") & 
+                         results$means$station %in% stations, ]
+rownames(means) <- c(1:NROW(means))
+means$station <- factor(means$station, levels = stations)
 means$method <- factor(means$method, 
@@
        data = means, auto.key = TRUE, 
-       type = "o", lwd = 2, lty = 1, layout = c(3,6))
+       type = "o", lwd = 2, lty = 1,  
+       drop.unused.levels = TRUE, layout = c(3,5),
+       as.table = TRUE, ylab = "CRPS", xlab = "Distribution")
 @
-\caption{\label{fig:lattice}CRPS values of all 4 methods, once applied with each of the three distributions for each of the 15 observation stations.}
+\caption{\label{fig:lattice}CRPS values of all four methods, once applied with 
+each of the three distributions for each of the 15 observation stations.}
 \end{figure}
 
-\begin{figure}[h]
+\begin{figure}
 \setkeys{Gin}{width=0.99\textwidth}
@@
 means_sel <- means[means$distribution %in% c("logistic", "hgaussian"),]
-  means_sel$distribution <- factor(means_sel$distribution, levels(means_sel$distribution)[c(2:3)])
+means_sel$distribution <- factor(means_sel$distribution, 
+                                 levels(means_sel$distribution)[c(2:3)])
 
-  par(mfrow = c(1,4))
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "Distributional forest"), main = "Distributional forest",
-          ylim = c(-0.12,0.07), las = 2, ylab = "CRPS skill score")
+# CRPS skill score for distributional forest
+s_df <- matrix(ncol = 2, nrow = 15)
+s_df[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
+s_df[,2] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "logistic", "crps_ss_dist"]
+
+# CRPS skill score for prespecified GAMLSS
+s_g <- matrix(ncol = 2, nrow = 15)
+s_g[,1] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
+s_g[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "logistic", "crps_ss_dist"]
+
+# CRPS skill score for boosted GAMLSS
+s_gb <- matrix(ncol = 2, nrow = 15)
+s_gb[,1] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
+s_gb[,2] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "logistic", "crps_ss_dist"]
+
+# CRPS skill score for EMOS
+s_emos <- matrix(ncol = 2, nrow = 15)
+s_emos[,1] <- means_sel[means_sel$method == "EMOS" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
+s_emos[,2] <- means_sel[means_sel$method == "EMOS" & means_sel$distribution == "logistic", "crps_ss_dist"]
+
+colnames(s_df) <- colnames(s_g) <- colnames(s_gb) <- colnames(s_emos) <- c("hgaussian", "logistic")
+
+# boxplots with matplot
+par(mfrow = c(1,4), mar = c(4,4,3,0.7))
+matplot(t(s_df[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "Distributional forest",
+        xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 2.5),
+        ylim = c(-0.10, 0.05))
+boxplot(s_df, add = TRUE, col = "transparent")
 abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "Prespecified GAMLSS"), main = "Prespecified GAMLSS",
-          ylim = c(-0.12,0.07), las = 2)
+#boxplot(crps_ss_dist~distribution, data = means_sel, 
+#        subset = (method == "Distributional forest"), main = "Distributional forest",
+#        ylim = c(-0.12,0.07), las = 2, ylab = "CRPS skill score")
+#abline(h = 0, col = pal[5], lwd = 2)
+par(mar = c(4,2.9,3,1.8))
+matplot(t(s_g[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE, main = "Prespecified GAMLSS",
+        xlab = "", ylab = "", xlim = c(0.5, 2.5),
+        ylim = c(-0.10, 0.05))
+boxplot(s_g, add = TRUE, col = "transparent", yaxt = 'n')
 abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "Boosted GAMLSS"), main = "Boosted GAMLSS",
-          ylim = c(-0.12,0.07), las = 2)
+#boxplot(crps_ss_dist~distribution, data = means_sel, 
+#        subset = (method == "Prespecified GAMLSS"), main = "Prespecified GAMLSS",
+#        ylim = c(-0.12,0.07), las = 2)
+#abline(h = 0, col = pal[5], lwd = 2)
+par(mar = c(4,1.8,3,2.9))
+matplot(t(s_gb[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE, main = "Boosted GAMLSS",
+        xlab = "", ylab = "", xlim = c(0.5, 2.5),
+        ylim = c(-0.10, 0.05))
+boxplot(s_gb, add = TRUE, col = "transparent", yaxt = 'n')
 abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_dist~distribution, data = means_sel, 
-          subset = (method == "EMOS"), main = "EMOS",
-          ylim = c(-0.12,0.07), las = 2)
+#boxplot(crps_ss_dist~distribution, data = means_sel, 
+#        subset = (method == "Boosted GAMLSS"), main = "Boosted GAMLSS",
+#        ylim = c(-0.12,0.07), las = 2)
+#abline(h = 0, col = pal[5], lwd = 2)
+par(mar = c(4,0.7,3,4))
+matplot(t(s_emos[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "EMOS",
+        xlab = "", ylab = "", xlim = c(0.5, 2.5),
+        ylim = c(-0.10, 0.05))
+boxplot(s_emos, add = TRUE, col = "transparent", yaxt = 'n')
 abline(h = 0, col = pal[5], lwd = 2)
+#boxplot(crps_ss_dist~distribution, data = means_sel, 
+#        subset = (method == "EMOS"), main = "EMOS",
+#        ylim = c(-0.12,0.07), las = 2)
+#abline(h = 0, col = pal[5], lwd = 2)
 @
-\caption{\label{fig:box_crpss_gaussian}(Mean) CRPS skill score values of each of the 15 observation stations, separated by method and distribution. Gaussian is chosen as reference distribution.}
+\caption{\label{fig:box_crpss_gaussian} CRPS skill score values of each of 
+the 15 observation stations, separated by method and distribution. Left-censored 
+at zero Gaussian is chosen as reference distribution.}
 \end{figure}
 
-\begin{figure}[h]
+Figures~\ref{fig:box_crpss_EMOS} and \ref{fig:box_crps} compare the methods separately 
+for each distribution, once based on CRPS skill score values with reference model EMOS 
+and once based on CRPS values. Both confirm the conclusion that the censored Gaussian 
+model and the hurdle model lead to almost the same results for all methods while only 
+the logistic model shows small variations.
+
+Altogether it can be stated that the left-censored at zero Gaussian distribution seems 
+to be an appropriate choice while the other two distributions, especially the hurdle model, 
+are also reasonable alternatives. However, they do not lead to an overall improvement in 
+performance for this application.
+
+\begin{figure}
 \setkeys{Gin}{width=0.99\textwidth}
@@
 
-  par(mfrow = c(1,3))
-  boxplot(crps_ss_method~method, data = means_sel, 
-          subset = (distribution == "gaussian"), main = "gaussian",
-          ylim = c(-0.12,0.23), axes = FALSE, ylab = "CRPS skill score")
-  axis(2, seq(-0.15, 0.25, 0.05), c(seq(-0.15, 0, 0.05), seq(0.05, 0.25, 0.05)), las = 2)
+# CRPS skill score for gaussian
+s_gaussian <- matrix(ncol = 3, nrow = 15)
+s_gaussian[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "gaussian", "crps_ss_method"]
+s_gaussian[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "gaussian", "crps_ss_method"]
+s_gaussian[,3] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "gaussian", "crps_ss_method"]
+
+# CRPS skill score for boosted hgaussian
+s_hgaussian <- matrix(ncol = 3, nrow = 15)
+s_hgaussian[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "hgaussian", "crps_ss_method"]
+s_hgaussian[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "hgaussian", "crps_ss_method"]
+s_hgaussian[,3] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "hgaussian", "crps_ss_method"]
+
+# CRPS skill score for prespecified logistic
+s_logistic <- matrix(ncol = 3, nrow = 15)
+s_logistic[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "logistic", "crps_ss_method"]
+s_logistic[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "logistic", "crps_ss_method"]
+s_logistic[,3] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "logistic", "crps_ss_method"]
+
+colnames(s_gaussian) <- colnames(s_hgaussian) <- colnames(s_logistic) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
+
+par(mfrow = c(1,3), mar = c(5,4,3,0))
+matplot(t(s_gaussian[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "gaussian", ylim = c(-0.12,0.23), 
+        xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
+boxplot(s_gaussian, add = TRUE, col = "transparent", axes = FALSE)
+abline(h = 0, col = pal[5], lwd = 2)
+#boxplot(crps_ss_method~method, data = means_sel, 
+#        subset = (distribution == "gaussian"), main = "gaussian",
+#        ylim = c(-0.12,0.23), axes = FALSE, ylab = "CRPS skill score")
+axis(1, 0:4, c("",
+               "Distr.
+                 forest",
+               "Presp.
+                 GAMLSS",
+               "Boosted
+                 GAMLSS", ""), las=2)
+axis(2, seq(-0.15, 0.25, 0.05), c(seq(-0.15, 0, 0.05), seq(0.05, 0.25, 0.05)), las = 0)
+axis(3, 0:4, lwd.ticks = 0, labels = FALSE)
+axis(4, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
+
+par(mar = c(5,2,3,2))
+matplot(t(s_hgaussian[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "hgaussian", ylim = c(-0.12,0.23), 
+        xlab = "", ylab = "", xlim = c(0.5, 3.5))
+boxplot(s_hgaussian, add = TRUE, col = "transparent", axes = FALSE)
+abline(h = 0, col = pal[5], lwd = 2)
+#boxplot(crps_ss_method~method, data = means_sel, 
+#        subset = (distribution == "hgaussian"), main = "hgaussian",
+#        ylim = c(-0.12,0.23), axes = FALSE)
+axis(2, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
 axis(1, 0:4, c("",
@@
 axis(4, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
+
+par(mar = c(5,0,3,4))
+matplot(t(s_logistic[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "logistic", ylim = c(-0.12,0.23), 
+        xlab = "", ylab = "", xlim = c(0.5, 3.5))
+boxplot(s_logistic, add = TRUE, col = "transparent", axes = FALSE)
 abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_method~method, data = means_sel, 
-          subset = (distribution == "logistic"), main = "logistic",
-          ylim = c(-0.12,0.23), axes = FALSE)
-  axis(2, seq(-0.15, 0.25, 0.05), c(seq(-0.15, 0, 0.05), seq(0.05, 0.25, 0.05)), las = 2)
+#boxplot(crps_ss_method~method, data = means_sel, 
+#        subset = (distribution == "logistic"), main = "logistic",
+#        ylim = c(-0.12,0.23), axes = FALSE)
+axis(2, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
 axis(1, 0:4, c("",
@@
 axis(4, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
-  abline(h = 0, col = pal[5], lwd = 2)
-  boxplot(crps_ss_method~method, data = means_sel, 
-          subset = (distribution == "hgaussian"), main = "hgaussian",
-          ylim = c(-0.12,0.23), axes = FALSE)
-  axis(2, seq(-0.15, 0.25, 0.05), c(seq(-0.15, 0, 0.05), seq(0.05, 0.25, 0.05)), las = 2)
-  axis(1, 0:4, c("",
-                 "Distr.
-                 forest",
-                 "Presp.
-                 GAMLSS",
-                 "Boosted
-                 GAMLSS", ""), las=2)
-  axis(3, 0:4, lwd.ticks = 0, labels = FALSE)
-  axis(4, seq(-0.15, 0.25, 0.05), lwd.ticks = 0, labels = FALSE)
-  abline(h = 0, col = pal[5], lwd = 2)
+
 @
-\caption{\label{fig:box_crpss_EMOS}(Mean) CRPS skill score values of each of the 15 observation stations, separated by method and distribution. EMOS is chosen as reference model.}
+\caption{\label{fig:box_crpss_EMOS}CRPS skill score values of each of 
+the 15 observation stations, separated by method and distribution. 
+EMOS is chosen as reference model.}
 \end{figure}
 
-\begin{figure}[h]
+\begin{figure}
 \setkeys{Gin}{width=0.99\textwidth}
@@
 
-  # crps
-  par(mfrow = c(1,3))
-  boxplot(crps~method, data = means, subset = distribution == "gaussian", main = "gaussian",
-          ylim = c(0.65,1.15), axes = FALSE, ylab = "CRPS")
-  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
+# CRPS for gaussian
+c_gaussian <- matrix(ncol = 4, nrow = 15)
+c_gaussian[,1] <- means[means$method == "Distributional forest" & means$distribution == "gaussian", "crps"]
+c_gaussian[,2] <- means[means$method == "Prespecified GAMLSS" & means$distribution == "gaussian", "crps"]
+c_gaussian[,3] <- means[means$method == "Boosted GAMLSS" & means$distribution == "gaussian", "crps"]
+c_gaussian[,4] <- means[means$method == "EMOS" & means$distribution == "gaussian", "crps"]
+
+# CRPS skill score for boosted hgaussian
+c_hgaussian <- matrix(ncol = 4, nrow = 15)
+c_hgaussian[,1] <- means[means$method == "Distributional forest" & means$distribution == "hgaussian", "crps"]
+c_hgaussian[,2] <- means[means$method == "Prespecified GAMLSS" & means$distribution == "hgaussian", "crps"]
+c_hgaussian[,3] <- means[means$method == "Boosted GAMLSS" & means$distribution == "hgaussian", "crps"]
+c_hgaussian[,4] <- means[means$method == "EMOS" & means$distribution == "hgaussian", "crps"]
+
+# CRPS skill score for prespecified logistic
+c_logistic <- matrix(ncol = 4, nrow = 15)
+c_logistic[,1] <- means[means$method == "Distributional forest" & means$distribution == "logistic", "crps"]
+c_logistic[,2] <- means[means$method == "Prespecified GAMLSS" & means$distribution == "logistic", "crps"]
+c_logistic[,3] <- means[means$method == "Boosted GAMLSS" & means$distribution == "logistic", "crps"]
+c_logistic[,4] <- means[means$method == "EMOS" & means$distribution == "logistic", "crps"]
+
+colnames(c_gaussian) <- colnames(c_hgaussian) <- colnames(c_logistic) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS")
+
+par(mfrow = c(1,3), mar = c(5,4,3,0))
+matplot(t(c_gaussian[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "gaussian", 
+        xlab = "", ylab = "CRPS", xlim = c(0.5, 4.5),
+        ylim = c(0.65,1.15))
+boxplot(c_gaussian, add = TRUE, col = "transparent", axes = FALSE)
+#boxplot(crps~method, data = means, subset = distribution == "gaussian", main = "gaussian",
+#        ylim = c(0.65,1.15), axes = FALSE, ylab = "CRPS")
+axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 0)
 axis(1, 0:5, c("",
@@
 
-  boxplot(crps~method, data = means, subset = distribution == "hgaussian", main = "hgaussian",
-          ylim = c(0.65,1.15), las = 2, axes = FALSE)
-  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
+par(mar = c(5,2,3,2))
+matplot(t(c_hgaussian[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "hgaussian", 
+        xlab = "", ylab = "", xlim = c(0.5, 4.5),
+        ylim = c(0.65,1.15))
+boxplot(c_hgaussian, add = TRUE, col = "transparent", axes = FALSE)
+#boxplot(crps~method, data = means, subset = distribution == "hgaussian", main = "hgaussian",
+#        ylim = c(0.65,1.15), las = 2, axes = FALSE)
+axis(2, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
 axis(1, 0:5, c("",
@@
 axis(4, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
-  boxplot(crps~method, data = means, subset = distribution == "logistic", main = "logistic",
-          ylim = c(0.65,1.15), las = 2, axes = FALSE)
-  axis(2, seq(0.6, 1.2, 0.1), seq(0.6, 1.2, 0.1), las = 2)
+
+par(mar = c(5,0,3,4))
+matplot(t(c_logistic[,]), type = "l", lwd = 2, 
+        col = gray(0.5, alpha = 0.2),
+        lty = 1, axes = FALSE,  main = "logistic", 
+        xlab = "", ylab = "CRPS", xlim = c(0.5, 4.5),
+        ylim = c(0.65,1.15))
+boxplot(c_logistic, add = TRUE, col = "transparent", axes = FALSE)
+#boxplot(crps~method, data = means, subset = distribution == "logistic", main = "logistic",
+#        ylim = c(0.65,1.15), las = 2, axes = FALSE)
+axis(2, seq(0.6, 1.2, 0.1), lwd.ticks = 0, labels = FALSE)
 axis(1, 0:5, c("",
@@
 @
-\caption{\label{fig:box_crps}(Mean) CRPS values of each of the 15 observation stations, separated by method and distribution.}
+\caption{\label{fig:box_crps}CRPS values of each of the 15 observation stations, 
+separated by method and distribution.}
 \end{figure}
 
-\section{Discussion}
-Alltogether it can be stated that the left-censored at zero Gaussian distribution seems to be an appropriate choice while the other two distributions are possible and mostly also reasonable alternatives.
-However, they do not lead to an overall improvement in performance for this application.
-
-As explained above the the parameter $\nu$ is modeled separately from $\mu$ and $\sigma$ 
+As explained above, the parameter $\nu$ is modeled separately from $\mu$ and $\sigma$ 
 in the GAMLSS framework which is why two separate models can be built. For the forest model
 an alternative approach would be to build just one model employing the three-parametric
-distribution family. However, testing both versions of the forest model in this application study shows that they both perform almost equally. Figure~\ref{fig:box_crpss_forests} shows
+distribution family. However, testing both versions of the forest model in this application 
+study shows that they both perform almost equally. Figure~\ref{fig:box_crpss_forests} shows
 the CRPS skill score of the three-parametric forest model over the considered 15 observation
@@
 of the forest model seems to perform better on average, according to the median this 
-improvement is not even one percent, hence neglectably small. Therefore, the two foreset version can be stated to perform almost equally.
+improvement is not even one percent, hence neglectably small. Therefore, the two forest 
+versions can be stated to perform almost equally well.
 
-\begin{figure}[h]
+\begin{figure}
 \center
@@
 axis(1, 0:2, c("","Three-parametric forest", ""), las=1)
-  #sum(crps_forests[,"forest_2p"] > crps_forests[,"forest_h"])
-  #summary(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"])
-  #summary(abs(crps_forests[,"forest_h"] - crps_forests[,"forest_2p"]))
-  #colMeans(crps_forests)  
 @
-\caption{\label{fig:box_crpss_forests}(Mean) CRPS skill score values of each of the 15 observation stations for the three-parametric forest model with the two part forest model as reference model.}
+\caption{\label{fig:box_crpss_forests}CRPS skill score values of each of 
+the 15 observation stations for the three-parametric forest model with 
+the two-part forest model as reference model.}
 \end{figure}
 
+\newpage
 %bibliographystyle{jss}

### 2018-09-21 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
revised versions of the two supplements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2288 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4fd840560f1fff3def8ab5426de9972e78214d72

@@
 \usepackage{Sweave}
-\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,pdfpages,hyperref}
+\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,pdfpages,hyperref,bookmark}
 
@@
 
-\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
+\Keywords{random forests, GAMLSS, evaluation, meteorological stations}
 
@@
 \section{Introduction}
-This application study is supplementary material to the paper ''Distributional 
-Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''. 
-In Section~3 the presented distributional forest model is employed to obtain 
+This application study is supplementary material to the application in Section~3 of 
+the paper where the presented distributional forest model is employed to obtain 
 probabilistic precipitation forecasts in a mountainous region.
@@
 Gaussian distribution is evaluated together with three other zero-censored 
-Gaussian models. Further details on all four models can be found in Section~3.2.. 
-In particular, Table~2 gives an overview over the models and their specifications. 
-As a brief reminder the approaches are summarized below.
-
+Gaussian models. Further details on all four models can be found in Section~3.2. of
+the paper. In particular, Table~2 gives an overview over the models and their 
+specifications. As a brief reminder the approaches are summarized below.
+%
 \begin{itemize}
@@
 \item \emph{Distributional forest:} All predictor variables are
-considered for learning a forest where bootstrap
-sampling is employed for each tree. Parameters are estimated by
+considered for learning a forest where subsampling is employed 
+for each tree. Parameters are estimated by
 adaptive local likelihood based on forest weights.
@@
 %\subsection{Subset selection}
-The models are evaluated in two different scenarios.
+The models are always learned on data from 24 years and tested on the 
+remaining 4 years. Splitting the data into these two subsets has been done
+in two different ways leading to the two following evaluation settings.
+%
 \begin{itemize}
-\item Ordered/Base model: The learning data consists of the first 24 years (1985--2008) and the testing data of the following 4 years (2009--2012). 
-\item Cross-validation: The 28 years are randomly split in 7 blocks of 4 years which are not necessarily successive.
+
+\item Single split: The data is time-ordered. The learning data consists of the first 24 years (1985--2008) and the testing data of the following 4 years (2009--2012).
+
+\item Cross-validation: In order to not just split the data at one splitpoint the 28 years are randomly split in 7 blocks of 4 years which are not necessarily successive.
 Each of the blocks is chosen once as testing data while the remaining 6 blocks are the learning data.
@@
 
-The performance of the models are compared based on the following evaluations.
+The performances of the models are compared based on numerical and graphical evaluations.
 
+\textit{Numerical:}
+For the single split setting one value is calculated for each observation in the test data and the mean of these values is considered.
+For the cross-validation this mean value over all observations in the test data is retrieved from all 70 evaluations and then again the mean of them is considered.
+This holds for the following measurements.
+%
 \begin{itemize}
-\item CRPS: The continuous ranked probability score (CRPS) is calculated
-for each observation in the testing data set and then averaged. For the cross-validation
-setup the mean of the resulting 70 CRPS average values is listed and all 70 average values
-are illustrated in a boxplot where the EMOS model is chosen as reference model, 
-represented by the horizontal line at 0.
 
-\item Fit time: It is measured how long it takes to fit the model to the learning data.
+\item CRPS: The continuous ranked probability score (CRPS) is calculated to assess the predictive performance in terms of full probabilistic forecasts.
 
-\item Predict time: It is measured how long it takes the model to make predictions.
-For the ordered model the predict time refers to the time needed to make one single prediction.
-The mean predict time over all observations in the test data is considered.
-For the cross-validation the predict time refers to the time needed to make all predictions for the test data. The mean predict time over all 70 evaluations in the cross-validation is considered.
+\item Time: For all models fitting times and predictions times are measured.
+In particular, \textit{fit time} refers to the time needed to fit the model to the whole learn data.
+Depending on the setting \textit{predict time} refers to the time needed to make one prediction in the single split setting and to make predictions for the whole test data in the cross-validation.
 
-\item Variable importance: For the forest model it is investigated which variables have the strongest influence on the model and its goodness of fit.
+\end{itemize}
+
+\textit{Graphical:} The following methods of visualization are applied to provide further means of comparison of all models but also to investigate on the distributional forest in particular.
+%
+\begin{itemize}
+%
+\item Boxplot of the CRPS skill score: With the EMOS as reference model the CRPS skill scores are calculated for the cross-validation setting and illustrated in a boxplot.
+%
+\item Barplot of variable importance: For the forest model it is investigated which variables have the strongest influence on the model and its goodness of fit.
 For each variable this is measured based on mean decrease in CRPS after permutation.
-The 10 variables showing the highest values are listed in a barplot.
-
-\item Residual QQ plots: For each of the models an out-of-sample residual QQ plot for the ordered model provides a visualization of goodness of fit.
+The 10 variables showing the highest values in the single split setting are listed in a barplot.
+%
+\item Residual QQ plots: For each of the models an out-of-sample residual QQ plot for the single split setting provides a visualization of goodness of fit.
 Randomized residual quantiles are compared to standard normal quantiles.
-
-\item PIT histograms: As an additional visualization out-of-sample probability integral transform histograms illustrated how well-calibrated the distributional fits are for the ordered model.
-
+%
+\item PIT histograms: As an additional visualization out-of-sample probability integral transform histograms illustrate how well-calibrated the distributional fits are for the single split setting.
+%
 \end{itemize}
@@
 
-Stations: \hyperlink{sumAxams.1}{Axams},
+Selected stations: \hyperlink{sumAxams.1}{Axams},
 \hyperlink{sumLech.1}{Lech},
@@
 \hyperlink{sumSeeimPaznaun.1}{See im Paznaun},
+\hyperlink{sumJungholz.1}{Jungholz},
+\hyperlink{sumLadisNeuegg.1}{Ladis-Neuegg},
 \hyperlink{sumOetz.1}{Oetz},
-\hyperlink{sumJungholz.1}{Jungholz},
-\hyperlink{sumKoessen.1}{Koessen},
-\hyperlink{sumWalchsee.1}{Walchsee},
+\hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut},
 \hyperlink{sumGinzling.1}{Ginzling},
 \hyperlink{sumRotholz.1}{Rotholz},
+\hyperlink{sumWalchsee.1}{Walchsee},
+\hyperlink{sumKoessen.1}{Koessen},
 \hyperlink{sumInnervillgraten.1}{Innervillgraten},
 \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol},
-\hyperlink{sumStJohannimWalde.1}{St.Johann im Walde},
-\hyperlink{sumLadisNeuegg.1}{Ladis-Neuegg},
-\hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut},
+\hyperlink{sumStJohannimWalde.1}{St.Johann im Walde}.
 
@@
 \section{Results}
-Comparing CRPS skill score values from the cross-validation the distributional forest
-and the two GAMLSS versions show clear improvements compared to the reference model EMOS.
-At many stations the forest and either the prespecified GAMLSS (e.g.
+
+\subsection{Forecasting skills} 
+Comparing CRPS values of the single split setting it can be stated that the forest model works well over all stations.
+Often it is even the best performing method (e.g.
+at stations \hyperlink{sumLech.1}{Lech}, \hyperlink{sumSeeimPaznaun.1}{See im Paznaun} and \hyperlink{sumGinzling.1}{Ginzling}), but there are also stations where one of the other models leads to better results.
+For example at \hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut} the boosted GAMLSS performs best, at \hyperlink{sumRotholz.1}{Rotholz} the EMOS model and at \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol} the prespecified GAMLSS is the best model, however, they do not clearly outperform the forest model.
+Moreover, these advantages often even out or turn into disadvantages when looking at the CRPS values from the cross-validation.
+For example at station \hyperlink{sumInnervillgraten.1}{Innervillgraten}, the prespecified GAMLSS is clearly the best in the single split setting, however, in the cross-validation the forest model is slightly ahead.
+An exception is given by station \hyperlink{sumSt.JohannimWalde.1}{St.Johann im Walde} as it is the only station where the prespecified GAMLSS clearly outperforms all other models in both settings.
+But overall, the results from the cross validation also support the statement that the forest model performs well on all stations being either the best or among the best performing methods at the majority of stations.
+Moreover, it can be concluded that Axams is an observation station without any unusual features that would lead to outstanding performances of any of the methods.
+Therefore, it is a reasonable choice as a representative for all 95 station.
+Examples for stations where the forest model performs better or worse can be found easily, however, in general the methods perform similarly as for station Axams.
+
+%Comparing CRPS skill score values from the cross-validation the distributional forest
+%and the two GAMLSS versions show clear improvements compared to the reference model EMOS.
+%At many stations the forest and either the prespecified GAMLSS (e.g.
 at station Innervillgraten) or the boosted GAMLSS (e.g.
@@
 But overall it can be stated that all three models perform well with only small variations over stations.
-Even if one of the evaluated models shows clear advantages in the ordered model (such as the EMOS at station Rotholz or the prespecified GAMLSS at station Innervillgraten) this is evened out within the cross-validation.
+Even if one of the evaluated models shows clear advantages in the single split setting (such as the EMOS at station Rotholz or the prespecified GAMLSS at station Innervillgraten) this is evened out within the cross-validation.
 
-Concerning fitting times either the prespecified GAMLSS (e.g.
-St.Johann im Walde, Innervillgraten) or the distributional forest (e.g.
-Zuers, Walchsee, Jungholz, Oetz, Koessen, Ginzling) are the fastest while the boosted GAMLSS usually takes more time, especially if the time required for the selection of the optimal value for \code{mstop} a multiple of the times of the other models is needed.
+\subsection{Computation time}
+The forest model and the two GAMLSS models achieve similar results in both time measurements, fit and predict time, while the EMOS model is clearly the fastest.
+However, for the boosted GAMLSS the cross-validation applied to find the optimal stopping value \code{mstop} is extremely time consuming.
+This additional time can not be neglected since the stopping value is crucial for the complexity of the model and the selection of covariates.
 
-As to prediction times the EMOS is by far the fastest due to its simple linear structure.
-The boosted GAMLSS is the second fastest followed by the prespecified GAMLSS and the distributional forest on the 4th place.
-This order holds for all stations and also for both subsample settings.
-However, while all four models are comparably fast making one single prediction the forest needs about ten times more time to make predictions for the whole learning data set.
-This is due to the maximum-likelihood estimation based on observation-specific weights matrices which have to be calculated separately for each observation.
+As to prediction times it has to be mentioned that the forest can keep up with the other models if only one prediction is made.
+However, to make predictions for a whole set of new observations the forest clearly needs more time.
+This is due to the fact that the forest has to computes each single prediction separately as the maximum-likelihood estimation is based on observation-specific weight matrix.
 
-Concerning variable importance three covariates based on total precipitation (\code{tp\_min}, \code{tp\_max}, \code{tp\_mean}) are at the top of the list for almost all stations (except for the two stations Matrei in Osttirol and Ochsengarten-Obergut where a covariate describing total column-integrated condensate is on the third place).
-This was to be expected since the aim is to predict amounts of precipitation.
+\subsection{Variable importance}
+Concerning variable importance, two covariates based on total precipitation (\code{tp\_max} and \code{tp\_mean}) are at the top of the list for all stations, often clearly ahead of the other covariates.
 The other members of the top ten variables vary slightly over stations, however, further covariates concerning total precipitation, total column-integrated condensate and temperature differences are often among them.
 
-\section{Discussion}
-Overall, the results presented in this supplement show that Axams is an observation station without any unusual features that would lead to outstanding performances of any of the methods.
-Therefore, it is a reasonable choice as a representative for all 95 station.
-Examples for stations where the forest model performs better or worse can be found easily, however, in general the methods perform similarly as for station Axams.
-Also at most observation stations in East Tyrol where the prespecified GAMLSS performed clearly better in the ordered subsample setting the cross-validation confirms that the forest model performs at almost equally well over varying subsample selections.
+\subsection{Calibration}
+The PIT histograms and residual QQ plots show that overall all four models are reasonably well calibrated at the majority of stations while at some stations they all fit better (e.g.
+at station \hyperlink{sumOetz.1}{Oetz}) or worse (e.g.
+at station \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol}) than on average.
+One feature of the boosting method that can be detected in the residual QQ plots for many stations is that heavy upper tails are not captured very well.
+This is also supported by the PIT histogram.
 
-As to fit times, all models achieve similar results, however, the cross-validation applied to find the optimal stopping value \code{mstop} for the boosted GAMLSS is extremely time consuming. 
-Looking at prediction times for one single observation, the distributional forest is on average
-the slowest model, however, the difference to the predict times of the other models is not too big.
-This changes regarding predictions for a whole set of observation due to the fact that the forest computes each single prediction separately as it is based on an observation-specific
-weight matrix.
-
-As mentioned by \cite{Hofner+Mayr+Schmid:2016} the methods provided by GAMLSS ``can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters''.
-These problems have also occured in a few situations for the
+\textit{Note:} As mentioned by \cite{Hofner+Mayr+Schmid:2016} the methods provided by GAMLSS ``can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters''.
+These problems have also occurred in a few situations for the
 prespecified GAMLSS model within the cross-validation. In case of this method failing to 
@@
 
-%\bibliography{ref_supplement_stationwise.bib}
-
 \includepdf[pages=-, link, linkname=sumAxams]{Axams.pdf}
+\bookmark[dest=sumAxams.1]{Axams}
 \includepdf[pages=-, link, linkname=sumLech]{Lech.pdf}
+\bookmark[dest=sumLech.1]{Lech}
 \includepdf[pages=-, link, linkname=sumZuers]{Zuers.pdf}
+\bookmark[dest=sumZuers.1]{Zuers}
 \includepdf[pages=-, link, linkname=sumSeeimPaznaun]{SeeimPaznaun.pdf}
+\bookmark[dest=sumSeeimPaznaun.1]{See im Paznaun}
+\includepdf[pages=-, link, linkname=sumJungholz]{Jungholz.pdf}
+\bookmark[dest=sumJungholz.1]{Jungholz}
+\includepdf[pages=-, link, linkname=sumLadisNeuegg]{LadisNeuegg.pdf}
+\bookmark[dest=sumLadisNeuegg.1]{Ladis-Neuegg}
 \includepdf[pages=-, link, linkname=sumOetz]{Oetz.pdf}
-\includepdf[pages=-, link, linkname=sumJungholz]{Jungholz.pdf}
+\bookmark[dest=sumOetz.1]{Oetz}
+\includepdf[pages=-, link, linkname=sumOchsengartenObergut]{OchsengartenObergut.pdf}
+\bookmark[dest=sumOchsengartenObergut.1]{Ochsengarten-Obergut}
+\includepdf[pages=-, link, linkname=sumGinzling]{Ginzling.pdf}
+\bookmark[dest=sumGinzling.1]{Ginzling}
+\includepdf[pages=-, link, linkname=sumRotholz]{Rotholz.pdf}
+\bookmark[dest=sumRotholz.1]{Rotholz}
+\includepdf[pages=-, link, linkname=sumWalchsee]{Walchsee.pdf}
+\bookmark[dest=sumWalchsee.1]{Walchsee}
 \includepdf[pages=-, link, linkname=sumKoessen]{Koessen.pdf}
-\includepdf[pages=-, link, linkname=sumWalchsee]{Walchsee.pdf}
-\includepdf[pages=-, link, linkname=sumGinzling]{Ginzling.pdf}
-\includepdf[pages=-, link, linkname=sumRotholz]{Rotholz.pdf}
+\bookmark[dest=sumKoessen.1]{Koessen}
 \includepdf[pages=-, link, linkname=sumInnervillgraten]{Innervillgraten.pdf}
+\bookmark[dest=sumInnervillgraten.1]{Innervillgraten}
 \includepdf[pages=-, link, linkname=sumMatreiinOsttirol]{MatreiinOsttirol.pdf}
+\bookmark[dest=sumMatreiinOsttirol.1]{Matrei in Osttirol}
 \includepdf[pages=-, link, linkname=sumStJohannimWalde]{StJohannimWalde.pdf}
-\includepdf[pages=-, link, linkname=sumLadisNeuegg]{LadisNeuegg.pdf}
-\includepdf[pages=-, link, linkname=sumOchsengartenObergut]{OchsengartenObergut.pdf}
+\bookmark[dest=sumStJohannimWalde.1]{St.Johann im Walde}
 

### 2018-09-22 pkg/disttree/inst/draft/supplement/Rain_stationwise_results.Rnw
revised versions of supplements with corresponding reproduction files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2289 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 443c37f89e7b5b3420a6e5a3862fb100989d9f2d

\documentclass[a4paper,11pt]{article}

%% layout
\usepackage[a4paper, total={15cm, 26.5cm}]{geometry}
\usepackage{graphicx}
\pagestyle{empty}
\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
\setlength{\parindent}{0em}
\usepackage{Sweave}

%% fonts
\renewcommand{\rmdefault}{phv}
\renewcommand{\sfdefault}{phv}

<<echo = FALSE, results=hide>>=
## station
station <- "_STATION_"
if(station == "SeeimPaznaun") station <- "See im Paznaun"
if(station == "MatreiinOsttirol") station <- "Matrei in Osttirol"
if(station == "LadisNeuegg") station <- "Ladis-Neuegg"
if(station == "StJohannimWalde") station <- "St.Johann im Walde"
if(station == "OchsengartenObergut") station <- "Ochsengarten-Obergut"

## results data
load("stationwise.rda")
results <- stationwise[[station]]

#station_rda <- function(name, root = getwd()) {
#  for(i in c("~", "-", ".", " ")) name <- gsub(i, "", name, fixed = TRUE)  
#  file.path(root, paste0("res_", name, ".rda"))
#}
#load(station_rda(station))

## (i) map
data("StationsTyrol", package = "RainTyrol")
data("MapTyrol", package = "RainTyrol")
library("sp")
sp <- SpatialPointsDataFrame(subset(StationsTyrol, select = c(lon, lat)),
  data = subset(StationsTyrol, select = -c(lon, lat)),
  proj4string = raster::crs(MapTyrol$RasterLayer))

## convenience functions to unify results
cross <- Reduce("c", results$res_cross)

get_cross <- function(what = "crps", which = NULL) {
  ## methods
  if(is.null(which)) which <- c("distforest", "gamlss", "gamboostLSS", "gamboostLSS_cvr", "emos_log")
  if(what != "evaltime") which <- which[which != "gamboostLSS_cvr"]

  ## extract desired quantity (crps, evaltime, predtime)
  rval <- lapply(cross, "[[", what)
  rval <- if(what == "crps") {
    sapply(rval, function(x) x[which])
  } else {
    sapply(rval, function(x) x[which, "elapsed"])
  }

  ## reshape
  rval <- t(rval)
  rval <- aggregate(rval, list(rep(1:10, each = 7)), mean, na.rm = TRUE)[, -1]
  rval <- as.matrix(rval)
  which[which == "emos_log"] <- "EMOS"
  colnames(rval) <- which

  return(rval)
}

fmt <- function(x) {
  ## select and aggregate (if necessary)
  if(!is.null(dim(x))) {
    x <- if("elapsed" %in% colnames(x)) x[, "elapsed"] else colMeans(x, na.rm = TRUE)
  }

  ## round and format to three digits
  x <- format(round(x, digits = 3), nsmall = 3)

  ## collect results across methods (if necessary)
  if(length(x) == 1L) return(x)
  nam <- c("distforest", "gamlss", "gamboostLSS", "gamboostLSS_cvr", "EMOS")
  x <- x[nam]
  x[is.na(x)] <- ""
  names(x) <- nam
  return(x)
}

## (ii) summary table
tab <- cbind(
  "crps_24to4" =     fmt(results$crps),
  "fittime_24to4" =  fmt(results$fit_time),
  "predtime_24to4" = fmt(results$pred_time),
  "crps_cv" =        fmt(get_cross("crps")),
  "fittime_cv" =     fmt(get_cross("evaltime")),
  "predtime_cv" =    fmt(get_cross("predtime"))
)

## (iii) crps skill score
crps_cross <- get_cross("crps")
crps_cross_mean <- colMeans(crps_cross, na.rm = TRUE)
crpss_cross <- 1 - crps_cross[, c("distforest", "gamlss", "gamboostLSS")] / crps_cross[, "EMOS"]
crpss_24to4 <- 1 - colMeans(results$crps[, c("distforest", "gamlss", "gamboostLSS")], na.rm = TRUE) / mean(results$crps[, "EMOS"], na.rm = TRUE)

## (iv) variable importance
varimp10 <- rev(head(sort(results$vimp_crps, decreasing = TRUE), 10))
names(varimp10) <- gsub("pow", "", names(varimp10), fixed = TRUE)
@

\begin{document}
\SweaveOpts{concordance=TRUE}

\section*{\huge \Sexpr{station}}

Longitude: \Sexpr{fmt(StationsTyrol[StationsTyrol$name == station, "lon"])}

Latitude: \Sexpr{fmt(StationsTyrol[StationsTyrol$name == station, "lat"])}

Altitude: \Sexpr{fmt(StationsTyrol[StationsTyrol$name == station, "alt"])}

\setkeys{Gin}{width=0.7\textwidth}
\vspace*{-4.1cm}\hspace*{4cm}%
<<map, fig=TRUE, echo = FALSE, height = 5, width=7=>>=
plot(MapTyrol$SpatialPolygons)
points(sp, pch = 21, col = "black", las = 1, cex = 1)
points(sp[which(sp$name == station), ], pch = 21, bg = hcl(325, 100, 70), cex = 1.5)
@
\vspace*{-1.6cm}

\subsection*{Overview}

\begin{tabular}{lrrrrrr}
\hline
      & \multicolumn{3}{l}{Single split} & \multicolumn{3}{l}{10 times 7-fold cross-validation} \\
Model & CRPS & Fit time & Predict time                 & CRPS & Fit time & Predict time                       \\ \hline
<<tab, echo=FALSE, results=tex>>=
tab <- apply(tab, 1, paste, collapse = " & ")
tab <- paste(c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "\\emph{(mstop selection)}", "EMOS"), "&", tab, "\\\\")
writeLines(tab)
@
\hline
\end{tabular}

\subsection*{CRPS skill score (cross-validation, reference: EMOS)}

\setkeys{Gin}{width=0.73\textwidth}
<<boxplot_crps_skillscore, fig=TRUE, echo=FALSE, height=4, width=6.4>>=
par(mar = c(2.5, 4, 1, 2))
boxplot(crpss_cross, 
        ylim = c(min(-0.01, min(crpss_cross, na.rm = TRUE) - 0.01), 
                 max(-0.01, max(crpss_cross, na.rm = TRUE) + 0.01)),
        main = "",
        names = c("Distributional\nforest", "Prespecified\nGAMLSS", "Boosted\nGAMLSS"),
        ylab = "CRPS skill score", col = "lightgray", las = 1) 
#axis(1, 0:4, c("", "Distributional\nforest", "Prespecified\nGAMLSS", "Boosted\nGAMLSS", ""))
abline(h = 0, col = hcl(50, 100, 50), lwd = 2)
# lines(crpss_24to4, col = hcl(128, 100, 50), lwd = 2, type = "o", pch = 19)
@

\subsection*{Variable importance}

<<varimp, fig=TRUE, echo=FALSE, height=4, width=6.4>>=
par(mar = c(5, 10, 1, 2))
barplot(varimp10, horiz = TRUE, las = 1, axes = FALSE,
        xlab = "Mean decrease in CRPS")
axis(1, at = seq(0, ceiling(max(varimp10) * 5)/5, 0.2), las = 1, mgp = c(0, 1, 0))
@

\newpage

\subsection*{Residual QQ plots}

\setkeys{Gin}{width=0.93\textwidth}
<<qq, fig=TRUE, echo=FALSE, height=6, width = 7, png=TRUE, pdf=FALSE>>=
library("countreg")
set.seed(7)
par(mfrow = c(2, 2))
qqrplot(results$pit_df, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(results$pit_ml, nsim = 100, main = "EMOS",		    ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
if(!all(is.na(results$pit_g))) qqrplot(results$pit_g,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(results$pit_gb, nsim = 100, main = "Boosted GAMLSS",	    ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
@

\subsection*{PIT histograms}

<<pit, fig=TRUE, echo=FALSE, height=6, width = 7>>=
set.seed(4)
par(mfrow = c(2, 2))
pithist(results$pit_df, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0, 1.9))
pithist(results$pit_ml, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0, 1.9))
if(!all(is.na(results$pit_g))) pithist(results$pit_g,  nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0, 1.9))
pithist(results$pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAMLSS", ylim = c(0, 1.9))
@

\end{document}

### 2018-09-22 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
revised versions of supplements with corresponding reproduction files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2289 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 443c37f89e7b5b3420a6e5a3862fb100989d9f2d

@@
 as reference represented by the horizontal line at zero. Again, the forest model 
-and the EMOS show only slight variations (on average not even 1\%) over different 
+and the EMOS show only slight variations (on average not even one percent) over different 
 distributions while the two GAMLSS models perform even or slightly better applying 
@@
 
-Altogether it can be stated that the left-censored at zero Gaussian distribution seems 
+Altogether it can be stated that for this application the left-censored at zero Gaussian distribution seems 
 to be an appropriate choice while the other two distributions, especially the hurdle model, 
@@
 
-# CRPS skill score for boosted hgaussian
+# CRPS for hgaussian
 c_hgaussian <- matrix(ncol = 4, nrow = 15)
@@
 
-# CRPS skill score for prespecified logistic
+# CRPS for logistic
 c_logistic <- matrix(ncol = 4, nrow = 15)

### 2018-09-22 pkg/disttree/inst/draft/supplement/Rain_stationwise.R

@@ -1,0 +1,577 @@
+#######################################################
+### Probabilistic Forecasting on Precipitation Data ###
+#######################################################
+
+## Replication material for supplement of: 
+## Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain (2018)
+## by Lisa Schlosser and Torsten Hothorn and Reto Stauffer and Achim Zeileis
+## URL: http://arxiv.org/abs/1804.02921
+
+## This demo includes the application on selected stations
+## Full replication of all other results can be obtained with
+## demo("RainTyrol", package = "disttree")
+
+## Computation time: approximately .. minutes (on our machines, using .. kernels)
+
+
+library("disttree")
+
+#####
+# further packages
+library("gamlss")
+library("gamlss.dist")
+library("gamboostLSS")
+library("crch")
+library("scoringRules")
+library("RainTyrol")
+
+# if gamlss.cens family object should be used as family
+library("gamlss.cens")
+gen.cens(LO, type = "left")
+
+assign("NO",  gamlss.dist::NO,  pos = ".GlobalEnv")
+assign("dNO", gamlss.dist::dNO, pos = ".GlobalEnv")
+assign("pNO", gamlss.dist::pNO, pos = ".GlobalEnv")
+assign("qNO", gamlss.dist::qNO, pos = ".GlobalEnv")
+assign("rNO", gamlss.dist::rNO, pos = ".GlobalEnv")
+gamlss.cens::gen.cens(NO, type = "left")
+assign("NOlc",  NOlc,  pos = ".GlobalEnv")
+assign("dNOlc", dNOlc, pos = ".GlobalEnv")
+assign("pNOlc", pNOlc, pos = ".GlobalEnv")
+assign("qNOlc", qNOlc, pos = ".GlobalEnv")
+
+
+# set function for parallelization
+applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, detectCores() - 1))
+
+
+# HCL palette
+pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
+
+
+#####
+
+# formula 
+{
+  
+  # tree and forest formula
+  dt.formula <- df.formula <- 
+    robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
+    tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
[…]

### 2018-09-22 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
revised versions of supplements with corresponding reproduction files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2289 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 443c37f89e7b5b3420a6e5a3862fb100989d9f2d

@@
 The models are always learned on data from 24 years and tested on the 
-remaining 4 years. Splitting the data into these two subsets has been done
+remaining 4 years. Splitting the data into these two subsets is done
 in two different ways leading to the two following evaluation settings.
@@
 An exception is given by station \hyperlink{sumSt.JohannimWalde.1}{St.Johann im Walde} as it is the only station where the prespecified GAMLSS clearly outperforms all other models in both settings.
-But overall, the results from the cross validation also support the statement that the forest model performs well on all stations being either the best or among the best performing methods at the majority of stations.
+But overall, the results from the cross-validation also support the statement that the forest model performs well on all stations being either the best or among the best performing methods at the majority of stations.
 Moreover, it can be concluded that Axams is an observation station without any unusual features that would lead to outstanding performances of any of the methods.

### 2018-09-22 pkg/disttree/inst/draft/supplement/Rain_hurdleversions_forest.R

@@ -1,0 +1,396 @@
+#######################################################
+### Probabilistic Forecasting on Precipitation Data ###
+#######################################################
+
+## Replication material for supplement of: 
+## Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain (2018)
+## by Lisa Schlosser and Torsten Hothorn and Reto Stauffer and Achim Zeileis
+## URL: http://arxiv.org/abs/1804.02921
+
+## This demo includes the application of two different 
+## versions of the forest hurdle model on selected stations
+## (models learned on 24 years and evaluated on 4 years)
+## Full replication of all other results can be obtained with
+## demo("RainTyrol", package = "disttree")
+
+## Computation time: approximately ... (on our machines, using .. kernel)
+
+
+library("disttree")
+library("RainTyrol")
+library("crch")
+
+##### 
+# HCL palette
+pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
+
+
+#####
+# formula for truncated model
+{
+  
+  # tree and forest formula
+  dt.formula <- df.formula <- 
+    robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
+    tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
+    tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
+    capepow_mean + capepow_sprd + capepow_min + capepow_max + 
+    capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
+    capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
+    dswrf_mean_mean + dswrf_mean_max + 
+    dswrf_sprd_mean + dswrf_sprd_max +
+    msl_mean_mean + msl_mean_min + msl_mean_max + 
+    msl_sprd_mean + msl_sprd_min + msl_sprd_max +
+    pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
+    pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
+    tmax_mean_mean + tmax_mean_min + tmax_mean_max +
+    tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
+    tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
[…]

### 2018-09-22 pkg/disttree/inst/draft/draft.Rnw
small corrections and recalculation of model evaluations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2294 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1407f041d74a07199a243ac20d7dbabf4a01161a

@@
 is typically more relevant or, in some situations, a prediction interval of
-expected precipitation (say from the expected \textit{10\%} to \textit{90\%} quantile). Similar
+expected precipitation (say from the expected 10\% to 90\% quantiles). Similar
 considerations apply for other meteorological quantities and hence attention
@@
 $\bm{\theta} = (\mu, \sigma)$ is
-$$
+\begin{equation}
 \ell(\mu, \sigma; Y) = 
@@
 \end{cases}
-$$
+\end{equation}
 
@@
 points. Each of the distributional trees is grown on a different data set
-obtained through bootstrap sampling (or subsampling) and in each node only a
+obtained through resampling (bootstrap or subsampling) and in each node only a
 random subset of the covariates $\bold{Z}$ is considered. As usual in random
@@
 \item \emph{Distributional forest:} All 80~predictor variables are
-  considered for learning a forest of 100 trees. Bootstrap
-  sampling is employed for each tree using a third of the predictors
+  considered for learning a forest of 100 trees. Subsampling is employed 
+  for each tree using a third of the predictors
   in each split of the tree (\code{mtry}). Parameters are estimated by
@@
   \hl{As stopping criteria the minimal number
-  of observations to perform a split} (\code{minsplit}) \hl{is set to 50, 
+  of observations to perform a split} (\code{minsplit}) \hl{is set to $50$, 
   the minimal number of observations in a terminal node} (\code{minbucket})
-  \hl{is set to 20 and the significance level for variable selection} (\code{alpha}) 
+  \hl{is set to $20$ and the significance level for variable selection} (\code{alpha}) 
   \hl{is kept at the default value $1$.}
@@
   out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
-  the number of boosting iterations (``mstop'') is: $50, 75, \dots, 975, 1000$.
+  the number of boosting iterations (\code{mstop}) is: $50, 75, \dots, 975, 1000$.
 

### 2018-09-23 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
enhancements in stationwise supplement
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2295 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 80df211ca44fd0cbd1a862b185b25559c2c1615b

@@
 
-\Abstract{In the application discussed in the paper ''Distributional 
-Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
-the presented method is evaluated on 95 observation stations. More extensive
-evaluations including a cross-validation are preformed for one station, Axams.
-In this supplement this exact same analysis as for station Axams is
-made for 14 further observation stations.}
+\Abstract{
+  In the case study presented in the main manuscript ``Distributional Regression
+  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+  the novel distributional regression forests are evaluated and compared with
+  other distributional regression approaches on 95 meteorological stations.
+  Moreover, further details -- including a cross-validation analysis -- are
+  preformed for only one station, Axams. To show that Axams is a fairly typical
+  station and similar insights can be obtained for other stations as well,
+  this supplements presents the same analysis as in Section~3.3 of the main
+  manuscript for 14~further meteorological stations.
+}
 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
 
-\section{Introduction}
-This application study is supplementary material to the application in Section~3 of 
-the paper where the presented distributional forest model is employed to obtain 
-probabilistic precipitation forecasts in a mountainous region.
-Next to an evaluation over all 95 observation stations further investigations
-including a cross-validation have been performed for one selected station, Axams.
-Here, this more extensive evaluation is carried out, applying the exact same settings 
-as in the paper, for Axams and 14 further observation stations. Additionally, computation
-times are measured.
+\section{Setup}
 
-\section{Models and evaluation}
-As in the paper a distributional forest applying a left-censored at zero 
-Gaussian distribution is evaluated together with three other zero-censored 
-Gaussian models. Further details on all four models can be found in Section~3.2. of
-the paper. In particular, Table~2 gives an overview over the models and their 
-specifications. As a brief reminder the approaches are summarized below.
+\subsection{Models}
+
+As introduced in Section~3.2 of the main manuscript, a distributional forest
+based on a Gaussian distribution, left-censored at zero, is evaluated together
+with three other zero-censored Gaussian models. Table~2 of the manuscript provides
+an overview of all models considered and their specifications. As a reminder
+the approaches are briefly summarized below.
 %
@@
 \item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
-Standard ensemble model output statistics models use the ensemble 
+Standard ensemble model output statistics uses the ensemble 
 mean of total precipitation as regressor in the location submodel 
@@
 following \cite{Stauffer+Umlauf+Messner:2017}. The model is estimated by maximum
-penalized likelihood using a backfitting algorithm (\cite{Stasinopoulos+Rigby:2007}).
+penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.
 
@@
 automatically from all available variables, using non-cyclic boosting
-for parameter estimation (\cite{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}).
-The optimal stopping point for the iteration is evaluated based on 
+for parameter estimation \citep{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}.
+The optimal stopping point for the iterations is determined based on 
 a (computationally intensive) out-of-bag bootstrap estimate of the log-likelihood.
@@
 
-%\subsection{Subset selection}
+\subsection{Learning and test samples}
+
 The models are always learned on data from 24 years and tested on the 
@@
 
-\item Single split: The data is time-ordered. The learning data consists of the first 24 years (1985--2008) and the testing data of the following 4 years (2009--2012).
+\item \emph{Single split}: The data is time-ordered. The learning data consists of the
+first 24 years (1985--2008) and the testing data of the following 4 years
+(2009--2012).
 
-\item Cross-validation: In order to not just split the data at one splitpoint the 28 years are randomly split in 7 blocks of 4 years which are not necessarily successive.
-Each of the blocks is chosen once as testing data while the remaining 6 blocks are the learning data.
-This procedure is repeated 10 times resulting in a 10 times 7-fold cross-validation framework.
+\item \emph{Cross-validation}: In order to not just split the data at one splitpoint
+the 28 years are randomly split into 7 blocks of 4 years which are not
+necessarily successive. Each of the blocks is chosen once as testing data while
+the remaining 6 blocks are the learning data. This procedure is repeated 10
+times resulting in a 10 times 7-fold cross-validation framework.
+
+\end{itemize}
+%
+The performances of the models are compared based on numerical and graphical evaluations.
+
+\subsection{Numerical evaluation}
+
+Several numerical summary statistics are computed in both
+settings. For the single-split setting one value is calculated for each
+observation in the test data and then their mean is reported. For the
+cross-validation setting, the same is done for all 70~splits considered and
+again their mean is reported. The following summary statistics are used.
+%
+\begin{itemize}
+
+\item \emph{CRPS}: The continuous ranked probability score (CRPS) is calculated to
+assess the predictive performance in terms of full probabilistic forecasts.
+
+\item \emph{Computation time}: For all models fitting times and predictions times
+are measured on the same 16-core Linux computer. In particular,
+\textit{fit time} refers to the time needed to fit the model to the learning
+data in a particular split. The \textit{predict time} refers to the time needed
+to compute the out-of-sample prediction. In the single-split setting, the
+predictions are computed observation by observation, i.e., emulating the
+realistic setup where predictions are made once the new NWP model outputs become
+available. In the cross-validation setting, the predictions for the full 4-year
+out-of-sample period in one go. Although this is not realistic in this
+particular application, it might be an interesting quantity for comparing the
+competing regression approaches.
+
+\end{itemize}
+%
+\textit{Note:} For a few splits in the cross-validation, the prespecified GAMLSS
+(as computed by the \proglang{R} package \pkg{gamlss}) failed, leading to \code{NA}s
+in the corresponding summary statistics. As mentioned by \cite{Hofner+Mayr+Schmid:2016}
+such problems are not unusual because GAMLSS ``can be unstable, especially when it comes
+to selecting possibly different sets of variables for multiple distribution parameters''.
+
+\subsection{Graphical evaluation}
+
+The following methods of visualization are applied to
+provide further means of comparison of all models but also to investigate on the
+distributional forest in particular.
+%
+\begin{itemize}
+
+\item \emph{Boxplot of the CRPS skill score:} With EMOS as the reference model
+the CRPS skill scores are calculated on each of the 70~splits considered in the
+cross-validation setting and illustrated in a boxplot.
+
+\item \emph{Barplot of variable importance:} For the distributional forest it is
+investigated which variables have the strongest influence on the model and its
+goodness of fit. For each variable this is measured based on mean decrease in
+CRPS after permutation. The 10 variables showing the highest values in the
+single-split setting are listed in a barplot.
+
+\item \emph{Residual QQ plots:} For each of the models an out-of-sample residual
+QQ plot for the single-split setting provides a graphical goodness-of-fit
+assessment.
+
+\item \emph{PIT histograms:} Similarly, the out-of-sample probability integral
+transform (PIT) in the single-split setting is amployed as an alternative
+goodness-of-fit visualization.
 
@@
 
-The performances of the models are compared based on numerical and graphical evaluations.
+\section{Stations}
 
-\textit{Numerical:}
-For the single split setting one value is calculated for each observation in the test data and the mean of these values is considered.
-For the cross-validation this mean value over all observations in the test data is retrieved from all 70 evaluations and then again the mean of them is considered.
-This holds for the following measurements.
-%
-\begin{itemize}
-
-\item CRPS: The continuous ranked probability score (CRPS) is calculated to assess the predictive performance in terms of full probabilistic forecasts.
-
-\item Time: For all models fitting times and predictions times are measured.
-In particular, \textit{fit time} refers to the time needed to fit the model to the whole learn data.
-Depending on the setting \textit{predict time} refers to the time needed to make one prediction in the single split setting and to make predictions for the whole test data in the cross-validation.
-
-\end{itemize}
-
-\textit{Graphical:} The following methods of visualization are applied to provide further means of comparison of all models but also to investigate on the distributional forest in particular.
-%
-\begin{itemize}
-%
-\item Boxplot of the CRPS skill score: With the EMOS as reference model the CRPS skill scores are calculated for the cross-validation setting and illustrated in a boxplot.
-%
-\item Barplot of variable importance: For the forest model it is investigated which variables have the strongest influence on the model and its goodness of fit.
-For each variable this is measured based on mean decrease in CRPS after permutation.
-The 10 variables showing the highest values in the single split setting are listed in a barplot.
-%
-\item Residual QQ plots: For each of the models an out-of-sample residual QQ plot for the single split setting provides a visualization of goodness of fit.
-Randomized residual quantiles are compared to standard normal quantiles.
-%
-\item PIT histograms: As an additional visualization out-of-sample probability integral transform histograms illustrate how well-calibrated the distributional fits are for the single split setting.
-%
-\end{itemize}
-
-\section{Stations}
-Out of the 95 observation stations 15 are considered for this 
-benchmark study. They have been selected based on the region
-they are in, their altitude and the performance of the 
-different models in the application in the paper such that
-all of these features are represented equally.
+Out of the 95~observation stations 15 are considered for this 
+supplemental study. They have been selected to cover a wide range regarding
+geographical location, altitude, and which of the competing models performed
+best in the single-split setting reported in Section~3.4 of the main manuscript.
 
@@
 \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol},
-\hyperlink{sumStJohannimWalde.1}{St.Johann im Walde}.
+\hyperlink{sumStJohannimWalde.1}{St.~Johann im Walde}.
 
@@
 
-\subsection{Forecasting skills} 
-Comparing CRPS values of the single split setting it can be stated that the forest model works well over all stations.
-Often it is even the best performing method (e.g.
-at stations \hyperlink{sumLech.1}{Lech}, \hyperlink{sumSeeimPaznaun.1}{See im Paznaun} and \hyperlink{sumGinzling.1}{Ginzling}), but there are also stations where one of the other models leads to better results.
-For example at \hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut} the boosted GAMLSS performs best, at \hyperlink{sumRotholz.1}{Rotholz} the EMOS model and at \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol} the prespecified GAMLSS is the best model, however, they do not clearly outperform the forest model.
-Moreover, these advantages often even out or turn into disadvantages when looking at the CRPS values from the cross-validation.
-For example at station \hyperlink{sumInnervillgraten.1}{Innervillgraten}, the prespecified GAMLSS is clearly the best in the single split setting, however, in the cross-validation the forest model is slightly ahead.
-An exception is given by station \hyperlink{sumSt.JohannimWalde.1}{St.Johann im Walde} as it is the only station where the prespecified GAMLSS clearly outperforms all other models in both settings.
-But overall, the results from the cross-validation also support the statement that the forest model performs well on all stations being either the best or among the best performing methods at the majority of stations.
-Moreover, it can be concluded that Axams is an observation station without any unusual features that would lead to outstanding performances of any of the methods.
-Therefore, it is a reasonable choice as a representative for all 95 station.
-Examples for stations where the forest model performs better or worse can be found easily, however, in general the methods perform similarly as for station Axams.
+\subsection{Forecasting skill}
 
-%Comparing CRPS skill score values from the cross-validation the distributional forest
-%and the two GAMLSS versions show clear improvements compared to the reference model EMOS.
-%At many stations the forest and either the prespecified GAMLSS (e.g.
-at station Innervillgraten) or the boosted GAMLSS (e.g.
-at station Lech or Zuers) are ahead, almost on the same level, while the second GAMLSS is slightly behind.
-At stations Axams, See im Paznaun and Ginzling the forest clearly leads to the highest improvement in terms of CRPS.
-Station St.Johann im Walde stands out as it is the only of the 15 stations where the prespecified GAMLSS is clearly ahead of the other models.
-But overall it can be stated that all three models perform well with only small variations over stations.
-Even if one of the evaluated models shows clear advantages in the single split setting (such as the EMOS at station Rotholz or the prespecified GAMLSS at station Innervillgraten) this is evened out within the cross-validation.
+Based on the out-of-sample CRPS results it can be concluded that the
+distributional forest generally performs well on all stations and is typically
+close to (e.g., \hyperlink{sumLech.1}{Lech}) or somewhat better than
+the (next-)best method (e.g., \hyperlink{sumSeeimPaznaun.1}{See im Paznaun} and
+\hyperlink{sumGinzling.1}{Ginzling}). Naturally, there is more variation
+in the single-split results with another models performing better than the
+distributional forest. However, in the cross-validation such differences
+often even out, showing that distributional forests are on par. Examples for
+this include \hyperlink{sumRotholz.1}{Rotholz},
+\hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol}, and 
+\hyperlink{sumOchsengartenObergut.1}{Ochsengarten-Obergut}, where the 
+best models in the single split are EMOS, prespecified GAMLSS and boosted
+GAMLSS, respectively. The only station, where another model performs
+clearly better than distributional forests in both settings, is
+\hyperlink{sumSt.JohannimWalde.1}{St.~Johann im Walde}.
+
+Overall, the results support the conclusion that distributional forests
+perform well for this prediction task -- they are often on par and sometimes
+even better than the other models considered. Moreover, it is shown that the
+results for Axams are fairly typical and not particularly unusual compared
+to the other stations considered.
 
 \subsection{Computation time}
-The forest model and the two GAMLSS models achieve similar results in both time measurements, fit and predict time, while the EMOS model is clearly the fastest.
-However, for the boosted GAMLSS the cross-validation applied to find the optimal stopping value \code{mstop} is extremely time consuming.
-This additional time can not be neglected since the stopping value is crucial for the complexity of the model and the selection of covariates.
 
-As to prediction times it has to be mentioned that the forest can keep up with the other models if only one prediction is made.
-However, to make predictions for a whole set of new observations the forest clearly needs more time.
-This is due to the fact that the forest has to computes each single prediction separately as the maximum-likelihood estimation is based on observation-specific weight matrix.
+Distributional forests are comparable with respect to the computation times
+to the other two flexible GAMLSS models. Only the basic EMOS model is clearly
+the fastest. Moreover, unlike the other models, the boosted GAMLSS requires
+a time-consuming cross-validation to determine the optimal stopping value
+\code{mstop}. This is crucial for selecting the appropriate model complexity
+and relevant covariates and hence cannot be neglected in the analysis.
+
+The observation-wise prediction times in the single-split setting are in the same
+order of magnitude for the three flexible distributional models. However, when
+many predictions are computed simultaneously in the cross-validation setting,
+then distributional forests are slower than the GAMLSS models. The reason for this
+is that the GAMLSS can set up the model matrices once and then carry out a matrix
+product while the distributional forest necessitates fitting a separate set of
+parameters for each observation-specific weight vector.
 
 \subsection{Variable importance}
-Concerning variable importance, two covariates based on total precipitation (\code{tp\_max} and \code{tp\_mean}) are at the top of the list for all stations, often clearly ahead of the other covariates.
-The other members of the top ten variables vary slightly over stations, however, further covariates concerning total precipitation, total column-integrated condensate and temperature differences are often among them.
+
+Two NWP model outputs based on total precipitation (\code{tp\_max} and
+\code{tp\_mean}) are clearly the most important predictors for precipitation
+at all stations considered. The remaining variables typically follow with
+some margin and vary slightly in their exact order. However, they typically
+encompass some NWP model outputs for total precipitation, total column-integrated
+condensate, and temperature differences.
 
 \subsection{Calibration}
-The PIT histograms and residual QQ plots show that overall all four models are reasonably well calibrated at the majority of stations while at some stations they all fit better (e.g.
-at station \hyperlink{sumOetz.1}{Oetz}) or worse (e.g.
-at station \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol}) than on average.
-One feature of the boosting method that can be detected in the residual QQ plots for many stations is that heavy upper tails are not captured very well.
-This is also supported by the PIT histogram.
 
-\textit{Note:} As mentioned by \cite{Hofner+Mayr+Schmid:2016} the methods provided by GAMLSS ``can be unstable, especially when it comes to selecting possibly different sets of variables for multiple distribution parameters''.
-These problems have also occurred in a few situations for the
-prespecified GAMLSS model within the cross-validation. In case of this method failing to 
-build a model it is represented by \code{NA}s in the evaluation.
+The PIT histograms and residual QQ plots show that overall all four models are
+reasonably well-calibrated at the majority of stations while at some stations
+they all fit better (e.g., at station \hyperlink{sumOetz.1}{Oetz}) or worse (e.g.,
+at station \hyperlink{sumMatreiinOsttirol.1}{Matrei in Osttirol}) than on
+average. For several stations, however, the boosted GAMLSS does not capture
+the heavy upper tail as well as the other methods which is best seen in the
+residual QQ plots but also the PIT histograms.
 
@@
 \includepdf[pages=-, link, linkname=sumStJohannimWalde]{StJohannimWalde.pdf}
-\bookmark[dest=sumStJohannimWalde.1]{St.Johann im Walde}
+\bookmark[dest=sumStJohannimWalde.1]{St.~Johann im Walde}
 

### 2018-09-26 pkg/disttree/inst/draft/draft.Rnw
correction of notational mistakes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2296 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ee2aa82b3f7ffeeb68707860c4f9ce9ad6029a18

@@
 the value of $Z_{li}$ is $1$. In this case the statistic $T_l \in \mathbb{R}^{H \cdot k}$
-as the ``vec'' operator converts the $H \times k$ matrix into a $k$ column vector 
+as the ``vec'' operator converts the $H \times k$ matrix into a $H \cdot k$ column vector 
 by column-wise combination.} Observations with missing values are excluded from the sums.
@@
 \hl{The observed multivariate linear statistic $t_l$ which is either an
-$k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of Y,
+$k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of $Z_l$,
 is mapped onto the real line by a univariate test statistic $c$.

### 2018-09-28 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
some touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2297 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cf6ac1fc0569166b41731ec77e0c53af34018b88

@@
 
-\title{Supplement 1: Testing different distributions}
+\title{Supplement 1: Different Response Distributions}
 
@@
 
-\Abstract{In the application discussed in the paper ''Distributional 
-Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
-the underlying distribution is assumed to be a left-censored at zero Gaussian distribution.
-In this supplement the exact same evaluations as in the paper are performed employing
-other distribution assumptions. In particular, a left-censored at zero logistic distribution 
-and a hurdle-model are applied and their results are compared to those of the left-censored 
-at zero Gaussian distribution presented in the paper.}
+\Abstract{
+  In the case study presented in the main manuscript ``Distributional Regression
+  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+  the novel distributional regression forests are based on a Gaussian response distribution,
+  left-censored at zero to accomodate both the non-negativity of precipitation and
+  the point mass at zero. To assess the goodness of fit of this response distribution
+  this supplement employs the same evaluations as in the main manuscript but based on
+  two other distributional assumptions: A logistic distribution, left-censored at zero,
+  is employed to potentially better capture heavy tails -- and a two-part hurdle model
+  combining a binary model for zero vs.\ positive precipitation and a Gaussian model,
+  truncated at zero, for the positive precipitation observations. The latter could
+  potentially better discriminate determinants of the occurrence of precipitation
+  vs.\ the amount of precipitation given that it occurs.
+}
 
@@
 \section{Introduction}
-This application study is supplementary material to the application presented
-in Section~3 of the paper.
-There, the presented distributional forest model is employed to obtain 
-probabilistic precipitation forecasts in a mountainous region assuming that the 
-response variable (the power-transformed daily precipitation sum) follows a 
-left-censored Gaussian distribution with censoring point~$0$.
 
-To assess how appropriate this distributional assumption is we test the robustness
-of the results with respect to two questions: (1)~Are the tails of the distribution
-captured appropriately by a Gaussian distribution or are \emph{heavier tails} required?
+This supplement extends the case study presented in Section~3.4 of the main
+manuscript by carrying out the same evaluation at 15~stations using two
+alternative response distributions for the power-transformed daily precipitation
+sums.
+
+To assess how appropriate the distributional assumption (left-censored Gaussian distribution
+with censoring point zerol; or zero-censored for short) is, two questions are considered:
+(1)~Are the tails of the distribution captured appropriately by a Gaussian distribution or
+are \emph{heavier tails} required?
 (2)~Is the point mass at zero (= no rain) driven by the same determinants as the
 positive observations (= rain) or should this be captured by a \emph{separate parameter}?
-For (1) we compare the censored Gaussian with a \emph{censored logistic} distribution and
-for (2) a Gaussian \emph{hurdle model} (or two-part model) is employed instead of the
-censored Gaussian.
+For (1) we compare the zero-censored Gaussian with a \emph{zero-censored logistic} distribution and
+for (2) a Gaussian \emph{hurdle model} (or two-part model with binary and zero-truncated parts)
+is employed instead of the zero-censored Gaussian.
 
-To compare the performance of the distributional forest to other commonly used
-methods the exact same models as in the paper are evaluated on the same data set 
-and applying the same settings while only the distribution is exchanged. Further 
-details on the models can be found in Section~3.2. of the paper. In particular, 
-Table~2 gives an overview over the models and their specifications. As a brief 
-reminder the approaches are summarized below.
-
+Just as in the main manuscript the performance of the distributional forest is also
+compared with other commonly-used distributional regression models using the same
+data set. Table~2 of the manuscript provides
+an overview of all models considered and their specifications. As a reminder
+the approaches are briefly summarized below.
+%
 \begin{itemize}
 
-\item \emph{Distributional forest:} All predictor variables are considered for 
-learning a forest where subsampling is employed for each tree. Parameters are 
-estimated by adaptive local likelihood based on forest weights.
+\item \emph{Distributional forest:} All predictor variables are
+considered for learning a forest where subsampling is employed 
+for each tree. Parameters are estimated by
+adaptive local likelihood based on forest weights.
 
 \item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
-Standard ensemble model output statistics models use the ensemble 
+Ensemble model output statistics uses the ensemble 
 mean of total precipitation as regressor in the location submodel 
@@
 
-\section{Distributions applied / Distributional approaches}
+\section{Alternative response distributions}
 
-The distribution applied in the paper is based on a Gaussian distribution 
-for which the probability density function $f$ and the distribution 
-function $F$ are defined by
+The main manuscript employes a Gaussian distribution with probability
+density function $f$ (PDF) and cumulative distribution function $F$ (CDF)
+given by
 %
@@
 \end{align}
-where $\phi$ and $\Phi$ are the probability density function and the 
-distribution function of the standard normal distribution $\mathcal{N}(0,1)$.
+where $\phi$ and $\Phi$ are the PDF and the CDF of the standard normal
+distribution $\mathcal{N}(0,1)$.
 
-Alternatively one could also choose a distribution with heavier tails 
-such as a logistic distribution with corresponding probability density 
-function and distribution function
+An alternative distribution with several similar properties but with
+heavier tails is the logistic distribution. The PDF and CDF are given by
 %
@@
 %
-Both distributions have to be adapted in our application to capture that
+Both distributions have to be adapted in our case study to accomodate that
 precipitation is non-negative with a point mass at zero (= no rain).
@@
 or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
-a truncated (positive) part \citep[see][for an overview]{Winkelmann:2006}.
+a truncated (positive) part \citep[see][for an overview of both techniques]{Winkelmann:2006}.
 \begin{itemize}
-\item Censored distribution \citep{Winkelmann:2006}: 
-%In general, censoring data usually leads 
-%to a point mass at the censoring point which has to be considered in 
-%the modeling process. Therefore, censored distributions are applied as 
-%they are able to capture this point mass. 
-%In precipitation data, observations can not be negative, thus, left-censoring
-%it at zero does not change the data. However, there is a naturally given
-%point mass at zero (days without rain) making a censored distribution an 
-%appropriate choice to model precipitation data.
-For a chosen distribution with probability density function $f$ and
-distribution function $F$ the log-likelihood function of the corresponding 
-censored distribution is
+\item \emph{Censored distribution:}
+For a chosen distribution with PDF $f$ and CDF $F$ the log-likelihood function of the corresponding 
+zero-censored distribution is
 %
@@
 %
-as seen in Equation~2.1.1. of the paper for a Gaussian distribution.
-For positive values of $Y$ the censored version of the density function
+as stated in Equation~2.1.1 of the main manuscript for the Gaussian special case.
+For positive values of $Y$ the censored version of the PDF
 equals the original non-censored version. For $Y=0$ the censored density
@@
 censoring point. Therefore, the parameters $\mu$ and $\sigma$ control
-both parts, the point mass and the positive part of the distribution.
+both aspects, the point mass and the positive part of the distribution.
 
-\item Hurdle model / Two-part model (\cite{Cragg:1971}, \cite{Mullahy:1986}): 
+\item \emph{Hurdle model \citep[or two-part model,][]{Cragg:1971,Mullahy:1986}:}
 An additional parameter $\nu$ describing the probability that there is 
 any precipitation at all, i.e. $\nu = \mathbb{P}(Y>0)$, is included in the model.
-The three-parametric log-likelihood function for the full hurdle model 
+The three-parameter log-likelihood function for the full hurdle model 
 with parameter vector $\bm{\theta} = (\mu, \sigma, \nu)$ is
@@
 \end{equation}
-For $Y=0$ the probability that it won't rain is expressed by $1-\nu$. 
+For $Y=0$ the probability for no rain is expressed by $1-\nu$. 
 For $Y>0$ the probability that it will rain $\nu$ is multiplied with the 
-left-truncated at zero density function. Contrary to the censored model, 
-here it is only the parameter $\nu$ controlling the point mass while $\mu$ 
-and $\sigma$ only influence the distribution of the positive observations.
+left-truncated at zero density function. In contrast to the censored model, 
+there is thus a dedicated parameter $\nu$ controlling the point mass while 
+the other parameters $\mu$ and $\sigma$ only influence the distribution of
+the positive observations.
 

### 2018-09-29 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
more work on supplements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2299 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 42375fce433cf91b5ec4e546daaa344ee5df0bc4

@@
 \item \emph{Hurdle model \citep[or two-part model,][]{Cragg:1971,Mullahy:1986}:}
-An additional parameter $\nu$ describing the probability that there is 
-any precipitation at all, i.e. $\nu = \mathbb{P}(Y>0)$, is included in the model.
+An additional parameter~$\nu$ describing the probability that there is 
+any precipitation at all, i.e., $\nu = \mathbb{P}(Y>0)$, is included in the model.
 The three-parameter log-likelihood function for the full hurdle model 
@@
 \end{equation}
-For $Y=0$ the probability for no rain is expressed by $1-\nu$. 
-For $Y>0$ the probability that it will rain $\nu$ is multiplied with the 
-left-truncated at zero density function. In contrast to the censored model, 
+For $Y=0$ the probability for no rain is $1-\nu$. 
+For $Y>0$ the probability $\nu$ for no rain is multiplied with the 
+PDF, left-truncated at zero. In contrast to the censored model, 
 there is thus a dedicated parameter $\nu$ controlling the point mass while 
@@
 \end{itemize}
+%
+Note that, due to the logs in the first part of Equation~\ref{eq:hurdle},
+the log-likelihood of the hurdle model can be separated additively into
+a part that depends only on $\nu$ (for all observations) and a part that
+depends on $\mu$ and $\sigma$ (for only the positive observations). Therefore,
+in a GAMLSS approach, the log-likelihood of the $\nu$ submodel can be
+maximized separately from the log-likelihood of the $\mu$ and $\sigma$
+submodels to maximize the overall log-likelihood. Hence in all GAMLSS
+models considered subsequently a binary logit model is employed for
+$Y = 0$ vs.\ $Y > 0$ and a zero-truncated Gaussian model in the case
+of $Y > 0$. Similarly, for the distributional forest two separate forests
+are employed with a binary and a zero-truncated Gaussian response,
+respectively. For comparison with the zero-censored distributions
+the combined three-parameter hurdle model as formulated in Equation~\ref{eq:hurdle}
+is employed.
 
-In the paper a left-censored at zero Gaussian distribution is chosen.
-Here, we additionally apply a left-censored at zero logistic distribution
-and a hurdle model based on a Gaussian distribution.
-
-In the log-likelihood function in Equation~\ref{eq:hurdle} properties of the
-logarithmic function can be applied to split the term for $Y>0$ 
-into two additive terms separating the parameters $\mu$ and $\sigma$
-from the additional parameter $\nu$. Therefore, $\nu$ can be modeled 
-by a separate model in the GAMLSS framework. The same approach is
-chosen for the distributional forest model in this application study.
-
-The models are compared based on the continuous ranked probability 
-score (CRPS). For the censored models the closed form following 
-\cite{Jordan:2018} is used while a numerical approximation is 
-calculated for the hurdle models.
+All models are evaluated using the continuous ranked probability 
+score (CRPS). For the censored models the closed form formulas provided by
+\cite{Jordan:2018} are used while a numerical approximation is 
+employed for the hurdle models.
 
 \section{Stations}
-Out of the 95 observation stations 15 are considered for this benchmark 
-study (see Figure~\ref{fig:map}). 
-They have been selected based on the region they are in, their altitude 
-and the performance of the different models in the application in the 
-paper such that all of these features are represented equally.
+
+Out of the 95~observation stations 15 are considered for this 
+supplemental study (see Figure~\ref{fig:map}). They have been selected to cover a wide range regarding
+geographical location, altitude, and which of the competing models performed
+best in the single-split setting reported in Section~3.4 of the main manuscript.
 
@@
 Oetz, Ochsengarten-Obergut, Ginzling, Rotholz, Walchsee, Koessen, 
-Innervillgraten, Matrei in Osttirol, St.Johann im Walde.
+Innervillgraten, Matrei in Osttirol, St.~Johann im Walde. See also Supplement~2
+(``Stationwise Evaluation'') for more details.
 
-\begin{figure}[h]
-\vspace*{-3cm}
+\begin{figure}[t!]
+\vspace*{-3.8cm}
 \setkeys{Gin}{width=0.99\textwidth}
@@
 @
-\vspace*{-3.5cm}
+\vspace*{-3.8cm}
 \caption{\label{fig:map}Map of Tyrol with all 95 observation stations and 
@@
 
+\pagebreak
+
 \section{Results}
-In Figure~\ref{fig:lattice} each panel belongs to one of the 15 observation stations. 
-The three distributions \textit{gaussian} (left-censored at zero Gaussian distribution), 
-\textit{hgaussian} (hurdle model based on the Gaussian distribution) and 
-\textit{logistic} (left-censored at zero logistic distribution) are compared based 
-on the resulting CRPS values for each of the four applied methods that are represented 
-by different colors and symbols. 
-Overall, the forest model performs almost equally well for all three distributions. 
-Therefore it seems to be very robust against missspecifications of the distribution.
-For the two GAMLSS models the censored Gaussian and the hurdle model perform similarly 
-at most of the stations while applying the censored logistic distribution leads to an 
-increase in CRPS, i.e. to a decline in performance, especially for the prespecified 
-GAMLSS (for example at stations Zuers and Jungholz).
-For the EMOS model exchanging the distribution does not have a strong effect on its 
-performance, similar to the forest model.
-All in all, exchanging the underlying distribution does not lead to clear 
-improvements of any of the methods, especially the censored Gaussian and the 
-hurdle model achieve almost the same results. However, the robustness against 
-missspecification of the distribution varies slightly over stations. For example 
-at station Matrei in Osttirol all methods seem to be very stable while stronger 
-influences can be seen at station Jungholz.
 
-\textit{Note:} When applying a hurdle model for the stations Oetz and 
-Innvervillgraten the prespecified GAMLSS computationally failed to build/fit 
-a model. Therefore this method is not represented in the corresponding panels 
-of Figure~\ref{fig:lattice} for these two cases.
+\subsubsection*{CRPS for all distributions by model and station}
 
-In Figure~\ref{fig:box_crpss_gaussian} the tested distributions are compared 
-for each method based on CRPS skill score with the censored Gaussian distribution 
-as reference represented by the horizontal line at zero. Again, the forest model 
-and the EMOS show only slight variations (on average not even one percent) over different 
-distributions while the two GAMLSS models perform even or slightly better applying 
-a hurdle model and loose strength when employing a logistic distribution compared 
-to a Gaussian distribution.
+Each panel of Figure~\ref{fig:lattice} compares the CRPS across the three distributions
+considered: \textit{cgaussian} (zero-censored Gaussian), \textit{hgaussian} (Gaussian hurdle model),
+and \textit{clogistic} (zero-censored logistic). The four different models considered
+are coded by both colors and symbols and connected by lines for easier comparison between
+models. As the average CRPS (i.e., the difficulty of the prediction problem) varies
+substantially by station, separate panels are employed for each station.
+%
+\begin{itemize}
 
-\begin{figure}
+\item For the distributional forest all three distributions perform very similarly for all
+stations. Thus, it appears to be very robust against missspecifications of the distribution.
+
+\item Similarly, for the EMOS model all distributions lead to comparable CRPS for all
+stations. However, there is some more variation in CRPS across distributions compared to
+the distributional forest.
+
+\item In contrast, the two GAMLSS models typically perform better (i.e., have lower CRPS) for
+the Gaussian distributions compared to the logistic distribution. The preference of
+censored vs.\ hurdle specification is not unanimous across stations.
+
+\end{itemize}
+%
+\textit{Note:} When applying the hurdle model at stations Oetz and 
+Innvervillgraten the prespecified GAMLSS could not be estimated (using the \proglang{R}
+package \pkg{gamlss}) due to numerical issues in the fitting algorithm.
+Therefore, this method is not represented in the two corresponding panels 
+of Figure~\ref{fig:lattice}.
+
+\begin{figure}[p!]
 \setkeys{Gin}{width=0.99\textwidth}
@@
 load("Rain_distributions.rda")
+levels(results$means$distribution) <- c("cgaussian", "hgaussian", "clogistic")
+levels(results$means$station)[16] <- "St. Johann im Walde"
+
 stations <- c("Axams", "Lech", "Zuers", "See im Paznaun", "Jungholz", 
@@
               "Innervillgraten", "Matrei in Osttirol", 
-              "St.Johann im Walde")
+              "St. Johann im Walde")
 
@@
 @
-\caption{\label{fig:lattice}CRPS values of all four methods, once applied with 
-each of the three distributions for each of the 15 observation stations.}
+\caption{\label{fig:lattice}CRPS values (lower = better) across the three
+distributions for all four models and 15~stations.}
 \end{figure}
 
-\begin{figure}
+\subsubsection*{CRPS skill score in reference to zero-censored Gaussian}
+
+To aggregate the CRPS results across stations, CRPS skill scores are visualized
+with boxplots in Figure~\ref{fig:box_crpss_gaussian}, using the zero-censored Gaussian distribution
+as the reference distribution (orange horizontal line at zero). The underlying skill
+scores for each station are also included in the graphic by gray lines.
+%
+\begin{itemize}
+
+\item As before, there are only slight variations for the distributional forest, mostly
+  between $\pm 1$\%. While the median CRPS skill score over the 15 stations for the 
+  Gaussian hurdle model is slightly better than for the simpler zero-censored Gaussian
+  model, this advantage is very small and not systematic across stations.
+
+\item For EMOS, the median skill score for both Gaussian hurdle and zero-censored logistic
+  are very close to the zero-censored Gaussian reference. However, the variation for the
+  hurdle model is much higher than for the censored model.
+
+\item For the two GAMLSS models the Gaussian hurdle is quite close to the zero-censored
+  Gaussian reference, while the zero-censored logistic clearly performs worse (especially
+  in the prespecified GAMLSS).
+
+\end{itemize}
+
+\begin{figure}[t!]
 \setkeys{Gin}{width=0.99\textwidth}
-<<results_box_crpss_gaussian, fig=TRUE, echo=FALSE, height=3, width=7>>=
+<<results_box_crpss_gaussian, fig=TRUE, echo=FALSE, height=3, width=7.2>>=
 # crps skill score by distribution (reference: gaussian)
 means$crps_ss_dist <- means$crps
-means$crps_ss_dist[means$distribution == "gaussian"] <- 
-  1 - means$crps[means$distribution == "gaussian"] / means$crps[means$distribution == "gaussian"]
-means$crps_ss_dist[means$distribution == "logistic"] <- 
-  1 - means$crps[means$distribution == "logistic"] / means$crps[means$distribution == "gaussian"]
+means$crps_ss_dist[means$distribution == "cgaussian"] <- 
+  1 - means$crps[means$distribution == "cgaussian"] / means$crps[means$distribution == "cgaussian"]
+means$crps_ss_dist[means$distribution == "clogistic"] <- 
+  1 - means$crps[means$distribution == "clogistic"] / means$crps[means$distribution == "cgaussian"]
 means$crps_ss_dist[means$distribution == "hgaussian"] <- 
-  1 - means$crps[means$distribution == "hgaussian"] / means$crps[means$distribution == "gaussian"]
+  1 - means$crps[means$distribution == "hgaussian"] / means$crps[means$distribution == "cgaussian"]
 
-means_sel <- means[means$distribution %in% c("logistic", "hgaussian"),]
+means_sel <- means[means$distribution %in% c("clogistic", "hgaussian"),]
 means_sel$distribution <- factor(means_sel$distribution, 
@@
 s_df[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
-s_df[,2] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "logistic", "crps_ss_dist"]
+s_df[,2] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "clogistic", "crps_ss_dist"]
 
@@
 s_g[,1] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
-s_g[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "logistic", "crps_ss_dist"]
+s_g[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "clogistic", "crps_ss_dist"]
 
@@
 s_gb[,1] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
-s_gb[,2] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "logistic", "crps_ss_dist"]
+s_gb[,2] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "clogistic", "crps_ss_dist"]
 
@@
 s_emos[,1] <- means_sel[means_sel$method == "EMOS" & means_sel$distribution == "hgaussian", "crps_ss_dist"]
-s_emos[,2] <- means_sel[means_sel$method == "EMOS" & means_sel$distribution == "logistic", "crps_ss_dist"]
+s_emos[,2] <- means_sel[means_sel$method == "EMOS" & means_sel$distribution == "clogistic", "crps_ss_dist"]
 
-colnames(s_df) <- colnames(s_g) <- colnames(s_gb) <- colnames(s_emos) <- c("hgaussian", "logistic")
+colnames(s_df) <- colnames(s_g) <- colnames(s_gb) <- colnames(s_emos) <- c("hgaussian", "clogistic")
 
@@
 @
-\caption{\label{fig:box_crpss_gaussian} CRPS skill score values of each of 
-the 15 observation stations, separated by method and distribution. Left-censored 
-at zero Gaussian is chosen as reference distribution.}
+\caption{\label{fig:box_crpss_gaussian} CRPS skill score (higher = better)
+for hurdle Gaussian and censored logistic distributions in reference to the
+censored Gaussian distribution. Lines correspond to the 15 stations and are
+aggregated in the boxplot. The y-axis range is limited
+to $-10$\%, lower values for the two GAMLSS models are not displayed.}
 \end{figure}
@@
 \setkeys{Gin}{width=0.99\textwidth}
-<<results_box_crpss_EMOS, fig=TRUE, echo=FALSE, height=3, width=7>>=
+<<results_box_crpss_EMOS, fig=TRUE, echo=FALSE, height=3, width=7.2>>=
 # crps skill score by method (reference: EMOS)
@@
 s_gaussian <- matrix(ncol = 3, nrow = 15)
-s_gaussian[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "gaussian", "crps_ss_method"]
-s_gaussian[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "gaussian", "crps_ss_method"]
-s_gaussian[,3] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "gaussian", "crps_ss_method"]
+s_gaussian[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "cgaussian", "crps_ss_method"]
+s_gaussian[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "cgaussian", "crps_ss_method"]
+s_gaussian[,3] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "cgaussian", "crps_ss_method"]
 
@@
 s_logistic <- matrix(ncol = 3, nrow = 15)
-s_logistic[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "logistic", "crps_ss_method"]
-s_logistic[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "logistic", "crps_ss_method"]
-s_logistic[,3] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "logistic", "crps_ss_method"]
+s_logistic[,1] <- means_sel[means_sel$method == "Distributional forest" & means_sel$distribution == "clogistic", "crps_ss_method"]
+s_logistic[,2] <- means_sel[means_sel$method == "Prespecified GAMLSS" & means_sel$distribution == "clogistic", "crps_ss_method"]
+s_logistic[,3] <- means_sel[means_sel$method == "Boosted GAMLSS" & means_sel$distribution == "clogistic", "crps_ss_method"]
 
@@
         col = gray(0.5, alpha = 0.2),
-        lty = 1, axes = FALSE,  main = "gaussian", ylim = c(-0.12,0.23), 
+        lty = 1, axes = FALSE,  main = "cgaussian", ylim = c(-0.12,0.23), 
         xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
@@
 #boxplot(crps_ss_method~method, data = means_sel, 
-#        subset = (distribution == "gaussian"), main = "gaussian",
+#        subset = (distribution == "cgaussian"), main = "cgaussian",
 #        ylim = c(-0.12,0.23), axes = FALSE, ylab = "CRPS skill score")
@@
         col = gray(0.5, alpha = 0.2),
-        lty = 1, axes = FALSE,  main = "logistic", ylim = c(-0.12,0.23), 
+        lty = 1, axes = FALSE,  main = "clogistic", ylim = c(-0.12,0.23), 
         xlab = "", ylab = "", xlim = c(0.5, 3.5))
@@
 #boxplot(crps_ss_method~method, data = means_sel, 
-#        subset = (distribution == "logistic"), main = "logistic",
+#        subset = (distribution == "clogistic"), main = "clogistic",
 #        ylim = c(-0.12,0.23), axes = FALSE)
@@
 \setkeys{Gin}{width=0.99\textwidth}
-<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7>>=
+<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7.2>>=
 
@@
 c_gaussian <- matrix(ncol = 4, nrow = 15)
-c_gaussian[,1] <- means[means$method == "Distributional forest" & means$distribution == "gaussian", "crps"]
-c_gaussian[,2] <- means[means$method == "Prespecified GAMLSS" & means$distribution == "gaussian", "crps"]
-c_gaussian[,3] <- means[means$method == "Boosted GAMLSS" & means$distribution == "gaussian", "crps"]
-c_gaussian[,4] <- means[means$method == "EMOS" & means$distribution == "gaussian", "crps"]
+c_gaussian[,1] <- means[means$method == "Distributional forest" & means$distribution == "cgaussian", "crps"]
+c_gaussian[,2] <- means[means$method == "Prespecified GAMLSS" & means$distribution == "cgaussian", "crps"]
+c_gaussian[,3] <- means[means$method == "Boosted GAMLSS" & means$distribution == "cgaussian", "crps"]
+c_gaussian[,4] <- means[means$method == "EMOS" & means$distribution == "cgaussian", "crps"]
 
@@
 c_logistic <- matrix(ncol = 4, nrow = 15)
-c_logistic[,1] <- means[means$method == "Distributional forest" & means$distribution == "logistic", "crps"]
-c_logistic[,2] <- means[means$method == "Prespecified GAMLSS" & means$distribution == "logistic", "crps"]
-c_logistic[,3] <- means[means$method == "Boosted GAMLSS" & means$distribution == "logistic", "crps"]
-c_logistic[,4] <- means[means$method == "EMOS" & means$distribution == "logistic", "crps"]
+c_logistic[,1] <- means[means$method == "Distributional forest" & means$distribution == "clogistic", "crps"]
+c_logistic[,2] <- means[means$method == "Prespecified GAMLSS" & means$distribution == "clogistic", "crps"]
+c_logistic[,3] <- means[means$method == "Boosted GAMLSS" & means$distribution == "clogistic", "crps"]
+c_logistic[,4] <- means[means$method == "EMOS" & means$distribution == "clogistic", "crps"]
 
@@
         col = gray(0.5, alpha = 0.2),
-        lty = 1, axes = FALSE,  main = "gaussian", 
+        lty = 1, axes = FALSE,  main = "cgaussian", 
         xlab = "", ylab = "CRPS", xlim = c(0.5, 4.5),
@@
 boxplot(c_gaussian, add = TRUE, col = "transparent", axes = FALSE)
-#boxplot(crps~method, data = means, subset = distribution == "gaussian", main = "gaussian",
+#boxplot(crps~method, data = means, subset = distribution == "cgaussian", main = "cgaussian",
 #        ylim = c(0.65,1.15), axes = FALSE, ylab = "CRPS")
@@
         col = gray(0.5, alpha = 0.2),
-        lty = 1, axes = FALSE,  main = "logistic", 
+        lty = 1, axes = FALSE,  main = "clogistic", 
         xlab = "", ylab = "CRPS", xlim = c(0.5, 4.5),
@@
 boxplot(c_logistic, add = TRUE, col = "transparent", axes = FALSE)
-#boxplot(crps~method, data = means, subset = distribution == "logistic", main = "logistic",
+#boxplot(crps~method, data = means, subset = distribution == "clogistic", main = "clogistic",
 #        ylim = c(0.65,1.15), las = 2, axes = FALSE)

### 2018-09-29 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
more work on supplements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2299 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 42375fce433cf91b5ec4e546daaa344ee5df0bc4

@@
 \item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
-Standard ensemble model output statistics uses the ensemble 
+Ensemble model output statistics uses the ensemble 
 mean of total precipitation as regressor in the location submodel 

### 2018-09-29 pkg/disttree/inst/draft/draft.Rnw
small revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2300 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6deaeee5a1598e0c3e75ed98600da222164f22f5

@@
 at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
-Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}}
+Observations are modelled by a left-censored at 0 distribution. The corresponding predicted point mass is shown at the censoring point (0).}}
 \end{figure}
@@
 \caption{\label{fig:axams-crps}CRPS skill score from the 10 times 
-7-fold cross validation at station Axams (1985--2012). The horizontal orange 
+7-fold cross-validation at station Axams (1985--2012). The horizontal orange 
 line pertains to the reference model EMOS.}
@@
 If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
-is chosen as splitting variable. \hl{Otherwise no further split is made is in this node.}
+is chosen as splitting variable. \hl{Otherwise no further split is made 
+in this node as the stopping criterion of no $p$-values being below the 
+significance level is fulfilled.}
 

### 2018-09-29 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
finished supplement improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2301 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 655ba80a0f84d600cb1a8ca78549694632c2781d

@@
 for hurdle Gaussian and censored logistic distributions in reference to the
-censored Gaussian distribution. Lines correspond to the 15 stations and are
+censored Gaussian distribution (orange horizontal line). Gray lines correspond to the 15~stations and are
 aggregated in the boxplot. The y-axis range is limited
@@
 \end{figure}
+
+\subsubsection*{FIXME: FURTHER GRAPHICS?}
+
+\emph{The same data already shown in Figures~\ref{fig:lattice} and~\ref{fig:box_crpss_gaussian}
+can be displayed in further ways. However, these do not really add much and hence should probably
+be omitted.}
 
@@
 
-\begin{figure}
+\begin{figure}[t!]
 \setkeys{Gin}{width=0.99\textwidth}
@@
 
-As explained above, the parameter $\nu$ is modeled separately from $\mu$ and $\sigma$ 
-in the GAMLSS framework which is why two separate models can be built. For the forest model
-an alternative approach would be to build just one model employing the three-parametric
-distribution family. However, testing both versions of the forest model in this application 
-study shows that they both perform almost equally. Figure~\ref{fig:box_crpss_forests} shows
-the CRPS skill score of the three-parametric forest model over the considered 15 observation
-stations with the two-part forest model as reference. Even though this alternative version
-of the forest model seems to perform better on average, according to the median this 
-improvement is not even one percent, hence neglectably small. Therefore, the two forest 
-versions can be stated to perform almost equally well.
+\section{Discussion}
 
-\begin{figure}
-\center
+The comparison of the zero-censored Gaussian models from the main manuscript with the
+alternative distributional specifications in this supplement reveals that a Gaussian hurdle
+may lead to small improvements. However, especially for EMOS and the boosted GAMLSS model
+the outcome is also considerably more variable. Therefore, overall there is no convincing
+enough improvement over the zero-censored Gaussian model which is therefore the focus of
+the main manuscript. Also, the distributional forest appears to be more robust against
+distributional (mis-)specifications compared to the other models.
+
+As a final consideration, we briefly remark that the distributional forest also allows
+for a different hurdle model specification. Previously, we argued that for GAMLSS/EMOS
+the log-likelihood of the hurdle model can be separated additively and hence the maximum
+likelihood fit can be obtained by fitting two separate models for $\nu$ on one hand and
+$\mu$ and $\sigma$ on the other. And therefore the same two-part approach was also adopted
+for the distributional forests. However, an additional alternative is to fit a single
+distributional forest based on the three-parameter log-likelihood from Equation~\ref{eq:hurdle}.
+When all three parameters change simultaneously across the covariates, then this specification
+can be expected to perform somewhat better. Whereas when different variables/effects drive
+the parameters of the model, then the two-part approach implemented in this supplement
+should perform somewhat better. In this particular case study, though, neither approach
+performs clearly better as Figure~\ref{fig:box_crpss_forests} reveals. The figure depicts
+CRPS skill scores for the 15~stations of the one-part hurdle models with the two-part
+hurdle model as the reference. Clearly, both hurdle specifications perform very
+similarly with skill scores being mostly below 1\%.
+
+\begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.5\textwidth}
@@
 abline(h = 0, col = pal[5], lwd = 2)
-axis(1, 0:2, c("","Three-parametric forest", ""), las=1)
+axis(1, 0:2, c("","One-part vs. two-part hurdle forest", ""), las=1)
 @
-\caption{\label{fig:box_crpss_forests}CRPS skill score values of each of 
-the 15 observation stations for the three-parametric forest model with 
-the two-part forest model as reference model.}
+\caption{\label{fig:box_crpss_forests}CRPS skill score (higher = better) for the one-part
+hurdle distributional forest in reference to the two part hurdle distributional forest
+(orange horizontal line).
+}
 \end{figure}
 
-\newpage
-%bibliographystyle{jss}
 \bibliography{ref_supplement.bib}

### 2018-09-30 pkg/disttree/inst/draft/rejoinder.tex
first draft of rejoinder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2302 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3ccf8a68000484f182ed31433f38867e2d70f113

\documentclass[american,foldmarks=false,noconfig]{uibklttr}
%\documentclass{article}
\usepackage{sfmath,hyperref,color,enumitem}

\let\code=\texttt
\let\pkg=\emph
\let\proglang=\textsf
\newenvironment{review}{\fontshape{\itdefault}\fontseries{\bfdefault} \selectfont \smallskip}{\par}

\setkomavar{signature}{Lisa Schlosser}

\makeatletter
\@setplength{refvpos}{75mm}
\makeatother
\setlength{\textheight}{25cm}
\definecolor{quotecolor}{rgb}{0.4,0.4,0.5}

\newenvironment{reto}{
    \begin{color}{uibkorange}
    \textbf{Reto:~}
        \itshape
}{
    \end{color}
}
\usepackage{amsmath}

\begin{document}
\begin{letter}{
Nicoleta Serban\\
Area Editor\\
The Annals of Applied Statistics}

\setkomavar{subject}{Revision of AOAS1804-015: Distributional 
regression forests for probabilistic precipitation forecasting 
in complex terrain}

\opening{Dear Professor Serban,}

please find attached the revised version of our manuscript 
``Distributional regression forests for probabilistic precipitation 
forecasting in complex terrain'' (AOAS1804-015).

Thank you very much for the constructive and helpful feedback from you, 
the associate editor and the two referees. The manuscript has been 
improved and additional results haven been obtained from further evaluations.

The most important changes are:
\begin{itemize}

\item The more extensive evaluations for station Axams have been 
carried out for 14 other observation stations. Results are 
provided in an online supplement which comes with our revised manuscript.

\item The evaluations performed for all 95 observation stations 
have been carried out under different distribution assumptions 
for the response variable in order to test for robustness against
distributional misspecifications. Results are provided in an 
online supplement.

\item The mathematical notation in the Appendix is now more precise.

\end{itemize}

All changes and additons are explained in much more detail in the
point-to-point reply on the next pages.

\closing{Best regards,
\vspace*{1cm}
}

\vspace*{-2.8cm}
\hspace*{-0.3cm}signature %\includegraphics{zeileis-signature}

\end{letter}

\newpage

\textbf{\LARGE Associate Editor}

\begin{review}
This manuscript proposes a framework for distributional regression 
forests which combine the idea of random forests with distributional 
regression in order to obtain a flexible class of forecast models 
where the full forecast distribution depends on the predictors. 
This new approach is demonstrated in the context of probabilistic 
precipitation forecasting where it is used to construct full 
predictive distributions based on the output of an ensemble 
weather prediction model. The paper is very well written, the 
proposed methodology is useful, and the data example used to 
illustrate this method is practically relevant and discussed 
thoroughly.
\end{review}

Thank you very much for your advice and the positive feedback, 
it is highly appreciated!

\begin{review}
A typical limitation of very flexible modeling approaches like 
the one proposed here is that a sufficiently large sample is 
required to fit the model. Can the authors provide some 
guidelines as to how large the training sample needs to be for 
this method to be effective? Does the stopping criterion 
guarantee that the model is not overfitted? 
Is it obvious how to define the specific stopping criterion and 
if not, can general guidelines for doing so be provided?
\end{review}

It is difficult to set a minimal size of the learning data for
the method to be effective, however, the distributional forest
performed quite well in different applications with at least
50 observations. 
The stopping criteria that are based on the number of 
observations in a node need to be adapted to the size of the
learning data, especially if the data set is small.
For example, in the application presented in this paper the 
learning data consists of 744 observations and each tree is built
on a subsample of around 470 observations (63.2 percent, as 
this is the mean fraction for N-out-of-N bootstrapping with 
replacement and therefore commonly used for subsampling
in forest models). 
The stopping criteria are set such that no split will be performed 
if there are either less than 50 observations 
in the considered node or if there were less than 20 observations in 
one of the resulting subgroups of the considered possible split.
(This information is now also added in Section~3.2., 'Distributional forest').
Generally, these values (and also the significance level for
the statistical tests in the tree algorithm) should be kept 
rather low for a forest model as it is desirable to grow very 
large trees. Therefore, each of the trees in the forest will 
most likely overfit the subsample it is learned on, however, this 
is evened out by combining them to a forest model (with an adequate 
number of trees). For that reason, setting the stopping criteria too 
low does not lead to a decrease in goodness of fit, however, the 
computational costs might increase, especially for large data sets. 
On the other hand, if they are too high, splits might not be detected 
such that the corresponding effects are not captured by the model.

\textit{FIX ME: too long?}

\begin{review}
The equations are not numbered but the authors refer to equations 
by number (e.g. bottom of p7)
\end{review}

We apologize for the confusion the missing equation numbers 
might have caused. They are now added.

\begin{review}
In the last section where computational details are discussed, 
can you also provide some information on the computational cost? 
In a weather forecasting contest like the one discussed here, 
forecasts are often provided on a high-resolution grid, 
so it would be important to understand if the approach 
proposed here is computationally affordable. This is not clear 
to me because the nearest neighbor weights are specific 
to each new observation, so in a forecasting context these 
computations have to be done in realtime.
\end{review}

We have measured fitting time and prediction time in our 
application. For single predictions the distributional forest 
can keep up with the other models. However, as you already 
stated correctly, each prediction made by the forest is based 
on an observation-specific weight matrix and therefore has to 
be calculated separately. For that reason prediction time for a 
set of observations is clearly higher for the forest model than 
the GAMLSS models or the EMOS. The exact results and times can 
be found in an online supplement.

\begin{review}
abstract: the acronym 'GAMLSS' should be spelled out the first 
time it is used
\end{review}

\begin{review}
p3: I'm not sure if people outside the weather forecasting 
community is familiar with the notion of an 'ensemble', so 
this should be explained briefly
\end{review}

\begin{reto}
Wo baut man das am besten ein?
\end{reto}

\begin{review}
p7: What is meant by 'domain knowledge'?
\end{review}

With `domain knowledge' we are referring to expert knowledge 
in the field of the considered application. 
If variable selection and further model specifications have 
to be made in advance this additional knowledge is required 
in order to make reasonable decisions.

\begin{reto}
Im paper entsprechend adressieren (tbd).
\end{reto}

\begin{review}
p14: Was there any particular rationale for choosing Axams 
as the station to be studied in detail
\end{review}

Axams was chosen as the observation station being closest 
to Innsbruck, the capital of Tyrol and the work place of 
three of the authors, but not based on any other particular 
features. Furthermore, it has been checked, whether it 
is an adequate representative of all stations regarding 
the results of our evaluations. This is now also 
confirmed by further analysis in the online supplement which
comes along with our manuscript.

\begin{review}
Fig. 4: Are these the plots for residuals of precipitation 
at the original or at the transformed scale?
\end{review}

These are the residuals of precipitation on the transformed scale.

\begin{review}
p17: Is there any explanation why the prespecified GAMLSS 
method is the best-performing method in East Tyrol? Any 
explanation why the distributional regression forests 
performed less well than the simple EMOS method in this area?
\end{review}

East Tyrol is located south of the main Alpine Ridge and exhibits different
climatological characteristics than the part north of the Alpine Ridge.  It
seems that the choice of covariates and their interactions we made for the
prespecified GAMLSS method is particularly adequate for the region of East
Tyrol for this specific analysis. However, the advantages of the prespecified
GAMLSS models in this particular setting (training: 1985--2008, validation:
2009--2012) even out in a cross-validation framework for most of the stations
in East Tyrol. These results are presented in the online supplement which comes
along with our manuscript showing that the forest performs equally well as the
prespecified GAMLSS and better than the EMOS at most of the stations.

\newpage

\textbf{\LARGE Reviewer 1}

\bigskip

\textbf{General comments:}

\begin{review}
The article presents a new parametric methodology, called 
distributional forests, to estimate the probability 
distribution of some random variables, based on information 
from covariates. The methodology is new in that it mixes the
automatic variable selection properties of random forests 
with a parametric assumption on the forecast distribution. 
This allows to keep the interpretability of parametric 
distributions with the flexibility of random forests in 
dealing with local fitting, non-linearity, interactions 
and variable selection.

This approach may be used in applied statistics, as 
illustrated with its application to forecasting daily 
precipitation amount in the mountainous region of Tyrol, 
Austria. The mountains make it very difficult for numerical 
weather prediction models to forecast local precipitation. 
Statistical post-processing of weather forecasts allows 
downscaling the prediction.

The paper is well written and is a pleasant reading. 
The methodology is introduced clearly and well grounded 
in statistical practices, but some points need to be 
made clearer (see later).

The article shows on the Tyrol rain forecasts that the 
proposed method performs better than three other methods, 
among which is the current state-of-the-art in statistical 
post-processing in meteorology (EMOS). Thus distributional
forests may be added to the several statistical post-processing 
techniques, as an efficient and easy-to-use parametric 
probabilistic forecast.

I recommend the paper to be published after the following 
minor revisions have been taken into account.
\end{review}

Thank you very much for the positive and constructive 
comments regarding the manuscript. These helped to 
substantially improve the manuscript.

\bigskip

\textbf{Specific comments:}

\begin{review}
1. {\color{quotecolor} page 2: exceedence}\\
This spelling is not wrong, but less common than ``exceedance''.
Please change into \textit{exceedance}
\end{review}

\begin{review}
2. {\color{quotecolor}\textit{page 2:} 10$\%$ to 90\% quantile}\\
Please change into \textit{10\% to 90\% quantiles}
\end{review}

\begin{review}
3. {\color{quotecolor}\textit{page 5:} Fig 3, caption: Observations 
are left-censored at 0}\\
Since observations are precipitation amount, they are truncated, not 
left-censored (which may let think that they could be negative). 
Please change into \textit{Observations are truncated at 0}
\end{review}

You are right, the observations can not be negative and are 
therefore left-truncated at 0 and not left-censored (we have adapted 
this in the caption of Figure~3). However, due to days without 
rain there is a naturally given point mass at zero in precipitation 
data, similar to the point mass that results from censoring 
possibly negative data.
Thus, we use the concept of \textit{censoring} which allows us to directly
account for the large fraction of observations at zero and the non-negative
part of the observations at once.
The online supplement which comes along with our manuscript includes the
results for a hurdle model where the occurrence of precipitation is addressed
using a binary model while the non-negative part is
modeled by a \textit{truncated} Gaussian distribution.

\begin{review}
4. {\color{quotecolor}\textit{page 5:} Fig 3, caption: censoring point}\\
Please change into \textit{truncation point}
\end{review}

Since we are applying a censored distribution as explained above we 
prefer using the term censoring point, even though the data itself 
is not censored.

\begin{review}
5. {\color{quotecolor}\textit{page 5:} Fig 3, legend: Point mass at 
censoring point}\\
Please change into \textit{Point mass at truncation point}
\end{review}

See points 3 and 4.

\begin{review}
6. {\color{quotecolor}\textit{page 5:} Fig 3, x label: $mm^{(1/1.6)}$/24\textit{h}}\\
This is not a very common unit, which make the figure not easily 
interpretable. Furthermore, since you have not yet said that 
observations have been power transformed, this may be surprising 
here at first reading. Please redraw this figure with 
observations in $mm$/24$h$.
\end{review}

Thank you for the advice, the figure is now redrawn on the scale $mm$/24$h$.

\begin{review}
7. {\color{quotecolor}\textit{page 7:} see Equation 2.1}\\
Your equations have no numbering. Please add one.
\end{review}

We apologize for the confusion the missing equation numbers 
might have caused. They are now added.

\begin{review}
8. {\color{quotecolor}\textit{page 8:} $\hat{\Theta}(z) = \cdots$}\\
This expression may lead to think that the parameter is estimated 
for each new prediction. If I am right, this is not the case, 
since the weights $w_i^{tree}$ are either $0$ or $1$. Thus the 
estimated parameter vector is the same as the one for the 
observations in the same segments as the new observation.
Please make this clear in the text.
\end{review}

For each new observation $z$ the corresponding 
parameter vector $\hat{\Theta}(z)$ can be calculated following 
the given formula. But you are right, one can also just evaluate 
the predicted terminal node for $z$ if the predicted parameter
vector for this node is already known as there is one unique 
parameter vector for all observations in the same terminal node.
This is now also explained at the end of Section~2.2..

\begin{review}
9. {\color{quotecolor}\textit{page 9:} Each of the distributional trees}\\
Please change into \textit{Each of the distributional tree}
\end{review}

\textit{FIX ME: ?}

\begin{review}
10. {\color{quotecolor}\textit{page 9:} $w_i^{forest}$}\\
The normalization with $|B_b^t|$ makes the weight $w_i^{forest}$ 
take their value in $[0; 1]$. If I am not wrong, this means the 
parameter vector has to be estimated for each new prediction. 
How long does it take? For operational applications, the time 
to produce post-processed forecast may be a constraint. It may 
be interesting to compare the prediction time with distributional 
forest and with GAMLSS, since you claim in the abstract that
GAMLSS are computationally demanding. Please add some information
about the computation time to train your models and produce predictions.
\end{review}

Yes, in the forest model the parameter vector has to be estimated 
separately for each new prediction as it is based on an 
observation-specific weight matrix. For that reason, making 
predictions for a set of new observations is clearly more 
time consuming for the forest model than the GAMLSS models 
or the EMOS model. However, the time needed to make one single 
prediction is similar for all methods. The exact results 
concerning fitting and prediction times can be found in an online 
supplement.

\begin{review}
11. {\color{quotecolor}\textit{page 10:} While this approach 
alone is already highly effective in the plains, it typically 
does not perform as well in complex terrain due to unresolved 
effects in the NWP system. For example, in the Tyrolean 
Alps -- considered in the following case study -- the NWP grid 
cells of $50 \times 50$ km$^2$ are too coarse to capture single 
mountains, narrow valleys, etc.}\\
Do you have references about this lack of performance of NWP 
in complex terrains? If so, please add them.
\end{review}

\textit{TO DO}

\begin{reto}
Es gibt da ein klassisches MOS-Paper in welchem Postprocessing
mit einem Globalmodell und einem Lokalmodell gemacht werden.
Das ist nicht die Referenz die der Reviewer will.
Das Problem ist, dass es explizit dazu nix gibt. Das ist einfach
jedem von uns klar. M\"uller zeigt aber in seinen Figures (f\"ur
Wind und Temperatur), dass ein Modell mit h\"oherer Aufl\"osung
einen geringeren Fehler aufweist und sagt folgendes:

``\textit{The largest differences are found in the raw
model forecast for stations in complex terrain. For wind
speed, the increase in resolution is more beneficial than
for temperature but is again negligible for MOS forecasts.
[\dots]
A higher resolution will
further reduce the model bias, especially in complex
terrain, but this can also be achieved with statistical
postprocessing if observations are available. Furthermore,
a postprocessed temperature forecast at 12-km
resolution outperforms a raw forecast at 3-km, regardless
of the complexity of the terrain}''.

\begin{verbatim}
@article{mueller2011,
    author = {M{\"u}ller, M. D.},
    title = {Effects of Model Resolution and Statistical Postprocessing on Shelter Temperature and Wind Forecasts},
    journal = {Journal of Applied Meteorology and Climatology},
    volume = {50},
    number = {8},
    pages = {1627--1636},
    year = {2011},
    doi = {10.1175/2011JAMC2615.1},
}
\end{verbatim}

Oder Meyers and Steenburgh (2013) der aufzeigt, dass auch der Mensch sich mit
der Wettervorhersage im Gebirge schwer tut bzw. dort die groessten Challenges
zu finden sind mit all den lokalen und kleinraeumigen Eigenschaften und Einfluessen.
Habe das Buch grad nicht rumliegen, ist wohl aber auch nicht zwingend notwendig.

\begin{verbatim}
@Inbook{meyers2013,
    author = {Meyers, Michael P. and Steenburgh, W. James},
    editor = {Chow, Fotini K. and De Wekker, Stephan F.J. and Snyder, Bradley J.},
    title = {Mountain Weather Prediction: Phenomenological Challenges and Forecast Methodology},
    bookTitle = {Mountain Weather Research and Forecasting: Recent Progress and Current Challenges},
    year = {2013},
    publisher = {Springer Netherlands},
    pages = {1--34},
    isbn = {978-94-007-4098-3},
    doi = {10.1007/978-94-007-4098-3_1},
}
\end{verbatim}
\end{reto}

\begin{review}
12. {\color{quotecolor}\textit{page 11:} Table 1, caption}\\
Please move the caption above the table.
\end{review}

\begin{review}
13. {\color{quotecolor}\textit{page 12:} Parameters are estimated 
by adaptive local likelihood based on the forest weights as 
described in Section 2.}\\
Please explain what is your stopping criterion.
\end{review}

As stopping criteria the minimal number of observations to perform 
a split (\texttt{minsplit}) is set to 50,  the minimal number of 
observations in a terminal node (\texttt{minbucket}) is set to 20 
and the significance level for variable selection (\texttt{alpha}) 
is kept at the default value 1. We have added this information in 
the manuscript in Section~3.2., 'Distributional forest'.

\begin{review}
14. {\color{quotecolor}\textit{page 13:} Table 2, caption}\\
Please move the caption above the table.
\end{review}

\begin{review}
15. {\color{quotecolor}\textit{page 17:} Thus, while the covariates 
themselves are not surprising, selecting a GAMLSS with a particular 
combination of all the transformations would be much more challenging.}\\
Did you compare the most important covariates in distributional forests
with the ones in preselected GAMLSS? If so, is there some insightful
information you could add?
\end{review}

Figure 6 shows the variable importance of the distributional forest
for one specific case (similar for other sites and cross-folds) 
using a CRPS based variable importance measure. The subset of 
most important covariates is similar to the set of covariates 
used for the prespecified GAMLSS but not equivalent.
The use of different variations of the same covariate (cf., Table 2)
is not surprising as the covariates are often highly correlated.
Compared to the prespecified GAMLSS the forest seems to depict some
features via differences among the different variations of a specific
covariate (e.g., use \code{tp\_max}, \code{tp\_mean}, and \code{tp\_min}).
Unfortunately the analysis does not allow to get more insights
of (the importance of) interactions between different covariates.

\begin{reto}
Oder sowas \dots Oder g\"abe es daf\"ur tests in W\"aldern?
\end{reto}

\begin{review}
16. {\color{quotecolor}\textit{page 20:} If $Z_l$ is numeric 
then $v_l$ is simply the identity function $v_l(z_l^i) = z_l^i$. 
If $Z_l$ is a categorical variable with $H$ categories then 
$v_l(z_l^i) = e_H(z_l^i) = (I(z_j^i = 1), \dots , I(z_l^i = H))$ 
such that $v_l$ is a unit vector}\\
Based on this sentence, $v_l$ can be a vector or a scalar, 
which makes not very clear what $T_l$ is: is it a vector 
or a scalar? Please clarify.
\end{review}

Apologies for the unclear notation in Appendix A. This has been 
revised in the manuscript, also regarding your comments 17--20,
and we hope that it now clarifies the content.

Depending on whether the split variable $Z_l$ is a numeric 
or a categorical variable, $t_l$ is either a
$k$~-~or~$k \cdot H$~-~dimensional vector.

\begin{review}
17. {\color{quotecolor}\textit{page 20:} the observed 
multivariate linear statistic $t$}\\
What is $t$? Is it the vector with $m$ components $T_l$? 
Please clarify.
\end{review}

Apologies again for the confusing notation. Here we are 
actually referring to the linear statistic $t_l$ (instead of $t$) 
which is the observed value of $T_l$, the test statistic for 
the $l$-th split variable $Z_l$.

\begin{review}
18. {\color{quotecolor}\textit{page 20:} $c_{quad}(t, \mu, \Sigma)$}\\
$\mu$ and $\Sigma$ have not been defined. What is their relationship 
with $\mu_l$ and $\Sigma_l$? Please clarify.
\end{review}

As in the foregoing point this was just a general notation, but 
actually the variables corresponding to the $l$-th split variable 
are considered, i.e. $t_l$, $\mu_l$ and $\sigma_l$. Thank you for
pointing this out, now the missing indices are set.

\begin{review}
19. {\color{quotecolor}\textit{page 21:} it has to be assessed 
whether any of the resulting p-values is beneath the selected 
significance level. If so, the partitioning variable $Z_l$
with the lowest p-value is chosen as splitting variable.}\\
What if no p-value is beneath the selected significance level. 
Is it a stopping criterion? Please clarify.
\end{review}

Yes, one stopping criterion is that no p-value is beneath the 
selected significance level. Therefore, no further 
split is performed in the considered node in this case.

\begin{review}
20. {\color{quotecolor}\textit{page 21:} $B_{1r}$ is the first 
of the two new subgroups that are defined by splitting in split 
point $r$ of variable $Z_{l^{\ast}}$}\\
What do you mean by ``first of the two new subgroups''? Is it 
the subgroup with values $Z_{l^{\ast}} > r$? I don't see how 
such an ordering can be achieved if $Z_{l^{\ast}}$ is an 
unordered categorical variable. Please clarify how the two
subgroups are ordered.
\end{review}

Again, we are sorry for a confusing explanation. The two subgroups 
are not ordered which is also not necessary as only the sum of 
scores over the two subgroups is considered for the linear 
statistic $T_{l^{\ast}}^{qr}$.

\newpage

\textbf{\LARGE Reviewer 2}

\begin{review}
The authors propose an extension of non-homogeneous Gaussian models 
for the use in weather forecasting based on parametric distributional 
models (GAMLSS). While the latter have proved to be applicable to a 
wide range of empirical questions, the authors focus on precipitation 
forecasts in the Alps. The approach combines a distributional model 
with a random forest, which they call distributional random forests, 
in order to include variable selection and interactions or 
non-linearities of predictors. I have a number of major concerns 
given in the following. My impression is that the method is rather
an extension of existing methodology on random forests to censored 
normal problems without important with a lack in discussing 
advantages and disadvantages of distributional regression models 
and only a rather minor improvement over certain competitors.
\end{review}

Thank you very much for your feedback and helpful suggestions. We hope 
the improvements we made on the paper can clear your doubts and concerns.

\bigskip

\textbf{Main comments:}

\begin{review}
1. It is obvious in distributional regression that not only the 
variables influence the predictive performance but rather that 
the exact distributional assumption has a big impact on both the 
fit and the prediction. The authors do not consider any other
than the censored normal distribution, which I found odd. 
In earlier references, they used the same distribution but I 
did not find any justification why not to use a truncated normal, 
a zero-adjusted model or a continuous part for $y > 0$ other
than the normal. In any distributional model, checking the 
assumption on the error density is a crucial step which in 
my opinion should not be ignored or assumed not to be a problem.
\end{review}

Thank you for pointing this out. We have now rerun the exact 
same evaluations as presented in the paper with different 
distributional assumptions, including a logistic model and 
a hurdle model with a truncated part. The exact results can be 
found in an online supplement.

\begin{review}
2. The response and some predictors are power transformations 
of originally measured variables like total precipitation as 
it is known that skewness is usually present. Why? As far as I 
understand, one major advantage of distributional models is
that the response does not have to be normal but to model scale, 
skewness and the shape in general as a function of the covariates. 
By applying the transformation in advance and hence marginally, 
the dependence of covariates on further moments of the response 
is assumed to be not existing and simply ignored.
\end{review}

The skewness results from the characteristics of precipitation sums. 
Most of the time no or only moderate precipitation amounts are forecasted 
and observed. In some situations (e.g., days with thunderstorms, cold and 
warm fronts, \dots) large amounts of precipitation can be observed at 
specific sites. This is often also depict by the ensemble prediction 
system where a subset of all members predicts rather large amounts 
(compared to the overall ensemble mean or median). This yields strongly 
skewed ensemble forecasts which has to be addressed in some sort.\\
As you mention one way would be to directly apply a
distributional family with an additional parameter to capture
skewness. Another frequently used approach is to transform the 
data to remove large parts of the skewness. The latter has the 
advantage that the model assumptions get simpler as no estimate 
for the skewness parameter is required. 
Moreover, applying a transformation together with a censored 
at zero Gaussian distribution allows for a better comparison 
to the other models as they can all be built under the same 
settings.
In this study we are thus following Stauffer et al. (2017) 
using the power transformation which turned out to be optimal 
for a similar application for the very same study area.

\begin{review}
3. Is a random forest really the right thing to do? In the 
end, the analyst or meteorologist does not obtain an 
interpretable model and no explanation which variable 
affects the distribution in which direction is possible. 
I think, the reason that focussing on predictions should 
be made clearer and well justified.
\end{review}

This is the trade-off between interpretability and predictive 
skill. Regression-type models might be easier to interpret, 
however, if one is mainly interested in predictive skill and
a good performing fully automated weather forecasting tool the 
interpretability is not implicitly needed in some cases 
(machine learning performance vs. interpretability).
We adapted our manuscript as we definitely want to emphasize that 
in our application the focus is mainly on prediction but not on 
interpretation. We hope we could point this out clearly.

\begin{review}
4. The method itself is hard to follow in particular since 
authors reference to equations without using equation numbers. 
The part on random forests for instance is quite a black box 
without looking into a number of earlier works of some of 
the authors. From an applied perspective, there should be a 
better motivation and connection of the methodology and the 
data application.
\end{review}

We apologize for the confusion the missing equation numbers 
might have caused. They are now added.

\textit{FIX ME: comment on 'there should be a 
better motivation and connection of the methodology and the 
data application'?}

\begin{review}
5. In the introduction it is said that ``while requiring no 
expert knowledge for the model specification''. I do not think 
it is true given that the authors do not consider a possible 
distributional misspecification at all. In addition, If I am 
not wrong, the functional forms in the distributional 
regression models do not have to be prespecified. Applying a 
mixed model type representation it should be possible to
have for instance interaction surfaces separated into the pure 
interaction effects and the main effects. For instance, 
Wood et al. (2013) discuss a generic approach for defining 
well-identified and interpretable effect decompositions in 
tensor product interactions again based on the mixed model 
representation of penalised model components, while Goicoa 
et al. (2017); Ugarte et al. (2017) determine the identifiability
constraints from the null space of the penalties associated 
with the spatio-temporal interaction effect.
\end{review}

Thank you for your suggestions and the references. 
The prespecification of the distribution is expert knowledge in 
this context as in some of our previous publications we have been 
working on different response distributions and the censored normal 
distribution turned out to be one of the best performing. In this article 
we have not proven whether or not the chosen distribution is the overall
best for this specific application, however, the main part of the article 
is and should be the distributional forest and not the application itself. 
However, we are sure that, given our previous work, the selected methods 
and model assumptions are well performing and well suited for the practical 
example presented.
Moreover, as mentioned before we have tested robustness against 
distributional misspecification. The results show that applying a 
left-censored at zero logistic distribution or a Gaussian hurdle model 
instead of the left-censored at zero Gaussian distribution, as presented 
in the paper, does not lead to any significant improvements in goodness of 
fit of the applied models. The exact results are provided in an online 
supplement which comes with our manuscript.

\textit{FIX ME: refer to Wood, Goicoa, Ugarte}

\begin{review}
6. A simulation study designed according to the data set should 
be used in order to give an empirical evidence that the random 
forests are generally better able to perform predictions compared 
to the competing methods. The figures indicate that advantages 
for the chosen station are rather minor and one should keep in 
mind that GAMLSS and boosting yield interpretable results which 
the random forests do not do. also, what happens when choosing 
another station than Axams? The figure 8 indicates that the 
distributional forest does not perform better in all stations 
and I am wondering whether a meteorologist would really do a 
better job on a relative scale of precipitation.
\end{review}

You are right, there are clearly some stations where other models perform
better than the forest. However, at the majority of stations used in our study
it can at least keep up with them and therefore seems to provide a good and
reasonable alternative that is very easy to apply as no expert knowledge or
additional work in advance is required and the computational effort is kept
relatively small compared to a boosting method. In an online supplement the
same evaluations as for station Axams have been carried out for 14 other
stations, including some of those where the forest does not perform better than
the other models in the setting presented in Figure~8.  It shows that the
differences in performance even out in the cross-validation framework. There,
the forest performs at least equally well compared to the best of the other
models at most stations.

\textit{FIX ME: comments on simulation study? comments on use of
method by meteorologists? (last sentence)}

\begin{reto}
``[\dots] I am wondering whether a meteorologist would really do a
better job on a relative scale of precipitation.'' ist wohl eher eine
rethorische Frage und tut nix zur Sache wenn du mich fragst.
Unkommentiert lassen? Oder Z hat vielleicht eine Idee?
\end{reto}

\begin{review}
7. How are the folds chosen for performing the predictions? 
is it 4 subsequent years or randomly chosen 4 years out of 
the 28? I expect this to have a strong impact on the predictions 
and sensitivity with respect to this needs to be considered both 
in simulations and the real study.
\end{review}

The folds of the cross-validation are chosen randomly. The data 
set of 28 years is split randomly into 7 blocks of 4 years which 
are not necessarily successive. This is now also explained in the 
description of the cross-validation setup third paragraph of 
Section~3.3..

\begin{review}
8. I found the competing models to be chosen pretty unfair. 
First, for the prespecified GAMLSS, the authors use 
”relevant effects based on expert knowledge”. Why not all? 
Why not making a decision on some information criteria such as AIC/BIC? 
The latter would be done if the analyst decided to use a prespecified 
GAMLSS. I guess, the GAMLSS could improve a lot by this . 
Second, what does a basic EMOS exactly mean? Third, why not to include 
interactions in the boosting algorithm as well? this should be 
conceptually straight forward? Fourth, why are 100 trees chosen 
in the forest? Is this a a default value? May this number not also 
be dependent on the complexity of the model? Why not to choose 
the number of trees similar to what is done in boosting? this 
would also increase the computation time of the forest a lot.
\end{review}

We decided to evaluate two different versions of a GAMLSS: 
the prespecified GAMLSS where the covariates are selected 
in advance based on expert knowledge and the boosted GAMLSS
where all covariates are included. For the latter covariates 
are selected within the boosting algorithm. As stated by 
Hofner et al. (2016) the AIC-based variable selection 
methods implemented in the \textsf{R} package \textbf{gamlss} 
``can be unstable, especially when it comes to selecting 
possibly different sets of variables for multiple 
distribution parameters.''

By a `basic EMOS' we are referring to the linear distributional 
regression model based on the idea presented by Gneiting et al. (2005) 
where only the ensemble mean is included for the location parameter 
and the ensemble standard deviation for the scale parameter.

In the boosted GAMLSS we did not include interactions as this would
lead to an extremely high number of covariates that would be hardly
manageable. Even if we included only the pairwise interactions of all 
80 covariates we would end up with 3160 additional covariates.

% factorial(80)/factorial(80-2)/factorial(2)

\begin{reto}
K\"onnte man eventuell auch im Paper so erw\"ahnen?

\begin{equation}
    \frac{n!}{k! \cdot (n-k)!}~~\mbox{bei}~n=80~\mbox{und}~k=2~\mbox{die}~3160~\mbox{Kombinationen.}
\end{equation}
\end{reto}

By default the number of trees in a distributional forest
is set to 500. However, to keep computation time at a 
reasonable level we decided  to use only 100 trees. 
But of course, the more trees, the better the complexity of 
the model can be captured, especially smooth transitions can 
be modeled better by a higher number of trees.

\textit{FIX ME: comment on 'Why not to choose the number of trees 
similar to what is done in boosting? this would also increase the 
computation time of the forest a lot.'}

\bigskip

\textbf{Minor comments:}

\begin{review}
1. Given the length of the paper, I suggest to move the 
appendix to an online supplement.
\end{review}

\begin{review}
2. Does the EMOS method employ ensemble post-processing 
techniques? There are recent references using this in weather 
forecasting (Chen et al.; 2014)?
\end{review}

\textit{FIX ME: add formula for EMOS?}

\begin{reto}
Chen ist auf jeden Fall non-sense. Es gibt viele Papers die
sowas machen, die Frage ist, ob es 10 Referenzen braucht (unter
anderem macht das Jakob, Manu, ich, die Gneiting-Gruppe bzw.
Hamill-Gruppe hat auch was in die Richtung aber mit Gamma \dots).

Effizienter waere es wohl kurz das hier EMOS genannte Modell (was
eigentlich einfach ein non-homogeneous regression Modell ist)
im Appendix kurz hinzuschreiben?

    \begin{equation}
        \begin{split}
            y^{1/p} = \mathcal{N}_0(\mu, \sigma) \\
            \mu = \beta_0 + \beta_1 \cdot \bar{\mbox{tp}^{1/p}} + \mbox{split?} \\
            \log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{1/p}) + \mbox{split?}
        \end{split}
    \end{equation}

\dots damit k\"onnte man auch auf Kommentar 8 von Reviewer 2 reagieren.
\end{reto}

\begin{review}
3. Looking at the PIT in the appendix, they seem to provide a 
worse fit as compared to the quantile residuals. Also, both 
just give a graphical device. If the focus is on prediction 
then my impression is that one should rather plot PIT and 
residuals for the predictions.
\end{review}

\textit{TO DO}

\begin{reto}
Beide? Verstehe ich den Kollegen da richtig?
\end{reto}

\begin{review}
4. Figure 5: Which most important variables are included in the 
boosting algorithm? Why are the 10 most important covariates chosen?
\end{review}

Figure 6 shows the 10 most important covariates of the forest model
in one particular setting (station Axams, learned on 1985--2008). 
Variable importance has been calculated for all covariates, however,
we decided to present only the top 10 as all following covariates 
are on a rather low and similar level not providing further useful
information. But this top 10 list is not considered for any 
specification of the boosted GAMLSS for which all available 
covariates are included. Out of them, in the exact same setting 
(Axams, 1985--2008), the boosting algorithm selects 28 covariates 
for the location parameter, mainly related to precipitation (7) 
and temperature (11), and 46 covariates for the scale parameter, 
mainly related to precipitation (10), convective available potential 
energy (8) and temperature (15).
% mu: 7 tppow, 1 cape, 2 dswrf, 2 msl, 2 pwat, 3 tcolc, 11 temperature
% sigma: 10 tppow, 8 cape, 1 dswrf, 5 msl, 5 pwat, 2 tcolc, 15 temperature

% R> names(coef(gb, parameter = "mu"))
% [1] "bbs(tppow_min)"        "bbs(tppow_max)"        "bbs(tppow_mean1218)"  
% [4] "bbs(tppow_mean1824)"   "bbs(tppow_mean2430)"   "bbs(tppow_sprd0612)"  
% [7] "bbs(tppow_sprd2430)"   "bbs(capepow_sprd0612)" "bbs(dswrf_mean_mean)" 
%[10] "bbs(dswrf_sprd_mean)"  "bbs(msl_sprd_min)"     "bbs(msl_sprd_max)"    
%[13] "bbs(pwat_sprd_min)"    "bbs(pwat_sprd_max)"    "bbs(tcolc_mean_mean)" 
%[16] "bbs(tcolc_mean_max)"   "bbs(tcolc_sprd_mean)"  "bbs(t500_mean_min)"   
%[19] "bbs(t850_mean_min)"    "bbs(t500_sprd_mean)"   "bbs(t500_sprd_max)"   
%[22] "bbs(t700_sprd_mean)"   "bbs(t700_sprd_min)"    "bbs(t700_sprd_max)"   
%[25] "bbs(t850_sprd_min)"    "bbs(t850_sprd_max)"    "bbs(tdiff700850_min)" 
%[28] "bbs(tdiff500700_max)"

% R> names(coef(gb, parameter = "sigma"))
% [1] "bbs(tppow_mean)"       "bbs(tppow_sprd)"       "bbs(tppow_min)"       
% [4] "bbs(tppow_max)"        "bbs(tppow_mean1824)"   "bbs(tppow_mean2430)"  
% [7] "bbs(tppow_sprd0612)"   "bbs(tppow_sprd1218)"   "bbs(tppow_sprd1824)"  
%[10] "bbs(tppow_sprd2430)"   "bbs(capepow_sprd)"     "bbs(capepow_min)"     
%[13] "bbs(capepow_max)"      "bbs(capepow_mean1218)" "bbs(capepow_mean1224)"
%[16] "bbs(capepow_mean1230)" "bbs(capepow_sprd0612)" "bbs(capepow_sprd1218)"
%[19] "bbs(dswrf_sprd_mean)"  "bbs(msl_mean_mean)"    "bbs(msl_mean_max)"    
%[22] "bbs(msl_sprd_mean)"    "bbs(msl_sprd_min)"     "bbs(pwat_mean_mean)"  
%[25] "bbs(pwat_mean_min)"    "bbs(pwat_mean_max)"    "bbs(pwat_sprd_min)"   
%[28] "bbs(pwat_sprd_max)"    "bbs(tmax_mean_mean)"   "bbs(tmax_mean_min)"   
%[31] "bbs(tcolc_mean_min)"   "bbs(tcolc_sprd_max)"   "bbs(t500_mean_mean)"  
%[34] "bbs(t700_mean_mean)"   "bbs(t700_mean_max)"    "bbs(t850_mean_min)"   
%[37] "bbs(t850_mean_max)"    "bbs(t500_sprd_max)"    "bbs(t850_sprd_min)"   
%[40] "bbs(t850_sprd_max)"    "bbs(tdiff700850_min)"  "bbs(tdiff700850_max)" 
%[43] "bbs(tdiff500700_mean)" "bbs(tdiff500700_min)"  "bbs(tdiff500700_max)" 
%[46] "bbs(msl_diff)"

Applying the stability selection procedure by Meinshausen et al. (2010) and
Hofner et al. (2015) on the same model shows that the 10 covariates with 
maximal selection probability are six covariates based on precipitation and 
one each based on mean sea level pressure, downwards short wave radiation 
flux, temperature and precipitable water respectively.

% load("~/svn/partykit/pkg/disttree/inst/draft/Axams_24to4.rda")
% sgb <- stabsel(gb, cutoff = 0.6, q = 30)

% R> sgb$selected
% tppow_min.mu         tppow_max.mu    tppow_mean1824.mu   dswrf_mean_mean.mu 
%      3                    4                    7                   25 
% tppow_min.sigma      tppow_max.sigma tppow_sprd2430.sigma  pwat_mean_max.sigma 
%      83                   84                   92                  117 
% t700_mean_max.sigma   msl_diff.sigma 
%      138                  160

% R> head(sort(sgb$max, decreasing = TRUE), 10)
% tppow_max.mu    tppow_mean1824.mu      tppow_max.sigma  t700_mean_max.sigma 
%    1.00                 0.94                 0.91                 0.87 
% dswrf_mean_mean.mu  pwat_mean_max.sigma  tppow_min.sigma tppow_sprd2430.sigma 
%    0.85                 0.84                 0.81                 0.74 
% msl_diff.sigma      tppow_min.mu 
%    0.70                 0.63

% R> head(sort(rowMeans(sgb$phat), decreasing = TRUE), 10)
% tppow_max.mu    tppow_mean1824.mu  t700_mean_max.sigma      tppow_max.sigma 
%    0.99516              0.86018              0.84936              0.84857 
% tppow_min.sigma  pwat_mean_max.sigma   dswrf_mean_mean.mu tppow_sprd2430.sigma 
%    0.79779              0.75813              0.74074              0.65091 
% msl_diff.sigma         tppow_min.mu 
%    0.63928              0.55443

Overall it can be stated that the most important covariates of the boosted 
GAMLSS are similar to the ones of the distributional forest even though they
are not exactly the same ones.

\bigskip

\noindent{\large\textbf{References}}

\begin{itemize}[leftmargin=*]

\item[] Chen, J., Brissette, F. P. and Lhi, Z. (2014). Postprocessing 
of ensemble weather forecasts using a stochastic weather generator, 
\textit{Monthly Weather Review}.

\item[] Goicoa, T., Adin, A., Ugarte, M. D. and Hodges, J. S. (2017). 
In spatio-temporal disease mapping models, identifiability constraints 
affet PQL and INLA results, \textit{Stochastic Environmental Research 
and Risk Assessment} p. to appear.

\item[] Ugarte, M. D., Adin, A. and Goicoa, T. (2017). One-dimensional, 
two-dimensional, and three-dimensional B-splines to specify space-time 
interations in bayesian disease mapping: Model fitting and model 
identifiability, \textit{Spatial Statistics} p. to appear.

\item[] Wood, S. N., Scheipl, F. and Faraway, J. J. (2013). 
Straightforward intermediate rank tensor product smoothing in mixed models, 
\textit{Statistics and Computing} \textbf{23}: 341--360.

%%%%%%%%%
% references used in answers
\item[] Gneiting, T., Raftery, A. E., Westveld III, A. H. and Goldman, T. (2005). 
Calibrated Probabilistic Forecasting Using Ensemble Model Output Statistics and 
Minimum CRPS Estimation,
\textit{Monthly Weather Review} \textbf{133, 5}: 1098--1118.

\item[] Stauffer, R., Mayr, G. J., Messner, J. W., Umlauf, N. and Zeileis, A. (2017). 
Spatio-Temporal Precipitation Climatology over Complex Terrain Using a Censored Additive Regression Model,
\textit{International Journal of Climatology} \textbf{37, 7}: 3264--3275.

\item[] Hofner, B., Mayr, A. and Schmid, M. (2016). 
\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and 
Variable Selection in the {GAMLSS} Framework,
\textit{Journal of Statistical Software} \textbf{74, 1}: 1--31.

\item[] Hofner, B., Boccuto, L. and Goeker, M. (2015). 
Controlling false discoveries in high-dimensional situations: Boosting 
with stability selection,
\textit{BMC Bioinformatics} \textbf{16:144}.

\item[] Meinshausen, N. and Buehlmann, P. (2010). 
Stability selection,
\textit{Journal of the Royal Statistical Society, Series B} 
\textbf{72}: 417--473.

\end{itemize}

\end{document}

### 2018-10-01 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
added information on learning and testing data
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2303 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2b8fc790aec696d98018ff8a7425bdf5a51be78d

@@
 To assess how appropriate the distributional assumption (left-censored Gaussian distribution
-with censoring point zerol; or zero-censored for short) is, two questions are considered:
+with censoring point zero; or zero-censored for short) is, two questions are considered:
 (1)~Are the tails of the distribution captured appropriately by a Gaussian distribution or
@@
 compared with other commonly-used distributional regression models using the same
-data set. Table~2 of the manuscript provides
-an overview of all models considered and their specifications. As a reminder
-the approaches are briefly summarized below.
+data set. The models are learned on data from the first 24 years (1985--2008) and 
+evaluated on the following 4 years (2009--2012). Table~2 of the manuscript provides 
+an overview of all models considered and their specifications. As a reminder the 
+approaches are briefly summarized below.
 %

### 2018-10-03 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
added explanation of PIT histograms and residual QQ plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2304 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: aabc77b81fc38653de06e4077285ccca60181093

@@
 \item \emph{PIT histograms:} Similarly, the out-of-sample probability integral
-transform (PIT) in the single-split setting is amployed as an alternative
+transform (PIT) in the single-split setting is employed as an alternative
 goodness-of-fit visualization.
@@
 \end{itemize}
+
+The Resiudal QQ plots and the PIT histograms visualize the same values,
+however, on different scales. For each observation $y$ of the test sample 
+the corresponding PIT value is calculated by applying the probability function 
+$F(y; \hat{\mu}, \hat{\sigma})$
+with the estimated distribution parameters plugged in. The resulting values
+can either be plotted in a histogram (PIT histogram) or be plotted against
+the corresponding quantiles of a standard normal distribution (residual QQ plot)
+for a continuous variable $Y$. To account for the point mass at 0, for each 
+observation equal to 0 the corresponding PIT value is multiplied with a randomized
+value following a uniform distribution on \[0,1\]. In that way these values
+are apportioned on those bins of the histogram that would not contain any
+values because no observation is beneath 0.
 

### 2018-10-03 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
revision of discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2305 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ec2df34a22dfa64682a2d73fc15437c31eb03126

@@
 
-The comparison of the zero-censored Gaussian models from the main manuscript with the
-alternative distributional specifications in this supplement reveals that a Gaussian hurdle
-may lead to small improvements. However, especially for EMOS and the boosted GAMLSS model
-the outcome is also considerably more variable. Therefore, overall there is no convincing
-enough improvement over the zero-censored Gaussian model which is therefore the focus of
-the main manuscript. Also, the distributional forest appears to be more robust against
-distributional (mis-)specifications compared to the other models.
+This supplement shows that the zero-censored Gaussian models from the main
+manuscript and the alternative distributional specifications in this
+supplement perform at least similarly. For some combinations of models and
+stations, the Gaussian hurdle leads to small improvements on average but
+typically with more variability (especially for EMOS and the boosted
+GAMLSS). Therefore, the main manuscript focuses on the zero-censored
+Gaussian model which is well-established in the probabilistic
+precipitation forecasting literature. Moreover, the distributional forest
+can be shown to be rather robust against distributional
+(mis-)specifications compared to the other models.
+
+% The comparison of the zero-censored Gaussian models from the main manuscript with the
+% alternative distributional specifications in this supplement reveals that a Gaussian hurdle
+% may lead to small improvements. However, especially for EMOS and the boosted GAMLSS model
+% the outcome is also considerably more variable. Therefore, overall there is no convincing
+% enough improvement over the zero-censored Gaussian model which is therefore the focus of
+% the main manuscript. Also, the distributional forest appears to be more robust against
+% distributional (mis-)specifications compared to the other models.
 

### 2018-10-03 pkg/disttree/inst/draft/draft.Rnw
revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2306 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d9572df8a9de22eab2f427c279fa463032ea6162

@@
 lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
-legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
+legend("topright", c("Predicted distribution", "Predicted point mass at 0", "Observation"),
        bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)
@@
 @
-\caption{\label{fig:axams-dist}\hl{Total precipitation predictions by a distributional forest
-at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
-Observations are modelled by a left-censored at 0 distribution. The corresponding predicted point mass is shown at the censoring point (0).}}
+\caption{\label{fig:axams-dist}\hl{Total precipitation predictions by a 
+distributional forest at station Axams for July 24 in 2009, 2010, 2011 and 
+2012 learned on data from 1985--2008. Observations are non-negative and 
+modeled by a Gaussian distribution, left-censored at 0.
+The observations are depicted by crosses and the predicted point mass
+from the model by filled circles.}}
 \end{figure}
@@
 additive terms and their functional forms in advance which can be challenging in
-practice and potentially require domain knowledge, especially if the number of covariates $m$ is large.
+practice and potentially require expert knowledge in the application domain, 
+especially if the number of covariates $m$ is large.
 
@@
 that pick up changes in the location of the response across the covariates,
-\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. These
-pick up changes in the ``direction'' of any distribution parameter by
+\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. 
+These pick up changes in the ``direction'' of any distribution parameter by
 considering the full score vector for choosing splitting variables and split
-points. Each of the distributional trees is grown on a different data set
-obtained through resampling (bootstrap or subsampling) and in each node only a
+points. \hl{For each distributional tree a different learning data set is obtained} 
+through resampling (bootstrap or subsampling) and in each node only a
 random subset of the covariates $\bold{Z}$ is considered. As usual in random
@@
 \caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
-for station Axams based on models learned on data from 1985--2008. 
-(Residuals of precipitation are on the powertransformed scale.)}
+for station Axams based on models learned on data from 1985--2008.}
 \end{figure}
@@
 diagonal), indicating that all four models are sufficiently well calibrated. 
-This is also supported by the corresponding probability integral transform
-(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) histograms in
-Appendix~\ref{app:pit}.
+\hl{This is also supported by the corresponding probability integral transform}
+(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) \hl{histograms for station
+Axams in Supplement 2 (Stationwise Evaluation) which contains a more 
+detailed explanation of residual QQ plots and PIT histograms.}
 
@@
         font.axis = 3, #list(family="HersheySerif", face=3),
+        xlim = c(0,1.1),
         names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))

### 2018-10-03 pkg/disttree/inst/draft/rejoinder.tex
further revisions, also adapted to changes in the manuscript and the supplements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2307 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 75ba5d90d4c34e226ab5d13fdb064c2e4136d22e

@@
 
-It is difficult to set a minimal size of the learning data for
-the method to be effective, however, the distributional forest
-performed quite well in different applications with at least
-50 observations. 
-The stopping criteria that are based on the number of 
-observations in a node need to be adapted to the size of the
-learning data, especially if the data set is small.
-For example, in the application presented in this paper the 
-learning data consists of 744 observations and each tree is built
-on a subsample of around 470 observations (63.2 percent, as 
-this is the mean fraction for N-out-of-N bootstrapping with 
-replacement and therefore commonly used for subsampling
-in forest models). 
-The stopping criteria are set such that no split will be performed 
-if there are either less than 50 observations 
-in the considered node or if there were less than 20 observations in 
-one of the resulting subgroups of the considered possible split.
-(This information is now also added in Section~3.2., 'Distributional forest').
-Generally, these values (and also the significance level for
-the statistical tests in the tree algorithm) should be kept 
-rather low for a forest model as it is desirable to grow very 
-large trees. Therefore, each of the trees in the forest will 
-most likely overfit the subsample it is learned on, however, this 
-is evened out by combining them to a forest model (with an adequate 
-number of trees). For that reason, setting the stopping criteria too 
-low does not lead to a decrease in goodness of fit, however, the 
-computational costs might increase, especially for large data sets. 
-On the other hand, if they are too high, splits might not be detected 
+The required size of the training data depends heavily on the complexity 
+of the model, however, in different applications which we have tested the 
+distributional forest on it performed quite well with at least 50 observations. 
+Generally, the stopping criteria \texttt{minbucket} and \texttt{minsplit} (both 
+determining the minimal number of observations in a node) and \texttt{alpha} 
+(the significance level for the statistical tests in the tree algorithm) should 
+be kept rather low for a forest model as it is desirable to grow very large trees. 
+(FIX ME: reference?)
+Therefore, each of the trees in the forest will most likely overfit the subsample 
+it is learned on, however, this is evened out by combining them to a forest 
+model (with an adequate number of trees). (FIX ME: reference?) 
+For that reason, setting the stopping criteria too low does not lead to a decrease 
+in goodness of fit, however, the computational costs might increase, especially for 
+large data sets. On the other hand, if they are too high, splits might not be detected 
 such that the corresponding effects are not captured by the model.
@@
 
-With `domain knowledge' we are referring to expert knowledge 
-in the field of the considered application. 
-If variable selection and further model specifications have 
-to be made in advance this additional knowledge is required 
-in order to make reasonable decisions.
-
-\begin{reto}
-Im paper entsprechend adressieren (tbd).
-\end{reto}
+We have rephrased this as 'expert knowledge in the application domain'
+in Section~2.2..
 
@@
 
-These are the residuals of precipitation on the transformed scale.
+The residual QQ plots (Dunn et al. 1996) illustrate the probability 
+integral transform (PIT, Gneiting et al. 2007) values of the observations 
+in the testing data and are therefore scale invariant. A more detailed 
+explanation can be found in Supplement 2 (Stationwise Evaluation).
 
@@
 
-You are right, the observations can not be negative and are 
-therefore left-truncated at 0 and not left-censored (we have adapted 
-this in the caption of Figure~3). However, due to days without 
-rain there is a naturally given point mass at zero in precipitation 
-data, similar to the point mass that results from censoring 
-possibly negative data.
-Thus, we use the concept of \textit{censoring} which allows us to directly
-account for the large fraction of observations at zero and the non-negative
-part of the observations at once.
-The online supplement which comes along with our manuscript includes the
-results for a hurdle model where the occurrence of precipitation is addressed
-using a binary model while the non-negative part is
-modeled by a \textit{truncated} Gaussian distribution.
+You are right, our explanation was incorrect, thank you for pointing
+this out! We have now clarified that the model employs a censored
+distribution but the observations are not censored, just non-negative
+with a point mass at 0.
 
@@
 
-Since we are applying a censored distribution as explained above we 
-prefer using the term censoring point, even though the data itself 
-is not censored.
+We have rephrased this part of the caption to 'Observations are non-negative 
+and modeled by a Gaussian distribution, left-censored at 0.
+The observations are depicted by crosses and the predicted point mass
+from the model by filled circles.' to clearly distinguish between the model
+and the observations.
 
@@
 
-See points 3 and 4.
+See point 4.
 
@@
 
-\textit{FIX ME: ?}
+We have rephrased this to 'For each distributional tree a different learning 
+data set is obtained ... '
 
@@
 
-\textit{TO DO}
-
-\begin{reto}
-Beide? Verstehe ich den Kollegen da richtig?
-\end{reto}
+The PIT histograms and the residual QQ plots visualize the same
+results on different scales. The out-of-sample PIT values are 
+either plotted in a histogram or plotted against the corresponding
+quantiles of the standard normal distribution. A more detailed 
+explanation can be found in Supplement 2 (Stationwise Evaluation) 
+which contains both, PIT histograms and residual QQ plots, 
+for 15 stations. In the manuscript we decided to show only the 
+residual QQ plot as it allows for a better analysis of tail 
+behaviour than the PIT histogram.
 
@@
 % load("~/svn/partykit/pkg/disttree/inst/draft/Axams_24to4.rda")
+% gb <- Axams_24to4$gb
 % sgb <- stabsel(gb, cutoff = 0.6, q = 30)
@@
 
+\item[] Dunn, P. K. and Smyth, G. K. (1996).
+Randomized Quantile Residuals,
+\textit{Journal of Computational and Graphical Statistics}
+\textbf{5, 3}: 236--244.
+
+\item[] Gneiting, T., Balabdaoui, F. and Raftery, A. E. (2007).
+Probabilistic Forecasts, Calibration and Sharpness,
+\textit{Journal of the Royal Statistical Society B}
+\textbf{69, 2}: 243--268.
+
 \end{itemize}

### 2018-10-06 pkg/disttree/inst/draft/supplement/Rain_stationwise_results.Rnw
omit concordance
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2291 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0c81e683da54475febf987bef6b3c53106d5c7bf
adapted to changes in variable importance barplot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2309 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 535230c8ec202851f14de5ae50cac85c6cce076b

@@
 station <- "_STATION_"
+
 if(station == "SeeimPaznaun") station <- "See im Paznaun"
@@
 barplot(varimp10, horiz = TRUE, las = 1, axes = FALSE,
-        xlab = "Mean decrease in CRPS")
-axis(1, at = seq(0, ceiling(max(varimp10) * 5)/5, 0.2), las = 1, mgp = c(0, 1, 0))
+        xlab = "Mean increase in CRPS",
+        xlim = c(0,0.09))
+axis(1, at = seq(0, ceiling(max(varimp10) * 5)/5, 0.02), las = 1, mgp = c(0, 1, 0))
 @

### 2018-10-08 pkg/disttree/inst/draft/supplement/Rain_stationwise_results.Rnw
adapted to new results of variable importance
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2315 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: adebd8a052398e7be1f9bc0a7e9e68719d88191b

@@
         xlab = "Mean increase in CRPS",
-        xlim = c(0,0.09))
+        xlim = c(0,max(varimp10)))
 axis(1, at = seq(0, ceiling(max(varimp10) * 5)/5, 0.02), las = 1, mgp = c(0, 1, 0))

### 2018-10-10 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2320 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9bb15fcbae5d21f01b86034d67b0086639e10c93

@@
 library("RainTyrol")
+library("scoringRules")
 
@@
 transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
+
+# set function for parallelization
+applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))
+
 @
@@
 
-## get predicted parameter
-# for learndata and testdata
+## get PIT
 {
@@
   set.seed(7)
-  # locally redefine logLik.distforest to use crps in varimp()
-  logLik.distforest <- function(object, newdata = NULL, ...) {
-    if(is.null(newdata)) {
-      newdata <- object$data
+  nperm <- 50
+
+  # function to permutate chosen variable and then calculate mean crps
+  meancrps <- function(permute = NULL, newdata = Axams_24to4$testdata) {
+    if(!is.null(permute)) newdata[[permute]] <- sample(newdata[[permute]])
+    p <- predict(Axams_24to4$df, newdata = newdata, type = "parameter")
+    mean(crps_cnorm(newdata$robs, location = p$mu, scale = p$sigma, lower = 0))
   }
 
-    # predict parameter
-    pdf <- predict(object, newdata = newdata, type = "parameter")
-    df_mu <- pdf$mu
-    df_sigma <- pdf$sigma
-
-    # calculate CRPS
-    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)
-
-    return(structure(crps, df = NA, class = "logLik"))
-  }
-
-  vimp_crps <- varimp(Axams_24to4$df, nperm = 1L)
+  # apply for all covariates except for dswrf_mean_min and 
+  # dswrf_sprd_min (columns 30 and 33) as they are always 0
+
+  # risk <- sapply(c(5:29, 31, 32, 34: ncol(testdata)), meancrps)
+
+  # using only one core
+  # risk_all <- replicate(nperm, sapply(c(5:29, 31, 32, 34: ncol(testdata)), meancrps))
+  # risk <- rowMeans(risk_all)
+
+  # or parallel
+  risklist <- applyfun(1:nperm, 
+                       function(i){
+                         set.seed(i)
+                         sapply(c(5:29, 31, 32, 34: ncol(Axams_24to4$testdata)), meancrps)
+                       })
+  risk <- Reduce("+", risklist) / length(risklist)
+
+  names(risk) <- names(Axams_24to4$testdata)[c(5:29, 31, 32, 34: ncol(Axams_24to4$testdata))]
+  vimp_crps <- risk - meancrps(newdata = Axams_24to4$testdata)
+  vimp_crps <- sort(vimp_crps, decreasing = TRUE)
 
   save(vimp_crps, file = "vimp_crps.rda")
-
-  rm(logLik.distforest) 
 }
@@
 Compared to three alternative zero-censored Gaussian models distributional forests
-perform at least on par and sometimes clearly better while requiring no expert
-knowledge for the model specification. The three alternatives are: a standard ensemble model
+perform at least on par and sometimes clearly better while requiring no meteorological
+knowledge about the athmospheric processes which drive formation of precipitation
+for the model specification. The three alternatives are: a standard ensemble model
 output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
@@
 
-While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
-that pick up changes in the location of the response across the covariates,
-\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. 
-These pick up changes in the ``direction'' of any distribution parameter by
-considering the full score vector for choosing splitting variables and split
-points. \hl{For each distributional tree a different learning data set is obtained} 
-through resampling (bootstrap or subsampling) and in each node only a
-random subset of the covariates $\bold{Z}$ is considered. As usual in random
-forests, this reduces the correlation among the trees and stabilizes the
-variance of the model.
+\hl{The idea of random forests is to learn an ensemble of trees, each on a different
+learning data obtained through resampling (bootstrap or subsampling). In each node 
+only a random subset of the covariates $\bold{Z}$ is considered to reduce the 
+correlation among the trees and to stabilizes the variance of the model.
+For a simple regression random forest the mean of predictions over
+all trees is considered. In that way changes in the location  of the response 
+across the covariates are detected} (e.g. in Breiman and Cutler's random forests, 
+\cite{Breiman:2001}).
+\hl{This idea is now taken one step further by embedding it in a distributional 
+framework based on loglikelihood estimation. \emph{Distributional forests} 
+employ an ensemble of $T$~\emph{distributional trees} which pick up changes in 
+the ``direction'' of any distribution parameter by considering the full score 
+vector for choosing splitting variables and split points.}
 
@@
   in Section~\ref{sec:methods}. 
-  \hl{As stopping criteria the minimal number
-  of observations to perform a split} (\code{minsplit}) \hl{is set to $50$, 
-  the minimal number of observations in a terminal node} (\code{minbucket})
-  \hl{is set to $20$ and the significance level for variable selection} (\code{alpha}) 
-  \hl{is kept at the default value $1$.}
+  \hl{As stopping criteria the minimal number of observations to perform a split} 
+  (\code{minsplit}) \hl{is set to $50$, the minimal number of observations in a 
+  terminal node} (\code{minbucket})\hl{is set to $20$ and the significance level 
+  for variable selection} (\code{alpha}) \hl{is kept at the default value $1$.}
   % stopping criteria:
@@
 <<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
-par(mar = c(2.5,10,1,2))
+par(mar = c(4,10,1,2))
 barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
         horiz = TRUE, las = 1, axes = FALSE,
-        xlab = "Variable importance: mean decrease in CRPS",
+        xlab = "Variable importance: mean increase in CRPS",
         font.axis = 3, #list(family="HersheySerif", face=3),
-        xlim = c(0,1.1),
+        xlim = c(0,0.08),
         names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
-axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
+axis(1, at = seq(0,1,0.02), las = 1, mgp=c(0,1,0))
 @
@@
 % sum(is.na(crps[,3]))
-%<<eval = TRUE, echo = FALSE>>=
 
@@
 for disciplines such as meteorology.
+
+\subsubsection*{Variable Selection}
+\hl{Generally, there are many possibilities how to specify the variables that are
+to be included in a model. Especially for a low number of variables all types
+of interactions can be employed and mixed effect models and separations into 
+main effects and interaction effects can be considered allowing for easier 
+interpretations} (see for example \cite{Wood+Scheipl+Faraway:2013}, 
+\cite{Goicoa+Adin+Ugarte:2018}, \cite{Ugarte+Adin+Goicoa:2017}).
+However, in the application presented in Section~\ref{sec:precipitation}
+alltogether 80 variables are available. Even if only paired interactions were
+considered this would already lead to 3160 additional covariates in the model
+making it very difficult to keep the model manageable regarding computational
+costs but also interpretability. Therefore, only main effects are considered
+for the distributional forest and the boosted GAMLSS which both automatically
+select covariates while meteorological expert knowledge was applied to 
+select covariates and specify interactions for the prespecified GAMLSS.
+Therefore,
+
+\subsubsection*{Tuning parameters}
+Setting tuning parameters of a forest model can have a major influence
+not only on the accuracy of the model but also on the computational 
+complexity. 
+For the application in Section~\ref{sec:precipitation} we have 
+selected the values of these parameters (listed in Section~\ref{sec:evaluation}) 
+based on advice from literature and our own experience.
+
+\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} recommend to 
+build full-grown trees. Therefore, low values should be chosen for the stopping 
+criteria \code{minsplit}, \code{minbucket} and \code{alpha}.
+However, in a distributional model it also has to be considered that a 
+minimal number of observations is required to allow for a reasonable 
+estimation of parameters in each node. For that reason the minimal number
+of observations for which statistical tests are performed in the 
+distributional forest in Section~\ref{sec:precipitation} is set to
+50 (\code{minbucket}) and the minimal number of observations on which
+parameters are estimated is set to 20 (\code{minsplit}).
+
+As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
+\hl{random forests do not overfit, not even for an increasing number of 
+trees} (\code{ntree}). \hl{Hence, it is recommended to build forests consisting 
+of a large number of trees} (\code{ntree}) \hl{as this generally leads to an increase 
+in accuracy. However, this also comes along with increasing computational costs. 
+Therefore, ``a good choice results from a trade-off between computational 
+complexity and accuracy''} \citep{Biau+Scornet:2016}.
+\hl{Following this advice we chose to build forests consisting of 100 trees. 
+However, increasing this tuning parameter could improve the performance
+of the distributional forest without the problem of overfitting.}
+
+%% TO DOs:
+%% East Tyrol
+\subsubsection*{East Tyrol}
+Figure~\ref{fig:all-map} shows that, contrary to the rest of Tyrol, 
+for the majority of stations in East Tyrol the prespecified GAMLSS performs
+best in this particular setting (learning data: 1985--2008, testing data:
+2009--2012). East Tyrol is located south of the main Alpine Ridge and 
+exhibits different climatological characteristics than the part north of 
+the Alpine Ridge. It seems that the choice of covariates and their interactions 
+we made for the prespecified GAMLSS method is particularly adequate for the 
+region of East Tyrol for this specific analysis. However, the advantages of 
+the prespecified GAMLSS models in this particular setting (training: 1985--2008, 
+validation: 2009--2012) even out in a cross-validation framework for most of 
+the stations in East Tyrol. These results are presented in the online supplement 
+which comes along with our manuscript. There, more extensive evaluations
+are carried out for 15 observation stations, including a cross-validation
+as presented for station Axams in this paper.
+
+\subsubsection*{Choice of station Axams}
+%% choice of station Axams mit Verweis auf Supplement stationwise
+Axams was chosen for the more extensive evaluations presented
+in this paper as it is the observation station being closest 
+to Innsbruck, the capital of Tyrol and the work place of 
+three of the authors, but not based on any other particular 
+features. Furthermore, it has been checked, whether it 
+is an adequate representative of all stations regarding 
+the results of our evaluations. This is now also 
+confirmed by further analysis in the online supplement which
+comes along with our manuscript.
+There, the same evaluations as for station Axams have been carried out for 14 other
+stations, including some of those where the forest does not perform better than
+the other models in the setting as presented in Figure~\ref{fig:all-map}.  
+It shows that the differences in performance even out in the cross-validation 
+framework. There, the forest performs at least equally well compared to the 
+best of the other models at most stations.
+
+\subsubsection{Distributional specifications}
+%% choice of distribution mit Verweis auf Supplement distributions
+Moreover, as mentioned before we have tested robustness against 
+distributional misspecification. The results show that applying a 
+left-censored at zero logistic distribution or a Gaussian hurdle model 
+instead of the left-censored at zero Gaussian distribution, as presented 
+in the paper, does not lead to any significant improvements in goodness of 
+fit of the applied models. The exact results are provided in an online 
+supplement which comes with our manuscript.
+
+\subsubsection*{Computational difficulties}
+%% GAMLSS errors with references gamboostLSS Hofner et al.
+As stated by \cite{Hofner+Mayr+Schmid:2016} the AIC-based variable selection 
+methods implemented in the \textsf{R} package \textbf{gamlss} ``can be unstable, 
+especially when it comes to selecting possibly different sets of variables for 
+multiple distribution parameters.'' We have noticed computational problems when
+applying the function \textbf{gamlss} in certain settings within the cross-validation
+framework as it did not succeed in fitting a model. In this cases the method
+was not taken into consideration in the comparison of all applied models.
 

### 2018-10-10 pkg/disttree/inst/draft/rejoinder.tex
some more revision based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2321 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: db8f0d9ca0385c319b96352e1c0dfe43b43026ff

@@
 %\documentclass{article}
-\usepackage{sfmath,hyperref,color,enumitem}
+\usepackage{sfmath,hyperref,color,enumitem,soul}
 
@@
 
-All changes and additons are explained in much more detail in the
+All changes and additions are explained in much more detail in the
 point-to-point reply on the next pages.
@@
 of the model, however, in different applications which we have tested the 
-distributional forest on it performed quite well with at least 50 observations. 
-Generally, the stopping criteria \texttt{minbucket} and \texttt{minsplit} (both 
+distributional forest on it performed quite well with at least 50 observations. \\
+Applying the Law of Large Numbers it can be shown that random forests do not overfit, 
+not even for an increasing number of trees (Breiman (2001), Hastie et al. (2001), 
+Biau et al. (2016)). Therefore, it is recommended to build a large number of trees
+as this generally leads to more accurate predictions. However, ``the computational cost 
+for inducing a forest increases linearly with the number of trees, so a good choice 
+results from a trade-off between computational complexity and accuracy'' 
+(Biau et al. (2016)). Following this advice, we decided to build forests consisting 
+of 100 trees.\\
+Moreover, Hastie et al. (2001) recommend to build trees of maximal depth:
+``Our experience is that using full-grown trees seldom costs much, and 
+results in one less tuning parameter.''
+Thus, the stopping criteria \texttt{minbucket} and \texttt{minsplit} (both 
 determining the minimal number of observations in a node) and \texttt{alpha} 
 (the significance level for the statistical tests in the tree algorithm) should 
-be kept rather low for a forest model as it is desirable to grow very large trees. 
-(FIX ME: reference?)
-Therefore, each of the trees in the forest will most likely overfit the subsample 
-it is learned on, however, this is evened out by combining them to a forest 
-model (with an adequate number of trees). (FIX ME: reference?) 
-For that reason, setting the stopping criteria too low does not lead to a decrease 
-in goodness of fit, however, the computational costs might increase, especially for 
-large data sets. On the other hand, if they are too high, splits might not be detected 
-such that the corresponding effects are not captured by the model.
-
-\textit{FIX ME: too long?}
+be kept rather low. However, in a distributional framework it also has to be 
+considered that the minimal number of observations in a node needs to be high 
+enough in order to allow for estimating the distribution parameters reasonably well.
+In our application we have set \texttt{minsplit} to 50 and \texttt{minbucket} to 20.
+These tuning parameters are now also listed in the main manuscript in Sections~3.2.
+and discussed in Section~4..
 
@@
 confirmed by further analysis in the online supplement which
-comes along with our manuscript.
+comes along with our manuscript and discussed in Section~4..
 
@@
 prespecified GAMLSS and better than the EMOS at most of the stations.
+Moreover, this point is now also discussed in Section~4. in the main manuscript.
 
@@
 
-We have rephrased this to 'For each distributional tree a different learning 
-data set is obtained ... '
+We have rephrased this the second paragraph of Section~2.3. which
+included this sentence.
 
@@
 is kept at the default value 1. We have added this information in 
-the manuscript in Section~3.2., 'Distributional forest'.
+the manuscript in Section~3.2., 'Distributional forest' and
+also discussed the choice of these values in Section~4. and
+in the answer to the second point of the associate editor.
 
@@
 
-Thank you for pointing this out. We have now rerun the exact 
-same evaluations as presented in the paper with different 
-distributional assumptions, including a logistic model and 
-a hurdle model with a truncated part. The exact results can be 
-found in an online supplement.
+We have applied the left-censored at zero Gaussian distribution
+as it has proven to be an adequate choice for this type of 
+application in earlier publications.
+Moreover, the residual QQ plots and PIT histograms illustrate
+that the models are well calibrated employing this distribution.
+But thank you very much for the advice to consider a hurdle model!
+This is definitely a good idea and therefore we have carried out the
+same evaluations as presented in the paper but employing a
+Gaussian hurdle model. Additionally, a left-censored at zero logistic
+distribution has been applied as well in the same setting in
+order to account for heavier tails. The results of these evaluation
+can be found in the online supplement 1 (``Different Response Distributions''). 
+Overall, it is shown that under the logistic distribution the methods perform 
+slightly worse while the hurdle model leads to small improvement in some settings,
+however, this comes along with broader variability. The distributional 
+forest is proven to be very robust against distributional misspecifications.
+Based on these results we decided to keep the distributional assumption
+considered in the main manuscript but we have also discussed the tested
+alternatives in Section~4..
 
@@
 affects the distribution in which direction is possible. 
-I think, the reason that focussing on predictions should 
+I think, the reason that focusing on predictions should 
 be made clearer and well justified.
@@
 might have caused. They are now added.
-
-\textit{FIX ME: comment on 'there should be a 
-better motivation and connection of the methodology and the 
-data application'?}
+We have also rephrased Section~2.3. which now first introduces random 
+forests and then distributional forests as an extension of classical 
+random forests.
 
@@
 
-Thank you for your suggestions and the references. 
-The prespecification of the distribution is expert knowledge in 
-this context as in some of our previous publications we have been 
-working on different response distributions and the censored normal 
-distribution turned out to be one of the best performing. In this article 
-we have not proven whether or not the chosen distribution is the overall
-best for this specific application, however, the main part of the article 
-is and should be the distributional forest and not the application itself. 
-However, we are sure that, given our previous work, the selected methods 
-and model assumptions are well performing and well suited for the practical 
-example presented.
-Moreover, as mentioned before we have tested robustness against 
-distributional misspecification. The results show that applying a 
-left-censored at zero logistic distribution or a Gaussian hurdle model 
-instead of the left-censored at zero Gaussian distribution, as presented 
-in the paper, does not lead to any significant improvements in goodness of 
-fit of the applied models. The exact results are provided in an online 
-supplement which comes with our manuscript.
-
-\textit{FIX ME: refer to Wood, Goicoa, Ugarte}
+Thank you for your suggestions and the references. These are all
+good ideas, however, in our application the number of covariates
+is too large in order to keep a model with all interactions manageable.
+We have also discussed this in Section~4 of the main manuscript and
+rephrased the corresponding sentence in the introduction to ``while 
+requiring no meteorological knowledge about the atmospheric processes 
+which drive formation of precipitation'' in order to clarify that
+this refers only to the variable prespecification in this application.
 
@@
 
-You are right, there are clearly some stations where other models perform
-better than the forest. However, at the majority of stations used in our study
-it can at least keep up with them and therefore seems to provide a good and
-reasonable alternative that is very easy to apply as no expert knowledge or
-additional work in advance is required and the computational effort is kept
-relatively small compared to a boosting method. In an online supplement the
-same evaluations as for station Axams have been carried out for 14 other
-stations, including some of those where the forest does not perform better than
-the other models in the setting presented in Figure~8.  It shows that the
-differences in performance even out in the cross-validation framework. There,
-the forest performs at least equally well compared to the best of the other
-models at most stations.
-
-\textit{FIX ME: comments on simulation study? comments on use of
-method by meteorologists? (last sentence)}
-
-\begin{reto}
-``[\dots] I am wondering whether a meteorologist would really do a
-better job on a relative scale of precipitation.'' ist wohl eher eine
-rethorische Frage und tut nix zur Sache wenn du mich fragst.
-Unkommentiert lassen? Oder Z hat vielleicht eine Idee?
-\end{reto}
+In order to investigate on these questions we have carried out the full
+analysis as presented for station Axams in the main manuscript for 14
+further observation stations. It shows that the distributional
+forest is very robust against distributional misspecifications
+and almost always competitive with the best among the other methods 
+or even better. The exact results are provided in 
+Supplement 2 (Stationwise Evaluation) and discussed in Section~4. 
+Especially for meteorologists intending to apply model output 
+statistics (MOS) for several hundreds of observation stations 
+(\hl{\texttt{FIXME: reference Reto}})
+the distributional forest provides a good and robust alternative. 
+Even though computational costs are higher than for an EMOS model, 
+they are similar as for a prespecified model and lower than for 
+a model applying variable selection based on cross-validation.
 
@@
 Section~3.3..
+We have chosen two different setups: First we learned the model
+on the first 24 years and evaluated on the following 4 years
+in order to provide a classic time ordered setup. 
+However, following (\hl{\texttt{FIXME:} reference Reto})
+no influence of time is to be expected. This is also supported by an 
+autocorrelation coefficient of -0.007 of the out-of-bag quantile 
+residuals of the distributional forest model for station Axams for
+a time lag of 1 day.
+% ## calculation of autocorrelation:
+% load("Axams_24to4.rda") 
+% set.seed(4)
+% pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata, type = "parameter")
+% pit_df <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = pdf$mu, sd = pdf$sigma))
+% pit_df[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_df[Axams_24to4$testdata[,"robs"]>0, 2]
+% ac <- acf(qqrplot(pit_df)$residuals)
+Then, to evaluate the models in a broader framework a cross-validation 
+with the above described choice of folds was applied, for station
+Axams in the main manuscript and for 14 further stations in 
+Supplement 2 (Stationwise Evaluation).
 
@@
 GAMLSS. I guess, the GAMLSS could improve a lot by this . 
-Second, what does a basic EMOS exactly mean? Third, why not to include 
-interactions in the boosting algorithm as well? this should be 
-conceptually straight forward? Fourth, why are 100 trees chosen 
-in the forest? Is this a a default value? May this number not also 
-be dependent on the complexity of the model? Why not to choose 
-the number of trees similar to what is done in boosting? this 
-would also increase the computation time of the forest a lot.
-\end{review}
-
-We decided to evaluate two different versions of a GAMLSS: 
-the prespecified GAMLSS where the covariates are selected 
-in advance based on expert knowledge and the boosted GAMLSS
-where all covariates are included. For the latter covariates 
-are selected within the boosting algorithm. As stated by 
-Hofner et al. (2016) the AIC-based variable selection 
-methods implemented in the \textsf{R} package \textbf{gamlss} 
-``can be unstable, especially when it comes to selecting 
-possibly different sets of variables for multiple 
-distribution parameters.''
-
-By a `basic EMOS' we are referring to the linear distributional 
-regression model based on the idea presented by Gneiting et al. (2005) 
-where only the ensemble mean is included for the location parameter 
-and the ensemble standard deviation for the scale parameter.
-
-In the boosted GAMLSS we did not include interactions as this would
-lead to an extremely high number of covariates that would be hardly
-manageable. Even if we included only the pairwise interactions of all 
-80 covariates we would end up with 3160 additional covariates.
-
+\end{review}
+
+As mentioned in point 5 and also discussed in Section~4, the high
+number of covariates (80) restricts the possibilities of variable selection
+and specification. For an AIC or BIC based selection an evaluation
+of 80 main effects for two parameters would make this a very complex
+procedure. Moreover, Hofner et al. (2016) stated that the 
+AIC-based variable selection methods implemented in the \textsf{R} 
+package \textbf{gamlss} ``can be unstable, especially when it comes 
+to selecting possibly different sets of variables for 
+multiple distribution parameters.'' For that reason we decided
+not to choose such a selection method.
+
+\begin{review}
+Second, what does a basic EMOS exactly mean? 
+\end{review}
+
+We have adapted the introduction of a basic EMOS in the main manuscript
+in Section~ ...
+% By a `basic EMOS' we are referring to the linear distributional 
+% regression model based on the idea presented by Gneiting et al. (2005) 
+% where only the ensemble mean is included for the location parameter 
+% and the ensemble standard deviation for the scale parameter.
+
+\begin{review}
+Third, why not to include interactions in the boosting algorithm 
+as well? this should be conceptually straight forward? 
+\end{review}
+
+As now discussed in Section~4 including only all pairwise interactions
+of the 80 available variables would already lead to 3160 additional
+covariates which would make the model too complex and hardly manageable
+regarding computational costs. Therefore, we decided not to include
+any interactions in the boosting algorithm.
 % factorial(80)/factorial(80-2)/factorial(2)
 
-\begin{reto}
-K\"onnte man eventuell auch im Paper so erw\"ahnen?
-
-\begin{equation}
-    \frac{n!}{k! \cdot (n-k)!}~~\mbox{bei}~n=80~\mbox{und}~k=2~\mbox{die}~3160~\mbox{Kombinationen.}
-\end{equation}
-\end{reto}
-
-By default the number of trees in a distributional forest
-is set to 500. However, to keep computation time at a 
-reasonable level we decided  to use only 100 trees. 
-But of course, the more trees, the better the complexity of 
-the model can be captured, especially smooth transitions can 
-be modeled better by a higher number of trees.
-
-\textit{FIX ME: comment on 'Why not to choose the number of trees 
-similar to what is done in boosting? this would also increase the 
-computation time of the forest a lot.'}
+\begin{review}
+Fourth, why are 100 trees chosen in the forest? Is this a a default 
+value? May this number not also be dependent on the complexity of 
+the model? Why not to choose the number of trees similar to what is 
+done in boosting? this would also increase the computation time of 
+the forest a lot.
+\end{review}
+
+Following Breiman (2001) random forests are very robust and do not overfit, 
+not even for a high number of trees. However, computational costs increase 
+linearly with the number of trees (Biau et al. (2016)).
+Therefore, we followed these recommendations of choosing a high number of
+trees, but still not too high in order to keep the computation time at
+a reasonable level. Moreover, 100 is a commonly used number of trees that
+seems to work very well. Of course, more trees could be applied but 
+the performance is usually not significantly worse with 100 trees.
+Tuning parameters of forest models, including the number of 
+trees (\texttt{ntree}), are now discussed in Section~4 
+of the main manuscript.
 
@@
 \end{review}
+
+Thank you for your advice. We have kept only the details regarding
+the presented methodology in the appendix of the main manuscript
+(Appendix A) and extended Appendix B to an online supplement
+(Supplement 2: Stationwise Evaluation).
 
@@
 \end{review}
+
+We used ``EMOS'' as an expression for a standard post-processing
+model output statistics (MOS, Gneiting et al. (2005)) with a censored 
+distribution. We have revised the introduction of the EMOS model in 
+the main manuscript and hope this now explains the model better.
 
@@
 residual QQ plot as it allows for a better analysis of tail 
-behaviour than the PIT histogram.
+behavior than the PIT histogram.
 
@@
 specification of the boosted GAMLSS for which all available 
-covariates are included. Out of them, in the exact same setting 
-(Axams, 1985--2008), the boosting algorithm selects 28 covariates 
-for the location parameter, mainly related to precipitation (7) 
-and temperature (11), and 46 covariates for the scale parameter, 
-mainly related to precipitation (10), convective available potential 
-energy (8) and temperature (15).
+covariates are included. As for the distributional forest, 
+we have also calculated variable importance based on mean increase 
+of out-of-bag CRPS for the gamboostLSS model for the exact same 
+setting (station Axams, learned on 1985--2008). The resulting 
+top 10 variables are listed in the following barplot:
+
+\begin{center}
+\includegraphics[width = 0.4\textwidth]{varimp_gb.jpeg}
+\end{center}
+
+Overall it can be stated that the most important covariates of the boosted 
+GAMLSS are similar to the ones of the distributional forest even though they
+are not exactly the same ones and the influence of \texttt{tp\_max} is clearly
+stronger in the boosted GAMLSS.
+
+% Out of them, in the exact same setting (Axams, 1985--2008), 
+% the boosting algorithm selects 28 covariates 
+% for the location parameter, mainly related to precipitation (7) 
+% and temperature (11), and 46 covariates for the scale parameter, 
+% mainly related to precipitation (10), convective available potential 
+% energy (8) and temperature (15).
 % mu: 7 tppow, 1 cape, 2 dswrf, 2 msl, 2 pwat, 3 tcolc, 11 temperature
@@
 
-Applying the stability selection procedure by Meinshausen et al. (2010) and
-Hofner et al. (2015) on the same model shows that the 10 covariates with 
-maximal selection probability are six covariates based on precipitation and 
-one each based on mean sea level pressure, downwards short wave radiation 
-flux, temperature and precipitable water respectively.
-
-% load("~/svn/partykit/pkg/disttree/inst/draft/Axams_24to4.rda")
-% gb <- Axams_24to4$gb
-% sgb <- stabsel(gb, cutoff = 0.6, q = 30)
-
-% R> sgb$selected
-% tppow_min.mu         tppow_max.mu    tppow_mean1824.mu   dswrf_mean_mean.mu 
-%      3                    4                    7                   25 
-% tppow_min.sigma      tppow_max.sigma tppow_sprd2430.sigma  pwat_mean_max.sigma 
-%      83                   84                   92                  117 
-% t700_mean_max.sigma   msl_diff.sigma 
-%      138                  160
-
-% R> head(sort(sgb$max, decreasing = TRUE), 10)
-% tppow_max.mu    tppow_mean1824.mu      tppow_max.sigma  t700_mean_max.sigma 
-%    1.00                 0.94                 0.91                 0.87 
-% dswrf_mean_mean.mu  pwat_mean_max.sigma  tppow_min.sigma tppow_sprd2430.sigma 
-%    0.85                 0.84                 0.81                 0.74 
-% msl_diff.sigma      tppow_min.mu 
-%    0.70                 0.63
-
-% R> head(sort(rowMeans(sgb$phat), decreasing = TRUE), 10)
-% tppow_max.mu    tppow_mean1824.mu  t700_mean_max.sigma      tppow_max.sigma 
-%    0.99516              0.86018              0.84936              0.84857 
-% tppow_min.sigma  pwat_mean_max.sigma   dswrf_mean_mean.mu tppow_sprd2430.sigma 
-%    0.79779              0.75813              0.74074              0.65091 
-% msl_diff.sigma         tppow_min.mu 
-%    0.63928              0.55443
-
-Overall it can be stated that the most important covariates of the boosted 
-GAMLSS are similar to the ones of the distributional forest even though they
-are not exactly the same ones.
-
 \bigskip
@@
 two-dimensional, and three-dimensional B-splines to specify space-time 
-interations in bayesian disease mapping: Model fitting and model 
+interactions in bayesian disease mapping: Model fitting and model 
 identifiability, \textit{Spatial Statistics} p. to appear.
@@
 
+\item[] Biau, G. and Scornet, E. (2016).
+A Random Forest Guided Tour,
+\textit{TEST}
+\textbf{25}: 197--227.
+
+\item[] Hastie, T., Tibshirani, R. and Friedman, J. (2001).
+The Elements of Statistical Learning,
+\textit{Springer New York Inc.}.
+
+\item[] Breiman, L. (2001).
+Random Forests,
+\textit{Machine Learning}
+\textbf{45, 1}: 5--32.
+
 \end{itemize}

### 2018-10-11 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2322 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0dfc728e986b921f0e6d64e66120d4f400f19891

@@
 rather than relying on just the one fixed test set for the years 2009--2012.
-\hl{To do so, a 10 times 7-fold cross-validation is carried out where each
-splits the available 28~years into 7~subsets of 4~randomly selected and 
+\hl{For each setting in the considered 10 times 7-fold cross-validation the 
+available 28~years are split into 7~subsets of 4~randomly selected and 
 therefore mostly not subsequent years.}
@@
 \cite{Goicoa+Adin+Ugarte:2018}, \cite{Ugarte+Adin+Goicoa:2017}).
-However, in the application presented in Section~\ref{sec:precipitation}
-alltogether 80 variables are available. Even if only paired interactions were
+\hl{However, in the application presented in} Section~\ref{sec:precipitation}
+\hl{alltogether 80 variables are available. Even if only paired interactions were
 considered this would already lead to 3160 additional covariates in the model
@@
 select covariates while meteorological expert knowledge was applied to 
-select covariates and specify interactions for the prespecified GAMLSS.
-Therefore,
+select covariates and specify interactions for the prespecified GAMLSS.}
 
 \subsubsection*{Tuning parameters}
-Setting tuning parameters of a forest model can have a major influence
+\hl{Setting tuning parameters of a forest model can have a major influence
 not only on the accuracy of the model but also on the computational 
 complexity. 
-For the application in Section~\ref{sec:precipitation} we have 
-selected the values of these parameters (listed in Section~\ref{sec:evaluation}) 
+For the application in} Section~\ref{sec:precipitation} \hl{we have 
+selected the values of these parameters} (listed in Section~\ref{sec:evaluation}) 
 based on advice from literature and our own experience.
 
-\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} recommend to 
+\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hl{recommend to 
 build full-grown trees. Therefore, low values should be chosen for the stopping 
-criteria \code{minsplit}, \code{minbucket} and \code{alpha}.
-However, in a distributional model it also has to be considered that a 
+criteria} \code{minsplit}, \code{minbucket} and \code{alpha}.
+\hl{However, in a distributional model it also has to be considered that a 
 minimal number of observations is required to allow for a reasonable 
@@
 of observations for which statistical tests are performed in the 
-distributional forest in Section~\ref{sec:precipitation} is set to
-50 (\code{minbucket}) and the minimal number of observations on which
-parameters are estimated is set to 20 (\code{minsplit}).
+distributional forest in} Section~\ref{sec:precipitation} is set to
+50 (\code{minbucket}) \hl{and the minimal number of observations on which
+parameters are estimated is set to 20} (\code{minsplit}).
 
@@
 trees} (\code{ntree}). \hl{Hence, it is recommended to build forests consisting 
-of a large number of trees} (\code{ntree}) \hl{as this generally leads to an increase 
-in accuracy. However, this also comes along with increasing computational costs. 
+of a large number of trees as this generally leads to an increase 
+in accuracy. However, it also comes along with increasing computational costs. 
 Therefore, ``a good choice results from a trade-off between computational 
 complexity and accuracy''} \citep{Biau+Scornet:2016}.
-\hl{Following this advice we chose to build forests consisting of 100 trees. 
-However, increasing this tuning parameter could improve the performance
-of the distributional forest without the problem of overfitting.}
-
-%% TO DOs:
-%% East Tyrol
+\hl{Following this advice we chose to build forests consisting of 100 trees.}
+
 \subsubsection*{East Tyrol}
-Figure~\ref{fig:all-map} shows that, contrary to the rest of Tyrol, 
+Figure~\ref{fig:all-map} \hl{shows that, contrary to the rest of Tyrol, 
 for the majority of stations in East Tyrol the prespecified GAMLSS performs
@@
 validation: 2009--2012) even out in a cross-validation framework for most of 
-the stations in East Tyrol. These results are presented in the online supplement 
-which comes along with our manuscript. There, more extensive evaluations
-are carried out for 15 observation stations, including a cross-validation
-as presented for station Axams in this paper.
+the stations in East Tyrol. These results are presented in Supplement 2 
+(Stationwise Evaluation). There, the exact same evaluations as in} 
+Section~\ref{sec:axams} \hl{for station Axams have been carried out 
+for 14 other stations.}
 
@@
 %% choice of station Axams mit Verweis auf Supplement stationwise
-Axams was chosen for the more extensive evaluations presented
+\hl{Axams was chosen for the more extensive evaluations presented
 in this paper as it is the observation station being closest 
@@
 is an adequate representative of all stations regarding 
-the results of our evaluations. This is now also 
-confirmed by further analysis in the online supplement which
-comes along with our manuscript.
-There, the same evaluations as for station Axams have been carried out for 14 other
-stations, including some of those where the forest does not perform better than
-the other models in the setting as presented in Figure~\ref{fig:all-map}.  
-It shows that the differences in performance even out in the cross-validation 
-framework. There, the forest performs at least equally well compared to the 
-best of the other models at most stations.
+the results of our evaluations. This is also confirmed by 
+the extended analysis in Supplement 2 (Stationwise Evaluation)
+where the 15 considered stations also include some of those where 
+the forest does not perform better than the other models in the setting 
+presented in} Figure~\ref{fig:all-map}. \hl{It shows that the differences in 
+performance even out in the cross-validation framework where the forest 
+performs at least equally well compared to the best of the other models 
+at most stations.}
 
@@
 %% choice of distribution mit Verweis auf Supplement distributions
-Moreover, as mentioned before we have tested robustness against 
-distributional misspecification. The results show that applying a 
-left-censored at zero logistic distribution or a Gaussian hurdle model 
-instead of the left-censored at zero Gaussian distribution, as presented 
-in the paper, does not lead to any significant improvements in goodness of 
-fit of the applied models. The exact results are provided in an online 
-supplement which comes with our manuscript.
+\hl{Choosing an adequate distributional family is a crucial step and forms
+the basis required to build a well fitting model. Therefore, this
+decision should be made thoughtfuly. 
+In the presented application we applied a Gaussian left-censored at
+zero distribution as this has proven to be a well working choice in
+earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
+\hl{But to test for robustness against distributional misspecifation 
+different distributional assumptions have been considered as well. 
+In Supplement 1 (Different Response Distributions) the exact
+same evaluations as in} Section~\ref{sec:all} \hl{are carried out for 15
+observation stations, once employing a Gaussian hurdle model and once
+with a left-censored at zero logistic distribution in order to account
+for heavier tails. Under both assumptions the obtained results 
+are similar to those presented in the paper applying a left-censored
+at zero Gaussian distribution. However, comparing the two censored distributions,
+at many stations the models perform not as well under the logisctic distribution 
+as under the Gaussian distribution. The hurdle model leads to slight improvements 
+in some settings, but this comes with more variablity (especially for EMOS and the 
+boosted GAMLSS).
+Alltogether, the distributional forest has proven to be very robust against
+distributional misspecifications.}
 
@@
 %% GAMLSS errors with references gamboostLSS Hofner et al.
-As stated by \cite{Hofner+Mayr+Schmid:2016} the AIC-based variable selection 
-methods implemented in the \textsf{R} package \textbf{gamlss} ``can be unstable, 
+As stated by \cite{Hofner+Mayr+Schmid:2016} \hl{the AIC-based variable selection 
+methods implemented in the} \textsf{R} package \textbf{gamlss} \hl{``can be unstable, 
 especially when it comes to selecting possibly different sets of variables for 
 multiple distribution parameters.'' We have noticed computational problems when
-applying the function \textbf{gamlss} in certain settings within the cross-validation
+applying the function} \textbf{gamlss} \hl{in certain settings within the cross-validation
 framework as it did not succeed in fitting a model. In this cases the method
-was not taken into consideration in the comparison of all applied models.
+was not taken into consideration in the comparison of all applied models.}
 

### 2018-10-12 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2323 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4370c954bbfdcdb0fe859110bb476a265775f6a2

@@
 
+\definecolor{changes}{rgb}{1,1,0.7}
+\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}
+
 \numberwithin{equation}{subsection}
@@
 @
-\caption{\label{fig:axams-dist}\hl{Total precipitation predictions by a 
+\caption{\label{fig:axams-dist}\hlc{Total precipitation predictions by a 
 distributional forest at station Axams for July 24 in 2009, 2010, 2011 and 
@@
 $$
-\hl{Even though this parameter depends on the given $z$ it has to be noted that
+\hlc{Even though this parameter depends on the given $z$ it has to be noted that
 it is the same for all other observations that are in the same terminal node of 
@@
 
-\hl{The idea of random forests is to learn an ensemble of trees, each on a different
+\hlc{The idea of random forests is to learn an ensemble of trees, each on a different
 learning data obtained through resampling (bootstrap or subsampling). In each node 
@@
 \cite{Breiman:2001}).
-\hl{This idea is now taken one step further by embedding it in a distributional 
+\hlc{This idea is now taken one step further by embedding it in a distributional 
 framework based on loglikelihood estimation. \emph{Distributional forests} 
@@
 Many statistical weather forecasting models leverage the strengths of modern
-numerical weather prediction (NWP) systems
-\citep[see][]{Bauer+Thorpe+Brunet:2015}.  
-One frequently used method based on
-distributional regression models is the ensemble model output statistics 
-(EMOS) approach first proposed by
+numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}.  
+One frequently used method based on distributional regression models is the 
+ensemble model output statistics (EMOS) approach first proposed by
 \cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
 specific quantities and sites.  
-In case of precipitation forecasting EMOS typically uses the ensemble mean of 
-``total precipitation'' (\emph{tp}) forecasts as predictor for the location parameter and
-the corresponding ensemble standard deviation for the scale part of the statistical model
-to correct for possible errors of the ensemble in both, the expectation but also
-the uncertainty of a specific forecast.
+\hlc{In case of precipitation forecasting EMOS typically uses the ensemble mean of 
+``total precipitation'' (\emph{tp}) forecasts as predictor for the location 
+parameter $\mu$ and the corresponding ensemble standard deviation for the 
+scale parameter $\sigma$ of the statistical model based on the assumption
+of the response following a left-censored at zero Gaussian distribution.
+This distributional approach of modeling both parameters allows to correct 
+for possible errors of the ensemble in both, the expectation but also the 
+uncertainty of a specific forecast. In order to assure positivity of the 
+scale parameter a log-link is applied. Hence, the two distribution parameters 
+are modeled by the linear functions} 
+$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
+$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
+with power transformation parameter $\frac{1}{p}$ and 
+regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$.  
+\hlc{(Hereafter, this model is going to be referred to as a basic EMOS model.)}
+
+\hlc{\textit{FIXME: notation with or withoug power transformation parameter?}}
 
@@
   in Section~\ref{sec:methods}. 
-  \hl{As stopping criteria the minimal number of observations to perform a split} 
-  (\code{minsplit}) \hl{is set to $50$, the minimal number of observations in a 
-  terminal node} (\code{minbucket})\hl{is set to $20$ and the significance level 
-  for variable selection} (\code{alpha}) \hl{is kept at the default value $1$.}
+  \hlc{As stopping criteria the minimal number of observations to perform a split} 
+  (\code{minsplit}) \hlc{is set to $50$, the minimal number of observations in a 
+  terminal node} (\code{minbucket})\hlc{is set to $20$ and the significance level 
+  for variable selection} (\code{alpha}) \hlc{is kept at the default value $1$.}
   % stopping criteria:
@@
 diagonal), indicating that all four models are sufficiently well calibrated. 
-\hl{This is also supported by the corresponding probability integral transform}
-(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) \hl{histograms for station
+\hlc{This is also supported by the corresponding probability integral transform}
+(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) \hlc{histograms for station
 Axams in Supplement 2 (Stationwise Evaluation) which contains a more 
@@
 rather than relying on just the one fixed test set for the years 2009--2012.
-\hl{For each setting in the considered 10 times 7-fold cross-validation the 
+\hlc{For each setting in the considered 10 times 7-fold cross-validation the 
 available 28~years are split into 7~subsets of 4~randomly selected and 
@@
 \subsubsection*{Variable Selection}
-\hl{Generally, there are many possibilities how to specify the variables that are
+\hlc{Generally, there are many possibilities how to specify the variables that are
 to be included in a model. Especially for a low number of variables all types
@@
 \cite{Goicoa+Adin+Ugarte:2018}, \cite{Ugarte+Adin+Goicoa:2017}).
-\hl{However, in the application presented in} Section~\ref{sec:precipitation}
-\hl{alltogether 80 variables are available. Even if only paired interactions were
+\hlc{However, in the application presented in} Section~\ref{sec:precipitation}
+\hlc{alltogether 80 variables are available. Even if only paired interactions were
 considered this would already lead to 3160 additional covariates in the model
@@
 \subsubsection*{Tuning parameters}
-\hl{Setting tuning parameters of a forest model can have a major influence
+\hlc{Setting tuning parameters of a forest model can have a major influence
 not only on the accuracy of the model but also on the computational 
 complexity. 
-For the application in} Section~\ref{sec:precipitation} \hl{we have 
+For the application in} Section~\ref{sec:precipitation} \hlc{we have 
 selected the values of these parameters} (listed in Section~\ref{sec:evaluation}) 
@@
 
-\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hl{recommend to 
+\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hlc{recommend to 
 build full-grown trees. Therefore, low values should be chosen for the stopping 
 criteria} \code{minsplit}, \code{minbucket} and \code{alpha}.
-\hl{However, in a distributional model it also has to be considered that a 
+\hlc{However, in a distributional model it also has to be considered that a 
 minimal number of observations is required to allow for a reasonable 
@@
 distributional forest in} Section~\ref{sec:precipitation} is set to
-50 (\code{minbucket}) \hl{and the minimal number of observations on which
+50 (\code{minbucket}) \hlc{and the minimal number of observations on which
 parameters are estimated is set to 20} (\code{minsplit}).
@@
 As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
-\hl{random forests do not overfit, not even for an increasing number of 
-trees} (\code{ntree}). \hl{Hence, it is recommended to build forests consisting 
+\hlc{random forests do not overfit, not even for an increasing number of 
+trees} (\code{ntree}). \hlc{Hence, it is recommended to build forests consisting 
 of a large number of trees as this generally leads to an increase 
@@
 complexity and accuracy''} \citep{Biau+Scornet:2016}.
-\hl{Following this advice we chose to build forests consisting of 100 trees.}
+\hlc{Following this advice we chose to build forests consisting of 100 trees.}
 
 \subsubsection*{East Tyrol}
-Figure~\ref{fig:all-map} \hl{shows that, contrary to the rest of Tyrol, 
+Figure~\ref{fig:all-map} \hlc{shows that, contrary to the rest of Tyrol, 
 for the majority of stations in East Tyrol the prespecified GAMLSS performs
@@
 (Stationwise Evaluation). There, the exact same evaluations as in} 
-Section~\ref{sec:axams} \hl{for station Axams have been carried out 
+Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
 for 14 other stations.}
@@
 %% choice of station Axams mit Verweis auf Supplement stationwise
-\hl{Axams was chosen for the more extensive evaluations presented
+\hlc{Axams was chosen for the more extensive evaluations presented
 in this paper as it is the observation station being closest 
@@
 the forest does not perform better than the other models in the setting 
-presented in} Figure~\ref{fig:all-map}. \hl{It shows that the differences in 
+presented in} Figure~\ref{fig:all-map}. \hlc{It shows that the differences in 
 performance even out in the cross-validation framework where the forest 
@@
 %% choice of distribution mit Verweis auf Supplement distributions
-\hl{Choosing an adequate distributional family is a crucial step and forms
+\hlc{Choosing an adequate distributional family is a crucial step and forms
 the basis required to build a well fitting model. Therefore, this
@@
 earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
-\hl{But to test for robustness against distributional misspecifation 
+\hlc{But to test for robustness against distributional misspecifation 
 different distributional assumptions have been considered as well. 
 In Supplement 1 (Different Response Distributions) the exact
-same evaluations as in} Section~\ref{sec:all} \hl{are carried out for 15
+same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
 observation stations, once employing a Gaussian hurdle model and once
@@
 %% GAMLSS errors with references gamboostLSS Hofner et al.
-As stated by \cite{Hofner+Mayr+Schmid:2016} \hl{the AIC-based variable selection 
-methods implemented in the} \textsf{R} package \textbf{gamlss} \hl{``can be unstable, 
+As stated by \cite{Hofner+Mayr+Schmid:2016} \hlc{the AIC-based variable selection 
+methods implemented in the} \textsf{R} package \textbf{gamlss} \hlc{``can be unstable, 
 especially when it comes to selecting possibly different sets of variables for 
 multiple distribution parameters.'' We have noticed computational problems when
-applying the function} \textbf{gamlss} \hl{in certain settings within the cross-validation
+applying the function} \textbf{gamlss} \hlc{in certain settings within the cross-validation
 framework as it did not succeed in fitting a model. In this cases the method
@@
 $$
-\hl{where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
+\hlc{where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
 type of the transformation function $v_l$ depends on the type of the split 
@@
 \cite{Strasser+Weber:1999} the test statistic can be standardized. 
-\hl{The observed multivariate linear statistic $t_l$ which is either an
+\hlc{The observed multivariate linear statistic $t_l$ which is either an
 $k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of $Z_l$,
@@
 $$
-\hl{where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.} Alternatively,
+\hlc{where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.} Alternatively,
 the maximum of the absolute values of the standardized linear statistic
@@
 If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
-is chosen as splitting variable. \hl{Otherwise no further split is made 
+is chosen as splitting variable. \hlc{Otherwise no further split is made 
 in this node as the stopping criterion of no $p$-values being below the 
@@
 $$
-\hl{for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
+\hlc{for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
 two new subgroups, without any particular ordering, that are defined by 

### 2018-10-12 pkg/disttree/inst/draft/rejoinder.tex
some more revision based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2324 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 70cc66cef334356c8f237acd0a6507e9902cb99b

@@
 
-We have adapted the introduction of a basic EMOS in the main manuscript
-in Section~ ...
-% By a `basic EMOS' we are referring to the linear distributional 
-% regression model based on the idea presented by Gneiting et al. (2005) 
-% where only the ensemble mean is included for the location parameter 
-% and the ensemble standard deviation for the scale parameter.
+By a `basic EMOS' we are referring to the linear distributional 
+regression model based on the idea presented by Gneiting et al. (2005) 
+applying a left-censored at zero Gaussian distribution where only the 
+ensemble mean is included for the location parameter and the ensemble 
+standard deviation for the scale parameter.
+We have introduced the model which we are referring to as a basic EMOS 
+in detail in the main manuscript in the first paragraph of Section~3.
 
@@
 distribution. We have revised the introduction of the EMOS model in 
-the main manuscript and hope this now explains the model better.
-
-\textit{FIX ME: add formula for EMOS?}
-
-\begin{reto}
-Chen ist auf jeden Fall non-sense. Es gibt viele Papers die
-sowas machen, die Frage ist, ob es 10 Referenzen braucht (unter
-anderem macht das Jakob, Manu, ich, die Gneiting-Gruppe bzw.
-Hamill-Gruppe hat auch was in die Richtung aber mit Gamma \dots).
-
-Effizienter waere es wohl kurz das hier EMOS genannte Modell (was
-eigentlich einfach ein non-homogeneous regression Modell ist)
-im Appendix kurz hinzuschreiben?
-
-    \begin{equation}
-        \begin{split}
-            y^{1/p} = \mathcal{N}_0(\mu, \sigma) \\
-            \mu = \beta_0 + \beta_1 \cdot \bar{\mbox{tp}^{1/p}} + \mbox{split?} \\
-            \log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{1/p}) + \mbox{split?}
-        \end{split}
-    \end{equation}
-
-\dots damit k\"onnte man auch auf Kommentar 8 von Reviewer 2 reagieren.
-\end{reto}
+the main manuscript in Section~3 and hope this now explains the model better.
+
+\textit{FIX ME: refer to better explanation in the paper}
 

### 2018-10-15 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2326 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0f09150666914e1d76d6048a151708759a87441a

@@
 mean and variance of a Gaussian response and is particularly
-popular in weather forecasting. More generally, the GAMLSS framework
-allows to establish generalized additive models for location, scale, and shape
-with smooth linear or nonlinear effects.
+popular in weather forecasting. 
+% More generally, the GAMLSS framework allows to establish generalized additive 
+% models for location, scale, and shape with smooth linear or nonlinear effects.
+Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.
 %
@@
 2012 learned on data from 1985--2008. Observations are non-negative and 
-modeled by a Gaussian distribution, left-censored at 0.
+modeled by a Gaussian distribution left-censored at zero.
 The observations are depicted by crosses and the predicted point mass
@@
 Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
-More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
+More specifically, a Gaussian distribution left-censored at zero, is employed to model 24-hour total precipitation so that the
 zero-censored point mass describes the probability of observing no precipitation 
@@
 The general distributional notation is exemplified in all three steps for the
-Gaussian distribution, left-censored at zero. The latter is employed in the empirical 
+Gaussian distribution left-censored at zero. The latter is employed in the empirical 
 case study in Section~\ref{sec:precipitation} to model power-transformed daily 
@@
 scale parameter $\sigma$ of the statistical model based on the assumption
-of the response following a left-censored at zero Gaussian distribution.
+of the response following a Gaussian distribution left-censored at zero.
 This distributional approach of modeling both parameters allows to correct 
-for possible errors of the ensemble in both, the expectation but also the 
+for possible errors of the NWP ensemble in both, the expectation but also the 
 uncertainty of a specific forecast. In order to assure positivity of the 
@@
 with power transformation parameter $\frac{1}{p}$ and 
-regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$.  
-\hlc{(Hereafter, this model is going to be referred to as a basic EMOS model.)}
+regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
+\hlc{hereafter referred to as a basic EMOS model.}
 
@@
 and spreads of various NWP outputs. The predictive performance of the forest is
-compared to three other zero-censored Gaussian models: (a)~a standard basic
+compared to three other zero-censored Gaussian models: (a)~a basic
 EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
@@
 
-\item \emph{EMOS:} Standard ensemble model output statistics
+\item \emph{EMOS:} The basic ensemble model output statistics
   models use the ensemble mean of total precipitation as regressor in the location
@@
 the stations in East Tyrol. These results are presented in Supplement 2 
-(Stationwise Evaluation). There, the exact same evaluations as in} 
+(Stationwise Evaluation) where the exact same evaluations as in} 
 Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
@@
 decision should be made thoughtfuly. 
-In the presented application we applied a Gaussian left-censored at
-zero distribution as this has proven to be a well working choice in
+In the presented application we applied a Gaussian distribution left-censored 
+at zero as this has proven to be a well working choice in
 earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
-\hlc{But to test for robustness against distributional misspecifation 
+\hlc{To test for robustness against distributional misspecifation 
 different distributional assumptions have been considered as well. 
@@
 observation stations, once employing a Gaussian hurdle model and once
-with a left-censored at zero logistic distribution in order to account
+with a logistic distribution left-censored at zero in order to account
 for heavier tails. Under both assumptions the obtained results 
-are similar to those presented in the paper applying a left-censored
-at zero Gaussian distribution. However, comparing the two censored distributions,
+are similar to those presented in the paper applying a Gaussian distribution
+left-censored at zero. However, comparing the two censored distributions,
 at many stations the models perform not as well under the logisctic distribution 
@@
 As stated by \cite{Hofner+Mayr+Schmid:2016} \hlc{the AIC-based variable selection 
-methods implemented in the} \textsf{R} package \textbf{gamlss} \hlc{``can be unstable, 
-especially when it comes to selecting possibly different sets of variables for 
+methods implemented in the} \textsf{R} package \textbf{gamlss} \hlc{``[...] can be 
+unstable, especially when it comes to selecting possibly different sets of variables for 
 multiple distribution parameters.'' We have noticed computational problems when
 applying the function} \textbf{gamlss} \hlc{in certain settings within the cross-validation
-framework as it did not succeed in fitting a model. In this cases the method
+framework as it did not succeed in fitting the model. In these cases the method
 was not taken into consideration in the comparison of all applied models.}
@@
 
-\section{PIT histograms}
-\label{app:pit}
-
-\begin{figure}[b!]
-\centering
-%\begin{subfigure}{0.8\textwidth}
-<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7>>=
+%\section{PIT histograms}
+%\label{app:pit}
+
+%\begin{figure}[b!]
+%\centering
+%%\begin{subfigure}{0.8\textwidth}
+<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7, eval=FALSE>>=
 
@@
 @
-%\end{subfigure}
-\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
-\end{figure}
-
-As an alternative to QQ~plots based on randomized quantile residuals,
-probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
-histograms are a commonly used tool to assess how well-calibrated distributional 
-fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
-probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
-some deviations the overall calibration of all models is acceptable.
+%%\end{subfigure}
+%\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
+%\end{figure}
+
+%As an alternative to QQ~plots based on randomized quantile residuals,
+%probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
+%histograms are a commonly used tool to assess how well-calibrated distributional 
+%fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
+%probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
+%some deviations the overall calibration of all models is acceptable.
 

### 2018-10-15 pkg/disttree/inst/draft/rejoinder.tex
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2326 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0f09150666914e1d76d6048a151708759a87441a

@@
 \usepackage{sfmath,hyperref,color,enumitem,soul}
+%% additional style file to include bibliograph in class uibklttr
+%\usepackage{letterbib}
+%\usepackage{natbib,hyperref}
 
@@
 
+\item The evaluations performed for all 95 observation stations 
+in the main manuscript have been carried out under different 
+distribution assumptions for 15 observation stations in order to 
+test for robustness against distributional misspecifications. 
+Results are provided in an online supplement 
+(Supplement 1: Different Response Distributions) 
+which comes with our revised manuscript.
+
 \item The more extensive evaluations for station Axams have been 
 carried out for 14 other observation stations. Results are 
-provided in an online supplement which comes with our revised manuscript.
-
-\item The evaluations performed for all 95 observation stations 
-have been carried out under different distribution assumptions 
-for the response variable in order to test for robustness against
-distributional misspecifications. Results are provided in an 
-online supplement.
+provided in an online supplement 
+(Supplement 2: Stationwise Evaluation)
+which comes with our revised manuscript.
 
 \item The mathematical notation in the Appendix is now more precise.
+
+\item In the calculation of variable importance we have detected some inconsistencies.
+Thus, we have revised the applied method to evaluate this measurement and recalculated
+variable importance for the distributional forest.
 
@@
 
-The required size of the training data depends heavily on the complexity 
-of the model, however, in different applications which we have tested the 
-distributional forest on it performed quite well with at least 50 observations. \\
+The required size of the training data heavily depends on the complexity 
+of the model, however, in different applications where we have tested the 
+distributional forest it performed quite well with at least 50 observations. \\
 Applying the Law of Large Numbers it can be shown that random forests do not overfit, 
-not even for an increasing number of trees (Breiman (2001), Hastie et al. (2001), 
-Biau et al. (2016)). Therefore, it is recommended to build a large number of trees
-as this generally leads to more accurate predictions. However, ``the computational cost 
-for inducing a forest increases linearly with the number of trees, so a good choice 
-results from a trade-off between computational complexity and accuracy'' 
-(Biau et al. (2016)). Following this advice, we decided to build forests consisting 
-of 100 trees.\\
-Moreover, Hastie et al. (2001) recommend to build trees of maximal depth:
+not even for an increasing number of trees (Breiman 2001, Hastie et al. 2001, 
+Biau et al. 2016). Therefore, it is recommended to build forests with a large number 
+of trees as this generally yields more accurate predictions. 
+However, ``[\dots] the computational cost for inducing a forest increases linearly 
+with the number of trees, so a good choice results from a trade-off between 
+computational complexity and accuracy'' (Biau et al. 2016, p.~205). Following this 
+advice, we decided to build forests consisting of 100 trees.\\
+Moreover, Hastie et al. (2001, p.~596) recommend to build trees of maximal depth:
 ``Our experience is that using full-grown trees seldom costs much, and 
@@
 These tuning parameters are now also listed in the main manuscript in Sections~3.2.
-and discussed in Section~4..
+and discussed in Section~4.
 
@@
 
-We have measured fitting time and prediction time in our 
+We have measured fitting time and prediction time for our 
 application. For single predictions the distributional forest 
@@
 the GAMLSS models or the EMOS. The exact results and times can 
-be found in an online supplement.
+be found in Supplement 2 (Stationwise Evaluation).
 
@@
 We have rephrased this as 'expert knowledge in the application domain'
-in Section~2.2..
+in Section~2.2.
 
@@
 three of the authors, but not based on any other particular 
-features. Furthermore, it has been checked, whether it 
+features. Furthermore, it has been checked whether it 
 is an adequate representative of all stations regarding 
 the results of our evaluations. This is now also 
-confirmed by further analysis in the online supplement which
-comes along with our manuscript and discussed in Section~4..
+confirmed by further analysis in Supplement 2 (Stationwise Evaluation) 
+and discussed in Section~4.
 
@@
 
-East Tyrol is located south of the main Alpine Ridge and exhibits different
-climatological characteristics than the part north of the Alpine Ridge.  It
-seems that the choice of covariates and their interactions we made for the
-prespecified GAMLSS method is particularly adequate for the region of East
-Tyrol for this specific analysis. However, the advantages of the prespecified
-GAMLSS models in this particular setting (training: 1985--2008, validation:
-2009--2012) even out in a cross-validation framework for most of the stations
-in East Tyrol. These results are presented in the online supplement which comes
-along with our manuscript showing that the forest performs equally well as the
-prespecified GAMLSS and better than the EMOS at most of the stations.
-Moreover, this point is now also discussed in Section~4. in the main manuscript.
+This seems to result from the one specific training and validation period 
+(training: 1985--2008, validation: 2009-2012).
+The advantages of the GAMLSS model even out at most of the stations when looking 
+on the results of a cross-validation. We have extended Section~4 to discuss this 
+issue. The results of a the extended cross-validation for a set of $15$ stations 
+can be found in Supplement 2 (Stationwise Evaluation).
 
@@
 
-We have rephrased this part of the caption to 'Observations are non-negative 
+We have rephrased this part of the caption to ``Observations are non-negative 
 and modeled by a Gaussian distribution, left-censored at 0.
 The observations are depicted by crosses and the predicted point mass
-from the model by filled circles.' to clearly distinguish between the model
+from the model by filled circles.'' to clearly distinguish between the model
 and the observations.
@@
 
-We have rephrased this the second paragraph of Section~2.3. which
-included this sentence.
+We have rephrased the second paragraph of Section~2.3.
 
@@
 prediction is similar for all methods. The exact results 
-concerning fitting and prediction times can be found in an online 
-supplement.
+concerning fitting and prediction times can now be found in
+Supplement 2 (Stationwise Evaluation).
 
@@
 the manuscript in Section~3.2., 'Distributional forest' and
-also discussed the choice of these values in Section~4. and
+also discussed the choice of these tuning parameter in Section~4. and
 in the answer to the second point of the associate editor.
@@
 used for the prespecified GAMLSS but not equivalent.
-The use of different variations of the same covariate (cf., Table 2)
+The selection of different variations of the same covariate (cf., Table 2)
 is not surprising as the covariates are often highly correlated.
+Moreover, in the distributional forest for each node only a randomly
+drawn subset of covariates is considered. Hence, if for example the most 
+influential covariate is not included in this subset for a specific node 
+a similar covariate is very likely to be selected instead.
 Compared to the prespecified GAMLSS the forest seems to depict some
@@
 of (the importance of) interactions between different covariates.
-
-\begin{reto}
-Oder sowas \dots Oder g\"abe es daf\"ur tests in W\"aldern?
-\end{reto}
 
@@
 
-Apologies for the unclear notation in Appendix A. This has been 
-revised in the manuscript, also regarding your comments 17--20,
-and we hope that it now clarifies the content.
-
+Apologies for the unclear notation in Appendix A.
 Depending on whether the split variable $Z_l$ is a numeric 
@@
 $k$~-~or~$k \cdot H$~-~dimensional vector.
+We have revised the corresponding parts in the manuscript (also 
+regarding your comments 17--20) and we hope that it now clarifies 
+the content.
 
@@
 
-Apologies again for the confusing notation. Here we are 
-actually referring to the linear statistic $t_l$ (instead of $t$) 
-which is the observed value of $T_l$, the test statistic for 
-the $l$-th split variable $Z_l$.
+Here we are actually referring to the linear statistic $t_l$ (instead of $t$) 
+which is the observed value of $T_l$, the test statistic for the $l$-th split 
+variable $Z_l$. This has been adjusted in the Appendix of the manuscript.
 
@@
 are considered, i.e. $t_l$, $\mu_l$ and $\sigma_l$. Thank you for
-pointing this out, now the missing indices are set.
+pointing this out, the missing indices have been added in the
+Appendix of the manuscript.
 
@@
 
-Again, we are sorry for a confusing explanation. The two subgroups 
-are not ordered which is also not necessary as only the sum of 
-scores over the two subgroups is considered for the linear 
-statistic $T_{l^{\ast}}^{qr}$.
+The two subgroups are not ordered which is also not necessary as only 
+the sum of scores over the two subgroups is considered for the linear 
+statistic $T_{l^{\ast}}^{qr}$. We have adjusted the Appendix of
+the manuscript to clarify this issue.
 
@@
 order to account for heavier tails. The results of these evaluation
-can be found in the online supplement 1 (``Different Response Distributions''). 
-Overall, it is shown that under the logistic distribution the methods perform 
-slightly worse while the hurdle model leads to small improvement in some settings,
-however, this comes along with broader variability. The distributional 
-forest is proven to be very robust against distributional misspecifications.
-Based on these results we decided to keep the distributional assumption
-considered in the main manuscript but we have also discussed the tested
-alternatives in Section~4..
+can be found in Supplement 1 (Different Response Distributions). 
+Overall, the distributional forest is proven to be very robust against 
+distributional misspecifications. Therefore, we decided to keep the 
+distributional assumption considered in the main manuscript but we have 
+also discussed the tested alternatives in Section~4.
 
@@
 requiring no meteorological knowledge about the atmospheric processes 
-which drive formation of precipitation'' in order to clarify that
-this refers only to the variable prespecification in this application.
+which drive formation of precipitation for the model specification.'' 
+in order to clarify that this refers only to the variable prespecification 
+in this application.
 
@@
 statistics (MOS) for several hundreds of observation stations 
-(\hl{\texttt{FIXME: reference Reto}})
+(e.g., Wahl 2014, Rasp and Lerch 2018)
 the distributional forest provides a good and robust alternative. 
@@
 a model applying variable selection based on cross-validation.
+Execution times have been measured and can be found
+in Supplement~2 (Stationwise Evaluation).
 
@@
 description of the cross-validation setup third paragraph of 
-Section~3.3..
+Section~3.3.
+
 We have chosen two different setups: First we learned the model
@@
 in order to provide a classic time ordered setup. 
-However, following (\hl{\texttt{FIXME:} reference Reto})
-no influence of time is to be expected. This is also supported by an 
-autocorrelation coefficient of -0.007 of the out-of-bag quantile 
+However, no influence of time is to be expected which is also supported by 
+an autocorrelation coefficient of -0.007 of the out-of-bag quantile 
 residuals of the distributional forest model for station Axams for
@@
 % ac <- acf(qqrplot(pit_df)$residuals)
-Then, to evaluate the models in a broader framework a cross-validation 
-with the above described choice of folds was applied, for station
-Axams in the main manuscript and for 14 further stations in 
-Supplement 2 (Stationwise Evaluation).
+To evaluate the models in a broader framework a cross-validation 
+with the above described choice of folds was applied. The results
+for station Axams are included in the main manuscript, results
+for 14 additional stations can be found in
+Supplement~2 (Stationwise Evaluation).
 
@@
 
-As mentioned in point 5 and also discussed in Section~4, the high
-number of covariates (80) restricts the possibilities of variable selection
-and specification. For an AIC or BIC based selection an evaluation
-of 80 main effects for two parameters would make this a very complex
-procedure. Moreover, Hofner et al. (2016) stated that the 
+As mentioned in our response to your comment 5, and also discussed in Section~4, 
+the high number of covariates (80) restricts the possibilities of variable 
+selection and model specification. For an AIC or BIC based selection an 
+evaluation of 80 main effects for two parameters would make this a very complex
+procedure. Moreover, Hofner et al. (2016, p.~3) stated that the 
 AIC-based variable selection methods implemented in the \textsf{R} 
-package \textbf{gamlss} ``can be unstable, especially when it comes 
+package \textbf{gamlss} ``[...] can be unstable, especially when it comes 
 to selecting possibly different sets of variables for 
 multiple distribution parameters.'' For that reason we decided
-not to choose such a selection method.
+not to apply such a selection method.
 
@@
 standard deviation for the scale parameter.
-We have introduced the model which we are referring to as a basic EMOS 
+We now introduce the model which we are referring to as a `basic EMOS'
 in detail in the main manuscript in the first paragraph of Section~3.
@@
 
-As now discussed in Section~4 including only all pairwise interactions
+Including only pairwise interactions
 of the 80 available variables would already lead to 3160 additional
@@
 any interactions in the boosting algorithm.
+This issue is now addressed and discussed in Section~4.
 % factorial(80)/factorial(80-2)/factorial(2)
@@
 not even for a high number of trees. However, computational costs increase 
-linearly with the number of trees (Biau et al. (2016)).
+linearly with the number of trees (Biau et al. 2016).
 Therefore, we followed these recommendations of choosing a high number of
-trees, but still not too high in order to keep the computation time at
+trees, but still small enough to keep the computation time at
 a reasonable level. Moreover, 100 is a commonly used number of trees that
-seems to work very well. Of course, more trees could be applied but 
-the performance is usually not significantly worse with 100 trees.
-Tuning parameters of forest models, including the number of 
+seems to work very well. A larger number of trees could be used but 
+the performance is usually not significantly better compared to a forest with 
+100 trees. Tuning parameters of forest models, including the number of 
 trees (\texttt{ntree}), are now discussed in Section~4 
@@
 
-We used ``EMOS'' as an expression for a standard post-processing
-model output statistics (MOS, Gneiting et al. (2005)) with a censored 
-distribution. We have revised the introduction of the EMOS model in 
-the main manuscript in Section~3 and hope this now explains the model better.
-
-\textit{FIX ME: refer to better explanation in the paper}
+The EMOS as originally suggested by Gneiting et al. (2005) can be seen
+as one of the standard ensemble model output statistics (or post-processing)
+approaches. This concept is well known within the post-processing community.
+We have revised the introduction of our manuscript and introduce the `basic EMOS'
+in more explicitly and in more detail (cf., Section~3).
 
@@
 which contains both, PIT histograms and residual QQ plots, 
-for 15 stations. In the manuscript we decided to show only the 
-residual QQ plot as it allows for a better analysis of tail 
+for 15 stations. In the manuscript we decided to only show the 
+residual QQ plot as it allows for a better analysis of the tail 
 behavior than the PIT histogram.
@@
 
-Figure 6 shows the 10 most important covariates of the forest model
-in one particular setting (station Axams, learned on 1985--2008). 
-Variable importance has been calculated for all covariates, however,
-we decided to present only the top 10 as all following covariates 
-are on a rather low and similar level not providing further useful
-information. But this top 10 list is not considered for any 
-specification of the boosted GAMLSS for which all available 
-covariates are included. As for the distributional forest, 
-we have also calculated variable importance based on mean increase 
-of out-of-bag CRPS for the gamboostLSS model for the exact same 
-setting (station Axams, learned on 1985--2008). The resulting 
-top 10 variables are listed in the following barplot:
+Figure 6 is only a visualization of the 10 most important covariates 
+of the forest model in one particular setting (station Axams, learned 
+on 1985--2008). However, this top 10 list is not considered for any specification 
+of the boosted GAMLSS for which all available covariates are included. As 
+for the distributional forest, we have also calculated variable importance 
+based on mean increase of out-of-bag CRPS for the gamboostLSS model for 
+the exact same setting (station Axams, learned on 1985--2008). The resulting 
+top 10 variables are listed in the following barplot.
 
 \begin{center}
-\includegraphics[width = 0.4\textwidth]{varimp_gb.jpeg}
+\includegraphics[width = 0.7\textwidth]{varimp_gb.jpeg}
 \end{center}
@@
 are not exactly the same ones and the influence of \texttt{tp\_max} is clearly
-stronger in the boosted GAMLSS.
+stronger in the boosted GAMLSS. This can be explained by the random selection
+of possible split variables in each node of the distributional forest. Thus,
+if \texttt{tp\_max} is not available, a similar covariate such as \texttt{tp\_min}
+or \texttt{tp\_mean} are very likely to be selected instead while in the boosting
+algorithm \texttt{tp\_max} is always a possible choice.
 
@@
 
+%\bibliographystyle{jss}
+%\bibliography{ref.bib}
+
 \noindent{\large\textbf{References}}
@@
 \item[] Chen, J., Brissette, F. P. and Lhi, Z. (2014). Postprocessing 
-of ensemble weather forecasts using a stochastic weather generator, 
+of Ensemble Weather Forecasts Using a Stochastic Weather Generator, 
 \textit{Monthly Weather Review}.
 
-\item[] Goicoa, T., Adin, A., Ugarte, M. D. and Hodges, J. S. (2017). 
-In spatio-temporal disease mapping models, identifiability constraints 
-affet PQL and INLA results, \textit{Stochastic Environmental Research 
-and Risk Assessment} p. to appear.
-
-\item[] Ugarte, M. D., Adin, A. and Goicoa, T. (2017). One-dimensional, 
-two-dimensional, and three-dimensional B-splines to specify space-time 
-interactions in bayesian disease mapping: Model fitting and model 
-identifiability, \textit{Spatial Statistics} p. to appear.
+%\item[] Goicoa, T., Adin, A., Ugarte, M. D. and Hodges, J. S. (2017). 
+%In Spatio-Temporal Disease Mapping Models, Identifiability Constraints 
+%Affet PQL and INLA Results, \textit{Stochastic Environmental Research 
+%and Risk Assessment} p. to appear.
+
+%\item[] Ugarte, M. D., Adin, A. and Goicoa, T. (2017). One-Dimensional, 
+%Two-Dimensional, and Three-Dimensional B-Splines to Specify Space-Time 
+%Interactions in Bayesian Disease Mapping: Model Fitting and Model 
+%Identifiability, \textit{Spatial Statistics} p. to appear.
+
+\item[] Goicoa, T., Adin, A., Ugarte, M. D. and Hodges, J. S. (2018). 
+In Spatio-Temporal Disease Mapping Models, Identifiability Constraints 
+Affet PQL and INLA Results, \textit{Stochastic Environmental Research 
+and Risk Assessment} \textbf{32, 3}: 749--770.
+
+\item[] Ugarte, M. D., Adin, A. and Goicoa, T. (2017). One-Dimensional, 
+Two-Dimensional, and Three-Dimensional B-Splines to Specify Space-Time 
+Interactions in Bayesian Disease Mapping: Model Fitting and Model 
+Identifiability, \textit{Spatial Statistics} \textbf{22}: 451--468.
 
 \item[] Wood, S. N., Scheipl, F. and Faraway, J. J. (2013). 
-Straightforward intermediate rank tensor product smoothing in mixed models, 
+Straightforward Intermediate Rank Tensor Product Smoothing in Mixed Models, 
 \textit{Statistics and Computing} \textbf{23}: 341--360.
@@
 
-\item[] Hofner, B., Boccuto, L. and Goeker, M. (2015). 
-Controlling false discoveries in high-dimensional situations: Boosting 
-with stability selection,
-\textit{BMC Bioinformatics} \textbf{16:144}.
+%\item[] Hofner, B., Boccuto, L. and Goeker, M. (2015). 
+%Controlling false discoveries in high-dimensional situations: Boosting 
+%with stability selection,
+%\textit{BMC Bioinformatics} \textbf{16:144}.
 
@@
 \item[] Hastie, T., Tibshirani, R. and Friedman, J. (2001).
-The Elements of Statistical Learning,
+The Elements of Statistical Learning, Second Edition,
 \textit{Springer New York Inc.}.
@@
 
+\item[] Wahl, S. (2015). Uncertainty in Mesoscale Numerical Weather Prediction:
+Probabilistic Forecasting of Precipitation, PhD thesis, 
+\textit{University of Bonn}, p. 1--120.% \url{http://hss.ulb.uni-bonn.de/2015/4190/4190.pdf}.
+
+\item[] Rasp, S., Lerch, S. (2018). Neural Networks for Post-Processing Ensemble
+Weather Forecasts, \textit{Montly Weather Review}, NA.
+%doi:\href{https://doi.org/10.1175/MWR-D-18-0187.1}{10.1175/MWR-D-18-0187.1}.
+
 \end{itemize}

### 2018-10-16 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2327 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c4f83dac6f05d1a0d58f5b78b00dd29d8faf8b49

@@
 perform at least on par and sometimes clearly better while requiring no meteorological
-knowledge about the athmospheric processes which drive formation of precipitation
+knowledge about the atmospheric processes which drive formation of precipitation
 for the model specification. The three alternatives are: a standard ensemble model
@@
 Many statistical weather forecasting models leverage the strengths of modern
-numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}.  
+numerical ensemble prediction systems (EPSs; see \cite{Bauer+Thorpe+Brunet:2015}).
+\hlc{EPSs not only predict the most likely future state of the atmosphere but provide
+information about the uncertainty for a specific quantity and weather situation.
+This is done by running the NWP model several times using slightly perturbed 
+initial conditions and model specifications to account for uncertainties in both,
+the initial state and the model and its parametrizations.
 One frequently used method based on distributional regression models is the 
-ensemble model output statistics (EMOS) approach first proposed by
-\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
+ensemble model output statistics (EMOS) approach first proposed by}
+\cite{Gneiting+Raftery+Westveld:2005} \hlc{to produce high-quality forecasts for
 specific quantities and sites.  
-\hlc{In case of precipitation forecasting EMOS typically uses the ensemble mean of 
-``total precipitation'' (\emph{tp}) forecasts as predictor for the location 
+In case of precipitation forecasting EMOS typically uses the ensemble mean of 
+``total precipitation''} (\emph{tp}) \hlc{forecasts as predictor for the location 
 parameter $\mu$ and the corresponding ensemble standard deviation for the 
@@
 are modeled by the linear functions} 
-$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
-$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
-with power transformation parameter $\frac{1}{p}$ and 
-regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
-\hlc{hereafter referred to as a basic EMOS model.}
-
-\hlc{\textit{FIXME: notation with or withoug power transformation parameter?}}
+$\mu = \beta_0 + \beta_1 \cdot \emph{mean}(\emph{tp})$ and  
+$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \emph{sd}(\emph{tp})$
+with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
+\hlc{hereafter referred to as a basic EMOS model} \citep{Gebetsberger+Messner+Mayr:2017}.
+%$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
+%$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
+%with power transformation parameter $\frac{1}{p}$ and 
+%regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
+%\hlc{hereafter referred to as a basic EMOS model \citep{Gebetsberger+Messner+Mayr:2017}.}
+
+\hlc{\textit{FIXME: notation with or without power transformation parameter?}}
 
@@
 it typically does not perform as well in complex terrain due to unresolved
-effects in the NWP system. For example, in the Tyrolean Alps -- considered
+effects in the NWP system \citep{Bauer+Thorpe+Brunet:2015}. 
+For example, in the Tyrolean Alps -- considered
 in the following case study -- the NWP grid cells of $50 \times 50$
@@
 \hlc{However, in the application presented in} Section~\ref{sec:precipitation}
-\hlc{alltogether 80 variables are available. Even if only paired interactions were
+\hlc{altogether 80 variables are available. Even if only paired interactions were
 considered this would already lead to 3160 additional covariates in the model
@@
 the basis required to build a well fitting model. Therefore, this
-decision should be made thoughtfuly. 
+decision should be made thoughtfully. 
 In the presented application we applied a Gaussian distribution left-censored 
@@
 left-censored at zero. However, comparing the two censored distributions,
-at many stations the models perform not as well under the logisctic distribution 
+at many stations the models perform not as well under the logistic distribution 
 as under the Gaussian distribution. The hurdle model leads to slight improvements 
-in some settings, but this comes with more variablity (especially for EMOS and the 
+in some settings, but this comes with more variability (especially for EMOS and the 
 boosted GAMLSS).
-Alltogether, the distributional forest has proven to be very robust against
+Altogether, the distributional forest has proven to be very robust against
 distributional misspecifications.}
@@
 Gaussian models considered in this paper and generating the corresponding 
-QQ~plots (Figure~\ref{fig:axams-qq}) and
-PIT histograms (Figure~\ref{fig:axams-pit}).
+QQ~plots (Figure~\ref{fig:axams-qq}).
 Full replication of all results can be obtained with

### 2018-10-16 pkg/disttree/inst/draft/rejoinder.tex
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2327 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c4f83dac6f05d1a0d58f5b78b00dd29d8faf8b49

@@
 
-\begin{reto}
-Wo baut man das am besten ein?
-\end{reto}
+Ensemble models are now introduced in more detail in the
+first paragraph of Section~3.
 
@@
 
-\textit{TO DO}
-
-\begin{reto}
-Es gibt da ein klassisches MOS-Paper in welchem Postprocessing
-mit einem Globalmodell und einem Lokalmodell gemacht werden.
-Das ist nicht die Referenz die der Reviewer will.
-Das Problem ist, dass es explizit dazu nix gibt. Das ist einfach
-jedem von uns klar. M\"uller zeigt aber in seinen Figures (f\"ur
-Wind und Temperatur), dass ein Modell mit h\"oherer Aufl\"osung
-einen geringeren Fehler aufweist und sagt folgendes:
-
-``\textit{The largest differences are found in the raw
-model forecast for stations in complex terrain. For wind
-speed, the increase in resolution is more beneficial than
-for temperature but is again negligible for MOS forecasts.
-[\dots]
-A higher resolution will
-further reduce the model bias, especially in complex
-terrain, but this can also be achieved with statistical
-postprocessing if observations are available. Furthermore,
-a postprocessed temperature forecast at 12-km
-resolution outperforms a raw forecast at 3-km, regardless
-of the complexity of the terrain}''.
-
-\begin{verbatim}
-@article{mueller2011,
-    author = {M{\"u}ller, M. D.},
-    title = {Effects of Model Resolution and Statistical Postprocessing on Shelter Temperature and Wind Forecasts},
-    journal = {Journal of Applied Meteorology and Climatology},
-    volume = {50},
-    number = {8},
-    pages = {1627--1636},
-    year = {2011},
-    doi = {10.1175/2011JAMC2615.1},
-}
-\end{verbatim}
-
-Oder Meyers and Steenburgh (2013) der aufzeigt, dass auch der Mensch sich mit
-der Wettervorhersage im Gebirge schwer tut bzw. dort die groessten Challenges
-zu finden sind mit all den lokalen und kleinraeumigen Eigenschaften und Einfluessen.
-Habe das Buch grad nicht rumliegen, ist wohl aber auch nicht zwingend notwendig.
-
-\begin{verbatim}
-@Inbook{meyers2013,
-    author = {Meyers, Michael P. and Steenburgh, W. James},
-    editor = {Chow, Fotini K. and De Wekker, Stephan F.J. and Snyder, Bradley J.},
-    title = {Mountain Weather Prediction: Phenomenological Challenges and Forecast Methodology},
-    bookTitle = {Mountain Weather Research and Forecasting: Recent Progress and Current Challenges},
-    year = {2013},
-    publisher = {Springer Netherlands},
-    pages = {1--34},
-    isbn = {978-94-007-4098-3},
-    doi = {10.1007/978-94-007-4098-3_1},
-}
-\end{verbatim}
-\end{reto}
+This is itself a complex topic with an own scientific community
+focusing on the development and improvement of numerical weather
+forecasts. Bauer, Thorphe, and Brunet (2015) give a good overview
+of the history and open scientific challenges of NWP ensemble forecast
+models including a discussion about unresolved processes and possible
+solutions beyond the current state of NWP models.
+We have added this reference next to the statement you are referring to.
+
+The following figure shows a comparison of the true local topography
+(Robinson et~al. 2014; left)
+and the model topography of the GEFS reforecast model
+($0.5^\circ \times 0.5^\circ$, approx. $50 \times 50~km^2$; right) to illustrate
+the extent of simplification needed in such global weather forecast models.
+
+\begin{center}
+\includegraphics[width = .8\textwidth]{topo-GFS.pdf}
+\end{center}
+
+%\begin{verbatim}
+%@Article{Bauer2015,
+%    author   = {Bauer, Peter and Thorpe, Alan and Brunet, Gilbert},
+%    title    = {The Quiet Revolution of Numerical Weather Prediction},
+%    journal  = {Nature},
+%    year     = {2015},
+%    month    = {Sep},
+%    day      = {02},
+%    volume   = {525},
+%    pages    = {47--55},
+%    doi      = {10.1038/nature14956}
+%}
+%\end{verbatim}
+%Bauer and Thorpe (2015, p.~52): ``\textit{The scientific challenges:
+%[\dots] the uncertainties inherent to physical parameterizations, either
+%from incomplete process understanding or the dilemma of representing
+%the impact of unresolved processes on the resolved scales, may require a
+%fundamentally different approach.''
 
@@
 \begin{center}
-\includegraphics[width = 0.7\textwidth]{varimp_gb.jpeg}
+\includegraphics[width = .7\textwidth]{varimp_gb.pdf}
 \end{center}
@@
 
+\item[] M{\"u}ller, M. D. (2011): Effects of Model Resolution and Statistical
+Postprocessing on Shelter Temperature and Wind Forecasts,
+\textit{Journal of Applied Meteorology and Climatology} 
+\textbf{50 8}: 1627--1636.
+%doi:10.1175/2011JAMC2615.1
+
+\item[] Bauer, P. and Thorpe, A. (2015). The Quiet Revolution of Numerical 
+Weather Prediction,
+\textit{Nature} \textbf{525}: 47--55.
+%doi:10.1038/nature14956
+
 \end{itemize}

### 2018-10-16 pkg/disttree/inst/draft/varimp_gb.R

@@ -61,12 +61,15 @@ vimp_crps_gb <- sort(vimp_crps_gb, decreasing = TRUE)
 
 save(vimp_crps_gb, file = "~/svn/partykit/pkg/disttree/inst/draft/reviews/vimp_crps_gb.rda")
 
+
+## plot
+load("~/svn/partykit/pkg/disttree/inst/draft/reviews/vimp_crps_gb.rda")
 top10_gb <- vimp_crps_gb[1:10]
 
-par(mar = c(2.5,10,1,2))
+par(mar = c(4,10,1,2))
 barplot(rev(top10_gb), 
         horiz = TRUE, las = 1, axes = FALSE,
-        # xlab = "Variable importance: mean decrease in CRPS",
+        xlab = "Variable importance: mean decrease in CRPS",
         font.axis = 3, #list(family="HersheySerif", face=3),
         # xlim = c(0,1.1),
         xlim = c(0,0.4),

### 2018-10-16 pkg/disttree/inst/draft/rejoinder.tex
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2329 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9f936232cf3367f90db1542e75ad04c56fd59c3c

@@
 carried out for 14 other observation stations. Results are 
-provided in an online supplement 
-(Supplement 2: Stationwise Evaluation)
-which comes with our revised manuscript.
-
-\item The mathematical notation in the Appendix is now more precise.
+provided in an online supplement (Supplement 2: Stationwise Evaluation)
+which comes with our revised manuscript. As this also contains
+the results that were presented in Appendix B we have removed this
+section of the appendix.
+
+\item The mathematical notation in Appendix A is now more precise.
 
@@
 
+We have rephrased this part of the abstract now spelling out 'GAMLSS' at 
+its first appearance.
+
 \begin{review}
@@
 on the results of a cross-validation. We have extended Section~4 to discuss this 
-issue. The results of a the extended cross-validation for a set of $15$ stations 
-can be found in Supplement 2 (Stationwise Evaluation).
+issue. The results of an extended evaluation including a cross-validation for a 
+set of $15$ stations can be found in Supplement 2 (Stationwise Evaluation).
 
@@
 Thank you very much for the positive and constructive 
-comments regarding the manuscript. These helped to 
+comments regarding the paper. These helped to 
 substantially improve the manuscript.
@@
 
+This has been changed in the manuscript.
+
 \begin{review}
@@
 \end{review}
+
+This has been changed in the manuscript.
 
@@
 distribution but the observations are not censored, just non-negative
-with a point mass at 0.
+with a point mass at zero.
 
@@
 We have rephrased this part of the caption to ``Observations are non-negative 
-and modeled by a Gaussian distribution, left-censored at 0.
+and modeled by a Gaussian distribution left-censored at zero.
 The observations are depicted by crosses and the predicted point mass
@@
 parameter vector for all observations in the same terminal node.
-This is now also explained at the end of Section~2.2..
+This is now also explained at the end of Section~2.2.
 
@@
 is kept at the default value 1. We have added this information in 
-the manuscript in Section~3.2., 'Distributional forest' and
-also discussed the choice of these tuning parameter in Section~4. and
+the manuscript in Section~3.2, 'Distributional forest', and
+also discussed the choice of these tuning parameter in Section~4 and
 in the answer to the second point of the associate editor.
@@
 a similar covariate is very likely to be selected instead.
-Compared to the prespecified GAMLSS the forest seems to depict some
-features via differences among the different variations of a specific
-covariate (e.g., use \code{tp\_max}, \code{tp\_mean}, and \code{tp\_min}).
-Unfortunately the analysis does not allow to get more insights
-of (the importance of) interactions between different covariates.
 
@@
 which is the observed value of $T_l$, the test statistic for the $l$-th split 
-variable $Z_l$. This has been adjusted in the Appendix of the manuscript.
+variable $Z_l$. This has been adjusted in the appendix of the manuscript.
 
@@
 pointing this out, the missing indices have been added in the
-Appendix of the manuscript.
+appendix of the manuscript.
 
@@
 the sum of scores over the two subgroups is considered for the linear 
-statistic $T_{l^{\ast}}^{qr}$. We have adjusted the Appendix of
+statistic $T_{l^{\ast}}^{qr}$. We have adjusted the appendix of
 the manuscript to clarify this issue.
@@
 
-We have applied the left-censored at zero Gaussian distribution
+We have applied a Gaussian distribution left-censored at zero
 as it has proven to be an adequate choice for this type of 
@@
 same evaluations as presented in the paper but employing a
-Gaussian hurdle model. Additionally, a left-censored at zero logistic
-distribution has been applied as well in the same setting in
+Gaussian hurdle model. Additionally, a logistic distribution 
+left-censored at zero has been applied as well in the same setting in
 order to account for heavier tails. The results of these evaluation
@@
 for the skewness parameter is required. 
-Moreover, applying a transformation together with a censored 
-at zero Gaussian distribution allows for a better comparison 
-to the other models as they can all be built under the same 
-settings.
+Moreover, applying a transformation together with a Gaussian distribution
+censored at zero allows for a better comparison to the other models as 
+they can all be built under the same settings.
 In this study we are thus following Stauffer et al. (2017) 
@@
 We have revised the introduction of our manuscript and introduce the `basic EMOS'
-in more explicitly and in more detail (cf., Section~3).
+more explicitly and in more detail (cf., Section~3).
 
@@
 for the distributional forest, we have also calculated variable importance 
-based on mean increase of out-of-bag CRPS for the gamboostLSS model for 
+based on mean increase of out-of-bag CRPS for the boosted GAMLSS model for 
 the exact same setting (station Axams, learned on 1985--2008). The resulting 

### 2018-10-16 pkg/disttree/inst/draft/rejoinder.Rnw
rejoinder now as .Rnw-file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2330 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3720b1820e7da2193c2409d02aacb1f3a4c841d0

--- 2018-10-16 pkg/disttree/inst/draft/rejoinder.tex
@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 \begin{letter}{
@@
 \begin{center}
-\includegraphics[width = .8\textwidth]{topo-GFS.pdf}
+\setkeys{Gin}{width=0.8\textwidth}
+<<topography, fig=TRUE, echo=FALSE, height=3.5, width=8>>=
+
+# Preamble
+library("sp")
+library("raster")
+library("ncdf4")
+data(StationsTyrol, package = "RainTyrol")
+data(MapTyrol, package = "RainTyrol")
+# data(MapTyrol_border, package = "RainTyrol")
+# Create SpatialPointsDataFrame from station list
+sp <- SpatialPointsDataFrame(subset(StationsTyrol,
+                                    select=c(lon,lat)),
+                             data = subset(StationsTyrol,
+                                           select = -c(lon,lat)),
+                             proj4string = crs(MapTyrol$RasterLayer))
+
+## plot map of Tyrol with all 95 observations
+par(mfrow = c(1, 2), mar = rep(0.1, 4), oma = rep(3, 4))
+
+# limits and axis config
+zlim <- c(400, 3900)
+xat  <- pretty(coordinates(MapTyrol$RasterLayer)[,1], 5)
+yat  <- pretty(coordinates(MapTyrol$RasterLayer)[,2], 5)
+cols <- gray.colors(100, start = 0.95, end = 0.05)
+
+# Plotting "real" topography
+raster::image(MapTyrol$RasterLayer, col = cols,
+              main=NA, ylab = "Latitude", xlab = "Longitude",
+              xlim = c(9.8,13.2), ylim = c(46.6, 47.87), zlim = zlim,
+              xaxt = "n", yaxt = "n")
+box()
+plot(MapTyrol$SpatialPolygons, add = TRUE)
+points(sp, pch = 19, cex = .5)
+axis(side = 1, at = xat); axis(side = 2, at = yat)
+mtext(side = 2, line = 2, cex = .8, "Latitude")
+
+# Loading and plotting regular_ll GFS topography
+##gfs <- raster("gfs_topo.nc")
+##gfs_resampled <- raster::resample(gfs, MapTyrol$RasterLayer)
+raster::image(raster("gfs_topo.nc"), col = cols,
+              main=NA, ylab = "Latitude", xlab = "Longitude",
+              xlim = c(9.8,13.2), ylim = c(46.6, 47.87), zlim = zlim,
+              xaxt = "n", yaxt = "n")
+box()
+plot(MapTyrol$SpatialPolygons, add = TRUE)
+points(sp, pch = 19, cex = .5)
+axis(side = 1, at = xat); axis(side = 4, at = yat)
+mtext(side = 4, line = 2, cex = .8, "Latitude")
+
+mtext(side = 1, line = 2, cex = .8, outer = TRUE, "Longitude")
+# Title
+mtext(side = 3, line = 1, cex = 1.2, font = 2, outer = TRUE,
+      "Comparison of Real Topography and GFS Model Topography")
+@
 \end{center}
@@
 \begin{center}
-\includegraphics[width = .7\textwidth]{varimp_gb.pdf}
+\setkeys{Gin}{width=0.6\textwidth}
+<<varimp_gb, fig=TRUE, echo=FALSE, height=4, width=6>>=
+if(file.exists("vimp_crps_gb.rda")){
+  load("vimp_crps_gb.rda")
+} else {
+
+  nperm <- 50
+  library("gamboostLSS")
+  library("crch")
+  library("scoringRules")
+  library("RainTyrol")
+
+  # set function for parallelization
+  applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, detectCores() - 1))
+
+  # model
+  load("Axams_24to4.rda")
+  gb <- Axams_24to4$gb
+
+  # data
+  data("RainTyrol")
+  RainData <- RainTyrol[RainTyrol$station == "Axams", ]
+  rownames(RainData) <- c(1:NROW(RainData))
+
+  # learning data: 24 years (1985 - 2008, both inlcuded)
+  # testing data: 4 successive years (2009, 2010, 2011, 2012)
+  # learndata <- RainData[RainData$year < 2009,]
+  testdata <- RainData[RainData$year %in% c(2009, 2010, 2011, 2012),]
+
+  ## compute variable importance for the gamboostLSS object 'gb'
+
+  meancrps <- function(permute = NULL, newdata = testdata) {
+    if(!is.null(permute)) newdata[[permute]] <- sample(newdata[[permute]])
+    m <- predict(gb, newdata = newdata, parameter = "mu", type = "response")
+    s <- predict(gb, newdata = newdata, parameter = "sigma", type = "response")
+    mean(crps_cnorm(newdata$robs, location = m, scale = s, lower = 0))
+  }
+
+  # apply for all covariates except for dswrf_mean_min and 
+  # dswrf_sprd_min (columns 30 and 33) as they are always 0
+
+  risklist <- applyfun(1:nperm, 
+                       function(i){
+                         set.seed(i)
+                         sapply(c(5:29, 31, 32, 34: ncol(testdata)), meancrps)
+                       })
+  risk <- Reduce("+", risklist) / length(risklist)
+
+  names(risk) <- names(testdata)[c(5:29, 31, 32, 34: ncol(testdata))]
+  vimp_crps_gb <- risk - meancrps(newdata = testdata)
+  vimp_crps_gb <- sort(vimp_crps_gb, decreasing = TRUE)
+
+  save(vimp_crps_gb, file = "vimp_crps_gb.rda")
+
+}
+
+top10_gb <- vimp_crps_gb[1:10]
+
+par(mar = c(4,10,1,2))
+barplot(rev(top10_gb), 
+        horiz = TRUE, las = 1, axes = FALSE,
+        xlab = "Variable importance: mean decrease in CRPS",
+        font.axis = 3, #list(family="HersheySerif", face=3),
+        # xlim = c(0,1.1),
+        xlim = c(0,0.4),
+        names.arg = gsub("pow", "", names(rev(top10_gb))))
+axis(1, at = seq(0,1.6,0.1), las = 1, mgp=c(0,1,0))
+@
 \end{center}

### 2018-10-17 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2328 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fb83d2e9bc0fca7c239ad56c9b0d75e6f91707ab
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2331 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1ae54badfd71bcdb3a85535a769dfad222b51798

@@
 % models for location, scale, and shape with smooth linear or nonlinear effects.
-Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.
+\hlc{Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.}
 %
@@
 \caption{\label{fig:axams-dist}\hlc{Total precipitation predictions by a 
-distributional forest at station Axams for July 24 in 2009, 2010, 2011 and 
+distributional forest at station Axams for July~24 in 2009, 2010, 2011 and 
 2012 learned on data from 1985--2008. Observations are non-negative and 
@@
 $$
-\hlc{Even though this parameter depends on the given $z$ it has to be noted that
+\hlc{Even though this parameter depends on the given $\bold{z}$ it has to be noted that
 it is the same for all other observations that are in the same terminal node of 
-the tree as $z$ and therefore does not necessarily have to be recalculated for
+the tree as $\bold{z}$ and therefore does not necessarily have to be recalculated for
 each of them.}
@@
 only a random subset of the covariates $\bold{Z}$ is considered to reduce the 
-correlation among the trees and to stabilizes the variance of the model.
+correlation among the trees and to stabilize the variance of the model.
 For a simple regression random forest the mean of predictions over
@@
 \hlc{hereafter referred to as a basic EMOS model} \citep{Gebetsberger+Messner+Mayr:2017}.
+
+% FIXME: notation with or without power transformation parameter?
+
+% with:
 %$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
@@
 
-\hlc{\textit{FIXME: notation with or without power transformation parameter?}}
-
 While this approach alone is already highly effective in the plains,
 it typically does not perform as well in complex terrain due to unresolved
-effects in the NWP system \citep{Bauer+Thorpe+Brunet:2015}. 
+effects in the \hlc{NWP system} \citep{Bauer+Thorpe+Brunet:2015}. 
 For example, in the Tyrolean Alps -- considered
@@
 performance of the EMOS by including additional predictor variables, either 
-from local meteorological observations or an NWP model. Unfortunately, it is typically unknown 
-which variables are relevant for improving the predictions. Simply including
-all available variables may be computationally burdensome and
+from local meteorological observations or an NWP model. Unfortunately, it is 
+typically unknown which variables are relevant for improving the predictions. 
+Simply including all available variables may be computationally burdensome and
 can lead to overfitting but, on the other hand, excluding too many variables
@@
 
-\item \emph{EMOS:} The basic ensemble model output statistics
+\item \emph{EMOS:} \hcl{The basic} ensemble model output statistics
   models use the ensemble mean of total precipitation as regressor in the location
@@
 Therefore, ``a good choice results from a trade-off between computational 
-complexity and accuracy''} \citep{Biau+Scornet:2016}.
+complexity and accuracy''} (\cite{Biau+Scornet:2016}, p.~205).
 \hlc{Following this advice we chose to build forests consisting of 100 trees.}
@@
 Gaussian models considered in this paper and generating the corresponding 
-QQ~plots (Figure~\ref{fig:axams-qq}).
+QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
+(Supplement 2: Stationwise Evaluation).
 Full replication of all results can be obtained with

### 2018-10-17 pkg/disttree/inst/draft/rejoinder.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2331 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1ae54badfd71bcdb3a85535a769dfad222b51798

@@
 }
+
+\definecolor{fixmecol}{rgb}{1,1,0.7}
+\newcommand{\fixme}[1]{{\sethlcolor{fixmecol}\hl{#1}}}
+
 \usepackage{amsmath}
@@
 skewed ensemble forecasts which has to be addressed in some sort.\\
-As you mention one way would be to directly apply a
-distributional family with an additional parameter to capture
-skewness. Another frequently used approach is to transform the 
-data to remove large parts of the skewness. The latter has the 
-advantage that the model assumptions get simpler as no estimate 
-for the skewness parameter is required. 
-Moreover, applying a transformation together with a Gaussian distribution
-censored at zero allows for a better comparison to the other models as 
-they can all be built under the same settings.
-In this study we are thus following Stauffer et al. (2017) 
+\fixme{\textit{FIXME: Should this explanation why there is skewness be included?}}
+
+As you mention one way how to deal with skewness would be to 
+apply a distributional family with an additional parameter to capture
+skewness. However, this can be very complex and thus lead to less well 
+fitting estimators of the location and scale parameter.
+Moreover, the focus of this paper is on the features of the presented 
+distributional forest regarding variable selection and the capability 
+to also model non-linear effects rather than optimizing the model 
+specification for this particular application.
+Transforming the data to remove large parts of the skewness is
+a frequently used approach. Therefore, employing such a transformation 
+allows for a better way of comparing our results to those presented in 
+literature. In this study we are thus following Stauffer et al. (2017) 
 using the power transformation which turned out to be optimal 
@@
 
-This is the trade-off between interpretability and predictive 
-skill. Regression-type models might be easier to interpret, 
-however, if one is mainly interested in predictive skill and
-a good performing fully automated weather forecasting tool the 
-interpretability is not implicitly needed in some cases 
-(machine learning performance vs. interpretability).
-We adapted our manuscript as we definitely want to emphasize that 
-in our application the focus is mainly on prediction but not on 
-interpretation. We hope we could point this out clearly.
+We completely agree with you as that other models such as
+GAMLSS are clearly advantageous regarding interpretability. 
+These models provide a very powerful framework that works
+well in a wide range of applications and research fields, 
+including meteorology. Therefore, our intention is not to present 
+distributional forests as a replacement of these commonly used methods 
+but rather as a well working alternative. Especially if one is mainly 
+interested in predictive skill and a good performing fully automated 
+weather forecasting tool the interpretability is not implicitly needed 
+in some cases (machine learning performance vs. interpretability).
+We adapted our manuscript as we definitely want to emphasize that in our 
+application the focus is mainly on prediction but not on interpretation. 
+We hope we could point this out clearly. \\
+\fixme{\textit{FIXME: Where should this be pointed out in the manuscript? Discussion?}}
 
@@
 \item[] Hastie, T., Tibshirani, R. and Friedman, J. (2001).
-The Elements of Statistical Learning, Second Edition,
+The Elements of Statistical Learning, %Second Edition,
 \textit{Springer New York Inc.}.

### 2018-10-17 pkg/disttree/inst/draft/rejoinder.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2332 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4e2046f29218da7c2bc2cc048097a096f6a0157b
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2333 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 23f1b1b3a505c4e3312e7057cde6de6cbeedeb4a

@@
 
-The skewness results from the characteristics of precipitation sums. 
-Most of the time no or only moderate precipitation amounts are forecasted 
-and observed. In some situations (e.g., days with thunderstorms, cold and 
-warm fronts, \dots) large amounts of precipitation can be observed at 
-specific sites. This is often also depict by the ensemble prediction 
-system where a subset of all members predicts rather large amounts 
-(compared to the overall ensemble mean or median). This yields strongly 
-skewed ensemble forecasts which has to be addressed in some sort.\\
-\fixme{\textit{FIXME: Should this explanation why there is skewness be included?}}
-
-As you mention one way how to deal with skewness would be to 
-apply a distributional family with an additional parameter to capture
-skewness. However, this can be very complex and thus lead to less well 
-fitting estimators of the location and scale parameter.
+As you mention one way to deal with skewness is to use a distribution
+with an additional skewness parameter (e.g., Scheuerer and Hamill 2015,
+Baran and Nemoda 2016) which increases the complexity of the statistical
+models.
+Another frequently used approach is to apply a
+power-transformation (Stidd 1973, Hutchinson 1998, Stauffer et al. 2017) to
+remove the majority of the skewness from the data. The optimal transformation
+depends on the application and data set, in our case we are following the work
+of Stauffer et al. (2017) who presented a similar application on the very same
+data set. As our results show (Figure~3 in the manuscript) this approach works 
+well for the application presented in our manuscript.
 Moreover, the focus of this paper is on the features of the presented
@@
 specification for this particular application.
-Transforming the data to remove large parts of the skewness is
-a frequently used approach. Therefore, employing such a transformation 
-allows for a better way of comparing our results to those presented in 
-literature. In this study we are thus following Stauffer et al. (2017) 
-using the power transformation which turned out to be optimal 
-for a similar application for the very same study area.
+
+%The skewness results from the characteristics of precipitation sums. 
+%Most of the time no or only moderate precipitation amounts are forecasted 
+%and observed. In some situations (e.g., days with thunderstorms, cold and 
+%warm fronts, \dots) large amounts of precipitation can be observed at 
+%specific sites. This is often also depict by the ensemble prediction 
+%system where a subset of all members predicts rather large amounts 
+%(compared to the overall ensemble mean or median). This yields strongly 
+%skewed ensemble forecasts which has to be addressed in some sort.\\
+%\fixme{\textit{FIXME: Should this explanation why there is skewness be included?}}
+
+%As you mention one way how to deal with skewness would be to 
+%apply a distributional family with an additional parameter to capture
+%skewness. However, this can be very complex and thus lead to less well 
+%fitting estimators of the location and scale parameter.
+%Moreover, the focus of this paper is on the features of the presented 
+%distributional forest regarding variable selection and the capability 
+%to also model non-linear effects rather than optimizing the model 
+%specification for this particular application.
+%Transforming the data to remove large parts of the skewness is
+%a frequently used approach. Therefore, employing such a transformation 
+%allows for a better way of comparing our results to those presented in 
+%literature. In this study we are thus following Stauffer et al. (2017) 
+%using the power transformation which turned out to be optimal 
+%for a similar application for the very same study area.
 
@@
 weather forecasting tool the interpretability is not implicitly needed
-in some cases (machine learning performance vs. interpretability).
-We adapted our manuscript as we definitely want to emphasize that in our 
-application the focus is mainly on prediction but not on interpretation. 
+in some cases (machine learning performance vs. interpretability, e.g.,
+Rasp and Lerch 2018).
+We adapted our manuscript as we want to emphasize that in our
+application the focus is mainly on the predictive performance and less on
+the interpretation of the different effects.
 We hope we could point this out clearly. \\
-\fixme{\textit{FIXME: Where should this be pointed out in the manuscript? Discussion?}}
+\fixme{\textit{FIXME: Where should this be pointed out in the manuscript? 
+In the discussion or after introducing trees and forests? (trade-off between
+flexibility of the model and interpretability)}}
 
@@
 8. I found the competing models to be chosen pretty unfair. 
-First, for the prespecified GAMLSS, the authors use 
-”relevant effects based on expert knowledge”. Why not all? 
+First, for the prespecified GAMLSS, the authors use ``relevant effects based on expert knowledge''. Why not all? 
 Why not making a decision on some information criteria such as AIC/BIC? 
@@
 
+\item[] Scheuerer, M. and Hamill T. M. (2015). Statistical Postprocessing of
+Ensemble Precipitation Forecasts by Fitting Censored, Shifted Gamma
+Distributions,
+\textit{Monthly Weather Review} \textbf{143, 11}: 4578--4596.
+%doi:10.1175/MWR-D-15-0061.1
+
+\item[] Baran, S., and Nemoda, D. (2016). Censored and Shifted Gamma Distribution Based EMOS
+Model for Probabilistic Quantitative Precipitation Forecasting,
+\textit{Environmetrics}, \textbf{27}: 280--292.
+%doi: 10.1002/env.2391.
+
+\item[] Hutchinson M. F. (1998). Interpolation of Rainfall Data with Thin
+Plate Smoothing Splines -- Part II: Analysis of Topographic Dependence,
+\textit{Journal of Geographic Information and
+Decision Analysis} \textbf{2, 2}: 152--167.
+%doi:10.1080/02693799508902045
+
+\item[] Stidd, C. K. (1973). Estimating the Precipitation Climate,
+\textit{Water Resources Research}, \textbf{9, 5}: 1235--1241.
+%doi:10.1029/wr009i005p01235
 \end{itemize}

### 2018-10-19 pkg/disttree/inst/draft/rejoinder.Rnw
now using data from MapTyrol to plot topography
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2337 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 55a805b19e96c8f92ba57febb79c0e09cb46e517

@@
 library("raster")
-library("ncdf4")
+#library("ncdf4")
 data(StationsTyrol, package = "RainTyrol")
@@
 ##gfs_resampled <- raster::resample(gfs, MapTyrol$RasterLayer)
-raster::image(raster("gfs_topo.nc"), col = cols,
+raster::image(MapTyrol$RasterLayerGFS, col = cols,
               main=NA, ylab = "Latitude", xlab = "Longitude",

### 2018-10-19 pkg/disttree/inst/draft/rejoinder.Rnw
Z updates -> please re-read
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2339 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a6f5c0c53c028af27844d28645e082498d06eed2

@@
 \documentclass[american,foldmarks=false,noconfig]{uibklttr}
-%\documentclass{article}
-\usepackage{sfmath,hyperref,color,enumitem,soul}
-%% additional style file to include bibliograph in class uibklttr
-%\usepackage{letterbib}
-%\usepackage{natbib,hyperref}
+\usepackage{amsmath,sfmath,hyperref,color,enumitem,soul}
 
@@
 \makeatletter
-\@setplength{refvpos}{75mm}
+\@setplength{refvpos}{68mm}
 \makeatother
@@
 
-\newenvironment{reto}{
-    \begin{color}{uibkorange}
-    \textbf{Reto:~}
-        \itshape
-}{
-    \end{color}
-}
-
 \definecolor{fixmecol}{rgb}{1,1,0.7}
-\newcommand{\fixme}[1]{{\sethlcolor{fixmecol}\hl{#1}}}
-
-\usepackage{amsmath}
+\newcommand{\fixme}[1]{{\sethlcolor{fixmecol}\hl{\textit{FIXME: #1}}}}
 
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+
 \begin{letter}{
@@
 
-\setkomavar{subject}{Revision of AOAS1804-015: Distributional 
-regression forests for probabilistic precipitation forecasting 
-in complex terrain}
+\setkomavar{subject}{Revision of AOAS1804-015}
 
@@
 Thank you very much for the constructive and helpful feedback from you, 
-the associate editor and the two referees. The manuscript has been 
-improved and additional results haven been obtained from further evaluations.
-
+the associate editor, and the two referees. Based on this feedback, the
+manuscript has been revised and embellished with additional results
+from further evaluations. We feel that this substantially improved our
+contribution.
+%
 The most important changes are:
@@
 
-\item The evaluations performed for all 95 observation stations 
-in the main manuscript have been carried out under different 
-distribution assumptions for 15 observation stations in order to 
-test for robustness against distributional misspecifications. 
-Results are provided in an online supplement 
-(Supplement 1: Different Response Distributions) 
-which comes with our revised manuscript.
-
-\item The more extensive evaluations for station Axams have been 
-carried out for 14 other observation stations. Results are 
-provided in an online supplement (Supplement 2: Stationwise Evaluation)
-which comes with our revised manuscript. As this also contains
-the results that were presented in Appendix B we have removed this
-section of the appendix.
-
-\item The mathematical notation in Appendix A is now more precise.
-
-\item In the calculation of variable importance we have detected some inconsistencies.
-Thus, we have revised the applied method to evaluate this measurement and recalculated
-variable importance for the distributional forest.
+\item \emph{Supplement 1: Different response distributions.}\\
+To assess the appropriateness and robustness of the distributional
+assumption (censored Gaussian) made in the main manuscript, further
+response distributions (two-part Gaussian hurdle; censored
+logistic) are evaluated to potentially better capture lower
+and upper tails, respectively. The supplement follows the
+evaluations performed for all 95 observation stations 
+in the main manuscript using all three response distributions
+for a subset of 15 stations.
+
+\item \emph{Supplement 2: Stationwise evaluation.}\\
+The show that the more extensive cross-validation results -- obtained only
+for one station (Axams) in the main manuscript -- are fairly typical,
+they are extended to 14 other observation stations. This also incorporates
+results that were previously presented in Appendix B which has consequently
+been omitted from the main manuscript.
+
+\item \emph{Section~4: Discussion.}\\
+The concluding section of the main manuscript has been rewritten and
+extended to regarding discussion of: variable selection, tuning parameters,
+distributional specifications (referring to Supplement~1), and regional
+differences (referring to Supplement~2).
+
+\item \emph{Appendix~A: Tree algorithm.}\\
+The exposition has been extended and embellished, in particular regarding
+mathematical notation.
+
+% \item In the calculation of variable importance we have detected some inconsistencies.
+% Thus, we have revised the applied method to evaluate this measurement and recalculated
+% variable importance for the distributional forest.
 
@@
 
-\vspace*{-2.8cm}
-\hspace*{-0.3cm}signature %\includegraphics{zeileis-signature}
+% \vspace*{-2.8cm}
+% \hspace*{-0.3cm} %\includegraphics{zeileis-signature}
 
@@
 The required size of the training data heavily depends on the complexity 
-of the model, however, in different applications where we have tested the 
-distributional forest it performed quite well with at least 50 observations. \\
-Applying the Law of Large Numbers it can be shown that random forests do not overfit, 
-not even for an increasing number of trees (Breiman 2001, Hastie et al. 2001, 
-Biau et al. 2016). Therefore, it is recommended to build forests with a large number 
-of trees as this generally yields more accurate predictions. 
+of the underlying effects. However, in different applications we have tested the 
+distributional forest performed quite well starting from about 50--100 observations.
+
+Applying the Law of Large Numbers it can be shown that random forests do not overfit 
+as the number of trees increases (Breiman 2001, Hastie et al.\ 2001, 
+Biau et al.\ 2016). Therefore, in principle, forests can be built with a very
+large number of trees as this cannot deteriorate the predictions.
 However, ``[\dots] the computational cost for inducing a forest increases linearly 
 with the number of trees, so a good choice results from a trade-off between 
-computational complexity and accuracy'' (Biau et al. 2016, p.~205). Following this 
-advice, we decided to build forests consisting of 100 trees.\\
-Moreover, Hastie et al. (2001, p.~596) recommend to build trees of maximal depth:
+computational complexity and accuracy'' (Biau et al.\ 2016, p.~205). Following this 
+advice, we decided to build forests consisting of 100 trees. Increasing this number
+further would not change the results much.
+
+Moreover, Hastie et al.\ (2001, p.~596) recommend to build trees of maximal depth:
 ``Our experience is that using full-grown trees seldom costs much, and 
@@
 Thus, the stopping criteria \texttt{minbucket} and \texttt{minsplit} (both 
-determining the minimal number of observations in a node) and \texttt{alpha} 
-(the significance level for the statistical tests in the tree algorithm) should 
-be kept rather low. However, in a distributional framework it also has to be 
-considered that the minimal number of observations in a node needs to be high 
-enough in order to allow for estimating the distribution parameters reasonably well.
-In our application we have set \texttt{minsplit} to 50 and \texttt{minbucket} to 20.
+determining the minimal number of observations in a node) should be kept low
+-- but in our context large enough for estimating the distribution parameters
+reasonably well. Thus, in our application we have set \texttt{minsplit} to 50 and
+\texttt{minbucket} to 20. The significance level \texttt{alpha} (for early stopping
+upon non-significance) is set to 1, thus disabling any early stopping.
 These tuning parameters are now also listed in the main manuscript in Sections~3.2.
@@
 
-We apologize for the confusion the missing equation numbers 
-might have caused. They are now added.
+We apologize for the confusion that the missing equation numbers 
+have caused. They have now been added.
 
@@
 We have measured fitting time and prediction time for our 
-application. For single predictions the distributional forest 
-can keep up with the other models. However, as you already 
-stated correctly, each prediction made by the forest is based 
-on an observation-specific weight matrix and therefore has to 
-be calculated separately. For that reason prediction time for a 
-set of observations is clearly higher for the forest model than 
-the GAMLSS models or the EMOS. The exact results and times can 
-be found in Supplement 2 (Stationwise Evaluation).
-
-\begin{review}
-abstract: the acronym 'GAMLSS' should be spelled out the first 
+application. For observation-by-observation predictions -- as
+done in real-time -- the computational cost for the distributional
+forest is comparable to the GAMLSS approaches. The former have
+to re-fit the model using the observation-specific weights
+while the latter has to set up the model matrices for all the
+additive terms. However, when predicting a set of observations
+the costs for the GAMLSS are lower (due to leveraging matrix
+operations) and for the distributional forest higher (as it
+still operates observation-wise) but not prohibitively so.
+The exact results and times can be found in Supplement 2
+(Stationwise Evaluation).
+
+\begin{review}
+abstract: the acronym `GAMLSS' should be spelled out the first 
 time it is used
@@
 
-We have rephrased this part of the abstract now spelling out 'GAMLSS' at 
-its first appearance.
+We have rephrased this part of the abstract now spelling out
+`GAMLSS' more explicitly at its first appearance.
 
@@
 p3: I'm not sure if people outside the weather forecasting 
-community is familiar with the notion of an 'ensemble', so 
+community is familiar with the notion of an `ensemble', so 
 this should be explained briefly
@@
 \begin{review}
-p7: What is meant by 'domain knowledge'?
-\end{review}
-
-We have rephrased this as 'expert knowledge in the application domain'
+p7: What is meant by `domain knowledge'?
+\end{review}
+
+We have rephrased this to `expert knowledge in the application domain'
 in Section~2.2.
@@
 
-The residual QQ plots (Dunn et al. 1996) illustrate the probability 
-integral transform (PIT, Gneiting et al. 2007) values of the observations 
-in the testing data and are therefore scale invariant. A more detailed 
+The residual QQ plots (Dunn et al.\ 1996) are built on top of
+the probability integral transform (PIT, Gneiting et al.\ 2007).
+Therefore they are invariant to transformations that preserve
+the PIT, as is the case here. A more detailed 
 explanation can be found in Supplement 2 (Stationwise Evaluation).
@@
 This seems to result from the one specific training and validation period 
-(training: 1985--2008, validation: 2009-2012).
+(training: 1985--2008, validation: 2009-2012) and from East Tyrol having
+somewhat different weather than the rest of Tyrol.
+
 The advantages of the GAMLSS model even out at most of the stations when looking 
@@
 You are right, our explanation was incorrect, thank you for pointing
-this out! We have now clarified that the model employs a censored
-distribution but the observations are not censored, just non-negative
-with a point mass at zero.
+this out. We have now clarified that the \emph{model} employs a censored
+distribution while the \emph{observations} are not censored, just non-negative
+with many ties at zero.
 
@@
 The observations are depicted by crosses and the predicted point mass
-from the model by filled circles.'' to clearly distinguish between the model
-and the observations.
+from the model by filled circles.'' to clearly distinguish between the
+\emph{model} and the \emph{observations}.
 
@@
 parameter vector $\hat{\Theta}(z)$ can be calculated following 
-the given formula. But you are right, one can also just evaluate 
-the predicted terminal node for $z$ if the predicted parameter
-vector for this node is already known as there is one unique 
-parameter vector for all observations in the same terminal node.
-This is now also explained at the end of Section~2.2.
+the given formula. But you are right, this is equivalent to simply
+obtaining the paramter vector that has been estimated in the 
+corresponding node when learning the tree. This is now also
+explained at the end of Section~2.2.
 
@@
 
-Yes, in the forest model the parameter vector has to be estimated 
-separately for each new prediction as it is based on an 
-observation-specific weight matrix. For that reason, making 
-predictions for a set of new observations is clearly more 
-time consuming for the forest model than the GAMLSS models 
-or the EMOS model. However, the time needed to make one single 
-prediction is similar for all methods. The exact results 
-concerning fitting and prediction times can now be found in
+Yes, in distributional forests the parameter vector has to be estimated 
+separately for each new prediction as it is based on
+observation-specific weights. For that reason, making 
+predictions for \emph{a set of} new observations is more 
+time consuming for the forest (where predictions are obtained
+separately) than the GAMLSS models (where matrix operations
+can be leveraged). However, when making predictions observation
+by observation -- as done in real-time -- the time required is
+comparable for the forest and GAMLSS. The exact results 
+regarding fitting and prediction times can now be found in
 Supplement 2 (Stationwise Evaluation).
@@
 
-This is itself a complex topic with an own scientific community
-focusing on the development and improvement of numerical weather
-forecasts. Bauer, Thorphe, and Brunet (2015) give a good overview
-of the history and open scientific challenges of NWP ensemble forecast
-models including a discussion about unresolved processes and possible
-solutions beyond the current state of NWP models.
-We have added this reference next to the statement you are referring to.
-
-The following figure shows a comparison of the true local topography
+A reference to Bauer, Thorphe, and Brunet (2015) has been added. They
+give a good overview of the history and open scientific challenges
+of NWP ensemble forecast models including a discussion about
+unresolved processes and possible solutions beyond the current state
+of NWP models.
+
+To illustrate the issue in our forecast domain,
+the following figure shows a comparison of the true local topography
 (Robinson et~al. 2014; left)
 and the model topography of the GEFS reforecast model
-($0.5^\circ \times 0.5^\circ$, approx. $50 \times 50~km^2$; right) to illustrate
-the extent of simplification needed in such global weather forecast models.
+($0.5^\circ \times 0.5^\circ$, approx. $50 \times 50~km^2$; right). 
+This conveys the extent of topographical simplification needed in
+such global weather forecast models.
 
@@
 observations in a terminal node (\texttt{minbucket}) is set to 20 
-and the significance level for variable selection (\texttt{alpha}) 
-is kept at the default value 1. We have added this information in 
-the manuscript in Section~3.2, 'Distributional forest', and
+and early stopping (or ``pre-pruning'') based on the significance level is disabled
+(by setting \texttt{alpha} to 1). We have added this information in 
+the manuscript in Section~3.2, `Distributional forest', and
 also discussed the choice of these tuning parameter in Section~4 and
@@
 \end{review}
+
+\fixme{This reply needs to be rewritten. It should be listed first
+which variables correspond to each other between the models. Then
+you can explain that the greater variation in the forest is not
+surprising. Finally, it should be pointed out where in the manuscript
+this is also discussed.}
 
@@
 
-Apologies for the unclear notation in Appendix A.
+Apologies for the unclear notation in Appendix~A.
 Depending on whether the split variable $Z_l$ is a numeric 
-or a categorical variable, $t_l$ is either a
-$k$~-~or~$k \cdot H$~-~dimensional vector.
+or a categorical variable, $t_l$ is a vector either of
+dimension $k$ or $k \cdot H$.
 We have revised the corresponding parts in the manuscript (also 
-regarding your comments 17--20) and we hope that it now clarifies 
-the content.
+regarding your comments 17--20) and hope that these aspects
+are now conveyed clearly.
 
@@
 
-As in the foregoing point this was just a general notation, but 
+As in the previous point this was just a general notation, but 
 actually the variables corresponding to the $l$-th split variable 
@@
 
-Yes, one stopping criterion is that no p-value is beneath the 
-selected significance level. Therefore, no further 
-split is performed in the considered node in this case.
+Yes, significance can be used as one of the stopping criteria,
+so that partitioning is stopped early upon non-signifcance
+of the tests. This is sometimes also called ``pre-pruning'' in
+the tree literature.
+
+However, in our applications pre-pruning is disabled which
+is explained in \dots \fixme{Improve Appendix~A to better
+reflect this. Also align with the other comments about
+alpha.}
 
@@
 Thank you very much for your feedback and helpful suggestions. We hope 
-the improvements we made on the paper can clear your doubts and concerns.
+the improvements we made address your doubts and concerns.
 
@@
 
-We have applied a Gaussian distribution left-censored at zero
-as it has proven to be an adequate choice for this type of 
-application in earlier publications.
-Moreover, the residual QQ plots and PIT histograms illustrate
-that the models are well calibrated employing this distribution.
-But thank you very much for the advice to consider a hurdle model!
-This is definitely a good idea and therefore we have carried out the
-same evaluations as presented in the paper but employing a
-Gaussian hurdle model. Additionally, a logistic distribution 
-left-censored at zero has been applied as well in the same setting in
-order to account for heavier tails. The results of these evaluation
-can be found in Supplement 1 (Different Response Distributions). 
-Overall, the distributional forest is proven to be very robust against 
-distributional misspecifications. Therefore, we decided to keep the 
-distributional assumption considered in the main manuscript but we have 
-also discussed the tested alternatives in Section~4.
+Thank you for emphasizing this issue. In the initial submission
+we did include assessments of calibration by graphical means
+(QQ plots and PIT histograms). But we completely agree that a
+more thorough assessment of the distributional specification is
+preferable in this context.
+
+Therefore, we followed your advice and considered both a different
+distribution for the continuous part and a two-part hurdle model.
+For the former, the logistic distribution (also left-censored at zero)
+is used as it has heavier tails and was previously used in the
+precipitation forecasting literature \fixme{add reference}.
+For the latter, a two-part model is used combining a binary
+zero hurdle ($y = 0$ vs.\ $y > 0$) and an adjusted Gaussian distribution
+(left-truncated at zero).
+
+We have carried out the
+same evaluations as presented in the main manuscript and present
+the full results in Supplement 1 (Different Response Distributions). 
+The main insights from this are also discussed in Section~4:
+At most stations the censored Gaussian manuscript from the main
+manuscript performs at least as well as the two alternative
+distributions, if not better. Moreover, the distributional forest
+proves to be rather robust against distributional misspecifications.
+Therefore, the discussion in the main manuscript is still based
+on the censored Gaussian distribution.
 
@@
 
-As you mention one way to deal with skewness is to use a distribution
-with an additional skewness parameter (e.g., Scheuerer and Hamill 2015,
-Baran and Nemoda 2016) which increases the complexity of the statistical
-models.
-Another frequently used approach is to apply a
-power-transformation (Stidd 1973, Hutchinson 1998, Stauffer et al. 2017) to
-remove the majority of the skewness from the data. The optimal transformation
-depends on the application and data set, in our case we are following the work
-of Stauffer et al. (2017) who presented a similar application on the very same
-data set. As our results show (Figure~3 in the manuscript) this approach works 
-well for the application presented in our manuscript.
-Moreover, the focus of this paper is on the features of the presented
-distributional forest regarding variable selection and the capability
-to also model non-linear effects rather than optimizing the model
-specification for this particular application.
-
-%The skewness results from the characteristics of precipitation sums. 
-%Most of the time no or only moderate precipitation amounts are forecasted 
-%and observed. In some situations (e.g., days with thunderstorms, cold and 
-%warm fronts, \dots) large amounts of precipitation can be observed at 
-%specific sites. This is often also depict by the ensemble prediction 
-%system where a subset of all members predicts rather large amounts 
-%(compared to the overall ensemble mean or median). This yields strongly 
-%skewed ensemble forecasts which has to be addressed in some sort.\\
-%\fixme{\textit{FIXME: Should this explanation why there is skewness be included?}}
-
-%As you mention one way how to deal with skewness would be to 
-%apply a distributional family with an additional parameter to capture
-%skewness. However, this can be very complex and thus lead to less well 
-%fitting estimators of the location and scale parameter.
-%Moreover, the focus of this paper is on the features of the presented 
-%distributional forest regarding variable selection and the capability 
-%to also model non-linear effects rather than optimizing the model 
-%specification for this particular application.
-%Transforming the data to remove large parts of the skewness is
-%a frequently used approach. Therefore, employing such a transformation 
-%allows for a better way of comparing our results to those presented in 
-%literature. In this study we are thus following Stauffer et al. (2017) 
-%using the power transformation which turned out to be optimal 
-%for a similar application for the very same study area.
+Our manuscript follows the frequently-used approach of applying a
+power-transformation (Stidd 1973, Hutchinson 1998, Stauffer et al.\ 2017)
+to be more easily comparable to previous publications in the field.
+These focus on predicting location and scale based on mean and
+variance from the NWP ensemble. Hence, we also follow that approach
+but adopt a different strategy for estimating selecting variables
+and estimating nonlinear effects and interactions.
+
+As you correctly point out, another way to deal with skewness is to
+use an extended distributional specification (e.g., Scheuerer and Hamill 2015,
+Baran and Nemoda 2016). This increases the complexity of the statistical
+models and makes the comparison to publications based on power transformations
+more difficult. Therefore, we feel that this exceeds the scope of this
+contribution but we do discuss it in Section~4 as being of interest for
+future research. \fixme{Include one or two sentences about this as an
+outlook in the discussion of distributional specifications.}
 
@@
 
-We completely agree with you as that other models such as
-GAMLSS are clearly advantageous regarding interpretability.
-These models provide a very powerful framework that works
-well in a wide range of applications and research fields,
-including meteorology. Therefore, our intention is not to present
-distributional forests as a replacement of these commonly used methods
-but rather as a well working alternative. Especially if one is mainly
-interested in predictive skill and a good performing fully automated
-weather forecasting tool the interpretability is not implicitly needed
-in some cases (machine learning performance vs. interpretability, e.g.,
-Rasp and Lerch 2018).
-We adapted our manuscript as we want to emphasize that in our
-application the focus is mainly on the predictive performance and less on
-the interpretation of the different effects.
-We hope we could point this out clearly. \\
-\fixme{\textit{FIXME: Where should this be pointed out in the manuscript? 
-In the discussion or after introducing trees and forests? (trade-off between
-flexibility of the model and interpretability)}}
+We completely agree with you that GAMLSS are tremendously useful,
+especially in situations where an interpretable but still
+well-performing model is needed. The intention of our work is not
+to present distributional forests as a replacement of GAMLSS but
+rather as a complementing technique for the distributional regression
+toolbox. Much like random forests and GAMs are complements in
+the conditional mean regression toolbox.
+
+In this sense, distributional forests are appealing for weather
+forecasting tasks because the focus is not so much on interpretation
+but often much more on: (a)~forecasting skill and (b)~(semi-)automatic
+application of models on a larger domain. See also the discussion
+in, e.g., Rasp and Lerch 2018).
+
+We have adapted our manuscript to reflect that we focus on forecasting
+skill in a large number of models -- and not so much on interpretation.
+
+\fixme{Where should this be pointed out in the manuscript? 
+Z: Both in the introduction when the ``zoo'' of regression techniques
+is introduced -- and again in the discussion (possibly along with
+variable selection?).}
 
@@
 
-We apologize for the confusion the missing equation numbers 
-might have caused. They are now added.
+We apologize for the confusion that the missing equation numbers 
+have caused. They have now been added.
 We have also rephrased Section~2.3. which now first introduces random 
@@
 interaction effects and the main effects. For instance, 
-Wood et al. (2013) discuss a generic approach for defining 
+Wood et al.\ (2013) discuss a generic approach for defining 
 well-identified and interpretable effect decompositions in 
@@
 representation of penalised model components, while Goicoa 
-et al. (2017); Ugarte et al. (2017) determine the identifiability
+et al.\ (2017); Ugarte et al.\ (2017) determine the identifiability
 constraints from the null space of the penalties associated 
@@
 
-Thank you for your suggestions and the references. These are all
-good ideas, however, in our application the number of covariates
-is too large in order to keep a model with all interactions manageable.
-We have also discussed this in Section~4 of the main manuscript and
+Thank you for your suggestions and the references. Indeed we did
+not discuss sufficiently the possibilities for including interactions
+in GAMLSS. This has been remedied, see Section~4.
+
+However, in our application the number of covariates is too large
+(80) to even consider all pairwise interactions (3160 combinations)
+using the approaches mentioned above, let alone interactions of higher orders.
+
+We now discuss this in Section~4 of the main manuscript and
 rephrased the corresponding sentence in the introduction to ``while 
 requiring no meteorological knowledge about the atmospheric processes 
-which drive formation of precipitation for the model specification.'' 
-in order to clarify that this refers only to the variable prespecification 
-in this application.
+which drive formation of precipitation for the model specification'' 
+in order to clarify that this pertains to the variable selection.
 
@@
 mind that GAMLSS and boosting yield interpretable results which 
-the random forests do not do. also, what happens when choosing 
+the random forests do not do. Also, what happens when choosing 
 another station than Axams? The figure 8 indicates that the 
@@
 
-In order to investigate on these questions we have carried out the full
+In order to investigate these questions we have carried out the full
 analysis as presented for station Axams in the main manuscript for 14
 further observation stations. It shows that the distributional
-forest is very robust against distributional misspecifications
+forest is very robust against distributional (mis)specifications
 and almost always competitive with the best among the other methods 
@@
 Especially for meteorologists intending to apply model output 
-statistics (MOS) for several hundreds of observation stations 
+statistics (MOS) for dozens or hundreds of observation stations 
 (e.g., Wahl 2014, Rasp and Lerch 2018)
 the distributional forest provides a good and robust alternative. 
-Even though computational costs are higher than for an EMOS model, 
-they are similar as for a prespecified model and lower than for 
-a model applying variable selection based on cross-validation.
-Execution times have been measured and can be found
-in Supplement~2 (Stationwise Evaluation).
+Computational costs are clearly higher than for a basic EMOS model, 
+but similar to a prespecified GAMLSS and lower than for GAMLSS
+using variable selection by cross-validation. Fitting and forecasting
+times are reported now in Supplement~2 (Stationwise Evaluation).
+
+Regarding forecasting tasks other than the one presented in
+the manuscript, we would \emph{not} expect that distributional
+forests are \emph{``generally better''}. As for (boosted) GAMs and
+random forests we would expect that sometimes (boosted) GAMLSS and
+sometimes distributional forests perform better. As already argued
+above we see the methods as complements not substitutes.
+
+\fixme{Refer again to the corresponding point in the manuscript.}
 
@@
 7. How are the folds chosen for performing the predictions? 
-is it 4 subsequent years or randomly chosen 4 years out of 
+Is it 4 subsequent years or randomly chosen 4 years out of 
 the 28? I expect this to have a strong impact on the predictions 
@@
 are not necessarily successive. This is now also explained in the 
-description of the cross-validation setup third paragraph of 
-Section~3.3.
-
-We have chosen two different setups: First we learned the model
-on the first 24 years and evaluated on the following 4 years
-in order to provide a classic time ordered setup. 
-However, no influence of time is to be expected which is also supported by 
-an autocorrelation coefficient of -0.007 of the out-of-bag quantile 
-residuals of the distributional forest model for station Axams for
-a time lag of 1 day.
+description of the cross-validation setup in the third paragraph of 
+Section~3.3. As already argued previously, the same cross-validation
+approach is not only applied to Axams but also 14 further stations
+in Supplement~2 to check for robustness across stations.
+
+In addition to the cross-validation setup there is also the setup
+where models are learned on the first 24 years and evaluated on the
+final 4 years. Thus, the classic time-ordered setup is employed there,
+again for Axams and 14 further stations.
+
+Moreover, no influence of time is to be expected as autocorrelations of
+daily lags are typically very low. For example, the autocorrelation
+coefficient of the out-of-bag quantile residuals from the distributional
+forest for station Axams is only $-0.007$.
 % ## calculation of autocorrelation:
@@
 % ac <- acf(qqrplot(pit_df)$residuals)
-To evaluate the models in a broader framework a cross-validation 
-with the above described choice of folds was applied. The results
-for station Axams are included in the main manuscript, results
-for 14 additional stations can be found in
-Supplement~2 (Stationwise Evaluation).
 
@@
 
-As mentioned in our response to your comment 5, and also discussed in Section~4, 
-the high number of covariates (80) restricts the possibilities of variable 
-selection and model specification. For an AIC or BIC based selection an 
-evaluation of 80 main effects for two parameters would make this a very complex
-procedure. Moreover, Hofner et al. (2016, p.~3) stated that the 
-AIC-based variable selection methods implemented in the \textsf{R} 
+As mentioned in our response to your comment 5 above (and now also discussed in Section~4),
+the high number of covariates (80) prevents simply including all variables
+(or even all interactions) without further penalization. For an AIC- or BIC-based selection an 
+evaluation of 80 main effects for both distributional parameters would make this a very complex
+procedure. Moreover, Hofner et al.\ (2016, p.~3) stated that the 
+selection methods implemented in the \textsf{R} 
 package \textbf{gamlss} ``[...] can be unstable, especially when it comes 
 to selecting possibly different sets of variables for 
-multiple distribution parameters.'' For that reason we decided
-not to apply such a selection method.
+multiple distribution parameters.'' For that reason we considered
+the boosting-based approach as an alternative approach to variable selection
+in GAMLSS.
 
@@
 
-By a `basic EMOS' we are referring to the linear distributional 
-regression model based on the idea presented by Gneiting et al. (2005) 
-applying a left-censored at zero Gaussian distribution where only the 
-ensemble mean is included for the location parameter and the ensemble 
-standard deviation for the scale parameter.
+By a `basic EMOS' we are referring to the idea of Gneiting et al.\ (2005)
+that employs a distributional regression with linear predictors for
+location and scale, respectively, based on the corresponding empirical
+mean and variance from the NWP ensemble.
 We now introduce the model which we are referring to as a `basic EMOS'
@@
 Third, why not to include interactions in the boosting algorithm 
-as well? this should be conceptually straight forward? 
-\end{review}
-
-Including only pairwise interactions
+as well? This should be conceptually straight forward? 
+\end{review}
+
+As argued above, including only pairwise interactions
 of the 80 available variables would already lead to 3160 additional
-covariates which would make the model too complex and hardly manageable
-regarding computational costs. Therefore, we decided not to include
-any interactions in the boosting algorithm.
-This issue is now addressed and discussed in Section~4.
-% factorial(80)/factorial(80-2)/factorial(2)
-
-\begin{review}
-Fourth, why are 100 trees chosen in the forest? Is this a a default 
+covariates which would make the model even more complex and substantially
+increase (the already high) computational costs. Therefore, we decided not to include
+any interactions in the boosting algorithm which is now addressed and
+discussed in Section~4.
+% choose(80, 2)
+
+\begin{review}
+Fourth, why are 100 trees chosen in the forest? Is this a default 
 value? May this number not also be dependent on the complexity of 
 the model? Why not to choose the number of trees similar to what is 
-done in boosting? this would also increase the computation time of 
+done in boosting? This would also increase the computation time of 
 the forest a lot.
@@
 
-Following Breiman (2001) random forests are very robust and do not overfit, 
+As pointed out by Breiman (2001), random forests are very robust and do not overfit, 
 not even for a high number of trees. However, computational costs increase 
-linearly with the number of trees (Biau et al. 2016).
-Therefore, we followed these recommendations of choosing a high number of
+linearly with the number of trees (Biau et al.\ 2016).
+Therefore, we followed their recommendations of choosing a high number of
 trees, but still small enough to keep the computation time at
-a reasonable level. Moreover, 100 is a commonly used number of trees that
-seems to work very well. A larger number of trees could be used but 
-the performance is usually not significantly better compared to a forest with 
-100 trees. Tuning parameters of forest models, including the number of 
+a reasonable level. For this, 100 is a commonly-used number of trees.
+Tuning parameters of distributional forests, including the number of 
 trees (\texttt{ntree}), are now discussed in Section~4 
@@
 
-The EMOS as originally suggested by Gneiting et al. (2005) can be seen
+The EMOS as originally suggested by Gneiting et al.\ (2005) can be seen
 as one of the standard ensemble model output statistics (or post-processing)
@@
 for 15 stations. In the manuscript we decided to only show the 
-residual QQ plot as it allows for a better analysis of the tail 
-behavior than the PIT histogram.
+residual QQ plot as it allows for a better analysis of the 
+behavior in the upper tail compared to the PIT histogram.
 
@@
 
+\fixme{This also needs to be rewritten (probably best by Reto).
+It is a bit frustrating to start the response with comments how instable
+all of this is. Instead focus on the insights that are quite
+stable across stations.}
+
 Figure 6 is only a visualization of the 10 most important covariates 
-of the forest model in one particular setting (station Axams, learned 
+of the distributional forest in one particular setting (station Axams, learned 
 on 1985--2008). However, this top 10 list is not considered for any specification 

### 2018-10-21 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
recalculation of Rain_distributions.R after bugfix and adaption of the corresponding supplement
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2345 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e6432f985fe0140900ab2a8615af671e9020aa26

@@
 
-\item Similarly, for the EMOS model all distributions lead to comparable CRPS for all
-stations. However, there is some more variation in CRPS across distributions compared to
-the distributional forest.
+\item Also for the prespecified GAMLSS all three distributions lead to almost the same 
+results and no clear advantage of one or the other distribution can be detected.
 
-\item In contrast, the two GAMLSS models typically perform better (i.e., have lower CRPS) for
-the Gaussian distributions compared to the logistic distribution. The preference of
-censored vs.\ hurdle specification is not unanimous across stations.
+\item Similarly, for the EMOS model and the boosted GAMLSS all distributions 
+lead to comparable CRPS for all stations. However, there is some more variation 
+in CRPS across distributions compared to the distributional forest.
 
@@
 levels(results$means$distribution) <- c("cgaussian", "hgaussian", "clogistic")
-levels(results$means$station)[16] <- "St. Johann im Walde"
+levels(results$means$station)[15] <- "St. Johann im Walde"
 
@@
 
+\item Again, for the prespecified GAMLSS the results are very similar as for the 
+  distributional forest showing only a slight advantage of the Gaussian hurdle model, 
+  however, not for all stations. 
+  %The zero-censored logistic distribution performs on average equally
+  %well as the the zero-censored Gaussian distribution.
+
+\item For the boosted GAMLSS, the Gaussian hurdle is quite close to the zero-censored
+  Gaussian reference but shows a high variation, while the zero-censored logistic 
+  performs slightly better.
+
 \item For EMOS, the median skill score for both Gaussian hurdle and zero-censored logistic
-  are very close to the zero-censored Gaussian reference. However, the variation for the
-  hurdle model is much higher than for the censored model.
-
-\item For the two GAMLSS models the Gaussian hurdle is quite close to the zero-censored
-  Gaussian reference, while the zero-censored logistic clearly performs worse (especially
-  in the prespecified GAMLSS).
+  are very close to the zero-censored Gaussian reference. However, as for the boosted GAMLSS
+  the variation for the hurdle model is much higher than for the censored model.
 
@@
         xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 2.5),
-        ylim = c(-0.10, 0.05))
+        ylim = c(-0.05, 0.05))
 boxplot(s_df, add = TRUE, col = "transparent")
@@
         xlab = "", ylab = "", xlim = c(0.5, 2.5),
-        ylim = c(-0.10, 0.05))
+        ylim = c(-0.05, 0.05))
 boxplot(s_g, add = TRUE, col = "transparent", yaxt = 'n')
@@
         xlab = "", ylab = "", xlim = c(0.5, 2.5),
-        ylim = c(-0.10, 0.05))
+        ylim = c(-0.05, 0.05))
 boxplot(s_gb, add = TRUE, col = "transparent", yaxt = 'n')
@@
         xlab = "", ylab = "", xlim = c(0.5, 2.5),
-        ylim = c(-0.10, 0.05))
+        ylim = c(-0.05, 0.05))
 boxplot(s_emos, add = TRUE, col = "transparent", yaxt = 'n')
@@
 for hurdle Gaussian and censored logistic distributions in reference to the
-censored Gaussian distribution (orange horizontal line). Gray lines correspond to the 15~stations and are
-aggregated in the boxplot. The y-axis range is limited
-to $-10$\%, lower values for the two GAMLSS models are not displayed.}
+censored Gaussian distribution (orange horizontal line). Gray lines correspond to the 15~stations and are aggregated in the boxplot.}
 \end{figure}
@@
 
-\emph{The same data already shown in Figures~\ref{fig:lattice} and~\ref{fig:box_crpss_gaussian}
-can be displayed in further ways. However, these do not really add much and hence should probably
-be omitted.}
+\emph{The same data already shown in Figures~\ref{fig:lattice} 
+and~\ref{fig:box_crpss_gaussian} can be displayed in further ways. However, these do 
+not really add much and hence should probably be omitted.}
 
@@
 for each distribution, once based on CRPS skill score values with reference model EMOS 
-and once based on CRPS values. Both confirm the conclusion that the censored Gaussian 
-model and the hurdle model lead to almost the same results for all methods while only 
-the logistic model shows small variations.
+and once based on CRPS values. Both confirm the conclusion that all three distributions
+lead to almost the same results for all methods.
 
-Altogether it can be stated that for this application the left-censored at zero Gaussian distribution seems 
-to be an appropriate choice while the other two distributions, especially the hurdle model, 
-are also reasonable alternatives. However, they do not lead to an overall improvement in 
+Altogether it can be stated that for this application the left-censored at zero Gaussian
+distribution seems to be an appropriate choice while the other two distributions are also
+reasonable alternatives. However, they do not lead to an overall improvement in 
 performance for this application.
@@
 typically with more variability (especially for EMOS and the boosted
-GAMLSS). Therefore, the main manuscript focuses on the zero-censored
+GAMLSS). Furthermore, for the boosted GAMLSS the zero-censored logistic
+distribution shows slightly better results, however, only within a range of
+less than $2\%$ and not for all stations.
+Therefore, the main manuscript focuses on the zero-censored
 Gaussian model which is well-established in the probabilistic

### 2018-10-21 pkg/disttree/inst/draft/draft.Rnw
adaption to new results of Rain_distribution and further revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2346 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7ebcc9e62debed3abf93f2f926af1ebb44bc0415

@@
   (\code{minsplit}) \hlc{is set to $50$, the minimal number of observations in a 
-  terminal node} (\code{minbucket})\hlc{is set to $20$ and the significance level 
-  for variable selection} (\code{alpha}) \hlc{is kept at the default value $1$.}
+  terminal node} (\code{minbucket}) \hlc{is set to $20$ and the significance level 
+  for variable selection} (\code{alpha}) \hlc{is kept at the default value $1$ such 
+  that no early-stopping (sometimes also refered to as ``pre-pruning'') is applied
+  based on results of the statistical tests.}
   % stopping criteria:
@@
 are similar to those presented in the paper applying a Gaussian distribution
-left-censored at zero. However, comparing the two censored distributions,
-at many stations the models perform not as well under the logistic distribution 
-as under the Gaussian distribution. The hurdle model leads to slight improvements 
-in some settings, but this comes with more variability (especially for EMOS and the 
-boosted GAMLSS).
-Altogether, the distributional forest has proven to be very robust against
-distributional misspecifications.}
+left-censored at zero. In some settings the hurdle model leads to slight 
+improvements, but this comes with more variability (especially for EMOS and the 
+boosted GAMLSS). For the boosted GAMLSS the left-censored logistic distribution
+shows slightly better results, however, only within a range of less than $2\%$
+and not for all stations. Altogether, the distributional forest has proven to be 
+very robust against distributional misspecifications.
+
+Alternatively, a distribution including an addidtional parameter capturing skewness
+\citep{Scheuerer+Hamill:2015, Baran+Nemoda:2016} could be applied. For example, 
+in transformation forests} \citep{Hothorn+Zeileis:2017} \hlc{a very flexible 
+five-parametric distribution family is employed such that skewness can be modeled 
+explicitly.
+Evaluating further distributions might be of interest but this would go beyond the
+scope of this manuscript.
+}
 
@@
 in this node as the stopping criterion of no $p$-values being below the 
-significance level is fulfilled.}
+significance level is fulfilled. This type of early-stopping in building
+a tree is sometimes also referred to as ``pre-pruning''. For forest models
+pre-pruning is often switched off by setting the significance level to $1$.}
 

### 2018-10-21 pkg/disttree/inst/draft/rejoinder.Rnw
adaption to new results of Rain_distribution and further revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2346 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7ebcc9e62debed3abf93f2f926af1ebb44bc0415

@@
 \item \emph{Supplement 2: Stationwise evaluation.}\\
-The show that the more extensive cross-validation results -- obtained only
+To show that the more extensive cross-validation results -- obtained only
 for one station (Axams) in the main manuscript -- are fairly typical,
@@
 However, in our applications pre-pruning is disabled which
-is explained in \dots \fixme{Improve Appendix~A to better
-reflect this. Also align with the other comments about
-alpha.}
+is explained in Section~3.2 and in the appendix.
 
@@
 is used as it has heavier tails and was previously used in the
-precipitation forecasting literature \fixme{add reference}.
+precipitation forecasting literature (Stauffer et al.\ 2017b).
 For the latter, a two-part model is used combining a binary
@@
 
-We have carried out the
-same evaluations as presented in the main manuscript and present
-the full results in Supplement 1 (Different Response Distributions). 
+We have carried out the same evaluations as presented in the main manuscript and 
+present the full results in Supplement 1 (Different Response Distributions). 
 The main insights from this are also discussed in Section~4:
-At most stations the censored Gaussian manuscript from the main
+At most stations the censored Gaussian distribution from the main
 manuscript performs at least as well as the two alternative
-distributions, if not better. Moreover, the distributional forest
-proves to be rather robust against distributional misspecifications.
-Therefore, the discussion in the main manuscript is still based
+distributions. %, if not better. 
+Moreover, the distributional forest proves to be rather robust against distributional
+misspecifications. Therefore, the discussion in the main manuscript is still based
 on the censored Gaussian distribution.
@@
 Our manuscript follows the frequently-used approach of applying a
-power-transformation (Stidd 1973, Hutchinson 1998, Stauffer et al.\ 2017)
+power-transformation (Stidd 1973, Hutchinson 1998, Stauffer et al.\ 2017a)
 to be more easily comparable to previous publications in the field.
@@
 
-\item[] Stauffer, R., Mayr, G. J., Messner, J. W., Umlauf, N. and Zeileis, A. (2017). 
+\item[] Stauffer, R., Mayr, G. J., Messner, J. W., Umlauf, N. and Zeileis, A. (2017a). 
 Spatio-Temporal Precipitation Climatology over Complex Terrain Using a Censored Additive Regression Model,
 \textit{International Journal of Climatology} \textbf{37, 7}: 3264--3275.
+
+\item[] Stauffer, R., Umlauf, N., Messner, J. W., Mayr, G. J. and Zeileis, A. (2017b). 
+Ensemble Post-Processing of Daily Precipitation Sums over Complex Terrain Using Censored High-Resolution Standardized Anomalies,
+\textit{Monthly Weather Review} \textbf{45, 3}: 955--969.
+%doi = 10.1175/mwr-d-16-0260.1
 

### 2018-10-21 pkg/disttree/inst/draft/draft.Rnw
adaption to new results of Rain_distribution and further revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2347 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bd7c6bbd46a5fc06879c99fb98375489d745a676

@@
   for variable selection} (\code{alpha}) \hlc{is kept at the default value $1$ such 
-  that no early-stopping (sometimes also refered to as ``pre-pruning'') is applied
+  that no early-stopping (sometimes also referred to as ``pre-pruning'') is applied
   based on results of the statistical tests.}
@@
 
-Alternatively, a distribution including an addidtional parameter capturing skewness
+Alternatively, a distribution including an additional parameter capturing skewness
 \citep{Scheuerer+Hamill:2015, Baran+Nemoda:2016} could be applied. For example, 

### 2018-10-21 pkg/disttree/inst/draft/rejoinder.Rnw
adaption to new results of Rain_distribution and further revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2347 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bd7c6bbd46a5fc06879c99fb98375489d745a676

@@
 the given formula. But you are right, this is equivalent to simply
-obtaining the paramter vector that has been estimated in the 
+obtaining the parameter vector that has been estimated in the 
 corresponding node when learning the tree. This is now also
@@
 Yes, significance can be used as one of the stopping criteria,
-so that partitioning is stopped early upon non-signifcance
+so that partitioning is stopped early upon non-significance
 of the tests. This is sometimes also called ``pre-pruning'' in

### 2018-10-21 pkg/disttree/inst/draft/draft.Rnw
further revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2348 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f085d88cc0b20e11401fa7b30d38ad5601ed0a2f

@@
 \end{itemize}
-In the following, particular focus is given to distributional forests as a method for obtaining
-probabilistic forecasts by leveraging the strengths of random forests: the ability to
-capture both smooth and abruptly changing functions along with simultaneous selection of
-variables and possibly complex interactions. Thus, these properties make the method particularly
-appealing in case of many covariates with unknown effects and interactions where it would be 
-challenging to specify a distributional regression model like GAMLSS.
+
+\hlc{However, distributional forests should not be considered as a 
+replacement of GAMLSS but rather as a complementing method in the field of
+distributional models, similar to the way how GAM and random forests are complements 
+in the conditional mean regression toolbox.}
+
+In the following, particular focus is given to distributional forests as a method for
+obtaining probabilistic forecasts by leveraging the strengths of random forests: the 
+ability to capture both smooth and abruptly changing functions along with simultaneous selection of variables and possibly complex interactions.
+Thus, these properties make the method particularly appealing in case of many covariates with unknown effects and
+interactions where it would be challenging to specify a distributional regression model 
+like GAMLSS.
 
@@
 select covariates while meteorological expert knowledge was applied to 
-select covariates and specify interactions for the prespecified GAMLSS.}
+select covariates and specify interactions for the prespecified GAMLSS.
+
+As mentioned before, distributional forests should not be seen as a replacement
+for GAMLSS but rather as a complementary method. GAMLSS provide a powerful framework
+in the field of distributional modeling as they allow for the combination of 
+interpretability and a high goodnes-of-fit. For distributional forests, the focus
+is mainly on forecasting skill, automatic variable selection and capturing also 
+non-additive effects rather than on interpretability. Therefore, both approaches
+provide valuable features and advantages that can complement each other in many 
+applications as it can not be expected that one of them is in general better than 
+the other. 
+}
 
@@
 
-Alternatively, a distribution including an additional parameter capturing skewness
-\citep{Scheuerer+Hamill:2015, Baran+Nemoda:2016} could be applied. For example, 
+Alternatively, a distribution including an additional parameter capturing skewness}
+\citep{Scheuerer+Hamill:2015, Baran+Nemoda:2016} \hlc{could be applied. For example, 
 in transformation forests} \citep{Hothorn+Zeileis:2017} \hlc{a very flexible 

### 2018-10-22 pkg/model4you/inst/JORS/man_model4you.Rnw
nice header
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1843 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 57dc10457bec608c981aa10c7dd792a152d7425f
update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2174 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b7e23f0d093fd036a6f75a8abcfc0edb5815a2b2
update manuals
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2352 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0f27fc1e71b261c31a167c24228d74a278b7a34c

@@
 \documentclass{josr}
+
+%% revised text
+\newcommand{\new}[1]{{\color{blue} #1}}
 
@@
 
-To complete this template, please replace the blue text with your own. The paper has three main sections: (1) Overview; (2) Availability; (3) Reuse potential. \\
+%To complete this template, please replace the blue text with your own. The paper has three main sections: (1) Overview; (2) Availability; (3) Reuse potential. \\
 
-Please submit the completed paper to: editor.jors@ubiquitypress.com
+%Please submit the completed paper to: editor.jors@ubiquitypress.com
 
@@
 estimate personalised models.  The R package \emph{model4you} implements these
-stratified and personalised models with a focus on ease of use and
+stratified and personalised models \new{in the setting with two randomly
+assigned treatments} with a focus on ease of use and
 interpretability so that clinicians and other users can take the model they
@@
 (individual in first group) and $1$ (individual in second group).
+
+Currently the R package \emph{model4you} can handle only
 
@@
 The R package \emph{model4you} is focused on ease of use and interpretability.
-Users can take a simple model that they know and understand as basis and simply
-plug it into \Sexpr{'pmtree()'} or \Sexpr{'pmforest()'} depending on whether
-they want subgroup wise or personalised models. The basis for these
+Users can take a simple model that they know and understand as basis.
+\new{
+Models currently available for use are linear models (function \Sexpr{'lm()'}
+in R), generalised
+linear models (\Sexpr{'glm()'}) and survival models (\Sexpr{'survreg()'} and
+\Sexpr{'coxph()'} from the R package \emph{survival}). Models are restricted
+to those with a single binary covariate (e.g. two treatment options).
+The user can simply plug the model object into \Sexpr{'pmtree()'} or
+\Sexpr{'pmforest()'} depending on whether they want subgroup wise or
+personalised models.
+}
+
+The basis for these
 functionalities is provided by the \emph{partykit}

### 2018-10-22 pkg/model4you/R/pmforest.R
update manuals
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2352 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0f27fc1e71b261c31a167c24228d74a278b7a34c

@@ -2,7 +2,7 @@
 #'
 #' Input a parametric model and get a forest.
 #'
-#' @param model a model object.
+#' @param model a model object. The model can be a parametric model with a single binary covariate.
 #' @param data data. If \code{NULL} the data from the model object are used.
 #' @param zformula formula describing which variable should be used for partitioning.
 #' Default is to use all variables in data that are not in the model (i.e. \code{~ .}).

### 2018-10-22 pkg/model4you/R/pmtree.R

@@ -1,8 +1,8 @@
 #' Compute model-based tree from model.
 #'
-#' Input a parametric model and get a forest.
+#' Input a parametric model and get a model-based tree.
 #'
-#' @param model a model object.
+#' @param model a model object. The model can be a parametric model with a binary covariate. 
 #' @param data data. If NULL (default) the data from the model object are used.
 #' @param zformula formula describing which variable should be used for partitioning.
 #' Default is to use all variables in data that are not in the model (i.e. \code{~ .}).

### 2018-10-22 pkg/model4you/inst/JORS/man_model4you.Rnw
revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2353 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 50505e1e661eedd0d20c5ada30d78fc811700cc3

@@
 different. This is achieved by finding instabilities in the model parameters
-with respect to a variable (characteristicum) and recursively partitioning the
+with respect to a variable (characteristic) and recursively partitioning the
 data into groups. If, for example the algorithm finds that men and women have
@@
 and personalised models. For this we need to load the package and -- to ensure
-reproducibility -- set a random seed. Also for visualisations we need packages
-\emph{ggplot2}, \emph{ggbeeswarm} and \emph{gridExtra}.
+reproducibility -- set a random seed.
+\new{
+Also for visualisations we need packages
+\emph{ggplot2} \citep{ggplot2}, \emph{ggbeeswarm} \citep{ggbeeswarm} and
+\emph{gridExtra} \citep{gridExtra}. The data used is
+available in the \emph{psychotools} package \citep{psychotools}.
+}
 <<prep>>=
@@
 In 2014 first-year business and economics students at the University of
-Innsbruck were divided into two examination groups. Group 1 wrote the exam in
+Innsbruck were divided into two examination groups. Group 1 took the exam in
 the morning and group 2 started after the first group finished. The exams for
@@
 @
-The estimtates and confidence intervals of this model can be computed via
+The estimates and confidence intervals of this model can be computed via
 <<>>=
@@
 for differences in exam performance).
+\new{Note that confidence intervals shown in the tree are not adjusted and should
+hence be interpreted as a measure for variablity and not be used for inference.}
 
@@
 student characteristics and the treatment. Since the percentage of successfull
-online tests is measured on a grid, a bee plot possibly shows the relationships
+online tests is measured on a grid, a \new{beeswarm} plot possibly shows the relationships
 even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1}).
-<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
+<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Beeswarm plot.")>>=
 dpdat_math <- cbind(pmods_math, MathExam)
@@
 For the number of previous attempts to pass the exam and the gender box plots,
-bee plots or a combination thereof can be used (Figure~\ref{fig:math_dp2}).
+\new{beeswarm} plots or a combination thereof can be used (Figure~\ref{fig:math_dp2}).
 <<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=
@@
 very simple to create understandable stratified and personalised models and
-compelling visualisations that can be used to communicate theses models.
+compelling visualisations that can be used to communicate these models.
 

### 2018-10-22 pkg/disttree/inst/draft/draft.Rnw
added computational details for supplements and corrected a few typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2354 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 782ebafa479aad5324d6874ee64c2dff4006a85a

@@
 and not for all stations. Altogether, the distributional forest has proven to be 
-very robust against distributional misspecifications.
-
-Alternatively, a distribution including an additional parameter capturing skewness}
-\citep{Scheuerer+Hamill:2015, Baran+Nemoda:2016} \hlc{could be applied. For example, 
-in transformation forests} \citep{Hothorn+Zeileis:2017} \hlc{a very flexible 
-five-parametric distribution family is employed such that skewness can be modeled 
-explicitly.
+very robust against distributional misspecifications.}
+
+\hlc{Alternatively, a distribution including an additional parameter capturing 
+skewness could be applied} \citep{Scheuerer+Hamill:2015, Baran+Nemoda:2016}. 
+\hlc{For example, in transformation forests} \citep{Hothorn+Zeileis:2017} 
+\hlc{a very flexible five-parametric distribution family is employed such 
+that skewness can be modeled explicitly.
 Evaluating further distributions might be of interest but this would go beyond the
@@
 also available within the \textsf{R}-Forge~project.
+The results presented in Supplement 1 (Different Response Distributions) and 
+Supplement 2 (Stationwise Evaluation) can be reproduced using
+\code{demo("RainDistributions", package = "disttree")} and 
+\code{demo("RainStationwise", package = "disttree")} respectively.
 

### 2018-10-22 pkg/disttree/inst/draft/rejoinder.Rnw
further revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2348 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f085d88cc0b20e11401fa7b30d38ad5601ed0a2f
added computational details for supplements and corrected a few typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2354 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 782ebafa479aad5324d6874ee64c2dff4006a85a

@@
 the manuscript in Section~3.2, `Distributional forest', and
-also discussed the choice of these tuning parameter in Section~4 and
+also discussed the choice of these tuning parameters in Section~4 and
 in the answer to the second point of the associate editor.
@@
 contribution but we do discuss it in Section~4 as being of interest for
-future research. \fixme{Include one or two sentences about this as an
-outlook in the discussion of distributional specifications.}
+future research. 
+%\fixme{Include one or two sentences about this as an
+%outlook in the discussion of distributional specifications.}
 
@@
 application of models on a larger domain. See also the discussion
-in, e.g., Rasp and Lerch 2018).
-
-We have adapted our manuscript to reflect that we focus on forecasting
-skill in a large number of models -- and not so much on interpretation.
-
-\fixme{Where should this be pointed out in the manuscript? 
-Z: Both in the introduction when the ``zoo'' of regression techniques
-is introduced -- and again in the discussion (possibly along with
-variable selection?).}
+in, e.g., Rasp and Lerch (2018).
+
+We have adapted our manuscript in Section~1 and Section~4, \textit{Variable Selection}, 
+to reflect that we focus on forecasting skill in a large number of 
+models -- and not so much on interpretation.
+
+%\fixme{Where should this be pointed out in the manuscript? 
+%Z: Both in the introduction when the ``zoo'' of regression techniques
+%is introduced -- and again in the discussion (possibly along with
+%variable selection?).}
 
@@
 further observation stations. It shows that the distributional
-forest is very robust against distributional (mis)specifications
-and almost always competitive with the best among the other methods 
+forest is %very robust against distributional (mis)specifications and (FIXME: this is shown in Supplement 1)
+almost always competitive with the best among the other methods 
 or even better. The exact results are provided in 
@@
 above we see the methods as complements not substitutes.
-
-\fixme{Refer again to the corresponding point in the manuscript.}
+This is now also mentioned in Section~1 and discussed in Section~4,
+\textit{Variable Selection}.
+
+%\fixme{Refer again to the corresponding point in the manuscript.}
 

### 2018-10-23 pkg/model4you/inst/JORS/man_model4you.Rnw
revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2355 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b13909c37992ab18cd7253fbec58437392562903

@@
 1. PhD student; Biostatistics Department, Epidemiology, Biostatistics \&
-Prevention Institute, University of Zurich
+Prevention Institute, University of Zurich; \new{Postdoc; Institute for Medical
+Information Processing, Biometry, and Epidemiology, LMU Munich}
 
@@
 individuals in the \Sexpr{'data'} and allow for predicting the response of
-other individuals given they do or don't receive the treatment of interest.
+other individuals given they do or do not receive the treatment of interest.
 
@@
 Again here the potential effect-modifying variables are taken by default
-as all variable not given in the model formula and can be defined using
+as all variables not given in the model formula and can be defined using
 the \Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.
@@
 @
-
-To investigate whether the exam was fair, we asses whether the two groups
-differ in terms of the percentage of correctly answered exam questions. This
-can be done using a simple linear model regressing the percentage points of
-correct anwers on the exam group.
+\new{To investigate the correlation between exam group and exam performance,
+we compute a simple linear model regressing the percentage points of
+correct answers on the exam group.}
 <<math_bmod>>=
@@
 @
-
+\new{
+We restrict the depth of the tree to two (\Sexpr{'maxdepth = 2'}) for
+illustration purposes. If this setting is not used a more complex tree would
+be estimated.
+}
 The tree (see Figure~\ref{fig:math_tree}) divides students based on the
@@
 
-Dependence plots with the group effect (treatment effect) on the y-axsis and
+Dependence plots with the group effect (treatment effect) on the y-axis and
 the student characteristics on the x-axis are a good way of visualising the
 personalised models and for getting knowledge about the interactions between
-student characteristics and the treatment. Since the percentage of successfull
-online tests is measured on a grid, a \new{beeswarm} plot possibly shows the relationships
-even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1}).
-<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Beeswarm plot.")>>=
+student characteristics and the treatment.
+\new{
+Note that the plot is related to but not the same as the classical partial
+dependence plot \citep{friedman2001greedy}, as each point in the figure is one
+patient in the given data set.
+}
+Since the percentage of successful online tests is measured on a grid, a
+\new{beeswarm} plot possibly shows the relationships
+even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1},
+\new{
+Figure~\ref{fig:math_dp11} includes a loess curve). We see a nonlinear relation
+between the percentage of successful online \Sexpr{'tests'} and the exam group
+effect. Students with great performance during the semenster are estimated to
+have the strongest effect.
+}
+<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot with loess curve.", "Beeswarm plot.")>>=
 dpdat_math <- cbind(pmods_math, MathExam)
@@
   theme(legend.position = "none") +
-  ylab("estmated individual\nexam group effect")
+  ylab("estimated individual\nexam group effect")
 
@@
   theme(legend.position = "none") +
-  ylab("estmated individual\nexam group effect")
+  ylab("estimated individual\nexam group effect")
 @
 
-For the number of previous attempts to pass the exam and the gender box plots,
-\new{beeswarm} plots or a combination thereof can be used (Figure~\ref{fig:math_dp2}).
-<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=
-
-ggplot(dpdat_math, aes(x = attempt, y = group2, color = attempt)) +
-  geom_quasirandom(alpha = 0.5) +
-  theme(legend.position = "none") +
-  ylab("estmated individual\nexam group effect")
-
+\new{
+For categorical variables such as the number of previous attempts to pass the
+exam (\Sexpr{'attempt'}) and \Sexpr{'gender'} we can use box plots,
+\new{beeswarm} plots or a combination thereof can be used (as in
+Figure~\ref{fig:math_dp21}). Figure~\ref{fig:math_dp22} shows the number of
+previous attempts on the x-axis and the color indicates the gender of each student.
+This way we can see that for students writing the exam for the first
+time, gender seems to play a bigger role in terms of estimated individual exam
+group effect than for others.
+}
+<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Gender.', 'Number of previous attempts and gender.')>>=
 ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
@@
   theme(legend.position = "none") +
-  ylab("estmated individual\nexam group effect")
+  ylab("estimated individual\nexam group effect")
+
+ggplot(dpdat_math, aes(x = attempt, y = group2, color = gender)) +
+  geom_quasirandom(alpha = 0.5) +
+  ylab("estimated individual\nexam group effect")
 @

### 2018-10-23 pkg/disttree/inst/draft/draft.Rnw
small corrections and change of order of subsections in the discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2358 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a3e7487d01efd56a103371034be8cf0acb55e46d

@@
 in the field of distributional modeling as they allow for the combination of 
-interpretability and a high goodnes-of-fit. For distributional forests, the focus
+interpretability and a high goodness-of-fit. For distributional forests, the focus
 is mainly on forecasting skill, automatic variable selection and capturing also 
@@
 }
-
-\subsubsection*{Tuning parameters}
-\hlc{Setting tuning parameters of a forest model can have a major influence
-not only on the accuracy of the model but also on the computational 
-complexity. 
-For the application in} Section~\ref{sec:precipitation} \hlc{we have 
-selected the values of these parameters} (listed in Section~\ref{sec:evaluation}) 
-based on advice from literature and our own experience.
-
-\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hlc{recommend to 
-build full-grown trees. Therefore, low values should be chosen for the stopping 
-criteria} \code{minsplit}, \code{minbucket} and \code{alpha}.
-\hlc{However, in a distributional model it also has to be considered that a 
-minimal number of observations is required to allow for a reasonable 
-estimation of parameters in each node. For that reason the minimal number
-of observations for which statistical tests are performed in the 
-distributional forest in} Section~\ref{sec:precipitation} is set to
-50 (\code{minbucket}) \hlc{and the minimal number of observations on which
-parameters are estimated is set to 20} (\code{minsplit}).
-
-As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
-\hlc{random forests do not overfit, not even for an increasing number of 
-trees} (\code{ntree}). \hlc{Hence, it is recommended to build forests consisting 
-of a large number of trees as this generally leads to an increase 
-in accuracy. However, it also comes along with increasing computational costs. 
-Therefore, ``a good choice results from a trade-off between computational 
-complexity and accuracy''} (\cite{Biau+Scornet:2016}, p.~205).
-\hlc{Following this advice we chose to build forests consisting of 100 trees.}
-
-\subsubsection*{East Tyrol}
-Figure~\ref{fig:all-map} \hlc{shows that, contrary to the rest of Tyrol, 
-for the majority of stations in East Tyrol the prespecified GAMLSS performs
-best in this particular setting (learning data: 1985--2008, testing data:
-2009--2012). East Tyrol is located south of the main Alpine Ridge and 
-exhibits different climatological characteristics than the part north of 
-the Alpine Ridge. It seems that the choice of covariates and their interactions 
-we made for the prespecified GAMLSS method is particularly adequate for the 
-region of East Tyrol for this specific analysis. However, the advantages of 
-the prespecified GAMLSS models in this particular setting (training: 1985--2008, 
-validation: 2009--2012) even out in a cross-validation framework for most of 
-the stations in East Tyrol. These results are presented in Supplement 2 
-(Stationwise Evaluation) where the exact same evaluations as in} 
-Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
-for 14 other stations.}
-
-\subsubsection*{Choice of station Axams}
-%% choice of station Axams mit Verweis auf Supplement stationwise
-\hlc{Axams was chosen for the more extensive evaluations presented
-in this paper as it is the observation station being closest 
-to Innsbruck, the capital of Tyrol and the work place of 
-three of the authors, but not based on any other particular 
-features. Furthermore, it has been checked, whether it 
-is an adequate representative of all stations regarding 
-the results of our evaluations. This is also confirmed by 
-the extended analysis in Supplement 2 (Stationwise Evaluation)
-where the 15 considered stations also include some of those where 
-the forest does not perform better than the other models in the setting 
-presented in} Figure~\ref{fig:all-map}. \hlc{It shows that the differences in 
-performance even out in the cross-validation framework where the forest 
-performs at least equally well compared to the best of the other models 
-at most stations.}
 
@@
 boosted GAMLSS). For the boosted GAMLSS the left-censored logistic distribution
-shows slightly better results, however, only within a range of less than $2\%$
-and not for all stations. Altogether, the distributional forest has proven to be 
+shows slightly better results regarding CRPS skill score, however, only within 
+a range of less than $2\%$ and not for all stations. 
+Altogether, the distributional forest has proven to be 
 very robust against distributional misspecifications.}
@@
 }
+
+\subsubsection*{Tuning parameters}
+\hlc{Setting tuning parameters of a forest model can have a major influence
+not only on the accuracy of the model but also on the computational 
+complexity. 
+For the application in} Section~\ref{sec:precipitation} \hlc{we have 
+selected the values of these parameters} (listed in Section~\ref{sec:evaluation}) 
+based on advice from literature and our own experience.
+
+\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hlc{recommend to 
+build full-grown trees. Therefore, low values should be chosen for the stopping 
+criteria} \code{minsplit} and \code{minbucket} \hcl{together with a high significance 
+level} \code{alpha}.
+\hlc{However, in a distributional model it also has to be considered that a 
+minimal number of observations is required to allow for a reasonable 
+estimation of parameters in each node. For that reason, the minimal number
+of observations for which statistical tests are performed in the 
+distributional forest in} Section~\ref{sec:precipitation} is set to
+50 (\code{minbucket}) \hlc{and the minimal number of observations on which
+parameters are estimated is set to 20} (\code{minsplit}).
+
+As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
+\hlc{random forests do not overfit, not even for an increasing number of 
+trees} (\code{ntree}). \hlc{Hence, it is recommended to build forests consisting 
+of a large number of trees as this generally leads to an increase 
+in accuracy. However, it also comes along with increasing computational costs. 
+Therefore, ``a good choice results from a trade-off between computational 
+complexity and accuracy''} (\cite{Biau+Scornet:2016}, p.~205).
+\hlc{Following this advice we chose to build forests consisting of 100 trees which
+is a commonly used value.}
+
+\subsubsection*{East Tyrol}
+Figure~\ref{fig:all-map} \hlc{shows that, contrary to the rest of Tyrol, 
+for the majority of stations in East Tyrol the prespecified GAMLSS performs
+best in this particular setting (learning data: 1985--2008, testing data:
+2009--2012). East Tyrol is located south of the main Alpine Ridge and 
+exhibits different climatological characteristics than the part north of 
+the Alpine Ridge. It seems that the choice of covariates and their interactions 
+we made for the prespecified GAMLSS method is particularly adequate for the 
+region of East Tyrol for this specific analysis. However, the advantages of 
+the prespecified GAMLSS models in this particular setting (training: 1985--2008, 
+validation: 2009--2012) even out in a cross-validation framework for most of 
+the stations in East Tyrol. These results are presented in Supplement 2 
+(Stationwise Evaluation) where the exact same evaluations as in} 
+Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
+for 14 other stations.}
+
+\subsubsection*{Choice of station Axams}
+%% choice of station Axams mit Verweis auf Supplement stationwise
+\hlc{Axams was chosen for the more extensive evaluations presented
+in this paper as it is the observation station being closest 
+to Innsbruck, the capital of Tyrol and the work place of 
+three of the authors, but not based on any other particular 
+features. Furthermore, it has been checked, whether it 
+is an adequate representative of all stations regarding 
+the results of our evaluations. This is also confirmed by 
+the extended analysis in Supplement 2 (Stationwise Evaluation)
+where the 15 considered stations also include some of those where 
+the forest does not perform better than the other models in the setting 
+presented in} Figure~\ref{fig:all-map}. \hlc{It shows that the differences in 
+performance even out in the cross-validation framework where the forest 
+performs at least equally well compared to the best of the other models 
+at most stations.}
 

### 2018-10-23 pkg/disttree/inst/draft/rejoinder.Rnw
small correction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2360 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f420e47ef5a77795180de80f135055d7d05089ea

@@
 analysis as presented for station Axams in the main manuscript for 14
-further observation stations. It shows that the distributional
-forest is %very robust against distributional (mis)specifications and (FIXME: this is shown in Supplement 1)
-almost always competitive with the best among the other methods 
-or even better. The exact results are provided in 
+further observation stations. It shows that at most stations the 
+distributional forest is competitive with the best among the other 
+methods or even better. The exact results are provided in 
 Supplement 2 (Stationwise Evaluation) and discussed in Section~4. 

### 2018-10-24 pkg/model4you/inst/JORS/man_model4you.Rnw
revision JORS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2362 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5edb103fec7acbeff2896ea7d7abcb248dff1aa9

@@
 stratified and personalised models \new{in the setting with two randomly
-assigned treatments} with a focus on ease of use and
-interpretability so that clinicians and other users can take the model they
-usually use for the estimation of the average treatment effect and with a few
-lines of code get a visualisation that is easy to understand and interpret.
+assigned treatments} with a focus on ease of use and interpretability so that
+clinicians and other users can take the model they usually use for the
+estimation of the average treatment effect and with a few lines of code get a
+visualisation that is easy to understand and interpret.
 
@@
 the two groups as \emph{treatment groups} and to the group indicator as
-\emph{treatment indicator}, which always takes values $0$
-(individual in first group) and $1$ (individual in second group).
+\emph{treatment indicator}, which always takes values $0$ (individual in first
+group) and $1$ (individual in second group).
 
@@
 @
-Note that \Sexpr{'pmtree()'} uses the data given in the call of the base
-model. It automatically uses variables not used in the model formula
-(in the example above \Sexpr{'response ~ treatment'}) as potential subgroup
-defining variables. This can be edited using the \Sexpr{'zformula'} argument.
+Note that \Sexpr{'pmtree()'} uses the data given in the call of the base model.
+It automatically uses variables not used in the model formula (in the example
+above \Sexpr{'response ~ treatment'}) as potential subgroup defining variables.
+This can be edited using the \Sexpr{'zformula'} argument.
 
@@
 data, where the similarity measure corresponds to the weight. Details on the
-personalised models can be found in \citep{seibold_individual_2017}.
-Computing personalised models for all observations in the training data is
-simple:
+personalised models can be found in \citep{seibold_individual_2017}.  Computing
+personalised models for all observations in the training data is simple:
 <<eval=FALSE>>=
@@
 @
-Again here the potential effect-modifying variables are taken by default
-as all variables not given in the model formula and can be defined using
-the \Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.
+Again here the potential effect-modifying variables are taken by default as all
+variables not given in the model formula and can be defined using the
+\Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.
 
@@
 @
-\new{To investigate the correlation between exam group and exam performance,
-we compute a simple linear model regressing the percentage points of
-correct answers on the exam group.}
+\new{To investigate the correlation between exam group and exam performance, we
+compute a simple linear model regressing the percentage points of correct
+answers on the exam group.}
 <<math_bmod>>=
@@
 @
-Both the estimates and confidence intervals and the density curves suggest
-that there is almost no difference between the two groups. But does this
-really hold for all types of students?
+Both the estimates and confidence intervals and the density curves suggest that
+there is almost no difference between the two groups. But does this really hold
+for all types of students?
 
@@
 We restrict the depth of the tree to two (\Sexpr{'maxdepth = 2'}) for
-illustration purposes. If this setting is not used a more complex tree would
-be estimated.
+illustration purposes. If this setting is not used a more complex tree would be
+estimated. Also the cutpoints and effect estimates are rounded in the plot.  To
+view the tree in the R console, the command \Sexpr{'print()'} function can be
+used.
 }
+<<math_printtree>>=
+print(tr_math)
+@
 The tree (see Figure~\ref{fig:math_tree}) divides students based on the
@@
 for differences in exam performance).
-\new{Note that confidence intervals shown in the tree are not adjusted and should
-hence be interpreted as a measure for variablity and not be used for inference.}
+\new{Note that confidence intervals shown in the tree are not adjusted and
+should hence be interpreted as a measure for variablity and not be used for
+inference.}
 
@@
 Since the percentage of successful online tests is measured on a grid, a
-\new{beeswarm} plot possibly shows the relationships
-even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1},
+\new{beeswarm} plot possibly shows the relationships even better than the
+scatter plot (both shown in Figure~\ref{fig:math_dp1},
 \new{
@@
 Figure~\ref{fig:math_dp21}). Figure~\ref{fig:math_dp22} shows the number of
-previous attempts on the x-axis and the color indicates the gender of each student.
-This way we can see that for students writing the exam for the first
+previous attempts on the x-axis and the color indicates the gender of each
+student.  This way we can see that for students writing the exam for the first
 time, gender seems to play a bigger role in terms of estimated individual exam
@@
 
-With the tools provided by the \emph{model4you} package it is
-very simple to create understandable stratified and personalised models and
-compelling visualisations that can be used to communicate these models.
+With the tools provided by the \emph{model4you} package it is very simple to
+create understandable stratified and personalised models and compelling
+visualisations that can be used to communicate these models.
 
@@
 Models currently available for use are linear models (function \Sexpr{'lm()'}
-in R), generalised
-linear models (\Sexpr{'glm()'}) and survival models (\Sexpr{'survreg()'} and
-\Sexpr{'coxph()'} from the R package \emph{survival}). Models are restricted
-to those with a single binary covariate (e.g. two treatment options).
-The user can simply plug the model object into \Sexpr{'pmtree()'} or
-\Sexpr{'pmforest()'} depending on whether they want subgroup wise or
-personalised models.
+in R), generalised linear models (\Sexpr{'glm()'}) and survival models
+(\Sexpr{'survreg()'} and \Sexpr{'coxph()'} from the R package \emph{survival}).
+Models are restricted to those with a single binary covariate (e.g. two
+treatment options).  The user can simply plug the model object into
+\Sexpr{'pmtree()'} or \Sexpr{'pmforest()'} depending on whether they want
+subgroup wise or personalised models.
 }
 
-The basis for these
-functionalities is provided by the \emph{partykit}
-package which is a widely
-used R package for trees and forests \citep{hothorn_partykit_2015, hothorn_partykit_2017}.
-The \emph{model4you} package provides
-wrappers for the well implemented and tested functions
+The basis for these functionalities is provided by the \emph{partykit} package
+which is a widely used R package for trees and forests
+\citep{hothorn_partykit_2015, hothorn_partykit_2017}.  The \emph{model4you}
+package provides wrappers for the well implemented and tested functions
 \Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::cforest()'} and extends the

### 2018-10-24 pkg/disttree/inst/draft/draft.Rnw
added supplements as references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2363 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 21db45aee2f2a8bfcf6b109c313cb0ef96f1d2f9

@@
 @
-\caption{\label{fig:axams-dist}\hlc{Total precipitation predictions by a 
+\caption{\label{fig:axams-dist}\protect\hlc{Total precipitation predictions by a 
 distributional forest at station Axams for July~24 in 2009, 2010, 2011 and 
@@
 
-\item \emph{EMOS:} \hcl{The basic} ensemble model output statistics
+\item \emph{EMOS:} \hlc{The basic} ensemble model output statistics
   models use the ensemble mean of total precipitation as regressor in the location
@@
 (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) \hlc{histograms for station
-Axams in Supplement 2 (Stationwise Evaluation) which contains a more 
-detailed explanation of residual QQ plots and PIT histograms.}
+Axams in Supplement 2} \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{which 
+contains a more detailed explanation of residual QQ plots and PIT histograms.}
 
@@
 \subsubsection{Distributional specifications}
-%% choice of distribution mit Verweis auf Supplement distributions
 \hlc{Choosing an adequate distributional family is a crucial step and forms
@@
 different distributional assumptions have been considered as well. 
-In Supplement 1 (Different Response Distributions) the exact
+In Supplement 1} \citep{Schlosser+Hothorn+Stauffer:2018a} \hlc{the exact
 same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
@@
 build full-grown trees. Therefore, low values should be chosen for the stopping 
-criteria} \code{minsplit} and \code{minbucket} \hcl{together with a high significance 
+criteria} \code{minsplit} and \code{minbucket} \hlc{together with a high significance 
 level} \code{alpha}.
@@
 validation: 2009--2012) even out in a cross-validation framework for most of 
-the stations in East Tyrol. These results are presented in Supplement 2 
-(Stationwise Evaluation) where the exact same evaluations as in} 
+the stations in East Tyrol. These results are presented in Supplement 2} 
+\citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{where the exact same evaluations as in} 
 Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
@@
 the results of our evaluations. This is also confirmed by 
-the extended analysis in Supplement 2 (Stationwise Evaluation)
-where the 15 considered stations also include some of those where 
+the extended analysis in Supplement 2} \citep{Schlosser+Hothorn+Stauffer:2018b}
+\hlc{where the 15 considered stations also include some of those where 
 the forest does not perform better than the other models in the setting 
@@
 QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
-(Supplement 2: Stationwise Evaluation).
+\citep[Supplement 2]{Schlosser+Hothorn+Stauffer:2018b}.
 Full replication of all results can be obtained with
@@
 also available within the \textsf{R}-Forge~project.
-The results presented in Supplement 1 (Different Response Distributions) and 
-Supplement 2 (Stationwise Evaluation) can be reproduced using
+\hlc{The results presented in Supplement 1} \citep{Schlosser+Hothorn+Stauffer:2018a} and 
+Supplement 2 \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{can be reproduced using}
 \code{demo("RainDistributions", package = "disttree")} and 

### 2018-10-25 pkg/disttree/inst/draft/rejoinder.Rnw
revision of the two remaining fixmes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2364 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 209e44913af253c0f60e60c82301c4bb03a76f9b

@@
 
-\fixme{This reply needs to be rewritten. It should be listed first
-which variables correspond to each other between the models. Then
-you can explain that the greater variation in the forest is not
-surprising. Finally, it should be pointed out where in the manuscript
-this is also discussed.}
-
-Figure 6 shows the variable importance of the distributional forest
-for one specific case (similar for other sites and cross-folds) 
-using a CRPS based variable importance measure. The subset of 
-most important covariates is similar to the set of covariates 
-used for the prespecified GAMLSS but not equivalent.
-The selection of different variations of the same covariate (cf., Table 2)
-is not surprising as the covariates are often highly correlated.
-Moreover, in the distributional forest for each node only a randomly
-drawn subset of covariates is considered. Hence, if for example the most 
-influential covariate is not included in this subset for a specific node 
-a similar covariate is very likely to be selected instead.
+The distributional forest selects very similar covariates as included in the
+prespecified GAMLSS model. For both models, total precipitation (\texttt{tp})
+and total column-integrated condensate (\texttt{tcolc}) are the two most important
+covariates. Due to the randomization within the nodes, the distributional forest selects different variations of the same covariate (e.g., \texttt{tp\_max}, \texttt{tp\_mean},
+\texttt{tp\_min}; see Figure~6 in the main manuscript and Supplement 2) while
+the prespecified GAMLSS mainly only uses one specific variation (\texttt{tp\_max}).
+The same is true for variations of \texttt{tcolc\_mean\_*}, \texttt{tcolc\_sprd\_*}
+and \texttt{tcolc\_sprd\_*}.
+
+Overall, the covariates selected by the distributional forest do not provide too 
+much insights but show that the prespecified GAMLSS model contains an appropriate 
+subset of all available covariates.
 
@@
 future research.
-%\fixme{Include one or two sentences about this as an
-%outlook in the discussion of distributional specifications.}
 
@@
 models -- and not so much on interpretation.
-
-%\fixme{Where should this be pointed out in the manuscript? 
-%Z: Both in the introduction when the ``zoo'' of regression techniques
-%is introduced -- and again in the discussion (possibly along with
-%variable selection?).}
 
@@
 
-%\fixme{Refer again to the corresponding point in the manuscript.}
-
 \begin{review}
@@
 
-\fixme{This also needs to be rewritten (probably best by Reto).
-It is a bit frustrating to start the response with comments how instable
-all of this is. Instead focus on the insights that are quite
-stable across stations.}
-
-Figure 6 is only a visualization of the 10 most important covariates 
-of the distributional forest in one particular setting (station Axams, learned 
-on 1985--2008). However, this top 10 list is not considered for any specification 
-of the boosted GAMLSS for which all available covariates are included. As 
-for the distributional forest, we have also calculated variable importance 
-based on mean increase of out-of-bag CRPS for the boosted GAMLSS model for 
-the exact same setting (station Axams, learned on 1985--2008). The resulting 
-top 10 variables are listed in the following barplot.
+The covariates selected in the distributional forest strongly overlap with the 
+ones selected in the boosted GAMLSS. Both methods select variations of the 
+forecasted total precipitation (\texttt{tp}) and the total column-integrated 
+condensate (\texttt{tcolc}). Due to the randomization in the nodes of the 
+trees, different variations of the same covariate (e.g., \texttt{tp\_max}, 
+\texttt{tp\_mean}, \texttt{tp\_min}) are selected by the distributional 
+forest while the boosted GAMLSS only uses one specific variation (\texttt{tp\_max}). 
+The following figure show the CRPS variable importance of the boosted GAMLSS 
+model for the very same training and validation period as in Figure~6 in the
+manuscript.
 
@@
         horiz = TRUE, las = 1, axes = FALSE,
-        xlab = "Variable importance: mean decrease in CRPS",
+        xlab = "Variable importance: mean increase in CRPS",
         font.axis = 3, #list(family="HersheySerif", face=3),
@@
 
-Overall it can be stated that the most important covariates of the boosted 
-GAMLSS are similar to the ones of the distributional forest even though they
-are not exactly the same ones and the influence of \texttt{tp\_max} is clearly
-stronger in the boosted GAMLSS. This can be explained by the random selection
-of possible split variables in each node of the distributional forest. Thus,
-if \texttt{tp\_max} is not available, a similar covariate such as \texttt{tp\_min}
-or \texttt{tp\_mean} are very likely to be selected instead while in the boosting
-algorithm \texttt{tp\_max} is always a possible choice.
-
-% Out of them, in the exact same setting (Axams, 1985--2008), 
-% the boosting algorithm selects 28 covariates 
-% for the location parameter, mainly related to precipitation (7) 
-% and temperature (11), and 46 covariates for the scale parameter, 
-% mainly related to precipitation (10), convective available potential 
-% energy (8) and temperature (15).
-% mu: 7 tppow, 1 cape, 2 dswrf, 2 msl, 2 pwat, 3 tcolc, 11 temperature
-% sigma: 10 tppow, 8 cape, 1 dswrf, 5 msl, 5 pwat, 2 tcolc, 15 temperature
-
-% R> names(coef(gb, parameter = "mu"))
-% [1] "bbs(tppow_min)"        "bbs(tppow_max)"        "bbs(tppow_mean1218)"  
-% [4] "bbs(tppow_mean1824)"   "bbs(tppow_mean2430)"   "bbs(tppow_sprd0612)"  
-% [7] "bbs(tppow_sprd2430)"   "bbs(capepow_sprd0612)" "bbs(dswrf_mean_mean)" 
-%[10] "bbs(dswrf_sprd_mean)"  "bbs(msl_sprd_min)"     "bbs(msl_sprd_max)"    
-%[13] "bbs(pwat_sprd_min)"    "bbs(pwat_sprd_max)"    "bbs(tcolc_mean_mean)" 
-%[16] "bbs(tcolc_mean_max)"   "bbs(tcolc_sprd_mean)"  "bbs(t500_mean_min)"   
-%[19] "bbs(t850_mean_min)"    "bbs(t500_sprd_mean)"   "bbs(t500_sprd_max)"   
-%[22] "bbs(t700_sprd_mean)"   "bbs(t700_sprd_min)"    "bbs(t700_sprd_max)"   
-%[25] "bbs(t850_sprd_min)"    "bbs(t850_sprd_max)"    "bbs(tdiff700850_min)" 
-%[28] "bbs(tdiff500700_max)"
-
-% R> names(coef(gb, parameter = "sigma"))
-% [1] "bbs(tppow_mean)"       "bbs(tppow_sprd)"       "bbs(tppow_min)"       
-% [4] "bbs(tppow_max)"        "bbs(tppow_mean1824)"   "bbs(tppow_mean2430)"  
-% [7] "bbs(tppow_sprd0612)"   "bbs(tppow_sprd1218)"   "bbs(tppow_sprd1824)"  
-%[10] "bbs(tppow_sprd2430)"   "bbs(capepow_sprd)"     "bbs(capepow_min)"     
-%[13] "bbs(capepow_max)"      "bbs(capepow_mean1218)" "bbs(capepow_mean1224)"
-%[16] "bbs(capepow_mean1230)" "bbs(capepow_sprd0612)" "bbs(capepow_sprd1218)"
-%[19] "bbs(dswrf_sprd_mean)"  "bbs(msl_mean_mean)"    "bbs(msl_mean_max)"    
-%[22] "bbs(msl_sprd_mean)"    "bbs(msl_sprd_min)"     "bbs(pwat_mean_mean)"  
-%[25] "bbs(pwat_mean_min)"    "bbs(pwat_mean_max)"    "bbs(pwat_sprd_min)"   
-%[28] "bbs(pwat_sprd_max)"    "bbs(tmax_mean_mean)"   "bbs(tmax_mean_min)"   
-%[31] "bbs(tcolc_mean_min)"   "bbs(tcolc_sprd_max)"   "bbs(t500_mean_mean)"  
-%[34] "bbs(t700_mean_mean)"   "bbs(t700_mean_max)"    "bbs(t850_mean_min)"   
-%[37] "bbs(t850_mean_max)"    "bbs(t500_sprd_max)"    "bbs(t850_sprd_min)"   
-%[40] "bbs(t850_sprd_max)"    "bbs(tdiff700850_min)"  "bbs(tdiff700850_max)" 
-%[43] "bbs(tdiff500700_mean)" "bbs(tdiff500700_min)"  "bbs(tdiff500700_max)" 
-%[46] "bbs(msl_diff)"
+Due to graphical reasons, the plots only show the 10 \textit{most important}
+covariates based on mean increase in CRPS, however, the models (distributional
+forest and boosted GAMLSS) are allowed to use as many covariates as required.
 

### 2018-10-25 pkg/model4you/inst/JORS/man_model4you.Rnw
Torsten's comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2365 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: adf2266b1eaf897b98f98d22a1b1f79b7322b49b

@@
 \section*{Paper Author Roles and Affiliations}
-1. PhD student; Biostatistics Department, Epidemiology, Biostatistics \&
-Prevention Institute, University of Zurich; \new{Postdoc; Institute for Medical
+1. \new{Postdoc; Institute for Medical
 Information Processing, Biometry, and Epidemiology, LMU Munich}
@@
 characteristics of the individuals.  Model-based random forests allow for a
-similarity estimation between individuals in terms of model parameters (e.g.\
+similarity estimation between individuals in terms of model parameters (e.g.~
 intercept and treatment effect). The similarity measure can then be used to
@@
 
-Currently the R package \emph{model4you} can handle only
-
 Treatment effect estimation is often done using simple models with the binary
@@
 data into groups. If, for example the algorithm finds that men and women have
-differing treatment effects, the data is partitioned into two subgroups.
+differing treatment effects, the data are partitioned into two subgroups.
 Details on model-based trees in general can be found in
@@
 
-With the tools provided by the \emph{model4you} package it is very simple to
-create understandable stratified and personalised models and compelling
-visualisations that can be used to communicate these models.
+With the tools provided by the \emph{model4you} package it is very
+simple to create understandable stratified and personalised models
+and compelling visualisations that can be used to communicate these
+models.
 
 \section*{Implementation and architecture}
-The R package \emph{model4you} is focused on ease of use and interpretability.
-Users can take a simple model that they know and understand as basis.
+The R package \emph{model4you} is focused on ease of use and
+interpretability.  Users can take a simple model that they know and
+understand as basis.
 \new{
-Models currently available for use are linear models (function \Sexpr{'lm()'}
-in R), generalised linear models (\Sexpr{'glm()'}) and survival models
-(\Sexpr{'survreg()'} and \Sexpr{'coxph()'} from the R package \emph{survival}).
-Models are restricted to those with a single binary covariate (e.g. two
-treatment options).  The user can simply plug the model object into
-\Sexpr{'pmtree()'} or \Sexpr{'pmforest()'} depending on whether they want
-subgroup wise or personalised models.
+Models currently available for use are linear models (function
+\Sexpr{'lm()'} in R), generalised linear models (\Sexpr{'glm()'}) and
+survival models (\Sexpr{'survreg()'} and \Sexpr{'coxph()'} from the R
+package \emph{survival}).  Models are restricted to those with a
+single binary covariate (e.g.~ two treatment options).  The user can
+simply plug the model object into \Sexpr{'pmtree()'} or
+\Sexpr{'pmforest()'} depending on whether they want subgroup wise or
+personalised models.
 }
 
-The basis for these functionalities is provided by the \emph{partykit} package
-which is a widely used R package for trees and forests
-\citep{hothorn_partykit_2015, hothorn_partykit_2017}.  The \emph{model4you}
-package provides wrappers for the well implemented and tested functions
-\Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::cforest()'} and extends the
-functionalities to allow for the computation of personalised models and to
-improve usability and interpretability.
+The basis for these functionalities is provided by the
+\emph{partykit} package which is a widely used R package for trees
+and forests \citep{hothorn_partykit_2015, hothorn_partykit_2017}.
+The \emph{model4you} package provides wrappers for the well
+implemented and tested functions \Sexpr{'partykit::ctree()'} and
+\Sexpr{'partykit::cforest()'} and extends the functionalities to
+allow for the computation of personalised models and to improve
+usability and interpretability.
 
-The \emph{partykit} package provides the basis for functionalities in other
-packages namely \emph{glmertree}, \emph{psychotree}, \emph{betareg} (all on
-CRAN), \emph{trtf}, \emph{disttree}, \emph{lagsarlmtree} and \emph{palmtree}
-(all available on R-Forge, publishing on CRAN planned).
+The \emph{partykit} package provides the basis for functionalities in
+other packages namely \emph{glmertree}, \emph{psychotree},
+\emph{betareg}, \emph{disttree}, \emph{lagsarlmtree}, \emph{palmtree}
+(all on CRAN), and \emph{trtf} (currently avilable on R-Forge).
 
@@
 treatment effect for all patients is too strict and the efficacy of the
-treatment depends on patient characteristics (e.g. gender, biomarkers, etc.).
+treatment depends on patient characteristics (e.g.~ gender, biomarkers, etc.).
 For subgroup analyses (stratified treatment effects) model-based trees

### 2018-10-27 pkg/disttree/inst/draft/draft.Rnw
small corrections in references and revisions based on Torsten's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2366 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0015264e7bafac16500e253e54cd5a13dad363e6

@@
 making it very difficult to keep the model manageable regarding computational
-costs but also interpretability. Therefore, only main effects are considered
-for the distributional forest and the boosted GAMLSS which both automatically
-select covariates while meteorological expert knowledge was applied to 
-select covariates and specify interactions for the prespecified GAMLSS.
+costs but also interpretability. 
+Therefore, only additive effects are considered for the boosted GAMLSS which
+automatically selects covariates while meteorological expert knowledge was 
+applied to select covariates and specify interactions for the prespecified 
+GAMLSS. For the distributional forest no prespecification is required as it
+automatically selects covariates and also detects interaction effects.
 
@@
 \hlc{For example, in transformation forests} \citep{Hothorn+Zeileis:2017} 
-\hlc{a very flexible five-parametric distribution family is employed such 
-that skewness can be modeled explicitly.
+\hlc{a very flexible parametric distribution characterised by the transformation 
+function is employed such that higher moments can be modeled implicitly.
 Evaluating further distributions might be of interest but this would go beyond the
-scope of this manuscript.
-}
+scope of this manuscript.}
 
@@
 
+%\begin{supplement}[id=supp1]
+%  \sname{Supplement 1}
+%  \stitle{Different Response Distributions}
+%  \slink[doi]{10.1214/00-AOASXXXXSUPP}
+%  \sdatatype{.pdf}" 
+%  \sdescription{
+%  In the case study presented in the main manuscript ``Distributional Regression
+%  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+%  the novel distributional regression forests are based on a Gaussian response distribution,
+%  left-censored at zero to accomodate both the non-negativity of precipitation and
+%  the point mass at zero. To assess the goodness of fit of this response distribution
+%  this supplement employs the same evaluations as in the main manuscript but based on
+%  two other distributional assumptions: A logistic distribution, left-censored at zero,
+%  is employed to potentially better capture heavy tails -- and a two-part hurdle model
+%  combining a binary model for zero vs.\ positive precipitation and a Gaussian model,
+%  truncated at zero, for the positive precipitation observations. The latter could
+%  potentially better discriminate determinants of the occurrence of precipitation
+%  vs.\ the amount of precipitation given that it occurs.
+%}
+%\end{supplement}
+
+%\begin{supplement}[id=supp2]
+%  \sname{Supplement 2}
+%  \stitle{Stationwise Evaluation}
+%  \slink[doi]{10.1214/00-AOASXXXXSUPP}
+%  \sdatatype{.pdf}" 
+%  \sdescription{
+%  In the case study presented in the main manuscript ``Distributional Regression
+%  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+%  the novel distributional regression forests are evaluated and compared with
+%  other distributional regression approaches on 95 meteorological stations.
+%  Moreover, further details -- including a cross-validation analysis -- are
+%  preformed for only one station, Axams. To show that Axams is a fairly typical
+%  station and similar insights can be obtained for other stations as well,
+%  this supplements presents the same analysis as in Section~3.3 of the main
+%  manuscript for 14~further meteorological stations.
+%}
+%\end{supplement}
+
 \newpage

### 2018-10-31 pkg/disttree/inst/draft/draft.Rnw
changed numbering of supplements (A and B instead of 1 and 2)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2369 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9093ae989a286d620c85bcfcb92b68ad87d3c504

@@
   \item \emph{Generalized additive models for location, scale, and shape}
-    (GAMLSS, \citealt{Rigby+Stasinopoulos:2005}) adopting a probabilistic modeling
+    (GAMLSS, \citealp{Rigby+Stasinopoulos:2005}) adopting a probabilistic modeling
     approach. In GAMLSS, each parameter of a statistical distribution can depend
@@
 \begin{itemize}
-  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
+  \item \emph{Regression trees} (\citealp{Breiman+Friedman+Stone:1984}) recursively
     split the data into more homogeneous subgroups and can thus capture abrupt shifts
@@
     their interactions. 
-  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
+  \item \emph{Random forests} (\citealp{Breiman:2001}) average the predictions of
     an ensemble of trees fitted to resampled versions of the learning data. This
@@
 daily precipitation sums provided by the National Hydrographical Service 
-(\citealt{ehyd}) and numerical weather forecasts from the U.S.~National
+(\citealp{ehyd}) and numerical weather forecasts from the U.S.~National
 Oceanic and Atmospheric Administration (NOAA).
@@
 The predictive performance in terms of full probabilistic forecasts is
-assessed using the continuous ranked probability score (CRPS, \citealt{Hersbach:2000}).
+assessed using the continuous ranked probability score (CRPS, \citealp{Hersbach:2000}).
 For each of the models this assesses the discrepancy of the
@@
 \hlc{This is also supported by the corresponding probability integral transform}
-(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) \hlc{histograms for station
-Axams in Supplement 2} \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{which 
+(PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) \hlc{histograms for station
+Axams in Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{which 
 contains a more detailed explanation of residual QQ plots and PIT histograms.}
@@
 different distributional assumptions have been considered as well. 
-In Supplement 1} \citep{Schlosser+Hothorn+Stauffer:2018a} \hlc{the exact
+In Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a} \hlc{the exact
 same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
@@
 validation: 2009--2012) even out in a cross-validation framework for most of 
-the stations in East Tyrol. These results are presented in Supplement 2} 
+the stations in East Tyrol. These results are presented in Supplement B} 
 \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{where the exact same evaluations as in} 
@@
 the results of our evaluations. This is also confirmed by 
-the extended analysis in Supplement 2} \citep{Schlosser+Hothorn+Stauffer:2018b}
+the extended analysis in Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b}
 \hlc{where the 15 considered stations also include some of those where 
@@
 GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
-(\citealt{Stasinopoulos+Rigby:2007}, version~\Sexpr{packageVersion("gamlss.dist")})
+(\citealp{Stasinopoulos+Rigby:2007}, version~\Sexpr{packageVersion("gamlss.dist")})
 or with custom lists containing all required information about the distribution family.
@@
 QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
-\citep[Supplement 2]{Schlosser+Hothorn+Stauffer:2018b}.
+\citep[Supplement B]{Schlosser+Hothorn+Stauffer:2018b}.
 Full replication of all results can be obtained with
@@
 also available within the \textsf{R}-Forge~project.
-\hlc{The results presented in Supplement 1} \citep{Schlosser+Hothorn+Stauffer:2018a} and 
-Supplement 2 \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{can be reproduced using}
+\hlc{The results presented in Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a} and 
+Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{can be reproduced using}
 \code{demo("RainDistributions", package = "disttree")} and 
@@
 %\begin{supplement}[id=supp1]
-%  \sname{Supplement 1}
+%  \sname{Supplement A}
 %  \stitle{Different Response Distributions}
@@
 %\begin{supplement}[id=supp2]
-%  \sname{Supplement 2}
+%  \sname{Supplement B}
 %  \stitle{Stationwise Evaluation}
@@
 A different framework to build a tree is provided by the MOB algorithm 
-which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).
+which is based on M-fluctuation tests (\citealp{Zeileis+Hothorn+Hornik:2008}).
 
@@
 %As an alternative to QQ~plots based on randomized quantile residuals,
-%probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
+%probability integral transform (PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) 
 %histograms are a commonly used tool to assess how well-calibrated distributional 

### 2018-10-31 pkg/disttree/inst/draft/rejoinder.Rnw
changed numbering of supplements (A and B instead of 1 and 2)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2369 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9093ae989a286d620c85bcfcb92b68ad87d3c504

@@
 
-\item \emph{Supplement 1: Different response distributions.}\\
+\item \emph{Supplement A: Different response distributions.}\\
 To assess the appropriateness and robustness of the distributional
@@
 
-\item \emph{Supplement 2: Stationwise evaluation.}\\
+\item \emph{Supplement B: Stationwise evaluation.}\\
 To show that the more extensive cross-validation results -- obtained only
@@
 
-% \vspace*{-2.8cm}
-% \hspace*{-0.3cm} %\includegraphics{zeileis-signature}
+\vspace*{-2.8cm}
+\hspace*{-0.3cm} 
+\includegraphics[width = 0.23\textwidth]{signature-Schlosser}
 
@@
 still operates observation-wise) but not prohibitively so.
-The exact results and times can be found in Supplement 2
+The exact results and times can be found in Supplement B
 (Stationwise Evaluation).
@@
 the results of our evaluations. This is now also 
-confirmed by further analysis in Supplement 2 (Stationwise Evaluation) 
+confirmed by further analysis in Supplement B (Stationwise Evaluation) 
 and discussed in Section~4.
@@
 the PIT, as is the case here. A more detailed 
-explanation can be found in Supplement 2 (Stationwise Evaluation).
+explanation can be found in Supplement B (Stationwise Evaluation).
 
@@
 issue. The results of an extended evaluation including a cross-validation for a 
-set of $15$ stations can be found in Supplement 2 (Stationwise Evaluation).
+set of $15$ stations can be found in Supplement B (Stationwise Evaluation).
 
@@
 regarding fitting and prediction times can now be found in
-Supplement 2 (Stationwise Evaluation).
+Supplement B (Stationwise Evaluation).
 
@@
 covariates. Due to the randomization within the nodes, the distributional forest selects different variations of the same covariate (e.g., \texttt{tp\_max}, \texttt{tp\_mean},
-\texttt{tp\_min}; see Figure~6 in the main manuscript and Supplement 2) while
+\texttt{tp\_min}; see Figure~6 in the main manuscript and Supplement B) while
 the prespecified GAMLSS mainly only uses one specific variation (\texttt{tp\_max}).
@@
 We have carried out the same evaluations as presented in the main manuscript and 
-present the full results in Supplement 1 (Different Response Distributions). 
+present the full results in Supplement A (Different Response Distributions). 
 The main insights from this are also discussed in Section~4:
@@
 methods or even better. The exact results are provided in 
-Supplement 2 (Stationwise Evaluation) and discussed in Section~4. 
+Supplement B (Stationwise Evaluation) and discussed in Section~4. 
 Especially for meteorologists intending to apply model output 
@@
 (Appendix A) and extended Appendix B to an online supplement
-(Supplement 2: Stationwise Evaluation).
+(Supplement B: Stationwise Evaluation).
 
@@
 quantiles of the standard normal distribution. A more detailed 
-explanation can be found in Supplement 2 (Stationwise Evaluation) 
+explanation can be found in Supplement B (Stationwise Evaluation) 
 which contains both, PIT histograms and residual QQ plots, 

### 2018-10-31 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2367 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0d580b958d1f6ff70bb8f8a5afe8157b8006a68d
changed numbering of supplements (A and B instead of 1 and 2)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2369 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9093ae989a286d620c85bcfcb92b68ad87d3c504

@@
 
-\title{Supplement 2: Stationwise Evaluation}
+\title{Supplement B: Stationwise Evaluation}
 
@@
   station and similar insights can be obtained for other stations as well,
-  this supplements presents the same analysis as in Section~3.3 of the main
+  this supplement presents the same analysis as in Section~3.3 of the main
   manuscript for 14~further meteorological stations.
@@
 \hyperlink{sumGinzling.1}{Ginzling}). Naturally, there is more variation
-in the single-split results with another models performing better than the
+in the single-split results with another model performing better than the
 distributional forest. However, in the cross-validation such differences

### 2018-10-31 pkg/disttree/inst/draft/draft.Rnw
small revisions: added information on supplementary material; now Appendix before references; typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2371 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1c9e1be81d8df61a522af0f5397b6a7b825a7557

@@
 %% packages
-\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}
+\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref}
 
@@
 segment:
-$$
+\begin{equation}
 w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
-$$
+\end{equation}
 where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
@@
 $\bm{\hat \theta}(\bold{z})$ where
-$$
+\begin{equation}
 \bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
-$$
+\end{equation}
 \hlc{Even though this parameter depends on the given $\bold{z}$ it has to be noted that
@@
 all trees is considered. In that way changes in the location  of the response 
-across the covariates are detected} (e.g. in Breiman and Cutler's random forests, 
-\cite{Breiman:2001}).
+across the covariates are detected} (e.g., in Breiman and Cutler's random forests, 
+\citealp{Breiman:2001}).
 \hlc{This idea is now taken one step further by embedding it in a distributional 
@@
 \citep{Lin+Jeon:2006} from the $T$~trees in the forest
-$$
+\begin{equation}
 w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
 \frac{\mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
-$$
+\end{equation}
 where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
@@
 observation and can be obtained by
-$$
+\begin{equation}
 \bm{\hat \theta}(\bold{z}) =  \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
-$$
+\end{equation}
 In summary, this yields a parametric distributional regression model 
@@
 Many statistical weather forecasting models leverage the strengths of modern
-numerical ensemble prediction systems (EPSs; see \cite{Bauer+Thorpe+Brunet:2015}).
+numerical ensemble prediction systems (EPSs; see \citealp{Bauer+Thorpe+Brunet:2015}).
 \hlc{EPSs not only predict the most likely future state of the atmosphere but provide
@@
 predicted distribution function $F$ from the observation $y$ by
-$$
+\begin{equation}
 \text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
-$$
+\end{equation}
 where \(\mathbf{1}(\cdot)\) is the indicator function. In the subsequent
@@
 the reference method is computed:
-$$
+\begin{equation}
 \text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{EMOS}}}.
-$$
+\end{equation}
 
@@
 main effects and interaction effects can be considered allowing for easier 
-interpretations} (see for example \cite{Wood+Scheipl+Faraway:2013}, 
-\cite{Goicoa+Adin+Ugarte:2018}, \cite{Ugarte+Adin+Goicoa:2017}).
+interpretations} (see for example \citealp{Wood+Scheipl+Faraway:2013}; 
+\citealp{Goicoa+Adin+Ugarte:2018}; \citealp{Ugarte+Adin+Goicoa:2017}).
 \hlc{However, in the application presented in} Section~\ref{sec:precipitation}
@@
 \hlc{For example, in transformation forests} \citep{Hothorn+Zeileis:2017} 
-\hlc{a very flexible parametric distribution characterised by the transformation 
+\hlc{a very flexible parametric distribution characterized by the transformation 
 function is employed such that higher moments can be modeled implicitly.
@@
 Therefore, ``a good choice results from a trade-off between computational 
-complexity and accuracy''} (\cite{Biau+Scornet:2016}, p.~205).
+complexity and accuracy''} (\citealp{Biau+Scornet:2016}, p.~205).
 \hlc{Following this advice we chose to build forests consisting of 100 trees which
@@
 
-\bibliography{ref.bib}
-
-%\begin{supplement}[id=supp1]
-%  \sname{Supplement A}
-%  \stitle{Different Response Distributions}
-%  \slink[doi]{10.1214/00-AOASXXXXSUPP}
-%  \sdatatype{.pdf}" 
-%  \sdescription{
-%  In the case study presented in the main manuscript ``Distributional Regression
-%  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
-%  the novel distributional regression forests are based on a Gaussian response distribution,
-%  left-censored at zero to accomodate both the non-negativity of precipitation and
-%  the point mass at zero. To assess the goodness of fit of this response distribution
-%  this supplement employs the same evaluations as in the main manuscript but based on
-%  two other distributional assumptions: A logistic distribution, left-censored at zero,
-%  is employed to potentially better capture heavy tails -- and a two-part hurdle model
-%  combining a binary model for zero vs.\ positive precipitation and a Gaussian model,
-%  truncated at zero, for the positive precipitation observations. The latter could
-%  potentially better discriminate determinants of the occurrence of precipitation
-%  vs.\ the amount of precipitation given that it occurs.
-%}
-%\end{supplement}
-
-%\begin{supplement}[id=supp2]
-%  \sname{Supplement B}
-%  \stitle{Stationwise Evaluation}
-%  \slink[doi]{10.1214/00-AOASXXXXSUPP}
-%  \sdatatype{.pdf}" 
-%  \sdescription{
-%  In the case study presented in the main manuscript ``Distributional Regression
-%  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
-%  the novel distributional regression forests are evaluated and compared with
-%  other distributional regression approaches on 95 meteorological stations.
-%  Moreover, further details -- including a cross-validation analysis -- are
-%  preformed for only one station, Axams. To show that Axams is a fairly typical
-%  station and similar insights can be obtained for other stations as well,
-%  this supplements presents the same analysis as in Section~3.3 of the main
-%  manuscript for 14~further meteorological stations.
-%}
-%\end{supplement}
-
-\newpage
+\section*{Appendix}
 \begin{appendix}
@@
 dependencies between the score values
-$$
+\begin{equation}
 s(\bm{\hat{\theta}}, y) = 
@@
 \end{pmatrix}
-$$
+\end{equation}
 and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
-\begin{align*}
+\begin{align}
 H_0^l:  s(\bm{\hat{\theta}}, Y) \qquad \bot \qquad Z_l 
-\end{align*}
+\end{align}
 
@@
 the multivariate linear statistic
-$$
+\begin{equation}
 T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\bm{\hat{\theta}}, Y_i)\right)
-$$
+\end{equation}
 \hlc{where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
@@
 In the application of this paper a quadratic form is chosen, such that}
-$$
+\begin{equation}
 c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
-$$
+\end{equation}
 \hlc{where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.} Alternatively,
@@
 This is measured by the linear statistic
-$$
+\begin{equation}
 T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\bm{\hat{\theta}}, Y_i)
-$$
+\end{equation}
 \hlc{for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
@@
 The split point is then chosen as follows:}
-$$
+\begin{equation}
 r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
-$$
+\end{equation}
 One repeats the testing and splitting procedure in each of the resulting 
@@
 
+\section*{Supplementary material}
+\subsection*{Supplement A: Different Response Distributions}
+(doi: \href{http://dx.doi.org/10.1214/00-AOASXXXXSUPPA}{10.1214/00-AOASXXXXSUPPA}).
+  %In the case study presented in the main manuscript ``Distributional Regression
+  %Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+  %the novel distributional regression forests are based on a Gaussian response distribution,
+  %left-censored at zero to accomodate both the non-negativity of precipitation and
+  %the point mass at zero. 
+  To assess the goodness of fit of the Gaussian distribution, left-censored at zero,
+  this supplement employs the same evaluations as in the main manuscript but based on
+  two other distributional assumptions: A logistic distribution, left-censored at zero,
+  is employed to potentially better capture heavy tails -- and a two-part hurdle model
+  combining a binary model for zero vs.\ positive precipitation and a Gaussian model,
+  truncated at zero, for the positive precipitation observations. 
+  %The latter could
+  %potentially better discriminate determinants of the occurrence of precipitation
+  %vs.\ the amount of precipitation given that it occurs.
+
+\subsection*{Supplement B: Stationwise Evaluation}
+(doi: \href{http://dx.doi.org/10.1214/00-AOASXXXXSUPPB}{10.1214/00-AOASXXXXSUPPB}).
+  %In the case study presented in the main manuscript ``Distributional Regression
+  %Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
+  %the novel distributional regression forests are evaluated and compared with
+  %other distributional regression approaches on 95 meteorological stations.
+  %Moreover, further details -- including a cross-validation analysis -- are
+  %preformed for only one station, Axams. 
+  To show that Axams is a fairly typical
+  station and similar insights can be obtained for other stations as well,
+  this supplement presents the same analysis as in Section~3.3 of the main
+  manuscript for 14~further meteorological stations.
+
+\bibliography{ref.bib}
+
 \end{document}

### 2018-10-31 pkg/disttree/inst/draft/rejoinder.Rnw
small revisions: added information on supplementary material; now Appendix before references; typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2371 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1c9e1be81d8df61a522af0f5397b6a7b825a7557

@@
 extended to regarding discussion of: variable selection, tuning parameters,
-distributional specifications (referring to Supplement~1), and regional
-differences (referring to Supplement~2).
+distributional specifications (referring to Supplement~A), and regional
+differences (referring to Supplement~B).
 
@@
 using variable selection by cross-validation. Fitting and forecasting
-times are reported now in Supplement~2 (Stationwise Evaluation).
+times are reported now in Supplement~B (Stationwise Evaluation).
 
@@
 approach is not only applied to Axams but also 14 further stations
-in Supplement~2 to check for robustness across stations.
+in Supplement~B to check for robustness across stations.
 

### 2018-10-31 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
small improvement
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2349 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 149a8be8f8fa3bbed0d3f0b6c11a9868e3144036
changed numbering of supplements (A and B instead of 1 and 2)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2369 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9093ae989a286d620c85bcfcb92b68ad87d3c504
small revisions: added information on supplementary material; now Appendix before references; typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2371 255f4249-2263-4417-b1ea-8498

@@
 
-\title{Supplement 1: Different Response Distributions}
+\title{Supplement A: Different Response Distributions}
 
@@
   the novel distributional regression forests are based on a Gaussian response distribution,
-  left-censored at zero to accomodate both the non-negativity of precipitation and
+  left-censored at zero to accommodate both the non-negativity of precipitation and
   the point mass at zero. To assess the goodness of fit of this response distribution
@@
 
-The main manuscript employes a Gaussian distribution with probability
+The main manuscript employs a Gaussian distribution with probability
 density function $f$ (PDF) and cumulative distribution function $F$ (CDF)
@@
 %
-Both distributions have to be adapted in our case study to accomodate that
+Both distributions have to be adapted in our case study to accommodate that
 precipitation is non-negative with a point mass at zero (= no rain).
@@
 for hurdle Gaussian and censored logistic distributions in reference to the
-censored Gaussian distribution (orange horizontal line). Gray lines correspond to the 15~stations and are aggregated in the boxplot.}
+censored Gaussian distribution (orange horizontal line). Gray lines
+correspond to the 15~stations that are aggregated in the boxplot.}
 \end{figure}
 
-\subsubsection*{FIXME: FURTHER GRAPHICS?}
+%\subsubsection*{FIXME: FURTHER GRAPHICS?}
 
-\emph{The same data already shown in Figures~\ref{fig:lattice} 
-and~\ref{fig:box_crpss_gaussian} can be displayed in further ways. However, these do 
-not really add much and hence should probably be omitted.}
+%\emph{The same data already shown in Figures~\ref{fig:lattice} 
+%and~\ref{fig:box_crpss_gaussian} can be displayed in further ways. However, these do 
+%not really add much and hence should probably be omitted.}
 
-Figures~\ref{fig:box_crpss_EMOS} and \ref{fig:box_crps} compare the methods separately 
-for each distribution, once based on CRPS skill score values with reference model EMOS 
-and once based on CRPS values. Both confirm the conclusion that all three distributions
-lead to almost the same results for all methods.
+%Figures~\ref{fig:box_crpss_EMOS} and \ref{fig:box_crps} compare the methods separately 
+%for each distribution, once based on CRPS skill score values with reference model EMOS 
+%and once based on CRPS values. Both confirm the conclusion that all three distributions
+%lead to almost the same results for all methods.
 
@@
 
-\begin{figure}[t!]
-\setkeys{Gin}{width=0.99\textwidth}
-<<results_box_crpss_EMOS, fig=TRUE, echo=FALSE, height=3, width=7.2>>=
+%\begin{figure}[t!]
+%\setkeys{Gin}{width=0.99\textwidth}
+<<results_box_crpss_EMOS, fig=TRUE, echo=FALSE, height=3, width=7.2, eval=FALSE>>=
 # crps skill score by method (reference: EMOS)
@@
 @
-\caption{\label{fig:box_crpss_EMOS}CRPS skill score values of each of 
-the 15 observation stations, separated by method and distribution. 
-EMOS is chosen as reference model.}
-\end{figure}
+%\caption{\label{fig:box_crpss_EMOS}CRPS skill score values of each of 
+%the 15 observation stations, separated by method and distribution. 
+%EMOS is chosen as reference model.}
+%\end{figure}
 
-\begin{figure}[t!]
-\setkeys{Gin}{width=0.99\textwidth}
-<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7.2>>=
+%\begin{figure}[t!]
+%\setkeys{Gin}{width=0.99\textwidth}
+<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7.2, eval=FALSE>>=
 
@@
 @
-\caption{\label{fig:box_crps}CRPS values of each of the 15 observation stations, 
-separated by method and distribution.}
-\end{figure}
+%\caption{\label{fig:box_crps}CRPS values of each of the 15 observation stations, 
+%separated by method and distribution.}
+%\end{figure}
 

### 2018-10-31 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
small revisions: added information on supplementary material; now Appendix before references; typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2371 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1c9e1be81d8df61a522af0f5397b6a7b825a7557

@@
 observation equal to 0 the corresponding PIT value is multiplied with a randomized
-value following a uniform distribution on \[0,1\]. In that way these values
+value following a uniform distribution on $[0,1]$. In that way these values
 are apportioned on those bins of the histogram that would not contain any

### 2018-11-01 pkg/disttree/inst/draft/draft.Rnw
small touch-ups in abstract/intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2372 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4eb56ab7907fb1c328e898785cfe442533ec4c10

@@
 popular in weather forecasting. 
-% More generally, the GAMLSS framework allows to establish generalized additive 
-% models for location, scale, and shape with smooth linear or nonlinear effects.
-\hlc{Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.}
+\hlc{Moreover, generalized additive models for location, scale, and shape (GAMLSS)
+provide a framework where each distribution parameter is modeled separately
+capturing smooth linear or nonlinear effects.}
 %
@@
 
-\hlc{However, distributional forests should not be considered as a 
-replacement of GAMLSS but rather as a complementing method in the field of
-distributional models, similar to the way how GAM and random forests are complements 
-in the conditional mean regression toolbox.}
-
-In the following, particular focus is given to distributional forests as a method for
-obtaining probabilistic forecasts by leveraging the strengths of random forests: the 
-ability to capture both smooth and abruptly changing functions along with simultaneous selection of variables and possibly complex interactions.
-Thus, these properties make the method particularly appealing in case of many covariates with unknown effects and
-interactions where it would be challenging to specify a distributional regression model 
-like GAMLSS.
+In the following, particular focus is given to distributional forests as a
+method for obtaining probabilistic forecasts by leveraging the strengths of
+random forests: the  ability to capture both smooth and abruptly changing
+functions along with simultaneous selection of variables and possibly complex
+interactions. Thus, these properties make the method particularly appealing in
+case of many covariates with unknown effects and  interactions where it would be
+challenging to specify a distributional regression model  like GAMLSS.
+%
+\hlc{However, distributional forests should not be considered as a replacement
+of GAMLSS but rather as a complementing technique for flexible distributional
+regression -- much like GAMs and random forests are complements for conditional
+mean regression.}
 
@@
 
-In weather forecasting, these properties are especially appealing in 
+In weather forecasting, the flexibility of distributional forests is especially appealing in 
 mountainous regions and complex terrain where a wide range of local-scale

### 2018-11-02 pkg/model4you/inst/JORS/man_model4you.Rnw
consistent font of R code in text and chunks
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2373 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b971c2b79875d0ea5072932a285afc40ae63a35d

@@
 @
-with \Sexpr{'response'} being the response measured, \Sexpr{'treatment'} being
-the treatment indicator and \Sexpr{'data'} being the data set containing these
-variables.  The function \Sexpr{'model()'} can be replaced for example by
-\Sexpr{'lm()'} to estimate a linear model, \Sexpr{'glm()'} to estimate a
-generalised linear model or \Sexpr{'survreg()'} to estimate a parametric
+with \texttt{\Sexpr{'response'}} being the response measured, \texttt{\Sexpr{'treatment'}} being
+the treatment indicator and \texttt{\Sexpr{'data'}} being the data set containing these
+variables.  The function \texttt{\Sexpr{'model()'}} can be replaced for example by
+\texttt{\Sexpr{'lm()'}} to estimate a linear model, \texttt{\Sexpr{'glm()'}} to estimate a
+generalised linear model or \texttt{\Sexpr{'survreg()'}} to estimate a parametric
 survival model.  These models estimate intercept and treatment effect for all
-individuals in the \Sexpr{'data'} and allow for predicting the response of
+individuals in the \texttt{\Sexpr{'data'}} and allow for predicting the response of
 other individuals given they do or do not receive the treatment of interest.
@@
 @
-Note that \Sexpr{'pmtree()'} uses the data given in the call of the base model.
+Note that \texttt{\Sexpr{'pmtree()'}} uses the data given in the call of the base model.
 It automatically uses variables not used in the model formula (in the example
-above \Sexpr{'response ~ treatment'}) as potential subgroup defining variables.
-This can be edited using the \Sexpr{'zformula'} argument.
+above \texttt{\Sexpr{'response ~ treatment'}}) as potential subgroup defining variables.
+This can be edited using the \texttt{\Sexpr{'zformula'}} argument.
 
@@
 variables not given in the model formula and can be defined using the
-\Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.
+\texttt{\Sexpr{'zformula'}} argument in \texttt{\Sexpr{'pmforest()'}}.
 
@@
 \new{
-We restrict the depth of the tree to two (\Sexpr{'maxdepth = 2'}) for
+We restrict the depth of the tree to two (\texttt{\Sexpr{'maxdepth = 2'}}) for
 illustration purposes. If this setting is not used a more complex tree would be
 estimated. Also the cutpoints and effect estimates are rounded in the plot.  To
-view the tree in the R console, the command \Sexpr{'print()'} function can be
+view the tree in the R console, the command \texttt{\Sexpr{'print()'}} function can be
 used.
@@
 The tree (see Figure~\ref{fig:math_tree}) divides students based on the
-percentage of successful online \Sexpr{'tests'}. These online tests were
+percentage of successful online \texttt{\Sexpr{'tests'}}. These online tests were
 conducted biweekly throughout the semester.  The largest difference between the
 two exam groups is in the students who did very well in the online tests (more
-than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
+than \texttt{\Sexpr{round(tr_math[[5]]$node$split$breaks, 1)}} percent correct). The
 tree thus gives us much more information on the fairness of the exam than the
@@
 Figure~\ref{fig:math_dp11} includes a loess curve). We see a nonlinear relation
-between the percentage of successful online \Sexpr{'tests'} and the exam group
+between the percentage of successful online \texttt{\Sexpr{'tests'}} and the exam group
 effect. Students with great performance during the semenster are estimated to
@@
 For categorical variables such as the number of previous attempts to pass the
-exam (\Sexpr{'attempt'}) and \Sexpr{'gender'} we can use box plots,
+exam (\texttt{\Sexpr{'attempt'}}) and \texttt{\Sexpr{'gender'}} we can use box plots,
 \new{beeswarm} plots or a combination thereof can be used (as in
@@
 Models currently available for use are linear models (function
-\Sexpr{'lm()'} in R), generalised linear models (\Sexpr{'glm()'}) and
-survival models (\Sexpr{'survreg()'} and \Sexpr{'coxph()'} from the R
+\texttt{\Sexpr{'lm()'}} in R), generalised linear models (\texttt{\Sexpr{'glm()'}}) and
+survival models (\texttt{\Sexpr{'survreg()'}} and \texttt{\Sexpr{'coxph()'}} from the R
 package \emph{survival}).  Models are restricted to those with a
 single binary covariate (e.g.~ two treatment options).  The user can
-simply plug the model object into \Sexpr{'pmtree()'} or
-\Sexpr{'pmforest()'} depending on whether they want subgroup wise or
+simply plug the model object into \texttt{\Sexpr{'pmtree()'}} or
+\texttt{\Sexpr{'pmforest()'}} depending on whether they want subgroup wise or
 personalised models.
@@
 The \emph{model4you} package provides wrappers for the well
-implemented and tested functions \Sexpr{'partykit::ctree()'} and
-\Sexpr{'partykit::cforest()'} and extends the functionalities to
+implemented and tested functions \texttt{\Sexpr{'partykit::ctree()'}} and
+\texttt{\Sexpr{'partykit::cforest()'}} and extends the functionalities to
 allow for the computation of personalised models and to improve
@@
 For subgroup analyses (stratified treatment effects) model-based trees
-(\Sexpr{'pmtree()'}) can be used; For personalised treatment effects
-model-based forests (\Sexpr{'pmforest()'}) provide a way of estimating
+(\texttt{\Sexpr{'pmtree()'}}) can be used; For personalised treatment effects
+model-based forests (\texttt{\Sexpr{'pmforest()'}}) provide a way of estimating
 similarity between patients and using this similarity measure to estimate
-personalised models (\Sexpr{'pmodel()'}).  The target audience are people who
+personalised models (\texttt{\Sexpr{'pmodel()'}}).  The target audience are people who
 deal with heterogeneous treatment effects, such as medical researchers,

### 2018-11-02 pkg/model4you/inst/JORS/man_model4you.Rnw
comments Achim JORS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2374 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 33b6c34c30ee9921f076ef0bf033fcac905317a6

@@
 model-based trees one can compute treatment effects for different strata of
-individuals. The strata are found in a data driven fashion and depend on
+individuals. The strata are found in a data-driven fashion and depend on
 characteristics of the individuals.  Model-based random forests allow for a
@@
 estimated. Also the cutpoints and effect estimates are rounded in the plot.  To
-view the tree in the R console, the command \texttt{\Sexpr{'print()'}} function can be
+view the tree in the R console, the \texttt{\Sexpr{'print()'}} function can be
 used.
@@
 for differences in exam performance).
-\new{Note that confidence intervals shown in the tree are not adjusted and
+\new{Note that confidence intervals shown in the tree are not adjusted for
+selection of the cutpoints in the tree and
 should hence be interpreted as a measure for variablity and not be used for

### 2018-11-02 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
final Z revision of main manuscript, adapting much of the discussion further touch-ups in remaining materials
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2375 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a394dd57fb692b04399df40ba94b573455f36c1f

@@
 or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
-a truncated (positive) part \citep[see][for an overview of both techniques]{Winkelmann:2006}.
+a truncated (positive) part (see \citealp{Long:1997} or \citealp{Winkelmann:2006}
+for an overview of both techniques).
+%
 \begin{itemize}
@@
 
-Altogether it can be stated that for this application the left-censored at zero Gaussian
+Altogether it can be stated that for this application the zero-censored Gaussian
 distribution seems to be an appropriate choice while the other two distributions are also
-reasonable alternatives. However, they do not lead to an overall improvement in 
+reasonable alternatives. However, they do not lead to a pronounced overall improvement in 
 performance for this application.
@@
 can be expected to perform somewhat better. Whereas when different variables/effects drive
-the parameters of the model, then the two-part approach implemented in this supplement
+the different model parameters, then the two-part approach implemented in this supplement
 should perform somewhat better. In this particular case study, though, neither approach

### 2018-11-02 pkg/disttree/inst/draft/draft.Rnw
final Z revision of main manuscript, adapting much of the discussion further touch-ups in remaining materials
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2375 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a394dd57fb692b04399df40ba94b573455f36c1f

@@
 
-The general distributional notation is exemplified in all three steps for the
-Gaussian distribution left-censored at zero. The latter is employed in the empirical 
-case study in Section~\ref{sec:precipitation} to model power-transformed daily 
-precipitation amounts.
+The general distributional notation is exemplified in all three steps using the
+Gaussian distribution left-censored at zero (for short: zero-censored Gaussian).
+The latter is employed in the empirical case study in Section~\ref{sec:precipitation}
+to model power-transformed daily precipitation amounts.
 
@@
 while accounting for a point mass at zero (i.e., dry days without
-rain), a Gaussian distribution left-censored at zero (for short: zero-censored Gaussian)
+rain), a zero-censored Gaussian distribution
 with location parameter $\mu$ and scale parameter $\sigma$ is employed. Therefore,
@@
 
-where $\phi$ and $\Phi$ are the probability density function and the 
+where $\phi$ and $\Phi$ are the probability density function and the cumulative
 distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
@@
 Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
-for reasonably representing the response distribution and covariates
+for reasonably representing the response distribution. Therefore, covariates
 $\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
@@
 $\mathcal{Z}$ recursively into disjoint segments so that a homogenous
-distributional model for the response $Y$ can be found with segment-specific
+distributional model for the response $Y$ can be found for each segment with segment-specific
 parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
@@
   in the current subsample.
-\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
+\item Test for associations (or instabilities) of the scores $s(\bm{\hat \theta}, y_i)$ 
   and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
@@
 be ``sent down'' the tree and the corresponding segment-specific MLE has to be
-obtained. This can also be understood as a weighted MLE where the weights
+obtained. \hlc{Thus, in practice $\bm{\hat \theta}(\bold{z})$ does not have to be recalculated
+for each new $\bold{z}$ but one can simply extract the parameter estimates for
+the corresponding segment which have been computed already while
+learning the tree. However, to understand this estimator conceptually it is useful
+to denote it as a weighted MLE} where the weights
 select those observations from the learning sample that fall into the same
@@
 \end{equation}
-\hlc{Even though this parameter depends on the given $\bold{z}$ it has to be noted that
-it is the same for all other observations that are in the same terminal node of 
-the tree as $\bold{z}$ and therefore does not necessarily have to be recalculated for
-each of them.}
 
@@
 learning data obtained through resampling (bootstrap or subsampling). In each node 
-only a random subset of the covariates $\bold{Z}$ is considered to reduce the 
+only a random subset of the covariates $\bold{Z}$ is considered for splitting to reduce the 
 correlation among the trees and to stabilize the variance of the model.
@@
 \hlc{This idea is now taken one step further by embedding it in a distributional 
-framework based on loglikelihood estimation. \emph{Distributional forests} 
+framework based on maximum-likelihood estimation. \emph{Distributional forests} 
 employ an ensemble of $T$~\emph{distributional trees} which pick up changes in 
-the ``direction'' of any distribution parameter by considering the full score 
-vector for choosing splitting variables and split points.}
+the ``direction'' of any distribution parameter by considering the full likelihood
+and corresponding score function for choosing splitting variables and split points.}
 
@@
 (e.g., log rank scores in survival trees, model residuals in transformation 
-or generalized forests). Distributional forests, in contrast to these 
+or generalized forests). Distributional forests, in contrast to 
 nonparametric approaches, provide a compromise between model flexibility 
@@
 initial conditions and model specifications to account for uncertainties in both,
-the initial state and the model and its parametrizations.
-One frequently used method based on distributional regression models is the 
+the initial atmospheric state and the NWP model (and its parametrizations).
+One frequently-used method based on distributional regression models is the 
 ensemble model output statistics (EMOS) approach first proposed by}
@@
 specific quantities and sites.  
-In case of precipitation forecasting EMOS typically uses the ensemble mean of 
-``total precipitation''} (\emph{tp}) \hlc{forecasts as predictor for the location 
+In case of precipitation forecasting, EMOS typically uses the ensemble mean of 
+``total precipitation'' (\emph{tp}) forecasts as the predictor for the location 
 parameter $\mu$ and the corresponding ensemble standard deviation for the 
-scale parameter $\sigma$ of the statistical model based on the assumption
-of the response following a Gaussian distribution left-censored at zero.
+scale parameter $\sigma$, e.g., assuming the observations to follow a zero-censored
+Gaussian distribution.
 This distributional approach of modeling both parameters allows to correct 
 for possible errors of the NWP ensemble in both, the expectation but also the 
-uncertainty of a specific forecast. In order to assure positivity of the 
-scale parameter a log-link is applied. Hence, the two distribution parameters 
-are modeled by the linear functions} 
-$\mu = \beta_0 + \beta_1 \cdot \emph{mean}(\emph{tp})$ and  
-$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \emph{sd}(\emph{tp})$
-with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
-\hlc{hereafter referred to as a basic EMOS model} \citep{Gebetsberger+Messner+Mayr:2017}.
+uncertainty of a specific forecast. Thus, a basic EMOS specification typically
+models the two distribution parameters by two linear predictors, e.g.,} 
+$\mu = \beta_0 + \beta_1 \cdot \mathrm{mean}(\emph{tp})$ and  
+$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \log(\mathrm{sd}(\emph{tp}))$
+with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$, and $\gamma_1$
+\citep[where the log link assures positivity of the scale parameter, following][]{Gebetsberger+Messner+Mayr:2017}.
 
 % FIXME: notation with or without power transformation parameter?
-
-% with:
+%
 %$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
 %$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
-%with power transformation parameter $\frac{1}{p}$ and 
-%regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
-%\hlc{hereafter referred to as a basic EMOS model \citep{Gebetsberger+Messner+Mayr:2017}.}
+%with power transformation parameter $\frac{1}{p}$
 
@@
 Therefore, it is often possible to substantially improve the predictive 
-performance of the EMOS by including additional predictor variables, either 
+performance of a basic EMOS by including additional predictor variables, either 
 from local meteorological observations or an NWP model. Unfortunately, it is 
@@
 In the following, it is illustrated how distributional forests can
-solve this problem without requiring prior expert knowledge.  For fitting the
-forest only the response distribution and the list of potential predictor
+solve this problem without requiring prior expert knowledge about the meteorological covariates.
+For fitting the forest only the response distribution and the list of potential predictor
 variables need to be specified (along with a few algorithmic details) and then
@@
 initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
-0600\,UTC to 0600\,UTC of the next day). $^*$Minimum values of \emph{dswrf}
+0600\,UTC to 0600\,UTC of the next day). Note: $^*$Minimum values of \emph{dswrf}
 over 24\,h are always zero and thus neglected.}
@@
 \emph{tp}: total precipitation,           & 12  & ensemble mean of sums over 24h, \\
-\hspace*{0.5cm} power transformed (by $1.6^{-1}$) &    & ensemble std. deviation of sums over 24h, \\ 
+\hspace*{0.5cm} power transformed (by $1.6^{-1}$) &    & ensemble std.\ deviation of sums over 24h, \\ 
 \emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
@@
 \hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
-\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
+\hspace*{\fill}                            &    & ensemble std.\ deviation of sums over 6h\\
 \hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
@@
 \emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
-\emph{pwat}: precipitable water         &   & ensemble std. deviation of mean values,\\
-\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimum values$^*$,\\
-\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
+\emph{pwat}: precipitable water         &   & ensemble std.\ deviation of mean values,\\
+\emph{tmax}: 2m maximum temperature     &   & ensemble std.\ deviation of minimum values$^*$,\\
+\hspace*{1.15cm}                          &   & ensemble std.\ deviation of maximal values,\\
 \emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 \\
@@
 quality-checked by the National Hydrographical Service. NWP outputs
-are obtained from the second generation reforecast data set of the
+are obtained from the second-generation reforecast data set of the
 global ensemble forecast system \citep[GEFS,][]{Hamill+Bates+Whitaker:2013}. 
@@
 aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
-in their choice of $1.6^{-1}$ as a suitable power parameter for the
-region of Tyrol. The same power transformation is applied to both the
+in their choice of $1.6^{-1}$ as a suitable power parameter for precipitation
+in the region of Tyrol. The same power transformation is applied to both the
 observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
@@
   in Section~\ref{sec:methods}. 
-  \hlc{As stopping criteria the minimal number of observations to perform a split} 
-  (\code{minsplit}) \hlc{is set to $50$, the minimal number of observations in a 
-  terminal node} (\code{minbucket}) \hlc{is set to $20$ and the significance level 
-  for variable selection} (\code{alpha}) \hlc{is kept at the default value $1$ such 
-  that no early-stopping (sometimes also referred to as ``pre-pruning'') is applied
+  \hlc{The stopping criteria are the minimal number of observations to perform a split 
+  (\texttt{minsplit} $= 50$), the minimal number of observations in a 
+  segment (\texttt{minbucket} $= 20$), and the significance level 
+  for variable selection (\texttt{alpha} $= 1$). The latter means
+  that no early stopping (or ``pre-pruning'') is applied
   based on results of the statistical tests.}
@@
 interested in the overall predictive performance and calibration of the
-distributional fits.
-
-To assess calibration Figure~\ref{fig:axams-qq} shows residual QQ~plots for
+distributional fits. \hlc{More details of this assessment as well as
+an application to 14~further meteorological stations is provided in
+Supplement~B} \citep{Schlosser+Hothorn+Stauffer:2018b}.
+
+To assess calibration, Figure~\ref{fig:axams-qq} shows residual QQ~plots for
 out-of-sample predictions (2009--2012) from the different models trained on
@@
 
-To assess the predictive performance, a full cross-validation is carried out
-rather than relying on just the one fixed test set for the years 2009--2012.
-\hlc{For each setting in the considered 10 times 7-fold cross-validation the 
-available 28~years are split into 7~subsets of 4~randomly selected and 
-therefore mostly not subsequent years.}
-The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
-(= 4~years) using the average CRPS across all observations.
-The resulting 10~CRPS skill scores are displayed by boxplots in
-Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
-line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
-perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
-of around 4~percent, the distributional forest has a slightly higher improvement
-of around 5.5~percent in median.
-
 \begin{figure}[t!]
@@
 
+To assess the predictive performance, a full cross-validation is carried out
+rather than relying on just the one fixed test set for the years 2009--2012.
+To do so, a 10 times 7-fold cross-validation is carried out where each repetition
+splits the available 28~years into 7~subsets of 4~randomly-selected 
+\hlc{(and thus not necessarily consecutive)} years.
+The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
+(= 4~years) using the average CRPS across all observations.
+The resulting 10~CRPS skill scores are displayed by boxplots in
+Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
+line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
+perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
+of around 4~percent, the distributional forest has a slightly higher improvement
+of around 5.5~percent in median.
+
 Finally, it is of interest how this improvement in predictive performance
@@
 # HCL colors (relatively flashy, essentially CARTO Tropic)
-clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
+clr <- colorspace::diverge_hcl(5, h = c(130, 320), c = 70, l = c(50, 90), power = 1.3)
 
@@
 
-\begin{figure}[p!]
+\begin{figure}[t!]
 \centering
@@
 learned on 1985--2008 and validated for 2009--2012.}
-
-\vspace*{0.4cm}
-
+\end{figure}
+
+\begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.99\textwidth}
@@
 \caption{\label{fig:all-map}Map of Tyrol coding the best-performing model
-for each station (type of symbol). The color codes whether the distributional 
+for each station (type of symbol) when learned on 1985--2008 and validated for 2009--2012.
+The color codes whether the distributional 
 forest had higher (green) or lower (red) CRPS compared to the best of the 
@@
 for a few stations in the north-east EMOS performs best, and in East Tyrol
-the prespecified GAMLSS performs particularly well.
+the prespecified GAMLSS performs particularly well in the validation period
+(2009--2012).
 
@@
 
-\subsubsection*{Variable Selection}
+\subsubsection*{Variable selection}
+
 \hlc{Generally, there are many possibilities how to specify the variables that are
-to be included in a model. Especially for a low number of variables all types
-of interactions can be employed and mixed effect models and separations into 
-main effects and interaction effects can be considered allowing for easier 
-interpretations} (see for example \citealp{Wood+Scheipl+Faraway:2013}; 
+to be included in a distributional regression model. Especially for a low number
+of covariates, the GAMLSS approach offers a powerful framework in which
+penalized estimation of both smooth main effects and corresponding interaction surfaces
+yields models that often balance good predictive performance with high
+interpretability} (see for example \citealp{Wood+Scheipl+Faraway:2013}; 
 \citealp{Goicoa+Adin+Ugarte:2018}; \citealp{Ugarte+Adin+Goicoa:2017}).
-\hlc{However, in the application presented in} Section~\ref{sec:precipitation}
-\hlc{altogether 80 variables are available. Even if only paired interactions were
-considered this would already lead to 3160 additional covariates in the model
-making it very difficult to keep the model manageable regarding computational
-costs but also interpretability. 
-Therefore, only additive effects are considered for the boosted GAMLSS which
-automatically selects covariates while meteorological expert knowledge was 
-applied to select covariates and specify interactions for the prespecified 
-GAMLSS. For the distributional forest no prespecification is required as it
-automatically selects covariates and also detects interaction effects.
-
-As mentioned before, distributional forests should not be seen as a replacement
-for GAMLSS but rather as a complementary method. GAMLSS provide a powerful framework
-in the field of distributional modeling as they allow for the combination of 
-interpretability and a high goodness-of-fit. For distributional forests, the focus
-is mainly on forecasting skill, automatic variable selection and capturing also 
-non-additive effects rather than on interpretability. Therefore, both approaches
-provide valuable features and advantages that can complement each other in many 
-applications as it can not be expected that one of them is in general better than 
-the other. 
-}
+\hlc{However, if the number of covariates is high, including all (or many)
+main effects and interactions in a GAMLSS typically becomes challenging both in terms
+of interpretability and computational complexity/stability}
+\citep[see also][]{Hofner+Mayr+Schmid:2016}.
+
+\hlc{In the precipitation forecasting application, as presented in} Section~\ref{sec:precipitation},
+\hlc{80~covariates are considered which corresponds to 3160~potential pairwise
+interactions (and even more higher-order interactions). 
+Therefore, only main effects are considered for the boosted GAMLSS while
+the prespecified GAMLSS also includes selected interactions chosen based on
+meteorological expert knowledge. In contrast, the distributional forest requires no
+prespecification as covariates and corresponding interactions are selected automatically.
+Thus, distributional forests are an appealing alternative to (boosted) GAMLSS
+in weather forecasting tasks as the main concern is typically not so much
+interpretability but forecasting skill and (semi-)automatic application on a
+larger domain} \citep[see also the discussion in][]{Rasp+Lerch:2018}.
 
 \subsubsection{Distributional specifications}
-\hlc{Choosing an adequate distributional family is a crucial step and forms
-the basis required to build a well fitting model. Therefore, this
-decision should be made thoughtfully. 
-In the presented application we applied a Gaussian distribution left-censored 
-at zero as this has proven to be a well working choice in
-earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
-\hlc{To test for robustness against distributional misspecifation 
-different distributional assumptions have been considered as well. 
-In Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a} \hlc{the exact
-same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
-observation stations, once employing a Gaussian hurdle model and once
-with a logistic distribution left-censored at zero in order to account
-for heavier tails. Under both assumptions the obtained results 
-are similar to those presented in the paper applying a Gaussian distribution
-left-censored at zero. In some settings the hurdle model leads to slight 
-improvements, but this comes with more variability (especially for EMOS and the 
-boosted GAMLSS). For the boosted GAMLSS the left-censored logistic distribution
-shows slightly better results regarding CRPS skill score, however, only within 
-a range of less than $2\%$ and not for all stations. 
-Altogether, the distributional forest has proven to be 
-very robust against distributional misspecifications.}
-
-\hlc{Alternatively, a distribution including an additional parameter capturing 
-skewness could be applied} \citep{Scheuerer+Hamill:2015, Baran+Nemoda:2016}. 
-\hlc{For example, in transformation forests} \citep{Hothorn+Zeileis:2017} 
-\hlc{a very flexible parametric distribution characterized by the transformation 
-function is employed such that higher moments can be modeled implicitly.
-Evaluating further distributions might be of interest but this would go beyond the
-scope of this manuscript.}
+
+\hlc{Choosing an adequate distributional family is an important step for
+establishing a well-fitting model. A zero-censored Gaussian distribution is
+employed in this manuscript as this has been found to be an appropriate choice
+for precipitation modeling in earlier literature} \citep[e.g.,][]{Stauffer+Mayr+Messner:2017}. 
+\hlc{To test for robustness against distributional misspecifation, 
+two alternative distributional specifications have been considered
+in Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a}: \hlc{Using the
+same evaluations as in} Section~\ref{sec:all}, \hlc{all models are
+additionally fitted for 15~meteorological stations using a zero-censored
+logistic distribution in order to account for heavier tails and a two-part
+Gaussian hurdle model combining a binary model for zero vs.~positive
+precipitation and a separate Gaussian model, truncated at zero, for the
+positive precipitation observations. Both specifications yield qualitatively 
+similar results as for the zero-censored Gaussian distribution. For some statons
+the two-part hurdle model leads to small improvements, however at the expense
+of increased variability across stations (especially for EMOS and the 
+boosted GAMLSS). Overall, the results from this manuscript are quite
+robust across these distributional specifications, especially for the
+distributional forests.}
+
+\hlc{Moreover, one could consider a distribution including an additional parameter
+for capturing skewness} \citep[as in][]{Scheuerer+Hamill:2015, Baran+Nemoda:2016}. 
+\hlc{However, this would go beyond the mean/variance specification of the NGR
+that is widely used in ensemble post-processing. Therefore, this contribution
+investigates the effects of using the same distributional family with a novel
+strategy for specifying dependence on covariates. Considering more flexible
+parametric specifications as well as transformation model specifications that
+allow to capture dependence of higher moments} \citep[see e.g.,][]{Hothorn+Zeileis:2017}
+\hlc{are extensions of interest for future research.}
+
+\subsubsection*{Axams vs.~other meteorological stations}
+
+\hlc{Axams was chosen as the meteorological station for the more extensive evaluations
+in} Section~\ref{sec:axams} \hlc{as it yields fairly typical results
+and is geographically in the center of the study area and closest to
+Innsbruck, the capital of Tyrol and the work place of three of the authors.
+To show that qualitatively similar results are obtained for other
+meteorological stations, Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b}
+\hlc{carries out the same evaluation for 14~further stations. These
+cover a wide range of geographical locations/altitudes and a mix of
+different best-performing models in the single-split setting reported in}
+Section~\ref{sec:all}.
+
+\hlc{The supplement shows that some of the differences in forecast skill from}
+Figure~\ref{fig:all-map} \hlc{even out in the cross-validation with distributional
+forests typically performing at least as well as the best of the other models at most
+stations. In particular, this also includes three stations in East Tyrol
+where the prespecified GAMLSS performs best in the single-split setting
+(learning based on 1985--2008 and validation for 2009--2012). While part of
+the performance difference can also be attributed to different
+climatological characteristics in East Tyrol (that, unlike North Tyrol, is located
+south of the main Alpine Ridge), the differences are greatly reduced for several
+stations in East Tyrol by considering more splits in the full cross-validation.}
 
 \subsubsection*{Tuning parameters}
-\hlc{Setting tuning parameters of a forest model can have a major influence
-not only on the accuracy of the model but also on the computational 
-complexity. 
-For the application in} Section~\ref{sec:precipitation} \hlc{we have 
-selected the values of these parameters} (listed in Section~\ref{sec:evaluation}) 
-based on advice from literature and our own experience.
-
-\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hlc{recommend to 
-build full-grown trees. Therefore, low values should be chosen for the stopping 
-criteria} \code{minsplit} and \code{minbucket} \hlc{together with a high significance 
-level} \code{alpha}.
-\hlc{However, in a distributional model it also has to be considered that a 
-minimal number of observations is required to allow for a reasonable 
-estimation of parameters in each node. For that reason, the minimal number
-of observations for which statistical tests are performed in the 
-distributional forest in} Section~\ref{sec:precipitation} is set to
-50 (\code{minbucket}) \hlc{and the minimal number of observations on which
-parameters are estimated is set to 20} (\code{minsplit}).
-
-As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
-\hlc{random forests do not overfit, not even for an increasing number of 
-trees} (\code{ntree}). \hlc{Hence, it is recommended to build forests consisting 
-of a large number of trees as this generally leads to an increase 
-in accuracy. However, it also comes along with increasing computational costs. 
-Therefore, ``a good choice results from a trade-off between computational 
-complexity and accuracy''} (\citealp{Biau+Scornet:2016}, p.~205).
-\hlc{Following this advice we chose to build forests consisting of 100 trees which
-is a commonly used value.}
-
-\subsubsection*{East Tyrol}
-Figure~\ref{fig:all-map} \hlc{shows that, contrary to the rest of Tyrol, 
-for the majority of stations in East Tyrol the prespecified GAMLSS performs
-best in this particular setting (learning data: 1985--2008, testing data:
-2009--2012). East Tyrol is located south of the main Alpine Ridge and 
-exhibits different climatological characteristics than the part north of 
-the Alpine Ridge. It seems that the choice of covariates and their interactions 
-we made for the prespecified GAMLSS method is particularly adequate for the 
-region of East Tyrol for this specific analysis. However, the advantages of 
-the prespecified GAMLSS models in this particular setting (training: 1985--2008, 
-validation: 2009--2012) even out in a cross-validation framework for most of 
-the stations in East Tyrol. These results are presented in Supplement B} 
-\citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{where the exact same evaluations as in} 
-Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
-for 14 other stations.}
-
-\subsubsection*{Choice of station Axams}
-%% choice of station Axams mit Verweis auf Supplement stationwise
-\hlc{Axams was chosen for the more extensive evaluations presented
-in this paper as it is the observation station being closest 
-to Innsbruck, the capital of Tyrol and the work place of 
-three of the authors, but not based on any other particular 
-features. Furthermore, it has been checked, whether it 
-is an adequate representative of all stations regarding 
-the results of our evaluations. This is also confirmed by 
-the extended analysis in Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b}
-\hlc{where the 15 considered stations also include some of those where 
-the forest does not perform better than the other models in the setting 
-presented in} Figure~\ref{fig:all-map}. \hlc{It shows that the differences in 
-performance even out in the cross-validation framework where the forest 
-performs at least equally well compared to the best of the other models 
-at most stations.}
+
+\hlc{Selecting tuning parameters for flexible regression models is important
+not only in terms of predictive accuracy but also computational complexity. 
+For the application in} Section~\ref{sec:precipitation} \hlc{tuning parameters
+are selected based on advice from the literature as well as our own experiences.}
+As \cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hlc{recommend to 
+build full-grown trees, early stopping upon non-significance is disabled
+(\texttt{alpha} $= 1$) and low values are used for \texttt{minsplit} ($ = 50$)
+and \texttt{minbucket} ($ = 20$), while assuring that \texttt{minsplit}
+is sufficiently large for reasonably obtaining MLEs of all parameters in each
+segment of the tree.}
+
+\hlc{Applying the Law of Large Numbers it can be shown that random forests do not overfit 
+as the number of trees increases}
+\citep{Breiman:2001,Hastie+Tibshirani+Friedman:2001,Biau+Scornet:2016}.
+\hlc{Therefore, in principle, forests can be built with a very large number of trees
+(\texttt{ntree}) as this cannot deteriorate the predictions. However, ``[\dots] the
+computational cost for inducing a forest increases linearly with the number of trees,
+so a good choice results from a trade-off between computational complexity and accuracy''}
+\citep[][p.~205]{Biau+Scornet:2016}.
+\hlc{Following this advice, we decided to build forests consisting of 100 trees.}
 
@@
 multiple distribution parameters.'' We have noticed computational problems when
-applying the function} \textbf{gamlss} \hlc{in certain settings within the cross-validation
-framework as it did not succeed in fitting the model. In these cases the method
-was not taken into consideration in the comparison of all applied models.}
+applying} \textbf{gamlss} \hlc{in certain settings within the cross-validation
+framework as it did not succeed in fitting the model. In these cases the prespecified
+GAMLSS was not taken into consideration in the comparison of all applied models.}
 
@@
 \code{demo("RainDistributions", package = "disttree")} and 
-\code{demo("RainStationwise", package = "disttree")} respectively.
+\code{demo("RainStationwise", package = "disttree")}, respectively.
 
@@
 s(\bm{\hat{\theta}}, y) = 
-%\begin{pmatrix} 
-%s(\bm{\hat{\theta}}, y_1) \\
-%s(\bm{\hat{\theta}}, y_2) \\
-%\vdots \\
-%s(\bm{\hat{\theta}}, y_n)
-%\end{pmatrix} =
 \begin{pmatrix} 
@@
 and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
+More formally, the following hypotheses are assessed with permutation tests:
 \begin{align}
-H_0^l:  s(\bm{\hat{\theta}}, Y) \qquad \bot \qquad Z_l 
+H_0^l:  s(\bm{\hat{\theta}}, Y) \qquad \bot \qquad Z_l.
 \end{align}
-
-These hypotheses are assessed with permutation tests based on
-the multivariate linear statistic
+%
+The permutation tests are based on the multivariate linear statistic
 \begin{equation}
@@
 If $Z_l$ is a categorical variable with $H$ categories then 
-$v_l(Z_{li}) = e_H(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$ 
+$v_l(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$ 
 such that $v_l$ is a $H$-dimensional unit vector where the element corresponding to 
@@
 significance level is fulfilled. This type of early-stopping in building
-a tree is sometimes also referred to as ``pre-pruning''. For forest models
+a tree is sometimes also referred to as ``pre-pruning''. For random forests
 pre-pruning is often switched off by setting the significance level to $1$.}
@@
 
-This permutation test based tree algorithm is presented in 
+This permutation-test-based tree algorithm is presented in 
 \cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
-A different framework to build a tree is provided by the MOB algorithm 
+A different framework to build a likelihood-based tree is provided by the MOB algorithm 
 which is based on M-fluctuation tests (\citealp{Zeileis+Hothorn+Hornik:2008}).
@@
 \subsection*{Supplement A: Different Response Distributions}
-(doi: \href{http://dx.doi.org/10.1214/00-AOASXXXXSUPPA}{10.1214/00-AOASXXXXSUPPA}).
-  %In the case study presented in the main manuscript ``Distributional Regression
-  %Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
-  %the novel distributional regression forests are based on a Gaussian response distribution,
-  %left-censored at zero to accomodate both the non-negativity of precipitation and
-  %the point mass at zero. 
   To assess the goodness of fit of the Gaussian distribution, left-censored at zero,
@@
   truncated at zero, for the positive precipitation observations.
-  %The latter could
-  %potentially better discriminate determinants of the occurrence of precipitation
-  %vs.\ the amount of precipitation given that it occurs.
 
 \subsection*{Supplement B: Stationwise Evaluation}
-(doi: \href{http://dx.doi.org/10.1214/00-AOASXXXXSUPPB}{10.1214/00-AOASXXXXSUPPB}).
-  %In the case study presented in the main manuscript ``Distributional Regression
-  %Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
-  %the novel distributional regression forests are evaluated and compared with
-  %other distributional regression approaches on 95 meteorological stations.
-  %Moreover, further details -- including a cross-validation analysis -- are
-  %preformed for only one station, Axams. 
   To show that Axams is a fairly typical

### 2018-11-02 pkg/disttree/inst/draft/rejoinder.Rnw
small touch-ups in abstract/intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2372 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4eb56ab7907fb1c328e898785cfe442533ec4c10
final Z revision of main manuscript, adapting much of the discussion further touch-ups in remaining materials
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2375 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a394dd57fb692b04399df40ba94b573455f36c1f

@@
 The concluding section of the main manuscript has been rewritten and
-extended to regarding discussion of: variable selection, tuning parameters,
+extended regarding discussion of: variable selection, tuning parameters,
 distributional specifications (referring to Supplement~A), and regional
@@
 \hspace*{-0.3cm} 
-\includegraphics[width = 0.23\textwidth]{signature-Schlosser}
+%\includegraphics[width = 0.23\textwidth]{signature-Schlosser}
 
@@
 as the number of trees increases (Breiman 2001, Hastie et al.\ 2001, 
-Biau et al.\ 2016). Therefore, in principle, forests can be built with a very
+Biau and Scornet 2016). Therefore, in principle, forests can be built with a very
 large number of trees as this cannot deteriorate the predictions.
@@
 with the number of trees, so a good choice results from a trade-off between 
-computational complexity and accuracy'' (Biau et al.\ 2016, p.~205). Following this 
+computational complexity and accuracy'' (Biau and Scornet 2016, p.~205). Following this 
 advice, we decided to build forests consisting of 100 trees. Increasing this number
@@
 \texttt{minbucket} to 20. The significance level \texttt{alpha} (for early stopping
-upon non-significance) is set to 1, thus disabling any early stopping.
+upon non-significance) is set to 1, thus disabling such early stopping.
 These tuning parameters are now also listed in the main manuscript in Sections~3.2.
@@
 obtaining the parameter vector that has been estimated in the 
-corresponding node when learning the tree. This is now also
-explained at the end of Section~2.2.
+corresponding node when learning the tree. This is now emphasized
+in the last paragraph of Section~2.2.
 
@@
 
-Overall, the covariates selected by the distributional forest do not provide too 
-much insights but show that the prespecified GAMLSS model contains an appropriate 
+Overall, the covariates selected by the distributional forest do not provide 
+new meteorological insights but show that the prespecified GAMLSS model contains an appropriate 
 subset of all available covariates.
@@
 rather as a complementing technique for the distributional regression
-toolbox. Much like random forests and GAMs are complements in
-the conditional mean regression toolbox.
+toolbox. Much like random forests and GAMs are complements for
+conditional mean regression.
 
@@
 
-We have adapted our manuscript in Section~1 and Section~4, \textit{Variable Selection}, 
-to reflect that we focus on forecasting skill in a large number of 
-models -- and not so much on interpretation.
+We have adapted our introduction (Section~1) and discussion (Section~4)
+to better reflect these aspects.
 
@@
 not even for a high number of trees. However, computational costs increase 
-linearly with the number of trees (Biau et al.\ 2016).
+linearly with the number of trees (Biau and Scornet, 2016).
 Therefore, we followed their recommendations of choosing a high number of
@@
 
+\item[] Baran, S., and Nemoda, D. (2016). Censored and Shifted Gamma Distribution Based EMOS
+Model for Probabilistic Quantitative Precipitation Forecasting,
+\textit{Environmetrics}, \textbf{27}: 280--292.
+%doi: 10.1002/env.2391.
+
+\item[] Bauer, P. and Thorpe, A. (2015). The Quiet Revolution of Numerical 
+Weather Prediction,
+\textit{Nature} \textbf{525}: 47--55.
+%doi:10.1038/nature14956
+
+\item[] Biau, G. and Scornet, E. (2016).
+A Random Forest Guided Tour,
+\textit{TEST}
+\textbf{25}: 197--227.
+
+\item[] Breiman, L. (2001).
+Random Forests,
+\textit{Machine Learning}
+\textbf{45, 1}: 5--32.
+
 \item[] Chen, J., Brissette, F. P. and Lhi, Z. (2014). Postprocessing 
@@
 
-%\item[] Goicoa, T., Adin, A., Ugarte, M. D. and Hodges, J. S. (2017). 
-%In Spatio-Temporal Disease Mapping Models, Identifiability Constraints 
-%Affet PQL and INLA Results, \textit{Stochastic Environmental Research 
-%and Risk Assessment} p. to appear.
-
-%\item[] Ugarte, M. D., Adin, A. and Goicoa, T. (2017). One-Dimensional, 
-%Two-Dimensional, and Three-Dimensional B-Splines to Specify Space-Time 
-%Interactions in Bayesian Disease Mapping: Model Fitting and Model 
-%Identifiability, \textit{Spatial Statistics} p. to appear.
+\item[] Dunn, P. K. and Smyth, G. K. (1996).
+Randomized Quantile Residuals,
+\textit{Journal of Computational and Graphical Statistics}
+\textbf{5, 3}: 236--244.
+
+\item[] Gneiting, T., Balabdaoui, F. and Raftery, A. E. (2007).
+Probabilistic Forecasts, Calibration and Sharpness,
+\textit{Journal of the Royal Statistical Society B}
+\textbf{69, 2}: 243--268.
+
+\item[] Gneiting, T., Raftery, A. E., Westveld III, A. H. and Goldman, T. (2005). 
+Calibrated Probabilistic Forecasting Using Ensemble Model Output Statistics and 
+Minimum CRPS Estimation,
+\textit{Monthly Weather Review} \textbf{133, 5}: 1098--1118.
 
@@
 
-\item[] Ugarte, M. D., Adin, A. and Goicoa, T. (2017). One-Dimensional, 
-Two-Dimensional, and Three-Dimensional B-Splines to Specify Space-Time 
-Interactions in Bayesian Disease Mapping: Model Fitting and Model 
-Identifiability, \textit{Spatial Statistics} \textbf{22}: 451--468.
-
-\item[] Wood, S. N., Scheipl, F. and Faraway, J. J. (2013). 
-Straightforward Intermediate Rank Tensor Product Smoothing in Mixed Models, 
-\textit{Statistics and Computing} \textbf{23}: 341--360.
-
-%%%%%%%%%
-% references used in answers
-\item[] Gneiting, T., Raftery, A. E., Westveld III, A. H. and Goldman, T. (2005). 
-Calibrated Probabilistic Forecasting Using Ensemble Model Output Statistics and 
-Minimum CRPS Estimation,
-\textit{Monthly Weather Review} \textbf{133, 5}: 1098--1118.
-
-\item[] Stauffer, R., Mayr, G. J., Messner, J. W., Umlauf, N. and Zeileis, A. (2017a). 
-Spatio-Temporal Precipitation Climatology over Complex Terrain Using a Censored Additive Regression Model,
-\textit{International Journal of Climatology} \textbf{37, 7}: 3264--3275.
-
-\item[] Stauffer, R., Umlauf, N., Messner, J. W., Mayr, G. J. and Zeileis, A. (2017b). 
-Ensemble Post-Processing of Daily Precipitation Sums over Complex Terrain Using Censored High-Resolution Standardized Anomalies,
-\textit{Monthly Weather Review} \textbf{45, 3}: 955--969.
-%doi = 10.1175/mwr-d-16-0260.1
+\item[] Hastie, T., Tibshirani, R. and Friedman, J. (2001).
+The Elements of Statistical Learning, %Second Edition,
+\textit{Springer New York Inc.}.
 
@@
 
-%\item[] Hofner, B., Boccuto, L. and Goeker, M. (2015). 
-%Controlling false discoveries in high-dimensional situations: Boosting 
-%with stability selection,
-%\textit{BMC Bioinformatics} \textbf{16:144}.
+\item[] Hutchinson M. F. (1998). Interpolation of Rainfall Data with Thin
+Plate Smoothing Splines -- Part II: Analysis of Topographic Dependence,
+\textit{Journal of Geographic Information and
+Decision Analysis} \textbf{2, 2}: 152--167.
+%doi:10.1080/02693799508902045
 
@@
 
-\item[] Dunn, P. K. and Smyth, G. K. (1996).
-Randomized Quantile Residuals,
-\textit{Journal of Computational and Graphical Statistics}
-\textbf{5, 3}: 236--244.
-
-\item[] Gneiting, T., Balabdaoui, F. and Raftery, A. E. (2007).
-Probabilistic Forecasts, Calibration and Sharpness,
-\textit{Journal of the Royal Statistical Society B}
-\textbf{69, 2}: 243--268.
-
-\item[] Biau, G. and Scornet, E. (2016).
-A Random Forest Guided Tour,
-\textit{TEST}
-\textbf{25}: 197--227.
-
-\item[] Hastie, T., Tibshirani, R. and Friedman, J. (2001).
-The Elements of Statistical Learning, %Second Edition,
-\textit{Springer New York Inc.}.
-
-\item[] Breiman, L. (2001).
-Random Forests,
-\textit{Machine Learning}
-\textbf{45, 1}: 5--32.
-
-\item[] Wahl, S. (2015). Uncertainty in Mesoscale Numerical Weather Prediction:
-Probabilistic Forecasting of Precipitation, PhD thesis, 
-\textit{University of Bonn}, p. 1--120.% \url{http://hss.ulb.uni-bonn.de/2015/4190/4190.pdf}.
-
-\item[] Rasp, S., Lerch, S. (2018). Neural Networks for Post-Processing Ensemble
-Weather Forecasts, \textit{Montly Weather Review}, NA.
-%doi:\href{https://doi.org/10.1175/MWR-D-18-0187.1}{10.1175/MWR-D-18-0187.1}.
-
 \item[] M{\"u}ller, M. D. (2011): Effects of Model Resolution and Statistical
@@
 \textit{Journal of Applied Meteorology and Climatology} 
-\textbf{50 8}: 1627--1636.
+\textbf{50, 8}: 1627--1636.
 %doi:10.1175/2011JAMC2615.1
 
-\item[] Bauer, P. and Thorpe, A. (2015). The Quiet Revolution of Numerical 
-Weather Prediction,
-\textit{Nature} \textbf{525}: 47--55.
-%doi:10.1038/nature14956
+\item[] Rasp, S., Lerch, S. (2018). Neural Networks for Post-Processing Ensemble
+Weather Forecasts, \textit{Montly Weather Review}, \textbf{146, 11}:, 3885--3900.
+%doi:\href{https://doi.org/10.1175/MWR-D-18-0187.1}{10.1175/MWR-D-18-0187.1}.
 
@@
 
-\item[] Baran, S., and Nemoda, D. (2016). Censored and Shifted Gamma Distribution Based EMOS
-Model for Probabilistic Quantitative Precipitation Forecasting,
-\textit{Environmetrics}, \textbf{27}: 280--292.
-%doi: 10.1002/env.2391.
-
-\item[] Hutchinson M. F. (1998). Interpolation of Rainfall Data with Thin
-Plate Smoothing Splines -- Part II: Analysis of Topographic Dependence,
-\textit{Journal of Geographic Information and
-Decision Analysis} \textbf{2, 2}: 152--167.
-%doi:10.1080/02693799508902045
+\item[] Stauffer, R., Mayr, G. J., Messner, J. W., Umlauf, N. and Zeileis, A. (2017a). 
+Spatio-Temporal Precipitation Climatology over Complex Terrain Using a Censored Additive Regression Model,
+\textit{International Journal of Climatology} \textbf{37, 7}: 3264--3275.
+
+\item[] Stauffer, R., Umlauf, N., Messner, J. W., Mayr, G. J. and Zeileis, A. (2017b). 
+Ensemble Post-Processing of Daily Precipitation Sums over Complex Terrain Using Censored High-Resolution Standardized Anomalies,
+\textit{Monthly Weather Review} \textbf{45, 3}: 955--969.
+%doi = 10.1175/mwr-d-16-0260.1
 
@@
 %doi:10.1029/wr009i005p01235
+
+\item[] Ugarte, M. D., Adin, A. and Goicoa, T. (2017). One-Dimensional, 
+Two-Dimensional, and Three-Dimensional B-Splines to Specify Space-Time 
+Interactions in Bayesian Disease Mapping: Model Fitting and Model 
+Identifiability, \textit{Spatial Statistics} \textbf{22}: 451--468.
+
+\item[] Wahl, S. (2015). Uncertainty in Mesoscale Numerical Weather Prediction:
+Probabilistic Forecasting of Precipitation, PhD thesis, 
+\textit{University of Bonn}, p. 1--120.% \url{http://hss.ulb.uni-bonn.de/2015/4190/4190.pdf}.
+
+\item[] Wood, S. N., Scheipl, F. and Faraway, J. J. (2013). 
+Straightforward Intermediate Rank Tensor Product Smoothing in Mixed Models, 
+\textit{Statistics and Computing} \textbf{23}: 341--360.
+
 \end{itemize}

### 2018-11-03 pkg/disttree/inst/draft/draft.Rnw
removed highlighting of changed/revised parts; typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2376 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 16004c6499fa53e2d5192a1c5cfc9f6db6a312de

@@
 \usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref}
+\usepackage[all]{hypcap}
 
@@
 regression approaches modeling all parameters including location, scale, and
-shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
+shape. Notably, so-called non-homogeneous Gaussian regression (NGR) models both
 mean and variance of a Gaussian response and is particularly
 popular in weather forecasting. 
-\hlc{Moreover, generalized additive models for location, scale, and shape (GAMLSS)
+Moreover, generalized additive models for location, scale, and shape (GAMLSS)
 provide a framework where each distribution parameter is modeled separately
-capturing smooth linear or nonlinear effects.}
+capturing smooth linear or nonlinear effects.
 %
@@
 %
-\hlc{However, distributional forests should not be considered as a replacement
+However, distributional forests should not be considered as a replacement
 of GAMLSS but rather as a complementing technique for flexible distributional
 regression -- much like GAMs and random forests are complements for conditional
-mean regression.}
+mean regression.
 
@@
 @
-\caption{\label{fig:axams-dist}\protect\hlc{Total precipitation predictions by a 
+\caption{\label{fig:axams-dist}Total precipitation predictions by a 
 distributional forest at station Axams for July~24 in 2009, 2010, 2011 and 
@@
 The observations are depicted by crosses and the predicted point mass
-from the model by filled circles.}}
+from the model by filled circles.}
 \end{figure}
@@
 \citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
-\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
-80~available regressors.
+\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection 
+among all 80~available regressors.
 
@@
 distributional models the goal is to partition the covariate space
-$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
+$\mathcal{Z}$ recursively into disjoint segments so that a homogeneous
 distributional model for the response $Y$ can be found for each segment with segment-specific
@@
 
-To find the segments $\mathcal{B}_b$ that are (approximately) homogenous
+To find the segments $\mathcal{B}_b$ that are (approximately) homogeneous
 with respect to the distributional model with given parameters, the idea
@@
 be ``sent down'' the tree and the corresponding segment-specific MLE has to be
-obtained. \hlc{Thus, in practice $\bm{\hat \theta}(\bold{z})$ does not have to be recalculated
-for each new $\bold{z}$ but one can simply extract the parameter estimates for
-the corresponding segment which have been computed already while
+obtained. Thus, in practice $\bm{\hat \theta}(\bold{z})$ does not have to be 
+recalculated for each new $\bold{z}$ but one can simply extract the parameter estimates 
+for the corresponding segment which have been computed already while
 learning the tree. However, to understand this estimator conceptually it is useful
-to denote it as a weighted MLE} where the weights
-select those observations from the learning sample that fall into the same
-segment:
+to denote it as a weighted MLE where the weights select those observations from 
+the learning sample that fall into the same segment:
 \begin{equation}
@@
 
-\hlc{The idea of random forests is to learn an ensemble of trees, each on a different
+The idea of random forests is to learn an ensemble of trees, each on a different
 learning data obtained through resampling (bootstrap or subsampling). In each node 
@@
 all trees is considered. In that way changes in the location  of the response 
-across the covariates are detected} (e.g., in Breiman and Cutler's random forests, 
+across the covariates are detected (e.g., in Breiman and Cutler's random forests, 
 \citealp{Breiman:2001}).
-\hlc{This idea is now taken one step further by embedding it in a distributional 
+This idea is now taken one step further by embedding it in a distributional 
 framework based on maximum-likelihood estimation. \emph{Distributional forests} 
@@
 the ``direction'' of any distribution parameter by considering the full likelihood
-and corresponding score function for choosing splitting variables and split points.}
+and corresponding score function for choosing splitting variables and split points.
 
@@
 numerical ensemble prediction systems (EPSs; see \citealp{Bauer+Thorpe+Brunet:2015}).
-\hlc{EPSs not only predict the most likely future state of the atmosphere but provide
+EPSs not only predict the most likely future state of the atmosphere but provide
 information about the uncertainty for a specific quantity and weather situation.
@@
 One frequently-used method based on distributional regression models is the 
-ensemble model output statistics (EMOS) approach first proposed by}
-\cite{Gneiting+Raftery+Westveld:2005} \hlc{to produce high-quality forecasts for
+ensemble model output statistics (EMOS) approach first proposed by
+\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
 specific quantities and sites.  
@@
 uncertainty of a specific forecast. Thus, a basic EMOS specification typically
-models the two distribution parameters by two linear predictors, e.g.,} 
+models the two distribution parameters by two linear predictors, e.g., 
 $\mu = \beta_0 + \beta_1 \cdot \mathrm{mean}(\emph{tp})$ and  
@@
 it typically does not perform as well in complex terrain due to unresolved
-effects in the \hlc{NWP system} \citep{Bauer+Thorpe+Brunet:2015}. 
+effects in the NWP system \citep{Bauer+Thorpe+Brunet:2015}. 
 For example, in the Tyrolean Alps -- considered
@@
 In the following, it is illustrated how distributional forests can
-solve this problem without requiring prior expert knowledge about the meteorological covariates.
+solve this problem without requiring prior expert knowledge about the 
+meteorological covariates.
 For fitting the forest only the response distribution and the list of potential predictor
@@
 \begin{minipage}{\textwidth}
-\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
+\caption[Table caption text]{Basic covariates together with the number ({\#}) 
+and the type of variations.
 Time periods indicate aggregation time periods in hours after NWP model
@@
   in Section~\ref{sec:methods}. 
-  \hlc{The stopping criteria are the minimal number of observations to perform a split 
+  The stopping criteria are the minimal number of observations to perform a split 
   (\texttt{minsplit} $= 50$), the minimal number of observations in a 
@@
   that no early stopping (or ``pre-pruning'') is applied
-  based on results of the statistical tests.}
+  based on results of the statistical tests.
   % stopping criteria:
@@
 
-\item \emph{EMOS:} \hlc{The basic} ensemble model output statistics
+\item \emph{EMOS:} The basic ensemble model output statistics
   models use the ensemble mean of total precipitation as regressor in the location
@@
 for one observation site, Axams in Tyrol (in the center of the study area,
-see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation observations and
-numerical weather predictions are available for the month of July from 1985
-through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
-forecasts from the distributional forest, trained on 1985--2008, for July 24
-in 2009--2012 have already been shown as a motivational example. The figure
-shows that the model properly
-depicts the point masses at zero (i.e., the probability of a dry day) and
-the forecasted probability density function for the total amount of precipitation.
-The four sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
+see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation 
+observations and numerical weather predictions are available for the month of July 
+from 1985 through 2012. In Figure~\ref{fig:axams-dist} in the introduction the 
+probabilistic forecasts from the distributional forest, trained on 1985--2008, 
+for July 24 in 2009--2012 have already been shown as a motivational example. 
+The figure shows that the model properly depicts the point masses at zero 
+(i.e., the probability of a dry day) and the forecasted probability density 
+function for the total amount of precipitation. The four sample forecasts differ 
+considerably in location $\mu$, scale $\sigma$, and the amount of
 censoring while conforming quite well with the actual observations
@@
 interested in the overall predictive performance and calibration of the
-distributional fits. \hlc{More details of this assessment as well as
+distributional fits. More details of this assessment as well as
 an application to 14~further meteorological stations is provided in
-Supplement~B} \citep{Schlosser+Hothorn+Stauffer:2018b}.
+Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b}.
 
@@
 diagonal), indicating that all four models are sufficiently well calibrated. 
-\hlc{This is also supported by the corresponding probability integral transform}
-(PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) \hlc{histograms for station
-Axams in Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{which 
-contains a more detailed explanation of residual QQ plots and PIT histograms.}
+This is also supported by the corresponding probability integral transform
+(PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) histograms for station
+Axams in Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} which 
+contains a more detailed explanation of residual QQ plots and PIT histograms.
 
@@
 splits the available 28~years into 7~subsets of 4~randomly-selected 
-\hlc{(and thus not necessarily consecutive)} years.
+(and thus not necessarily consecutive) years.
 The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
@@
 
-\hlc{Generally, there are many possibilities how to specify the variables that are
+Generally, there are many possibilities how to specify the variables that are
 to be included in a distributional regression model. Especially for a low number
@@
 yields models that often balance good predictive performance with high
-interpretability} (see for example \citealp{Wood+Scheipl+Faraway:2013}; 
+interpretability (see for example \citealp{Wood+Scheipl+Faraway:2013}; 
 \citealp{Goicoa+Adin+Ugarte:2018}; \citealp{Ugarte+Adin+Goicoa:2017}).
-\hlc{However, if the number of covariates is high, including all (or many)
+However, if the number of covariates is high, including all (or many)
 main effects and interactions in a GAMLSS typically becomes challenging both in terms
-of interpretability and computational complexity/stability}
+of interpretability and computational complexity/stability
 \citep[see also][]{Hofner+Mayr+Schmid:2016}.
 
-\hlc{In the precipitation forecasting application, as presented in} Section~\ref{sec:precipitation},
-\hlc{80~covariates are considered which corresponds to 3160~potential pairwise
-interactions (and even more higher-order interactions). 
+In the precipitation forecasting application, as presented in 
+Section~\ref{sec:precipitation}, 80~covariates are considered which corresponds 
+to 3160~potential pairwise interactions (and even more higher-order interactions). 
 Therefore, only main effects are considered for the boosted GAMLSS while
@@
 interpretability but forecasting skill and (semi-)automatic application on a
-larger domain} \citep[see also the discussion in][]{Rasp+Lerch:2018}.
+larger domain \citep[see also the discussion in][]{Rasp+Lerch:2018}.
 
@@
 
-\hlc{Choosing an adequate distributional family is an important step for
+Choosing an adequate distributional family is an important step for
 establishing a well-fitting model. A zero-censored Gaussian distribution is
 employed in this manuscript as this has been found to be an appropriate choice
-for precipitation modeling in earlier literature} \citep[e.g.,][]{Stauffer+Mayr+Messner:2017}. 
-\hlc{To test for robustness against distributional misspecifation, 
+for precipitation modeling in earlier literature
+\citep[e.g.,][]{Stauffer+Mayr+Messner:2017}. 
+To test for robustness against distributional misspecifation, 
 two alternative distributional specifications have been considered
-in Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a}: \hlc{Using the
-same evaluations as in} Section~\ref{sec:all}, \hlc{all models are
+in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a}: Using the
+same evaluations as in Section~\ref{sec:all}, all models are
 additionally fitted for 15~meteorological stations using a zero-censored
@@
 positive precipitation observations. Both specifications yield qualitatively 
-similar results as for the zero-censored Gaussian distribution. For some statons
+similar results as for the zero-censored Gaussian distribution. For some stations
 the two-part hurdle model leads to small improvements, however at the expense
@@
 robust across these distributional specifications, especially for the
-distributional forests.}
-
-\hlc{Moreover, one could consider a distribution including an additional parameter
-for capturing skewness} \citep[as in][]{Scheuerer+Hamill:2015, Baran+Nemoda:2016}. 
-\hlc{However, this would go beyond the mean/variance specification of the NGR
+distributional forests.
+
+Moreover, one could consider a distribution including an additional parameter
+for capturing skewness \citep[as in][]{Scheuerer+Hamill:2015, Baran+Nemoda:2016}. 
+However, this would go beyond the mean/variance specification of the NGR
 that is widely used in ensemble post-processing. Therefore, this contribution
@@
 parametric specifications as well as transformation model specifications that
-allow to capture dependence of higher moments} \citep[see e.g.,][]{Hothorn+Zeileis:2017}
-\hlc{are extensions of interest for future research.}
+allow to capture dependence of higher moments \citep[see e.g.,][]{Hothorn+Zeileis:2017}
+are extensions of interest for future research.
 
@@
 
-\hlc{Axams was chosen as the meteorological station for the more extensive evaluations
-in} Section~\ref{sec:axams} \hlc{as it yields fairly typical results
+Axams was chosen as the meteorological station for the more extensive evaluations
+in Section~\ref{sec:axams} as it yields fairly typical results
 and is geographically in the center of the study area and closest to
@@
 To show that qualitatively similar results are obtained for other
-meteorological stations, Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b}
-\hlc{carries out the same evaluation for 14~further stations. These
+meteorological stations, Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b}
+carries out the same evaluation for 14~further stations. These
 cover a wide range of geographical locations/altitudes and a mix of
-different best-performing models in the single-split setting reported in}
+different best-performing models in the single-split setting reported in
 Section~\ref{sec:all}.
 
-\hlc{The supplement shows that some of the differences in forecast skill from}
-Figure~\ref{fig:all-map} \hlc{even out in the cross-validation with distributional
+The supplement shows that some of the differences in forecast skill from
+Figure~\ref{fig:all-map} even out in the cross-validation with distributional
 forests typically performing at least as well as the best of the other models at most
@@
 south of the main Alpine Ridge), the differences are greatly reduced for several
-stations in East Tyrol by considering more splits in the full cross-validation.}
+stations in East Tyrol by considering more splits in the full cross-validation.
 
@@
 
-\hlc{Selecting tuning parameters for flexible regression models is important
+Selecting tuning parameters for flexible regression models is important
 not only in terms of predictive accuracy but also computational complexity. 
-For the application in} Section~\ref{sec:precipitation} \hlc{tuning parameters
-are selected based on advice from the literature as well as our own experiences.}
-As \cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} \hlc{recommend to 
+For the application in Section~\ref{sec:precipitation} tuning parameters
+are selected based on advice from the literature as well as our own experiences.
+As \cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} recommend to 
 build full-grown trees, early stopping upon non-significance is disabled
@@
 is sufficiently large for reasonably obtaining MLEs of all parameters in each
-segment of the tree.}
-
-\hlc{Applying the Law of Large Numbers it can be shown that random forests do not overfit 
-as the number of trees increases}
+segment of the tree.
+
+Applying the Law of Large Numbers it can be shown that random forests do not overfit 
+as the number of trees increases
 \citep{Breiman:2001,Hastie+Tibshirani+Friedman:2001,Biau+Scornet:2016}.
-\hlc{Therefore, in principle, forests can be built with a very large number of trees
+Therefore, in principle, forests can be built with a very large number of trees
 (\texttt{ntree}) as this cannot deteriorate the predictions. However, ``[\dots] the
 computational cost for inducing a forest increases linearly with the number of trees,
-so a good choice results from a trade-off between computational complexity and accuracy''}
+so a good choice results from a trade-off between computational complexity and accuracy''
 \citep[][p.~205]{Biau+Scornet:2016}.
-\hlc{Following this advice, we decided to build forests consisting of 100 trees.}
+Following this advice, we decided to build forests consisting of 100 trees.
 
@@
 %% GAMLSS errors with references gamboostLSS Hofner et al.
-As stated by \cite{Hofner+Mayr+Schmid:2016} \hlc{the AIC-based variable selection 
-methods implemented in the} \textsf{R} package \textbf{gamlss} \hlc{``[...] can be 
+As stated by \cite{Hofner+Mayr+Schmid:2016} the AIC-based variable selection 
+methods implemented in the \textsf{R} package \textbf{gamlss} ``[...] can be 
 unstable, especially when it comes to selecting possibly different sets of variables for 
 multiple distribution parameters.'' We have noticed computational problems when
-applying} \textbf{gamlss} \hlc{in certain settings within the cross-validation
+applying \textbf{gamlss} in certain settings within the cross-validation
 framework as it did not succeed in fitting the model. In these cases the prespecified
-GAMLSS was not taken into consideration in the comparison of all applied models.}
+GAMLSS was not taken into consideration in the comparison of all applied models.
 
@@
 \textsf{R} package \textbf{crch} 
-\citep[version~\Sexpr{packageVersion("crch")}]{Messner+Mayr+Zeileis:2016} for the EMOS models,
-\textbf{gamlss}
-\citep[version~\Sexpr{packageVersion("gamlss")}]{Stasinopoulos+Rigby:2007} for the prespecified GAMLSS, and
-\textbf{gamboostLSS}
+\citep[version~\Sexpr{packageVersion("crch")}]{Messner+Mayr+Zeileis:2016} for the EMOS 
+models, \textbf{gamlss}
+\citep[version~\Sexpr{packageVersion("gamlss")}]{Stasinopoulos+Rigby:2007} for the 
+prespecified GAMLSS, and \textbf{gamboostLSS}
 \citep[version~\Sexpr{packageVersion("gamboostLSS")}]{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
@@
 observation station Axams (including Figure~\ref{fig:axams-dist})
-is reproducible using 
-\code{demo("RainAxams", package = "disttree")}.
+is reproducible using \code{demo("RainAxams", package = "disttree")}.
 This also includes fitting the other zero-censored
@@
 \code{demo("RainTyrol", package = "disttree")}
-requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
+requiring the companion \textsf{R} package \textbf{RainTyrol} 
+(version~\Sexpr{packageVersion("RainTyrol")}),
 also available within the \textsf{R}-Forge~project.
-\hlc{The results presented in Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a} and 
-Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{can be reproduced using}
+The results presented in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a} and 
+Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
 \code{demo("RainDistributions", package = "disttree")} and 
@@
 \end{equation}
-\hlc{where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
+where $s(\bm{\hat{\theta}}, Y_i) \in \mathbb{R}^{1\times k}$ and the 
 type of the transformation function $v_l$ depends on the type of the split 
@@
 as the ``vec'' operator converts the $H \times k$ matrix into a $H \cdot k$ column vector 
-by column-wise combination.} Observations with missing values are excluded from the sums.
+by column-wise combination. Observations with missing values are excluded from the sums.
 
@@
 \cite{Strasser+Weber:1999} the test statistic can be standardized. 
-\hlc{The observed multivariate linear statistic $t_l$ which is either an
+The observed multivariate linear statistic $t_l$ which is either an
 $k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of $Z_l$,
 is mapped onto the real line by a univariate test statistic $c$.
-In the application of this paper a quadratic form is chosen, such that}
+In the application of this paper a quadratic form is chosen, such that
 \begin{equation}
@@
 \end{equation}
-\hlc{where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.} Alternatively,
+where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$. Alternatively,
 the maximum of the absolute values of the standardized linear statistic
@@
 If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
-is chosen as splitting variable. \hlc{Otherwise no further split is made 
+is chosen as splitting variable. Otherwise no further split is made 
 in this node as the stopping criterion of no $p$-values being below the 
@@
 a tree is sometimes also referred to as ``pre-pruning''. For random forests
-pre-pruning is often switched off by setting the significance level to $1$.}
+pre-pruning is often switched off by setting the significance level to $1$.
 
@@
 \end{equation}
-\hlc{for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
+for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
 two new subgroups, without any particular ordering, that are defined by 
 splitting in split point $r$ of variable $Z_{l^{\ast}}$.
-The split point is then chosen as follows:}
+The split point is then chosen as follows:
 \begin{equation}

### 2018-11-03 pkg/disttree/inst/draft/draft_AOAS.tex
draft in AOAS format
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2377 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4a0dc4a9a1926a0cd376361f2b456085878127c4

--- 2018-11-03 pkg/disttree/inst/draft/draft.Rnw
@@
-\documentclass[nojss]{jss}
-\shortcites{Hamill+Bates+Whitaker:2013}
+% Template for the submission to:
+%   The Annals of Probability           [aop]
+%   The Annals of Applied Probability   [aap]
+%   The Annals of Statistics            [aos] 
+%   The Annals of Applied Statistics    [aoas]
+%   Stochastic Systems                  [ssy]
+%
+%Author: In this template, the places where you need to add information
+%        (or delete line) are indicated by {???}.  Mostly the information
+%        required is obvious, but some explanations are given in lines starting
+%Author:
+%All other lines should be ignored.  After editing, there should be
+%no instances of ??? after this line.
+
+% use option [preprint] to remove info line at bottom
+% journal options: aop,aap,aos,aoas,ssy
+% natbib option: authoryear
+
+\documentclass[aoas, preprint]{imsart}
+%\shortcites{Hamill+Bates+Whitaker:2013}
 
@@
 %% packages
-\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref}
+\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}
+%\usepackage{amsthm,amsmath,natbib}
+\usepackage{array,makecell,tikz}
+\RequirePackage[OT1]{fontenc}
+\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
 \usepackage[all]{hypcap}
+\RequirePackage{natbib}
 
 %% tikz
-\usepackage{array,makecell,tikz,color,soul}
 \usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-
-%% need no \usepackage{Sweave}
-\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
-
-%% commands
+%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
+
+%% colors
+\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
+\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
+
+% provide arXiv number if available:
+\arxiv{arXiv:1804.02921}
+
+% put your definitions there:
+\startlocaldefs
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
 \newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\renewcommand{\Prob}{\mathbb{P} }
-\renewcommand{\E}{\mathbb{E}}
-\newcommand{\V}{\mathbb{V}}
-\newcommand{\Var}{\mathbb{V}}
-\newcommand{\R}{\mathbb{R} }
-\newcommand{\N}{\mathbb{N} }
-\newcommand{\C}{\mathbb{C} }
-\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
-\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
-\newcommand{\LS}{\mathcal{L}_n}
-\newcommand{\TS}{\mathcal{T}_n}
-\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
-\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
-\newcommand{\F}{\mathcal{F}}
-\newcommand{\A}{\mathcal{A}}
-\newcommand{\yn}{y_{\text{new}}}
-\newcommand{\z}{\mathbf{z}}
-\newcommand{\X}{\mathbf{X}}
-\newcommand{\Y}{\mathbf{Y}}
-\newcommand{\sX}{\mathcal{X}}
-\newcommand{\sY}{\mathcal{Y}}
-\newcommand{\T}{\mathbf{T}}
-\newcommand{\x}{\mathbf{x}}
-\renewcommand{\a}{\mathbf{a}}
-\newcommand{\xn}{\mathbf{x}_{\text{new}}}
-\newcommand{\y}{\mathbf{y}}
-\newcommand{\w}{\mathbf{w}}
-\newcommand{\ws}{\mathbf{w}_\cdot}
-\renewcommand{\t}{\mathbf{t}}
-\newcommand{\M}{\mathbf{M}}
-\renewcommand{\vec}{\text{vec}}
-\newcommand{\B}{\mathbf{B}}
-\newcommand{\K}{\mathbf{K}}
-\newcommand{\W}{\mathbf{W}}
-\newcommand{\D}{\mathbf{D}}
 \newcommand{\I}{\mathbf{I}}
-\newcommand{\bS}{\mathbf{S}}
-\newcommand{\cellx}{\pi_n[\x]}
-\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
-\newcommand{\err}{\text{Err}}
-\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
-\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
-\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
-\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
-\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
-\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
-\newcommand{\bft}{\mathbf{t}}
-
-\definecolor{changes}{rgb}{1,1,0.7}
-\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}
-
+\endlocaldefs
 \numberwithin{equation}{subsection}
 
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-library("disttree")
-library("crch")
-library("countreg")
-library("gamlss")
-library("gamlss.cens")
-gen.cens(NO, type = "left")
-library("RainTyrol")
-library("scoringRules")
-
-## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-# set function for parallelization
-applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))
-
-@
-
+\begin{document}
+
+\begin{frontmatter}
+
+% "Title of the paper"
 \title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
-\Shorttitle{Distributional Regression Forests for Probabilistic Precipitation Forecasting}
-
-\author{Lisa Schlosser\\Universit\"at Innsbruck
-   \And Torsten Hothorn\\Universit\"at Z\"urich
-   \And Reto Stauffer\\Universit\"at Innsbruck
-   \And Achim Zeileis\\Universit\"at Innsbruck}
-\Plainauthor{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
-
-\Abstract{
+%\runtitle{Distributional Regression Forests for Probabilistic Precipitation Forecasting}
+\runtitle{Distributional Regression Forests}
+
+% indicate corresponding author with \corref{}
+% \author{\fnms{John} \snm{Smith}\corref{}\ead[label=e1]{smith@foo.com}\thanksref{t1}}
+% \thankstext{t1}{Thanks to somebody} 
+% \address{line 1\\ line 2\\ printead{e1}}
+% \affiliation{Some University}
+
+\begin{aug}
+\author{\fnms{Lisa} \snm{Schlosser}\thanksref{m1}\corref{}\ead[label=e1]{Lisa.Schlosser@uibk.ac.at}
+\ead[label=u1,url]{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/}},
+\author{\fnms{Torsten} \snm{Hothorn}\thanksref{m2}\ead[label=e2]{Torsten.Hothorn@R-project.org}
+\ead[label=u2,url]{http://user.math.uzh.ch/hothorn/}},
+\author{\fnms{Reto} \snm{Stauffer}\thanksref{m1}\ead[label=e3]{Reto.Stauffer@uibk.ac.at}
+\ead[label=u3,url]{https://retostauffer.org/}}
+\and
+\author{\fnms{Achim} \snm{Zeileis}\thanksref{m1}\ead[label=e4]{Achim.Zeileis@R-project.org}
+\ead[label=u4,url]{https://eeecon.uibk.ac.at/\~{}zeileis/}}
+
+\runauthor{L. Schlosser, T. Hothorn, R. Stauffer, A. Zeileis}
+
+\affiliation{Universit\"at Innsbruck\thanksmark{m1} and Universit\"at Z\"urich\thanksmark{m2}}
+
+\address{Universit\"at Innsbruck \\
+  Department of Statistics \\
+  Faculty of Economics and Statistics \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
+\printead{e1}\\
+\phantom{E-mail:\ }\printead*{e3}\\
+\phantom{E-mail:\ }\printead*{e4}}
+%\printead{u1}\\
+%\phantom{URL:\ }\printead*{u3}\\
+%\phantom{URL:\ }\printead*{u4}}
+
+\address{Universit\"at Z\"urich \\
+Institut f\"ur Epidemiologie,\\
+Biostatistik und Pr\"avention \\
+Hirschengraben 84\\
+CH-8001 Z\"urich, Switzerland \\
+\printead{e2}}
+%\printead{u2}}
+\end{aug}
+
+\begin{abstract}
+
 To obtain a probabilistic model for a dependent variable based on some set of
@@
 computationally more demanding boosting approach.
-}
-\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
-
-\Address{
-  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
-  Universit\"at Innsbruck \\
-  Department of Statistics \\
-  Faculty of Economics and Statistics \\
-  Universit\"atsstr.~15 \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
-  \email{Achim.Zeileis@R-project.org} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
-  \phantom{URL: }\url{https://retostauffer.org/},\\
-  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
-
-  Torsten Hothorn\\
-  Universit\"at Z\"urich \\
-  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
-  Hirschengraben 84\\
-  CH-8001 Z\"urich, Switzerland \\
-  E-mail: \email{Torsten.Hothorn@R-project.org}\\
-  URL: \url{http://user.math.uzh.ch/hothorn/}\\
-
-}
-
-\begin{document}
+\end{abstract}
+
+%parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS
+
+%\begin{keyword}[class=MSC]
+%\kwd[Primary ]{60K35}
+%\kwd{60K35}
+%\kwd[; secondary ]{60K35}
+%\end{keyword}
+
+\begin{keyword}
+\kwd{parametric models}
+\kwd{regression trees}
+\kwd{random forests}
+\kwd{recursive partitioning}
+\kwd{probabilistic forecasting}
+\kwd{GAMLSS}
+\end{keyword}
+
+\end{frontmatter}
 
@@
 
-\pagebreak
+%\pagebreak
 
@@
 \centering
-<<motivation_GLM, echo=FALSE, results=hide>>=
-nobs <- 200
-## GLM
-set.seed(7)
-x <- c(1:nobs)/nobs
-ytrue <- 1+x*1.5
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-
-<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal[1], lwd = 7, main = "")
-@
+\setkeys{Gin}{width=0.9\textwidth}
+\includegraphics{draft-plot_motivation_GLM}
 \endminipage};
-\node (b) at (5,0) 
+\node (b) at (4.5,0) 
 {
@@
 \centering
-<<motivation_GAM, echo=FALSE, results=hide>>=
-## GAM
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3 
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-
-<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal[1], lwd = 7, main = "")
-@
+\setkeys{Gin}{width=0.9\textwidth}
+\includegraphics{draft-plot_motivation_GAM}
 \endminipage};
-\node (c) at (10,0) 
+\node (c) at (9,0) 
 {
@@
 \centering
-<<motivation_GAMLSS, echo=FALSE, results=hide>>=
-## GAMLSS
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3
-var <- exp(-(2*x)^2)/2
-y <- ytrue + rnorm(nobs, 0, 0.1 + var)
-@
-
-<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
-polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
-  col = pallight[1], border = "transparent")
-lines(x, ytrue, col = pal[1], lwd=7)
-points(x, y, col = "slategray", pch = 19)
-box(lwd = 5)
-@
+\setkeys{Gin}{width=0.9\textwidth}
+\includegraphics{draft-plot_motivation_GAMLSS}
 \endminipage};
@@
 \endminipage};
-\node (e) at (5,-2) 
+\node (e) at (4.5,-2) 
 {
@@
 \endminipage};
-\node at (10,-2)
+\node at (9,-2)
 {
@@
 \endminipage};
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
+\draw[-{>[scale=3, length=2, width=3]},line width=0.4pt](1.85,0)--(2.65,0);
+\draw[-{>[scale=3, length=2, width=3]},line width=0.4pt](6.35,0)--(7.15,0);
 \end{tikzpicture}
@@
 \centering
-<<motivation_regtree, echo=FALSE, results=hide>>=
-## Reg. Tree
-set.seed(7)
-kappa <- 12
-x <- c(1:nobs)/nobs
-ytrue <- ytree <- yforest <- numeric(length = length(x))
-for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
-y <- ytrue + rnorm(nobs,0,0.3)
-for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
-@
-
-<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
-lines(x = x, y = ytree, col = pal[1], lwd=7)
-@
+\setkeys{Gin}{width=0.9\textwidth}
+\includegraphics{draft-plot_motivation_regtree}
 \endminipage};
-\node (b) at (5,0) 
+\node (b) at (4.5,0) 
 {
@@
 \centering
-<<motivation_randforest, echo=FALSE, results=hide>>=
-## Random Forest
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-@
-
-<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
-lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
-@
+\setkeys{Gin}{width=0.9\textwidth}
+\includegraphics{draft-plot_motivation_randforest}
 \endminipage};
-\node (c) at (10,0) 
+\node (c) at (9,0) 
 {
@@
 \centering
-<<motivation_distforest, echo=FALSE, results=hide>>=
-## distforest
-set.seed(4723)
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-xt <- 2*(x-0.5)
-var <- exp(-(2*xt)^4)/2
-y <- ytrue + rnorm(nobs, 0, 0.1 + var)
-x <- c(1:nobs)/nobs
-@
-
-<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
-par(mar = c(2, 0, 2, 0))
-plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
-usr <- par("usr")
-rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
-polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
-  col = pallight[1], border = "transparent")
-lines(x = x, y = yforest, col = pal[1], lwd=7, main = "")
-points(x, y, col = "slategray", pch = 19)
-box(lwd=8)
-@
+\setkeys{Gin}{width=0.9\textwidth}
+\includegraphics{draft-plot_motivation_distforest}
 \endminipage};
@@
 \endminipage};
-\node (g) at (0,-5.4) 
+\node (g) at (0,-4.78) 
 {
@@
 \centering
-<<motivation_disttree, echo=FALSE, results=hide>>=
-## disttree
-set.seed(74)
-x <- c(1:nobs)/nobs
-var <- ytree/3
-y <- ytrue + rnorm(nobs,0,var)
-@
-
-<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
-par(mar = c(2, 0, 2, 0))
-plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
-usr <- par("usr")
-rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
-lines(x = x, y = ytree, col = pal[1], lwd=7, main = "")
-polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
-  col = pallight[1], border = "transparent")
-points(x, y, col = "slategray", pch = 19)
-box(lwd=8)
-@
+\setkeys{Gin}{width=0.9\textwidth}
+\includegraphics{draft-plot_motivation_disttree}
 \vspace{-0.2cm}%\\
@@
 \endminipage};
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](0,-2.4)--(0,-3.2);
-\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.2,-5.4)--(7.9,-2.5);
+\draw[-{>[scale=3, length=2, width=3]},line width=0.3pt](1.85,0)--(2.65,0);
+\draw[-{>[scale=3, length=2, width=3]},line width=0.3pt](6.35,0)--(7.15,0);
+\draw[-{>[scale=3, length=2, width=3]},line width=0.3pt](0,-2.35)--(0,-3);
+\draw[-{>[scale=3, length=2, width=3]},line width=0.3pt](2.2,-4.78)--(7.15,-2.5);
 \end{tikzpicture}
@@
 \begin{itemize}
-  \item \emph{Regression trees} (\citealp{Breiman+Friedman+Stone:1984}) recursively
+  \item \emph{Regression trees} \citep{Breiman+Friedman+Stone:1984} recursively
     split the data into more homogeneous subgroups and can thus capture abrupt shifts
@@
     their interactions. 
-  \item \emph{Random forests} (\citealp{Breiman:2001}) average the predictions of
+  \item \emph{Random forests} \citep{Breiman:2001} average the predictions of
     an ensemble of trees fitted to resampled versions of the learning data. This
@@
 
-<<echo=FALSE, results=hide>>=
-#### Axams prediction 24 - 4
-if(file.exists("Axams_24to4.rda")){
-  load("Axams_24to4.rda")
-} else {
-  source("Axams_24to4.R")
-  Axams_24to4 <- Axams_24to4()
-  save(Axams_24to4, file = "Axams_24to4.rda")
-}
-
-#### prepare data for plot of estimated density functions
-# predictions for one day (in each of the four years) 
-# (19th of July 2011 is missing)
-pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
-
-pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
-pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata[pdays,], type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, Axams_24to4$testdata[pdays,"robs"])
-
-# plot predicted distributions together with observations
-set.seed(7)
-x <- c(0.01, sort(runif(500,0.01,8)))
-y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
-y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
-y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
-y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
-dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
-
-# switch x-axis back to untransformed scale
-x <- x^(1.6)
-
-# point mass (slightly shifted)
-pm1 <- c(0.15, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.05
-pm2 <- c(0.05, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.01
-pm3 <- c(-0.05, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.03
-pm4 <- c(-0.15, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.07
-
-# observations
-obs1 <- c(Axams_24to4$testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-obs2 <- c(Axams_24to4$testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-obs3 <- c(Axams_24to4$testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-obs4 <- c(Axams_24to4$testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(Axams_24to4$testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
-
-#legendheight
-lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
-lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
-lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
-
-############
-# get predicted parameter of all models for testdata
-
-# distributional tree
-pdt <- predict(Axams_24to4$dt, newdata = Axams_24to4$testdata, type = "parameter")
-dt_mu <- pdt$mu
-dt_sigma <- pdt$sigma
-dt_exp <- pnorm(dt_mu/dt_sigma) * (dt_mu + dt_sigma * (dnorm(dt_mu/dt_sigma) / pnorm(dt_mu/dt_sigma)))
-
-# distributional forest
-pdf <- predict(Axams_24to4$df, newdata = Axams_24to4$testdata, type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-
-# prespecified GAM
-g_mu <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "mu", type = "response", data = Axams_24to4$g_learndata)
-g_sigma <- predict(Axams_24to4$g, newdata = Axams_24to4$testdata, what = "sigma", type = "response", data = Axams_24to4$g_learndata)
-g_exp <- pnorm(g_mu/g_sigma) * (g_mu + g_sigma * (dnorm(g_mu/g_sigma) / pnorm(g_mu/g_sigma)))
-
-# boosted GAM
-pgb <- predict(Axams_24to4$gb, newdata = Axams_24to4$testdata, parameter = c("mu","sigma"), type = "response")
-gb_mu <- pgb$mu
-gb_sigma <- pgb$sigma
-gb_exp <- pnorm(gb_mu/gb_sigma) * (gb_mu + gb_sigma * (dnorm(gb_mu/gb_sigma) / pnorm(gb_mu/gb_sigma)))
-
-# EMOS
-ml_mu <- predict(Axams_24to4$ml, type = "location", newdata = Axams_24to4$testdata)
-ml_sigma <- predict(Axams_24to4$ml, type = "scale", newdata = Axams_24to4$testdata)
-ml_exp <- pnorm(ml_mu/ml_sigma) * (ml_mu + ml_sigma * (dnorm(ml_mu/ml_sigma) / pnorm(ml_mu/ml_sigma)))
-
-############
-## PIT histograms and qqrplots
-
-## get PIT
-{
-  set.seed(7)
-
-  # distributional tree
-  pit_dt <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = dt_mu, sd = dt_sigma))
-  pit_dt[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_dt[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # distributional forest
-  pit_df <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = df_mu, sd = df_sigma))
-  pit_df[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_df[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # prespecified GAM
-  pit_g <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = g_mu, sd = g_sigma))
-  pit_g[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_g[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # boosted GAM
-  pit_gb <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = gb_mu, sd = gb_sigma))
-  pit_gb[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_gb[Axams_24to4$testdata[,"robs"]>0, 2]
-
-  # EMOS
-  pit_ml <- cbind(0, pnorm(Axams_24to4$testdata[,"robs"], mean = ml_mu, sd = ml_sigma))
-  pit_ml[Axams_24to4$testdata[,"robs"]>0, 1] <- pit_ml[Axams_24to4$testdata[,"robs"]>0, 2]
-
-}
-
-## Variable importance
-if(file.exists("vimp_crps.rda")){
-  load("vimp_crps.rda")
-} else {
-
-  set.seed(7)
-  nperm <- 50
-
-  # function to permutate chosen variable and then calculate mean crps
-  meancrps <- function(permute = NULL, newdata = Axams_24to4$testdata) {
-    if(!is.null(permute)) newdata[[permute]] <- sample(newdata[[permute]])
-    p <- predict(Axams_24to4$df, newdata = newdata, type = "parameter")
-    mean(crps_cnorm(newdata$robs, location = p$mu, scale = p$sigma, lower = 0))
-  }
-
-  # apply for all covariates except for dswrf_mean_min and 
-  # dswrf_sprd_min (columns 30 and 33) as they are always 0
-
-  # risk <- sapply(c(5:29, 31, 32, 34: ncol(testdata)), meancrps)
-
-  # using only one core
-  # risk_all <- replicate(nperm, sapply(c(5:29, 31, 32, 34: ncol(testdata)), meancrps))
-  # risk <- rowMeans(risk_all)
-
-  # or parallel
-  risklist <- applyfun(1:nperm, 
-                       function(i){
-                         set.seed(i)
-                         sapply(c(5:29, 31, 32, 34: ncol(Axams_24to4$testdata)), meancrps)
-                       })
-  risk <- Reduce("+", risklist) / length(risklist)
-
-  names(risk) <- names(Axams_24to4$testdata)[c(5:29, 31, 32, 34: ncol(Axams_24to4$testdata))]
-  vimp_crps <- risk - meancrps(newdata = Axams_24to4$testdata)
-  vimp_crps <- sort(vimp_crps, decreasing = TRUE)
-
-  save(vimp_crps, file = "vimp_crps.rda")
-}
-
-@
-
 \begin{figure}[t!]
 \centering
-\setkeys{Gin}{width=0.75\textwidth}
-<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=3.8>>=
-par(mar = c(3.8, 4, 2.3, 1.5))
-plot(x = x, y = y1, type = "l", col = pal[1], lwd = 1.3, 
-     main = paste0("July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
-     # xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
-     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-2,28))     #  xlim = c(-1.5,8))
-
-lines(x = x, y = y2, type = "l", col = pal[3], lwd = 1.3)
-lines(x = x, y = y3, type = "l", col = pal[2], lwd = 1.3)
-lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
-legend("topright", c("Predicted distribution", "Predicted point mass at 0", "Observation"),
-       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)
-
-# plot point mass
-lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = pal[1], type = "l", lwd = 1.3)
-lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = pal[3], type = "l", lwd = 1.3)
-lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = pal[2], type = "l", lwd = 1.3)
-lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = pal[4], type = "l", lwd = 1.3)
-
-points(x = pm1[1], y = pm1[2], col = pal[1], pch = 19)
-points(x = pm2[1], y = pm2[2], col = pal[3], pch = 19)
-points(x = pm3[1], y = pm3[2], col = pal[2], pch = 19)
-points(x = pm4[1], y = pm4[2], col = pal[4], pch = 19)
-
-# plot observations
-points(x = obs1[1], y = obs1[2], col = pal[1], pch = 4, cex = 1.4)
-points(x = obs2[1], y = obs2[2], col = pal[3], pch = 4, cex = 1.4)
-points(x = obs3[1], y = obs3[2], col = pal[2], pch = 4, cex = 1.4)
-points(x = obs4[1], y = obs4[2], col = pal[4], pch = 4, cex = 1.4)
-
-lines(x = c(obs1[1], obs1[1]), y = c(obs1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(obs2[1], obs2[1]), y = c(obs2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(obs3[1], obs3[1]), y = c(obs3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-lines(x = c(obs4[1], obs4[1]), y = c(obs4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
-
-# add labels
-text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
-text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
-text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
-text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
-@
+\setkeys{Gin}{width=0.8\textwidth}
+\includegraphics{draft-rain_Axams_pred}
 \caption{\label{fig:axams-dist}Total precipitation predictions by a 
@@
 
-In weather forecasting, the flexibility of distributional forests is especially appealing in 
-mountainous regions and complex terrain where a wide range of local-scale
+In weather forecasting, the flexibility of distributional forests is especially 
+appealing in mountainous regions and complex terrain where a wide range of local-scale
 effects are not yet resolved by the NWP models. 
@@
 Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
-More specifically, a Gaussian distribution left-censored at zero, is employed to model 24-hour total precipitation so that the
+More specifically, a Gaussian distribution left-censored at zero, is employed 
+to model 24-hour total precipitation so that the
 zero-censored point mass describes the probability of observing no precipitation 
@@
 
+\pagebreak
 \section{Methodology}
@@
 variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
-\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$ and
+\textit{k}-dimensional parameter vector $\bm{\theta} \in \bm{\Theta}$ and
 corresponding log-likelihood function $\ell(\bm{\theta}; Y)$. The GAMLSS
@@
 \label{eq:mle}
-\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
+\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
 \end{equation}
@@
 for reasonably representing the response distribution. Therefore, covariates
-$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
+$\bm{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
 employed to capture differences in the distribution parameters $\bm{\theta}$.
@@
 they are considered as regressors in additive predictors
-$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
+$g_j(\theta_j) = f_{j,1}(\bm{Z}) + f_{j,2}(\bm{Z}) + \dots$
 in GAMLSS.
@@
 For obtaining probabilistic predictions from the tree for a (possibly new) set
-of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
+of covariates $\bm{z} = (z_1, \ldots, z_m)$, the observation simply has to
 be ``sent down'' the tree and the corresponding segment-specific MLE has to be
-obtained. Thus, in practice $\bm{\hat \theta}(\bold{z})$ does not have to be 
-recalculated for each new $\bold{z}$ but one can simply extract the parameter estimates 
+obtained. Thus, in practice $\bm{\hat \theta}(\bm{z})$ does not have to be 
+recalculated for each new $\bm{z}$ but one can simply extract the parameter estimates 
 for the corresponding segment which have been computed already while
@@
 \begin{equation}
-w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
+w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
 \end{equation}
 where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
-for a given $\bold{z}$ is then fully specified by the estimated parameter
-$\bm{\hat \theta}(\bold{z})$ where
-\begin{equation}
-\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
+for a given $\bm{z}$ is then fully specified by the estimated parameter
+$\bm{\hat \theta}(\bm{z})$ where
+\begin{equation}
+\bm{\hat \theta}(\bm{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\bm{\theta}; y_i).
 \end{equation}
@@
 learning data obtained through resampling (bootstrap or subsampling). In each node 
-only a random subset of the covariates $\bold{Z}$ is considered for splitting to reduce the 
+only a random subset of the covariates $\bm{Z}$ is considered for splitting to reduce the 
 correlation among the trees and to stabilize the variance of the model.
@@
 needs to be specified how to compute the parameter estimates
-$\bm{\hat \theta}(\bold{z})$ for a (potentially new) set of covariates $\bold{z}$.
+$\bm{\hat \theta}(\bm{z})$ for a (potentially new) set of covariates $\bm{z}$.
 Following \cite{Hothorn+Zeileis:2017} we interpret random forests as adaptive
@@
 \begin{equation}
-w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
-\frac{\mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
+w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
+\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
 \end{equation}
@@
 segment of the $t$-th tree.
-Thus, these $w^{\text{forest}}_i(\bold{z}) \in [0, 1]$ whereas
-$w^{\text{tree}}_i(\bold{z}) \in \{0, 1\}$. Hence, weights cannot only be $0$
+Thus, these $w^{\text{forest}}_i(\bm{z}) \in [0, 1]$ whereas
+$w^{\text{tree}}_i(\bm{z}) \in \{0, 1\}$. Hence, weights cannot only be $0$
 or $1$ but change more smoothly, giving high weight to those observations $i$
 from the learning sample that co-occur in the same segment $\mathcal{B}_b^t$
-as the new observation $\bold{z}$ for many of the trees $t = 1, \dots, T$.
+as the new observation $\bm{z}$ for many of the trees $t = 1, \dots, T$.
 Consequently, the parameter estimates may, in principle, change for every
@@
 \begin{equation}
-\bm{\hat \theta}(\bold{z}) =  \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
+\bm{\hat \theta}(\bm{z}) =  \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{forest}}_i(\bm{z}) \cdot \ell(\bm{\theta}; y_i).
 \end{equation}
@@
 
+%\pagebreak
 \subsection{Data}
@@
   for each tree using a third of the predictors
-  in each split of the tree (\code{mtry}). Parameters are estimated by
+  in each split of the tree (\texttt{mtry}). Parameters are estimated by
   adaptive local likelihood based on the forest weights as described
@@
   out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
-  the number of boosting iterations (\code{mstop}) is: $50, 75, \dots, 975, 1000$.
+  the number of boosting iterations~(\texttt{mstop}) is: $50, 75, \dots, 975, 1000$.
 
@@
 \setkeys{Gin}{width=0.99\textwidth}
-<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
-# out of sample
-set.seed(7)
-par(mfrow = c(2, 2))
-qqrplot(pit_df, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_ml, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_g,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-qqrplot(pit_gb, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
-@
+\includegraphics{draft-rain_axams_qqr_oosample}
 \caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
@@
 
-<<echo=FALSE, results=hide>>=
-#### cross validation rain
-if(file.exists("crps_cross.rda")){
-  load("crps_cross.rda")
-} else {
-
-  nrep_cross <- 10
-  seed <- 7
-
-  res_cross <- mclapply(1:nrep_cross,
-                        function(i){
-
-                          set.seed(seed*i)
-
-                          # randomly split data in 7 parts each including 4 years
-                          years <- 1985:2012
-                          testyears <- list()
-                          for(j in 1:7){
-                            testyears[[j]] <- sample(years, 4, replace = FALSE)
-                            years <- years[!(years %in% testyears[[j]])]
-                          }
-
-                          #crps <- matrix(nrow = 7, ncol = 7)
-                          reslist <- list()
-                          for(k in 1:7){
-                            test <- testyears[[k]]
-                            train <- c(1985:2012)[!c(1985:2012) %in% test]
-
-                            res <- evalmodels(station = "Axams",
-                                              train = train,
-                                              test = test,
-                                              gamboost_cvr = TRUE)
-
-                            #crps[k,] <- res$crps
-                            reslist[[k]] <- res
-                          }
-
-                          #colnames(crps) <- names(res$crps)
-                          return(reslist)
-                        },
-                        mc.cores = detectCores() - 1
-  )
-
-  # extract CRPS
-  crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
-  # loop over all repetitions
-  for(i in 1:length(res_cross)){
-    #loop over all 7 folds (for 7 methods)
-    crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
-    for(j in 1:length(res_cross[[1]])){
-      crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
-    }
-    crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
-  }
-  colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps)
-
-  save(crps_cross, file = "crps_cross.rda")
-  save(res_cross, file = "res_cross.rda")
-
-}
-@
-
 \begin{figure}[t!]
 \centering
-<<rain_cross_axams_crps_skill_score, fig=TRUE, echo=FALSE, height=4.5, width = 7>>=
-par(mar = c(2.5,2,1,2))
-boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
-        names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
-        ylab = "CRPS skill score", col = "lightgray") 
-abline(h = 0, col = pal[5], lwd = 2)
-@
+\setkeys{Gin}{width=0.95\textwidth}
+\includegraphics{draft-rain_cross_axams_crps_skill_score}
 \caption{\label{fig:axams-crps}CRPS skill score from the 10 times 
@@
 \setkeys{Gin}{width=0.7\textwidth}
-<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
-par(mar = c(4,10,1,2))
-barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
-        horiz = TRUE, las = 1, axes = FALSE,
-        xlab = "Variable importance: mean increase in CRPS",
-        font.axis = 3, #list(family="HersheySerif", face=3),
-        xlim = c(0,0.08),
-        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
-axis(1, at = seq(0,1,0.02), las = 1, mgp=c(0,1,0))
-@
+\includegraphics{draft-rain_axams_varim}
 \caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
@@
 (and thus not necessarily consecutive) years.
-The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
+The models are learned on 6~folds (=~24~years) and evaluated on the 7-th fold
 (= 4~years) using the average CRPS across all observations.
@@
 
-<<echo = FALSE>>=
-#### prediction over all stations 24 - 4
-if(file.exists("crps_24to4_all.rda")){
-  load("crps_24to4_all.rda")
-} else {
-
-  data("StationsTyrol")
-  stations <- StationsTyrol$name
-  test <- 2009:2012
-  train <- 1985:2008
-
-  res_24to4_all <- mclapply(1:length(stations),
-                            function(i){
-
-                              set.seed(7)
-
-                              res <- evalmodels(station = stations[i],
-                                                train = train,
-                                                test = test,
-                                                gamboost_cvr = TRUE)
-
-                              return(res)
-                            },
-                            mc.cores = detectCores() - 1
-  )
-
-  # extract crps
-  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
-  # loop over all stations
-  for(i in 1:length(stations)){
-    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
-  }
-
-  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
-  rownames(crps_24to4_all) <- stations
-
-  save(crps_24to4_all, file = "crps_24to4_all.rda")
-  save(res_24to4_all, file = "res_24to4_all.rda")
-
-}
-
-# skill score
-s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
-colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
-
-## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
-
-crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]
-
-# best method
-bst <- apply(crps_map, 1, which.min)
-
-# distance of forest to best other method
-dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]
-
-# breaks/groups
-brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
-#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
-grp <- cut(dst, breaks = brk)
-
-# HCL colors (relatively flashy, essentially CARTO Tropic)
-clr <- colorspace::diverge_hcl(5, h = c(130, 320), c = 70, l = c(50, 90), power = 1.3)
-
-library("raster") # dem (digital elevation model)
-library("sp")     # gadm www.gadm.org/country
-
-#load("plot_map_rain/data/tirol.rda")
-#load("plot_map_rain/data/dem.rda")
-
-data(StationsTyrol, package = "RainTyrol")
-data(MapTyrol, package = "RainTyrol")
-# data(MapTyrol_border, package = "RainTyrol")
-# Create SpatialPointsDataFrame from station list
-sp <- SpatialPointsDataFrame(subset(StationsTyrol,
-                                    select=c(lon,lat)),
-                             data = subset(StationsTyrol,
-                                           select = -c(lon,lat)),
-                             proj4string = crs(MapTyrol$RasterLayer))
-@
-
 \begin{figure}[t!]
 \centering
-<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, height = 4.5, width = 7>>=
-par(mar = c(3.1, 4.1, 1.1, 2.1))
-matplot(t(s[,]), type = "l", lwd = 2, 
-          col = gray(0.5, alpha = 0.2),
-          lty = 1, axes = FALSE, 
-          xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
-lines(s[70,], col = pal[2], type = "o", pch = 19, lwd = 2)  
-# Station Axams is the 77th station which is the 70th in the list of complete stations
-  boxplot(s, add = TRUE, col = "transparent")
-  abline(h = 0, col = pal[5], lwd = 2)
-
-@
+\setkeys{Gin}{width=0.99\textwidth}
+\includegraphics{draft-rain_all_crps_skill_score}
 \caption{\label{fig:all-crps}CRPS skill score for each station (gray lines
@@
 \setkeys{Gin}{width=0.99\textwidth}
-<<map, fig=TRUE, echo=FALSE, height=6.1, width=9>>=
-
-  ## plot map of Tyrol with all 95 observations
-  layout(cbind(1, 2), width = c(9, 1))
-  par(mar = c(5,4,4,0.1))
-  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
-                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
-                xlim = c(9.8,13.2), 
-                ylim = c(46.6, 47.87))
-  plot(MapTyrol$SpatialPolygons, add = TRUE)
-  points(sp[70,], pch = 19, col = "black", cex = 1.85)
-  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
-  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
-  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
-  par(mar = c(0.5,0.2,0.5,2.3))
-  ## legend
-  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
-       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
-  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
-  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))
-
-@
+\includegraphics{draft-map}
 \caption{\label{fig:all-map}Map of Tyrol coding the best-performing model
@@
 
-\subsubsection{Distributional specifications}
+\subsubsection*{Distributional specifications}
 
@@
 
+%\pagebreak
 \subsubsection*{Computational difficulties}
@@
 The proposed methods are in the \textsf{R} package \textbf{disttree}
-(version~\Sexpr{packageVersion("disttree")}) based on the \textbf{partykit}
-package (version~\Sexpr{packageVersion("partykit")}), both available 
+(version~0.1.0) based on the \textbf{partykit}
+package (version~1.2.3), both available 
 on \textsf{R}-Forge at
 (\url{https://R-Forge.R-project.org/projects/partykit/}). The function
-\code{distfit} fits distributional models by maximum likelihood, which is
-used as the basis for the tree-building function \code{disttree}, upon which
-the \code{distforest} is built. All functions can either be used with
+\texttt{distfit} fits distributional models by maximum likelihood, which is
+used as the basis for the tree-building function \texttt{disttree}, upon which
+the \texttt{distforest} is built. All functions can either be used with
 GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
-(\citealp{Stasinopoulos+Rigby:2007}, version~\Sexpr{packageVersion("gamlss.dist")})
+(\citealp{Stasinopoulos+Rigby:2007}, version~5.0.6)
 or with custom lists containing all required information about the distribution family.
@@
 \textsf{R} package \textbf{crch} 
-\citep[version~\Sexpr{packageVersion("crch")}]{Messner+Mayr+Zeileis:2016} for the EMOS 
+\citep[version~1.0.1]{Messner+Mayr+Zeileis:2016} for the EMOS 
 models, \textbf{gamlss}
-\citep[version~\Sexpr{packageVersion("gamlss")}]{Stasinopoulos+Rigby:2007} for the 
+\citep[version~5.1.0]{Stasinopoulos+Rigby:2007} for the 
 prespecified GAMLSS, and \textbf{gamboostLSS}
-\citep[version~\Sexpr{packageVersion("gamboostLSS")}]{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
+\citep[version~2.0.1]{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.
 
@@
 observation station Axams (including Figure~\ref{fig:axams-dist})
-is reproducible using \code{demo("RainAxams", package = "disttree")}.
+is reproducible using \texttt{demo("RainAxams", package = "disttree")}.
 This also includes fitting the other zero-censored
@@
 Full replication of all results can be obtained with
-\code{demo("RainTyrol", package = "disttree")}
+\texttt{demo("RainTyrol", package = "disttree")}
 requiring the companion \textsf{R} package \textbf{RainTyrol} 
-(version~\Sexpr{packageVersion("RainTyrol")}),
+(version~0.1.0),
 also available within the \textsf{R}-Forge~project.
@@
 Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
-\code{demo("RainDistributions", package = "disttree")} and 
-\code{demo("RainStationwise", package = "disttree")}, respectively.
-
-\section*{Appendix}
+\texttt{demo("RainDistributions", package = "disttree")} and 
+\texttt{demo("RainStationwise", package = "disttree")}, respectively.
+
+\pagebreak
 \begin{appendix}
-
 \section{Tree algorithm}
@@
 $\bm{\hat{\theta}} = 
-(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
+(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$, 
 can be plugged in the score function $s(\bm{\theta}, Y)$.
@@
 a tree is sometimes also referred to as ``pre-pruning''. For random forests
-pre-pruning is often switched off by setting the significance level to $1$.
+pre-pruning is often switched off by setting the significance level to~$1$.
 
@@
 two new subgroups, without any particular ordering, that are defined by 
-splitting in split point $r$ of variable $Z_{l^{\ast}}$.
+splitting in split point $r$ of variable~$Z_{l^{\ast}}$.
 The split point is then chosen as follows:
@@
 %%\begin{subfigure}{0.8\textwidth}
-<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7, eval=FALSE>>=
-
-# out of sample
-set.seed(4)
-
-par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
-pithist(pit_df, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
-pithist(pit_ml, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
-pithist(pit_g, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
-pithist(pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
-
-@
 %%\end{subfigure}
@@
 
-\section*{Supplementary material}
-\subsection*{Supplement A: Different Response Distributions}
+\pagebreak
+% AOS,AOAS: If there are supplements please fill:
+\begin{supplement}[id=suppA]
+  \sname{Supplement A}
+  \stitle{Different Response Distributions}
+  \slink[doi]{10.1214/00-AOASXXXXSUPPA}
+  \sdatatype{.pdf} 
+  \sdescription{
   To assess the goodness of fit of the Gaussian distribution, left-censored at zero,
@@
   truncated at zero, for the positive precipitation observations. 
-
-\subsection*{Supplement B: Stationwise Evaluation}
+}
+\end{supplement}
+
+\begin{supplement}[id=suppB]
+  \sname{Supplement B}
+  \stitle{Stationwise Evaluation}
+  \slink[doi]{10.1214/00-AOASXXXXSUPPB}
+  \sdatatype{.pdf} 
+  \sdescription{
   To show that Axams is a fairly typical
   station and similar insights can be obtained for other stations as well,
-  this supplement presents the same analysis as in Section~3.3 of the main
-  manuscript for 14~further meteorological stations.
-
+  this supplement presents the same analysis as in Section~\ref{sec:axams} 
+  of the main manuscript for 14~further meteorological stations.
+}
+\end{supplement}
+
+\bibliographystyle{imsart-nameyear}
 \bibliography{ref.bib}

### 2018-11-04 pkg/disttree/inst/draft/draft.Rnw
final formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2378 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4503d2d73b067eabaa07a9184943d75e94a802fc

@@
 
+\newpage
 \section*{Appendix}
@@
   station and similar insights can be obtained for other stations as well,
-  this supplement presents the same analysis as in Section~3.3 of the main
+  this supplement presents the same analysis as in Section~\ref{sec:axams} of the main
   manuscript for 14~further meteorological stations.
 
+\newpage
 \bibliography{ref.bib}

### 2018-11-04 pkg/disttree/inst/draft/rejoinder.Rnw
removed highlighting of changed/revised parts; typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2376 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 16004c6499fa53e2d5192a1c5cfc9f6db6a312de
final formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2378 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4503d2d73b067eabaa07a9184943d75e94a802fc

@@
 
-\item \emph{Supplement A: Different response distributions.}\\
+\item \emph{Supplement A: Different Response Distributions.}\\
 To assess the appropriateness and robustness of the distributional
@@
 
-\item \emph{Supplement B: Stationwise evaluation.}\\
+\item \emph{Supplement B: Stationwise Evaluation.}\\
 To show that the more extensive cross-validation results -- obtained only
@@
 \hspace*{-0.3cm} 
-%\includegraphics[width = 0.23\textwidth]{signature-Schlosser}
+\includegraphics[width = 0.23\textwidth]{signature-Schlosser}
 

### 2018-11-04 pkg/disttree/inst/draft/supplement/Rain_stationwise_results.Rnw
changed numbering of supplements (A and B instead of 1 and 2)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2369 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9093ae989a286d620c85bcfcb92b68ad87d3c504
final formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2378 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4503d2d73b067eabaa07a9184943d75e94a802fc

@@
 if(!all(is.na(results$pit_g))) pithist(results$pit_g,  nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0, 1.9))
-pithist(results$pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "boosted GAMLSS", ylim = c(0, 1.9))
+pithist(results$pit_gb, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0, 1.9))
 @

### 2018-11-08 pkg/model4you/inst/JORS/man_model4you.Rnw
comments Achim JORS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2382 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 699094a147e87fe7824050639f16b5755f9a557c

@@
 \begin{document}
+\hypersetup{pageanchor=false}
 
@@
 personalised models and for getting knowledge about the interactions between
-student characteristics and the treatment.
+student characteristics and the exam group.
 \new{
@@
 dependence plot \citep{friedman2001greedy}, as each point in the figure is one
-patient in the given data set.
-}
-Since the percentage of successful online tests is measured on a grid, a
-\new{beeswarm} plot possibly shows the relationships even better than the
-scatter plot (both shown in Figure~\ref{fig:math_dp1},
-\new{
-Figure~\ref{fig:math_dp11} includes a loess curve). We see a nonlinear relation
-between the percentage of successful online \texttt{\Sexpr{'tests'}} and the exam group
-effect. Students with great performance during the semenster are estimated to
-have the strongest effect.
+patient in the given data set.  Since the percentage of successful online tests
+is measured on a grid, a beeswarm plot (a variant of jittered scatter plot)
+possibly shows the relationships even better than the scatter plot (both shown
+in Figure~\ref{fig:math_dp1}, Figure~\ref{fig:math_dp11} includes a loess
+curve). We see a nonlinear relation between the percentage of successful online
+\texttt{\Sexpr{'tests'}} and the exam group effect. Students with great
+performance during the semenster are estimated to have the strongest effect.
 }
@@
 exam (\texttt{\Sexpr{'attempt'}}) and \texttt{\Sexpr{'gender'}} we can use box plots,
-\new{beeswarm} plots or a combination thereof can be used (as in
+\new{beeswarm} plots or a combination thereof (as in
 Figure~\ref{fig:math_dp21}). Figure~\ref{fig:math_dp22} shows the number of
@@
 variables.
+\new{
+Currently the packages supports a limited set of model types (linear and generalised
+linear models and survival models). Further models can be added. 
+}
 

### 2018-12-16 pkg/partykit/inst/guideliketest/draft/draft.Rnw
restructuring of chapter on unbiased recursive partitioning and on the unifying framework; selection of presented results; further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2393 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c81d78a543e39934aa64777e9b14cacec5c1d449

@@
 %% packages
-\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern}
+\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
+\usepackage[all]{hypcap}
 
-%% need no \usepackage{Sweave}
+%% need no \usepackage{Sweave} ?
+%\usepackage{Sweave}
 \SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
@@
 library("lattice")
+library("scales")
+# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
 source("../guidelike.R")
 source("../tests.R")
-load("../sim/simres20180422_1step.rda")
-sim1step <- simres
+#load("../sim/simres20180422_1step.rda")
+load("../sim/simres20181122_stump.rda")
+sim_stump <- simres$res
 #load("../sim/simres20180422_2step.rda")
@@
 rm(simres)
+load("../sim/simres20181122_tree.rda")
+sim_tree <- simres$res
+rm(simres)
+
+## HCL palette
+pal <- hcl(c(140, 80, 260, 50, 10), 300, 60)
+names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_full")
+
 @
 
-\title{Comparing Testing Strategies of Unbiased Tree Algorithms}
-%\Shorttitle{Comparing Different Testing Approaches in Tree Algorithms}
+\title{The Power of Unbiased Recursive Partitioning:\\
+A Unifying View of CTree, MOB, and GUIDE}
+\Shorttitle{The Power of Unbiased Recursive Partitioning}
+%Comparing Testing Strategies of Unbiased Tree Algorithms
 
@@
 
-\Abstract{Comparing Testing Strategies of Unbiased Tree Algorithms}
+\Abstract{
+A core step of every algorithm for learning regression trees is the decision if, 
+how, and where to split the underlying data - or, in other words, the selection 
+of the best splitting variable from the available covariates and the corresponding 
+split point. Early tree algorithms (e.g., AID, CART) employ greedy search strategies 
+that directly compare all possible split points in all available covariates. However, 
+subsequent research showed that this is biased towards selection of covariates with 
+more potential split points. Therefore, unbiased recursive partitioning algorithms 
+have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
+based on statistical inference using p~-~values that are appropriately adjusted for 
+the possible split points. In a second step a split point optimizing some objective 
+function is selected in the chosen split variable. However, different unbiased tree 
+algorithms employ different inference frameworks for computing these p~-~values and 
+their relative advantages or disadvantages are not well understood, yet.
+
+Therefore, three different approaches are considered here and embedded into a common 
+modeling framework with special emphasis to linear model trees: classical categorical 
+association tests (GUIDE), conditional inference (CTree), parameter instability 
+tests (MOB). It is assessed how different building blocks affect the power of the 
+tree algorithms to select the appropriate covariates for splitting: residuals vs. full 
+model scores, binarization of residuals/scores at zero, binning of covariates, 
+conditional vs. unconditional approximations of the null distribution.
+}
 \Keywords{classification and regression trees, independence tests, recursive partitioning}
@@
 
-\section{Introduction}
-In many situations fitting one global model to a data set can be very challenging, especially if this data contains lots of different features with high/extreme variations or complex interactions.
-Therefore, separating the data into more homogeneous subgroups based on (a set of) covariates first can simplify the task and fitting a local model to each of the resulting subgroups often leads to better results.
-This separation of the data can be done by applying a tree algorithm.
-Many different approaches and algorithms have been developed in this research field.
-While they all follow the idea of splitting the data in a way such that some objective function is minimized/maximized, they differ in their approach to the crucial decision that has to made before each split: In which variable and at which break point should a split be performed.
-(Early tree algorithms rely on exhaustive search procedures to find the best split point and split variable in one step.
-However, it has been shown that this is not only computatitionaly expensive but also biased towards split variables with many possible split points.
-Therefore, succeeding algorithms select a split variable in a first step and then search for the best split point within this variable only in a second step.
-In this paper the focus will be put on the task of selecting the (right/best) split variable. 
-In particular the strategy of the GUIDE \citep{Loh:2002} algorithm is compared to the one of the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} and the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008}.
-Moreover, adapted versions of these algorithms are evaluated as well in order to compare effects of specific differences separately.
-After introducing the setting for the simulation study in Section~\ref{sec:settings} and a general scheme on how to build a tree in Section~\ref{sec:tree}, the three main differences will be discussed in Sections~\ref{sec:scores}, \ref{sec:categorization} and \ref{sec:binarization}.
+\section{Introduction}\label{sec:introduction}
+In many situations fitting one global model to a data set can be very challenging, 
+especially if this data contains lots of different features with high/extreme variations 
+or complex interactions. Therefore, separating the data into more homogeneous subgroups 
+based on (a set of) covariates first can simplify the task and fitting a local model to 
+each of the resulting subgroups often leads to better results. This separation of the 
+data can be done by applying a tree algorithm. Many different algorithms have been developed
+in this research field. While they all follow the general idea of splitting the data such 
+that some objective function is minimized/maximized, they differ in their 
+specific/particular approaches to selecting a split variable and the corresponding split point. 
+Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
+CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
+the best split point and split variable in one step by directly comparing all possible split
+points in all possible split variables. However, it has been shown that this is not only
+computationally expensive but also biased towards split variables with many possible split
+points \citep{Doyle:1973, Kim+Loh:2001}. 
+Therefore, selecting a split variable in a first step and then searching for the 
+best split point only within this variable in a separated second step is a more 
+promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
+CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
+and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
+For the first step of selecting a split variable they all follow the idea of 
+choosing the covariate which shows the highest association to/influence on the response variable 
+based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
+tests for contingency tables, CTree applies permutation tests in a conditional inference
+framework and MOB uses fluctuation tests based on asymptotic theory. 
+In this paper the focus will be put on this first step, i.e. the task of selecting 
+the right/best split variable. In particular, the approach of the GUIDE algorithm is
+compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
+unifying framework for unbiased model based tree algorithms is presented such that each 
+of the three algorithms can be obtained by a specific combination of the available building 
+blocks. In that way, the effects of each of these characteristic features of the GUIDE,
+CTree and MOB algorithm can be investigated separately by varying the switches in
+the tool box provided by this general framework.
 
-\section{Simulation Settings}
-\label{sec:settings}
-\subsection{Data Generating Process}
-The different approaches are evaluated on generated data sets.
-Each data set consists of a response variable, one regressor, one split variable and nine noise variables.
-In particular, the variables listed in Table~\ref{tab:variables} together with the corresponding distributions are included in the data generating process.
+In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
+Then, after explaining the investigated testing strategies more in detail in 
+Section~\ref{sec:strategies} the unifying framework for model based unbiased
+recursive partitioning is presented in Section~\ref{sec:unify}.
+The setting for the simulation study is introduced in Section~\ref{sec:simulation}
+and the results are discussed in Section~\ref{sec:results}.
 
-\vspace{0.2cm}
+\section{Unbiased Recursive Partitioning}
+\label{sec:urp}
+The basic idea of building a regression tree model is to partition the data into smaller and more 
+homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
+have been developed, all following the same general structure:
+\begin{enumerate}
+\item For the whole underlying data set in the current node some form of discrepancy 
+measure is calculated. 
+%This can for example be the sum of deviations from a typical or average value of the 
+%data set or it can be based on a fitted model.
+\item Among all available covariates one is selected as split variable and a value 
+within the range/measurable space of this variable is chosen as split point such that 
+after splitting the data the discrepancy measure is minimal within the resulting subgroups.
+\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criterion
+is reached or the subgroup consists of only one or all equal observations. (In the second
+case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
+\end{enumerate}
+
+The employed discrepancy measure can for example be the sum of deviations from a typical 
+or average value, but it can also be constructed based on a statistical model. For instance,
+different types of residuals or their signs can be applied (see e.g., \cite{Loh:2002}) as well as
+rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
+(In the simulation study of this paper trees with linear models fitted in each node are 
+investigated with the residual sum of squares (RSS) employed as discrepancy measure. However, 
+the discussed methodology and the presented results also apply for a wide range of other 
+model types.)
+
+As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
+GUIDE first select the split variable and then the split point in a separate step. 
+For the first step of selecting a split variable they all apply statistical (independence) tests
+following the same general strategy: 
+Based on the discrepancy measure an $n \times k$ - matrix $D$ is obtained giving information
+on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
+current node and $k \in \mathbb{N}$.
+Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
+variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
+of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
+values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
+necessary to calculate the corresponding p~-~values in order to allow for an unbiased
+variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
+p~-~value is selected as split variable as it has shown to be the variable with the strongest
+influence on the goodness of fit of the model.
+% Hinweis: pruning-Strategien
+%Once the split variable is chosen the corresponding split point is then selected such that it 
+%minimizes the discrepancy measure.
+
+While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
+in their strategies on how to calculate test statistics. In order to point out these specific 
+characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
+explained in detail in Section~\ref{sec:strategies}.
+
+%To emphasize the differences between these three algorithms in how they select split variables,
+%their testing strategies are explained in detail in Section~\ref{sec:strategies} 
+%using the example of tree models with a linear model fitted in each node as discussed
+%in the simulation study. Again, this is done only for means of simplification and the same 
+%methodology can be applied for a wide variety of models.
+
+\subsection{Testing strategies}\label{sec:strategies}
+
+\subsubsection{CTree}\label{sec:ctree}
+The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
+a conditional inference framework by applying permutation tests. In its basic form linear test 
+statistics are applied to measure the dependency between a covariate and the response variable. 
+Optionally, maximally selected statistics can be employed as well. By allowing for transformations 
+of the covariates and the application of so~-~called influence functions on the response variable 
+within the test statistic a model-based approach is also possible. For example, the influence
+function can calculate a model-based measure of goodness of fit as a discrepancy measure 
+for each response value (in a permutation symmetric way) such that the test statistic evaluates 
+the dependency of a covariate on the goodness of fit.
+Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
+statistic given all permutations of the response variable can be calculated and used to standardize 
+the test statistic. In order to map multivariate test statistics into the real line two options are 
+given: either the maximum of the absolute values of the standardized statistic or a quadratic form 
+is considered. The resulting values can then be transformed to the scale of p~-~values where they 
+can be compared in an unbiased way to select a split variable.
+In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
+observed values or optionally at an average value lying between two observations. To select the
+best split point two-sample linear statistics of the same structure as in the independence test 
+are evaluated for all possible subsets of the sample space of the selected split variable.
+
+\subsubsection{MOB}\label{sec:mob}
+While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
+was developed with the aim of embedding parametric models into the framework of regression trees
+and is hence purely intended for a model-based approach.
+
+An objective function $\psi(Y,X,\theta)$ depending on observations $(Y,X)$, with a response $Y$ and 
+covariate(s) $X$, and a $k$-dimensional vector of parameters $\theta$ is minimized to obtain the 
+parameter estimate $\hat{\theta}$ for the considered parametric model fitted in the current node of 
+the tree. For this purpose many different M-type estimators can be applied including maximum likelihood 
+or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
+algorithm can be employed as objective function. 
+To obtain additional information on the goodness of fit of the parametric model so~-~called scores are
+calculated. Scores are the values of the derivative of the objective function with respect to the parameters 
+with the estimated parameters and the observed values plugged in:
+$$
+S(y, x, \hat{\theta}) = \frac{\partial \psi(y, x, \hat{\theta})}{\partial \theta}
+$$
+which is a $n \times k$ - matrix for a $k$-dimensional parameter vector $\theta$ and
+a data set of $n$ observations containing the response vector $y = (y_1,\ldots,y_n)^\top$
+and regressor(s) $x = (x_1,\ldots,x_n)^\top$. 
+In that way a score value indicating how well the model fits the data is given separately
+for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
+and $l \in \{1,\ldots,k\}$. 
+After calculating these scores the aim is to find out whether there is any instability in the parameters
+over each particular ordering induced by the possible partitioning variables. If a covariate has
+no specific influence on the parameters the scores would always fluctuate randomly around their 
+mean 0. Therefore, these deviations are investigated by measuring them using an empirical fluctuation 
+process. Applying a scalar function which captures the fluctuation in this process yields the
+test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
+the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
+In this parametric unconditional framework p~-~values can be calculated and the covariate
+with the smallest p~-~value is selected as split variable. Then an exhaustive search over all possible
+partitions can be applied to find the best split point. Hence, the number of possible breakpoints is
+not fixed unlike in the CTree algorithm which can make this operation computationally expensive.
+
+\subsubsection{GUIDE}\label{sec:guide}
+Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
+From the model fitted to the data in the current node residuals are calculated. But before applying
+a statistical test two additional steps are performed (to prepare(?) the data):
+The residuals are binarized, i.e. depending on whether the sign of a residual is 
+positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
+and for each possible split variable the observed values are categorized into 4 groups 
+separated by the corresponding 1st, 2nd and 3rd quartile.
+Then, a $\chi^2$ test of independence is performed for the binarized residuals and
+each categorized split variable.
+After choosing the split variable showing the highest dependency by yielding the lowest p~-~value,
+the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
+is determined by the categorization of the possible split variables such that the quartiles are
+the only three possible split points.
+
+\bigskip
+
+\textit{Note:}
+While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
+method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
+independence tests such that no split is performed if the lowest p~-~value is not smaller than 
+the significance level.
+
+\section{Unifying Framework}\label{sec:unify}
+As all three algorithms follow the same basic structure their approaches can be presented together
+in a unifying framework for a model based tree algorithm from which each specific algorithm can be 
+obtained by switching on or off certain features. By applying such a general framework each of the 
+characteristic features of the three algorithms and its impact can be investigated in detail. 
+These switches are explained in this Section and their setting for the algorithms
+CTree, MOB, and GUIDE are listed in Table~\ref{tab:switches}.
+
+\subsubsection{Residuals vs full model scores}
+Following the basic tree building strategy in Section~\ref{sec:urp}, a $n \times k$ - matrix is 
+obtained based on the discrepancy measure. In the here discussed unifying framework this matrix 
+is the score matrix $S$ as presented in Section~\ref{sec:mob} providing information on the 
+goodness of fit for each observation and model parameter.
+Since the first column of the score matrix is the vector of residuals the decision of using either 
+residuals or full model scores for the test statistic is equivalent to the decision of using only 
+the first column or the whole matrix $S$.
+
+By including all available scores the goodness of fit of the model is investigated regarding each 
+parameter rather than just regarding the mean/location parameter. In that way changes in parameters
+other than the mean/location parameter should be detected more easily.
+
+\textit{Example LM:} To exemplify this feature a linear model tree is considered as this allows 
+for an easy and well structured description of the investigated subjects/issues. Therefore, from
+here on this example will be used for illustrations.
+(However, the conclusions drawn for this example also apply for a wide range of other model
+types.)
+The models fitted in the nodes of the considered linear tree model are of the form
+$$
+Y = \beta_0 + \beta_1 \cdot X + \epsilon
+$$
+for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
+The coefficient vector $\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$ is estimated
+via minimizing the residual sum of squares (RSS) for the learning data set in 
+the current node, for notational simplicity generally denoted by 
+$\{(y_i, x_i)\}_{i=1\dots n}$. Hence, the discrepancy measure, i.e. the objective function is
+$$
+RSS(y_i, x_i, \boldsymbol{\beta}) = \sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2
+$$ 
+such that the corresponding score matrix is defined by
+$$
+s(y_i, x_i, \hat{\boldsymbol{\beta}}) = \frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
+$$
+which is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is two dimensional. 
+(In that way a score value indicating how well the model fits the data is given separately
+for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
+and $j \in \{1,2\}$.)
+In this example considering only residuals corresponds to including the derivatives of $RSS$
+only regarding the intercept parameter $\beta_0$, i.e. the first column of $S$.
+On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
+the derivatives of $RSS$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
+$\beta_1$ (second column).
+
+\subsubsection{Binarization of scores}
+Rather than applying the independence test directly on the selected (columns of) scores a
+binarized version of them can be used as well. In particular, the score values are binarized
+at 0 such that only their signs are considered.
+
+\subsubsection{Categorization of covariates}
+Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd and 
+3rd quartile are used to bin each possible split variable into four categories.
+
+%\subsubsection{Conditional vs unconditional}
+
+\begin{table}[t!]
+\centering
+\caption[Table caption text]{CTree, MOB, and GUIDE with the corresponding setting of
+the switches in the unifying framework and the approximation type of the underlying null distribution.}
+\label{tab:switches}
+\begin{tabular}{ l c c c c }
+\hline 
+& Scores            & Binarization  & Categorization  & Null distribution \\ 
+\hline
+CTree & full model scores & off           & off             & conditional\\
+MOB   & full model scores & off           & off             & unconditional\\
+GUIDE & residuals         & on            & on              & unconditional\\
+\hline
+\end{tabular}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%All three algorithms apply statistical tests yielding p~-~values based on which
+%the most appropriate split variable is chosen. One main difference lies in
+%the way how scores are included. While CTree and MOB directly include the values
+%of all available scores the GUIDE algorithm only uses residuals, in particular,
+%the signs of residuals and binned versions of the covariates.
+%The advantages of including all scores or only residuals as well as the impact
+%of the binarization of scores at zero and the binning of covariates is
+%investigated in detail in Section~\ref{sec:results}.
+%Hence, the GUIDE algorithm differs mainly in the setting of the applied test
+%in which the MOB and CTree algorithms are very similar. However, one essential difference
+%between the testing strategies of MOB and CTree lies in the underlying approximation 
+%of the null distribution where MOB employs an unconditional approach contrary
+%to the conditional framework of CTree.
+
+%Once the split variable is found they choose the split point minimizing the discrepancy measure, hence, 
+%the RSS in this case.
+
+%Once the split variable is chosen the search area for the optimal split point 
+%is reduced to the range of this split variable. A possible but computationally 
+%demanding approach to finding the best split point is an exhaustive search over 
+%all possible split points in the split variable. However, other
+%strategies have been developed as well, for example based on two-sample statistics such as 
+%applied in the CTree algorithm.
+
+\section{Simulation Setting and Evaluation}
+\label{sec:simulation}
+
+Two different scenarios are considered in this simulation study. First, the underlying
+tree structure based on which the data is generated is a stump, i.e. a tree with only one 
+split. By keeping the complexity of the model as low as possible the features/advantages and
+disadvantages of the different testing strategies can be elaborated more precisely and 
+the focus can be put on their power in terms of selecting the correct split variable.
+
+In the second scenario the 'true' tree structure contains two splits in two different variables
+yielding a tree with three terminal nodes. After having investigated the characteristics of 
+the employed algorithms in detail for the stump scenario this additional evaluation aims at
+showing that the same conclusions can be drawn for more complex structures.
+
+In the following the data generating process is explained in detail for the first scenario.
+The second, more complex scenario employs the same basic structure simply adding another split
+as illustrated in Figure~\ref{fig:true_tree}.
+
+\subsection{Data generating process}
+Each generated data set consists of a response variable, one regressor, one split 
+variable and nine noise variables. In particular, the variables listed in 
+Table~\ref{tab:variables} together with the corresponding distributions are 
+included in the data generating process.
 
@@
 \begin{center}
+\caption[Table caption text]{Variables included in the data generating process.}
+\label{tab:variables}
 \begin{tabular}{l l l}
@@
 variance of $Y$         & $\sigma^2$                 & 1\\
-regressor               & $X$                        & $\mathcal{U}([-1,1])$  or $(-1)^{\mathcal{B}(0.5)}$ \\
+regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\ %or $(-1)^{\mathcal{B}(0.5)}$ \\
 true split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
@@
 \end{tabular}
-\caption[Table caption text]{Variables included in the data generating process.}
-\label{tab:variables}
 \end{center}
@@
 
-\vspace{0.4cm}
-
-\noindent The location parameter $\mu$ of the normally distributed response variable $Y$ depends linearly on the regressor variable $X$ where the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation (can) depend on the true split variable $Z_1$ as explained below.
+The location parameter $\mu$ of the normally distributed response variable $Y$ depends
+linearly on the regressor variable $X$ which is a uniformly distributed variable taking 
+values between $-1$ and $1$.
+% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
+% variable (taking values between $-1$ and $1$).
+Moreover, the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation
+(can) depend on the true split variable $Z_1$ in the following way:
 \begin{center}
-$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$
+$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
 \end{center}
 
-\noindent For the coefficients $\beta_0$ and $\beta_1$ the following three different scenarios are considered:
+In particular, three different cases are considered for the coefficients $\beta_0$ 
+and $\beta_1$:
 \begin{enumerate}
-\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at 1).
-\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed (at 0).
+\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
+\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
+(at $0$).
 \item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
@@
 
-\noindent The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, can either be binary, e.g. a step function, or a continuous function. In the binary case the following definition is applied
+The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, are binary taking either the positive or
+the negative value of the effectsize, i.e. $\delta$ or $-\delta$.
+% can either be binary (taking either $\delta$ or $-\delta$), or a continuous function. 
+% In the binary case the following definition is applied for the true split point $\xi$
 \begin{align*}
@@
 \end{align*}
-while in the continuous case a simple linear function is considered in the form of
-$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
-In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in oposite directions.
-
-%\noindent Depending on whether the coefficients are binary or continuous, in the case of only one coefficient varying it is defined by 
-%\begin{align*}
-%\beta_{0/1}(Z_1) = \begin{cases}
-%+\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
-%-\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
-%Z_1 \cdot \delta \qquad \text{if } \texttt{binary\_beta=FALSE}
-%\end{cases}
-%\end{align*}
-
-%\noindent while in the case of both coefficients varying they are defined by:
-%\begin{align*}
-%\beta_{0}(Z_1) = \begin{cases}
-%-\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
-%+\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
-%Z_1 \cdot \delta \qquad \text{if } \texttt{binary\_beta=FALSE}
-%\end{cases}
-%\end{align*}
-%\begin{align*}
-%\beta_{1}(Z_1) = \begin{cases}
-%+\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
-%-\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
-%- Z_1 \cdot  \delta \hspace{0.5cm} \text{if } \texttt{binary\_beta=FALSE}
-%\end{cases}
-%\end{align*}
+%while in the continuous case a simple linear function is considered in the form of
+%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
+In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
+opposite directions.
+An illustration of the three different scenarios 
+%for binary coefficients and a continuous regressor 
+can be found in Figure~\ref{fig:dgp}.
 
 %\newpage
-\begin{figure}%[t!]
+\begin{figure}%[h!]
 \setkeys{Gin}{width=1.0\linewidth}
@@
 
-nobs <- 300
+nobs <- 250
 sigma <- 0.5
@@
 par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_0")
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
-legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
-plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_0")
-plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_0")
+plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("vary"~beta[0]),
+     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
+points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
+plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1))
 
@@
 par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_1")
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
-legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
-plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_1")
-plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_1")
+plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("vary"~beta[1]),
+     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
+points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
+plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1))
 
@@
 par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary all")
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], col = "red")
-legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
-plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary all")
-plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary all")
+plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x",
+     main = expression("vary"~beta[0]~"and"~beta[1]),
+     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
+points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], pch = 19, col = alpha("red", 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
+plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
 @
-\caption{\label{fig:dgp}{Generated data of 300 observations for binary coefficients $\beta_0$ and $\beta_1$, continuous regressor $X$, variance parameter $\sigma = 0.5$, effect size $\delta = 1$ and split point $\xi = 0$.
-Left panels: response $y$ on regressor $X$; middle panels: intercept $\beta_0$ on split variable $Z_1$; middle panels: slope coefficients $\beta_1$ on split variable $Z_1$; each for varying $\beta_0$ (top), varying $\beta_1$ (middle) and both coefficients varying (bottom).}}
+\caption{\label{fig:dgp}{Generated data set of 250 observations for binary coefficients 
+$\beta_0$ and $\beta_1$, continuous regressor $X$, variance parameter $\sigma = 0.5$, 
+effect size $\delta = 1$ and split point $\xi = 0$. 
+Left panels: response $Y$ on regressor $X$; middle panels: intercept $\beta_0$ on split 
+variable $Z_1$; middle panels: slope coefficients $\beta_1$ on split variable $Z_1$; each 
+for varying $\beta_0$ (top), varying $\beta_1$ (middle) and both coefficients varying 
+(bottom).}}
 \end{figure}
 
-%\newpage
-\begin{figure}%[t!]
-\setkeys{Gin}{width=1.0\linewidth}
-<<dgp_cont_beta_bin_reg, fig=TRUE, echo=FALSE>>=
-set.seed(7)
+For the second simulation scenario the same basic setting with the same variables is used
+with an additional split in $Z_2$ resulting in a tree with three terminal nodes. The first
+split (in $Z_1$, at $\xi$) induces a change in the slope parameter $\beta_1$ while the second
+split (in$Z_1$, also at $\xi$) corresponds to a change in the intercept $\beta_0$.
+The underlying tree structure for the data generating process is illustrated in 
+Figures~\ref{fig:true_tree} and \ref{fig:true_tree_biv}.
 
-nobs <- 300
-sigma <- 0.5
-xi <- 0
-delta <- 1
-beta0 <- 0
-beta1 <- 1
+<<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=
+d <- dgp_tree(nobs = 250)
 
-par(mfrow = c(3,3))
+sp_xi2 <- partysplit(4L, breaks = 0)
+sp_xi1 <- partysplit(3L, breaks = 0)
 
-d <- dgp_stump(nobs = nobs, binary_beta = FALSE, binary_regressor = TRUE, only_intercept = FALSE,
-          vary_beta = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)
+pn <- partynode(1L, split = sp_xi2, 
+                kids = list(
+                  partynode(2L, info = c("true parameters:", 
+                                         expression(beta[0] == 0), 
+                                         expression(beta[1] == delta))),
+                  partynode(3L, split = sp_xi1, 
+                            kids = list(
+                              partynode(4L, info = c("true parameters:", 
+                                                     expression(beta[0] == -delta), 
+                                                     expression(beta[1] == -delta))),
+                              partynode(5L, info = c("true parameters:", 
+                                                     expression(beta[0] == delta),
 
-par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_0")
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
-legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
-plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_0")
-plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_0")
+                                                     expression(beta[1] == -delta)))))))
 
-d <- dgp_stump(nobs = nobs, binary_beta = FALSE, binary_regressor = TRUE, only_intercept = FALSE,
-          vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)
+p <- party(pn, d)
 
-par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_1")
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
-legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
-plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_1")
-plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_1")
+set.seed(747523)
+d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
+              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.001)
+ctest <- evaltests(y~x|z1+z2+z3+z4+z5+z6+z7+z8+z9+z10, data = d, testfun = "ctree", stump = FALSE)
+@
 
-d <- dgp_stump(nobs = nobs, binary_beta = FALSE, binary_regressor = TRUE, only_intercept = FALSE,
-          vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
-par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary all")
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], col = "red")
-legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
-plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary all")
-plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary all")
+\begin{figure}[h!]
+\begin{center}
+\setkeys{Gin}{width=0.6\linewidth}
+<<true_tree, fig=TRUE, echo=FALSE, width=5.4, height=4>>=
+par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
+#paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
+plot(p,
+     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]),
+     drop_terminal = TRUE, tnex = 1)
 @
-\caption{\label{fig:dgp}{Generated data of 300 observations for continuous coefficients $\beta_0$ and $\beta_1$, binary regressor $X$, variance parameter $\sigma = 0.5$, effect size $\delta = 1$.
-Left panels: response $y$ on regressor $X$; middle panels: intercept $\beta_0$ on split variable $Z_1$; middle panels: slope coefficients $\beta_1$ on split variable $Z_1$; each for varying $\beta_0$ (top), varying $\beta_1$ (middle) and both coefficients varying (bottom).}}
+\caption{\label{fig:true_tree}True tree structure for split point $\xi = 0$} with the
+true model parameters $\beta_0$ and $\beta_1$ for each terminal node.
+\end{center}
 \end{figure}
 
-\newpage
-\section{Building a Tree}
-\label{sec:tree}
-The aim of building a tree model is to partition the data into smaller and more homogeneous subgroups based on (a set of) covariates. 
-Various different tree building algorithms have been developed, all following the same basic structure:
-\begin{enumerate}
-\item Starting with the whole data set, some form of discrepancy measure is calculated.
-This can for example be the sum of deviations from a typical or average value of the data set or it can be based on a fitted model such as the residual sum of squares (RSS).
-\item Among all available covariates one is selected as split variable and a value within the range/measurable space of this variable is chosen as split point such that after splitting the data the discrepancy measure is minimal within the resulting subgroups.
-% One of all available covariates together with a value within the range of this variable is chosen as split variable with the corresponding split point such that after splitting the data the discrepancy measure is minimal within the resulting subgroups.
-\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criteria is reached or the subgroup consists of only one or all equal observations.
-(In the second case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
-\end{enumerate}
+\begin{figure}[h!]
+\begin{center}
+\setkeys{Gin}{width=0.6\linewidth}
+<<true_tree_biv, fig=TRUE, echo=FALSE, width=5.4, height=4>>=
+plot(ctest, terminal_panel = node_bivplot, drop_terminal = TRUE,
+     tp_args = list(margins = c(0,1,0,1), pointcol = "white", pointcex = 0.7, lwd = 2),
+     ip_args = list(pval = FALSE, fill = paltrees[c(2, 3)]),
+     ep_args = list(digits = 1))
+@
+\caption{\label{fig:true_tree_biv}True tree structure for split point $\xi = 0$ and 
+effect size $\delta = 1$} with the corresponding linear functions being plotted
+in each terminal node.
+\end{center}
+\end{figure}
 
-% For notational simplicity the following generic tree algorithm will be explained for the root node containing observations ... 
-% At first the whole data set is considered.
-In order to find the best way of splitting the data some kind of discrepancy measure is calculated, often to a mean % or average value of the data.
-Then the data is split into subgroups such that the decrease in the discrepancy measure is maximal.
-This procedure is then
-% repeated in each of the subgroups until some stopping criterion is met.
-% In a first step a global model is fit to the whole data set and its goodness of fit is evaluated.
-This can for example be done by calculating the residual sum of squares (RSS) or any other error measurement.
+\subsection{Evaluation}
+The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
+on 100 data sets per step each consisting of 250 observations.
 
-The crucial part in such an algorithm is the selection of a split variable and the corresponding split point.
-Algorithms such as CART make this decision in one step by performing an exhaustive search over all possible split points in all available split variables.
-However, this method is not only computationally demanding but also biased as it favors variables with a high number of possible splits.
-Therefore, other algorithms, such as GUIDE, QUEST, CTree and MOB tackle the splitting decision in two separate steps: First, a split variable is chosen and, second, a split point in the chosen split variable is selected.
-One way to search for the best split variable is to apply independence tests between the values of the discrepancy measurement for each observation and the corresponding values of each of the possible split variables and select the one showing the highest dependence.
-Once the split variable is chosen the search area for the optimal split point is reduced to the range of this split variable.
-A possible but computationally demanding approach to finding the best split point is an exhaustive search over all possible split points in the split variable.
-However, other strategies have been developed as well, for example based on two sample statistics such as applied in the CTree algorithm.
+By applying the unifying framework presented in Section~\ref{sec:unify} the effect of
+each switch can be investigated on its own. In Table~\ref{tab:switches} the setting of these 
+switches is listed for the algorithms CTree, MOB and GUIDE in their original form.
+However, in order tobe able to focus on particular features rather than the whole algorithms
+other combinations as presented in Table~\ref{tab:switches}, hence variations of the original 
+algorithms, are evaluated as well.
 
-% After splitting the data into subgroups it has to be checked whether a predefined stopping criterion has already been fulfilled.
-Otherwise, the foregoing steps are repeated for each of the new subgroups, e.g.
-a model is fitted in each subgroup, goodness of fit is evaluated and split variables and split points are selected.
-This procedure is repeated until a stopping criterion is met or, if no criterion has been set in advance, until each subgroup consists of only one value or only equal values.
-In the second case the resulting tree is of maximal size and can be reduced by (post-)pruning.
+Moreover, all of the considered statistical tests are applied as significance tests answering 
+two questions at once: (1) Should a split be performed at all? (2) If so, in which variable?
+For the first question the p~-~values regarding all available covariates are compared 
+to a predefined level of significance $\alpha$. Only if the smallest p~-~value is smaller than
+$\alpha$ a split is performed and the covariate corresponding to this p~-~value is selected.
+Including this criterion of significance in the building process of a tree is often refereed
+to as \textit{pre-pruning}, however, this is not always applied in practice as significance
+is often not considered. As mentioned before, the GUIDE algorithm usually builds large 
+trees which are then pruned afterwards. However, for the purpose of comparing the testing
+strategies rather than the whole tree algorithms in this simulation study a significance
+level is considered for all statistical tests, including the $\xi^2$-test in the GUIDE algorithm
+such that all algorithms apply pre-pruning rather than post-pruning.
 
-%Summed up, the steps of building a model-based tree are:
-%\begin{enumerate}
-%\item Fit a global model to the (whole) data set.
-%\item Evaluate the goodness of fit of this model.
-%\item Select the split variable showing the highest dependence to the values of the goodness of fit-measurement and the split point leading to the highest improvement in goodness of fit and divide the data into subgroups.
-%\item Repeat steps 1--3 for each resulting subgroup until a predefined stopping criterion is met.
-%\end{enumerate}
+To compare the performance of the evaluated testing strategies for each step the proportion
+of data sets for which the true splitting variable $Z_1$ is detected and also selected 
+(for a given significance level $\alpha = 0.05$) is considered for the first scenario (stump).
+For the second scenario (tree with 2 splits) the adjusted Rand index (ARI) is calculated 
+as a measure of similarity between the true tree structure and the model tree.
 
-\newpage
-\section{Which scores should be used?}
+\section{Results}
+\label{sec:results}
+
+To illustrate the effects of the switches in the unifying framework presented in 
+Section~\ref{sec:unify} each of them is investigated separately for the first
+simulation scenario (stump) in Sections~\ref{sec:scores}, \ref{sec:categorization}
+and \ref{sec:binarization}. Then the testing strategies are evaluated on a more
+complex tree structure in Section~\ref{sec:tree}. The corresponding results are shown 
+in the following plots.
+
+\subsection{Residuals vs full model scores}
 \label{sec:scores}
+To assess the impact of using only residuals or the full score matrix four testing
+strategies are compared: CTree, GUIDE, MOB (all three in their original form),
+and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_full).
+Thus, the four strategies are obtained from the unifying framework by setting the switches 
+as listed in Table~\ref{tab:switches_scores}.
 
-In this simulation study the models that are fit to the data in each node of the tree are linear models.
-The fitting is done by estimating the coefficient vector $\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$ via minimizing an objective function $\Psi(y_i,x_i,\boldsymbol{\beta})$ summed up over all observations of the learning data set that are in the current node, for notational simplicity generally denoted by $\{(y_i, x_i)\}_{i=1\dots n}$.
-Here, this sum is the residual sum of squares.
-Information about the goodness of fit of the resulting model can be obtained by calculating so called scores.
-These are the values of the objective function derived by the coefficients with the given data and the estimated coefficients $\hat{\boldsymbol{\beta}}$ plugged in:
+\begin{table}[h!]
+\centering
+\caption[Table caption text]{CTree, MOB, GUIDE, and GUIDE\_full with the corresponding setting of
+the switches in the unifying framework and the approximation type of the underlying null distribution.}
+\label{tab:switches_scores}
+\begin{tabular}{ l c c c c }
+\hline 
+& Scores            & Binarization  & Categorization  & Null distribution \\ \hline
+CTree       & full model scores & off           & off             & conditional\\
+MOB         & full model scores & off           & off             & unconditional\\
+GUIDE       & residuals         & on            & on              & unconditional\\
+GUIDE\_full & full model scores & on            & on              & unconditional\\ \hline
+\end{tabular}
+\end{table}
 
-$$
-s(y_i, x_i, \hat{\boldsymbol{\beta}}) = \frac{\partial \Psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
-$$
+Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
+scenarios of varying only the intercept $\beta_0$ or only the slope parameter $\beta_1$
+or both of them. If the intercept varies (left and right panel) all strategies perform 
+similarly with CTree and MOB being slightly ahead. However, this advantage increases if 
+the slope parameter varies and the intercept is kept at a fixed value (middle panel). 
+In particular, the original strategy of GUIDE fails to detect any changes in $\beta_1$ 
+while GUIDE\_full performs reasonably well in this scenario.
+These results clearly illustrate that overall the decision to include all available scores
+when testing for dependencies in order to select split variables is advantageous over 
+considering only the scores regarding the intercept as changes not only in the mean but also 
+in all other parameters can be detected in that way.
 
-which is in this case a vector of two dimensions as $\boldsymbol{\beta}$ is two-dimensional.
-In that way a score value indicating how well the model fits the data is given separately for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$ and $j \in \{1,2\}$.
-
-In order to choose a split variable dependency tests are employed on the score values and each of the possible split variables.
-The variable showing the highest dependency is selected.
-Now the question arises which of the score values should be considered for these tests.
-Different approaches have been established.
-While the GUIDE-algorithm \citep{Loh:2002} only includes the scores of the intercept coefficient $\beta_0$, the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} and the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} consider all available score values.
-In Figure~\ref{fig:which_scores} the results of these two strategies applied on the before described simulation setting (for \texttt{binary\_beta = FALSE} and  \texttt{binary\_regressor = TRUE}) can be compared.
-Additionally, an adapted version of the GUIDE algorithm including all available score values has been employed as well.
-For changes only in the intercept (left panel) the correct splitting variable ($Z_1$) is detected successfully by all methods with increasing effect size.
-While mfluc performs best, CTree follows second, slightly ahead of the original GUIDE version which performs better than the adapted version.
-In case of changes in the slope coefficient (middle panel) $\beta_1$ mfluc, CTree and the adapted GUIDE version again lead to similar results while the original GUIDE version fails in detecting the correct splitting variable as it was to be expected.
-In the third scenario where both coefficients vary over $Z_1$ (right panel) the adapted GUIDE version slightly outperforms the original version while CTree and mfluc are again ahead.
-These results clearly illustrate that overall the decision to include all available scores when testing for dependencies in order to select split variables is advantageous over considering only the scores regarding the intercept.
-Only in situations where changes appear solely in the intercept the original GUIDE version can lead to slighter better results than the adapted version.
-However, since this particular type of scenario usually can not be assumed to be the case in advance, its strength in this particular situation is clearly outweighed by the obvious disadvantages in all other situations.
-
-%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
 \begin{figure}
 \setkeys{Gin}{width=\linewidth}
-<<which_scores_1step_z1_binreg, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim1step, test %in% c("ctree_max","mfluc",
+<<scores_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                          "guide_sum_12", "guide_sum_1_cor"))
-subdata$test <- factor(subdata$test)
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree","mfluc",
+                                  "guide_sum_12", "guide_sum_1_cor"),
+                       labels = c("CTree","MOB",
+                                  "GUIDE_full", "GUIDE"))
 
-xyplot(prop_z1 ~ delta | vary_beta, groups = ~ test, 
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
+
+xyplot(prop_Tsplit ~ delta | vary_beta, groups = ~ test, 
        data = subdata, 
-       subset = binary_beta == FALSE & binary_regressor == TRUE & 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
          only_intercept == FALSE & xi == 0, 
-       type = "b", auto.key = TRUE, layout= c(3,1),
-       index.cond = list(c(2,3,1)))
+       type = "b", lty = c(1,1,1,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,1,1,1), type = "b", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+       layout= c(3,1),
+       index.cond = list(c(2,3,1)),
+       xlab = expression(delta),
+       ylab = expression("Proportion of selecting"~Z[1]),
+       strip = strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
+                                              expression("vary"~beta[0]), 
+                                              expression("vary"~beta[1]))),
+       par.settings = par.settings)
 @
-\caption{\label{fig:which_scores}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for continuous coefficients, binary regressor, $\sigma = 1$, 250 observations and 100 repetitions per step.
-Each panel represents one scenario for the coefficients: varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and both coefficients varying (right panel).
-The performances of 4 algorithms are compared: mfluc, tree, adapted GUIDE, original GUIDE.}
+\caption{\label{fig:scores_stump}Proportion of cases where the true split variable $Z_1$ is
+selected over increasing effect size $\delta$ for $\sigma = 1$, 250 observations and 100 data 
+sets per step. Each panel 
+represents one scenario for the coefficients: varying intercept 
+$\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and both 
+coefficients varying (right panel). The performances of four algorithms are compared: 
+CTree, MOB, GUIDE, GUIDE\_full.}
 \end{figure}
 
-\newpage
-\noindent Within the GUIDE algorithm two additional steps are made before testing for dependencies between scores and split variables. 
-\begin{itemize}
-\item The scores are binarized, i.e. depending on whether the actual values are positive or negative they are replaced by $+1$ or $-1$. 
-\item The observed values of the possible split variables are categorized into 4 groups separated by the 1st, 2nd and 3rd quartile.
-\end{itemize}
-The effects of these two steps are going to be investigated in detail in the following two sections.
+\subsection{Categorization of splitting variables}
+\label{sec:categorization}
+To find out whether a categorization of the possible splitting variables leads to 
+an improvement or a deterioration of performance the two algorithms CTree and MOB are 
+applied, once in their original version without a categorization and once in an adapted 
+version with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
+Moreover, they are compared to GUIDE\_full which includes all available scores.
 
-\section{Categorize the splitting variables?}
-\label{sec:categorization}
-To find out whether a categorization of the possible splitting variables leads to an improvement or a deterioration of performance the two algorithms CTree and MOB are applied, once in their original version without a categorization and once in an adapted version with a categorization.
-Moreover, they are compared to the adapted GUIDE version which includes scores of all available scores such as CTree and MOB do.
+Hence, the five strategies are obtained from the unifying framework by setting the switches 
+as listed in Table~\ref{tab:switches_cat}.
 
-In Figure~\ref{fig:cat_ctree} the effect of the categorization on the performance of the CTree algorithm is illustrated.
-Only for the split point $\xi = 0.5$ it leads to an improvement which was to be expected since this is one of the quartiles used for the categorization.
-However, apart from this point categorizing the values of the split variable has a negative effect on the performance, even at $\xi = 0$ which is also one of the used quartiles.
-Similar conclusions can be drawn looking at Figure~\ref{fig:cat_mfluc} where the effect on the MOB algorithm is shown.
-In this case, the categorization has almost no effect at the quartiles $\xi = 0$ and $\xi=0.5$, but again, leads to a decrease in the proportion of cases where the true splitting variable is detected.
+\begin{table}[h!]
+\centering
+\caption[Table caption text]{CTree, CTree\_cat, MOB, MOB\_cat, and GUIDE\_full with the 
+corresponding setting of the switches in the unifying framework and the approximation type 
+of the underlying null distribution.}
+\label{tab:switches_cat}
+\begin{tabular}{ l c c c c }
+\hline 
+& Scores            & Binarization  & Categorization  & Null distribution \\ 
+\hline
+CTree       & full model scores & off           & off             & conditional\\
+CTree\_cat  & full model scores & off           & on              & conditional\\
+MOB         & full model scores & off           & off             & unconditional\\
+MOB\_cat    & full model scores & off           & on              & unconditional\\
+GUIDE\_full & full model scores & on            & on              & unconditional\\
+\hline
+\end{tabular}
+\end{table}
 
-Therefore, it can be summarized that a categorization of the values of the split variable does not lead to any advantages unless the true split point corresponds to one of the quartiles used for the categorization.
-In most situations it even causes the performance to get worse.
+In Figure~\ref{fig:cat_stump} the effect of categorizing split variables on the performance 
+is illustrated for four different values of the true split point $\xi$. 
+For both, CTree and MOB, it can be stated that overall they perform better in their
+original form without any categorization. Only if the true split point $\xi$
+is close to the quartiles used as breakpoints for the categorizations both versions
+lead to a similar proportion of cases where the true splitting variable is detected
+(e.g., for CTree and $\xi=0.5$)
 
-%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
-\begin{figure}[h!]
+Therefore, it can be summarized that a categorization of the values of the split variable
+does not lead to any advantages unless the true split point corresponds to one of the 
+quartiles used for the categorization. In most situations it even causes the performance to 
+get worse.
+
+\begin{figure}%[h!]
 \setkeys{Gin}{width=\linewidth}
-<<cat_ctree_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim1step, test %in% c("ctree","ctree_cat", 
+<<cat_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
+                                         #"ctree_max", "ctree_max_cat",
+                                         "mfluc", "mfluc_cat",
                                          "guide_sum_12"))
-subdata$test <- factor(subdata$test)
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_cat", 
+                                  #"ctree_max","ctree_max_cat",
+                                  "mfluc", "mfluc_cat",
+                                  "guide_sum_12"),
+                       labels = c("CTree","CTree_cat",
+                                  #"CTree_max","CTree_max_cat",
+                                  "MOB", "MOB_cat",
+                                  "GUIDE_full"))
 
-xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
+                                                         #"CTree_max", "CTree_max", 
+                                                         "MOB", "MOB",
+                                                         "GUIDE_full")], 
+                                             fill = pal[c("CTree", "CTree", 
+                                                          #"CTree_max", "CTree_max", 
+                                                          "MOB", "MOB",
+                                                          "GUIDE_full")]), 
+                     superpose.line = list(col = pal[c("CTree", "CTree", 
+                                                       #"CTree_max", "CTree_max", 
+                                                       "MOB", "MOB",
+                                                       "GUIDE_full")]))
+
+xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
        data = subdata, 
@@
          only_intercept == FALSE & vary_beta == "beta1", 
-       type = "b", auto.key = TRUE , layout= c(4,1))
+       type = "b", 
+       lty = c(1,2,#1,2,
+               1,2,1),
+       layout= c(4,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,2,#1,2,
+                                        1,2,1), 
+                                type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "CTree", 
+                                            #"CTree_max", "CTree_max", 
+                                            "MOB", "MOB",
+                                            "GUIDE_full")], lwd = 1)), 
+       #index.cond = list(c(1,2,3,4)),
+       xlab = expression(delta),
+       ylab = expression("Proportion of selecting"~Z[1]),
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings)
 
 @
-\caption{\label{fig:cat_ctree}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step.
-Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right).
-The performances of 3 algorithms are compared: CTree, categorized versions of CTree, adapted GUIDE.}
+\caption{\label{fig:cat_stump}Proportion of cases where the true split variable $Z_1$ is
+selected over increasing effect size $\delta$ for a
+varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 data sets per 
+step. Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ 
+and $0.8$ from left to right). The performances of five testing strategies are compared: 
+CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_full.}
+%\end{figure}
+%\begin{figure}[h!]
+\setkeys{Gin}{width=\linewidth}
+<<cat_stump_all, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
+                                         #"ctree_max", "ctree_max_cat",
+                                         "mfluc", "mfluc_cat",
+                                         "guide_sum_12"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_cat", 
+                                  #"ctree_max","ctree_max_cat",
+                                  "mfluc", "mfluc_cat",
+                                  "guide_sum_12"),
+                       labels = c("CTree","CTree_cat",
+                                  #"CTree_max","CTree_max_cat",
+                                  "MOB", "MOB_cat",
+                                  "GUIDE_full"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
+                                                         #"CTree_max", "CTree_max", 
+                                                         "MOB", "MOB",
+                                                         "GUIDE_full")], 
+                                             fill = pal[c("CTree", "CTree", 
+                                                          #"CTree_max", "CTree_max", 
+                                                          "MOB", "MOB",
+                                                          "GUIDE_full")]), 
+                     superpose.line = list(col = pal[c("CTree", "CTree", 
+                                                       #"CTree_max", "CTree_max", 
+                                                       "MOB", "MOB",
+                                                       "GUIDE_full")]))
+
+xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
+         only_intercept == FALSE & vary_beta == "all", 
+       type = "b", 
+       lty = c(1,2,#1,2,
+               1,2,1),
+       layout= c(4,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,2,#1,2,
+                                        1,2,1), 
+                                type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "CTree", 
+                                            #"CTree_max", "CTree_max", 
+                                            "MOB", "MOB",
+                                            "GUIDE_full")], lwd = 1)), 
+       #index.cond = list(c(1,2,3,4)),
+       xlab = expression(delta),
+       ylab = expression("Proportion of selecting"~Z[1]),
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings)
+
+@
+\caption{\label{fig:cat_stump_all}Proportion of cases where the true split variable $Z_1$ is
+selected over increasing effect size $\delta$ for both parameters varying, $\sigma = 1$, 
+250 observations and 100 data sets per step. Each panel represents a different value of the 
+split point $\xi$ ($0$, $0.2$, $0.5$ and $0.8$ from left to right). The performances of five 
+testing strategies are compared: CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_full.}
 \end{figure}
+
+\subsection{Binarization of scores}
+\label{sec:binarization}
+To find out whether a binarization of the scores at 0 leads to an improvement or a deterioration
+of performance the two algorithms CTree and MOB are applied, once in their original version
+without a binarization and once in an adapted version with a binarization of scores at 0 
+(CTree\_bin, MOB\_bin). 
+Moreover, they are compared to the adapted GUIDE version which includes all available 
+scores (GUIDE\_full).
+
+Hence, the five strategies are obtained from the unifying framework by setting the switches 
+as listed in Table~\ref{tab:switches_bin}.
+
+\begin{table}[h!]
+\centering
+\caption[Table caption text]{CTree, CTree\_bin, MOB, MOB\_bin, and GUIDE\_full with the 
+corresponding setting of the switches in the unifying framework and the approximation type 
+of the underlying null distribution.}
+\label{tab:switches_bin}
+\begin{tabular}{ l c c c c }
+\hline 
+& Scores            & Binarization  & Categorization  & Null distribution \\ 
+\hline
+CTree       & full model scores & off           & off             & conditional\\
+CTree\_bin  & full model scores & on            & off             & conditional\\
+MOB         & full model scores & off           & off             & unconditional\\
+MOB\_bin    & full model scores & on            & off             & unconditional\\
+GUIDE\_full & full model scores & on            & on              & unconditional\\
+\hline
+\end{tabular}
+\end{table}
+
+Figure~\ref{fig:bin_stump} shows the effect of binarizing the score varlues over
+different values for the true split point $\xi$, however, all four situations
+lead to the same conclusions. Binarizing the score values decreases the proportion
+of cases where the split variable was selected correctly.
+
+%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
+\begin{figure}%[h!]
+\setkeys{Gin}{width=\linewidth}
+<<bin_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
+                                         #"ctree_max", "ctree_max_bin",
+                                         "mfluc", "mfluc_bin",
+                                         "guide_sum_12"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_bin", 
+                                  #"ctree_max", "ctree_max_bin",
+                                  "mfluc", "mfluc_bin",
+                                  "guide_sum_12"),
+                       labels = c("CTree","CTree_bin",
+                                  #"CTree_max","CTree_max_bin",
+                                  "MOB", "MOB_bin",
+                                  "GUIDE_full"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
+                                                         #"CTree_max", "CTree_max", 
+                                                         "MOB", "MOB",
+                                                         "GUIDE_full")], 
+                                             fill = pal[c("CTree", "CTree", 
+                                                          #"CTree_max", "CTree_max", 
+                                                          "MOB", "MOB",
+                                                          "GUIDE_full")]), 
+                     superpose.line = list(col = pal[c("CTree", "CTree", 
+                                                       #"CTree_max", "CTree_max", 
+                                                       "MOB", "MOB",
+                                                       "GUIDE_full")]))
+
+xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
+         only_intercept == FALSE & vary_beta == "beta1", 
+       type = "b",
+       lty = c(1,2,#1,2,
+               1,2,1),
+       layout= c(4,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,2,#1,2,
+                                        1,2,1), 
+                                type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "CTree", 
+                                            #"CTree_max", "CTree_max", 
+                                            "MOB", "MOB",
+                                            "GUIDE_full")], lwd = 1)), 
+       #index.cond = list(c(1,2,3,4)),
+       xlab = expression(delta),
+       ylab = expression("Proportion of selecting"~Z[1]),
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings)
+
+@
+\caption{\label{fig:bin_stump}Proportion of cases where the true split variable $Z_1$ is
+detected over increasing effect size $\delta$ for a 
+varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 data sets per 
+step. Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8
+from left to right). The performances of five testing strategies are compared: CTree, CTree\_bin,
+MOB, MOB\_bin, GUIDE\_full.}
+%\end{figure}
+%
+%\begin{figure}[h!]
+\setkeys{Gin}{width=\linewidth}
+<<bin_stump_all, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
+                                         #"ctree_max", "ctree_max_bin",
+                                         "mfluc", "mfluc_bin",
+                                         "guide_sum_12"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_bin", 
+                                  #"ctree_max", "ctree_max_bin",
+                                  "mfluc", "mfluc_bin",
+                                  "guide_sum_12"),
+                       labels = c("CTree","CTree_bin",
+                                  #"CTree_max","CTree_max_bin",
+                                  "MOB", "MOB_bin",
+                                  "GUIDE_full"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
+                                                         #"CTree_max", "CTree_max", 
+                                                         "MOB", "MOB",
+                                                         "GUIDE_full")], 
+                                             fill = pal[c("CTree", "CTree", 
+                                                          #"CTree_max", "CTree_max", 
+                                                          "MOB", "MOB",
+                                                          "GUIDE_full")]), 
+                     superpose.line = list(col = pal[c("CTree", "CTree", 
+                                                       #"CTree_max", "CTree_max", 
+                                                       "MOB", "MOB",
+                                                       "GUIDE_full")]))
+
+xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
+         only_intercept == FALSE & vary_beta == "all", 
+       type = "b",
+       lty = c(1,2,#1,2,
+               1,2,1),
+       layout= c(4,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,2,#1,2,
+                                        1,2,1), 
+                                type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "CTree", 
+                                            #"CTree_max", "CTree_max", 
+                                            "MOB", "MOB",
+                                            "GUIDE_full")], lwd = 1)), 
+       #index.cond = list(c(1,2,3,4)),
+       xlab = expression(delta),
+       ylab = expression("Proportion of selecting"~Z[1]),
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings)
+
+@
+\caption{\label{fig:bin_stump_all}Proportion of cases where the true split variable $Z_1$ is
+detected over increasing effect size $\delta$ for both parameter varying, $\sigma = 1$, 
+250 observations and 100 data sets per step. Each panel represents a different value of the 
+split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
+strategies are compared: CTree, CTree\_bin, MOB, MOB\_bin, GUIDE\_full.}
+\end{figure}
+
+\subsection{Tree}
+\label{sec:tree}
+To compare the performance of testing strategies in a more complex scenario they have also
+been evaluated on data sets based on a tree structure with two different split variables 
+($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
+Figures~\ref{fig:true_tree} and \ref{fig:true_tree_biv}.
+
+Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
+almost equally well for the true split points $0$, $0.2$ and $0.5$ while GUIDE\_full
+a bit worse but still achieves an ARI above 0.5 for the highest considered effect size
+of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
+due to the fact that it doesn't detect the first split in $Z_2$ corresponding to a changing
+slope parameter as it only consideres residuals in its testing strategy.
+For a split point farther away from the center such as for $\xi = 0.8$ MOB outperforms all other
+methods and both GUIDE versions fail to find the correct tree structure.
+Based on these results it can be concluded once again that including all available score in the
+testing strategy leads to a better performance while binarizing scores and categorizing
+split variables both do not cause any improvement but can even have a negative effect in many situations.
 
@@
 \setkeys{Gin}{width=\linewidth}
-<<cat_mfluc_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim1step, test %in% c("mfluc","mfluc_cat", 
-                                        "guide_sum_12"))
-subdata$test <- factor(subdata$test)
+<<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
+                                        "guide_sum_12", "guide_sum_1_cor"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "mfluc",
+                                  "guide_sum_12", "guide_sum_1_cor"),
+                       labels = c("CTree","MOB",
+                                  "GUIDE_full", "GUIDE"))
 
-xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
+
+xyplot(ari ~ delta | xi, groups = ~ test, 
        data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-                only_intercept == FALSE & vary_beta == "beta1", 
-       type = "b", auto.key = TRUE , layout= c(4,1))
+       type = "b",
+       lty = c(1,1,1,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,1,1,1), type = "b", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+       layout= c(4,1),
+       #index.cond = list(c(1,2,3,4)),
+       xlab = expression(delta),
+       ylab = "Adjusted Rand Index",
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings)
 
 @
-\caption{\label{fig:cat_mfluc}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step.
-Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right).
-The performances of 3 algorithms are compared: mfluc, categorized version of mfluc, adapted GUIDE.}
+\caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ for
+250 observations and 100 data sets per step.
+Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ and 
+$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
+MOB, GUIDE, GUIDE\_full.}
 \end{figure}
 
-\newpage
-\section{Binarize the response variable?}
-\label{sec:binarization}
-To find out whether a binarization of the scores leads to an improvement or a deterioration of performance the two algorithms CTree and MOB are applied, once in their original version without a binarization and once in an adapted version with a binarization.
-Moreover, they are compared to the adapted GUIDE version which includes scores of all available scores such as CTree and MOB do.
+\section{Discussion}
+\label{sec:discussion}
+FIX ME: only pre-pruning applied, but same results if significance level is ignored (maybe show one plot).
+Optionally compare GUIDE with prepruning to GUIDE with postpruning on tree structure?
+Check implementation of pruning.
 
-%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
-\begin{figure}[h!]
-\setkeys{Gin}{width=\linewidth}
-<<bin_ctree_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim1step, test %in% c("ctree","ctree_bin", 
-                                        "guide_sum_12"))
-subdata$test <- factor(subdata$test)
+FIX ME: LM-specific results
 
-xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-                only_intercept == FALSE & vary_beta == "beta1", 
-       type = "b", auto.key = TRUE , layout= c(4,1))
-
-@
-\caption{\label{fig:bin_ctree}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step.
-Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right).
-The performances of 3 algorithms are compared: CTree, categorized versions of CTree, adapted GUIDE.}
-\end{figure}
-
-\begin{figure}[h!]
-\setkeys{Gin}{width=\linewidth}
-<<bin_mfluc_1step_z1, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim1step, test %in% c("mfluc","mfluc_bin", 
-                                        "guide_sum_12"))
-subdata$test <- factor(subdata$test)
-
-xyplot(prop_z1 ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-                only_intercept == FALSE & vary_beta == "beta1", 
-       type = "b", auto.key = TRUE , layout= c(4,1))
-
-@
-\caption{\label{fig:bin_mfluc}Proportion of cases where the true split variable $Z_1$ is detected over increasing effect size $\delta$ for binary coefficients, continuous regressor, varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 repetitions per step.
-Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right).
-The performances of 3 algorithms are compared: mfluc, categorized version of mfluc, adapted GUIDE.}
-\end{figure}
+FIX ME: results without loss of generalization
 

### 2019-01-02 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised sections 2 and 3 and fixed typos in references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2399 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ffc676ddf3697959995a7bf4905fea229d54cf55

@@
 #load("../sim/simres20180422_1step.rda")
-load("../sim/simres20181122_stump.rda")
+load("../sim/simres20181223_stump.rda")
 sim_stump <- simres$res
@@
 rm(simres)
-load("../sim/simres20181122_tree.rda")
+load("../sim/simres20181223_tree.rda")
 sim_tree <- simres$res
@@
 ## HCL palette
-pal <- hcl(c(140, 80, 260, 50, 10), 300, 60)
+pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
 names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_full")
@@
 have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
-based on statistical inference using p~-~values that are appropriately adjusted for 
+based on statistical inference using $p$-values that are appropriately adjusted for 
 the possible split points. In a second step a split point optimizing some objective 
 function is selected in the chosen split variable. However, different unbiased tree 
-algorithms employ different inference frameworks for computing these p~-~values and 
+algorithms employ different inference frameworks for computing these $p$-values and 
 their relative advantages or disadvantages are not well understood, yet.
@@
 model scores, binarization of residuals/scores at zero, binning of covariates, 
-conditional vs. unconditional approximations of the null distribution.
+conditional vs.\ unconditional approximations of the null distribution.
 }
@@
 framework and MOB uses fluctuation tests based on asymptotic theory.
+
 In this paper the focus will be put on this first step, i.e. the task of selecting 
@@
 the tool box provided by this general framework. 
+Depending on the setup different effects can be observed for the investigated characteristics.
+However, none of the results contradicts the conclusion that in general including all available 
+model scores is at least as efficient but mostly clearly advantageous over considering only 
+residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
+the values of possible split variables are investigated and the impact of applying a conditional
+or an unconditional approximation of the null distribution is analyzed.
 
@@
 Then, after explaining the investigated testing strategies more in detail in 
-Section~\ref{sec:strategies} the unifying framework for model based unbiased
-recursive partitioning is presented in Section~\ref{sec:unify}.
+Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for model 
+based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
 The setting for the simulation study is introduced in Section~\ref{sec:simulation}
@@
 \label{sec:urp}
-The basic idea of building a regression tree model is to partition the data into smaller and more 
-homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
+
+\subsection{Generic Algorithm}
+The basic idea of building a regression tree model is to partition the data into smaller and 
+more homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
 have been developed, all following the same general structure:
@@
 values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
-necessary to calculate the corresponding p~-~values in order to allow for an unbiased
+necessary to calculate the corresponding $p$-values in order to allow for an unbiased
 variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
-p~-~value is selected as split variable as it has shown to be the variable with the strongest
+$p$-value is selected as split variable as it has shown to be the variable with the strongest
 influence on the goodness of fit of the model.
@@
 characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
-explained in detail in Section~\ref{sec:strategies}.
+explained in detail in Sections~\ref{sec:ctree}.
 
@@
 
-\subsection{Testing strategies}\label{sec:strategies}
-
-\subsubsection{CTree}\label{sec:ctree}
+\subsection{CTree}\label{sec:ctree}
+
 The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
-a conditional inference framework by applying permutation tests. In its basic form linear test 
-statistics are applied to measure the dependency between a covariate and the response variable. 
-Optionally, maximally selected statistics can be employed as well. By allowing for transformations 
-of the covariates and the application of so~-~called influence functions on the response variable 
-within the test statistic a model-based approach is also possible. For example, the influence
-function can calculate a model-based measure of goodness of fit as a discrepancy measure 
-for each response value (in a permutation symmetric way) such that the test statistic evaluates 
-the dependency of a covariate on the goodness of fit.
+a conditional inference framework by applying permutation tests. 
+To select a split variable it is tested whether there is any association between the transformed 
+response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split variable $Z$.
+For different types of response variables different functions $h$ which depend on $Y$ in a 
+permutation-symmetric way and might also be multidimensional can be selected.
+For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
+response can be mapped to a unity vector indicating the corresponding category by an indicator
+function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$. Alternatively, the function $h$ can return a location
+and scale parameter of $Y$, e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
+If a parametric model is fit to the values of the response variable~$Y$ and regressor variable(s)~$X$
+the corresponding model scores can be obtained by setting $h(Y) = s(Y,X,\hat{\theta})$
+where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function:
+$$
+s(Y, X, \hat{\theta}) = \frac{\partial \psi(Y, X, \hat{\theta})}{\partial \theta}
+$$
+with $\psi$ being the objective function which is minimized to estimate the model parameter
+$\theta$. Hence, for a data set of $n$ observations containing the response vector 
+$y = (y_1,\ldots,y_n)^\top$ and regressor(s) $x = (x_1,\ldots,x_n)^\top$ and a $k$-dimensional 
+parameter vector $\theta$ the score function $s$ returns a $n \times k$ - matrix.
+In that way a score value indicating how well the model fits the data is given separately
+for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
+and $l \in \{1,\ldots,k\}$. 
+(To estimate $\theta$ many different M-type estimators can be applied including maximum likelihood 
+or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
+algorithm can be employed as objective function.)
+
+Similarly, different types of functions can be chosen for the influence function $g$ depending
+on a possible split variable $Z$. A simple choice in case of $Z$ being a numeric variable is again
+the identity function $g(Z) = Z$. For categorical variables $g$ can also map its values to the 
+corresponding unity vectors by $g(Z) = (0,\ldots,1,\ldots,0)$.
+
+To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
 Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
@@
 given: either the maximum of the absolute values of the standardized statistic or a quadratic form 
-is considered. The resulting values can then be transformed to the scale of p~-~values where they 
+is considered. The resulting values can then be transformed to the scale of $p$-values where they 
 can be compared in an unbiased way to select a split variable.
+
+If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
+correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
+is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
+provides a very general framework where the choice of the type of variables and the corresponding 
+functions $g$ and $h$ determine the type of the applied statistical test.
+
 In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
@@
 
-\subsubsection{MOB}\label{sec:mob}
+\subsection{MOB}\label{sec:mob}
 While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
 was developed with the aim of embedding parametric models into the framework of regression trees
-and is hence purely intended for a model-based approach.
-
-An objective function $\psi(Y,X,\theta)$ depending on observations $(Y,X)$, with a response $Y$ and 
-covariate(s) $X$, and a $k$-dimensional vector of parameters $\theta$ is minimized to obtain the 
-parameter estimate $\hat{\theta}$ for the considered parametric model fitted in the current node of 
-the tree. For this purpose many different M-type estimators can be applied including maximum likelihood 
-or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
-algorithm can be employed as objective function. 
-To obtain additional information on the goodness of fit of the parametric model so~-~called scores are
-calculated. Scores are the values of the derivative of the objective function with respect to the parameters 
-with the estimated parameters and the observed values plugged in:
-$$
-S(y, x, \hat{\theta}) = \frac{\partial \psi(y, x, \hat{\theta})}{\partial \theta}
-$$
-which is a $n \times k$ - matrix for a $k$-dimensional parameter vector $\theta$ and
-a data set of $n$ observations containing the response vector $y = (y_1,\ldots,y_n)^\top$
-and regressor(s) $x = (x_1,\ldots,x_n)^\top$. 
-In that way a score value indicating how well the model fits the data is given separately
-for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
-and $l \in \{1,\ldots,k\}$. 
-After calculating these scores the aim is to find out whether there is any instability in the parameters
-over each particular ordering induced by the possible partitioning variables. If a covariate has
-no specific influence on the parameters the scores would always fluctuate randomly around their 
-mean 0. Therefore, these deviations are investigated by measuring them using an empirical fluctuation 
-process. Applying a scalar function which captures the fluctuation in this process yields the
-test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
-the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
-In this parametric unconditional framework p~-~values can be calculated and the covariate
-with the smallest p~-~value is selected as split variable. Then an exhaustive search over all possible
-partitions can be applied to find the best split point. Hence, the number of possible breakpoints is
-not fixed unlike in the CTree algorithm which can make this operation computationally expensive.
-
-\subsubsection{GUIDE}\label{sec:guide}
+and is hence purely intended for a model-based approach. The testing strategy for selecting a split
+variable is based on a fitted model, i.e. on an estimated parameter (vector) $\hat{\theta}$ for each
+node of the tree, and on the application of the Central Limit Theorem. However, the fitted parametric 
+models can be of a very basic and simple form such as an average value. The algorithm only requires the 
+corresponding model scores $s(Y,X,\hat{\theta})$ in order to test for any instability in the model 
+parameters over each particular ordering induced by each of the possible partitioning variables $Z$.
+If a variable $Z$ has no specific influence on the parameters the scores would always fluctuate randomly
+around their mean 0. Therefore, these deviations are investigated by measuring them using an 
+empirical fluctuation process. Applying a scalar function which captures the fluctuation in this process 
+yields the test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
+the resulting test statistic asymptotic theory, in particular the Central Limit Theorem is applied.
+In this parametric unconditional framework $p$-values can be calculated and the covariate
+with the smallest $p$-value is selected as split variable. 
+For $Z$ being a numeric variable a maximally selected split point search is employed while for a 
+categorical variable all possible splits are evaluated to find the best one.
+Hence, the number of possible breakpoints is not fixed unlike in the CTree algorithm which can make 
+this operation computationally expensive.
+
+\subsection{GUIDE}\label{sec:guide}
 Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
-From the model fitted to the data in the current node residuals are calculated. But before applying
-a statistical test two additional steps are performed (to prepare(?) the data):
+For the statistical test applied to select a split variable in each node it requires some type of 
+residuals $r(Y,X,\hat{\theta})$ from the models such as for example 
+$r(Y,X,\hat{\theta}) = Y - X^{\top}\hat{\theta}$. 
+Before applying a statistical test two additional steps are performed:
 The residuals are binarized, i.e. depending on whether the sign of a residual is 
@@
 separated by the corresponding 1st, 2nd and 3rd quartile.
-Then, a $\chi^2$ test of independence is performed for the binarized residuals and
+Then, a $\chi^2$-test of independence is performed for the binarized residuals and
 each categorized split variable.
-After choosing the split variable showing the highest dependency by yielding the lowest p~-~value,
+After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
 the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
@@
 method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
-independence tests such that no split is performed if the lowest p~-~value is not smaller than 
-the significance level.
+independence tests such that no split is performed if the lowest $p$-value is not smaller than 
+the significance level. The impact of the choice of pruning strategy is further discussed in 
+Section~\ref{sec:discussion}.
 
@@
 
-\subsubsection{Residuals vs full model scores}
-Following the basic tree building strategy in Section~\ref{sec:urp}, a $n \times k$ - matrix is 
-obtained based on the discrepancy measure. In the here discussed unifying framework this matrix 
-is the score matrix $S$ as presented in Section~\ref{sec:mob} providing information on the 
-goodness of fit for each observation and model parameter.
-Since the first column of the score matrix is the vector of residuals the decision of using either 
-residuals or full model scores for the test statistic is equivalent to the decision of using only 
-the first column or the whole matrix $S$.
-
-By including all available scores the goodness of fit of the model is investigated regarding each 
-parameter rather than just regarding the mean/location parameter. In that way changes in parameters
-other than the mean/location parameter should be detected more easily.
-
-\textit{Example LM:} To exemplify this feature a linear model tree is considered as this allows 
-for an easy and well structured description of the investigated subjects/issues. Therefore, from
-here on this example will be used for illustrations.
-(However, the conclusions drawn for this example also apply for a wide range of other model
-types.)
-The models fitted in the nodes of the considered linear tree model are of the form
-$$
-Y = \beta_0 + \beta_1 \cdot X + \epsilon
-$$
-for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
-The coefficient vector $\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$ is estimated
-via minimizing the residual sum of squares (RSS) for the learning data set in 
-the current node, for notational simplicity generally denoted by 
-$\{(y_i, x_i)\}_{i=1\dots n}$. Hence, the discrepancy measure, i.e. the objective function is
-$$
-RSS(y_i, x_i, \boldsymbol{\beta}) = \sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2
-$$ 
-such that the corresponding score matrix is defined by
-$$
-s(y_i, x_i, \hat{\boldsymbol{\beta}}) = \frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
-$$
-which is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is two dimensional. 
-(In that way a score value indicating how well the model fits the data is given separately
-for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
-and $j \in \{1,2\}$.)
-In this example considering only residuals corresponds to including the derivatives of $RSS$
-only regarding the intercept parameter $\beta_0$, i.e. the first column of $S$.
-On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
-the derivatives of $RSS$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
-$\beta_1$ (second column).
-
-\subsubsection{Binarization of scores}
+\subsection{Switches}
+\begin{itemize}
+\item  Transformations of Y:\\
+The selection of a split variable is based on a statistical test investigating the dependency of the
+response $Y$ on each possible split variables $Z$. However, before evaluating such a test a model-based
+transformation of $Y$ is applied to obtain a discrepancy measure. This can be residuals 
+$r(Y,X,\hat{\theta})$ as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm
+or, more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
+Since the first column of the score matrix $S = s(Y,X,\hat{\theta})$ is the vector of residuals the 
+GUIDE algorithm selects only this first column while the MOB algorithm uses all available score values,
+hence the full model matrix $S$.
+%providing information on the goodness of fit for each observation and model parameter.
+(By including all available scores the goodness of fit of the model is investigated regarding 
+each parameter rather than just regarding the mean/location parameter. In that way changes in 
+parameters other than the mean/location parameter should be detected more easily.)
+
+\item Binarization of scores:\\
 Rather than applying the independence test directly on the selected (columns of) scores a
 binarized version of them can be used as well. In particular, the score values are binarized
-at 0 such that only their signs are considered.
-
-\subsubsection{Categorization of covariates}
-Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd and 
-3rd quartile are used to bin each possible split variable into four categories.
-
+at $0$ such that only their signs are considered.
+
+\item Categorization of covariates:\\
+Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
+and 3rd quartile are used to bin each possible split variable into four categories. In that way
+a $\chi^2$-test can be applied and the quartiles are the only possible split points while otherwise maximally selected splits over $Z$ are considered.
+
+\end{itemize}
 %\subsubsection{Conditional vs unconditional}
@@
 \hline
-CTree & full model scores & off           & off             & conditional\\
-MOB   & full model scores & off           & off             & unconditional\\
-GUIDE & residuals         & on            & on              & unconditional\\
+CTree & full model scores & -           & -             & conditional\\
+MOB   & full model scores & -           & -             & unconditional\\
+GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
 \hline
@@
 
+\subsection{Linear model tree}
+As an example a linear model tree is considered as this allows 
+for an easy and well structured description of the investigated subjects/issues. 
+For that reason, this model is also considered for the simulation study in Section~\ref{sec:simulation}.
+However, the conclusions drawn for this example also apply for a wide range of other model
+types. The models fitted in the nodes of the considered linear tree model are of the form
+$$
+Y = \beta_0 + \beta_1 \cdot X + \epsilon
+$$
+for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
+As for each node the type of the fitted model is the same no node-specific features are considered
+and the observations of the current (arbitrarily chosen) node are denoted by 
+$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.
+
+A very basic form of a discrepency measure can be obtained by simply calculating deviations of the 
+response values from the mean value within the node, for example by $\sum_{i=1}^n (\bar{y} - y_i)^2$.
+However, this includes only the response variable $Y$ but not the fitted model.
+For a model-based approach residual can be considered such as in the residual sum of squares (RSS)
+$$
+r(y,x,\boldsymbol{\beta}) = RSS(y, x, \boldsymbol{\beta}) = 
+\sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2.
+$$ 
+which is also the objective function that is minimized for the estimation of the coefficient vector 
+$\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$, 
+i.e. to obtain $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
+The corresponding scores are defined by
+$$
+s(y_i, x_i, \hat{\boldsymbol{\beta}}) = 
+\frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
+$$
+such that the score matrix $S$ is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is 
+two dimensional. In that way a score value indicating how well the model fits the data is given separately
+for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
+and $j \in \{1,2\}$.
+
+In this example considering only residuals, i.e. only the first column of the score matrix $S$,
+corresponds to including the derivatives of $RSS$ only regarding the intercept parameter~$\beta_0$.
+On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
+the derivatives of $RSS$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
+$\beta_1$ (second column).
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%All three algorithms apply statistical tests yielding p~-~values based on which
+%All three algorithms apply statistical tests yielding $p$-values based on which
 %the most appropriate split variable is chosen. One main difference lies in
@@
 
-In the second scenario the 'true' tree structure contains two splits in two different variables
+In the second scenario the true tree structure contains two splits in two different variables
 yielding a tree with three terminal nodes. After having investigated the characteristics of 
@@
 The second, more complex scenario employs the same basic structure simply adding another split
-as illustrated in Figure~\ref{fig:true_tree}.
+as illustrated in Figure~\ref{fig:dgp_tree}.
 
@@
 \begin{center}
-\caption[Table caption text]{Variables included in the data generating process.}
+\caption[Table caption text]{Variables included in the data generating process of 
+the first scenario (stump).}
 \label{tab:variables}
 \begin{tabular}{l l l}
+\hline
 Name                    & Notation                   & Specification \\
 \hline
-response                & $Y$                        & $\mathcal{N}(\mu(X, Z_1), \sigma^2)$ \\
-mean of $Y$             & $\mu$                      & linear function of $X$ and $Z_1$ \\
-variance of $Y$         & $\sigma^2$                 & 1\\
-regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\ %or $(-1)^{\mathcal{B}(0.5)}$ \\
-true split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
-noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
-coefficients            & $\beta_0$, $\beta_1$       & functions of $Z_1$\\
-true split point        & $\xi$                      & $\in \{0, 0.2, 0.5, 0.8\}$\\
-effect size             & $\delta$                   & $\in \{0, 0.25, 0.5, 0.75, 1\}$
+\textit{Variables:}     &                            & \\ 
+Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
+Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
+Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
+True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
+Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
+\hline
+\textit{Parameters/functions:} &            & \\ 
+Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
+Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
+True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
+Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
+\hline
 \end{tabular}
@@
 Moreover, the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation
-(can) depend on the true split variable $Z_1$ in the following way:
-\begin{center}
-$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
-\end{center}
-
+(can) depend on the true split variable $Z_1$. 
+%in the following way:
+%\begin{center}
+%$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
+%\end{center}
+%
 In particular, three different cases are considered for the coefficients $\beta_0$ 
@@
 The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, are binary taking either the positive or
-the negative value of the effectsize, i.e. $\delta$ or $-\delta$.
+the negative value of the effect size, i.e. $\delta$ or $-\delta$.
 % can either be binary (taking either $\delta$ or $-\delta$), or a continuous function. 
@@
 %for binary coefficients and a continuous regressor 
-can be found in Figure~\ref{fig:dgp}.
+can be found in Figure~\ref{fig:dgp_stump}.
 
 %\newpage
-\begin{figure}%[h!]
-\setkeys{Gin}{width=1.0\linewidth}
-<<dgp_bin_beta_cont_reg, fig=TRUE, echo=FALSE>>=
+<<dgp, echo=FALSE>>=
 set.seed(7)
@@
 
-par(mfrow = c(3,3))
-
 d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
@@
 
+sp_xi1 <- partysplit(3L, breaks = 0)
+
+pn <- partynode(1L, split = sp_xi1, 
+                kids = list(
+                  #partynode(2L, info = c("true parameters:", 
+                  #                       " ",
+                  #                       expression(beta[0] == 0~" , "~beta[1] == -delta),
+                  #                       "or",
+                  #                       expression(beta[0] == +delta~" , "~beta[1] == 1),
+                  #                       "or",
+                  #                       expression(beta[0] == +delta~" , "~beta[1] == -delta))),
+                  #partynode(3L, info = c("true parameters:", 
+                  #                       " ",
+                  #                       expression(beta[0] == 0~" , "~beta[1] == +delta),
+                  #                       "or",
+                  #                       expression(beta[0] == -delta~" , "~beta[1] == 1),
+                  #                       "or",
+                  #                      expression(beta[0] == -delta~" , "~beta[1] == +delta)))))
+
+                  partynode(2L, info = c("true parameters:", 
+                                         expression(beta[0] ==~" "~0~" or "~-delta), 
+                                         expression(beta[1] ==~" "~1~" or "~+delta))),
+                  partynode(3L, info = c("true parameters:", 
+                                         expression(beta[0] ==~" "~0~" or "~+delta), 
+                                         expression(beta[1] ==~" "~1~" or "~-delta)))))
+
+p <- party(pn, d)
+
+#par(mfrow = c(2,2))
+#par(mar=c(1,1,1,1), oma=c(1,1,1,1))
+paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
+
+ep <- function (obj, digits = 3, abbreviate = FALSE, justmin = Inf, 
+    just = c("alternate", "increasing", "decreasing", "equal"), 
+    fill = "white") 
+{
+  meta <- obj$data
+  justfun <- function(i, split) {
+    myjust <- if (mean(nchar(split)) > justmin) {
+      match.arg(just, c("alternate", "increasing", "decreasing", "equal"))
+    } else {
+      "equal"
+    }
+    k <- length(split)
+    rval <- switch(myjust, equal = rep.int(0, k), 
+                   alternate = rep(c(0.5, -0.5), length.out = k), 
+                   increasing = seq(from = -k/2, to = k/2, by = 1), 
+                   decreasing = seq(from = k/2, to = -k/2, by = -1))
+    unit(0.5, "npc") + unit(rval[i], "lines")
+  }
+  function(node, i) {
+    split <-  c("<= xi", "> xi")
+    y <- justfun(i, split)
+    split <- split[i]
+    if (any(grep(">", split) > 0) | any(grep("<", split) > 0)) {
+      tr <- suppressWarnings(try(parse(text = paste("phantom(0)", split)), silent = TRUE))
+      if (!inherits(tr, "try-error")) split <- tr
+    }
+    grid.rect(y = y, gp = gpar(fill = fill, col = 0), width = unit(1, "strwidth", split))
+    grid.text(split, y = y, just = "center")
+  }
+}
+class(ep) <- "grapcon_generator"
+@
+
+\begin{figure}%[h!]
+\setkeys{Gin}{width=1\linewidth}
+\minipage{0.45\textwidth}
+<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=4>>=
+#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
+plot(p, 
+     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]),
+     drop_terminal = TRUE, tnex = 1, edge_panel = ep)
+@
+\endminipage
+\minipage{0.45\textwidth}
+<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=4>>=
 par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("vary"~beta[0]),
+plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
+     main = expression("varying"~beta[0]),
      pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
-plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1))
-
+abline(coef = c(-1,1), col = "blue", lwd = 2)
+abline(coef = c(1,1), col = "red", lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+@
+\endminipage
+
+\minipage{0.45\textwidth}
+<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=4>>=
 d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
@@
 par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("vary"~beta[1]),
+plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
      pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
-plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1))
-
+abline(coef = c(0,1), col = "blue", lwd = 2)
+abline(coef = c(0,-1), col = "red", lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+@
+\endminipage
+\minipage{0.45\textwidth}
+<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=4>>=
 d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
@@
 par(mar = c(4,4,2,1))
-plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x",
-     main = expression("vary"~beta[0]~"and"~beta[1]),
+plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
+     main = expression("varying"~beta[0]~"and"~beta[1]),
      pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
-plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
-@
-\caption{\label{fig:dgp}{Generated data set of 250 observations for binary coefficients 
-$\beta_0$ and $\beta_1$, continuous regressor $X$, variance parameter $\sigma = 0.5$, 
-effect size $\delta = 1$ and split point $\xi = 0$. 
-Left panels: response $Y$ on regressor $X$; middle panels: intercept $\beta_0$ on split 
-variable $Z_1$; middle panels: slope coefficients $\beta_1$ on split variable $Z_1$; each 
-for varying $\beta_0$ (top), varying $\beta_1$ (middle) and both coefficients varying 
-(bottom).}}
+abline(coef = c(-1,1), col = "blue", lwd = 2)
+abline(coef = c(1,-1), col = "red", lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha("red", 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+@
+\endminipage
+\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$ and $\beta_1$, either varying or being fixed at $0$ or $1$.
+Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
+with effect size $\delta = 1$ and for three scenarios: 
+varying $\beta_0$ and fixed $\beta_1$ (top right),
+fixed $\beta_0$ and varying $\beta_1$ (bottom left),
+varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
 \end{figure}
@@
 with an additional split in $Z_2$ resulting in a tree with three terminal nodes. The first
-split (in $Z_1$, at $\xi$) induces a change in the slope parameter $\beta_1$ while the second
-split (in$Z_1$, also at $\xi$) corresponds to a change in the intercept $\beta_0$.
+split (in $Z_2$, at~$\xi$) induces a change in the slope parameter $\beta_1$ while the second
+split (in $Z_1$, also at~$\xi$) corresponds to a change in the intercept $\beta_0$.
 The underlying tree structure for the data generating process is illustrated in 
-Figures~\ref{fig:true_tree} and \ref{fig:true_tree_biv}.
+Figure~\ref{fig:dgp_tree}.
 
 <<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=
-d <- dgp_tree(nobs = 250)
+
+# d <- dgp_tree(nobs = 250)
+set.seed(7)
+d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
+              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.5)
 
@@
                                          expression(beta[0] == 0), 
-                                         expression(beta[1] == delta))),
+                                         expression(beta[1] == +delta))),
                   partynode(3L, split = sp_xi1, 
@@
                               partynode(5L, info = c("true parameters:", 
-                                                     expression(beta[0] == delta),
-
+                                                     expression(beta[0] == +delta), 
                                                      expression(beta[1] == -delta)))))))
@@
 
-set.seed(747523)
-d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
-              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.001)
-ctest <- evaltests(y~x|z1+z2+z3+z4+z5+z6+z7+z8+z9+z10, data = d, testfun = "ctree", stump = FALSE)
+#ctest <- evaltests(y~x|z1+z2+z3+z4+z5+z6+z7+z8+z9+z10, data = d, testfun = "ctree", stump = FALSE)
 @
@@
 \begin{center}
-\setkeys{Gin}{width=0.6\linewidth}
-<<true_tree, fig=TRUE, echo=FALSE, width=5.4, height=4>>=
+\setkeys{Gin}{width=1\linewidth}
+\minipage{0.5\textwidth}
+<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
 par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
@@
 paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
-plot(p,
+plot(p, edge_panel = ep,
      tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
@@
 @
-\caption{\label{fig:true_tree}True tree structure for split point $\xi = 0$} with the
-true model parameters $\beta_0$ and $\beta_1$ for each terminal node.
+\endminipage
+\minipage{0.45\textwidth}
+<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
+par(mar = c(4,4,2,1))
+plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
+     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
+abline(coef = c(0,1), col = "blue", lwd = 2)
+abline(coef = c(-1,-1), col = "red", lwd = 2)
+abline(coef = c(1,-1), col = "green", lwd = 2)
+points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
+       col = alpha("red", 0.4), cex = 0.7)
+points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
+       col = alpha("green", 0.4), cex = 0.7)
+legend("top", c(expression(z[2]<=xi), 
+                expression(z[2]>xi~" and "~z[1]<=xi),
+                expression(z[2]>xi~" and "~z[1]>xi)), 
+       col = c("blue", "red", "green"), pch = c(19), bty = "n", cex = 0.9)
+@
+\endminipage
+\caption{\label{fig:dgp_tree}Left panel: True tree structure with the true model parameters 
+$\beta_0$ and $\beta_1$ for each terminal panel.
+Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
+with effect size $\delta = 1$.}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\setkeys{Gin}{width=0.6\linewidth}
-<<true_tree_biv, fig=TRUE, echo=FALSE, width=5.4, height=4>>=
-plot(ctest, terminal_panel = node_bivplot, drop_terminal = TRUE,
-     tp_args = list(margins = c(0,1,0,1), pointcol = "white", pointcex = 0.7, lwd = 2),
-     ip_args = list(pval = FALSE, fill = paltrees[c(2, 3)]),
-     ep_args = list(digits = 1))
-@
-\caption{\label{fig:true_tree_biv}True tree structure for split point $\xi = 0$ and 
-effect size $\delta = 1$} with the corresponding linear functions being plotted
-in each terminal node.
-\end{center}
-\end{figure}
+%\begin{figure}[h!]
+%\begin{center}
+%\setkeys{Gin}{width=0.6\linewidth}
+%<<true_tree_biv, fig=TRUE, echo=FALSE, width=5.4, height=4>>=
+%plot(ctest, terminal_panel = node_bivplot, drop_terminal = TRUE, edge_panel = ep,
+%     tp_args = list(margins = c(0,1,0,1), pointcol = "white", pointcex = 0.7, lwd = 2),
+%     ip_args = list(pval = FALSE, fill = paltrees[c(2, 3)]))
+%@
+%\caption{\label{fig:true_tree_biv}True tree structure for split point $\xi = 0$ and 
+%effect size $\delta = 1$} with the corresponding linear functions being plotted
+%in each terminal node.
+%\end{center}
+%\end{figure}
 
@@
 switches is listed for the algorithms CTree, MOB and GUIDE in their original form.
-However, in order tobe able to focus on particular features rather than the whole algorithms
+However, in order to be able to focus on particular features rather than the whole algorithms
 other combinations as presented in Table~\ref{tab:switches}, hence variations of the original 
@@
 two questions at once: (1) Should a split be performed at all? (2) If so, in which variable?
-For the first question the p~-~values regarding all available covariates are compared 
-to a predefined level of significance $\alpha$. Only if the smallest p~-~value is smaller than
-$\alpha$ a split is performed and the covariate corresponding to this p~-~value is selected.
-Including this criterion of significance in the building process of a tree is often refereed
-to as \textit{pre-pruning}, however, this is not always applied in practice as significance
-is often not considered. As mentioned before, the GUIDE algorithm usually builds large 
-trees which are then pruned afterwards. However, for the purpose of comparing the testing
-strategies rather than the whole tree algorithms in this simulation study a significance
-level is considered for all statistical tests, including the $\xi^2$-test in the GUIDE algorithm
-such that all algorithms apply pre-pruning rather than post-pruning.
+For the first question the $p$-values regarding all available covariates are compared 
+to a predefined level of significance $\alpha$. Only if the smallest $p$-value is smaller
+than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is 
+selected. Including this criterion of significance in the building process of a tree is often
+referred to as \textit{pre-pruning}, however, this is not always applied in practice as 
+significance is often not considered. As mentioned before, the GUIDE algorithm usually builds
+large trees which are then pruned afterwards. However, for the purpose of comparing the
+testing strategies rather than the whole tree algorithms in this simulation study a 
+significance level is considered for all statistical tests, including the $\xi^2$-test in 
+the GUIDE algorithm such that all algorithms apply pre-pruning rather than post-pruning.
 
@@
 of data sets for which the true splitting variable $Z_1$ is detected and also selected 
-(for a given significance level $\alpha = 0.05$) is considered for the first scenario (stump).
+(for a given significance level $\alpha = 0.05$) is considered for the first scenario 
+(stump).
 For the second scenario (tree with 2 splits) the adjusted Rand index (ARI) is calculated 
@@
 & Scores            & Binarization  & Categorization  & Null distribution \\ \hline
-CTree       & full model scores & off           & off             & conditional\\
-MOB         & full model scores & off           & off             & unconditional\\
-GUIDE       & residuals         & on            & on              & unconditional\\
-GUIDE\_full & full model scores & on            & on              & unconditional\\ \hline
+CTree       & full model scores & -           & -               & conditional\\
+MOB         & full model scores & -           & -               & unconditional\\
+GUIDE       & residuals         & \checkmark  & \checkmark      & unconditional\\
+GUIDE\_full & full model scores & \checkmark  & \checkmark      & unconditional\\ \hline
 \end{tabular}
@@
          only_intercept == FALSE & xi == 0, 
-       type = "b", lty = c(1,1,1,1),
+       type = "l", lty = c(1,1,1,1),
        key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,1,1,1), type = "b", pch = 1, cex = 0.8,
+                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                 col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
@@
 
-\subsection{Categorization of splitting variables}
-\label{sec:categorization}
-To find out whether a categorization of the possible splitting variables leads to 
-an improvement or a deterioration of performance the two algorithms CTree and MOB are 
-applied, once in their original version without a categorization and once in an adapted 
-version with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
-Moreover, they are compared to GUIDE\_full which includes all available scores.
+\subsection{Binarization of scores}
+\label{sec:binarization}
+To find out whether a binarization of the scores at 0 leads to an improvement or a deterioration
+of performance the two algorithms CTree and MOB are applied, once in their original version
+without a binarization and once in an adapted version with a binarization of scores at 0 
+(CTree\_bin, MOB\_bin). 
+Moreover, they are compared to the adapted GUIDE version which includes all available 
+scores (GUIDE\_full).
 
 Hence, the five strategies are obtained from the unifying framework by setting the switches 
-as listed in Table~\ref{tab:switches_cat}.
+as listed in Table~\ref{tab:switches_bin}.
 
@@
 \centering
-\caption[Table caption text]{CTree, CTree\_cat, MOB, MOB\_cat, and GUIDE\_full with the 
+\caption[Table caption text]{CTree, CTree\_bin, MOB, MOB\_bin, and GUIDE\_full with the 
 corresponding setting of the switches in the unifying framework and the approximation type 
 of the underlying null distribution.}
-\label{tab:switches_cat}
+\label{tab:switches_bin}
 \begin{tabular}{ l c c c c }
@@
 \hline
-CTree       & full model scores & off           & off             & conditional\\
-CTree\_cat  & full model scores & off           & on              & conditional\\
-MOB         & full model scores & off           & off             & unconditional\\
-MOB\_cat    & full model scores & off           & on              & unconditional\\
-GUIDE\_full & full model scores & on            & on              & unconditional\\
+CTree       & full model scores & -           & -             & conditional\\
+CTree\_bin  & full model scores & \checkmark  & -             & conditional\\
+MOB         & full model scores & -           & -             & unconditional\\
+MOB\_bin    & full model scores & \checkmark  & -             & unconditional\\
+GUIDE\_full & full model scores & \checkmark  & \checkmark    & unconditional\\
 \hline
@@
 
-In Figure~\ref{fig:cat_stump} the effect of categorizing split variables on the performance 
-is illustrated for four different values of the true split point $\xi$. 
-For both, CTree and MOB, it can be stated that overall they perform better in their
-original form without any categorization. Only if the true split point $\xi$
-is close to the quartiles used as breakpoints for the categorizations both versions
-lead to a similar proportion of cases where the true splitting variable is detected
-(e.g., for CTree and $\xi=0.5$)
-
-Therefore, it can be summarized that a categorization of the values of the split variable
-does not lead to any advantages unless the true split point corresponds to one of the 
-quartiles used for the categorization. In most situations it even causes the performance to 
-get worse.
-
+Figure~\ref{fig:bin_stump} shows the effect of binarizing the score values over
+different values for the true split point $\xi$, however, all four situations
+lead to the same conclusions. Binarizing the score values decreases the proportion
+of cases where the split variable was selected correctly.
+
+%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
 \begin{figure}%[h!]
 \setkeys{Gin}{width=\linewidth}
-<<cat_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
-                                         #"ctree_max", "ctree_max_cat",
-                                         "mfluc", "mfluc_cat",
+<<bin_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
+                                         #"ctree_max", "ctree_max_bin",
+                                         "mfluc", "mfluc_bin",
                                          "guide_sum_12"))
+
+subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
 subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "ctree_cat", 
-                                  #"ctree_max","ctree_max_cat",
-                                  "mfluc", "mfluc_cat",
+                       levels = c("ctree", "ctree_bin", 
+                                  #"ctree_max", "ctree_max_bin",
+                                  "mfluc", "mfluc_bin",
                                   "guide_sum_12"),
-                       labels = c("CTree","CTree_cat",
-                                  #"CTree_max","CTree_max_cat",
-                                  "MOB", "MOB_cat",
+                       labels = c("CTree","CTree_bin",
+                                  #"CTree_max","CTree_max_bin",
+                                  "MOB", "MOB_bin",
                                   "GUIDE_full"))
@@
        subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & vary_beta == "beta1", 
+         only_intercept == FALSE & vary_beta == "all", 
        type = "b",
@@
                                         1,2,1), 
-                                type = "l", pch = 1, cex = 0.8,
+                                type = "b", pch = 1, cex = 0.8,
                                 col = pal[c("CTree", "CTree", 
@@
 @
-\caption{\label{fig:cat_stump}Proportion of cases where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ for a
-varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 data sets per 
-step. Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ 
-and $0.8$ from left to right). The performances of five testing strategies are compared: 
-CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_full.}
-%\end{figure}
-%\begin{figure}[h!]
+\caption{\label{fig:bin_stump}Proportion of cases where the true split variable $Z_1$ is
+detected over increasing effect size $\delta$ for both parameter varying, $\sigma = 1$, 
+250 observations and 100 data sets per step. Each panel represents a different value of the 
+split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
+strategies are compared: CTree, CTree\_bin, MOB, MOB\_bin, GUIDE\_full.}
+\end{figure}
+
+\subsection{Categorization of splitting variables}
+\label{sec:categorization}
+To find out whether a categorization of the possible splitting variables leads to 
+an improvement or a deterioration of performance the two algorithms CTree and MOB are 
+applied, once in their original version without a categorization and once in an adapted 
+version with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
+Moreover, they are compared to GUIDE\_full which includes all available scores.
+
+Hence, the five strategies are obtained from the unifying framework by setting the switches 
+as listed in Table~\ref{tab:switches_cat}.
+
+\begin{table}[h!]
+\centering
+\caption[Table caption text]{CTree, CTree\_cat, MOB, MOB\_cat, and GUIDE\_full with the 
+corresponding setting of the switches in the unifying framework and the approximation type 
+of the underlying null distribution.}
+\label{tab:switches_cat}
+\begin{tabular}{ l c c c c }
+\hline 
+& Scores            & Binarization  & Categorization  & Null distribution \\ 
+\hline
+CTree       & full model scores & -           & -               & conditional\\
+CTree\_cat  & full model scores & -           & \checkmark      & conditional\\
+MOB         & full model scores & -           &                 & unconditional\\
+MOB\_cat    & full model scores & -           & \checkmark      & unconditional\\
+GUIDE\_full & full model scores & \checkmark  & \checkmark      & unconditional\\
+\hline
+\end{tabular}
+\end{table}
+
+In Figure~\ref{fig:cat_stump} the effect of categorizing split variables on the performance 
+is illustrated for four different values of the true split point $\xi$. 
+For both, CTree and MOB, it can be stated that overall they perform better in their
+original form without any categorization. Only if the true split point $\xi$
+is close to the quartiles used as breakpoints for the categorizations both versions
+lead to a similar proportion of cases where the true splitting variable is detected
+(e.g., for CTree and $\xi=0.5$)
+
+Therefore, it can be summarized that a categorization of the values of the split variable
+does not lead to any advantages unless the true split point corresponds to one of the 
+quartiles used for the categorization. In most situations it even causes the performance to 
+get worse.
+
+\begin{figure}%[h!]
 \setkeys{Gin}{width=\linewidth}
-<<cat_stump_all, fig=TRUE, echo=FALSE, height=4, width=8>>=
+<<cat_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
 subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
@@
                                          "guide_sum_12"))
+subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
 subdata$test <- factor(subdata$test, 
@@
          only_intercept == FALSE & vary_beta == "all", 
+       #xlim = c(0,0.1,0.5),
        type = "b", 
@@
 @
-\caption{\label{fig:cat_stump_all}Proportion of cases where the true split variable $Z_1$ is
+\caption{\label{fig:cat_stump}Proportion of cases where the true split variable $Z_1$ is
 selected over increasing effect size $\delta$ for both parameters varying, $\sigma = 1$, 
@@
 
-\subsection{Binarization of scores}
-\label{sec:binarization}
-To find out whether a binarization of the scores at 0 leads to an improvement or a deterioration
-of performance the two algorithms CTree and MOB are applied, once in their original version
-without a binarization and once in an adapted version with a binarization of scores at 0 
-(CTree\_bin, MOB\_bin). 
-Moreover, they are compared to the adapted GUIDE version which includes all available 
-scores (GUIDE\_full).
-
-Hence, the five strategies are obtained from the unifying framework by setting the switches 
-as listed in Table~\ref{tab:switches_bin}.
-
-\begin{table}[h!]
-\centering
-\caption[Table caption text]{CTree, CTree\_bin, MOB, MOB\_bin, and GUIDE\_full with the 
-corresponding setting of the switches in the unifying framework and the approximation type 
-of the underlying null distribution.}
-\label{tab:switches_bin}
-\begin{tabular}{ l c c c c }
-\hline 
-& Scores            & Binarization  & Categorization  & Null distribution \\ 
-\hline
-CTree       & full model scores & off           & off             & conditional\\
-CTree\_bin  & full model scores & on            & off             & conditional\\
-MOB         & full model scores & off           & off             & unconditional\\
-MOB\_bin    & full model scores & on            & off             & unconditional\\
-GUIDE\_full & full model scores & on            & on              & unconditional\\
-\hline
-\end{tabular}
-\end{table}
-
-Figure~\ref{fig:bin_stump} shows the effect of binarizing the score varlues over
-different values for the true split point $\xi$, however, all four situations
-lead to the same conclusions. Binarizing the score values decreases the proportion
-of cases where the split variable was selected correctly.
-
-%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
-\begin{figure}%[h!]
-\setkeys{Gin}{width=\linewidth}
-<<bin_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
-                                         #"ctree_max", "ctree_max_bin",
-                                         "mfluc", "mfluc_bin",
-                                         "guide_sum_12"))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "ctree_bin", 
-                                  #"ctree_max", "ctree_max_bin",
-                                  "mfluc", "mfluc_bin",
-                                  "guide_sum_12"),
-                       labels = c("CTree","CTree_bin",
-                                  #"CTree_max","CTree_max_bin",
-                                  "MOB", "MOB_bin",
-                                  "GUIDE_full"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
-                                                         #"CTree_max", "CTree_max", 
-                                                         "MOB", "MOB",
-                                                         "GUIDE_full")], 
-                                             fill = pal[c("CTree", "CTree", 
-                                                          #"CTree_max", "CTree_max", 
-                                                          "MOB", "MOB",
-                                                          "GUIDE_full")]), 
-                     superpose.line = list(col = pal[c("CTree", "CTree", 
-                                                       #"CTree_max", "CTree_max", 
-                                                       "MOB", "MOB",
-                                                       "GUIDE_full")]))
-
-xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & vary_beta == "beta1", 
-       type = "b",
-       lty = c(1,2,#1,2,
-               1,2,1),
-       layout= c(4,1),
-       key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,2,#1,2,
-                                        1,2,1), 
-                                type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "CTree", 
-                                            #"CTree_max", "CTree_max", 
-                                            "MOB", "MOB",
-                                            "GUIDE_full")], lwd = 1)), 
-       #index.cond = list(c(1,2,3,4)),
-       xlab = expression(delta),
-       ylab = expression("Proportion of selecting"~Z[1]),
-       strip = strip.custom(factor.levels = c(expression(xi==0), 
-                                              expression(xi==0.2), 
-                                              expression(xi==0.5), 
-                                              expression(xi==0.8))),
-       par.settings = par.settings)
-
-@
-\caption{\label{fig:bin_stump}Proportion of cases where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ for a 
-varying slope parameter $\beta_1$, $\sigma = 1$, 250 observations and 100 data sets per 
-step. Each panel represents a different value of the split point $\xi$ (0, 0.2, 0.5 and 0.8
-from left to right). The performances of five testing strategies are compared: CTree, CTree\_bin,
-MOB, MOB\_bin, GUIDE\_full.}
-%\end{figure}
-%
-%\begin{figure}[h!]
-\setkeys{Gin}{width=\linewidth}
-<<bin_stump_all, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
-                                         #"ctree_max", "ctree_max_bin",
-                                         "mfluc", "mfluc_bin",
-                                         "guide_sum_12"))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "ctree_bin", 
-                                  #"ctree_max", "ctree_max_bin",
-                                  "mfluc", "mfluc_bin",
-                                  "guide_sum_12"),
-                       labels = c("CTree","CTree_bin",
-                                  #"CTree_max","CTree_max_bin",
-                                  "MOB", "MOB_bin",
-                                  "GUIDE_full"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
-                                                         #"CTree_max", "CTree_max", 
-                                                         "MOB", "MOB",
-                                                         "GUIDE_full")], 
-                                             fill = pal[c("CTree", "CTree", 
-                                                          #"CTree_max", "CTree_max", 
-                                                          "MOB", "MOB",
-                                                          "GUIDE_full")]), 
-                     superpose.line = list(col = pal[c("CTree", "CTree", 
-                                                       #"CTree_max", "CTree_max", 
-                                                       "MOB", "MOB",
-                                                       "GUIDE_full")]))
-
-xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & vary_beta == "all", 
-       type = "b",
-       lty = c(1,2,#1,2,
-               1,2,1),
-       layout= c(4,1),
-       key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,2,#1,2,
-                                        1,2,1), 
-                                type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "CTree", 
-                                            #"CTree_max", "CTree_max", 
-                                            "MOB", "MOB",
-                                            "GUIDE_full")], lwd = 1)), 
-       #index.cond = list(c(1,2,3,4)),
-       xlab = expression(delta),
-       ylab = expression("Proportion of selecting"~Z[1]),
-       strip = strip.custom(factor.levels = c(expression(xi==0), 
-                                              expression(xi==0.2), 
-                                              expression(xi==0.5), 
-                                              expression(xi==0.8))),
-       par.settings = par.settings)
-
-@
-\caption{\label{fig:bin_stump_all}Proportion of cases where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ for both parameter varying, $\sigma = 1$, 
-250 observations and 100 data sets per step. Each panel represents a different value of the 
-split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
-strategies are compared: CTree, CTree\_bin, MOB, MOB\_bin, GUIDE\_full.}
-\end{figure}
-
 \subsection{Tree}
@@
 ($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
-Figures~\ref{fig:true_tree} and \ref{fig:true_tree_biv}.
+Figure~\ref{fig:dgp_tree}.
 
@@
 due to the fact that it doesn't detect the first split in $Z_2$ corresponding to a changing
-slope parameter as it only consideres residuals in its testing strategy.
+slope parameter as it only considers residuals in its testing strategy.
 For a split point farther away from the center such as for $\xi = 0.8$ MOB outperforms all other
@@
        data = subdata, 
-       type = "b",
+       type = "l",
        lty = c(1,1,1,1),
        key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,1,1,1), type = "b", pch = 1, cex = 0.8,
+                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                 col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
@@
 FIX ME: only pre-pruning applied, but same results if significance level is ignored (maybe show one plot).
-Optionally compare GUIDE with prepruning to GUIDE with postpruning on tree structure?
+Optionally compare GUIDE with pre-pruning to GUIDE with post-pruning on tree structure?
 Check implementation of pruning.
@@
 FIX ME: results without loss of generalization
+
+\subsection{Pruning}
+<<pruning_data, echo=FALSE, results=hide>>=
+load("../sim/simres20181227_pruning.rda")
+pr <- simres$res
+## extract data for xyplot comparing ARI and ARI_p
+pr <- pr[,c(1,2,7,8,9,12,15)]
+## extended data set with post-pruning results being represented as separated tests
+prex <- pr[,c(1,2,3,4,6)]
+prex <- rbind(prex,prex)
+
+prex$test <- factor(prex$test,
+                    levels = c("ctree","mfluc",
+                              "guide_sum_12", "guide_sum_1_cor",
+                              "ctree_p","mfluc_p",
+                              "guide_sum_12_p", "guide_sum_1_cor_p"),
+                    labels = c("CTree","MOB",
+                               "GUIDE_full", "GUIDE",
+                               "CTree_p","MOB_p",
+                               "GUIDE_full_p", "GUIDE_p"))
+
+for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
+  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
+  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
+  if(prex$test[i] == "GUIDE_full") prex$test[i] <- "GUIDE_full_p"
+  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
+}
+
+prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
+prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
+
+@
+
+\begin{figure}[h!]
+\setkeys{Gin}{width=\linewidth}
+<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
+xyplot(ari ~ delta | xi, groups = ~ test, 
+       data = prex, 
+       type = "l", lty = c(1,1,1,1,2,2,2,2),
+       key =  list(text = list(levels(prex$test)[1:4]),
+                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+       layout= c(4,1),
+       xlab = expression(delta),
+       ylab = "Adjusted Rand Index",
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings
+)
+@
+\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ for
+250 observations and 100 data sets per step.
+Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ and 
+$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
+MOB, GUIDE, GUIDE\_full.}
+\end{figure}
+
+\begin{figure}[h!]
+\setkeys{Gin}{width=\linewidth}
+<<prune_nrsubgr, fig=TRUE, echo=FALSE, height=4, width=8>>=
+xyplot(nrsubgr ~ delta | xi, groups = ~ test, 
+       data = prex, 
+       type = "l", lty = c(1,1,1,1,2,2,2,2),
+       key =  list(text = list(levels(prex$test)[1:4]),
+                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+       layout= c(4,1),
+       xlab = expression(delta),
+       ylab = "Adjusted Rand Index",
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings
+)
+
+@
+\caption{\label{fig:pruning_nrsubgr}Adjusted Rand Index over increasing effect size $\delta$ for
+250 observations and 100 data sets per step.
+Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ and 
+$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
+MOB, GUIDE, GUIDE\_full. Each of them is once applied with pre-pruning as before and once with 
+post-pruning (dashed lines)}
+\end{figure}
 

### 2019-01-04 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small corrections, mainly in sections 2 and 3; structured discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2401 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9f7d5a8831aafb19fea0a98bde255980387338dd

@@
 source("../guidelike.R")
-source("../tests.R")
+source("../tests_guideprune.R")
 #load("../sim/simres20180422_1step.rda")
@@
 In this paper the focus will be put on this first step, i.e. the task of selecting 
-the right/best split variable. In particular, the approach of the GUIDE algorithm is
+the best split variable. In particular, the approach of the GUIDE algorithm is
 compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
-unifying framework for unbiased model based tree algorithms is presented such that each 
+unifying framework for unbiased model-based tree algorithms is presented such that each 
 of the three algorithms can be obtained by a specific combination of the available building 
@@
 Depending on the setup different effects can be observed for the investigated characteristics.
-However, none of the results contradicts the conclusion that in general including all available 
+However, none of the results contradicts the conclusion that in general using all available 
 model scores is at least as efficient but mostly clearly advantageous over considering only 
@@
 Then, after explaining the investigated testing strategies more in detail in 
-Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for model 
-based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
+Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
+model-based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
 The setting for the simulation study is introduced in Section~\ref{sec:simulation}
@@
 more homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
-have been developed, all following the same general structure:
+have been developed, all following the same general structure starting at the root of the tree
+which is simply the first node containing all available data:
 \begin{enumerate}
@@
 within the range/measurable space of this variable is chosen as split point such that 
-after splitting the data the discrepancy measure is minimal within the resulting subgroups.
-\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criterion
-is reached or the subgroup consists of only one or all equal observations. (In the second
-case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
+after splitting the data into subgroups the discrepancy measure is minimal within the 
+resulting nodes.
+\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
+is reached or the data in the node consists of only one or all equal observations. 
 \end{enumerate}
@@
 (In the simulation study of this paper trees with linear models fitted in each node are 
-investigated with the residual sum of squares (RSS) employed as discrepancy measure. However, 
-the discussed methodology and the presented results also apply for a wide range of other 
-model types.)
+investigated with the residual sum of squares (RSS) employed as discrepancy measure.)
 
@@
 GUIDE first select the split variable and then the split point in a separate step. 
-For the first step of selecting a split variable they all apply statistical (independence) tests
+For the first step of selecting a split variable they all apply statistical tests
 following the same general strategy: 
-Based on the discrepancy measure an $n \times k$ - matrix $D$ is obtained giving information
+Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
 on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
@@
 influence on the goodness of fit of the model.
-% Hinweis: pruning-Strategien
-%Once the split variable is chosen the corresponding split point is then selected such that it 
-%minimizes the discrepancy measure.
 
@@
 For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
-response can be mapped to a unity vector indicating the corresponding category by an indicator
-function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$. Alternatively, the function $h$ can return a location
-and scale parameter of $Y$, e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
+response can be mapped to a unity vector by an indicator function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
+Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
+e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
 If a parametric model is fit to the values of the response variable~$Y$ and regressor variable(s)~$X$
-the corresponding model scores can be obtained by setting $h(Y) = s(Y,X,\hat{\theta})$
-where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function:
+the function $h$ can be a model-based transformation returning the corresponding model scores by 
+setting $h(Y) = s(Y,X,\hat{\theta})$
+where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function
 $$
@@
 $\theta$. Hence, for a data set of $n$ observations containing the response vector 
-$y = (y_1,\ldots,y_n)^\top$ and regressor(s) $x = (x_1,\ldots,x_n)^\top$ and a $k$-dimensional 
-parameter vector $\theta$ the score function $s$ returns a $n \times k$ - matrix.
+$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
+regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
+a $k$-dimensional parameter vector $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_k)^\top$ the 
+score function $s$ returns a $n \times k$-matrix.
 In that way a score value indicating how well the model fits the data is given separately
-for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
+for each tuple of observation and parameter element $(y_i, x_i, \theta_l)$ with $i \in \{1,\dots,n\}$
 and $l \in \{1,\ldots,k\}$. 
-(To estimate $\theta$ many different M-type estimators can be applied including maximum likelihood 
-or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
-algorithm can be employed as objective function.)
+To estimate $\boldsymbol{\theta}$ many different M-type estimators can be applied including 
+maximum likelihood or ordinary least squares. Therefore, these objective functions
+are also possible options to obtain a model-based discrepancy measure in the tree algorithm.
 
@@
 the identity function $g(Z) = Z$. For categorical variables $g$ can also map its values to the 
-corresponding unity vectors by $g(Z) = (0,\ldots,1,\ldots,0)$.
+corresponding unity vectors by an indicator function $g(Z) = (0,\ldots,1,\ldots,0)$.
 
@@
 variable is based on a fitted model, i.e. on an estimated parameter (vector) $\hat{\theta}$ for each
-node of the tree, and on the application of the Central Limit Theorem. However, the fitted parametric 
+node of the tree, and on the application of the central limit theorem. However, the fitted parametric 
 models can be of a very basic and simple form such as an average value. The algorithm only requires the 
@@
 yields the test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
-the resulting test statistic asymptotic theory, in particular the Central Limit Theorem is applied.
+the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
 In this parametric unconditional framework $p$-values can be calculated and the covariate
 with the smallest $p$-value is selected as split variable. 
-For $Z$ being a numeric variable a maximally selected split point search is employed while for a 
-categorical variable all possible splits are evaluated to find the best one.
-Hence, the number of possible breakpoints is not fixed unlike in the CTree algorithm which can make 
-this operation computationally expensive.
+For a categorical variable all possible splits are evaluated to find the best one while for $Z$ being 
+a numeric variable a maximally selected split point search is employed. Hence, the number 
+of possible breakpoints is not fixed unlike in the CTree algorithm which can make this operation 
+computationally expensive.
 
@@
 Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
-For the statistical test applied to select a split variable in each node it requires some type of 
-residuals $r(Y,X,\hat{\theta})$ from the models such as for example 
+For the statistical test applied to select a split variable in each node some type of 
+residuals from the models is required such as for example 
 $r(Y,X,\hat{\theta}) = Y - X^{\top}\hat{\theta}$. 
@@
 independence tests such that no split is performed if the lowest $p$-value is not smaller than 
-the significance level. The impact of the choice of pruning strategy is further discussed in 
-Section~\ref{sec:discussion}.
+the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
+in Section~\ref{sec:simulation} all three algorithms are applied using pre-pruning in order to put 
+the focus on the power of the algorithms in terms of selecting the best split variable. 
+However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.
 
 \section{Unifying Framework}\label{sec:unify}
-As all three algorithms follow the same basic structure their approaches can be presented together
-in a unifying framework for a model based tree algorithm from which each specific algorithm can be 
+Since all three algorithms follow the same basic structure their approaches can be presented together
+in a unifying framework for a model-based tree algorithm from which each specific algorithm can be 
 obtained by switching on or off certain features. By applying such a general framework each of the 
@@
 \subsection{Switches}
+As explained before, the selection of a split variable is based on a statistical test 
+investigating the dependency of the response $Y$ on each possible split variables $Z$.
+The three following switches all refer to the way how the given data is prepared/transformed
+before employing such a statistical test.
 \begin{itemize}
 \item  Transformations of Y:\\
-The selection of a split variable is based on a statistical test investigating the dependency of the
-response $Y$ on each possible split variables $Z$. However, before evaluating such a test a model-based
-transformation of $Y$ is applied to obtain a discrepancy measure. This can be residuals 
-$r(Y,X,\hat{\theta})$ as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm
-or, more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
+A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
+information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\theta})$
+as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm or, 
+more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
 Since the first column of the score matrix $S = s(Y,X,\hat{\theta})$ is the vector of residuals the 
 GUIDE algorithm selects only this first column while the MOB algorithm uses all available score values,
-hence the full model matrix $S$.
+hence the full model score matrix $S$.
 %providing information on the goodness of fit for each observation and model parameter.
@@
 Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
-and 3rd quartile are used to bin each possible split variable into four categories. In that way
-a $\chi^2$-test can be applied and the quartiles are the only possible split points while otherwise maximally selected splits over $Z$ are considered.
+and 3rd quartile are used to bin each possible split variable into four categories. In that way,
+depending on whether the scores are binarized or not, a $\chi^2$-test or 1-way ANOVA can be applied 
+where the quartiles are the only possible split points while otherwise maximally selected splits 
+over $Z$ are considered.
 
 \end{itemize}
-%\subsubsection{Conditional vs unconditional}
+
+\textit{Note:}
+Apart from these three switches, the algorithms differ in their approximation of the underlying
+null distribution of the test statistic. While the permutation tests in the CTree algorithm
+work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional approach.
 
@@
 For that reason, this model is also considered for the simulation study in Section~\ref{sec:simulation}.
-However, the conclusions drawn for this example also apply for a wide range of other model
+However, the conclusions drawn from this example also apply for a wide range of other model
 types. The models fitted in the nodes of the considered linear tree model are of the form
@@
 
-A very basic form of a discrepency measure can be obtained by simply calculating deviations of the 
+A very basic form of a discrepancy measure can be obtained by simply calculating deviations of the 
 response values from the mean value within the node, for example by $\sum_{i=1}^n (\bar{y} - y_i)^2$.
 However, this includes only the response variable $Y$ but not the fitted model.
-For a model-based approach residual can be considered such as in the residual sum of squares (RSS)
+For a model-based approach residuals can be considered such as in the residual sum of squares (RSS)
 $$
 r(y,x,\boldsymbol{\beta}) = RSS(y, x, \boldsymbol{\beta}) = 
-\sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2.
+\sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2
 $$ 
@@
 $$
-such that the score matrix $S$ is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is 
+such that the score matrix $S$ is in this case a $n \times 2$-matrix as $\boldsymbol{\beta}$ is 
 two dimensional. In that way a score value indicating how well the model fits the data is given separately
@@
 The location parameter $\mu$ of the normally distributed response variable $Y$ depends
-linearly on the regressor variable $X$ which is a uniformly distributed variable taking 
-values between $-1$ and $1$.
+linearly on the regressor variable $X$.
+% which is a uniformly distributed variable taking values between $-1$ and $1$.
 % can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
@@
 
-\begin{figure}%[h!]
+\begin{figure}%[t]
 \setkeys{Gin}{width=1\linewidth}
@@
 
-\begin{figure}[h!]
+\begin{figure}%[t]
 \begin{center}
@@
 \endminipage
-\caption{\label{fig:dgp_tree}Left panel: True tree structure with the true model parameters 
-$\beta_0$ and $\beta_1$ for each terminal panel.
+\caption{\label{fig:dgp_tree}Left panel: True tree structure for the second scenario with the 
+true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
 Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
@@
 
-%\begin{figure}[h!]
+%\begin{figure}%[t]
 %\begin{center}
@@
 \subsection{Evaluation}
+\label{sec:evaluation}
 The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
@@
 testing strategies rather than the whole tree algorithms in this simulation study a 
-significance level is considered for all statistical tests, including the $\xi^2$-test in 
+significance level is considered for all statistical tests, including the $\chi^2$-test in 
 the GUIDE algorithm such that all algorithms apply pre-pruning rather than post-pruning.
+The effect of these two pruning strategies are investigated in Section~\ref{sec:discussion}.
 
@@
 (stump).
-For the second scenario (tree with 2 splits) the adjusted Rand index (ARI) is calculated 
-as a measure of similarity between the true tree structure and the model tree.
+For the second scenario (tree with 2 splits) the Adjusted Rand Index (ARI) is calculated 
+as a measure of similarity between the true tree structure and the fitted model tree.
 
@@
 simulation scenario (stump) in Sections~\ref{sec:scores}, \ref{sec:categorization}
-and \ref{sec:binarization}. Then the testing strategies are evaluated on a more
-complex tree structure in Section~\ref{sec:tree}. The corresponding results are shown 
-in the following plots.
+and \ref{sec:binarization}. Then the testing strategies are evaluated on the more
+complex tree structure of the second scenario in Section~\ref{sec:tree}. 
+The corresponding results are shown in the following plots.
 
@@
 
-\begin{table}[h!]
+\begin{table}%[t]
 \centering
@@
 In particular, the original strategy of GUIDE fails to detect any changes in $\beta_1$ 
-while GUIDE\_full performs reasonably well in this scenario.
+while GUIDE\_full performs reasonably well in this scenario. This is due to the fact that
+the GUIDE algorithm only includes residuals, i.e. only the scores regarding the intercept,
+such that changes in the slope parameter are not detected unless the corresponding scores 
+are als included as in the GUIDE\_full algorithm.
 These results clearly illustrate that overall the decision to include all available scores
@@
 \caption{\label{fig:scores_stump}Proportion of cases where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ for $\sigma = 1$, 250 observations and 100 data 
-sets per step. Each panel 
-represents one scenario for the coefficients: varying intercept 
-$\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and both 
-coefficients varying (right panel). The performances of four algorithms are compared: 
+selected over increasing effect size $\delta$ for $\sigma = 1$, $\xi=0$, 250 observations and 
+100 data sets per step. Each panel represents one scenario for the coefficients: 
+varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and 
+both coefficients varying (right panel). The performances of four algorithms are compared: 
 CTree, MOB, GUIDE, GUIDE\_full.}
@@
 \label{sec:binarization}
-To find out whether a binarization of the scores at 0 leads to an improvement or a deterioration
+To find out whether a binarization of the scores at $0$ leads to an improvement or a deterioration
 of performance the two algorithms CTree and MOB are applied, once in their original version
-without a binarization and once in an adapted version with a binarization of scores at 0 
-(CTree\_bin, MOB\_bin). 
-Moreover, they are compared to the adapted GUIDE version which includes all available 
-scores (GUIDE\_full).
-
+without a binarization and once in an adapted version with a binarization of scores at $0$ 
+(CTree\_bin, MOB\_bin). Moreover, they are compared to the adapted GUIDE version which includes 
+all available scores (GUIDE\_full).
 Hence, the five strategies are obtained from the unifying framework by setting the switches 
@@
 %% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
-\begin{figure}%[h!]
+\begin{figure}%[t]
 \setkeys{Gin}{width=\linewidth}
@@
 \label{sec:categorization}
-To find out whether a categorization of the possible splitting variables leads to 
-an improvement or a deterioration of performance the two algorithms CTree and MOB are 
-applied, once in their original version without a categorization and once in an adapted 
-version with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
-Moreover, they are compared to GUIDE\_full which includes all available scores.
-
-Hence, the five strategies are obtained from the unifying framework by setting the switches 
-as listed in Table~\ref{tab:switches_cat}.
-
-\begin{table}[h!]
+To investigate the effect of categorizing the possible splitting variables the two algorithms 
+CTree and MOB are applied, once in their original version without a categorization and once 
+in an adapted version with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
+Moreover, they are also compared to GUIDE\_full which includes all available scores.
+The corresponding setting of the switches in the unifying framework is listed in 
+Table~\ref{tab:switches_cat}.
+
+\begin{table}%[t]
 \centering
@@
 
-In Figure~\ref{fig:cat_stump} the effect of categorizing split variables on the performance 
+In Figure~\ref{fig:cat_stump} the impact of categorizing split variables on the performance 
 is illustrated for four different values of the true split point $\xi$. 
@@
 
-\begin{figure}%[h!]
+\begin{figure}%%[t]
 \setkeys{Gin}{width=\linewidth}
@@
 \label{sec:tree}
-To compare the performance of testing strategies in a more complex scenario they have also
-been evaluated on data sets based on a tree structure with two different split variables 
-($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
-Figure~\ref{fig:dgp_tree}.
+To compare the performance of testing strategies in a more complex scenario they have
+been evaluated on data sets based on the tree structure with two different split variables 
+($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in Figure~\ref{fig:dgp_tree}.
 
 Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
-almost equally well for the true split points $0$, $0.2$ and $0.5$ while GUIDE\_full
-a bit worse but still achieves an ARI above 0.5 for the highest considered effect size
+almost equally well for the true split points $0$, $0.2$ and $0.5$ while GUIDE\_full performs
+a bit worse but still achieves an ARI above $0.5$ for the highest considered effect size
 of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
@@
 methods and both GUIDE versions fail to find the correct tree structure.
-Based on these results it can be concluded once again that including all available score in the
+Based on these results it can be concluded once again that including all available scores in the
 testing strategy leads to a better performance while binarizing scores and categorizing
@@
 
-\begin{figure}[h!]
+\begin{figure}%[t]
 \setkeys{Gin}{width=\linewidth}
@@
 
+\newpage
+
 \section{Discussion}
 \label{sec:discussion}
-FIX ME: only pre-pruning applied, but same results if significance level is ignored (maybe show one plot).
-Optionally compare GUIDE with pre-pruning to GUIDE with post-pruning on tree structure?
-Check implementation of pruning.
-
-FIX ME: LM-specific results
-
-FIX ME: results without loss of generalization
+
+\subsection{Conclusions}
+From comparing the effects of the three switches three main conclusions can be drawn: 
+(1) In most cases including all available score values rather than just residuals leads to 
+clearly better results and performs in general at least equally well.
+(2) Binarizing scores/residuals does not lead to any improvement in performance.
+(3) Categorizing possible split variables also doesn't improve the performance 
+unless the breakpoints used for the categorization are very close to the true split points.
+However, even then the results are very similar to those achieved without any categorization.
+
+\subsection{Significance level}
+In the simulation study presented in Section~\ref{sec:simulation} a significance level is 
+included in all of the algorithms such that a split is only performed if the lowest $p$-value
+is smaller than this significance level. This strategy, usually referred to as pre-pruning, aims at 
+avoiding overfitting.
+However, ignoring this significance level within the tree algorithms yields the same results 
+which is shown in Figure~\ref{fig:scores_stump_wosign} where the proportion of cases 
+in which a split is performed in the true splitting variable, but not necessarily with a $p$-value 
+smaller than the significance level, is illustrated.
+
+\begin{figure}
+\setkeys{Gin}{width=\linewidth}
+<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
+                                         "guide_sum_12", "guide_sum_1_cor"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree","mfluc",
+                                  "guide_sum_12", "guide_sum_1_cor"),
+                       labels = c("CTree","MOB",
+                                  "GUIDE_full", "GUIDE"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
+
+xyplot(prop_z1 ~ delta | vary_beta, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
+         only_intercept == FALSE & xi == 0, 
+       type = "l", lty = c(1,1,1,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+       layout= c(3,1),
+       index.cond = list(c(2,3,1)),
+       xlab = expression(delta),
+       ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
+       strip = strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
+                                              expression("vary"~beta[0]), 
+                                              expression("vary"~beta[1]))),
+       par.settings = par.settings)
+@
+\caption{\label{fig:scores_stump_wosign}Proportion of cases where the true split variable $Z_1$ 
+is selected over increasing effect size $\delta$ for $\sigma = 1$, $\xi=0$, 250 observations and 
+100 data sets per step without any significance level. 
+Each panel represents one scenario for the coefficients: 
+varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and 
+both coefficients varying (right panel). The performances of four algorithms are compared: 
+CTree, MOB, GUIDE, GUIDE\_full.}
+\end{figure}
+
+%FIX ME: Check implementation of pruning.
+
+%FIX ME: LM-specific results
+
+%FIX ME: results without loss of generalization
 
 \subsection{Pruning}
+
+To show that similar results are obtained without any significance level but with the application
+of post-pruning (building very large trees and employing cost-complexity pruning as in the CART algorithm) the models built in the second simulation scenario are built once with a pre-pruning and once
+with a post-pruning approach. The results are compared based on the Adjusted Rand Index in 
+Figure~\ref{fig:pruning_ari}.
+
 <<pruning_data, echo=FALSE, results=hide>>=
@@
        xlab = expression(delta),
-       ylab = "Adjusted Rand Index",
+       ylab = "Nr. of terminal nodes",
        strip = strip.custom(factor.levels = c(expression(xi==0), 
@@
 @
-\caption{\label{fig:pruning_nrsubgr}Adjusted Rand Index over increasing effect size $\delta$ for
+\caption{\label{fig:pruning_nrsubgr}Number of terminal nodes over increasing effect size $\delta$ for
 250 observations and 100 data sets per step.

### 2019-01-05 pkg/partykit/inst/guideliketest/draft/draft.Rnw
now results are recalculated if not already stored
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2402 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 164d038dfb371f0c5a9c548590ebfde74be694b7

@@
 source("../tests_guideprune.R")
-#load("../sim/simres20180422_1step.rda")
-load("../sim/simres20181223_stump.rda")
+
+if(file.exists("stump.rda")){
+  load("stump.rda")
+} else {
+  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
+                         nrsplitvar = 1, nrsteps = 1, nrlevels = 2,
+                         delta = seq(from = 0, to = 1, by = 0.1),
+                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = c("all", "beta0", "beta1"),
+                         binary_regressor = FALSE, binary_beta = TRUE, only_intercept = FALSE,
+                         test = c("ctree", "mfluc", "ctree_max",
+                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
+                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
+                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
+                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
+                         beta0 = 0, beta1 = 1,
+                         stump = TRUE, z1dist = "unif", sigma = 1, alpha = 0.05)
+
+  save(simres, file = "stump.rda")
+}
 sim_stump <- simres$res
-#load("../sim/simres20180422_2step.rda")
-#sim2step <- simres
 rm(simres)
-load("../sim/simres20181223_tree.rda")
+
+if(file.exists("tree.rda")){
+  load("tree.rda")
+} else {
+  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
+                       nrsplitvar = 2, nrsteps = 1, nrlevels = 2,
+                       delta = seq(from = 0, to = 1, by = 0.1),
+                       xi = c(0, 0.2, 0.5, 0.8), vary_beta = "all",
+                       binary_regressor = FALSE, binary_beta = TRUE,
+                       only_intercept = FALSE,
+                       test = c("ctree", "mfluc", "ctree_max",
+                                "ctree_cat", "mfluc_cat", "ctree_max_cat",
+                                "ctree_bin", "mfluc_bin", "ctree_max_bin",
+                                "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
+                                "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
+                       beta0 = NULL, beta1 = NULL,
+                       stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05)
+
+  save(simres, file = "tree.rda")
+}
 sim_tree <- simres$res
@@
 <<pruning_data, echo=FALSE, results=hide>>=
-load("../sim/simres20181227_pruning.rda")
+if(file.exists("pruning.rda")){
+  load("pruning.rda")
+} else {
+  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7, 
+                         nrsteps = 1, nrlevels = 2,
+                         delta = seq(from = 0, to = 1, by = 0.1),
+                         xi = c(0, 0.2, 0.5, 0.8), 
+                         vary_beta = "all", binary_regressor = FALSE,
+                         binary_beta = TRUE, only_intercept = FALSE,
+                         compare_pruning = TRUE,
+                         test = c("ctree", "mfluc",
+                                  "guide_sum_12", 
+                                  "guide_sum_1_cor"),
+                         beta0 = NULL, beta1 = NULL,
+                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05,
+                         nrsplitvar = 2)
+
+  save(simres, file = "pruning.rda")  
+}
+
 pr <- simres$res

### 2019-01-16 pkg/partykit/inst/guideliketest/tests_guideprune.R
further revisions, added references and added function to prepare data set for 3-way facotrial analysis
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2405 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 94e669fcfa7e6c18b775c493b98b53d55df26910

@@ -1866,6 +1866,211 @@ simwrapper_p <- function(nobs = 200, nrep = 100, seed = 7, nrsteps = 1, nrlevels
 
 
 
+### function to prepare data set for 3-way factorial analysis
+prep_3way <- function(simres)
+{
+  pval<- simres$pval
+  # prepare data.frame
+  {
+    pval_T <- pval
+    pval_T[which(simres$sv !="z1")] <- 1
+    
+    head(pval_T)
+    
+    d <- rep(colnames(pval)[1], NROW(pval))                 
+    for(i in 2:NCOL(pval)){
+      d <- c(d, rep(colnames(pval)[i], NROW(pval)))
+    }
+    
+    d <- cbind(d, 
+               rep(0, length(d)), 
+               rep(0, length(d)), 
+               rep(0, length(d)))    
+    
+    colnames(d) <- c("strategy", "res_scores", "bin", "cat")
+    
+    ## set values for factor variables representing the three switches
+    {
+      d[d[,"strategy"] == "ctree", "res_scores"] <- "scores"
+      d[d[,"strategy"] == "ctree", "bin"] <- "lin"
+      d[d[,"strategy"] == "ctree", "cat"] <- "lin"
+      
+      d[d[,"strategy"] == "ctree_max", "res_scores"] <- "scores"
+      d[d[,"strategy"] == "ctree_max", "bin"] <- "lin"
+      d[d[,"strategy"] == "ctree_max", "cat"] <- "max"
+      
+      d[d[,"strategy"] == "mfluc", "res_scores"] <- "scores"
+      d[d[,"strategy"] == "mfluc", "bin"] <- "lin"
+      d[d[,"strategy"] == "mfluc", "cat"] <- "max"
+      
+      d[d[,"strategy"] == "ctree_cat", "res_scores"] <- "scores"
+      d[d[,"strategy"] == "ctree_cat", "bin"] <- "lin"
+      d[d[,"strategy"] == "ctree_cat", "cat"] <- "cat"
+      
+      d[d[,"strategy"] == "ctree_max_cat", "res_scores"] <- "scores"
+      d[d[,"strategy"] == "ctree_max_cat", "bin"] <- "lin"
+      d[d[,"strategy"] == "ctree_max_cat", "cat"] <- "cat"
+      
[…]

### 2019-01-16 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions, added references and added function to prepare data set for 3-way facotrial analysis
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2405 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 94e669fcfa7e6c18b775c493b98b53d55df26910

@@
 \usepackage[all]{hypcap}
+\usepackage{enumitem}
 
@@
 pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
-names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_full")
+names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")
 
@@
 tests (MOB). It is assessed how different building blocks affect the power of the 
-tree algorithms to select the appropriate covariates for splitting: residuals vs. full 
+tree algorithms to select the appropriate covariates for splitting: residuals vs.\ full 
 model scores, binarization of residuals/scores at zero, binning of covariates, 
@@
 in this research field. While they all follow the general idea of splitting the data such 
-that some objective function is minimized/maximized, they differ in their 
-specific/particular approaches to selecting a split variable and the corresponding split point. 
+that some objective function is optimized, they differ in their 
+specific approaches to selecting a split variable and the corresponding split point. 
 Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
@@
 and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
-For the first step of selecting a split variable they all follow the idea of 
-choosing the covariate which shows the highest association to/influence on the response variable 
+For the first step of selecting a split variable they all share the same basic concept of 
+choosing the covariate which shows the highest association to the response variable 
 based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
 tests for contingency tables, CTree applies permutation tests in a conditional inference
-framework and MOB uses fluctuation tests based on asymptotic theory.
-
-In this paper the focus will be put on this first step, i.e. the task of selecting 
-the best split variable. In particular, the approach of the GUIDE algorithm is
-compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
-unifying framework for unbiased model-based tree algorithms is presented such that each 
-of the three algorithms can be obtained by a specific combination of the available building 
-blocks. In that way, the effects of each of these characteristic features of the GUIDE,
-CTree and MOB algorithm can be investigated separately by varying the switches in
-the tool box provided by this general framework. 
-Depending on the setup different effects can be observed for the investigated characteristics.
+framework and MOB uses fluctuation tests based on asymptotic theory. All these approaches
+have shown to work well for various situations, however, the specific advantages and
+disadvantages of each of the testing strategies have not yet been investigated and compared
+in detail.
+
+Therefore, in this paper the focus will be put on that first step of tree algorithms,
+i.e. the task of selecting the best split variable. In particular, the approach of the 
+GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm. 
+For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
+algorithms is presented such that each of the three strategies can be obtained by a specific 
+combination of the available building blocks. In that way, the effects of each of these 
+features can be investigated separately by varying the building blocks in the tool box 
+provided by this general framework.
+
+Depending on the setup different effects can be observed for the investigated building blocks.
 However, none of the results contradicts the conclusion that in general using all available 
@@
 residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
-the values of possible split variables are investigated and the impact of applying a conditional
-or an unconditional approximation of the null distribution is analyzed.
+the values of possible split variables are investigated.
+Moreover, the impact of applying either a pre- or postpruning strategy is analyzed.
 
 In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
-Then, after explaining the investigated testing strategies more in detail in 
+Then, after explaining the investigated algorithms CTree, MOB, and GUIDE more in detail in 
 Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
-model-based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
+testing strategies in model-based unbiased recursive partitioning is presented in 
+Section~\ref{sec:unify}.
 The setting for the simulation study is introduced in Section~\ref{sec:simulation}
@@
 The basic idea of building a regression tree model is to partition the data into smaller and 
-more homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
+more homogeneous subgroups based on a set of covariates. Various different tree algorithms 
 have been developed, all following the same general structure starting at the root of the tree
@@
 measure is calculated. 
-%This can for example be the sum of deviations from a typical or average value of the 
-%data set or it can be based on a fitted model.
 \item Among all available covariates one is selected as split variable and a value 
@@
 The employed discrepancy measure can for example be the sum of deviations from a typical 
-or average value, but it can also be constructed based on a statistical model. For instance,
-different types of residuals or their signs can be applied (see e.g., \cite{Loh:2002}) as well as
-rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
-(In the simulation study of this paper trees with linear models fitted in each node are 
-investigated with the residual sum of squares (RSS) employed as discrepancy measure.)
+or average value, but it can also be constructed based on a parametric model. For instance,
+different types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
+well as rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
+%(In the simulation study of this paper trees with linear models fitted in each node are 
+%investigated with the residual sum of squares (RSS) employed as discrepancy measure.)
 
 As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
-GUIDE first select the split variable and then the split point in a separate step. 
+GUIDE first select the split variable and then in a separate step the split point. 
 For the first step of selecting a split variable they all apply statistical tests
 following the same general strategy: 
-Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
+\begin{enumerate}
+\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
 on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
 current node and $k \in \mathbb{N}$.
-Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
-variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
-of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
-values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
-necessary to calculate the corresponding $p$-values in order to allow for an unbiased
-variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
-$p$-value is selected as split variable as it has shown to be the variable with the strongest
-influence on the goodness of fit of the model.
+\item Using this matrix, the algorithm-specific test statistic is evaluated for each possible 
+split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
+of $Z_j$ on $D$ and in that way the dependency of $Z_j$ on the response $Y$ as the discrepancy
+measure (and therefore also $D$) depends on $Y$. 
+\item For the resulting values of the test statistics the corresponding $p$-values are calculated
+in order to allow for their comparison on a standardized/unified scale and in that way for
+an unbiased variable selection.
+%As the variables are usually not all measured at the same scale the resulting
+%values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
+%necessary to calculate the corresponding $p$-values in order to allow for an unbiased
+%variable selection on this standardized/unified scale. 
+\item The variable corresponding to the smallest $p$-value is selected as split variable as 
+it has shown to be the variable with the strongest influence on the goodness of fit of the model.
+\end{enumerate}
 
@@
 characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
-explained in detail in Sections~\ref{sec:ctree}.
+explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide} .
 
@@
 
-The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
-a conditional inference framework by applying permutation tests. 
+The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
+non-parametric regression tree models in a conditional inference framework by applying 
+permutation tests. 
 To select a split variable it is tested whether there is any association between the transformed 
-response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split variable $Z$.
+response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
+variable $Z$.
 For different types of response variables different functions $h$ which depend on $Y$ in a 
@@
 For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
-response can be mapped to a unity vector by an indicator function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
+response can be mapped to a unity vector by an indicator function 
+$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
 Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
 e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
-If a parametric model is fit to the values of the response variable~$Y$ and regressor variable(s)~$X$
-the function $h$ can be a model-based transformation returning the corresponding model scores by 
-setting $h(Y) = s(Y,X,\hat{\theta})$
-where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function
+If a parametric model is fit to the values of the response variable~$Y$ and regressor 
+variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
+corresponding model scores by setting $h(Y) = s(Y,X,\hat{\beta})$
+where $\hat{\beta}$ is the estimated model parameter and $s$ is the score function
 $$
-s(Y, X, \hat{\theta}) = \frac{\partial \psi(Y, X, \hat{\theta})}{\partial \theta}
+s(Y, X, \hat{\beta}) = \frac{\partial \psi(Y, X, \hat{\beta})}{\partial \beta}
 $$
-with $\psi$ being the objective function which is minimized to estimate the model parameter
-$\theta$. Hence, for a data set of $n$ observations containing the response vector 
-$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
+with the sum of $\psi$ over all observations being the objective function that is optimized 
+to estimate the model parameter $\beta$. For this purpose many different M-type estimators can 
+be applied including maximum likelihood or ordinary least squares. Therefore, these objective 
+functions are also possible options to obtain a model-based discrepancy measure for the tree 
+algorithm.
+Applying the score function $s$ on a data set of $n$ observations containing the 
+response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
 regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
-a $k$-dimensional parameter vector $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_k)^\top$ the 
-score function $s$ returns a $n \times k$-matrix.
+on a $k$-dimensional estimated parameter vector 
+$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
+yields a $n \times k$-matrix $S$.
 In that way a score value indicating how well the model fits the data is given separately
-for each tuple of observation and parameter element $(y_i, x_i, \theta_l)$ with $i \in \{1,\dots,n\}$
+for each tuple of observation and estimated parameter element 
+$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
 and $l \in \{1,\ldots,k\}$.
-To estimate $\boldsymbol{\theta}$ many different M-type estimators can be applied including 
-maximum likelihood or ordinary least squares. Therefore, these objective functions
-are also possible options to obtain a model-based discrepancy measure in the tree algorithm.
 
@@
 Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
-statistic given all permutations of the response variable can be calculated and used to standardize 
-the test statistic. In order to map multivariate test statistics into the real line two options are 
-given: either the maximum of the absolute values of the standardized statistic or a quadratic form 
-is considered. The resulting values can then be transformed to the scale of $p$-values where they 
-can be compared in an unbiased way to select a split variable.
+statistic given all permutations of the response variable can be calculated and used to 
+standardize the test statistic. In order to map multivariate test statistics into the real 
+line two options are given: either the maximum of the absolute values of the standardized 
+statistic or a quadratic form is considered. The resulting values can then be transformed to 
+the scale of $p$-values where they can be compared in an unbiased way to select a split variable.
 
@@
 provides a very general framework where the choice of the type of variables and the corresponding 
-functions $g$ and $h$ determine the type of the applied statistical test.
+functions $g$ and $h$ determine the type of the applied statistical test 
+(see \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox).
 
@@
 \subsection{MOB}\label{sec:mob}
-While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
-was developed with the aim of embedding parametric models into the framework of regression trees
-and is hence purely intended for a model-based approach. The testing strategy for selecting a split
-variable is based on a fitted model, i.e. on an estimated parameter (vector) $\hat{\theta}$ for each
-node of the tree, and on the application of the central limit theorem. However, the fitted parametric 
-models can be of a very basic and simple form such as an average value. The algorithm only requires the 
-corresponding model scores $s(Y,X,\hat{\theta})$ in order to test for any instability in the model 
-parameters over each particular ordering induced by each of the possible partitioning variables $Z$.
-If a variable $Z$ has no specific influence on the parameters the scores would always fluctuate randomly
-around their mean 0. Therefore, these deviations are investigated by measuring them using an 
-empirical fluctuation process. Applying a scalar function which captures the fluctuation in this process 
-yields the test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
-the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
-In this parametric unconditional framework $p$-values can be calculated and the covariate
-with the smallest $p$-value is selected as split variable. 
-For a categorical variable all possible splits are evaluated to find the best one while for $Z$ being 
-a numeric variable a maximally selected split point search is employed. Hence, the number 
-of possible breakpoints is not fixed unlike in the CTree algorithm which can make this operation 
-computationally expensive.
+
+While CTree offers to apply model-based discrepancy measures as one possible choice, the 
+MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was developed with the aim of embedding 
+parametric models into the framework of regression trees and is hence purely intended for 
+a model-based approach.
+
+The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
+an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
+application of the central limit theorem. 
+The fitted model can be of a very basic and simple form such as 
+an average value of the response vector
+%a constant fit where the parameter $\beta$ is an
+%average value of the response, hence 
+%$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
+%$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ 
+but can also be a more complex parametric model.
+In \cite{Zeileis+Hothorn+Hornik:2008} GLMs and survival regression models are used to 
+exemplify the algorithm. Subsequently various other types of models have been applied as well,
+for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
+psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
+mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
+
+Regardless of the applied model type the algorithm only requires the corresponding model 
+score matrix $S$ in order to test for any instability in the model parameters over each 
+particular ordering induced by each of the possible partitioning variables $Z$.
+If a variable $Z$ has no specific influence on the parameters the scores would always 
+fluctuate randomly around their mean 0. Therefore, these deviations are investigated by 
+measuring them using an empirical fluctuation process. Applying a scalar function which 
+captures the fluctuation in this process yields the test statistic of this so called 
+generalized M-fluctuation test \citep{Zeileis+Hornik:2007}. 
+To obtain the distribution of the resulting test statistic asymptotic theory, in particular 
+the central limit theorem is applied. In this parametric unconditional framework $p$-values 
+can be calculated and the covariate with the smallest $p$-value is selected as split variable.
+
+For a categorical variable all possible splits are evaluated to find the best one while 
+for $Z$ being a numeric variable a maximally selected split point search is employed. Hence, 
+the number of possible breakpoints is not fixed unlike in the CTree algorithm.
 
 \subsection{GUIDE}\label{sec:guide}
-Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
-For the statistical test applied to select a split variable in each node some type of 
-residuals from the models is required such as for example 
-$r(Y,X,\hat{\theta}) = Y - X^{\top}\hat{\theta}$. 
+Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
+the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
+a constant fit where the parameter $\beta$ is the mean of the response, hence 
+$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
+$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, can be employed as well as more complex models.
+For the statistical test applied to select a split variable some type of model-based
+residuals is required. In \cite{Loh:2002} linear models are used as an example where the
+residuals are defined by
+$$
+r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
+$$
+In related literature other models together with an appropriate choice of residuals
+have been applied such as 
+regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, %(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
+Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))
+and quantile regression models \citep{Chaudhuri+Loh:2002} among many others.
+
 Before applying a statistical test two additional steps are performed:
-The residuals are binarized, i.e. depending on whether the sign of a residual is 
-positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
-and for each possible split variable the observed values are categorized into 4 groups 
-separated by the corresponding 1st, 2nd and 3rd quartile.
+(1)~The residuals are binarized, i.e. depending on whether the sign of a residual is 
+positive or not it is replaced by $1$ or $2$.  % FIX ME: +1 and -1 in the implementation? 
+(2)~For each possible split variable the observed values are categorized into four groups 
+separated by the corresponding 1st, 2nd and 3rd quartile unless the split variable is already
+categorical.
 Then, a $\chi^2$-test of independence is performed for the binarized residuals and
-each categorized split variable.
+each categorized/categorical split variable.
 After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
 the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
-is determined by the categorization of the possible split variables such that the quartiles are
-the only three possible split points.
-
-\bigskip
-
-\textit{Note:}
-While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
-method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
-independence tests such that no split is performed if the lowest $p$-value is not smaller than 
-the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
-in Section~\ref{sec:simulation} all three algorithms are applied using pre-pruning in order to put 
-the focus on the power of the algorithms in terms of selecting the best split variable. 
-However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.
+is determined by the categorization of the possible split variables and is in that way fixed.
+
+\newpage
+\subsection{Pruning}
+\label{sec:pruning}
+In early developed methods of unbiased recursive partitioning
+such as FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
+CRUISE \citep{Kim+Loh:2001} or GUIDE \citep{Loh:2002}
+the $p$-values of applied significance tests are used purely for the selection of a split variable 
+based on a unified scale. All other steps are performed in the same way as in classic tree 
+algorithms, such as finding the best split point by optimizing an objective function.
+
+In subsequent approaches as in \cite{Hothorn+Hornik+Zeileis:2006}
+it is shown that these $p$-values can also be used as a basis for a
+pruning strategy namely prepruning: Only if the $p$-value corresponding to the selected split
+variable (hence the smallest $p$-value) is smaller than a given significance level a split is
+performed while otherwise no further splits are made in the current node.
+In that way the information provided by the significance test is exploited in a more profound way.
+However, prepruning requires the applied test to perform reasonably well as a significance test, 
+i.e. it has to have power but also avoid overfitting.
+
+For postpruning no significance level is included leading to very large trees which are then 
+pruned afterwards, for example by employing cost-complexity pruning as in the 
+CART algorithm \citep{Breiman+Friedman+Stone:1984}. 
+Thus, for postpruning the test only needs to be able to select the best split variable but does 
+not have to be effective as a significance test.
+
+Originally the GUIDE algorithm was introduced applying postpruning while CTree and MOB
+were both presented employing prepruning (see Table~\ref{tab:combinations} for a more detailed
+overview). Alternatively, for model-based partitioning a pruning strategy can also use model-based 
+information criteria such as AIC or BIC \citep[see for example][]{Su+Wang+Fan:2004}.
+However, all testing strategies can be combined with each of these pruning strategies.
+If a test performs well the choice of a pruning strategy has hardly any impact on the results, 
+but as prepruning only requires already provided information it is usually less computationally
+expensive than the additional calculations needed for postpruning.
+However, if the test performs poorly in terms of power and overfitting, prepruning suffers from 
+this deficiency while postpruning can compensate it at least partly and improve the results.
+For this reason we are focusing on the testing strategies applied in tree algorithms and 
+investigate how effective they are as significance tests.
+
+%\bigskip
+%\textit{Note:}
+%While GUIDE originally constructs large trees and applies postpruning with the cross-validation 
+%method of CART, CTree and MOB employ prepruning by setting a significance level for the 
+%independence tests such that no split is performed if the lowest $p$-value is not smaller than 
+%the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
+%in Section~\ref{sec:simulation} all three algorithms are applied using prepruning in order to put 
+%the focus on the power of the algorithms in terms of selecting the best split variable. 
+%However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.
 
 \section{Unifying Framework}\label{sec:unify}
-Since all three algorithms follow the same basic structure their approaches can be presented together
-in a unifying framework for a model-based tree algorithm from which each specific algorithm can be 
-obtained by switching on or off certain features. By applying such a general framework each of the 
-characteristic features of the three algorithms and its impact can be investigated in detail. 
-These switches are explained in this Section and their setting for the algorithms
-CTree, MOB, and GUIDE are listed in Table~\ref{tab:switches}.
-
-\subsection{Switches}
-As explained before, the selection of a split variable is based on a statistical test 
-investigating the dependency of the response $Y$ on each possible split variables $Z$.
-The three following switches all refer to the way how the given data is prepared/transformed
-before employing such a statistical test.
+Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
+of the type of fitted models, the type of applied tests and the employed pruning strategy
+as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
+original form. However, in principle all other combinations of the listed features are possible 
+as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
+strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
+of parametric models together with prepruning as the properties of this pruning strategy
+are already explained/determined by the testing strategy. Subsequently combinations
+with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
+From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
+testing strategies rather than the complete algorithms.
+
+\begin{table}[t]
+\centering
+\caption[Table caption text]{Combinations of fitted model type, test type and pruning strategy
+for the algorithms CTree, MOB, and GUIDE.}
+\label{tab:combinations}
+\begin{tabular}{ l c c c }
+\hline 
+      & Fit             & Test                     & Pruning \\ 
+\hline
+CTree & non-parametric  & conditional inference    & prepruning\\
+MOB   & parametric      & score-based fluctuation  & prepruning (or AIC/BIC-based pruning)\\
+GUIDE & parametric      & residual-based $\chi^2$  & postpruning (cost-complexity pruning)\\
+\hline
+\end{tabular}
+\end{table}
+
+\subsection{Unification of Testing Strategies}
+
+Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
+Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
+basic structure. Thus they can be presented together in a unifying framework for testing strategies
+in model-based unbiased recursive partitioning. 
+In this framework each specific testing strategy can be obtained by combining the provided 
+building blocks accordingly. 
+This general toolbox covers all features of the here discussed testing strategies CTree, MOB, and
+GUIDE. Furthermore, it allows for all possible combinations of these features and in that way also
+for variations of the commonly known testing strategies. Moreover, by exchanging only one building
+block its impact can be investigated in detail.
+These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
+for CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which gives a more 
+detailed overview of the differences in the second column of Table~\ref{tab:combinations}.
+
+\subsection{Building Blocks of Testing Strategies}\label{sec:buildingblocks}
+The three following building blocks all refer to the way how the given data is prepared/transformed
+before employing a statistical test to select a split variable.
 \begin{itemize}
@@
 A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
-information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\theta})$
-as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm or, 
-more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
-Since the first column of the score matrix $S = s(Y,X,\hat{\theta})$ is the vector of residuals the 
-GUIDE algorithm selects only this first column while the MOB algorithm uses all available score values,
-hence the full model score matrix $S$.
-%providing information on the goodness of fit for each observation and model parameter.
-(By including all available scores the goodness of fit of the model is investigated regarding 
-each parameter rather than just regarding the mean/location parameter. In that way changes in 
-parameters other than the mean/location parameter should be detected more easily.)
-
-\item Binarization of scores:\\
-Rather than applying the independence test directly on the selected (columns of) scores a
-binarized version of them can be used as well. In particular, the score values are binarized
+information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
+as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
+more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
+cases.
+By including all available scores rather than just residuals the goodness of fit of the model is 
+investigated regarding each parameter instead of just the mean/location parameter. 
+Therefore, this facilitates the detection of changes in parameters other than the mean/location
+parameter.
+
+\item Binarization of residuals/scores:\\
+Rather than applying the independence test directly on the residuals/scores a
+binarized version of them can be used as well. In particular, the values are binarized
 at $0$ such that only their signs are considered.
@@
 \textit{Note:}
-Apart from these three switches, the algorithms differ in their approximation of the underlying
-null distribution of the test statistic. While the permutation tests in the CTree algorithm
-work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional approach.
-
-\begin{table}[t!]
+Apart from these three building blocks, the algorithms differ in their approximation of the 
+underlying null distribution of the test statistic. While the permutation tests in the CTree 
+algorithm work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional 
+approach.
+
+\begin{table}[t]
 \centering
-\caption[Table caption text]{CTree, MOB, and GUIDE with the corresponding setting of
-the switches in the unifying framework and the approximation type of the underlying null distribution.}
-\label{tab:switches}
+\caption[Table caption text]{Testing strategies of CTree, MOB, and GUIDE with the corresponding
+setting of the building blocks in the unifying framework and the approximation type of the
+underlying null distribution.}
+\label{tab:buildingblocks}
 \begin{tabular}{ l c c c c }
@@
 \hline
-CTree & full model scores & -           & -             & conditional\\
-MOB   & full model scores & -           & -             & unconditional\\
+CTree & full model scores & --          & --            & conditional\\
+MOB   & full model scores & --          & --            & unconditional\\
 GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
@@
 
-\subsection{Linear model tree}
+%\newpage
+
+\subsection{Linear Model Tree}\label{sec:lmtrees}
 As an example a linear model tree is considered as this allows 
 for an easy and well structured description of the investigated subjects/issues. 
-For that reason, this model is also considered for the simulation study in Section~\ref{sec:simulation}.
+For that reason, this model is also considered for the simulation study in 
+Section~\ref{sec:simulation}.
 However, the conclusions drawn from this example also apply for a wide range of other model
-types. The models fitted in the nodes of the considered linear tree model are of the form
+types. The models fitted in the nodes of the considered linear model tree are of the form
 $$
@@
 
-A very basic form of a discrepancy measure can be obtained by simply calculating deviations of the 
-response values from the mean value within the node, for example by $\sum_{i=1}^n (\bar{y} - y_i)^2$.
+A very basic form of a discrepancy measure can be obtained for each observation by simply 
+calculating deviations of the response values from the mean value within the node, for example 
+by $(\bar{y} - y_i)$.
 However, this includes only the response variable $Y$ but not the fitted model.
-For a model-based approach residuals can be considered such as in the residual sum of squares (RSS)
+For a model-based approach residuals can be considered defined by
 $$
-r(y,x,\boldsymbol{\beta}) = RSS(y, x, \boldsymbol{\beta}) = 
-\sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2
+r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
 $$
-which is also the objective function that is minimized for the estimation of the coefficient vector 
+The appropriate choice of the function $\psi$ is then
+$$
+\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2.
+$$
+such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
+which is minimized for the estimation of the coefficient vector
 $\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$, 
 i.e. to obtain $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
-The corresponding scores are defined by
+The corresponding score function is defined by
 $$
-s(y_i, x_i, \hat{\boldsymbol{\beta}}) = 
-\frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
+s(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
+\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
+2 \cdot r(y_i,x_i,\hat{\beta_0},\hat{\beta_1}) \cdot (1,x_i)^\top
 $$
-such that the score matrix $S$ is in this case a $n \times 2$-matrix as $\boldsymbol{\beta}$ is 
-two dimensional. In that way a score value indicating how well the model fits the data is given separately
-for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
-and $j \in \{1,2\}$.
+where the factor $2$ is often omitted such that the first column of the score matrix $S$ 
+corresponds to the residuals. $S$ is in this case a $n \times 2$-matrix
+as $\boldsymbol{\beta}$ is two dimensional. In that way a score value indicating how well the model
+fits the data is given separately for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$
+with $i \in \{1,\dots,n\}$ and $j \in \{1,2\}$.
 
 In this example considering only residuals, i.e. only the first column of the score matrix $S$,
-corresponds to including the derivatives of $RSS$ only regarding the intercept parameter~$\beta_0$.
+corresponds to including the derivatives of $\psi$ only regarding the intercept 
+parameter~$\beta_0$.
 On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
-the derivatives of $RSS$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
+the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
 $\beta_1$ (second column).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%All three algorithms apply statistical tests yielding $p$-values based on which
-%the most appropriate split variable is chosen. One main difference lies in
-%the way how scores are included. While CTree and MOB directly include the values
-%of all available scores the GUIDE algorithm only uses residuals, in particular,
-%the signs of residuals and binned versions of the covariates.
-%The advantages of including all scores or only residuals as well as the impact
-%of the binarization of scores at zero and the binning of covariates is
-%investigated in detail in Section~\ref{sec:results}.
-%Hence, the GUIDE algorithm differs mainly in the setting of the applied test
-%in which the MOB and CTree algorithms are very similar. However, one essential difference
-%between the testing strategies of MOB and CTree lies in the underlying approximation 
-%of the null distribution where MOB employs an unconditional approach contrary
-%to the conditional framework of CTree.
-
-%Once the split variable is found they choose the split point minimizing the discrepancy measure, hence, 
-%the RSS in this case.
-
-%Once the split variable is chosen the search area for the optimal split point 
-%is reduced to the range of this split variable. A possible but computationally 
-%demanding approach to finding the best split point is an exhaustive search over 
-%all possible split points in the split variable. However, other
-%strategies have been developed as well, for example based on two-sample statistics such as 
-%applied in the CTree algorithm.
 
@@
 
-Two different scenarios are considered in this simulation study. First, the underlying
-tree structure based on which the data is generated is a stump, i.e. a tree with only one 
-split. By keeping the complexity of the model as low as possible the features/advantages and
+In this simulation study two different scenarios are considered for the linear model trees
+presented in Section~\ref{sec:lmtrees}. First, the underlying tree structure based on which the
+data is generated is a stump, i.e. a tree with only one split. 
+By keeping the complexity of the model as low as possible the advantages and
 disadvantages of the different testing strategies can be elaborated more precisely and 
@@
 
-\subsection{Data generating process}
+\subsection{Data Generating Process}
 Each generated data set consists of a response variable, one regressor, one split 
@@
 opposite directions.
-An illustration of the three different scenarios 
+An illustration of the three different cases 
 %for binary coefficients and a continuous regressor 
@@
 \minipage{0.45\textwidth}
-<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=4>>=
+<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
 #par(mar=c(3,3,0,0), oma = c(3,3,0,0))
@@
 \minipage{0.45\textwidth}
-<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=4>>=
-par(mar = c(4,4,2,1))
+<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
+par(mar = c(4,4,2,4))
 plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
@@
 legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = 3.2)
+mtext(expression(beta[1]== 1), side = 4, col = "red", las = 1, line = 0.5, at = 2.2)
+mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.2)
+mtext(expression(beta[1]== 1), side = 4, col = "blue", las = 1, line = 0.5, at = -0.8)
 @
@@
 \minipage{0.45\textwidth}
-<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=4>>=
+<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
 d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
@@
 
-par(mar = c(4,4,2,1))
+par(mar = c(4,4,2,4))
 plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
@@
 legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==0), side = 4, col = "red", las = 1, line = 0.5, at = -1)
+mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
+mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
+mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1)
 @
@@
 \minipage{0.45\textwidth}
-<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=4>>=
+<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
 d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
                vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
-par(mar = c(4,4,2,1))
+par(mar = c(4,4,2,4))
 plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
      main = expression("varying"~beta[0]~"and"~beta[1]),
+     xlim = c(-1,1.1),
      pch = 19, col = alpha("blue", 0.4), cex = 0.7)
@@
 legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.7)
+mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.7)
+mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = -0.7)
+mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -1.7)
 @
 \endminipage
-\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$ and $\beta_1$, either varying or being fixed at $0$ or $1$.
+\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$ and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
 Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
-with effect size $\delta = 1$ and for three scenarios: 
+with effect size $\delta = 1$ and for three cases: 
 varying $\beta_0$ and fixed $\beta_1$ (top right),
@@
 <<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
-par(mar = c(4,4,2,1))
+par(mar = c(4,4,2,4))
 plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
@@
 abline(coef = c(-1,-1), col = "red", lwd = 2)
-abline(coef = c(1,-1), col = "green", lwd = 2)
+abline(coef = c(1,-1), col = "darkgreen", lwd = 2)
 points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
@@
 points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
-       col = alpha("green", 0.4), cex = 0.7)
+       col = alpha("darkgreen", 0.4), cex = 0.7)
 legend("top", c(expression(z[2]<=xi), 
@@
                 expression(z[2]>xi~" and "~z[1]>xi)), 
-       col = c("blue", "red", "green"), pch = c(19), bty = "n", cex = 0.9)
+       col = c("blue", "red", "darkgreen"), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
+mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.3)
+mtext(expression(beta[0]==+delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = 0.1)
+mtext(expression(beta[1]==-delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = -0.6)
+mtext(expression(beta[0]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
+mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2.7)
 @
@@
 
-%\begin{figure}%[t]
-%\begin{center}
-%\setkeys{Gin}{width=0.6\linewidth}
-%<<true_tree_biv, fig=TRUE, echo=FALSE, width=5.4, height=4>>=
-%plot(ctest, terminal_panel = node_bivplot, drop_terminal = TRUE, edge_panel = ep,
-%     tp_args = list(margins = c(0,1,0,1), pointcol = "white", pointcex = 0.7, lwd = 2),
-%     ip_args = list(pval = FALSE, fill = paltrees[c(2, 3)]))
-%@
-%\caption{\label{fig:true_tree_biv}True tree structure for split point $\xi = 0$ and 
-%effect size $\delta = 1$} with the corresponding linear functions being plotted
-%in each terminal node.
-%\end{center}
-%\end{figure}
-
 \subsection{Evaluation}
@@
 on 100 data sets per step each consisting of 250 observations. 
+To compare the performance of the evaluated testing strategies for each step 
+the $p$-values corresponding to the true split variable $Z_1$ and the proportion
+of data sets for which $Z_1$ is detected and also selected 
+%(for a given significance level $\alpha = 0.05$) 
+is considered for the first scenario (stump).
+For the second scenario (tree with 2 splits) the Adjusted Rand Index~(ARI) is calculated 
+as a measure of similarity between the true tree structure and the fitted model tree.
 
 By applying the unifying framework presented in Section~\ref{sec:unify} the effect of
-each switch can be investigated on its own. In Table~\ref{tab:switches} the setting of these 
-switches is listed for the algorithms CTree, MOB and GUIDE in their original form.
-However, in order to be able to focus on particular features rather than the whole algorithms
-other combinations as presented in Table~\ref{tab:switches}, hence variations of the original 
-algorithms, are evaluated as well.
-
-Moreover, all of the considered statistical tests are applied as significance tests answering 
-two questions at once: (1) Should a split be performed at all? (2) If so, in which variable?
+each building block can be investigated on its own. In Table~\ref{tab:buildingblocks} the setting
+of these building blocks is listed for CTree, MOB and GUIDE in their original form.
+However, in order to be able to focus on particular features rather than the whole testing
+strategies other combinations as presented in Table~\ref{tab:buildingblocks}, hence variations of
+the original testing strategies, are evaluated as well.
+
+Moreover, as prepruning is employed all of the considered statistical tests are applied as significance tests answering two questions at once: 
+(1)~Should a split be performed at all? 
+(2)~If so, in which variable?
 For the first question the $p$-values regarding all available covariates are compared 
-to a predefined level of significance $\alpha$. Only if the smallest $p$-value is smaller
+to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
 than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is 
-selected. Including this criterion of significance in the building process of a tree is often
-referred to as \textit{pre-pruning}, however, this is not always applied in practice as 
-significance is often not considered. As mentioned before, the GUIDE algorithm usually builds
-large trees which are then pruned afterwards. However, for the purpose of comparing the
-testing strategies rather than the whole tree algorithms in this simulation study a 
-significance level is considered for all statistical tests, including the $\chi^2$-test in 
-the GUIDE algorithm such that all algorithms apply pre-pruning rather than post-pruning.
-The effect of these two pruning strategies are investigated in Section~\ref{sec:discussion}.
-
-To compare the performance of the evaluated testing strategies for each step the proportion
-of data sets for which the true splitting variable $Z_1$ is detected and also selected 
-(for a given significance level $\alpha = 0.05$) is considered for the first scenario 
-(stump).
-For the second scenario (tree with 2 splits) the Adjusted Rand Index (ARI) is calculated 
-as a measure of similarity between the true tree structure and the fitted model tree.
-
+selected.
+
+%As mentioned before, the GUIDE algorithm usually builds large trees which are then pruned 
+%afterwards. However, for the purpose of comparing the testing strategies rather than the whole
+%tree algorithms in this simulation study a significance level is considered for all statistical 
+%tests, including the $\chi^2$-test in GUIDE such that all algorithms apply prepruning rather 
+%than postpruning.
+%The effect of these two pruning strategies are investigated in Section~\ref{sec:discussion}.
+
+\newpage
 \section{Results}
@@
 
-To illustrate the effects of the switches in the unifying framework presented in 
-Section~\ref{sec:unify} each of them is investigated separately for the first
-simulation scenario (stump) in Sections~\ref{sec:scores}, \ref{sec:categorization}
-and \ref{sec:binarization}. Then the testing strategies are evaluated on the more
+To illustrate the effects of the building blocks in the unifying framework presented in 
+Section~\ref{sec:unify} different combinations of them are investigated for the first
+simulation scenario (stump) in Sections ~\ref{sec:scores} and \ref{sec:3way}. 
+Then the testing strategies are evaluated on the more
 complex tree structure of the second scenario in Section~\ref{sec:tree}. 
@@
 
-\subsection{Residuals vs full model scores}
+\subsection{Residuals vs. Full Model Scores}
 \label{sec:scores}
@@
 strategies are compared: CTree, GUIDE, MOB (all three in their original form),
-and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_full).
-Thus, the four strategies are obtained from the unifying framework by setting the switches 
-as listed in Table~\ref{tab:switches_scores}.
-
-\begin{table}%[t]
-\centering
-\caption[Table caption text]{CTree, MOB, GUIDE, and GUIDE\_full with the corresponding setting of
-the switches in the unifying framework and the approximation type of the underlying null distribution.}
-\label{tab:switches_scores}
-\begin{tabular}{ l c c c c }
-\hline 
-& Scores            & Binarization  & Categorization  & Null distribution \\ \hline
-CTree       & full model scores & -           & -               & conditional\\
-MOB         & full model scores & -           & -               & unconditional\\
-GUIDE       & residuals         & \checkmark  & \checkmark      & unconditional\\
-GUIDE\_full & full model scores & \checkmark  & \checkmark      & unconditional\\ \hline
-\end{tabular}
-\end{table}
-
-Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
-scenarios of varying only the intercept $\beta_0$ or only the slope parameter $\beta_1$
-or both of them. If the intercept varies (left and right panel) all strategies perform 
-similarly with CTree and MOB being slightly ahead. However, this advantage increases if 
-the slope parameter varies and the intercept is kept at a fixed value (middle panel). 
-In particular, the original strategy of GUIDE fails to detect any changes in $\beta_1$ 
-while GUIDE\_full performs reasonably well in this scenario. This is due to the fact that
-the GUIDE algorithm only includes residuals, i.e. only the scores regarding the intercept,
-such that changes in the slope parameter are not detected unless the corresponding scores 
-are als included as in the GUIDE\_full algorithm.
-These results clearly illustrate that overall the decision to include all available scores
-when testing for dependencies in order to select split variables is advantageous over 
-considering only the scores regarding the intercept as changes not only in the mean but also 
-in all other parameters can be detected in that way.
+and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_scores).
+%Thus, the four strategies are obtained from the unifying framework by setting the switches 
+%as listed in Table~\ref{tab:buildingblocks_scores}.
+
+%\begin{table}%[t]
+%\centering
+%\caption[Table caption text]{CTree, MOB, GUIDE, and GUIDE\_scores with the corresponding setting of
+%the switches in the unifying framework and the approximation type of the underlying null 
+%distribution.}
+%\label{tab:buildingblocks_scores}
+%\begin{tabular}{ l c c c c }
+%\hline 
+%            & Scores            & Binarization  & Categorization  & Null distribution \\ \hline
+%CTree       & full model scores & --          & --              & conditional\\
+%MOB         & full model scores & --          & --              & unconditional\\
+%GUIDE       & residuals         & \checkmark  & \checkmark      & unconditional\\
+%GUIDE\_scores & full model scores & \checkmark  & \checkmark      & unconditional\\ \hline
+%\end{tabular}
+%\end{table}
 
@@
 \setkeys{Gin}{width=\linewidth}
-<<scores_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
+<<scores_stump, fig=TRUE, echo=FALSE, height=5, width=8>>=
 subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
@@
                        labels = c("CTree","MOB",
-                                  "GUIDE_full", "GUIDE"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
-
-xyplot(prop_Tsplit ~ delta | vary_beta, groups = ~ test, 
-       data = subdata, 
+                                  "GUIDE_scores", "GUIDE"))
+
+subdataxi <- subset(subdata, xi %in% c(0,0.8))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
+
+library("latticeExtra")
+useOuterStrips(xyplot(prop_Tsplit ~ delta | vary_beta + xi, groups = ~ test, 
+                      data = subdataxi, 
                       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & xi == 0, 
+                        only_intercept == FALSE,# & xi == 0, 
                       type = "l", lty = c(1,1,1,1),
-       key =  list(text = list(levels(subdata$test)),
+                      scales=list(x=list(at=seq(1,11,2))),
+                      key =  list(text = list(levels(subdataxi$test)),
                                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
-       layout= c(3,1),
-       index.cond = list(c(2,3,1)),
+                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
+                      layout= c(3,2),
+                      index.cond = list(c(2,3,1),c(2,1)),
                       xlab = expression(delta),
                       ylab = expression("Proportion of selecting"~Z[1]),
-       strip = strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
+                      par.settings = par.settings),
+
+               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
                                                      expression("vary"~beta[0]), 
                                                      expression("vary"~beta[1]))),
-       par.settings = par.settings)
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8))))
+
 @
@@
 varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and 
-both coefficients varying (right panel). The performances of four algorithms are compared: 
-CTree, MOB, GUIDE, GUIDE\_full.}
+both coefficients varying (right panel). The performances of four testing strategies are compared: 
+CTree, MOB, GUIDE, GUIDE\_scores.}
 \end{figure}
 
-\subsection{Binarization of scores}
-\label{sec:binarization}
-To find out whether a binarization of the scores at $0$ leads to an improvement or a deterioration
-of performance the two algorithms CTree and MOB are applied, once in their original version
-without a binarization and once in an adapted version with a binarization of scores at $0$ 
-(CTree\_bin, MOB\_bin). Moreover, they are compared to the adapted GUIDE version which includes 
-all available scores (GUIDE\_full).
-Hence, the five strategies are obtained from the unifying framework by setting the switches 
-as listed in Table~\ref{tab:switches_bin}.
-
-\begin{table}%[t]
+Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
+scenarios of varying only the intercept $\beta_0$ or only the slope parameter $\beta_1$
+or both of them, once for a true split point $\xi=0$ and once for $\xi=0.8$. 
+As explained before, the tests are applied as significance tests with significance 
+level $\alpha=0.05$. Therefore, the proportion of cases in which the true split variable $Z_1$
+is detected with a $p$-value smaller than $\alpha$ is compared.
+(To show that in general the drawn conclusions do not change if no significance level is included 
+the same setting is evaluated again without any significance level and the results are presented
+in Appendix~\ref{app:significance}.)
+
+For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
+intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
+this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
+(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
+in $\beta_1$. 
+This is due to the fact that GUIDE only includes residuals, i.e. only the scores 
+regarding the intercept, such that changes in the slope parameter are not detected unless the 
+corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
+only scores has already been discussed and elaborated in the literature of structural change
+analysis \citep[see e.g.][]{Ploberger+Kramer:1992}.
+
+Looking at the results for the split point being close to the upper edge of the considered
+interval ($\xi=0.8$, lower panels) all of the applied strategies perform worse than for the
+central split point. In this case MOB is clearly ahead leading to the best results. CTree 
+performs second best while both GUIDE versions struggle to detect the correct split variable 
+even for a high effect size.
+
+Altogether these results clearly illustrate that overall the decision to include all available
+scores when testing for dependencies in order to select split variables is advantageous over
+considering only residuals, i.e. the scores regarding only the intercept, as changes not only in
+the mean but also in all other parameters can be detected in that way. This is particularly obvious
+when comparing GUIDE and GUIDE\_scores. Moreover, from the fact that these two version are both
+performing worse than MOB and CTree it can be deduced that categorizing split variables and
+binarizing scores has a negative effect on the ability to detect the right split variable. 
+To elaborate these effects more in detail the results for all possible combinations of the 
+building blocks are discussed in a full factorial 3-way analysis.
+
+\subsection{Full Factorial Analysis}\label{sec:3way}
+In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
+case where the intercept and slope parameter are both varying and the true split point is in the
+center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
+$\delta = 0.3$ is considered making the differences particularly well observable. 
+(Additional results over increasing $\delta$ as in Section~\ref{sec:scores} 
+can be found in Appendix~\ref{app:increasingdelta}.)
+The power of the different combinations is measured by the mean $p$-value for the true split 
+variable~$Z_1$.
+The possible values for each of the three building blocks are listed in Table~\ref{tab:levels}.
+In order to evaluate all 12 combinations the original testing strategies CTree, MOB, and GUIDE
+are applied together with variations of them. For each combination the corresponding employed
+strategy can be found in Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.
+
+\begin{table}[h!]
 \centering
-\caption[Table caption text]{CTree, CTree\_bin, MOB, MOB\_bin, and GUIDE\_full with the 
-corresponding setting of the switches in the unifying framework and the approximation type 
-of the underlying null distribution.}
-\label{tab:switches_bin}
-\begin{tabular}{ l c c c c }
+\caption[Table caption text]{Available levels of the building blocks.}
+\label{tab:levels}
+\begin{tabular}{ l l l }
 \hline
-            & Scores            & Binarization  & Categorization  & Null distribution \\ 
+Building Block         & Levels              & Effect \\
 \hline
-CTree       & full model scores & -           & -             & conditional\\
-CTree\_bin  & full model scores & \checkmark  & -             & conditional\\
-MOB         & full model scores & -           & -             & unconditional\\
-MOB\_bin    & full model scores & \checkmark  & -             & unconditional\\
-GUIDE\_full & full model scores & \checkmark  & \checkmark    & unconditional\\
+Residuals vs.\ Scores  & \textit{residuals}  & use only residuals\\
+                       & \textit{scores}     & use full score matrix\\
+Categorization         & \textit{cat}        & categorize split variables\\
+                       & \textit{max}        & selected split variables maximally w.o. categorization\\
+                       & \textit{lin}        & selected split variables linearly w.o. categorization\\
+Binarization           & \textit{bin}        & binarize scores\\
+                       & \textit{lin}        & use residuals/score values without binarization\\
 \hline
@@
 
-Figure~\ref{fig:bin_stump} shows the effect of binarizing the score values over
-different values for the true split point $\xi$, however, all four situations
-lead to the same conclusions. Binarizing the score values decreases the proportion
-of cases where the split variable was selected correctly.
-
-%% FIX ME: ctree or ctree_max (1 step: ctree better, 2 steps: ctree_max better)
+<<3way_data, results=hide, echo=FALSE>>=
+if(file.exists("3way_xi00.rda")){
+  load("3way_xi00.rda")
+} else {
+  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
+                delta = 0.3,
+                xi = 0, 
+                binary_regressor = FALSE,
+                binary_beta = TRUE,
+                only_intercept = FALSE,
+                nrlevels = 2, nrsplitvar = 1,
+                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
+                return_matrices = TRUE,
+                test = c("guide_sum_1_cor",
+                         "mfluc_resid_bin",
+                         "ctree_resid_bin",
+                         "mfluc_resid_cat",
+                         "mfluc_resid",
+                         "ctree_resid",
+                         "guide_sum_12",
+                         "mfluc_bin",
+                         "ctree_bin",
+                         "mfluc_cat",
+                         "mfluc",
+                         "ctree"))
+  d00 <- prep_3way(simres)
+  save(d00, file = "3way_xi00.rda")
+}
+
+if(file.exists("3way_xi08.rda")){
+  load("3way_xi08.rda")
+} else {
+  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
+                delta = 0.5,
+                xi = 0.8, 
+                binary_regressor = FALSE,
+                binary_beta = TRUE,
+                only_intercept = FALSE,
+                nrlevels = 2, nrsplitvar = 1,
+                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
+                return_matrices = TRUE,
+                test = c("guide_sum_1_cor",
+                         "mfluc_resid_bin",
+                         "ctree_resid_bin",
+                         "mfluc_resid_cat",
+                         "mfluc_resid",
+                         "ctree_resid",
+                         "guide_sum_12",
+                         "mfluc_bin",
+                         "ctree_bin",
+                         "mfluc_cat",
+                         "mfluc",
+                         "ctree"))
+  d08 <- prep_3way(simres)
+  save(d08, file = "3way_xi08_delta05.rda")
+}
+@
+
+\begin{figure}%[t]
+\setkeys{Gin}{width=1\linewidth}
+\minipage{0.55\textwidth}
+<<3way_xi00, fig=TRUE, echo=FALSE, height=4, width=4>>=
+par(mar = c(3,5,1,5.7))
+with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
+                           ylab = expression("Mean"~p~"-value for"~Z[1])))
+@
+\endminipage
+\minipage{0.45\textwidth}
+<<3way_xi08, fig=TRUE, echo=FALSE, height=4, width=3.3>>=
+par(mar = c(3,2,1,5.8))
+with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
+                           ylab = expression("Mean"~p~"-value for"~Z[1])))
+@
+\endminipage
+\caption{\label{fig:3way00_08} Mean $p$-values corresponding to the true split variable $Z_1$
+for the true split point $\xi=0$ (left panel) and $\xi=0.8$ (right panel),
+with both coefficients varying, effect size $\delta = 0.3$, 250 observations and 100 data sets.
+All 12 combinations of the three building blocks are evaluated.}
+\end{figure}
+
 \begin{figure}%[t]
 \setkeys{Gin}{width=\linewidth}
-<<bin_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
-                                         #"ctree_max", "ctree_max_bin",
-                                         "mfluc", "mfluc_bin",
-                                         "guide_sum_12"))
-
-subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "ctree_bin", 
-                                  #"ctree_max", "ctree_max_bin",
-                                  "mfluc", "mfluc_bin",
-                                  "guide_sum_12"),
-                       labels = c("CTree","CTree_bin",
-                                  #"CTree_max","CTree_max_bin",
-                                  "MOB", "MOB_bin",
-                                  "GUIDE_full"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
-                                                         #"CTree_max", "CTree_max", 
-                                                         "MOB", "MOB",
-                                                         "GUIDE_full")], 
-                                             fill = pal[c("CTree", "CTree", 
-                                                          #"CTree_max", "CTree_max", 
-                                                          "MOB", "MOB",
-                                                          "GUIDE_full")]), 
-                     superpose.line = list(col = pal[c("CTree", "CTree", 
-                                                       #"CTree_max", "CTree_max", 
-                                                       "MOB", "MOB",
-                                                       "GUIDE_full")]))
-
-xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & vary_beta == "all", 
-       type = "b",
-       lty = c(1,2,#1,2,
-               1,2,1),
-       layout= c(4,1),
-       key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,2,#1,2,
-                                        1,2,1), 
-                                type = "b", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "CTree", 
-                                            #"CTree_max", "CTree_max", 
-                                            "MOB", "MOB",
-                                            "GUIDE_full")], lwd = 1)), 
-       #index.cond = list(c(1,2,3,4)),
-       xlab = expression(delta),
-       ylab = expression("Proportion of selecting"~Z[1]),
-       strip = strip.custom(factor.levels = c(expression(xi==0), 
-                                              expression(xi==0.2), 
-                                              expression(xi==0.5), 
-                                              expression(xi==0.8))),
-       par.settings = par.settings)
-
-@
-\caption{\label{fig:bin_stump}Proportion of cases where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ for both parameter varying, $\sigma = 1$, 
-250 observations and 100 data sets per step. Each panel represents a different value of the 
-split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
-strategies are compared: CTree, CTree\_bin, MOB, MOB\_bin, GUIDE\_full.}
+<<3way_xi00solo, fig=TRUE, echo=FALSE, height=5, width=9.2>>=
+par(mar = c(3,5,1,5.7))
+with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1,
+                           ylab = expression("Mean"~p~"-value for"~Z[1]),
+                           xpd = TRUE))
+@
+\caption{\label{fig:3way} Mean $p$-values corresponding to the true split variable $Z_1$
+for the true split point $\xi=0$,
+with both coefficients varying, effect size $\delta = 0.3$, 250 observations and 100 data sets.
+All 12 combinations of the three building blocks are evaluated.}
 \end{figure}
 
-\subsection{Categorization of splitting variables}
-\label{sec:categorization}
-To investigate the effect of categorizing the possible splitting variables the two algorithms 
-CTree and MOB are applied, once in their original version without a categorization and once 
-in an adapted version with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
-Moreover, they are also compared to GUIDE\_full which includes all available scores.
-The corresponding setting of the switches in the unifying framework is listed in 
-Table~\ref{tab:switches_cat}.
-
-\begin{table}%[t]
-\centering
-\caption[Table caption text]{CTree, CTree\_cat, MOB, MOB\_cat, and GUIDE\_full with the 
-corresponding setting of the switches in the unifying framework and the approximation type 
-of the underlying null distribution.}
-\label{tab:switches_cat}
-\begin{tabular}{ l c c c c }
-\hline 
-& Scores            & Binarization  & Categorization  & Null distribution \\ 
-\hline
-CTree       & full model scores & -           & -               & conditional\\
-CTree\_cat  & full model scores & -           & \checkmark      & conditional\\
-MOB         & full model scores & -           &                 & unconditional\\
-MOB\_cat    & full model scores & -           & \checkmark      & unconditional\\
-GUIDE\_full & full model scores & \checkmark  & \checkmark      & unconditional\\
-\hline
-\end{tabular}
-\end{table}
-
-In Figure~\ref{fig:cat_stump} the impact of categorizing split variables on the performance 
-is illustrated for four different values of the true split point $\xi$. 
-For both, CTree and MOB, it can be stated that overall they perform better in their
-original form without any categorization. Only if the true split point $\xi$
-is close to the quartiles used as breakpoints for the categorizations both versions
-lead to a similar proportion of cases where the true splitting variable is detected
-(e.g., for CTree and $\xi=0.5$)
-
-Therefore, it can be summarized that a categorization of the values of the split variable
-does not lead to any advantages unless the true split point corresponds to one of the 
-quartiles used for the categorization. In most situations it even causes the performance to 
-get worse.
-
-\begin{figure}%%[t]
-\setkeys{Gin}{width=\linewidth}
-<<cat_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
-                                         #"ctree_max", "ctree_max_cat",
-                                         "mfluc", "mfluc_cat",
-                                         "guide_sum_12"))
-subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "ctree_cat", 
-                                  #"ctree_max","ctree_max_cat",
-                                  "mfluc", "mfluc_cat",
-                                  "guide_sum_12"),
-                       labels = c("CTree","CTree_cat",
-                                  #"CTree_max","CTree_max_cat",
-                                  "MOB", "MOB_cat",
-                                  "GUIDE_full"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
-                                                         #"CTree_max", "CTree_max", 
-                                                         "MOB", "MOB",
-                                                         "GUIDE_full")], 
-                                             fill = pal[c("CTree", "CTree", 
-                                                          #"CTree_max", "CTree_max", 
-                                                          "MOB", "MOB",
-                                                          "GUIDE_full")]), 
-                     superpose.line = list(col = pal[c("CTree", "CTree", 
-                                                       #"CTree_max", "CTree_max", 
-                                                       "MOB", "MOB",
-                                                       "GUIDE_full")]))
-
-xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & vary_beta == "all", 
-       #xlim = c(0,0.1,0.5),
-       type = "b", 
-       lty = c(1,2,#1,2,
-               1,2,1),
-       layout= c(4,1),
-       key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,2,#1,2,
-                                        1,2,1), 
-                                type = "b", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "CTree", 
-                                            #"CTree_max", "CTree_max", 
-                                            "MOB", "MOB",
-                                            "GUIDE_full")], lwd = 1)), 
-       #index.cond = list(c(1,2,3,4)),
-       xlab = expression(delta),
-       ylab = expression("Proportion of selecting"~Z[1]),
-       strip = strip.custom(factor.levels = c(expression(xi==0), 
-                                              expression(xi==0.2), 
-                                              expression(xi==0.5), 
-                                              expression(xi==0.8))),
-       par.settings = par.settings)
-
-@
-\caption{\label{fig:cat_stump}Proportion of cases where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ for both parameters varying, $\sigma = 1$, 
-250 observations and 100 data sets per step. Each panel represents a different value of the 
-split point $\xi$ ($0$, $0.2$, $0.5$ and $0.8$ from left to right). The performances of five 
-testing strategies are compared: CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_full.}
-\end{figure}
-
+The results presented in Figure~\ref{fig:3way} confirm the already discussed advantage of
+including all available scores rather than just residuals as this leads to a lower mean $p$-value 
+of $Z_1$ for all testing strategies.
+Moreover, it can be observed that all combinations that binarize scores/residuals perform
+worse than those who do not employ any binarization. 
+Regardless of the transformation of scores/residuals categorizing split variables also leads
+to higher $p$-values and therefore has a negative effect on the power of a testing strategy.
+Consequentially, without categorization the $p$-values are lower, particularly for a linear
+selection of split variables.
+
+\newpage
 \subsection{Tree}
@@
 been evaluated on data sets based on the tree structure with two different split variables 
-($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in Figure~\ref{fig:dgp_tree}.
+($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
+Figure~\ref{fig:dgp_tree}.
 
 Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
-almost equally well for the true split points $0$, $0.2$ and $0.5$ while GUIDE\_full performs
-a bit worse but still achieves an ARI above $0.5$ for the highest considered effect size
+almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE\_scores performs
+a bit worse but it still achieves an ARI above $0.5$ for the highest considered effect size
 of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
@@
 testing strategy leads to a better performance while binarizing scores and categorizing
-split variables both do not cause any improvement but can even have a negative effect in many situations.
+split variables both do not cause any improvement but in many situations even have a negative
+effect on the power of a testing strategies.
 
@@
                        labels = c("CTree","MOB",
-                                  "GUIDE_full", "GUIDE"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
+                                  "GUIDE_scores", "GUIDE"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
 
@@
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
        layout= c(4,1),
        #index.cond = list(c(1,2,3,4)),
+       scales=list(x=list(at=seq(1,11,2))),
        xlab = expression(delta),
@@
 $0.8$ from left to right). The performances of four testing strategies are compared: CTree,
-MOB, GUIDE, GUIDE\_full.}
+MOB, GUIDE, GUIDE\_scores.}
 \end{figure}
 
-\newpage
-
-\section{Discussion}
-\label{sec:discussion}
-
-\subsection{Conclusions}
-From comparing the effects of the three switches three main conclusions can be drawn: 
-(1) In most cases including all available score values rather than just residuals leads to 
-clearly better results and performs in general at least equally well.
-(2) Binarizing scores/residuals does not lead to any improvement in performance.
-(3) Categorizing possible split variables also doesn't improve the performance 
-unless the breakpoints used for the categorization are very close to the true split points.
-However, even then the results are very similar to those achieved without any categorization.
-
-\subsection{Significance level}
-In the simulation study presented in Section~\ref{sec:simulation} a significance level is 
-included in all of the algorithms such that a split is only performed if the lowest $p$-value
-is smaller than this significance level. This strategy, usually referred to as pre-pruning, aims at 
-avoiding overfitting.
-However, ignoring this significance level within the tree algorithms yields the same results 
-which is shown in Figure~\ref{fig:scores_stump_wosign} where the proportion of cases 
-in which a split is performed in the true splitting variable, but not necessarily with a $p$-value 
-smaller than the significance level, is illustrated.
-
-\begin{figure}
-\setkeys{Gin}{width=\linewidth}
-<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
-                                         "guide_sum_12", "guide_sum_1_cor"))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree","mfluc",
-                                  "guide_sum_12", "guide_sum_1_cor"),
-                       labels = c("CTree","MOB",
-                                  "GUIDE_full", "GUIDE"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
-
-xyplot(prop_z1 ~ delta | vary_beta, groups = ~ test, 
-       data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & xi == 0, 
-       type = "l", lty = c(1,1,1,1),
-       key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
-       layout= c(3,1),
-       index.cond = list(c(2,3,1)),
-       xlab = expression(delta),
-       ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
-       strip = strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
-                                              expression("vary"~beta[0]), 
-                                              expression("vary"~beta[1]))),
-       par.settings = par.settings)
-@
-\caption{\label{fig:scores_stump_wosign}Proportion of cases where the true split variable $Z_1$ 
-is selected over increasing effect size $\delta$ for $\sigma = 1$, $\xi=0$, 250 observations and 
-100 data sets per step without any significance level. 
-Each panel represents one scenario for the coefficients: 
-varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and 
-both coefficients varying (right panel). The performances of four algorithms are compared: 
-CTree, MOB, GUIDE, GUIDE\_full.}
-\end{figure}
-
-%FIX ME: Check implementation of pruning.
-
-%FIX ME: LM-specific results
-
-%FIX ME: results without loss of generalization
-
-\subsection{Pruning}
-
-To show that similar results are obtained without any significance level but with the application
-of post-pruning (building very large trees and employing cost-complexity pruning as in the CART algorithm) the models built in the second simulation scenario are built once with a pre-pruning and once
-with a post-pruning approach. The results are compared based on the Adjusted Rand Index in 
-Figure~\ref{fig:pruning_ari}.
+\subsection{Prepruning vs. Postpruning}
+
+To elaborate the impact of the choice of a pruning strategy
+the tree models presented in Figure~\ref{fig:tree_ari} are built again, but this time
+applying postpruning (building very large trees without any significance level and employing 
+cost-complexity pruning as in the CART algorithm). 
+In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with prepruning,
+again based on the Adjusted Rand Index.
+
+For MOB and CTree it has been shown that the tests work very well as significance tests.
+Therefore, it can be recommended to take advantage of this power and use the information
+given by the tests for prepruning. As it can be seen in Figure~\ref{fig:pruning_ari}
+in most cases the results of these two testing strategies do not vary significantly over 
+changing the applied pruning strategy.
+On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
+strategy should not be based on their results which argues for the application of postpruning.
+This can also be observed in Figure~\ref{fig:pruning_ari} showing that postpruning improves
+the results for both GUIDE versions.
 
@@
 pr <- pr[,c(1,2,7,8,9,12,15)]
-## extended data set with post-pruning results being represented as separated tests
+## extended data set with postpruning results being represented as separated tests
 prex <- pr[,c(1,2,3,4,6)]
@@
                     labels = c("CTree","MOB",
-                               "GUIDE_full", "GUIDE",
+                               "GUIDE_scores", "GUIDE",
                                "CTree_p","MOB_p",
-                               "GUIDE_full_p", "GUIDE_p"))
+                               "GUIDE_scores_p", "GUIDE_p"))
 
@@
   if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
-  if(prex$test[i] == "GUIDE_full") prex$test[i] <- "GUIDE_full_p"
+  if(prex$test[i] == "GUIDE_scores") prex$test[i] <- "GUIDE_scores_p"
   if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
@@
 
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")]))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
 
@@
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
        layout= c(4,1),
+       scales=list(x=list(at=seq(1,11,2))),
        xlab = expression(delta),
@@
 $0.8$ from left to right). The performances of four testing strategies are compared: CTree,
-MOB, GUIDE, GUIDE\_full.}
+MOB, GUIDE, GUIDE\_scores. The dashed lines represent the results from applying postpruning
+and the continuous lines the results from applying prepruning.}
 \end{figure}
+
+\newpage
+
+\section{Discussion}
+\label{sec:discussion}
+
+\subsection{Conclusions}
+From comparing the effects of the building blocks three main conclusions can be drawn: 
+(1)~In most cases including all available score values rather than just residuals leads to 
+clearly better results and performs in general at least equally well.
+(2)~Binarizing scores/residuals does not lead to any improvement in performance.
+(3)~Categorizing possible split variables also doesn't improve the performance 
+unless the breakpoints used for the categorization are very close to the true split points.
+However, even then the results are very similar to those achieved without any categorization.
+
+%FIX ME: Check implementation of pruning.
+
+%FIX ME: LM-specific results
+
+%FIX ME: results without loss of generalization
+
+\newpage
+\section*{Appendix}
+\begin{appendix}
+
+\section{Applied Combinations of Building Blocks}\label{app:combinations}
+For the evaluation of all 12 possible combinations of the presented building blocks
+the original testing strategies CTree, MOB, and GUIDE have been applied together
+with modified versions of them. The employed versions and the corresponding combination
+of building blocks is listed in Table~\ref{tab:combinations_appendix}.
+Each suffix represents a change of the original form of the considered testing strategy.
+In particular, ``\_res'' refers to using only residuals (instead of all scores),
+``\_scores'' to using all scores (instead of only residuals),
+``\_cat'' to applying a categorization of split variables,
+``\_bin'' to applying a binarization of scores/residuals.
+
+\begin{table}%[t]
+\centering
+\caption[Table caption text]{Applied testing strategies representing the 12 combinations
+of building blocks}
+\label{tab:combinations_appendix}
+\begin{tabular}{ l c c c }
+\hline 
+Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
+\hline
+GUIDE             & residuals          & bin          & cat \\
+MOB\_bin\_res     & residuals          & bin          & max \\
+CTree\_bin\_res   & residuals          & bin          & lin \\
+MOB\_cat\_res    & residuals          & lin          & cat \\
+MOB\_res          & residuals          & lin          & max \\
+CTree\_res        & residuals          & lin          & lin \\
+GUIDE\_scores     & scores             & bin          & cat \\
+MOB\_bin          & scores             & bin          & max \\
+CTree\_bin        & scores             & bin          & lin \\
+MOB\_cat          & scores             & lin          & cat \\
+MOB               & scores             & lin          & max \\
+CTree             & scores             & lin          & lin \\
+\hline
+\end{tabular}
+\end{table}
+
+\section{Significance Level}\label{app:significance}
+In Section~\ref{sec:scores} the investigated testing strategies are compared based
+on the number of cases in which the true split variable $Z_1$ is detected with a $p$-value
+smaller than the predefined significance level $\alpha = 0.05$. This choice of measurement
+is due to the aim of investigating how well the tests perform as significance test.
+However, ignoring this significance level yields the same conclusions 
+which can be seen when comparing Figures~\ref{fig:scores_stump_wosign}
+and \ref{fig:scores_stump}.
+In Figure~\ref{fig:scores_stump_wosign} the proportion of cases 
+in which the true splitting variable $Z_1$ is detected, but not necessarily with a $p$-value 
+smaller than the significance level, is illustrated. In Figure~\ref{fig:scores_stump} 
+the exact same simulation setting is applied with a significance level.
+
+\begin{figure}
+\setkeys{Gin}{width=\linewidth}
+<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
+                                         "guide_sum_12", "guide_sum_1_cor"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree","mfluc",
+                                  "guide_sum_12", "guide_sum_1_cor"),
+                       labels = c("CTree","MOB",
+                                  "GUIDE_scores", "GUIDE"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
+
+xyplot(prop_z1 ~ delta | vary_beta, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
+         only_intercept == FALSE & xi == 0, 
+       type = "l", lty = c(1,1,1,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
+       layout= c(3,1),
+       index.cond = list(c(2,3,1)),
+       scales=list(x=list(at=seq(1,11,2))),
+       xlab = expression(delta),
+       ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
+       strip = strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
+                                              expression("vary"~beta[0]), 
+                                              expression("vary"~beta[1]))),
+       par.settings = par.settings)
+@
+\caption{\label{fig:scores_stump_wosign}Proportion of cases where the true split variable $Z_1$ 
+is selected over increasing effect size $\delta$ for $\xi=0$, 250 observations and 
+100 data sets per step without any significance level. 
+Each panel represents one scenario for the coefficients: 
+varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and 
+both coefficients varying (right panel). The performances of four testing strategies are compared: 
+CTree, MOB, GUIDE, GUIDE\_scores.}
+\end{figure}
+
+\newpage
+\section{Results Over Increasing Effect Size}
+\label{app:increasingdelta}
+
+\subsection{Binarization of Residuals/Scores}
+\label{app:binarization}
+To elaborate over an increasing effect size whether a binarization of the residuals/scores at $0$ 
+leads to an improvement or a deterioration of performance CTree and MOB are applied, once in their 
+original version without a binarization and once in an adapted version with a binarization of 
+scores at $0$ (CTree\_bin, MOB\_bin). Moreover, they are compared to the adapted GUIDE version 
+which includes all available scores (GUIDE\_scores).
+%Hence, the five strategies are obtained from the unifying framework by setting the switches 
+%as listed in Table~\ref{tab:buildingblocks_appendix}.
+
+%\begin{table}%[t]
+%\centering
+%\caption[Table caption text]{CTree, CTree\_bin, MOB, MOB\_bin, and GUIDE\_scores with the 
+%corresponding setting of the switches in the unifying framework and the approximation type 
+%of the underlying null distribution.}
+%\label{tab:buildingblocks_appendix}
+%\begin{tabular}{ l c c c c }
+%\hline 
+%& Scores            & Binarization  & Categorization  & Null distribution \\ 
+%\hline
+%CTree       & full model scores & --          & --            & conditional\\
+%CTree\_bin  & full model scores & \checkmark  & --            & conditional\\
+%MOB         & full model scores & --          & --            & unconditional\\
+%MOB\_bin    & full model scores & \checkmark  & --            & unconditional\\
+%GUIDE\_scores & full model scores & \checkmark  & \checkmark    & unconditional\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+Figure~\ref{fig:bin_stump} shows the effect of binarizing the score values over
+different values for the true split point $\xi$, however, all four situations
+lead to the same conclusions. Binarizing the score values decreases the proportion
+of cases where the split variable was selected correctly.
 
@@
 \setkeys{Gin}{width=\linewidth}
-<<prune_nrsubgr, fig=TRUE, echo=FALSE, height=4, width=8>>=
-xyplot(nrsubgr ~ delta | xi, groups = ~ test, 
-       data = prex, 
-       type = "l", lty = c(1,1,1,1,2,2,2,2),
-       key =  list(text = list(levels(prex$test)[1:4]),
-                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_full", "GUIDE")], lwd = 1)),
+<<bin_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
+                                         #"ctree_max", "ctree_max_bin",
+                                         "mfluc", "mfluc_bin",
+                                         "guide_sum_12"))
+
+subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_bin", 
+                                  #"ctree_max", "ctree_max_bin",
+                                  "mfluc", "mfluc_bin",
+                                  "guide_sum_12"),
+                       labels = c("CTree","CTree_bin",
+                                  #"CTree_max","CTree_max_bin",
+                                  "MOB", "MOB_bin",
+                                  "GUIDE_scores"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
+                                                         #"CTree_max", "CTree_max", 
+                                                         "MOB", "MOB",
+                                                         "GUIDE_scores")], 
+                                             fill = pal[c("CTree", "CTree", 
+                                                          #"CTree_max", "CTree_max", 
+                                                          "MOB", "MOB",
+                                                          "GUIDE_scores")]), 
+                     superpose.line = list(col = pal[c("CTree", "CTree", 
+                                                       #"CTree_max", "CTree_max", 
+                                                       "MOB", "MOB",
+                                                       "GUIDE_scores")]))
+
+xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
+         only_intercept == FALSE & vary_beta == "all", 
+       type = "b",
+       lty = c(1,2,#1,2,
+               1,2,1),
        layout= c(4,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,2,#1,2,
+                                        1,2,1), 
+                                type = "b", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "CTree", 
+                                            #"CTree_max", "CTree_max", 
+                                            "MOB", "MOB",
+                                            "GUIDE_scores")], lwd = 1)), 
+       #index.cond = list(c(1,2,3,4)),
        xlab = expression(delta),
-       ylab = "Nr. of terminal nodes",
+       ylab = expression("Proportion of selecting"~Z[1]),
        strip = strip.custom(factor.levels = c(expression(xi==0), 
@@
                                               expression(xi==0.8))),
-       par.settings = par.settings
-)
-
-@
-\caption{\label{fig:pruning_nrsubgr}Number of terminal nodes over increasing effect size $\delta$ for
-250 observations and 100 data sets per step.
-Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ and 
-$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
-MOB, GUIDE, GUIDE\_full. Each of them is once applied with pre-pruning as before and once with 
-post-pruning (dashed lines)}
+       par.settings = par.settings)
+
+@
+\caption{\label{fig:bin_stump}Proportion of cases where the true split variable $Z_1$ is
+detected over increasing effect size $\delta$ for both parameter varying, 
+250 observations and 100 data sets per step. Each panel represents a different value of the 
+split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
+strategies are compared: CTree, CTree\_bin, MOB, MOB\_bin, GUIDE\_scores.}
 \end{figure}
+
+\subsection{Categorization of Splitting Variables}
+\label{app:categorization}
+To investigate the effect of categorizing the possible splitting variables CTree and MOB are
+applied, once in their original version without a categorization and once in an adapted version
+with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
+Moreover, they are also compared to GUIDE\_scores which includes all available scores.
+%The corresponding setting of the switches in the unifying framework is listed in 
+%Table~\ref{tab:buildingblocks_cat}.
+
+%\begin{table}%[t]
+%\centering
+%\caption[Table caption text]{CTree, CTree\_cat, MOB, MOB\_cat, and GUIDE\_scores with the 
+%corresponding setting of the switches in the unifying framework and the approximation type 
+%of the underlying null distribution.}
+%\label{tab:buildingblocks_cat}
+%\begin{tabular}{ l c c c c }
+%\hline 
+%& Scores            & Binarization  & Categorization  & Null distribution \\ 
+%\hline
+%CTree       & full model scores & --          & --              & conditional\\
+%CTree\_cat  & full model scores & --          & \checkmark      & conditional\\
+%MOB         & full model scores & --          &                 & unconditional\\
+%MOB\_cat    & full model scores & --          & \checkmark      & unconditional\\
+%GUIDE\_scores & full model scores & \checkmark  & \checkmark      & unconditional\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+In Figure~\ref{fig:cat_stump} the impact of categorizing split variables on the performance 
+is illustrated over increasing effect size for four different values of the true split 
+point $\xi$. For both, CTree and MOB, it can be stated that overall they perform better in their
+original form without any categorization. Only if the true split point $\xi$
+is close to the quartiles used as breakpoints for the categorizations both versions
+lead to a similar proportion of cases where the true splitting variable is detected
+(e.g., for CTree and $\xi=0.5$)
+
+Therefore, it can be summarized that a categorization of the values of the split variable
+does not lead to any advantages unless the true split point corresponds to one of the 
+quartiles used for the categorization. In most situations it even causes the performance to 
+get worse.
+
+\begin{figure}%%[t]
+\setkeys{Gin}{width=\linewidth}
+<<cat_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
+                                         #"ctree_max", "ctree_max_cat",
+                                         "mfluc", "mfluc_cat",
+                                         "guide_sum_12"))
+subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_cat", 
+                                  #"ctree_max","ctree_max_cat",
+                                  "mfluc", "mfluc_cat",
+                                  "guide_sum_12"),
+                       labels = c("CTree","CTree_cat",
+                                  #"CTree_max","CTree_max_cat",
+                                  "MOB", "MOB_cat",
+                                  "GUIDE_scores"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
+                                                         #"CTree_max", "CTree_max", 
+                                                         "MOB", "MOB",
+                                                         "GUIDE_scores")], 
+                                             fill = pal[c("CTree", "CTree", 
+                                                          #"CTree_max", "CTree_max", 
+                                                          "MOB", "MOB",
+                                                          "GUIDE_scores")]), 
+                     superpose.line = list(col = pal[c("CTree", "CTree", 
+                                                       #"CTree_max", "CTree_max", 
+                                                       "MOB", "MOB",
+                                                       "GUIDE_scores")]))
+
+xyplot(prop_Tsplit ~ delta | xi, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == TRUE & binary_regressor == FALSE & 
+         only_intercept == FALSE & vary_beta == "all", 
+       #xlim = c(0,0.1,0.5),
+       type = "b", 
+       lty = c(1,2,#1,2,
+               1,2,1),
+       layout= c(4,1),
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,2,#1,2,
+                                        1,2,1), 
+                                type = "b", pch = 1, cex = 0.8,
+                                col = pal[c("CTree", "CTree", 
+                                            #"CTree_max", "CTree_max", 
+                                            "MOB", "MOB",
+                                            "GUIDE_scores")], lwd = 1)), 
+       #index.cond = list(c(1,2,3,4)),
+       xlab = expression(delta),
+       ylab = expression("Proportion of selecting"~Z[1]),
+       strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                              expression(xi==0.2), 
+                                              expression(xi==0.5), 
+                                              expression(xi==0.8))),
+       par.settings = par.settings)
+
+@
+\caption{\label{fig:cat_stump}Proportion of cases where the true split variable $Z_1$ is
+selected over increasing effect size $\delta$ for both parameters varying, $\sigma = 1$, 
+250 observations and 100 data sets per step. Each panel represents a different value of the 
+split point $\xi$ ($0$, $0.2$, $0.5$ and $0.8$ from left to right). The performances of five 
+testing strategies are compared: CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_scores.}
+\end{figure}
+
+\end{appendix}
 

### 2019-01-24 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
first draft of abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2407 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a9a1d635cf9c615ac8348d695ca12e1b1b9c4f6e

%***********************************************************************

% This is a template to be used for the preparation of
% papers submitted to the 34th International Workshop on
% Statistical Modelling, to be held in Guimar?es, Portugal,
% July 7-12, 2019.

% Please follow the following general guidelines:
%
% (1) Do not specify any definitions, commands or style parameters.
%     Upon submission, your file will be checked for presence of
%     \newcommand or \def statements and if found, error message will be reported
%     by the submission form.
%
% (2) Follow the template below very tightly.
%
% (3) Include .pdf figures using the \includegraphics
%      command, an example of which are given below.
%
% (4) Use file names which begin with the surname of the first author.
%
% (5) When creating labels for cross-references, please start always
%     by surname of the first author, e.g., \label{smith:likelihood}
%
% (6) The template below contains some example materials
%      to guide you through the preparation of your paper.  However,
%      remove all the redundant material from your final document
%      before submitting.

% The guidelines above are needed in order to be able to combine all
% the papers into a single proceedings book of acceptable quality.
% Please follow the guidelines as strictly as possible. Deviations may
% result in papers being either refused by the registration form
% or sent back to the authors with the request to change
% their documents according to the guidelines.

% Special characters:
% Please do not use special characters (e.g., accents).
% Use TeX composition instead, such as \~n, \'a, \`e, \v{s}, \r{u} etc.

% Changes as of IWSM 2013:
%  * \usepackage{booktabs} added which allows \toprule et al. in the tabular environment
%    (\hline\hline is not longer used)
%  * '^\T' added in iwsm.sty to denote transposed vectors and matrices within math (see example below)
%  * \usepackage{amsmath, amssymb} introduced since IWSM 2012 is allowed (allowing usage of boldsymbols
%    and other handy constructions (align, pmatrix etc.) within math)
%  * \usepackage{psfrag} introduced since IWSM 2012 is NOT allowed
%
%

%***********************************************************************
% PLEASE LEAVE THIS PART UNCHANGED
%***********************************************************************

\documentclass[twoside]{report}
\usepackage{iwsm}
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}

% Please do not specify any new definitions, any new commands,
% and do not specify any style parameters.
% The preamble of the document should be left unchanged.

\begin{document}

%***********************************************************************
% PLEASE INSERT YOUR CONTENT FROM HERE
%***********************************************************************

% Title and running title to be used as left header:
\title{Distributional Trees for Circular Data}
\titlerunning{Distributional Trees for Circular Data}

% Authors and running list of authors to be used as right header:
\author{Lisa Schlosser\inst{1}, Moritz N. Lang\inst{1}, Achim Zeileis\inst{1}}
\authorrunning{Schlosser et al.}    %% use \authorrunning{Surname 1} if only 1 author
                                    %% use \authorrunning{Surname 1 and Surname2} if two authors
                                    %% use \authorrunning{Surname 1 et al.} if more than two authors

% Institutes of all authors
% Include city and country of each institute, do not include the full address.
\institute{University of Innsbruck, Austria}

% E-mail of presenting author for correspondence
\email{Lisa.Schlosser@uibk.ac.at}

% Brief abstract of the paper:
\abstract{\textit{TO DO}}

% Keywords (at most 5):
\keywords{Distributional trees; Circular response; von Mises distribution.}

% Produce the title:
\maketitle

%***********************************************************************

% Sections and subsections (do not use lower levels):

\section{Motivation}
Circular response variables occur in a variety of applications and subject areas. E.g., gun crime
data on a $24$-hour scale is analysed in the social-economics, animal orientation or gene-structure
analysis are often subject of examination in biology, and wind data is one of the most important weather variables in meteorology.

\section{Methodology}

\subsection{von Mises Distribution}
A commonly chosen distribution to model circular data is the von Misis distribution
which is also known as the circular normal distribution. Its density function is given by
\begin{equation}
  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
modified Bessel function of the first kind and order $0$.
For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
to the variance $\sigma^2$ of the normal distribution.

\subsection{Distributional Trees}
Distributional trees (Schlosser et. al, 2018) embed distributional modeling such as
provided by generalized additive models for location, scale, and shape 
(GAMLSS, Rigby et. al, 2005) into the framework of regression trees (Breiman et. al, 1984)
by applying one of the unbiased recursive partitioning algorithms MOB (Hothorn et. al, 2006) 
and CTree (Zeileis et. al, 2008).

In the here discussed application the prespecified distribution family is the von Mises distribution. Starting at the root of the tree, i.e. the first node containing the complete data, the corresponding distribution parameter vector $\theta = (\mu,\kappa)$ is estimated via maximum likelihood, i.e. by maximizing $l(y, \theta) = log(f_{vM}(y;\theta))$. To evaluate the goodness of fit model scores are calculated by $s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. For each possible split variable $Z$ the association between the scores and $Z$ is evaluated by a permutation or M-fluctuation test such that the resulting $p$-values allow for an unbiased split variable selection on a unified scale. The variable showing the highest association to the scores is chosen as a split variable. The corresponding split point is then selected based on a greedy search over all possible split points. After splitting the data into subgroups the procedure is repeated in each of the resulting child nodes until eather a stopping criteria is reached or only equal observations are left in a node. 
In that way a tree model is provided specifying a distributional model in each of its nodes.

\section{Data}

\textit{TO DO}

\bigskip

\section{Application}

\textit{TO DO}

\bigskip

%***********************************************************************

% Tables can be included at any place in the text.
% As general format, use tables with horizontal lines only
% (i.e., no vertical lines separating columns).
% Start and end each table with a double horizontal line.
% Tables are incorporated using a table environment:

%\begin{table}[!ht]\centering
% A caption is put above the table and a label is defined
% to be used for reference to this specific table.
% Use labels which are very unlikely to be used by
% other contributors; for example, use labels starting
% with the surname of the first author.
%\caption{\label{smith:tab1} Caption text \textbf{ABOVE} the table.}
% A table with three columns, where the first is left aligned,
% the second has centred entries, and the third is right aligned,
% is generated as follows (note: if requested, use \cmidrule{} instead of \cline{})
%\medskip
%\begin{tabular}{lcr}
% First line:
%\toprule[0.09 em]
% The body of the table:
%Title col1 & Title col2 & Title col3 \\
%\midrule
%row1 col1  & row1 col2  & row1 col3  \\
%row2 col1  & row2 col2  & row2 col3  \\ %
%row3 col1  & row3 col2  & row3 col3  \\
% last line:
%\bottomrule[0.09 em]
%\end{tabular}
%\end{table}

% In the text, reference to the Table can be made as follows:
%We refer to Table~\ref{smith:tab1} for a summary of our main results. Have a look to Table~\ref{smith:tab2} for
%an additional example.

%\begin{table}[!ht]\centering
%\caption{\label{smith:tab2} Caption text \textbf{ABOVE} the table.}
%\medskip
%\begin{tabular}{lcr}
% First line:
%\toprule[0.09 em]
% The body of the table:
%  &\multicolumn{2}{c}{Title  for cols 2 -3} \\
%\cmidrule{2-3} %
%Title1 & Title2 & Title3 \\
%\midrule
%& $a$  & $c$ \\
%& $b$  & $d$ \\ %
%\midrule[0 em]
%Total  & $a+b$  & $n$  \\
% last line:
%\bottomrule[0.09 em]
%\end{tabular}
%\end{table}

%***********************************************************************

% Figures can be included at any place in the text.
% The only allowed formats for figures are pdf files.
%
% Please, do not include figures in any other format.
%
% Use file names which are very unlikely to be used by
% other contributors; for example, use file names starting
% with the surname of the first author.
% Figures are incorporated using a figure environment:
% Make sure you specify the extension of the file (pdf)

%Finally a figure (in \verb|.pdf|!)

%\begin{figure}[!ht]\centering
% You can pre-specify the width of the graph:
%\includegraphics[width=8cm]{circmax-004.pdf}
% Below the figure, a caption is put, and a label is defined
% to be used for reference to this specific figure.
% Use labels which are very unlikely to be used by
% other contributors; for example, use labels starting
% with the surname of the first author.
%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}
%\end{figure}

% In the text, reference to the Figure can be made as follows:
%We refer to Figure~\ref{smith:fig1} for a~graphical representation.

%***********************************************************************

% Acknowledgments, if needed:
%\acknowledgments{Special Thanks to ... }

%***********************************************************************

% References should be placed in the text in author (year) form.
% The list of references should be placed below IN ALPHABETICAL ORDER.
% (Please follow the format of the examples very tightly).

\references
\begin{description}
\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     {\it working paper}.
\item[Hothorn, T., Hornik, K., and Zeileis, A.] (2006).
     Unbiased Recursive Partitioning: A Conditional Inference Framework.
     {\it Journal of Computational and Graphical Statistics}, {\bf 15}, 3,
     651\,--\,674. 
\item[Zeileis, A., Hothorn, T., and Hornik, K.] (2008).
     Model-Based Recursive Partitioning.
     {\it Journal of Computational and Graphical Statistics}, {\bf 17}, 2,
     492\,--\,514.

\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
     Classification and Regression Trees.
     {\it Wadsworth}.

\item[Rigby, R. A., and Stasinopoulos, D. M.] (2005).
     Generalized Additive Models for Location Scale and Shape (with Discussion).
     {\it Journal of the Royal Statistical Society C}, {\bf 54}, 3,
     507\,--\,554.

\end{description}

\end{document}

### 2019-01-24 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions and additional references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2408 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a4e4e6e9d7ec28bf7326cf7326bc6e6657caf7b5

@@
 In many situations fitting one global model to a data set can be very challenging, 
-especially if this data contains lots of different features with high/extreme variations 
+especially if this data contains lots of different features with a strong variation 
 or complex interactions. Therefore, separating the data into more homogeneous subgroups 
-based on (a set of) covariates first can simplify the task and fitting a local model to 
+based on a set of covariates first can simplify the task and fitting a local model to 
 each of the resulting subgroups often leads to better results. This separation of the 
@@
 specific approaches to selecting a split variable and the corresponding split point. 
-Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
-CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
+Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
+CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
 the best split point and split variable in one step by directly comparing all possible split
@@
 the values of possible split variables are investigated.
-Moreover, the impact of applying either a pre- or postpruning strategy is analyzed.
-
-In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
+Additionaly, the impact of applying either a pre- or postpruning strategy is analyzed.
+
+In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree model is given. 
 Then, after explaining the investigated algorithms CTree, MOB, and GUIDE more in detail in 
 Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
-testing strategies in model-based unbiased recursive partitioning is presented in 
+testing strategies in model-based unbiased recursive partitioning algorithms is presented in 
 Section~\ref{sec:unify}.
@@
 As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
-GUIDE first select the split variable and then in a separate step the split point. 
+GUIDE first select the split variable and then the split point in a separate step. 
 For the first step of selecting a split variable they all apply statistical tests
@@
 split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
-of $Z_j$ on $D$ and in that way the dependency of $Z_j$ on the response $Y$ as the discrepancy
-measure (and therefore also $D$) depends on $Y$. 
+of $Z_j$ on $D$.
 \item For the resulting values of the test statistics the corresponding $p$-values are calculated
@@
 
-%To emphasize the differences between these three algorithms in how they select split variables,
-%their testing strategies are explained in detail in Section~\ref{sec:strategies} 
-%using the example of tree models with a linear model fitted in each node as discussed
-%in the simulation study. Again, this is done only for means of simplification and the same 
-%methodology can be applied for a wide variety of models.
-
 \subsection{CTree}\label{sec:ctree}
@@
 response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
-variable $Z$.
-For different types of response variables different functions $h$ which depend on $Y$ in a 
-permutation-symmetric way and might also be multidimensional can be selected.
-For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
-response can be mapped to a unity vector by an indicator function 
+variable~$Z$.
+The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
+and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
+choice, while a categorical response can be mapped to a unity vector by an indicator function 
 $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
@@
 algorithm.
+
 Applying the score function $s$ on a data set of $n$ observations containing the 
@@
 line two options are given: either the maximum of the absolute values of the standardized 
-statistic or a quadratic form is considered. The resulting values can then be transformed to 
+statistic or a quadratic form. The resulting values can then be transformed to 
 the scale of $p$-values where they can be compared in an unbiased way to select a split variable.
@@
 provides a very general framework where the choice of the type of variables and the corresponding 
-functions $g$ and $h$ determine the type of the applied statistical test 
-(see \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox).
+functions $g$ and $h$ determine the type of the applied statistical test.
+(See \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox.)
 
@@
 application of the central limit theorem. 
-The fitted model can be of a very basic and simple form such as 
-an average value of the response vector
-%a constant fit where the parameter $\beta$ is an
-%average value of the response, hence 
-%$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
-%$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ 
-but can also be a more complex parametric model.
-In \cite{Zeileis+Hothorn+Hornik:2008} GLMs and survival regression models are used to 
-exemplify the algorithm. Subsequently various other types of models have been applied as well,
+The fitted model can be of a very basic and simple form such as an average value of the response 
+vector but can also be a more complex parametric model.
+In \cite{Zeileis+Hothorn+Hornik:2008} generalized linear models (GLMs) and survival regression models 
+are used to exemplify the algorithm. Subsequently various other types of models have been applied as well,
 for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
@@
 
-For a categorical variable all possible splits are evaluated to find the best one while 
-for $Z$ being a numeric variable a maximally selected split point search is employed. Hence, 
+For a categorical split variable all possible splits are evaluated to find the best one while 
+for numeric split variable a maximally selected split point search is employed. Hence, 
 the number of possible breakpoints is not fixed unlike in the CTree algorithm.
@@
 In related literature other models together with an appropriate choice of residuals
-have been applied such as 
-regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, %(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
-Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))
-and quantile regression models \citep{Chaudhuri+Loh:2002} among many others.
+have been applied such as in
+regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
+%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
+quantile regression models \citep{Chaudhuri+Loh:2002} 
+and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
+%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))
 
@@
 \label{sec:pruning}
-In early developed methods of unbiased recursive partitioning
-such as FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
+In early developed methods of unbiased recursive partitioning such as 
+FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
 CRUISE \citep{Kim+Loh:2001} or GUIDE \citep{Loh:2002}
@@
 
-%\bigskip
-%\textit{Note:}
-%While GUIDE originally constructs large trees and applies postpruning with the cross-validation 
-%method of CART, CTree and MOB employ prepruning by setting a significance level for the 
-%independence tests such that no split is performed if the lowest $p$-value is not smaller than 
-%the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
-%in Section~\ref{sec:simulation} all three algorithms are applied using prepruning in order to put 
-%the focus on the power of the algorithms in terms of selecting the best split variable. 
-%However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.
-
 \section{Unifying Framework}\label{sec:unify}
@@
 \label{tab:combinations}
-\begin{tabular}{ l c c c }
+\begin{tabular}{ l l l l }
 \hline 
@@
 \hline
-CTree & non-parametric  & conditional inference    & prepruning\\
-MOB   & parametric      & score-based fluctuation  & prepruning (or AIC/BIC-based pruning)\\
-GUIDE & parametric      & residual-based $\chi^2$  & postpruning (cost-complexity pruning)\\
+CTree & non-parametric  & conditional inference    & pre\\
+MOB   & parametric      & score-based fluctuation  & pre (or post with AIC/BIC)\\
+GUIDE & parametric      & residual-based $\chi^2$  & post (cost-complexity pruning)\\
 \hline
@@
                 kids = list(
-                  #partynode(2L, info = c("true parameters:", 
-                  #                       " ",
-                  #                       expression(beta[0] == 0~" , "~beta[1] == -delta),
-                  #                       "or",
-                  #                       expression(beta[0] == +delta~" , "~beta[1] == 1),
-                  #                       "or",
-                  #                       expression(beta[0] == +delta~" , "~beta[1] == -delta))),
-                  #partynode(3L, info = c("true parameters:", 
-                  #                       " ",
-                  #                       expression(beta[0] == 0~" , "~beta[1] == +delta),
-                  #                       "or",
-                  #                       expression(beta[0] == -delta~" , "~beta[1] == 1),
-                  #                       "or",
-                  #                      expression(beta[0] == -delta~" , "~beta[1] == +delta)))))
-
                   partynode(2L, info = c("true parameters:", 
@@
 
-#par(mfrow = c(2,2))
-#par(mar=c(1,1,1,1), oma=c(1,1,1,1))
 paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
@@
 
-%As mentioned before, the GUIDE algorithm usually builds large trees which are then pruned 
-%afterwards. However, for the purpose of comparing the testing strategies rather than the whole
-%tree algorithms in this simulation study a significance level is considered for all statistical 
-%tests, including the $\chi^2$-test in GUIDE such that all algorithms apply prepruning rather 
-%than postpruning.
-%The effect of these two pruning strategies are investigated in Section~\ref{sec:discussion}.
-
 \newpage
@@
 <<3way_data, results=hide, echo=FALSE>>=
-if(file.exists("3way_xi00.rda")){
-  load("3way_xi00.rda")
+if(file.exists("3way_xi00_delta03.rda")){
+  load("3way_xi00_delta03.rda")
 } else {
@@
   d00 <- prep_3way(simres)
-  save(d00, file = "3way_xi00.rda")
+  save(d00, file = "3way_xi00_delta03.rda")
 }
 
-if(file.exists("3way_xi08.rda")){
-  load("3way_xi08.rda")
+if(file.exists("3way_xi08_delta1.rda")){
+  load("3way_xi08_delta1.rda")
 } else {
   simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
-                delta = 0.5,
+                delta = 1,
                 xi = 0.8, 
@@
   d08 <- prep_3way(simres)
-  save(d08, file = "3way_xi08_delta05.rda")
+  save(d08, file = "3way_xi08_delta1.rda")
 }
+
+d00$xi <- 0
+d08$xi <- 0.8
+
+d3way <- rbind(d00, d08)
+d3way$xi <- factor(d3way$xi, levels = c(0,0.8), labels = c(0,0.8))
 @
@@
 \setkeys{Gin}{width=1\linewidth}
-\minipage{0.55\textwidth}
-<<3way_xi00, fig=TRUE, echo=FALSE, height=4, width=4>>=
-par(mar = c(3,5,1,5.7))
-with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
-                           ylab = expression("Mean"~p~"-value for"~Z[1])))
-@
-\endminipage
-\minipage{0.45\textwidth}
-<<3way_xi08, fig=TRUE, echo=FALSE, height=4, width=3.3>>=
-par(mar = c(3,2,1,5.8))
-with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
-                           ylab = expression("Mean"~p~"-value for"~Z[1])))
-@
-\endminipage
-\caption{\label{fig:3way00_08} Mean $p$-values corresponding to the true split variable $Z_1$
-for the true split point $\xi=0$ (left panel) and $\xi=0.8$ (right panel),
-with both coefficients varying, effect size $\delta = 0.3$, 250 observations and 100 data sets.
-All 12 combinations of the three building blocks are evaluated.}
+<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
+xyplot(pval_z1 ~ cat | xi, 
+       groups = interaction(bin, res_scores), 
+       data = d3way, 
+       ylim = c(0,0.5),
+       type = c("a"), 
+       col = c("red", "blue", "red", "blue"),
+       lty = c(1,1,2,2),
+       xlab = "Categorization",
+       ylab = expression("p-value of"~Z[1]),
+       #index.cond = list(c(1,2,3,4)),
+       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
+                                              expression(xi==0.8~"and"~delta==1))),
+       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
+                  lines = list(lty = c(1,1,2,2), 
+                               type = "l",
+                               col = c("red", "blue", "red", "blue"))))
+
+#par(mar = c(3,5,1,5.7))
+#with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
+#                           ylab = expression("Mean"~p~"-value for"~Z[1])))
+#with(d00, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
+#with(d00, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
+#with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1)))
+#with(d08, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
+#with(d08, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
+@
+\caption{\label{fig:3way} Mean $p$-values corresponding to the true split variable $Z_1$
+over different levels of the building block \textit{Categorization}. 
+Each line represents
+a combination of the levels of the other two building blocks \textit{Binarization} and 
+\textit{Residuals vs.\ Scores}.
+Left panel: true split point $\xi=0$ and effect size $\delta = 0.3$.
+Right panel: true split point $\xi=0.8$ and effect size $\delta = 1$.}
 \end{figure}
 
-\begin{figure}%[t]
-\setkeys{Gin}{width=\linewidth}
-<<3way_xi00solo, fig=TRUE, echo=FALSE, height=5, width=9.2>>=
-par(mar = c(3,5,1,5.7))
-with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1,
-                           ylab = expression("Mean"~p~"-value for"~Z[1]),
-                           xpd = TRUE))
-@
-\caption{\label{fig:3way} Mean $p$-values corresponding to the true split variable $Z_1$
-for the true split point $\xi=0$,
-with both coefficients varying, effect size $\delta = 0.3$, 250 observations and 100 data sets.
-All 12 combinations of the three building blocks are evaluated.}
-\end{figure}
-
-The results presented in Figure~\ref{fig:3way} confirm the already discussed advantage of
-including all available scores rather than just residuals as this leads to a lower mean $p$-value 
-of $Z_1$ for all testing strategies.
-Moreover, it can be observed that all combinations that binarize scores/residuals perform
-worse than those who do not employ any binarization. 
-Regardless of the transformation of scores/residuals categorizing split variables also leads
-to higher $p$-values and therefore has a negative effect on the power of a testing strategy.
-Consequentially, without categorization the $p$-values are lower, particularly for a linear
-selection of split variables.
+The results presented in Figure~\ref{fig:3way} clearly show that binarizing scores/residuals
+reduces the power of the testing strategy as it leads to higher $p$-values corresponding
+to the true split variable $Z_1$. 
+This effect can be observed for a central split point (left panel, $xi=0$) as well as for a
+a split point closer to upper end of the considered interval (right panel, $xi=0.8$).
+However, the effect of the building block \textit{Categorization} varies over these two settings.
+For a central split point all three levels perform similarly. The linear selection of split variables
+is slightly ahead even though the true parameter function is a step function. 
+On the contrary, clear differences can be seen for the outward split point. Here, the maximal selection
+performs best while the linear selection shows its disadvantage in detecting abrupt changes
+and the categorized split variables lead to even higher $p$-values and thus lower power.
+Overall, it depends on the situation whether a linear or a maximal selection of a split variable
+is advantageous. However, categorizing possible split variables does not lead to any improvements
+but even declines the performance of a testing strategy unless the true split point is close
+to a break point used for the binarization.
 
@@
 
-\subsection{Conclusions}
+\fixme
 From comparing the effects of the building blocks three main conclusions can be drawn: 
@@
 
-%FIX ME: Check implementation of pruning.
-
-%FIX ME: LM-specific results
-
 %FIX ME: results without loss of generalization
@@
 of building blocks is listed in Table~\ref{tab:combinations_appendix}.
-Each suffix represents a change of the original form of the considered testing strategy.
-In particular, ``\_res'' refers to using only residuals (instead of all scores),
+Each suffix indicates a change of the original form of the considered testing strategy.
+In particular, it relates to the transformations of the data before applying a statistical
+test as represented by the different building blocks. In that way,
+``\_res'' refers to using only residuals (instead of all scores),
 ``\_scores'' to using all scores (instead of only residuals),

### 2019-01-27 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
added references and section 'outlook', revised section 'motivation'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2409 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cd6f94e40251eb9d46c3803549e9595f0df4269d

@@
 \section{Motivation}
-Circular response variables occur in a variety of applications and subject areas. E.g., gun crime
-data on a $24$-hour scale is analysed in the social-economics, animal orientation or gene-structure
-analysis are often subject of examination in biology, and wind data is one of the most important weather variables in meteorology.
+Circular response variables occur in a variety of applications and subject areas. 
+For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
+animal orientation or gene-structure analysis are often subject of examination in biology, 
+and wind direction is one of the most important weather variables in meteorology.
+
+One way to represent such a two-dimensional direction is by considering angles in a
+unity circle. This requires specifying a starting point (angle 0) and defining
+either a clockwise or anti-clockwise rotation. However, these settings can vary
+and hence should not influence statistical inference from a fitted model.
+For that reason circular data demands for purpose-built statistical methods. 
+One common way of modelling circular data is to apply the von Mises distribution 
+which is often regarded as the circular analog of the normal distribution. 
+Estimating its two distribution parameter by linear functions of regressors is
+a possible approach, however, might not be able to detect abrupt effects.
+As an alternative  we propose the application of distributional trees to
+model wind directions based on a set of regression variables.
 
@@
 provided by generalized additive models for location, scale, and shape 
-(GAMLSS, Rigby et. al, 2005) into the framework of regression trees (Breiman et. al, 1984)
-by applying one of the unbiased recursive partitioning algorithms MOB (Hothorn et. al, 2006) 
-and CTree (Zeileis et. al, 2008).
+(GAMLSS, Rigby et. al, 2005) into the framework of regression trees (Breiman et. al, 1984).
+In particular one of the unbiased recursive partitioning algorithms MOB (Hothorn et. al, 2006) 
+and CTree (Zeileis et. al, 2008) is applied and a distributional model is fitted in each node 
+by estimating the parameters of a specified distribution family via maximum likeklihood.
+For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
+tests applied on model scores and each possible split variable. After choosing the split variable 
+showing the highest association to the goodness of fit of the model the corresponding split point 
+which optimizes an objective function is selected in a separate step.
 
 In the here discussed application the prespecified distribution family is the von Mises distribution. 
-Starting at the root of the tree, i.e.
-the first node containing the complete data, the corresponding distribution parameter vector $\theta = (\mu,\kappa)$ is estimated via maximum likelihood, i.e.
-by maximizing $l(y, \theta) = log(f_{vM}(y;\theta))$.
-To evaluate the goodness of fit model scores are calculated by $s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$.
-For each possible split variable $Z$ the association between the scores and $Z$ is evaluated by a permutation or M-fluctuation test such that the resulting $p$-values allow for an unbiased split variable selection on a unified scale.
+The corresponding distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
+by maximizing $l(y, \theta) = log(f_{vM}(y;\theta))$ and model scores are calculated 
+by $s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$.
+
+%For each possible split variable $Z$ the association between the scores and $Z$ is evaluated by a permutation or M-fluctuation test such that the resulting $p$-values allow for an unbiased split variable selection on a unified scale.
 The variable showing the highest association to the scores is chosen as a split variable.
@@
 After splitting the data into subgroups the procedure is repeated in each of the resulting child nodes until eather a stopping criteria is reached or only equal observations are left in a node.
-In that way a tree model is provided specifying a distributional model in each of its nodes.
 
-\section{Data}
+\section{Application}
 
@@
 
-\section{Application}
+\section{Outlook}
 
-\textit{TO DO}
+While the application of the von Mises distribution allows for modelling a circular response variable,
+additional adaptions have to made when including circular regressor variables in a distributional tree
+model. In particular, if the selected split variable is a circular variable this demands for finding
+not only one split point but a ``split intervall''. This extension is of interest for many applications,
+for example in order to include day time as a regressor variable when modelling wind directions.
 
@@
 \begin{description}
-\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
-     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
-     {\it working paper}.
+
+\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
+     Classification and Regression Trees.
+     {\it Wadsworth}.
+
 \item[Hothorn, T., Hornik, K., and Zeileis, A.] (2006).
@@
      651\,--\,674.
-\item[Zeileis, A., Hothorn, T., and Hornik, K.] (2008).
-     Model-Based Recursive Partitioning.
-     {\it Journal of Computational and Graphical Statistics}, {\bf 17}, 2,
-     492\,--\,514.
 
-\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
-     Classification and Regression Trees.
-     {\it Wadsworth}.
+\item[Jammalamadaka, S. R., and Sengupta, A.] (2001).
+     Topics in Circular Statistics.
+     {\it World Scientific}, {\bf 5}.
+
+\item[Mardia, K. V., and Jupp, P. E.] (2009).
+     Directional Statistics.
+     {\it John Wiley \& Sons}, {\bf 494}.
 
@@
 
+\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
+     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
+     {\it working paper}.
+
+\item[Zeileis, A., Hothorn, T., and Hornik, K.] (2008).
+     Model-Based Recursive Partitioning.
+     {\it Journal of Computational and Graphical Statistics}, {\bf 17}, 2,
+     492\,--\,514.
+
 \end{description}

### 2019-01-29 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2412 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e3783c19a54ce64b2e08630bc05e6b1f16cc9a91

@@
 and wind direction is one of the most important weather variables in meteorology.
-
-One way to represent such a two-dimensional direction is by considering angles in a
-unity circle. This requires specifying a starting point (angle 0) and defining
-either a clockwise or anti-clockwise rotation. However, these settings can vary
-and hence should not influence statistical inference from a fitted model.
+Representing such a two-dimensional direction by an angle in a unity circle requires 
+the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
+rotation. However, these settings can vary but should not influence statistical 
+inference from a fitted model.
 For that reason circular data demands for purpose-built statistical methods. 
@@
 which is often regarded as the circular analog of the normal distribution. 
-Estimating its two distribution parameter by linear functions of regressors is
-a possible approach, however, might not be able to detect abrupt effects.
+Estimating its two distribution parameter by linear functions of regressor variables 
+is a possible approach, however, might not be able to detect abrupt effects.
 As an alternative  we propose the application of distributional trees to

### 2019-01-30 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2413 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0cbd347ba07043908f377c831625e83ed810129f

@@
 % Authors and running list of authors to be used as right header:
-\author{Lisa Schlosser\inst{1}, Moritz N. Lang\inst{1}, Achim Zeileis\inst{1}}
+\author{Lisa Schlosser\inst{1}, 
+Moritz N. Lang\inst{1},
+Torsten Hothorn\inst{2},
+Georg J. Mayr\inst{1},
+Reto Stauffer\inst{1},
+Achim Zeileis\inst{1}}
 \authorrunning{Schlosser et al.}    %% use \authorrunning{Surname 1} if only 1 author
@@
 % Include city and country of each institute, do not include the full address.
-\institute{University of Innsbruck, Austria}
+\institute{University of Innsbruck, Austria
+\and University of Zurich, Switzerland}
 
@@
 % Keywords (at most 5):
-\keywords{Distributional trees; Circular response; von Mises distribution.}
+\keywords{Distributional trees; Circular response; Von Mises distribution.}
 
@@
 inference from a fitted model.
-For that reason circular data demands for purpose-built statistical methods. 
+For that reason circular data demand for purpose-built statistical methods. 
 One common way of modelling circular data is to apply the von Mises distribution 
 which is often regarded as the circular analog of the normal distribution. 
-Estimating its two distribution parameter by linear functions of regressor variables 
-is a possible approach, however, might not be able to detect abrupt effects.
-As an alternative  we propose the application of distributional trees to
-model wind directions based on a set of regression variables.
+Estimating its two distribution parameters by generalized additive models 
+using a set of regressor variables is one possible approach that can be successful 
+in various settings.
+As an alternative we propose the application of distributional trees 
+which are capable of detecting also non-additive effects and provide easily interpretable 
+models allowing for a wide range of inference procedures.
+Moreover, covariates and their possible interactions are selected automatically
+such that no expert knowledge is required to specify the model in advance.
 
@@
 
-\subsection{von Mises Distribution}
-A commonly chosen distribution to model circular data is the von Misis distribution
-which is also known as the circular normal distribution. Its density function is given by
+\subsection{Von Mises Distribution}
+A commonly chosen distribution to model circular data is the von Mises distribution
+which is also known as the circular normal distribution 
+(Jammalamadaka and Sengupta, 2001; Mardia an dJupp, 2009). 
+Its density function is given by
 \begin{equation}
@@
 provided by generalized additive models for location, scale, and shape 
-(GAMLSS, Rigby et. al, 2005) into the framework of regression trees (Breiman et. al, 1984).
-In particular one of the unbiased recursive partitioning algorithms MOB (Hothorn et. al, 2006) 
-and CTree (Zeileis et. al, 2008) is applied and a distributional model is fitted in each node 
-by estimating the parameters of a specified distribution family via maximum likeklihood.
+(GAMLSS, Rigby and Stasinopoulos, 2005) into the framework of regression trees (Breiman et. al, 1984).
+In particular one of the unbiased recursive partitioning algorithms MOB (Zeileis et. al, 2008) 
+and CTree (Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
+by estimating the parameters of a specified distribution family via maximum likelihood.
 For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
@@
 
-In the here discussed application the prespecified distribution family is the von Mises distribution. 
+In the here discussed application the MOB algorithm is applied and the prespecified 
+distribution family is the von Mises distribution. 
 The corresponding distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
 by maximizing $l(y, \theta) = log(f_{vM}(y;\theta))$ and model scores are calculated 
-by $s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$.
+by $s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. In that way
+a value indicating the goodness of fit is provided for each parameter and observation
+such that changes can be detected not only in the location parameter $\mu$ but also in the 
+dispersion parameter $\kappa$.
 
@@
 While the application of the von Mises distribution allows for modelling a circular response variable,
-additional adaptions have to made when including circular regressor variables in a distributional tree
+additional adaptions have to be made when including circular regressor variables in a distributional tree
 model. In particular, if the selected split variable is a circular variable this demands for finding
-not only one split point but a ``split intervall''. This extension is of interest for many applications,
-for example in order to include day time as a regressor variable when modelling wind directions.
+not only one split point but a ``split interval''. This consideration is of interest for many applications,
+for example when modelling wind directions it might be desirable to include day time or earlier 
+observations of wind directions as regressor variables.
+
+Another possible extension of the here presented methodology is to build not just one tree model but
+an ensemble of trees to regularize and stabilize the model 
+(see e.g. distributional forests, Schlosser et. al, 2018).
 

### 2019-01-30 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Include application part (w/o fig)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2414 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7f9ff234d28d19800e05dfcff65ac1d78f2f7411

@@
 
-\textit{TO DO}
+%\begin{figure}[!ht]\centering
+%\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
+%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
+%\end{figure}
 
-\bigskip
+Wind is a classical circular quantity and accurate forecasts of wind direction
+are of great importance for decision making processes and risk management,
+e.g., in air traffic management and renewable energy production. As a suitable
+and useful application of circular regression trees, we therefore forecast the
+wind direction at the airport of Innsbruck. As potential split variables we
+employ several 1-hourly lagged observed weather variables (e.g., wind speed,
+relative humidity, pressure) and derived spatial and temporal difference of
+these lagged observations.
+
+%Innsbruck lies with the west-east oriented Inn Valley with an approx. 2300\,m
+%high mountain ridge on the northern side and with the south-north oriented Wipp
+%Valley between the approx. 2300\,m high mountains on the southern side.
+%Therefore, skill full covariates are pressure differences along the Inn Valley
+%as indices for up- or down-valley winds. For the typical dry gap wind coming
+%from the south, indices are, e.g. low relative humidity or temperature
+%differences between Innsbruck and stations up the Wipp Valley.
+
+The weather station of interest, Innsbruck Airport, lies at the bottom of the
+Inn Valley, a deep mountain valley in the Eastern European Alps orientated
+parallel to the Alpine crest. At Innsbruck, the Inn Valley is bordered in the
+North with a approx. 2300\,m high mountain ridge and opens in the South into
+the crest-perpendicular Wipp Valley. Therefore, mainly three different wind
+regimes exist: Up-valley winds blowing from the valley mouth towards the upper
+valley (from east to west), down-valley winds blowing in the direction of the
+valley mouth (from west to east), and dry and warm downslope winds blowing
+across the Alpine crest along the Wipp Valley towards Innsbruck (from
+south-east to north-west).
+
+As an illustration of tree based forecasts of wind direction for Innsbruck Airport, a
+model based regression tree of depth two is shown in Fig.~\ref{fig:tree}.
+Winds typically blow from areas of high pressure to areas of low pressure. The
+first chosen split variable is therefore the pressure difference between
+Innsbruck and a station located near the valley mouth. In the left branch of
+the tree, with lower pressure in Innsbruck than at the valley mouth, the second
+split is performed on the pressure difference between Innsbruck a station
+further up the valley. This leads to the final \emph{Node~3} with low pressure
+gradients within the Inn Valley and, therefore, no clear partition in the wind
+direction. The final \emph{Node~4} contains measurements with a strong pressure
+gradient towards the upper Inn Valley leading to up-valley winds from east to
+west. The right branch of the tree is split in relative humidity as an index
+for the dry downslope winds. Therefore, for a low relative humidity winds from
+south-east to north-west are predominant in the final \emph{Node~6}. Whereas,
+the final \emph{Node~7} contains the cases with down-valley winds towards the
+valley mouth in the East.
 

### 2019-01-31 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2415 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ccd205def5da9487d04e8ac86b6c3618a1186e61

@@
 For that reason circular data demand for purpose-built statistical methods. 
-One common way of modelling circular data is to apply the von Mises distribution 
-which is often regarded as the circular analog of the normal distribution. 
-Estimating its two distribution parameters by generalized additive models 
-using a set of regressor variables is one possible approach that can be successful 
-in various settings.
+A commonly chosen distribution for modeling circular data is the von Mises distribution.
+Estimating its two distribution parameters by generalized additive models using a set of 
+regressor variables is one possible approach that can be successful in various settings.
 As an alternative we propose the application of distributional trees 
@@
 
-\subsection{Von Mises Distribution}
-A commonly chosen distribution to model circular data is the von Mises distribution
-which is also known as the circular normal distribution 
-(Jammalamadaka and Sengupta, 2001; Mardia an dJupp, 2009). 
-Its density function is given by
+%\subsection{Von Mises Distribution}
+%The von Mises distribution is also known as the circular normal distribution 
+%and its density function is given by
+%\begin{equation}
+%  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
+%\end{equation}
+%with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
+%modified Bessel function of the first kind and order $0$.
+%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
+%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
+%to the variance $\sigma^2$ of the normal distribution. 
+%See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.
+
+%\subsection{Distributional Trees}
+%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
+%of regression trees.
+Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
+provided by generalized additive models for location, scale, and shape~(GAMLSS, 
+Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
+In particular one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
+and CTree~(Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
+by estimating the parameters of a specified distribution family via maximum likelihood.
+For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
+tests evaluating associations between model scores and each possible split variable. After choosing 
+the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step by optimizing an objective function.
+
+In the here discussed application the MOB algorithm is applied and the prespecified 
+distribution family is the von Mises distribution which is also known as the circular 
+normal distribution. Its density function is given by
 \begin{equation}
@@
 to the variance $\sigma^2$ of the normal distribution. 
-
-\subsection{Distributional Trees}
-Distributional trees (Schlosser et. al, 2018) embed distributional modeling such as
-provided by generalized additive models for location, scale, and shape 
-(GAMLSS, Rigby and Stasinopoulos, 2005) into the framework of regression trees (Breiman et. al, 1984).
-In particular one of the unbiased recursive partitioning algorithms MOB (Zeileis et. al, 2008) 
-and CTree (Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
-by estimating the parameters of a specified distribution family via maximum likelihood.
-For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
-tests applied on model scores and each possible split variable. After choosing the split variable 
-showing the highest association to the goodness of fit of the model the corresponding split point 
-which optimizes an objective function is selected in a separate step.
-
-In the here discussed application the MOB algorithm is applied and the prespecified 
-distribution family is the von Mises distribution. 
-The corresponding distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
-by maximizing $l(y, \theta) = log(f_{vM}(y;\theta))$ and model scores are calculated 
-by $s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. In that way
+(See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.)
+The distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
+by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated 
+by 
+$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \mu},
+\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \kappa}\right)$. 
+%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
+In that way
 a value indicating the goodness of fit is provided for each parameter and observation
@@
 
-%\begin{figure}[!ht]\centering
-%\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
-%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
-%\end{figure}
+\begin{figure}[!ht]\centering
+\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
+\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
+\end{figure}
 
@@
 
+%\bigskip
 \section{Outlook}
 
-While the application of the von Mises distribution allows for modelling a circular response variable,
+While the application of the von Mises distribution allows for modeling a circular response variable,
 additional adaptions have to be made when including circular regressor variables in a distributional tree
@@
 not only one split point but a ``split interval''. This consideration is of interest for many applications,
-for example when modelling wind directions it might be desirable to include day time or earlier 
+for example when modeling wind directions it might be desirable to include day time or earlier 
 observations of wind directions as regressor variables.
@@
      Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
-     {\it working paper}.
+     {\it arXiv.org E-Print Archive}, arXiv 1804.02921, \url{http://arxiv.org/abs/1804.02921}
 

### 2019-01-31 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2417 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 42f98be05571c49204ae08d5ef330278e73d2db0

@@
 % Brief abstract of the paper:
-\abstract{\textit{TO DO}}
+\abstract{Circular response variables occur in a variety of applications, e.g., when
+geographical directions are considered or measurements are taken on a $24$-hour scale. 
+A commonly chosen distribution for modeling circular data is the von Mises distribution.
+As an alternative to estimating its distribution parameters by generalized additive models 
+we propose the application of distributional trees which are capable of detecting also 
+non-additive effects, select covariates and their possible interactions automatically and 
+provide easily interpretable models allowing for a wide range of inference procedures. 
+This approach is illustrated in an application to wind directions with a set of meteorological
+covariates, all measured at the airport of Innsbruck, Austria.}
+
+%which are capable of detecting also non-additive effects and provide easily interpretable 
+%models allowing for a wide range of inference procedures.
+%Moreover, covariates and their possible interactions are selected automatically
+%such that no expert knowledge is required to specify the model in advance.
 
@@
 
-\section{Motivation}
-Circular response variables occur in a variety of applications and subject areas. 
-For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
-animal orientation or gene-structure analysis are often subject of examination in biology, 
-and wind direction is one of the most important weather variables in meteorology.
-Representing such a two-dimensional direction by an angle in a unity circle requires 
-the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
-rotation. However, these settings can vary but should not influence statistical 
-inference from a fitted model.
-For that reason circular data demand for purpose-built statistical methods. 
-A commonly chosen distribution for modeling circular data is the von Mises distribution.
-Estimating its two distribution parameters by generalized additive models using a set of 
-regressor variables is one possible approach that can be successful in various settings.
-As an alternative we propose the application of distributional trees 
-which are capable of detecting also non-additive effects and provide easily interpretable 
-models allowing for a wide range of inference procedures.
-Moreover, covariates and their possible interactions are selected automatically
-such that no expert knowledge is required to specify the model in advance.
+%\section{Motivation}
+%Circular response variables occur in a variety of applications and subject areas. 
+%For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
+%animal orientation or gene-structure analysis are often subject of examination in biology, 
+%and wind direction is one of the most important weather variables in meteorology.
+%Representing such a two-dimensional direction by an angle in a unity circle requires 
+%the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
+%rotation. However, these settings can vary but should not influence statistical 
+%inference from a fitted model.
+%For that reason circular data demand for purpose-built statistical methods. 
+%A commonly chosen distribution for modeling circular data is the von Mises distribution.
+%Estimating its two distribution parameters by generalized additive models using a set of 
+%regressor variables is one possible approach that can be successful in various settings.
+%As an alternative we propose the application of distributional trees 
+%which are capable of detecting also non-additive effects and provide easily interpretable 
+%models allowing for a wide range of inference procedures.
+%Moreover, covariates and their possible interactions are selected automatically
+%such that no expert knowledge is required to specify the model in advance.
 
@@
 %\subsection{Distributional Trees}
-%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
-%of regression trees.
-Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
-provided by generalized additive models for location, scale, and shape~(GAMLSS, 
-Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
+Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
+of regression trees.
+%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
+%provided by generalized additive models for location, scale, and shape~(GAMLSS, 
+%Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
 In particular one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
@@
 \end{equation}
-with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
+with location parameter $\mu$, dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
 modified Bessel function of the first kind and order $0$.
-For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
-a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
-to the variance $\sigma^2$ of the normal distribution. 
-(See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.)
-The distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
-by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated 
-by 
-$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \mu},
-\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \kappa}\right)$. 
+%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
+%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
+%to the variance $\sigma^2$ of the normal distribution. 
+(See e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview.)
+
+In each node the distribution parameters $\mu$ and $\kappa$ are 
+%vector $\theta = (\mu,\kappa)$ is 
+estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated by 
+$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
+\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$. 
 %$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
-In that way
-a value indicating the goodness of fit is provided for each parameter and observation
+In that way a value indicating the goodness of fit is provided for each parameter and observation
 such that changes can be detected not only in the location parameter $\mu$ but also in the 
@@
 
-\begin{figure}[!ht]\centering
-\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
-\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
-\end{figure}
+%\begin{figure}
+%\includegraphics[height = 0.5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
+%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises %distribution.}\label{fig:tree}
+%\end{figure}
 
@@
 
-%Innsbruck lies with the west-east oriented Inn Valley with an approx. 2300\,m
-%high mountain ridge on the northern side and with the south-north oriented Wipp
-%Valley between the approx. 2300\,m high mountains on the southern side.
-%Therefore, skill full covariates are pressure differences along the Inn Valley
-%as indices for up- or down-valley winds. For the typical dry gap wind coming
-%from the south, indices are, e.g. low relative humidity or temperature
-%differences between Innsbruck and stations up the Wipp Valley.
+Innsbruck lies with the west-east oriented Inn Valley with an approx. 2300\,m
+high mountain ridge on the northern side and with the south-north oriented Wipp
+Valley between the approx. 2300\,m high mountains on the southern side.
+Therefore, skill full covariates are pressure differences along the Inn Valley
+as indices for up- or down-valley winds. For the typical dry gap wind coming
+from the south, indices are, e.g. low relative humidity or temperature
+differences between Innsbruck and stations up the Wipp Valley.
 
@@
 
-While the application of the von Mises distribution allows for modeling a circular response variable,
-additional adaptions have to be made when including circular regressor variables in a distributional tree
-model. In particular, if the selected split variable is a circular variable this demands for finding
-not only one split point but a ``split interval''. This consideration is of interest for many applications,
-for example when modeling wind directions it might be desirable to include day time or earlier 
-observations of wind directions as regressor variables.
+This application shows that including covariates such as day time or earlier observations 
+of wind directions as (unordered) factors provides valuable information. However, in order 
+to properly account not only for the circularity of the response but also of regressors, 
+additional adaptions of distributional trees are required. 
+In particular, if the selected split variable is a circular variable this demands for 
+finding a ``split interval'' instead of a split point.
+%While the application of the von Mises distribution allows for modeling a circular response variable,
+%additional adaptions have to be made when including circular regressor variables in a distributional tree
+%model. In particular, if the selected split variable is a circular variable this demands for finding
+%not only one split point but a ``split interval''. This consideration is of interest for many applications,
+%for example when modeling wind directions it might be desirable to include day time or earlier 
+%observations of wind directions as regressor variables.
 
-Another possible extension of the here presented methodology is to build not just one tree model but
-an ensemble of trees to regularize and stabilize the model 
+Another possible extension of the here presented methodology is to build not just one tree 
+model but an ensemble of trees to regularize and stabilize the model 
 (see e.g. distributional forests, Schlosser et. al, 2018).
@@
 
-\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
-     Classification and Regression Trees.
-     {\it Wadsworth}.
+%\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
+%     Classification and Regression Trees.
+%     {\it Wadsworth}.
 
@@
 
-\item[Mardia, K. V., and Jupp, P. E.] (2009).
-     Directional Statistics.
-     {\it John Wiley \& Sons}, {\bf 494}.
+%\item[Mardia, K. V., and Jupp, P. E.] (2009).
+%     Directional Statistics.
+%     {\it John Wiley \& Sons}, {\bf 494}.
 
-\item[Rigby, R. A., and Stasinopoulos, D. M.] (2005).
-     Generalized Additive Models for Location Scale and Shape (with Discussion).
-     {\it Journal of the Royal Statistical Society C}, {\bf 54}, 3,
-     507\,--\,554.
+%\item[Rigby, R. A., and Stasinopoulos, D. M.] (2005).
+%     Generalized Additive Models for Location Scale and Shape (with Discussion).
+%     {\it Journal of the Royal Statistical Society C}, {\bf 54}, 3,
+%     507\,--\,554.
 
@@
      Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
-     {\it arXiv.org E-Print Archive}, arXiv 1804.02921, \url{http://arxiv.org/abs/1804.02921}
+     {\it arXiv.org E-Print Archive}, https://arxiv.org/abs/1804.02921v2.
 

### 2019-01-31 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Merge conflict
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2418 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 93087120b186ea5ca5a2be2365f3a0f4085b54f5
Update application to old version
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2419 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: efa02a275e938382e721d33da40155e69efa2e74

@@
 %\begin{figure}
-%\includegraphics[height = 0.5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises %distribution.}\label{fig:tree}
+%\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
+%\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
+%\label{fig:tree}
 %\end{figure}
@@
 wind direction at the airport of Innsbruck. As potential split variables we
-employ several 1-hourly lagged observed weather variables (e.g., wind speed,
-relative humidity, pressure) and derived spatial and temporal difference of
-these lagged observations.
-
-Innsbruck lies with the west-east oriented Inn Valley with an approx. 2300\,m
-high mountain ridge on the northern side and with the south-north oriented Wipp
-Valley between the approx. 2300\,m high mountains on the southern side.
-Therefore, skill full covariates are pressure differences along the Inn Valley
-as indices for up- or down-valley winds. For the typical dry gap wind coming
-from the south, indices are, e.g. low relative humidity or temperature
-differences between Innsbruck and stations up the Wipp Valley.
+employ several 1-hourly lagged observed weather variables (e.g., wind
+direction, wind speed, relative humidity, pressure) and derived spatial and
+temporal difference of these lagged observations.
 
@@
 Inn Valley, a deep mountain valley in the Eastern European Alps orientated
-parallel to the Alpine crest. At Innsbruck, the Inn Valley is bordered in the
+parallel to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered in the
 North with a approx. 2300\,m high mountain ridge and opens in the South into
@@
 
-%\bigskip
 \section{Outlook}

### 2019-01-31 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Adapt application to larger tree (first draft)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2420 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b78cdd4a3f32c689a6e40362553f09643946a09b

@@
 and useful application of circular regression trees, we therefore forecast the
-wind direction at the airport of Innsbruck. As potential split variables we
-employ several 1-hourly lagged observed weather variables (e.g., wind
-direction, wind speed, relative humidity, pressure) and derived spatial and
-temporal difference of these lagged observations.
+wind direction at the airport of Innsbruck. Innsbruck lies at the bottom of the
+Inn Valley, a deep mountain valley in the Eastern European Alps orientated
+parallel to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is
+bordered in the North with a approx. 2300\,m high mountain ridge and opens in
+the South into the crest-perpendicular Wipp Valley. As potential split
+variables we employ several 1-hourly lagged observed weather variables (e.g.,
+wind direction, wind speed, relative humidity, pressure gradients along the
+valley) and derived spatial and temporal difference of these lagged
+observations.
 
-The weather station of interest, Innsbruck Airport, lies at the bottom of the
-Inn Valley, a deep mountain valley in the Eastern European Alps orientated
-parallel to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered in the
-North with a approx. 2300\,m high mountain ridge and opens in the South into
-the crest-perpendicular Wipp Valley. Therefore, mainly three different wind
-regimes exist: Up-valley winds blowing from the valley mouth towards the upper
-valley (from east to west), down-valley winds blowing in the direction of the
-valley mouth (from west to east), and dry and warm downslope winds blowing
-across the Alpine crest along the Wipp Valley towards Innsbruck (from
-south-east to north-west).
-
-As an illustration of tree based forecasts of wind direction for Innsbruck Airport, a
-model based regression tree of depth two is shown in Fig.~\ref{fig:tree}.
-Winds typically blow from areas of high pressure to areas of low pressure. The
-first chosen split variable is therefore the pressure difference between
-Innsbruck and a station located near the valley mouth. In the left branch of
-the tree, with lower pressure in Innsbruck than at the valley mouth, the second
-split is performed on the pressure difference between Innsbruck a station
-further up the valley. This leads to the final \emph{Node~3} with low pressure
-gradients within the Inn Valley and, therefore, no clear partition in the wind
-direction. The final \emph{Node~4} contains measurements with a strong pressure
-gradient towards the upper Inn Valley leading to up-valley winds from east to
-west. The right branch of the tree is split in relative humidity as an index
-for the dry downslope winds. Therefore, for a low relative humidity winds from
-south-east to north-west are predominant in the final \emph{Node~6}. Whereas,
-the final \emph{Node~7} contains the cases with down-valley winds towards the
-valley mouth in the East.
+As an illustration for the tree based forecast performance of predicting wind
+directions at Innsbruck airport, a BIC based post-pruned regression tree of
+depth three is shown in Fig.~\ref{fig:tree}. According to the fitted von Mises
+distributions three distinct wind regimes: Firstly, up-valley winds blowing
+from the valley mouth towards the upper valley (from east to west) as shown in
+the final nodes 4 and 5. Secondly, downslope winds blowing across the Alpine
+crest along the Wipp Valley towards Innsbruck (from south-east to north-west)
+as predicted in nodes 7 and 8. Thirdly, down-valley winds blowing in the
+direction of the valley mouth (from west to east) present in nodes 12, 14, and
+15. The final node 11 with a density parameter $\kappa$ of $0.84$ captures the
+cases with rather low wind speeds and therefore no clear wind direction.
+Generally, the splits in the lagged wind direction perform a clustering into
+the different wind regimes. As winds typically blow from areas of high pressure
+to areas of low pressure and have larger wind speeds in case of stronger
+pressure gradients, these respective splits in pressure gradients and wind
+speed divide the predictions according to the stability of the lagged wind
+direction. As an example node 3 differentiates in the pressure difference along
+the valley, and the cases with a stronger pressure gradient have a higher
+fitted $\kappa$ as indicator for the stability of the forecast. In other words,
+in final nodes with a higher $\kappa$ most training cases are classified
+correctly and therefore the density parameter is an indicator for the
+uncertainty of the forecast.
 

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Adapt application
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2421 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 479ef4939d3a38033574fef93486317a6a839943

@@
 
-%\begin{figure}
-%\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-%\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
-%\label{fig:tree}
-%\end{figure}
+\begin{figure}
+\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
+\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
+\label{fig:tree}
+\end{figure}
 
@@
 
-As an illustration for the tree based forecast performance of predicting wind
-directions at Innsbruck airport, a BIC based post-pruned regression tree of
-depth three is shown in Fig.~\ref{fig:tree}. According to the fitted von Mises
-distributions three distinct wind regimes: Firstly, up-valley winds blowing
-from the valley mouth towards the upper valley (from east to west) as shown in
-the final nodes 4 and 5. Secondly, downslope winds blowing across the Alpine
-crest along the Wipp Valley towards Innsbruck (from south-east to north-west)
-as predicted in nodes 7 and 8. Thirdly, down-valley winds blowing in the
-direction of the valley mouth (from west to east) present in nodes 12, 14, and
-15. The final node 11 with a density parameter $\kappa$ of $0.84$ captures the
-cases with rather low wind speeds and therefore no clear wind direction.
-Generally, the splits in the lagged wind direction perform a clustering into
-the different wind regimes. As winds typically blow from areas of high pressure
-to areas of low pressure and have larger wind speeds in case of stronger
-pressure gradients, these respective splits in pressure gradients and wind
-speed divide the predictions according to the stability of the lagged wind
-direction. As an example node 3 differentiates in the pressure difference along
-the valley, and the cases with a stronger pressure gradient have a higher
-fitted $\kappa$ as indicator for the stability of the forecast. In other words,
-in final nodes with a higher $\kappa$ most training cases are classified
-correctly and therefore the density parameter is an indicator for the
-uncertainty of the forecast.
+As an illustration of the tree based wind direction forecasts at Innsbruck airport, a
+BIC based post-pruned regression tree of depth three is shown in
+Fig.~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
+nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
+valley (from east to west, as predicted in the final nodes 4 and 5). Secondly,
+downslope winds blowing across the Alpine crest along the Wipp Valley towards
+Innsbruck (from south-east to north-west, as predicted in nodes 7 and 8).
+Thirdly, down-valley winds blowing in the direction of the valley mouth (from
+west to east, as predicted in nodes 12, 14, and 15). The final node 11 with a
+concentration parameter $\kappa$ of only $0.84$ captures the cases with rather low
+wind speeds and typically no clear wind direction or distinct wind regime.
+In general, winds typically blow from areas of high pressure to areas of low
+pressure, and a strong pressure gradient leads to large wind speeds and a stable
+wind direction. Therefore, the splits in the lagged wind direction perform a
+clustering into the different wind regimes and the further splits in pressure
+gradients or wind speed divide the predictions according to the consistency of
+the lagged wind direction. This can be seen in the final nodes with strong
+pressure gradients or high wind speeds where most training cases are classified
+correctly and therefore the concentration parameter as an indicator of the
+certainty of the forecast is rather large.
 

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Include Reto's input (application)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2422 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 447dfddb4da7617048853054c6ca37a7168c18e4

@@
 
-\begin{figure}
-\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
-\label{fig:tree}
-\end{figure}
+%\begin{figure}
+%\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
+%\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
+%\label{fig:tree}
+%\end{figure}
 
@@
 are of great importance for decision making processes and risk management,
-e.g., in air traffic management and renewable energy production. As a suitable
-and useful application of circular regression trees, we therefore forecast the
-wind direction at the airport of Innsbruck. Innsbruck lies at the bottom of the
-Inn Valley, a deep mountain valley in the Eastern European Alps orientated
-parallel to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is
-bordered in the North with a approx. 2300\,m high mountain ridge and opens in
-the South into the crest-perpendicular Wipp Valley. As potential split
-variables we employ several 1-hourly lagged observed weather variables (e.g.,
-wind direction, wind speed, relative humidity, pressure gradients along the
-valley) and derived spatial and temporal difference of these lagged
-observations.
+e.g., in air traffic management or renewable energy production. This study
+presents an application of circular regression trees  on hourly direction
+forecasts at Innsbruck airport.  Innsbruck lies at the bottom of the Inn
+Valley, a deep mountain valley in the Eastern European Alps orientated parallel
+to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
+in the North with a approx. 2300\,m high mountain ridge and opens to the South
+into the crest-perpendicular Wipp Valley. As potential split variables we
+employ several 1-hourly lagged observed weather variables (e.g., wind
+direction, wind speed, relative humidity, pressure gradients along the valley)
+and derived spatial and temporal difference of these lagged observations.
 
-As an illustration of the tree based wind direction forecasts at Innsbruck airport, a
-BIC based post-pruned regression tree of depth three is shown in
+As an illustration of the tree based wind direction forecasts at Innsbruck
+airport, a BIC based post-pruned regression tree of depth three is shown in
 Fig.~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
 nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
-valley (from east to west, as predicted in the final nodes 4 and 5). Secondly,
-downslope winds blowing across the Alpine crest along the Wipp Valley towards
-Innsbruck (from south-east to north-west, as predicted in nodes 7 and 8).
-Thirdly, down-valley winds blowing in the direction of the valley mouth (from
-west to east, as predicted in nodes 12, 14, and 15). The final node 11 with a
-concentration parameter $\kappa$ of only $0.84$ captures the cases with rather low
-wind speeds and typically no clear wind direction or distinct wind regime.
-In general, winds typically blow from areas of high pressure to areas of low
-pressure, and a strong pressure gradient leads to large wind speeds and a stable
-wind direction. Therefore, the splits in the lagged wind direction perform a
-clustering into the different wind regimes and the further splits in pressure
-gradients or wind speed divide the predictions according to the consistency of
-the lagged wind direction. This can be seen in the final nodes with strong
-pressure gradients or high wind speeds where most training cases are classified
-correctly and therefore the concentration parameter as an indicator of the
-certainty of the forecast is rather large.
+valley (from east to west, nodes 4 and 5). Secondly, downslope winds blowing
+across the Alpine crest along the Wipp Valley towards Innsbruck (from
+south-east to north-west, nodes 7 and 8). Thirdly, down-valley winds blowing
+in the direction of the valley mouth (from west to east, nodes 12, 14, and 15).
+The final node 11 with a concentration parameter $\kappa$ of only $0.84$
+captures the cases with rather low wind speeds which typically don't show a
+clear wind direction or distinct wind regime. In general, winds typically blow
+from areas of high pressure to areas of low pressure where strong pressure
+gradients yield high wind speeds and a rather well-defined wind direction.
+Thus, the splits in the lagged wind direction perform a clustering into the
+different wind regimes and the additional splits in pressure gradients or wind
+speed divide the predictions according to the expected consistency of the wind
+direction. This can be seen in the final nodes with strong pressure gradients
+or high wind speeds where most training cases are classified correctly and
+therefore the concentration parameter as an indicator of the certainty of the
+forecast is rather large.
 

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2423 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 734ae7b5450a9cbbef3a325e1bd648788ea7fdf5

@@
 % Brief abstract of the paper:
-\abstract{Circular response variables occur in a variety of applications, e.g., when
-geographical directions are considered or measurements are taken on a $24$-hour scale. 
+\abstract{
+%Circular response variables occur in a variety of applications, e.g., when
+%geographical directions are considered or measurements are taken on a $24$-hour scale. 
 A commonly chosen distribution for modeling circular data is the von Mises distribution.
@@
 covariates, all measured at the airport of Innsbruck, Austria.}
-
-%which are capable of detecting also non-additive effects and provide easily interpretable 
-%models allowing for a wide range of inference procedures.
-%Moreover, covariates and their possible interactions are selected automatically
-%such that no expert knowledge is required to specify the model in advance.
 
@@
 %Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
-In particular one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
-and CTree~(Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
-by estimating the parameters of a specified distribution family via maximum likelihood.
+The basic idea is to partition the covariate space recursively into subgroups 
+such that a homogeneous distributional model can be fit to the response in each resulting subgroup.
+In particular, one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
+and CTree~(Hothorn et. al, 2006) is applied and in each node the parameters of a specified 
+distribution family are estimated via maximum likelihood.
 For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
 tests evaluating associations between model scores and each possible split variable. After choosing 
-the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step by optimizing an objective function.
+the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step, either by optimizing the log-likelihood function (MOB)
+or by evaluating a linear two-sample statistic (CTree) over all possible partitions.
 
@@
 
-In each node the distribution parameters $\mu$ and $\kappa$ are 
+In each node of the tree the distribution parameters $\mu$ and $\kappa$ are 
 %vector $\theta = (\mu,\kappa)$ is 
-estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated by 
+estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$. 
+For the resulting estimators $\hat{\mu}$ and $\hat{\kappa}$ model scores are calculated by 
 $s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
-\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$. 
+\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$
+yielding a $n \times 2$-matrix for a data set of $n$ observations.
 %$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
@@
 
-%\begin{figure}
-%\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-%\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
-%\label{fig:tree}
-%\end{figure}
+\begin{figure}[!ht]\centering
+\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
+\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
+\label{fig:tree}
+\end{figure}
 
@@
 
-\section{Outlook}
+\section{Discussion / Summary and Outlook}
+Applying distributional trees on circular data provides easily interpretable models clearly
+illustrating the impact of given covariates and their possible interactions on the distribution
+of the dependent variable. Moreover, covariates are selected automatically such that no 
+prespecification of the model is required which can be a challenging task, particularly for a 
+meteorlogical quantity such as wind direction with high variation and many possible influencing 
+factors.
 
-This application shows that including covariates such as day time or earlier observations 
-of wind directions as (unordered) factors provides valuable information. However, in order 
-to properly account not only for the circularity of the response but also of regressors, 
-additional adaptions of distributional trees are required. 
-In particular, if the selected split variable is a circular variable this demands for 
-finding a ``split interval'' instead of a split point.
+This application shows that including covariates such as lagged observations of wind direction
+as linear regressors provides valuable information. However, in order to properly account not 
+only for the circularity of the response but also of regressors, additional adaptions of 
+distributional trees are required. In particular, if the selected split variable is a circular 
+variable this demands for finding a ``split interval'' instead of a split point.
+This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
+the linear regressor \textit{dd} (lagged wind direction) are required in order to represent its effect.
+
 %While the application of the von Mises distribution allows for modeling a circular response variable,
@@
 model but an ensemble of trees to regularize and stabilize the model 
-(see e.g. distributional forests, Schlosser et. al, 2018).
+(see e.g. distributional forests, Schlosser et. al, 2018). In that way, it should not have a 
+significant impact on the results whether circular covariates are included as linear regressors
+yielding a larger number of splits or as circular regressors with ``intervall splits''.
 

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2424 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: af56616cddd904fb55a07715460f4b2eb289ca1c

@@
 provide easily interpretable models allowing for a wide range of inference procedures. 
-This approach is illustrated in an application to wind directions with a set of meteorological
-covariates, all measured at the airport of Innsbruck, Austria.}
+This approach is illustrated by an application of distributional trees on hourly wind 
+direction forecasts with a set of meteorological covariates, all measured at the airport 
+of Innsbruck, Austria.
+}
 
@@
 %\end{equation}
-%with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
+%with a location parameter $\mu$, a conentration parameter $\kappa$, and $I_0(\kappa)$ being the 
 %modified Bessel function of the first kind and order $0$.
@@
 
-In the here discussed application the MOB algorithm is applied and the prespecified 
-distribution family is the von Mises distribution which is also known as the circular 
+In this study distributional trees are applied to circular response data employing the MOB algorithm
+and the von Mises distribution which is also known as the circular 
 normal distribution. Its density function is given by
@@
 \end{equation}
-with location parameter $\mu$, dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
-modified Bessel function of the first kind and order $0$.
+with location parameter $\mu$, conentration parameter $\kappa$, and $I_0(\kappa)$ being the 
+modified Bessel function of the first kind and order $0$
 %For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
@@
 %to the variance $\sigma^2$ of the normal distribution. 
-(See e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview.)
+(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).
 
@@
 For the resulting estimators $\hat{\mu}$ and $\hat{\kappa}$ model scores are calculated by  
-$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
-\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$
+%$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
+%\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$
+$s(y,\hat{\mu}, \hat{\kappa}) = (\partial_{\mu} l(y, \hat{\mu}, \hat{\kappa}),
+\partial_{\kappa}l(y, \hat{\mu}, \hat{\kappa}))$
 yielding a $n \times 2$-matrix for a data set of $n$ observations.
 %$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
-In that way a value indicating the goodness of fit is provided for each parameter and observation
-such that changes can be detected not only in the location parameter $\mu$ but also in the 
-dispersion parameter $\kappa$.
+In that way for each parameter a value indicating the goodness of fit with respect to one single 
+observation is provided such that changes can be detected not only in the location parameter $\mu$ 
+but also in the concentration parameter $\kappa$.
 
@@
 airport, a BIC based post-pruned regression tree of depth three is shown in
-Fig.~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
+Figure~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
 nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
@@
 \section{Discussion / Summary and Outlook}
-Applying distributional trees on circular data provides easily interpretable models clearly
-illustrating the impact of given covariates and their possible interactions on the distribution
-of the dependent variable. Moreover, covariates are selected automatically such that no 
-prespecification of the model is required which can be a challenging task, particularly for a 
-meteorlogical quantity such as wind direction with high variation and many possible influencing 
-factors.
+This study shows that applying distributional trees on circular data provides easily 
+interpretable models clearly illustrating the impact of given covariates and their possible 
+interactions on the distribution of the dependent variable. Moreover, covariates are selected
+automatically such that no prespecification of the model is required which can be a challenging 
+task, particularly for a meteorlogical quantity such as wind direction with high variation and 
+many possible influencing factors.
 
-This application shows that including covariates such as lagged observations of wind direction
-as linear regressors provides valuable information. However, in order to properly account not 
-only for the circularity of the response but also of regressors, additional adaptions of 
-distributional trees are required. In particular, if the selected split variable is a circular 
-variable this demands for finding a ``split interval'' instead of a split point.
+Based on the results it can be assumed that including covariates such as lagged observations of 
+wind direction as linear regressors provides valuable information. 
+However, in order to properly account not only for the circular properties of the response but 
+also of regressors, additional adaptions of distributional trees are required. In particular, 
+if the selected split variable is a circular variable, e.g. wind direction, this demands for finding
+two split point defining a ``split interval'' instead of a single split point.
 This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
-the linear regressor \textit{dd} (lagged wind direction) are required in order to represent its effect.
+the linearly employed regressor \textit{dd} (lagged wind direction) are required in order to 
+cluster the training data in corresponding wind directions.
 
@@
 
-Another possible extension of the here presented methodology is to build not just one tree 
-model but an ensemble of trees to regularize and stabilize the model 
-(see e.g. distributional forests, Schlosser et. al, 2018). In that way, it should not have a 
-significant impact on the results whether circular covariates are included as linear regressors
-yielding a larger number of splits or as circular regressors with ``intervall splits''.
+Another possible extension of the method as presented is to build an ensemble of 
+distributional trees to regularize and stabilize the model 
+(see, e.g., distributional forests, Schlosser et. al, 2018). 
+%In that way, it should not have a 
+%significant impact on the results whether circular covariates are included as linear regressors
+%yielding a larger number of splits or as circular regressors with ``intervall splits''.
 

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
some rewriting in abstract/methodology
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2426 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d9bbe07d4966a972d86026540fe7914f4426cdeb

@@
 % Include city and country of each institute, do not include the full address.
-\institute{University of Innsbruck, Austria
-\and University of Zurich, Switzerland}
+\institute{Universit\"at Innsbruck, Austria
+\and Universit\"at Z\"urich, Switzerland}
 
@@
 \abstract{
-%Circular response variables occur in a variety of applications, e.g., when
-%geographical directions are considered or measurements are taken on a $24$-hour scale.
-A commonly chosen distribution for modeling circular data is the von Mises distribution.
-As an alternative to estimating its distribution parameters by generalized additive models 
-we propose the application of distributional trees which are capable of detecting also 
-non-additive effects, select covariates and their possible interactions automatically and 
-provide easily interpretable models allowing for a wide range of inference procedures. 
-This approach is illustrated by an application of distributional trees on hourly wind 
-direction forecasts with a set of meteorological covariates, all measured at the airport 
-of Innsbruck, Austria.
+For probabilistic modeling of circular data the von Mises distribution is widely used.
+To capture how its parameters change with covariates, a regression tree model
+is proposed as an alternative to more commonly-used additive models. The resulting
+distributional trees are easy to interpret, can detect non-additive effects, and select
+covariates along with interactions automatically. For illustration, hourly wind 
+direction forecasts are obtained at Innsbruck Airport based on a set of meteorological
+measurements.
 }
@@
 %\begin{equation}
-%  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
+%  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
 %\end{equation}
@@
 
-%\subsection{Distributional Trees}
-Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
-of regression trees.
-%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
-%provided by generalized additive models for location, scale, and shape~(GAMLSS, 
-%Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
-The basic idea is to partition the covariate space recursively into subgroups 
-such that a homogeneous distributional model can be fit to the response in each resulting subgroup.
-In particular, one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
-and CTree~(Hothorn et. al, 2006) is applied and in each node the parameters of a specified 
-distribution family are estimated via maximum likelihood.
-For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
-tests evaluating associations between model scores and each possible split variable. After choosing 
-the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step, either by optimizing the log-likelihood function (MOB)
-or by evaluating a linear two-sample statistic (CTree) over all possible partitions.
+Distributional trees~(Schlosser et.~al, 2018) fuse distributional regression modeling with
+regression trees based on the unbiased recursive partitioning algorithms MOB~(Zeileis et.~al, 2008)
+or CTree~(Hothorn et.~al, 2006). The basic idea is to partition the covariate space recursively into subgroups 
+such that an (approximately) homogeneous distributional model can be fitted to the response in each resulting subgroup.
+To capture dependence on covariates, the association between the model's scores and each available
+covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
+In each partitioning step, the covariate with the highest significant association (i.e., lowest
+significant $p$-value, if any) is selected for splitting the data. The corresponding split point
+is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
+over all possible partitions.
 
-In this study distributional trees are applied to circular response data employing the MOB algorithm
-and the von Mises distribution which is also known as the circular 
-normal distribution. Its density function is given by
+In this study distributional trees are adapted to circular responses by employing the von Mises
+distribution, also known as ``the circular normal distribution''. Based on a location parameter
+$\mu \in \mathbb{R}$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2 \pi]$
+is given by:
 \begin{equation}
-  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
+  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
 \end{equation}
-with location parameter $\mu$, conentration parameter $\kappa$, and $I_0(\kappa)$ being the 
-modified Bessel function of the first kind and order $0$
-%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
-%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
-%to the variance $\sigma^2$ of the normal distribution. 
+where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
 (see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).
 
-In each node of the tree the distribution parameters $\mu$ and $\kappa$ are 
-%vector $\theta = (\mu,\kappa)$ is 
-estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$. 
-For the resulting estimators $\hat{\mu}$ and $\hat{\kappa}$ model scores are calculated by  
-%$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
-%\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$
-$s(y,\hat{\mu}, \hat{\kappa}) = (\partial_{\mu} l(y, \hat{\mu}, \hat{\kappa}),
-\partial_{\kappa}l(y, \hat{\mu}, \hat{\kappa}))$
-yielding a $n \times 2$-matrix for a data set of $n$ observations.
-%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
-In that way for each parameter a value indicating the goodness of fit with respect to one single 
-observation is provided such that changes can be detected not only in the location parameter $\mu$ 
-but also in the concentration parameter $\kappa$.
-
-%For each possible split variable $Z$ the association between the scores and $Z$ is evaluated by a permutation or M-fluctuation test such that the resulting $p$-values allow for an unbiased split variable selection on a unified scale.
-The variable showing the highest association to the scores is chosen as a split variable.
-The corresponding split point is then selected based on a greedy search over all possible split points.
-After splitting the data into subgroups the procedure is repeated in each of the resulting child nodes until eather a stopping criteria is reached or only equal observations are left in a node.
+In each subgroup maximum likelihood estimators $\hat \mu$ and $\hat \kappa$ are obtained
+by maximizing the corresponding log-likelihood $\ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))$. 
+The model scores are calculated by  $s(y; \mu, \kappa) = (\partial_{\mu} \ell(\mu, \kappa; y),
+\partial_{\kappa} \ell(\mu, \kappa; y))$. In a subgroup of size $n$, evaluating the scores
+at the individual observations and parameter estimates $s(y_i; \hat{\mu}, \hat{\kappa})$
+yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
+a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, respectively.
+Hence MOB can test for parameter instabilities in the model by assessing whether the scores
+change along with the available covariates. If so, by maximimizing a partitioned likelihood
+the parameter instabilities are incorporated into the model. This procedure is repeated recursively
+until there are no significant parameter instabilities or until a another stopping criterion
+is met (e.g., concerning subgroup sizes or depth of the tree).
 
@@
 presents an application of circular regression trees  on hourly direction
-forecasts at Innsbruck airport.  Innsbruck lies at the bottom of the Inn
+forecasts at Innsbruck Airport.  Innsbruck lies at the bottom of the Inn
 Valley, a deep mountain valley in the Eastern European Alps orientated parallel
@@
 distributional trees to regularize and stabilize the model 
-(see, e.g., distributional forests, Schlosser et. al, 2018). 
+(see, e.g., distributional forests, Schlosser et.~al, 2018). 
 %In that way, it should not have a 

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2427 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d37dc5fa62a5fde539e2ca4d2744abd872207fd4

@@
 regression trees based on the unbiased recursive partitioning algorithms MOB~(Zeileis et.~al, 2008)
-or CTree~(Hothorn et.~al, 2006). The basic idea is to partition the covariate space recursively into subgroups 
-such that an (approximately) homogeneous distributional model can be fitted to the response in each resulting subgroup.
+or CTree~(Hothorn et.~al, 2006). The basic idea is to partition the covariate space recursively into 
+subgroups such that an (approximately) homogeneous distributional model can be fitted to the response 
+in each resulting subgroup.
 To capture dependence on covariates, the association between the model's scores and each available
@@
 yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
-a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, respectively.
+a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
+respectively.
 Hence MOB can test for parameter instabilities in the model by assessing whether the scores
@@
 the parameter instabilities are incorporated into the model. This procedure is repeated recursively
-until there are no significant parameter instabilities or until a another stopping criterion
+until there are no significant parameter instabilities or until another stopping criterion
 is met (e.g., concerning subgroup sizes or depth of the tree).
@@
 
-\section{Discussion / Summary and Outlook}
+\section{Discussion and Outlook}
 This study shows that applying distributional trees on circular data provides easily 
 interpretable models clearly illustrating the impact of given covariates and their possible 
-interactions on the distribution of the dependent variable. Moreover, covariates are selected
-automatically such that no prespecification of the model is required which can be a challenging 
-task, particularly for a meteorlogical quantity such as wind direction with high variation and 
-many possible influencing factors.
+interactions on the distribution of the dependent variable. 
+Moreover, covariates are selected automatically such that no prespecification of the model is 
+required which can be a challenging task, particularly for a meteorlogical quantity such as wind 
+direction with high variation and many possible influencing factors.
+%Another valuable feature is the capability of capturing also non-additive and abrupt changes, 
+Additionally, abrupt changes are captured, e.g., when the impact of a covariate such as 
+\textit{air pressure difference} determines either up- or down-valley wind directions.
 
-Based on the results it can be assumed that including covariates such as lagged observations of 
-wind direction as linear regressors provides valuable information. 
-However, in order to properly account not only for the circular properties of the response but 
-also of regressors, additional adaptions of distributional trees are required. In particular, 
-if the selected split variable is a circular variable, e.g. wind direction, this demands for finding
-two split point defining a ``split interval'' instead of a single split point.
-This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
-the linearly employed regressor \textit{dd} (lagged wind direction) are required in order to 
-cluster the training data in corresponding wind directions.
+A possible extension of the method as presented is to build an ensemble of 
+distributional trees to regularize and stabilize the model 
+(see, e.g., distributional forests, Schlosser et.~al, 2018). 
+%In that way, it should not have a 
+%significant impact on the results whether circular covariates are included as linear regressors
+%yielding a larger number of splits or as circular regressors with ``intervall splits''.
+
+Note that the circular variable $dd$ (lagged observations of wind) is employed as a linear
+regressor. Therefore, several splits are performed in $dd$ in order to properly cluster the training 
+data in corresponding wind directions (see Figure~\ref{fig:tree}). 
+In this case searching for two split points defining a split interval on the circular scale instead 
+of a single split point on the linear scale could reduce the complexity of the model and possibly 
+improve the results.
+%Alternatively, in the case of a circular covariate 
+%the tree building algorithm could search for two split points defining a split interval instead of 
+%a single split point. This would reduce the complexity of the model and possibly improve the results.
+%Based on the results it can be assumed that including covariates such as lagged observations of 
+%wind direction as linear regressors provides valuable information. 
+%However, in order to properly account not only for the circular properties of the response but 
+%also of regressors, additional adaptions of distributional trees are required. In particular, 
+%if the selected split variable is a circular variable, e.g. wind direction, this demands for finding
+%two split point defining a ``split interval'' instead of a single split point.
+%This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
+%the linearly employed regressor \textit{dd} (lagged wind direction) are required in order to 
+%cluster the training data in corresponding wind directions.  
+\textit{Computational Details:} The employed \textsf{R} packages are available on \textsf{R}-Forge at
+https://R-Forge.R-project.org/projects/partykit/.
 
@@
 %observations of wind directions as regressor variables.
-
-Another possible extension of the method as presented is to build an ensemble of 
-distributional trees to regularize and stabilize the model 
-(see, e.g., distributional forests, Schlosser et.~al, 2018). 
-%In that way, it should not have a 
-%significant impact on the results whether circular covariates are included as linear regressors
-%yielding a larger number of splits or as circular regressors with ``intervall splits''.
 

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Update application and include caption (first draft)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2428 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3435cd96e290c5557cc826f003664d337ed4b705

@@
 \includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
-\label{fig:tree}
-\end{figure}
+
+\caption{Fitted distribution tree for a circular response employing the von
+Mises distribution. For each terminal node, the empirical histogram and the
+fitted density distribution is displayed in gray and red, respectively. The
+location parameters are depicted as red hands and convey the direction from
+which the wind is blowing. The chosen split variables are wind direction
+(dd~[$\text{rad}$]), wind speed (ff~[$\text{ms}^{-1}$]), and pressure gradients
+along the upper (pdiff\_up~[$\text{hPa}$]) and lower
+(pdiff\_low[$\text{hPa}$]) Inn Valley where positive values point in the
+direction of Innsbruck.}
+\label{fig:tree} \end{figure}
 
@@
 into the crest-perpendicular Wipp Valley. As potential split variables we
-employ several 1-hourly lagged observed weather variables (e.g., wind
-direction, wind speed, relative humidity, pressure gradients along the valley)
-and derived spatial and temporal difference of these lagged observations.
+employ several 1-hourly lagged observed weather variables, amongst others, the
+pressure gradient along the upper and lower Inn Valley, the wind direction, and
+the wind speed at Innsbruck airport.
 
 As an illustration of the tree based wind direction forecasts at Innsbruck
-airport, a BIC based post-pruned regression tree of depth three is shown in
-Figure~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
-nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
-valley (from east to west, nodes 4 and 5). Secondly, downslope winds blowing
-across the Alpine crest along the Wipp Valley towards Innsbruck (from
-south-east to north-west, nodes 7 and 8). Thirdly, down-valley winds blowing
-in the direction of the valley mouth (from west to east, nodes 12, 14, and 15).
-The final node 11 with a concentration parameter $\kappa$ of only $0.84$
-captures the cases with rather low wind speeds which typically don't show a
-clear wind direction or distinct wind regime. In general, winds typically blow
-from areas of high pressure to areas of low pressure where strong pressure
-gradients yield high wind speeds and a rather well-defined wind direction.
-Thus, the splits in the lagged wind direction perform a clustering into the
-different wind regimes and the additional splits in pressure gradients or wind
-speed divide the predictions according to the expected consistency of the wind
-direction. This can be seen in the final nodes with strong pressure gradients
-or high wind speeds where most training cases are classified correctly and
-therefore the concentration parameter as an indicator of the certainty of the
-forecast is rather large.
+airport, a regression tree of depth three is depicted in Figure~\ref{fig:tree}
+showing the empirical histogram and fitted response distribution for each terminal node.
+According to the fitted location parameters the following wind regimes can be summarized:
+Firstly, up-valley winds blowing from the valley mouth towards the upper valley
+(from east to west, nodes 4 and 5). Secondly, downslope winds blowing across
+the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
+north-west, nodes 7 and 8). Thirdly, down-valley winds blowing in the direction
+of the valley mouth (from west to east, nodes 12, 14, and 15). The terminal node
+11 with a concentration parameter $\kappa$ of only $0.84$ captures the cases
+with rather low wind speeds which typically do not show a clear wind direction
+or distinct wind regime. In general, winds typically blow from areas of high
+pressure to areas of low pressure where strong pressure gradients yield high
+wind speeds and a rather well-defined wind direction. Thus, the splits in the
+lagged wind direction perform a clustering into the different wind regimes
+employed by the fitted location parameters, and the additional splits in
+pressure gradients or wind speed divide the predictions according to the
+expected consistency of the wind direction expressed by the fitted
+concentration parameters. This can be seen in the terminal nodes with strong
+pressure gradients or high wind speeds where most training cases are classified
+correctly and therefore the concentration parameter as an indicator of the
+certainty of the forecast is rather large.
 

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Improve applications
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2429 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4e8a4104cf810e134fe6e21c7948fd1170b8a846

@@
 
-\caption{Fitted distribution tree for a circular response employing the von
+\caption{Fitted distributional tree for a circular response employing the von
 Mises distribution. For each terminal node, the empirical histogram and the
@@
 which the wind is blowing. The chosen split variables are wind direction
-(dd~[$\text{rad}$]), wind speed (ff~[$\text{ms}^{-1}$]), and pressure gradients
-along the upper (pdiff\_up~[$\text{hPa}$]) and lower
-(pdiff\_low[$\text{hPa}$]) Inn Valley where positive values point in the
+(dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$), and pressure gradients
+along the upper (pdiff\_up; $\text{hPa}$) and lower
+(pdiff\_low; $\text{hPa}$) Inn Valley where positive values point in the
 direction of Innsbruck.}
@@
 employ several 1-hourly lagged observed weather variables, amongst others, the
-pressure gradient along the upper and lower Inn Valley, the wind direction, and
-the wind speed at Innsbruck airport.
+pressure gradient along the upper and lower Inn Valley, and the wind direction and
+wind speed at Innsbruck airport.
 
@@
 showing the empirical histogram and fitted response distribution for each terminal node.
-According to the fitted location parameters the following wind regimes can be summarized:
+According to the fitted location parameters $\mu$ the following wind regimes can be summarized:
 Firstly, up-valley winds blowing from the valley mouth towards the upper valley
@@
 lagged wind direction perform a clustering into the different wind regimes
-employed by the fitted location parameters, and the additional splits in
+identified by the fitted location parameters, and the additional splits in
 pressure gradients or wind speed divide the predictions according to the

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2431 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bae0c8c64bf67fb2fa011fce6b8a27ad432ab088

--- 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
@@
 direction with high variation and many possible influencing factors.
-%Another valuable feature is the capability of capturing also non-additive and abrupt changes, 
 Additionally, abrupt changes are captured, e.g., when the impact of a covariate such as 
@@
 (see, e.g., distributional forests, Schlosser et.~al, 2018). 
-%In that way, it should not have a 
-%significant impact on the results whether circular covariates are included as linear regressors
-%yielding a larger number of splits or as circular regressors with ``intervall splits''.
+% In that way the impact of the chosen scale on which circular regressors are considered should be reduced.
 
-Note that the circular variable $dd$ (lagged observations of wind) is employed as a linear
-regressor. Therefore, several splits are performed in $dd$ in order to properly cluster the training 
-data in corresponding wind directions (see Figure~\ref{fig:tree}). 
+Note that circular variables are employed as linear regressors which leads to multiple splits
+in these variables in order to properly cluster the training data.
+%Therefore, considering them on a circular scale could reduce the complexity of the model and
+%possibly improve the results.
+%Note that the circular variable $dd$ (lagged observations of wind) is employed as a linear
+%regressor. Therefore, several splits are performed in $dd$ in order to properly cluster the training 
+%data in corresponding wind directions (see Figure~\ref{fig:tree}). 
 In this case searching for two split points defining a split interval on the circular scale instead 
@@
 %cluster the training data in corresponding wind directions.
+
+\medskip
+
 \textit{Computational Details:} The employed \textsf{R} packages are available on \textsf{R}-Forge at

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Include Moritz' adaptions (again)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2432 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ea110b5755f99ed397a9cac06ff3772385bbbd64

--- 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
@@
 \includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-
 \caption{Fitted distributional tree for a circular response employing the von

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Include suggestion Moritz
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2433 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: afd1dbcdef7567f58865d5bb97d25db77e6c2e08
further revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2434 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dd970367c56d12abee4e8f2c3dad362a283ece4e

--- 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
@@
 Hence MOB can test for parameter instabilities in the model by assessing whether the scores
-change along with the available covariates. If so, by maximimizing a partitioned likelihood
+change along with the available covariates. If so, by maximizing a partitioned likelihood
 the parameter instabilities are incorporated into the model. This procedure is repeated recursively
@@
 \includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-
 \caption{Fitted distributional tree for a circular response employing the von
-Mises distribution. For each terminal node, the empirical histogram and the
+Mises distribution. For each terminal node, the empirical density and the 
 fitted density distribution is displayed in gray and red, respectively. The
-location parameters are depicted as red hands and convey the direction from
+estimated location parameter is depicted as red hand and conveys the direction from
 which the wind is blowing. The chosen split variables are wind direction
@@
 
-As an illustration of the tree based wind direction forecasts at Innsbruck
-airport, a regression tree of depth three is depicted in Figure~\ref{fig:tree}
-showing the empirical histogram and fitted response distribution for each terminal node.
-According to the fitted location parameters $\mu$ the following wind regimes can be summarized:
+As an illustration of the tree-based wind direction forecasts at Innsbruck
+airport, a distributional tree of depth three is depicted in Figure~\ref{fig:tree}
+showing the circular histogram and fitted response distribution for each terminal node.
+According to the fitted location parameter $\mu$ the following wind regimes can be summarized:
 Firstly, up-valley winds blowing from the valley mouth towards the upper valley
@@
 of the valley mouth (from west to east, nodes 12, 14, and 15). The terminal node
-11 with a concentration parameter $\kappa$ of only $0.84$ captures the cases
-with rather low wind speeds which typically do not show a clear wind direction
-or distinct wind regime. In general, winds typically blow from areas of high
+11 captures the cases with rather low wind speeds which typically do not show a clear 
+wind direction or distinct wind regime as implied by a rather low concentration 
+parameter $\kappa$.
+In general, winds typically blow from areas of high
 pressure to areas of low pressure where strong pressure gradients yield high
@@
 lagged wind direction perform a clustering into the different wind regimes
-identified by the fitted location parameters, and the additional splits in
+identified by the fitted location parameter, and the additional splits in
 pressure gradients or wind speed divide the predictions according to the
 expected consistency of the wind direction expressed by the fitted
-concentration parameters. This can be seen in the terminal nodes with strong
+concentration parameter. This can be seen in the terminal nodes with strong
 pressure gradients or high wind speeds where most training cases are classified
@@
 \section{Discussion and Outlook}
-This study shows that applying distributional trees on circular data provides easily 
-interpretable models clearly illustrating the impact of given covariates and their possible 
-interactions on the distribution of the dependent variable. 
-Moreover, covariates are selected automatically such that no prespecification of the model is 
-required which can be a challenging task, particularly for a meteorlogical quantity such as wind 
-direction with high variation and many possible influencing factors.
-%Another valuable feature is the capability of capturing also non-additive and abrupt changes, 
-Additionally, abrupt changes are captured, e.g., when the impact of a covariate such as 
-\textit{air pressure difference} determines either up- or down-valley wind directions.
+
+This study shows that applying distributional trees on circular data provides
+easily interpretable models clearly illustrating the impact of given covariates
+and their possible interactions on the distribution of the dependent variable.
+Moreover, covariates are selected automatically such that no prespecification
+of the model is required which can be a challenging task, particularly for a
+meteorological quantity such as wind direction with high variation and many
+possible influencing factors. Additionally, the employed distributional trees
+allow for abrupt changes in the response variable, such as e.g., a wind shift due to
+a turn of the the pressure gradient within the Inn Valley.
 
 A possible extension of the method as presented is to build an ensemble of
-distributional trees to regularize and stabilize the model 
-(see, e.g., distributional forests, Schlosser et.~al, 2018). 
-%In that way, it should not have a 
-%significant impact on the results whether circular covariates are included as linear regressors
-%yielding a larger number of splits or as circular regressors with ``intervall splits''.
+distributional trees to regularize and stabilize the model (see, e.g.,
+distributional forests, Schlosser et.~al, 2018). 
+Furthermore, in order to properly account not only for the circular properties of 
+the response but also of the regressors, extensions in the search of split points 
+would be feasible. For example, searching for two split points defining a split 
+interval on the circular scale instead of a single split point on the linear scale 
+could reduce the complexity of the model and possibly improve the results.
 
-Note that the circular variable $dd$ (lagged observations of wind) is employed as a linear
-regressor. Therefore, several splits are performed in $dd$ in order to properly cluster the training 
-data in corresponding wind directions (see Figure~\ref{fig:tree}). 
-In this case searching for two split points defining a split interval on the circular scale instead 
-of a single split point on the linear scale could reduce the complexity of the model and possibly 
-improve the results.
-%Alternatively, in the case of a circular covariate 
-%the tree building algorithm could search for two split points defining a split interval instead of 
-%a single split point. This would reduce the complexity of the model and possibly improve the results.
-%Based on the results it can be assumed that including covariates such as lagged observations of 
-%wind direction as linear regressors provides valuable information. 
-%However, in order to properly account not only for the circular properties of the response but 
-%also of regressors, additional adaptions of distributional trees are required. In particular, 
-%if the selected split variable is a circular variable, e.g. wind direction, this demands for finding
-%two split point defining a ``split interval'' instead of a single split point.
-%This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
-%the linearly employed regressor \textit{dd} (lagged wind direction) are required in order to 
-%cluster the training data in corresponding wind directions.  
+\medskip
+
 \textit{Computational Details:} The employed \textsf{R} packages are available on \textsf{R}-Forge at
@@
 
-%While the application of the von Mises distribution allows for modeling a circular response variable,
-%additional adaptions have to be made when including circular regressor variables in a distributional tree
-%model. In particular, if the selected split variable is a circular variable this demands for finding
-%not only one split point but a ``split interval''. This consideration is of interest for many applications,
-%for example when modeling wind directions it might be desirable to include day time or earlier 
-%observations of wind directions as regressor variables.
+%% alternative version:
+%This study shows that applying distributional trees on circular data provides easily 
+%interpretable models clearly illustrating the impact of given covariates and their possible 
+%interactions on the distribution of the dependent variable. 
+%Moreover, covariates are selected automatically such that no prespecification of the model is 
+%required which can be a challenging task, particularly for a meteorological quantity such as wind 
+%direction with high variation and many possible influencing factors.
+%Additionally, abrupt changes are captured, e.g., when the impact of a covariate such as 
+%\textit{air pressure difference} determines either up- or down-valley wind directions.
+
+%A possible extension of the method as presented is to build an ensemble of 
+%distributional trees to regularize and stabilize the model 
+%(see, e.g., distributional forests, Schlosser et.~al, 2018). 
+%% In that way the impact of the chosen scale on which circular regressors are considered should be reduced.
+
+%Note that circular variables are employed as linear regressors which leads to multiple splits
+%in these variables in order to properly cluster the training data.
+%In this case searching for two split points defining a split interval on the circular scale instead 
+%of a single split point on the linear scale could reduce the complexity of the model and possibly 
+%improve the results.
 

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Adapt minor spelling errors
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2435 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e70b4fe94523e8ec2f220b5ef0a08b3e8e3d6c7e

@@
 e.g., in air traffic management or renewable energy production. This study
-presents an application of circular regression trees  on hourly direction
+presents an application of circular regression trees  on hourly wind direction
 forecasts at Innsbruck Airport. Innsbruck lies at the bottom of the Inn
@@
 to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
-in the North with a approx. 2300\,m high mountain ridge and opens to the South
+in the North with an approx. 2300\,m high mountain ridge and opens to the South
 into the crest-perpendicular Wipp Valley. As potential split variables we
@@
 pressure gradient along the upper and lower Inn Valley, and the wind direction and
-wind speed at Innsbruck airport.
+wind speed at Innsbruck Airport.
 
 As an illustration of the tree-based wind direction forecasts at Innsbruck
-airport, a distributional tree of depth three is depicted in Figure~\ref{fig:tree}
+Airport, a distributional tree of depth three is depicted in Figure~\ref{fig:tree}
 showing the circular histogram and fitted response distribution for each terminal node.
-According to the fitted location parameter $\mu$ the following wind regimes can be summarized:
+According to the fitted location parameter the following wind regimes can be summarized:
 Firstly, up-valley winds blowing from the valley mouth towards the upper valley
@@
 11 captures the cases with rather low wind speeds which typically do not show a clear 
-wind direction or distinct wind regime as implied by a rather low concentration 
-parameter $\kappa$.
+wind direction or distinct wind regime as implied by a small concentration 
+parameter.
 In general, winds typically blow from areas of high

### 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
some further streamlining
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2436 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 92af93b3c57911048ea93b0d2293237ee0d34f5a

@@
 distributional trees are easy to interpret, can detect non-additive effects, and select
-covariates along with interactions automatically. For illustration, hourly wind 
+covariates and their interactions automatically. For illustration, hourly wind 
 direction forecasts are obtained at Innsbruck Airport based on a set of meteorological
@@
 by maximizing the corresponding log-likelihood $\ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))$. 
-The model scores are calculated by  $s(y; \mu, \kappa) = (\partial_{\mu} \ell(\mu, \kappa; y),
+The model scores are given by  $s(y; \mu, \kappa) = (\partial_{\mu} \ell(\mu, \kappa; y),
 \partial_{\kappa} \ell(\mu, \kappa; y))$. In a subgroup of size $n$, evaluating the scores
@@
 respectively.
-Hence MOB can test for parameter instabilities in the model by assessing whether the scores
-change along with the available covariates. If so, by maximizing a partitioned likelihood
+Hence MOB or CTree can assess whether the scores change along with the available covariates.
+If so, by maximizing a partitioned likelihood
 the parameter instabilities are incorporated into the model. This procedure is repeated recursively
 until there are no significant parameter instabilities or until another stopping criterion
-is met (e.g., concerning subgroup sizes or depth of the tree).
+is met (e.g., subgroup size or tree depth).
 
@@
 \includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
-\caption{Fitted distributional tree for a circular response employing the von
-Mises distribution. For each terminal node, the empirical density and the 
-fitted density distribution is displayed in gray and red, respectively. The
-estimated location parameter is depicted as red hand and conveys the direction from
-which the wind is blowing. The chosen split variables are wind direction
-(dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$), and pressure gradients
-along the upper (pdiff\_up; $\text{hPa}$) and lower
-(pdiff\_low; $\text{hPa}$) Inn Valley where positive values point in the
-direction of Innsbruck.}
+\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
+In each terminal node the empirical histogram (gray) and fitted density (red line)
+are depicted along with the estimated location parameter (red hand). The covariates employed are
+wind direction (dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$),
+and pressure gradients along the upper (pdiff\_up; $\text{hPa}$) and lower
+(pdiff\_low; $\text{hPa}$) Inn Valley.}
 \label{fig:tree} \end{figure}
@@
 Wind is a classical circular quantity and accurate forecasts of wind direction
-are of great importance for decision making processes and risk management,
+are of great importance for decision-making processes and risk management,
 e.g., in air traffic management or renewable energy production. This study
-presents an application of circular regression trees  on hourly wind direction
+employs circular regression trees to obtain hourly wind direction
 forecasts at Innsbruck Airport. Innsbruck lies at the bottom of the Inn
@@
 to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
-in the North with an approx. 2300\,m high mountain ridge and opens to the South
-into the crest-perpendicular Wipp Valley. As potential split variables we
-employ several 1-hourly lagged observed weather variables, amongst others, the
-pressure gradient along the upper and lower Inn Valley, and the wind direction and
+in the North by an approx.~2300\,m high mountain ridge and opens to the South
+into the crest-perpendicular Wipp Valley. Hence, pressure gradients along the
+upper and lower Inn Valley are considered as covariates along with other
+meteorological measurements (lagged by one hour) such as wind direction and
 wind speed at Innsbruck Airport.
@@
 As an illustration of the tree-based wind direction forecasts at Innsbruck
-Airport, a distributional tree of depth three is depicted in Figure~\ref{fig:tree}
-showing the circular histogram and fitted response distribution for each terminal node.
-According to the fitted location parameter the following wind regimes can be summarized:
-Firstly, up-valley winds blowing from the valley mouth towards the upper valley
-(from east to west, nodes 4 and 5). Secondly, downslope winds blowing across
+Airport, a distributional tree of depth three is depicted in
+
+Figure~\ref{fig:tree} depicts the resulting distributional tree, including
+both the empirical (gray) and fitted von Mises (red) distribution of wind direction
+in each terminal node. Based on the fitted location parameters $\hat \mu$, the subgroups
+can be distinguished into the following wind regimes:
+(1)~Up-valley winds blowing from the valley mouth towards the upper valley
+(from east to west, nodes 4 and 5). (2) Downslope winds blowing across
 the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
-north-west, nodes 7 and 8). Thirdly, down-valley winds blowing in the direction
-of the valley mouth (from west to east, nodes 12, 14, and 15). The terminal node
-11 captures the cases with rather low wind speeds which typically do not show a clear 
-wind direction or distinct wind regime as implied by a small concentration 
-parameter.
-In general, winds typically blow from areas of high
-pressure to areas of low pressure where strong pressure gradients yield high
-wind speeds and a rather well-defined wind direction. Thus, the splits in the
-lagged wind direction perform a clustering into the different wind regimes
-identified by the fitted location parameter, and the additional splits in
-pressure gradients or wind speed divide the predictions according to the
-expected consistency of the wind direction expressed by the fitted
-concentration parameter. This can be seen in the terminal nodes with strong
-pressure gradients or high wind speeds where most training cases are classified
-correctly and therefore the concentration parameter as an indicator of the
-certainty of the forecast is rather large.
+north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
+of the valley mouth (from west to east, nodes 12, 14, and 15). Node~11
+captures observations with rather low wind speeds that cannot be distinguished
+clearly into wind regimes and consequently are associated with a very low
+estimated concentration $\hat \kappa$. In terms of covariates, the lagged
+wind direction (``persistence'') is mostly responsible for distinguishing
+the broad wind regimes listed above while the pressure gradients and wind
+speed separate between subgroups with high vs.\ low precision.
 
@@
 
-This study shows that applying distributional trees on circular data provides
-easily interpretable models clearly illustrating the impact of given covariates
-and their possible interactions on the distribution of the dependent variable.
-Moreover, covariates are selected automatically such that no prespecification
-of the model is required which can be a challenging task, particularly for a
-meteorological quantity such as wind direction with high variation and many
-possible influencing factors. Additionally, the employed distributional trees
-allow for abrupt changes in the response variable, such as e.g., a wind shift due to
-a turn of the the pressure gradient within the Inn Valley.
+Distributional trees for circular responses are established by coupling
+model-based recursive partitioning with the von Mises distribution.
+The resulting trees can capture nonlinear changes, shifts, and potential interactions
+in covariates without prespecification of such effects. This is modeling
+wind direction in mountainous terrain where wind shifts can occur due to turns
+a turn of the pressure gradient along a valley.
 
-A possible extension of the method as presented is to build an ensemble of
-distributional trees to regularize and stabilize the model (see, e.g.,
-distributional forests, Schlosser et.~al, 2018). 
-Furthermore, in order to properly account not only for the circular properties of 
-the response but also of the regressors, extensions in the search of split points 
-would be feasible. For example, searching for two split points defining a split 
-interval on the circular scale instead of a single split point on the linear scale 
-could reduce the complexity of the model and possibly improve the results.
+A natural extension are ensembles (or forests) of such circular trees
+that can improve forecasts by regularizing and stabilizing the model
+(see Schlosser et.~al, 2018 for a general discussion of distributional forests).
 
-\medskip
+Another possible improvement for obtaining more parsimonious and more stable trees is to
+consider splitting circular covariates into two circle segments by searching two
+split points simultaneously (rather than sequentially at different depths as
+for variable ``dd'' in Figure~\ref{fig:tree}).
 
-\textit{Computational Details:} The employed \textsf{R} packages are available on \textsf{R}-Forge at
-https://R-Forge.R-project.org/projects/partykit/.
+\textsf{R} packages implementing the proposed methods are under development at
+\texttt{https://R-Forge.R-project.org/projects/partykit/}.
 
@@
      Topics in Circular Statistics.
-     {\it World Scientific}, {\bf 5}.
+     {\it World Scientific}.
 
@@
      Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
-     {\it arXiv.org E-Print Archive}, https://arxiv.org/abs/1804.02921v2.
+     arXiv:1804.02921, {\it arXiv.org E-Print Archive}.
 

### 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
figure placement 'p' rather than 'ht'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2437 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b0656aaab27d43a3b76cc3c11e51277bf7c1a375
Correct minor errors and update domain of mu
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2438 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6da81f5f674e38e5ce1f31cdafc30d945ba87dc0

@@
 distribution, also known as ``the circular normal distribution''. Based on a location parameter
-$\mu \in \mathbb{R}$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2 \pi]$
+$\mu \in [0, 2 \pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2 \pi]$
 is given by:
@@
 
-\begin{figure}[!ht]\centering
+\begin{figure}[p!]\centering
 \includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
@@
 
-As an illustration of the tree-based wind direction forecasts at Innsbruck
-Airport, a distributional tree of depth three is depicted in
-
 Figure~\ref{fig:tree} depicts the resulting distributional tree, including
@@
 The resulting trees can capture nonlinear changes, shifts, and potential interactions
-in covariates without prespecification of such effects. This is modeling
+in covariates without prespecification of such effects. This allows to model
 wind direction in mountainous terrain where wind shifts can occur due to turns
-a turn of the pressure gradient along a valley.
+of the pressure gradients along a valley.
 

### 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
adaptions following formatting guidelines
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2440 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ec3f093bb1e6fb42108e018bde25d018195c8fc0

--- 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
@@
 distribution, also known as ``the circular normal distribution''. Based on a location parameter
-$\mu \in [0, 2 \pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2 \pi]$
+$\mu \in [-\pi, \pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [-\pi, \pi]$
 is given by:
@@
 The resulting trees can capture nonlinear changes, shifts, and potential interactions
-in covariates without prespecification of such effects. This allows to model
-wind direction in mountainous terrain where wind shifts can occur due to turns
-of the pressure gradients along a valley.
+in covariates without prespecification of such effects. This is particularly
+useful for modeling wind direction in mountainous terrain where wind shifts
+can occur due to turns of the pressure gradients along a valley.
 

### 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
adaptions following formatting guidelines
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2441 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: eb2f6fbdb34d0043eaed171a3ad052491561e4aa

@@
 % Keywords (at most 5):
-\keywords{Distributional trees; Circular response; Von Mises distribution.}
+\keywords{Distributional Trees; Circular Response; Von Mises Distribution.}
 
@@
 \begin{equation}
-  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
+  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
 \end{equation}
@@
 \begin{figure}[p!]\centering
-\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
+\includegraphics[height = .5\textheight,angle=90,origin=c]{schlosser-circtree_plot.pdf}
 \caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
@@
 (pdiff\_low; $\text{hPa}$) Inn Valley.}
-\label{fig:tree} \end{figure}
+\label{schlosser:fig_tree} \end{figure}
 
@@
 
-Figure~\ref{fig:tree} depicts the resulting distributional tree, including
+Figure~\ref{schlosser:fig_tree} depicts the resulting distributional tree, including
 both the empirical (gray) and fitted von Mises (red) distribution of wind direction
@@
 that can improve forecasts by regularizing and stabilizing the model
-(see Schlosser et.~al, 2018 for a general discussion of distributional forests).
+(see Schlosser et.~al~(2018) for a general discussion of distributional forests).
 
@@
 split points simultaneously (rather than sequentially at different depths as
-for variable ``dd'' in Figure~\ref{fig:tree}).
+for variable ``dd'' in Figure~\ref{schlosser:fig_tree}).
 
@@
 
-\item[Jammalamadaka, S. R., and Sengupta, A.] (2001).
-     Topics in Circular Statistics.
-     {\it World Scientific}.
+\item[Jammalamadaka, S. R. and Sengupta, A.] (2001).
+     {\it Topics in Circular Statistics}.
+     World Scientific.
 

### 2019-02-06 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
feedback from Georg, prepared changing variable names
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2443 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 861ac95cb09ac928da2664d762938bc4d1b33d9a

@@
 
-%\subsection{Von Mises Distribution}
-%The von Mises distribution is also known as the circular normal distribution 
-%and its density function is given by
-%\begin{equation}
-%  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
-%\end{equation}
-%with a location parameter $\mu$, a conentration parameter $\kappa$, and $I_0(\kappa)$ being the 
-%modified Bessel function of the first kind and order $0$.
-%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
-%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
-%to the variance $\sigma^2$ of the normal distribution. 
-%See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.
-
 Distributional trees~(Schlosser et.~al, 2018) fuse distributional regression modeling with
@@
 In each terminal node the empirical histogram (gray) and fitted density (red line)
-are depicted along with the estimated location parameter (red hand). The covariates employed are
-wind direction (dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$),
-and pressure gradients along the upper (pdiff\_up; $\text{hPa}$) and lower
-(pdiff\_low; $\text{hPa}$) Inn Valley.}
+are depicted along with the estimated location parameter (red hand). The covariates
+employed are wind direction ($\text{rad}$), wind speed ($\text{ms}^{-1}$),
+and pressure gradients ($\text{hPa}$) west and east of the airport,
+all lagged by one hour.}
 \label{schlosser:fig_tree} \end{figure}
@@
 employs circular regression trees to obtain hourly wind direction
-forecasts at Innsbruck Airport. Innsbruck lies at the bottom of the Inn
-Valley, a deep mountain valley in the Eastern European Alps orientated parallel
-to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
-in the North by an approx.~2300\,m high mountain ridge and opens to the South
-into the crest-perpendicular Wipp Valley. Hence, pressure gradients along the
-upper and lower Inn Valley are considered as covariates along with other
-meteorological measurements (lagged by one hour) such as wind direction and
-wind speed at Innsbruck Airport.
+forecasts at Innsbruck Airport.
+Innsbruck lies at the bottom of a deep valley in the Alps. Topography
+channels wind along the west-east valley axis or along a tributary valley
+intersecting from the south. Hence, pressure gradients to which valley wind
+regimes react both west and east of the airport are considered as covariates
+along with other meteorological measurements at the airport (lagged by one hour),
+such as wind direction and wind speed at Innsbruck Airport.
 
@@
 split points simultaneously (rather than sequentially at different depths as
-for variable ``dd'' in Figure~\ref{schlosser:fig_tree}).
+for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
 
@@
 
-%% alternative version:
-%This study shows that applying distributional trees on circular data provides easily 
-%interpretable models clearly illustrating the impact of given covariates and their possible 
-%interactions on the distribution of the dependent variable. 
-%Moreover, covariates are selected automatically such that no prespecification of the model is 
-%required which can be a challenging task, particularly for a meteorological quantity such as wind 
-%direction with high variation and many possible influencing factors.
-%Additionally, abrupt changes are captured, e.g., when the impact of a covariate such as 
-%\textit{air pressure difference} determines either up- or down-valley wind directions.
-
-%A possible extension of the method as presented is to build an ensemble of 
-%distributional trees to regularize and stabilize the model 
-%(see, e.g., distributional forests, Schlosser et.~al, 2018). 
-%% In that way the impact of the chosen scale on which circular regressors are considered should be reduced.
-
-%Note that circular variables are employed as linear regressors which leads to multiple splits
-%in these variables in order to properly cluster the training data.
-%In this case searching for two split points defining a split interval on the circular scale instead 
-%of a single split point on the linear scale could reduce the complexity of the model and possibly 
-%improve the results.
-
 \bigskip
-
-%***********************************************************************
-
-% Tables can be included at any place in the text.
-% As general format, use tables with horizontal lines only
-% (i.e., no vertical lines separating columns).
-% Start and end each table with a double horizontal line.
-% Tables are incorporated using a table environment:
-
-%\begin{table}[!ht]\centering
-% A caption is put above the table and a label is defined
-% to be used for reference to this specific table.
-% Use labels which are very unlikely to be used by
-% other contributors; for example, use labels starting
-% with the surname of the first author.
-%\caption{\label{smith:tab1} Caption text \textbf{ABOVE} the table.}
-% A table with three columns, where the first is left aligned,
-% the second has centred entries, and the third is right aligned,
-% is generated as follows (note: if requested, use \cmidrule{} instead of \cline{})
-%\medskip
-%\begin{tabular}{lcr}
-% First line:
-%\toprule[0.09 em]
-% The body of the table:
-%Title col1 & Title col2 & Title col3 \\
-%\midrule
-%row1 col1  & row1 col2  & row1 col3  \\
-%row2 col1  & row2 col2  & row2 col3  \\ %
-%row3 col1  & row3 col2  & row3 col3  \\
-% last line:
-%\bottomrule[0.09 em]
-%\end{tabular}
-%\end{table}
-
-% In the text, reference to the Table can be made as follows:
-%We refer to Table~\ref{smith:tab1} for a summary of our main results. Have a look to Table~\ref{smith:tab2} for
-%an additional example.
-
-%\begin{table}[!ht]\centering
-%\caption{\label{smith:tab2} Caption text \textbf{ABOVE} the table.}
-%\medskip
-%\begin{tabular}{lcr}
-% First line:
-%\toprule[0.09 em]
-% The body of the table:
-%  &\multicolumn{2}{c}{Title  for cols 2 -3} \\
-%\cmidrule{2-3} %
-%Title1 & Title2 & Title3 \\
-%\midrule
-%& $a$  & $c$ \\
-%& $b$  & $d$ \\ %
-%\midrule[0 em]
-%Total  & $a+b$  & $n$  \\
-% last line:
-%\bottomrule[0.09 em]
-%\end{tabular}
-%\end{table}
-
-%***********************************************************************
-
-% Figures can be included at any place in the text.
-% The only allowed formats for figures are pdf files.
-%
-% Please, do not include figures in any other format.
-%
-% Use file names which are very unlikely to be used by
-% other contributors; for example, use file names starting
-% with the surname of the first author.
-% Figures are incorporated using a figure environment:
-% Make sure you specify the extension of the file (pdf)
-
-%Finally a figure (in \verb|.pdf|!)
-
-%\begin{figure}[!ht]\centering
-% You can pre-specify the width of the graph:
-%\includegraphics[width=8cm]{circmax-004.pdf}
-% Below the figure, a caption is put, and a label is defined
-% to be used for reference to this specific figure.
-% Use labels which are very unlikely to be used by
-% other contributors; for example, use labels starting
-% with the surname of the first author.
-%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}
-%\end{figure}
-
-% In the text, reference to the Figure can be made as follows:
-%We refer to Figure~\ref{smith:fig1} for a~graphical representation.
-
-%***********************************************************************
-
-% Acknowledgments, if needed:
-%\acknowledgments{Special Thanks to ... }
-
-%***********************************************************************
-
-% References should be placed in the text in author (year) form.
-% The list of references should be placed below IN ALPHABETICAL ORDER.
-% (Please follow the format of the examples very tightly).
 

### 2019-02-07 pkg/disttree/inst/draft/rejoinder2.Rnw
first revisions based on reviews; documentation of changes in rejoinder2.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2448 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b45f4b30e679e49be7210ebdd93fc64574908019

\documentclass[american,foldmarks=false,noconfig]{uibklttr}
\usepackage{amsmath,sfmath,hyperref,color,enumitem,soul}

\let\code=\texttt
\let\pkg=\emph
\let\proglang=\textsf
\newenvironment{review}{\fontshape{\itdefault}\fontseries{\bfdefault} \selectfont \smallskip}{\par}

\setkomavar{signature}{Lisa Schlosser}

\makeatletter
\@setplength{refvpos}{68mm}
\makeatother
\setlength{\textheight}{25cm}
\definecolor{quotecolor}{rgb}{0.4,0.4,0.5}

\definecolor{fixmecol}{rgb}{1,1,0.7}
\newcommand{\fixme}[1]{{\sethlcolor{fixmecol}\hl{\textit{FIXME: #1}}}}

\begin{document}
\SweaveOpts{concordance=TRUE}

\begin{letter}{
Nicoleta Serban\\
Area Editor\\
The Annals of Applied Statistics}

\setkomavar{subject}{Revision of AOAS1804-015}

\opening{Dear Professor Serban,}

please find attached the revised version of our manuscript 
``Distributional regression forests for probabilistic precipitation 
forecasting in complex terrain'' (AOAS1804-015).

Thank you very much for the positive and constructive feedback from you, 
the associate editor, and the two referees. Based on your comments, the
manuscript has been revised and we feel that this substantially improved our
contribution.
%
The most important changes are:
\begin{itemize}

\item \emph{Discussion.}\\
TO DO

\item \emph{\ldots}\\

% \item In the calculation of variable importance we have detected some inconsistencies.
% Thus, we have revised the applied method to evaluate this measurement and recalculated
% variable importance for the distributional forest.

\end{itemize}

All changes and additions are explained in much more detail in the
point-to-point reply on the next pages.

\closing{Best regards,
\vspace*{1cm}
}

\vspace*{-2.8cm}
\hspace*{-0.3cm} 
\includegraphics[width = 0.23\textwidth]{signature-Schlosser}

\end{letter}

\newpage

\textbf{\LARGE Associate Editor}

\begin{review}
{\color{quotecolor} \textit{page 19, last paragraph:} 
To assess whether these differences in predictive performance 
are due to differences in the topography \ldots .}\\
It is not clear to me how conclusions 
about topography-related differences in predictive perfomance can be drawn from Fig. 8. 
The terrain seems to be rather complex everywhere in the study area. What is different 
in the south-eastern part of Tyrol? And why do these differences negatively affect 
the performance of distributional regression forests, the most flexible of all four 
methods compared here? Is it possible that the patterns in this figure are due to 
(spatially correlated) sampling variability, i.e. not statistically significant? 
\end{review}

TO DO: adaption in the manuscript

Thank you very much for your comment! East Tyrol lies in a different climat zone than the 
rest of Tyrol indicating different meteorological characteristics. For example, a covariate
might have a linear effect on the expected amount of rain in East Tyrol while it has a
non-additive effect in Northern Tyrol. In this case, in East Tyrol an additive model such as 
GAMLSS will capture this influence better than a tree model, particularly if this covariate
is correctly specified in the model (which might be the case in the prespecified GAMLSS).
Of course, these are just assumptions and there could be many other reasons. However,
looking at the results of the cross-validation in Supplement~B it can be observed that
the difference in performance between the GAMLSS and the distributional tree models as
seen in the single evaluation evens out, hence the specific choice of learning years seems
to have an impact as well.

\newpage

\textbf{\LARGE Reviewer 1}

\bigskip

\textbf{General Comments:}

\begin{review}
The authors have well addressed the reviewers’ comments. The two added supplementary 
materials give a more solid ground to the proposed post-processing method by showing 
it is robustness to some choice of forecast distribution and that the results presented 
in the main manuscript does not hold only for the specific station.
I recommend the paper to be accepted after the minor revisions proposed
hereafter have been adressed.
\end{review}

Thank you very much for the positive feedback and the helpful comments regarding the paper.

\bigskip

\textbf{Specific comments:}

\textbf{Main Article:}

\begin{review}
1. {\color{quotecolor} \textit{page 14:} mtry, minsplit, minbucket, alpha.}\\
You have not yet given any detail about the statistical software you used. 
Please refer here to the ``computational~details'' section at the end of article. 
Also, please indicate to which function these parameters pertain to (ctree?).
\end{review}

The missing information and reference have been added in the 
manuscript in Section~3.2 \textit{Models and evaluation: Distributional forest}.
The described parameters pertain to the forest-building
function \code{distforest}. But you are right, within the function \code{distforest} 
the arguments \code{minsplit}, \code{minbucket} and \code{alpha} are handed over to the 
tree-building function \code{disttree} which in this case applies the function \code{ctree}.

\begin{review}
2. {\color{quotecolor}\textit{page 15:} The figure shows that the model properly 
depicts the point masses at zero (i.e., the probability of a dry day) and the forecasted 
probability density function for the total amount of precipitation.}\\
It is not possible to assess a probabilistic forecast with only one example. 
Please erase or rephrase this sentence.
\end{review}

Thank you for pointing this out. We have rephrased this sentence in Section 3.3. of the manuscript.

\begin{review}
3. {\color{quotecolor}\textit{page 25:} 
The smaller the $\boldsymbol{p}$-value corresponding to the standardized test statistic 
$\boldsymbol{c(t_l, \mu_l, \Sigma_l)}$}\\
Please specify the distribution of the test statistic under the null hypothesis.
\end{review}

Depending on the whether a maximum or a quadratic form of the test statistic is considered
for the standardized test statistic its asymptotic conditional distribution is either a 
normal ($c_{max}$) or a $\chi^2$ distribution ($c_{quad}$). 
We have added this information to the addressed section of Appendix~A.

\textbf{Supplement B:}

\begin{review}
4. {\color{quotecolor}\textit{page 1 (Abstract):} Moreover, further details including a 
crossvalidation analysis are preformed for only one station, Axams.}\\
Please change ``preformed'' into ``performed''.
\end{review}

This has been corrected in Supplement B.

\begin{review}
5. {\color{quotecolor}\textit{page 3:} The Resiudal QQ plots and the PIT histograms 
visualize the same values, however, on different scales.}\\
Please change into \textit{the Residual}
\end{review}

This has been corrected in Supplement B.

\newpage

\textbf{\LARGE Reviewer 2}

\textbf{Comments:}

\begin{review}
I think the authors tried to address my comments and invested considerable additional work 
on simulations and predictions on more stations. I appreciate this but am still no completely 
convinced that the models can provide benefits in distributional regression problems in 
particular when the distribution is more complex. The reason is that in all models considered, 
the distribution parameters are "orthogonal" in the sense that the mean and the variance 
parameter separate well.
This is not the case for many other distributions like the Dagum, Zero-inflated models, etc. 
I am wondering how well the distributional forest in these cases works with respect to the 
splits in the different distribution parameters. For instance, the expectation of the response 
for a Dagum distribution is a function of all distribution parameters and I have doubts that 
splitting the forests works without identification issues such that the forests may get very 
sensitive with respect to the test and training data.

In addition, overall I would say the method performs as well as the competitors but loses 
interpretation and for the prespecified GAMLSS also uncertainty estimates. With uncertainty 
I mean uncertainty of the estimates no the variance of the response.

Overall I found the paper interesting but not innovative and general enough to make a 
contribution qualified for AOAS but rather for instance JRSSC.
\end{review}

TO DO

Thank you very much for the positive and constructive 
comments regarding the paper.

\end{document}

### 2019-02-07 pkg/disttree/inst/draft/draft.Rnw
first revisions based on reviews; documentation of changes in rejoinder2.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2448 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b45f4b30e679e49be7210ebdd93fc64574908019

@@
   for each tree using a third of the predictors
-  in each split of the tree (\code{mtry}). Parameters are estimated by
+  in each split of the tree (\hlc{specified by the argument} \code{mtry} 
+  \hlc{in the applied function} \code{distforest}; 
+  \hlc{more information on the employed software is provided in Section}
+  ``Computational Details''). 
+  Parameters are estimated by
   adaptive local likelihood based on the forest weights as described
@@
 for July 24 in 2009--2012 have already been shown as a motivational example. 
-The figure shows that the model properly depicts the point masses at zero 
-(i.e., the probability of a dry day) and the forecasted probability density 
-function for the total amount of precipitation. The four sample forecasts differ 
-considerably in location $\mu$, scale $\sigma$, and the amount of
-censoring while conforming quite well with the actual observations
-from these days. While this is a nice illustrative example we are  
+\hlc{In particular, the figure displays the point mass at zero (i.e., the probability 
+of a dry day) depicted by the model together with the forecasted probability 
+density function for the total amount of precipitation. Based on this illustration 
+it can be observed that} the four sample forecasts differ considerably in 
+location $\mu$, scale $\sigma$, and the amount of censoring while conforming quite 
+well with the actual observations from these days. 
+While this is a nice illustrative example we are  
 interested in the overall predictive performance and calibration of the
@@
 \section*{Computational details}
+\label{sec:comp}
 
@@
 \code{demo("RainStationwise", package = "disttree")}, respectively.
+
+\section*{Acknowledgements}
+\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
+2017 to January 2018) was financially supported by the Swiss National
+Science Foundation, grant number SNF IZSEZ0 177091.}
 
@@
 the maximum of the absolute values of the standardized linear statistic
-can be considered.
+can be considered ($c_{\text{max}}$).
+
+\hlc{Since} \cite{Strasser+Weber:1999} \hlc{showed that the asymptotic conditional 
+distribution of a linear statistics with conditional expectation} $\mu$ 
+\hlc{and covariance} $\Sigma$ 
+\hlc{is a multivariate normal distribution with parameters}
+$\mu$ and $\Sigma$ 
+\hlc{the asymptotic conditional distribution of the standardized test statistic}
+$c(t_{l},\mu_{l},\Sigma_{l})$ 
+\hlc{is either normal}
+($c_{\text{max}}$) 
+\hlc{or follows a} 
+$\chi^2$ distribution ($c_{\text{quad}}$).
 

### 2019-02-07 pkg/disttree/inst/draft/supplement/supplement_stationwise.Rnw
final formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2378 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4503d2d73b067eabaa07a9184943d75e94a802fc
first revisions based on reviews; documentation of changes in rejoinder2.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2448 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b45f4b30e679e49be7210ebdd93fc64574908019

@@
 \usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,pdfpages,hyperref,bookmark}
+\usepackage[all]{hypcap}
 
@@
   Moreover, further details -- including a cross-validation analysis -- are
-  preformed for only one station, Axams. To show that Axams is a fairly typical
+  performed for only one station, Axams. To show that Axams is a fairly typical
   station and similar insights can be obtained for other stations as well,
@@
 
-The Resiudal QQ plots and the PIT histograms visualize the same values,
+The Residual QQ plots and the PIT histograms visualize the same values,
 however, on different scales. For each observation $y$ of the test sample 

### 2019-02-07 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
extended information on authors' intstitutes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2449 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cc4c8e6115cdbcae47a3b44112f52319eff46c2e

@@
 \author{Lisa Schlosser\inst{1}, 
-Moritz N. Lang\inst{1},
-Torsten Hothorn\inst{2},
-Georg J. Mayr\inst{1},
+Moritz N. Lang\inst{{1,2}},
+Torsten Hothorn\inst{3},
+Georg~J.~Mayr\inst{2},
 Reto Stauffer\inst{1},
@@
 
-% Institutes of all authors
-% Include city and country of each institute, do not include the full address.
-\institute{Universit\"at Innsbruck, Austria
-\and Universit\"at Z\"urich, Switzerland}
+%% Institutes of all authors
+%% Include city and country of each institute, do not include the full address.
+%\institute{Universit\"at Innsbruck, Austria
+%\and Universit\"at Z\"urich, Switzerland}
+
+%%%%%%%% IF DEPARTMENTS ARE LISTED AS WELL CHANGE NUMBERING!!!  %%%%%%%%%%%%%%
+\institute{Department of Statistics, Universit\"at Innsbruck, Innsbruck, Austria
+\and Department of Atmospheric and Cryospheric Science, Universit\"at Innsbruck, Innsbruck, Austria
+\and Epidemiology, Biostatistics and Prevention Institute, Universit\"at Z\"urich, Zurich, Switzerland}
 
@@
 are depicted along with the estimated location parameter (red hand). The covariates
-employed are wind direction ($\text{rad}$), wind speed ($\text{ms}^{-1}$),
+employed are wind direction ($degree$), wind speed ($\text{ms}^{-1}$),
 and pressure gradients ($\text{hPa}$) west and east of the airport,
@@
 (from east to west, nodes 4 and 5). (2) Downslope winds blowing across
-the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
+the Alpine crest along the intersecting valley towards Innsbruck (from south-east to
 north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
@@
 \texttt{https://R-Forge.R-project.org/projects/partykit/}.
+
+\bigskip
+
+% Acknowledgments, if needed:
+\acknowledgments{Special Thanks to ... }
 

### 2019-02-07 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
Update fig and include acknowledgment
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2450 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a3edb39fd4e8a07f484632470769773bdece6407

@@
 distribution, also known as ``the circular normal distribution''. Based on a location parameter
-$\mu \in [-\pi, \pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [-\pi, \pi]$
+$\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2\pi]$
 is given by:
@@
 are depicted along with the estimated location parameter (red hand). The covariates
-employed are wind direction ($degree$), wind speed ($\text{ms}^{-1}$),
+employed are wind direction ($\text{degree}$), wind speed ($\text{ms}^{-1}$),
 and pressure gradients ($\text{hPa}$) west and east of the airport,
@@
 % Acknowledgments, if needed:
-\acknowledgments{Special Thanks to ... }
+\acknowledgments{This project was partially funded by the Austrian Research Promotion 
+Agency~(FFG) grant no.~$858537$.}
 

### 2019-02-07 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
adapted names of covariates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2451 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 075eeb86197634ee8c5ec854f61125cb34819c46

@@
 are depicted along with the estimated location parameter (red hand). The covariates
-employed are wind direction ($\text{degree}$), wind speed ($\text{ms}^{-1}$),
-and pressure gradients ($\text{hPa}$) west and east of the airport,
+employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
+and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
 all lagged by one hour.}

### 2019-02-07 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
Update affilitation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2452 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 87344947a7c4c486c38d7a84bb2b527b2fe3fd4a

@@
 %%%%%%%% IF DEPARTMENTS ARE LISTED AS WELL CHANGE NUMBERING!!!  %%%%%%%%%%%%%%
-\institute{Department of Statistics, Universit\"at Innsbruck, Innsbruck, Austria
-\and Department of Atmospheric and Cryospheric Science, Universit\"at Innsbruck, Innsbruck, Austria
-\and Epidemiology, Biostatistics and Prevention Institute, Universit\"at Z\"urich, Zurich, Switzerland}
+\institute{Department of Statistics, University of Innsbruck, Innsbruck, Austria
+\and Department of Atmospheric and Cryospheric Science, University of Innsbruck, Innsbruck, Austria
+\and Epidemiology, Biostatistics and Prevention Institute, University of Zurich, Zurich, Switzerland}
 

### 2019-02-08 pkg/disttree/inst/draft/rejoinder2.Rnw
further comments and revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2454 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7bafc7f1c6ded64c069588aad011b416b311048d

@@
 
-\setkomavar{subject}{Revision of AOAS1804-015}
+\setkomavar{subject}{Revision of AOAS1811-010}
 
@@
 the associate editor, and the two referees. Based on your comments, the
-manuscript has been revised and we feel that this substantially improved our
-contribution.
-%
-The most important changes are:
-\begin{itemize}
-
-\item \emph{Discussion.}\\
-TO DO
-
-\item \emph{\ldots}\\
-
-% \item In the calculation of variable importance we have detected some inconsistencies.
-% Thus, we have revised the applied method to evaluate this measurement and recalculated
-% variable importance for the distributional forest.
-
-\end{itemize}
-
-All changes and additions are explained in much more detail in the
+manuscript has been revised. In particular, an additional subsection
+has been added to the discussion commenting on the complexity of the
+applied distributions.
+%The most important changes are:
+%\begin{itemize}
+%\item \emph{Discussion.}\\
+%\item \emph{\ldots}\\
+%\end{itemize}
+All changes and extensions are explained in much more detail in the
 point-to-point reply on the next pages.
@@
 
-\vspace*{-2.8cm}
-\hspace*{-0.3cm} 
-\includegraphics[width = 0.23\textwidth]{signature-Schlosser}
+%\vspace*{-2.8cm}
+%\hspace*{-0.3cm} 
+%\includegraphics[width = 0.23\textwidth]{signature-Schlosser}
 
@@
 
-Thank you very much for your comment! East Tyrol lies in a different climat zone than the 
+Thank you very much for your comment! East Tyrol lies in a different climate zone than the 
 rest of Tyrol indicating different meteorological characteristics. For example, a covariate
@@
 looking at the results of the cross-validation in Supplement~B it can be observed that
-the difference in performance between the GAMLSS and the distributional tree models as
+the difference in performance between the GAMLSS and the distributional forest models as
 seen in the single evaluation evens out, hence the specific choice of learning years seems
@@
 \begin{review}
-The authors have well addressed the reviewers’ comments. The two added supplementary 
+The authors have well addressed the reviewers' comments. The two added supplementary 
 materials give a more solid ground to the proposed post-processing method by showing 
@@
 I recommend the paper to be accepted after the minor revisions proposed
-hereafter have been adressed.
+hereafter have been addressed.
 \end{review}
@@
 
-TO DO
-
-Thank you very much for the positive and constructive 
-comments regarding the paper.
+Thank you very much for your feedback and the constructive comments regarding the paper. 
+You are right that it still has to be tested how distributional trees and forests perform when
+employing very complex distributions such as the Dagum distribution. 
+However, contrary to the uncensored Gaussian distribution the parameters of the here applied 
+censored Gaussian distribution are not orthogonal as changing the mean also influences the variance,
+especially close-by the censoring point. Moreover, the three-parametric two-part hurdle model
+employed in Supplement A is almost equivalent to the zero-inflated model you suggested. 
+Hence, the distributional trees and forests have been tested for distributions with dependencies
+among their parameters and have shown to perform well. But of course further investigation is
+required in order to confirm these findings in general. This is now also mentioned in an additional 
+subsection of the discussion in the manuscript.
 

### 2019-02-08 pkg/disttree/inst/draft/draft.Rnw
further comments and revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2455 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 040ae125f359c8e60b9c8e376d9569c786f4c006

@@
 
+\subsubsection*{Distributional complexity / Orthogonality of distribution parameters}
+\hlc{Allowing for splits in each distribution parameter rather than only in the location parameter
+is particularly advantageous if the parameters are orthogonal, i.e. do not depend on each other,
+as changes in parameters other than the location parameter would be overlooked otherwise.
+If there is a dependency among parameters it can be sufficient considering only one parameter 
+in order to detect all changes, but this depends on the type of dependency.
+
+In the case of the here applied censored Gaussian distribution a shift in the mean also influences 
+the variance, hence the parameters are not orthogonal. The same holds for the 
+distributions employed in Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a}. 
+\hlc{Therefore, distributional trees and forests have shown to perform reasonably well also for  
+distributions with non-orthogonal parameters. However, testing a broader variety of 
+distributions, particularly very complex distributions such as the Dagum distribution}
+(see, e.g., \cite{Klein+Kneib+Lang:2015}) 
+\hlc{are required in order to confirm these findings in general.}
+
 \subsubsection*{Axams vs.~other meteorological stations}

### 2019-02-08 pkg/disttree/inst/draft/rejoinder2.Rnw
adapted response to Reviewer 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2456 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 314499749df9878932fd123c778d122447e73e95

@@
 Thank you very much for the positive and constructive feedback from you, 
-the associate editor, and the two referees. Based on your comments, the
-manuscript has been revised. In particular, an additional subsection
-has been added to the discussion commenting on the complexity of the
-applied distributions.
-%The most important changes are:
-%\begin{itemize}
-%\item \emph{Discussion.}\\
-%\item \emph{\ldots}\\
-%\end{itemize}
-All changes and extensions are explained in much more detail in the
-point-to-point reply on the next pages.
+the associate editor, and the two referees. We have incorporated all
+the remaining comments and, in particular, added a new subsection
+discussing more complex distributions with non-orthogonal parameters.
+All changes are explained in more detail in the point-to-point reply
+on the next pages.
 
@@
 
-The missing information and reference have been added in the 
-manuscript in Section~3.2 \textit{Models and evaluation: Distributional forest}.
-The described parameters pertain to the forest-building
-function \code{distforest}. But you are right, within the function \code{distforest} 
-the arguments \code{minsplit}, \code{minbucket} and \code{alpha} are handed over to the 
-tree-building function \code{disttree} which in this case applies the function \code{ctree}.
+We now point out that the arguments pertain to our function \code{distforest}
+and include a forward reference to the ``computational details''.
+
+You are also correct, though, that \code{ctree} takes these arguments. Internally,
+\code{distforest}/\code{disttree} pass the arguments on to \code{cforest}/\code{ctree}.
+This is now pointed out more clearly in the ``computational details''.
 
@@
 
-Thank you for pointing this out. We have rephrased this sentence in Section 3.3. of the manuscript.
+We have rephrased this sentence, emphasizing that the \emph{forecasted} distributions
+are depicted and avoiding an assessment.
 
@@
 
-Depending on the whether a maximum or a quadratic form of the test statistic is considered
-for the standardized test statistic its asymptotic conditional distribution is either a 
-normal ($c_{max}$) or a $\chi^2$ distribution ($c_{quad}$). 
-We have added this information to the addressed section of Appendix~A.
+Depending on whether a maximum or a quadratic form of the test statistic is considered,
+the asymptotic conditional distribution is either normal ($c_{max}$) or $\chi^2$ ($c_{quad}$). 
+We have added this information in Appendix~A.
 

### 2019-02-08 pkg/disttree/inst/draft/draft.Rnw
adapted response to Reviewer 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2456 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 314499749df9878932fd123c778d122447e73e95

@@
   considered for learning a forest of 100 trees. Subsampling is employed 
-  for each tree using a third of the predictors
-  in each split of the tree (\hlc{specified by the argument} \code{mtry} 
-  \hlc{in the applied function} \code{distforest}; 
-  \hlc{more information on the employed software is provided in Section}
-  ``Computational Details''). 
+  for each tree using a third of the predictors in each split of the tree
+  (\hlc{argument} \code{mtry} \hlc{in our implementation} \code{distforest},
+  \hlc{with more ``computational details'' provided at the end of the manuscript}).
   Parameters are estimated by
@@
 for July 24 in 2009--2012 have already been shown as a motivational example. 
-\hlc{In particular, the figure displays the point mass at zero (i.e., the probability 
-of a dry day) depicted by the model together with the forecasted probability 
+\hlc{In particular, the figure depicts the forecasted point mass at zero
+(i.e., the probability of a dry day) along with the forecasted probability 
 density function for the total amount of precipitation. Based on this illustration 
@@
 on \textsf{R}-Forge at
-(\url{https://R-Forge.R-project.org/projects/partykit/}). The function
-\code{distfit} fits distributional models by maximum likelihood, which is
-used as the basis for the tree-building function \code{disttree}, upon which
-the \code{distforest} is built. All functions can either be used with
-GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
+(\url{https://R-Forge.R-project.org/projects/partykit/}). 
+The function \code{distforest} learns the distributional forests proposed
+in this manuscript by combining the general \code{cforest} function from
+\textbf{partykit} with the function \code{distfit} for fitting distributional
+models by maximum likelihood. Analogously, \code{disttree} can learn a single
+distributional tree by combining \code{ctree} with \code{distfit}.
+All functions can either be used with GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
 (\citealp{Stasinopoulos+Rigby:2007}, version~\Sexpr{packageVersion("gamlss.dist")})
@@
 \cite{Strasser+Weber:1999} the test statistic can be standardized. 
-The observed multivariate linear statistic $t_l$ which is either an
-$k$~-~or~$k \cdot H$~-~dimensional vector, depending on the scale of $Z_l$,
+The observed multivariate linear statistic $t_l$ which is either a
+$k$- or $k \cdot H$-dimensional vector, depending on the scale of $Z_l$,
 is mapped onto the real line by a univariate test statistic $c$.
@@
 
-\hlc{Since} \cite{Strasser+Weber:1999} \hlc{showed that the asymptotic conditional 
-distribution of a linear statistics with conditional expectation} $\mu$ 
-\hlc{and covariance} $\Sigma$ 
-\hlc{is a multivariate normal distribution with parameters}
-$\mu$ and $\Sigma$ 
-\hlc{the asymptotic conditional distribution of the standardized test statistic}
+\cite{Strasser+Weber:1999} \hlc{showed that the asymptotic conditional 
+distribution of the linear statistic} $t_l$ \hlc{is a multivariate normal with parameters}
+$\mu$ and $\Sigma$. \hlc{Hence, the asymptotic conditional distribution of}
 $c(t_{l},\mu_{l},\Sigma_{l})$ 
 \hlc{is either normal}
-($c_{\text{max}}$) 
-\hlc{or follows a} 
-$\chi^2$ distribution ($c_{\text{quad}}$).
+(for $c_{\text{max}}$) 
+\hlc{or} 
+$\chi^2$ (for $c_{\text{quad}}$).
 

### 2019-02-09 pkg/disttree/inst/draft/draft.Rnw
- revised rejoinder and corresponding modifications in the manuscript - especially the discussion of more complex distributions should be checked again - also rejoinder on East Tyrol aspects should be improved
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2457 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d3a61fb932ac7bc551babc5d498793cb0b08415b

@@
 and generalized random forests for causal inferences 
-\citep{Athey+Tibshirani+Wager:2017}.
+\citep{Athey+Tibshirani+Wager:2019}.
 These procedures aggregate over trees fitted to specific score functions 
@@
 (PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) histograms for station
-Axams in Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} which 
+Axams in Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b} which 
 contains a more detailed explanation of residual QQ plots and PIT histograms.
@@
 
-\subsubsection*{Distributional specifications}
+\subsubsection*{Distributional specifications for precipitation modeling}
 
@@
 two alternative distributional specifications have been considered
-in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a}: Using the
+in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2018a}: Using the
 same evaluations as in Section~\ref{sec:all}, all models are
@@
 investigates the effects of using the same distributional family with a novel
-strategy for specifying dependence on covariates. Considering more flexible
-parametric specifications as well as transformation model specifications that
-allow to capture dependence of higher moments \citep[see e.g.,][]{Hothorn+Zeileis:2017}
-are extensions of interest for future research.
-
-\subsubsection*{Distributional complexity / Orthogonality of distribution parameters}
-\hlc{Allowing for splits in each distribution parameter rather than only in the location parameter
-is particularly advantageous if the parameters are orthogonal, i.e. do not depend on each other,
-as changes in parameters other than the location parameter would be overlooked otherwise.
-If there is a dependency among parameters it can be sufficient considering only one parameter 
-in order to detect all changes, but this depends on the type of dependency.
-
-In the case of the here applied censored Gaussian distribution a shift in the mean also influences 
-the variance, hence the parameters are not orthogonal. The same holds for the 
-distributions employed in Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a}. 
-\hlc{Therefore, distributional trees and forests have shown to perform reasonably well also for  
-distributions with non-orthogonal parameters. However, testing a broader variety of 
-distributions, particularly very complex distributions such as the Dagum distribution}
-(see, e.g., \cite{Klein+Kneib+Lang:2015}) 
-\hlc{are required in order to confirm these findings in general.}
+strategy for specifying dependence on covariates.
+
+\subsubsection*{More general distributional specifications}
+
+\hlc{Beyond the task of modeling precipitation it is of interest how well distributional
+forests perform in combination with other more general distributional specifications.
+It has been shown previously in the literature that using a score- or gradient-based
+selection of splitting variables outperforms a mean-based selection with subsequent
+flexible distributional modeling: For example, both} \citet[][Figure~2]{Athey+Tibshirani+Wager:2019}
+\hlc{and} \citet[][Figure~1]{Hothorn+Zeileis:2017} \hlc{demonstrated (independently) that their
+respective score-based random forest algorithms outperform the mean-based quantile
+regression forests of} \cite{Meinshausen:2006} \hlc{in a setup where only the variance of a normal
+response variable changes across the considered covariates. However, if all distribution
+parameters are closely correlated with the distribution mean the forests with different
+splitting strategies all perform similarly, provided a sufficiently flexible distribution
+is employed for the final predictions} \citep[see][Section~7]{Hothorn+Zeileis:2017}.
+
+\hlc{Similarly, the score-based distributional forests introduced in this manuscript
+proved to be quite robust to the different distributional specifications considered.
+While all specifications focus on capturing mean-variance effects note that these
+parameters are never fully orthogonal but can actually become quite closely correlated
+due to the censoring (or trunctation and/or zero-inflation considered in Supplement~A).
+
+However, exploring extensions to more flexible parametric distributions}
+\citep[e.g., such as the Dagum distribution considered by][in GAMLSS-type models]{Klein+Kneib+Lang:2015}
+\hlc{as well as transformation model specifications} \citep[e.g., as in][]{Hothorn+Zeileis:2017}
+\hlc{are of interest for future research.}
 
@@
 To show that qualitatively similar results are obtained for other
-meteorological stations, Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b}
+meteorological stations, Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b}
 carries out the same evaluation for 14~further stations. These
@@
 QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
-\citep[Supplement B]{Schlosser+Hothorn+Stauffer:2018b}.
+\citep[Supplement~B]{Schlosser+Hothorn+Stauffer:2018b}.
 Full replication of all results can be obtained with
@@
 also available within the \textsf{R}-Forge~project.
-The results presented in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a} and 
-Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
+The results presented in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2018a} and 
+Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
 \code{demo("RainDistributions", package = "disttree")} and 
@@
 \section*{Supplementary material}
-\subsection*{Supplement A: Different Response Distributions}
+\subsection*{Supplement~A: Different Response Distributions}
   To assess the goodness of fit of the Gaussian distribution, left-censored at zero,
@@
 
-\subsection*{Supplement B: Stationwise Evaluation}
+\subsection*{Supplement~B: Stationwise Evaluation}
   To show that Axams is a fairly typical

### 2019-02-09 pkg/disttree/inst/draft/rejoinder2.Rnw
- revised rejoinder and corresponding modifications in the manuscript - especially the discussion of more complex distributions should be checked again - also rejoinder on East Tyrol aspects should be improved
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2457 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d3a61fb932ac7bc551babc5d498793cb0b08415b

@@
 
-TO DO: adaption in the manuscript
+FIXME: adaption in the manuscript
+
+FIXME: I think we should present the two arguments more side by side, i.e., mention both
+early on and subsequently discuss both in more detail. (1) One factor is that only a single
+split into 28 years of training and 4 years of test data is considered in the main manuscript.
+When doing the full cross-validation the pronounced differences between the methods are much
+diminished. (2) East Tyrol is in a different climate zone and hence effects could be more
+additive and with less interactions than in North Tyrol.
 
@@
 particular when the distribution is more complex. The reason is that in all models considered, 
-the distribution parameters are "orthogonal" in the sense that the mean and the variance 
+the distribution parameters are ``orthogonal'' in the sense that the mean and the variance 
 parameter separate well.
+\end{review}
+
+Thank you for emphasizing this issue. We have now included some explicit pointers in Section~4
+to the previous random forest literature where it is indeed shown that the advantages of score-based
+split selection are particularly pronounced when the parameters are orthogonal. In contrast,
+when parameters are closely correlated with the distribution mean then the advantages are
+diminished and different random forest flavors typically perform similarly.
+
+Our results confirm these findings in the context of parametric distributional modeling,
+showing that results are quite robust to the distributional specification. Note also that
+in our models mean and variance are typically quite closely correlated due to the substantial
+censoring at zero (or truncation/zero-inflation in Supplement~A).
+
+\begin{review}
 This is not the case for many other distributions like the Dagum, Zero-inflated models, etc. 
@@
 sensitive with respect to the test and training data.
+\end{review}
 
+We do not expect identification issues but agree that the advantages over a method like
+Meinshausen's quantile regression forests will likely decrease or even disappear. However,
+we do not speculate about these properties in the manuscript but do point out that this
+deserves exploration in future research.
+
+\begin{review}
 In addition, overall I would say the method performs as well as the competitors but loses 
@@
 
-Thank you very much for your feedback and the constructive comments regarding the paper. 
-You are right that it still has to be tested how distributional trees and forests perform when
-employing very complex distributions such as the Dagum distribution. 
-However, contrary to the uncensored Gaussian distribution the parameters of the here applied 
-censored Gaussian distribution are not orthogonal as changing the mean also influences the variance,
-especially close-by the censoring point. Moreover, the three-parametric two-part hurdle model
-employed in Supplement A is almost equivalent to the zero-inflated model you suggested. 
-Hence, the distributional trees and forests have been tested for distributions with dependencies
-among their parameters and have shown to perform well. But of course further investigation is
-required in order to confirm these findings in general. This is now also mentioned in an additional 
-subsection of the discussion in the manuscript.
+FIXME: Torsten, should we comment on this? In a predictive setup like the precipitation
+forecasting the uncertainty of the estimates is not of so much concern. Also, the 
+uncertainty due to variation in the response is typically much larger.
 

### 2019-02-09 pkg/disttree/inst/draft/draft.Rnw
further revisions regarding the comment of the AE in the main manuscript and the rejoinder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2458 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3d50f9000476a0f25edaf2e24265d953cb86472f

@@
 (2009--2012).
+\hlc{Note that differences in the results of East Tyrol compared to the rest of 
+Tyrol could be (partly) due to East Tyrol lying in a different climate zone but 
+could also be caused by the specific selection of learning and validation period
+as only one single time split is considered. This will be discussed more
+in detail in the following section.}
 
@@
 parameters are never fully orthogonal but can actually become quite closely correlated
-due to the censoring (or trunctation and/or zero-inflation considered in Supplement~A).
+due to the censoring (or truncation and/or zero-inflation considered in Supplement~A).
 
@@
 where the prespecified GAMLSS performs best in the single-split setting
-(learning based on 1985--2008 and validation for 2009--2012). While part of
-the performance difference can also be attributed to different
-climatological characteristics in East Tyrol (that, unlike North Tyrol, is located
-south of the main Alpine Ridge), the differences are greatly reduced for several
-stations in East Tyrol by considering more splits in the full cross-validation.
+(learning based on 1985--2008 and validation for 2009--2012). 
+\hlc{The differences are greatly reduced for several stations in East Tyrol by considering 
+more splits in the full cross-validation. However, part of the performance difference can also 
+be attributed to varying climatological characteristics in East Tyrol that, unlike North Tyrol,
+is located south of the main Alpine Ridge. Therefore, effects could be more additive
+and with less interactions than in North Tyrol and thus be captured more precisely by
+an additive model such as GAMLSS, particularly with an appropriate prespecification of covariates.}
 

### 2019-02-09 pkg/disttree/inst/draft/rejoinder2.Rnw
further revisions regarding the comment of the AE in the main manuscript and the rejoinder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2458 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3d50f9000476a0f25edaf2e24265d953cb86472f

@@
 
-FIXME: adaption in the manuscript
+% FIXME: adaption in the manuscript
 
-FIXME: I think we should present the two arguments more side by side, i.e., mention both
-early on and subsequently discuss both in more detail. (1) One factor is that only a single
-split into 28 years of training and 4 years of test data is considered in the main manuscript.
-When doing the full cross-validation the pronounced differences between the methods are much
-diminished. (2) East Tyrol is in a different climate zone and hence effects could be more
-additive and with less interactions than in North Tyrol.
+% FIXME: I think we should present the two arguments more side by side, i.e., mention both
+% early on and subsequently discuss both in more detail. (1) One factor is that only a single
+% split into 28 years of training and 4 years of test data is considered in the main manuscript.
+% When doing the full cross-validation the pronounced differences between the methods are much
+% diminished. (2) East Tyrol is in a different climate zone and hence effects could be more
+% additive and with less interactions than in North Tyrol.
 
-Thank you very much for your comment! East Tyrol lies in a different climate zone than the 
-rest of Tyrol indicating different meteorological characteristics. For example, a covariate
+Thank you very much for your comment! One important fact that could be a reason for the
+different results is that East Tyrol does not lie in the same climate zone as the rest of Tyrol 
+indicating different meteorological characteristics. For example, a covariate
 might have a linear effect on the expected amount of rain in East Tyrol while it has a
@@
 is correctly specified in the model (which might be the case in the prespecified GAMLSS).
-Of course, these are just assumptions and there could be many other reasons. However,
-looking at the results of the cross-validation in Supplement~B it can be observed that
-the difference in performance between the GAMLSS and the distributional forest models as
-seen in the single evaluation evens out, hence the specific choice of learning years seems
-to have an impact as well.
+Of course, this is just an assumption and there could be other reasons for the performance
+difference between the GAMLSS and the distributional forest models in East Tyrol. However, 
+looking at the results of the cross-validation in Supplement~B it can be observed that this 
+difference as seen in the single evaluation evens out for several stations in East Tyrol, 
+hence the specific choice of learning and validation period seems to have an impact as well.
+These two aspects are now also discussed more in detail in the main manuscript.
 

### 2019-02-10 pkg/disttree/inst/draft/draft.Rnw
move all new comments about Easty Tyrol to Section 3.4, some streamlining
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2459 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 222b3af3a649bb2f808931a750f1359e9e91411d

--- 2019-02-09 pkg/disttree/inst/draft/draft.Rnw
@@
 forest had higher (green) or lower (red) CRPS compared to the best of the 
-other three models. The background shows the local topography \citep{Robinson:2014}.
+other three models. The gray background shows the local topography \citep{Robinson:2014}.
 Station Axams is highlighted in bold.}
@@
 the prespecified GAMLSS performs particularly well in the validation period
-(2009--2012).
+(2009--2012). \hlc{Partially, this can be attributed to random variation as
+the differences at several stations are mitigated when considering a full
+cross-validation rather than a single split into learning and validation period}
+\citep[see Supplement~A,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2018a}.
+\hlc{Further differences are possibly due to East Tyrol lying in a different
+climate zone, south of the main Alpine Ridge. Hence, long-term climatological
+characteristics as well as the precipitation patterns in 2009--2012 differ
+from North Tyrol, conforming particularly well with the additive effects from
+the prespecified GAMLSS.}
 
@@
 where the prespecified GAMLSS performs best in the single-split setting
-(learning based on 1985--2008 and validation for 2009--2012). While part of
-the performance difference can also be attributed to different
-climatological characteristics in East Tyrol (that, unlike North Tyrol, is located
-south of the main Alpine Ridge), the differences are greatly reduced for several
-stations in East Tyrol by considering more splits in the full cross-validation.
+(learning based on 1985--2008 and validation for 2009--2012).
 

### 2019-02-10 pkg/disttree/inst/draft/rejoinder2.Rnw
rephrased rejoinder comments regarding East Tyrol
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2460 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 18cee89968efc59466a8e2a36bda56a31e5f2230

--- 2019-02-09 pkg/disttree/inst/draft/rejoinder2.Rnw
@@
 
-FIXME: adaption in the manuscript
+East Tyrol differs from North Tyrol as it is located south of the main Alpine Ridge and thus
+in a different climate zone. Consequently, both long-term climatological characteristics
+and precipitation patterns in 2009--2012 differ. This is not visible from Figure~8 directly,
+though, and hence we added a short explanation at the end of Section~3.4
 
-FIXME: I think we should present the two arguments more side by side, i.e., mention both
-early on and subsequently discuss both in more detail. (1) One factor is that only a single
-split into 28 years of training and 4 years of test data is considered in the main manuscript.
-When doing the full cross-validation the pronounced differences between the methods are much
-diminished. (2) East Tyrol is in a different climate zone and hence effects could be more
-additive and with less interactions than in North Tyrol.
+Moreover, sampling variability clearly also plays a role as the differences from Figure~8
+are somewhat evened out in the cross-validation, specifically for some of the East Tyrolian 
+stations. This is shown in Supplement~B and was already discussed in the corresponding
+paragraph in Section~4 (Axams vs.\ other meteorological stations). Additionally, we now
+include some comments on this at the end of Section~3.4 where this was admittedly lacking
+in the previous version of the manuscript.
 
-Thank you very much for your comment! East Tyrol lies in a different climate zone than the 
-rest of Tyrol indicating different meteorological characteristics. For example, a covariate
-might have a linear effect on the expected amount of rain in East Tyrol while it has a
-non-additive effect in Northern Tyrol. In this case, in East Tyrol an additive model such as 
-GAMLSS will capture this influence better than a tree model, particularly if this covariate
-is correctly specified in the model (which might be the case in the prespecified GAMLSS).
-Of course, these are just assumptions and there could be many other reasons. However,
-looking at the results of the cross-validation in Supplement~B it can be observed that
-the difference in performance between the GAMLSS and the distributional forest models as
-seen in the single evaluation evens out, hence the specific choice of learning years seems
-to have an impact as well.
+% Thank you very much for your comment! One important fact that could be a reason for the
+% different results is that East Tyrol does not lie in the same climate zone as the rest of Tyrol 
+% indicating different meteorological characteristics. For example, a covariate
+% might have a linear effect on the expected amount of rain in East Tyrol while it has a
+% non-additive effect in Northern Tyrol. In this case, in East Tyrol an additive model such as 
+% GAMLSS will capture this influence better than a tree model, particularly if this covariate
+% is correctly specified in the model (which might be the case in the prespecified GAMLSS).
+% Of course, this is just an assumption and there could be other reasons for the performance
+% difference between the GAMLSS and the distributional forest models in East Tyrol. However, 
+% looking at the results of the cross-validation in Supplement~B it can be observed that this 
+% difference as seen in the single evaluation evens out for several stations in East Tyrol, 
+% hence the specific choice of learning and validation period seems to have an impact as well.
+% These two aspects are now also discussed more in detail in the main manuscript.
 

### 2019-02-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added section on computation details; further text revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2461 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 581da9ae784df1de924164984aff000212702893

@@
 \usepackage{enumitem}
+\usepackage{color,soul}
 
@@
 
+\definecolor{changes}{rgb}{1,1,0.7}
+\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}
+
 <<preliminaries, echo=FALSE, results=hide>>=
@@
 \section{Introduction}\label{sec:introduction}
-In many situations fitting one global model to a data set can be very challenging, 
-especially if this data contains lots of different features with a strong variation 
+In many situations fitting one global model to a given data set can be very challenging, 
+especially if the data contains lots of different features with a strong variation 
 or complex interactions. Therefore, separating the data into more homogeneous subgroups 
 based on a set of covariates first can simplify the task and fitting a local model to 
-each of the resulting subgroups often leads to better results. This separation of the 
-data can be done by applying a tree algorithm. Many different algorithms have been developed
-in this research field. While they all follow the general idea of splitting the data such 
-that some objective function is optimized, they differ in their 
-specific approaches to selecting a split variable and the corresponding split point. 
+each of the resulting subgroups often leads to better results. This separation can be 
+done by applying a tree algorithm. While all of the many developed algorithms 
+follow the general idea of splitting the data such that some objective function is optimized, 
+they differ in their specific approaches to selecting a split variable and the corresponding 
+split point. 
 Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
@@
 choosing the covariate which shows the highest association to the response variable 
-based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
-tests for contingency tables, CTree applies permutation tests in a conditional inference
-framework and MOB uses fluctuation tests based on asymptotic theory. All these approaches
-have shown to work well for various situations, however, the specific advantages and
-disadvantages of each of the testing strategies have not yet been investigated and compared
-in detail.
+based on $p$-values provided by a statistical test. While QUEST, CHAID and GUIDE employ 
+statistical significance tests for contingency tables, CTree applies permutation tests in 
+a conditional inference framework and MOB uses fluctuation tests based on asymptotic theory. 
+All these approaches have shown to work well for various situations, however, the specific 
+advantages and disadvantages of each of the testing strategies have not yet been 
+investigated and compared in detail.
 
@@
 i.e. the task of selecting the best split variable. In particular, the approach of the 
-GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm. 
+GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
+by investigating the building blocks of their testing strategies in which they differ:
+(1)~Transformations of the response: using residuals or full model scores
+(2)~Binarization of residuals/scores
+(3)~Categorization of covariates.
 For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
-algorithms is presented such that each of the three strategies can be obtained by a specific 
-combination of the available building blocks. In that way, the effects of each of these 
-features can be investigated separately by varying the building blocks in the tool box 
-provided by this general framework.
+algorithms is presented such that each of the three strategies GUIDE, CTree, and MOB can be 
+obtained by a specific combination of the available building blocks. In that way, the 
+effects of each of these features can be investigated separately by varying the building 
+blocks in the tool box provided by this general framework. 
+Additionally, the impact of applying either a pre- or postpruning strategy is analyzed.
 
@@
 model scores is at least as efficient but mostly clearly advantageous over considering only 
-residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
-the values of possible split variables are investigated.
-Additionaly, the impact of applying either a pre- or postpruning strategy is analyzed.
+residuals. Also regarding the effects of binarizing residuals or scores at $0$ or categorizing
+the values of possible split variables clear recommendations can be made based on the presented 
+results.
 
 In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree model is given. 
-Then, after explaining the investigated algorithms CTree, MOB, and GUIDE more in detail in 
-Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
-testing strategies in model-based unbiased recursive partitioning algorithms is presented in 
-Section~\ref{sec:unify}.
+After explaining the investigated algorithms CTree (Section~\ref{sec:ctree}), 
+MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}) more in detail the applied pruning strategies 
+(\ref{sec:pruning}) are discussed.
+In Section~\ref{sec:unify} the unifying framework for testing strategies in model-based unbiased 
+recursive partitioning algorithms is presented.
 The setting for the simulation study is introduced in Section~\ref{sec:simulation}
-and the results are discussed in Section~\ref{sec:results}.
+and the results are illustrated and discussed in Section~\ref{sec:results}.
 
@@
 \subsection{Generic Algorithm}
-The basic idea of building a regression tree model is to partition the data into smaller and 
-more homogeneous subgroups based on a set of covariates. Various different tree algorithms 
-have been developed, all following the same general structure starting at the root of the tree
-which is simply the first node containing all available data:
+\label{sec:generic_alg}
+The basic idea of building a regression tree model is to partition the data into smaller 
+and more homogeneous subgroups based on a set of covariates. Various different tree 
+algorithms have been developed, all following the same general structure starting at 
+the root of the tree which is simply the first node containing all available data:
 \begin{enumerate}
@@
 The employed discrepancy measure can for example be the sum of deviations from a typical 
-or average value, but it can also be constructed based on a parametric model. For instance,
-different types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
-well as rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
+or average value, but it can also be constructed based on a parametric model fitted in each 
+node. For instance, different types of residuals or their signs can be applied 
+\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores as 
+in \cite{Hothorn+Hornik+Wiel:2006}.
 %(In the simulation study of this paper trees with linear models fitted in each node are 
@@
 As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
-GUIDE first select the split variable and then the split point in a separate step. 
+GUIDE first select the split variable and then, in a separate step, the split point. 
 For the first step of selecting a split variable they all apply statistical tests
@@
 on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
-current node and $k \in \mathbb{N}$.
-\item Using this matrix, the algorithm-specific test statistic is evaluated for each possible 
+current node and the parameter $k \in \mathbb{N}$ defined by the algorithm.
+\item Using this matrix the algorithm-specific test statistic is evaluated for each possible 
 split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
-of $Z_j$ on $D$.
+of $D$ on $Z_j$.
 \item For the resulting values of the test statistics the corresponding $p$-values are calculated
@@
 and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
-choice, while a categorical response can be mapped to a unity vector by an indicator function 
+choice while a categorical response can be mapped to a unity vector by an indicator function 
 $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
@@
 variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
-corresponding model scores by setting $h(Y) = s(Y,X,\hat{\beta})$
-where $\hat{\beta}$ is the estimated model parameter and $s$ is the score function
+corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
+where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
 $$
-s(Y, X, \hat{\beta}) = \frac{\partial \psi(Y, X, \hat{\beta})}{\partial \beta}
+s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
 $$
 with the sum of $\psi$ over all observations being the objective function that is optimized 
-to estimate the model parameter $\beta$. For this purpose many different M-type estimators can 
-be applied including maximum likelihood or ordinary least squares. Therefore, these objective 
-functions are also possible options to obtain a model-based discrepancy measure for the tree 
-algorithm.
-
-Applying the score function $s$ on a data set of $n$ observations containing the 
-response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
-regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
-on a $k$-dimensional estimated parameter vector 
-$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
-yields a $n \times k$-matrix $S$.
+to obtain the estimated model parameter $\hat{\beta}$. For this purpose many different M-type 
+estimators can be applied including maximum likelihood or ordinary least squares such that the 
+corresponding objective function can be employed as model-based discrepancy measure for the 
+tree algorithm. Evaluationg the score function $s$ on a data set of $n$ observations 
+%containing the 
+%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
+%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
+and a $k$-dimensional estimated parameter vector 
+%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
+yields a $n \times k$-matrix.
 In that way a score value indicating how well the model fits the data is given separately
@@
 In \cite{Zeileis+Hothorn+Hornik:2008} generalized linear models (GLMs) and survival regression models 
-are used to exemplify the algorithm. Subsequently various other types of models have been applied as well,
+are used to exemplify the algorithm. Subsequently, various other types of models have been applied as well,
 for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
@@
 Regardless of the applied model type the algorithm only requires the corresponding model 
-score matrix $S$ in order to test for any instability in the model parameters over each 
+score matrix in order to test for any instability in the model parameters over each 
 particular ordering induced by each of the possible partitioning variables $Z$.
-If a variable $Z$ has no specific influence on the parameters the scores would always 
-fluctuate randomly around their mean 0. Therefore, these deviations are investigated by 
-measuring them using an empirical fluctuation process. Applying a scalar function which 
+If a variable $Z$ has no specific influence on the parameters the ordered scores would 
+always fluctuate randomly around their mean 0. Therefore, these deviations are investigated 
+by measuring them using an empirical fluctuation process. Applying a scalar function which 
 captures the fluctuation in this process yields the test statistic of this so called 
@@
 the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
-a constant fit where the parameter $\beta$ is the mean of the response, hence 
-$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ for a response vector 
-$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, can be employed as well as more complex models.
+a constant fit whith the parameter $\beta$ being the mean of the response, hence 
+$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ 
+%for a response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, 
+can be employed as well as more complex models.
 For the statistical test applied to select a split variable some type of model-based
-residuals is required. In \cite{Loh:2002} linear models are used as an example where the
+residual is required. In \cite{Loh:2002} linear models are used as an example where the
 residuals are defined by
@@
 Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
-basic structure. Thus they can be presented together in a unifying framework for testing strategies
+basic structure as explained in Section~\ref{sec:generic_alg}. 
+Thus they can be presented together in a unifying framework for testing strategies
 in model-based unbiased recursive partitioning. 
 In this framework each specific testing strategy can be obtained by combining the provided 
-building blocks accordingly. 
-This general toolbox covers all features of the here discussed testing strategies CTree, MOB, and
-GUIDE. Furthermore, it allows for all possible combinations of these features and in that way also
-for variations of the commonly known testing strategies. Moreover, by exchanging only one building
-block its impact can be investigated in detail.
+building blocks accordingly. But apart from that, variations of the commonly known testing 
+strategies can be obtained as this general toolbox allows for all possible combinations of the 
+building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
 These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
@@
 By including all available scores rather than just residuals the goodness of fit of the model is 
-investigated regarding each parameter instead of just the mean/location parameter. 
-Therefore, this facilitates the detection of changes in parameters other than the mean/location
-parameter.
+investigated regarding each parameter instead of just the mean. 
+%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.
 
@@
 \subsection{Linear Model Tree}\label{sec:lmtrees}
-As an example a linear model tree is considered as this allows 
-for an easy and well structured description of the investigated subjects/issues. 
-For that reason, this model is also considered for the simulation study in 
-Section~\ref{sec:simulation}.
-However, the conclusions drawn from this example also apply for a wide range of other model
-types. The models fitted in the nodes of the considered linear model tree are of the form
+As an example a linear model tree is considered as this allows for an easy and well structured 
+description of the investigated subjects. However, the conclusions drawn from this example also 
+apply for a wide range of other model types. 
+The models fitted in the nodes of the linear model tree as considered for the simulation 
+study in Section~\ref{sec:simulation} are of the form
 $$
@@
 for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
-As for each node the type of the fitted model is the same no node-specific features are considered
-and the observations of the current (arbitrarily chosen) node are denoted by 
+For each node the type of the fitted model is the same such that no node-specific features are 
+considered and the observations of the current (arbitrarily chosen) node are denoted by 
 $\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.
@@
 by $(\bar{y} - y_i)$.
-However, this includes only the response variable $Y$ but not the fitted model.
+However, this includes only the response variable $Y$ but not the fitted linear model.
 For a model-based approach residuals can be considered defined by
@@
 $$
-\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2.
+\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
 $$
 such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
-which is minimized for the estimation of the coefficient vector
-$\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$, 
-i.e. to obtain $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
+which is minimized to obtain the estimated coefficient vector
+$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
 The corresponding score function is defined by
@@
 $$
-where the factor $2$ is often omitted such that the first column of the score matrix $S$ 
-corresponds to the residuals. $S$ is in this case a $n \times 2$-matrix
-as $\boldsymbol{\beta}$ is two dimensional. In that way a score value indicating how well the model
-fits the data is given separately for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$
-with $i \in \{1,\dots,n\}$ and $j \in \{1,2\}$.
-
-In this example considering only residuals, i.e. only the first column of the score matrix $S$,
+where the factor $2$ is often omitted such that the first column of the $n \times 2$-dimensional
+score matrix corresponds to the residuals. 
+Therefore, including only residuals (i.e. only the first column of the score matrix)
 corresponds to including the derivatives of $\psi$ only regarding the intercept 
 parameter~$\beta_0$.
-On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
-the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
-$\beta_1$ (second column).
+On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
+the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
+parameter $\beta_1$ (second column) are considered.
 
@@
 Each generated data set consists of a response variable, one regressor, one split 
-variable and nine noise variables. In particular, the variables listed in 
-Table~\ref{tab:variables} together with the corresponding distributions are 
-included in the data generating process.
+variable and nine noise variables as listed in Table~\ref{tab:variables} 
+together with the corresponding distributions.
 
@@
 
-By applying the unifying framework presented in Section~\ref{sec:unify} the effect of
-each building block can be investigated on its own. In Table~\ref{tab:buildingblocks} the setting
-of these building blocks is listed for CTree, MOB and GUIDE in their original form.
-However, in order to be able to focus on particular features rather than the whole testing
-strategies other combinations as presented in Table~\ref{tab:buildingblocks}, hence variations of
-the original testing strategies, are evaluated as well.
-
-Moreover, as prepruning is employed all of the considered statistical tests are applied as significance tests answering two questions at once: 
+As explained before, the aim of this simulation study is to investigate the effects
+of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
+rather than the whole testing strategies. Therefore, other combinations as presented in 
+Table~\ref{tab:buildingblocks}, hence variations of GUIDE, CTREE, and MOB are evaluated as well
+as their original versions. Moreover, as prepruning is employed all of the considered statistical 
+tests are applied as significance tests answering two questions at once: 
 (1)~Should a split be performed at all? 
@@
 to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
-than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is 
-selected.
-
-\newpage
+than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.
+
+%\newpage
 \section{Results}
@@
 to the true split variable $Z_1$. 
-This effect can be observed for a central split point (left panel, $xi=0$) as well as for a
-a split point closer to upper end of the considered interval (right panel, $xi=0.8$).
+This effect can be observed for a central split point (left panel, $\xi=0$) as well as for a
+a split point closer to upper end of the considered interval (right panel, $\xi=0.8$).
 However, the effect of the building block \textit{Categorization} varies over these two settings.
@@
 
-\fixme
 From comparing the effects of the building blocks three main conclusions can be drawn: 
@@
 However, even then the results are very similar to those achieved without any categorization.
-
-%FIX ME: results without loss of generalization
+But it has to be noted that a categorization of possible split variables can facilitate
+evaluations. Moreover, depending on the focus of interest, the results can be more
+straightforward, for example in case of clearly separated subgroups being defined by the values 
+of a continuous split variable. Therefore, categorizing possible split variable can have
+advantages in practice, however, it has to be considered that in general
+important effects might be neglected.
+
+Linear models as presented in this study have been selected as they allow for a simple illustration
+of the investigated subjects. However, the obtained results also hold for a variety of other models
+as the introduced building blocks can be applied in the same way for all model types.
+Furthermore, with only two parameters determining a linear model the importance of including scores 
+regarding each parameter could already be pointed out clearly. But for an increasing number of 
+parameters in a model this aspect becomes even more important as it allows for the detection of changes 
+in each parameter.
+
+\section*{Computational Details}
+\label{sec:comp}
+The applied implementation is based on the \textsf{R} package \textbf{partykit}
+(version~\Sexpr{packageVersion("partykit")}) which is available 
+on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/}).
+The code to reproduce the simulation study can also be found on \textsf{R}-Forge
+at (\url{https://R-Forge.R-project.org/projects/partykit/inst/guideliketests}).
+
+The functions \code{ctree} and \code{mob} provide an implementation of the
+two tree algorithms in their original form. For their adapted versions additional
+modifications have been applied within these functions allowing for a categorization
+of possible split variables and a binarization of scores.
+To evaluate the GUIDE algorithm a reimplementation of this algorithm has been 
+established/constructed using the basic framework of \code{ctree} and \code{mob}.
+
+\section*{Acknowledgements}
+\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
+2017 to January 2018) was financially supported by the Swiss National
+Science Foundation, grant number SNF IZSEZ0 177091.}
 

### 2019-02-12 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further text revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2462 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 00f139ebb2799a8a337f5ea1a28509e978aaea20

@@
 In many situations fitting one global model to a given data set can be very challenging, 
-especially if the data contains lots of different features with a strong variation 
-or complex interactions. Therefore, separating the data into more homogeneous subgroups 
+especially if the data contains lots of different features with strong variation 
+and complex interactions. Therefore, separating the data into more homogeneous subgroups 
 based on a set of covariates first can simplify the task and fitting a local model to 
@@
 residuals. Also regarding the effects of binarizing residuals or scores at $0$ or categorizing
-the values of possible split variables clear recommendations can be made based on the presented 
+the values of possible split variables clear recommendations can be given based on the presented 
 results.
@@
 corresponding objective function can be employed as model-based discrepancy measure for the 
-tree algorithm. Evaluationg the score function $s$ on a data set of $n$ observations 
+tree algorithm. Evaluating the score function $s$ on a data set of $n$ observations 
 %containing the 
@@
 To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
-Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
-statistic given all permutations of the response variable can be calculated and used to 
+Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the linear
+test statistic given all permutations of the response variable can be calculated and used to 
 standardize the test statistic. In order to map multivariate test statistics into the real 
@@
 statistic or a quadratic form. The resulting values can then be transformed to 
-the scale of $p$-values where they can be compared in an unbiased way to select a split variable.
+the scale of $p$-values allowing for an unbiased comparison in order to select a split variable.
 
@@
 correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
-is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
-provides a very general framework where the choice of the type of variables and the corresponding 
+is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree 
+algorithm provides a very general framework where the type of variables and the corresponding 
 functions $g$ and $h$ determine the type of the applied statistical test.
@@
 the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
-a constant fit whith the parameter $\beta$ being the mean of the response, hence 
-$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ 
+a constant fit with the parameter $\beta$ being the mean of the response, hence 
+$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$, 
 %for a response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, 
@@
 it is shown that these $p$-values can also be used as a basis for a
-pruning strategy namely prepruning: Only if the $p$-value corresponding to the selected split
+pruning strategy namely \textit{prepruning}: Only if the $p$-value corresponding to the selected split
 variable (hence the smallest $p$-value) is smaller than a given significance level a split is
@@
 
-For postpruning no significance level is included leading to very large trees which are then 
-pruned afterwards, for example by employing cost-complexity pruning as in the 
+On the contrary, when employing postpruning no significance level is included leading to very large 
+trees which are then pruned afterwards, for example by applying cost-complexity pruning as in the 
 CART algorithm \citep{Breiman+Friedman+Stone:1984}. 
@@
 but as prepruning only requires already provided information it is usually less computationally
-expensive than the additional calculations needed for postpruning.
+expensive than postpruning needing additional calculations. 
 However, if the test performs poorly in terms of power and overfitting, prepruning suffers from 
 this deficiency while postpruning can compensate it at least partly and improve the results.
-For this reason we are focusing on the testing strategies applied in tree algorithms and 
-investigate how effective they are as significance tests.
+Therefore, being aware of the power of the selected testing strategy is essential for choosing 
+an appropriate pruning strategy.
+For that reason we are focusing on the testing strategies applied in tree algorithms 
+and investigate how effective they are as significance tests.
 
@@
 of parametric models together with prepruning as the properties of this pruning strategy
-are already explained/determined by the testing strategy. Subsequently combinations
+are already determined by the testing strategy. Subsequently combinations
 with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
@@
 Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
-and 3rd quartile are used to bin each possible split variable into four categories. In that way,
+and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
 depending on whether the scores are binarized or not, a $\chi^2$-test or 1-way ANOVA can be applied
@@
 description of the investigated subjects. However, the conclusions drawn from this example also 
-apply for a wide range of other model types. 
+hold for a wide range of other model types. 
 The models fitted in the nodes of the linear model tree as considered for the simulation 
@@
 $$
-s(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
-\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
-2 \cdot r(y_i,x_i,\hat{\beta_0},\hat{\beta_1}) \cdot (1,x_i)^\top
+s(y_i, x_i, \beta_0, \beta_1) = 
+\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
+2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
 $$
@@
 parameter $\beta_1$ (second column) are considered.
+
+\newpage
 
@@
 rather than the whole testing strategies. Therefore, other combinations as presented in 
-Table~\ref{tab:buildingblocks}, hence variations of GUIDE, CTREE, and MOB are evaluated as well
+Table~\ref{tab:buildingblocks}, hence adapted versions of GUIDE, CTREE, and MOB are evaluated as well
 as their original versions. Moreover, as prepruning is employed all of the considered statistical 
@@
 Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
-scenarios of varying only the intercept $\beta_0$ or only the slope parameter $\beta_1$
-or both of them, once for a true split point $\xi=0$ and once for $\xi=0.8$. 
+scenarios (varying only the intercept $\beta_0$, only the slope parameter $\beta_1$
+or both of them) once for a true split point $\xi=0$ and once for $\xi=0.8$. 
 As explained before, the tests are applied as significance tests with significance 
@@
 center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
-$\delta = 0.3$ is considered making the differences particularly well observable. 
-(Additional results over increasing $\delta$ as in Section~\ref{sec:scores} 
-can be found in Appendix~\ref{app:increasingdelta}.)
+$\delta = 0.3$ is considered making the differences particularly well observable 
+(additional results over increasing $\delta$ as in Section~\ref{sec:scores} 
+can be found in Appendix~\ref{app:increasingdelta}).
 The power of the different combinations is measured by the mean $p$-value for the true split 
 variable~$Z_1$.
-The possible values for each of the three building blocks are listed in Table~\ref{tab:levels}.
-In order to evaluate all 12 combinations the original testing strategies CTree, MOB, and GUIDE
-are applied together with variations of them. For each combination the corresponding employed
-strategy can be found in Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.
+The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
+In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
+and variations of them. A more detailed overview of all 12 combinations can be found in 
+Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.
 
@@
 
-\newpage
+%\newpage
 \subsection{Tree}
@@
 
-\newpage
+%\newpage
 
@@
 (version~\Sexpr{packageVersion("partykit")}) which is available 
-on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/}).
-The code to reproduce the simulation study can also be found on \textsf{R}-Forge
-at (\url{https://R-Forge.R-project.org/projects/partykit/inst/guideliketests}).
+on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
+where the code to reproduce the simulation study can be found as well.
 

### 2019-02-13 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revision on discussion; correct paths in simulation code
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2463 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d1f8ff4b6250a557ba335c0c39dc0d75c0087fc8

@@
 
+Note that considering full model scores is particularly advantageous if the model parameters do not
+depend on each other, hence are orthogonal, as it is the case for the intercept and slope parameter
+of a linear model. On the other hand, if all parameters are closely correlated with the location parameter
+a residual-based approach might be able to detect changes in other parameters than the mean as well,
+however, the performance of a score-based search for split variables is not negatively effected.
+
+\newpage
+
 \section*{Computational Details}

### 2019-02-13 pkg/partykit/inst/guideliketest/tests_guideprune.R

@@ -1,12 +1,12 @@
 library("partykit")
 library("Formula")
 library("parallel")
-source("~/svn/partykit/pkg/partykit/inst/guideliketest/guidelike.R")
+# source("~/svn/partykit/pkg/partykit/inst/guideliketest/guidelike.R")
 # source("~/partykit/inst/guideliketest/guidelike.R")
-# source("guidelike.R")
-source("~/svn/partykit/pkg/partykit/inst/guideliketest/ccprune.R")
+source("guidelike.R")
+# source("~/svn/partykit/pkg/partykit/inst/guideliketest/ccprune.R")
 # source("~/partykit/inst/guideliketest/ccprune.R")
-# source("ccprune.R")
+source("ccprune.R")
 
 # function to compute adjusted Rand Index
 adj_rand_index <- function(x, y) {

### 2019-02-13 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further text revisions and correct path to source code
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2465 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4a9e11b2144c6f6045be53186dd6de2db3dd93eb

@@
 
-%% need no \usepackage{Sweave} ?
-%\usepackage{Sweave}
+%% need no \usepackage{Sweave} 
 \SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
@@
 options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
+
 library("partykit")
@@
 library("scales")
-# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
+
 source("../guidelike.R")
 source("../tests_guideprune.R")
+source("../ccprune.R")
 
@@
 an unbiased variable selection.
-%As the variables are usually not all measured at the same scale the resulting
-%values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
-%necessary to calculate the corresponding $p$-values in order to allow for an unbiased
-%variable selection on this standardized/unified scale. 
 \item The variable corresponding to the smallest $p$-value is selected as split variable as 
@@
 
-%\newpage
 <<dgp, echo=FALSE>>=

### 2019-02-18 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revision of discussion regarding linear models
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2468 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 724c1d12ecdff22908887a991eb10f6768d10b9f

@@
 From comparing the effects of the building blocks three main conclusions can be drawn: 
-(1)~In most cases including all available score values rather than just residuals leads to 
+(1)~In many cases including all available score values rather than just residuals leads to 
 clearly better results and performs in general at least equally well.
@@
 as the introduced building blocks can be applied in the same way for all model types.
+One specific feature of linear models is the orthogonality of the conditional expectation
+$\mathbb{E}(Y|X)$ and the regressor $X$ which affects the capability of tests using only residuals
+to capture shifts in the parameters. In particular, \cite{Ploberger+Kramer:1992} have shown that
+residual-based tests are insensitive to parameter changes orthogonal to the mean regressor.
+For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
+$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
+Testing procedures purely based on residuals perform best in situations with shifts only in the
+intercept, however, the costs in terms of power that come with including all scores are 
+negligibly small. 
 Furthermore, with only two parameters determining a linear model the importance of including scores 
-regarding each parameter could already be pointed out clearly. But for an increasing number of 
-parameters in a model this aspect becomes even more important as it allows for the detection of changes 
-in each parameter.
-
-Note that considering full model scores is particularly advantageous if the model parameters do not
-depend on each other, hence are orthogonal, as it is the case for the intercept and slope parameter
-of a linear model. On the other hand, if all parameters are closely correlated with the location parameter
-a residual-based approach might be able to detect changes in other parameters than the mean as well,
-however, the performance of a score-based search for split variables is not negatively effected.
-
-\newpage
+regarding each parameter could already be pointed out clearly while for an increasing number of 
+parameters in a model this aspect becomes even more important.
+
+In general residual-based and score-based procedures are expected to perform equally well if all
+parameters are highly correlated. On the other hand, if parameters are orthogonal changes that are 
+not connected to the mean might be missed when only considering residuals. This can be avoided by
+including scores which are a standard measurement of goodness of fit that can be easily employed.
+Even though it also depends on the type of the applied residuals 
+(see, e.g., \cite{Loh+Zheng:2013}, \cite{Chaudhuri+Loh:2002} 
+and \cite{Loh+He+Man:2015})
+whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
+fit such as residuals provides less power than full model scores. 
+Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
+determined by the way how residuals are connected to parameter shifts as well as the type of parameter
+shift and the type of model. Therefore, a general statement for all types of models and residuals can 
+hardly be made.
+
+%\newpage
 

### 2019-02-18 pkg/disttree/inst/draft/rejoinder2.Rnw
final revision and formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2470 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bc69d6b1df1a94a14c4954e27896487bb7cffb6f

@@
 ``Distributional regression forests for probabilistic precipitation 
-forecasting in complex terrain'' (AOAS1804-015).
+forecasting in complex terrain'' (AOAS1811-010).
 
@@
 and precipitation patterns in 2009--2012 differ. This is not visible from Figure~8 directly,
-though, and hence we added a short explanation at the end of Section~3.4
+though, and hence we added a short explanation at the end of Section~3.4.
 
@@
 
-FIXME: Torsten, should we comment on this? In a predictive setup like the precipitation
-forecasting the uncertainty of the estimates is not of so much concern. Also, the 
-uncertainty due to variation in the response is typically much larger.
-
 \end{document}

### 2019-02-18 pkg/disttree/inst/draft/draft.Rnw
final revision and formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2470 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bc69d6b1df1a94a14c4954e27896487bb7cffb6f

@@
 \begin{equation}
-w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
+w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
 \end{equation}
 where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
-for a given $\bold{z}$ is then fully specified by the estimated parameter
-$\bm{\hat \theta}(\bold{z})$ where
+for a given $\bm{z}$ is then fully specified by the estimated parameter
+$\bm{\hat \theta}(\bm{z})$ where
 \begin{equation}
-\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
+\bm{\hat \theta}(\bm{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\bm{\theta}; y_i).
 \end{equation}
@@
 an application to 14~further meteorological stations is provided in
-Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b}.
+Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b}.
 
@@
 (PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) histograms for station
-Axams in Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b} which 
+Axams in Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b} which 
 contains a more detailed explanation of residual QQ plots and PIT histograms.
@@
 cross-validation rather than a single split into learning and validation period}
-\citep[see Supplement~A,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2018a}.
+\citep[see Supplement~A,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019a}.
 \hlc{Further differences are possibly due to East Tyrol lying in a different
@@
 two alternative distributional specifications have been considered
-in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2018a}: Using the
+in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a}: Using the
 same evaluations as in Section~\ref{sec:all}, all models are
@@
 To show that qualitatively similar results are obtained for other
-meteorological stations, Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b}
+meteorological stations, Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b}
 carries out the same evaluation for 14~further stations. These
@@
 \section*{Computational details}
-\label{sec:comp}
 
@@
 QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
-\citep[Supplement~B]{Schlosser+Hothorn+Stauffer:2018b}.
+\citep[Supplement~B]{Schlosser+Hothorn+Stauffer:2019b}.
 Full replication of all results can be obtained with
@@
 also available within the \textsf{R}-Forge~project.
-The results presented in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2018a} and 
-Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
+The results presented in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a} and 
+Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b} can be reproduced using
 \code{demo("RainDistributions", package = "disttree")} and 
@@
   station and similar insights can be obtained for other stations as well,
-  this supplement presents the same analysis as in Section~\ref{sec:axams} of the main
-  manuscript for 14~further meteorological stations.
+  this supplement presents the same analysis as in Section~\ref{sec:axams} 
+  of the main manuscript for 14~further meteorological stations.
 

### 2019-02-18 pkg/disttree/inst/draft/draft_AOAS.tex
final formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2378 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4503d2d73b067eabaa07a9184943d75e94a802fc
- revised rejoinder and corresponding modifications in the manuscript - especially the discussion of more complex distributions should be checked again - also rejoinder on East Tyrol aspects should be improved
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2457 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d3a61fb932ac7bc551babc5d498793cb0b08415b
final revision and formatting
git-svn-id: svn:/

@@
 \usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern}
-%\usepackage{amsthm,amsmath,natbib}
 \usepackage{array,makecell,tikz}
@@
 \usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
 
@@
 \title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
-%\runtitle{Distributional Regression Forests for Probabilistic Precipitation Forecasting}
 \runtitle{Distributional Regression Forests}
@@
 computationally more demanding boosting approach.
+
 \end{abstract}
-
-%parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS
 
@@
 \begin{itemize}
-  \item \emph{Regression trees} \citep{Breiman+Friedman+Stone:1984} recursively
+  \item \emph{Regression trees} (\citealp{Breiman+Friedman+Stone:1984}) recursively
     split the data into more homogeneous subgroups and can thus capture abrupt shifts
@@
     their interactions. 
-  \item \emph{Random forests} \citep{Breiman:2001} average the predictions of
+  \item \emph{Random forests} (\citealp{Breiman:2001}) average the predictions of
     an ensemble of trees fitted to resampled versions of the learning data. This
@@
   considered for learning a forest of 100 trees. Subsampling is employed 
-  for each tree using a third of the predictors
-  in each split of the tree (\texttt{mtry}). Parameters are estimated by
+  for each tree using a third of the predictors in each split of the tree
+  (argument \texttt{mtry} in our implementation \texttt{distforest},
+  with more ``computational details'' provided at the end of the manuscript).
+  Parameters are estimated by
   adaptive local likelihood based on the forest weights as described
@@
 for July 24 in 2009--2012 have already been shown as a motivational example. 
-The figure shows that the model properly depicts the point masses at zero 
-(i.e., the probability of a dry day) and the forecasted probability density 
-function for the total amount of precipitation. The four sample forecasts differ 
-considerably in location $\mu$, scale $\sigma$, and the amount of
-censoring while conforming quite well with the actual observations
-from these days. While this is a nice illustrative example we are  
+In particular, the figure depicts the forecasted point mass at zero
+(i.e., the probability of a dry day) along with the forecasted probability 
+density function for the total amount of precipitation. Based on this illustration 
+it can be observed that the four sample forecasts differ considerably in 
+location $\mu$, scale $\sigma$, and the amount of censoring while conforming quite 
+well with the actual observations from these days. 
+While this is a nice illustrative example we are  
 interested in the overall predictive performance and calibration of the
@@
 (PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) histograms for station
-Axams in Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} which 
+Axams in Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b} which 
 contains a more detailed explanation of residual QQ plots and PIT histograms.
@@
 the prespecified GAMLSS performs particularly well in the validation period
-(2009--2012).
+(2009--2012). Partially, this can be attributed to random variation as
+the differences at several stations are mitigated when considering a full
+cross-validation rather than a single split into learning and validation period
+\citep[see Supplement~A,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019a}.
+Further differences are possibly due to East Tyrol lying in a different
+climate zone, south of the main Alpine Ridge. Hence, long-term climatological
+characteristics as well as the precipitation patterns in 2009--2012 differ
+from North Tyrol, conforming particularly well with the additive effects from
+the prespecified GAMLSS.
 
@@
 two alternative distributional specifications have been considered
-in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a}: Using the
+in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a}: Using the
 same evaluations as in Section~\ref{sec:all}, all models are
@@
 investigates the effects of using the same distributional family with a novel
-strategy for specifying dependence on covariates. Considering more flexible
-parametric specifications as well as transformation model specifications that
-allow to capture dependence of higher moments \citep[see e.g.,][]{Hothorn+Zeileis:2017}
-are extensions of interest for future research.
+strategy for specifying dependence on covariates.
+
+\subsubsection*{More general distributional specifications}
+
+Beyond the task of modeling precipitation it is of interest how well distributional
+forests perform in combination with other more general distributional specifications.
+It has been shown previously in the literature that using a score- or gradient-based
+selection of splitting variables outperforms a mean-based selection with subsequent
+flexible distributional modeling: For example, both \citet[][Figure~2]{Athey+Tibshirani+Wager:2019}
+and \citet[][Figure~1]{Hothorn+Zeileis:2017} demonstrated (independently) that their
+respective score-based random forest algorithms outperform the mean-based quantile
+regression forests of \cite{Meinshausen:2006} in a setup where only the variance of a normal
+response variable changes across the considered covariates. However, if all distribution
+parameters are closely correlated with the distribution mean the forests with different
+splitting strategies all perform similarly, provided a sufficiently flexible distribution
+is employed for the final predictions \citep[see][Section~7]{Hothorn+Zeileis:2017}.
+
+Similarly, the score-based distributional forests introduced in this manuscript
+proved to be quite robust to the different distributional specifications considered.
+While all specifications focus on capturing mean-variance effects note that these
+parameters are never fully orthogonal but can actually become quite closely correlated
+due to the censoring (or truncation and/or zero-inflation considered in Supplement~A).
+
+However, exploring extensions to more flexible parametric distributions
+\citep[e.g., such as the Dagum distribution considered by][in GAMLSS-type models]{Klein+Kneib+Lang:2015}
+as well as transformation model specifications \citep[e.g., as in][]{Hothorn+Zeileis:2017}
+are of interest for future research.
 
@@
 To show that qualitatively similar results are obtained for other
-meteorological stations, Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b}
+meteorological stations, Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b}
 carries out the same evaluation for 14~further stations. These
@@
 
-%\pagebreak
 \subsubsection*{Computational difficulties}
@@
 on \textsf{R}-Forge at
-(\url{https://R-Forge.R-project.org/projects/partykit/}). The function
-\texttt{distfit} fits distributional models by maximum likelihood, which is
-used as the basis for the tree-building function \texttt{disttree}, upon which
-the \texttt{distforest} is built. All functions can either be used with
-GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
+(\url{https://R-Forge.R-project.org/projects/partykit/}). 
+The function \texttt{distforest} learns the distributional forests proposed
+in this manuscript by combining the general \texttt{cforest} function from
+\textbf{partykit} with the function \texttt{distfit} for fitting distributional
+models by maximum likelihood. Analogously, \texttt{disttree} can learn a single
+distributional tree by combining \texttt{ctree} with \texttt{distfit}.
+All functions can either be used with GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
 (\citealp{Stasinopoulos+Rigby:2007}, version~5.0.6)
@@
 QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
-\citep[Supplement B]{Schlosser+Hothorn+Stauffer:2018b}.
+\citep[Supplement~B]{Schlosser+Hothorn+Stauffer:2019b}.
 Full replication of all results can be obtained with
@@
 also available within the \textsf{R}-Forge~project.
-The results presented in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a} and 
-Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
+The results presented in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a} and 
+Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b} can be reproduced using
 \texttt{demo("RainDistributions", package = "disttree")} and 
 \texttt{demo("RainStationwise", package = "disttree")}, respectively.
+
+\section*{Acknowledgements}
+An extended research stay of Torsten Hothorn in Innsbruck (August
+2017 to January 2018) was financially supported by the Swiss National
+Science Foundation, grant number SNF IZSEZ0 177091.
 
@@
 $\bm{\hat{\theta}} = 
-(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$, 
+(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
 can be plugged in the score function $s(\bm{\theta}, Y)$.
@@
 the maximum of the absolute values of the standardized linear statistic
-can be considered.
+can be considered ($c_{\text{max}}$).
+
+\cite{Strasser+Weber:1999} showed that the asymptotic conditional 
+distribution of the linear statistic $t_l$ is a multivariate normal with parameters
+$\mu$ and $\Sigma$. Hence, the asymptotic conditional distribution of
+$c(t_{l},\mu_{l},\Sigma_{l})$ 
+is either normal
+(for $c_{\text{max}}$) 
+or 
+$\chi^2$ (for $c_{\text{quad}}$).
 
@@
   \stitle{Different Response Distributions}
-  \slink[doi]{10.1214/00-AOASXXXXSUPPA}
+  %\slink[doi]{...}
+  \slink[doi]{(to be completed by the typesetter)}
+  %\slink[doi]{10.1214/00-AOASXXXXSUPP}
   \sdatatype{.pdf} 
@@
   \stitle{Stationwise Evaluation}
-  \slink[doi]{10.1214/00-AOASXXXXSUPPB}
+  %\slink[doi]{...}
+  \slink[doi]{(to be completed by the typesetter)}
+  %\slink[doi]{10.1214/00-AOASXXXXSUPP}
   \sdatatype{.pdf} 

### 2019-02-19 pkg/disttree/inst/draft/draft.Rnw
final revision and formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2471 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 30a154308024c44cf21f9f0809cf599e54fbbba1

@@
 cross-validation rather than a single split into learning and validation period}
-\citep[see Supplement~A,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019a}.
+\citep[see Supplement~B,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019b}.
 \hlc{Further differences are possibly due to East Tyrol lying in a different
@@
 \citep[e.g.,][]{Stauffer+Mayr+Messner:2017}. 
-To test for robustness against distributional misspecifation, 
+To test for robustness against distributional misspecification, 
 two alternative distributional specifications have been considered
@@
 parameters are never fully orthogonal but can actually become quite closely correlated
-due to the censoring (or truncation and/or zero-inflation considered in Supplement~A).
-
-However, exploring extensions to more flexible parametric distributions}
+due to the censoring (or truncation and/or zero-inflation considered in} 
+Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a}).
+
+\hlc{However, exploring extensions to more flexible parametric distributions}
 \citep[e.g., such as the Dagum distribution considered by][in GAMLSS-type models]{Klein+Kneib+Lang:2015}

### 2019-02-19 pkg/disttree/inst/draft/draft_AOAS.tex
final revision and formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2471 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 30a154308024c44cf21f9f0809cf599e54fbbba1

@@
 cross-validation rather than a single split into learning and validation period
-\citep[see Supplement~A,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019a}.
+\citep[see Supplement~B,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019b}.
 Further differences are possibly due to East Tyrol lying in a different
@@
 parameters are never fully orthogonal but can actually become quite closely correlated
-due to the censoring (or truncation and/or zero-inflation considered in Supplement~A).
+due to the censoring (or truncation and/or zero-inflation considered in 
+Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a}).
 
@@
 
-\section*{Acknowledgements}
-An extended research stay of Torsten Hothorn in Innsbruck (August
-2017 to January 2018) was financially supported by the Swiss National
-Science Foundation, grant number SNF IZSEZ0 177091.
-
 \pagebreak
@@
 the maximum of the absolute values of the standardized linear statistic
-can be considered ($c_{\text{max}}$).
+can be considered~($c_{\text{max}}$).
 
@@
 \pagebreak
+
+\section*{Acknowledgments}
+An extended research stay of Torsten Hothorn in Innsbruck (August
+2017 to January 2018) was financially supported by the Swiss National
+Science Foundation, grant number SNF IZSEZ0 177091.
+
 % AOS,AOAS: If there are supplements please fill:

### 2019-02-19 pkg/disttree/inst/draft/rejoinder2.Rnw
final revision and formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2471 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 30a154308024c44cf21f9f0809cf599e54fbbba1

@@
 Moreover, sampling variability clearly also plays a role as the differences from Figure~8
-are somewhat evened out in the cross-validation, specifically for some of the East Tyrolian 
+are somewhat evened out in the cross-validation, specifically for some of the East Tyrolean 
 stations. This is shown in Supplement~B and was already discussed in the corresponding
@@
 
+In a predictive setup like the precipitation forecasting the uncertainty of the estimates is not 
+of so much concern. Also, the uncertainty due to variation in the response is typically much larger.
+
 \end{document}

### 2019-02-20 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2473 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bcdd074e435d195cd5e1ea3b6c88485e988b70ba

@@
 been evaluated on data sets based on the tree structure with two different split variables 
-($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
+($Z_1$ and $Z_2$) both with the same split point $\xi$ as illustrated in 
 Figure~\ref{fig:dgp_tree}.
@@
 almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE\_scores performs
-a bit worse but it still achieves an ARI above $0.5$ for the highest considered effect size
+slightly worse but it still achieves an ARI above $0.5$ for the highest considered effect size
 of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
@@
 split variables both do not cause any improvement but in many situations even have a negative
-effect on the power of a testing strategies.
+effect on the power of a testing strategy.
 
@@
 applying postpruning (building very large trees without any significance level and employing 
-cost-complexity pruning as in the CART algorithm). 
+cost-complexity pruning as in the CART algorithm \citep{Breiman+Friedman+Stone:1984}). 
 In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with prepruning,
@@
 as the introduced building blocks can be applied in the same way for all model types.
-One specific feature of linear models is the orthogonality of the conditional expectation
-$\mathbb{E}(Y|X)$ and the regressor $X$ which affects the capability of tests using only residuals
-to capture shifts in the parameters. In particular, \cite{Ploberger+Kramer:1992} have shown that
-residual-based tests are insensitive to parameter changes orthogonal to the mean regressor.
+Specific characteristics of linear models regarding structural changes have been investigated
+in detail in previous literature.
+\cite{Ploberger+Kramer:1992} have shown that residual-based tests can detect a change in the parameters
+only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
+This is not the case if changes are orthogonal to the mean regressor $(1,0)^{\top}$, hence, if
+the parameter vector $\bm{\beta}$ is shifted 
+%to $\bm{\beta} + \bm{\Delta}$ 
+by $\bm{\Delta} = (0, \Delta_1)^{\top}$.
 For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
 $\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
-Testing procedures purely based on residuals perform best in situations with shifts only in the
-intercept, however, the costs in terms of power that come with including all scores are 
-negligibly small. 
-Furthermore, with only two parameters determining a linear model the importance of including scores 
-regarding each parameter could already be pointed out clearly while for an increasing number of 
-parameters in a model this aspect becomes even more important.
-
-In general residual-based and score-based procedures are expected to perform equally well if all
-parameters are highly correlated. On the other hand, if parameters are orthogonal changes that are 
-not connected to the mean might be missed when only considering residuals. This can be avoided by
-including scores which are a standard measurement of goodness of fit that can be easily employed.
+While this might be an extreme example similar scenarios can appear in many applications, see, e.g., 
+Figure~2 in \cite{Loh+He+Man:2015} where an additional separation of the data in each node is
+required in order to fit a reasonable model. This can be avoided by including model scores
+allowing for the detection of changes not only in the location parameter but also in all other 
+parameters, i.e., not only in the expected value but also in the variance.
+In situations where shifts appear only in the intercept testing procedures purely based on residuals 
+perform best. However, the cost in terms of power that comes with including all scores, in particular
+with testing more than one parameter, is negligibly small.
+
+Also in more general models residual-based and score-based procedures are expected to perform equally 
+well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
+that are not connected to the mean might be missed when only considering residuals. This can be prevented 
+by including scores which are a standard measurement of goodness of fit that can be easily employed for 
+all model types.
 Even though it also depends on the type of the applied residuals 
@@
 shift and the type of model. Therefore, a general statement for all types of models and residuals can 
-hardly be made.
+hardly be made. However, the results from previous literature such as \cite{Ploberger+Kramer:1992} and
+the here presented simulation show that considering only residuals provides small gains in certain 
+situations but remarkable deficits in general. On the contrary, the benefits of including full model 
+scores clearly outweigh the price of of testing all parameters.
+
+%Furthermore, with only two parameters determining a linear model the importance of including scores 
+%regarding each parameter could already be pointed out clearly while for an increasing number of 
+%parameters in a model this aspect becomes even more important.
 

### 2019-02-25 pkg/disttree/inst/draft/draft.Rnw
removed highlighting of changes and unneeded comments; final formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2474 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0be09dda9fd42d8a35ac2a25b130c9625d684d82

@@
   for each tree using a third of the predictors in each split of the tree
-  (\hlc{argument} \code{mtry} \hlc{in our implementation} \code{distforest},
-  \hlc{with more ``computational details'' provided at the end of the manuscript}).
+  (argument \code{mtry} in our implementation \code{distforest},
+  with more ``computational details'' provided at the end of the manuscript).
   Parameters are estimated by
@@
   The stopping criteria are the minimal number of observations to perform a split 
-  (\texttt{minsplit} $= 50$), the minimal number of observations in a 
-  segment (\texttt{minbucket} $= 20$), and the significance level 
-  for variable selection (\texttt{alpha} $= 1$). The latter means
+  (\texttt{minsplit}~$= 50$), the minimal number of observations in a 
+  segment (\texttt{minbucket}~$= 20$), and the significance level 
+  for variable selection (\texttt{alpha}~$= 1$). The latter means
   that no early stopping (or ``pre-pruning'') is applied
@@
 for July 24 in 2009--2012 have already been shown as a motivational example. 
-\hlc{In particular, the figure depicts the forecasted point mass at zero
+In particular, the figure depicts the forecasted point mass at zero
 (i.e., the probability of a dry day) along with the forecasted probability 
 density function for the total amount of precipitation. Based on this illustration 
-it can be observed that} the four sample forecasts differ considerably in 
+it can be observed that the four sample forecasts differ considerably in 
 location $\mu$, scale $\sigma$, and the amount of censoring while conforming quite 
@@
 the prespecified GAMLSS performs particularly well in the validation period
-(2009--2012). \hlc{Partially, this can be attributed to random variation as
+(2009--2012). Partially, this can be attributed to random variation as
 the differences at several stations are mitigated when considering a full
-cross-validation rather than a single split into learning and validation period}
+cross-validation rather than a single split into learning and validation period
 \citep[see Supplement~B,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019b}.
-\hlc{Further differences are possibly due to East Tyrol lying in a different
+Further differences are possibly due to East Tyrol lying in a different
 climate zone, south of the main Alpine Ridge. Hence, long-term climatological
@@
 from North Tyrol, conforming particularly well with the additive effects from
-the prespecified GAMLSS.}
+the prespecified GAMLSS.
 
@@
 
-\hlc{Beyond the task of modeling precipitation it is of interest how well distributional
+Beyond the task of modeling precipitation it is of interest how well distributional
 forests perform in combination with other more general distributional specifications.
@@
 selection of splitting variables outperforms a mean-based selection with subsequent
-flexible distributional modeling: For example, both} \citet[][Figure~2]{Athey+Tibshirani+Wager:2019}
-\hlc{and} \citet[][Figure~1]{Hothorn+Zeileis:2017} \hlc{demonstrated (independently) that their
+flexible distributional modeling: For example, both \citet[][Figure~2]{Athey+Tibshirani+Wager:2019}
+and \citet[][Figure~1]{Hothorn+Zeileis:2017} demonstrated (independently) that their
 respective score-based random forest algorithms outperform the mean-based quantile
-regression forests of} \cite{Meinshausen:2006} \hlc{in a setup where only the variance of a normal
+regression forests of \cite{Meinshausen:2006} in a setup where only the variance of a normal
 response variable changes across the considered covariates. However, if all distribution
@@
 splitting strategies all perform similarly, provided a sufficiently flexible distribution
-is employed for the final predictions} \citep[see][Section~7]{Hothorn+Zeileis:2017}.
-
-\hlc{Similarly, the score-based distributional forests introduced in this manuscript
+is employed for the final predictions \citep[see][Section~7]{Hothorn+Zeileis:2017}.
+
+Similarly, the score-based distributional forests introduced in this manuscript
 proved to be quite robust to the different distributional specifications considered.
@@
 parameters are never fully orthogonal but can actually become quite closely correlated
-due to the censoring (or truncation and/or zero-inflation considered in} 
+due to the censoring (or truncation and/or zero-inflation considered in
 Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a}).
 
-\hlc{However, exploring extensions to more flexible parametric distributions}
+However, exploring extensions to more flexible parametric distributions
 \citep[e.g., such as the Dagum distribution considered by][in GAMLSS-type models]{Klein+Kneib+Lang:2015}
-\hlc{as well as transformation model specifications} \citep[e.g., as in][]{Hothorn+Zeileis:2017}
-\hlc{are of interest for future research.}
+as well as transformation model specifications \citep[e.g., as in][]{Hothorn+Zeileis:2017}
+are of interest for future research.
 
@@
 build full-grown trees, early stopping upon non-significance is disabled
-(\texttt{alpha} $= 1$) and low values are used for \texttt{minsplit} ($ = 50$)
-and \texttt{minbucket} ($ = 20$), while assuring that \texttt{minsplit}
+(\texttt{alpha}~$= 1$) and low values are used for \texttt{minsplit}~($ = 50$)
+and \texttt{minbucket}~($ = 20$), while assuring that \texttt{minsplit}
 is sufficiently large for reasonably obtaining MLEs of all parameters in each
@@
 
-\section*{Acknowledgements}
-\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
-2017 to January 2018) was financially supported by the Swiss National
-Science Foundation, grant number SNF IZSEZ0 177091.}
-
-\newpage
+%\newpage
 \section*{Appendix}
@@
 
-\cite{Strasser+Weber:1999} \hlc{showed that the asymptotic conditional 
-distribution of the linear statistic} $t_l$ \hlc{is a multivariate normal with parameters}
-$\mu$ and $\Sigma$. \hlc{Hence, the asymptotic conditional distribution of}
+\cite{Strasser+Weber:1999} showed that the asymptotic conditional 
+distribution of the linear statistic $t_l$ is a multivariate normal with parameters
+$\mu$ and $\Sigma$. Hence, the asymptotic conditional distribution of
 $c(t_{l},\mu_{l},\Sigma_{l})$ 
-\hlc{is either normal}
+is either normal
 (for $c_{\text{max}}$) 
-\hlc{or} 
+or 
 $\chi^2$ (for $c_{\text{quad}}$).
@@
 
+\section*{Acknowledgments}
+An extended research stay of Torsten Hothorn in Innsbruck (August
+2017 to January 2018) was financially supported by the Swiss National
+Science Foundation, grant number SNF IZSEZ0 177091.
+
 \section*{Supplementary material}
@@
 
-\newpage
+%\newpage
 \bibliography{ref.bib}

### 2019-02-27 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
first draft of slides for DAGStat2019
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2475 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 992aeda67b4440f5b1c1d25e589623e15df2451b

\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
\usetheme[nototalframenumber, noslidenumber]{uibk}

\title{The Power of Unbiased Recursive Partitioning:\\
A Unifying View of CTree, MOB, and GUIDE}
\author{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
%\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
\URL{https://R-Forge.R-project.org/projects/partykit/} 
%\\
%https://partykit.org}
\headerimage{3}

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
\usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
%\graphicspath{{plots/}}

\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
set.seed(7)
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/DAGStat2019/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
#source("../guidelike.R")
source("../tests_guideprune.R")
#source("../ccprune.R")

load("../draft/stump.rda")
sim_stump <- simres$res
rm(simres)

load("../draft/tree.rda")
sim_tree <- simres$res
rm(simres)

load("../draft/pruning.rda")
pr <- simres$res
## extract data for xyplot comparing ARI and ARI_p
pr <- pr[,c(1,2,7,8,9,12,15)]
## extended data set with postpruning results being represented as separated tests
prex <- pr[,c(1,2,3,4,6)]
prex <- rbind(prex,prex)

prex$test <- factor(prex$test,
                    levels = c("ctree","mfluc",
                               "guide_sum_12", "guide_sum_1_cor",
                               "ctree_p","mfluc_p",
                               "guide_sum_12_p", "guide_sum_1_cor_p"),
                    labels = c("CTree","MOB",
                               "GUIDE_scores", "GUIDE",
                               "CTree_p","MOB_p",
                               "GUIDE_scores_p", "GUIDE_p"))

for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
  if(prex$test[i] == "GUIDE_scores") prex$test[i] <- "GUIDE_scores_p"
  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
}

prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p

load("../draft/3way_xi00_delta03.rda")
load("../draft/3way_xi08_delta1.rda")
d00$xi <- 0
d08$xi <- 0.8
d3way <- rbind(d00, d08)
d3way$xi <- factor(d3way$xi, levels = c(0,0.8), labels = c(0,0.8))

## HCL palette
pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")

## HCL palette (light)
pallight <- hcl(c(140, 80, 260, 70, 10), 300, 60, alpha = 0.25)
names(pallight) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")

#produce plots
set.seed(7)
x <- runif(50,-1,1)
e1 <- rnorm(50,0,1)
e2 <- rnorm(50,0,1)
y1 <- x*6 + e1 
y2 <- x*1 + e2
y <- c(y1,y2)

jpeg("lm1.jpeg")
plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
points(x, y2)
abline(lm(y~c(x,x)), col = "red", lwd = 2)
dev.off()

jpeg("lm2.jpeg")
plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
abline(lm(y1~x), col = "red", lwd = 2)
dev.off()

jpeg("lm3.jpeg")
plot(x,y2, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
abline(lm(y2~x), col = "red", lwd = 2)
dev.off()

@

\begin{document}
\SweaveOpts{concordance=TRUE}

\section{Unbiased Recursive Partitioning}

\begin{frame}%[fragile]
\frametitle{Motivation}
\center
\begin{tikzpicture}
\visible<1>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (n0) at (4, 5) {$Y,X,Z$};
}
\visible<2>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (n0) at (4, 5) {$\mathcal{M}(Y,X;\hat{\beta}$)};
}
\visible<3->{
\node[inner sep=0pt] (lm1) at (4, 4.5)
    {\includegraphics[width=0.25\textwidth]{lm1.jpeg}};
}
\visible<4>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {?};
\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6);
\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6);
}
\visible<5-6>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {$Y_1,X_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {$Y_2,X_2$};
\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6) node [midway, left] {\scriptsize $Z \leq \xi$};
\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6) node [midway, right] {\scriptsize $Z > \xi$};
}
\visible<6>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {$\mathcal{M}(Y_1,X_1;\hat{\beta_1}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {$\mathcal{M}(Y_2,X_2;\hat{\beta_2}$)};
}
\visible<7->{
\node[inner sep=0pt] (lm2) at (2,0.5)
    {\includegraphics[width=0.25\textwidth]{lm2.jpeg}};
\node[inner sep=0pt] (lm3) at (6,0.5)
    {\includegraphics[width=0.25\textwidth]{lm3.jpeg}};
\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6) node [midway, left] {\scriptsize $Z \leq \xi$};
\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6) node [midway, right] {\scriptsize $Z > \xi$};
}
\end{tikzpicture}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Unbiased Recursive Partitioning}
{\bf Basic tree algorithm:}
\medskip
\begin{enumerate}
\item Fit a model to the response $Y$ (and possible covariates $X$).
\item Assess association of $Y$ (or a model-based transformation)
and each possible split variable~$Z_j$ and select the split~variable~$Z_{j^\ast}$ showing the strongest association.
\item Choose the corresponding split point leading to the highest improvement of model fit and split the data.
\item Repeat steps 1--3 recursively in each of the resulting subgroups until some stopping criterion is met.
\end{enumerate}

\vspace{1.2cm}

{\bf Here:} Focus on split variable selection (step 2).

\end{frame}

\begin{frame}%[fragile]
\frametitle{Split Variable Selection}

{\bf General strategy:}
\begin{enumerate}
\item Evaluate a discrepancy measure giving information on the goodness of fit of the model for each observation and each model parameter (transformation of $Y$).
\item Apply a statistical test assessing dependency of the discrepancy measure on each possible split variable $Z_j$.
\item Select the split variable $Z_j^\ast$ showing the smallest $p$-value.
\end{enumerate}

\bigskip

{\bf Possible discrepancy measures:}
\begin{itemize}
\item Absolute deviations from an average value
\item Residuals
\item Model scores
\item \ldots
\end{itemize}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Discrepancy Measures}
As an example linear models $\mathcal{LM}(Y,X;\beta_0, \beta_1)$
fitted via ordinary least squares (OLS)
are considered with the following discrepancy measures:

\bigskip

{\bf Residuals:}
\begin{flalign*}
\hspace{0.7cm}
r(Y,X,\hat{\beta_0},\hat{\beta_1}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X &&
\end{flalign*}

\bigskip

{\bf Model scores:}
\begin{flalign*}
\hspace{0.7cm}
s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
\left(\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_0} \quad , \quad  
\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_1}\right) 
&&
\end{flalign*}

\pause

$${\color{red}\Downarrow}$$
$$
{\color{red} 2 \cdot r(Y,X,\hat{\beta_0},\hat{\beta_1})}
$$

\end{frame}

\begin{frame}%[fragile]
\frametitle{General Framework}

{\bf{Building blocks:}}

\begin{itemize}
\item {\only<4->{\color{lightgray}} Residuals vs.\ full model scores}
\item {\only<1-3,7->{\color{lightgray}} Binarization of residuals/scores}
\item {\only<1-6>{\color{lightgray}} Categorization of possible split variables}
\end{itemize}

\bigskip

\only<2>{
\small
$$
s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
\begin{pmatrix} 
\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
\vdots & {\color{lightgray} \vdots} \\
\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
\end{pmatrix}
$$
}
\only<3>{
\small
$$
s(Y, X, \beta_0, \beta_1) = 
\begin{pmatrix} 
\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
\vdots & \vdots \\
\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
\end{pmatrix}
$$
}
\only<5-6>{
\small
$$
r(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
\begin{pmatrix}
r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
\vdots \\
r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
\end{pmatrix}
%
\visible<6>{
\quad \Rightarrow \quad
\small
\begin{pmatrix}
1\\
0\\
\vdots \\
1
\end{pmatrix}
}
$$
}
\only<7-8>{
\small
$$
Z_j = 
\begin{pmatrix}
Z_{j1}\\
Z_{j2}\\
\vdots \\
Z_{jn}\\
\end{pmatrix}
%
\visible<8>{
\quad \Rightarrow \quad
\small
\begin{pmatrix}
3\\
1\\
\vdots \\
2
\end{pmatrix}
}
$$
}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Pruning}

{\bf Goal:} Avoid overfitting.

\bigskip

{\bf Two Strategies:}

\begin{itemize}
\item Pre-pruning: \\
Internal stopping criterion based on Bonferroni-corrected $p$-values of the applied tests. Stop splitting when there is no significant association. 
\item Post-pruning: \\
First grow a very large tree and afterwards prune splits that do not improve the model fit, 
either via cross-validation (e.g.,~cost-complexity pruning as in CART) or 
based on information criteria (e.g.,~AIC or BIC). 
\end{itemize}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Simulation Setting}

\begin{table}
\begin{center}
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\end{center}
\end{table}

\end{frame}

\begin{frame}%[fragile]
\frametitle{Simulation Setting (Scenario 1)}
True tree structure:

<<dgp, echo=FALSE>>=
set.seed(7)

nobs <- 250
sigma <- 0.5
xi <- 0
delta <- 1
beta0 <- 0
beta1 <- 1

d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)

sp_xi1 <- partysplit(3L, breaks = 0)

pn <- partynode(1L, split = sp_xi1, 
                kids = list(
                  partynode(2L, info = c("true parameters:", 
                                         expression(beta[0] ==~" "~0~" or "~-delta), 
                                         expression(beta[1] ==~" "~1~" or "~+delta))),
                  partynode(3L, info = c("true parameters:", 
                                         expression(beta[0] ==~" "~0~" or "~+delta), 
                                         expression(beta[1] ==~" "~1~" or "~-delta)))))

p <- party(pn, d)

paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))

ep <- function (obj, digits = 3, abbreviate = FALSE, justmin = Inf, 
                just = c("alternate", "increasing", "decreasing", "equal"), 
                fill = "white") 
{
  meta <- obj$data
  justfun <- function(i, split) {
    myjust <- if (mean(nchar(split)) > justmin) {
      match.arg(just, c("alternate", "increasing", "decreasing", "equal"))
    } else {
      "equal"
    }
    k <- length(split)
    rval <- switch(myjust, equal = rep.int(0, k), 
                   alternate = rep(c(0.5, -0.5), length.out = k), 
                   increasing = seq(from = -k/2, to = k/2, by = 1), 
                   decreasing = seq(from = k/2, to = -k/2, by = -1))
    unit(0.5, "npc") + unit(rval[i], "lines")
  }
  function(node, i) {
    split <-  c("<= xi", "> xi")
    y <- justfun(i, split)
    split <- split[i]
    if (any(grep(">", split) > 0) | any(grep("<", split) > 0)) {
      tr <- suppressWarnings(try(parse(text = paste("phantom(0)", split)), silent = TRUE))
      if (!inherits(tr, "try-error")) split <- tr
    }
    grid.rect(y = y, gp = gpar(fill = fill, col = 0), width = unit(1, "strwidth", split))
    grid.text(split, y = y, just = "center")
  }
}
class(ep) <- "grapcon_generator"

tp <- function (obj, digits = 3, abbreviate = FALSE, 
                fill = c("lightgray", "white"), id = TRUE, 
                just = c("center", "top"), top = 0.85, 
                align = c("center", "left", "right"), gp = NULL, FUN = NULL, 
                height = NULL, width = NULL) 
{
  nam <- names(obj)
  extract_label <- function(node) formatinfo_node(node, FUN = FUN, 
                                                  default = c("terminal", "node"))
  maxstr <- function(node) {
    lab <- extract_label(node)
    klab <- if (is.terminal(node)) 
      ""
    else unlist(lapply(kids_node(node), maxstr))
    lab <- c(lab, klab)
    lab <- try(unlist(lapply(lab, function(x) strsplit(x, 
                                                       "\n"))), silent = TRUE)
    if (inherits(lab, "try-error")) {
      paste(rep("a", 9L), collapse = "")
    }
    else {
      return(lab[which.max(nchar(lab))])
    }
  }
  nstr <- if (is.null(width)) 
    maxstr(node_party(obj))
  else paste(rep("a", width), collapse = "")
  just <- match.arg(just[1L], c("center", "centre", "top"))
  if (just == "centre") 
    just <- "center"
  align <- match.arg(align[1L], c("center", "centre", "left", 
                                  "right"))
  if (align == "centre") 
    align <- "center"
  rval <- function(node) {
    if(node$id == 2) fill <- c("white", "blue")
    if(node$id == 3 | node$id == 4) fill <- c("white", "red")
    if(node$id == 5) fill <- c("white", "darkgreen")
    fill <- rep(fill, length.out = 2)
    lab <- extract_label(node)
    if (!is.null(gp)) {
      outer_vp <- viewport(gp = gp)
      pushViewport(outer_vp)
    }
    if (is.null(height)) 
      height <- length(lab) + 1L
    node_vp <- viewport(x = unit(0.5, "npc"), y = unit(if (just == 
                                                           "top") 
      top
      else 0.5, "npc"), just = c("center", just), width = unit(1, 
                                                               "strwidth", nstr) * 1.1, height = unit(height, "lines"), 
      name = paste("node_terminal", id_node(node), sep = ""), 
      gp = if (is.null(gp)) 
        gpar()
      else gp)
    pushViewport(node_vp)
    grid.rect(gp = gpar(fill = fill[1]))
    for (i in seq_along(lab)) grid.text(x = switch(align, 
                                                   center = unit(0.5, "npc"), left = unit(1, "strwidth", 
                                                                                          "a"), right = unit(1, "npc") - unit(1, "strwidth", 
                                                                                                                              "a")), y = unit(length(lab) - i + 1, "lines"), 
                                        lab[i], just = align)
    if (id) {
      nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
                                                          "npc"), width = max(unit(1, "lines"), unit(1.3, 
                                                                                                     "strwidth", nam[id_node(node)])), height = max(unit(1, 
                                                                                                                                                         "lines"), unit(1.3, "strheight", nam[id_node(node)])))
      pushViewport(nodeIDvp)
      grid.rect(gp = gpar(fill = fill[2], lty = "solid"))
      grid.text(nam[id_node(node)])
      popViewport()
    }
    if (is.null(gp)) 
      upViewport()
    else upViewport(2)
  }
  return(rval)
}

class(tp) <- "grapcon_generator"
@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(-1,1), col = "blue", lwd = 2)
abline(coef = c(1,1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = 3.2)
mtext(expression(beta[1]== 1), side = 4, col = "red", las = 1, line = 0.5, at = 2.2)
mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.2)
mtext(expression(beta[1]== 1), side = 4, col = "blue", las = 1, line = 0.5, at = -0.8)
@
\endminipage

\minipage{0.45\textwidth}
<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(0,1), col = "blue", lwd = 2)
abline(coef = c(0,-1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = "red", las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(-1,1), col = "blue", lwd = 2)
abline(coef = c(1,-1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -1.7)
@
\endminipage
\end{figure}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Results}
Residuals vs.\ full model scores:

\begin{figure}
\setkeys{Gin}{width=\linewidth}
<<scores_stump, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                         "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree","mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE_scores", "GUIDE"))

subdataxi <- subset(subdata, xi %in% c(0,0.8))

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))

library("latticeExtra")
useOuterStrips(xyplot(prop_Tsplit ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1),
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Proportion of selecting"~Z[1]),
                      par.settings = par.settings),

               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
                                                     expression("vary"~beta[0]), 
                                                     expression("vary"~beta[1]))),
               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8))))

@
\end{figure}

\end{frame}

\begin{frame}%[fragile]
\frametitle{Results}
Full factorial analysis (with both parameters varying):
\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = d3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c("red", "blue", "red", "blue"),
       lty = c(1,1,2,2),
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       #index.cond = list(c(1,2,3,4)),
       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
                                              expression(xi==0.8~"and"~delta==1))),
       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
                  lines = list(lty = c(1,1,2,2), 
                               type = "l",
                               col = c("red", "blue", "red", "blue"))))
@
\end{figure}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Simulation Setting (Scenario 2)}
True tree structure:

<<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=

# d <- dgp_tree(nobs = 250)
set.seed(7)
d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.5)

sp_xi2 <- partysplit(4L, breaks = 0)
sp_xi1 <- partysplit(3L, breaks = 0)

pn <- partynode(1L, split = sp_xi2, 
                kids = list(
                  partynode(2L, info = c("true parameters:", 
                                         expression(beta[0] == 0), 
                                         expression(beta[1] == +delta))),
                  partynode(3L, split = sp_xi1, 
                            kids = list(
                              partynode(4L, info = c("true parameters:", 
                                                     expression(beta[0] == -delta), 
                                                     expression(beta[1] == -delta))),
                              partynode(5L, info = c("true parameters:", 
                                                     expression(beta[0] == +delta), 
                                                     expression(beta[1] == -delta)))))))

p <- party(pn, d)

#ctest <- evaltests(y~x|z1+z2+z3+z4+z5+z6+z7+z8+z9+z10, data = d, testfun = "ctree", stump = FALSE)
@

\begin{figure}%[t]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
#paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(0,1), col = "blue", lwd = 2)
abline(coef = c(-1,-1), col = "red", lwd = 2)
abline(coef = c(1,-1), col = "darkgreen", lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha("red", 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha("darkgreen", 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c("blue", "red", "darkgreen"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2.7)
@
\endminipage
\end{center}
\end{figure}

\end{frame}

\begin{frame}%[fragile]
\frametitle{Results}

Residuals vs.\ full model scores:

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
                                        "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE_scores", "GUIDE"))

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))

xyplot(ari ~ delta | xi, groups = ~ test, 
       data = subdata, 
       type = "l",
       lty = c(1,1,1,1),
       key =  list(text = list(levels(subdata$test)),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8))),
       par.settings = par.settings)

@

\end{figure}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Results}
Pre-pruning vs.\ post-pruning:

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))

xyplot(ari ~ delta | xi, groups = ~ test, 
       data = prex, 
       type = "l", lty = c(1,1,1,1,2,2,2,2),
       key =  list(text = list(levels(prex$test)[1:4]),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8))),
       par.settings = par.settings
)
@
\end{figure}

\end{frame}

\begin{frame}%[fragile]
\frametitle{Recommendations}
Based on the simulation results the following general conclusions can be drawn:
\begin{itemize}
\item Including full model scores is advantageous over using only residuals.
\item Binarizing scores/residuals does not lead to any improvements in performance.
\item Categorizing split variables can facilitate calculations, however, important
effects might be missed unless a binning point is close to the true split point.
\item If the applied statistical test performs well as significance test pre-pruning
is preferable. Otherwise post-pruning can compensate this deficiency at least partly.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Software}
\vspace{0.4cm}
\textbf{Package:} \emph{partykit} available on CRAN\\

\bigskip

\textbf{Reproduction code:} available on R-Forge at\\

\url{https://R-Forge.R-project.org/projects/partykit/}\\

\bigskip

\textbf{Applied functions:}

\begin{itemize}
\item \texttt{ctree}
\item \texttt{mob}
\item Adaptions of both of them
\item Reimplementation of GUIDE
\end{itemize}
\end{frame}

\subsection{References}

\begin{frame}
\frametitle{References}

\vspace{-0.2cm}

\footnotesize
Schlosser L, Hothorn T, Zeileis A (2019).
 \dquote{The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE.}
 \emph{arXiv ...}, arXiv.org E-Print Archive.
 \doi{...}

\medskip

Breiman L, Friedman J H, Olshen R A, Stone C J (1984).
 \textit{Classification and Regression Trees.}
 Wadsworth,
 California.

\medskip

Hothorn T, Hornik K, Zeileis A (2006).
 \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
 \textit{Journal of Computational and Graphical Statistics},
 \textbf{15}(3), 651--674.
 \doi{10.1198/106186006X133933}

\medskip

Hothorn T, Zeileis A (2015).
 \dquote{\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
 \textit{Journal of Machine Learning Research},
 \textbf{16}, 3905--3909.
 \url{http://www.jmlr.org/papers/v16/hothorn15a.html}

\medskip

Loh W-Y (2002).
 \dquote{Regression Trees with Unbiased Variable Selection and Interaction Detection.}
 \textit{Statistica Sinica},
 \textbf{12}(2), 361--386.
 \url{http://www.jstor.org/stable/24306967}

\medskip

Zeileis A, Hothorn T, Hornik K (2008).
 \dquote{Model-Based Recursive Partitioning.}
  \textit{Journal of Computational and Graphical Statistics},
  \textbf{17}(2), 492--514.
  \doi{10.1198/106186008X319331}

\end{frame}

%\begin{frame}
%\frametitle{References (2)}

%\footnotesize

%\end{frame}

\end{document}

### 2019-02-27 pkg/partykit/inst/guideliketest/draft/draft.Rnw
edited references and illustrations of data generating process in the simulation study
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2476 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 477f23216c3441f0932a463db8d563d8f054e703

@@
 class(ep) <- "grapcon_generator"
+
+tp <- function (obj, digits = 3, abbreviate = FALSE, 
+                fill = c("lightgray", "white"), id = TRUE, 
+                just = c("center", "top"), top = 0.85, 
+                align = c("center", "left", "right"), gp = NULL, FUN = NULL, 
+                height = NULL, width = NULL) 
+{
+  nam <- names(obj)
+  extract_label <- function(node) formatinfo_node(node, FUN = FUN, 
+                                                  default = c("terminal", "node"))
+  maxstr <- function(node) {
+    lab <- extract_label(node)
+    klab <- if (is.terminal(node)) 
+      ""
+    else unlist(lapply(kids_node(node), maxstr))
+    lab <- c(lab, klab)
+    lab <- try(unlist(lapply(lab, function(x) strsplit(x, 
+                                                       "\n"))), silent = TRUE)
+    if (inherits(lab, "try-error")) {
+      paste(rep("a", 9L), collapse = "")
+    }
+    else {
+      return(lab[which.max(nchar(lab))])
+    }
+  }
+  nstr <- if (is.null(width)) 
+    maxstr(node_party(obj))
+  else paste(rep("a", width), collapse = "")
+  just <- match.arg(just[1L], c("center", "centre", "top"))
+  if (just == "centre") 
+    just <- "center"
+  align <- match.arg(align[1L], c("center", "centre", "left", 
+                                  "right"))
+  if (align == "centre") 
+    align <- "center"
+  rval <- function(node) {
+    if(node$id == 2) fill <- c("white", "blue")
+    if(node$id == 3 | node$id == 4) fill <- c("white", "red")
+    if(node$id == 5) fill <- c("white", "darkgreen")
+    fill <- rep(fill, length.out = 2)
+    lab <- extract_label(node)
+    if (!is.null(gp)) {
+      outer_vp <- viewport(gp = gp)
+      pushViewport(outer_vp)
+    }
+    if (is.null(height)) 
+      height <- length(lab) + 1L
+    node_vp <- viewport(x = unit(0.5, "npc"), y = unit(if (just == 
+                                                           "top") 
+      top
+      else 0.5, "npc"), just = c("center", just), width = unit(1, 
+                                                               "strwidth", nstr) * 1.1, height = unit(height, "lines"), 
+      name = paste("node_terminal", id_node(node), sep = ""), 
+      gp = if (is.null(gp)) 
+        gpar()
+      else gp)
+    pushViewport(node_vp)
+    grid.rect(gp = gpar(fill = fill[1]))
+    for (i in seq_along(lab)) grid.text(x = switch(align, 
+                                                   center = unit(0.5, "npc"), left = unit(1, "strwidth", 
+                                                                                          "a"), right = unit(1, "npc") - unit(1, "strwidth", 
+                                                                                                                              "a")), y = unit(length(lab) - i + 1, "lines"), 
+                                        lab[i], just = align)
+    if (id) {
+      nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
+                                                          "npc"), width = max(unit(1, "lines"), unit(1.3, 
+                                                                                                     "strwidth", nam[id_node(node)])), height = max(unit(1, 
+                                                                                                                                                         "lines"), unit(1.3, "strheight", nam[id_node(node)])))
+      pushViewport(nodeIDvp)
+      grid.rect(gp = gpar(fill = fill[2], lty = "solid"))
+      grid.text(nam[id_node(node)])
+      popViewport()
+    }
+    if (is.null(gp)) 
+      upViewport()
+    else upViewport(2)
+  }
+  return(rval)
+}
+
+class(tp) <- "grapcon_generator"
+
 @
@@
 plot(p, 
-     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
+     tp_args = list(FUN = identity, width = 15), 
      ip_args = list(fill = paltrees[c(2, 3)]),
-     drop_terminal = TRUE, tnex = 1, edge_panel = ep)
+     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
 @
@@
 paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
-plot(p, edge_panel = ep,
-     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
+plot(p, edge_panel = ep, terminal_panel = tp,
+     tp_args = list(FUN = identity, width = 15), 
      ip_args = list(fill = paltrees[c(2, 3)]),

### 2019-02-27 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
some touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2478 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d0468c79a26da1f358c42b0dcb7ac6b194b835d5

@@
-\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber, noslidenumber]{uibk}
+\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
+\usetheme[license, url]{uibk}
 
@@
 %\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
-\URL{https://R-Forge.R-project.org/projects/partykit/} 
-%\\
-%https://partykit.org}
+\URL{http://www.partykit.org/} 
 \headerimage{3}
@@
 %% includes a replacement for \usepackage{Sweave}
-\usepackage{Sweave}
 \usepackage{changepage}
@@
 \begin{frame}%[fragile]
-\frametitle{Simulation Setting}
+\frametitle{Simulation}
 
@@
 \begin{frame}%[fragile]
-\frametitle{Simulation Setting (Scenario 1)}
-True tree structure:
+\frametitle{Simulation 1: True tree structure}
 
@@
 \begin{frame}%[fragile]
-\frametitle{Results}
-Residuals vs.\ full model scores:
+\frametitle{Simulation 1: Residuals vs.\ full model scores}
 
@@
 \begin{frame}%[fragile]
-\frametitle{Results}
-Full factorial analysis (with both parameters varying):
+\frametitle{Simulation 1: Full factorial analysis (vary both)}
+
 \begin{figure}%[t]
@@
 \begin{frame}%[fragile]
-\frametitle{Simulation Setting (Scenario 2)}
-True tree structure:
+\frametitle{Simulation 2: True tree structure}
 
@@
 \begin{frame}%[fragile]
-\frametitle{Results}
-
-Residuals vs.\ full model scores:
+\frametitle{Simulation 2: Residuals vs.\ full model scores}
 
@@
 \begin{frame}%[fragile]
-\frametitle{Results}
-Pre-pruning vs.\ post-pruning:
+\frametitle{Simulation 2: Pre-pruning vs.\ post-pruning}
 
@@
 \frametitle{Recommendations}
+
 Based on the simulation results the following general conclusions can be drawn:
@@
 \vspace{0.4cm}
-\textbf{Package:} \emph{partykit} available on CRAN\\
+\textbf{Package:} Available on CRAN at\\
+
+\url{https://CRAN.R-project.org/package=partykit}\\
 
@@
 
-\textbf{Reproduction code:} available on R-Forge at\\
+\textbf{Replication code:} Available on R-Forge at\\
 

### 2019-03-01 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
further revisions, changed color palettes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2480 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 10c2bbe1c2c2de545727cb9189d0869b8b8ca57f

@@
 library("scales")
+library("colorspace")
 #source("../guidelike.R")
@@
 ## HCL palette
-pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
-names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")
+pal_dgp <- qualitative_hcl(3, "Set 2")
 
 ## HCL palette (light)
-pallight <- hcl(c(140, 80, 260, 70, 10), 300, 60, alpha = 0.25)
-names(pallight) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")
+pal_results <- qualitative_hcl(4, "Dark 3")
 
@@
 set.seed(7)
-x <- runif(50,-1,1)
-e1 <- rnorm(50,0,1)
-e2 <- rnorm(50,0,1)
+x <- runif(70,-1,1)
+e1 <- rnorm(70,0,1)
+e2 <- rnorm(70,0,1)
 y1 <- x*6 + e1 
@@
 jpeg("lm1.jpeg")
+par(mar=c(3.2,3.2,0,0))
 plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
+mtext("Y",side = 2, las = 2, line = 2, cex = 2.5, col = gray.colors(1,0.3))
+mtext("X",side = 1, las = 1, line = 2, cex = 2.5, col = gray.colors(1,0.3))
 points(x, y2)
-abline(lm(y~c(x,x)), col = "red", lwd = 2)
+abline(lm(y~c(x,x)), col = "red", lwd = 3)
 dev.off()
@@
 jpeg("lm2.jpeg")
+par(mar=c(3.2,3.2,0,0))
 plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
-abline(lm(y1~x), col = "red", lwd = 2)
+mtext("Y",side = 2, las = 2, line = 2, cex = 2.5, col = gray.colors(1,0.3))
+mtext("X",side = 1, las = 1, line = 2, cex = 2.5, col = gray.colors(1,0.3))
+abline(lm(y1~x), col = "red", lwd = 3)
 dev.off()
@@
 jpeg("lm3.jpeg")
+par(mar=c(3.2,3.2,0,0))
 plot(x,y2, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
-abline(lm(y2~x), col = "red", lwd = 2)
+mtext("Y",side = 2, las = 2, line = 2, cex = 2.5, col = gray.colors(1,0.3))
+mtext("X",side = 1, las = 1, line = 2, cex = 2.5, col = gray.colors(1,0.3))
+abline(lm(y2~x), col = "red", lwd = 3)
 dev.off()
@@
 \frametitle{Motivation}
+
+\vspace{-0.5cm}
+
 \center
 \begin{tikzpicture}
-\visible<1>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (n0) at (4, 5) {$Y,X,Z$};
+\visible<1-3>{
+\node[inner sep=0pt] (lm1) at (4, 6)
+    {\includegraphics[width=0.2\textwidth]{lm1.jpeg}};
 }
 \visible<2>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (n0) at (4, 5) {$\mathcal{M}(Y,X;\hat{\beta}$)};
+\node[scale=1] at (4.15, 3.7) {other covariates $Z_1, \ldots, Z_p$?};
 }
-\visible<3->{
-\node[inner sep=0pt] (lm1) at (4, 4.5)
-    {\includegraphics[width=0.25\textwidth]{lm1.jpeg}};
+\visible<3>{
+\node[inner sep=0pt] (lm2) at (2,1.7)
+    {\includegraphics[width=0.2\textwidth]{lm2.jpeg}};
+\node[inner sep=0pt] (lm3) at (6,1.7)
+    {\includegraphics[width=0.2\textwidth]{lm3.jpeg}};
+\draw[-, line width=0.2pt] (3.5,4.85) -- (2.5,3.1) node [midway, left] {\scriptsize $Z_j \leq \xi$};
+\draw[-, line width=0.2pt] (4.7,4.85) -- (5.7,3.1) node [midway, right] {\scriptsize $Z_j > \xi$};
 }
-\visible<4>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {?};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {?};
-\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6);
-\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6);
+\visible<4->{
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (model1) at (4.15, 5.35) {$\mathcal{M}(Y,X;\hat{\beta}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model2) at (2.15, 2.55) {$\mathcal{M}(Y_1,X_1;\hat{\beta}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model3) at (6.15, 2.55) {$\mathcal{M}(Y_2,X_2;\hat{\beta}$)};
+\draw[-, line width=0.2pt] (3.5,4.85) -- (2.5,3.1) node [midway, left] {\scriptsize $Z_j \leq \xi$};
+\draw[-, line width=0.2pt] (4.7,4.85) -- (5.7,3.1) node [midway, right] {\scriptsize $Z_j > \xi$};
 }
-\visible<5-6>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {$Y_1,X_1$};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {$Y_2,X_2$};
-\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6) node [midway, left] {\scriptsize $Z \leq \xi$};
-\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6) node [midway, right] {\scriptsize $Z > \xi$};
-}
-\visible<6>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {$\mathcal{M}(Y_1,X_1;\hat{\beta_1}$)};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {$\mathcal{M}(Y_2,X_2;\hat{\beta_2}$)};
-}
-\visible<7->{
-\node[inner sep=0pt] (lm2) at (2,0.5)
-    {\includegraphics[width=0.25\textwidth]{lm2.jpeg}};
-\node[inner sep=0pt] (lm3) at (6,0.5)
-    {\includegraphics[width=0.25\textwidth]{lm3.jpeg}};
-\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6) node [midway, left] {\scriptsize $Z \leq \xi$};
-\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6) node [midway, right] {\scriptsize $Z > \xi$};
+\visible<5>{
+\node[scale=1] at (4.15, 1) {$\mathcal{M}$ can also be a more general model (possibly without $X$).};
 }
@@
 \begin{enumerate}
-\item Fit a model to the response $Y$ (and possible covariates $X$).
+\item Fit a model $\mathcal{M}(Y,X;\hat{\beta})$ to the response $Y$ (and possible covariates $X$).
 \item Assess association of $Y$ (or a model-based transformation)
@@
 
-\vspace{1.2cm}
+\vspace{0.8cm}
 
@@
 \begin{enumerate}
-\item Evaluate a discrepancy measure giving information on the goodness of fit of the model for each observation and each model parameter (transformation of $Y$).
-\item Apply a statistical test assessing dependency of the discrepancy measure on each possible split variable $Z_j$.
+\item Evaluate a discrepancy measure giving information on the goodness of fit of the model for each observation and each model parameter (transformation~of~$Y$).
+\item Apply a statistical test assessing dependency of the discrepancy measure to each possible split variable $Z_j$.
 \item Select the split variable $Z_j^\ast$ showing the smallest $p$-value.
@@
 \begin{itemize}
-\item Absolute deviations from an average value
+\item (Ranks of) Y
 \item Residuals
 \item Model scores
+\item Absolute deviations from an average value
 \item \ldots
@@
 \frametitle{Discrepancy Measures}
-As an example linear models $\mathcal{LM}(Y,X;\beta_0, \beta_1)$
-fitted via ordinary least squares (OLS)
+As an example linear models $\mathcal{M}(Y,X;\beta_0, \beta_1)$
+fitted via ordinary least squares~(OLS)
 are considered with the following discrepancy measures:
@@
 \hspace{0.7cm}
-r(Y,X,\hat{\beta_0},\hat{\beta_1}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X &&
+r(y,x,\hat{\beta_0},\hat{\beta_1}) = y - \hat{\beta_0} - \hat{\beta_1} \cdot x &&
 \end{flalign*}
@@
 
-{\bf Model scores:}
+{\bf Model scores} (based on least squares):
 \begin{flalign*}
 \hspace{0.7cm}
-s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
-\left(\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_0} \quad , \quad  
-\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_1}\right) 
+s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
+\left(\frac{\partial r^2(y, x, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_0} \quad , \quad  
+\frac{\partial r^2(y, x, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_1}\right) 
 &&
@@
 
-\pause
-
-$${\color{red}\Downarrow}$$
+\visible<1>{
+\medskip
+Alternatively: Based on log-likelihood.
+}
+\visible<2>{
+\vspace{-1cm}
+$$\hspace{-0.3cm}{\color{red}\Downarrow} \hspace{3.4cm} {\color{red}\Downarrow} $$
 $$
-{\color{red} 2 \cdot r(Y,X,\hat{\beta_0},\hat{\beta_1})}
+{\color{red} -2 \cdot r(y,x,\hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot x \cdot r(y,x,\hat{\beta_0},\hat{\beta_1})}
 $$
-
+}
 \end{frame}
@@
 \item {\only<4->{\color{lightgray}} Residuals vs.\ full model scores}
-\item {\only<1-3,7->{\color{lightgray}} Binarization of residuals/scores}
-\item {\only<1-6>{\color{lightgray}} Categorization of possible split variables}
+\item {\only<1-3,6->{\color{lightgray}} Binarization of residuals/scores}
+\item {\only<1-5>{\color{lightgray}} Categorization of possible split variables}
 \end{itemize}
@@
 $$
-s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
-\begin{pmatrix} 
-\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
-\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
+s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
+-2 \cdot \begin{pmatrix} 
+r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})}\\
+r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})}\\
 \vdots & {\color{lightgray} \vdots} \\
-\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
+r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})}
 \end{pmatrix}
+%\begin{pmatrix}
+%\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
+%\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
+%\vdots & {\color{lightgray} \vdots} \\
+%\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
+%\end{pmatrix}
 $$
@@
 $$
-s(Y, X, \beta_0, \beta_1) = 
-\begin{pmatrix} 
-\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
-\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
+s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
+-2 \cdot \begin{pmatrix} 
+r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})\\
+r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})\\
 \vdots & \vdots \\
-\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
+r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})
 \end{pmatrix}
+%\begin{pmatrix} 
+%\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
+%\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
+%\vdots & \vdots \\
+%\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
+%\end{pmatrix}
 $$
 }
-\only<5-6>{
+\only<4-5>{
 \small
 $$
-r(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
+r(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
 \begin{pmatrix}
-r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
-r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
+r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})\\
+r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})\\
 \vdots \\
-r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
+r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})
 \end{pmatrix}
 %
-\visible<6>{
+\visible<5>{
 \quad \Rightarrow \quad
@@
 \begin{pmatrix}
-1\\
-0\\
+\text{pos}\\
+\text{non-pos}\\
 \vdots \\
-1
+\text{pos}
 \end{pmatrix}
@@
 }
-\only<7-8>{
+\only<6-7>{
 \small
 $$
-Z_j = 
+z_j = 
 \begin{pmatrix}
-Z_{j1}\\
-Z_{j2}\\
+z_{j1}\\
+z_{j2}\\
 \vdots \\
-Z_{jn}\\
+z_{jn}\\
 \end{pmatrix}
 %
-\visible<8>{
+\visible<7>{
 \quad \Rightarrow \quad
@@
 \begin{pmatrix}
-3\\
-1\\
+\text{Q}3\\
+\text{Q}1\\
 \vdots \\
-2
+\text{Q}2
 \end{pmatrix}
@@
 }
+\end{frame}
+
+\begin{frame}
+\frametitle{Combinations}
+
+The testing strategies of CTree, MOB, and GUIDE can be obtained from the general 
+framework by specific combinations of the building blocks together with the 
+corresponding approximation type of the underlying null distribution.
+
+\bigskip
+
+\begin{table}[t]
+\begin{tabular}{ l c c c c }
+\hline 
+      & Scores            & Binarization  & Categorization  & Null distribution \\ 
+\hline
+CTree & full model scores & --          & --            & conditional\\
+MOB   & full model scores & --          & --            & unconditional\\
+GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
+\hline
+\end{tabular}
+\end{table}
 \end{frame}
@@
   rval <- function(node) {
-    if(node$id == 2) fill <- c("white", "blue")
-    if(node$id == 3 | node$id == 4) fill <- c("white", "red")
-    if(node$id == 5) fill <- c("white", "darkgreen")
+    if(node$id == 2) fill <- c("white", pal_dgp[1])
+    if(node$id == 3 | node$id == 4) fill <- c("white", pal_dgp[2])
+    if(node$id == 5) fill <- c("white", pal_dgp[3])
     fill <- rep(fill, length.out = 2)
@@
 
+\vspace{-0.2cm}
+
 \begin{figure}%[t]
 \setkeys{Gin}{width=1\linewidth}
-\minipage{0.45\textwidth}
+\minipage{0.36\textwidth}
 <<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
@@
 \endminipage
-\minipage{0.45\textwidth}
+\minipage{0.36\textwidth}
 <<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
 par(mar = c(4,4,2,4))
-plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
+plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "", xlab = "", 
      main = expression("varying"~beta[0]),
-     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-abline(coef = c(-1,1), col = "blue", lwd = 2)
-abline(coef = c(1,1), col = "red", lwd = 2)
-points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = 3.2)
-mtext(expression(beta[1]== 1), side = 4, col = "red", las = 1, line = 0.5, at = 2.2)
-mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.2)
-mtext(expression(beta[1]== 1), side = 4, col = "blue", las = 1, line = 0.5, at = -0.8)
+     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
+mtext("y", side = 2, line = 3)
+mtext("x", side = 1, line = 2.5)
+abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
+abline(coef = c(1,1), col = pal_dgp[2], lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 3.2)
+mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 2.2)
+mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.2)
+mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = -0.8)
 @
@@
 
-\minipage{0.45\textwidth}
+\minipage{0.36\textwidth}
 <<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
@@
 par(mar = c(4,4,2,4))
-plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
-     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-abline(coef = c(0,1), col = "blue", lwd = 2)
-abline(coef = c(0,-1), col = "red", lwd = 2)
-points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==0), side = 4, col = "red", las = 1, line = 0.5, at = -1)
-mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
-mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
-mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1)
+plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "", xlab = "", main = expression("varying"~beta[1]),
+     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
+mtext("y", side = 2, line = 3)
+mtext("x", side = 1, line = 2.5)
+abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
+abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
+mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
+mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
 @
 \endminipage
-\minipage{0.45\textwidth}
+\minipage{0.36\textwidth}
 <<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
@@
 par(mar = c(4,4,2,4))
-plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
+plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "", xlab = "",
      main = expression("varying"~beta[0]~"and"~beta[1]),
      xlim = c(-1,1.1),
-     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-abline(coef = c(-1,1), col = "blue", lwd = 2)
-abline(coef = c(1,-1), col = "red", lwd = 2)
-points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.7)
-mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.7)
-mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = -0.7)
-mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -1.7)
+     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
+mtext("y", side = 2, line = 3)
+mtext("x", side = 1, line = 2.5)
+abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
+abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
+mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
+mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
 @
@@
 \begin{figure}
-\setkeys{Gin}{width=\linewidth}
+\setkeys{Gin}{width=0.8\linewidth}
 <<scores_stump, fig=TRUE, echo=FALSE, height=5, width=8>>=
@@
 
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
+par.settings <- list(superpose.symbol = list(col = pal_results, 
+                                             fill = pal_results), 
+                     superpose.line = list(col = pal_results))
 
@@
                                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
+                                               col = pal_results, lwd = 1)),
                       layout= c(3,2),
@@
                                                      expression("vary"~beta[0]), 
-                                                     expression("vary"~beta[1]))),
-               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8))))
+                                                     expression("vary"~beta[1])),
+                                   bg = "lightgray"),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
+                                       bg = "lightgray"))
 
@@
 \begin{figure}%[t]
-\setkeys{Gin}{width=1\linewidth}
+\setkeys{Gin}{width=0.8\linewidth}
 <<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
@@
        xlab = "Categorization",
-       ylab = expression("p-value of"~Z[1]),
+       ylab = expression(p~"-value of"~Z[1]),
        #index.cond = list(c(1,2,3,4)),
@@
 \setkeys{Gin}{width=1\linewidth}
-\minipage{0.5\textwidth}
+\minipage{0.45\textwidth}
 <<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
@@
 \endminipage
-\minipage{0.45\textwidth}
+\minipage{0.4\textwidth}
 <<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
@@
 plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
-     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-abline(coef = c(0,1), col = "blue", lwd = 2)
-abline(coef = c(-1,-1), col = "red", lwd = 2)
-abline(coef = c(1,-1), col = "darkgreen", lwd = 2)
+     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
+abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
+abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
+abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
 points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
-       col = alpha("red", 0.4), cex = 0.7)
+       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
 points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
-       col = alpha("darkgreen", 0.4), cex = 0.7)
+       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
 legend("top", c(expression(z[2]<=xi), 
@@
                 expression(z[2]>xi~" and "~z[1]>xi)), 
-       col = c("blue", "red", "darkgreen"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
-mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.3)
-mtext(expression(beta[0]==+delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = 0.1)
-mtext(expression(beta[1]==-delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = -0.6)
-mtext(expression(beta[0]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
-mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2.7)
+       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
+mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
+mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
+mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
 @
@@
 \begin{figure}%[t]
-\setkeys{Gin}{width=\linewidth}
+\setkeys{Gin}{width=0.8\linewidth}
 <<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
@@
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
+                                col = pal_results, lwd = 1)),
        layout= c(4,1),
@@
 \begin{figure}%[t]
-\setkeys{Gin}{width=\linewidth}
+\setkeys{Gin}{width=0.8\linewidth}
 <<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
+par.settings <- list(superpose.symbol = list(col = pal_results, 
+                                             fill = pal_results), 
+                     superpose.line = list(col = pal_results))
 
@@
 
-\bigskip
+%\bigskip
 
-\textbf{Applied functions:}
+%\textbf{Applied functions:}
 
-\begin{itemize}
-\item \texttt{ctree}
-\item \texttt{mob}
-\item Adaptions of both of them
-\item Reimplementation of GUIDE
-\end{itemize}
+%\begin{itemize}
+%\item \texttt{ctree}
+%\item \texttt{mob}
+%\item Adaptions of both of them
+%\item Reimplementation of GUIDE
+%\end{itemize}
 \end{frame}

### 2019-03-05 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
small corrections in notation and illustrations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2482 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cca523f738777eba594966e8394465dd42c01660

@@
 
-{\bf General strategy:}
+{\bf General testing strategy:}
 \begin{enumerate}
@@
 \hspace{0.7cm}
-r(y,x,\hat{\beta_0},\hat{\beta_1}) = y - \hat{\beta_0} - \hat{\beta_1} \cdot x &&
+r(Y,X,\hat{\beta_0},\hat{\beta_1}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X &&
 \end{flalign*}
@@
 \hspace{0.7cm}
-s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
-\left(\frac{\partial r^2(y, x, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_0} \quad , \quad  
-\frac{\partial r^2(y, x, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_1}\right) 
+s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
+\left(\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_0} \quad , \quad  
+\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_1}\right) 
 &&
@@
 $$
-{\color{red} -2 \cdot r(y,x,\hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot x \cdot r(y,x,\hat{\beta_0},\hat{\beta_1})}
+{\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot X \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}
 $$
@@
 $$
-s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
+s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
 -2 \cdot \begin{pmatrix} 
-r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})}\\
-r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})}\\
+r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} X \cdot r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
+r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} X \cdot r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
 \vdots & {\color{lightgray} \vdots} \\
-r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})}
+r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} X \cdot r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
 \end{pmatrix}
@@
 $$
-s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
+s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
 -2 \cdot \begin{pmatrix} 
-r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})\\
-r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})\\
+r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & X \cdot r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
+r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & X \cdot r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
 \vdots & \vdots \\
-r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})
+r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & X \cdot r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
 \end{pmatrix}
@@
 $$
-r(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
+r(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
 \begin{pmatrix}
-r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})\\
-r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})\\
+r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
+r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
 \vdots \\
-r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})
+r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
 \end{pmatrix}
@@
 $$
-z_j = 
+Z_j = 
 \begin{pmatrix}
-z_{j1}\\
-z_{j2}\\
+Z_{j1}\\
+Z_{j2}\\
 \vdots \\
-z_{jn}\\
+Z_{jn}\\
 \end{pmatrix}
@@
      pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
-mtext("y", side = 2, line = 3)
-mtext("x", side = 1, line = 2.5)
+mtext("Y", side = 2, line = 2.5)
+mtext("X", side = 1, line = 2.5)
 abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
@@
      pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
-mtext("y", side = 2, line = 3)
-mtext("x", side = 1, line = 2.5)
+mtext("Y", side = 2, line = 2.5)
+mtext("X", side = 1, line = 2.5)
 abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
@@
 par(mar = c(4,4,2,4))
-plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
+plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "Y", xlab = "X",
      pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
@@
 \item Binarizing scores/residuals does not lead to any improvements in performance.
-\item Categorizing split variables can facilitate calculations, however, important
+\item Categorizing split variables can simplify calculations, however, important
 effects might be missed unless a binning point is close to the true split point.

### 2019-03-06 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
some small improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2483 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5a0fa24666bf7e1c93e3bbf1bb40498c3e7cea53

@@
 \visible<2>{
-\node[scale=1] at (4.15, 3.7) {other covariates $Z_1, \ldots, Z_p$?};
+\node[scale=1] at (4.15, 3.7) {Other covariates $Z_1, \ldots, Z_p$?};
 }
@@
 \begin{enumerate}
-\item Fit a model $\mathcal{M}(Y,X;\hat{\beta})$ to the response $Y$ (and possible covariates $X$).
-\item Assess association of $Y$ (or a model-based transformation)
+\item Fit a model $\mathcal{M}(Y,X;\hat{\beta})$ to the response $Y$ and possible covariates $X$.
+\item Assess association of $\mathcal{M}(Y,X;\hat{\beta})$
 and each possible split variable~$Z_j$ and select the split~variable~$Z_{j^\ast}$ showing the strongest association.
@@
 \begin{enumerate}
-\item Evaluate a discrepancy measure giving information on the goodness of fit of the model for each observation and each model parameter (transformation~of~$Y$).
+\item Evaluate a discrepancy measure capturing the obervation-wise goodness of fit of $\mathcal{M}(Y,X;\hat{\beta})$.
 \item Apply a statistical test assessing dependency of the discrepancy measure to each possible split variable $Z_j$.
@@
 
-{\bf Possible discrepancy measures:}
+{\bf Discrepancy measures:} (Model-based) transformations of $Y$ (and $X$, if any), possibly for each model parameter.
 \begin{itemize}
-\item (Ranks of) Y
-\item Residuals
-\item Model scores
-\item Absolute deviations from an average value
+\item (Ranks of) $Y$.
+\item (Absolute) deviations $Y - \bar Y$.
+\item Residuals of $\mathcal{M}(Y,X;\hat{\beta})$.
+\item Score matrix of $\mathcal{M}(Y,X;\hat{\beta})$.
 \item \ldots
@@
 \frametitle{Discrepancy Measures}
-As an example linear models $\mathcal{M}(Y,X;\beta_0, \beta_1)$
-fitted via ordinary least squares~(OLS)
-are considered with the following discrepancy measures:
+
+\textbf{Example:} Simple linear regression $\mathcal{M}(Y,X;\beta_0, \beta_1)$,
+fitted via ordinary least squares~(OLS).
 
@@
 
-{\bf Model scores} (based on least squares):
+{\bf Model scores:} Based on log-likelihood or residual sum of squares.
 \begin{flalign*}
@@
 
-\visible<1>{
-\medskip
-Alternatively: Based on log-likelihood.
-}
 \visible<2>{
-\vspace{-1cm}
+\vspace{-0.3cm}
 $$\hspace{-0.3cm}{\color{red}\Downarrow} \hspace{3.4cm} {\color{red}\Downarrow} $$
 $$
-{\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot X \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}
+{\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1}) \cdot X}
 $$
 }
+\end{frame}
+
+\begin{frame}
+\frametitle{A Unifying view}
+
+The testing strategies of CTree, MOB, and GUIDE can be obtained from the general 
+framework by specific combinations of the building blocks together with the 
+corresponding approximation type of the underlying null distribution.
+
+\bigskip
+
+\begin{table}[t]
+\begin{tabular}{lllll}
+\hline 
+      & Scores       & Binarization  & Categorization  & Statistic \\ 
+\hline
+CTree & Model scores & --	   & -- 	   & Sum of squares\\
+MOB   & Model scores & --	   & -- 	   & Maximally selected\\
+GUIDE & Residuals    & \checkmark  & \checkmark    & Sum of squares\\
+\hline
+\end{tabular}
+\end{table}
 \end{frame}
@@
 -2 \cdot \begin{pmatrix} 
-r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} X \cdot r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
-r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} X \cdot r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
+r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X}\\
+r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X}\\
 \vdots & {\color{lightgray} \vdots} \\
-r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} X \cdot r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
+r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X}
 \end{pmatrix}
@@
 -2 \cdot \begin{pmatrix} 
-r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & X \cdot r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
-r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & X \cdot r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
+r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X\\
+r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X\\
 \vdots & \vdots \\
-r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & X \cdot r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
+r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X
 \end{pmatrix}
@@
 \begin{pmatrix}
-\text{pos}\\
-\text{non-pos}\\
+> 0\\
+\le 0\\
 \vdots \\
-\text{pos}
+> 0
 \end{pmatrix}
@@
 }
-\end{frame}
-
-\begin{frame}
-\frametitle{Combinations}
-
-The testing strategies of CTree, MOB, and GUIDE can be obtained from the general 
-framework by specific combinations of the building blocks together with the 
-corresponding approximation type of the underlying null distribution.
-
-\bigskip
-
-\begin{table}[t]
-\begin{tabular}{ l c c c c }
-\hline 
-      & Scores            & Binarization  & Categorization  & Null distribution \\ 
-\hline
-CTree & full model scores & --          & --            & conditional\\
-MOB   & full model scores & --          & --            & unconditional\\
-GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
-\hline
-\end{tabular}
-\end{table}
 \end{frame}

### 2019-03-06 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
revision of recommendations and further small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2484 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 02da08f193b42d809a6aa3269f6ff390613395a9

@@
 ## HCL palette
-pal_dgp <- qualitative_hcl(3, "Set 2")
+pal_dgp_light <- qualitative_hcl(3, "Set 2")
+pal_dgp <- qualitative_hcl(3, "Dark 3")
 
@@
 \begin{enumerate}
-\item Evaluate a discrepancy measure capturing the obervation-wise goodness of fit of $\mathcal{M}(Y,X;\hat{\beta})$.
+\item Evaluate a discrepancy measure capturing the observation-wise goodness of fit of $\mathcal{M}(Y,X;\hat{\beta})$.
 \item Apply a statistical test assessing dependency of the discrepancy measure to each possible split variable $Z_j$.
@@
 \begin{frame}
-\frametitle{A Unifying view}
+\frametitle{A Unifying View}
 
-The testing strategies of CTree, MOB, and GUIDE can be obtained from the general 
-framework by specific combinations of the building blocks together with the 
-corresponding approximation type of the underlying null distribution.
+The testing strategies of CTree, MOB, and GUIDE can be obtained from a\\
+general framework by combining building blocks as follows:
 
@@
 \end{table}
+
+\vspace{0.5cm}
+\textit{Note:}\\
+\begin{itemize}\item Adaptions of standard versions are provided.\\
+\item Further differences (e.g., approximation of null distribution) remain.
+\end{itemize}
 \end{frame}
@@
 \begin{itemize}
-\item {\only<4->{\color{lightgray}} Residuals vs.\ full model scores}
-\item {\only<1-3,6->{\color{lightgray}} Binarization of residuals/scores}
-\item {\only<1-5>{\color{lightgray}} Categorization of possible split variables}
+\item {\only<4->{\color{lightgray}} Residuals vs.\ full model scores.}
+\item {\only<1-3,6->{\color{lightgray}} Binarization of residuals/scores.}
+\item {\only<1-5>{\color{lightgray}} Categorization of possible split variables.}
 \end{itemize}
@@
 -2 \cdot \begin{pmatrix} 
-r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X}\\
-r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X}\\
+r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X_1}\\
+r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X_2}\\
 \vdots & {\color{lightgray} \vdots} \\
-r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X}
+r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X_n}
 \end{pmatrix}
@@
 -2 \cdot \begin{pmatrix} 
-r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X\\
-r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X\\
+r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X_1\\
+r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X_2\\
 \vdots & \vdots \\
-r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X
+r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X_n
 \end{pmatrix}
@@
   rval <- function(node) {
-    if(node$id == 2) fill <- c("white", pal_dgp[1])
-    if(node$id == 3 | node$id == 4) fill <- c("white", pal_dgp[2])
-    if(node$id == 5) fill <- c("white", pal_dgp[3])
+    if(node$id == 2) fill <- c("white", pal_dgp_light[1])
+    if(node$id == 3 | node$id == 4) fill <- c("white", pal_dgp_light[2])
+    if(node$id == 5) fill <- c("white", pal_dgp_light[3])
     fill <- rep(fill, length.out = 2)
@@
                                                      expression("vary"~beta[1])),
-                                   bg = "lightgray"),
+                                   bg = "gray90"),
                strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
-                                       bg = "lightgray"))
+                                       bg = "gray90"))
 
@@
 
-\begin{frame}%[fragile]
-\frametitle{Simulation 1: Full factorial analysis (vary both)}
+%\begin{frame}%[fragile]
+%\frametitle{Simulation 1: Full factorial analysis (vary both)}
 
-\begin{figure}%[t]
-\setkeys{Gin}{width=0.8\linewidth}
-<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
-xyplot(pval_z1 ~ cat | xi, 
-       groups = interaction(bin, res_scores), 
-       data = d3way, 
-       ylim = c(0,0.5),
-       type = c("a"), 
-       col = c("red", "blue", "red", "blue"),
-       lty = c(1,1,2,2),
-       xlab = "Categorization",
-       ylab = expression(p~"-value of"~Z[1]),
-       #index.cond = list(c(1,2,3,4)),
-       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
-                                              expression(xi==0.8~"and"~delta==1))),
-       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
-                  lines = list(lty = c(1,1,2,2), 
-                               type = "l",
-                               col = c("red", "blue", "red", "blue"))))
-@
-\end{figure}
-\end{frame}
+%\begin{figure}%[t]
+%\setkeys{Gin}{width=0.8\linewidth}
+%<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
+%xyplot(pval_z1 ~ cat | xi, 
+%       groups = interaction(bin, res_scores), 
+%       data = d3way, 
+%       ylim = c(0,0.5),
+%       type = c("a"), 
+%       col = c("red", "blue", "red", "blue"),
+%       lty = c(1,1,2,2),
+%       xlab = "Categorization",
+%       ylab = expression(p~"-value of"~Z[1]),
+%       #index.cond = list(c(1,2,3,4)),
+%       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
+%                                              expression(xi==0.8~"and"~delta==1))),
+%       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
+%                  lines = list(lty = c(1,1,2,2), 
+%                               type = "l",
+%                               col = c("red", "blue", "red", "blue"))))
+%@
+%\end{figure}
+%\end{frame}
 
@@
                                         "guide_sum_12", "guide_sum_1_cor"))
+#subdata <- subset(subdata, xi %in% c(0,0.5,0.8))
 subdata$test <- factor(subdata$test, 
@@
                                               expression(xi==0.5), 
-                                              expression(xi==0.8))),
+                                              expression(xi==0.8)),
+                            bg = "gray90"),
        par.settings = par.settings)
@@
 
+#prex <- subset(prex, xi %in% c(0,0.5,0.8))
+
 xyplot(ari ~ delta | xi, groups = ~ test, 
@@
                                               expression(xi==0.5), 
-                                              expression(xi==0.8))),
+                                              expression(xi==0.8)),
+                            bg = "gray90"),
        par.settings = par.settings
@@
 
-Based on the simulation results the following general conclusions can be drawn:
+\vspace{0.4cm}
+
+In this setting:
+
+\vspace{0.2cm}
+
 \begin{itemize}
-\item Including full model scores is advantageous over using only residuals.
-\item Binarizing scores/residuals does not lead to any improvements in performance.
-\item Categorizing split variables can simplify calculations, however, important
-effects might be missed unless a binning point is close to the true split point.
-\item If the applied statistical test performs well as significance test pre-pruning
-is preferable. Otherwise post-pruning can compensate this deficiency at least partly.
+\setlength\itemsep{0.2cm}
+\item Include full model scores better than residuals only.
+\item Use original values of scores/residuals better than binarized values.
+\item Categorization is simpler, but less powerful in margins.
+\item If the significance test performs well pre-pruning works well,\\
+otherwise post-pruning might be needed.
+%\item Including full model scores is advantageous over using only residuals.
+%\item Binarizing scores/residuals does not improve the performance.
+%\item Categorizing split variables can simplify calculations, however,
+%effects might be missed unless a binning point is close to the true split point.
+%\item If the applied test performs well as significance test pre-pruning
+%is preferable, otherwise post-pruning.
 \end{itemize}
@@
  \dquote{The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE.}
- \emph{arXiv ...}, arXiv.org E-Print Archive.
- \doi{...}
+ \emph{paper in preparation}.
 

### 2019-03-06 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
feedback from TH
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2485 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 54ba039a7f30477bf8dd5011e3d507d24fee69ff

@@
 \node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (model1) at (4.15, 5.35) {$\mathcal{M}(Y,X;\hat{\beta}$)};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model2) at (2.15, 2.55) {$\mathcal{M}(Y_1,X_1;\hat{\beta}$)};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model3) at (6.15, 2.55) {$\mathcal{M}(Y_2,X_2;\hat{\beta}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model2) at (2.15, 2.55) {$\mathcal{M}(Y_1,X_1;\hat{\beta_1}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model3) at (6.15, 2.55) {$\mathcal{M}(Y_2,X_2;\hat{\beta_2}$)};
 \draw[-, line width=0.2pt] (3.5,4.85) -- (2.5,3.1) node [midway, left] {\scriptsize $Z_j \leq \xi$};
@@
 
-The testing strategies of CTree, MOB, and GUIDE can be obtained from a\\
-general framework by combining building blocks as follows:
+\textbf{Algorithms:} CTree, MOB, GUIDE are all `flavors' of the general framework.
 
@@
 
-\begin{table}[t]
+\textbf{Building blocks:} For standard setup.
+
+\bigskip
+
 \begin{tabular}{lllll}
@@
 \end{tabular}
-\end{table}
 
-\vspace{0.5cm}
-\textit{Note:}\\
-\begin{itemize}\item Adaptions of standard versions are provided.\\
-\item Further differences (e.g., approximation of null distribution) remain.
+\bigskip
+
+\textbf{Remarks:}
+\begin{itemize}
+  \item All three algorithms allow for certain modifications of standard setup.
+  \item Further differences, e.g., null distribution, pruning strategy, etc.
 \end{itemize}
+
 \end{frame}
@@
 
-{\bf Two Strategies:}
+{\bf Two strategies:}
 
 \begin{itemize}
-\item Pre-pruning: \\
+\item \emph{Pre-pruning:}
 Internal stopping criterion based on Bonferroni-corrected $p$-values of the applied tests. Stop splitting when there is no significant association. 
-\item Post-pruning: \\
+\item \emph{Post-pruning:}
 First grow a very large tree and afterwards prune splits that do not improve the model fit, 
@@
 
-\vspace{0.4cm}
-
-In this setting:
-
-\vspace{0.2cm}
+\textbf{In this setting:}
 
@@
 \setlength\itemsep{0.2cm}
-\item Include full model scores better than residuals only.
-\item Use original values of scores/residuals better than binarized values.
+\item Full model scores better than residuals only.
+\item Original values of scores/residuals better than binarized values.
 \item Categorization is simpler, but less powerful in margins.
@@
 \frametitle{Software}
-\vspace{0.4cm}
-\textbf{Package:} Available on CRAN at\\
+
+\textbf{Package:} Available on CRAN at
 
@@
 
-\textbf{Replication code:} Available on R-Forge at\\
+\textbf{Replication code:} Available on R-Forge at
 
@@
  \dquote{The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE.}
- \emph{paper in preparation}.
+ \emph{Working paper in preparation}.
 

### 2019-03-29 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added evaluation with continuous coefficients to appendix; changed color palletes; further minor revisions, also based on feedback from DAGStat2019 presentation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2506 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b1baf05c6a93261edf18f4ce93de579b7dde02cd

--- 2019-02-20 pkg/partykit/inst/guideliketest/draft/draft.Rnw
@@
 library("scales")
+library("colorspace")
 
@@
 ## HCL palette
-pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
-names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")
+pal <- qualitative_hcl(5, "Dark 3")
+names(pal) <- c("CTree", "MOB", "GUIDE", "GUIDE_scores", "CTree_max")
+
+## HCL palette for illustrations of data generating process
+pal_dgp_light <- qualitative_hcl(3, "Set 2")
+pal_dgp <- qualitative_hcl(3, "Dark 3")
+
+## HCL palette for tree panels
+paltrees <- rgb(c(0.9, 1), c(0.9, 1), c(0.9, 1))
 
@@
 \begin{enumerate}
-\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
+\item A discrepancy measure is obtained in the form of an $n \times k$-matrix $D$ giving information
 on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
@@
 These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
-for CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which gives a more 
-detailed overview of the differences in the second column of Table~\ref{tab:combinations}.
+for the standard setup of CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which 
+gives a more detailed overview of the differences in the second column of Table~\ref{tab:combinations}.
 
@@
 \textit{Note:}
-Apart from these three building blocks, the algorithms differ in their approximation of the 
-underlying null distribution of the test statistic. While the permutation tests in the CTree 
-algorithm work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional 
-approach.
+Apart from these three building blocks, the algorithms CTree, MOB, and GUIDE differ in the type of test 
+statistic they apply and their approximation of the underlying null distribution of the corresponding 
+test statistic. While the permutation tests in CTree work in a conditional framework both MOB (M-fluctuation tests) 
+and GUIDE ($\chi^2$-tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
+introduced employing prepruning while GUIDE originally applies a postpruning strategy.
+However, these differences refer to the standard versions of the three algorithms and all of them
+allow for certain modifications of this standard setup. For example, a different form of test statistic
+can be selected in CTree considering a maximum value as in MOB.
 
@@
 \caption[Table caption text]{Testing strategies of CTree, MOB, and GUIDE with the corresponding
-setting of the building blocks in the unifying framework and the approximation type of the
-underlying null distribution.}
+setting of the building blocks in the unifying framework and the type of test statistic.}
 \label{tab:buildingblocks}
@@
 \hline 
-      & Scores            & Binarization  & Categorization  & Null distribution \\ 
+      & Scores       & Binarization  & Categorization  & Statistic\\
 \hline
-CTree & full model scores & --          & --            & conditional\\
-MOB   & full model scores & --          & --            & unconditional\\
-GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
+CTree & Model scores & --	           & -- 	           & Sum of squares\\
+MOB   & Model scores & --	           & -- 	           & Maximally selected\\
+GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
 \hline
@@
 For each node the type of the fitted model is the same such that no node-specific features are 
-considered and the observations of the current (arbitrarily chosen) node are denoted by 
+considered and the observations of the (arbitrarily chosen) current node are denoted by 
 $\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.
@@
 \frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
+-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
 $$
-where the factor $2$ is often omitted such that the first column of the $n \times 2$-dimensional
+where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
 score matrix corresponds to the residuals. 
@@
 p <- party(pn, d)
-
-paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
 
@@
 class(ep) <- "grapcon_generator"
+
+tp <- function (obj, digits = 3, abbreviate = FALSE, 
+                fill = c("lightgray", "white"), id = TRUE, 
+                just = c("center", "top"), top = 0.85, 
+                align = c("center", "left", "right"), gp = NULL, FUN = NULL, 
+                height = NULL, width = NULL) 
+{
+  nam <- names(obj)
+  extract_label <- function(node) formatinfo_node(node, FUN = FUN, 
+                                                  default = c("terminal", "node"))
+  maxstr <- function(node) {
+    lab <- extract_label(node)
+    klab <- if (is.terminal(node)) 
+      ""
+    else unlist(lapply(kids_node(node), maxstr))
+    lab <- c(lab, klab)
+    lab <- try(unlist(lapply(lab, function(x) strsplit(x, 
+                                                       "\n"))), silent = TRUE)
+    if (inherits(lab, "try-error")) {
+      paste(rep("a", 9L), collapse = "")
+    }
+    else {
+      return(lab[which.max(nchar(lab))])
+    }
+  }
+  nstr <- if (is.null(width)) 
+    maxstr(node_party(obj))
+  else paste(rep("a", width), collapse = "")
+  just <- match.arg(just[1L], c("center", "centre", "top"))
+  if (just == "centre") 
+    just <- "center"
+  align <- match.arg(align[1L], c("center", "centre", "left", 
+                                  "right"))
+  if (align == "centre") 
+    align <- "center"
+  rval <- function(node) {
+    if(node$id == 2) fill <- c("white", pal_dgp_light[1])
+    if(node$id == 3 | node$id == 4) fill <- c("white", pal_dgp_light[2])
+    if(node$id == 5) fill <- c("white", pal_dgp_light[3])
+    fill <- rep(fill, length.out = 2)
+    lab <- extract_label(node)
+    if (!is.null(gp)) {
+      outer_vp <- viewport(gp = gp)
+      pushViewport(outer_vp)
+    }
+    if (is.null(height)) 
+      height <- length(lab) + 1L
+    node_vp <- viewport(x = unit(0.5, "npc"), y = unit(if (just == 
+                                                           "top") 
+      top
+      else 0.5, "npc"), just = c("center", just), width = unit(1, 
+                                                               "strwidth", nstr) * 1.1, height = unit(height, "lines"), 
+      name = paste("node_terminal", id_node(node), sep = ""), 
+      gp = if (is.null(gp)) 
+        gpar()
+      else gp)
+    pushViewport(node_vp)
+    grid.rect(gp = gpar(fill = fill[1]))
+    for (i in seq_along(lab)) grid.text(x = switch(align, 
+                                                   center = unit(0.5, "npc"), left = unit(1, "strwidth", 
+                                                                                          "a"), right = unit(1, "npc") - unit(1, "strwidth", 
+                                                                                                                              "a")), y = unit(length(lab) - i + 1, "lines"), 
+                                        lab[i], just = align)
+    if (id) {
+      nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
+                                                          "npc"), width = max(unit(1, "lines"), unit(1.3, 
+                                                                                                     "strwidth", nam[id_node(node)])), height = max(unit(1, 
+                                                                                                                                                         "lines"), unit(1.3, "strheight", nam[id_node(node)])))
+      pushViewport(nodeIDvp)
+      grid.rect(gp = gpar(fill = fill[2], lty = "solid"))
+      grid.text(nam[id_node(node)])
+      popViewport()
+    }
+    if (is.null(gp)) 
+      upViewport()
+    else upViewport(2)
+  }
+  return(rval)
+}
+
+class(tp) <- "grapcon_generator"
+
 @
@@
 plot(p, 
-     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]),
-     drop_terminal = TRUE, tnex = 1, edge_panel = ep)
+     tp_args = list(FUN = identity, width = 15), 
+     ip_args = list(fill = paltrees[c(1, 2)]),
+     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
 @
@@
      main = expression("varying"~beta[0]),
-     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-abline(coef = c(-1,1), col = "blue", lwd = 2)
-abline(coef = c(1,1), col = "red", lwd = 2)
-points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = 3.2)
-mtext(expression(beta[1]== 1), side = 4, col = "red", las = 1, line = 0.5, at = 2.2)
-mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.2)
-mtext(expression(beta[1]== 1), side = 4, col = "blue", las = 1, line = 0.5, at = -0.8)
+     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
+abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
+abline(coef = c(1,1), col = pal_dgp[2], lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 3.2)
+mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 2.2)
+mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.2)
+mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = -0.8)
 @
@@
 plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
-     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-abline(coef = c(0,1), col = "blue", lwd = 2)
-abline(coef = c(0,-1), col = "red", lwd = 2)
-points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==0), side = 4, col = "red", las = 1, line = 0.5, at = -1)
-mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
-mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
-mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1)
+     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
+abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
+abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
+mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
+mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
 @
@@
      xlim = c(-1,1.1),
-     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
-abline(coef = c(-1,1), col = "blue", lwd = 2)
-abline(coef = c(1,-1), col = "red", lwd = 2)
-points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha("red", 0.4), cex = 0.7)
-legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.7)
-mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.7)
-mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = -0.7)
-mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -1.7)
+     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
+abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
+abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
+points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
+legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
+mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
+mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
 @
@@
 par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
-#paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
-plot(p, edge_panel = ep,
-     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]),
+plot(p, edge_panel = ep, terminal_panel = tp,
+     tp_args = list(FUN = identity, width = 15), 
+     ip_args = list(fill = paltrees[c(1, 2)]),
      drop_terminal = TRUE, tnex = 1)
@@
                 expression(z[2]>xi~" and "~z[1]>xi)), 
-       col = c("blue", "red", "darkgreen"), pch = c(19), bty = "n", cex = 0.9)
-mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
-mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.3)
-mtext(expression(beta[0]==+delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = 0.1)
-mtext(expression(beta[1]==-delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = -0.6)
-mtext(expression(beta[0]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
-mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2.7)
+       col = c(pal[1], pal[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
+mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
+mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
+mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
+mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
+mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
 @
 \endminipage
-\caption{\label{fig:dgp_tree}Left panel: True tree structure for the second scenario with the 
-true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
+\caption{\label{fig:dgp_tree}Left panel: True tree structure for the second simulation scenario 
+with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
 Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
@@
                       type = "l", lty = c(1,1,1,1),
+                      lwd = 2,
                       scales=list(x=list(at=seq(1,11,2))),
@@
                                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
+                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
                       layout= c(3,2),
@@
                                                      expression("vary"~beta[0]), 
-                                                     expression("vary"~beta[1]))),
-               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8))))
+                                                     expression("vary"~beta[1])),
+                                   bg = "gray90"),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
+                                   bg = "gray90"))
 
@@
 \caption{\label{fig:scores_stump}Proportion of cases where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ for $\sigma = 1$, $\xi=0$, 250 observations and 
-100 data sets per step. Each panel represents one scenario for the coefficients: 
-varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and 
-both coefficients varying (right panel). The performances of four testing strategies are compared: 
+selected over increasing effect size $\delta$ (100 data sets of 250 observations per step),
+once for a true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row). 
+Each column represents one scenario for the coefficients: 
+varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
+both coefficients varying (right). The performances of four testing strategies are compared: 
 CTree, MOB, GUIDE, GUIDE\_scores.}
@@
 even for a high effect size. 
+(In order to show that the here observed advantage of MOB over CTree is mainly due to the 
+abrupt shift in the model parameters $\beta_0$ and $\beta_1$ an additional evaluation is 
+presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)
 
@@
 $\delta = 0.3$ is considered making the differences particularly well observable 
-(additional results over increasing $\delta$ as in Section~\ref{sec:scores} 
+(additional easing $\delta$ as in Section~\ref{sec:scores} 
 can be found in Appendix~\ref{app:increasingdelta}).
@@
        type = c("a"), 
-       col = c("red", "blue", "red", "blue"),
+       col = c(pal[1], pal[3], pal[1], pal[3]),
        lty = c(1,1,2,2),
+       lwd = 2,
        xlab = "Categorization",
@@
        strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
-                                              expression(xi==0.8~"and"~delta==1))),
+                                              expression(xi==0.8~"and"~delta==1)),
+                                   bg = "gray90"),
        key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
                   lines = list(lty = c(1,1,2,2), 
-                               type = "l",
-                               col = c("red", "blue", "red", "blue"))))
+                               type = "l", lwd = 2,
+                               col = c(pal[1], pal[3], pal[1], pal[3]))))
 
@@
        lty = c(1,1,1,1),
+       lwd = 2,
        key =  list(text = list(levels(subdata$test)),
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
+                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
        layout= c(4,1),
@@
                                               expression(xi==0.5), 
-                                              expression(xi==0.8))),
+                                              expression(xi==0.8)),
+                                   bg = "gray90"),
        par.settings = par.settings)
@@
 @
-\caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ for
-250 observations and 100 data sets per step.
-Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ and 
+\caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ 
+(100 data sets of 250 observations per step).
+Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
 $0.8$ from left to right). The performances of four testing strategies are compared: CTree,
@@
        data = prex, 
-       type = "l", lty = c(1,1,1,1,2,2,2,2),
+       type = "l", lty = c(2,2,2,2,1,1,1,1),
+       lwd = 2,
        key =  list(text = list(levels(prex$test)[1:4]),
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
+                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
        layout= c(4,1),
@@
                                               expression(xi==0.5), 
-                                              expression(xi==0.8))),
+                                              expression(xi==0.8)),
+                                   bg = "gray90"),
        par.settings = par.settings
@@
 @
-\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ for
-250 observations and 100 data sets per step.
-Each panel represents a different value of the split point $\xi$ ($0$, $0.2$, $0.5$ and 
-$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
-MOB, GUIDE, GUIDE\_scores. The dashed lines represent the results from applying postpruning
-and the continuous lines the results from applying prepruning.}
+\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
+(100 data sets of 250 observations per step).
+Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
+$0.8$ from left to right). 
+The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE\_scores. 
+The dashed lines represent the results from applying prepruning and the continuous lines 
+the results from applying postpruning.}
 \end{figure}
@@
 However, even then the results are very similar to those achieved without any categorization.
-But it has to be noted that a categorization of possible split variables can facilitate
+But it has to be noted that a categorization of possible split variables can simplify
 evaluations. Moreover, depending on the focus of interest, the results can be more
@@
 \caption[Table caption text]{Applied testing strategies representing the 12 combinations
-of building blocks}
+of building blocks.}
 \label{tab:combinations_appendix}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \setkeys{Gin}{width=\linewidth}
-<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=4, width=8>>=
+<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
 subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
@@
 
+subdataxi <- subset(subdata, xi %in% c(0,0.8))
+
 par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
@@
 
-xyplot(prop_z1 ~ delta | vary_beta, groups = ~ test, 
-       data = subdata, 
+library("latticeExtra")
+useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
+                      data = subdataxi, 
                       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & xi == 0, 
+                        only_intercept == FALSE,# & xi == 0, 
                       type = "l", lty = c(1,1,1,1),
-       key =  list(text = list(levels(subdata$test)),
+                      lwd = 2,
+                      scales=list(x=list(at=seq(1,11,2))),
+                      key =  list(text = list(levels(subdataxi$test)),
                                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 1)),
-       layout= c(3,1),
-       index.cond = list(c(2,3,1)),
-       scales=list(x=list(at=seq(1,11,2))),
+                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
+                      layout= c(3,2),
+                      index.cond = list(c(2,3,1),c(2,1)),
                       xlab = expression(delta),
                       ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
-       strip = strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
+                      par.settings = par.settings),
+
+               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
                                                      expression("vary"~beta[0]), 
-                                              expression("vary"~beta[1]))),
+                                                     expression("vary"~beta[1])),
+                                   bg = "gray90"),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
+                                   bg = "gray90"))
+
+@
+\caption{\label{fig:scores_stump_wosign}Proportion of cases where the true split variable $Z_1$ 
+shows the smallest $p$-value, hence, where $Z_1$ is selected without any significance level, 
+over increasing effect size $\delta$ (100 data sets of 250 observations per step), once for a 
+true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row).
+Each column represents one scenario for the coefficients: 
+varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
+both coefficients varying (right). The performances of four testing strategies are compared: 
+CTree, MOB, GUIDE, GUIDE\_scores.}
+\end{figure}
+
+\section{Simulation Results for Continuous Model Parameters}
+\label{app:contbeta}
+In the simulation study presented in Sections~\ref{sec:simulation} and \ref{sec:results} the intercept
+and slope parameters $\beta_0$ and $\beta_1$ are both either fixed or binary variables taking
+either the positive or the negative value of the effect size $\delta$. Therefore, these parameters
+are step functions for which MOB has shown to perform better than CTree. However, in case of
+monotonous functions CTree is advantageous as it is constructed to detect monotonous effects.
+To point out that the difference in performance of CTree and MOB depends on the type of effect
+the same setting of which the results are presented in Figure~\ref{fig:scores_stump} is
+evaluated again but this time with the varying parameter(s) changing continuously.
+In particular, the parameters $\beta_0$ and $\beta_1$ are linear functions of the true
+split variable $Z_1$:
+\begin{align*}
+\beta_{i}(Z_1) = \begin{cases}
+-\delta \cdot (-1)^{i} \cdot Z_1\\
++\delta \cdot (-1)^{i} \cdot Z_1.
+\end{cases}
+\end{align*}
+Moreover, a modified version of CTree considering maximum values in the test statistic is evaluated
+(denoted by CTree\_max). 
+Looking at the results presented in Figure~\ref{fig:stump_contbeta} it can be observed that in this setting the 
+original version of CTree is ahead while CTree\_max and MOB perform almost equally well. Hence, the comparison of 
+Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} points out that it clearly depends on the type of 
+effect whether the maximal selection (MOB, CTree\_max) or the linear selection (CTree) is advantageous.
+
+\begin{figure}%%[t]
+\setkeys{Gin}{width=\linewidth}
+<<contbeta, fig=TRUE, echo=FALSE, height=4, width=8>>=
+load("../sim/simres20181122_stump.rda")
+sim_stump_cont <- simres$res
+subdata <- subset(sim_stump_cont, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
+#subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"),
+                       labels = c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores"))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")], 
+                                             fill = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")]), 
+                     superpose.line = list(col = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")]))
+
+xyplot(prop_Tsplit ~ delta | vary_beta, groups = ~ test, 
+       data = subdata, 
+       subset = binary_beta == FALSE & binary_regressor == FALSE & 
+         only_intercept == FALSE, 
+       #xlim = c(0,0.1,0.5),
+       type = "l", 
+       layout= c(3,1),
+       lwd = 2,
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,1,1,1,1), 
+                                type = "l", 
+                                col = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")], 
+                                lwd = 2)), 
+       xlab = expression(delta),
+       ylab = expression("Proportion of selecting"~Z[1]),
+       index.cond = list(c(2,3,1)),
+       strip = strip.custom(factor.levels = c(expression("vary "~beta[0]~" and "~beta[1]),
+                                              expression("vary "~beta[0]), 
+                                              expression("vary "~beta[1])),
+                                    bg = "gray90"),
        par.settings = par.settings)
 @
-\caption{\label{fig:scores_stump_wosign}Proportion of cases where the true split variable $Z_1$ 
-is selected over increasing effect size $\delta$ for $\xi=0$, 250 observations and 
-100 data sets per step without any significance level. 
+\caption{\label{fig:stump_contbeta}
+Proportion of cases where the true split variable $Z_1$ is
+selected over increasing effect size $\delta$ (100 data sets of 250 observations per step)
+with the varying coefficient(s) being continuous. 
 Each panel represents one scenario for the coefficients: 
-varying intercept $\beta_0$ (left panel), varying slope parameter $\beta_1$ (middle panel) and 
-both coefficients varying (right panel). The performances of four testing strategies are compared: 
-CTree, MOB, GUIDE, GUIDE\_scores.}
+varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
+both coefficients varying (right). The performances of five testing strategies are compared: 
+CTree, CTree\_max, MOB, GUIDE, GUIDE\_scores.}
 \end{figure}
 
-\newpage
 \section{Results Over Increasing Effect Size}
@@
          only_intercept == FALSE & vary_beta == "all", 
-       type = "b",
+       type = "l",
        lty = c(1,2,#1,2,
                1,2,1),
+       lwd = 2,
        layout= c(4,1),
@@
                                         1,2,1), 
-                                type = "b", pch = 1, cex = 0.8,
+                                type = "l", pch = 1, cex = 0.8,
                                 col = pal[c("CTree", "CTree", 
@@
                                             "MOB", "MOB",
-                                            "GUIDE_scores")], lwd = 1)), 
+                                            "GUIDE_scores")], lwd = 2)), 
        #index.cond = list(c(1,2,3,4)),
@@
 \caption{\label{fig:bin_stump}Proportion of cases where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ for both parameter varying, 
-250 observations and 100 data sets per step. Each panel represents a different value of the 
+detected over increasing effect size $\delta$ (100 data sets of 250 observations per step)
+with both parameter varying. Each panel represents a different value of the true 
 split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
@@
        #xlim = c(0,0.1,0.5),
-       type = "b", 
+       type = "l", 
        lty = c(1,2,#1,2,
                1,2,1),
+       lwd = 2,
        layout= c(4,1),
@@
 \caption{\label{fig:cat_stump}Proportion of cases where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ for both parameters varying, $\sigma = 1$, 
-250 observations and 100 data sets per step. Each panel represents a different value of the 
-split point $\xi$ ($0$, $0.2$, $0.5$ and $0.8$ from left to right). The performances of five 
-testing strategies are compared: CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_scores.}
+detected over increasing effect size $\delta$ (100 data sets of 250 observations per step)
+with both parameter varying. Each panel represents a different value of the true 
+split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
+strategies are compared: CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_scores.}
 \end{figure}

### 2019-04-03 pkg/partykit/inst/pkgdown/glmtree-titanic.Rmd
started playing around with possible pkgdown materials -> in separate directory inst/pkgdown and .Rbuildignore for the moment
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2513 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 611bc8b35daac863e6f9ed1c29e2591b144e23a7

---
title: "Logistic Regression Tree: Titanic Survival Data"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: party.bib
vignette: >
  %\VignetteIndexEntry{Logistic Regression Tree: Titanic Survival Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{partykit}
  %\VignetteKeywords{GLM, tree, recursive partitioning, logit, Titanic}
  %\VignettePackage{partykit}
---

```{r preliminaries, echo = FALSE, message = FALSE}
library("partykit")
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
  dev = "png", dpi = 100, out.width = "100%",
  fig.height = 6, fig.width = 6, fig.align = "center", fig.show = "hold")
```

## Motivation: Women and children first?

To illustrate how differences in treatment effects can be captured by MOB [@Zeileis+Hothorn+Hornik:2008],
the Titanic survival data [see @Friendly+Symanzik+Onder:2019 for an overview]
is considered: The question is whether "women and children first"
is applied in the same way for all subgroups of the
passengers of the Titanic. Or, in other words, whether the effectiveness of
preferential treatment for women/children differed across subgroups.

The `Titanic` data is provided in base R as a contingency
table and transformed here to a `data.frame` for use with `glmtree()`:

```{r Titanic-data}
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
ttnc <- transform(ttnc, Treatment = factor(
  Gender == "Female" | Age == "Child", levels = c(FALSE, TRUE),
  labels = c("Male&Adult", "Female|Child")))
```

The resulting `ttnc` data provides factors `Survived` (yes/no), `Class` (1st,
2nd, 3rd, crew), `Gender` (male, female), and `Age` (child, adult).
Additionally, a factor `Treatment` is added that distinguishes women/children
from male adults.

## Treatment-subgroup interactions

To investigate how the preferential treatment effect (`Survived ~ Treatment`)
differs across the remaining explanatory variables, the following
logistic-regression-based tree is considered. The significance
level of `alpha = 0.01` is employed here to avoid overfitting and separation
problems in the logistic regression.

```{r Titanic-tree}
ttnc_tree <- glmtree(Survived ~ Treatment | Class + Gender + Age,
  data = ttnc, family = binomial, alpha = 0.01)
ttnc_tree
```

The corresponding tree can be visualized with `plot(ttnc_tree)` where
spinograms in the leaves show survival status versus preferential treatment
(women or children). This shows that the treatment differs strongly across passengers
classes. The plot with some small tweaks in the layout is given by:

```{r Titanic-plot, fig = TRUE, fig.height = 6, fig.width = 10}
plot(ttnc_tree, tp_args = list(ylines = 1, margins = c(1.5, 1.5, 1.5, 2.5)))
```

The treatment effect is much lower in the 3rd class where women/children still have
higher survival rates than adult men but the odds ratio is much lower compared to
all remaining classes. The split between the 2nd and the remaining two classes (1st, crew)
is due to a lower overall survival rate (intercepts of `r round(coef(ttnc_tree)[2, 1], digits = 2)` and 
`r round(coef(ttnc_tree)[3, 1], digits = 2)`, respectively) while the odds ratios
associated with the preferential treatment are rather similar
(`r round(coef(ttnc_tree)[2, 2], digits = 2)` and `r round(coef(ttnc_tree)[3, 2], digits = 2)`,
respectively).

Another option for assessing the class effect would be to immediately split into
all four classes rather than using recursive binary splits. This can be obtained
by setting `catsplit = "multiway"` in the `glmtree()` call above. This
yields a tree with just a single split but four kid nodes.

## References

### 2019-04-05 pkg/partykit/inst/guideliketest/draft/draft.Rnw
results of extended simulation for stump scenario and adaptions in loading the corresponding data in the draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2514 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4d4d6abb0b2f5796f73c3626d6d6dcf2f04838fe

@@
 <<contbeta, fig=TRUE, echo=FALSE, height=4, width=8>>=
-load("../sim/simres20181122_stump.rda")
-sim_stump_cont <- simres$res
-subdata <- subset(sim_stump_cont, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
 #subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
@@
          only_intercept == FALSE, 
-       #xlim = c(0,0.1,0.5),
+       scales=list(x=list(at=seq(1,11,2))),
        type = "l", 

### 2019-04-05 pkg/partykit/inst/guideliketest/draft/draft.Rnw
results of extended simulation for stump scenario and adaptions in loading the corresponding data in the draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2515 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4d579fb4e3e23f13f408842f240026319fd0b45b

@@
        data = d3way, 
-       ylim = c(0,0.5),
+       ylim = c(0,0.65),
        type = c("a"), 

### 2019-04-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
first draft of extended abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2516 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 82bb06a3b27bdc2fb6e10702f8102edf95e55ef2

@@
 
-%\section{Motivation}
-%Circular response variables occur in a variety of applications and subject areas. 
-%For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
-%animal orientation or gene-structure analysis are often subject of examination in biology, 
-%and wind direction is one of the most important weather variables in meteorology.
-%Representing such a two-dimensional direction by an angle in a unity circle requires 
-%the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
-%rotation. However, these settings can vary but should not influence statistical 
-%inference from a fitted model.
-%For that reason circular data demand for purpose-built statistical methods. 
+\section{Motivation}
+\textit{FIXME: should we include this section 'Motivation'?}
+
+Circular response variables occur in a variety of applications and subject areas. 
+For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
+animal orientation or gene-structure analysis are often subject of examination in biology, 
+and wind direction is one of the most important weather variables in meteorology.
+Representing such a two-dimensional direction by an angle in a unity circle requires 
+the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
+rotation. However, these settings can vary but should not influence statistical 
+inference from a fitted model.
+For that reason circular data demand for purpose-built statistical methods. 
 %A commonly chosen distribution for modeling circular data is the von Mises distribution.
@@
 
-Distributional trees~(Schlosser et.~al, 2018) fuse distributional regression modeling with
+Distributional trees~(Schlosser et.~al, 2019) fuse distributional regression modeling with
 regression trees based on the unbiased recursive partitioning algorithms MOB~(Zeileis et.~al, 2008)
@@
 
+%% exchange file name of figure: schlosser-circtree_plot.pdf -> schlosserfig1.pdf
 \begin{figure}[p!]\centering
@@
 
-A natural extension are ensembles (or forests) of such circular trees
-that can improve forecasts by regularizing and stabilizing the model
-(see Schlosser et.~al~(2018) for a general discussion of distributional forests).
-
-Another possible improvement for obtaining more parsimonious and more stable trees is to
-consider splitting circular covariates into two circle segments by searching two
-split points simultaneously (rather than sequentially at different depths as
-for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
-
 \textsf{R} packages implementing the proposed methods are under development at
 \texttt{https://R-Forge.R-project.org/projects/partykit/}.
+
+\subsection{Ensembles/Random Forests}
+A natural extension are ensembles (or forests) of such circular trees
+that can improve forecasts by regularizing and stabilizing the model.
+Random forests introduced by Breiman~et.~al~(1984) average the
+predictions of an ensemble of trees, each built on a subsample 
+or bootstrap of the original data. 
+A different strategy of combining trees is employed in distributional forests 
+(Schlosser~et.~al, 2019) which can also be seen as adaptive local likelihood models. 
+For a possibly new observation $x$ weights $w_i(x)$ are obtained by counting the 
+number of trees in which it is assigned to the same terminal node as a learning
+observation $y_i, i \in \{1,\ldots,n\}$.
+The parameters $\mu$ and $\kappa$ can then be estimated specifically for the (new) 
+observation $x$ by applying the weighted maximum likelihood estimator
+$$
+\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
+$$
+In that way predictions are made mainly based on similar observations rather than the 
+complete learning data.
+
+\subsection{Splits in Circular Variables}
+Another possible improvement for obtaining more parsimonious and more stable trees is to
+consider splitting circular covariates into two circle segments by searching two split 
+points simultaneously (rather than sequentially at different depths as for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
+In particular, the search for the best separation of the data space into two intervals is based on the optimization problem
+$$
+\max \sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)
+$$
+where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters for the corresponding subsets of the response as defined by the intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
+In case of a linear split variable these intervals can be defined by a single split point. 
+By treating a circular variable such as wind direction linearly as done in the presented 
+application the considered space~$[0,2\pi)$ is then split by a split point $\nu$ into
+$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing an unnatural separation at $0$. 
+Moreover, further splits are often required to obtain properly fitting subintervals of~$[0,2\pi)$.
+This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
+process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$.
+The required test statistic to maximally select two split points simultaneously is readily 
+available in the \textsf{R} package \texttt{coin} and could be enhanced as proposed by Hothorn 
+and Zeileis~(2008). For a score-based approach the general strategy of double maximum tests 
+introduced by Bai and Perron~(2003) can be considered. 
+Hence, the splitting idea can be extended naturally to a two point-search, however, for 
+an unbiased and inference based selection the corresponding tests employing a proper
+asymptotic null distribution still need to be implemented.
 
@@
 
-%\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
-%     Classification and Regression Trees.
-%     {\it Wadsworth}.
+\item[Bai, J., and Perron, P.] (2003).
+     Computation and Analysis of Multiple Structural Change Models.
+     {\it Journal of Applied Econometrics}, {\bf 18},
+     1\,--\,22.
+
+\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
+     Classification and Regression Trees.
+     {\it Wadsworth}.
 
@@
      651\,--\,674.
+
+\item[Hothorn, T., and Zeileis, A.] (2008).
+     Generalized Maximally Selected Statistics.
+     {\it Biometrics}, {\bf 64}, 4,
+     1263\,--\,1269.
 
@@
 
-\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
+\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2019).
      Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.

### 2019-04-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
some improvements in abstract/intro, plus various small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2518 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 248811aa6e84cef6892744882d16aeb2f650775e

@@
 of the best splitting variable from the available covariates and the corresponding 
-split point. Early tree algorithms (e.g., AID, CART) employ greedy search strategies 
+split point. Early tree algorithms (e.g., AID, CART) employed greedy search strategies 
 that directly compare all possible split points in all available covariates. However, 
@@
 
-Therefore, three different approaches are considered here and embedded into a common 
-modeling framework with special emphasis to linear model trees: classical categorical 
-association tests (GUIDE), conditional inference (CTree), parameter instability 
-tests (MOB). It is assessed how different building blocks affect the power of the 
-tree algorithms to select the appropriate covariates for splitting: residuals vs.\ full 
-model scores, binarization of residuals/scores at zero, binning of covariates, 
-conditional vs.\ unconditional approximations of the null distribution.
+Therefore, three different popular approaches are considered here: classical categorical 
+association tests (GUIDE), conditional inference (CTree), and parameter instability 
+tests (MOB). First, these are embedded into a common inference framework encompassing
+parametric model trees, in particular linear model trees. Second, it is assessed how
+different building blocks from this common framework affect the power of the 
+tree algorithms to select the appropriate covariates for splitting: observation-wise
+goodness-of-fit measure (residuals vs.\ model scores), binarization of residuals/scores
+at zero, binning of covariates, conditional vs.\ unconditional approximations of the
+null distribution. This shows that specifically the goodness-of-fit measure is crucial
+for the power of the procedures, with model scores without binarization performing
+much better in many scenarios.
 }
@@
 each of the resulting subgroups often leads to better results. This separation can be 
-done by applying a tree algorithm. While all of the many developed algorithms 
+done by applying a tree algorithm. While almost all algorithms proposed in the literature
 follow the general idea of splitting the data such that some objective function is optimized, 
@@
 choosing the covariate which shows the highest association to the response variable 
-based on $p$-values provided by a statistical test. While QUEST, CHAID and GUIDE employ 
+based on $p$-values provided by a statistical test. While QUEST, CHAID, and GUIDE employ 
 statistical significance tests for contingency tables, CTree applies permutation tests in 
-a conditional inference framework and MOB uses fluctuation tests based on asymptotic theory. 
-All these approaches have shown to work well for various situations, however, the specific 
-advantages and disadvantages of each of the testing strategies have not yet been 
-investigated and compared in detail.
+a conditional inference framework and MOB uses fluctuation tests based on central limit theorems
+for the parameter estimators. 
+All these approaches have shown to work well for various situations, however, the relative
+(dis)advantages of the testing strategies have not yet been investigated and compared in detail.
 
@@
 by investigating the building blocks of their testing strategies in which they differ:
-(1)~Transformations of the response: using residuals or full model scores
-(2)~Binarization of residuals/scores
+(1)~Transformations of a goodness-of-fit measure for the response: using residuals or full model scores.
+(2)~Binarization of these residuals or scores.
 (3)~Categorization of covariates.
+(4)~Approximations of the null distribution: conditional vs.\ unconditional.
 For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
-algorithms is presented such that each of the three strategies GUIDE, CTree, and MOB can be 
-obtained by a specific combination of the available building blocks. In that way, the 
-effects of each of these features can be investigated separately by varying the building 
-blocks in the tool box provided by this general framework. 
-Additionally, the impact of applying either a pre- or postpruning strategy is analyzed.
+algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
+obtained by a specific combination of the available building blocks. This allows to systematically
+vary the building blocks and assess the power of the resulting inference procedure. Moreover,
+it is investigated whether the performance of the inference impacts the performance of the
+trees differently under pre- vs.\ post-pruning.
 
@@
 
-In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree model is given. 
-After explaining the investigated algorithms CTree (Section~\ref{sec:ctree}), 
-MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}) more in detail the applied pruning strategies 
+The remainder of this manuscript is structured as follows:
+Section~\ref{sec:urp} reviews unbiased tree models, starting with the general algorithmic idea
+(Section~\ref{sec:generic_alg}) followed by the specific algorithms CTree (Section~\ref{sec:ctree}), 
+MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}), before possible pruning strategies 
 (\ref{sec:pruning}) are discussed.
@@
 
-\section{Unbiased Recursive Partitioning}
+\section{Unbiased recursive partitioning}
 \label{sec:urp}
 
-\subsection{Generic Algorithm}
+\subsection{Generic algorithm}
 \label{sec:generic_alg}
@@
 \begin{enumerate}
-\item For the whole underlying data set in the current node some form of discrepancy 
+\item For the data set in the current node some form of discrepancy 
 measure is calculated. 
@@
 \item Steps 1 and 2 are repeated within each new node until either a stopping criterion
-is reached or the data in the node consists of only one or all equal observations. 
+is reached. 
 \end{enumerate}
@@
 
-\section{Unifying Framework}\label{sec:unify}
+\section{Unifying framework}\label{sec:unify}
 Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
@@
 
-\begin{table}[t]
+\begin{table}[t!]
 \centering
-\caption[Table caption text]{Combinations of fitted model type, test type and pruning strategy
-for the algorithms CTree, MOB, and GUIDE.}
-\label{tab:combinations}
 \begin{tabular}{ l l l l }
@@
 \hline
-CTree & non-parametric  & conditional inference    & pre\\
-MOB   & parametric      & score-based fluctuation  & pre (or post with AIC/BIC)\\
-GUIDE & parametric      & residual-based $\chi^2$  & post (cost-complexity pruning)\\
+CTree & Non-parametric  & Conditional inference    & Pre\\
+MOB   & Parametric      & Score-based fluctuation  & Pre (or post with AIC/BIC)\\
+GUIDE & Parametric      & Residual-based $\chi^2$  & Post (cost-complexity pruning)\\
 \hline
 \end{tabular}
+\caption{Combinations of fitted model type, test type and pruning strategy
+for the algorithms CTree, MOB, and GUIDE.
+\label{tab:combinations}}
 \end{table}
 
-\subsection{Unification of Testing Strategies}
+\subsection{Unification of testing strategies}
 
@@
 
-\subsection{Building Blocks of Testing Strategies}\label{sec:buildingblocks}
+\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
 The three following building blocks all refer to the way how the given data is prepared/transformed
@@
 
-\begin{table}[t]
+\begin{table}[t!]
 \centering
-\caption[Table caption text]{Testing strategies of CTree, MOB, and GUIDE with the corresponding
-setting of the building blocks in the unifying framework and the type of test statistic.}
-\label{tab:buildingblocks}
-\begin{tabular}{ l c c c c }
+\begin{tabular}{ l l l l l }
 \hline 
@@
 \end{tabular}
+\caption{Testing strategies of CTree, MOB, and GUIDE with the corresponding
+setting of the building blocks in the unifying framework and the type of test statistic.
+\label{tab:buildingblocks}}
 \end{table}
@@
 
-\subsection{Linear Model Tree}\label{sec:lmtrees}
+\subsection{Linear model tree}\label{sec:lmtrees}
 As an example a linear model tree is considered as this allows for an easy and well structured 
@@
 
-\section{Simulation Setting and Evaluation}
+\section{Simulation setting and evaluation}
 \label{sec:simulation}
@@
 
-\subsection{Data Generating Process}
+\subsection{Data generating process}
 Each generated data set consists of a response variable, one regressor, one split 
@@
 \begin{table}
-\begin{center}
-\caption[Table caption text]{Variables included in the data generating process of 
-the first scenario (stump).}
-\label{tab:variables}
+\centering
 \begin{tabular}{l l l}
@@
 \end{tabular}
-\end{center}
+\caption{Variables included in the data generating process of 
+the first scenario (stump).
+\label{tab:variables}}
 \end{table}
@@
 
-%\newpage
 \section{Results}
@@
 
-\subsection{Residuals vs. Full Model Scores}
+\subsection[Residuals vs. full model scores]{Residuals vs.\ full model scores}
 \label{sec:scores}
@@
 and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_scores).
-%Thus, the four strategies are obtained from the unifying framework by setting the switches 
-%as listed in Table~\ref{tab:buildingblocks_scores}.
-
-%\begin{table}%[t]
-%\centering
-%\caption[Table caption text]{CTree, MOB, GUIDE, and GUIDE\_scores with the corresponding setting of
-%the switches in the unifying framework and the approximation type of the underlying null 
-%distribution.}
-%\label{tab:buildingblocks_scores}
-%\begin{tabular}{ l c c c c }
-%\hline 
-%            & Scores            & Binarization  & Categorization  & Null distribution \\ \hline
-%CTree       & full model scores & --          & --              & conditional\\
-%MOB         & full model scores & --          & --              & unconditional\\
-%GUIDE       & residuals         & \checkmark  & \checkmark      & unconditional\\
-%GUIDE\_scores & full model scores & \checkmark  & \checkmark      & unconditional\\ \hline
-%\end{tabular}
-%\end{table}
 
@@
 
-\subsection{Full Factorial Analysis}\label{sec:3way}
+\subsection{Full factorial analysis}\label{sec:3way}
 In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
@@
 \centering
-\caption[Table caption text]{Available levels of the building blocks.}
-\label{tab:levels}
 \begin{tabular}{ l l l }
@@
 \hline
-Residuals vs.\ Scores  & \textit{residuals}  & use only residuals\\
-                       & \textit{scores}     & use full score matrix\\
-Categorization         & \textit{cat}        & categorize split variables\\
-                       & \textit{max}        & selected split variables maximally w.o. categorization\\
-                       & \textit{lin}        & selected split variables linearly w.o. categorization\\
-Binarization           & \textit{bin}        & binarize scores\\
-                       & \textit{lin}        & use residuals/score values without binarization\\
+Residuals vs.\ scores  & \textit{residuals}  & Use only residuals\\
+                       & \textit{scores}     & Use full score matrix\\
+Categorization         & \textit{cat}        & Categorize split variables\\
+                       & \textit{max}        & Selected split variables maximally w.o. categorization\\
+                       & \textit{lin}        & Selected split variables linearly w.o. categorization\\
+Binarization           & \textit{bin}        & Binarize scores\\
+                       & \textit{lin}        & Use residuals/score values without binarization\\
 \hline
 \end{tabular}
+\caption{Available levels of the building blocks.
+\label{tab:levels}}
 \end{table}
@@
 
-\subsection{Prepruning vs. Postpruning}
+\subsection[Pre-pruning vs. post-pruning]{Pre-pruning vs.\ post-pruning}
 
@@
 
-\section*{Computational Details}
+\section*{Computational details}
 \label{sec:comp}
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 \hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
@@
 
-\section{Applied Combinations of Building Blocks}\label{app:combinations}
+\section{Combinations of building blocks applied}\label{app:combinations}
 For the evaluation of all 12 possible combinations of the presented building blocks
@@
 
-\begin{table}%[t]
+\begin{table}[t!]
 \centering
-\caption[Table caption text]{Applied testing strategies representing the 12 combinations
-of building blocks.}
-\label{tab:combinations_appendix}
-\begin{tabular}{ l c c c }
+\begin{tabular}{ l l l l }
 \hline 
@@
 \end{tabular}
+\caption{Testing strategies applied (representing the 12 combinations
+of building blocks).
+\label{tab:combinations_appendix}}
 \end{table}
 
-\section{Significance Level}\label{app:significance}
+\section{Significance level}\label{app:significance}
 In Section~\ref{sec:scores} the investigated testing strategies are compared based
@@
 
-\section{Simulation Results for Continuous Model Parameters}
+\section{Simulation results for continuous model parameters}
 \label{app:contbeta}
@@
 
-\section{Results Over Increasing Effect Size}
+\section{Results for increasing effect size}
 \label{app:increasingdelta}
 
-\subsection{Binarization of Residuals/Scores}
+\subsection{Binarization of residuals/scores}
 \label{app:binarization}
@@
 %as listed in Table~\ref{tab:buildingblocks_appendix}.
-
-%\begin{table}%[t]
-%\centering
-%\caption[Table caption text]{CTree, CTree\_bin, MOB, MOB\_bin, and GUIDE\_scores with the 
-%corresponding setting of the switches in the unifying framework and the approximation type 
-%of the underlying null distribution.}
-%\label{tab:buildingblocks_appendix}
-%\begin{tabular}{ l c c c c }
-%\hline 
-%& Scores            & Binarization  & Categorization  & Null distribution \\ 
-%\hline
-%CTree       & full model scores & --          & --            & conditional\\
-%CTree\_bin  & full model scores & \checkmark  & --            & conditional\\
-%MOB         & full model scores & --          & --            & unconditional\\
-%MOB\_bin    & full model scores & \checkmark  & --            & unconditional\\
-%GUIDE\_scores & full model scores & \checkmark  & \checkmark    & unconditional\\
-%\hline
-%\end{tabular}
-%\end{table}
 
@@
 
-\subsection{Categorization of Splitting Variables}
+\subsection{Categorization of splitting variables}
 \label{app:categorization}
@@
 Moreover, they are also compared to GUIDE\_scores which includes all available scores.
-%The corresponding setting of the switches in the unifying framework is listed in 
-%Table~\ref{tab:buildingblocks_cat}.
-
-%\begin{table}%[t]
-%\centering
-%\caption[Table caption text]{CTree, CTree\_cat, MOB, MOB\_cat, and GUIDE\_scores with the 
-%corresponding setting of the switches in the unifying framework and the approximation type 
-%of the underlying null distribution.}
-%\label{tab:buildingblocks_cat}
-%\begin{tabular}{ l c c c c }
-%\hline 
-%& Scores            & Binarization  & Categorization  & Null distribution \\ 
-%\hline
-%CTree       & full model scores & --          & --              & conditional\\
-%CTree\_cat  & full model scores & --          & \checkmark      & conditional\\
-%MOB         & full model scores & --          &                 & unconditional\\
-%MOB\_cat    & full model scores & --          & \checkmark      & unconditional\\
-%GUIDE\_scores & full model scores & \checkmark  & \checkmark      & unconditional\\
-%\hline
-%\end{tabular}
-%\end{table}
 

### 2019-04-08 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
further revisions on extended abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2520 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 96a3aaedbf6b3c4a4f08c16cc4e3c29aead080ab

--- 2019-02-07 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
@@
 
-%\section{Motivation}
-%Circular response variables occur in a variety of applications and subject areas. 
-%For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
-%animal orientation or gene-structure analysis are often subject of examination in biology, 
-%and wind direction is one of the most important weather variables in meteorology.
+\section{Motivation}
+%\textit{FIXME: should we include this section 'Motivation'?}
+
+Circular data can be found in a variety of applications and subject areas, 
+e.g., hourly crime rate in the social-economics, 
+animal movement direction or gene-structure in biology, 
+and wind direction as one of the most important weather variables in meteorology.
+Circular regression models were first introduced by Fisher and Lee (1992) and 
+further extended by Jammalamadaka and Sengupta (2001) and 
+Mulder and Klugkist (2017) among others.
+While most of these approaches are built on additive regression models, 
+we propose an adaption/extension of regression trees to circular data by employing 
+distributional trees.
+
 %Representing such a two-dimensional direction by an angle in a unity circle requires 
@@
 
+\subsection{Ensembles/Random Forests}
 A natural extension are ensembles (or forests) of such circular trees
-that can improve forecasts by regularizing and stabilizing the model
-(see Schlosser et.~al~(2018) for a general discussion of distributional forests).
+that can improve forecasts by regularizing and stabilizing the model.
+Random forests introduced by Breiman~et.~al~(1984) average the
+predictions of an ensemble of trees, each built on a subsample 
+or bootstrap of the original data. 
+A different strategy of combining trees is employed in distributional forests 
+(Schlosser~et.~al, 2019) which can also be seen as adaptive local likelihood models. 
+For a possibly new observation $x$ weights $w_i(x)$ are obtained by counting the 
+number of trees in which it is assigned to the same terminal node as a learning
+observation $y_i, i \in \{1,\ldots,n\}$.
+The parameters $\mu$ and $\kappa$ are then estimated specifically for the (new) 
+observation $x$ by applying the weighted maximum likelihood estimator
+$$
+\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
+$$
+In that way predictions are mainly based on similar observations rather than the 
+complete learning data.
 
-Another possible improvement for obtaining more parsimonious and more stable trees is to
-consider splitting circular covariates into two circle segments by searching two
-split points simultaneously (rather than sequentially at different depths as
-for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
+\subsection{Splits in Circular Variables}
+Another planned extension of the introduced circular distributional trees 
+for obtaining more parsimonious and more stable trees is to
+consider splitting circular covariates into two flexible circle segments by searching two split 
+points simultaneously.
+In general, the search for the best separation of the data space into two intervals is based on the optimization problem
+$$
+\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
+$$
+where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters for the corresponding subsets of the response as defined by the intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
+For a linear split variable these intervals are defined by a single split point.
+By treating a circular variable such as wind direction linearly
+this single split point $\nu$ devides the considered space~$[0,2\pi)$ into
+$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing a potentially unnatural 
+separation at zero. 
+Moreover, further sequential splits in the circular variable are often required to obtain 
+data driven subintervals of~$[0,2\pi)$ (as seen for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
+This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
+process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$
+where both interval borders are determined purely by the data.
 
-\textsf{R} packages implementing the proposed methods are under development at
-\texttt{https://R-Forge.R-project.org/projects/partykit/}.
+The required test statistic to maximally select two split points simultaneously is readily 
+available in the \textsf{R} package \texttt{coin} but would need to be further enhanced as proposed 
+by Hothorn and Zeileis~(2008). For a score-based approach the general strategy of double maximum tests introduced by Bai and Perron~(2003) can be considered. 
+Hence, the splitting idea can be naturally extended to a two-point search, however, for 
+an unbiased and inference based selection the corresponding tests employing a proper
+asymptotic null distribution still need to be implemented.
 
 \bigskip
+
+\subsubsection*{Computational Details:}
+\textsf{R} packages implementing the proposed methods are currently under development at
+\texttt{https://R-Forge.R-project.org/} \texttt{projects/partykit/}.
 
@@
 
-%\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
-%     Classification and Regression Trees.
-%     {\it Wadsworth}.
+\item[Bai, J., and Perron, P.] (2003).
+     Computation and Analysis of Multiple Structural Change Models.
+     {\it Journal of Applied Econometrics}, {\bf 18},
+     1\,--\,22.
+
+\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
+     Classification and Regression Trees.
+     {\it Wadsworth}.
+
+\item[Fisher, N. I., and Lee, A. J.] (1992).
+     Regression Models for an Angular Response.
+     {\it Biometrics}, {\bf 48}, 3,
+     665\,--\,677.
 
@@
 
+\item[Hothorn, T., and Zeileis, A.] (2008).
+     Generalized Maximally Selected Statistics.
+     {\it Biometrics}, {\bf 64}, 4,
+     1263\,--\,1269.
+
 \item[Jammalamadaka, S. R. and Sengupta, A.] (2001).
@@
      World Scientific.
+
+\item[Mulder, K., and Klugkist, I.] (2017).
+     Bayesian Estimation and Hypothesis Tests for a Circular Generalized Linear Model.
+     {\it Journal of Mathematical Psychology}, {\bf 80},
+     4\,--\,14.
 

### 2019-04-09 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
further revisions on extended abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2521 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e1ab82ca1c016aa6e46d2c39bf7127efb1bfb047

@@
 Mulder and Klugkist (2017) among others.
-While most of these approaches are built on additive regression models, 
+While most of the already existing approaches are built on additive regression models, 
 we propose an adaption/extension of regression trees to circular data by employing 
@@
 
-\subsection{Ensembles/Random Forests}
+\subsection{Ensembles/Random forests}
 A natural extension are ensembles (or forests) of such circular trees
@@
 A different strategy of combining trees is employed in distributional forests 
-(Schlosser~et.~al, 2019) which can also be seen as adaptive local likelihood models. 
-For a possibly new observation $x$ weights $w_i(x)$ are obtained by counting the 
-number of trees in which it is assigned to the same terminal node as a learning
-observation $y_i, i \in \{1,\ldots,n\}$.
+(Schlosser~et.~al, 2019) which can also be interpreted as adaptive local likelihood models. 
+For a possibly new observation $x$ averaged nearest neighbor weights $w_i(x)$ are obtained 
+by counting the number of trees in which it is assigned to the same terminal node as a 
+learning observation $y_i, i \in \{1,\ldots,n\}$ divided by the total number of trees.
 The parameters $\mu$ and $\kappa$ are then estimated specifically for the (new) 
 observation $x$ by applying the weighted maximum likelihood estimator
-$$
+\begin{equation}
 \operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
-$$
+\end{equation}
 In that way predictions are mainly based on similar observations rather than the 
@@
 
-\subsection{Splits in Circular Variables}
-Another planned extension of the introduced circular distributional trees 
-for obtaining more parsimonious and more stable trees is to
-consider splitting circular covariates into two flexible circle segments by searching two split 
-points simultaneously.
-In general, the search for the best separation of the data space into two intervals is based on the optimization problem
-$$
+\subsection{Splits in circular variables}
+In order to obtain more parsimonious and more stable trees another possible extension of the 
+introduced circular distributional trees is to split circular covariates into two 
+flexible circle segments by searching two split points simultaneously.
+In general, the search for the best separation of the data space into two intervals is based 
+on the optimization problem
+\begin{equation}
 \max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
-$$
-where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters for the corresponding subsets of the response as defined by the intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
+\end{equation}
+where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
+of the von Mises distribution for the corresponding subsets of the response as defined by the
+intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
 For a linear split variable these intervals are defined by a single split point.
 By treating a circular variable such as wind direction linearly
-this single split point $\nu$ devides the considered space~$[0,2\pi)$ into
+this single split point $\nu$ divides the considered space~$[0,2\pi)$ into
 $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing a potentially unnatural 

### 2019-04-09 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
small touch-ups in Sections 1-3
some reformulation in Section 4.1

some more reformulation Section 4.2 with "covariate" rather than
"variable", emphsizing circular _covariate_ (as opposed to _response_),
and some details re inference


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2522 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 29b93241118b95a8e23f4593b82b10c4ea6f687d

@@
 %%%%%%%% IF DEPARTMENTS ARE LISTED AS WELL CHANGE NUMBERING!!!  %%%%%%%%%%%%%%
-\institute{Department of Statistics, University of Innsbruck, Innsbruck, Austria
-\and Department of Atmospheric and Cryospheric Science, University of Innsbruck, Innsbruck, Austria
-\and Epidemiology, Biostatistics and Prevention Institute, University of Zurich, Zurich, Switzerland}
+\institute{Department of Statistics, Universit\"at Innsbruck, Innsbruck, Austria
+\and Department of Atmospheric and Cryospheric Science, Universit\"at Innsbruck, Innsbruck, Austria
+\and Epidemiology, Biostatistics and Prevention Institute, Universit\"at Z\"urich, Z\"urich, Switzerland}
 
@@
 where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
-(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).
+(see, e.g., Jammalamadaka and Sengupta 2001, for a more detailed overview).
 
@@
 
-\subsection{Ensembles/Random forests}
-A natural extension are ensembles (or forests) of such circular trees
+\subsection{Ensembles and random forests}
+A natural extension are ensembles or forests of such circular trees
 that can improve forecasts by regularizing and stabilizing the model.
@@
 or bootstrap of the original data. 
-A different strategy of combining trees is employed in distributional forests 
-(Schlosser~et.~al, 2019) which can also be interpreted as adaptive local likelihood models. 
-For a possibly new observation $x$ averaged nearest neighbor weights $w_i(x)$ are obtained 
-by counting the number of trees in which it is assigned to the same terminal node as a 
-learning observation $y_i, i \in \{1,\ldots,n\}$ divided by the total number of trees.
-The parameters $\mu$ and $\kappa$ are then estimated specifically for the (new) 
-observation $x$ by applying the weighted maximum likelihood estimator
+A generalization of this strategy is to obtain weighted predictions
+by adaptive local likelihood estimation of the distributional parameters
+(Schlosser~et.~al, 2019). More specificially, for each possibly new
+observation $x$ a set of ``nearest neighbor'' weights $w_i(x)$ is obtained 
+that is based on how often $x$ is assigned to the same terminal node as
+each learning observation $y_i, i \in \{1,\ldots,n\}$.
+
+The parameters $\mu$ and $\kappa$ are then estimated for each (new) 
+observation $x$ by weighted maximum likelihood estimator based on the
+adaptive nearest neighbor weights:
 \begin{equation}
@@
 \end{equation}
-In that way predictions are mainly based on similar observations rather than the 
-complete learning data.
+Therefore, the resulting parameter estimates can smoothly adapt to the given
+covariates $x$ whereas $w_i(x) = 1$ would correspond to the unweighted
+full-sample estimates and $w_i(x) \in \{0, 1\}$ corresponds to the abrupt
+splits from the tree.
 
-\subsection{Splits in circular variables}
-In order to obtain more parsimonious and more stable trees another possible extension of the 
-introduced circular distributional trees is to split circular covariates into two 
-flexible circle segments by searching two split points simultaneously.
-In general, the search for the best separation of the data space into two intervals is based 
-on the optimization problem
+\subsection{Splits in circular covariates}
+
+In order to obtain more parsimonious and more stable trees another possible
+extension for \emph{circular covariates} (with or without a \emph{circular response})
+is to consider their circular nature when searching the best split into two segments.
+In general, searching the best separation of a covariate into a ``left'' and ``right''
+daugther node tries to maximize the segmented log-likelihood:
 \begin{equation}
@@
 where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
-of the von Mises distribution for the corresponding subsets of the response as defined by the
-intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
-For a linear split variable these intervals are defined by a single split point.
-By treating a circular variable such as wind direction linearly
-this single split point $\nu$ divides the considered space~$[0,2\pi)$ into
-$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing a potentially unnatural 
-separation at zero. 
-Moreover, further sequential splits in the circular variable are often required to obtain 
-data driven subintervals of~$[0,2\pi)$ (as seen for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
-This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
-process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$
-where both interval borders are determined purely by the data.
+of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a circular covariate $\in [0, 2 \pi)$
+only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
+thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
+for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
+and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$, encompassing zero. The latter
+strategy is invariant to the (often arbitrary) definition of the direction at zero.
 
-The required test statistic to maximally select two split points simultaneously is readily 
-available in the \textsf{R} package \texttt{coin} but would need to be further enhanced as proposed 
-by Hothorn and Zeileis~(2008). For a score-based approach the general strategy of double maximum tests introduced by Bai and Perron~(2003) can be considered. 
+When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
+far away, a simple linear split typically suffices to capture such a split (as seen for the
+lagged wind direction in Figure~\ref{schlosser:fig_tree}). If both $\nu$ and $\tau$ differ
+clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
+fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
+might miss such a pattern.
+
+The required test statistic to maximally select two split points simultaneously is straightforward
+to accomodate in the CTree framework by providing all binary indicators corresponding
+to the splints into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
+increasingly slow for larger sample sizes but it might be possible to speed up computations by
+exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
+MOB framework the extension is not quite as straightforward but one strategy could be to
+adapt double maximum tests \`a la Bai and Perron~(2003).
+
 Hence, the splitting idea can be naturally extended to a two-point search, however, for 
-an unbiased and inference based selection the corresponding tests employing a proper
-asymptotic null distribution still need to be implemented.
+an unbiased and inference-based selection the corresponding testing strategies might need
+further adaption.
 

### 2019-04-09 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
note on scale of wind direction and correction of typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2523 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 11b45c1d109939aab524a814cff592115f89740f

@@
 While most of the already existing approaches are built on additive regression models, 
-we propose an adaption/extension of regression trees to circular data by employing 
+we propose an adaption of regression trees to circular data by employing 
 distributional trees.
@@
 such as wind direction and wind speed at Innsbruck Airport.
+Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
 
@@
 
-\section{Discussion and Outlook}
+\section{Discussion and outlook}
 
@@
 by adaptive local likelihood estimation of the distributional parameters
-(Schlosser~et.~al, 2019). More specificially, for each possibly new
-observation $x$ a set of ``nearest neighbor'' weights $w_i(x)$ is obtained 
+(Schlosser~et.~al, 2019). More specifically, for each possibly new
+observation~$x$ a set of ``nearest neighbor'' weights $w_i(x)$ is obtained 
 that is based on how often $x$ is assigned to the same terminal node as
@@
 The parameters $\mu$ and $\kappa$ are then estimated for each (new) 
-observation $x$ by weighted maximum likelihood estimator based on the
+observation $x$ by weighted maximum likelihood based on the
 adaptive nearest neighbor weights:
@@
 In general, searching the best separation of a covariate into a ``left'' and ``right''
-daugther node tries to maximize the segmented log-likelihood:
+daughter node tries to maximize the segmented log-likelihood:
 \begin{equation}
@@
 The required test statistic to maximally select two split points simultaneously is straightforward
-to accomodate in the CTree framework by providing all binary indicators corresponding
-to the splints into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
+to accommodate in the CTree framework by providing all binary indicators corresponding
+to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
 increasingly slow for larger sample sizes but it might be possible to speed up computations by
@@
 
-\subsubsection*{Computational Details:}
+\subsubsection*{Computational details:}
 \textsf{R} packages implementing the proposed methods are currently under development at

### 2019-04-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions on section 2, notation of goodness-of-fit measure and other small corrections (including typos in the references)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2524 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3115ec7f726a60cbd05c39501ecca12baee96081

@@
   simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
-                         nrsplitvar = 1, nrsteps = 1, nrlevels = 2,
+                         nrsplitvar = 1, 
+                         nrsteps = 1, nrlevels = 2,
                          delta = seq(from = 0, to = 1, by = 0.1),
-                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = c("all", "beta0", "beta1"),
-                         binary_regressor = FALSE, binary_beta = TRUE, only_intercept = FALSE,
+                         xi = c(0, 0.2, 0.5, 0.8), 
+                         vary_beta = c("all", "beta0", "beta1"),
+                         binary_regressor = FALSE,
+                         binary_beta = c(TRUE, FALSE),
+                         only_intercept = FALSE,
                          test = c("ctree", "mfluc", "ctree_max",
@@
 goodness-of-fit measure (residuals vs.\ model scores), binarization of residuals/scores
-at zero, binning of covariates, conditional vs.\ unconditional approximations of the
-null distribution. This shows that specifically the goodness-of-fit measure is crucial
-for the power of the procedures, with model scores without binarization performing
+at zero, binning of covariates. This shows that specifically the goodness-of-fit measure 
+is crucial for the power of the procedures, with model scores without binarization performing
 much better in many scenarios.
@@
 by investigating the building blocks of their testing strategies in which they differ:
-(1)~Transformations of a goodness-of-fit measure for the response: using residuals or full model scores.
+(1)~Variations of a goodness-of-fit measure for the response: using residuals or full model scores.
 (2)~Binarization of these residuals or scores.
 (3)~Categorization of covariates.
-(4)~Approximations of the null distribution: conditional vs.\ unconditional.
+Apart from these three main factors further aspects such as the approximation of the null
+distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
+selected vs.\ sum of squares) will be considered as well.
 For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
@@
 
-Depending on the setup different effects can be observed for the investigated building blocks.
-However, none of the results contradicts the conclusion that in general using all available 
-model scores is at least as efficient but mostly clearly advantageous over considering only 
-residuals. Also regarding the effects of binarizing residuals or scores at $0$ or categorizing
-the values of possible split variables clear recommendations can be given based on the presented 
-results.
+In many of the considered scenarios it can be observed that the choice of goodness-of-fit
+measure heavily influences the performance of testing strategies as overall using model 
+scores leads to clearly better results than employing residuals only. Moreover, the original
+values of the goodness-of-fit measure are to be preferred over binarized versions of them.
+Also regarding the effects of categorizing possible split variables and the selection
+of a pruning strategy clear recommendations can be given based on the presented results.
 
@@
 \begin{enumerate}
-\item For the data set in the current node some form of discrepancy 
-measure is calculated. 
-\item Among all available covariates one is selected as split variable and a value 
-within the range/measurable space of this variable is chosen as split point such that 
-after splitting the data into subgroups the discrepancy measure is minimal within the 
-resulting nodes.
-\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
-is reached. 
+\item For the data set in the current node a model is fitted by optimizing
+some objective function that reflects the goodness of fit.
+\item Among all available covariates one is selected as the split variable,
+choosing the split point such that goodness of fit is maximized in the
+resulting subgroups.
+\item Steps 1 and 2 are repeated within each subgroup until some stopping
+criterion is attained.
 \end{enumerate}
 
-The employed discrepancy measure can for example be the sum of deviations from a typical 
-or average value, but it can also be constructed based on a parametric model fitted in each 
-node. For instance, different types of residuals or their signs can be applied 
-\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores as 
-in \cite{Hothorn+Hornik+Wiel:2006}.
-%(In the simulation study of this paper trees with linear models fitted in each node are 
-%investigated with the residual sum of squares (RSS) employed as discrepancy measure.)
-
-As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
-GUIDE first select the split variable and then, in a separate step, the split point. 
-For the first step of selecting a split variable they all apply statistical tests
-following the same general strategy: 
+The term ``model'' is used here in a very broad sense and encompasses not
+only least-squares or maximum-likelihood models but also simple constant
+fits such as means or average proportions.
+Therefore, depending on the type of model, the employed objective function can 
+for example be the sum of deviations from a typical or average value, but it can 
+also be constructed based on the model fitted in each node and depend not only on 
+the response but also on possible covariates. For instance, different 
+types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
+well as rank sums or logrank scores as in \cite{Hothorn+Hornik+VanDeWiel:2006}.
+
+As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, 
+MOB, and GUIDE first select the split variable and then, in a separate step, the 
+split point. For the first step of selecting a split variable they all apply statistical 
+tests following the same general strategy: 
 \begin{enumerate}
-\item A discrepancy measure is obtained in the form of an $n \times k$-matrix $D$ giving information
-on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
-current node and the parameter $k \in \mathbb{N}$ defined by the algorithm.
-\item Using this matrix the algorithm-specific test statistic is evaluated for each possible 
-split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
-of $D$ on $Z_j$.
-\item For the resulting values of the test statistics the corresponding $p$-values are calculated
-in order to allow for their comparison on a standardized/unified scale and in that way for
-an unbiased variable selection.
-\item The variable corresponding to the smallest $p$-value is selected as split variable as 
-it has shown to be the variable with the strongest influence on the goodness of fit of the model.
+\item To capture how the objective function changes with the observations in
+the current subgroup, a disaggregated, observation-wise goodness-of-fit
+measure is obtained. More formally, this is an $n \times k$ matrix $G$
+where $n \in \mathbb{N}$ is the number of observations and $k \in \mathbb{N}$ 
+the number of measures per observation.
+\item The dependency or association of this goodness-of-fit matrix $G$ with
+each possible split variable $Z_j$, $j \in \{1, \dots, m\}$, is assessed
+using some suitable test statistic. The corresponding $p$-values allow for
+a comparison of all $m$ covariates on a standardized or unified scale and in
+that way for an unbiased split variable selection.
+\item The covariate corresponding to the smallest $p$-value -- and thus the
+highest influence on the goodness of fit of the model -- is selected for 
+splitting the data into subgroups.
 \end{enumerate}
+
+As explained in more detail below, the aggregated overall objective
+function in the linear regression model would be the residual sum of
+squares and the corresponding observation-wise goodness-of-fit measure is
+given by the residuals or scores (gradient contributions). Analogously,
+the log-likelihood and corresponding score function could be used.
 
@@
 characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
-explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide} .
+explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide}.
 
@@
 $$
-with the sum of $\psi$ over all observations being the objective function that is optimized 
-to obtain the estimated model parameter $\hat{\beta}$. For this purpose many different M-type 
-estimators can be applied including maximum likelihood or ordinary least squares such that the 
-corresponding objective function can be employed as model-based discrepancy measure for the 
-tree algorithm. Evaluating the score function $s$ on a data set of $n$ observations 
+with the sum of $\psi$ over all observations being the aggregated overall objective 
+function that is optimized to obtain the estimated model parameter $\hat{\beta}$. 
+For this purpose many different M-type estimators can be applied including maximum 
+likelihood or ordinary least squares such that the corresponding objective functions 
+can also be employed for the tree algorithm. Evaluating the score function $s$ on a 
+data set of $n$ observations 
 %containing the 
@@
 
-While CTree offers to apply model-based discrepancy measures as one possible choice, the 
+While CTree offers to apply model-based goodness-of-fit measures as one possible choice, the 
 MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was developed with the aim of embedding 
@@
 each categorized/categorical split variable.
-After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
-the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
-is determined by the categorization of the possible split variables and is in that way fixed.
-
-\newpage
+After choosing the split variable showing the highest dependency by yielding the lowest 
+$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
+The possible number of breakpoints is determined by the categorization of the possible 
+split variables and is in that way fixed.
+
+%\newpage
 \subsection{Pruning}
@@
 based on a unified scale. All other steps are performed in the same way as in classic tree 
-algorithms, such as finding the best split point by optimizing an objective function.
+algorithms, such as finding the best split point by optimizing an overall goodness-of-fit measure.
 
@@
 \begin{itemize}
-\item  Transformations of Y:\\
-A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
-information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\beta})$
-as in the GUIDE algorithm or scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, 
-more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing 
-cases.
-By including all available scores rather than just residuals the goodness of fit of the model is 
-investigated regarding each parameter instead of just the mean. 
+\item Variations of the goodness-of-fit measure:\\
+Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
+model that is obtained for the response $Y$ and possible covariates $X$ can be 
+considered. This can be residuals $r(Y,X,\hat{\beta})$ as in the GUIDE algorithm or 
+scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, more generally, a 
+transformation function $h$ as in the CTree algorithm including both of the foregoing cases.
+By including all available scores rather than just residuals the goodness of fit of the 
+model is investigated regarding each parameter instead of just the mean. 
 %Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.
@@
 
-A very basic form of a discrepancy measure can be obtained for each observation by simply 
+A very basic form of goodness-of-fit measure can be obtained for each observation by simply 
 calculating deviations of the response values from the mean value within the node, for example 
@@
 However, this includes only the response variable $Y$ but not the fitted linear model.
-For a model-based approach residuals can be considered defined by
+For a model-based approach one can consider residuals defined by
 $$
@@
 $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
-The corresponding score function is defined by
+The corresponding disaggregated, observation-wise goodness-of-fit measure is defined by
+the score function 
 $$
@@
 corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
-only scores has already been discussed and elaborated in the literature of structural change
-analysis \citep[see e.g.][]{Ploberger+Kramer:1992}.
+only scores has already been elaborated in the literature of structural change
+analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
+in Section~\ref{sec:discussion}.
 
@@
 In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
-case where the intercept and slope parameter are both varying and the true split point is in the
-center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
-$\delta = 0.3$ is considered making the differences particularly well observable 
-(additional easing $\delta$ as in Section~\ref{sec:scores} 
+case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e. $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
+This choice has been made based on the results shown in Figure~\ref{fig:scores_stump} 
+in order to make the differences particularly well observable.
+In the right panel the true split point is close to the upper end of the considered interval
+($\xi=0.8$) requiring a higher effect size $\delta = 1$ in order to visualize results properly.
+(Additional evaluations for increasing $\delta$ as in Section~\ref{sec:scores} 
 can be found in Appendix~\ref{app:increasingdelta}).
@@
        data = d3way, 
-       ylim = c(0,0.65),
+       ylim = c(0,0.5),
        type = c("a"), 

### 2019-04-24 pkg/glmertree/R/glmertree.R
Added new plot method (called plot.glmertree2 for now). Updated vignette. Reduced size of MHserviceDemo. Added separate testfile+results.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2537 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5b4a56e6eb9346094da1787d050365fa6d64c87f

@@ -3,7 +3,7 @@ utils::globalVariables(c(".tree", ".ranef", ".weights", ".cluster"))
 lmertree <- function(formula, data, weights = NULL, cluster = NULL,
                      ranefstart = NULL, offset = NULL, joint = TRUE,
                      abstol = 0.001, maxit = 100L, dfsplit = TRUE, 
-                     verbose = FALSE, plot = FALSE, 
+                     verbose = FALSE, plot = FALSE, REML = TRUE,
                      lmer.control = lmerControl(), ...)
 {
   ## remember call
@@ -43,7 +43,6 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
     data$.weights <- rep(1L, times = nrow(data))
   }
   
-  
   ## formula processing (full, tree, random)
   ff <- Formula::as.Formula(formula)
   tf <- formula(ff, lhs = 1L, rhs = c(1L, 3L))
@@ -68,7 +67,7 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
   } else if (ranefstart && length(ranefstart) == 1L) {
     ## generate ranefstart from lme null model: 
     predict(lmer(formula(ff, lhs = 1L, rhs = 2L),
-                 data = data, weights = .weights, 
+                 data = data, weights = .weights, REML = REML,
                  offset = offset, control = lmer.control),
             newdata = data)
   } else {
@@ -96,7 +95,7 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
                      cluster = .cluster, dfsplit = dfsplit, ...)
     }
     
-    if (plot) plot(tree)
+    if (plot) plot(tree, main = paste("Iteration", iteration))
     
     data$.tree <- if (joint) {
       factor(predict(tree, newdata = data, type = "node"))
@@ -116,43 +115,39 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
                           lhs = 1L, rhs = c(1L, 2L), collapse = TRUE)
         if (is.null(offset)) {
           lme <- lmer(rf.alt, data = data, weights = .weights, 
-                      control = lmer.control)          
+                      REML = REML, control = lmer.control)          
         } else {
           lme <- lmer(rf.alt, data = data, weights = .weights, 
-                      offset = offset, control = lmer.control)
+                      REML = REML, offset = offset, control = lmer.control)
         }
       } else { # a tree was grown
         if (is.null(offset)) {
[…]

### 2019-05-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small touch-ups in abstract/intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2543 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c01a77a32e540b42f3ebc909d905291c2ab8bde8

@@
 A core step of every algorithm for learning regression trees is the decision if, 
-how, and where to split the underlying data - or, in other words, the selection 
+how, and where to split the underlying data -- or, in other words, the selection 
 of the best splitting variable from the available covariates and the corresponding 
@@
 Therefore, three different popular approaches are considered here: classical categorical 
-association tests (GUIDE), conditional inference (CTree), and parameter instability 
-tests (MOB). First, these are embedded into a common inference framework encompassing
+association tests (as in GUIDE), conditional inference (as in CTree), and parameter instability 
+tests (as in MOB). First, these are embedded into a common inference framework encompassing
 parametric model trees, in particular linear model trees. Second, it is assessed how
@@
 promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
-CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
+%% CHAID \citep{Kass:1980},
+GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
 and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
@@
 choosing the covariate which shows the highest association to the response variable 
-based on $p$-values provided by a statistical test. While QUEST, CHAID, and GUIDE employ 
+based on $p$-values provided by a statistical test. While QUEST
+%%, CHAID,
+and GUIDE employ 
 statistical significance tests for contingency tables, CTree applies permutation tests in 
@@
 Therefore, in this paper the focus will be put on that first step of tree algorithms,
-i.e. the task of selecting the best split variable. In particular, the approach of the 
+i.e. the task of selecting the best split variable in a given (sub)sample. In particular, the approach of the 
 GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
 by investigating the building blocks of their testing strategies in which they differ:
-(1)~Variations of a goodness-of-fit measure for the response: using residuals or full model scores.
+(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
 (2)~Binarization of these residuals or scores.
@@
 
-In many of the considered scenarios it can be observed that the choice of goodness-of-fit
-measure heavily influences the performance of testing strategies as overall using model 
-scores leads to clearly better results than employing residuals only. Moreover, the original
-values of the goodness-of-fit measure are to be preferred over binarized versions of them.
+In many of the considered scenarios the choice of goodness-of-fit measure heavily influences
+the performance of testing strategies with overall using model scores leading to clearly better
+results than employing residuals only. Moreover, the original
+values of the goodness-of-fit measure are preferred over binarized versions of them.
 Also regarding the effects of categorizing possible split variables and the selection
@@
 (\ref{sec:pruning}) are discussed.
-In Section~\ref{sec:unify} the unifying framework for testing strategies in model-based unbiased 
+In Section~\ref{sec:unify} the unifying framework for testing strategies in unbiased model-based
 recursive partitioning algorithms is presented.

### 2019-05-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
worked through Section 2 (except 'pruning')
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2544 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8ccfb2cdfda6b953addda71640d46027293c7b0b

@@
 The basic idea of building a regression tree model is to partition the data into smaller 
-and more homogeneous subgroups based on a set of covariates. Various different tree 
-algorithms have been developed, all following the same general structure starting at 
-the root of the tree which is simply the first node containing all available data:
+and more homogeneous subgroups based on a set of covariates. Various tree 
+algorithms have been developed, following essentially the same general structure.
+Starting at the root of the tree, pertaining to the full available data sample, the
+algorithms proceed in the following steps:
+%
 \begin{enumerate}
-\item For the data set in the current node a model is fitted by optimizing
-some objective function that reflects the goodness of fit.
+  \item For the current sample a (possibly simple) model is fitted by optimizing
+        some objective function (or loss function) that reflects the goodness of fit.
   \item Among all available covariates one is selected as the split variable,
@@
 \end{enumerate}
-
+%
 The term ``model'' is used here in a very broad sense and encompasses not
@@
 Therefore, depending on the type of model, the employed objective function can 
-for example be the sum of deviations from a typical or average value, but it can 
-also be constructed based on the model fitted in each node and depend not only on 
-the response but also on possible covariates. For instance, different 
-types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
-well as rank sums or logrank scores as in \cite{Hothorn+Hornik+VanDeWiel:2006}.
+for example be the sum of deviations from a typical/average value, but it can 
+also be based on a model fro a response along with potential covariates.
+For instance, different types of residuals (or the signs thereof) can be employd
+\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores
+\citep[as in][]{Hothorn+Hornik+VanDeWiel:2006}.
 
@@
 split point. For the first step of selecting a split variable they all apply statistical 
-tests following the same general strategy: 
+tests following the same basic strategy: 
+%
 \begin{enumerate}
@@
         the current subgroup, a disaggregated, observation-wise goodness-of-fit
-measure is obtained. More formally, this is an $n \times k$ matrix $G$
+        measure is obtained. More formally, this is an $n \times k$ matrix
         where $n \in \mathbb{N}$ is the number of observations and $k \in \mathbb{N}$ 
-the number of measures per observation.
-\item The dependency or association of this goodness-of-fit matrix $G$ with
+        the number of goodness-of-fit measures per observation.
+  \item The dependency or association of this goodness-of-fit matrix with
         each possible split variable $Z_j$, $j \in \{1, \dots, m\}$, is assessed
@@
 \end{enumerate}
-
-As explained in more detail below, the aggregated overall objective
-function in the linear regression model would be the residual sum of
-squares and the corresponding observation-wise goodness-of-fit measure is
-given by the residuals or scores (gradient contributions). Analogously,
-the log-likelihood and corresponding score function could be used.
-
-While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
+%
+As an example -- and explained in more detail below -- consider a linear regression tree.
+Thus, a linear regression model is fitted in each subgroup, minimizing the
+residual sum of squares as the aggregated goodness-of-fit measure. The corresponding
+observation-wise goodness-of-fit measure can be given by the residuals or the scores
+(gradient contributions). Analogously, the log-likelihood and corresponding score function could be used.
+
+While this basic approach is the same for the GUIDE, CTree, and MOB algorithms, they differ 
 in their strategies on how to calculate test statistics. In order to point out these specific 
 characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
-explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide}.
+explained in more detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide}.
 
@@
 To select a split variable it is tested whether there is any association between the transformed 
-response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
-variable~$Z$.
-The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
-and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
+response $h(Y)$ and each possible transformed split variable $g(Z_j)$ ($j = 1, \dots, m$).
+The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
+this encompasses ranks, scores, indicator functions, etc.\ and can also be multidimensional.
+For a numeric response the identity function $h(Y) = Y$ is a common 
 choice while a categorical response can be mapped to a unity vector by an indicator function 
 $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
-Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
-e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
-If a parametric model is fit to the values of the response variable~$Y$ and regressor 
-variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
-corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
-where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
+Alternatively, the function $h$ can capture location and scale of $Y$ via
+$h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
+If a parametric model is fitted to the response $Y$ with regressor variable(s)~$X$,
+then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
+the residuals in a linear model with regression coefficients $\beta$. Moreover,
+$s$ can be the score function pertaining to the objective function (or loss function)
+$\ell$:
 $$
-s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
+s(Y, X, \beta) = \frac{\partial \ell(Y, X, \beta)}{\partial \beta}
 $$
-with the sum of $\psi$ over all observations being the aggregated overall objective 
-function that is optimized to obtain the estimated model parameter $\hat{\beta}$. 
-For this purpose many different M-type estimators can be applied including maximum 
-likelihood or ordinary least squares such that the corresponding objective functions 
-can also be employed for the tree algorithm. Evaluating the score function $s$ on a 
-data set of $n$ observations 
-%containing the 
-%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
-%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
-and a $k$-dimensional estimated parameter vector 
-%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
-yields a $n \times k$-matrix.
-In that way a score value indicating how well the model fits the data is given separately
-for each tuple of observation and estimated parameter element 
-$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
-and $l \in \{1,\ldots,k\}$.
+The estimate of the model parameters $\hat{\beta}$ is obtained by optimizing
+the sum of the objective function $\ell$, aggregated over all observations.
+This framework includes many different M-type estimators as special cases, including maximum 
+likelihood and ordinary least squares estimation. For a $k$-dimensional parameter
+$\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
+is also a $k$dimensional vector, i.e., the gradient contribution of the $i$-th observation.
+Thus, the $n \times k$-matrix consisting of these scores or gradient contributions for all 
+$i = 1, \dots, n$ observations is a natural candidate for the observation-wise goodness-of-fit
+measure as described in the previous Section~\ref{sec:generic_alg}.
 
@@
 
-To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
-Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the linear
-test statistic given all permutations of the response variable can be calculated and used to 
-standardize the test statistic. In order to map multivariate test statistics into the real 
-line two options are given: either the maximum of the absolute values of the standardized 
-statistic or a quadratic form. The resulting values can then be transformed to 
-the scale of $p$-values allowing for an unbiased comparison in order to select a split variable.
-
-If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
-correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
-is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree 
-algorithm provides a very general framework where the type of variables and the corresponding 
-functions $g$ and $h$ determine the type of the applied statistical test.
-(See \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox.)
-
-In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
-observed values or optionally at an average value lying between two observations. To select the
-best split point two-sample linear statistics of the same structure as in the independence test 
-are evaluated for all possible subsets of the sample space of the selected split variable.
+To test for independence of $h(Y)$ and $g(Z)$ CTree calculates a linear association test statistic,
+following the framework of \cite{Strasser+Weber:1999}. The corresponding conditional expectation and
+covariance given all permutations of the response variable can be calculated and used to 
+standardize the test statistic. This standardized statistic has an asymptotic normal distribution
+which is in fact multivariate if either of the transformation $h(Y)$ and/or $g(Z)$ is
+multivariate. The actual test is carried out be mapping this standardized statistic to
+the real line either by taking the absolute maximum or using a quadratic form -- with $p$-values
+being computed from the analogous transformation of the normal distribution
+\citep[see][for a hands-on introduction]{Hothorn+Hornik+VanDeWiel:2006}.
+
+If both variables $Y$ and $Z$ are numeric the default independence test corresponds to a Pearson 
+correlation test. For one numeric and one categorical variable essentially a 1-way ANOVA
+(analysis of variance) is employed while for two categorical variables a $\chi^2$~test is performed.
+Thus, in the general CTree framework many types of tests can be specified by selecting suitable
+transformations $g$ and $h$. While originally
+conceived for nonparametric models, it is easy to adapt CTree to model-based testing and 
+recursive partitioning by choosing a model-based $h$ transformation as argued above
+\citep[see the concrete examples in][]{Zeileis+Hothorn:2013,Seibold+Zeileis+Hothorn:2016}.
+
+% In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
+% observed values or optionally at an average value lying between two observations. To select the
+% best split point two-sample linear statistics of the same structure as in the independence test 
+% are evaluated for all possible subsets of the sample space of the selected split variable.
 
@@
 
-While CTree offers to apply model-based goodness-of-fit measures as one possible choice, the 
-MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was developed with the aim of embedding 
-parametric models into the framework of regression trees and is hence purely intended for 
-a model-based approach.
-
-The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
-an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
-application of the central limit theorem. 
-The fitted model can be of a very basic and simple form such as an average value of the response 
-vector but can also be a more complex parametric model.
-In \cite{Zeileis+Hothorn+Hornik:2008} generalized linear models (GLMs) and survival regression models 
-are used to exemplify the algorithm. Subsequently, various other types of models have been applied as well,
-for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
-psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
-mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
-
-Regardless of the applied model type the algorithm only requires the corresponding model 
-score matrix in order to test for any instability in the model parameters over each 
-particular ordering induced by each of the possible partitioning variables $Z$.
-If a variable $Z$ has no specific influence on the parameters the ordered scores would 
-always fluctuate randomly around their mean 0. Therefore, these deviations are investigated 
-by measuring them using an empirical fluctuation process. Applying a scalar function which 
-captures the fluctuation in this process yields the test statistic of this so called 
-generalized M-fluctuation test \citep{Zeileis+Hornik:2007}. 
-To obtain the distribution of the resulting test statistic asymptotic theory, in particular 
-the central limit theorem is applied. In this parametric unconditional framework $p$-values 
-can be calculated and the covariate with the smallest $p$-value is selected as split variable.
-
-For a categorical split variable all possible splits are evaluated to find the best one while 
-for numeric split variable a maximally selected split point search is employed. Hence, 
-the number of possible breakpoints is not fixed unlike in the CTree algorithm.
+In contrast to CTree the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was explicitly
+designed for a model-based goodness-of-fit measure in order to embed parametric models
+into a regression tree framework. Thus, MOB is based on an objective/loss function $\ell$
+and corresponding score function $s$. The original paper considered generalized linear models
+(GLMs) and survival regression models but subsequently various other models have been applied as well,
+including beta regression \citep{Gruen+Kosmidis+Zeileis:2012}, psychometric item response theory
+models \citep{Strobl+Kopf+Zeileis:2015}, or mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
+But just like CTree can be applied to parametric models, MOB conversely also encompasses
+simple regression and classification trees, e.g., by choosing an intercept-only model.
+
+For selecting a split variable MOB employs a score-based test that relies on the central limit
+theorem for the parameter estimate $\hat{\beta}$. The test assesses whether the scores
+-- when ordered by the potential split variable $Z$ -- fluctuate randomly around their
+zero mean or differ systematically in certain subgroups. The latter would indicate a
+parameter instability that could be captured by fitting separate models (optimizing $\ell$)
+in the resulting subgroups. In case of a numeric split variable $Z$ both the score-based statistic
+and the partitioned objective function $\ell$ are maximized over all possible splits in $Z$
+(subject to certain minimal subgroup size constraints). Unlike the CTree framework, MOB
+relies on classical unconditional inference. For more details see \cite{Zeileis+Hornik:2007}.
 
 \subsection{GUIDE}\label{sec:guide}
-Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
-the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
-a constant fit with the parameter $\beta$ being the mean of the response, hence 
-$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$, 
-%for a response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, 
-can be employed as well as more complex models.
-For the statistical test applied to select a split variable some type of model-based
-residual is required. In \cite{Loh:2002} linear models are used as an example where the
-residuals are defined by
+
+Building on earlier work for the QUEST algorithm \citep{Loh+Shih:1997}, \citep{Loh:2002}
+proposed the GUIDE algorithm blending trees with parametric regression models and
+encompassing simpler classification and regression trees as special cases. Thus,
+linear regression models could be fitted in the nodes of tree as well as
+constant fits such as simple mean response $\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$,.
+The tests for selecting the split variable are then based on the corresponding residuals,
+e.g., for a simple linear regression \citep[as used in][]{Loh:2002}:
 $$
@@
 $$
-In related literature other models together with an appropriate choice of residuals
+In subsequent work other models together with an appropriate choice of residuals
 have been applied such as in
@@
 %(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
-quantile regression models \citep{Chaudhuri+Loh:2002} 
-and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
-%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))
-
-Before applying a statistical test two additional steps are performed:
-(1)~The residuals are binarized, i.e. depending on whether the sign of a residual is 
-positive or not it is replaced by $1$ or $2$.  % FIX ME: +1 and -1 in the implementation? 
-(2)~For each possible split variable the observed values are categorized into four groups 
-separated by the corresponding 1st, 2nd and 3rd quartile unless the split variable is already
-categorical.
-Then, a $\chi^2$-test of independence is performed for the binarized residuals and
+quantile regression models \citep{Chaudhuri+Loh:2002}, 
+and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015}, among others.
+
+To construct a statistical test two additional transformation are carried out:
+(1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
+(2)~All possible split variable $Z$ are categorized, i.e., unless $Z$ is already categorical
+it is split at its quartiles into four bins.
+Subsequently, a $\chi^2$~test of independence is performed for the binarized residuals and
 each categorized/categorical split variable.
@@
 $p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
-The possible number of breakpoints is determined by the categorization of the possible 
-split variables and is in that way fixed.
+Note that the split point in numeric variables $Z$ is searched over all possible splits,
+not just the four bins that were constructed for the $\chi^2$~test.
 
@@
 and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
-depending on whether the scores are binarized or not, a $\chi^2$-test or 1-way ANOVA can be applied
+depending on whether the scores are binarized or not, a $\chi^2$~test or 1-way ANOVA can be applied
 where the quartiles are the only possible split points while otherwise maximally selected splits 
@@
 test statistic. While the permutation tests in CTree work in a conditional framework both MOB (M-fluctuation tests) 
-and GUIDE ($\chi^2$-tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
+and GUIDE ($\chi^2$~tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
 introduced employing prepruning while GUIDE originally applies a postpruning strategy.
@@
 $$
-The appropriate choice of the function $\psi$ is then
+The appropriate choice of the function $\ell$ is then
 $$
-\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
+\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
 $$
-such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
+such that $\sum_{i=1}^n \ell(y_i,x_i,\beta_0,\beta_1)$ is the objective function
 which is minimized to obtain the estimated coefficient vector
@@
 s(y_i, x_i, \beta_0, \beta_1) = 
-\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
+\frac{\partial \ell}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
 -2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
@@
 Therefore, including only residuals (i.e. only the first column of the score matrix)
-corresponds to including the derivatives of $\psi$ only regarding the intercept 
+corresponds to including the derivatives of $\ell$ only regarding the intercept 
 parameter~$\beta_0$.
 On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
-the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
+the derivatives of $\ell$ regarding the intercept $\beta_0$ (first column) and the slope 
 parameter $\beta_1$ (second column) are considered.

### 2019-05-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
rewrote pruning section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2546 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 53a831f790efcf703d5b8d9663cad912d7e00d42

@@
 \label{sec:pruning}
-In early developed methods of unbiased recursive partitioning such as 
-FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
-CRUISE \citep{Kim+Loh:2001} or GUIDE \citep{Loh:2002}
-the $p$-values of applied significance tests are used purely for the selection of a split variable 
-based on a unified scale. All other steps are performed in the same way as in classic tree 
-algorithms, such as finding the best split point by optimizing an overall goodness-of-fit measure.
-
-In subsequent approaches as in \cite{Hothorn+Hornik+Zeileis:2006}
-it is shown that these $p$-values can also be used as a basis for a
-pruning strategy namely \textit{prepruning}: Only if the $p$-value corresponding to the selected split
-variable (hence the smallest $p$-value) is smaller than a given significance level a split is
-performed while otherwise no further splits are made in the current node.
-In that way the information provided by the significance test is exploited in a more profound way.
-However, prepruning requires the applied test to perform reasonably well as a significance test, 
-i.e. it has to have power but also avoid overfitting.
-
-On the contrary, when employing postpruning no significance level is included leading to very large 
-trees which are then pruned afterwards, for example by applying cost-complexity pruning as in the 
-CART algorithm \citep{Breiman+Friedman+Stone:1984}. 
-Thus, for postpruning the test only needs to be able to select the best split variable but does 
-not have to be effective as a significance test.
-
-Originally the GUIDE algorithm was introduced applying postpruning while CTree and MOB
-were both presented employing prepruning (see Table~\ref{tab:combinations} for a more detailed
-overview). Alternatively, for model-based partitioning a pruning strategy can also use model-based 
-information criteria such as AIC or BIC \citep[see for example][]{Su+Wang+Fan:2004}.
-However, all testing strategies can be combined with each of these pruning strategies.
-If a test performs well the choice of a pruning strategy has hardly any impact on the results, 
-but as prepruning only requires already provided information it is usually less computationally
-expensive than postpruning needing additional calculations. 
-However, if the test performs poorly in terms of power and overfitting, prepruning suffers from 
-this deficiency while postpruning can compensate it at least partly and improve the results.
-Therefore, being aware of the power of the selected testing strategy is essential for choosing 
-an appropriate pruning strategy.
-For that reason we are focusing on the testing strategies applied in tree algorithms 
-and investigate how effective they are as significance tests.
+
+To avoid overfitting recursive partitioning algorithms need to assure that trees do not grow
+too large. Apart from certain minimal subgroup size or maximal tree depth constraints this is
+classically accomplished by so-called ``pruning'' approaches. As the three tree algorithms
+considered here (CTree, MOB, and GUIDE) differ in their default choice of pruning approach,
+we briefly discuss these here. However, as all three algorithms can in principle be combined
+with any of the pruning approaches this is done only relatively briefly.
+
+The classical CART algorithm \citep{Breiman+Friedman+Stone:1984} proposed to first grow a large
+tree and then prune those splits in the tree that did not increase predictive performance in a
+cross-validation. This is also know as \emph{postpruning} (after growing the initial tree) and
+more specifically cost-complexity pruning.
+
+In the unbiased recursive partitioning literature this postpruning approach is also usef frequently
+\citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
+from the association tests are only employed for selecting the split variable on a unified scale.
+However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
+so-called \emph{prepruning} strategy which stops growing the tree as soon as no significant association
+can be found in a given subgroup. This approach is the default in CTree and also in
+MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
+postpruning in model-based partitioning is to use information criteria such as AIC (Akaike information
+criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.
+
+Clearly, for inference-based prepruning it is crucial that the association tests employed for
+split selection work well as statistical significance test, i.e., conform with their nominal size
+and have high power. In contrast, when using postpruning (either based on cross-validation or
+information criteria) it might not be as crucial that the significance test works well and has
+high power.
+
+Due to these considerations we first evaluate the significance tests underlying CTree, MOB,
+and GUIDE by themselves, i.e., without growing an actual tree in combination with a pruning strategy.
+Subsequently we combine the tests with a cost-complexity postpruning approach in order to assess
+whether shortcomings of the tests are mitigated by pruning.
 

### 2019-05-08 pkg/partykit/inst/guideliketest/draft/draft.Rnw
corrected minor typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2547 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 19dd5ea266a9117a17589d924a4a468c7f6e5a84

@@
 for example be the sum of deviations from a typical/average value, but it can 
-also be based on a model fro a response along with potential covariates.
-For instance, different types of residuals (or the signs thereof) can be employd
+also be based on a model for a response along with potential covariates.
+For instance, different types of residuals (or the signs thereof) can be employed
 \citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores
@@
 $\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
-is also a $k$dimensional vector, i.e., the gradient contribution of the $i$-th observation.
+is also a $k$-dimensional vector, i.e., the gradient contribution of the $i$-th observation.
 Thus, the $n \times k$-matrix consisting of these scores or gradient contributions for all 
@@
 which is in fact multivariate if either of the transformation $h(Y)$ and/or $g(Z)$ is
-multivariate. The actual test is carried out be mapping this standardized statistic to
+multivariate. The actual test is carried out by mapping this standardized statistic to
 the real line either by taking the absolute maximum or using a quadratic form -- with $p$-values
@@
 
-Building on earlier work for the QUEST algorithm \citep{Loh+Shih:1997}, \citep{Loh:2002}
+Building on earlier work for the QUEST algorithm \citep{Loh+Shih:1997}, \cite{Loh:2002}
 proposed the GUIDE algorithm blending trees with parametric regression models and
 encompassing simpler classification and regression trees as special cases. Thus,
-linear regression models could be fitted in the nodes of tree as well as
-constant fits such as simple mean response $\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$,.
+linear regression models could be fitted in the nodes of a tree as well as
+constant fits such as simple mean response $\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$.
 The tests for selecting the split variable are then based on the corresponding residuals,
@@
 
-To construct a statistical test two additional transformation are carried out:
+To construct a statistical test two additional transformations are carried out:
 (1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
-(2)~All possible split variable $Z$ are categorized, i.e., unless $Z$ is already categorical
+(2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
 it is split at its quartiles into four bins.
@@
 
-In the unbiased recursive partitioning literature this postpruning approach is also usef frequently
+In the unbiased recursive partitioning literature this postpruning approach is also used frequently
 \citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
@@
 of parametric models together with prepruning as the properties of this pruning strategy
-are already determined by the testing strategy. Subsequently combinations
+are already determined by the testing strategy. Subsequently, combinations
 with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
@@
 basic structure as explained in Section~\ref{sec:generic_alg}. 
-Thus they can be presented together in a unifying framework for testing strategies
+Thus, they can be presented together in a unifying framework for testing strategies
 in model-based unbiased recursive partitioning. 
@@
 scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, more generally, a 
-transformation function $h$ as in the CTree algorithm including both of the foregoing cases.
+transformation function $h(Y)$ as in the CTree algorithm including both of the foregoing cases.
 By including all available scores rather than just residuals the goodness of fit of the 
@@
 \item Categorization of covariates:\\
-Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
+Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd, 
 and 3rd quartile are considered to bin each possible split variable into four categories. In that way,

### 2019-05-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised Section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2549 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1f3bd10cffd26eae331ed6541b2a399ae9e918da

@@
 Therefore, in this paper the focus will be put on that first step of tree algorithms,
-i.e. the task of selecting the best split variable in a given (sub)sample. In particular, the approach of the 
+i.e., the task of selecting the best split variable in a given (sub)sample. In particular, the approach of the 
 GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
@@
 tree and then prune those splits in the tree that did not increase predictive performance in a
-cross-validation. This is also know as \emph{postpruning} (after growing the initial tree) and
+cross-validation. This is also know as \emph{post-pruning} (after growing the initial tree) and
 more specifically cost-complexity pruning.
 
-In the unbiased recursive partitioning literature this postpruning approach is also used frequently
+In the unbiased recursive partitioning literature this post-pruning approach is also used frequently
 \citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
@@
 However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
-so-called \emph{prepruning} strategy which stops growing the tree as soon as no significant association
+so-called \emph{pre-pruning} strategy which stops growing the tree as soon as no significant association
 can be found in a given subgroup. This approach is the default in CTree and also in
 MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
-postpruning in model-based partitioning is to use information criteria such as AIC (Akaike information
+post-pruning in model-based partitioning is to use information criteria such as AIC (Akaike information
 criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.
 
-Clearly, for inference-based prepruning it is crucial that the association tests employed for
+Clearly, for inference-based pre-pruning it is crucial that the association tests employed for
 split selection work well as statistical significance test, i.e., conform with their nominal size
-and have high power. In contrast, when using postpruning (either based on cross-validation or
+and have high power. In contrast, when using post-pruning (either based on cross-validation or
 information criteria) it might not be as crucial that the significance test works well and has
@@
 and GUIDE by themselves, i.e., without growing an actual tree in combination with a pruning strategy.
-Subsequently we combine the tests with a cost-complexity postpruning approach in order to assess
+Subsequently we combine the tests with a cost-complexity post-pruning approach in order to assess
 whether shortcomings of the tests are mitigated by pruning.
@@
 \section{Unifying framework}\label{sec:unify}
+
 Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
-of the type of fitted models, the type of applied tests and the employed pruning strategy
-as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
-original form. However, in principle all other combinations of the listed features are possible 
-as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
-strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
-of parametric models together with prepruning as the properties of this pruning strategy
-are already determined by the testing strategy. Subsequently, combinations
-with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
-From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
-testing strategies rather than the complete algorithms.
+of the type of model fits, tests, and pruning strategy employed to
+grow the tree. Table~\ref{tab:combinations} provides an overview of the default
+combinations.
+However, as discussed above, subsequent publications have emphasized that all three
+algorithms can be combined with different model fitting approaches and to some degree
+different pruning strategies have been explored as well. Thus, the class of tests
+employed for the unbiased splitting variable selection forms the core of each of the
+algorithms: conditional inference (CTree) vs.\ score-based fluctuation tests (MOB)
+vs.\ residual-based $\chi^2$ tests (GUIDE). Therefore, we consider a standard class
+of model fits (namely, linear regression trees) and investigate the relative
+advantages and disadvantages of the tests themselves (which are essential to pre-pruning)
+as well as the combination with post-pruning. Thus, subsequently
+the names CTree, MOB, and GUIDE distinguish the test-based variable and split selection
+rather than the entire algorithm with all default settings.
 
@@
 \end{tabular}
-\caption{Combinations of fitted model type, test type and pruning strategy
+\caption{Default combinations of fitted model type, test type and pruning strategy
 for the algorithms CTree, MOB, and GUIDE.
@@
 
-\subsection{Unification of testing strategies}
-
-Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
-Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
-basic structure as explained in Section~\ref{sec:generic_alg}. 
-Thus, they can be presented together in a unifying framework for testing strategies
-in model-based unbiased recursive partitioning. 
-In this framework each specific testing strategy can be obtained by combining the provided 
-building blocks accordingly. But apart from that, variations of the commonly known testing 
-strategies can be obtained as this general toolbox allows for all possible combinations of the 
-building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
-These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
-for the standard setup of CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which 
-gives a more detailed overview of the differences in the second column of Table~\ref{tab:combinations}.
-
 \subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
-The three following building blocks all refer to the way how the given data is prepared/transformed
-before employing a statistical test to select a split variable.
+
+Even though CTree, MOB, and GUIDE differ in the specific tests they apply, their approaches for split
+variable selection follow the same basic structure as explained in Section~\ref{sec:generic_alg}. 
+In fact, the tests can be embedded in a unifying conceptual framework that yields the
+different tests by combining various building blocks.
+These mostly differ in the way the model for the dependent variable on the one hand and
+the splitting variables on the other are prepared or transformed:
+%
 \begin{itemize}
-\item Variations of the goodness-of-fit measure:\\
+  \item \emph{Goodness-of-fit measure:}\\
         Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
-model that is obtained for the response $Y$ and possible covariates $X$ can be 
-considered. This can be residuals $r(Y,X,\hat{\beta})$ as in the GUIDE algorithm or 
-scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, more generally, a 
-transformation function $h(Y)$ as in the CTree algorithm including both of the foregoing cases.
-By including all available scores rather than just residuals the goodness of fit of the 
-model is investigated regarding each parameter instead of just the mean. 
-%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.
-
-\item Binarization of residuals/scores:\\
-Rather than applying the independence test directly on the residuals/scores a
-binarized version of them can be used as well. In particular, the values are binarized
-at $0$ such that only their signs are considered.
-
-\item Categorization of covariates:\\
-Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd, 
-and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
-depending on whether the scores are binarized or not, a $\chi^2$~test or 1-way ANOVA can be applied
-where the quartiles are the only possible split points while otherwise maximally selected splits 
-over $Z$ are considered.
-
+        model for the response $Y$ and possible covariates $X$ can be considered. Either residuals
+	$r(Y,X,\hat{\beta})$ can be used as proposed for GUIDE or model scores $s(Y,X,\hat{\beta})$
+        as proposed for MOB. All three algorithms can, in principle, use both goodness-of-fit
+	measures though, which is probably brought out most clearly in the CTree algorithm that
+        explicitly allowed for different transformations $h(Y)$ in its original description already.
+  \item \emph{Binarization of residuals/scores:}\\
+        Rather than testing independence between the split variables and the residuals/scores
+	tmeselves, it is possible	
+	to binarize residuals/scores at $0$ so that only their signs are assessed (as proposed for GUIDE).
+  \item \emph{Categorization of split variables:}\\
+        Similarly, the split variables can also be categorized (for testing only). This was proposed for
+	GUIDE, employing binning at the quartiles yielding four categories of approximately
+	equal size.
 \end{itemize}
-
-\textit{Note:}
-Apart from these three building blocks, the algorithms CTree, MOB, and GUIDE differ in the type of test 
-statistic they apply and their approximation of the underlying null distribution of the corresponding 
-test statistic. While the permutation tests in CTree work in a conditional framework both MOB (M-fluctuation tests) 
-and GUIDE ($\chi^2$~tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
-introduced employing prepruning while GUIDE originally applies a postpruning strategy.
-However, these differences refer to the standard versions of the three algorithms and all of them
-allow for certain modifications of this standard setup. For example, a different form of test statistic
-can be selected in CTree considering a maximum value as in MOB.
+%
+The three algorithms combine these building blocks in different ways as shown in Table~\ref{tab:buildingblocks}:
+When applying CTree for unbiased model-based recursive partitioning it has been suggested to use
+the model scores without binarization and assess their association with the untransformed split
+variables using a conditional inference test. This is similar to a squared correlation test statistic.
+MOB also employs the scores without
+binarization and maximally selects a score statistic over all potential split points in the
+split variable. GUIDE employs the binarized residuals and assesses their association with
+the categorized split variable in a classical (unconditional) $\chi^2$ test.
+
+But the building blocks could be easily re-combined to yield new types of tests. For example,
+in the GUIDE approach, a 1-way ANOVA can be used for assessing the association of the residuals
+(without binarization) with the categorized split variable. Or alternatively, a multivariate
+1-way ANOVA can be used for the non-binarized scores as opposed to the residuals etc.
+
+Note that there are further differences in the testing strategies between the three algorithms, e.g.,
+using conditional vs.\ unconditional approximations of the null distributions. However, this difference
+has relatively little influence compared to the other building blocks considered in detail here.
+Moreover, both similarities and relative differences between these approaches have been previously
+discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.
 
@@
 \subsection{Linear model tree}\label{sec:lmtrees}
-As an example a linear model tree is considered as this allows for an easy and well structured 
-description of the investigated subjects. However, the conclusions drawn from this example also 
-hold for a wide range of other model types. 
-The models fitted in the nodes of the linear model tree as considered for the simulation 
-study in Section~\ref{sec:simulation} are of the form
+
+To focus on the unified testing framework, as described in the previous section, we employ
+the same model fits for all three algorithms. To do so, we employ linear regression models
+because it is such a basic and widely used model and linear model trees were the leading
+illustrations in both the original MOB \citep{Zeileis+Hothorn+Hornik:2008} and GUIDE \citep{Loh:2002} papers.
+However, the conclusions drawn from this example also hold for many other model types.
+
+To fix notation, we consider the following models for the simulation 
+study in Section~\ref{sec:simulation}:
 $$
@@
 $$
-for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
-For each node the type of the fitted model is the same such that no node-specific features are 
-considered and the observations of the (arbitrarily chosen) current node are denoted by 
-$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.
-
-A very basic form of goodness-of-fit measure can be obtained for each observation by simply 
-calculating deviations of the response values from the mean value within the node, for example 
-by $(\bar{y} - y_i)$.
-However, this includes only the response variable $Y$ but not the fitted linear model.
-For a model-based approach one can consider residuals defined by
+with response variable $Y$, regressor variable $X$, and error term $\epsilon$. This model is
+fitted, as usual, by ordinary least squares (OLS) to the observations in each notation,
+yielding the parameter estimates $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
+To keep notation simple we present the following equations for the root node with all observations
+$\{(y_i,x_i)\}_{i=1,\ldots,n}$ and $n \in \mathbb{N}$. In subsequent nodes the same equations
+are used but just for a smaller subgroup.
+
+The aggregated goodness-of-fit measure (or objective or loss function) in OLS estimation
+is the sum of squared residuals: $\sum_{i=1}^n \ell(y_i,x_i,\beta_0,\beta_1)$ where
+$$
+\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
+$$
+is the squared residual which is defined as
 $$
@@
 $$
-The appropriate choice of the function $\ell$ is then
-$$
-\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
-$$
-such that $\sum_{i=1}^n \ell(y_i,x_i,\beta_0,\beta_1)$ is the objective function
-which is minimized to obtain the estimated coefficient vector
-$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
-The corresponding disaggregated, observation-wise goodness-of-fit measure is defined by
-the score function 
+These residuals can also intuitively be used as the corresponding disaggregated
+observation-wise goodness-of-fit measure. Another natural candidate for this is the
+score function for the $i$-th observation:
 $$
@@
 $$
-where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
-score matrix corresponds to the residuals. 
-Therefore, including only residuals (i.e. only the first column of the score matrix)
-corresponds to including the derivatives of $\ell$ only regarding the intercept 
-parameter~$\beta_0$.
-On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
-the derivatives of $\ell$ regarding the intercept $\beta_0$ (first column) and the slope 
-parameter $\beta_1$ (second column) are considered.
-
-%\newpage
+Thus, up to a constant scaling factor of $-2$ (that could also be omitted) the first component
+of the scores is in fact the residual. However, it is complemented by a second component
+that captures the slope effect of $x_i$. Therefore, when computing the score for all $n$ observations
+this yields an $n \times 2$ score matrix whose first column corresponds to the $n$ residuals.
+As this is the derivative with respect to the intercept parameter~$\beta_0$, it
+captures changes in the intercept. Moreover, the second column of the score matrix contains
+derivatives with respect to the slope parameter~$\beta_1$ and thus captures changes in this.
 
@@
 presented in Section~\ref{sec:lmtrees}. First, the underlying tree structure based on which the
-data is generated is a stump, i.e. a tree with only one split. 
+data is generated is a stump, i.e., a tree with only one split. 
 By keeping the complexity of the model as low as possible the advantages and
-disadvantages of the different testing strategies can be elaborated more precisely and 
-the focus can be put on their power in terms of selecting the correct split variable.
+disadvantages of the different testing strategies can be elaborated in more detail and 
+focus can be given to their power in terms of selecting the correct split variable.
 
@@
 \subsection{Data generating process}
-Each generated data set consists of a response variable, one regressor, one split 
-variable and nine noise variables as listed in Table~\ref{tab:variables} 
+Each generated data set consists of the response and regressor variables, one true split 
+variable and nine noise split variables as listed in Table~\ref{tab:variables} 
 together with the corresponding distributions.
@@
 % variable (taking values between $-1$ and $1$).
-Moreover, the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation
-(can) depend on the true split variable $Z_1$. 
-%in the following way:
-%\begin{center}
-%$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
-%\end{center}
+Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$ of this linear equation
+(can) depend on the true split variable $Z_1$. More specifically, three different scenarios
+are considered for the coefficients $\beta_0$ and $\beta_1$:
 %
-In particular, three different cases are considered for the coefficients $\beta_0$ 
-and $\beta_1$:
 \begin{enumerate}
@@
 \end{enumerate}
-
-The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, are binary taking either the positive or
-the negative value of the effect size, i.e. $\delta$ or $-\delta$.
+%
+The varying coefficient(s) $\beta_j$, $j\in\{0,1\}$, are binary taking either the positive or
+the negative value of the effect size, i.e., $\delta$ or $-\delta$.
 % can either be binary (taking either $\delta$ or $-\delta$), or a continuous function. 
@@
 \beta_{i}(Z_1) = \begin{cases}
--\delta \cdot (-1)^{i}  \quad \text{if } Z_1 < \xi \\
-+\delta \cdot (-1)^{i}  \quad \text{if } Z_1 \geq \xi
+-\delta \cdot (-1)^{j}  \quad \text{if } Z_1 < \xi \\
++\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \geq \xi
 \end{cases}
@@
 \endminipage
-\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$ and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
+\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$
+and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
 Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
@@
 Table~\ref{tab:buildingblocks}, hence adapted versions of GUIDE, CTREE, and MOB are evaluated as well
-as their original versions. Moreover, as prepruning is employed all of the considered statistical 
+as their original versions. Moreover, as pre-pruning is employed all of the considered statistical 
 tests are applied as significance tests answering two questions at once: 
@@
 in $\beta_1$. 
-This is due to the fact that GUIDE only includes residuals, i.e. only the scores 
+This is due to the fact that GUIDE only includes residuals, i.e., only the scores 
 regarding the intercept, such that changes in the slope parameter are not detected unless the 
@@
 scores when testing for dependencies in order to select split variables is advantageous over
-considering only residuals, i.e. the scores regarding only the intercept, as changes not only in
+considering only residuals, i.e., the scores regarding only the intercept, as changes not only in
 the mean but also in all other parameters can be detected in that way. This is particularly obvious
@@
 In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
-case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e. $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
+case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e., $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
 This choice has been made based on the results shown in Figure~\ref{fig:scores_stump} 
@@
 the tree models presented in Figure~\ref{fig:tree_ari} are built again, but this time
-applying postpruning (building very large trees without any significance level and employing 
+applying post-pruning (building very large trees without any significance level and employing 
 cost-complexity pruning as in the CART algorithm \citep{Breiman+Friedman+Stone:1984}). 
-In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with prepruning,
+In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with pre-pruning,
 again based on the Adjusted Rand Index.
@@
 Therefore, it can be recommended to take advantage of this power and use the information
-given by the tests for prepruning. As it can be seen in Figure~\ref{fig:pruning_ari}
+given by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:pruning_ari}
 in most cases the results of these two testing strategies do not vary significantly over 
@@
 On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
-strategy should not be based on their results which argues for the application of postpruning.
-This can also be observed in Figure~\ref{fig:pruning_ari} showing that postpruning improves
+strategy should not be based on their results which argues for the application of post-pruning.
+This can also be observed in Figure~\ref{fig:pruning_ari} showing that post-pruning improves
 the results for both GUIDE versions.
@@
 pr <- pr[,c(1,2,7,8,9,12,15)]
-## extended data set with postpruning results being represented as separated tests
+## extended data set with post-pruning results being represented as separated tests
 prex <- pr[,c(1,2,3,4,6)]
@@
 The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE\_scores. 
-The dashed lines represent the results from applying prepruning and the continuous lines 
-the results from applying postpruning.}
+The dashed lines represent the results from applying pre-pruning and the continuous lines 
+the results from applying post-pruning.}
 \end{figure}

### 2019-05-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised Section 4.1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2550 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6fdf7acbc7096d32e525b693997aba6b59e83f37

@@
 ## HCL palette for illustrations of data generating process
-pal_dgp_light <- qualitative_hcl(3, "Set 2")
+pal_dgp_light <- qualitative_hcl(3, "Dark 3", c1 = 50, l1 = 75)
 pal_dgp <- qualitative_hcl(3, "Dark 3")
@@
 presented in Section~\ref{sec:lmtrees}. First, the underlying tree structure based on which the
-data is generated is a stump, i.e., a tree with only one split. 
-By keeping the complexity of the model as low as possible the advantages and
-disadvantages of the different testing strategies can be elaborated in more detail and 
-focus can be given to their power in terms of selecting the correct split variable.
+data is generated is a stump, i.e., a tree with only one split (see Figure~\ref{fig:dgp_stump}). 
+By keeping the tree so simple we can focus on the testing strategy only giving focus to their
+power in terms of selecting the correct split variable.
 
 In the second scenario the true tree structure contains two splits in two different variables
-yielding a tree with three terminal nodes. After having investigated the characteristics of 
-the employed algorithms in detail for the stump scenario this additional evaluation aims at
-showing that the same conclusions can be drawn for more complex structures.
-
-In the following the data generating process is explained in detail for the first scenario.
-The second, more complex scenario employs the same basic structure simply adding another split
-as illustrated in Figure~\ref{fig:dgp_tree}.
+yielding a tree with three terminal nodes (see Figure~\ref{fig:dgp_tree}). It employs the same
+basic structure as the first scenario but simply adds another split. This allows to evaluate
+the power of the three testing strategies in a more complex setting, in combination with using
+a post-pruning strategy.
 
 \subsection{Data generating process}
+
 Each generated data set consists of the response and regressor variables, one true split 
@@
 % variable (taking values between $-1$ and $1$).
-Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$ of this linear equation
-(can) depend on the true split variable $Z_1$. More specifically, three different scenarios
+Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
+can depend on the true split variable $Z_1$. More specifically, three different scenarios
 are considered for the coefficients $\beta_0$ and $\beta_1$:
@@
 %
-The varying coefficient(s) $\beta_j$, $j\in\{0,1\}$, are binary taking either the positive or
-the negative value of the effect size, i.e., $\delta$ or $-\delta$.
-% can either be binary (taking either $\delta$ or $-\delta$), or a continuous function. 
-% In the binary case the following definition is applied for the true split point $\xi$
+See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
+changes ($\beta_j$ with $j\in\{0,1\}$) switches between two values at the split point $\xi$:
+%
 \begin{align*}
@@
 \end{align*}
+%
 %while in the continuous case a simple linear function is considered in the form of
@@
 opposite directions.
-An illustration of the three different cases 
-%for binary coefficients and a continuous regressor 
-can be found in Figure~\ref{fig:dgp_stump}.
 
@@
                 expression(z[2]>xi~" and "~z[1]>xi)), 
-       col = c(pal[1], pal[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
+       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
 mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)

### 2019-05-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised Section 4.2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2551 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 82b94f655d8fb580fb2a901ac49e6b99045444e7

@@
 The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
-on 100 data sets per step each consisting of 250 observations. 
+on 100 replications per step each consisting of 250 observations. 
 To compare the performance of the evaluated testing strategies for each step 
-the $p$-values corresponding to the true split variable $Z_1$ and the proportion
-of data sets for which $Z_1$ is detected and also selected 
-%(for a given significance level $\alpha = 0.05$) 
-is considered for the first scenario (stump).
-For the second scenario (tree with 2 splits) the Adjusted Rand Index~(ARI) is calculated 
+in the first scenario (stump with 1 split) the following criteria are considered:
+the $p$-values pertaining only to the true split variable $Z_1$;
+and the proportion of data sets for which the $p$-value of $Z_1$
+is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
+selected for splitting in a pre-pruning approach).
+For the second scenario (tree with 2 splits) the adjusted Rand index~(ARI) is calculated 
 as a measure of similarity between the true tree structure and the fitted model tree.

### 2019-05-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
assorted touch-ups in Section 5, still various TODOs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2552 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 093b0a157596e2b471cab114ffb16029144d5208

@@
 pal <- qualitative_hcl(5, "Dark 3")
-names(pal) <- c("CTree", "MOB", "GUIDE", "GUIDE_scores", "CTree_max")
+names(pal) <- c("CTree", "MOB", "GUIDE", "GUIDE+scores", "CTree+max")
 
@@
 \begin{document}
+
+\vspace*{0.4cm}
 
@@
 the $p$-values pertaining only to the true split variable $Z_1$;
-and the proportion of data sets for which the $p$-value of $Z_1$
+and the proportion of replications for which the $p$-value of $Z_1$
 is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
@@
 
-To illustrate the effects of the building blocks in the unifying framework presented in 
-Section~\ref{sec:unify} different combinations of them are investigated for the first
-simulation scenario (stump) in Sections ~\ref{sec:scores} and \ref{sec:3way}. 
-Then the testing strategies are evaluated on the more
-complex tree structure of the second scenario in Section~\ref{sec:tree}. 
-The corresponding results are shown in the following plots.
+% To illustrate the effects of the building blocks in the unifying framework presented in 
+% Section~\ref{sec:unify} different combinations of them are investigated for the first
+% simulation scenario (stump) in Sections~\ref{sec:scores} and~\ref{sec:3way}. 
+% Then the testing strategies are evaluated on the more
+% complex tree structure of the second scenario in Section~\ref{sec:tree}. 
+% The corresponding results are shown in the following plots.
 
@@
 \label{sec:scores}
+
 To assess the impact of using only residuals or the full score matrix four testing
-strategies are compared: CTree, GUIDE, MOB (all three in their original form),
-and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_scores).
+strategies are compared: CTree, GUIDE, MOB -- all three using the default specification
+as shown in Table~\ref{tab:buildingblocks} -- and additionally the GUIDE strategy using
+the full scores instead (denoted by GUIDE+scores).
 
@@
                        labels = c("CTree","MOB",
-                                  "GUIDE_scores", "GUIDE"))
+                                  "GUIDE+scores", "GUIDE"))
 
@@
 
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
 
@@
                                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
+                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                       layout= c(3,2),
@@
 
-               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
-                                                     expression("vary"~beta[0]), 
-                                                     expression("vary"~beta[1])),
+               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
+                                                     expression("varying"~beta[0]), 
+                                                     expression("varying"~beta[1])),
                                    bg = "gray90"),
@@
 @
-\caption{\label{fig:scores_stump}Proportion of cases where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ (100 data sets of 250 observations per step),
+\caption{\label{fig:scores_stump}Proportion of replications where the true split variable $Z_1$ is
+selected over increasing effect size $\delta$ (100 replications of 250 observations per step),
 once for a true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row). 
@@
 both coefficients varying (right). The performances of four testing strategies are compared: 
-CTree, MOB, GUIDE, GUIDE\_scores.}
+CTree, MOB, GUIDE, GUIDE+scores.}
 \end{figure}
@@
 As explained before, the tests are applied as significance tests with significance 
-level $\alpha=0.05$. Therefore, the proportion of cases in which the true split variable $Z_1$
+level $\alpha=0.05$. Therefore, the proportion of replications in which the true split variable $Z_1$
 is detected with a $p$-value smaller than $\alpha$ is compared.
@@
 regarding the intercept, such that changes in the slope parameter are not detected unless the 
-corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
+corresponding scores are also included as in GUIDE+scores. This disadvantage of using 
 only scores has already been elaborated in the literature of structural change
@@
 the mean but also in all other parameters can be detected in that way. This is particularly obvious
-when comparing GUIDE and GUIDE\_scores. Moreover, from the fact that these two version are both
+when comparing GUIDE and GUIDE+scores. Moreover, from the fact that these two version are both
 performing worse than MOB and CTree it can be deduced that categorizing split variables and
@@
 \hline
-Building Block         & Levels              & Effect \\
+Building block         & Levels              & Effect \\
 \hline
@@
 Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
-almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE\_scores performs
+almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE+scores performs
 slightly worse but it still achieves an ARI above $0.5$ for the highest considered effect size
@@
                        labels = c("CTree","MOB",
-                                  "GUIDE_scores", "GUIDE"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
+                                  "GUIDE+scores", "GUIDE"))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
 
@@
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
+                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
        layout= c(4,1),
@@
 \caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ 
-(100 data sets of 250 observations per step).
+(100 replications of 250 observations per step).
 Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
 $0.8$ from left to right). The performances of four testing strategies are compared: CTree,
-MOB, GUIDE, GUIDE\_scores.}
+MOB, GUIDE, GUIDE+scores.}
 \end{figure}
@@
                     labels = c("CTree","MOB",
-                               "GUIDE_scores", "GUIDE",
+                               "GUIDE+scores", "GUIDE",
                                "CTree_p","MOB_p",
-                               "GUIDE_scores_p", "GUIDE_p"))
+                               "GUIDE+scores_p", "GUIDE_p"))
 
@@
   if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
-  if(prex$test[i] == "GUIDE_scores") prex$test[i] <- "GUIDE_scores_p"
+  if(prex$test[i] == "GUIDE+scores") prex$test[i] <- "GUIDE+scores_p"
   if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
@@
 \caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
-(100 data sets of 250 observations per step).
+(100 replications of 250 observations per step).
 Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
 $0.8$ from left to right). 
-The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE\_scores. 
-The dashed lines represent the results from applying pre-pruning and the continuous lines 
+The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE+scores. 
+The dashed lines represent the results from applying pre-pruning and the solid lines 
 the results from applying post-pruning.}
@@
 Even though it also depends on the type of the applied residuals 
-(see, e.g., \cite{Loh+Zheng:2013}, \cite{Chaudhuri+Loh:2002} 
-and \cite{Loh+He+Man:2015})
+(see, e.g., \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002} 
+and \citealp{Loh+He+Man:2015})
 whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
@@
 
+\bibliography{ref.bib}
+
 \newpage
-\section*{Appendix}
+
 \begin{appendix}
@@
 \section{Combinations of building blocks applied}\label{app:combinations}
+
 For the evaluation of all 12 possible combinations of the presented building blocks
@@
 test as represented by the different building blocks. In that way,
-``\_res'' refers to using only residuals (instead of all scores),
-``\_scores'' to using all scores (instead of only residuals),
-``\_cat'' to applying a categorization of split variables,
-``\_bin'' to applying a binarization of scores/residuals.
-
-\begin{table}[t!]
+``+res'' refers to using only residuals (instead of all scores),
+``+scores'' to using all scores (instead of only residuals),
+``+cat'' to applying a categorization of split variables,
+``+bin'' to applying a binarization of scores/residuals.
+
+\begin{table}[h!]
 \centering
@@
 GUIDE             & residuals          & bin          & cat \\
-MOB\_bin\_res     & residuals          & bin          & max \\
-CTree\_bin\_res   & residuals          & bin          & lin \\
-MOB\_cat\_res    & residuals          & lin          & cat \\
-MOB\_res          & residuals          & lin          & max \\
-CTree\_res        & residuals          & lin          & lin \\
-GUIDE\_scores     & scores             & bin          & cat \\
-MOB\_bin          & scores             & bin          & max \\
-CTree\_bin        & scores             & bin          & lin \\
-MOB\_cat          & scores             & lin          & cat \\
+MOB+bin+res     & residuals          & bin          & max \\
+CTree+bin+res   & residuals          & bin          & lin \\
+MOB+cat+res    & residuals          & lin          & cat \\
+MOB+res          & residuals          & lin          & max \\
+CTree+res        & residuals          & lin          & lin \\
+GUIDE+scores     & scores             & bin          & cat \\
+MOB+bin          & scores             & bin          & max \\
+CTree+bin        & scores             & bin          & lin \\
+MOB+cat          & scores             & lin          & cat \\
 MOB               & scores             & lin          & max \\
@@
 \section{Significance level}\label{app:significance}
+
+\begin{figure}[t]
+\setkeys{Gin}{width=\linewidth}
+<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_max", "mfluc",
+                                  "guide_sum_12", "guide_sum_1_cor"),
+                       labels = c("CTree", "CTree+max", "MOB",
+                                  "GUIDE+scores", "GUIDE"))
+
+subdataxi <- subset(subdata, xi %in% c(0,0.8))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]))
+
+library("latticeExtra")
+useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
+                      data = subdataxi, 
+                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
+                        only_intercept == FALSE,# & xi == 0, 
+                      type = "l", lty = c(1,1,1,1,1),
+                      lwd = 2,
+                      scales=list(x=list(at=seq(1,11,2))),
+                      key =  list(text = list(levels(subdataxi$test)),
+                                  lines = list(lty = c(1,1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                               col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
+                      layout= c(3,2),
+                      index.cond = list(c(2,3,1),c(2,1)),
+                      xlab = expression(delta),
+                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
+                      par.settings = par.settings),
+
+               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
+                                                     expression("varying"~beta[0]), 
+                                                     expression("varying"~beta[1])),
+                                   bg = "gray90"),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
+                                   bg = "gray90"))
+
+@
+\caption{\label{fig:scores_stump_wosign}Proportion of replications where the true split variable $Z_1$ 
+shows the smallest $p$-value, hence, where $Z_1$ is selected without any significance level, 
+over increasing effect size $\delta$ (100 replications of 250 observations per step), once for a 
+true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row).
+Each column represents one scenario for the coefficients: 
+varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
+both coefficients varying (right). The performances of four testing strategies are compared: 
+CTree, MOB, GUIDE, GUIDE+scores.}
+\end{figure}
+
 In Section~\ref{sec:scores} the investigated testing strategies are compared based
-on the number of cases in which the true split variable $Z_1$ is detected with a $p$-value
+on the number of replications in which the true split variable $Z_1$ is detected with a $p$-value
 smaller than the predefined significance level $\alpha = 0.05$. This choice of measurement
@@
 and \ref{fig:scores_stump}.
-In Figure~\ref{fig:scores_stump_wosign} the proportion of cases 
+In Figure~\ref{fig:scores_stump_wosign} the proportion of replications 
 in which the true splitting variable $Z_1$ is detected, but not necessarily with a $p$-value 
@@
 
-\begin{figure}[t]
-\setkeys{Gin}{width=\linewidth}
-<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
-                                         "guide_sum_12", "guide_sum_1_cor"))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree","mfluc",
-                                  "guide_sum_12", "guide_sum_1_cor"),
-                       labels = c("CTree","MOB",
-                                  "GUIDE_scores", "GUIDE"))
-
-subdataxi <- subset(subdata, xi %in% c(0,0.8))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))
-
-library("latticeExtra")
-useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
-                      data = subdataxi, 
-                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
-                        only_intercept == FALSE,# & xi == 0, 
-                      type = "l", lty = c(1,1,1,1),
-                      lwd = 2,
-                      scales=list(x=list(at=seq(1,11,2))),
-                      key =  list(text = list(levels(subdataxi$test)),
-                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
-                      layout= c(3,2),
-                      index.cond = list(c(2,3,1),c(2,1)),
-                      xlab = expression(delta),
-                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
-                      par.settings = par.settings),
-
-               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
-                                                     expression("vary"~beta[0]), 
-                                                     expression("vary"~beta[1])),
-                                   bg = "gray90"),
-               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
-                                   bg = "gray90"))
-
-@
-\caption{\label{fig:scores_stump_wosign}Proportion of cases where the true split variable $Z_1$ 
-shows the smallest $p$-value, hence, where $Z_1$ is selected without any significance level, 
-over increasing effect size $\delta$ (100 data sets of 250 observations per step), once for a 
-true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row).
-Each column represents one scenario for the coefficients: 
-varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
-both coefficients varying (right). The performances of four testing strategies are compared: 
-CTree, MOB, GUIDE, GUIDE\_scores.}
-\end{figure}
-
 \section{Simulation results for continuous model parameters}
 \label{app:contbeta}
+
 In the simulation study presented in Sections~\ref{sec:simulation} and \ref{sec:results} the intercept
@@
 Moreover, a modified version of CTree considering maximum values in the test statistic is evaluated
-(denoted by CTree\_max). 
+(denoted by CTree+max). 
 Looking at the results presented in Figure~\ref{fig:stump_contbeta} it can be observed that in this setting the 
-original version of CTree is ahead while CTree\_max and MOB perform almost equally well. Hence, the comparison of 
+original version of CTree is ahead while CTree+max and MOB perform almost equally well. Hence, the comparison of 
 Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} points out that it clearly depends on the type of 
-effect whether the maximal selection (MOB, CTree\_max) or the linear selection (CTree) is advantageous.
+effect whether the maximal selection (MOB, CTree+max) or the linear selection (CTree) is advantageous.
 
@@
                        levels = c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"),
-                       labels = c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores"))
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")], 
-                                             fill = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")]), 
-                     superpose.line = list(col = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")]))
+                       labels = c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores"))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")], 
+                                             fill = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]), 
+                     superpose.line = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]))
 
@@
                                 type = "l", 
-                                col = pal[c("CTree","CTree_max","MOB", "GUIDE","GUIDE_scores")], 
+                                col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")], 
                                 lwd = 2)), 
@@
        index.cond = list(c(2,3,1)),
-       strip = strip.custom(factor.levels = c(expression("vary "~beta[0]~" and "~beta[1]),
-                                              expression("vary "~beta[0]), 
-                                              expression("vary "~beta[1])),
+       strip = strip.custom(factor.levels = c(expression("varying"~beta[0]~" and "~beta[1]),
+                                              expression("varying"~beta[0]), 
+                                              expression("varying"~beta[1])),
                                     bg = "gray90"),
@@
 \caption{\label{fig:stump_contbeta}
-Proportion of cases where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ (100 data sets of 250 observations per step)
+Proportion of replications where the true split variable $Z_1$ is
+selected over increasing effect size $\delta$ (100 replications of 250 observations per step)
 with the varying coefficient(s) being continuous. 
@@
 both coefficients varying (right). The performances of five testing strategies are compared: 
-CTree, CTree\_max, MOB, GUIDE, GUIDE\_scores.}
+CTree, CTree+max, MOB, GUIDE, GUIDE+scores.}
 \end{figure}
@@
 original version without a binarization and once in an adapted version with a binarization of 
-scores at $0$ (CTree\_bin, MOB\_bin). Moreover, they are compared to the adapted GUIDE version 
-which includes all available scores (GUIDE\_scores).
+scores at $0$ (CTree+bin, MOB+bin). Moreover, they are compared to the adapted GUIDE version 
+which includes all available scores (GUIDE+scores).
 %Hence, the five strategies are obtained from the unifying framework by setting the switches 
@@
 lead to the same conclusions. Binarizing the score values decreases the proportion
-of cases where the split variable was selected correctly.
+of replications where the split variable was selected correctly.
 
@@
                        labels = c("CTree","CTree_bin",
-                                  #"CTree_max","CTree_max_bin",
+                                  #"CTree+max","CTree+max_bin",
                                   "MOB", "MOB_bin",
-                                  "GUIDE_scores"))
+                                  "GUIDE+scores"))
 
 par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
-                                                         #"CTree_max", "CTree_max", 
+                                                         #"CTree+max", "CTree+max", 
                                                          "MOB", "MOB",
-                                                         "GUIDE_scores")], 
+                                                         "GUIDE+scores")], 
                                              fill = pal[c("CTree", "CTree", 
-                                                          #"CTree_max", "CTree_max", 
+                                                          #"CTree+max", "CTree+max", 
                                                           "MOB", "MOB",
-                                                          "GUIDE_scores")]), 
+                                                          "GUIDE+scores")]), 
                      superpose.line = list(col = pal[c("CTree", "CTree", 
-                                                       #"CTree_max", "CTree_max", 
+                                                       #"CTree+max", "CTree+max", 
                                                        "MOB", "MOB",
-                                                       "GUIDE_scores")]))
+                                                       "GUIDE+scores")]))
 
@@
                                 col = pal[c("CTree", "CTree", 
-                                            #"CTree_max", "CTree_max", 
+                                            #"CTree+max", "CTree+max", 
                                             "MOB", "MOB",
-                                            "GUIDE_scores")], lwd = 2)), 
+                                            "GUIDE+scores")], lwd = 2)), 
        #index.cond = list(c(1,2,3,4)),
@@
 @
-\caption{\label{fig:bin_stump}Proportion of cases where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ (100 data sets of 250 observations per step)
+\caption{\label{fig:bin_stump}Proportion of replications where the true split variable $Z_1$ is
+detected over increasing effect size $\delta$ (100 replications of 250 observations per step)
 with both parameter varying. Each panel represents a different value of the true 
 split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
-strategies are compared: CTree, CTree\_bin, MOB, MOB\_bin, GUIDE\_scores.}
+strategies are compared: CTree, CTree+bin, MOB, MOB+bin, GUIDE+scores.}
 \end{figure}
@@
 applied, once in their original version without a categorization and once in an adapted version
-with a categorization of the possible split variables (CTree\_cat and MOB\_cat). 
-Moreover, they are also compared to GUIDE\_scores which includes all available scores.
+with a categorization of the possible split variables (CTree+cat and MOB+cat). 
+Moreover, they are also compared to GUIDE+scores which includes all available scores.
 
@@
 is close to the quartiles used as breakpoints for the categorizations both versions
-lead to a similar proportion of cases where the true splitting variable is detected
+lead to a similar proportion of replications where the true splitting variable is detected
 (e.g., for CTree and $\xi=0.5$)
@@
                        labels = c("CTree","CTree_cat",
-                                  #"CTree_max","CTree_max_cat",
+                                  #"CTree+max","CTree+max_cat",
                                   "MOB", "MOB_cat",
-                                  "GUIDE_scores"))
+                                  "GUIDE+scores"))
 
 par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
-                                                         #"CTree_max", "CTree_max", 
+                                                         #"CTree+max", "CTree+max", 
                                                          "MOB", "MOB",
-                                                         "GUIDE_scores")], 
+                                                         "GUIDE+scores")], 
                                              fill = pal[c("CTree", "CTree", 
-                                                          #"CTree_max", "CTree_max", 
+                                                          #"CTree+max", "CTree+max", 
                                                           "MOB", "MOB",
-                                                          "GUIDE_scores")]), 
+                                                          "GUIDE+scores")]), 
                      superpose.line = list(col = pal[c("CTree", "CTree", 
-                                                       #"CTree_max", "CTree_max", 
+                                                       #"CTree+max", "CTree+max", 
                                                        "MOB", "MOB",
-                                                       "GUIDE_scores")]))
+                                                       "GUIDE+scores")]))
 
@@
 @
-\caption{\label{fig:cat_stump}Proportion of cases where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ (100 data sets of 250 observations per step)
+\caption{\label{fig:cat_stump}Proportion of replications where the true split variable $Z_1$ is
+detected over increasing effect size $\delta$ (100 replications of 250 observations per step)
 with both parameter varying. Each panel represents a different value of the true 
 split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
-strategies are compared: CTree, CTree\_cat, MOB, MOB\_cat, GUIDE\_scores.}
+strategies are compared: CTree, CTree+cat, MOB, MOB+cat, GUIDE+scores.}
 \end{figure}
@@
 
-%\newpage
-\bibliography{ref.bib}
-
 \end{document}

### 2019-05-12 pkg/partykit/inst/guideliketest/draft/draft.Rnw
rewrote sections 5.1. and 5.2. and edited captions of section 5.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2553 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7d65c57349137404aee06a59161a71fced4ca478

@@
         Rather than testing independence between the split variables and the residuals/scores
-	tmeselves, it is possible	
+	themselves, it is possible	
 	to binarize residuals/scores at $0$ so that only their signs are assessed (as proposed for GUIDE).
@@
 To compare the performance of the evaluated testing strategies for each step 
-in the first scenario (stump with 1 split) the following criteria are considered:
+in the first scenario (stump with one split) the following criteria are considered:
 the $p$-values pertaining only to the true split variable $Z_1$;
@@
 selected for splitting in a pre-pruning approach).
-For the second scenario (tree with 2 splits) the adjusted Rand index~(ARI) is calculated 
+For the second scenario (tree with two splits) the adjusted Rand index~(ARI) is calculated 
 as a measure of similarity between the true tree structure and the fitted model tree.
@@
 @
-\caption{\label{fig:scores_stump}Proportion of replications where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ (100 replications of 250 observations per step),
-once for a true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row). 
-Each column represents one scenario for the coefficients: 
-varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
-both coefficients varying (right). The performances of four testing strategies are compared: 
-CTree, MOB, GUIDE, GUIDE+scores.}
+\caption{\label{fig:scores_stump}
+CTree, MOB, GUIDE, and GUIDE+scores evaluated on the first simulation scenario (stump with one split). 
+Their performances are compared based on the proportion of replications where the true split 
+variable $Z_1$ is selected over increasing effect size $\delta$ (100 replications of 
+250 observations per step), once for a true split point $\xi = 0$ (top row) and once 
+for $\xi = 0.8$ (bottom row). 
+Left: varying intercept $\beta_0$. Middle: varying slope parameter $\beta_1$.
+Right: both coefficients varying.}
 \end{figure}
 
-Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
-scenarios (varying only the intercept $\beta_0$, only the slope parameter $\beta_1$
-or both of them) once for a true split point $\xi=0$ and once for $\xi=0.8$. 
-As explained before, the tests are applied as significance tests with significance 
-level $\alpha=0.05$. Therefore, the proportion of replications in which the true split variable $Z_1$
-is detected with a $p$-value smaller than $\alpha$ is compared.
-(To show that in general the drawn conclusions do not change if no significance level is included 
-the same setting is evaluated again without any significance level and the results are presented
-in Appendix~\ref{app:significance}.)
-
-For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
-intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
-this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
-(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
-in $\beta_1$. 
-This is due to the fact that GUIDE only includes residuals, i.e., only the scores 
-regarding the intercept, such that changes in the slope parameter are not detected unless the 
-corresponding scores are also included as in GUIDE+scores. This disadvantage of using 
-only scores has already been elaborated in the literature of structural change
-analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
-in Section~\ref{sec:discussion}.
-
-Looking at the results for the split point being close to the upper edge of the considered
-interval ($\xi=0.8$, lower panels) all of the applied strategies perform worse than for the
-central split point. In this case MOB is clearly ahead leading to the best results. CTree 
-performs second best while both GUIDE versions struggle to detect the correct split variable 
-even for a high effect size. 
+Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
+model scores) perform clearly better than the original GUIDE version (using residuals 
+only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
+of replications. Even though the advantage of CTree and MOB evens out in certain situations
+such as for a high effect size (see, e.g., for $\xi=0$ and varying $\beta_0$), GUIDE is 
+never ahead of them in any of the considered settings. Moreover, the positive effect of 
+including all available scores can be observed particularly well in the top middle panel 
+where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
+GUIDE. The fact that residual-based methods fail to detect changes that do not effect the
+expected value has already been elaborated in the literature of structural change
+analysis \citep[see, e.g.,][]{Ploberger+Kramer:1992} and is also discussed in 
+Section~\ref{sec:discussion}. Moreover, the advantage of including full models scores 
+clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
+to GUIDE+scores in case of a fixed slope parameter (left panels).
+
+For a central split point ($\xi=0$, upper panels) all strategies perform similarly
+well as long as the intercept varies (left and right panel) with CTree and MOB 
+being only slightly ahead. However, for the split point being close to the upper 
+edge of the considered interval ($\xi=0.8$, lower panels) the performance of all of 
+the applied strategies decreases. 
+While both GUIDE versions struggle to detect the correct split variable even for a 
+high effect size, MOB is clearly ahead with CTree leading to the second best results.
 (In order to show that the here observed advantage of MOB over CTree is mainly due to the 
@@
 presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)
-
-Altogether these results clearly illustrate that overall the decision to include all available
-scores when testing for dependencies in order to select split variables is advantageous over
-considering only residuals, i.e., the scores regarding only the intercept, as changes not only in
-the mean but also in all other parameters can be detected in that way. This is particularly obvious
-when comparing GUIDE and GUIDE+scores. Moreover, from the fact that these two version are both
-performing worse than MOB and CTree it can be deduced that categorizing split variables and
-binarizing scores has a negative effect on the ability to detect the right split variable. 
-To elaborate these effects more in detail the results for all possible combinations of the 
-building blocks are discussed in a full factorial 3-way analysis.
+From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
+can be deduced that categorizing split variables and binarizing scores has a negative effect 
+on the ability to detect the right split variable.
+To elaborate these effects more in detail the results for all possible combinations of the building blocks are discussed in a full factorial 3-way analysis in the following section.
+
+\textit{Note:} As explained before, the tests are applied as significance tests with 
+significance level $\alpha=0.05$. To show that in general the drawn conclusions do not 
+change if no significance level is included the same setting is evaluated again without 
+any significance level and the results are presented in Appendix~\ref{app:significance}.
 
 \subsection{Full factorial analysis}\label{sec:3way}
-In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
-case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e., $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
-This choice has been made based on the results shown in Figure~\ref{fig:scores_stump} 
-in order to make the differences particularly well observable.
-In the right panel the true split point is close to the upper end of the considered interval
-($\xi=0.8$) requiring a higher effect size $\delta = 1$ in order to visualize results properly.
-(Additional evaluations for increasing $\delta$ as in Section~\ref{sec:scores} 
-can be found in Appendix~\ref{app:increasingdelta}).
-The power of the different combinations is measured by the mean $p$-value for the true split 
-variable~$Z_1$.
-The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
-In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
-and variations of them. A more detailed overview of all 12 combinations can be found in 
-Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.
+To investigate the effects of each of the building blocks separately the most general 
+case of the first simulation scenario (stump with one split) where the intercept and 
+slope parameter are both varying is considered. In the underlying evaluation all 
+possible combinations of the building blocks have been included. The different levels of 
+each of the three building blocks are listed in Table~\ref{tab:levels}. A more detailed
+overview of all 12 combinations can be found in Appendix~\ref{app:combinations}, 
+Table~\ref{tab:combinations_appendix}.
 
@@
 \end{table}
+
+Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing residuals/scores decreases the performance 
+as it leads to higher $p$-values corresponding to the true split variable $Z_1$ 
+and thus to lower power in the presented settings. 
+For a central true split point (left panel) this effect remains at an almost constant level 
+regardless of whether possible split variables are categorized or not. However, in case 
+of the true split point being closer to the upper margin ($\xi=0.8$, right panel), categorizing
+split variables increases the $p$-value of $Z_1$ even more. 
+Here, the maximal selection performs best while the linear selection shows its disadvantage in 
+detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximal
+selection of a split variable is advantageous. But unless one of the breakpoints used for binning possible split variables is close to the true split point (as for $\xi=0$, left panel),
+this categorization weakens the performance of a testing strategy.
+
+While the effect of both of these transformations of the data (binarization of 
+residuals/scores and categorization of split variables) can be observed
+separately in Figure~\ref{fig:3way}, combining them increases the negative impact 
+remarkably.
+As already shown in Section~\ref{sec:scores} the use of scores vs.\ residuals has a minor 
+effect if there is a change in the intercept which is
+supported by the small differences between the dashed lines (scores) and the solid
+lines (residuals) in Figure~\ref{fig:3way}.
+
+Note that the choice of the two specific settings 
+(left panel: true split point $\xi=0$,  effect size $\delta = 0.3$; 
+right panel: $\xi=0.8$ requiring a higher effect size $\delta = 1$)
+has been made based on the results shown in Figure~\ref{fig:scores_stump} in order to make 
+the differences particularly well observable. (Additional evaluations for increasing $\delta$ 
+as in Section~\ref{sec:scores} can be found in Appendix~\ref{app:increasingdelta}.)
 
@@
 @
-\caption{\label{fig:3way} Mean $p$-values corresponding to the true split variable $Z_1$
-over different levels of the building block \textit{Categorization}. 
-Each line represents
-a combination of the levels of the other two building blocks \textit{Binarization} and 
-\textit{Residuals vs.\ Scores}.
+\caption{\label{fig:3way} Effects of the three building blocks \textit{residuals vs.\ scores},
+\textit{binarization}, and \textit{categorization}. All possible combinations have been 
+evaluated in two different settings of the first simulation study (stump with one split) 
+and their performances are compared based on the mean $p$-values corresponding to the 
+true split variable $Z_1$. 
 Left panel: true split point $\xi=0$ and effect size $\delta = 0.3$.
@@
 \end{figure}
-
-The results presented in Figure~\ref{fig:3way} clearly show that binarizing scores/residuals
-reduces the power of the testing strategy as it leads to higher $p$-values corresponding
-to the true split variable $Z_1$. 
-This effect can be observed for a central split point (left panel, $\xi=0$) as well as for a
-a split point closer to upper end of the considered interval (right panel, $\xi=0.8$).
-However, the effect of the building block \textit{Categorization} varies over these two settings.
-For a central split point all three levels perform similarly. The linear selection of split variables
-is slightly ahead even though the true parameter function is a step function. 
-On the contrary, clear differences can be seen for the outward split point. Here, the maximal selection
-performs best while the linear selection shows its disadvantage in detecting abrupt changes
-and the categorized split variables lead to even higher $p$-values and thus lower power.
-Overall, it depends on the situation whether a linear or a maximal selection of a split variable
-is advantageous. However, categorizing possible split variables does not lead to any improvements
-but even declines the performance of a testing strategy unless the true split point is close
-to a break point used for the binarization.
 
@@
 @
-\caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ 
-(100 replications of 250 observations per step).
+\caption{\label{fig:tree_ari}
+CTree, MOB, GUIDE, and GUIDE+scores evaluated on the second simulation scenario (tree with two splits), all
+applying pre-pruning. Their performances are compared based on the Adjusted Rand Index over
+increasing effect size $\delta$ (100 replications of 250 observations per step).
 Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
-$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
-MOB, GUIDE, GUIDE+scores.}
+$0.8$ from left to right).}
 \end{figure}
@@
 @
-\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
-(100 replications of 250 observations per step).
+\caption{\label{fig:pruning_ari}
+CTree, MOB, GUIDE, and GUIDE+scores evaluated on the second simulation scenario (tree), 
+once applying pre-pruning (dashed lines) and once post-pruning (solid lines). 
+Their performances are compared based on the Adjusted Rand Index over increasing effect 
+size $\delta$ (100 replications of 250 observations per step).
 Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
-$0.8$ from left to right). 
-The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE+scores. 
-The dashed lines represent the results from applying pre-pruning and the solid lines 
-the results from applying post-pruning.}
+$0.8$ from left to right).}
 \end{figure}
@@
 with modified versions of them. The employed versions and the corresponding combination
-of building blocks is listed in Table~\ref{tab:combinations_appendix}.
+of building blocks are listed in Table~\ref{tab:combinations_appendix}.
 Each suffix indicates a change of the original form of the considered testing strategy.

### 2019-05-13 pkg/partykit/inst/guideliketest/draft/draft.Rnw
Kramer -> Kraemer
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2554 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1f450d9448bba5f693d63a378f8be08f3e18ba7c

@@
 expected value has already been elaborated in the literature of structural change
-analysis \citep[see, e.g.,][]{Ploberger+Kramer:1992} and is also discussed in 
+analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
 Section~\ref{sec:discussion}. Moreover, the advantage of including full models scores 
@@
 in detail in previous literature.
-\cite{Ploberger+Kramer:1992} have shown that residual-based tests can detect a change in the parameters
+\cite{Ploberger+Kraemer:1992} have shown that residual-based tests can detect a change in the parameters
 only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
@@
 shift and the type of model. Therefore, a general statement for all types of models and residuals can 
-hardly be made. However, the results from previous literature such as \cite{Ploberger+Kramer:1992} and
+hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
 the here presented simulation show that considering only residuals provides small gains in certain 

### 2019-05-17 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added section in the appendix on the test statistics of CTree, MOB, and GUIDE
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2572 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2b4f72d8595a97b0ff2b2ea20e7783e2db0f11b1

@@
 
+\section{Test statistics}
+This section provides further information on the test statistics applied in CTree, MOB, 
+and GUIDE. The presented details have been extracted from the original papers 
+\cite{Hothorn+Hornik+Zeileis:2006}, \cite{Zeileis+Hothorn+Hornik:2008}, 
+and \cite{Loh:2002}, however, notation is adapted to the main manuscript
+in order to allow for better comparison.
+
+\subsection{CTree}
+To measure the association of a response $Y$ and each possible split variable $Z$ the
+CTree algorithm applies a linear test statistic $T$ which is excerpted from Section~3.1.
+(``Variable selection and stopping criteria'') of the original paper and is of the following form:
+$$
+T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i, (Y_1,\ldots,Y_n))^{\top}\right)
+$$
+where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom transformation of the split variable $Z$ and the influence function $h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the responses $(Y_1, \ldots, Y_n)$ in a permutation symmetric way.
+Moreover, by applying the ``vec'' operator the resulting $p~\times~q$ matrix is converted into a $p q$ column vector.
+Following \cite{Strasser+Weber:1999}
+the conditional expectation $\mu$ and covariance $\Sigma$ of a linear test statistic can 
+be calculated and used to standardize an observed linear test statistic $t$ within a 
+function $c$ mapping into the real line.
+For example, the maximum of the absolute values of the standardized linear statistic
+$$
+c_{\max}(t,\mu,\Sigma) = \max_{k=1,\ldots,pq} \left| \frac{(t-\mu)}{\sqrt{(\Sigma)_{kk}}} \right|
+$$
+or a quadratic form
+$$
+c_{\text{quad}}(t, \mu, \Sigma) = (t - \mu)\Sigma^{+} (t - \mu)^{\top}
+$$
+can be considered where $\Sigma^{+}$ is the Moore-Penrose inverse of $\Sigma$.
+Since the asymptotic conditional distribution of a linear test statistic $T$ is a 
+multivariate normal with parameters $\mu$ and $\Sigma$ \citep{Strasser+Weber:1999}, 
+the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
+$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
+the corresponding $p$-values can be calculated easily.
+
+% FIXME: adapt notation from here on (split variable Z and response Y, scores s)
+\subsection{MOB}
+The MOB algorithm employs an empirical fluctuation process $W$ to measure deviations of
+the model scores $s$ from zero. As described in detail in Section~3.2. (``Testing for 
+parameter instability'') of the original paper this process is of the following form:
+$$
+W(t) = \hat{J}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s_{\sigma(Z_{i})} \qquad (0 \leq t \leq 1)
+$$
+where the model scores $s$ are sorted by the split variable Z by including the ordering permutation $\sigma(Z_{i})$. 
+To scale this partial sum process an estimate $\hat{J}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
+Following \cite{Zeileis+Hornik:2007} $W(t)$ converges to a Brownian bridge under the null
+hypothesis of parameter stability. To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied.
+For more details, particularly on calculating the corresponding $p$-values see Section~3.2. of the original paper.
+
+\subsection{GUIDE}
+The test statistic of the $\chi^2$~test as applied in the GUIDE algorithm is
+$$
+X^2 = \frac{(O_{11} - E_{11})^2}{E_{11}} + \ldots + \frac{(O_{11} - E_{11})^2}{E_{11}}
+$$
+where $O_{ij}$ are the observed frequencies of observations in a certain combination of binarized residuals ($i=1,2$) and categorized split variables ($j = 1,\ldots,4$).
+$E_{ij}$ are the corresponding expected frequencies under independence $E_{ij} = \frac{O_{i \cdot} \cdot O_{\cdot j}}{n}$.
+Under the null hypothesis of independence the asymptotic distribution of $X^2$ is a $\chi^2$~distribution allowing for a straight-forward calculation of $p$-values.
+If full model scores are used instead of residuals a sum of the test statistic over the number of distribution parameters is considered such that each 
+summand corresponds to one column of the score matrix.
+
 \section{Combinations of building blocks applied}\label{app:combinations}

### 2019-05-19 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions as discussed (small changes in figures and captions, tables 3 and 4, formula for beta, consistent use of unifying terms for scenarios, structure of section 5, and some text revisions in sections 4 and 5), still to do: further text revision in section 5
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2575 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cc4b21dc7789b8a0c13678111c8619bed8200665

@@
 %% packages
-\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
+\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref,dsfont}
 \usepackage[all]{hypcap}
@@
 
+\subsubsection{``Stump'' scenario}
 Each generated data set consists of the response and regressor variables, one true split 
@@
 Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
-True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
-Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
+True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$\\
+Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
 \hline
@@
 \end{tabular}
-\caption{Variables included in the data generating process of 
-the first scenario (stump).
-\label{tab:variables}}
+\caption{\label{tab:variables}Variables included in the data generating process 
+as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a split variable, 
+not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_i(Z_1,Z_2)$, $i=1,2$.}
 \end{table}
@@
 Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
-can depend on the true split variable $Z_1$. More specifically, three different scenarios
+can depend on the true split variable $Z_1$. More specifically, three different variations
 are considered for the coefficients $\beta_0$ and $\beta_1$:
@@
 \beta_{i}(Z_1) = \begin{cases}
--\delta \cdot (-1)^{j}  \quad \text{if } Z_1 < \xi \\
-+\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \geq \xi
+-\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \leq \xi \\
++\delta \cdot (-1)^{j}  \quad \text{if } Z_1 > \xi
 \end{cases}
@@
 \endminipage
-\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$
+\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
 and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
@@
 
-For the second simulation scenario the same basic setting with the same variables is used
-with an additional split in $Z_2$ resulting in a tree with three terminal nodes. The first
-split (in $Z_2$, at~$\xi$) induces a change in the slope parameter $\beta_1$ while the second
-split (in $Z_1$, also at~$\xi$) corresponds to a change in the intercept $\beta_0$.
+\subsubsection{``Tree'' scenario}
+For the second simulation scenario (``tree'' scenario) the same basic setting with the same 
+variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
+However, not only $Z_1$ but also $Z_2$ is used as a split variable. Therefore, an additional split
+is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
+change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
+to a change in the intercept $\beta_0$.
+Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
+$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
+the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
+%
+\begin{align*}
+\beta_{0}(Z_1, Z_2) = \begin{cases}
+0       \qquad \text{if } Z_2 \leq \xi \\
+-\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
++\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
+\end{cases}
+\end{align*}
+\begin{align*}
+\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
++\delta \quad \text{if } Z_2 \leq \xi \\
+-\delta \quad \text{if } Z_2 > \xi
+\end{cases}
+\end{align*}
+%
 The underlying tree structure for the data generating process is illustrated in 
@@
 \endminipage
-\caption{\label{fig:dgp_tree}Left panel: True tree structure for the second simulation scenario 
+\caption{\label{fig:dgp_tree}Left panel: True tree structure for the ``tree'' scenario 
 with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
@@
 To compare the performance of the evaluated testing strategies for each step 
-in the first scenario (stump with one split) the following criteria are considered:
+in the ``stump'' scenario the following criteria are considered:
 the $p$-values pertaining only to the true split variable $Z_1$;
@@
 is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
-selected for splitting in a pre-pruning approach).
-For the second scenario (tree with two splits) the adjusted Rand index~(ARI) is calculated 
+selected for splitting in a pre-pruning approach), denoted by the 
+``selection probability''.
+For the ``tree'' scenario the adjusted Rand index~(ARI) is calculated 
 as a measure of similarity between the true tree structure and the fitted model tree.
@@
 
+In the following we first investigate the properties of the testing
+strategies in the ``stump'' scenario, focusing on the tests' $p$-values
+for the true split variable $Z_1$ and its corresponding selection
+probability (i.e., $Z_1$ being significant at 5\% level and having the
+lowest $p$-value among all split variables). Section~\ref{sec:scores} begins by
+highlighting the importance of using full model scores vs.\ residuals only
+before Section~\ref{sec:3way} considers all building blocks (scores vs.\ residuals,
+binarization of these, and categorization of the split variable).
+Subsequently, the ``tree'' scenario is employed to investigate how the
+performance of the tests affects growing the trees overall. This is
+evaluated using the adjusted Rand index for trees grown by pre-pruning and 
+cost-complexity post-pruning (Section~\ref{sec:tree}).
+
 % To illustrate the effects of the building blocks in the unifying framework presented in 
@@
 
-\subsection[Residuals vs. full model scores]{Residuals vs.\ full model scores}
+\subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
 \label{sec:scores}
 
-To assess the impact of using only residuals or the full score matrix four testing
-strategies are compared: CTree, GUIDE, MOB -- all three using the default specification
-as shown in Table~\ref{tab:buildingblocks} -- and additionally the GUIDE strategy using
-the full scores instead (denoted by GUIDE+scores).
+A crucial difference between the testing strategies of CTree/MOB and GUIDE
+is the difference between using only the residuals vs. the full model
+scores. In the literature on structural change tests there is evidence
+that residual-based tests can only capture parameter differences that
+affect the conditional mean \citep[see e.g.,][]{Ploberger+Kraemer:1992}.
+Hence we compare the CTree, MOB, and GUIDE algorithms 
+-- all three using the default specification as shown in Table~\ref{tab:buildingblocks} -- 
+and additionally consider a new GUIDE flavor, denoted GUIDE+scores, that uses
+binarized scores rather than residuals.
+
+%To assess the impact of using only residuals or the full score matrix four testing
+%strategies are compared: CTree, GUIDE, MOB -- all three using the default specification
+%as shown in Table~\ref{tab:buildingblocks} -- and additionally the GUIDE strategy using
+%the full scores instead (denoted by GUIDE+scores).
 
@@
                                    bg = "gray90"),
-               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0~" (50%)"),                                                              expression(xi == 0.8~" (90%)" )),
                                    bg = "gray90"))
@@
 \caption{\label{fig:scores_stump}
-CTree, MOB, GUIDE, and GUIDE+scores evaluated on the first simulation scenario (stump with one split). 
-Their performances are compared based on the proportion of replications where the true split 
-variable $Z_1$ is selected over increasing effect size $\delta$ (100 replications of 
-250 observations per step), once for a true split point $\xi = 0$ (top row) and once 
-for $\xi = 0.8$ (bottom row). 
-Left: varying intercept $\beta_0$. Middle: varying slope parameter $\beta_1$.
-Right: both coefficients varying.}
+Selection probability of the true split variable $Z_1$ for
+testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the ``stump''
+scenario. Probabilities are estimated over increasing effect size $\delta$
+with 100 replications of 250 observations per step. The true split point
+in $Z_1$ is either the median 0 (top) or the 90\% quantile 0.8 (bottom)
+for either varying intercept $\beta_0$ (left), varying slope $\beta_1$
+(middle), or both coefficients varying (right).}
 \end{figure}
@@
 of replications. Even though the advantage of CTree and MOB evens out in certain situations
-such as for a high effect size (see, e.g., for $\xi=0$ and varying $\beta_0$), GUIDE is 
+such as for a high effect size (see, e.g., for the median $\xi=0$ and varying $\beta_0$), GUIDE is 
 never ahead of them in any of the considered settings. Moreover, the positive effect of 
@@
 where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
-GUIDE. The fact that residual-based methods fail to detect changes that do not effect the
-expected value has already been elaborated in the literature of structural change
-analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
-Section~\ref{sec:discussion}. Moreover, the advantage of including full models scores 
+GUIDE. 
+%The fact that residual-based methods fail to detect changes that do not effect the
+%expected value has already been elaborated in the literature of structural change
+%analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
+%Section~\ref{sec:discussion}. 
+Furthermore, the advantage of including full models scores 
 clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
@@
 
-For a central split point ($\xi=0$, upper panels) all strategies perform similarly
-well as long as the intercept varies (left and right panel) with CTree and MOB 
+For a central split point ($\xi=0$ corresponding to the median, upper panels) all strategies 
+perform similarly well as long as the intercept varies (left and right panel) with CTree and MOB 
 being only slightly ahead. However, for the split point being close to the upper 
-edge of the considered interval ($\xi=0.8$, lower panels) the performance of all of 
-the applied strategies decreases. 
+edge of the considered interval ($\xi=0.8$ corresponding to the $90$\%-quantile, lower panels) 
+the performance of all of the applied strategies decreases. 
 While both GUIDE versions struggle to detect the correct split variable even for a 
@@
 
-\subsection{Full factorial analysis}\label{sec:3way}
+\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
 To investigate the effects of each of the building blocks separately the most general 
-case of the first simulation scenario (stump with one split) where the intercept and 
+case of the ``stump'' scenario where the intercept and 
 slope parameter are both varying is considered. In the underlying evaluation all 
@@
 \hline
-Building block         & Levels              & Effect \\
+Building block               & Levels              & Transformation \\
 \hline
-Residuals vs.\ scores  & \textit{residuals}  & Use only residuals\\
-                       & \textit{scores}     & Use full score matrix\\
-Categorization         & \textit{cat}        & Categorize split variables\\
-                       & \textit{max}        & Selected split variables maximally w.o. categorization\\
-                       & \textit{lin}        & Selected split variables linearly w.o. categorization\\
-Binarization           & \textit{bin}        & Binarize scores\\
-                       & \textit{lin}        & Use residuals/score values without binarization\\
+Residuals vs.\ scores        & \textit{residuals}  & $r(Y,X,\hat{\beta})$\\
+\quad ($h$ transformation)   & \textit{scores}     & $s(Y,X,\hat{\beta})$\\
+\hline
+Binarization                 & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
+\quad (of $h$)               & \textit{no}         & $h$ without binarization\\
+\hline
+Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)$\\
+\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
+                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)$\\
+                             &                     & \quad indicating the crossed quantile\\
+                             & \textit{lin}        & $g(Z)=Z$\\
 \hline
@@
 regardless of whether possible split variables are categorized or not. However, in case 
-of the true split point being closer to the upper margin ($\xi=0.8$, right panel), categorizing
-split variables increases the $p$-value of $Z_1$ even more. 
+of the true split point being closer to the upper margin ($90$\%-quantile $\xi=0.8$, right panel),
+categorizing split variables increases the $p$-value of $Z_1$ even more. 
 Here, the maximal selection performs best while the linear selection shows its disadvantage in 
 detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximal
-selection of a split variable is advantageous. But unless one of the breakpoints used for binning possible split variables is close to the true split point (as for $\xi=0$, left panel),
+selection of a split variable is advantageous. But unless one of the breakpoints used for binning possible split variables is close to the true split point (as for the median $\xi=0$, left panel),
 this categorization weakens the performance of a testing strategy.
@@
 Note that the choice of the two specific settings 
-(left panel: true split point $\xi=0$,  effect size $\delta = 0.3$; 
-right panel: $\xi=0.8$ requiring a higher effect size $\delta = 1$)
+(left panel: true split point at the median $\xi=0$,  effect size $\delta = 0.3$; 
+right panel: true split point at the $90$\%-quantile $\xi=0.8$ requiring a higher effect 
+size $\delta = 1$)
 has been made based on the results shown in Figure~\ref{fig:scores_stump} in order to make 
@@
        lwd = 2,
+       scales = list(alternating=FALSE),
        xlab = "Categorization",
@@
        #index.cond = list(c(1,2,3,4)),
-       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
-                                              expression(xi==0.8~"and"~delta==1)),
+       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
+                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
                                    bg = "gray90"),
-       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
-                  lines = list(lty = c(1,1,2,2), 
+       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Binarization:", "  yes", "  no")),
+                  columns = 2,
+                  lines = list(lty = c(0,1,2,0,1,1), 
                                type = "l", lwd = 2,
-                               col = c(pal[1], pal[3], pal[1], pal[3]))))
+                               col = c("black","black","black","black", pal[1], pal[3]))))
 
@@
 \caption{\label{fig:3way} Effects of the three building blocks \textit{residuals vs.\ scores},
-\textit{binarization}, and \textit{categorization}. All possible combinations have been 
-evaluated in two different settings of the first simulation study (stump with one split) 
-and their performances are compared based on the mean $p$-values corresponding to the 
-true split variable $Z_1$. 
-Left panel: true split point $\xi=0$ and effect size $\delta = 0.3$.
-Right panel: true split point $\xi=0.8$ and effect size $\delta = 1$.}
+\textit{binarization}, and \textit{categorization} in the ``stump'' scenario with the intercept
+and the slope parameter both varying. All possible combinations have been 
+evaluated in two different settings and their performances are compared based on the 
+mean $p$-values corresponding to the true split variable $Z_1$. 
+Left panel: true split point at the median $\xi=0$ and 
+effect size $\delta = 0.3$.
+Right panel: true split point at the $90$\%-quantile $\xi=0.8$ and effect 
+size $\delta = 1$.}
 \end{figure}
@@
 %\newpage
-\subsection{Tree}
+\subsection[``Tree'' scenario: Pre-pruning vs. post-pruning]{``Tree'' scenario: Pre-pruning vs.\ post-pruning}
 \label{sec:tree}
@@
 slope parameter as it only considers residuals in its testing strategy.
-For a split point farther away from the center such as for $\xi = 0.8$ MOB outperforms all other
+For a split point farther away from the center such as for the $90$\%-quantile $\xi = 0.8$ MOB 
+outperforms all other
 methods and both GUIDE versions fail to find the correct tree structure.
@@
 \caption{\label{fig:tree_ari}
-CTree, MOB, GUIDE, and GUIDE+scores evaluated on the second simulation scenario (tree with two splits), all
-applying pre-pruning. Their performances are compared based on the Adjusted Rand Index over
-increasing effect size $\delta$ (100 replications of 250 observations per step).
+Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
+``tree'' scenario, all applying pre-pruning. The ARI is evaluated over
+increasing effect size $\delta$ with 100 replications of 250 observations per step.
 Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
@@
 \end{figure}
-
-\subsection[Pre-pruning vs. post-pruning]{Pre-pruning vs.\ post-pruning}
 
@@
 In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with pre-pruning,
-again based on the Adjusted Rand Index.
+again based on the adjusted Rand index.
 
@@
 \caption{\label{fig:pruning_ari}
-CTree, MOB, GUIDE, and GUIDE+scores evaluated on the second simulation scenario (tree), 
-once applying pre-pruning (dashed lines) and once post-pruning (solid lines). 
-Their performances are compared based on the Adjusted Rand Index over increasing effect 
-size $\delta$ (100 replications of 250 observations per step).
+Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
+``tree'' scenario, once applying pre-pruning (dashed lines) and once post-pruning (solid lines). 
+The ARI is evaluated over
+increasing effect size $\delta$ with 100 replications of 250 observations per step.
 Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
 $0.8$ from left to right).}
+\end{figure}
+
+<<tree_all_data, echo=FALSE, results=hide>>=
+if(file.exists("pruning.rda")){
+  load("pruning.rda")
+} else {
+  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7, 
+                         nrsteps = 1, nrlevels = 2,
+                         delta = seq(from = 0, to = 1, by = 0.1),
+                         xi = c(0, 0.2, 0.5, 0.8), 
+                         vary_beta = "all", binary_regressor = FALSE,
+                         binary_beta = TRUE, only_intercept = FALSE,
+                         compare_pruning = TRUE,
+                         test = c("ctree", "mfluc",
+                                  "guide_sum_12", 
+                                  "guide_sum_1_cor"),
+                         beta0 = NULL, beta1 = NULL,
+                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05,
+                         nrsplitvar = 2)
+
+  save(simres, file = "pruning.rda")  
+}
+
+pr <- simres$res
+## extract data for xyplot comparing ARI and ARI_p
+pr <- pr[,c(1,2,7,8,9,12,15)]
+## extended data set with post-pruning results being represented as separated tests
+prex <- pr[,c(1,2,3,4,6)]
+prex <- rbind(prex,prex)
+
+prex$test <- factor(prex$test,
+                    levels = c("ctree","mfluc",
+                               "guide_sum_12", "guide_sum_1_cor"),
+                    labels = c("CTree","MOB",
+                               "GUIDE+scores", "GUIDE"))
+
+prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
+prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
+prex$pruning <- c(rep("pre", NROW(pr)), rep("post", NROW(pr)))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
+
+@
+
+\begin{figure}%[t]
+\setkeys{Gin}{width=\linewidth}
+<<tree_all, fig=TRUE, echo=FALSE, height=5, width=8>>=
+library("latticeExtra")
+useOuterStrips(xyplot(ari ~ delta | xi + pruning, groups = ~ test, 
+                      data = prex, 
+                      type = "l", lty = c(1,1,1,1),
+                      lwd = 2,
+                      key =  list(text = list(levels(prex$test)[1:4]),
+                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
+                      layout= c(4,2),
+                      scales=list(x=list(at=seq(1,11,2))),
+                      xlab = expression(delta),
+                      ylab = "Adjusted Rand Index",
+                      #index.cond = list(c(2,3,1),c(2,1)),
+                      par.settings = par.settings),
+               strip = strip.custom(factor.levels = c(expression(xi==0), 
+                                                      expression(xi==0.2), 
+                                                      expression(xi==0.5), 
+                                                      expression(xi==0.8)),
+                                    bg = "gray90"),
+               strip.left=strip.custom(factor.levels = c("post-pruning", "pre-pruning"),
+                                   bg = "gray90"))
+@
+\caption{\label{fig:tree_all}
+Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
+``tree'' scenario, once applying pre-pruning (top row) and once post-pruning (bottom row). 
+The ARI is evaluated over
+increasing effect size $\delta$ with 100 replications of 250 observations per step.
+Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
+$0.8$ from left to right).
+}
 \end{figure}
@@
 CTree algorithm applies a linear test statistic $T$ which is excerpted from Section~3.1.
-(``Variable selection and stopping criteria'') of the original paper and is of the following form:
+(``Variable selection and stopping criteria'') of the original paper \citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
 $$
@@
 
-% FIXME: adapt notation from here on (split variable Z and response Y, scores s)
 \subsection{MOB}
@@
 the model scores $s$ from zero. As described in detail in Section~3.2. (``Testing for 
-parameter instability'') of the original paper this process is of the following form:
+parameter instability'') of the original paper \citep{Zeileis+Hothorn+Hornik:2008}
+this process is of the following form:
 $$
@@
 To scale this partial sum process an estimate $\hat{J}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
-Following \cite{Zeileis+Hornik:2007} $W(t)$ converges to a Brownian bridge under the null
-hypothesis of parameter stability. To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied.
-For more details, particularly on calculating the corresponding $p$-values see Section~3.2. of the original paper.
+Following \cite{Zeileis+Hornik:2007} $W(t)$ converges to a Brownian bridge $W^0$ under the null
+hypothesis of parameter stability.
+To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied and the corresponding asymptotic distribution of $\lambda(W(t))$ can be obtained by employing the same functional to the limit process, i.e., $\lambda(W^0)$.
+For more details, particularly on calculating the corresponding $p$-values see Section~3.2. of the original paper \citep{Zeileis+Hothorn+Hornik:2008}.
 
@@
                                    bg = "gray90"),
-               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0~" (50%)"),                                                              expression(xi == 0.8~" (90%)")),
                                    bg = "gray90"))
@@
 true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row).
-Each column represents one scenario for the coefficients: 
+Each column represents one variation for the coefficients: 
 varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
@@
 with the varying coefficient(s) being continuous. 
-Each panel represents one scenario for the coefficients: 
+Each panel represents one variation for the coefficients: 
 varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
@@
 lead to a similar proportion of replications where the true splitting variable is detected
-(e.g., for CTree and $\xi=0.5$)
+(e.g., for CTree and the median $\xi=0.5$)
 

### 2019-05-22 pkg/partykit/inst/guideliketest/draft/draft.Rnw
results of pre- and post-pruning now illustrated in one combined plot; further text revisions, mainly in sections 4 and 5
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2585 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1c8a91c149b9ea0c7379246f2af8a36a08b12967

@@
 presented in Section~\ref{sec:lmtrees}. First, the underlying tree structure based on which the
-data is generated is a stump, i.e., a tree with only one split (see Figure~\ref{fig:dgp_stump}). 
+data is generated is a stump, i.e., a tree with only one split (``stump'' scenario, see 
+Figure~\ref{fig:dgp_stump}). 
 By keeping the tree so simple we can focus on the testing strategy only giving focus to their
@@
 In the second scenario the true tree structure contains two splits in two different variables
-yielding a tree with three terminal nodes (see Figure~\ref{fig:dgp_tree}). It employs the same
-basic structure as the first scenario but simply adds another split. This allows to evaluate
-the power of the three testing strategies in a more complex setting, in combination with using
-a post-pruning strategy.
+yielding a tree with three terminal nodes (``tree'' scenario, see Figure~\ref{fig:dgp_tree}). 
+It employs the same basic structure as the first scenario but simply adds another split. This 
+allows to evaluate the power of the three testing strategies in a more complex setting, in 
+combination with using a post-pruning strategy.
 
@@
 \caption{\label{tab:variables}Variables included in the data generating process 
-as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a split variable, 
+as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
 not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_i(Z_1,Z_2)$, $i=1,2$.}
@@
 Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
-with effect size $\delta = 1$ and for three cases: 
+with effect size $\delta = 1$ and for three variations: 
 varying $\beta_0$ and fixed $\beta_1$ (top right),
@@
 variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
-However, not only $Z_1$ but also $Z_2$ is used as a split variable. Therefore, an additional split
+However, not only $Z_1$ but also $Z_2$ is used as a true split variable. Therefore, an additional split
 is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
@@
 \endminipage
-\caption{\label{fig:dgp_tree}Left panel: True tree structure for the ``tree'' scenario 
+\caption{\label{fig:dgp_tree}Left panel: True tree structure
+applied in the data generating process for the ``tree'' scenario 
 with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
@@
 selected for splitting in a pre-pruning approach), denoted by the 
-``selection probability''.
-For the ``tree'' scenario the adjusted Rand index~(ARI) is calculated 
-as a measure of similarity between the true tree structure and the fitted model tree.
-
-As explained before, the aim of this simulation study is to investigate the effects
-of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
-rather than the whole testing strategies. Therefore, other combinations as presented in 
-Table~\ref{tab:buildingblocks}, hence adapted versions of GUIDE, CTREE, and MOB are evaluated as well
-as their original versions. Moreover, as pre-pruning is employed all of the considered statistical 
-tests are applied as significance tests answering two questions at once: 
+``selection probability''. In that way, the power of the
+considered statistical tests can be compared as they are all applied
+as significance tests answering two questions at once:
 (1)~Should a split be performed at all? 
@@
 than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.
+
+For the ``tree'' scenario the adjusted Rand index~(ARI) is calculated 
+as a measure of similarity between the true tree structure and the fitted model tree.
+
+As explained before, the aim of this simulation study is to investigate the effects
+of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
+rather than the whole testing strategies. Therefore, other combinations as presented in 
+Table~\ref{tab:buildingblocks}, hence adapted versions of GUIDE, CTREE, and MOB are evaluated as well
+as their original versions.
 
@@
 that residual-based tests can only capture parameter differences that
-affect the conditional mean \citep[see e.g.,][]{Ploberger+Kraemer:1992}.
-Hence we compare the CTree, MOB, and GUIDE algorithms 
--- all three using the default specification as shown in Table~\ref{tab:buildingblocks} -- 
-and additionally consider a new GUIDE flavor, denoted GUIDE+scores, that uses
+affect the conditional mean \citep[see e.g.,][ and further discussion in 
+Section~\ref{sec:discussion}]{Ploberger+Kraemer:1992}.
+Hence we compare the CTree, MOB, and GUIDE algorithms  -- all three using the 
+default specification as shown in Table~\ref{tab:buildingblocks} -- and 
+additionally consider a new GUIDE flavor, denoted GUIDE+scores, that uses
 binarized scores rather than residuals.
 
-%To assess the impact of using only residuals or the full score matrix four testing
-%strategies are compared: CTree, GUIDE, MOB -- all three using the default specification
-%as shown in Table~\ref{tab:buildingblocks} -- and additionally the GUIDE strategy using
-%the full scores instead (denoted by GUIDE+scores).
+In Figure~\ref{fig:scores_stump} the performances of the testing strategies are 
+represented by the corresponding selection probability, hence, a significance level 
+is incorporated. To show that in general the drawn conclusions do not change if no 
+significance level is included, a different illustration of the exact same evaluation 
+is provided in Appendix~\ref{app:significance} where the performances are compared based 
+on the smallest $p$-values, regardless of whether they are beneath any significance level.
+
+Moreover, as defined by the true tree structure used in the data generating process, the
+parameter functions $\beta_i(Z_1)$, $i=1,2$, are step functions which is particularly 
+advantageous for the maximum selection in the test statistic applied by MOB while the linear 
+selection in CTree is preferable for detecting monotonous dependencies. To show that employing 
+a continuous parameter function instead of a step function only has an impact on whether CTree or
+MOB performs better the exact same simulation setting as used for Figure~\ref{fig:scores_stump} 
+but with continuously changing parameters has been evaluated and is illustrated in 
+Appendix~\ref{app:contbeta}.
 
@@
                       xlab = expression(delta),
-                      ylab = expression("Proportion of selecting"~Z[1]),
+                      ylab = expression("Selection probability of"~Z[1]),
                       par.settings = par.settings),
@@
 of replications. Even though the advantage of CTree and MOB evens out in certain situations
-such as for a high effect size (see, e.g., for the median $\xi=0$ and varying $\beta_0$), GUIDE is 
-never ahead of them in any of the considered settings. Moreover, the positive effect of 
-including all available scores can be observed particularly well in the top middle panel 
-where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
-GUIDE. 
+such as for a high effect size (see, e.g., for the median $0$ and varying $\beta_0$), 
+GUIDE is never ahead of them in any of the considered settings. Moreover, the positive 
+effect of including all available scores can be observed particularly well in the top middle 
+panel where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to GUIDE. 
 %The fact that residual-based methods fail to detect changes that do not effect the
@@
 %Section~\ref{sec:discussion}. 
-Furthermore, the advantage of including full models scores 
+Furthermore, this advantage of including full models scores 
 clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
@@
 For a central split point ($\xi=0$ corresponding to the median, upper panels) all strategies 
-perform similarly well as long as the intercept varies (left and right panel) with CTree and MOB 
-being only slightly ahead. However, for the split point being close to the upper 
-edge of the considered interval ($\xi=0.8$ corresponding to the $90$\%-quantile, lower panels) 
-the performance of all of the applied strategies decreases. 
-While both GUIDE versions struggle to detect the correct split variable even for a 
-high effect size, MOB is clearly ahead with CTree leading to the second best results.
-(In order to show that the here observed advantage of MOB over CTree is mainly due to the 
-abrupt shift in the model parameters $\beta_0$ and $\beta_1$ an additional evaluation is 
-presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)
+perform similarly well as long as the intercept varies (left and right panel) with CTree and 
+MOB being only slightly ahead. However, for the split point at the $90$\% quantile $0.8$ 
+(lower panels) the performance of all of the applied strategies decreases. 
+While both GUIDE versions struggle to detect the correct split variable even for a high effect 
+size, MOB is clearly ahead with CTree leading to the second best results.
+As discussed before, this advantage of MOB over CTree is mainly due to the abrupt shift in the 
+model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
+continuously changing parameters (see Appendix~\ref{app:contbeta}).
 From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
 can be deduced that categorizing split variables and binarizing scores has a negative effect 
-on the ability to detect the right split variable.
-To elaborate these effects more in detail the results for all possible combinations of the building blocks are discussed in a full factorial 3-way analysis in the following section.
-
-\textit{Note:} As explained before, the tests are applied as significance tests with 
-significance level $\alpha=0.05$. To show that in general the drawn conclusions do not 
-change if no significance level is included the same setting is evaluated again without 
-any significance level and the results are presented in Appendix~\ref{app:significance}.
+on the ability to detect the right split variable. To elaborate these effects more in detail 
+the results for all possible combinations of the building blocks are discussed in a full 
+factorial 3-way analysis in the following section.
 
@@
 possible combinations of the building blocks have been included. The different levels of 
-each of the three building blocks are listed in Table~\ref{tab:levels}. A more detailed
-overview of all 12 combinations can be found in Appendix~\ref{app:combinations}, 
-Table~\ref{tab:combinations_appendix}.
+each of the three building blocks are listed in Table~\ref{tab:levels} where 
+$h$ and $g$ refer to the transformation functions applied to the response $Y$ or
+a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
+(See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
+for a more detailed overview of all 12 combinations.)
+If quartiles are used as breakpoints to categorize a split variable the employed
+function $g$ maps to a four-dimensional vector of zeros except for one value $1$
+indication the bin to which an observation is assigned to. For a maximum selection
+of a split variable $g$ maps to a vector containing only zeros until a certain
+changepoint from which on the values are replace by $1$. This breakpoint indicates
+for example a crossed quantile. In linear selection $g$ is simply the identity function.
 
@@
 
-Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing residuals/scores decreases the performance 
-as it leads to higher $p$-values corresponding to the true split variable $Z_1$ 
-and thus to lower power in the presented settings. 
-For a central true split point (left panel) this effect remains at an almost constant level 
-regardless of whether possible split variables are categorized or not. However, in case 
-of the true split point being closer to the upper margin ($90$\%-quantile $\xi=0.8$, right panel),
-categorizing split variables increases the $p$-value of $Z_1$ even more. 
-Here, the maximal selection performs best while the linear selection shows its disadvantage in 
-detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximal
-selection of a split variable is advantageous. But unless one of the breakpoints used for binning possible split variables is close to the true split point (as for the median $\xi=0$, left panel),
+Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing 
+residuals/scores decreases the performance as it leads to higher $p$-values corresponding 
+to the true split variable $Z_1$ and thus to lower power in the presented settings. 
+For a central true split point ($\xi=0$, left panel) this effect remains at an almost constant 
+level regardless of whether possible split variables are categorized or not. However, in case 
+of the true split point at the $90$\% quantile $0.8$ (right panel), categorizing split 
+variables increases the $p$-value of $Z_1$ even more. 
+Here, the maximum selection performs best while the linear selection shows its disadvantage in 
+detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximum
+selection of a split variable is advantageous. But unless one of the breakpoints used for binning 
+possible split variables is close to the true split point (as for the median $0$, left panel),
 this categorization weakens the performance of a testing strategy. 
-
 While the effect of both of these transformations of the data (binarization of 
@@
 remarkably.
+
 As already shown in Section~\ref{sec:scores} the use of scores vs.\ residuals has a minor 
@@
 Note that the choice of the two specific settings 
-(left panel: true split point at the median $\xi=0$,  effect size $\delta = 0.3$; 
-right panel: true split point at the $90$\%-quantile $\xi=0.8$ requiring a higher effect 
-size $\delta = 1$)
+(left panel: true split point at the median $0$,  effect size $0.3$; 
+right panel: true split point at the $90$\% quantile $0.8$ requiring a higher effect 
+size $1$)
 has been made based on the results shown in Figure~\ref{fig:scores_stump} in order to make 
@@
 mean $p$-values corresponding to the true split variable $Z_1$. 
-Left panel: true split point at the median $\xi=0$ and 
-effect size $\delta = 0.3$.
-Right panel: true split point at the $90$\%-quantile $\xi=0.8$ and effect 
-size $\delta = 1$.}
+Left panel: true split point at the median $0$ and with
+effect size $0.3$.
+Right panel: true split point at the $90$\% quantile $0.8$ and with effect 
+size $1$.}
 \end{figure}
@@
 Figure~\ref{fig:dgp_tree}.
-
-Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
-almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE+scores performs
-slightly worse but it still achieves an ARI above $0.5$ for the highest considered effect size
-of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
-due to the fact that it doesn't detect the first split in $Z_2$ corresponding to a changing
-slope parameter as it only considers residuals in its testing strategy.
-For a split point farther away from the center such as for the $90$\%-quantile $\xi = 0.8$ MOB 
-outperforms all other
-methods and both GUIDE versions fail to find the correct tree structure.
-Based on these results it can be concluded once again that including all available scores in the
-testing strategy leads to a better performance while binarizing scores and categorizing
-split variables both do not cause any improvement but in many situations even have a negative
-effect on the power of a testing strategy.
-
-\begin{figure}%[t]
-\setkeys{Gin}{width=\linewidth}
-<<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
-                                        "guide_sum_12", "guide_sum_1_cor"))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "mfluc",
-                                  "guide_sum_12", "guide_sum_1_cor"),
-                       labels = c("CTree","MOB",
-                                  "GUIDE+scores", "GUIDE"))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
-
-xyplot(ari ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       type = "l",
-       lty = c(1,1,1,1),
-       lwd = 2,
-       key =  list(text = list(levels(subdata$test)),
-                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
-       layout= c(4,1),
-       #index.cond = list(c(1,2,3,4)),
-       scales=list(x=list(at=seq(1,11,2))),
-       xlab = expression(delta),
-       ylab = "Adjusted Rand Index",
-       strip = strip.custom(factor.levels = c(expression(xi==0), 
-                                              expression(xi==0.2), 
-                                              expression(xi==0.5), 
-                                              expression(xi==0.8)),
-                                   bg = "gray90"),
-       par.settings = par.settings)
-
-@
-\caption{\label{fig:tree_ari}
-Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
-``tree'' scenario, all applying pre-pruning. The ARI is evaluated over
-increasing effect size $\delta$ with 100 replications of 250 observations per step.
-Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
-$0.8$ from left to right).}
-\end{figure}
-
-To elaborate the impact of the choice of a pruning strategy
-the tree models presented in Figure~\ref{fig:tree_ari} are built again, but this time
-applying post-pruning (building very large trees without any significance level and employing 
-cost-complexity pruning as in the CART algorithm \citep{Breiman+Friedman+Stone:1984}). 
-In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with pre-pruning,
-again based on the adjusted Rand index.
-
-For MOB and CTree it has been shown that the tests work very well as significance tests.
-Therefore, it can be recommended to take advantage of this power and use the information
-given by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:pruning_ari}
-in most cases the results of these two testing strategies do not vary significantly over 
-changing the applied pruning strategy.
-On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
-strategy should not be based on their results which argues for the application of post-pruning.
-This can also be observed in Figure~\ref{fig:pruning_ari} showing that post-pruning improves
-the results for both GUIDE versions.
-
-<<pruning_data, echo=FALSE, results=hide>>=
-if(file.exists("pruning.rda")){
-  load("pruning.rda")
-} else {
-  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7, 
-                         nrsteps = 1, nrlevels = 2,
-                         delta = seq(from = 0, to = 1, by = 0.1),
-                         xi = c(0, 0.2, 0.5, 0.8), 
-                         vary_beta = "all", binary_regressor = FALSE,
-                         binary_beta = TRUE, only_intercept = FALSE,
-                         compare_pruning = TRUE,
-                         test = c("ctree", "mfluc",
-                                  "guide_sum_12", 
-                                  "guide_sum_1_cor"),
-                         beta0 = NULL, beta1 = NULL,
-                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05,
-                         nrsplitvar = 2)
-
-  save(simres, file = "pruning.rda")  
-}
-
-pr <- simres$res
-## extract data for xyplot comparing ARI and ARI_p
-pr <- pr[,c(1,2,7,8,9,12,15)]
-## extended data set with post-pruning results being represented as separated tests
-prex <- pr[,c(1,2,3,4,6)]
-prex <- rbind(prex,prex)
-
-prex$test <- factor(prex$test,
-                    levels = c("ctree","mfluc",
-                               "guide_sum_12", "guide_sum_1_cor",
-                               "ctree_p","mfluc_p",
-                               "guide_sum_12_p", "guide_sum_1_cor_p"),
-                    labels = c("CTree","MOB",
-                               "GUIDE+scores", "GUIDE",
-                               "CTree_p","MOB_p",
-                               "GUIDE+scores_p", "GUIDE_p"))
-
-for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
-  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
-  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
-  if(prex$test[i] == "GUIDE+scores") prex$test[i] <- "GUIDE+scores_p"
-  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
-}
-
-prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
-prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
-
-@
-
-\begin{figure}%[t]
-\setkeys{Gin}{width=\linewidth}
-<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
-xyplot(ari ~ delta | xi, groups = ~ test, 
-       data = prex, 
-       type = "l", lty = c(2,2,2,2,1,1,1,1),
-       lwd = 2,
-       key =  list(text = list(levels(prex$test)[1:4]),
-                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
-       layout= c(4,1),
-       scales=list(x=list(at=seq(1,11,2))),
-       xlab = expression(delta),
-       ylab = "Adjusted Rand Index",
-       strip = strip.custom(factor.levels = c(expression(xi==0), 
-                                              expression(xi==0.2), 
-                                              expression(xi==0.5), 
-                                              expression(xi==0.8)),
-                                   bg = "gray90"),
-       par.settings = par.settings
-)
-@
-\caption{\label{fig:pruning_ari}
-Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
-``tree'' scenario, once applying pre-pruning (dashed lines) and once post-pruning (solid lines). 
-The ARI is evaluated over
-increasing effect size $\delta$ with 100 replications of 250 observations per step.
-Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
-$0.8$ from left to right).}
-\end{figure}
+The results presented in the top row of Figure~\ref{fig:tree_all} have been obtained applying pre-pruning.
+Therefore, based on the corresponding ARI values conclusion can again be drawn on the power of the testing strategies 
+as significance tests. CTree and MOB perform almost equally well for the true split points $0$, $0.2$ and $0.5$ 
+($50$\%, $60$\%, and $75$\% quantile). GUIDE+scores performs slightly worse but it still achieves an ARI above 
+$0.5$ for the highest considered effect size of $\delta = 1$. 
+However, the original version of GUIDE is clearly behind which is mainly due to the fact that it doesn't detect 
+the first split in $Z_2$ corresponding to a changing slope parameter as it only considers residuals in its testing 
+strategy. 
+For a split point farther away from the center such as the $90$\% quantile $0.8$ MOB 
+outperforms all other methods and both GUIDE versions fail to find the correct tree structure.
+Hence, the insights provided by the top row of Figure~\ref{fig:tree_all} conform with the conclusions
+drawn from the ``stump'' scenario: Including all available scores in the testing strategy leads to a better performance 
+while binarizing scores and categorizing split variables both do not cause any improvement but in many situations even 
+have a negative effect on the power of a testing strategy.
 
@@
                                     bg = "gray90"),
-               strip.left=strip.custom(factor.levels = c("post-pruning", "pre-pruning"),
+               strip.left=strip.custom(factor.levels = c("Post-pruning", "Pre-pruning"),
                                    bg = "gray90"))
@@
 Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
-$0.8$ from left to right).
+$0.8$ from left to right, corresponding to the $50$\%, $60$\%, $75$\%, and $90$\% quantile).
 }
 \end{figure}
+
+To elaborate the impact of the choice of a pruning strategy the presented tree models
+are built again in the exact same simulation setting but employing post-pruning (i.e., building very 
+large trees without any significance level and applying cost-complexity pruning as in the CART 
+algorithm \citep{Breiman+Friedman+Stone:1984}). 
+The corresponding ARI values are illustrated in the bottom row of Figure~\ref{fig:tree_all}
+allowing for a comparison to those obtained with pre-pruning (top row).
+
+For MOB and CTree it has been shown that the tests work very well as significance tests.
+Therefore, it can be recommended to take advantage of this power and use the information
+given by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:tree_all}
+in most cases the results of these two testing strategies do not vary significantly over 
+a change of the applied pruning strategy.
+On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
+strategy should not be based on their results which argues for the application of post-pruning.
+This is also supported by the results in Figure~\ref{fig:tree_all} showing that post-pruning improves
+the performance of both GUIDE versions. Particularly if the true split point is close to a binning 
+point (such that categorizing split variables has no negative effect) the post-pruning strategy 
+compensates the lack of power of GUIDE+scores compared to CTree and MOB almost completely (bottom 
+row, first and third panel).
+In conclusion, the choice of a pruning strategy should be based on knowledge about the 
+power of the applied testing strategy. With a well working significance test additional expensive
+calculations such as needed for post-pruning may be avoided being one of the main advantages of 
+pre-pruning. However, deficiencies of a testing strategy in terms of power can be at least partly 
+evened out by employing a post-pruning strategy.
 
@@
 @
-\caption{\label{fig:scores_stump_wosign}Proportion of replications where the true split variable $Z_1$ 
-shows the smallest $p$-value, hence, where $Z_1$ is selected without any significance level, 
-over increasing effect size $\delta$ (100 replications of 250 observations per step), once for a 
-true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row).
-Each column represents one variation for the coefficients: 
-varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
-both coefficients varying (right). The performances of four testing strategies are compared: 
-CTree, MOB, GUIDE, GUIDE+scores.}
+\caption{\label{fig:scores_stump_wosign}
+Proportion of replications where the true split variable $Z_1$ shows the smallest $p$-value, 
+hence, where $Z_1$ is selected without any significance level in the ``stump'' scenario.
+The testing strategies CTree, CTree+max, MOB, GUIDE, and GUIDE+scores are evaluated 
+over increasing effect size $\delta$ with 100 replications of 250 observations per step. 
+The true split point in $Z_1$ is either the median 0 (top) or the 90\% quantile 0.8 (bottom)
+for either varying intercept $\beta_0$ (left), varying slope $\beta_1$
+(middle), or both coefficients varying (right).}
 \end{figure}
@@
        xlab = expression(delta),
-       ylab = expression("Proportion of selecting"~Z[1]),
+       ylab = expression("Selection probability of"~Z[1]),
        index.cond = list(c(2,3,1)),
@@
 \caption{\label{fig:stump_contbeta}
-Proportion of replications where the true split variable $Z_1$ is
-selected over increasing effect size $\delta$ (100 replications of 250 observations per step)
-with the varying coefficient(s) being continuous. 
+Selection probability of the true split variable $Z_1$ for testing strategies CTree, CTree+max,
+MOB, GUIDE, and GUIDE+scores for continuous parameter functions $\beta_0(Z_1)$ and $\beta_1(Z_1)$.
+Probabilities are estimated over increasing effect size $\delta$
+with 100 replications of 250 observations per step. 
 Each panel represents one variation for the coefficients: 
 varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
-both coefficients varying (right). The performances of five testing strategies are compared: 
-CTree, CTree+max, MOB, GUIDE, GUIDE+scores.}
+both coefficients varying (right).}
 \end{figure}
@@
        xlab = expression(delta),
-       ylab = expression("Proportion of selecting"~Z[1]),
+       ylab = expression("Selection probability of"~Z[1]),
        strip = strip.custom(factor.levels = c(expression(xi==0), 
@@
 @
-\caption{\label{fig:bin_stump}Proportion of replications where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ (100 replications of 250 observations per step)
-with both parameter varying. Each panel represents a different value of the true 
-split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
-strategies are compared: CTree, CTree+bin, MOB, MOB+bin, GUIDE+scores.}
+\caption{\label{fig:bin_stump}
+Selection probability of the true split variable $Z_1$ for
+testing strategies CTree, CTree+bin, MOB, MOB+bin, and GUIDE+scores in the ``stump''
+scenario with both parameter varying. 
+Probabilities are estimated over increasing effect size $\delta$
+with 100 replications of 250 observations per step. 
+Each panel represents a different value of the true split point $\xi$ (0, 0.2, 0.5 
+and 0.8 from left to right corresponding to the 50\%, 60\%, 75\%, and 90\% quantile).
+}
 \end{figure}
@@
 lead to a similar proportion of replications where the true splitting variable is detected
-(e.g., for CTree and the median $\xi=0.5$)
+(e.g., for CTree and $\xi=0.5$)
 
@@
 @
-\caption{\label{fig:cat_stump}Proportion of replications where the true split variable $Z_1$ is
-detected over increasing effect size $\delta$ (100 replications of 250 observations per step)
-with both parameter varying. Each panel represents a different value of the true 
-split point $\xi$ (0, 0.2, 0.5 and 0.8 from left to right). The performances of five testing 
-strategies are compared: CTree, CTree+cat, MOB, MOB+cat, GUIDE+scores.}
+\caption{\label{fig:cat_stump}
+Selection probability of the true split variable $Z_1$ for
+testing strategies CTree, CTree+cat, MOB, MOB+cat, and GUIDE+scores in the ``stump''
+scenario with both parameter varying. 
+Probabilities are estimated over increasing effect size $\delta$
+with 100 replications of 250 observations per step. 
+Each panel represents a different value of the true split point $\xi$ (0, 0.2, 0.5 
+and 0.8 from left to right corresponding to the 50\%, 60\%, 75\%, and 90\% quantile).
+}
 \end{figure}
@@
 \end{document}
+
+%\begin{figure}%[t]
+%\setkeys{Gin}{width=\linewidth}
+%<<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
+%subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
+%                                        "guide_sum_12", "guide_sum_1_cor"))
+%subdata$test <- factor(subdata$test, 
+%                       levels = c("ctree", "mfluc",
+%                                  "guide_sum_12", "guide_sum_1_cor"),
+%                       labels = c("CTree","MOB",
+%                                  "GUIDE+scores", "GUIDE"))
+%
+%par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+%                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+%                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
+%
+%
+%xyplot(ari ~ delta | xi, groups = ~ test, 
+%       data = subdata, 
+%       type = "l",
+%       lty = c(1,1,1,1),
+%       lwd = 2,
+%      key =  list(text = list(levels(subdata$test)),
+%                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+%                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
+%       layout= c(4,1),
+%       #index.cond = list(c(1,2,3,4)),
+%       scales=list(x=list(at=seq(1,11,2))),
+%       xlab = expression(delta),
+%       ylab = "Adjusted Rand Index",
+%       strip = strip.custom(factor.levels = c(expression(xi==0), 
+%                                              expression(xi==0.2), 
+%                                              expression(xi==0.5), 
+%                                              expression(xi==0.8)),
+%                                   bg = "gray90"),
+%       par.settings = par.settings)
+%
+%@
+%\caption{\label{fig:tree_ari}
+%Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
+%``tree'' scenario, all applying pre-pruning. The ARI is evaluated over
+%increasing effect size $\delta$ with 100 replications of 250 observations per step.
+%Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
+%$0.8$ from left to right).}
+%\end{figure}
+
+%<<pruning_data, echo=FALSE, results=hide>>=
+%if(file.exists("pruning.rda")){
+%  load("pruning.rda")
+%} else {
+%  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7, 
+%                         nrsteps = 1, nrlevels = 2,
+%                         delta = seq(from = 0, to = 1, by = 0.1),
+%                         xi = c(0, 0.2, 0.5, 0.8), 
+%                         vary_beta = "all", binary_regressor = FALSE,
+%                         binary_beta = TRUE, only_intercept = FALSE,
+%                         compare_pruning = TRUE,
+%                         test = c("ctree", "mfluc",
+%                                  "guide_sum_12", 
+%                                  "guide_sum_1_cor"),
+%                         beta0 = NULL, beta1 = NULL,
+%                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05,
+%                         nrsplitvar = 2)
+%  
+%  save(simres, file = "pruning.rda")  
+%}
+%
+%pr <- simres$res
+%## extract data for xyplot comparing ARI and ARI_p
+%pr <- pr[,c(1,2,7,8,9,12,15)]
+%## extended data set with post-pruning results being represented as separated tests
+%prex <- pr[,c(1,2,3,4,6)]
+%prex <- rbind(prex,prex)
+%
+%prex$test <- factor(prex$test,
+%                    levels = c("ctree","mfluc",
+%                               "guide_sum_12", "guide_sum_1_cor",
+%                               "ctree_p","mfluc_p",
+%                               "guide_sum_12_p", "guide_sum_1_cor_p"),
+%                    labels = c("CTree","MOB",
+%                               "GUIDE+scores", "GUIDE",
+%                               "CTree_p","MOB_p",
+%                               "GUIDE+scores_p", "GUIDE_p"))
+%
+%for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
+%  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
+%  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
+%  if(prex$test[i] == "GUIDE+scores") prex$test[i] <- "GUIDE+scores_p"
+%  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
+%}
+%
+%prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
+%prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
+%
+%par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+%                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+%                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
+%
+%@
+
+%\begin{figure}%[t]
+%\setkeys{Gin}{width=\linewidth}
+%<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
+%xyplot(ari ~ delta | xi, groups = ~ test, 
+%       data = prex, 
+%       type = "l", lty = c(2,2,2,2,1,1,1,1),
+%       lwd = 2,
+%       key =  list(text = list(levels(prex$test)[1:4]),
+%                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+%                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
+%       layout= c(4,1),
+%       scales=list(x=list(at=seq(1,11,2))),
+%       xlab = expression(delta),
+%       ylab = "Adjusted Rand Index",
+%       strip = strip.custom(factor.levels = c(expression(xi==0), 
+%                                              expression(xi==0.2), 
+%                                              expression(xi==0.5), 
+%                                              expression(xi==0.8)),
+%                                   bg = "gray90"),
+%       par.settings = par.settings
+%)
+%@
+%\caption{\label{fig:pruning_ari}
+%Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
+%``tree'' scenario, once applying pre-pruning (dashed lines) and once post-pruning (solid lines). 
+%The ARI is evaluated over
+%increasing effect size $\delta$ with 100 replications of 250 observations per step.
+%Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
+%$0.8$ from left to right).}
+%\end{figure}

### 2019-05-23 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions, mainly on section 5 and the appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2586 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b7890df49716d07a5e3ffb8c27c221676de1855a

@@
 as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
-not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_i(Z_1,Z_2)$, $i=1,2$.}
+not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_j(Z_1,Z_2)$, $j=0,1$.}
 \end{table}
@@
 \begin{align*}
-\beta_{i}(Z_1) = \begin{cases}
+\beta_{j}(Z_1) = \begin{cases}
 -\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \leq \xi \\
@@
 variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
-However, not only $Z_1$ but also $Z_2$ is used as a true split variable. Therefore, an additional split
+However, not only $Z_1$ but also $Z_2$ is used as a true split variable following a uniform distribution
+on $[-1,1]$. Therefore, an additional split
 is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
@@
 Moreover, as defined by the true tree structure used in the data generating process, the
-parameter functions $\beta_i(Z_1)$, $i=1,2$, are step functions which is particularly 
+parameter functions $\beta_j(Z_1)$, $j=0,1$, are step functions which is particularly 
 advantageous for the maximum selection in the test statistic applied by MOB while the linear 
@@
 From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
-can be deduced that categorizing split variables and binarizing scores has a negative effect 
-on the ability to detect the right split variable. To elaborate these effects more in detail 
+can be deduced that categorizing split variables and binarizing scores also decreases the
+ability to detect the right split variable. To elaborate these effects more in detail 
 the results for all possible combinations of the building blocks are discussed in a full 
@@
 \subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
-To investigate the effects of each of the building blocks separately the most general 
-case of the ``stump'' scenario where the intercept and 
-slope parameter are both varying is considered. In the underlying evaluation all 
-possible combinations of the building blocks have been included. The different levels of 
-each of the three building blocks are listed in Table~\ref{tab:levels} where 
-$h$ and $g$ refer to the transformation functions applied to the response $Y$ or
-a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
+To investigate the impact of each of the building blocks separately the most general 
+case of the ``stump'' scenario where the intercept and slope parameter are both varying is 
+considered. In the underlying evaluation all possible combinations of the building blocks 
+have been included. The different levels of each of the three building blocks are listed in 
+Table~\ref{tab:levels} where $h$ and $g$ refer to the transformation functions applied to the 
+response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
+If quartiles are used as breakpoints to categorize a split variable the employed
+function $g$ maps to a four-dimensional vector of zeros except for one value $1$
+indication the bin to which an observation is assigned. For a maximum selection
+of a split variable $g$ maps to a vector containing only zeros until a certain
+changepoint from which on the values are replaced by $1$. This breakpoint indicates
+for example a crossed quantile. In a linear selection strategy $g$ is simply the identity function.
 (See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
 for a more detailed overview of all 12 combinations.)
-If quartiles are used as breakpoints to categorize a split variable the employed
-function $g$ maps to a four-dimensional vector of zeros except for one value $1$
-indication the bin to which an observation is assigned to. For a maximum selection
-of a split variable $g$ maps to a vector containing only zeros until a certain
-changepoint from which on the values are replace by $1$. This breakpoint indicates
-for example a crossed quantile. In linear selection $g$ is simply the identity function.
 
@@
 Here, the maximum selection performs best while the linear selection shows its disadvantage in 
-detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximum
-selection of a split variable is advantageous. But unless one of the breakpoints used for binning 
+detecting abrupt changes, but overall it depends on the situation whether a linear or a maximum
+selection of a split variable is advantageous. 
+Comparing the two panels suggests that unless one of the breakpoints used for binning 
 possible split variables is close to the true split point (as for the median $0$, left panel), 
-this categorization weakens the performance of a testing strategy. 
-While the effect of both of these transformations of the data (binarization of 
+a categorization weakens the performance of a testing strategy. 
+Moreover, while the effect of both of these transformations of the data (binarization of 
 residuals/scores and categorization of split variables) can be observed
@@
 
-For MOB and CTree it has been shown that the tests work very well as significance tests.
+For MOB and CTree it has been shown that their tests work very well as significance tests.
 Therefore, it can be recommended to take advantage of this power and use the information
-given by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:tree_all}
+provided by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:tree_all}
 in most cases the results of these two testing strategies do not vary significantly over 
@@
 situations but remarkable deficits in general. On the contrary, the benefits of including full model 
-scores clearly outweigh the price of of testing all parameters.
+scores clearly outweigh the price of testing all parameters.
 
@@
 To evaluate the GUIDE algorithm a reimplementation of this algorithm has been 
-established/constructed using the basic framework of \code{ctree} and \code{mob}.
+built using the basic framework of \code{ctree} and \code{mob}.
 
@@
 This section provides further information on the test statistics applied in CTree, MOB, 
-and GUIDE. The presented details have been extracted from the original papers 
+and GUIDE. Most of the presented details have been extracted from the original papers 
 \cite{Hothorn+Hornik+Zeileis:2006}, \cite{Zeileis+Hothorn+Hornik:2008}, 
@@
 $$
-where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom transformation of the split variable $Z$ and the influence function $h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the responses $(Y_1, \ldots, Y_n)$ in a permutation symmetric way.
+where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom transformation of the split variable $Z$ and the influence function $h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the response $Y = (Y_1, \ldots, Y_n)$ in a permutation symmetric way.
 Moreover, by applying the ``vec'' operator the resulting $p~\times~q$ matrix is converted into a $p q$ column vector.
@@
 test as represented by the different building blocks. In that way,
-``+res'' refers to using only residuals (instead of all scores),
-``+scores'' to using all scores (instead of only residuals),
+``+res'' refers to using only residuals (instead of all model scores),
+``+scores'' to using all model scores (instead of only residuals),
 ``+cat'' to applying a categorization of split variables,
@@
 \hline
-GUIDE             & residuals          & bin          & cat \\
-MOB+bin+res     & residuals          & bin          & max \\
-CTree+bin+res   & residuals          & bin          & lin \\
-MOB+cat+res    & residuals          & lin          & cat \\
-MOB+res          & residuals          & lin          & max \\
-CTree+res        & residuals          & lin          & lin \\
-GUIDE+scores     & scores             & bin          & cat \\
-MOB+bin          & scores             & bin          & max \\
-CTree+bin        & scores             & bin          & lin \\
-MOB+cat          & scores             & lin          & cat \\
-MOB               & scores             & lin          & max \\
-CTree             & scores             & lin          & lin \\
+GUIDE             & residuals          & yes          & cat \\
+MOB+bin+res       & residuals          & yes          & max \\
+CTree+bin+res     & residuals          & yes          & lin \\
+MOB+cat+res       & residuals          & no           & cat \\
+MOB+res           & residuals          & no           & max \\
+CTree+res         & residuals          & no           & lin \\
+GUIDE+scores      & scores             & yes          & cat \\
+MOB+bin           & scores             & yes          & max \\
+CTree+bin         & scores             & yes          & lin \\
+MOB+cat           & scores             & no           & cat \\
+MOB               & scores             & no           & max \\
+CTree             & scores             & no           & lin \\
 \hline
@@
 <<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
-subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
+subdata <- subset(sim_stump, test %in% c("ctree", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
 subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "ctree_max", "mfluc",
+                       levels = c("ctree", "mfluc",
                                   "guide_sum_12", "guide_sum_1_cor"),
-                       labels = c("CTree", "CTree+max", "MOB",
+                       labels = c("CTree", "MOB",
                                   "GUIDE+scores", "GUIDE"))
@@
 
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
 
@@
                         only_intercept == FALSE,# & xi == 0, 
-                      type = "l", lty = c(1,1,1,1,1),
+                      type = "l", lty = c(1,1,1,1),
                       lwd = 2,
@@
                       key =  list(text = list(levels(subdataxi$test)),
-                                  lines = list(lty = c(1,1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                               col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
+                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                       layout= c(3,2),
@@
 hence, where $Z_1$ is selected without any significance level in the ``stump'' scenario.
-The testing strategies CTree, CTree+max, MOB, GUIDE, and GUIDE+scores are evaluated 
+The testing strategies CTree, MOB, GUIDE, and GUIDE+scores are evaluated 
 over increasing effect size $\delta$ with 100 replications of 250 observations per step. 
@@
 In Section~\ref{sec:scores} the investigated testing strategies are compared based
-on the number of replications in which the true split variable $Z_1$ is detected with a $p$-value
-smaller than the predefined significance level $\alpha = 0.05$. This choice of measurement
-is due to the aim of investigating how well the tests perform as significance test.
-However, ignoring this significance level yields the same conclusions 
-which can be seen when comparing Figures~\ref{fig:scores_stump_wosign}
-and \ref{fig:scores_stump}.
-In Figure~\ref{fig:scores_stump_wosign} the proportion of replications 
-in which the true splitting variable $Z_1$ is detected, but not necessarily with a $p$-value 
-smaller than the significance level, is illustrated. In Figure~\ref{fig:scores_stump} 
-the exact same simulation setting is applied with a significance level.
-
-\section{Simulation results for continuous model parameters}
+on the selection probability of $Z_1$, hence, the number of replications in which the true split 
+variable $Z_1$ is detected with a $p$-value smaller than the predefined significance level 
+$\alpha = 0.05$. This choice of measurement is due to the aim of investigating how well the 
+tests perform as significance test.
+However, ignoring this significance level yields the same conclusions which can be seen when 
+comparing Figures~\ref{fig:scores_stump_wosign} and \ref{fig:scores_stump}, both being based 
+on the exact same evaluation, but in Figure~\ref{fig:scores_stump_wosign} the proportion of 
+replications in which the true splitting variable $Z_1$ is detected by showing the smallest $p$-value, 
+but not necessarily smaller than the significance level, is illustrated.
+
+\section{Continuous ``stump'' scenario}
 \label{app:contbeta}
@@
 either the positive or the negative value of the effect size $\delta$. Therefore, these parameters
-are step functions for which MOB has shown to perform better than CTree. However, in case of
-monotonous functions CTree is advantageous as it is constructed to detect monotonous effects.
+are step functions for which the maximum selection employed in the testing strategy of MOB has shown 
+to perform better than the linear selection in CTree. However, this changes in case of
+monotonous functions where CTree is advantageous as it is constructed to detect monotonous effects.
 To point out that the difference in performance of CTree and MOB depends on the type of effect
-the same setting of which the results are presented in Figure~\ref{fig:scores_stump} is
+the same setting for which the results are presented in Figure~\ref{fig:scores_stump} is
 evaluated again but this time with the varying parameter(s) changing continuously.
@@
 \begin{align*}
-\beta_{i}(Z_1) = \begin{cases}
--\delta \cdot (-1)^{i} \cdot Z_1\\
-+\delta \cdot (-1)^{i} \cdot Z_1.
+\beta_{j}(Z_1) = \begin{cases}
+-\delta \cdot (-1)^{j} \cdot Z_1\\
++\delta \cdot (-1)^{j} \cdot Z_1
 \end{cases}
 \end{align*}
-Moreover, a modified version of CTree considering maximum values in the test statistic is evaluated
-(denoted by CTree+max). 
-Looking at the results presented in Figure~\ref{fig:stump_contbeta} it can be observed that in this setting the 
-original version of CTree is ahead while CTree+max and MOB perform almost equally well. Hence, the comparison of 
-Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} points out that it clearly depends on the type of 
-effect whether the maximal selection (MOB, CTree+max) or the linear selection (CTree) is advantageous.
+for $j = 0,1$.
+Additionally, a modified version of CTree employing a maximum selection, such as MOB, is evaluated 
+(denoted by CTree+max). Looking at the selection probability illustrated in Figure~\ref{fig:stump_contbeta} 
+it can be observed that in this setting the original version of CTree is ahead while CTree+max and MOB perform 
+almost equally well. Hence, the comparison of Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} 
+points out that it clearly depends on the type of effect whether the maximum selection (MOB, CTree+max) 
+or the linear selection (CTree) is to be preferred.
 
@@
 
-\section{Results for increasing effect size}
+\section{Results over increasing effect size}
 \label{app:increasingdelta}
@@
 \label{app:binarization}
-To elaborate over an increasing effect size whether a binarization of the residuals/scores at $0$ 
-leads to an improvement or a deterioration of performance CTree and MOB are applied, once in their 
-original version without a binarization and once in an adapted version with a binarization of 
-scores at $0$ (CTree+bin, MOB+bin). Moreover, they are compared to the adapted GUIDE version 
+To elaborate over an increasing effect size whether a binarization of the residuals/scores at zero 
+leads to an improvement or a deterioration of performance, CTree and MOB are applied, once in their 
+original version without binarization and once in an adapted version with a binarization of 
+scores at zero (CTree+bin, MOB+bin). Moreover, they are compared to the adapted GUIDE version 
 which includes all available scores (GUIDE+scores).
@@
 different values for the true split point $\xi$, however, all four situations
-lead to the same conclusions. Binarizing the score values decreases the proportion
-of replications where the split variable was selected correctly.
+lead to the same conclusions: Binarizing the score values decreases the selection probability
+of the true split variable $Z_1$, and hence reduces the power of the testing strategy.
 
@@
                                   "guide_sum_12"),
-                       labels = c("CTree","CTree_bin",
+                       labels = c("CTree","CTree+bin",
                                   #"CTree+max","CTree+max_bin",
-                                  "MOB", "MOB_bin",
+                                  "MOB", "MOB+bin",
                                   "GUIDE+scores"))
@@
 testing strategies CTree, CTree+bin, MOB, MOB+bin, and GUIDE+scores in the ``stump''
-scenario with both parameter varying. 
+scenario with both parameters varying. 
 Probabilities are estimated over increasing effect size $\delta$
@@
 In Figure~\ref{fig:cat_stump} the impact of categorizing split variables on the performance 
-is illustrated over increasing effect size for four different values of the true split 
-point $\xi$. For both, CTree and MOB, it can be stated that overall they perform better in their
-original form without any categorization. Only if the true split point $\xi$
-is close to the quartiles used as breakpoints for the categorizations both versions
-lead to a similar proportion of replications where the true splitting variable is detected
-(e.g., for CTree and $\xi=0.5$)
-
-Therefore, it can be summarized that a categorization of the values of the split variable
-does not lead to any advantages unless the true split point corresponds to one of the 
-quartiles used for the categorization. In most situations it even causes the performance to 
-get worse.
+is illustrated by the selection probability of $Z_1$ over increasing effect size and for four 
+different values of the true split point $\xi$. For both, CTree and MOB, it can be stated that 
+overall they perform better in their original form without categorization. Only if the true 
+split point $\xi$ is close to the quartiles used as breakpoints for the categorizations both 
+versions lead to a selection probability of $Z_1$ (e.g., for CTree and $\xi=0.5$)
+
+Therefore, these results support the conclusions drawn in the main manuscript: Categorizing the values 
+of the split variable does not lead to any advantages unless the true split point corresponds (or is at
+least close) to one of the quartiles used for the categorization. In most situations it even causes the 
+power of the testing strategy to decrease.
 
@@
                                   "guide_sum_12"),
-                       labels = c("CTree","CTree_cat",
+                       labels = c("CTree","CTree+cat",
                                   #"CTree+max","CTree+max_cat",
-                                  "MOB", "MOB_cat",
+                                  "MOB", "MOB+cat",
                                   "GUIDE+scores"))
@@
 testing strategies CTree, CTree+cat, MOB, MOB+cat, and GUIDE+scores in the ``stump''
-scenario with both parameter varying. 
+scenario with both parameters varying. 
 Probabilities are estimated over increasing effect size $\delta$

### 2019-05-23 pkg/partykit/inst/guideliketest/draft/draft.Rnw
minor revisions, mainly in the appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2587 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0d11ed92ff4d807656c2080c21605e0b2e9daee8

@@
 being computed from the analogous transformation of the normal distribution
-\citep[see][for a hands-on introduction]{Hothorn+Hornik+VanDeWiel:2006}.
+\citep[see][for a hands-on introduction and Appendix~\ref{app:teststat_ctree} for more details on the linear test statistic]{Hothorn+Hornik+VanDeWiel:2006}.
 
@@
 (subject to certain minimal subgroup size constraints). Unlike the CTree framework, MOB
-relies on classical unconditional inference. For more details see \cite{Zeileis+Hornik:2007}.
+relies on classical unconditional inference. For more details see \cite{Zeileis+Hornik:2007} and 
+Appendix~\ref{app:teststat_mob}.
 
@@
 Note that the split point in numeric variables $Z$ is searched over all possible splits,
-not just the four bins that were constructed for the $\chi^2$~test.
+not just the four bins that were constructed for the $\chi^2$~test. More details on the
+applied test statistic can be found in Appendix~\ref{app:teststat_guide}.
 
@@
 \hline
-Residuals vs.\ scores        & \textit{residuals}  & $r(Y,X,\hat{\beta})$\\
-\quad ($h$ transformation)   & \textit{scores}     & $s(Y,X,\hat{\beta})$\\
+Residuals vs.\ scores        & \textit{residuals}  & $h(Y) = r(Y,X,\hat{\beta})$\\
+\quad ($h$ transformation)   & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
 \hline
@@
 \section*{Acknowledgments}
-\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
+An extended research stay of Torsten Hothorn in Innsbruck (August
 2017 to January 2018) was financially supported by the Swiss National
-Science Foundation, grant number SNF IZSEZ0 177091.}
+Science Foundation, grant number SNF IZSEZ0 177091.
 
@@
 
-\section{Test statistics}
+\section{Test statistics}\label{app:teststat}
 This section provides further information on the test statistics applied in CTree, MOB, 
@@
 
-\subsection{CTree}
+\subsection{CTree}\label{app:teststat_ctree}
 To measure the association of a response $Y$ and each possible split variable $Z$ the
@@
 $$
-T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i, (Y_1,\ldots,Y_n))^{\top}\right)
-$$
-where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom transformation of the split variable $Z$ and the influence function $h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the response $Y = (Y_1, \ldots, Y_n)$ in a permutation symmetric way.
-Moreover, by applying the ``vec'' operator the resulting $p~\times~q$ matrix is converted into a $p q$ column vector.
+T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i)^{\top}\right)
+$$
+where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom 
+transformation of the split variable $Z$ and the influence function 
+$h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the response 
+$Y = (Y_1, \ldots, Y_n)$ in a permutation symmetric way and is set to 
+$h(Y_i) = s(Y_i, X_i, \hat{\beta})$ for a score-based approach. 
+Moreover, by applying the ``vec'' operator the resulting
+$p~\times~q$ matrix is converted into a $p q$ column vector.
 Following \cite{Strasser+Weber:1999}
@@
 
-\subsection{MOB}
+\subsection{MOB}\label{app:teststat_mob}
 The MOB algorithm employs an empirical fluctuation process $W$ to measure deviations of
@@
 $$
-where the model scores $s$ are sorted by the split variable Z by including the ordering permutation $\sigma(Z_{i})$. 
+with the model scores $s_i = s(Y_i, X_i, \hat{\beta}$ being sorted by the split variable Z by including the ordering permutation $\sigma(Z_{i})$. 
 To scale this partial sum process an estimate $\hat{J}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
@@
 To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied and the corresponding asymptotic distribution of $\lambda(W(t))$ can be obtained by employing the same functional to the limit process, i.e., $\lambda(W^0)$.
-For more details, particularly on calculating the corresponding $p$-values see Section~3.2. of the original paper \citep{Zeileis+Hothorn+Hornik:2008}.
-
-\subsection{GUIDE}
+
+One possible and intuitive choice for a functional in order to asses instabilities over a numerical 
+split variable is the following:
+$$\lambda_{supLM}(W) = \max_{i = \underline{i},\ldots,\overline{i}}\left(\frac{i}{n}\cdot \frac{n-i}{n}\right)^{-1} \left|\left|W\left(\frac{i}{n}\right)\right|\right|_2^2
+$$
+where a minimal segment size $\underline{i}$ and then $\overline{i} = n - \underline{i}$ are used to 
+define the interval $[\underline{i},\overline{i}]$. Other possible functionals, for example also
+for categorical split variables, and more details,
+particularly on calculating the corresponding $p$-values can be found in Section~3.2. of the 
+original paper \citep{Zeileis+Hothorn+Hornik:2008}.
+
+\subsection{GUIDE}\label{app:teststat_guide}
 The test statistic of the $\chi^2$~test as applied in the GUIDE algorithm is
 $$
-X^2 = \frac{(O_{11} - E_{11})^2}{E_{11}} + \ldots + \frac{(O_{11} - E_{11})^2}{E_{11}}
-$$
-where $O_{ij}$ are the observed frequencies of observations in a certain combination of binarized residuals ($i=1,2$) and categorized split variables ($j = 1,\ldots,4$).
-$E_{ij}$ are the corresponding expected frequencies under independence $E_{ij} = \frac{O_{i \cdot} \cdot O_{\cdot j}}{n}$.
+X^2 = \frac{(O_{11} - E_{11})^2}{E_{11}} + \ldots + \frac{(O_{24} - E_{24})^2}{E_{24}}
+$$
+where $O_{kl}$ are the observed frequencies of observations in a certain combination of binarized residuals ($k=1,2$) and categorized split variables ($l = 1,\ldots,4$).
+$E_{kl}$ are the corresponding expected frequencies under independence $E_{kl} = \frac{O_{k \cdot} \cdot O_{\cdot l}}{n}$.
 Under the null hypothesis of independence the asymptotic distribution of $X^2$ is a $\chi^2$~distribution allowing for a straight-forward calculation of $p$-values.

### 2019-05-30 pkg/partykit/inst/guideliketest/draft/draft.Rnw
first revisions based on feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2597 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 64636afb6a878945825813378d9b3fd805d69500

@@
 goodness-of-fit measure (residuals vs.\ model scores), binarization of residuals/scores
-at zero, binning of covariates. This shows that specifically the goodness-of-fit measure 
+at zero, and binning of covariates. This shows that specifically the goodness-of-fit measure 
 is crucial for the power of the procedures, with model scores without binarization performing
@@
 done by applying a tree algorithm. While almost all algorithms proposed in the literature
-follow the general idea of splitting the data such that some objective function is optimized, 
-they differ in their specific approaches to selecting a split variable and the corresponding 
-split point. 
+follow the general idea of splitting the data such that some objective function is optimized
+locally, they differ in their specific approaches to selecting a split variable and the 
+corresponding split point. 
 Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
@@
 In many of the considered scenarios the choice of goodness-of-fit measure heavily influences
-the performance of testing strategies with overall using model scores leading to clearly better
-results than employing residuals only. Moreover, the original
+the performance of testing strategies. In particular, using model scores leads to
+overall clearly better results than employing residuals only. Moreover, the original
 values of the goodness-of-fit measure are preferred over binarized versions of them.
@@
         the current subgroup, a disaggregated, observation-wise goodness-of-fit
-        measure is obtained. More formally, this is an $n \times k$ matrix
-        where $n \in \mathbb{N}$ is the number of observations and $k \in \mathbb{N}$ 
+        measure is obtained. More formally, this is an $N \times K$ matrix
+        where $N \in \mathbb{N}$ is the number of observations and $K \in \mathbb{N}$ 
         the number of goodness-of-fit measures per observation.
   \item The dependency or association of this goodness-of-fit matrix with
-        each possible split variable $Z_j$, $j \in \{1, \dots, m\}$, is assessed
+        each possible split variable $Z_j$, $j \in \{1, \dots, J\}$, is assessed
         using some suitable test statistic. The corresponding $p$-values allow for
-        a comparison of all $m$ covariates on a standardized or unified scale and in
+        a comparison of all $J$ covariates on a standardized or unified scale and in
         that way for an unbiased split variable selection.
@@
 observation-wise goodness-of-fit measure can be given by the residuals or the scores
-(gradient contributions). Analogously, the log-likelihood and corresponding score function could be used.
+(gradient contributions). Analogously, the log-likelihood and corresponding score function 
+could be used.
 
@@
 To select a split variable it is tested whether there is any association between the transformed 
-response $h(Y)$ and each possible transformed split variable $g(Z_j)$ ($j = 1, \dots, m$).
+response $h(Y)$ and each possible transformed split variable $g(Z_j)$ ($j = 1, \dots, J$).
 The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
@@
 This framework includes many different M-type estimators as special cases, including maximum 
-likelihood and ordinary least squares estimation. For a $k$-dimensional parameter
+likelihood and ordinary least squares estimation. For a $K$-dimensional parameter
 $\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
-is also a $k$-dimensional vector, i.e., the gradient contribution of the $i$-th observation.
-Thus, the $n \times k$-matrix consisting of these scores or gradient contributions for all 
-$i = 1, \dots, n$ observations is a natural candidate for the observation-wise goodness-of-fit
+is also a $K$-dimensional vector, i.e., the gradient contribution of the $i$-th observation.
+Thus, the $N \times K$-matrix consisting of these scores or gradient contributions for all 
+$i = 1, \dots, N$ observations is a natural candidate for the observation-wise goodness-of-fit
 measure as described in the previous Section~\ref{sec:generic_alg}.
@@
 the identity function $g(Z) = Z$. For categorical variables $g$ can also map its values to the 
-corresponding unity vectors by an indicator function $g(Z) = (0,\ldots,1,\ldots,0)$.
+corresponding unity vectors by an indicator function $g(Z) = (0,\ldots,1,\ldots,0)^{\top}$.
 
@@
 If both variables $Y$ and $Z$ are numeric the default independence test corresponds to a Pearson 
-correlation test. For one numeric and one categorical variable essentially a 1-way ANOVA
+correlation test. For one numeric and one categorical variable essentially a one-way ANOVA
 (analysis of variance) is employed while for two categorical variables a $\chi^2$~test is performed.
@@
 linear regression models could be fitted in the nodes of a tree as well as
-constant fits such as simple mean response $\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$.
+constant fits such as simple mean response~\mbox{$\hat{\beta} = \bar{y} = \frac{1}{N}\sum_{i=1}^N y_i$}.
 The tests for selecting the split variable are then based on the corresponding residuals,
@@
 But the building blocks could be easily re-combined to yield new types of tests. For example,
-in the GUIDE approach, a 1-way ANOVA can be used for assessing the association of the residuals
+in the GUIDE approach, a one-way ANOVA can be used for assessing the association of the residuals
 (without binarization) with the categorized split variable. Or alternatively, a multivariate
-1-way ANOVA can be used for the non-binarized scores as opposed to the residuals etc.
+one-way ANOVA can be used for the non-binarized scores as opposed to the residuals etc.
 
@@
 $$
-with response variable $Y$, regressor variable $X$, and error term $\epsilon$. This model is
-fitted, as usual, by ordinary least squares (OLS) to the observations in each notation,
+with response variable $Y$, regressor variable $X$, and error term $\epsilon$.
+In particular, in the investigated tree models the coefficients $\beta_0$ and $\beta_1$
+can depend on the possible split variables $Z_j$, $j=1,\ldots,J$, such that
+$$
+Y = \beta_0(Z_1, \ldots, Z_J) + \beta_1(Z_1, \ldots, Z_J) \cdot X + \epsilon.
+$$
+This model is fitted, as usual, by ordinary least squares (OLS) to the observations in each notation,
 yielding the parameter estimates $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
 To keep notation simple we present the following equations for the root node with all observations
-$\{(y_i,x_i)\}_{i=1,\ldots,n}$ and $n \in \mathbb{N}$. In subsequent nodes the same equations
+$\{(y_i,x_i)\}_{i=1,\ldots,N}$ and $N \in \mathbb{N}$. In subsequent nodes the same equations
 are used but just for a smaller subgroup.
@@
 The aggregated goodness-of-fit measure (or objective or loss function) in OLS estimation
-is the sum of squared residuals: $\sum_{i=1}^n \ell(y_i,x_i,\beta_0,\beta_1)$ where
+is the sum of squared residuals: $\sum_{i=1}^N \ell(y_i,x_i,\beta_0,\beta_1)$ where
 $$
@@
 s(y_i, x_i, \beta_0, \beta_1) = 
-\frac{\partial \ell}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
+\frac{\partial \ell}{\partial (\beta_0, \beta_1)^{f\top}}(y_i, x_i, \beta_0, \beta_1) = 
 -2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
@@
 of the scores is in fact the residual. However, it is complemented by a second component
-that captures the slope effect of $x_i$. Therefore, when computing the score for all $n$ observations
-this yields an $n \times 2$ score matrix whose first column corresponds to the $n$ residuals.
+that captures the slope effect of $x_i$. Therefore, when computing the score for all $N$ observations
+this yields an $N \times 2$ score matrix whose first column corresponds to the $N$ residuals.
 As this is the derivative with respect to the intercept parameter~$\beta_0$, it
@@
 as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
-not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_j(Z_1,Z_2)$, $j=0,1$.}
+not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_{k-1}(Z_1,Z_2)$, $k=1,2=K$.}
 \end{table}
@@
 See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
-changes ($\beta_j$ with $j\in\{0,1\}$) switches between two values at the split point $\xi$:
+changes~($\beta_{k-1}$ with $k=1,2=K$) switches between two values at the split point $\xi$:
 %
 \begin{align*}
-\beta_{j}(Z_1) = \begin{cases}
--\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \leq \xi \\
-+\delta \cdot (-1)^{j}  \quad \text{if } Z_1 > \xi
+\beta_{k-1}(Z_1) = \begin{cases}
+-\delta \cdot (-1)^{k-1}  \quad \text{if } Z_1 \leq \xi \\
++\delta \cdot (-1)^{k-1}  \quad \text{if } Z_1 > \xi
 \end{cases}
 \end{align*}
-%
-%while in the continuous case a simple linear function is considered in the form of
-%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
 In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
@@
 for the true split variable $Z_1$ and its corresponding selection
-probability (i.e., $Z_1$ being significant at 5\% level and having the
-lowest $p$-value among all split variables). Section~\ref{sec:scores} begins by
-highlighting the importance of using full model scores vs.\ residuals only
+probability (i.e., the association of $Y$ and $Z_1$ being significant at 5\% level 
+and having the lowest $p$-value among all split variables). Section~\ref{sec:scores} 
+begins by highlighting the importance of using full model scores vs.\ residuals only
 before Section~\ref{sec:3way} considers all building blocks (scores vs.\ residuals,
@@
 Moreover, as defined by the true tree structure used in the data generating process, the
-parameter functions $\beta_j(Z_1)$, $j=0,1$, are step functions which is particularly 
+parameter functions $\beta_{k-1}(Z_1)$, $k=1,2=K$, are step functions which is particularly 
 advantageous for the maximum selection in the test statistic applied by MOB while the linear 
@@
 the results for all possible combinations of the building blocks are discussed in a full 
-factorial 3-way analysis in the following section.
+factorial three-way analysis in the following section.
 
@@
 \hline
-Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)$\\
+Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)^{\top}$\\
 \quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
-                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)$\\
+                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)^{\top}$\\
                              &                     & \quad indicating the crossed quantile\\
@@
 point (such that categorizing split variables has no negative effect) the post-pruning strategy 
-compensates the lack of power of GUIDE+scores compared to CTree and MOB almost completely (bottom 
-row, first and third panel).
+compensates the lack of power of GUIDE+scores compared to CTree and MOB almost completely 
+(bottom row, first and third panel).
 In conclusion, the choice of a pruning strategy should be based on knowledge about the 
@@
 Linear models as presented in this study have been selected as they allow for a simple illustration 
-of the investigated subjects. However, the obtained results also hold for a variety of other models
-as the introduced building blocks can be applied in the same way for all model types.
+of the investigated subjects. However, the obtained results also hold for a variety of other 
+models as the introduced building blocks can be applied in the same way for all model types.
 Specific characteristics of linear models regarding structural changes have been investigated
@@
 \section*{Acknowledgments}
-An extended research stay of Torsten Hothorn in Innsbruck (August
-2017 to January 2018) was financially supported by the Swiss National
-Science Foundation, grant number SNF IZSEZ0 177091.
+Torsten Hothorn received funding from the Swiss National Science
+Foundation, grant number 200021\_184603.
 
@@
 \subsection{CTree}\label{app:teststat_ctree}
-To measure the association of a response $Y$ and each possible split variable $Z$ the
-CTree algorithm applies a linear test statistic $T$ which is excerpted from Section~3.1.
-(``Variable selection and stopping criteria'') of the original paper \citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
-$$
-T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i)^{\top}\right)
-$$
-where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom 
-transformation of the split variable $Z$ and the influence function 
-$h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the response 
-$Y = (Y_1, \ldots, Y_n)$ in a permutation symmetric way and is set to 
+To measure the association of a response $Y$ and each possible split variable $Z_j$, $j=1,\ldots,J$,
+the CTree algorithm applies a linear test statistic $T_j$ which is excerpted from Section~3.1.
+(``Variable selection and stopping criteria'') of the original paper 
+\citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
+$$
+T_j(Y, Z_j, w) = \text{vec}\left( \sum_{i=1}^N w_i g_j(Z_{ji})h(Y_i)^{\top}\right)
+$$
+where $w$ are optional weights, $g_j: \mathcal{Z}_j \rightarrow \mathbb{R}^{P}$ is a nonrandom 
+transformation of the split variable $Z_j$ and the influence function 
+$h: \mathcal{Y} \times \mathcal{Y}^N \rightarrow \mathbb{R}^Q$ depends on the response 
+$Y = (Y_1, \ldots, Y_N)$ in a permutation symmetric way and is set to 
 $h(Y_i) = s(Y_i, X_i, \hat{\beta})$ for a score-based approach. 
 Moreover, by applying the ``vec'' operator the resulting
-$p~\times~q$ matrix is converted into a $p q$ column vector.
+$P~\times~Q$ matrix is converted into a $PQ$ column vector.
 Following \cite{Strasser+Weber:1999}
-the conditional expectation $\mu$ and covariance $\Sigma$ of a linear test statistic can 
-be calculated and used to standardize an observed linear test statistic $t$ within a 
+the conditional expectation $\mu_j$ and covariance $\Sigma_j$ of a linear test statistic
+$T_j$ can be calculated and used to standardize an observed linear test statistic $t_j$ within a 
 function $c$ mapping into the real line.
@@
 $$
-c_{\max}(t,\mu,\Sigma) = \max_{k=1,\ldots,pq} \left| \frac{(t-\mu)}{\sqrt{(\Sigma)_{kk}}} \right|
+c_{\max}(t_j,\mu_j,\Sigma_j) = \max \left| \frac{(t_j-\mu_j)}{\text{diag}(\Sigma_j)} \right|
 $$
@@
 $$
-c_{\text{quad}}(t, \mu, \Sigma) = (t - \mu)\Sigma^{+} (t - \mu)^{\top}
-$$
-can be considered where $\Sigma^{+}$ is the Moore-Penrose inverse of $\Sigma$.
-Since the asymptotic conditional distribution of a linear test statistic $T$ is a 
-multivariate normal with parameters $\mu$ and $\Sigma$ \citep{Strasser+Weber:1999}, 
+c_{\text{quad}}(t_j, \mu_j, \Sigma_j) = (t_j - \mu_j)\Sigma_j^{+} (t_j - \mu_j)^{\top}
+$$
+can be considered where $\Sigma_j^{+}$ is the Moore-Penrose inverse of $\Sigma_j$.
+Since the asymptotic conditional distribution of a linear test statistic $T_j$ is a 
+multivariate normal with parameters $\mu_j$ and $\Sigma_j$ \citep{Strasser+Weber:1999}, 
 the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
@@
 \subsection{MOB}\label{app:teststat_mob}
-The MOB algorithm employs an empirical fluctuation process $W$ to measure deviations of
-the model scores $s$ from zero. As described in detail in Section~3.2. (``Testing for 
+The MOB algorithm employs an empirical fluctuation process $W_j$ to measure deviations of
+the model scores $s$ from zero with respect to an ordering induced by the possible split variable
+$Z_j$, $j = 1,\ldots,J$. As described in detail in Section~3.2. (``Testing for 
 parameter instability'') of the original paper \citep{Zeileis+Hothorn+Hornik:2008}
@@
 $$
-W(t) = \hat{J}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s_{\sigma(Z_{i})} \qquad (0 \leq t \leq 1)
-$$
-with the model scores $s_i = s(Y_i, X_i, \hat{\beta}$ being sorted by the split variable Z by including the ordering permutation $\sigma(Z_{i})$. 
-To scale this partial sum process an estimate $\hat{J}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
-Following \cite{Zeileis+Hornik:2007} $W(t)$ converges to a Brownian bridge $W^0$ under the null
+W_j(t) = \hat{\Sigma}^{-1/2} N^{-1/2} \sum_{i=1}^{\lfloor Nt \rfloor} s_{\pi(Z_{ji})} \qquad (0 \leq t \leq 1)
+$$
+with the model scores $s_i = s(Y_i, X_i, \hat{\beta})$ being sorted by the possible split variable $Z_j$ by including the ordering permutation $\pi(Z_{ji})$. 
+To scale this partial sum process an estimate $\hat{\Sigma}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
+Following \cite{Zeileis+Hornik:2007} $W_j(t)$ converges to a Brownian bridge $W^0$ under the null
 hypothesis of parameter stability.
-To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied and the corresponding asymptotic distribution of $\lambda(W(t))$ can be obtained by employing the same functional to the limit process, i.e., $\lambda(W^0)$.
+To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied and the corresponding asymptotic distribution of $\lambda(W_j(t))$ can be obtained by employing the same functional to the limit process, i.e., $\lambda(W^0)$.
 
@@
 split variable is the following:
-$$\lambda_{supLM}(W) = \max_{i = \underline{i},\ldots,\overline{i}}\left(\frac{i}{n}\cdot \frac{n-i}{n}\right)^{-1} \left|\left|W\left(\frac{i}{n}\right)\right|\right|_2^2
-$$
-where a minimal segment size $\underline{i}$ and then $\overline{i} = n - \underline{i}$ are used to 
+$$\lambda_\text{supLM}(W_j) = \max_{i = \underline{i},\ldots,\overline{i}}\left(\frac{i}{N}\cdot \frac{N-i}{N}\right)^{-1} \left|\left|W_j\left(\frac{i}{N}\right)\right|\right|_2^2
+$$
+where a minimal segment size $\underline{i}$ and then $\overline{i} = N - \underline{i}$ are used to 
 define the interval $[\underline{i},\overline{i}]$. Other possible functionals, for example also
@@
 $$
-where $O_{kl}$ are the observed frequencies of observations in a certain combination of binarized residuals ($k=1,2$) and categorized split variables ($l = 1,\ldots,4$).
-$E_{kl}$ are the corresponding expected frequencies under independence $E_{kl} = \frac{O_{k \cdot} \cdot O_{\cdot l}}{n}$.
+where $O_{lm}$ are the observed frequencies of observations in a certain combination of binarized residuals ($l=1,2$) and categorized split variables ($m = 1,\ldots,4$).
+$E_{lm}$ are the corresponding expected frequencies under independence $E_{lm} = \frac{O_{l \cdot} \cdot O_{\cdot m}}{N}$.
 Under the null hypothesis of independence the asymptotic distribution of $X^2$ is a $\chi^2$~distribution allowing for a straight-forward calculation of $p$-values.
@@
 \begin{align*}
-\beta_{j}(Z_1) = \begin{cases}
--\delta \cdot (-1)^{j} \cdot Z_1\\
-+\delta \cdot (-1)^{j} \cdot Z_1
+\beta_{k-1}(Z_1) = \begin{cases}
+-\delta \cdot (-1)^{k-1} \cdot Z_1\\
++\delta \cdot (-1)^{k-1} \cdot Z_1
 \end{cases}
 \end{align*}
-for $j = 0,1$.
+for $k = 1,2=K$.
 Additionally, a modified version of CTree employing a maximum selection, such as MOB, is evaluated 
@@
 Therefore, these results support the conclusions drawn in the main manuscript: Categorizing the values
-of the split variable does not lead to any advantages unless the true split point corresponds (or is at
-least close) to one of the quartiles used for the categorization. In most situations it even causes the 
-power of the testing strategy to decrease.
+of the split variable does not lead to any advantages unless the true split point corresponds (or is 
+at least close) to one of the quartiles used for the categorization. In most situations it even causes
+the power of the testing strategy to decrease.
 

### 2019-05-31 pkg/partykit/inst/guideliketest/draft/draft.Rnw
adapted tab:combinations_appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2598 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f303946b2469daa1cfbec4e7ee3139deccdc0cf8

@@
 \centering
-\begin{tabular}{ l l l l }
-\hline 
-Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
-\hline
-GUIDE             & residuals          & yes          & cat \\
-MOB+bin+res       & residuals          & yes          & max \\
-CTree+bin+res     & residuals          & yes          & lin \\
-MOB+cat+res       & residuals          & no           & cat \\
-MOB+res           & residuals          & no           & max \\
-CTree+res         & residuals          & no           & lin \\
-GUIDE+scores      & scores             & yes          & cat \\
-MOB+bin           & scores             & yes          & max \\
-CTree+bin         & scores             & yes          & lin \\
-MOB+cat           & scores             & no           & cat \\
-MOB               & scores             & no           & max \\
-CTree             & scores             & no           & lin \\
-\hline
+\begin{tabular}{ l l l l l}
+\hline\noalign{\smallskip} 
+Residuals/Scores   & Dichotomization & Categorization & \multicolumn{2}{l}{Via testing strategy}\\
+\noalign{\smallskip}\hline\noalign{\smallskip}
+residuals	   & yes	     & cat            & GUIDE & {(default)}\\
+residuals	   & yes	     & max            & MOB   & {(with modifications)}\\
+residuals	   & yes	     & lin            & CTree & {(with modifications)}\\
+residuals	   & no 	     & cat            & MOB   & {(with modifications)}\\
+residuals	   & no 	     & max            & MOB   & {(with modifications)}\\
+residuals	   & no 	     & lin            & CTree & {(with modifications)}\\
+scores  	   & yes	     & cat            & GUIDE & {(with modifications)}\\
+scores  	   & yes	     & max            & MOB   & {(with modifications)}\\
+scores  	   & yes	     & lin            & CTree & {(with modifications)}\\
+scores  	   & no 	     & cat            & MOB   & {(with modifications)}\\
+scores  	   & no 	     & max            & MOB   & {(default)}\\
+scores  	   & no 	     & lin            & CTree & {(default)}\\
+\noalign{\smallskip}\hline
 \end{tabular}
-\caption{Testing strategies applied (representing the 12 combinations
-of building blocks).
+\caption{Testing strategies applied for all 12 factorial combinations of the building blocks.
+For each combination one of the three testing strategies CTree/MOB/GUIDE is applied,
+possibly after modifying the assessed variables as indicated (i.e., dichotomized or categorized).
 \label{tab:combinations_appendix}}

### 2019-05-31 pkg/partykit/inst/guideliketest/draft/draft.Rnw
distinguish covariates into split variables and regressors, starting from Section 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2599 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7fa53cd4ed1e2a6b26e0ced8823012e73583fe67

@@
 and more homogeneous subgroups based on a set of covariates. Various tree 
-algorithms have been developed, following essentially the same general structure.
+algorithms have been developed, following essentially the same general structure,
+employing the covariates as \emph{split variables} in the tree induction.
 Starting at the root of the tree, pertaining to the full available data sample, the
@@
         some objective function (or loss function) that reflects the goodness of fit.
-  \item Among all available covariates one is selected as the split variable,
+  \item Among all available split variables one is selected as the split variable,
         choosing the split point such that goodness of fit is maximized in the
@@
 for example be the sum of deviations from a typical/average value, but it can 
-also be based on a model for a response along with potential covariates.
+also be based on a model for a response along with potential split variables.
 For instance, different types of residuals (or the signs thereof) can be employed
@@
         using some suitable test statistic. The corresponding $p$-values allow for
-        a comparison of all $J$ covariates on a standardized or unified scale and in
+        a comparison of all $J$ split variables on a standardized or unified scale and in
         that way for an unbiased split variable selection.
-  \item The covariate corresponding to the smallest $p$-value -- and thus the
+  \item The split variable corresponding to the smallest $p$-value -- and thus the
         highest influence on the goodness of fit of the model -- is selected for 
@@
 $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
-If a parametric model is fitted to the response $Y$ with regressor variable(s)~$X$,
+If a parametric model is fitted to the response $Y$ with some covariate(s)~$X$ employed as regressor variable(s),
 then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
@@
         Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
-        model for the response $Y$ and possible covariates $X$ can be considered. Either residuals
+        model for the response $Y$ and possible regressors $X$ can be considered. Either residuals
 	$r(Y,X,\hat{\beta})$ can be used as proposed for GUIDE or model scores $s(Y,X,\hat{\beta})$
@@
 (2)~If so, in which variable?
-For the first question the $p$-values regarding all available covariates are compared 
+For the first question the $p$-values regarding all available split variables are compared 
 to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
-than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.
+than $\alpha$ a split is performed and the split variable corresponding to this $p$-value is selected.
 

### 2019-05-31 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revision based on feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2600 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9c942a21810fead5e10876b57e76591ac8b0a04e

@@
 tree algorithms to select the appropriate covariates for splitting: observation-wise
-goodness-of-fit measure (residuals vs.\ model scores), binarization of residuals/scores
-at zero, and binning of covariates. This shows that specifically the goodness-of-fit measure 
-is crucial for the power of the procedures, with model scores without binarization performing
+goodness-of-fit measure (residuals vs.\ model scores), dichotomization of residuals/scores
+at zero, and binning of possible split variables. This shows that specifically the goodness-of-fit 
+measure is crucial for the power of the procedures, with model scores without dichotomization performing
 much better in many scenarios.
@@
 (1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
-(2)~Binarization of these residuals or scores.
-(3)~Categorization of covariates.
+(2)~Dichotomization of these residuals or scores.
+(3)~Categorization of possible split variables.
 Apart from these three main factors further aspects such as the approximation of the null
@@
 overall clearly better results than employing residuals only. Moreover, the original
-values of the goodness-of-fit measure are preferred over binarized versions of them.
+values of the goodness-of-fit measure are preferred over dichotomized versions of them.
 Also regarding the effects of categorizing possible split variables and the selection
@@
 To select a split variable it is tested whether there is any association between the transformed 
-response $h(Y)$ and each possible transformed split variable $g(Z_j)$ ($j = 1, \dots, J$).
+response $h(Y)$ and each possible transformed split variable $g(Z_j)$, $j = 1, \dots, J$.
 The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
@@
 $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
-If a parametric model is fitted to the response $Y$ with some covariate(s)~$X$ employed as regressor variable(s),
+If a parametric model is fitted to the response $Y$ with some covariate(s)~$X$ employed as 
+\textit{regressor} variable(s),
 then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
@@
 To construct a statistical test two additional transformations are carried out:
-(1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
+(1)~The residuals $r(Y,X,\hat{\beta})$ are dichotomized at zero, yielding an indicator for positive vs.\ negative residuals.
 (2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
 it is split at its quartiles into four bins.
-Subsequently, a $\chi^2$~test of independence is performed for the binarized residuals and
+Subsequently, a $\chi^2$~test of independence is performed for the dichotomized residuals and
 each categorized/categorical split variable.
@@
         explicitly allowed for different transformations $h(Y)$ in its original description already.
-  \item \emph{Binarization of residuals/scores:}\\
+  \item \emph{Dichotomization of residuals/scores:}\\
         Rather than testing independence between the split variables and the residuals/scores
 	themselves, it is possible	
-	to binarize residuals/scores at $0$ so that only their signs are assessed (as proposed for GUIDE).
+	to dichotomize residuals/scores at $0$ so that only their signs are assessed (as proposed for GUIDE).
   \item \emph{Categorization of split variables:}\\
@@
 When applying CTree for unbiased model-based recursive partitioning it has been suggested to use
-the model scores without binarization and assess their association with the untransformed split
+the model scores without dichotomization and assess their association with the untransformed split
 variables using a conditional inference test. This is similar to a squared correlation test statistic.
 MOB also employs the scores without
-binarization and maximally selects a score statistic over all potential split points in the
-split variable. GUIDE employs the binarized residuals and assesses their association with
+dichotomization and maximally selects a score statistic over all potential split points in the
+split variable. GUIDE employs the dichotomized residuals and assesses their association with
 the categorized split variable in a classical (unconditional) $\chi^2$ test.
@@
 in the GUIDE approach, a one-way ANOVA can be used for assessing the association of the residuals
-(without binarization) with the categorized split variable. Or alternatively, a multivariate
-one-way ANOVA can be used for the non-binarized scores as opposed to the residuals etc.
+(without dichotomization) with the categorized split variable. Or alternatively, a multivariate
+one-way ANOVA can be used for the non-dichotomized scores as opposed to the residuals etc.
 
@@
 \hline 
-      & Scores       & Binarization  & Categorization  & Statistic\\
+      & Scores       & Dichotomization  & Categorization  & Statistic\\
 \hline
@@
 This model is fitted, as usual, by ordinary least squares (OLS) to the observations in each notation,
-yielding the parameter estimates $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
+yielding the parameter estimates $\hat{\beta} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
 To keep notation simple we present the following equations for the root node with all observations
@@
 $$
-Thus, up to a constant scaling factor of $-2$ (that could also be omitted) the first component
-of the scores is in fact the residual. However, it is complemented by a second component
+Thus, up to a constant scaling factor of $-2$ (that could also be omitted) the first 
+component of the scores is in fact the residual. 
+However, it is complemented by a second component
 that captures the slope effect of $x_i$. Therefore, when computing the score for all $N$ observations
@@
 derivatives with respect to the slope parameter~$\beta_1$ and thus captures changes in this.
+Hence, GUIDE can be seen as a special case of CTree in which only a subset of the full model
+scores, in particular the derivatives only with respect to the intercept parameter, is used.
 
@@
 before Section~\ref{sec:3way} considers all building blocks (scores vs.\ residuals,
-binarization of these, and categorization of the split variable).
+dichotomization of these, and categorization of the split variable).
 Subsequently, the ``tree'' scenario is employed to investigate how the
@@
 additionally consider a new GUIDE flavor, denoted GUIDE+scores, that uses
-binarized scores rather than residuals.
+dichotomized scores rather than residuals.
 
@@
 From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
-can be deduced that categorizing split variables and binarizing scores also decreases the
+can be deduced that categorizing split variables and dichotomizing scores also decreases the
 ability to detect the right split variable. To elaborate these effects more in detail 
@@
 \hline
-Binarization                 & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
-\quad (of $h$)               & \textit{no}         & $h$ without binarization\\
+Dichotomization              & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
+\quad (of $h$)               & \textit{no}         & $h$ without dichotomization\\
 \hline
@@
 
-Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing 
+Based on the results displayed in Figure~\ref{fig:3way} it can be stated that dichotomizing 
 residuals/scores decreases the performance as it leads to higher $p$-values corresponding 
@@
 a categorization weakens the performance of a testing strategy. 
-Moreover, while the effect of both of these transformations of the data (binarization of 
+Moreover, while the effect of both of these transformations of the data (dichotomization of 
 residuals/scores and categorization of split variables) can be observed
@@
                                    bg = "gray90"),
-       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Binarization:", "  yes", "  no")),
+       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
                   columns = 2,
@@
 \caption{\label{fig:3way} Effects of the three building blocks \textit{residuals vs.\ scores},
-\textit{binarization}, and \textit{categorization} in the ``stump'' scenario with the intercept
+\textit{dichotomization}, and \textit{categorization} in the ``stump'' scenario with the intercept
 and the slope parameter both varying. All possible combinations have been 
@@
 drawn from the ``stump'' scenario: Including all available scores in the testing strategy leads to a better performance 
-while binarizing scores and categorizing split variables both do not cause any improvement but in many situations even 
+while dichotomizing scores and categorizing split variables both do not cause any improvement but in many situations even 
 have a negative effect on the power of a testing strategy.
@@
 clearly better results and performs in general at least equally well.
-(2)~Binarizing scores/residuals does not lead to any improvement in performance.
+(2)~Dichotomizing scores/residuals does not lead to any improvement in performance.
 (3)~Categorizing possible split variables also doesn't improve the performance 
@@
 that are not connected to the mean might be missed when only considering residuals. This can be prevented 
-by including scores which are a standard measurement of goodness of fit that can be easily employed for 
-all model types.
+by including scores which are a standard measurement of goodness of fit that can be easily employed 
+for all model types. Moreover, the scores function $s$ can simply be seen as a  transformation to 
+a different space in order to allow for a well structured analysis of dependencies, regardless
+of the type of model.
 Even though it also depends on the type of the applied residuals 
@@
 modifications have been applied within these functions allowing for a categorization
-of possible split variables and a binarization of scores.
+of possible split variables and a dichotomization of scores.
 To evaluate the GUIDE algorithm a reimplementation of this algorithm has been 
@@
 $$
-where $O_{lm}$ are the observed frequencies of observations in a certain combination of binarized residuals ($l=1,2$) and categorized split variables ($m = 1,\ldots,4$).
+where $O_{lm}$ are the observed frequencies of observations in a certain combination of dichotomized residuals ($l=1,2$) and categorized split variables ($m = 1,\ldots,4$).
 $E_{lm}$ are the corresponding expected frequencies under independence $E_{lm} = \frac{O_{l \cdot} \cdot O_{\cdot m}}{N}$.
@@
 of building blocks are listed in Table~\ref{tab:combinations_appendix}.
-Each suffix indicates a change of the original form of the considered testing strategy.
-In particular, it relates to the transformations of the data before applying a statistical
-test as represented by the different building blocks. In that way,
-``+res'' refers to using only residuals (instead of all model scores),
-``+scores'' to using all model scores (instead of only residuals),
-``+cat'' to applying a categorization of split variables,
-``+bin'' to applying a binarization of scores/residuals.
+%Each suffix indicates a change of the original form of the considered testing strategy.
+%In particular, it relates to the transformations of the data before applying a statistical
+%test as represented by the different building blocks. In that way,
+%``+res'' refers to using only residuals (instead of all model scores),
+%``+scores'' to using all model scores (instead of only residuals),
+%``+cat'' to applying a categorization of split variables,
+%``+bin'' to applying a dichotomization of scores/residuals.
 
@@
 
-\subsection{Binarization of residuals/scores}
-\label{app:binarization}
-To elaborate over an increasing effect size whether a binarization of the residuals/scores at zero 
+\subsection{Dichotomization of residuals/scores}
+\label{app:dichotomization}
+To elaborate over an increasing effect size whether a dichotomization of the residuals/scores at zero 
 leads to an improvement or a deterioration of performance, CTree and MOB are applied, once in their 
-original version without binarization and once in an adapted version with a binarization of 
-scores at zero (CTree+bin, MOB+bin). Moreover, they are compared to the adapted GUIDE version 
+original version without dichotomization and once in an adapted version with a dichotomization of 
+scores at zero (CTree+dich, MOB+dich). Moreover, they are compared to the adapted GUIDE version 
 which includes all available scores (GUIDE+scores).
@@
 
-Figure~\ref{fig:bin_stump} shows the effect of binarizing the score values over
+Figure~\ref{fig:bin_stump} shows the effect of dichotomizing the score values over
 different values for the true split point $\xi$, however, all four situations
-lead to the same conclusions: Binarizing the score values decreases the selection probability
+lead to the same conclusions: Dichotomizing the score values decreases the selection probability
 of the true split variable $Z_1$, and hence reduces the power of the testing strategy.
@@
                                   "guide_sum_12"),
-                       labels = c("CTree","CTree+bin",
+                       labels = c("CTree","CTree+dich",
                                   #"CTree+max","CTree+max_bin",
-                                  "MOB", "MOB+bin",
+                                  "MOB", "MOB+dich",
                                   "GUIDE+scores"))
@@
 Selection probability of the true split variable $Z_1$ for
-testing strategies CTree, CTree+bin, MOB, MOB+bin, and GUIDE+scores in the ``stump''
-scenario with both parameters varying. 
+testing strategies CTree, CTree+dich, MOB, MOB+dich, and GUIDE+scores in the ``stump''
+scenario with both parameters varying. CTree+dich and MOB+dich refer to modifications
+of CTree and MOB applying a dichotomization of residuals/scores.
 Probabilities are estimated over increasing effect size $\delta$

### 2019-05-31 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revision based on feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2601 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a6d983645486a120a96628e6f2c30bb8773c46a3

@@
 for the parameter estimators. 
-All these approaches have shown to work well for various situations, however, the relative
+All these approaches have been shown to work well for various situations, however, the relative
 (dis)advantages of the testing strategies have not yet been investigated and compared in detail.
@@
 
-% In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
+% In the CTree algorithm the number of possible breaks is fixed as splits can only be made at
 % observed values or optionally at an average value lying between two observations. To select the
@@
 derivatives with respect to the slope parameter~$\beta_1$ and thus captures changes in this.
-Hence, GUIDE can be seen as a special case of CTree in which only a subset of the full model
-scores, in particular the derivatives only with respect to the intercept parameter, is used.
+
+Hence, tests based on the full scores (as in CTree and MOB) include residual-based tests (as 
+in GUIDE) as a special case. Therefore, score-based tests can in principle capture all changes 
+in the objective function that residual-based tests can capture -- but the reverse is not
+necessarily true. In the next sections we will investigate how relevant this is in practice 
+and how much it depends on the concrete test statistics employed.
 
@@
 response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
-If quartiles are used as breakpoints to categorize a split variable the employed
-function $g$ maps to a four-dimensional vector of zeros except for one value $1$
-indication the bin to which an observation is assigned. For a maximum selection
-of a split variable $g$ maps to a vector containing only zeros until a certain
-changepoint from which on the values are replaced by $1$. This breakpoint indicates
-for example a crossed quantile. In a linear selection strategy $g$ is simply the identity function.
+In the case of categorization, the split variable $Z$ is binned at the
+quartiles. This corresponds to a four-dimensional 0/1 transformation
+function $g(Z)$ that indicates into which of the bins each observation falls.
+Maximum selection across potential split points in a variable $Z$ also
+corresponds to a multivariate 0/1 transformation function $g(Z)$. However,
+in this case for each potential split point an indicator is used that is 0
+before and 1 after the respective split point.
 (See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
@@
                              & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)^{\top}$\\
-                             &                     & \quad indicating the crossed quantile\\
+                             &                     & \quad indicating the potential split point\\
                              & \textit{lin}        & $g(Z)=Z$\\
@@
 selection of a split variable is advantageous. 
-Comparing the two panels suggests that unless one of the breakpoints used for binning 
+Comparing the two panels suggests that unless one of the breaks used for binning 
 possible split variables is close to the true split point (as for the median $0$, left panel), 
@@
 \label{sec:discussion}
-
 From comparing the effects of the building blocks three main conclusions can be drawn: 
@@
 (3)~Categorizing possible split variables also doesn't improve the performance 
-unless the breakpoints used for the categorization are very close to the true split points.
+unless the breaks used for the categorization are very close to the true split points.
 However, even then the results are very similar to those achieved without any categorization.
@@
 straightforward, for example in case of clearly separated subgroups being defined by the values 
-of a continuous split variable. Therefore, categorizing possible split variable can have
+of a continuous split variable. Therefore, categorizing possible split variables can have
 advantages in practice, however, it has to be considered that in general
 important effects might be neglected.
+
+Both of the transformation functions $h$ and $g$ (applied to the response~$Y$ or a split 
+variable~$Z$ respectively) can also be multivariate. For this case, the corresponding
+statistical tests are available, in particular the \textsf{R} package \textbf{coin}
+which is employed for the implementation of MOB and CTree provides 
+the required toolbox for multivariate dependency tests, including a $\chi^2$~test.
 
@@
 This is not the case if changes are orthogonal to the mean regressor $(1,0)^{\top}$, hence, if
-the parameter vector $\bm{\beta}$ is shifted 
-%to $\bm{\beta} + \bm{\Delta}$ 
-by $\bm{\Delta} = (0, \Delta_1)^{\top}$.
+the parameter vector $\beta$ is shifted 
+%to $\beta + \Delta$ 
+by $\Delta = (0, \Delta_1)^{\top}$.
 For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
@@
 In situations where shifts appear only in the intercept testing procedures purely based on residuals 
-perform best. However, the cost in terms of power that comes with including all scores, in particular
-with testing more than one parameter, is negligibly small.
-
-Also in more general models residual-based and score-based procedures are expected to perform equally 
-well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
-that are not connected to the mean might be missed when only considering residuals. This can be prevented 
-by including scores which are a standard measurement of goodness of fit that can be easily employed 
-for all model types. Moreover, the scores function $s$ can simply be seen as a  transformation to 
-a different space in order to allow for a well structured analysis of dependencies, regardless
-of the type of model.
-Even though it also depends on the type of the applied residuals 
-(see, e.g., \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002} 
-and \citealp{Loh+He+Man:2015})
-whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
-fit such as residuals provides less power than full model scores. 
+perform best. However, the cost in terms of power that comes with including all scores, in 
+particular with testing more than one parameter, is negligibly small.
+
+Also in more general models residual-based and score-based procedures are expected to perform
+equally well if all parameters are highly correlated. On the other hand, if parameters are 
+orthogonal changes that are not connected to the mean might be missed when only considering
+residuals. This can be prevented by including scores which are a standard measurement of goodness 
+of fit that can be easily employed for all model types. Moreover, the score function $s$ can 
+simply be seen as a  transformation to a different space in order to allow for a well structured
+analysis of dependencies, regardless of the type of model and therefore applicable in a
+wide variety of situations.
+In order to benefit from the advantages of using scores 
+some already existing models have been adapted, e.g., in \cite{Hothorn+Zeileis:2017} 
+and \cite{Schlosser+Hothorn+Stauffer:2019}. Similarly, variations of the GUIDE algorithm
+using different types of residuals are provided, e.g., in \citealp{Loh+Zheng:2013}, 
+\citealp{Chaudhuri+Loh:2002} and \citealp{Loh+He+Man:2015}.
+But even though it also depends on the type of the applied residuals whether parameter changes 
+can be detected successfully, a one-dimensional measurement of goodness of fit such as residuals
+provides less power than full model scores. 
 Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
@@
 $$
-W_j(t) = \hat{\Sigma}^{-1/2} N^{-1/2} \sum_{i=1}^{\lfloor Nt \rfloor} s_{\pi(Z_{ji})} \qquad (0 \leq t \leq 1)
+W_j(t) = \hat{V}^{-1/2} N^{-1/2} \sum_{i=1}^{\lfloor Nt \rfloor} s_{\pi(Z_{ji})} \qquad (0 \leq t \leq 1)
 $$
 with the model scores $s_i = s(Y_i, X_i, \hat{\beta})$ being sorted by the possible split variable $Z_j$ by including the ordering permutation $\pi(Z_{ji})$. 
-To scale this partial sum process an estimate $\hat{\Sigma}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
+To scale this partial sum process an estimate $\hat{V}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
 Following \cite{Zeileis+Hornik:2007} $W_j(t)$ converges to a Brownian bridge $W^0$ under the null
@@
 $$
-where $O_{lm}$ are the observed frequencies of observations in a certain combination of dichotomized residuals ($l=1,2$) and categorized split variables ($m = 1,\ldots,4$).
-$E_{lm}$ are the corresponding expected frequencies under independence $E_{lm} = \frac{O_{l \cdot} \cdot O_{\cdot m}}{N}$.
-Under the null hypothesis of independence the asymptotic distribution of $X^2$ is a $\chi^2$~distribution allowing for a straight-forward calculation of $p$-values.
-If full model scores are used instead of residuals a sum of the test statistic over the number of distribution parameters is considered such that each 
-summand corresponds to one column of the score matrix.
+where $O_{lm}$ are the observed frequencies of observations in a certain combination of 
+dichotomized residuals ($l=1,2$) and categorized split variables ($m = 1,\ldots,4$). $E_{lm}$ 
+are the corresponding expected frequencies under independence 
+$E_{lm} = \frac{O_{l \cdot} \cdot O_{\cdot m}}{N}$. 
+Under the null hypothesis of independence the asymptotic distribution of $X^2$ is a 
+$\chi^2$~distribution allowing for a straight-forward calculation of $p$-values.
+If full model scores are used instead of residuals a sum of the test statistic over the 
+number of distribution parameters is considered such that each summand corresponds to one 
+column of the score matrix. For this case the degrees of freedom of the resulting 
+$\chi^2$~distribution is then also the corresponding sum of degrees of freedom over the 
+number of distribution parameters.
 
@@
 overall they perform better in their original form without categorization. Only if the true 
-split point $\xi$ is close to the quartiles used as breakpoints for the categorizations both 
+split point $\xi$ is close to the quartiles used as breaks for the categorizations both 
 versions lead to a selection probability of $Z_1$ (e.g., for CTree and $\xi=0.5$)

### 2019-05-31 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revision based on feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2603 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9618f197678a5f0698391a05193202ce650cc61a

@@
 (See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
-for a more detailed overview of all 12 combinations.)
+for a more detailed overview of all 12 combinations of the building blocks.)
 
@@
 variable~$Z$ respectively) can also be multivariate. For this case, the corresponding
-statistical tests are available, in particular the \textsf{R} package \textbf{coin}
-which is employed for the implementation of MOB and CTree provides 
+statistical tests are available, in particular in the \textsf{R} package \textbf{coin}
+which is employed for the implementation of MOB and CTree and provides 
 the required toolbox for multivariate dependency tests, including a $\chi^2$~test.
@@
 residuals. This can be prevented by including scores which are a standard measurement of goodness 
-of fit that can be easily employed for all model types. Moreover, the score function $s$ can 
+of fit that can be easily employed for all model types. Moreover, the score function~$s$ can 
 simply be seen as a transformation to a different space in order to allow for a well structured

### 2019-06-04 pkg/partykit/inst/guideliketest/draft/draft.Rnw
fine-tuning of layout and pagebreaks
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2607 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7662f70e5bc36c48644010a8983fb7e82dda480a

@@
 
-\begin{table}
+\begin{table}[t!]
 \centering
@@
 
-\begin{figure}%[t]
+\begin{figure}[p!]
 \setkeys{Gin}{width=1\linewidth}
 \minipage{0.45\textwidth}
-<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
+<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
 #par(mar=c(3,3,0,0), oma = c(3,3,0,0))
@@
 \minipage{0.45\textwidth}
-<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
+<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
 par(mar = c(4,4,2,4))
@@
 \minipage{0.45\textwidth}
-<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
+<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
 d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
@@
 \minipage{0.45\textwidth}
-<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
+<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
 d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
@@
 
-\subsubsection{``Tree'' scenario}
-For the second simulation scenario (``tree'' scenario) the same basic setting with the same 
-variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
-However, not only $Z_1$ but also $Z_2$ is used as a true split variable following a uniform distribution
-on $[-1,1]$. Therefore, an additional split
-is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
-change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
-to a change in the intercept $\beta_0$.
-Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
-$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
-the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
-%
-\begin{align*}
-\beta_{0}(Z_1, Z_2) = \begin{cases}
-0       \qquad \text{if } Z_2 \leq \xi \\
--\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
-+\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
-\end{cases}
-\end{align*}
-\begin{align*}
-\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
-+\delta \quad \text{if } Z_2 \leq \xi \\
--\delta \quad \text{if } Z_2 > \xi
-\end{cases}
-\end{align*}
-%
-The underlying tree structure for the data generating process is illustrated in 
-Figure~\ref{fig:dgp_tree}.
-
 <<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=
-
 # d <- dgp_tree(nobs = 250)
@@
 
-\begin{figure}%[t]
+\begin{figure}[p!]
 \begin{center}
@@
 \minipage{0.5\textwidth}
-<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
+<<dgp_tree1, fig=TRUE, echo=FALSE, width=5.2, height=4.4>>=
 par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
@@
 
+\subsubsection{``Tree'' scenario}
+
+For the second simulation scenario (``tree'' scenario, see Figure~\ref{fig:dgp_tree}) the same basic setting with the same 
+variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
+However, not only $Z_1$ but also $Z_2$ is used as a true split variable following a uniform distribution
+on $[-1,1]$. Therefore, an additional split
+is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
+change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
+to a change in the intercept $\beta_0$.
+Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
+$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
+the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
+%
+\begin{align*}
+\beta_{0}(Z_1, Z_2) = \begin{cases}
+0       \qquad \text{if } Z_2 \leq \xi \\
+-\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
++\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
+\end{cases}
+\end{align*}
+\begin{align*}
+\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
++\delta \quad \text{if } Z_2 \leq \xi \\
+-\delta \quad \text{if } Z_2 > \xi
+\end{cases}
+\end{align*}
+
 \subsection{Evaluation}
@@
 
-\begin{table}[h!]
+\begin{table}[t!]
 \centering
@@
 
-\begin{table}[h!]
+\begin{table}[p!]
 \centering
@@
 
-\begin{figure}[t]
+\begin{figure}[p!]
 \setkeys{Gin}{width=\linewidth}
@@
 
-In the simulation study presented in Sections~\ref{sec:simulation} and \ref{sec:results} the intercept
-and slope parameters $\beta_0$ and $\beta_1$ are both either fixed or binary variables taking
-either the positive or the negative value of the effect size $\delta$. Therefore, these parameters
-are step functions for which the maximum selection employed in the testing strategy of MOB has shown 
-to perform better than the linear selection in CTree. However, this changes in case of
-monotonous functions where CTree is advantageous as it is constructed to detect monotonous effects.
-To point out that the difference in performance of CTree and MOB depends on the type of effect
-the same setting for which the results are presented in Figure~\ref{fig:scores_stump} is
-evaluated again but this time with the varying parameter(s) changing continuously.
-In particular, the parameters $\beta_0$ and $\beta_1$ are linear functions of the true
-split variable $Z_1$:
-\begin{align*}
-\beta_{k-1}(Z_1) = \begin{cases}
--\delta \cdot (-1)^{k-1} \cdot Z_1\\
-+\delta \cdot (-1)^{k-1} \cdot Z_1
-\end{cases}
-\end{align*}
-for $k = 1,2=K$.
-Additionally, a modified version of CTree employing a maximum selection, such as MOB, is evaluated 
-(denoted by CTree+max). Looking at the selection probability illustrated in Figure~\ref{fig:stump_contbeta} 
-it can be observed that in this setting the original version of CTree is ahead while CTree+max and MOB perform 
-almost equally well. Hence, the comparison of Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} 
-points out that it clearly depends on the type of effect whether the maximum selection (MOB, CTree+max) 
-or the linear selection (CTree) is to be preferred.
-
-\begin{figure}%%[t]
+\begin{figure}[t!]
 \setkeys{Gin}{width=\linewidth}
@@
 
+In the simulation study presented in Sections~\ref{sec:simulation} and \ref{sec:results} the intercept
+and slope parameters $\beta_0$ and $\beta_1$ are both either fixed or binary variables taking
+either the positive or the negative value of the effect size $\delta$. Therefore, these parameters
+are step functions for which the maximum selection employed in the testing strategy of MOB has shown 
+to perform better than the linear selection in CTree. However, this changes in case of
+monotonous functions where CTree is advantageous as it is constructed to detect monotonous effects.
+To point out that the difference in performance of CTree and MOB depends on the type of effect
+the same setting for which the results are presented in Figure~\ref{fig:scores_stump} is
+evaluated again but this time with the varying parameter(s) changing continuously.
+In particular, the parameters $\beta_0$ and $\beta_1$ are linear functions of the true
+split variable $Z_1$:
+\begin{align*}
+\beta_{k-1}(Z_1) = \begin{cases}
+-\delta \cdot (-1)^{k-1} \cdot Z_1\\
++\delta \cdot (-1)^{k-1} \cdot Z_1
+\end{cases}
+\end{align*}
+for $k = 1,2=K$.
+Additionally, a modified version of CTree employing a maximum selection, such as MOB, is evaluated 
+(denoted by CTree+max). Looking at the selection probability illustrated in Figure~\ref{fig:stump_contbeta} 
+it can be observed that in this setting the original version of CTree is ahead while CTree+max and MOB perform 
+almost equally well. Hence, the comparison of Figures \ref{fig:scores_stump} and \ref{fig:stump_contbeta} 
+points out that it clearly depends on the type of effect whether the maximum selection (MOB, CTree+max) 
+or the linear selection (CTree) is to be preferred.
+
 \section{Results over increasing effect size}
@@
 
-\begin{figure}%[t]
+\subsection{Categorization of splitting variables}
+\label{app:categorization}
+To investigate the effect of categorizing the possible splitting variables CTree and MOB are
+applied, once in their original version without a categorization and once in an adapted version
+with a categorization of the possible split variables (CTree+cat and MOB+cat). 
+Moreover, they are also compared to GUIDE+scores which includes all available scores.
+
+In Figure~\ref{fig:cat_stump} the impact of categorizing split variables on the performance 
+is illustrated by the selection probability of $Z_1$ over increasing effect size and for four 
+different values of the true split point $\xi$. For both, CTree and MOB, it can be stated that 
+overall they perform better in their original form without categorization. Only if the true 
+split point $\xi$ is close to the quartiles used as breaks for the categorizations both 
+versions lead to a selection probability of $Z_1$ (e.g., for CTree and $\xi=0.5$)
+
+Therefore, these results support the conclusions drawn in the main manuscript: Categorizing the values
+of the split variable does not lead to any advantages unless the true split point corresponds (or is 
+at least close) to one of the quartiles used for the categorization. In most situations it even causes
+the power of the testing strategy to decrease.
+
+\newpage
+
+\begin{figure}[h!]
 \setkeys{Gin}{width=\linewidth}
@@
 }
-\end{figure}
-
-\subsection{Categorization of splitting variables}
-\label{app:categorization}
-To investigate the effect of categorizing the possible splitting variables CTree and MOB are
-applied, once in their original version without a categorization and once in an adapted version
-with a categorization of the possible split variables (CTree+cat and MOB+cat). 
-Moreover, they are also compared to GUIDE+scores which includes all available scores.
-
-In Figure~\ref{fig:cat_stump} the impact of categorizing split variables on the performance 
-is illustrated by the selection probability of $Z_1$ over increasing effect size and for four 
-different values of the true split point $\xi$. For both, CTree and MOB, it can be stated that 
-overall they perform better in their original form without categorization. Only if the true 
-split point $\xi$ is close to the quartiles used as breaks for the categorizations both 
-versions lead to a selection probability of $Z_1$ (e.g., for CTree and $\xi=0.5$)
-
-Therefore, these results support the conclusions drawn in the main manuscript: Categorizing the values
-of the split variable does not lead to any advantages unless the true split point corresponds (or is 
-at least close) to one of the quartiles used for the categorization. In most situations it even causes
-the power of the testing strategy to decrease.
-
-\begin{figure}%%[t]
+
 \setkeys{Gin}{width=\linewidth}
@@
 \end{figure}
+
+\clearpage
 

### 2019-06-05 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised Section 5
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2608 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1f12521d29aa3827abb6a1d208e3d10c344e7607

@@
 Hirschengraben 84\\
-CH-8001 Z\"urich, Switzerland \\
+8001 Z\"urich, Switzerland \\
 E-mail: \email{Torsten.Hothorn@R-project.org}\\
@@
 A crucial difference between the testing strategies of CTree/MOB and GUIDE
-is the difference between using only the residuals vs. the full model
-scores. In the literature on structural change tests there is evidence
+is the difference between using only the residuals vs.\ the full model
+scores. In the literature on structural change tests it is well-established
 that residual-based tests can only capture parameter differences that
@@
 In Figure~\ref{fig:scores_stump} the performances of the testing strategies are 
-represented by the corresponding selection probability, hence, a significance level 
-is incorporated. To show that in general the drawn conclusions do not change if no 
-significance level is included, a different illustration of the exact same evaluation 
-is provided in Appendix~\ref{app:significance} where the performances are compared based 
-on the smallest $p$-values, regardless of whether they are beneath any significance level.
-
-Moreover, as defined by the true tree structure used in the data generating process, the
-parameter functions $\beta_{k-1}(Z_1)$, $k=1,2=K$, are step functions which is particularly 
-advantageous for the maximum selection in the test statistic applied by MOB while the linear 
-selection in CTree is preferable for detecting monotonous dependencies. To show that employing 
-a continuous parameter function instead of a step function only has an impact on whether CTree or
-MOB performs better the exact same simulation setting as used for Figure~\ref{fig:scores_stump} 
-but with continuously changing parameters has been evaluated and is illustrated in 
-Appendix~\ref{app:contbeta}.
+represented by the corresponding selection probability, i.e., a significance level 
+is incorporated. (Appendix~\ref{app:significance} shows that the same qualitative
+conclusions can be drawn when the significance level is \emph{not} included.)
+For a split point at the median ($\xi=0$, top row) all testing strategies 
+perform similarly well as long as the intercept varies (left and right panel) with CTree and 
+MOB being only slightly ahead. However, for the split point at the $90$\% quantile $\xi = 0.8$ 
+(bottom row) the performance of all of the applied strategies decreases. 
+While both GUIDE versions struggle to detect the correct split variable even for a high effect 
+size, MOB is clearly ahead with CTree leading to the second best results.
+This advantage of MOB over CTree is mainly due to the abrupt shift in the 
+model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
+a smooth transition with continuously-changing parameters (see Appendix~\ref{app:contbeta}).
+
+However, in the scenario where only $\beta_1$ (but not the intercept $\beta_0$) is affected by the split,
+the residual-based GUIDE approach has no power at all even for a true split at the median (top middle panel).
+It is easily possible, though, to substantially mitigate this problem by using scores
+(sensitive to changes in all parameters) rather than residuals only (sensitive to changes
+in the conditional mean). The remaining difference between MOB/CTree and GUIDE+scores
+is due to dichotomizing the scores at zero and due to categorizing the split variables
+which are investigated in more detail in the following section.
 
@@
 
-Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
-model scores) perform clearly better than the original GUIDE version (using residuals 
-only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
-of replications. Even though the advantage of CTree and MOB evens out in certain situations
-such as for a high effect size (see, e.g., for the median $0$ and varying $\beta_0$), 
-GUIDE is never ahead of them in any of the considered settings. Moreover, the positive 
-effect of including all available scores can be observed particularly well in the top middle 
-panel where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to GUIDE. 
-%The fact that residual-based methods fail to detect changes that do not effect the
-%expected value has already been elaborated in the literature of structural change
-%analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
-%Section~\ref{sec:discussion}. 
-Furthermore, this advantage of including full models scores 
-clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
-to GUIDE+scores in case of a fixed slope parameter (left panels).
-
-For a central split point ($\xi=0$ corresponding to the median, upper panels) all strategies 
-perform similarly well as long as the intercept varies (left and right panel) with CTree and 
-MOB being only slightly ahead. However, for the split point at the $90$\% quantile $0.8$ 
-(lower panels) the performance of all of the applied strategies decreases. 
-While both GUIDE versions struggle to detect the correct split variable even for a high effect 
-size, MOB is clearly ahead with CTree leading to the second best results.
-As discussed before, this advantage of MOB over CTree is mainly due to the abrupt shift in the 
-model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
-continuously changing parameters (see Appendix~\ref{app:contbeta}).
-From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
-can be deduced that categorizing split variables and dichotomizing scores also decreases the
-ability to detect the right split variable. To elaborate these effects more in detail 
-the results for all possible combinations of the building blocks are discussed in a full 
-factorial three-way analysis in the following section.
-
 \subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
+
 To investigate the impact of each of the building blocks separately the most general 
 case of the ``stump'' scenario where the intercept and slope parameter are both varying is 
-considered. In the underlying evaluation all possible combinations of the building blocks 
+considered. In this evaluation all possible combinations of the building blocks 
 have been included. The different levels of each of the three building blocks are listed in 
@@
 response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}.
+
 In the case of categorization, the split variable $Z$ is binned at the
@@
 before and 1 after the respective split point.
-(See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
+(See also Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
 for a more detailed overview of all 12 combinations of the building blocks.)
@@
 Based on the results displayed in Figure~\ref{fig:3way} it can be stated that dichotomizing 
-residuals/scores decreases the performance as it leads to higher $p$-values corresponding 
-to the true split variable $Z_1$ and thus to lower power in the presented settings. 
-For a central true split point ($\xi=0$, left panel) this effect remains at an almost constant 
-level regardless of whether possible split variables are categorized or not. However, in case 
-of the true split point at the $90$\% quantile $0.8$ (right panel), categorizing split 
+residuals/scores decreases the performance as it leads to higher $p$-values for
+the true split variable $Z_1$ and thus to lower power in the settings considered. 
+For a true split at the median ($\xi=0$, left panel) this effect is almost constant
+across the three types of categorization considered. However, in case 
+of the true split point at the $90$\% quantile $\xi = 0.8$ (right panel), categorizing split 
 variables increases the $p$-value of $Z_1$ even more. 
-Here, the maximum selection performs best while the linear selection shows its disadvantage in 
-detecting abrupt changes, but overall it depends on the situation whether a linear or a maximum
-selection of a split variable is advantageous. 
-Comparing the two panels suggests that unless one of the breaks used for binning 
-possible split variables is close to the true split point (as for the median $0$, left panel), 
-a categorization weakens the performance of a testing strategy. 
-Moreover, while the effect of both of these transformations of the data (dichotomization of 
-residuals/scores and categorization of split variables) can be observed
-separately in Figure~\ref{fig:3way}, combining them increases the negative impact 
-remarkably.
+Not surprisingly, maximum selection is most advantageous in this case (i.e., for a late abrupt shift)
+which is harder to detect based on a linear statistic. But overall it depends on the situation
+whether a linear or a maximum selection of a split variable leads to lower $p$-values.
+
+Comparing the two panels suggests that a categorization weakens the performance of the
+tests unless the true split point is close to one of the breaks from the binning
+(as in the left panel).
+Moreover, while the effect of both transformations (dichotomization of residuals/scores and
+categorization of split variables) can be observed separately in Figure~\ref{fig:3way},
+combining them increases the negative impact remarkably.
 
@@
 
-Note that the choice of the two specific settings 
-(left panel: true split point at the median $0$,  effect size $0.3$; 
-right panel: true split point at the $90$\% quantile $0.8$ requiring a higher effect 
-size $1$)
-has been made based on the results shown in Figure~\ref{fig:scores_stump} in order to make 
-the differences particularly well observable. (Additional evaluations for increasing $\delta$ 
-as in Section~\ref{sec:scores} can be found in Appendix~\ref{app:increasingdelta}.)
+Note that the effect size $\delta$ in the results from Figure~\ref{fig:scores_stump} has been chosen
+so that $p$-values for the non-dichotomized tests are roughly comparable: $\delta = 0.3$
+for the true split point at the median $\xi = 0$ (left) vs.\ a stronger effect of $\delta = 1$
+for the true split point at the 90\% quantile $\xi = 0.8$ (right).
+Additional evaluations for varying effect size $\delta$ can be found in Appendix~\ref{app:increasingdelta}.
 
@@
 
-\begin{figure}%[t]
+\begin{figure}[t!]
 \setkeys{Gin}{width=1\linewidth}
@@
 
-%\newpage
 \subsection[``Tree'' scenario: Pre-pruning vs. post-pruning]{``Tree'' scenario: Pre-pruning vs.\ post-pruning}
 \label{sec:tree}
-To compare the performance of testing strategies in a more complex scenario they have
-been evaluated on data sets based on the tree structure with two different split variables 
-($Z_1$ and $Z_2$) both with the same split point $\xi$ as illustrated in 
-Figure~\ref{fig:dgp_tree}.
-The results presented in the top row of Figure~\ref{fig:tree_all} have been obtained applying pre-pruning.
-Therefore, based on the corresponding ARI values conclusion can again be drawn on the power of the testing strategies 
-as significance tests. CTree and MOB perform almost equally well for the true split points $0$, $0.2$ and $0.5$ 
-($50$\%, $60$\%, and $75$\% quantile). GUIDE+scores performs slightly worse but it still achieves an ARI above 
-$0.5$ for the highest considered effect size of $\delta = 1$. 
-However, the original version of GUIDE is clearly behind which is mainly due to the fact that it doesn't detect 
-the first split in $Z_2$ corresponding to a changing slope parameter as it only considers residuals in its testing 
-strategy. 
-For a split point farther away from the center such as the $90$\% quantile $0.8$ MOB 
-outperforms all other methods and both GUIDE versions fail to find the correct tree structure.
-Hence, the insights provided by the top row of Figure~\ref{fig:tree_all} conform with the conclusions
-drawn from the ``stump'' scenario: Including all available scores in the testing strategy leads to a better performance 
-while dichotomizing scores and categorizing split variables both do not cause any improvement but in many situations even 
-have a negative effect on the power of a testing strategy.
+
+So far the testing strategies underlying the different tree algorithms have only been considered
+as classical significance tests, i.e., in terms of power and $p$-values. However, one could argue
+that for a tree this is practically not really relevant -- at least when combined with a post-pruning
+strategy such as cost-complexity pruning \citep{Breiman+Friedman+Stone:1984}. In the latter case it only matters that the relevant split
+variables have the lowest $p$-value among all potential split variables -- but it is irrelevant whether
+this is significant or not. To investigate to which extent this is actually true we evaluate the
+different tree algorithms in the more complex ``tree'' scenario (see Figure~\ref{fig:dgp_tree}):
+once with significance-based pre-pruning and once with cost-complexity post-pruning.
+
+Recall that the true split structure is composed of splits in two different variables, both at
+the same split point $\xi$. First, the split in $Z_2$ changes the slope from $+\delta$ to
+$-\delta$. Second, the split in $Z_1$ changes the intercept in the negative slope group from
+$-\delta$ to $+\delta$. In the simulation the effect size $\delta$ is increased from 0 to 1
+for different split points $\xi$ from 0 to 0.8. Here, the performance is not evaluated in terms
+of test properties but only in terms of tree properties, namely the adjusted Rand index (ARI)
+in comparison to the true partition of the data.
+
+The top row of Figure~\ref{fig:tree_all} presents the results for pre-pruning which, not surprisingly,
+reflect the results from previous sections. Thus, CTree and MOB perform similarly and rather well
+as the underlying significance tests have good power properties. Only for late abrupt shifts the
+linear test statistics in CTree clearly have lower power than the maximally-selected statistics in MOB.
+Again, this picture would reverse for smooth changes instead of abrupt shifts. Compared to CTree/MOB
+both GUIDE flavors clearly perform worse as they underlying tests are less powerful; notably
+for the residual-based GUIDE which again has problems picking up the slope change associated with $Z_2$.
 
@@
 
-\begin{figure}%[t]
+\begin{figure}[t!]
 \setkeys{Gin}{width=\linewidth}
@@
 
-To elaborate the impact of the choice of a pruning strategy the presented tree models
-are built again in the exact same simulation setting but employing post-pruning (i.e., building very 
-large trees without any significance level and applying cost-complexity pruning as in the CART 
-algorithm \citep{Breiman+Friedman+Stone:1984}). 
-The corresponding ARI values are illustrated in the bottom row of Figure~\ref{fig:tree_all}
-allowing for a comparison to those obtained with pre-pruning (top row).
-
-For MOB and CTree it has been shown that their tests work very well as significance tests.
-Therefore, it can be recommended to take advantage of this power and use the information
-provided by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:tree_all}
-in most cases the results of these two testing strategies do not vary significantly over 
-a change of the applied pruning strategy.
-On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
-strategy should not be based on their results which argues for the application of post-pruning.
-This is also supported by the results in Figure~\ref{fig:tree_all} showing that post-pruning improves
-the performance of both GUIDE versions. Particularly if the true split point is close to a binning 
-point (such that categorizing split variables has no negative effect) the post-pruning strategy 
-compensates the lack of power of GUIDE+scores compared to CTree and MOB almost completely 
-(bottom row, first and third panel).
-In conclusion, the choice of a pruning strategy should be based on knowledge about the 
-power of the applied testing strategy. With a well working significance test additional expensive
-calculations such as needed for post-pruning may be avoided being one of the main advantages of 
-pre-pruning. However, deficiencies of a testing strategy in terms of power can be at least partly
-evened out by employing a post-pruning strategy.
-
-%\newpage
+When switching from pre-pruning to post-pruning (bottom row of Figure~\ref{fig:tree_all})
+it is indeed shown that many of the problems stemming from the low power of the two GUIDE
+flavors are indeed mitigated. Thus, by first growing a large tree without stopping upon
+non-significance and then pruning back based on predictive performance instead substantially
+improves the fit of the two GUIDE flavors. However, using the residuals only in GUIDE still
+performs clearly worse compared to the GUIDE+scores. The latter is essentially on par with
+CTree and MOB when the true split matches one of the categorization bins (i.e., $xi = 0$
+and $\xi = 0.5$, respectively) while CTree/MOB still perform somewhat better in the other
+cases ($\xi = 0.2$ and $\xi = 0.8$).
+
+In summary, there is clear support for the conventional wisdom that the power of the testing
+strategy in an unbiased tree algorithm is not so important when combined with post-pruning.
+However, there are limits to this. Consequently, when tests can be made more powerful --
+e.g., by using scores instead of residuals -- then this improves the corresponding tree
+algorithm. Finally, the simulation also supports using pre-pruning in an unbiased tree algorithm
+when the underlying testing strategy that also works well as a classical significance test.
 

### 2019-06-05 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised discussion in Section 6
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2609 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2104dba9c547e39a2f7bcbc087e3119bb8390278

@@
 \label{sec:discussion}
-From comparing the effects of the building blocks three main conclusions can be drawn: 
-(1)~In many cases including all available score values rather than just residuals leads to 
-clearly better results and performs in general at least equally well.
-(2)~Dichotomizing scores/residuals does not lead to any improvement in performance.
-(3)~Categorizing possible split variables also doesn't improve the performance 
-unless the breaks used for the categorization are very close to the true split points.
-However, even then the results are very similar to those achieved without any categorization.
-But it has to be noted that a categorization of possible split variables can simplify
-evaluations. Moreover, depending on the focus of interest, the results can be more
-straightforward, for example in case of clearly separated subgroups being defined by the values 
-of a continuous split variable. Therefore, categorizing possible split variables can have
-advantages in practice, however, it has to be considered that in general
-important effects might be neglected.
-
-Both of the transformation functions $h$ and $g$ (applied to the response~$Y$ or a split 
-variable~$Z$ respectively) can also be multivariate. For this case, the corresponding
-statistical tests are available, in particular in the \textsf{R} package \textbf{coin}
-which is employed for the implementation of MOB and CTree and provides 
-the required toolbox for multivariate dependency tests, including a $\chi^2$~test.
-
-Linear models as presented in this study have been selected as they allow for a simple illustration 
-of the investigated subjects. However, the obtained results also hold for a variety of other 
-models as the introduced building blocks can be applied in the same way for all model types.
-Specific characteristics of linear models regarding structural changes have been investigated
-in detail in previous literature.
-\cite{Ploberger+Kraemer:1992} have shown that residual-based tests can detect a change in the parameters
-only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
-This is not the case if changes are orthogonal to the mean regressor $(1,0)^{\top}$, hence, if
-the parameter vector $\beta$ is shifted 
-%to $\beta + \Delta$ 
-by $\Delta = (0, \Delta_1)^{\top}$.
-For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
-$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
-While this might be an extreme example similar scenarios can appear in many applications, see, e.g., 
-Figure~2 in \cite{Loh+He+Man:2015} where an additional separation of the data in each node is
-required in order to fit a reasonable model. This can be avoided by including model scores
-allowing for the detection of changes not only in the location parameter but also in all other 
-parameters, i.e., not only in the expected value but also in the variance.
-In situations where shifts appear only in the intercept testing procedures purely based on residuals 
-perform best. However, the cost in terms of power that comes with including all scores, in 
-particular with testing more than one parameter, is negligibly small.
-
-Also in more general models residual-based and score-based procedures are expected to perform
-equally well if all parameters are highly correlated. On the other hand, if parameters are 
-orthogonal changes that are not connected to the mean might be missed when only considering
-residuals. This can be prevented by including scores which are a standard measurement of goodness 
-of fit that can be easily employed for all model types. Moreover, the score function~$s$ can 
-simply be seen as a transformation to a different space in order to allow for a well structured
-analysis of dependencies, regardless of the type of model and therefore applicable in a
-wide variety of situations.
-In order to benefit from the advantages of using scores 
-some already existing models have been adapted, e.g., in \cite{Hothorn+Zeileis:2017} 
-and \cite{Schlosser+Hothorn+Stauffer:2019}. Similarly, variations of the GUIDE algorithm
-using different types of residuals are provided, e.g., in \citealp{Loh+Zheng:2013}, 
-\citealp{Chaudhuri+Loh:2002} and \citealp{Loh+He+Man:2015}.
-But even though it also depends on the type of the applied residuals whether parameter changes 
-can be detected successfully, a one-dimensional measurement of goodness of fit such as residuals
-provides less power than full model scores. 
-Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
-determined by the way how residuals are connected to parameter shifts as well as the type of parameter
-shift and the type of model. Therefore, a general statement for all types of models and residuals can 
-hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
-the here presented simulation show that considering only residuals provides small gains in certain 
-situations but remarkable deficits in general. On the contrary, the benefits of including full model 
-scores clearly outweigh the price of testing all parameters.
-
-%Furthermore, with only two parameters determining a linear model the importance of including scores 
-%regarding each parameter could already be pointed out clearly while for an increasing number of 
-%parameters in a model this aspect becomes even more important.
-
-%\newpage
+
+The testing strategies underlying the unbiased recursive partitioning algorithms
+CTree, MOB, and GUIDE have been embedded in a common inference framework, highlighting
+what the tests have in common and what sets them apart. Concerning the effects 
+of the corresponding building blocks for the tests, three main conclusions can be drawn: 
+%
+\begin{itemize}
+  \item \emph{Goodness-of-fit measure:} Assessing all dimensions of a model via the full
+        scores is to be preferred over assessing only a subset with the residuals. The former
+	can substantially improve performance while leading only to minor deteriorations
+	when it is not necessary.
+  \item \emph{Dichotomization of residuals/scores:} No scenarios could be found where this
+        is beneficial and tests without dichotomization performed clearly better in several
+	scenarios.
+  \item \emph{Categorization of split variables:} The effects of categorization are not so
+        clear-cut and depend on the true data structure. If there are indeed abrupt shifts
+	close to the breaks from the categorization, it works well. However, for splits
+	close to the margins performance can deteriorate and a maximally-selected test is
+	preferable. Finally, a linear statistic performs better for smooth rather than
+	abrupt changes.
+\end{itemize}
+%
+Thus, also when categorization as in GUIDE is used we would recommend to employ the non-dichotomized
+scores instead of the dichotomized residuals. Note that such a test corresponds to a
+multiple ANOVAs (for the score components) and is easily available in statistical software.
+Moreover, in the \proglang{R} system the \pkg{coin} package provides a convenient toolbox that
+encompasses multivariate transformation functions $h$ and/or $g$.
+
+Linear models as presented in this study have been chosen as they are highly relevant in practice,
+allow for simple illustrations, and theoretical insights are available for the testing strategies.
+However, the results can be easily extended to a wide variety of other models where the introduced building
+blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
+some changes in the parameters do not lead to shifts in the residuals (but in other components of the
+scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
+of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
+This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
+Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
+is of type $(0, \delta)^\top$ and thus orhtogonal to the mean regressor. Due to this residual-based
+tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
+have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
+investigations) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.
+
+Also in more general models, residual-based and score-based procedures are expected to perform
+equally well if all model parameters are highly correlated. But if parameters do change orthogonally
+this might again be missed when only considering residuals -- and full model scores are typically
+easily available as the appropriate remedy. Note that the score function~$s$ can also
+simply be seen as a transformation of the response variable $Y$ (and potentially regressors $X$)
+to a different space in order to allow for a well structured analysis of dependencies. This has
+been exploited in several tree-based approaches previously published in the literature,
+e.g., in \cite{Hothorn+Zeileis:2017} and \cite{Schlosser+Hothorn+Stauffer:2019}. Similarly, it would
+be of interest to investigate a score-based version of the extended GUIDE algorithms beyond the
+linear model, e.g., in \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002}, and \citealp{Loh+He+Man:2015}.
 
@@
 \label{sec:comp}
-The applied implementation is based on the \textsf{R} package \textbf{partykit}
+The applied implementation is based on the \proglang{R} package \pkg{partykit}
 (version~\Sexpr{packageVersion("partykit")}) which is available 
-on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
+on \proglang{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
 where the code to reproduce the simulation study can be found as well.
@@
 
-\section{Results over increasing effect size}
+\section{Results for increasing effect size}
 \label{app:increasingdelta}

### 2019-06-06 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small corrections; added information on where to find code to reproduce simulation study (arxiv)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2613 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: abfe84cfdb0c2cf1b1f6c774ee588cb4039acfc1

@@
 
-if(file.exists("tree.rda")){
-  load("tree.rda")
-} else {
-  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
-                         nrsplitvar = 2, nrsteps = 1, nrlevels = 2,
-                         delta = seq(from = 0, to = 1, by = 0.1),
-                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = "all",
-                         binary_regressor = FALSE, binary_beta = TRUE,
-                         only_intercept = FALSE,
-                         test = c("ctree", "mfluc", "ctree_max",
-                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
-                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
-                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
-                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
-                         beta0 = NULL, beta1 = NULL,
-                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05)
-
-  save(simres, file = "tree.rda")
-}
-sim_tree <- simres$res
-rm(simres)
+#if(file.exists("tree.rda")){
+#  load("tree.rda")
+#} else {
+#  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
+#                         nrsplitvar = 2, nrsteps = 1, nrlevels = 2,
+#                         delta = seq(from = 0, to = 1, by = 0.1),
+#                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = "all",
+#                         binary_regressor = FALSE, binary_beta = TRUE,
+#                         only_intercept = FALSE,
+#                         test = c("ctree", "mfluc", "ctree_max",
+#                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
+#                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
+#                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
+#                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
+#                         beta0 = NULL, beta1 = NULL,
+#                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05)
+#  
+#  save(simres, file = "tree.rda")
+#}
+#sim_tree <- simres$res
+#rm(simres)
 
@@
 Again, this picture would reverse for smooth changes instead of abrupt shifts. Compared to CTree/MOB
-both GUIDE flavors clearly perform worse as they underlying tests are less powerful; notably
+both GUIDE flavors clearly perform worse as the underlying tests are less powerful; notably
 for the residual-based GUIDE which again has problems picking up the slope change associated with $Z_2$.
@@
 performs clearly worse compared to the GUIDE+scores. The latter is essentially on par with
-CTree and MOB when the true split matches one of the categorization bins (i.e., $xi = 0$
+CTree and MOB when the true split matches one of the categorization bins (i.e., $\xi = 0$
 and $\xi = 0.5$, respectively) while CTree/MOB still perform somewhat better in the other
@@
 algorithm. Finally, the simulation also supports using pre-pruning in an unbiased tree algorithm
-when the underlying testing strategy that also works well as a classical significance test.
+when the underlying testing strategy also works well as a classical significance test.
 
@@
 blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
-some changes in the parameters do not lead to shifts in the residuals (but in other components of the
+changes in the parameters do not lead to shifts in the residuals (but in other components of the
 scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
@@
 Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
-is of type $(0, \delta)^\top$ and thus orhtogonal to the mean regressor. Due to this residual-based
+is of type $(0, \delta)^\top$ and thus orthogonal to the mean regressor. Due to this residual-based
 tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
@@
 (version~\Sexpr{packageVersion("partykit")}) which is available 
-on \proglang{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
-where the code to reproduce the simulation study can be found as well.
+on \proglang{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.
+The code to reproduce the simulation study is available in the supplement for this 
+paper on arXiv.org E-Print Archive (\url{https://arxiv.org/}).
 

### 2019-06-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
adapted draft.Rnw to restructured replication material and deleted old results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2617 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3a7ac63ca25bf4009378d3b0946c1aa9f7481a68

@@
 
-source("../guidelike.R")
-source("../tests_guideprune.R")
-source("../ccprune.R")
-
-if(file.exists("stump.rda")){
-  load("stump.rda")
+source("../simulation/guidelike.R")
+source("../simulation/tests_guideprune.R")
+source("../simulation/ccprune.R")
+
+if(file.exists("../simulation/results/sim_stump.rda")){
+  load("../simulation/results/sim_stump.rda")
 } else {
-  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
-                         nrsplitvar = 1, 
-                         nrsteps = 1, nrlevels = 2,
+  sim_stump <- simwrapper(scenario = "stump",
+                          nobs = 250, nrep = 100, seed = 7,
                           delta = seq(from = 0, to = 1, by = 0.1),
                           xi = c(0, 0.2, 0.5, 0.8), 
-                         vary_beta = c("all", "beta0", "beta1"),
-                         binary_regressor = FALSE,
-                         binary_beta = c(TRUE, FALSE),
-                         only_intercept = FALSE,
+                          variation = c("all", "beta0", "beta1"),
+                          changetype = c("abrupt", "continuous"),
                           test = c("ctree", "mfluc", "ctree_max",
@@
                                    "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
-                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
-                       beta0 = 0, beta1 = 1,
-                       stump = TRUE, z1dist = "unif", sigma = 1, alpha = 0.05)
-
-  save(simres, file = "stump.rda")
+                                   "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"))
+
+  sim_stump <- sim_stump$res
+
+  save(sim_stump, file = "../simulation/results/sim_stump")
 }
-sim_stump <- simres$res
-rm(simres)
-
-#if(file.exists("tree.rda")){
-#  load("tree.rda")
-#} else {
-#  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
-#                         nrsplitvar = 2, nrsteps = 1, nrlevels = 2,
-#                         delta = seq(from = 0, to = 1, by = 0.1),
-#                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = "all",
-#                         binary_regressor = FALSE, binary_beta = TRUE,
-#                         only_intercept = FALSE,
-#                         test = c("ctree", "mfluc", "ctree_max",
-#                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
-#                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
-#                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
-#                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
-#                         beta0 = NULL, beta1 = NULL,
-#                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05)
-#  
-#  save(simres, file = "tree.rda")
-#}
-#sim_tree <- simres$res
-#rm(simres)
 
@@
 
-d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
-               vary_beta = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)
+d <- dgp_stump(nobs = nobs, changetype= "abrupt",
+               variation = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)
 
@@
 <<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
-d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
-               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)
+d <- dgp_stump(nobs = nobs, changetype= "abrupt",
+               variation = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)
 
@@
 <<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
-d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
-               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
+d <- dgp_stump(nobs = nobs, changetype= "abrupt", 
+               variation = "all", delta = delta, xi = xi, sigma = sigma)
 par(mar = c(4,4,2,4))
@@
 set.seed(7)
-d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
-              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.5)
+d <- dgp_tree(nobs = 250, xi = 0, variation = "all", 
+              changetype= "abrupt", sigma = 0.5)
 
@@
 library("latticeExtra")
-useOuterStrips(xyplot(prop_Tsplit ~ delta | vary_beta + xi, groups = ~ test, 
+useOuterStrips(xyplot(prop_Tsplit ~ delta | variation + xi, groups = ~ test, 
                       data = subdataxi, 
-                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
-                        only_intercept == FALSE,# & xi == 0, 
+                      subset = changetype == "abrupt", 
                       type = "l", lty = c(1,1,1,1),
@@
 <<3way_data, results=hide, echo=FALSE>>=
-if(file.exists("3way_xi00_delta03.rda")){
-  load("3way_xi00_delta03.rda")
+if(file.exists("../simulation/results/sim_3way.rda")){
+  load("../simulation/results/sim_3way.rda")
 } else {
-  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
+  # for delta = 0.3 and xi = 0
+  sim3way1 <- sim(scenario = "stump",
+                  nobs = 250, nrep = 100, seed = 7,
                   delta = 0.3,
                   xi = 0, 
-                binary_regressor = FALSE,
-                binary_beta = TRUE,
-                only_intercept = FALSE,
-                nrlevels = 2, nrsplitvar = 1,
-                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
+                  changetype = "abrupt",
+                  variation = "all", 
                   return_matrices = TRUE,
@@
                            "ctree"))
-  d00 <- prep_3way(simres)
-  save(d00, file = "3way_xi00_delta03.rda")
-}
-
-if(file.exists("3way_xi08_delta1.rda")){
-  load("3way_xi08_delta1.rda")
-} else {
-  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
+  # prepare results for illustrations
+  sim_3way1 <- prep_3way(sim_3way1)
+
+  # for delta = 1 and xi = 0.8
+  sim_3way2 <- sim(scenario = "stump",
+                   nobs = 250, nrep = 100, seed = 7,
                    delta = 1,
                    xi = 0.8, 
-                binary_regressor = FALSE,
-                binary_beta = TRUE,
-                only_intercept = FALSE,
-                nrlevels = 2, nrsplitvar = 1,
-                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
+                   changetype = "abrupt",
+                   variation = "all",
                    return_matrices = TRUE,
@@
                             "ctree"))
-  d08 <- prep_3way(simres)
-  save(d08, file = "3way_xi08_delta1.rda")
+  # prepare results for illustrations
+  sim_3way2 <- prep_3way(sim_3way2)
+
+  sim_3way1$xi <- 0
+  sim_3way2$xi <- 0.8
+
+  # combine results for illustration in two panels as presented in Section 5.2. (Figure 4)
+  sim_3way <- rbind(sim_3way1, sim_3way2)
+  sim_3way$xi <- factor(sim_3way$xi, levels = c(0,0.8), labels = c(0,0.8))
+
+  save(sim_3way, file = "../simulation/results/sim_3way.rda")
 }
-
-d00$xi <- 0
-d08$xi <- 0.8
-
-d3way <- rbind(d00, d08)
-d3way$xi <- factor(d3way$xi, levels = c(0,0.8), labels = c(0,0.8))
 @
@@
        groups = interaction(bin, res_scores), 
-       data = d3way, 
+       data = sim_3way, 
        ylim = c(0,0.5),
@@
        ylab = expression("p-value of"~Z[1]),
-       #index.cond = list(c(1,2,3,4)),
        strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
@@
 
-#par(mar = c(3,5,1,5.7))
-#with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
-#                           ylab = expression("Mean"~p~"-value for"~Z[1])))
-#with(d00, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
-#with(d00, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
-#with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1)))
-#with(d08, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
-#with(d08, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
 @
@@
 <<tree_all_data, echo=FALSE, results=hide>>=
-if(file.exists("pruning.rda")){
-  load("pruning.rda")
+if(file.exists("../simulation/results/sim_tree.rda")){
+  load("../simulation/results/sim_tree.rda")
 } else {
-  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7, 
-                         nrsteps = 1, nrlevels = 2,
+  sim_tree <- simwrapper(scenario = "tree",
+                         nobs = 250, nrep = 100, seed = 7, 
                          delta = seq(from = 0, to = 1, by = 0.1),
                          xi = c(0, 0.2, 0.5, 0.8), 
-                         vary_beta = "all", binary_regressor = FALSE,
-                         binary_beta = TRUE, only_intercept = FALSE,
+                         variation = "all", 
+                         changetype = "abrupt",
                          compare_pruning = TRUE,
@@
                                   "guide_sum_12", 
-                                  "guide_sum_1_cor"),
-                         beta0 = NULL, beta1 = NULL,
-                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05,
-                         nrsplitvar = 2)
-
-  save(simres, file = "pruning.rda")  
-}
-
-pr <- simres$res
-## extract data for xyplot comparing ARI and ARI_p
-pr <- pr[,c(1,2,7,8,9,12,15)]
-## extended data set with post-pruning results being represented as separated tests
-prex <- pr[,c(1,2,3,4,6)]
-prex <- rbind(prex,prex)
-
-prex$test <- factor(prex$test,
+                                  "guide_sum_1_cor"))
+
+  # prepare data for illustration as presented in Section 5.3. (Figure 5)
+  sim_tree <- sim_tree$res
+  # extract data for xyplot comparing ARI and ARI_p
+  sim_tree <- sim_tree[,c(1,2,5,6,7,10,13)]
+  # extended data set with post-pruning results being represented as separated tests
+  sim_tree_ex <- sim_tree[,c(1,2,3,4,6)]
+  sim_tree_ex <- rbind(sim_tree_ex, sim_tree_ex)
+
+  sim_tree_ex$test <- factor(sim_tree_ex$test,
                              levels = c("ctree","mfluc",
@@
 
-prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
-prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
-prex$pruning <- c(rep("pre", NROW(pr)), rep("post", NROW(pr)))
-
-par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
-                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
-                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
+  sim_tree_ex$nrsubgr[(length(sim_tree$nrsubgr)+1) : (2*length(sim_tree$nrsubgr))] <- sim_tree$nrsubgr_p
+  sim_tree_ex$ari[(length(sim_tree$ari)+1) : (2*length(sim_tree$ari))] <- sim_tree$ari_p
+  sim_tree_ex$pruning <- c(rep("pre", NROW(sim_tree)), rep("post", NROW(sim_tree)))
+  sim_tree <- sim_tree_ex
+  rm(sim_tree_ex)
+  save(sim_tree, file = "../simulation/results/sim_tree.rda")  
+}
 
@@
 library("latticeExtra")
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
 useOuterStrips(xyplot(ari ~ delta | xi + pruning, groups = ~ test, 
-                      data = prex, 
+                      data = sim_tree, 
                       type = "l", lty = c(1,1,1,1),
                       lwd = 2,
-                      key =  list(text = list(levels(prex$test)[1:4]),
+                      key =  list(text = list(levels(sim_tree$test)[1:4]),
                                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
@@
                                        bg = "gray90"))
+
 @
@@
 
-library("latticeExtra")
-useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
+useOuterStrips(xyplot(prop_z1 ~ delta | variation + xi, groups = ~ test, 
                       data = subdataxi, 
-                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
-                        only_intercept == FALSE,# & xi == 0, 
+                      subset = changetype== "abrupt",
                       type = "l", lty = c(1,1,1,1),
@@
                                    bg = "gray90"),
-               strip.left=strip.custom(factor.levels = c(expression(xi == 0~" (50%)"),                                                              expression(xi == 0.8~" (90%)")),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0~" (50%)"),                                                              
+                                                         expression(xi == 0.8~" (90%)")),
                                        bg = "gray90"))
@@
 subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
-#subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
 subdata$test <- factor(subdata$test, 
@@
 
-xyplot(prop_Tsplit ~ delta | vary_beta, groups = ~ test, 
+xyplot(prop_Tsplit ~ delta | variation, groups = ~ test, 
        data = subdata, 
-       subset = binary_beta == FALSE & binary_regressor == FALSE & 
-         only_intercept == FALSE, 
+       subset = changetype == "continuous", 
        scales=list(x=list(at=seq(1,11,2))),
@@
        par.settings = par.settings)
+
 @
@@
 subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
-                                         #"ctree_max", "ctree_max_bin",
                                          "mfluc", "mfluc_bin",
@@
                        levels = c("ctree", "ctree_bin", 
-                                  #"ctree_max", "ctree_max_bin",
                                   "mfluc", "mfluc_bin",
@@
                        labels = c("CTree","CTree+dich",
-                                  #"CTree+max","CTree+max_bin",
                                   "MOB", "MOB+dich",
@@
 par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree", 
-                                                         #"CTree+max", "CTree+max", 
                                                          "MOB", "MOB",
@@
                                              fill = pal[c("CTree", "CTree", 
-                                                          #"CTree+max", "CTree+max", 
                                                           "MOB", "MOB",
@@
                      superpose.line = list(col = pal[c("CTree", "CTree", 
-                                                       #"CTree+max", "CTree+max", 
                                                        "MOB", "MOB",
@@
        data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & vary_beta == "all", 
+       subset = changetype== "abrupt" & variation == "all", 
        type = "l",
@@
                                 col = pal[c("CTree", "CTree", 
-                                            #"CTree+max", "CTree+max", 
                                             "MOB", "MOB",
                                             "GUIDE+scores")], lwd = 2)), 
-       #index.cond = list(c(1,2,3,4)),
        xlab = expression(delta),
@@
        par.settings = par.settings)
-
 @
@@
 subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
-                                         #"ctree_max", "ctree_max_cat",
                                          "mfluc", "mfluc_cat",
@@
                        levels = c("ctree", "ctree_cat", 
-                                  #"ctree_max","ctree_max_cat",
                                   "mfluc", "mfluc_cat",
@@
                        labels = c("CTree","CTree+cat",
-                                  #"CTree+max","CTree+max_cat",
                                   "MOB", "MOB+cat",
@@
        data = subdata, 
-       subset = binary_beta == TRUE & binary_regressor == FALSE & 
-         only_intercept == FALSE & vary_beta == "all", 
-       #xlim = c(0,0.1,0.5),
+       subset = changetype== "abrupt" & variation == "all", 
        type = "l", 
@@
 \end{document}
-
-%\begin{figure}%[t]
-%\setkeys{Gin}{width=\linewidth}
-%<<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
-%subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
-%                                        "guide_sum_12", "guide_sum_1_cor"))
-%subdata$test <- factor(subdata$test, 
-%                       levels = c("ctree", "mfluc",
-%                                  "guide_sum_12", "guide_sum_1_cor"),
-%                       labels = c("CTree","MOB",
-%                                  "GUIDE+scores", "GUIDE"))
-%
-%par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
-%                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
-%                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
-%
-%
-%xyplot(ari ~ delta | xi, groups = ~ test, 
-%       data = subdata, 
-%       type = "l",
-%       lty = c(1,1,1,1),
-%       lwd = 2,
-%      key =  list(text = list(levels(subdata$test)),
-%                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-%                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
-%       layout= c(4,1),
-%       #index.cond = list(c(1,2,3,4)),
-%       scales=list(x=list(at=seq(1,11,2))),
-%       xlab = expression(delta),
-%       ylab = "Adjusted Rand Index",
-%       strip = strip.custom(factor.levels = c(expression(xi==0), 
-%                                              expression(xi==0.2), 
-%                                              expression(xi==0.5), 
-%                                              expression(xi==0.8)),
-%                                   bg = "gray90"),
-%       par.settings = par.settings)
-%
-%@
-%\caption{\label{fig:tree_ari}
-%Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
-%``tree'' scenario, all applying pre-pruning. The ARI is evaluated over
-%increasing effect size $\delta$ with 100 replications of 250 observations per step.
-%Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
-%$0.8$ from left to right).}
-%\end{figure}
-
-%<<pruning_data, echo=FALSE, results=hide>>=
-%if(file.exists("pruning.rda")){
-%  load("pruning.rda")
-%} else {
-%  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7, 
-%                         nrsteps = 1, nrlevels = 2,
-%                         delta = seq(from = 0, to = 1, by = 0.1),
-%                         xi = c(0, 0.2, 0.5, 0.8), 
-%                         vary_beta = "all", binary_regressor = FALSE,
-%                         binary_beta = TRUE, only_intercept = FALSE,
-%                         compare_pruning = TRUE,
-%                         test = c("ctree", "mfluc",
-%                                  "guide_sum_12", 
-%                                  "guide_sum_1_cor"),
-%                         beta0 = NULL, beta1 = NULL,
-%                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05,
-%                         nrsplitvar = 2)
-%  
-%  save(simres, file = "pruning.rda")  
-%}
-%
-%pr <- simres$res
-%## extract data for xyplot comparing ARI and ARI_p
-%pr <- pr[,c(1,2,7,8,9,12,15)]
-%## extended data set with post-pruning results being represented as separated tests
-%prex <- pr[,c(1,2,3,4,6)]
-%prex <- rbind(prex,prex)
-%
-%prex$test <- factor(prex$test,
-%                    levels = c("ctree","mfluc",
-%                               "guide_sum_12", "guide_sum_1_cor",
-%                               "ctree_p","mfluc_p",
-%                               "guide_sum_12_p", "guide_sum_1_cor_p"),
-%                    labels = c("CTree","MOB",
-%                               "GUIDE+scores", "GUIDE",
-%                               "CTree_p","MOB_p",
-%                               "GUIDE+scores_p", "GUIDE_p"))
-%
-%for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
-%  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
-%  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
-%  if(prex$test[i] == "GUIDE+scores") prex$test[i] <- "GUIDE+scores_p"
-%  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
-%}
-%
-%prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
-%prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
-%
-%par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
-%                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
-%                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
-%
-%@
-
-%\begin{figure}%[t]
-%\setkeys{Gin}{width=\linewidth}
-%<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
-%xyplot(ari ~ delta | xi, groups = ~ test, 
-%       data = prex, 
-%       type = "l", lty = c(2,2,2,2,1,1,1,1),
-%       lwd = 2,
-%       key =  list(text = list(levels(prex$test)[1:4]),
-%                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-%                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
-%       layout= c(4,1),
-%       scales=list(x=list(at=seq(1,11,2))),
-%       xlab = expression(delta),
-%       ylab = "Adjusted Rand Index",
-%       strip = strip.custom(factor.levels = c(expression(xi==0), 
-%                                              expression(xi==0.2), 
-%                                              expression(xi==0.5), 
-%                                              expression(xi==0.8)),
-%                                   bg = "gray90"),
-%       par.settings = par.settings
-%)
-%@
-%\caption{\label{fig:pruning_ari}
-%Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
-%``tree'' scenario, once applying pre-pruning (dashed lines) and once post-pruning (solid lines). 
-%The ARI is evaluated over
-%increasing effect size $\delta$ with 100 replications of 250 observations per step.
-%Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
-%$0.8$ from left to right).}
-%\end{figure}

### 2019-06-12 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small revisions (formating of tables, shorter abstract, additional keyword 'simulation' and minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2621 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d5452c86478035ad9f6254fcb2a1090a193413b9

@@
 \Shorttitle{The Power of Unbiased Recursive Partitioning}
-%Comparing Testing Strategies of Unbiased Tree Algorithms
 
@@
 \Abstract{
-A core step of every algorithm for learning regression trees is the decision if, 
-how, and where to split the underlying data -- or, in other words, the selection 
+A core step of every algorithm for learning regression trees is the selection 
 of the best splitting variable from the available covariates and the corresponding 
-split point. Early tree algorithms (e.g., AID, CART) employed greedy search strategies 
-that directly compare all possible split points in all available covariates. However, 
-subsequent research showed that this is biased towards selection of covariates with 
+split point. Early tree algorithms (e.g., AID, CART) employed greedy search strategies, 
+directly comparing all possible split points in all available covariates. However, 
+subsequent research showed that this is biased towards selecting covariates with 
 more potential split points. Therefore, unbiased recursive partitioning algorithms 
 have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
-based on statistical inference using $p$-values that are appropriately adjusted for 
-the possible split points. In a second step a split point optimizing some objective 
-function is selected in the chosen split variable. However, different unbiased tree 
-algorithms employ different inference frameworks for computing these $p$-values and 
-their relative advantages or disadvantages are not well understood, yet.
+based on statistical inference using $p$-values that are adjusted for the possible split 
+points. In a second step a split point optimizing some objective function is selected in 
+the chosen split variable. However, different unbiased tree algorithms obtain these 
+$p$-values from different inference frameworks and their relative advantages or 
+disadvantages are not well understood, yet.
 
 Therefore, three different popular approaches are considered here: classical categorical 
-association tests (as in GUIDE), conditional inference (as in CTree), and parameter instability 
-tests (as in MOB). First, these are embedded into a common inference framework encompassing
-parametric model trees, in particular linear model trees. Second, it is assessed how
-different building blocks from this common framework affect the power of the 
-tree algorithms to select the appropriate covariates for splitting: observation-wise
+association tests (as in GUIDE), conditional inference (as in CTree), and parameter 
+instability tests (as in MOB). First, these are embedded into a common inference framework
+encompassing parametric model trees, in particular linear model trees. Second, it is 
+assessed how different building blocks from this common framework affect the power of the 
+algorithms to select the appropriate covariates for splitting: observation-wise
 goodness-of-fit measure (residuals vs.\ model scores), dichotomization of residuals/scores
-at zero, and binning of possible split variables. This shows that specifically the goodness-of-fit 
-measure is crucial for the power of the procedures, with model scores without dichotomization performing
-much better in many scenarios.
+at zero, and binning of possible split variables. This shows that specifically the 
+goodness-of-fit measure is crucial for the power of the procedures, with model scores 
+without dichotomization performing much better in many scenarios.
 }
-\Keywords{classification and regression trees, independence tests, recursive partitioning}
+\Keywords{classification and regression trees, independence tests, recursive partitioning,
+simulation}
 
@@
 
-% In the CTree algorithm the number of possible breaks is fixed as splits can only be made at
-% observed values or optionally at an average value lying between two observations. To select the
-% best split point two-sample linear statistics of the same structure as in the independence test 
-% are evaluated for all possible subsets of the sample space of the selected split variable.
-
 \subsection{MOB}\label{sec:mob}
@@
 \begin{tabular}{ l l l l }
-\hline 
+\hline\noalign{\smallskip} 
 & Fit             & Test                     & Pruning \\ 
-\hline
+\noalign{\smallskip}\hline\noalign{\smallskip}
 CTree & Non-parametric  & Conditional inference    & Pre\\
@@
 GUIDE & Parametric      & Residual-based $\chi^2$  & Post (cost-complexity pruning)\\
-\hline
+\noalign{\smallskip}\hline
 \end{tabular}
@@
 \begin{tabular}{ l l l l l }
-\hline 
-& Scores       & Dichotomization  & Categorization  & Statistic\\
-\hline
+\hline\noalign{\smallskip} 
+& Scores       & Dich.  & Cat.  & Statistic\\
+\noalign{\smallskip}\hline\noalign{\smallskip}
 CTree & Model scores & --	     & -- 	       & Sum of squares\\
@@
 GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
-\hline
+\noalign{\smallskip}\hline
 \end{tabular}
@@
 s(y_i, x_i, \beta_0, \beta_1) = 
-\frac{\partial \ell}{\partial (\beta_0, \beta_1)^{f\top}}(y_i, x_i, \beta_0, \beta_1) = 
+\frac{\partial \ell}{\partial (\beta_0, \beta_1)^{\top}}(y_i, x_i, \beta_0, \beta_1) = 
 -2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
@@
 \begin{tabular}{l l l}
-\hline
+\hline\noalign{\smallskip} 
 Name                    & Notation                   & Specification \\
-\hline
+\noalign{\smallskip}\hline\noalign{\smallskip} 
 \textit{Variables:}     &                        & \\ 
@@
 True split variable     & $Z_1$                  & $\mathcal{U}([-1,1])$\\
-Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
-\hline
-\textit{Parameters/functions:} &            & \\ 
+Noise split variables   & $Z_2$--$Z_{10}$   & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
+                        &                        & (alternating) \\
+\noalign{\smallskip}\hline\noalign{\smallskip} 
+\multicolumn{3}{l}{\textit{Parameters/functions:}} \\ 
 Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
@@
 Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
-\hline
+\noalign{\smallskip}\hline 
 \end{tabular}
@@
 
+\newpage
 \subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
@@
 
-\begin{figure}
+\begin{figure}[t!]
 \setkeys{Gin}{width=\linewidth}
@@
 
-\begin{table}[t!]
+\begin{table}
 \centering
 \begin{tabular}{ l l l }
-\hline
+\hline\noalign{\smallskip} 
 Building block               & Levels              & Transformation \\
-\hline
+\noalign{\smallskip}\hline\noalign{\smallskip} 
 Residuals vs.\ scores        & \textit{residuals}  & $h(Y) = r(Y,X,\hat{\beta})$\\
-\quad ($h$ transformation)   & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
-\hline
+($h$ transformation)         & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
+\noalign{\smallskip}\hline\noalign{\smallskip} 
 Dichotomization              & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
-\quad (of $h$)               & \textit{no}         & $h$ without dichotomization\\
-\hline
+(of $h$)                     & \textit{no}         & $h$ without\\
+                             &                     & \quad dichotomization\\
+\noalign{\smallskip}\hline\noalign{\smallskip} 
 Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)^{\top}$\\
-\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
+($g$ transformation)         &                     & \quad indicating the\\
+                             &                     & \quad assigned bin\\
 & \textit{max}               & $g(Z)=(0,\ldots,0,1,\ldots,1)^{\top}$\\
-&                     & \quad indicating the potential split point\\
+&                            & \quad indicating the\\
+&                            & \quad potential split point\\
 & \textit{lin}        & $g(Z)=Z$\\
-\hline
+\noalign{\smallskip}\hline
 \end{tabular}
@@
 \hline\noalign{\smallskip} 
-Residuals/Scores   & Dichotomization & Categorization & \multicolumn{2}{l}{Via testing strategy}\\
+Residuals/Scores   & Dich. & Cat. & \multicolumn{2}{l}{Via testing strategy}\\
 \noalign{\smallskip}\hline\noalign{\smallskip}
 residuals	   & yes	     & cat            & GUIDE & {(default)}\\
-residuals	   & yes	     & max            & MOB   & {(with modifications)}\\
-residuals	   & yes	     & lin            & CTree & {(with modifications)}\\
-residuals	   & no 	     & cat            & MOB   & {(with modifications)}\\
-residuals	   & no 	     & max            & MOB   & {(with modifications)}\\
-residuals	   & no 	     & lin            & CTree & {(with modifications)}\\
-scores  	   & yes	     & cat            & GUIDE & {(with modifications)}\\
-scores  	   & yes	     & max            & MOB   & {(with modifications)}\\
-scores  	   & yes	     & lin            & CTree & {(with modifications)}\\
-scores  	   & no 	     & cat            & MOB   & {(with modifications)}\\
+residuals	   & yes	     & max            & MOB   & {(modified)}\\
+residuals	   & yes	     & lin            & CTree & {(modified)}\\
+residuals	   & no 	     & cat            & MOB   & {(modified)}\\
+residuals	   & no 	     & max            & MOB   & {(modified)}\\
+residuals	   & no 	     & lin            & CTree & {(modified)}\\
+scores  	   & yes	     & cat            & GUIDE & {(modified)}\\
+scores  	   & yes	     & max            & MOB   & {(modified)}\\
+scores  	   & yes	     & lin            & CTree & {(modified)}\\
+scores  	   & no 	     & cat            & MOB   & {(modified)}\\
 scores  	   & no 	     & max            & MOB   & {(default)}\\

### 2019-06-24 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2623 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b7aadd24c53e287906b69410e3240668da6c0d44
small revision on layout of Figure 4
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2625 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 10f9d5783845ade5302ee1d16498e7561ee963c2

@@
 
-In Figure~\ref{fig:scores_stump} the performances of the testing strategies are 
-represented by the corresponding selection probability, i.e., a significance level 
-is incorporated. (Appendix~\ref{app:significance} shows that the same qualitative
-conclusions can be drawn when the significance level is \emph{not} included.)
-For a split point at the median ($\xi=0$, top row) all testing strategies 
-perform similarly well as long as the intercept varies (left and right panel) with CTree and 
-MOB being only slightly ahead. However, for the split point at the $90$\% quantile $\xi = 0.8$ 
-(bottom row) the performance of all of the applied strategies decreases. 
-While both GUIDE versions struggle to detect the correct split variable even for a high effect 
-size, MOB is clearly ahead with CTree leading to the second best results.
-This advantage of MOB over CTree is mainly due to the abrupt shift in the 
-model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
-a smooth transition with continuously-changing parameters (see Appendix~\ref{app:contbeta}).
-
-However, in the scenario where only $\beta_1$ (but not the intercept $\beta_0$) is affected by the split,
-the residual-based GUIDE approach has no power at all even for a true split at the median (top middle panel).
-It is easily possible, though, to substantially mitigate this problem by using scores
-(sensitive to changes in all parameters) rather than residuals only (sensitive to changes
-in the conditional mean). The remaining difference between MOB/CTree and GUIDE+scores
-is due to dichotomizing the scores at zero and due to categorizing the split variables
-which are investigated in more detail in the following section.
-
 \begin{figure}[t!]
@@
 \end{figure}
+
+In Figure~\ref{fig:scores_stump} the performances of the testing strategies are 
+represented by the corresponding selection probability, i.e., a significance level 
+is incorporated. (Appendix~\ref{app:significance} shows that the same qualitative
+conclusions can be drawn when the significance level is \emph{not} included.)
+For a split point at the median ($\xi=0$, top row) all testing strategies 
+perform similarly well as long as the intercept varies (left and right panel) with CTree and 
+MOB being only slightly ahead. However, for the split point at the $90$\% quantile $\xi = 0.8$ 
+(bottom row) the performance of all of the applied strategies decreases. 
+While both GUIDE versions struggle to detect the correct split variable even for a high effect 
+size, MOB is clearly ahead with CTree leading to the second best results.
+This advantage of MOB over CTree is mainly due to the abrupt shift in the 
+model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
+a smooth transition with continuously-changing parameters (see Appendix~\ref{app:contbeta}).
+
+However, in the scenario where only $\beta_1$ (but not the intercept $\beta_0$) is affected by the split,
+the residual-based GUIDE approach has no power at all even for a true split at the median (top middle panel).
+It is easily possible, though, to substantially mitigate this problem by using scores
+(sensitive to changes in all parameters) rather than residuals only (sensitive to changes
+in the conditional mean). The remaining difference between MOB/CTree and GUIDE+scores
+is due to dichotomizing the scores at zero and due to categorizing the split variables
+which are investigated in more detail in the following section.
 
@@
        lwd = 2,
+       layout = c(2,1),
        scales = list(alternating=FALSE),
@@
 in comparison to the true partition of the data.
-
-The top row of Figure~\ref{fig:tree_all} presents the results for pre-pruning which, not surprisingly,
-reflect the results from previous sections. Thus, CTree and MOB perform similarly and rather well
-as the underlying significance tests have good power properties. Only for late abrupt shifts the
-linear test statistics in CTree clearly have lower power than the maximally-selected statistics in MOB.
-Again, this picture would reverse for smooth changes instead of abrupt shifts. Compared to CTree/MOB
-both GUIDE flavors clearly perform worse as the underlying tests are less powerful; notably
-for the residual-based GUIDE which again has problems picking up the slope change associated with $Z_2$.
 
@@
 \end{figure}
+
+The top row of Figure~\ref{fig:tree_all} presents the results for pre-pruning which, not surprisingly,
+reflect the results from previous sections. Thus, CTree and MOB perform similarly and rather well
+as the underlying significance tests have good power properties. Only for late abrupt shifts the
+linear test statistics in CTree clearly have lower power than the maximally-selected statistics in MOB.
+Again, this picture would reverse for smooth changes instead of abrupt shifts. Compared to CTree/MOB
+both GUIDE flavors clearly perform worse as the underlying tests are less powerful; notably
+for the residual-based GUIDE which again has problems picking up the slope change associated with $Z_2$.
 

### 2019-06-28 pkg/disttree/inst/IWSM2019/slides.Rnw
first draft of slides for IWSM 2019
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2626 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 227c0bb756d4097710ef42ed50c0b6a74658c2bd

--- 2018-07-15 pkg/disttree/inst/JSM2018/slides.Rnw
@@
 
-\title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
+\title{Distributional Trees 
+%and Forests \\
+for Circular Data}
+\author{Lisa Schlosser, Moritz N. Lang, Torsten Hothorn, \\
+Georg~J.~Mayr, Reto Stauffer, Achim Zeileis}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
 %\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
-\URL{https://R-Forge.R-project.org/projects/partykit/}
+\URL{https://R-Forge.R-project.org/projects/partykit/ \\
+https://partykit.org}
 \headerimage{3}
@@
 %% includes a replacement for \usepackage{Sweave}
+\usepackage{Sweave}
 \usepackage{amsmath,tikz}
@@
 
-# setwd("~/svn/partykit/pkg/disttree/inst/JSM2018")
-library("ggplot2")
-theme_set(theme_bw(base_size = 18))
+# setwd("~/svn/partykit/pkg/disttree/inst/IWSM2019")
 library("disttree")
 library("Formula")
-#library("partykitR1")
-library("gamlss")
-library("lattice")
-library("crch")
 library("latex2exp")
-library("parallel")
-library("gamlss.cens")
-gen.cens(NO, type = "left")
 
@@
 
-######## data generating process based on a predefined tree or a parameter function
-# if tree: 2 splits => 3 terminal nodes
-# 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
-#
-# input: n            ..... nr of observations
-#        family       ..... distribution of the generated observations
-#        split.matrix ..... indices of the variables used for the splits together with the corresponding split points
-#        parm.matrix  ..... set of distribution parameters for each subgroup
-#        fun          ..... parameter function
-#
-# output: data.frame with generated observations y (generated seperatly in each subgroups with given distributions parameters),
-#                         the given split variables x1, ..., x10 for each observation and
-#                         the index of the subgroup for each observation or
-#                         the distribution parameter for each observation
-
-# FIX: until now only complete lists can be handed over (dist_list...)
-dgp <- function(n, family = dist_list_normal, 
-                fun = NULL,
-                split.matrix = matrix(nrow = 2, ncol = 2), 
-                par.matrix = matrix(nrow = 3, ncol = 2), 
-                round.sp = 3)
-{
-
-  # generating the possible split variables
-  x1 <- runif(n,-0.4,1)
-  x2 <- runif(n,-10,10)
-  x3 <- runif(n,0,100)
-  x4 <- x1 + rnorm(n, sd = 0.1)
-  x5 <- x1 + rnorm(n, sd = 0.3)
-  x6 <- rbinom(n,1,0.5)
-  x7 <- rbinom(n,1,0.5)
-  x8 <- sample(1:4, n, replace = TRUE) 
-  x9 <- sample(1:5, n, replace = TRUE) 
-  x10 <- sample(1:7, n, replace = TRUE)
-  #x11 <- runif(n,-0.5,1)
-  x <- cbind(x1,x2,x3,x4,x5,x6,x7,x8,x9,x10)
-  # reduce nr of possible split points by rounding values of split variables
-  x <- round(x, digits = round.sp)
-
-  y <- vector(mode = "numeric", length = n)
-
-  # getting the random function
-  if(is.function(family)) family <- family()
-  if(inherits(family, "gamlss.family")) {
-    rfun <- get(paste0("r",family$family[[1]]))
-  } else {
-    if(family$family.name == "Normal Distribution") rfun <- rNO
-    if(family$family.name == "censored Normal Distribution") rfun <- rNO
-    if(family$family.name == "Poisson Distribution") rfun <- rPO
-    if(family$family.name == "Weibull Distribution") rfun <- survival:::rsurvreg 
-  }
-
-  if(is.null(fun)) {
-    index <- vector(mode = "numeric", length = n)
-    sv = split.matrix[,1]    # index of split variables
-    sp = split.matrix[,2]    # split points
-    par = par.matrix
-
-    # splitting based on the given split variables and split points and
-    # generating the observations seperatly for each subgroup (with the corresponding given distribution parameters)
-    for(k in 1:n){
-      if(x[k, sv[1]] <= sp[1]){
-        y[k] <- do.call(rfun, as.list(c(1,par[1,])))
-        index[k] <- 2L
-      } else {
-        if(x[k, sv[2]] <= sp[2]){
-          y[k] <- do.call(rfun, as.list(c(1,par[2,]))) 
-          index[k] <- 4L
-        } else {
-          y[k] <- do.call(rfun, as.list(c(1,par[3,])))
-          index[k] <- 5L
-        }
-      }
-    }
-    d <- as.data.frame(cbind(y, x, index))
-    colnames(d) <- c("y", paste0("x", c(1:10)), "index")
-
-  } else {
-
-    dpar <- fun(x)
-
-    if(!(family$family.name == "Poisson Distribution")){
-      y <- rfun(n, dpar[,1], dpar[,2])
-      d <- as.data.frame(cbind(y, x, dpar))
-      colnames(d) <- c("y", paste0("x", c(1:10)), "mu", "sigma")
-    } else {
-      y <- rfun(n, dpar)
-      d <- as.data.frame(cbind(y, x, dpar))
-      colnames(d) <- c("y", paste0("x", c(1:10)), "lambda")
-    }
-  }
-
-  if(family$family.name == "censored Normal Distribution") {
-    #d$ystar <- d$y
-    d$y <- pmax(d$y, 0)
-  }
-
-  return(d)
-}
-
-############################################ 
-# parameter functions
-
-### plotting function for one data set
-plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE, all.var = FALSE) 
-{
-  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
-                    df$fitted.par,
-                    dt$fitted.par,
-                    #coef(dt)[paste(dt$fitted[,1]),],
-                    g$mu.fv, g$sigma.fv)
-
-  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
-                          "fitted.mu.df","fitted.sigma.df",
-                          "fitted.mu.dt","fitted.sigma.dt",
-                          "fitted.mu.g","fitted.sigma.g")
-  sp <- plotdata[order(plotdata["x1"]),]
-
-  if(!(any(dt.par, df.par, g.par, all.par, all.var))){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-  col = transpgrey, border = "transparent")
-    #lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
-    #       col = c('black', transpgrey), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # disttree
-  if(dt.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", col.main = pal["tree"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
-  col = pallight["tree"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-    #       col = c('black','grey', pal["tree"], pallight["tree"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # distforest
-  if(df.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", col.main = pal["forest"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
-  col = pallight["forest"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
-    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-    #       col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # gamlss
-  if(g.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
-  col = pallight["gamlss"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
-    #       col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # disttree vs. distforest vs. gamlss
-  if(all.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
-    #                    TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
-    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 13.95, expression(mu), bty = "n")
-  }
-
-  # variance: disttree vs. distforest vs. gamlss
-  if(all.var){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'black')
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
-  col = pallight["tree"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
-  col = pallight["forest"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
-  col = pallight["gamlss"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
-    #                    TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
-    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-  legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-}
-
 @
@@
 \begin{document}
-
-\section{Distributional Trees and Forests}
+\SweaveOpts{concordance=TRUE}
+
+\section{Distributional Trees and Forests for Circular Data}
 
@@
 \begin{frame}[fragile]
-\frametitle{Motivation}
-\vspace{-0.41cm}
-\begin{figure}[!htb]
-\minipage{0.29\textwidth}
-\begin{center}
-<<motivation_GLM, echo=FALSE, results=hide>>=
-nobs <- 200
-## GLM
-set.seed(7)
-x <- c(1:nobs)/nobs
-ytrue <- 1+x*1.5
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-\visible<2->{
-<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
-@
-}
-\end{center}
-\endminipage
-\visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
-\begin{center}
-<<motivation_GAM, echo=FALSE, results=hide>>=
-## GAM
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3 
-y <- ytrue + rnorm(nobs,0,0.3)
-@
-\visible<3->{
-<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
-@
-}
-\end{center}
-\endminipage
-\visible<4->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
-\begin{center}
-<<motivation_GAMLSS, echo=FALSE, results=hide>>=
-## GAMLSS
-set.seed(7)
-x <- c(1:nobs)/nobs
-x <- 2*(x-0.5)
-ytrue <- x^3
-var <- exp(-(2*x)^2)/2
-y <- ytrue + rnorm(nobs, 0, 0.1 + var)
-@
-\visible<4->{
-<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
-polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
-  col = pallight["forest"], border = "transparent")
-lines(x, ytrue, col = pal["forest"], lwd=7)
-points(x, y, col = "slategray", pch = 19)
-box(lwd = 5)
-@
-
-}
-\end{center}
-\endminipage
-
-\vspace{0.5cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\visible<2->{
-LM, GLM\\
-\vspace{0.5cm}
-\code{lm}\\
-\code{glm}\\
-\vspace{1.5cm}}
-\end{center}
-\endminipage
-\hspace{1.1cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\visible<3->{
-GAM\\
-\vspace{0.5cm}
-\code{mgcv}\\
-\code{VGAM}\\
-$\ldots$\\
-\vspace{1cm}}
-\end{center}
-\endminipage
-\hspace{1.1cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\visible<4->{
-GAMLSS\\
-\vspace{0.5cm}
-\code{gamlss}\\
-\code{mgcv}\\
-\code{VGAM}\\
-\code{gamboostLSS}\\
-$\ldots$}
-\end{center}
-\endminipage
-\end{figure}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Motivation}
-\vspace{-0.2cm}
-\begin{figure}[!htb]
-\minipage{0.29\textwidth}
-\begin{center}
-<<motivation_regtree, echo=FALSE, results=hide>>=
-## Reg. Tree
-set.seed(7)
-kappa <- 12
-x <- c(1:nobs)/nobs
-ytrue <- ytree <- yforest <- numeric(length = length(x))
-for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
-y <- ytrue + rnorm(nobs,0,0.3)
-for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
-@
-\visible<1->{
-<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = ytree, col = pal["forest"], lwd=7)
-@
-}
-\end{center}
-
-\endminipage
-\visible<2->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
- \begin{center}
-<<motivation_randforest, echo=FALSE, results=hide>>=
-## Random Forest
-for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
-@
-\visible<2->{
-<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
-par(mar=c(2,0,2,0))
-plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
-box(lwd=5)
-#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
-lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
-@
-}
-\end{center}
-
-\endminipage
-\visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
-  \visible<3->{
-  \begin{center}
-  \begin{tikzpicture}
-  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
-  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
-  \end{tikzpicture}
-  \end{center}}
-\endminipage
-
-\minipage{0.29\textwidth}
-\begin{center}
-\vspace{0.0cm}
-\visible<1->{
-Regression Tree\\
-\vspace{0.5cm}
-\resizebox{0.3\textwidth}{!}{
-\begin{tikzpicture}
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (1, 2) {};
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0.5, 1) {};
-  \draw[-, line width=1pt] (n0) -- (n1);
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (1.5, 1) {};
-  \draw[-, line width=1pt] (n0) -- (n2);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (1, 0) {};
-  \draw[-, line width=1pt] (n2) -- (n3);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (2, 0) {};
-  \draw[-, line width=1pt] (n2) -- (n4);
-\end{tikzpicture}}
-\vspace{0.3cm}\\
-\code{rpart}\\
-\code{party(kit)}\\
-\vspace{1cm}}
-\end{center}
-\endminipage
-\hspace{0.65cm}
-\minipage{0.29\textwidth}
-\begin{center}
-\vspace{0.0cm}
-\visible<2->{
-Random Forest\\
-\vspace{0.5cm}
-\resizebox{0.9\textwidth}{!}{
-\begin{tikzpicture}
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n00) at (1, 2) {};
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n01) at (0.5, 1) {};
-  \draw[-, line width=1pt] (n00) -- (n01);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n02) at (1.5, 1) {};
-  \draw[-, line width=1pt] (n00) -- (n02);
-
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n10) at (3, 2) {};
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n11) at (2.5, 1) {};
-  \draw[-, line width=1pt] (n10) -- (n11);
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n12) at (3.5, 1) {};
-  \draw[-, line width=1pt] (n10) -- (n12);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n13) at (2, 0) {};
-  \draw[-, line width=1pt] (n11) -- (n13);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n14) at (2.8, 0) {};
-  \draw[-, line width=1pt] (n11) -- (n14);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n15) at (3.2, 0) {};
-  \draw[-, line width=1pt] (n12) -- (n15);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n16) at (4, 0) {};
-  \draw[-, line width=1pt] (n12) -- (n16);
-
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
-  \draw[-, line width=1pt] (n20) -- (n21);
-  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
-  \draw[-, line width=1pt] (n20) -- (n22);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
-  \draw[-, line width=1pt] (n22) -- (n23);
-  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
-  \draw[-, line width=1pt] (n22) -- (n24);
-\end{tikzpicture}
-}
-\vspace{0.5cm}\\
-\code{randomForest}\\
-\code{ranger}\\
-\code{party(kit)}\\
-$\ldots$}
-\end{center}
-\endminipage
-\hspace{0.65cm}
-\minipage{0.28\textwidth}
-\begin{center}
-\vspace{0.68cm}
-\visible<3->{
-Distributional trees and forests\\
-\vspace{1.7cm}
-\code{disttree}\\
-based on \code{partykit}\\
-\vspace{1cm}}
-\end{center}
-\endminipage
-\end{figure}
+\frametitle{Distributional Tree}
+
+\vspace{1cm}
+
+{\bf Starting point:}
+\begin{itemize}
+\item Response $Y$
+\item Covariate(s) $X$ $\rightarrow$ split variable(s) $X$
+\item Distribution family $\mathcal{D}(\theta)$
+\end{itemize}
+
+\vspace{1cm}
+
+{\bf Goal:} Probabilistic model
+
+%Idea: maximum-likelihood method
+%\begin{enumerate}
+%\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
+%\item Estimate $\hat{\theta}$ via maximum likelihood.
+%\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
+%\item Assess whether the \emph{model scores} are associated with
+%    (or change along) any of the available covariates -- e.g.,
+%    using parameter instability tests (\emph{strucchange}) or
+%    conditional inference (\emph{coin}).
+%\item Split the sample along the partitioning variable with the strongest association or instability.
+%    Choose breakpoint with highest improvement in log-likelihood.
+%\item Repeat steps 2--4 recursively in the subgroups
+%    until some stopping criterion is met.
+%    %-- e.g., for significance or sample size.
+%\item Choose the variable with the strongest association
+%\item Choose the split point which leads to the highest improvement of the model
+%\item Split and repeat 2-6 in each node until a stopping criterion is met
+%\end{enumerate}
+
 \end{frame}
@@
 \begin{frame}
-\frametitle{Goals}
-
-\textbf{Distributional:} \\
-\begin{itemize}
-\item Specify the complete probability distribution (including location, scale and shape).
-\end{itemize}
-
-\medskip
-
-{\bf Tree:}
-\begin{itemize}
-  \item Automatic variable selection and detection of steps and abrupt changes. %(data driven)
-  \item Capture non-linear and non-additive effects and interactions.
-\end{itemize}
-
-\medskip
-
-{\bf Forest:}
-\begin{itemize}
-  \item Smoother effects.
-  \item Stabilization and regularization of the model.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Distributional trees}
+\frametitle{Distributional Tree}
 \vspace*{-0.12cm}
@@
 \begin{frame}[fragile]
-\frametitle{Distributional trees}
+\frametitle{Distributional Tree}
 \begin{center}
@@
 \end{frame}
-%\begin{frame}
-%\frametitle{Distributional trees}
-%\center
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{tree_box}
-%\end{frame}
 
@@
 
-%\begin{frame}
-%\frametitle{Distributional trees}
-%\center
-%\vspace*{-0.7cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{tree_xyplot}
-%\end{frame}
-
-\subsection{Building Distributional Trees and Forests}
 \begin{frame}
 \frametitle{Building Distributional Trees and Forests}
+\begin{minipage}{0.7\textwidth}
 {\bf Tree:}
 \begin{enumerate}
-\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\hat{\theta}$ via maximum likelihood.
-\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable.
-%\item Assess whether the \emph{model scores} are associated with
-%    (or change along) any of the available covariates -- e.g.,
-%    using parameter instability tests (\emph{strucchange}) or
-%    conditional inference (\emph{coin}).
-\item Choose the partitioning variable with the strongest association or instability as split variable.
-\item Choose the breakpoint along this variable with highest improvement in log-likelihood and split the sample.
-\item Repeat steps 2--5 recursively in the subgroups
-    until some stopping criterion is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
-    %-- e.g., for significance or sample size.
-%\item Choose the variable with the strongest association
-%\item Choose the split point which leads to the highest improvement of the model
-%\item Split and repeat 2-6 in each node until a stopping criterion is met
+\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
+Estimate $\hat{\theta}$ via maximum likelihood\\
+$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
+\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
 \end{enumerate}
-
-\vspace{0.2cm}
+\end{minipage}
+\begin{minipage}{0.23\textwidth}
+%\vspace{-0.1cm}
+\begin{tikzpicture}
+\visible<2-3>{
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
+}
+\visible<4>{
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{D}(Y;\hat{\theta}$)};
+}
+\visible<5->{
+\node[inner sep=0pt] (density1) at (0.8, 1.7)
+    {\includegraphics[width=0.6\textwidth]{density1.jpeg}};
+}
+\visible<6-8>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
+\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1);
+\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2);
+}
+\visible<9-10>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
+\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<11>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
+}
+\visible<11->{
+\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<12->{
+\node[inner sep=0pt] (density2) at (-0.1,-0.1)
+    {\includegraphics[width=0.5\textwidth]{density2.jpeg}};
+}
+\visible<12->{
+\node[inner sep=0pt] (density3) at (1.7,-0.1)
+    {\includegraphics[width=0.5\textwidth]{density3.jpeg}};
+}
+\end{tikzpicture}
+\end{minipage}
+\vspace{0.1cm}
+%\begin{adjustwidth}{-0.0em}{-1em}
+\begin{enumerate}
+\setcounter{enumi}{2}
+\item<8-> Split along the covariate $X$ with strongest association or instability and at breakpoint $p$ with highest improvement \\
+in log-likelihood.
+\item<10-> Repeat steps 1--3 recursively until some stopping criterion \\
+is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
+\end{enumerate}
+%\end{adjustwidth}
+\vspace{0.4cm}
+\visible<13->{
 {\bf Forest:} Ensemble of $T$ trees.
@@
 \end{itemize}
+}
 \end{frame}
@@
 \begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.4cm}
-\textbf{Goal:} Predict daily precipitation amount in alpine regions.\\
-\medskip
-\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
-\medskip
-\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
-\vspace{1.3cm}
-%\bigskip
-
-%\textbf{Idea:} Statistical model with numerical predictions as covariates.\\
-%\bigskip
-%\textbf{Distribution assumption:}\\
-%\vspace{-0.25cm}
-%\hspace{1.5cm}
-%\center
-%$$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
-
-\end{frame}
-
-<<echo = FALSE, results=hide>>=
-library("zoo")
-
-Sys.setenv("TZ"="UTC")
-
-# -------------------------------------------------------------------
-# GFS ensemble forecast
-# -------------------------------------------------------------------
-
-GFS   <- read.table("ensemble_plot/GENS_00_innsbruck-flughafen.txt", header = TRUE)
-init  <- min(as.POSIXct(GFS$timestamp-GFS$step*3600,origin="1970-01-01"))
-title <- sprintf("%s\n%s",
-                 "Global Forecast System (GFS) Ensemble Forecast for Innsbruck, Airport",
-                 sprintf("Forecast initialized %s",strftime(init,"%Y-%m-%d %H:%M UTC")))
-
-# observations
-load("STAGEobs_tawes_11121.rda")
-# "2018-03-14 00:00:00" in line 321 (first entry for temperature)
-# "2018-03-13 06:00:00" in line 317 (first entry for rain -> aggregated to 24h sums)
-# "2018-03-23 00:00:00" in line 357
-obs <- obs[318:357,]
-
-getZoo <- function( x, variable ) {
-  x <- subset( x, varname == variable )
-  if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
-  # Else create zoo
-  x <- zoo( subset(x,select=-c(varname,timestamp,step)),
-            as.POSIXct(x$timestamp,origin="1970-01-01") )
-  x
-}
-GFS_t2m  <- getZoo( GFS, "tmp2m" )
-# temperature bias: model simulates temperature for height 1675 (~1070 m higher than true height)
-# => add 10.7 degrees to account for bias (approx. 1 degree / 100 m)
-GFS_t2m <- GFS_t2m + 10.7
-GFS_t2m$obs <- obs[,"obs$tl"]
-# first value of rain is 18 hours later than temperature 
-#(both start at 6:00 but rain is cumulated over 24 hours)
-# -> drop first 3 values of temperature
-GFS_t2m <- GFS_t2m[-c(1:3),]
-
-GFS_rain <- getZoo( GFS, "apcpsfc" )
-GFS_rain$obs <- obs[,"rr6"]
-# precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
-GFS_rain24 <- GFS_rain[c((1:10)*4),]
-for(i in 1:NROW(GFS_rain24)){
-  GFS_rain24[i,] <- colSums(GFS_rain[c((4*i-3):(4*i)),])
-}
-
-# Two POSIXct vectors for the axis
-main  <- seq(min(as.POSIXct(as.Date(index(GFS_t2m)))),max(index(GFS_t2m)),by=86400)
-minor <- seq(min(index(GFS_t2m)),max(index(GFS_t2m)),by=3*3600)
-
-# GFS_t2m_mean <- rowMeans(GFS_t2m)
-GFS_t2m_mean <- GFS_t2m
-GFS_t2m_mean[,1] <- rowMeans(GFS_t2m[,-NCOL(GFS_t2m)])
-GFS_t2m_mean <- GFS_t2m_mean[,1]
-
-# GFS_rain_mean <- rowMeans(GFS_rain)
-GFS_rain24_mean <- GFS_rain24
-GFS_rain24_mean[,1] <- rowMeans(GFS_rain24[,-NCOL(GFS_rain24)])
-GFS_rain24_mean <- GFS_rain24_mean[,1]
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_1_plot, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05, col = "slategray" )
-#plot( GFS_t2m_mean, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-#plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
-#plot( GFS_rain_mean, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_1_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_2_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,c(1,2)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_3_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_20_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,-NCOL(GFS_t2m)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,-NCOL(GFS_rain24)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.4cm}
-\textbf{Goal:} Predict daily precipitation amount in alpine regions.\\
-\medskip
-\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
-\medskip
-\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
-\vspace{1.3cm}
-%\bigskip
-
-\only<2->{
-\textbf{Idea:} Statistical model with numerical predictions as covariates.\\
-\bigskip
-\textbf{Distribution assumption:}\\
-\vspace{-0.25cm}
-%\hspace{1.5cm}
-%\center
-$$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
-}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.4cm}
-\textbf{Observation data} (National Hydrographical Service):\\
+\frametitle{Circular Data}
+
+\bigskip
+
+\textbf{Examples/Applications:} 
 \begin{itemize}
-\item Daily 24h precipitation sums from July over 28 years (1985--2012).\\
-\item 95 observation stations in Tyrol, Austria.
+\item Wind directions: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$
+\item Hourly crime data: $[0\text{h}, 24\text{h})$ with $0\text{h} = 24\text{h}$
+\item Directional behavior of animals: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$
+\item \ldots
 \end{itemize}
-\vspace{1cm}
-%\bigskip
-\textbf{NWP model output} (Global Ensemble Forecast System):\\
-\begin{itemize}
-\item Ensemble predictions of various meteorological quantities related to precipitation, temperature,
-air pressure, \dots.\\
-\item 11 members each using slightly different perturbed initial conditions.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
-\textbf{Base variables:}
-\begin{itemize}
-\item Total precipitation
-\item Convective available potential energy
-\item Downwards short wave radiation flux (``sunshine'')
-\item Mean sea level pressure
-\item Preciptable water
-\item 2m maximum temperature
-\item Total column-integrated condensate
-\item Temperature 
-\item Temperature differences in altitude
-\end{itemize}
-\medskip
-\textbf{Variations:} Ensemble min/max/mean/standard deviation\\
-\vspace{0.1cm}
-$\rightarrow$ 80 covariates.
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
-\textbf{Application for one station:}
-\begin{itemize}
-\item Learn forest model on data from 24 years (1985--2008).
-\item Evaluate on 4 years (2009--2012).
-\end{itemize}
-\end{frame}
-
-<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
-
-#####
-# load observations and covariates 
-data("RainAxams")
-
-# tree and forest formula
-{
-  dt.formula <- df.formula <- 
-      robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
-      tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
-      tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
-      capepow_mean + capepow_sprd + capepow_min + capepow_max + 
-      capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
-      capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
-      dswrf_mean_mean + dswrf_mean_max +  
-      dswrf_sprd_mean + dswrf_sprd_max + 
-      msl_mean_mean + msl_mean_min + msl_mean_max + 
-      msl_sprd_mean + msl_sprd_min + msl_sprd_max +
-      pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
-      pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
-      tmax_mean_mean + tmax_mean_min + tmax_mean_max +
-      tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
-      tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
-      tcolc_sprd_mean + tcolc_sprd_min + tcolc_sprd_max +
-      t500_mean_mean + t500_mean_min + t500_mean_max +
-      t700_mean_mean + t700_mean_min + t700_mean_max +
-      t850_mean_mean + t850_mean_min + t850_mean_max +
-      t500_sprd_mean + t500_sprd_min + t500_sprd_max +
-      t700_sprd_mean + t700_sprd_min + t700_sprd_max +
-      t850_sprd_mean + t850_sprd_min + t850_sprd_max +
-      tdiff500850_mean + tdiff500850_min + tdiff500850_max +
-      tdiff700850_mean + tdiff700850_min + tdiff700850_max +
-      tdiff500700_mean + tdiff500700_min + tdiff500700_max +
-      msl_diff
-
-  }
-
-  # learning data: 24 years (1985 - 2008, both inlcuded)
-  # testing data: 4 successive years (2009, 2010, 2011, 2012)
-  learndata <- RainAxams[RainAxams$year < 2009,]
-  testdata <- RainAxams[RainAxams$year %in% c(2009, 2010, 2011, 2012),]
-
-##############################################################
-# fitting the models
-set.seed(7)
-
-df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
-                 type.tree = "ctree", ntree = 100, censtype = "left", censpoint = 0,
-                 control = ctree_control(teststat = "quad", testtype = "Univ",
-                                         intersplit = TRUE,
-                                         mincriterion = 0, minsplit = 50,
-                                         minbucket = 20), mtry = 27)
-
-#### prepare data for plot of estimated density functions
-# predictions for one day (in each of the four years) 
-# (19th of July 2011 is missing)
-pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
-
-pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
-pdf <- predict(df, newdata = testdata[pdays,], type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, testdata[pdays,"robs"])
-
-# plot predicted distributions together with observations
-set.seed(7)
-x <- c(0.01, sort(runif(500,0.01,8)))
-y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
-y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
-y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
-y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
-dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
-
-# switch x-axis back to untransformed scale
-x <- x^(1.6)
-
-# point mass (slightly shifted)
-pm1 <- c(0.15, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.05
-pm2 <- c(0.05, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.01
-pm3 <- c(-0.05, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.03
-pm4 <- c(-0.15, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.07
-
-# predictions
-pred1 <- c(testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
-
-#legendheight
-lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
-lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
-lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\begin{center}
-\vspace*{-0.2cm}
-<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=5.2>>=
-plot(x = x, y = y1, type = "l", col = "red", 
-     main = paste0("Axams, July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
-     # xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
-     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-2,28))     #  xlim = c(-1.5,8))
-
-lines(x = x, y = y2, type = "l", col = "blue")
-lines(x = x, y = y3, type = "l", col = "darkgreen")
-lines(x = x, y = y4, type = "l", col = "purple")
-legend('topright', c("Predicted distribution", "Point mass at censoring point", "Observation"),
-       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
-
-# plot point mass
-lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
-lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
-lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
-lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)
-
-points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
-points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
-points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
-points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)
-
-# plot predictions
-points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
-points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
-points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
-points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
-
-lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)
-
-# add labels
-text(x = -1.7, y = lh1, labels = "2009", col = "red", cex = 0.8)         # -0.8
-text(x = -1.7, y = lh2, labels = "2010", col = "blue", cex = 0.8)        # -0.8
-text(x = -1.7, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)   # -0.8
-text(x = -1.7, y = lh4, labels = "2012", col = "purple", cex = 0.8)      # -0.8
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.2cm}
-\textbf{Application for one station:}
-\begin{itemize}
-\item Learn forest model on data from 24 years.
-\item Evaluate on 4 years.
-\item 10 times 7-fold cross validation.
-\item Compare to commonly used heteroscedastic censored Gaussian models:
-\end{itemize}
-%\begin{itemize}
-%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-%\item Boosted GAM: automatic variable selection.
-%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
-%\end{itemize}
-\vspace{0.1cm}
-\begin{table}[t!]
-\footnotesize
-
-\hskip-0.4cm\begin{tabular}{ l l l }
-\hline
-Model & Type & Variable Selection \\
-\hline
-\vspace{-0.15cm}
-\textbf{Forest} & recursive    & automatic \\ 
-                & partitioning &           \\ 
-\hline
-\vspace{-0.15cm}
-\textbf{prespecified GAM} & spline  & based on expert \\
-                          & in each & knowledge      \\
-\hline
-\vspace{-0.15cm}
-\textbf{boosted GAM} & spline  & automatic  \\
-                     & in each &            \\
-\hline
-\vspace{-0.15cm}
-\textbf{EMOS} & linear & total precipitation mean \\
-              &        & and standard deviation   \\
-\hline
-\end{tabular}
-\end{table}
-
-\end{frame}
-
-<<echo=FALSE, results=hide>>=
-#### cross validation rain
-if(file.exists("crps_cross.rda")){
-  load("crps_cross.rda")
-} else {
-
-  nrep_cross <- 10
-  seed <- 7
-
-  res_cross <- mclapply(1:nrep_cross,
-                        function(i){
-
-                          set.seed(seed*i)
-
-                          # randomly split data in 7 parts each including 4 years
-                          years <- 1985:2012
-                          testyears <- list()
-                          for(j in 1:7){
-                            testyears[[j]] <- sample(years, 4, replace = FALSE)
-                            years <- years[!(years %in% testyears[[j]])]
-                          }
-
-                          #crps <- matrix(nrow = 7, ncol = 7)
-                          reslist <- list()
-                          for(k in 1:7){
-                            test <- testyears[[k]]
-                            train <- c(1985:2012)[!c(1985:2012) %in% test]
-
-                            res <- evalmodels(station = "Axams",
-                                              train = train,
-                                              test = test,
-                                              gamboost_cvr = TRUE)
-
-                            #crps[k,] <- res$crps
-                            reslist[[k]] <- res
-                          }
-
-                          #colnames(crps) <- names(res$crps)
-                          return(reslist)
-                        },
-                        mc.cores = detectCores() - 1
-  )
-
-  # extract CRPS
-  crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
-  # loop over all repetitions
-  for(i in 1:length(res_cross)){
-    #loop over all 7 folds (for 7 methods)
-    crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
-    for(j in 1:length(res_cross[[1]])){
-      crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
-    }
-    crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
-  }
-  colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps)
-
-  save(crps_cross, file = "crps_cross.rda")
-}
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\begin{center}
-\vspace*{-0.2cm}
-<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.6, width=6.8>>=
-#par(mar = c(2.5,2,1,2))
-boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
-        names = c("Distributional Forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
-        main = "Cross validation (with reference model EMOS)",
-        cex.main=1.4,
-        cex.lab=1.2,
-        ylab = "CRPS skill score", col = "lightgray") 
-abline(h = 0, col = pal["EMOS"], lwd = 2)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.2cm}
-\textbf{Application for all 95 stations:}
-\begin{itemize}
-\item Learn forest model on data from 24 years (1985--2008).
-\item Evaluate on 4 years (2009--2012).
-\item Compare to commonly used heteroscedastic censored Gaussian models. %:\\
-\end{itemize}
-%\begin{itemize}
-%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-%\item Boosted GAM: automatic variable selection.
-%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
-%\end{itemize}
-
-%\vspace{0.1cm}
-%\begin{table}[t!]
-%\footnotesize
-%\nodindent
-%\hskip-0.4cm\begin{tabular}{ l l l }
-%\hline
-%Model & Type & Variable Selection \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{Forest} & recursive    & automatic \\ 
-%                & partitioning &           \\ 
-%\hline
-%\vspace{-0.15cm}
-%\textbf{prespecified GAM} & spline  & based on expert \\
-%                          & in each & knowledge      \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{boosted GAM} & spline  & automatic  \\
-%                     & in each &            \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{EMOS} & linear & total precipitation mean \\
-%              &        & and standard deviation   \\
-%\hline
-%\end{tabular}
-%\end{table}
-\end{frame}
-
-<<echo = FALSE>>=
-#### prediction over all stations 24 - 4
-if(file.exists("crps_24to4_all.rda")){
-  load("crps_24to4_all.rda")
-} else {
-
-  data("StationsTyrol")
-  stations <- StationsTyrol$name
-  test <- 2009:2012
-  train <- 1985:2008
-
-  res_24to4_all <- mclapply(1:length(stations),
-                            function(i){
-
-                              set.seed(7)
-
-                              res <- evalmodels(station = stations[i],
-                                                train = train,
-                                                test = test,
-                                                gamboost_cvr = TRUE)
-
-                              return(res)
-                            },
-                            mc.cores = detectCores() - 1
-  )
-
-  # extract crps
-  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
-  # loop over all stations
-  for(i in 1:length(stations)){
-    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
-  }
-
-  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
-  rownames(crps_24to4_all) <- stations
-
-  save(crps_24to4_all, file = "crps_24to4_all.rda")
-}
-
-# skill score
-s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
-colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
-
-## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
-
-crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]
-
-# best method
-bst <- apply(crps_map, 1, which.min)
-
-# distance of forest to best other method
-dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]
-
-# breaks/groups
-brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
-#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
-grp <- cut(dst, breaks = brk)
-
-# HCL colors (relatively flashy, essentially CARTO Tropic)
-clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
-
-library("raster") # dem (digital elevation model)
-library("sp")     # gadm www.gadm.org/country
-
-#load("plot_map_rain/data/tirol.rda")
-#load("plot_map_rain/data/dem.rda")
-
-data(StationsTyrol, package = "RainTyrol")
-data(MapTyrol, package = "RainTyrol")
-# data(MapTyrol_border, package = "RainTyrol")
-# Create SpatialPointsDataFrame from station list
-sp <- SpatialPointsDataFrame(subset(StationsTyrol,
-                                    select=c(lon,lat)),
-                             data = subset(StationsTyrol,
-                                           select = -c(lon,lat)),
-                             proj4string = crs(MapTyrol$RasterLayer))
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\begin{center}
-%\vspace*{-2cm}
-
-\setkeys{Gin}{width=0.95\textwidth}
-<<map, fig=TRUE, echo=FALSE, width=10, height=6.5>>=
-
-  ## plot map of Tyrol with all 95 observations
-  layout(cbind(1, 2), width = c(9, 1))
-  par(mar = c(5,4,4,0.1))
-  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
-                main="Stations in Tyrol", cex.main=1.7,
-                ylab = "Latitude", xlab = "Longitude",
-                cex.lab = 1.3, 
-                xlim = c(9.8,13.2), 
-                ylim = c(46.6, 47.87))
-  plot(MapTyrol$SpatialPolygons, add = TRUE)
-  #points(sp[70,], pch = 19, col = "black", cex = 1.85)
-  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
-  text(x = 10.25, y = 47.82, labels = "Models with lowest CRPS")
-  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
-  par(mar = c(0.5,0.2,0.5,2.3))
-  ## legend
-  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
-       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
-  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
-  axis(4, at = brk, las = 1, mgp=c(0,-0.4,-1.2))
-
-@
-\end{center}
-\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
+
+\bigskip
+
+\textbf{Task:} Model periodic response variable measured on the unit circle.
+
+\medskip
+
+\quad $\rightarrow$ Circular distribution
+
 %\begin{center}
@@
 %\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_ll}
+%\includegraphics{}
 %\end{center}
-%\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Von Mises distribution}
+Also known as ``the circular normal distribution'' with
+\begin{itemize}
+\item location parameter $\mu \in [0, 2\pi)$ 
+\item concentration parameter $\kappa > 0$ 
+\end{itemize}
+
+\medskip
+
+The density for $y \in [0, 2\pi)$ is given by:
+$$
+f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
+$$
+where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$.
+
+\bigskip
+
+\textit{FIXME: include a plot of the density function on the unity circle}
 %\begin{center}
@@
 %\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_RMSE_par}
+%\includegraphics{}
 %\end{center}
-%\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Wind Directions}
+
+\textbf{Goal:} Predict wind directions at Innsbruck Airport (Austria).
+
+\medskip
+
+\textbf{Input:} Observation data of various meteorological quantities of the previous hour(s) from Innsbruck and 5 nearby weather stations.
+
+\medskip
+
+\textbf{Base variables:}
+\begin{itemize}
+\item Wind direction
+\item Temperature
+\item Air pressure
+\item \dots 
+\item \textit{FIXME: list of all included covariates}
+\end{itemize}
 %\begin{center}
@@
 %\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_RMSE_exp}
+%\includegraphics{}
 %\end{center}
-%\end{frame}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Wind Directions}
+
+\textbf{Distributional tree model:}
+
+\begin{center}
+%\vspace*{1cm}
+\setkeys{Gin}{width=\linewidth}
+\includegraphics{schlosser-circtree_plot.pdf}
+\end{center}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Wind Directions}
+
+\textbf{Cross validation:}
+
+\bigskip
+
+\textit{FIXME: explain applied setting}
+
+%\begin{center}
+%\vspace*{-1cm}
+%\setkeys{Gin}{width=0.8\linewidth}
+%\includegraphics{}
+%\end{center}
+\end{frame}
 
@@
 \frametitle{Software}
-\vspace{0.4cm}
-\textbf{Package:} \emph{disttree} available on R-Forge at\\
-
-\medskip
-
-\url{https://R-Forge.R-project.org/projects/partykit/}\\
+%\vspace{0.4cm}
+\textbf{Packages:} \emph{disttree} and \emph{circmax} available on R-Forge at\\
+
+\medskip
+
+\small{\url{https://R-Forge.R-project.org/projects/partykit/}\\ 
+and\\
+\url{https://r-forge.r-project.org/projects/uibk-rprog-2018/}}\\
 
@@
 \begin{tabular}{ll}
-\code{distfit}    & Distributional fits (ML, \code{gamlss.family}/custom \code{list}).\\
-                  & No covariates. \\
+
 \code{disttree}   & Distributional trees (\code{ctree}/\code{mob} + \code{distfit}).\\
@@
 
+\bigskip
+\bigskip
+
+\textbf{Distribution family:} \code{dist\_vonmises}
+
 \end{frame}
@@
 \begin{frame}
-\frametitle{References}
-
-\vspace{-0.2cm}
-
-\footnotesize
-Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
-  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
-  \emph{arXiv 1804.02921}, arXiv.org E-Print Archive.
-  \url{http://arxiv.org/abs/1804.02921}
+\frametitle{References (1)}
+
+%\vspace{-0.4cm}
+
+\small
+Bai J, Perron P (2003).
+ \dquote{Computation and Analysis of Multiple Structural Change Models.}
+ \textit{Journal of Applied Econometrics}, 
+ \textbf{18}, 1--22.
+ \doi{}
+
+\medskip
+
+Breiman L (2001).
+ \dquote{Random Forests.}
+ \textit{Machine Learning}, 
+ \textbf{45}(1), 5--32.
+ \doi{10.1023/A:1010933404324}
+
+\medskip
+
+Fisher N I, Lee A J (1992).
+ \dquote{Regression Models for an Angular Response.}
+ \textit{Biometrics}, 
+ \textbf{48}(3), 665--677. 
+ \doi{}
 
@@
 
+Hothorn T, Zeileis A (2008).
+ \dquote{Generalized Maximally Selected Statistics.}
+ \textit{Biometrics}, 
+ \textbf{64}(4), 1263--1269. 
+ \doi{}
+
+\medskip
+
+Jammalamadaka S R, Sengupta A (2001).
+ \textit{Topics in Circular Statistics}.
+ World Scientific.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{References (2)}
+
+%\vspace{-0.4cm}
+
+\small
+Mulder K, Klugkist I (2017).
+ \dquote{Bayesian Estimation and Hypothesis Tests for a Circular Generalized Linear Model.}
+ \textit{Journal of Mathematical Psychology}, 
+ \textbf{80}, 4--14. 
+ \doi{}
+
+\medskip
+
+Schlosser L, Hothorn T, Stauffer R, Zeileis A (2019).
+  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
+  \emph{arXiv 1804.02921}, arXiv.org E-Print Archive.
+  \url{http://arxiv.org/abs/1804.02921}
+
+\medskip
+
 Zeileis A, Hothorn T, Hornik K (2008).
@@
 
-\medskip
-
-Hothorn T, Zeileis A (2015).
- \dquote{\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
- \textit{Journal of Machine Learning Research},
- \textbf{16}, 3905--3909.
- \url{http://www.jmlr.org/papers/v16/hothorn15a.html}
-
-\medskip
-
-Stasinopoulos DM, Rigby RA (2007).
-  \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in \textsf{R}.}
-  \textit{Journal of Statistical Software}, 
-  \textbf{23}(7), 1--46.
-  \doi{10.18637/jss.v023.i07}
-
 %\medskip
 
-%Seibold H, Zeileis A, Hothorn T (2017).
-%  \dquote{Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.}
-%  \textit{Statistical Methods in Medical Research}, 
-%  \textbf{12}(1), 45--63.
-%  \doi{10.1177/0962280217693034}
-
-%\medskip
-
-%Hothorn T, Zeileis A (2017).
-%  \dquote{Transformation Forests.}
-%  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
-%  \url{http://arxiv.org/abs/1701.02110}
-
-\medskip
-
-Hofner B, Mayr A, Schmid M (2016).
-  \dquote{\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and Variable Selection in the {GAMLSS} Framework}
-  \textit{Journal of Statistical Software},
-  \textbf{74}(1), 1--31.
-  \doi{10.18637/jss.v074.i01}
-
-\end{frame}
-
-%\begin{frame}
-%\frametitle{References (2)}
-
-%\vspace{-0.2cm}
-
-%\footnotesize
-
-%Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-%  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and %Minimum CRPS Estimation}
-%  \textit{Monthly Weather Review}, 
-%  \textbf{133}(5), 1098--1118.
-%  \doi{10.1175/MWR2904.1}
-
-%\medskip
-
-% Reference for the observation data set
-%BMLFUW (2016).
-%  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtsch%aft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
-%  \text{Available at \url{http://ehyd.gv.at}}, 
-%  \text{Accessed: 2016--02--29}.
-
-%\medskip
-
-% Reference for the GFS reforecast version 2 data set 
-%Hamill T M, Bates G T, Whitaker J S, Murray D R, Fiorino M, Galarneau T J Jr., Zhu Y, Lapenta %W (2013).
-%  \dquote{{NOAA}'s Second-Generation Global Medium-Range Ensemble Reforecast Dataset}
-%  \textit{Bulletin of the American Meteorological Society}, 
-%  \textbf{94}(10), 1553--1565.
-%  \doi{10.1175/BAMS-D-12-00014.1}
-
-%\medskip
-
-%Breiman L (2001).
-%  \dquote{Random {Forests}}
-%  \textit{Machine Learning}, 
-%  \textbf{45}(1), 5--32.
-%  \doi{10.1023/A:1010933404324}
-
-%Messner J W,  Mayr G J, Zeileis A (2016).
-%  \dquote{Heteroscedastic Censored and Truncated Regression with \textbf{crch}}
-%  \textit{The \textsf{R} Journal}, 
-%  \textbf{8}(1), 173--181.\\
-%  \scriptsize
-%  \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
-%  \footnotesize
-
-%\medskip
-
-%Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
-%  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored %Additive Regression Model}
-%  \textit{International Journal of Climatology}, 
-%  \textbf{37}(7), 3264--3275.
-%  \doi{10.1002/joc.4913}
-
-%\medskip
-
-%Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
-%  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: %Unanimous Predictions, Heavy Tails, and Link Functions}
-%  \textit{Monthly Weather Review}, 
-%  \text{Forthcoming}.
-%  \doi{10.1175/MWR-D-16-0388.1}
-
-%\medskip
-
-%{CGIAR-CSI} (2016).
-%  \dquote{{SRTM} 90m Digital Elevation Database {v4.1}}
-%  \text{Available at \url{http://srtm.csi.cgiar.org}}, 
-%  \text{Accessed: 2016--07--10}.
-
-%\end{frame}
+%Hothorn T, Zeileis A (2015).
+% \dquote{\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
+% \textit{Journal of Machine Learning Research},
+% \textbf{16}, 3905--3909.
+% \url{http://www.jmlr.org/papers/v16/hothorn15a.html}
+
+\end{frame}
 

### 2019-06-28 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
first draft of slides for IWSM 2019
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2626 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 227c0bb756d4097710ef42ed50c0b6a74658c2bd

@@
 that can improve forecasts by regularizing and stabilizing the model.
-Random forests introduced by Breiman~et.~al~(1984) average the
+Random forests introduced by Breiman~(2001) average the
 predictions of an ensemble of trees, each built on a subsample 
@@
 
-\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
-     Classification and Regression Trees.
-     {\it Wadsworth}.
+\item[Breiman, L.] (2001).
+     Random Forests.
+     {\it Machine Learning}, {\bf 45}, 1,
+     5\,--\,32.
 
@@
 
-\item[Jammalamadaka, S. R. and Sengupta, A.] (2001).
+\item[Jammalamadaka, S. R., and Sengupta, A.] (2001).
      {\it Topics in Circular Statistics}.
@@
 
-\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2019).
+\item[Schlosser, L., Hothorn, T., Stauffer, R., and Zeileis, A.] (2019).
      Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.

### 2019-06-28 pkg/disttree/inst/IWSM2019/slides.Rnw
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2627 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f9e30d2e14d3e8de24d86a859fdee4caad228dbc

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber, noslidenumber]{uibk}
+\usetheme[url, license]{uibk}
 
@@
 for Circular Data}
-\author{Lisa Schlosser, Moritz N. Lang, Torsten Hothorn, \\
+\author{Lisa Schlosser, Moritz N. Lang, Torsten~Hothorn,\\
 Georg~J.~Mayr, Reto Stauffer, Achim Zeileis}
@@
 %\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
-\URL{https://R-Forge.R-project.org/projects/partykit/ \\
-https://partykit.org}
+\URL{https://R-Forge.R-project.org/projects/partykit/}
 \headerimage{3}
@@
 $$
-f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
+f_\text{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
 $$
@@
 and\\
-\url{https://r-forge.r-project.org/projects/uibk-rprog-2018/}}\\
+\url{https://R-Forge.R-project.org/projects/uibk-rprog-2018/}}\\
 
@@
 \begin{frame}
-\frametitle{References (1)}
+\frametitle{References}
 
@@
 \begin{frame}
-\frametitle{References (2)}
+\frametitle{References}
 

### 2019-06-30 pkg/disttree/inst/IWSM2019/slides.Rnw
resvision of slides for IWSM 2019: application Ibk airport on motivation slide, adaption of theory to specific case of von Mises distribution (including graphics)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2629 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bbf0613f62020166b886de5796855654002acbae

@@
-\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[url, license]{uibk}
+\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
+\usetheme[nototalframenumber, url, license]{uibk}
 
-\title{Distributional Trees 
-%and Forests \\
-for Circular Data}
-\author{Lisa Schlosser, Moritz N. Lang, Torsten~Hothorn,\\
-Georg~J.~Mayr, Reto Stauffer, Achim Zeileis}
+\title{Distributional Trees and Forests for Circular Data}
+\author{Lisa Schlosser, Moritz N. Lang, \\
+Torsten~Hothorn, Georg~J.~Mayr, Reto Stauffer, Achim Zeileis}
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
@@
 %% includes a replacement for \usepackage{Sweave}
-\usepackage{Sweave}
+% \usepackage{Sweave}
 \usepackage{amsmath,tikz}
 \usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-\graphicspath{{plots/}}
+\usepackage{xcolor}
+\usepackage{BOONDOX-cal}
+%\graphicspath{{plots/}}
 
@@
 library("disttree")
-library("Formula")
+#library("Formula")
 library("latex2exp")
-
-## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
-
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
-
-transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)
 
@@
 \begin{frame}[fragile]
-\frametitle{Distributional Tree}
+\frametitle{Motivation}
+\textbf{Goal:} Regression model for a circular response variable%, \\
+%\hspace{1.05cm} i.e. a periodic response variable measured on the unit circle.
 
-\vspace{1cm}
+\medskip
 
-{\bf Starting point:}
+\textbf{Examples of cirular data:} 
 \begin{itemize}
-\item Response $Y$
-\item Covariate(s) $X$ $\rightarrow$ split variable(s) $X$
-\item Distribution family $\mathcal{D}(\theta)$
+\item Wind directions: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$
+\item Hourly crime data: $[0\text{h}, 24\text{h})$ with $0\text{h} = 24\text{h}$
+\item Directional behavior of animals: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$
+%\item \ldots
 \end{itemize}
 
-\vspace{1cm}
+\medskip
 
-{\bf Goal:} Probabilistic model
+\textbf{Application:} Probabilistic forecasting of wind directions for the next hour at Innsbruck airport (Austria, \textcolor{gray}{\url{https://innsbruck-airport.panomax.com}}).
 
-%Idea: maximum-likelihood method
-%\begin{enumerate}
-%\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-%\item Estimate $\hat{\theta}$ via maximum likelihood.
-%\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
-%\item Assess whether the \emph{model scores} are associated with
-%    (or change along) any of the available covariates -- e.g.,
-%    using parameter instability tests (\emph{strucchange}) or
-%    conditional inference (\emph{coin}).
-%\item Split the sample along the partitioning variable with the strongest association or instability.
-%    Choose breakpoint with highest improvement in log-likelihood.
-%\item Repeat steps 2--4 recursively in the subgroups
-%    until some stopping criterion is met.
-%    %-- e.g., for significance or sample size.
-%\item Choose the variable with the strongest association
-%\item Choose the split point which leads to the highest improvement of the model
-%\item Split and repeat 2-6 in each node until a stopping criterion is met
-%\end{enumerate}
+\medskip
+
+\textbf{Previous knowledge:} Possibly abrupt changes due to geographical position.
+
+\medskip
+
+\textbf{Idea:} Regression trees
+
+\bigskip
+
+$\rightarrow$ \textbf{Circular distributional trees}
 
@@
 
-\begin{frame}
-\frametitle{Distributional Tree}
-\vspace*{-0.12cm}
-\begin{center} 
-DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
-\vspace*{-0.21cm}
-<<dgp_tree, echo=FALSE, results=hide>>=
-data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
-names(data) <- c("x","x","x")
-fig <- party(
-  partynode(1L,
-            split = partysplit(2L, breaks = 0.4),
-            kids = list(
-              partynode(2L, info = c(
-                "n = 200",
-                "   True parameters:   ",
-                expression(mu == '4'),
-                expression(sigma == '1')
-              )),
-              partynode(3L,
-                        split = partysplit(3L, breaks = 0.8),
-                        kids = list(
-                          partynode(4L, info = c(
-                            "n = 200",
-                            "   True parameters:   ",
-                            expression(mu == '12'),
-                            expression(sigma == '3')
-                          )),
-                          partynode(5L, info = c(
-                            "n = 100",
-                            "   True parameters:   ",
-                            expression(mu == '4'),
-                            expression(sigma == '3')
-                          )))))),
-  data
-)
+\begin{frame}[fragile]
+\frametitle{Von Mises distribution}
+\begin{minipage}{0.5\textwidth}
+``Circular normal distribution'' with
+\begin{itemize}
+\item location parameter $\mu \in [0, 2\pi)$ 
+\item concentration parameter $\kappa > 0$ 
+\end{itemize}
+\end{minipage}
+\hspace{1cm}
+\begin{minipage}{0.4\textwidth}
+%\begin{center}
+%\vspace*{-1cm}
+\setkeys{Gin}{width=0.6\linewidth}
+\includegraphics{density_right.pdf}
+%\end{center}
+\end{minipage}
 
-node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
-    gp = gpar()) 
-{
-    meta <- obj$data
-    nam <- names(obj)
-    extract_label <- function(node) {
-        if (is.terminal(node)) 
-            return(rep.int("", 2L))
-        varlab <- character_split(split_node(node), meta)$name
-        if (abbreviate > 0L) 
-            varlab <- abbreviate(varlab, as.integer(abbreviate))
-        if (pval) {
-            nullna <- function(x) is.null(x) || is.na(x)
-            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
-                silent = TRUE))
-            pval <- if (inherits(pval, "try-error")) 
-                FALSE
-            else pval
-        }
-        if (pval) {
-            pvalue <- node$info$p.value
-            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
-                paste("p =", round(pvalue, digits = 3L)))
-        }
-        else {
-            plab <- ""
-        }
-        return(c(varlab, plab))
-    }
-    maxstr <- function(node) {
-        lab <- extract_label(node)
-        klab <- if (is.terminal(node)) 
-            ""
-        else unlist(lapply(kids_node(node), maxstr))
-        lab <- c(lab, klab)
-        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
-        lab <- lab[which.max(nchar(lab))]
-        if (length(lab) < 1L) 
-            lab <- ""
-        return(lab)
-    }
-    nstr <- maxstr(node_party(obj))
-    if (nchar(nstr) < 6) 
-        nstr <- "aAAAAa"
-    rval <- function(node) {
-        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
-            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
-            height = unit(3, "lines"), name = paste("node_inner", 
-                id_node(node), sep = ""), gp = gp)
-        pushViewport(node_vp)
-        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
-            seq(0.8, 1, by = 0.01))
-        yell <- sqrt(xell * (1 - xell))
-        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
-        lab <- extract_label(node)
-        fill <- rep(fill, length.out = 2L)
-        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
-            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
-        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
-            "lines"))
-        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
-        #    "lines"))
-        grid.text(lab[2L], y = unit(1, "lines"))
-        if (id) {
-            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
-                "npc"), width = max(unit(1, "lines"), unit(1.3, 
-                "strwidth", nam[id_node(node)])), height = max(unit(1, 
-                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
-            pushViewport(nodeIDvp)
-            grid.rect(gp = gpar(fill = fill[2]))
-            grid.text(nam[id_node(node)])
-            popViewport()
-        }
-        upViewport()
-    }
-    return(rval)
-}
+\medskip
 
-class(node_inner_ext) <- "grapcon_generator"
+The density for $y \in [0, 2\pi)$ and parameter vector $\theta = (\mu, \kappa)$ is given by
+$$
+f_\mathcal{vM}(y; \theta) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
+$$
+where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$.
 
-@
-
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_dgp, fig=TRUE, echo=FALSE>>=
-paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-plot(fig, inner_panel = node_inner_ext,
-     tp_args = list(FUN = identity, width = 18, fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]),
-     drop_terminal = TRUE, tnex = 1.7)
-@
-\end{center}
-\end{frame}
-
-<<tree, echo=FALSE, results=hide>>=
-set.seed(7)
-nobs <- 500
-x <- runif(nobs, 0, 1)
-mu <- sigma <- ytrue <- numeric(length = nobs)
-for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
-for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
-y <- rnorm(nobs, mean = mu, sd = sigma)
-#y <- rcnorm(nobs, mean = mu, sd = sigma, left = 0)
-ytrue <- mu
-data <- data.frame(cbind(y,x, ytrue))
-tree <- disttree(y ~ x, data = data, family = NO(), type.tree = "mob")
-#tree <- disttree(y ~ x, data = data, family = dist_list_cens_normal)
-@
-
-\begin{frame}[fragile]
-\frametitle{Distributional Tree}
-\begin{center}
-\vspace*{-0.12cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_estpar, fig=TRUE, echo=FALSE>>=
-# function for output in terminal panels
-FUN <- function (x) 
-{
-  cf <- x$coefficients
-  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
-  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
-                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
-}
-
-paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
-
-## plot version using FUN and tree of class 'disttree'
-plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
-     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
-     ip_args = list(fill = paltrees[c(2, 3)]))
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Distributional Tree}
-\begin{center}
-\vspace*{-0.12cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_box, fig=TRUE, echo=FALSE>>=
-plot(as.constparty(tree), tnex = 1.7, drop = TRUE,
-     tp_args = list(fill = paltrees[c(1, 3)], ylines = 1.5), 
-     ip_args = list(fill = paltrees[c(2, 3)]))
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Distributional Tree}
-\begin{center}
-\vspace*{-0.12cm}
-Model: \code{disttree(y ~ x)}\\
-\vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
-<<plottree_dens, fig=TRUE, echo=FALSE>>=
-node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
-                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
-                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
-{
-  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
-  ylines <- 1.5
-  if (is.null(xscale)) xscale <- c(-5.1,22.5)
-  if (is.null(yscale)) yscale <- c(-0.05,0.45)
-  xr <- xscale
-  yr <- yscale
-
-  if (horizontal) {
-    yyy <- xscale
-    xscale <- yscale
-    yscale <- yyy
-  }
-
-  rval <- function(node) {
-    yrange <- seq(from = -20, to = 90)/4
-    ydens <- node$info$object$ddist(yrange)
-
-    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3, 
-                                            widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")), 
-                                            heights = unit(c(1, 1), c("lines", "null"))), 
-                       width = unit(1, "npc"), 
-                       height = unit(1, "npc") - unit(2, "lines"), 
-                       name = paste("node_density",node$id, sep = ""))
-    pushViewport(top_vp)
-    grid.rect(gp = gpar(fill = "white", col = 0))
-    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
-    pushViewport(top)
-    mainlab <- paste(ifelse(id, paste("Node", node$id, "(n = "), "n = "), node$info$nobs, ifelse(id, ")", ""), sep = "")
-
-    grid.text(mainlab)
-    popViewport()
-    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2, 
-                     xscale = xscale, yscale = yscale, 
-                     name = paste("node_density",  node$id, "plot", sep = ""))
-    pushViewport(plot)
-    yd <- ydens
-    xd <- yrange
-    if (horizontal) {
-      yyy <- xd
-      xd <- yd
-      yd <- yyy
-      yyy <- xr
-      xr <- yr
-      yr <- yyy
-      rxd <- rep(0, length(xd))
-      ryd <- rev(yd)
-    } else {
-      rxd <- rev(xd)
-      ryd <- rep(0, length(yd))
-    }
-
-    if (rug) {
-      nodeobs <- node$info$object$y
-      if (horizontal) {
-        grid.rect(x = xscale[1], y = nodeobs , height = 0, width = xscale[1], 
-                  default.units = "native", just = c("right", "bottom"),
-		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
-      } else {
-        grid.rect(x = nodeobs, y = yscale[1], 
-                  width = 0, height = abs(yscale[1]), default.units = "native", 
-                  just = c("center", "bottom"),
-		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
-        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
-        #           default.units = "native")
-        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
-        #           default.units = "native")
-      }
-    }
-
-    grid.polygon(x = c(xd, rxd), y = c(yd, ryd), default.units = "native",
-              gp = gpar(col = "black", fill = fill, lwd = lwd))
-    #grid.lines(x = xd, y = yd, default.units = "native", 
-    #           gp = gpar(col = col, lwd = lwd))
-    grid.xaxis()
-    grid.yaxis()
-    grid.rect(gp = gpar(fill = "transparent"))
-    upViewport(2)
-  }
-  return(rval)
-}
-
-class(node_density) <- "grapcon_generator"
-
-plot(tree, tnex = 1.7, drop = TRUE,
-     terminal_panel = node_density,
-     tp_args = list(fill = paltrees[c(1, 3)]), 
-     ip_args = list(fill = paltrees[c(2, 3)]))
-@
-\end{center}
-\end{frame}
-
-<<treedata, echo=FALSE, results=hide>>=
-set.seed(54)
-nobs <- 500
-x <- runif(nobs, 0, 1)
-mu <- sigma <- ytrue <- numeric(length = nobs)
-for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
-for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
-y <- rnorm(nobs, mean = mu, sd = sigma)
-ytrue <- mu
-data <- data.frame(cbind(y,x, ytrue))
-
-alldata <- cbind(data, mu, sigma)
-odata <- alldata[order(alldata["x"]),]
-@
-
-\begin{frame}[fragile]
-\frametitle{Distributional Tree}
-\begin{center}
-\vspace*{-1.03cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plottree_xyplot, fig=TRUE, echo=FALSE>>=
-par(mar=c(5.1,4.1,4.1,3.1))
-plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "grey")
-#plot(y=odata$y, x=odata$x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray")
-lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
-polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
-  col = pallight["forest"], border = "transparent")
-legend(x = -0.0715, y = 21.62, expression(mu  %+-%  sigma), bty = "n")
-#lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
-#lines(x = odata$x, y = odata$mu - odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
-@
-\end{center}
+The corresponding log-likelihood function is defined by
+$$
+\ell(\theta; y) =  \log(f_\mathcal{vM}(y;\theta)).
+$$
 \end{frame}
@@
 \begin{enumerate}
-\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
+\item<3-> Fit global distributional model $\mathcal{vM}(Y; \theta)$: \\ % to the whole data set:\\
 Estimate $\hat{\theta}$ via maximum likelihood\\
@@
 \visible<4>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{D}(Y;\hat{\theta}$)};
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{vM}(Y;\hat{\theta}$)};
 }
 \visible<5->{
-\node[inner sep=0pt] (density1) at (0.8, 1.7)
-    {\includegraphics[width=0.6\textwidth]{density1.jpeg}};
+\node[inner sep=0pt] (density_root) at (0.8, 1.7)
+    {\includegraphics[width=0.6\textwidth]{density_root.pdf}};
 }
@@
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
-\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1);
-\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2);
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
+\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1);
+\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2);
 }
@@
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
-\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
+\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (1, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
 }
 \visible<11>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{vM}(Y_1;\hat{\theta_1}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {$\mathcal{vM}(Y_2;\hat{\theta_2}$)};
 }
 \visible<11->{
-\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
-\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
+\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
 }
 \visible<12->{
-\node[inner sep=0pt] (density2) at (-0.1,-0.1)
-    {\includegraphics[width=0.5\textwidth]{density2.jpeg}};
+\node[inner sep=0pt] (density2) at (-0.1,-0.3)
+    {\includegraphics[width=0.5\textwidth]{density_left.pdf}};
 }
 \visible<12->{
-\node[inner sep=0pt] (density3) at (1.7,-0.1)
-    {\includegraphics[width=0.5\textwidth]{density3.jpeg}};
+\node[inner sep=0pt] (density3) at (1.7,-0.3)
+    {\includegraphics[width=0.5\textwidth]{density_right.pdf}};
 }
@@
 \setcounter{enumi}{2}
-\item<8-> Split along the covariate $X$ with strongest association or instability and at breakpoint $p$ with highest improvement \\
-in log-likelihood.
-\item<10-> Repeat steps 1--3 recursively until some stopping criterion \\
-is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
+\item<8-> Split along the covariate $X$ with strongest association or instability and\\ at breakpoint $p$ with highest improvement in log-likelihood.
+\item<10-> Repeat steps 1--3 recursively until some stopping criterion is met,\\
+yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
 \end{enumerate}
 %\end{adjustwidth}
-\vspace{0.4cm}
+\vspace{0.3cm}
 \visible<13->{
@@
 \textbf{Parameter estimator} for \only<1-3>{\textbf{a global}}\only<4->{\textbf{an adaptive local}}\\ model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$} :
-
 \[
@@
 
-\begin{frame}[fragile]
-\frametitle{Circular Data}
-
-\bigskip
-
-\textbf{Examples/Applications:} 
-\begin{itemize}
-\item Wind directions: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$
-\item Hourly crime data: $[0\text{h}, 24\text{h})$ with $0\text{h} = 24\text{h}$
-\item Directional behavior of animals: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$
-\item \ldots
-\end{itemize}
-
-\bigskip
-
-\textbf{Task:} Model periodic response variable measured on the unit circle.
-
-\medskip
-
-\quad $\rightarrow$ Circular distribution
-
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{}
-%\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Von Mises distribution}
-Also known as ``the circular normal distribution'' with
-\begin{itemize}
-\item location parameter $\mu \in [0, 2\pi)$ 
-\item concentration parameter $\kappa > 0$ 
-\end{itemize}
-
-\medskip
-
-The density for $y \in [0, 2\pi)$ is given by:
-$$
-f_\text{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
-$$
-where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$.
-
-\bigskip
-
-\textit{FIXME: include a plot of the density function on the unity circle}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{}
-%\end{center}
-\end{frame}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Application
 
@@
 
-\textbf{Goal:} Predict wind directions at Innsbruck Airport (Austria).
+%\textbf{Goal:} Predict wind directions for the next hour at Innsbruck airport (Austria).
 
-\medskip
+%\medskip
 
-\textbf{Input:} Observation data of various meteorological quantities of the previous hour(s) from Innsbruck and 5 nearby weather stations.
+\textbf{Input:} Observation data of various meteorological quantities of the previous hour(s) from Innsbruck airport and 5 nearby weather stations.
 
@@
 \item Wind direction
+\item Wind speed
 \item Temperature
@@
 \end{frame}
+
+%\begin{frame}[fragile]
+%\includegraphics{airport_ibk.jpg}
+%\end{frame}
 
@@
 
+\textbf{Applied models and reference models:}
+\begin{itemize}
+\item Circular distributional tree
+\item Circular distributional forest
+\item Persistence
+\item Climatology 
+\end{itemize}
+
+\medskip
+
+\textbf{Data:} .. observations of response variables and .. covariates
+
+\medskip
+
 \textbf{Cross validation:} 
+\begin{itemize}
+\item 5-fold setting 
+\item learning data: .. observations
+\item testing data: .. observations
+\end{itemize}
 
-\bigskip
+\medskip
 
-\textit{FIXME: explain applied setting}
+\textbf{Evaluation:} Results are compared based on circular CRPS.
 
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Wind Directions}
+
+boxplot CRPS cross validation
 %\begin{center}
@@
 %\vspace{0.4cm}
-\textbf{Packages:} \emph{disttree} and \emph{circmax} available on R-Forge at\\
+\textbf{Packages:} \emph{disttree} %and \emph{circmax} 
+available on R-Forge at\\
 
@@
 
-\small{\url{https://R-Forge.R-project.org/projects/partykit/}\\ 
-and\\
-\url{https://R-Forge.R-project.org/projects/uibk-rprog-2018/}}\\
+\small{\url{https://R-Forge.R-project.org/projects/partykit/}}\\ 
+%and\\
+%\url{https://R-Forge.R-project.org/projects/uibk-rprog-2018/}}\\
 
@@
 
-%\vspace{-0.4cm}
+\vspace{-0.3cm}
 
-\small
-Bai J, Perron P (2003).
- \dquote{Computation and Analysis of Multiple Structural Change Models.}
- \textit{Journal of Applied Econometrics}, 
- \textbf{18}, 1--22.
- \doi{}
-
-\medskip
+\footnotesize
+%Bai J, Perron P (2003).
+% \dquote{Computation and Analysis of Multiple Structural Change Models.}
+% \textit{Journal of Applied Econometrics}, 
+% \textbf{18}, 1--22.
+% \doi{10.1002/jae.659}
+% 
+%\medskip
 
@@
  \textbf{48}(3), 665--677. 
- \doi{}
+ \doi{10.2307/2532334}
 
@@
 
-Hothorn T, Zeileis A (2008).
- \dquote{Generalized Maximally Selected Statistics.}
- \textit{Biometrics}, 
- \textbf{64}(4), 1263--1269. 
- \doi{}
-
-\medskip
+%Hothorn T, Zeileis A (2008).
+% \dquote{Generalized Maximally Selected Statistics.}
+% \textit{Biometrics}, 
+% \textbf{64}(4), 1263--1269. 
+% \doi{10.1111/j.1541-0420.2008.00995.x}
+% 
+%\medskip
 
@@
 
-\end{frame}
+\medskip
 
-\begin{frame}
-\frametitle{References}
-
-%\vspace{-0.4cm}
-
-\small
 Mulder K, Klugkist I (2017).
@@
  \textbf{80}, 4--14. 
- \doi{}
+ \doi{10.1016/j.jmp.2017.07.001}
 

### 2019-07-02 pkg/disttree/inst/CMStatistics2019/abstract/abstract_lang_cmstatistics2019.tex
Intital commit of cmstatistics abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2632 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0c40c365276c0f014929765e25b9979776651761

\documentclass[a4paper,nojss]{jss}
\usepackage{amsmath,amssymb,amsfonts,thumbpdf, caption, subcaption, lipsum}
%\usepackage[utf8]{inputenc}

\newcommand{\CRANpkg}[1]{\href{http://CRAN.R-project.org/package=#1}{\pkg{#1}}}%
\newcommand{\samp}[1]{`\code{#1}'}%
\DefineVerbatimEnvironment{example}{Verbatim}{}
\setkeys{Gin}{width=\textwidth}

%% for internal use
\newcommand{\fixme}[1]{\marginpar{\color{black}fixme} {\color{red}#1}}
\newcommand{\readme}[1]{\marginpar{\color{black}readme} {\emph{\color{red}(#1)}}}
\usepackage{chngcntr}

%%% jss stuff
%\let\proglang=\textsf
%\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
%\bibliographystyle{jss}
%\makeatletter
%\newcommand\code{\bgroup\@makeother\_\@makeother\~\@makeother\$\@codex}
%\def\@codex#1{{\normalfont\ttfamily\hyphenchar\font=-1 #1}\egroup}
%\makeatother

\title{Distributional Trees and Forests for Circular Data}

\author{Moritz N. Lang\\University of Innsbruck \And
        Lisa Schlosser\\University of Innsbruck \And
        Torsten Hothorn\\University of Zurich \And
        Georg J. Mayr\\University of Innsbruck \And
        Reto Stauffer\\University of Innsbruck \And
        Achim Zeileis\\University of Innsbruck}

\Abstract{
Circular data can be found in a variety of applications and subject areas,  e.g., hourly crime rate in the social-economics, animal movement direction or gene-structure in biology, and wind direction as one of the most important weather variables in meteorology. For probabilistic modeling of circular data the von Mises distribution is widely used. While most of the already existing approaches are built on additive regression models, we propose an adaption of regression trees to circular data by employing distributional trees. In comparison to the more commonly-used additive models, the resulting distributional trees are easy to interpret, can detect non-additive effects, and select covariates and their interactions automatically. In addition, as natural extension ensembles or forests of such circular trees are introduced that can further improve the forecasts by regularizing and stabilizing the model. For illustration, short-term probabilistic wind direction forecasts at airports are obtained to assure a safe landing of airplanes which are sensitive to wind shifts. To benchmark the performance of distributional trees and forests, their predictive skill is compared to a persistence model, a climatology model and to a additive regression model, employing the continuous ranked probability score. Model code for circular distributional trees and forest is available in the newly developed R package "disttree".
}
\Keywords{}
\Plainkeywords{}

\Address{
  Moritz N. Lang\\
  University of Innsbruck\\
  6020 Innsbruck, Austria\\
  E-mail: \email{moritz.lang@uibk.ac.at}
}

\begin{document}
\end{document}

### 2019-07-02 pkg/disttree/inst/CMStatistics2019/abstract/abstract_lang_cmstatistics2019.tex
Update abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2633 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b13158ad32026ffb87455116cef32fa0f004eca8

@@
 \Abstract{
-Circular data can be found in a variety of applications and subject areas,  e.g., hourly crime rate in the social-economics, animal movement direction or gene-structure in biology, and wind direction as one of the most important weather variables in meteorology.
-For probabilistic modeling of circular data the von Mises distribution is widely used.
-While most of the already existing approaches are built on additive regression models, we propose an adaption of regression trees to circular data by employing distributional trees.
-In comparison to the more commonly-used additive models, the resulting distributional trees are easy to interpret, can detect non-additive effects, and select covariates and their interactions automatically.
-In addition, as natural extension ensembles or forests of such circular trees are introduced that can further improve the forecasts by regularizing and stabilizing the model.
-For illustration, short-term probabilistic wind direction forecasts at airports are obtained to assure a safe landing of airplanes which are sensitive to wind shifts.
-To benchmark the performance of distributional trees and forests, their predictive skill is compared to a persistence model, a climatology model and to a additive regression model, employing the continuous ranked probability score.
-Model code for circular distributional trees and forest is available in the newly developed R package "disttree".
+Circular data can be found in a variety of applications and subject areas,
+e.g., hourly crime rate in the social-economics, animal movement direction or
+gene-structure in biology, and wind direction as one of the most important
+weather variables in meteorology. For probabilistic modeling of circular data
+the von Mises distribution is widely used. While most of the already existing
+approaches are built on additive regression models, we propose an adaption of
+regression trees to circular data by employing distributional trees. In
+comparison to the more commonly-used additive models, the resulting
+distributional trees are easy to interpret, can detect non-additive effects,
+and select covariates and their interactions automatically. In addition, as
+natural extension ensembles or forests of such circular trees are introduced
+that can further improve the forecasts by regularizing and stabilizing the
+model. For illustration, short-term probabilistic wind direction forecasts at
+airports are obtained in order to direct airplanes to a a safe landing. To
+benchmark the performance of distributional trees and forests, their predictive
+skill is compared to a persistence model, a climatology model and to a additive
+regression model, employing the continuous ranked probability score. Model code
+for circular distributional trees and forest is available in the R package "disttree".
 }
+
 \Keywords{}
+
 \Plainkeywords{}

### 2019-07-03 pkg/disttree/inst/IWSM2019/slides.Rnw
revision of slides for IWSM 2019, added map and boxplots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2638 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8a30f663555be7c4a2f53bc2c8cb2a058d97fee7

@@
 \usepackage{xcolor}
-\usepackage{BOONDOX-cal}
+%\usepackage[cal=boondoxo]{mathalfa}
 %\graphicspath{{plots/}}
@@
 library("latex2exp")
-
+library("circmax")
 @
@@
 \frametitle{Motivation}
-\textbf{Goal:} Regression model for a circular response variable%, \\
+\textbf{Goal:} Regression model for a circular response variable.%, \\
 %\hspace{1.05cm} i.e. a periodic response variable measured on the unit circle.
@@
 $$
-f_\mathcal{vM}(y; \theta) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
+f_\text{vM}(y; \theta) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
 $$
@@
 $$
-\ell(\theta; y) =  \log(f_\mathcal{vM}(y;\theta)).
+\ell(\theta; y) =  \log(f_\text{vM}(y;\theta)).
 $$
@@
 \begin{enumerate}
-\item<3-> Fit global distributional model $\mathcal{vM}(Y; \theta)$: \\ % to the whole data set:\\
+\item<3-> Fit global distributional model $\text{vM}(Y; \theta)$: \\ % to the whole data set:\\
 Estimate $\hat{\theta}$ via maximum likelihood\\
@@
 \visible<2-3>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
 }
 \visible<4>{
-\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{vM}(Y;\hat{\theta}$)};
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\text{vM}(Y;\hat{\theta}$)};
 }
@@
 \visible<6-8>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {?};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
 \draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1);
@@
 \visible<9-10>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
 \draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
@@
 \visible<11>{
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{vM}(Y_1;\hat{\theta_1}$)};
-\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {$\mathcal{vM}(Y_2;\hat{\theta_2}$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$\text{vM}(Y_1;\hat{\theta}_1$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$\text{vM}(Y_2;\hat{\theta}_2$)};
 }
@@
 
-%\textbf{Goal:} Predict wind directions for the next hour at Innsbruck airport (Austria).
+\textbf{Response:} Wind direction in the next hour at Innsbruck airport (Austria).
 
-%\medskip
+\only<2>{
+%\vspace{-0.4cm}
+\begin{center}
+\setkeys{Gin}{width=0.65\linewidth}
+\includegraphics{ibk_airport.png}
+\end{center}
+}
 
-\textbf{Input:} Observation data of various meteorological quantities of the previous hour(s) from Innsbruck airport and 5 nearby weather stations.
+\only<3->{
+\medskip
+
+\textbf{Input:} Observation data of various meteorological quantities from Innsbruck airport and 5 nearby weather stations.
 
@@
 \item Air pressure
-\item \dots 
-\item \textit{FIXME: list of all included covariates}
+\item Humidity
+% "dd"   "ff"   "ffx"  "p"    "pred" "rf"   "tl"  
 \end{itemize}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{}
-%\end{center}
+
+\medskip
+
+\textbf{Variations:}
+Minima, maxima, changes over time or over location, \ldots
+
+\medskip
+
+\quad $\rightarrow$ 354 covariates
+}
 \end{frame}
-
-%\begin{frame}[fragile]
-%\includegraphics{airport_ibk.jpg}
-%\end{frame}
 
@@
 
-\textbf{Distributional tree model:}
+\textbf{Circular distributional tree:}
+
+\vspace{-1cm}
+
+\setkeys{Gin}{width=1.0\linewidth}
 
 \begin{center}
-%\vspace*{1cm}
-\setkeys{Gin}{width=\linewidth}
 \includegraphics{schlosser-circtree_plot.pdf}
@@
 
-\textbf{Data:} .. observations of response variables and .. covariates
-
-\medskip
-
-\textbf{Cross validation:} 
+\textbf{Evaluation setting:} 
 \begin{itemize}
-\item 5-fold setting 
-\item learning data: .. observations
-\item testing data: .. observations
+\item Data set of $40\ 000$ observations with 354 covariates
+\item 5-fold cross validation
+\item Results are compared based on circular CRPS (Grimit et al., 2006)
 \end{itemize}
-
-\medskip
-
-\textbf{Evaluation:} Results are compared based on circular CRPS.
 
@@
 
-boxplot CRPS cross validation
-%\begin{center}
+\begin{center}
 %\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{}
-%\end{center}
+\setkeys{Gin}{width=0.8\linewidth}
+\includegraphics{boxplot_crpsraw.pdf}
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Probabilistic Forecasting of Wind Directions}
+\begin{center}
+%\vspace*{-1cm}
+\setkeys{Gin}{width=0.8\linewidth}
+\includegraphics{boxplot_crpsskill.pdf}
+\end{center}
 \end{frame}
@@
 
-Breiman L (2001).
- \dquote{Random Forests.}
- \textit{Machine Learning}, 
- \textbf{45}(1), 5--32.
- \doi{10.1023/A:1010933404324}
-
-\medskip
+%Breiman L (2001).
+% \dquote{Random Forests.}
+% \textit{Machine Learning}, 
+% \textbf{45}(1), 5--32.
+% \doi{10.1023/A:1010933404324}
+% 
+%\medskip
 
@@
  \doi{10.2307/2532334}
+
+\medskip
+
+Grimit E P, Gneiting T, Berrocal V J, Johnson N A (2006).
+  \dquote{The Continuous Ranked Probability Score for Circular Variables and its Application to Mesoscale Forecast Ensemble Verification.}
+  \textit{Q.J.R. Meteorol. Soc.}, 
+  \textbf{132}, 2925--2942. 
+  \doi{10.1256/qj.05.235}
 
@@
 
-Mulder K, Klugkist I (2017).
- \dquote{Bayesian Estimation and Hypothesis Tests for a Circular Generalized Linear Model.}
- \textit{Journal of Mathematical Psychology}, 
- \textbf{80}, 4--14. 
- \doi{10.1016/j.jmp.2017.07.001}
-
-\medskip
+%Mulder K, Klugkist I (2017).
+% \dquote{Bayesian Estimation and Hypothesis Tests for a Circular Generalized Linear Model.}
+% \textit{Journal of Mathematical Psychology}, 
+% \textbf{80}, 4--14. 
+% \doi{10.1016/j.jmp.2017.07.001}
+%
+%\medskip
 

### 2019-07-04 pkg/disttree/inst/IWSM2019/slides.Rnw
small improvementes (adapted appearance order in tree theory slide, selected different example plot)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2639 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4687e5057b1e0198dc8d928d1985800d9811b222
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2640 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c3d79d7112537c5b373926c3178ebfc81431adb1

@@
 library("latex2exp")
-library("circmax")
+#library("circmax")
 @
@@
 
-\textbf{Application:} Probabilistic forecasting of wind directions for the next hour at Innsbruck airport (Austria, \textcolor{gray}{\url{https://innsbruck-airport.panomax.com}}).
+\textbf{Application:} Probabilistic forecasting of wind directions for the next hour at Innsbruck airport (Austria, \textcolor{gray}{\url{https://innsbruck-airport.panomax.com/}}).
 
@@
 \setkeys{Gin}{width=0.6\linewidth}
-\includegraphics{density_right.pdf}
+\includegraphics{density_left.pdf}
 %\end{center}
@@
 }
-\visible<9-10>{
+\visible<9>{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
@@
 }
-\visible<11>{
+\visible<10>{
 \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$\text{vM}(Y_1;\hat{\theta}_1$)};
@@
 }
-\visible<11->{
+\visible<10->{
 \draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
@@
 }
-\visible<12->{
+\visible<11->{
 \node[inner sep=0pt] (density2) at (-0.1,-0.3)
@@
 }
-\visible<12->{
+\visible<11->{
 \node[inner sep=0pt] (density3) at (1.7,-0.3)
@@
 \item<8-> Split along the covariate $X$ with strongest association or instability and\\ at breakpoint $p$ with highest improvement in log-likelihood.
-\item<10-> Repeat steps 1--3 recursively until some stopping criterion is met,\\
+\item<12-> Repeat steps 1--3 recursively until some stopping criterion is met,\\
 yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
@@
 
-\textbf{Applied models and reference models:}
+\textbf{Novel vs.\ reference models:}
 \begin{itemize}
@@
 \item 5-fold cross validation
-\item Results are compared based on circular CRPS (Grimit et al., 2006)
+\item Results are compared based on circular CRPS
 \end{itemize}
@@
 %\vspace{0.4cm}
-\textbf{Packages:} \emph{disttree} %and \emph{circmax} 
+\textbf{Software:} \emph{disttree} %and \emph{circmax} 
 available on R-Forge at\\
@@
   \dquote{The Continuous Ranked Probability Score for Circular Variables and its Application to Mesoscale Forecast Ensemble Verification.}
-  \textit{Q.J.R. Meteorol. Soc.}, 
+  \textit{Quarterly Journal of the Royal Meteorological Society}, 
   \textbf{132}, 2925--2942. 

### 2019-07-31 pkg/circtree/R/circtree.R
Initial commit for circtree package
- Copy of pkg 'circmax' in project 'uibk-rprog-2018' on R-Forge
- Revision Number 136


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2651 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e1aef9510366a6a334598836b2b86235129efefd

@@ -1,0 +1,110 @@
+## High-level convenience interface to mob() + circfit()
+circtree <- function(formula, data, start, subset, na.action, weights, offset,
+                     control = partykit::mob_control(), fit_control = circfit_control(...), ...){
+
+  ## Keep call
+  cl <- match.call(expand.dots = TRUE)
+
+  ## Data
+  if(missing(data)) data <- environment(formula)
+
+  ## Formula
+  oformula <- as.formula(formula)
+  formula <- Formula::as.Formula(formula)
+  if(length(formula)[2L] > 1L) {
+    formula <- Formula::Formula(formula(formula, rhs = 2L))
+    warning("formula must not have more than one RHS parts (only partitioning variables allowed)")
+  }
+
+  ## Call mob (with reordered arguments)
+  m <- match.call(expand.dots = FALSE)
+  morder <- match(c("formula", "data", "start", "subset", "na.action", "weights", "offset"), names(m), 0L)
+  m <- m[c(1L, morder)]
+  m$fit <- circfit
+  # FIXME: circfit is called with all default values. How can I change estfun and object as user 
+    # (tried control arguments but then two parameters were in the function call, as circfit is called with all defaults) ?!
+  m$formula <- formula
+  m$control <- control
+  m$fit_control <- fit_control
+  #for(n in names(control)) if(!is.null(control[[n]])) m[[n]] <- control[[n]]
+  if("..." %in% names(m)) m[["..."]] <- NULL
+  m[[1L]] <- as.call(quote(partykit::mob))
+  rval <- eval(m, parent.frame())
+  
+  ## Extend class and keep original call
+  rval$info$call <- cl
+  rval$info$fit_control <- fit_control
+  class(rval) <- c("circtree", class(rval))
+  return(rval)
+}
+
+
+## Control function for circfit
+circfit_control <- function(solve_kappa = "Newton-Fourier", useC = FALSE, ncores = 1,...) {
+  ctrl <- c(
+    list(solve_kappa = solve_kappa, useC = useC, ncores = ncores), 
+    list(...)
+  )
+  ctrl
+}
+
+
+## Print method
+print.circtree <- function(x, title = "Circular distribution tree, employing the Von Mises Distribution", 
+                           objfun = "negative log-likelihood", ...) {
[…]

### 2019-07-31 pkg/circtree/R/crps_vonmises.R

@@ -1,0 +1,122 @@
+# -------------------------------------------------------------------
+# - NAME:   crps_vonmises.R
+# - AUTHOR: Moritz N. Lang, Lisa Schlosser
+# - DATE:   2019-05-21
+# -------------------------------------------------------------------
+# - PURPOSE: Circular CRPS (von Mises) based on numeric integration using
+#            the charististic equation
+# -------------------------------------------------------------------
+# - L@ST MODIFIED: 2019-07-31 on thinkmoritz
+# -------------------------------------------------------------------
+
+### Function
+#crps_vonmises <- function(y, mu, kappa) {
+#
+#  require(CharFun)
+#  
+#  int_fun <- function(x) { 
+#    1 / pi * abs(cfC_vonMises(x, mu, kappa) - exp((0+1i) * x * y))^2 / x^2
+#  } 
+#    
+#  rval <- integrate(int_fun, 0, Inf)$value
+#  return(rval)
+#}
+
+##YEAR: 2017
+##COPYRIGHT HOLDER: Ludmila Simkova
+## Function cfC_vonMises() from Package 'CharFun'
+cfC_vonMises <- function(t, mu = 0, kappa = 1){
+  szt <- dim(t)
+  t <- c(t)
+  cf <- unlist(lapply(t, function(t) tryCatch((Bessel::BesselI(kappa, 
+      abs(t), TRUE)/Bessel::BesselI(kappa, 0, TRUE)) * exp((0+1i) * 
+      t * mu), error = function(e) 0)))
+  cf[t == 0] <- 1
+  dim(cf) <- szt
+  return(cf)
+}
+
+
+
+## Function
+crps_vonmises <- function(y, mu, kappa, sum = FALSE) {
+
+  if(any(y < -pi) || any(y > pi) || any(mu < -pi) || any(mu > pi) || any(kappa < 0 ))
+    stop("y and mu must be in the interval of [-pi, pi], and kappa must be non negative!") 
+
+  if(!inherits(y, c("numeric", "integer")) || !inherits(mu, c("numeric", "integer")) ||
+    !inherits(kappa, c("numeric", "integer"))) {
+    stop("Input 'y', 'mu', and 'kappa' must be numeric vectors...")
[…]

### 2019-07-31 pkg/circtree/tests/testthat/test_Croutines.R

@@ -1,0 +1,77 @@
+context("C routines for kappa estimation")
+
+## Test results
+test_that("computed values are correct", {
+  expect_equal(
+    solve_kappa_Newton_Fourier(0.1,100,useC = FALSE),
+    solve_kappa_Newton_Fourier(0.1,100,useC = TRUE)
+  )
+  expect_equal(
+    .Call("Ginv", 0.1, 0.2, 0.3),
+    movMF:::Ginv(0.1,0.2,0.3)
+  )
+  expect_equal(
+    .Call("Rinv_lower_Amos_bound", 0.3, 0.3),
+    movMF:::Rinv_lower_Amos_bound(0.3, 0.3)
+  )
+  expect_equal(
+    .Call("Rinv_upper_Amos_bound", 0.3, 0.3),
+    movMF:::Rinv_upper_Amos_bound(0.3, 0.3)
+  )
+  expect_equal(
+    .Call("Rinv_upper_Amos_bound", 0.3, 2),
+    movMF:::Rinv_upper_Amos_bound(0.3, 2)
+  )
+  expect_equal(
+    .Call("A_PCF", 0.3, 1.1),
+    movMF:::A(0.3, 1.1)
+  )
+  expect_equal(
+    .Call("A_PCF", 0.3, 2),
+    movMF:::A(0.3, 2)
+  )
+  expect_equal(
+    .Call("Aprime_PCF", 0.1, 1.1, 0.3),
+    movMF:::Aprime(0.1, 1.1, A = 0.3)
+  )
+  expect_equal(
+    .Call("Aprime_PCF", 0.1, 0.1, 2),
+    movMF:::Aprime(0.1, 0.1, A = 2)
+  )
+})
+
+## Small helper function
+compare_times <- function(t_R, t_C, label = "R vs C", output = c("text", "numeric")) {
+    output <- match.arg(output)
+    t_R <- as.numeric(t_R[3L], units = "secs")
+    t_C <- as.numeric(t_C[3L], units = "secs")
+    inc <- 100 * (t_R/t_C - 1);
+    if(output == "text"){
+      cat(sprintf("\nTiming for \"%s\"\n", label))
+      cat(sprintf("   R:  %10.5f sek   C: %10.5f sek   (%4.1f percent %s)\n",
+                  t_R, t_C, inc, ifelse(inc < 0, "slower in C", "faster in C")))
+    } else {
+      inc
+    }
+}
+
+## Test performance
+r <- rnorm(5)
[…]

### 2019-07-31 pkg/circtree/tests/testthat/test_Csdist.R

@@ -1,0 +1,86 @@
+context("C routines for sdist")
+
+## Small helper function
+compare_times <- function(t_R, t_C, label = "R vs C", output = c("text", "numeric")) {
+    output <- match.arg(output)
+    t_R <- as.numeric(t_R[3L], units = "secs")
+    t_C <- as.numeric(t_C[3L], units = "secs")
+    inc <- 100 * (t_R/t_C - 1);
+    if(output == "text"){
+      cat(sprintf("\nTiming for \"%s\"\n", label))
+      cat(sprintf("   R:  %10.5f sek   C: %10.5f sek   (%4.1f percent %s)\n",
+                  t_R, t_C, inc, ifelse(inc < 0, "slower in C", "faster in C")))
+    } else {
+      inc
+    }
+}
+
+## Set up families functions
+dR  <- dist_vonmises(useC = FALSE, ncores = 1)
+dC  <- dist_vonmises(useC = TRUE, ncores = 1)
+dC2 <- dist_vonmises(useC = TRUE, ncores = 2)
+dC5 <- dist_vonmises(useC = TRUE, ncores = 5)
+
+## Test results
+set.seed(111) 
+eta <- c(tan(0.8/2), log(0.8))
+n <- 1000000
+y <- rnorm(n)
+
+dR_out <- dR$sdist(y, eta)
+dC_out <- dC$sdist(y, eta)
+C2_out <- dC2$sdist(y, eta)
+C5_out <- dC5$sdist(y, eta)
+
+test_that("scores (sum = TRUE) are the same", {
+  expect_equal(
+    dR_out,
+    dC_out
+  )
+  expect_equal(
+    dC_out,
+    C2_out
+  )
+  expect_equal(
+    C2_out,
+    C5_out
+  )
+})
+
+dR_out <- dR$sdist(y, eta, sum = TRUE)
+dC_out <- dC$sdist(y, eta, sum = TRUE)
+C2_out <- dC2$sdist(y, eta, sum = TRUE)
+C5_out <- dC5$sdist(y, eta, sum = TRUE)
+
+test_that("scores (sum = FALSE) are the same", {
+  expect_equal(
+    dR_out,
+    dC_out
+  )
+  expect_equal(
+    dC_out,
+    C2_out
+  )
+  expect_equal(
+    C2_out,
+    C5_out
+  )
+})
+
+
+## Test performance 
+t_R  <- system.time(dR$sdist(y, eta))
[…]

### 2019-09-18 pkg/circtree/inst/wind_paper/draft/draft.Rnw
first draft of wind paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2763 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c0dc031b8e012486a74ce9965691ac0598821ac0

\documentclass[nojss]{jss}

%% packages
\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}

%% tikz
\usepackage{array,makecell,tikz,color,soul}
\usetikzlibrary{arrows.meta,positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}

%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}

%% commands
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

\definecolor{changes}{rgb}{1,1,0.7}
\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}

\numberwithin{equation}{subsection}

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("disttree")
library("circtree")

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)

# set function for parallelization
applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))

@

\title{Distributional Trees and Forests for Circular Data}
%\Shorttitle{Distributional Trees and Forests for Circular Data}

\author{Moritz N. Lang\\Universit\"at Innsbruck
   \And Lisa Schlosser\\Universit\"at Innsbruck
   \And Torsten Hothorn\\Universit\"at Z\"urich
   \AND Georg~J.~Mayr\\Universit\"at Innsbruck
   \And Reto Stauffer\\Universit\"at Innsbruck
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{M. N. Lang, L. Schlosser, T. Hothorn, 
G. J. Mayr, R. Stauffer, A. Zeileis}

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabelize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at Innsbruck Airport based on a set 
of meteorological measurements from Innsbruck and surrounding weather stations.
}
\Keywords{Distributional Trees; Circular Response; Von Mises Distribution}

\Address{
  Moritz N. Lang, Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
  Universit\"at Innsbruck \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Moritz.Lang@uibk.ac.at}, \email{Lisa.Schlosser@uibk.ac.at},\\
  \email{Reto.Stauffer@uibk.ac.at},\email{Achim.Zeileis@R-project.org} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/moritz-lang/},\\
  \phantom{URL: }\url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
  \phantom{URL: }\url{https://retostauffer.org/},\\
  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\

  Georg~J.~Mayr\\
  Universit\"at Innsbruck \\
  Department of Atmospheric and Cryospheric Science \\
  Faculty of Geo- and Atmospheric Sciences \\
  Innrain~52f \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Georg.Mayr@uibk.ac.at} \\
  URL: \url{http://acinn.uibk.ac.at/persons/georg_mayr}\\

  Torsten Hothorn\\
  Universit\"at Z\"urich \\
  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
  Hirschengraben 84\\
  CH-8001 Z\"urich, Switzerland \\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://user.math.uzh.ch/hothorn/}\\

}

\begin{document}

\section{Introduction}
\label{sec:introduction}

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, animal movement direction or 
gene-structure in biology, and wind direction as one of the most important weather 
variables in meteorology.
Circular regression models were first introduced by \cite{Fisher+Lee:1992} and 
further extended by \cite{Jammalamadaka+Sengupta:2001} and 
\cite{Mulder+Klugkist:2017} among others.
While most of the already existing approaches are built on additive regression models, 
we propose an adaption of regression trees to circular data by employing 
distributional trees.

\section{Probabilistic Circular Modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle applying the modulo operation on the 
interval length $2\pi$ \citep{Mardia+Jupp:2009}.
In that way, for example the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in [0, 2\pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

The corresponding log-likelihood function is defined by
\begin{equation}
\begin{aligned}
\ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
&= -\log(2 \pi I_0(\kappa)) +  \kappa \cos(y - \mu).
\end{aligned}
\end{equation}

Once an appropriate distribution family is selected fitting a probabilistic model corresponds to
estimating the distribution parameters. Thus, when applying the von Mises distribution to fit a 
probabilistic model $\text{vM}(Y, \mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ 
the distribution parameters $\mu$ and $\kappa$ are to be estimated.
For a learning sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by 
maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
\begin{equation}\label{eq:optim}
(\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i),
\end{equation}
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that a fully specified 
distributional model is fit to the learning data.

Employing the score function
\begin{equation}
\begin{aligned}
s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
\smallskip &,
\quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
\\
&= \bigg(\kappa \sin(y-\mu)
\smallskip &,
\quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
\end{aligned}
\end{equation}
the optimization problem in Equation~\ref{eq:optim} can also be defined by 
%setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
\begin{equation}
\sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand).}
%TODO: refer to data, Innsbruck airport
\label{fig:density_vM} 
\end{figure}

Figure~\ref{fig:density_vM} illustrates a fitted distributional model employing
the von Mises distribution for observed wind direction data from the airport of Innsbruck.
This example shows that reasonably well fitting probabilistic models can
be obtained for circular data by specifying an appropriate distribution family 
such as the von Mises distribution and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Regression Trees and Forests for Circular Data}

\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
While this strategy works well for smooth effects difficulties can arise in case of 
non-additive changes. Moreover, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
One way how to deal with these problems is to apply a tree-structured model.

\subsection{Circular Distributional Tree}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
In the presented distributional trees a full distributional model is specified in each node of the 
tree. For this purpose a distribution family has to be chosen in advance. Hence, selecting the 
von Mises distribution allows for an application of distributional trees to circular data following 
the tree building algorithm as introduced by \cite{Schlosser+Hothorn+Stauffer:2019}.

In particular, in each node the crucial decision on how and where to split the data is based on model scores which are obtained by evaluating the score function $s$ at the individual observations and 
parameter estimates. For the von Mises distribution and a data set of $n$ observations
this yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

%In particular, starting with the full learning sample, the steps for building a circular 
%distributional tree are:
%\begin{enumerate}
%\item Estimate the parameter pair $(\hat{\mu}, \hat{\kappa})$ via maximum likelihood for the observations
%  in the current (sub)sample.
%\item Test for associations (or instabilities) of the scores $s(\hat{\mu}, \hat{\kappa}, y_i)$ 
%  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
%\item Split the sample along the partitioning variable $Z_l^*$ with the
%  strongest association or instability. Choose the breakpoint leading to the highest
%  improvement in model fit.
%  %with the highest improvement in the log-likelihood or the highest discrepancy.
%\item Repeat steps 1--3 recursively in the subsamples until these become too
%  small or there is no significant association/instability (or some other
%  stopping criterion is reached).
%\end{enumerate}

%To test for associations between scores and covariates in step~3 different strategies can be applied
%such as permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} or asymptotic M-fluctuation
%tests for parameter instability \citep{Zeileis+Hornik:2007}. In order to select the covariate with the %highest significant association the one covariate that corresponds to the lowest significant $p$-value (if %any) is chosen for splitting the data. Within the range of this split variable the split point leading to %the highest improvement in model fit is chosen either by optimizing the log-likelihood \citep[as in the %MOB algorithm,][]{Zeileis+Hothorn+Hornik:2008} or a two-sample test statistic  over all possible partition%s \citep[as in the CTree algorithm][]{Hothorn+Hornik+Zeileis:2006}.

Once a circular distributional tree model is fit it can be employed to obtain probabilistic predictions
for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node depending on the values of the covariates and in that way to the corresponding estimated parameter pair $(\hat{\mu}, \hat{\kappa})$ of this node. Therefore, the same parameter pair is estimated for all observations
ending up in the same terminal node. This facilitates the application as the parameter estimates do not need to be recalculated for each (new) observation via maximum likelihood but can be extracted directly
from the learning sample and the fitted model. This can also be expressed by employing weights

\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

While tree models can capture non-additive effects they do not allow for purely smooth effects and therefore might impose abrupt changes even in case of rather smooth transitions. This can be avoided
by combining an ensemble of trees in order to obtain a more stable forest model.

\subsection{Circular Distributional Forest}
\label{sec:circforest}
A natural extension of circular distributional trees are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions
by adaptive local likelihood estimation of the distributional parameters
\citep{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on how often $\bm{z}$ is assigned to the same terminal node as
each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

In that way a specific set of weights can be calculated for each observation yielding
its specific parameter estimates
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the unweighted
full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$ corresponds to the abrupt
splits from the tree.
Thus circular distributional forests can capture both, smooth and abrupt changes, while
covariates and possible interactions are selected automatically allowing for an easy to
apply methodology to fully specify the von Mises distribution for circular data.

\section{Wind Direction Forecasts}
Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production.

%% TODO: available approaches to wind direction forecasting

This study employs the above introduced methodology of circular distributional
trees and forests to obtain hourly wind direction forecasts at Innsbruck Airport.
Innsbruck lies at the bottom of a deep valley in the Alps. Topography
channels wind along the west-east valley axis or along a tributary valley
intersecting from the south. Hence, pressure gradients to which valley wind
regimes react both west and east of the airport are considered as covariates
along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

\subsection{Data}
Observations of various meteorological quantities from the airport of Innsbruck and .. surrounding
weather stations are used as covariates in order to predict the wind direction at Innsbruck
airport in the following hour.
In particular .. covariates are included in the model. Next to the current wind direction,
temperature, air pressure, ... at the each of the stations also differences in these quantities
between the considered stations are included. For example, differences in air pressure
between Innsbruck and either a western or eastern location are expected to have a high influence 
on wind directions.

\subsection{Reference Models}
Next to circular distributional trees and forests three other models have been applied and their 
performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).

\paragraph{Persistence}
Assuming that weather conditions will remain the same in the following hours/days is the
basic idea of persistence in meteorology. Hence, fitting a distributional model as presented in
Section~\ref{sec:prob_circ} to the observations of the response variable from the previous hour(s) 
contained in the learning data provides a distributional approach to persistence.

\paragraph{Climatology}
If there are no climatological changes it can be expected that for a selected day of the year weather conditions will always be the same. Therefore, in order to predict wind direction for a specified day
and time observations for this selected timespace from all years contained in the learning data are
used to fit a distributional model.

\paragraph{Linear Model}
Another commonly applied model ...

\subsection{Results}

\begin{figure}[t]
\centering
%\includegraphics[height = .5\textheight,angle=90,origin=c]{schlosserfig1.pdf}
\includegraphics[width = \textwidth]{tree_ibk.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

Figure~\ref{fig:tree_ibk} depicts the resulting distributional tree, including
both the empirical (gray) and fitted von Mises (red) distribution of wind direction
in each terminal node. Based on the fitted location parameters $\hat \mu$, the subgroups
can be distinguished into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley
(from east to west, nodes 4 and 5). (2) Downslope winds blowing across
the Alpine crest along the intersecting valley towards Innsbruck (from south-east to
north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
of the valley mouth (from west to east, nodes 12, 14, and 15). Node~11
captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.

\section{Discussion and outlook}

Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is particularly
useful for modeling wind direction in mountainous terrain where wind shifts
can occur due to turns of the pressure gradients along a valley.

\subsection{Splits in circular covariates}

In order to obtain more parsimonious and more stable trees another possible
extension for \emph{circular covariates} (with or without a \emph{circular response})
is to consider their circular nature when searching the best split into two segments.
In general, searching the best separation of a covariate into a ``left'' and ``right''
daughter node tries to maximize the segmented log-likelihood:
\begin{equation}
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
\end{equation}
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a circular covariate $\in [0, 2 \pi)$
only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi)$, encompassing zero. The latter
strategy is invariant to the (often arbitrary) definition of the direction at zero.

When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
far away, a simple linear split typically suffices to capture such a split (as seen for the
lagged wind direction in Figure~\ref{fig:tree_ibk}). If both $\nu$ and $\tau$ differ
clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
might miss such a pattern.

The required test statistic to maximally select two split points simultaneously is straightforward
to accommodate in the CTree framework by providing all binary indicators corresponding
to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
increasingly slow for larger sample sizes but it might be possible to speed up computations by
exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
MOB framework the extension is not quite as straightforward but one strategy could be to
adapt double maximum tests \`a la \cite{Bai+Perron:2003}.

Hence, the splitting idea can be naturally extended to a two-point search, however, for 
an unbiased and inference-based selection the corresponding testing strategies might need
further adaption.

\section*{Acknowledgments}
This project was partially funded by the Austrian Research Promotion 
Agency~(FFG) grant number~$858537$.

Torsten Hothorn received funding from the Swiss National Science
Foundation, grant number~$200021\_184603$.

\section*{Computational details}

The proposed methods are in the \textsf{R} package \textbf{circtree}
(version~\Sexpr{packageVersion("circtree")}) based on the \textbf{disttree}
package (version~\Sexpr{packageVersion("disttree")}) which applies the main
tree building function from the  \textbf{partykit} package 
(version~\Sexpr{packageVersion("partykit")}), all three available 
on \textsf{R}-Forge at
(\url{https://R-Forge.R-project.org/projects/partykit/}).

%\newpage
%\section*{Appendix}
%\begin{appendix}
%\end{appendix}

%\newpage
\bibliography{ref.bib}

\end{document}

### 2019-09-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small revisions on methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2764 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f2129efc5b46c474e76a5dc39bf2784b5d30a4f9

@@
 Generally, this feature can be obtained by ``wrapping'' the probability density function
-of any continuous distribtuion around the unit circle applying the modulo operation on the 
-interval length $2\pi$ \citep{Mardia+Jupp:2009}.
-In that way, for example the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
+of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
+applying the modulo operation on the interval length $2\pi$.
+In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
 A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
@@
 particularly for a high number of covariates and no information about possible interactions.
-One way how to deal with these problems is to apply a tree-structured model.
+An alternative approach that can deal with these problems is to apply a tree-structured model.
 
@@
 this yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
-a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
+each given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
 respectively.
@@
 for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
-Starting at the root node the tree structure leads the observation to a terminal node depending on the values of the covariates and in that way to the corresponding estimated parameter pair $(\hat{\mu}, \hat{\kappa})$ of this node.
-Therefore, the same parameter pair is estimated for all observations
-ending up in the same terminal node.
-This facilitates the application as the parameter estimates do not need to be recalculated for each (new) observation via maximum likelihood but can be extracted directly
-from the learning sample and the fitted model. This can also be expressed by employing weights
-
+Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
+the corresponding subset of learning observations. 
+This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
+node:
 \begin{equation}
@@
 \end{equation}
+
+Therefore, the same parameter pair is estimated for all observations
+belonging to the same terminal node.
+This facilitates the application as the parameter estimates do not need to be recalculated for each (new) observation via maximum likelihood but can be extracted directly
+from the learning sample and the fitted model.
 

### 2019-09-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small revisions on methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2765 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 38da6d9010bf8be5be44b393f8bcea77bddff7b8

@@
 
-\subsection{Circular Distributional Tree}
+\subsection{Distributional Trees for Circular Data}
 \label{sec:circtree}
@@
 
-Once a circular distributional tree model is fit it can be employed to obtain probabilistic predictions
+Once a distributional tree model is fit it can be employed to obtain probabilistic predictions
 for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
@@
 
-\subsection{Circular Distributional Forest}
+\subsection{Distributional Forests for Circular Data}
 \label{sec:circforest}
-A natural extension of circular distributional trees are ensembles or forests 
+A natural extension of distributional trees for circular data are ensembles or forests 
 that can improve forecasts by regularizing and stabilizing the model.
@@
 %\citep{Lin+Jeon:2006}
-is obtained that is based on how often $\bm{z}$ is assigned to the same terminal node as
-each learning observation $y_i, i \in \{1,\ldots,n\}$.
+is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
+node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
 Hence, for a forest of $T$ trees the weights are calculated via
@@
 In that way a specific set of weights can be calculated for each observation yielding
-its specific parameter estimates
+its specific parameter estimates for the von Mises distribution
 \begin{equation}
@@
 splits from the tree.
-Thus circular distributional forests can capture both, smooth and abrupt changes, while
+Thus distributional forests for circular data can capture both, smooth and abrupt changes, while
 covariates and possible interactions are selected automatically allowing for an easy to
@@
 \subsection{Reference Models}
-Next to circular distributional trees and forests three other models have been applied and their 
+Next to distributional trees for circular data and forests three other models have been applied and their 
 performances have been compared based on a circular version of the continuous ranked probability 

### 2019-09-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
restructured application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2766 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e6b118f6c4b11cbf173bf9802a19e401325a85bb

@@
 \section{Wind Direction Forecasts}
+
+\subsection{Motivation}
 Wind is a classical circular quantity and accurate forecasts of wind direction
@@
 Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
-
-\subsection{Data}
-Observations of various meteorological quantities from the airport of Innsbruck and .. surrounding
-weather stations are used as covariates in order to predict the wind direction at Innsbruck
-airport in the following hour.
-In particular .. covariates are included in the model. Next to the current wind direction,
-temperature, air pressure, ... at the each of the stations also differences in these quantities
-between the considered stations are included. For example, differences in air pressure
-between Innsbruck and either a western or eastern location are expected to have a high influence 
-on wind directions.
-
-\subsection{Reference Models}
-Next to distributional trees for circular data and forests three other models have been applied and their 
-performances have been compared based on a circular version of the continuous ranked probability 
-score (CRPS).
-
-\paragraph{Persistence}
-Assuming that weather conditions will remain the same in the following hours/days is the
-basic idea of persistence in meteorology. Hence, fitting a distributional model as presented in
-Section~\ref{sec:prob_circ} to the observations of the response variable from the previous hour(s) 
-contained in the learning data provides a distributional approach to persistence.
-
-\paragraph{Climatology}
-If there are no climatological changes it can be expected that for a selected day of the year weather conditions will always be the same.
-Therefore, in order to predict wind direction for a specified day
-and time observations for this selected timespace from all years contained in the learning data are
-used to fit a distributional model.
-
-\paragraph{Linear Model}
-Another commonly applied model ...
-
-\subsection{Results}
 
@@
 speed separate between subgroups with high vs.\ low precision.
+
+\subsection{Data}
+Observations of various meteorological quantities from the airport of Innsbruck and .. surrounding
+weather stations are used as covariates in order to predict the wind direction at Innsbruck
+airport in the following hour.
+In particular .. covariates are included in the model. Next to the current wind direction,
+temperature, air pressure, ... at the each of the stations also differences in these quantities
+between the considered stations are included. For example, differences in air pressure
+between Innsbruck and either a western or eastern location are expected to have a high influence 
+on wind directions.
+
+\subsection{Models and Evaluation}
+Next to distributional trees for circular data and forests three other models have been applied and their 
+performances have been compared based on a circular version of the continuous ranked probability 
+score (CRPS).
+
+\paragraph{Persistence}
+Assuming that weather conditions will remain the same in the following hours/days is the
+basic idea of persistence in meteorology. Hence, fitting a distributional model as presented in
+Section~\ref{sec:prob_circ} to the observations of the response variable from the previous hour(s) 
+contained in the learning data provides a distributional approach to persistence.
+
+\paragraph{Climatology}
+If there are no climatological changes it can be expected that for a selected day of the year weather conditions will always be the same.
+Therefore, in order to predict wind direction for a specified day
+and time observations for this selected timespace from all years contained in the learning data are
+used to fit a distributional model.
+
+\paragraph{Linear Model}
+Another commonly applied model ...
+
+Compare fitted models based on a circular version of CRPS ...
+
+\subsection{Results}
+Ibk/Vie, 1h/3h
 

### 2019-10-03 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision, particularly on introduction, reference models and motivational example in the application section; added further TODOs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2784 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0002dfac5c8c9b2ebc8d2514568fb92c067fd839

@@
 gene-structure in biology, and wind direction as one of the most important weather 
-variables in meteorology.
-Circular regression models were first introduced by \cite{Fisher+Lee:1992} and 
-further extended by \cite{Jammalamadaka+Sengupta:2001} and 
-\cite{Mulder+Klugkist:2017} among others.
-While most of the already existing approaches are built on additive regression models, 
-we propose an adaption of regression trees to circular data by employing 
-distributional trees.
+variables in meteorology. Fitting a statistical model to this type of data requires
+the incorporation of its specific feature of periodicity, i.e., in case of angular
+data the fact that values are restricted to an interval such as $[0,2\pi)$ with
+$0$ being equivalent to $2\pi$.
+
+In many approaches to model circular data it is assumed that the circular variable of interest 
+follows a circular distribution, in particular the von Mises distribution (which will be discussed 
+in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
+models for circular data were built on as they regress the distribution parameter(s) on covariates.
+One of the first regression models with a circular response variable and linear covariates
+was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
+the circular mean, hence the location parameter of the von Mises distribtuion. 
+\cite{Johnson+Wehrly:1978} took this idea one step further 
+by plugging in a link function such that the values of the linear predictor are transformed 
+to an restricted interval of length $2\pi$. 
+This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
+allowing for the mean direction or the dispersion to be modeled not only in separate models 
+but also in a mixed model. However, applying these models has shown to be very challenging, 
+mainly due to the complexity encountered in optimizing the log-likelihood function.
+Moreover the named approaches and also its further developments 
+\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
+are built on additive regression models which may not be able to capture all types of effects.
+For example abrupt changes might not be detected properly owing to the fixed structure enforced
+by model assumptions such as the type of link function and the type of linear combination.
+Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
+distributional trees and distributional forests. 
+In that way no model assumptions are required as covariates and their possible interactions are
+selected automatically. Non-additive effects are captured by the tree structure while combining an ensemble
+of trees to a forest model allows for modeling smooth changes as well. 
+Therefore, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
+These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
+distributional models for circular data in Section~\ref{sec:prob_circ}.
+In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
+at the airports Innsbruck and Vienna.
 
@@
 \section{Regression Trees and Forests for Circular Data}
-
+\label{sec:tree_forest}
 \subsection{Motivation}
@@
 \section{Wind Direction Forecasts}
-
+\label{sec:wind}
 \subsection{Motivation}
@@
 
-This study employs the above introduced methodology of circular distributional
-trees and forests to obtain hourly wind direction forecasts at Innsbruck Airport.
-Innsbruck lies at the bottom of a deep valley in the Alps. Topography
-channels wind along the west-east valley axis or along a tributary valley
-intersecting from the south. Hence, pressure gradients to which valley wind
-regimes react both west and east of the airport are considered as covariates
-along with other meteorological measurements at the airport (lagged by one hour),
+As a motivational example Figure~\ref{fig:tree_ibk} illustrates a distributional
+tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
+Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
+west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
+gradients to which valley wind regimes react both west and east of the airport are considered 
+as covariates along with other meteorological measurements at the airport (lagged by one hour),
 such as wind direction and wind speed at Innsbruck Airport.
-Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
+
+Next to the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
+and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
+fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
+wind regimes:
+(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
+east to west, nodes 4 and 5). 
+(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
+Innsbruck (from south-east to north-west, nodes 7 and 8). 
+(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
+nodes 12, 14 and 15). 
+Node~11 captures observations with rather low wind speeds that cannot be distinguished
+clearly into wind regimes and consequently are associated with a very low
+estimated concentration $\hat \kappa$. In terms of covariates, the lagged
+wind direction (``persistence'') is mostly responsible for distinguishing
+the broad wind regimes listed above while the pressure gradients and wind
+speed separate between subgroups with high vs.\ low precision.
+Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
+increases clockwise from North ($0$ degree).
 
@@
 
-Figure~\ref{fig:tree_ibk} depicts the resulting distributional tree, including
-both the empirical (gray) and fitted von Mises (red) distribution of wind direction
-in each terminal node. Based on the fitted location parameters $\hat \mu$, the subgroups
-can be distinguished into the following wind regimes:
-(1)~Up-valley winds blowing from the valley mouth towards the upper valley
-(from east to west, nodes 4 and 5). (2) Downslope winds blowing across
-the Alpine crest along the intersecting valley towards Innsbruck (from south-east to
-north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
-of the valley mouth (from west to east, nodes 12, 14, and 15). Node~11
-captures observations with rather low wind speeds that cannot be distinguished
-clearly into wind regimes and consequently are associated with a very low
-estimated concentration $\hat \kappa$. In terms of covariates, the lagged
-wind direction (``persistence'') is mostly responsible for distinguishing
-the broad wind regimes listed above while the pressure gradients and wind
-speed separate between subgroups with high vs.\ low precision.
+Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
+a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
+area, is considered in this study.
+In that way completely different geographic features and their possible influences
+on wind direction are included which allows to test for a high flexibility of the applied methods.
+For the two locations the above introduced methodology of circular distributional trees and forests is
+employed to wind direction data and additional meteorological measurements in order to obtain hourly 
+wind direction forecasts, separately for Innsbruck and Vienna.
 
 \subsection{Data}
-Observations of various meteorological quantities from the airport of Innsbruck and .. surrounding
-weather stations are used as covariates in order to predict the wind direction at Innsbruck
-airport in the following hour.
-In particular .. covariates are included in the model. Next to the current wind direction,
-temperature, air pressure, ... at the each of the stations also differences in these quantities
-between the considered stations are included. For example, differences in air pressure
-between Innsbruck and either a western or eastern location are expected to have a high influence 
-on wind directions.
+For both airports observations of various meteorological quantities measured
+either directly at the airports or at surrounding weather stations are used as 
+covariates in order to predict the wind direction at the corresponding airport 
+in the following hour.
+Next to the current wind direction, temperature, air pressure, humidity, ... at each 
+of the weather stations also differences in these quantities between the considered 
+stations are included. For example, differences in air pressure between Innsbruck and 
+either a western or eastern location are expected to have a high influence on wind 
+directions. Moreover, changes of these quantities over time are included as well as 
+daytime. In total, 550 covariates are provided for Innsbruck and 522 covariates for Vienna. 
+A list of all covariates together with their abbreviations as used in this paper
+can be found in Table~\ref{tab:covariates}.
+The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
+After eliminating observations with at least one missing value a total number of 26861 time points
+for Innsbruck and 33853 for Vienna are available.
+
+\begin{table}[t!]
+%\resizebox{1.4\textwidth}{!}{
+\begin{minipage}{\textwidth}
+\caption[Table caption text]{Basic covariates together with the type of variations.}
+\label{tab:covariates}
+\begin{tabular}{l  l}
+\hline
+Basic covariates & Variations\\
+\hline
+\emph{dd}: wind direction       & different locations, \\
+                                & changes over time(?)\\
+\emph{ff}: wind speed           & different locations, \\
+                                & changes over time(?), \\
+                                & (hourly) maxima\\
+\emph{p}: pressure              & different locations, \\
+                                & differences between locations, \\
+                                & changes over time(?)\\
+\emph{t}: temperature           & different locations, \\
+                                & changes over time(?)\\
+\emph{rf}: relative humidity    & different locations\\
+\hline
+\end{tabular}
+\end{minipage} 
+%}
+\end{table}
 
 \subsection{Models and Evaluation}
-Next to distributional trees for circular data and forests three other models have been applied and their 
-performances have been compared based on a circular version of the continuous ranked probability 
+Along with distributional trees for circular data and forests three other models have been applied 
+and their performances have been compared based on a circular version of the continuous ranked probability 
 score (CRPS).
@@
 \paragraph{Persistence}
-Assuming that weather conditions will remain the same in the following hours/days is the
-basic idea of persistence in meteorology. Hence, fitting a distributional model as presented in
-Section~\ref{sec:prob_circ} to the observations of the response variable from the previous hour(s) 
-contained in the learning data provides a distributional approach to persistence.
+Assuming that weather conditions will remain the same after a certain period of time is the
+basic idea of persistence in meteorology. While this simple way of making predictions can
+be very effective particularly under stable atmospheric circumstances it can fail remarkably
+as soon as any changes take place, especially in case of rapid changes.
+Therefore, it heavily depends on the specific features and the variability of the variable
+of interest whether this approach can provide overall reliable predictions.
+% TODO: references
+To establish a distributional model of persistence for this study a set of six observations
+is considered to make a prediction for a selected day and time point. In particular,
+the exact same time points of the previous six days are considered. In order to account
+for a stronger influence of observations closer to the prediction time point exponentially
+increasing weights are included with the highest weight being assigned to the time point of the
+previous day. For this selected set of observations the distribution parameters of the von Mises 
+distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
+parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.
 
 \paragraph{Climatology}
-If there are no climatological changes it can be expected that for a selected day of the year weather conditions will always be the same.
-Therefore, in order to predict wind direction for a specified day
-and time observations for this selected timespace from all years contained in the learning data are
-used to fit a distributional model.
+If there are no climatological changes within a certain number of years it can be expected that 
+for a selected day of the year weather conditions will always be the same over all years 
+considered.
+Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
+meteorological variable of interest that can be very useful but its performance can also suffer 
+dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.
+% TODO: references
+Similar to the applied persistence model observations are selected to fit a distributional climatology
+model. For a specified day and time point observations from the exact same time point and day but all 
+other years contained in the learning data are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.
 
-\paragraph{Linear Model}
-Another commonly applied model ...
+\paragraph{Generalized Linear Model}
+Another commonly applied approach is to regress the parameter(s) of the von Mises distribution
+on covariates by a linear combination of selected covariates connected to the parameter(s) via
+a specified link function. The toolbox of generalized linear models ...
+% TODO: references
 
+% TODO: CRPS
+\vspace{0.5cm}
 Compare fitted models based on a circular version of CRPS ...

### 2019-10-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small revisions on reference models and structure of application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2786 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5bfb7c033c7e056ec72f8786c730574c2280d3dd

@@
 forest which allows for modeling smooth effects as well.
-For illustration, hourly wind direction forecasts are obtained at Innsbruck Airport based on a set 
-of meteorological measurements from Innsbruck and surrounding weather stations.
+For illustration, hourly wind direction forecasts are obtained at two airports in Austria
+(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
+airports and surrounding weather stations.
 }
@@
 of trees to a forest model allows for modeling smooth changes as well. 
-Therefore, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
+In that way, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
 These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
@@
 In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
-at the airports Innsbruck and Vienna.
+at the airports of Innsbruck and Vienna.
 
@@
 
-\subsection{Models and Evaluation}
+\subsection{Reference Models}
 Along with distributional trees for circular data and forests three other models have been applied 
@@
 Similar to the applied persistence model observations are selected to fit a distributional climatology
-model. For a specified day and time point observations from the exact same time point and day but all 
-other years contained in the learning data are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.
+model. For a specified day and time point observations from the exact same time point and the same day 
+as well as the 30 surrounding days from all other years contained in the learning data are selected and
+parameters are estimated following again Section~\ref{sec:prob_circ}.
 
@@
 
-% TODO: CRPS
-\vspace{0.5cm}
+\subsection{Evaluation}
+Five-fold cross validation (four years as learning data, one year as testing data) ...
+
 Compare fitted models based on a circular version of CRPS ...

### 2019-10-06 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revisions and restructuring of section 'Evaluation Setting and Models Applied'; small changes in motivation for distributional trees and forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2789 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 386e7c5ecf000263b56206a789b1888806aa3e14

@@
 This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
-allowing for the mean direction or the dispersion to be modeled not only in separate models 
-but also in a mixed model. However, applying these models has shown to be very challenging, 
-mainly due to the complexity encountered in optimizing the log-likelihood function.
-Moreover the named approaches and also its further developments 
+introducing another specific transformation function which maps the values of the real line to the 
+unit circle. Moreover, the presented methodology allows for the mean direction or the dispersion 
+to be modeled not only in separate models but also in a mixed model. However, applying these models 
+has shown to be very challenging, mainly due to the complexity encountered in optimizing the 
+log-likelihood function.
+Additionally, the named approaches and also its further developments 
 \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
@@
 In that way no model assumptions are required as covariates and their possible interactions are
-selected automatically. Non-additive effects are captured by the tree structure while combining an ensemble
-of trees to a forest model allows for modeling smooth changes as well. 
-In that way, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
+selected automatically. Non-additive effects are captured by the tree structure while combining an 
+ensemble of trees to a forest model allows for modeling smooth changes as well. 
+In that way, a high level of flexibility is provided while a full distributional model is specified 
+by distributional trees and forests such that a wide range of inference is made available.
 These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
@@
 and commonly applied for probabilistic modeling of circular data.
-Based on a location parameter $\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ 
-the density of the von Mises distribution for an observation $y \in [0, 2\pi]$ is given by:
+Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
+the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
 \begin{equation}
@@
 There, each distribution parameter is linked to a linear predictor by an additive function.
-While this strategy works well for smooth effects difficulties can arise in case of 
-non-additive changes. Moreover, specifying a model properly can be very challenging,
+For circular data the mixed model by \cite{Fisher:1993} follows the idea of modeling both
+parameters of the von Mises distribution, however, in an iterative optimization procedure
+requiring well fitting initial values for the parameters that are recommended to be calculated 
+in advance in two separate models. Using the obtained initial value the model parameters are 
+estimated and improved stepwise until convergence. While providing a well elaborated theory 
+this methodology and the contained optimization problem have shown to be very complex and 
+bear the risk of not converging if not applied thoughtfully \textit{FIXME: citation}. %\citealp{}.
+Furthermore, these additive strategies work well for smooth effects but difficulties can arise 
+in case of non-additive changes. Additionally, specifying a model properly can be very challenging,
 particularly for a high number of covariates and no information about possible interactions.
-An alternative approach that can deal with these problems is to apply a tree-structured model.
+An alternative regression approach that can deal with these problems is to apply a tree-structured 
+model.
 
@@
 
-\subsection{Reference Models}
-Along with distributional trees for circular data and forests three other models have been applied 
-and their performances have been compared based on a circular version of the continuous ranked probability 
-score (CRPS).
+\subsection{Evaluation Setting and Models Applied}
+Along with distributional trees and forests for circular data three other models have been applied
+in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
+the five years contained in the data set being used as testing data once after the models where 
+learned on the remaining four years. 
+In order to compare the performances of the employed models a circular version of the continuous 
+ranked probability score (CRPS) is calculated and summed up over each observation contained in
+the learning data and the corresponding predicted distribution parameters of each of the models.
+
+To test whether distributional trees and forests can be applied as a reasonable alternative to already
+existing approaches to modeling circular data, particularly in the field of meteorological forecasting,
+three commonly used models are selected as reference models.
+The exact settings of the tree and forest models as well as a specification of the reference models
+is provided in the remainder of this section.
+
+\paragraph{Distributional Tree}
+For the distributional tree as discussed in Section~\ref{sec:circtree} all covariates provided in the 
+learning data are considered. The tree is obtained by employing the CTree algorithm with a minimal 
+number of 2000 observations in each terminal node serving as a stopping criterion 
+(argument \code{minbucket} of our implementation).
+
+\paragraph{Distributional Forest}
+Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
+As for the distributional tree all available covariates are used to learn a distributional 
+forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
+containing a fraction of 30 percent of the original learning data. As it can be expected that 
+the lagged response variable (as included as a covariate) has a strong influence on the 
+response variable it is desirable to employ it in each tree of the forest. Therefore, all 
+covariates are considered for each tree
+(by setting the argument \code{mtry} of our implementation to the total number of covariates)
+in order to guarantee for the lagged response to be available.
+Since a high number of possible split points lead to a high computational effort that is 
+particularly noteworthy for forest models the covariates are binned allowing for a maximum of
+$50$ classes.
+% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
+As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
+is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
+is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
+maximum value of 1 in order to build trees as large as possible as commonly done for forest models.
 
@@
 parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.
+As this approach only considers data from the previous six days a cross validation is not 
+applicable, however, the same evaluation framework as for the other four models can be employed 
+by obtaining predictions only for the testing data of the cross validation without using 
+the corresponding learning data.
 
@@
 % TODO: references
-
-\subsection{Evaluation}
-Five-fold cross validation (four years as learning data, one year as testing data) ...
-
-Compare fitted models based on a circular version of CRPS ...
 

### 2019-10-16 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added reference on circular CRPS, information on subset and exponential weights in persistence model; added further fixmes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2795 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0a63e551fbc3b892f5ebd7e2cb4e732580afe6cf

@@
 \caption{Empirical histogram (gray) and fitted density (red line)
-are depicted along with the estimated location parameter (red hand).}
-%TODO: refer to data, Innsbruck airport
+are depicted along with the estimated location parameter (red hand).
+\fixme{refer to data, Innsbruck airport}
+}
+
 \label{fig:density_vM} 
@@
 this methodology and the contained optimization problem have shown to be very complex and 
-bear the risk of not converging if not applied thoughtfully \textit{FIXME: citation}. %\citealp{}.
+bear the risk of not converging if not applied thoughtfully. \fixme{citation} %\citealp{}.
 Furthermore, these additive strategies work well for smooth effects but difficulties can arise 
@@
 
-%% TODO: available approaches to wind direction forecasting
+\fixme{information on already available approaches to wind direction forecasting}
 
@@
 
-Next to the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
+Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
 and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
@@
 
-\subsection{Data}
-For both airports observations of various meteorological quantities measured
-either directly at the airports or at surrounding weather stations are used as 
-covariates in order to predict the wind direction at the corresponding airport 
-in the following hour.
-Next to the current wind direction, temperature, air pressure, humidity, ... at each 
-of the weather stations also differences in these quantities between the considered 
-stations are included. For example, differences in air pressure between Innsbruck and 
-either a western or eastern location are expected to have a high influence on wind 
-directions. Moreover, changes of these quantities over time are included as well as 
-daytime. In total, 550 covariates are provided for Innsbruck and 522 covariates for Vienna. 
-A list of all covariates together with their abbreviations as used in this paper
-can be found in Table~\ref{tab:covariates}.
-The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
-After eliminating observations with at least one missing value a total number of 26861 time points
-for Innsbruck and 33853 for Vienna are available.
-
-\begin{table}[t!]
-%\resizebox{1.4\textwidth}{!}{
-\begin{minipage}{\textwidth}
-\caption[Table caption text]{Basic covariates together with the type of variations.}
-\label{tab:covariates}
-\begin{tabular}{l  l}
-\hline
-Basic covariates & Variations\\
-\hline
-\emph{dd}: wind direction       & different locations, \\
-                                & changes over time(?)\\
-\emph{ff}: wind speed           & different locations, \\
-                                & changes over time(?), \\
-                                & (hourly) maxima\\
-\emph{p}: pressure              & different locations, \\
-                                & differences between locations, \\
-                                & changes over time(?)\\
-\emph{t}: temperature           & different locations, \\
-                                & changes over time(?)\\
-\emph{rf}: relative humidity    & different locations\\
-\hline
-\end{tabular}
-\end{minipage} 
-%}
-\end{table}
-
 \subsection{Evaluation Setting and Models Applied}
@@
 In order to compare the performances of the employed models a circular version of the continuous 
-ranked probability score (CRPS) is calculated and summed up over each observation contained in
-the learning data and the corresponding predicted distribution parameters of each of the models.
+ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
+As the linear version of the CRPS it measures the difference between an observation
+and the corresponding predicted distribution function in order to assess the probabilistic goodness 
+of fit. In order to adjust this general methodology to circular data an additional auxiliary function
+is employed which measures the angular distance between two circular variables. 
+The resulting values are summed up over each observation contained in the learning data and the 
+corresponding predicted distribution parameters of each of the models.
 
@@
 of interest whether this approach can provide overall reliable predictions.
-% TODO: references
-To establish a distributional model of persistence for this study a set of six observations
+
+\fixme{references}
+
+To establish a distributional model of persistence in this study a set of six observations
 is considered to make a prediction for a selected day and time point. In particular,
-the exact same time points of the previous six days are considered. In order to account
-for a stronger influence of observations closer to the prediction time point exponentially
-increasing weights are included with the highest weight being assigned to the time point of the
-previous day. For this selected set of observations the distribution parameters of the von Mises 
+the six previous time points are considered. In order to allow for a stronger influence of 
+observations closer to the prediction time point exponential smoothing is employed, i.e.,
+exponentially decaying weights are included with a smoothing factor of $0.5$ such that
+in each prediction step the current prediction and the previous observation both have an 
+equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
+as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
+For this selected set of weighted observations the distribution parameters of the von Mises 
 distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
@@
 dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.
-% TODO: references
+
+\fixme{add references}
+
 Similar to the applied persistence model observations are selected to fit a distributional climatology
@@
 a specified link function. The toolbox of generalized linear models ...
-% TODO: references
+
+\fixme{add references}
+
+\fixme{information on applied implementation}
+
+\subsection{Data}
+For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
+such as wind direction, temperature, air pressure and humidity
+measured either directly at the airports or at surrounding weather stations are used as 
+covariates in order to predict the wind direction at the corresponding airport 
+in the following hour.
+Next to the current measurements at each of the weather stations also differences in these 
+quantities between the considered stations are included. For example, differences in air 
+pressure between Innsbruck and either a western or eastern location are expected to have a 
+high influence on wind directions. Moreover, changes of these quantities over time are 
+included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
+covariates for Vienna. 
+A list of all covariates together with their abbreviations as used in this paper
+can be found in Table~\ref{tab:covariates}.
+The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
+After eliminating observations with at least one missing value a total number of 26861 time points
+for Innsbruck and 33853 for Vienna are available.
+As winds with very low speed tend to vary their direction strongly they are not expected to be very 
+useful/meaningful for predictions. Moreover, the performance and the stability of the employed reference 
+models might suffer from the influence of these ``noise'' observations. Therefore, only observations
+corresponding to a wind speed of at least $1$ m/s are considered.
+
+\fixme{information on NAs}
+
+\begin{table}[t!]
+%\resizebox{1.4\textwidth}{!}{
+\begin{minipage}{\textwidth}
+\caption[Table caption text]{Basic covariates together with the type of variations.}
+\label{tab:covariates}
+\begin{tabular}{l  l}
+\hline
+Basic covariates & Variations\\
+\hline
+\emph{dd}: wind direction       & different locations, \\
+                                & changes over time(?)\\
+\emph{ff}: wind speed           & different locations, \\
+                                & changes over time(?), \\
+                                & (hourly) maxima\\
+\emph{p}: pressure              & different locations, \\
+                                & differences between locations, \\
+                                & changes over time(?)\\
+\emph{t}: temperature           & different locations, \\
+                                & changes over time(?)\\
+\emph{rf}: relative humidity    & different locations\\
+\hline
+\end{tabular}
+\end{minipage} 
+%}
+\end{table}
 

### 2019-10-18 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added references and information on CRPS skill score, restructured introduction and application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2801 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dad6a845b2a7ac82a1cdd7d220fa4b8048b0adb3

@@
 
-\title{Distributional Trees and Forests for Circular Data}
-%\Shorttitle{Distributional Trees and Forests for Circular Data}
+\title{Distributional Trees and Forests for Circular Data with an Application to Probabilistic Wind Direction Forecasting}
+\Shorttitle{Distributional Trees and Forests for Circular Data}
 
@@
 In many approaches to model circular data it is assumed that the circular variable of interest 
-follows a circular distribution, in particular the von Mises distribution (which will be discussed 
-in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
-models for circular data were built on as they regress the distribution parameter(s) on covariates.
+follows a circular distribution, in particular the von Mises distribution which is also known 
+as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
+This was also the basis which early regression models for circular data were built on as they 
+regress the distribution parameter(s) on covariates.
+
 One of the first regression models with a circular response variable and linear covariates
-was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
-the circular mean, hence the location parameter of the von Mises distribtuion. 
-\cite{Johnson+Wehrly:1978} took this idea one step further 
-by plugging in a link function such that the values of the linear predictor are transformed 
-to an restricted interval of length $2\pi$. 
+was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
+the von Mises distribtuion, is predicted by a linear combination of covariates.
+\cite{Johnson+Wehrly:1978} took this idea one step further by employing generalized linear 
+model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
+linear predictor are transformed to a restricted interval of length $2\pi$. 
 This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
 introducing another specific transformation function which maps the values of the real line to the 
-unit circle. Moreover, the presented methodology allows for the mean direction or the dispersion 
-to be modeled not only in separate models but also in a mixed model. However, applying these models 
-has shown to be very challenging, mainly due to the complexity encountered in optimizing the 
-log-likelihood function.
-Additionally, the named approaches and also its further developments 
-\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
-are built on additive regression models which may not be able to capture all types of effects.
-For example abrupt changes might not be detected properly owing to the fixed structure enforced
-by model assumptions such as the type of link function and the type of linear combination.
-Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
-distributional trees and distributional forests. 
+unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter, to be modeled not only in separate models but also in a 
+mixed model. There, within an iterative optimization algorithm the two parameters are estimated and 
+improved stepwise until convergence. To obtain proper initial values it is recommended to apply
+two separate models where either one of the parameters is fixed while the other is modeled depending
+on covariates. While providing a well elaborated theory this methodology and the contained optimization 
+problem have shown to be very challenging, mainly due to the complexity encountered in optimizing the 
+log-likelihood function that is not globally concave, and thus bear a high risk of not converging if 
+not applied thoughtfully \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
+An alternative approach to regression modeling of circular data that avoids this strong dependence 
+on choosing well-fitting initial values is presented by \cite{Mulder+Klugkist:2017} employing a GLM 
+in a Bayesian framework.
+
+The named methods and also many of their further developments 
+\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
+While they provide a powerful toolbox, particularly for modeling smooth effects, they may not be 
+able to capture all types of effects. For example, abrupt changes might not be detected properly owing 
+to the fixed structure enforced by model assumptions such as the type of link function and the 
+form of linear combinations.
+
+Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
+forests to circular data, specifically by employing distributional trees and distributional forests
+following \cite{Schlosser+Hothorn+Stauffer:2019}. 
 In that way no model assumptions are required as covariates and their possible interactions are
@@
 by distributional trees and forests such that a wide range of inference is made available.
-These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
-distributional models for circular data in Section~\ref{sec:prob_circ}.
+These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after
+discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
 In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
@@
 are depicted along with the estimated location parameter (red hand).
-\fixme{refer to data, Innsbruck airport}
 }
-
 \label{fig:density_vM} 
@@
 
-Figure~\ref{fig:density_vM} illustrates a fitted distributional model employing
-the von Mises distribution for observed wind direction data from the airport of Innsbruck.
+Figure~\ref{fig:density_vM} illustrates a distributional model fitted to circular data
+employing the von Mises distribution.
+% on observed wind direction data from the airport of Innsbruck.
 This example shows that reasonably well fitting probabilistic models can
@@
 \label{sec:tree_forest}
-\subsection{Motivation}
-A commonly used approach to including covariates as regressors in distributional models is 
-provided by generalized additive models for location, scale, and shape 
-\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
-There, each distribution parameter is linked to a linear predictor by an additive function.
-For circular data the mixed model by \cite{Fisher:1993} follows the idea of modeling both
-parameters of the von Mises distribution, however, in an iterative optimization procedure
-requiring well fitting initial values for the parameters that are recommended to be calculated 
-in advance in two separate models. Using the obtained initial value the model parameters are 
-estimated and improved stepwise until convergence. While providing a well elaborated theory 
-this methodology and the contained optimization problem have shown to be very complex and 
-bear the risk of not converging if not applied thoughtfully. \fixme{citation} %\citealp{}.
-Furthermore, these additive strategies work well for smooth effects but difficulties can arise 
-in case of non-additive changes. Additionally, specifying a model properly can be very challenging,
-particularly for a high number of covariates and no information about possible interactions.
+%\subsection{Motivation}
+A common way to including covariates as regressors in distributional models is to apply
+generalized additive models for location, scale, and shape 
+\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
+by its own generalized additive model (GAM).
+As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
+for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
+specifying a model properly can be very challenging, particularly for a high number of covariates 
+and no information about possible interactions.
 An alternative regression approach that can deal with these problems is to apply a tree-structured 
@@
 
-\section{Wind Direction Forecasts}
+\section{Case Study: Wind Direction Forecasts}
 \label{sec:wind}
-\subsection{Motivation}
+
 Wind is a classical circular quantity and accurate forecasts of wind direction
@@
 
-As a motivational example Figure~\ref{fig:tree_ibk} illustrates a distributional
+\subsection{Motivational Example}
+As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
 tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
@@
 
+\subsection{Data}
+For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
+such as wind direction, temperature, air pressure and humidity
+measured either directly at the airports or at surrounding weather stations are used as 
+covariates in order to predict the wind direction at the corresponding airport 
+in the following hour.
+Next to the current measurements at each of the weather stations also differences in these 
+quantities between the considered stations are included. For example, differences in air 
+pressure between Innsbruck and either a western or eastern location are expected to have a 
+high influence on wind directions. Moreover, changes of these quantities over time are 
+included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
+covariates for Vienna. 
+A list of all covariates together with their abbreviations as used in this paper
+can be found in Table~\ref{tab:covariates}.
+The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
+After eliminating observations with at least one missing value a total number of 26861 time points
+for Innsbruck and 33853 for Vienna are available.
+As winds with very low speed tend to vary their direction strongly they are not expected to be very 
+useful/meaningful for predictions. Moreover, the performance and the stability of the employed reference 
+models might suffer from the influence of these ``noise'' observations. Therefore, only observations
+corresponding to a wind speed of at least $1$ m/s are considered.
+
+\fixme{information on NAs}
+
+\begin{table}[t!]
+%\resizebox{1.4\textwidth}{!}{
+\begin{minipage}{\textwidth}
+\caption[Table caption text]{Basic covariates together with the type of variations.}
+\label{tab:covariates}
+\begin{tabular}{l  l}
+\hline
+Basic covariates & Variations\\
+\hline
+\emph{dd}: wind direction       & different locations, \\
+                                & changes over time(?)\\
+\emph{ff}: wind speed           & different locations, \\
+                                & changes over time(?), \\
+                                & (hourly) maxima\\
+\emph{p}: pressure              & different locations, \\
+                                & differences between locations, \\
+                                & changes over time(?)\\
+\emph{t}: temperature           & different locations, \\
+                                & changes over time(?)\\
+\emph{rf}: relative humidity    & different locations\\
+\hline
+\end{tabular}
+\end{minipage} 
+%}
+\end{table}
+
 \subsection{Evaluation Setting and Models Applied}
@@
 ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
-As the linear version of the CRPS it measures the difference between an observation
-and the corresponding predicted distribution function in order to assess the probabilistic goodness 
-of fit. In order to adjust this general methodology to circular data an additional auxiliary function
-is employed which measures the angular distance between two circular variables. 
-The resulting values are summed up over each observation contained in the learning data and the 
-corresponding predicted distribution parameters of each of the models.
+Just as the linear version of the CRPS \citep[see][for more details]{Hersbach:2000} it measures the 
+difference between an observation and the corresponding predicted distribution function in order 
+to assess the probabilistic goodness of fit of the assessed model. Hence, the lower the CRPS value 
+the better the performance. In order to adjust the general methodology of the CRPS to circular data 
+\cite{Grimit+Gneiting+Berrocal:2006} employ an auxiliary function which measures the angular distance 
+between two circular variables. 
+The circular CRPS is evaluated out of sample and the resulting values are averaged over 
+each observation contained in the testing data and the corresponding predicted distribution 
+parameters of each of the models.
+To measure the improvement of the trees, forests, persistence and linear models 
+over the climatology model the skill score based on the circular CRPS is evaluated
+with the climatology model used as reference:
+\begin{equation}
+\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
+\end{equation}
 
@@
 \paragraph{Generalized Linear Model}
-Another commonly applied approach is to regress the parameter(s) of the von Mises distribution
-on covariates by a linear combination of selected covariates connected to the parameter(s) via
-a specified link function. The toolbox of generalized linear models ...
+Another commonly applied approach is to linearly regress the circular variable of interest on
+its lagged values. For the here discussed application this corresponds to establishing a linear model
+with wind direction as the outcome variable and wind direction of the previous time point as covariate.
+Following \cite{Mulder+Klugkist:2017}, the employed model plugs in a link function in order
+to guarantee values on the interval $[-\pi,\pi]$ and is of the form
+\begin{align}
+dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
+\end{align}
+with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
+$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
+Using the corresponding implementation provided in the \textsf{R} package \textbf{circglmbayes}
+by \cite{Mulder+Klugkist:2017} the function \textbf{circGLM} estimates the
+intercept and regression coefficient along with the the concentration parameter $\kappa$.
 
-\fixme{add references}
+\fixme{add references and further information on the estimation process within circGLM}
 
-\fixme{information on applied implementation}
-
-\subsection{Data}
-For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
-such as wind direction, temperature, air pressure and humidity
-measured either directly at the airports or at surrounding weather stations are used as 
-covariates in order to predict the wind direction at the corresponding airport 
-in the following hour.
-Next to the current measurements at each of the weather stations also differences in these 
-quantities between the considered stations are included. For example, differences in air 
-pressure between Innsbruck and either a western or eastern location are expected to have a 
-high influence on wind directions. Moreover, changes of these quantities over time are 
-included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
-covariates for Vienna. 
-A list of all covariates together with their abbreviations as used in this paper
-can be found in Table~\ref{tab:covariates}.
-The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
-After eliminating observations with at least one missing value a total number of 26861 time points
-for Innsbruck and 33853 for Vienna are available.
-As winds with very low speed tend to vary their direction strongly they are not expected to be very 
-useful/meaningful for predictions. Moreover, the performance and the stability of the employed reference 
-models might suffer from the influence of these ``noise'' observations. Therefore, only observations
-corresponding to a wind speed of at least $1$ m/s are considered.
-
-\fixme{information on NAs}
-
-\begin{table}[t!]
-%\resizebox{1.4\textwidth}{!}{
-\begin{minipage}{\textwidth}
-\caption[Table caption text]{Basic covariates together with the type of variations.}
-\label{tab:covariates}
-\begin{tabular}{l  l}
-\hline
-Basic covariates & Variations\\
-\hline
-\emph{dd}: wind direction       & different locations, \\
-                                & changes over time(?)\\
-\emph{ff}: wind speed           & different locations, \\
-                                & changes over time(?), \\
-                                & (hourly) maxima\\
-\emph{p}: pressure              & different locations, \\
-                                & differences between locations, \\
-                                & changes over time(?)\\
-\emph{t}: temperature           & different locations, \\
-                                & changes over time(?)\\
-\emph{rf}: relative humidity    & different locations\\
-\hline
-\end{tabular}
-\end{minipage} 
-%}
-\end{table}
+\fixme{information on applied implementation here or in separate chapter?}
 
@@
 
-\section{Discussion and outlook}
+\section{Discussion and Outlook}
 
@@
 
-\subsection{Splits in circular covariates}
+\fixme{Summary of results}
 
-In order to obtain more parsimonious and more stable trees another possible
-extension for \emph{circular covariates} (with or without a \emph{circular response})
-is to consider their circular nature when searching the best split into two segments.
-In general, searching the best separation of a covariate into a ``left'' and ``right''
-daughter node tries to maximize the segmented log-likelihood:
-\begin{equation}
-\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
-\end{equation}
-where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
-of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a circular covariate $\in [0, 2 \pi)$
-only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
-thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
-for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
-and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi)$, encompassing zero. The latter
-strategy is invariant to the (often arbitrary) definition of the direction at zero.
+\fixme{Possible extensions of models, use of NWP data}
 
-When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
-far away, a simple linear split typically suffices to capture such a split (as seen for the
-lagged wind direction in Figure~\ref{fig:tree_ibk}). If both $\nu$ and $\tau$ differ
-clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
-fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
-might miss such a pattern.
+%\subsection{Splits in circular covariates}
 
-The required test statistic to maximally select two split points simultaneously is straightforward
-to accommodate in the CTree framework by providing all binary indicators corresponding
-to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
-increasingly slow for larger sample sizes but it might be possible to speed up computations by
-exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
-MOB framework the extension is not quite as straightforward but one strategy could be to
-adapt double maximum tests \`a la \cite{Bai+Perron:2003}.
+%In order to obtain more parsimonious and more stable trees another possible
+%extension for \emph{circular covariates} (with or without a \emph{circular response})
+%is to consider their circular nature when searching the best split into two segments.
+%In general, searching the best separation of a covariate into a ``left'' and ``right''
+%daughter node tries to maximize the segmented log-likelihood:
+%\begin{equation}
+%\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} %\ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
+%\end{equation}
+%where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
+%of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a %circular covariate $\in [0, 2 \pi)$
+%only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
+%thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
+%for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
+%and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi)$, encompassing zero. The latter
+%strategy is invariant to the (often arbitrary) definition of the direction at zero.
 
-Hence, the splitting idea can be naturally extended to a two-point search, however, for 
-an unbiased and inference-based selection the corresponding testing strategies might need
-further adaption.
+%When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
+%far away, a simple linear split typically suffices to capture such a split (as seen for the
+%lagged wind direction in Figure~\ref{fig:tree_ibk}). If both $\nu$ and $\tau$ differ
+%clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
+%fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
+%might miss such a pattern.
+
+%The required test statistic to maximally select two split points simultaneously is straightforward
+%to accommodate in the CTree framework by providing all binary indicators corresponding
+%to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
+%increasingly slow for larger sample sizes but it might be possible to speed up computations by
+%exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
+%MOB framework the extension is not quite as straightforward but one strategy could be to
+%adapt double maximum tests \`a la \cite{Bai+Perron:2003}.
+
+%Hence, the splitting idea can be naturally extended to a two-point search, however, for 
+%an unbiased and inference-based selection the corresponding testing strategies might need
+%further adaption.
 

### 2019-10-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2803 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5526f6db8a1ab44ea03cd4f8f2a21594e5e3b2a8

@@
 }
-\Keywords{Distributional Trees; Circular Response; Von Mises Distribution}
+\Keywords{Distributional Trees and Forests; Circular Response; Von Mises Distribution}
 
@@
 
-In many approaches to model circular data it is assumed that the circular variable of interest 
+In many approaches to modeling circular data it is assumed that the circular variable of interest 
 follows a circular distribution, in particular the von Mises distribution which is also known 
 as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
-This was also the basis which early regression models for circular data were built on as they 
-regress the distribution parameter(s) on covariates.
+This is also the basis which early regression models for circular data are built on as they 
+regress the parameter(s) of the von Mises distribution on covariates.
 
@@
 the von Mises distribtuion, is predicted by a linear combination of covariates.
-\cite{Johnson+Wehrly:1978} took this idea one step further by employing generalized linear 
+\cite{Johnson+Wehrly:1978} took this idea one step further by employing a generalized linear 
 model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
@@
 two separate models where either one of the parameters is fixed while the other is modeled depending
-on covariates. While providing a well elaborated theory this methodology and the contained optimization 
-problem have shown to be very challenging, mainly due to the complexity encountered in optimizing the 
-log-likelihood function that is not globally concave, and thus bear a high risk of not converging if 
-not applied thoughtfully \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
+on covariates. While being built on well elaborated theory this methodology and the contained 
+optimization problem have shown to be very challenging in their application, mainly due to the 
+complexity encountered in optimizing the log-likelihood function that is not globally concave, 
+and thus bear a high risk of not converging if not applied thoughtfully 
+\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
 An alternative approach to regression modeling of circular data that avoids this strong dependence 
@@
 \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
-While they provide a powerful toolbox, particularly for modeling smooth effects, they may not be 
-able to capture all types of effects. For example, abrupt changes might not be detected properly owing 
-to the fixed structure enforced by model assumptions such as the type of link function and the 
-form of linear combinations.
+While they provide a powerful toolbox for modeling circular data, particularly in case of
+smooth effects, they may not be able to capture all types of effects. For example, abrupt changes 
+might not be detected properly owing to the enforced additive structure and model assumptions 
+such as the type of link function and the form of linear combinations.
 
 Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
-forests to circular data, specifically by employing distributional trees and distributional forests
-following \cite{Schlosser+Hothorn+Stauffer:2019}. 
+forests to circular data, specifically by employing the von Mises distribution within the methodology
+of distributional trees and distributional forests following \cite{Schlosser+Hothorn+Stauffer:2019}. 
 In that way no model assumptions are required as covariates and their possible interactions are
@@
 ensemble of trees to a forest model allows for modeling smooth changes as well. 
-In that way, a high level of flexibility is provided while a full distributional model is specified 
+Thus, a high level of flexibility is provided while a full distributional model is specified 
 by distributional trees and forests such that a wide range of inference is made available.
-These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after
+The corresponding methodology and its features are presented in Section~\ref{sec:tree_forest} after
 discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
-In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
+In Section~\ref{sec:wind} the presented models are applied to probabilistic forecasting of wind direction
 at the airports of Innsbruck and Vienna.
@@
 
-Employing the score function
+By means of the score function
 \begin{equation}
+\label{eq:scores}
 \begin{aligned}
@@
 \end{equation}
-the optimization problem in Equation~\ref{eq:optim} can also be defined by 
+an indicator of deviation and in that way a measure of goodness of fit of the model can be obtained 
+for each observation $y_i$. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
+be defined by 
 %setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
@@
 \includegraphics[width=4cm]{density_vM.pdf}
-\caption{Empirical histogram (gray) and fitted density (red line)
-are depicted along with the estimated location parameter (red hand).
+\caption{Illustration of a distributional model fitted to circular data by maximum likelihood-estimation
+employing the von Mises distribution. Empirical histogram (gray) and fitted density (red line) are depicted
+along with the estimated location parameter (red hand).
 }
@@
 This example shows that reasonably well fitting probabilistic models can
-be obtained for circular data by specifying an appropriate distribution family 
+be established for circular data by specifying an appropriate distribution family 
 such as the von Mises distribution and estimating its parameters via maximum likelihood. 
@@
 %\subsection{Motivation}
-A common way to including covariates as regressors in distributional models is to apply
+A common way of including covariates as regressors in distributional models is to apply
 generalized additive models for location, scale, and shape 
@@
 specifying a model properly can be very challenging, particularly for a high number of covariates 
-and no information about possible interactions.
+and no information on possible interactions.
 An alternative regression approach that can deal with these problems is to apply a tree-structured 
@@
 algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
-In the presented distributional trees a full distributional model is specified in each node of the 
-tree. For this purpose a distribution family has to be chosen in advance. Hence, selecting the 
-von Mises distribution allows for an application of distributional trees to circular data following 
-the tree building algorithm as introduced by \cite{Schlosser+Hothorn+Stauffer:2019}.
+By applying the presented distributional trees a full distributional model is specified in each 
+node of the tree. For this purpose a distribution family has to be chosen in advance. Hence, 
+selecting the von Mises distribution allows for an application of distributional trees to circular
+data following the tree building algorithm as introduced in Section~2.2. of 
+\cite{Schlosser+Hothorn+Stauffer:2019}.
 
-In particular, in each node the crucial decision on how and where to split the data is based on model scores which are obtained by evaluating the score function $s$ at the individual observations and 
-parameter estimates. For the von Mises distribution and a data set of $n$ observations
-this yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
-each given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
-respectively.
+In particular, in each node the crucial decision on how and where to split the data is based on model
+scores which are obtained by evaluating the score function $s$ as defined in Equation~\ref{eq:scores} 
+at the individual observations and parameter estimates. For the von Mises distribution and a data set 
+of $n$ observations this yields an $n \times 2$ matrix that can be employed as a kind of residual,
+capturing how well each given observation conforms with the estimated location $\hat{\mu}$ and precision
+$\hat{\kappa}$, respectively.
 To capture dependence on covariates, the association between the model's scores and each available
-covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
+covariate is assessed using either a parameter instability test~(MOB) or a permutation test~(CTree).
 In each partitioning step, the covariate with the highest significant association (i.e., lowest
 significant $p$-value, if any) is selected for splitting the data. The corresponding split point
-is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
+is chosen either by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
 over all possible partitions.
@@
 Therefore, the same parameter pair is estimated for all observations
-belonging to the same terminal node.
-This facilitates the application as the parameter estimates do not need to be recalculated for each (new) observation via maximum likelihood but can be extracted directly
-from the learning sample and the fitted model.
+belonging to the same terminal node. This facilitates the application as the parameter estimates do 
+not need to be recalculated for each (new) observation via maximum likelihood but can be extracted 
+directly from the learning sample and the fitted model.
 
-While tree models can capture non-additive effects they do not allow for purely smooth effects and therefore might impose abrupt changes even in case of rather smooth transitions. This can be avoided
+While tree models can capture non-additive effects their structure and the consequential strict 
+separation of data into subgroups hinders an adequate depiction of smooth effects. Therefore, 
+they might impose abrupt changes even in case of rather smooth transitions. This can be avoided 
 by combining an ensemble of trees in order to obtain a more stable forest model.
@@
 of trees, each built on a subsample or bootstrap of the original data. 
-A generalization of this strategy is to obtain weighted predictions
-by adaptive local likelihood estimation of the distributional parameters
-\citep{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
-More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights $w_i^{forest}(\bm{z})$ 
+A generalization of this strategy is to obtain weighted predictions by adaptive local 
+likelihood estimation of the distributional parameters 
+\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
+More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
+weights $w_i^{forest}(\bm{z})$ 
 %\citep{Lin+Jeon:2006}
@@
 for Innsbruck and 33853 for Vienna are available.
-As winds with very low speed tend to vary their direction strongly they are not expected to be very 
-useful/meaningful for predictions. Moreover, the performance and the stability of the employed reference 
-models might suffer from the influence of these ``noise'' observations. Therefore, only observations
-corresponding to a wind speed of at least $1$ m/s are considered.
+As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
+expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
+especially of the employed reference models might suffer from the influence of these ``noise'' 
+observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
+considered.
 
@@
 learned on the remaining four years.
+
 In order to compare the performances of the employed models a circular version of the continuous 
 ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
-Just as the linear version of the CRPS \citep[see][for more details]{Hersbach:2000} it measures the 
+Just as the linear version of the CRPS \citep[for more details see][]{Hersbach:2000} it measures the 
 difference between an observation and the corresponding predicted distribution function in order 
-to assess the probabilistic goodness of fit of the assessed model. Hence, the lower the CRPS value 
+to assess the probabilistic goodness of fit of the fitted model. Hence, the lower the CRPS value 
 the better the performance. In order to adjust the general methodology of the CRPS to circular data 
@@
 between two circular variables.
+
 The circular CRPS is evaluated out of sample and the resulting values are averaged over 
@@
 parameters of each of the models.
-To measure the improvement of the trees, forests, persistence and linear models 
-over the climatology model the skill score based on the circular CRPS is evaluated
-with the climatology model used as reference:
+
+\fixme{Comment on aggregation}
+
+To measure the improvement of one of the applied methods over a climatology model (which will 
+be described in detail subsequently) the skill score based on the circular 
+CRPS of the assessed method is evaluated with the climatology model used as reference:
 \begin{equation}
@@
 
-To test whether distributional trees and forests can be applied as a reasonable alternative to already
-existing approaches to modeling circular data, particularly in the field of meteorological forecasting,
-three commonly used models are selected as reference models.
+Three commonly used methods are selected as reference models in order to investigate whether 
+distributional trees and forests can be applied as a reasonable alternative to already
+existing approaches to modeling circular data, particularly in the field of meteorological forecasting.
 The exact settings of the tree and forest models as well as a specification of the reference models
@@
 in order to guarantee for the lagged response to be available.
-Since a high number of possible split points lead to a high computational effort that is 
+Since a high number of possible split points leads to a high computational effort that is 
 particularly noteworthy for forest models the covariates are binned allowing for a maximum of
-$50$ classes.
+$50$ classes. In that way the number of split points that have to be considered is reduced
+in a reasonably.
 % control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
@@
 distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
-parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.
+parametric model is specified allowing for probabilistic predictions and an evaluation of the 
+circular CRPS.
 As this approach only considers data from the previous six days a cross validation is not 
@@
 If there are no climatological changes within a certain number of years it can be expected that 
-for a selected day of the year weather conditions will always be the same over all years 
-considered.
+for a selected time point and day of the year weather conditions will always be the same over all 
+years considered.
 Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
@@
 
-Similar to the applied persistence model observations are selected to fit a distributional climatology
-model. For a specified day and time point observations from the exact same time point and the same day 
-as well as the 30 surrounding days from all other years contained in the learning data are selected and
-parameters are estimated following again Section~\ref{sec:prob_circ}.
+Similar to the applied persistence model a set of observations is selected to fit a distributional
+climatology model. For a specified day and time point observations from the exact same time point of 
+the same day as well as of the 30 surrounding days from all other years contained in the learning data 
+are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.
 
@@
 Another commonly applied approach is to linearly regress the circular variable of interest on
-its lagged values. For the here discussed application this corresponds to establishing a linear model
-with wind direction as the outcome variable and wind direction of the previous time point as covariate.
-Following \cite{Mulder+Klugkist:2017}, the employed model plugs in a link function in order
+its lagged values. For the here discussed application this corresponds to establishing a generalized
+linear model with wind direction as the outcome variable and wind direction of the previous time point 
+as covariate. 
+Following \cite{Mulder+Klugkist:2017} the employed model plugs in a link function in order
 to guarantee values on the interval $[-\pi,\pi]$ and is of the form
@@
 $g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
-Using the corresponding implementation provided in the \textsf{R} package \textbf{circglmbayes}
-by \cite{Mulder+Klugkist:2017} the function \textbf{circGLM} estimates the
-intercept and regression coefficient along with the the concentration parameter $\kappa$.
+The intercept and the regression coefficient are estimated in a Bayesian framework where an 
+MCMC algorithm is used to obtain posterior samples. Moreover, a constant concentration
+parameter is fitted to the full learning sample.
 
-\fixme{add references and further information on the estimation process within circGLM}
+\fixme{add references and further information on the estimation process within circGLM?}
 
@@
 
+Lisa Schlosser recieved a PhD scholarship granted from the University of Innsbruck.
+
 \section*{Computational details}
@@
 
+For the circular GLM considered as reference model the corresponding implementation is provided 
+in the \textsf{R} package \textbf{circglmbayes} by \cite{Mulder+Klugkist:2017}. In particular the 
+function \textbf{circGLM} is applied to estimates the intercept and regression coefficient along with 
+the concentration parameter $\kappa$.
+
 %\newpage

### 2019-11-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
first draft of result description
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2811 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b9943e69d7ce0f93299aabf666976eb308c1e76f

@@
 
-\subsection{Motivational Example}
+\subsection{Motivational Example}\label{sec:mot_ex}
 As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
@@
 covariates in order to predict the wind direction at the corresponding airport 
-in the following hour.
+in the following hour(s).
 Next to the current measurements at each of the weather stations also differences in these 
@@
 \subsection{Results}
-Ibk/Vie, 1h/3h
+In order to test and compare the models over varying settings not only 1-hour but also 3-hour
+forecasts are considered at both airports yielding four different scenarios of which the results
+are discussed in this section.
+
+Figure~\ref{fig:boxplot_crpsraw} illustrates the CRPS values of the employed models with a lower 
+CRPS indicating better performance. 
+In all four settings it can be observed that on average the distributional forest provides the best 
+fitting probabilistic predictions followed by the distributional tree. Especially in Vienna the 
+persistence (hour) model is only slightly behind while the linear model and the climatology
+show clearly higher CRPS values.
+
+Comparing the performances over the different locations the narrower boxes in Vienna represent
+a lower variation. This can be explained by more abrupt changes of wind directions in Innsbruck
+due to the surrounding mountains channeling wind and in that way allowing only for a limited
+number of possible wind directions (mainly the three wind regimes as discussed in the motivational
+example in Section~\ref{sec:mot_ex}). For that reason, relatively stable models such as persistency
+produce well fitting forecasts as long as wind directions remain in one of these possible states,
+however, once a change takes place it can only be an abrupt and major shift leading to high
+prediction errors as supported by the wide boxes of the persistence model in Innsbruck. In Vienna,
+on the other hand, changes can also appear in a smaller extend as the geographic surroundings do not
+constrain wind directions. This wider range of possible observations might also weaken the performance
+of the climatology as expressed by the higher CRPS values in Vienna.
+
+The impact of the different forecasting time windows of 1 and 3 hours can hardly be noticed, especially
+as it does not seem to effect the order of which method performs best. But taking a closer look
+at the CRPS values of the tree, forest and persistency reveals lower scores and hence a better 
+performance for the 1-hour forecasts as it could have been expected due to increasing uncertainties
+in a longer time window.
+
+\begin{figure}[t]
+\centering
+\setkeys{Gin}{width=1.0\textwidth}
+%\hspace{-2cm}
+\includegraphics{boxplot_crpsraw.pdf}
+\caption{}
+\label{fig:boxplot_crpsraw} 
+\end{figure}
+
+With the aim of excluding any possible influence of the selection of testing and training data
+the performances of the methods are also compared based on the CRPS skills score as illustrated in 
+Figure~\ref{fig:boxplot_crpsskill}. Using the climatology as reference model the values on the y-axis
+indicate the percentage of improvement of each model compared to the reference model. Overall
+the same conclusions can be drawn as from Figure~\ref{fig:boxplot_crpsraw} with the distribution
+forest leading to the highest overall improvement in model fit.
+
+\begin{figure}[t]
+\centering
+\setkeys{Gin}{width=1.0\textwidth}
+%\hspace{-2cm}
+\includegraphics{boxplot_crpsskill.pdf}
+\caption{}
+\label{fig:boxplot_crpsskill} 
+\end{figure}
 
 \section{Discussion and Outlook}
-
 Distributional trees for circular responses are established by coupling

### 2019-11-05 pkg/extree_partykit/index.Rmd
create extree improvement branch
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2813 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6ea65118b722de54b5770b6466add9909cf65a11

## A Toolkit for Recursive Partytioning

```{r ttnc, echo = FALSE, message = FALSE, results = "hide"}
library("partykit")
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
ttnc <- transform(ttnc, Treatment = factor(
  Gender == "Female" | Age == "Child", levels = c(FALSE, TRUE),
  labels = c("Normal\nMale&Adult", "Preferential\nFemale|Child")
))

```{r ttnc-mob, echo = FALSE, message = FALSE, out.width = "100%", fig.height = 6, fig.width = 10, fig.align = "center", dev = "png", dpi = 48}
ttnc_mob <- glmtree(Survived ~ Treatment | Class + Gender + Age,
  data = ttnc, family = binomial, alpha = 0.01)
plot(ttnc_mob, tp_args = list(ylines = 1.3, margins = c(1.5, 1, 1.5, 2.2)))
```

**Tree algorithms:** Conditional inference trees (`ctree()`) and model-based
recursive partitioning (`mob()`) along with interfaces `lmtree()` and `glmtree()`
for (generalized) linear model trees.

**Infrastructure:** Unified classes and methods for representing, summarizing, and
visualizing tree-structured regression and classification models. Common print/plot/predict methods,
enabling reading/coercing tree models from different sources (`rpart`, `RWeka`, `PMML`).

* `partysplit`: Split with integer ID for the splitting variable,
  breakpoint(s), indexes for the kids.
* `partynode`: Node specification with integer ID, a `partysplit`,
  and a list of kids (if any) that are `partynode` objects again.
* `party`: Tree with a recursive `partynode` and a `data.frame`
  (optionally empty), potentially plus information about fitted values
  and `terms` allowing to preprocess new data for predictions.

### 2019-11-05 pkg/extree_partykit/R/cforest.R

@@ -1,0 +1,304 @@
+
+### constructor for forest objects
+constparties <- function(nodes, data, weights, fitted = NULL, terms = NULL, info = NULL) {
+
+    stopifnot(all(sapply(nodes, function(x) inherits(x, "partynode"))))
+    stopifnot(inherits(data, "data.frame"))
+    stopifnot(inherits(weights, "list"))
+
+    if(!is.null(fitted)) {
+        stopifnot(inherits(fitted, "data.frame"))
+        stopifnot(nrow(data) == 0L | nrow(data) == nrow(fitted))
+        if (nrow(data) == 0L)
+            stopifnot("(response)" %in% names(fitted))
+    } else {
+        stopifnot(nrow(data) > 0L)
+        stopifnot(!is.null(terms))
+        fitted <- data.frame("(response)" = model.response(model.frame(terms, data = data, 
+                                                                       na.action = na.pass)),
+                             check.names = FALSE)
+    }
+
+    ret <- list(nodes = nodes, data = data, weights = weights, fitted = fitted)
+    class(ret) <- c("constparties", "parties")
+
+    if(!is.null(terms)) {
+        stopifnot(inherits(terms, "terms"))
+        ret$terms <- terms
+    }
+
+    if (!is.null(info))
+        ret$info <- info
+
+    ret
+}
+
+.perturb <- function(replace = FALSE, fraction = .632) {
+    ret <- function(prob) {
+        if (replace) {
+            rw <- rmultinom(1, size = length(prob), prob = prob)
+        } else {
+            rw <- integer(length(prob))
+            i <- sample(1:length(prob), ceiling(fraction * length(prob)), prob = prob)
+            rw[i] <- 1L
+        }
+        as.integer(rw)
+    }
+    ret
+}
+
+cforest <- function
+(
+    formula,
+    data,   
+    weights,
+    subset, 
+    offset, 
+    cluster,
+    strata,
+    na.action = na.pass,
+    control = ctree_control(
+        teststat = "quad", testtype = "Univ", mincriterion = 0,
+        saveinfo = FALSE, ...),
+    ytrafo = NULL, 
+    scores = NULL, 
+    ntree = 500L, 
+    perturb = list(replace = FALSE, fraction = 0.632),
+    mtry = ceiling(sqrt(nvar)), 
+    applyfun = NULL,
+    cores = NULL, 
+    trace = FALSE,
+    ...
+) {
+   
+    ### get the call and the calling environment for .urp_tree
[…]

### 2019-11-05 pkg/extree_partykit/inst/guideliketest/logLik_newdata.R

@@ -1,0 +1,75 @@
+# logLik method for lm with optional argument newdata
+logLik_lm_nd <- function (object, REML = FALSE, newdata = NULL, ...) 
+{
+  if (inherits(object, "mlm")) 
+    stop("'logLik.lm' does not support multiple responses")
+  if(!is.null(newdata)){
+    ## FIX ME: better way of choosing response variable in newdata
+    respname <- as.character(object$terms[[2]])
+    res <- predict(object = object, newdata = newdata) - newdata[,respname]
+  } 
+    else res <- object$residuals
+  p <- object$rank
+  N <- length(res)
+  if (is.null(w <- object$weights)) {
+    w <- rep.int(1, N)
+  }
+  else {
+    excl <- w == 0
+    if (any(excl)) {
+      res <- res[!excl]
+      N <- length(res)
+      w <- w[!excl]
+    }
+  }
+  N0 <- N
+  if (REML) 
+    N <- N - p
+  val <- 0.5 * (sum(log(w)) - N * (log(2 * pi) + 1 - log(N) + 
+                                     log(sum(w * res^2))))
+  if (REML) 
+    val <- val - sum(log(abs(diag(object$qr$qr)[1L:p])))
+  attr(val, "nall") <- N0
+  attr(val, "nobs") <- N
+  attr(val, "df") <- p + 1
+  class(val) <- "logLik"
+  val
+}
+
+# logLik method for modelparty and lm in nodes with optional argument newdata
+logLik_modelparty_lm_nd <- function (object, dfsplit = NULL, newdata = NULL, ...) 
+{
+  if (is.null(dfsplit)) 
+    dfsplit <- object$info$control$dfsplit
+  dfsplit <- as.integer(dfsplit)
+  ids <- partykit::nodeids(object, terminal = TRUE)
+  
+  if(!is.null(newdata)){
+    ids_newdata <- predict(object, newdata = newdata, type = "node")
+    newdata$ids <- ids_newdata
+    ll <- list()
+    for(i in ids){
+      if(i %in% ids_newdata){
+        ll <- c(ll, partykit:::apply_to_models(object, node = i, 
+                                               FUN = function(object, REML = FALSE, ...) 
+                                                 logLik_lm_nd(object, REML, 
+                                                              newdata = newdata[newdata$ids == i,], ...)))
+      }
+    }
+  } else {
+    ll <- apply_to_models(object, node = ids, FUN = logLik)
+  }
+  
+  
[…]

### 2019-11-05 pkg/extree_partykit/inst/guideliketest/simulation/ccprune.R

@@ -1,0 +1,263 @@
+#################################
+## Source file for replication of the simulation results presented in:
+## The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE (2019)
+## by Lisa Schlosser and Torsten Hothorn and Achim Zeileis
+##
+## Content: reimplementation of cost-complexity pruning
+#################################
+
+
+ccprune <- function(object, costfunction = NULL, nrfolds = 10, SE1rule = FALSE, seed = 7) 
+{
+  
+  cl <- match.call()
+  
+  # per default use residual sum of squares
+  if(is.null(costfunction)) costfunction <- "RSS"   
+  
+  if(costfunction == "RSS"){
+    cf <- function(predcoef, data, mean_error = FALSE){
+      if(mean_error) return(mean((data$y - (predcoef[1] + predcoef[2] * data$x))^2)) 
+      return(sum((data$y - (predcoef[1] + predcoef[2] * data$x))^2))
+    }
+  }
+  
+  if(is.function(costfunction)) cf <- costfunction
+  
+  ## function to evaluate cf for given node IDs (in the original tree 'object')
+  eval_cf <- function(object, nids, mean_error = FALSE){
+    cfsum <- 0
+    for(i in nids) {
+      node_cf <- cf(predcoef = coef(nodeprune(object[[i]], ids = 1)),
+                    data = object[[i]]$data, mean_error = mean_error)
+      if(mean_error) node_cf <- node_cf * NROW(object[[i]]$data) / NROW(object$data)
+      cfsum <- cfsum + node_cf
+    }
+    return(cfsum)
+  }
+  
+  
+  ###############################################################
+  ## 1st step: select finite step of subtrees which are candidates for the optimal tree
+  
+  alpha_all <- c(0)
+  tnids_all <- inids_all <- list()
+  
+  # ids of terminal nodes, all nodes and inner nodes
+  tnids <- nodeids(object, terminal = TRUE)
+  anids <- nodeids(object)
+  inids <- nodeids(object)[!(anids %in% tnids)]
+  tnids_all[[1]] <- tnids
+  inids_all[[1]] <- inids
+  
+  
+  while(length(tnids) > 1){
+    
+    alpha_min <- Inf
+    t_min_id <- integer()
+    
+    for(i in inids){
+      
+      # ids of terminal nodes of branch with root node i
+      bnids <- anids[anids %in% names(object[[i]]) & anids %in% tnids]
+      
[…]

### 2019-11-05 pkg/extree_partykit/inst/guideliketest/simulation/guidelike.R

@@ -1,0 +1,1211 @@
+#################################
+## Source file for replication of the simulation results presented in:
+## The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE (2019)
+## by Lisa Schlosser and Torsten Hothorn and Achim Zeileis
+##
+## Content: test functions (modifications of CTree and MOB, reimplementation of GUIDE, also with modifications)
+#################################
+
+
+##### reimplementation of testfunctions for GUIDE
+.guide_select <- function(...) {
+  function(model, trafo, data, subset, weights, whichvar, ctrl) {
+    args <- list(...)
+    ctrl[names(args)] <- args
+    
+    if(!"guide_parm" %in% names(ctrl)) ctrl$guide_parm <- NULL
+    if(!"guide_testtype" %in% names(ctrl)) ctrl$guide_testtype <- c("max", "sum", "coin")
+    if(!"interaction" %in% names(ctrl)) ctrl$interaction <- FALSE
+    if(!"weighted.scores" %in% names(ctrl)) ctrl$weighted.scores <- FALSE
+    if(!"xgroups" %in% names(ctrl)) ctrl$xgroups <- NULL
+    if(!"ygroups" %in% names(ctrl)) ctrl$ygroups <- NULL
+    if(!"guide_decorrelate" %in% names(ctrl)) ctrl$guide_decorrelate <- "vcov"
+    
+    if(ctrl$interaction){
+      partykit:::.select_int(model, trafo, data, subset, weights, whichvar, ctrl, FUN = .guide_test_int)
+    } else {
+      partykit:::.select(model, trafo, data, subset, weights, whichvar, ctrl, FUN = .guide_test)
+    }
+  }
+}
+
+
+
+.guide_test <- function(model, trafo, data, subset, weights, j, SPLITONLY = FALSE, ctrl) {
+  
+  # test function returning 'p.value' = log(1-pval) and 'statistic' = log(stat) of independence tests
+  
+  Y <- model$estfun  
+  
+  ## check control arguments
+  
+  if(length(ctrl$guide_testtype) > 1) ctrl$guide_testtype <- ctrl$guide_testtype[1]
+  if(!ctrl$guide_testtype %in% c("max", "sum", "coin")) stop("guide_testtype has to be one of the following options: max, sum, coin")
[…]

### 2019-11-05 pkg/extree_partykit/tests/bugfixes.R

@@ -1,0 +1,862 @@
+suppressWarnings(RNGversion("3.5.2"))
+
+set.seed(290875)
+
+datLB <-
+structure(list(Site = c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
+1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
+3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 
+4L, 4L, 4L, 4L, 5L, 5L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 6L, 6L, 6L, 
+6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
+7L, 7L, 7L, 7L, 7L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 
+8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 
+9L, 9L, 9L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 
+2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 
+3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
+4L, 5L, 5L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 
+6L, 6L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
+7L, 7L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 
+8L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 
[…]

### 2019-11-05 pkg/extree_partykit/inst/ERCIM-2011/titanic.R
create extree improvement branch
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2813 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6ea65118b722de54b5770b6466add9909cf65a11

@@ -1,0 +1,68 @@
+## packages
+library("partykit")
+library("rpart")
+library("RWeka")
+library("CHAID")
+library("evtree")
+
+## data
+data("Titanic", package = "datasets")
+ttnc <- as.data.frame(Titanic)
+ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
+names(ttnc)[2] <- "Gender"
+ttnc <- transform(ttnc, First = factor(Gender == "Female" | Age == "Child",
+  levels = c(FALSE, TRUE), labels = c("Male&Adult", "Female|Child")))
+
+## rpart
+rp <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
+plot(as.party(rp))
+
+## ctree
+ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
+  control = ctree_control(mincrit = 0.99, teststat = "max"))
+plot(ct)
+
+## J4.8
+j48 <- J48(Survived ~ Gender + Age + Class, data = ttnc)
+plot(as.party(j48))
+
+## QUEST
+qu <- pmmlTreeModel(file.path(system.file("pmml", package = "partykit"), "ttnc.pmml"))
+plot(qu)
+
+## CHAID
+chd <- chaid(Survived ~ Gender + Age + Class, data = ttnc)
+plot(chd)
+
+## evtree
+set.seed(1071)
+ev <- evtree(Survived ~ Gender + Age + Class, data = ttnc, maxdepth = 3)
+plot(ev)
+
+## performance
+mc <- function(obj) 1 - mean(predict(obj) == ttnc$Survived)
+evalfun <- function(obj) 2 * nrow(ttnc) * mc(obj) + width(obj) * log(nrow(ttnc))
+trees <- list(evtree = ev, rpart = as.party(rp), ctree = ct, J48 = as.party(j48))
+round(sapply(trees, function(obj) c(
+  "complexity" = width(obj),
+  "misclassification" = mc(obj),
+  "evaluation function" = evalfun(obj))), digits = 3)
+
+## compare predictions
+nd <- ttnc[rep(1:nrow(ttnc), 100), ]
+system.time(p1_rp <- predict(rp, newdata = nd, type = "class"))
+system.time(p2_rp <- predict(as.party(rp), newdata = nd))
+table(p1_rp, p2_rp)
+
+system.time(p1_j48 <- predict(j48, newdata = nd))
[…]

### 2019-11-05 pkg/extree_partykit/inst/guideliketest/simulation/simulation.R

@@ -1,0 +1,139 @@
+#################################
+## Replication code for the simulation results presented in:
+## The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE (2019)
+## by Lisa Schlosser and Torsten Hothorn and Achim Zeileis
+## (corresponding figures can be reproduced by function calls provided in figures.R)
+#################################
+
+library("partykit")
+library("Formula")
+library("parallel")
+
+# load source files
+source("guidelike.R")
+source("tests_guideprune.R")
+source("ccprune.R")
+
+
+
+#############
+### reproduction of simulation results
+
+# computation time for full reproduction of results (on our machines, using 3 kernels): 
+# 296 hours and 20 min consisting of
+# approximately 21 h for the "stump" scenario 
+# + 20 min for the full factorial analysis of the building blocks in the "stump" scenario 
+# + 275 h for the "tree" scenario.
+
+
+### "stump" scenario as discussed in Section 5.1. and in the Appendix
+sim_stump <- simwrapper(scenario = "stump",
+                        nobs = 250, nrep = 100, seed = 7,
+                        delta = seq(from = 0, to = 1, by = 0.1),
+                        xi = c(0, 0.2, 0.5, 0.8), 
+                        variation = c("all", "beta0", "beta1"),
+                        changetype = c("abrupt", "continuous"),
+                        test = c("ctree", "mfluc", "ctree_max",
+                                 "ctree_cat", "mfluc_cat", "ctree_max_cat",
+                                 "ctree_bin", "mfluc_bin", "ctree_max_bin",
+                                 "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
+                                 "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"))
+
+sim_stump <- sim_stump$res
+
+save(sim_stump, file = "sim_stump.rda")
+
+
+### "stump" scenario, full factorial analysis as discussed in Section 5.2.
+
+# for delta = 0.3 and xi = 0
+sim_3way1 <- sim(scenario = "stump",
+                 nobs = 250, nrep = 100, seed = 7,
+                 delta = 0.3,
+                 xi = 0, 
+                 changetype = "abrupt",
+                 variation = "all", 
+                 return_matrices = TRUE,
+                 test = c("guide_sum_1_cor",
[…]

### 2019-11-05 pkg/extree_partykit/inst/guideliketest/tests.R

@@ -1,0 +1,1966 @@
+library("partykit")
+library("Formula")
+library("parallel")
+source("~/svn/partykit/pkg/partykit/inst/guideliketest/guidelike.R")
+# source("~/partykit/inst/guideliketest/guidelike.R")
+
+# function to compute adjusted Rand Index
+adj_rand_index <- function(x, y) {
+  
+  tab <- table(x, y)
+  a <- rowSums(tab)
+  b <- colSums(tab)
+  
+  M <- sum(choose(tab, 2))
+  N <- choose(length(x), 2)
+  A <- sum(choose(a, 2))
+  B <- sum(choose(b, 2))
+  
+  c(ARI = (M - (A * B) / N) / (0.5 * (A + B) - (A * B) / N))
+}
+
+
+###########
+# DGP
+
+## FIX ME: 
+# for vary_beta = "all":
+# beta1 includes the effect of 2*delta, because otherwise (1*delta)
+# the values for z1>0 would all be smaller than for z1>0 and the other way round
+# -> similar scenario as for vary_beta = "beta0"
+# -> guide1 could easily find split
+
+dgp_stump <- function(nobs = 100, delta = 1, xi = 0.3,  
+                      # xi can also be a vector with 2 elements for nrsteps = 2
+                      sigma = 1, seed = 7, only_intercept = FALSE, 
+                      binary_regressor = FALSE, binary_beta = TRUE,
+                      vary_beta = c("all", "beta0", "beta1"),
+                      beta0 = NULL, beta1 = NULL,
+                      nrsteps = 1, nrlevels = 2, shift = 0,
+                      z1dist = c("unif", "norm"),
+                      nrsplitvar = 1 ){ # nrsplitvar not needed but to match form of dgp_tree
+  
+  if(nrsteps == 1 & nrlevels > 2) stop("for 1 step only two levels are possible")
+  if(nrsteps == 1 & length(xi)>1) stop("for 1 step only one split point xi can be handed over")
[…]

### 2019-11-05 pkg/extree_partykit/tests/ctree-ctree.R

@@ -1,0 +1,96 @@
+suppressWarnings(RNGversion("3.5.2"))
+
+### compare partykit::ctree and party::ctree wrt predictions
+### and computing time. But not during R CMD check
+
+library("partykit")
+library("party")
+
+set.seed(29)
+
+n <- 200
+dgp <- function(n, response = c("numeric", "ordered", "binary", "factor"), na = FALSE) {
+
+    response <- match.arg(response)
+    x <- runif(n * 10)
+    x <- matrix(x, ncol = 10)
+    x <- cbind(x, x[,1:2] + cbind(rnorm(n, sd = 0.1), rnorm(n, sd = 0.1)))
+    x1 <- sample(gl(4, n / 4))
+    x2 <- sample(ordered(gl(4, n / 4)))
+    y <- as.integer(interaction(x[,1] > 0.5, x[,2] > 0.25))
+    y <- rnorm(n, mean = y)
+    if (na) x[sample(1:(n * 10), n)] <- NA
+    if (response == "numeric")
+        return(data.frame(x, x1, x2, y = y))
+    if (response == "binary") {
+        y <- factor(y < median(y))
+        return(data.frame(x, x1, x2, y = y))
+    } else {
+        y <- cut(y, 
+            breaks = c(-Inf, quantile(y, c(0.25, 0.5, 0.75)), Inf))
+    }
+    if (response == "ordered")
+        return(data.frame(x, x1, x2, y = ordered(y)))
+    if (response == "factor")
+        return(data.frame(x, x1, x2, y = y))
+}
+
+args <- list(response = c("numeric", "ordered", "binary", "factor"),
+             na = c(FALSE, TRUE),
+             teststat = c("quad", "max"),
+             testtype = c("Bonferroni", "Univariate", "Teststatistic"),
+             mincriterion = c(0.8, 0.9, 0.95),
+             maxsurrogate = 0:2, maxdepth = 3)
+
+gr <- do.call("expand.grid", args)
+tgr <- gr
+gr[["response"]] <- as.character(gr[["response"]])
+gr[["teststat"]] <- as.character(gr[["teststat"]])
+gr[["testtype"]] <- as.character(gr[["testtype"]])
+
+### random splits are done differently in party::ctree
+gr <- subset(gr, !na | (na & maxsurrogate > 0))
+
+
+fun <- function(args) {
+
[…]

### 2019-11-12 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update draft (restructure and fixmes)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2865 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 39f5f0cd12ef4f88b3e71d82673e607ddb4ac692

@@
 
-\numberwithin{equation}{subsection}
+\numberwithin{equation}{section}
 
@@
 
-\title{Distributional Trees and Forests for Circular Data with an Application to Probabilistic Wind Direction Forecasting}
+\title{Circular Regression Trees and Forests with an Application to 
+    Probabilistic Wind Direction Forecasting}
 \Shorttitle{Distributional Trees and Forests for Circular Data}
@@
 distributional trees are easy to interpret, can detect non-additive effects, and select
-covariates and their interactions automatically. In order to stabelize and regularize
+covariates and their interactions automatically. In order to stabilize and regularize
 the model an ensemble of distributional trees is combined yielding a distributional
@@
 
+\fixme{revise literature survey with focus on transition to GLM}
+
 The named methods and also many of their further developments 
@@
 such as the type of link function and the form of linear combinations.
+
+\fixme{revise transition to trees/forests}
 
@@
 by distributional trees and forests such that a wide range of inference is made available.
+
+\fixme{include and revise ``motivating'' example}
+
+As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
+tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
+Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
+west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
+gradients to which valley wind regimes react both west and east of the airport are considered 
+as covariates along with other meteorological measurements at the airport (lagged by one hour),
+such as wind direction and wind speed at Innsbruck Airport.
+
+Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
+and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
+fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
+wind regimes:
+(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
+east to west, nodes 4 and 5). 
+(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
+Innsbruck (from south-east to north-west, nodes 7 and 8). 
+(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
+nodes 12, 14 and 15). 
+Node~11 captures observations with rather low wind speeds that cannot be distinguished
+clearly into wind regimes and consequently are associated with a very low
+estimated concentration $\hat \kappa$. In terms of covariates, the lagged
+wind direction (``persistence'') is mostly responsible for distinguishing
+the broad wind regimes listed above while the pressure gradients and wind
+speed separate between subgroups with high vs.\ low precision.
+Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
+increases clockwise from North ($0$ degree).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
+\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
+In each terminal node the empirical histogram (gray) and fitted density (red line)
+are depicted along with the estimated location parameter (red hand). The covariates
+employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
+and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
+all lagged by one hour.}
+\label{fig:tree_ibk} 
+\end{figure}
+
 The corresponding methodology and its features are presented in Section~\ref{sec:tree_forest} after
@@
 
-\section{Probabilistic Circular Modeling}
+\section{Probabilistic circular modeling}
 \label{sec:prob_circ}
@@
 Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
+\fixme{check domain of $\mu$ in literature}
 the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
@@
 \end{figure}
+\fixme{update figure 2d to 3d, using data of example tree}
 
@@
 
-\section{Regression Trees and Forests for Circular Data}
+\section{Circular regression trees and forests}
 \label{sec:tree_forest}
-%\subsection{Motivation}
+\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
 A common way of including covariates as regressors in distributional models is to apply
@@
 
-\subsection{Distributional Trees for Circular Data}
+\subsection{Circular trees}
 \label{sec:circtree}
@@
 
-\subsection{Distributional Forests for Circular Data}
+\subsection{Circular forests}
 \label{sec:circforest}
@@
 
-\section{Case Study: Wind Direction Forecasts}
+\section{Case study: Probabilistic wind direction forecasting}
 \label{sec:wind}
@@
 \fixme{information on already available approaches to wind direction forecasting}
-
-\subsection{Motivational Example}\label{sec:mot_ex}
-As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
-tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
-Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
-west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
-gradients to which valley wind regimes react both west and east of the airport are considered 
-as covariates along with other meteorological measurements at the airport (lagged by one hour),
-such as wind direction and wind speed at Innsbruck Airport.
-
-Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
-and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
-fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
-wind regimes:
-(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
-east to west, nodes 4 and 5). 
-(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
-Innsbruck (from south-east to north-west, nodes 7 and 8). 
-(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
-nodes 12, 14 and 15). 
-Node~11 captures observations with rather low wind speeds that cannot be distinguished
-clearly into wind regimes and consequently are associated with a very low
-estimated concentration $\hat \kappa$. In terms of covariates, the lagged
-wind direction (``persistence'') is mostly responsible for distinguishing
-the broad wind regimes listed above while the pressure gradients and wind
-speed separate between subgroups with high vs.\ low precision.
-Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
-increases clockwise from North ($0$ degree).
-
-\begin{figure}[t]
-\centering
-%\includegraphics[height = .5\textheight,angle=90,origin=c]{schlosserfig1.pdf}
-\includegraphics[width = \textwidth]{tree_ibk.pdf}
-\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
-In each terminal node the empirical histogram (gray) and fitted density (red line)
-are depicted along with the estimated location parameter (red hand). The covariates
-employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
-and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
-all lagged by one hour.}
-\label{fig:tree_ibk} 
-\end{figure}
 
@@
 
-\fixme{information on NAs}
+\fixme{information on NAs, update data information, include overview}
 
@@
 
-\subsection{Evaluation Setting and Models Applied}
+\subsection{Evaluation and models}
 Along with distributional trees and forests for circular data three other models have been applied
@@
 
+\fixme{condense evaluation}
+
 The circular CRPS is evaluated out of sample and the resulting values are averaged over 
@@
 
-\fixme{Comment on aggregation}
+\fixme{comment on aggregation}
 
@@
 is provided in the remainder of this section.
+
+\fixme{condense model descriptions}
 
@@
 
-\fixme{references}
+\fixme{add references}
 
@@
 
-\fixme{information on applied implementation here or in separate chapter?}
+\fixme{add information on applied implementation here (!) or in separate chapter?}
 
 \subsection{Results}
+\fixme{update results section}
 In order to test and compare the models over varying settings not only 1-hour but also 3-hour
@@
 %\hspace{-2cm}
-\includegraphics{boxplot_crpsraw.pdf}
+%\includegraphics{boxplot_crpsraw.pdf}
+\includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
 \caption{}
@@
 %\hspace{-2cm}
-\includegraphics{boxplot_crpsskill.pdf}
+%\includegraphics{boxplot_crpsskill.pdf}
+\includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
 \caption{}
@@
 
-\section{Discussion and Outlook}
+\section{Summary and conclusion}
 Distributional trees for circular responses are established by coupling
@@
 
-\fixme{Summary of results}
+\fixme{add summary of results}
 
-\fixme{Possible extensions of models, use of NWP data}
+\fixme{outlook on possible extensions of models, use of NWP data}
 

### 2019-11-13 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include captions, start to revise results, modify makefile
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2868 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 72fe968c18f11973e700543ec34dcf6356a417bf

@@
 \fixme{update results section}
-In order to test and compare the models over varying settings not only 1-hour but also 3-hour
-forecasts are considered at both airports yielding four different scenarios of which the results
-are discussed in this section.
+This section provides a detailed analysis of the different proposed statistal models with focus
+on their predictive performance. For a comprehensive comparison of the models, $1$\,hour and 
+$3$\,h wind direction forecasts are evaluated at two airports, namely Innsbruck and Vienna, with
+different meteorological site characteristics.
 
@@
   \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
-\caption{}
+  \caption{CRPS skill scores for wind direction forecasts based on the full
+    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
+    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
+    per $12$~months averaged over the $5$~validation years; this yields a total of
+    $288$~yearly mean values. The scores are shown for the reference models climatology, 
+    persistence and the linear model, as well as for the circular regression tree and
+    forest.}
   \label{fig:boxplot_crpsraw} 
@@
   \includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
-\caption{}
+  \caption{As Figure~\ref{fig:boxplot_crpsraw}, but showing CRPS skill scores
+    with the climatology model as reference. Skill scores are in percent; positive
+    values indicate improvements over the reference.} 
   \label{fig:boxplot_crpsskill} 

### 2019-11-13 pkg/circtree/inst/wind_paper/draft/draft_2019_11_12.tex
Add tex file for 2019-11-12 for 'make diff'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2869 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 23f3420a322e3e1a175904949ddad9b9d769a6ad

--- 2019-11-12 pkg/circtree/inst/wind_paper/draft/draft.Rnw
@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
 
@@
 \numberwithin{equation}{section}
-
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-library("disttree")
-library("circtree")
-
-## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-# set function for parallelization
-applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))
-
-@
 
@@
 The proposed methods are in the \textsf{R} package \textbf{circtree}
-(version~\Sexpr{packageVersion("circtree")}) based on the \textbf{disttree}
-package (version~\Sexpr{packageVersion("disttree")}) which applies the main
+(version~0.1.0) based on the \textbf{disttree}
+package (version~0.2.0) which applies the main
 tree building function from the  \textbf{partykit} package 
-(version~\Sexpr{packageVersion("partykit")}), all three available 
+(version~1.2.5), all three available 
 on \textsf{R}-Forge at

### 2019-11-13 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update plot, makefile, revise draft (partly 4.2 and 4.3)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2873 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 56644a1584918bfeda066dd1e229c8159def3c3a

@@
 
-\subsection{Evaluation and models}
-Along with distributional trees and forests for circular data three other models have been applied
-in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
-the five years contained in the data set being used as testing data once after the models where 
-learned on the remaining four years.
+\subsection{Models and evaluation}\label{sec:wind:models}
+For a fair evaluation of circular regression trees and forest and to
+investigate whether circular trees and forests can be applied as a reasonable
+alternative to already existing approaches, three additional statistical models are
+applied in this study to probabilistically forecast wind directions. Two
+approaches are based on existing approaches in the field of meteorological
+forecasting, one approach is a state-of-the-art model to forecast circular
+variables.  The next paragraphs introduce the alternative approaches and
+provide for all statistical models, including the circular tree and forest, the
+model specification employed in this study.
 
-In order to compare the performances of the employed models a circular version of the continuous 
-ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
-Just as the linear version of the CRPS \citep[for more details see][]{Hersbach:2000} it measures the 
-difference between an observation and the corresponding predicted distribution function in order 
-to assess the probabilistic goodness of fit of the fitted model. Hence, the lower the CRPS value 
-the better the performance. In order to adjust the general methodology of the CRPS to circular data 
-\cite{Grimit+Gneiting+Berrocal:2006} employ an auxiliary function which measures the angular distance 
-between two circular variables.
-
-\fixme{condense evaluation}
-
-The circular CRPS is evaluated out of sample and the resulting values are averaged over 
-each observation contained in the testing data and the corresponding predicted distribution 
-parameters of each of the models.
-
-\fixme{comment on aggregation}
-
-To measure the improvement of one of the applied methods over a climatology model (which will 
-be described in detail subsequently) the skill score based on the circular 
-CRPS of the assessed method is evaluated with the climatology model used as reference:
-\begin{equation}
-\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
-\end{equation}
-
-Three commonly used methods are selected as reference models in order to investigate whether 
-distributional trees and forests can be applied as a reasonable alternative to already
-existing approaches to modeling circular data, particularly in the field of meteorological forecasting.
-The exact settings of the tree and forest models as well as a specification of the reference models
-is provided in the remainder of this section.
-
-\fixme{condense model descriptions}
-
-\paragraph{Distributional Tree}
+\paragraph{Circular regression tree}
 For the distributional tree as discussed in Section~\ref{sec:circtree} all covariates provided in the 
@@
 
-\paragraph{Distributional Forest}
+\paragraph{Circular regression forest}
 Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
@@
 
-\paragraph{Persistence}
+\paragraph{Persistency model}
 Assuming that weather conditions will remain the same after a certain period of time is the
@@
 the corresponding learning data.
+\fixme{not done like this, always using the whole data set. wrong?}
 
-\paragraph{Climatology}
+\paragraph{Climatological model}
 If there are no climatological changes within a certain number of years it can be expected that 
@@
 
-\paragraph{Generalized Linear Model}
+\paragraph{Generalized linear model}
 Another commonly applied approach is to linearly regress the circular variable of interest on
@@
 
-\fixme{add references and further information on the estimation process within circGLM?}
+%\fixme{add references and further information on the estimation process within circGLM?}
 
-\fixme{add information on applied implementation here (!) or in separate chapter?}
+%\fixme{add information on applied implementation here (!) or in separate chapter?}
+
+To compare the predictive performances of all proposed models a circular
+analogue of the continuous ranked probability score (CRPS) as introduced by
+\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
+the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
+rule \citep{Gneiting+Raftery:2007} and measures the difference between an
+observation and the corresponding predicted distribution function in order to
+assess the probabilistic goodness of fit for the estimated model. Hence, the lower
+the CRPS value the better the performance.  Contrary to the linear version, the
+circular CRPS reduces not to the absolute error but to the angular distance
+when the forecast is deterministic.
+
+In addition to the raw CRPS, to assess differences in the
+improvement of the various statistical models over the climatological model
+used as a reference method, CRPS-based skill scores are computed according to
+\begin{equation} 
+  \text{CRPSS}_{\text{method}} = 1 -
+\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
+\end{equation}
+
+All scores presented in the next section are out-of-sample: For the persistency
+model only time points prior to the time of interest are used, for all other
+models a five-fold cross-validation is employed using up to four calendar years
+for model training and the remaining single calendar year for validation. Due to the large sample
+size of 24-hourly values per day over five years, some kind of temporal
+aggregation is needed to ensure a correct visual comparison of the individual
+methods. As for the employed models the variablilty of the predictive performance over the five years
+is lower than over a single day or over a single year, CRPS and CRPS skill scores are
+aggregrated over the respective five validation years; this yields 24 hourly
+scores per 12 months averaged over the five validation years.
 
 \subsection{Results} 
-\fixme{update results section}
-This section provides a detailed analysis of the different proposed statistal models with focus
-on their predictive performance. For a comprehensive comparison of the models, $1$\,hour and 
-$3$\,h wind direction forecasts are evaluated at two airports, namely Innsbruck and Vienna, with
-different meteorological site characteristics.
+This section provides a detailed analysis on the predictive performance of the
+different proposed statistal models applied to probabilistic wind direction
+forecasting. To ensure a comprehensive comparison of the models, wind direction
+forecasts are evaluated for two different lead times at two airports with
+different climatological site characteristics.
+Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
+at forecast steps 1\,h and 3\,h for the airports Innsbruck (panels a,\,c) and
+Vienna (panels b,\,d). The scores are aggregated over the $5$~validation years,
+yielding yearly mean values for every hour per calendar month, with a lower
+score indicating better performance.  For both stations at both forecast steps,
+the circular regression forest overall provides the best predictive
+performance, followed by the circular regression tree and, except for the
+$3$\,h forecast at Innsbruck, the persistency model.  In comparison to the
+circular tree and forest, for both stations and forecast steps, the
+climatological model and the linear model show clearly higher CRPS scores and
+hence a lower predictive performance.
 
-Figure~\ref{fig:boxplot_crpsraw} illustrates the CRPS values of the employed models with a lower 
-CRPS indicating better performance. 
-In all four settings it can be observed that on average the distributional forest provides the best 
-fitting probabilistic predictions followed by the distributional tree. Especially in Vienna the 
-persistence (hour) model is only slightly behind while the linear model and the climatology
-show clearly higher CRPS values.
+The different site characteristics between the airports Innsbruck
+(Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
+(Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
+level of the respective model performances and on their respective performance
+variation. At Innsbruck, due to the surrounding mountains a limited number of
+possible wind directions exists, namely the three wind regimes discussed for
+Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
+Innsbruck the wind direction remains rather constant in one of these possible
+states, but once a change takes place it is mostly a major wind direction
+shift.  Hence, due to the few wind regimes the rather unflexible climatological
+and linear models perform relatively similar to the other models
+(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the potential
+high prediction errors in case of a paradigma change seem to lead to an higher
+variation in the predictive performance for all models in comparison to Vienna;
+this variation is especially high for the persistency model due to its strong
+vulnerability to abrupt wind shifts. On the contrary, at Vienna the geographic
+surroundings do not constrain any wind direction which leads to smaller and
+less abrupt changes in the wind direction as well as less pronounced wind
+regimes. This seems to weaken the predictive performance of the climatological
+and linear models, and to reduce the performance variability for all models
+(Figure~\ref{fig:boxplot_crpsraw}b,\,d).
 
-Comparing the performances over the different locations the narrower boxes in Vienna represent
-a lower variation. This can be explained by more abrupt changes of wind directions in Innsbruck
-due to the surrounding mountains channeling wind and in that way allowing only for a limited
-number of possible wind directions (mainly the three wind regimes as discussed in the motivational
-example in Section~\ref{sec:mot_ex}). For that reason, relatively stable models such as persistency
-produce well fitting forecasts as long as wind directions remain in one of these possible states,
-however, once a change takes place it can only be an abrupt and major shift leading to high
-prediction errors as supported by the wide boxes of the persistence model in Innsbruck. In Vienna,
-on the other hand, changes can also appear in a smaller extend as the geographic surroundings do not
-constrain wind directions. This wider range of possible observations might also weaken the performance
-of the climatology as expressed by the higher CRPS values in Vienna.
-
-The impact of the different forecasting time windows of 1 and 3 hours can hardly be noticed, especially
-as it does not seem to effect the order of which method performs best. But taking a closer look
-at the CRPS values of the tree, forest and persistency reveals lower scores and hence a better 
-performance for the 1-hour forecasts as it could have been expected due to increasing uncertainties
-in a longer time window.
+The different forecast steps seem to have only a minor effect on the predictive
+performance of the climatological model and the linear model for both stations.
+However, for the persistency model, at both stations, higher scores for the
+$3$\,h forecast (panels c,\,d) reveal a lower performance for longer lead times. This is
+clearly due to at least 3-hourly instead of 1-hourly lagged response values
+employed as covariates in the persistency model. Apparently, the circular
+regression tree and forest can partially compensate the lower skill of
+the lagged response values by other covariates, hence their predictive
+perfomance only slightly decreases for the 3-hourly in comparison to the
+1-hourly wind direction forecast. This compensation is especially evident for
+Innsbruck, where the performance difference between the persistency model and
+the tree-based methods significantly increases for the 3-hourly forecast, with
+a significantly better performance for the circular tree and forest.
 
@@
 
-With the aim of excluding any possible influence of the selection of testing and training data
-the performances of the methods are also compared based on the CRPS skills score as illustrated in 
-Figure~\ref{fig:boxplot_crpsskill}. Using the climatology as reference model the values on the y-axis
-indicate the percentage of improvement of each model compared to the reference model. Overall
-the same conclusions can be drawn as from Figure~\ref{fig:boxplot_crpsraw} with the distribution
-forest leading to the highest overall improvement in model fit.
+In addition to the raw CRPS, Figure~\ref{fig:boxplot_crpsskill} provides CRPS
+skill scores with the climatological model as a reference. Skill scores are in
+percent, and positive values indicate an improvement in the predictive
+performance over the reference. For all setups, the circular regression forest
+has the highest skill scores with a mean performance gain of 13--25\% and 58--71\% for
+Innsbruck and Vienna, respectively. As discussed for
+Figure~\ref{fig:boxplot_crpsraw}, this improvement over the climatological
+model is lower for Innsbruck due to the few number of predominat wind regimes
+and hence a relatively good performance of the climatological model.
+Additionally, Figure~\ref{fig:boxplot_crpsskill} shows that even if the
+persistency model's performance is lower than the reference (panel c) the
+tree-based models are significantly superior to the climatolical model used as
+reference.
 

### 2019-11-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Spell check, minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2874 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 829528fe7fe5d19d183731dd82fe3532ff29c345

@@
 observation and the corresponding predicted distribution function in order to
-assess the probabilistic goodness of fit for the estimated model. Hence, the lower
-the CRPS value the better the performance.  Contrary to the linear version, the
-circular CRPS reduces not to the absolute error but to the angular distance
-when the forecast is deterministic.
+assess the probabilistic goodness of fit for the estimated model. Hence, the
+lower the CRPS value the better the predictive performance.  Contrary to the
+linear version, the circular CRPS reduces not to the absolute error but to the
+angular distance when the forecast is deterministic.
 
@@
 improvement of the various statistical models over the climatological model
-used as a reference method, CRPS-based skill scores are computed according to
+used as a reference, CRPS-based skill scores are computed according to
 \begin{equation} 
@@
 models a five-fold cross-validation is employed using up to four calendar years
-for model training and the remaining single calendar year for validation. Due to the large sample
-size of 24-hourly values per day over five years, some kind of temporal
-aggregation is needed to ensure a correct visual comparison of the individual
-methods. As for the employed models the variablilty of the predictive performance over the five years
-is lower than over a single day or over a single year, CRPS and CRPS skill scores are
-aggregrated over the respective five validation years; this yields 24 hourly
-scores per 12 months averaged over the five validation years.
+for model training and the remaining single calendar year for validation. Due
+to the large sample size of 24-hourly values per day over five years, some kind
+of temporal aggregation is needed to ensure a correct visual comparison of the
+individual methods. As for the employed models the variability of the
+predictive performance over the five years is lower than over a single day or
+over a single year, CRPS and CRPS skill scores are aggregated over the
+respective five validation years; this yields 24 hourly scores per 12 months
+averaged over the five validation years.
 
@@
 This section provides a detailed analysis on the predictive performance of the
-different proposed statistal models applied to probabilistic wind direction
+different proposed statistical models applied to probabilistic wind direction
 forecasting. To ensure a comprehensive comparison of the models, wind direction
@@
 states, but once a change takes place it is mostly a major wind direction
-shift.  Hence, due to the few wind regimes the rather unflexible climatological
-and linear models perform relatively similar to the other models
-(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the potential
-high prediction errors in case of a paradigma change seem to lead to an higher
-variation in the predictive performance for all models in comparison to Vienna;
-this variation is especially high for the persistency model due to its strong
-vulnerability to abrupt wind shifts. On the contrary, at Vienna the geographic
-surroundings do not constrain any wind direction which leads to smaller and
-less abrupt changes in the wind direction as well as less pronounced wind
-regimes. This seems to weaken the predictive performance of the climatological
-and linear models, and to reduce the performance variability for all models
-(Figure~\ref{fig:boxplot_crpsraw}b,\,d).
+shift.  Hence, due to the few wind regimes the rather inflexible climatological
+and linear models score relatively well with similar values as the other models
+(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the
+potential high prediction errors in case of a paradigm shift seem to lead to
+an higher variation in the predictive performance for all models in comparison
+to Vienna; this variation is especially high for the persistency model due to
+its strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
+geographic surroundings do not constrain any wind direction which leads to
+smaller and less abrupt changes in the wind direction as well as less
+pronounced wind regimes. This seems to weaken the predictive performance of the
+climatological and linear models, and to reduce the performance variability for
+all models (Figure~\ref{fig:boxplot_crpsraw}b,\,d).
 
 The different forecast steps seem to have only a minor effect on the predictive
-performance of the climatological model and the linear model for both stations.
+performance of the climatological model and the linear model at both stations.
 However, for the persistency model, at both stations, higher scores for the
-$3$\,h forecast (panels c,\,d) reveal a lower performance for longer lead times. This is
-clearly due to at least 3-hourly instead of 1-hourly lagged response values
-employed as covariates in the persistency model. Apparently, the circular
-regression tree and forest can partially compensate the lower skill of
+$3$\,h forecast (panels c,\,d) reveal a lower performance for longer lead
+times; this is probably due to minimum 3-hourly instead of 1-hourly lagged
+response values employed as covariates in the persistency model. The circular
+regression tree and forest seem to partially compensate for the lower skill of
 the lagged response values by other covariates, hence their predictive
-perfomance only slightly decreases for the 3-hourly in comparison to the
-1-hourly wind direction forecast. This compensation is especially evident for
-Innsbruck, where the performance difference between the persistency model and
-the tree-based methods significantly increases for the 3-hourly forecast, with
-a significantly better performance for the circular tree and forest.
+performance only slightly decreases for the longer lead time.  This compensation
+is especially evident for Innsbruck, where the performance difference between
+the persistency model and the tree-based methods significantly increases from
+the 1-hourly to the 3-hourly forecast.
 
@@
 Figure~\ref{fig:boxplot_crpsraw}, this improvement over the climatological
-model is lower for Innsbruck due to the few number of predominat wind regimes
+model is lower for Innsbruck due to the few number of predominant wind regimes
 and hence a relatively good performance of the climatological model.
@@
 persistency model's performance is lower than the reference (panel c) the
-tree-based models are significantly superior to the climatolical model used as
+tree-based models are significantly superior to the climatological model used as
 reference.

### 2019-11-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revisions on motivational example and introduction to trees
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2875 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 03ce88331d21eef3a914f85febae014619c07f17

@@
 
-As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
+As a motivational example Figure~\ref{fig:tree_ibk} displays a distributional
 tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
@@
 (2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
-Innsbruck (from south-east to north-west, nodes 7 and 8). 
+Innsbruck (from south-east to north-west, node 8). 
 (3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
-nodes 12, 14 and 15). 
-Node~11 captures observations with rather low wind speeds that cannot be distinguished
+nodes 10, 12 and 13). 
+Node~7 captures observations with rather low wind speeds that cannot be distinguished
 clearly into wind regimes and consequently are associated with a very low
@@
 
-The corresponding methodology and its features are presented in Section~\ref{sec:tree_forest} after
-discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
-In Section~\ref{sec:wind} the presented models are applied to probabilistic forecasting of wind direction
-at the airports of Innsbruck and Vienna.
+A more extensive case study of circular regression trees and forests applied to probabilistic 
+forecasting of wind direction at the airports of Innsbruck and Vienna is presented in 
+Section~\ref{sec:wind} where their performance is also compared to other commonly used models.
+Before that, distributional models for circular data are discussed in Section~\ref{sec:prob_circ} 
+and the methodology of circular regression trees and forest and its features are presented in 
+Section~\ref{sec:tree_forest}.
 
@@
 \fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
-A common way of including covariates as regressors in distributional models is to apply
-generalized additive models for location, scale, and shape 
-\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
-by its own generalized additive model (GAM).
-As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
-for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
-specifying a model properly can be very challenging, particularly for a high number of covariates 
-and no information on possible interactions.
+A common way of including covariates is to regress the model parameters on the covariates.
+In the case of a probabilistic model for a circular response this corresponds to expressing the 
+location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
+by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
+appropriate link function plugged in as in GLMs \citep{Fisher+Lee:1992, Fisher:1993}.
+In a more general setting this idea can be extended by generalized additive models for location, scale, 
+and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
+by its own generalized additive model (GAM). However, an adaption of this methodology to circular data
+can not be obtained by the straight forward approach of simply wrapping the modelled additive effects
+around the unit circle since the parameters of the von Mises distribution and their dependence/interaction 
+differs strongly from the linear case and therefore requires specific handling.
+Moreover, difficulties might arise in case of non-additive changes and specifying a model properly can 
+be very challenging, particularly for a high number of covariates and no information on possible 
+interactions.
 An alternative regression approach that can deal with these problems is to apply a tree-structured 

### 2019-11-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Small revisions (Lisa and Moritz)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2876 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 78d452c84de6c440bff6b61bad3f8ad865d95fa7

@@
 
-As a motivational example Figure~\ref{fig:tree_ibk} displays a distributional
-tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
-Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
-west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
-gradients to which valley wind regimes react both west and east of the airport are considered 
-as covariates along with other meteorological measurements at the airport (lagged by one hour),
-such as wind direction and wind speed at Innsbruck Airport.
+As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional
+tree applied to predict the wind direction at the airport of Innsbruck in the
+following hour.  Innsbruck lies at the bottom of a deep valley in the Alps.
+Topography channels wind along the west-east valley axis or along a tributary
+valley intersecting from the south. Hence, pressure gradients to which valley
+wind regimes react are considered as covariates along with other meteorological
+measurements at the airport (lagged by one hour), such as wind direction and
+wind speed at the airport.
 
@@
 
-A more extensive case study of circular regression trees and forests applied to probabilistic 
-forecasting of wind direction at the airports of Innsbruck and Vienna is presented in 
-Section~\ref{sec:wind} where their performance is also compared to other commonly used models.
-Before that, distributional models for circular data are discussed in Section~\ref{sec:prob_circ} 
-and the methodology of circular regression trees and forest and its features are presented in 
-Section~\ref{sec:tree_forest}.
+A more extensive case study of circular regression trees and forests applied to
+probabilistic forecasting of wind direction at the airports of Innsbruck and
+Vienna is presented in Section~\ref{sec:wind} where their performances are
+benchmarked to alternative commonly used approaches.  Beforehand,
+distributional models for circular data are discussed in
+Section~\ref{sec:prob_circ} and the methodology of circular regression trees
+and forest and its features are presented in Section~\ref{sec:tree_forest}.
+After the case study, a comprehensive summary and conlusion is given in
+Section~\ref{sec:summary}.
 
@@
 \subsection{Data}
-For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
-such as wind direction, temperature, air pressure and humidity
-measured either directly at the airports or at surrounding weather stations are used as 
-covariates in order to predict the wind direction at the corresponding airport 
-in the following hour(s).
-Next to the current measurements at each of the weather stations also differences in these 
+As response variables, $10$\,min mean wind direction observations are employed at
+the airports Innsbruck and Vienna on an hourly temporal resolution. As
+predictor variables, we use observations of various meteorological quantities
+such as wind direction, temperature, air pressure and humidity directly at the
+respective airports, as well as within the vicinity of the airports. For
+Innsbruck, meteorological variables are observed at XXX different locations
+along the intersecting valleys; whereas for Vienna, located in the flat,
+meteorological variables are observed at XXX different locations within an area
+of about $10$\,km$^2$. In addition, we employ temporal differences, averages, minima, maxima over the last 3\,h,
+
+of ias well as tempat each of the weather stations also differences in these 
 quantities between the considered stations are included. For example, differences in air 
@@
 For a fair evaluation of circular regression trees and forest and to
-investigate whether circular trees and forests can be applied as a reasonable
-alternative to already existing approaches, three additional statistical models are
-applied in this study to probabilistically forecast wind directions. Two
-approaches are based on existing approaches in the field of meteorological
-forecasting, one approach is a state-of-the-art model to forecast circular
-variables.  The next paragraphs introduce the alternative approaches and
-provide for all statistical models, including the circular tree and forest, the
-model specification employed in this study.
+investigate whether they can be applied as a reasonable alternative to already
+existing approaches, three additional statistical models are employed in this
+study for probabilistic forecasting of wind directions. Two of them are based
+on existing approaches in the field of meteorological forecasting while the
+third is a state-of-the-art model to forecast circular response variables. The
+next paragraphs introduce the reference methods and provide the corresponding
+model specifications as employed in this study for all considered model
+including the circular tree and forest.
 
@@
 
-In addition to the raw CRPS, to assess differences in the
-improvement of the various statistical models over the climatological model
-used as a reference, CRPS-based skill scores are computed according to
+In addition to the raw CRPS, CRPS-based skill scores are computed to assess
+differences in the improvement of the various statistical models over the
+climatological model used as a reference:
 \begin{equation} 
@@
 for model training and the remaining single calendar year for validation. Due
-to the large sample size of 24-hourly values per day over five years, some kind
+to the large sample size of 24 hourly values per day over five years, some kind
 of temporal aggregation is needed to ensure a correct visual comparison of the
-individual methods. As for the employed models the variability of the
+individual methods. As observed for the employed models, the variability of the
 predictive performance over the five years is lower than over a single day or
-over a single year, CRPS and CRPS skill scores are aggregated over the
-respective five validation years; this yields 24 hourly scores per 12 months
+over a single year. Hence, CRPS and CRPS skill scores are aggregated over the
+respective five validation years which yields 24 hourly scores per month
 averaged over the five validation years.
@@
 (Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
-level of the respective model performances and on their respective performance
+level of the model performances and on their respective performance
 variation. At Innsbruck, due to the surrounding mountains a limited number of
@@
 potential high prediction errors in case of a paradigm shift seem to lead to
-an higher variation in the predictive performance for all models in comparison
+a higher variation in the predictive performance for all models in comparison
 to Vienna; this variation is especially high for the persistency model due to
@@
     airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
-    per $12$~months averaged over the $5$~validation years; this yields a total of
-    $288$~yearly mean values. The scores are shown for the reference models climatology, 
-    persistence and the linear model, as well as for the circular regression tree and
+    per for each of the $12$~months averaged over the $5$~validation years which yields a total of
+    $288$~yearly mean values. The scores are shown for the climatological, 
+    persistency and the linear model as well as for the circular regression tree and
     forest.}
@@
 
-\section{Summary and conclusion}
+\section{Summary and conclusion}\label{sec:summary}
 Distributional trees for circular responses are established by coupling

### 2019-11-18 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added two references on circular trees; revision of section 1 and introductions of sections 3 and 4
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2877 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 59239c84ae2c865e9bf371494133fec4fbb77d28

@@
 regress the parameter(s) of the von Mises distribution on covariates.
-
 One of the first regression models with a circular response variable and linear covariates
@@
 introducing another specific transformation function which maps the values of the real line to the 
-unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter, to be modeled not only in separate models but also in a 
-mixed model. There, within an iterative optimization algorithm the two parameters are estimated and 
-improved stepwise until convergence. To obtain proper initial values it is recommended to apply
-two separate models where either one of the parameters is fixed while the other is modeled depending
-on covariates. While being built on well elaborated theory this methodology and the contained 
+unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
+on covariates, not only in separate models but also in a mixed model. There, within an iterative 
+optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
+To obtain proper initial values it is recommended to apply two separate models where either one of 
+the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
+While being built on well elaborated theory this methodology and the contained 
 optimization problem have shown to be very challenging in their application, mainly due to the 
-complexity encountered in optimizing the log-likelihood function that is not globally concave, 
-and thus bear a high risk of not converging if not applied thoughtfully 
-\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
-An alternative approach to regression modeling of circular data that avoids this strong dependence 
-on choosing well-fitting initial values is presented by \cite{Mulder+Klugkist:2017} employing a GLM 
+complexity encountered in optimizing the log-likelihood function which is not globally concave 
+\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, the selection of 
+starting values is crucial and bears a high risk of the model not to converge if not done thoughtfully. 
+In order to avoid this strong dependence on choosing well-fitting initial values \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM
 in a Bayesian framework.
 
-\fixme{revise literature survey with focus on transition to GLM}
+%\fixme{revise literature survey with focus on transition to GLM}
 
@@
 \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
-While they provide a powerful toolbox for modeling circular data, particularly in case of
-smooth effects, they may not be able to capture all types of effects. For example, abrupt changes 
-might not be detected properly owing to the enforced additive structure and model assumptions 
-such as the type of link function and the form of linear combinations.
+While they generally provide a powerful toolbox for modeling particularly smooth effects, 
+it has to be considered that representing these smooth effects on the unit circle
+can be challenging. Moreover, the discussed additive models may not be able to capture all 
+types of effects adequately. For example, abrupt changes might not be detected properly owing 
+to the enforced additive structure and model assumptions such as the type of link 
+function and the form of linear combinations.
 
-\fixme{revise transition to trees/forests}
+An alternative modeling approach that can deal with non-additive effects is provided
+by tree-based methods. \cite{Lund:2002} introduced the concept of circular regression trees 
+which was further explored by \cite{Larraondo+Inza+Lozano:2018}. This methodology
+enables splits leading to subsets which contain the ``origin point'' $0$ equal to $2\pi$
+and in that way allows for a proper representation of periodicity.
 
-Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
-forests to circular data, specifically by employing the von Mises distribution within the methodology
-of distributional trees and distributional forests following \cite{Schlosser+Hothorn+Stauffer:2019}. 
+Following the idea of employing tree-based methods but including also the probabilistic concept
+of regressing distribution parameters on covariates we propose a very flexible approach 
+by applying the von Mises distribution within the methodology of distributional trees and 
+distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular data.
 In that way no model assumptions are required as covariates and their possible interactions are
@@
 
-\fixme{include and revise ``motivating'' example}
+As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional tree applied to 
+predict wind directions at the airport of Innsbruck in the following hour. Wind direction is 
+a classical circular quantity and accurate forecasts are of great importance for decision-making 
+processes and risk management in air traffic management. Innsbruck lies at the bottom of a 
+deep valley in the Alps where topography channels wind along the west-east valley axis or along 
+a tributary valley intersecting from the south. Hence, pressure gradients to which valley wind 
+regimes react are considered as covariates along with other meteorological measurements at the 
+airport (lagged by one hour), such as wind direction and wind speed at the airport.
 
-As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional
-tree applied to predict the wind direction at the airport of Innsbruck in the
-following hour.  Innsbruck lies at the bottom of a deep valley in the Alps.
-Topography channels wind along the west-east valley axis or along a tributary
-valley intersecting from the south. Hence, pressure gradients to which valley
-wind regimes react are considered as covariates along with other meteorological
-measurements at the airport (lagged by one hour), such as wind direction and
-wind speed at the airport.
-
-Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
-and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
-fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
-wind regimes:
+Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the 
+empirical (gray) and fitted von Mises (red) distribution of wind direction in each terminal 
+node. Based on the fitted location parameters $\hat \mu$, the subgroups can be distinguished 
+into the following wind regimes:
 (1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
@@
 Node~7 captures observations with rather low wind speeds that cannot be distinguished
-clearly into wind regimes and consequently are associated with a very low
-estimated concentration $\hat \kappa$. In terms of covariates, the lagged
-wind direction (``persistence'') is mostly responsible for distinguishing
-the broad wind regimes listed above while the pressure gradients and wind
-speed separate between subgroups with high vs.\ low precision.
-Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
-increases clockwise from North ($0$ degree).
+clearly into wind regimes and consequently are associated with a very low estimated 
+concentration $\hat \kappa$. In terms of covariates, the lagged wind direction 
+(``persistence'') is mostly responsible for distinguishing the broad wind regimes 
+listed above while the pressure gradients and wind speed separate between subgroups 
+with high vs.\ low precision. Note that in the meteorological context wind direction 
+is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
 
@@
 \includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
-\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
+\caption{Fitted tree based on the von Mises distribution for wind direction forecasting 
+at the airport of Innsbruck for the following hour.
 In each terminal node the empirical histogram (gray) and fitted density (red line)
@@
 employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
-and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
+and pressure gradients ($\text{dpressure; hPa}$) west, east and south of the airport,
 all lagged by one hour.}
@@
 A more extensive case study of circular regression trees and forests applied to 
-probabilistic forecasting of wind direction at the airports of Innsbruck and
-Vienna is presented in Section~\ref{sec:wind} where their performances are
-benchmarked to alternative commonly used approaches.  Beforehand,
-distributional models for circular data are discussed in
-Section~\ref{sec:prob_circ} and the methodology of circular regression trees
-and forest and its features are presented in Section~\ref{sec:tree_forest}.
-After the case study, a comprehensive summary and conlusion is given in
+probabilistic forecasting of wind direction at the airports of Innsbruck and Vienna 
+is presented in Section~\ref{sec:wind} where their performances are benchmarked to 
+alternative commonly used approaches.  Beforehand, distributional models for circular 
+data are discussed in Section~\ref{sec:prob_circ} and the methodology of circular 
+regression trees and forest and its features are presented in Section~\ref{sec:tree_forest}.
+After the case study, a comprehensive summary and conclusion is given in
 Section~\ref{sec:summary}.
@@
 \label{sec:tree_forest}
-\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
-A common way of including covariates is to regress the model parameters on the covariates.
+%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
+A common way of including covariates is to regress the model parameters on covariates.
 In the case of a probabilistic model for a circular response this corresponds to expressing the 
@@
 by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
-appropriate link function plugged in as in GLMs \citep{Fisher+Lee:1992, Fisher:1993}.
-In a more general setting this idea can be extended by generalized additive models for location, scale, 
-and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
-by its own generalized additive model (GAM). However, an adaption of this methodology to circular data
-can not be obtained by the straight forward approach of simply wrapping the modelled additive effects
-around the unit circle since the parameters of the von Mises distribution and their dependence/interaction 
-differs strongly from the linear case and therefore requires specific handling.
-Moreover, difficulties might arise in case of non-additive changes and specifying a model properly can 
-be very challenging, particularly for a high number of covariates and no information on possible 
-interactions.
+appropriate link function plugged in as in GLMs. This kind of approach is presented for example
+in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. While applying the mixed model 
+requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
+\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
+a circular response variable.
+Apart from the methodology developed specifically for circular data more general approaches to 
+regressing model parameters on covariates are commonly used such as for example generalized additive 
+models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
+where each distribution parameter is estimated by its own generalized additive model (GAM). 
+However, an adaption of this methodology to circular data can be very complex due to a
+challenging representation of smooth effects on the unit circle.
+%However, an adaption of this methodology to circular data can not be obtained by the straight 
+%forward approach of simply wrapping the modeled additive effects
+%around the unit circle since the parameters of the von Mises distribution and their 
+%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
+Moreover, difficulties might arise in case of non-additive changes and specifying a model 
+properly can be a very tough task, particularly for a high number of covariates and no information
+on possible interactions.
 An alternative regression approach that can deal with these problems is to apply a tree-structured 
@@
 \label{sec:wind}
-
-Wind is a classical circular quantity and accurate forecasts of wind direction
-are of great importance for decision-making processes and risk management,
-e.g., in air traffic management or renewable energy production.
-
-\fixme{information on already available approaches to wind direction forecasting}
-
+The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
+employed for probabilistic forecasting of wind directions at the airport of Innsbruck. As explained
+and also supported by the results illustrated in the terminal nodes of the tree in 
+Figure~\ref{fig:tree_ibk} the geographic location of Innsbruck and particularly the surrounding
+mountain chains are a strongly influential component of the behavior of wind directions.
+While this provides an interesting and challenging environment for testing different
+statistical models it has to be considered as a rather unusual location for an airport.
+Therefore, in order to allow for a broad and more general evaluation and also to test
+for a high flexibility of the applied methods a second site with quite converse geographic 
+features is included in this study as well.
 Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
 a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
-area, is considered in this study.
-In that way completely different geographic features and their possible influences
-on wind direction are included which allows to test for a high flexibility of the applied methods.
+area, is considered.
 For the two locations the above introduced methodology of circular distributional trees and forests is
@@
 
-Lisa Schlosser recieved a PhD scholarship granted from the University of Innsbruck.
+Lisa Schlosser received a PhD scholarship granted from the University of Innsbruck.
 

### 2019-11-18 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Merge Lisa and Moritz
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2878 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 863de85c66db6ac7a6aae305563de0c493feaf43

@@
 \ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
-&= -\log(2 \pi I_0(\kappa)) +  \kappa \cos(y - \mu).
+&= -\log(2 \pi I_0(\kappa)) +  \kappa \cos(y - \mu).\label{equ:vM:loglik}
 \end{aligned}
@@
 
-\subsection{Data}
-As response variables, $10$\,min mean wind direction observations are employed at
-the airports Innsbruck and Vienna on an hourly temporal resolution. As
-predictor variables, we use observations of various meteorological quantities
-such as wind direction, temperature, air pressure and humidity directly at the
-respective airports, as well as within the vicinity of the airports. For
-Innsbruck, meteorological variables are observed at XXX different locations
-along the intersecting valleys; whereas for Vienna, located in the flat,
-meteorological variables are observed at XXX different locations within an area
-of about $10$\,km$^2$. In addition, we employ temporal differences, averages, minima, maxima over the last 3\,h,
+\subsection{Data}\label{sec:wind:data}
+For the case study, as circular response variables we employ $10$\,min mean
+wind direction measurements at the airports Innsbruck and Vienna on an hourly
+temporal resolution. As predictor variables, we use $1$-hourly resolved
+$10$\,min mean observations of various meteorological quantities such as wind
+direction, wind speed, temperature, air pressure and humidity, all
+lagged by one or three hours according to the respective forecasting step. The
+meteorological variables are measured either directly at the airports or within
+the vicinity of the respective airports. For Innsbruck, measurements of
+21~different locations at the airport and along the intersecting valleys are
+used, whereas, for Vienna, measurements of 22~locations at the airport and
+within an area of about $10$\,km$^2$ are used. A topographical overview of the
+airports and their surrounding areas with the station sites employed in this
+study is provided in Figure~XXX. In addition, we use derived quantities such
+as 3~hourly means, minima and maxima, as well as 1~and 3~hourly temporal
+changes and spatial differences towards the airport of the respective
+quantities. A summary of all predictor variables can be found in
+Table~\ref{tab:covariates}.
 
-of ias well as tempat each of the weather stations also differences in these 
-quantities between the considered stations are included. For example, differences in air 
-pressure between Innsbruck and either a western or eastern location are expected to have a 
-high influence on wind directions. Moreover, changes of these quantities over time are 
-included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
-covariates for Vienna. 
-A list of all covariates together with their abbreviations as used in this paper
-can be found in Table~\ref{tab:covariates}.
-The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
-After eliminating observations with at least one missing value a total number of 26861 time points
-for Innsbruck and 33853 for Vienna are available.
-As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
-expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
-especially of the employed reference models might suffer from the influence of these ``noise'' 
-observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
-considered.
-
-\fixme{information on NAs, update data information, include overview}
+The data used in this study consists of 5~total years from January 2015 to
+December 2018. After first eliminating predictor variables with more than 5\% missing
+values and then cases with missing observations, the
+data set consists of 41979 time points and 260 covariates for Innsbruck, and of
+38985 times points and 494 covariates for Vienna, respectively.
 
@@
 \subsection{Models and evaluation}\label{sec:wind:models}
-For a fair evaluation of circular regression trees and forest and to
+For a fair evaluation of circular regression trees and forests, and to
 investigate whether they can be applied as a reasonable alternative to already
@@
 study for probabilistic forecasting of wind directions. Two of them are based
-on existing approaches in the field of meteorological forecasting while the
-third is a state-of-the-art model to forecast circular response variables. The
-next paragraphs introduce the reference methods and provide the corresponding
-model specifications as employed in this study for all considered model
-including the circular tree and forest.
+on existing approaches in the meteorological field while the third is a
+state-of-the-art model to forecast circular response variables. The next
+paragraphs introduce the reference methods and provide the corresponding model
+specifications as employed in this study for all considered models including the
+circular tree and forest.
+
+\paragraph{Climatological model} 
+Accurate knowledge of weather quantities' climatologies can be important for a
+wide range of applications and are often also used as a baseline for the
+validation of newly developed forecasting systems
+\citep{simon.etal:2017,stauffer.etal:2017}. Usually, the performance of
+climatological models strongly depends on the temporal dependency of the
+quantity itself, either on a daily or yearly basis. Wind direction at a single
+location is often quite unstable and shows only a daily or annual
+seasonality in case of predominant local wind systems like the ones at
+Innsbruck introduced for the analysis of Figure~\ref{fig:tree_ibk}. Hence,
+the climatological model in this study provides a very simple implementation of
+a baseline model and might experience serious performance losses in case of
+random temporal variabilities or abrupt changes in the employed response
+variable wind direction.
+
+For simplicity, rather to fit one single climatological model, we perform the
+model fitting for each time point separately with varying training data. For
+each time point, the training data set consists of the same daytime as the time
+of interest within 31~days centered around the same calender day over all years
+except the year of interest. To gain full probabilistic climatological
+forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
+the distribution parameters of the von Mises distribution via maximum
+likelihood estimation. The construction of a climatological reference model
+was similarly pursued by \citet{vogel.etal:2018} and is discussed in a
+comprehensive summary on different time-adaptive training scheme in
+\citet{lang.etal:2019c}.
+
+\paragraph{Persistency model} 
+Persistence describes the previous value of a single weather quantity in a time
+series; hence, it is another objective measure in the verification of
+weather forecasts and also often applied as a reference model
+\citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
+atmospheric conditions, persistence can be a very good estimate, but as soon as
+any major change takes place in the response variable during the forecast
+period, forecasts based on persistence may also fail remarkably. Therefore, it
+heavily depends on the variability of the variable of interest and the length
+of the forecasting step whether persistence can provide overall reliable
+predictions.
+%If a certain weather quantity remains constant over a certain time period,
+%persistence can be a very good estimate.
+
+To gain a full probabilistic persistency model, we use as for the
+climatological model maximimum likelihood estimation and fit the distribution
+parameters of the von Mises distribution conditional on lagged response values
+according to the description in Section~\ref{sec:prob_circ}. We fit one model
+for every time point in the validation data set employing the previous six
+lagged response values as training data. In order to allow for a stronger
+influence of observations closer to the time of interest an exponential
+smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
+prediction the current observation and all previous five observations together
+have an equal influence rate of 50 percent. Observations with longer time lags
+have exponential weights below $0.01$ and are therefore omitted
+from the training data.
+
+\paragraph{Generalized linear model}
+Traditional approaches to forecast circular response variables are often based on
+circular GLMs \citep{Fisher:1993}. As discussed in
+Section~\ref{sec:introduction}, circular regression models often experience
+the problem that the likelihood function can be strongly irregular which makes
+optimization rather difficult. Hence, they often do not converge if no
+appropriate initial values are provided \citep{Pewsey+Neuhaeuser+Ruxton:2013,
+Gill+Hangartner:2010}. In this study, to be able to employ a GLM out of the box
+as a reference, we use the Bayesian implementation of
+\cite{Mulder+Klugkist:2017} which does not dependent so much on initial values
+due to an MCMC sampling algorithm using weakly- or non-informative priors.
+
+Following \cite{Mulder+Klugkist:2017} the employed model plugs in an
+appropriate link function~$g$ in order to guarantee response values on the
+interval $[-\pi,\pi]$. As the implementation cannot handle circular covariates,
+we use the components of the lagged 2-dimensional wind vector $vec$ and the lagged
+wind speed $spd$ as predictor variables. The model formula for the location parameter
+$mu$ of the von Mises distribution can be written as:
+%%
+\begin{align}
+\mu = \beta_0 + g(\beta_1 \cdot vec_{1} + \beta_2 \cdot vec_{2} + \beta_3 \cdot spd
+\end{align}
+%%
+with $\beta_0$ being a circular intercept, $\beta_\bullet$ the regression
+coefficients and $g$ the link function $g(x) = 2 \cdot \arctan(x)$. In addition, a
+constant concentration parameter $\kappa$ is fitted to the full learning sample.
 
 \paragraph{Circular regression tree}
-For the distributional tree as discussed in Section~\ref{sec:circtree} all covariates provided in the 
-learning data are considered. The tree is obtained by employing the CTree algorithm with a minimal 
-number of 2000 observations in each terminal node serving as a stopping criterion 
-(argument \code{minbucket} of our implementation).
+For the circular regression tree introduced in Section~\ref{sec:circtree}, all
+covariates provided in the learning data, as described in
+Section~\ref{sec:wind:data}, can be considered due to an intrinsic automatic
+variable selection performed in the tree estimation. The tree building is
+performed by our newly developed R-package \pkg{circtree} employing the CTree
+algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a minimal number of 2000
+observations in each terminal node serving as a stopping criterion (argument
+\code{minbucket}).
 
-\paragraph{Circular regression forest}
-Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
-As for the distributional tree all available covariates are used to learn a distributional 
-forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
-containing a fraction of 30 percent of the original learning data. As it can be expected that 
-the lagged response variable (as included as a covariate) has a strong influence on the 
-response variable it is desirable to employ it in each tree of the forest. Therefore, all 
-covariates are considered for each tree
-(by setting the argument \code{mtry} of our implementation to the total number of covariates)
-in order to guarantee for the lagged response to be available.
-Since a high number of possible split points leads to a high computational effort that is 
-particularly noteworthy for forest models the covariates are binned allowing for a maximum of
-$50$ classes. In that way the number of split points that have to be considered is reduced
-in a reasonably.
-% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
-As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
-is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
-is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
-maximum value of 1 in order to build trees as large as possible as commonly done for forest models.
-
-\paragraph{Persistency model}
-Assuming that weather conditions will remain the same after a certain period of time is the
-basic idea of persistence in meteorology. While this simple way of making predictions can
-be very effective particularly under stable atmospheric circumstances it can fail remarkably
-as soon as any changes take place, especially in case of rapid changes.
-Therefore, it heavily depends on the specific features and the variability of the variable
-of interest whether this approach can provide overall reliable predictions.
-
-\fixme{add references}
-
-To establish a distributional model of persistence in this study a set of six observations
-is considered to make a prediction for a selected day and time point. In particular,
-the six previous time points are considered. In order to allow for a stronger influence of 
-observations closer to the prediction time point exponential smoothing is employed, i.e.,
-exponentially decaying weights are included with a smoothing factor of $0.5$ such that
-in each prediction step the current prediction and the previous observation both have an 
-equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
-as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
-For this selected set of weighted observations the distribution parameters of the von Mises 
-distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
-parametric model is specified allowing for probabilistic predictions and an evaluation of the 
-circular CRPS.
-As this approach only considers data from the previous six days a cross validation is not 
-applicable, however, the same evaluation framework as for the other four models can be employed 
-by obtaining predictions only for the testing data of the cross validation without using 
-the corresponding learning data.
-\fixme{not done like this, always using the whole data set. wrong?}
-
-\paragraph{Climatological model}
-If there are no climatological changes within a certain number of years it can be expected that 
-for a selected time point and day of the year weather conditions will always be the same over all 
-years considered.
-Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
-meteorological variable of interest that can be very useful but its performance can also suffer 
-dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.
-
-\fixme{add references}
-
-Similar to the applied persistence model a set of observations is selected to fit a distributional
-climatology model. For a specified day and time point observations from the exact same time point of 
-the same day as well as of the 30 surrounding days from all other years contained in the learning data 
-are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.
-
-\paragraph{Generalized linear model}
-Another commonly applied approach is to linearly regress the circular variable of interest on
-its lagged values. For the here discussed application this corresponds to establishing a generalized
-linear model with wind direction as the outcome variable and wind direction of the previous time point 
-as covariate. 
-Following \cite{Mulder+Klugkist:2017} the employed model plugs in a link function in order
-to guarantee values on the interval $[-\pi,\pi]$ and is of the form
-\begin{align}
-dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
-\end{align}
-with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
-$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
-The intercept and the regression coefficient are estimated in a Bayesian framework where an 
-MCMC algorithm is used to obtain posterior samples. Moreover, a constant concentration
-parameter is fitted to the full learning sample.
-
-%\fixme{add references and further information on the estimation process within circGLM?}
-
-%\fixme{add information on applied implementation here (!) or in separate chapter?}
+\paragraph{Circular regression forest} Following the description in
+Section~\ref{sec:circforest}, a circular regression forest is constructed based
+on 100 individual trees employing our R-package \pkg{circtree}. Each of these
+trees is again built by the CTree algorithm on a subsample containing a
+fraction of 30 percent of the original learning data. As it can be expected
+that the lagged response variable is one of the most informative predictor
+variables for wind direction forecasting, all covariates are considered for
+building each tree. This `bagging' ensures that the lagged response
+variable is included in each tree; in our implementation this can be done by
+setting the argument \code{mtry} to the total number of covariates. Since a
+high number of possible split points leads to high computational costs, the
+covariates are automatically binned in a maximum of $50$ classes (argument
+\code{nmax = c("yx" = Inf, "z" = 50)}). Contrary to building a single tree,
+forests usually consists of large trees as they are not prone to overfitting
+the data after some kind of averaging over the individual trees. Therefore, we
+use the following control arguments to build rather long trees: The minimal number 
+of observations to perform a split is set to 20
+(argument \texttt{minsplit}), the minimal number of observations in each
+terminal node is set to 7 (argument \texttt{minbucket}), and the significance
+level for variable selection is kept at its maximum value of $1$
+(argument \texttt{alpha}).
 

### 2019-11-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Revision of methodology by Lisa/Moritz
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2879 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3cf061283c286759eccefb6dbfb3453a6683138e

@@
 
-In many approaches to modeling circular data it is assumed that the circular variable of interest 
+In many approaches to model circular data it is assumed that the circular variable of interest 
 follows a circular distribution, in particular the von Mises distribution which is also known 
@@
 One of the first regression models with a circular response variable and linear covariates
-was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
-the von Mises distribtuion, is predicted by a linear combination of covariates.
-\cite{Johnson+Wehrly:1978} took this idea one step further by employing a generalized linear 
-model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
+was presented by \cite{Gould:1969} where the circular mean, i.e., the location parameter of 
+the von Mises distribtuion is predicted by a linear combination of covariates.
+\cite{Johnson+Wehrly:1978} took this idea one step further by plugging in a link function such that the values of the 
 linear predictor are transformed to a restricted interval of length $2\pi$. 
-This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
+This generalized linear model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
 introducing another specific transformation function which maps the values of the real line to the 
 unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
-on covariates, not only in separate models but also in a mixed model. There, within an iterative 
-optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
-To obtain proper initial values it is recommended to apply two separate models where either one of 
-the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
-While being built on well elaborated theory this methodology and the contained 
-optimization problem have shown to be very challenging in their application, mainly due to the 
+on covariates, not only in separate models but also in a mixed model. 
+%There, within an iterative 
+%optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
+%To obtain proper initial values it is recommended to apply two separate models where either one of 
+%the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
+While being built on well elaborated theory this methodology 
+%and the contained optimization problem have 
+has shown to be very challenging in their application, mainly due to the 
 complexity encountered in optimizing the log-likelihood function which is not globally concave 
-\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, the selection of 
-starting values is crucial and bears a high risk of the model not to converge if not done thoughtfully. 
-In order to avoid this strong dependence on choosing well-fitting initial values \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM
+\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, 
+highly informative starting values are crucial for the model to converge at all.
+In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM 
 in a Bayesian framework. 
-
 %\fixme{revise literature survey with focus on transition to GLM}
-
-The named methods and also many of their further developments 
-\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
-While they generally provide a powerful toolbox for modeling particularly smooth effects, 
-it has to be considered that representing these smooth effects on the unit circle
-can be challenging. Moreover, the discussed additive models may not be able to capture all 
-types of effects adequately. For example, abrupt changes might not be detected properly owing 
-to the enforced additive structure and model assumptions such as the type of link 
-function and the form of linear combinations.
+The named methods and also many of their further developments \citep[see,
+e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
+While generally providing a powerful toolbox for modeling particularly smooth
+effects, they may not be able to capture all types of effects adequately.
+Especially abrupt changes might not be detected properly owing to the enforced
+additive structure and model assumptions such as a possibly highly non-linear link function and
+the form of linear combinations.
 
@@
 by tree-based methods. \cite{Lund:2002} introduced the concept of circular regression trees 
-which was further explored by \cite{Larraondo+Inza+Lozano:2018}. This methodology
-enables splits leading to subsets which contain the ``origin point'' $0$ equal to $2\pi$
-and in that way allows for a proper representation of periodicity.
-
-Following the idea of employing tree-based methods but including also the probabilistic concept
-of regressing distribution parameters on covariates we propose a very flexible approach 
+which was further extended by \cite{Larraondo+Inza+Lozano:2018}. 
+Following this idea of employing tree-based methods but also including the probabilistic concept
+of regressing distribution parameters on covariates, we propose a very flexible approach 
 by applying the von Mises distribution within the methodology of distributional trees and 
@@
 
-As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional tree applied to 
-predict wind directions at the airport of Innsbruck in the following hour. Wind direction is 
-a classical circular quantity and accurate forecasts are of great importance for decision-making 
-processes and risk management in air traffic management. Innsbruck lies at the bottom of a 
-deep valley in the Alps where topography channels wind along the west-east valley axis or along 
-a tributary valley intersecting from the south. Hence, pressure gradients to which valley wind 
-regimes react are considered as covariates along with other meteorological measurements at the 
-airport (lagged by one hour), such as wind direction and wind speed at the airport.
+As a motivating example, a circular distributional tree is employed to
+probabilistic wind direction forecasting. Wind direction is a classical
+circular quantity and accurate forecasts are of great importance for
+decision-making processes especially in air traffic management.
+Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at
+the airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
+Topography channels wind along the west-east valley axis or along a tributary
+valley intersecting from the south. Hence, pressure gradients to which valley
+wind regimes react are considered as covariates along with other meteorological
+measurements at the airport (lagged by one hour), such as wind direction and
+wind speed at the airport.
 
-Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the 
+Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the 
 empirical (gray) and fitted von Mises (red) distribution of wind direction in each terminal 
@@
 A more extensive case study of circular regression trees and forests applied to
-probabilistic forecasting of wind direction at the airports of Innsbruck and Vienna 
-is presented in Section~\ref{sec:wind} where their performances are benchmarked to 
-alternative commonly used approaches.  Beforehand, distributional models for circular 
-data are discussed in Section~\ref{sec:prob_circ} and the methodology of circular 
-regression trees and forest and its features are presented in Section~\ref{sec:tree_forest}.
-After the case study, a comprehensive summary and conclusion is given in
+probabilistic forecasting of wind direction at the airports of Innsbruck and
+Vienna will be presented in this paper. Moreover, their performances are
+benchmarked to alternative commonly used approaches.  The theory on
+probabilistic circular modeling introducing the von Mises distribution will be
+discussed in Section~\ref{sec:prob_circ}. The methodology of circular
+distributional trees and forests and their features are introcuded in
+Section~\ref{sec:tree_forest}.  After the case study presented in
+\ref{sec:wind}, a comprehensive summary and conclusion is given in
 Section~\ref{sec:summary}.
@@
 Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
-\fixme{check domain of $\mu$ in literature}
 the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
@@
 appropriate link function plugged in as in GLMs. This kind of approach is presented for example
-in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. While applying the mixed model 
-requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
-\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
-a circular response variable.
+in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
+%While applying the mixed model 
+%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
+%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
+%a circular response variable.
 Apart from the methodology developed specifically for circular data more general approaches to 
@@
 where each distribution parameter is estimated by its own generalized additive model (GAM). 
-However, an adaption of this methodology to circular data can be very complex due to a
-challenging representation of smooth effects on the unit circle.
+%However, an adaption of this methodology to circular data can be very complex due to a
+%challenging representation of smooth effects on the unit circle.
 %However, an adaption of this methodology to circular data can not be obtained by the straight 
@@
 %dependence/interaction differs strongly from the linear case and therefore requires specific handling.
-Moreover, difficulties might arise in case of non-additive changes and specifying a model 
-properly can be a very tough task, particularly for a high number of covariates and no information
+However, difficulties might arise in case of non-additive changes and specifying a model 
+properly can be a very challenging, particularly for a high number of covariates and no information
 on possible interactions.
@@
 \label{sec:wind}
-The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
-employed for probabilistic forecasting of wind directions at the airport of Innsbruck. As explained
-and also supported by the results illustrated in the terminal nodes of the tree in 
-Figure~\ref{fig:tree_ibk} the geographic location of Innsbruck and particularly the surrounding
-mountain chains are a strongly influential component of the behavior of wind directions.
-While this provides an interesting and challenging environment for testing different
-statistical models it has to be considered as a rather unusual location for an airport.
-Therefore, in order to allow for a broad and more general evaluation and also to test
-for a high flexibility of the applied methods a second site with quite converse geographic 
-features is included in this study as well.
-Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
-a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
-area, is considered.
-For the two locations the above introduced methodology of circular distributional trees and forests is
-employed to wind direction data and additional meteorological measurements in order to obtain hourly 
-wind direction forecasts, separately for Innsbruck and Vienna.
+Accurate forecasts of wind directions are of great importance for risk
+management in various fields such as agriculture, industry or energy
+production. Thus, for example in order to direct airplanes to a safe landing,
+precise knowledge of wind direction at the respective airport for the next hour(s) is highly
+desirable. This section provides a comprehensive case study on probabilistic
+wind direction forecasting at two airports in Austria with fundamentally
+different site characteristics: Innsbruck with a rather unusual location for an
+airport, located at the bottom of a deep alpine valley, and Vienna located
+within flat terrain in the East of Austria. In addition to the newly
+introduced circular distributional trees and forests, the case study employs various
+alternative approaches to forecast wind direction. The study is based on
+$1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
+the airports.
+
+%
+%highly fluctating, intersting variable for trees,,, requires high flexibility
+%many covariables, few variables have influence, hihgly non-linear non-known interactions
+%airports vienna and innsbruck
+%lead times 1 and 3 hours
+%
+% plus figures table and fixmes
+%The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
+%employed for probabilistic forecasting of wind directions at the airport of Innsbruck. As explained
+%and also supported by the results illustrated in the terminal nodes of the tree in 
+%Figure~\ref{fig:tree_ibk} the geographic location of Innsbruck and particularly the surrounding
+%mountain chains are a strongly influential component of the behavior of wind directions.
+%While this provides an interesting and challenging environment for testing different
+%statistical models it has to be considered as a rather unusual location for an airport.
+%Therefore, in order to allow for a broad and more general evaluation and also to test
+%for a high flexibility of the applied methods a second site with quite converse geographic 
+%features is included in this study as well.
+%Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
+%a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
+%area, is considered.
+%For the two locations the above introduced methodology of circular distributional trees and forests is
+%employed to wind direction data and additional meteorological measurements in order to obtain hourly 
+%wind direction forecasts, separately for Innsbruck and Vienna.
 
 \subsection{Data}\label{sec:wind:data}
-For the case study, as circular response variables we employ $10$\,min mean
-wind direction measurements at the airports Innsbruck and Vienna on an hourly
+The circular response variable considered in this case study is a $10$\,min mean
+of wind direction measurements at the airports Innsbruck and Vienna on an hourly
 temporal resolution. As predictor variables, we use $1$-hourly resolved
@@
 meteorological variables are measured either directly at the airports or within
-the vicinity of the respective airports. For Innsbruck, measurements of
+the their vicinity. For Innsbruck, measurements of
 21~different locations at the airport and along the intersecting valleys are
@@
 study is provided in Figure~XXX. In addition, we use derived quantities such
-as 3~hourly means, minima and maxima, as well as 1~and 3~hourly temporal
+as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
 changes and spatial differences towards the airport of the respective
@@
 December 2018. After first eliminating predictor variables with more than 5\% missing
-values and then cases with missing observations, the
+values and then time points with missing observations, the
 data set consists of 41979 time points and 260 covariates for Innsbruck, and of
@@
 
-\paragraph{Climatological model} 
+\begin{itemize}
+\item \emph{Climatological model:} 
 Accurate knowledge of weather quantities' climatologies can be important for a
@@
 variable wind direction.
-
+\newline
 For simplicity, rather to fit one single climatological model, we perform the
@@
 \citet{lang.etal:2019c}.
-
-\paragraph{Persistency model} 
+%%
+\item \emph{Persistency model:}
 Persistence describes the previous value of a single weather quantity in a time
-series; hence, it is another objective measure in the verification of
+series. Hence, it is another objective measure in the verification of
 weather forecasts and also often applied as a reference model
@@
 %persistence can be a very good estimate.
-
-To gain a full probabilistic persistency model, we use as for the
-climatological model maximimum likelihood estimation and fit the distribution
+\newline
+To gain a full probabilistic persistency model, we proceed similarly as for the
+climatological model by using maximimum likelihood estimation and fitting the distribution
 parameters of the von Mises distribution conditional on lagged response values
@@
 lagged response values as training data. In order to allow for a stronger
-influence of observations closer to the time of interest an exponential
+influence of observations closer to the time of interest exponential
 smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
@@
 from the training data.
-
-\paragraph{Generalized linear model}
+%%
+\item \emph{Generalized linear model:}
 Traditional approaches to forecast circular response variables are often based on
@@
 \cite{Mulder+Klugkist:2017} which does not dependent so much on initial values
-due to an MCMC sampling algorithm using weakly- or non-informative priors.
-
+due to an MCMC sampling algorithm using weakly informative priors.
+\newline
 Following \cite{Mulder+Klugkist:2017} the employed model plugs in an
@@
 wind speed $spd$ as predictor variables. The model formula for the location parameter
-$mu$ of the von Mises distribution can be written as:
+$\mu$ of the von Mises distribution can be written as:
 %%
 \begin{align}
-\mu = \beta_0 + g(\beta_1 \cdot vec_{1} + \beta_2 \cdot vec_{2} + \beta_3 \cdot spd
+\mu = \beta_0 + g(\beta_1 \cdot vec_{1} + \beta_2 \cdot vec_{2} + \beta_3 \cdot spd)
 \end{align}
@@
 constant concentration parameter $\kappa$ is fitted to the full learning sample.
-
-\paragraph{Circular regression tree}
+%%
+\item \emph{Circular regression tree:}
 For the circular regression tree introduced in Section~\ref{sec:circtree}, all
@@
 variable selection performed in the tree estimation. The tree building is
-performed by our newly developed R-package \pkg{circtree} employing the CTree
+performed by the newly developed R-package \pkg{circtree} employing the CTree
 algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a minimal number of 2000
@@
 \code{minbucket}).
-
-\paragraph{Circular regression forest} Following the description in
+%%
+\item \emph{Circular regression forest:} Following the description in
 Section~\ref{sec:circforest}, a circular regression forest is constructed based
-on 100 individual trees employing our R-package \pkg{circtree}. Each of these
+on 100 individual trees employing the R-package \pkg{circtree}. Each of these
 trees is again built by the CTree algorithm on a subsample containing a
@@
 that the lagged response variable is one of the most informative predictor
-variables for wind direction forecasting, all covariates are considered for
-building each tree. This `bagging' ensures that the lagged response
-variable is included in each tree; in our implementation this can be done by
+variables for wind direction forecasting, all covariates are included for
+building each tree in order to ensure that the lagged response
+variable is always considered. This bagging approach can be done in \pkg{circtree} implementation by
 setting the argument \code{mtry} to the total number of covariates. Since a
@@
 the data after some kind of averaging over the individual trees. Therefore, we
-use the following control arguments to build rather long trees: The minimal number 
+use the following control arguments to build rather large trees: The minimal number 
 of observations to perform a split is set to 20
@@
 (argument \texttt{alpha}).
+\end{itemize}
 
@@
 
-In addition to the raw CRPS, CRPS-based skill scores are computed to assess
+In addition to the raw CRPS, also CRPS-based skill scores are computed to assess
 differences in the improvement of the various statistical models over the
@@
 \begin{equation} 
-  \text{CRPSS}_{\text{method}} = 1 -
-\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
+  \text{CRPSS}_{\text{model}} = 1 -
+\frac{\text{CRPS}_{\text{model}}}{\text{CRPS}_{\text{climatology}}}.
 \end{equation}
 
-All scores presented in the next section are out-of-sample: For the persistency
-model only time points prior to the time of interest are used, for all other
-models a five-fold cross-validation is employed using up to four calendar years
-for model training and the remaining single calendar year for validation. Due
-to the large sample size of 24 hourly values per day over five years, some kind
-of temporal aggregation is needed to ensure a correct visual comparison of the
-individual methods. As observed for the employed models, the variability of the
-predictive performance over the five years is lower than over a single day or
-over a single year. Hence, CRPS and CRPS skill scores are aggregated over the
+All scores presented in the next section are out-of-sample based on five years
+of observations: For the persistency model only time points prior to the time
+of interest are used and the validation is performed rolling over all
+observations. For all other models a five-fold cross-validation is employed
+using up to four calendar years for model training and the remaining single
+calendar year for validation. Due to the large sample size of 24 hourly values
+per day over five years, some kind of temporal aggregation is needed to ensure
+a correct visual comparison of the individual methods.  Performed analyses have
+shown that the for the employed models the variability of the predictive
+performance over the five years is lower than over a single day or over a
+single year. Hence, CRPS and CRPS skill scores are aggregated over the
 respective five validation years which yields 24 hourly scores per month
@@
 Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
-at forecast steps 1\,h and 3\,h for the airports Innsbruck (panels a,\,c) and
-Vienna (panels b,\,d). The scores are aggregated over the $5$~validation years,
+at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
+Vienna (Panels~b,\,d). The scores are aggregated over the $5$~validation years,
 yielding yearly mean values for every hour per calendar month, with a lower
@@
 circular tree and forest, for both stations and forecast steps, the
-climatological model and the linear model show clearly higher CRPS scores and
+climatological model and the linear model show clearly higher CRPS values and
 hence a lower predictive performance.
 
-The different site characteristics between the airports Innsbruck
+The different site characteristics of the airports Innsbruck
 (Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
 (Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
-level of the model performances and on their respective performance
-variation. At Innsbruck, due to the surrounding mountains a limited number of
-possible wind directions exists, namely the three wind regimes discussed for
+level of the model performances and on their respective performance variation.
+At Innsbruck, due to the surrounding mountains a limited number of possible
+wind directions exists, namely the three wind regimes discussed for
 Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
@@
 (Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the
-potential high prediction errors in case of a paradigm shift seem to lead to
-a higher variation in the predictive performance for all models in comparison
-to Vienna; this variation is especially high for the persistency model due to
-its strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
+potential high prediction errors in case of a paradigm shift seem to lead to a
+higher variation in the predictive performance for all models in comparison to
+Vienna; this variation is especially high for the persistency model due to its
+strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
 geographic surroundings do not constrain any wind direction which leads to
@@
 However, for the persistency model, at both stations, higher scores for the
-$3$\,h forecast (panels c,\,d) reveal a lower performance for longer lead
+$3$\,h forecast (Panels~c,\,d) reveal a lower performance for longer lead
 times; this is probably due to minimum 3-hourly instead of 1-hourly lagged
@@
 Figure~\ref{fig:boxplot_crpsraw}, this improvement over the climatological
-model is lower for Innsbruck due to the few number of predominant wind regimes
+model is lower for Innsbruck due to the low number of predominant wind regimes
 and hence a relatively good performance of the climatological model.
 Additionally, Figure~\ref{fig:boxplot_crpsskill} shows that even if the
-persistency model's performance is lower than the reference (panel c) the
+persistency model's performance is lower than the reference (Panel~c) the
 tree-based models are significantly superior to the climatological model used as

### 2019-11-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of introduction to section 4; minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2880 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5d368970440aac070259cddf23b8bbf386990728

@@
 
-\title{Circular Regression Trees and Forests with an Application to 
+\title{Circular Trees and Forests with an Application to 
     Probabilistic Wind Direction Forecasting}
-\Shorttitle{Distributional Trees and Forests for Circular Data}
+\Shorttitle{Circular Trees and Forests}
 
@@
 For probabilistic modeling of circular data the von Mises distribution is widely used.
-To capture how its parameters change with covariates, a regression tree model
+To capture how its parameters change with covariates, a distributional regression tree model
 is proposed as an alternative to more commonly-used additive models. The resulting
@@
 
-A more extensive case study of circular regression trees and forests applied to
+A more extensive case study of circular distributional trees and forests applied to
 probabilistic forecasting of wind direction at the airports of Innsbruck and
@@
 discussed in Section~\ref{sec:prob_circ}. The methodology of circular
-distributional trees and forests and their features are introcuded in
+distributional trees and forests and their features are introduced in
 Section~\ref{sec:tree_forest}.  After the case study presented in
@@
 
-\section{Circular regression trees and forests}
+\section{Circular distributional trees and forests}
 \label{sec:tree_forest}
@@
 Accurate forecasts of wind directions are of great importance for risk
-management in various fields such as agriculture, industry or energy
-production. Thus, for example in order to direct airplanes to a safe landing,
-precise knowledge of wind direction at the respective airport for the next hour(s) is highly
-desirable. This section provides a comprehensive case study on probabilistic
-wind direction forecasting at two airports in Austria with fundamentally
+management in various fields such as agriculture, energy production or air safety control. 
+For example in order to direct airplanes to a safe landing, precise knowledge of wind 
+direction at the respective airport for the next hour(s) is highly desirable and adequate 
+prediction methods are required. In this section, a comprehensive case study on probabilistic
+wind direction forecasting is provided for two airports in Austria with fundamentally
 different site characteristics: Innsbruck with a rather unusual location for an
@@
 quantities. A summary of all predictor variables can be found in
-Table~\ref{tab:covariates}.
+Table~\ref{tab:data}.
 
@@
 \begin{minipage}{\textwidth}
-\caption[Table caption text]{Basic covariates together with the type of variations.}
-\label{tab:covariates}
+\caption[Table caption text]{Overview of the employed data set.}
+\label{tab:data}
 \begin{tabular}{l  l}
 \hline
-Basic covariates & Variations\\
+Data components              & Description \\
 \hline
-\emph{dd}: wind direction       & different locations, \\
-                                & changes over time(?)\\
-\emph{ff}: wind speed           & different locations, \\
-                                & changes over time(?), \\
-                                & (hourly) maxima\\
-\emph{p}: pressure              & different locations, \\
-                                & differences between locations, \\
-                                & changes over time(?)\\
-\emph{t}: temperature           & different locations, \\
-                                & changes over time(?)\\
-\emph{rf}: relative humidity    & different locations\\
+Meteorological variables     & Wind direction, wind speed, air pressure, \\
+                             & temperature, relative humidity \\
+Derived quantities           & 3-hourly means, minima, maxima; \\
+                             & 1-hourly and 3-hourly temporal changes; \\
+                             & spatial differences towards the airport\\
+Weather stations (Innsbruck) & 21 stations at the airport, \\
+                             & Landeck, Kufstein, Steinach, \dots \\
+Weather stations (Vienna)    & 22 stations at the airport, \\
+                             & \ldots \\
+%\emph{dd}: wind direction       & different locations, \\
+%                                & changes over time(?)\\
+%\emph{ff}: wind speed           & different locations, \\
+%                                & changes over time(?), \\
+%                                & (hourly) maxima\\
+%\emph{p}: pressure              & different locations, \\
+%                                & differences between locations, \\
+%                                & changes over time(?)\\
+%\emph{t}: temperature           & different locations, \\
+%                                & changes over time(?)\\
+%\emph{rf}: relative humidity    & different locations\\
 \hline
@@
 \subsection{Models and evaluation}\label{sec:wind:models}
-For a fair evaluation of circular regression trees and forests, and to
+For a fair evaluation of circular distributional trees and forests, and to
 investigate whether they can be applied as a reasonable alternative to already
@@
 To gain a full probabilistic persistency model, we proceed similarly as for the
-climatological model by using maximimum likelihood estimation and fitting the distribution
+climatological model by using maximum likelihood estimation and fitting the distribution
 parameters of the von Mises distribution conditional on lagged response values
@@
 %%
-\item \emph{Circular regression tree:}
-For the circular regression tree introduced in Section~\ref{sec:circtree}, all
+\item \emph{Circular distributional tree:}
+For the circular distributional tree introduced in Section~\ref{sec:circtree}, all
 covariates provided in the learning data, as described in
@@
 the circular regression forest overall provides the best predictive
-performance, followed by the circular regression tree and, except for the
+performance, followed by the circular distributional tree and, except for the
 $3$\,h forecast at Innsbruck, the persistency model. In comparison to the
@@
 response values employed as covariates in the persistency model. The circular
-regression tree and forest seem to partially compensate for the lower skill of
+distributional tree and forest seem to partially compensate for the lower skill of
 the lagged response values by other covariates, hence their predictive
@@
     $288$~yearly mean values. The scores are shown for the climatological, 
-    persistency and the linear model as well as for the circular regression tree and
+    persistency and the linear model as well as for the circular distributional tree and
     forest.}

### 2019-11-19 pkg/circtree/inst/wind_paper/R_scripts/circforest_plot_exampletree.R
Minor modifications in plot nad updated example tree
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2881 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: da29af6c8add406f14141edeb255443a54d4e246

@@ -5,7 +5,7 @@
 # -------------------------------------------------------------------
 # - PURPOSE:
 # -------------------------------------------------------------------
-# - L@ST MODIFIED: 2019-11-12 on thinkmoritz
+# - L@ST MODIFIED: 2019-11-19 on thinkmoritz
 # -------------------------------------------------------------------
 
 # -------------------------------------------------------------------
@@ -161,8 +161,25 @@ d$doy <- tmp_time$yday + 1
 
 rm(tmp_time); gc()
 
+d$steinach.pred.diff_resp <- round((d$innsbruck.pred - mean(d$innsbruck.pred)) - 
+  (d$steinach.pred - mean(d$steinach.pred)), 1)
 d_example <- subset(d, select = c(response.dd, innsbruck.dd, innsbruck.ff, landeck.pred.diff_resp,
     kufstein.pred.diff_resp, steinach.pred.diff_resp))
+
+#par(mfrow = c(2, 1))
+#plot(d$steinach.p[1:100], type = "l", col = 2, ylim = 
+#  range(c(d$steinach.p[1:100], d$steinach.pred[1:100])), main = "Steinach")
+#lines(d$steinach.pred[1:100], col = 3)
+#legend("topleft", lty = 1, col = c(2, 3), legend = c(paste0("mean(p) = ", 
+#  round(mean(d$steinach.p), 2)), 
+#  paste0("mean(pred) = ", round(mean(d$steinach.pred), 2))))
+#
+#plot(d$innsbruck.p[1:100], type = "l", col = 2, ylim = 
+#  range(c(d$innsbruck.p[1:100], d$innsbruck.pred[1:100])), main = "Innsbruck")
+#lines(d$innsbruck.pred[1:100], col = 3)
+#legend("topleft", lty = 1, col = c(2, 3), legend = c(paste0("mean(p) = ", 
+#  round(mean(d$innsbruck.p), 2)), 
+#  paste0("mean(pred) = ", round(mean(d$innsbruck.pred), 2))))
 
 names(d_example)[names(d_example) == "innsbruck.dd"] <- "direction"
 names(d_example)[names(d_example) == "innsbruck.ff"] <- "speed"

### 2019-11-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor modifications in plot nad updated example tree
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2881 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: da29af6c8add406f14141edeb255443a54d4e246
Update title
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2883 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 06c6edcaf85dc58a1875050563ba326df1176af9

@@
 
-\title{Circular Trees and Forests with an Application to 
-    Probabilistic Wind Direction Forecasting}
+\title{Circular Distributional Trees and Forests with an Application to 
+    Wind Direction Forecasting}
 \Shorttitle{Circular Trees and Forests}
@@
 $1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
-the airports.
+the airports as predictor variables.
 

### 2019-11-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added linear illustration of fitted von Mises distribution to the circular illustration
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2885 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1c302863dcc272c1b1ac574541263e64aa11f764

@@
 \begin{figure}[t]
-\centering
-\includegraphics[width=4cm]{density_vM.pdf}
-\caption{Illustration of a distributional model fitted to circular data by maximum likelihood-estimation
-employing the von Mises distribution. Empirical histogram (gray) and fitted density (red line) are depicted
-along with the estimated location parameter (red hand).
+%\centering
+\hspace{-1cm}
+\begin{minipage}{.5\textwidth}
+  %\centering
+  \includegraphics[width=1.2\linewidth]{density_linear.pdf}
+  \end{minipage}%
+\begin{minipage}{.5\textwidth}
+  %\centering
+  \includegraphics[width=1.2\linewidth]{density_circular.pdf}
+\end{minipage}
+\caption{Illustration of a distributional model fitted to circular data by 
+  maximum likelihood-estimation employing the von Mises distribution. 
+  Left panel: linear illustration. Right panel: circular illustration.
+  In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
+  with the estimated location parameter (red hand).
   }
-\label{fig:density_vM} 
+  \label{fig:densities} 
+  %\includegraphics[width=4cm]{density_vM.pdf}
 \end{figure}
@@
 
-Figure~\ref{fig:density_vM} illustrates a distributional model fitted to circular data
+Figure~\ref{fig:densities} illustrates a distributional model fitted to circular data
 employing the von Mises distribution.

### 2019-11-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include 'new draft' for abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2887 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b68ab7b74a4372cd65aa4a31e380ba115b497d33

@@
 \Abstract{
-As circular data can be found in many different subject areas accounting for periodicity has 
-become an important aspect in establishing appropriate statistical models.
-For probabilistic modeling of circular data the von Mises distribution is widely used.
-To capture how its parameters change with covariates, a distributional regression tree model
-is proposed as an alternative to more commonly-used additive models. The resulting
-distributional trees are easy to interpret, can detect non-additive effects, and select
-covariates and their interactions automatically. In order to stabilize and regularize
-the model an ensemble of distributional trees is combined yielding a distributional
-forest which allows for modeling smooth effects as well.
-For illustration, hourly wind direction forecasts are obtained at two airports in Austria
-(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
-airports and surrounding weather stations.
+As circular data can be found in many different areas, the development of
+appropriate statistical models accounting, i.a., for periodicity has become a
+challenging and important research field. For probabilistic forecasting
+typically the von Mises distribution is employed in an additive model
+framework. Combining their underlying distributional setup with tree-based
+estimation methods, circular distributional trees will be presented. The resulting
+distributional trees are easy to interpret, can detect non-additive effects,
+and select covariates and their interactions automatically. By combining
+ensembles of circular trees a stabilized circular forest can build, which
+automatically regularizes the estimation and, hence, allows for modeling smooth effects.
+
+To validate the proposed circular trees and forests, a case study on
+probabilistic forecasting of wind direction will be shown for two airports in
+Austria. In addition to the tree-based models, alternative commonly used
+approaches will be employed to compare their predictive performances to the
+proposed circular trees and forest.
+
+%As circular data can be found in many different subject areas accounting for periodicity has 
+%become an important aspect in establishing appropriate statistical models.
+%For probabilistic modeling of circular data the von Mises distribution is widely used.
+%To capture how its parameters change with covariates, a distributional regression tree model
+%is proposed as an alternative to more commonly-used additive models. The resulting
+%distributional trees are easy to interpret, can detect non-additive effects, and select
+%covariates and their interactions automatically. In order to stabilize and regularize
+%the model an ensemble of distributional trees is combined yielding a distributional
+%forest which allows for modeling smooth effects as well.
+%For illustration, hourly wind direction forecasts are obtained at two airports in Austria
+%(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
+%airports and surrounding weather stations.
 }
-\Keywords{Distributional Trees and Forests; Circular Response; Von Mises Distribution}
+\Keywords{Distributional Trees and Forests; Circular Response; von Mises Distribution}
 
@@
 %\centering
-\hspace{-1cm}
-\begin{minipage}{.5\textwidth}
+%\hspace{-1cm}
+\begin{minipage}{.05\textwidth}
+\hfill
+\end{minipage}%
+\begin{minipage}{.45\textwidth}
   %\centering
-  \includegraphics[width=1.2\linewidth]{density_linear.pdf}
+  \vspace{-1em}
+  \includegraphics[width=1.05\linewidth]{density_linear.pdf}
   \end{minipage}%
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.45\textwidth}
   %\centering
@@
 \end{minipage}
+\begin{minipage}{.05\textwidth}
+\hfill
+\end{minipage}%
 \caption{Illustration of a distributional model fitted to circular data by 

### 2019-11-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revised abstract, small corrections in description/caption of Figure 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2888 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 365d174d6d1ff9c8426d47689da099be6dc29348

@@
 \Abstract{
-As circular data can be found in many different areas, the development of
-appropriate statistical models accounting, i.a., for periodicity has become a
-challenging and important research field. For probabilistic forecasting
-typically the von Mises distribution is employed in an additive model
-framework. Combining their underlying distributional setup with tree-based
-estimation methods, circular distributional trees will be presented. The resulting
-distributional trees are easy to interpret, can detect non-additive effects,
-and select covariates and their interactions automatically. By combining
-ensembles of circular trees a stabilized circular forest can build, which
-automatically regularizes the estimation and, hence, allows for modeling smooth effects.
+As circular data appears in many different areas, the development of appropriate statistical 
+models which also account for periodicity has become an important but challenging research field. 
+For probabilistic modeling of circular data typically the von Mises distribution is employed.
+In order to regress its parameters on covariates, a circular distributional tree is proposed as 
+an alternative to more commonly-used additive models. The resulting circular trees are 
+easy to interpret, can detect non-additive effects, and select covariates and their 
+interactions automatically.
+By combining an ensemble of circular trees a more stable and regularized circular distributional forest
+can be built which allows for modeling smooth effects as well.
+%To stabilize and regularize the model an ensemble of distributional trees can be combined 
+%yielding a distributional forest which allows for modeling smooth effects as well.
 
-To validate the proposed circular trees and forests, a case study on
-probabilistic forecasting of wind direction will be shown for two airports in
-Austria. In addition to the tree-based models, alternative commonly used
-approaches will be employed to compare their predictive performances to the
-proposed circular trees and forest.
+To evaluate the proposed circular trees and forests, a case study on probabilistic 
+forecasting of wind direction for two airports in Austria is presented. 
+In addition to the tree-based models, alternative common approaches are 
+used as a benchmark in order to assess the predictive performances of circular trees 
+and forests.
+
+%In order to assess the predictive performance of the proposed circular trees and forests,
+%a case study on probabilistic forecasting of wind direction for two airports in Austria is presented
+%with alternative commonly used approaches being employed as benchmarks.
+
+%As circular data can be found in many different areas, the development of
+%appropriate statistical models accounting, i.a., for periodicity has become a
+%challenging and important research field. For probabilistic forecasting
+%typically the von Mises distribution is employed in an additive model
+%framework. Combining their underlying distributional setup with tree-based
+%estimation methods, circular distributional trees will be presented. The resulting
+%distributional trees are easy to interpret, can detect non-additive effects,
+%and select covariates and their interactions automatically. By combining
+%ensembles of circular trees a stabilized circular forest can build, which
+%automatically regularizes the estimation and, hence, allows for modeling smooth effects.
+
+%To validate the proposed circular trees and forests, a case study on
+%probabilistic forecasting of wind direction will be shown for two airports in
+%Austria. In addition to the tree-based models, alternative commonly used
+%approaches will be employed to compare their predictive performances to the
+%proposed circular trees and forest.
+
+%%%%%%
 
@@
   maximum likelihood-estimation employing the von Mises distribution. 
-  Left panel: linear illustration. Right panel: circular illustration.
+  Left panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. 
+  Right panel: circular illustration.
   In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
@@
 \end{figure}
-\fixme{update figure 2d to 3d, using data of example tree}
+%\fixme{update figure 2d to 3d, using data of example tree}
 
-Figure~\ref{fig:densities} illustrates a distributional model fitted to circular data
-employing the von Mises distribution.
+Figure~\ref{fig:densities} depicts a distributional model fitted to circular data
+employing the von Mises distribution. While the left panel only serves as an illustration 
+of the corresponding  ``linear'' description restricted to an interval of length $2\pi$ but 
+consequently not accounting for periodicity the right panel displays an appropriate 
+representation of a circular quantity on the unit circle. 
 % on observed wind direction data from the airport of Innsbruck.
-This example shows that reasonably well fitting probabilistic models can
-be established for circular data by specifying an appropriate distribution family 
-such as the von Mises distribution and estimating its parameters via maximum likelihood. 
+Looking at the fitted density (red line) and estimated location (red hand) this example 
+shows that reasonably well fitting probabilistic models can be established for circular 
+data by specifying an appropriate distribution family such as the von Mises distribution
+and estimating its parameters via maximum likelihood. 
 However, this procedure considers only the response variable $Y$. Including covariates 

### 2019-11-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of Table 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2889 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4cce6f566164abae43d4ae5bd11b83ab7d31a0f2

@@
 \begin{table}[t!]
-%\resizebox{1.4\textwidth}{!}{
-\begin{minipage}{\textwidth}
-\caption[Table caption text]{Overview of the employed data set.}
+\caption[Table caption text]{Overview of the employed data sets for the case study. For Innsbruck and Vienna,
+ various meteorological variables and derived quantities of these are considered at the respective stations, located directly  at the airports or in their vicinities.}
 \label{tab:data}
+\begin{center}
 \begin{tabular}{l  l}
+%\noalign{\vskip 1mm}                  
 \hline
@@
 \hline
-Meteorological variables     & Wind direction, wind speed, air pressure, \\
-                             & temperature, relative humidity \\
-Derived quantities           & 3-hourly means, minima, maxima; \\
-                             & 1-hourly and 3-hourly temporal changes; \\
+Meteorological variables:     & Wind direction, wind (gust) speed, \\
+                              & (reduced) air pressure, relative humidity, \\
+                              & temperature\\ 
+\noalign{\vskip 1mm}                           
+Derived quantities:           & 3-hourly means/minima/maxima, \\
+                              & 1-hourly and 3-hourly temporal changes, \\
                               & spatial differences towards the airport\\
-Weather stations (Innsbruck) & 21 stations at the airport, \\
-                             & Landeck, Kufstein, Steinach, \dots \\
-Weather stations (Vienna)    & 22 stations at the airport, \\
-                             & \ldots \\
-%\emph{dd}: wind direction       & different locations, \\
-%                                & changes over time(?)\\
-%\emph{ff}: wind speed           & different locations, \\
-%                                & changes over time(?), \\
-%                                & (hourly) maxima\\
-%\emph{p}: pressure              & different locations, \\
-%                                & differences between locations, \\
-%                                & changes over time(?)\\
-%\emph{t}: temperature           & different locations, \\
-%                                & changes over time(?)\\
-%\emph{rf}: relative humidity    & different locations\\
+\noalign{\vskip 1mm}                              
+Weather stations (Innsbruck): & 4 stations at the airport, \\
+                              & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
+                              & Steinach\\
+\noalign{\vskip 1mm}
+Weather stations (Vienna):    & 9 stations at the airport, \\
+                              & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
+                              & Gro{\ss}-Enzersdorf, Gumpoldskirchen, Hohe Warte,\\
+                              & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
+                              & Seibersdorf, Unterlaa, Wolkersdorf\\
 \hline
 \end{tabular}
-\end{minipage} 
-%}
+\end{center}
 \end{table}

### 2019-11-21 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of Table 1; resolve FIXMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2890 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1117b24c4e4d8609763cdeab7469e40c9ebf3b80
Update draft (Lisa and Moritz 'Proofread')
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2892 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8e73c665f95f6ef08f1cc7e70cc0d9cf65a86ea6

@@
 As circular data appears in many different areas, the development of appropriate statistical 
-models which also account for periodicity has become an important but challenging research field. 
+models which can account for periodicity has become an important but challenging research field. 
 For probabilistic modeling of circular data typically the von Mises distribution is employed.
@@
 can be built which allows for modeling smooth effects as well.
-%To stabilize and regularize the model an ensemble of distributional trees can be combined 
-%yielding a distributional forest which allows for modeling smooth effects as well.
 
@@
 and forests.
-
-%In order to assess the predictive performance of the proposed circular trees and forests,
-%a case study on probabilistic forecasting of wind direction for two airports in Austria is presented
-%with alternative commonly used approaches being employed as benchmarks.
 
@@
 has shown to be very challenging in their application, mainly due to the
-complexity encountered in optimizing the log-likelihood function which is not globally concave 
-\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, 
-highly informative starting values are crucial for the model to converge at all.
-In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM 
-in a Bayesian framework. 
-%\fixme{revise literature survey with focus on transition to GLM}
-The named methods and also many of their further developments \citep[see,
-e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
-While generally providing a powerful toolbox for modeling particularly smooth
-effects, they may not be able to capture all types of effects adequately.
-Especially abrupt changes might not be detected properly owing to the enforced
-additive structure and model assumptions such as a possibly highly non-linear link function and
-the form of linear combinations.
+complexity encountered in optimizing the log-likelihood function which is not
+globally concave.  Therefore, highly informative starting values are crucial
+for the model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
+Gill+Hangartner:2010}.  In order to avoid this strong dependence on appropriate
+initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
+to regression modeling of circular data by employing a GLM in a Bayesian
+framework.  The listed methods and most of their further developments
+\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
+regression models.  While these generally provide a powerful toolbox for modeling
+particularly smooth effects, they may not be able to capture all types of
+effects adequately.  Especially abrupt changes might not be detected properly
+owing to the enforced additive structure and model assumptions, such as a
+possibly highly non-linear link function and the form of linear combinations.
 
-An alternative modeling approach that can deal with non-additive effects is provided
-by tree-based methods. \cite{Lund:2002} introduced the concept of circular regression trees 
-which was further extended by \cite{Larraondo+Inza+Lozano:2018}. 
-Following this idea of employing tree-based methods but also including the probabilistic concept
-of regressing distribution parameters on covariates, we propose a very flexible approach 
-by applying the von Mises distribution within the methodology of distributional trees and 
-distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular data.
-In that way no model assumptions are required as covariates and their possible interactions are
-selected automatically. Non-additive effects are captured by the tree structure while combining an 
-ensemble of trees to a forest model allows for modeling smooth changes as well. 
-Thus, a high level of flexibility is provided while a full distributional model is specified 
-by distributional trees and forests such that a wide range of inference is made available.
+An alternative modeling approach that can deal with non-additive effects is
+provided by tree-based methods. \cite{Lund:2002} introduced the concept of
+circular regression trees which was further extended by
+\cite{Larraondo+Inza+Lozano:2018}.  Following this idea of employing tree-based
+methods but also including the probabilistic concept of regressing distribution
+parameters on covariates, we propose a very flexible approach by applying the
+von Mises distribution within the methodology of distributional trees and
+distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
+data.  The resulting circular distributional trees and forests do not require
+any model assumptions as covariates and their possible interactions are
+selected automatically. Non-additive effects are captured by the tree structure
+while circular forests allow for modeling smooth changes as well.  Thus, a high
+level of flexibility is provided while a full distributional model is specified offering
+a wide range of inference.
 
@@
 \includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
-\caption{Fitted tree based on the von Mises distribution for wind direction forecasting 
-at the airport of Innsbruck for the following hour.
+\caption{Fitted tree based on the von Mises distribution for 1-hourly wind direction forecasts 
+at the airport of Innsbruck.
 In each terminal node the empirical histogram (gray) and fitted density (red line)
@@
 
-A more extensive case study of circular distributional trees and forests applied to
-probabilistic forecasting of wind direction at the airports of Innsbruck and
-Vienna will be presented in this paper. Moreover, their performances are
-benchmarked to alternative commonly used approaches.  The theory on
-probabilistic circular modeling introducing the von Mises distribution will be
-discussed in Section~\ref{sec:prob_circ}. The methodology of circular
-distributional trees and forests and their features are introduced in
-Section~\ref{sec:tree_forest}.  After the case study presented in
-\ref{sec:wind}, a comprehensive summary and conclusion is given in
-Section~\ref{sec:summary}.
+A more extensive case study of circular distributional trees and forests
+applied to probabilistic forecasting of wind direction at the airports of
+Innsbruck and Vienna will be presented in this paper. Moreover, their
+performances are benchmarked to alternative commonly used approaches.  The
+remainder of the paper is structured as follows: The theory on probabilistic
+circular modeling introducing the von Mises distribution will be discussed in
+Section~\ref{sec:prob_circ}. The methodology of circular distributional trees
+and forests and their features are introduced in Section~\ref{sec:tree_forest}.
+After the case study presented in \ref{sec:wind}, a comprehensive summary and
+conclusion is given in Section~\ref{sec:summary}.
 
@@
 Generally, this feature can be obtained by ``wrapping'' the probability density function
-of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
-applying the modulo operation on the interval length $2\pi$.
+of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}.
 In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
@@
 estimating the distribution parameters. Thus, when applying the von Mises distribution to fit a 
-probabilistic model $\text{vM}(Y, \mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ 
+probabilistic model $\text{vM}(Y; \mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ 
 the distribution parameters $\mu$ and $\kappa$ are to be estimated.
@@
 \begin{equation}\label{eq:optim}
-(\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i),
+(\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i)
 \end{equation}
@@
 an indicator of deviation and in that way a measure of goodness of fit of the model can be obtained 
-for each observation $y_i$. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
+for each observation $y_i$ and fitted parameter. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
 be defined by 
@@
 \caption{Illustration of a distributional model fitted to circular data by 
-  maximum likelihood-estimation employing the von Mises distribution. 
+  maximum likelihood estimation employing the von Mises distribution. 
   Left panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. 
@@
 \end{figure}
-%\fixme{update figure 2d to 3d, using data of example tree}
 
@@
 of the corresponding  ``linear'' description restricted to an interval of length $2\pi$ but 
-consequently not accounting for periodicity the right panel displays an appropriate 
+consequently not accounting for periodicity, the right panel displays an appropriate 
 representation of a circular quantity on the unit circle. 
 % on observed wind direction data from the airport of Innsbruck.
-Looking at the fitted density (red line) and estimated location (red hand) this example 
+Looking at the fitted density (red line) and estimated location parameter (red hand) this example 
 shows that reasonably well fitting probabilistic models can be established for circular 
@@
 and estimating its parameters via maximum likelihood. 
-However, this procedure considers only the response variable $Y$. Including covariates 
+However, this procedure considers only the response variable. Including covariates 
 can improve any model remarkably as they might provide valuable additional information.
@@
 \label{sec:tree_forest}
-%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
 A common way of including covariates is to regress the model parameters on covariates.
@@
 %dependence/interaction differs strongly from the linear case and therefore requires specific handling.
-However, difficulties might arise in case of non-additive changes and specifying a model 
-properly can be a very challenging, particularly for a high number of covariates and no information
+However, difficulties might arise in case of non-additive changes and a proper model 
+specification can be very challenging, particularly for a high number of covariates and no information
 on possible interactions.
-An alternative regression approach that can deal with these problems is to apply a tree-structured 
-model.
+As an alternative regression approach which can deal with these problems, we introduce a circular probabilistic tree-structured model in this section.
 
@@
 subgroups based on covariates before fitting a local model in each of these subgroups allows 
-to capture (possibly) group specific effects more precisely and hence can result in a better 
-fitting overall model.
-This is the general idea of regression trees which are combined with distributional regression
+to capture (possibly) group specific effects more precisely and hence can result in a overall better 
+fitting model.
+This is the general idea of regression trees which are combined with distributional 
 modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
 algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
-By applying the presented distributional trees a full distributional model is specified in each 
-node of the tree. For this purpose a distribution family has to be chosen in advance. Hence, 
+By applying the presented distributional trees, a full distributional model is specified in each 
+node of the tree for a specific distribution family chosen in advance. Hence, 
 selecting the von Mises distribution allows for an application of distributional trees to circular
@@
 
-In particular, in each node the crucial decision on how and where to split the data is based on model
-scores which are obtained by evaluating the score function $s$ as defined in Equation~\ref{eq:scores} 
-at the individual observations and parameter estimates. For the von Mises distribution and a data set 
-of $n$ observations this yields an $n \times 2$ matrix that can be employed as a kind of residual,
+In the applied tree-building algorithm, in each node the crucial decision on how and where to split the data is based on model
+scores which are obtained by evaluating the score function $s$ 
+for the individual observations and parameter estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its two distribution parameters and a data set 
+of $n$ observations this yields an $n \times 2$ matrix that can be employed as a discrepancy measure,
 capturing how well each given observation conforms with the estimated location $\hat{\mu}$ and precision
@@
 covariate is assessed using either a parameter instability test~(MOB) or a permutation test~(CTree).
-In each partitioning step, the covariate with the highest significant association (i.e., lowest
+By doing so in each partitioning step, the covariate with the highest significant association (i.e., lowest
 significant $p$-value, if any) is selected for splitting the data. The corresponding split point
@@
 
-Once a distributional tree model is fit it can be employed to obtain probabilistic predictions
-for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
+Once a distributional tree model is fitted it can be employed to obtain probabilistic predictions
+for a possibly new set of observed covariates $\bm{z} = (z_1, \ldots, z_m)$.
 Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
@@
 \label{sec:circforest}
-A natural extension of distributional trees for circular data are ensembles or forests 
+A natural extension of circular distributional trees are ensembles or forests 
 that can improve forecasts by regularizing and stabilizing the model.
@@
 segment of the $t$-th tree.
-Hence, similar observations ending up more often in the same terminal node have higher weights
+Therefore, similar observations ending up more often in the same terminal node have higher weights
 and in that way a stronger influence in the estimation process.
@@
 full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$ corresponds to the abrupt
-splits from the tree.
-Thus distributional forests for circular data can capture both, smooth and abrupt changes, while
-covariates and possible interactions are selected automatically allowing for an easy to
-apply methodology to fully specify the von Mises distribution for circular data.
+splits from a tree.
+Thus, circular distributional forests can capture both smooth and abrupt changes, while
+covariates and possible interactions are selected automatically allowing for an easy-to-apply 
+methodology with a full specification of the von Mises distribution.
 
@@
 management in various fields such as agriculture, energy production or air safety control. 
-For example in order to direct airplanes to a safe landing, precise knowledge of wind 
-direction at the respective airport for the next hour(s) is highly desirable and adequate 
+For example, in order to direct airplanes to a safe landing, precise knowledge of wind 
+direction for the next hour(s) at the respective airport is highly desirable and adequate 
 prediction methods are required. In this section, a comprehensive case study on probabilistic
@@
 different site characteristics: Innsbruck with a rather unusual location for an
-airport, located at the bottom of a deep alpine valley, and Vienna located
+airport, located at the bottom of a deep Alpine valley, and Vienna located
 within flat terrain in the East of Austria. In addition to the newly
 introduced circular distributional trees and forests, the case study employs various
-alternative approaches to forecast wind direction. The study is based on
+alternative approaches to forecast wind direction as a benchmark. The study is based on
 $1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
@@
 
-%
-%highly fluctating, intersting variable for trees,,, requires high flexibility
-%many covariables, few variables have influence, hihgly non-linear non-known interactions
-%airports vienna and innsbruck
-%lead times 1 and 3 hours
-%
-% plus figures table and fixmes
 %The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
@@
 meteorological variables are measured either directly at the airports or within
-the their vicinity. For Innsbruck, measurements of
-21~different locations at the airport and along the intersecting valleys are
-used, whereas, for Vienna, measurements of 22~locations at the airport and
+the their vicinity. For Innsbruck, measurements at the airport and along the intersecting valleys are
+used, whereas, for Vienna, measurements at the airport and
 within an area of about $10$\,km$^2$ are used. A topographical overview of the
@@
 changes and spatial differences towards the airport of the respective
-quantities. A summary of all predictor variables can be found in
+quantities. An overview of the employed data sets can be found in
 Table~\ref{tab:data}.
@@
 \begin{table}[t!]
-\caption[Table caption text]{Overview of the employed data sets for the case study. For Innsbruck and Vienna,
- various meteorological variables and derived quantities of these are considered at the respective stations, located directly  at the airports or in their vicinities.}
+\caption[Table caption text]{Overview of the employed data sets for the case study: For Innsbruck and Vienna,
+ various meteorological variables and derived quantities of these are considered at the respective stations, located either directly at the airports or in their vicinities.}
 \label{tab:data}
@@
 \begin{tabular}{l  l}
-%\noalign{\vskip 1mm}                  
 \hline
@@
 study for probabilistic forecasting of wind directions. Two of them are based
-on existing approaches in the meteorological field while the third is a
+on existing approaches in the meteorological field, while the third is a
 state-of-the-art model to forecast circular response variables. The next
 paragraphs introduce the reference methods and provide the corresponding model
-specifications as employed in this study for all considered models including the
+specifications used in this study for all considered models including the
 circular tree and forest.
@@
 the climatological model in this study provides a very simple implementation of
-a baseline model and might experience serious performance losses in case of
+a baseline model, which might experience serious performance losses in case of
 random temporal variabilities or abrupt changes in the employed response
@@
 For simplicity, rather to fit one single climatological model, we perform the
-model fitting for each time point separately with varying training data. For
+model fitting separately rolling over all time points. For
 each time point, the training data set consists of the same daytime as the time
@@
 was similarly pursued by \citet{vogel.etal:2018} and is discussed in a
-comprehensive summary on different time-adaptive training scheme in
+comprehensive summary on different time-adaptive training schemes in
 \citet{lang.etal:2019c}.
@@
 Traditional approaches to forecast circular response variables are often based on
-circular GLMs \citep{Fisher:1993}. As discussed in
-Section~\ref{sec:introduction}, circular regression models often experience
+circular GLM-typed models \citep{Fisher:1993}. As discussed in
+Section~\ref{sec:introduction}, circular regression models can experience
 the problem that the likelihood function can be strongly irregular which makes
@@
 as a reference, we use the Bayesian implementation of
-\cite{Mulder+Klugkist:2017} which does not dependent so much on initial values
+\cite{Mulder+Klugkist:2017} which less depends on initial values
 due to an MCMC sampling algorithm using weakly informative priors.
@@
 appropriate link function~$g$ in order to guarantee response values on the
-interval $[-\pi,\pi]$. As the implementation cannot handle circular covariates,
+interval $(-\pi,\pi]$. As the implementation cannot handle circular covariates,
 we use the components of the lagged 2-dimensional wind vector $vec$ and the lagged
@@
 For the circular distributional tree introduced in Section~\ref{sec:circtree}, all
-covariates provided in the learning data, as described in
-Section~\ref{sec:wind:data}, can be considered due to an intrinsic automatic
+covariates provided in the learning data can be considered due to an intrinsic automatic
 variable selection performed in the tree estimation. The tree building is
@@
 %%
-\item \emph{Circular regression forest:} Following the description in
-Section~\ref{sec:circforest}, a circular regression forest is constructed based
+\item \emph{Circular distribution forest:} Following the description in
+Section~\ref{sec:circforest}, a circular forest is constructed based
 on 100 individual trees employing the R-package \pkg{circtree}. Each of these
@@
 variables for wind direction forecasting, all covariates are included for
-building each tree in order to ensure that the lagged response
-variable is always considered. This bagging approach can be done in \pkg{circtree} implementation by
+building each tree which ensures that the lagged response
+variable is always considered for splitting. This bagging approach can be applied in \pkg{circtree} by
 setting the argument \code{mtry} to the total number of covariates. Since a
 high number of possible split points leads to high computational costs, the
-covariates are automatically binned in a maximum of $50$ classes (argument
-\code{nmax = c("yx" = Inf, "z" = 50)}). Contrary to building a single tree,
-forests usually consists of large trees as they are not prone to overfitting
-the data after some kind of averaging over the individual trees. Therefore, we
+covariates are binned in a maximum of $50$ classes (argument
+\code{nmax = c("yx" = Inf, "z" = 50)}). Contrary to a single-tree model,
+forests usually consist of very large trees as they are not prone to overfitting
+the data due to the stabilization obtained by combining the individual trees. Therefore, we
 use the following control arguments to build rather large trees: The minimal number 
@@
 
-All scores presented in the next section are out-of-sample based on five years
-of observations: For the persistency model only time points prior to the time
+All scores presented in the next section are computed out-of-sample based on five years
+of observations. For the persistency model only time points prior to the time
 of interest are used and the validation is performed rolling over all
@@
 a correct visual comparison of the individual methods.  Performed analyses have
-shown that the for the employed models the variability of the predictive
+shown that for the employed models the variability of the predictive
 performance over the five years is lower than over a single day or over a
@@
 score indicating better performance. For both stations at both forecast steps,
-the circular regression forest overall provides the best predictive
+the circular forest overall provides the best predictive
 performance, followed by the circular distributional tree and, except for the
@@
 The different site characteristics of the airports Innsbruck
-(Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
-(Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
-level of the model performances and on their respective performance variation.
+(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
+(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the absolute
+level of the model performances and on their respective predictive performance variances.
 At Innsbruck, due to the surrounding mountains a limited number of possible
@@
 states, but once a change takes place it is mostly a major wind direction
-shift. Hence, due to the few wind regimes the rather inflexible climatological
-and linear models score relatively well with similar values as the other models
-(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the
+shift. Due to the few wind regimes the rather inflexible climatological
+and linear models score relatively well with similar CRPS values as the other models
+(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
 potential high prediction errors in case of a paradigm shift seem to lead to a
@@
 climatological and linear models, and to reduce the performance variability for
-all models (Figure~\ref{fig:boxplot_crpsraw}b,\,d).
+all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).
 
-The different forecast steps seem to have only a minor effect on the predictive
+The different forecast steps have apparently only a minor effect on the predictive
 performance of the climatological model and the linear model at both stations.
 However, for the persistency model, at both stations, higher scores for the
-$3$\,h forecast (Panels~c,\,d) reveal a lower performance for longer lead
+$3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a lower performance for longer lead
 times; this is probably due to minimum 3-hourly instead of 1-hourly lagged
@@
 
-In addition to the raw CRPS, Figure~\ref{fig:boxplot_crpsskill} provides CRPS
-skill scores with the climatological model as a reference. Skill scores are in
+In addition to the raw CRPS (Figure~\ref{fig:boxplot_crpsraw}), CRPS
+skill scores with the climatological model as a reference are provided in
+Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in
 percent, and positive values indicate an improvement in the predictive
-performance over the reference. For all setups, the circular regression forest
+performance over the reference. For all setups, the circular forest
 has the highest skill scores with a mean performance gain of 13--25\% and 58--71\% for
@@
 \section{Summary and conclusion}\label{sec:summary}
-Distributional trees for circular responses are established by coupling
+Circular distributional trees for circular responses are established by coupling
 model-based recursive partitioning with the von Mises distribution.
-The resulting trees can capture nonlinear changes, shifts, and potential interactions
+The resulting trees can capture non-linear changes, shifts, and potential interactions
 in covariates without prespecification of such effects. This is particularly
-useful for modeling wind direction in mountainous terrain where wind shifts
-can occur due to turns of the pressure gradients along a valley.
+useful for modeling a highly fluctating response such as wind direction. \dots
 
 \fixme{add summary of results}
-
 \fixme{outlook on possible extensions of models, use of NWP data}
@@
 This project was partially funded by the Austrian Research Promotion
-Agency~(FFG) grant number~$858537$.
-
-Torsten Hothorn received funding from the Swiss National Science
-Foundation, grant number~$200021\_184603$.
-
-Lisa Schlosser received a PhD scholarship granted from the University of Innsbruck.
+Agency~(FFG) grant number~$858537$. Torsten Hothorn received funding from the
+Swiss National Science Foundation, grant number~$200021\_184603$. Lisa
+Schlosser received a PhD scholarship granted from the University of Innsbruck.
 
@@
 
-The proposed methods are in the \textsf{R} package \textbf{circtree}
-(version~\Sexpr{packageVersion("circtree")}) based on the \textbf{disttree}
+The proposed circular tree and forest models are implemented in the \textsf{R} package \textbf{circtree}
+(version~\Sexpr{packageVersion("circtree")}). The package is based on the \textbf{disttree}
 package (version~\Sexpr{packageVersion("disttree")}) which applies the main
 tree building function from the  \textbf{partykit} package 
-(version~\Sexpr{packageVersion("partykit")}), all three available 
-on \textsf{R}-Forge at
-(\url{https://R-Forge.R-project.org/projects/partykit/}).
+(version~\Sexpr{packageVersion("partykit")}). All three packages are available 
+on \textsf{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.
 

### 2019-11-21 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Remove all comments and clean up Rnw file
- additionally spell check
- Add draft version 2019-11-20
- Update makefile (for diff)


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2894 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f82750fb5fb8c4263ddf38abe3b4a195471ef852

@@
 \Abstract{
-As circular data appears in many different areas, the development of appropriate statistical 
-models which can account for periodicity has become an important but challenging research field. 
-For probabilistic modeling of circular data typically the von Mises distribution is employed.
-In order to regress its parameters on covariates, a circular distributional tree is proposed as 
-an alternative to more commonly-used additive models. The resulting circular trees are 
-easy to interpret, can detect non-additive effects, and select covariates and their 
-interactions automatically.
-By combining an ensemble of circular trees a more stable and regularized circular distributional forest
-can be built which allows for modeling smooth effects as well.
+As circular data appears in many different areas, the development of
+appropriate statistical models which can account for periodicity has become an
+important but challenging research field.  For probabilistic modeling of
+circular data typically the von Mises distribution is employed.  In order to
+regress its parameters on covariates, a circular distributional tree is
+proposed as an alternative to more commonly-used additive models. The resulting
+circular trees are easy to interpret, can detect non-additive effects, and
+select covariates and their interactions automatically.  By combining an
+ensemble of circular trees a more stable and regularized circular
+distributional forest can be built which allows for modeling smooth effects as
+well.
 
-To evaluate the proposed circular trees and forests, a case study on probabilistic 
-forecasting of wind direction for two airports in Austria is presented. 
-In addition to the tree-based models, alternative common approaches are 
-used as a benchmark in order to assess the predictive performances of circular trees 
-and forests.
-
-%As circular data can be found in many different areas, the development of
-%appropriate statistical models accounting, i.a., for periodicity has become a
-%challenging and important research field. For probabilistic forecasting
-%typically the von Mises distribution is employed in an additive model
-%framework. Combining their underlying distributional setup with tree-based
-%estimation methods, circular distributional trees will be presented. The resulting
-%distributional trees are easy to interpret, can detect non-additive effects,
-%and select covariates and their interactions automatically. By combining
-%ensembles of circular trees a stabilized circular forest can build, which
-%automatically regularizes the estimation and, hence, allows for modeling smooth effects.
-
-%To validate the proposed circular trees and forests, a case study on
-%probabilistic forecasting of wind direction will be shown for two airports in
-%Austria. In addition to the tree-based models, alternative commonly used
-%approaches will be employed to compare their predictive performances to the
-%proposed circular trees and forest.
-
-%%%%%%
-
-%As circular data can be found in many different subject areas accounting for periodicity has 
-%become an important aspect in establishing appropriate statistical models.
-%For probabilistic modeling of circular data the von Mises distribution is widely used.
-%To capture how its parameters change with covariates, a distributional regression tree model
-%is proposed as an alternative to more commonly-used additive models. The resulting
-%distributional trees are easy to interpret, can detect non-additive effects, and select
-%covariates and their interactions automatically. In order to stabilize and regularize
-%the model an ensemble of distributional trees is combined yielding a distributional
-%forest which allows for modeling smooth effects as well.
-%For illustration, hourly wind direction forecasts are obtained at two airports in Austria
-%(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
-%airports and surrounding weather stations.
+To evaluate the proposed circular trees and forests, a case study on
+probabilistic forecasting of wind direction for two airports in Austria is
+presented.  In addition to the tree-based models, alternative common approaches
+are used as a benchmark in order to assess the predictive performances of
+circular trees and forests. 
 }
@@
 
+% -------------------------------------------------------------------
+% Introduction
+% -------------------------------------------------------------------
 \begin{document}
-
 \section{Introduction}
 \label{sec:introduction}
-
 Circular data can be found in a variety of applications and subject areas,
 e.g., hourly crime rate in the social-economics, animal movement direction or
-gene-structure in biology, and wind direction as one of the most important weather 
-variables in meteorology. Fitting a statistical model to this type of data requires
-the incorporation of its specific feature of periodicity, i.e., in case of angular
-data the fact that values are restricted to an interval such as $[0,2\pi)$ with
-$0$ being equivalent to $2\pi$.
+gene-structure in biology, and wind direction as one of the most important
+weather variables in meteorology. Fitting a statistical model to this type of
+data requires the incorporation of its specific feature of periodicity, i.e.,
+in case of angular data the fact that values are restricted to an interval such
+as $[0,2\pi)$ with $0$ being equivalent to $2\pi$.
 
-In many approaches to model circular data it is assumed that the circular variable of interest 
-follows a circular distribution, in particular the von Mises distribution which is also known 
-as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
-This is also the basis which early regression models for circular data are built on as they 
-regress the parameter(s) of the von Mises distribution on covariates.
-One of the first regression models with a circular response variable and linear covariates
-was presented by \cite{Gould:1969} where the circular mean, i.e., the location parameter of 
-the von Mises distribtuion is predicted by a linear combination of covariates.
-\cite{Johnson+Wehrly:1978} took this idea one step further by plugging in a link function such that the values of the 
-linear predictor are transformed to a restricted interval of length $2\pi$. 
-This generalized linear model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
-introducing another specific transformation function which maps the values of the real line to the 
-unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
-on covariates, not only in separate models but also in a mixed model. 
-%There, within an iterative 
-%optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
-%To obtain proper initial values it is recommended to apply two separate models where either one of 
-%the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
-While being built on well elaborated theory this methodology 
-%and the contained optimization problem have 
-has shown to be very challenging in their application, mainly due to the
-complexity encountered in optimizing the log-likelihood function which is not
-globally concave.  Therefore, highly informative starting values are crucial
-for the model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
+In many approaches to model circular data it is assumed that the circular
+variable of interest follows a circular distribution, in particular the von
+Mises distribution which is also known as the ``circular normal distribution''
+(see Section~\ref{sec:prob_circ} for more details). This is also the basis
+which early regression models for circular data are built on as they regress
+the parameter(s) of the von Mises distribution on covariates. One of the first
+regression models with a circular response variable and linear covariates was
+presented by \cite{Gould:1969} where the circular mean, i.e., the location
+parameter of the von Mises distribution is predicted by a linear combination of
+covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
+plugging in a link function such that the values of the linear predictor are
+transformed to a restricted interval of length $2\pi$. This generalized linear
+model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
+subsequently by \cite{Fisher:1993} introducing another specific transformation
+function which maps the values of the real line to the unit circle. Moreover,
+the presented methodology allows for both distribution parameters, the mean
+direction and the concentration parameter of the von Mises distribution, to be
+modeled depending on covariates, not only in separate models but also in a
+mixed model. While being built on well elaborated theory this methodology has
+shown to be very challenging in their application, mainly due to the complexity
+encountered in optimizing the log-likelihood function which is not globally
+concave. Therefore, highly informative starting values are crucial for the
+model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
 Gill+Hangartner:2010}. In order to avoid this strong dependence on appropriate
@@
 \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
-regression models.  While these generally provide a powerful toolbox for modeling
-particularly smooth effects, they may not be able to capture all types of
-effects adequately.  Especially abrupt changes might not be detected properly
-owing to the enforced additive structure and model assumptions, such as a
-possibly highly non-linear link function and the form of linear combinations.
+regression models. While these generally provide a powerful toolbox for
+modeling particularly smooth effects, they may not be able to capture all types
+of effects adequately. Especially abrupt changes might not be detected
+properly owing to the enforced additive structure and model assumptions, such
+as a possibly highly non-linear link function and the form of linear
+combinations.
 
@@
 while circular forests allow for modeling smooth changes as well. Thus, a high
-level of flexibility is provided while a full distributional model is specified offering
-a wide range of inference.
+level of flexibility is provided while a full distributional model is specified
+offering a wide range of inference.
 
@@
 decision-making processes especially in air traffic management.
-Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at
-the airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
+Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
+airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
 Topography channels wind along the west-east valley axis or along a tributary
@@
 Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
-empirical (gray) and fitted von Mises (red) distribution of wind direction in each terminal 
-node. Based on the fitted location parameters $\hat \mu$, the subgroups can be distinguished 
-into the following wind regimes:
-(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
-east to west, nodes 4 and 5). 
-(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
-Innsbruck (from south-east to north-west, node 8). 
-(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
-nodes 10, 12 and 13). 
-Node~7 captures observations with rather low wind speeds that cannot be distinguished
-clearly into wind regimes and consequently are associated with a very low estimated 
-concentration $\hat \kappa$. In terms of covariates, the lagged wind direction 
-(``persistence'') is mostly responsible for distinguishing the broad wind regimes 
-listed above while the pressure gradients and wind speed separate between subgroups 
-with high vs.\ low precision. Note that in the meteorological context wind direction 
+empirical (gray) and fitted von Mises (red) distribution of wind direction in
+each terminal node. Based on the fitted location parameters $\hat \mu$, the
+subgroups can be distinguished into the following wind regimes: (1)~Up-valley
+winds blowing from the valley mouth towards the upper valley (from east to
+west, nodes 4 and 5). (2) Downslope winds blowing across the Alpine crest
+along the intersecting valley towards Innsbruck (from south-east to north-west,
+node 8). (3) Down-valley winds blowing in the direction of the valley mouth
+(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
+rather low wind speeds that cannot be distinguished clearly into wind regimes
+and consequently are associated with a very low estimated concentration $\hat
+\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
+mostly responsible for distinguishing the broad wind regimes listed above while
+the pressure gradients and wind speed separate between subgroups with high vs.\
+low precision. Note that in the meteorological context wind direction 
 is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
-
+%%
 \begin{figure}[t]
@@
   \includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
-\caption{Fitted tree based on the von Mises distribution for 1-hourly wind direction forecasts 
-at the airport of Innsbruck.
-In each terminal node the empirical histogram (gray) and fitted density (red line)
-are depicted along with the estimated location parameter (red hand). The covariates
-employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
-and pressure gradients ($\text{dpressure; hPa}$) west, east and south of the airport,
-all lagged by one hour.}
+  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
+  direction forecasts at the airport of Innsbruck. In each terminal node the
+  empirical histogram (gray) and fitted density (red line) are depicted along
+  with the estimated location parameter (red hand). The covariates employed are
+  wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
+  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
+  one hour.}
   \label{fig:tree_ibk} 
 \end{figure}
+%%
 
@@
 
+% -------------------------------------------------------------------
+% Probabilistic circular modeling
+% -------------------------------------------------------------------
 \section{Probabilistic circular modeling}
 \label{sec:prob_circ}
-Probabilistic modeling of circular data requires the selection of a proper circular
-probability distribution, i.e., a probability distribution which accounts for the periodicity
-of circular data.
-Generally, this feature can be obtained by ``wrapping'' the probability density function
-of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}.
-In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
-A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
-to use \citep{Fisher:1993} is provided by the von Mises distribution,
-a purely circular distribution which is also known as ``the circular normal distribution''
-and commonly applied for probabilistic modeling of circular data.
-Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
-the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
+Probabilistic modeling of circular data requires the selection of a proper
+circular probability distribution, i.e., a probability distribution which
+accounts for the periodicity of circular data. Generally, this feature can be
+obtained by ``wrapping'' the probability density function of any continuous
+distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
+wrapped Cauchy distribution or the wrapped normal distribution can be employed
+to model symmetric unimodal circular data. A close approximation to the wrapped
+normal distribution that is mathematically simpler and hence easier to use
+\citep{Fisher:1993} is provided by the von Mises distribution, a purely
+circular distribution which is also known as ``the circular normal
+distribution'' and commonly applied for probabilistic modeling of circular
+data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
+parameter $\kappa > 0$ the density of the von Mises distribution for an
+observation $y \in (-\pi, \pi]$ is given by:
+%%
 \begin{equation}
@@
 \end{equation}
-where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
-\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.
+%%
+where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
+$0$ \citep[see, e.g.,][for a more detailed
+overview]{Jammalamadaka+Sengupta:2001}.
 
 The corresponding log-likelihood function is defined by
+%%
 \begin{equation}
@@
 \end{equation}
+%%
 
-Once an appropriate distribution family is selected fitting a probabilistic model corresponds to
-estimating the distribution parameters. Thus, when applying the von Mises distribution to fit a 
-probabilistic model $\text{vM}(Y; \mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ 
-the distribution parameters $\mu$ and $\kappa$ are to be estimated.
-For a learning sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by 
+Once an appropriate distribution family is selected fitting a probabilistic
+model corresponds to estimating the distribution parameters. Thus, when
+applying the von Mises distribution to fit a probabilistic model $\text{vM}(Y;
+\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ the
+distribution parameters $\mu$ and $\kappa$ are to be estimated. For a learning
+sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by
 maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
-\begin{equation}\label{eq:optim}
-(\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i)
-\end{equation}
-yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that a fully specified 
-distributional model is fit to the learning data.
+\begin{equation}\label{eq:optim} (\hat{\mu},\hat{\kappa}) =
+\argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) \end{equation}
+yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
+a fully specified distributional model is fit to the learning data.
 
 By means of the score function
+%%
 \begin{equation}
@@
 \end{equation}
-an indicator of deviation and in that way a measure of goodness of fit of the model can be obtained 
-for each observation $y_i$ and fitted parameter. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
-be defined by 
-%setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
+%%
+an indicator of deviation and in that way a measure of goodness of fit of the
+model can be obtained for each observation $y_i$ and fitted parameter.
+Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
+defined by 
+%%
 \begin{equation}
@@
 \end{equation}
-
+%%
+%%
 \begin{figure}[t]
-%\centering
-%\hspace{-1cm}
   \begin{minipage}{.05\textwidth}
@@
   \begin{minipage}{.45\textwidth}
-  %\centering
     \vspace{-1em}
@@
   \begin{minipage}{.45\textwidth}
-  %\centering
     \includegraphics[width=1.2\linewidth]{density_circular.pdf}
@@
   \caption{Illustration of a distributional model fitted to circular data by
-  maximum likelihood estimation employing the von Mises distribution. 
-  Left panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. 
-  Right panel: circular illustration.
-  In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
-  with the estimated location parameter (red hand).
-  }
+    maximum likelihood estimation employing the von Mises distribution. Left
+    panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. Right
+    panel: circular illustration. In both panels the empirical histogram (gray)
+    and fitted density (red line) are depicted along with the estimated location
+    parameter (red hand).}
   \label{fig:densities} 
-  %\includegraphics[width=4cm]{density_vM.pdf}
 \end{figure}
+%%
 
-Figure~\ref{fig:densities} depicts a distributional model fitted to circular data
-employing the von Mises distribution. While the left panel only serves as an illustration 
-of the corresponding  ``linear'' description restricted to an interval of length $2\pi$ but 
-consequently not accounting for periodicity, the right panel displays an appropriate 
-representation of a circular quantity on the unit circle. 
-% on observed wind direction data from the airport of Innsbruck.
-Looking at the fitted density (red line) and estimated location parameter (red hand) this example 
-shows that reasonably well fitting probabilistic models can be established for circular 
-data by specifying an appropriate distribution family such as the von Mises distribution
-and estimating its parameters via maximum likelihood. 
-However, this procedure considers only the response variable. Including covariates 
-can improve any model remarkably as they might provide valuable additional information.
-Therefore, distributional regression models are of high interest for all types of data,
-including circular data.
+Figure~\ref{fig:densities} depicts a distributional model fitted to circular
+data employing the von Mises distribution. While the left panel only serves as
+an illustration of the corresponding ``linear'' description restricted to an
+interval of length $2\pi$ but consequently not accounting for periodicity, the
+right panel displays an appropriate representation of a circular quantity on
+the unit circle. Looking at the fitted density (red line) and estimated
+location parameter (red hand) this example shows that reasonably well fitting
+probabilistic models can be established for circular data by specifying an
+appropriate distribution family such as the von Mises distribution and
+estimating its parameters via maximum likelihood. However, this procedure
+considers only the response variable. Including covariates can improve any
+model remarkably as they might provide valuable additional information.
+Therefore, distributional regression models are of high interest for all types
+of data, including circular data.
 
+% -------------------------------------------------------------------
+% Circular distributional trees and forests
+% -------------------------------------------------------------------
 \section{Circular distributional trees and forests}
 \label{sec:tree_forest}
-A common way of including covariates is to regress the model parameters on covariates.
-In the case of a probabilistic model for a circular response this corresponds to expressing the 
-location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
-by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
-appropriate link function plugged in as in GLMs. This kind of approach is presented for example
-in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
-%While applying the mixed model 
-%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
-%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
-%a circular response variable.
-Apart from the methodology developed specifically for circular data more general approaches to 
-regressing model parameters on covariates are commonly used such as for example generalized additive 
-models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
-where each distribution parameter is estimated by its own generalized additive model (GAM). 
-%However, an adaption of this methodology to circular data can be very complex due to a
-%challenging representation of smooth effects on the unit circle.
-%However, an adaption of this methodology to circular data can not be obtained by the straight 
-%forward approach of simply wrapping the modeled additive effects
-%around the unit circle since the parameters of the von Mises distribution and their 
-%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
-However, difficulties might arise in case of non-additive changes and a proper model 
-specification can be very challenging, particularly for a high number of covariates and no information
-on possible interactions.
-As an alternative regression approach which can deal with these problems, we introduce a circular probabilistic tree-structured model in this section.
+A common way of including covariates is to regress the model parameters on
+covariates. In the case of a probabilistic model for a circular response this
+corresponds to expressing the location parameter $\mu$ and the concentration
+parameter $\kappa$ of the von Mises distribution by means of the available
+covariates, e.g., in the form of a linear combination, possibly with an
+appropriate link function plugged in as in GLMs. This kind of approach is
+presented for example in the mixed model of \cite{Fisher+Lee:1992} and
+\cite{Fisher:1993}. Apart from the methodology developed specifically for
+circular data more general approaches to regressing model parameters on
+covariates are commonly used such as for example generalized additive models
+for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
+where each distribution parameter is estimated by its own generalized additive
+model (GAM). However, difficulties might arise in case of non-additive changes
+and a proper model specification can be very challenging, particularly for a
+high number of covariates and no information on possible interactions. As an
+alternative regression approach which can deal with these problems, we
+introduce a circular probabilistic tree-structured model in this section.
 
@@
 \label{sec:circtree}
-Fitting a global model to a full data set can be very challenging, particularly for 
-complex data with strong variations. Therefore, separating the data set into more homogeneous 
-subgroups based on covariates before fitting a local model in each of these subgroups allows 
-to capture (possibly) group specific effects more precisely and hence can result in a overall better 
-fitting model.
-This is the general idea of regression trees which are combined with distributional 
-modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
-algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
-By applying the presented distributional trees, a full distributional model is specified in each 
-node of the tree for a specific distribution family chosen in advance. Hence, 
-selecting the von Mises distribution allows for an application of distributional trees to circular
-data following the tree building algorithm as introduced in Section~2.2. of 
+Fitting a global model to a full data set can be very challenging, particularly
+for complex data with strong variations. Therefore, separating the data set
+into more homogeneous subgroups based on covariates before fitting a local
+model in each of these subgroups allows to capture (possibly) group specific
+effects more precisely and hence can result in a overall better fitting model.
+This is the general idea of regression trees which are combined with
+distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
+unbiased recursive partitioning algorithms MOB
+\citep{Zeileis+Hothorn+Hornik:2008} or CTree
+\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
+trees, a full distributional model is specified in each node of the tree for a
+specific distribution family chosen in advance. Hence, selecting the von Mises
+distribution allows for an application of distributional trees to circular data
+following the tree building algorithm as introduced in Section~2.2. of
 \cite{Schlosser+Hothorn+Stauffer:2019}.
 
-In the applied tree-building algorithm, in each node the crucial decision on how and where to split the data is based on model
-scores which are obtained by evaluating the score function $s$ 
-for the individual observations and parameter estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its two distribution parameters and a data set 
-of $n$ observations this yields an $n \times 2$ matrix that can be employed as a discrepancy measure,
-capturing how well each given observation conforms with the estimated location $\hat{\mu}$ and precision
-$\hat{\kappa}$, respectively.
-To capture dependence on covariates, the association between the model's scores and each available
-covariate is assessed using either a parameter instability test~(MOB) or a permutation test~(CTree).
-By doing so in each partitioning step, the covariate with the highest significant association (i.e., lowest
-significant $p$-value, if any) is selected for splitting the data. The corresponding split point
-is chosen either by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
-over all possible partitions.
-%and the covariate corresponding to the strongest change is selected as split varialbe.
-%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
-This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).
+In the applied tree-building algorithm, in each node the crucial decision on
+how and where to split the data is based on model scores which are obtained by
+evaluating the score function $s$ for the individual observations and parameter
+estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
+two distribution parameters and a data set of $n$ observations this yields an
+$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
+how well each given observation conforms with the estimated location
+$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
+on covariates, the association between the model's scores and each available
+covariate is assessed using either a parameter instability test~(MOB) or a
+permutation test~(CTree). By doing so in each partitioning step, the covariate
+with the highest significant association (i.e., lowest significant $p$-value,
+if any) is selected for splitting the data. The corresponding split point is
+chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
+statistic~(CTree) over all possible partitions. This procedure is repeated
+recursively until there are no significant parameter instabilities or until
+another stopping criterion is met (e.g., subgroup size or tree depth).
 
-%In particular, starting with the full learning sample, the steps for building a circular 
-%distributional tree are:
-%\begin{enumerate}
-%\item Estimate the parameter pair $(\hat{\mu}, \hat{\kappa})$ via maximum likelihood for the observations
-%  in the current (sub)sample.
-%\item Test for associations (or instabilities) of the scores $s(\hat{\mu}, \hat{\kappa}, y_i)$ 
-%  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
-%\item Split the sample along the partitioning variable $Z_l^*$ with the
-%  strongest association or instability. Choose the breakpoint leading to the highest
-%  improvement in model fit.
-%  %with the highest improvement in the log-likelihood or the highest discrepancy.
-%\item Repeat steps 1--3 recursively in the subsamples until these become too
-%  small or there is no significant association/instability (or some other
-%  stopping criterion is reached).
-%\end{enumerate}
+Once a distributional tree model is fitted it can be employed to obtain
+probabilistic predictions for a possibly new set of observed covariates $\bm{z}
+= (z_1, \ldots, z_m)$. Starting at the root node the tree structure leads the
+observation to a terminal node, depending on the values of the covariates,
+where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for the
+corresponding subset of learning observations. This can also be expressed by
+employing weights which indicate whether the $i$-th learning observation and
+the observation $\bm{z}$ belong to the same terminal node: \begin{equation}
+w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
+\mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), \end{equation} where
+$\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th
+out of $B$ segments partitioning the covariate space in disjoint subsets. In
+that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$
+specifying the predicted von Mises distribution for a given $\bm{z}$ is
+obtained by 
+%%
+\begin{equation} 
+  (\hat{\mu},\hat{\kappa})(\bm{z}) =
+  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot
+  \ell(\mu,\kappa; y_i).
+\end{equation}
+%%
+Therefore, the same parameter pair is estimated for all observations belonging
+to the same terminal node. This facilitates the application as the parameter
+estimates do not need to be recalculated for each (new) observation via maximum
+likelihood but can be extracted directly from the learning sample and the
+fitted model.
 
-%To test for associations between scores and covariates in step~3 different strategies can be applied
-%such as permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} or asymptotic M-fluctuation
-%tests for parameter instability \citep{Zeileis+Hornik:2007}.
-In order to select the covariate with the %highest significant association the one covariate that corresponds to the lowest significant $p$-value (if %any) is chosen for splitting the data.
-Within the range of this split variable the split point leading to %the highest improvement in model fit is chosen either by optimizing the log-likelihood \citep[as in the %MOB algorithm,][]{Zeileis+Hothorn+Hornik:2008} or a two-sample test statistic  over all possible partition%s \citep[as in the CTree algorithm][]{Hothorn+Hornik+Zeileis:2006}.
-
-Once a distributional tree model is fitted it can be employed to obtain probabilistic predictions
-for a possibly new set of observed covariates $\bm{z} = (z_1, \ldots, z_m)$.
-Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
-the corresponding subset of learning observations. 
-This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
-node:
-\begin{equation}
-w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
-\end{equation}
-where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
-partitioning the covariate space in disjoint subsets. 
-In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
-von Mises distribution for a given $\bm{z}$ is obtained by
-\begin{equation}
-(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
-\end{equation}
-
-Therefore, the same parameter pair is estimated for all observations
-belonging to the same terminal node. This facilitates the application as the parameter estimates do 
-not need to be recalculated for each (new) observation via maximum likelihood but can be extracted 
-directly from the learning sample and the fitted model.
-
-While tree models can capture non-additive effects their structure and the consequential strict 
-separation of data into subgroups hinders an adequate depiction of smooth effects. Therefore, 
-they might impose abrupt changes even in case of rather smooth transitions. This can be avoided 
-by combining an ensemble of trees in order to obtain a more stable forest model.
+While tree models can capture non-additive effects their structure and the
+consequential strict separation of data into subgroups hinders an adequate
+depiction of smooth effects. Therefore, they might impose abrupt changes even
+in case of rather smooth transitions. This can be avoided by combining an
+ensemble of trees in order to obtain a more stable forest model.
 
@@
 A natural extension of circular distributional trees are ensembles or forests
-that can improve forecasts by regularizing and stabilizing the model.
-Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
-of trees, each built on a subsample or bootstrap of the original data. 
-A generalization of this strategy is to obtain weighted predictions by adaptive local 
-likelihood estimation of the distributional parameters 
-\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
-More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
-weights $w_i^{forest}(\bm{z})$ 
-%\citep{Lin+Jeon:2006}
-is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
-node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
-Hence, for a forest of $T$ trees the weights are calculated via
+that can improve forecasts by regularizing and stabilizing the model. Random
+forests introduced by \cite{Breiman:2001} average the predictions of an
+ensemble of trees, each built on a subsample or bootstrap of the original data.
+A generalization of this strategy is to obtain weighted predictions by adaptive
+local likelihood estimation of the distributional parameters
+\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
+Hothorn+Zeileis:2017}. More specifically, for each possibly new
+observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
+$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
+which $\bm{z}$ is assigned to the same terminal node as each learning
+observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
+weights are calculated via
+%%
 \begin{equation}
@@
 \end{equation}
+%%
 where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
-segment of the $t$-th tree.
-Therefore, similar observations ending up more often in the same terminal node have higher weights
-and in that way a stronger influence in the estimation process.
+segment of the $t$-th tree. Therefore, similar observations ending up more
+often in the same terminal node have higher weights and in that way a stronger
+influence in the estimation process.
 
-In that way a specific set of weights can be calculated for each observation yielding
-its specific parameter estimates for the von Mises distribution
+In that way a specific set of weights can be calculated for each observation
+yielding its specific parameter estimates for the von Mises distribution
+%%
 \begin{equation}
-(\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
+  (\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu,
+  \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
 \end{equation}
+%%
 
 Therefore, the resulting parameter estimates can smoothly adapt to the given
-covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the unweighted
-full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$ corresponds to the abrupt
-splits from a tree.
-Thus, circular distributional forests can capture both smooth and abrupt changes, while
-covariates and possible interactions are selected automatically allowing for an easy-to-apply 
+covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the
+unweighted full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$
+corresponds to the abrupt splits from a tree. Thus, circular distributional
+forests can capture both smooth and abrupt changes, while covariates and
+possible interactions are selected automatically allowing for an easy-to-apply
 methodology with a full specification of the von Mises distribution.
 
+% -------------------------------------------------------------------
+% Case study
+% -------------------------------------------------------------------
 \section{Case study: Probabilistic wind direction forecasting}
@@
 Accurate forecasts of wind directions are of great importance for risk
-management in various fields such as agriculture, energy production or air safety control. 
-For example, in order to direct airplanes to a safe landing, precise knowledge of wind 
-direction for the next hour(s) at the respective airport is highly desirable and adequate 
-prediction methods are required. In this section, a comprehensive case study on probabilistic
-wind direction forecasting is provided for two airports in Austria with fundamentally
+management in various fields such as agriculture, energy production or air
+safety control. For example, in order to direct airplanes to a safe landing,
+precise knowledge of wind direction for the next hour(s) at the respective
+airport is highly desirable and adequate prediction methods are required. In
+this section, a comprehensive case study on probabilistic wind direction
+forecasting is provided for two airports in Austria with fundamentally
 different site characteristics: Innsbruck with a rather unusual location for an
 airport, located at the bottom of a deep Alpine valley, and Vienna located
-within flat terrain in the East of Austria. In addition to the newly
-introduced circular distributional trees and forests, the case study employs various
-alternative approaches to forecast wind direction as a benchmark. The study is based on
-$1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
-the airports as predictor variables.
+within flat terrain in the East of Austria. In addition to the newly introduced
+circular distributional trees and forests, the case study employs various
+alternative approaches to forecast wind direction as a benchmark. The study is
+based on $1$\,h and $3$\,h forecasts employing lagged observations in the
+vicinity of the airports as predictor variables.
 
-%The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
-%employed for probabilistic forecasting of wind directions at the airport of Innsbruck. As explained
-%and also supported by the results illustrated in the terminal nodes of the tree in 
-%Figure~\ref{fig:tree_ibk} the geographic location of Innsbruck and particularly the surrounding
-%mountain chains are a strongly influential component of the behavior of wind directions.
-%While this provides an interesting and challenging environment for testing different
-%statistical models it has to be considered as a rather unusual location for an airport.
-%Therefore, in order to allow for a broad and more general evaluation and also to test
-%for a high flexibility of the applied methods a second site with quite converse geographic 
-%features is included in this study as well.
-%Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
-%a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
-%area, is considered.
-%For the two locations the above introduced methodology of circular distributional trees and forests is
-%employed to wind direction data and additional meteorological measurements in order to obtain hourly 
-%wind direction forecasts, separately for Innsbruck and Vienna.
-
-\subsection{Data}\label{sec:wind:data}
-The circular response variable considered in this case study is a $10$\,min mean
-of wind direction measurements at the airports Innsbruck and Vienna on an hourly
-temporal resolution. As predictor variables, we use $1$-hourly resolved
+\subsection{Data}
+\label{sec:wind:data}
+The circular response variable considered in this case study is a $10$\,min
+mean of wind direction measurements at the airports Innsbruck and Vienna on an
+hourly temporal resolution. As predictor variables, we use $1$-hourly resolved
 $10$\,min mean observations of various meteorological quantities such as wind
-direction, wind speed, temperature, air pressure and humidity, all
-lagged by one or three hours according to the respective forecasting step. The
+direction, wind speed, temperature, air pressure and humidity, all lagged by
+one or three hours according to the respective forecasting step. The
 meteorological variables are measured either directly at the airports or within
-the their vicinity. For Innsbruck, measurements at the airport and along the intersecting valleys are
-used, whereas, for Vienna, measurements at the airport and
-within an area of about $10$\,km$^2$ are used. A topographical overview of the
-airports and their surrounding areas with the station sites employed in this
-study is provided in Figure~XXX. In addition, we use derived quantities such
-as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
+the their vicinity. For Innsbruck, measurements at the airport and along the
+intersecting valleys are used, whereas, for Vienna, measurements at the airport
+and within an area of about $10$\,km$^2$ are used. A topographical overview of
+the airports and their surrounding areas with the station sites employed in
+this study is provided in Figure~XXX. In addition, we use derived quantities
+such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
 changes and spatial differences towards the airport of the respective
@@
 The data used in this study consists of 5~total years from January 2015 to
-December 2018. After first eliminating predictor variables with more than 5\% missing
-values and then time points with missing observations, the
-data set consists of 41979 time points and 260 covariates for Innsbruck, and of
-38985 times points and 494 covariates for Vienna, respectively.
+December 2018. After first eliminating predictor variables with more than 5\%
+missing values and then time points with missing observations, the data set
+consists of 41979 time points and 260 covariates for Innsbruck, and of 38985
+times points and 494 covariates for Vienna, respectively.
 
+%%
 \begin{table}[t!]
-\caption[Table caption text]{Overview of the employed data sets for the case study: For Innsbruck and Vienna,
- various meteorological variables and derived quantities of these are considered at the respective stations, located either directly at the airports or in their vicinities.}
+  \caption[Table caption text]{Overview of the employed data sets for the case
+    study: For Innsbruck and Vienna, various meteorological variables and derived
+    quantities of these are considered at the respective stations, located either
+    directly at the airports or in their vicinities.}
   \label{tab:data}
@@
 \end{table}
+%%
 
-\subsection{Models and evaluation}\label{sec:wind:models}
+\subsection{Models and evaluation}
+\label{sec:wind:models}
 For a fair evaluation of circular distributional trees and forests, and to
@@
     quantity itself, either on a daily or yearly basis. Wind direction at a single
-location is often quite unstable and shows only a daily or annual
-seasonality in case of predominant local wind systems like the ones at
-Innsbruck introduced for the analysis of Figure~\ref{fig:tree_ibk}. Hence,
-the climatological model in this study provides a very simple implementation of
-a baseline model, which might experience serious performance losses in case of
-random temporal variabilities or abrupt changes in the employed response
-variable wind direction.
+    location is often quite unstable and shows only a daily or annual seasonality
+    in case of predominant local wind systems like the ones at Innsbruck introduced
+    for the analysis of Figure~\ref{fig:tree_ibk}. Hence, the climatological model
+    in this study provides a very simple implementation of a baseline model, which
+    might experience serious performance losses in case of random temporal
+    variabilities or abrupt changes in the employed response variable wind
+    direction.
     \newline
     For simplicity, rather to fit one single climatological model, we perform the
-model fitting separately rolling over all time points. For
-each time point, the training data set consists of the same daytime as the time
-of interest within 31~days centered around the same calender day over all years
-except the year of interest. To gain full probabilistic climatological
-forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
-the distribution parameters of the von Mises distribution via maximum
-likelihood estimation. The construction of a climatological reference model
-was similarly pursued by \citet{vogel.etal:2018} and is discussed in a
-comprehensive summary on different time-adaptive training schemes in
-\citet{lang.etal:2019c}.
+    model fitting separately rolling over all time points. For each time point, the
+    training data set consists of the same daytime as the time of interest within
+    31~days centered around the same calender day over all years except the year of
+    interest. To gain full probabilistic climatological forecasts, we employ the
+    steps described in Section~\ref{sec:prob_circ} and fit the distribution
+    parameters of the von Mises distribution via maximum likelihood estimation. The
+    construction of a climatological reference model was similarly pursued by
+    \citet{vogel.etal:2018} and is discussed in a comprehensive summary on
+    different time-adaptive training schemes in \citet{lang.etal:2019c}.
     %%
@@
     Persistence describes the previous value of a single weather quantity in a time
-series. Hence, it is another objective measure in the verification of
-weather forecasts and also often applied as a reference model
+    series. Hence, it is another objective measure in the verification of weather
+    forecasts and also often applied as a reference model
     \citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
@@
     predictions.
-%If a certain weather quantity remains constant over a certain time period,
-%persistence can be a very good estimate.
     \newline
     To gain a full probabilistic persistency model, we proceed similarly as for the
-climatological model by using maximum likelihood estimation and fitting the distribution
-parameters of the von Mises distribution conditional on lagged response values
-according to the description in Section~\ref{sec:prob_circ}. We fit one model
-for every time point in the validation data set employing the previous six
-lagged response values as training data. In order to allow for a stronger
-influence of observations closer to the time of interest exponential
+    climatological model by using maximum likelihood estimation and fitting the
+    distribution parameters of the von Mises distribution conditional on lagged
+    response values according to the description in Section~\ref{sec:prob_circ}. We
+    fit one model for every time point in the validation data set employing the
+    previous six lagged response values as training data. In order to allow for a
+    stronger influence of observations closer to the time of interest exponential
     smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
@@
     have an equal influence rate of 50 percent. Observations with longer time lags
-have exponential weights below $0.01$ and are therefore omitted
-from the training data.
+    have exponential weights below $0.01$ and are therefore omitted from the
+    training data.
     %%
   \item \emph{Generalized linear model:}
-Traditional approaches to forecast circular response variables are often based on
-circular GLM-typed models \citep{Fisher:1993}. As discussed in
-Section~\ref{sec:introduction}, circular regression models can experience
-the problem that the likelihood function can be strongly irregular which makes
+    Traditional approaches to forecast circular response variables are often based
+    on circular GLM-typed models \citep{Fisher:1993}. As discussed in
+    Section~\ref{sec:introduction}, circular regression models can experience the
+    problem that the likelihood function can be strongly irregular which makes
     optimization rather difficult. Hence, they often do not converge if no
@@
     as a reference, we use the Bayesian implementation of
-\cite{Mulder+Klugkist:2017} which less depends on initial values
-due to an MCMC sampling algorithm using weakly informative priors.
+    \cite{Mulder+Klugkist:2017} which less depends on initial values due to an MCMC
+    sampling algorithm using weakly informative priors.
     \newline
@@
     interval $(-\pi,\pi]$. As the implementation cannot handle circular covariates,
-we use the components of the lagged 2-dimensional wind vector $vec$ and the lagged
-wind speed $spd$ as predictor variables. The model formula for the location parameter
-$\mu$ of the von Mises distribution can be written as:
+    we use the components of the lagged 2-dimensional wind vector $vec$ and the
+    lagged wind speed $spd$ as predictor variables. The model formula for the
+    location parameter $\mu$ of the von Mises distribution can be written as:
     %%
@@
     with $\beta_0$ being a circular intercept, $\beta_\bullet$ the regression
-coefficients and $g$ the link function $g(x) = 2 \cdot \arctan(x)$. In addition, a
-constant concentration parameter $\kappa$ is fitted to the full learning sample.
+    coefficients and $g$ the link function $g(x) = 2 \cdot \arctan(x)$. In
+    addition, a constant concentration parameter $\kappa$ is fitted to the full
+    learning sample.
     %%
   \item \emph{Circular distributional tree:}
-For the circular distributional tree introduced in Section~\ref{sec:circtree}, all
-covariates provided in the learning data can be considered due to an intrinsic automatic
-variable selection performed in the tree estimation. The tree building is
-performed by the newly developed R-package \pkg{circtree} employing the CTree
-algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a minimal number of 2000
-observations in each terminal node serving as a stopping criterion (argument
-\code{minbucket}).
+    For the circular distributional tree introduced in Section~\ref{sec:circtree},
+    all covariates provided in the learning data can be considered due to an
+    intrinsic automatic variable selection performed in the tree estimation. The
+    tree building is performed by the newly developed R-package \pkg{circtree}
+    employing the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a
+    minimal number of 2000 observations in each terminal node serving as a stopping
+    criterion (argument \code{minbucket}).
     %%
   \item \emph{Circular distribution forest:} Following the description in
-Section~\ref{sec:circforest}, a circular forest is constructed based
-on 100 individual trees employing the R-package \pkg{circtree}. Each of these
-trees is again built by the CTree algorithm on a subsample containing a
-fraction of 30 percent of the original learning data. As it can be expected
-that the lagged response variable is one of the most informative predictor
-variables for wind direction forecasting, all covariates are included for
-building each tree which ensures that the lagged response
-variable is always considered for splitting. This bagging approach can be applied in \pkg{circtree} by
-setting the argument \code{mtry} to the total number of covariates. Since a
-high number of possible split points leads to high computational costs, the
-covariates are binned in a maximum of $50$ classes (argument
-\code{nmax = c("yx" = Inf, "z" = 50)}). Contrary to a single-tree model,
-forests usually consist of very large trees as they are not prone to overfitting
-the data due to the stabilization obtained by combining the individual trees. Therefore, we
-use the following control arguments to build rather large trees: The minimal number 
-of observations to perform a split is set to 20
-(argument \texttt{minsplit}), the minimal number of observations in each
-terminal node is set to 7 (argument \texttt{minbucket}), and the significance
-level for variable selection is kept at its maximum value of $1$
-(argument \texttt{alpha}).
+    Section~\ref{sec:circforest}, a circular forest is constructed based on 100
+    individual trees employing the R-package \pkg{circtree}. Each of these trees is
+    again built by the CTree algorithm on a subsample containing a fraction of 30
+    percent of the original learning data. As it can be expected that the lagged
+    response variable is one of the most informative predictor variables for wind
+    direction forecasting, all covariates are included for building each tree which
+    ensures that the lagged response variable is always considered for splitting.
+    This bagging approach can be applied in \pkg{circtree} by setting the argument
+    \code{mtry} to the total number of covariates. Since a high number of possible
+    split points leads to high computational costs, the covariates are binned in a
+    maximum of $50$ classes (argument \code{nmax = c("yx" = Inf, "z" = 50)}).
+    Contrary to a single-tree model, forests usually consist of very large trees as
+    they are not prone to overfitting the data due to the stabilization obtained by
+    combining the individual trees. Therefore, we use the following control
+    arguments to build rather large trees: The minimal number of observations to
+    perform a split is set to 20 (argument \texttt{minsplit}), the minimal number
+    of observations in each terminal node is set to 7 (argument
+    \texttt{minbucket}), and the significance level for variable selection is kept
+    at its maximum value of $1$ (argument \texttt{alpha}).
 \end{itemize}
@@
 
-In addition to the raw CRPS, also CRPS-based skill scores are computed to assess
-differences in the improvement of the various statistical models over the
-climatological model used as a reference:
+In addition to the raw CRPS, also CRPS-based skill scores are computed to
+assess differences in the improvement of the various statistical models over
+the climatological model used as a reference:
+%%
 \begin{equation} 
@@
 \end{equation}
+%%
 
-All scores presented in the next section are computed out-of-sample based on five years
-of observations. For the persistency model only time points prior to the time
-of interest are used and the validation is performed rolling over all
+All scores presented in the next section are computed out-of-sample based on
+five years of observations. For the persistency model only time points prior to
+the time of interest are used and the validation is performed rolling over all
 observations. For all other models a five-fold cross-validation is employed
@@
 score indicating better performance. For both stations at both forecast steps,
-the circular forest overall provides the best predictive
-performance, followed by the circular distributional tree and, except for the
-$3$\,h forecast at Innsbruck, the persistency model. In comparison to the
-circular tree and forest, for both stations and forecast steps, the
-climatological model and the linear model show clearly higher CRPS values and
-hence a lower predictive performance.
+the circular forest overall provides the best predictive performance, followed
+by the circular distributional tree and, except for the $3$\,h forecast at
+Innsbruck, the persistency model. In comparison to the circular tree and
+forest, for both stations and forecast steps, the climatological model and the
+linear model show clearly higher CRPS values and hence a lower predictive
+performance.
 
@@
 (Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
-(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the absolute
-level of the model performances and on their respective predictive performance variances.
-At Innsbruck, due to the surrounding mountains a limited number of possible
-wind directions exists, namely the three wind regimes discussed for
-Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
-Innsbruck the wind direction remains rather constant in one of these possible
-states, but once a change takes place it is mostly a major wind direction
-shift. Due to the few wind regimes the rather inflexible climatological
-and linear models score relatively well with similar CRPS values as the other models
+(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
+absolute level of the model performances and on their respective predictive
+performance variances. At Innsbruck, due to the surrounding mountains a
+limited number of possible wind directions exists, namely the three wind
+regimes discussed for Figure~\ref{fig:tree_ibk} in
+Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
+remains rather constant in one of these possible states, but once a change
+takes place it is mostly a major wind direction shift. Due to the few wind
+regimes the rather inflexible climatological and linear models score relatively
+well with similar CRPS values as the other models
 (Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
@@
 
-The different forecast steps have apparently only a minor effect on the predictive
-performance of the climatological model and the linear model at both stations.
-However, for the persistency model, at both stations, higher scores for the
-$3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a lower performance for longer lead
-times; this is probably due to minimum 3-hourly instead of 1-hourly lagged
-response values employed as covariates in the persistency model. The circular
-distributional tree and forest seem to partially compensate for the lower skill of
-the lagged response values by other covariates, hence their predictive
-performance only slightly decreases for the longer lead time. This compensation
-is especially evident for Innsbruck, where the performance difference between
-the persistency model and the tree-based methods significantly increases from
-the 1-hourly to the 3-hourly forecast.
-
+The different forecast steps have apparently only a minor effect on the
+predictive performance of the climatological model and the linear model at both
+stations. However, for the persistency model, at both stations, higher scores
+for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
+lower performance for longer lead times; this is probably due to minimum
+3-hourly instead of 1-hourly lagged response values employed as covariates in
+the persistency model. The circular distributional tree and forest seem to
+partially compensate for the lower skill of the lagged response values by other
+covariates, hence their predictive performance only slightly decreases for the
+longer lead time. This compensation is especially evident for Innsbruck, where
+the performance difference between the persistency model and the tree-based
+methods significantly increases from the 1-hourly to the 3-hourly forecast.
+%%
 \begin{figure}[t]
@@
   \setkeys{Gin}{width=1.0\textwidth}
-  %\hspace{-2cm}
-  %\includegraphics{boxplot_crpsraw.pdf}
   \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
@@
     airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
-    per for each of the $12$~months averaged over the $5$~validation years which yields a total of
-    $288$~yearly mean values. The scores are shown for the climatological, 
-    persistency and the linear model as well as for the circular distributional tree and
-    forest.}
+    per for each of the $12$~months averaged over the $5$~validation years which
+    yields a total of $288$~yearly mean values. The scores are shown for the
+    climatological, persistency and the linear model as well as for the circular
+    distributional tree and forest.}
   \label{fig:boxplot_crpsraw} 
 \end{figure}
+%%
 
-In addition to the raw CRPS (Figure~\ref{fig:boxplot_crpsraw}), CRPS
-skill scores with the climatological model as a reference are provided in
-Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in
-percent, and positive values indicate an improvement in the predictive
-performance over the reference. For all setups, the circular forest
-has the highest skill scores with a mean performance gain of 13--25\% and 58--71\% for
-Innsbruck and Vienna, respectively. As discussed for
-Figure~\ref{fig:boxplot_crpsraw}, this improvement over the climatological
-model is lower for Innsbruck due to the low number of predominant wind regimes
-and hence a relatively good performance of the climatological model.
-Additionally, Figure~\ref{fig:boxplot_crpsskill} shows that even if the
-persistency model's performance is lower than the reference (Panel~c) the
-tree-based models are significantly superior to the climatological model used as
-reference.
-
+In addition to the raw CRPS (Figure~\ref{fig:boxplot_crpsraw}), CRPS skill
+scores with the climatological model as a reference are provided in
+Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in percent, and positive
+values indicate an improvement in the predictive performance over the
+reference. For all setups, the circular forest has the highest skill scores
+with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and Vienna,
+respectively. As discussed for Figure~\ref{fig:boxplot_crpsraw}, this
+improvement over the climatological model is lower for Innsbruck due to the low
+number of predominant wind regimes and hence a relatively good performance of
+the climatological model. Additionally, Figure~\ref{fig:boxplot_crpsskill}
+shows that even if the persistency model's performance is lower than the
+reference (Panel~c) the tree-based models are significantly superior to the
+climatological model used as reference.
+%%
 \begin{figure}[t]
@@
   \setkeys{Gin}{width=1.0\textwidth}
-  %\hspace{-2cm}
-  %\includegraphics{boxplot_crpsskill.pdf}
   \includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
@@
 \end{figure}
+%%
 
-\section{Summary and conclusion}\label{sec:summary}
-Circular distributional trees for circular responses are established by coupling
-model-based recursive partitioning with the von Mises distribution.
-The resulting trees can capture non-linear changes, shifts, and potential interactions
-in covariates without prespecification of such effects. This is particularly
-useful for modeling a highly fluctating response such as wind direction. \dots
+% -------------------------------------------------------------------
+% Summary
+% -------------------------------------------------------------------
+\section{Summary and conclusion}
+\label{sec:summary}
+Circular distributional trees for circular responses are established by
+coupling model-based recursive partitioning with the von Mises distribution.
+The resulting trees can capture non-linear changes, shifts, and potential
+interactions in covariates without prespecification of such effects. This is
+particularly useful for modeling a highly fluctating response such as wind
+direction. \dots
 
-\fixme{add summary of results}
-\fixme{outlook on possible extensions of models, use of NWP data}
-
-%\subsection{Splits in circular covariates}
-
-%In order to obtain more parsimonious and more stable trees another possible
-%extension for \emph{circular covariates} (with or without a \emph{circular response})
-%is to consider their circular nature when searching the best split into two segments.
-%In general, searching the best separation of a covariate into a ``left'' and ``right''
-%daughter node tries to maximize the segmented log-likelihood:
-%\begin{equation}
-%\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} %\ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
-%\end{equation}
-%where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
-%of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a %circular covariate $\in [0, 2 \pi)$
-%only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
-%thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
-%for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
-%and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi)$, encompassing zero. The latter
-%strategy is invariant to the (often arbitrary) definition of the direction at zero.
-
-%When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
-%far away, a simple linear split typically suffices to capture such a split (as seen for the
-%lagged wind direction in Figure~\ref{fig:tree_ibk}). If both $\nu$ and $\tau$ differ
-%clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
-%fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
-%might miss such a pattern.
-
-%The required test statistic to maximally select two split points simultaneously is straightforward
-%to accommodate in the CTree framework by providing all binary indicators corresponding
-%to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
-%increasingly slow for larger sample sizes but it might be possible to speed up computations by
-%exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
-%MOB framework the extension is not quite as straightforward but one strategy could be to
-%adapt double maximum tests \`a la \cite{Bai+Perron:2003}.
-
-%Hence, the splitting idea can be naturally extended to a two-point search, however, for 
-%an unbiased and inference-based selection the corresponding testing strategies might need
-%further adaption.
+\begin{itemize}
+  \item Add summary/conclusion of case study;
+  \item Add summary/conclusion of circular trees and forests.
+  \item Outlook on possible extensions (use of NWP data, circular splits, ...).
+\end{itemize}
 
@@
 \section*{Computational details}
-
-The proposed circular tree and forest models are implemented in the \textsf{R} package \textbf{circtree}
-(version~\Sexpr{packageVersion("circtree")}). The package is based on the \textbf{disttree}
-package (version~\Sexpr{packageVersion("disttree")}) which applies the main
-tree building function from the  \textbf{partykit} package 
+The proposed circular tree and forest models are implemented in the \textsf{R}
+package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The
+package is based on the \textbf{disttree} package
+(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
+building functions from the \textbf{partykit} package
 (version~\Sexpr{packageVersion("partykit")}). All three packages are available
@@
 
-For the circular GLM considered as reference model the corresponding implementation is provided 
-in the \textsf{R} package \textbf{circglmbayes} by \cite{Mulder+Klugkist:2017}. In particular the 
-function \textbf{circGLM} is applied to estimates the intercept and regression coefficient along with 
-the concentration parameter $\kappa$.
+For the circular GLM considered as reference model the corresponding
+implementation is provided in the \textsf{R} package \textbf{circglmbayes} by
+\cite{Mulder+Klugkist:2017}. In particular the function \textbf{circGLM} is
+applied to estimates the intercept and regression coefficient along with the
+concentration parameter $\kappa$.
 

### 2019-11-21 pkg/circtree/inst/wind_paper/draft/draft_2019_11_20.tex
Remove all comments and clean up Rnw file
- additionally spell check
- Add draft version 2019-11-20
- Update makefile (for diff)


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2894 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f82750fb5fb8c4263ddf38abe3b4a195471ef852

--- 2019-11-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
@@
 
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-library("disttree")
-library("circtree")
-
-## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-# set function for parallelization
-applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))
-
-@
-
 \title{Circular Distributional Trees and Forests with an Application to 
@@
 in a Bayesian framework. 
-%\fixme{revise literature survey with focus on transition to GLM}
 The named methods and also many of their further developments \citep[see,

### 2019-11-22 pkg/circtree/inst/wind_paper/draft/draft.Rnw
minor revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2895 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e4133dbdd6289ad239ad1444e5b20980384adf54

@@
 mixed model. While being built on well elaborated theory this methodology has
-shown to be very challenging in their application, mainly due to the complexity
+shown to be very challenging in its application, mainly due to the complexity
 encountered in optimizing the log-likelihood function which is not globally
@@
 to regression modeling of circular data by employing a GLM in a Bayesian
-framework. The listed methods and most of their further developments
+framework. All listed methods and most of their further developments
 \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
@@
 % -------------------------------------------------------------------
+\newpage
 \section{Probabilistic circular modeling}
@@
 an indicator of deviation and in that way a measure of goodness of fit of the
-model can be obtained for each observation $y_i$ and fitted parameter.
+model can be obtained for each observation and fitted parameter.
 Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
@@
 model in each of these subgroups allows to capture (possibly) group specific
-effects more precisely and hence can result in a overall better fitting model.
+effects more precisely and hence can result in an overall better fitting model.
 This is the general idea of regression trees which are combined with
@@
 estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
-two distribution parameters and a data set of $n$ observations this yields an
+two distribution parameters and a data set of $n$ observations, this yields an
 $n \times 2$ matrix that can be employed as a discrepancy measure, capturing
@@
 meteorological variables are measured either directly at the airports or within
-the their vicinity. For Innsbruck, measurements at the airport and along the
+their vicinities. For Innsbruck, measurements at the airport and along the
 intersecting valleys are used, whereas, for Vienna, measurements at the airport
-and within an area of about $10$\,km$^2$ are used. A topographical overview of
+and within an area of about XXX\,km$^2$ are used. A topographical overview of
 the airports and their surrounding areas with the station sites employed in
@@
 
-The data used in this study consists of 5~total years from January 2015 to
+The data used in this study consists of five total years from January 2015 to
 December 2018. After first eliminating predictor variables with more than 5\%
@@
 \begin{table}[t!]
-  \caption[Table caption text]{Overview of the employed data sets for the case
+  \caption[Table caption text]{Overview of the data sets employed in the case
     study: For Innsbruck and Vienna, various meteorological variables and derived
@@
     Accurate knowledge of weather quantities' climatologies can be important for a
-    wide range of applications and are often also used as a baseline for the
+    wide range of applications and climatologies are often also used as a baseline for the
     validation of newly developed forecasting systems
@@
     on circular GLM-typed models \citep{Fisher:1993}. As discussed in
-    Section~\ref{sec:introduction}, circular regression models can experience the
+    Section~\ref{sec:introduction}, circular regression models often experience the
     problem that the likelihood function can be strongly irregular which makes
@@
     Following \cite{Mulder+Klugkist:2017} the employed model plugs in an
-    appropriate link function~$g$ in order to guarantee response values on the
-    interval $(-\pi,\pi]$. As the implementation cannot handle circular covariates,
+    appropriate link function~$g$ in order to guarantee response values within an
+    interval of length $2\pi$. As the implementation cannot handle circular covariates,
     we use the components of the lagged 2-dimensional wind vector $vec$ and the
@@
     %%
-  \item \emph{Circular distribution forest:} Following the description in
+  \item \emph{Circular distributional forest:} Following the description in
     Section~\ref{sec:circforest}, a circular forest is constructed based on 100
@@
 at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
-Vienna (Panels~b,\,d). The scores are aggregated over the $5$~validation years,
+Vienna (Panels~b,\,d). The scores are aggregated over the five validation years,
 yielding yearly mean values for every hour per calendar month, with a lower
@@
 performance.
-
+%%
+\begin{figure}[t]
+  \centering
+  \setkeys{Gin}{width=1.0\textwidth}
+  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
+  \caption{CRPS skill scores for wind direction forecasts based on the full
+    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
+    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
+    for each of the $12$~months averaged over the $5$~validation years which
+    yields a total of $288$~yearly mean values. The scores are shown for the
+    climatological, persistency and the linear model as well as for the circular
+    distributional tree and forest.}
+  \label{fig:boxplot_crpsraw} 
+\end{figure}
+%%
 The different site characteristics of the airports Innsbruck
@@
 for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
-lower performance for longer lead times; this is probably due to minimum
-3-hourly instead of 1-hourly lagged response values employed as covariates in
-the persistency model. The circular distributional tree and forest seem to
+lower performance for longer lead times; this is probably due to the lower 
+information content of minimum 3-hourly instead of 1-hourly lagged response 
+values employed as covariates in the persistency model. 
+The circular distributional tree and forest seem to
 partially compensate for the lower skill of the lagged response values by other
@@
 methods significantly increases from the 1-hourly to the 3-hourly forecast.
-%%
-\begin{figure}[t]
-  \centering
-  \setkeys{Gin}{width=1.0\textwidth}
-  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
-  \caption{CRPS skill scores for wind direction forecasts based on the full
-    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
-    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
-    per for each of the $12$~months averaged over the $5$~validation years which
-    yields a total of $288$~yearly mean values. The scores are shown for the
-    climatological, persistency and the linear model as well as for the circular
-    distributional tree and forest.}
-  \label{fig:boxplot_crpsraw} 
-\end{figure}
-%%
 
@@
 the climatological model. Additionally, Figure~\ref{fig:boxplot_crpsskill}
-shows that even if the persistency model's performance is lower than the
-reference (Panel~c) the tree-based models are significantly superior to the
-climatological model used as reference.
+shows that while the persistency model's performance is lower than the
+reference (Panel~c) the tree-based models can compensate for the low skill of the 
+lagged response values employed as covariates and, hence, are still significantly 
+superior to the reference.
+
 %%
@@
 interactions in covariates without prespecification of such effects. This is
-particularly useful for modeling a highly fluctating response such as wind
+particularly useful for modeling a highly fluctuating response such as wind
 direction. \dots
@@
 \section*{Computational details}
-The proposed circular tree and forest models are implemented in the \textsf{R}
-package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The
-package is based on the \textbf{disttree} package
+The corresponding implementation of the proposed methodology for circular
+trees and forests is provided in the \textsf{R}
+package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}).
+% The proposed circular tree and forest models are implemented in the \textsf{R}
+% package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). 
+The package is based on the \textbf{disttree} package
 (version~\Sexpr{packageVersion("disttree")}) which applies the main tree

### 2019-11-22 pkg/circtree/inst/wind_paper/draft/draft.Rnw
minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2896 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2e854dae3aac7aa72a0c4f5590324f9171eb015e

@@
 winds blowing from the valley mouth towards the upper valley (from east to
-west, nodes 4 and 5). (2) Downslope winds blowing across the Alpine crest
+west, nodes~4 and 5). (2) Downslope winds blowing across the Alpine crest
 along the intersecting valley towards Innsbruck (from south-east to north-west,
@@
     \smallskip &,
-    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
+    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
     \\
@@
 applied to estimates the intercept and regression coefficient along with the
-concentration parameter $\kappa$.
+concentration parameter.
 

### 2019-11-25 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revisions on abstract and introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2897 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 13d860415908960772a21bb81941abb8ebaa64eb

@@
 
-\title{Circular Distributional Trees and Forests with an Application to 
-    Wind Direction Forecasting}
-\Shorttitle{Circular Trees and Forests}
+\title{Circular Regression Trees and Forests with an Application to
+Probabilistic Wind Direction Forecasting}
+\Shorttitle{Circular Regression Trees and Forests}
 
@@
 \Abstract{
-As circular data appears in many different areas, the development of
-appropriate statistical models which can account for periodicity has become an
-important but challenging research field. For probabilistic modeling of
-circular data typically the von Mises distribution is employed. In order to
-regress its parameters on covariates, a circular distributional tree is
-proposed as an alternative to more commonly-used additive models. The resulting
-circular trees are easy to interpret, can detect non-additive effects, and
-select covariates and their interactions automatically. By combining an
-ensemble of circular trees a more stable and regularized circular
-distributional forest can be built which allows for modeling smooth effects as
+Establishing a proper regression model for circular data such as animal movement
+directions, periodic measurements of neuronal activity or wind directions has shown
+to be very challenging. Commonly applied generalized linear and additive models 
+can provide a useful toolbox, however, non-additive effects might not be detected 
+and difficulties may arise in modeling smooth effects due to highly non-linear link 
+functions and increasing irregularities when moving closer to the ``origin point'' 
+(0 which is equal to $2\pi$ in the radiant case).
+%% FIXME: clear explanation of difficulties
+Therefore, we propose a new approach by applying distributional trees and 
+forests to circular data. In particular, the parameters of the employed von Mises 
+distribution are regressed on covariates by a tree-based model yielding a constant set 
+of estimated distribution parameters for each resulting subgroup, regardless of any 
+``circular shift''/the ``circular position''.
+%As circular data appears in many different areas, the development of
+%appropriate statistical models which can account for periodicity has become an
+%important but challenging research field. For probabilistic modeling of
+%circular data typically the von Mises distribution is employed. In order to
+%regress its parameters on covariates, a circular regression tree is
+%proposed as an alternative to more commonly-used generalized linear and additive 
+%models. 
+The resulting circular regression trees are easy to interpret, can detect non-additive effects, 
+and select covariates and their interactions automatically. By combining an
+ensemble of circular regression trees a more stable and regularized circular
+regression forest can be built which allows for modeling smooth effects as
 well.
 
-To evaluate the proposed circular trees and forests, a case study on
+To evaluate the proposed circular regression trees and forests, a case study on
 probabilistic forecasting of wind direction for two airports in Austria is
@@
 are used as a benchmark in order to assess the predictive performances of
-circular trees and forests. 
+circular regression trees and forests. 
 }
-\Keywords{Distributional Trees and Forests; Circular Response; von Mises Distribution}
+\Keywords{Distributional Trees, Random Forests; Circular Data; von Mises Distribution}
 
@@
 subsequently by \cite{Fisher:1993} introducing another specific transformation
-function which maps the values of the real line to the unit circle. Moreover,
-the presented methodology allows for both distribution parameters, the mean
-direction and the concentration parameter of the von Mises distribution, to be
-modeled depending on covariates, not only in separate models but also in a
-mixed model. While being built on well elaborated theory this methodology has
-shown to be very challenging in its application, mainly due to the complexity
-encountered in optimizing the log-likelihood function which is not globally
-concave. Therefore, highly informative starting values are crucial for the
-model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
-Gill+Hangartner:2010}. In order to avoid this strong dependence on appropriate
-initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
-to regression modeling of circular data by employing a GLM in a Bayesian
-framework. All listed methods and most of their further developments
-\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
-regression models. While these generally provide a powerful toolbox for
-modeling particularly smooth effects, they may not be able to capture all types
-of effects adequately. Especially abrupt changes might not be detected
-properly owing to the enforced additive structure and model assumptions, such
-as a possibly highly non-linear link function and the form of linear
-combinations.
+function which maps the values of the real line to the unit circle. 
+Moreover, the presented methodology does not only allow for the mean direction to be modeled 
+depending on covariates while a constant estimate is obtained for the concentration parameter 
+but also for the vice versa version. For each of these two approaches a separate model is introduced, 
+however, they are also combined in a joint model (referred to as ``mixed model'').
+%Moreover, the presented methodology allows for both distribution parameters, the mean
+%direction and the concentration parameter of the von Mises distribution, to be
+%modeled depending on covariates, not only separately but also in a
+%combined model (also referred to ``mixed model''). 
+In the employed iterative optimization algorithm the two sets of regression coefficients used
+to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
+%To obtain proper initial values it is recommended to apply two separate models 
+%where either one of the parameters is fixed while the other is predicted employing covariates 
+%in a GLM-type model.
+While the combined model but also the two separate models are built on well elaborated theory 
+the employed methodology has shown to be very challenging in its application, mainly due to the 
+complexity encountered in optimizing the log-likelihood function which is not globally concave. 
+Therefore, highly informative starting values are crucial for the model to converge at all 
+\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
+In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
+present an alternative approach to regression modeling of circular data by employing a GLM in a 
+Bayesian framework, however, only the mean direction is regressed on covariates.
+All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
+are built on additive regression models.
+Generally these provide a powerful toolbox for modeling particularly smooth effects,
+however, in the case of circular data a possibly highly non-linear link function
+can induce a more complex behavior of the modeled parameters as the location
+parameter shifts closely towards the ``origin point'' $0$, also depending on the 
+direction of the shift.
+Moreover, not all types effects of may be modeled adequately. Especially abrupt changes 
+might not be detected properly owing to the enforced additive structure and selected model 
+specification such as the form of linear combinations.
+%% FIXME: clear explanation of difficulties close to 0
 
-An alternative modeling approach that can deal with non-additive effects is
-provided by tree-based methods. \cite{Lund:2002} introduced the concept of
-circular regression trees which was further extended by
-\cite{Larraondo+Inza+Lozano:2018}. Following this idea of employing tree-based
-methods but also including the probabilistic concept of regressing distribution
-parameters on covariates, we propose a very flexible approach by applying the
-von Mises distribution within the methodology of distributional trees and
-distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
-data. The resulting circular distributional trees and forests do not require
-any model assumptions as covariates and their possible interactions are
-selected automatically. Non-additive effects are captured by the tree structure
-while circular forests allow for modeling smooth changes as well. Thus, a high
-level of flexibility is provided while a full distributional model is specified
-offering a wide range of inference.
+In order to avoid these difficulties in probabilistic regression modeling of circular data 
+(and also the requirement of initial values) we propose an alternative and very flexible 
+approach by applying the von Mises distribution within the methodology of distributional 
+trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
+The resulting circular regression trees and forests 
+(for short: \emph{circular trees and forests}) 
+do not require any model assumptions as covariates and their possible interactions are
+selected automatically. Non-additive effects are captured by the tree structure while 
+circular forests allow for modeling smooth changes as well. Thus, a high level of 
+flexibility is provided while a full distributional model is specified in each terminal 
+node offering a wide range of inference. Moreover, due to the parameters being estimated 
+as constants for each subgroup difficulties connected to ``circular shifts'' close to $0$ 
+are avoided.
 
-As a motivating example, a circular distributional tree is employed to
-probabilistic wind direction forecasting. Wind direction is a classical
-circular quantity and accurate forecasts are of great importance for
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+This novel approach to circular regression trees and forests complements the literature 
+on tree-based circular modeling.
+%In particular, 
+\cite{Lund:2002} have already introduced the concept of circular regression 
+trees where binary splits are made based on a circular distance which is evaluated between 
+the observed response values and the conditional mean direction of the circular response variable
+and in that way serves as a measure of homogeneity within subgroups. 
+%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
+%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
+%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
+%subset of observations.
+While this approach and also further extensions \citealp{Larraondo+Inza+Lozano:2018} allow not only for 
+linear but also for a circular covariate, the tree splits are performed based on effects in the expected 
+value only, i.e., only the location parameter $\mu$ is regressed on covariates.
+%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
+%along with the predicted mean value in each terminal node
+Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
+of tree-based circular models by incorporating the framework of distributional modeling and in that way
+considering splits in all distribution parameters and providing a fully specified distribution fitted 
+in each terminal node.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%An alternative modeling approach that can deal with non-additive effects is
+%provided by tree-based methods. \cite{Lund:2002} introduced the concept of
+%circular regression trees which was further extended by
+%\cite{Larraondo+Inza+Lozano:2018}. Following this idea of employing tree-based
+%methods but also including the probabilistic concept of regressing distribution
+%parameters on covariates, we propose a very flexible approach by applying the
+%von Mises distribution within the methodology of distributional trees and
+%distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
+%data. The resulting circular distributional trees and forests do not require
+%any model assumptions as covariates and their possible interactions are
+%selected automatically. Non-additive effects are captured by the tree structure
+%while circular forests allow for modeling smooth changes as well. Thus, a high
+%level of flexibility is provided while a full distributional model is specified
+%offering a wide range of inference.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsubsection*{Motivating example}
+To provide a first impression of the presented methodology a circular regression 
+tree is employed to probabilistic wind direction forecasting. Wind direction is a 
+classical circular quantity and accurate forecasts are of great importance for
 decision-making processes especially in air traffic management.
@@
 wind speed at the airport.
-
-Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
-empirical (gray) and fitted von Mises (red) distribution of wind direction in
-each terminal node. Based on the fitted location parameters $\hat \mu$, the
-subgroups can be distinguished into the following wind regimes: (1)~Up-valley
-winds blowing from the valley mouth towards the upper valley (from east to
-west, nodes~4 and 5). (2) Downslope winds blowing across the Alpine crest
-along the intersecting valley towards Innsbruck (from south-east to north-west,
-node 8). (3) Down-valley winds blowing in the direction of the valley mouth
-(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
-rather low wind speeds that cannot be distinguished clearly into wind regimes
-and consequently are associated with a very low estimated concentration $\hat
-\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
-mostly responsible for distinguishing the broad wind regimes listed above while
-the pressure gradients and wind speed separate between subgroups with high vs.\
-low precision. Note that in the meteorological context wind direction 
-is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
 %%
-\begin{figure}[t]
+\begin{figure}[p!]
   \centering
-  \includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
+  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
   \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
@@
 %%
+Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
+empirical (gray) and fitted von Mises (red) distribution of wind direction in
+each terminal node. Based on the fitted location parameters $\hat \mu$, the
+subgroups can be distinguished into the following wind regimes: (1)~Up-valley
+winds blowing from the valley mouth towards the upper valley (from east to
+west, nodes~4 and 5). (2) Downslope winds blowing across the Alpine crest
+along the intersecting valley towards Innsbruck (from south-east to north-west,
+node~8). (3) Down-valley winds blowing in the direction of the valley mouth
+(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
+rather low wind speeds that cannot be distinguished clearly into wind regimes
+and consequently are associated with a very low estimated concentration $\hat
+\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
+mostly responsible for distinguishing the broad wind regimes listed above while
+the pressure gradients and wind speed separate between subgroups with high vs.\
+low precision. Note that in the meteorological context wind direction 
+is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
 
-A more extensive case study of circular distributional trees and forests
+A more extensive case study of circular regression trees and forests
 applied to probabilistic forecasting of wind direction at the airports of
@@
 circular modeling introducing the von Mises distribution will be discussed in
-Section~\ref{sec:prob_circ}. The methodology of circular distributional trees
+Section~\ref{sec:prob_circ}. The methodology of circular regression trees
 and forests and their features are introduced in Section~\ref{sec:tree_forest}.
@@
 % -------------------------------------------------------------------
-\newpage
+%\newpage
 \section{Probabilistic circular modeling}
@@
 of data, including circular data.
+Already existing generalized linear or additive approaches to circular regression 
+models employ linear combinations of the covariates to connect them to one (or more) 
+distribution parameters. 
+%For the mean direction~$\mu$ \cite{Gould:1969} proposes the model 
+%\begin{equation}
+%\mu = \mu_0 + \sum_j \beta_j\cdot x_j(\textit{mod } 2\pi)
+%\end{equation}
+%with a circular intercept $\mu_0$ and $x$ being a vector of covariates
+%with the corresponding vector of coefficients $\beta$.
+By following this idea and including a link function $g$ many other models
+have been developed in the general form 
+\begin{equation}
+\mu = \mu_0 + g(\beta^{\top} x).
+\end{equation}
+For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
+on an intervall of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
+distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
+$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. The concentration parameter
+
+%% FIXME: information on already existing generalized linear and additive approaches
+%% with formulas (\mu = g(X \beta) + \mu_0)   FIXME: add mu_0?
+%% clear description of difficulties
 
 % -------------------------------------------------------------------
-% Circular distributional trees and forests
+% Circular regression trees and forests
 % -------------------------------------------------------------------
-\section{Circular distributional trees and forests}
+\section{Circular regression trees and forests}
 \label{sec:tree_forest}
+
+%% FIXME: what are we aiming at? which problems do we want to avoid/solve?
+%% mu and kappa as constants in subgroups, no initial values
+%% Lund: what do they present (including formulas)
+%% compare, what is different, how do we improve this approach
+
 A common way of including covariates is to regress the model parameters on
@@
 
-\subsection{Circular trees}
+\subsection{Circular regression trees}
 \label{sec:circtree}
@@
 another stopping criterion is met (e.g., subgroup size or tree depth).
+A more detailed description of the applied tree-building algorithm,
+particularly of CTree, can be found in Appendix~\ref{app:tree}.
 
@@
 \label{sec:circforest} 
-A natural extension of circular distributional trees are ensembles or forests
+A natural extension of circular regression trees are ensembles or forests
 that can improve forecasts by regularizing and stabilizing the model. Random
@@
 unweighted full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$
-corresponds to the abrupt splits from a tree. Thus, circular distributional
+corresponds to the abrupt splits from a tree. Thus, circular regression
 forests can capture both smooth and abrupt changes, while covariates and
@@
 within flat terrain in the East of Austria. In addition to the newly introduced
-circular distributional trees and forests, the case study employs various
+circular regression trees and forests, the case study employs various
 alternative approaches to forecast wind direction as a benchmark. The study is
@@
 \label{sec:wind:models}
-For a fair evaluation of circular distributional trees and forests, and to
+For a fair evaluation of circular regression trees and forests, and to
 investigate whether they can be applied as a reasonable alternative to already
@@
 specifications used in this study for all considered models including the
-circular tree and forest.
+circular regression tree and forest.
 
@@
     validation of newly developed forecasting systems
-    \citep{simon.etal:2017,stauffer.etal:2017}. Usually, the performance of
+    \citep{Simon+Umlauf+Zeileis:2017, Stauffer+Mayr+Messner:2017}. Usually, the performance of
     climatological models strongly depends on the temporal dependency of the
@@
     construction of a climatological reference model was similarly pursued by
-    \citet{vogel.etal:2018} and is discussed in a comprehensive summary on
-    different time-adaptive training schemes in \citet{lang.etal:2019c}.
+    \citet{Vogel+Knippertz+Fink:2018} and is discussed in a comprehensive summary on
+    different time-adaptive training schemes in \citet{Lang+Lerch+Mayr:2019}.
     %%
@@
     %%
-  \item \emph{Circular distributional tree:}
-    For the circular distributional tree introduced in Section~\ref{sec:circtree},
+  \item \emph{Circular regression tree:}
+    For the circular regression tree introduced in Section~\ref{sec:circtree},
     all covariates provided in the learning data can be considered due to an
@@
     %%
-  \item \emph{Circular distributional forest:} Following the description in
+  \item \emph{Circular regression forest:} Following the description in
     Section~\ref{sec:circforest}, a circular forest is constructed based on 100
@@
 the circular forest overall provides the best predictive performance, followed
-by the circular distributional tree and, except for the $3$\,h forecast at
-Innsbruck, the persistency model. In comparison to the circular tree and
+by the circular regression tree and, except for the $3$\,h forecast at
+Innsbruck, the persistency model. In comparison to the circular regression tree and
 forest, for both stations and forecast steps, the climatological model and the
@@
 values employed as covariates in the persistency model. 
-The circular distributional tree and forest seem to
+The circular regression tree and forest seem to
 partially compensate for the lower skill of the lagged response values by other
@@
 \label{sec:summary}
-Circular distributional trees for circular responses are established by
+Circular regression trees for circular responses are established by
 coupling model-based recursive partitioning with the von Mises distribution.
@@
   \item Add summary/conclusion of case study;
-  \item Add summary/conclusion of circular trees and forests.
+  \item Add summary/conclusion of circular regression trees and forests.
   \item Outlook on possible extensions (use of NWP data, circular splits, ...).
@@
 implementation is provided in the \textsf{R} package \textbf{circglmbayes} by
-\cite{Mulder+Klugkist:2017}. In particular the function \textbf{circGLM} is
+\cite{Mulder+Klugkist:2017}. In particular the function \code{circGLM} is
 applied to estimates the intercept and regression coefficient along with the
@@
 
+\begin{appendix}
+\section{Tree-building algorithms}
+\label{app:tree}
+CTree, MOB
+\end{appendix}
 \end{document}

### 2019-11-25 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revisions on abstract and introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2898 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 008c34f7c84d6000f50baf27532300881343f061

@@
 parameter of the von Mises distribution is predicted by a linear combination of
-covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
+covariates. 
+\cite{Johnson+Wehrly:1978} took this idea one step further by
 plugging in a link function such that the values of the linear predictor are
-transformed to a restricted interval of length $2\pi$. This generalized linear
+transformed to a restricted interval of length $2\pi$. 
+%This generalized linear
+%model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
+%subsequently by \cite{Fisher:1993} introducing another specific transformation
+%function which maps the values of the real line to the unit circle. 
+This generalized linear
 model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
-subsequently by \cite{Fisher:1993} introducing another specific transformation
-function which maps the values of the real line to the unit circle. 
-Moreover, the presented methodology does not only allow for the mean direction to be modeled 
-depending on covariates while a constant estimate is obtained for the concentration parameter 
-but also for the vice versa version. For each of these two approaches a separate model is introduced, 
-however, they are also combined in a joint model (referred to as ``mixed model'').
-%Moreover, the presented methodology allows for both distribution parameters, the mean
-%direction and the concentration parameter of the von Mises distribution, to be
-%modeled depending on covariates, not only separately but also in a
-%combined model (also referred to ``mixed model''). 
-In the employed iterative optimization algorithm the two sets of regression coefficients used
-to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
-%To obtain proper initial values it is recommended to apply two separate models 
-%where either one of the parameters is fixed while the other is predicted employing covariates 
-%in a GLM-type model.
-While the combined model but also the two separate models are built on well elaborated theory 
+subsequently by \cite{Fisher:1993} introducing independent GLMs for either the 
+location or scale parameter with appropriate link functions while keeping the other 
+parameter constant. Additionally, they introduce a combined hetereoscedastic version 
+where the location and scale parameter are alternately reestimated conditional on 
+respective sets of covariates until convergence.
+%Moreover, the presented methodology does not only allow for the mean direction to be modeled 
+%depending on covariates while a constant estimate is obtained for the concentration parameter 
+%but also for the vice versa version. For each of these two approaches a separate model is introduced, 
+%however, they are also combined in a joint model (referred to as ``mixed model'').
+%%Moreover, the presented methodology allows for both distribution parameters, the mean
+%%direction and the concentration parameter of the von Mises distribution, to be
+%%modeled depending on covariates, not only separately but also in a
+%%combined model (also referred to ``mixed model''). 
+%In the employed iterative optimization algorithm the two sets of regression coefficients used
+%to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
+%%To obtain proper initial values it is recommended to apply two separate models 
+%%where either one of the parameters is fixed while the other is predicted employing covariates 
+%%in a GLM-type model.
+%While the combined model but also the two separate models are built on well elaborated theory 
+While all of these models are built on well elaborated theory 
 the employed methodology has shown to be very challenging in its application, mainly due to the 
@@
 direction of the shift.
-Moreover, not all types effects of may be modeled adequately. Especially abrupt changes 
+Moreover, not all types of effects may be modeled adequately. Especially abrupt changes 
 might not be detected properly owing to the enforced additive structure and selected model 
@@
 model remarkably as they might provide valuable additional information.
-Therefore, distributional regression models are of high interest for all types
+Therefore, (distributional) regression models are of high interest for all types
 of data, including circular data.
-Already existing generalized linear or additive approaches to circular regression 
-models employ linear combinations of the covariates to connect them to one (or more) 
-distribution parameters. 
-%For the mean direction~$\mu$ \cite{Gould:1969} proposes the model 
-%\begin{equation}
-%\mu = \mu_0 + \sum_j \beta_j\cdot x_j(\textit{mod } 2\pi)
-%\end{equation}
-%with a circular intercept $\mu_0$ and $x$ being a vector of covariates
-%with the corresponding vector of coefficients $\beta$.
-By following this idea and including a link function $g$ many other models
-have been developed in the general form 
+
+In already existing generalized linear or additive approaches to circular regression modeling 
+usually the location parameter $\mu$ of is regressed on 
+covariates in the general form 
+%models employ linear combinations of the covariates to connect them to distribution 
+%parameters. By plugging in a link function $g$ 
 \begin{equation}
@@
 \end{equation}
+with a circular intercept $\mu_0$ and $x$ being a vector of covariates
+with the corresponding vector of coefficients $\beta$ that are to be estimated.
 For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
-on an intervall of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
+on an interval of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
 distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
-$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. The concentration parameter
-
-%% FIXME: information on already existing generalized linear and additive approaches
-%% with formulas (\mu = g(X \beta) + \mu_0)   FIXME: add mu_0?
-%% clear description of difficulties
+$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
+The concentration parameter $\kappa$ is mostly estimated as a constant, however,
+\cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on covariates
+employing the exponential function as a link function while $\mu$ is assumed to be a fixed
+value. But regardless of whether $\mu$ or $\kappa$ depend on covariates,
+both variations have shown to be very complex in their optimization procedure and hence
+challenging in their application. Consequently, these difficulties intensify even more in 
+a combined model where the sets of regression 
+coefficients for the two distribution parameters $\mu$ and $\kappa$ respectively are reestimated 
+alternatingly in an iterative optimization algorithm. Hence, regardless of difficulties in practice, 
+the presented theory provides a first approach to probabilistic regression modeling of circular data, 
+however, the parameters are not regressed simultaneously on covariates as for example in a more general
+case by generalized additive models for location, scale, and shape 
+\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
+Moreover, the additive structure of the discussed models might impose a smooth effect even if the 
+true underlying effect is an abrupt shift.
+One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
+is to employ tree-based models as presented in the following Section.
+%% FIXME: better description of difficulties?
 
@@
 \label{sec:tree_forest}
-
 %% FIXME: what are we aiming at? which problems do we want to avoid/solve?

### 2019-11-26 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update draft (intro)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2899 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5026cbd44fb6b5144594f5f292b335b0a2eac778

@@
 location or scale parameter with appropriate link functions while keeping the other 
-parameter constant. Additionally, they introduce a combined hetereoscedastic version 
+parameter constant. Additionally, they developed a combined hetereoscedastic version 
 where the location and scale parameter are alternately reestimated conditional on 
@@
 the employed methodology has shown to be very challenging in its application, mainly due to the 
-complexity encountered in optimizing the log-likelihood function which is not globally concave. 
-Therefore, highly informative starting values are crucial for the model to converge at all 
+complexity encountered in optimizing the corresponding log-likelihood function which is not globally concave. 
+Therefore, highly informative starting values are crucial for these circular GLMs to converge at all 
 \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
 In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
-present an alternative approach to regression modeling of circular data by employing a GLM in a 
-Bayesian framework, however, only the mean direction is regressed on covariates.
+present a Bayesian alternative of a homoscedastic GLM for circular data.
 All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
 are built on additive regression models.
-Generally these provide a powerful toolbox for modeling particularly smooth effects,
-however, in the case of circular data a possibly highly non-linear link function
-can induce a more complex behavior of the modeled parameters as the location
-parameter shifts closely towards the ``origin point'' $0$, also depending on the 
-direction of the shift.
-Moreover, not all types of effects may be modeled adequately. Especially abrupt changes 
+Generally these provide a powerful toolbox for modeling particularly smooth effects.
+However, in the case of modeling circular data 
+addititve effects conditional on linear covariates are not intuitive
+due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
+can be obtained in either positive or negative direction on the circle.
+%a possibly highly non-linear link function
+%can induce a more complex behavior of the modeled parameters as the location
+%parameter shifts closely towards the ``origin point'' $0$, also depending on the 
+%direction of the shift.
+Moreover, abrupt changes 
 might not be detected properly owing to the enforced additive structure and selected model 
 specification such as the form of linear combinations.
-%% FIXME: clear explanation of difficulties close to 0
-
-In order to avoid these difficulties in probabilistic regression modeling of circular data 
-(and also the requirement of initial values) we propose an alternative and very flexible 
-approach by applying the von Mises distribution within the methodology of distributional 
-trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
+
+As a very intuitive and data-driven alternative, we propose a flexible
+tree-based regression approach for modeling circular data by applying the von
+Mises distribution within the methodology of distributional trees and
+distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
+%In order to avoid these difficulties in regression modeling of circular data 
+%(and also the requirement of initial values) 
+%we propose an alternative and very flexible 
+%approach by applying the von Mises distribution within the methodology of distributional 
+%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
 The resulting circular regression trees and forests 
-(for short: \emph{circular trees and forests}) 
-do not require any model assumptions as covariates and their possible interactions are
+do not require a pre-specified link function or any other model assumptions as covariates and their possible interactions are
 selected automatically. Non-additive effects are captured by the tree structure while 
-circular forests allow for modeling smooth changes as well. Thus, a high level of 
+forests allow for modeling smooth changes as well. Thus, a high level of 
 flexibility is provided while a full distributional model is specified in each terminal 
-node offering a wide range of inference. Moreover, due to the parameters being estimated 
-as constants for each subgroup difficulties connected to ``circular shifts'' close to $0$ 
-are avoided.
+node offering a wide range of inference. Moreover, as covariates are employed for splitting only and not
+considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.
 
@@
 %In particular, 
-\cite{Lund:2002} have already introduced the concept of circular regression 
-trees where binary splits are made based on a circular distance which is evaluated between 
-the observed response values and the conditional mean direction of the circular response variable
-and in that way serves as a measure of homogeneity within subgroups. 
-%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
-%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
-%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
-%subset of observations.
-While this approach and also further extensions \citealp{Larraondo+Inza+Lozano:2018} allow not only for 
-linear but also for a circular covariate, the tree splits are performed based on effects in the expected 
-value only, i.e., only the location parameter $\mu$ is regressed on covariates.
+\cite{Lund:2002} has already introduced the concept of circular regression 
+trees where binary splits are made based on an angular distance measure employed to evalute node homogeneity.
+%which is evaluated between 
+%the observed response values and the conditional mean direction of the circular response variable
+%and in that way serves as a measure of homogeneity within subgroups. 
+%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
+%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
+%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
+%%subset of observations.
+While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not only for 
+linear but also for a circular covariate, the tree only considers effects in the mean direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
 %Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
 %along with the predicted mean value in each terminal node
-Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
-of tree-based circular models by incorporating the framework of distributional modeling and in that way
-considering splits in all distribution parameters and providing a fully specified distribution fitted 
-in each terminal node.
+By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.
+
+%Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
+%of tree-based circular models by incorporating the framework of distributional modeling and in that way
+%considering splits in all distribution parameters and providing a fully specified distribution fitted 
+%in each terminal node.
 

### 2019-11-26 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision on final part of section 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2900 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f14c174edd245f1dbd8e13c5b368822026a633b8

@@
 location or scale parameter with appropriate link functions while keeping the other 
-parameter constant. Additionally, they developed a combined hetereoscedastic version 
+parameter constant. Additionally, they developed a combined heteroscedastic version 
 where the location and scale parameter are alternately reestimated conditional on 
@@
 However, in the case of modeling circular data 
-addititve effects conditional on linear covariates are not intuitive
+additive effects conditional on linear covariates are not intuitive
 due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
@@
 \cite{Lund:2002} has already introduced the concept of circular regression 
-trees where binary splits are made based on an angular distance measure employed to evalute node homogeneity.
+trees where binary splits are made based on an angular distance measure employed to evaluate node homogeneity.
 %which is evaluated between 
@@
 In already existing generalized linear or additive approaches to circular regression modeling 
-usually the location parameter $\mu$ of is regressed on 
-covariates in the general form 
+usually the location parameter $\mu$ is regressed on covariates in the general form 
 %models employ linear combinations of the covariates to connect them to distribution 
@@
 \begin{equation}
-\mu = \mu_0 + g(\beta^{\top} x).
+\mu = \mu_0 + g(\beta^{\top} z).
 \end{equation}
-with a circular intercept $\mu_0$ and $x$ being a vector of covariates
-with the corresponding vector of coefficients $\beta$ that are to be estimated.
+with a circular intercept $\mu_0$, a vector of covariates $z$
+with the corresponding vector of coefficients $\beta$ that are to be estimated,
+and a link function $g$.
 For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
@@
 $g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
-The concentration parameter $\kappa$ is mostly estimated as a constant, however,
-\cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on covariates
-employing the exponential function as a link function while $\mu$ is assumed to be a fixed
-value. But regardless of whether $\mu$ or $\kappa$ depend on covariates,
-both variations have shown to be very complex in their optimization procedure and hence
-challenging in their application. Consequently, these difficulties intensify even more in 
-a combined model where the sets of regression 
-coefficients for the two distribution parameters $\mu$ and $\kappa$ respectively are reestimated 
-alternatingly in an iterative optimization algorithm. Hence, regardless of difficulties in practice, 
-the presented theory provides a first approach to probabilistic regression modeling of circular data, 
-however, the parameters are not regressed simultaneously on covariates as for example in a more general
-case by generalized additive models for location, scale, and shape 
+Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
+as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
+covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
+constant value. Additionally to these two model variations where only one of the two distribution 
+parameters depends on covariates a combination of them is introduced in a joint model. There, the 
+two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
+in alternation within an iterative optimization algorithm. 
+Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
+data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
+more general case by generalized additive models for location, scale, and shape 
 \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
+However, in practice the incorporated optimization procedure has shown to be very complex, already
+for only one parameter being regressed on covariates and consequently even more so in the combined model.
+This is mainly due to characteristics of the log-likelihood function of the employed von Mises distribution
+with the plugged-in link function. In particular, the function is not globally concave and possibly very 
+sharp local maxima can prevent the algorithm from converging.
 Moreover, the additive structure of the discussed models might impose a smooth effect even if the 
 true underlying effect is an abrupt shift.
+%% FIXME: difficulties in modeling smooth effects on the unit circle
 One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
 is to employ tree-based models as presented in the following Section.
-%% FIXME: better description of difficulties?
 

### 2019-11-27 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision on the final part of section 2 and the beginning of section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2901 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ecc1143bfc3733cc8672ddd486aa0b8fc102f3bb

@@
 For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
-on an interval of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
-distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
+on an interval of length $2\pi$ by setting 
+$g(x) = 2\pi \cdot F(x)$ 
+for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
 $g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
@@
 constant value. Additionally to these two model variations where only one of the two distribution 
-parameters depends on covariates a combination of them is introduced in a joint model. There, the 
+parameters depends on covariates a combination of them is introduced. In this joint model the 
 two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
@@
 However, in practice the incorporated optimization procedure has shown to be very complex, already
-for only one parameter being regressed on covariates and consequently even more so in the combined model.
-This is mainly due to characteristics of the log-likelihood function of the employed von Mises distribution
-with the plugged-in link function. In particular, the function is not globally concave and possibly very 
-sharp local maxima can prevent the algorithm from converging.
-Moreover, the additive structure of the discussed models might impose a smooth effect even if the 
-true underlying effect is an abrupt shift.
-%% FIXME: difficulties in modeling smooth effects on the unit circle
+for only one parameter being regressed on covariates and consequently even more so in the joint model.
+This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
+distribution with the plugged-in link function. In particular, the function is not globally concave and
+possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
+Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
+is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
+covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
+compression of the estimated density function.
+\newline
+Referring to additive models in general and hence also to their application on circular data, it has to
+be considered that a proper model specification can be very challenging, particularly for a high number 
+of covariates and no information on possible interactions. Moreover, the additive structure might impose 
+a smooth effect even  if the true underlying effect is an abrupt shift.
+\newline
 One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
@@
 \label{sec:tree_forest}
-%% FIXME: what are we aiming at? which problems do we want to avoid/solve?
-%% mu and kappa as constants in subgroups, no initial values
-%% Lund: what do they present (including formulas)
-%% compare, what is different, how do we improve this approach
-
-A common way of including covariates is to regress the model parameters on
-covariates. In the case of a probabilistic model for a circular response this
-corresponds to expressing the location parameter $\mu$ and the concentration
-parameter $\kappa$ of the von Mises distribution by means of the available
-covariates, e.g., in the form of a linear combination, possibly with an
-appropriate link function plugged in as in GLMs. This kind of approach is
-presented for example in the mixed model of \cite{Fisher+Lee:1992} and
-\cite{Fisher:1993}. Apart from the methodology developed specifically for
-circular data more general approaches to regressing model parameters on
-covariates are commonly used such as for example generalized additive models
-for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
-where each distribution parameter is estimated by its own generalized additive
-model (GAM). However, difficulties might arise in case of non-additive changes
-and a proper model specification can be very challenging, particularly for a
-high number of covariates and no information on possible interactions. As an
-alternative regression approach which can deal with these problems, we
-introduce a circular probabilistic tree-structured model in this section.
+The main goal of employing tree-based models to circular data is to provide a flexible and 
+easy-to-apply modeling approach which allows to detect additive as well as abrupt changes while
+accounting for a proper representation on the unit circle. 
+%A very flexible framework should allow for detecting a wide range of 
+%effects, from abrupt changes to smooth transitions, while for the latter difficulties related
+%to a circular representation should be avoided as well as any dependence on initial values.
+\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
+variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
+and where to split the data an angular distance measure is evaluated between the mean direction 
+and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
+Hence, only the location parameter is considered for splitting such that effects in other parameters 
+might be missed. Therefore, we introduce a distributional approach to circular regression trees 
+where a fully specified distributional model is fitted in each node of the tree which enables the 
+detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
+corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
+quantities.
+Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
+within each node is fitted without employing covariates.
+%A common way of including covariates is to regress the model parameters on
+%covariates. In the case of a probabilistic model for a circular response this
+%corresponds to expressing the location parameter $\mu$ and the concentration
+%parameter $\kappa$ of the von Mises distribution by means of the available
+%covariates, e.g., in the form of a linear combination, possibly with an
+%appropriate link function plugged in as in GLMs. This kind of approach is
+%presented for example in the mixed model of \cite{Fisher+Lee:1992} and
+%\cite{Fisher:1993}. Apart from the methodology developed specifically for
+%circular data more general approaches to regressing model parameters on
+%covariates are commonly used such as for example generalized additive models
+%for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
+%where each distribution parameter is estimated by its own generalized additive
+%model (GAM). However, difficulties might arise in case of non-additive changes
+%and a proper model specification can be very challenging, particularly for a
+%high number of covariates and no information on possible interactions. As an
+%alternative regression approach which can deal with these problems, we
+%introduce a circular probabilistic tree-structured model in this section.
 

### 2019-11-28 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update draft according to lisa/moritz morning revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2902 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 89f3a2f64a6d5bb6acb9b16a7ac4d28fbd7dde2b

@@
 variable of interest follows a circular distribution, in particular the von
-Mises distribution which is also known as the ``circular normal distribution''
-(see Section~\ref{sec:prob_circ} for more details). This is also the basis
-which early regression models for circular data are built on as they regress
-the parameter(s) of the von Mises distribution on covariates. One of the first
+Mises distribution which is also known as the ``circular normal distribution''.
+%This is also the basis
+%which early regression models for circular data are built on as they regress
+%the parameter(s) of the von Mises distribution on covariates. 
+One of the first
 regression models with a circular response variable and linear covariates was
@@
 covariates. 
-\cite{Johnson+Wehrly:1978} took this idea one step further by
-plugging in a link function such that the values of the linear predictor are
-transformed to a restricted interval of length $2\pi$. 
+\cite{Johnson+Wehrly:1978} refined this idea by
+plugging in a link function transforming the linear predictor to a 
+restricted interval of length $2\pi$. 
+%%%
+%%%In order to restrict the circular response to an interval of length $2\,pi\
+%%%\cite{Johnson+Wehrly:1978} refined this idea by linintroducing link functions transforming
+%%%the linear predictor to the restricted interval of length $2\pi$. 
+%\cite{Johnson+Wehrly:1978} took this idea one step further by
+%plugging in a link function such that the values of the linear predictor are
+%transformed to a restricted interval of length $2\pi$. 
 %This generalized linear
@@
 present a Bayesian alternative of a homoscedastic GLM for circular data.
-All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
-are built on additive regression models.
-Generally these provide a powerful toolbox for modeling particularly smooth effects.
-However, in the case of modeling circular data 
-additive effects conditional on linear covariates are not intuitive
-due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
+However, apart from potential difficulties in the optimization procedure of circular GLMs, 
+additive effects and their interpretation are not intuitive for circular responses.
+Due to the highly non-linear link function, the representation of smooth transitions on the unit circle
+is not straightforward; additionally, the same rotation
 can be obtained in either positive or negative direction on the circle.
+%%%All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
+%%%are built on additive regression models.
+%%%Generally these provide a powerful toolbox for modeling particularly smooth effects.
+%%%however, in the case of modeling circular data 
+%%%additive effects conditional on linear covariates are not intuitive
+%%%due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
+%%%can be obtained in either positive or negative direction on the circle.
 %a possibly highly non-linear link function
@@
 %direction of the shift.
-Moreover, abrupt changes 
-might not be detected properly owing to the enforced additive structure and selected model 
-specification such as the form of linear combinations.
 
@@
 considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.
+Moreover, abrupt changes 
+might not be detected properly owing to the enforced additive structure and selected model 
+specification such as the form of linear combinations.
 

### 2019-11-29 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2903 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2d0f5072b8edbdfe0e03245dccf1b82b60388198

@@
 However, apart from potential difficulties in the optimization procedure of circular GLMs, 
-additive effects and their interpretation are not intuitive for circular responses.
-Due to the highly non-linear link function, the representation of smooth transitions on the unit circle
-is not straightforward; additionally, the same rotation
-can be obtained in either positive or negative direction on the circle.
+additive effects and their interpretation are not intuitive for circular responses
+due to the highly non-linear link function. Consequently, the representation of smooth transitions 
+on the unit circle is not straightforward. In addition, the same rotation can be obtained in either 
+positive or negative direction on the circle leading to an ambiguous interpretation.
+
 %%%All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
@@
 %trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
-The resulting circular regression trees and forests 
-do not require a pre-specified link function or any other model assumptions as covariates and their possible interactions are
-selected automatically. Non-additive effects are captured by the tree structure while 
-forests allow for modeling smooth changes as well. Thus, a high level of 
-flexibility is provided while a full distributional model is specified in each terminal 
-node offering a wide range of inference. Moreover, as covariates are employed for splitting only and not
-considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.
-Moreover, abrupt changes 
-might not be detected properly owing to the enforced additive structure and selected model 
-specification such as the form of linear combinations.
+The resulting circular regression trees and forests avoid the discussed difficulties 
+of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
+simple intercept-only model for each resulting subset.
+Therefore, no covariates and, thus, no link function must be employed in the estimation
+of the group-specific distribution parameters.
+Moreover, by incorporating the framework of distributional modeling we allow for detecting 
+splits in all distribution parameters and provide a fully specified circular response distribution 
+for each terminal node offering a wide range of inference.
+In addition, the employed tree structure captures non-additive effects while forests allow for 
+modeling smooth changes as well.
+Furhtermore, covariates and their possible interactions do not need to specified in advance as they are
+selected automatically.
+
+%Moreover, non-additive effects are captured by the tree structure while forests allow for 
+%modeling smooth changes as well. Thus, a high level of flexibility is provided while a full 
+%distributional model is specified in each terminal node offering a wide range of inference.
+
+%The resulting circular regression trees and forests do not require a pre-specified link function 
+%or any other model assumptions as covariates and their possible interactions are
+%selected automatically.
+
+%Moreover, as covariates are employed for splitting only and not
+%considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.
 
@@
 %In particular, 
-\cite{Lund:2002} has already introduced the concept of circular regression 
-trees where binary splits are made based on an angular distance measure employed to evaluate node homogeneity.
+\cite{Lund:2002} has already introduced the concept of circular regression trees where 
+binary splits are made based on an angular distance measure employed to evaluate node 
+homogeneity.
+While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not 
+only for linear but also for a circular covariate, the tree only considers effects in the mean 
+direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
+Therefore, the introduced distributional approach to circular regression trees extends the concept
+of previous circular trees by considering splits in all distribution parameters and 
+yielding a full probabilistic model. In addition, as a natural further development ensembles or forests 
+of circular regression trees are presented in this study. 
+% that can improve forecasts by regularizing and stabilizing the model.
+
 %which is evaluated between 
@@
 %%subset of observations.
-While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not only for 
-linear but also for a circular covariate, the tree only considers effects in the mean direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
 %Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
 %along with the predicted mean value in each terminal node
-By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.
+%By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.
 

### 2019-11-29 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of last part of section 2 and introduction of section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2904 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4e43a915a9643696e80909a94261c890ff5867ea

@@
 of the group-specific distribution parameters.
-Moreover, by incorporating the framework of distributional modeling we allow for detecting 
+By incorporating the framework of distributional modeling we allow for detecting 
 splits in all distribution parameters and provide a fully specified circular response distribution 
@@
 modeling smooth changes as well.
-Furhtermore, covariates and their possible interactions do not need to specified in advance as they are
+Furthermore, covariates and their possible interactions do not need to specified in advance as they are
 selected automatically.
@@
 
-In already existing generalized linear or additive approaches to circular regression modeling 
-usually the location parameter $\mu$ is regressed on covariates in the general form 
+In most generalized linear or additive approaches to circular regression modeling 
+the location parameter $\mu$ is regressed on covariates $\mathbf{z}$ in the form of 
 %models employ linear combinations of the covariates to connect them to distribution 
@@
 \begin{equation}
-\mu = \mu_0 + g(\beta^{\top} z).
+\mu = \mu_0 + g(\mathbf{\beta}^{\top} \mathbf{z})
 \end{equation}
-with a circular intercept $\mu_0$, a vector of covariates $z$
-with the corresponding vector of coefficients $\beta$ that are to be estimated,
-and a link function $g$.
-For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
-on an interval of length $2\pi$ by setting 
-$g(x) = 2\pi \cdot F(x)$ 
-for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
-$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
-Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
-as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
-covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
-constant value. Additionally to these two model variations where only one of the two distribution 
-parameters depends on covariates a combination of them is introduced. In this joint model the 
-two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
-in alternation within an iterative optimization algorithm. 
-Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
-data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
-more general case by generalized additive models for location, scale, and shape 
+with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$; whereby the 
+the additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 
+As an appropriate link function typically $g(x) = 2 \cdot arctan(x)$ is employed, 
+as suggested by \citet{Fisher+Lee:1992}.
+While most of the proposed models estimate a constant concentration parameter, hence, assume
+homogeneity, \citet{Fisher+Lee:1992} also provide a heteroscedastic version by combining two  
+individual GLMs, each for one of the parameters $\mu$ and $\kappa$. 
+This provides a first approach to a fully probabilistic regression model for circular 
+data, however, the parameters are not regressed simultaneously on covariates as known from the 
+more general framework provided by generalized additive models for location, scale, and shape 
 \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. 
-However, in practice the incorporated optimization procedure has shown to be very complex, already
-for only one parameter being regressed on covariates and consequently even more so in the joint model.
-This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
-distribution with the plugged-in link function. In particular, the function is not globally concave and
-possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
-Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
-is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
-covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
-compression of the estimated density function.
-\newline
-Referring to additive models in general and hence also to their application on circular data, it has to
-be considered that a proper model specification can be very challenging, particularly for a high number 
-of covariates and no information on possible interactions. Moreover, the additive structure might impose 
-a smooth effect even  if the true underlying effect is an abrupt shift.
-\newline
-One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
-is to employ tree-based models as presented in the following Section.
+However, all circular additive models incorporate the previously discussed difficulties induced by
+the characteristics of the log-likelihood function and the strongly non-linear link function.
+%For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
+%on an interval of length $2\pi$ by setting 
+%$g(x) = 2\pi \cdot F(x)$ 
+%for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
+%$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
+%Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
+%as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
+%covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
+%constant value. Additionally to these two model variations where only one of the two distribution 
+%parameters depends on covariates a combination of them is introduced. In this joint model the 
+%two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
+%in alternation within an iterative optimization algorithm. 
+%Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
+%data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
+%more general case by generalized additive models for location, scale, and shape 
+%\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
+%However, in practice the incorporated optimization procedure has shown to be very complex, already
+%for only one parameter being regressed on covariates and consequently even more so in the joint model.
+%This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
+%distribution with the plugged-in link function. In particular, the function is not globally concave and
+%possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
+%Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
+%is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
+%covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
+%compression of the estimated density function.
+%\newline
+Referring to additive models in general, it has to be considered that a proper model specification 
+can be very challenging, particularly for a high number of covariates and no information on possible 
+interactions. Moreover, the additive structure might impose a smooth effect even if the true underlying 
+effect is an abrupt shift.
+
+An alternative to additive models, which can deal or even 
+avoid the mentioned difficulties in circular regression modeling, is to employ tree-based models. 
+For the particular case of circular data, the methodology of circular regression trees and forests is
+presented in the following section.
 
@@
 \label{sec:tree_forest}
+
 The main goal of employing tree-based models to circular data is to provide a flexible and 
-easy-to-apply modeling approach which allows to detect additive as well as abrupt changes while
-accounting for a proper representation on the unit circle. 
+easy-to-apply modeling approach which accounts for a proper representation on the unit circle.
+Based on ideas of \citet{Lund:2002}, we introduce probabilistic circular regression trees and forests
+employing the von Mises distribution. The resulting tree-based models provide a very intuitive and data-driven 
+alternative to commonly used circular regression models for circular data.
+
 %A very flexible framework should allow for detecting a wide range of 
@@
 %to a circular representation should be avoided as well as any dependence on initial values.
-\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
-variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
-and where to split the data an angular distance measure is evaluated between the mean direction 
-and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
-Hence, only the location parameter is considered for splitting such that effects in other parameters 
-might be missed. Therefore, we introduce a distributional approach to circular regression trees 
-where a fully specified distributional model is fitted in each node of the tree which enables the 
-detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
-corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
-quantities.
-Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
-within each node is fitted without employing covariates.
+%\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
+%variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
+%and where to split the data an angular distance measure is evaluated between the mean direction 
+%and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
+%Hence, only the location parameter is considered for splitting such that effects in other parameters 
+%might be missed. Therefore, we introduce a distributional approach to circular regression trees 
+%where a fully specified distributional model is fitted in each node of the tree which enables the 
+%detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
+%corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
+%quantities.
+%Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
+%within each node is fitted without employing covariates.
 %A common way of including covariates is to regress the model parameters on

### 2019-11-29 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revised abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2905 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5b3b5277e9fa1e6f4167208a94c2646237fe0d8d

@@
 \Abstract{
-Establishing a proper regression model for circular data such as animal movement
-directions, periodic measurements of neuronal activity or wind directions has shown
-to be very challenging. Commonly applied generalized linear and additive models 
-can provide a useful toolbox, however, non-additive effects might not be detected 
-and difficulties may arise in modeling smooth effects due to highly non-linear link 
-functions and increasing irregularities when moving closer to the ``origin point'' 
-(0 which is equal to $2\pi$ in the radiant case).
-%% FIXME: clear explanation of difficulties
-Therefore, we propose a new approach by applying distributional trees and 
-forests to circular data. In particular, the parameters of the employed von Mises 
-distribution are regressed on covariates by a tree-based model yielding a constant set 
-of estimated distribution parameters for each resulting subgroup, regardless of any 
-``circular shift''/the ``circular position''.
+Despite the occurrence of circular data in a wide range of scientific
+fields, the existing methodology for probabilistically forecasting
+circular data
+is highly limited. Most of the existing methods are built on the
+framework of
+generalized linear and additive models, which can incorporate
+difficulties in
+both the optimization and the interpretation. As a very intuitive
+alternative
+to additive circular regression models, we extend previous ideas of circular
+trees and introduce probabilistic circular regression trees and forests
+employing the von Mises distribution. The resulting tree-based models
+simplify
+the estimation process by separating the data into more homogeneous
+subgroups
+and fitting a simple intercept-only model for each resulting subset. By
+incorporating the framework of distributional modeling, splits in all
+distribution parameters are considered and a fully specified distribution is
+provided for each node.  The proposed circular regression trees are easy to
+interpret, can detect non-additive effects, and select covariates and their
+interactions automatically. By combining an ensemble of circular regression
+trees a more stable and regularized circular regression forest can be built
+which allows for modeling smooth effects as well.
+
+To evaluate the proposed circular regression trees and forests, a case
+study on
+probabilistic forecasting of wind direction for two airports in Austria is
+presented. In addition to the tree-based models, alternative common
+approaches
+are used as a benchmark in order to assess the predictive performances of
+circular regression trees and forests.
+%Establishing a proper regression model for circular data such as animal movement
+%directions, periodic measurements of neuronal activity or wind directions,
+%requires specific attention regarding its representation on the unit circle.
+%Commonly applied generalized linear and additive models can provide a useful toolbox, 
+%however, highly non-linear link functions can complicate interpretations while
+%a non-concave log-likelihood function of the usually employed von Mises distribution 
+%leads to difficulties in the incorporated optimization procedure.
+%Moreover, an adequate model specification can be very challenging and non-additive 
+%effects might not be detected.
+
+%Therefore, an alternative approach to modeling circular data is proposed by
+%employing the von Mises distribution within the methodology of distributional 
+%trees and forests. Exploiting the tree-based separation of the data set into 
+%subgroups an intercept-model is fitted within each subset such that no covariates 
+%are included simplifying ....
+%difficulties may arise in 
+%modeling smooth effects due to highly non-linear link 
+%functions and increasing irregularities when moving closer to the ``origin point'' 
+%(0 which is equal to $2\pi$ in the radiant case).
+%%% FIXME: clear explanation of difficulties
+%Therefore, we propose a new approach by applying distributional trees and 
+%forests to circular data. In particular, the parameters of the employed von Mises 
+%distribution are regressed on covariates by a tree-based model yielding a constant set 
+%of estimated distribution parameters for each resulting subgroup, regardless of any 
+%``circular shift''/the ``circular position''.
 %As circular data appears in many different areas, the development of
@@
 %models. 
-The resulting circular regression trees are easy to interpret, can detect non-additive effects, 
-and select covariates and their interactions automatically. By combining an
-ensemble of circular regression trees a more stable and regularized circular
-regression forest can be built which allows for modeling smooth effects as
-well.
-
-To evaluate the proposed circular regression trees and forests, a case study on
-probabilistic forecasting of wind direction for two airports in Austria is
-presented. In addition to the tree-based models, alternative common approaches
-are used as a benchmark in order to assess the predictive performances of
-circular regression trees and forests. 
+%Moreover, the resulting circular regression trees are easy to interpret, 
+%can detect non-additive effects, and select covariates and their interactions automatically. 
+%By combining an ensemble of circular regression trees a more stable and regularized circular
+%regression forest can be built which allows for modeling smooth effects as
+%well.
+
+%To evaluate the proposed circular regression trees and forests, a case study on
+%probabilistic forecasting of wind direction for two airports in Austria is
+%presented. In addition to the tree-based models, alternative common approaches
+%are used as a benchmark in order to assess the predictive performances of
+%circular regression trees and forests. 
 }
@@
 additive effects and their interpretation are not intuitive for circular responses
-due to the highly non-linear link function. Consequently, the representation of smooth transitions 
-on the unit circle is not straightforward. In addition, the same rotation can be obtained in either 
+due to the highly non-linear link function and the representation of smooth transitions 
+on the unit circle is not straightforward. For example, the same rotation can be obtained in either 
 positive or negative direction on the circle leading to an ambiguous interpretation.
@@
 Therefore, no covariates and, thus, no link function must be employed in the estimation
-of the group-specific distribution parameters.
-By incorporating the framework of distributional modeling we allow for detecting 
+of the group-specific distribution parameters which also simplifies the optimization process.
+By incorporating the framework of distributional modeling, we allow for detecting 
 splits in all distribution parameters and provide a fully specified circular response distribution 
@@
 modeling smooth changes as well.
-Furthermore, covariates and their possible interactions do not need to specified in advance as they are
-selected automatically.
+Furthermore, covariates and their possible interactions do not need to be specified in advance as 
+they are selected automatically.
 
@@
 of previous circular trees by considering splits in all distribution parameters and 
-yielding a full probabilistic model. In addition, as a natural further development ensembles or forests 
+yielding a full probabilistic model. In addition, as a natural further development, ensembles or forests 
 of circular regression trees are presented in this study. 
@@
 \begin{equation}
-\mu = \mu_0 + g(\mathbf{\beta}^{\top} \mathbf{z})
+\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
 \end{equation}
 with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$; whereby the 
-the additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 
+additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 
 As an appropriate link function typically $g(x) = 2 \cdot arctan(x)$ is employed, 
@@
 While most of the proposed models estimate a constant concentration parameter, hence, assume
-homogeneity, \citet{Fisher+Lee:1992} also provide a heteroscedastic version by combining two  
+homoscedasticity, \citet{Fisher+Lee:1992} also developed a heteroscedastic version by combining two  
 individual GLMs, each for one of the parameters $\mu$ and $\kappa$. 
@@
 employing the von Mises distribution. The resulting tree-based models provide a very intuitive and data-driven 
-alternative to commonly used circular regression models for circular data.
+alternative to commonly used regression models for circular data.
 
@@
 
-\subsection{Circular forests}
+\subsection{Circular regression forests}
 \label{sec:circforest} 

### 2019-11-30 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor stuff in draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2906 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6b87a7f1b1391b1c967a1f2dc8478066691c1840

@@
 distribution parameters are considered and a fully specified distribution is
-provided for each node.  The proposed circular regression trees are easy to
+provided for each terminal node.  The proposed circular regression trees are easy to
 interpret, can detect non-additive effects, and select covariates and their
@@
 In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
-present a Bayesian alternative of a homoscedastic GLM for circular data.
+have presented a Bayesian alternative of a homoscedastic GLM for circular data.
 However, apart from potential difficulties in the optimization procedure of circular GLMs, 
-additive effects and their interpretation are not intuitive for circular responses
+additive effects and their interpretation are often not intuitive for circular responses
 due to the highly non-linear link function and the representation of smooth transitions 

### 2019-12-01 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added appendix on tree-building algorithm (still needs to be revised)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2907 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 440b0d7f12318fbef4291da3d4e45f88d6e00225

@@
 another stopping criterion is met (e.g., subgroup size or tree depth).
-A more detailed description of the applied tree-building algorithm,
-particularly of CTree, can be found in Appendix~\ref{app:tree}.
+A more detailed description of the applied tree-building algorithm
+can be found in Appendix~\ref{app:algorithm}.
 
@@
 %\newpage
-%\section*{Appendix}
-%\begin{appendix}
-%\end{appendix}
-
-%\newpage
 \bibliography{ref.bib}
@@
 \begin{appendix}
-\section{Tree-building algorithms}
-\label{app:tree}
-CTree, MOB
+
+\section{Tree algorithm}
+\label{app:algorithm}
+This section provides a more detailed overview on the permutation-test-based CTree algorithm 
+\citep{Hothorn+Hornik+Zeileis:2006}, 
+specifically for circular data as applied to build the circular regression trees and forests 
+in the presented case study. 
+Alternatively, also the MOB algorithm, which is based on M-fluctuation tests, can be employed 
+\citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.
+
+\subsection*{CTree}
+In the following the testing and splitting strategy is described for the root node of the tree
+which contains the entire learning sample.
+%consisting of circular observations denoted by $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
+For a complete tree model, the same procedure is applied to all resulting child nodes with the 
+corresponding subsamples.
+
+First, employing the von Mises distribution, a distributional model $\text{vM}(Y; \mu,\kappa)$ 
+is fitted to the learning sample of circular observations $\{y_i\}_{i = 1,\ldots,n}$ as explained 
+in Section~\ref{sec:prob_circ}.
+In order to obtain a goodness-of-fit measurement of $\text{vM}(Y; \mu,\kappa)$ for each parameter 
+and each observation the score function $s(\mu, \kappa, Y)$ is evaluated at the estimated location 
+and concentration parameter $\hat{\mu}$ and $\hat{\kappa}$.
+In a next step, a statistical test is employed to detect dependencies between the calculated score 
+values
+\begin{equation}
+s(\hat{\mu}, \hat{\kappa}, y) = 
+\begin{pmatrix} 
+s(\hat{\mu}, \hat{\kappa}, y_1)_1 & s(\hat{\mu}, \hat{\kappa}, y_1)_2\\
+\vdots & \vdots\\
+s(\hat{\mu}, \hat{\kappa}, y_n)_1 & s(\hat{\mu}, \hat{\kappa}, y_n)_2
+\end{pmatrix}
+\end{equation}
+and each possible split variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
+In particular, a permutation test is applied to asses the following hypotheses:
+\begin{align}
+H_0^l:  s(\hat{\mu}, \hat{\kappa}, Y) \qquad \bot \qquad Z_l.
+\end{align}
+%
+The permutation test is based on the multivariate linear statistic
+\begin{equation}
+T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\hat{\mu}, \hat{\kappa}, Y_i)\right)
+\end{equation}
+where $s(\hat{\mu}, \hat{\kappa}, Y_i) \in \mathbb{R}^{1\times 2}$. The form 
+of the transformation function $v_l$ depends on the type of the split 
+variable $Z_l$. If $Z_l$ is numeric then $v_l$ is simply the identity 
+function $v_l(Z_{li}) = Z_{li}$ and therefore $T_l \in \mathbb{R}^2$ as 
+the ``vec'' operator converts the $1 \times 2$ matrix into a $2$ column vector. 
+If $Z_l$ is a categorical variable with $H$ categories then 
+$v_l(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$ 
+such that $v_l$ is a $H$-dimensional unit vector where the element corresponding to 
+the value of $Z_{li}$ is $1$. In this case the statistic $T_l \in \mathbb{R}^{H \cdot 2}$
+as the ``vec'' operator converts the $H \times 2$ matrix into a $H \cdot 2$ column vector 
+by column-wise combination. Observations with missing values are excluded from the sums.
+
+Using the conditional expectation $\mu_l$ and the covariance 
+$\Sigma_l$ of $T_l$ as derived by \cite{Strasser+Weber:1999} 
+the test statistic can be standardized. 
+The observed multivariate linear statistic $t_l$ which is either a
+$2$- or $2 \cdot H$-dimensional vector, depending on the scale of $Z_l$,
+is mapped onto the real line by a univariate test statistic $c$.
+In the application of this paper a quadratic form is chosen, such that
+\begin{equation}
+c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
+\end{equation}
+where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$. Alternatively,
+the maximum of the absolute values of the standardized linear statistic
+can be considered ($c_{\text{max}}$).
+
+\cite{Strasser+Weber:1999} showed that the asymptotic conditional 
+distribution of the linear statistic $t_l$ is a multivariate normal with parameters
+$\mu_l$ and $\Sigma_l$. Hence, the asymptotic conditional distribution of
+$c(t_{l},\mu_{l},\Sigma_{l})$ 
+is either normal
+(for $c_{\text{max}}$) 
+or 
+$\chi^2$ (for $c_{\text{quad}}$).
+
+The smaller the $p$-value corresponding to the standardized test 
+statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
+from the assumption of independence between the scores and the split 
+variable $Z_l$.
+%
+After Bonferroni-adjusting the $p$-values it has to be assessed whether
+any of the resulting $p$-values is beneath the selected significance level. 
+If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
+is chosen as splitting variable. Otherwise no further split is made 
+in this node as the stopping criterion of no $p$-values being below the 
+significance level is fulfilled. This type of early-stopping in building
+a tree is sometimes also referred to as ``pre-pruning''. For random forests
+pre-pruning is often switched off by setting the significance level to $1$.
+
+The breakpoint that leads to the highest discrepancy between score functions 
+in the two resulting subgroups is selected as split point. 
+This is measured by the linear statistic
+\begin{equation}
+T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\hat{\mu}, \hat{\kappa}, Y_i)
+\end{equation}
+for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
+two new subgroups, without any particular ordering, that are defined by 
+splitting in split point $r$ of variable $Z_{l^{\ast}}$.
+The split point is then chosen as follows:
+\begin{equation}
+r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
+\end{equation}
+One repeats the testing and splitting procedure in each of the resulting 
+subgroups until some stopping criterion is reached. This criterion 
+can for example be a minimal number of observations in a node or a 
+minimal $p$-value for the statistical tests. In that way pre-pruning
+is applied in order to find right-sized trees and hence 
+avoid overfitting.
+
+%This permutation-test-based tree algorithm is presented in 
+%\cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
+%A different framework to build a likelihood-based tree is provided by the MOB algorithm 
+%which is based on M-fluctuation tests (\citealp{Zeileis+Hothorn+Hornik:2008}).
+
+%\subsection{Teststatistic: CTree}\label{app:teststat_ctree}
+%To measure the association of a response $Y$ and each possible split variable $Z_j$, $j=1,\ldots,J$,
+%the CTree algorithm applies a linear test statistic $T_j$ which is excerpted from Section~3.1.
+%(``Variable selection and stopping criteria'') of the original paper 
+%\citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
+%$$
+%T_j(Y, Z_j, w) = \text{vec}\left( \sum_{i=1}^N w_i g_j(Z_{ji})h(Y_i)^{\top}\right)
+%$$
+%where $w$ are optional weights, $g_j: \mathcal{Z}_j \rightarrow \mathbb{R}^{P}$ is a nonrandom 
+%transformation of the split variable $Z_j$ and the influence function 
+%$h: \mathcal{Y} \times \mathcal{Y}^N \rightarrow \mathbb{R}^Q$ depends on the response 
+%$Y = (Y_1, \ldots, Y_N)$ in a permutation symmetric way and is set to 
+%$h(Y_i) = s(Y_i, X_i, \hat{\beta})$ for a score-based approach. 
+%Moreover, by applying the ``vec'' operator the resulting
+%$P~\times~Q$ matrix is converted into a $PQ$ column vector.
+%Following \cite{Strasser+Weber:1999}
+%the conditional expectation $\mu_j$ and covariance $\Sigma_j$ of a linear test statistic
+%$T_j$ can be calculated and used to standardize an observed linear test statistic $t_j$ within a 
+%function $c$ mapping into the real line.
+%For example, the maximum of the absolute values of the standardized linear statistic
+%$$
+%c_{\max}(t_j,\mu_j,\Sigma_j) = \max \left| \frac{(t_j-\mu_j)}{\text{diag}(\Sigma_j)} \right|
+%$$
+%or a quadratic form
+%$$
+%c_{\text{quad}}(t_j, \mu_j, \Sigma_j) = (t_j - \mu_j)\Sigma_j^{+} (t_j - \mu_j)^{\top}
+%$$
+%can be considered where $\Sigma_j^{+}$ is the Moore-Penrose inverse of $\Sigma_j$.
+%Since the asymptotic conditional distribution of a linear test statistic $T_j$ is a 
+%multivariate normal with parameters $\mu_j$ and $\Sigma_j$ \citep{Strasser+Weber:1999}, 
+%the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
+%$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
+%the corresponding $p$-values can be calculated easily.
+
 \end{appendix}
+
 \end{document}

### 2019-12-02 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revised appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2908 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1b62dcc7e55225c10ad74f65aab35bf100fbcb8d

--- 2019-11-30 pkg/circtree/inst/wind_paper/draft/draft.Rnw
@@
 \begin{appendix}
-\section{Tree-building algorithms}
-\label{app:tree}
-CTree, MOB
+
+\section{Tree algorithm}
+\label{app:algorithm}
+This section provides a more detailed overview on the permutation-test-based CTree algorithm 
+\citep{Hothorn+Hornik+Zeileis:2006}, 
+specifically for circular data as applied to build the circular regression trees and forests 
+in the presented case study. 
+Alternatively, also the MOB algorithm, which is based on M-fluctuation tests, can be employed 
+\citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.
+
+\subsection*{CTree}
+In the following, the testing and splitting strategy is described for the root node of the tree
+which contains the entire learning sample.
+%consisting of circular observations denoted by $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
+For a complete tree model, the same procedure is applied iteratively to all resulting child nodes 
+with the corresponding subsamples.
+
+First, employing the von Mises distribution, a distributional model $\text{vM}(Y; \mu,\kappa)$ 
+is fitted to the learning sample of circular observations $\{y_i\}_{i = 1,\ldots,n}$ as explained 
+in Section~\ref{sec:prob_circ}.
+In a next step, a goodness-of-fit measurement %of $\text{vM}(Y; \mu,\kappa)$ 
+is obtained for each parameter and each observation by evaluating the score 
+function $s(\mu, \kappa, Y)$ at the estimated location and concentration 
+parameter $\hat{\mu}$ and $\hat{\kappa}$.
+To detect dependencies between the resulting score values
+\begin{equation}
+s(\hat{\mu}, \hat{\kappa}, y) = 
+\begin{pmatrix} 
+s(\hat{\mu}, \hat{\kappa}, y_1)_1 & s(\hat{\mu}, \hat{\kappa}, y_1)_2\\
+\vdots & \vdots\\
+s(\hat{\mu}, \hat{\kappa}, y_n)_1 & s(\hat{\mu}, \hat{\kappa}, y_n)_2
+\end{pmatrix}
+\end{equation}
+and each possible split variable $Z_l \in \{Z_1, \ldots, Z_m\}$ a permutation test is applied. 
+In particular, the hypotheses
+\begin{align}
+H_0^l:  s(\hat{\mu}, \hat{\kappa}, Y) \qquad \bot \qquad Z_l
+\end{align}
+is assessed by employing the multivariate linear statistic
+\begin{equation}
+T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\hat{\mu}, \hat{\kappa}, Y_i)\right)
+\end{equation}
+with $s(\hat{\mu}, \hat{\kappa}, Y_i) \in \mathbb{R}^{1\times 2}$. 
+For a numeric split variable $Z_l$ the transformation function $v_l$
+is simply the identity function $v_l(Z_{li}) = Z_{li}$ such that $T_l \in \mathbb{R}^2$ 
+as the ``vec'' operator converts the matrix of dimension $1 \times 2$ into a $2$ column vector.
+If $Z_l$ is a categorical variable with $H$ categories then 
+$v_l(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$, hence, $v_l$ returns a unit vector
+of dimension $H$ where the entry $1$ indicates the category of $Z_{li}$. 
+In this case the ``vec'' operator converts the $H \times 2$ matrix into a $H \cdot 2$ column vector 
+by column-wise combination such that $T_l \in \mathbb{R}^{H \cdot 2}$. 
+If there are any observations with missing values these are not included in the calculation of $T_l$.
+
+To map the observed multivariate linear statistic $t_l$ onto the real line a univariate test statistic
+$c$ is employed, for example in a quadratic form
+\begin{equation}
+c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
+\end{equation}
+where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance of $T_l$,
+as derived by \cite{Strasser+Weber:1999} and used to standardize $t_l$, and $\Sigma_l^+$ 
+is the Moore-Penrose inverse of $\Sigma_l$.
+As an alternative, also a maximum form ($c_{\text{max}}$) can be considered such that
+the maximum of the absolute values of the standardized linear statistic is returned.
+
+The asymptotic conditional distribution of $c(t_{l},\mu_{l},\Sigma_{l})$ 
+is either normal (for $c_{\text{max}}$) or $\chi^2$ (for $c_{\text{quad}}$) owing to 
+the asymptotic conditional distribution of the linear statistic $t_l$ being a multivariate 
+normal with parameters $\mu_l$ and $\Sigma_l$ \citep{Strasser+Weber:1999}.
+With this knowledge at hand, the corresponding $p$-values can be calculated and used
+to select the best splitting variable. A small $p$-value corresponding to 
+$c(t_{l},\mu_{l},\Sigma_{l})$ indicates a strong discrepancy from the assumption of independence 
+between the scores and the split variable $Z_l$.
+Therefore, if any of the Bonferroni-adjusted $p$-values is beneath the selected significance level,
+the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value is selected as split variable, otherwise 
+no split is performed. This early stopping induced by the significance level is referred to as 
+``pre-pruning'' which is often avoided for forest models by setting the significance level to $1$.
+
+To select the best split point within the already chosen split variable, again, a linear test statistic
+is employed. In particular, for a breakpoint $r$ of the variable $Z_{l^{\ast}}$ leading to two subgroups 
+$\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ the discrepancy between score functions in the subgroups
+is measured by evaluating
+\begin{equation}
+T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\hat{\mu}, \hat{\kappa}, Y_i)
+\end{equation}
+for $q \in \{1,2\}$.
+The breakpoint that leads to the highest discrepancy is then selected as split point as defined by
+\begin{equation}
+r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
+\end{equation}
+
+Subsequently, the same testing and splitting procedure is repeated in each of the resulting 
+subgroups until some stopping criterion is reached. Next to the already mentioned 
+significance-level-based stopping criterion, i.e., a minimal $p$-value for the statistical tests,
+also a maximal tree depth or a minimal number of observations in a node can be employed as stopping
+criteria.
+
+%This permutation-test-based tree algorithm is presented in 
+%\cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
+%A different framework to build a likelihood-based tree is provided by the MOB algorithm 
+%which is based on M-fluctuation tests (\citealp{Zeileis+Hothorn+Hornik:2008}).
+
+%\subsection{Teststatistic: CTree}\label{app:teststat_ctree}
+%To measure the association of a response $Y$ and each possible split variable $Z_j$, $j=1,\ldots,J$,
+%the CTree algorithm applies a linear test statistic $T_j$ which is excerpted from Section~3.1.
+%(``Variable selection and stopping criteria'') of the original paper 
+%\citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
+%$$
+%T_j(Y, Z_j, w) = \text{vec}\left( \sum_{i=1}^N w_i g_j(Z_{ji})h(Y_i)^{\top}\right)
+%$$
+%where $w$ are optional weights, $g_j: \mathcal{Z}_j \rightarrow \mathbb{R}^{P}$ is a nonrandom 
+%transformation of the split variable $Z_j$ and the influence function 
+%$h: \mathcal{Y} \times \mathcal{Y}^N \rightarrow \mathbb{R}^Q$ depends on the response 
+%$Y = (Y_1, \ldots, Y_N)$ in a permutation symmetric way and is set to 
+%$h(Y_i) = s(Y_i, X_i, \hat{\beta})$ for a score-based approach. 
+%Moreover, by applying the ``vec'' operator the resulting
+%$P~\times~Q$ matrix is converted into a $PQ$ column vector.
+%Following \cite{Strasser+Weber:1999}
+%the conditional expectation $\mu_j$ and covariance $\Sigma_j$ of a linear test statistic
+%$T_j$ can be calculated and used to standardize an observed linear test statistic $t_j$ within a 
+%function $c$ mapping into the real line.
+%For example, the maximum of the absolute values of the standardized linear statistic
+%$$
+%c_{\max}(t_j,\mu_j,\Sigma_j) = \max \left| \frac{(t_j-\mu_j)}{\text{diag}(\Sigma_j)} \right|
+%$$
+%or a quadratic form
+%$$
+%c_{\text{quad}}(t_j, \mu_j, \Sigma_j) = (t_j - \mu_j)\Sigma_j^{+} (t_j - \mu_j)^{\top}
+%$$
+%can be considered where $\Sigma_j^{+}$ is the Moore-Penrose inverse of $\Sigma_j$.
+%Since the asymptotic conditional distribution of a linear test statistic $T_j$ is a 
+%multivariate normal with parameters $\mu_j$ and $\Sigma_j$ \citep{Strasser+Weber:1999}, 
+%the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
+%$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
+%the corresponding $p$-values can be calculated easily.
+
 \end{appendix}
+
 \end{document}

### 2019-12-03 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision based on Reto's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2909 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0cc0d39aebda2df31d14542bffdc40f3d968a4f8

@@
 trees a more stable and regularized circular regression forest can be built
-which allows for modeling smooth effects as well.
+which also allows for modeling smooth effects.
 
@@
 are used as a benchmark in order to assess the predictive performances of
-circular regression trees and forests.
+the novel methodology of circular regression trees and forests.
 %Establishing a proper regression model for circular data such as animal movement
@@
 
-In many approaches to model circular data it is assumed that the circular
+Many approaches to model circular data assume that the circular
 variable of interest follows a circular distribution, in particular the von
@@
 presented by \cite{Gould:1969} where the circular mean, i.e., the location
-parameter of the von Mises distribution is predicted by a linear combination of
+parameter of the von Mises distribution, is predicted by a linear combination of
 covariates. 
@@
 complexity encountered in optimizing the corresponding log-likelihood function which is not globally concave. 
-Therefore, highly informative starting values are crucial for these circular GLMs to converge at all 
+Therefore, highly informative starting values are crucial for these circular GLMs to converge 
 \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
@@
 additive effects and their interpretation are often not intuitive for circular responses
-due to the highly non-linear link function and the representation of smooth transitions 
-on the unit circle is not straightforward. For example, the same rotation can be obtained in either 
+due to the highly non-linear link function and the not-straightforward representation of smooth 
+transitions on the unit circle. For example, the same rotation can be obtained in either 
 positive or negative direction on the circle leading to an ambiguous interpretation.
@@
 simple intercept-only model for each resulting subset.
-Therefore, no covariates and, thus, no link function must be employed in the estimation
+Therefore, no covariates and, thus, no link functions must be employed in the estimation
 of the group-specific distribution parameters which also simplifies the optimization process.
@@
 splits in all distribution parameters and provide a fully specified circular response distribution 
-for each terminal node offering a wide range of inference.
-In addition, the employed tree structure captures non-additive effects while forests allow for 
-modeling smooth changes as well.
+for each terminal node offering a wide range of statistical inference.
+In addition, the employed tree structure allows to capture non-additive effects while forests 
+enable the modeling of smooth changes.
 Furthermore, covariates and their possible interactions do not need to be specified in advance as 
@@
 homogeneity.
-While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not 
-only for linear but also for a circular covariate, the tree only considers effects in the mean 
+While this approach and further extensions \citep{Larraondo+Inza+Lozano:2018} do not only
+allow for linear, but also for circular covariates, the tree solely considers effects in the mean 
 direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
@@
 of previous circular trees by considering splits in all distribution parameters and 
-yielding a full probabilistic model. In addition, as a natural further development, ensembles or forests 
+providing a full probabilistic model. In addition, as a natural extension, ensembles or forests 
 of circular regression trees are presented in this study. 
@@
 To provide a first impression of the presented methodology a circular regression 
-tree is employed to probabilistic wind direction forecasting. Wind direction is a 
+tree is employed for probabilistic wind direction forecasting. Wind direction is a 
 classical circular quantity and accurate forecasts are of great importance for
-decision-making processes especially in air traffic management.
+decision-making processes, e.g., in air traffic management as considered in this study.
 Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
-airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
-Topography channels wind along the west-east valley axis or along a tributary
+airport of Innsbruck which lies at the bottom of a narro valley within the European Alps.
+Topography channels the wind along the west-east valley axis or along a tributary
 valley intersecting from the south. Hence, pressure gradients to which valley
 wind regimes react are considered as covariates along with other meteorological
-measurements at the airport (lagged by one hour), such as wind direction and
-wind speed at the airport.
+measurements (lagged by one hour) and their derivatives, such as 
+wind direction and wind speed at the airport or spatial and temporal differences.
 %%
@@
   empirical histogram (gray) and fitted density (red line) are depicted along
-  with the estimated location parameter (red hand). The covariates employed are
-  wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
+  with the estimated location parameter (red hand). The covariates used for splitting
+  are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
   ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
-  one hour.}
+  one hour.
+  Note that in the meteorological context wind direction is defined on the 
+  scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
+}
   \label{fig:tree_ibk} 
@@
 winds blowing from the valley mouth towards the upper valley (from east to
-west, nodes~4 and 5). (2) Downslope winds blowing across the Alpine crest
+west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest
 along the intersecting valley towards Innsbruck (from south-east to north-west,
-node~8). (3) Down-valley winds blowing in the direction of the valley mouth
+node~8); (3) Down-valley winds blowing in the direction of the valley mouth
 (from west to east, nodes 10, 12 and 13). Node~7 captures observations with
-rather low wind speeds that cannot be distinguished clearly into wind regimes
-and consequently are associated with a very low estimated concentration $\hat
-\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
-mostly responsible for distinguishing the broad wind regimes listed above while
-the pressure gradients and wind speed separate between subgroups with high vs.\
-low precision. Note that in the meteorological context wind direction 
-is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
-
+rather low wind speeds that cannot be clearly distinguished into specific wind 
+regimes and are consequently associated with a very low estimated concentration 
+parameter~$\kappa$, i.e. a high estimated variance. 
+In terms of covariates, the lagged wind direction (``persistence'') is mostly 
+responsible for distinguishing the broad range of wind regimes listed above while
+the pressure gradients and wind speed separate the data into subgroups with high 
+vs.\ low precision. 
 A more extensive case study of circular regression trees and forests
@@
 Innsbruck and Vienna will be presented in this paper. Moreover, their
-performances are benchmarked to alternative commonly used approaches. The
-remainder of the paper is structured as follows: The theory on probabilistic
+performances are benchmarked to alternative commonly used approaches.
+
+The remainder of the paper is structured as follows: The theory on probabilistic
 circular modeling introducing the von Mises distribution will be discussed in
@@
 circular distribution which is also known as ``the circular normal
-distribution'' and commonly applied for probabilistic modeling of circular
-data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
+distribution'' and is a common choice for probabilistic modeling of circular
+data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
 parameter $\kappa > 0$ the density of the von Mises distribution for an
-observation $y \in (-\pi, \pi]$ is given by:
+observation $y \in [0, 2\pi)$ is given by:
 %%
@@
 
-The corresponding log-likelihood function is defined by
+The corresponding log-likelihood function is defined as
 %%
@@
 
-Once an appropriate distribution family is selected fitting a probabilistic
+Once an appropriate distribution family is selected, fitting a probabilistic
 model corresponds to estimating the distribution parameters. Thus, when
@@
 yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
-a fully specified distributional model is fit to the learning data.
+a fully specified distributional model is fitted to the learning data.
 
@@
     maximum likelihood estimation employing the von Mises distribution. Left
-    panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. Right
+    panel: linearized illustration restricted to the interval $[0, 2\pi)$. Right
     panel: circular illustration. In both panels the empirical histogram (gray)
@@
 An alternative to additive models, which can deal or even 
-avoid the mentioned difficulties in circular regression modeling, is to employ tree-based models. 
-For the particular case of circular data, the methodology of circular regression trees and forests is
-presented in the following section.
+avoid the mentioned difficulties in circular regression modeling, are 
+tree-based models. For the particular case of circular data, the methodology 
+of circular regression trees and forests is presented in the following section.
 
@@
 distribution allows for an application of distributional trees to circular data
-following the tree building algorithm as introduced in Section~2.2. of
+following the tree building algorithm as introduced in Section~2.2 of
 \cite{Schlosser+Hothorn+Stauffer:2019}.
@@
 how and where to split the data is based on model scores which are obtained by
-evaluating the score function $s$ for the individual observations and parameter
+evaluating the score function~$s$ for the individual observations and parameter
 estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
-two distribution parameters and a data set of $n$ observations, this yields an
-$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
+two distribution parameters ($\mu$ and $\kappa$) and a data set of $n$ observations, 
+this yields an $n \times 2$ matrix that can be employed as a discrepancy measure, capturing
 how well each given observation conforms with the estimated location
@@
 probabilistic predictions for a possibly new set of observed covariates $\bm{z}
-= (z_1, \ldots, z_m)$. Starting at the root node the tree structure leads the
+= (z_1, \ldots, z_m)$. Starting at the root node, the tree structure leads the
 observation to a terminal node, depending on the values of the covariates,
@@
 
-While tree models can capture non-additive effects their structure and the
+While tree models can capture non-additive effects, their structure and the
 consequential strict separation of data into subgroups hinders an adequate
@@
 \label{sec:circforest} 
-A natural extension of circular regression trees are ensembles or forests
+A natural extension of (circular) regression trees are ensembles or forests
 that can improve forecasts by regularizing and stabilizing the model. Random
@@
 which $\bm{z}$ is assigned to the same terminal node as each learning
-observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
+observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees, the
 weights are calculated via
@@
 \label{sec:wind}
-Accurate forecasts of wind directions are of great importance for risk
+As motivated in Section~\ref{sec:introduction},
+accurate forecasts of wind directions are of great importance for risk
 management in various fields such as agriculture, energy production or air
@@
 different site characteristics: Innsbruck with a rather unusual location for an
-airport, located at the bottom of a deep Alpine valley, and Vienna located
+airport, located at the bottom of a narrow Alpine valley, and Vienna located
 within flat terrain in the East of Austria. In addition to the newly introduced
 circular regression trees and forests, the case study employs various
-alternative approaches to forecast wind direction as a benchmark. The study is
+alternative approaches to forecast wind direction as benchmarks. The study is
 based on $1$\,h and $3$\,h forecasts employing lagged observations in the
-vicinity of the airports as predictor variables.
+vicinity of the airports as possible predictor variables.
 
@@
 The circular response variable considered in this case study is a $10$\,min
-mean of wind direction measurements at the airports Innsbruck and Vienna on an
-hourly temporal resolution. As predictor variables, we use $1$-hourly resolved
-$10$\,min mean observations of various meteorological quantities such as wind
-direction, wind speed, temperature, air pressure and humidity, all lagged by
+mean of wind direction measurements at the airport for Innsbruck and Vienna on an
+hourly temporal resolution. As predictor variables, we use temporal information
+and $1$-hourly resolved $10$\,min mean observations of various meteorological quantities such as 
+wind direction, wind speed, temperature, air pressure and humidity, all lagged by
 one or three hours according to the respective forecasting step. The
@@
 
-The data used in this study consists of five total years from January 2015 to
+The data used in this study consists of five subsequent years from January 2015 to
 December 2018. After first eliminating predictor variables with more than 5\%
 missing values and then time points with missing observations, the data set
-consists of 41979 time points and 260 covariates for Innsbruck, and of 38985
+consists of 41\,979 time points and 260 covariates for Innsbruck, and of 38\,985
 times points and 494 covariates for Vienna, respectively.
@@
       Data components               & Description \\
+      \hline
+      Temporal information:         & Daytime, day of the year \\
       \hline
@@
 study for probabilistic forecasting of wind directions. Two of them are based
-on existing approaches in the meteorological field, while the third is a
+on existing approaches used in the meteorological field, while the third is a
 state-of-the-art model to forecast circular response variables. The next
@@
     quantity itself, either on a daily or yearly basis. Wind direction at a single
-    location is often quite unstable and shows only a daily or annual seasonality
+    location is often quite unstable and only shows a daily cycle or an annual seasonality
     in case of predominant local wind systems like the ones at Innsbruck introduced
-    for the analysis of Figure~\ref{fig:tree_ibk}. Hence, the climatological model
+    in the analysis of Figure~\ref{fig:tree_ibk}. Hence, the climatological model
     in this study provides a very simple implementation of a baseline model, which
@@
     \newline
-    For simplicity, rather to fit one single climatological model, we perform the
-    model fitting separately rolling over all time points. For each time point, the
-    training data set consists of the same daytime as the time of interest within
-    31~days centered around the same calender day over all years except the year of
-    interest. To gain full probabilistic climatological forecasts, we employ the
+    For simplicity, rather to fit one single climatological model, a single model is estimated 
+    rolling over all observation time points, 
+    using the $31$~days centered around the day of interest for all available years in the training data.
+    %For simplicity, rather to fit one single climatological model, we perform the
+    %model fitting separately rolling over all time points. For each time point, the
+    %training data set consists of the same daytime as the time of interest within
+    %31~days centered around the same calender day over all years except the year of
+    %interest. 
+    To gain full probabilistic climatological forecasts, we employ the
     steps described in Section~\ref{sec:prob_circ} and fit the distribution

### 2019-12-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision based on Reto's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2910 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f7af6bd8b52fdf72ff99966ca9556124b490479f

@@
 where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
-$0$ \citep[see, e.g.,][for a more detailed
-overview]{Jammalamadaka+Sengupta:2001}.
+$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} or \citealt{Ley+Verdebout:2017} for a more detailed overview).
 
@@
       Temporal information:         & Daytime, day of the year \\
-      \hline
+      \noalign{\vskip 1mm}   
       Meteorological variables:     & Wind direction, wind (gust) speed, \\
@@
     \newline
-    For simplicity, rather to fit one single climatological model, a single model is estimated 
+    For simplicity, rather to fit one single climatological model, separate models are estimated 
     rolling over all observation time points, 
-    using the $31$~days centered around the day of interest for all available years in the training data.
+    using the $31$~days centered around the day of interest over all available years in the training data.
     %For simplicity, rather to fit one single climatological model, we perform the
@@
   \item \emph{Persistency model:}
-    Persistence describes the previous value of a single weather quantity in a time
+    The persistence describes the previous value of a single weather quantity in a time
     series. Hence, it is another objective measure in the verification of weather
@@
     atmospheric conditions, persistence can be a very good estimate, but as soon as
-    any major change takes place in the response variable during the forecast
+    the regime changes takes place in the response variable during the forecast
     period, forecasts based on persistence may also fail remarkably. Therefore, it
     heavily depends on the variability of the variable of interest and the length
-    of the forecasting step whether persistence can provide overall reliable
-    predictions.
+    of the forecasting step whether the persistence can provide overall reliable
+    information.
     \newline
@@
     distribution parameters of the von Mises distribution conditional on lagged
-    response values according to the description in Section~\ref{sec:prob_circ}. We
-    fit one model for every time point in the validation data set employing the
+    response values according to the description in Section~\ref{sec:prob_circ}. 
+    We fit one model for every hour throughout 
+    the validation data set employing the
     previous six lagged response values as training data. In order to allow for a
-    stronger influence of observations closer to the time of interest exponential
+    stronger influence of observations closer to the time of interest, exponential
     smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
-    prediction the current observation and all previous five observations together
-    have an equal influence rate of 50 percent. Observations with longer time lags
-    have exponential weights below $0.01$ and are therefore omitted from the
+    prediction an equal influence rate of 50 percent is assigned both to the current 
+    observation and to the previous five observations together. Observations with longer 
+    time lags have exponential weights below $0.01$ and are therefore omitted from the
     training data.
@@
     interval of length $2\pi$. As the implementation cannot handle circular covariates,
-    we use the components of the lagged 2-dimensional wind vector $vec$ and the
+    we use the components of the lagged 2-dimensional wind vector $\bm{vec} = (vec_1, vec_2)^{\top}$ and the
     lagged wind speed $spd$ as predictor variables. The model formula for the
@@
     tree building is performed by the newly developed R-package \pkg{circtree}
-    employing the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a
+    employing the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} using a
     minimal number of 2000 observations in each terminal node serving as a stopping
@@
   \item \emph{Circular regression forest:} Following the description in
-    Section~\ref{sec:circforest}, a circular forest is constructed based on 100
-    individual trees employing the R-package \pkg{circtree}. Each of these trees is
-    again built by the CTree algorithm on a subsample containing a fraction of 30
-    percent of the original learning data. As it can be expected that the lagged
+    Section~\ref{sec:circforest}, the circular forest used in this study is constructed 
+    based on $100$~individual trees employing the R-package \pkg{circtree}. Each of these 
+    trees is again built by the CTree algorithm on a subsample containing a fraction of 
+    $30$~percent of the original learning data. As it can be expected that the lagged
     response variable is one of the most informative predictor variables for wind
@@
 
-To compare the predictive performances of all proposed models a circular
+To compare the predictive performance of all proposed models, a circular
 analogue of the continuous ranked probability score (CRPS) as introduced by
@@
 
-In addition to the raw CRPS, also CRPS-based skill scores are computed to
+In addition to the raw CRPS, corresponding skill scores are computed to
 assess differences in the improvement of the various statistical models over
@@
 All scores presented in the next section are computed out-of-sample based on
-five years of observations. For the persistency model only time points prior to
+five years of data. For the persistency model only time points prior to
 the time of interest are used and the validation is performed rolling over all
@@
 per day over five years, some kind of temporal aggregation is needed to ensure
-a correct visual comparison of the individual methods. Performed analyses have
+a correct visual comparison of the individual methods. The analyses performed have
 shown that for the employed models the variability of the predictive
@@
 the circular forest overall provides the best predictive performance, followed
-by the circular regression tree and, except for the $3$\,h forecast at
-Innsbruck, the persistency model. In comparison to the circular regression tree and
+by the circular regression tree and the persistency model; except for the $3$\,h~forecast 
+at Innsbruck, where the persistency model is outperformed by all others. 
+In comparison to the circular regression tree and
 forest, for both stations and forecast steps, the climatological model and the
@@
   \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
-  \caption{CRPS skill scores for wind direction forecasts based on the full
+  \caption{CRPS skill scores of wind direction forecasts based on the full
     predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
@@
 absolute level of the model performances and on their respective predictive
-performance variances. At Innsbruck, due to the surrounding mountains a
+performance variances. At Innsbruck, due to the surrounding mountains only a
 limited number of possible wind directions exists, namely the three wind
@@
 remains rather constant in one of these possible states, but once a change
-takes place it is mostly a major wind direction shift. Due to the few wind
+takes place it is mostly a major wind direction shift, 
+e.g., from up-valley to down-valley. Due to the few wind
 regimes the rather inflexible climatological and linear models score relatively
@@
 (Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
-potential high prediction errors in case of a paradigm shift seem to lead to a
-higher variation in the predictive performance for all models in comparison to
+potential high prediction errors in case of a change of the wind regime seem to 
+lead to a higher variation in the predictive performance for all models in comparison to
 Vienna; this variation is especially high for the persistency model due to its
@@
 predictive performance of the climatological model and the linear model at both
-stations. However, for the persistency model, at both stations, higher scores
+stations. As expected, for the persistency model, at both stations, higher scores
 for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
-lower performance for longer lead times; this is probably due to the lower 
+lower performance for longer lead times; this is due to the lower 
 information content of minimum 3-hourly instead of 1-hourly lagged response 
@@
 scores with the climatological model as a reference are provided in
-Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in percent, and positive
+Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in percent, where positive
 values indicate an improvement in the predictive performance over the
@@
 \label{sec:summary}
-Circular regression trees for circular responses are established by
-coupling model-based recursive partitioning with the von Mises distribution.
-The resulting trees can capture non-linear changes, shifts, and potential
-interactions in covariates without prespecification of such effects. This is
-particularly useful for modeling a highly fluctuating response such as wind
-direction. \dots
+Circular regression trees and forests for circular responses are established by coupling 
+model-based recursive partitioning with the von Mises distribution.
+
+The resulting 
+trees can capture non-linear changes, shifts, and potential interactions in covariates 
+without prespecification of such effects. This is particularly useful for modeling a 
+highly fluctuating response such as wind direction. \dots
 

### 2019-12-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added first draft of summary
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2911 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8ffe9f07bc007fff7273a37fdb0da500017ae494

@@
 }
-\Keywords{Distributional Trees, Random Forests; Circular Data; von Mises Distribution}
+\Keywords{Distributional Trees; Random Forests; Circular Data; von Mises Distribution}
 
@@
 \label{sec:summary}
-Circular regression trees and forests for circular responses are established by coupling 
-model-based recursive partitioning with the von Mises distribution.
-
-The resulting 
-trees can capture non-linear changes, shifts, and potential interactions in covariates 
-without prespecification of such effects. This is particularly useful for modeling a 
-highly fluctuating response such as wind direction. \dots
-
-\begin{itemize}
-  \item Add summary/conclusion of case study;
-  \item Add summary/conclusion of circular regression trees and forests.
-  \item Outlook on possible extensions (use of NWP data, circular splits, ...).
-\end{itemize}
+Extending the toolbox for modeling circular data, circular regression trees and forests 
+are established by coupling model-based recursive partitioning with the von Mises distribution. 
+By separating the data into more homogeneous subgroups, possible difficulties in circular regression
+are bypassed as covariates are solely considered for splitting and 
+group-specific models are fitted as intercept-only models.
+%distributional models are fitted as intercept-only models in each node. 
+In addition, by specifying the von Mises distribution for each node and
+allowing for splits in both distribution parameters $\mu$ and $\kappa$, 
+fully probabilistic forecasts are provided.
+
+The performance of the novel circular trees and forests are assessed in a real case study for shortterm
+probabilistic wind direction forecasting at two airports with different site characteristics.
+As benchline models, probabilistic climatology and perstistency models, as well as a state-of-the-art
+circular GLM are evaluated based on proper scoring rules.
+In summary, the circular trees and forests have the highest predictive performance in this setting.
+For cases without changes in the wind regime, lagged response values provide already
+highly skillful estimates leading to good performance of the persistency model for shortterm wind direction forecasts.
+While in these cases the trees and forests also benefit from the highly informative lagged response, they can compensate for a lower information of this covariate by incorporating other quantities and possible interactions of these, contrary to the persistency model (see Figure~\ref{fig:boxplot_crpsskill}).
+Hence, the tree-based models provide reliable forecasts in all tested meterological settings.
+For operational use, a possible extension could be the incorporation of numerical weather predictions as (additional) covariates.
+While this probably only slightly improves the predictive skill for short leadtimes, it possibly extends the potential forecast range of the different methods.
+
+% TODO: circular splits (see first version of IWSM, 4 pages)
+
+To conlcude, the tree structure can capture non-linear changes, shifts, and potential interactions in 
+covariates without prespecification of such effects. As supported by the presented case study,
+this can be particularly useful for modeling a highly fluctuating response, such as typically 
+observed for wind direction, or/and in case of a large number of possible covariates.
+Moreover, the case study shows that building ensembles of circular trees can even improve the forecasting performance, as the resulting circular forests allow for modeling smooth effects and stabelize the model.
+
+%The comparison of the novel circular trees with simple climatological and persistency models, depicted
+%for probabilistic wind direction forecasts at two airports, shows that even in case of no obvious 
+%skillful covariates the trees have an high predictive performance.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\begin{itemize}
+%  \item Add summary/conclusion of case study;
+%  \item Add summary/conclusion of circular regression trees and forests.
+%  \item Outlook on possible extensions (use of NWP data, circular splits, ...).
+%\end{itemize}
 

### 2019-12-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Inlcude circular splits
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2912 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d790f4ba845d72736312c0db11d98237da9fd2e2

@@
 
-% TODO: circular splits (see first version of IWSM, 4 pages)
+Another possible improvement for obtaining more parsimonious circular trees is
+to consider splitting circular covariates into two circle segments by searching
+two split points simultaneously rather than sequentially at different depths.
+While this might slightly improve the predictive performance of circular trees,
+this should not affect the performance  of circular forest, as these still typically 
+consist of very large trees with many different splits.
 

### 2019-12-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2913 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0766b25b739f23cd7417072fe9b2f94fea1f79ee

@@
 While this probably only slightly improves the predictive skill for short leadtimes, it possibly extends the potential forecast range of the different methods.
-
 Another possible improvement for obtaining more parsimonious circular trees is
@@
 While this might slightly improve the predictive performance of circular trees,
-this should not affect the performance  of circular forest, as these still typically 
+this should not affect the performance  of circular forest, as these
 consist of very large trees with many different splits.

### 2019-12-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2914 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1d6ee5ea7dcf6004aa962e05092f79a10362c5af

@@
 Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
-airport of Innsbruck which lies at the bottom of a narro valley within the European Alps.
+airport of Innsbruck which lies at the bottom of a narrow valley within the European Alps.
 Topography channels the wind along the west-east valley axis or along a tributary
@@
 probabilistic wind direction forecasting at two airports with different site characteristics.
-As benchline models, probabilistic climatology and perstistency models, as well as a state-of-the-art
+As benchline models, probabilistic climatology and persistency models, as well as a state-of-the-art
 circular GLM are evaluated based on proper scoring rules.
@@
 For cases without changes in the wind regime, lagged response values provide already
-highly skillful estimates leading to good performance of the persistency model for shortterm wind direction forecasts.
+highly skillful estimates leading to a good performance of the persistency model as observed for shortterm wind direction forecasts in this study.
 While in these cases the trees and forests also benefit from the highly informative lagged response, they can compensate for a lower information of this covariate by incorporating other quantities and possible interactions of these, contrary to the persistency model (see Figure~\ref{fig:boxplot_crpsskill}).
-Hence, the tree-based models provide reliable forecasts in all tested meterological settings.
+Hence, the tree-based models provide reliable forecasts in all tested meteorological settings.
 For operational use, a possible extension could be the incorporation of numerical weather predictions as (additional) covariates.
@@
 While this might slightly improve the predictive performance of circular trees,
-this should not affect the performance  of circular forest, as these
+this should not affect the performance  of circular forests, as these
 consist of very large trees with many different splits.
 
-To conlcude, the tree structure can capture non-linear changes, shifts, and potential interactions in 
+To conclude, in general the tree structure can capture non-linear changes, shifts, and potential interactions in 
 covariates without prespecification of such effects. As supported by the presented case study,

### 2019-12-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Final proofreading before given to Thorsten
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2915 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c299d00ef83671e7ca69229fc32e04fbb01483ea

@@
 Mises distribution within the methodology of distributional trees and
-distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
+forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
 %In order to avoid these difficulties in regression modeling of circular data 
@@
 homogeneity.
-While this approach and further extensions \citep{Larraondo+Inza+Lozano:2018} do not only
+While this approach does not only
 allow for linear, but also for circular covariates, the tree solely considers effects in the mean 
@@
   \centering
-  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
+  \includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
   \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
@@
   empirical histogram (gray) and fitted density (red line) are depicted along
-  with the estimated location parameter (red hand). The covariates used for splitting
+  with the estimated location parameter (red hand). The covariates selected for splitting
   are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
@@
 regimes and are consequently associated with a very low estimated concentration 
-parameter~$\kappa$, i.e. a high estimated variance. 
+parameter~$\hat{\kappa}$, i.e., a high estimated variance. 
 In terms of covariates, the lagged wind direction (``persistence'') is mostly 
@@
 The remainder of the paper is structured as follows: The theory on probabilistic
-circular modeling introducing the von Mises distribution will be discussed in
+circular modeling introducing the von Mises distribution, and circular regression
+models will be discussed in
 Section~\ref{sec:prob_circ}. The methodology of circular regression trees
@@
 parameter $\kappa > 0$ the density of the von Mises distribution for an
-observation $y \in [0, 2\pi)$ is given by:
+observation $y \in [0, 2\pi)$ is given by
 %%
 \begin{equation}
-  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
+  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
 \end{equation}
@@
 where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
-$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} or \citealt{Ley+Verdebout:2017} for a more detailed overview).
+$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and \citealt{Ley+Verdebout:2017} for a more detailed overview).
 
@@
 right panel displays an appropriate representation of a circular quantity on
-the unit circle. Looking at the fitted density (red line) and estimated
+the unit circle. Looking at the empirical histogram along with 
+the fitted density (red line) and estimated
 location parameter (red hand) this example shows that reasonably well fitting
@@
 \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. 
-However, all circular additive models incorporate the previously discussed difficulties induced by
+Nevertheless, all circular additive models incorporate the previously discussed difficulties induced by
 the characteristics of the log-likelihood function and the strongly non-linear link function.
@@
 An alternative to additive models, which can deal or even 
-avoid the mentioned difficulties in circular regression modeling, are 
+avoid the mentioned difficulties in circular regression modeling are 
 tree-based models. For the particular case of circular data, the methodology 
@@
 intersecting valleys are used, whereas, for Vienna, measurements at the airport
-and within an area of about XXX\,km$^2$ are used. A topographical overview of
+and within an area of approximately 30\,km$^2$ are used. A topographical overview of
 the airports and their surrounding areas with the station sites employed in
@@
     rolling over all observation time points, 
-    using the $31$~days centered around the day of interest over all available years in the training data.
+    using the same daytime of $31$~days centered around the day of interest over all available years in the training data.
     %For simplicity, rather to fit one single climatological model, we perform the
@@
     atmospheric conditions, persistence can be a very good estimate, but as soon as
-    the regime changes takes place in the response variable during the forecast
+    the regime changes during the forecast
     period, forecasts based on persistence may also fail remarkably. Therefore, it
@@
     split points leads to high computational costs, the covariates are binned in a
-    maximum of $50$ classes (argument \code{nmax = c("yx" = Inf, "z" = 50)}).
+    maximum of $50$ classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}).
     Contrary to a single-tree model, forests usually consist of very large trees as
@@
 the circular forest overall provides the best predictive performance, followed
-by the circular regression tree and the persistency model; except for the $3$\,h~forecast 
-at Innsbruck, where the persistency model is outperformed by all others. 
+by the circular regression tree and the persistency model; except for the $3$\,h~forecasts
+at Innsbruck where the persistency model is outperformed by all others. 
 In comparison to the circular regression tree and
@@
   \caption{CRPS skill scores of wind direction forecasts based on the full
-    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
+    predictive von~Mises distribution for $+1$\,h and $+3$\,h forecasts at the
     airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
@@
 \citep{Hothorn+Hornik+Zeileis:2006}, 
-specifically for circular data as applied to build the circular regression trees and forests 
-in the presented case study. 
-Alternatively, also the MOB algorithm, which is based on M-fluctuation tests, can be employed 
-\citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.
+specifically for circular data as applied for building circular regression trees and forests 
+as presented in this case study. 
+An alternative tree-building framework is provided by the MOB algorithm, which is based on 
+M-fluctuation tests \citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.
 
@@
 
-To map the observed multivariate linear statistic $t_l$ onto the real line a univariate test statistic
-$c$ is employed, for example in a quadratic form
+To map the observed multivariate linear statistic $t_l$ onto the real line a univariate test statistic~$c$ is employed, for example in a quadratic form
 \begin{equation}

### 2019-12-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2917 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3c87c57b37758f4a74fded0955ca83045888ca7a

@@
 
+\newpage
 \begin{appendix}
@@
 specifically for circular data as applied for building circular regression trees and forests 
-as presented in this case study. 
+presented in this case study. 
 An alternative tree-building framework is provided by the MOB algorithm, which is based on 

### 2019-12-06 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Add info to trainingsset for linear model
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2919 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 386973da67f622605284e048ce717cb10600c9d2

@@
 \Abstract{
-Despite the occurrence of circular data in a wide range of scientific
-fields, the existing methodology for probabilistically forecasting
-circular data
-is highly limited. Most of the existing methods are built on the
-framework of
-generalized linear and additive models, which can incorporate
-difficulties in
-both the optimization and the interpretation. As a very intuitive
-alternative
+Despite the occurrence of circular data in a wide range of scientific fields,
+the existing methodology for probabilistically forecasting circular data is
+highly limited. Most of the existing methods are built on the framework of
+generalized linear and additive models, which can incorporate difficulties in
+both the optimization and the interpretation. As a very intuitive alternative
 to additive circular regression models, we extend previous ideas of circular
 trees and introduce probabilistic circular regression trees and forests
-employing the von Mises distribution. The resulting tree-based models
-simplify
-the estimation process by separating the data into more homogeneous
-subgroups
+employing the von Mises distribution. The resulting tree-based models simplify
+the estimation process by separating the data into more homogeneous subgroups
 and fitting a simple intercept-only model for each resulting subset. By
@@
 distribution parameters are considered and a fully specified distribution is
-provided for each terminal node. The proposed circular regression trees are easy to
-interpret, can detect non-additive effects, and select covariates and their
-interactions automatically. By combining an ensemble of circular regression
-trees a more stable and regularized circular regression forest can be built
-which also allows for modeling smooth effects.
-
-To evaluate the proposed circular regression trees and forests, a case
-study on
+provided for each terminal node. The proposed circular regression trees are
+easy to interpret, can detect non-additive effects, and select covariates and
+their interactions automatically. By combining an ensemble of circular
+regression trees a more stable and regularized circular regression forest can
+be built which also allows for modeling smooth effects.
+
+To evaluate the proposed circular regression trees and forests, a case study on
 probabilistic forecasting of wind direction for two airports in Austria is
-presented. In addition to the tree-based models, alternative common
-approaches
-are used as a benchmark in order to assess the predictive performances of
-the novel methodology of circular regression trees and forests.
-%Establishing a proper regression model for circular data such as animal movement
-%directions, periodic measurements of neuronal activity or wind directions,
-%requires specific attention regarding its representation on the unit circle.
-%Commonly applied generalized linear and additive models can provide a useful toolbox, 
-%however, highly non-linear link functions can complicate interpretations while
-%a non-concave log-likelihood function of the usually employed von Mises distribution 
-%leads to difficulties in the incorporated optimization procedure.
-%Moreover, an adequate model specification can be very challenging and non-additive 
-%effects might not be detected.
-
-%Therefore, an alternative approach to modeling circular data is proposed by
-%employing the von Mises distribution within the methodology of distributional 
-%trees and forests. Exploiting the tree-based separation of the data set into 
-%subgroups an intercept-model is fitted within each subset such that no covariates 
-%are included simplifying ....
-%difficulties may arise in 
-%modeling smooth effects due to highly non-linear link 
-%functions and increasing irregularities when moving closer to the ``origin point'' 
-%(0 which is equal to $2\pi$ in the radiant case).
-%%% FIXME: clear explanation of difficulties
-%Therefore, we propose a new approach by applying distributional trees and 
-%forests to circular data. In particular, the parameters of the employed von Mises 
-%distribution are regressed on covariates by a tree-based model yielding a constant set 
-%of estimated distribution parameters for each resulting subgroup, regardless of any 
-%``circular shift''/the ``circular position''.
-%As circular data appears in many different areas, the development of
-%appropriate statistical models which can account for periodicity has become an
-%important but challenging research field. For probabilistic modeling of
-%circular data typically the von Mises distribution is employed. In order to
-%regress its parameters on covariates, a circular regression tree is
-%proposed as an alternative to more commonly-used generalized linear and additive 
-%models. 
-%Moreover, the resulting circular regression trees are easy to interpret, 
-%can detect non-additive effects, and select covariates and their interactions automatically. 
-%By combining an ensemble of circular regression trees a more stable and regularized circular
-%regression forest can be built which allows for modeling smooth effects as
-%well.
-
-%To evaluate the proposed circular regression trees and forests, a case study on
-%probabilistic forecasting of wind direction for two airports in Austria is
-%presented. In addition to the tree-based models, alternative common approaches
-%are used as a benchmark in order to assess the predictive performances of
-%circular regression trees and forests. 
+presented. In addition to the tree-based models, alternative common approaches
+are used as a benchmark in order to assess the predictive performances of the
+novel methodology of circular regression trees and forests.
 }
+
 \Keywords{Distributional Trees; Random Forests; Circular Data; von Mises Distribution}
@@
     addition, a constant concentration parameter $\kappa$ is fitted to the full
-    learning sample.
+    learning sample. To allow for seasonally varying error characteristics in both bias and slope
+    coefficients, and to allow for seasonal heteroscedasticity captured by the concentration
+    parameter, we use the same time-adaptive trainings approach as for the
+    climatological model; hence, separate models are estimated over all observation time points,
+    using the same daytime of $31$~days centered around the day of interest over
+    all available years in the training data \citep{Lang+Lerch+Mayr:2019}.
     %%

### 2019-12-09 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Remove all comments, add overview, adapt bibfile
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2920 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7ebd0a65b072806dfd1b1b818aa9217f1ba3d815

@@
 
-Many approaches to model circular data assume that the circular
-variable of interest follows a circular distribution, in particular the von
-Mises distribution which is also known as the ``circular normal distribution''.
-%This is also the basis
-%which early regression models for circular data are built on as they regress
-%the parameter(s) of the von Mises distribution on covariates. 
-One of the first
-regression models with a circular response variable and linear covariates was
-presented by \cite{Gould:1969} where the circular mean, i.e., the location
-parameter of the von Mises distribution, is predicted by a linear combination of
-covariates. 
-\cite{Johnson+Wehrly:1978} refined this idea by
-plugging in a link function transforming the linear predictor to a 
-restricted interval of length $2\pi$. 
-%%%
-%%%In order to restrict the circular response to an interval of length $2\,pi\
-%%%\cite{Johnson+Wehrly:1978} refined this idea by linintroducing link functions transforming
-%%%the linear predictor to the restricted interval of length $2\pi$. 
-%\cite{Johnson+Wehrly:1978} took this idea one step further by
-%plugging in a link function such that the values of the linear predictor are
-%transformed to a restricted interval of length $2\pi$. 
-%This generalized linear
-%model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
-%subsequently by \cite{Fisher:1993} introducing another specific transformation
-%function which maps the values of the real line to the unit circle. 
-This generalized linear
-model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
-subsequently by \cite{Fisher:1993} introducing independent GLMs for either the 
-location or scale parameter with appropriate link functions while keeping the other 
-parameter constant. Additionally, they developed a combined heteroscedastic version 
-where the location and scale parameter are alternately reestimated conditional on 
-respective sets of covariates until convergence.
-%Moreover, the presented methodology does not only allow for the mean direction to be modeled 
-%depending on covariates while a constant estimate is obtained for the concentration parameter 
-%but also for the vice versa version. For each of these two approaches a separate model is introduced, 
-%however, they are also combined in a joint model (referred to as ``mixed model'').
-%%Moreover, the presented methodology allows for both distribution parameters, the mean
-%%direction and the concentration parameter of the von Mises distribution, to be
-%%modeled depending on covariates, not only separately but also in a
-%%combined model (also referred to ``mixed model''). 
-%In the employed iterative optimization algorithm the two sets of regression coefficients used
-%to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
-%%To obtain proper initial values it is recommended to apply two separate models 
-%%where either one of the parameters is fixed while the other is predicted employing covariates 
-%%in a GLM-type model.
-%While the combined model but also the two separate models are built on well elaborated theory 
-While all of these models are built on well elaborated theory 
-the employed methodology has shown to be very challenging in its application, mainly due to the 
-complexity encountered in optimizing the corresponding log-likelihood function which is not globally concave. 
-Therefore, highly informative starting values are crucial for these circular GLMs to converge 
-\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
-In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
-have presented a Bayesian alternative of a homoscedastic GLM for circular data.
-However, apart from potential difficulties in the optimization procedure of circular GLMs, 
-additive effects and their interpretation are often not intuitive for circular responses
-due to the highly non-linear link function and the not-straightforward representation of smooth 
-transitions on the unit circle. For example, the same rotation can be obtained in either 
-positive or negative direction on the circle leading to an ambiguous interpretation.
-
-%%%All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
-%%%are built on additive regression models.
-%%%Generally these provide a powerful toolbox for modeling particularly smooth effects.
-%%%however, in the case of modeling circular data 
-%%%additive effects conditional on linear covariates are not intuitive
-%%%due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
-%%%can be obtained in either positive or negative direction on the circle.
-%a possibly highly non-linear link function
-%can induce a more complex behavior of the modeled parameters as the location
-%parameter shifts closely towards the ``origin point'' $0$, also depending on the 
-%direction of the shift.
+Many approaches to model circular data assume that the circular variable of
+interest follows a circular distribution, in particular the von Mises
+distribution which is also known as the ``circular normal distribution''. One
+of the first regression models with a circular response variable and linear
+covariates was presented by \cite{Gould:1969} where the circular mean, i.e.,
+the location parameter of the von Mises distribution, is predicted by a linear
+combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
+plugging in a link function transforming the linear predictor to a restricted
+interval of length $2\pi$. This generalized linear model (GLM)-type approach
+was further extended by \cite{Fisher+Lee:1992} and subsequently by
+\cite{Fisher:1993} introducing independent GLMs for either the location or
+scale parameter with appropriate link functions while keeping the other
+parameter constant. Additionally, they developed a combined heteroscedastic
+version where the location and scale parameter are alternately reestimated
+conditional on respective sets of covariates until convergence. While all of
+these models are built on well elaborated theory the employed methodology has
+shown to be very challenging in its application, mainly due to the complexity
+encountered in optimizing the corresponding log-likelihood function which is
+not globally concave. Therefore, highly informative starting values are
+crucial for these circular GLMs to converge
+\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
+this strong dependence on appropriate initial values,
+\cite{Mulder+Klugkist:2017} have presented a Bayesian alternative of a
+homoscedastic GLM for circular data. However, apart from potential
+difficulties in the optimization procedure of circular GLMs, additive effects
+and their interpretation are often not intuitive for circular responses due to
+the highly non-linear link function and the not-straightforward representation
+of smooth transitions on the unit circle. For example, the same rotation can be
+obtained in either positive or negative direction on the circle leading to an
+ambiguous interpretation.
 
@@
 tree-based regression approach for modeling circular data by applying the von
-Mises distribution within the methodology of distributional trees and
-forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
-%In order to avoid these difficulties in regression modeling of circular data 
-%(and also the requirement of initial values) 
-%we propose an alternative and very flexible 
-%approach by applying the von Mises distribution within the methodology of distributional 
-%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
-The resulting circular regression trees and forests avoid the discussed difficulties 
-of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
-simple intercept-only model for each resulting subset.
-Therefore, no covariates and, thus, no link functions must be employed in the estimation
-of the group-specific distribution parameters which also simplifies the optimization process.
-By incorporating the framework of distributional modeling, we allow for detecting 
-splits in all distribution parameters and provide a fully specified circular response distribution 
-for each terminal node offering a wide range of statistical inference.
-In addition, the employed tree structure allows to capture non-additive effects while forests 
-enable the modeling of smooth changes.
-Furthermore, covariates and their possible interactions do not need to be specified in advance as 
-they are selected automatically.
-
-%Moreover, non-additive effects are captured by the tree structure while forests allow for 
-%modeling smooth changes as well. Thus, a high level of flexibility is provided while a full 
-%distributional model is specified in each terminal node offering a wide range of inference.
-
-%The resulting circular regression trees and forests do not require a pre-specified link function 
-%or any other model assumptions as covariates and their possible interactions are
-%selected automatically.
-
-%Moreover, as covariates are employed for splitting only and not
-%considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-This novel approach to circular regression trees and forests complements the literature 
-on tree-based circular modeling.
-%In particular, 
-\cite{Lund:2002} has already introduced the concept of circular regression trees where 
-binary splits are made based on an angular distance measure employed to evaluate node 
-homogeneity.
-While this approach does not only
-allow for linear, but also for circular covariates, the tree solely considers effects in the mean 
-direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
-Therefore, the introduced distributional approach to circular regression trees extends the concept
-of previous circular trees by considering splits in all distribution parameters and 
-providing a full probabilistic model. In addition, as a natural extension, ensembles or forests 
-of circular regression trees are presented in this study. 
-% that can improve forecasts by regularizing and stabilizing the model.
-
-%which is evaluated between 
-%the observed response values and the conditional mean direction of the circular response variable
-%and in that way serves as a measure of homogeneity within subgroups. 
-%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
-%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
-%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
-%%subset of observations.
-%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
-%along with the predicted mean value in each terminal node
-%By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.
-
-%Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
-%of tree-based circular models by incorporating the framework of distributional modeling and in that way
-%considering splits in all distribution parameters and providing a fully specified distribution fitted 
-%in each terminal node.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%An alternative modeling approach that can deal with non-additive effects is
-%provided by tree-based methods. \cite{Lund:2002} introduced the concept of
-%circular regression trees which was further extended by
-%\cite{Larraondo+Inza+Lozano:2018}. Following this idea of employing tree-based
-%methods but also including the probabilistic concept of regressing distribution
-%parameters on covariates, we propose a very flexible approach by applying the
-%von Mises distribution within the methodology of distributional trees and
-%distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
-%data. The resulting circular distributional trees and forests do not require
-%any model assumptions as covariates and their possible interactions are
-%selected automatically. Non-additive effects are captured by the tree structure
-%while circular forests allow for modeling smooth changes as well. Thus, a high
-%level of flexibility is provided while a full distributional model is specified
-%offering a wide range of inference.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Mises distribution within the methodology of distributional trees and forests
+\citep{Schlosser+Hothorn+Stauffer:2019}. The resulting circular regression
+trees and forests avoid the discussed difficulties of circular GLMs by
+separating the data into more homogeneous subgroups and fitting a simple
+intercept-only model for each resulting subset. Therefore, no covariates and,
+thus, no link functions must be employed in the estimation of the
+group-specific distribution parameters which also simplifies the optimization
+process. By incorporating the framework of distributional modeling, we allow
+for detecting splits in all distribution parameters and provide a fully
+specified circular response distribution for each terminal node offering a wide
+range of statistical inference. In addition, the employed tree structure
+allows to capture non-additive effects while forests enable the modeling of
+smooth changes. Furthermore, covariates and their possible interactions do not
+need to be specified in advance as they are selected automatically.
+
+This novel approach to circular regression trees and forests complements the
+literature on tree-based circular modeling. \cite{Lund:2002} has already
+introduced the concept of circular regression trees where binary splits are
+made based on an angular distance measure employed to evaluate node
+homogeneity. While this approach does not only allow for linear, but also for
+circular covariates, the tree solely considers effects in the mean direction
+for splitting, i.e., only the location parameter $\mu$ is regressed on
+covariates. Therefore, the introduced distributional approach to circular
+regression trees extends the concept of previous circular trees by considering
+splits in all distribution parameters and providing a full probabilistic model.
+In addition, as a natural extension, ensembles or forests of circular
+regression trees are presented in this study.
 
 \subsubsection*{Motivating example}
-To provide a first impression of the presented methodology a circular regression 
-tree is employed for probabilistic wind direction forecasting. Wind direction is a 
-classical circular quantity and accurate forecasts are of great importance for
-decision-making processes, e.g., in air traffic management as considered in this study.
-Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
-airport of Innsbruck which lies at the bottom of a narrow valley within the European Alps.
-Topography channels the wind along the west-east valley axis or along a tributary
-valley intersecting from the south. Hence, pressure gradients to which valley
-wind regimes react are considered as covariates along with other meteorological
-measurements (lagged by one hour) and their derivatives, such as 
-wind direction and wind speed at the airport or spatial and temporal differences.
-%%
-\begin{figure}[p!]
-  \centering
+To provide a first impression of the presented methodology a circular
+regression tree is employed for probabilistic wind direction forecasting. Wind
+direction is a classical circular quantity and accurate forecasts are of great
+importance for decision-making processes, e.g., in air traffic management as
+considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
+for 1-hourly forecasts at the airport of Innsbruck which lies at the bottom of
+a narrow valley within the European Alps. Topography channels the wind along
+the west-east valley axis or along a tributary valley intersecting from the
+south. Hence, pressure gradients to which valley wind regimes react are
+considered as covariates along with other meteorological measurements (lagged
+by one hour) and their derivatives, such as wind direction and wind speed at
+the airport or spatial and temporal differences.
+%%
+\begin{figure}[p!] \centering
 \includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
@@
 empirical histogram (gray) and fitted density (red line) are depicted along
-  with the estimated location parameter (red hand). The covariates selected for splitting
-  are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
-  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
-  one hour.
-  Note that in the meteorological context wind direction is defined on the 
-  scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
-}
-  \label{fig:tree_ibk} 
-\end{figure}
+with the estimated location parameter (red hand). The covariates selected for
+splitting are wind direction (degree), wind speed ($\text{ms}^{-1}$), and
+pressure gradients ($\text{dpressure; hPa}$) west, east and south of the
+airport, all lagged by one hour. Note that in the meteorological context wind
+direction is defined on the scale $[0,360]$ degree and increases clockwise from
+North ($0$ degree). } \label{fig:tree_ibk} \end{figure}
 %%
@@
 winds blowing from the valley mouth towards the upper valley (from east to
-west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest
-along the intersecting valley towards Innsbruck (from south-east to north-west,
+west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest along
+the intersecting valley towards Innsbruck (from south-east to north-west,
 node~8); (3) Down-valley winds blowing in the direction of the valley mouth
@@
 regimes and are consequently associated with a very low estimated concentration
-parameter~$\hat{\kappa}$, i.e., a high estimated variance. 
-In terms of covariates, the lagged wind direction (``persistence'') is mostly 
-responsible for distinguishing the broad range of wind regimes listed above while
-the pressure gradients and wind speed separate the data into subgroups with high 
-vs.\ low precision. 
-A more extensive case study of circular regression trees and forests
-applied to probabilistic forecasting of wind direction at the airports of
-Innsbruck and Vienna will be presented in this paper. Moreover, their
-performances are benchmarked to alternative commonly used approaches.
-
-The remainder of the paper is structured as follows: The theory on probabilistic
-circular modeling introducing the von Mises distribution, and circular regression
-models will be discussed in
-Section~\ref{sec:prob_circ}. The methodology of circular regression trees
-and forests and their features are introduced in Section~\ref{sec:tree_forest}.
-After the case study presented in \ref{sec:wind}, a comprehensive summary and
-conclusion is given in Section~\ref{sec:summary}.
+parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
+covariates, the lagged wind direction (``persistence'') is mostly responsible
+for distinguishing the broad range of wind regimes listed above while the
+pressure gradients and wind speed separate the data into subgroups with high
+vs.\ low precision. A more extensive case study of circular regression trees
+and forests applied to probabilistic forecasting of wind direction at the
+airports of Innsbruck and Vienna will be presented in this paper. Moreover,
+their performances are benchmarked to alternative commonly used approaches.
+
+The remainder of the paper is structured as follows: The theory on
+probabilistic circular modeling introducing the von Mises distribution, and
+circular regression models will be discussed in Section~\ref{sec:prob_circ}.
+The methodology of circular regression trees and forests and their features are
+introduced in Section~\ref{sec:tree_forest}. After the case study presented in
+\ref{sec:wind}, a comprehensive summary and conclusion is given in
+Section~\ref{sec:summary}.
 
@@
 % -------------------------------------------------------------------
-%\newpage
 \section{Probabilistic circular modeling}
@@
 where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
-$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and \citealt{Ley+Verdebout:2017} for a more detailed overview).
+$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and
+\citealt{Ley+Verdebout:2017} for a more detailed overview).
 
@@
 an indicator of deviation and in that way a measure of goodness of fit of the
-model can be obtained for each observation and fitted parameter.
-Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
-defined by 
+model can be obtained for each observation and fitted parameter. Moreover, the
+optimization problem in Equation~\ref{eq:optim} can also be defined by 
 %%
@@
 right panel displays an appropriate representation of a circular quantity on
-the unit circle. Looking at the empirical histogram along with 
-the fitted density (red line) and estimated
-location parameter (red hand) this example shows that reasonably well fitting
-probabilistic models can be established for circular data by specifying an
-appropriate distribution family such as the von Mises distribution and
-estimating its parameters via maximum likelihood. However, this procedure
-considers only the response variable. Including covariates can improve any
-model remarkably as they might provide valuable additional information.
-Therefore, (distributional) regression models are of high interest for all types
-of data, including circular data.
-
-In most generalized linear or additive approaches to circular regression modeling 
-the location parameter $\mu$ is regressed on covariates $\mathbf{z}$ in the form of 
-%models employ linear combinations of the covariates to connect them to distribution 
-%parameters. By plugging in a link function $g$ 
+the unit circle. Looking at the empirical histogram along with the fitted
+density (red line) and estimated location parameter (red hand) this example
+shows that reasonably well fitting probabilistic models can be established for
+circular data by specifying an appropriate distribution family such as the von
+Mises distribution and estimating its parameters via maximum likelihood.
+However, this procedure considers only the response variable. Including
+covariates can improve any model remarkably as they might provide valuable
+additional information.  Therefore, (distributional) regression models are of
+high interest for all types of data, including circular data.
+
+In most generalized linear or additive approaches to circular regression
+modeling the location parameter $\mu$ is regressed on covariates $\mathbf{z}$
+in the form of 
 \begin{equation}
@@
 \end{equation}
-with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$; whereby the 
-additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 
-As an appropriate link function typically $g(x) = 2 \cdot arctan(x)$ is employed, 
-as suggested by \citet{Fisher+Lee:1992}.
-While most of the proposed models estimate a constant concentration parameter, hence, assume
-homoscedasticity, \citet{Fisher+Lee:1992} also developed a heteroscedastic version by combining two  
-individual GLMs, each for one of the parameters $\mu$ and $\kappa$. 
-This provides a first approach to a fully probabilistic regression model for circular 
-data, however, the parameters are not regressed simultaneously on covariates as known from the 
-more general framework provided by generalized additive models for location, scale, and shape 
-\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. 
-Nevertheless, all circular additive models incorporate the previously discussed difficulties induced by
-the characteristics of the log-likelihood function and the strongly non-linear link function.
-%For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
-%on an interval of length $2\pi$ by setting 
-%$g(x) = 2\pi \cdot F(x)$ 
-%for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
-%$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
-%Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
-%as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
-%covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
-%constant value. Additionally to these two model variations where only one of the two distribution 
-%parameters depends on covariates a combination of them is introduced. In this joint model the 
-%two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
-%in alternation within an iterative optimization algorithm. 
-%Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
-%data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
-%more general case by generalized additive models for location, scale, and shape 
-%\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
-%However, in practice the incorporated optimization procedure has shown to be very complex, already
-%for only one parameter being regressed on covariates and consequently even more so in the joint model.
-%This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
-%distribution with the plugged-in link function. In particular, the function is not globally concave and
-%possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
-%Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
-%is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
-%covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
-%compression of the estimated density function.
-%\newline
-Referring to additive models in general, it has to be considered that a proper model specification 
-can be very challenging, particularly for a high number of covariates and no information on possible 
-interactions. Moreover, the additive structure might impose a smooth effect even if the true underlying 
-effect is an abrupt shift.
-
-An alternative to additive models, which can deal or even 
-avoid the mentioned difficulties in circular regression modeling are 
-tree-based models. For the particular case of circular data, the methodology 
-of circular regression trees and forests is presented in the following section.
+with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$;
+whereby the additive predictor is transformed to an interval of length $2\pi$
+by a link function $g$. As an appropriate link function typically $g(x) = 2
+\cdot arctan(x)$ is employed, as suggested by \citet{Fisher+Lee:1992}. While
+most of the proposed models estimate a constant concentration parameter, hence,
+assume homoscedasticity, \citet{Fisher+Lee:1992} also developed a
+heteroscedastic version by combining two individual GLMs, each for one of the
+parameters $\mu$ and $\kappa$. This provides a first approach to a fully
+probabilistic regression model for circular data, however, the parameters are
+not regressed simultaneously on covariates as known from the more general
+framework provided by generalized additive models for location, scale, and
+shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular
+additive models incorporate the previously discussed difficulties induced by
+the characteristics of the log-likelihood function and the strongly non-linear
+link function. Referring to additive models in general, it has to be
+considered that a proper model specification can be very challenging,
+particularly for a high number of covariates and no information on possible
+interactions. Moreover, the additive structure might impose a smooth effect
+even if the true underlying effect is an abrupt shift.
+
+An alternative to additive models, which can deal or even avoid the mentioned
+difficulties in circular regression modeling are tree-based models. For the
+particular case of circular data, the methodology of circular regression trees
+and forests is presented in the following section.
 
@@
 
-The main goal of employing tree-based models to circular data is to provide a flexible and 
-easy-to-apply modeling approach which accounts for a proper representation on the unit circle.
-Based on ideas of \citet{Lund:2002}, we introduce probabilistic circular regression trees and forests
-employing the von Mises distribution. The resulting tree-based models provide a very intuitive and data-driven 
-alternative to commonly used regression models for circular data.
-
-%A very flexible framework should allow for detecting a wide range of 
-%effects, from abrupt changes to smooth transitions, while for the latter difficulties related
-%to a circular representation should be avoided as well as any dependence on initial values.
-%\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
-%variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
-%and where to split the data an angular distance measure is evaluated between the mean direction 
-%and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
-%Hence, only the location parameter is considered for splitting such that effects in other parameters 
-%might be missed. Therefore, we introduce a distributional approach to circular regression trees 
-%where a fully specified distributional model is fitted in each node of the tree which enables the 
-%detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
-%corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
-%quantities.
-%Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
-%within each node is fitted without employing covariates.
-%A common way of including covariates is to regress the model parameters on
-%covariates. In the case of a probabilistic model for a circular response this
-%corresponds to expressing the location parameter $\mu$ and the concentration
-%parameter $\kappa$ of the von Mises distribution by means of the available
-%covariates, e.g., in the form of a linear combination, possibly with an
-%appropriate link function plugged in as in GLMs. This kind of approach is
-%presented for example in the mixed model of \cite{Fisher+Lee:1992} and
-%\cite{Fisher:1993}. Apart from the methodology developed specifically for
-%circular data more general approaches to regressing model parameters on
-%covariates are commonly used such as for example generalized additive models
-%for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
-%where each distribution parameter is estimated by its own generalized additive
-%model (GAM). However, difficulties might arise in case of non-additive changes
-%and a proper model specification can be very challenging, particularly for a
-%high number of covariates and no information on possible interactions. As an
-%alternative regression approach which can deal with these problems, we
-%introduce a circular probabilistic tree-structured model in this section.
+The main goal of employing tree-based models to circular data is to provide a
+flexible and easy-to-apply modeling approach which accounts for a proper
+representation on the unit circle. Based on ideas of \citet{Lund:2002}, we
+introduce probabilistic circular regression trees and forests employing the von
+Mises distribution. The resulting tree-based models provide a very intuitive
+and data-driven alternative to commonly used regression models for circular
+data.
 
@@
 estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
-two distribution parameters ($\mu$ and $\kappa$) and a data set of $n$ observations, 
-this yields an $n \times 2$ matrix that can be employed as a discrepancy measure, capturing
-how well each given observation conforms with the estimated location
-$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
-on covariates, the association between the model's scores and each available
-covariate is assessed using either a parameter instability test~(MOB) or a
-permutation test~(CTree). By doing so in each partitioning step, the covariate
-with the highest significant association (i.e., lowest significant $p$-value,
-if any) is selected for splitting the data. The corresponding split point is
-chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
-statistic~(CTree) over all possible partitions. This procedure is repeated
-recursively until there are no significant parameter instabilities or until
-another stopping criterion is met (e.g., subgroup size or tree depth).
-A more detailed description of the applied tree-building algorithm
-can be found in Appendix~\ref{app:algorithm}.
+two distribution parameters ($\mu$ and $\kappa$) and a data set of $n$
+observations, this yields an $n \times 2$ matrix that can be employed as a
+discrepancy measure, capturing how well each given observation conforms with
+the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, respectively.
+To capture dependence on covariates, the association between the model's scores
+and each available covariate is assessed using either a parameter instability
+test~(MOB) or a permutation test~(CTree). By doing so in each partitioning
+step, the covariate with the highest significant association (i.e., lowest
+significant $p$-value, if any) is selected for splitting the data. The
+corresponding split point is chosen either by optimizing the
+log-likelihood~(MOB) or a two-sample test statistic~(CTree) over all possible
+partitions. This procedure is repeated recursively until there are no
+significant parameter instabilities or until another stopping criterion is met
+(e.g., subgroup size or tree depth). A more detailed description of the
+applied tree-building algorithm can be found in Appendix~\ref{app:algorithm}.
 
@@
 unweighted full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$
-corresponds to the abrupt splits from a tree. Thus, circular regression
-forests can capture both smooth and abrupt changes, while covariates and
-possible interactions are selected automatically allowing for an easy-to-apply
+corresponds to the abrupt splits from a tree. Thus, circular regression forests
+can capture both smooth and abrupt changes, while covariates and possible
+interactions are selected automatically allowing for an easy-to-apply
 methodology with a full specification of the von Mises distribution.
@@
 \label{sec:wind}
-As motivated in Section~\ref{sec:introduction},
-accurate forecasts of wind directions are of great importance for risk
-management in various fields such as agriculture, energy production or air
-safety control. For example, in order to direct airplanes to a safe landing,
-precise knowledge of wind direction for the next hour(s) at the respective
-airport is highly desirable and adequate prediction methods are required. In
-this section, a comprehensive case study on probabilistic wind direction
-forecasting is provided for two airports in Austria with fundamentally
-different site characteristics: Innsbruck with a rather unusual location for an
-airport, located at the bottom of a narrow Alpine valley, and Vienna located
-within flat terrain in the East of Austria. In addition to the newly introduced
-circular regression trees and forests, the case study employs various
-alternative approaches to forecast wind direction as benchmarks. The study is
-based on $1$\,h and $3$\,h forecasts employing lagged observations in the
-vicinity of the airports as possible predictor variables.
+As motivated in Section~\ref{sec:introduction}, accurate forecasts of wind
+directions are of great importance for risk management in various fields such
+as agriculture, energy production or air safety control. For example, in order
+to direct airplanes to a safe landing, precise knowledge of wind direction for
+the next hour(s) at the respective airport is highly desirable and adequate
+prediction methods are required. In this section, a comprehensive case study on
+probabilistic wind direction forecasting is provided for two airports in
+Austria with fundamentally different site characteristics: Innsbruck with a
+rather unusual location for an airport, located at the bottom of a narrow
+Alpine valley, and Vienna located within flat terrain in the East of Austria.
+In addition to the newly introduced circular regression trees and forests, the
+case study employs various alternative approaches to forecast wind direction as
+benchmarks. The study is based on $1$\,h and $3$\,h forecasts employing lagged
+observations in the vicinity of the airports as possible predictor variables.
 
 \subsection{Data}
+%%
+\begin{figure}[t]
+  \setkeys{Gin}{width=1.0\textwidth}
+  \centering
+  \includegraphics{_plot_circforest_stationmap_combined.pdf}
+  \caption{Overview of the study area for the airports Innsbruck and Vienna. 
+    For Innsbruck, $4$~stations at the airport and $6$~along the
+    intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
+    and $13$~within an area of approximately 30\,km$^2$ are used.
+    Elevation data are obtained from the TandDEM-X digital elevation model with a 
+    horizontal resolution of $90$~m \citep{Wessel:2018}.}
+  \label{fig:overview} 
+\end{figure}
+%%
 \label{sec:wind:data}
 The circular response variable considered in this case study is a $10$\,min
-mean of wind direction measurements at the airport for Innsbruck and Vienna on an
-hourly temporal resolution. As predictor variables, we use temporal information
-and $1$-hourly resolved $10$\,min mean observations of various meteorological quantities such as 
-wind direction, wind speed, temperature, air pressure and humidity, all lagged by
-one or three hours according to the respective forecasting step. The
-meteorological variables are measured either directly at the airports or within
-their vicinities. For Innsbruck, measurements at the airport and along the
-intersecting valleys are used, whereas, for Vienna, measurements at the airport
-and within an area of approximately 30\,km$^2$ are used. A topographical overview of
-the airports and their surrounding areas with the station sites employed in
-this study is provided in Figure~XXX. In addition, we use derived quantities
+mean of wind direction measurements at the airport for Innsbruck and Vienna on
+an hourly temporal resolution. As predictor variables, we use temporal
+information and $1$-hourly resolved $10$\,min mean observations of various
+meteorological quantities such as wind direction, wind speed, temperature, air
+pressure and humidity, all lagged by one or three hours according to the
+respective forecasting step. The meteorological variables are measured either
+directly at the airports or within their vicinities. For Innsbruck,
+measurements at the airport and along the intersecting valleys are used,
+whereas, for Vienna, measurements at the airport and within an area of
+approximately 30\,km$^2$ are used. A topographical overview of the airports and
+their surrounding areas with the station sites employed in this study is
+provided in Figure~\ref{fig:overview}. In addition, we use derived quantities
 such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
@@
 
-The data used in this study consists of five subsequent years from January 2015 to
-December 2018. After first eliminating predictor variables with more than 5\%
-missing values and then time points with missing observations, the data set
-consists of 41\,979 time points and 260 covariates for Innsbruck, and of 38\,985
-times points and 494 covariates for Vienna, respectively.
+The data used in this study consists of five subsequent years from January 2015
+to December 2018. After first eliminating predictor variables with more than
+5\% missing values and then time points with missing observations, the data set
+consists of 41\,979 time points and 260 covariates for Innsbruck, and of
+38\,985 times points and 494 covariates for Vienna, respectively.
 
@@
     \newline
-    For simplicity, rather to fit one single climatological model, separate models are estimated 
-    rolling over all observation time points, 
-    using the same daytime of $31$~days centered around the day of interest over all available years in the training data.
-    %For simplicity, rather to fit one single climatological model, we perform the
-    %model fitting separately rolling over all time points. For each time point, the
-    %training data set consists of the same daytime as the time of interest within
-    %31~days centered around the same calender day over all years except the year of
-    %interest. 
-    To gain full probabilistic climatological forecasts, we employ the
-    steps described in Section~\ref{sec:prob_circ} and fit the distribution
-    parameters of the von Mises distribution via maximum likelihood estimation. The
-    construction of a climatological reference model was similarly pursued by
-    \citet{Vogel+Knippertz+Fink:2018} and is discussed in a comprehensive summary on
-    different time-adaptive training schemes in \citet{Lang+Lerch+Mayr:2019}.
+    For simplicity, rather to fit one single climatological model, separate
+    models are estimated rolling over all observation time points, using the same
+    daytime of $31$~days centered around the day of interest over all available
+    years in the training data.  To gain full probabilistic climatological
+    forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
+    the distribution parameters of the von Mises distribution via maximum
+    likelihood estimation. The construction of a climatological reference model was
+    similarly pursued by \citet{Vogel+Knippertz+Fink:2018} and is discussed in a
+    comprehensive summary on different time-adaptive training schemes in
+    \citet{Lang+Lerch+Mayr:2019}.
     %%
@@
     \newline
-    To gain a full probabilistic persistency model, we proceed similarly as for the
-    climatological model by using maximum likelihood estimation and fitting the
+    To gain a full probabilistic persistency model, we proceed similarly as for
+    the climatological model by using maximum likelihood estimation and fitting the
     distribution parameters of the von Mises distribution conditional on lagged
     response values according to the description in Section~\ref{sec:prob_circ}.
-    We fit one model for every hour throughout 
-    the validation data set employing the
-    previous six lagged response values as training data. In order to allow for a
-    stronger influence of observations closer to the time of interest, exponential
-    smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
-    prediction an equal influence rate of 50 percent is assigned both to the current 
-    observation and to the previous five observations together. Observations with longer 
-    time lags have exponential weights below $0.01$ and are therefore omitted from the
-    training data.
+    We fit one model for every hour throughout the validation data set employing
+    the previous six lagged response values as training data. In order to allow for
+    a stronger influence of observations closer to the time of interest,
+    exponential smoothing is employed with a smoothing factor of $0.5$;
+    accordingly, for every prediction an equal influence rate of 50 percent is
+    assigned both to the current observation and to the previous five observations
+    together. Observations with longer time lags have exponential weights below
+    $0.01$ and are therefore omitted from the training data.
     %%
   \item \emph{Generalized linear model:}
-    Traditional approaches to forecast circular response variables are often based
-    on circular GLM-typed models \citep{Fisher:1993}. As discussed in
+    Traditional approaches to forecast circular response variables are often
+    based on circular GLM-typed models \citep{Fisher:1993}. As discussed in
     Section~\ref{sec:introduction}, circular regression models often experience the
@@
     addition, a constant concentration parameter $\kappa$ is fitted to the full
-    learning sample. To allow for seasonally varying error characteristics in both bias and slope
-    coefficients, and to allow for seasonal heteroscedasticity captured by the concentration
-    parameter, we use the same time-adaptive trainings approach as for the
-    climatological model; hence, separate models are estimated over all observation time points,
-    using the same daytime of $31$~days centered around the day of interest over
-    all available years in the training data \citep{Lang+Lerch+Mayr:2019}.
+    learning sample. To allow for seasonally varying error characteristics in both
+    bias and slope coefficients, and to allow for seasonal heteroscedasticity
+    captured by the concentration parameter, we use the same time-adaptive
+    trainings approach as for the climatological model; hence, separate models are
+    estimated over all observation time points, using the same daytime of $31$~days
+    centered around the day of interest over all available years in the training
+    data \citep{Lang+Lerch+Mayr:2019}.
     %%
@@
   \item \emph{Circular regression forest:} Following the description in
-    Section~\ref{sec:circforest}, the circular forest used in this study is constructed 
-    based on $100$~individual trees employing the R-package \pkg{circtree}. Each of these 
-    trees is again built by the CTree algorithm on a subsample containing a fraction of 
-    $30$~percent of the original learning data. As it can be expected that the lagged
-    response variable is one of the most informative predictor variables for wind
-    direction forecasting, all covariates are included for building each tree which
-    ensures that the lagged response variable is always considered for splitting.
-    This bagging approach can be applied in \pkg{circtree} by setting the argument
-    \code{mtry} to the total number of covariates. Since a high number of possible
-    split points leads to high computational costs, the covariates are binned in a
-    maximum of $50$ classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}).
-    Contrary to a single-tree model, forests usually consist of very large trees as
-    they are not prone to overfitting the data due to the stabilization obtained by
-    combining the individual trees. Therefore, we use the following control
-    arguments to build rather large trees: The minimal number of observations to
-    perform a split is set to 20 (argument \texttt{minsplit}), the minimal number
-    of observations in each terminal node is set to 7 (argument
-    \texttt{minbucket}), and the significance level for variable selection is kept
-    at its maximum value of $1$ (argument \texttt{alpha}).
+    Section~\ref{sec:circforest}, the circular forest used in this study is
+    constructed based on $100$~individual trees employing the R-package
+    \pkg{circtree}. Each of these trees is again built by the CTree algorithm on a
+    subsample containing a fraction of $30$~percent of the original learning data.
+    As it can be expected that the lagged response variable is one of the most
+    informative predictor variables for wind direction forecasting, all covariates
+    are included for building each tree which ensures that the lagged response
+    variable is always considered for splitting. This bagging approach can be
+    applied in \pkg{circtree} by setting the argument \code{mtry} to the total
+    number of covariates. Since a high number of possible split points leads to
+    high computational costs, the covariates are binned in a maximum of $50$
+    classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}). Contrary to a
+    single-tree model, forests usually consist of very large trees as they are not
+    prone to overfitting the data due to the stabilization obtained by combining
+    the individual trees. Therefore, we use the following control arguments to
+    build rather large trees: The minimal number of observations to perform a split
+    is set to 20 (argument \texttt{minsplit}), the minimal number of observations
+    in each terminal node is set to 7 (argument \texttt{minbucket}), and the
+    significance level for variable selection is kept at its maximum value of $1$
+    (argument \texttt{alpha}).
 \end{itemize}
@@
 
-In addition to the raw CRPS, corresponding skill scores are computed to
-assess differences in the improvement of the various statistical models over
-the climatological model used as a reference:
+In addition to the raw CRPS, corresponding skill scores are computed to assess
+differences in the improvement of the various statistical models over the
+climatological model used as a reference:
 %%
@@
 All scores presented in the next section are computed out-of-sample based on
-five years of data. For the persistency model only time points prior to
-the time of interest are used and the validation is performed rolling over all
+five years of data. For the persistency model only time points prior to the
+time of interest are used and the validation is performed rolling over all
 observations. For all other models a five-fold cross-validation is employed
@@
 per day over five years, some kind of temporal aggregation is needed to ensure
-a correct visual comparison of the individual methods. The analyses performed have
-shown that for the employed models the variability of the predictive
+a correct visual comparison of the individual methods. The analyses performed
+have shown that for the employed models the variability of the predictive
 performance over the five years is lower than over a single day or over a
@@
 at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
-Vienna (Panels~b,\,d). The scores are aggregated over the five validation years,
-yielding yearly mean values for every hour per calendar month, with a lower
-score indicating better performance. For both stations at both forecast steps,
-the circular forest overall provides the best predictive performance, followed
-by the circular regression tree and the persistency model; except for the $3$\,h~forecasts
-at Innsbruck where the persistency model is outperformed by all others. 
-In comparison to the circular regression tree and
-forest, for both stations and forecast steps, the climatological model and the
-linear model show clearly higher CRPS values and hence a lower predictive
-performance.
+Vienna (Panels~b,\,d). The scores are aggregated over the five validation
+years, yielding yearly mean values for every hour per calendar month, with a
+lower score indicating better performance. For both stations at both forecast
+steps, the circular forest overall provides the best predictive performance,
+followed by the circular regression tree and the persistency model; except for
+the $3$\,h~forecasts at Innsbruck where the persistency model is outperformed
+by all others. In comparison to the circular regression tree and forest, for
+both stations and forecast steps, the climatological model and the linear model
+show clearly higher CRPS values and hence a lower predictive performance.
 %%
@@
 remains rather constant in one of these possible states, but once a change
-takes place it is mostly a major wind direction shift, 
-e.g., from up-valley to down-valley. Due to the few wind
-regimes the rather inflexible climatological and linear models score relatively
-well with similar CRPS values as the other models
-(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
+takes place it is mostly a major wind direction shift, e.g., from up-valley to
+down-valley. Due to the few wind regimes the rather inflexible climatological
+and linear models score relatively well with similar CRPS values as the other
+models (Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
 potential high prediction errors in case of a change of the wind regime seem to
-lead to a higher variation in the predictive performance for all models in comparison to
-Vienna; this variation is especially high for the persistency model due to its
-strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
-geographic surroundings do not constrain any wind direction which leads to
-smaller and less abrupt changes in the wind direction as well as less
+lead to a higher variation in the predictive performance for all models in
+comparison to Vienna; this variation is especially high for the persistency
+model due to its strong vulnerability to abrupt wind shifts. On the contrary,
+at Vienna the geographic surroundings do not constrain any wind direction which
+leads to smaller and less abrupt changes in the wind direction as well as less
 pronounced wind regimes. This seems to weaken the predictive performance of the
@@
 predictive performance of the climatological model and the linear model at both
-stations. As expected, for the persistency model, at both stations, higher scores
-for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
-lower performance for longer lead times; this is due to the lower 
-information content of minimum 3-hourly instead of 1-hourly lagged response 
-values employed as covariates in the persistency model. 
-The circular regression tree and forest seem to
-partially compensate for the lower skill of the lagged response values by other
-covariates, hence their predictive performance only slightly decreases for the
-longer lead time. This compensation is especially evident for Innsbruck, where
-the performance difference between the persistency model and the tree-based
-methods significantly increases from the 1-hourly to the 3-hourly forecast.
+stations. As expected, for the persistency model, at both stations, higher
+scores for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal
+a lower performance for longer lead times; this is due to the lower information
+content of minimum 3-hourly instead of 1-hourly lagged response values employed
+as covariates in the persistency model. The circular regression tree and
+forest seem to partially compensate for the lower skill of the lagged response
+values by other covariates, hence their predictive performance only slightly
+decreases for the longer lead time. This compensation is especially evident for
+Innsbruck, where the performance difference between the persistency model and
+the tree-based methods significantly increases from the 1-hourly to the
+3-hourly forecast.
 
@@
 shows that while the persistency model's performance is lower than the
-reference (Panel~c) the tree-based models can compensate for the low skill of the 
-lagged response values employed as covariates and, hence, are still significantly 
-superior to the reference.
+reference (Panel~c) the tree-based models can compensate for the low skill of
+the lagged response values employed as covariates and, hence, are still
+significantly superior to the reference.
 
@@
 \label{sec:summary}
-Extending the toolbox for modeling circular data, circular regression trees and forests 
-are established by coupling model-based recursive partitioning with the von Mises distribution. 
-By separating the data into more homogeneous subgroups, possible difficulties in circular regression
-are bypassed as covariates are solely considered for splitting and 
-group-specific models are fitted as intercept-only models.
-%distributional models are fitted as intercept-only models in each node. 
-In addition, by specifying the von Mises distribution for each node and
-allowing for splits in both distribution parameters $\mu$ and $\kappa$, 
-fully probabilistic forecasts are provided.
-
-The performance of the novel circular trees and forests are assessed in a real case study for shortterm
-probabilistic wind direction forecasting at two airports with different site characteristics.
-As benchline models, probabilistic climatology and persistency models, as well as a state-of-the-art
-circular GLM are evaluated based on proper scoring rules.
-In summary, the circular trees and forests have the highest predictive performance in this setting.
-For cases without changes in the wind regime, lagged response values provide already
-highly skillful estimates leading to a good performance of the persistency model as observed for shortterm wind direction forecasts in this study.
-While in these cases the trees and forests also benefit from the highly informative lagged response, they can compensate for a lower information of this covariate by incorporating other quantities and possible interactions of these, contrary to the persistency model (see Figure~\ref{fig:boxplot_crpsskill}).
-Hence, the tree-based models provide reliable forecasts in all tested meteorological settings.
-For operational use, a possible extension could be the incorporation of numerical weather predictions as (additional) covariates.
-While this probably only slightly improves the predictive skill for short leadtimes, it possibly extends the potential forecast range of the different methods.
-Another possible improvement for obtaining more parsimonious circular trees is
-to consider splitting circular covariates into two circle segments by searching
-two split points simultaneously rather than sequentially at different depths.
-While this might slightly improve the predictive performance of circular trees,
-this should not affect the performance  of circular forests, as these
-consist of very large trees with many different splits.
-
-To conclude, in general the tree structure can capture non-linear changes, shifts, and potential interactions in 
-covariates without prespecification of such effects. As supported by the presented case study,
-this can be particularly useful for modeling a highly fluctuating response, such as typically 
-observed for wind direction, or/and in case of a large number of possible covariates.
-Moreover, the case study shows that building ensembles of circular trees can even improve the forecasting performance, as the resulting circular forests allow for modeling smooth effects and stabelize the model.
-
-%The comparison of the novel circular trees with simple climatological and persistency models, depicted
-%for probabilistic wind direction forecasts at two airports, shows that even in case of no obvious 
-%skillful covariates the trees have an high predictive performance.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{itemize}
-%  \item Add summary/conclusion of case study;
-%  \item Add summary/conclusion of circular regression trees and forests.
-%  \item Outlook on possible extensions (use of NWP data, circular splits, ...).
-%\end{itemize}
+Extending the toolbox for modeling circular data, circular regression trees and
+forests are established by coupling model-based recursive partitioning with the
+von Mises distribution. By separating the data into more homogeneous
+subgroups, possible difficulties in circular regression are bypassed as
+covariates are solely considered for splitting and group-specific models are
+fitted as intercept-only models. In addition, by specifying the von Mises
+distribution for each node and allowing for splits in both distribution
+parameters $\mu$ and $\kappa$, fully probabilistic forecasts are provided.
+
+The performance of the novel circular trees and forests are assessed in a real
+case study for shortterm probabilistic wind direction forecasting at two
+airports with different site characteristics. As benchline models,
+probabilistic climatology and persistency models, as well as a state-of-the-art
+circular GLM are evaluated based on proper scoring rules. In summary, the
+circular trees and forests have the highest predictive performance in this
+setting. For cases without changes in the wind regime, lagged response values
+provide already highly skillful estimates leading to a good performance of the
+persistency model as observed for shortterm wind direction forecasts in this
+study. While in these cases the trees and forests also benefit from the highly
+informative lagged response, they can compensate for a lower information of
+this covariate by incorporating other quantities and possible interactions of
+these, contrary to the persistency model (see
+Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
+reliable forecasts in all tested meteorological settings. For operational use,
+a possible extension could be the incorporation of numerical weather
+predictions as (additional) covariates. While this probably only slightly
+improves the predictive skill for short leadtimes, it possibly extends the
+potential forecast range of the different methods. Another possible
+improvement for obtaining more parsimonious circular trees is to consider
+splitting circular covariates into two circle segments by searching two split
+points simultaneously rather than sequentially at different depths. While this
+might slightly improve the predictive performance of circular trees, this
+should not affect the performance of circular forests, as these consist of
+very large trees with many different splits.
+
+To conclude, in general the tree structure can capture non-linear changes,
+shifts, and potential interactions in covariates without prespecification of
+such effects. As supported by the presented case study, this can be
+particularly useful for modeling a highly fluctuating response, such as
+typically observed for wind direction, or/and in case of a large number of
+possible covariates. Moreover, the case study shows that building ensembles of
+circular trees can even improve the forecasting performance, as the resulting
+circular forests allow for modeling smooth effects and stabelize the model.
 

### 2019-12-09 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include main issues of Georg's comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2921 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 91c5c1244c11c460873fa3a9b5e4cce158213c61

@@
 \Abstract{
-Despite the occurrence of circular data in a wide range of scientific fields,
-the existing methodology for probabilistically forecasting circular data is
+While circular data occur in a wide range of scientific fields,
+the methodology for probabilistically forecasting them is
 highly limited. Most of the existing methods are built on the framework of
@@
 distribution parameters are considered and a fully specified distribution is
-provided for each terminal node. The proposed circular regression trees are
+provided for each terminal node. These circular regression trees are
 easy to interpret, can detect non-additive effects, and select covariates and
@@
 regression trees a more stable and regularized circular regression forest can
-be built which also allows for modeling smooth effects.
+be built, which also allows for modeling smooth effects.
 
@@
 weather variables in meteorology. Fitting a statistical model to this type of
-data requires the incorporation of its specific feature of periodicity, i.e.,
-in case of angular data the fact that values are restricted to an interval such
+data requires the incorporation of its specific feature of periodicity. For example,
+angular data are restricted to an interval such
 as $[0,2\pi)$ with $0$ being equivalent to $2\pi$.
@@
 this strong dependence on appropriate initial values,
-\cite{Mulder+Klugkist:2017} have presented a Bayesian alternative of a
+\cite{Mulder+Klugkist:2017} presented a Bayesian alternative of a
 homoscedastic GLM for circular data. However, apart from potential
@@
 This novel approach to circular regression trees and forests complements the
-literature on tree-based circular modeling. \cite{Lund:2002} has already
+literature on tree-based circular modeling. \cite{Lund:2002} already
 introduced the concept of circular regression trees where binary splits are
@@
 circular covariates, the tree solely considers effects in the mean direction
-for splitting, i.e., only the location parameter $\mu$ is regressed on
+for splitting so that only the location parameter $\mu$ is regressed on
 covariates. Therefore, the introduced distributional approach to circular
@@
 \subsubsection*{Motivating example}
-To provide a first impression of the presented methodology a circular
+To provide a first impression of the presented methodology, a circular
 regression tree is employed for probabilistic wind direction forecasting. Wind
@@
 %%
-Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
-empirical (gray) and fitted von Mises (red) distribution of wind direction in
-each terminal node. Based on the fitted location parameters $\hat \mu$, the
+Figure~\ref{fig:tree_ibk} illustrates the obtained tree structure together with
+the empirical (gray) and fitted von Mises (red) distribution of wind direction
+in each terminal node. Based on the fitted location parameters $\hat \mu$, the
 subgroups can be distinguished into the following wind regimes: (1)~Up-valley
@@
 and forests applied to probabilistic forecasting of wind direction at the
-airports of Innsbruck and Vienna will be presented in this paper. Moreover,
-their performances are benchmarked to alternative commonly used approaches.
+airports of Innsbruck and Vienna will be presented in Section~\ref{sec:wind} of
+the paper. Moreover, their performances are benchmarked to alternative commonly
+used approaches.
 
@@
 introduced in Section~\ref{sec:tree_forest}. After the case study presented in
-\ref{sec:wind}, a comprehensive summary and conclusion is given in
+Section~\ref{sec:wind}, a comprehensive summary and conclusions are given in
 Section~\ref{sec:summary}.
@@
 \label{sec:prob_circ}
-Probabilistic modeling of circular data requires the selection of a proper
-circular probability distribution, i.e., a probability distribution which
-accounts for the periodicity of circular data. Generally, this feature can be
-obtained by ``wrapping'' the probability density function of any continuous
-distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
-wrapped Cauchy distribution or the wrapped normal distribution can be employed
-to model symmetric unimodal circular data. A close approximation to the wrapped
-normal distribution that is mathematically simpler and hence easier to use
-\citep{Fisher:1993} is provided by the von Mises distribution, a purely
-circular distribution which is also known as ``the circular normal
-distribution'' and is a common choice for probabilistic modeling of circular
-data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
-parameter $\kappa > 0$ the density of the von Mises distribution for an
-observation $y \in [0, 2\pi)$ is given by
+Probabilistic modeling of circular data requires the selection of a probability
+distribution which accounts for the periodicity of circular data. Generally,
+this feature can be obtained by ``wrapping'' the probability density function
+of any continuous distribution around the unit circle \citep{Mardia+Jupp:2009}.
+In that way, the wrapped Cauchy distribution or the wrapped normal distribution
+can be employed to model symmetric unimodal circular data. A close
+approximation to the wrapped normal distribution that is mathematically simpler
+and hence easier to use \citep{Fisher:1993} is provided by the von Mises
+distribution, a purely circular distribution which is also known as ``the
+circular normal distribution'' and is a common choice for probabilistic
+modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
+and a concentration parameter $\kappa > 0$ the density of the von Mises
+distribution for an observation $y \in [0, 2\pi)$ is given by
 %%
@@
 
-Once an appropriate distribution family is selected, fitting a probabilistic
-model corresponds to estimating the distribution parameters. Thus, when
-applying the von Mises distribution to fit a probabilistic model $\text{vM}(Y;
-\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ the
-distribution parameters $\mu$ and $\kappa$ are to be estimated. For a learning
-sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by
-maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
-\begin{equation}\label{eq:optim} (\hat{\mu},\hat{\kappa}) =
-\argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) \end{equation}
+To fit a probabilistic model $\text{vM}(Y;
+\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$, the
+distribution parameters $\mu$ and $\kappa$ are to be estimated.
+This can be done by
+maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
+%%
+\begin{equation}\label{eq:optim}
+  (\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) 
+\end{equation}
+%%
 yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
-a fully specified distributional model is fitted to the learning data.
-
-By means of the score function
+a fully specified distributional model is fitted to the learning data $\{y_i\}_{i=1,\ldots,n}$.
+
+The score function
 %%
@@
 %%
-an indicator of deviation and in that way a measure of goodness of fit of the
-model can be obtained for each observation and fitted parameter. Moreover, the
-optimization problem in Equation~\ref{eq:optim} can also be defined by 
+provides a way to obtain a measure of goodness of fit of the model for
+each observation and fitted parameter. Then, the optimization problem in
+Equation~\ref{eq:optim} can alternatively be specified as 
 %%
@@
 
-Figure~\ref{fig:densities} depicts a distributional model fitted to circular
-data employing the von Mises distribution. While the left panel only serves as
-an illustration of the corresponding ``linear'' description restricted to an
-interval of length $2\pi$ but consequently not accounting for periodicity, the
-right panel displays an appropriate representation of a circular quantity on
-the unit circle. Looking at the empirical histogram along with the fitted
-density (red line) and estimated location parameter (red hand) this example
-shows that reasonably well fitting probabilistic models can be established for
-circular data by specifying an appropriate distribution family such as the von
-Mises distribution and estimating its parameters via maximum likelihood.
-However, this procedure considers only the response variable. Including
-covariates can improve any model remarkably as they might provide valuable
-additional information.  Therefore, (distributional) regression models are of
-high interest for all types of data, including circular data.
+Figure~\ref{fig:densities} depicts two representations of the same
+distributional model fitted to circular data employing the von Mises
+distribution: While the left panel only serves as an illustration of the
+corresponding ``linear'' description restricted to an interval of length $2\pi$
+but consequently not accounting for periodicity, the right panel displays an
+appropriate representation of a circular quantity on the unit circle. Looking
+at the empirical histogram along with the fitted density (red line) and
+estimated location parameter (red hand) this example shows that reasonably well
+fitting probabilistic models can be established for circular data by specifying
+an appropriate distribution family such as the von Mises distribution and
+estimating its parameters via maximum likelihood.  However, this procedure
+considers only the response variable. Including covariates can improve any
+model remarkably as they might provide valuable additional information.
+Therefore, (distributional) regression models are of high interest for all
+types of data, including circular data.
 
@@
 \end{equation}
-with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$;
-whereby the additive predictor is transformed to an interval of length $2\pi$
-by a link function $g$. As an appropriate link function typically $g(x) = 2
-\cdot arctan(x)$ is employed, as suggested by \citet{Fisher+Lee:1992}. While
-most of the proposed models estimate a constant concentration parameter, hence,
-assume homoscedasticity, \citet{Fisher+Lee:1992} also developed a
-heteroscedastic version by combining two individual GLMs, each for one of the
-parameters $\mu$ and $\kappa$. This provides a first approach to a fully
-probabilistic regression model for circular data, however, the parameters are
-not regressed simultaneously on covariates as known from the more general
-framework provided by generalized additive models for location, scale, and
-shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular
-additive models incorporate the previously discussed difficulties induced by
-the characteristics of the log-likelihood function and the strongly non-linear
-link function. Referring to additive models in general, it has to be
-considered that a proper model specification can be very challenging,
-particularly for a high number of covariates and no information on possible
-interactions. Moreover, the additive structure might impose a smooth effect
-even if the true underlying effect is an abrupt shift.
+with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$.  A
+link function $g$ transforms the additive predictor to an interval of length
+$2\pi$.  Typically $g(x) = 2\cdot arctan(x)$ is employed, as suggested by
+\citet{Fisher+Lee:1992}.  They also developed a heteroscedastic version by
+combining two individual GLMs, each for one of the parameters $\mu$ and
+$\kappa$. This provides a first approach to a fully probabilistic regression
+model for circular data. However, the parameters are not regressed
+simultaneously on covariates as in the more general framework provided by
+generalized additive models for location, scale, and shape \citep[GAMLSS,
+][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular additive models
+incorporate the previously discussed difficulties induced by the
+characteristics of the log-likelihood function and the strongly non-linear link
+function. Referring to additive models in general, it has to be considered that
+a proper model specification can be very challenging, particularly for a high
+number of covariates and no information on possible interactions. Moreover, the
+additive structure might impose a smooth effect even if the true underlying
+effect is an abrupt shift, which occurs, e.g., in atmospheric wind fields.
 
 An alternative to additive models, which can deal or even avoid the mentioned
-difficulties in circular regression modeling are tree-based models. For the
-particular case of circular data, the methodology of circular regression trees
-and forests is presented in the following section.
+difficulties in circular regression modeling are tree-based models.
 
@@
 
-The main goal of employing tree-based models to circular data is to provide a
-flexible and easy-to-apply modeling approach which accounts for a proper
-representation on the unit circle. Based on ideas of \citet{Lund:2002}, we
-introduce probabilistic circular regression trees and forests employing the von
-Mises distribution. The resulting tree-based models provide a very intuitive
-and data-driven alternative to commonly used regression models for circular
-data.
+Based on ideas of \citet{Lund:2002}, we introduce circular regression trees and
+forests considering splits in all distribution parameters of the von Mises
+distribution and providing a full probabilistic model.  Moreover, the resulting
+tree-based models provide a very intuitive and data-driven alternative to
+commonly used GLMs for circular data.
 
@@
 This is the general idea of regression trees which are combined with
-distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
-unbiased recursive partitioning algorithms MOB
-\citep{Zeileis+Hothorn+Hornik:2008} or CTree
-\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
-trees, a full distributional model is specified in each node of the tree for a
-specific distribution family chosen in advance. Hence, selecting the von Mises
-distribution allows for an application of distributional trees to circular data
-following the tree building algorithm as introduced in Section~2.2 of
-\cite{Schlosser+Hothorn+Stauffer:2019}.
-
-In the applied tree-building algorithm, in each node the crucial decision on
-how and where to split the data is based on model scores which are obtained by
-evaluating the score function~$s$ for the individual observations and parameter
-estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
-two distribution parameters ($\mu$ and $\kappa$) and a data set of $n$
-observations, this yields an $n \times 2$ matrix that can be employed as a
-discrepancy measure, capturing how well each given observation conforms with
-the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, respectively.
-To capture dependence on covariates, the association between the model's scores
-and each available covariate is assessed using either a parameter instability
-test~(MOB) or a permutation test~(CTree). By doing so in each partitioning
-step, the covariate with the highest significant association (i.e., lowest
-significant $p$-value, if any) is selected for splitting the data. The
-corresponding split point is chosen either by optimizing the
+distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019}. Specifying a
+full distributional model in each node of the tree yields a distributional
+regression tree, where selecting the von Mises distribution enables an
+application to circular data. The crucial step of how and where to split the
+data can be accomplished with the unbiased recursive partitioning algorithms
+MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree
+\citep{Hothorn+Hornik+Zeileis:2006}. For this purpose, model scores are
+obtained by evaluating the score function~$s$ for the individual observations
+and parameter estimates (Equation~\ref{eq:scores}). For the von Mises
+distribution with its two distribution parameters ($\mu$ and $\kappa$) and a
+data set of $n$ observations, this yields an $n \times 2$ matrix that can be
+employed as a discrepancy measure, capturing how well each given observation
+conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$,
+respectively.  To capture dependence on covariates, the association between the
+model's scores and each available covariate is assessed using either a
+parameter instability test~(MOB) or a permutation test~(CTree). By doing so in
+each partitioning step, the covariate with the highest significant association
+(i.e., lowest significant $p$-value, if any) is selected for splitting the
+data. The corresponding split point is chosen either by optimizing the
 log-likelihood~(MOB) or a two-sample test statistic~(CTree) over all possible
@@
 significant parameter instabilities or until another stopping criterion is met
-(e.g., subgroup size or tree depth). A more detailed description of the
-applied tree-building algorithm can be found in Appendix~\ref{app:algorithm}.
+(e.g., subgroup size or tree depth). A more detailed description of the applied
+tree-building algorithm can be found in Appendix~\ref{app:algorithm}.
 
@@
 = (z_1, \ldots, z_m)$. Starting at the root node, the tree structure leads the
-observation to a terminal node, depending on the values of the covariates,
-where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for the
-corresponding subset of learning observations. This can also be expressed by
-employing weights which indicate whether the $i$-th learning observation and
-the observation $\bm{z}$ belong to the same terminal node: \begin{equation}
+observation to a terminal node where the parameter pair $(\hat{\mu},
+\hat{\kappa})$ is estimated for the corresponding subset of learning
+observations. This can also be expressed by employing weights which indicate
+whether the $i$-th learning observation and the observation $\bm{z}$ belong to
+the same terminal node: 
+%%
+\begin{equation}
   w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
-\mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), \end{equation} where
-$\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th
-out of $B$ segments partitioning the covariate space in disjoint subsets. In
-that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$
-specifying the predicted von Mises distribution for a given $\bm{z}$ is
-obtained by 
+  \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), 
+\end{equation} 
+%%
+where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the
+$b$-th out of $B$ segments partitioning the covariate space in disjoint
+subsets. Then the estimated parameter pair
+$(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted von Mises
+distribution for a given $\bm{z}$ is obtained by 
 %%
@@
 Therefore, the same parameter pair is estimated for all observations belonging
-to the same terminal node. This facilitates the application as the parameter
+to the same terminal node, which speeds up computation since the parameter
 estimates do not need to be recalculated for each (new) observation via maximum
@@
 consequential strict separation of data into subgroups hinders an adequate
-depiction of smooth effects. Therefore, they might impose abrupt changes even
-in case of rather smooth transitions. This can be avoided by combining an
-ensemble of trees in order to obtain a more stable forest model.
+depiction of smooth effects. They can be included by combining an ensemble of
+trees in order to obtain a regression forest, which also stabilizes the model.
 
@@
 segment of the $t$-th tree. Therefore, similar observations ending up more
-often in the same terminal node have higher weights and in that way a stronger
+often in the same terminal node have higher weights and thus a stronger
 influence in the estimation process.
@@
 interactions are selected automatically allowing for an easy-to-apply
-methodology with a full specification of the von Mises distribution.
+methodology.
 
@@
 directions are of great importance for risk management in various fields such
-as agriculture, energy production or air safety control. For example, in order
-to direct airplanes to a safe landing, precise knowledge of wind direction for
-the next hour(s) at the respective airport is highly desirable and adequate
-prediction methods are required. In this section, a comprehensive case study on
-probabilistic wind direction forecasting is provided for two airports in
-Austria with fundamentally different site characteristics: Innsbruck with a
-rather unusual location for an airport, located at the bottom of a narrow
-Alpine valley, and Vienna located within flat terrain in the East of Austria.
-In addition to the newly introduced circular regression trees and forests, the
-case study employs various alternative approaches to forecast wind direction as
-benchmarks. The study is based on $1$\,h and $3$\,h forecasts employing lagged
-observations in the vicinity of the airports as possible predictor variables.
+as agriculture, energy production or aviation. For example, in order to direct
+airplanes to a safe landing, precise knowledge of wind direction for the next
+hour(s) at the respective airport is highly desirable and adequate prediction
+methods are required. This section exemplifies the use of circular regression
+trees and forests with wind direction forecasts for two Austrian airports --
+one in flat terrain, the other one in mountainous terrain. The results will be
+benchmarked against alternative probabilistic forecasting methods. The study
+is based on $1$\,h and $3$\,h forecasts employing lagged observations in the
+vicinity of the airports as possible predictor variables.
 
@@
   \caption{Overview of the study area for the airports Innsbruck and Vienna. 
-    For Innsbruck, $4$~stations at the airport and $6$~along the
+    For Innsbruck, $4$~stations at the airport and $6$~stations along the
     intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
-    and $13$~within an area of approximately 30\,km$^2$ are used.
+    and $13$~stations within the vicinity of approximately 30\,km are used.
     Elevation data are obtained from the TandDEM-X digital elevation model with a 
@@
 \label{sec:wind:data}
-The circular response variable considered in this case study is a $10$\,min
-mean of wind direction measurements at the airport for Innsbruck and Vienna on
+The circular response variable considered in this case study is a $10$\,min-average
+of wind direction measurements at the airport for Innsbruck and Vienna on
 an hourly temporal resolution. As predictor variables, we use temporal
@@
 measurements at the airport and along the intersecting valleys are used,
-whereas, for Vienna, measurements at the airport and within an area of
-approximately 30\,km$^2$ are used. A topographical overview of the airports and
+whereas, for Vienna, measurements at the airport and within its vicinity of
+approximately 30\,km are used. A topographical overview of the airports and
 their surrounding areas with the station sites employed in this study is
@@
 to December 2018. After first eliminating predictor variables with more than
-5\% missing values and then time points with missing observations, the data set
-consists of 41\,979 time points and 260 covariates for Innsbruck, and of
-38\,985 times points and 494 covariates for Vienna, respectively.
+5\% missing values and then points with missing observations, the data set
+consists of 41\,979 dates and 260 covariates for Innsbruck, and of
+38\,985 dates and 494 covariates for Vienna, respectively.
 
@@
       \hline
-      Temporal information:         & Daytime, day of the year \\
+      Temporal information:         & Time of the day, day of the year \\
       \noalign{\vskip 1mm}   
@@
 on existing approaches used in the meteorological field, while the third is a
-state-of-the-art model to forecast circular response variables. The next
+state-of-the-art GLM to forecast circular response variables. The next
 paragraphs introduce the reference methods and provide the corresponding model
@@
     \newline
-    For simplicity, rather to fit one single climatological model, separate
-    models are estimated rolling over all observation time points, using the same
-    daytime of $31$~days centered around the day of interest over all available
-    years in the training data.  To gain full probabilistic climatological
-    forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
-    the distribution parameters of the von Mises distribution via maximum
-    likelihood estimation. The construction of a climatological reference model was
-    similarly pursued by \citet{Vogel+Knippertz+Fink:2018} and is discussed in a
+    Instead of simply fitting a climatological model over all dates, separate
+    models are estimated rolling over all observation dates, using the same time of
+    $31$~days centered around the day of interest over all available years in the
+    training data.  To gain full probabilistic climatological forecasts, we employ
+    the steps described in Section~\ref{sec:prob_circ} and fit the distribution
+    parameters of the von Mises distribution via maximum likelihood estimation.
+    This approach follows \citet{Vogel+Knippertz+Fink:2018} and is discussed in a
     comprehensive summary on different time-adaptive training schemes in
@@
   \item \emph{Persistency model:}
-    The persistence describes the previous value of a single weather quantity in a time
-    series. Hence, it is another objective measure in the verification of weather
-    forecasts and also often applied as a reference model
+    The persistence describes the previous value of a single weather quantity
+    in a time series. Hence, it is another objective measure in the verification of
+    weather forecasts and also often applied as a reference model
     \citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
     atmospheric conditions, persistence can be a very good estimate, but as soon as
-    the regime changes during the forecast
-    period, forecasts based on persistence may also fail remarkably. Therefore, it
-    heavily depends on the variability of the variable of interest and the length
-    of the forecasting step whether the persistence can provide overall reliable
-    information.
+    the regime changes during the forecast period, forecasts based on persistence
+    may also fail remarkably. Therefore, whether the persistence can provide
+    overall reliable information, heavily depends on the variability of the
+    variable of interest and the length of the forecasting step.
     \newline
@@
     as a reference, we use the Bayesian implementation of
-    \cite{Mulder+Klugkist:2017} which less depends on initial values due to an MCMC
-    sampling algorithm using weakly informative priors.
-    \newline
-    Following \cite{Mulder+Klugkist:2017} the employed model plugs in an
-    appropriate link function~$g$ in order to guarantee response values within an
-    interval of length $2\pi$. As the implementation cannot handle circular covariates,
-    we use the components of the lagged 2-dimensional wind vector $\bm{vec} = (vec_1, vec_2)^{\top}$ and the
-    lagged wind speed $spd$ as predictor variables. The model formula for the
-    location parameter $\mu$ of the von Mises distribution can be written as:
+    \cite{Mulder+Klugkist:2017} which depends less on initial values due to an MCMC
+    sampling algorithm using weakly informative priors.  \newline Following
+    \cite{Mulder+Klugkist:2017} the model uses a link function~$g$ to keep the
+    response values within an interval of length $2\pi$. As the implementation
+    cannot handle circular covariates, we use the components of the lagged
+    2-dimensional wind vector $\bm{vec} = (vec_1, vec_2)^{\top}$ and the lagged
+    wind speed $spd$ as predictor variables. The model formula for the location
+    parameter $\mu$ of the von Mises distribution can be written as:
     %%
@@
     trainings approach as for the climatological model; hence, separate models are
-    estimated over all observation time points, using the same daytime of $31$~days
+    estimated over all observation dates, using the same time of $31$~days
     centered around the day of interest over all available years in the training
@@
     intrinsic automatic variable selection performed in the tree estimation. The
-    tree building is performed by the newly developed R-package \pkg{circtree}
-    employing the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} using a
-    minimal number of 2000 observations in each terminal node serving as a stopping
-    criterion (argument \code{minbucket}).
+    tree is built with the newly developed R-package \pkg{circtree} employing the
+    CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} using a minimal number of
+    2000 observations in each terminal node (argument \code{minbucket}).
     %%
@@
     \pkg{circtree}. Each of these trees is again built by the CTree algorithm on a
-    subsample containing a fraction of $30$~percent of the original learning data.
-    As it can be expected that the lagged response variable is one of the most
-    informative predictor variables for wind direction forecasting, all covariates
-    are included for building each tree which ensures that the lagged response
-    variable is always considered for splitting. This bagging approach can be
-    applied in \pkg{circtree} by setting the argument \code{mtry} to the total
+    subsample containing $30$~percent of the original learning data.  All
+    covariates are included for building each tree which ensures that the lagged
+    response variable is always considered for splitting. This bagging approach can
+    be applied in \pkg{circtree} by setting the argument \code{mtry} to the total
     number of covariates. Since a high number of possible split points leads to
@@
 All scores presented in the next section are computed out-of-sample based on
-five years of data. For the persistency model only time points prior to the
+five years of data. For the persistency model only dates prior to the
 time of interest are used and the validation is performed rolling over all
@@
 years, yielding yearly mean values for every hour per calendar month, with a
-lower score indicating better performance. For both stations at both forecast
-steps, the circular forest overall provides the best predictive performance,
-followed by the circular regression tree and the persistency model; except for
-the $3$\,h~forecasts at Innsbruck where the persistency model is outperformed
-by all others. In comparison to the circular regression tree and forest, for
-both stations and forecast steps, the climatological model and the linear model
-show clearly higher CRPS values and hence a lower predictive performance.
+lower score indicating better performance. The circular forest overall provides
+the best predictive performance, followed by the circular regression tree and
+the persistency model for both stations at both forecast steps; except for the
+$3$\,h~forecasts at Innsbruck where the persistency model is outperformed by
+all others. In comparison to the circular regression tree and forest, for both
+stations and forecast steps, the climatological model and the linear model show
+clearly higher CRPS values and hence a lower predictive performance.
 %%
@@
     climatological, persistency and the linear model as well as for the circular
-    distributional tree and forest.}
+    regression tree and forest.}
   \label{fig:boxplot_crpsraw} 
@@
 a lower performance for longer lead times; this is due to the lower information
-content of minimum 3-hourly instead of 1-hourly lagged response values employed
+content of 3-hourly instead of 1-hourly lagged response values employed
 as covariates in the persistency model. The circular regression tree and
@@
 
-The performance of the novel circular trees and forests are assessed in a real
+The performance of the novel circular trees and forests is assessed in a 
 case study for shortterm probabilistic wind direction forecasting at two

### 2019-12-09 pkg/circtree/inst/wind_paper/talks/CMStatistics2019/abstract/abstract_lang_cmstatistics2019.tex
Initial commit for CMStatistics2019 (abstract, slides, figs)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2923 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4facf55885e7815d09efa4f5955432718f95ad6d

--- 2019-07-02 pkg/disttree/inst/CMStatistics2019/abstract/abstract_lang_cmstatistics2019.tex
@@
 
-\author{Moritz N. Lang\\University of Innsbruck \And
-        Lisa Schlosser\\University of Innsbruck \And
-        Torsten Hothorn\\University of Zurich \And
-        Georg J. Mayr\\University of Innsbruck \And
-        Reto Stauffer\\University of Innsbruck \And
-        Achim Zeileis\\University of Innsbruck}
+\author{Moritz N. Lang\\University of Innsbruck 
+\And Lisa Schlosser\\University of Innsbruck 
+\And Torsten Hothorn\\University of Zurich 
+\AND Georg J. Mayr\\University of Innsbruck 
+\And Reto Stauffer\\University of Innsbruck 
+\And Achim Zeileis\\University of Innsbruck}
+
+\Plainauthor{Moritz N. Lang}
 
@@
 Circular data can be found in a variety of applications and subject areas,
-e.g., hourly crime rate in the social-economics, animal movement direction or
-gene-structure in biology, and wind direction as one of the most important
-weather variables in meteorology. For probabilistic modeling of circular data
-the von Mises distribution is widely used. While most of the already existing
-approaches are built on additive regression models, we propose an adaption of
-regression trees to circular data by employing distributional trees. In
-comparison to the more commonly-used additive models, the resulting
-distributional trees are easy to interpret, can detect non-additive effects,
-and select covariates and their interactions automatically. In addition, as
-natural extension ensembles or forests of such circular trees are introduced
-that can further improve the forecasts by regularizing and stabilizing the
-model. For illustration, short-term probabilistic wind direction forecasts at
-airports are obtained in order to direct airplanes to a a safe landing. To
-benchmark the performance of distributional trees and forests, their predictive
-skill is compared to a persistence model, a climatology model and to a additive
-regression model, employing the continuous ranked probability score. Model code
-for circular distributional trees and forest is available in the R package "disttree".
+e.g., hourly crime rates in the social-economics, animal movement direction in
+ecology, and wind direction as one of the most important weather variables in
+meteorology. For probabilistic modeling of circular data the von Mises
+distribution is widely used. While most existing approaches are built on
+additive regression models, we propose an adaption of regression trees to
+circular data by employing distributional trees. In comparison to the more
+commonly-used additive models, the resulting distributional trees are easy to
+interpret, can detect non-additive effects, and automatically select covariates
+and their interactions. In addition, as a natural extension, ensembles or
+forests of such circular trees are introduced that can further improve the
+forecasts by regularizing and stabilizing the model. For illustration,
+short-term probabilistic wind direction nowcasts at different airports are
+obtained in order to direct airplanes to a safe landing. The predictive skill
+of the novel approaches is benchmarked with an additive regression model plus a
+persistency and a climatology model, employing the circular continuous ranked
+probability score. The proposed methods for circular distributional trees and
+forests are available in the R package ``disttree'' from R-Forge.
 }
 
-\Keywords{}
-
-\Plainkeywords{}
+\Keywords{Distributional Trees; Distributional Forest; Circular Response; Von Mises Distribution}
 

### 2019-12-10 pkg/circtree/inst/wind_paper/draft/draft.Rnw
first revisions based on Trosten's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2924 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8ee43fc23222f72289c8702535f870efc6e7c362

@@
 
-To fit a probabilistic model $\text{vM}(Y;
-\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$, the
+To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response $y$, the
 distribution parameters $\mu$ and $\kappa$ are to be estimated.
@@
 link function $g$ transforms the additive predictor to an interval of length
-$2\pi$.  Typically $g(x) = 2\cdot arctan(x)$ is employed, as suggested by
+$2\pi$.  Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
 \citet{Fisher+Lee:1992}.  They also developed a heteroscedastic version by
@@
 observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
-$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
+$w_i^\text{forest}(\bm{z})$ is obtained that is based on the number of trees in
 which $\bm{z}$ is assigned to the same terminal node as each learning
@@
   (\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu,
-  \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
+  \kappa} \sum_{i=1}^n w_i^\text{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
 \end{equation}
@@
 Therefore, the resulting parameter estimates can smoothly adapt to the given
-covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the
-unweighted full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$
+covariates $\bm{z}$ whereas $w_i^\text{forest}(\bm{z}) = 1$ would correspond to the
+unweighted full-sample estimates and $w_i^\text{forest}(\bm{z}) \in \{0, 1\}$
 corresponds to the abrupt splits from a tree. Thus, circular regression forests
@@
     with $\beta_0$ being a circular intercept, $\beta_\bullet$ the regression
-    coefficients and $g$ the link function $g(x) = 2 \cdot \arctan(x)$. In
+    coefficients and $g$ the link function $g(x) = 2 \cdot \text{arctan}(x)$. In
     addition, a constant concentration parameter $\kappa$ is fitted to the full
@@
 
-\subsection*{CTree}
+%\subsection*{CTree}
 In the following, the testing and splitting strategy is described for the root node of the tree
@@
 
-First, employing the von Mises distribution, a distributional model $\text{vM}(Y; \mu,\kappa)$ 
+First, employing the von Mises distribution, a distributional model $\text{vM}(y; \mu,\kappa)$ 
 is fitted to the learning sample of circular observations $\{y_i\}_{i = 1,\ldots,n}$ as explained 
@@
 is obtained for each parameter and each observation by evaluating the score 
-function $s(\mu, \kappa, Y)$ at the estimated location and concentration 
+function $s(\mu, \kappa, y)$ at the estimated location and concentration 
 parameter $\hat{\mu}$ and $\hat{\kappa}$.
-To detect dependencies between the resulting score values
-\begin{equation}
-s(\hat{\mu}, \hat{\kappa}, y) = 
+To detect dependencies between the resulting scoring matrix
+\begin{equation}
 \begin{pmatrix} 
-s(\hat{\mu}, \hat{\kappa}, y_1)_1 & s(\hat{\mu}, \hat{\kappa}, y_1)_2\\
+s(\hat{\mu}, \hat{\kappa}, y_1)_1 \quad, & s(\hat{\mu}, \hat{\kappa}, y_1)_2\\
 \vdots & \vdots\\
-s(\hat{\mu}, \hat{\kappa}, y_n)_1 & s(\hat{\mu}, \hat{\kappa}, y_n)_2
+s(\hat{\mu}, \hat{\kappa}, y_n)_1 \quad,& s(\hat{\mu}, \hat{\kappa}, y_n)_2
 \end{pmatrix}
 \end{equation}
-and each possible split variable $Z_l \in \{Z_1, \ldots, Z_m\}$ a permutation test is applied. 
-In particular, the hypotheses
-\begin{align}
-H_0^l:  s(\hat{\mu}, \hat{\kappa}, Y) \qquad \bot \qquad Z_l
-\end{align}
+and each possible split variable $z_l \in \{z_1, \ldots, z_m\}$ a permutation test is applied. 
+In particular, the null hypotheses
+of indipendence of each split variable and the scores
+%\begin{align}
+%H_0^l:  s(\hat{\mu}, \hat{\kappa}, Y) \qquad \bot \qquad Z_l
+%\end{align}
 is assessed by employing the multivariate linear statistic
 \begin{equation}
-T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\hat{\mu}, \hat{\kappa}, Y_i)\right)
-\end{equation}
-with $s(\hat{\mu}, \hat{\kappa}, Y_i) \in \mathbb{R}^{1\times 2}$. 
-For a numeric split variable $Z_l$ the transformation function $v_l$
-is simply the identity function $v_l(Z_{li}) = Z_{li}$ such that $T_l \in \mathbb{R}^2$ 
+t_l = vec\left(\sum_{i=1}^n v_l(z_{li}) \cdot s(\hat{\mu}, \hat{\kappa}, y_i)\right)
+\end{equation}
+with $s(\hat{\mu}, \hat{\kappa}, y_i) \in \mathbb{R}^{1\times 2}$. 
+For a numeric split variable $z_l$ the transformation function $v_l$
+is simply the identity function $v_l(z_{li}) = z_{li}$ such that $t_l \in \mathbb{R}^2$ 
 as the ``vec'' operator converts the matrix of dimension $1 \times 2$ into a $2$ column vector.
-If $Z_l$ is a categorical variable with $H$ categories then 
-$v_l(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$, hence, $v_l$ returns a unit vector
-of dimension $H$ where the entry $1$ indicates the category of $Z_{li}$. 
-In this case the ``vec'' operator converts the $H \times 2$ matrix into a $H \cdot 2$ column vector 
-by column-wise combination such that $T_l \in \mathbb{R}^{H \cdot 2}$. 
-If there are any observations with missing values these are not included in the calculation of $T_l$.
-
-To map the observed multivariate linear statistic $t_l$ onto the real line a univariate test statistic~$c$ is employed, for example in a quadratic form
+If $z_l$ is a categorical variable with $h$ categories then 
+$v_l(z_{li}) = (\I(z_{li} = 1), \ldots, \I(z_{li} = h))$, hence, $v_l$ returns a unit vector
+of dimension $h$ where the entry $1$ indicates the category of $z_{li}$. 
+In this case the ``vec'' operator converts the $h \times 2$ matrix into a $h \cdot 2$ column vector 
+by column-wise combination such that $t_l \in \mathbb{R}^{h \cdot 2}$. 
+If there are any observations with missing values these are not included in the calculation of $t_l$.
+
+To map the %observed 
+multivariate linear statistic $t_l$ onto the real line a univariate test statistic~$c$ is employed, for example in a quadratic form
 \begin{equation}
@@
 \end{equation}
-where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance of $T_l$,
-as derived by \cite{Strasser+Weber:1999} and used to standardize $t_l$, and $\Sigma_l^+$ 
+where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance of $t_l$,
+as derived by \cite{Strasser+Weber:1999} and used for standardization, and $\Sigma_l^+$ 
 is the Moore-Penrose inverse of $\Sigma_l$.
@@
 $c(t_{l},\mu_{l},\Sigma_{l})$ indicates a strong discrepancy from the assumption of independence 
-between the scores and the split variable $Z_l$.
+between the scores and the split variable $z_l$.
 Therefore, if any of the Bonferroni-adjusted $p$-values is beneath the selected significance level,
-the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value is selected as split variable, otherwise 
+the partitioning variable $z_{l^\ast}$ with the lowest $p$-value is selected as split variable, otherwise 
 no split is performed. This early stopping induced by the significance level is referred to as 
@@
 To select the best split point within the already chosen split variable, again, a linear test statistic
-is employed. In particular, for a breakpoint $r$ of the variable $Z_{l^{\ast}}$ leading to two subgroups 
+is employed. In particular, for a breakpoint $r$ of the variable $z_{l^{\ast}}$ leading to two subgroups 
 $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ the discrepancy between score functions in the subgroups
@@
 \begin{equation}
-T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\hat{\mu}, \hat{\kappa}, Y_i)
+t_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\hat{\mu}, \hat{\kappa}, y_i)
 \end{equation}

### 2019-12-10 pkg/circtree/inst/wind_paper/draft/draft.Rnw
first revisions based on Trosten's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2925 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 00f1e5f758c41d667125d36382be63e19e674777
revisions based on Torsten's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2926 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d6ff57627a0b0b179d6b0f115eb44c3af77d4d86

@@
     cannot handle circular covariates, we use the components of the lagged
-    2-dimensional wind vector $\bm{vec} = (vec_1, vec_2)^{\top}$ and the lagged
+    2-dimensional wind vector~$(u, v)^{\top}$ and the lagged
     wind speed $spd$ as predictor variables. The model formula for the location
@@
     \begin{align}
-      \mu = \beta_0 + g(\beta_1 \cdot vec_{1} + \beta_2 \cdot vec_{2} + \beta_3 \cdot spd)
+      \mu = \beta_0 + g(\beta_1 \cdot u + \beta_2 \cdot v + \beta_3 \cdot spd)
     \end{align}
@@
     captured by the concentration parameter, we use the same time-adaptive
-    trainings approach as for the climatological model; hence, separate models are
+    training approach as for the climatological model; hence, separate models are
     estimated over all observation dates, using the same time of $31$~days
@@
 In particular, the null hypotheses
-of indipendence of each split variable and the scores
+of independence of each split variable and the scores
 %\begin{align}

### 2019-12-12 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2928 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cb5a7e31fea719a1eb8678bcbefde44289bd01d9

@@
     intrinsic automatic variable selection performed in the tree estimation. The
-    tree is built with the newly developed R-package \pkg{circtree} employing the
+    tree is built with the newly developed \textsf{R} package \pkg{circtree} employing the
     CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} using a minimal number of
@@
     Section~\ref{sec:circforest}, the circular forest used in this study is
-    constructed based on $100$~individual trees employing the R-package
+    constructed based on $100$~individual trees employing the \textsf{R} package
     \pkg{circtree}. Each of these trees is again built by the CTree algorithm on a
@@
 The corresponding implementation of the proposed methodology for circular
-trees and forests is provided in the \textsf{R}
-package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}).
+trees and forests is provided in the \textsf{R} package \textbf{circtree} 
+(version~\Sexpr{packageVersion("circtree")}).
 % The proposed circular tree and forest models are implemented in the \textsf{R}

### 2019-12-13 pkg/circtree/inst/wind_paper/draft/draft.Rnw
improvements in abstract, intro, and references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2931 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f32c56538a675dde962ebdb688e2bb609eee9486

@@
 While circular data occur in a wide range of scientific fields,
-the methodology for probabilistically forecasting them is
-highly limited. Most of the existing methods are built on the framework of
-generalized linear and additive models, which can incorporate difficulties in
-both the optimization and the interpretation. As a very intuitive alternative
-to additive circular regression models, we extend previous ideas of circular
-trees and introduce probabilistic circular regression trees and forests
-employing the von Mises distribution. The resulting tree-based models simplify
-the estimation process by separating the data into more homogeneous subgroups
-and fitting a simple intercept-only model for each resulting subset. By
-incorporating the framework of distributional modeling, splits in all
-distribution parameters are considered and a fully specified distribution is
-provided for each terminal node. These circular regression trees are
-easy to interpret, can detect non-additive effects, and select covariates and
-their interactions automatically. By combining an ensemble of circular
-regression trees a more stable and regularized circular regression forest can
-be built, which also allows for modeling smooth effects.
-
-To evaluate the proposed circular regression trees and forests, a case study on
-probabilistic forecasting of wind direction for two airports in Austria is
-presented. In addition to the tree-based models, alternative common approaches
-are used as a benchmark in order to assess the predictive performances of the
-novel methodology of circular regression trees and forests. 
+the methodology for distributional modeling and probabilistic forecasting of circular
+response variables is rather limited. Most of the existing methods are built on the framework of
+generalized linear and additive models, which are often challenging to optimize
+and to interpret. Therefore, we suggest circular regression trees and random forests
+as an inuitive alternative approach that is relatively easy to fit. Building on
+previous ideas for trees modeling circular means, we suggest a distributional
+approach for both trees and forests yielding probabilistic forecasts based on the
+von Mises distribution. The resulting tree-based models simplify
+the estimation process by using the available covariates for partitioning the
+data into sufficiently homogeneous subgroups so that a simple von Mises distribution
+without further covariates can be fitted to the circular response in each subgroup.
+These circular regression trees are straightforward to interpret, can capture
+nonlinear effects and interactions, and automatically select the relevant
+covariates that are associated with either location and/or scale changes in the
+von Mises distribution. Combining an ensemble of circular regression trees
+to a circular regression forest yields a local adaptive likelihood estimator
+for the von Mises distribution that can regularize and smooth the covariate effects.
+The new methods are evaluated in a case study on probabilistic wind direction
+forecasting at two Austrian airports, considering other common approaches as
+a benchmark.
 }
 
-\Keywords{Distributional Trees; Random Forests; Circular Data; von Mises Distribution}
+\Keywords{circular data, regression trees, random forests, distributional regression, von Mises distribution}
 
@@
 
+\begin{document}
+
 % -------------------------------------------------------------------
@@
 % -------------------------------------------------------------------
-\begin{document}
 \section{Introduction}
 \label{sec:introduction}
+
 Circular data can be found in a variety of applications and subject areas,
-e.g., hourly crime rate in the social-economics, animal movement direction or
+e.g., hourly crime rate in socio-economics, animal movement direction or
 gene-structure in biology, and wind direction as one of the most important
@@
 angular data are restricted to an interval such
-as $[0,2\pi)$ with $0$ being equivalent to $2\pi$.
+as $[0,2\pi)$ with $0$ being equivalent~to~$2\pi$.
+
+\subsection[Circular regression: Conditional mean vs. distributional models]{Circular regression: Conditional mean vs.\ distributional models}
 
@@
 of the first regression models with a circular response variable and linear
-covariates was presented by \cite{Gould:1969} where the circular mean, i.e.,
-the location parameter of the von Mises distribution, is predicted by a linear
-combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
+covariates was presented by \cite{Gould:1969} where the circular mean
+%% , i.e., the location parameter of the von Mises distribution, 
+is predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
 plugging in a link function transforming the linear predictor to a restricted
-interval of length $2\pi$. This generalized linear model (GLM)-type approach
+interval of length $2\pi$. This generalized linear model (GLM) type approach
 was further extended by \cite{Fisher+Lee:1992} and subsequently by
@@
 parameter constant. Additionally, they developed a combined heteroscedastic
-version where the location and scale parameter are alternately reestimated
-conditional on respective sets of covariates until convergence. While all of
-these models are built on well elaborated theory the employed methodology has
-shown to be very challenging in its application, mainly due to the complexity
+or distributional version with alternating reestimation of the location and scale parameters
+conditional on the respective sets of covariates until convergence. While all of
+these models are built on well-elaborated theory, their application in practice
+remains very challenging, mainly due to the complexity
 encountered in optimizing the corresponding log-likelihood function which is
-not globally concave. Therefore, highly informative starting values are
-crucial for these circular GLMs to converge
+not globally concave. Therefore, highly-informative starting values are
+crucial for such circular GLMs to converge
 \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
 this strong dependence on appropriate initial values,
-\cite{Mulder+Klugkist:2017} presented a Bayesian alternative of a
+\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a
 homoscedastic GLM for circular data. However, apart from potential
-difficulties in the optimization procedure of circular GLMs, additive effects
-and their interpretation are often not intuitive for circular responses due to
-the highly non-linear link function and the not-straightforward representation
-of smooth transitions on the unit circle. For example, the same rotation can be
+difficulties in the optimization procedure of circular GLMs, the interpretation
+of the underlying additive effects is often challenging as well because the link function
+is highly nonlinear and the representation of smooth transitions on the unit circle
+is not straightforward. For example, the same rotation can be
 obtained in either positive or negative direction on the circle leading to an
@@
 trees and forests avoid the discussed difficulties of circular GLMs by
-separating the data into more homogeneous subgroups and fitting a simple
-intercept-only model for each resulting subset. Therefore, no covariates and,
-thus, no link functions must be employed in the estimation of the
-group-specific distribution parameters which also simplifies the optimization
-process. By incorporating the framework of distributional modeling, we allow
-for detecting splits in all distribution parameters and provide a fully
-specified circular response distribution for each terminal node offering a wide
+using the available covariates for partitioning the
+data into sufficiently homogeneous subgroups so that a simple von Mises distribution
+without further covariates can be fitted to the circular response in each subgroup.
+This obviates the need for a link function or for iterating between models for the
+separate distribution parameters. By leveraging the distributional modeling approach, the
+trees can automatically detect and capture differences in both
+distribution parameters, providing a fully
+specified circular response distribution in each terminal node offering a wide
 range of statistical inference. In addition, the employed tree structure
@@
 smooth changes. Furthermore, covariates and their possible interactions do not
-need to be specified in advance as they are selected automatically.
+need to be specified in advance as they are selected automatically in the
+recursive partitioning algorithm.
 
@@
 literature on tree-based circular modeling. \cite{Lund:2002} already
-introduced the concept of circular regression trees where binary splits are
-made based on an angular distance measure employed to evaluate node
-homogeneity. While this approach does not only allow for linear, but also for
-circular covariates, the tree solely considers effects in the mean direction
-for splitting so that only the location parameter $\mu$ is regressed on
-covariates. Therefore, the introduced distributional approach to circular
-regression trees extends the concept of previous circular trees by considering
-splits in all distribution parameters and providing a full probabilistic model.
+introduced a circular regression tree algorithms where binary splits are
+made based on an angular distance measure capturing node
+homogeneity. However, this only models changes in the conditional mean
+but not the conditional variance or full probabilistic distribution.
+In contrast, here we introduce a distributional tree approach that considers
+splits in all distribution parameters and yields a full probabilistic predictive model.
 In addition, as a natural extension, ensembles or forests of circular
-regression trees are presented in this study.
-
-\subsubsection*{Motivating example}
-To provide a first impression of the presented methodology, a circular
-regression tree is employed for probabilistic wind direction forecasting. Wind
-direction is a classical circular quantity and accurate forecasts are of great
-importance for decision-making processes, e.g., in air traffic management as
-considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
-for 1-hourly forecasts at the airport of Innsbruck which lies at the bottom of
-a narrow valley within the European Alps. Topography channels the wind along
-the west-east valley axis or along a tributary valley intersecting from the
-south. Hence, pressure gradients to which valley wind regimes react are
-considered as covariates along with other meteorological measurements (lagged
-by one hour) and their derivatives, such as wind direction and wind speed at
-the airport or spatial and temporal differences.
-%%
-\begin{figure}[p!] \centering
+regression trees are presented in this study, showing that these can further
+improve predictive power.
+
+\begin{figure}[p!]
+\centering
 \includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
-\caption{Fitted tree based on the von Mises distribution for 1-hourly wind
+\caption{\label{fig:tree_ibk} Fitted tree based on the von Mises distribution for 1-hourly wind
 direction forecasts at the airport of Innsbruck. In each terminal node the
@@
 direction is defined on the scale $[0,360]$ degree and increases clockwise from
-North ($0$ degree). } \label{fig:tree_ibk} \end{figure}
-%%
-Figure~\ref{fig:tree_ibk} illustrates the obtained tree structure together with
-the empirical (gray) and fitted von Mises (red) distribution of wind direction
+North ($0$ degree).}
+\end{figure}
+
+\subsection{Motivating example}
+
+To provide a first impression of the presented methodology, a circular
+regression tree is employed for probabilistic wind direction forecasting. Wind
+direction is a classical circular quantity and accurate forecasts are of great
+importance for decision-making processes, e.g., in air traffic management as
+considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
+for 1-hourly forecasts at Innsbruck Airport, located at the bottom of
+a narrow valley within the European Alps. Topography channels the wind along
+the west-east valley axis or along a tributary valley intersecting from the
+south. Hence, pressure gradients to which valley wind regimes react are
+considered as covariates along with other meteorological measurements (lagged
+by one hour) and their derivatives, such as wind direction and wind speed at
+the airport as well as spatial and temporal differences.
+
+Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with
+the empirical (gray) and fitted von Mises (red) wind direction distribution
 in each terminal node. Based on the fitted location parameters $\hat \mu$, the
@@
 vs.\ low precision. A more extensive case study of circular regression trees
-and forests applied to probabilistic forecasting of wind direction at the
-airports of Innsbruck and Vienna will be presented in Section~\ref{sec:wind} of
-the paper. Moreover, their performances are benchmarked to alternative commonly
-used approaches.
+and forests applied to probabilistic wind direction forecasting at
+Innsbruck Airport and Vienna International Airport will be presented in Section~\ref{sec:wind},
+along with a benchmark against commonly-used alternative approaches.
 
@@
 this feature can be obtained by ``wrapping'' the probability density function
-of any continuous distribution around the unit circle \citep{Mardia+Jupp:2009}.
+of any continuous distribution around the unit circle \citep{Mardia+Jupp:1999}.
 In that way, the wrapped Cauchy distribution or the wrapped normal distribution
@@
 approximation to the wrapped normal distribution that is mathematically simpler
-and hence easier to use \citep{Fisher:1993} is provided by the von Mises
-distribution, a purely circular distribution which is also known as ``the
+and hence easier to use is provided by the von Mises
+distribution \citep{Fisher:1993}, a purely circular distribution which is also known as ``the
 circular normal distribution'' and is a common choice for probabilistic
@@
 incorporate the previously discussed difficulties induced by the
-characteristics of the log-likelihood function and the strongly non-linear link
+characteristics of the log-likelihood function and the strongly nonlinear link
 function. Referring to additive models in general, it has to be considered that
@@
     Traditional approaches to forecast circular response variables are often
-    based on circular GLM-typed models \citep{Fisher:1993}. As discussed in
+    based on circular GLM-type models \citep{Fisher:1993}. As discussed in
     Section~\ref{sec:introduction}, circular regression models often experience the
@@
 
-To conclude, in general the tree structure can capture non-linear changes,
+To conclude, in general the tree structure can capture nonlinear changes,
 shifts, and potential interactions in covariates without prespecification of

### 2019-12-13 pkg/circtree/inst/wind_paper/draft/draft.Rnw
some touch-ups in modeling sections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2932 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 58cba3bd614a9b159e5111708f6455fb2d05b45d

@@
 
-\title{Circular Regression Trees and Forests with an Application to
-Probabilistic Wind Direction Forecasting}
+\title{Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting}
 \Shorttitle{Circular Regression Trees and Forests}
@@
    \And Achim Zeileis\\Universit\"at Innsbruck}
-\Plainauthor{M. N. Lang, L. Schlosser, T. Hothorn, 
-G. J. Mayr, R. Stauffer, A. Zeileis}
+\Plainauthor{Lang, Schlosser, Hothorn, Mayr, Stauffer, Zeileis}
 
@@
 and forests applied to probabilistic wind direction forecasting at
-Innsbruck Airport and Vienna International Airport will be presented in Section~\ref{sec:wind},
+Innsbruck Airport and Vienna International Airport is presented in Section~\ref{sec:wind},
 along with a benchmark against commonly-used alternative approaches.
@@
 probabilistic circular modeling introducing the von Mises distribution, and
-circular regression models will be discussed in Section~\ref{sec:prob_circ}.
+circular regression models are discussed in Section~\ref{sec:prob_circ}.
 The methodology of circular regression trees and forests and their features are
@@
 where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
-$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and
+$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} or
 \citealt{Ley+Verdebout:2017} for a more detailed overview).
 
-The corresponding log-likelihood function is defined as
-%%
-\begin{equation}
-  \begin{aligned}
-    \ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
-    &= -\log(2 \pi I_0(\kappa)) + \kappa \cos(y - \mu).\label{equ:vM:loglik}
-  \end{aligned}
-\end{equation}
-%%
-
-To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response $y$, the
-distribution parameters $\mu$ and $\kappa$ are to be estimated.
-This can be done by
-maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
-%%
-\begin{equation}\label{eq:optim}
-  (\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) 
-\end{equation}
-%%
-yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
-a fully specified distributional model is fitted to the learning data $\{y_i\}_{i=1,\ldots,n}$.
-
-The score function
-%%
-\begin{equation}
-  \label{eq:scores}
-  \begin{aligned}
-    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
-    \smallskip &,
-    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
-    \\
-    &= \bigg(\kappa \sin(y-\mu)
-    \smallskip &,
-    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
-  \end{aligned}
-\end{equation}
-%%
-provides a way to obtain a measure of goodness of fit of the model for
-each observation and fitted parameter. Then, the optimization problem in
-Equation~\ref{eq:optim} can alternatively be specified as 
-%%
-\begin{equation}
-  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
-\end{equation}
-%%
-%%
 \begin{figure}[t]
@@
   \end{minipage}%
-  \caption{Illustration of a distributional model fitted to circular data by
-    maximum likelihood estimation employing the von Mises distribution. Left
-    panel: linearized illustration restricted to the interval $[0, 2\pi)$. Right
-    panel: circular illustration. In both panels the empirical histogram (gray)
+  \caption{\label{fig:densities} Illustration of a von Mises model for circular data in
+    the interval $[0, 2\pi)$ fitted by maximum likelihood. Left: Linearized
+    scale. Right: Circular scale. In both panels the empirical histogram (gray bars)
     and fitted density (red line) are depicted along with the estimated location
     parameter (red hand).}
-  \label{fig:densities} 
 \end{figure}
-%%
-
-Figure~\ref{fig:densities} depicts two representations of the same
-distributional model fitted to circular data employing the von Mises
-distribution: While the left panel only serves as an illustration of the
-corresponding ``linear'' description restricted to an interval of length $2\pi$
-but consequently not accounting for periodicity, the right panel displays an
-appropriate representation of a circular quantity on the unit circle. Looking
-at the empirical histogram along with the fitted density (red line) and
-estimated location parameter (red hand) this example shows that reasonably well
-fitting probabilistic models can be established for circular data by specifying
-an appropriate distribution family such as the von Mises distribution and
-estimating its parameters via maximum likelihood.  However, this procedure
-considers only the response variable. Including covariates can improve any
-model remarkably as they might provide valuable additional information.
-Therefore, (distributional) regression models are of high interest for all
-types of data, including circular data.
-
-In most generalized linear or additive approaches to circular regression
-modeling the location parameter $\mu$ is regressed on covariates $\mathbf{z}$
-in the form of 
-\begin{equation}
-\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
-\end{equation}
-with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$.  A
-link function $g$ transforms the additive predictor to an interval of length
+
+The corresponding log-likelihood function is defined as
+%%
+\begin{equation}
+    \ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))
+    = -\log(2 \pi I_0(\kappa)) + \kappa \cos(y - \mu).\label{equ:vM:loglik}
+\end{equation}
+%%
+To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response $y$, the
+distribution parameters $\mu$ and $\kappa$ need to be estimated.
+This can be done by
+maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
+%%
+\begin{equation}\label{eq:optim}
+  (\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) 
+\end{equation}
+%%
+yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
+a fully specified distributional model is fitted to the learning data $\{y_i\}_{i=1,\ldots,n}$.
+
+The score function
+%%
+\begin{equation}
+  \label{eq:scores}
+  \begin{aligned}
+    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
+    \smallskip &,
+    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
+    \\
+    &= \bigg(\kappa \sin(y-\mu)
+    \smallskip &,
+    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
+  \end{aligned}
+\end{equation}
+%%
+provides a way to obtain a measure of goodness of fit of the model for
+each observation and fitted parameter. Then, the optimization problem in
+Equation~\ref{eq:optim} can alternatively be specified as 
+%%
+\begin{equation}
+  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
+\end{equation}
+%%
+Figure~\ref{fig:densities} depicts von Mises model for circular data in
+$[0, 2 \pi)$ fitted by maximum likelihood, either using a linearized
+(left) or circular (right) scale. In both cases,
+the empirical histogram (gray bars) is shown along with the fitted density (red line) and
+estimated location parameter (red hand). However, this distributional model
+only considers the circular response variable but no covariate. Of course,
+including covariates is of interest in a regression setup for forecasting.
+
+In most generalized linear model (GLM) or generalized additive model (GAM)
+approaches to circular regression the location parameter $\mu$ depends on covariates $\mathbf{z}$
+through a link function $g(\cdot)$, circular intercept $\mu_0$ and
+coefficient vector $\mathbf{\beta}$:
+\begin{equation}
+\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right).
+\end{equation}
+The link function transforms the additive predictor to an interval of length
 $2\pi$.  Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
@@
 $\kappa$. This provides a first approach to a fully probabilistic regression
-model for circular data. However, the parameters are not regressed
+model for circular data, albeit the parameters are not regressed
 simultaneously on covariates as in the more general framework provided by
 generalized additive models for location, scale, and shape \citep[GAMLSS,
-][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular additive models
-incorporate the previously discussed difficulties induced by the
+][]{Rigby+Stasinopoulos:2005}. Nevertheless, other circular additive models
+share the previously discussed difficulties induced by the
 characteristics of the log-likelihood function and the strongly nonlinear link
@@
 
-An alternative to additive models, which can deal or even avoid the mentioned
-difficulties in circular regression modeling are tree-based models.
+In contrast, the tree-based circular regression models proposed in the next sections
+largely avoid the problems above by employing recursive partitioning in combination
+with local adaptive likelihood estimation.
 
@@
 
-Based on ideas of \citet{Lund:2002}, we introduce circular regression trees and
+Starting out from the ideas of \citet{Lund:2002}, we introduce circular regression trees and
 forests considering splits in all distribution parameters of the von Mises
@@
 Fitting a global model to a full data set can be very challenging, particularly
-for complex data with strong variations. Therefore, separating the data set
+for complex data with substantial variations. Therefore, separating the data set
 into more homogeneous subgroups based on covariates before fitting a local
-model in each of these subgroups allows to capture (possibly) group specific
-effects more precisely and hence can result in an overall better fitting model.
+model in each of these subgroups allows to capture (potential) group-specific
+effects more precisely and hence can result in an overall better-fitting model.
 This is the general idea of regression trees which are combined with
@@
 \citep{Hothorn+Hornik+Zeileis:2006}. For this purpose, model scores are
-obtained by evaluating the score function~$s$ for the individual observations
-and parameter estimates (Equation~\ref{eq:scores}). For the von Mises
+obtained by evaluating the score function~$s(\cdot)$ for each individual observation
+at the parameter estimates (Equation~\ref{eq:scores}). For the von Mises
 distribution with its two distribution parameters ($\mu$ and $\kappa$) and a
@@
   w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
-  \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), 
-\end{equation} 
-%%
-where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the
+  \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)).
+\end{equation} 
+%%
+Here, $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the
 $b$-th out of $B$ segments partitioning the covariate space in disjoint
@@
 $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted von Mises
-distribution for a given $\bm{z}$ is obtained by 
+distribution for a given $\bm{z}$ is obtained by a weighted maximum likelihood estimator:
 %%
@@
 \citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
-Hothorn+Zeileis:2017}. More specifically, for each possibly new
+Hothorn+Zeileis:2017}. More specifically, for each (possibly new)
 observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
@@
 often in the same terminal node have higher weights and thus a stronger
-influence in the estimation process.
+influence in the weighted maximum likelihood estimation.
 
@@
 Therefore, the resulting parameter estimates can smoothly adapt to the given
-covariates $\bm{z}$ whereas $w_i^\text{forest}(\bm{z}) = 1$ would correspond to the
-unweighted full-sample estimates and $w_i^\text{forest}(\bm{z}) \in \{0, 1\}$
-corresponds to the abrupt splits from a tree. Thus, circular regression forests
+covariates $\bm{z}$ whereas $w_i(\bm{z}) = 1$ would correspond to the
+unweighted full-sample estimates and $w_i(\bm{z}) \in \{0, 1\}$
+corresponds to the subgroup selection from a tree. Thus, circular regression forests
 can capture both smooth and abrupt changes, while covariates and possible
-interactions are selected automatically allowing for an easy-to-apply
-methodology.
+interactions are selected automatically and do not explicitly need to be
+specified beforehand.
 
@@
 trees and forests with wind direction forecasts for two Austrian airports --
-one in flat terrain, the other one in mountainous terrain. The results will be
+one in flat terrain, the other one in mountainous terrain. The results are
 benchmarked against alternative probabilistic forecasting methods. The study
@@
   \includegraphics{_plot_circforest_stationmap_combined.pdf}
-  \caption{Overview of the study area for the airports Innsbruck and Vienna. 
+  \caption{Overview of the study area for Innsbruck Airports (left) and Vienna International Airport (right). 
     For Innsbruck, $4$~stations at the airport and $6$~stations along the
@@
 The circular response variable considered in this case study is a $10$\,min-average
-of wind direction measurements at the airport for Innsbruck and Vienna on
+of wind direction measurements at Innsbruck Airport (INN) and Vienna International Airport (VIE) on
 an hourly temporal resolution. As predictor variables, we use temporal

### 2019-12-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
some touch-ups in Sections 4 and 5
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2933 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e596d25751143dde1f4ed92025bbd31262fd447e

@@
 directions are of great importance for risk management in various fields such
-as agriculture, energy production or aviation. For example, in order to direct
+as agriculture, energy production, or aviation. For example, in order to direct
 airplanes to a safe landing, precise knowledge of wind direction for the next
@@
 of wind direction measurements at Innsbruck Airport (INN) and Vienna International Airport (VIE) on
-an hourly temporal resolution. As predictor variables, we use temporal
-information and $1$-hourly resolved $10$\,min mean observations of various
-meteorological quantities such as wind direction, wind speed, temperature, air
+an hourly temporal resolution. Temporal information and $1$-hourly resolved $10$\,min mean observations of various
+meteorological quantities are used as predictor variables, including wind direction, wind speed, temperature, air
 pressure and humidity, all lagged by one or three hours according to the
@@
 whereas, for Vienna, measurements at the airport and within its vicinity of
-approximately 30\,km are used. A topographical overview of the airports and
-their surrounding areas with the station sites employed in this study is
-provided in Figure~\ref{fig:overview}. In addition, we use derived quantities
+approximately 30\,km are used. Figure~\ref{fig:overview} provides a topographical
+overview of the airports and their surrounding areas with the station sites employed
+in this study. In addition, we use derived quantities
 such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
@@
 
-The data used in this study consists of five subsequent years from January 2015
+The data used in this study consists of four subsequent years from January 2015
 to December 2018. After first eliminating predictor variables with more than
-5\% missing values and then points with missing observations, the data set
-consists of 41\,979 dates and 260 covariates for Innsbruck, and of
-38\,985 dates and 494 covariates for Vienna, respectively.
+5\% missing values and then time points with any missing observations, the data set
+consists of 41\,979 time points and 260 covariates for Innsbruck, and of
+38\,985 time points and 494 covariates for Vienna, respectively.
 
@@
 on existing approaches used in the meteorological field, while the third is a
-state-of-the-art GLM to forecast circular response variables. The next
-paragraphs introduce the reference methods and provide the corresponding model
-specifications used in this study for all considered models including the
-circular regression tree and forest.
+state-of-the-art GLM-type model to forecast circular response variables.
 
@@
     Accurate knowledge of weather quantities' climatologies can be important for a
-    wide range of applications and climatologies are often also used as a baseline for the
-    validation of newly developed forecasting systems
-    \citep{Simon+Umlauf+Zeileis:2017, Stauffer+Mayr+Messner:2017}. Usually, the performance of
-    climatological models strongly depends on the temporal dependency of the
-    quantity itself, either on a daily or yearly basis. Wind direction at a single
-    location is often quite unstable and only shows a daily cycle or an annual seasonality
-    in case of predominant local wind systems like the ones at Innsbruck introduced
-    in the analysis of Figure~\ref{fig:tree_ibk}. Hence, the climatological model
-    in this study provides a very simple implementation of a baseline model, which
-    might experience serious performance losses in case of random temporal
-    variabilities or abrupt changes in the employed response variable wind
-    direction.
+    wide range of applications. While forecasts based on climatologies, by
+    construction, do not adapt to the current weather situation they are still a
+    useful baseline for the validation of newly developed forecasting systems
+    \citep{Simon+Umlauf+Zeileis:2017, Stauffer+Mayr+Messner:2017}. 
     \newline
-    Instead of simply fitting a climatological model over all dates, separate
-    models are estimated rolling over all observation dates, using the same time of
-    $31$~days centered around the day of interest over all available years in the
-    training data.  To gain full probabilistic climatological forecasts, we employ
-    the steps described in Section~\ref{sec:prob_circ} and fit the distribution
-    parameters of the von Mises distribution via maximum likelihood estimation.
-    This approach follows \citet{Vogel+Knippertz+Fink:2018} and is discussed in a
-    comprehensive summary on different time-adaptive training schemes in
+    Specifically, the climatology employed in the following uses all observations
+    at the same time (to adapt to daily cycles) in a window of $31$~days centered
+    around the day of interest (to adapt to seasonal cycles)
+    in all available years in the sample. Based on these observations a probabilistic
+    model is obtained by maximum likelihood estimation as described in
+    Section~\ref{sec:prob_circ}. This approach follows \citet{Vogel+Knippertz+Fink:2018}
+    and is discussed in a comprehensive summary on different time-adaptive training schemes in
     \citet{Lang+Lerch+Mayr:2019}.
@@
     The persistence describes the previous value of a single weather quantity
-    in a time series. Hence, it is another objective measure in the verification of
-    weather forecasts and also often applied as a reference model
-    \citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
-    atmospheric conditions, persistence can be a very good estimate, but as soon as
-    the regime changes during the forecast period, forecasts based on persistence
-    may also fail remarkably. Therefore, whether the persistence can provide
-    overall reliable information, heavily depends on the variability of the
-    variable of interest and the length of the forecasting step.
+    in a time series. Like the climatology it is a very basic prediction model that
+    is often applied as a baseline reference in weather forecasts
+    \citep{noaasnationalweatherservice:2019}. Especially in nowcasting tasks with very short
+    forecasting steps, the persistence can provide very good estimates.
     \newline
@@
     high computational costs, the covariates are binned in a maximum of $50$
-    classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}). Contrary to a
+    classes (argument \code{nmax = c(yx = Inf, z = 50)}). Contrary to a
     single-tree model, forests usually consist of very large trees as they are not
@@
 Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
-at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
+at forecast steps 1\,h and 3\,h for Innsbruck (Panels~a,\,c) and
 Vienna (Panels~b,\,d). The scores are aggregated over the five validation
@@
     predictive von~Mises distribution for $+1$\,h and $+3$\,h forecasts at the
-    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
+    Innsbruck Airport and Vienna International Airport. Each box-and-whisker contains $24$~hourly scores
     for each of the $12$~months averaged over the $5$~validation years which
@@
 covariates are solely considered for splitting and group-specific models are
-fitted as intercept-only models. In addition, by specifying the von Mises
+fitted without further covariates. In addition, by specifying the von Mises
 distribution for each node and allowing for splits in both distribution
@@
 probabilistic climatology and persistency models, as well as a state-of-the-art
-circular GLM are evaluated based on proper scoring rules. In summary, the
+circular GLM-type model are evaluated based on proper scoring rules. In summary, the
 circular trees and forests have the highest predictive performance in this

### 2019-12-16 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small corrections based on Reto's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2934 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 480646ef57b78d0bdcb9bb68dd1961b4b47dd54a

@@
 and to interpret. Therefore, we suggest circular regression trees and random forests
-as an inuitive alternative approach that is relatively easy to fit. Building on
+as an intuitive alternative approach that is relatively easy to fit. Building on
 previous ideas for trees modeling circular means, we suggest a distributional
@@
 \Address{
-  Moritz N. Lang, Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
+  Moritz N. Lang, Lisa Schlosser, Achim Zeileis \\
   Universit\"at Innsbruck \\
@@
   6020 Innsbruck, Austria \\
-  E-mail: \email{Moritz.Lang@uibk.ac.at}, \email{Lisa.Schlosser@uibk.ac.at},\\
-  \email{Reto.Stauffer@uibk.ac.at},\email{Achim.Zeileis@R-project.org} \\
+  E-mail: \email{Moritz.Lang@uibk.ac.at}, \\
+  \phantom{E-mail: }\email{Lisa.Schlosser@uibk.ac.at},\\
+  \phantom{E-mail: }\email{Achim.Zeileis@R-project.org} \\
   URL: \url{https://www.uibk.ac.at/statistics/personal/moritz-lang/},\\
   \phantom{URL: }\url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
-  \phantom{URL: }\url{https://retostauffer.org/},\\
   \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
+
+  Reto Stauffer \\
+  Universit\"at Innsbruck \\
+  Digital Science Center and \\ 
+  Department of Statistics \\
+  Faculty of Economics and Statistics \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
+  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
+  URL: \url{https://retostauffer.org/}\\
 
@@
 data into sufficiently homogeneous subgroups so that a simple von Mises distribution
-without further covariates can be fitted to the circular response in each subgroup.
+without further covariates can be fitted to the circular response in each of these subgroups.
 This obviates the need for a link function or for iterating between models for the
@@
 literature on tree-based circular modeling. \cite{Lund:2002} already
-introduced a circular regression tree algorithms where binary splits are
+introduced a circular regression tree algorithm where binary splits are
 made based on an angular distance measure capturing node
@@
 but not the conditional variance or full probabilistic distribution.
-In contrast, here we introduce a distributional tree approach that considers
+In contrast, we introduce a distributional tree approach that considers
 splits in all distribution parameters and yields a full probabilistic predictive model.
@@
 with the estimated location parameter (red hand). The covariates selected for
-splitting are wind direction (degree), wind speed ($\text{ms}^{-1}$), and
+splitting are wind direction (meteorological degree), wind speed ($\text{ms}^{-1}$), and
 pressure gradients ($\text{dpressure; hPa}$) west, east and south of the
@@
 by one hour) and their derivatives, such as wind direction and wind speed at
-the airport as well as spatial and temporal differences.
+the airport itself as well as spatial and temporal differences.
 
@@
 benchmarked against alternative probabilistic forecasting methods. The study
-is based on $1$\,h and $3$\,h forecasts employing lagged observations in the
+is based on $+1$\,h and $+3$\,h forecasts employing lagged observations in the
 vicinity of the airports as possible predictor variables.
@@
 
-The data used in this study consists of four subsequent years from January 2015
+The data used in this study consists of five subsequent years from January 2014
 to December 2018. After first eliminating predictor variables with more than
@@
       \noalign{\vskip 1mm}                              
-      Weather stations (Innsbruck): & 4 stations at the airport, \\
+      Weather stations (Innsbruck): & 4 stations at the airport, as well as\\
                                     & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
-                                    & Steinach\\
+                                    & and Steinach\\
       \noalign{\vskip 1mm}
-      Weather stations (Vienna):    & 9 stations at the airport, \\
+      Weather stations (Vienna):    & 9 stations at the airport, as well as\\
                                     & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
@@
                                     & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
-                                    & Seibersdorf, Unterlaa, Wolkersdorf\\
+                                    & Seibersdorf, Unterlaa, and Wolkersdorf\\
       \hline
@@
     sampling algorithm using weakly informative priors.  \newline Following
-    \cite{Mulder+Klugkist:2017} the model uses a link function~$g$ to keep the
+    \cite{Mulder+Klugkist:2017} the model uses a link function~$g(\cdot)$ to keep the
     response values within an interval of length $2\pi$. As the implementation
@@
 Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
-at forecast steps 1\,h and 3\,h for Innsbruck (Panels~a,\,c) and
+at forecast steps $+1$\,h and $+3$\,h for Innsbruck (Panels~a,\,c) and
 Vienna (Panels~b,\,d). The scores are aggregated over the five validation
@@
 model due to its strong vulnerability to abrupt wind shifts. On the contrary,
-at Vienna the geographic surroundings do not constrain any wind direction which
-leads to smaller and less abrupt changes in the wind direction as well as less
-pronounced wind regimes. This seems to weaken the predictive performance of the
-climatological and linear models, and to reduce the performance variability for
-all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).
+at Vienna smaller and less abrupt changes in the wind direction as well as less pronounced 
+wind regimes are observed due to the less mountainous surrounding. This seems to 
+weaken the predictive performance of the climatological and linear models, and to 
+reduce the performance variability for all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).
 
@@
 case study for shortterm probabilistic wind direction forecasting at two
-airports with different site characteristics. As benchline models,
+airports with different site characteristics. As benchmark models,
 probabilistic climatology and persistency models, as well as a state-of-the-art
@@
 is simply the identity function $v_l(z_{li}) = z_{li}$ such that $t_l \in \mathbb{R}^2$ 
-as the ``vec'' operator converts the matrix of dimension $1 \times 2$ into a $2$ column vector.
+as the ``$vec$'' operator converts the matrix of dimension $1 \times 2$ into a $2$ column vector.
 If $z_l$ is a categorical variable with $h$ categories then 
@@
 of dimension $h$ where the entry $1$ indicates the category of $z_{li}$. 
-In this case the ``vec'' operator converts the $h \times 2$ matrix into a $h \cdot 2$ column vector 
+In this case the ``$vec$'' operator converts the $h \times 2$ matrix into a $h \cdot 2$ column vector 
 by column-wise combination such that $t_l \in \mathbb{R}^{h \cdot 2}$. 

### 2019-12-16 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added further references and first attempts to include them in the draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2935 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8f128b169066c390627cca179fd10c03430556e3

@@
 homogeneity. However, this only models changes in the conditional mean
-but not the conditional variance or full probabilistic distribution.
+but not the conditional variance or full probabilistic distribution 
+%% FIXME: first attempt to include reference Gneiting+Balabdaoui+Raftery:2007
+which would allow for also considering uncertainty in forecasts \citep{Gneiting+Balabdaoui+Raftery:2007}.
+%%
 In contrast, we introduce a distributional tree approach that considers
@@
 where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
-$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} or
+$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001}, \citealt{Mardia+Zemroch:1975b}, or
 \citealt{Ley+Verdebout:2017} for a more detailed overview).
@@
 improves the predictive skill for short leadtimes, it possibly extends the
-potential forecast range of the different methods. Another possible
+potential forecast range of the different methods. 
+%% FIXME: first draft to include reference on wind speed (Thorarinsdottir+Gneiting:2010)
+For the specific case of wind direction forecasting it has to be considered
+that this quantity is highly related to wind speed. For example, low wind speeds
+usually leads to a higher fluctuation in wind direction. One way how to account for
+this influence is by modeling both quantities in a bivariate model. %% FIXME: add reference?
+Alternatively, the approaches of \cite{Thorarinsdottir+Gneiting:2010} could be 
+incorporated in circular regression trees and forests.
+%%
+Another possible
 improvement for obtaining more parsimonious circular trees is to consider

### 2019-12-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include suggestions for outlook (two-step model)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2937 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3758631ef5650fbaf593d89c4c8d13a9c4698d99

@@
 but not the conditional variance or full probabilistic distribution 
-%% FIXME: first attempt to include reference Gneiting+Balabdaoui+Raftery:2007
-which would allow for also considering uncertainty in forecasts \citep{Gneiting+Balabdaoui+Raftery:2007}.
+%% FIXME: first attempt to include reference Gneiting+Balabdaoui+Raftery:2007/ changed to Gneiting:2008
+which would allow for also considering uncertainty in forecasts \citep{Gneiting:2008}.
 %%
@@
 For the specific case of wind direction forecasting it has to be considered
-that this quantity is highly related to wind speed. For example, low wind speeds
-usually leads to a higher fluctuation in wind direction. One way how to account for
-this influence is by modeling both quantities in a bivariate model. %% FIXME: add reference?
-Alternatively, the approaches of \cite{Thorarinsdottir+Gneiting:2010} could be 
-incorporated in circular regression trees and forests.
+that this quantity is highly related to wind speed. For example, low wind
+speeds usually lead to a high fluctuation in wind direction. Hence, to account
+for both quantities simultaneously, distributional tree-based models could be
+built for wind vectors employing the bivariate normal distribution as performed
+for additive regression models by \citet{Lang+Mayr+Stauffer:2019}.
+Alternatively, also accounting for wind speed and direction but extending the
+presented methodology, circular trees and forests could be combined with
+distributional models for wind speed employing, e.g., the truncated normal
+distribution \citep{Thorarinsdottir+Gneiting:2010}. This could be pursued
+either in a two-step approach with two independent tree-based distributional
+models or in a more sophisticated approach employing a mixture response
+distribution.
+%One way how to account for
+%this influence is by modeling both quantities in a bivariate model. %% FIXME: add reference?
+%Alternatively, the approaches of \cite{Thorarinsdottir+Gneiting:2010} could be 
+%incorporated in circular regression trees and forests.
 %%

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor spelling correction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2938 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 62aa9b00d70146bc5fa6b8bdaa24107481e3d4f0

@@
 %% FIXME: first draft to include reference on wind speed (Thorarinsdottir+Gneiting:2010)
-For the specific case of wind direction forecasting it has to be considered
+For the specific case of wind direction forecasting, it has to be considered
 that this quantity is highly related to wind speed. For example, low wind
@@
 presented methodology, circular trees and forests could be combined with
-distributional models for wind speed employing, e.g., the truncated normal
+distributional models for wind speed employing the truncated normal
 distribution \citep{Thorarinsdottir+Gneiting:2010}. This could be pursued

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update conclusion (L+M+R)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2939 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ffa39c2b5ad2fc1ec6b667ddcee5f8b55a2314da

--- 2019-12-16 pkg/circtree/inst/wind_paper/draft/draft.Rnw
@@
 but not the conditional variance or full probabilistic distribution 
-%% FIXME: first attempt to include reference Gneiting+Balabdaoui+Raftery:2007
-which would allow for also considering uncertainty in forecasts \citep{Gneiting+Balabdaoui+Raftery:2007}.
-%%
+which would allow for also considering uncertainty in forecasts \citep{Gneiting:2008}.
 In contrast, we introduce a distributional tree approach that considers
@@
 where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
-$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001}, \citealt{Mardia+Zemroch:1975b}, or
+$0$ (see, e.g., \citealt{Mardia+Zemroch:1975b}, \citealt{Jammalamadaka+Sengupta:2001}, or
 \citealt{Ley+Verdebout:2017} for a more detailed overview).
@@
   \item \emph{Circular regression forest:} Following the description in
-    Section~\ref{sec:circforest}, the circular forest used in this study is
+    Section~\ref{sec:circforest}, the circular regression forest used in this study is
     constructed based on $100$~individual trees employing the \textsf{R} package
@@
 years, yielding yearly mean values for every hour per calendar month, with a
-lower score indicating better performance. The circular forest overall provides
+lower score indicating better performance. The circular regression forest overall provides
 the best predictive performance, followed by the circular regression tree and
@@
 values indicate an improvement in the predictive performance over the
-reference. For all setups, the circular forest has the highest skill scores
+reference. For all setups, the circular regression forest has the highest skill scores
 with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and Vienna,
@@
 
-The performance of the novel circular trees and forests is assessed in a 
-case study for shortterm probabilistic wind direction forecasting at two
-airports with different site characteristics. As benchmark models,
-probabilistic climatology and persistency models, as well as a state-of-the-art
-circular GLM-type model are evaluated based on proper scoring rules. In summary, the
-circular trees and forests have the highest predictive performance in this
+The performance of the novel circular regression trees and forests is assessed in a case
+study for shortterm probabilistic wind direction forecasting at two airports
+with different site characteristics. As benchmark models, probabilistic
+climatology and persistency models, as well as a state-of-the-art circular
+GLM-type model are evaluated based on proper scoring rules. In summary, the
+circular regression trees and forests have the highest predictive performance in this
 setting. For cases without changes in the wind regime, lagged response values
@@
 improves the predictive skill for short leadtimes, it possibly extends the
-potential forecast range of the different methods. 
-%% FIXME: first draft to include reference on wind speed (Thorarinsdottir+Gneiting:2010)
-For the specific case of wind direction forecasting it has to be considered
-that this quantity is highly related to wind speed. For example, low wind speeds
-usually leads to a higher fluctuation in wind direction. One way how to account for
-this influence is by modeling both quantities in a bivariate model. %% FIXME: add reference?
-Alternatively, the approaches of \cite{Thorarinsdottir+Gneiting:2010} could be 
-incorporated in circular regression trees and forests.
-%%
-Another possible
-improvement for obtaining more parsimonious circular trees is to consider
-splitting circular covariates into two circle segments by searching two split
-points simultaneously rather than sequentially at different depths. While this
-might slightly improve the predictive performance of circular trees, this
-should not affect the performance of circular forests, as these consist of
-very large trees with many different splits.
+potential forecast range of the different methods. For the specific task of
+wind forecasting not only the wind direction, but also the wind speed and
+especially their combination is of high interest. To account for both
+quantities simultaneously, distributional tree-based models could be built for
+wind vectors employing the bivariate normal distribution as performed by
+\citet{Lang+Mayr+Stauffer:2019} in an additive regression framework.
+Alternatively, circular regression trees could be combined with distributional models for
+wind speed employing the truncated normal distribution
+\citep{Thorarinsdottir+Gneiting:2010}. This could be pursued either in a
+two-step approach with two independent distributional models or in a more
+sophisticated way employing a mixed response distribution in the presented
+tree-building algorithm. Another possible improvement for obtaining more
+parsimonious circular regression trees is to consider splitting circular covariates into
+two circle segments by searching two split points simultaneously rather than
+sequentially at different depths. While this might slightly improve the
+predictive performance of circular regression trees, this should not affect the
+performance of the forests, as they consist of very large trees with many
+different splits.
 
@@
 possible covariates. Moreover, the case study shows that building ensembles of
-circular trees can even improve the forecasting performance, as the resulting
-circular forests allow for modeling smooth effects and stabelize the model.
+circular regression trees can even improve the forecasting performance, as the resulting
+forests allow for modeling smooth effects and stabelize the model.
 
@@
 The corresponding implementation of the proposed methodology for circular
-trees and forests is provided in the \textsf{R} package \textbf{circtree} 
+regression trees and forests is provided in the \textsf{R} package \textbf{circtree} 
 (version~\Sexpr{packageVersion("circtree")}).
-% The proposed circular tree and forest models are implemented in the \textsf{R}
+% The proposed circular regression tree and forest models are implemented in the \textsf{R}
 % package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). 

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
spell check and 'von~Mises'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2940 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2fd01d599a3b46feaf0f98614874ebb08e78f40d

@@
 approach for both trees and forests yielding probabilistic forecasts based on the
-von Mises distribution. The resulting tree-based models simplify
+von~Mises distribution. The resulting tree-based models simplify
 the estimation process by using the available covariates for partitioning the
-data into sufficiently homogeneous subgroups so that a simple von Mises distribution
+data into sufficiently homogeneous subgroups so that a simple von~Mises distribution
 without further covariates can be fitted to the circular response in each subgroup.
@@
 covariates that are associated with either location and/or scale changes in the
-von Mises distribution. Combining an ensemble of circular regression trees
+von~Mises distribution. Combining an ensemble of circular regression trees
 to a circular regression forest yields a local adaptive likelihood estimator
-for the von Mises distribution that can regularize and smooth the covariate effects.
+for the von~Mises distribution that can regularize and smooth the covariate effects.
 The new methods are evaluated in a case study on probabilistic wind direction
@@
 
-\Keywords{circular data, regression trees, random forests, distributional regression, von Mises distribution}
+\Keywords{circular data, regression trees, random forests, distributional regression, von~Mises distribution}
 
@@
 Many approaches to model circular data assume that the circular variable of
-interest follows a circular distribution, in particular the von Mises
+interest follows a circular distribution, in particular the von~Mises
 distribution which is also known as the ``circular normal distribution''. One
@@
 covariates was presented by \cite{Gould:1969} where the circular mean
-%% , i.e., the location parameter of the von Mises distribution, 
+%% , i.e., the location parameter of the von~Mises distribution, 
 is predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
@@
 As a very intuitive and data-driven alternative, we propose a flexible
-tree-based regression approach for modeling circular data by applying the von
-Mises distribution within the methodology of distributional trees and forests
+tree-based regression approach for modeling circular data by applying the von~Mises 
+distribution within the methodology of distributional trees and forests
 \citep{Schlosser+Hothorn+Stauffer:2019}. The resulting circular regression
@@
 using the available covariates for partitioning the
-data into sufficiently homogeneous subgroups so that a simple von Mises distribution
+data into sufficiently homogeneous subgroups so that a simple von~Mises distribution
 without further covariates can be fitted to the circular response in each of these subgroups.
@@
 \includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
-\caption{\label{fig:tree_ibk} Fitted tree based on the von Mises distribution for 1-hourly wind
+\caption{\label{fig:tree_ibk} Fitted tree based on the von~Mises distribution for 1-hourly wind
 direction forecasts at the airport of Innsbruck. In each terminal node the
@@
 Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with
-the empirical (gray) and fitted von Mises (red) wind direction distribution
+the empirical (gray) and fitted von~Mises (red) wind direction distribution
 in each terminal node. Based on the fitted location parameters $\hat \mu$, the
@@
 The remainder of the paper is structured as follows: The theory on
-probabilistic circular modeling introducing the von Mises distribution, and
+probabilistic circular modeling introducing the von~Mises distribution, and
 circular regression models are discussed in Section~\ref{sec:prob_circ}.
@@
 approximation to the wrapped normal distribution that is mathematically simpler
-and hence easier to use is provided by the von Mises
+and hence easier to use is provided by the von~Mises
 distribution \citep{Fisher:1993}, a purely circular distribution which is also known as ``the
@@
 modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
-and a concentration parameter $\kappa > 0$ the density of the von Mises
+and a concentration parameter $\kappa > 0$ the density of the von~Mises
 distribution for an observation $y \in [0, 2\pi)$ is given by
@@
   \end{minipage}%
-  \caption{\label{fig:densities} Illustration of a von Mises model for circular data in
+  \caption{\label{fig:densities} Illustration of a von~Mises model for circular data in
     the interval $[0, 2\pi)$ fitted by maximum likelihood. Left: Linearized
@@
 %%
-Figure~\ref{fig:densities} depicts von Mises model for circular data in
+Figure~\ref{fig:densities} depicts von~Mises model for circular data in
 $[0, 2 \pi)$ fitted by maximum likelihood, either using a linearized
@@
 Starting out from the ideas of \citet{Lund:2002}, we introduce circular regression trees and
-forests considering splits in all distribution parameters of the von Mises
+forests considering splits in all distribution parameters of the von~Mises
 distribution and providing a full probabilistic model.  Moreover, the resulting
@@
 full distributional model in each node of the tree yields a distributional
-regression tree, where selecting the von Mises distribution enables an
+regression tree, where selecting the von~Mises distribution enables an
 application to circular data. The crucial step of how and where to split the
@@
 obtained by evaluating the score function~$s(\cdot)$ for each individual observation
-at the parameter estimates (Equation~\ref{eq:scores}). For the von Mises
+at the parameter estimates (Equation~\ref{eq:scores}). For the von~Mises
 distribution with its two distribution parameters ($\mu$ and $\kappa$) and a
@@
 subsets. Then the estimated parameter pair
-$(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted von Mises
+$(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted von~Mises
 distribution for a given $\bm{z}$ is obtained by a weighted maximum likelihood estimator:
@@
 In that way a specific set of weights can be calculated for each observation
-yielding its specific parameter estimates for the von Mises distribution
+yielding its specific parameter estimates for the von~Mises distribution
 %%
@@
     the climatological model by using maximum likelihood estimation and fitting the
-    distribution parameters of the von Mises distribution conditional on lagged
+    distribution parameters of the von~Mises distribution conditional on lagged
     response values according to the description in Section~\ref{sec:prob_circ}.
@@
     wind speed $spd$ as predictor variables. The model formula for the location
-    parameter $\mu$ of the von Mises distribution can be written as:
+    parameter $\mu$ of the von~Mises distribution can be written as:
     %%
@@
 forests are established by coupling model-based recursive partitioning with the
-von Mises distribution. By separating the data into more homogeneous
+von~Mises distribution. By separating the data into more homogeneous
 subgroups, possible difficulties in circular regression are bypassed as
 covariates are solely considered for splitting and group-specific models are
-fitted without further covariates. In addition, by specifying the von Mises
+fitted without further covariates. In addition, by specifying the von~Mises
 distribution for each node and allowing for splits in both distribution
@@
 
-First, employing the von Mises distribution, a distributional model $\text{vM}(y; \mu,\kappa)$ 
+First, employing the von~Mises distribution, a distributional model $\text{vM}(y; \mu,\kappa)$ 
 is fitted to the learning sample of circular observations $\{y_i\}_{i = 1,\ldots,n}$ as explained 

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small touch-ups and re-formulations in summary
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2941 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 19def0c8742f8f7e869c5227c4f31e0ee1102b0e

@@
 improves the predictive skill for short leadtimes, it possibly extends the
-potential forecast range of the different methods. For the specific task of
-wind forecasting not only the wind direction, but also the wind speed and
-especially their combination is of high interest. To account for both
-quantities simultaneously, distributional tree-based models could be built for
-wind vectors employing the bivariate normal distribution as performed by
-\citet{Lang+Mayr+Stauffer:2019} in an additive regression framework.
-Alternatively, circular regression trees could be combined with distributional models for
-wind speed employing the truncated normal distribution
-\citep{Thorarinsdottir+Gneiting:2010}. This could be pursued either in a
-two-step approach with two independent distributional models or in a more
-sophisticated way employing a mixed response distribution in the presented
-tree-building algorithm. Another possible improvement for obtaining more
+potential forecast range of the different methods.
+
+For the specific task of wind forecasting, the wind direction is often only
+relevant if the wind speed is sufficiently high. Hence, it is of interest
+to account for both quantities simultaneously, e.g., by considering a 
+bivariate normal distribution for the wind vectors (from which wind speed
+and wind direction can be obtained). The parameters of this bivariate normal
+distribution could then be linked to the available covariate using 
+an additive regression framework \citep[as proposed by][]{Lang+Mayr+Stauffer:2019}
+or using a tree-based approach, similar to the one proposed in this paper.
+Moreover, a rather different approach for a combined response of wind speed
+and wind direction would be a two-step or hurdle model: In the first step
+this could build on the truncated normal model of
+\cite{Thorarinsdottir+Gneiting:2010} to capture wind speed. And in a second
+step a circular wind direction model is leveraged provided that a certain
+hurdle for the wind speed is crossed.
+
+Another possible improvement for obtaining more
 parsimonious circular regression trees is to consider splitting circular covariates into

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor spelling corrections and formating
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2942 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 795d4259043e111819d42077e695670f25a38d07

@@
 \Abstract{
-While circular data occur in a wide range of scientific fields,
-the methodology for distributional modeling and probabilistic forecasting of circular
-response variables is rather limited. Most of the existing methods are built on the framework of
-generalized linear and additive models, which are often challenging to optimize
-and to interpret. Therefore, we suggest circular regression trees and random forests
-as an intuitive alternative approach that is relatively easy to fit. Building on
-previous ideas for trees modeling circular means, we suggest a distributional
-approach for both trees and forests yielding probabilistic forecasts based on the
-von~Mises distribution. The resulting tree-based models simplify
-the estimation process by using the available covariates for partitioning the
-data into sufficiently homogeneous subgroups so that a simple von~Mises distribution
-without further covariates can be fitted to the circular response in each subgroup.
-These circular regression trees are straightforward to interpret, can capture
-nonlinear effects and interactions, and automatically select the relevant
-covariates that are associated with either location and/or scale changes in the
-von~Mises distribution. Combining an ensemble of circular regression trees
-to a circular regression forest yields a local adaptive likelihood estimator
-for the von~Mises distribution that can regularize and smooth the covariate effects.
+While circular data occur in a wide range of scientific fields, the methodology
+for distributional modeling and probabilistic forecasting of circular response
+variables is rather limited. Most of the existing methods are built on the
+framework of generalized linear and additive models, which are often
+challenging to optimize and to interpret. Therefore, we suggest circular
+regression trees and random forests as an intuitive alternative approach that
+is relatively easy to fit. Building on previous ideas for trees modeling
+circular means, we suggest a distributional approach for both trees and forests
+yielding probabilistic forecasts based on the von~Mises distribution. The
+resulting tree-based models simplify the estimation process by using the
+available covariates for partitioning the data into sufficiently homogeneous
+subgroups so that a simple von~Mises distribution without further covariates
+can be fitted to the circular response in each subgroup. These circular
+regression trees are straightforward to interpret, can capture nonlinear
+effects and interactions, and automatically select the relevant covariates that
+are associated with either location and/or scale changes in the von~Mises
+distribution. Combining an ensemble of circular regression trees to a circular
+regression forest yields a local adaptive likelihood estimator for the
+von~Mises distribution that can regularize and smooth the covariate effects.
 The new methods are evaluated in a case study on probabilistic wind direction
-forecasting at two Austrian airports, considering other common approaches as
-a benchmark.
+forecasting at two Austrian airports, considering other common approaches as a
+benchmark.
 }
@@
 weather variables in meteorology. Fitting a statistical model to this type of
-data requires the incorporation of its specific feature of periodicity. For example,
-angular data are restricted to an interval such
-as $[0,2\pi)$ with $0$ being equivalent~to~$2\pi$.
+data requires the incorporation of its specific feature of periodicity. For
+example, angular data are restricted to an interval such as $[0,2\pi)$ with $0$
+being equivalent~to~$2\pi$.
 
@@
 of the first regression models with a circular response variable and linear
-covariates was presented by \cite{Gould:1969} where the circular mean
-%% , i.e., the location parameter of the von~Mises distribution, 
-is predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
-plugging in a link function transforming the linear predictor to a restricted
-interval of length $2\pi$. This generalized linear model (GLM) type approach
-was further extended by \cite{Fisher+Lee:1992} and subsequently by
-\cite{Fisher:1993} introducing independent GLMs for either the location or
-scale parameter with appropriate link functions while keeping the other
-parameter constant. Additionally, they developed a combined heteroscedastic
-or distributional version with alternating reestimation of the location and scale parameters
-conditional on the respective sets of covariates until convergence. While all of
-these models are built on well-elaborated theory, their application in practice
-remains very challenging, mainly due to the complexity
-encountered in optimizing the corresponding log-likelihood function which is
-not globally concave. Therefore, highly-informative starting values are
-crucial for such circular GLMs to converge
+covariates was presented by \cite{Gould:1969} where the circular mean is
+predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978}
+refined this idea by plugging in a link function transforming the linear
+predictor to a restricted interval of length $2\pi$. This generalized linear
+model (GLM) type approach was further extended by \cite{Fisher+Lee:1992} and
+subsequently by \cite{Fisher:1993} introducing independent GLMs for either the
+location or scale parameter with appropriate link functions while keeping the
+other parameter constant. Additionally, they developed a combined
+heteroscedastic or distributional version with alternating reestimation of the
+location and scale parameters conditional on the respective sets of covariates
+until convergence. While all of these models are built on well-elaborated
+theory, their application in practice remains very challenging, mainly due to
+the complexity encountered in optimizing the corresponding log-likelihood
+function which is not globally concave. Therefore, highly-informative starting
+values are crucial for such circular GLMs to converge
 \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
 this strong dependence on appropriate initial values,
-\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a
-homoscedastic GLM for circular data. However, apart from potential
-difficulties in the optimization procedure of circular GLMs, the interpretation
-of the underlying additive effects is often challenging as well because the link function
-is highly nonlinear and the representation of smooth transitions on the unit circle
-is not straightforward. For example, the same rotation can be
-obtained in either positive or negative direction on the circle leading to an
-ambiguous interpretation.
+\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a homoscedastic
+GLM for circular data. However, apart from potential difficulties in the
+optimization procedure of circular GLMs, the interpretation of the underlying
+additive effects is often challenging as well because the link function is
+highly nonlinear and the representation of smooth transitions on the unit
+circle is not straightforward. For example, the same rotation can be obtained
+in either positive or negative direction on the circle leading to an ambiguous
+interpretation.
 
 As a very intuitive and data-driven alternative, we propose a flexible
-tree-based regression approach for modeling circular data by applying the von~Mises 
-distribution within the methodology of distributional trees and forests
-\citep{Schlosser+Hothorn+Stauffer:2019}. The resulting circular regression
-trees and forests avoid the discussed difficulties of circular GLMs by
-using the available covariates for partitioning the
-data into sufficiently homogeneous subgroups so that a simple von~Mises distribution
-without further covariates can be fitted to the circular response in each of these subgroups.
-This obviates the need for a link function or for iterating between models for the
-separate distribution parameters. By leveraging the distributional modeling approach, the
-trees can automatically detect and capture differences in both
-distribution parameters, providing a fully
-specified circular response distribution in each terminal node offering a wide
-range of statistical inference. In addition, the employed tree structure
-allows to capture non-additive effects while forests enable the modeling of
-smooth changes. Furthermore, covariates and their possible interactions do not
-need to be specified in advance as they are selected automatically in the
-recursive partitioning algorithm.
+tree-based regression approach for modeling circular data by applying the
+von~Mises distribution within the methodology of distributional trees and
+forests \citep{Schlosser+Hothorn+Stauffer:2019}. The resulting circular
+regression trees and forests avoid the discussed difficulties of circular GLMs
+by using the available covariates for partitioning the data into sufficiently
+homogeneous subgroups so that a simple von~Mises distribution without further
+covariates can be fitted to the circular response in each of these subgroups.
+This obviates the need for a link function or for iterating between models for
+the separate distribution parameters. By leveraging the distributional modeling
+approach, the trees can automatically detect and capture differences in both
+distribution parameters, providing a fully specified circular response
+distribution in each terminal node offering a wide range of statistical
+inference. In addition, the employed tree structure allows to capture
+non-additive effects while forests enable the modeling of smooth changes.
+Furthermore, covariates and their possible interactions do not need to be
+specified in advance as they are selected automatically in the recursive
+partitioning algorithm.
 
 This novel approach to circular regression trees and forests complements the
-literature on tree-based circular modeling. \cite{Lund:2002} already
-introduced a circular regression tree algorithm where binary splits are
-made based on an angular distance measure capturing node
-homogeneity. However, this only models changes in the conditional mean
-but not the conditional variance or full probabilistic distribution 
-which would allow for also considering uncertainty in forecasts \citep{Gneiting:2008}.
-In contrast, we introduce a distributional tree approach that considers
-splits in all distribution parameters and yields a full probabilistic predictive model.
-In addition, as a natural extension, ensembles or forests of circular
-regression trees are presented in this study, showing that these can further
-improve predictive power.
+literature on tree-based circular modeling. \cite{Lund:2002} already introduced
+a circular regression tree algorithm where binary splits are made based on an
+angular distance measure capturing node homogeneity. However, this only models
+changes in the conditional mean but not the conditional variance or full
+probabilistic distribution which would allow for also considering uncertainty
+in forecasts \citep{Gneiting:2008}. In contrast, we introduce a distributional
+tree approach that considers splits in all distribution parameters and yields a
+full probabilistic predictive model. In addition, as a natural extension,
+ensembles or forests of circular regression trees are presented in this study,
+showing that these can further improve predictive power.
 
@@
 \includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
-\caption{\label{fig:tree_ibk} Fitted tree based on the von~Mises distribution for 1-hourly wind
-direction forecasts at the airport of Innsbruck. In each terminal node the
-empirical histogram (gray) and fitted density (red line) are depicted along
-with the estimated location parameter (red hand). The covariates selected for
-splitting are wind direction (meteorological degree), wind speed ($\text{ms}^{-1}$), and
-pressure gradients ($\text{dpressure; hPa}$) west, east and south of the
-airport, all lagged by one hour. Note that in the meteorological context wind
-direction is defined on the scale $[0,360]$ degree and increases clockwise from
-North ($0$ degree).}
+\caption{\label{fig:tree_ibk} Fitted tree based on the von~Mises distribution
+for 1-hourly wind direction forecasts at the airport of Innsbruck. In each
+terminal node the empirical histogram (gray) and fitted density (red line) are
+depicted along with the estimated location parameter (red hand). The covariates
+selected for splitting are wind direction (meteorological degree), wind speed
+($\text{ms}^{-1}$), and pressure gradients ($\text{dpressure; hPa}$) west, east
+and south of the airport, all lagged by one hour. Note that in the
+meteorological context wind direction is defined on the scale $[0,360]$ degree
+and increases clockwise from North ($0$ degree).}
 \end{figure}
@@
 importance for decision-making processes, e.g., in air traffic management as
-considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
-for 1-hourly forecasts at Innsbruck Airport, located at the bottom of
-a narrow valley within the European Alps. Topography channels the wind along
-the west-east valley axis or along a tributary valley intersecting from the
-south. Hence, pressure gradients to which valley wind regimes react are
-considered as covariates along with other meteorological measurements (lagged
-by one hour) and their derivatives, such as wind direction and wind speed at
-the airport itself as well as spatial and temporal differences.
-
-Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with
-the empirical (gray) and fitted von~Mises (red) wind direction distribution
-in each terminal node. Based on the fitted location parameters $\hat \mu$, the
+considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree for
+1-hourly forecasts at Innsbruck Airport, located at the bottom of a narrow
+valley within the European Alps. Topography channels the wind along the
+west-east valley axis or along a tributary valley intersecting from the south.
+Hence, pressure gradients to which valley wind regimes react are considered as
+covariates along with other meteorological measurements (lagged by one hour)
+and their derivatives, such as wind direction and wind speed at the airport
+itself as well as spatial and temporal differences.
+
+Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with the
+empirical (gray) and fitted von~Mises (red) wind direction distribution in each
+terminal node. Based on the fitted location parameters $\hat \mu$, the
 subgroups can be distinguished into the following wind regimes: (1)~Up-valley
 winds blowing from the valley mouth towards the upper valley (from east to
-west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest along
+west, nodes~4 and 5); (2)~Downslope winds blowing across the Alpine crest along
 the intersecting valley towards Innsbruck (from south-east to north-west,
-node~8); (3) Down-valley winds blowing in the direction of the valley mouth
-(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
+node~8); (3)~Down-valley winds blowing in the direction of the valley mouth
+(from west to east, nodes 10, 12 and~13). Node~7 captures observations with
 rather low wind speeds that cannot be clearly distinguished into specific wind
@@
 vs.\ low precision. A more extensive case study of circular regression trees
-and forests applied to probabilistic wind direction forecasting at
-Innsbruck Airport and Vienna International Airport is presented in Section~\ref{sec:wind},
-along with a benchmark against commonly-used alternative approaches.
+and forests applied to probabilistic wind direction forecasting at Innsbruck
+Airport and Vienna International Airport is presented in
+Section~\ref{sec:wind}, along with a benchmark against commonly-used
+alternative approaches.
 
@@
 probabilistic circular modeling introducing the von~Mises distribution, and
-circular regression models are discussed in Section~\ref{sec:prob_circ}.
-The methodology of circular regression trees and forests and their features are
+circular regression models are discussed in Section~\ref{sec:prob_circ}. The
+methodology of circular regression trees and forests and their features are
 introduced in Section~\ref{sec:tree_forest}. After the case study presented in
@@
 approximation to the wrapped normal distribution that is mathematically simpler
-and hence easier to use is provided by the von~Mises
-distribution \citep{Fisher:1993}, a purely circular distribution which is also known as ``the
-circular normal distribution'' and is a common choice for probabilistic
+and hence easier to use is provided by the von~Mises distribution
+\citep{Fisher:1993}, a purely circular distribution which is also known as
+``the circular normal distribution'' and is a common choice for probabilistic
 modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
@@
 %%
-To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response $y$, the
-distribution parameters $\mu$ and $\kappa$ need to be estimated.
-This can be done by
-maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
+To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response
+$y$, the distribution parameters $\mu$ and $\kappa$ need to be estimated. This
+can be done by maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
 %%
@@
 yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
-a fully specified distributional model is fitted to the learning data $\{y_i\}_{i=1,\ldots,n}$.
+a fully specified distributional model is fitted to the learning data
+$\{y_i\}_{i=1,\ldots,n}$.
 
@@
 %%
-provides a way to obtain a measure of goodness of fit of the model for
-each observation and fitted parameter. Then, the optimization problem in
+provides a way to obtain a measure of goodness of fit of the model for each
+observation and fitted parameter. Then, the optimization problem in
 Equation~\ref{eq:optim} can alternatively be specified as 
@@
 %%
-Figure~\ref{fig:densities} depicts von~Mises model for circular data in
-$[0, 2 \pi)$ fitted by maximum likelihood, either using a linearized
-(left) or circular (right) scale. In both cases,
-the empirical histogram (gray bars) is shown along with the fitted density (red line) and
-estimated location parameter (red hand). However, this distributional model
-only considers the circular response variable but no covariate. Of course,
-including covariates is of interest in a regression setup for forecasting.
+Figure~\ref{fig:densities} depicts von~Mises model for circular data in $[0, 2
+\pi)$ fitted by maximum likelihood, either using a linearized (left) or
+circular (right) scale. In both cases, the empirical histogram (gray bars) is
+shown along with the fitted density (red line) and estimated location parameter
+(red hand). However, this distributional model only considers the circular
+response variable but no covariate. Of course, including covariates is of
+interest in a regression setup for forecasting.
 
 In most generalized linear model (GLM) or generalized additive model (GAM)
-approaches to circular regression the location parameter $\mu$ depends on covariates $\mathbf{z}$
-through a link function $g(\cdot)$, circular intercept $\mu_0$ and
-coefficient vector $\mathbf{\beta}$:
-\begin{equation}
-\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right).
-\end{equation}
-The link function transforms the additive predictor to an interval of length
-$2\pi$.  Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
+approaches to circular regression the location parameter $\mu$ depends on
+covariates $\mathbf{z}$ through a link function $g(\cdot)$, circular intercept
+$\mu_0$ and coefficient vector $\mathbf{\beta}$: \begin{equation} \mu = \mu_0 +
+g\left(\mathbf{\beta}^{\top} \mathbf{z}\right). \end{equation} The link
+function transforms the additive predictor to an interval of length $2\pi$.
+Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
 \citet{Fisher+Lee:1992}. They also developed a heteroscedastic version by
@@
 $\kappa$. This provides a first approach to a fully probabilistic regression
-model for circular data, albeit the parameters are not regressed
-simultaneously on covariates as in the more general framework provided by
-generalized additive models for location, scale, and shape \citep[GAMLSS,
+model for circular data, albeit the parameters are not regressed simultaneously
+on covariates as in the more general framework provided by generalized additive
+models for location, scale, and shape \citep[GAMLSS,
 ][]{Rigby+Stasinopoulos:2005}. Nevertheless, other circular additive models
-share the previously discussed difficulties induced by the
-characteristics of the log-likelihood function and the strongly nonlinear link
-function. Referring to additive models in general, it has to be considered that
-a proper model specification can be very challenging, particularly for a high
-number of covariates and no information on possible interactions. Moreover, the
-additive structure might impose a smooth effect even if the true underlying
-effect is an abrupt shift, which occurs, e.g., in atmospheric wind fields.
-
-In contrast, the tree-based circular regression models proposed in the next sections
-largely avoid the problems above by employing recursive partitioning in combination
-with local adaptive likelihood estimation.
+share the previously discussed difficulties induced by the characteristics of
+the log-likelihood function and the strongly nonlinear link function. Referring
+to additive models in general, it has to be considered that a proper model
+specification can be very challenging, particularly for a high number of
+covariates and no information on possible interactions. Moreover, the additive
+structure might impose a smooth effect even if the true underlying effect is an
+abrupt shift, which occurs, e.g., in atmospheric wind fields.
+
+In contrast, the tree-based circular regression models proposed in the next
+sections largely avoid the problems above by employing recursive partitioning
+in combination with local adaptive likelihood estimation.
 
@@
 
-Starting out from the ideas of \citet{Lund:2002}, we introduce circular regression trees and
-forests considering splits in all distribution parameters of the von~Mises
-distribution and providing a full probabilistic model.  Moreover, the resulting
-tree-based models provide a very intuitive and data-driven alternative to
-commonly used GLMs for circular data.
+Starting out from the ideas of \citet{Lund:2002}, we introduce circular
+regression trees and forests considering splits in all distribution parameters
+of the von~Mises distribution and providing a full probabilistic model.
+Moreover, the resulting tree-based models provide a very intuitive and
+data-driven alternative to commonly used GLMs for circular data.
 
@@
 Fitting a global model to a full data set can be very challenging, particularly
-for complex data with substantial variations. Therefore, separating the data set
-into more homogeneous subgroups based on covariates before fitting a local
+for complex data with substantial variations. Therefore, separating the data
+set into more homogeneous subgroups based on covariates before fitting a local
 model in each of these subgroups allows to capture (potential) group-specific
@@
 \citep{Hothorn+Hornik+Zeileis:2006}. For this purpose, model scores are
-obtained by evaluating the score function~$s(\cdot)$ for each individual observation
-at the parameter estimates (Equation~\ref{eq:scores}). For the von~Mises
-distribution with its two distribution parameters ($\mu$ and $\kappa$) and a
-data set of $n$ observations, this yields an $n \times 2$ matrix that can be
-employed as a discrepancy measure, capturing how well each given observation
-conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$,
-respectively.  To capture dependence on covariates, the association between the
-model's scores and each available covariate is assessed using either a
-parameter instability test~(MOB) or a permutation test~(CTree). By doing so in
-each partitioning step, the covariate with the highest significant association
-(i.e., lowest significant $p$-value, if any) is selected for splitting the
-data. The corresponding split point is chosen either by optimizing the
-log-likelihood~(MOB) or a two-sample test statistic~(CTree) over all possible
-partitions. This procedure is repeated recursively until there are no
-significant parameter instabilities or until another stopping criterion is met
-(e.g., subgroup size or tree depth). A more detailed description of the applied
-tree-building algorithm can be found in Appendix~\ref{app:algorithm}.
+obtained by evaluating the score function~$s(\cdot)$ for each individual
+observation at the parameter estimates (Equation~\ref{eq:scores}). For the
+von~Mises distribution with its two distribution parameters ($\mu$ and
+$\kappa$) and a data set of $n$ observations, this yields an $n \times 2$
+matrix that can be employed as a discrepancy measure, capturing how well each
+given observation conforms with the estimated location $\hat{\mu}$ and
+precision $\hat{\kappa}$, respectively. To capture dependence on covariates,
+the association between the model's scores and each available covariate is
+assessed using either a parameter instability test~(MOB) or a permutation
+test~(CTree). By doing so in each partitioning step, the covariate with the
+highest significant association (i.e., lowest significant $p$-value, if any) is
+selected for splitting the data. The corresponding split point is chosen either
+by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
+over all possible partitions. This procedure is repeated recursively until
+there are no significant parameter instabilities or until another stopping
+criterion is met (e.g., subgroup size or tree depth). A more detailed
+description of the applied tree-building algorithm can be found in
+Appendix~\ref{app:algorithm}.
 
@@
 $b$-th out of $B$ segments partitioning the covariate space in disjoint
-subsets. Then the estimated parameter pair
-$(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted von~Mises
-distribution for a given $\bm{z}$ is obtained by a weighted maximum likelihood estimator:
+subsets. Then the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$
+specifying the predicted von~Mises distribution for a given $\bm{z}$ is
+obtained by a weighted maximum likelihood estimator:
 %%
@@
 covariates $\bm{z}$ whereas $w_i(\bm{z}) = 1$ would correspond to the
-unweighted full-sample estimates and $w_i(\bm{z}) \in \{0, 1\}$
-corresponds to the subgroup selection from a tree. Thus, circular regression forests
-can capture both smooth and abrupt changes, while covariates and possible
+unweighted full-sample estimates and $w_i(\bm{z}) \in \{0, 1\}$ corresponds to
+the subgroup selection from a tree. Thus, circular regression forests can
+capture both smooth and abrupt changes, while covariates and possible
 interactions are selected automatically and do not explicitly need to be
@@
 one in flat terrain, the other one in mountainous terrain. The results are
-benchmarked against alternative probabilistic forecasting methods. The study
-is based on $+1$\,h and $+3$\,h forecasts employing lagged observations in the
+benchmarked against alternative probabilistic forecasting methods. The study is
+based on $+1$\,h and $+3$\,h forecasts employing lagged observations in the
 vicinity of the airports as possible predictor variables.
@@
 \label{sec:wind:data}
-The circular response variable considered in this case study is a $10$\,min-average
-of wind direction measurements at Innsbruck Airport (INN) and Vienna International Airport (VIE) on
-an hourly temporal resolution. Temporal information and $1$-hourly resolved $10$\,min mean observations of various
-meteorological quantities are used as predictor variables, including wind direction, wind speed, temperature, air
-pressure and humidity, all lagged by one or three hours according to the
-respective forecasting step. The meteorological variables are measured either
-directly at the airports or within their vicinities. For Innsbruck,
-measurements at the airport and along the intersecting valleys are used,
-whereas, for Vienna, measurements at the airport and within its vicinity of
-approximately 30\,km are used. Figure~\ref{fig:overview} provides a topographical
-overview of the airports and their surrounding areas with the station sites employed
-in this study. In addition, we use derived quantities
-such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
-changes and spatial differences towards the airport of the respective
-quantities. An overview of the employed data sets can be found in
-Table~\ref{tab:data}.
+The circular response variable considered in this case study is a
+$10$\,min-average of wind direction measurements at Innsbruck Airport (INN) and
+Vienna International Airport (VIE) on an hourly temporal resolution. Temporal
+information and $1$-hourly resolved $10$\,min mean observations of various
+meteorological quantities are used as predictor variables, including wind
+direction, wind speed, temperature, air pressure and humidity, all lagged by
+one or three hours according to the respective forecasting step. The
+meteorological variables are measured either directly at the airports or within
+their vicinities. For Innsbruck, measurements at the airport and along the
+intersecting valleys are used, whereas, for Vienna, measurements at the airport
+and within its vicinity of approximately 30\,km are used.
+Figure~\ref{fig:overview} provides a topographical overview of the airports and
+their surrounding areas with the station sites employed in this study. In
+addition, we use derived quantities such as 3-hourly means, minima and maxima,
+as well as 1-and 3-hourly temporal changes and spatial differences towards the
+airport of the respective quantities. An overview of the employed data sets can
+be found in Table~\ref{tab:data}.
 
@@
 to December 2018. After first eliminating predictor variables with more than
-5\% missing values and then time points with any missing observations, the data set
-consists of 41\,979 time points and 260 covariates for Innsbruck, and of
+5\% missing values and then time points with any missing observations, the data
+set consists of 41\,979 time points and 260 covariates for Innsbruck, and of
 38\,985 time points and 494 covariates for Vienna, respectively.
@@
 All scores presented in the next section are computed out-of-sample based on
-five years of data. For the persistency model only dates prior to the
-time of interest are used and the validation is performed rolling over all
+five years of data. For the persistency model only dates prior to the time of
+interest are used and the validation is performed rolling over all
 observations. For all other models a five-fold cross-validation is employed
@@
 Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
-at forecast steps $+1$\,h and $+3$\,h for Innsbruck (Panels~a,\,c) and
-Vienna (Panels~b,\,d). The scores are aggregated over the five validation
-years, yielding yearly mean values for every hour per calendar month, with a
-lower score indicating better performance. The circular regression forest overall provides
-the best predictive performance, followed by the circular regression tree and
-the persistency model for both stations at both forecast steps; except for the
-$3$\,h~forecasts at Innsbruck where the persistency model is outperformed by
-all others. In comparison to the circular regression tree and forest, for both
-stations and forecast steps, the climatological model and the linear model show
-clearly higher CRPS values and hence a lower predictive performance.
+at forecast steps $+1$\,h and $+3$\,h for Innsbruck (Panels~a,\,c) and Vienna
+(Panels~b,\,d). The scores are aggregated over the five validation years,
+yielding yearly mean values for every hour per calendar month, with a lower
+score indicating better performance. The circular regression forest overall
+provides the best predictive performance, followed by the circular regression
+tree and the persistency model for both stations at both forecast steps; except
+for the $3$\,h~forecasts at Innsbruck where the persistency model is
+outperformed by all others. In comparison to the circular regression tree and
+forest, for both stations and forecast steps, the climatological model and the
+linear model show clearly higher CRPS values and hence a lower predictive
+performance.
 %%
@@
 model due to its strong vulnerability to abrupt wind shifts. On the contrary,
-at Vienna smaller and less abrupt changes in the wind direction as well as less pronounced 
-wind regimes are observed due to the less mountainous surrounding. This seems to 
-weaken the predictive performance of the climatological and linear models, and to 
-reduce the performance variability for all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).
+at Vienna smaller and less abrupt changes in the wind direction as well as less
+pronounced wind regimes are observed due to the less mountainous surrounding.
+This seems to weaken the predictive performance of the climatological and
+linear models, and to reduce the performance variability for all models
+(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).
 
@@
 a lower performance for longer lead times; this is due to the lower information
-content of 3-hourly instead of 1-hourly lagged response values employed
-as covariates in the persistency model. The circular regression tree and
-forest seem to partially compensate for the lower skill of the lagged response
-values by other covariates, hence their predictive performance only slightly
-decreases for the longer lead time. This compensation is especially evident for
+content of 3-hourly instead of 1-hourly lagged response values employed as
+covariates in the persistency model. The circular regression tree and forest
+seem to partially compensate for the lower skill of the lagged response values
+by other covariates, hence their predictive performance only slightly decreases
+for the longer lead time. This compensation is especially evident for
 Innsbruck, where the performance difference between the persistency model and
@@
 values indicate an improvement in the predictive performance over the
-reference. For all setups, the circular regression forest has the highest skill scores
-with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and Vienna,
-respectively. As discussed for Figure~\ref{fig:boxplot_crpsraw}, this
+reference. For all setups, the circular regression forest has the highest skill
+scores with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and
+Vienna, respectively. As discussed for Figure~\ref{fig:boxplot_crpsraw}, this
 improvement over the climatological model is lower for Innsbruck due to the low
@@
 forests are established by coupling model-based recursive partitioning with the
-von~Mises distribution. By separating the data into more homogeneous
-subgroups, possible difficulties in circular regression are bypassed as
-covariates are solely considered for splitting and group-specific models are
-fitted without further covariates. In addition, by specifying the von~Mises
-distribution for each node and allowing for splits in both distribution
-parameters $\mu$ and $\kappa$, fully probabilistic forecasts are provided.
-
-The performance of the novel circular regression trees and forests is assessed in a case
-study for shortterm probabilistic wind direction forecasting at two airports
-with different site characteristics. As benchmark models, probabilistic
-climatology and persistency models, as well as a state-of-the-art circular
-GLM-type model are evaluated based on proper scoring rules. In summary, the
-circular regression trees and forests have the highest predictive performance in this
-setting. For cases without changes in the wind regime, lagged response values
-provide already highly skillful estimates leading to a good performance of the
-persistency model as observed for shortterm wind direction forecasts in this
-study. While in these cases the trees and forests also benefit from the highly
-informative lagged response, they can compensate for a lower information of
-this covariate by incorporating other quantities and possible interactions of
-these, contrary to the persistency model (see
+von~Mises distribution. By separating the data into more homogeneous subgroups,
+possible difficulties in circular regression are bypassed as covariates are
+solely considered for splitting and group-specific models are fitted without
+further covariates. In addition, by specifying the von~Mises distribution for
+each node and allowing for splits in both distribution parameters $\mu$ and
+$\kappa$, fully probabilistic forecasts are provided.
+
+The performance of the novel circular regression trees and forests is assessed
+in a case study for shortterm probabilistic wind direction forecasting at two
+airports with different site characteristics. As benchmark models,
+probabilistic climatology and persistency models, as well as a state-of-the-art
+circular GLM-type model are evaluated based on proper scoring rules. In
+summary, the circular regression trees and forests have the highest predictive
+performance in this setting. For cases without changes in the wind regime,
+lagged response values provide already highly skillful estimates leading to a
+good performance of the persistency model as observed for shortterm wind
+direction forecasts in this study. While in these cases the trees and forests
+also benefit from the highly informative lagged response, they can compensate
+for a lower information of this covariate by incorporating other quantities and
+possible interactions of these, contrary to the persistency model (see
 Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
@@
 For the specific task of wind forecasting, the wind direction is often only
-relevant if the wind speed is sufficiently high. Hence, it is of interest
-to account for both quantities simultaneously, e.g., by considering a 
-bivariate normal distribution for the wind vectors (from which wind speed
-and wind direction can be obtained). The parameters of this bivariate normal
-distribution could then be linked to the available covariate using 
-an additive regression framework \citep[as proposed by][]{Lang+Mayr+Stauffer:2019}
-or using a tree-based approach, similar to the one proposed in this paper.
-Moreover, a rather different approach for a combined response of wind speed
-and wind direction would be a two-step or hurdle model: In the first step
-this could build on the truncated normal model of
-\cite{Thorarinsdottir+Gneiting:2010} to capture wind speed. And in a second
-step a circular wind direction model is leveraged provided that a certain
-hurdle for the wind speed is crossed.
-
-Another possible improvement for obtaining more
-parsimonious circular regression trees is to consider splitting circular covariates into
-two circle segments by searching two split points simultaneously rather than
-sequentially at different depths. While this might slightly improve the
-predictive performance of circular regression trees, this should not affect the
+relevant if the wind speed is sufficiently high. Hence, it is of interest to
+account for both quantities simultaneously, e.g., by considering a bivariate
+normal distribution for wind vectors (from which wind speed and wind direction
+can be obtained). The parameters of this bivariate normal distribution could
+then be linked to available covariates using an additive regression framework
+\citep[as proposed by][]{Lang+Mayr+Stauffer:2019} or using a tree-based
+approach, similar to the one proposed in this paper. Moreover, a rather
+different approach for a combined response of wind speed and wind direction
+would be a two-step or hurdle model: In the first step this could build on the
+truncated normal model of \cite{Thorarinsdottir+Gneiting:2010} to capture wind
+speed; in the second step a circular wind direction model is leveraged given
+that a certain hurdle for the wind speed is crossed.
+
+Another possible improvement for obtaining more parsimonious circular
+regression trees is to consider splitting circular covariates into two circle
+segments by searching two split points simultaneously rather than sequentially
+at different depths. While this might slightly improve the predictive
+performance of circular regression trees, this should not affect the
 performance of the forests, as they consist of very large trees with many
@@
 possible covariates. Moreover, the case study shows that building ensembles of
-circular regression trees can even improve the forecasting performance, as the resulting
-forests allow for modeling smooth effects and stabelize the model.
+circular regression trees can even improve the forecasting performance, as the
+resulting forests allow for modeling smooth effects and stabelize the model.
 
@@
 The corresponding implementation of the proposed methodology for circular
-regression trees and forests is provided in the \textsf{R} package \textbf{circtree} 
-(version~\Sexpr{packageVersion("circtree")}).
-% The proposed circular regression tree and forest models are implemented in the \textsf{R}
-% package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). 
-The package is based on the \textbf{disttree} package
+regression trees and forests is provided in the \textsf{R} package
+\textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The package is
+based on the \textbf{disttree} package
 (version~\Sexpr{packageVersion("disttree")}) which applies the main tree
@@
 \label{app:algorithm}
-This section provides a more detailed overview on the permutation-test-based CTree algorithm 
-\citep{Hothorn+Hornik+Zeileis:2006}, 
-specifically for circular data as applied for building circular regression trees and forests 
-presented in this case study. 
-An alternative tree-building framework is provided by the MOB algorithm, which is based on 
-M-fluctuation tests \citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.
-
-%\subsection*{CTree}
-In the following, the testing and splitting strategy is described for the root node of the tree
-which contains the entire learning sample.
-%consisting of circular observations denoted by $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
-For a complete tree model, the same procedure is applied iteratively to all resulting child nodes 
-with the corresponding subsamples.
-
-First, employing the von~Mises distribution, a distributional model $\text{vM}(y; \mu,\kappa)$ 
-is fitted to the learning sample of circular observations $\{y_i\}_{i = 1,\ldots,n}$ as explained 
-in Section~\ref{sec:prob_circ}.
-In a next step, a goodness-of-fit measurement %of $\text{vM}(Y; \mu,\kappa)$ 
-is obtained for each parameter and each observation by evaluating the score 
+This section provides a more detailed overview on the permutation-test-based
+CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006}, specifically for circular
+data as applied for building circular regression trees and forests presented in
+this case study. An alternative tree-building framework is provided by the MOB
+algorithm, which is based on M-fluctuation tests \citep[see][for more
+details]{Zeileis+Hothorn+Hornik:2008}.
+
+In the following, the testing and splitting strategy is described for the root
+node of the tree which contains the entire learning sample. For a complete
+tree model, the same procedure is applied iteratively to all resulting child
+nodes with the corresponding subsamples.
+
+First, employing the von~Mises distribution, a distributional model
+$\text{vM}(y; \mu,\kappa)$ is fitted to the learning sample of circular
+observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in
+Section~\ref{sec:prob_circ}. In a next step, a goodness-of-fit measurement is
+obtained for each parameter and each observation by evaluating the score
 function $s(\mu, \kappa, y)$ at the estimated location and concentration
-parameter $\hat{\mu}$ and $\hat{\kappa}$.
-To detect dependencies between the resulting scoring matrix
+parameter $\hat{\mu}$ and $\hat{\kappa}$. To detect dependencies between the
+resulting scoring matrix
 \begin{equation}
@@
 \end{equation}
-and each possible split variable $z_l \in \{z_1, \ldots, z_m\}$ a permutation test is applied. 
-In particular, the null hypotheses
-of independence of each split variable and the scores
-%\begin{align}
-%H_0^l:  s(\hat{\mu}, \hat{\kappa}, Y) \qquad \bot \qquad Z_l
-%\end{align}
-is assessed by employing the multivariate linear statistic
+and each possible split variable $z_l \in \{z_1, \ldots, z_m\}$ a permutation
+test is applied. In particular, the null hypotheses of independence of each
+split variable and the scores is assessed by employing the multivariate linear
+statistic
 \begin{equation}
@@
 \end{equation}
-with $s(\hat{\mu}, \hat{\kappa}, y_i) \in \mathbb{R}^{1\times 2}$. 
-For a numeric split variable $z_l$ the transformation function $v_l$
-is simply the identity function $v_l(z_{li}) = z_{li}$ such that $t_l \in \mathbb{R}^2$ 
-as the ``$vec$'' operator converts the matrix of dimension $1 \times 2$ into a $2$ column vector.
-If $z_l$ is a categorical variable with $h$ categories then 
-$v_l(z_{li}) = (\I(z_{li} = 1), \ldots, \I(z_{li} = h))$, hence, $v_l$ returns a unit vector
-of dimension $h$ where the entry $1$ indicates the category of $z_{li}$. 
-In this case the ``$vec$'' operator converts the $h \times 2$ matrix into a $h \cdot 2$ column vector 
-by column-wise combination such that $t_l \in \mathbb{R}^{h \cdot 2}$. 
-If there are any observations with missing values these are not included in the calculation of $t_l$.
-
-To map the %observed 
-multivariate linear statistic $t_l$ onto the real line a univariate test statistic~$c$ is employed, for example in a quadratic form
+with $s(\hat{\mu}, \hat{\kappa}, y_i) \in \mathbb{R}^{1\times 2}$. For a
+numeric split variable $z_l$ the transformation function $v_l$ is simply the
+identity function $v_l(z_{li}) = z_{li}$ such that $t_l \in \mathbb{R}^2$ as
+the ``$vec$'' operator converts the matrix of dimension $1 \times 2$ into a $2$
+column vector. If $z_l$ is a categorical variable with $h$ categories then
+$v_l(z_{li}) = (\I(z_{li} = 1), \ldots, \I(z_{li} = h))$, hence, $v_l$ returns
+a unit vector of dimension $h$ where the entry $1$ indicates the category of
+$z_{li}$. In this case the ``$vec$'' operator converts the $h \times 2$ matrix
+into a $h \cdot 2$ column vector by column-wise combination such that $t_l \in
+\mathbb{R}^{h \cdot 2}$. If there are any observations with missing values
+these are not included in the calculation of $t_l$.
+
+To map the multivariate linear statistic $t_l$ onto the real line a univariate
+test statistic~$c$ is employed, for example in a quadratic form
 \begin{equation}
@@
 \end{equation}
-where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance of $t_l$,
-as derived by \cite{Strasser+Weber:1999} and used for standardization, and $\Sigma_l^+$ 
-is the Moore-Penrose inverse of $\Sigma_l$.
-As an alternative, also a maximum form ($c_{\text{max}}$) can be considered such that
-the maximum of the absolute values of the standardized linear statistic is returned.
-
-The asymptotic conditional distribution of $c(t_{l},\mu_{l},\Sigma_{l})$ 
-is either normal (for $c_{\text{max}}$) or $\chi^2$ (for $c_{\text{quad}}$) owing to 
-the asymptotic conditional distribution of the linear statistic $t_l$ being a multivariate 
-normal with parameters $\mu_l$ and $\Sigma_l$ \citep{Strasser+Weber:1999}.
-With this knowledge at hand, the corresponding $p$-values can be calculated and used
-to select the best splitting variable. A small $p$-value corresponding to 
-$c(t_{l},\mu_{l},\Sigma_{l})$ indicates a strong discrepancy from the assumption of independence 
-between the scores and the split variable $z_l$.
-Therefore, if any of the Bonferroni-adjusted $p$-values is beneath the selected significance level,
-the partitioning variable $z_{l^\ast}$ with the lowest $p$-value is selected as split variable, otherwise 
-no split is performed. This early stopping induced by the significance level is referred to as 
-``pre-pruning'' which is often avoided for forest models by setting the significance level to $1$.
-
-To select the best split point within the already chosen split variable, again, a linear test statistic
-is employed. In particular, for a breakpoint $r$ of the variable $z_{l^{\ast}}$ leading to two subgroups 
-$\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ the discrepancy between score functions in the subgroups
-is measured by evaluating
+where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance
+of $t_l$, as derived by \cite{Strasser+Weber:1999} and used for
+standardization, and $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.
+As an alternative, also a maximum form ($c_{\text{max}}$) can be considered
+such that the maximum of the absolute values of the standardized linear
+statistic is returned.
+
+The asymptotic conditional distribution of $c(t_{l},\mu_{l},\Sigma_{l})$ is
+either normal (for $c_{\text{max}}$) or $\chi^2$ (for $c_{\text{quad}}$) owing
+to the asymptotic conditional distribution of the linear statistic $t_l$ being
+a multivariate normal with parameters $\mu_l$ and $\Sigma_l$
+\citep{Strasser+Weber:1999}. With this knowledge at hand, the corresponding
+$p$-values can be calculated and used to select the best splitting variable. A
+small $p$-value corresponding to $c(t_{l},\mu_{l},\Sigma_{l})$ indicates a
+strong discrepancy from the assumption of independence between the scores and
+the split variable $z_l$. Therefore, if any of the Bonferroni-adjusted
+$p$-values is beneath the selected significance level, the partitioning
+variable $z_{l^\ast}$ with the lowest $p$-value is selected as split variable,
+otherwise no split is performed. This early stopping induced by the
+significance level is referred to as ``pre-pruning'' which is often avoided for
+forest models by setting the significance level to $1$.
+
+To select the best split point within the already chosen split variable, again,
+a linear test statistic is employed. In particular, for a breakpoint $r$ of the
+variable $z_{l^{\ast}}$ leading to two subgroups $\mathcal{B}_{1r}$ and
+$\mathcal{B}_{2r}$ the discrepancy between score functions in the subgroups is
+measured by evaluating
 \begin{equation}
@@
 \end{equation}
-for $q \in \{1,2\}$.
-The breakpoint that leads to the highest discrepancy is then selected as split point as defined by
+for $q \in \{1,2\}$. The breakpoint that leads to the highest discrepancy is
+then selected as split point as defined by
 \begin{equation}
@@
 
-Subsequently, the same testing and splitting procedure is repeated in each of the resulting 
-subgroups until some stopping criterion is reached. Next to the already mentioned 
-significance-level-based stopping criterion, i.e., a minimal $p$-value for the statistical tests,
-also a maximal tree depth or a minimal number of observations in a node can be employed as stopping
-criteria.
+Subsequently, the same testing and splitting procedure is repeated in each of
+the resulting subgroups until some stopping criterion is reached. Next to the
+already mentioned significance-level-based stopping criterion, i.e., a minimal
+$p$-value for the statistical tests, also a maximal tree depth or a minimal
+number of observations in a node can be employed as stopping criteria.
 

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Ready for arxiv submission
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2943 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5333f7d17803dcb8035df4d6925907fde0eafcf7

@@
 data requires the incorporation of its specific feature of periodicity. For
-example, angular data are restricted to an interval such as $[0,2\pi)$ with $0$
+example, angular data are restricted to an interval such as~$[0,2\pi)$ with~$0$
 being equivalent~to~$2\pi$.
@@
 refined this idea by plugging in a link function transforming the linear
-predictor to a restricted interval of length $2\pi$. This generalized linear
+predictor to a restricted interval of length~$2\pi$. This generalized linear
 model (GLM) type approach was further extended by \cite{Fisher+Lee:1992} and
@@
 and south of the airport, all lagged by one hour. Note that in the
-meteorological context wind direction is defined on the scale $[0,360]$ degree
+meteorological context wind direction is defined on the scale $[0,360]$~degree
 and increases clockwise from North ($0$ degree).}
@@
 Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with the
-empirical (gray) and fitted von~Mises (red) wind direction distribution in each
-terminal node. Based on the fitted location parameters $\hat \mu$, the
+empirical (gray) and fitted von~Mises~(red) wind direction distribution in each
+terminal node. Based on the fitted location parameters~$\hat \mu$, the
 subgroups can be distinguished into the following wind regimes: (1)~Up-valley
@@
 %%
-Figure~\ref{fig:densities} depicts von~Mises model for circular data in $[0, 2
+Figure~\ref{fig:densities} depicts a von~Mises model for circular data in $[0, 2
 \pi)$ fitted by maximum likelihood, either using a linearized (left) or
@@
 approaches to circular regression the location parameter $\mu$ depends on
-covariates $\mathbf{z}$ through a link function $g(\cdot)$, circular intercept
+covariates $\mathbf{z}$ through a link function~$g(\cdot)$, circular intercept
 $\mu_0$ and coefficient vector $\mathbf{\beta}$: \begin{equation} \mu = \mu_0 +
@@
 \citet{Fisher+Lee:1992}. They also developed a heteroscedastic version by
-combining two individual GLMs, each for one of the parameters $\mu$ and
-$\kappa$. This provides a first approach to a fully probabilistic regression
+combining two individual GLMs, each for one of the parameters~$\mu$ and~$\kappa$. 
+This provides a first approach to a fully probabilistic regression
 model for circular data, albeit the parameters are not regressed simultaneously
@@
 observations. This can also be expressed by employing weights which indicate
-whether the $i$-th learning observation and the observation $\bm{z}$ belong to
+whether the $i$-th learning observation and the observation~$\bm{z}$ belong to
 the same terminal node: 
@@
   \includegraphics{_plot_circforest_stationmap_combined.pdf}
-  \caption{Overview of the study area for Innsbruck Airports (left) and Vienna International Airport (right). 
+  \caption{Overview of the study area for Innsbruck Airport (left) and Vienna International Airport (right). 
     For Innsbruck, $4$~stations at the airport and $6$~stations along the
@@
 
-The data used in this study consists of five subsequent years from January 2014
-to December 2018. After first eliminating predictor variables with more than
+The data used in this study consists of five subsequent years from January~2014
+to December~2018. After first eliminating predictor variables with more than
 5\% missing values and then time points with any missing observations, the data
-set consists of 41\,979 time points and 260 covariates for Innsbruck, and of
-38\,985 time points and 494 covariates for Vienna, respectively.
+set consists of 41\,979~time points and $260$~covariates for Innsbruck, and of
+38\,985~time points and $494$~covariates for Vienna, respectively.
 
@@
     with $\beta_0$ being a circular intercept, $\beta_\bullet$ the regression
-    coefficients and $g$ the link function $g(x) = 2 \cdot \text{arctan}(x)$. In
+    coefficients and the link function $g(x) = 2 \cdot \text{arctan}(x)$. In
     addition, a constant concentration parameter $\kappa$ is fitted to the full
@@
 (version~\Sexpr{packageVersion("partykit")}). All three packages are available
-on \textsf{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.
+on \textsf{R}-Forge at {\url{https://R-Forge.R-project.org/projects/partykit/}}.
 
@@
 function $s(\mu, \kappa, y)$ at the estimated location and concentration
-parameter $\hat{\mu}$ and $\hat{\kappa}$. To detect dependencies between the
+parameter~$\hat{\mu}$ and $\hat{\kappa}$. To detect dependencies between the
 resulting scoring matrix
@@
 a unit vector of dimension $h$ where the entry $1$ indicates the category of
-$z_{li}$. In this case the ``$vec$'' operator converts the $h \times 2$ matrix
+$z_{li}$. In this case the ``$vec$'' operator converts the $h \times 2$~matrix
 into a $h \cdot 2$ column vector by column-wise combination such that $t_l \in

### 2019-12-23 pkg/circtree/inst/wind_paper/draft/draft.Rnw
corrected typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2944 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6db935ca10449570355f41c7a36cc63f0760daec

@@
 circular regression trees can even improve the forecasting performance, as the
-resulting forests allow for modeling smooth effects and stabelize the model.
+resulting forests allow for modeling smooth effects and stabilize the model.
 

### 2020-01-02 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update url of georg
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2945 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 03adb4d7b24dda3bedff0c29144fd91df8dc3114

@@
   E-mail: \email{Georg.Mayr@uibk.ac.at} \\
-  URL: \url{http://acinn.uibk.ac.at/persons/georg_mayr}\\
+  URL: \url{https://www.uibk.ac.at/acinn/people/georg-mayr}\\
 

### 2020-01-08 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2946 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d409f23c44bb961f6411bdd6d1aa6471cccf40a2

@@
 \Abstract{
+%While circular data occur in a wide range of scientific fields, the methodology
+%for distributional modeling and probabilistic forecasting of circular response
+%variables is rather limited. Most of the existing methods are built on the
+%framework of generalized linear and additive models, which are often
+%challenging to optimize and to interpret. Therefore, we suggest circular
+%regression trees and random forests as an intuitive alternative approach that
+%is relatively easy to fit. Building on previous ideas for trees modeling
+%circular means, we suggest a distributional approach for both trees and forests
+%yielding probabilistic forecasts based on the von~Mises distribution. The
+%resulting tree-based models simplify the estimation process by using the
+%available covariates for partitioning the data into sufficiently homogeneous
+%subgroups so that a simple von~Mises distribution without further covariates
+%can be fitted to the circular response in each subgroup. These circular
+%regression trees are straightforward to interpret, can capture nonlinear
+%effects and interactions, and automatically select the relevant covariates that
+%are associated with either location and/or scale changes in the von~Mises
+%distribution. Combining an ensemble of circular regression trees to a circular
+%regression forest yields a local adaptive likelihood estimator for the
+%von~Mises distribution that can regularize and smooth the covariate effects.
+%The new methods are evaluated in a case study on probabilistic wind direction
+%forecasting at two Austrian airports, considering other common approaches as a
+%benchmark.
 While circular data occur in a wide range of scientific fields, the methodology
@@
 framework of generalized linear and additive models, which are often
-challenging to optimize and to interpret. Therefore, we suggest circular
-regression trees and random forests as an intuitive alternative approach that
-is relatively easy to fit. Building on previous ideas for trees modeling
-circular means, we suggest a distributional approach for both trees and forests
-yielding probabilistic forecasts based on the von~Mises distribution. The
-resulting tree-based models simplify the estimation process by using the
-available covariates for partitioning the data into sufficiently homogeneous
-subgroups so that a simple von~Mises distribution without further covariates
-can be fitted to the circular response in each subgroup. These circular
-regression trees are straightforward to interpret, can capture nonlinear
-effects and interactions, and automatically select the relevant covariates that
-are associated with either location and/or scale changes in the von~Mises
-distribution. Combining an ensemble of circular regression trees to a circular
-regression forest yields a local adaptive likelihood estimator for the
-von~Mises distribution that can regularize and smooth the covariate effects.
-The new methods are evaluated in a case study on probabilistic wind direction
-forecasting at two Austrian airports, considering other common approaches as a
-benchmark.
+challenging to optimize and interpret. Therefore, building on previous ideas
+for trees modeling circular means, we suggest a distributional approach for
+regression trees and random forests yielding probabilistic forecasts based on
+the von~Mises distribution. The resulting tree-based models simplify the
+estimation process by using the available covariates for partitioning the data
+into sufficiently homogeneous subgroups so that a simple von~Mises distribution
+without further covariates can be fitted to the circular response in each
+subgroup. These circular regression trees are straightforward to interpret, can
+capture nonlinear effects and interactions, and automatically select the
+relevant covariates that are associated with either location and/or scale
+changes in the von~Mises distribution. Combining an ensemble of circular
+regression trees to a circular regression forest can regularize and smooth the
+covariate effects. The new methods are evaluated in a case study on
+probabilistic wind direction forecasting at two Austrian airports, considering
+other common approaches as a benchmark.
 }

### 2020-01-09 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Add shortnames for citation and move fig
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2948 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8fde74ac6ef08646c865ce688e9e176f941204ae
Final touch ups for submission
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2950 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5b556e128245df53eb232f7d643994a6dba911e4

@@
-\documentclass[nojss]{jss}
+\documentclass[nojss,shortnames]{jss}
 
@@
 
-\Keywords{circular data, regression trees, random forests, distributional regression, von~Mises distribution}
+\Keywords{circular data, regression trees, random forests, distributional regression, von~Mises distribution, probabilistic forecasting}
 
@@
 being equivalent~to~$2\pi$.
+
+\newpage
 
@@
 
+\subsection{Motivating example}
+
 \begin{figure}[p!]
@@
 \end{figure}
-
-\subsection{Motivating example}
 
@@
 and a concentration parameter $\kappa > 0$ the density of the von~Mises
-distribution for an observation $y \in [0, 2\pi)$ is given by
+distribution for an observation~\mbox{$y \in [0, 2\pi)$} is given by
 %%
@@
 To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response
-$y$, the distribution parameters $\mu$ and $\kappa$ need to be estimated. This
-can be done by maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
+$y$, the distribution parameters~$\mu$ and~$\kappa$ need to be estimated. This
+can be done by maximizing the log-likelihood function~$\ell(\mu, \kappa; y)$
 %%
@@
 which $\bm{z}$ is assigned to the same terminal node as each learning
-observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees, the
+observation~\mbox{$y_i, i \in \{1,\ldots,n\}$}. Hence, for a forest of $T$ trees, the
 weights are calculated via
@@
 covariates $\bm{z}$ whereas $w_i(\bm{z}) = 1$ would correspond to the
-unweighted full-sample estimates and $w_i(\bm{z}) \in \{0, 1\}$ corresponds to
+unweighted full-sample estimates and \mbox{$w_i(\bm{z}) \in \{0, 1\}$} corresponds to
 the subgroup selection from a tree. Thus, circular regression forests can
@@
 \end{equation}
-and each possible split variable $z_l \in \{z_1, \ldots, z_m\}$ a permutation
+and each possible split variable~\mbox{$z_l \in \{z_1, \ldots, z_m\}$} a permutation
 test is applied. In particular, the null hypotheses of independence of each

### 2020-01-26 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2963 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3382c44655d0a3c6388be51f8c9bbc8eda620998

@@
 have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
-investigations) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.
+interactions, see \citealp{Doove+Dusseldorp+VanDeun:2014}) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.
+%% FIXME: add reference to Seibold+Hothorn+Zeileis:2019
+Testing strategies being capable of detecting all kinds of changes including those orthogonal to the 
+mean are also employed in an extension of (generalised) linear model trees introduced by 
+\cite{Seibold+Hothorn+Zeileis:2019}
+which allows for selected model parameters to be kept fixed at a globally estimated value while only 
+changes in the remaining parameters are considered for splitting.
 

### 2020-01-26 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
adapted slides to new structure of simulation results and simulation functions; added slides for continuously changing parameters and full factorial analysis
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2965 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: af9c8b553e4b3bcf72a03ad9d1f029444d4cfbb3

@@
 library("lattice")
+library("latticeExtra")
 library("scales")
 library("colorspace")
-#source("../guidelike.R")
-source("../tests_guideprune.R")
-#source("../ccprune.R")
+#source("../simulation/guidelike.R")
+source("../simulation/tests_guideprune.R")
+#source("../simulation/ccprune.R")
 
-load("../draft/stump.rda")
-sim_stump <- simres$res
-rm(simres)
-
-load("../draft/tree.rda")
-sim_tree <- simres$res
-rm(simres)
-
-load("../draft/pruning.rda")
-pr <- simres$res
-## extract data for xyplot comparing ARI and ARI_p
-pr <- pr[,c(1,2,7,8,9,12,15)]
-## extended data set with postpruning results being represented as separated tests
-prex <- pr[,c(1,2,3,4,6)]
-prex <- rbind(prex,prex)
-
-prex$test <- factor(prex$test,
-                    levels = c("ctree","mfluc",
-                               "guide_sum_12", "guide_sum_1_cor",
-                               "ctree_p","mfluc_p",
-                               "guide_sum_12_p", "guide_sum_1_cor_p"),
-                    labels = c("CTree","MOB",
-                               "GUIDE_scores", "GUIDE",
-                               "CTree_p","MOB_p",
-                               "GUIDE_scores_p", "GUIDE_p"))
-
-for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
-  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
-  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
-  if(prex$test[i] == "GUIDE_scores") prex$test[i] <- "GUIDE_scores_p"
-  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
-}
-
-prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
-prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
-
-load("../draft/3way_xi00_delta03.rda")
-load("../draft/3way_xi08_delta1.rda")
-d00$xi <- 0
-d08$xi <- 0.8
-d3way <- rbind(d00, d08)
-d3way$xi <- factor(d3way$xi, levels = c(0,0.8), labels = c(0,0.8))
+load("../simulation/results/sim_stump.rda")
+load("../simulation/results/sim_tree.rda")
+load("../simulation/results/sim_3way.rda")
 
 ## HCL palette
-pal_dgp_light <- qualitative_hcl(3, "Set 2")
+pal <- qualitative_hcl(5, "Dark 3")
+names(pal) <- c("CTree", "MOB", "GUIDE", "GUIDE+scores", "CTree+max")
+
+## HCL palette for illustrations of data generating process
+pal_dgp_light <- qualitative_hcl(3, "Dark 3", c1 = 50, l1 = 75)
 pal_dgp <- qualitative_hcl(3, "Dark 3")
 
-## HCL palette (light)
-pal_results <- qualitative_hcl(4, "Dark 3")
+## HCL palette for tree panels
+paltrees <- rgb(c(0.9, 1), c(0.9, 1), c(0.9, 1))
 
@@
 class(tp) <- "grapcon_generator"
+
 @
@@
      tp_args = list(FUN = identity, width = 15), 
-     ip_args = list(fill = paltrees[c(2, 3)]),
+     ip_args = list(fill = paltrees[c(1, 2)]),
      drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
@@
 <<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
-d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
-               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)
+d <- dgp_stump(nobs = nobs, changetype= "abrupt",
+               variation = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)
 
@@
 <<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
-d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
-               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
+d <- dgp_stump(nobs = nobs, changetype= "abrupt", 
+               variation = "all", delta = delta, xi = xi, sigma = sigma)
 par(mar = c(4,4,2,4))
@@
 
-par.settings <- list(superpose.symbol = list(col = pal_results, 
-                                             fill = pal_results), 
-                     superpose.line = list(col = pal_results))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
 
-library("latticeExtra")
-useOuterStrips(xyplot(prop_Tsplit ~ delta | vary_beta + xi, groups = ~ test, 
+useOuterStrips(xyplot(prop_Tsplit ~ delta | variation + xi, groups = ~ test, 
                       data = subdataxi, 
-                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
-                        only_intercept == FALSE,# & xi == 0, 
+                      subset = changetype == "abrupt", 
                       type = "l", lty = c(1,1,1,1),
+                      lwd = 2,
                       scales=list(x=list(at=seq(1,11,2))),
@@
                                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                               col = pal_results, lwd = 1)),
+                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                       layout= c(3,2),
@@
                       xlab = expression(delta),
-                      ylab = expression("Proportion of selecting"~Z[1]),
+                      ylab = expression("Selection probability of"~Z[1]),
                       par.settings = par.settings),
 
-               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
-                                                     expression("vary"~beta[0]), 
-                                                     expression("vary"~beta[1])),
+               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
+                                                     expression("varying"~beta[0]), 
+                                                     expression("varying"~beta[1])),
                                    bg = "gray90"),
-               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0~" (50%)"),                                                              expression(xi == 0.8~" (90%)" )),
                                        bg = "gray90"))
@@
 \end{figure}
-
 \end{frame}
 
-%\begin{frame}%[fragile]
-%\frametitle{Simulation 1: Full factorial analysis (vary both)}
+\begin{frame}%[fragile]
+\frametitle{Simulation 1: Continuously changing parameters}
 
-%\begin{figure}%[t]
-%\setkeys{Gin}{width=0.8\linewidth}
-%<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
-%xyplot(pval_z1 ~ cat | xi, 
-%       groups = interaction(bin, res_scores), 
-%       data = d3way, 
-%       ylim = c(0,0.5),
-%       type = c("a"), 
-%       col = c("red", "blue", "red", "blue"),
-%       lty = c(1,1,2,2),
-%       xlab = "Categorization",
-%       ylab = expression(p~"-value of"~Z[1]),
-%       #index.cond = list(c(1,2,3,4)),
-%       strip = strip.custom(factor.levels = c(expression(xi==0~"and"~delta==0.3), 
-%                                              expression(xi==0.8~"and"~delta==1))),
-%       key = list(text = list(levels(interaction(d3way$bin, d3way$res_scores))),
-%                  lines = list(lty = c(1,1,2,2), 
-%                               type = "l",
-%                               col = c("red", "blue", "red", "blue"))))
-%@
-%\end{figure}
-%\end{frame}
+\begin{figure}%[t]
+\setkeys{Gin}{width=0.8\linewidth}
+<<contbeta, fig=TRUE, echo=FALSE, height=4, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max","mfluc","guide_sum_12","guide_sum_1_cor"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_max","mfluc","guide_sum_12","guide_sum_1_cor"),
+                       labels = c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE"))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE")], 
+                                             fill = pal[c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE")]))
+
+xyplot(prop_Tsplit ~ delta | variation, groups = ~ test, 
+       data = subdata, 
+       subset = changetype == "continuous", 
+       scales=list(x=list(at=seq(1,11,2))),
+       type = "l", 
+       layout= c(3,1),
+       lwd = 2,
+       key =  list(text = list(levels(subdata$test)),
+                   lines = list(lty = c(1,1,1,1,1), 
+                                type = "l", 
+                                col = pal[c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE")], 
+                                lwd = 2)), 
+       xlab = expression(delta),
+       ylab = expression("Selection probability of"~Z[1]),
+       index.cond = list(c(2,3,1)),
+       strip = strip.custom(factor.levels = c(expression("varying"~beta[0]~" and "~beta[1]),
+                                              expression("varying"~beta[0]), 
+                                              expression("varying"~beta[1])),
+                            bg = "gray90"),
+       par.settings = par.settings)
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}%[fragile]
+\frametitle{Simulation 1: Full factorial analysis of building blocks}
+
+\begin{figure}%[t]
+\setkeys{Gin}{width=0.8\linewidth}
+<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
+xyplot(pval_z1 ~ cat | xi, 
+       groups = interaction(bin, res_scores), 
+       data = sim_3way, 
+       ylim = c(0,0.5),
+       type = c("a"), 
+       col = c(pal[1], pal[3], pal[1], pal[3]),
+       lty = c(1,1,2,2),
+       lwd = 2,
+       layout = c(2,1),
+       scales = list(alternating=FALSE),
+       xlab = "Categorization",
+       ylab = expression("p-value of"~Z[1]),
+       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
+                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
+                            bg = "gray90"),
+       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
+                  columns = 2,
+                  lines = list(lty = c(0,1,2,0,1,1), 
+                               type = "l", lwd = 2,
+                               col = c("black","black","black","black", pal[1], pal[3]))))
+
+@
+\end{figure}
+\end{frame}
 
@@
 <<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=
-
 # d <- dgp_tree(nobs = 250)
 set.seed(7)
-d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
-              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.5)
+d <- dgp_tree(nobs = 250, xi = 0, variation = "all", 
+              changetype= "abrupt", sigma = 0.5)
 
@@
 \end{figure}
-
 \end{frame}
@@
 <<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
-subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
-                                        "guide_sum_12", "guide_sum_1_cor"))
-#subdata <- subset(subdata, xi %in% c(0,0.5,0.8))
-subdata$test <- factor(subdata$test, 
-                       levels = c("ctree", "mfluc",
-                                  "guide_sum_12", "guide_sum_1_cor"),
-                       labels = c("CTree","MOB",
-                                  "GUIDE_scores", "GUIDE"))
-
-par.settings <- list(superpose.symbol = list(col = pal_results, 
-                                             fill = pal_results), 
-                     superpose.line = list(col = pal_results))
-
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
 xyplot(ari ~ delta | xi, groups = ~ test, 
-       data = subdata, 
-       type = "l",
-       lty = c(1,1,1,1),
-       key =  list(text = list(levels(subdata$test)),
+       data = sim_tree[sim_tree$pruning == "pre",], 
+       type = "l", lty = c(1,1,1,1),
+       lwd = 2,
+       key =  list(text = list(levels(sim_tree$test)[1:4]),
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal_results, lwd = 1)),
+                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
        layout= c(4,1),
@@
        ylab = "Adjusted Rand Index",
+       #index.cond = list(c(2,3,1),c(2,1)),
+       par.settings = par.settings,
        strip = strip.custom(factor.levels = c(expression(xi==0), 
@@
                                               expression(xi==0.8)),
-                            bg = "gray90"),
-       par.settings = par.settings)
+                            bg = "gray90"))
 
@@
 <<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
-par.settings <- list(superpose.symbol = list(col = pal_results, 
-                                             fill = pal_results), 
-                     superpose.line = list(col = pal_results))
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
+                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
 
-#prex <- subset(prex, xi %in% c(0,0.5,0.8))
+# add new variable that combines test and pruning strategy
+sim_tree$test_pruning <- paste(sim_tree$test, sim_tree$pruning, sep="_")
+sim_tree$test_pruning <- factor(sim_tree$test_pruning,
+                                levels = c("CTree_pre", "MOB_pre", "GUIDE+scores_pre", "GUIDE_pre",
+                                           "CTree_post", "MOB_post", "GUIDE+scores_post", "GUIDE_post"),
+                                labels = c("CTree_pre", "MOB_pre", "GUIDE+scores_pre", "GUIDE_pre",
+                                           "CTree_post", "MOB_post", "GUIDE+scores_post", "GUIDE_post"))
 
-xyplot(ari ~ delta | xi, groups = ~ test, 
-       data = prex, 
+xyplot(ari ~ delta | xi, groups = ~ test_pruning, 
+       data = sim_tree, 
        type = "l", lty = c(1,1,1,1,2,2,2,2),
-       key =  list(text = list(levels(prex$test)[1:4]),
+       lwd = 2,
+       key =  list(text = list(levels(sim_tree$test)[1:4]),
                    lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                col = pal_results, lwd = 1)),
+                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
        layout= c(4,1),

### 2020-01-27 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
added slide for maximally selecting split variables
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2966 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 49e2087ff18d766596aa64aebdfa36202fd2b804

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
-\usetheme[license, url]{uibk}
+\usetheme[url,license]{uibk}
 
@@
                       scales=list(x=list(at=seq(1,11,2))),
+                      key =  list(text = list(c(levels(subdataxi$test)," ")),
+                                  lines = list(lty = c(1,1,1,1,0), type = "l", pch = 1, cex = 0.8,
+                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
+                      layout= c(3,2),
+                      index.cond = list(c(2,3,1),c(2,1)),
+                      xlab = expression(delta),
+                      ylab = expression("Selection probability of"~Z[1]),
+                      par.settings = par.settings),
+
+               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
+                                                     expression("varying"~beta[0]), 
+                                                     expression("varying"~beta[1])),
+                                   bg = "gray90"),
+               strip.left=strip.custom(factor.levels = c(expression(xi == 0~" (50%)"),                                                              expression(xi == 0.8~" (90%)" )),
+                                       bg = "gray90"))
+
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}%[fragile]
+\frametitle{Simulation 1: Maximum vs.\ linear selection}
+
+\begin{figure}
+\setkeys{Gin}{width=0.8\linewidth}
+<<scores_stump_max, fig=TRUE, echo=FALSE, height=5, width=8>>=
+subdata <- subset(sim_stump, test %in% c("ctree","ctree_max","mfluc","guide_sum_12","guide_sum_1_cor"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree", "ctree_max","mfluc","guide_sum_12","guide_sum_1_cor"),
+                       labels = c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE"))
+
+subdataxi <- subset(subdata, xi %in% c(0,0.8))
+
+par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE")], 
+                                             fill = pal[c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE")]), 
+                     superpose.line = list(col = pal[c("CTree","CTree+max","MOB","GUIDE+scores","GUIDE")]))
+
+useOuterStrips(xyplot(prop_Tsplit ~ delta | variation + xi, groups = ~ test, 
+                      data = subdataxi, 
+                      subset = changetype == "abrupt", 
+                      type = "l", lty = c(1,1,1,1,1),
+                      lwd = 2,
+                      scales=list(x=list(at=seq(1,11,2))),
                       key =  list(text = list(levels(subdataxi$test)),
-                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
-                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
+                                  lines = list(lty = c(1,1,1,1,1), type = "l", pch = 1, cex = 0.8,
+                                               col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                       layout= c(3,2),

### 2020-02-09 pkg/partykit/inst/guideliketest/draft/draft.Rnw
first revisions based on reviewer's comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2976 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 43a79bfd0fe422b02dc3879c75c6800ba912bd11

@@
 \usepackage{enumitem}
-\usepackage{color,soul}
+\usepackage{color,xcolor,soul}
 
@@
 
-Therefore, in this paper the focus will be put on that first step of tree algorithms,
-i.e., the task of selecting the best split variable in a given (sub)sample. In particular, the approach of the 
-GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
-by investigating the building blocks of their testing strategies in which they differ:
+\fixme{``Why CTree, GUIDE, and MOB?''}
+
+\hlc{Therefore, in this paper the focus will be put on that first step of tree algorithms,
+i.e., the task of selecting the best split variable in a given (sub)sample. 
+For this purpose, three leading unbiased recursive partitioning algorithms are considered, 
+namely GUIDE~(Google Scholar: 484), CTree~(2444) and MOB~(421). Moreover, through this 
+selection the approach of QUEST (1306) is included as well due to GUIDE having been introduced 
+as a descendant/extension of QUEST (and others). %\citep{} User Manual for GUIDE ver. 31.0.
+In particular, GUIDE, CTree, and MOB are compared by investigating the building blocks of 
+their testing strategies in which they differ:}
 (1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
@@
 
+\subsection{Selecting a split point}
+\fixme{``Why is the choice of the split point not part of this paper?''}
+
+\hlc{GUIDE and MOB, which are both model-based approaches, employ the model estimated 
+in the considered node for an exhaustive search of the best split point over all possible 
+split points in the already selected split variable.
+CTree has not been developed specifically for model-based partitioning, hence, a 
+different approach to selecting a split point is suggested in the original paper. 
+However, as an adaption of CTree employing models for the split variable selection 
+is considered in this study, it is also combined with the same exhaustive search 
+strategy as GUIDE and MOB.}
 %\newpage
+
 \subsection{Pruning}
@@
 
-In the unbiased recursive partitioning literature this post-pruning approach is also used frequently
-\citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
-from the association tests are only employed for selecting the split variable on a unified scale.
+In the unbiased recursive partitioning literature this post-pruning approach is also used 
+frequently \citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} 
+and the $p$-values from the association tests are only employed for selecting the split 
+variable on a unified scale.
 However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
-so-called \emph{pre-pruning} strategy which stops growing the tree as soon as no significant association
-can be found in a given subgroup. This approach is the default in CTree and also in
+so-called \emph{pre-pruning} strategy which stops growing the tree as soon as no significant
+association can be found in a given subgroup. This approach is the default in CTree and also in
 MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
-post-pruning in model-based partitioning is to use information criteria such as AIC (Akaike information
-criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.
+post-pruning in model-based partitioning is to use information criteria such as AIC (Akaike 
+information criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.
 
 Clearly, for inference-based pre-pruning it is crucial that the association tests employed for
-split selection work well as statistical significance test, i.e., conform with their nominal size
-and have high power. In contrast, when using post-pruning (either based on cross-validation or
-information criteria) it might not be as crucial that the significance test works well and has
-high power.
+split selection work well as statistical significance test, i.e., conform with their nominal
+size and have high power. In contrast, when using post-pruning (either based on 
+cross-validation or information criteria) it might not be as crucial that the significance 
+test works well and has high power.
 
 Due to these considerations we first evaluate the significance tests underlying CTree, MOB,
-and GUIDE by themselves, i.e., without growing an actual tree in combination with a pruning strategy.
-Subsequently we combine the tests with a cost-complexity post-pruning approach in order to assess
-whether shortcomings of the tests are mitigated by pruning.
+and GUIDE by themselves, i.e., without growing an actual tree in combination with a pruning 
+strategy. Subsequently we combine the tests with a cost-complexity post-pruning approach in 
+order to assess whether shortcomings of the tests are mitigated by pruning.
 
@@
 Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
-of the type of model fits, tests, and pruning strategy employed to
-grow the tree. Table~\ref{tab:combinations} provides an overview of the default
-combinations.
+of the type of model fits, tests, and pruning strategy employed to grow the tree. 
+Table~\ref{tab:combinations} provides an overview of the default combinations.
 However, as discussed above, subsequent publications have emphasized that all three
@@
 
+\fixme{``Do the building blocks cover all possible options?''}
+
 Note that there are further differences in the testing strategies between the three algorithms, e.g.,
@@
 discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.
+\hlc{Therefore, in this study the focus is put on the selected building blocks as they are assumed 
+(and have shown) to have the highest influence on the difference in performance between the testing 
+strategies.}
 
@@
 
-% To illustrate the effects of the building blocks in the unifying framework presented in 
-% Section~\ref{sec:unify} different combinations of them are investigated for the first
-% simulation scenario (stump) in Sections~\ref{sec:scores} and~\ref{sec:3way}. 
-% Then the testing strategies are evaluated on the more
-% complex tree structure of the second scenario in Section~\ref{sec:tree}. 
-% The corresponding results are shown in the following plots.
-
-\newpage
+%\newpage
 \subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
@@
 is due to dichotomizing the scores at zero and due to categorizing the split variables
-which are investigated in more detail in the following section.
+which are investigated in more detail in the next section.
+
+\fixme{Type I error}
+
+Figure~\ref{fig:scores_stump} \hlc{also shows that for $\delta = 0$, i.e., no difference
+in the parameters and consequently a random split variable selection, the probability of 
+selecting $Z_1$ as split variable is (close to)~$0$ (with the employed Bonferroni-adjustment of 
+$p$-values keeping this value particularly low).
+Even though it has to be noted that the number of replications in this simulation study is of
+course limited, the results for this setting indicate that all testing strategies conform well with 
+the predefined level of significance $\alpha = 0.05$ as the probability of splitting in any 
+of the available covariates is between 4 and 5~percent for $\delta = 0$ 
+(the exact values can be found in} Table~\ref{tab:typeIerror} in Appendix~\ref{app:typeIerror})
+\hlc{with the tests for structural changes being slightly more conservative} 
+\citep[see, e.g.,][for more details on this typical behaviour]{Hothorn+Zeileis:2008, Zeileis+Hothorn:2013, Hansen:1997}.
 
@@
 interactions, see \citealp{Doove+Dusseldorp+VanDeun:2014}) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.
-%% FIXME: add reference to Seibold+Hothorn+Zeileis:2019
 Testing strategies being capable of detecting all kinds of changes including those orthogonal to the 
@@
 Foundation, grant number 200021\_184603.
+Lisa Schlosser received a PhD scholarship granted from the University of Innsbruck.
 
@@
 of building blocks are listed in Table~\ref{tab:combinations_appendix}.
-%Each suffix indicates a change of the original form of the considered testing strategy.
-%In particular, it relates to the transformations of the data before applying a statistical
-%test as represented by the different building blocks. In that way,
-%``+res'' refers to using only residuals (instead of all model scores),
-%``+scores'' to using all model scores (instead of only residuals),
-%``+cat'' to applying a categorization of split variables,
-%``+bin'' to applying a dichotomization of scores/residuals.
 
@@
 
+\section{Type I error}
+\label{app:typeIerror}
+<<typeIerror, echo=FALSE, results=hide>>=
+# Table: Type I error
+subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
+                                         "guide_sum_12", "guide_sum_1_cor"))
+subdata$test <- factor(subdata$test, 
+                       levels = c("ctree","mfluc",
+                                  "guide_sum_12", "guide_sum_1_cor"),
+                       labels = c("CTree","MOB",
+                                  "GUIDE+scores", "GUIDE"))
+
+sd <- subset(subdata, changetype == "abrupt" & delta == 0 & xi %in% c("0","0.8"))
+sd$xi <- factor(sd$xi)
+# sd[, c("test","xi", "variation", "prop_split")]
+
+xtbl <- xtabs(prop_split ~ (xi + test) + variation, data = sd)
+
+# sd$prop_split[sd$test == "CTree" & sd$variation == "all" & sd$xi == "0"]
+@
+
+\begin{table}[h]
+\centering
+\begin{tabular}{ l l c c c }
+\hline\noalign{\smallskip} 
+Split point  & Strategy     & Varying $\beta_0$  & Varying $\beta_1$ & Varying $\beta_0$ and $\beta_1$ \\ 
+\noalign{\smallskip}\hline\noalign{\smallskip} 
+$\xi = 0$   & CTree         &  0.05 & 0.05 & 0.05 \\ 
+($50$\%)    & MOB           &  0.04 & 0.04 & 0.04 \\
+            & GUIDE+scores  &  0.05 & 0.05 & 0.05 \\
+            & GUIDE         &  0.04 & 0.04 & 0.04 \\
+\noalign{\smallskip}\hline\noalign{\smallskip} 
+$\xi = 0.8$ & CTree         &  0.05 & 0.05 & 0.05 \\
+($90$\%)    & MOB           &  0.04 & 0.04 & 0.04 \\
+            & GUIDE+scores  &  0.05 & 0.05 & 0.05 \\
+            & GUIDE         &  0.04 & 0.04 & 0.04 \\
+\noalign{\smallskip}\hline
+\end{tabular}
+\caption{Probability of splitting in any of the available split variables,
+i.e., probability of any of the available split variables showing a $p$-value
+below the predefined level of significance $\alpha = 0.05$. 
+The testing strategies CTree, MOB, GUIDE, and GUIDE+scores are considered in 
+the ``stump'' scenario. Probabilities are estimated for effect size $\delta = 0$ 
+(i.e., no difference in the parameters) with 100 replications of 250 observations.
+The true split point $\xi$ in $Z_1$ is either the median~$0$~(top row) or the 90\%~quantile~$0.8$~(bottom row)
+for either varying intercept~$\beta_0$, varying slope~$\beta_1$, or both coefficients varying.
+\label{tab:typeIerror}}
+\end{table}
+
 \end{appendix}

### 2020-02-09 pkg/partykit/inst/guideliketest/draft/draft.Rnw
adapted the 'why' modifications
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2977 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2ea030793b2ad1b12fccdaf48a906a7db00cd316

@@
 Therefore, selecting a split variable in a first step and then searching for the 
-best split point only within this variable in a separated second step is a more 
-promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
+best split point only within this variable in a separate second step is a more 
+promising strategy. \hlc{The most popular algorithms adopting such an unbiased recursive
+partitioning strategy are} QUEST \citep{Loh+Shih:1997},
 %% CHAID \citep{Kass:1980},
 GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
-and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
-For the first step of selecting a split variable they all share the same basic concept of 
-choosing the covariate which shows the highest association to the response variable 
-based on $p$-values provided by a statistical test. While QUEST
+and MOB \citep{Zeileis+Hothorn+Hornik:2008}. \hlc{QUEST and CTree have been suggested as
+nonparametric decision trees initially and have acquired more than 1,300 and 2,400 citations
+on Google Scholar, respectively. GUIDE and MOB have been designed for general model-based
+trees and have both received more than 400 citations.
+All of these share the same basic concept of selecting that covariate for splitting which shows the
+highest association to the response variable as captured by the $p$-value in the corresponding
+statistical test.} While QUEST
 %%, CHAID,
-and GUIDE employ 
+and its successor GUIDE employ 
 statistical significance tests for contingency tables, CTree applies permutation tests in 
@@
 
-\hlc{Therefore, in this paper the focus will be put on that first step of tree algorithms,
+Therefore, in this paper the focus will be put on that first step of tree algorithms,
 i.e., the task of selecting the best split variable in a given (sub)sample. 
-For this purpose, three leading unbiased recursive partitioning algorithms are considered, 
-namely GUIDE~(Google Scholar: 484), CTree~(2444) and MOB~(421). Moreover, through this 
-selection the approach of QUEST (1306) is included as well due to GUIDE having been introduced 
-as a descendant/extension of QUEST (and others). %\citep{} User Manual for GUIDE ver. 31.0.
-In particular, GUIDE, CTree, and MOB are compared by investigating the building blocks of 
-their testing strategies in which they differ:}
+In particular, the approach from the GUIDE algorithm \hlc{(which essentially encompasses QUEST)}
+is compared to the one of the CTree and MOB algorithms by investigating the building blocks of 
+their testing strategies in which they differ:
 (1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.

### 2020-02-09 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further touch-ups to new paragraphs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2978 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 523cb0dd86a3ae532f555401d0acbf36aeddddd0

@@
 
-\hlc{GUIDE and MOB, which are both model-based approaches, employ the model estimated 
-in the considered node for an exhaustive search of the best split point over all possible 
-split points in the already selected split variable.
-CTree has not been developed specifically for model-based partitioning, hence, a 
-different approach to selecting a split point is suggested in the original paper. 
-However, as an adaption of CTree employing models for the split variable selection 
-is considered in this study, it is also combined with the same exhaustive search 
-strategy as GUIDE and MOB.}
-%\newpage
+\hlc{Once a split variable has been selected, the selection of the corresponding
+split point is less controversial. The most natural strategy is to use an exhaustive
+search over all possible split points and then select the one that optimizes the
+objective function (or loss function). In fact, this is the strategy adopted by both
+model-based approaches, GUIDE and MOB. Only CTree uses a slightly different approach,
+based on maximizing a two-sample test statistic, because the algorithm as originally
+proposed did not employ an explicit objective function. However, it can be shown
+that in some settings the two approaches coincide} \citep{Zeileis+Hothorn:2013}
+\hlc{while the differ somewhat more generally. As we focus on model-based partitioning
+here we employ the exhaustive model-based search for split point selection in
+all three algorithms considered.}
 
@@
 discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.
-\hlc{Therefore, in this study the focus is put on the selected building blocks as they are assumed 
-(and have shown) to have the highest influence on the difference in performance between the testing 
-strategies.}
+\hlc{Therefore, this aspect is not considered again here but focus is given to the building
+blocks selected above.}
 
@@
 $p$-values keeping this value particularly low).
-Even though it has to be noted that the number of replications in this simulation study is of
-course limited, the results for this setting indicate that all testing strategies conform well with 
-the predefined level of significance $\alpha = 0.05$ as the probability of splitting in any 
-of the available covariates is between 4 and 5~percent for $\delta = 0$ 
-(the exact values can be found in} Table~\ref{tab:typeIerror} in Appendix~\ref{app:typeIerror})
-\hlc{with the tests for structural changes being slightly more conservative} 
-\citep[see, e.g.,][for more details on this typical behaviour]{Hothorn+Zeileis:2008, Zeileis+Hothorn:2013, Hansen:1997}.
+Even with the limited number of replications in this simulation study the empirical size
+of all testing strategies (i.e., for $\delta = 0$) is between 4 and 5~percent and thus
+conforms well with the nominal significance level of $\alpha = 0.05$. The exact values
+can be found in} Table~\ref{tab:typeIerror} in Appendix~\ref{app:typeIerror})
+\hlc{which also shows that the asymptotic structural change tests are slightly more conservative}
+\citep[confirming previous results in][]{Hothorn+Zeileis:2008, Zeileis+Hothorn:2013}.
 

### 2020-02-14 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small corrections (typos)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2979 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e08cb5f9d03cf42da2d8fa5818c7964441a08742
final touch-ups; removed fixmes and highlighting of changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2981 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0178b6bb6aea3308d524bc0fafb3f4e4e2306cfb

@@
 
-\vspace*{0.4cm}
+%\vspace*{0.4cm}
 
@@
 best split point only within this variable in a separate second step is a more 
-promising strategy. \hlc{The most popular algorithms adopting such an unbiased recursive
-partitioning strategy are} QUEST \citep{Loh+Shih:1997},
+promising strategy. The most popular algorithms adopting such an unbiased recursive
+partitioning strategy are QUEST \citep{Loh+Shih:1997},
 %% CHAID \citep{Kass:1980},
-GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
-and MOB \citep{Zeileis+Hothorn+Hornik:2008}. \hlc{QUEST and CTree have been suggested as
-nonparametric decision trees initially and have acquired more than 1,300 and 2,400 citations
+GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006}, 
+and MOB \citep{Zeileis+Hothorn+Hornik:2008}. QUEST and CTree have been suggested as
+nonparametric decision trees initially and have acquired more than 1,300 and 2,400~citations
 on Google Scholar, respectively. GUIDE and MOB have been designed for general model-based
@@
 highest association to the response variable as captured by the $p$-value in the corresponding
-statistical test.} While QUEST
+statistical test. While QUEST
 %%, CHAID,
@@
 
-\fixme{``Why CTree, GUIDE, and MOB?''}
-
+%\fixme{``Why CTree, GUIDE, and MOB?''}
+%
 Therefore, in this paper the focus will be put on that first step of tree algorithms,
 i.e., the task of selecting the best split variable in a given (sub)sample. 
-In particular, the approach from the GUIDE algorithm \hlc{(which essentially encompasses QUEST)}
+In particular, the approach from the GUIDE algorithm (which essentially encompasses QUEST)
 is compared to the one of the CTree and MOB algorithms by investigating the building blocks of 
@@
 The remainder of this manuscript is structured as follows:
-Section~\ref{sec:urp} reviews unbiased tree models, starting with the general algorithmic idea
-(Section~\ref{sec:generic_alg}) followed by the specific algorithms CTree (Section~\ref{sec:ctree}), 
-MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}), before possible pruning strategies 
-(\ref{sec:pruning}) are discussed.
+Section~\ref{sec:urp} reviews unbiased tree models, starting with the general algorithmic idea~(Section~\ref{sec:generic_alg}) followed by the specific algorithms CTree~(Section~\ref{sec:ctree}), 
+MOB~(\ref{sec:mob}), and GUIDE~(\ref{sec:guide}), before possible pruning strategies~(\ref{sec:pruning}) are discussed.
 In Section~\ref{sec:unify} the unifying framework for testing strategies in unbiased model-based
@@
 likelihood and ordinary least squares estimation. For a $K$-dimensional parameter
-$\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
-is also a $K$-dimensional vector, i.e., the gradient contribution of the $i$-th observation.
+$\beta$ the score function evaluated for the $i$-th~observation~$s(y_i, x_i, \beta)$
+is also a $K$-dimensional vector, i.e., the gradient contribution of the $i$-th~observation.
 Thus, the $N \times K$-matrix consisting of these scores or gradient contributions for all 
@@
 
-If both variables $Y$ and $Z$ are numeric the default independence test corresponds to a Pearson 
-correlation test. For one numeric and one categorical variable essentially a one-way ANOVA
-(analysis of variance) is employed while for two categorical variables a $\chi^2$~test is performed.
+If both variables $Y$ and $Z$ are numeric the default independence test corresponds to a \mbox{Pearson} correlation test. For one numeric and one categorical variable essentially a one-way 
+ANOVA~(analysis of variance) is employed while for two categorical variables a $\chi^2$~test is performed.
 Thus, in the general CTree framework many types of tests can be specified by selecting suitable
@@
 (GLMs) and survival regression models but subsequently various other models have been applied as well,
-including beta regression \citep{Gruen+Kosmidis+Zeileis:2012}, psychometric item response theory
-models \citep{Strobl+Kopf+Zeileis:2015}, or mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
+including beta regression \citep{Gruen+Kosmidis+Zeileis:2012}, psychometric item response theory models~\citep{Strobl+Kopf+Zeileis:2015}, or mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
 But just like CTree can be applied to parametric models, MOB conversely also encompasses
@@
 To construct a statistical test two additional transformations are carried out:
-(1)~The residuals $r(Y,X,\hat{\beta})$ are dichotomized at zero, yielding an indicator for positive vs.\ negative residuals.
+(1)~The residuals~$r(Y,X,\hat{\beta})$ are dichotomized at zero, yielding an indicator for positive vs.\ negative residuals.
 (2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
@@
 \subsection{Selecting a split point}
-\fixme{``Why is the choice of the split point not part of this paper?''}
-
-\hlc{Once a split variable has been selected, the selection of the corresponding
+%\fixme{``Why is the choice of the split point not part of this paper?''}
+
+Once a split variable has been selected, the selection of the corresponding
 split point is less controversial. The most natural strategy is to use an exhaustive
@@
 proposed did not employ an explicit objective function. However, it can be shown
-that in some settings the two approaches coincide} \citep{Zeileis+Hothorn:2013}
-\hlc{while the differ somewhat more generally. As we focus on model-based partitioning
+that in some settings the two approaches coincide \citep{Zeileis+Hothorn:2013}
+while they differ somewhat more generally. As we focus on model-based partitioning
 here we employ the exhaustive model-based search for split point selection in
-all three algorithms considered.}
+all three algorithms considered.
 
@@
 
-\fixme{``Do the building blocks cover all possible options?''}
-
+%\fixme{``Do the building blocks cover all possible options?''}
+%
 Note that there are further differences in the testing strategies between the three algorithms, e.g.,
@@
 discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.
-\hlc{Therefore, this aspect is not considered again here but focus is given to the building
-blocks selected above.}
+Therefore, this aspect is not considered again here but focus is given to the building
+blocks selected above.
 
@@
 can depend on the true split variable $Z_1$. More specifically, three different variations
-are considered for the coefficients $\beta_0$ and $\beta_1$:
+are considered for the coefficients $\beta_0$ and $\beta_1$ (see Figure~\ref{fig:dgp_stump} for an illustration):
 %
@@
 %
-See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
-changes~($\beta_{k-1}$ with $k=1,2=K$) switches between two values at the split point $\xi$:
+More precisely, the coefficient that
+changes~($\beta_{k-1}$~with $k=1,2=K$) switches between two values at the split point $\xi$:
 %
@@
 and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
-Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
+Top right and bottom panels: Bivariate plot of the response~$Y$ on regressor~$X$ 
 with effect size $\delta = 1$ and for three variations: 
@@
 with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
-Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
+Right panel:~Bivariate plot of the response $Y$ on regressor~$X$ 
 with effect size $\delta = 1$.}
@@
 The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
-on 100 replications per step each consisting of 250 observations. 
+on 100~replications per step each consisting of 250 observations. 
 To compare the performance of the evaluated testing strategies for each step 
@@
 the $p$-values pertaining only to the true split variable $Z_1$;
-and the proportion of replications for which the $p$-value of $Z_1$
+and the proportion of replications~for which the $p$-value of $Z_1$
 is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
@@
 as significance tests answering two questions at once:
-(1)~Should a split be performed at all? 
+(1)~Should a split be performed~at~all? 
 (2)~If so, in which variable?
@@
 in $Z_1$ is either the median 0 (top) or the 90\% quantile 0.8 (bottom)
-for either varying intercept $\beta_0$ (left), varying slope $\beta_1$
+for either varying intercept~$\beta_0$ (left), varying slope~$\beta_1$
 (middle), or both coefficients varying (right).}
@@
 
-\fixme{Type I error}
-
-Figure~\ref{fig:scores_stump} \hlc{also shows that for $\delta = 0$, i.e., no difference
+%\fixme{Type I error}
+%
+Figure~\ref{fig:scores_stump} also shows that for $\delta = 0$, i.e., no difference
 in the parameters and consequently a random split variable selection, the probability of 
@@
 $p$-values keeping this value particularly low).
-Even with the limited number of replications in this simulation study the empirical size
-of all testing strategies (i.e., for $\delta = 0$) is between 4 and 5~percent and thus
-conforms well with the nominal significance level of $\alpha = 0.05$. The exact values
-can be found in} Table~\ref{tab:typeIerror} in Appendix~\ref{app:typeIerror})
-\hlc{which also shows that the asymptotic structural change tests are slightly more conservative}
+Even with the limited number of replications in this simulation study the empirical~size~of 
+all testing strategies (i.e., for $\delta = 0$) is between 4 and 5~percent and thus
+conforms well~with~the nominal significance level of $\alpha = 0.05$. The exact values
+can be found in Table~\ref{tab:typeIerror} in Appendix~\ref{app:typeIerror}
+which also shows that the asymptotic structural change tests are slightly more conservative
 \citep[confirming previous results in][]{Hothorn+Zeileis:2008, Zeileis+Hothorn:2013}.
@@
 different tree algorithms in the more complex ``tree'' scenario (see Figure~\ref{fig:dgp_tree}):
-once with significance-based pre-pruning and once with cost-complexity post-pruning.
+once with significance-based pre-pruning and once \mbox{with cost-complexity post-pruning.}
 
@@
 performs clearly worse compared to the GUIDE+scores. The latter is essentially on par with
-CTree and MOB when the true split matches one of the categorization bins (i.e., $\xi = 0$
+CTree~and~MOB when the true split matches one of the categorization bins (i.e., $\xi = 0$
 and $\xi = 0.5$, respectively) while CTree/MOB still perform somewhat better in the other
@@
 
+\newpage
 \section*{Computational details}
@@
 Lisa Schlosser received a PhD scholarship granted from the University of Innsbruck.
+
+\section*{Competing interests statement}
+The authors declare that they have no competing interests.
 
@@
 with the model scores $s_i = s(Y_i, X_i, \hat{\beta})$ being sorted by the possible split variable $Z_j$ by including the ordering permutation $\pi(Z_{ji})$. 
-To scale this partial sum process an estimate $\hat{V}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
+To scale this partial sum process an estimate~$\hat{V}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
 Following \cite{Zeileis+Hornik:2007} $W_j(t)$ converges to a Brownian bridge $W^0$ under the null
@@
 The true split point in $Z_1$ is either the median 0 (top) or the 90\% quantile 0.8 (bottom)
-for either varying intercept $\beta_0$ (left), varying slope $\beta_1$
+for either varying intercept~$\beta_0$ (left), varying slope~$\beta_1$
 (middle), or both coefficients varying (right).}

### 2020-02-14 pkg/partykit/inst/guideliketest/draft/draft.Rnw
corrected link to simulation code
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2982 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5bd3dddf3ce29d54fcf8792ffed22efeb02440e0

@@
 The applied implementation is based on the \proglang{R} package \pkg{partykit}
-(version~\Sexpr{packageVersion("partykit")}) which is available 
-on \proglang{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.
-The code to reproduce the simulation study is available in the supplement for this 
-paper on arXiv.org E-Print Archive (\url{https://arxiv.org/}).
+%(version~\Sexpr{packageVersion("partykit")}) 
+(version~1.2.4) 
+which is available on \proglang{R}-Forge at 
+\url{https://R-Forge.R-project.org/projects/partykit/}.
+The code to reproduce the simulation study is available at \url{https://eeecon.uibk.ac.at/~zeileis/news/power_partitioning/}.
 

### 2020-02-27 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
added overview slide for GUIDE/CTree/MOB; omit full factorial results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2990 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c9a2fc377370b72cb0f452ce43082b6efa818f8b

@@
 %\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
-\URL{http://www.partykit.org/} 
-\headerimage{3}
+\URL{http://www.partykit.org/partykit} 
+\headerimage{4}
 
@@
 \definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
+\definecolor{dred}{RGB}{119,16,26}
 
@@
 
-\section{Unbiased Recursive Partitioning}
+\section{The Power of Unbiased Recursive Partitioning}
+
+\subsection{Motivation}
 
@@
 
-\begin{frame}%[fragile]
-\frametitle{Unbiased Recursive Partitioning}
+\subsection{Unbiased recursive partitioning}
+
+\begin{frame}
+\frametitle{Unbiased recursive partitioning}
+
+\textbf{GUIDE:} Loh (2002, \emph{Statistica Sinica}).
+\begin{itemize}
+  \item First \emph{unbiased} algorithm for recursive partitioning of linear models.
+  \item Separation of split variable and split point selection.
+  \item Based on $\chi^2$ tests.
+\end{itemize}
+
+\medskip
+
+\textbf{CTree:} Hothorn, Hornik, Zeileis (2006, \emph{JCGS}).
+\begin{itemize}
+  \item Proposed as unbiased recursive partitioning for nonparametric modeling.
+  \item Based on conditional inference (or permutation tests).
+  \item Can be model-based via model scores as the response transformation.
+\end{itemize}
+
+\medskip
+
+\textbf{MOB:} Zeileis, Hothorn, Hornik (2008, \emph{JCGS}).
+\begin{itemize}
+  \item Model-based recursive partitioning using M-estimation (ML, OLS, CRPS, \dots).
+  \item Based on parameter instability tests.
+  \item Adapted to various psychometric models: Rasch, PCM, Bradley-Terry, MPT, SEM, networks, \dots.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Unbiased recursive partitioning}
 {\bf Basic tree algorithm:}
@@
 \begin{frame}%[fragile]
-\frametitle{Split Variable Selection}
+\frametitle{Split variable selection}
 
@@
 \begin{frame}%[fragile]
-\frametitle{Discrepancy Measures}
+\frametitle{Discrepancy measures}
 
@@
 \bigskip
+\pause
 
@@
 
-\visible<2>{
+\pause
+
 \vspace{-0.3cm}
-$$\hspace{-0.3cm}{\color{red}\Downarrow} \hspace{3.4cm} {\color{red}\Downarrow} $$
+$$\hspace{-0.3cm}{\color{dred}\Downarrow} \hspace{3.4cm} {\color{dred}\Downarrow} $$
 $$
-{\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1}) \cdot X}
+{\color{dred} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{dred} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1}) \cdot X}
 $$
-}
-\end{frame}
+
+\end{frame}
+
+\subsection{A unifying view}
 
 \begin{frame}
-\frametitle{A Unifying View}
+\frametitle{A unifying view}
 
@@
 \begin{frame}%[fragile]
-\frametitle{General Framework}
+\frametitle{General framework}
 
@@
 \item \emph{Pre-pruning:}
-Internal stopping criterion based on Bonferroni-corrected $p$-values of the applied tests. Stop splitting when there is no significant association. 
+Internal stopping criterion based on Bonferroni-corrected $p$-values of the underlying tests. Stop splitting when there is no significant association. 
 \item \emph{Post-pruning:}
@@
 \end{frame}
+
+\subsection{Simulation}
 
@@
 
-\begin{frame}%[fragile]
-\frametitle{Simulation 1: Full factorial analysis of building blocks}
-
-\begin{figure}%[t]
-\setkeys{Gin}{width=0.8\linewidth}
-<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
-xyplot(pval_z1 ~ cat | xi, 
-       groups = interaction(bin, res_scores), 
-       data = sim_3way, 
-       ylim = c(0,0.5),
-       type = c("a"), 
-       col = c(pal[1], pal[3], pal[1], pal[3]),
-       lty = c(1,1,2,2),
-       lwd = 2,
-       layout = c(2,1),
-       scales = list(alternating=FALSE),
-       xlab = "Categorization",
-       ylab = expression("p-value of"~Z[1]),
-       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
-                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
-                            bg = "gray90"),
-       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
-                  columns = 2,
-                  lines = list(lty = c(0,1,2,0,1,1), 
-                               type = "l", lwd = 2,
-                               col = c("black","black","black","black", pal[1], pal[3]))))
-
-@
-\end{figure}
-\end{frame}
+%% \begin{frame}%[fragile]
+%% \frametitle{Simulation 1: Full factorial analysis of building blocks}
+%% 
+%% \begin{figure}%[t]
+%% \setkeys{Gin}{width=0.8\linewidth}
+%% <<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
+%% xyplot(pval_z1 ~ cat | xi, 
+%%        groups = interaction(bin, res_scores), 
+%%        data = sim_3way, 
+%%        ylim = c(0,0.5),
+%%        type = c("a"), 
+%%        col = c(pal[1], pal[3], pal[1], pal[3]),
+%%        lty = c(1,1,2,2),
+%%        lwd = 2,
+%%        layout = c(2,1),
+%%        scales = list(alternating=FALSE),
+%%        xlab = "Categorization",
+%%        ylab = expression("p-value of"~Z[1]),
+%%        strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
+%%                                               expression(xi==0.8~"(90%)"~" and"~delta==1)),
+%%                             bg = "gray90"),
+%%        key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
+%%                   columns = 2,
+%%                   lines = list(lty = c(0,1,2,0,1,1), 
+%%                                type = "l", lwd = 2,
+%%                                col = c("black","black","black","black", pal[1], pal[3]))))
+%% 
+%% @
+%% \end{figure}
+%% \end{frame}
 
@@
 
+\subsection{Recommendations}
+
 \begin{frame}%[fragile]
@@
 \item Categorization is simpler, but less powerful in margins.
-\item If the significance test performs well pre-pruning works well,\\
+\item Maximally-selected statistics (as in MOB) more powerful for abrupt shifts.
+\item Linear statistics (default in CTree) more powerful for linear changes.
+\item If the significance tests perform well pre-pruning works well,\\
       otherwise post-pruning might be needed.
-%\item Including full model scores is advantageous over using only residuals.
-%\item Binarizing scores/residuals does not improve the performance.
-%\item Categorizing split variables can simplify calculations, however,
-%effects might be missed unless a binning point is close to the true split point.
-%\item If the applied test performs well as significance test pre-pruning
-%is preferable, otherwise post-pruning.
 \end{itemize}
@@
 
-\begin{frame}
-\frametitle{Software}
-
-\textbf{Package:} Available on CRAN at
-
-\url{https://CRAN.R-project.org/package=partykit}\\
-
-\bigskip
-
-\textbf{Replication code:} Available on R-Forge at
-
-\url{https://R-Forge.R-project.org/projects/partykit/}\\
-
+%% \begin{frame}
+%% \frametitle{Software}
+%% 
+%% \textbf{Package:} Available on CRAN at
+%% 
+%% \url{https://CRAN.R-project.org/package=partykit}\\
+%% 
+%% \bigskip
+%% 
+%% \textbf{Replication code:} Available on R-Forge at
+%% 
+%% \url{https://R-Forge.R-project.org/projects/partykit/}\\
+%% 
 %\bigskip
@@
 %\end{itemize}
-\end{frame}
+%% \end{frame}
 
@@
   \dquote{The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE.}
- \emph{Working paper in preparation}.
-
-\medskip
-
-Breiman L, Friedman J H, Olshen R A, Stone C J (1984).
- \textit{Classification and Regression Trees.}
- Wadsworth,
- California.
+  arXiv:1906.10179, \emph{arXiv.org E-Print Archive}.
+  \url{https://arxiv.org/abs/1906.10179}.
+
+\bigskip
+\bigskip
+
+Loh W-Y (2002).
+  \dquote{Regression Trees with Unbiased Variable Selection and Interaction Detection.}
+  \textit{Statistica Sinica},
+  \textbf{12}(2), 361--386.
+  \url{http://www.jstor.org/stable/24306967}
 
@@
 
-Hothorn T, Zeileis A (2015).
- \dquote{\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
- \textit{Journal of Machine Learning Research},
- \textbf{16}, 3905--3909.
- \url{http://www.jmlr.org/papers/v16/hothorn15a.html}
-
-\medskip
-
-Loh W-Y (2002).
- \dquote{Regression Trees with Unbiased Variable Selection and Interaction Detection.}
- \textit{Statistica Sinica},
- \textbf{12}(2), 361--386.
- \url{http://www.jstor.org/stable/24306967}
-
-\medskip
-
 Zeileis A, Hothorn T, Hornik K (2008).

### 2020-03-22 pkg/circtree/inst/wind_paper/talks/ACINN_graduateseminar2020/slides/slides.Rnw
Add RNW for talk at ACINN
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3000 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 51bc558f5ee3a861c5857f44883c50b46073638d

--- 2018-07-15 pkg/disttree/inst/JSM2018/slides.Rnw
@@
-\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
-\usetheme[nototalframenumber, noslidenumber]{uibk}
-
-\title{Distributional Trees and Forests}
-\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
+\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
+\usetheme[url,license]{uibk}
+
+\title[Circular Regression Trees and Forests with an Application to
+Probabilistic Wind Direction Forecasting]{Circular Regression Trees and Forests 
+with an\\\vspace{0.2em} Application to Probabilistic Wind Direction Forecasting}
+\author{Moritz N. Lang, Lisa Schlosser, \\
+Torsten~Hothorn, Georg~J.~Mayr, Reto Stauffer, Achim Zeileis}
+
+%-------------------------------------------------------------------
+% Preamble
+%-------------------------------------------------------------------
 \setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
-%\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
-\URL{https://R-Forge.R-project.org/projects/partykit/}
-\headerimage{3}
+\headerimage{1}
+
+\graphicspath{{images/}{./}}
+\usepackage[utf8]{inputenc}
 
@@
 %% includes a replacement for \usepackage{Sweave}
+% \usepackage{Sweave}
+\usepackage{changepage}
 \usepackage{amsmath,tikz}
-\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-\graphicspath{{plots/}}
-
+\usepackage{calc}
+\usepackage{graphicx}
+\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
+\usepackage{xcolor}
 \newcommand{\argmax}{\operatorname{argmax}\displaylimits}
+
+\newcommand{\backupbegin}{
+   \newcounter{finalframe}
+   \setcounter{finalframe}{\value{framenumber}}
+}
+\newcommand{\backupend}{
+   \setcounter{framenumber}{\value{finalframe}}
+}
 
@@
 \definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
+\definecolor{forestred}{RGB}{206,73,81}
+\definecolor{treegreen}{RGB}{0,143,0}
+\definecolor{lightblue}{RGB}{34,151,230}
+\definecolor{lightorange}{RGB}{255,165,0}
 
@@
 
-# setwd("~/svn/partykit/pkg/disttree/inst/JSM2018")
-library("ggplot2")
-theme_set(theme_bw(base_size = 18))
+invisible(.Call(grDevices:::C_palette, grDevices::hcl(
+  h = c(0,   5, 125, 245, 195, 315,  65,   0),
+  c = c(0, 100,  90,  85,  63, 105,  90,   0),
+  l = c(0,  55,  75,  60,  82,  48,  80,  65)
+)))
+
+# setwd("~/svn/partykit/pkg/disttree/inst/dissertation_overview/slides/")
 library("disttree")
 library("Formula")
-#library("partykitR1")
+library("latex2exp")
 library("gamlss")
@@
 library("crch")
-library("latex2exp")
+library("RainTyrol")
 library("parallel")
@@
 gen.cens(NO, type = "left")
+library("ggplot2")
+theme_set(theme_bw(base_size = 18))
 
@@
 
-transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-######## data generating process based on a predefined tree or a parameter function
-# if tree: 2 splits => 3 terminal nodes
-# 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
-#
-# input: n            ..... nr of observations
-#        family       ..... distribution of the generated observations
-#        split.matrix ..... indices of the variables used for the splits together with the corresponding split points
-#        parm.matrix  ..... set of distribution parameters for each subgroup
-#        fun          ..... parameter function
-#
-# output: data.frame with generated observations y (generated seperatly in each subgroups with given distributions parameters),
-#                         the given split variables x1, ..., x10 for each observation and
-#                         the index of the subgroup for each observation or
-#                         the distribution parameter for each observation
-
-# FIX: until now only complete lists can be handed over (dist_list...)
-dgp <- function(n, family = dist_list_normal, 
-                fun = NULL,
-                split.matrix = matrix(nrow = 2, ncol = 2), 
-                par.matrix = matrix(nrow = 3, ncol = 2), 
-                round.sp = 3)
-{
-
-  # generating the possible split variables
-  x1 <- runif(n,-0.4,1)
-  x2 <- runif(n,-10,10)
-  x3 <- runif(n,0,100)
-  x4 <- x1 + rnorm(n, sd = 0.1)
-  x5 <- x1 + rnorm(n, sd = 0.3)
-  x6 <- rbinom(n,1,0.5)
-  x7 <- rbinom(n,1,0.5)
-  x8 <- sample(1:4, n, replace = TRUE) 
-  x9 <- sample(1:5, n, replace = TRUE) 
-  x10 <- sample(1:7, n, replace = TRUE)
-  #x11 <- runif(n,-0.5,1)
-  x <- cbind(x1,x2,x3,x4,x5,x6,x7,x8,x9,x10)
-  # reduce nr of possible split points by rounding values of split variables
-  x <- round(x, digits = round.sp)
-
-  y <- vector(mode = "numeric", length = n)
-
-  # getting the random function
-  if(is.function(family)) family <- family()
-  if(inherits(family, "gamlss.family")) {
-    rfun <- get(paste0("r",family$family[[1]]))
-  } else {
-    if(family$family.name == "Normal Distribution") rfun <- rNO
-    if(family$family.name == "censored Normal Distribution") rfun <- rNO
-    if(family$family.name == "Poisson Distribution") rfun <- rPO
-    if(family$family.name == "Weibull Distribution") rfun <- survival:::rsurvreg 
-  }
-
-  if(is.null(fun)) {
-    index <- vector(mode = "numeric", length = n)
-    sv = split.matrix[,1]    # index of split variables
-    sp = split.matrix[,2]    # split points
-    par = par.matrix
-
-    # splitting based on the given split variables and split points and
-    # generating the observations seperatly for each subgroup (with the corresponding given distribution parameters)
-    for(k in 1:n){
-      if(x[k, sv[1]] <= sp[1]){
-        y[k] <- do.call(rfun, as.list(c(1,par[1,])))
-        index[k] <- 2L
-      } else {
-        if(x[k, sv[2]] <= sp[2]){
-          y[k] <- do.call(rfun, as.list(c(1,par[2,]))) 
-          index[k] <- 4L
-        } else {
-          y[k] <- do.call(rfun, as.list(c(1,par[3,])))
-          index[k] <- 5L
-        }
-      }
-    }
-    d <- as.data.frame(cbind(y, x, index))
-    colnames(d) <- c("y", paste0("x", c(1:10)), "index")
-
-  } else {
-
-    dpar <- fun(x)
-
-    if(!(family$family.name == "Poisson Distribution")){
-      y <- rfun(n, dpar[,1], dpar[,2])
-      d <- as.data.frame(cbind(y, x, dpar))
-      colnames(d) <- c("y", paste0("x", c(1:10)), "mu", "sigma")
-    } else {
-      y <- rfun(n, dpar)
-      d <- as.data.frame(cbind(y, x, dpar))
-      colnames(d) <- c("y", paste0("x", c(1:10)), "lambda")
-    }
-  }
-
-  if(family$family.name == "censored Normal Distribution") {
-    #d$ystar <- d$y
-    d$y <- pmax(d$y, 0)
-  }
-
-  return(d)
-}
-
-############################################ 
-# parameter functions
-
-### plotting function for one data set
-plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE, all.var = FALSE) 
-{
-  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
-                    df$fitted.par,
-                    dt$fitted.par,
-                    #coef(dt)[paste(dt$fitted[,1]),],
-                    g$mu.fv, g$sigma.fv)
-
-  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
-                          "fitted.mu.df","fitted.sigma.df",
-                          "fitted.mu.dt","fitted.sigma.dt",
-                          "fitted.mu.g","fitted.sigma.g")
-  sp <- plotdata[order(plotdata["x1"]),]
-
-  if(!(any(dt.par, df.par, g.par, all.par, all.var))){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-  col = transpgrey, border = "transparent")
-    #lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
-    #       col = c('black', transpgrey), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # disttree
-  if(dt.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", col.main = pal["tree"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
-  col = pallight["tree"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-    #       col = c('black','grey', pal["tree"], pallight["tree"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # distforest
-  if(df.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", col.main = pal["forest"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
-  col = pallight["forest"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
-    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
-    #       col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # gamlss
-  if(g.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2.5)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
-  col = pallight["gamlss"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
-    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
-    #       col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-  # disttree vs. distforest vs. gamlss
-  if(all.par){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-
-    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
-    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
-    #                    TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
-    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-    legend(x = -0.5, y = 13.95, expression(mu), bty = "n")
-  }
-
-  # variance: disttree vs. distforest vs. gamlss
-  if(all.var){
-    par(mar=c(5.1,4.1,4.1,3.1))
-    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
-    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
-    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
-    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
-    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)
-
-    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
-    #col = transpgrey, border = "transparent")
-    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'black')
-    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'black')
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
-  col = pallight["tree"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
-  col = pallight["forest"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
-    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
-  col = pallight["gamlss"], border = "transparent")
-    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
-    #legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
-    #                    TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
-    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
-  legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
-  }
-
-}
-
-@
-
+transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
+@
+
+%-------------------------------------------------------------------
+% Main
+%-------------------------------------------------------------------
 \begin{document}
-
-\section{Distributional Trees and Forests}
-
+\section{Distributional Trees and Forests for Circular Data}
+
+%-------------------------------------------------------------------
 \subsection{Motivation}
-
-\begin{frame}[fragile]
+%-------------------------------------------------------------------
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Motivation}
+
+\pause
+
+\textbf{Goal:} 
+\begin{itemize}
+  \item Statistical modeling of circular response variables.
+  \item Probabilistic forecasting.
+\end{itemize}
+
+\medskip
+
+\textbf{Examples:} Circular data.
+\begin{itemize}
+\item Hourly crime data: $[0\text{h}, 24\text{h})$ with $0\text{h} = 24\text{h}$.
+\item Directional behavior of animals: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
+\item Wind directions: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
+\end{itemize}
+
+\end{frame}
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Motivation}
+
+\textbf{Application:} 
+\begin{itemize}
+  \item Probabilistic forecasting of wind directions for the next hours
+  at\\Innsbruck and Vienna Airport, Austria.
+\end{itemize}
+
+\medskip
+
+\textbf{Challenges:}
+\begin{itemize}
+  %\item Circular response in $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
+  \item Possibly abrupt and short-term changes due to geographical position.
+  \item Numerical weather predictions less useful due to short lead time.
+  \item Large number of unknown interactions between covariates.
+\end{itemize}
+
+\bigskip
+\pause
+\begin{center}
+$\boldsymbol{\Rightarrow}$ \textbf{What kind of statistical model should be used?}
+\end{center}
+
+\end{frame}
+
+%% SLIDE 
+\begin{frame}%[fragile]
 \frametitle{Motivation}
@@
 \begin{figure}[!htb]
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 @
-\visible<2->{
+\visible<1->{
 <<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
@@
 \endminipage
-\visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\visible<2->{{\LARGE$\rightarrow$}}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 @
-\visible<3->{
+\visible<2->{
 <<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
@@
 \endminipage
-\visible<4->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\visible<3->{{\LARGE$\rightarrow$}}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 @
-\visible<4->{
+\visible<3->{
 <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
@@
 @
-
 }
@@
 
-\vspace{0.5cm}
+\vspace{0.0cm}
 \minipage{0.25\textwidth}
-\begin{center}
-\visible<2->{
-LM, GLM\\
-\vspace{0.5cm}
-\code{lm}\\
-\code{glm}\\
-\vspace{1.5cm}}
+\vspace{-0.3cm}
+\begin{center}
+\visible<1->{
+(Generalized) Linear model\\
+\vspace{0.8cm}
+{\small$\mathbb{E}(Y|X) =  \alpha_0 + \alpha_1 \cdot X$}
+\vspace{0.3cm}
+}
 \end{center}
@@
 \begin{center}
-\visible<3->{
-GAM\\
-\vspace{0.5cm}
-\code{mgcv}\\
-\code{VGAM}\\
-$\ldots$\\
-\vspace{1cm}}
+\vspace{-0.3cm}
+\visible<2->{
+Generalized additive model\\
+\vspace{0.8cm}
+{\small$\mathbb{E}(Y|X) = \alpha + f(X)$}\\
+\vspace{0.3cm}
+}
 \end{center}
@@
 \hspace{1.1cm}
-\minipage{0.25\textwidth}
-\begin{center}
-\visible<4->{
-GAMLSS\\
-\vspace{0.5cm}
-\code{gamlss}\\
-\code{mgcv}\\
-\code{VGAM}\\
-\code{gamboostLSS}\\
-$\ldots$}
+\minipage{0.27\textwidth}
+\begin{center}
+\visible<3->{
+Generalized additive model for location, scale, and shape\\
+\vspace{0.3cm}
+{\small $\mathbb{E}(Y|X) = \alpha + f_1(X)$}\\\vspace{0.2em}
+{\small $\sqrt{Var(Y|X)} = \beta + f_2(X)$}
+}
 \end{center}
@@
 
+%% SLIDE
 \begin{frame}[fragile]
 \frametitle{Motivation}
-\vspace{-0.2cm}
+\vspace{-0.41cm}
 \begin{figure}[!htb]
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 \visible<2->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
  \begin{center}
@@
 \visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
+\begin{center}
 \visible<3->{
-  \begin{center}
-  \begin{tikzpicture}
-  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
-  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
-  \end{tikzpicture}
-  \end{center}}
+<<plot_motivation_question, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
+box(lwd=5)
+text(x = mean(range(x)), y = mean(range(y)), "?", cex = 12)
+@
+}
+\end{center}
+
 \endminipage
 
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 \visible<1->{
-Regression Tree\\
-\vspace{0.5cm}
-\resizebox{0.3\textwidth}{!}{
+\vspace{-0.05cm}
+Regression tree\\
+\vspace{0.7cm}
+\resizebox{0.2\textwidth}{!}{
 \begin{tikzpicture}
@@
 \end{tikzpicture}}
-\vspace{0.3cm}\\
-\code{rpart}\\
-\code{party(kit)}\\
+\vspace{0.2cm}\\
 \vspace{1cm}}
@@
 \hspace{0.65cm}
-\minipage{0.29\textwidth}
-\begin{center}
-\vspace{0.0cm}
+\minipage{0.285\textwidth}
+\begin{center}
+\vspace{-1.0cm}
 \visible<2->{
-Random Forest\\
-\vspace{0.5cm}
-\resizebox{0.9\textwidth}{!}{
+Random forest\\
+\vspace{0.75cm}
+\resizebox{0.6\textwidth}{!}{
 \begin{tikzpicture}
@@
 }
-\vspace{0.5cm}\\
-\code{randomForest}\\
-\code{ranger}\\
-\code{party(kit)}\\
-$\ldots$}
+\vspace{0.3cm}\\
+}
 \end{center}
@@
 \hspace{0.65cm}
-\minipage{0.28\textwidth}
-\begin{center}
-\vspace{0.68cm}
+\minipage{0.285\textwidth}
+\begin{center}
+\vspace{-0.4cm}
 \visible<3->{
 Distributional trees and forests\\
-\vspace{1.7cm}
-\code{disttree}\\
-based on \code{partykit}\\
+\vspace{1.15cm}
 \vspace{1cm}}
@@
 
+%% SLIDE
 \begin{frame}
-\frametitle{Goals}
-
-\textbf{Distributional:} \\
+\frametitle{Distributional trees}
+
+\textbf{Distributional:}
 \begin{itemize}
-\item Specify the complete probability distribution (including location, scale and shape).
+  \item Specify the complete probability distribution (location, scale, shape, \dots).
+  \item Probabilistic / distributional forecasts.
 \end{itemize}
@@
 
-{\bf Tree:}
+\textbf{Tree:}
 \begin{itemize}
-  \item Automatic variable selection and detection of steps and abrupt changes. %(data driven)
-  \item Capture non-linear and non-additive effects and interactions.
+  \item Automatic detection of steps and abrupt changes. %(data driven)
+  \item Capture non-linear and non-additive effects, and interactions.
 \end{itemize}
@@
 \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Distributional trees}
-\vspace*{-0.12cm}
-\begin{center} 
-DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
-\vspace*{-0.21cm}
+
+\medskip
+\pause
+\begin{center}
+$\boldsymbol{\Rightarrow}$ \textbf{Ideal for capturing short-term and abrupt changes in wind direction.}
+\end{center}
+
+\end{frame}
+
+%-------------------------------------------------------------------
+\subsection{Distributional trees}
+%-------------------------------------------------------------------
+
+<<treedata, echo=FALSE, results=hide>>=
+set.seed(54)
+nobs <- 500
+x <- runif(nobs, 0, 1)
+mu <- sigma <- ytrue <- numeric(length = nobs)
+for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
+for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
+y <- rnorm(nobs, mean = mu, sd = sigma)
+ytrue <- mu
+data <- data.frame(cbind(y,x, ytrue))
+
+alldata <- cbind(data, mu, sigma)
+odata <- alldata[order(alldata["x"]),]
+@
+
 <<dgp_tree, echo=FALSE, results=hide>>=
@@
 
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\vspace*{-0.12cm}
+\begin{center} 
+DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
+
+\vspace*{-0.21cm}
+\setkeys{Gin}{width=0.58\linewidth}
+<<plottree_xyplot, fig=TRUE, echo=FALSE,width=7>>=
+par(mar=c(5.1,4.1,2.4,1.1))
+plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "gray")
+lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
+polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
+  col = pallight["forest"], border = "transparent")
+legend("topleft", expression(mu  %+-%  sigma), bty = "n")
+@
+\end{center}
+\end{frame}
+
+%% SLIDE
+\begin{frame}
+\frametitle{Distributional trees}
+\vspace*{-0.12cm}
+\begin{center} 
+DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
+
+\vspace*{-0.21cm}
+\setkeys{Gin}{width=0.5\linewidth}
 <<plottree_dgp, fig=TRUE, echo=FALSE>>=
@@
 
+%% SLIDE
 <<tree, echo=FALSE, results=hide>>=
@@
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+\setkeys{Gin}{width=0.5\linewidth}
 <<plottree_estpar, fig=TRUE, echo=FALSE>>=
@@
 
+%% SLIDE
 \begin{frame}[fragile]
@@
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+\setkeys{Gin}{width=0.5\linewidth}
 <<plottree_box, fig=TRUE, echo=FALSE>>=
@@
 \vspace*{-0.2cm}
-\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
+\setkeys{Gin}{width=0.5\linewidth}
 <<plottree_dens, fig=TRUE, echo=FALSE>>=
@@
 \end{frame}
-%\begin{frame}
-%\frametitle{Distributional trees}
-%\center
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{tree_box}
-%\end{frame}
-
-<<treedata, echo=FALSE, results=hide>>=
-set.seed(54)
-nobs <- 500
-x <- runif(nobs, 0, 1)
-mu <- sigma <- ytrue <- numeric(length = nobs)
-for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
-for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
-y <- rnorm(nobs, mean = mu, sd = sigma)
-ytrue <- mu
-data <- data.frame(cbind(y,x, ytrue))
-
-alldata <- cbind(data, mu, sigma)
-odata <- alldata[order(alldata["x"]),]
-@
-
-\begin{frame}[fragile]
-\frametitle{Distributional trees}
-\begin{center}
-\vspace*{-1.03cm}
-\setkeys{Gin}{width=0.8\linewidth}
-<<plottree_xyplot, fig=TRUE, echo=FALSE>>=
-par(mar=c(5.1,4.1,4.1,3.1))
-plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "grey")
-#plot(y=odata$y, x=odata$x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray")
-lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
-polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
-  col = pallight["forest"], border = "transparent")
-legend(x = -0.0715, y = 21.62, expression(mu  %+-%  sigma), bty = "n")
-#lines(x = odata$x, y = odata$mu + odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
-#lines(x = odata$x, y = odata$mu - odata$sigma, col = hcl(10, 80, 70), lwd = 1.5, lty = 1)
-@
-\end{center}
-\end{frame}
-
-%\begin{frame}
-%\frametitle{Distributional trees}
-%\center
-%\vspace*{-0.7cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{tree_xyplot}
-%\end{frame}
-
-\subsection{Building Distributional Trees and Forests}
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Circular response distribution}
+\pause
+
+\vspace*{-0.3cm}
+
+\begin{minipage}{0.5\textwidth}
+\textbf{Distribution assumption:} Von Mises.
+\begin{itemize}
+\item Circular normal distribution.
+\item Location parameter $\mu \in [0, 2 \pi)$.
+\item Concentration parameter $\kappa > 0$.
+\end{itemize}
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\only<2>{\includegraphics[width=\linewidth]{density_linear.pdf}}%
+\only<3>{\includegraphics[width=\linewidth]{density_circular.pdf}}%
+\end{minipage}
+
+\bigskip
+\medskip
+
+\textbf{Log-likelihood:} $y \in [0, 2 \pi)$ and parameter vector $\theta = (\mu, \kappa)$.
+$$
+\ell(\theta; y) = \log \left\{ \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)} \right\}
+$$
+where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$.
+
+\end{frame}
+
+%-------------------------------------------------------------------
+\subsection{Learning}
+%-------------------------------------------------------------------
+
+%% SLIDE
 \begin{frame}
-\frametitle{Building Distributional Trees and Forests}
+\frametitle{Learning circular trees and forests}
+      \begin{minipage}{0.7\textwidth}
       {\bf Tree:}
       \begin{enumerate}
-\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
-\item Estimate $\hat{\theta}$ via maximum likelihood.
-\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable.
-%\item Assess whether the \emph{model scores} are associated with
-%    (or change along) any of the available covariates -- e.g.,
-%    using parameter instability tests (\emph{strucchange}) or
-%    conditional inference (\emph{coin}).
-\item Choose the partitioning variable with the strongest association or instability as split variable.
-\item Choose the breakpoint along this variable with highest improvement in log-likelihood and split the sample.
-\item Repeat steps 2--5 recursively in the subgroups
-    until some stopping criterion is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
-    %-- e.g., for significance or sample size.
-%\item Choose the variable with the strongest association
-%\item Choose the split point which leads to the highest improvement of the model
-%\item Split and repeat 2-6 in each node until a stopping criterion is met
+      \item<3-> Fit global distributional model $\text{vM}(Y; \theta)$: \\ % to the whole data set:\\
+      Estimate model parameters $\hat{\theta}$.
+      \item<7-> Evaluate goodness of fit \\
+      (for each parameter and each observation).
+      \item<8-> Choose covariate $X$ with strongest influence on goodness of fit
+        of $\text{vM}(Y; \hat{\theta})$ as split variable.
+      \item<8-> Find the split point $p$ which leads to the highest improvement.
+      \item<10-> Repeat steps 1--4 recursively in the subgroups
+        until some stopping criterion is met.
       \end{enumerate}
-
-\vspace{0.2cm}
-{\bf Forest:} Ensemble of $T$ trees.
+      \end{minipage}
+      \begin{minipage}{0.23\textwidth}
+      %\vspace{-0.1cm}
+      \begin{tikzpicture}
+      \visible<2-3>{
+      \node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
+      }
+      \visible<4>{
+      \node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\text{vM}(Y;\hat{\theta}$)};
+      }
+      \visible<5->{
+      \node[inner sep=0pt] (density_root) at (0.8, 1.7)
+          {\includegraphics[width=0.6\textwidth]{density_root.pdf}};
+      }
+      \visible<6-8>{
+      \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {?};
+      \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
+      \draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1);
+      \draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2);
+      }
+      \visible<9-10>{
+      \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
+      \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
+      \draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+      \draw[-, gray, line width=0.5pt] (1, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
+      }
+      \visible<11>{
+      \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$\text{vM}(Y_1;\hat{\theta}_1$)};
+      \node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$\text{vM}(Y_2;\hat{\theta}_2$)};
+      }
+      \visible<11->{
+      \draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+      \draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
+      }
+      \visible<12->{
+      \node[inner sep=0pt] (density2) at (-0.1,-0.3)
+          {\includegraphics[width=0.5\textwidth]{density_left.pdf}};
+      }
+      \visible<12->{
+      \node[inner sep=0pt] (density3) at (1.7,-0.3)
+          {\includegraphics[width=0.5\textwidth]{density_right.pdf}};
+      }
+      \end{tikzpicture}
+      \end{minipage}
+      \vspace{0.4cm}
+\end{frame}
+
+%% SLIDE
+\begin{frame}
+\frametitle{Learning circular trees and forests}
+      \begin{minipage}{0.7\textwidth}
+      {\bf Forest:} 
+      \begin{enumerate}
+      \item Learn an ensemble of $T$ trees:\\
+       Each tree is built on a bootstrap or subsample of the data, with
+      random input variable sampling.
+      \item<2-> Combine / average over estimated $T$ trees:\\
+      To stabilize and regularize the model.
+      \end{enumerate}
+      \end{minipage}
+
+\end{frame}
+
+%-------------------------------------------------------------------
+\subsection{Wind forecasting}
+%-------------------------------------------------------------------
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Wind forecasting}
+
+\textbf{Here:} Nowcasting (1--3 hours ahead) of wind direction at Innsbruck Airport.
+
+\bigskip
+\pause
+
+\textbf{Inputs:} Observation data only (41\,979 data points).
 \begin{itemize}
-\item Bootstrap or subsamples.
-\item Random input variable sampling.
+  \item 4 stations at Innsbruck Airport, 6 nearby weather stations. 
+  \item Base variables: Wind direction, wind (gust) speed, temperature, (reduced) air pressure, relative humidity.
+  \item 260 covariates based on means/minima/maxima, temporal changes, spatial differences towards the airport.
 \end{itemize}
-\end{frame}
-
+
+\end{frame}
+
+%% SLIDE
 \begin{frame}
-\frametitle{Adaptive Local Likelihood Estimation}
-
-\textbf{Parameter estimator} for \only<1-3>{\textbf{a global}}\only<4->{\textbf{an adaptive local}}\\ model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$} :
-
-\[
-\hat{\theta}\visible<2->{(\bold{x})} =  \argmax_{\theta \in \Theta} \sum_{i=1}^n \visible<2->{w_i(\bold{x}) \cdot} \ell(\theta; y_i)
-\]
-
-\bigskip
-
-\visible<3->{
-\textbf{Weights:}
-\begin{eqnarray*}
-w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.2cm]
-\visible<4->{
-w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.1cm]
-\visible<5->{
-w^{\text{forest}}_i(\bold{x}) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
-\end{eqnarray*}
-}}}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.4cm}
-\textbf{Goal:} Predict daily precipitation amount in alpine regions.\\
-\medskip
-\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
-\medskip
-\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
-\vspace{1.3cm}
-%\bigskip
-
-%\textbf{Idea:} Statistical model with numerical predictions as covariates.\\
-%\bigskip
-%\textbf{Distribution assumption:}\\
-%\vspace{-0.25cm}
-%\hspace{1.5cm}
-%\center
-%$$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
-
-\end{frame}
-
-<<echo = FALSE, results=hide>>=
-library("zoo")
-
-Sys.setenv("TZ"="UTC")
-
-# -------------------------------------------------------------------
-# GFS ensemble forecast
-# -------------------------------------------------------------------
-
-GFS   <- read.table("ensemble_plot/GENS_00_innsbruck-flughafen.txt", header = TRUE)
-init  <- min(as.POSIXct(GFS$timestamp-GFS$step*3600,origin="1970-01-01"))
-title <- sprintf("%s\n%s",
-                 "Global Forecast System (GFS) Ensemble Forecast for Innsbruck, Airport",
-                 sprintf("Forecast initialized %s",strftime(init,"%Y-%m-%d %H:%M UTC")))
-
-# observations
-load("STAGEobs_tawes_11121.rda")
-# "2018-03-14 00:00:00" in line 321 (first entry for temperature)
-# "2018-03-13 06:00:00" in line 317 (first entry for rain -> aggregated to 24h sums)
-# "2018-03-23 00:00:00" in line 357
-obs <- obs[318:357,]
-
-getZoo <- function( x, variable ) {
-  x <- subset( x, varname == variable )
-  if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
-  # Else create zoo
-  x <- zoo( subset(x,select=-c(varname,timestamp,step)),
-            as.POSIXct(x$timestamp,origin="1970-01-01") )
-  x
-}
-GFS_t2m  <- getZoo( GFS, "tmp2m" )
-# temperature bias: model simulates temperature for height 1675 (~1070 m higher than true height)
-# => add 10.7 degrees to account for bias (approx. 1 degree / 100 m)
-GFS_t2m <- GFS_t2m + 10.7
-GFS_t2m$obs <- obs[,"obs$tl"]
-# first value of rain is 18 hours later than temperature 
-#(both start at 6:00 but rain is cumulated over 24 hours)
-# -> drop first 3 values of temperature
-GFS_t2m <- GFS_t2m[-c(1:3),]
-
-GFS_rain <- getZoo( GFS, "apcpsfc" )
-GFS_rain$obs <- obs[,"rr6"]
-# precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
-GFS_rain24 <- GFS_rain[c((1:10)*4),]
-for(i in 1:NROW(GFS_rain24)){
-  GFS_rain24[i,] <- colSums(GFS_rain[c((4*i-3):(4*i)),])
-}
-
-# Two POSIXct vectors for the axis
-main  <- seq(min(as.POSIXct(as.Date(index(GFS_t2m)))),max(index(GFS_t2m)),by=86400)
-minor <- seq(min(index(GFS_t2m)),max(index(GFS_t2m)),by=3*3600)
-
-# GFS_t2m_mean <- rowMeans(GFS_t2m)
-GFS_t2m_mean <- GFS_t2m
-GFS_t2m_mean[,1] <- rowMeans(GFS_t2m[,-NCOL(GFS_t2m)])
-GFS_t2m_mean <- GFS_t2m_mean[,1]
-
-# GFS_rain_mean <- rowMeans(GFS_rain)
-GFS_rain24_mean <- GFS_rain24
-GFS_rain24_mean[,1] <- rowMeans(GFS_rain24[,-NCOL(GFS_rain24)])
-GFS_rain24_mean <- GFS_rain24_mean[,1]
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_1_plot, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05, col = "slategray" )
-#plot( GFS_t2m_mean, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-#plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
-#plot( GFS_rain_mean, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_1_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_2_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,c(1,2)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,c(1,2)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_3_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,c(1,2,3)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{-0.3cm}
-\setkeys{Gin}{width=1.1\linewidth}
-<<ensemble_20_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
-par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
-layout( matrix(1:2,ncol=1) )
-plot( GFS_t2m[,-NCOL(GFS_t2m)], screen=1, xaxs="i", xaxt="n", 
-      ylim = range(GFS_t2m)*1.05, col = "slategray" )
-lines( GFS_t2m[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
-plot( GFS_rain24[,-NCOL(GFS_rain24)], screen=1, xaxs="i", xaxt="n", yaxs="i", 
-      ylim=range(GFS_rain24)*c(0,1.05), type = "o", col = "slategray" )
-lines( GFS_rain24[,"obs"], col = "red", lwd = 2)
-abline( v = main, lty = 2 )
-mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
-
-#axis( side = 1, at = minor, strftime(minor,"%H") )
-axis( side = 1, at = main + 42300,  strftime(main,"%b %d"),
-      line = 0, col.ticks=NA, col.axis="gray30", col = NA )
-#     line = 2, col.ticks=NA, col.axis="gray30", col = NA )
-
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.4cm}
-\textbf{Goal:} Predict daily precipitation amount in alpine regions.\\
-\medskip
-\textbf{Input:} Numerical Weather Prediction (NWP) based on a global physical model numerically simulating atmospheric processes on a $50\times 50 \text{km}^2$ grid.\\
-\medskip
-\textbf{Problem:} Uncertain initial conditions, unresolved processes.\\ %$\rightarrow$ requires post-processing.\\
-\vspace{1.3cm}
-%\bigskip
-
-\only<2->{
-\textbf{Idea:} Statistical model with numerical predictions as covariates.\\
-\bigskip
-\textbf{Distribution assumption:}\\
-\vspace{-0.25cm}
-%\hspace{1.5cm}
-%\center
-$$(\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \quad \text{censored at 0.}$$
-}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.4cm}
-\textbf{Observation data} (National Hydrographical Service):\\
+\frametitle{Wind forecasting}
+
+\vspace*{-0.4cm}
+
+\setkeys{Gin}{width=0.75\linewidth}
+\only<1>{
+  \vspace*{0.2cm}
+  \begin{center}
+  \includegraphics{_plot_circforest_stationmap_innsbruck.pdf}
+  \end{center}
+}
+\only<2>{
+  \vspace*{0.2cm}
+  \begin{center}
+  \includegraphics{_plot_circforest_stationmap_innsbruck_box.pdf}
+  \end{center}
+}
+\only<3>{
+  \framesubtitle{~}
+  \vspace*{-0.55cm}
+  \begin{center}
+  \includegraphics{ibk_airport.png}
+  \end{center}
+}
+\only<4>{
+  \framesubtitle{Up-valley wind}
+  \vspace*{-0.55cm}
+  \begin{center}
+  \includegraphics{ibk_airport_east.png}
+  \end{center}
+}
+\only<5>{
+  \framesubtitle{Down-valley wind}
+  \vspace*{-0.55cm}
+  \begin{center}
+  \includegraphics{ibk_airport_west.png}
+  \end{center}
+}
+\only<6>{
+  \framesubtitle{Southerly wind}
+  \vspace*{-0.55cm}
+  \begin{center}
+  \includegraphics{ibk_airport_south.png}
+  \end{center}
+}
+
+\end{frame}
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Wind forecasting}
+
+\vspace{-0.7cm}
+
+\setkeys{Gin}{width=1\linewidth}
+\begin{center}
+\includegraphics{circtree_ibk.pdf}
+\end{center}
+
+\end{frame}
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Wind forecasting}
+
+\textbf{Benchmark:} Against other naive and circular models.
 \begin{itemize}
-\item Daily 24h precipitation sums from July over 28 years (1985--2012).\\
-\item 95 observation stations in Tyrol, Austria.
+\item Climatology: Without covariates.
+\item Persistency: Based on current wind direction.
+\item Circular linear model: Based on current wind speed and wind vector $(u, v)^{\top}$.
 \end{itemize}
-\vspace{1cm}
-%\bigskip
-\textbf{NWP model output} (Global Ensemble Forecast System):\\
+
+\end{frame}
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Wind forecasting}
+
+\textbf{Evaluation:} CRPS skill score for \only<1>{1}\only<2>{3}-hourly predictions (5-fold cross validation).
+\begin{center}
+\setkeys{Gin}{width=0.88\linewidth}
+\only<1>{\includegraphics{circforest_crpsskill_ibk_lag1.pdf}}%
+\only<2>{\includegraphics{circforest_crpsskill_ibk_lag3.pdf}}
+\end{center}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Wind forecasting}
+
+   \only<1-2>{
+      \textbf{Summary}:
       \begin{itemize}
-\item Ensemble predictions of various meteorological quantities related to precipitation, temperature,
-air pressure, \dots.\\
-\item 11 members each using slightly different perturbed initial conditions.
+         \item Trees allow to capture abrupt changes, perform variable selection,
+               and account for possible interactions.
+         %\item Trees capture abrupt changes and possibly non-linear interactions.
+         \item Forests additionally allow for modeling smooth changes.
+         \item Circular trees are easy to interpret, in comparison to additive models.
+          \pause
+         \item Circular trees and forests outperform all other models.
+         \item Even when persistence has no skill, trees and forests outperform climatology.
       \end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
-\textbf{Base variables:}
+   }
+
+   \only<3->{
+      \textbf{Outlook:} Wind shifts.
       \begin{itemize}
-\item Total precipitation
-\item Convective available potential energy
-\item Downwards short wave radiation flux (``sunshine'')
-\item Mean sea level pressure
-\item Preciptable water
-\item 2m maximum temperature
-\item Total column-integrated condensate
-\item Temperature 
-\item Temperature differences in altitude
+      \item Adapt already developed circular regression trees and forests to probabilistic nowcasting 
+        of wind shifts.
+      \item Build two step-model to predict (1) probability of occurrence of a wind shift
+        and (2) possibly direction and magnitude.
       \end{itemize}
-\medskip
-\textbf{Variations:} Ensemble min/max/mean/standard deviation\\
-\vspace{0.1cm}
-$\rightarrow$ 80 covariates.
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace*{0.2cm}
-\textbf{Application for one station:}
-\begin{itemize}
-\item Learn forest model on data from 24 years (1985--2008).
-\item Evaluate on 4 years (2009--2012).
-\end{itemize}
-\end{frame}
-
-<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
-
-#####
-# load observations and covariates 
-data("RainAxams")
-
-# tree and forest formula
-{
-  dt.formula <- df.formula <- 
-      robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
-      tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
-      tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
-      capepow_mean + capepow_sprd + capepow_min + capepow_max + 
-      capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
-      capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
-      dswrf_mean_mean + dswrf_mean_max +  
-      dswrf_sprd_mean + dswrf_sprd_max + 
-      msl_mean_mean + msl_mean_min + msl_mean_max + 
-      msl_sprd_mean + msl_sprd_min + msl_sprd_max +
-      pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
-      pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
-      tmax_mean_mean + tmax_mean_min + tmax_mean_max +
-      tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
-      tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
-      tcolc_sprd_mean + tcolc_sprd_min + tcolc_sprd_max +
-      t500_mean_mean + t500_mean_min + t500_mean_max +
-      t700_mean_mean + t700_mean_min + t700_mean_max +
-      t850_mean_mean + t850_mean_min + t850_mean_max +
-      t500_sprd_mean + t500_sprd_min + t500_sprd_max +
-      t700_sprd_mean + t700_sprd_min + t700_sprd_max +
-      t850_sprd_mean + t850_sprd_min + t850_sprd_max +
-      tdiff500850_mean + tdiff500850_min + tdiff500850_max +
-      tdiff700850_mean + tdiff700850_min + tdiff700850_max +
-      tdiff500700_mean + tdiff500700_min + tdiff500700_max +
-      msl_diff
-
-  }
-
-  # learning data: 24 years (1985 - 2008, both inlcuded)
-  # testing data: 4 successive years (2009, 2010, 2011, 2012)
-  learndata <- RainAxams[RainAxams$year < 2009,]
-  testdata <- RainAxams[RainAxams$year %in% c(2009, 2010, 2011, 2012),]
-
-##############################################################
-# fitting the models
-set.seed(7)
-
-df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
-                 type.tree = "ctree", ntree = 100, censtype = "left", censpoint = 0,
-                 control = ctree_control(teststat = "quad", testtype = "Univ",
-                                         intersplit = TRUE,
-                                         mincriterion = 0, minsplit = 50,
-                                         minbucket = 20), mtry = 27)
-
-#### prepare data for plot of estimated density functions
-# predictions for one day (in each of the four years) 
-# (19th of July 2011 is missing)
-pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
-
-pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
-pdf <- predict(df, newdata = testdata[pdays,], type = "parameter")
-df_mu <- pdf$mu
-df_sigma <- pdf$sigma
-df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
-cbind(df_exp, testdata[pdays,"robs"])
-
-# plot predicted distributions together with observations
-set.seed(7)
-x <- c(0.01, sort(runif(500,0.01,8)))
-y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
-y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
-y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
-y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
-dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})
-
-# switch x-axis back to untransformed scale
-x <- x^(1.6)
-
-# point mass (slightly shifted)
-pm1 <- c(0.15, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.05
-pm2 <- c(0.05, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.01
-pm3 <- c(-0.05, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.03
-pm4 <- c(-0.15, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.07
-
-# predictions
-pred1 <- c(testdata[pdays,"robs"][1]^(1.6), crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
-pred2 <- c(testdata[pdays,"robs"][2]^(1.6), crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
-pred3 <- c(testdata[pdays,"robs"][3]^(1.6), crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
-pred4 <- c(testdata[pdays,"robs"][4]^(1.6), crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
-
-#legendheight
-lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
-lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
-lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
-lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\begin{center}
-\vspace*{-0.2cm}
-<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=5.2>>=
-plot(x = x, y = y1, type = "l", col = "red", 
-     main = paste0("Axams, July ", pday), ylab = "Density", 
-     xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
-     # xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
-     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
-     xlim = c(-2,28))     #  xlim = c(-1.5,8))
-
-lines(x = x, y = y2, type = "l", col = "blue")
-lines(x = x, y = y3, type = "l", col = "darkgreen")
-lines(x = x, y = y4, type = "l", col = "purple")
-legend('topright', c("Predicted distribution", "Point mass at censoring point", "Observation"),
-       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)
-
-# plot point mass
-lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
-lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
-lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
-lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)
-
-points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
-points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
-points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
-points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)
-
-# plot predictions
-points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
-points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
-points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
-points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)
-
-lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
-lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)
-
-# add labels
-text(x = -1.7, y = lh1, labels = "2009", col = "red", cex = 0.8)         # -0.8
-text(x = -1.7, y = lh2, labels = "2010", col = "blue", cex = 0.8)        # -0.8
-text(x = -1.7, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)   # -0.8
-text(x = -1.7, y = lh4, labels = "2012", col = "purple", cex = 0.8)      # -0.8
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.2cm}
-\textbf{Application for one station:}
-\begin{itemize}
-\item Learn forest model on data from 24 years.
-\item Evaluate on 4 years.
-\item 10 times 7-fold cross validation.
-\item Compare to commonly used heteroscedastic censored Gaussian models:
-\end{itemize}
-%\begin{itemize}
-%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-%\item Boosted GAM: automatic variable selection.
-%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
-%\end{itemize}
-\vspace{0.1cm}
-\begin{table}[t!]
-\footnotesize
-
-\hskip-0.4cm\begin{tabular}{ l l l }
-\hline
-Model & Type & Variable Selection \\
-\hline
-\vspace{-0.15cm}
-\textbf{Forest} & recursive    & automatic \\ 
-                & partitioning &           \\ 
-\hline
-\vspace{-0.15cm}
-\textbf{prespecified GAM} & spline  & based on expert \\
-                          & in each & knowledge      \\
-\hline
-\vspace{-0.15cm}
-\textbf{boosted GAM} & spline  & automatic  \\
-                     & in each &            \\
-\hline
-\vspace{-0.15cm}
-\textbf{EMOS} & linear & total precipitation mean \\
-              &        & and standard deviation   \\
-\hline
-\end{tabular}
-\end{table}
-
-\end{frame}
-
-<<echo=FALSE, results=hide>>=
-#### cross validation rain
-if(file.exists("crps_cross.rda")){
-  load("crps_cross.rda")
-} else {
-
-  nrep_cross <- 10
-  seed <- 7
-
-  res_cross <- mclapply(1:nrep_cross,
-                        function(i){
-
-                          set.seed(seed*i)
-
-                          # randomly split data in 7 parts each including 4 years
-                          years <- 1985:2012
-                          testyears <- list()
-                          for(j in 1:7){
-                            testyears[[j]] <- sample(years, 4, replace = FALSE)
-                            years <- years[!(years %in% testyears[[j]])]
-                          }
-
-                          #crps <- matrix(nrow = 7, ncol = 7)
-                          reslist <- list()
-                          for(k in 1:7){
-                            test <- testyears[[k]]
-                            train <- c(1985:2012)[!c(1985:2012) %in% test]
-
-                            res <- evalmodels(station = "Axams",
-                                              train = train,
-                                              test = test,
-                                              gamboost_cvr = TRUE)
-
-                            #crps[k,] <- res$crps
-                            reslist[[k]] <- res
-                          }
-
-                          #colnames(crps) <- names(res$crps)
-                          return(reslist)
-                        },
-                        mc.cores = detectCores() - 1
-  )
-
-  # extract CRPS
-  crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
-  # loop over all repetitions
-  for(i in 1:length(res_cross)){
-    #loop over all 7 folds (for 7 methods)
-    crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
-    for(j in 1:length(res_cross[[1]])){
-      crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
-    }
-    crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
-  }
-  colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps)
-
-  save(crps_cross, file = "crps_cross.rda")
-}
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\begin{center}
-\vspace*{-0.2cm}
-<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.6, width=6.8>>=
-#par(mar = c(2.5,2,1,2))
-boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
-        names = c("Distributional Forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
-        main = "Cross validation (with reference model EMOS)",
-        cex.main=1.4,
-        cex.lab=1.2,
-        ylab = "CRPS skill score", col = "lightgray") 
-abline(h = 0, col = pal["EMOS"], lwd = 2)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\vspace{0.2cm}
-\textbf{Application for all 95 stations:}
-\begin{itemize}
-\item Learn forest model on data from 24 years (1985--2008).
-\item Evaluate on 4 years (2009--2012).
-\item Compare to commonly used heteroscedastic censored Gaussian models. %:\\
-\end{itemize}
-%\begin{itemize}
-%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-%\item Boosted GAM: automatic variable selection.
-%\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
-%\end{itemize}
-
-%\vspace{0.1cm}
-%\begin{table}[t!]
-%\footnotesize
-%\nodindent
-%\hskip-0.4cm\begin{tabular}{ l l l }
-%\hline
-%Model & Type & Variable Selection \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{Forest} & recursive    & automatic \\ 
-%                & partitioning &           \\ 
-%\hline
-%\vspace{-0.15cm}
-%\textbf{prespecified GAM} & spline  & based on expert \\
-%                          & in each & knowledge      \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{boosted GAM} & spline  & automatic  \\
-%                     & in each &            \\
-%\hline
-%\vspace{-0.15cm}
-%\textbf{EMOS} & linear & total precipitation mean \\
-%              &        & and standard deviation   \\
-%\hline
-%\end{tabular}
-%\end{table}
-\end{frame}
-
-<<echo = FALSE>>=
-#### prediction over all stations 24 - 4
-if(file.exists("crps_24to4_all.rda")){
-  load("crps_24to4_all.rda")
-} else {
-
-  data("StationsTyrol")
-  stations <- StationsTyrol$name
-  test <- 2009:2012
-  train <- 1985:2008
-
-  res_24to4_all <- mclapply(1:length(stations),
-                            function(i){
-
-                              set.seed(7)
-
-                              res <- evalmodels(station = stations[i],
-                                                train = train,
-                                                test = test,
-                                                gamboost_cvr = TRUE)
-
-                              return(res)
-                            },
-                            mc.cores = detectCores() - 1
-  )
-
-  # extract crps
-  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
-  # loop over all stations
-  for(i in 1:length(stations)){
-    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
-  }
-
-  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
-  rownames(crps_24to4_all) <- stations
-
-  save(crps_24to4_all, file = "crps_24to4_all.rda")
-}
-
-# skill score
-s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
-colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
-
-## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
-
-crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]
-
-# best method
-bst <- apply(crps_map, 1, which.min)
-
-# distance of forest to best other method
-dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]
-
-# breaks/groups
-brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
-#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
-grp <- cut(dst, breaks = brk)
-
-# HCL colors (relatively flashy, essentially CARTO Tropic)
-clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)
-
-library("raster") # dem (digital elevation model)
-library("sp")     # gadm www.gadm.org/country
-
-#load("plot_map_rain/data/tirol.rda")
-#load("plot_map_rain/data/dem.rda")
-
-data(StationsTyrol, package = "RainTyrol")
-data(MapTyrol, package = "RainTyrol")
-# data(MapTyrol_border, package = "RainTyrol")
-# Create SpatialPointsDataFrame from station list
-sp <- SpatialPointsDataFrame(subset(StationsTyrol,
-                                    select=c(lon,lat)),
-                             data = subset(StationsTyrol,
-                                           select = -c(lon,lat)),
-                             proj4string = crs(MapTyrol$RasterLayer))
-@
-
-\begin{frame}[fragile]
-\frametitle{Probabilistic Forecasting of Precipitation}
-\begin{center}
-%\vspace*{-2cm}
-
-\setkeys{Gin}{width=0.95\textwidth}
-<<map, fig=TRUE, echo=FALSE, width=10, height=6.5>>=
-
-  ## plot map of Tyrol with all 95 observations
-  layout(cbind(1, 2), width = c(9, 1))
-  par(mar = c(5,4,4,0.1))
-  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
-                main="Stations in Tyrol", cex.main=1.7,
-                ylab = "Latitude", xlab = "Longitude",
-                cex.lab = 1.3, 
-                xlim = c(9.8,13.2), 
-                ylim = c(46.6, 47.87))
-  plot(MapTyrol$SpatialPolygons, add = TRUE)
-  #points(sp[70,], pch = 19, col = "black", cex = 1.85)
-  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
-  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
-  text(x = 10.25, y = 47.82, labels = "Models with lowest CRPS")
-  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
-  par(mar = c(0.5,0.2,0.5,2.3))
-  ## legend
-  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
-       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
-  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
-  axis(4, at = brk, las = 1, mgp=c(0,-0.4,-1.2))
-
-@
-\end{center}
-\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_ll}
-%\end{center}
-%\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_RMSE_par}
-%\end{center}
-%\end{frame}
-
-%\begin{frame}[fragile]
-%\frametitle{Simulation}
-%\begin{center}
-%\vspace*{-1cm}
-%\setkeys{Gin}{width=0.8\linewidth}
-%\includegraphics{sim_RMSE_exp}
-%\end{center}
-%\end{frame}
-
+   }
+
+\end{frame}
+
+%-------------------------------------------------------------------
+\subsection{Software}
+%-------------------------------------------------------------------
+
+%% SLIDE
 \begin{frame}
 \frametitle{Software}
-\vspace{0.4cm}
-\textbf{Package:} \emph{disttree} available on R-Forge at\\
-
-\medskip
-
-\url{https://R-Forge.R-project.org/projects/partykit/}\\
+%\vspace{0.4cm}
+\textbf{Software:} \emph{circtree}
+available on R-Forge at\\
+
+\medskip
+
+\small{\url{https://R-Forge.R-project.org/projects/partykit/}}\\
 
 \bigskip
-\bigskip
+\medskip
 
@@
 \begin{tabular}{ll}
-\code{distfit}    & Distributional fits (ML, \code{gamlss.family}/custom \code{list}).\\
+\code{circfit}    & Circular distributional fits (ML).\\
 & No covariates. \\
-\code{disttree}   & Distributional trees (\code{ctree}/\code{mob} + \code{distfit}).\\
+\code{circtree}   & Circular distributional trees (\code{ctree}/\code{mob} + \code{circfit}).\\
 & Covariates as partitioning variables. \\
-\code{distforest} & Distributional forests (ensemble of \code{disttree}s).\\
+\code{circforest} & Circular distributional forests (ensemble of \code{circtree}s).\\
 & Covariates as partitioning variables.
@@
 
-\end{frame}
-
+\bigskip
+\medskip
+
+\textbf{Correspondingly:} \emph{disttree} for any other response distribution (e.g., \code{gamlss.family}).
+
+\end{frame}
+
+%-------------------------------------------------------------------
 \subsection{References}
-
+%-------------------------------------------------------------------
+
+%% SLIDE
 \begin{frame}
@@
 \vspace{-0.2cm}
-
 \footnotesize
-Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
+
+Lang MN, Schlosser L, Hothorn T, Mayr GJ, Stauffer R, Zeileis A (2020).
+\dquote{Circular Regression Trees and Forests with an
+Application to Probabilistic Wind Direction Forecasting.}
+arXiv:12001.0041, {\it arXiv.org E-Print Archive}, in review, 2020.
+
+\medskip
+
+Schlosser L, Lang MN, Hothorn T, Mayr GJ, Stauffer R, Zeileis A (2019).
+\dquote{Distributional Trees for Circular Data.}
+\emph{Proceedings of the 34th International Workshop on Statistical Modelling}, 
+\textbf{1}, 226--231.
+\url{https://eeecon.uibk.ac.at/~zeileis/papers/Schlosser+Lang+Hothorn-2019.pdf}
+
+\medskip
+
+Schlosser L, Hothorn T, Stauffer R, Zeileis A (2019).
 \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
-  \emph{arXiv 1804.02921}, arXiv.org E-Print Archive.
-  \url{http://arxiv.org/abs/1804.02921}
-
+\emph{The Annals of Applied Statistics}, \textbf{13}(3), 1564--1589.
+\doi{10.1214/19-AOAS1247}
+
+\bigskip
 \medskip
@@
  \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
- \textit{Journal of Computational and Graphical Statistics},
+ \emph{Journal of Computational and Graphical Statistics},
  \textbf{15}(3), 651--674.
@@
  \dquote{Model-Based Recursive Partitioning.}
-  \textit{Journal of Computational and Graphical Statistics},
+  \emph{Journal of Computational and Graphical Statistics},
   \textbf{17}(2), 492--514.
@@
 
-\medskip
-
-Hothorn T, Zeileis A (2015).
- \dquote{\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
- \textit{Journal of Machine Learning Research},
- \textbf{16}, 3905--3909.
- \url{http://www.jmlr.org/papers/v16/hothorn15a.html}
-
-\medskip
-
-Stasinopoulos DM, Rigby RA (2007).
-  \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in \textsf{R}.}
-  \textit{Journal of Statistical Software}, 
-  \textbf{23}(7), 1--46.
-  \doi{10.18637/jss.v023.i07}
-
-%\medskip
-
-%Seibold H, Zeileis A, Hothorn T (2017).
-%  \dquote{Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.}
-%  \textit{Statistical Methods in Medical Research}, 
-%  \textbf{12}(1), 45--63.
-%  \doi{10.1177/0962280217693034}
-
-%\medskip
-
-%Hothorn T, Zeileis A (2017).
-%  \dquote{Transformation Forests.}
-%  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
-%  \url{http://arxiv.org/abs/1701.02110}
-
-\medskip
-
-Hofner B, Mayr A, Schmid M (2016).
-  \dquote{\textbf{gamboostLSS}: An \textsf{R} Package for Model Building and Variable Selection in the {GAMLSS} Framework}
-  \textit{Journal of Statistical Software},
-  \textbf{74}(1), 1--31.
-  \doi{10.18637/jss.v074.i01}
-
-\end{frame}
-
-%\begin{frame}
-%\frametitle{References (2)}
-
-%\vspace{-0.2cm}
-
-%\footnotesize
-
-%Gneiting T, Raftery A E, Westveld III A H, Goldman T (2005).
-%  \dquote{Calibrated Probabilistic Forecasting using Ensemble Model Output Statistics and %Minimum CRPS Estimation}
-%  \textit{Monthly Weather Review}, 
-%  \textbf{133}(5), 1098--1118.
-%  \doi{10.1175/MWR2904.1}
-
-%\medskip
-
-% Reference for the observation data set
-%BMLFUW (2016).
-%  \dquote{{B}undesministerium f\"ur {L}and und {F}orstwirtschaft, {U}mwelt und {W}asserwirtsch%aft ({BMLFUW}), {A}bteilung {IV/4} -- {W}asserhaushalt}
-%  \text{Available at \url{http://ehyd.gv.at}}, 
-%  \text{Accessed: 2016--02--29}.
-
-%\medskip
-
-% Reference for the GFS reforecast version 2 data set 
-%Hamill T M, Bates G T, Whitaker J S, Murray D R, Fiorino M, Galarneau T J Jr., Zhu Y, Lapenta %W (2013).
-%  \dquote{{NOAA}'s Second-Generation Global Medium-Range Ensemble Reforecast Dataset}
-%  \textit{Bulletin of the American Meteorological Society}, 
-%  \textbf{94}(10), 1553--1565.
-%  \doi{10.1175/BAMS-D-12-00014.1}
-
-%\medskip
-
-%Breiman L (2001).
-%  \dquote{Random {Forests}}
-%  \textit{Machine Learning}, 
-%  \textbf{45}(1), 5--32.
-%  \doi{10.1023/A:1010933404324}
-
-%Messner J W,  Mayr G J, Zeileis A (2016).
-%  \dquote{Heteroscedastic Censored and Truncated Regression with \textbf{crch}}
-%  \textit{The \textsf{R} Journal}, 
-%  \textbf{8}(1), 173--181.\\
-%  \scriptsize
-%  \url{https://journal.R-project.org/archive/2016-1/messner-mayr-zeileis.pdf}
-%  \footnotesize
-
-%\medskip
-
-%Stauffer R, Mayr G J, Messner J W, Umlauf N, Zeileis A (2017).
-%  \dquote{Spatio-Temporal Precipitation Climatology Over Complex Terrain Using a Censored %Additive Regression Model}
-%  \textit{International Journal of Climatology}, 
-%  \textbf{37}(7), 3264--3275.
-%  \doi{10.1002/joc.4913}
-
-%\medskip
-
-%Gebetsberger M, Messner J W, Mayr G J, Zeileis A (2017).
-%  \dquote{Fine-Tuning Non-Homogeneous Regression for Probabilistic Precipitation Forecasts: %Unanimous Predictions, Heavy Tails, and Link Functions}
-%  \textit{Monthly Weather Review}, 
-%  \text{Forthcoming}.
-%  \doi{10.1175/MWR-D-16-0388.1}
-
-%\medskip
-
-%{CGIAR-CSI} (2016).
-%  \dquote{{SRTM} 90m Digital Elevation Database {v4.1}}
-%  \text{Available at \url{http://srtm.csi.cgiar.org}}, 
-%  \text{Accessed: 2016--07--10}.
-
-%\end{frame}
-
+\end{frame}
+
+%-------------------------------------------------------------------
+\subsection{Appendix}
+%-------------------------------------------------------------------
+\backupbegin
+
+%% SLIDE
+\begin{frame}
+~~~
+\end{frame}
+
+%% SLIDE
+\begin{frame}
+\frametitle{Wind forecasting: Vienna airport}
+
+\vspace*{-0.2cm}
+
+\setkeys{Gin}{width=0.75\linewidth}
+\begin{center}
+\includegraphics{_plot_circforest_stationmap_wien.pdf}
+\end{center}
+\end{frame}
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Wind forecasting: Vienna airport}
+
+\vspace{-0.7cm}
+
+\setkeys{Gin}{width=1\linewidth}
+\begin{center}
+\includegraphics{circtree_vie3h.pdf}
+\end{center}
+
+\end{frame}
+
+%% SLIDE
+\begin{frame}[fragile]
+\frametitle{Wind forecasting: Vienna airport}
+
+\textbf{Evaluation:} CRPS skill score for \only<1>{1}\only<2>{3}-hourly predictions (5-fold cross validation).
+\begin{center}
+\setkeys{Gin}{width=0.88\linewidth}
+\only<1>{\includegraphics{circforest_crpsskill_vie_lag1.pdf}}%
+\only<2>{\includegraphics{circforest_crpsskill_vie_lag3.pdf}}
+\end{center}
+
+\end{frame}
+\backupend
 \end{document}

### 2020-03-23 pkg/circtree/inst/wind_paper/talks/ACINN_graduateseminar2020/slides/slides.Rnw
Include Achim's Input to slides (ACINN graduate seminar)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3001 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a226746d60d62d3b29c747d0a714a10a4e5baa07

@@
 \begin{itemize}
-\item Hourly crime data: $[0\text{h}, 24\text{h})$ with $0\text{h} = 24\text{h}$.
-\item Directional behavior of animals: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
-\item Wind directions: $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
+\item Hourly crime data: $[0\text{h}, 24\text{h})$.
+\item Directional behavior of animals: $[0^{\circ}, 360^{\circ})$.
+\item Wind directions: $[0^{\circ}, 360^{\circ})$.
 \end{itemize}
@@
 \frametitle{Learning circular trees and forests}
-      \begin{minipage}{0.7\textwidth}
+      \begin{minipage}{1\textwidth}
       {\bf Forest:} 
@@
       random input variable sampling.
-      \item<2-> Combine / average over estimated $T$ trees:\\
+      \item<2-> Aggregate / average over estimated $T$ trees:\\
       To stabilize and regularize the model.
+      %\item<3-> Only observations $\bold{x}_i$ in the same subgroup as a 
+      %(new) observation $\bold{x}$ enter the parameter estimation.
+      \item<2-> Weighted parameter estimation:\\The weight of observation $i$ for the parameter 
+        estimatation for a 
+        (new) observation $j$ is the mean of assignments to the same node.
       \end{enumerate}
       \end{minipage}
+\begin{center}
+\minipage{0.38\textwidth}
+\vspace{0.5em}
+\resizebox{0.94\textwidth}{!}{
+\begin{tikzpicture}
+  \visible<3->{
+  \draw (11.1,5) node{\phantom{.}};
+  \draw (18.5,9) node{\phantom{.}};
+
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (12, 9) {};
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (11.5, 8) {\only<4->{$i, j$}};
+  \draw[-, line width=1pt] (n0) -- (n1);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (12.5, 8) {};
+  \draw[-, line width=1pt] (n0) -- (n2);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (12, 7) {};
+  \draw[-, line width=1pt] (n2) -- (n3);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (13, 7) {};
+  \draw[-, line width=1pt] (n2) -- (n4);
+  }
+
+  \visible<5->{
+
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n1_0) at (12+3.25, 9) {};
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n1_1) at (11.5+3.25, 8) {};
+  \draw[-, line width=1pt] (n1_0) -- (n1_1);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1_2) at (12.5+3.25, 8) {\only<6->{$i, j$}};
+  \draw[-, line width=1pt] (n1_0) -- (n1_2);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n1_3) at (11+3.25, 7) {};
+  \draw[-, line width=1pt] (n1_1) -- (n1_3);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1_4) at (12+3.25, 7) {};
+  \draw[-, line width=1pt] (n1_1) -- (n1_4);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1_5) at (10.5+3.25, 6) {};
+  \draw[-, line width=1pt] (n1_3) -- (n1_5);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n1_6) at (11.5+3.25, 6) {};
+  \draw[-, line width=1pt] (n1_3) -- (n1_6);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1_7) at (11+3.25, 5) {};
+  \draw[-, line width=1pt] (n1_6) -- (n1_7);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1_8) at (12+3.25, 5) {};
+  \draw[-, line width=1pt] (n1_6) -- (n1_8);
+
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n2_0) at (12+6, 9) {};
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n2_1) at (11.5+6, 8) {};
+  \draw[-, line width=1pt] (n2_0) -- (n2_1);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n2_2) at (12.5+6, 8) {};
+  \draw[-, line width=1pt] (n2_0) -- (n2_2);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n2_3) at (11+6, 7) {\only<6->{$i\vphantom{j}$}};
+  \draw[-, line width=1pt] (n2_1) -- (n2_3);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n2_4) at (12+6, 7) {\only<6->{$j$}};
+  \draw[-, line width=1pt] (n2_1) -- (n2_4);
+  }
+\end{tikzpicture}
+}
+\endminipage
+\end{center}
 

### 2020-04-10 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Add draft of rebuttal
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3005 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9e2f1d523d2c40ff3ba678a482988c9ddf5b97b7

\documentclass[english, noconfig]{uibklttr}

\newcommand{\section}[1]{{\Large{\textbf{#1}}}}
\newcommand{\subsection}[1]{{\large \textbf{#1}}}
\newcommand{\subsubsection}[1]{{\large \textbf{#1}}}
\newcommand{\tightlist}[0]{}

% Some colors
\definecolor{myblue}{rgb}{0.2,0.2,0.9}
\definecolor{mygray}{rgb}{0.7,0.7,0.7}

\makeatletter
\newenvironment{thebibliography}[1]
     {\list{\@biblabel{\@arabic\c@enumiv}}%
           {\settowidth\labelwidth{\@biblabel{#1}}%
            \leftmargin\labelwidth
            \advance\leftmargin\labelsep
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\newcommand\newblock{\hskip .11em\@plus.33em\@minus.07em}
\makeatother
\usepackage{natbib}

%\usepackage{subcaption}
%\captionsetup{font={bf,small},skip=0.25\baselineskip}
%\captionsetup[subfigure]{font={bf,small}, skip=1pt, singlelinecheck=false}

\usepackage{amsmath}

\usepackage{float} % For figure placements
\usepackage{enumitem}
\setitemize{itemsep=0em}
\setlist{itemsep = 0em}

\usepackage{parskip}
\setlength{\parskip}{.4em}

% fake-float with caption for figures
\usepackage{caption}
\usepackage{newfloat}
\DeclareFloatingEnvironment[fileext=lot]{figure}
\DeclareCaptionFont{myblue}{\color{myblue}}
\captionsetup{labelfont={bf},textfont={it}}

% short re
\newcommand{\sre}[1]{
    \begin{color}{uibkgraym}#1\end{color}%
}
% long re
\newenvironment{re}{
    \begin{color}{uibkgraym}
        \itshape
}{
    \end{color}
}
\newcommand{\todo}[1]{
    \begin{color}{red}\textbf{TODO:} #1\end{color}%
}
\newcommand{\note}[1]{
    \begin{color}{orange}\textbf{Reto's comment:} #1\end{color}%
}
\newcommand{\see}[2]{Page \textbf{#1}, line \textbf{#2}.}
\newcommand{\msee}[3]{Page \textbf{#1}, lines \textbf{#2--#3}.}
\newcommand{\fig}[1]{\textit{now Fig~#1}}
\newcommand{\nsec}[1]{\textit{now Sec~#1}}

\begin{document}

\setkomavar{subject}{Revision of `npg-2019-49'}
\setkomavar{signature}{Moritz Lang\\Corresponding Author}

\setkomavar{fromname}[Name]{Moritz Lang}
\setkomavar{fromemail}{Moritz.Lang@uibk.ac.at}

\setkomavar{fromaddress}{University of Innsbruck, Department of Statistics, Universit\"atsstrasse
15, 6020 Innsbruck, Austria}

%\begin{letter}[fromemail]{To Maxime Taillardat}
%\opening{Dear Christopher Paciorek}
%
%We thank you, as well as both reviewers, for the detailed and fruitful feedback
%regarding our manuscript ``\textbf{Bivariate Gaussian models for wind vectors in a
%distributional regression framework}''.
%
%\vspace{5mm}
%
%We have carefully revised our manuscript according to your comments and
%suggestions. The most substantial changes are the following:
%
%\begin{itemize}
%\item
%An additional comparison to the work of \citet{schuhen.etal:2012} and of
%\citet{pinson:2012} has been included. Both comparisons nicely confirm our
%previous results. To keep the manuscript as clear as possible, we describe the
%comparisons in the appendix and provide short summaries of the main finding
%in the discussion section.
%
%\item
%The introduction of the logarithmic score has been
%extended. We now explicitly derive the employed formulation of the logarithmic
%score and provide additional references.
%
%\item
%We have added more information on the estimation procedure and describe the key
%steps involved in the model fitting. Please also remark that for consistency we
%have changed `mean and variance` to `location and scale parameter` within the
%abstract and the introductory part of the discussion.
%\end{itemize}
%
%\vspace{5mm}
%
%On the following pages a point-to-point response to both reviewers and to the
%editor will be given. The attached manuscript highlights the changes in the
%text in blue color.
%
%\closing{Your sincerely,}
%
%\end{letter}
%
%\newpage

\section{Editor: Prof. Peter Smith - Comments to the Author}

\begin{itemize}

\item Please can you rewrite the Summary to give the application more prominence.
Currently, the last sentence reads to me like an afterthought. Furthermore, I
would prefer you to use the term ‘application’ rather than ‘case study’
throughout the paper, since in Applied Statistics papers the methodological
developments should be motivated by the application.

\item Page 2, line 46: ‘… structure can capture non-additive …’?

\item Page 2, lines 54 and 55: ‘… mean, whereas modelling the conditional … distribution would also allow uncertainty in the forecast to be estimated.’?

\item Page 6, line 16 and Equation (2.6): bold the vector $\beta$.

\item Page 6, lines 29 and 30: ‘In general, for additive models a proper model …’?

\item Equations (3.1) to (3.4) and elsewhere: I find the use of multiple letters for a single superscript or subscript awkward and prefer to keep lower case letters for indices, so please consider replacing them, for example, T for tree and F for forest. Also for Equation (4.2), how about using parentheses rather than subscripts?

\item Equation (3.4): bold vector $z$.

\item Page 9, line 39: delete ‘respectively’.

\item Page 15, line 23: replace ‘leveraged’ with ‘used’?

\end{itemize}

\newpage

\section{Associate Editor - Comments to the Author}

Both referees agree this is a good paper, and have only a few suggestions for
changes, so I recommend minor revisions to address these.

I have two short suggestions in addition to this:

\begin{itemize}
\item Can you comment on the typical computational requirements of the various
methods used in your application, in terms of hardware required and length of
computing time.

\item Is it possible to include a link to computer code to reproduce your
analysis? This is not essential for publication but would be helpful for
readers who wish to perform similar analyses. The R package you have provided
for your method is very helpful in this regard.  
\end{itemize}

\newpage

\section{Referee: 1 - Comments to the Author}

I found Circular Regression Trees and Forests to be a very nicely written
paper, with a meaningful contribution to the literature on regression models
for a circular response variable.

I have very few suggestions to the authors; but here are two:

\begin{itemize}
\item On page 2, what is the reference for the phrase "Additionally, they
developed a combined ... version ...".

\item On page 11, regarding "As the implementation cannot handle circular
covariates ...": It could handle circular covariates in a fashion similar to
Lund (1999), using a rotational or reflective relationship, or modeling the
association with a trigonometric polynomial. Adding this to their comparison
model, or at least mentioning the possibility of doing so, would be more
appropriate than stating that circular covariates cannot be accommodated.
\end{itemize}

\vspace{0.5cm}
References

Lund UJ (1999). ``Least Circular Distance Regression for Directional Data.''
\emph{Journal of Applied Statistics}, \textbf{26}\,(6), 723--733.
doi:10.1080/02664769922160.

\newpage

\section{Referee: 2 - Comments to the Author}

This is a well-written paper on using circular regression trees for analyzing a
specific type of circular data. The probabilistic approach is well-founded. The
experiments on wind direction data are well organized and the results are
thoroughly analyzed.

\begin{itemize}
\item The authors could address model order determination problem using the
Bayes Information Criterion (Schwartz, 1978, Annals of Statistics).

\item I would like to point out related works on circular regression models for
shape representation (Kashyap and Chellappa) and circular random forests for
vehicle direction estimation (Hara and Chellappa) given below. I feel there is
some overlap between this paper and the work of Hara and Chellappa on using
circular statistics for vehicle orientation estimation. The authors may include
some comparisons with Hara, Chellappa work.
\end{itemize}

\vspace{0.5cm}
References

Hara K, Chellappa R (2017). ``Growing Regression Tree Forests by
Classification for Continuous Object Pose Estimation.'' \emph{International
Journal of Computer Vision}, \textbf{122}\,(2), 292--312.
doi:10.1007/s11263-016-0942-1.

Kashyap R, Chellappa R (1981). ``Stochastic Models for Closed Boundary
Analysis: Representation and Reconstruction.'' \emph{IEEE Transactions on
Information Theory}, \textbf{27}\,(5), 627--637. doi:10.1109/TIT.1981.1056390.

\end{document}

### 2020-04-10 pkg/circtree/inst/wind_paper/draft/JRSSC/draft_jrssc_submitted_2020_01_09.tex
Create diff option in makefile
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3006 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7afc18d26503bf349a371078d37fa6d921a0e74c

--- 2020-01-08 pkg/circtree/inst/wind_paper/draft/draft.Rnw
@@
-\documentclass[nojss]{jss}
-
-%% packages
-\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref}
-\usepackage[all]{hypcap}
+\documentclass{statsoc}
+
+\usepackage[a4paper]{geometry}
+%\usepackage{graphicx}
+
+%%% packages
+\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref,natbib}
+%\usepackage[all]{hypcap} %% FAILS IN JRSSC
 
@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}
 
 %% commands
-\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
-\renewcommand{\Prob}{\mathbb{P} }
-\renewcommand{\E}{\mathbb{E}}
-\newcommand{\V}{\mathbb{V}}
-\newcommand{\Var}{\mathbb{V}}
-\newcommand{\R}{\mathbb{R} }
-\newcommand{\N}{\mathbb{N} }
-\newcommand{\C}{\mathbb{C} }
+%\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
+%\newcommand{\Prob}{\mathbb{P} }
+%\newcommand{\E}{\mathbb{E}}
+%\newcommand{\V}{\mathbb{V}}
+%\newcommand{\Var}{\mathbb{V}}
+%\newcommand{\R}{\mathbb{R} }
+%\newcommand{\N}{\mathbb{N} }
+%\newcommand{\C}{\mathbb{C} }
 \newcommand{\argmin}{\operatorname{argmin}\displaylimits}
 \newcommand{\argmax}{\operatorname{argmax}\displaylimits}
-\newcommand{\LS}{\mathcal{L}_n}
-\newcommand{\TS}{\mathcal{T}_n}
-\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
-\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
-\newcommand{\F}{\mathcal{F}}
-\newcommand{\A}{\mathcal{A}}
-\newcommand{\yn}{y_{\text{new}}}
-\newcommand{\z}{\mathbf{z}}
-\newcommand{\X}{\mathbf{X}}
-\newcommand{\Y}{\mathbf{Y}}
-\newcommand{\sX}{\mathcal{X}}
-\newcommand{\sY}{\mathcal{Y}}
-\newcommand{\T}{\mathbf{T}}
-\newcommand{\x}{\mathbf{x}}
-\renewcommand{\a}{\mathbf{a}}
-\newcommand{\xn}{\mathbf{x}_{\text{new}}}
-\newcommand{\y}{\mathbf{y}}
-\newcommand{\w}{\mathbf{w}}
-\newcommand{\ws}{\mathbf{w}_\cdot}
-\renewcommand{\t}{\mathbf{t}}
-\newcommand{\M}{\mathbf{M}}
-\renewcommand{\vec}{\text{vec}}
-\newcommand{\B}{\mathbf{B}}
-\newcommand{\K}{\mathbf{K}}
-\newcommand{\W}{\mathbf{W}}
-\newcommand{\D}{\mathbf{D}}
+%\newcommand{\LS}{\mathcal{L}_n}
+%\newcommand{\TS}{\mathcal{T}_n}
+%\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
+%\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
+%\newcommand{\F}{\mathcal{F}}
+%\newcommand{\A}{\mathcal{A}}
+%\newcommand{\yn}{y_{\text{new}}}
+%\newcommand{\z}{\mathbf{z}}
+%\newcommand{\X}{\mathbf{X}}
+%\newcommand{\Y}{\mathbf{Y}}
+%\newcommand{\sX}{\mathcal{X}}
+%\newcommand{\sY}{\mathcal{Y}}
+%\newcommand{\T}{\mathbf{T}}
+%\newcommand{\x}{\mathbf{x}}
+%\renewcommand{\a}{\mathbf{a}}
+%\newcommand{\xn}{\mathbf{x}_{\text{new}}}
+%\newcommand{\y}{\mathbf{y}}
+%\newcommand{\w}{\mathbf{w}}
+%\newcommand{\ws}{\mathbf{w}_\cdot}
+%\renewcommand{\t}{\mathbf{t}}
+%\newcommand{\M}{\mathbf{M}}
+%\renewcommand{\vec}{\text{vec}}
+%\newcommand{\B}{\mathbf{B}}
+%\newcommand{\K}{\mathbf{K}}
+%\newcommand{\W}{\mathbf{W}}
+%\newcommand{\D}{\mathbf{D}}
 \newcommand{\I}{\mathbf{I}}
-\newcommand{\bS}{\mathbf{S}}
-\newcommand{\cellx}{\pi_n[\x]}
-\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
-\newcommand{\err}{\text{Err}}
-\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
-\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
-\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
-\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
-\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
-\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
-\newcommand{\bft}{\mathbf{t}}
-
-\definecolor{changes}{rgb}{1,1,0.7}
-\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}
-
-\numberwithin{equation}{section}
-
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-library("disttree")
-library("circtree")
-
-## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-# set function for parallelization
-applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))
-
-@
-
-\title{Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting}
-\Shorttitle{Circular Regression Trees and Forests}
-
-\author{Moritz N. Lang\\Universit\"at Innsbruck
-   \And Lisa Schlosser\\Universit\"at Innsbruck
-   \And Torsten Hothorn\\Universit\"at Z\"urich
-   \AND Georg~J.~Mayr\\Universit\"at Innsbruck
-   \And Reto Stauffer\\Universit\"at Innsbruck
-   \And Achim Zeileis\\Universit\"at Innsbruck}
-\Plainauthor{Lang, Schlosser, Hothorn, Mayr, Stauffer, Zeileis}
-
-\Abstract{
+%\newcommand{\bS}{\mathbf{S}}
+%\newcommand{\cellx}{\pi_n[\x]}
+%\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
+%\newcommand{\err}{\text{Err}}
+%\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
+%\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
+%\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
+%\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
+%\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
+%\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
+%\newcommand{\bft}{\mathbf{t}}
+
+%\definecolor{changes}{rgb}{1,1,0.7}
+%\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}
+%
+%\numberwithin{equation}{section}
+
+\graphicspath{{../}}
+
+\title[Circular Regression Trees and Forests]{Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting}
+
+\author{Moritz N. Lang}
+\address{
+  Universit\"at Innsbruck,
+  Innsbruck, Austria.}
+\email{Moritz.Lang@uibk.ac.at}
+\author{Lisa Schlosser}
+\address{
+  Universit\"at Innsbruck,
+  Innsbruck, Austria}
+\author{Torsten Hothorn.}
+\address{
+  Universit\"at Z\"urich,
+  Z\"urich, Switzerland.}
+\author{Georg~J.~Mayr}
+\address{
+  Universit\"at Innsbruck,
+  Innsbruck, Austria.}
+\author{Reto Stauffer}
+\address{
+  Universit\"at Innsbruck,
+  Innsbruck, Austria.}
+\author[Lang, Schlosser, Hothorn, Mayr, Stauffer, Zeileis]{Achim Zeileis}
+\address{
+  Universit\"at Innsbruck,
+  Innsbruck, Austria.}
+
+\begin{document}
+
+\begin{abstract}
 While circular data occur in a wide range of scientific fields, the methodology
@@
 other common approaches as a benchmark.
-}
-
-\Keywords{circular data, regression trees, random forests, distributional regression, von~Mises distribution}
-
-\Address{
-  Moritz N. Lang, Lisa Schlosser, Achim Zeileis \\
-  Universit\"at Innsbruck \\
-  Department of Statistics \\
-  Faculty of Economics and Statistics \\
-  Universit\"atsstr.~15 \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Moritz.Lang@uibk.ac.at}, \\
-  \phantom{E-mail: }\email{Lisa.Schlosser@uibk.ac.at},\\
-  \phantom{E-mail: }\email{Achim.Zeileis@R-project.org} \\
-  URL: \url{https://www.uibk.ac.at/statistics/personal/moritz-lang/},\\
-  \phantom{URL: }\url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
-  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\
-
-  Reto Stauffer \\
-  Universit\"at Innsbruck \\
-  Digital Science Center and \\ 
-  Department of Statistics \\
-  Faculty of Economics and Statistics \\
-  Universit\"atsstr.~15 \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
-  URL: \url{https://retostauffer.org/}\\
-
-  Georg~J.~Mayr\\
-  Universit\"at Innsbruck \\
-  Department of Atmospheric and Cryospheric Science \\
-  Faculty of Geo- and Atmospheric Sciences \\
-  Innrain~52f \\
-  6020 Innsbruck, Austria \\
-  E-mail: \email{Georg.Mayr@uibk.ac.at} \\
-  URL: \url{https://www.uibk.ac.at/acinn/people/georg-mayr}\\
-
-  Torsten Hothorn\\
-  Universit\"at Z\"urich \\
-  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
-  Hirschengraben 84\\
-  CH-8001 Z\"urich, Switzerland \\
-  E-mail: \email{Torsten.Hothorn@R-project.org}\\
-  URL: \url{http://user.math.uzh.ch/hothorn/}\\
-
-}
-
-\begin{document}
+\end{abstract}
+
+\keywords{Circular data; Regression trees; Random forests; Distributional regression; von~Mises distribution}
 
@@
   \label{tab:data}
-  \begin{center}
+  \centering
     \begin{tabular}{l  l}
@@
     \end{tabular}
-  \end{center}
 \end{table}
@@
     intrinsic automatic variable selection performed in the tree estimation. The
-    tree is built with the newly developed \textsf{R} package \pkg{circtree} employing the
+    tree is built with the newly developed \textsf{R} package \textbf{circtree} employing the
     CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} using a minimal number of
-    2000 observations in each terminal node (argument \code{minbucket}).
+    2000 observations in each terminal node (argument \texttt{minbucket}).
     %%
@@
     constructed based on $100$~individual trees employing the \textsf{R} package
-    \pkg{circtree}. Each of these trees is again built by the CTree algorithm on a
+    \textbf{circtree}. Each of these trees is again built by the CTree algorithm on a
     subsample containing $30$~percent of the original learning data. All
@@
     response variable is always considered for splitting. This bagging approach can
-    be applied in \pkg{circtree} by setting the argument \code{mtry} to the total
+    be applied in \textbf{circtree} by setting the argument \texttt{mtry} to the total
     number of covariates. Since a high number of possible split points leads to
     high computational costs, the covariates are binned in a maximum of $50$
-    classes (argument \code{nmax = c(yx = Inf, z = 50)}). Contrary to a
+    classes (argument \texttt{nmax = c(yx = Inf, z = 50)}). Contrary to a
     single-tree model, forests usually consist of very large trees as they are not
@@
 regression trees and forests is provided in the \textsf{R} package
-\textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The package is
+\textbf{circtree} (version~0.1.0). The package is
 based on the \textbf{disttree} package
-(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
+(version~0.2.0) which applies the main tree
 building functions from the \textbf{partykit} package
-(version~\Sexpr{packageVersion("partykit")}). All three packages are available
+(version~1.2.6). All three packages are available
 on \textsf{R}-Forge at {\url{https://R-Forge.R-project.org/projects/partykit/}}.
@@
 implementation is provided in the \textsf{R} package \textbf{circglmbayes} by
-\cite{Mulder+Klugkist:2017}. In particular the function \code{circGLM} is
+\cite{Mulder+Klugkist:2017}. In particular the function \texttt{circGLM} is
 applied to estimates the intercept and regression coefficient along with the
@@
 %\newpage
+\bibliographystyle{rss}
 \bibliography{ref.bib}

### 2020-04-10 pkg/circtree/inst/wind_paper/draft/draft_submitted_2020_01_09.tex
Update make diff for draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3007 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 735d1b2c68023e00f6d83f0584a7386adb699856

--- 2020-01-09 pkg/circtree/inst/wind_paper/draft/draft.Rnw
@@
 
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
-library("disttree")
-library("circtree")
-
-## HCL palette
-pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-# set function for parallelization
-applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))
-
-@
-
 \title{Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting}
@@
 regression trees and forests is provided in the \textsf{R} package
-\textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The package is
+\textbf{circtree} (version~0.1.0). The package is
 based on the \textbf{disttree} package
-(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
+(version~0.2.0) which applies the main tree
 building functions from the \textbf{partykit} package
-(version~\Sexpr{packageVersion("partykit")}). All three packages are available
+(version~1.2.7). All three packages are available
 on \textsf{R}-Forge at {\url{https://R-Forge.R-project.org/projects/partykit/}}.

### 2020-04-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include comments by reviewer (all w/o abstract) and write rebuttal
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3008 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 36023b3e82b13a8dd846c7059e17af522cc31db5

@@
 location and scale parameters conditional on the respective sets of covariates
-until convergence. While all of these models are built on well-elaborated
+until convergence \citep{Fisher+Lee:1992, Fisher:1993}. While all of these models are built on well-elaborated
 theory, their application in practice remains very challenging, mainly due to
@@
 distribution in each terminal node offering a wide range of statistical
-inference. In addition, the employed tree structure allows to capture
+inference. In addition, the employed tree structure can capture
 non-additive effects while forests enable the modeling of smooth changes.
@@
 This novel approach to circular regression trees and forests complements the
-literature on tree-based circular modeling. \cite{Lund:2002} already introduced
-a circular regression tree algorithm where binary splits are made based on an
+literature on tree-based circular modeling. \cite{Lund:2002} already introduced a
+circular regression tree algorithm where binary splits are made based on an
 angular distance measure capturing node homogeneity. However, this only models
-changes in the conditional mean but not the conditional variance or full
-probabilistic distribution which would allow for also considering uncertainty
-in forecasts \citep{Gneiting:2008}. In contrast, we introduce a distributional
-tree approach that considers splits in all distribution parameters and yields a
-full probabilistic predictive model. In addition, as a natural extension,
-ensembles or forests of circular regression trees are presented in this study,
-showing that these can further improve predictive power.
+changes in the conditional mean, whereas modeling the conditional variance or
+full probabilistic distribution would also allow uncertainty in the forecast to
+be estimated \citep{Gneiting:2008}. 
+%However, this only models
+%changes in the conditional mean but not the conditional variance or full
+%probabilistic distribution which would allow for also considering uncertainty
+%in forecasts \citep{Gneiting:2008}. 
+In the field of image based object direction estimation,
+\citet{Hara+Chellappa:2017} presented a distribution-based tree
+building algorithm by applying $k$-means clustering to the target space where
+the number of clusters $k$ is adaptively determined by the Bayes
+information criterion employing the von~Mises distribution. Despite their
+distribution-based splitting approach to circular response variables, they
+solely predict the mode of the distribution, thus, providing no full
+probabilistic forecasts. Combining these previous ideas for circular trees with the
+concept of distributional modeling, we introduce a fully distributional tree
+approach that considers both splits in all distribution parameters and yields a
+full probabilistic predictive model, as proposed by \citet{Gneiting:2008}. In
+addition, as a natural extension, ensembles or forests of circular regression
+trees are presented in this study, showing that these can further improve
+predictive power.
 
@@
 pressure gradients and wind speed separate the data into subgroups with high
-vs.\ low precision. A more extensive case study of circular regression trees
-and forests applied to probabilistic wind direction forecasting at Innsbruck
-Airport and Vienna International Airport is presented in
-Section~\ref{sec:wind}, along with a benchmark against commonly-used
-alternative approaches.
+vs.\ low precision. A more comprehensive application of circular regression
+trees and forests on probabilistic wind direction forecasting is presented in
+Section~\ref{sec:wind} for Innsbruck Airport and Vienna International Airport,
+together with a comparison with commonly-used alternative approaches.
 
@@
 methodology of circular regression trees and forests and their features are
-introduced in Section~\ref{sec:tree_forest}. After the case study presented in
+introduced in Section~\ref{sec:tree_forest}. After the application presented in
 Section~\ref{sec:wind}, a comprehensive summary and conclusions are given in
@@
 covariates $\mathbf{z}$ through a link function~$g(\cdot)$, circular intercept
-$\mu_0$ and coefficient vector $\mathbf{\beta}$: \begin{equation} \mu = \mu_0 +
-g\left(\mathbf{\beta}^{\top} \mathbf{z}\right). \end{equation} The link
+$\mu_0$ and coefficient vector $\boldsymbol{\beta}$: 
+%
+\begin{equation} \mu = \mu_0 +
+g\left(\boldsymbol{\beta}^{\top} \mathbf{z}\right). 
+\end{equation} 
+%
+The link
 function transforms the additive predictor to an interval of length $2\pi$.
@@
 share the previously discussed difficulties induced by the characteristics of
-the log-likelihood function and the strongly nonlinear link function. Referring
-to additive models in general, it has to be considered that a proper model
-specification can be very challenging, particularly for a high number of
+the log-likelihood function and the strongly nonlinear link function. In general,
+for additive models a proper model specification can be very challenging, particularly for a high number of
 covariates and no information on possible interactions. Moreover, the additive
@@
 \begin{equation}
-  w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
+  w^{\text{T}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
   \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)).
@@
   (\hat{\mu},\hat{\kappa})(\bm{z}) =
-  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot
+  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{T}}_i(\bm{z}) \cdot
   \ell(\mu,\kappa; y_i).
@@
 observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
-$w_i^\text{forest}(\bm{z})$ is obtained that is based on the number of trees in
+$w_i^\text{F}(\bm{z})$ is obtained that is based on the number of trees in
 which $\bm{z}$ is assigned to the same terminal node as each learning
@@
 \begin{equation}
-  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
+  w^{\text{F}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
   \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
@@
   (\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu,
-  \kappa} \sum_{i=1}^n w_i^\text{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
+  \kappa} \sum_{i=1}^n w_i^\text{F}(\bm{z}) \cdot \ell(\mu, \kappa; y_i). 
 \end{equation}
@@
 % -------------------------------------------------------------------
-% Case study
+% Application
 % -------------------------------------------------------------------
-\section{Case study: Probabilistic wind direction forecasting}
+\section{Application: Probabilistic wind direction forecasting}
 \label{sec:wind}
@@
 \label{sec:wind:data}
-The circular response variable considered in this case study is a
+The circular response variable considered in this application is a
 $10$\,min-average of wind direction measurements at Innsbruck Airport (INN) and
@@
 set consists of 41\,979~time points and $260$~covariates for Innsbruck, and of
-38\,985~time points and $494$~covariates for Vienna, respectively.
+38\,985~time points and $494$~covariates for Vienna.
 
@@
 \begin{table}[t!]
- \caption[Table caption text]{Overview of the data sets employed in the case
-    study: For Innsbruck and Vienna, various meteorological variables and derived
-    quantities of these are considered at the respective stations, located either
-    directly at the airports or in their vicinities.}
+ \caption[Table caption text]{Overview of the data sets employed in the
+    application on probabilistic wind direction forecasting: For Innsbruck and
+    Vienna, various meteorological variables and derived quantities of these are
+    considered at the respective stations, located either directly at the airports
+    or in their vicinities.}
   \label{tab:data}
@@
     sampling algorithm using weakly informative priors. \newline Following
-    \cite{Mulder+Klugkist:2017} the model uses a link function~$g(\cdot)$ to keep the
-    response values within an interval of length $2\pi$. As the implementation
-    cannot handle circular covariates, we use the components of the lagged
-    2-dimensional wind vector~$(u, v)^{\top}$ and the lagged
-    wind speed $spd$ as predictor variables. The model formula for the location
+    \cite{Fisher+Lee:1992} the model uses a link function~$g(\cdot)$ to keep the
+    response values within an interval of length $2\pi$. Since in the
+    implementation of \cite{Mulder+Klugkist:2017} the accommodation of circular covariates would require
+    an additional manual adjustment, such as an approximation by a finite Fourier expansion
+    (cf., \citealt{Lund:1999}), we use only the linear components of the lagged
+    2-dimensional wind vector~$(u, v)^{\top}$ and the lagged wind speed $spd$ as
+    predictor variables. Thus, the model formula for the location
     parameter $\mu$ of the von~Mises distribution can be written as:
@@
 \begin{equation} 
-  \text{CRPSS}_{\text{model}} = 1 -
-  \frac{\text{CRPS}_{\text{model}}}{\text{CRPS}_{\text{climatology}}}.
+  \text{CRPSS}(\text{model}) = 1 -
+  \frac{\text{CRPS}(\text{model})}{\text{CRPS}(\text{climatology})}.
 \end{equation}
@@
 The performance of the novel circular regression trees and forests is assessed
-in a case study for shortterm probabilistic wind direction forecasting at two
+in an application on shortterm probabilistic wind direction forecasting at two
 airports with different site characteristics. As benchmark models,
@@
 truncated normal model of \cite{Thorarinsdottir+Gneiting:2010} to capture wind
-speed; in the second step a circular wind direction model is leveraged given
+speed; in the second step a circular wind direction model is used given
 that a certain hurdle for the wind speed is crossed.
@@
 shifts, and potential interactions in covariates without prespecification of
-such effects. As supported by the presented case study, this can be
-particularly useful for modeling a highly fluctuating response, such as
-typically observed for wind direction, or/and in case of a large number of
-possible covariates. Moreover, the case study shows that building ensembles of
-circular regression trees can even improve the forecasting performance, as the
-resulting forests allow for modeling smooth effects and stabilize the model.
+such effects. As supported by the application on probabilistic wind direction
+forecasting, this can be particularly useful for modeling a highly fluctuating
+response, such as typically observed for wind direction, or/and in case of a
+large number of possible covariates. Moreover, the application shows that
+building ensembles of circular regression trees can even improve the
+forecasting performance, as the resulting forests allow for modeling smooth
+effects and stabilize the model.
 
@@
 data as applied for building circular regression trees and forests presented in
-this case study. An alternative tree-building framework is provided by the MOB
+this application. An alternative tree-building framework is provided by the MOB
 algorithm, which is based on M-fluctuation tests \citep[see][for more

### 2020-04-14 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Include comments by reviewer (all w/o abstract) and write rebuttal
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3008 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 36023b3e82b13a8dd846c7059e17af522cc31db5

@@
 
-\setkomavar{subject}{Revision of `npg-2019-49'}
+\setkomavar{subject}{Revision of `JRSS-SC-Jan-20-0007'}
 \setkomavar{signature}{Moritz Lang\\Corresponding Author}
@@
 
-%\begin{letter}[fromemail]{To Maxime Taillardat}
-%\opening{Dear Christopher Paciorek}
-%
-%We thank you, as well as both reviewers, for the detailed and fruitful feedback
-%regarding our manuscript ``\textbf{Bivariate Gaussian models for wind vectors in a
-%distributional regression framework}''.
-%
-%\vspace{5mm}
-%
-%We have carefully revised our manuscript according to your comments and
-%suggestions. The most substantial changes are the following:
-%
-%\begin{itemize}
-%\item
-%An additional comparison to the work of \citet{schuhen.etal:2012} and of
-%\citet{pinson:2012} has been included. Both comparisons nicely confirm our
-%previous results. To keep the manuscript as clear as possible, we describe the
-%comparisons in the appendix and provide short summaries of the main finding
-%in the discussion section.
-%
-%\item
-%The introduction of the logarithmic score has been
-%extended. We now explicitly derive the employed formulation of the logarithmic
-%score and provide additional references.
-%
-%\item
-%We have added more information on the estimation procedure and describe the key
-%steps involved in the model fitting. Please also remark that for consistency we
-%have changed `mean and variance` to `location and scale parameter` within the
-%abstract and the introductory part of the discussion.
-%\end{itemize}
-%
-%\vspace{5mm}
-%
-%On the following pages a point-to-point response to both reviewers and to the
-%editor will be given. The attached manuscript highlights the changes in the
-%text in blue color.
-%
-%\closing{Your sincerely,}
-%
-%\end{letter}
-%
-%\newpage
+\begin{letter}[fromemail]{To Prof. Peter Smith}
+\opening{Dear Prof. Peter Smith}
+
+We thank you, as well as the associate editor and both reviewers, for the
+positive and constructive feedback regarding our manuscript ``\textbf{Circular
+Regression Trees and Forests with an Application to Probabilistic Wind
+Direction Forecasting}''.
+
+\vspace{5mm}
+
+We have carefully revised our manuscript according to given comments.
+The most substantial changes in the manuscript are the following:
+
+\begin{itemize}
+\item We have fully revised the abstract in order to put a larger
+focus on the application. In addition, we now use the term `application' 
+instead of `case study' throughout the manuscript and have also renamed Sect.~4.
+
+\item In the section computational details, we have additionally listed the needed
+computing time of the different methods on a standard desktop computer.
+
+\item We now discuss the work of Hara and Chellappa~(2017) and compare their distribution-based 
+tree-building approach to our implementation of circular distributional trees within Sect.~1.1.
+
+\item For the employed GLM implemented by Mulder and Klugkist~(2017), we now
+explicitly state that the accommodation of circular covariates would be possible
+but require an additional manual adjustment, e.g., an approximation by a finite
+Fourier expansion (cf., Lund 1999).
+
+\end{itemize}
+
+\vspace{5mm}
+
+On the following pages a point-to-point response to all comments will be given.
+The attached manuscript highlights the changes in the text in blue color.
+
+\closing{Your sincerely,}
+
+\end{letter}
+
+\newpage
 
@@
 
-\begin{itemize}
-
-\item Please can you rewrite the Summary to give the application more prominence.
+\begin{re}
+Please can you rewrite the Summary to give the application more prominence.
 Currently, the last sentence reads to me like an afterthought. Furthermore, I
-would prefer you to use the term ‘application’ rather than ‘case study’
+would prefer you to use the term `application' rather than `case study'
 throughout the paper, since in Applied Statistics papers the methodological
 developments should be motivated by the application.
+\end{re}
 
-\item Page 2, line 46: ‘… structure can capture non-additive …’?
+\vspace{-1em}
+First of all we want to thank you for your valuable input.
 
-\item Page 2, lines 54 and 55: ‘… mean, whereas modelling the conditional … distribution would also allow uncertainty in the forecast to be estimated.’?
+We agree with your comment and have rewritten the abstract to put a larger
+focus on the application. Throughout the manuscript we now use the term
+`application' instead of `case study' and have also renamed Sect.~4
+accordingly. \todo{Lisa: Revise abstract}
 
-\item Page 6, line 16 and Equation (2.6): bold the vector $\beta$.
+\vspace{0.5em}
+\begin{re}
+Page 2, line 46: `… structure can capture non-additive …'?
+\end{re}
 
-\item Page 6, lines 29 and 30: ‘In general, for additive models a proper model …’?
+Thank you, we have rephrased the sentence according to your suggestion.
 
-\item Equations (3.1) to (3.4) and elsewhere: I find the use of multiple letters for a single superscript or subscript awkward and prefer to keep lower case letters for indices, so please consider replacing them, for example, T for tree and F for forest.
+\vspace{0.5em}
+\begin{re}
+Page 2, lines 54 and 55: `… mean, whereas modelling the conditional …
+distribution would also allow uncertainty in the forecast to be estimated.'?
+\end{re}
+
+We have rephrased the sentence accordingly.
+
+\vspace{0.5em}
+\begin{re}
+Page 6, line 16 and Equation (2.6): bold the vector $\beta$.
+\end{re}
+
+Thank you for pointing that out, we have corrected that accordingly.
+
+\vspace{0.5em}
+\begin{re}
+Page 6, lines 29 and 30: `In general, for additive models a proper model …'?
+\end{re}
+
+We have rephrased the sentence according to your suggestion.
+
+\vspace{0.5em}
+\begin{re}
+Equations (3.1) to (3.4) and elsewhere: I find the use of multiple letters for a single superscript or subscript awkward and prefer to keep lower case letters for indices, so please consider replacing them, for example, T for tree and F for forest.
 Also for Equation (4.2), how about using parentheses rather than subscripts?
+\end{re}
 
-\item Equation (3.4): bold vector $z$.
+We have followed your suggestions and now use only single letters for super-
+and subscripts in the main body of the manuscript. In the appendix, we keep the
+notation as it is in order to be consistent with the reference Hothorn~\emph{et al.}~2006.
 
-\item Page 9, line 39: delete ‘respectively’.
+\vspace{0.5em}
+\begin{re}
+Equation (3.4): bold vector $z$.
+\end{re}
 
-\item Page 15, line 23: replace ‘leveraged’ with ‘used’?
+Thank you for pointing that out.
 
-\end{itemize}
+\vspace{0.5em}
+\begin{re}
+Page 9, line 39: delete `respectively'.
+\end{re}
+
+We have corrected the sentence accordingly.
+
+\vspace{0.5em}
+\begin{re}
+Page 15, line 23: replace `leveraged' with `used'?
+\end{re}
+
+We have rephrased the sentence accordingly.
+
+\vspace{0.5cm}
+References
+
+Hothorn T, Hornik K, Zeileis A (2006). ``Unbiased Recursive Partitioning: A Conditional
+Inference Framework.'' \emph{Journal of Computational and Graphical Statistics}, \textbf{15}\,(3), 651--674.
+doi:10.1198/106186006x133933.
 
@@
 
+\begin{re}
 Both referees agree this is a good paper, and have only a few suggestions for
-changes, so I recommend minor revisions to address these.
+changes, so I recommend minor revisions to address these. I have two short suggestions in addition to this:
 
-I have two short suggestions in addition to this:
-
-\begin{itemize}
-\item Can you comment on the typical computational requirements of the various
+Can you comment on the typical computational requirements of the various
 methods used in your application, in terms of hardware required and length of
 computing time.
+\end{re}
 
-\item Is it possible to include a link to computer code to reproduce your
+First of all we want to thank you for your valuable input.
+
+We agree with your comment that additional information on the calculation
+requirements is valuable additional information for the user.  We have
+therefore listed the needed computing time of the different methods on a
+standard desktop computer within the section computational details.
+\todo{Moritz: Run models}
+
+\vspace{0.5em}
+\begin{re}
+Is it possible to include a link to computer code to reproduce your
 analysis? This is not essential for publication but would be helpful for
@@
 for your method is very helpful in this regard.  
-\end{itemize}
+\end{re}
+
+As we have used data from the Austrian air navigation service provider, which
+we are not entitled to distribute, we are unfortunately not able to provide
+self-contained computer code in order to reproduce our application. However, as you have already
+pointed out the implementation of circular regression trees and forests is
+provided in a freely available R-package with some minimal examples.
+Independent of this publication, we are planning to provide an online
+platform for the whole `partykit' project in the future.
 
@@
 
+\begin{re}
 I found Circular Regression Trees and Forests to be a very nicely written
@@
 for a circular response variable. 
+\end{re}
 
+\begin{re}
 I have very few suggestions to the authors; but here are two:
 
-\begin{itemize}
-\item On page 2, what is the reference for the phrase "Additionally, they
+On page 2, what is the reference for the phrase "Additionally, they
 developed a combined ... version ...".
+\end{re}
 
-\item On page 11, regarding "As the implementation cannot handle circular
+First of all we want to thank you for your review and your valuable input.
+
+Thanks for pointing out that the citation here is not clearly assigned. We have therefore
+listed the references below the statement again.
+
+\vspace{0.5em}
+\begin{re}
+On page 11, regarding "As the implementation cannot handle circular
 covariates ...": It could handle circular covariates in a fashion similar to
@@
 appropriate than stating that circular covariates cannot be accommodated.
-\end{itemize}
+\end{re}
+
+A very good remark. We have adapted the statement accordingly and now
+explicitly state that the accommodation of circular covariates would be
+possible but require an additional manual adjustment, e.g., an approximation by
+a finite Fourier expansion.
 
@@
 
+\begin{re}
 This is a well-written paper on using circular regression trees for analyzing a
@@
 thoroughly analyzed.
+\end{re}
 
-\begin{itemize}
-\item The authors could address model order determination problem using the
+\begin{re}
+The authors could address model order determination problem using the
 Bayes Information Criterion (Schwartz, 1978, Annals of Statistics).  
+\end{re}
 
-\item I would like to point out related works on circular regression models for
+First of all we want to thank you for your review and your valuable input.
+
+Regarding your suggestion, when building
+a circular distributional tree, the package \textbf{circtree} uses per default
+\emph{p}-values from association tests both for selecting the split variable
+and for a so-called pre-pruning strategy which stops growing the tree as soon as
+no significant associations can be found in a given subgroup. This is the
+default in both used unbiased recursive partitioning algorithms CTree
+and MOB. For the latter, as discussed by Zeileis~\emph{et al.}~(2008), an
+alternative, already implemented strategy  is to
+use information criteria such as AIC (Akaike information criterion) or BIC
+(Bayes information criterion) for post-pruning (after growing the initial tree). 
+As we use neither MOB no post-pruning in the
+presented application, we refrain from discussing this aspect any further in
+the manuscript. However, we want to refer to the submitted manuscript by
+Schlosser~\emph{et al.}~2019 which focuses on the advantages of different
+unbiased recursive partitioning algorithm for tree building as suggested in the literature.
+
+\vspace{0.5em}
+\begin{re}
+I would like to point out related works on circular regression models for
 shape representation (Kashyap and Chellappa) and circular random forests for
@@
 some comparisons with Hara, Chellappa work.
-\end{itemize}
+\end{re}
+
+Thank you very much for the very interesting and to us unfortunately so far unknown references.
+We now review the work of Hara and Chellappa~(2017) and briefly compare their approach
+to our implementation of circular distributional trees in Sect.~1.1.
 
@@
 
+Schlosser L, Hothorn T, Zeileis A (2019). ``The Power of Unbiased Recursive
+Partitioning: A Unifying View of CTree, MOB, and GUIDE.'' arXiv:1906.10179,
+arXiv.org E-Print Archive.
+
+Zeileis A, Hothorn T, Hornik K (2008). ``Model-Based Recursive Partitioning.''
+\emph{Journal of Computational and Graphical Statistics}, \textbf{17}\,(2),
+492--514. doi:10.1198/106186008x319331.
+
 \end{document}

### 2020-04-14 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Update rebuttal and draft (Lisa Input and new Abstract)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3009 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4bc02d09fa6af89983ac9884ec756193c0a9a91a

@@
 \begin{itemize}
-\item We have fully revised the abstract in order to put a larger
+\item We have fully revised the `Abstract' in order to put a larger
 focus on the application. In addition, we now use the term `application' 
@@
 
-\item In the section computational details, we have additionally listed the needed
+\item In the section `Computational details', we have additionally listed the needed
 computing time of the different methods on a standard desktop computer.
@@
 
-\item For the employed GLM implemented by Mulder and Klugkist~(2017), we now
+\item Regarding the employed GLM implemented by Mulder and Klugkist~(2017), we now
 explicitly state that the accommodation of circular covariates would be possible
-but require an additional manual adjustment, e.g., an approximation by a finite
+but would require an additional manual adjustment, e.g., an approximation by a finite
 Fourier expansion (cf., Lund 1999).
@@
 On the following pages a point-to-point response to all comments will be given.
-The attached manuscript highlights the changes in the text in blue color.
+The attached manuscript highlights the changes in the main body in blue color,
+the `Abstract' is fully revised but displayed in black color.
 
-\closing{Your sincerely,}
+\closing{Yours sincerely,}
 
@@
 
-We agree with your comment and have rewritten the abstract to put a larger
+We agree with your comment and have rewritten the `Abstract' to put a larger
 focus on the application. Throughout the manuscript we now use the term
 `application' instead of `case study' and have also renamed Sect.~4
-accordingly. \todo{Lisa: Revise abstract}
+accordingly.
 
@@
 
-Thank you for pointing that out.
+Thank you for pointing that out, the font style has been adapted.
 
@@
 
-First of all we want to thank you for your valuable input.
+We first want to thank you for the positive feedback and your valuable input.
 
 We agree with your comment that additional information on the calculation
-requirements is valuable additional information for the user.  We have
-therefore listed the needed computing time of the different methods on a
-standard desktop computer within the section computational details.
-\todo{Moritz: Run models}
+requirements is valuable for the user. We have therefore listed the needed
+computing time of the different methods on a standard desktop computer within
+the section 'Computational details'.  \todo{Moritz: Run models}
 
@@
 provided in a freely available R-package with some minimal examples.
-Independent of this publication, we are planning to provide an online
+Independent of this publication, we are planning to provide an updated online
 platform for the whole `partykit' project in the future.
@@
 
-First of all we want to thank you for your review and your valuable input.
+First of all we want to thank you for your positive and constructive feedback.
 
@@
 explicitly state that the accommodation of circular covariates would be
-possible but require an additional manual adjustment, e.g., an approximation by
+possible but would require an additional manual adjustment, e.g., an approximation by
 a finite Fourier expansion.
@@
 
-First of all we want to thank you for your review and your valuable input.
+First of all we want to thank you for your positive feedback and valuable input.
 
-Regarding your suggestion, when building
-a circular distributional tree, the package \textbf{circtree} uses per default
-\emph{p}-values from association tests both for selecting the split variable
-and for a so-called pre-pruning strategy which stops growing the tree as soon as
-no significant associations can be found in a given subgroup. This is the
-default in both used unbiased recursive partitioning algorithms CTree
-and MOB. For the latter, as discussed by Zeileis~\emph{et al.}~(2008), an
-alternative, already implemented strategy  is to
-use information criteria such as AIC (Akaike information criterion) or BIC
-(Bayes information criterion) for post-pruning (after growing the initial tree). 
-As we use neither MOB no post-pruning in the
-presented application, we refrain from discussing this aspect any further in
-the manuscript. However, we want to refer to the submitted manuscript by
-Schlosser~\emph{et al.}~2019 which focuses on the advantages of different
-unbiased recursive partitioning algorithm for tree building as suggested in the literature.
+Regarding your suggestion, when building a circular distributional tree, the
+package \textbf{circtree} uses per default \emph{p}-values from association
+tests both for selecting the split variable and for a so-called pre-pruning
+strategy which stops growing the tree as soon as no significant associations
+can be found in a given subgroup. This is the default in both available
+unbiased recursive partitioning algorithms CTree and MOB. For the latter, as
+discussed by Zeileis~\emph{et al.}~(2008), an alternative, already implemented
+strategy  is to use information criteria such as AIC (Akaike information
+criterion) or BIC (Bayes information criterion) for post-pruning (after growing
+the initial tree).  As we use neither MOB nor post-pruning in the presented
+application, we refrain from discussing this aspect any further in the
+manuscript. However, we want to refer to the submitted manuscript by
+Schlosser~\emph{et al.}~2019 which investigates the performance of different 
+testing and pruning strategies employed in unbiased recursive partitioning algorithms.
 

### 2020-04-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update rebuttal and draft (Lisa Input and new Abstract)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3009 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4bc02d09fa6af89983ac9884ec756193c0a9a91a

@@
 \Abstract{
-While circular data occur in a wide range of scientific fields, the methodology
+%SUBMITTED VERSION:
+%While circular data occur in a wide range of scientific fields, the methodology
+%for distributional modeling and probabilistic forecasting of circular response
+%variables is rather limited. Most of the existing methods are built on the
+%framework of generalized linear and additive models, which are often
+%challenging to optimize and interpret. Therefore, building on previous ideas
+%for trees modeling circular means, we suggest a distributional approach for
+%regression trees and random forests yielding probabilistic forecasts based on
+%the von~Mises distribution. The resulting tree-based models simplify the
+%estimation process by using the available covariates for partitioning the data
+%into sufficiently homogeneous subgroups so that a simple von~Mises distribution
+%without further covariates can be fitted to the circular response in each
+%subgroup. These circular regression trees are straightforward to interpret, can
+%capture nonlinear effects and interactions, and automatically select the
+%relevant covariates that are associated with either location and/or scale
+%changes in the von~Mises distribution. Combining an ensemble of circular
+%regression trees to a circular regression forest can regularize and smooth the
+%covariate effects. The new methods are evaluated in a case study on
+%probabilistic wind direction forecasting at two Austrian airports, considering
+%other common approaches as a benchmark.
+%
+%NEW VERSION:
+Circular data occur in a wide range of scientific fields, but the methodology
 for distributional modeling and probabilistic forecasting of circular response
-variables is rather limited. Most of the existing methods are built on the
-framework of generalized linear and additive models, which are often
-challenging to optimize and interpret. Therefore, building on previous ideas
-for trees modeling circular means, we suggest a distributional approach for
-regression trees and random forests yielding probabilistic forecasts based on
-the von~Mises distribution. The resulting tree-based models simplify the
-estimation process by using the available covariates for partitioning the data
-into sufficiently homogeneous subgroups so that a simple von~Mises distribution
-without further covariates can be fitted to the circular response in each
-subgroup. These circular regression trees are straightforward to interpret, can
-capture nonlinear effects and interactions, and automatically select the
-relevant covariates that are associated with either location and/or scale
-changes in the von~Mises distribution. Combining an ensemble of circular
-regression trees to a circular regression forest can regularize and smooth the
-covariate effects. The new methods are evaluated in a case study on
-probabilistic wind direction forecasting at two Austrian airports, considering
-other common approaches as a benchmark.
-}
+variables is rather limited. While most of the existing methods are built on
+the framework of generalized linear and additive models, which are often
+challenging to optimize and interpret, we propose a distributional approach for
+regression trees yielding probabilistic forecasts based on the von~Mises
+distribution. The resulting tree-based models use the available covariates only
+for partitioning the data into sufficiently homogeneous subgroups, allowing for
+a simple circular distributional fit in each subgroup. These circular
+regression trees are straightforward to interpret, can capture nonlinear
+effects and interactions, and automatically select the relevant covariates that
+are associated with changes in any distribution parameter. Ensembles or
+forests of such circular trees can further improve the forecasts by
+regularizing and stabilizing the model. The newly developed methodology is
+illustrated in an application on short-term probabilistic wind direction
+forecasting for Innsbruck and Vienna Airport in order to direct airplanes to a
+safe landing. The predictive skill of the novel approaches is benchmarked with
+an additive regression model plus a persistency and a climatology model,
+employing the circular continuous ranked probability score.}
 
@@
 In the field of image based object direction estimation,
-\citet{Hara+Chellappa:2017} presented a distribution-based tree
-building algorithm by applying $k$-means clustering to the target space where
+\citet{Hara+Chellappa:2017} presented a distribution-based tree-building 
+algorithm by applying $k$-means clustering to the target space where
 the number of clusters $k$ is adaptively determined by the Bayes
@@
 concept of distributional modeling, we introduce a fully distributional tree
-approach that considers both splits in all distribution parameters and yields a
-full probabilistic predictive model, as proposed by \citet{Gneiting:2008}. In
+approach that not only considers splits in all distribution parameters, but also yields a
+full probabilistic predictive model as proposed by \citet{Gneiting:2008}. In
 addition, as a natural extension, ensembles or forests of circular regression
@@
 Section~\ref{sec:wind} for Innsbruck Airport and Vienna International Airport,
-together with a comparison with commonly-used alternative approaches.
+together with a comparison to commonly-used alternative approaches.
 
@@
 one in flat terrain, the other one in mountainous terrain. The results are
-benchmarked against alternative probabilistic forecasting methods. The study is
+benchmarked against alternative probabilistic forecasting methods. The application is
 based on $+1$\,h and $+3$\,h forecasts employing lagged observations in the
@@
 Figure~\ref{fig:overview} provides a topographical overview of the airports and
-their surrounding areas with the station sites employed in this study. In
+their surrounding areas with the station sites employed in the application. In
 addition, we use derived quantities such as 3-hourly means, minima and maxima,
@@
 
-The data used in this study consists of five subsequent years from January~2014
+The data used in the application consists of five subsequent years from January~2014
 to December~2018. After first eliminating predictor variables with more than
@@
 investigate whether they can be applied as a reasonable alternative to already
-existing approaches, three additional statistical models are employed in this
-study for probabilistic forecasting of wind directions. Two of them are based
+existing approaches, three additional statistical models are employed in the
+application on probabilistic wind direction forecasting. Two of them are based
 on existing approaches used in the meteorological field, while the third is a
@@
 good performance of the persistency model as observed for shortterm wind
-direction forecasts in this study. While in these cases the trees and forests
+direction forecasts in the application. While in these cases the trees and forests
 also benefit from the highly informative lagged response, they can compensate
@@
 based on the \textbf{disttree} package
-(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
-building functions from the \textbf{partykit} package
+(version~\Sexpr{packageVersion("disttree")}) which applies the main tree-building 
+functions from the \textbf{partykit} package
 (version~\Sexpr{packageVersion("partykit")}). All three packages are available
@@
 data as applied for building circular regression trees and forests presented in
-this application. An alternative tree-building framework is provided by the MOB
+this study. An alternative tree-building framework is provided by the MOB
 algorithm, which is based on M-fluctuation tests \citep[see][for more

### 2020-04-16 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update computational details
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3010 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fd4ab622ddc4a10971a21c6bffe15b344f0c3bd7

@@
 
-For the circular GLM considered as reference model the corresponding
+For the circular generalized linear model considered as reference model the corresponding
 implementation is provided in the \textsf{R} package \textbf{circglmbayes} by
@@
 concentration parameter.
+
+The computation time on a standard computer is in the order of 10\,milliseconds
+for the circular fit (used for the persistency and the climatological model),
+seconds for the estimation of the linear model and the circular tree, and from
+a few minutes to an hour for the estimation of the circular forest; the latter
+strongly depends on the number of covariates employed and, hence, varies
+between Innsbruck and Vienna Airport. However, for a single station and
+forecast step, the persistency and linear models must be re-estimated for each
+time of interest, and independent climatological models have to be fitted for
+all times of interest within a single calendar year, whereas for all times of interest only one 
+circular tree and circular forest model is required.
+
+%The computation time on a standard computer is in the order of 10\,milliseconds
+%for the circular fit (used for the persistency and the climatological model),
+%seconds for the estimation of the linear model and the circular tree, and from
+%a few minutes to an hour for the estimation of the circular forest; the latter
+%strongly depends on the number of covariates employed and, hence, varies
+%between Innsbruck and Vienna Airport. However, the persistency and linear
+%models must be re-estimated for each time point in the validation data set, and
+%independent climatological models have to be estimated for all the time points
+%within one calendar year which exist in the validation period, whereas only one
+%model is needed for the circular tree and the circular forest valid for all
+%times points in the validation data set.
 

### 2020-04-26 pkg/disttree/inst/PhD_thesis_defense/slides_169.Rnw
slides for PhD thesis defense
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3011 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2ac42af760b377bd2d56a4a05978e1eca5f1f0e9

\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
\usetheme[nototalframenumber, url,license]{uibk}

\title{Distributional Trees and Forests}
\author{\small{Lisa Schlosser\\
\vspace{0.4cm}
Supervisors: Achim Zeileis, Torsten~Hothorn, Michael Oberguggenberger\\}
\vspace{0.4cm}
\footnotesize{PhD thesis defense, 2020-04-24}
%Committee: Achim Zeileis, Michael Oberguggenberger, Georg Moser
}
%\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
%\URL{http://www.partykit.org/partykit/}
\renewcommand{\headerimage}[1]{%
   \IfStrEqCase{#1}{%
      {1}{%
         \gdef\myheaderimageid{#1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{forest.jpg}%
      }}[%
         \gdef\myheaderimageid{1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{forest.jpg}%
      ]%
}
\headerimage{1}

\graphicspath{{Figures/}}
\usepackage[utf8]{inputenc}
\usepackage[export]{adjustbox}

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
% \usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usepackage{calc}
\usepackage{graphicx}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
\usepackage{xcolor}
%\usepackage[cal=boondoxo]{mathalfa}
%\graphicspath{{plots/}}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
\definecolor{forestred}{rgb}{0.8078431, 0.2862745, 0.3176471}
\definecolor{treegreen}{rgb}{0.0000000, 0.5607843, 0.0000000}
\definecolor{lightblue}{RGB}{34,151,230}
\definecolor{lightorange}{RGB}{255,165,0}

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)

set.seed(7)

invisible(.Call(grDevices:::C_palette, grDevices::hcl(
  h = c(0,   5, 125, 245, 195, 315,  65,   0),
  c = c(0, 100,  90,  85,  63, 105,  90,   0),
  l = c(0,  55,  75,  60,  82,  48,  80,  65)
)))

# setwd("~/svn/partykit/pkg/disttree/inst/defensio/")
library("disttree")
library("Formula")
library("latex2exp")
library("gamlss")
library("lattice")
library("crch")
library("RainTyrol")
library("parallel")
library("gamlss.cens")
gen.cens(NO, type = "left")
library("ggplot2")
theme_set(theme_bw(base_size = 18))
library(colorspace)
## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)

lightblue <- "#2297E6"
lightorange <- "#FFA500"
@

\begin{document}

\section{Distributional Trees and Forests}

\subsection{Statistical modeling}

\begin{frame}[fragile]
\frametitle{Statistical modeling}
\textbf{Assumption:}
\vspace{-0.543cm}
\begin{center}
\begin{tikzpicture}
\node[] at (0,0.4) {X};
\draw[->] (0.4,0.4)--(0.9,0.4);
\draw[] (1,0) rectangle (2.5,0.8);
\node[font=\scriptsize] at (1.75,0.4) {nature};
\draw[->] (2.6,0.4)--(3.1,0.4);

\node[] at (3.5,0.4) {Y};

\end{tikzpicture}
\end{center}
\vspace{0.3cm}
\textbf{Goal:} Approximation of reality based on a set of observations.\\
\pause

\vspace{0.5cm}
\textbf{Motivation:}
\begin{itemize}
\item Prediction.
\item Capturing dependencies and interactions.
\end{itemize}
\pause

\vspace{0.5cm}
\textbf{Two cultures:} Breiman (2001, \emph{Statistical Science})
\begin{itemize}
\item Data modeling.
\item Algorithmic modeling.
\end{itemize}
\end{frame}

\subsubsection{Data modeling}

\begin{frame}[fragile]
\frametitle{Data modeling}
\vspace{-0.1cm}
\begin{center}
\begin{tikzpicture}
\node[] at (0,0.4) {X};
\draw[->] (0.4,0.4)--(0.9,0.4);
\draw[] (1,0) rectangle (2.5,0.8);
\node[font=\scriptsize] at (1.75,0.4) {f(X; par)};
\draw[->] (2.6,0.4)--(3.1,0.4);
\node[] at (3.5,0.4) {Y};
\end{tikzpicture}
\end{center}
\vspace{-0.4cm}
\only<1->{
\begin{itemize}
\item Approximate data generating process up to a few unknown processes.
\item Mathematical equations, model assumptions.
\item Data: estimation of parameters.
\end{itemize}
}
\only<1>{\vspace{1cm}}

\vspace{0.1cm}

\visible<2->{
\begin{minipage}{0.65\textwidth}
\textbf{Example:} 
\only<2-3>{Linear model}\only<4>{Generalized additive model}\only<5>{Generalized additive model\\ for location, scale, and shape}
\\
\only<2-3>{
$$\visible<3>{\mathbb{E}(Y|X) = } f(X; \alpha) = \alpha_0 + \alpha_1 \cdot X$$  
\vspace{0.15cm}
}
\only<4>{
$$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
\vspace{0.15cm}
}
\only<5>{
\vspace{-1.29cm}

\begin{align*}
\visible<5>{\mathbb{E}(Y|X) = }\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
\visible<5>{\sqrt{Var(Y|X)} = }\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
\end{align*}
}
\end{minipage}
%\vspace{-0.6cm}
%\hfill
\begin{minipage}{0.3\textwidth}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@
\only<2-3>{
<<plot_motivation_GLM_single, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(1.74*x)^2)/2.5
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\only<4>{
<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(1.74*x)^2)/2.5
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\only<5>{
<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
        col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
}
\end{minipage}
}
\end{frame}

<<motivation_tree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- numeric(length = length(x))
for(i in 1:(nobs)) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
# more points for precise illustration of true function
x100 <- c(1:(100*nobs))/(100*nobs)
ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- yforest <- numeric(length = length(x100))
for(i in 1:(nobs*100)) ytrue[i] <- if(x100[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x100[i]-0.2)-1)))
for(i in 1:(nobs*100)) ytree[i] <- if(x100[i]<1/3) 0.5 else {if(x100[i]<2/3) 2 else 1}
for(i in 1:(nobs*100)) ytree2[i] <- if(x100[i]<0.31) 0.47 else {if(x100[i]<0.69) 1.9 else 1.1}
for(i in 1:(nobs*100)) ytree3[i] <- if(x100[i]<0.34) 0.52 else {if(x100[i]<0.54) 2.2 else {if(x100[i]<0.74) 1.74 else 0.9}}
for(i in 1:(nobs*100)) ytree4[i] <- if(x100[i]<0.32) 0.51 else {if(x100[i]<0.63) 1.8 else 1.15}
for(i in 1:(nobs*100)) yforest[i] <- if(x100[i]<=0.33) 0.52 else {if(x100[i]<=0.34) 1.5 else {if(x100[i]<=0.54) 2 else {if(x100[i]<=0.6) 1.9 else {if(x100[i]<=0.64) 1.7 else {if(x100[i]<=0.7) 1.4 else {if(x100[i]<=0.74) 1.2 else 1}}}}}}
@

\subsubsection{Algorithmic modeling}

\begin{frame}[fragile]
\frametitle{Algorithmic modeling}
\vspace{-0.1cm}
\begin{center}
\begin{tikzpicture}
\node[] at (0,0.4) {X};
\draw[->] (0.4,0.4)--(0.9,0.4);
\draw[] (1,0) rectangle (2.5,0.8);
\node[font=\scriptsize] at (1.75,0.4) {nature};
\draw[->] (2.6,0.4)--(3.1,0.4);
\node[] at (3.5,0.4) {Y};
\draw[<->] (0.1,0.1) .. controls (0.2, -0.4) and (0.4, -0.5) .. (0.8,-0.5);
\draw[<->] (2.7,-0.5) .. controls (3.1, -0.5) and (3.3, -0.4) .. (3.4,0.1);
\node[font=\scriptsize] at (1.75,-0.5) {algorithm};
\end{tikzpicture}
\end{center}
\vspace{-0.4cm}
\visible<2->{
\begin{itemize}
\item Flexible approximation with no/few assumptions. %\\
%$\Rightarrow$ no fixed $f$.
\item Data-driven model fitting.
\end{itemize}
}
\vspace{0.25cm}
\visible<3->{
\begin{minipage}{0.5425\textwidth}
\textbf{Example:} \only<3>{Tree model}\only<4-10>{Subsampling}\only<11->{Forest model}
%\vspace{-0.5cm}
%\vspace{-0.8cm}
\begin{center}
\only<3>{
%\vspace{0.2cm}
\resizebox{0.45\textwidth}{!}{
\begin{tikzpicture}
  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (2, 3) {};
  \node[] (n00) at (1, 1.5) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0, 0) {$\hat{Y}_1$};
  \draw[-] (n0) -- (n00) node [midway, left] {\small $X \leq p_1$};
  \draw[-] (n0) -- (n1);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (3, 1.5) {};
  \draw[-] (n0) -- (n2) node [midway, right] {\small $X > p_1$};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (2, 0) {$\hat{Y}_2$};
  \draw[-] (n2) -- (n3) node [midway, left] {\small $X \leq p_2$};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (4, 0) {$\hat{Y}_3$};
  \draw[-] (n2) -- (n4) node [midway, right] {\small $X > p_2$};
\end{tikzpicture}}
%\vspace{0.445cm}
}}
\only<4->{
%\vspace{0.2cm}
\resizebox{0.347\textwidth}{!}{
\begin{tikzpicture}
%\draw[ellipse, draw=black, align=center, scale = 0.7, minimum width=170pt, minimum height = 35pt] (set) at (1, 3) {};
\draw[gray,rounded corners=10pt] (-1.05,4.35) -- (4.05,4.35) -- (4.05,2.7) -- (-2.05,2.7) -- (-2.05,4.35) -- (0.05,4.35);
%\draw[gray] (1,3.5) ellipse (3 and 0.8);
%\visible<4>{
\begin{scope}
\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\clip (1,3.5) ellipse (3 and 0.8);
\pgfmathsetseed{7}
\foreach \p in {1,...,200} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
\end{scope}
%}
%%% color points
%\visible<5-6>{
%\begin{scope}
%\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\pgfmathsetseed{7}
%\foreach \p in {1,...,100} {\fill[treegreen] (1+3*rand,3.5+0.8*rand) circle (0.07);}
%\end{scope}
%}
\visible<5->{
\node[ellipse, draw=treegreen, align=center, minimum width=80pt, minimum height = 40pt, line width = 3pt] (subset1) at (0.1, 3.5) {};
}
%\visible<7-8>{
%\begin{scope}
%\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\pgfmathsetseed{7}
%\foreach \p in {1,...,49} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
%\foreach \p in {50,...,150} {\fill[lightblue] (1+3*rand,3.5+0.8*rand) circle (0.07);}
%\end{scope}
%}
\visible<7->{
\node[ellipse, draw = lightblue, align=center, minimum width=100pt, minimum height = 30pt, line width = 3pt] (subset2) at (1.2, 3.6) {};
}
%\visible<9-10>{
%\begin{scope}
%\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\pgfmathsetseed{7}
%\foreach \p in {1,...,99} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
%\foreach \p in {100,...,200} {\fill[lightorange] (1+3*rand,3.5+0.8*rand) circle (0.07);}
%\end{scope}
%}
\visible<9->{
\node[ellipse, draw = lightorange, align=center, minimum width=120pt, minimum height = 25pt, line width = 3pt] (subset3) at (1.6, 3.3) {};
}
\visible<6->{
\node[ellipse, fill=treegreen, align=center] (n00) at (-1, 2) {};
\node[ellipse, fill=treegreen, align=center] (n01) at (-1.25, 1.25) {};
\draw[-, line width=1pt] (n00) -- (n01);
\node[rectangle, fill=treegreen, align=center] (n02) at (-1.5, 0.5) {};
\draw[-, line width=1pt] (n01) -- (n02);
\node[rectangle, fill=treegreen, align=center] (n03) at (-1, 0.5) {};
\draw[-, line width=1pt] (n01) -- (n03);
\node[rectangle, fill=treegreen, align=center] (n04) at (-0.5, 0.5) {};
\draw[-, line width=1pt] (n00) -- (n04);
}
\visible<8->{
\node[ellipse, fill=lightblue, align=center] (n10) at (1, 2) {};
\node[ellipse, fill=lightblue, align=center] (n11) at (0.5, 1.25) {};
\draw[-, line width=1pt] (n10) -- (n11);
\node[rectangle, fill=lightblue, align=center] (n12) at (0.25, 0.5) {};
\draw[-, line width=1pt] (n11) -- (n12);
\node[rectangle, fill=lightblue, align=center] (n13) at (0.75, 0.5) {};
\draw[-, line width=1pt] (n11) -- (n13);
\node[ellipse, fill=lightblue, align=center] (n14) at (1.5, 1.25) {};
\draw[-, line width=1pt] (n10) -- (n14);
\node[rectangle, fill=lightblue, align=center] (n15) at (1.75, 0.5) {};
\draw[-, line width=1pt] (n14) -- (n15);
\node[rectangle, fill=lightblue, align=center] (n16) at (1.25, 0.5) {};
\draw[-, line width=1pt] (n14) -- (n16);

}
\visible<10->{
\node[ellipse, fill=lightorange, align=center] (n20) at (3, 2) {};
\node[rectangle, fill=lightorange, align=center] (n21) at (2.5, 0.5) {};
\draw[-, line width=1pt] (n20) -- (n21);
\node[ellipse, fill=lightorange, align=center] (n22) at (3.25, 1.25) {};
\draw[-, line width=1pt] (n20) -- (n22);
\node[rectangle, fill=lightorange, align=center] (n23) at (3, 0.5) {};
\draw[-, line width=1pt] (n22) -- (n23);
\node[rectangle, fill=lightorange, align=center] (n24) at (3.5, 0.5) {};
\draw[-, line width=1pt] (n22) -- (n24);
}
\visible<11->{
\draw[-, line width=1pt] (-0.5, -0.1) -- (0.2,-0.5);
\draw[-, line width=1pt] (1, -0.1) -- (1,-0.5);
\draw[-, line width=1pt] (2.5, -0.1) -- (1.8,-0.5);
\node[rectangle, align=center] (combine) at (1, -0.9) {\Large average};
}
\end{tikzpicture}
}
}
\end{center}
\end{minipage}
\hspace{0.7cm}
\begin{minipage}{0.3\textwidth}
%\begin{center}
%\vspace{0.1cm}
\only<3-5>{
%\vspace{0.01cm}
<<plot_motivation_tree, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(2,0,2,0))
#par(mar=c(3,3,2,5.5))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x100, y = ytree, col = pal["forest"], lwd=7)
@
% mtext("rain", side = 4, las = TRUE, at = 16, line = 0.5, cex = 2.5, col = pal["forest"])
% mtext("now", side = 4, las = TRUE, at = 14.7, line = 0.5, cex = 2.5, col = pal["forest"])
% mtext("no", side = 4, las = TRUE, at = 10.3, line = 0.5, cex = 2.5, col = pal["forest"])
% mtext("rain", side = 4, las = TRUE, at = 9, line = 0.5, cex = 2.5, col = pal["forest"])
% mtext("now", side = 4, las = TRUE, at = 7.7, line = 0.5, cex = 2.5, col = pal["forest"])
% mtext("rain 24h", side = 2, las = FALSE, line = 1, cex = 3)
% mtext("humidity", side = 1, line = 1.5, cex = 3)
}
\only<6-7>{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest1, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(2,0,2,0))
#par(mar=c(3,3,2,5.5))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x100, y = ytree, col = pal["forest"], lwd=7)
lines(x = x100, y = ytree2, col = pal["tree"], lwd=7)
@
}
\only<8-9>{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest2, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(2,0,2,0))
#par(mar=c(3,3,2,5.5))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x100, y = ytree, col = pal["forest"], lwd=7)
lines(x = x100, y = ytree2, col = pal["tree"], lwd=7)
lines(x = x100, y = ytree3, col = lightblue, lwd=7)
@
}
\only<10>{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest3, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(2,0,2,0))
#par(mar=c(3,3,2,5.5))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x100, y = ytree, col = pal["forest"], lwd=7)
lines(x = x100, y = ytree2, col = pal["tree"], lwd=7)
lines(x = x100, y = ytree3, col = lightblue, lwd=7)
lines(x = x100, y = ytree4, col = lightorange, lwd=7)
@
}
\only<11->{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest4, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x100, y = yforest, col = pal["forest"], lwd=7)
@
}
%\end{center}
\end{minipage}
\end{frame}

% \subsection{Motivation}
% 
% \begin{frame}%[fragile]
% \frametitle{Motivation}
% \vspace{-0.41cm}
% \begin{figure}[!htb]
% \minipage{0.285\textwidth}
% \begin{center}
% <<motivation_GLM, echo=FALSE, results=hide>>=
% nobs <- 200
% ## GLM
% set.seed(7)
% x <- c(1:nobs)/nobs
% ytrue <- 1+x*1.5
% y <- ytrue + rnorm(nobs,0,0.3)
% @
% \visible<2->{
% <<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
% par(mar=c(2,0,2,0))
% plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
% box(lwd=5)
% lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
% @
% }
% \end{center}
% \endminipage
% \visible<3->{{\LARGE$\rightarrow$}}
% \minipage{0.285\textwidth}
% \begin{center}
% <<motivation_GAM, echo=FALSE, results=hide>>=
% ## GAM
% set.seed(7)
% x <- c(1:nobs)/nobs
% x <- 2*(x-0.5)
% ytrue <- x^3 
% y <- ytrue + rnorm(nobs,0,0.3)
% @
% \visible<3->{
% <<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
% par(mar=c(2,0,2,0))
% plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
% box(lwd=5)
% lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
% @
% }
% \end{center}
% \endminipage
% \visible<4->{{\LARGE$\rightarrow$}}
% \minipage{0.285\textwidth}
% \begin{center}
% <<motivation_GAMLSS, echo=FALSE, results=hide>>=
% ## GAMLSS
% set.seed(7)
% x <- c(1:nobs)/nobs
% x <- 2*(x-0.5)
% ytrue <- x^3
% var <- exp(-(2*x)^2)/2
% y <- ytrue + rnorm(nobs, 0, 0.1 + var)
% @
% \visible<4->{
% <<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
% par(mar=c(2,0,2,0))
% plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
% polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
%   col = pallight["forest"], border = "transparent")
% lines(x, ytrue, col = pal["forest"], lwd=7)
% points(x, y, col = "slategray", pch = 19)
% box(lwd = 5)
% @
% }
% \end{center}
% \endminipage
% 
% \vspace{0.5cm}
% \minipage{0.25\textwidth}
% \begin{center}
% \visible<2->{
% LM, GLM\\
% \vspace{0.5cm}
% \code{lm}\\
% \code{glm}\\
% \vspace{1.5cm}}
% \end{center}
% \endminipage
% \hspace{1.1cm}
% \minipage{0.25\textwidth}
% \begin{center}
% \visible<3->{
% GAM\\
% \vspace{0.5cm}
% \code{mgcv}\\
% \code{VGAM}\\
% \vspace{1.5cm}}
% \end{center}
% \endminipage
% \hspace{1.1cm}
% \minipage{0.25\textwidth}
% \begin{center}
% \visible<4->{
% GAMLSS\\
% \vspace{0.5cm}
% \code{gamlss}\\
% \code{mgcv}\\
% \code{VGAM}\\
% \code{gamboostLSS}\\
% \code{bamlss}}
% \end{center}
% \endminipage
% \end{figure}
% \end{frame}
% 
% 
% 
% 
% \begin{frame}[fragile]
% \frametitle{Motivation}
% \vspace{-0.41cm}
% \begin{figure}[!htb]
% \minipage{0.285\textwidth}
% \begin{center}
% <<motivation_regtree, echo=FALSE, results=hide>>=
% ## Reg. Tree
% set.seed(7)
% kappa <- 12
% x <- c(1:nobs)/nobs
% ytrue <- ytree <- yforest <- numeric(length = length(x))
% for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
% y <- ytrue + rnorm(nobs,0,0.3)
% for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
% @
% \visible<1->{
% <<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
% par(mar=c(2,0,2,0))
% plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
% box(lwd=5)
% #lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
% lines(x = x, y = ytree, col = pal["forest"], lwd=7)
% @
% }
% \end{center}
% 
% \endminipage
% \visible<2->{{\LARGE$\rightarrow$}}
% \minipage{0.285\textwidth}
%  \begin{center}
% <<motivation_randforest, echo=FALSE, results=hide>>=
% ## Random Forest
% for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
% @
% \visible<2->{
% <<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
% par(mar=c(2,0,2,0))
% plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
% box(lwd=5)
% #lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
% lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
% @
% }
% \end{center}
% 
% \endminipage
% \visible<3->{{\LARGE$\rightarrow$}}
% \minipage{0.285\textwidth}
% \begin{center}
% \visible<3->{
% <<plot_motivation_question, fig=TRUE, echo=FALSE>>=
% par(mar=c(2,0,2,0))
% plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
% box(lwd=5)
% text(x = mean(range(x)), y = mean(range(y)), "?", cex = 12)
% @
% }
% \end{center}
% 
% \endminipage
% 
% \minipage{0.285\textwidth}
% \begin{center}
% \vspace{0.0cm}
% \visible<1->{
% Regression tree\\
% \vspace{0.3cm}
% \resizebox{0.2\textwidth}{!}{
% \begin{tikzpicture}
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (1, 2) {};
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0.5, 1) {};
%   \draw[-, line width=1pt] (n0) -- (n1);
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (1.5, 1) {};
%   \draw[-, line width=1pt] (n0) -- (n2);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (1, 0) {};
%   \draw[-, line width=1pt] (n2) -- (n3);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (2, 0) {};
%   \draw[-, line width=1pt] (n2) -- (n4);
% \end{tikzpicture}}
% \vspace{0.2cm}\\
% \code{rpart}\\
% \code{party(kit)}\\
% \vspace{1cm}}
% \end{center}
% \endminipage
% \hspace{0.65cm}
% \minipage{0.285\textwidth}
% \begin{center}
% \vspace{-0.4cm}
% \visible<2->{
% Random forest\\
% \vspace{0.4cm}
% \resizebox{0.6\textwidth}{!}{
% \begin{tikzpicture}
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n00) at (1, 2) {};
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n01) at (0.5, 1) {};
%   \draw[-, line width=1pt] (n00) -- (n01);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n02) at (1.5, 1) {};
%   \draw[-, line width=1pt] (n00) -- (n02);
%   
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n10) at (3, 2) {};
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n11) at (2.5, 1) {};
%   \draw[-, line width=1pt] (n10) -- (n11);
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n12) at (3.5, 1) {};
%   \draw[-, line width=1pt] (n10) -- (n12);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n13) at (2, 0) {};
%   \draw[-, line width=1pt] (n11) -- (n13);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n14) at (2.8, 0) {};
%   \draw[-, line width=1pt] (n11) -- (n14);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n15) at (3.2, 0) {};
%   \draw[-, line width=1pt] (n12) -- (n15);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n16) at (4, 0) {};
%   \draw[-, line width=1pt] (n12) -- (n16);
%   
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
%   \draw[-, line width=1pt] (n20) -- (n21);
%   \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
%   \draw[-, line width=1pt] (n20) -- (n22);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
%   \draw[-, line width=1pt] (n22) -- (n23);
%   \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
%   \draw[-, line width=1pt] (n22) -- (n24);
% \end{tikzpicture}
% }
% \vspace{0.3cm}\\
% \code{randomForest}\\
% \code{ranger}\\
% \code{party(kit)}}
% \end{center}
% \endminipage
% \hspace{0.65cm}
% \minipage{0.285\textwidth}
% \begin{center}
% \vspace{0.15cm}
% \visible<3->{
% Distributional trees and forests\\
% \vspace{1.15cm}
% \code{disttree}\\
% based on \code{partykit}\\
% \vspace{1cm}}
% \end{center}
% \endminipage
% \end{figure}
% \end{frame}

\begin{frame}
\frametitle{Building blocks}
%\vspace{0.5cm}

\begin{minipage}{0.79\textwidth}
\vspace{-0.5cm}
\textbf{Distributional modeling:}
\begin{itemize}
\vspace{-0.05cm}
  \item Specify the complete probability distribution \\
  \vspace{-0.05cm}
  (location, scale, shape, \dots).
\end{itemize}

\smallskip

\textbf{Tree:}
\begin{itemize}
\vspace{-0.05cm}
  \item Capture non-linear and non-additive effects.
  \vspace{-0.1cm}
  \item Automatic selection of covariates and interactions. %detection of steps and abrupt changes. %(data driven)
\end{itemize}

\smallskip

\textbf{Forest:}
\begin{itemize}
\vspace{-0.05cm}
  \item Smoother effects.
  \vspace{-0.1cm}
  \item Stabilization and regularization of the model.
\end{itemize}

\bigskip

\bigskip

$\Rightarrow$ \textbf{Synthesis:} Distributional Trees and Forests
\end{minipage}
<<motivation_GAMLSS_syn, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
nobs <- 200
x_g <- c(1:nobs)/nobs
x_g <- 2*(x_g-0.5)
ytrue_g <- x_g^3
var_g <- exp(-(1.74*x_g)^2)/2.5
y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var_g)
@
<<motivation_regtree_syn, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- var <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
for(i in 1:nobs) var[i] <-  if(x[i]<1/3) 0.1 else 0.6 *(4/3 - x[i])
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@
<<motivation_randforest_syn, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@
\begin{minipage}{0.2\textwidth}
\vspace{-0.05cm}
<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g)),
  col = pallight["forest"], border = "transparent")
lines(x_g, ytrue_g, col = pal["forest"], lwd=10)
points(x_g, y_g, col = "slategray", pch = 19)
box(lwd = 5)
@

\vspace{0.15cm}

<<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
     ylim = c(min(y), max(y)))
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=10)
box(lwd=5)
@

\vspace{0.15cm}

<<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
     ylim = c(min(y), max(y)))
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=10, main = "")
box(lwd=5)
@

\vspace{0.15cm}

<<plot_motivation_distforest, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
#plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
#box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
#lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
     ylim = c(min(y), max(y)))
polygon(c(x, rev(x)), c(yforest + 0.1 + var, rev(yforest - (0.1 + var))),
       # c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g))
  col = pallight["forest"], border = "transparent")
box(lwd=5)
lines(x = x, y = yforest, col = pal["forest"], lwd=10)
@
\end{minipage}
\end{frame}

\subsection{Distributional trees}

<<treedata, echo=FALSE, results=hide>>=
set.seed(54)
nobs <- 500
x <- runif(nobs, 0, 1)
mu <- sigma <- ytrue <- numeric(length = nobs)
for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
y <- rnorm(nobs, mean = mu, sd = sigma)
ytrue <- mu
data <- data.frame(cbind(y,x, ytrue))

alldata <- cbind(data, mu, sigma)
odata <- alldata[order(alldata["x"]),]
@

<<dgp_tree, echo=FALSE, results=hide>>=
data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
names(data) <- c("x","x","x")
fig <- party(
  partynode(1L,
            split = partysplit(2L, breaks = 0.4),
            kids = list(
              partynode(2L, info = c(
                "n = 200",
                "   True parameters:   ",
                expression(mu == '4'),
                expression(sigma == '1')
              )),
              partynode(3L,
                        split = partysplit(3L, breaks = 0.8),
                        kids = list(
                          partynode(4L, info = c(
                            "n = 200",
                            "   True parameters:   ",
                            expression(mu == '12'),
                            expression(sigma == '3')
                          )),
                          partynode(5L, info = c(
                            "n = 100",
                            "   True parameters:   ",
                            expression(mu == '4'),
                            expression(sigma == '3')
                          )))))),
  data
)

node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
    gp = gpar()) 
{
    meta <- obj$data
    nam <- names(obj)
    extract_label <- function(node) {
        if (is.terminal(node)) 
            return(rep.int("", 2L))
        varlab <- character_split(split_node(node), meta)$name
        if (abbreviate > 0L) 
            varlab <- abbreviate(varlab, as.integer(abbreviate))
        if (pval) {
            nullna <- function(x) is.null(x) || is.na(x)
            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
                silent = TRUE))
            pval <- if (inherits(pval, "try-error")) 
                FALSE
            else pval
        }
        if (pval) {
            pvalue <- node$info$p.value
            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
                paste("p =", round(pvalue, digits = 3L)))
        }
        else {
            plab <- ""
        }
        return(c(varlab, plab))
    }
    maxstr <- function(node) {
        lab <- extract_label(node)
        klab <- if (is.terminal(node)) 
            ""
        else unlist(lapply(kids_node(node), maxstr))
        lab <- c(lab, klab)
        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
        lab <- lab[which.max(nchar(lab))]
        if (length(lab) < 1L) 
            lab <- ""
        return(lab)
    }
    nstr <- maxstr(node_party(obj))
    if (nchar(nstr) < 6) 
        nstr <- "aAAAAa"
    rval <- function(node) {
        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
            height = unit(3, "lines"), name = paste("node_inner", 
                id_node(node), sep = ""), gp = gp)
        pushViewport(node_vp)
        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
            seq(0.8, 1, by = 0.01))
        yell <- sqrt(xell * (1 - xell))
        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
        lab <- extract_label(node)
        fill <- rep(fill, length.out = 2L)
        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
            "lines"))
        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
        #    "lines"))
        grid.text(lab[2L], y = unit(1, "lines"))
        if (id) {
            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
                "npc"), width = max(unit(1, "lines"), unit(1.3, 
                "strwidth", nam[id_node(node)])), height = max(unit(1, 
                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
            pushViewport(nodeIDvp)
            grid.rect(gp = gpar(fill = fill[2]))
            grid.text(nam[id_node(node)])
            popViewport()
        }
        upViewport()
    }
    return(rval)
}

class(node_inner_ext) <- "grapcon_generator"

@

\begin{frame}[fragile]
\frametitle{Distributional trees}
\vspace*{-0.12cm}
\begin{center} 
DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$

\vspace*{-0.21cm}
\setkeys{Gin}{width=0.58\linewidth}
<<plottree_xyplot, fig=TRUE, echo=FALSE,width=7>>=
par(mar=c(5.1,4.1,2.4,1.1))
plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "gray")
lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
  col = pallight["forest"], border = "transparent")
legend("topleft", expression(mu  %+-%  sigma), bty = "n")
@
\end{center}
\end{frame}

% \begin{frame}
% \frametitle{Distributional trees}
% \vspace*{-0.12cm}
% \begin{center} 
% DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
% 
% \vspace*{-0.21cm}
% \setkeys{Gin}{width=0.5\linewidth}
% <<plottree_dgp, fig=TRUE, echo=FALSE>>=
% paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
% plot(fig, inner_panel = node_inner_ext,
%      tp_args = list(FUN = identity, width = 18, fill = paltrees[c(1, 3)]), 
%      ip_args = list(fill = paltrees[c(2, 3)]),
%      drop_terminal = TRUE, tnex = 1.7)
% @
% \end{center}
% \end{frame}

<<tree, echo=FALSE, results=hide>>=
set.seed(7)
nobs <- 500
x <- runif(nobs, 0, 1)
mu <- sigma <- ytrue <- numeric(length = nobs)
for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
y <- rnorm(nobs, mean = mu, sd = sigma)
#y <- rcnorm(nobs, mean = mu, sd = sigma, left = 0)
ytrue <- mu
data <- data.frame(cbind(y,x, ytrue))
tree <- disttree(y ~ x, data = data, family = NO(), type.tree = "mob")
#tree <- disttree(y ~ x, data = data, family = dist_list_cens_normal)
@

\begin{frame}[fragile]
\frametitle{Distributional trees}
\begin{center}
\vspace*{-0.12cm}
Model: \code{disttree(y ~ x)}\\
\vspace*{-0.2cm}
\setkeys{Gin}{width=0.5\linewidth}
<<plottree_estpar, fig=TRUE, echo=FALSE>>=
# function for output in terminal panels
FUN <- function (x) 
{
  cf <- x$coefficients
  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
}

paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))

## plot version using FUN and tree of class 'disttree'
plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Distributional trees}
\begin{center}
\vspace*{-0.12cm}
Model: \code{disttree(y ~ x)}\\
\vspace*{-0.2cm}
\setkeys{Gin}{width=0.5\linewidth}
<<plottree_box, fig=TRUE, echo=FALSE>>=
plot(as.constparty(tree), tnex = 1.7, drop = TRUE,
     tp_args = list(fill = paltrees[c(1, 3)], ylines = 1.5), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Distributional trees}
\begin{center}
\vspace*{-0.12cm}
Model: \code{disttree(y ~ x)}\\
\vspace*{-0.2cm}
\setkeys{Gin}{width=0.5\linewidth}
<<plottree_dens, fig=TRUE, echo=FALSE>>=
node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
{
  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
  ylines <- 1.5
  if (is.null(xscale)) xscale <- c(-5.1,22.5)
  if (is.null(yscale)) yscale <- c(-0.05,0.45)
  xr <- xscale
  yr <- yscale

  if (horizontal) {
    yyy <- xscale
    xscale <- yscale
    yscale <- yyy
  }

  rval <- function(node) {
    yrange <- seq(from = -20, to = 90)/4
    ydens <- node$info$object$ddist(yrange)

    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3, 
                                            widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")), 
                                            heights = unit(c(1, 1), c("lines", "null"))), 
                       width = unit(1, "npc"), 
                       height = unit(1, "npc") - unit(2, "lines"), 
                       name = paste("node_density",node$id, sep = ""))
    pushViewport(top_vp)
    grid.rect(gp = gpar(fill = "white", col = 0))
    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
    pushViewport(top)
    mainlab <- paste(ifelse(id, paste("Node", node$id, "(n = "), "n = "), node$info$nobs, ifelse(id, ")", ""), sep = "")

    grid.text(mainlab)
    popViewport()
    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2, 
                     xscale = xscale, yscale = yscale, 
                     name = paste("node_density",  node$id, "plot", sep = ""))
    pushViewport(plot)
    yd <- ydens
    xd <- yrange
    if (horizontal) {
      yyy <- xd
      xd <- yd
      yd <- yyy
      yyy <- xr
      xr <- yr
      yr <- yyy
      rxd <- rep(0, length(xd))
      ryd <- rev(yd)
    } else {
      rxd <- rev(xd)
      ryd <- rep(0, length(yd))
    }

    if (rug) {
      nodeobs <- node$info$object$y
      if (horizontal) {
        grid.rect(x = xscale[1], y = nodeobs , height = 0, width = xscale[1], 
                  default.units = "native", just = c("right", "bottom"),
		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
      } else {
        grid.rect(x = nodeobs, y = yscale[1], 
                  width = 0, height = abs(yscale[1]), default.units = "native", 
                  just = c("center", "bottom"),
		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
        #           default.units = "native")
        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
        #           default.units = "native")
      }
    }

    grid.polygon(x = c(xd, rxd), y = c(yd, ryd), default.units = "native",
              gp = gpar(col = "black", fill = fill, lwd = lwd))
    #grid.lines(x = xd, y = yd, default.units = "native", 
    #           gp = gpar(col = col, lwd = lwd))
    grid.xaxis()
    grid.yaxis()
    grid.rect(gp = gpar(fill = "transparent"))
    upViewport(2)
  }
  return(rval)
}

class(node_density) <- "grapcon_generator"

plot(tree, tnex = 1.7, drop = TRUE,
     terminal_panel = node_density,
     tp_args = list(fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
\end{center}
\end{frame}
%\begin{frame}
%\frametitle{Distributional trees}
%\center
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{tree_box}
%\end{frame}

\subsection{Learning}

\begin{frame}
\frametitle{Learning distributional trees and forests}
\begin{minipage}{0.7\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
Estimate $\hat{\theta}$ via maximum likelihood \\
$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
\item<8-> Test for associations/instabilities of the \\
scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each possible split variable $X_j$.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.23\textwidth}
%\vspace{-0.1cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
}
\visible<4>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{D}(Y;\hat{\theta}$)};
}
\visible<5>{
\node[inner sep=0pt] (density_all) at (0.8, 1.7)
{\includegraphics[width=0.7\textwidth]{density_all}};
}
\visible<6->{
\node[inner sep=0pt] (density_all) at (0.8, 1.7)
{\includegraphics[width=0.7\textwidth]{density_all_hist}};
}
\visible<7-9>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0.2) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0.2) {?};
\draw[-, gray, line width=0.5pt] (0.7, 1.16) -- (n1);
\draw[-, gray, line width=0.5pt] (0.9, 1.16) -- (n2);
}
\visible<10-11>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0.2) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0.2) {$Y_2$};
\draw[-, gray, line width=0.5pt] (0.7, 1.16) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (0.9, 1.16) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<12>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0.2) {$\mathcal{D}(Y_1;\hat{\theta}_1$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0.2) {$\mathcal{D}(Y_2;\hat{\theta}_2$)};
}
\visible<12->{
\draw[-, gray, line width=0.5pt] (0.7, 1.16) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (0.9, 1.16) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<13->{
\node[inner sep=0pt] (density_1_hist) at (-0.2,-0.04)
{\includegraphics[width=0.6\textwidth]{density_2_hist}};
}
\visible<13->{
\node[inner sep=0pt] (density_2_hist) at (1.8,-0.04)
{\includegraphics[width=0.6\textwidth]{density_1_hist}};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
%\begin{adjustwidth}{-0.0em}{-1em}
\begin{enumerate}
\setcounter{enumi}{2}
\item<9-> Split along the covariate $X$ with strongest association or instability \\
and at breakpoint $p$ with highest improvement in log-likelihood.
\item<11-> Repeat steps 1--3 recursively until some stopping criterion \\
is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}
%\end{adjustwidth}
\vspace{0.4cm}
\visible<14->{
{\bf Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
}
\end{frame}

\begin{frame}
\frametitle{Adaptive local likelihood estimation}

\textbf{Parameter estimator} for \only<1-3>{\textbf{a global}}\only<4->{\textbf{an adaptive local}} model\\
with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$} :

\[
\hat{\theta}\visible<2->{(\bold{x})} =  \argmax_{\theta \in \Theta} \sum_{i=1}^n \visible<2->{w_i(\bold{x}) \cdot} \ell(\theta; y_i)
\]

\medskip

\visible<3->{
\textbf{Weights:}
\begin{eqnarray*}
w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.2cm]
\visible<4->{
w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.1cm]
\visible<5->{
w^{\text{forest}}_i(\bold{x}) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} \frac{I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))}{\left|\mathcal{B}^t_b\right|}
\end{eqnarray*}
}}}
\end{frame}

\subsection{Unbiased recursive partitioning}

\begin{frame}%[fragile]
\frametitle{}
\vspace{1.5cm}
\center
{\bf \large The Power of Unbiased Recursive Partitioning: \\
\smallskip
A Unifying View of CTree, MOB, and GUIDE}

\bigskip
\bigskip

{\small
Lisa Schlosser, Torsten Hothorn, Achim Zeileis}

\bigskip
\bigskip

{\footnotesize
Submitted to\\
\emph{Computational Statistics \& Data Analysis}

\medskip

Preprint version at \url{https://arxiv.org/abs/1906.10179}}

\end{frame}

\begin{frame}
\frametitle{Unbiased recursive partitioning}

\medskip

\textbf{Idea:} 
\begin{itemize}
\smallskip
  \item Separation of split variable and split point selection.
\smallskip
  \item Application of a statistical test to assess associations between the response~$Y$ (or corresponding scores) and each possible split variable~$X_j$.
\smallskip
  \item Split variable selection based on resulting $p$-values.
\end{itemize}
\end{frame}
% \begin{frame}%[fragile]
% \frametitle{Unbiased recursive partitioning}
% 
% {\bf General testing strategy for split variable selection:}
% \begin{enumerate}
% \item Evaluate a discrepancy measure capturing the observation-wise goodness of fit of
% the considered model.
% %$\mathcal{M}(Y,X;\hat{\beta})$.
% \item Apply a statistical test assessing dependency of the discrepancy measure to each possible split variable $X_j$.
% \item Select the split variable $X$ %$X_j^\ast$ 
% showing the smallest $p$-value.
% \end{enumerate}
% 
% \bigskip
% 
% {\bf Discrepancy measures:} (Model-based) transformations of $Y$, possibly for each model parameter.
% \begin{itemize}
% \item (Ranks of) $Y$.
% \item (Absolute) deviations $Y - \bar Y$.
% \item Residuals of the fitted model. %$\mathcal{M}(Y,X;\hat{\beta})$.
% \item Score matrix of the fitted model. %$\mathcal{M}(Y,X;\hat{\beta})$.
% \item \ldots
% \end{itemize}
% \end{frame}

\begin{frame}
\frametitle{Unbiased recursive partitioning algorithms}

\textbf{GUIDE:} Loh (2002, \emph{Statistica Sinica}).
\begin{itemize}
  \item First \emph{unbiased} algorithm for recursive partitioning of linear models.
  \item Based on categorical association tests. 
\end{itemize}

\medskip

\textbf{CTree:} Hothorn, Hornik, Zeileis (2006, \emph{JCGS}).
\begin{itemize}
  \item Proposed as unbiased recursive partitioning for \emph{nonparametric} modeling.
  \item Can be model-based via model scores as the response transformation.
  \item Based on conditional permutation tests.
\end{itemize}

\medskip

\textbf{MOB:} Zeileis, Hothorn, Hornik (2008, \emph{JCGS}).
\begin{itemize}
  \item Model-based unbiased recursive partitioning. % using M-estimation (ML, OLS, CRPS, \ldots)
  \item Based on asymptotic parameter instability tests.
\end{itemize}

\end{frame}

\subsection{A unifying view}
\begin{frame}
\frametitle{A unifying view}

\textbf{Algorithms:} CTree, MOB, GUIDE are all `flavors' of a general framework.

\bigskip

\textbf{Building blocks:} For standard setup.

\bigskip

\begin{tabular}{llccl}
\hline 
      & Scores  & Binarization  & Categorization  & Test statistic \\ 
\hline
CTree & Model scores & --	           & -- 	           & Sum of squares\\
MOB   & Model scores & --	           & -- 	           & Maximally selected\\
GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
\hline
\end{tabular}

\bigskip

\textbf{Remarks:}
\begin{itemize}
  \item All three algorithms allow for certain modifications of standard setup.
  \item Further differences, e.g., null distribution, pruning strategy, etc.
\end{itemize}

\end{frame}

% \begin{frame}%[fragile]
% \frametitle{General Framework}
% 
% {\bf{Building blocks:}}
% 
% \begin{itemize}
% \item {\only<4->{\color{lightgray}} Residuals vs.\ full model scores.}
% \item {\only<1-3,6->{\color{lightgray}} Binarization of residuals/scores.}
% \item {\only<1-5>{\color{lightgray}} Categorization of possible split variables.}
% \end{itemize}
% 
% \bigskip
% 
% 
% \only<2>{
% \small
% $$
% s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
% -2 \cdot \begin{pmatrix} 
% r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X_1}\\
% r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X_2}\\
% \vdots & {\color{lightgray} \vdots} \\
% r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X_n}
% \end{pmatrix}
% %\begin{pmatrix}
% %\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
% %\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
% %\vdots & {\color{lightgray} \vdots} \\
% %\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
% %\end{pmatrix}
% $$
% }
% \only<3>{
% \small
% $$
% s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
% -2 \cdot \begin{pmatrix} 
% r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) \cdot X_1\\
% r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) \cdot X_2\\
% \vdots & \vdots \\
% r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) \cdot X_n
% \end{pmatrix}
% %\begin{pmatrix} 
% %\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
% %\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
% %\vdots & \vdots \\
% %\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
% %\end{pmatrix}
% $$
% }
% \only<4-5>{
% \small
% $$
% r(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
% \begin{pmatrix}
% r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
% r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
% \vdots \\
% r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
% \end{pmatrix}
% %
% \visible<5>{
% \quad \Rightarrow \quad
% \small
% \begin{pmatrix}
% > 0\\
% \le 0\\
% \vdots \\
% > 0
% \end{pmatrix}
% }
% $$
% }
% \only<6-7>{
% \small
% $$
% Z_j = 
% \begin{pmatrix}
% Z_{j1}\\
% Z_{j2}\\
% \vdots \\
% Z_{jn}\\
% \end{pmatrix}
% %
% \visible<7>{
% \quad \Rightarrow \quad
% \small
% \begin{pmatrix}
% \text{Q}3\\
% \text{Q}1\\
% \vdots \\
% \text{Q}2
% \end{pmatrix}
% }
% $$
% }
% \end{frame}

% \begin{frame}%[fragile]
% \frametitle{Pruning}
% 
% {\bf Goal:} Avoid overfitting.
% 
% \bigskip
% 
% {\bf Two strategies:}
% 
% \begin{itemize}
% \item \emph{Pre-pruning:}
% Internal stopping criterion based on Bonferroni-corrected $p$-values of the underlying tests. Stop splitting when there is no significant association.
% \item \emph{Post-pruning:}
% First grow a very large tree and afterwards prune splits that do not improve the model fit, 
% either via cross-validation (e.g.,~cost-complexity pruning as in CART) or 
% based on information criteria (e.g.,~AIC or BIC). 
% \end{itemize}
% \end{frame}
% 
% \subsection{Recommendations}
% \begin{frame}%[fragile]
% \frametitle{Recommendations}
% 
% \vspace{0.2cm} 
% 
% \textbf{Based on evaluations in our simulation study:}
% 
% \vspace{0.2cm} 
% 
% \begin{itemize}
% \setlength\itemsep{0.2cm}
% \item Full model scores better than residuals only. %: Using scores allows for capturing changes in all model parameters.
% \item Original values of scores/residuals better than binarized values.
% \item Categorization is simpler, but less powerful in margins.
% \item Maximally-selected statistics (as in MOB) more powerful for abrupt shifts.
% \item Linear statistics (default in CTree) more powerful for linear changes.
% \item If the significance test performs well pre-pruning works well,\\
% otherwise post-pruning might be needed.
% %\item Including full model scores is advantageous over using only residuals.
% %\item Binarizing scores/residuals does not improve the performance.
% %\item Categorizing split variables can simplify calculations, however,
% %effects might be missed unless a binning point is close to the true split point.
% %\item If the applied test performs well as significance test pre-pruning
% %is preferable, otherwise post-pruning.
% \end{itemize}
% \end{frame}

\subsection{Recommendations}
\begin{frame}%[fragile]
\frametitle{Recommendations}

\vspace{0.34cm}

\textbf{Based on evaluations in our simulation study:}

\vspace{0.24cm}

\begin{itemize}
\setlength\itemsep{0.23cm}
\item Best performance with full model scores, without binarization, and without categorization of possible split variables.
\item Adapting GUIDE to employing full model scores improves its performance.
\item Preferable test statistic depends on the type of effects (maximally-selected for abrupt shifts, linear statistics for linear changes).
%\item If the significance test performs well pre-pruning works well,\\
%otherwise post-pruning might be needed.
%\item Including full model scores is advantageous over using only residuals.
%\item Binarizing scores/residuals does not improve the performance.
%\item Categorizing split variables can simplify calculations, however,
%effects might be missed unless a binning point is close to the true split point.
%\item If the applied test performs well as significance test pre-pruning
%is preferable, otherwise post-pruning.
\end{itemize}
\end{frame}

\subsection{Weather forecasting}

\begin{frame}%[fragile]
\frametitle{}
\vspace{1.5cm}
\center
{\bf \large Distributional Regression Forests for Probabilistic\\
\smallskip
Precipitation Forecasting in Complex Terrain}

\bigskip
\bigskip

{\small
Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}

\bigskip
\bigskip

{\footnotesize
Published in\\
\emph{The Annals of Applied Statistics}

\medskip

\doi{10.1214/19-AOAS1247}}

\end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting}

\textbf{Goal:}

\medskip

\begin{center}
\begin{tikzpicture}
\draw[->] (0.4,0)--(0.9,0);
\draw[] (1,-0.4) rectangle (2.5,0.4);
\node[font=\scriptsize] at (1.75,0) {nature};
\draw[->] (2.6,0)--(3.1,0);
\visible<1>{%
\node[] at (0.0,0) {X};
\node[] at (3.5,0) {Y};}
\visible<2->{%
\node[inner sep=0pt] (today) at (-2.3,-0.4)
{\includegraphics[width=.365\textwidth]{airport_20200412.jpg}};
\node[font=\scriptsize] at (-2.3,-2.4) {2020-04-12};
\node[inner sep=0pt] (future) at (5.8,-0.4)
{\includegraphics[width=.365\textwidth]{airport_20200413.jpg}};
\node[font=\scriptsize] at (5.8,-2.4) {2020-04-13};}
\end{tikzpicture}
\end{center}

%% https://innsbruck-airport.panomax.com/

\only<1-2>{%
\textbf{Data:}
\begin{itemize}
\item X: State of the atmosphere now (temperature, precipitation, wind, \dots).
\item Y: State of the atmosphere in the future (hours, days, weeks, \dots).
\end{itemize}}%
\only<3->{%
\textbf{Two stages:}
\begin{itemize}
\item Physical model: Numerical weather prediction (NWP).
\item Statistical model: Model output statistics (MOS).
\end{itemize}}

\end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting}

\textbf{Numerical weather prediction (NWP):}
\begin{itemize}
  \item Based on a physical model.
  \item Massive numerical simulation of atmospheric processes.
  \item Here: Global model on a $50\times 50 \text{ km}^2$ grid.
\end{itemize}

\medskip

\textbf{Problem:} Uncertain initial conditions, unresolved processes.

\medskip

\textbf{Solution:} Ensemble of simulation runs under perturbed conditions.

\end{frame}

<<echo = FALSE, results=hide>>=
library("zoo")

Sys.setenv("TZ"="UTC")

# -------------------------------------------------------------------
# GFS ensemble forecast
# -------------------------------------------------------------------

GFS   <- read.table("Data/GENS_00_innsbruck_20200412.dat", header = TRUE)
init  <- min(as.POSIXct(GFS$timestamp-GFS$step*3600,origin="1970-01-01"))
title <- sprintf("%s\n%s",
                 "Global Forecast System (GFS) Ensemble Forecast for Innsbruck, Airport",
                 sprintf("Forecast initialized %s", strftime(init, "%Y-%m-%d %H:%M UTC")))

# observations
load("Data/STAGEobs_tawes_11121_defense.rda")
#load("Data/STAGEobs_tawes_11121.rda")
# "2018-03-14 00:00:00" in line 321 (first entry for temperature)
# "2018-03-13 06:00:00" in line 317 (first entry for rain -> aggregated to 24h sums)
# "2018-03-23 00:00:00" in line 357

getZoo <- function( x, variable ) {
  x <- subset( x, varname == variable )
  if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
  # Else create zoo
  x <- zoo( subset(x,select=-c(varname,timestamp,step)),
            as.POSIXct(x$timestamp,origin="1970-01-01") )
  x
}
GFS_t2m  <- getZoo( GFS, "tmp2m" ) - 273.15
# temperature bias: model simulates temperature for height 1675 (~1070 m higher than true height)
# => add 10.7 degrees to account for bias (approx. 1 degree / 100 m)
# Custom bias
GFS_t2m <- GFS_t2m + 6.5

# Subsetting obs
obs <- subset(obs, index(obs) %in% index(GFS_t2m))

GFS_t2m <- merge(GFS_t2m, obs = obs[, grep("tl", names(obs))])
###GFS_t2m$obs <- obs[,"obs$tl"]
# first value of rain is 18 hours later than temperature 
#(both start at 6:00 but rain is cumulated over 24 hours)
# -> drop first 3 values of temperature
GFS_t2m <- GFS_t2m[-c(1:3),]

GFS_rain <- getZoo( GFS, "apcpsfc" )
GFS_rain$obs <- obs[,"rr6"]
# precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
GFS_rain24 <- GFS_rain[c((1:10)*4),]
for(i in 1:NROW(GFS_rain24)){
  GFS_rain24[i,] <- colSums(GFS_rain[c((4*i-3):(4*i)),])
}

# Two POSIXct vectors for the axis
main  <- seq(min(as.POSIXct(as.Date(index(GFS_t2m)))),max(index(GFS_t2m)),by=86400)
minor <- seq(min(index(GFS_t2m)),max(index(GFS_t2m)),by=3*3600)

# GFS_t2m_mean <- rowMeans(GFS_t2m)
GFS_t2m_mean <- GFS_t2m
GFS_t2m_mean[,1] <- rowMeans(GFS_t2m[,-NCOL(GFS_t2m)])
GFS_t2m_mean <- GFS_t2m_mean[,1]

# GFS_rain_mean <- rowMeans(GFS_rain)
GFS_rain24_mean <- GFS_rain24
GFS_rain24_mean[,1] <- rowMeans(GFS_rain24[,-NCOL(GFS_rain24)])
GFS_rain24_mean <- GFS_rain24_mean[,1]
@

<<t2m, echo=FALSE, eval=FALSE>>=
par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
layout( matrix(1:2,ncol=1) )
plot(GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m, na.rm = TRUE)*1.05, col = "slategray" )
abline(v = main, lty = 2 )
mtext(side = 3, line = 1, cex = 1.2, font = 2, title )
mtext(side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
@

<<rain24, echo=FALSE, eval=FALSE>>=
plot(GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24, na.rm = TRUE)*c(0,1.05), col = "slategray" )
abline(v = main, lty = 2 )
mtext(side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
axis(side = 1, at = main + 42300,  strftime(main,"%b %d"),
line = 0, col.ticks=NA, col.axis="gray30", col = NA )
@

\begin{frame}[fragile]
\frametitle{Weather forecasting}

\vspace{-0.6cm}
\setkeys{Gin}{width=0.95\linewidth}
\begin{center}
\only<1>{%
<<gfs1, fig=TRUE, echo=FALSE, width=9>>=
<<t2m>>
<<rain24>>
@
}%
\only<2>{%
<<gfs2, fig=TRUE, echo=FALSE, width=9>>=
<<t2m>>
lines(GFS_t2m[,3], col = "slategray")
<<rain24>>
lines(GFS_rain24[,3], col = "slategray")
@
}%
\only<3>{%
<<gfs3, fig=TRUE, echo=FALSE, width=9>>=
<<t2m>>
lines(GFS_t2m[,2], col = "slategray")
lines(GFS_t2m[,3], col = "slategray")
<<rain24>>
lines(GFS_rain24[,2], col = "slategray")
lines(GFS_rain24[,3], col = "slategray")
@
}%
\only<4>{%
<<gfs4, fig=TRUE, echo=FALSE, width=9>>=
<<t2m>>
for(j in 2:(NCOL(GFS_t2m) - 1)) lines(GFS_t2m[,j], col = "slategray")
<<rain24>>
for(j in 2:(NCOL(GFS_rain24) - 1)) lines(GFS_rain24[,j], col = "slategray")
@
}%
\only<5>{%
<<gfs5, fig=TRUE, echo=FALSE, width=9>>=
<<t2m>>
for(j in 2:(NCOL(GFS_t2m) - 1)) lines(GFS_t2m[,j], col = "slategray")
lines(GFS_t2m[,"obs"], col = 2, lwd = 2.3)
<<rain24>>
for(j in 2:(NCOL(GFS_rain24) - 1)) lines(GFS_rain24[,j], col = "slategray")
lines(GFS_rain24[,"obs"], col = 2, lwd = 2.3)
@
}%
\end{center}

\end{frame}

\subsection{Weather forecasting: precipitation}

\begin{frame}[fragile]
\frametitle{Weather forecasting: precipitation}

\textbf{Goal:} Predict daily precipitation amount in Tyrol. %complex terrain.

\bigskip
\pause

\textbf{Observation data:} National Hydrographical Service.
\begin{itemize}
\item Daily 24h precipitation sums from July over 28 years (1985--2012).
\item 95 observation stations in Tyrol.
\end{itemize}

\bigskip
\pause

\textbf{NWP:} Global Ensemble Forecast System.
\begin{itemize}
\item Model outputs: Precipitation, temperature, air pressure, 
  convective available potential energy, downwards short wave radiation flux, \dots
\item 80 covariates based on ensemble min/max/mean/standard deviation.
\end{itemize}

\bigskip
\pause

\textbf{Distribution assumption:} Power-transformed Gaussian, censored at 0.
\[ (\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \]

\end{frame}

%% \begin{frame}
%% \frametitle{Weather forecasting: precipitation}
%% 
%% \textbf{Base variables:}
%% \begin{itemize}
%% \item Total precipitation.
%% \item Convective available potential energy.
%% \item Downwards short wave radiation flux (``sunshine'').
%% \item Mean sea level pressure.
%% \item Preciptable water.
%% \item 2m maximum temperature.
%% \item Total column-integrated condensate.
%% \item Temperature.
%% \item Temperature differences in altitude.
%% \end{itemize}
%% 
%% \medskip
%% 
%% \textbf{Variations:} 80 covariates based on ensemble min/max/mean/standard deviation.
%% 
%% \end{frame}

<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
if(file.exists("Data/Axams_pred.rda") & file.exists("Data/Axams_testdata.rda")){
  load("Data/Axams_pred.rda")
  load("Data/Axams_testdata.rda")
} else {
  #####
  # load observations and covariates 
  data("RainAxams")

  # tree and forest formula
  {
    dt.formula <- df.formula <- 
      robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
      tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
      tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
      capepow_mean + capepow_sprd + capepow_min + capepow_max + 
      capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
      capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
      dswrf_mean_mean + dswrf_mean_max +  
      dswrf_sprd_mean + dswrf_sprd_max + 
      msl_mean_mean + msl_mean_min + msl_mean_max + 
      msl_sprd_mean + msl_sprd_min + msl_sprd_max +
      pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
      pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
      tmax_mean_mean + tmax_mean_min + tmax_mean_max +
      tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
      tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
      tcolc_sprd_mean + tcolc_sprd_min + tcolc_sprd_max +
      t500_mean_mean + t500_mean_min + t500_mean_max +
      t700_mean_mean + t700_mean_min + t700_mean_max +
      t850_mean_mean + t850_mean_min + t850_mean_max +
      t500_sprd_mean + t500_sprd_min + t500_sprd_max +
      t700_sprd_mean + t700_sprd_min + t700_sprd_max +
      t850_sprd_mean + t850_sprd_min + t850_sprd_max +
      tdiff500850_mean + tdiff500850_min + tdiff500850_max +
      tdiff700850_mean + tdiff700850_min + tdiff700850_max +
      tdiff500700_mean + tdiff500700_min + tdiff500700_max +
      msl_diff

  }

  # learning data: 24 years (1985 - 2008, both inlcuded)
  # testing data: 4 successive years (2009, 2010, 2011, 2012)
  learndata <- RainAxams[RainAxams$year < 2009,]
  testdata <- RainAxams[RainAxams$year %in% c(2009, 2010, 2011, 2012),]
  save(file = "Data/Axams_testdata.rda", testdata)

  ##############################################################
  # fitting the models
  set.seed(7)

  df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
                   ntree = 100, censtype = "left", censpoint = 0,
                   control = disttree_control(teststat = "quad", testtype = "Univ",
                                              type.tree = "ctree",
                                              intersplit = TRUE,
                                              mincriterion = 0, minsplit = 50,
                                              minbucket = 20), mtry = 27)

  #### prepare data for plot of estimated density functions
  # predictions for one day (in each of the four years) 
  # (19th of July 2011 is missing)
  pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),

  pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
  pdf <- predict(df, newdata = testdata[pdays,], type = "parameter")
  pdf$pdays <- pdays
  save(file = "Data/Axams_pred.rda", pdf)
}

pdays <- pdf$pdays
pday <- pdays[1]
df_mu <- pdf$mu
df_sigma <- pdf$sigma

# plot predicted distributions together with observations
set.seed(7)
x <- c(0.01, sort(runif(500,0.01,8)))
y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)

# switch x-axis back to untransformed scale
# x <- x^(1.6)

# point mass (slightly shifted)
pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.15
pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.05
pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.05
pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.15

# predictions
pred1 <- c(testdata[pdays,"robs"][1], # ^(1.6), 
           crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
pred2 <- c(testdata[pdays,"robs"][2], # ^(1.6), 
           crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
pred3 <- c(testdata[pdays,"robs"][3], # ^(1.6), 
           crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
pred4 <- c(testdata[pdays,"robs"][4], # ^(1.6), 
           crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))

#legendheight
lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
@

\begin{frame}[fragile]
\frametitle{Weather forecasting: precipitation}

\textbf{Application for one station:} Axams.
\begin{itemize}
%\vspace{-0.05cm}
\item Learn forest model on data from 24 years (1985--2008).
%\vspace{-0.05cm}
\item Evaluate on 4 years (2009--2012). Here: July \Sexpr{24}.
\end{itemize}

\vspace*{-0.4cm}
\begin{center}
\setkeys{Gin}{width=0.54\textwidth}
<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=4.3, width=7>>=
par(mar = c(4, 4, 1, 1))
plot(x = x, y = y1, type = "l", col = 2, ylab = "Density", lwd = 2,
     # xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
     xlim = c(-1.5,8))     #  xlim = c(-2,28))

lines(x = x, y = y2, type = "l", lwd = 2, col = 4)
lines(x = x, y = y3, type = "l", lwd = 2, col = 3)
lines(x = x, y = y4, type = "l", lwd = 2, col = 6)
legend('topright', c("Predicted distribution", "Point mass at censoring point", "Observation"),
       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 1)

# plot point mass
lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = 2, type = "l", lwd = 2)
lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = 4, type = "l", lwd = 2)
lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = 3, type = "l", lwd = 2)
lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = 6, type = "l", lwd = 2)

points(x = pm1[1], y = pm1[2], col = 2, pch = 19, cex = 1.4)
points(x = pm2[1], y = pm2[2], col = 4, pch = 19, cex = 1.4)
points(x = pm3[1], y = pm3[2], col = 3, pch = 19, cex = 1.4)
points(x = pm4[1], y = pm4[2], col = 6, pch = 19, cex = 1.4)

# plot predictions
points(x = pred1[1], y = pred1[2], col = 2, pch = 4, cex = 1.4, lwd = 2)
points(x = pred2[1], y = pred2[2], col = 4, pch = 4, cex = 1.4, lwd = 2)
points(x = pred3[1], y = pred3[2], col = 3, pch = 4, cex = 1.4, lwd = 2)
points(x = pred4[1], y = pred4[2], col = 6, pch = 4, cex = 1.4, lwd = 2)

lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)
lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)
lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)
lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)

# add labels
text(x = -0.8, y = lh1, labels = "2009", col = 2, cex = 1)      # -1.7 
text(x = -0.8, y = lh2, labels = "2010", col = 4, cex = 1)      # -1.7
text(x = -0.8, y = lh3, labels = "2011", col = 3, cex = 1)      # -1.7
text(x = -0.8, y = lh4, labels = "2012", col = 6, cex = 1)      # -1.7
@
\end{center}
\end{frame}

% \begin{frame}[fragile]
% \frametitle{Weather forecasting: precipitation}
% 
% \textbf{Application for one station:} Axams.
% \begin{itemize}
% %\vspace{-0.05cm}
% \item Learn forest model on data from 24 years.
% %\vspace{-0.05cm}
% \item Evaluate on 4 years.
% %\vspace{-0.05cm}
% \item 10 times 7-fold cross validation.
% \end{itemize}
% 
% \bigskip
% 
% \textbf{Benchmark:} Against other heteroscedastic censored Gaussian models.
% \begin{itemize}
% \item \emph{Ensemble MOS:} Linear predictors using only total precipitation.
% \item \emph{Prespecified GAMLSS:} Variable selection based on expert knowledge.
% \item \emph{Boosted GAMLSS:} Automatic variable selection.
% \end{itemize}
% 
% \bigskip
% 
% \textbf{Evaluation:} Continuous ranked probability skill score.
% 
% %% \vspace{0.1cm}
% %% \begin{table}[t!]
% %% \footnotesize
% %% 
% %% \hskip-0.4cm\begin{tabular}{ l l l }
% %% \hline
% %% Model & Type & Variable Selection \\
% %% \hline
% %% \vspace{-0.15cm}
% %% \textbf{Distributional forest} & recursive    & automatic \\ 
% %%                                & partitioning &           \\ 
% %% \vspace{-0.15cm}
% %% \textbf{Prespecified GAMLSS}   & spline  & based on expert \\
% %%                                & in each & knowledge      \\
% %% \vspace{-0.15cm}
% %% \textbf{Boosted GAMLSS}        & spline  & automatic  \\
% %%                                & in each &            \\
% %% \vspace{-0.15cm}
% %% \textbf{EMOS}                  & linear & total precipitation mean \\
% %%                                &        & and standard deviation   \\
% %% \hline
% %% \end{tabular}
% %% \end{table}
% 
% \end{frame}
% 
% 
% 
% 
% <<echo=FALSE, results=hide>>=
% #### cross validation rain
% if(file.exists("Data/crps_cross.rda")){
%   load("Data/crps_cross.rda")
% } else {
%   
%   nrep_cross <- 10
%   seed <- 7
%   
%   res_cross <- mclapply(1:nrep_cross,
%                         function(i){
%                           
%                           set.seed(seed*i)
%                           
%                           # randomly split data in 7 parts each including 4 years
%                           years <- 1985:2012
%                           testyears <- list()
%                           for(j in 1:7){
%                             testyears[[j]] <- sample(years, 4, replace = FALSE)
%                             years <- years[!(years %in% testyears[[j]])]
%                           }
%                           
%                           #crps <- matrix(nrow = 7, ncol = 7)
%                           reslist <- list()
%                           for(k in 1:7){
%                             test <- testyears[[k]]
%                             train <- c(1985:2012)[!c(1985:2012) %in% test]
%                             
%                             res <- evalmodels(station = "Axams",
%                                               train = train,
%                                               test = test,
%                                               gamboost_cvr = TRUE)
%                             
%                             #crps[k,] <- res$crps
%                             reslist[[k]] <- res
%                           }
%                           
%                           #colnames(crps) <- names(res$crps)
%                           return(reslist)
%                         },
%                         mc.cores = detectCores() - 1
%   )
%   
%   # extract CRPS
%   crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
%   # loop over all repetitions
%   for(i in 1:length(res_cross)){
%     #loop over all 7 folds (for 7 methods)
%     crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
%     for(j in 1:length(res_cross[[1]])){
%       crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
%     }
%     crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
%   }
%   colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps) 
%   
%   save(crps_cross, file = "Data/crps_cross.rda")
% }
% @
% 
% 
% 
% 
% \begin{frame}[fragile]
% \frametitle{Weather forecasting: precipitation}
% \begin{center}
% \vspace*{-0.2cm}
% \setkeys{Gin}{width=0.65\textwidth}
% <<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.6, width=7>>=
% #par(mar = c(2.5,2,1,2))
% boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
%         names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
%         main = "Cross validation (with reference model EMOS)",
%         cex.main=1.4,
%         cex.lab=1.2,
%         ylab = "CRPS skill score", col = "lightgray") 
% abline(h = 0, col = pal["EMOS"], lwd = 2)
% @
% \end{center}
% \end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting: precipitation}
%\vspace{0.2cm}
\textbf{Application for all 95 stations:}
\begin{itemize}
\item Learn forest model on data from 24 years (1985--2008).
\item Evaluate on 4 years (2009--2012).
\item Benchmark against other heteroscedastic censored Gaussian models.
\end{itemize}

\end{frame}

<<echo = FALSE>>=
#### prediction over all stations 24 - 4
if(file.exists("Data/crps_24to4_all.rda")){
  load("Data/crps_24to4_all.rda")
} else {

  data("StationsTyrol")
  stations <- StationsTyrol$name
  test <- 2009:2012
  train <- 1985:2008

  res_24to4_all <- mclapply(1:length(stations),
                            function(i){

                              set.seed(7)

                              res <- evalmodels(station = stations[i],
                                                train = train,
                                                test = test,
                                                gamboost_cvr = TRUE)

                              return(res)
                            },
                            mc.cores = detectCores() - 1
  )

  # extract crps
  crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
  # loop over all stations
  for(i in 1:length(stations)){
    crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
  }

  colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
  rownames(crps_24to4_all) <- stations

  save(crps_24to4_all, file = "Data/crps_24to4_all.rda")
}

# skill score
s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")

## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps

crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]

# best method
bst <- apply(crps_map, 1, which.min)

# distance of forest to best other method
dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]

# breaks/groups
brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
grp <- cut(dst, breaks = brk)

# HCL colors (relatively flashy, essentially CARTO Tropic)
clr <- colorspace::diverging_hcl(5, h = c(130, 320), c = 70, l = c(50, 90), power = 1.3)

library("raster") # dem (digital elevation model)
library("sp")     # gadm www.gadm.org/country

data("StationsTyrol", package = "RainTyrol")
data("MapTyrol", package = "RainTyrol")
# data(MapTyrol_border, package = "RainTyrol")
# Create SpatialPointsDataFrame from station list
sp <- SpatialPointsDataFrame(subset(StationsTyrol,
                                    select=c(lon,lat)),
                             data = subset(StationsTyrol,
                                           select = -c(lon,lat)),
                             proj4string = crs(MapTyrol$RasterLayer))
@

\begin{frame}[fragile]
\frametitle{Weather forecasting: precipitation}
\begin{center}
\vspace*{-0.5cm}

\setkeys{Gin}{width=0.84\textwidth}
<<map, fig=TRUE, echo=FALSE, width=10, height=6.5>>=

  ## plot map of Tyrol with all 95 observations
  layout(cbind(1, 2), width = c(9, 1))
  par(mar = c(5,4,4,0.1))
  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
                xlim = c(9.8,13.2), 
                ylim = c(46.6, 47.87))
  plot(MapTyrol$SpatialPolygons, add = TRUE)
  points(sp[70,], pch = 19, col = "black", cex = 1.85)
  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
  par(mar = c(0.5,0.2,0.5,2.3))
  ## legend
  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))

@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting: precipitation}
\begin{center}
\vspace*{-0.5cm}

\setkeys{Gin}{width=0.47\textwidth}
<<map, fig=TRUE, echo=FALSE, width=10, height=6.5>>=

  ## plot map of Tyrol with all 95 observations
  layout(cbind(1, 2), width = c(9, 1))
  par(mar = c(5,4,4,0.1))
  raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
                xlim = c(9.8,13.2), 
                ylim = c(46.6, 47.87))
  plot(MapTyrol$SpatialPolygons, add = TRUE)
  points(sp[70,], pch = 19, col = "black", cex = 1.85)
  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
  par(mar = c(0.5,0.2,0.5,2.3))
  ## legend
  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))

@
\end{center}
\vspace{-0.2cm}
\textbf{Conclusions:}
\begin{itemize}
\item Distributional forests provide an easy-to-apply alternative to already existing methods.
\item Do not require expert knowledge for pre-specifications.
\item Competitive performance: mainly en par or even slightly better.
\end{itemize}

\end{frame}

\subsection{Weather forecasting: wind direction}

\begin{frame}%[fragile]
\frametitle{}
\vspace{1.5cm}
\center
{\bf \large 
Circular Regression Trees and Forests with an \\
\smallskip
Application to Probabilistic Wind Direction Forecasting}

\bigskip
\bigskip

\small{
Moritz N. Lang, Lisa Schlosser, Torsten Hothorn, \\
Georg J. Mayr, Reto Stauffer, Achim Zeileis}

\bigskip
\bigskip

{\footnotesize
Minor revision in\\
\emph{Journal of the Royal Statistical Society C}

\medskip

Preprint version at \url{https://arxiv.org/abs/2001.00412}}

\end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting: wind direction}

\textbf{Goal:} Nowcasting (1--3 hours ahead) of wind direction at Innsbruck Airport.

\bigskip
\pause

\only<2>{
\vspace{-0.4cm}
\setkeys{Gin}{width=0.7\linewidth}
\begin{center}
\includegraphics{ibk_airport.png}
\end{center}
}

\only<3>{
\vspace{-0.5cm}
\setkeys{Gin}{width=0.7\linewidth}
\begin{center}
\includegraphics{ibk_airport_east.png}
\end{center}

\vspace{-1.4cm}
Up-valley wind
}

\only<4>{
\vspace{-0.5cm}
\setkeys{Gin}{width=0.7\linewidth}
\begin{center}
\includegraphics{ibk_airport_west.png}
\end{center}

\vspace{-1.4cm}
Down-valley wind
}

\only<5>{
\vspace{-0.5cm}
\setkeys{Gin}{width=0.7\linewidth}
\begin{center}
\includegraphics{ibk_airport_south.png}
\end{center}

\vspace{-1.4cm}
Southerly wind
}

\only<6->{
\textbf{Challenges:}
\begin{itemize}
  \item Possibly abrupt changes due to geographical position.
  \item Circular response in $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
  \item NWP outputs are less useful due to short lead time.
\end{itemize}

\bigskip

\visible<7->{
\textbf{Inputs:} Observation data only (41,979 data points).
\begin{itemize}
  \item 4 stations at Innsbruck Airport, 6 nearby weather stations. 
  \item Base variables: Wind direction, wind (gust) speed, temperature, \\
  (reduced) air pressure, relative humidity.
  \item 260 covariates based on means/minima/maxima, temporal changes, spatial differences towards the airport.
\end{itemize}
}}
\end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting: wind direction}

\vspace*{-0.3cm}

\begin{minipage}{0.5\textwidth}
\textbf{Distribution assumption:} Von Mises.
\begin{itemize}
\item Circular normal distribution.
\item Location parameter $\mu \in [0, 2 \pi)$.
\item Concentration parameter $\kappa > 0$.
\end{itemize}
\end{minipage}
\begin{minipage}{0.49\textwidth}
\only<1>{\includegraphics[width=\linewidth]{density_linear.pdf}}%
\only<2>{\includegraphics[width=\linewidth]{density_circular.pdf}}%
\end{minipage}

\bigskip
\medskip

\textbf{Log-likelihood:} $y \in [0, 2 \pi)$ and parameter vector $\theta = (\mu, \kappa)$.
$$
\ell(\theta; y) = \log \left\{ \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)} \right\}
$$
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$.

\end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting: wind direction}

\vspace{-0.7cm}

\setkeys{Gin}{width=1\linewidth}
\begin{center}
\includegraphics{circtree_ibk.pdf}
\end{center}

\end{frame}

\begin{frame}[fragile]
\frametitle{Weather forecasting: wind direction}

\vspace{-0.54cm}

\setkeys{Gin}{width=0.47\linewidth}
\begin{figure}
\includegraphics[center]{circtree_ibk.pdf}
\end{figure}

%\vspace{-0.3cm}
\textbf{Conclusions:}
\begin{itemize}
\item Circular trees are an easy-to-apply method with high interpretability. 
\item A single circular tree provides reliable predictions which can be improved by employing a circular forest model.
\item Competitive performance: better than persistence and linear model.
\end{itemize}

\end{frame}

% \begin{frame}[fragile]
% \frametitle{Weather forecasting: wind direction}
% 
% \textbf{Benchmark:} Against other naive and circular models.
% \begin{itemize}
% \item Climatology: Without covariates.
% \item Persistency: Based on current wind direction.
% \item Circular GLM: Based on current wind speed and wind vectors $(u, v)$.
% \end{itemize}
% 
% \end{frame}
% 
% 
% \begin{frame}[fragile]
% \frametitle{Weather forecasting: wind direction}
% 
% %\includegraphics{circforest_crpsraw_ibk_lag1.pdf}
% %\includegraphics{circforest_crpsraw_ibk_lag3.pdf}
% 
% \textbf{Evaluation:} CRPS skill score for \only<1>{1}\only<2>{3}-hourly predictions (5-fold cross validation).
% \begin{center}
% \setkeys{Gin}{width=0.88\linewidth}
% \only<1>{\includegraphics{circforest_crpsskill_ibk_lag1.pdf}}%
% \only<2>{\includegraphics{circforest_crpsskill_ibk_lag3.pdf}}
% \end{center}
% 
% \end{frame}

%\subsection{Transformation models}
%
% \begin{frame}
% \frametitle{Transformation models}
% 
% \textbf{Alternative:} When no obvious classic distribution assumption is available.
% 
% \medskip
% 
% \textbf{Advantages:}
% \begin{itemize}
% \item Does not require specification of distribution family.
% \item More flexible framework.
% \end{itemize}
% 
% \medskip
% 
% \textbf{Distribution function:}
% \[
% F(y; \theta) = \Phi(\bold{a}_{Bs,d}(y)^\top \theta) 
% \]
% \begin{itemize} 
% \item $\bold{a}_{Bs,d}(y)^\top \theta$ is a smooth, monotone Bernstein polynomial of degree $d$.
% \item $d=1$ corresponds to $\mathcal{N}(\mu,\sigma^2)$.
% \item $d=5$ is surprisingly flexible.
% \end{itemize}
% 
% \medskip
% 
% \textbf{Example:} Body Mass Index explained by lifestyle factors (Switzerland).
% 
% \end{frame}
% 
% \begin{frame}
% \frametitle{Transformation models}
% 
% \vspace*{0.2cm}
% \includegraphics[width=0.84\textwidth]{cmpx-tree.pdf}
% 
% \end{frame}

\subsection{Software}

\begin{frame}
\frametitle{Software}
%\vspace{0.4cm}
\textbf{Software:} \emph{disttree} and \emph{circtree} 
available on R-Forge at\\

\medskip

\small{\url{https://R-Forge.R-project.org/projects/partykit/}}\\ 
%and\\
%\url{https://R-Forge.R-project.org/projects/uibk-rprog-2018/}}\\

\bigskip
\medskip

\textbf{Main functions:}

\medskip

\begin{tabular}{ll}
\code{distfit}    & Distributional fits (ML, \code{gamlss.family}/custom \code{list}).\\
& No covariates. \\
\code{disttree}   & Distributional trees (\code{ctree}/\code{mob} + \code{distfit}).\\
& Covariates as partitioning variables. \\
\code{distforest} & Distributional forests (ensemble of \code{disttree}s).\\
& Covariates as partitioning variables.
\end{tabular}

\bigskip
\medskip

\textbf{Correspondingly:} \code{circtree}, \code{circforest}

\end{frame}

\subsection{Papers}

\begin{frame}
\frametitle{Papers}

%\vspace{-0.3cm}

\footnotesize
Schlosser L, Hothorn T, Stauffer R, Zeileis A (2019).
\dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
\emph{The Annals of Applied Statistics}, \textbf{13}(3), 1564--1589.
\doi{10.1214/19-AOAS1247}

\bigskip

Schlosser L, Hothorn T, Zeileis A (2019). 
\dquote{The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE}
\emph{arXiv 1906.10179}, arXiv.org E-Print Archive.
\url{https://arxiv.org/abs/1906.10179}.
Submitted to \emph{Computational Statistics \& Data Analysis}.

\bigskip

Lang MN, Schlosser L, Hothorn T, Mayr GJ, Stauffer R, Zeileis A (2020).
\dquote{Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting.}
Minor revision in \emph{Journal of the Royal Statistical Society C}.
Preprint version at \url{https://arxiv.org/abs/2001.00412}

% % \medskip
% % 
% % Hothorn T, Zeileis A (2017).
% %   \dquote{Transformation Forests.}
% %   \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
% %   \url{http://arxiv.org/abs/1701.02110}
%   
% \bigskip
% \bigskip
% %\medskip
%   
% Hothorn T, Hornik K, Zeileis A (2006).
%  \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
%  \emph{Journal of Computational and Graphical Statistics},
%  \textbf{15}(3), 651--674.
%  \doi{10.1198/106186006X133933}
%  
% \medskip
% 
% Zeileis A, Hothorn T, Hornik K (2008).
%  \dquote{Model-Based Recursive Partitioning.}
%   \emph{Journal of Computational and Graphical Statistics},
%   \textbf{17}(2), 492--514.
%   \doi{10.1198/106186008X319331}
% 
% \medskip
% 
% Hothorn T, Zeileis A (2015).
%  \dquote{{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
%  \emph{Journal of Machine Learning Research},
%  \textbf{16}, 3905--3909.
%  \url{http://www.jmlr.org/papers/v16/hothorn15a}
% 
% 
% 
% %% \medskip
% %% 
% %% Stasinopoulos DM, Rigby RA (2007).
% %%   \dquote{Generalized Additive Models for Location Scale and Shape (GAMLSS) in \textsf{R}.}
% %%   \emph{Journal of Statistical Software}, 
% %%   \textbf{23}(7), 1--46.
% %%   \doi{10.18637/jss.v023.i07}
% %%   
% %% \medskip
% %% 
% %% Breiman L (2001).
% %%   \dquote{Random {Forests}}
% %%   \emph{Machine Learning}, 
% %%   \textbf{45}(1), 5--32.
% %%   \doi{10.1023/A:1010933404324}

\end{frame}

\end{document}

### 2020-04-26 pkg/disttree/inst/PhD_thesis_defense/plot_densities.R

@@ -1,0 +1,83 @@
+set.seed(7)
+nobs <- 1000
+x <- c(1:nobs)/nobs
+mu1 <- 0.7
+sigma1 <- 0.05
+mu2 <- 0.4
+sigma2 <- 0.1
+y1 <- rnorm(nobs, mu1, sigma1)
+y2 <- rnorm(nobs, mu2, sigma2)
+mu <- mean(c(y1,y2))
+sigma <- sd(c(y1,y2))
+yd <- rnorm(10000*nobs, mean = mu, sd = sigma)
+yd1 <- rnorm(10000*nobs, mean = mu1, sd = sigma1)
+yd2 <- rnorm(10000*nobs, mean = mu2, sd = sigma2)
+
+setwd("~/svn/partykit/pkg/disttree/inst/defensio/")
+par(mar=c(0,0,0,0))
+min(c(y1, y2))
+max(c(y1, y2))
+width <- 11
+lwd <- 14
+
+## HCL palette
+pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
+names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
+pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
+names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
+transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
+
+pdf(file = "density_all.pdf", width = width)
+hist(c(y1,y2), breaks = c(0:50)/50, freq = FALSE, 
+     xaxt="n", yaxt="n", ann=FALSE, border = "white", col = "white",
+     ylim = c(0,4.5))
+lines(density(yd), col = pal["forest"], lwd = lwd)
+box(lwd=1)
+dev.off()
+
+pdf(file = "density_all_hist.pdf", width = width)
+hist(c(y1,y2), breaks = c(0:50)/50, freq = FALSE, 
+     xaxt="n", yaxt="n", ann=FALSE, border = transpgray, col = transpgray,
+     ylim = c(0,4.5))
+lines(density(yd), col = pal["forest"], lwd = lwd)
+box(lwd=1)
+dev.off()
+
+pdf(file = "density_1.pdf", width = width)
+hist(y1, breaks = c(0:50)/50, freq = FALSE, 
+     xaxt="n", yaxt="n", ann=FALSE, border = "white", col = "white",
+     ylim = c(0,9))
[…]

### 2020-04-26 pkg/disttree/inst/Psychoco2018/slides.Rnw
slides for PhD thesis defense
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3011 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2ac42af760b377bd2d56a4a05978e1eca5f1f0e9

@@
 learndata <- dgp(300, family = dist_list_normal, round.sp = 4, fun = fun)
-control <-  ctree_control(teststat = "quad", testtype = "Bonferroni", mincriterion = 0.7, minbucket = 20L)
-dt <- disttree(y~x1, data=learndata, family=dist_list_normal, type.tree = "ctree", control = control)
+control <-  disttree_control(type.tree = "ctree", teststat = "quad", testtype = "Bonferroni", mincriterion = 0.7, minbucket = 20L)
+dt <- disttree(y~x1, data=learndata, family=dist_list_normal, control = control)
 
@@
 } else {
-  control <-  ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 7L)
-  df <- distforest(y~x1, data=learndata, family=dist_list_normal, ntree = 500L, type.tree = "ctree", control = control)
+  control <-  disttree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 7L,type.tree = "ctree")
+  df <- distforest(y~x1, data=learndata, family=dist_list_normal, ntree = 500L, control = control)
   df$info <- df$fitted <- df$weights <- df$nodes <- df$data <- df$terms <- df$loglik <- df$trafo <- NULL
@@
 
+\textbf{Alternative:} When no obvious classic distribution assumption is available.
+
 \textbf{Advantages:}

### 2020-04-26 pkg/disttree/inst/Psychoco2018/slides.Rnw
small corrections to adapt for changes and updates in the disttree implementation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3012 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5622db15384bfba47becde3a04582ef1bfc97701

@@
 \endminipage
-\hspace{1.1cm}
+\hspace{1.cm}
 \minipage{0.25\textwidth}
@@
 dt <- disttree(y~x1, data=learndata, family=dist_list_normal, control = control)
+dt$fitted.par <- predict(dt, type = "parameter")
 
@@
   df <- distforest(y~x1, data=learndata, family=dist_list_normal, ntree = 500L, control = control)
+  df$fitted.par <- predict(df, type = "parameter")
   df$info <- df$fitted <- df$weights <- df$nodes <- df$data <- df$terms <- df$loglik <- df$trafo <- NULL

### 2020-04-26 pkg/disttree/inst/useR2017/slides.Rnw
adaptation to latest uibk beamer theme
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1499 255f4249-2263-4417-b1ea-8498779e27a7
first draft of slides for PhD-Seminar
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2053 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: be6eb4965a7015ec7f3014964e62ad0fce63684d
slides for PhD thesis defense
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3011 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2ac42af760b377bd2d56a4a05978e1eca5f1f0e9
small corrections to adapt for changes and updates in the disttr

@@
 %% includes a replacement for \usepackage{Sweave}
+%\usepackage{Sweave}
 \usepackage{amsmath,tikz}
 \usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
-%\usepackage[style=authoryear, backend=bibtex]{biblatex}
 \graphicspath{{plots/}}
@@
 transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)
-
-## define distribution list:
-# dist_list_normal
-{
-
-  dist_list_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {
-
-    val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
-    if(!log) val <- exp(val)
-
-    # par <- c(eta[1], exp(eta[2]))
-    # val <- dnorm(y, mean = par[1], sd = par[2], log = log)
-
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE) {
-
-    score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
-                   -1 + exp(-2*eta[2] + log((y-eta[1])^2)))
-
-    # par <- c(eta[1], exp(eta[2])) 
-    # score <- cbind(1/par[2]^2 * (y-par[1]), 
-    #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))
-
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y))
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
-    d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-eta[1]) * exp(-2*eta[2]), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    d2ld.etasigma2 <- sum(weights * (-2)*exp(log((y-eta[1])^2) - 2*eta[2]), na.rm = TRUE)
-
-    # par <- c(eta[1], exp(eta[2]))                           
-    # d2ld.etamu2 <- sum(weights * rep.int(-1/par[2]^2, ny))
-    # d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-par[1])/par[2]^2), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
-    # d2ld.etasigma2 <- sum(weights * (-2)*(y-par[1])^2/par[2]^2, na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etamu.d.etasigma, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- pnorm
-  qdist <- qnorm
-  rdist <- rnorm
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    if(is.null(weights)) {
-      mu <- mean(y)
-      sigma <- sqrt(1/length(y) * sum((y - mu)^2))
-    } else {
-      mu <- weighted.mean(y, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (y - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- TRUE
-
-  dist_list_normal <- list(family.name = "Normal Distribution",
-                           ddist = ddist, 
-                           sdist = sdist, 
-                           hdist = hdist,
-                           pdist = pdist,
-                           qdist = qdist,
-                           rdist = rdist,
-                           link = link, 
-                           linkfun = linkfun, 
-                           linkinv = linkinv, 
-                           linkinvdr = linkinvdr,
-                           startfun = startfun,
-                           mle = mle
-  )
-}
-
-# dist_list_cens_normal
-{
-
-  dist_list_cens_normal <- list()
-
-  parnames <- c("mu", "sigma")
-  etanames <- c("mu", "log(sigma)")
-
-  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE, left = 0, right = Inf) {     
-    par <- c(eta[1], exp(eta[2]))
-    val <- crch::dcnorm(x = y, mean = par[1], sd = par[2], left = left, right = right, log = log)
-    if(sum) {
-      if(is.null(weights)) weights <- if(is.matrix(y)) rep.int(1, dim(y)[1]) else rep.int(1, length(y))
-      val <- sum(weights * val, na.rm = TRUE)
-    }
-    return(val)
-  }
-
-  sdist <- function(y, eta, weights = NULL, sum = FALSE, left = 0, right = Inf) {   
-    par <- c(eta[1], exp(eta[2]))
-    # y[y==0] <- 1e-323
-
-    score_m <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    score_s <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right) * exp(eta[2]) # inner derivation exp(eta[2])
-    score <- cbind(score_m, score_s)
-    score <- as.matrix(score)
-    colnames(score) <- etanames
-    if(sum) {
-      if(is.null(weights)) weights <- rep.int(1, length(y)[1])
-      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN (0 in weights)
-      score[score==Inf] = 1.7e308
-      score <- colSums(weights * score, na.rm = TRUE)
-      #if(any(is.nan(score))) print(c(eta, "y", y))
-    }
-    return(score)
-  }
-
-  hdist <- function(y, eta, weights = NULL, left = 0, right = Inf) {    
-    ny <- length(y)
-    if(is.null(weights)) weights <- rep.int(1, ny)
-
-    par <- c(eta[1], exp(eta[2]))                           
-    # y[y==0] <- 1e-323
-
-    d2mu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu", left = left, right = right)
-    d2sigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-    dmudsigma <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "mu.sigma", left = left, right = right) # FIX: order?
-    dsigmadmu <- crch:::hcnorm(x = y, mean = par[1], sd = par[2], which = "sigma.mu", left = left, right = right) # FIX: order?
-    dsigma <- crch:::scnorm(x = y, mean = par[1], sd = par[2], which = "sigma", left = left, right = right)
-
-    d2ld.etamu2 <- sum(weights * d2mu, na.rm = TRUE)
-    d2ld.etamu.d.etasigma <- sum(weights * dmudsigma * par[2], na.rm = TRUE)
-    d2ld.etasigma.d.etamu <- sum(weights * dsigmadmu * par[2], na.rm = TRUE)
-    d2ld.etasigma2 <- sum(weights * (d2sigma * exp(2*eta[2]) + dsigma * par[2]), na.rm = TRUE)
-
-    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etasigma.d.etamu, d2ld.etasigma2), nrow = 2)
-    colnames(hess) <- rownames(hess) <-  etanames
-
-    return(hess)
-  }
-
-  ## additional functions pdist, qdist, rdist
-  pdist <- function(q, eta, lower.tail = TRUE, log.p = FALSE) crch:::pcnorm(q, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  qdist <- function(p, eta, lower.tail = TRUE, log.p = FALSE) crch:::qcnorm(p, mean = eta[1], sd = eta[2], 
-                                                                            lower.tail = lower.tail, log.p = log.p, 
-                                                                            left = left, right = right)
-  rdist <- function(n, eta) crch:::rcnorm(n, mean = eta[1], sd = eta[2], left = left, right = right)
-
-  link <- c("identity", "log")
-
-  linkfun <- function(par) {
-    eta <- c(par[1], log(par[2]))
-    names(eta) <- etanames
-    return(eta)
-  }
-
-  linkinv <- function(eta) {
-    par <- c(eta[1], exp(eta[2]))
-    names(par) <- parnames
-    return(par)
-  }
-
-  linkinvdr <- function(eta) {
-    dpardeta <- c(1, exp(eta[2]))
-    names(dpardeta) <- parnames
-    return(dpardeta)
-  }
-
-  startfun <- function(y, weights = NULL){
-    yc <- pmax(0,y)  # optional ?
-    if(is.null(weights)) {
-      mu <- mean(yc)
-      sigma <- sqrt(1/length(yc) * sum((yc - mu)^2))
-    } else {
-      mu <- weighted.mean(yc, weights)
-      sigma <- sqrt(1/sum(weights) * sum(weights * (yc - mu)^2))
-    }
-    starteta <- c(mu, log(sigma))
-    names(starteta) <- etanames
-    return(starteta)
-  }
-
-  mle <- FALSE
-
-  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
-                                ddist = ddist, 
-                                sdist = sdist, 
-                                hdist = hdist,
-                                pdist = pdist,
-                                qdist = qdist,
-                                rdist = rdist,
-                                link = link, 
-                                linkfun = linkfun, 
-                                linkinv = linkinv, 
-                                linkinvdr = linkinvdr,
-                                startfun = startfun,
-                                mle = mle
-  )
-}
 
@@
 \endminipage
-\hspace{1.1cm}
+\hspace{0.95cm}
 \minipage{0.25\textwidth}
@@
 \endminipage
-\hspace{0.65cm}
+\hspace{0.5cm}
 \minipage{0.29\textwidth}
@@
 \endminipage
-\hspace{0.65cm}
+\hspace{0.5cm}
 \minipage{0.28\textwidth}
@@
 learndata <- dgp(300, family = dist_list_normal, round.sp = 4, fun = fun)
-control <-  ctree_control(teststat = "quad", testtype = "Bonferroni", mincriterion = 0.7, minbucket = 20L)
-dt <- disttree(y~x1, data=learndata, family=dist_list_normal, type.tree = "ctree", control = control)
+control <-  disttree_control(type.tree = "ctree", 
+                             teststat = "quad", testtype = "Bonferroni", 
+                             mincriterion = 0.7, minbucket = 20L)
+dt <- disttree(y~x1, data=learndata, family = dist_list_normal, control = control)
+dt$fitted.par <- predict(dt, type = "parameter")
 
@@
 } else {
-  control <-  ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 7L)
-  df <- distforest(y~x1, data=learndata, family=dist_list_normal, ntree = 500L, type.tree = "ctree", control = control)
+  control <-  disttree_control(type.tree = "ctree",
+                               teststat = "quad", testtype = "Univ",
+                               mincriterion = 0, minbucket = 7L)
+  df <- distforest(y~x1, data=learndata, family = dist_list_normal, ntree = 500L, control = control)
+  df$fitted.par <- predict(df, type = "parameter")
   df$info <- df$fitted <- df$weights <- df$nodes <- df$data <- df$terms <- df$loglik <- df$trafo <- NULL

### 2020-04-26 pkg/disttree/inst/JSM2018/slides.Rnw
slides for PhD thesis defense
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3011 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2ac42af760b377bd2d56a4a05978e1eca5f1f0e9
small corrections to adapt for changes and updates in the disttree implementation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3012 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5622db15384bfba47becde3a04582ef1bfc97701

@@
 %% includes a replacement for \usepackage{Sweave}
+% \usepackage{Sweave}
 \usepackage{amsmath,tikz}
@@
 \visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 \visible<4->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 \visible<3->{{\LARGE$\rightarrow$}}
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
   \visible<3->{
@@
   \begin{tikzpicture}
-  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
-  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
+  \draw[line width=0.2mm, black] (0,0) rectangle (2.44,2.12);
+  \node[rectangle, align=center] (sq) at (1.22, 1.06) {\Huge{?}};
   \end{tikzpicture}
@@
 
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 \hspace{0.65cm}
-\minipage{0.29\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 \hspace{0.65cm}
-\minipage{0.28\textwidth}
+\minipage{0.285\textwidth}
 \begin{center}
@@
 df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
-                 type.tree = "ctree", ntree = 100, censtype = "left", censpoint = 0,
-                 control = ctree_control(teststat = "quad", testtype = "Univ",
+                 ntree = 100, censtype = "left", censpoint = 0,
+                 control = disttree_control(type.tree = "ctree",
+                                            teststat = "quad", testtype = "Univ",
                                             intersplit = TRUE,
@@
 %\begin{itemize}
-%\item Prespecified GAM: variable selection based on meteorological expert knowledge.
-%\item Boosted GAM: automatic variable selection.
+%\item Prespecified GAMLSS: variable selection based on meteorological expert knowledge.
+%\item Boosted GAMLSS: automatic variable selection.
 %\item Ensemble Model Output Statistics (EMOS): linear predictors using only total precipitation predictions as covariates.
@@
 \vspace{-0.15cm}
-\textbf{Prespecified GAM} & spline  & based on expert \\
+\textbf{Prespecified GAMLSS} & spline  & based on expert \\
                           & in each & knowledge      \\
@@
 \vspace{-0.15cm}
-\textbf{Boosted GAM} & spline  & automatic  \\
+\textbf{Boosted GAMLSS} & spline  & automatic  \\
                      & in each &            \\
@@
 %\vspace{-0.15cm}
-%\textbf{prespecified GAM} & spline  & based on expert \\
+%\textbf{Prespecified GAMLSS} & spline  & based on expert \\
 %                          & in each & knowledge      \\
@@
 %\vspace{-0.15cm}
-%\textbf{boosted GAM} & spline  & automatic  \\
+%\textbf{Boosted GAMLSS} & spline  & automatic  \\
 %                     & in each &            \\

### 2020-04-26 pkg/disttree/inst/MIP-Seminar/slides.Rnw
slides for PhD thesis defense
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3011 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2ac42af760b377bd2d56a4a05978e1eca5f1f0e9
small corrections to adapt for changes and updates in the disttree implementation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3012 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5622db15384bfba47becde3a04582ef1bfc97701

@@
 %% includes a replacement for \usepackage{Sweave}
+\usepackage{Sweave}
 \usepackage{changepage}
@@
                  type.tree = "ctree", ntree = 100, censtype = "left", censpoint = 0,
-                 control = ctree_control(teststat = "quad", testtype = "Univ",
+                 control = disttree_control(type.tree = "ctree", 
+                                            teststat = "quad", testtype = "Univ",
                                             intersplit = TRUE,

### 2020-04-26 pkg/disttree/inst/eeecon_workshop/slides.Rnw
slides for PhD thesis defense
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3011 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2ac42af760b377bd2d56a4a05978e1eca5f1f0e9
small corrections to adapt for changes and updates in the disttree implementation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3012 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5622db15384bfba47becde3a04582ef1bfc97701

@@
 \visible<6->{
-\textbf{Application/Motivation:}
+\textbf{Application / Motivation:}
 \begin{itemize}
@@
 \begin{itemize}
-\item 95 observation stations in Tyrol, Austria.\\
-\item Daily 24h precipitation sums (July, 1985--2012).
-%\item Daily 24h precipitation sums from July over 28 years (1985--2012).
+\item Daily 24h precipitation sums (July, 1985--2012).\\
+\item 95 observation stations in Tyrol, Austria.
 \end{itemize}
@@
 df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
-                 type.tree = "ctree", ntree = 100, censtype = "left", censpoint = 0,
-                 control = ctree_control(teststat = "quad", testtype = "Univ",
+                 ntree = 100, censtype = "left", censpoint = 0,
+                 control = disttree_control(type.tree = "ctree", ,
+                                            teststat = "quad", testtype = "Univ",
                                             intersplit = TRUE,

### 2020-06-14 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added explanations/comments based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3017 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4c2b1bea2e4ed05de981fa68b5503c7ea74db959

@@
 %
-Therefore, in this paper the focus will be put on that first step of tree algorithms,
+\fixme{Focus on testing strategies}
+
+Therefore, \hlc{instead of comparing the complete algorithms,} in this paper the focus will be put on that first step of tree algorithms,
 i.e., the task of selecting the best split variable in a given (sub)sample. 
@@
 
+%% FIXME
+\fixme{Provided theory, finite samples}
+
+\hlc{As the underlying theory of the different testing strategies and the resulting properties and skills but also limitations are already discussed in previous literature} \citep[see, e.g.,][]{Ploberger+Kraemer:1992, Hothorn+Hornik+Zeileis:2006, Zeileis+Hothorn+Hornik:2008, Loh:2002}, \hlc{the presented simulation study aims at investigating the corresponding consequences in finite samples.}
 In many of the considered scenarios the choice of goodness-of-fit measure heavily influences
@@
 
+\fixme{again, focus on testing strategies}
+
+\hlc{For both scenarios the employed testing framework and the considered evaluations allow for a direct comparison of the testing strategies only opposed to the complete tree algorithms.}
+
 %\newpage
@@
 linear model, e.g., in \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002}, and \citealp{Loh+He+Man:2015}.
+
+\fixme{Outlook: non-tree structure}
+
+\hlc{To exploit the flexibility of tree models future investigations could also include situations where the underlying data generating process does not follow a (clear) tree structure.
+In this case, assessing the effects of different building blocks in the testing strategies should in general show similar results, however, they might vary depending on how well the true functional relationships can be approximated by a tree structure.}
+
+\fixme{Reference on Fu and Simonoff (2015)}
+
+As \cite{Fu+Simonoff:2015} \hlc{show, unbiased recursive partitioning algorithms can have a bias on their own.
+In particular, given that there are two possible split variables showing the same significant association to the (transformed) response the algorithms tend to select the one providing less possible split points.
+Therefore, this could be an additional specific setting in which the power of testing strategies and their building blocks could be compared.}
 

### 2020-06-29 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
material for introductory course on trees and forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3018 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 388e124f82296b07a34c1e60b134613ce0d783e6

\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
\usetheme[nototalframenumber,url,license]{uibk}

\title{Random Forests}
\author{Achim Zeileis, Lisa Schlosser}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
%\URL{http://www.partykit.org/partykit/}
\renewcommand{\headerimage}[1]{%
   \IfStrEqCase{#1}{%
      {1}{%
         \gdef\myheaderimageid{#1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{forest.jpg}%
      }}[%
         \gdef\myheaderimageid{1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{forest.jpg}%
      ]%
}
\headerimage{1}

\usepackage[utf8]{inputenc}

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
%\usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usepackage{calc}
\usepackage{graphicx}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
\usepackage{xcolor}
%\usepackage[cal=boondoxo]{mathalfa}
%\graphicspath{{plots/}}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
\definecolor{forestred}{RGB}{206,73,81}
\definecolor{treegreen}{RGB}{0,143,0}
\definecolor{lightblue}{RGB}{34,151,230}
\definecolor{lightorange}{RGB}{255,165,0}

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)

set.seed(7)

invisible(.Call(grDevices:::C_palette, grDevices::hcl(
  h = c(0,   5, 125, 245, 195, 315,  65,   0),
  c = c(0, 100,  90,  85,  63, 105,  90,   0),
  l = c(0,  55,  75,  60,  82,  48,  80,  65)
)))

library("disttree")
library("Formula")
library("latex2exp")
library("lattice")
library("MASS")
library(colorspace)

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)

lightblue <- "#2297E6"
lightorange <- "#FFA500"
@

<<motivation_tree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
nobs <- 200
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- numeric(length = length(x))
for(i in 1:(nobs)) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
# more points for precise illustration of true function
x100 <- c(1:(100*nobs))/(100*nobs)
ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- yforest <- numeric(length = length(x100))
for(i in 1:(nobs*100)) ytrue[i] <- if(x100[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x100[i]-0.2)-1)))
for(i in 1:(nobs*100)) ytree[i] <- if(x100[i]<1/3) 0.5 else {if(x100[i]<2/3) 2 else 1}
for(i in 1:(nobs*100)) ytree2[i] <- if(x100[i]<0.31) 0.47 else {if(x100[i]<0.69) 1.9 else 1.1}
for(i in 1:(nobs*100)) ytree3[i] <- if(x100[i]<0.34) 0.52 else {if(x100[i]<0.54) 2.2 else {if(x100[i]<0.74) 1.74 else 0.9}}
for(i in 1:(nobs*100)) ytree4[i] <- if(x100[i]<0.32) 0.51 else {if(x100[i]<0.63) 1.8 else 1.15}
for(i in 1:(nobs*100)) yforest[i] <- if(x100[i]<=0.33) 0.52 else {if(x100[i]<=0.34) 1.5 else {if(x100[i]<=0.54) 2 else {if(x100[i]<=0.63) 1.9 else {if(x100[i]<=0.69) 1.6 else {if(x100[i]<=0.74) 1.3 else 1}}}}}
@

\begin{document}

\section{Forests models}

\subsection{Motivation: Forest models}
\begin{frame}[fragile]
\frametitle{Motivation: Forest models}

{\bf Idea:} \\
\begin{itemize}
\item Combination of an ensemble of trees. 
\item Each tree can capture non-linear and non-additive effects and select covariates and possible interactions automatically.
\item Combining trees to a forest model allows for an approximation of smooth effects.
\item A forest can regularize and stabilize the model.
\end{itemize}

%\bigskip

\end{frame}

\begin{frame}
\frametitle{Motivation: Forest models}

\vspace{0.2cm}

\begin{minipage}{0.4\textwidth}
\only<1>{\textbf{Tree model:\\}}\only<2-8>{\textbf{Subsampling:\\}}\only<9->{\textbf{Forest model:\\}}
\vspace{-0.4cm}
%\vspace{-0.8cm}
\begin{center}
\only<1>{
%\vspace{0.2cm}
\resizebox{0.63\textwidth}{!}{
\begin{tikzpicture}
  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (2, 3) {};
  \node[] (n00) at (1, 1.5) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0, 0) {$\hat{Y}_1$};
  \draw[-] (n0) -- (n00) node [midway, left] {\small $X \leq p_1$};
  \draw[-] (n0) -- (n1);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (3, 1.5) {};
  \draw[-] (n0) -- (n2) node [midway, right] {\small $X > p_1$};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (2, 0) {$\hat{Y}_2$};
  \draw[-] (n2) -- (n3) node [midway, left] {\small $X \leq p_2$};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (4, 0) {$\hat{Y}_3$};
  \draw[-] (n2) -- (n4) node [midway, right] {\small $X > p_2$};
\end{tikzpicture}}

\vspace{0.475cm}

}
\only<2->{
%\vspace{0.2cm}
\resizebox{0.6\textwidth}{!}{
\begin{tikzpicture}
%\draw[ellipse, draw=black, align=center, scale = 0.7, minimum width=170pt, minimum height = 35pt] (set) at (1, 3) {};
\draw[gray,rounded corners=10pt] (-1.05,4.35) -- (4.05,4.35) -- (4.05,2.7) -- (-2.05,2.7) -- (-2.05,4.35) -- (0.05,4.35);
%\draw[gray] (1,3.5) ellipse (3 and 0.8);
%\visible<4>{
\begin{scope}
\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\clip (1,3.5) ellipse (3 and 0.8);
\pgfmathsetseed{7}
\foreach \p in {1,...,200} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
\end{scope}
%}
%%% color points
%\visible<5-6>{
%\begin{scope}
%\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\pgfmathsetseed{7}
%\foreach \p in {1,...,100} {\fill[treegreen] (1+3*rand,3.5+0.8*rand) circle (0.07);}
%\end{scope}
%}
\visible<3->{
\node[ellipse, draw=treegreen, align=center, minimum width=80pt, minimum height = 40pt, line width = 3pt] (subset1) at (0.1, 3.5) {};
}
%\visible<7-8>{
%\begin{scope}
%\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\pgfmathsetseed{7}
%\foreach \p in {1,...,49} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
%\foreach \p in {50,...,150} {\fill[lightblue] (1+3*rand,3.5+0.8*rand) circle (0.07);}
%\end{scope}
%}
\visible<5->{
\node[ellipse, draw = lightblue, align=center, minimum width=100pt, minimum height = 30pt, line width = 3pt] (subset2) at (1.2, 3.6) {};
}
%\visible<9-10>{
%\begin{scope}
%\clip (-1,4.3) -- (4,4.3) -- (4,2.7) -- (-2,2.7) -- (-2,4.3) -- (0,4.3);
%\pgfmathsetseed{7}
%\foreach \p in {1,...,99} {\fill[gray] (1+3*rand,3.5+0.8*rand) circle (0.04);}
%\foreach \p in {100,...,200} {\fill[lightorange] (1+3*rand,3.5+0.8*rand) circle (0.07);}
%\end{scope}
%}
\visible<7->{
\node[ellipse, draw = lightorange, align=center, minimum width=120pt, minimum height = 25pt, line width = 3pt] (subset3) at (1.6, 3.3) {};
}
\visible<4->{
\node[ellipse, fill=treegreen, align=center] (n00) at (-1, 2) {};
\node[ellipse, fill=treegreen, align=center] (n01) at (-1.25, 1.25) {};
\draw[-, line width=1pt] (n00) -- (n01);
\node[rectangle, fill=treegreen, align=center] (n02) at (-1.5, 0.5) {};
\draw[-, line width=1pt] (n01) -- (n02);
\node[rectangle, fill=treegreen, align=center] (n03) at (-1, 0.5) {};
\draw[-, line width=1pt] (n01) -- (n03);
\node[rectangle, fill=treegreen, align=center] (n04) at (-0.5, 0.5) {};
\draw[-, line width=1pt] (n00) -- (n04);
}
\visible<6->{
\node[ellipse, fill=lightblue, align=center] (n10) at (1, 2) {};
\node[ellipse, fill=lightblue, align=center] (n11) at (0.5, 1.25) {};
\draw[-, line width=1pt] (n10) -- (n11);
\node[rectangle, fill=lightblue, align=center] (n12) at (0.25, 0.5) {};
\draw[-, line width=1pt] (n11) -- (n12);
\node[rectangle, fill=lightblue, align=center] (n13) at (0.75, 0.5) {};
\draw[-, line width=1pt] (n11) -- (n13);
\node[ellipse, fill=lightblue, align=center] (n14) at (1.5, 1.25) {};
\draw[-, line width=1pt] (n10) -- (n14);
\node[rectangle, fill=lightblue, align=center] (n15) at (1.75, 0.5) {};
\draw[-, line width=1pt] (n14) -- (n15);
\node[rectangle, fill=lightblue, align=center] (n16) at (1.25, 0.5) {};
\draw[-, line width=1pt] (n14) -- (n16);
}
\visible<8->{
\node[ellipse, fill=lightorange, align=center] (n20) at (3, 2) {};
\node[rectangle, fill=lightorange, align=center] (n21) at (2.5, 0.5) {};
\draw[-, line width=1pt] (n20) -- (n21);
\node[ellipse, fill=lightorange, align=center] (n22) at (3.25, 1.25) {};
\draw[-, line width=1pt] (n20) -- (n22);
\node[rectangle, fill=lightorange, align=center] (n23) at (3, 0.5) {};
\draw[-, line width=1pt] (n22) -- (n23);
\node[rectangle, fill=lightorange, align=center] (n24) at (3.5, 0.5) {};
\draw[-, line width=1pt] (n22) -- (n24);
}
\visible<10->{
\node[ellipse, draw = black, align=center, minimum width=15pt, minimum height = 15pt, line width = 2pt] (B1) at (-1.5, 0.5) {};
\node[ellipse, align=center, line width = 3pt, text=treegreen] (Y1) at (-1.5, -0.5) {\LARGE $\hat{y}_1$};
}
\visible<11->{
\node[ellipse, draw = black, align=center, minimum width=15pt, minimum height = 15pt, line width = 2pt] (B2) at (1.25, 0.5) {};
\node[ellipse, align=center, line width = 3pt, text=lightblue] (Y2) at (1.25, -0.5) {\LARGE $\hat{y}_2$};
}
\visible<12->{
\node[ellipse, draw = black, align=center, minimum width=15pt, minimum height = 15pt, line width = 2pt] (B2) at (3.5, 0.5) {};
\node[ellipse, align=center, line width = 3pt, text=lightorange] (Y3) at (3.5, -0.5) {\LARGE $\hat{y}_3$};(3.5, 0.5)
}
\end{tikzpicture}
}
}
\end{center}
\end{minipage}
\hspace{0.7cm}
\begin{minipage}{0.36\textwidth}
%\begin{center}
%\vspace{0.1cm}
\only<1-2>{
%\vspace{0.01cm}
<<plot_motivation_tree, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(3,3,2,0))
#par(mar=c(3,3,2,5.5))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x100, y = ytree, col = pal["forest"], lwd=7)
mtext(text = "X", side = 1, cex = 2.5, line = 2)  
mtext(text = "Y", side = 2, cex = 2.5, line = 1)  
@
}
\only<3>{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest_data, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(3,3,2,0))
#par(mar=c(3,3,2,5.5))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
mtext(text = "X", side = 1, cex = 2.5, line = 2)  
mtext(text = "Y", side = 2, cex = 2.5, line = 1)  
@
}
\only<4-5>{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest1, fig=TRUE, echo=FALSE, width=7>>=
<<plot_motivation_randforest_data>>
lines(x = x100, y = ytree2, col = pal["tree"], lwd=7)
@
}
\only<6-7>{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest2, fig=TRUE, echo=FALSE, width=7>>=
<<plot_motivation_randforest1>>
lines(x = x100, y = ytree3, col = lightblue, lwd=7)
@
}
\only<8>{
%\vspace{0.01cm}
%\hspace{-0.1cm}
<<plot_motivation_randforest3, fig=TRUE, echo=FALSE, width=7>>=
<<plot_motivation_randforest2>>
lines(x = x100, y = ytree4, col = lightorange, lwd=7)
@
}
\only<9-12>{
<<plot_motivation_randforest_x, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(3,3,2,0))
#par(mar=c(3,3,2,5.5))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x100, y = ytree2, col = pal["tree"], lwd=7)
lines(x = x100, y = ytree3, col = lightblue, lwd=7)
lines(x = x100, y = ytree4, col = lightorange, lwd=7)
lines(x = c(0.6,0.6), y = c(-1, 1.8), type = "l", lty = 2, lwd = 4)
mtext(text = "Y", side = 2, cex = 2.5, line = 1)  
mtext(text = "x", side = 1, cex = 4, line = 2, at=0.6)  
@
}
\only<13->{
<<plot_motivation_randforest4, fig=TRUE, echo=FALSE, width=7>>=
par(mar=c(3,3,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x100, y = yforest, col = pal["forest"], lwd=7)
lines(x = c(0.6,0.6), y = c(-1, 1.8), type = "l", lty = 2, lwd = 4)
mtext(text = "x", side = 1, cex = 4, line = 2, at=0.6)  
mtext(text = "Y", side = 2, cex = 2.5, line = 1)  
@
}
\end{minipage}

\vspace{0.5cm}

\only<13->{
$\hat{y}= \frac{1}{3} (\hat{y}_1 + \hat{y}_2 + \hat{y}_3)$
}

\vspace{0.3cm}

\visible<13->{
for $\hat{y}_t$ being the average response value in the corresponding segment $\mathcal{B}^t_x$ of the $t$-th tree for learning data $\{y_i\}_{i=1,\ldots,n}$:

\vspace{0.3cm}

$\hat{y}_t= \frac{1}{|\mathcal{B}^t_x|} \sum_{i: y_i \in \mathcal{B}^t_x} y_i$
\visible<14->{
$\ =\ \sum_{i=1}^n \frac{I(x_i \in \mathcal{B}^t_x)}{|\mathcal{B}^t_x|} \cdot y_i$
}
\visible<15->{
$\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$
}
}
\end{frame}

\subsection{Aggregation of trees}

\begin{frame}
\frametitle{Aggregation of trees}
{\bf Strategy:} Combine trees to a forest model by employing tree-based weights.

The predicted response $\hat{y}$ or, more generally, the predicted model parameter (vector) $\hat{\theta}$ is the argument of the parameter space $\Theta$ that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{y_i\}_{i=1,\ldots,n}$:

\vspace{-0.2cm}

\[
\hat{\theta}(x) =  \argmin_{\theta \in \Theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)
\]

\medskip

\only<2-4>{

\vspace{0.13cm}

\textbf{Weights:}

\vspace{-1.14cm}

\begin{eqnarray*}
w^{\text{base}}_i(x)   & = & 1 \\[0.2cm]
\visible<3-4>{
w^{\text{tree}}_i(x)   & = & \sum_{b=1}^B I((x_i \in \mathcal{B}_b) \land (x \in \mathcal{B}_b)) \\[0.1cm]
\visible<4>{
w^{\text{forest}}_i(x) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} \frac{I((x_i \in \mathcal{B}^t_b) \land (x \in \mathcal{B}^t_b))}{\left|\mathcal{B}^t_b\right|}
\end{eqnarray*}
}}}

\only<5->{
%\vspace{-3cm}
{\bf Note:} This general notation holds for classic random forests averaging over mean values of tree nodes by setting $f(\theta, y_i) = (\theta - y_i)^2$\\
but also allows for combining quantiles (quantile regression forests) or full distributions as in distributional forests with
$f(\theta, y_i) = log(likelihood(\theta, y_i))$.\\
}

\end{frame}

\subsection{Subsampling}

\begin{frame}
\frametitle{Subsampling}
{\bf Idea:} Build each tree on a different subset of the full data set.

\medskip

{\bf Types of subsampling:}
\begin{itemize}
\item Subsets (recommended size: 63.2\% of the full data set)
\item Bootstrapping (resampling with replacement)
\end{itemize}

\bigskip

{\bf Random input variable sampling:} In order to reduce the correlation among trees only a subset of possible split variables is provided for each split (recommended size: $\sqrt{\text{\#Var}}$).

\bigskip

{\bf Note:} Each tree is built as large as possible which can lead to overfitting for the single tree. However, this is compensated by aggregating the trees to a forest model.

\end{frame}

\subsection{Software}

\begin{frame}[fragile]
\frametitle{R software for forest models}

{\bf randomForest:}
\begin{itemize}
  \item R package for classic random forests.
  \item Based on implementations of CART.
\end{itemize}

\medskip

<<call_randomForest, eval=FALSE, echo=TRUE>>=
randomForest(formula, data=NULL, ..., subset, na.action=na.fail)
randomForest(x, y=NULL,  ...)
@

\medskip

Optional control arguments:
\begin{itemize}
\item \code{ntree}: nr. of trees, 
\item \code{mtry}: nr. of randomly sampled split variable candidates, 
\item \code{replace}: type of subsetting, 
\item \code{sampsize}: size of drawn subsamples, 
\item \code{nodesize}: minimum size of terminal nodes, 
\item \code{maxnodes}: maximum number of terminal nodes, 
\item \ldots 
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{R software for forest models}

{\bf partykit:}
\begin{itemize}
  \item R package offering a toolkit for unbiased recursive partitioning. 
  \item Provides the forest-building function \fct{cforest}, based on conditional inference trees.
\end{itemize}

\medskip

<<call_cforest, eval = FALSE>>=
cforest(formula, data, subset, ...)
@

\medskip

\only<1>{
Optional forest-specific control arguments:
\begin{itemize}
\item \code{ytrafo}: transformation function applied to the response,
\item \code{ntree}: nr. of trees, 
\item \code{mtry}: nr. of randomly sampled split variable candidates, 
\item \code{perturb}: type and size of subsetting, 
\item \ldots 
\end{itemize}
\vspace{0.9cm}
}
\only<2>{
Optional tree-specific arguments handed over via \fct{ctree\_control}:
\begin{itemize}
\item \code{minsplit}: minimum nr. of observations to perform a split, 
\item \code{minbucket}: minimum nr. of observations in a node after splitting, 
\item \code{maxdepth}: maximum depth of the tree, 
\item \code{alpha}: significance level for split variable selection,
\item \ldots 
\end{itemize}
\vspace{0.9cm}
}
\end{frame}

\begin{frame}[fragile]
\frametitle{R software for forest models}

{\bf Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.

\bigskip

<<call_predict, eval=FALSE, echo=TRUE>>=
predict(object, newdata, type = "response", ...)
@

\end{frame}

\begin{frame}
\frametitle{R software for forest models}

\textbf{Further R packages:}
%
\begin{itemize}
  \item ranger (classic random forests),
  \item grf (generalized random forests),
  \item disttree (distributional forests, based on partykit),
  \item \ldots
\end{itemize}

\end{frame}

\subsection{Variable importance}

\begin{frame}
\frametitle{Variable importance}
{\bf Idea:} Asses the influence of a covariate by evaluating the change in performance of the model (mean decrease in accuracy) after dropping or permutating the selected covariate.

\bigskip

Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, nr. of misclassifications, \ldots).

\bigskip

{\bf Implementations:}
\begin{itemize}
\item \fct{importance} in the R package randomForest.
\item \fct{importance.ranger} in the R package ranger.
\item \fct{varimp} in the R package partykit.
\end{itemize}

\end{frame}

\subsection{References}

\begin{frame}
\frametitle{References}

\vspace{-0.2cm}

\scriptsize

Breiman L (2001).
  \dquote{Random {Forests}.}
  \emph{Machine Learning}, 
  \textbf{45}(1), 5--32.
  \doi{10.1023/A:1010933404324}

\medskip

Breiman L, Cutler A (2004).
 \dquote{Random Forests.},
 \url{https://www.stat.berkeley.edu/~breiman/RandomForests}
 (Accessed: 2018-02-22)

\medskip

%Hothorn T, Zeileis A (2017).
%  \dquote{Transformation Forests.}
%  \emph{arXiv 1701.02110}, arXiv.org E-Print Archive.
%  \url{http://arxiv.org/abs/1701.02110}
%  
%\smallskip

Hothorn T, Hornik K, Zeileis A (2006).
 \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
 \emph{Journal of Computational and Graphical Statistics},
 \textbf{15}(3), 651--674.
 \doi{10.1198/106186006X133933}

\medskip

Hothorn T, Zeileis A (2015).
 \dquote{{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
 \emph{Journal of Machine Learning Research},
 \textbf{16}, 3905--3909.
 \url{http://www.jmlr.org/papers/v16/hothorn15a.html}

\medskip

Liaw A, Wiener M (2002).
 \dquote{Classification and Regression by randomForest.}
 \emph{R News},
 \textbf{2}(3), 18--22.
 \url{https://CRAN.R-project.org/doc/Rnews/}

\medskip

% Zeileis A, Hothorn T, Hornik K (2008).
%  \dquote{Model-Based Recursive Partitioning.}
%   \emph{Journal of Computational and Graphical Statistics},
%   \textbf{17}(2), 492--514.
%   \doi{10.1198/106186008X319331}
% 
% \medskip

Athey S, Tibshirani J, Wager S (2019). 
 \dquote{Generalized Random Forests.} 
 \emph{Annals of Statistics}, \textbf{47}{2}, 1148--1178.
 \doi{10.1214/18-AOS1709"}

\medskip

Schlosser L, Hothorn T, Stauffer R, Zeileis A (2019).
\dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
\emph{The Annals of Applied Statistics}, \textbf{13}(3), 1564--1589.
\doi{10.1214/19-AOAS1247}

\medskip

Wright M N, Ziegler A (2017).
\dquote{{ranger}: A Fast Implementation of Random Forests for High
Dimensional Data in {C++} and {R}.}
\emph{Journal of Statistical Software}, \textbf{77}{1}, 1--17.
\doi{10.18637/jss.v077.i01}

\end{frame}

\end{document}

### 2020-06-29 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
material for introductory course on trees and forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3018 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 388e124f82296b07a34c1e60b134613ce0d783e6

\documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
\usetheme[nototalframenumber,url,license]{uibk}

\title{Tree models}
\subtitle{Supervised Learning: Algorithmic Modeling}
\author{Achim Zeileis, Lisa Schlosser}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
%\URL{http://www.partykit.org/partykit/}
\renewcommand{\headerimage}[1]{%
   \IfStrEqCase{#1}{%
      {1}{%
         \gdef\myheaderimageid{#1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{forest.jpg}%
      }}[%
         \gdef\myheaderimageid{1}%
         \gdef\myheaderimageposition{nw}%
         \gdef\myheaderimage{forest.jpg}%
      ]%
}
\headerimage{1}

%\graphicspath{{Figures/}}
\usepackage[utf8]{inputenc}

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
% \usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usepackage{calc}
\usepackage{graphicx}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
\usepackage{color}
\usepackage{xcolor}
\usepackage{fancyvrb}
%\usepackage[cal=boondoxo]{mathalfa}
%\graphicspath{{plots/}}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

\setbeamertemplate{subsection page}
{\begin{center}
\vspace{2cm}
% {\Large \insertsection\par}
% \vspace{0.5cm}
Section~\thesubsection\par
\vspace{0.5cm}
{\LARGE \insertsubsection\par}
\end{center}}
\AtBeginSubsection{\frame{\subsectionpage}}

%% colors
% \definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
% \definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
% \definecolor{forestred}{RGB}{206,73,81}
% \definecolor{treegreen}{RGB}{0,143,0}
% \definecolor{lightblue}{RGB}{34,151,230}
% \definecolor{lightorange}{RGB}{255,165,0}

%\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)

set.seed(7)

invisible(.Call(grDevices:::C_palette, grDevices::hcl(
  h = c(0,   5, 125, 245, 195, 315,  65,   0),
  c = c(0, 100,  90,  85,  63, 105,  90,   0),
  l = c(0,  55,  75,  60,  82,  48,  80,  65)
)))

# setwd("~/Documents/ULG/slides_tree/")
library("disttree")
library("Formula")
library("latex2exp")
library("AER")
#library("gamlss")
#library("lattice")
#library("crch")
#library("RainTyrol")
#library("parallel")
library("rpart")
#library("gamlss.cens")
#gen.cens(NO, type = "left")
#library("ggplot2")
#theme_set(theme_bw(base_size = 18))
library("MASS")
#library("Ecdat")
#library("wooldridge")
library("partykit")
#library("RWeka")
#library("evtree")
library("coin")

## HCL palette
# pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
# names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
# 
# pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
# names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
# 
# transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
@

<<data>>=
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
# ttnc <- transform(ttnc, Treatment = factor(Gender == "Female" | Age == "Child", 
#                                            levels = c(FALSE, TRUE),
#                                            labels = c("Male&Adult", "Female|Child")))
##   labels = c("Normal\n(Male&Adult)", "Preferential\n(Female|Child)")
@

\begin{document}

\section{Tree models}

\subsection{Motivation}
%\subsectionpage

\begin{frame}[fragile]
\frametitle{Motivation}

\smallskip

{\bf Idea:} \\
\smallskip
\quad Separate data into smaller and more homogeneous subgroups \\
\quad based on information provided by covariates.
%in order to obtain better fitting \\
%\quad models in the resulting subsets.

\bigskip

\bigskip

{\bf Formally:} \\
\smallskip
\quad Modeling of a dependent variable $Y$ by ``learning'' a recursive partition \\
\quad w.r.t explanatory variables $X_1$, \dots, $X_m$.

\bigskip

\bigskip

{\bf Key features:} \\
\begin{itemize}
\item Predictive power in nonlinear regression relationships.
\item Interpretability (enhanced by visualization), i.e., no ``black box'' methods.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Titanic:  \quad $Survived \; \sim  \; Gender + Age + Class$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<ctree_Titanic, echo=FALSE, fig=TRUE, width=11>>=
ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
            control = ctree_control(alpha = 0.01))
plot(ct)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
{\bf Data:}
The Titanic data set provides information on the survival of 2201~passengers on the Titanic with the additional variables Class, Sex and Age.

\bigskip

The variables and their levels are as follows:

\bigskip

\begin{tabular}{l l l}
\hline 
No & Name &	Levels\\
\hline 
1 &	Class &	1st, 2nd, 3rd, Crew\\
2 &	Sex &	Male, Female\\
3 &	Age &	Child, Adult\\
4 &	Survived &	No, Yes \\
\hline 
\end{tabular}

\end{frame}
\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.99\textwidth}
<<ctree_Wages1985, echo=FALSE, fig=TRUE, width=13>>=
data("CPS1985")
plot(ctree(log(wage)~education+experience+age+ethnicity+gender+union,data=CPS1985,
           control = ctree_control(alpha = 0.01)))
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education \ | \ experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<lmtree_Wages1985, echo=FALSE, fig=TRUE, width=11>>=
data("CPS1985")
plot(lmtree(log(wage)~education|experience+age+ethnicity+gender+union,data=CPS1985))
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}

Wind direction forecasting\\
at Innsbruck airport

\vspace{-1.4cm}

\setkeys{Gin}{width=0.95\linewidth}
\begin{center}
\includegraphics{circtree_ibk.pdf}
\end{center}

\end{frame}

% \begin{frame}[fragile]
% \frametitle{Motivation}
% {\small Wage data:  \quad $log(wage) \; \sim  \; education + experience + ethnicity + region + parttime$}
% \begin{center}
% \setkeys{Gin}{width=0.99\textwidth}
% <<ctree_Wages1988, echo=FALSE, fig=TRUE, width=13>>=
% data("CPS1988")
% ## ctree
% plot(ctree(log(wage)~education+experience+ethnicity+region+parttime,data=CPS1988,
%            control = ctree_control(alpha = 0.01, minsplit = 7000)))
% @
% \end{center}
% \end{frame}

% \begin{frame}[fragile]
% \frametitle{Motivation}
% {\small Boston housing:  \quad $medv \; \sim  \; lstat + crim + rm + age + black$}
% \begin{center}
% \setkeys{Gin}{width=0.99\textwidth}
% <<ctree_Boston, echo=FALSE, fig=TRUE, width=13>>=
% data("Boston")
% plot(ctree(medv~lstat + crim + rm + age + black, 
%            data = Boston, control = ctree_control(alpha = 0.01, minsplit = 150)))
% @
% \end{center}
% \end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}

{\bf Homogeneity:} Partitioning the data into subgroups aims at allowing for a better model fit in each resulting subset.

\bigskip

{\bf Model:} The model fitted to a response $Y$ in each subgroup optimizes an objective function, i.e., it can be a simple constant fit (e.g. average value or proportion, minimizing deviations) but also a more complex model (possibly including additional regressors, and, e.g., minimizing the residual sum of squares or the negative log-likelihood).

\bigskip

{\bf Trade-off:} Complexity in the tree structure vs.\ complexity in the fitted models.\\
Simple models in the nodes (such as constant fits) often lead to large trees while more complex models can compensate a higher variation within the nodes allowing for less complex trees.

% \begin{itemize}
% \item Choose an indicator or a measure of homogeneity within subgroups.
% %% FIXME: discrepancy measure, loss function
% \item Find the split which optimizes this measure in the resulting subgroups.
% \end{itemize}
% 
% \bigskip
% 
% \bigskip
% 
% {\bf Examples of discrepancy measures:} 
% \begin{itemize}
% \item For regression tree (numeric response): Deviations from an average value, residual sum of squares.
% \item For classification tree (categorical response): Number of misclassifications.
% \end{itemize}
 \end{frame}

\subsection{Tree algorithm}

\begin{frame}[fragile]
\frametitle{Tree algorithm}

{\bf Base algorithm:}
\begin{enumerate}
\item Fit a model to the response $Y$.
\item Assess association of $Y$ (or a corresponding transformation / goodness-of-fit measure) and each possible split variable $X_j$.
\item Split sample along the $X_{j^{\ast}}$ with strongest association: Choose breakpoint with highest improvement of the model fit.
\item Repeat steps 1--3 recursively in the subsamples until some stopping criterion is met.
\item Optionally: Reduce the size of the tree through post-pruning, i.e., by cutting off branches that do not improve the performance significantly.
\end{enumerate}

\end{frame}

\begin{frame}[fragile]
\frametitle{Tree algorithm: Building blocks}

{\bf Examples of possible transformations of the response $Y$:} 
\begin{itemize}
\item Numeric response: 
\begin{itemize}
\item Ranks of $Y$.
\item (Absolute) deviations $Y - \bar Y$.
\item Residual sum of squares $\sum (Y - \hat{Y})^2$.
\end{itemize}
\item Categorical response: 
\begin{itemize}
\item Number of misclassifications (Gini impurity).
\end{itemize}
\end{itemize}

\medskip

{\bf Possible association tests:} 
$\chi^2$-test, Pearson correlation test, two-sample t-test, ANOVA, signed-rank tests, rank-sum tests, \ldots (depending on the scale and type of variables)

\medskip

{\bf Split point selection:}
Exhaustive search over all possible split points or two-sample test statistics.

\medskip

{\bf Stopping criteria:}
Minimal number of observations per node, tree depth, minimal $p$-value, \ldots

\end{frame}

\begin{frame}[fragile]
\frametitle{Tree algorithm}
\vspace{0.2cm}

{\bf How and where to split?}

\vspace{0.4cm}

\begin{enumerate}
\item Selection of the best split variable.
\vspace{0.2cm}
% \item \textcolor{gray}{Selection of the corresponding best split point.}
\item Selection of the corresponding best split point.
\end{enumerate}
\end{frame}

\subsection{Split variable selection}
\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Gender

\vspace{0.5cm}

\begin{minipage}{0.5\linewidth}
\vspace{-2.2cm}
<<echo=TRUE, results=verbatim>>=
table(ttnc$Survived, ttnc$Gender)
@
\end{minipage}
\begin{minipage}{0.45\linewidth}
<<echo=TRUE, eval=FALSE>>=
plot(Survived~Gender, data = ttnc)
@
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(Survived~Gender, data = ttnc)
@
\end{minipage}

\pause

<<echo=TRUE, results=verbatim>>=
chisq.test(ttnc$Survived, ttnc$Gender)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Age

\vspace{0.5cm}

\begin{minipage}{0.5\linewidth}
\vspace{-2.2cm}
<<echo=TRUE, results=verbatim>>=
table(ttnc$Survived, ttnc$Age)
@
\end{minipage}
\begin{minipage}{0.45\linewidth}
<<echo=TRUE, eval=FALSE>>=
plot(Survived~Age, data = ttnc)
@
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(Survived~Age, data = ttnc)
@
\end{minipage}

\pause

<<echo=TRUE, results=verbatim>>=
chisq.test(ttnc$Survived, ttnc$Age)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Class

\vspace{0.5cm}

\begin{minipage}{0.5\linewidth}
\vspace{-2.2cm}
<<echo=TRUE, results=verbatim>>=
table(ttnc$Survived, ttnc$Class)
@
\end{minipage}
\begin{minipage}{0.45\linewidth}
<<echo=TRUE, eval=FALSE>>=
plot(Survived~Class, data = ttnc)
@
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(Survived~Class, data = ttnc)
@
\end{minipage}

\pause

<<echo=TRUE, results=verbatim>>=
chisq.test(ttnc$Survived, ttnc$Class)
@
\end{frame}

% 
% \begin{frame}[fragile]
% \frametitle{Split variable selection} 
% {\bf Compare pairwise associations} (Titanic): 
% Based on independence tests, e.g., $\chi^2$-test for categorical variables.
% \vspace{0.4cm}
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived~Gender, data = ttnc)
% @
% 
% \vspace{0.2cm}
% 
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived~Age, data = ttnc)
% @
% 
% \vspace{0.7cm}
% 
% $\Rightarrow$ Select the split variable showing the lowest $p$-value.
% 
% \end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\small Titanic:  \quad $Survived \; \sim  \; Gender + Age + Class$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<echo=FALSE, fig=TRUE, width=11>>=
<<ctree_Titanic>>
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}

%\vspace{0.2cm}

{\bf Idea:}
Employ statistical tests for the split variable selection.\\
Evaluating associations on the unified $p$-value scale allows for an unbiased comparison of possible split variables on different scales and of different types.

\bigskip

{\bf Unbiased recursive partitioning:} Separation of split variable selection (based on $p$-values) and split point selection (only within the already chosen split variable).

\bigskip

{\bf Alternative:} Directly compare all available split points in all possible split variables. In order to allow for an unbiased selection within this strategy, it has to be accounted for differences such as a varying number of available split points, different scales or variable types by correcting the measurement considered for comparison accordingly. However, this has often not been included in classic tree algorithms.
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ education

\vspace{-0.7cm}

\begin{minipage}{0.4\linewidth}
<<echo=TRUE, results=verbatim>>=
cor(log(CPS1985$wage), 
    CPS1985$education)
@
<<echo=TRUE, eval=FALSE>>=
plot(log(wage)~education, 
     data = CPS1985)
@
\end{minipage}
\hfill
\begin{minipage}{0.5\linewidth}
%\vspace{-0.2cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(log(wage)~education, data = CPS1985)
@
\end{minipage}

\pause

\vspace{-0.7cm}

<<echo=TRUE, results=verbatim>>=
cor.test(log(CPS1985$wage), CPS1985$education)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender

\vspace{0.2cm}

<<echo=TRUE, results=verbatim>>=
tapply(log(CPS1985$wage), CPS1985$gender, summary)
@
\begin{minipage}{0.37\linewidth}
\vspace{-2cm}
<<echo=TRUE, eval=FALSE>>=
boxplot(log(wage)~gender, 
        data = CPS1985)
@
\end{minipage}
\hfill
\begin{minipage}{0.55\linewidth}
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.4\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
boxplot(log(wage)~gender, data = CPS1985)
@
\end{minipage}
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender

\vspace{0.2cm}

<<echo=TRUE, results=verbatim>>=
oneway.test(log(wage)~gender, data = CPS1985, var.equal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Split variable selection}
{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.99\textwidth}
<<echo=FALSE, fig=TRUE, width=13>>=
<<ctree_Wages1985>> 
@
\end{center}
\end{frame}

% \begin{frame}[fragile]
% \frametitle{Tree algorithm}
% \vspace{0.2cm}
% 
% {\bf How and where to split?}
% 
% \vspace{0.4cm}
% 
% \begin{enumerate}
% %\item \textcolor{gray}{Selection of the best split variable.} \checkmark
% \item Selection of the best split variable.
% \vspace{0.2cm}
% \item Selection of the corresponding best split point.
% \end{enumerate}
% \end{frame}

\subsection{Split point selection\\ 
\&\\ 
pruning}

\begin{frame}[fragile]
\frametitle{Split point selection}
{\bf Exhaustive search:} For each possible split point evaluate a defined objective function in the resulting subgroups and select the split point that minimizes the objective function, i.e., leads to the highest improvement of the model fit.

\vspace{0.4cm}

{\bf Alternative:} Two-sample test statistics evaluating the discrepancy between resulting subgroups. 
\end{frame}

<<split_point_search>>=

evalsplit <- function(sp, type = c("rmse","teststatistic")){

  if(type == "rmse"){
    d1 <- CPS1985[CPS1985$education <= sp,]
    d2 <- CPS1985[CPS1985$education > sp,]

    y1 <- mean(log(d1$wage))
    y2 <- mean(log(d2$wage))

    rmse1 <- sum(sqrt((log(d1$wage) - y1)^2))
    rmse2 <- sum(sqrt((log(d2$wage) - y2)^2))

    rmse <- rmse1 + rmse2

    return(rmse)
  }

  if(type == "teststatistic"){

    itest <- independence_test(log(wage) ~ factor(education <= sp), data = CPS1985) 
    pval <- as.numeric(pvalue(itest))
    teststatistic <- statistic(itest)
    return(list(pval = pval,
                teststatistic = teststatistic))
  }
}

evalpoints <- sort(unique(CPS1985$education))
evalpoints <- evalpoints[-length(evalpoints)]
plotpoints_rmse <- numeric(length = length(evalpoints))
for(j in 1: length(plotpoints_rmse))
  plotpoints_rmse[j] <- evalsplit(evalpoints[j], type = "rmse")

plotpoints_statistic <- plotpoints_pval <- numeric(length = length(evalpoints))
for(k in 1: length(plotpoints_statistic)){
  test <- evalsplit(evalpoints[k], type = "teststatistic")
  plotpoints_statistic[k] <- test$teststatistic
  plotpoints_pval[k] <- test$pval
}

@

\begin{frame}[fragile]
\frametitle{Split point selection}

%\setkeys{Gin}{width=0.8\textwidth}
\begin{minipage}{0.49\linewidth}
<<plot_split_point_search_rmse, fig=TRUE, width = 4.5, height = 4>>=
plot(x = evalpoints, y = plotpoints_rmse, type = 'o',
     xlab = "education", ylab = "RMSE")
@
\end{minipage}
\hfill
\begin{minipage}{0.49\linewidth}
<<plot_split_point_search_statistic, fig=TRUE, width = 4.5, height = 4>>=
plot(x = evalpoints, y = plotpoints_statistic, type = 'o',
     xlab = "education", ylab = "Two-sample test statistic")

# plot(x = evalpoints, y = plotpoints_pval, type = 'o',
#      xlab = "education", ylab = "twosample-teststatistic",
#      ylim = c(0,0.1))

@
\end{minipage}
\end{frame}

\begin{frame}[fragile]
\frametitle{Split point selection}
{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.99\textwidth}
<<echo=FALSE, fig=TRUE, width=13>>=
<<ctree_Wages1985>> 
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Pruning}

{\bf Goal:} Avoid overfitting.

\medskip

{\bf Pre-pruning:} Internal stopping criterium. Stop splitting when there is no significant association to any of the possible split variable.

\medskip

{\bf Post-pruning:} Grow large tree and prune splits that do not improve the model fit (e.g., via cross-validation or information criteria).

\end{frame}

\begin{frame}[fragile]
\frametitle{Pruning}

{\bf Pre-pruning:}
\begin{itemize}
\item[+] Does not require additional calculations as only already provided information from statistical test is used.
\item[+] Computationally less expensive as trees are prevented from getting too large.
\item[-] Performance depends on the power of the selected statistical test.
\end{itemize}

\vspace{0.4cm}

{\bf Post-pruning:}
\begin{itemize}
\item[+] Stable method to avoid overfitting, regardless of the power of the employed statistical test.
\item[+] Can employ information criteria such as AIC or BIC for model-based partitioning. 
\item[-] Computationally more expensive as very large trees are grown and additional (out-of-bag) evaluations are required.
\end{itemize}

\end{frame}

\subsection{Algorithms \\
\& \\
implementations}

\begin{frame}
\frametitle{CTree}

\begin{itemize} 
\item Conditional inference trees.
\item Flexible and general framework for unbiased recursive partitioning,
originally proposed for nonparametric modeling but also allowing for a model based approach via transformations of the response. 
\item Split variable selection based on conditional permutation tests.
\item Split point selection based on two-sample test statistics.
\item Employs pre-pruning.
\item Implemented in the R package partykit.
\end{itemize}

\bigskip

{\bf Note:} The CTree algorithm has been employed for all previous tree illustrations.

\end{frame}

\begin{frame}[fragile]
\frametitle{CTree}

Apply \fct{ctree} to preprocessed \code{ttnc} data:

<<ctree, echo=TRUE, eval=FALSE>>=
ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc)
@

\medskip

Plot:

<<ctree-plot, echo=TRUE, eval=FALSE>>=
plot(ct)
@

\medskip

Add control arguments via \fct{ctree\_control}:

<<ctree_control, echo=TRUE, eval=FALSE>>=
ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
            control = ctree_control(alpha = 0.01,
                                    minsplit = 20, 
                                    maxdepth = 3))
@
\end{frame}

\begin{frame}[fragile]
\frametitle{CTree}

\setkeys{Gin}{width=0.9\linewidth}
<<ctree_control_plot, echo=TRUE, eval=TRUE, fig=TRUE, width=10, height=5>>=
ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
            control = ctree_control(alpha = 0.01, minsplit = 20, maxdepth = 3))
plot(ct)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{CTree}

{\bf Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.

<<ctree_predict, echo=TRUE, eval=TRUE, results=verbatim>>=
nd <- data.frame(Class = "2nd", Gender = c("Female", "Male"), Age = "Child")
nd
predict(ct, newdata = nd)
@

\end{frame}

\begin{frame}
\frametitle{rpart}

\begin{itemize} 
\item Recursive PARTitioning.
\item Based on ideas of CART (Classification And Regression Trees).
\item Split variable and split point selection based on optimizing an objective function.
\item Classification trees: reduce number of misclassifications.
\item Regression trees: reduce objective function for numeric response.
\item Post-pruning can be employed.
\item Implemented in the R package rpart.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

Apply \fct{rpart} to preprocessed \code{ttnc} data:

<<rpart, echo=TRUE, eval=TRUE>>=
rp <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
@

\medskip

Standard plot:

<<rpart-plot, echo=TRUE, eval=FALSE>>=
plot(rp)
text(rp)
@

Visualization via partykit:

<<rpart-party-plot, echo=TRUE, eval=FALSE>>=
plot(as.party(rp))
@
\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

<<rpart-plot_eval, echo=TRUE, eval=FALSE>>=
plot(rp)
text(rp)
@

\vspace{-0.7cm}

\setkeys{Gin}{width=0.9\linewidth}
<<rpart-plot_eval, echo=FALSE, eval=TRUE, fig=TRUE, width=10, height = 6>>=
plot(rp)
text(rp)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

<<rpart-party-plot_eval, echo=TRUE, eval=FALSE>>=
plot(as.party(rp))
@

%\vspace{-0.4cm}

\setkeys{Gin}{width=0.9\linewidth}
<<rpart-party-plot_eval, echo=FALSE, eval=TRUE, fig=TRUE, width = 12>>=
plot(as.party(rp))
@
\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

Standard output: 
<<rpart-print, echo=TRUE, results=verbatim>>=
rp
@

\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

Output via partykit: 
<<rpart-party_print, echo=TRUE, results=verbatim>>=
as.party(rp)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

{\bf Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.

%\medskip

<<rpart_pruning, echo=TRUE, eval=TRUE, results=verbatim>>=
rp_pruned <- prune(rp, cp = 0.01)
@

\setkeys{Gin}{width=0.6\linewidth}
<<rpart_pruning_plot, echo=TRUE, eval=TRUE, fig=TRUE, width = 8, height=5>>=
plot(as.party(rp_pruned))
@
% plot(rp_pruned)
% text(rp_pruned)

\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

{\bf Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.

%\medskip

<<rpart_pruning2, echo=TRUE, eval=TRUE, results=verbatim>>=
rp_pruned <- prune(rp, cp = 0.02)
@

\setkeys{Gin}{width=0.6\linewidth}
<<rpart_pruning2_plot, echo=TRUE, eval=TRUE, fig=TRUE, width = 8, height=5>>=
plot(as.party(rp_pruned))
@
% plot(rp_pruned)
% text(rp_pruned)

\end{frame}

\begin{frame}[fragile]
\frametitle{rpart}

{\bf Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.

%\medskip

<<rpart_pruning3, echo=TRUE, eval=TRUE, results=verbatim>>=
rp_pruned <- prune(rp, cp = 0.03)
@

\setkeys{Gin}{width=0.6\linewidth}
<<rpart_pruning3_plot, echo=TRUE, eval=TRUE, fig=TRUE, width =8, height=5>>=
plot(as.party(rp_pruned))
@
% plot(rp_pruned)
% text(rp_pruned)

\end{frame}

\begin{frame}
\frametitle{MOB}

\begin{itemize} 
\item Model-based recursive partitioning using M-estimation (ML, OLS, CRPS, \ldots).
\item Split variable selection based on asymptotic parameter instability tests.
\item Split point selection based on exhaustive search.
\item Employs pre-pruning, but can also use post-pruning based on AIC or BIC.
\item Implemented in the R package partykit.
\end{itemize}

\end{frame}

% \begin{frame}
% \frametitle{R software for tree models}
% 
% {\bf rpart:} 
% \begin{itemize}
% \item R package for Recursive PARTitioning.
% \item Implementation based on ideas of CART (Classification And Regression Trees).
% \item Classification trees: reduce number of misclassifications.
% \item Regression trees: reduce objective function for numeric response.
% \end{itemize}
% 
% \vspace{0.4cm}
% 
% {\bf partykit:} 
% \begin{itemize} 
% \item R package offering a toolkit for unbiased recursive partitioning. 
% \item Implementations of the tree algorithms CTree (Conditional Inference Trees) and MOB (Model-based Recursive Partitioning).
% \item Provides a very general framework with a unifying infrastructure.
% \item Can import rpart.
% \end{itemize}
% 
% \end{frame}

\begin{frame}[fragile]
\frametitle{MOB}

Is ``women and children first'' applied equally for all passenger classes?

\bigskip

Add a factor variable distinguishing women and children from male adults:
<<ttnc_treatment, echo=TRUE, eval=TRUE>>=
ttnc <- transform(ttnc, Treatment = factor(Gender == "Female" | Age == "Child", 
                                           levels = c(FALSE, TRUE),
                                           labels = c("Male&Adult", "Female|Child")))
@

\bigskip

Assess differences in the preferential treatment effect over the remaining covariates by applying \fct{glmtree} which employs the MOB algorithm for generalized linear models in the tree nodes:

<<glmtree, echo=TRUE, eval=TRUE, results=verbatim>>=
glmt <- glmtree(Survived ~ Treatment | Class + Gender + Age,
                data = ttnc, family = binomial, alpha = 0.01)
@

<<glmtree_print, echo=TRUE, eval=FALSE>>=
glmt
@
\end{frame}

\begin{frame}[fragile]
\frametitle{MOB}
\vspace{-0.4cm}
\small
<<glmtree_print_eval, echo=FALSE, eval=TRUE, results=verbatim>>=
glmt
@
\end{frame}

\begin{frame}[fragile]
\frametitle{MOB}
\setkeys{Gin}{width=0.99\textwidth}
<<glmtree-plot, echo=TRUE, eval=TRUE, fig=TRUE, width=11, height=5>>=
plot(glmt)
@
\end{frame}

\begin{frame}
\frametitle{R software for tree models}

{\bf Further R packages based on partykit:}
\begin{itemize}
  \item evtree (evolutionary trees),
  \item model4you (personalised trees),
  \item disttree (distributional trees),
  \item stablelearner (stability assessment),
  \item \ldots
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{R software for tree models}

{\bf Further R packages:}
\begin{itemize}
  \item mvpart (multivariate CART),
  \item RWeka (J4.8, M5', LMT),
  \item and many more (C50, quint, stima, \dots).
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{R software for tree models}

\textbf{Furthermore:} Tree algorithms/software without R interface, e.g.,
%
\begin{itemize}
  \item QUEST,
  \item GUIDE,
  \item LOTUS,
  \item CRUISE,
  \item \dots
\end{itemize}

\end{frame}

% \subsection*{References}

\begin{frame}
\frametitle{References}

%\vspace{-0.2cm}

\small

Breiman L, Friedman JH, Olshen RA, Stone CJ (1984).
  \dquote{Classification and Regression Trees.}
  \emph{Wadsworth, California.},

\medskip

% Breiman L (2001).
%   \dquote{Random {Forests}.}
%   \emph{Machine Learning}, 
%   \textbf{45}(1), 5--32.
%   \doi{10.1023/A:1010933404324}
% 
% \medskip

Hothorn T, Hornik K, Zeileis A (2006).
 \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
 \emph{Journal of Computational and Graphical Statistics},
 \textbf{15}(3), 651--674.
 \doi{10.1198/106186006X133933}

\medskip

Zeileis A, Hothorn T, Hornik K (2008).
 \dquote{Model-Based Recursive Partitioning.}
  \emph{Journal of Computational and Graphical Statistics},
  \textbf{17}(2), 492--514.
  \doi{10.1198/106186008X319331}

\medskip

Hothorn T, Zeileis A (2015).
 \dquote{{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
 \emph{Journal of Machine Learning Research},
 \textbf{16}, 3905--3909.
 \url{http://www.jmlr.org/papers/v16/hothorn15a.html}

\medskip

Therneau T, Atkinson B (2019). 
  \dquote{{rpart}: Recursive Partitioning and Regression Trees.}
  \url{https://CRAN.R-project.org/package=rpart}

% medskip
%  
% Schlosser L, Hothorn T, Stauffer R, Zeileis A (2019).
% \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
% \emph{The Annals of Applied Statistics}, \textbf{13}(3), 1564--1589.
% \doi{10.1214/19-AOAS1247}

% \medskip

% Breiman L (2001).
%   \dquote{Statistical Modeling: The Two Cultures.}
%   \emph{Statistical Science}, 
%   \textbf{16}(3), 199--231.
%   \doi{10.1214/ss/1009213726}

\end{frame}

\end{document}

### 2020-06-30 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
changed title changed author order streamlined LaTeX header
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3019 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b392afeb40779fe25199dc93b3e0adc981db2198

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
-\usetheme[nototalframenumber,url,license]{uibk}
-
-\title{Tree models}
+\usetheme[nototalframenumber,license]{uibk}
+
+\title{Classification and Regression Trees and Beyond}
 \subtitle{Supervised Learning: Algorithmic Modeling}
-\author{Achim Zeileis, Lisa Schlosser}
-\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
-%\URL{http://www.partykit.org/partykit/}
+\author{Lisa Schlosser, Achim Zeileis}
+
+%% forest header image
 \renewcommand{\headerimage}[1]{%
@@
 
-%\graphicspath{{Figures/}}
-\usepackage[utf8]{inputenc}
-
-\setbeamertemplate{caption}{\insertcaption} 
-%% includes a replacement for \usepackage{Sweave}
-% \usepackage{Sweave}
-\usepackage{changepage}
-\usepackage{amsmath,tikz}
-\usepackage{calc}
-\usepackage{graphicx}
-\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
-\usepackage{color}
-\usepackage{xcolor}
-\usepackage{fancyvrb}
-%\usepackage[cal=boondoxo]{mathalfa}
-%\graphicspath{{plots/}}
-\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
-
-\setbeamertemplate{subsection page}
-{\begin{center}
-\vspace{2cm}
-% {\Large \insertsection\par}
-% \vspace{0.5cm}
-Section~\thesubsection\par
-\vspace{0.5cm}
-{\LARGE \insertsubsection\par}
-\end{center}}
-\AtBeginSubsection{\frame{\subsectionpage}}
-
-%% colors
-% \definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
-% \definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
-% \definecolor{forestred}{RGB}{206,73,81}
-% \definecolor{treegreen}{RGB}{0,143,0}
-% \definecolor{lightblue}{RGB}{34,151,230}
-% \definecolor{lightorange}{RGB}{255,165,0}
-
-%\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
+%% custom subsection slides
+\setbeamercolor*{subsectionfade}{use={normal text},parent={normal text},fg=structure.fg!30!normal text.bg}
+\AtBeginSubsection[]{%
+  \begin{frame}[c]
+    \begin{center}
+      \usebeamercolor[fg]{subsectionfade}
+      \Large \insertsection \\[2ex]
+      \usebeamercolor[fg]{structure}
+      \huge\bfseries\insertsubsection
+    \end{center}
+  \end{frame}
+}
+
+%% for \usepackage{Sweave}
 \SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}
@@
 
+library("rpart")
+library("partykit")
+library("coin")
+
 set.seed(7)
-
-invisible(.Call(grDevices:::C_palette, grDevices::hcl(
-  h = c(0,   5, 125, 245, 195, 315,  65,   0),
-  c = c(0, 100,  90,  85,  63, 105,  90,   0),
-  l = c(0,  55,  75,  60,  82,  48,  80,  65)
-)))
-
-# setwd("~/Documents/ULG/slides_tree/")
-library("disttree")
-library("Formula")
-library("latex2exp")
-library("AER")
-#library("gamlss")
-#library("lattice")
-#library("crch")
-#library("RainTyrol")
-#library("parallel")
-library("rpart")
-#library("gamlss.cens")
-#gen.cens(NO, type = "left")
-#library("ggplot2")
-#theme_set(theme_bw(base_size = 18))
-library("MASS")
-#library("Ecdat")
-#library("wooldridge")
-library("partykit")
-#library("RWeka")
-#library("evtree")
-library("coin")
-
-## HCL palette
-# pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
-# names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
-# 
-# pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
-# names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")
-# 
-# transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
-@
-
-<<data>>=
+@
+
+<<Titanic-data>>=
 data("Titanic", package = "datasets")
@@
 names(ttnc)[2] <- "Gender"
-# ttnc <- transform(ttnc, Treatment = factor(Gender == "Female" | Age == "Child", 
-#                                            levels = c(FALSE, TRUE),
-#                                            labels = c("Male&Adult", "Female|Child")))
-##   labels = c("Normal\n(Male&Adult)", "Preferential\n(Female|Child)")
 @
@@
 
-\section{Tree models}
+\section{Classification and regression trees and beyond}
 
@@
 <<ctree_Wages1985, echo=FALSE, fig=TRUE, width=13>>=
-data("CPS1985")
+data("CPS1985", package = "AER")
 plot(ctree(log(wage)~education+experience+age+ethnicity+gender+union,data=CPS1985,
@@
 <<lmtree_Wages1985, echo=FALSE, fig=TRUE, width=11>>=
-data("CPS1985")
-plot(lmtree(log(wage)~education|experience+age+ethnicity+gender+union,data=CPS1985))
+data("CPS1985", package = "AER")
+plot(lmtree(log(wage) ~ education | experience + age + ethnicity + gender + union, data = CPS1985))
 @
@@
 % <<ctree_Wages1988, echo=FALSE, fig=TRUE, width=13>>=
-% data("CPS1988")
+% data("CPS1988", package = "AER")
 % ## ctree

### 2020-06-30 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
streamline R code
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3020 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 80c1c0b24be6d1e76e628c9e970c540b56484e29

@@
 
-<<Titanic-data>>=
+<<data>>=
+data("CPS1985", package = "AER")
 data("Titanic", package = "datasets")
@@
 <<ctree_Titanic, echo=FALSE, fig=TRUE, width=11>>=
-ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
-            control = ctree_control(alpha = 0.01))
-plot(ct)
+ct_ttnc <- ctree(Survived ~ Gender + Age + Class, data = ttnc, alpha = 0.01)
+plot(ct_ttnc)
 @
@@
 <<ctree_Wages1985, echo=FALSE, fig=TRUE, width=13>>=
-data("CPS1985", package = "AER")
-plot(ctree(log(wage)~education+experience+age+ethnicity+gender+union,data=CPS1985,
-           control = ctree_control(alpha = 0.01)))
+ct_cps <- ctree(log(wage) ~ education + experience + age + ethnicity + gender + union, data = CPS1985, alpha = 0.01)
+plot(ct_cps)
 @
@@
 <<lmtree_Wages1985, echo=FALSE, fig=TRUE, width=11>>=
-data("CPS1985", package = "AER")
-plot(lmtree(log(wage) ~ education | experience + age + ethnicity + gender + union, data = CPS1985))
+mob_cps <- lmtree(log(wage) ~ education | experience + age + ethnicity + gender + union, data = CPS1985)
+plot(mob_cps)
 @
@@
 
-\vspace{0.5cm}
-
-\begin{minipage}{0.5\linewidth}
-\vspace{-2.2cm}
+\begin{minipage}[t]{0.43\linewidth}
+<<survival-gender, echo=TRUE, eval=FALSE>>=
+plot(Survived ~ Gender, data = ttnc)
+@
+\vspace{-0.7cm}
+<<survival-gender-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
+<<survival-gender>>
+@
+\end{minipage}
+\hfill
+\pause
+\begin{minipage}[t]{0.5\linewidth}
 <<echo=TRUE, results=verbatim>>=
-table(ttnc$Survived, ttnc$Gender)
+xtabs(~ Survived + Gender, data = ttnc)
 @
 \end{minipage}
-\begin{minipage}{0.45\linewidth}
-<<echo=TRUE, eval=FALSE>>=
-plot(Survived~Gender, data = ttnc)
-@
-\vspace{-0.7cm}
-%\setkeys{Gin}{width=0.5\linewidth}
-<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
-plot(Survived~Gender, data = ttnc)
-@
-\end{minipage}
 
@@
 <<echo=TRUE, results=verbatim>>=
-chisq.test(ttnc$Survived, ttnc$Gender)
-@
+chisq.test(xtabs(~ Survived + Gender, data = ttnc))
+@
+
 \end{frame}
@@
 
-\vspace{0.5cm}
-
-\begin{minipage}{0.5\linewidth}
-\vspace{-2.2cm}
+\begin{minipage}[t]{0.43\linewidth}
+<<survival-age, echo=TRUE, eval=FALSE>>=
+plot(Survived ~ Age, data = ttnc)
+@
+\vspace{-0.7cm}
+<<survival-age-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
+<<survival-age>>
+@
+\end{minipage}
+\hfill
+\pause
+\begin{minipage}[t]{0.5\linewidth}
 <<echo=TRUE, results=verbatim>>=
-table(ttnc$Survived, ttnc$Age)
+xtabs(~ Survived + Age, data = ttnc)
 @
 \end{minipage}
-\begin{minipage}{0.45\linewidth}
-<<echo=TRUE, eval=FALSE>>=
-plot(Survived~Age, data = ttnc)
-@
-\vspace{-0.7cm}
-%\setkeys{Gin}{width=0.5\linewidth}
-<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
-plot(Survived~Age, data = ttnc)
-@
-\end{minipage}
-
-\pause
 
 <<echo=TRUE, results=verbatim>>=
-chisq.test(ttnc$Survived, ttnc$Age)
-@
+chisq.test(xtabs(~ Survived + Age, data = ttnc))
+@
+
 \end{frame}
@@
 
-\vspace{0.5cm}
-
-\begin{minipage}{0.5\linewidth}
-\vspace{-2.2cm}
+\begin{minipage}[t]{0.43\linewidth}
+<<survival-class, echo=TRUE, eval=FALSE>>=
+plot(Survived ~ Class, data = ttnc)
+@
+\vspace{-0.7cm}
+<<survival-class-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
+<<survival-class>>
+@
+\end{minipage}
+\hfill
+\pause
+\begin{minipage}[t]{0.5\linewidth}
 <<echo=TRUE, results=verbatim>>=
-table(ttnc$Survived, ttnc$Class)
+xtabs(~ Survived + Class, data = ttnc)
 @
 \end{minipage}
-\begin{minipage}{0.45\linewidth}
-<<echo=TRUE, eval=FALSE>>=
-plot(Survived~Class, data = ttnc)
-@
-\vspace{-0.7cm}
-%\setkeys{Gin}{width=0.5\linewidth}
-<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
-plot(Survived~Class, data = ttnc)
-@
-\end{minipage}
-
-\pause
 
 <<echo=TRUE, results=verbatim>>=
-chisq.test(ttnc$Survived, ttnc$Class)
-@
+chisq.test(xtabs(~ Survived + Class, data = ttnc))
+@
+
 \end{frame}
@@
 % <<echo=TRUE, results=verbatim>>=
-% chisq_test(Survived~Gender, data = ttnc)
+% chisq_test(Survived ~ Gender, data = ttnc)
 % @
@@
 % <<echo=TRUE, results=verbatim>>=
-% chisq_test(Survived~Age, data = ttnc)
+% chisq_test(Survived ~ Age, data = ttnc)
 % @
@@
 
+\begin{minipage}[t]{0.39\linewidth}
+<<logwage-education, echo=TRUE, eval=FALSE>>=
+plot(log(wage) ~ education,
+  data = CPS1985)
+@
 \vspace{-0.7cm}
-
-\begin{minipage}{0.4\linewidth}
-<<echo=TRUE, results=verbatim>>=
-cor(log(CPS1985$wage), 
-    CPS1985$education)
-@
-<<echo=TRUE, eval=FALSE>>=
-plot(log(wage)~education, 
-     data = CPS1985)
+\setkeys{Gin}{width=\textwidth}
+<<logwage-education-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
+<<logwage-education>>
 @
@@
 \hfill
-\begin{minipage}{0.5\linewidth}
-%\vspace{-0.2cm}
-%\setkeys{Gin}{width=0.5\linewidth}
-<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
-plot(log(wage)~education, data = CPS1985)
+\pause
+\begin{minipage}[t]{0.59\linewidth}
+<<echo=TRUE, eval=FALSE>>=
+cor.test(~ log(wage) + education, data = CPS1985)
+@
+<<echo=FALSE, results=verbatim>>=
+out <- capture.output(cor.test(~ log(wage) + education, data = CPS1985))
+out <- gsub("alternative hypothesis: ", "alternative hypothesis:\n", out, fixed = TRUE)
+writeLines(out)
 @
 \end{minipage}
-
-\pause
-
-\vspace{-0.7cm}
-
-<<echo=TRUE, results=verbatim>>=
-cor.test(log(CPS1985$wage), CPS1985$education)
-@
 
@@
 
-\vspace{0.2cm}
-
-<<echo=TRUE, results=verbatim>>=
-tapply(log(CPS1985$wage), CPS1985$gender, summary)
-@
-\begin{minipage}{0.37\linewidth}
-\vspace{-2cm}
-<<echo=TRUE, eval=FALSE>>=
-boxplot(log(wage)~gender, 
+\begin{minipage}[t]{0.39\linewidth}
+<<logwage-gender, echo=TRUE, eval=FALSE>>=
+plot(log(wage) ~ gender,
   data = CPS1985)
+@
+\vspace{-0.7cm}
+\setkeys{Gin}{width=\textwidth}
+<<logwage-gender-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
+<<logwage-gender>>
 @
@@
 \hfill
-\begin{minipage}{0.55\linewidth}
-\vspace{-0.7cm}
-%\setkeys{Gin}{width=0.4\linewidth}
-<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
-boxplot(log(wage)~gender, data = CPS1985)
+\pause
+\begin{minipage}[t]{0.59\linewidth}
+<<echo=TRUE, eval=FALSE>>=
+t.test(log(wage) ~ gender, data = CPS1985)
+@
+<<echo=FALSE, results=verbatim>>=
+out <- capture.output(t.test(log(wage) ~ gender, data = CPS1985))
+out <- gsub("alternative hypothesis: ", "alternative hypothesis:\n", out, fixed = TRUE)
+writeLines(out)
 @
 \end{minipage}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Split variable selection}
-{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender
-
-\vspace{0.2cm}
-
-<<echo=TRUE, results=verbatim>>=
-oneway.test(log(wage)~gender, data = CPS1985, var.equal = TRUE)
-@
+
 \end{frame}
@@
 <<split_point_search>>=
-
-evalsplit <- function(sp, type = c("rmse","teststatistic")){
-
-  if(type == "rmse"){
-    d1 <- CPS1985[CPS1985$education <= sp,]
-    d2 <- CPS1985[CPS1985$education > sp,]
-
-    y1 <- mean(log(d1$wage))
-    y2 <- mean(log(d2$wage))
-
-    rmse1 <- sum(sqrt((log(d1$wage) - y1)^2))
-    rmse2 <- sum(sqrt((log(d2$wage) - y2)^2))
-
-    rmse <- rmse1 + rmse2
-
-    return(rmse)
+evalsplit <- function(sp) {
+  m <- lm(log(wage) ~ factor(education <= sp), data = CPS1985)
+  s <- independence_test(log(wage) ~ factor(education <= sp), data = CPS1985, teststat = "quadratic") 
+  c(
+    rss = deviance(m),
+    loglik = logLik(m),
+    logpval = pchisq(statistic(s), df = 1, log = TRUE, lower.tail = FALSE),
+    teststatistic = statistic(s)
+  )
 }  
-
-  if(type == "teststatistic"){
-
-    itest <- independence_test(log(wage) ~ factor(education <= sp), data = CPS1985) 
-    pval <- as.numeric(pvalue(itest))
-    teststatistic <- statistic(itest)
-    return(list(pval = pval,
-                teststatistic = teststatistic))
-  }
-}
-
-evalpoints <- sort(unique(CPS1985$education))
-evalpoints <- evalpoints[-length(evalpoints)]
-plotpoints_rmse <- numeric(length = length(evalpoints))
-for(j in 1: length(plotpoints_rmse))
-  plotpoints_rmse[j] <- evalsplit(evalpoints[j], type = "rmse")
-
-plotpoints_statistic <- plotpoints_pval <- numeric(length = length(evalpoints))
-for(k in 1: length(plotpoints_statistic)){
-  test <- evalsplit(evalpoints[k], type = "teststatistic")
-  plotpoints_statistic[k] <- test$teststatistic
-  plotpoints_pval[k] <- test$pval
-}
-
+sp <- head(sort(unique(CPS1985$education)), -1)
+eval <- sapply(sp, evalsplit)
 @
@@
 
-%\setkeys{Gin}{width=0.8\textwidth}
-\begin{minipage}{0.49\linewidth}
-<<plot_split_point_search_rmse, fig=TRUE, width = 4.5, height = 4>>=
-plot(x = evalpoints, y = plotpoints_rmse, type = 'o',
-     xlab = "education", ylab = "RMSE")
-@
-\end{minipage}
-\hfill
-\begin{minipage}{0.49\linewidth}
-<<plot_split_point_search_statistic, fig=TRUE, width = 4.5, height = 4>>=
-plot(x = evalpoints, y = plotpoints_statistic, type = 'o',
-     xlab = "education", ylab = "Two-sample test statistic")
-
-# plot(x = evalpoints, y = plotpoints_pval, type = 'o',
-#      xlab = "education", ylab = "twosample-teststatistic",
-#      ylim = c(0,0.1))
-
-@
-\end{minipage}
+<<plot_split_point_search1, fig=TRUE, width = 7.5, height = 5>>=
+par(mar = c(5, 3, 1, 3))
+plot(sp, eval["loglik",], type = "n", xlab = "education", ylab = "", axes = FALSE)
+abline(v = 13, lty = 2, lwd = 1.5)
+lines(sp, eval["loglik", ], type = "o", col = 2, lwd = 1.5)
+axis(2, col.ticks = 2, col.axis = 2)
+par(new = TRUE)
+plot(sp, eval["teststatistic",], type = "o",
+  col = 4, lwd = 1.5,
+  xlab = "", ylab = "", axes = FALSE)
+axis(4, col.ticks = 4, col.axis = 4)
+axis(1)
+axis(1, at = 13)
+box()
+legend("topleft", c("Log-likelihood", "Test statistic"), col = c(2, 4), lwd = 1.5, pch = 1, bty = "n")
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Split point selection}
+
+<<plot_split_point_search2, fig=TRUE, width = 7.5, height = 5>>=
+par(mar = c(5, 3, 1, 3))
+plot(sp, eval["rss",], type = "n", xlab = "education", ylab = "", axes = FALSE)
+abline(v = 13, lty = 2, lwd = 1.5)
+lines(sp, eval["rss", ], type = "o", col = 2, lwd = 1.5)
+axis(2, col.ticks = 2, col.axis = 2)
+par(new = TRUE)
+plot(sp, eval["logpval",], type = "o",
+  col = 4, lwd = 1.5,
+  xlab = "", ylab = "", axes = FALSE)
+axis(4, col.ticks = 4, col.axis = 4)
+axis(1)
+axis(1, at = 13)
+box()
+legend("bottomleft", c("Residual sum of squares", "Log-p-value"), col = c(2, 4), lwd = 1.5, pch = 1, bty = "n")
+@
+
 \end{frame}

### 2020-06-30 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
almost finished motivation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3021 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c4d4400c491c5d2ee06160f4ef354a5d2b4f8bfe

@@
 
-{\bf Idea:} \\
-\smallskip
-\quad Separate data into smaller and more homogeneous subgroups \\
-\quad based on information provided by covariates.
-%in order to obtain better fitting \\
-%\quad models in the resulting subsets.
-
-\bigskip
-
-\bigskip
-
-{\bf Formally:} \\
-\smallskip
-\quad Modeling of a dependent variable $Y$ by ``learning'' a recursive partition \\
-\quad w.r.t explanatory variables $X_1$, \dots, $X_m$.
-
-\bigskip
-
-\bigskip
-
-{\bf Key features:} \\
+\textbf{Idea:} ``Divide and conquer.''
+\begin{itemize}
+  \item \emph{Goal:} Split the data into small(er) and (rather) homogenous subgroups.
+  \item \emph{Inputs:} Explanatory variables (or covariates/regressors) used for splitting.
+  \item \emph{Output:} Prediction for dependent (or target) variable(s).
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Formally:}
+\begin{itemize}
+  \item Dependent variable $Y$ (possibly multivariate).
+  \item Based on explanatory variables $X_1, \dots, X_m$.
+  \item ``Learn'' subgroups of data by combining splits in $X_1, \dots, X_m$.
+  \item Predict $Y$ with (simple) model in the subgroups, often simply the mean.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Key features:}
 \begin{itemize}
   \item Predictive power in nonlinear regression relationships.
-\item Interpretability (enhanced by visualization), i.e., no ``black box'' methods.
+  \item Interpretability (enhanced by tree visualization), i.e., no ``black box''.
 \end{itemize}
@@
 \frametitle{Motivation}
-{\small Titanic:  \quad $Survived \; \sim  \; Gender + Age + Class$}
+
+\textbf{Example:} Survival (yes/no) on the Titanic. Women and children first?
+
 \begin{center}
@@
 \end{center}
+
 \end{frame}
@@
 \frametitle{Motivation}
-{\bf Data:}
-The Titanic data set provides information on the survival of 2201~passengers on the Titanic with the additional variables Class, Sex and Age.
-
-\bigskip
-
-The variables and their levels are as follows:
-
-\bigskip
-
-\begin{tabular}{l l l}
+
+\textbf{Model formula:} \code{Survived ~ Gender + Age + Class}.
+
+\bigskip
+
+\textbf{Data:} Information on the survival of 2201~passengers on board the
+ill-fated maiden voyage of the RMS~Titanic in 1912. A data frame containing
+\Sexpr{nrow(ttnc)} observations on \Sexpr{ncol(ttnc)} variables.
+
+\bigskip
+
+\begin{tabular}{ll}
 \hline 
-No & Name &	Levels\\
+Variable        & Description\\
 \hline 
-1 &	Class &	1st, 2nd, 3rd, Crew\\
-2 &	Sex &	Male, Female\\
-3 &	Age &	Child, Adult\\
-4 &	Survived &	No, Yes \\
+\code{Class}    & Factor: 1st, 2nd, 3rd, or Crew.\\
+\code{Gender}   & Factor: Male, Female.\\
+\code{Age}      & Factor: Child, Adult.\\
+\code{Survived} & Factor: No, Yes.\\
 \hline 
@@
 \end{frame}
+
 \begin{frame}[fragile]
 \frametitle{Motivation}
-{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
+
+\textbf{Example:} Determinants of wages and returns to education.
+
 \begin{center}
@@
 \frametitle{Motivation}
-{\small Wages:  \quad $log(wage) \; \sim  \; education \ | \ experience + age + ethnicity + gender + union$}
+
+\textbf{Model formula:} \code{log(wage) ~ education + experience + age + ethnicity + gender + union}.
+
+\bigskip
+
+\textbf{Data:} Random sample from the May 1985 US Current Population Survey.
+A data frame containing \Sexpr{nrow(CPS1985)} observations on \Sexpr{ncol(CPS1985)} variables.
+
+\bigskip
+
+\begin{tabular}{ll}
+\hline 
+Variable & Description\\
+\hline 
+\code{wage}       & Wage (in US dollars per hour). \\
+\code{education}  & Education (in years).  \\
+\code{experience} & Potential work experience (in years, \code{age - education - 6}). \\
+\code{age}        & Age (in years). \\
+\code{ethnicity}  & Factor: Caucasian, Hispanic, Other. \\
+\code{gender}     & Factor: Male, Female. \\
+\code{union}      & Factor. Does the individual work on a union job? \\
+\hline 
+\end{tabular}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Motivation}
+
+\textbf{Alternatively:} Linear regression tree.
+
 \begin{center}
@@
 
-Wind direction forecasting\\
-at Innsbruck airport
-
-\vspace{-1.4cm}
-
-\setkeys{Gin}{width=0.95\linewidth}
+\textbf{Model formula:} \code{log(wage) ~ education | experience + age + ethnicity + gender + union}.
+
+\bigskip
+
+\textbf{Model fit:}
+\begin{itemize}
+  \item Not just a group-specific mean.
+  \item But group-specific intercept and slope (i.e., returns to education).
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Motivation}
+
+\textbf{Example:} Nowcasting (1--3 hours ahead) of wind direction at Innsbruck Airport.
+
+\setkeys{Gin}{width=0.87\linewidth}
 \begin{center}
@@
 \end{center}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Motivation}
+
+\textbf{Data:} 41,979 data points for various weather observations.
+\begin{itemize}
+  \item Dependent variable: Wind direction 1--3 hours ahead.
+  \item Explanatory variables: Current weather observations including
+    wind direction, wind speed, temperature, (reduced) air pressure, relative humidity.
+  \item Circular response in $[0^{\circ}, 360^{\circ})$ with $0^{\circ} = 360^{\circ}$.
+\end{itemize}
+
+\bigskip
+
+\textbf{Model fit:} Circular distribution (von Mises), fitted by maximum likelihood.
 
@@
 
-{\bf Homogeneity:} Partitioning the data into subgroups aims at allowing for a better model fit in each resulting subset.
-
-\bigskip
-
-{\bf Model:} The model fitted to a response $Y$ in each subgroup optimizes an objective function, i.e., it can be a simple constant fit (e.g.
+\textbf{Homogeneity:} Partitioning the data into subgroups aims at allowing for a better model fit in each resulting subset.
+
+\bigskip
+
+\textbf{Model:} The model fitted to a response $Y$ in each subgroup optimizes an objective function, i.e., it can be a simple constant fit (e.g.
 average value or proportion, minimizing deviations) but also a more complex model (possibly including additional regressors, and, e.g., minimizing the residual sum of squares or the negative log-likelihood).
@@
 
-{\bf Trade-off:} Complexity in the tree structure vs.\ complexity in the fitted models.\\
+\textbf{Trade-off:} Complexity in the tree structure vs.\ complexity in the fitted models.\\
 Simple models in the nodes (such as constant fits) often lead to large trees while more complex models can compensate a higher variation within the nodes allowing for less complex trees.
@@
 % 
-% {\bf Examples of discrepancy measures:} 
+% \textbf{Examples of discrepancy measures:} 
 % \begin{itemize}
@@
 
-{\bf Base algorithm:}
+\textbf{Base algorithm:}
 \begin{enumerate}
@@
 
-{\bf Examples of possible transformations of the response $Y$:} 
+\textbf{Examples of possible transformations of the response $Y$:} 
 \begin{itemize}
@@
 
-{\bf Possible association tests:} 
+\textbf{Possible association tests:} 
 $\chi^2$-test, Pearson correlation test, two-sample t-test, ANOVA, signed-rank tests, rank-sum tests, \ldots (depending on the scale and type of variables)
@@
 
-{\bf Split point selection:}
+\textbf{Split point selection:}
 Exhaustive search over all possible split points or two-sample test statistics.
@@
 
-{\bf Stopping criteria:}
+\textbf{Stopping criteria:}
 Minimal number of observations per node, tree depth, minimal $p$-value, \ldots
@@
 
-{\bf How and where to split?}
+\textbf{How and where to split?}
 
@@
 \frametitle{Split variable selection}
-{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Gender
+\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Gender
 
@@
 \frametitle{Split variable selection}
-{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Age
+\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Age
 
@@
 \frametitle{Split variable selection}
-{\bf Assess pairwise associations} (Titanic): \ Survived $\sim$ Class
+\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Class
 
@@
 % \frametitle{Split variable selection} 
-% {\bf Compare pairwise associations} (Titanic): 
+% \textbf{Compare pairwise associations} (Titanic): 
 % Based on independence tests, e.g., $\chi^2$-test for categorical variables.
@@
 
-{\bf Idea:}
+\textbf{Idea:}
 Employ statistical tests for the split variable selection.\\
@@
 
-{\bf Unbiased recursive partitioning:} Separation of split variable selection (based on $p$-values) and split point selection (only within the already chosen split variable).
-
-\bigskip
-
-{\bf Alternative:} Directly compare all available split points in all possible split variables.
+\textbf{Unbiased recursive partitioning:} Separation of split variable selection (based on $p$-values) and split point selection (only within the already chosen split variable).
+
+\bigskip
+
+\textbf{Alternative:} Directly compare all available split points in all possible split variables.
 In order to allow for an unbiased selection within this strategy, it has to be accounted for differences such as a varying number of available split points, different scales or variable types by correcting the measurement considered for comparison accordingly.
@@
 \frametitle{Split variable selection}
-{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ education
+\textbf{Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ education
 
@@
 \frametitle{Split variable selection}
-{\bf Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender
+\textbf{Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender
 
@@
 % 
-% {\bf How and where to split?}
+% \textbf{How and where to split?}
 % 
@@
 \frametitle{Split point selection}
-{\bf Exhaustive search:} For each possible split point evaluate a defined objective function in the resulting subgroups and select the split point that minimizes the objective function, i.e., leads to the highest improvement of the model fit.
+\textbf{Exhaustive search:} For each possible split point evaluate a defined objective function in the resulting subgroups and select the split point that minimizes the objective function, i.e., leads to the highest improvement of the model fit.
 
@@
 
-{\bf Alternative:} Two-sample test statistics evaluating the discrepancy between resulting subgroups. 
+\textbf{Alternative:} Two-sample test statistics evaluating the discrepancy between resulting subgroups. 
 \end{frame}
@@
 
-{\bf Goal:} Avoid overfitting.
+\textbf{Goal:} Avoid overfitting.
 
@@
 
-{\bf Pre-pruning:} Internal stopping criterium. Stop splitting when there is no significant association to any of the possible split variable.
+\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when there is no significant association to any of the possible split variable.
 
@@
 
-{\bf Post-pruning:} Grow large tree and prune splits that do not improve the model fit (e.g., via cross-validation or information criteria).
+\textbf{Post-pruning:} Grow large tree and prune splits that do not improve the model fit (e.g., via cross-validation or information criteria).
 
@@
 
-{\bf Pre-pruning:}
+\textbf{Pre-pruning:}
 \begin{itemize}
@@
 
-{\bf Post-pruning:}
+\textbf{Post-pruning:}
 \begin{itemize}
@@
 
-{\bf Note:} The CTree algorithm has been employed for all previous tree illustrations.
+\textbf{Note:} The CTree algorithm has been employed for all previous tree illustrations.
 
@@
 
-{\bf Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.
+\textbf{Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.
 
@@
 
-{\bf Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.
+\textbf{Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.
 
@@
 % 
-% {\bf rpart:} 
+% \textbf{rpart:} 
 % \begin{itemize}
@@
 % 
-% {\bf partykit:} 
+% \textbf{partykit:} 
 % \begin{itemize} 
@@
 
-{\bf Further R packages based on partykit:}
+\textbf{Further R packages based on partykit:}
 \begin{itemize}
@@
 
-{\bf Further R packages:}
+\textbf{Further R packages:}
 \begin{itemize}

### 2020-06-30 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Update computational details
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3010 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fd4ab622ddc4a10971a21c6bffe15b344f0c3bd7
Minor corrections by me and Reto
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3022 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e49422c5085c0dbf88805718c78f6726899ff00a

@@
 focus on the application. In addition, we now use the term `application' 
-instead of `case study' throughout the manuscript and have also renamed Sect.~4.
+instead of `case study' throughout the manuscript and have also renamed Sect.~4 accordingly.
 
@@
 On the following pages a point-to-point response to all comments will be given.
-The attached manuscript highlights the changes in the main body in blue color,
-the `Abstract' is fully revised but displayed in black color.
+The attached manuscript highlights the changes in the text in blue color.
 
@@
 and subscripts in the main body of the manuscript. In the appendix, we keep the
-notation as it is in order to be consistent with the reference Hothorn~\emph{et al.}~2006.
+notation as it is in order to be consistent with the referenced article by Hothorn~\emph{et al.}~2006.
 
@@
 computing time of the different methods on a standard desktop computer within
-the section 'Computational details'.  \todo{Moritz: Run models}
+the section 'Computational details'.
 
@@
 we are not entitled to distribute, we are unfortunately not able to provide
-self-contained computer code in order to reproduce our application. However, as you have already
+code and data to reproduce the application. However, as you have already
 pointed out the implementation of circular regression trees and forests is
-provided in a freely available R-package with some minimal examples.
+provided in a freely available R-package with examples included in the documentation.
 Independent of this publication, we are planning to provide an updated online
-platform for the whole `partykit' project in the future.
+platform for the whole `partykit' project in the near future.
 
@@
 Thanks for pointing out that the citation here is not clearly assigned. We have therefore
-listed the references below the statement again.
+listed the references at the end of the sentence once more.
 

### 2020-06-30 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor corrections by me and Reto
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3022 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e49422c5085c0dbf88805718c78f6726899ff00a

--- 2020-04-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
@@
 %NEW VERSION:
-Circular data occur in a wide range of scientific fields, but the methodology
+Circular data occur in a wide range of scientific fields, however the methodology
 for distributional modeling and probabilistic forecasting of circular response
-variables is rather limited. While most of the existing methods are built on
+variables is rather limited. While most of the existing methods are built within
 the framework of generalized linear and additive models, which are often
@@
 regression trees yielding probabilistic forecasts based on the von~Mises
-distribution. The resulting tree-based models use the available covariates only
+distribution. The resulting tree-based models use the available covariates 
 for partitioning the data into sufficiently homogeneous subgroups, allowing for
@@
 illustrated in an application on short-term probabilistic wind direction
-forecasting for Innsbruck and Vienna Airport in order to direct airplanes to a
+forecasting for Innsbruck and Vienna Airport required to direct airplanes to a
 safe landing. The predictive skill of the novel approaches is benchmarked with
-an additive regression model plus a persistency and a climatology model,
+an additive regression model plus a persistency and a climatological model,
 employing the circular continuous ranked probability score.}
@@
 predictor to a restricted interval of length~$2\pi$. This generalized linear
-model (GLM) type approach was further extended by \cite{Fisher+Lee:1992} and
+model~(GLM) type approach was further extended by \cite{Fisher+Lee:1992} and
 subsequently by \cite{Fisher:1993} introducing independent GLMs for either the
@@
 This novel approach to circular regression trees and forests complements the
-literature on tree-based circular modeling. \cite{Lund:2002} already introduced a
-circular regression tree algorithm where binary splits are made based on an
+literature on tree-based circular modeling. \cite{Lund:2002} already introduced
+a circular regression tree algorithm where binary splits are made based on an
 angular distance measure capturing node homogeneity. However, this only models
@@
 full probabilistic distribution would also allow uncertainty in the forecast to
-be estimated \citep{Gneiting:2008}. 
-%However, this only models
-%changes in the conditional mean but not the conditional variance or full
-%probabilistic distribution which would allow for also considering uncertainty
-%in forecasts \citep{Gneiting:2008}. 
-In the field of image based object direction estimation,
-\citet{Hara+Chellappa:2017} presented a distribution-based tree-building 
-algorithm by applying $k$-means clustering to the target space where
-the number of clusters $k$ is adaptively determined by the Bayes
-information criterion employing the von~Mises distribution. Despite their
-distribution-based splitting approach to circular response variables, they
-solely predict the mode of the distribution, thus, providing no full
-probabilistic forecasts. Combining these previous ideas for circular trees with the
-concept of distributional modeling, we introduce a fully distributional tree
-approach that not only considers splits in all distribution parameters, but also yields a
-full probabilistic predictive model as proposed by \citet{Gneiting:2008}. In
-addition, as a natural extension, ensembles or forests of circular regression
-trees are presented in this study, showing that these can further improve
-predictive power.
+be estimated \citep{Gneiting:2008}.  In the field of image based object
+direction estimation, \citet{Hara+Chellappa:2017} presented a
+distribution-based tree-building algorithm by applying $k$-means clustering to
+the target space where the number of clusters $k$ is adaptively determined by
+the Bayes information criterion employing the von~Mises distribution. Despite
+their distribution-based splitting approach to circular response variables,
+they solely predict the mode of the distribution, thus, providing no full
+probabilistic forecasts. Combining these previous ideas for circular trees with
+the concept of distributional modeling, we introduce a fully distributional
+tree approach that not only considers splits in all distribution parameters,
+but also yields a full probabilistic predictive model as proposed by
+\citet{Gneiting:2008}. In addition, as a natural extension, ensembles or
+forests of circular regression trees are presented in this study, showing that
+these can further improve predictive power.
 
@@
 
-In most generalized linear model (GLM) or generalized additive model (GAM)
+In most generalized linear model~(GLM) or generalized additive model~(GAM)
 approaches to circular regression the location parameter $\mu$ depends on
@@
     an additional manual adjustment, such as an approximation by a finite Fourier expansion
-    (cf., \citealt{Lund:1999}), we use only the linear components of the lagged
+    (cf., \citealt{Lund:1999}), we only use the linear components of the lagged
     2-dimensional wind vector~$(u, v)^{\top}$ and the lagged wind speed $spd$ as
@@
 To compare the predictive performance of all proposed models, a circular
-analogue of the continuous ranked probability score (CRPS) as introduced by
+analogue of the continuous ranked probability score~(CRPS) as introduced by
 \cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
@@
   \caption{As Figure~\ref{fig:boxplot_crpsraw}, but showing CRPS skill scores
-    with the climatology model as reference. Skill scores are in percent; positive
+    with the climatological model as reference. Skill scores are in percent; positive
     values indicate improvements over the reference.} 
@@
 shifts, and potential interactions in covariates without prespecification of
-such effects. As supported by the application on probabilistic wind direction
+such effects. As supported by the presented application on probabilistic wind direction
 forecasting, this can be particularly useful for modeling a highly fluctuating
@@
 
-%\newpage
+The computation time on a standard computer is in the order of 10\,milliseconds
+for the circular fit (used for the persistency and the climatological model),
+seconds for the estimation of the linear model and the circular tree, and from
+a few minutes up to an hour for the estimation of the circular forest; the latter
+strongly depends on the number of covariates employed and, hence, varies
+between the two locations included in the application. However, for a single station and
+forecast step, the persistency and linear models must be re-estimated for each
+time of interest, and independent climatological models have to be fitted for
+all times of interest within a single calendar year, whereas for all times of interest only a single 
+circular tree and a single circular forest model is required.
+
 \bibliography{ref.bib}

### 2020-06-30 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Touchups and put Equation in brackets
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3023 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a8d1b637a940e6959d8d30df4a83a3805ff774b9

@@
 %NEW VERSION:
-Circular data occur in a wide range of scientific fields, however the methodology
-for distributional modeling and probabilistic forecasting of circular response
-variables is rather limited. While most of the existing methods are built within
-the framework of generalized linear and additive models, which are often
-challenging to optimize and interpret, we propose a distributional approach for
-regression trees yielding probabilistic forecasts based on the von~Mises
-distribution. The resulting tree-based models use the available covariates 
-for partitioning the data into sufficiently homogeneous subgroups, allowing for
-a simple circular distributional fit in each subgroup. These circular
-regression trees are straightforward to interpret, can capture nonlinear
-effects and interactions, and automatically select the relevant covariates that
-are associated with changes in any distribution parameter. Ensembles or
-forests of such circular trees can further improve the forecasts by
-regularizing and stabilizing the model. The newly developed methodology is
+Circular data occur in a wide range of scientific fields, however the
+methodology for distributional modeling and probabilistic forecasting of
+circular response variables is rather limited. While most of the existing
+methods are built within the framework of generalized linear and additive
+models, which are often challenging to optimize and interpret, we propose a
+distributional approach for regression trees yielding probabilistic forecasts
+based on the von~Mises distribution. The resulting tree-based models use the
+available covariates for partitioning the data into sufficiently homogeneous
+subgroups, allowing for a simple circular distributional fit in each subgroup.
+These circular regression trees are straightforward to interpret, can capture
+nonlinear effects and interactions, and automatically select the relevant
+covariates that are associated with changes in any distribution parameter.
+Ensembles or forests of such circular trees can further improve the forecasts
+by regularizing and stabilizing the model. The newly developed methodology is
 illustrated in an application on short-term probabilistic wind direction
@@
 location and scale parameters conditional on the respective sets of covariates
-until convergence \citep{Fisher+Lee:1992, Fisher:1993}. While all of these models are built on well-elaborated
-theory, their application in practice remains very challenging, mainly due to
-the complexity encountered in optimizing the corresponding log-likelihood
-function which is not globally concave. Therefore, highly-informative starting
-values are crucial for such circular GLMs to converge
-\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
-this strong dependence on appropriate initial values,
-\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a homoscedastic
-GLM for circular data. However, apart from potential difficulties in the
-optimization procedure of circular GLMs, the interpretation of the underlying
-additive effects is often challenging as well because the link function is
-highly nonlinear and the representation of smooth transitions on the unit
-circle is not straightforward. For example, the same rotation can be obtained
-in either positive or negative direction on the circle leading to an ambiguous
-interpretation.
+until convergence \citep{Fisher+Lee:1992, Fisher:1993}. While all of these
+models are built on well-elaborated theory, their application in practice
+remains very challenging, mainly due to the complexity encountered in
+optimizing the corresponding log-likelihood function which is not globally
+concave. Therefore, highly-informative starting values are crucial for such
+circular GLMs to converge \citep{Pewsey+Neuhaeuser+Ruxton:2013,
+Gill+Hangartner:2010}. In order to avoid this strong dependence on appropriate
+initial values, \cite{Mulder+Klugkist:2017} present a Bayesian alternative of a
+homoscedastic GLM for circular data. However, apart from potential difficulties
+in the optimization procedure of circular GLMs, the interpretation of the
+underlying additive effects is often challenging as well because the link
+function is highly nonlinear and the representation of smooth transitions on
+the unit circle is not straightforward. For example, the same rotation can be
+obtained in either positive or negative direction on the circle leading to an
+ambiguous interpretation.
 
@@
 distribution in each terminal node offering a wide range of statistical
-inference. In addition, the employed tree structure can capture
-non-additive effects while forests enable the modeling of smooth changes.
-Furthermore, covariates and their possible interactions do not need to be
-specified in advance as they are selected automatically in the recursive
-partitioning algorithm.
+inference. In addition, the employed tree structure can capture non-additive
+effects while forests enable the modeling of smooth changes.  Furthermore,
+covariates and their possible interactions do not need to be specified in
+advance as they are selected automatically in the recursive partitioning
+algorithm.
 
@@
 observation and fitted parameter. Then, the optimization problem in
-Equation~\ref{eq:optim} can alternatively be specified as 
+Equation~(\ref{eq:optim}) can alternatively be specified as 
 %%
@@
 %%
-Figure~\ref{fig:densities} depicts a von~Mises model for circular data in $[0, 2
-\pi)$ fitted by maximum likelihood, either using a linearized (left) or
-circular (right) scale. In both cases, the empirical histogram (gray bars) is
-shown along with the fitted density (red line) and estimated location parameter
-(red hand). However, this distributional model only considers the circular
-response variable but no covariate. Of course, including covariates is of
-interest in a regression setup for forecasting.
+Figure~\ref{fig:densities} depicts a von~Mises model for circular data in $[0,
+2 \pi)$ fitted by maximum likelihood, either using a linearized~(left) or
+circular~(right) scale. In both cases, the empirical histogram (gray bars) is
+shown along with the fitted density (red line) and estimated location
+parameter~(red hand). However, this distributional model only considers the
+circular response variable but no covariate. Of course, including covariates is
+of interest in a regression setup for forecasting.
 
@@
 %
-The link
-function transforms the additive predictor to an interval of length $2\pi$.
-Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
-\citet{Fisher+Lee:1992}. They also developed a heteroscedastic version by
-combining two individual GLMs, each for one of the parameters~$\mu$ and~$\kappa$. 
-This provides a first approach to a fully probabilistic regression
-model for circular data, albeit the parameters are not regressed simultaneously
-on covariates as in the more general framework provided by generalized additive
-models for location, scale, and shape \citep[GAMLSS,
+The link function transforms the additive predictor to an interval of length
+$2\pi$. Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested
+by \citet{Fisher+Lee:1992}. They also developed a heteroscedastic version by
+combining two individual GLMs, each for one of the parameters~$\mu$
+and~$\kappa$. This provides a first approach to a fully probabilistic
+regression model for circular data, albeit the parameters are not regressed
+simultaneously on covariates as in the more general framework provided by
+generalized additive models for location, scale, and shape \citep[GAMLSS,
 ][]{Rigby+Stasinopoulos:2005}. Nevertheless, other circular additive models
 share the previously discussed difficulties induced by the characteristics of
-the log-likelihood function and the strongly nonlinear link function. In general,
-for additive models a proper model specification can be very challenging, particularly for a high number of
-covariates and no information on possible interactions. Moreover, the additive
-structure might impose a smooth effect even if the true underlying effect is an
-abrupt shift, which occurs, e.g., in atmospheric wind fields.
+the log-likelihood function and the strongly nonlinear link function. In
+general, for additive models a proper model specification can be very
+challenging, particularly for a high number of covariates and no information on
+possible interactions. Moreover, the additive structure might impose a smooth
+effect even if the true underlying effect is an abrupt shift, which occurs,
+e.g., in atmospheric wind fields.
 
@@
 obtained by evaluating the score function~$s(\cdot)$ for each individual
-observation at the parameter estimates (Equation~\ref{eq:scores}). For the
+observation at the parameter estimates (Equation~(\ref{eq:scores})). For the
 von~Mises distribution with its two distribution parameters ($\mu$ and
@@
   \includegraphics{_plot_circforest_stationmap_combined.pdf}
-  \caption{Overview of the study area for Innsbruck Airport (left) and Vienna International Airport (right). 
-    For Innsbruck, $4$~stations at the airport and $6$~stations along the
-    intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
-    and $13$~stations within the vicinity of approximately 30\,km are used.
-    Elevation data are obtained from the TandDEM-X digital elevation model with a 
-    horizontal resolution of $90$~m \citep{Wessel:2018}.}
+  \caption{Overview of the study area for Innsbruck Airport (left) and Vienna
+    International Airport (right). For Innsbruck, $4$~stations at the airport and
+    $6$~stations along the intersecting valleys are used, whereas, for Vienna,
+    $9$~stations at the airport and $13$~stations within the vicinity of
+    approximately 30\,km are used. Elevation data are obtained from the TandDEM-X
+    digital elevation model with a horizontal resolution of $90$~m
+    \citep{Wessel:2018}.}
   \label{fig:overview} 
@@
     response values within an interval of length $2\pi$. Since in the
-    implementation of \cite{Mulder+Klugkist:2017} the accommodation of circular covariates would require
-    an additional manual adjustment, such as an approximation by a finite Fourier expansion
-    (cf., \citealt{Lund:1999}), we only use the linear components of the lagged
-    2-dimensional wind vector~$(u, v)^{\top}$ and the lagged wind speed $spd$ as
-    predictor variables. Thus, the model formula for the location
-    parameter $\mu$ of the von~Mises distribution can be written as:
+    implementation of \cite{Mulder+Klugkist:2017} the accommodation of circular
+    covariates would require an additional manual adjustment, such as an
+    approximation by a finite Fourier expansion (cf., \citealt{Lund:1999}), we only
+    use the linear components of the lagged 2-dimensional wind vector~$(u,
+    v)^{\top}$ and the lagged wind speed $spd$ as predictor variables. Thus, the
+    model formula for the location parameter $\mu$ of the von~Mises distribution
+    can be written as:
     %%
@@
     predictive von~Mises distribution for $+1$\,h and $+3$\,h forecasts at the
-    Innsbruck Airport and Vienna International Airport. Each box-and-whisker contains $24$~hourly scores
-    for each of the $12$~months averaged over the $5$~validation years which
-    yields a total of $288$~yearly mean values. The scores are shown for the
-    climatological, persistency and the linear model as well as for the circular
-    regression tree and forest.}
+    Innsbruck Airport and Vienna International Airport. Each box-and-whisker
+    contains $24$~hourly scores for each of the $12$~months averaged over the
+    $5$~validation years which yields a total of $288$~yearly mean values. The
+    scores are shown for the climatological, persistency and the linear model as
+    well as for the circular regression tree and forest.}
   \label{fig:boxplot_crpsraw} 
@@
 The performance of the novel circular regression trees and forests is assessed
-in an application on shortterm probabilistic wind direction forecasting at two
+in an application on short-term probabilistic wind direction forecasting at two
 airports with different site characteristics. As benchmark models,
@@
 lagged response values provide already highly skillful estimates leading to a
-good performance of the persistency model as observed for shortterm wind
+good performance of the persistency model as observed for short-term wind
 direction forecasts in the application. While in these cases the trees and forests
@@
 seconds for the estimation of the linear model and the circular tree, and from
-a few minutes up to an hour for the estimation of the circular forest; the latter
-strongly depends on the number of covariates employed and, hence, varies
-between the two locations included in the application. However, for a single station and
-forecast step, the persistency and linear models must be re-estimated for each
-time of interest, and independent climatological models have to be fitted for
-all times of interest within a single calendar year, whereas for all times of interest only a single 
-circular tree and a single circular forest model is required.
+a few minutes up to an hour for the estimation of the circular forest; the
+latter strongly depends on the number of covariates employed and, hence, varies
+between the two locations included in the application. However, for a single
+station and forecast step, the persistency and linear models must be
+re-estimated for each time of interest, and independent climatological models
+have to be fitted for all times of interest within a single calendar year,
+whereas for all times of interest only a single circular tree and a single
+circular forest model is required.
 

### 2020-07-01 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
finished algorithm and split selection
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3025 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e30fa4f5005d0781528a6a4ea14c06bd2ed8b4c6

@@
 
-\textbf{Homogeneity:} Partitioning the data into subgroups aims at allowing for a better model fit in each resulting subset.
-
-\bigskip
-
-\textbf{Model:} The model fitted to a response $Y$ in each subgroup optimizes an objective function, i.e., it can be a simple constant fit (e.g.
-average value or proportion, minimizing deviations) but also a more complex model (possibly including additional regressors, and, e.g., minimizing the residual sum of squares or the negative log-likelihood).
-
-\bigskip
-
-\textbf{Trade-off:} Complexity in the tree structure vs.\ complexity in the fitted models.\\
-Simple models in the nodes (such as constant fits) often lead to large trees while more complex models can compensate a higher variation within the nodes allowing for less complex trees.
-
-% \begin{itemize}
-% \item Choose an indicator or a measure of homogeneity within subgroups.
-% %% FIXME: discrepancy measure, loss function
-% \item Find the split which optimizes this measure in the resulting subgroups.
-% \end{itemize}
-% 
-% \bigskip
-% 
-% \bigskip
-% 
-% \textbf{Examples of discrepancy measures:} 
-% \begin{itemize}
-% \item For regression tree (numeric response): Deviations from an average value, residual sum of squares.
-% \item For classification tree (categorical response): Number of misclassifications.
-% \end{itemize}
+\textbf{Original idea:} Trees are purely algorithmic models without assumptions.
+\begin{itemize}
+  \item Data-driven ``learning'' of homogenous subgroups.
+  \item Simple constant fit in each group, e.g., an average or proportion.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Subsequently:} Trees are well-suited for combination with classical models.
+\begin{itemize}
+  \item Group-wise models, e.g., fitted by least squares or maximum likelihood.
+  \item Model-based learning accounts for model differences across subgroups.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Trade-off:} 
+\begin{itemize}
+  \item Assume simple model and learn larger tree \emph{vs.}
+  \item More complex model and potentially smaller tree.
+\end{itemize}
+
 \end{frame}
@@
 \item Fit a model to the response $Y$.
-\item Assess association of $Y$ (or a corresponding transformation / goodness-of-fit measure) and each possible split variable $X_j$.
-\item Split sample along the $X_{j^{\ast}}$ with strongest association: Choose breakpoint with highest improvement of the model fit.
+\item Assess association of $Y$ (or a corresponding transformation/goodness-of-fit measure) and each possible split variable $X_j$.
+\item Split sample along the $X_{j^{\ast}}$ with strongest association: Choose split point with highest improvement of the model fit.
 \item Repeat steps 1--3 recursively in the subsamples until some stopping criterion is met.
-\item Optionally: Reduce the size of the tree through post-pruning, i.e., by cutting off branches that do not improve the performance significantly.
+\item \emph{Optionally:} Reduce size of the tree by pruning branches of splits that do not improve the model fit sufficiently.
 \end{enumerate}
@@
 \begin{frame}[fragile]
-\frametitle{Tree algorithm: Building blocks}
-
-\textbf{Examples of possible transformations of the response $Y$:} 
+\frametitle{Tree algorithm}
+
+\textbf{Specific algorithms:} Many (albeit not all) can be derived from the base algorithm
+by combining suitable building blocks.
+
+\bigskip
+\pause
+
+\textbf{Models for $Y$:} Simple constant fits vs.\ more complex statistical models.
+
+\bigskip
+\pause
+
+\textbf{Goodness of fit:} Suitable measure depends on type of response variable(s) $Y$
+and the corresponding model.
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Tree algorithm}
+
+\textbf{Goodness of fit:}
 \begin{itemize}
@@
   \begin{itemize}
-\item Ranks of $Y$.
+    \item $Y$ or ranks of $Y$.
     \item (Absolute) deviations $Y - \bar Y$.
@@
   \end{itemize}
+  \pause
   \item Categorical response: 
   \begin{itemize}
-\item Number of misclassifications (Gini impurity).
-\end{itemize}
-\end{itemize}
-
-\medskip
-
-\textbf{Possible association tests:} 
-$\chi^2$-test, Pearson correlation test, two-sample t-test, ANOVA, signed-rank tests, rank-sum tests, \ldots (depending on the scale and type of variables)
-
-\medskip
-
-\textbf{Split point selection:}
-Exhaustive search over all possible split points or two-sample test statistics.
-
-\medskip
+    \item Dummy variables for categories.
+    \item Number of misclassifications.
+    \item Gini impurity.
+  \end{itemize}
+  \pause
+  \item Survival response:
+  \begin{itemize}
+    \item Log-rank scores.
+  \end{itemize}
+  \pause
+  \item Parametric model:
+  \begin{itemize}
+    \item Residuals.
+    \item Model scores (gradient contributions).
+  \end{itemize}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Tree algorithm}
+
+\textbf{Split variable selection:} Optimize some criterion over all $X_j$ ($j = 1, \dots, m$).
+\begin{itemize}
+  \item Objective function (residual sum of squares, log-likelihood, misclassification rate, impurity, \dots).
+  \item Test statistic or corresponding $p$-value.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Split point selection:} Optimize some criterion over all (binary) splits in $X_{j*}$.
+\begin{itemize}
+  \item Objective function.
+  \item Two-sample test statistic or corresponding $p$-value.
+\end{itemize}
+
+\bigskip
+\pause
 
 \textbf{Stopping criteria:}
-Minimal number of observations per node, tree depth, minimal $p$-value, \ldots
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Tree algorithm}
-\vspace{0.2cm}
-
-\textbf{How and where to split?}
-
-\vspace{0.4cm}
-
-\begin{enumerate}
-\item Selection of the best split variable.
-\vspace{0.2cm}
-% \item \textcolor{gray}{Selection of the corresponding best split point.}
-\item Selection of the corresponding best split point.
-\end{enumerate}
+\begin{itemize}
+  \item Constraints: Number of observations per node, tree depth, \dots
+  \item Lack of improvement: Significance, information criteria, \dots
+\end{itemize}
+
 \end{frame}
@@
 \subsection{Split variable selection}
+
 \begin{frame}[fragile]
 \frametitle{Split variable selection}
-\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Gender
+
+\textbf{Idea:}
+\begin{itemize}
+  \item Select variable $X_j$ ($j = 1, \dots, m$) most associated with heterogeneity in $Y$.
+  \item Heterogeneity captured by goodness-of-fit measure.
+  \item \emph{Often:} Maximum association over all possible binary splits.
+  \item \emph{Alternatively:} Overall association.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Split variable selection}
+
+\textbf{Potential bias:} Variables with many potential splits may yield greater
+association ``by chance'', e.g., continuous $X_j$ or categorical with many levels.
+
+\bigskip
+\pause
+
+\textbf{Unbiased recursive partitioning:} Accounts for potential random variation by
+employing $p$-values from appropriate statistical tests.
+
+\bigskip
+\pause
+
+\textbf{Possible tests:} Depend on scales of $Y$, $X_j$, and the adopted model.
+\begin{itemize}
+  \item $\chi^2$ test, Pearson correlation test, two-sample t-test, ANOVA.
+  \item Maximally-selected two-sample tests.
+  \item Parameter instability tests.
+  \item \dots
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Split variable selection}
+
+\textbf{Examples:}
+\begin{itemize}
+  \item Titanic survival and wage determinants data.
+  \item Selection of first split variable in root node.
+  \item Employ classical statistical tests.
+  \item Does not exactly match a particular tree algorithm but similar to CTree.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Split variable selection}
+\textbf{Assess pairwise associations:} \code{Survived ~ Gender} (Titanic).
 
@@
 \frametitle{Split variable selection}
-\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Age
+\textbf{Assess pairwise associations:} \code{Survived ~ Age} (Titanic).
 
@@
 \frametitle{Split variable selection}
-\textbf{Assess pairwise associations} (Titanic): \ Survived $\sim$ Class
+\textbf{Assess pairwise associations:} \code{Survived ~ Class} (Titanic).
 
@@
 \frametitle{Split variable selection}
-{\small Titanic:  \quad $Survived \; \sim  \; Gender + Age + Class$}
+
 \begin{center}
-\setkeys{Gin}{width=0.9\textwidth}
+\setkeys{Gin}{width=0.95\textwidth}
 <<echo=FALSE, fig=TRUE, width=11>>=
@@
 \frametitle{Split variable selection}
-
-%\vspace{0.2cm}
-
-\textbf{Idea:}
-Employ statistical tests for the split variable selection.\\
-Evaluating associations on the unified $p$-value scale allows for an unbiased comparison of possible split variables on different scales and of different types.
-
-\bigskip
-
-\textbf{Unbiased recursive partitioning:} Separation of split variable selection (based on $p$-values) and split point selection (only within the already chosen split variable).
-
-\bigskip
-
-\textbf{Alternative:} Directly compare all available split points in all possible split variables.
-In order to allow for an unbiased selection within this strategy, it has to be accounted for differences such as a varying number of available split points, different scales or variable types by correcting the measurement considered for comparison accordingly.
-However, this has often not been included in classic tree algorithms.
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Split variable selection}
-\textbf{Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ education
+\textbf{Assess pairwise associations:} \code{log(wage) ~ education} (Wages).
 
@@
 \frametitle{Split variable selection}
-\textbf{Assess pairwise associations} (Wages, CPS1985): \ log(wage) $\sim$ gender
+\textbf{Assess pairwise associations:} \code{log(wage) ~ gender} (Wages).
 
@@
 \frametitle{Split variable selection}
-{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
+
 \begin{center}
@@
 \end{center}
-\end{frame}
-
-% \begin{frame}[fragile]
-% \frametitle{Tree algorithm}
-% \vspace{0.2cm}
-% 
-% \textbf{How and where to split?}
-% 
-% \vspace{0.4cm}
-% 
-% \begin{enumerate}
-% %\item \textcolor{gray}{Selection of the best split variable.} \checkmark
-% \item Selection of the best split variable.
-% \vspace{0.2cm}
-% \item Selection of the corresponding best split point.
-% \end{enumerate}
-% \end{frame}
-
-\subsection{Split point selection\\ 
-\&\\ 
-pruning}
+
+\end{frame}
+
+\subsection{Split point selection \& pruning}
 
@@
 \frametitle{Split point selection}
-\textbf{Exhaustive search:} For each possible split point evaluate a defined objective function in the resulting subgroups and select the split point that minimizes the objective function, i.e., leads to the highest improvement of the model fit.
-
-\vspace{0.4cm}
-
-\textbf{Alternative:} Two-sample test statistics evaluating the discrepancy between resulting subgroups. 
+
+\textbf{Idea:}
+\begin{itemize}
+  \item Split $Y$ into most homogenous subgroups with respect to selected $X_{j*}$.
+  \item Homogeneity captured by goodness-of-fit measure.
+  \item \emph{Often:} Consider only binary splits.
+  \item Trivial for binary $X_{j*}$.
+  \item Otherwise typically exhaustive search over all binary splits.
+  \item Possibly already done in split variable selection.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Goodness-of-fit measure:} Depends on scale of $Y$ and the adopted model.
+\begin{itemize}
+  \item Objective function (residual sum of squares, log-likelihood, misclassification rate, impurity, \dots).
+  \item Two-sample test statistic or corresponding $p$-value.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Split point selection}
+
+\textbf{Example:}
+\begin{itemize}
+  \item Wage determinants data.
+  \item Selection of best split point in education in root node.
+  \item Residual sum of squares vs.\ normal log-likelihood.
+  \item Two-sample ANOVA-type $\chi^2$ test and corresponding $p$-value.
+\end{itemize}
+
 \end{frame}
@@
 \frametitle{Split point selection}
-{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
+
 \begin{center}
@@
 
-\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when there is no significant association to any of the possible split variable.
+\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when there is no significant association to any of the possible split variables $X_j$.
 
@@
 
-\subsection{Algorithms \\
-\& \\
-implementations}
+\subsection{Algorithms \& implementations}
 

### 2020-07-01 pkg/partykit/inst/ULGcourse-2020/exercises/exercise_german_credit.Rmd
added exercises using Titanic and South German Credit data
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3026 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d1522a1cf76a7f94b18d993d46dfff3ac075fab2

---
title: "Exercise: South German Credit"
output: html_document
---

```{r, include = FALSE}
file <- "data/german.rds"
stopifnot(file.exists(file))
data <- readRDS(file)
head(data)
```

In this exercise we will use the 'South German Credit' data set. It contains a classification
of the credit risk of 1000 individuals into 'good' and 'bad' together with 20 additional attributes.

Simply download the file `r xfun::embed_file(file, text = "german.rds")` by clicking or
download it from OpenOLAT or from the corresponding homepage
<http://archive.ics.uci.edu/ml/datasets/South+German+Credit>
which also provides more detailed information on the data set.

We can import/read this file using `data <- readRDS(...)`. The file
contains the following information:

* `status`: status of the debtor's checking account with the bank (factor).
* `duration`: credit duration in months (integer).
* `credit_history`: history of compliance with previous or concurrent credit contracts (factor).
* `purpose`: purpose for which the credit is needed (factor).
* `amount`: credit amount in DM (integer).
* `savings`: debtor's savings (factor).
* `employment_duration`: duration of debtor's employment with current employer (factor; discretized quantitative).
* `installment_rate`: credit installments as a percentage of debtor's disposable income (ordered factor; discretized quantitative).
* `personal_status_sex`: combined information on sex and marital status (factor; sex cannot be recovered from the variable, because male singles and female non-singles are coded with the same code (2); female widows cannot be easily classified, because the code table does not list them in any of the female categories).
* `other_debtors`: Is there another debtor or a guarantor for the credit? (factor).
* `present_residence`: length of time (in years) the debtor lives in the present residence (ordered factor; discretized quantitative).
* `property`: the debtor's most valuable property, i.e. the highest possible code is used. Code 2 is used, if codes 3 or 4 are not applicable and there is a car or any other relevant property that does not fall under variable `savings` (factor).
* `age`: age in years (integer).
* `other_installment_plans`: installment plans from providers other than the credit-giving bank (factor).
* `housing`: type of housing the debtor lives in (factor)
* `number_credits`: number of credits including the current one the debtor has (or had) at this bank (ordered factor, discretized quantitative).
* `job`: quality of debtor's job (ordinal)
* `people_liable`: number of persons who financially depend on the debtor (i.e., are entitled to maintenance) (factor, discretized quantitative).
* `telephone`: Is there a telephone landline registered on the debtor's name? (factor; remember that the data are from the 1970s)
* `foreign_worker`: Is the debtor a foreign worker? (factor)
* `credit_risk`: Has the credit contract been complied with (good) or not (bad)? (factor)

# The Tasks

We would like to find out how the credit risk of a person depends on the provided additional attributes of the person and the considered credit itself.
By employing a tree model we are looking for a separation into homogeneous subgroups based on the additional information.

Our response in this case is the binary variable `credit_risk`, as covariates we have 20 additional variables (17 categorical, 3 numeric).

Apply the CTree algorithm to build the tree models described in the following steps:

* Load the data set `"german.rds"`.
* Build a tree using pre-pruning with a significance level of 0.04, with a maximum depth of 4 levels and the segment size of terminal nodes not being smaller than 15.
* Evaluate the performance on the learning data by calculating the corresponding confusion matrix. How large is the misclassification rate?
* Predict the credit risk of a new client who doesn't have a checking account, has never taken a credit before, is a 35-year-old married male who is a skilled employee working in a business in his home town for already 5 years now, and who plans to take one credit of 1200 DM for repairment in his own house where he moved in 2 years ago. The credit duration is 24 months, the installment rate is 30 %. There is no information provided on his savings and no registered telephone number on the client's name. There are no other installment plans or other debtors and there is no other person depending financially on the client.
Does it have an impact on the prediction if the duration is reduced to only 12 months?
* Separate the data set into a learning set (2/3 of the full data) and a testing set (1/3 of the full data). Build a tree on the learning data set and predict the credit risk on the testing data set. Evaluate the performance based on the number of misclassifications. How do parameters such as a minimal segment size or the significance level applied for pre-pruning influence the performance?

```{r, include = FALSE}
# data <- readRDS("data/german.rds")
formula <- credit_risk ~  status + duration + credit_history + purpose  + amount + savings + employment_duration + installment_rate + personal_status_sex + other_debtors + present_residence + property + age + other_installment_plans + housing + number_credits + job + people_liable + telephone + foreign_worker

library("partykit")
ct <- ctree(formula, data = data)
ct <- ctree(formula, data = data, control = ctree_control(alpha = 0.04, minbucket = 15, maxdepth = 4))

library("caret")
caret::confusionMatrix(data$credit_risk, predict(ct, newdata = data))

newclient <- data.frame(status = "no checking account",
                        duration = 24,
                        credit_history = "no credits taken/all credits paid back duly",
                        purpose = "repairs",
                        amount = 1200,
                        savings = "unknown/no savings account",
                        employment_duration = "4 <= ... < 7 yrs",
                        installment_rate =  "25 <= ... < 35",
                        personal_status_sex = "male : married/widowed",
                        other_debtors = "none",
                        present_residence = "1 <= ... < 4 yrs",
                        property = "real estate",
                        age = 35,
                        other_installment_plans = "none",
                        housing = "own",
                        number_credits = "1",
                        job = "skilled employee/official",
                        people_liable = "0 to 2",
                        telephone = "no",
                        foreign_worker = "no"
                        )
predict(ct, newdata = newclient)

newclient2 <- newclient
newclient2$duration <- 12
predict(ct, newdata = newclient2)
```

```{r, include = FALSE, echo = FALSE, out.width = "100%", fig.width = 10, fig.height = 5}
plot(ct)
```
```{r, include = FALSE}
set.seed(4)
trainid <- sample(1:NROW(data), size = 667, replace = FALSE)
train <- data[trainid,]
test <- data[-trainid,]

ctrain <- ctree(formula, data = train)
predtest <- predict(ctrain, newdata = test)

library("caret")
caret::confusionMatrix(test$credit_risk, predtest)

ctrain <- ctree(formula, data = train, control = ctree_control(alpha = 0.01))
plot(ctrain)
predtest <- predict(ctrain, newdata = test)
caret::confusionMatrix(test$credit_risk, predtest)
```

### 2020-07-01 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
small revisions and formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3027 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2c59e2eda706c3cec5c169f8a7d9832bf0c30e67

@@
 \documentclass[11pt,t,usepdftitle=false,aspectratio=169]{beamer}
-\usetheme[nototalframenumber,url,license]{uibk}
+\usetheme[nototalframenumber,license]{uibk}
 
 \title{Random Forests}
-\author{Achim Zeileis, Lisa Schlosser}
-\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
-%\URL{http://www.partykit.org/partykit/}
+\subtitle{Supervised Learning: Algorithmic Modeling}
+\author{Lisa Schlosser, Achim Zeileis}
+
+%% forest header image
 \renewcommand{\headerimage}[1]{%
@@
 
+%% custom subsection slides
+\setbeamercolor*{subsectionfade}{use={normal text},parent={normal text},fg=structure.fg!30!normal text.bg}
+\AtBeginSubsection[]{%
+  \begin{frame}[c]
+    \begin{center}
+      \usebeamercolor[fg]{subsectionfade}
+      \Large \insertsection \\[2ex]
+      \usebeamercolor[fg]{structure}
+      \huge\bfseries\insertsubsection
+    \end{center}
+  \end{frame}
+}
+
 \usepackage[utf8]{inputenc}
@@
 \usepackage{xcolor}
-%\usepackage[cal=boondoxo]{mathalfa}
-%\graphicspath{{plots/}}
 \newcommand{\argmin}{\operatorname{argmin}\displaylimits}
@@
 
-\section{Forests models}
+\section{Random forests}
 
-\subsection{Motivation: Forest models}
+\subsection{Motivation}
 \begin{frame}[fragile]
-\frametitle{Motivation: Forest models}
+\frametitle{Motivation}
 
@@
 \begin{frame}
-\frametitle{Motivation: Forest models}
+\frametitle{Motivation}
 
@@
 
-\subsection{Aggregation of trees}
+\subsection{Aggregation of trees\\
+\&\\
+supsampling}
 
@@
 
-The predicted response $\hat{y}$ or, more generally, the predicted model parameter (vector) $\hat{\theta}$ is the argument of the parameter space $\Theta$ that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{y_i\}_{i=1,\ldots,n}$:
+The predicted response $\hat{y}$ 
+%or, more generally, the predicted model parameter (vector) $\hat{\theta}$ 
+is the argument %of the parameter space $\Theta$ 
+that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{y_i\}_{i=1,\ldots,n}$:
 
@@
 \[
-\hat{\theta}(x) =  \argmin_{\theta \in \Theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)
+\hat{y}(x) =  \argmin_{y} \sum_{i=1}^n w_i(x) \cdot f(y; y_i)
 \]
@@
 %\vspace{-3cm}
-{\bf Note:} This general notation holds for classic random forests averaging over mean values of tree nodes by setting $f(\theta, y_i) = (\theta - y_i)^2$\\
+{\bf Note:} In model-based trees and forests a prediction is made not only for the response but for a model parameter (vector) $\theta$. 
+This more general approach holds for classic random forests averaging over mean values of tree nodes by setting $f(\theta, y_i) = (\theta - y_i)^2$\\
 but also allows for combining quantiles (quantile regression forests) or full distributions as in distributional forests with
@@
 \end{frame}
-
-\subsection{Subsampling}
 
@@
 
-\medskip
+% \medskip
+% 
+% <<call_randomForest, eval=FALSE, echo=TRUE>>=
+% randomForest(formula, data=NULL, ..., subset, na.action=na.fail)
+% randomForest(x, y=NULL,  ...)
+% @
+% 
+% \medskip
+% 
+% Optional control arguments:
+% \begin{itemize}
+% \item \code{ntree}: nr. of trees, 
+% \item \code{mtry}: nr. of randomly sampled split variable candidates, 
+% \item \code{replace}: type of subsetting, 
+% \item \code{sampsize}: size of drawn subsamples, 
+% \item \code{nodesize}: minimum size of terminal nodes, 
+% \item \code{maxnodes}: maximum number of terminal nodes, 
+% \item \ldots 
+% \end{itemize}
 
-<<call_randomForest, eval=FALSE, echo=TRUE>>=
-randomForest(formula, data=NULL, ..., subset, na.action=na.fail)
-randomForest(x, y=NULL,  ...)
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{R software for forest models}
+
+{\bf ranger}:
+\begin{itemize}
+\item RANdom  forest  GEneRator.
+\item R package providing a faster implementation of classic random forests.
+\end{itemize}
+
+<<call_ranger, eval=FALSE, echo=TRUE>>=
+ranger(formula, data)
 @
-
-\medskip
 
@@
 \begin{itemize}
-\item \code{ntree}: nr. of trees, 
-\item \code{mtry}: nr. of randomly sampled split variable candidates, 
-\item \code{replace}: type of subsetting, 
-\item \code{sampsize}: size of drawn subsamples, 
-\item \code{nodesize}: minimum size of terminal nodes, 
-\item \code{maxnodes}: maximum number of terminal nodes, 
+\item \code{num.trees}: number of trees, 
+\item \code{mtry}: number of randomly sampled split variable candidates, 
+\item \code{replace}: sample with replacement, 
+\item \code{sample.fraction}: fraction of observations to sample, 
+\item \code{min.node.size}: minimal size of terminal nodes, 
+\item \code{max.depth}: maximal tree depth, 
+\item \code{splitrule}: splitting rule, 
 \item \ldots 
 \end{itemize}
-
 \end{frame}
@@
 \begin{itemize}
-  \item ranger (classic random forests),
   \item grf (generalized random forests),
@@
 
-\subsection{References}
+%\subsection{References}
 

### 2020-07-01 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
finished revision of tree slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3028 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 937f161b6b4c726f83e9e0db39c1fcf3beaf8185

@@
 
-\textbf{Split point selection:} Optimize some criterion over all (binary) splits in $X_{j*}$.
+\textbf{Split point selection:} Optimize some criterion over all (binary) splits in $X_{j^*}$.
 \begin{itemize}
@@
 \begin{itemize}
-  \item Split $Y$ into most homogenous subgroups with respect to selected $X_{j*}$.
+  \item Split $Y$ into most homogenous subgroups with respect to selected $X_{j^*}$.
   \item Homogeneity captured by goodness-of-fit measure.
   \item \emph{Often:} Consider only binary splits.
-  \item Trivial for binary $X_{j*}$.
+  \item Trivial for binary $X_{j^*}$.
   \item Otherwise typically exhaustive search over all binary splits.
@@
 \item[+] Computationally less expensive as trees are prevented from getting too large.
-\item[-] Performance depends on the power of the selected statistical test.
+\item[--] Performance depends on the power of the selected statistical test.
 \end{itemize}
@@
 \item[+] Can employ information criteria such as AIC or BIC for model-based partitioning. 
-\item[-] Computationally more expensive as very large trees are grown and additional (out-of-bag) evaluations are required.
-\end{itemize}
-
-\end{frame}
-
-\subsection{Algorithms \& implementations}
+\item[--] Computationally more expensive as very large trees are grown and additional (out-of-bag) evaluations are required.
+\end{itemize}
+
+\end{frame}
+
+\subsection{Conditional inference trees}
 
 \begin{frame}
-\frametitle{CTree}
-
-\begin{itemize} 
-\item Conditional inference trees.
-\item Flexible and general framework for unbiased recursive partitioning,
-originally proposed for nonparametric modeling but also allowing for a model based approach via transformations of the response. 
-\item Split variable selection based on conditional permutation tests.
-\item Split point selection based on two-sample test statistics.
-\item Employs pre-pruning.
-\item Implemented in the R package partykit.
-\end{itemize}
-
-\bigskip
-
-\textbf{Note:} The CTree algorithm has been employed for all previous tree illustrations.
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{CTree}
-
-Apply \fct{ctree} to preprocessed \code{ttnc} data:
+\frametitle{Conditional inference trees}
+
+\textbf{CTree:}
+\begin{itemize} 
+  \item Employs general conditional inference (or permutation test) framework for association
+    of $h(Y)$ vs.\ $g(X_j)$.
+  \item Broadly applicable by choosing suitable transformations $h(\cdot)$ and $g(\cdot)$.
+  \item \emph{Examples:} Univariate $Y$ of arbitrary scale, multivariate, model-based, etc.
+  \item \emph{Default:} Pre-pruning based on direct asymptotic tests for split variable selection.
+    Maximally-selected two-sample statistics for split point selection.
+  \item \emph{Software:} R package \emph{partykit} (and previously: \emph{party}).
+  \item \emph{Reference:} Hothorn, Hornik, Zeileis (2006).
+\end{itemize}
+
+\bigskip
+
+\textbf{Note:} CTree used in previous illustrations of classification and regression trees.
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{Illustration:} Titanic survival data.
+
+\bigskip
+
+\textbf{Data:} Preprocessing from four-way contingency table \code{Titanic} in base R.
+
+<<echo=TRUE>>=
+data("Titanic", package = "datasets")
+ttnc <- as.data.frame(Titanic)
+ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
+names(ttnc)[2] <- "Gender"
+@
+
+\bigskip
+
+\textbf{CTree:} With default arguments.
 
 <<ctree, echo=TRUE, eval=FALSE>>=
-ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc)
-@
+library("partykit")
+ct_ttnc <- ctree(Survived ~ Gender + Age + Class, data = ttnc)
+plot(ct_ttnc)
+print(ct_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\setkeys{Gin}{width=0.9\linewidth}
+<<echo=FALSE, eval=TRUE, fig=TRUE, height=6, width=11>>=
+plot(ct_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+<<echo=FALSE, eval=TRUE, results=verbatim>>=
+print(ct_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{Predictions:} Different types.
+
+<<echo=TRUE, results=verbatim>>=
+ndm <- data.frame(Gender = "Male", Age = "Adult", Class = c("1st", "2nd", "3rd"))
+predict(ct_ttnc, newdata = ndm, type = "node")
+predict(ct_ttnc, newdata = ndm, type = "response")
+predict(ct_ttnc, newdata = ndm, type = "prob")
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{Predictions:} Women and children first?
+
+<<echo=TRUE, results=verbatim>>=
+ndf <- data.frame(Gender = "Female", Age = "Adult", Class = c("1st", "2nd", "3rd"))
+ndc <- data.frame(Gender = "Male", Age = "Child", Class = c("1st", "2nd", "3rd"))
+cbind(
+  Male   = predict(ct_ttnc, newdata = ndm, type = "prob")[, 2],
+  Female = predict(ct_ttnc, newdata = ndf, type = "prob")[, 2],
+  Child  = predict(ct_ttnc, newdata = ndc, type = "prob")[, 2]
+)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{Hyperparameters:} See \code{?ctree_control}.
+\begin{itemize}
+  \item \code{alpha}: Significance level for pre-pruning in split variable selection.
+  \item \code{minsplit}: Minimum number of observations in a node for splitting.
+  \item \code{minbucket}: Minimum number of observations in a terminal node.
+  \item \code{maxdepth}: Maximum depth of the tree.
+  \item \code{multiway}: Use multiway splits for unordered factors with $> 2$ levels?
+  \item \dots
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Example:}
+
+<<echo=TRUE, eval=TRUE>>=
+ct_ttnc2 <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
+  alpha = 0.01, minbucket = 5, minsplit = 15, maxdepth = 4)
+plot(ct_ttnc2)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\setkeys{Gin}{width=\linewidth}
+<<echo=FALSE, eval=TRUE, fig=TRUE, width=13, height=6.9>>=
+plot(ct_ttnc2)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{Predictions:} Male children in 1st and 2nd class are now in their own terminal nodes.
+
+<<ctree_predict, echo=TRUE, eval=TRUE, results=verbatim>>=
+predict(ct_ttnc2, newdata = ndc, type = "prob")
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{Evaluations:} ``As usual.''
+\begin{itemize}
+  \item In-sample vs. out-of-sample.
+  \item Scoring rules (e.g., misclassification rates or other loss functions).
+  \item Confusion matrix.
+  \item Receiver operator characteristic (ROC) curve.
+  \item \dots
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{In-sample:} Augment learning data.
+
+<<echo=TRUE, results=verbatim>>=
+ttnc$Fit <- predict(ct_ttnc2, type = "response")
+ttnc$Group <- factor(predict(ct_ttnc2, type = "node"))
+@
+
+\bigskip
+
+\textbf{Confusion matrix:}
+
+<<echo=TRUE, results=verbatim>>=
+xtabs(~ Fit + Survived, data = ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{Terminal nodes:} Recompute empirical survival rates.
+
+<<echo=TRUE, results=verbatim>>=
+tab <- xtabs(~ Group + Survived, data = ttnc)
+prop.table(tab, 1)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Conditional inference trees}
+
+\textbf{More graphics:} \emph{ggplot2} support via \code{autoplot()} method from \emph{ggparty}.
+
+\setkeys{Gin}{width=0.9\linewidth}
+<<echo=FALSE, eval=TRUE, fig=TRUE, width=13, height=6.9>>=
+library("ggparty")
+theme_set(theme_minimal())
+autoplot(ct_ttnc2)
+@
+
+\end{frame}
+
+\subsection{Recursive partitioning}
+\begin{frame}
+\frametitle{Recursive partitioning}
+
+\textbf{RPart:}
+\begin{itemize} 
+  \item Implements classic CART algorithm (classification and regression trees).
+  \item Split variable and split point selection via exhaustive search based on objective functions.
+  \item Hence biased towards split variables with many possible splits.
+  \item Cross-validation-based post-pruning and no pre-pruning.
+  \item \emph{Default:} Gini impurity for classification, residual sum of squares for regression.
+  \item \emph{Software:} R package \emph{rpart}. Similar implementations in other languages (e.g., \emph{scikit-learn}).
+  \item \emph{Reference:} Algorithm by Breiman \emph{et al.} (1984), open-source implementation by Therneau \& Atkinson (1997).
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\textbf{R package:} Part of every standard R installation.
+
+\bigskip
+
+\textbf{Hyperparameters:} \code{?rpart.control}.
+
+\bigskip
+
+\textbf{RPart:} With default arguments.
+
+<<rpart, echo=TRUE, eval=TRUE>>=
+library("rpart")
+rp_ttnc <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
+plot(rp_ttnc)
+text(rp_ttnc)
+print(rp_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\setkeys{Gin}{width=0.9\linewidth}
+<<echo=FALSE, eval=TRUE, fig=TRUE, width=10, height = 6>>=
+plot(rp_ttnc)
+text(rp_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\setkeys{Gin}{width=0.9\linewidth}
+<<echo=FALSE, eval=TRUE, results=verbatim>>=
+print(rp_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\textbf{Coercion:}
+\begin{itemize}
+  \item \code{as.party()} method to coerce \code{rpart} objects to \code{constparty} objects.
+  \item Based on generic infrastructure for recursive partitioning in \emph{partykit}.
+  \item Enables unified interface (print, plot, predict, \dots).
+\end{itemize}
 
@@
 
-Plot:
-
-<<ctree-plot, echo=TRUE, eval=FALSE>>=
-plot(ct)
-@
+<<echo=TRUE>>=
+py_ttnc <- as.party(rp_ttnc)
+plot(py_ttnc)
+print(py_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\setkeys{Gin}{width=0.9\linewidth}
+<<echo=FALSE, eval=TRUE, fig=TRUE, width=10, height = 6>>=
+plot(py_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\setkeys{Gin}{width=0.9\linewidth}
+<<echo=FALSE, eval=TRUE, results=verbatim>>=
+print(py_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\textbf{Post-pruning:}
+\begin{itemize}
+  \item Cost-complexity pruning, similar to using information criteria.
+  \item Complexity parameter \code{cp} controls trade-off between reduction of objective function and
+    tree size.  
+  \item Carried out automatically during fitting.
+  \item Additionally, ten-fold cross-validation carried out as well.
+  \item Results in \code{rpart} object as \code{cptable}.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Here:} Fitted object also has optimal cross-validation error (\code{xerror}).
 
@@
 
-Add control arguments via \fct{ctree\_control}:
-
-<<ctree_control, echo=TRUE, eval=FALSE>>=
-ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
-            control = ctree_control(alpha = 0.01,
-                                    minsplit = 20, 
-                                    maxdepth = 3))
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{CTree}
-
-\setkeys{Gin}{width=0.9\linewidth}
-<<ctree_control_plot, echo=TRUE, eval=TRUE, fig=TRUE, width=10, height=5>>=
-ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
-            control = ctree_control(alpha = 0.01, minsplit = 20, maxdepth = 3))
-plot(ct)
-@
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{CTree}
-
-\textbf{Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.
-
-<<ctree_predict, echo=TRUE, eval=TRUE, results=verbatim>>=
-nd <- data.frame(Class = "2nd", Gender = c("Female", "Male"), Age = "Child")
-nd
-predict(ct, newdata = nd)
-@
-
-\end{frame}
-
+<<echo=TRUE, results=verbatim>>=
+rp_ttnc$cptable
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\textbf{Example:} Employ complexity parameter \code{cp = 0.1} for illustration.
+
+\medskip
+
+<<echo=TRUE, results=verbatim>>=
+prune(rp_ttnc, cp = 0.1)
+@
+
+\end{frame}
+
+\subsection{Model-based recursive partitioning}
 \begin{frame}
-\frametitle{rpart}
-
-\begin{itemize} 
-\item Recursive PARTitioning.
-\item Based on ideas of CART (Classification And Regression Trees).
-\item Split variable and split point selection based on optimizing an objective function.
-\item Classification trees: reduce number of misclassifications.
-\item Regression trees: reduce objective function for numeric response.
-\item Post-pruning can be employed.
-\item Implemented in the R package rpart.
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-Apply \fct{rpart} to preprocessed \code{ttnc} data:
-
-<<rpart, echo=TRUE, eval=TRUE>>=
-rp <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
-@
+\frametitle{Model-based recursive partitioning}
+
+\textbf{MOB:}
+\begin{itemize}
+  \item Parametric model in subgroups (maximum likelihood, least squares, \dots).
+  \item Split variable selection based on asymptotic parameter instability tests.
+  \item Split point selection via exhaustive search based on objective function.
+  \item Significance-based pre-pruning (default) and optionally
+    post-pruning based on information criteria (AIC, BIC, \dots).  
+  \item \emph{Examples:} (Generalized) linear model tree, distributional trees, \dots
+  \item \emph{Software:} R package \emph{partykit} (and previously: \emph{party}), various ``mobsters'' in extension packages.
+  \item \emph{Hyperparameters:} \code{?mob\_control}.
+  \item \emph{Reference:} Zeileis, Hothorn, Hornik (2008).
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Model-based recursive partitioning}
+
+\textbf{Example:} Investigate treatment heterogeneity for ``women and children first.''
 
@@
 
-Standard plot:
-
-<<rpart-plot, echo=TRUE, eval=FALSE>>=
-plot(rp)
-text(rp)
-@
-
-Visualization via partykit:
-
-<<rpart-party-plot, echo=TRUE, eval=FALSE>>=
-plot(as.party(rp))
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-<<rpart-plot_eval, echo=TRUE, eval=FALSE>>=
-plot(rp)
-text(rp)
-@
-
-\vspace{-0.7cm}
-
-\setkeys{Gin}{width=0.9\linewidth}
-<<rpart-plot_eval, echo=FALSE, eval=TRUE, fig=TRUE, width=10, height = 6>>=
-plot(rp)
-text(rp)
-@
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-<<rpart-party-plot_eval, echo=TRUE, eval=FALSE>>=
-plot(as.party(rp))
-@
-
-%\vspace{-0.4cm}
-
-\setkeys{Gin}{width=0.9\linewidth}
-<<rpart-party-plot_eval, echo=FALSE, eval=TRUE, fig=TRUE, width = 12>>=
-plot(as.party(rp))
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-Standard output: 
-<<rpart-print, echo=TRUE, results=verbatim>>=
-rp
-@
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-Output via partykit: 
-<<rpart-party_print, echo=TRUE, results=verbatim>>=
-as.party(rp)
-@
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-\textbf{Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.
-
-%\medskip
-
-<<rpart_pruning, echo=TRUE, eval=TRUE, results=verbatim>>=
-rp_pruned <- prune(rp, cp = 0.01)
-@
-
-\setkeys{Gin}{width=0.6\linewidth}
-<<rpart_pruning_plot, echo=TRUE, eval=TRUE, fig=TRUE, width = 8, height=5>>=
-plot(as.party(rp_pruned))
-@
-% plot(rp_pruned)
-% text(rp_pruned)
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-\textbf{Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.
-
-%\medskip
-
-<<rpart_pruning2, echo=TRUE, eval=TRUE, results=verbatim>>=
-rp_pruned <- prune(rp, cp = 0.02)
-@
-
-\setkeys{Gin}{width=0.6\linewidth}
-<<rpart_pruning2_plot, echo=TRUE, eval=TRUE, fig=TRUE, width = 8, height=5>>=
-plot(as.party(rp_pruned))
-@
-% plot(rp_pruned)
-% text(rp_pruned)
-
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{rpart}
-
-\textbf{Post-pruning:} Cost-complexity pruning of an rpart-object with complexity parameter \texttt{cp}.
-
-%\medskip
-
-<<rpart_pruning3, echo=TRUE, eval=TRUE, results=verbatim>>=
-rp_pruned <- prune(rp, cp = 0.03)
-@
-
-\setkeys{Gin}{width=0.6\linewidth}
-<<rpart_pruning3_plot, echo=TRUE, eval=TRUE, fig=TRUE, width =8, height=5>>=
-plot(as.party(rp_pruned))
-@
-% plot(rp_pruned)
-% text(rp_pruned)
-
-\end{frame}
-
-\begin{frame}
-\frametitle{MOB}
-
-\begin{itemize} 
-\item Model-based recursive partitioning using M-estimation (ML, OLS, CRPS, \ldots).
-\item Split variable selection based on asymptotic parameter instability tests.
-\item Split point selection based on exhaustive search.
-\item Employs pre-pruning, but can also use post-pruning based on AIC or BIC.
-\item Implemented in the R package partykit.
-\end{itemize}
-
-\end{frame}
-
-% \begin{frame}
-% \frametitle{R software for tree models}
-% 
-% \textbf{rpart:} 
-% \begin{itemize}
-% \item R package for Recursive PARTitioning.
-% \item Implementation based on ideas of CART (Classification And Regression Trees).
-% \item Classification trees: reduce number of misclassifications.
-% \item Regression trees: reduce objective function for numeric response.
-% \end{itemize}
-% 
-% \vspace{0.4cm}
-% 
-% \textbf{partykit:} 
-% \begin{itemize} 
-% \item R package offering a toolkit for unbiased recursive partitioning. 
-% \item Implementations of the tree algorithms CTree (Conditional Inference Trees) and MOB (Model-based Recursive Partitioning).
-% \item Provides a very general framework with a unifying infrastructure.
-% \item Can import rpart.
-% \end{itemize}
-% 
-% \end{frame}
-
-\begin{frame}[fragile]
-\frametitle{MOB}
-
-Is ``women and children first'' applied equally for all passenger classes?
-
-\bigskip
-
-Add a factor variable distinguishing women and children from male adults:
 <<ttnc_treatment, echo=TRUE, eval=TRUE>>=
-ttnc <- transform(ttnc, Treatment = factor(Gender == "Female" | Age == "Child", 
-                                           levels = c(FALSE, TRUE),
-                                           labels = c("Male&Adult", "Female|Child")))
-@
-
-\bigskip
-
-Assess differences in the preferential treatment effect over the remaining covariates by applying \fct{glmtree} which employs the MOB algorithm for generalized linear models in the tree nodes:
-
-<<glmtree, echo=TRUE, eval=TRUE, results=verbatim>>=
-glmt <- glmtree(Survived ~ Treatment | Class + Gender + Age,
+ttnc <- transform(ttnc,
+  Treatment = factor(Gender == "Female" | Age == "Child", 
+    levels = c(FALSE, TRUE), labels = c("Male&Adult", "Female|Child")
+  )
+)
+@
+
+\bigskip
+\pause
+
+\textbf{Model:} Tree based on binary logit GLM for survival.
+
+\medskip
+
+<<glmtree, echo=TRUE>>=
+mob_ttnc <- glmtree(Survived ~ Treatment | Class + Gender + Age,
   data = ttnc, family = binomial, alpha = 0.01)
-@
-
-<<glmtree_print, echo=TRUE, eval=FALSE>>=
-glmt
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{MOB}
+plot(mob_ttnc)
+print(mob_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Model-based recursive partitioning}
+
+\setkeys{Gin}{width=0.83\textwidth}
+<<echo=FALSE, eval=TRUE, fig=TRUE, width=9, height=5>>=
+plot(mob_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Model-based recursive partitioning}
+
 \vspace{-0.4cm}
@@
 <<glmtree_print_eval, echo=FALSE, eval=TRUE, results=verbatim>>=
-glmt
-@
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{MOB}
-\setkeys{Gin}{width=0.99\textwidth}
-<<glmtree-plot, echo=TRUE, eval=TRUE, fig=TRUE, width=11, height=5>>=
-plot(glmt)
-@
-\end{frame}
-
+mob_ttnc
+@
+
+\end{frame}
+
+\subsection{Evolutionary learning of globally optimal trees}
 \begin{frame}
-\frametitle{R software for tree models}
-
-\textbf{Further R packages based on partykit:}
-\begin{itemize}
-  \item evtree (evolutionary trees),
-  \item model4you (personalised trees),
-  \item disttree (distributional trees),
-  \item stablelearner (stability assessment),
+\frametitle{Evolutionary learning of globally optimal trees}
+
+\textbf{EvTree:}
+\begin{itemize}
+  \item \emph{Goal:} Globally optimal partition rather than locally optimal split in each step.
+  \item NP-hard optimization problem, attempt solution via evolutionary learning.
+  \item Based on so-called ``fitness'' function.
+  \item Essentially a penalized objective function, similar to information criteria or cost-complexity.
+  \item Thus, no additional pruning step.
+  \item \emph{Software:} R package \emph{evtree} (leverages \emph{partykit}).
+  \item \emph{Hyperparameters:} \code{?evtree\_control}.
+  \item \emph{Reference:} Grubinger, Zeileis, Pfeiffer (2014).
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Evolutionary learning of globally optimal trees}
+
+\textbf{EvTree:} With default parameters.
+
+\medskip
+
+<<echo=TRUE, eval=FALSE>>=
+library("evtree")
+set.seed(1)
+ev_ttnc <- evtree(Survived ~ Gender + Age + class, data = ttnc)
+plot(ev_ttnc)
+print(ev_ttnc)
+@
+
+<<echo=FALSE, eval=TRUE>>=
+library("evtree")
+if(file.exists("ev_ttnc.rds")) {
+  ev_ttnc <- readRDS("ev_ttnc.rds")
+} else {
+  set.seed(1)
+  ev_ttnc <- evtree(Survived ~ Gender + Age + Class, data = ttnc)
+  saveRDS(ev_ttnc, file = "ev_ttnc.rds")
+}
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Evolutionary learning of globally optimal trees}
+
+\setkeys{Gin}{width=0.72\textwidth}
+<<echo=FALSE, eval=TRUE, fig=TRUE, width=8, height=5>>=
+plot(ev_ttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Evolutionary learning of globally optimal trees}
+
+<<echo=FALSE, eval=TRUE, results=verbatim>>=
+ev_ttnc
+@
+
+\end{frame}
+
+\subsection{Other algorithms}
+\begin{frame}
+\frametitle{Other algorithms}
+
+\textbf{Furthermore:} Based on \emph{partykit}.
+\begin{itemize}
+  \item \emph{model4you:} Model-based trees/forests with personalised treatment effects.
+  \item \emph{disttree:} Distributional modeling with regression trees and random forests.
+  \item \emph{circtree:} Circular distributional trees and forests.
+  \item \emph{glmertree:} Generalized linear mixed-effects model trees.
+  \item \emph{psychotree:} Psychometric model trees (Rasch, Bradley-Terry, \dots).
+  \item \emph{stablelearner:} Stability assessment of trees (and other learners).
   \item \ldots
@@
 \begin{frame}
-\frametitle{R software for tree models}
-
-\textbf{Further R packages:}
-\begin{itemize}
-  \item mvpart (multivariate CART),
-  \item RWeka (J4.8, M5', LMT),
-  \item and many more (C50, quint, stima, \dots).
-\end{itemize}
+\frametitle{Other algorithms}
+
+\textbf{Other R packages:}
+\begin{itemize}
+  \item \emph{RWeka:} Interface to \emph{Weka} machine learning library, provided open-source
+    implementations of C4.5 (J.48) and M5 (M5'), and LMT.
+  \item \emph{C50:} C5.0 (successor to C4.5) decision trees and rule-based models.
+  \item Many more: \emph{mvpart}, \emph{quint}, \emph{stima}, \dots.
+\end{itemize}
+
 \end{frame}
@@
 \begin{frame}
-\frametitle{R software for tree models}
-
-\textbf{Furthermore:} Tree algorithms/software without R interface, e.g.,
-%
-\begin{itemize}
-  \item QUEST,
-  \item GUIDE,
-  \item LOTUS,
-  \item CRUISE,
+\frametitle{Other algorithms}
+
+\textbf{Beyond R:}
+\begin{itemize}
+  \item Unbiased recursive partitioning software (standalone) by Loh and co-workers: \emph{QUEST}, \emph{GUIDE}, \dots
+  \item In Python: \emph{scikit-learn} implements classical machine learning algorithms,
+    very limited availability of statistical inference-based algorithms.
   \item \dots
@@
 
-% \subsection*{References}
+\subsection{References}
 
@@
 \frametitle{References}
-
-%\vspace{-0.2cm}
 

### 2020-07-01 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
use Therneau & Atkinson 1997 reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3029 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f78491190020eb28121edb7da05e7e24dbdb78bd

@@
 
+Therneau T, Atkinson B (1997). 
+  \dquote{An Introduction to Recursive Partitioning Using the {rpart} Routine.}
+  Technical Report~61, Section of Biostatistics, Mayo Clinic, Rochester.
+  \url{http://www.mayo.edu/hsr/techrpt/61.pdf}
+
+\medskip
+
 % Breiman L (2001).
@@
 
-\medskip
-
-Therneau T, Atkinson B (2019). 
-  \dquote{{rpart}: Recursive Partitioning and Regression Trees.}
-  \url{https://CRAN.R-project.org/package=rpart}
-
 % medskip

### 2020-07-01 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
revisions on section 'aggregation & sampling'; added example (wages)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3030 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 60f7cbeb1753c18ab5d34125d06830ac9e22d2dd

@@
 
-library("disttree")
+library("partykit")
 library("Formula")
@@
 library("MASS")
-library(colorspace)
+library("colorspace")
 
@@
 \begin{itemize}
-\item Combination of an ensemble of trees. 
+\item Combine an ensemble of trees. 
 \item Each tree can capture non-linear and non-additive effects and select covariates and possible interactions automatically.
@@
 \&\\
-supsampling}
+random sampling}
+
+\begin{frame}
+\frametitle{Aggregation of trees}
+{\bf Simple tree model:} Obtain predictions for the response $y$ by averaging over observations in the same terminal node $\mathcal{B}_x$: 
+$$
+\hat{y}= \frac{1}{|\mathcal{B}_x|} \sum_{i: y_i \in \mathcal{B}_x} y_i
+\visible<2->{
+\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$$
+}
+
+{\bf Model-based tree:} Obtain predictions for model parameter(s) $\theta$ by optimizing an objective function $f$ for all observations in the same terminal node $\mathcal{B}_x$:
+$$
+\hat{\theta}= \argmin_{\theta} \sum_{i: y_i \in \mathcal{B}_x} f(\theta; y_i)
+\visible<2->{
+\ =\ \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)$$
+}
+
+{\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (Y - \hat{Y})^2$.
+
+\end{frame}
 
@@
 
-The predicted response $\hat{y}$ 
-%or, more generally, the predicted model parameter (vector) $\hat{\theta}$ 
-is the argument %of the parameter space $\Theta$ 
+The predicted model parameter (vector) $\hat{\theta}$ 
+is the argument of the parameter space $\Theta$ 
 that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{y_i\}_{i=1,\ldots,n}$:
@@
 \[
-\hat{y}(x) =  \argmin_{y} \sum_{i=1}^n w_i(x) \cdot f(y; y_i)
+\hat{\theta}(x) =  \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)
 \]
@@
 
-\only<5->{
-%\vspace{-3cm}
-{\bf Note:} In model-based trees and forests a prediction is made not only for the response but for a model parameter (vector) $\theta$. 
-This more general approach holds for classic random forests averaging over mean values of tree nodes by setting $f(\theta, y_i) = (\theta - y_i)^2$\\
-but also allows for combining quantiles (quantile regression forests) or full distributions as in distributional forests with
-$f(\theta, y_i) = log(likelihood(\theta, y_i))$.\\
-}
-
 \end{frame}
@@
 \begin{frame}
-\frametitle{Subsampling}
+\frametitle{Random sampling}
 {\bf Idea:} Build each tree on a different subset of the full data set.
@@
 
-{\bf Types of subsampling:}
+{\bf Sampling observations:}
 \begin{itemize}
-\item Subsets (recommended size: 63.2\% of the full data set)
-\item Bootstrapping (resampling with replacement)
+\item Bootstrapping: resampling with replacement, usually ``n out of n''.
+\item Subsampling: draw a subset without replacement (typically of the size of 63.2\% of the full data set, but can be a smaller fraction for large data sets).
 \end{itemize}
@@
 
-{\bf Random input variable sampling:} In order to reduce the correlation among trees only a subset of possible split variables is provided for each split (recommended size: $\sqrt{\text{\#Var}}$).
+{\bf Sampling covariates:}\\
+\smallskip
+In order to reduce the correlation among trees random input variable sampling is employed, i.e., for each split only a subset of covariates is selected randomly and provided as possible split variables (typical sampling size: $\sqrt{\text{\#Var}}$).
+\end{frame}
+
+\begin{frame}
+\frametitle{Hyperparameter selection}
+
+Apart from the size and type of sampling two further control parameters have to be set for a forest model: 
+\begin{itemize}
+\item {\bf Number of trees:} A high number of trees allows for a more stable model and for a better approximation of smooth effects, however, at the cost of higher computational effort.
+\item {\bf Size of the trees:} The size of each single tree can be controlled by setting stopping criteria such as a minimal number of observations in a node, maximum tree depth or a significance level if a statistical test is employed in the tree algorithm.
+However, in a forest model each tree is usually built as large as possible which can lead to overfitting for a single tree, but this is compensated by aggregating the trees to a forest model.
+\end{itemize}
+\end{frame}
+
+<<data_wages, eval=TRUE, echo=FALSE, results=hide>>=
+data("CPS1985", package = "AER")
+@
+
+\begin{frame}[fragile]
+\frametitle{Example: wages}
+
+\textbf{Data:} Random sample from the May 1985 US Current Population Survey.
+A data frame containing \Sexpr{nrow(CPS1985)} observations on \Sexpr{ncol(CPS1985)} variables.
 
@@
 
-{\bf Note:} Each tree is built as large as possible which can lead to overfitting for the single tree. However, this is compensated by aggregating the trees to a forest model.
+\begin{tabular}{ll}
+\hline 
+Variable & Description\\
+\hline 
+\code{wage}       & Wage (in US dollars per hour). \\
+\code{education}  & Education (in years).  \\
+\code{experience} & Potential work experience (in years, \code{age - education - 6}). \\
+\code{age}        & Age (in years). \\
+\code{ethnicity}  & Factor: Caucasian, Hispanic, Other. \\
+\code{gender}     & Factor: Male, Female. \\
+\code{union}      & Factor. Does the individual work on a union job? \\
+\hline 
+\end{tabular}
+
+\bigskip
+
+\textbf{Model formula:} \code{log(wage) ~ education + experience + age + ethnicity + gender + union}.
 
@@
 
-\subsection{Software}
+<<example_wages, eval=TRUE, echo=FALSE, results=hide>>=
+set.seed(4)
+f <- log(wage) ~ education + experience + age + ethnicity + gender + union
+ct_cps <- ctree(f, data = CPS1985, alpha = 0.01)
+cf <- cforest(f, data = CPS1985, ntree = 5) #, control =ctree_control(maxdepth = 5))
+vimp <- varimp(cf , risk = "loglik")
+@
+
+\begin{frame}[fragile]
+\frametitle{Example: wages}
+Single tree model:
+\setkeys{Gin}{width=0.99\textwidth}
+<<example_wages_ctree, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
+plot(ct_cps)
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Example: wages}
+Forest model - tree I:
+\setkeys{Gin}{width=0.99\textwidth}
+<<example_wages_cforest1, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
+#par(mfrow=c(2,2))
+plot(party(cf[[1]][[1]], data = CPS1985), drop_terminal = TRUE)
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Example: wages}
+Forest model - tree II:
+\setkeys{Gin}{width=0.99\textwidth}
+<<example_wages_cforest2, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
+#par(mfrow=c(2,2))
+plot(party(cf[[1]][[2]], data = CPS1985), drop_terminal = TRUE)
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Example: wages}
+Forest model - tree III:
+\setkeys{Gin}{width=0.99\textwidth}
+<<example_wages_cforest3, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
+#par(mfrow=c(2,2))
+plot(party(cf[[1]][[3]], data = CPS1985), drop_terminal = TRUE)
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Example: wages}
+Forest model - tree IV:
+\setkeys{Gin}{width=0.99\textwidth}
+<<example_wages_cforest4, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
+#par(mfrow=c(2,2))
+plot(party(cf[[1]][[4]], data = CPS1985), drop_terminal = TRUE)
+@
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Example: wages}
+\vspace{-0.5cm}
+\setkeys{Gin}{width=0.94\textwidth}
+<<example_wages_forest_prediction, eval=TRUE, echo=FALSE, fig=TRUE, width = 10>>=
+nd <- data.frame(education = 13.02, #c(2:18),
+                 experience = 17.82,
+                 age = c((18:64)/1), #36.83,
+                 ethnicity = "cauc",
+                 region = "other",
+                 gender = "male",
+                 occupation = "worker",
+                 sector = "other",
+                 union = "no",
+                 married = "yes")
+
+{
+  par(mfrow = c(2,2))
+  set.seed(74)
+  cf <- cforest(f, data = CPS1985, ntree = 2)
+  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
+  colnames(predwage) <- c("age", "predicted log(wage)")
+  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
+       ylab = "predicted log(wage)", xlab = "age", main = "2 trees")
+
+  cf <- cforest(f, data = CPS1985, ntree = 4)
+  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
+  colnames(predwage) <- c("age", "predicted log(wage)")
+  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
+       ylab = "predicted log(wage)", xlab = "age", main = "4 trees")
+
+  cf <- cforest(f, data = CPS1985, ntree = 20)
+  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
+  colnames(predwage) <- c("age", "predicted log(wage)")
+  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
+       ylab = "predicted log(wage)", xlab = "age", main = "20 trees")
+
+  cf <- cforest(f, data = CPS1985, ntree = 200)
+  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
+  colnames(predwage) <- c("age", "predicted log(wage)")
+  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
+       ylab = "predicted log(wage)", xlab = "age", main = "200 trees")
+}
+@
+\end{frame}
+
+\begin{frame}
+\frametitle{Variable importance}
+{\bf Idea:} Asses the influence of a covariate by evaluating the change in performance of the model (mean decrease in accuracy) after dropping or permutating the selected covariate.
+
+\bigskip
+
+%\bigskip
+
+% {\bf Implementations:}
+% \begin{itemize}
+% \item \fct{importance} in the R package randomForest.
+% \item \fct{importance.ranger} in the R package ranger.
+% \item \fct{varimp} in the R package partykit.
+% \end{itemize}
+
+% \only<2->{
+% \setkeys{Gin}{width=0.5\textwidth}
+% <<example_wages_vimp, eval=TRUE, echo=FALSE, fig=TRUE, width=6, height=3>>=
+% par(mar = c(4,5.5,0,2))
+% barplot(sort(vimp, decreasing = FALSE),
+%         horiz = TRUE, las = 1, axes = FALSE,
+%         xlab = "Variable importance: mean decrease in log-likelihood")
+% axis(1, at = seq(0,1,0.02), las = 1, mgp=c(0,1,0))
+% @
+
+Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, nr.
+of misclassifications, \ldots).
+%}
+\end{frame}
+
+\begin{frame}
+\frametitle{Example: wages}
+Evaluation of variable importance in the fitted forest model:
+
+\bigskip
+
+\setkeys{Gin}{width=0.57\textwidth}
+<<example_wages_vimp, eval=TRUE, echo=FALSE, fig=TRUE, width=6, height=3>>=
+par(mar = c(4,5.4,1,2))
+barplot(sort(vimp, decreasing = FALSE),
+        horiz = TRUE, las = 1, axes = FALSE,
+        xlab = "Variable importance: mean decrease in log-likelihood")
+axis(1, at = seq(0,1,0.02), las = 1, mgp=c(0,1,0))
+@
+\end{frame}
+
+\subsection{Implementations}
 
@@
 <<call_ranger, eval=FALSE, echo=TRUE>>=
-ranger(formula, data)
+rf <- ranger(log(wage) ~ education + experience + age + ethnicity + gender + union,
+             data = CPS1985)
 @
@@
 <<call_cforest, eval = FALSE>>=
-cforest(formula, data, subset, ...)
+cf <- cforest(log(wage) ~ education + experience + age + ethnicity + gender + union,
+              data = CPS1985)
 @
@@
 
-<<call_predict, eval=FALSE, echo=TRUE>>=
-predict(object, newdata, type = "response", ...)
+For a ranger object:
+<<call_predict_rf, eval=FALSE, echo=TRUE>>=
+predict(rf, data = testdata)
+@
+
+\bigskip
+
+For a cforest object:
+<<call_predict_cf, eval=FALSE, echo=TRUE>>=
+predict(cf, newdata = testdata)
 @
@@
 
-\subsection{Variable importance}
-
-\begin{frame}
-\frametitle{Variable importance}
-{\bf Idea:} Asses the influence of a covariate by evaluating the change in performance of the model (mean decrease in accuracy) after dropping or permutating the selected covariate.
-
-\bigskip
-
-Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, nr.
-of misclassifications, \ldots).
-
-\bigskip
-
-{\bf Implementations:}
-\begin{itemize}
-\item \fct{importance} in the R package randomForest.
-\item \fct{importance.ranger} in the R package ranger.
-\item \fct{varimp} in the R package partykit.
-\end{itemize}
-
-\end{frame}
-
-%\subsection{References}
+\subsection{References}
 

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
small corrections in formula notations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3031 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ac8ccb92d32857780505b5b302a34c35ef883e34

@@
 \visible<13->{
-for $\hat{y}_t$ being the average response value in the corresponding segment $\mathcal{B}^t_x$ of the $t$-th tree for learning data $\{y_i\}_{i=1,\ldots,n}$:
+for $\hat{y}_t$ being the average response value in the segment $\mathcal{B}^t_x$ which the observed covariate $x$ is assigned to in the $t$-th tree with learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:
 
@@
 \frametitle{Aggregation of trees}
-{\bf Simple tree model:} Obtain predictions for the response $y$ by averaging over observations in the same terminal node $\mathcal{B}_x$: 
+{\bf Simple tree model:} Obtain the predicted response $\hat{y}$ for an observed covariate~$x$ by averaging over observations in the same terminal node $\mathcal{B}_x$: 
 $$
@@
 
-{\bf Model-based tree:} Obtain predictions for model parameter(s) $\theta$ by optimizing an objective function $f$ for all observations in the same terminal node $\mathcal{B}_x$:
+{\bf Model-based tree:} Obtain the predicted model parameter(s) $\hat{\theta}$ for an observed covariate~$x$ by optimizing an objective function $f$ for all observations in the same terminal node $\mathcal{B}_x$:
 $$
@@
 
-{\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (Y - \hat{Y})^2$.
+{\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (y_i - \theta)^2$.
 
@@
 is the argument of the parameter space $\Theta$ 
-that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{y_i\}_{i=1,\ldots,n}$:
+that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:
 

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
further revisions on section 'implementations'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3032 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 40e22e04e03b698bfab52584bd8165bbb1641efc

@@
   \item Based on implementations of CART.
+  \item First implementation of random forests in R.
 \end{itemize}
@@
 
+\medskip
+
+{\bf ranger}:
+\begin{itemize}
+\item RANdom  forest  GEneRator.
+\item R package providing a fast implementation of classic random forests.
+\end{itemize}
+
+\medskip
+
+{\bf partykit:}
+\begin{itemize}
+  \item R package offering a toolkit for unbiased recursive partitioning. 
+  \item Provides the forest-building function \fct{cforest}, based on conditional inference trees (Ctree).
+\end{itemize}
+
 \end{frame}
@@
 {\bf ranger}:
-\begin{itemize}
-\item RANdom  forest  GEneRator.
-\item R package providing a faster implementation of classic random forests.
-\end{itemize}
 
-<<call_ranger, eval=FALSE, echo=TRUE>>=
+<<call_ranger, eval=TRUE, echo=TRUE>>=
+library("ranger")
 rf <- ranger(log(wage) ~ education + experience + age + ethnicity + gender + union,
@@
 
+{\bf ranger}:
+
+<<call_ranger_predict, eval=TRUE, echo=TRUE, results=verbatim>>=
+newdata <- data.frame(education = 13, experience = 17,
+                      age = 36, ethnicity = "cauc",
+                      region = "other", gender = "male",
+                      occupation = "worker", sector = "other",
+                      union = "no", married = "yes")
+pred_rf <- predict(rf, data = newdata)
+pred_rf$predictions     
+exp(pred_rf$predictions)     
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{R software for forest models}
+
 {\bf partykit:}
-\begin{itemize}
-  \item R package offering a toolkit for unbiased recursive partitioning. 
-  \item Provides the forest-building function \fct{cforest}, based on conditional inference trees.
-\end{itemize}
 
-\medskip
-
-<<call_cforest, eval = FALSE>>=
+<<call_cforest, eval = FALSE, echo=TRUE>>=
+library("partykit")
 cf <- cforest(log(wage) ~ education + experience + age + ethnicity + gender + union,
@@
 @
-
-\medskip
 
@@
 \item \code{ytrafo}: transformation function applied to the response,
-\item \code{ntree}: nr. of trees, 
-\item \code{mtry}: nr. of randomly sampled split variable candidates, 
+\item \code{ntree}: number of trees, 
+\item \code{mtry}: number of randomly sampled split variable candidates, 
 \item \code{perturb}: type and size of subsetting, 
@@
 \end{itemize}
-\vspace{0.9cm}
+%\vspace{0.9cm}
 }
-\only<2>{
+\only<2->{
 Optional tree-specific arguments handed over via \fct{ctree\_control}:
 \begin{itemize}
-\item \code{minsplit}: minimum nr. of observations to perform a split, 
-\item \code{minbucket}: minimum nr. of observations in a node after splitting, 
+\item \code{minsplit}: minimum number of observations to perform a split, 
+\item \code{minbucket}: minimum number of observations in a node after splitting, 
 \item \code{maxdepth}: maximum depth of the tree, 
@@
 \end{itemize}
-\vspace{0.9cm}
+%\vspace{0.9cm}
 }
+
+% \only<3>{
+% <<call_cforest_control, eval = FALSE, echo=TRUE>>=
+% cf <- cforest(log(wage) ~ education + experience + age + ethnicity + gender + union,
+%              data = CPS1985, control = ctree_control(minsplit = 20))
+% @
+% }
+
 \end{frame}
@@
 
-{\bf Predictions:} For a (possibly new) set of observations the method \fct{predict} can be employed to obtain the corresponding predicted values of the response variable.
+{\bf partykit:}
 
-\bigskip
-
-For a ranger object:
-<<call_predict_rf, eval=FALSE, echo=TRUE>>=
-predict(rf, data = testdata)
-@
-
-\bigskip
-
-For a cforest object:
-<<call_predict_cf, eval=FALSE, echo=TRUE>>=
-predict(cf, newdata = testdata)
+<<call_cforest_predict, eval = TRUE, results=verbatim>>=
+newdata <- data.frame(education = 13, experience = 17,
+                      age = 36, ethnicity = "cauc",
+                      region = "other", gender = "male",
+                      occupation = "worker", sector = "other",
+                      union = "no", married = "yes")
+pred_cf <- predict(cf, newdata = newdata)
+pred_cf
+exp(pred_cf)
 @
@@
 \begin{itemize}
-  \item grf (generalized random forests),
-  \item disttree (distributional forests, based on partykit),
+  \item grf: generalized random forests,
+  \item disttree: distributional forests, based on partykit,
   \item \ldots

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
further revisions on section 'implementations'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3033 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d53b48c8616ccb0bbd4e12b91a21b8d609064ccc

@@
 \frametitle{Variable importance}
-{\bf Idea:} Asses the influence of a covariate by evaluating the change in performance of the model (mean decrease in accuracy) after dropping or permutating the selected covariate.
+{\bf Idea:} Asses the influence of a covariate on the model by evaluating the change in performance of the model (mean decrease in accuracy) after permutating the selected covariate.
 
@@
 
-Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, nr.
-of misclassifications, \ldots).
+Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, number of misclassifications, \ldots).
 %}
@@
 exp(pred_rf$predictions) 
+rf <- ranger(log(wage) ~ education + experience + age + ethnicity + gender + union,
+             data = CPS1985, importance = "impurity")
+importance(rf)
 @
@@
 \end{itemize}
-%\vspace{0.9cm}
 }
-
 % \only<3>{
+% 
 % <<call_cforest_control, eval = FALSE, echo=TRUE>>=
@@
 
+\begin{frame}[fragile]
+\frametitle{R software for forest models}
+
+{\bf partykit:}
+
+<<call_cforest, eval = FALSE, echo=TRUE>>=
+library("partykit")
+cf <- cforest(log(wage) ~ education + experience + age + ethnicity + gender + union,
+              data = CPS1985)
+@
+
+Optional tree-specific arguments handed over via \fct{ctree\_control}:
+\begin{itemize}
+\item \code{minsplit}: minimum number of observations to perform a split, 
+\item \code{minbucket}: minimum number of observations in a node after splitting, 
+\item \code{maxdepth}: maximum depth of the tree, 
+\item \code{alpha}: significance level for split variable selection,
+\item \ldots 
+\end{itemize}
+
+<<call_cforest_control, eval = FALSE, echo=TRUE>>=
+cf <- cforest(log(wage) ~ education + experience + age + ethnicity + gender + union,
+            data = CPS1985, control = ctree_control(minsplit = 20))
+@
+
+\end{frame}
 \begin{frame}[fragile]
@@
 exp(pred_cf)
+varimp(cf)
 @
@@
 
+Wright M N, Ziegler A (2017).
+\dquote{{ranger}: A Fast Implementation of Random Forests for High
+Dimensional Data in {C++} and {R}.}
+\emph{Journal of Statistical Software}, \textbf{77}{1}, 1--17.
+\doi{10.18637/jss.v077.i01}
+
+\medskip
+
 % Zeileis A, Hothorn T, Hornik K (2008).
@@
 
-\medskip
-
-Wright M N, Ziegler A (2017).
-\dquote{{ranger}: A Fast Implementation of Random Forests for High
-Dimensional Data in {C++} and {R}.}
-\emph{Journal of Statistical Software}, \textbf{77}{1}, 1--17.
-\doi{10.18637/jss.v077.i01}
-
 \end{frame}

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
minor adaptions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3034 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 666b75d7ae5b2ce734292dad650676a22614af71

@@
   colnames(predwage) <- c("age", "predicted log(wage)")
-  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
-       ylab = "predicted log(wage)", xlab = "age", main = "2 trees")
+  plot(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
+       ylab = "predicted wage", xlab = "age", main = "Forest with 2 trees")
 
@@
   colnames(predwage) <- c("age", "predicted log(wage)")
-  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
-       ylab = "predicted log(wage)", xlab = "age", main = "4 trees")
+  plot(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
+       ylab = "predicted wage", xlab = "age", main = "Forest with 4 trees")
 
@@
   colnames(predwage) <- c("age", "predicted log(wage)")
-  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
-       ylab = "predicted log(wage)", xlab = "age", main = "20 trees")
+  plot(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
+       ylab = "predicted wage", xlab = "age", main = "Forest with 20 trees")
 
@@
   colnames(predwage) <- c("age", "predicted log(wage)")
-  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
-       ylab = "predicted log(wage)", xlab = "age", main = "200 trees")
+  plot(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
+       ylab = "predicted wage", xlab = "age", main = "Forest with 200 trees")
 }
@@
 \frametitle{Variable importance}
-{\bf Idea:} Asses the influence of a covariate on the model by evaluating the change in performance of the model (mean decrease in accuracy) after permutating the selected covariate.
+{\bf Problem:} In a single tree model the effect of each covariate can easily be investigated based on a simple illustration of the tree structure.
+However, in a forest model this would require to investigate each tree on its own including its specific setting (induced by random sampling).
+
+\bigskip
+
+{\bf Idea:} Asses the influence of a covariate on the forest model by evaluating the change in performance of the model (mean decrease in accuracy) after permutating the selected covariate.
 

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/exercises/exercise_titanic.Rmd
small corrections in section 'aggregation'; added section on distributional forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3035 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c8e5510b3c8d99b30497ac311a8f39e92a900e30

---
title: "Exercise: Titanic"
output: html_document
---

```{r, include = FALSE}
file <- "data/titanic.rds"
stopifnot(file.exists(file))
data <- readRDS(file)
head(data)
```

In this exercise we will use the 'titanic' data set. As the data set presented on the slides, this version describes the survival status of individual passengers on the Titanic, however, it provides additional information on the passengers given by 10 covriates but does not include information on the crew.

Simply download the file `r xfun::embed_file(file, text = "titanic.rds")` by clicking or download it from OpenOLAT or from the corresponding homepage
<http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/titanic.html>
which also provides more detailed information on the data set.

We can import/read this file using `data <- readRDS(...)`. The file
contains the following information:

* `name`: passenger's name (character).
* `gender.`: 'male' or 'female' (factor) .
* `age`: age in years (numeric).
* `class`:  passenger class or the type of service aboard for crew members (factor).
* `embarked`: place of embarkment (factor; C = Cherbourg, Q = Queenstown, S = Southampthon).
* `country`: home country (factor).
* `ticketno`:  ticket number (integer; NA for crew members).
* `fare`: ticket price (numeric; NA for crew members, musicians and employees of the shipyard company).
* `sibsp`: number if siblings/spouses aboard (ordered factor).
* `parch`: number of parents/children aboard (ordered factor).
* `survived`: Did the passenger survive? (factor).

# The Tasks

We would like to find out how the survival status of a passenger on the Titanic depends on the provided additional attributes.
By employing a tree model we are looking for a separation into homogeneous subgroups based on the additional information.

Our response in this case is the binary variable `survived`, as covariates we use the additional variables `gender`, `age`, `class`, `embarked`, `fare`, `sibsp` and `parch`.

Apply the CTree algorithm to build the tree models described in the following steps:

* Load the data set `"titanic.rds"`.
* Build a tree using pre-pruning with a significance level of 0.01, with a maximum depth of 5 levels and the segment size of terminal nodes not being smaller than 20.
* Evaluate the performance on the learning data by calculating the corresponding confusion matrix. How large is the misclassification rate?
* Predict the survival status of a 30-years-old female passenger, travelling with her husband and her two parents who all embarked in Southampton and paid 25 Pounds each for a 2nd class ticket. Does the prediction change if she had a ticket for the 3rd class?
* Separate the data set into a learning set (2/3 of the full data) and a testing set (1/3 of the full data). Note that for observations in the test data which include NAs predictions can not be made. Build a tree on the learning data set and predict the survival status on the testing data set. Evaluate the performance based on the number of misclassifications. How do parameters such as a minimal segment size or the significance level applied for pre-pruning influence the performance?

```{r, include = FALSE}
formula <- survived ~  gender + age + class + embarked + fare + sibsp + parch

library("partykit")
ct <- ctree(formula, data = data)
ct <- ctree(formula, data = data, control = ctree_control(alpha = 0.01, minbucket = 20, maxdepth = 5))

library("caret")
caret::confusionMatrix(data$survived, predict(ct, newdata = data))

newpassenger <- data.frame(gender = "female",
                           age = 30,
                           class = "2nd",
                           embarked = "S",
                           fare = 25,
                           sibsp = "1",
                           parch = "2")
predict(ct, newdata = newpassenger)

newpassenger2 <- newpassenger
newpassenger2$class <- "3rd"
predict(ct, newdata = newpassenger2)
```

```{r, include = FALSE, echo = FALSE, out.width = "100%", fig.width = 10, fig.height = 5}
plot(ct)
```
```{r, include = FALSE}
set.seed(4)
trainid <- sample(1:NROW(data), size = 1471, replace = FALSE)
train <- data[trainid,]
test <- data[-trainid,]
test <- na.omit(test)

ctrain <- ctree(formula, data = train)
predtest <- predict(ctrain, newdata = test)

library("caret")
caret::confusionMatrix(test$survived, predtest)

ctrain <- ctree(formula, data = train, control = ctree_control(alpha = 0.01))
plot(ctrain)
predtest <- predict(ctrain, newdata = test)
caret::confusionMatrix(test$survived, predtest)
```

```{r, include = FALSE}
set.seed(4)
library("ranger")
rf <- ranger(formula, data = na.omit(train), num.trees = 20)
rf$confusion.matrix
rf <- ranger(formula, data = na.omit(train), num.trees = 500)
rf$confusion.matrix

cf <- cforest(formula, data = na.omit(train), ntree = 20)

pred_cf <- predict(cf, newdata = test)
pred_rf <- predict(rf, newdata = test)
```

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/exercises/exercise_german_credit.Rmd
small corrections in section 'aggregation'; added section on distributional forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3035 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c8e5510b3c8d99b30497ac311a8f39e92a900e30

@@
 ```
+
+```{r, include = FALSE}
+library("ranger")
+rf <- ranger(formula, data = train, num.trees = 20)
+rf$confusion.matrix
+rf <- ranger(formula, data = train, num.trees = 500)
+rf$confusion.matrix
+
+cf <- cforest(formula, data = na.omit(train), ntree = 20)
+
+pred_cf <- predict(cf, newdata = test)
+pred_rf <- predict(rf, data = test)
+```

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
small corrections in section 'aggregation'; added section on distributional forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3035 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c8e5510b3c8d99b30497ac311a8f39e92a900e30

@@
 \usepackage{xcolor}
+\usepackage{changepage}
+%\usepackage[cal=boondoxo]{mathalfa}
+%\graphicspath{{plots/}}
+\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
 \newcommand{\argmin}{\operatorname{argmin}\displaylimits}
@@
 library("colorspace")
-
+library("disttree")
+library("latex2exp")
+library("gamlss")
+library("crch")
+library("RainTyrol")
+library("ggplot2")
+theme_set(theme_bw(base_size = 18))
+library(colorspace)
 ## HCL palette
@@
 
-\visible<13->{
+\visible<14->{
 for $\hat{y}_t$ being the average response value in the segment $\mathcal{B}^t_x$ which the observed covariate $x$ is assigned to in the $t$-th tree with learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:
@@
 $\hat{y}_t= \frac{1}{|\mathcal{B}^t_x|} \sum_{i: y_i \in \mathcal{B}^t_x} y_i$
-\visible<14->{
+\visible<15->{
 $\ =\ \sum_{i=1}^n \frac{I(x_i \in \mathcal{B}^t_x)}{|\mathcal{B}^t_x|} \cdot y_i$
 }
-\visible<15->{
+\visible<16->{
 $\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$
@@
 }
-
-{\bf Model-based tree:} Obtain the predicted model parameter(s) $\hat{\theta}$ for an observed covariate~$x$ by optimizing an objective function $f$ for all observations in the same terminal node $\mathcal{B}_x$:
+\visible<3->{
+$$\ =\ \argmin_{\mu} \sum_{i=1}^n w_i(x) \cdot (y_i - \mu)^2$$
+}
+
+\visible<4->{
+{\bf Model-based tree:} Obtain the predicted model parameter(s) $\hat{\theta}$ for an observed covariate~$x$ by optimizing a loss function for all observations in the same terminal node $\mathcal{B}_x$:
 $$
-\hat{\theta}= \argmin_{\theta} \sum_{i: y_i \in \mathcal{B}_x} f(\theta; y_i)
-\visible<2->{
-\ =\ \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)$$
-}
-
-{\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (y_i - \theta)^2$.
+\hat{\theta}= \argmin_{\theta} \sum_{i: y_i \in \mathcal{B}_x} loss(\theta; y_i)
+\ =\ \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot loss(\theta; y_i)$$
+}
+
+% {\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (y_i - \theta)^2$.
 
@@
 is the argument of the parameter space $\Theta$ 
-that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:
+that minimizes the weighted sum of a loss function evaluated for learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:
 
@@
 \[
-\hat{\theta}(x) =  \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)
+\hat{\theta}(x) =  \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot loss(\theta; y_i)
 \]
@@
 \item Bootstrapping: resampling with replacement, usually ``n out of n''.
-\item Subsampling: draw a subset without replacement (typically of the size of 63.2\% of the full data set, but can be a smaller fraction for large data sets).
+\item Subsampling: draw a subset without replacement (typically of the size of 63.2\% of the full data set; for large data sets a smaller fraction can be advantageous).
 \end{itemize}
@@
 #par(mfrow=c(2,2))
-plot(party(cf[[1]][[1]], data = CPS1985), drop_terminal = TRUE)
+plot(party(cf[[1]][[1]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
 @
@@
 #par(mfrow=c(2,2))
-plot(party(cf[[1]][[2]], data = CPS1985), drop_terminal = TRUE)
+plot(party(cf[[1]][[2]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
 @
@@
 #par(mfrow=c(2,2))
-plot(party(cf[[1]][[3]], data = CPS1985), drop_terminal = TRUE)
+plot(party(cf[[1]][[3]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
 @
@@
 #par(mfrow=c(2,2))
-plot(party(cf[[1]][[4]], data = CPS1985), drop_terminal = TRUE)
+plot(party(cf[[1]][[4]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
 @
@@
        ylab = "predicted wage", xlab = "age", main = "Forest with 4 trees")
+  lines(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
+       col = "gray")
 
@@
 
-{\bf Idea:} Asses the influence of a covariate on the forest model by evaluating the change in performance of the model (mean decrease in accuracy) after permutating the selected covariate.
+{\bf Idea:} Asses the influence of a covariate on the forest model by 
+\begin{itemize} 
+\item permuting the selected covariate and evaluating the resulting change in performance of the model (mean decrease in accuracy, measured for example based on the employed loss function) or
+\item calculating the total decrease in node impurities from splitting in the selected covariate (measured for example by the Gini index for classification trees or the RMSE for regression trees), 
+\end{itemize}
+averaged over all trees.
 
@@
 
-Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, number of misclassifications, \ldots).
+% Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, number of misclassifications, \ldots).\\
+% To measure impurity, for example, the Gini index can be employed for classification trees or the RMSE for regression trees.
 %}
@@
   \item R package offering a toolkit for unbiased recursive partitioning. 
-  \item Provides the forest-building function \fct{cforest}, based on conditional inference trees (Ctree).
+  \item Provides the forest-building function \fct{cforest}, based on conditional inference trees (CTree).
 \end{itemize}
@@
   \item disttree: distributional forests, based on partykit,
+  \item circtree: circular distributional forests, based on partykit,
+  \item model4you: model-based random forests for personalized treatment effects,
+  \item randomSurvivalForest: random forests for the analysis of right-censored survival data, 
   \item \ldots
@@
 \end{frame}
+
+\subsection{Distributional forests}
+\begin{frame}%[fragile]
+\frametitle{Motivation}
+\vspace{-0.41cm}
+\begin{figure}[!htb]
+\minipage{0.285\textwidth}
+\begin{center}
+<<motivation_GLM, echo=FALSE, results=hide>>=
+nobs <- 200
+## GLM
+set.seed(7)
+x <- c(1:nobs)/nobs
+ytrue <- 1+x*1.5
+y <- ytrue + rnorm(nobs,0,0.3)
+@
+\visible<2->{
+<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
+@
+}
+\end{center}
+\endminipage
+\visible<3->{{\LARGE$\rightarrow$}}
+\minipage{0.285\textwidth}
+\begin{center}
+<<motivation_GAM, echo=FALSE, results=hide>>=
+## GAM
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3
+y <- ytrue + rnorm(nobs,0,0.3)
+@
+\visible<3->{
+<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
+@
+}
+\end{center}
+\endminipage
+\visible<4->{{\LARGE$\rightarrow$}}
+\minipage{0.285\textwidth}
+\begin{center}
+<<motivation_GAMLSS, echo=FALSE, results=hide>>=
+## GAMLSS
+set.seed(7)
+x <- c(1:nobs)/nobs
+x <- 2*(x-0.5)
+ytrue <- x^3
+var <- exp(-(2*x)^2)/2
+y <- ytrue + rnorm(nobs, 0, 0.1 + var)
+@
+\visible<4->{
+<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
+  col = pallight["forest"], border = "transparent")
+lines(x, ytrue, col = pal["forest"], lwd=7)
+points(x, y, col = "slategray", pch = 19)
+box(lwd = 5)
+@
+}
+\end{center}
+\endminipage
+
+\vspace{0.5cm}
+\minipage{0.25\textwidth}
+\begin{center}
+\visible<2->{
+LM, GLM\\
+\vspace{0.5cm}
+\code{lm}\\
+\code{glm}\\
+\vspace{1.5cm}}
+\end{center}
+\endminipage
+\hspace{1.1cm}
+\minipage{0.25\textwidth}
+\begin{center}
+\visible<3->{
+GAM\\
+\vspace{0.5cm}
+\code{mgcv}\\
+\code{VGAM}\\
+\vspace{1.5cm}}
+\end{center}
+\endminipage
+\hspace{1.1cm}
+\minipage{0.25\textwidth}
+\begin{center}
+\visible<4->{
+GAMLSS\\
+\vspace{0.5cm}
+\code{gamlss}\\
+\code{mgcv}\\
+\code{VGAM}\\
+\code{gamboostLSS}\\
+\code{bamlss}}
+\end{center}
+\endminipage
+\end{figure}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Motivation}
+\vspace{-0.41cm}
+\begin{figure}[!htb]
+\minipage{0.285\textwidth}
+\begin{center}
+<<motivation_regtree, echo=FALSE, results=hide>>=
+## Reg. Tree
+set.seed(7)
+kappa <- 12
+x <- c(1:nobs)/nobs
+ytrue <- ytree <- yforest <- numeric(length = length(x))
+for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
+y <- ytrue + rnorm(nobs,0,0.3)
+for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
+@
+\visible<1->{
+<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
+lines(x = x, y = ytree, col = pal["forest"], lwd=7)
+@
+}
+\end{center}
+
+\endminipage
+\visible<2->{{\LARGE$\rightarrow$}}
+\minipage{0.285\textwidth}
+ \begin{center}
+<<motivation_randforest, echo=FALSE, results=hide>>=
+## Random Forest
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
+@
+\visible<2->{
+<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+box(lwd=5)
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
+lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
+@
+}
+\end{center}
+
+\endminipage
+\visible<3->{{\LARGE$\rightarrow$}}
+\minipage{0.285\textwidth}
+\begin{center}
+\visible<3->{
+<<plot_motivation_question, fig=TRUE, echo=FALSE>>=
+par(mar=c(2,0,2,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
+box(lwd=5)
+text(x = mean(range(x)), y = mean(range(y)), "?", cex = 12)
+@
+}
+\end{center}
+
+\endminipage
+
+\minipage{0.285\textwidth}
+\begin{center}
+\vspace{0.0cm}
+\visible<1->{
+Regression tree\\
+\vspace{0.3cm}
+\resizebox{0.2\textwidth}{!}{
+\begin{tikzpicture}
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (1, 2) {};
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0.5, 1) {};
+  \draw[-, line width=1pt] (n0) -- (n1);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (1.5, 1) {};
+  \draw[-, line width=1pt] (n0) -- (n2);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (1, 0) {};
+  \draw[-, line width=1pt] (n2) -- (n3);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (2, 0) {};
+  \draw[-, line width=1pt] (n2) -- (n4);
+\end{tikzpicture}}
+\vspace{0.2cm}\\
+\code{rpart}\\
+\code{party(kit)}\\
+\vspace{1cm}}
+\end{center}
+\endminipage
+\hspace{0.65cm}
+\minipage{0.285\textwidth}
+\begin{center}
+\vspace{-0.4cm}
+\visible<2->{
+Random forest\\
+\vspace{0.4cm}
+\resizebox{0.6\textwidth}{!}{
+\begin{tikzpicture}
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n00) at (1, 2) {};
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n01) at (0.5, 1) {};
+  \draw[-, line width=1pt] (n00) -- (n01);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n02) at (1.5, 1) {};
+  \draw[-, line width=1pt] (n00) -- (n02);
+
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n10) at (3, 2) {};
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n11) at (2.5, 1) {};
+  \draw[-, line width=1pt] (n10) -- (n11);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n12) at (3.5, 1) {};
+  \draw[-, line width=1pt] (n10) -- (n12);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n13) at (2, 0) {};
+  \draw[-, line width=1pt] (n11) -- (n13);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n14) at (2.8, 0) {};
+  \draw[-, line width=1pt] (n11) -- (n14);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n15) at (3.2, 0) {};
+  \draw[-, line width=1pt] (n12) -- (n15);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n16) at (4, 0) {};
+  \draw[-, line width=1pt] (n12) -- (n16);
+
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
+  \draw[-, line width=1pt] (n20) -- (n21);
+  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
+  \draw[-, line width=1pt] (n20) -- (n22);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
+  \draw[-, line width=1pt] (n22) -- (n23);
+  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
+  \draw[-, line width=1pt] (n22) -- (n24);
+\end{tikzpicture}
+}
+\vspace{0.3cm}\\
+\code{randomForest}\\
+\code{ranger}\\
+\code{party(kit)}}
+\end{center}
+\endminipage
+\hspace{0.65cm}
+\minipage{0.285\textwidth}
+\begin{center}
+\vspace{0.15cm}
+\visible<3->{
+Distributional trees and forests\\
+\vspace{1.15cm}
+\code{disttree}\\
+based on \code{partykit}\\
+\vspace{1cm}}
+\end{center}
+\endminipage
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Building blocks}
+%\vspace{0.5cm}
+
+\begin{minipage}{0.79\textwidth}
+\vspace{-0.5cm}
+\textbf{Distributional modeling:}
+\begin{itemize}
+\vspace{-0.05cm}
+  \item Specify the complete probability distribution \\
+  \vspace{-0.05cm}
+  (location, scale, shape, \dots).
+\end{itemize}
+
+\smallskip
+
+\textbf{Tree:}
+\begin{itemize}
+\vspace{-0.05cm}
+  \item Capture non-linear and non-additive effects.
+  \vspace{-0.1cm}
+  \item Automatic selection of covariates and interactions. %detection of steps and abrupt changes. %(data driven)
+\end{itemize}
+
+\smallskip
+
+\textbf{Forest:}
+\begin{itemize}
+\vspace{-0.05cm}
+  \item Smoother effects.
+  \vspace{-0.1cm}
+  \item Stabilization and regularization of the model.
+\end{itemize}
+
+\bigskip
+
+\bigskip
+
+$\Rightarrow$ \textbf{Synthesis:} Distributional Trees and Forests
+\end{minipage}
+<<motivation_GAMLSS_syn, echo=FALSE, results=hide>>=
+## GAMLSS
+set.seed(7)
+nobs <- 200
+x_g <- c(1:nobs)/nobs
+x_g <- 2*(x_g-0.5)
+ytrue_g <- x_g^3
+var_g <- exp(-(1.74*x_g)^2)/2.5
+y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var_g)
+@
+<<motivation_regtree_syn, echo=FALSE, results=hide>>=
+## Reg. Tree
+set.seed(7)
+kappa <- 12
+x <- c(1:nobs)/nobs
+ytrue <- ytree <- yforest <- var <- numeric(length = length(x))
+for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
+for(i in 1:nobs) var[i] <-  if(x[i]<1/3) 0.1 else 0.6 *(4/3 - x[i])
+y <- ytrue + rnorm(nobs, 0, 0.1 + var)
+for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
+@
+<<motivation_randforest_syn, echo=FALSE, results=hide>>=
+## Random Forest
+for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
+@
+\begin{minipage}{0.2\textwidth}
+\vspace{-0.05cm}
+<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=8.5>>=
+par(mar=c(1,0,1,0))
+plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
+polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g)),
+  col = pallight["forest"], border = "transparent")
+lines(x_g, ytrue_g, col = pal["forest"], lwd=10)
+points(x_g, y_g, col = "slategray", pch = 19)
+box(lwd = 5)
+@
+
+\vspace{0.15cm}
+
+<<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=8.5>>=
+par(mar=c(1,0,1,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(y), max(y)))
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
+lines(x = x, y = ytree, col = pal["forest"], lwd=10)
+box(lwd=5)
+@
+
+\vspace{0.15cm}
+
+<<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=8.5>>=
+par(mar=c(1,0,1,0))
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(y), max(y)))
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
+lines(x = x, y = yforest, col = pal["forest"], lwd=10, main = "")
+box(lwd=5)
+@
+
+\vspace{0.15cm}
+
+<<plot_motivation_distforest, fig=TRUE, echo=FALSE, width=8.5>>=
+par(mar=c(1,0,1,0))
+#plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
+#box(lwd=5)
+#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
+#lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
+plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
+     ylim = c(min(y), max(y)))
+polygon(c(x, rev(x)), c(yforest + 0.1 + var, rev(yforest - (0.1 + var))),
+       # c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g))
+  col = pallight["forest"], border = "transparent")
+box(lwd=5)
+lines(x = x, y = yforest, col = pal["forest"], lwd=10)
+@
+\end{minipage}
+\end{frame}
+
+<<treedata, echo=FALSE, results=hide>>=
+set.seed(54)
+nobs <- 500
+x <- runif(nobs, 0, 1)
+mu <- sigma <- ytrue <- numeric(length = nobs)
+for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
+for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
+y <- rnorm(nobs, mean = mu, sd = sigma)
+ytrue <- mu
+data <- data.frame(cbind(y,x, ytrue))
+
+alldata <- cbind(data, mu, sigma)
+odata <- alldata[order(alldata["x"]),]
+@
+
+<<dgp_tree, echo=FALSE, results=hide>>=
+data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
+names(data) <- c("x","x","x")
+fig <- party(
+  partynode(1L,
+            split = partysplit(2L, breaks = 0.4),
+            kids = list(
+              partynode(2L, info = c(
+                "n = 200",
+                "   True parameters:   ",
+                expression(mu == '4'),
+                expression(sigma == '1')
+              )),
+              partynode(3L,
+                        split = partysplit(3L, breaks = 0.8),
+                        kids = list(
+                          partynode(4L, info = c(
+                            "n = 200",
+                            "   True parameters:   ",
+                            expression(mu == '12'),
+                            expression(sigma == '3')
+                          )),
+                          partynode(5L, info = c(
+                            "n = 100",
+                            "   True parameters:   ",
+                            expression(mu == '4'),
+                            expression(sigma == '3')
+                          )))))),
+  data
+)
+
+node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
+    gp = gpar()) 
+{
+    meta <- obj$data
+    nam <- names(obj)
+    extract_label <- function(node) {
+        if (is.terminal(node)) 
+            return(rep.int("", 2L))
+        varlab <- character_split(split_node(node), meta)$name
+        if (abbreviate > 0L) 
+            varlab <- abbreviate(varlab, as.integer(abbreviate))
+        if (pval) {
+            nullna <- function(x) is.null(x) || is.na(x)
+            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
+                silent = TRUE))
+            pval <- if (inherits(pval, "try-error")) 
+                FALSE
+            else pval
+        }
+        if (pval) {
+            pvalue <- node$info$p.value
+            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
+                paste("p =", round(pvalue, digits = 3L)))
+        }
+        else {
+            plab <- ""
+        }
+        return(c(varlab, plab))
+    }
+    maxstr <- function(node) {
+        lab <- extract_label(node)
+        klab <- if (is.terminal(node)) 
+            ""
+        else unlist(lapply(kids_node(node), maxstr))
+        lab <- c(lab, klab)
+        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
+        lab <- lab[which.max(nchar(lab))]
+        if (length(lab) < 1L) 
+            lab <- ""
+        return(lab)
+    }
+    nstr <- maxstr(node_party(obj))
+    if (nchar(nstr) < 6) 
+        nstr <- "aAAAAa"
+    rval <- function(node) {
+        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
+            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
+            height = unit(3, "lines"), name = paste("node_inner", 
+                id_node(node), sep = ""), gp = gp)
+        pushViewport(node_vp)
+        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
+            seq(0.8, 1, by = 0.01))
+        yell <- sqrt(xell * (1 - xell))
+        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
+        lab <- extract_label(node)
+        fill <- rep(fill, length.out = 2L)
+        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
+            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
+        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
+            "lines"))
+        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
+        #    "lines"))
+        grid.text(lab[2L], y = unit(1, "lines"))
+        if (id) {
+            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
+                "npc"), width = max(unit(1, "lines"), unit(1.3, 
+                "strwidth", nam[id_node(node)])), height = max(unit(1, 
+                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
+            pushViewport(nodeIDvp)
+            grid.rect(gp = gpar(fill = fill[2]))
+            grid.text(nam[id_node(node)])
+            popViewport()
+        }
+        upViewport()
+    }
+    return(rval)
+}
+
+class(node_inner_ext) <- "grapcon_generator"
+
+@
+
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\vspace*{-0.12cm}
+\begin{center} 
+DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
+
+\vspace*{-0.21cm}
+\setkeys{Gin}{width=0.58\linewidth}
+<<plottree_xyplot, fig=TRUE, echo=FALSE,width=7>>=
+par(mar=c(5.1,4.1,2.4,1.1))
+plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "gray")
+lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
+polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
+  col = pallight["forest"], border = "transparent")
+legend("topleft", expression(mu  %+-%  sigma), bty = "n")
+@
+\end{center}
+\end{frame}
+
+% \begin{frame}
+% \frametitle{Distributional trees}
+% \vspace*{-0.12cm}
+% \begin{center} 
+% DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
+% 
+% \vspace*{-0.21cm}
+% \setkeys{Gin}{width=0.5\linewidth}
+% <<plottree_dgp, fig=TRUE, echo=FALSE>>=
+% paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+% plot(fig, inner_panel = node_inner_ext,
+%      tp_args = list(FUN = identity, width = 18, fill = paltrees[c(1, 3)]), 
+%      ip_args = list(fill = paltrees[c(2, 3)]),
+%      drop_terminal = TRUE, tnex = 1.7)
+% @
+% \end{center}
+% \end{frame}
+
+<<tree, echo=FALSE, results=hide>>=
+set.seed(7)
+nobs <- 500
+x <- runif(nobs, 0, 1)
+mu <- sigma <- ytrue <- numeric(length = nobs)
+for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
+for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
+y <- rnorm(nobs, mean = mu, sd = sigma)
+#y <- rcnorm(nobs, mean = mu, sd = sigma, left = 0)
+ytrue <- mu
+data <- data.frame(cbind(y,x, ytrue))
+tree <- disttree(y ~ x, data = data, family = NO(), type.tree = "mob")
+#tree <- disttree(y ~ x, data = data, family = dist_list_cens_normal)
+@
+
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.5\linewidth}
+<<plottree_estpar, fig=TRUE, echo=FALSE>>=
+# function for output in terminal panels
+FUN <- function (x) 
+{
+  cf <- x$coefficients
+  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
+  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
+                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
+}
+
+paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
+
+## plot version using FUN and tree of class 'disttree'
+plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
+     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.5\linewidth}
+<<plottree_box, fig=TRUE, echo=FALSE>>=
+plot(as.constparty(tree), tnex = 1.7, drop = TRUE,
+     tp_args = list(fill = paltrees[c(1, 3)], ylines = 1.5), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Distributional trees}
+\begin{center}
+\vspace*{-0.12cm}
+Model: \code{disttree(y ~ x)}\\
+\vspace*{-0.2cm}
+\setkeys{Gin}{width=0.5\linewidth}
+<<plottree_dens, fig=TRUE, echo=FALSE>>=
+node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
+                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
+                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
+{
+  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
+  ylines <- 1.5
+  if (is.null(xscale)) xscale <- c(-5.1,22.5)
+  if (is.null(yscale)) yscale <- c(-0.05,0.45)
+  xr <- xscale
+  yr <- yscale
+
+  if (horizontal) {
+    yyy <- xscale
+    xscale <- yscale
+    yscale <- yyy
+  }
+
+  rval <- function(node) {
+    yrange <- seq(from = -20, to = 90)/4
+    ydens <- node$info$object$ddist(yrange)
+
+    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3, 
+                                            widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")), 
+                                            heights = unit(c(1, 1), c("lines", "null"))), 
+                       width = unit(1, "npc"), 
+                       height = unit(1, "npc") - unit(2, "lines"), 
+                       name = paste("node_density",node$id, sep = ""))
+    pushViewport(top_vp)
+    grid.rect(gp = gpar(fill = "white", col = 0))
+    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
+    pushViewport(top)
+    mainlab <- paste(ifelse(id, paste("Node", node$id, "(n = "), "n = "), node$info$nobs, ifelse(id, ")", ""), sep = "")
+
+    grid.text(mainlab)
+    popViewport()
+    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2, 
+                     xscale = xscale, yscale = yscale, 
+                     name = paste("node_density",  node$id, "plot", sep = ""))
+    pushViewport(plot)
+    yd <- ydens
+    xd <- yrange
+    if (horizontal) {
+      yyy <- xd
+      xd <- yd
+      yd <- yyy
+      yyy <- xr
+      xr <- yr
+      yr <- yyy
+      rxd <- rep(0, length(xd))
+      ryd <- rev(yd)
+    } else {
+      rxd <- rev(xd)
+      ryd <- rep(0, length(yd))
+    }
+
+    if (rug) {
+      nodeobs <- node$info$object$y
+      if (horizontal) {
+        grid.rect(x = xscale[1], y = nodeobs , height = 0, width = xscale[1], 
+                  default.units = "native", just = c("right", "bottom"),
+		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
+      } else {
+        grid.rect(x = nodeobs, y = yscale[1], 
+                  width = 0, height = abs(yscale[1]), default.units = "native", 
+                  just = c("center", "bottom"),
+		  gp = gpar(lwd = 2, col = gray(0, alpha = 0.18)))
+        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
+        #           default.units = "native")
+        #grid.lines(x = xr, y = yr, gp = gpar(col = "lightgray"), 
+        #           default.units = "native")
+      }
+    }
+
+    grid.polygon(x = c(xd, rxd), y = c(yd, ryd), default.units = "native",
+              gp = gpar(col = "black", fill = fill, lwd = lwd))
+    #grid.lines(x = xd, y = yd, default.units = "native", 
+    #           gp = gpar(col = col, lwd = lwd))
+    grid.xaxis()
+    grid.yaxis()
+    grid.rect(gp = gpar(fill = "transparent"))
+    upViewport(2)
+  }
+  return(rval)
+}
+
+class(node_density) <- "grapcon_generator"
+
+plot(tree, tnex = 1.7, drop = TRUE,
+     terminal_panel = node_density,
+     tp_args = list(fill = paltrees[c(1, 3)]), 
+     ip_args = list(fill = paltrees[c(2, 3)]))
+@
+\end{center}
+\end{frame}
+%\begin{frame}
+%\frametitle{Distributional trees}
+%\center
+%\vspace*{-1cm}
+%\setkeys{Gin}{width=0.8\linewidth}
+%\includegraphics{tree_box}
+%\end{frame}
+
+\begin{frame}
+\frametitle{Learning distributional trees and forests}
+\begin{minipage}{0.7\textwidth}
+{\bf Tree:}
+\begin{enumerate}
+\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
+Estimate $\hat{\theta}$ via maximum likelihood \\
+$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
+\item<8-> Test for associations/instabilities of the \\
+scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each possible split variable $X_j$.
+\end{enumerate}
+\end{minipage}
+\begin{minipage}{0.23\textwidth}
+%\vspace{-0.1cm}
+\begin{tikzpicture}
+\visible<2-3>{
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
+}
+\visible<4>{
+\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{D}(Y;\hat{\theta}$)};
+}
+\visible<5>{
+\node[inner sep=0pt] (density_all) at (0.8, 1.7)
+{\includegraphics[width=0.7\textwidth]{density_all}};
+}
+\visible<6->{
+\node[inner sep=0pt] (density_all) at (0.8, 1.7)
+{\includegraphics[width=0.7\textwidth]{density_all_hist}};
+}
+\visible<7-9>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0.2) {?};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0.2) {?};
+\draw[-, gray, line width=0.5pt] (0.7, 1.16) -- (n1);
+\draw[-, gray, line width=0.5pt] (0.9, 1.16) -- (n2);
+}
+\visible<10-11>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0.2) {$Y_1$};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0.2) {$Y_2$};
+\draw[-, gray, line width=0.5pt] (0.7, 1.16) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (0.9, 1.16) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<12>{
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0.2) {$\mathcal{D}(Y_1;\hat{\theta}_1$)};
+\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0.2) {$\mathcal{D}(Y_2;\hat{\theta}_2$)};
+}
+\visible<12->{
+\draw[-, gray, line width=0.5pt] (0.7, 1.16) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
+\draw[-, gray, line width=0.5pt] (0.9, 1.16) -- (n2) node [midway, right] {\scriptsize $X > p$};
+}
+\visible<13->{
+\node[inner sep=0pt] (density_1_hist) at (-0.2,-0.04)
+{\includegraphics[width=0.6\textwidth]{density_2_hist}};
+}
+\visible<13->{
+\node[inner sep=0pt] (density_2_hist) at (1.8,-0.04)
+{\includegraphics[width=0.6\textwidth]{density_1_hist}};
+}
+\end{tikzpicture}
+\end{minipage}
+\vspace{0.1cm}
+%\begin{adjustwidth}{-0.0em}{-1em}
+\begin{enumerate}
+\setcounter{enumi}{2}
+\item<9-> Split along the covariate $X$ with strongest association or instability \\
+and at breakpoint $p$ with highest improvement in log-likelihood.
+\item<11-> Repeat steps 1--3 recursively until some stopping criterion \\
+is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
+\end{enumerate}
+%\end{adjustwidth}
+\vspace{0.4cm}
+\visible<14->{
+{\bf Forest:} Ensemble of $T$ trees.
+\begin{itemize}
+\item Bootstrap or subsamples.
+\item Random input variable sampling.
+\end{itemize}
+}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Weather forecasting}
+
+\textbf{Goal:}
+
+\medskip
+
+\begin{center}
+\begin{tikzpicture}
+\draw[->] (0.4,0)--(0.9,0);
+\draw[] (1,-0.4) rectangle (2.5,0.4);
+\node[font=\scriptsize] at (1.75,0) {nature};
+\draw[->] (2.6,0)--(3.1,0);
+\visible<1>{%
+\node[] at (0.0,0) {X};
+\node[] at (3.5,0) {Y};}
+\visible<2->{%
+\node[inner sep=0pt] (today) at (-2.3,-0.4)
+{\includegraphics[width=.365\textwidth]{airport_20200412.jpg}};
+\node[font=\scriptsize] at (-2.3,-2.4) {2020-04-12};
+\node[inner sep=0pt] (future) at (5.8,-0.4)
+{\includegraphics[width=.365\textwidth]{airport_20200413.jpg}};
+\node[font=\scriptsize] at (5.8,-2.4) {2020-04-13};}
+\end{tikzpicture}
+\end{center}
+
+%% https://innsbruck-airport.panomax.com/
+
+\only<1-2>{%
+\textbf{Data:}
+\begin{itemize}
+\item X: State of the atmosphere now (temperature, precipitation, wind, \dots).
+\item Y: State of the atmosphere in the future (hours, days, weeks, \dots).
+\end{itemize}}%
+\only<3->{%
+\textbf{Two stages:}
+\begin{itemize}
+\item Physical model: Numerical weather prediction (NWP).
+\item Statistical model: Model output statistics (MOS).
+\end{itemize}}
+
+\end{frame}
+
+% \begin{frame}[fragile]
+% \frametitle{Weather forecasting}
+% 
+% \textbf{Numerical weather prediction (NWP):}
+% \begin{itemize}
+%   \item Based on a physical model.
+%   \item Massive numerical simulation of atmospheric processes.
+%   \item Here: Global model on a $50\times 50 \text{ km}^2$ grid.
+% \end{itemize}
+% 
+% \medskip
+% 
+% \textbf{Problem:} Uncertain initial conditions, unresolved processes.
+% 
+% \medskip
+% 
+% \textbf{Solution:} Ensemble of simulation runs under perturbed conditions.
+% 
+% \end{frame}
+% 
+% 
+% 
+% <<echo = FALSE, results=hide>>=
+% library("zoo")
+% 
+% Sys.setenv("TZ"="UTC")
+% 
+% # -------------------------------------------------------------------
+% # GFS ensemble forecast
+% # -------------------------------------------------------------------
+% 
+% GFS   <- read.table("Data/GENS_00_innsbruck_20200412.dat", header = TRUE)
+% init  <- min(as.POSIXct(GFS$timestamp-GFS$step*3600,origin="1970-01-01"))
+% title <- sprintf("%s\n%s",
+%                  "Global Forecast System (GFS) Ensemble Forecast for Innsbruck, Airport",
+%                  sprintf("Forecast initialized %s", strftime(init, "%Y-%m-%d %H:%M UTC")))
+% 
+% # observations
+% load("Data/STAGEobs_tawes_11121_defense.rda")
+% #load("Data/STAGEobs_tawes_11121.rda")
+% # "2018-03-14 00:00:00" in line 321 (first entry for temperature)
+% # "2018-03-13 06:00:00" in line 317 (first entry for rain -> aggregated to 24h sums)
+% # "2018-03-23 00:00:00" in line 357
+% 
+% getZoo <- function( x, variable ) {
+%   x <- subset( x, varname == variable )
+%   if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
+%   # Else create zoo
+%   x <- zoo( subset(x,select=-c(varname,timestamp,step)),
+%             as.POSIXct(x$timestamp,origin="1970-01-01") )
+%   x
+% }
+% GFS_t2m  <- getZoo( GFS, "tmp2m" ) - 273.15
+% # temperature bias: model simulates temperature for height 1675 (~1070 m higher than true height)
+% # => add 10.7 degrees to account for bias (approx. 1 degree / 100 m)
+% # Custom bias
+% GFS_t2m <- GFS_t2m + 6.5
+% 
+% # Subsetting obs
+% obs <- subset(obs, index(obs) %in% index(GFS_t2m))
+% 
+% GFS_t2m <- merge(GFS_t2m, obs = obs[, grep("tl", names(obs))])
+% ###GFS_t2m$obs <- obs[,"obs$tl"]
+% # first value of rain is 18 hours later than temperature 
+% #(both start at 6:00 but rain is cumulated over 24 hours)
+% # -> drop first 3 values of temperature
+% GFS_t2m <- GFS_t2m[-c(1:3),]
+% 
+% GFS_rain <- getZoo( GFS, "apcpsfc" )
+% GFS_rain$obs <- obs[,"rr6"]
+% # precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
+% GFS_rain24 <- GFS_rain[c((1:10)*4),]
+% for(i in 1:NROW(GFS_rain24)){
+%   GFS_rain24[i,] <- colSums(GFS_rain[c((4*i-3):(4*i)),])
+% }
+% 
+% 
+% # Two POSIXct vectors for the axis
+% main  <- seq(min(as.POSIXct(as.Date(index(GFS_t2m)))),max(index(GFS_t2m)),by=86400)
+% minor <- seq(min(index(GFS_t2m)),max(index(GFS_t2m)),by=3*3600)
+% 
+% # GFS_t2m_mean <- rowMeans(GFS_t2m)
+% GFS_t2m_mean <- GFS_t2m
+% GFS_t2m_mean[,1] <- rowMeans(GFS_t2m[,-NCOL(GFS_t2m)])
+% GFS_t2m_mean <- GFS_t2m_mean[,1]
+% 
+% # GFS_rain_mean <- rowMeans(GFS_rain)
+% GFS_rain24_mean <- GFS_rain24
+% GFS_rain24_mean[,1] <- rowMeans(GFS_rain24[,-NCOL(GFS_rain24)])
+% GFS_rain24_mean <- GFS_rain24_mean[,1]
+% @
+% 
+% <<t2m, echo=FALSE, eval=FALSE>>=
+% par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
+% layout( matrix(1:2,ncol=1) )
+% plot(GFS_t2m[,1], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m, na.rm = TRUE)*1.05, col = "slategray" )
+% abline(v = main, lty = 2 )
+% mtext(side = 3, line = 1, cex = 1.2, font = 2, title )
+% mtext(side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
+% @
+% 
+% <<rain24, echo=FALSE, eval=FALSE>>=
+% plot(GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24, na.rm = TRUE)*c(0,1.05), col = "slategray" )
+% abline(v = main, lty = 2 )
+% mtext(side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
+% axis(side = 1, at = main + 42300,  strftime(main,"%b %d"),
+% line = 0, col.ticks=NA, col.axis="gray30", col = NA )
+% @
+% 
+% \begin{frame}[fragile]
+% \frametitle{Weather forecasting}
+% 
+% \vspace{-0.6cm}
+% \setkeys{Gin}{width=0.95\linewidth}
+% \begin{center}
+% \only<1>{%
+% <<gfs1, fig=TRUE, echo=FALSE, width=9>>=
+% <<t2m>>
+% <<rain24>>
+% @
+% }%
+% \only<2>{%
+% <<gfs2, fig=TRUE, echo=FALSE, width=9>>=
+% <<t2m>>
+% lines(GFS_t2m[,3], col = "slategray")
+% <<rain24>>
+% lines(GFS_rain24[,3], col = "slategray")
+% @
+% }%
+% \only<3>{%
+% <<gfs3, fig=TRUE, echo=FALSE, width=9>>=
+% <<t2m>>
+% lines(GFS_t2m[,2], col = "slategray")
+% lines(GFS_t2m[,3], col = "slategray")
+% <<rain24>>
+% lines(GFS_rain24[,2], col = "slategray")
+% lines(GFS_rain24[,3], col = "slategray")
+% @
+% }%
+% \only<4>{%
+% <<gfs4, fig=TRUE, echo=FALSE, width=9>>=
+% <<t2m>>
+% for(j in 2:(NCOL(GFS_t2m) - 1)) lines(GFS_t2m[,j], col = "slategray")
+% <<rain24>>
+% for(j in 2:(NCOL(GFS_rain24) - 1)) lines(GFS_rain24[,j], col = "slategray")
+% @
+% }%
+% \only<5>{%
+% <<gfs5, fig=TRUE, echo=FALSE, width=9>>=
+% <<t2m>>
+% for(j in 2:(NCOL(GFS_t2m) - 1)) lines(GFS_t2m[,j], col = "slategray")
+% lines(GFS_t2m[,"obs"], col = 2, lwd = 2.3)
+% <<rain24>>
+% for(j in 2:(NCOL(GFS_rain24) - 1)) lines(GFS_rain24[,j], col = "slategray")
+% lines(GFS_rain24[,"obs"], col = 2, lwd = 2.3)
+% @
+% }%
+% \end{center}
+% 
+% \end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Weather forecasting: precipitation}
+
+\textbf{Goal:} Predict daily precipitation amount in Tyrol. %complex terrain.
+
+\bigskip
+\pause
+
+\textbf{Observation data:} National Hydrographical Service.
+\begin{itemize}
+\item Daily 24h precipitation sums from July over 28 years (1985--2012).
+\item 95 observation stations in Tyrol.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Numerical weather prediction:} Global Ensemble Forecast System.
+\begin{itemize}
+\item Model outputs: Precipitation, temperature, air pressure, 
+  convective available potential energy, downwards short wave radiation flux, \dots
+\item 80 covariates based on ensemble min/max/mean/standard deviation.
+\end{itemize}
+
+\bigskip
+\pause
+
+\textbf{Distribution assumption:} Power-transformed Gaussian, censored at 0.
+\[ (\text{precipitation})^\frac{1}{1.6} \sim \textit{c}\mathcal{N}(\mu,\sigma^2) \]
+
+\end{frame}
+
+%% \begin{frame}
+%% \frametitle{Weather forecasting: precipitation}
+%% 
+%% \textbf{Base variables:}
+%% \begin{itemize}
+%% \item Total precipitation.
+%% \item Convective available potential energy.
+%% \item Downwards short wave radiation flux (``sunshine'').
+%% \item Mean sea level pressure.
+%% \item Preciptable water.
+%% \item 2m maximum temperature.
+%% \item Total column-integrated condensate.
+%% \item Temperature.
+%% \item Temperature differences in altitude.
+%% \end{itemize}
+%% 
+%% \medskip
+%% 
+%% \textbf{Variations:} 80 covariates based on ensemble min/max/mean/standard deviation.
+%% 
+%% \end{frame}
+
+<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
+if(file.exists("Data/Axams_pred.rda") & file.exists("Data/Axams_testdata.rda")){
+  load("Data/Axams_pred.rda")
+  load("Data/Axams_testdata.rda")
+} else {
+  #####
+  # load observations and covariates 
+  data("RainAxams")
+
+  # tree and forest formula
+  {
+    dt.formula <- df.formula <- 
+      robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
+      tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
+      tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
+      capepow_mean + capepow_sprd + capepow_min + capepow_max + 
+      capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
+      capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
+      dswrf_mean_mean + dswrf_mean_max +  
+      dswrf_sprd_mean + dswrf_sprd_max + 
+      msl_mean_mean + msl_mean_min + msl_mean_max + 
+      msl_sprd_mean + msl_sprd_min + msl_sprd_max +
+      pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
+      pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
+      tmax_mean_mean + tmax_mean_min + tmax_mean_max +
+      tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
+      tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
+      tcolc_sprd_mean + tcolc_sprd_min + tcolc_sprd_max +
+      t500_mean_mean + t500_mean_min + t500_mean_max +
+      t700_mean_mean + t700_mean_min + t700_mean_max +
+      t850_mean_mean + t850_mean_min + t850_mean_max +
+      t500_sprd_mean + t500_sprd_min + t500_sprd_max +
+      t700_sprd_mean + t700_sprd_min + t700_sprd_max +
+      t850_sprd_mean + t850_sprd_min + t850_sprd_max +
+      tdiff500850_mean + tdiff500850_min + tdiff500850_max +
+      tdiff700850_mean + tdiff700850_min + tdiff700850_max +
+      tdiff500700_mean + tdiff500700_min + tdiff500700_max +
+      msl_diff
+
+  }
+
+  # learning data: 24 years (1985 - 2008, both inlcuded)
+  # testing data: 4 successive years (2009, 2010, 2011, 2012)
+  learndata <- RainAxams[RainAxams$year < 2009,]
+  testdata <- RainAxams[RainAxams$year %in% c(2009, 2010, 2011, 2012),]
+  save(file = "Data/Axams_testdata.rda", testdata)
+
+  ##############################################################
+  # fitting the models
+  set.seed(7)
+
+  df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
+                   ntree = 100, censtype = "left", censpoint = 0,
+                   control = disttree_control(teststat = "quad", testtype = "Univ",
+                                              type.tree = "ctree",
+                                              intersplit = TRUE,
+                                              mincriterion = 0, minsplit = 50,
+                                              minbucket = 20), mtry = 27)
+
+  #### prepare data for plot of estimated density functions
+  # predictions for one day (in each of the four years) 
+  # (19th of July 2011 is missing)
+  pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),
+
+  pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
+  pdf <- predict(df, newdata = testdata[pdays,], type = "parameter")
+  pdf$pdays <- pdays
+  save(file = "Data/Axams_pred.rda", pdf)
+}
+
+pdays <- pdf$pdays
+pday <- pdays[1]
+df_mu <- pdf$mu
+df_sigma <- pdf$sigma
+
+# plot predicted distributions together with observations
+set.seed(7)
+x <- c(0.01, sort(runif(500,0.01,8)))
+y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
+y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
+y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
+y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
+
+# switch x-axis back to untransformed scale
+# x <- x^(1.6)
+
+# point mass (slightly shifted)
+pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))     #  0.15
+pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))     #  0.05
+pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))    # -0.05
+pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))    # -0.15
+
+# predictions
+pred1 <- c(testdata[pdays,"robs"][1], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
+pred2 <- c(testdata[pdays,"robs"][2], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
+pred3 <- c(testdata[pdays,"robs"][3], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
+pred4 <- c(testdata[pdays,"robs"][4], # ^(1.6), 
+           crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))
+
+#legendheight
+lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
+lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
+lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
+lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
+@
+
+\begin{frame}[fragile]
+\frametitle{Weather forecasting: precipitation}
+
+\textbf{Application for one station:} Axams.
+\begin{itemize}
+%\vspace{-0.05cm}
+\item Learn forest model on data from 24 years (1985--2008).
+%\vspace{-0.05cm}
+\item Evaluate on 4 years (2009--2012). Here: July \Sexpr{24}.
+\end{itemize}
+
+\vspace*{-0.4cm}
+\begin{center}
+\setkeys{Gin}{width=0.54\textwidth}
+<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=4.3, width=7>>=
+par(mar = c(4, 4, 1, 1))
+plot(x = x, y = y1, type = "l", col = 2, ylab = "Density", lwd = 2,
+     # xlab = expression(Total~precipitation~"["~mm~"/"~"24h"~"]"),
+     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
+     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
+     xlim = c(-1.5,8))     #  xlim = c(-2,28))
+
+lines(x = x, y = y2, type = "l", lwd = 2, col = 4)
+lines(x = x, y = y3, type = "l", lwd = 2, col = 3)
+lines(x = x, y = y4, type = "l", lwd = 2, col = 6)
+legend('topright', c("Predicted distribution", "Point mass at censoring point", "Observation"),
+       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 1)
+
+# plot point mass
+lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = 2, type = "l", lwd = 2)
+lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = 4, type = "l", lwd = 2)
+lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = 3, type = "l", lwd = 2)
+lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = 6, type = "l", lwd = 2)
+
+points(x = pm1[1], y = pm1[2], col = 2, pch = 19, cex = 1.4)
+points(x = pm2[1], y = pm2[2], col = 4, pch = 19, cex = 1.4)
+points(x = pm3[1], y = pm3[2], col = 3, pch = 19, cex = 1.4)
+points(x = pm4[1], y = pm4[2], col = 6, pch = 19, cex = 1.4)
+
+# plot predictions
+points(x = pred1[1], y = pred1[2], col = 2, pch = 4, cex = 1.4, lwd = 2)
+points(x = pred2[1], y = pred2[2], col = 4, pch = 4, cex = 1.4, lwd = 2)
+points(x = pred3[1], y = pred3[2], col = 3, pch = 4, cex = 1.4, lwd = 2)
+points(x = pred4[1], y = pred4[2], col = 6, pch = 4, cex = 1.4, lwd = 2)
+
+lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)
+lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)
+lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)
+lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 2)
+
+# add labels
+text(x = -0.8, y = lh1, labels = "2009", col = 2, cex = 1)      # -1.7 
+text(x = -0.8, y = lh2, labels = "2010", col = 4, cex = 1)      # -1.7
+text(x = -0.8, y = lh3, labels = "2011", col = 3, cex = 1)      # -1.7
+text(x = -0.8, y = lh4, labels = "2012", col = 6, cex = 1)      # -1.7
+@
+\end{center}
+
+\only<2>{
+\vspace{-0.5cm}
+{\small Replication: \texttt{demo("RainAxams", package = "disttree")}}
+}
+\end{frame}
+
+% \begin{frame}[fragile]
+% \frametitle{Weather forecasting: precipitation}
+% 
+% \textbf{Application for one station:} Axams.
+% \begin{itemize}
+% %\vspace{-0.05cm}
+% \item Learn forest model on data from 24 years.
+% %\vspace{-0.05cm}
+% \item Evaluate on 4 years.
+% %\vspace{-0.05cm}
+% \item 10 times 7-fold cross validation.
+% \end{itemize}
+% 
+% \bigskip
+% 
+% \textbf{Benchmark:} Against other heteroscedastic censored Gaussian models.
+% \begin{itemize}
+% \item \emph{Ensemble MOS:} Linear predictors using only total precipitation.
+% \item \emph{Prespecified GAMLSS:} Variable selection based on expert knowledge.
+% \item \emph{Boosted GAMLSS:} Automatic variable selection.
+% \end{itemize}
+% 
+% \bigskip
+% 
+% \textbf{Evaluation:} Continuous ranked probability skill score.
+% 
+% %% \vspace{0.1cm}
+% %% \begin{table}[t!]
+% %% \footnotesize
+% %% 
+% %% \hskip-0.4cm\begin{tabular}{ l l l }
+% %% \hline
+% %% Model & Type & Variable Selection \\
+% %% \hline
+% %% \vspace{-0.15cm}
+% %% \textbf{Distributional forest} & recursive    & automatic \\ 
+% %%                                & partitioning &           \\ 
+% %% \vspace{-0.15cm}
+% %% \textbf{Prespecified GAMLSS}   & spline  & based on expert \\
+% %%                                & in each & knowledge      \\
+% %% \vspace{-0.15cm}
+% %% \textbf{Boosted GAMLSS}        & spline  & automatic  \\
+% %%                                & in each &            \\
+% %% \vspace{-0.15cm}
+% %% \textbf{EMOS}                  & linear & total precipitation mean \\
+% %%                                &        & and standard deviation   \\
+% %% \hline
+% %% \end{tabular}
+% %% \end{table}
+% 
+% \end{frame}
+% 
+% 
+% 
+% 
+% <<echo=FALSE, results=hide>>=
+% #### cross validation rain
+% if(file.exists("Data/crps_cross.rda")){
+%   load("Data/crps_cross.rda")
+% } else {
+%   
+%   nrep_cross <- 10
+%   seed <- 7
+%   
+%   res_cross <- mclapply(1:nrep_cross,
+%                         function(i){
+%                           
+%                           set.seed(seed*i)
+%                           
+%                           # randomly split data in 7 parts each including 4 years
+%                           years <- 1985:2012
+%                           testyears <- list()
+%                           for(j in 1:7){
+%                             testyears[[j]] <- sample(years, 4, replace = FALSE)
+%                             years <- years[!(years %in% testyears[[j]])]
+%                           }
+%                           
+%                           #crps <- matrix(nrow = 7, ncol = 7)
+%                           reslist <- list()
+%                           for(k in 1:7){
+%                             test <- testyears[[k]]
+%                             train <- c(1985:2012)[!c(1985:2012) %in% test]
+%                             
+%                             res <- evalmodels(station = "Axams",
+%                                               train = train,
+%                                               test = test,
+%                                               gamboost_cvr = TRUE)
+%                             
+%                             #crps[k,] <- res$crps
+%                             reslist[[k]] <- res
+%                           }
+%                           
+%                           #colnames(crps) <- names(res$crps)
+%                           return(reslist)
+%                         },
+%                         mc.cores = detectCores() - 1
+%   )
+%   
+%   # extract CRPS
+%   crps_cross <- matrix(nrow = nrep_cross, ncol = 7)
+%   # loop over all repetitions
+%   for(i in 1:length(res_cross)){
+%     #loop over all 7 folds (for 7 methods)
+%     crps_cross_int <- matrix(nrow = length(res_cross[[1]]), ncol = 7)
+%     for(j in 1:length(res_cross[[1]])){
+%       crps_cross_int[j,] <- res_cross[[i]][[j]]$crps
+%     }
+%     crps_cross[i,] <- colMeans(crps_cross_int, na.rm = TRUE)
+%   }
+%   colnames(crps_cross) <- names(res_cross[[1]][[1]]$crps) 
+%   
+%   save(crps_cross, file = "Data/crps_cross.rda")
+% }
+% @
+% 
+% 
+% 
+% 
+% \begin{frame}[fragile]
+% \frametitle{Weather forecasting: precipitation}
+% \begin{center}
+% \vspace*{-0.2cm}
+% \setkeys{Gin}{width=0.65\textwidth}
+% <<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.6, width=7>>=
+% #par(mar = c(2.5,2,1,2))
+% boxplot(1 - crps_cross[,c(2,3,4)] / crps_cross[,6], ylim = c(-0.005, 0.065),
+%         names = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS"),
+%         main = "Cross validation (with reference model EMOS)",
+%         cex.main=1.4,
+%         cex.lab=1.2,
+%         ylab = "CRPS skill score", col = "lightgray") 
+% abline(h = 0, col = pal["EMOS"], lwd = 2)
+% @
+% \end{center}
+% \end{frame}
+
+% \begin{frame}[fragile]
+% \frametitle{Weather forecasting: precipitation}
+% %\vspace{0.2cm}
+% \textbf{Application for all 95 stations:}
+% \begin{itemize}
+% \item Learn forest model on data from 24 years (1985--2008).
+% \item Evaluate on 4 years (2009--2012).
+% \item Benchmark against other heteroscedastic censored Gaussian models.
+% \end{itemize}
+% 
+% \end{frame}
+% 
+% 
+% <<echo = FALSE>>=
+% #### prediction over all stations 24 - 4
+% if(file.exists("Data/crps_24to4_all.rda")){
+%   load("Data/crps_24to4_all.rda")
+% } else {
+%   
+%   data("StationsTyrol")
+%   stations <- StationsTyrol$name
+%   test <- 2009:2012
+%   train <- 1985:2008
+%   
+%   
+%   res_24to4_all <- mclapply(1:length(stations),
+%                             function(i){
+%                               
+%                               set.seed(7)
+%                               
+%                               res <- evalmodels(station = stations[i],
+%                                                 train = train,
+%                                                 test = test,
+%                                                 gamboost_cvr = TRUE)
+%                               
+%                               return(res)
+%                             },
+%                             mc.cores = detectCores() - 1
+%   )
+%   
+%   # extract crps
+%   crps_24to4_all <- matrix(nrow = length(stations), ncol = 7)
+%   # loop over all stations
+%   for(i in 1:length(stations)){
+%     crps_24to4_all[i,] <- res_24to4_all[[i]]$crps
+%   }
+%   
+%   colnames(crps_24to4_all) <- names(res_24to4_all[[1]]$crps)
+%   rownames(crps_24to4_all) <- stations
+%   
+%   save(crps_24to4_all, file = "Data/crps_24to4_all.rda")
+% }
+% 
+% 
+% # skill score
+% s <- 1 - crps_24to4_all[, 2:4]/crps_24to4_all[,6] 
+% colnames(s) <- c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS")
+% 
+% 
+% ## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps
+% 
+% crps_map <- crps_24to4_all[,c("distforest", "gamlss", "gamboostLSS", "emos_log")]  
+% 
+% # best method
+% bst <- apply(crps_map, 1, which.min)
+% 
+% # distance of forest to best other method
+% dst <- crps_map[,1] - crps_map[cbind(1:nrow(crps_map), apply(crps_map[, -1], 1, which.min) + 1)]
+% 
+% # breaks/groups
+% brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
+% #brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
+% grp <- cut(dst, breaks = brk)
+% 
+% # HCL colors (relatively flashy, essentially CARTO Tropic)
+% clr <- colorspace::diverging_hcl(5, h = c(130, 320), c = 70, l = c(50, 90), power = 1.3)
+% 
+% 
+% library("raster") # dem (digital elevation model)
+% library("sp")     # gadm www.gadm.org/country
+% 
+% data("StationsTyrol", package = "RainTyrol")
+% data("MapTyrol", package = "RainTyrol")
+% # data(MapTyrol_border, package = "RainTyrol")
+% # Create SpatialPointsDataFrame from station list
+% sp <- SpatialPointsDataFrame(subset(StationsTyrol,
+%                                     select=c(lon,lat)),
+%                              data = subset(StationsTyrol,
+%                                            select = -c(lon,lat)),
+%                              proj4string = crs(MapTyrol$RasterLayer))
+% @
+% 
+% 
+% \begin{frame}[fragile]
+% \frametitle{Weather forecasting: precipitation}
+% \begin{center}
+% \vspace*{-0.5cm}
+% 
+% \setkeys{Gin}{width=0.84\textwidth}
+% <<map, fig=TRUE, echo=FALSE, width=10, height=6.5>>=
+% 
+%   ## plot map of Tyrol with all 95 observations
+%   layout(cbind(1, 2), width = c(9, 1))
+%   par(mar = c(5,4,4,0.1))
+%   raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
+%                 main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
+%                 xlim = c(9.8,13.2), 
+%                 ylim = c(46.6, 47.87))
+%   plot(MapTyrol$SpatialPolygons, add = TRUE)
+%   points(sp[70,], pch = 19, col = "black", cex = 1.85)
+%   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
+%   legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
+%   text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
+%   mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
+%   par(mar = c(0.5,0.2,0.5,2.3))
+%   ## legend
+%   plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
+%        xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
+%   rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
+%   axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))
+% 
+% 
+% @
+% \end{center}
+% \end{frame}
+% 
+% 
+% 
+% \begin{frame}[fragile]
+% \frametitle{Weather forecasting: precipitation}
+% \begin{center}
+% \vspace*{-0.5cm}
+% 
+% \setkeys{Gin}{width=0.47\textwidth}
+% <<map, fig=TRUE, echo=FALSE, width=10, height=6.5>>=
+% 
+%   ## plot map of Tyrol with all 95 observations
+%   layout(cbind(1, 2), width = c(9, 1))
+%   par(mar = c(5,4,4,0.1))
+%   raster::image(MapTyrol$RasterLayer, col = rev(gray.colors(100)),
+%                 main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
+%                 xlim = c(9.8,13.2), 
+%                 ylim = c(46.6, 47.87))
+%   plot(MapTyrol$SpatialPolygons, add = TRUE)
+%   points(sp[70,], pch = 19, col = "black", cex = 1.85)
+%   points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
+%   legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
+%   text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
+%   mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
+%   par(mar = c(0.5,0.2,0.5,2.3))
+%   ## legend
+%   plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
+%        xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
+%   rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
+%   axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))
+% 
+% 
+% @
+% \end{center}
+% \vspace{-0.2cm}
+% \textbf{Conclusions:}
+% \begin{itemize}
+% \item Distributional forests provide an easy-to-apply alternative to already existing methods.
+% \item Do not require expert knowledge for pre-specifications.
+% \item Competitive performance: mainly en par or even slightly better.
+% \end{itemize}
+% 
+% \end{frame}
 

### 2020-07-03 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update ref tease paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3024 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 47fc84bf7ba5483a3dfd947b1cb6fd5794ebcc0f
Mini-adaption in the computational details
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3039 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8e103c565f91bed9d689ce4f3456b6c820760968

@@
     and is discussed in a comprehensive summary on different time-adaptive training schemes in
-    \citet{Lang+Lerch+Mayr:2019}.
+    \citet{Lang+Lerch+Mayr:2020}.
     %%
@@
     centered around the day of interest over all available years in the training
-    data \citep{Lang+Lerch+Mayr:2019}.
+    data \citep{Lang+Lerch+Mayr:2020}.
     %%
@@
 
-The computation time on a standard computer is in the order of 10\,milliseconds
-for the circular fit (used for the persistency and the climatological model),
+The computation time on a standard computer is in the order of milliseconds for
+the circular fit (used for the persistency and the climatological model),
 seconds for the estimation of the linear model and the circular tree, and from
@@
 between the two locations included in the application. However, for a single
-station and forecast step, the persistency and linear models must be
-re-estimated for each time of interest, and independent climatological models
-have to be fitted for all times of interest within a single calendar year,
-whereas for all times of interest only a single circular tree and a single
-circular forest model is required.
+station and forecast step, the persistency model must be re-estimated for each
+time of interest, independent climatological and linear models have to be
+fitted for all times of interest within a single calendar year, whereas for all
+times of interest only a single circular tree and a single circular forest
+model is required.
 

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
rephrased cover letter somewhat
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3040 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 65954973e062106eb6d36d9683c2b2c9ac7d6d49

@@
 \begin{letter}[fromemail]{To Prof. Peter Smith}
-\opening{Dear Prof. Peter Smith}
+\opening{Dear Prof.\ Smith,}
 
-We thank you, as well as the associate editor and both reviewers, for the
+we thank you, as well as the associate editor and both reviewers, for the
 positive and constructive feedback regarding our manuscript ``\textbf{Circular
@@
 
-We have carefully revised our manuscript according to given comments.
-The most substantial changes in the manuscript are the following:
+Due to the COVID-19 pandemic our response took us longer than planned but
+we have finally carefully revised our manuscript, incorporating the feedback
+and suggestions. The most substantial changes in the manuscript are the following:
 
 \begin{itemize}
-\item We have fully revised the `Abstract' in order to put a larger
-focus on the application. In addition, we now use the term `application' 
-instead of `case study' throughout the manuscript and have also renamed Sect.~4 accordingly.
+\item We have fully revised the `Abstract' giving more emphasis to the application.
+In addition, we now use the term `application' instead of `case study' throughout
+the manuscript and have also renamed Sect.~4 accordingly.
 
-\item In the section `Computational details', we have additionally listed the needed
-computing time of the different methods on a standard desktop computer.
+\item In the section `Computational details', we have added information about the
+order of magnitude of the computing time for the different methods on a standard desktop computer.
 
-\item We now discuss the work of Hara and Chellappa~(2017) and compare their distribution-based 
-tree-building approach to our implementation of circular distributional trees within Sect.~1.1.
+\item We now discuss the work of Hara and Chellappa~(2017) and their distribution-based 
+tree algorithm in relation to our circular distributional trees within Sect.~1.1.
 
@@
 
-\section{Editor: Prof. Peter Smith - Comments to the Author}
+\section{Editor: Prof.\ Peter Smith -- Comments to the Author}
 

### 2020-07-05 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
small corrections in notation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3041 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 018bf82c7d3ba542ca6499e024ddb65f1576c5d3

@@
 \item Combine an ensemble of trees. 
-\item Each tree can capture non-linear and non-additive effects and select covariates and possible interactions automatically.
+\item A single tree can capture non-linear and non-additive effects and select covariates and possible interactions automatically.
 \item Combining trees to a forest model allows for an approximation of smooth effects.
@@
 
-$\hat{y}_t= \frac{1}{|\mathcal{B}^t_x|} \sum_{i: y_i \in \mathcal{B}^t_x} y_i$
+$\hat{y}_t= \frac{1}{|\mathcal{B}^t_x|} \sum_{i: x_i \in \mathcal{B}^t_x} y_i$
 \visible<15->{
@@
 \frametitle{Aggregation of trees}
-{\bf Simple tree model:} Obtain the predicted response $\hat{y}$ for an observed covariate~$x$ by averaging over observations in the same terminal node $\mathcal{B}_x$: 
+{\bf Simple tree model:} Obtain the predicted response $\hat{y}$ for an observed covariate~$x$ by averaging over observations in the terminal node~$\mathcal{B}_x$: 
 $$
-\hat{y}= \frac{1}{|\mathcal{B}_x|} \sum_{i: y_i \in \mathcal{B}_x} y_i
+\hat{y}= \frac{1}{|\mathcal{B}_x|} \sum_{i: x_i \in \mathcal{B}_x} y_i
 \visible<2->{
@@
 \visible<4->{
-{\bf Model-based tree:} Obtain the predicted model parameter(s) $\hat{\theta}$ for an observed covariate~$x$ by optimizing a loss function for all observations in the same terminal node $\mathcal{B}_x$:
+{\bf Model-based tree:} Obtain the predicted model parameter(s) $\hat{\theta}$ for an observed covariate~$x$ by optimizing a loss function for all observations in the terminal node~$\mathcal{B}_x$:
 $$
-\hat{\theta}= \argmin_{\theta} \sum_{i: y_i \in \mathcal{B}_x} loss(\theta; y_i)
+\hat{\theta}= \argmin_{\theta} \sum_{i: x_i \in \mathcal{B}_x} loss(\theta; y_i)
 \ =\ \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot loss(\theta; y_i)$$
@@
 \frametitle{Aggregation of trees}
-{\bf Strategy:} Combine trees to a forest model by employing tree-based weights.
+{\bf Strategy:} Combine T trees to a forest model by employing tree-based weights.
 
 The predicted model parameter (vector) $\hat{\theta}$ 
-is the argument of the parameter space $\Theta$ 
+is the argument %of the parameter space $\Theta$ 
 that minimizes the weighted sum of a loss function evaluated for learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:
@@
 \visible<3-4>{
-w^{\text{tree}}_i(x)   & = & \sum_{b=1}^B I((x_i \in \mathcal{B}_b) \land (x \in \mathcal{B}_b)) \\[0.1cm]
+w^{\text{tree}}_i(x)   & = & \frac{I(x_i \in \mathcal{B}_x)}{\left|\mathcal{B}_x\right|}  \\[0.1cm]
 \visible<4>{
-w^{\text{forest}}_i(x) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} \frac{I((x_i \in \mathcal{B}^t_b) \land (x \in \mathcal{B}^t_b))}{\left|\mathcal{B}^t_b\right|}
+w^{\text{forest}}_i(x) & = & \frac{1}{T} \sum_{t=1}^T \frac{I(x_i \in \mathcal{B}^t_x)}{\left|\mathcal{B}^t_x\right|}
 \end{eqnarray*}

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include Achim's newly written abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3042 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a3ef4b7b7d9b54c87f20174fa2e674c971566acb

@@
 %
-%NEW VERSION:
-Circular data occur in a wide range of scientific fields, however the
-methodology for distributional modeling and probabilistic forecasting of
-circular response variables is rather limited. While most of the existing
-methods are built within the framework of generalized linear and additive
-models, which are often challenging to optimize and interpret, we propose a
-distributional approach for regression trees yielding probabilistic forecasts
-based on the von~Mises distribution. The resulting tree-based models use the
-available covariates for partitioning the data into sufficiently homogeneous
-subgroups, allowing for a simple circular distributional fit in each subgroup.
-These circular regression trees are straightforward to interpret, can capture
-nonlinear effects and interactions, and automatically select the relevant
-covariates that are associated with changes in any distribution parameter.
-Ensembles or forests of such circular trees can further improve the forecasts
-by regularizing and stabilizing the model. The newly developed methodology is
-illustrated in an application on short-term probabilistic wind direction
-forecasting for Innsbruck and Vienna Airport required to direct airplanes to a
-safe landing. The predictive skill of the novel approaches is benchmarked with
-an additive regression model plus a persistency and a climatological model,
-employing the circular continuous ranked probability score.}
+%NEW LISA AND MORITZ VERSION
+%Circular data occur in a wide range of scientific fields, however the
+%methodology for distributional modeling and probabilistic forecasting of
+%circular response variables is rather limited. While most of the existing
+%methods are built within the framework of generalized linear and additive
+%models, which are often challenging to optimize and interpret, we propose a
+%distributional approach for regression trees yielding probabilistic forecasts
+%based on the von~Mises distribution. The resulting tree-based models use the
+%available covariates for partitioning the data into sufficiently homogeneous
+%subgroups, allowing for a simple circular distributional fit in each subgroup.
+%These circular regression trees are straightforward to interpret, can capture
+%nonlinear effects and interactions, and automatically select the relevant
+%covariates that are associated with changes in any distribution parameter.
+%Ensembles or forests of such circular trees can further improve the forecasts
+%by regularizing and stabilizing the model. The newly developed methodology is
+%illustrated in an application on short-term probabilistic wind direction
+%forecasting for Innsbruck and Vienna Airport required to direct airplanes to a
+%safe landing. The predictive skill of the novel approaches is benchmarked with
+%an additive regression model plus a persistency and a climatological model,
+%employing the circular continuous ranked probability score.
+While circular data occur in a wide range of scientific fields, the methodology 
+for distributional modeling and probabilistic forecasting of circular response
+variables is rather limited. Most of the existing methods are built on the
+framework of generalized linear and additive models, which are often challenging
+to optimize and interpret. Specifically, capturing abrupt changes or
+interactions is not straightforward but often relevant, e.g., for modeling wind
+directions subject to different wind regimes. Additionally, automatic covariate
+selection is desirable when many predictor variables are available, as often the
+case in weather forecasting.
+
+To address these challenges we suggest a general distributional approach using
+regression trees and random forests to obtain probabilistic forecasts for
+circular responses. Using trees simplifies model estimation as
+covariates are used only for partitioning the data and subsequently just a
+simple von~Mises distribution is fitted in the resulting subgroups.
+Circular regression trees are straightforward to interpret, can capture
+nonlinear effects and interactions, and automatically select covariates
+affecting location and/or scale in the von~Mises distribution. Circular random
+forests regularize and smooth the effects from an ensemble of trees.
+
+The new methods are applied to probabilistic wind direction forecasting at two
+Austrian airports, considering other common approaches as a benchmark.}
 

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include post-pruning and adapt computation times
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3044 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d8fd6d55b26bad4b64fd1299cd20da584bb3318a

@@
 times of interest only a single circular tree and a single circular forest
-model is required.
+model is required. In general with respect to other applications, the computation
+time for the estimation of circular regression trees and forests can be reduced by
+various implemented settings, such as, e.g., the use of additional CPU cores or
+binning of the input data (c.f.  Section~\ref{sec:wind:models}).
 
@@
 forest models by setting the significance level to $1$.
+
+An alternative split variable selection strategy is provided by cost-complexity
+``post-pruning'', where a large tree is first built and then pruned back based
+on the predictive performance \citep{Breiman+Friedman+Stone:1984}. As discussed
+in \citet{Zeileis+Hothorn+Hornik:2008}, for model-based partitioning by
+employing information criteria such as AIC~(Akaike information criterion) or
+BIC~(Bayes information criterion) this strategy can be built on the same
+underlying loss function as pre-pruning. However,
+\citet{Schlosser+Hothorn+Zeileis:2019} show that post-pruning in an unbiased
+tree algorithm is usually only superior to pre-pruning when the underlying
+significance tests have low power, for example, due to a large data set.
 

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Adapt answer in rebutall referring to BIC
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3045 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5b35ed48e7ca7c0efb73b9f43eeeeb9e7a38563a

@@
 
-Regarding your suggestion, when building a circular distributional tree, the
-package \textbf{circtree} uses per default \emph{p}-values from association
-tests both for selecting the split variable and for a so-called pre-pruning
-strategy which stops growing the tree as soon as no significant associations
-can be found in a given subgroup. This is the default in both available
-unbiased recursive partitioning algorithms CTree and MOB. For the latter, as
-discussed by Zeileis~\emph{et al.}~(2008), an alternative, already implemented
-strategy  is to use information criteria such as AIC (Akaike information
-criterion) or BIC (Bayes information criterion) for post-pruning (after growing
-the initial tree).  As we use neither MOB nor post-pruning in the presented
-application, we refrain from discussing this aspect any further in the
-manuscript. However, we want to refer to the submitted manuscript by
+A very good remark, indeed split variable selection based on BIC creates an
+alternative possible tree-building strategy. Per default when building a
+circular distributional tree, the package \textbf{circtree} uses
+\emph{p}-values from association tests both for selecting the split variable
+and for a so-called pre-pruning strategy which stops growing the tree as soon
+as no significant associations can be found in a given subgroup. However, as
+discussed by Zeileis~\emph{et al.}~(2008), an alternative strategy is to use
+information criteria such as AIC (Akaike information criterion) or BIC (Bayes
+information criterion) for cost-complexity post-pruning, where a large tree is
+first built and then pruned back. We now discuss this aspect in the Appendix~A
+of the manuscript and also refer to the submitted manuscript by
 Schlosser~\emph{et al.}~2019 which investigates the performance of different
-testing and pruning strategies employed in unbiased recursive partitioning algorithms.
+testing and pruning strategies employed in unbiased recursive partitioning
+algorithms.
 

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Minor adaptions in rebuttal
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3046 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 910a180b7431a990c81498e2710767a1e7f1bf87

@@
 as no significant associations can be found in a given subgroup. However, as
-discussed by Zeileis~\emph{et al.}~(2008), an alternative strategy is to use
+discussed in Zeileis~\emph{et al.}~(2008), an alternative strategy is to use
 information criteria such as AIC (Akaike information criterion) or BIC (Bayes
 information criterion) for cost-complexity post-pruning, where a large tree is
-first built and then pruned back. We now discuss this aspect in the Appendix~A
+first built and then pruned back. We now discuss post-pruning in the Appendix~A
 of the manuscript and also refer to the submitted manuscript by
-Schlosser~\emph{et al.}~2019 which investigates the performance of different
+Schlosser~\emph{et al.}~2019, which investigates the performance of different
 testing and pruning strategies employed in unbiased recursive partitioning

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
abstract shortened by 3 words
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3047 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 323199985b09c387192f18eeb11dd3542f1216b6
rewrote discussion of Hara & Chellappa, please check
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3048 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7fa73b6b321aa82eaaeb885da2b9c11bd7e22d07

@@
 for distributional modeling and probabilistic forecasting of circular response
-variables is rather limited. Most of the existing methods are built on the
-framework of generalized linear and additive models, which are often challenging
+variables is rather limited. Most of the existing methods are built on
+generalized linear and additive models, which are often challenging
 to optimize and interpret. Specifically, capturing abrupt changes or
@@
 full probabilistic distribution would also allow uncertainty in the forecast to
-be estimated \citep{Gneiting:2008}.  In the field of image based object
-direction estimation, \citet{Hara+Chellappa:2017} presented a
-distribution-based tree-building algorithm by applying $k$-means clustering to
-the target space where the number of clusters $k$ is adaptively determined by
-the Bayes information criterion employing the von~Mises distribution. Despite
-their distribution-based splitting approach to circular response variables,
-they solely predict the mode of the distribution, thus, providing no full
-probabilistic forecasts. Combining these previous ideas for circular trees with
-the concept of distributional modeling, we introduce a fully distributional
-tree approach that not only considers splits in all distribution parameters,
-but also yields a full probabilistic predictive model as proposed by
-\citet{Gneiting:2008}. In addition, as a natural extension, ensembles or
-forests of circular regression trees are presented in this study, showing that
-these can further improve predictive power.
+be estimated \citep{Gneiting:2008}.  Moreover, \citet{Hara+Chellappa:2017}
+introduced both regression trees and random forests for image-based object
+direction estimation. These extend classical conditional-mean regression trees
+\citep{Breiman+Friedman+Stone:1984} and corresponding random forests
+\citep{Breiman:2001} to conditional mean models for cicular data. But rather
+than considering only binary splits, they employed $k$-means clustering
+to determine multi-way splits with the number $k$ adaptivaly determined
+based on the Bayes information criterion with either a Gaussian or von~Mises
+kernel. While a distributional aspect is considered in this step of the
+algorithm, variances are always restricted to be shared across splits and, more
+importantly, no probabilistic forecasts are obtained. Therefore, our proposed
+circular regression trees and random forests take these ideas a step further:
+A fully distributional approach is adopted for all aspects of the algorithms
+from selection of the split variables, selection of the split points, and
+ensemble aggregation of the predictions across trees in a forest. Therefore,
+this yields a full probabilistic predictive model as proposed by
+\citet{Gneiting:2008}.
 

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
rewrote discussion of Hara & Chellappa, please check
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3048 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7fa73b6b321aa82eaaeb885da2b9c11bd7e22d07

@@
 
-\item We now discuss the work of Hara and Chellappa~(2017) and their distribution-based 
-tree algorithm in relation to our circular distributional trees within Sect.~1.1.
+\item We now discuss the work of Hara and Chellappa~(2017) in relation to our
+circular distributional trees within Sect.~1.1.
 
@@
 
-\section{Associate Editor - Comments to the Author}
+\section{Associate Editor -- Comments to the Author}
 
@@
 
-\section{Referee: 1 - Comments to the Author}
+\section{Referee 1 -- Comments to the Author}
 
@@
 
-\section{Referee: 2 - Comments to the Author}
+\section{Referee 2 -- Comments to the Author}
 
@@
 
-Thank you very much for the very interesting and to us unfortunately so far unknown references.
-We now review the work of Hara and Chellappa~(2017) and briefly compare their approach
-to our implementation of circular distributional trees in Sect.~1.1.
+Thank you very much for pointing out these interesting references that, unfortunately,
+we were not aware of before. We now briefly discuss the work of Hara and Chellappa~(2017),
+comparing their approach to ours in Sect.~1.1. The main difference is that their
+algorithm aims at conditional mean estimation for circular data and only adopts
+distributional aspects for determining the number of splits. Specifically, the
+variances are restricted to be shared across splits whereas in our proposal they
+are full model parameters estimated and tested along with the means.
 

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
shift post-pruning discussion to the end of the appendix and rephrase; adapted response to reviewer 2 correspondingly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3049 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a8deb148eb7ff01dca120b567ab2e4c139a72526

@@
 
-A very good remark, indeed split variable selection based on BIC creates an
-alternative possible tree-building strategy. Per default when building a
-circular distributional tree, the package \textbf{circtree} uses
-\emph{p}-values from association tests both for selecting the split variable
-and for a so-called pre-pruning strategy which stops growing the tree as soon
-as no significant associations can be found in a given subgroup. However, as
-discussed in Zeileis~\emph{et al.}~(2008), an alternative strategy is to use
-information criteria such as AIC (Akaike information criterion) or BIC (Bayes
-information criterion) for cost-complexity post-pruning, where a large tree is
-first built and then pruned back. We now discuss post-pruning in the Appendix~A
-of the manuscript and also refer to the submitted manuscript by
-Schlosser~\emph{et al.}~2019, which investigates the performance of different
-testing and pruning strategies employed in unbiased recursive partitioning
-algorithms.
+Indeed, information criteria like AIC or BIC are natural candidates for
+determining the model order in distributional models, i.e., the number of splits
+in distributional trees. While they could be used instead of the $p$-values in
+the ``internal stopping'' or ``pre-pruning'', they are typically considered
+for ``post-pruning'' after initially growing a large tree (e.g., in
+Su~\emph{et al.}~2004 or Zeileis~\emph{et al.}~2008). This is now discussed
+in Appendix~A along with the other details of the tree algorithm employed.
+There we also refer to the recent work of Schlosser~\emph{et al.}~2019
+which empirically investigates the performance of different pre- and post-pruning
+strategies in unbiased recursive partitioning. This work finds that post-pruning
+only improves the performance of the unbiased trees considered when the
+underlying tests do not work well, e.g., when the power of the tests is low
+or the sample size extremely large.
 

### 2020-07-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
shift post-pruning discussion to the end of the appendix and rephrase; adapted response to reviewer 2 correspondingly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3049 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a8deb148eb7ff01dca120b567ab2e4c139a72526

@@
 
-An alternative split variable selection strategy is provided by cost-complexity
-``post-pruning'', where a large tree is first built and then pruned back based
-on the predictive performance \citep{Breiman+Friedman+Stone:1984}. As discussed
-in \citet{Zeileis+Hothorn+Hornik:2008}, for model-based partitioning by
-employing information criteria such as AIC~(Akaike information criterion) or
-BIC~(Bayes information criterion) this strategy can be built on the same
-underlying loss function as pre-pruning. However,
-\citet{Schlosser+Hothorn+Zeileis:2019} show that post-pruning in an unbiased
-tree algorithm is usually only superior to pre-pruning when the underlying
-significance tests have low power, for example, due to a large data set.
-
 To select the best split point within the already chosen split variable, again,
@@
 Subsequently, the same testing and splitting procedure is repeated in each of
-the resulting subgroups until some stopping criterion is reached. Next to the
-already mentioned significance-level-based stopping criterion, i.e., a minimal
-$p$-value for the statistical tests, also a maximal tree depth or a minimal
-number of observations in a node can be employed as stopping criteria.
+the resulting subgroups until some stopping criterion is reached. Complementing
+the pre-pruning mentioned above, based on the significance level of the
+statistical tests, further criteria such as maximal tree depth or a minimal
+number of observations in a node can be employed.
+
+In addition to, or as an alternative to, pre-pruning, classical post-pruning
+can also be applied where a large tree is grown first and then pruned back based
+on the predictive performance \citep{Breiman+Friedman+Stone:1984}. As pointed out
+by \citet{Zeileis+Hothorn+Hornik:2008}, information criteria such as AIC~(Akaike
+information criterion) or BIC~(Bayes information criterion) are natural candidates
+for this when partitioning maximum likelihood models. However,
+\citet{Schlosser+Hothorn+Zeileis:2019} show empirically that post-pruning only improves
+upon pre-pruning in an unbiased tree algorithm when the underlying significance
+tests do not work well, e.g., when they have low power or when the number of observations
+is extremely large.
 

### 2020-07-06 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
tiny touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3050 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0b743b4a9f3baa367e37a693d37b7ff31eb8c764

@@
 
-As we have used data from the Austrian air navigation service provider, which
-we are not entitled to distribute, we are unfortunately not able to provide
+We have used data from the Austrian air navigation service provider, which
+we are not at liberty to share. Hence, we are unfortunately unable to provide
 code and data to reproduce the application. However, as you have already
 pointed out the implementation of circular regression trees and forests is
-provided in a freely available R-package with examples included in the documentation.
+provided in a freely available R package with examples included in the documentation.
 Independent of this publication, we are planning to provide an updated online

### 2020-07-06 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Minor adaptions and spell checking
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3051 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8adb6cdd8382022417f7a830ad58c451cdf6a467

@@
 variances are restricted to be shared across splits whereas in our proposal they
-are full model parameters estimated and tested along with the means.
+are full model parameters that are estimated and tested along with the means.
 
@@
 
+Su X, Wang M, Fan J (2004). ``Maximum Likelihood Regression Trees.'' 
+\emph{Journal of Computational and Graphical Statistics}, \textbf{13}\,(3), 586--598. 
+doi:10.1198/106186004X2165.
+
 Zeileis A, Hothorn T, Hornik K (2008). ``Model-Based Recursive Partitioning.''

### 2020-07-06 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor adaptions and spell checking
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3051 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8adb6cdd8382022417f7a830ad58c451cdf6a467

@@
 \citep{Breiman+Friedman+Stone:1984} and corresponding random forests
-\citep{Breiman:2001} to conditional mean models for cicular data. But rather
+\citep{Breiman:2001} to conditional mean models for circular data. But rather
 than considering only binary splits, they employed $k$-means clustering
-to determine multi-way splits with the number $k$ adaptivaly determined
+to determine multi-way splits with the number $k$ adaptively determined
 based on the Bayes information criterion with either a Gaussian or von~Mises
@@
 from selection of the split variables, selection of the split points, and
-ensemble aggregation of the predictions across trees in a forest. Therefore,
-this yields a full probabilistic predictive model as proposed by
-\citet{Gneiting:2008}.
+ensemble aggregation of the predictions across trees in a forest. This provides
+a full probabilistic predictive model, as proposed by \citet{Gneiting:2008}.
 

### 2020-07-06 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex
Include minor adaptions by Reto and Lisa
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3052 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6eed4d3a3cc88dd5c226902b9644142d414c116b

@@
 code and data to reproduce the application. However, as you have already
-pointed out the implementation of circular regression trees and forests is
+pointed out, the implementation of circular regression trees and forests is
 provided in a freely available R package with examples included in the documentation.
@@
 distributional aspects for determining the number of splits. Specifically, the
-variances are restricted to be shared across splits whereas in our proposal they
+variances are restricted to be shared across splits whereas in the method proposed they
 are full model parameters that are estimated and tested along with the means.

### 2020-07-06 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include minor adaptions by Reto and Lisa
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3052 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6eed4d3a3cc88dd5c226902b9644142d414c116b

@@
 time for the estimation of circular regression trees and forests can be reduced by
-various implemented settings, such as, e.g., the use of additional CPU cores or
-binning of the input data (c.f.  Section~\ref{sec:wind:models}).
+various implemented settings, such as, e.g., parallel estimation on multiple
+CPU cores or binning of the input data (c.f.  Section~\ref{sec:wind:models}).
 
@@
 the resulting subgroups until some stopping criterion is reached. Complementing
-the pre-pruning mentioned above, based on the significance level of the
-statistical tests, further criteria such as maximal tree depth or a minimal
+the pre-pruning based on the significance level of the
+statistical tests, as mentioned above, further criteria such as maximal tree depth or a minimal
 number of observations in a node can be employed.

### 2020-07-06 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor adaption
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3053 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 11a7ee400c14ccf7983b277af1bcadfb24b55209

@@
 Subsequently, the same testing and splitting procedure is repeated in each of
-the resulting subgroups until some stopping criterion is reached. Complementing
-the pre-pruning based on the significance level of the
-statistical tests, as mentioned above, further criteria such as maximal tree depth or a minimal
-number of observations in a node can be employed.
+the resulting subgroups until some stopping criterion is reached.
+Complementing the pre-pruning based on the significance level of the statistical tests, as described above, further criteria such as maximal tree depth or a minimal number of observations in a node can be employed.
 

### 2020-07-10 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor corrections by Torsten
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3060 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 15656e3b61cd80ff0ec5dada27411f85f054c865

@@
 obtained by evaluating the score function~$s(\cdot)$ for each individual
-observation at the parameter estimates (Equation~(\ref{eq:scores})). For the
+observation at the parameter estimates (Equation~\ref{eq:scores}). For the
 von~Mises distribution with its two distribution parameters ($\mu$ and
@@
     \cite{Fisher+Lee:1992} the model uses a link function~$g(\cdot)$ to keep the
-    response values within an interval of length $2\pi$. Since in the
+    response values within an interval of length $2\pi$. Because in the
     implementation of \cite{Mulder+Klugkist:2017} the accommodation of circular
@@
 \begin{equation}
-r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
+r^{\ast} = \argmax_{r} (\max_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr}))).
 \end{equation}

### 2020-07-10 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Make layout (incl. newpage, etc)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3061 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 06f7f3b2848eb52e139661b2e7bf5746b0cc1cb5

@@
 case in weather forecasting.
-
+% 
 To address these challenges we suggest a general distributional approach using
@@
 forests regularize and smooth the effects from an ensemble of trees.
-
+%
 The new methods are applied to probabilistic wind direction forecasting at two
@@
 
+\newpage
 \subsection{Data}
@@
 
+\newpage
 % -------------------------------------------------------------------

### 2020-12-01 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Clean up JRSSC Makefile Draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3132 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 23c9d8f885af8633eafa6144be118bba7a6b2a7c

@@
 vs.\ low precision. A more comprehensive application of circular regression
-trees and forests on probabilistic wind direction forecasting is presented in
+trees and forests to probabilistic wind direction forecasting is presented in
 Section~\ref{sec:wind} for Innsbruck Airport and Vienna International Airport,
@@
  \caption[Table caption text]{Overview of the data sets employed in the
-    application on probabilistic wind direction forecasting: For Innsbruck and
+    application to probabilistic wind direction forecasting: For Innsbruck and
     Vienna, various meteorological variables and derived quantities of these are
@@
 existing approaches, three additional statistical models are employed in the
-application on probabilistic wind direction forecasting. Two of them are based
+application to probabilistic wind direction forecasting. Two of them are based
 on existing approaches used in the meteorological field, while the third is a
@@
 The performance of the novel circular regression trees and forests is assessed
-in an application on short-term probabilistic wind direction forecasting at two
+in an application to short-term probabilistic wind direction forecasting at two
 airports with different site characteristics. As benchmark models,
@@
 shifts, and potential interactions in covariates without prespecification of
-such effects. As supported by the presented application on probabilistic wind direction
+such effects. As supported by the presented application to probabilistic wind direction
 forecasting, this can be particularly useful for modeling a highly fluctuating

### 2021-04-13 pkg/extree_partykit/inst/tinytest/test_extree_variable.R
extree_variable(variable = "yx", type = "original") returns original data not inum
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3203 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dc28139e581b6324bcc23483f572231e13fbc21a

@@ -25,13 +25,13 @@ edb <- extree_data(Species ~ Sepal.Width + Sepal.Length | Petal.Width + Petal.Le
 ev1 <- extree_variable(ed, 4, type = "original")
 expect_equal(ev1, ed$data$Petal.Width)
 
-ev2 <- extree_variable(ed, c(2, 4), type = "original") # FIXME: (SD) how to handle this if binning? --> returns unbinned vars currently
+ev2 <- extree_variable(ed, c(2, 4), type = "original")
 expect_equal(ev2, as.list(ed$data[, c("Sepal.Width", "Petal.Width")])) 
 
-ev3 <- extree_variable(ed, index = 1, type = "original") # FIXME: (SD) returns original --> should it return bins? 
+ev3 <- extree_variable(ed, index = 1, type = "original") 
 
 ## character index
-expect_null(extree_variable(ed, index = "y", type = "original")) # Question: (SD) expect error instead of NULL? 
+expect_null(extree_variable(ed, index = "y", type = "original")) # returns NULL because no variable named "y"
 ec1 <- extree_variable(ed, index = "Sepal.Length", type = "original")
 expect_equal(ec1, ed$data$Sepal.Length)
 ec2 <- extree_variable(ed, index = c("Sepal.Length", "Sepal.Width"), type = "original")
@@ -40,24 +40,24 @@ expect_equal(ec2, as.list(ed$data[, c("Sepal.Length", "Sepal.Width")]))
 ## y and x variables
 ey1 <- extree_variable(edb, variable = "y", type = "original")
 expect_equal(ey1, iris$Species)
-expect_equal(extree_variable(ed, variable = "y", type = "original"), ed$data$Species) # FIXME: (SD) currently does not return binned y 
-# --> if yx = "matrix" we can use x$yx$y for binned outcome otherwise not!! 
-# should it be equal to binned outcome?  --> attr(ed$yxindex, "levels")$Species
+expect_equal(extree_variable(ed, variable = "y", type = "original"), ed$data$Species) 
 exy <- extree_variable(ed, variable = "yx", type = "original")
-expect_equal(exy, attr(ed$yxindex, "levels")) # Question: (SD) Why not "original"
+expect_equal(data.frame(exy), ed$data[, c("Species", "Sepal.Width", "Sepal.Length")]) 
 expect_error(extree_variable(ed, variable = "x"), "'arg' should be one of")
[…]

### 2021-04-19 pkg/extree_partykit/inst/tinytest/test_extree_variable.R
Bug fix: vignette + Update extree_variable: docu, more efficient access to original and inum "y", tests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3206 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 778a19622c9371a248984e4434a3f3fa7c7e0b25

@@ -19,6 +19,9 @@ edm <- extree_data(Species + Sepal.Length ~ Sepal.Width | Petal.Width + Petal.Le
 # no binning
 edb <- extree_data(Species ~ Sepal.Width + Sepal.Length | Petal.Width + Petal.Length, 
   data = iris, yx = "matrix") # Question: (SD) why is edb$zindex$Petal.Width & Pegal.Length specified?  
+# only y - no x
+ednox <- extree_data(Species ~ Petal.Width + Petal.Length, data = iris, 
+  nmax = c("yx" = 25, "z" = 10))
 
 # Original 
 ## numeric index
@@ -46,6 +49,8 @@ expect_equal(data.frame(exy), ed$data[, c("Species", "Sepal.Width", "Sepal.Lengt
 expect_error(extree_variable(ed, variable = "x"), "'arg' should be one of")
 expect_equal(names(extree_variable(edm, variable = c("yx"))), c("Species", "Sepal.Length", "Sepal.Width"))
 expect_equal(extree_variable(ed, variable = "y"), extree_variable(ed, i = 1)) 
+# only y -  no x
+expect_equal(ey1, extree_variable(ednox, variable = "y"))
 
 # inum
 expect_equal(attr(extree_variable(ed, variable = "y", type = "inum"), "levels"), 
@@ -53,11 +58,13 @@ expect_equal(attr(extree_variable(ed, variable = "y", type = "inum"), "levels"),
 ediy <- extree_variable(ed, variable = "y", type = "inum")
 expect_equal(as.numeric(ediy), as.numeric(ed$yxindex))
 expect_inherits(ediy, "inumtotal")
-expect_equal(extree_variable(ed, variable = "yx", type = "inum"), ed$yxindex)
+expect_equal(dim(attr(ediy, "levels")), c(nrow(attr(ed$yxindex, "levels")), 1))
 extree_variable(ed, index = c(3, 4), type = "inum") # FIXME: (SD) how to handle this if binning? --> currently only z variables
 extree_variable(edb, index = c(3, 4), type = "inum") 
 expect_equivalent(class(extree_variable(edb, index = 5, type = "inum")), c("enum", "integer")) # FIXME: (SD) is this fine? 
 expect_null(extree_variable(edb, variable = "yx", type = "inum"))
+# only y no x
+expect_equal(unique(extree_variable(ednox, variable = "y", type = "inum")), c(1,2,3))
 
 # scores
[…]

### 2022-10-25 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3230 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 5f4f1afb87aba5c7be522eac1dfddfb21339cf1f

\documentclass[doc,floatsintext,natbib]{apa7}
%\usepackage[man,floatsintext]{apa6}
%\usepackage[doc,floatsintext]{apa6}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{doi}
\usepackage{booktabs}
%\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage{xcolor}
%\usepackage[colorinlistoftodos]{todonotes}
%%\todo[inline, color=green!40]{This is an inline comment.}

\linespread{1.3}

\title{Recursive partitioning of linear growth curve models}
\shorttitle{Partitioning linear trajectories with GLMM trees}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$University of Innbruck}}

\abstract{Growth curve models are popular tools for studying development over time. Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time. Recursive partitioning methods (RPMs) allow for detecting such subgroups. In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning LGCMs. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data. Here, we introduce recent adaptations of the algorithm that that are of particular relevance for partitioning longitudinal models, and growth curve models in particulat. Using simulated and existing data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart. We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for the analysis of trajectories with both discrete and continuous time, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

\usepackage{Sweave}
\begin{document}
\input{_Partitioning_GCMs_with_GLMM_trees-concordance}
\maketitle

\section{Introduction}
\label{sec:Introduction}

In longitudinal studies, (generalized) linear growth curve modeling is often used to model change over time, and inter individual differences in change. For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail. Inter individual heterogeneity is likely in such studies, where (groups of) participants show higher or lower rates of growth over time. Through the use of mixed-effects or latent-variable models, such heterogeneity can be captured, and explained by covariates of a-priori known relevance \citep{NeisyMatt18}.

Often, the covariates that predict or affect the trajectories may not be known a-priori, or researchers may have a specific interest in finding subgroups of participants with similar trajectories. Recursive partitioning methods (RPMs), also known as decision-tree methods, are particularly suited in such cases. Several existing RPMs allow for partitioning growth trajectories: For example, GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, Structural Equation Model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Partitioning based on non-linear growth curve models can be performed with the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods\footnote{Both IT methods specifically focus at detecting so-called \textit{treatment-subgroup interactions}: subgroups that respond more favorably to one treatment than another.}.

In this paper, we focus on generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}, which allow for subgroup detection in a wide range of mixed-effects models. Although originally not specifically developed for partitioning LGCMs, application of GLMM trees for partitioning LGCMs is straightforward. We propose two adaptations to the algorithm, which we expect to be of particular relevance for partitioning LGCMs.

The main difference between GLMM trees and the aforementioned RPMs, is that the latter all fit the full parametric model in each of the nodes. In contrast, GLMM trees uses the observations in each node to estimate the fixed-effects parameters, while using all observations to estimate the random-effects parameters. This may not only reduce computation time; it also reduces the total number of parameters to be estimated, which may improve stability and reduce overfitting. This local/global estimation approach was originally proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes; \cite{FokkySmit18} generalized it to model-based trees.

The longRPart, longRPart2, IT-LT and original SEM tree algorithms employ an exhaustive split detection procedure: For every possible cut point in the current node, the full parametric model is re-estimated in the two resulting nodes. To choose the optimal cutpoint, a likelihood ratio test is computed for every possible cut point. Drawbacks of this exhaustive search are a heavy computational burden, as well as a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}.

LongCART, MELT, GEE-based decision trees and score-based SEM trees fit full parametric models in each of the nodes as well, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces computational burden. Furthermore, these approaches allow for employing a two-step approach for split detection, which separates variable and cutpoint selection, thus preventing variable selection bias. The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.

The remainder of this paper is structured as follows: In the next section (Estimation) we explain how GLMM trees are estimated, and how estimation can be adjusted to (better) account for dependence structures encountered in LGCMs. Next, in Study I, we perform a simulation study to assess performance of GLMM trees and the proposed adaptations in partitioning LGCMs. In Study II, we use the same simulation design to compare the performance of GLMM trees with that of two other mixed-effects RPMs: SEM trees and LongCART.  In Study III, we compare the performance of GLMM trees and longRPart in partitioning children's trajectories of reading, math and science abilities. In the Discussion, we summarize our findings and discuss implications.

% Compared to RPMs that employ likelihood-ratio tests as a splitting criterion \citep{AbdoyLeBl02, StegyJaco18, BranyOert13, WeiyLiu20}, GLMM trees require less computation, do not present with a variable selection bias, and do not require partitioning variables to be measured at the highest level.

% Compared to GEE-based RPMs of \cite{Lee05, SuyMene11, CalhyLevi20}, GLMM trees also allow for a GEE-type approach to split selection, while explicitly modeling inter-individual variation through the estimation of random effects.

% Compared to RPMs which compute splitting criteria based on only the sign of residuals \cite{ChauyLo95, Loh02, Lee05}, the score-based tests employed by GLMM trees allow for capturing both sign and magnitude of misfit.

\section{Estimation of GLMM trees}

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
In the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors. Further, $X_i$ is the $N_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column for the observed timepoints of subject $i$. The set of observed timepoints may differ between subjects, and thus the number of rows of $X_i$ may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, their values depending on terminal node $j$ of which observation $i$ is part. Subscript $j$ is what distinguishes the GLMM tree model from a traditional GLMM. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$ and contains (a subset of) the columns of $X_i$. Further, $b_i$ is a vector of random effects of subject $i$. We assume that $b$ follows a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$. We assume Gaussian errors, with constant variance across subjects.

The parameters of a traditional GLMM can be estimated by, for example, (restricted) maximum likelihood. The GLMM tree model in Eq. \ref{eq:GLMMtree} is estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:

\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
	\item Set $r = r+1$. Estimate the partition using $Z_{i} \hat{b_i}_{(r-1)}$ as an offset.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with terminal node $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}

The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown. In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2. Note that the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset. The fixed effects are estimated \textit{locally}, using the observations in the current node. Convergence of the algorithm is monitored by computing the log-likelihood criterion of the mixed-effects model fitted in Step~3. Typically, this converges if the tree does not change from one iteration to the next.

\subsection{Initialization}

Studies to date indicate that initializing estimation with zero random effects yields an accurate final model \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initial values and found only minor impact on the final model, which became smaller with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional datasets. With longitudinal data (e.g., LGCMs), observations from the same unit tend to be more strongly correlated than in cross-sectional data; initialization with zero random effects may provide an unrealistic starting point which may be difficult to overcome in subsequent iterations. Initializing GLMM-tree estimation by estimating the random effects, assuming no grouping structure, may provide a better starting point. We thus hypothesize that for partitioning LGCMs, initializing GLMM-tree estimation with the random effects can improve subgroup recovery and predictive accuracy.

\subsection{Partitioning}

The partition (or tree) in Step~1 is estimated using model-based recursive partitioning \citep[MOB; ][]{ZeilyHoth08}. The MOB algorithm cycles iteratively through the following steps:

\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{\alph{enumi})}
	\item Fit the parametric model to all observations in the dataset. 
	\item Statistically test for parameter instability with respect to each of the partitioning variables.
	\item If there is some overall parameter instability, split the dataset with respect to the partitioning variable associated with the highest instability.
	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
\end{enumerate}

Parameter stability is tested in Step~(b), using the the \textit{scores} (gradient contributions) from the model fitted in Step~(a). Under mild regularity conditions, the scores have an expected value of 0. The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or if they exhibit systematic deviations when they are ordered by the values of one or more covariates. For continuous covariates (or ordered covariates with a large enough number of unique values), this involves computing the cumulative score process $W_k(t)$ with respect to each potential partitioning variable $U_k$ \citep{ZeilyHoth08}:

\begin{eqnarray}
\label{eq:efp}
W_{k}(t) = \hat{J}^{-1/2} n^{-1/2} \sum^{[nt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
\end{eqnarray}

where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in the current node, and $n$ gives the number of observations in the current node. Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes estimated scores (denoted $\hat{\psi}$), with the subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$. Note that $0 \leq t \leq 1$, thus $nt = 1$ for an observation with a unique minimum on the partitioning variable, and $nt = n$ for an observation with a unique maximum.

From the cumulative score process $W_k(t)$, a range of test statistics can be derived which capture non-random fluctuations. For numerical partitioning variables, a maximum Lagrange multiplier test statistic can be computed, which takes the maximum of the squared Euclidean norm of $W_k(t)$, weighted by it's variance \citep{ZeilyHorn07}. This statistic is referred to as the \textit{supLM} statistic, and is asymptotically equivalent to the maximum of likelihood-ratio statistics. Approximate asymptotic $p$-values for the \textit{supLM} statistic can be computed with the method of \cite{Hans97}. Categorical covariates do not provide an implicit ordering. Scores are therefore binned at each level of the covariate and a test statistic is computed that does not depend on the ordering of the levels \citep{MerkyFan14}.

When partitioning LGCMs, covariates will often not be measured at the lowest level, but at the subject level. The covariance matrix employed in Eq. \ref{eq:efp} can account for this clustering of observations within subjects \citep[e.g., ][]{ZeilyKoll20} by employing a clustered covariance matrix $\hat{J}$ in computation of the score process. \textcolor{orange}{@AZ: Thus, only computation of $\hat{J}$ is affected by cluster argument? Sscores are summed before computing $\hat{J}$, but computation of $\sum^{[nt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik}})$ is unaffected?} This resembles a GEE-type approach, where dependence between observations is accounted for in the variance structure. We thus hypothesize that for partitioning LGCMs, use of clustered covariances yields more accurate subgroup recovery and improves predictive accuracy.

\section{Study I: Performance of LM(M) trees}

To test whether initializing GLMM-tree estimation with the random effects and use of clustered covariances in computation of the parameter stability tests yields more accurate subgroup recovery when partitioning LGCMs, we performed a simulation study.

\subsection{Method}

\subsubsection{Data-generating design}

\begin{figure}[!ht]
\caption{Design of the subgroup-specific fixed effects in the simulation.}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-003}
\label{fig:design_tree}
\end{figure}

We simulated datasets according to the subgroup structure depicted in Figure~\ref{fig:design_tree}. Specifically, four non-overlapping subgroups corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree} were created. Observations in terminal nodes 3 and 4, and in terminal nodes 6 and 7 have the same fixed intercept ($\beta_{j0}$). To facilitate comparison with the SEM tree algorithm, the same timepoints were generated for all subjects (i.e., $t = 0, 1, 2, 3, 4$). The response was computed as:\\

$$y_{i} =  X_{i} \beta_{j} + Z_{i} b_i + \epsilon_{i},$$

\noindent where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part. The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints. The value for $b_i$ was generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next. Values in $\epsilon_{i}$ were independently generated from $N(\mu = 0, \sigma = \sqrt{5})$.

We varied the following data-generating characteristics:

\begin{itemize}
\setlength\itemsep{0em}
\setlength{\itemindent}{0.2in}
\item Number of subjects: small ($N = 100$) or large ($N = 250$).
\item Variance of the random intercept: small ($\sigma_{b_0}^2 = 1$) or large ($\sigma_{b_0}^2 = 2$).
\item Variance of the random slope: small ($\sigma_{b_1}^2 = .1$) or large ($\sigma_{b_1}^2 = .4$).
\item Number of noise partitioning variables: small ($p = 5$) or large ($p = 25$).
\item Intercorrelation between partitioning variables: absent ($\rho = 0$) or present ($\rho = .3$).
\end{itemize}

We employed a full factorial design, yielding $2^5 = 32$ cells.We performed 100 repetitions for each cell of the design. For generating and analysing data, we used R \citep[][version 4.1.2]{R22}.

\subsubsection{Partitioning methods}

We fitted two different LM trees to each dataset: One employing default settings (i.e., parameter stability tests performed at level I) and one where the parameter stability tests employed clustered covariance matrices (i.e., parameter stability tests performed at level 2). To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15][]{HothyZeil15}.

We fitted eight different LMM trees to each dataset: We estimated an LMM tree using default settings (i.e., parameter stability tests performed using observation-level covariances, estimation initialized with the tree structure); an LMM tree with parameter stability tests performed using cluster-level covariances; an LMM tree estimated by initializing estimation with the random effects; an LMM tree estimated by combining the latter two approaches. Each of the four approaches were applied with a specification comprising random intercepts only, as well as with a specification comprising both random intercepts and slopes, yielding eight different LMM trees. The different specifications allow for assessing the effect of mis-specification of the random-effects structure. To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29][]{BateyMach15} and to fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0][]{FokkySmit18}.

\subsubsection{Evaluation of performance}

We evaluated tree accuracy by counting the number of splits in each tree. Trees with $> 3$ splits are indicative of Type-I error, while trees with $> 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables). We also assessed which variable was selected for the first split in every tree.

\subsection{Results}

\begin{figure}[!ht]
\caption{Tree size distributions for LM(M) trees.}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-009}
\end{subfigure}
{\footnotesize \textit{Note.} The $y$-axis is on the log scale and represents tree size. Grey lines indicate true number of splits (3 nodes). cluster = cluster-level covariances employed in parameter stability tests; ranef init = estimation initialized with the random effects. Distances on $y$-axis are on the log scale.}
\label{fig:LMM_sizes}
\end{figure}

Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach. The default fitting approach overfits, and implemented $> 3$ splits in most datasets. This overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits 2.87), which appears most effective for LMM trees (mean number of splits 3.02 for trees estimating only random intercepts; 2.96 for trees estimating random intercepts and slopes).

Initializing estimation with the random effects also effectively reduces tree size, but too strongly when random slopes are estimated (mean number of splits without random slopes 3.15; with random slopes 1.15). Heterogeneity related to partitioning variables measured at level 2 can either be captured and explained by the tree, or captured by the random effects. By initializing estimation with the random effects effects, this heterogeneity can no longer be captured by the tree; this likely worsens with more complex random-effects specification. Combining the use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding a similar and too high number of splits, irrespective of the random-effects specification (mean number of splits 4.1).

Distributions of the number of splits, separated according to the levels of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ are depicted in Figure~\ref{fig:LMM_sizes_interact} (Appendix~\ref{sec:AppendixA}). These plots show a pattern similar to Figure~\ref{fig:LMM_sizes}, with use of cluster-level covariances providing the most accurate split recovery. In addition, the plots show that increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ yield higher numbers of splits, while the effect of $\rho$ is very minor. \textcolor{orange}{I moved plots to Appendix because of their size and because they are not of main interest. Not sure if following results should be discussed here or in the Appendix:} The main deviation from this pattern of main effects is that increased variance of the random effects (increases $\sigma_{b_0}$ and/or $\sigma_{b_1}$) yields lower numbers of splits if random slopes are estimated and estimation is initialized with the random effects.

Table~\ref{tab:first_splits} shows the variables selected for the first split. Only rarely is the wrong variable selected for the first split, and this occurs only with random-effects initialization and observation-level covariances in the parameter stability tests are employed. In the large majority of those cases, $U_2$ and $U_3$ were selected for the first split.

% latex table generated in R 4.1.3 by xtable 1.8-4 package
% Wed Oct 26 00:12:30 2022
\begin{table}[ht]
\caption{Variables selected for the first split by each LM(M) tree approach.} 
\label{tab:first_splits}
\begin{tabular}{llrrrrr}
  \hline
Random effects & Fitting approach & $U_1$ & $U_2$ & $U_3$ & $U_4$ - $U_{25}$ & no split \\ 
  \hline
$\sigma_{b_0} = \sigma_{b_1} = 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} = \sigma_{b_1} = 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
   \hline
$\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & ran.eff. init. & 0.996 & 0.002 & 0.002 & 0.001 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & both & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
   \hline
$\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & ran.eff. init. & 0.499 & 0.007 & 0.006 & 0.003 & 0.485 \\ 
  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & both & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
   \hline \multicolumn{7}{l}{\footnotesize\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are}\\  \multicolumn{7}{l}{\footnotesize continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\sigma_{b_0}$ and $\sigma_{b_1}$} are\\  \multicolumn{7}{l}{\footnotesize the standard deviations of the random intercept and slope, respectively.}\\ \end{tabular}
\end{table}
\FloatBarrier

\section{Study II: Comparison of LM(M) trees with SEM trees and LongCART}

In this second study, we compare performance of LM(M) trees with that of SEM trees and LongCART. This allows us to evaluate the possible (dis)advantages of global versus local estimation of random-effects parameters, as well as the performance of different splitting criteria. The same data-generating design as in Study I was employed. We compared performance of LM(M) trees employing clustered covariances only, which showed good performance in Study I.

\subsection{Method}

We fitted SEM trees using two different splitting criteria: First, we used the default "naive" splitting approach \citep{BranyOert13}, in which likelihood ratio tests (LRTs) are used as the splitting criterion. The likelihood of the SEM fitted to the observations in the current node is compared against the likelihood of a range of two-group SEMs, in which the two groups are defined by each possible candidate split. An LRT can thus be computed for each candidate split, which quantifies the improvement in fit that may be obtained by implementing the split. In each step, the candidate yielding the highest LRT is selected for splitting, and splitting is continued as long as a candidate split yields a $p$-value of the LRT above a pre-specified $\alpha$ level (0.05, by default).

Second, we fitted score-based SEM trees, as proposed by \cite{ArnoyVoel21}. This approach uses the MOB algorithm described in the Introduction, where the parametric model fitted in step (a) is a SEM. While for GLMM trees, parameter stability tests are computed for the fixed-effects parameters only, score-based SEM trees compute parameter stability tests based on both fixed- and random-effects parameters.

To specify the node-specific models for SEM trees, we employed an LGCM specification with the response at each timepoint regressed on a latent intercept and slope. Intercept loadings were fixed to 1; slope loadings were fixed to the value of $t$ at each timepoint. With both LRT- and score-based splitting criteria, we employed three different model specifications: a) variances of latent slope and intercept fixed to 0; b) variance of latent intercept freely estimated, variance of latent slope fixed to 0; c) variances of latent intercept and slope freely estimated. Errors were assumed uncorrelated between timepoints, error variances freely were estimated for each timepoint. To fit SEMs, we used package \textbf{lavaan} \citep[version 0.6-11][]{Ross12} and to fit SEM trees, we used package \textbf{semtree} \citep[version 0.9.17][]{BranyOert13}.

We fitted a LongCART tree to each dataset, employing default settings. The LongCART function estimates node-specific models comprising a random intercept term; this default cannot be overridden. A fixed-effects model was specified with the response regressed on time and a subject-specific random intercept. We used package \textbf{semtree} to fit SEM trees \citep[][version 0.9.17]{BranyOert13}. To fit LMMs, we used package \textbf{nlme} \citep[version 3.1-153][]{PinhyBate00} and to fit LongCART trees we used package \textbf{LongCART} \citep[version 3.1][]{Kund21}.

\subsection{Results}

\begin{figure}[!ht]
\caption{Tree size distributions for LM(M) trees with clustered covariances, SEM trees and LongCART.}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-014}
%\includegraphics{tree_size_plot.png}
\end{subfigure}
{\footnotesize \textit{Note.} Values on $x$-axis indicate whether random intercepts and/or slopes were estimated: no = random intercepts and slopes fixed to 0; int. = variance of random intercept freely estimated, variance of random slope fixed to 0; int. + slope = variance of random intercept and slope (and correlation between the two) freely estimated. Distances on $y$-axis are on the log scale. Grey lines indicate true number of splits (3 splits).}
\label{fig:tree_sizes}
\end{figure}

Figure~\ref{fig:tree_sizes} depicts tree size distributions for the different algorithms. LRT-based SEM trees appear somewhat overpowered with a mean number of splits of 3.61 ($SD = $1.71), while score-based SEM trees comprised the correct number of splits on average ($M = $3.09; $SD = $1.61). SEM trees appear more systematically affected by mis-specification of the random-effects structure than LMM trees and under-specification yields a higher number of splits. LongCART trees appear underpowered, with a mean number of splits of 2.01 ($SD = $1.45). This could be the result of under-specification of the random-effects structure: LongCART trees incorporate a random-intercept term only and this default cannot be overridden.

%We performed pairwise t-tests to assess differences between the approaches in Figure~\ref{fig:tree_sizes}. Bonferroni-corrected $p$-values indicated significant differences in tree size between most methods (all $p < .001$), with some notable exceptions: The number of splits implemented by LMM trees with clustered covariances was indistinguishable from that of LRT-based SEM trees with freely estimated random intercepts and slopes, and from the score-based SEM trees with freely estimated random intercepts.

Figure~\ref{fig:tree_sizes_interact} (Appendix \ref{sec:AppendixA}) presents tree size distributions, separated according to the levels of the data-generating parameters. Strongest effects were observed for $N$, followed by $\sigma_{b_0}$, $p_{noise}$, $\rho$ and $\sigma_{b_1}$. \textcolor{orange}{I moved plots to Appendix because of their size and because they are not of main interest. Not sure if following results should even be discussed here or in the Appendix:} Results for $N$ were as expected for all methods: with increasing sample size, more splits are implemented. Both LRT- and score-based SEM trees seem only affected by levels of $\sigma_{b_0}$, $p_{noise}$ and $\rho$ under misspecification of the random effects. Especially when both random intercept and slope variances are fixed to 0, higher levels of $\sigma_{b_0}$ and $p_{noise}$ yield more splits with both SEM tree approached. LRT-based SEM trees seem unaffected by levels of $\rho$, while score-based SEM trees seemed to implement a larger number of splits with increasing values of $\rho$. This pattern seemed reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects were correctly specified. LongCART implemented more splits with higher levels of $p_{noise}$ and $\rho$, but was unaffected by levels of $\sigma_{b_0}$ and $\sigma_{b_1}$.

% latex table generated in R 4.1.3 by xtable 1.8-4 package
% Wed Oct 26 00:12:30 2022
\begin{table}[!ht]
\caption{Variable selected for the first split by each of the partitioning methods.} 
\label{tab:first_splits2}
\begin{tabular}{llrrrrr}
  \hline
Algorithm & Random effects & $U_1$ & $U_2$ & $U_3$ & $U_4$ - $U_{25}$ & no split \\ 
  \hline
LM(M) tree & $\sigma_{b_0} = \sigma_{b_1} = 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
    (clustered & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
      covariances) & $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
   \hline
SEM tree & $\sigma_{b_0} = \sigma_{b_1} = 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
    (LRT-based) & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & 0.998 & 0.000 & 0.002 & 0.000 & 0.000 \\ 
    & $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & 0.992 & 0.001 & 0.001 & 0.002 & 0.004 \\ 
   \hline
SEM tree & $\sigma_{b_0} = \sigma_{b_1} = 0$ & 0.761 & 0.176 & 0.052 & 0.012 & 0.000 \\ 
    (score-based) & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & 0.879 & 0.077 & 0.044 & 0.000 & 0.000 \\ 
    & $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & 0.999 & 0.000 & 0.000 & 0.000 & 0.001 \\ 
   \hline
LongCART & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ &  & 0.419 & 0.297 & 0.092 & 0.192 \\ 
   \hline \multicolumn{7}{l}{\footnotesize\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables}\\  \multicolumn{7}{l}{\footnotesize are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\sigma_{b_0}$ and $\sigma_{b_1}$}\\  \multicolumn{7}{l}{\footnotesize are the standard deviations of the random intercept and slope, respectively.}\\ \end{tabular}
\end{table}

Table~\ref{tab:first_splits2} presents variable selection frequencies for the first split in the fitted trees. For SEM trees, the LRT criterion yields almost perfect accuracy for the first split. The score-based approach selected the wrong splitting variable for the first split in about 12\% of datasets, but only when the random-effects structure was mis-specified (i.e., $\sigma_{b_0}$ and/or $\sigma_{b_1}$ fixed to 0). When random effects were correctly specified, score-based SEM trees also provided perfect recovery of the first split. Closer inspection of parameter stability tests suggested that the tests are (much) more sensitive to instability in the fixed slope than in the fixed intercept.

LongCART trees exhibit strikingly low accuracy for recovering the first split, selecting the wrong variable in all datase. LongCART showed a more dramatic tendency than score-based SEM trees to use $U_2$ and $U_3$ for implementing the first split. Of note, in the LongCART trees where no split was implemented (19\% of datasets), 99\% $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not exceed the criterion of significance in those datasets. This suggests that either the parameter stability tests proposed by \cite{KundyHare19} are less sensitive to instabilities with respect to the fixed intercept, and/or with respect to categorical covariates.

\begin{figure}[!ht]
\caption{Computation time distributions for the different partitioning methods.}
\begin{subfigure}{.6\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-020}
\end{subfigure}
\label{fig:comp_times}
\end{figure}

Figure~\ref{fig:comp_times} presents computation time distributions for the partitioning algorithms. A clear computational advantage is observed for LM trees, with an average computation time of 0.141 seconds. Including random-effects estimation increased computation time: for LMM trees, average computation time was 0.65 seconds. LongCART and SEM trees yielded much longer computation times, with LongCART requiring 23.36 seconds, score-based SEM trees 4.4 seconds, and LRT-based SEM trees 2685.37 seconds computation time, on average.

\FloatBarrier

\section{Study III: Application}

\subsection{Method}

\subsubsection{Dataset}

We analyzed longitudinal assessments of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998-1999 \citep[ECLS-K; ][]{NCES10}. Data were collected from kindergarten 1998 through eighth grade 2007. In the current analyses, we focus on assessments during kindergarten, 1st, 3rd, 5th and 8th grade. In the ECLS-K study, data from 21,304 children from 1,018 schools across the USA were collected.

Response variables are reading, math and science ability. These were assessed using multi-item cognitive tests, from which latent ability estimates were computed with a mean of zero and variance of one. Reading and math abilities were assessed in all five rounds of data collection, while science knowledge was assessed in the 3rd, 5th and 8th grade only. We analyzed data from  children who completed all assessments ($N = 6,277$ for reading; $N = 6,512$ for math; $N = 6,625$ for science).

Our analyses focus on predicting ability trajectories, based on baseline characteristics. Time was measured as the number of months since the baseline assessment. In order to obtain approximately linear trajectories, used $\sqrt{\mathrm{months}}$ as the timing metric for reading and math trajectories, and $\mathrm{months}^{\frac{2}{3}}$ for science trajectories.

We used the following time-invariant covariates as potential partitioning variables:

\begin{itemize}
\setlength\itemsep{0em}
\setlength{\itemindent}{0.2in}
\item Gender (51.1\% male)
\item Age in months at baseline (range 53 to 96; M = 6.14 years at baseline).
\item Race (8 categories)
\item First time in kindergarten (yes/no)
\item Socio-economic status (range -5 to 3)
\item Fine motor skills (e.g., drawing figures; range 0 to 9)
\item Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8)
\item Interpersonal skills (range 1 to 4)
\item Self-control (range 1 to 4)
\item Internalizing problem behavior (range 1 to 4)
\item Externalizing problem behavior (range 1 to 4)
\end{itemize}

\subsubsection{Partitioning methods}

We fitted four LM(M) trees, employing the same approaches applied in Studies I and II, except for the default approaches, and the random-effects initialization approach when both random intercepts and slopes were estimated; approaches which showed far from optimal performance in Study I.

For comparison, we fitted longRPart \citep{AbdoyLeBl02} trees. We opted for longRPart, because neither SEM tree, LongCART, longRPart or longRPart2 implementations provide functionality to obtain predictions for new observations. We wrote custom functionality to allow for prediction with longRPart trees, but could not produce this for the other methods. We expect, however, that performance of longRPart will be representative of the other methods, in the sense that all fit the full mixed-effects model locally. Furthermore, SEM tree could not be employed because assessments at each wave did not take place on identical timepoints, which is a requirement for SEM-based growth curve modeling.

Using package \textbf{longRPart} \citep[version 1.0][]{StewyAbdo12}, we fitted two longRPart trees with different random effects specifications: One tree comprising a random intercept only, and one tree comprising both a random intercept and slope, with respect to student. In order to implement a split, longRPart uses a criterion based on the minimum decrease in deviance, which is set to 0.01, by default. This yielded a very large number of splits in the current analyses and poor predictive accuracy. We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = 32 terminal nodes, or 31 splits. Significance of the LRT, or a cross-validation approach to determine whether to implement a split or not may yield better results, but neither approach is implemented in longRPart (nor longRPart2).

\subsubsection{Evaluation of performance}

The ECLS-K datasets have exceptionally large sample sizes, which may not be representative of real-world applications. We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, and evaluated predictive accuracy using the responses of children not included in the training sample. For each combination of response variable and sample size, we performed 100 repetitions. Because we used disjunct samples of children for training and testing, only the fixed-effects parts of each fitted model were used for prediction; random effects were fixed to 0. Accuracy was assessed by computing the test MSE, by taking the mean squared difference between predicted and observed ability estimates of test observations. We measured tree size by counting the number of splits in each tree, and measured computation time in seconds.

\subsection{Results}

\begin{figure}[!ht]
\caption{Mean squared errors for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-021}
\end{subfigure}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-022}
\end{subfigure}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-023}
\end{subfigure}
{\footnotesize \textit{Note. }The $y$-axis on the right gives an estimate of the proportion of variance explained, computed as $1 - \frac{MSE}{var(y)}$. Distances on $y$-axis are on the log scale.}
\label{fig:application_MSEs}
\end{figure}

\begin{table}

\begin{threeparttable}
\caption{\label{tab:application_MSEs}Cross-validated mean squared errors for each of the response variables.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Math} & \multicolumn{2}{c}{Reading} & \multicolumn{2}{c}{Science} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{N = 250} & \multicolumn{1}{c}{N = 1,000} & \multicolumn{1}{c}{N = 250} & \multicolumn{1}{c}{N = 1,000} & \multicolumn{1}{c}{N = 250} & \multicolumn{1}{c}{N = 1,000} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7}
LM tree (c) & 0.1407 (0.004) & 0.1290 (0.002) & 0.1262 (0.003) & 0.1191 (0.001) & 0.4178 (0.012) & 0.3787 (0.007)\\
LMM tree (i,c) & 0.1398 (0.004) & 0.1283 (0.002) & 0.1261 (0.003) & 0.1191 (0.001) & 0.4151 (0.013) & 0.3777 (0.006)\\
LMM tree (i,s,c) & 0.1403 (0.005) & 0.1287 (0.002) & 0.1264 (0.003) & 0.1192 (0.001) & 0.4143 (0.012) & 0.3777 (0.006)\\
LMM tree (i,r) & 0.1441 (0.006) & 0.1285 (0.002) & 0.1302 (0.004) & 0.1202 (0.002) & 0.4124 (0.013) & 0.3762 (0.006)\\
longRPart (i) & 0.1463 (0.006) & 0.1318 (0.003) & 0.1335 (0.006) & 0.1213 (0.003) & 0.4197 (0.015) & 0.4050 (0.011)\\
longRPart (i,s) & 0.1430 (0.005) & 0.1306 (0.002) & 0.1291 (0.004) & 0.1199 (0.002) & 0.4289 (0.020) & 0.4296 (0.022)\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item [] \footnotesize \textit{Note.} Values represent means over 100 repetitions, parenthesized values represent SDs and can be interpreted as standard errors. c = cluster-level covariances; r = estimation initialized with random effects; i =  random-intercept variance freely estimated; s = random-slope variance freely estimated.
\end{tablenotes}
\end{threeparttable}
\end{table}

Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions for each of the partitioning approaches, separated by sample size\footnote{LongRPart trees comprising random intercepts and slopes ran into estimation errors in some datasets: In the Math ability data, 2 out of 200 datasets did not yield a tree. All tree estimations converged successfully in the Reading ability data. In the Science ability data, 40 out of 200 datasets did not yield a tree. Results are presented for iterations did yield a tree.}. Differences in performance between the partitioning approaches are of smaller magnitude than differences due to sample size. All methods show higher accuracy with larger sample sizes (with one exception: longRPart trees comprising random intercepts and slopes fitted on science ability data).

LM(M) trees tend to show better predictive accuracy than longRPart trees. Pairwise $t$-tests indicated significantly better performance for each of the LM(M) tree approaches employing clustered covariances, compared to both longRPart approaches (Bonferroni corrected $p$-values all $< .01$). However, it should be noted that differences in performance are small; for example, the standard errors presented in Table~\ref{tab:application_MSEs} indicate differences $> 1$ \textit{SE} only for Reading with $N=250$, and Science (both $N=250$ and $N=1,000$)\footnote{The significance of pairwise $t$-tests may seem in contrast with the but small differences in terms of standard errors presented in Table~\ref{tab:application_MSEs}, but the standard errors in the pairwise $t$-tests are sensitive to the number of replications, while the standard errors in Table~\ref{tab:application_MSEs} are not, and thus provide a more useful measure of the strength of the effect.}.

Table~\ref{tab:application_MSEs} indicates that LMM trees comprising a random intercept term and employing cluster-level covariances ("LMM tree (i,c)") performed best for predicting Math and Reading abilities. For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best, while this was the worst-performing approach for math and reading. For longRPart, a somewhat similar shift in performance occurs: While trees comprising both a random intercept and slope performed best for predicting Math and Reading, trees with a random intercept only performed best for predicting Science. This shift may be due to the lower number of Science ability assessments, compared to Math and Reading.

Interestingly, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with(out) random intercepts or slopes. Pairwise $t$-tests did not indicate significant differences between the three approaches for any of the response variables.

\begin{figure}[!b]
\caption{Tree sizes for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-028}
\end{subfigure}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-029}
\end{subfigure}
\begin{subfigure}{1.2\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-030}
\end{subfigure}
\label{fig:application_sizes}
\end{figure}

\begin{figure}[h]
\caption{Computation times for partitioning math, reading and science ability trajectories.}
\begin{subfigure}{.5\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-031}
\end{subfigure}
\label{fig:application_times}
\end{figure}

Figure~\ref{fig:application_sizes} present tree size distributions for each of the partitioning approaches. Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern appears reversed for longRPart trees. For math and reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small ($N=250$). This pattern reverses with large sample size ($N=1,000$), then the best-performing LM(M) tree approach yields a larger number of splits than the best-performing longRPart approach. For the math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.

Given the very minor differences in performance for LM(M) tree employing clustered covariances with different specifications of the random effects (Figure~\ref{fig:application_MSEs}), Figure~\ref{fig:application_sizes} suggests that LM trees with clustered covariances may provide very good complexity-accuracy trade-off.

Figure~\ref{fig:application_times} depicts the computation time distributions of the different partitioning algorithms. A clear computational advantage is observed for LM trees, with an average computation time of 0.47 seconds. Including estimation of random effects increases computation time, which was 5.62 seconds on average for LMM trees. LongRPart trees required longest computation times, with an average of 364.06 seconds.

\section{Discussion}

In Study I, we found the default estimation approach for GLMM trees proposed in \cite{FokkySmit18} yields high Type-I error rates when partitioning LGCMs. As hypothesized, performing parameter-stability tests using clustered covariances strongly improved performance. Also as hypothesized, we found initializing estimation with the random effects to be beneficial, but only when the random-effects specification comprised a random intercept only. Combining cluster-level covariances and random-effects initialization was not beneficial. The performance of of LM(M) trees using clustered covariances appeared largely unaffected by (mis-)specification of the random effects.

In Study II, we found good performance of SEM trees in partitioning LGCMs, but a stronger sensitivity to mis-specification of the random effects, where underspecification yielded higher Type-I errors. In accordance with results of \cite{ArnoyVoel21}, we found score-based SEM trees to have somewhat lower power than LRT-based SEM trees. LongCART tended to be underpowered, which could be due to the trees not incorporating random slopes, or the parameter stability test for categorical covariates proposed by \cite{KundyHare19} may be underpowered.

In Study III, we found higher predictive accuracy for LM(M) trees than for longRPart, but differences in performance were small. Similar to Studies I and II, we found LM(M) trees to be comparably insensitive to (mis-)specification of the random effects, which may provide an important practical advantage.

All three studies showed substantial computational advantage of GLM(M) trees. This is likely due to the estimation approach employed by GLMM trees, where fixed-effects parameters are estimated locally within a node while random-effects parameters are estimated globally, using all observations. In contrast, SEM trees, LongCART and longRPart fit the full mixed-effects model in each node, which may substantially increase computational burden.

\subsection{Possible points for further discussion}

\begin{itemize}

\item Lower performance of longRPart possibly due to lack of robust criterion for determining statistical significance of a split. Cross-validation procedures might mitigate this problem, but feasibility of such approaches currently impeded by lack of functionality for generating predictions and high computational burden.

\item MELT \citep{EoyCho14} employs random-effects predictions as a partitioning criterion, which is effective, as our results on random-effects initialization illustrate that random effects can capture effects of substantive interest that we want to use for explanation.

\item Future work:

\begin{itemize}
\item GLMM trees can be extended to allow for partitioning based on all mixed-effects model parameters simultaneously, by using the score-based tests for mixed-effects models proposed by \cite{WangyMerk18}.
\item GLMM trees can be extended to allow for partitioning non-linear growth curve models / GAMs. 
\item Performance of GLMM trees on non-gaussian responses (binomial or count responses) should be evaluated.
\end{itemize}
\end{itemize}

\bibliography{bib}

\appendix

\section{Effects of data-generating parameters on tree size}
\label{sec:AppendixA}

For LM(M) trees (Figure~\ref{fig:LMM_sizes_interact}), use of cluster-level covariances provided the most robust improvement in split recovery. For the data-generating parameters, the effects on tree size were strongest for $\sigma_{b_0}$, followed by $N$, $p$, $\sigma_{b_1}$ and $\rho$. Increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ tend to yield higher numbers of splits, and the effect of $\rho$ is minimal. The main exception to these main effects is that increased values of $\sigma_{b_0}$ and $\sigma_{b_1}$ yield lower numbers of splits when random slopes are estimated, and estimation is initialized with the random effects. In these cases, most variance may be captured by the random effects already in the first iteration of the algorithm, and will no longer be picked up by splits in the tree. Also, increased values of $p$ and $\sigma_{b_1}$ yield a slightly lower number of splits when cluster-level covariances are employed. Cluster-level covariances appear to be beneficial in accurately recovering the true subgroups, but with higher $p$, the Bonferroni correction will further reduce power to detect. Increased variance of $\sigma_{b_1}$ may reduce the power to detect the second and third split in the tree, which are driven by subgroup differences in slopes.

For SEM trees and LongCART (Figure~\ref{fig:tree_sizes_interact}), strongest effects were observed for $N$, followed by $\sigma_{b_0}$, $p_{noise}$, $\rho$ and $\sigma_{b_1}$. Results for $N$ were as expected for all methods: with increasing sample size, more splits are implemented. Both LRT- and score-based SEM trees seem only affected by levels of $\sigma_{b_0}$, $p_{noise}$ and $\rho$ under misspecification of the random effects. Especially when both random intercept and slope variances are fixed to 0, higher levels of $\sigma_{b_0}$ and $p_{noise}$ yield more splits with both SEM tree approached. LRT-based SEM trees seem unaffected by levels of $\rho$, while score-based SEM trees seemed to implement a larger number of splits with increasing values of $\rho$. This pattern seemed reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects were correctly specified. LongCART implemented more splits with higher levels of $p_{noise}$ and $\rho$, but was unaffected by levels of $\sigma_{b_0}$ and $\sigma_{b_1}$.

\begin{figure}[!ht]
\caption{Effects of data-generating parameters on tree size for LM(M) trees.}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-032}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-033}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-034}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-035}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-036}
\end{subfigure}
{\footnotesize \textit{Note.} Grey lines indicate true number of splits; distances on $y$-axis are on log scale. $\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $N$ = sample size at level 2; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
\label{fig:LMM_sizes_interact}
\end{figure}

\begin{figure}[!ht]
\caption{Effect of data-generating parameters on tree size for LM(M), SEM and LongCART trees.}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-037}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-038}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-039}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-040}
\end{subfigure}
\begin{subfigure}{1.25\textwidth}
\includegraphics{_Partitioning_GCMs_with_GLMM_trees-041}
\end{subfigure}
{\footnotesize \textit{Note.} Grey lines indicate true number of splits; distances on $y$-axis are on log scale. $N$ = sample size at level 2; $\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
\label{fig:tree_sizes_interact}
\end{figure}

\FloatBarrier

\end{document}

\section{Notes}

\subsection{Fully local mixed-effects RPMs}

The LongCART, longRPart, longPart2, non-linear longitudinal IT and SEM trees methods estimate all model parameters (fixed and random) locally, that is, separately in every (terminal) node. SEM trees provide a possible exception, by allowing the user to restrict one or more parameters to be equal across nodes. Such local estimation of the full parametric model increases the total number of estimated parameters. On the one hand, this allows the node-specific models to flexibly accommodate the observed datapoints, which may be useful for example in case of heteroscedasticity, but at the same time may increase the likelihood of overfitting.

LongRPart, non-linear longitudinal IT and SEM trees all employ likelihood-ratio tests for split selection. That is, in the current node, a likelihood-ratio is computed for every possible splitting variable and value pair. This brings a heavy computational load, as computing the likelihood ratios requires fitting the full parametric model for each of the possible splits, in each of the two resulting daughter nodes. This may also introduce variable selection bias towards variables with a larger number of possible split points \citep{ShihyTsai04, Shih04}. SEM trees \cite{BranyOert13} however allow for mitigating this bias through a two-stage procedure similar to that of \cite{LohyShih97} and \cite{KimyLoh01}.

MELT and LongCART also perform local estimation of all mixed-effects model parameters, but take a different approach to partitioning. MELT partitions the observations in the current node by fitting a basic and complex model: the basic model comprises an LMM with random intercepts and fixed slope of time (polynomial functions of time may also be included); the complex model additionally comprises random slope(s) of time. Node impurity is defined as the sum of the squared difference between the random slope from the complex model and the fixed slope from the basis model. The sign of the difference is used for splitting variable and value selection, following the approach of \cite{Loh09}. This likely reduces the computational burden compared to SEM trees, longRPart and IT, and also mitigates variable selection bias. Possible disadvantages are that MELT does not account for the precision of the random slope predictions, and does not allow for generating individual predictions, but only predicting the shape and magnitude of growth trajectories.

LongCART takes an approach related to that of \cite{ZeilyHoth08}. It fits the full mixed-effects model in the current node, and employs score-based parameter stability tests to assess the instability of the fixed-effects parameters of the mixed-effects model w.r.t. each of the partitioning variables \citep{KundyHare19}. They propose different parameter stability tests for continuous and categorical partitioning variables. The p-values of the parameter stability tests are used to determine whether splitting should be continued, and if so, which variable should be selected for splitting. In the next step, the cut-point for the variable is selected that will yield the largest decrease in AIC. This two-step approach aims to mitigate overfitting as well as a selection bias towards variables with a larger number of possible split points. In the simulation study of \cite{KundyHare19}, LongCART outperforms GLMM trees, MVRPART, and rpart- and ctree-based RE-EM Trees, in terms of predictive accuracy and tree size. \footnote{GLMM performs pretty badly, is outperformed by MVPART also. Probably, cluster argument was not employed for GLMM tree. Also, they did not evaluate variable selection bias. Also, LongCART package only supports continuous outcomes.}

Finally, the GUIDE algorithm for piecewise linear models as proposed \cite{Loh02} can also be applied for fitting piecewise (i.e., local) linear mixed-effects models. The algorithm proposed by \cite{Loh02} takes the residuals from a linear regression model fitted in the current node. $\chi^2$ tests are used to quantify the assocation between the partitioning variables and the sign of the residuals (numerical partitioning variables are converted to categorical variables by cutting the variable into groups at the sample quartiles). The partitioning variable with the lowest p-value is selected for splitting. Although there is currently no evidence available on the performance of GUIDE for partitioning LGCMs, the GUIDE approach can be employed by simply taking the residuals from a linear mixed-effects regression model, instead of a linear regression model.

\subsection{GEE-based RPMs}

Both longitudinal IT and GEE-based decision trees employ GEE instead of mixed-effects models. GEE-based decision trees \cite{Lee05} extend the generalized regression tree approach of \cite{ChauyLo95} to multivariate outcomes. A marginal GEE model is fitted in the current node and the average Pearson residual is computed for every level-2 unit. Units are grouped into two groups, one with non-negative and one with negative residuals, and two-sample t-test are performed to quantify differences between the two groups along each covariate axis. \footnote{The paper also mentions chi-square tests, but it is unclear if these are actually employed. They could be used for categorical covariates, but then p-values need to be used for selecting the splitting variable. \cite{ChauyLo95} transformed categorical covariates to ordered ones to allow for using t-tests.} The covariate selected to split the node is the one with the largest absolute t-statistic. The cut-point for the selected covariate is the weighted average of the two group means. Splitting is continued by repeating these steps in each of the resulting daughter nodes, until the p-value exceeds a pre-specified level.

Longitudinal IT \cite{SuyMene11} employs Wald tests for split selection. It fits a GEE model in the current node, and quantifies the strength of potential treatment-subgroup interactions for every possible split through a Wald statistic, which employs a sandwich estimator for the covariance matrix of the parameter estimates. The split with the highest Wald statistic is selected for splitting. After fitting a large initial tree, pruning is performed through cross validation to select a minimum value for the Wald statistic (i.e., a node's contribution to the interaction complexity of the tree).

Similar GEE approaches were developed for random forests. Both \cite{CalhyLevi20} and \cite{Mart15} found that the standard random-forest method favors splits based on covariates measured at level 2. \cite{CalhyLevi20} attribute this to the use of the Gini index, and successfully mitigate the bias by employing a robust Wald statistics for split selection instead.

\subsection{Partly global mixed-effects RPMs}

Several mixed-effects RPMs have been developed which take a partly global estimation approach \citep{HajjyBell11, HajjyLaro17, SelaySimo12, FuySimo15, SpeiyWolf18}. These methods estimate the random effects globally, using all observations, while estimating the fixed effects locally, i.e., separately in every subgroup. \cite{FokkySmit18} extended this approach to allow for GLMM-based recursive partitioning, which allows for estimating GLMs comprising one or more pre-specified predictor variables in the terminal nodes. This allows, for example, for partitioning LGCMs, while accounting for dependence of observations from the same subject through (global) estimation of random effects.

To estimate the full model, the partly global methods iterate between I) estimating the subgroup structure (recursive partition), given the current estimates of the random effects; and II) estimating the global effects, given the current subgroup structure. This procedure may require several iterations, but generally the computational load remains relatively low, as split selection does not require fitting a full parametric model for every possible split point. Note however, that if the partitioning algorithm employed in step I) exhibits variable selection bias, the full model will also exhibit this bias.

\subsection{List of decision-tree methods for longitudinal data}

\cite{HajjyBell11} proposed regression trees for clustered data.

\cite{HajjyBell14} proposed mixed-effects random forests for clustered data.

\cite{HajjyLaro17} proposed generalized regression trees for clustered data.

\cite{EoyCho14} proposed tree-structured mixed-effects regression modeling for longitudinal data.

\cite{Simo13} proposed goodness-of-fit tests for mixed-effects regression tree models that can be used to test for non-linearity of the fixed effects or heteroscedasticity of the errors.

\cite{SelaySimo12} proposed regression trees for clustered data.

\cite{FuySimo15} proposed unbiased regression trees for longitudinal and clustered data.

\cite{SpeiyWolf18} proposed a CART-based method for longitudinal and clustered data with binary outcomes. \cite{SpeiyWolf19} proposed random-forest method for longitudinal and clustered data with binary outcomes. The authors compared their tree and random forest methods also with GLMM trees, which were outperformed in the presence of strong random effects. However, this result may possibly be explained by the authors using the default settings of GLMM trees, where estimations is initialized by estimating the tree. Initializing estimation with the random effects may likely be beneficial in the presence of strong random effects.

For longitudinal data observed at very many times, \cite{YuyLamb99} treated each response vector as a random function and reduced the dimensionality of the data by fitting each trajectory with a spline curve. Then they used the estimated coefficients of the basis functions as multivariate responses to fit a regression tree model.

The method of \cite{StegyJaco18} allows for fitting non-linear mixed-effects models \cite{LindyBate90} in each of the tree nodes. The method of \cite{WeiyLiu20} allows for fitting splines \citep[i.e., natural cubic spline bases formed by B-splines, see][]{GyoryLasz06}.

\cite{Lee05} proposed generalized multivariate decision trees by using GEE.

From \cite{Loh14}: \cite{Sega92} first proposed a recursive partitioning method for longitudinal data by using as node impurity a function of the likelihood of an autoregressive or compound symmetry model. \cite{AbdoyLeBl02} used the same approach, but with a likelihood-ratio test statistic as impurity function.

\cite{Zhan98} extended \cite{Sega92} to multiple binary response variables, using as node impurity the log-likelihood of an exponential family distribution that depends only on the linear terms and the sum of second-order products of the responses. \cite{ZhanyYe08} applied the technique to ordinal responses. Their approach requires covariance matrices to be computed at every node.

\cite{WeiyLiu20} propose a tree-structured subgroup identification method IT-LT, that combines mixed-effects models with regression splines, in order to detect treatment subgroups with differential patterns of change over time. They find that in the absence of additive effects of partitioning variables, IT-LT and glmertree perform similarly. In the presence of additive effects, glmertree performs more poorly.

\cite{DeAt02} avoided the problem of covariance estimation by using as node impurity the total sum of squared deviations from the mean across the response variables. \cite{LarsySpec04} used the Mahalanobis distance, but estimated the covariance matrix from the whole data set. (And whereas \cite{DeAt02} does not refer to the work of \cite{Sega92}, \cite{LarsySpec04} do.)

"Hsiao \& Shih (2007) showed that multivariate extensions of CART are biased toward selecting  variables  that  allow  more  splits.  They  proposed  using  chi-squared  tests  of  conditional independence (conditioning on the components of the response vector) of residual signs versus grouped X values to select the split variables. The method may lack power if the effects of the X variables are not in the same direction across all the Y variables. Lee (2005) applied the GUIDE approach to multiple responses with ordered X variables by  fitting  a  generalized  estimating  equation  model  to  the  data  in  each  node  and  taking  the average  of  the  Pearson  residuals  over  the  responses  variables,  for  each  observation.  The observations  are  classified  into  two  groups  according  to  the  signs  of  the  average  residuals, and  the X with  the  smallest p -value  from  two-sample t -tests  is  chosen  to  split  the  node. Although  unbiased,  the  method  is  not  sensitive  to  all  response  trajectory  shapes.  Loh  \& Zheng (2013) solved this problem by using the residual vector patterns, rather than their averages,  to  choose  the  split  variables.  The  solution  is  applicable  to  data  observed  at  random time points."

Further, work of Ciampi needs to be mentioned?!

\cite{DeAt02} extended CART to multivariate continuous outcomes. But this was actually already proposed by \cite{GillyShel74} "a new inductive statistical technique, MAID-M, performs predictive modeling for a multivariate criterion from a set of predictor variables. Based on additive multivariate measures of association, it identifies the smallest combination of predictor variables accounting for a maximal proportion of the variation space of a given set of criterion variables."

As noted by \cite{SelaySimo12}, the methods of \cite{DeAt02} and \cite{Sega92} have a substantial drawback for the analysis of longitudinal data, in that they require the same timepoints have been observed for every subject, and do not allow for extrapolation to future timepoints, i.e., prediction on timepoints not observed in the training data.

\cite{BurgyRits15} propose tree-based varying coefficient regression. It builds on the MOB algorithm, in order to fit a varying coefficient model. This redesign involves two adjustments to MOB: 1) Inspired by the algorithms of Hajjem et al. (2011) and Sela and Simonoff (2012), our algorithm builds a closed model that consists of a tree-structured fixed-effects component and a global random effect component. By doing so, the observations of an individual are connected with the single set of corresponding random coefficients, regardless of in which nodes these observations fall, allowing for splits on time-varying covariates. 2) The coefficient constancy tests for the variable and tree size selection of MOB are adjusted, so that a separate partition is fitted for each fixed-effects coefficient.

The algorithm does not include auto-correlated errors.

\cite{BurgyRits15} distinguish between unconnected mixed effects tree methods (where separate mixed models are fitted in each node) and connected mixed-effects tree methods (where fixed effects are estimated locally and random effects globally).

\cite{BurgyRits17} do not have a mixed-effects part yet, only varying coefficients. The main difference with GLM trees is that for each coefficients of the GLM, a different model is fitted.

\cite{ChauyLo95} propose generalized regression trees, which fit a low-order polynomial in each node using maximum likelihood, and split each nodeusing a criterion based on the sign of the residuals. In principle, this allows for modeling longitudinal data, but the examples in the manuscript do not have repeated measures within subjects. Does, there is no correlated data.

\subsection{"Global" mixed-effects partitioning methods}

\cite{StegyJaco18} recently proposed an extension to the \cite{AbdoyLeBl02} to nonlinear mixed-effects models. Their method allows for partitioning based on cluster-level variables only. The implementation does currently not allow for specifying a within-group correlation structure, but defaults to assuming no within-group correlations.

\cite{StegyJaco18} note that a limitation of \cite{AbdoyLeBl02}, SEM and glmertree do not allow for observation-level covariates. They also note that the  algorithms of \cite{HajjyBell11} and \cite{SelaySimo12} extend the work of \cite{AbdoyLeBl02} to simultaneously handle observation-level and cluster-level predictors. This is incorrect, as glmertree allows for partitioning on both individual and cluster-level variables. Furthermore, glmertree, \cite{HajjyBell11} and \cite{SelaySimo12} are different from the algorithms of \cite{AbdoyLeBl02} and \cite{StegyJaco18}, in that the latter two estimate the random effects locally, within every node, whereas the former three estimate the random effects globally. I believe this global estimation of random effects is what allows for splitting on observation- and cluster level covariates. If the whole model is re-estimated locally, I am not sure whether one could or should partition based on observation-level covariates.

One advantage of RE-EMtree is that it also employs the nlme package, like \cite{AbdoyLeBl02} and \cite{StegyJaco18}. It thereby allows for specifying an autocorrelation structure within the errors (i.e., allows the error covariance matrix to be nondiagonal). glmertree employs lme4, which does not allow for specifying the correlation structure. The current implementation of \cite{StegyJaco18} employs nlme, but does not allow for specifying the correlation structure.

\cite{UsamyHaye17} performed a simulation study comparing the performance of LGCM-based SEM trees with LGCM mixtures, especially under model misspecification. They conclude that the Bonferroni method may outperform the 5-fold CV method for SEM trees, especially when the maximum number of possible partitions (subgroups?) and total sample size are small. Accurate recovery of the number of classes by SEM trees is strongly related to the agreement of the covariate with it true latent profile, and the influence of sample size is also notable (direction?). Agreement rates of .6 and .7 or up are needed for SEM trees to correctly detects classes, regardless of sample size. SEM trees might be very sensitive to model misspecification with respect to the template SEM. More separation between classes yields better class recovery for SEM trees. The impact of model misspecification on LGCMMs is smaller than that on SEM trees. This can be attributed to lower statistical power of oberved covariates in identifying classes with SEM trees.

\cite{UsamyJaco19} investigated the performance of LGC model-based SEM trees in simulated data. They assessed the performance of SEM trees to correctly identify classes using linear and quadratic LGCM. They conclude that: correct identification of the number of classes are most strongly related to the agreement rate of (observed) covariates with the true latent profile. If covariates are correlated .70 or stronger to the true latent profiles, in many cases SEM trees can uncover heterogeneity more precisely than LGCM mixtures. The larger the number of true classes, the lower the likelihood that the true number of classes is recovered by LGCM-based SEM trees. LGCM-based SEM trees were more robust against model misspecification than latent-change-score SEM trees.

In this context, the finding of \cite{MartyOert15} may be of interest: Growth mixture models outperform simpler clustering algorithms when detecting longitudinal heterogeneity, even with small sample sizes.

\subsection{Latent class growth trees}

\cite{BergySchm17} and \cite{BergyVerm18} used / proposed a method for latent class growth analysis to detect subpopulations that display different growth curves. The method recursively partitions observstions in a manner similar to divisive hierarchical clustering: classes are split until a certain criterion indicates that the fit can no longer be improved.

\subsection{Findings on random forests for multilevel data}

\cite{Mart15} proposed "an extension of the way in which variable importance measures are calculated for CART and random forests. In order to obtain more accurate estimates of variable importance, a simulated cross-validation sample for calculating variable importances is employed, rather than the out-of-bag sample. Simulation results \cite{Mart15} show that this indeed yields a more accurate ordering of variable importance than the traditional OOB approach." (from \cite{Finc15})

\cite{Mart15} Simulation phase results: "Both CART and conditional inference methods showed decreased performance in predictive accuracy and the identification of relevant variables when the ICC was moderate to large and predictors were measured at both levels of the analysis. In particular, both methods had a biased preference for level-2 variables, despite these variables having no simulated relationship with the outcome. While this is to be expected with conditional inference methods that utilize a permutation test framework built on independence assumptions, this finding is  unexpected for CART methods. If all variables are measured at the first level of analysis, however, both CART and conditional inference methods perform as expected, regardless of ICC values."

\cite{Mart15} Application phase results: "Results from three separate applications indicated  that forest methods did not massively outperform a main-effects only model in any application, but it did aid in the potential identification of small effects that deviated from linearity.

In the first application, a very small interaction effect was discovered between the SES and student minority status variables, such that non-minority students appeared to benefit more from having high SES compared to minority students, who appeared to benefit less.

The second dataset had no evidence for anything more complex than a main effect, which is not surprising given the fact that the dataset was small ($N < 200$), making it more difficult to identify more complex model specifications that were likely to generalize to a future sample.

The  third  dataset  found  a  small  non-linearity in student prior achievement predicting future achievement.While the deviation from nonlinearity was small, it contributed to the forest models outperforming the main effects only model due to the fact that the impact of prior achievement on future achievement was large. However, because prior achievement explained so much variation in future achievement, it left very little systematic variation to be explained by other measures"

\cite{KarpyHill09} "found that for multilevel data, the individual trees that make up the forest in RF are highly correlated with one another, and that this correlation increases concomitantly with increases in the ICC. In turn the inflated correlation among the trees results in an underestimate of the OOB error." (from \cite{Finc15})

\subsection{Regression trees with time-dependent covariates}

\cite{GaliyMont02} propose a CART-based approach to fitting regression trees (for continuous outcomes), where the split function (impurity measure) is adjusted so as to account for autocorrelation of the repeated observations on the same unit. The method allows for observations of the same unit to end up in different terminal nodes; the terminal nodes are piecewise-constant functions of (time-dependent) covariate.

\cite{PillayCalo03} combined this approach with radial basis function (RBF) networks, in order to smoothen the predictive model. In this approach, the location parameters of the RBF network are determined by fitting regression tree.

\cite{GaliyPill07} proposed alternative split criteria, based on loss functions minimized by M-estimators \cite{Hube64}. These split criteria are more robust, by downweighing outliers when calculating the measure of within-node impurity. (But this is not really for repeated measures data.)

\cite{NeisyMatt98} describe the use of mixed-effects models and latent-curve models to explore growth over time, and show that "one framework is often more advantageous to adopt". In a table, they provide an overview of when to prefer one approach over another. They summarize: "the ME approach tends to be most useful for straightforward models (e.g., a simple growth trajectory with one outcome variable), with complex data structures such as smaller samples, time-unstructured data, or multiple levels of nesting requiring more flexibility. Conversely, the LC approach is best suited to complex models with straightforward data structures, such as growth models embedded in larger models, assessments of global model fit, unconstrained time-varying covariates, and complex variance functions."

\subsection{Performance of traditional GEEs}

\cite{HubbyAher10}: "The one caveat to an otherwise straightforward approach is that the number of neighborhoods has to be sufficiently large; the inference is asymptotically correct but not necessarily accurate in smaller samples."

\cite{HubbyAher10} have pointed out that that mixed models involve unverifiable assumptions on the data-generating distribution, which lead to potentially misleading estimates and biased inference. They therefore favor a GEE-type approach, as it involves little assumptions, and just requires the number of level-2 observations to be sufficiently large for robust estimation of standard errors. They also point out the fact that GEE- or mixed-effects parameter estimates in the linear case are equivalent, but not in the binary case.

\subsection{Alternative application datasets}

\cite{Lee05} analyzed epileptic seizure data that is available in as the \verb|epil| dataset in package MASS. The data consist of the number of epileptic seizures in an eight-week baseline period before any treatment, and in each of four two-week treatment periods in which patients received either a placebo or the drug Progabide in addition to other therapy.

\cite{Lee05} analyzed RCT data to compare a test treatment and placebo for a respiratory disorder. Patients in each of two centers were randomly assigned to groups receiving the active treatment or a placebo. During treatment, respiratory status was determined at four visits. The data is available as the \verb|resp| dataset in package sanon.

\cite{BurgyRits15} analyzed data derived from the British Household Panel Study, to show how the effect of unemployment on self-reported happiness varies across individual life circumstances. The data is included as supplementary material to their paper?

\cite{FuySimo15} analyzed "the wages data, obtained from the UCLA Academic Technology Service website. It contains data on 888 individuals' hourly log wage (response variable) information and corresponding covariate values. It was previously studied by Singer and Willett (2003) and Eo and Cho (2014). The number of observations of each individual range from 1 to 13, so the data are highly unbalanced, with a total of 6402 observations. Eo and Cho (2014) were limited to time-invariant covariates in their applications of GUIDE and MELT to these data, using race (White, Black and Hispanic) and hgc (highest degree completed by each individual). The CART-based REEMtree is also given in Eo and Cho (2014), and it only splits once, on the hgc variable. Fig. 16 ... gives the unbiased REEM tree, which has a more complex structure, broadly similar to that of GUIDE (as presented in Eo and Cho, 2014)."

\subsection{Auto correlation structures}

https://bbolker.github.io/mixedmodels-misc/notes/corr\_braindump.html

%Table~\ref{tab:ECLSK_accuracy} shows the cross-validated predictive accuracies for all RPMs. Prediction errors were largest for the science data, which also had the largest sample variance (...), while prediction errors were lowest for the reading data, which also had the smallest sample variance (...). The sample variance of the math score was ....

%Several fitting attempts for longRPart trees resulted in estimation errors: For the trees with random intercepts estimated only, ... out of 30 tree-fitting attempts yielded an error, while for the trees with both random intercepts and slopes estimated, ... out of 30 tree-fitting attempts yielded an error. Results for longRPart are computed based on successfully fitted trees.

%The MSEs differ significantly only between the LM(M) trees and longRPart trees. LM(M) trees provide somewhat lower prediction errors, but the tree sizes in Table~\ref{tab:ECLSK_complexity} indicate that this comes at the cost of larger tree size.

%On average, LMM trees were more accurate but also larger than longRPart trees. In terms of average predictive accuracy, the tree which estimates random intercepts and initialized estimation with the random effects performed best for the LMM trees. For the longRPart trees, the model with both random intercepts and slopes performed best.

%However, the settings that perform best for predicting reading and math scores, did not perform best for predicting science scores. Cluster sizes (i.e., number of observations per student) are lower for the science scores (3) than for the reading and math scores (5), which may partly explain the difference. For LM(M) trees, the trees which estimate both random intercepts and slopes while performing parameter stability tests on the cluster level, perform best in the reading and math data. However, for the science data, these settings perform among the worst, while highest accuracy was obtained by the LMM tree in which only a random intercept was estimated, and estimation was initialized with the random effects. Note that both these settings were found to perform best in the earlier simulation study.

%For longPart, highest predictive accuracy was also obtained with trees in which both random intercepts and slopes were estimated for the reading and math data. For the science data, trees in which only random intercepts were estimated provided highest predictive accuracy. In all three datasets, the random-effects specification that yielded the largest longRPart trees, also yielded the highest predictive accuracy.

\begin{figure}
\caption{Computation time distributions for the different partitioning methods.}
\label{fig:comp_timeECLSK}
\end{figure}

%In terms of computation time, Figure~\ref{fig:comp_timeECLSK} indicates a clear computational advantage for LM(M) trees. The outlying, lower computation times for longRPart are due to estimation errors early in the fitting process, after which computations are halted completely.

### 2022-10-25 pkg/glmertree/R/glmertree.R

@@ -115,12 +115,23 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
     if (iteration > 2L) data$.tree <- NULL
     
     ## fit tree
-    if (is.null(q_cluster)) {
-      tree <- lmtree(tf, data = data, offset = .ranef, weights = .weights, 
-                     dfsplit = dfsplit, ...)
+    if (is.null(cl$vcov)) { ## Allow for use of ML estimation:
+      if (is.null(q_cluster)) {
+        tree <- lmtree(tf, data = data, offset = .ranef, weights = .weights, 
+                       dfsplit = dfsplit, ...)
+      } else {
+        tree <- lmtree(tf, data = data, offset = .ranef, weights = .weights, 
+                       cluster = .cluster, dfsplit = dfsplit, ...)
+      }
     } else {
-      tree <- lmtree(tf, data = data, offset = .ranef, weights = .weights, 
-                     cluster = .cluster, dfsplit = dfsplit, ...)
+      if (is.null(q_cluster)) {
+        tree <- glmtree(tf, data = data, offset = .ranef, weights = .weights, 
+                        dfsplit = dfsplit, family = "gaussian", ...)
+      } else {
+        tree <- glmtree(tf, data = data, offset = .ranef, weights = .weights, 
+                        cluster = .cluster, dfsplit = dfsplit, family = "gaussian", 
+                        ...)
+      }
     }
     
     if (plot) plot(tree, main = paste("Iteration", iteration))
@@ -157,8 +168,13 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
                       REML = REML, offset = offset, control = lmer.control)          
         }
       }
-      b <- structure(lme@beta, .Names = names(fixef(lme)))
-      b[substr(names(b), 1L, 5L) %in% c("(Inte", ".tree")] <- 0
+      if (length(tree) == 1L) { 
+        b <- structure(lme@beta, .Names = names(fixef(lme)))
+        b[names(b) %in% names(tree[[1]]$node$info$object$coefficients)] <- 0        
+      } else {
+        b <- structure(lme@beta, .Names = names(fixef(lme)))
+        b[substr(names(b), 1L, 5L) %in% c("(Inte", ".tree")] <- 0
+      }
       data$.ranef <- suppressWarnings(suppressMessages(
         predict(lme, newdata = data, newparams = list(beta = b))))
     } else {
[…]

### 2022-10-25 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3230 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 5f4f1afb87aba5c7be522eac1dfddfb21339cf1f

--- 2022-10-25 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.tex
@@
 
-\usepackage{Sweave}
 \begin{document}
-\input{_Partitioning_GCMs_with_GLMM_trees-concordance}
+\SweaveOpts{concordance=TRUE}
 \maketitle
@@
 \label{sec:Introduction}
+
+<<eval=FALSE,echo=FALSE>>=
+## If .Rnw will not compile bibliography, run tinytex on .tex file:
+tinytex::latexmk("Partitioning_GCMs_with_GLMM_trees.tex")
+@
 
@@
 
+<<echo=FALSE>>=
+library("partykit")
+set.seed(42)
+U1 <- sample(0:1, size = 1250, replace = TRUE) 
+U2 <- round(rnorm(1250))
+U3 <- round(rnorm(1250))
+time <- rep(0:4, each = 250)
+## left-most node: x1 == 0 & x2 <= 0
+## intercept = -1, slope = -1
+node3 <- as.numeric(U1 == 0 & U2 <= 0)
+## left-middle node: x1 == 0 & x2 > 0
+## intercept = -1, slope = 0
+node4 <- as.numeric(U1 == 0 & U2 > 0)
+## right-middle node: x1 == 1 & x3 =< 0
+## intercept = 1, slope = 0
+node6 <- as.numeric(U1 == 1 & U3 <= 0)
+## right-most node: x1 == 1 & x3 > 0
+## intercept = 1, slope = 1
+node7 <- as.numeric(U1 == 1 & U3 > 0)
+y <- (node3 * -1 - node3 * time) + (-node4) +
+ (node6) + (node7 * 1 + node7 * time) 
+data <- data.frame(U1 = as.factor(U1), U2, U3, time, y)   
+tree <- lmtree(y ~ time | U1 + U2 + U3, data = data, maxdepth = 3)
+
+fig4 <- party(
+  partynode(1L,
+            split = partysplit(1L, breaks = 1),
+            kids = list(
+              partynode(2L,
+                        split = partysplit(2L, breaks = 0),
+                        kids = list(partynode(3L, info = c(
+                            expression(beta[j0] == '-1.0'),
+                            expression(''),
+                            expression(beta[j1] == '-1.0'))),
+                          partynode(4L, info = c(
+                            expression(beta[j0] == '-1.0'),
+                            expression(''),
+                            expression(beta[j1] == '0.0'))))),
+              partynode(5L,
+                        split = partysplit(3L, breaks = 0),
+                        kids = list(
+                          partynode(6L, info = c(
+                            expression(beta[j0] == '1.0'),
+                            expression(''),
+                            expression(beta[j1] == '0.0'))),
+                          partynode(7L, info = c(
+                            expression(beta[j0] == '1.0'),
+                            expression(''),
+                            expression(beta[j1] == '1.0'))))))),
+  data.frame(data)
+)
+#plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
+
+## Panel-combining function:
+combine_panels <- function(panel1, panel2, party2 = NULL) {
+  function(node) {
+    nid <- id_node(node)
+    pushViewport(viewport(
+      layout = grid.layout(nrow = 2, ncol = 1, heights = c(1, 1)),
+      name = paste("node_mob_mypanel", nid, sep = "")))
+    grid.rect(gp = gpar(fill = "white", col = 0))
+    pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+    panel1(node)
+    popViewport()
+    pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+    node2 <- if(is.null(party2)) node else node_party(party2[nid])
+    panel2(node2)
+    popViewport(2)
+  }
+}
+@
+
 \begin{figure}[!ht]
 \caption{Design of the subgroup-specific fixed effects in the simulation.}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-003}
+<<fig=T, height=5, width=5, echo=F>>=
+#plot(tree, ip_args = list(pval = FALSE), gp = gpar(cex = .7),
+#     tp_args = list(pointcol = "white", xlab = "time",
+#                    mainlab = function(id, nobs) paste("node", id)))
+plot(tree, tnex = 3.5, terminal_panel = combine_panels(
+    panel1 = node_bivplot(tree, xlab = "time", pointcol = "white",
+                        mainlab = function(id, nobs) paste("node", id)),
+    panel2 = node_terminal(fig4, FUN = identity, height = 4, width = 12, id = FALSE),
+    party2 = fig4), 
+  gp = gpar(cex = .7), ip_args = list(pval = FALSE))
+@
 \label{fig:design_tree}
@@
 
+<<echo=FALSE>>=
+load("tree_sizes.Rda")
+tree_size <- data.frame(sizes, true = 7)
+N <- c(100, 250)
+sigma_int <- c(1, 2)
+sigma_slope <- c(sqrt(.1), sqrt(.4))
+p_noise <- c(5, 25)
+rho <- c(0, .3)
+design_matrix <- expand.grid(N = N, sigma_int = sigma_int,
+                             sigma_slope = sigma_slope,
+                             p_noise = p_noise, rho = rho)
+design_matrix[] <- lapply(design_matrix, factor)
+tree_sizes_long <- data.frame(stack(tree_size),
+                              dataset_id = factor(rep(1:nrow(tree_size), 
+                                                      times = ncol(tree_size))))
+names(tree_sizes_long)[1:2] <- c("tree_size", "method_id") 
+tree_sizes_long <- cbind(tree_sizes_long, design_matrix)
+tree_sizes_long$method_id <- relevel(tree_sizes_long$method_id, ref = "true")
+LMM_ids <- which(tree_sizes_long$method_id %in% c("true", "lm", "lm_c", "lmer", "lmer_c", 
+                                                  "lmer_cr", "lmer_r", "lmer_s",
+                                                  "lmer_s_c", "lmer_s_cr", "lmer_s_r"))
+stdCoef <- function(object) {
+  sdy <- sd(getME(object,"y"))
+  sdx <- apply(getME(object,"X"), 2, sd)
+  sc <- fixef(object)*sdx/sdy
+  se.fixef <- coef(summary(object))[,"Std. Error"]
+  se <- se.fixef*sdx/sdy
+  return(data.frame(stdcoef=sc, stdse=se))
+}
+@
+
+<<echo=FALSE, eval=FALSE>>=
+library("lmerTest")
+## LM(M) trees only
+mod_LMM <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
+  N + sigma_int + sigma_slope + p_noise + rho), 
+  data = tree_sizes_long[LMM_ids, ])
+#summary(mod_LMM)
+#fixef(mod_LMM)[order(abs(fixef(mod_LMM)))]
+coefs <- stdCoef(mod_LMM)[,1, drop = FALSE]^2
+
+## "Variance explained" by N, sigma_n0, sigma_b1, p_noise and rho:
+sum(coefs[grep("sigma_int", rownames(coefs)), ])
+sum(coefs[grep("N", rownames(coefs)), ])
+sum(coefs[grep("p_noise", rownames(coefs)), ])
+sum(coefs[grep("sigma_slope", rownames(coefs)), ])
+sum(coefs[grep("rho", rownames(coefs)), ])
+
+## Note, above is not really variance explained, but some measure 
+## of influence that does not add up to 1
+
+## Variance explained by main effects:
+coefs["N250", ] 
+coefs["sigma_int2", ]
+coefs["p_noise", ]
+coefs["sigma_slope0.632455532033676", ]
+coefs["rho0.3", ]
+
+round(coefs[order(coefs$stdcoef, decreasing = TRUE), , drop = FALSE], digits = 3)
+
+## Variance explained by random effects:
+varco <- as.data.frame(VarCorr(mod_LMM))$vcov /
+  var(tree_sizes_long[LMM_ids, "tree_size"])
+varco
+@
+
+<<echo=FALSE>>=
+library("lmerTest")
+## Compare means between settings
+tmp <-  tree_sizes_long[LMM_ids, ]
+tmp$tree_size <- (tmp$tree_size-1)/2
+tmp$default_ids <- tmp$method_id %in% c("lm", "lmer", "lmer_s")
+tmp$cluster_ids <- tmp$method_id %in% c("lm_c", "lmer_c", "lmer_s_c")
+tmp$ranefstart_ids <- tmp$method_id %in% c("lmer_r", "lmer_s_r")
+tmp$both_ids <- tmp$method_id %in% c("lmer_cr", "lmer_s_cr")
+
+M_default <- mean(tmp$tree_size[tmp$default_ids])
+M_cluster <- mean(tmp$tree_size[tmp$cluster_ids])
+M_ranefstart <- mean(tmp$tree_size[tmp$ranefstart_ids])
+M_both <- mean(tmp$tree_size[tmp$both_ids])
+@
+
+<<eval=FALSE, echo=FALSE>>=
+## t-tests for comparing no of splits with true number of splits
+round(apply((sizes-1)/2, 2, function(x) t.test(x, mu = 3)$p.value), digits = 3)
+
+## pairwise difference tests
+pairwise.t.test(tmp$tree_size, tmp$method_id, 
+               alternative = "two.sided", p.adjust = "bonferroni")
+##          lmer    lmer_c  lmer_r  lmer_cr lmer_s  lmer_s_c lmer_s_r lmer_s_cr lm
+## lmer_c    < 2e-16 -       -       -       -       -        -        -         -      
+## lmer_r    < 2e-16 1.0000  -       -       -       -        -        -         -      
+## lmer_cr   < 2e-16 < 2e-16 < 2e-16 -       -       -        -        -         -      
+## lmer_s    0.0773  < 2e-16 < 2e-16 < 2e-16 -       -        -        -         -      
+## lmer_s_c  < 2e-16 1.0000  0.0472  < 2e-16 < 2e-16 -        -        -         -      
+## lmer_s_r  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16  -        -         -      
+## lmer_s_cr < 2e-16 < 2e-16 < 2e-16 0.0027  < 2e-16 < 2e-16  < 2e-16  -         -      
+## lm        1.0000  < 2e-16 < 2e-16 < 2e-16 1.0000  < 2e-16  < 2e-16  < 2e-16   -      
+## lm_c      < 2e-16 2.4e-10 < 2e-16 < 2e-16 < 2e-16 4.6e-07  < 2e-16  < 2e-16   < 2e-16
+@
+
+<<echo=FALSE>>=
+means <- tapply(tmp$tree_size, tmp$method_id, mean)
+@
+
 \begin{figure}[!ht]
@@
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-009}
+<<echo=FALSE, fig=TRUE, height=3.5>>=
+## Prepare data
+tmp <-  tree_sizes_long
+tmp$tree_size <- (tmp$tree_size-1)/2
+tmp <- tmp[-which(tmp$method_id == "true") , ]
+tmp$method_id <- factor(tmp$method_id)
+
+## Create indicator for algorithm and random effects specification
+tmp$ranef <- NA
+tmp$ranef[tmp$method_id %in% c("lm", "lm_c")] <- "LM tree \n(no random effects)"
+tmp$ranef[tmp$method_id %in% c("lmer", "lmer_c", "lmer_r", "lmer_cr")] <- "LMM tree \n(random intercept)"
+tmp$ranef[tmp$method_id %in% c("lmer_s", "lmer_s_c", "lmer_s_r", "lmer_s_cr")] <- "LMM tree \n(random intercept + slope)"
+tmp$ranef[tmp$method_id %in% c("LongCART")] <- "LongCART"
+tmp$ranef[tmp$method_id %in% c("s_sem", "f_sem")] <- "SEM trees \n(no random effects)"
+tmp$ranef[tmp$method_id %in% c("s_sem_i", "f_sem_i")] <- "SEM tree \n(random intercept)"
+tmp$ranef[tmp$method_id %in% c("s_sem_s", "f_sem_s")] <- "SEM tree \n(random intercept + slope)"
+tmp$ranef <- factor(tmp$ranef, 
+                    levels = c("LM tree \n(no random effects)", "LMM tree \n(random intercept)", 
+                               "LMM tree \n(random intercept + slope)", 
+                               "SEM tree \n(no random effects)", "SEM tree \n(random intercept)", 
+                               "SEM tree \n(random intercept + slope)"))
+
+## Create indicator for additional settings
+tmp$setting <- NA
+tmp$setting[tmp$method_id %in% c("lmer", "lmer_s", "lm")] <- "default"
+tmp$setting[tmp$method_id %in% c("lmer_c", "lmer_s_c", "lm_c")] <- "cluster"
+tmp$setting[tmp$method_id %in% c("lmer_r", "lmer_s_r")] <- "ranef\ninit"
+tmp$setting[tmp$method_id %in% c("lmer_cr", "lmer_s_cr")] <- "cluster +\nranef"
+tmp$setting[tmp$method_id %in% c("LongCART")] <- "default"
+tmp$setting[tmp$method_id %in% c("s_sem", "s_sem_i", "s_sem_s")] <- "score"
+tmp$setting[tmp$method_id %in% c("f_sem", "f_sem_i", "f_sem_s")] <- "fair"
+tmp$setting <- factor(tmp$setting, levels = c("cluster +\nranef", "cluster", 
+                                              "default", "fair",
+                                              "ranef\ninit", "score"))
+levels(tmp$sigma_int) <- c("1", "4")
+levels(tmp$sigma_slope) <- c("0.1", "0.4")
+tmp$setting <- factor(tmp$setting, levels = c("default", "cluster",  "ranef\ninit", 
+                       "cluster +\nranef", "fair", "score"))
+
+## Create the plot
+library("ggplot2")
+library("gridExtra")
+library("colorspace")
+library("scales")
+
+## LM(M)
+LMM_ids <- which(tmp$method_id %in% c("lm", "lm_c", "lmer", "lmer_c", "lmer_cr", 
+                                      "lmer_r", "lmer_s", "lmer_s_c", 
+                                      "lmer_s_cr", "lmer_s_r"))
+breaks <- c(0, 1, 3, 7, 20)
+lim <- c(0, 25)
+plot0 <- ggplot(tmp[LMM_ids, ]) +
+  geom_boxplot(aes(x=setting, y=tree_size), 
+               position=position_dodge(1), show.legend = FALSE, 
+               alpha = 0.5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(sigma = 1, base = exp(1)), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits") + theme(legend.position = "none") +
+  geom_hline(yintercept=3, col = "darkgrey")
+plot0
+@
 \end{subfigure}
@@
 The default fitting approach overfits, and implemented $> 3$ splits in most datasets.
-This overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits 2.87), which appears most effective for LMM trees (mean number of splits 3.02 for trees estimating only random intercepts; 2.96 for trees estimating random intercepts and slopes).
-
-Initializing estimation with the random effects also effectively reduces tree size, but too strongly when random slopes are estimated (mean number of splits without random slopes 3.15; with random slopes 1.15).
+This overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits \Sexpr{round(M_cluster, digits = 2)}), which appears most effective for LMM trees (mean number of splits \Sexpr{round(means["lmer_c"], digits = 2)} for trees estimating only random intercepts; \Sexpr{round(means["lmer_s_c"], digits = 2)} for trees estimating random intercepts and slopes).
+
+Initializing estimation with the random effects also effectively reduces tree size, but too strongly when random slopes are estimated (mean number of splits without random slopes \Sexpr{round(means["lmer_r"], digits = 2)}; with random slopes \Sexpr{round(means["lmer_s_r"], digits = 2)}).
 Heterogeneity related to partitioning variables measured at level 2 can either be captured and explained by the tree, or captured by the random effects.
 By initializing estimation with the random effects effects, this heterogeneity can no longer be captured by the tree; this likely worsens with more complex random-effects specification.
-Combining the use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding a similar and too high number of splits, irrespective of the random-effects specification (mean number of splits 4.1).
+Combining the use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding a similar and too high number of splits, irrespective of the random-effects specification (mean number of splits \Sexpr{round(M_both, digits = 2)}).
 
@@
 
-% latex table generated in R 4.1.3 by xtable 1.8-4 package
-% Wed Oct 26 00:12:30 2022
-\begin{table}[ht]
-\caption{Variables selected for the first split by each LM(M) tree approach.} 
-\label{tab:first_splits}
-\begin{tabular}{llrrrrr}
-  \hline
-Random effects & Fitting approach & $U_1$ & $U_2$ & $U_3$ & $U_4$ - $U_{25}$ & no split \\ 
-  \hline
-$\sigma_{b_0} = \sigma_{b_1} = 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-  $\sigma_{b_0} = \sigma_{b_1} = 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-   \hline
-$\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & ran.eff. init. & 0.996 & 0.002 & 0.002 & 0.001 & 0.000 \\ 
-  $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & both & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-   \hline
-$\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ &  & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & clustered cov. & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & ran.eff. init. & 0.499 & 0.007 & 0.006 & 0.003 & 0.485 \\ 
-  $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & both & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-   \hline \multicolumn{7}{l}{\footnotesize\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are}\\  \multicolumn{7}{l}{\footnotesize continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3).
-$\sigma_{b_0}$ and $\sigma_{b_1}$} are\\  \multicolumn{7}{l}{\footnotesize the standard deviations of the random intercept and slope, respectively.}\\ \end{tabular}
-\end{table}
+<<echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
+library("xtable")
+load(file = "splitvars.Rda")
+splitvar <- apply(splits, 2, unlist)
+rownames(splitvar) <- NULL
+
+## LongCART registers best splitting variable if no split was implemented, overwrite
+#table(splitvar[tree_size$Lcrt == 1, "Lcrt"])
+# x1 x14 x16  x2 x28  x5  x9 
+#607   1   1   1   2   1   1 
+#
+## Note: variable selection bias may not be a problem; power seems to be a problem
+Lcrt_splitvars <- splitvar[, "Lcrt"] ## save for later use
+splitvar[tree_size$Lcrt == 1, "Lcrt"] <- NA
+
+## Get splitting proportions
+func <- function(x) table(x, useNA = "ifany") / sum(table(x, useNA = "ifany"))
+probs <- apply(splitvar, 2, func)
+probs <- sapply(probs, function(x) {
+  names(x)[is.na(names(x))] <- "no split"
+  return(x)
+})
+
+## Have column for u1, u2, u3, u4-up, no split
+splits <- matrix(sapply(probs, function(x) x["x1"]), ncol = 1,
+                 dimnames = list(names(probs), "x1"))
+for (u_set in list("x2","x3", paste0("x", 4:30), "no split")) {
+  splits <- cbind(splits, sapply(probs, function(x) sum(x[names(x) %in% u_set])))
+}
+splits <- data.frame(splits[c(9:10, 1:8, 11:17), ]) ## reorder rows so LM tree is 1st
+rows <- cbind(c("LM tree", " ", 
+                "LMM tree", " ", " ", " ", " ", " ", " ", " ", 
+                "SEM tree", " ", " ", " ", " ", " ", 
+                "LongCART"),
+                c("$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
+                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
+                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
+                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$"),
+             c("", "clustered cov.", "", "clustered cov.", 
+               "ran.eff. init.", "both", "", "clustered cov.", 
+               "ran.eff. init.", "both",
+               "LRT", "LRT", "LRT", "score-based", "score-based", "score-based", ""))
+splits <- cbind(rows, splits)
+colnames(splits) <- c("Algorithm", "Random effects", "Fitting approach", "$U_1$", 
+                      "$U_2$", "$U_3$", "$U_4$ - $U_{25}$", "no split")
+
+print(xtable(splits[1:10, -1], digits = 3,
+             caption = "Variables selected for the first split by each LM(M) tree approach.", 
+             label = "tab:first_splits"), 
+      include.rownames=FALSE, hline.after = c(-1, 0, 2, 6), #table.placement = "tb",
+      add.to.row = list(pos = list(10, 10, 10, 10), command = c(
+        "\\hline",
+        "\\multicolumn{7}{l}{\\footnotesize\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are}\\\\ ", 
+        "\\multicolumn{7}{l}{\\footnotesize continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\\sigma_{b_0}$ and $\\sigma_{b_1}$} are\\\\ ",
+        "\\multicolumn{7}{l}{\\footnotesize the standard deviations of the random intercept and slope, respectively.}\\\\ ")),
+      latex.environment = "", caption.placement = "top",
+      sanitize.text.function=function(x){x})
+@
+
 \FloatBarrier
@@
 
+<<echo=FALSE>>=
+tmp2 <- tmp
+LMM_c_ids <- which(tmp2$method_id %in% c("lm_c", "lmer_c", "lmer_s_c"))
+SEM_ids <- which(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s",
+                                      "s_sem", "s_sem_i", "s_sem_s"))
+Lcrt_ids <- which(tmp2$method_id %in% "Lcrt")
+tmp2 <-  tmp2[c(LMM_c_ids, SEM_ids, Lcrt_ids), ]
+tmp2$method_id <- factor(tmp2$method_id)
+@
+
+<<echo=FALSE, eval=FALSE>>=
+## SEM trees
+SEM_ids <- which(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s",
+                                      "s_sem", "s_sem_i", "s_sem_s"))
+mod_SEM <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
+  N + sigma_int + sigma_slope + p_noise + rho), 
+  data = tmp2[SEM_ids, ])
+summary(mod_SEM)
+coefs <- stdCoef(mod_SEM)[,1, drop = FALSE]^2
+r2_N <- sum(coefs[grep("N", rownames(coefs)), ])
+r2_si <- sum(coefs[grep("sigma_int", rownames(coefs)), ])
+r2_ss <- sum(coefs[grep("sigma_slope", rownames(coefs)), ])
+r2_p <- sum(coefs[grep("p_noise", rownames(coefs)), ])
+r2_rho <- sum(coefs[grep("rho", rownames(coefs)), ])
+
+## LongCART
+Lcrt_ids <- which(tmp2$method_id %in% "Lcrt")
+mod <- lm(tree_size ~ N + sigma_int + sigma_slope + p_noise + rho, 
+  data = tmp2[Lcrt_ids, ])
+summary(mod)
+coefs <- stdCoef(mod_LMM)[,1, drop = FALSE]^2
+r2_N <- sum(coefs[grep("N", rownames(coefs)), ])
+r2_si <- sum(coefs[grep("sigma_int", rownames(coefs)), ])
+r2_ss <- sum(coefs[grep("sigma_slope", rownames(coefs)), ])
+r2_p <- sum(coefs[grep("p_noise", rownames(coefs)), ])
+r2_rho <- sum(coefs[grep("rho", rownames(coefs)), ])
+@
+
+<<eval=FALSE, echo=FALSE>>=
+## LM(M) trees only
+mod_all <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
+  N + sigma_int + sigma_slope + p_noise + rho), data = tmp2)
+#summary(mod_LMM)
+#fixef(mod_LMM)[order(abs(fixef(mod_LMM)))]
+coefs <- stdCoef(mod_all)[,1, drop = FALSE]^2
+
+## "Variance explained" by N, sigma_n0, sigma_b1, p_noise and rho:
+sum(coefs[grep("N", rownames(coefs)), ])
+sum(coefs[grep("sigma_int", rownames(coefs)), ])
+sum(coefs[grep("p_noise", rownames(coefs)), ])
+sum(coefs[grep("rho", rownames(coefs)), ])
+sum(coefs[grep("sigma_slope", rownames(coefs)), ])
+
+## Note, above is not really variance explained, but some measure 
+## of influence that does not add up to 1
+
+## Variance explained by main effects:
+coefs["N250", ] 
+coefs["sigma_int2", ]
+coefs["p_noise", ]
+coefs["sigma_slope0.632455532033676", ]
+coefs["rho0.3", ]
+
+round(coefs[order(coefs$stdcoef, decreasing = TRUE), , drop = FALSE], digits = 3)
+
+## Variance explained by random effects:
+varco <- as.data.frame(VarCorr(mod_all))$vcov /
+  var(tmp2[, "tree_size"])
+varco
+@
+
 \begin{figure}[!ht]
@@
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-014}
+<<echo=FALSE, fig=TRUE, height=3.5>>=
+tmp2$method <- ifelse(
+  tmp2$method_id %in% c("lm_c", "lmer_c", "lmer_s_c"), "LM(M) tree", 
+      ifelse(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s"), "SEM tree (LRT-based)", 
+             ifelse(tmp2$method_id %in% c("s_sem", "s_sem_i", "s_sem_s"),
+                             "SEM tree (score-based)", "LongCART")))
+tmp2$random <- ifelse(tmp2$method_id %in% c("lm_c", "f_sem", "s_sem"), "no", 
+                      ifelse(tmp2$method_id %in% c("lmer_c", "f_sem_i", "s_sem_i", "Lcrt"),
+                             "int.", "int. + \nslope"))
+tmp2$random <- factor(tmp2$random, levels = c("no", "int.", "int. + \nslope"))
+tmp2$method <- factor(tmp2$method, levels = c("LM(M) tree", "SEM tree (LRT-based)", 
+                                              "SEM tree (score-based)", "LongCART"))
+breaks <- c(0, 1, 3, 7, 15)
+lim <- c(0, 15)
+plot1 <- ggplot(tmp2) +
+  geom_boxplot(aes(x=random, y=tree_size), alpha = .5, width = .6) + 
+  geom_hline(yintercept=3, col = "darkgrey") +
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~ method, scales = "free", space = "free") +
+  labs(x = "", y = "Tree size")
+plot1
+##ggsave(file = "tree_size_plot.png", height = 9.5, width=19, units = "cm")
+@
 %\includegraphics{tree_size_plot.png}
@@
 Figure~\ref{fig:tree_sizes} depicts tree size distributions for the different algorithms.
-LRT-based SEM trees appear somewhat overpowered with a mean number of splits of 3.61 ($SD = $1.71), while score-based SEM trees comprised the correct number of splits on average ($M = $3.09; $SD = $1.61).
+LRT-based SEM trees appear somewhat overpowered with a mean number of splits of \Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["SEM tree (LRT-based)"], digits = 2)} ($SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["SEM tree (LRT-based)"], digits = 2)}), while score-based SEM trees comprised the correct number of splits on average ($M = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["SEM tree (score-based)"], digits = 2)}; $SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["SEM tree (score-based)"], digits = 2)}).
 SEM trees appear more systematically affected by mis-specification of the random-effects structure than LMM trees and under-specification yields a higher number of splits.
-LongCART trees appear underpowered, with a mean number of splits of 2.01 ($SD = $1.45).
+LongCART trees appear underpowered, with a mean number of splits of \Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["LongCART"], digits = 2)} ($SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["LongCART"], digits = 2)}).
 This could be the result of under-specification of the random-effects structure: LongCART trees incorporate a random-intercept term only and this default cannot be overridden.
+
+<<eval=FALSE, echo=FALSE>>=
+pairwise.t.test(tmp2$tree_size, tmp2$method_id, 
+               alternative = "two.sided", p.adjust = "bonferroni")
+# 	Pairwise comparisons using t tests with pooled SD 
+# 
+# data:  tmp2$tree_size and tmp2$method_id 
+# 
+#          lmer_c  lmer_s_c lm_c    f_sem   f_sem_i f_sem_s s_sem   s_sem_i s_sem_s
+# lmer_s_c 1.00000 -        -       -       -       -       -       -       -      
+# lm_c     < 2e-16 < 2e-16  -       -       -       -       -       -       -      
+# f_sem    < 2e-16 < 2e-16  < 2e-16 -       -       -       -       -       -      
+# f_sem_i  0.00074 2.5e-08  < 2e-16 < 2e-16 -       -       -       -       -      
+# f_sem_s  1.00000 1.00000  < 2e-16 < 2e-16 9.6e-08 -       -       -       -      
+# s_sem    < 2e-16 < 2e-16  < 2e-16 < 2e-16 < 2e-16 < 2e-16 -       -       -      
+# s_sem_i  1.00000 1.00000  < 2e-16 < 2e-16 8.2e-09 1.00000 < 2e-16 -       -      
+# s_sem_s  < 2e-16 < 2e-16  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 -      
+# Lcrt     < 2e-16 < 2e-16  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 0.03511
+@
 
@@
 Bonferroni-corrected $p$-values indicated significant differences in tree size between most methods (all $p < .001$), with some notable exceptions: The number of splits implemented by LMM trees with clustered covariances was indistinguishable from that of LRT-based SEM trees with freely estimated random intercepts and slopes, and from the score-based SEM trees with freely estimated random intercepts.
+
+<<eval=FALSE, echo=FALSE>>=
+## t-tests for comparing no of splits with true number of splits
+round(apply((sizes-1)/2, 2, function(x) t.test(x, mu = 3)$p.value), digits = 3)
+@
 
@@
 
-% latex table generated in R 4.1.3 by xtable 1.8-4 package
-% Wed Oct 26 00:12:30 2022
-\begin{table}[!ht]
-\caption{Variable selected for the first split by each of the partitioning methods.} 
-\label{tab:first_splits2}
-\begin{tabular}{llrrrrr}
-  \hline
-Algorithm & Random effects & $U_1$ & $U_2$ & $U_3$ & $U_4$ - $U_{25}$ & no split \\ 
-  \hline
-LM(M) tree & $\sigma_{b_0} = \sigma_{b_1} = 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-    (clustered & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-      covariances) & $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-   \hline
-SEM tree & $\sigma_{b_0} = \sigma_{b_1} = 0$ & 1.000 & 0.000 & 0.000 & 0.000 & 0.000 \\ 
-    (LRT-based) & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & 0.998 & 0.000 & 0.002 & 0.000 & 0.000 \\ 
-    & $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & 0.992 & 0.001 & 0.001 & 0.002 & 0.004 \\ 
-   \hline
-SEM tree & $\sigma_{b_0} = \sigma_{b_1} = 0$ & 0.761 & 0.176 & 0.052 & 0.012 & 0.000 \\ 
-    (score-based) & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ & 0.879 & 0.077 & 0.044 & 0.000 & 0.000 \\ 
-    & $\sigma_{b_0} > 0$, $\sigma_{b_1} > 0$ & 0.999 & 0.000 & 0.000 & 0.000 & 0.001 \\ 
-   \hline
-LongCART & $\sigma_{b_0} > 0$, $\sigma_{b_1} = 0$ &  & 0.419 & 0.297 & 0.092 & 0.192 \\ 
-   \hline \multicolumn{7}{l}{\footnotesize\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables}\\  \multicolumn{7}{l}{\footnotesize are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3).
-$\sigma_{b_0}$ and $\sigma_{b_1}$}\\  \multicolumn{7}{l}{\footnotesize are the standard deviations of the random intercept and slope, respectively.}\\ \end{tabular}
-\end{table}
+<<echo=FALSE, eval=FALSE>>=
+tapply(tmp2$tree_size, list(tmp2$method_id, tmp2$N), mean)
+tapply(tmp2$tree_size, list(tmp2$method_id, tmp2$sigma_int), mean)
+tapply(tmp2$tree_size, list(tmp2$method_id, tmp2$p_noise), mean)
+tapply(tmp2$tree_size, list(tmp2$method_id, tmp2$rho), mean)
+tapply(tmp2$tree_size, list(tmp2$method_id, tmp2$sigma_slope), mean)
+@
+
+<<echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
+splits$Algorithm[2] <- "LM(M) tree"
+splits$Algorithm[4] <- "  (clustered"
+splits$Algorithm[8] <- "    covariances)"
+splits$Algorithm[12] <- "  (LRT-based)"
+splits$Algorithm[14] <- "SEM tree"
+splits$Algorithm[15] <- "  (score-based)"
+print(xtable(splits[c(2, 4, 8, 11:17), -3], digits = 3,
+             caption = "Variable selected for the first split by each of the partitioning methods.", 
+             label = "tab:first_splits2"), include.rownames=FALSE, 
+      hline.after = c(-1, 0, 3, 6, 9), table.placement = "!ht",
+      add.to.row = list(pos = list(10, 10, 10, 10), command = c(
+        "\\hline",
+        "\\multicolumn{7}{l}{\\footnotesize\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables}\\\\ ", 
+        "\\multicolumn{7}{l}{\\footnotesize are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\\sigma_{b_0}$ and $\\sigma_{b_1}$}\\\\ ",
+        "\\multicolumn{7}{l}{\\footnotesize are the standard deviations of the random intercept and slope, respectively.}\\\\ ")),
+      latex.environment = "", caption.placement = "top",
+      sanitize.text.function=function(x){x})
+
+@
 
@@
 For SEM trees, the LRT criterion yields almost perfect accuracy for the first split.
-The score-based approach selected the wrong splitting variable for the first split in about 12\% of datasets, but only when the random-effects structure was mis-specified (i.e., $\sigma_{b_0}$ and/or $\sigma_{b_1}$ fixed to 0).
+The score-based approach selected the wrong splitting variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$U_1$`[grep("s_sem", rownames(splits))])))}\% of datasets, but only when the random-effects structure was mis-specified (i.e., $\sigma_{b_0}$ and/or $\sigma_{b_1}$ fixed to 0).
 When random effects were correctly specified, score-based SEM trees also provided perfect recovery of the first split.
@@
 LongCART showed a more dramatic tendency than score-based SEM trees to use $U_2$ and $U_3$ for implementing the first split.
-Of note, in the LongCART trees where no split was implemented (19\% of datasets), 99\% $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not exceed the criterion of significance in those datasets.
+Of note, in the LongCART trees where no split was implemented (\Sexpr{round(100*prop.table(table(tree_size$Lcrt))[1])}\% of datasets), \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not exceed the criterion of significance in those datasets.
 This suggests that either the parameter stability tests proposed by \cite{KundyHare19} are less sensitive to instabilities with respect to the fixed intercept, and/or with respect to categorical covariates.
+
+<<echo=FALSE, eval=FALSE>>=
+## Check: How come that score-based SEM trees do not pick up x1 for splitting?
+## Which dataset yields the worst SEM tree(s)? Lots of nodes, wrong first splitting variable?
+which.max(tree_size$f_sem)
+which.max(tree_size$s_sem) ## seems worst
+
+splitvar[which.max(tree_size$s_sem),]
+which(!splitvar[ , "s_sem"] %in% c("x1", "x2", "x3"))
+## Size of score-based semtrees where it picks up noise for 1st split
+which(!splitvar[ , "s_sem"] %in% c("x1", "x2", "x3"))
+tree_size$s_sem[which(!splitvar[ , "s_sem"] %in% c("x1", "x2", "x3"))]
+## Size of score-based semtrees where it picks up x2 or x3 for 1st split
+which(splitvar[ , "s_sem"] %in% c("x2", "x3"))
+tree_size$s_sem[which(splitvar[ , "s_sem"] %in% c("x2", "x3"))]
+
+## Could it be that tests w.r.t. residuals are overpowered with misspecification?
+
+## It seems noise variables are mostly picked up when sample size is larger:
+table(which(!splitvar[ , "s_sem"] %in% c("x1", "x2", "x3")) %% 2 == 0)
+table(which(splitvar[ , "s_sem"] %in% c("x2", "x3")) %% 2 == 0)
+head(tmp$N)
+
+for (i in which(splitvar[ , "s_sem"] %in% c("x2", "x3"))[1:7]) {
+  print(table(which(splitvar[ , "s_sem"] %in% c("x2", "x3")) %% 32 == i))
+}
+## best chances at 1st and 4th value, that's rows 18 and 26 of design matrix
+tmp[18,]
+tmp[26,]
+
+## 1) Fit lavaan model
+## 2) Extract parameter stability tests using strucchange
+load("datasets1")
+p <- if (ncol(datasets[[18]]) > 12L) 28L else 8L
+
+data <- reshape(data = datasets[[18]], v.names = "y",
+                timevar = "time", idvar = "cluster_id",
+                direction = "wide")
+
+# data <- reshape(data = datasets[[26]], v.names = "y",
+#                 timevar = "time", idvar = "cluster_id",
+#                 direction = "wide")
+
+mod <- '
+        i =~ 1*y.0 + 1*y.1 + 1*y.2 + 1*y.3 + 1*y.4
+        s =~ 1*y.1 + 2*y.2 + 3*y.3 + 4*y.4
+        s ~~ 0*s
+        i ~~ 0*i
+        i ~~ 0*s
+      '
+fit <- growth(mod, data = data, do.fit = FALSE)
+
+mod.i <- '
+        i =~ 1*y.0 + 1*y.1 + 1*y.2 + 1*y.3 + 1*y.4
+        s =~ 1*y.1 + 2*y.2 + 3*y.3 + 4*y.4
+        s ~~ 0*s
+        i ~~ 0*s
+      '
+fit.i <- growth(mod.i, data = data, do.fit = FALSE)
+
+mod.s <- '
+        i =~ 1*y.0 + 1*y.1 + 1*y.2 + 1*y.3 + 1*y.4
+        s =~ 1*y.1 + 2*y.2 + 3*y.3 + 4*y.4
+      '
+fit.s <- growth(mod.s, data = data, do.fit = FALSE)
+
+st <- semtree(fit, predictors = paste0("x", 1L:p), data = data,
+              control = semtree.control(method = "score", 
+                                        bonferroni = TRUE))
+st_i <- semtree(fit.i, predictors = paste0("x", 1L:p), data = data,
+              control = semtree.control(method = "score", 
+                                        bonferroni = TRUE))
+st_s <- semtree(fit.s, predictors = paste0("x", 1L:p), data = data,
+              control = semtree.control(method = "score", 
+                                        bonferroni = TRUE))
+
+for (parm in 1:7) {
+for (i in c("x2")){#}, "x2", "x3", "x4", "x5")) {
+  efp <- gefp(st$model, ## lavaan model in first node
+              fit = NULL, 
+              scores = function(fm) estfun(fm), 
+              vcov = function(fm, order.by = order.by, data = data) fm@vcov$vcov, 
+              data = data, 
+              order.by = data[, i],
+              sandwich=FALSE,
+              parm = parm)
+  if (inherits(data[ , i], "factor")) {
+    LMuo <- catL2BB(efp)
+    plot(efp, functional = LMuo, 
+         xlab = paste(i, "(categorical partitioning variable)"))
+    print(sctest(efp, functional = LMuo))    
+  } else if (inherits(data[ , i], c("integer", "numeric"))) { 
+    maxLM <- supLM(0.1)
+    plot(efp, xlab = paste(i, "(numeric partitioning variable)"),
+         functional = maxLM)
+    print(sctest(efp, functional = maxLM))
+  }
+}
+}
+## Model without intercept and slope variances: Plots suggest all variables have very high test stats.
+## But x2 and x3 win by far, much larger test stats. x1 also has significant test stat, but:
+## Interestingly, the test with respect to the mean slope is much higher for 
+##   the test w.r.t. x1, compared to that of the intercept
+
+## Model with intercept but no slope variance: Again, all variables have very high stats but x2 and x3 win by far.
+## Test w.r.t. x1 and intercept mean is now much smaller compared to test w.r.t. slope mean.
+
+## Model with intercept and slope variance: Again, all variables have very high stats but x2 and x3 win by far.
+## Test w.r.t. x1 and intercept mean is now much smaller compared to test w.r.t. slope mean.
+
+## It a appears that the parameter test w.r.t. slope mean is much more sensitive than w.r.t. intercept mean
+## You can see this when you fill in: st, order.by = data[ , "x2"], parm = 7 versus parm = 6
+## Error variance tests behave well. But test w.r.t. slope mean is much higher than w.r.t. intercept mean.
+## Then again, this should be the case for x2, but NOT for x1.
+@
 
@@
 \begin{subfigure}{.6\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-020}
+<<fig=TRUE, echo=FALSE, width=4, height=4>>=
+load("comp_times.Rda")
+colnames(times) <- c(rep("LMM tree", times = 8),
+                     rep(" LM tree", times = 2), 
+                     rep("SEM tree\n(LRT)", times = 3),
+                     rep("SEM tree\n(score)", times = 3),
+                     "LongCART")
+times <- reshape2::melt(times)
+ggplot(times, aes(Var2, value)) + geom_boxplot(width=.5, alpha = .5) + 
+  labs(x = "", y = "Computation time (in seconds)") +
+  scale_y_continuous(trans = "log", breaks = c(.01, .1, 1, 10, 100, 1000, 10000))
+mean_times <- tapply(times$value, times$Var2, mean)
+@
 \end{subfigure}
@@
 Figure~\ref{fig:comp_times} presents computation time distributions for the partitioning algorithms.
-A clear computational advantage is observed for LM trees, with an average computation time of 0.141 seconds.
-Including random-effects estimation increased computation time: for LMM trees, average computation time was 0.65 seconds.
-LongCART and SEM trees yielded much longer computation times, with LongCART requiring 23.36 seconds, score-based SEM trees 4.4 seconds, and LRT-based SEM trees 2685.37 seconds computation time, on average.
+A clear computational advantage is observed for LM trees, with an average computation time of \Sexpr{round(mean_times[" LM tree"], digits = 3)} seconds.
+Including random-effects estimation increased computation time: for LMM trees, average computation time was \Sexpr{round(mean_times["LMM tree"], digits = 2)} seconds.
+LongCART and SEM trees yielded much longer computation times, with LongCART requiring \Sexpr{round(mean_times["LongCART"], digits = 2)} seconds, score-based SEM trees \Sexpr{round(mean_times["SEM tree\n(score)"], digits = 2)} seconds, and LRT-based SEM trees \Sexpr{round(mean_times["SEM tree\n(LRT)"], digits = 2)} seconds computation time, on average.
 
@@
 This yielded a very large number of splits in the current analyses and poor predictive accuracy.
-We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = 32 terminal nodes, or 31 splits.
+We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = \Sexpr{2^5} terminal nodes, or \Sexpr{(2^5)-1} splits.
 Significance of the LRT, or a cross-validation approach to determine whether to implement a split or not may yield better results, but neither approach is implemented in longRPart (nor longRPart2).
@@
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-021}
+<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+## Load datasets for computing variance of response variables
+load("Reading ability data.Rdata")
+var_read <- var(readdata$score)
+load("Math ability data.Rdata")
+var_math <- var(mathdata$score)
+load("Science ability data.Rdata")
+var_scie <- var(sciedata$score)
+
+## MSE
+load("MSEs, N = 250, Math.Rda")
+math_MSEs <- cbind(MSEs, N = 250)
+load("MSEs, N = 1000, Math.Rda")
+math_MSEs <- rbind(math_MSEs, cbind(MSEs, N = 1000))
+#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart
+#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart_s
+#sapply(math_MSEs, function(x) tapply(x, math_MSEs$N, mean, na.rm =TRUE))
+
+load("MSEs, N = 250, Reading.Rda")
+read_MSEs <- cbind(MSEs, N = 250)
+load("MSEs, N = 1000, Reading.Rda")
+read_MSEs <- rbind(read_MSEs, cbind(MSEs, N = 1000))
+#sapply(read_MSEs, function(x) table(is.na(x)))
+#sapply(read_MSEs, function(x) tapply(x, read_MSEs$N, mean))
+
+load("MSEs, N = 250, Science.Rda")
+scie_MSEs <- cbind(MSEs, N = 250)
+load("MSEs, N = 1000, Science.Rda")
+scie_MSEs <- rbind(scie_MSEs, cbind(MSEs, N = 1000))
+#sapply(scie_MSEs, function(x) table(is.na(x)))$longRPart
+#sapply(scie_MSEs, function(x) table(is.na(x)))$longRPart_s
+#sapply(scie_MSEs, function(x) tapply(x, scie_MSEs$N, mean))
+
+tmp3 <- data.frame(stack(math_MSEs[ , -c(4,6,7,10)]),
+                    dataset_id = factor(rep(1:nrow(math_MSEs), times = 12)),
+                   N = math_MSEs$N)
+names(tmp3)[1:2] <- c("MSE", "method") 
+tmp3$method2 <- as.character(tmp3$method)
+tmp3$method2[grepl("LMtree", tmp3$method2)] <- "LM tree\n\n(cluster)"
+tmp3$method2[tmp3$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
+tmp3$method2[tmp3$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
+tmp3$method2[tmp3$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
+tmp3$method2[grepl("longR", tmp3$method2)] <- "longRPart"
+tmp3$random <- ifelse(tmp3$method == "LMtree_c", "no", ifelse(
+  tmp3$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+  ifelse(tmp3$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+    "int. + \nslope", NA)))
+tmp3$N <- factor(tmp3$N)
+
+theme_set(theme_gray(base_size = 10))
+ggplot(tmp3) +
+  geom_boxplot(aes(x=random, y=MSE, fill = N), 
+               position=position_dodge(1), alpha = .5, width = .6) +
+  scale_y_continuous(trans = log_trans(),
+                     sec.axis = sec_axis(trans=~1-(./var_math))) +
+  facet_grid(~method2, scales = "free", space = "free") +
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(x = "", y = "MSE")
+@
 \end{subfigure}
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-022}
+<<echo=FALSE, fig=TRUE, height=1.6, width=8>>=
+tmp4 <- data.frame(stack(read_MSEs[ , -c(4,6,7,10)]),
+                    dataset_id = factor(rep(1:nrow(read_MSEs), times = 12)),
+                   N = read_MSEs$N)
+names(tmp4)[1:2] <- c("MSE", "method") 
+tmp4$method2 <- as.character(tmp4$method)
+tmp4$method2[grepl("LMtree", tmp4$method2)] <- "LM tree\n\n(cluster)"
+tmp4$method2[tmp4$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
+tmp4$method2[tmp4$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
+tmp4$method2[tmp4$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
+tmp4$method2[grepl("longR", tmp4$method2)] <- "longRPart"
+tmp4$random <- ifelse(tmp4$method == "LMtree_c", "no", ifelse(
+  tmp4$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+  ifelse(tmp4$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+    "int. + \nslope", NA)))
+tmp4$N <- factor(tmp4$N)
+ggplot(tmp4) +
+  geom_boxplot(aes(x=random, y=MSE, fill=N), 
+               position=position_dodge(1), alpha = .5, width = .6) +
+    scale_y_continuous(trans = log_trans(),
+                       sec.axis = sec_axis(trans=~1-(./var_read))) +
+  facet_grid(~method2, scales = "free", space = "free") +
+  theme(strip.background = element_blank(), strip.text.x = element_blank(),
+        axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(x = "", y = "MSE")
+@
 \end{subfigure}
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-023}
+<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+tmp5 <- data.frame(stack(scie_MSEs[ , -c(4,6,7,10)]),
+                    dataset_id = factor(rep(1:nrow(scie_MSEs), times = 12)),
+                   N = scie_MSEs$N)
+names(tmp5)[1:2] <- c("MSE", "method") 
+tmp5$method2 <- as.character(tmp5$method)
+tmp5$method2[grepl("LMtree", tmp5$method2)] <- "LM tree\n\n(cluster)"
+tmp5$method2[tmp5$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
+tmp5$method2[tmp5$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
+tmp5$method2[tmp5$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
+tmp5$method2[grepl("longR", tmp5$method2)] <- "longRPart"
+tmp5$random <- ifelse(tmp5$method == "LMtree_c", "no", ifelse(
+  tmp5$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+  ifelse(tmp5$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+    "int. + \nslope", NA)))
+tmp5$N <- factor(tmp5$N)
+ggplot(tmp5) +
+  geom_boxplot(aes(x=random, y=MSE, fill=N), 
+               position=position_dodge(1), alpha = .5, width = .6) +
+    scale_y_continuous(trans = log_trans(),
+                       sec.axis = sec_axis(trans=~1-(./var_scie))) +
+  facet_grid(~method2, scales = "free", space = "free") +
+  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(x = "", y = "MSE")
+@
 \end{subfigure}
@@
 
-\begin{table}
-
-\begin{threeparttable}
-\caption{\label{tab:application_MSEs}Cross-validated mean squared errors for each of the response variables.}
-\centering
-\fontsize{8}{10}\selectfont
-\begin{tabular}[t]{lcccccc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Math} & \multicolumn{2}{c}{Reading} & \multicolumn{2}{c}{Science} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
-\multicolumn{1}{c}{ } & \multicolumn{1}{c}{N = 250} & \multicolumn{1}{c}{N = 1,000} & \multicolumn{1}{c}{N = 250} & \multicolumn{1}{c}{N = 1,000} & \multicolumn{1}{c}{N = 250} & \multicolumn{1}{c}{N = 1,000} \\
-\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7}
-LM tree (c) & 0.1407 (0.004) & 0.1290 (0.002) & 0.1262 (0.003) & 0.1191 (0.001) & 0.4178 (0.012) & 0.3787 (0.007)\\
-LMM tree (i,c) & 0.1398 (0.004) & 0.1283 (0.002) & 0.1261 (0.003) & 0.1191 (0.001) & 0.4151 (0.013) & 0.3777 (0.006)\\
-LMM tree (i,s,c) & 0.1403 (0.005) & 0.1287 (0.002) & 0.1264 (0.003) & 0.1192 (0.001) & 0.4143 (0.012) & 0.3777 (0.006)\\
-LMM tree (i,r) & 0.1441 (0.006) & 0.1285 (0.002) & 0.1302 (0.004) & 0.1202 (0.002) & 0.4124 (0.013) & 0.3762 (0.006)\\
-longRPart (i) & 0.1463 (0.006) & 0.1318 (0.003) & 0.1335 (0.006) & 0.1213 (0.003) & 0.4197 (0.015) & 0.4050 (0.011)\\
-longRPart (i,s) & 0.1430 (0.005) & 0.1306 (0.002) & 0.1291 (0.004) & 0.1199 (0.002) & 0.4289 (0.020) & 0.4296 (0.022)\\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}
-\small
-\item [] \footnotesize \textit{Note.} Values represent means over 100 repetitions, parenthesized values represent SDs and can be interpreted as standard errors.
-c = cluster-level covariances; r = estimation initialized with random effects; i =  random-intercept variance freely estimated; s = random-slope variance freely estimated.
-\end{tablenotes}
-\end{threeparttable}
-\end{table}
-
-Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions for each of the partitioning approaches, separated by sample size\footnote{LongRPart trees comprising random intercepts and slopes ran into estimation errors in some datasets: In the Math ability data, 2 out of 200 datasets did not yield a tree.
+<<echo=FALSE, eval=FALSE>>=
+## Pairwise difference tests
+pairwise.t.test(tmp3$MSE, tmp3$method, alternative = "two.sided", 
+                p.adjust = "bonferroni")
+##             LMtree_c LMMtree_c LMMtree_r LMMtree_sc longRPart
+## LMMtree_c   1.00000  -         -         -          -        
+## LMMtree_r   0.08664  0.00033   -         -          -        
+## LMMtree_sc  1.00000  1.00000   0.00801   -          -        
+## longRPart   1.7e-13  < 2e-16   8.6e-06   5.8e-16    -        
+## longRPart_s 0.00789  1.1e-05   1.00000   0.00046    0.00029
+
+pairwise.t.test(tmp4$MSE, tmp4$method, alternative = "two.sided", 
+                p.adjust = "bonferroni")
+##            LMtree_c LMMtree_c LMMtree_r LMMtree_sc longRPart
+## LMMtree_c   1.00     -         -         -          -        
+## LMMtree_r   4.0e-10  1.2e-10   -         -          -        
+## LMMtree_sc  1.00     1.00      3.7e-09   -          -        
+## longRPart   < 2e-16  < 2e-16   1.9e-07   < 2e-16    -        
+## longRPart_s 1.9e-05  7.9e-06   0.98      1.0e-04    9.4e-13
+
+pairwise.t.test(tmp5$MSE, tmp5$method, alternative = "two.sided", 
+                p.adjust = "bonferroni")
+##            LMtree_c LMMtree_c LMMtree_r LMMtree_sc longRPart
+## LMMtree_c   1.000    -         -         -          -        
+## LMMtree_r   0.089    1.000     -         -          -        
+## LMMtree_sc  1.000    1.000     1.000     -          -        
+## longRPart   <2e-16   <2e-16    <2e-16    <2e-16     -        
+## longRPart_s <2e-16   <2e-16    <2e-16    <2e-16     <2e-16
+
+@
+
+<<application_MSE, echo=FALSE, results=tex>>=
+library("kableExtra")
+MSE_df <- data.frame(t(sapply(math_MSEs[ , -c(4,6,7,10)], 
+                              function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(math_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE)))))
+MSE_df <- cbind(paste0(MSE_df[,1], " (", MSE_df[,3], ")"),
+                paste0(MSE_df[,2], " (", MSE_df[,4], ")"),
+                paste0(MSE_df[,5], " (", MSE_df[,7], ")"), 
+                paste0(MSE_df[,6], " (", MSE_df[,8], ")"), 
+                paste0(MSE_df[,9], " (", MSE_df[,11], ")"),
+                paste0(MSE_df[,10], " (", MSE_df[,12], ")"))
+rownames(MSE_df) <- c("LM tree (c)", "LMM tree (i,c)", "LMM tree (i,r)", "LMM tree (i,s,c)",  
+                      "longRPart (i)", "longRPart (i,s)")
+MSE_df <- MSE_df[c(1,2,4,3,5,6), ]
+
+kable_styling(add_header_above(add_header_above(add_footnote(
+  kable(MSE_df, format = "latex", booktabs = TRUE, label = "application_MSEs", align = "c", 
+        escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
+        caption = "Cross-validated mean squared errors for each of the response variables."),
+  "\\footnotesize \\textit{Note.} Values represent means over 100 repetitions, parenthesized values represent SDs and can be interpreted as standard errors.
+c = cluster-level covariances; r = estimation initialized with random effects; i =  random-intercept variance freely estimated; s = random-slope variance freely estimated.",
+  notation="none", threeparttable = TRUE, escape = FALSE),
+  c(" ", "N = 250"=1, "N = 1,000"=1, "N = 250"=1, "N = 1,000"=1,
+    "N = 250"=1, "N = 1,000"=1), align = "c"),
+  c(" ", "Math" = 2, "Reading" = 2, "Science" = 2), align = "c"),
+  font_size = 8, full_width=FALSE)
+@
+
+<<echo=FALSE, eval=FALSE>>=
+
+###############################
+##
+## Check estimation problems
+##  for longRPart
+##
+
+load("notes, N = 250, Math.Rda")
+math_notes <- cbind(notes, N = 250)
+load("notes, N = 1000, Math.Rda")
+math_notes <- rbind(math_notes, cbind(notes, N = 1000))
+sapply(math_notes, function(x) table(is.na(x)))
+## 8 Warnings for LM(M)tree and longRPart: level of race removed from test data
+## 76 Warnings for longRPart_s: 
+sapply(math_notes, function(x) table(x, useNA = "ifany"))
+## 4 warnings for race omitted from test data
+## 76-4=72 warnings for 1, 2 or 3 nodes estimation not converged
+
+load("notes, N = 250, Reading.Rda")
+read_notes <- cbind(notes, N = 250)
+load("notes, N = 1000, Reading.Rda")
+read_notes <- rbind(read_notes, cbind(notes, N = 1000))
+sapply(read_notes, function(x) table(is.na(x)))
+## 18 Warnings for LM(M)tree and longRPart: level of race removed from test data
+## 80 Warnings for longRPart_s: 
+sapply(read_notes, function(x) table(x, useNA = "ifany"))
+## 11 warnings for race omitted from test data
+## 80-11=69 warnings for 1 or 2 nodes estimation not converged
+
+load("notes, N = 250, Science.Rda")
+scie_notes <- cbind(notes, N = 250)
+load("notes, N = 1000, Science.Rda")
+scie_notes <- rbind(scie_notes, cbind(notes, N = 1000))
+sapply(scie_notes, function(x) table(is.na(x)))
+## 18 Warnings for LM(M)tree and longRPart: level of race removed from test data
+## 82 Warnings for longRPart_s: 
+sapply(scie_notes, function(x) table(x, useNA = "ifany"))
+## 12 warnings for race omitted from test data
+## 82-12=70 warnings for 1, 2 or 3 nodes estimation not converged
+
+## How does this correspond to MSEs?
+sapply(math_MSEs, function(x) table(is.na(x), math_MSEs$N))
+sapply(read_MSEs, function(x) table(is.na(x), read_MSEs$N))
+sapply(scie_MSEs, function(x) table(is.na(x), scie_MSEs$N))
+## Oops, we got MSE anyway
+
+## Identify a tree where longRPart did not converge in a node
+tapply(math_sizes$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
+       table, useNA = "ifany")
+tapply(math_sizes$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
+       mean, na.rm = TRUE)
+## Non-converged nodes are in larger trees, this makes sense (smaller nodes)
+tapply(math_MSEs$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
+       table, useNA = "ifany")
+tapply(math_MSEs$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
+       mean, na.rm = TRUE)
+tapply(math_MSEs$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
+       sd, na.rm = TRUE)
+
+## [6] "NANode 0.221767719908376did not converge. "      
+math_sizes[6,]
+math_MSEs[6,]
+
+## Explanation: If lme model did not converge, no predictions were
+##   generated and written to preds. Thus, predictions from last lme
+##   were used for those nodes, which is always the lme with random 
+## intercepts only
+
+## Illustrate by refiting trees:
+load("bag_ids, N = 250, Math.Rda")
+library("longRPart")
+
+## Prepare formulas and data
+randomFormula <-  ~ 1 | CHILDID
+randomFormula_s <- ~ (1 + months) | CHILDID
+fixedFormula <- score ~ months
+rPartFormula <- ~ GENDER + RACE + WKSESL + C1GMOTOR + C1FMOTOR + 
+  T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE
+resp <- "Math"
+load(paste0(resp, " ability data.Rdata"))
+resp_short <- ifelse(resp == "Math", "math", ifelse(resp == "Reading", "read", "scie"))
+data <- get(paste0(resp_short, "data"))
+data$CHILDID <- factor(data$CHILDID)
+data$GENDER <- factor(data$GENDER)
+data$asmtmm <- factor(data$asmtmm)
+
+## Set up train and test data  
+train <- data[data$CHILDID %in% bag_ids[ , 6], ]
+test <- data[!data$CHILDID %in% bag_ids[ , 6], ]
+## Remove obs with levels of race in test are also in train
+levs <- unique(test$RACE) %in% unique(train$RACE)
+if (!all(levs)) {
+  test <- test[-which(!test$RACE %in% unique(test$RACE)[levs]), ]
+  print(paste0(notes[i, method], "Levels or race omitted from test data: ", 
+               unique(test$race)[!levs]))
+}
+
+## Fit longRPart
+lrpdata <- train[ , c("CHILDID", "months", "score", "GENDER", "RACE",
+                      "WKSESL", "C1GMOTOR", "C1FMOTOR", "T1INTERN",
+                      "T1EXTERN", "T1INTERP", "T1CONTRO", "P1FIRKDG",
+                      "AGEBASELINE")]
+lrpdata$CHILDID <- as.numeric(lrpdata$CHILDID)
+
+for (form in 1:2) {
+  rForm <- list(randomFormula, randomFormula_s)[[form]]
+  method <- c("longRPart", "longRPart_s")[form]
+  system.time(
+    tree <- try(longRPart(randomFormula = rForm, lmeFormula = fixedFormula,
+                    rPartFormula = rPartFormula, data = lrpdata, 
+                    control = rpart.control(maxdepth = 5))))["elapsed"]
+
+  ## Check size
+  (nrow(tree$frame)-1)/2
+  if (form == 1L) math_sizes$longRPart[6] else if (form == 2L) 
+    math_sizes$longRPart_s[6]
+
+  if (class(tree) != "try-error") {
+
+    ## Get predictions
+    train_nodes <- factor(predict(tree))
+    test_nodes <- factor(predict(tree, newdata = test),
+                         levels = levels(train_nodes))
+    preds <- rep(NA, times = nrow(test))
+    ## Fit lme to training observations in each node and predict test observations
+    for (j in unique(train_nodes)) {
+        lme_mod <- try(lme(random = tree$randomFormula,
+                           fixed = tree$lmeFormula,
+                           data = tree$data[train_nodes == j, ]))
+        if (class(lme_mod) != "try-error") {
+          preds[test_nodes == j] <- predict(lme_mod, 
+                                            newdata = test[test_nodes == j, ], 
+                                            level = 0)
+        } else {
+          print(paste0("Node ", j, "did not converge. "))
+        }
+    }
+    ## Check MSE:
+    mean((test$score - preds)^2)  
+    if (form == 1L) math_MSEs$longRPart[6] else if (form == 2L) math_MSEs$longRPart_s[6]
+  }
+}
+## MSEs and tree sizes match up
+
+## How many observations are affected?
+table(is.na(math_sizes$longRPart), math_sizes$N)
+table(is.na(math_sizes$longRPart_s), math_sizes$N)
+table(is.na(read_sizes$longRPart), read_sizes$N)
+table(is.na(read_sizes$longRPart_s), read_sizes$N)
+table(is.na(scie_sizes$longRPart), scie_sizes$N)
+table(is.na(scie_sizes$longRPart_s), scie_sizes$N)
+
+table(is.na(math_sizes$longRPart_s), is.na(math_MSEs$longRPart_s))
+table(is.na(read_sizes$longRPart_s), is.na(read_MSEs$longRPart_s))
+table(is.na(scie_sizes$longRPart_s), is.na(scie_MSEs$longRPart_s))
+
+## How strong is the influence?
+
+## Let's say on average, 2 nodes are affected (mean of 1:3)
+
+## Estimate total # of level-2 obs affected divide by total #:
+math_affected <- sum(2 * math_sizes$N[!is.na(math_notes$longRPart_s)] / 
+  (math_sizes$longRPart_s[!is.na(math_notes$longRPart_s)]+1))
+math_affected / sum(math_sizes$N)
+## [1] 0.06443202
+
+read_affected <- sum(2 * read_sizes$N[!is.na(read_notes$longRPart_s)] / 
+  (read_sizes$longRPart_s[!is.na(read_notes$longRPart_s)]+1))
+read_affected / sum(read_sizes$N)
+## [1] 0.111979
+
+scie_affected <- sum(2 * scie_sizes$N[!is.na(scie_notes$longRPart_s)] / 
+  (scie_sizes$longRPart_s[!is.na(scie_notes$longRPart_s)]+1), na.rm = TRUE)
+scie_affected / sum(scie_sizes$N)
+## [1] 0.2531524
+
+## Ideally, would refit all longRPart_s trees which had convergence errors
+## in one of the nodes. And then use NA for prediction, or refit a model
+## without random slopes. This might paint a slightly difference picture.
+## However, I do not expect dramatic differences, given that results
+## for longRPart and longRPart_s are quite similar, and seem to differ
+## most for Science, which was most strongly affected bu the mistake.
+
+@
+
+Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions for each of the partitioning approaches, separated by sample size\footnote{LongRPart trees comprising random intercepts and slopes ran into estimation errors in some datasets: In the Math ability data, \Sexpr{table(is.na(math_MSEs$longRPart_s))["TRUE"]} out of 200 datasets did not yield a tree.
 All tree estimations converged successfully in the Reading ability data.
-In the Science ability data, 40 out of 200 datasets did not yield a tree.
+In the Science ability data, \Sexpr{table(is.na(scie_MSEs$longRPart_s))["TRUE"]} out of 200 datasets did not yield a tree.
 Results are presented for iterations did yield a tree.}.
@@
 
+<<eval=FALSE, echo=FALSE>>=
+math_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = mathdata)
+(as.data.frame(VarCorr(math_LMM))$vcov / var_math)
+(as.data.frame(VarCorr(math_LMM))$vcov / as.data.frame(VarCorr(math_LMM))$vcov[4])
+
+read_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = readdata)
+(as.data.frame(VarCorr(read_LMM))$vcov / var_read)
+(as.data.frame(VarCorr(read_LMM))$vcov / as.data.frame(VarCorr(read_LMM))$vcov[4])
+
+scie_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = sciedata)
+(as.data.frame(VarCorr(scie_LMM))$vcov / var_scie)
+(as.data.frame(VarCorr(scie_LMM))$vcov / as.data.frame(VarCorr(scie_LMM))$vcov[4])
+@
+
 \begin{figure}[!b]
@@
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-028}
+<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+## Tree size
+load("sizes, N = 250, Math.Rda")
+math_sizes <- cbind(sizes, N = 250)
+load("sizes, N = 1000, Math.Rda")
+math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
+#sapply(math_sizes, function(x) table(is.na(x)))
+#sapply(math_sizes, function(x) tapply(x, math_sizes$N, mean, na.rm =TRUE))
+
+load("sizes, N = 250, Reading.Rda")
+read_sizes <- cbind(sizes, N = 250)
+load("sizes, N = 1000, Reading.Rda")
+read_sizes <- rbind(read_sizes, cbind(sizes, N = 1000))
+#sapply(read_sizes, function(x) table(is.na(x)))
+#sapply(read_sizes, function(x) tapply(x, read_sizes$N, mean, na.rm =TRUE))
+
+load("sizes, N = 250, Science.Rda")
+scie_sizes <- cbind(sizes, N = 250)
+load("sizes, N = 1000, Science.Rda")
+scie_sizes <- rbind(scie_sizes, cbind(sizes, N = 1000))
+#sapply(scie_sizes, function(x) table(is.na(x)))
+#sapply(scie_sizes, function(x) tapply(x, scie_sizes$N, mean, na.rm =TRUE))
+
+## Made a mistake in evaluating tree size in the Application scripts:
+## I took #splits = nrow(tree$frame)-1, assuming tree$frame contains a row for
+## each terminal node. It contains a row for each inner and terminal node.
+## Thus, #splits = (nrow(tree$frame)-1)/2; original number should be divided by 2:
+math_sizes$longRPart <- math_sizes$longRPart/2 
+math_sizes$longRPart_s <- math_sizes$longRPart_s/2 
+read_sizes$longRPart <- read_sizes$longRPart/2 
+read_sizes$longRPart_s <- read_sizes$longRPart_s/2 
+scie_sizes$longRPart <- scie_sizes$longRPart/2 
+scie_sizes$longRPart_s <- scie_sizes$longRPart_s/2
+
+tmp6 <- data.frame(stack(math_sizes[ , -c(4,6,7,10)]),
+                    dataset_id = factor(rep(1:nrow(math_sizes), times = 12)),
+                   N = math_sizes$N)
+names(tmp6)[1:2] <- c("splits", "method") 
+tmp6$method2 <- as.character(tmp6$method)
+tmp6$method2[grepl("LMtree", tmp6$method2)] <- "LM tree"
+tmp6$method2[tmp6$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
+tmp6$method2[tmp6$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
+tmp6$method2[tmp6$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
+tmp6$method2[grepl("longR", tmp6$method2)] <- "longRPart"
+tmp6$random <- ifelse(tmp6$method == "LMtree_c", "no", ifelse(
+  tmp6$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+  ifelse(tmp6$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+    "int. + \nslope", NA)))
+tmp6$N <- factor(tmp6$N)
+ggplot(tmp6) +
+  geom_boxplot(aes(x=random, y=splits, fill = N), 
+               position=position_dodge(1), alpha = .5, width = .6) +
+  facet_grid(~method2, scales = "free", space = "free") +
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(x = "", y = "# of splits")
+@
 \end{subfigure}
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-029}
+<<echo=FALSE, fig=TRUE, height=1.6, width=8>>=
+tmp7 <- data.frame(stack(read_sizes[ , -c(4,6,7,10)]),
+                    dataset_id = factor(rep(1:nrow(read_sizes), times = 12)),
+                   N = read_sizes$N)
+names(tmp7)[1:2] <- c("splits", "method") 
+tmp7$method2 <- as.character(tmp7$method)
+tmp7$method2[grepl("LMtree", tmp7$method2)] <- "LM tree"
+tmp7$method2[tmp7$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
+tmp7$method2[tmp7$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
+tmp7$method2[tmp7$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
+tmp7$method2[grepl("longR", tmp7$method2)] <- "longRPart"
+tmp7$random <- ifelse(tmp7$method == "LMtree_c", "no", ifelse(
+  tmp7$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+  ifelse(tmp7$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+    "int. + \nslope", NA)))
+tmp7$N <- factor(tmp7$N)
+ggplot(tmp7) +
+  geom_boxplot(aes(x=random, y=splits, fill=N), 
+               position=position_dodge(1), alpha = .5, width = .6) +
+  facet_grid(~method2, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits") +
+  theme(strip.background = element_blank(), strip.text.x = element_blank(),
+        axis.title.x=element_blank(), axis.text.x=element_blank())
+@
 \end{subfigure}
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-030}
+<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+tmp8 <- data.frame(stack(scie_sizes[ , -c(4,6,7,10)]),
+                    dataset_id = factor(rep(1:nrow(scie_sizes), times = 12)),
+                   N = scie_sizes$N)
+names(tmp8)[1:2] <- c("splits", "method") 
+tmp8$method2 <- as.character(tmp8$method)
+tmp8$method2[grepl("LMtree", tmp8$method2)] <- "LM tree"
+tmp8$method2[tmp8$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
+tmp8$method2[tmp8$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
+tmp8$method2[tmp8$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
+tmp8$method2[grepl("longR", tmp8$method2)] <- "longRPart"
+tmp8$random <- ifelse(tmp8$method == "LMtree_c", "no", ifelse(
+  tmp8$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+  ifelse(tmp8$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+    "int. + \nslope", NA)))
+tmp8$N <- factor(tmp8$N)
+ggplot(tmp8) +
+  geom_boxplot(aes(x=random, y=splits, fill=N), 
+               position=position_dodge(1), alpha = .5, width = .6) +
+  facet_grid(~method2, scales = "free", space = "free") +
+  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(x = "", y = "# of splits")
+@
 \end{subfigure}
@@
 \begin{subfigure}{.5\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-031}
+<<echo=FALSE, fig=TRUE, height=3, width=3>>=
+load("times, N = 250, Reading.Rda")
+read_times <- cbind(times, N = 250)
+load("times, N = 1000, Reading.Rda")
+read_times <- rbind(read_times, cbind(times, N = 1000))
+
+load("times, N = 250, Math.Rda")
+math_times <- cbind(times, N = 250)
+load("times, N = 1000, Math.Rda")
+math_times <- rbind(math_times, cbind(times, N = 1000))
+
+load("times, N = 250, Science.Rda")
+scie_times <- cbind(times, N = 250)
+#load("times, N = 1000, Science.Rda")
+#scie_times <- rbind(scie_times, cbind(times, N = 1000))
+
+## Combine all LMM trees
+ECLSK_times <- rbind(stack(math_times[ , -c(6, 10)]), 
+                     stack(read_times[ , -c(6, 10)]),
+                     stack(scie_times[ , -c(6, 10)]))
+ECLSK_times$ind <- as.character(ECLSK_times$ind)
+ECLSK_times$ind[grepl("LMMtree", ECLSK_times$ind)]<- "LMM tree"
+ECLSK_times$ind[grepl("longRPart", ECLSK_times$ind)]<- "longRPart"
+ECLSK_times$ind[ECLSK_times$ind == "LMtree_c"] <- "LM tree"
+ECLSK_times$ind <- factor(ECLSK_times$ind)
+ggplot(ECLSK_times) +
+  geom_boxplot(aes(x=ind, y=values), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = "log", 
+                     breaks = c(.01, .1, 1, 10, 100, 1000, 10000)) +
+  labs(x = "", y = "Computation time (in seconds)")
+mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
+@
 \end{subfigure}
@@
 Figure~\ref{fig:application_times} depicts the computation time distributions of the different partitioning algorithms.
-A clear computational advantage is observed for LM trees, with an average computation time of 0.47 seconds.
-Including estimation of random effects increases computation time, which was 5.62 seconds on average for LMM trees.
-LongRPart trees required longest computation times, with an average of 364.06 seconds.
+A clear computational advantage is observed for LM trees, with an average computation time of \Sexpr{round(mean_times["LM tree"], digits = 2)} seconds.
+Including estimation of random effects increases computation time, which was \Sexpr{round(mean_times["LMM tree"], digits = 2)} seconds on average for LMM trees.
+LongRPart trees required longest computation times, with an average of \Sexpr{round(mean_times["longRPart"], digits = 2)} seconds.
 
@@
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-032}
+<<echo=FALSE, fig=TRUE, height=1.7>>=
+theme_set(theme_gray(base_size = 8))
+cols <- rep(rainbow_hcl(10)[c(3,5)], times = 5)
+ggplot(tmp[LMM_ids, ]) +
+  geom_boxplot(aes(x=setting, y=tree_size, fill = sigma_int), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[0]))) + 
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(fill=expression(sigma[b[0]]^2))
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-033}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp[LMM_ids, ]) +
+  geom_boxplot(aes(x=setting, y=tree_size, fill = N), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(),
+                     breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits") +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank())
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-034}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp[LMM_ids, ]) +
+  geom_boxplot(aes(x=setting, y=tree_size, fill = p_noise), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(),
+                     breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col = expression(p[noise])) +
+  geom_hline(yintercept=3, col = "darkgrey") +
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(p[noise]))
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-035}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp[LMM_ids, ]) +
+  geom_boxplot(aes(x=setting, y=tree_size, fill = sigma_slope), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(),
+                     breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[1]))) +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(sigma[b[1]]^2))
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-036}
+<<echo=FALSE, fig=TRUE, height=1.8>>=
+ggplot(tmp[LMM_ids, ]) +
+  geom_boxplot(aes(x=setting, y=tree_size, fill = rho), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[1]))) +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(rho))
+theme_set(theme_gray(base_size = 11))
+@
 \end{subfigure}
@@
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-037}
+<<echo=FALSE, fig=TRUE, height=1.7>>=
+theme_set(theme_gray(base_size = 8))
+ggplot(tmp2) +
+  geom_boxplot(aes(x=random, y=tree_size, fill = N), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col=expression(N)) + 
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(fill=expression(N))
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-038}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2) +
+  geom_boxplot(aes(x=random, y=tree_size, fill = sigma_int), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[0]))) +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(sigma[b[0]]^2))
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-039}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2) +
+  geom_boxplot(aes(x=random, y=tree_size, fill = p_noise), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col = expression(p[noise])) +
+  geom_hline(yintercept=3, col = "darkgrey") +
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(p[noise]))
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-040}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2) +
+  geom_boxplot(aes(x=random, y=tree_size, fill = rho), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col=expression(rho)) +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(rho))
+@
 \end{subfigure}
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{_Partitioning_GCMs_with_GLMM_trees-041}
+<<echo=FALSE, fig=TRUE, height=1.8>>=
+ggplot(tmp2) +
+  geom_boxplot(aes(x=random, y=tree_size, fill = sigma_slope), 
+               position=position_dodge(1), alpha = .5, width = .6) + 
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "# of splits", col=expression(sigma[b[1]]^2)) +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(sigma[b[1]]^2))
+theme_set(theme_gray(base_size = 11))
+@
 \end{subfigure}
@@
 
+<<eval=FALSE, echo=FALSE, fig.height=4, fig.width=4>>=
+library("MASS")
+library("glmertree")
+gt1 <- glmertree(y ~ period | subject | base + age + trt, family = poisson,
+                data = epil)
+plot(gt1, "tree", fitted = "marginal")
+gt2 <- glmertree(y ~ period | subject | base + age + trt, family = poisson,
+                data = epil, ranefstart = TRUE)
+plot(gt2, "tree", fitted = "marginal")
+gt3 <- glmertree(y ~ period | subject | base + age + trt, family = poisson,
+                data = epil, cluster = subject)
+plot(gt3, "tree", fitted = "marginal")
+
+library("MASS")
+library("glmertree")
+gt1 <- glmtree(y ~ period | base + age + trt, family = poisson,
+                data = epil)
+plot(gt1, "tree", fitted = "marginal")
+gt2 <- glmtree(y ~ period | base + age + trt, family = poisson,
+                data = epil, ranefstart = TRUE)
+plot(gt2, "tree", fitted = "marginal")
+gt3 <- glmtree(y ~ period | base + age + trt, family = poisson,
+                data = epil, cluster = subject)
+plot(gt3, "tree", fitted = "marginal")
+
+@
+
 \cite{Lee05} analyzed RCT data to compare a test treatment and placebo for a respiratory disorder.
@@
 The data is available as the \verb|resp| dataset in package sanon.
+
+<<eval=FALSE, echo=FALSE, fig.height=4, fig.width=4>>=
+library("sanon")
+data("resp")
+resp$treatment <- factor(resp$treatment)
+resp$sex <- factor(resp$sex)
+Resp <- data.frame(y = with(resp, c(baseline, visit1, visit2, visit3, visit4)),
+                   time = rep(0:4, each = nrow(resp)), age = resp$age, 
+                   center = factor(resp$center), 
+                   treatment = factor(resp$treatment),
+                   sex = factor(resp$sex), subject = 1:nrow(resp))
+gt1 <- lmertree(y ~ time | subject | age + treatment + center + sex,
+                data = Resp)
+plot(gt1, "tree", fitted = "marginal")
+gt2 <- lmertree(y ~ time | subject | age + treatment + center + sex,
+                data = Resp, cluster = subject)
+plot(gt2, "tree", fitted = "marginal")
+gt3 <- lmertree(y ~ time | subject | age + treatment + center + sex,
+                data = Resp, ranefstart = TRUE)
+plot(gt3, "tree", fitted = "marginal")
+@
+
+<<eval=FALSE, echo=FALSE>>=
+library("semtree")
+library("lavaan")
+
+lav_mod <- '
+  intercept =~ 1*baseline + 1*visit1 + 1*visit2 + 1*visit3 + 1*visit4
+  slope =~ visit1 + 2*visit2 + 3*visit3 + 4*visit4
+'
+lav_fit <- growth(lav_mod, data = resp, do.fit = TRUE)
+tree_data <- resp[ , c("baseline", "visit1", "visit2", "visit3", 
+                       "visit4", "age", "center", "treatment", "sex")]
+st1 <- semtree(lav_fit, data = resp, 
+              predictors = c("age", "center", "treatment", "sex"))
+st1
+st2 <- semtree(lav_fit, data = resp, 
+              predictors = c("age", "center", "treatment", "sex"), 
+              control = semtree.control(method = "fair"))
+st2
+st3 <- semtree(lav_fit, data = resp, 
+              predictors = c("age", "center", "treatment", "sex"), 
+              control = semtree.control(method = "fair3"))
+st3
+st4 <- semtree(lav_fit, data = resp, 
+              predictors = c("age", "center", "treatment", "sex"), 
+              control = semtree.control(method = "naive"))
+st4
+st5 <- semtree(lav_fit, data = resp, 
+              predictors = c("age", "center", "treatment", "sex"), 
+              control = semtree.control(method = "cv"))
+st5
+st6 <- semtree(lav_fit, data = resp, 
+              predictors = c("age", "center", "treatment", "sex"), 
+              control = semtree.control(method = "score"))
+st6
+
+library("OpenMx")
+manifests <- c("baseline", "visit1", "visit2", "visit3", "visit4")
+latents <- c("I", "S")
+mx_mod <- mxModel(model = "LGC", type = "RAM", manifestVars = manifests,
+                  latentVars = latents,
+                  mxPath(from = "I", to=manifests, arrows = 1, free = FALSE, values = 1),
+                  mxPath(from = "S", to=manifests, arrows = 1, free = FALSE, values = 0:4),
+                  mxPath(from = latents, to = latents, arrows = 2, free = TRUE, 
+                           values = c(.8, .8), labels = c("VarI", "VarS")),
+                  mxPath(from = "I", to = "S", arrows = 2, free = TRUE, values = .6, 
+                         labels = "Cov_IS"),
+                  mxPath(from = manifests, to = manifests, arrows = 2, free = TRUE, values = .8,
+                         labels = c("base", "v1", "v2", "v3", "v4")),
+                  mxPath(from = "one", to = latents, arrows = 1, free = TRUE, values = .2, 
+                         labels = c("mI", "mS")),
+                  mxData(observed = resp, type = "raw"))
+mx_fit <- mxRun(mx_mod)  
+summary(mx_fit)
+st7 <- semtree(mx_fit, data = resp, predictors = c("age", "center", "treatment", "sex"), 
+               control = semtree.control(method = "score"))
+st7
+
+##undebug(semtree)
+##debug(growTree)
+##undebug(semtree:::naiveSplitScoreTest)
+
+@
 
@@
 
+<<eval=FALSE, echo=FALSE>>=
+## ECLSK are reading scores with minbucket = 500
+# load("predsECLSK")
+# load("tree_sizesECLSK")
+# MSEs <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
+# ses <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
+#          sqrt(sum(!is.na(x))))
+# Table <- data.frame(MSE = MSEs[-13], SE = ses[-13],
+#                     mean_tree_size = sapply(sizes, mean, na.rm = TRUE),
+#                     SD = sapply(sizes, sd, na.rm = TRUE))
+# names(Table) <- c("$MSE$", "$se_{MSE}$", "$M_{tree size}$", "$SD_{tree size}$")
+# rownames(Table) <- c("LMtree", "LMtree_c", "LMMtree", "LMMtree_c", "LMMtree_r",
+#                      "LMMtree_{cr}", "LMMtree_s", "LMMtree_{sc}", "LMMtree_{sr}",
+#                      "LMMtree_{scr}", "longRPart", "longRPart_s")
+# tmp_minbucket <- data.frame(ID = names(MSEs[-13]),
+#                             tree_size = colMeans(sizes, na.rm = TRUE),
+#                             MSE = MSEs[-13])
+@
+
+<<echo=FALSE>>=
+#' Print LaTeX tables with headings over two lines
+#' 
+#' @param .data A data frame with at least some of the column names having a \code{separator}.
+#' @param separator A regular expression that splits the top and bottom rows.
+If the separator is not found in the names of \code{.data}, the function returns an error (saying you should probably just use \code{print.xtable()})
+#' @param align Passed to \code{xtable}: Character vector of length equal to the number of columns of the resulting table, indicating the alignment of the corresponding columns.
+#' @param digits Passed to \code{xtable}
+#' @param label Passed to \code{xtable}
+#' @param caption Passed to \code{xtable}
+#' @param booktabs Should the tabular environment produced use booktabs? Set to TRUE for (my) convenience. This will cause an error if  \verb{\usepackage{booktabs}} has not been called in LaTeX.
+#' @param heading_command A (simple) LaTeX control sequence (properly escaped) to apply to each heading names.
+#' @param ... Arguments passed to \code{print.xtable}. You cannot pass \code{add.to.row}, \code{include.rownames}, or \code{include.colnames} as we make use of these options in this function.  
+#' @return Output intended for LaTeX. A table produced using xtable where groups of column names are put in the top row. 
+#' @author Hugh Parsonage
+#' @examples
+#' example_df <- 
+#' data.frame(yr = 2001:2005, 
+#'           Revenue__foo = 1:5, 
+#'           Revenue__bar = 11:15, 
+#'           Revenue__baz = 21:25, 
+#'           ordinary = 1:5,
+#'           Expense__foo = 1:5,
+#'           Expense__bar = 11:15, 
+#'           Expense__baz = 21:25, 
+#'           Last__foo = 1:5, 
+#'           Last__baz = 2:6,
+#'           last = 101:105)
+#' print_2heading_xtable(example_df, separator = "__")
+#' @export
+print_2heading_xtable <- function(.data, 
+                                  separator = "__", 
+                                  align = NULL, 
+                                  caption = NULL,
+                                  digits = NULL,
+                                  label = NULL,
+                                  booktabs = TRUE, 
+                                  heading_command = "\\textbf", ...){
+  orig_names <- names(.data)
+  if (!any(grepl(separator, orig_names))){
+    stop("No separator found in column names, so there is no point in using this function. Make sure you have specified the right separator; otherwise, just use print.xtable().")
+  }
+
+  if (any(c("add.to.row", "include.colnames", "include.rownames") %in% names(list(...)))){
+    stop("You should not pass add.to.row, include.colnames, or include.rownames to print.xtable() via this function.")
+  }
+
+  split_names <-  grep(separator, orig_names, value = TRUE)
+  split_positions <- grep(separator, orig_names, value = FALSE)
+
+  # get the names before the separator
+  top_headers <- gsub(paste0("^(.*)", separator, ".*$"), "\\1", split_names)
+  # Where in the original table is there a new top header?
+
+  orig_names_no_suffix <- 
+    gsub(paste0("^(.*)", separator, ".*$"), paste0("\\1", separator), orig_names)
+
+  # For cmidrule{}
+  position_of_header_instance <- 
+    # Need to test first column
+    which(orig_names_no_suffix == dplyr::lead(orig_names_no_suffix) & 
+            (orig_names_no_suffix != dplyr::lag(orig_names_no_suffix) | is.na(dplyr::lag(orig_names_no_suffix))))
+
+  position_of_header_final <- 
+    # Need to test final column
+    which((orig_names_no_suffix != dplyr::lead(orig_names_no_suffix) | is.na(dplyr::lead(orig_names_no_suffix))) &
+            orig_names_no_suffix == dplyr::lag(orig_names_no_suffix))
+
+  if (length(position_of_header_instance) != length(position_of_header_final)){
+    stop("This is a bug. Sorry. Please provide your data frame to the grattan package maintainer.")
+  }
+
+  double_row_column_names <- 
+    rbind(gsub("^(.*)__(.*)$", "\\1", orig_names), gsub("^(.*)__(.*)$", "\\2", orig_names))
+
+  # factor etc in table to preserve order
+  top_headers_widths <- 
+    as.data.frame(table(factor(double_row_column_names[1,], levels = unique(double_row_column_names[1,]))))
+
+  first_row <- 
+    unique(double_row_column_names[1,])
+
+  first_row_formatted <- 
+    paste0(heading_command, "{", first_row, "}")
+
+  top_row <- character(length(first_row))
+
+  # Could do paste0() directly but decided that it would 
+  # avoid the point which is to add \multicolumn only to the rows that call for it.
+  for (ii in seq_along(first_row)){
+    if (first_row[ii] %in% top_headers){
+      top_row[ii] <- paste0("\\multicolumn{", top_headers_widths$Freq[ii], "}{c}{", first_row_formatted[ii], "}")
+    }
+  }
+  rm(ii)
+
+  for_latex_top_row <- 
+    paste0(paste0(top_row, collapse = " & "), "\\\\")
+
+  if (booktabs){
+    # (lr) to avoid cmidrule touching adjacent groups
+    between_row <- paste0("\\cmidrule(lr){",  position_of_header_instance, "-", position_of_header_final, "}")
+  } else {
+    between_row <- paste0("\\cline{",  position_of_header_instance, "-", position_of_header_final, "}")
+  }
+  for_latex_between_row <- 
+    paste0(paste0(between_row, collapse = ""))
+
+  for_latex_second_row <- 
+    paste0(heading_command, "{", double_row_column_names[2,], "}")
+
+  for_latex_second_row <- 
+    paste0(paste0(for_latex_second_row, collapse = " & "), "\\\\")
+
+  addtorow <- list()
+  addtorow$pos <- list(0, 0, 0)
+  addtorow$command <- 
+    paste0(paste0(c(for_latex_top_row, for_latex_between_row, for_latex_second_row)), "\n")
+
+  xtable::print.xtable(xtable::xtable(.data, align = align, digits = digits, 
+                                      label = label, caption = caption), 
+                       type = "latex",
+                       add.to.row = addtorow, 
+                       include.colnames = FALSE, 
+                       include.rownames = FALSE,
+                       booktabs = booktabs,
+                       ...)
+}
+@
+
+<<echo=FALSE, eval=FALSE, results=tex>>=
+## ECLSK2 are reading scores with maxdepth = 4
+load("predsECLSK2")
+var_reading <- round(var(preds$observed), digits = 3)
+MSEs_read <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
+ses_read <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
+         sqrt(sum(!is.na(x))))
+load("preds_math")
+var_math <- round(var(preds$observed), digits = 3)
+MSEs_math <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
+ses_math <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
+         sqrt(sum(!is.na(x))))
+load("preds_science")
+var_science <- round(var(preds$observed), digits = 3)
+MSEs_scie <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
+ses_scie <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
+         sqrt(sum(!is.na(x))))
+Table2 <- data.frame(MSEs_read[-13], ses_read[-13],
+                     MSEs_math[-13], ses_math[-13],
+                     MSEs_scie[-13], ses_scie[-13])
+Table2$mean_rank <- 1/3*(rank(Table2$MSEs_read..13.) + rank(Table2$MSEs_math..13.) + 
+  rank(Table2$MSEs_scie..13.))
+names(Table2) <- c("Reading MSE", "SE", "Math MSE", "SE", "Science MSE", "SE", 
+                   "mean rank")
+Names <- cbind(sapply(strsplit(rownames(Table2), "_"), function(x) x[1]),
+                sapply(strsplit(sapply(strsplit(rownames(Table2), "_"), 
+                                       function(x) x[2]), ""), 
+                       function(x) paste(x, collapse = ", ")))
+Names[,2][Names[,2] == "NA"] <- ""
+Names[3:12, 2] <- paste("i,", Names[3:12, 2])
+
+Table3 <- cbind(Algorithm = paste(Names[,1], Names[,2]),
+                Table2, stringsAsFactors = FALSE)
+
+names(Table3) <- c("Algorithm",  "Reading__MSE", "Reading__SE", "Math__MSE", "Math__SE", 
+                   "Science__MSE", "Science__SE", "  __mean\nrank")
+
+print_2heading_xtable(Table3, separator = "__", 
+                      digits = c(0, 0, 4, 3, 4, 3, 4, 3, 1), 
+                      align = c("r", "l", rep("c", times = 7)), 
+                      caption = "Predictive accuracy for the ECLSK data.
+c indicates that parameter stabiliy tests were performed at the cluster level; r indicates that estimation was initialized with the random effects; i indicates that random intercepts were estimated; s indicates that random slopes were estimated.",
+                      label = "tab:ECLSK_accuracy",
+                      hline.after = c(-1, 0, nrow(Table3)))
+@
+
+<<echo=FALSE, eval=FALSE, results=tex>>=
+load("tree_sizesECLSK2")
+failed_longRpart_count <- sum(is.na(sizes$longRPart))
+failed_longRpart_s_count <- sum(is.na(sizes$longRPart_s))
+tree_sizes <- data.frame(
+  tree_size = colMeans(sizes, na.rm = TRUE),
+  SD = sapply(sizes, sd, na.rm = TRUE))
+load("sizes_math")
+failed_longRpart_count <- sum(failed_longRpart_count, sum(is.na(sizes$longRPart)))
+failed_longRpart_s_count <- sum(failed_longRpart_s_count, sum(is.na(sizes$longRPart_s)))
+tree_sizes <- cbind(tree_sizes,
+                    tree_size = colMeans(sizes, na.rm=TRUE),
+                    SD = sapply(sizes, sd, na.rm=TRUE))
+load("sizes_science")
+failed_longRpart_count <- sum(failed_longRpart_count, sum(is.na(sizes$longRPart)))
+failed_longRpart_s_count <- sum(failed_longRpart_s_count, sum(is.na(sizes$longRPart_s)))
+tree_sizes <- cbind(tree_sizes,
+                    tree_size = colMeans(sizes, na.rm=TRUE),
+                    SD = sapply(sizes, sd, na.rm=TRUE))
+tree_sizes$mean_rank <- 1/3*(rank(tree_sizes[,1]) + rank(tree_sizes[,3]) + rank(tree_sizes[,5]))
+
+tree_sizes <- cbind(Algorithm = paste(Names[,1], Names[,2]), tree_sizes, 
+                    stringsAsFactors = FALSE)
+names(tree_sizes) <- c("Algorithm",  "Reading__M", "Reading__SD", "Math__M", "Math__SD", 
+                   "Science__M", "Science__SD", "  __mean\nrank")
+
+print_2heading_xtable(tree_sizes, separator = "__", 
+                      digits = c(0, 0, 1, 1, 1, 1, 1, 1, 1), 
+                      align = c("r", "l", rep("c", times = 7)), 
+                      caption = "Tree sizes for the ECLSK data.
+c indicates that parameter stabiliy tests were performed at the cluster level; r indicates that estimation was initialized with the random effects; i indicates that random intercepts were estimated; s indicates that random slopes were estimated.",
+                      label = "tab:ECLSK_complexity",
+                      hline.after = c(-1, 0, nrow(tree_sizes)))
+@
+
 %Table~\ref{tab:ECLSK_accuracy} shows the cross-validated predictive accuracies for all RPMs.
@@
 \begin{figure}
+<<echo=FALSE, eval=FALSE, fig=TRUE>>=
+load("comp_timesECLSK2") ## Reading, maxdepth = 4
+times2 <- list(reading = times)
+load("times_math") 
+times2$math <- times
+load("times_science") 
+times2$science <- times
+times <- do.call(rbind, times2)
+comp_time <- reshape2::melt(times)
+p <- ggplot(comp_time, aes(variable, value))  
+p + geom_boxplot() + labs(x = "", y = "Computation time (in seconds)") +
+  scale_y_continuous(trans = "log", breaks = c(.1, 1, 10, 100, 1000, 10000)) +
+  theme(axis.text.x = element_text(angle = 90))
+@ 
 \caption{Computation time distributions for the different partitioning methods.}

### 2022-11-11 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.Rnw
improvements in abstract and intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3231 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ea3f202bd8f6cc54295e8a79410a244426cf5b26

@@
 \title{Recursive partitioning of linear growth curve models}
-\shorttitle{Partitioning linear trajectories with GLMM trees}
+\shorttitle{Recursive partitioning of linear growth curve models}
 \authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
@@
 
-\abstract{Growth curve models are popular tools for studying development over time.
+\abstract{Growth curve models are popular tools for studying {\color{red!80} the development of a response variable within the same observational units} over time.
 Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time.
 Recursive partitioning methods (RPMs) allow for detecting such subgroups.
-In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning LGCMs.
+In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning linear growth curve models (LGCMs).
 GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data.
-Here, we introduce recent adaptations of the algorithm that that are of particular relevance for partitioning longitudinal models, and growth curve models in particulat.
-Using simulated and existing data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart.
+{\color{red!80} Here, we introduce an extension of the algorithm for longitudinal (as opposed to clustered) data that can detect subgroups defined by subject-specific covariates on/or time-varying covariates.
+The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.} Using simulated and empirical data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart.
 We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart.
@@
 
-In longitudinal studies, (generalized) linear growth curve modeling is often used to model change over time, and inter individual differences in change.
+In longitudinal studies, (generalized) linear growth curve modeling is often used to model change over time, and inter-individual differences in change.
 For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail.
-Inter individual heterogeneity is likely in such studies, where (groups of) participants show higher or lower rates of growth over time.
+Inter-individual heterogeneity is likely in such studies, where (groups of) participants show higher or lower rates of growth over time.
 Through the use of mixed-effects or latent-variable models, such heterogeneity can be captured, and explained by covariates of a-priori known relevance \citep{NeisyMatt18}.
@@
 Recursive partitioning methods (RPMs), also known as decision-tree methods, are particularly suited in such cases.
-Several existing RPMs allow for partitioning growth trajectories: For example, GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, Structural Equation Model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
+Several existing RPMs allow for partitioning growth trajectories: For example, GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, structural equation model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
 Partitioning based on non-linear growth curve models can be performed with the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods\footnote{Both IT methods specifically focus at detecting so-called \textit{treatment-subgroup interactions}: subgroups that respond more favorably to one treatment than another.}.
@@
 In this paper, we focus on generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}, which allow for subgroup detection in a wide range of mixed-effects models.
-Although originally not specifically developed for partitioning LGCMs, application of GLMM trees for partitioning LGCMs is straightforward.
+{\color{red!80} FIXME: The rest of this paragraph needs to be a bit more forceful!} Although originally not specifically developed for partitioning LGCMs, application of GLMM trees for partitioning LGCMs is straightforward.
 We propose two adaptations to the algorithm, which we expect to be of particular relevance for partitioning LGCMs.
@@
 This may not only reduce computation time; it also reduces the total number of parameters to be estimated, which may improve stability and reduce overfitting.
-This local/global estimation approach was originally proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes; \cite{FokkySmit18} generalized it to model-based trees.
+This local/global estimation approach was originally proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant means in the terminal nodes; \cite{FokkySmit18} generalized it to trees containing generalized linear models in the nodes, i.e., potentially including regressors and/or using a non-Gaussian family.
 
@@
 LongCART, MELT, GEE-based decision trees and score-based SEM trees fit full parametric models in each of the nodes as well, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split.
-This reduces computational burden.
+This reduces the computational burden.
 Furthermore, these approaches allow for employing a two-step approach for split detection, which separates variable and cutpoint selection, thus preventing variable selection bias.
@@
 In Study III, we compare the performance of GLMM trees and longRPart in partitioning children's trajectories of reading, math and science abilities.
-In the Discussion, we summarize our findings and discuss implications.
+In the Discussion, we summarize our findings and their implications.
 

### 2022-11-11 pkg/glmertree/paper/glmm-tree/treatment_interactions.Rnw
moved first paper to glmm-tree folder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3235 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b616a7442d9e5ce4d2444cd3cb972adaa86cf16e

--- 2017-05-23 pkg/glmertree/paper/treatment_interactions.Rnw
@@
 \\
-\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees
+\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, mixed-effects models, classification and regression trees
 
@@
 Several tree-based methods take the mean of a continuous dependent variable or the majority class of a categorical dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}.
-Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM approach of \citeA{Ciam91} is the earliest example.
+Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM (recursive partition and amalgamation) approach of \citeA{Ciam91} is the earliest example.
 
@@
 Each of the splits in the tree is defined by a splitting variable and value.
-The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3.
-As the observations in node~2 are not further split, node~2 is a terminal node.
-Node~3 is an inner node, as the observations in this node are further split into nodes~4 and~5, based on the anxiety variable.
+The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of eight, yielding two rectangular regions, represented by node~2 and node~5.
+Node~2 is an inner node, as the observations in this node are further split into terminal nodes~3 and~4, based on the anxiety variable.
+The observations in node~5 are not further split and this is therefore a terminal node.
 
@@
 }
-\caption{\small Example of a recursive partition. In the left panel, the partition is depicted as a set of rectangular areas. In the right panel, the same partition is depicted as a decision tree.}
+\caption{\small Example recursive partition defined by two partitioning variables (duration and anxiety).
+In the left panel, the partition is depicted as a set of rectangular areas.
+In the right panel, the same partition is depicted as a tree.}
 \label{fig:exampletree}
@@
 In the fourth and last section we summarize the results and discuss limitations and directions for future research.
-A tutorial on how to fit GLMM trees using the R package \verb|glmertree| is included as supplementary material.
+In the Appendix, we provide a glossary explaining abbreviations and mathematical notation used in the current paper.
+Finally, a tutorial on how to fit GLMM trees using the R package \verb|glmertree| is included as supplementary material.
 In the tutorial, the artificial motivating dataset is used, allowing users to recreate the trees and models to be fitted in the next section.
@@
 \thickline
- & min & max & mean & sd\\
+ & min & max & $M$ & $SD$\\
 \hline
 depression & 3 & 16 & 9.12 & 2.66 \\
-age & 18 & 69 & 45 & 9.56 \\
+age & 18 & 69 & 45.00 & 9.56 \\
 anxeity & 3 & 18 &  10.26 & 3.05 \\
@@
 @
-\caption{\small Example of a global GLM for treatment outcomes, based on the artificial motivating dataset ($N=150$).
-The y-axis represent treatment outcome (post-treatment depression).
-The dot for Treatment~1 represents the first and the slope of the regression line represents the second element of $\beta$.}
+\caption{\small Example of a globally estimated GLM based on the artificial motivating dataset ($N=150$).
+The $x$-axis represents treatment group, the $y$-axis represents treatment outcome (post-treatment depression).
+The dot for Treatment~1 represents the first element of the fixed-effects coefficient vector $\beta$, the slope of the regression line represents the second element of $\beta$.}
 \label{fig:fixedeffects}
@@
 \caption{\small GLM tree grown on the artificial motivating dataset.
-Three additional covariates (pre-treatment anxiety, duration and age) were used as potential splitting variables.
-The y-axes in the terminal nodes represent the treatment outcome (post-treatment depression severity).}
+The $x$-axes in the terminal nodes represent the treatment group, the $y$-axes represent the treatment outcome (post-treatment depression).
+Three additional covariates (pre-treatment anxiety, duration and age) were used as potential splitting variables, of which two (duration and age) were selected.}
 \label{fig:example_mobtree}
@@
 The partition resulting from application of MOB can be depicted as a decision tree.
-If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
+If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node:
 %
@@
 %
-In the GLMM tree model, the fixed effects $\beta_j$ are local parameters, their value depending on subgroup $j$, but the random effects $b$ are global.
+In the GLMM tree model, the fixed effects $\beta_j$ are local parameters, their value depending on terminal node $j$, but the random effects $b$ are global.
 To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}.
@@
 \caption{\small GLMM tree of the motivating example dataset.
+The $x$-axes represent the treatment group, the $y$-axes represent the treatment outcome (post-treatment depression).
 Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months) and age were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
@@
 
-  \item[Step 2:] Fit the mixed-effects model $g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b$ with subgroups $j(r)$ from the GLM tree estimated in Step~1.
+  \item[Step 2:] Fit the mixed-effects model $g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b$ with terminal node $j(r)$ from the GLM tree estimated in Step~1.
 Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
@@
 \item Number of clusters: $M=5$, $M=10$, $M=25$.
-\item Population standard deviation (SD) of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
+\item Population standard deviation (SD) of the normal distribution from which the cluster-specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
 \item Intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
@@
 Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$.
-Means for other potential partitioning covariates were drawn from a discrete uniform distribution on the interval $[-70,70]$.
+The means of the other potential partitioning covariates ($U_3$ and, depending on the value of $K$, also $U_6$ through $U_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$.
 All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$.
@@
 The significance level $\alpha$ for the parameter instability tests was set to .05 for all trees, with a Bonferroni correction applied for multiple testing.
+The latter adjusts the $p$-values of the parameter stability tests, by multiplying these by the number of potential partitioning variables.
 The minimum number of observations per node in trees was set to 20 and maximum tree depth was set to three, thus limiting the number of terminal nodes to eight in every tree.
@@
 \caption{\small Data-generating model for treatment-subgroup interactions.
-Parameter $d_j$ denotes the node-specific standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
+$U_1$, $U_2$ and $U_5$ represent the (true) partitioning variables; $\beta$ represents the (true) fixed effects, with $\beta_{j0}$ representing the intercept and $\beta_{j1}$ representing the slope; parameter $d_j$ denotes the node-specific standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
 \label{fig:modelC}
@@
 For generating datasets with treatment-subgroup interactions, we used a design from \citeA{DussyMech14} which is depicted in Figure~\ref{fig:modelC}.
-Figure~\ref{fig:modelC} shows four subgroups, characterized by values of the partitioning variables $U_2$, and $U_1$ or $U_5$.
+Figure~\ref{fig:modelC} shows four terminal subgroups, characterized by values of the partitioning variables $U_2$, and $U_1$ or $U_5$.
 Two of the subgroups have mean differences in treatment outcome, indicated by a non-zero value of $\beta_{j1}$, and two subgroups do not have mean differences in treatment outcome, indicated by a $\beta_{j1}$ value of 0.
@@
 \caption{\small Average tree size of LM and LMM trees.
-Tree size is defined as the total number of nodes in a tree.
-Rows represent the correlation between the random intercepts and one of the partitioning variables, columns represent sample size.
-The horizontal line in each graph indicates the 'true' tree size (the total number of nodes in the tree used for generating the interactions).}
+The $y$-axes represent the average number of nodes in trees; $x$-axes represent the population standard deviation of the normal distribution from which the cluster-specific intercepts were drawn; rows represent the correlation between the random intercepts and one of the partitioning variables; columns represent sample size.
+The horizontal line in each graph indicates the 'true' tree size (total number of nodes in the tree used for generating the interactions).}
 \label{fig:xyplot_treesize}
@@
 @
-\caption{\small Tree accuracy of LM and LMM tree.
-Tree accuracy is defined as the proportion of datasets in which the true tree was accurately recovered.
-Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$, columns represent sample size.}
+\caption{\small Tree accuracy of LM and LMM trees in simulated datasets.
+The $y$-axes represent the proportion of datasets in which the true tree was accurately recovered; $x$-axes represent the population standard deviation of the normal distribution from which the cluster-specific intercepts were drawn; rows represent dependence between the random intercepts ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
 \label{fig:xyplot_treeaccuracy}
@@
 @
-\caption{\small Average predictive accuracy of LM and LMM trees.
-Predictive accuracy is defined as the correlation between the true and predicted treatment-effect differences.
-Rows represent absolute treatment-effect differences in subgroups with treatment-effect differences, columns represent sample size.}
+\caption{\small Average predictive accuracy of LM and LMM trees in simulated datasets.
+The $y$-axes represent the average correlation between the true and predicted treatment-effect differences; $x$-axes represent the population standard deviation of the normal distribution from which the cluster-specific intercepts were drawn; rows represent absolute treatment-effect differences in subgroups with treatment-effect differences; columns represent sample size.}
 \label{fig:xyplot_correlations}
@@
 @
-\caption{\small Type-I error rate of LM and LMM trees. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
+\caption{\small Type-I error rate of LM and LMM trees in simulated datasets.
+The $y$-axes represent the average Type-I error (or false positive) rate, $x$-axes represent the population standard deviation of the normal distribution from which the cluster-specific intercepts were drawn represent; rows represent dependence between random intercepts ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
 \label{fig:xyplot_treesize_nointeract}
@@
 \caption{\small Fixed-effects terms in simulations with continuous and combined continuous and piecewise interaction designs.}
-\begin{tabular}{lcccccc}
+\begin{tabular}{lccccccc}
 \thickline
-Term					&	$\beta_3$&	$\beta_4$		&	$\beta_6$&	$\beta_7$		&&	$\beta$		\\ \hline
-intercept				&	27		&	27		&	27		&	27		&&	27		\\
-$U_2$					&	.1		&	.1		&	.1		&	.1		&&	.1		\\
-$U_2 \cdot U_1$ 			&	$-$.357		&	0		&	0		&	0		&&	$-$.357		\\
-$U_2 \cdot U_5$	 			&	0		&	0		&	0		&	.357		&&	.357		\\
-$U_2 \cdot U_1 \cdot \mbox{treatment}$	&	$-$.151		&	0		&	0		&	0		&&	$-$.151		\\
-$U_2 \cdot U_5 \cdot \mbox{treatment}$	&	0		&	0		&	0		&	.151		&&	.151		\\ \hline
-\multicolumn{7}{l}{\footnotesize{\emph{Note:} Subscripted $\beta$ values refer to the terminal nodes}}\\	
-\multicolumn{7}{l}{\footnotesize in Figure~\ref{fig:modelC} for the combined piecewise and continuous}\\
-\multicolumn{7}{l}{\footnotesize interaction design; $\beta$ without subscript refers to the global}\\
-\multicolumn{7}{l}{\footnotesize coefficients in the continuous interaction design.}\\
+Term					                          &&$\beta$  &&$\beta_3$ &$\beta_4$&$\beta_6$&$\beta_7$	\\ 
+\hline
+intercept				                        && 27  && 27		& 27	&	27  & 27   	\\
+$U_2$					                          &&	 0.100  &&  0.100		&  0.100	&	 0.100	&	 0.100  	\\
+$U_2 \cdot U_1$ 			                  &&$-0.357$  &&$-0.357$	&  0	&	 0	&	 0  	\\
+$U_2 \cdot U_5$	 			                  &&	 0.357  &&  0		&  0	&	 0	&	 0.357   	\\
+$U_2 \cdot U_1 \cdot \mbox{treatment}$	&&$-0.151$  &&$-0.151$  &  0 	&	 0	&	 0	 	\\
+$U_2 \cdot U_5 \cdot \mbox{treatment}$	&&	 0.151  &&  0		&  0	&	 0	&	 0.151	 	\\ 
+\hline
+\multicolumn{8}{l}{\footnotesize{\emph{Note:} All values in the table represent fixed-effects linear regression}}\\
+\multicolumn{8}{l}{\footnotesize{coefficients. The intercept represent the first value of $\beta$ or $\beta_j$, values in}}\\
+\multicolumn{8}{l}{\footnotesize{the rows below represent the slopes for each of the main or interaction}}\\
+\multicolumn{8}{l}{\footnotesize{effect described in the 'Term' column. Subscripted $\beta$ values refer to}}\\
+\multicolumn{8}{l}{\footnotesize{the terminal nodes in Figure~\ref{fig:modelC} for the combined piecewise and}}\\
+\multicolumn{8}{l}{\footnotesize{continuous interaction design; $\beta$ without subscript refer to the global}}\\
+\multicolumn{8}{l}{\footnotesize{coefficients in the continuous interaction design.}}\\
 \end{tabular}
@@
 @
-\caption{\small Average predictive accuracy of LMMs and LMM trees.
-Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2.
-Columns represent sample size, rows represent the number of covariates.}
+\caption{\small Average predictive accuracy of LMMs and LMM trees in simulated datasets.
+$y$-axes represent the correlation between the true and predicted differences between Treatment~1 and~2; $x$-axes represent the type of interaction; rows represent the number of covariates; columns represent sample size.}
 \label{fig:xyplot_correlation_pw}
@@
 @
-\caption{\small LM tree for prediction of treatment outcomes.
-Upper terminal nodes: y-axes represent post-treatment HAM-D scores, x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
-Lower terminal nodes: Subgroup-specific descriptive statistics.}
+\caption{\small LM tree for prediction of treatment outcomes in the IPDMA dataset.
+The selected partitioning variables are level of education and the presence of a comorbid anxiety disorder.
+Upper terminal nodes: $y$-axes represent post-treatment HAM-D scores, $x$-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
+Lower terminal nodes represent subgroup-specific descriptive statistics.}
 \label{fig:lmtree_C&W}
@@
 
-Figure~\ref{fig:lmtree_C&W} and Figure~\ref{fig:lmertree_C&W} present the LM and LMM trees fitted on the IPDMA data.
-The LM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
-By taking into account study-specific intercepts, the LMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split in the LM tree may be spurious.
-The LMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
-The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$).
+Figures~\ref{fig:lmtree_C&W} and \ref{fig:lmertree_C&W} present the LM and LMM trees fitted on the IPDMA data.
+The LM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable.
+By taking into account study-specific intercepts, the LMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split in the LM tree may be spurious and selected presence of a comorbid anxiety disorder as the only partitioning variable.
+The terminal nodes of Figure~\ref{fig:lmertree_C&W} show a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$).
 For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$).
 The estimated intraclass correlation coefficient for the LMM tree was .05.
+
+The study-specific distributions of educational level and treatment outcome may explain why the LMM tree did not select level of education as a partitioning variable.
+Most (55) of the 74 observations with level of education $\leq 1$ were part of a single study, which showed a markedly lower mean level of education ($M = 2.57$, $SD = 1.02$; 128 observations) compared to the other studies ($M = 3.78$, $SD = 0.53$; 566 observations), as well as a markedly higher mean level of post-treatment HAM-D scores ($M = 11.20$, $SD = 6.87$) compared to the other studies ($M = 7.78$, $SD = 5.95$).
+
+%\begin{table}
+%\caption{Means (SDs) of educational level and post-treatment HAM-D scores by study}
+%\begin{tabular}{lcc}
+%\hline
+%&& post-treatment \\
+%study # & education & HAM-D \\
+%\hline
+%1 & 2.76 (0.54) & 8.05 (6.64) \\
+%2 & 2.66 (0.62) & 8.29 (6.30) \\
+%3 & 2.74 (0.52) & 8.01 (6.81) \\
+%4 & 1.57 (1.02) &11.20 (6.87) \\
+%5 & 2.98 (0.13) & 9.03 (5.41) \\
+%6 & 2.79 (0.55) & 6.68 (4.90) \\
+%7 & 2.86 (0.39) & 7.41 (5.81) \\
+%\hline
+%\end{tabular}
+%\end{table}
 
@@
 @
-\caption{\small LMM tree for prediction of treatment outcomes.
-Upper terminal nodes: y-axes represent post-treatment HAM-D scores, x-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
-Lower terminal nodes: Subgroup-specific descriptive statistics.}
+\caption{\small LMM tree for prediction of treatment outcomes in the IPDMA dataset.
+The selected partitioning variable is the presence of a comorbid anxiety disorder.
+Upper terminal nodes: $y$-axes represent post-treatment HAM-D scores, $x$-axes represent treatment levels (cognitive behavior therapy, CBT vs.\ pharmacotherapy, PHA).
+Lower terminal nodes represent subgroup-specific descriptive statistics.}
 \label{fig:lmertree_C&W}
@@
 	\small
-	\caption{\small Variable selection statistics}
+	\caption{\small Variable selection statistics for the LM and LMM trees in the IPDMA dataset}
 	\begin{tabular}{lcc}
@@
 LMM trees performed much better than LMMs when interactions were at least partly piecewise.
-However, the performance ofLMM trees deteriorated when the interactions were purely continuous functions of the predictor variables, in which case LMMs with pre-specified interactions performed very well.
+However, the performance of LMM trees deteriorated when the interactions were purely continuous functions of the predictor variables, in which case LMMs with pre-specified interactions performed very well.
 The performance of LMMs deteriorated with a larger number of predictor variables, whereas the performance of LMM trees was not affected by this, confirming our expectation that LMM trees are better suited for exploration than LMMs.
@@
 For recursive partitioning methods fitting local parametric models only, the danger of model misspecification may be limited \cite<e.g., >{Ciam91, SiciyMass08, SuyWang04}.
-GLMM trees, however, fit a global random-effect model in addition to local fixed-effects regression models and may therefore be less robust against model misspecification.
+GLMM trees, however, fit a global random-effects model in addition to local fixed-effects regression models and may therefore be less robust against model misspecification.
+For example, the global random-effects model assumes a single random-effects variance across terminal nodes.
+If the random-effects variance, however, does vary across terminal nodes, this may negatively affect the performance of the parameter stability tests.
 Further research on the effects of model misspecifications on the performance of GLMM trees is therefore needed.
@@
 
+\begin{appendix}
+
+\section{Glossary}
+
+\subsection{Abbreviations}
+
+\noindent CART: Classification and regression trees.\\
+GLM: Generalized linear model.\\
+GLMM: Generalized linear mixed model.\\
+IPDMA: Individual patient data meta analysis.\\
+LM: Linear model.\\
+LMM: Linear mixed model.\\ 
+MERT: Mixed-effects regression tree.\\
+ML: Maximum likelihood.\\
+MOB: Model-based recursive partitioning.\\
+RECPAM: Recursive partition and amalgamation.\\
+REML: Restricted maximum likelihood.\\
+
+\subsection{Notation}
+
+\NumTabs{12}
+\noindent $\beta_{j}$		\tab\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
+$b_{m}$ 		\tab\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
+$d_{j}$			\tab\tab\tab{standardized mean difference of treatment outcome in terminal node $j$}\\
+$\epsilon$		\tab\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
+$1,\dots,i,\dots,N$ \tab\tab{index for observation}\\
+$1,\dots,j,\dots,J$	\tab\tab{index for terminal node}\\
+$1,\dots,k,\dots,K$	\tab{index for partitioning variable}\\
+$1,\dots,m,\dots,M$	\tab{index for cluster}\\
+$r$			\tab\tab\tab{index for iteration}\\
+$U_k$			\tab\tab\tab{(potential) partitioning variable $k$}\\
+$x_i$ 			\tab\tab\tab{column vector of fixed-effects predictor variable values for}\\
+			\tab\tab\tab\tab{observation $i$}\\
+$y_i$ 			\tab\tab\tab{treatment outcome for observation $i$}\\
+$z_i$ 			\tab\tab\tab{column vector of random-effects predictor variable values for}\\
+			\tab\tab\tab\tab{observation $i$}\\
+
+\end{appendix}
+
 \end{document}

### 2022-11-11 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
move second paper to lgcm-tree folder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3236 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 872e6824c6a542582e905032ec1338bfa6c58b9c

--- 2022-11-11 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.Rnw
@@
 GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data.
-{\color{red!80} Here, we introduce an extension of the algorithm for longitudinal (as opposed to clustered) data that can detect subgroups defined by subject-specific covariates on/or time-varying covariates.
+{\color{red!80} Here, we introduce an extension of the algorithm for longitudinal (as opposed to clustered) data that can detect subgroups defined by subject-specific covariates and/or time-varying covariates.
 The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.} Using simulated and empirical data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart.

### 2022-12-02 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3240 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 6fd67765c7e27182cd2bbfc253c1c292279af2b0

--- 2022-10-25 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.Rnw
@@
 
-\abstract{Growth curve models are popular tools for studying development over time.
-Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time.
-Recursive partitioning methods (RPMs) allow for detecting such subgroups.
-In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning LGCMs.
-GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data.
-Here, we introduce recent adaptations of the algorithm that that are of particular relevance for partitioning longitudinal models, and growth curve models in particulat.
-Using simulated and existing data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart.
-We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart.
-In addition, GLMM trees allow for the analysis of trajectories with both discrete and continuous time, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}
+\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time.
+Between-subject heterogeneity is commonly encountered and often of specific research interest.
+We show how generalized linear mixed effects model (GLMM) trees can be used to identify heterogeneous subgroups in linear growth curve models.
+GLMM trees were originally proposed for clustered cross-sectional data; here we introduces extensions for longitudinal data.
+The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.
+We evaluate the performance of GLMM trees in partitioning growth curve models on simulated datasets and existing datasets on children's development of reading, math and science abilities.
+We compare the performance of GLMM trees with the SEM trees, LongCART and longRPart methods.
+GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart.
+In addition, GLMM trees allow for analysing discrete as well as continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}
 
@@
 
-In longitudinal studies, (generalized) linear growth curve modeling is often used to model change over time, and inter individual differences in change.
+Modeling development over time is of prime interest in psychological research.
 For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail.
-Inter individual heterogeneity is likely in such studies, where (groups of) participants show higher or lower rates of growth over time.
-Through the use of mixed-effects or latent-variable models, such heterogeneity can be captured, and explained by covariates of a-priori known relevance \citep{NeisyMatt18}.
-
-Often, the covariates that predict or affect the trajectories may not be known a-priori, or researchers may have a specific interest in finding subgroups of participants with similar trajectories.
-Recursive partitioning methods (RPMs), also known as decision-tree methods, are particularly suited in such cases.
-Several existing RPMs allow for partitioning growth trajectories: For example, GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, Structural Equation Model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
-Partitioning based on non-linear growth curve models can be performed with the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods\footnote{Both IT methods specifically focus at detecting so-called \textit{treatment-subgroup interactions}: subgroups that respond more favorably to one treatment than another.}.
-
-In this paper, we focus on generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}, which allow for subgroup detection in a wide range of mixed-effects models.
-Although originally not specifically developed for partitioning LGCMs, application of GLMM trees for partitioning LGCMs is straightforward.
-We propose two adaptations to the algorithm, which we expect to be of particular relevance for partitioning LGCMs.
-
-The main difference between GLMM trees and the aforementioned RPMs, is that the latter all fit the full parametric model in each of the nodes.
-In contrast, GLMM trees uses the observations in each node to estimate the fixed-effects parameters, while using all observations to estimate the random-effects parameters.
-This may not only reduce computation time; it also reduces the total number of parameters to be estimated, which may improve stability and reduce overfitting.
-This local/global estimation approach was originally proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes; \cite{FokkySmit18} generalized it to model-based trees.
-
-The longRPart, longRPart2, IT-LT and original SEM tree algorithms employ an exhaustive split detection procedure: For every possible cut point in the current node, the full parametric model is re-estimated in the two resulting nodes.
-To choose the optimal cutpoint, a likelihood ratio test is computed for every possible cut point.
-Drawbacks of this exhaustive search are a heavy computational burden, as well as a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}.
-
-LongCART, MELT, GEE-based decision trees and score-based SEM trees fit full parametric models in each of the nodes as well, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split.
-This reduces computational burden.
-Furthermore, these approaches allow for employing a two-step approach for split detection, which separates variable and cutpoint selection, thus preventing variable selection bias.
+Inter-individual heterogeneity is likely encountered in such studies.
+For example, some (groups of) participants may show higher or lower rates of growth over time.
+Development over time is generally studied with mixed-effects or latent-variable models, which allow for capturing inter-individual heterogeneity and explaining this heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}.
+
+However, the covariates that predict or affect developmental trajectories may not be known in advance and often researchers may have a specific interest in identifying these covariates.
+Recursive partitioning methods (RPMs) are exploratory methods that allow for identifying relevant predictors from a potentially (very) large number of covariates.
+At the same time, RPMs allow for direct identification of heterogeneous subgroups.
+Several existing RPMs allow for subgroup detection in growth-curve models: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, structural equation model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
+Further, longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models \footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.
+
+In this paper, we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for subgroup detection in LGCMs.
+GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas.
+Yet, GLMM trees provide a much more general method, allowing for subgroup detection in almost any type of mixed-effects model.
+Compared to clustered cross-sectional data, however, longitudinal data often come with two central characteristics that may require a different estimation approach: Higher random-effects variance (i.e., higher intra-class correlations) and predictors of interest measured at higher levels (e.g., time-constant covariates).
+In this paper, we introduce and test two recent extensions of GLMM trees that can account for these characteristics.
+We focus on the specific use case of partitioning LGCMs, but methods for dealing with stronger intra-class correlations and higher-level covariates of interest are obviously of relevance beyond the realm of LGCMs.
+
+\subsection{Recursive partitioning methods for longitudinal data}
+
+Compared to the aforementioned RPMs for partitioning LGCMs \citep{Loh02, AbdoyLeBl02, Lee05, SuyMene11, BranyOert13, ArnoyVoel21, EoyCho14, KundyHare19}, GLMM trees differ in that they do not fit a full parametric model in each of the terminal nodes.
+GLMM trees estimate the fixed-effects parameters locally, using only the observations within a terminal node, and estimate the random-effects parameters globally, using all observations.
+Not only does this reduce computation time, it also reduces model complexity, which in turn may reduce instability and overfitting.
+This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes.
+In effect, the approach replaces the fixed-effects part of the mixed-effects model with a decision tree with constant fits in the terminal nodes, while the random effects are estimated as usual.
+\cite{FokkySmit18} generalized this approach to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of interaction effects.
+
+The computational advantage of GLMM trees is highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees.
+These latter methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split.
+To choose the optimal split, a likelihood ratio test is computed for every possible cut point.
+In addition to a heavy computational burden, this also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}.
+LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split.
+This reduces the computational burden, and allows for separating variable and cutpoint selection, thus preventing variable selection bias.
 The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.
 
-The remainder of this paper is structured as follows: In the next section (Estimation) we explain how GLMM trees are estimated, and how estimation can be adjusted to (better) account for dependence structures encountered in LGCMs.
-Next, in Study I, we perform a simulation study to assess performance of GLMM trees and the proposed adaptations in partitioning LGCMs.
-In Study II, we use the same simulation design to compare the performance of GLMM trees with that of two other mixed-effects RPMs: SEM trees and LongCART.
-In Study III, we compare the performance of GLMM trees and longRPart in partitioning children's trajectories of reading, math and science abilities.
-In the Discussion, we summarize our findings and discuss implications.
+\subsubsection{Current paper}
+
+In what follows, the next section explains how GLMM trees are estimated, and how it can be adapted to better account for longitudinal data structures.
+In three empirical evaluations, we assess and compare performance of GLMM trees in partitioning LGCMs: In Simulation Study I, we assess performance of the proposed adaptations to GLMM trees.
+In Simulation Study II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
+In the Application, we compare the performance of GLMM trees and longRPart in partitioning existing datasets on children's development of reading, math and science abilities.
+In the Discussion, we summarize our findings and their implications.
 
@@
 
-\section{Estimation of GLMM trees}
+\section{Estimation}
+\label{sec:estimation}
 
@@
 %
-In the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors.
+In the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$.
+In principle, generalizations to other distributions within the GLM.
 Further, $X_i$ is the $N_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$).
-We assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column for the observed timepoints of subject $i$.
-The set of observed timepoints may differ between subjects, and thus the number of rows of $X_i$ may differ between subjects.
-$\beta_j$ is a column vector of node-specific fixed-effects parameters, their values depending on terminal node $j$ of which observation $i$ is part.
-Subscript $j$ is what distinguishes the GLMM tree model from a traditional GLMM.
-As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$ and contains (a subset of) the columns of $X_i$.
-Further, $b_i$ is a vector of random effects of subject $i$.
-We assume that $b$ follows a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.
-We assume Gaussian errors, with constant variance across subjects.
+We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises two columns: a column of 1s for the intercept and a column of observed timepoints for subject $i$.
+The number ($N_i$) and values of observed timepoints may differ between subjects.
+While in a traditional GLMM, the fixed-effects parameters are assumed constant over all observations, but in the GLMM tree model, $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part.
+As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$ and $b_i$ is a vector of random effects for subject $i$.
+Finally, we assume that $b$ follows a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.
 
 The parameters of a traditional GLMM can be estimated by, for example, (restricted) maximum likelihood.
-The GLMM tree model in Eq.
-\ref{eq:GLMMtree} is estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:
+The GLMM tree model is estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:
 
@@
 
-The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown.
+The algorithm initializes by assuming zero random effects.
 In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2.
-Note that the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset.
+Thus, the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset.
 The fixed effects are estimated \textit{locally}, using the observations in the current node.
-Convergence of the algorithm is monitored by computing the log-likelihood criterion of the mixed-effects model fitted in Step~3.
+Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3.
 Typically, this converges if the tree does not change from one iteration to the next.
@@
 Studies to date indicate that initializing estimation with zero random effects yields an accurate final model \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
-\cite{SelaySimo12} assessed the impact of different initial values and found only minor impact on the final model, which became smaller with increasing sample size.
-In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional datasets.
-With longitudinal data (e.g., LGCMs), observations from the same unit tend to be more strongly correlated than in cross-sectional data; initialization with zero random effects may provide an unrealistic starting point which may be difficult to overcome in subsequent iterations.
-Initializing GLMM-tree estimation by estimating the random effects, assuming no grouping structure, may provide a better starting point.
-We thus hypothesize that for partitioning LGCMs, initializing GLMM-tree estimation with the random effects can improve subgroup recovery and predictive accuracy.
+\cite{SelaySimo12} assessed found only minor impact of different initialization values on the final model, which became smaller with increasing sample size.
+In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data.
+With longitudinal data, repeated measures on the same subjects tend to be correlated more strongly than observations nested withiin the same unit in cross-sectional data.
+If the random effects are in fact sizable, assuming them to be zero could provide an unrealistic starting point that is difficult to overcome in subsequent iterations.
+Our first hypothesis is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure improves accuracy of GLMM trees.
 
@@
 
-The partition (or tree) in Step~1 is estimated using model-based recursive partitioning \citep[MOB; ][]{ZeilyHoth08}. The MOB algorithm cycles iteratively through the following steps:
+The subgroup structure in Step~1 is estimated using model-based recursive partitioning \citep[MOB; ][]{ZeilyHoth08}. The MOB algorithm cycles iteratively through the following steps:
 
@@
 	\item Fit the parametric model to all observations in the dataset. 
-	\item Statistically test for parameter instability with respect to each of the partitioning variables.
+	\item Test for parameter instability with respect to each of the partitioning variables.
 	\item If there is some overall parameter instability, split the dataset with respect to the partitioning variable associated with the highest instability.
@@
 
-Parameter stability is tested in Step~(b), using the the \textit{scores} (gradient contributions) from the model fitted in Step~(a).
+Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the model fitted in Step~(a).
 Under mild regularity conditions, the scores have an expected value of 0.
-The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or if they exhibit systematic deviations when they are ordered by the values of one or more covariates.
+The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or exhibit systematic deviations when ordered by the values of a covariate.
 For continuous covariates (or ordered covariates with a large enough number of unique values), this involves computing the cumulative score process $W_k(t)$ with respect to each potential partitioning variable $U_k$ \citep{ZeilyHoth08}:
@@
 
-where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in the current node, and $n$ gives the number of observations in the current node.
+where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in Step~(a), and $n$ gives the number of observations in the current node.
 Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes estimated scores (denoted $\hat{\psi}$), with the subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$.
 Note that $0 \leq t \leq 1$, thus $nt = 1$ for an observation with a unique minimum on the partitioning variable, and $nt = n$ for an observation with a unique maximum.
+
+\textcolor{red}{Perhaps replace $n$ with $N_j$? This is more line with notation of GLMM tree model, but might be confusing with cov mat $J$?}
 
@@
 
-When partitioning LGCMs, covariates will often not be measured at the lowest level, but at the subject level.
-The covariance matrix employed in Eq.
-\ref{eq:efp} can account for this clustering of observations within subjects \citep[e.g., ][]{ZeilyKoll20} by employing a clustered covariance matrix $\hat{J}$ in computation of the score process.
-\textcolor{orange}{@AZ: Thus, only computation of $\hat{J}$ is affected by cluster argument?
-Sscores are summed before computing $\hat{J}$, but computation of $\sum^{[nt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik}})$ is unaffected?} This resembles a GEE-type approach, where dependence between observations is accounted for in the variance structure.
-We thus hypothesize that for partitioning LGCMs, use of clustered covariances yields more accurate subgroup recovery and improves predictive accuracy.
-
-\section{Study I: Performance of LM(M) trees}
-
-To test whether initializing GLMM-tree estimation with the random effects and use of clustered covariances in computation of the parameter stability tests yields more accurate subgroup recovery when partitioning LGCMs, we performed a simulation study.
+When partitioning longitudinal data, covariates will often be measured at the subject level.
+This clustering of observations within subjects can be accounted for in computing a so-called \textit{clustered} covariance matrix $\hat{J}$ \citep[e.g., ][]{ZeilyKoll20}.
+This resembles a GEE-type approach with an independence correlation structure.
+Our second hypothesis is that for partitioning LGCMs, use of clustered covariances improves accuracy of GLMM trees.
+
+\section{Simulation Study I}
+
+We performed a simulation study to compare the performance GLM(M) trees estimated by initializing estimation with the subgroup structure versus the random effects, and employing observation- versus cluster-level covariances for the parameter stability tests.
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!t]
 \caption{Design of the subgroup-specific fixed effects in the simulation.}
@@
 
-We simulated datasets according to the subgroup structure depicted in Figure~\ref{fig:design_tree}.
-Specifically, four non-overlapping subgroups corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree} were created.
-Observations in terminal nodes 3 and 4, and in terminal nodes 6 and 7 have the same fixed intercept ($\beta_{j0}$).
-To facilitate comparison with the SEM tree algorithm, the same timepoints were generated for all subjects (i.e., $t = 0, 1, 2, 3, 4$).
+We simulated data according to the subgroup structure depicted in Figure~\ref{fig:design_tree}.
+Every dataset comprised four non-overlapping subgroups, corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree}.
 The response was computed as:\\
@@
 \noindent where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part.
+The bottom panels of Figure~\ref{fig:design_tree} present the subgroup-specific values of the fixed intercept and slope ($\beta_{j0}$ and $\beta_{j1}$, respectively).
 The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints.
-The value for $b_i$ was generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next.
+The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$ \footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees.}.
+Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next.
 Values in $\epsilon_{i}$ were independently generated from $N(\mu = 0, \sigma = \sqrt{5})$.
@@
 
-We employed a full factorial design, yielding $2^5 = 32$ cells.We performed 100 repetitions for each cell of the design. For generating and analysing data, we used R \citep[][version 4.1.2]{R22}.
-
-\subsubsection{Partitioning methods}
-
-We fitted two different LM trees to each dataset: One employing default settings (i.e., parameter stability tests performed at level I) and one where the parameter stability tests employed clustered covariance matrices (i.e., parameter stability tests performed at level 2).
-To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15][]{HothyZeil15}.
-
-We fitted eight different LMM trees to each dataset: We estimated an LMM tree using default settings (i.e., parameter stability tests performed using observation-level covariances, estimation initialized with the tree structure); an LMM tree with parameter stability tests performed using cluster-level covariances; an LMM tree estimated by initializing estimation with the random effects; an LMM tree estimated by combining the latter two approaches.
-Each of the four approaches were applied with a specification comprising random intercepts only, as well as with a specification comprising both random intercepts and slopes, yielding eight different LMM trees.
-The different specifications allow for assessing the effect of mis-specification of the random-effects structure.
-To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29][]{BateyMach15} and to fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0][]{FokkySmit18}.
+We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We used R \citep[version 4.1.2; ][]{R22} to generate and analyze all data.
+
+\subsubsection{Partitioning approaches}
+
+We fitted ten trees to each dataset:
+
+\begin{itemize}
+  \item Two linear model (LM) trees: 
+  \begin{itemize}
+    \item A tree employing default settings, where the parameter stability tests use observation-level covariances.
+    \item A tree where the parameter stability tests use clustered covariances. 
+  \end{itemize}
+  \item Eight linear mixed model (LMM) trees: 
+  \begin{itemize}
+    \item Two trees fitted using default settings, where parameter stability tests use observation-level covariances and estimation is initialized with the tree structure. 
+    \item Two trees where the parameter stability tests use cluster-level covariances.
+    \item Two trees where estimation is initialized with the random effects. 
+    \item Two trees where the parameter stability tests use cluster-level covariances and estimation is initialized with the random effects.
+  \end{itemize}
+\end{itemize}
+
+Each of the two LMM trees were fitted with different specifications of the random effects: One of the LMM trees estimated random intercepts only, one of the LMM trees estimated both random intercepts and slopes.
+Because variances of random intercepts and slopes were $> 0$ in all cells of the design, the three different random-effects specifications (LM trees assume 0 random effects) allow for assessing the effect of mis-specifying the random effects.
+To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}.
+To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29; ][]{BateyMach15}.
+To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}.
 
@@
 Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach.
-The default fitting approach overfits, and implemented $> 3$ splits in most datasets.
-This overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits \Sexpr{round(M_cluster, digits = 2)}), which appears most effective for LMM trees (mean number of splits \Sexpr{round(means["lmer_c"], digits = 2)} for trees estimating only random intercepts; \Sexpr{round(means["lmer_s_c"], digits = 2)} for trees estimating random intercepts and slopes).
-
-Initializing estimation with the random effects also effectively reduces tree size, but too strongly when random slopes are estimated (mean number of splits without random slopes \Sexpr{round(means["lmer_r"], digits = 2)}; with random slopes \Sexpr{round(means["lmer_s_r"], digits = 2)}).
-Heterogeneity related to partitioning variables measured at level 2 can either be captured and explained by the tree, or captured by the random effects.
-By initializing estimation with the random effects effects, this heterogeneity can no longer be captured by the tree; this likely worsens with more complex random-effects specification.
-Combining the use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding a similar and too high number of splits, irrespective of the random-effects specification (mean number of splits \Sexpr{round(M_both, digits = 2)}).
-
-Distributions of the number of splits, separated according to the levels of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ are depicted in Figure~\ref{fig:LMM_sizes_interact} (Appendix~\ref{sec:AppendixA}).
-These plots show a pattern similar to Figure~\ref{fig:LMM_sizes}, with use of cluster-level covariances providing the most accurate split recovery.
-In addition, the plots show that increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ yield higher numbers of splits, while the effect of $\rho$ is very minor.
-\textcolor{orange}{I moved plots to Appendix because of their size and because they are not of main interest.
-Not sure if following results should be discussed here or in the Appendix:} The main deviation from this pattern of main effects is that increased variance of the random effects (increases $\sigma_{b_0}$ and/or $\sigma_{b_1}$) yields lower numbers of splits if random slopes are estimated and estimation is initialized with the random effects.
+Irrespective of the random-effect specification, the default fitting approaches overfit, implementing $> 3$ splits in a large majority of datasets.
+Irrespective of the random-effects specification, the overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits \Sexpr{round(M_cluster, digits = 2)}), which appears most effective for LMM trees: the mean number of splits was \Sexpr{round(means["lmer_c"], digits = 2)} for LMM trees with random intercepts only and \Sexpr{round(means["lmer_s_c"], digits = 2)} for LMM trees estimating both random intercepts and slopes.
+
+Initializing estimation with the random effects was also effective in reducing tree size, bput the effect depended on the random-effects specification.
+When random slopes were estimated in addition to random intercepts, too few splits were implemented on average: the mean number of splits was \Sexpr{round(means["lmer_r"], digits = 2)} for LMM trees with random intercepts only, and \Sexpr{round(means["lmer_s_r"], digits = 2)}) for LMM trees estimating both random intercepts and slopes.
+Combined use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding too many splits, irrespective of the random-effects specification (mean number of splits \Sexpr{round(M_both, digits = 2)}).
+
+Distributions of the number of splits, separated according to the levels of the data-generating design are depicted in Figure~\ref{fig:LMM_sizes_interact} (Appendix~\ref{sec:AppendixA}).
+The plots indicate little interaction between data-generating parameters and model-fitting approached.
+Figure~\ref{fig:LMM_sizes_interact} shows a pattern very similar to Figure~\ref{fig:LMM_sizes}, where the use of cluster-level covariances provides the most accurate split recovery.
+Main effects of the data-generating design were as expected: Increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ yield higher numbers of splits.
+The effect of $\rho$ was minor.
+The only evidence of an interaction between data-generating parameters and model-fitting approaches is that increased variance of the random effects (higher $\sigma_{b_0}$ and/or $\sigma_{b_1}$) yields lower numbers of splits if random slopes are estimated and estimation is initialized with the random effects.
+Thus, with stronger random effects and a more complex random-effects specification, differences in fixed effects (which should be captured by the tree) tend to be captured by the random effects when estimation is intialized with the random effects.
 

### 2022-12-10 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Updated LGCM GLMM-tree paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3241 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ad9f2ca5e6d0d5ed9e683530438301d2e98e11c0

@@
 \usepackage{xcolor}
+\usepackage{setspace}
+
+\renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
+
 %\usepackage[colorinlistoftodos]{todonotes}
@@
 \abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time.
-Between-subject heterogeneity is commonly encountered and often of specific research interest.
+Heterogeneity between subjects is likely encountered in such models and researchers are often interested in explaining or predicting this heterogeneity.
 We show how generalized linear mixed effects model (GLMM) trees can be used to identify heterogeneous subgroups in linear growth curve models.
-GLMM trees were originally proposed for clustered cross-sectional data; here we introduces extensions for longitudinal data.
+GLMM trees were originally proposed for clustered cross-sectional data; here we extend the method to allow for longitudinal data.
 The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.
-We evaluate the performance of GLMM trees in partitioning growth curve models on simulated datasets and existing datasets on children's development of reading, math and science abilities.
-We compare the performance of GLMM trees with the SEM trees, LongCART and longRPart methods.
+We evaluate GLMM tree's performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities.
+We compare the performance of GLMM trees with three alternative methods: SEM trees, LongCART and longRPart.
 GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart.
-In addition, GLMM trees allow for analysing discrete as well as continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}
+In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}
 
@@
 
-Modeling development over time is of prime interest in psychological research.
-For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail.
-Inter-individual heterogeneity is likely encountered in such studies.
-For example, some (groups of) participants may show higher or lower rates of growth over time.
-Development over time is generally studied with mixed-effects or latent-variable models, which allow for capturing inter-individual heterogeneity and explaining this heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}.
-
-However, the covariates that predict or affect developmental trajectories may not be known in advance and often researchers may have a specific interest in identifying these covariates.
-Recursive partitioning methods (RPMs) are exploratory methods that allow for identifying relevant predictors from a potentially (very) large number of covariates.
-At the same time, RPMs allow for direct identification of heterogeneous subgroups.
-Several existing RPMs allow for subgroup detection in growth-curve models: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, structural equation model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
-Further, longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models \footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.
-
-In this paper, we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for subgroup detection in LGCMs.
+Development over time is of prime interest in psychological researchers.
+For example, researchers may want to model and predict student's reading or math development in educational studies \citep{MorgyFark09}, or patient's symptom reduction over the course of clinical trails \citep{Fein09}.
+Mixed-effects or latent-variable models can be used to model development over time, and allow for explaining possible heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}.
+However, the covariates that are predictive of development may often not be known in advance, and identifying them may be of prime interest.
+
+Recursive partitioning methods (RPMs) allow for identifying relevant predictors from a potentially (very) large number of covariates.
+Several RPMs that allow for partitioning growth-curve models have been developed: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
+Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.
+
+Here we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for partitioning LGCMs.
 GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas.
-Yet, GLMM trees provide a much more general method, allowing for subgroup detection in almost any type of mixed-effects model.
-Compared to clustered cross-sectional data, however, longitudinal data often come with two central characteristics that may require a different estimation approach: Higher random-effects variance (i.e., higher intra-class correlations) and predictors of interest measured at higher levels (e.g., time-constant covariates).
-In this paper, we introduce and test two recent extensions of GLMM trees that can account for these characteristics.
-We focus on the specific use case of partitioning LGCMs, but methods for dealing with stronger intra-class correlations and higher-level covariates of interest are obviously of relevance beyond the realm of LGCMs.
-
-\subsection{Recursive partitioning methods for longitudinal data}
-
-Compared to the aforementioned RPMs for partitioning LGCMs \citep{Loh02, AbdoyLeBl02, Lee05, SuyMene11, BranyOert13, ArnoyVoel21, EoyCho14, KundyHare19}, GLMM trees differ in that they do not fit a full parametric model in each of the terminal nodes.
-GLMM trees estimate the fixed-effects parameters locally, using only the observations within a terminal node, and estimate the random-effects parameters globally, using all observations.
-Not only does this reduce computation time, it also reduces model complexity, which in turn may reduce instability and overfitting.
+Yet, GLMM trees provide a more general method 2for subgroup detection in almost any type of mixed-effects model.
+Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates).
+In this paper, we propose and test two recent extensions of GLMM trees that account for these characteristics.
+We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wide range of settings where covariates are measured at higher levels and/or random-effects variance is substantial.
+
+The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the terminal nodes.
+Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations.
 This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes.
-In effect, the approach replaces the fixed-effects part of the mixed-effects model with a decision tree with constant fits in the terminal nodes, while the random effects are estimated as usual.
-\cite{FokkySmit18} generalized this approach to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of interaction effects.
-
-The computational advantage of GLMM trees is highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees.
-These latter methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split.
+With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \cite{FokkySmit18}.
+
+All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based the observations in that node only.
+This fully local estimation approach brings higher computational load and increased model complexity.
+GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model.
+Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.
+
+The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees.
+These methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split.
 To choose the optimal split, a likelihood ratio test is computed for every possible cut point.
-In addition to a heavy computational burden, this also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}.
+Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}.
 LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split.
-This reduces the computational burden, and allows for separating variable and cutpoint selection, thus preventing variable selection bias.
-The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.
-
-\subsubsection{Current paper}
-
-In what follows, the next section explains how GLMM trees are estimated, and how it can be adapted to better account for longitudinal data structures.
-In three empirical evaluations, we assess and compare performance of GLMM trees in partitioning LGCMs: In Simulation Study I, we assess performance of the proposed adaptations to GLMM trees.
-In Simulation Study II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
-In the Application, we compare the performance of GLMM trees and longRPart in partitioning existing datasets on children's development of reading, math and science abilities.
-In the Discussion, we summarize our findings and their implications.
+This not only reduces computational load, it also allows for separating variable and cutpoint selection, thus preventing selection bias.
+The GLMM tree algorithm also employs a two-step approach to split selection.
+
+In summary, compared to other methods for partitioning LGCMs, GLMM trees may provide lower computational load, lower model complexity and instability, and unbiased variable selection.
+
+In what follows, we first describe estimation of GLMM trees, and possible adjustments to better account for longitudinal data structures.
+Next, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees.
+In Simulation~Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
+In the Application, we compare the performance of GLMM trees and longRPart in existing datasets on children's development of reading, math and science abilities.
+In the Discussion, we summarize our findings and implications.
 
@@
 %
-In the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$.
-In principle, generalizations to other distributions within the GLM.
-Further, $X_i$ is the $N_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$).
-We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises two columns: a column of 1s for the intercept and a column of observed timepoints for subject $i$.
-The number ($N_i$) and values of observed timepoints may differ between subjects.
-While in a traditional GLMM, the fixed-effects parameters are assumed constant over all observations, but in the GLMM tree model, $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part.
-As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$ and $b_i$ is a vector of random effects for subject $i$.
-Finally, we assume that $b$ follows a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.
+Thus, in the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$.
+Generalizations to other response variable types within the GLM are straightforward.
+Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$).
+We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$.
+Note that the number and spacing of observed timepoints may differ between subjects.
+$\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part.
+As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$.
+Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.
 
@@
 
-Studies to date indicate that initializing estimation with zero random effects yields an accurate final model \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
-\cite{SelaySimo12} assessed found only minor impact of different initialization values on the final model, which became smaller with increasing sample size.
+Studies to date indicate that initializing estimation with zero random effects yields accurate final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
+\cite{SelaySimo12} assessed the impact of different initialization values and found minor impact that became smaller with increasing sample size.
 In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data.
-With longitudinal data, repeated measures on the same subjects tend to be correlated more strongly than observations nested withiin the same unit in cross-sectional data.
-If the random effects are in fact sizable, assuming them to be zero could provide an unrealistic starting point that is difficult to overcome in subsequent iterations.
-Our first hypothesis is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure improves accuracy of GLMM trees.
+With longitudinal data, however, random effects tend to be stronger: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data.
+If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that is difficult to overcome in subsequent iterations.
+Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy.
 
@@
 \label{eq:efp}
-W_{k}(t) = \hat{J}^{-1/2} n^{-1/2} \sum^{[nt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
+W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
 \end{eqnarray}
 
-where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in Step~(a), and $n$ gives the number of observations in the current node.
-Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes estimated scores (denoted $\hat{\psi}$), with the subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$.
-Note that $0 \leq t \leq 1$, thus $nt = 1$ for an observation with a unique minimum on the partitioning variable, and $nt = n$ for an observation with a unique maximum.
-
-\textcolor{red}{Perhaps replace $n$ with $N_j$? This is more line with notation of GLMM tree model, but might be confusing with cov mat $J$?}
+where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in Step~(a), and $n_j$ gives the number of observations in the current node.
+Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes the estimated scores, with subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$.
+Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
 
@@
 
-When partitioning longitudinal data, covariates will often be measured at the subject level.
-This clustering of observations within subjects can be accounted for in computing a so-called \textit{clustered} covariance matrix $\hat{J}$ \citep[e.g., ][]{ZeilyKoll20}.
+When partitioning longitudinal data, covariates will often be measured at the subject level, which should be accounted for in computing the estimated covariance matrix $\hat{J}$.
+In general, this computation makes use of the scores.
+By summing the scores within clusters prior to computation of the covariances, so-called \textit{clustered} covariances are obtained, which account for dependence between observations within the same cluster \citep{ZeilyKoll20}.
 This resembles a GEE-type approach with an independence correlation structure.
-Our second hypothesis is that for partitioning LGCMs, use of clustered covariances improves accuracy of GLMM trees.
+Our expectation is that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve accuracy of the tree.
+
+% $\textrm{Var} \hat{\beta}$ = (X^\top X)^{-1} X^\top \Omega X (X^\top X)^{-1}$
+
+% The $X^\top \Omega X$ is the meat, and the $(X^\top X)^{-1}$ is the bread.
+For computing robust standard errors, we compute a new kind of meat.
+$\Omega = \sigma^2 I_n$, a matrix with $\sigma^2$ on the diagonal and zeros elsewhere.
+$n$ can be the number of observations, or the number of clusters.
 
@@
 
-\subsubsection{Data-generating design}
+\subsubsection{Data generation}
 
@@
 node7 <- as.numeric(U1 == 1 & U3 > 0)
-y <- (node3 * -1 - node3 * time) + (-node4) +
- (node6) + (node7 * 1 + node7 * time) 
+y <- -(node3 + node3*time) - node4 + node6 + (node7 + node7*time) 
 data <- data.frame(U1 = as.factor(U1), U2, U3, time, y)   
 tree <- lmtree(y ~ time | U1 + U2 + U3, data = data, maxdepth = 3)
-
-fig4 <- party(
+fig <- party(
   partynode(1L,
@@
 )
-#plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
 
@@
 
-\begin{figure}[!t]
+\begin{figure}[bt]
 \caption{Design of the subgroup-specific fixed effects in the simulation.}
 <<fig=T, height=5, width=5, echo=F>>=
-#plot(tree, ip_args = list(pval = FALSE), gp = gpar(cex = .7),
-#     tp_args = list(pointcol = "white", xlab = "time",
-#                    mainlab = function(id, nobs) paste("node", id)))
 plot(tree, tnex = 3.5, terminal_panel = combine_panels(
@@
                         mainlab = function(id, nobs) paste("node", id)),
-    panel2 = node_terminal(fig4, FUN = identity, height = 4, width = 12, id = FALSE),
-    party2 = fig4), 
+    panel2 = node_terminal(fig, FUN = identity, height = 4, width = 12, id = FALSE),
+    party2 = fig), 
   gp = gpar(cex = .7), ip_args = list(pval = FALSE))
@@
 The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints.
-The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$ \footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees.}.
+The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$\footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees in the next section.}.
 Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next.
-Values in $\epsilon_{i}$ were independently generated from $N(\mu = 0, \sigma = \sqrt{5})$.
+Values of $\epsilon_{i}$ were independently generated from a normal distribution with a mean of zero and a variance of 5.
 
@@
 \item Variance of the random intercept: small ($\sigma_{b_0}^2 = 1$) or large ($\sigma_{b_0}^2 = 2$).
-\item Variance of the random slope: small ($\sigma_{b_1}^2 = .1$) or large ($\sigma_{b_1}^2 = .4$).
+\item Variance of the random slope: small ($\sigma_{b_1}^2 = 0.1$) or large ($\sigma_{b_1}^2 = 0.4$).
 \item Number of noise partitioning variables: small ($p = 5$) or large ($p = 25$).
@@
 
-We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We used R \citep[version 4.1.2; ][]{R22} to generate and analyze all data.
-
-\subsubsection{Partitioning approaches}
-
-We fitted ten trees to each dataset:
+We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We performed all data generation and analysis in R \citep[version 4.1.2; ][]{R22}.
+
+\subsubsection{Model fitting}
+
+We applied ten partitioning approaches to every generated dataset, each employing one of three different random-effects specifications:
 
   \begin{itemize}
-  \item Two linear model (LM) trees: 
+
+    \item $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$. That is, random effects and thus their variance were fixed to 0, and two fixed-effects linear model (LM) trees were fitted.
     \begin{itemize}
-    \item A tree employing default settings, where the parameter stability tests use observation-level covariances.
-    \item A tree where the parameter stability tests use clustered covariances. 
+        \item The default observation-level covariances were used to compute the parameter stability tests.
+        \item Clustered covariances were used to compute the parameter stability tests. 
     \end{itemize}
-  \item Eight linear mixed model (LMM) trees: 
+
+    \item $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$.
+That is, four linear mixed-effects model (LMM) tree were fitted in which the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0:
     \begin{itemize}
-    \item Two trees fitted using default settings, where parameter stability tests use observation-level covariances and estimation is initialized with the tree structure. 
-    \item Two trees where the parameter stability tests use cluster-level covariances.
-    \item Two trees where estimation is initialized with the random effects. 
-    \item Two trees where the parameter stability tests use cluster-level covariances and estimation is initialized with the random effects.
+      \item Default settings were used, in which parameter stability tests employed observation-level covariances and estimation is initialized with the tree structure. 
+    \item Instead, clustered covariances were used to compute the parameter stability tests.
+    \item Instead, estimation was initialized with the random effects. 
+    \item Instead, clustered covariances were used to compute the parameter stability tests and estimation was initialized with the random effects.
   \end{itemize}
+
+    \item $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$.
+That is, four LMM trees were fitted in which the variance of both the random intercept and slope were freely estimated.
+The same four approaches were applied (default, clustered covariances, random-effects initialization, clustered covariances and random-effects initialization) as with the previous LMM tree specification.
   \end{itemize}
 
-Each of the two LMM trees were fitted with different specifications of the random effects: One of the LMM trees estimated random intercepts only, one of the LMM trees estimated both random intercepts and slopes.
-Because variances of random intercepts and slopes were $> 0$ in all cells of the design, the three different random-effects specifications (LM trees assume 0 random effects) allow for assessing the effect of mis-specifying the random effects.
 To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}.
@@
 To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}.
+To compute clustered covariance, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}.
+We employed outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.
 
@@
 Trees with $> 3$ splits are indicative of Type-I error, while trees with $> 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables).
-We also assessed which variable was selected for the first split in every tree.
+We also assessed the variable selected for the first split in every tree.
 
@@
 <<echo=FALSE>>=
+###########################
+##
+## Load tree size data
+##
+
 load("tree_sizes.Rda")
@@
 M_both <- mean(tmp$tree_size[tmp$both_ids])
+
+means <- tapply(tmp$tree_size, tmp$method_id, mean)
+sds <- tapply(tmp$tree_size, tmp$method_id, sd)
+mads <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(abs(x-3)))
+props_more_splits <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(x > 3L))
+props_less_splits <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(x < 3L))
+props_corr_splits <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(x == 3L))
 @
@@
 
-<<echo=FALSE>>=
-means <- tapply(tmp$tree_size, tmp$method_id, mean)
-@
-
 \begin{figure}[!ht]
-\caption{Tree size distributions for LM(M) trees.}
-\begin{subfigure}{1.2\textwidth}
-<<echo=FALSE, fig=TRUE, height=3.5>>=
+\caption{Tree size distributions for LM trees (left panel) and LMM trees (middle and right panel).}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=4>>=
+library("dplyr")
+
 ## Prepare data
 tmp <-  tree_sizes_long
-tmp$tree_size <- (tmp$tree_size-1)/2
+tmp$tree_size <- (tmp$tree_size-1)/2 ## count splits instead of nodes
 tmp <- tmp[-which(tmp$method_id == "true") , ]
@@
 tmp$ranef <- NA
-tmp$ranef[tmp$method_id %in% c("lm", "lm_c")] <- "LM tree \n(no random effects)"
-tmp$ranef[tmp$method_id %in% c("lmer", "lmer_c", "lmer_r", "lmer_cr")] <- "LMM tree \n(random intercept)"
-tmp$ranef[tmp$method_id %in% c("lmer_s", "lmer_s_c", "lmer_s_r", "lmer_s_cr")] <- "LMM tree \n(random intercept + slope)"
+tmp$ranef[tmp$method_id %in% c("lm", "lm_c")] <- "{hat(sigma)[b[0]] == hat(sigma)[b[1]]} == 0"
+tmp$ranef[tmp$method_id %in% c("lmer", "lmer_c", "lmer_r", "lmer_cr")] <- "list(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)"
+tmp$ranef[tmp$method_id %in% c("lmer_s", "lmer_s_c", "lmer_s_r", "lmer_s_cr")] <- "list(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)"
 tmp$ranef[tmp$method_id %in% c("LongCART")] <- "LongCART"
@@
 tmp$ranef[tmp$method_id %in% c("s_sem_s", "f_sem_s")] <- "SEM tree \n(random intercept + slope)"
-tmp$ranef <- factor(tmp$ranef, 
-                    levels = c("LM tree \n(no random effects)", "LMM tree \n(random intercept)", 
-                               "LMM tree \n(random intercept + slope)", 
-                               "SEM tree \n(no random effects)", "SEM tree \n(random intercept)", 
-                               "SEM tree \n(random intercept + slope)"))
 
@@
 tmp$setting[tmp$method_id %in% c("lmer", "lmer_s", "lm")] <- "default"
-tmp$setting[tmp$method_id %in% c("lmer_c", "lmer_s_c", "lm_c")] <- "cluster"
-tmp$setting[tmp$method_id %in% c("lmer_r", "lmer_s_r")] <- "ranef\ninit"
-tmp$setting[tmp$method_id %in% c("lmer_cr", "lmer_s_cr")] <- "cluster +\nranef"
+tmp$setting[tmp$method_id %in% c("lmer_c", "lmer_s_c", "lm_c")] <- "cl.cov."
+tmp$setting[tmp$method_id %in% c("lmer_r", "lmer_s_r")] <- "ran.eff."
+tmp$setting[tmp$method_id %in% c("lmer_cr", "lmer_s_cr")] <- "cl.cov.+\nran.eff."
 tmp$setting[tmp$method_id %in% c("LongCART")] <- "default"
@@
 tmp$setting[tmp$method_id %in% c("f_sem", "f_sem_i", "f_sem_s")] <- "fair"
-tmp$setting <- factor(tmp$setting, levels = c("cluster +\nranef", "cluster", 
+tmp$setting <- factor(tmp$setting, levels = c("cl.cov.+\nran.eff.", "cl.cov.", 
                                               "default", "fair",
-                                              "ranef\ninit", "score"))
+                                              "ran.eff.", "score"))
 levels(tmp$sigma_int) <- c("1", "4")
 levels(tmp$sigma_slope) <- c("0.1", "0.4")
-tmp$setting <- factor(tmp$setting, levels = c("default", "cluster",  "ranef\ninit", 
-                       "cluster +\nranef", "fair", "score"))
+tmp$setting <- factor(tmp$setting, levels = c("default", "cl.cov.",  "ran.eff.", 
+                       "cl.cov.+\nran.eff.", "fair", "score"))
 
@@
 
+## Set up data.frame of summary stats
+summ_stats <- data.frame(M = round(means[c("lm", "lm_c", "lmer", "lmer_c", "lmer_cr", "lmer_r", 
+                   "lmer_s", "lmer_s_c", "lmer_s_cr", "lmer_s_r")], digits = 2), 
+                   SD = format(sds[c("lm", "lm_c", "lmer", "lmer_c", "lmer_cr", "lmer_r", 
+                   "lmer_s", "lmer_s_c", "lmer_s_cr", "lmer_s_r")], digits = 2),
+                   MAD = format(mads[c("lm", "lm_c", "lmer", "lmer_c", "lmer_cr", "lmer_r", 
+                   "lmer_s", "lmer_s_c", "lmer_s_cr", "lmer_s_r")], digits = 2))
+summ_stats$method <- rownames(summ_stats)
+summ_stats$y <- 50
+## Note!!! Labels of ran.eff. and cl.cov.+ran.eff. have been switched to have text labels
+## above the right boxplot / countplot. Setting it as a factor and releveling does not help. 
+summ_stats$setting <- c("default", "cl.cov.", 
+                        "default", "cl.cov.", "cl.cov.+\nran.eff.", "ran.eff.", 
+                        "default", "cl.cov.", "cl.cov.+\nran.eff.", "ran.eff.")
+
+summ_stats$ranef <- NA
+summ_stats$ranef[summ_stats$method %in% c("lm", "lm_c")] <- "{hat(sigma)[b[0]] == hat(sigma)[b[1]]} == 0"
+summ_stats$ranef[summ_stats$method %in% c("lmer", "lmer_c", "lmer_r", "lmer_cr")] <- "list(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)"
+summ_stats$ranef[summ_stats$method %in% c("lmer_s", "lmer_s_c", "lmer_s_r", "lmer_s_cr")] <- "list(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)"
+summ_stats$all <- with(summ_stats, paste(M, SD, MAD, sep = "\n"))
+
 ## LM(M)
@@
 breaks <- c(0, 1, 3, 7, 20)
-lim <- c(0, 25)
+lim <- c(0, 60)
+
+## Plot
 plot0 <- ggplot(tmp[LMM_ids, ]) +
@@
                position=position_dodge(1), show.legend = FALSE, 
-               alpha = 0.5, width = .6) + 
+               alpha = 0.5, width = .6, outlier.shape = NA, coef = NULL) + 
+  geom_count(aes(x=setting, y=tree_size), col = "grey", alpha = .75) +
   scale_y_continuous(trans = pseudo_log_trans(sigma = 1, base = exp(1)), 
                      breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits") + theme(legend.position = "none") +
-  geom_hline(yintercept=3, col = "darkgrey")
+  geom_text(data = summ_stats, aes(x=setting, y=y, label = all),
+            size = 3, col = "grey44") +
+  labs(x = "", y = "no. of splits") + theme(legend.position = "none") +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  facet_grid(~ranef, scales = "free", space = "free", 
+             labeller = label_parsed)  +
+  coord_cartesian(clip = 'off') +
+  labs(tag = "  M\n SD\nMAD") + theme(plot.tag.position = c(.04, .85),
+                                   plot.tag = element_text(size = 10))
 plot0
 @
-\end{subfigure}
-{\footnotesize \textit{Note.} The $y$-axis is on the log scale and represents tree size.
-Grey lines indicate true number of splits (3 nodes).
-cluster = cluster-level covariances employed in parameter stability tests; ranef init = estimation initialized with the random effects.
-Distances on $y$-axis are on the log scale.}
+\end{subfigure}%
+\vspace{-0.5cm}%
+
+{\textit{Note.} Grey circles represent counts, grey lines represent true number of splits.
+Panel labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.
+Distances on $y$-axis are on the log scale.
+cl.cov.
+= cluster-level covariances employed in parameter stability tests; ran.eff.
+= estimation initialized with the random effects.
+MAD = Mean absolute deviation from true tree size.}
 \label{fig:LMM_sizes}
@@
 Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach.
-Irrespective of the random-effect specification, the default fitting approaches overfit, implementing $> 3$ splits in a large majority of datasets.
-Irrespective of the random-effects specification, the overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits \Sexpr{round(M_cluster, digits = 2)}), which appears most effective for LMM trees: the mean number of splits was \Sexpr{round(means["lmer_c"], digits = 2)} for LMM trees with random intercepts only and \Sexpr{round(means["lmer_s_c"], digits = 2)} for LMM trees estimating both random intercepts and slopes.
-
-Initializing estimation with the random effects was also effective in reducing tree size, bput the effect depended on the random-effects specification.
-When random slopes were estimated in addition to random intercepts, too few splits were implemented on average: the mean number of splits was \Sexpr{round(means["lmer_r"], digits = 2)} for LMM trees with random intercepts only, and \Sexpr{round(means["lmer_s_r"], digits = 2)}) for LMM trees estimating both random intercepts and slopes.
-Combined use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding too many splits, irrespective of the random-effects specification (mean number of splits \Sexpr{round(M_both, digits = 2)}).
-
-Distributions of the number of splits, separated according to the levels of the data-generating design are depicted in Figure~\ref{fig:LMM_sizes_interact} (Appendix~\ref{sec:AppendixA}).
-The plots indicate little interaction between data-generating parameters and model-fitting approached.
-Figure~\ref{fig:LMM_sizes_interact} shows a pattern very similar to Figure~\ref{fig:LMM_sizes}, where the use of cluster-level covariances provides the most accurate split recovery.
-Main effects of the data-generating design were as expected: Increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ yield higher numbers of splits.
-The effect of $\rho$ was minor.
-The only evidence of an interaction between data-generating parameters and model-fitting approaches is that increased variance of the random effects (higher $\sigma_{b_0}$ and/or $\sigma_{b_1}$) yields lower numbers of splits if random slopes are estimated and estimation is initialized with the random effects.
-Thus, with stronger random effects and a more complex random-effects specification, differences in fixed effects (which should be captured by the tree) tend to be captured by the random effects when estimation is intialized with the random effects.
-
-Table~\ref{tab:first_splits} shows the variables selected for the first split.
-Only rarely is the wrong variable selected for the first split, and this occurs only with random-effects initialization and observation-level covariances in the parameter stability tests are employed.
-In the large majority of those cases, $U_2$ and $U_3$ were selected for the first split.
+Irrespective of the random-effect specification, the default fitting approaches (left-most box in every panel) tended to overfit, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average.
+This overfitting was successfully mitigated through use of cluster-level covariances in the parameter stability tests, irrespective of the random-effects specification.
+The average number of splits of LMM trees employing clustered covariances were closest to the true number of splits.
+In terms of mean absolute deviation (MAD) from the true tree size, LM trees with clustered covariances performed best, closely followed by the LMM trees employing clustered covariances.
+
+In terms of MAD, however, the LMM tree estimating random intercepts only performed even slightly better than the latter two approaches.
+Thus, initializing estimation with the random effects can successfully mitigate the overfitting observed for the default LMM trees, but the effect depens strongly on the random-effects specification.
+With estimation of both a random intercept and slope, random-effects initialization may lead to subgroup differences being captured by the random effects instead of the tree structure.
+Use of random-effects initialization and cluster-level covariances in tandem was less effective in reducing the number of splits than using either of the two approaches.
+
+Distributions of the number of splits, separated according to the levels of the data-generating design are presented and discussed in  Figure~\ref{fig:LMM_sizes_interact} and Appendix~\ref{sec:AppendixA}, because they show a pattern very similar to Figure~\ref{fig:LMM_sizes}.
+Use of cluster-level covariances provides the most accurate split recovery.
+Of the data-generating parameters, $\sigma_{b_1}$ and $N$ showed the strongest effects, with higher values resulting in a higher number of splits.
+The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, and were involved in minor interactions: With the default fitting approach, they had a positive effect on the number of implemented splits; with clustered covariance or random-effects initialization, they had a negative effect on the number of splits.
+The correlation between partitioning variables hardly affected the number of implemented splits.
+
+Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split: Only very rarely is $U_1$ not selected for the first split.
 
 <<echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
-library("xtable")
+library("kableExtra")
 load(file = "splitvars.Rda")
@@
                  dimnames = list(names(probs), "x1"))
+splits[is.na(splits)] <- 0
 for (u_set in list("x2","x3", paste0("x", 4:30), "no split")) {
@@
                 "LongCART"),
-                c("$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
-                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
-                  "$\\sigma_{b_0} = \\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} > 0$",
-                  "$\\sigma_{b_0} > 0$, $\\sigma_{b_1} = 0$"),
-             c("", "clustered cov.", "", "clustered cov.", 
-               "ran.eff. init.", "both", "", "clustered cov.", 
-               "ran.eff. init.", "both",
+                c("$\\hat{\\sigma}_{b_0} = \\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} = \\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} > 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} > 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} > 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} > 0$",
+                  "$\\hat{\\sigma}_{b_0} = \\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} > 0$",
+                  "$\\hat{\\sigma}_{b_0} = \\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} = 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} > 0$",
+                  "$\\hat{\\sigma}_{b_0} > 0$, $\\hat{\\sigma}_{b_1} = 0$"),
+             c("default", "cl.cov.", "default", "cl.cov.", 
+               "ran.eff.", "cl.cov. + ran.eff.", "default", "cl.cov.", 
+               "ran.eff.", "cl.cov. + ran.eff.",
                "LRT", "LRT", "LRT", "score-based", "score-based", "score-based", ""))
@@
 
-print(xtable(splits[1:10, -1], digits = 3,
-             caption = "Variables selected for the first split by each LM(M) tree approach.", 
-             label = "tab:first_splits"), 
-      include.rownames=FALSE, hline.after = c(-1, 0, 2, 6), #table.placement = "tb",
-      add.to.row = list(pos = list(10, 10, 10, 10), command = c(
-        "\\hline",
-        "\\multicolumn{7}{l}{\\footnotesize\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are}\\\\ ", 
-        "\\multicolumn{7}{l}{\\footnotesize continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\\sigma_{b_0}$ and $\\sigma_{b_1}$} are\\\\ ",
-        "\\multicolumn{7}{l}{\\footnotesize the standard deviations of the random intercept and slope, respectively.}\\\\ ")),
-      latex.environment = "", caption.placement = "top",
-      sanitize.text.function=function(x){x})
+## Print the table
+kable_styling(add_footnote(kable(splits[1:10, -1], format = "latex", booktabs = TRUE, 
+                                 label = "first_splits", align = "llccccc", 
+        escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
+        caption = "Variables selected for the first split by each LM tree (top two rows) and LMM tree (bottom eight rows) estimation approach.", row.names = FALSE, digits = 3L),
+  "\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3).
+$\\hat{\\sigma}_{b_0}$ and $\\hat{\\sigma}_{b_1}$ are the estimated standard deviations of the random intercept and slope, respectively.",
+  notation="none", threeparttable = TRUE, escape = FALSE),
+  font_size = 11, full_width=FALSE)
 @
@@
 
-\section{Study II: Comparison of LM(M) trees with SEM trees and LongCART}
-
-In this second study, we compare performance of LM(M) trees with that of SEM trees and LongCART.
-This allows us to evaluate the possible (dis)advantages of global versus local estimation of random-effects parameters, as well as the performance of different splitting criteria.
-The same data-generating design as in Study I was employed.
-We compared performance of LM(M) trees employing clustered covariances only, which showed good performance in Study I.
+\section{Simulation Study II}
+
+Next, we compared performance of LM(M) trees with that of SEM trees and LongCART.
+This allows us to evaluate the possible (dis)advantages of global versus local estimation of random-effects parameters, as well as the performance of the different splitting criteria employed by each method.
+The same data-generating design as in Simulation Study I was employed.
+To reduce the number of comparisons, we only include performance of LM(M) trees fitted using clustered covariances, because they showed good performance in Simulation Study I.
 
@@
 
-We fitted SEM trees using two different splitting criteria: First, we used the default "naive" splitting approach \citep{BranyOert13}, in which likelihood ratio tests (LRTs) are used as the splitting criterion.
-The likelihood of the SEM fitted to the observations in the current node is compared against the likelihood of a range of two-group SEMs, in which the two groups are defined by each possible candidate split.
+We fitted a total of six SEM trees to every dataset. We used two different splitting criteria:
+
+\begin{itemize}
+
+\item We used the default "naive" splitting approach which employs likelihood ratio tests (LRTs) as the splitting criterion \citep{BranyOert13}.
+That is, the likelihood of the SEM fitted to the observations in the current node is compared against the likelihood of a range of two-group SEMs, in which the two groups are defined by each possible candidate split.
 An LRT can thus be computed for each candidate split, which quantifies the improvement in fit that may be obtained by implementing the split.
@@
 
-Second, we fitted score-based SEM trees, as proposed by \cite{ArnoyVoel21}.
+\item We used the score-based splitting approach of \cite{ArnoyVoel21}.
 This approach uses the MOB algorithm described in the Introduction, where the parametric model fitted in step (a) is a SEM.
@@
 
+\end{itemize}
+
+For each splitting approach, three different random-effects specifications were employed: Variances of latent intercepts and slopes both fixed to 0; variance of the latent intercept freely estimated and variance of the latent slope fixed to 0; variances of latent intercepts and slopes both freely estimated.
 To specify the node-specific models for SEM trees, we employed an LGCM specification with the response at each timepoint regressed on a latent intercept and slope.
-Intercept loadings were fixed to 1; slope loadings were fixed to the value of $t$ at each timepoint.
-With both LRT- and score-based splitting criteria, we employed three different model specifications: a) variances of latent slope and intercept fixed to 0; b) variance of latent intercept freely estimated, variance of latent slope fixed to 0; c) variances of latent intercept and slope freely estimated.
-Errors were assumed uncorrelated between timepoints, error variances freely were estimated for each timepoint.
-To fit SEMs, we used package \textbf{lavaan} \citep[version 0.6-11][]{Ross12} and to fit SEM trees, we used package \textbf{semtree} \citep[version 0.9.17][]{BranyOert13}.
-
-We fitted a LongCART tree to each dataset, employing default settings.
+Intercept loadings were fixed to 1; slope loadings were fixed to ${0, 1, 2, 3, 4}$, respectively.
+Errors were assumed uncorrelated between timepoints and an error variance was freely estimated for each timepoint.
+We used package \textbf{lavaan} \citep[version 0.6-11; ][]{Ross12} to fit the SEMs and we used package \textbf{semtree} \citep[version 0.9.17; ][]{BranyOert13} to fit the SEM trees.
+
+We fitted a single LongCART tree to each dataset.
 The LongCART function estimates node-specific models comprising a random intercept term; this default cannot be overridden.
 A fixed-effects model was specified with the response regressed on time and a subject-specific random intercept.
-We used package \textbf{semtree} to fit SEM trees \citep[][version 0.9.17]{BranyOert13}.
-To fit LMMs, we used package \textbf{nlme} \citep[version 3.1-153][]{PinhyBate00} and to fit LongCART trees we used package \textbf{LongCART} \citep[version 3.1][]{Kund21}.
+We used package \textbf{LongCART} \citep[version 3.1][]{Kund21} to fit LongCART trees.
 
 \subsection{Results}
+
+\subsubsection{Tree size}
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!b]
 \caption{Tree size distributions for LM(M) trees with clustered covariances, SEM trees and LongCART.}
-\begin{subfigure}{1.2\textwidth}
-<<echo=FALSE, fig=TRUE, height=3.5>>=
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=4>>=
+## Note!!! White spaces added to LM(M) and SEM tree, b/c with use of geom_text panels get ordered
+## alphabetically
 tmp2$method <- ifelse(
-  tmp2$method_id %in% c("lm_c", "lmer_c", "lmer_s_c"), "LM(M) tree", 
-      ifelse(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s"), "SEM tree (LRT-based)", 
+  tmp2$method_id %in% c("lm_c", "lmer_c", "lmer_s_c"), " LM(M) tree ", 
+      ifelse(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s"), " SEM tree (LRT-based) ", 
              ifelse(tmp2$method_id %in% c("s_sem", "s_sem_i", "s_sem_s"),
-                             "SEM tree (score-based)", "LongCART")))
-tmp2$random <- ifelse(tmp2$method_id %in% c("lm_c", "f_sem", "s_sem"), "no", 
+                             " SEM tree (score-based) ", "LongCART")))
+tmp2$method <- factor(tmp2$method, levels = c(" LM(M) tree ", " SEM tree (LRT-based) ", 
+                                              " SEM tree (score-based) ", "LongCART"))
+
+tmp2$random <- ifelse(tmp2$method_id %in% c("lm_c", "f_sem", "s_sem"), 
+                      "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
                       ifelse(tmp2$method_id %in% c("lmer_c", "f_sem_i", "s_sem_i", "Lcrt"),
-                             "int.", "int. + \nslope"))
-tmp2$random <- factor(tmp2$random, levels = c("no", "int.", "int. + \nslope"))
-tmp2$method <- factor(tmp2$method, levels = c("LM(M) tree", "SEM tree (LRT-based)", 
-                                              "SEM tree (score-based)", "LongCART"))
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
+                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)"))
+
+## Set up data.frame of summary stats
+means <- tapply(tmp2$tree_size, tmp2$method_id, mean)
+sds <- tapply(tmp2$tree_size, tmp2$method_id, sd)
+mads <- tapply(tmp2$tree_size, tmp2$method_id, function(x) mean(abs(x - 3)))
+summ_stats <- data.frame(M = round(means[c("lm_c", "lmer_c", "lmer_s_c", 
+                                            "f_sem", "f_sem_i", "f_sem_s",
+                                            "s_sem", "s_sem_i", "s_sem_s", "Lcrt")], digits = 2), 
+                   SD = round(sds[c("lm_c", "lmer_c", "lmer_s_c", 
+                                            "f_sem", "f_sem_i", "f_sem_s",
+                                            "s_sem", "s_sem_i", "s_sem_s", "Lcrt")], digits = 2),
+                   MAD = sprintf(mads[c("lm_c", "lmer_c", "lmer_s_c", 
+                                            "f_sem", "f_sem_i", "f_sem_s",
+                                            "s_sem", "s_sem_i", "s_sem_s", "Lcrt")],  fmt = '%#.2f'))
+summ_stats$method_id <- factor(rownames(summ_stats))
+
+summ_stats$method <- ifelse(
+  summ_stats$method_id %in% c("lm_c", "lmer_c", "lmer_s_c"), " LM(M) tree ", 
+      ifelse(summ_stats$method_id %in% c("f_sem", "f_sem_i", "f_sem_s"), " SEM tree (LRT-based) ", 
+             ifelse(summ_stats$method_id %in% c("s_sem", "s_sem_i", "s_sem_s"),
+                             " SEM tree (score-based) ", "LongCART")))
+tmp2$method <- factor(tmp2$method, levels = c(" LM(M) tree ", " SEM tree (LRT-based) ", 
+                                              " SEM tree (score-based) ", "LongCART"))
+summ_stats$y <- 25.5
+summ_stats$random <- ifelse(summ_stats$method_id %in% c("lm_c", "f_sem", "s_sem"), 
+                      "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(summ_stats$method_id %in% c("lmer_c", "f_sem_i", "s_sem_i", "Lcrt"),
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
+                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)"))
+summ_stats$all <- with(summ_stats, paste(M, SD, MAD, sep = "\n"))
+
 breaks <- c(0, 1, 3, 7, 15)
-lim <- c(0, 15)
+lim <- c(0, 30)
 plot1 <- ggplot(tmp2) +
-  geom_boxplot(aes(x=random, y=tree_size), alpha = .5, width = .6) + 
+  geom_boxplot(aes(x=random, y=tree_size), alpha = .5, width = .6, coef = NULL) +
+  geom_count(aes(x=random, y=tree_size), col = "grey", alpha = .75) +
   geom_hline(yintercept=3, col = "darkgrey") +
+  geom_text(data = summ_stats, aes(x=random, y=y, label = all),
+            size = 3, col = "grey44") +
   scale_y_continuous(trans = pseudo_log_trans(), 
@@
   facet_grid(~ method, scales = "free", space = "free") +
-  labs(x = "", y = "Tree size")
+  labs(x = "", y = "no. of splits") + theme(legend.position = "none") +
+  scale_x_discrete(labels = scales::parse_format()) +
+  coord_cartesian(clip = 'off') +
+  labs(tag = "  M\n SD\nMAD") + theme(plot.tag.position = c(.03, .87),
+                                   plot.tag = element_text(size = 10))
 plot1
-##ggsave(file = "tree_size_plot.png", height = 9.5, width=19, units = "cm")
-@
-%\includegraphics{tree_size_plot.png}
-\end{subfigure}
-{\footnotesize \textit{Note.} Values on $x$-axis indicate whether random intercepts and/or slopes were estimated: no = random intercepts and slopes fixed to 0; int.
-= variance of random intercept freely estimated, variance of random slope fixed to 0; int.
-+ slope = variance of random intercept and slope (and correlation between the two) freely estimated.
-Distances on $y$-axis are on the log scale.
-Grey lines indicate true number of splits (3 splits).}
+@
+\end{subfigure}%
+\vspace{-0.5cm}%
+{
+\footnotesize \textit{Note.} Grey circles represent counts, grey lines indicate true number of splits.
+Distances on the $y$-axis are on the log scale.
+$x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.
+MAD = Mean absolute deviation from true tree size.}
 \label{fig:tree_sizes}
@@
 Figure~\ref{fig:tree_sizes} depicts tree size distributions for the different algorithms.
-LRT-based SEM trees appear somewhat overpowered with a mean number of splits of \Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["SEM tree (LRT-based)"], digits = 2)} ($SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["SEM tree (LRT-based)"], digits = 2)}), while score-based SEM trees comprised the correct number of splits on average ($M = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["SEM tree (score-based)"], digits = 2)}; $SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["SEM tree (score-based)"], digits = 2)}).
-SEM trees appear more systematically affected by mis-specification of the random-effects structure than LMM trees and under-specification yields a higher number of splits.
-LongCART trees appear underpowered, with a mean number of splits of \Sexpr{round(tapply(tmp2$tree_size, tmp2$method, mean)["LongCART"], digits = 2)} ($SD = $\Sexpr{round(tapply(tmp2$tree_size, tmp2$method, sd)["LongCART"], digits = 2)}).
-This could be the result of under-specification of the random-effects structure: LongCART trees incorporate a random-intercept term only and this default cannot be overridden.
+Comparing the performance of LM(M) trees (leftmost panel) with the performance of SEM trees (middle two panels) indicates that performance of SEM trees is more strongly and systematically affected by the complexity of the random-effects structure than LM(M) trees, with more complex structure yielding lower tree size.
+When random intercepts and slopes are fixed to zero, both LRT- and score based SEM trees tend to implement too many splits.
+With the correct random-effects specification (both random intercepts and slopes freely estimated), LRT-based SEM trees recover the correct tree size on average, but with higher variance than LMM trees.
+Score-based SEM trees perform best when only random intercepts are estimated, implementing too few splits when random slopes are estimated, too.
+LongCART trees appear underpowered, implementing two splits on average.
+
+On average, LMM trees estimating a random intercept and/or slope implement an average number of splits very close to the true number of splits; the same goes for LRT-based SEM trees estimating both random intercept and slope, and score-based SEM trees estimating a random intercept only.
+According to the mean absolute deviation from the true number of splits (MAD), however, LM trees perform best, closely followed by LMM trees estimating both random slope and intercept, LRT-based SEM trees estimating a random intercept only, LMM trees estimating only a random intercept and score-based SEM trees estimating only a random intercept.
 
@@
 # Lcrt     < 2e-16 < 2e-16  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 0.03511
-@
-
-%We performed pairwise t-tests to assess differences between the approaches in Figure~\ref{fig:tree_sizes}.
-Bonferroni-corrected $p$-values indicated significant differences in tree size between most methods (all $p < .001$), with some notable exceptions: The number of splits implemented by LMM trees with clustered covariances was indistinguishable from that of LRT-based SEM trees with freely estimated random intercepts and slopes, and from the score-based SEM trees with freely estimated random intercepts.
-
-<<eval=FALSE, echo=FALSE>>=
-## t-tests for comparing no of splits with true number of splits
-round(apply((sizes-1)/2, 2, function(x) t.test(x, mu = 3)$p.value), digits = 3)
-@
-
-Figure~\ref{fig:tree_sizes_interact} (Appendix \ref{sec:AppendixA}) presents tree size distributions, separated according to the levels of the data-generating parameters.
-Strongest effects were observed for $N$, followed by $\sigma_{b_0}$, $p_{noise}$, $\rho$ and $\sigma_{b_1}$.
-\textcolor{orange}{I moved plots to Appendix because of their size and because they are not of main interest.
-Not sure if following results should even be discussed here or in the Appendix:} Results for $N$ were as expected for all methods: with increasing sample size, more splits are implemented.
-Both LRT- and score-based SEM trees seem only affected by levels of $\sigma_{b_0}$, $p_{noise}$ and $\rho$ under misspecification of the random effects.
-Especially when both random intercept and slope variances are fixed to 0, higher levels of $\sigma_{b_0}$ and $p_{noise}$ yield more splits with both SEM tree approached.
-LRT-based SEM trees seem unaffected by levels of $\rho$, while score-based SEM trees seemed to implement a larger number of splits with increasing values of $\rho$.
-This pattern seemed reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects were correctly specified.
-LongCART implemented more splits with higher levels of $p_{noise}$ and $\rho$, but was unaffected by levels of $\sigma_{b_0}$ and $\sigma_{b_1}$.
+
+## t-tests for comparing no. of splits with true number of splits
+t_tests <- round(apply((sizes-1)/2, 2, function(x) t.test(x, mu = 3)$p.value), digits = 3)
+t_tests
+@
+
+Distributions of the number of splits, separated according to the levels of the data-generating parameters are depicted and discussed in Appendix~\ref{sec:AppendixA} and Figure~\ref{fig:tree_sizes_interact}, because they show a pattern very similar to Figure~\ref{fig:tree_sizes}.
+Of the data-generating parameters, $N$ and $\sigma_{b_0}^2$ showed the strongest effects and in the expected direction with higher values resulting in a higher number of splits.
+The number of splits implemented by SEM trees was most strongly affected by the data-generating parameters when the random effects were misspecified (i.e., random intercepts and/or slope fixed to 0).
 
@@
 
+\subsubsection{Split selection}
+
 <<echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
+library("kableExtra")
 splits$Algorithm[2] <- "LM(M) tree"
-splits$Algorithm[4] <- "  (clustered"
-splits$Algorithm[8] <- "    covariances)"
+splits$Algorithm[4] <- "  (cl.cov.)"
+splits$Algorithm[8] <- " "
 splits$Algorithm[12] <- "  (LRT-based)"
@@
 splits$Algorithm[15] <- "  (score-based)"
-print(xtable(splits[c(2, 4, 8, 11:17), -3], digits = 3,
-             caption = "Variable selected for the first split by each of the partitioning methods.", 
-             label = "tab:first_splits2"), include.rownames=FALSE, 
-      hline.after = c(-1, 0, 3, 6, 9), table.placement = "!ht",
-      add.to.row = list(pos = list(10, 10, 10, 10), command = c(
-        "\\hline",
-        "\\multicolumn{7}{l}{\\footnotesize\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables}\\\\ ", 
-        "\\multicolumn{7}{l}{\\footnotesize are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\\sigma_{b_0}$ and $\\sigma_{b_1}$}\\\\ ",
-        "\\multicolumn{7}{l}{\\footnotesize are the standard deviations of the random intercept and slope, respectively.}\\\\ ")),
-      latex.environment = "", caption.placement = "top",
-      sanitize.text.function=function(x){x})
-
+
+## Print table
+kable_styling(add_footnote(
+  kable(splits[c(2, 4, 8, 11:17), -3], format = "latex", booktabs = TRUE, 
+        label = "first_splits2", align = "llccccc", row.names = FALSE, digits = 3,
+        escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
+        caption = "Cross-validated mean squared errors for each of the response variables."),
+  "\\footnotesize \\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3).
+The first column indicates wheter the random intercept and/or slope were estimated or not.",
+  notation="none", threeparttable = TRUE, escape = FALSE),
+  font_size = 11, full_width=FALSE)
 @
@@
 For SEM trees, the LRT criterion yields almost perfect accuracy for the first split.
-The score-based approach selected the wrong splitting variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$U_1$`[grep("s_sem", rownames(splits))])))}\% of datasets, but only when the random-effects structure was mis-specified (i.e., $\sigma_{b_0}$ and/or $\sigma_{b_1}$ fixed to 0).
+The score-based approach selected the wrong variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$U_1$`[grep("s_sem", rownames(splits))])))}\% of datasets, but only when the random-effects structure was misspecified.
 When random effects were correctly specified, score-based SEM trees also provided perfect recovery of the first split.
-Closer inspection of parameter stability tests suggested that the tests are (much) more sensitive to instability in the fixed slope than in the fixed intercept.
-
-LongCART trees exhibit strikingly low accuracy for recovering the first split, selecting the wrong variable in all datase.
-LongCART showed a more dramatic tendency than score-based SEM trees to use $U_2$ and $U_3$ for implementing the first split.
-Of note, in the LongCART trees where no split was implemented (\Sexpr{round(100*prop.table(table(tree_size$Lcrt))[1])}\% of datasets), \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not exceed the criterion of significance in those datasets.
-This suggests that either the parameter stability tests proposed by \cite{KundyHare19} are less sensitive to instabilities with respect to the fixed intercept, and/or with respect to categorical covariates.
+Closer inspection of stability tests seperately per model parameter suggested that the the score-based tests for SEMs are more sensitive to instability in the fixed slope than in the fixed intercept.
+
+LongCART trees exhibit low accuracy for recovering the first split, selecting the wrong variable in all datasets where at least one split was implemented.
+LongCART showed an even stronger tendency than score-based SEM trees to select $U_2$ or $U_3$ for the first split.
+Of note, in the datasets where LongCART did not implement any splits, in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not reach the criterion for significance.
+This suggests the parameter stability tests proposed by \cite{KundyHare19} are less sensitive to instabilities with respect to the fixed intercept, or less sensitive to instability with respect to a categorical covariate.
 
@@
 
+\subsubsection{Computation time}
+
 \begin{figure}[!ht]
 \caption{Computation time distributions for the different partitioning methods.}
-\begin{subfigure}{.6\textwidth}
-<<fig=TRUE, echo=FALSE, width=4, height=4>>=
+\begin{subfigure}{.8\textwidth}
+<<fig=TRUE, echo=FALSE, width=4, height=3.25>>=
 load("comp_times.Rda")
@@
 times <- reshape2::melt(times)
-ggplot(times, aes(Var2, value)) + geom_boxplot(width=.5, alpha = .5) + 
+mean_times <- tapply(times$value, times$Var2, mean)
+summ_stats <- data.frame(M = sprintf(mean_times, fmt = '%#.2f'),
+                         Var2 = levels(times$Var2),
+                         y = 100000)
+ggplot(times, aes(Var2, value)) + 
+  geom_violin(, alpha = .5, fill = "grey", width=.5) +
   labs(x = "", y = "Computation time (in seconds)") +
-  scale_y_continuous(trans = "log", breaks = c(.01, .1, 1, 10, 100, 1000, 10000))
-mean_times <- tapply(times$value, times$Var2, mean)
+  geom_text(data = summ_stats, aes(x=Var2, y=y, label = M),
+            size = 3, col = "black") +
+  scale_y_continuous(trans = "log", limits = c(0.01, 100000), 
+                     breaks = c(.01, .1, 1, 10, 100, 1000, 10000)) +
+  labs(tag = "M") + theme(plot.tag.position = c(.12, .96),
+                          plot.tag = element_text(size = 10)) + 
+  geom_text(data = summ_stats, aes(x=Var2, y=y, label = M),
+            size = 3, col = "darkgrey") 
 @
@@
 Figure~\ref{fig:comp_times} presents computation time distributions for the partitioning algorithms.
-A clear computational advantage is observed for LM trees, with an average computation time of \Sexpr{round(mean_times[" LM tree"], digits = 3)} seconds.
-Including random-effects estimation increased computation time: for LMM trees, average computation time was \Sexpr{round(mean_times["LMM tree"], digits = 2)} seconds.
-LongCART and SEM trees yielded much longer computation times, with LongCART requiring \Sexpr{round(mean_times["LongCART"], digits = 2)} seconds, score-based SEM trees \Sexpr{round(mean_times["SEM tree\n(score)"], digits = 2)} seconds, and LRT-based SEM trees \Sexpr{round(mean_times["SEM tree\n(LRT)"], digits = 2)} seconds computation time, on average.
+A clear computational advantage is observed for LM trees.
+LMM trees require longer computation times because of the estimation of random effects.
+Yet, this increase is minor compared to the computation times required by LongCART and SEM trees: LRT-based SEM trees required strikingly computation times, while score-based SEM trees were computationally more efficient, as expected.
+Computation times for LongCART were in between LRT- and score-based SEM trees.
 
@@
 
-\section{Study III: Application}
+\section{Application}
 
@@
 We analyzed longitudinal assessments of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998-1999 \citep[ECLS-K; ][]{NCES10}.
-Data were collected from kindergarten 1998 through eighth grade 2007.
-In the current analyses, we focus on assessments during kindergarten, 1st, 3rd, 5th and 8th grade.
-In the ECLS-K study, data from 21,304 children from 1,018 schools across the USA were collected.
-
-Response variables are reading, math and science ability.
+Data were collected from 21,304 children from 1,018 schools across the USA were collected.
+Assessments took place from kindergarten in 1998 through 8th grade in 2007.
+In the current analyses, we focus on the assessments from kindergarten, 1st, 3rd, 5th and 8th grade.
+
+Response variables are reading, math and science abilities.
 These were assessed using multi-item cognitive tests, from which latent ability estimates were computed with a mean of zero and variance of one.
 Reading and math abilities were assessed in all five rounds of data collection, while science knowledge was assessed in the 3rd, 5th and 8th grade only.
-We analyzed data from  children who completed all assessments ($N = 6,277$ for reading; $N = 6,512$ for math; $N = 6,625$ for science).
-
-Our analyses focus on predicting ability trajectories, based on baseline characteristics.
+We analyzed data from children who completed all assessments yielding $N = 6,277$ for reading; $N = 6,512$ for math; $N = 6,625$ for science.
+
+Our analyses focus on predicting trajectories of reading, math and science abilities, based on baseline characteristics.
 Time was measured as the number of months since the baseline assessment.
-In order to obtain approximately linear trajectories, used $\sqrt{\mathrm{months}}$ as the timing metric for reading and math trajectories, and $\mathrm{months}^{\frac{2}{3}}$ for science trajectories.
-
-We used the following time-invariant covariates as potential partitioning variables:
-
-\begin{itemize}
-\setlength\itemsep{0em}
-\setlength{\itemindent}{0.2in}
-\item Gender (51.1\% male)
-\item Age in months at baseline (range 53 to 96; M = 6.14 years at baseline).
-\item Race (8 categories)
-\item First time in kindergarten (yes/no)
-\item Socio-economic status (range -5 to 3)
-\item Fine motor skills (e.g., drawing figures; range 0 to 9)
-\item Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8)
-\item Interpersonal skills (range 1 to 4)
-\item Self-control (range 1 to 4)
-\item Internalizing problem behavior (range 1 to 4)
-\item Externalizing problem behavior (range 1 to 4)
-\end{itemize}
+In order to obtain approximately linear trajectories, $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.
+
+We used eleven time-invariant covariates as potential partitioning variables: Gender (51.1\% male); Age in months at baseline (range 53 to 96; M = 6.14 years at baseline); Race (8 categories); First time in kindergarten (yes/no); Socio-economic status (range -5 to 3); Fine motor skills (e.g., drawing figures; range 0 to 9); Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8); Interpersonal skills (range 1 to 4); Self-control (range 1 to 4); Internalizing problem behavior (range 1 to 4); Externalizing problem behavior (range 1 to 4).
 
@@
 
-We fitted four LM(M) trees, employing the same approaches applied in Studies I and II, except for the default approaches, and the random-effects initialization approach when both random intercepts and slopes were estimated; approaches which showed far from optimal performance in Study I.
-
-For comparison, we fitted longRPart \citep{AbdoyLeBl02} trees.
-We opted for longRPart, because neither SEM tree, LongCART, longRPart or longRPart2 implementations provide functionality to obtain predictions for new observations.
-We wrote custom functionality to allow for prediction with longRPart trees, but could not produce this for the other methods.
-We expect, however, that performance of longRPart will be representative of the other methods, in the sense that all fit the full mixed-effects model locally.
-Furthermore, SEM tree could not be employed because assessments at each wave did not take place on identical timepoints, which is a requirement for SEM-based growth curve modeling.
-
-Using package \textbf{longRPart} \citep[version 1.0][]{StewyAbdo12}, we fitted two longRPart trees with different random effects specifications: One tree comprising a random intercept only, and one tree comprising both a random intercept and slope, with respect to student.
+We fitted four LM(M) trees using the approaches which showed most promise in the simulations: An LM tree ($\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$) using clustered covariances in the parameter-stability tests, two LMM trees using clustered covariances with different random-effects specifications ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$ and $\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$), and an LMM tree initializing estimation with the random effects and freely estimating a random intercept only ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$).
+
+To compare the performance of LM(M) trees on existing data, we could not employ SEM tree or LongCART, so we opted for longRPart \citep{AbdoyLeBl02}.
+Neither SEM trees, LongCART, longRPart or longRPart2 implementations provide functionality to obtain predictions for new observations.
+Writing custom functionality to allow for prediction was most straightforward for longRPart.
+Furthermore, performance of longRPart will be representative of the other methods, in the sense that all four methods fit full mixed-effects models locally.
+Like LongCART and longRPart2, longRPart fits a mixed-effects model in each of the terminal nodes.
+Like LRT-based SEM trees, longRPart uses an exhaustive LRT-based search in every node to find the optimal next split.
+
+Using package \textbf{longRPart} \citep[version 1.0; ][]{StewyAbdo12}, we fitted two trees with different random effects specifications: A tree comprising a random intercept only, and a tree comprising both a random intercept and slope with respect to student.
 In order to implement a split, longRPart uses a criterion based on the minimum decrease in deviance, which is set to 0.01, by default.
@@
 We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = \Sexpr{2^5} terminal nodes, or \Sexpr{(2^5)-1} splits.
-Significance of the LRT, or a cross-validation approach to determine whether to implement a split or not may yield better results, but neither approach is implemented in longRPart (nor longRPart2).
+Significance of the LRT, or a cross-validation approach to determine whether to implement a split or not may yield better results, but this has not been implemented.
 
@@
 
-The ECLS-K datasets have exceptionally large sample sizes, which may not be representative of real-world applications.
-We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, and evaluated predictive accuracy using the responses of children not included in the training sample.
-For each combination of response variable and sample size, we performed 100 repetitions.
-Because we used disjunct samples of children for training and testing, only the fixed-effects parts of each fitted model were used for prediction; random effects were fixed to 0.
-Accuracy was assessed by computing the test MSE, by taking the mean squared difference between predicted and observed ability estimates of test observations.
-We measured tree size by counting the number of splits in each tree, and measured computation time in seconds.
+The ECLS-K datasets have exceptionally large sample sizes.
+We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, likely more representative of real-world studies in psychology.
+For each combination of response variable (math, reading or science) and sample size, we performed 100 repetitions.
+We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample of the current repetition.
+Note that this measures the accuracy of the fixed-effects predictions only.
+Tree size was measured by the number of splits in each tree, and computation time was measured in seconds.
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!bt]
 \caption{Mean squared errors for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
@@
 
+## Compute summary stats
+MSE_df <- data.frame(t(sapply(math_MSEs[ , -c(4,6,7,10)], 
+                              function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(math_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE)))))
+MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE)))))
+summ_stats <- MSE_df[ , c(1:2,5:6,9:10)]
+names(summ_stats) <- c("math, N = 250", "math, N = 1,000", 
+                       "read, N = 250", "read, N = 250", 
+                       "scie, N = 250", "scie, N = 1,000")
+
 tmp3 <- data.frame(stack(math_MSEs[ , -c(4,6,7,10)]),
@@
 tmp3$method2 <- as.character(tmp3$method)
-tmp3$method2[grepl("LMtree", tmp3$method2)] <- "LM tree\n\n(cluster)"
-tmp3$method2[tmp3$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
-tmp3$method2[tmp3$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
-tmp3$method2[tmp3$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
+tmp3$method2[grepl("LMtree", tmp3$method2)] <- "LM tree\n\n(cl.cov.)"
+tmp3$method2[tmp3$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cl.cov.)"
+tmp3$method2[tmp3$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
+tmp3$method2[tmp3$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cl.cov.+ran.eff.)"
 tmp3$method2[grepl("longR", tmp3$method2)] <- "longRPart"
-tmp3$random <- ifelse(tmp3$method == "LMtree_c", "no", ifelse(
-  tmp3$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+tmp3$random <- ifelse(tmp3$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp3$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
                              ifelse(tmp3$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-    "int. + \nslope", NA)))
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
 tmp3$N <- factor(tmp3$N)
@@
 ggplot(tmp3) +
-  geom_boxplot(aes(x=random, y=MSE, fill = N), 
-               position=position_dodge(1), alpha = .5, width = .6) +
+  geom_violin(aes(x=random, y=MSE, fill = N), 
+               position=position_dodge(1), alpha = .5, width = .7) +
   scale_y_continuous(trans = log_trans(),
@@
 tmp4$method2[grepl("longR", tmp4$method2)] <- "longRPart"
-tmp4$random <- ifelse(tmp4$method == "LMtree_c", "no", ifelse(
-  tmp4$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+tmp4$random <- ifelse(tmp4$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp4$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
                              ifelse(tmp4$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-    "int. + \nslope", NA)))
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
 tmp4$N <- factor(tmp4$N)
 ggplot(tmp4) +
-  geom_boxplot(aes(x=random, y=MSE, fill=N), 
-               position=position_dodge(1), alpha = .5, width = .6) +
+  geom_violin(aes(x=random, y=MSE, fill=N), 
+               position=position_dodge(1), alpha = .5, width = .7) +
   scale_y_continuous(trans = log_trans(),
@@
 tmp5$method2[grepl("longR", tmp5$method2)] <- "longRPart"
-tmp5$random <- ifelse(tmp5$method == "LMtree_c", "no", ifelse(
-  tmp5$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+tmp5$random <- ifelse(tmp5$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp5$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
                              ifelse(tmp5$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-    "int. + \nslope", NA)))
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
 tmp5$N <- factor(tmp5$N)
-ggplot(tmp5) +
-  geom_boxplot(aes(x=random, y=MSE, fill=N), 
-               position=position_dodge(1), alpha = .5, width = .6) +
+library("ggforce")
+ggplot(tmp5, aes(x=random, y=MSE)) +
+  geom_violin(aes(fill=N), 
+               position=position_dodge(1), alpha = .5, width = .7) +
   scale_y_continuous(trans = log_trans(),
@@
   theme(strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(x = "", y = "MSE")
-@
-\end{subfigure}
-{\footnotesize \textit{Note. }The $y$-axis on the right gives an estimate of the proportion of variance explained, computed as $1 - \frac{MSE}{var(y)}$. Distances on $y$-axis are on the log scale.}
+  labs(x = "", y = "MSE") +
+  scale_x_discrete(labels = scales::parse_format()) +
+  coord_cartesian(clip = 'off') 
+@
+\end{subfigure}%
+\vspace{-0.5cm}%
+{\footnotesize \textit{Note.
+}The second $y$-axis (far right) gives an estimate of the proportion of variance explained, computed as $1 - \frac{MSE}{var(y)}$.
+Distances on $y$-axis are on the log scale.}
 \label{fig:application_MSEs}
@@
 
+\setlength{\tabcolsep}{3pt}
+
 <<application_MSE, echo=FALSE, results=tex>>=
 library("kableExtra")
-MSE_df <- data.frame(t(sapply(math_MSEs[ , -c(4,6,7,10)], 
-                              function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(math_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE)))))
 MSE_df <- cbind(paste0(MSE_df[,1], " (", MSE_df[,3], ")"),
@@
                 paste0(MSE_df[,10], " (", MSE_df[,12], ")"))
-rownames(MSE_df) <- c("LM tree (c)", "LMM tree (i,c)", "LMM tree (i,r)", "LMM tree (i,s,c)",  
-                      "longRPart (i)", "longRPart (i,s)")
+rownames(MSE_df) <- c("LM tree$^c$", "LMM tree$^{i,c}$", "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$",  
+                      "longRPart$^i$", "longRPart$^{i,s}$")
 MSE_df <- MSE_df[c(1,2,4,3,5,6), ]
@@
   c(" ", "Math" = 2, "Reading" = 2, "Science" = 2), align = "c"),
-  font_size = 8, full_width=FALSE)
+  font_size = 9, full_width=FALSE)
 @
@@
 
-## Illustrate by refiting trees:
+## Illustrate by refitting trees:
 load("bag_ids, N = 250, Math.Rda")
@@
 ## for longRPart and longRPart_s are quite similar, and seem to differ
-## most for Science, which was most strongly affected bu the mistake.
-
-@
-
-Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions for each of the partitioning approaches, separated by sample size\footnote{LongRPart trees comprising random intercepts and slopes ran into estimation errors in some datasets: In the Math ability data, \Sexpr{table(is.na(math_MSEs$longRPart_s))["TRUE"]} out of 200 datasets did not yield a tree.
-All tree estimations converged successfully in the Reading ability data.
-In the Science ability data, \Sexpr{table(is.na(scie_MSEs$longRPart_s))["TRUE"]} out of 200 datasets did not yield a tree.
-Results are presented for iterations did yield a tree.}.
-Differences in performance between the partitioning approaches are of smaller magnitude than differences due to sample size.
-All methods show higher accuracy with larger sample sizes (with one exception: longRPart trees comprising random intercepts and slopes fitted on science ability data).
-
+## most for Science, which was most strongly affected by the mistake.
+
+@
+
+\subsubsection{Predictive accuracy}
+
+LongRPart trees comprising both random intercepts and slopes yielded estimation errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the Math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the Science data.
+All other model-fitting attempts converged successfully.
+Results are presented for successful model estimations.
+
+Figure~\ref{fig:application_MSEs} presents MSE distributions for each of the partitioning approaches, separated by sample size.
+As expected, all methods show higher accuracy with larger sample sizes.
 LM(M) trees tend to show better predictive accuracy than longRPart trees.
 Pairwise $t$-tests indicated significantly better performance for each of the LM(M) tree approaches employing clustered covariances, compared to both longRPart approaches (Bonferroni corrected $p$-values all $< .01$).
-However, it should be noted that differences in performance are small; for example, the standard errors presented in Table~\ref{tab:application_MSEs} indicate differences $> 1$ \textit{SE} only for Reading with $N=250$, and Science (both $N=250$ and $N=1,000$)\footnote{The significance of pairwise $t$-tests may seem in contrast with the but small differences in terms of standard errors presented in Table~\ref{tab:application_MSEs}, but the standard errors in the pairwise $t$-tests are sensitive to the number of replications, while the standard errors in Table~\ref{tab:application_MSEs} are not, and thus provide a more useful measure of the strength of the effect.}.
-
-Table~\ref{tab:application_MSEs} indicates that LMM trees comprising a random intercept term and employing cluster-level covariances ("LMM tree (i,c)") performed best for predicting Math and Reading abilities.
-For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best, while this was the worst-performing approach for math and reading.
-For longRPart, a somewhat similar shift in performance occurs: While trees comprising both a random intercept and slope performed best for predicting Math and Reading, trees with a random intercept only performed best for predicting Science.
-This shift may be due to the lower number of Science ability assessments, compared to Math and Reading.
-
-Interestingly, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with(out) random intercepts or slopes.
-Pairwise $t$-tests did not indicate significant differences between the three approaches for any of the response variables.
+However, these performance differences are small; the effects of sample size are much more obvious than the differences between methods in Figure~\ref{fig:application_MSEs}.
+
+Table~\ref{tab:application_MSEs} presents means and standard deviations of the MSEs.
+Only few differences in performance between LM(M) and longRPart trees are larger $> 1$ standard error: For Reading with $N=250$ and for Science with both sample sizes.
+Note that the standard errors in the pairwise $t$-tests are get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not.
+The latter thus provide a more useful measure of effect size, while the former have higher power to detect differences between methods.
+
+Table~\ref{tab:application_MSEs} indicates that LMM trees estimating only a random intercept and employing cluster-level covariances performed best for predicting Math and Reading abilities.
+For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best.
+Random-effects initialization was the worst-performing LMM tree approach for Math and Reading.
+
+For longRPart, trees comprising both a random intercept and slope performed best for predicting Math and Reading, trees with a random intercept only performed best for predicting Science.
+The differences in performance between Math and Reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of Science abilities.
+
+Again, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with random intercepts, slopes or neither.
+Pairwise $t$-tests did not indicate significant differences between those three approaches for any of the response variables.
 
@@
 
-\begin{figure}[!b]
-\caption{Tree sizes for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
+\begin{figure}[!bt]
+\caption{Tree sizes for trees fitted to Math (top), Reading (middle) and Science (bottom) ability trajectories.}
 \begin{subfigure}{1.2\textwidth}
@@
 tmp6$method2[grepl("longR", tmp6$method2)] <- "longRPart"
-tmp6$random <- ifelse(tmp6$method == "LMtree_c", "no", ifelse(
-  tmp6$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+tmp6$random <- ifelse(tmp6$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp6$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
                              ifelse(tmp6$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-    "int. + \nslope", NA)))
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
 tmp6$N <- factor(tmp6$N)
 ggplot(tmp6) +
-  geom_boxplot(aes(x=random, y=splits, fill = N), 
-               position=position_dodge(1), alpha = .5, width = .6) +
+  geom_violin(aes(x=random, y=splits, fill = N), 
+               position=position_dodge(1), alpha = .5, width = .7) +
   facet_grid(~method2, scales = "free", space = "free") +
   theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
-  labs(x = "", y = "# of splits")
+  labs(x = "", y = "no. of splits")
 @
@@
 tmp7$method2[grepl("longR", tmp7$method2)] <- "longRPart"
-tmp7$random <- ifelse(tmp7$method == "LMtree_c", "no", ifelse(
-  tmp7$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+tmp7$random <- ifelse(tmp7$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp7$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
                              ifelse(tmp7$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-    "int. + \nslope", NA)))
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
 tmp7$N <- factor(tmp7$N)
 ggplot(tmp7) +
-  geom_boxplot(aes(x=random, y=splits, fill=N), 
-               position=position_dodge(1), alpha = .5, width = .6) +
+  geom_violin(aes(x=random, y=splits, fill=N), 
+               position=position_dodge(1), alpha = .5, width = .7) +
   facet_grid(~method2, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits") +
+  labs(x = "", y = "no. of splits") +
   theme(strip.background = element_blank(), strip.text.x = element_blank(),
@@
 tmp8$method2[grepl("longR", tmp8$method2)] <- "longRPart"
-tmp8$random <- ifelse(tmp8$method == "LMtree_c", "no", ifelse(
-  tmp8$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), "int.", 
+tmp8$random <- ifelse(tmp8$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp8$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
                              ifelse(tmp8$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-    "int. + \nslope", NA)))
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
 tmp8$N <- factor(tmp8$N)
 ggplot(tmp8) +
-  geom_boxplot(aes(x=random, y=splits, fill=N), 
-               position=position_dodge(1), alpha = .5, width = .6) +
+  geom_violin(aes(x=random, y=splits, fill=N), 
+               position=position_dodge(1), alpha = .5, width = .7) +
   facet_grid(~method2, scales = "free", space = "free") +
   theme(strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(x = "", y = "# of splits")
+  labs(x = "", y = "no. of splits") +
+  scale_x_discrete(labels = scales::parse_format()) +
+  coord_cartesian(clip = 'off')
 @
@@
 \begin{subfigure}{.5\textwidth}
-<<echo=FALSE, fig=TRUE, height=3, width=3>>=
+<<echo=FALSE, fig=TRUE, height=3.25, width=3>>=
 load("times, N = 250, Reading.Rda")
@@
 scie_times <- cbind(times, N = 250)
-#load("times, N = 1000, Science.Rda")
-#scie_times <- rbind(scie_times, cbind(times, N = 1000))
+load("times, N = 1000, Science.Rda")
+scie_times <- rbind(scie_times, cbind(times, N = 1000))
 
@@
 ECLSK_times$ind <- factor(ECLSK_times$ind)
+
+## Compute summary stats
+mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
+summ_stats <- data.frame(M = sprintf(mean_times, fmt = '%#.2f'),
+                         ind = names(mean_times),
+                         y = 11000)
+## Plot
 ggplot(ECLSK_times) +
-  geom_boxplot(aes(x=ind, y=values), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = "log", 
-                     breaks = c(.01, .1, 1, 10, 100, 1000, 10000)) +
-  labs(x = "", y = "Computation time (in seconds)")
-mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
+  geom_violin(aes(x=ind, y=values), alpha = .5, fill = "grey", width=.7) +
+  scale_y_continuous(trans = "log", limits = c(0.1, 12000),
+                     breaks = c(.01, .1, 1, 10, 100, 1000)) +
+  labs(x = "", y = "Computation time (in seconds)") +
+  geom_text(data = summ_stats, aes(x=ind, y=y, label = M),
+            size = 3, col = "grey44") +
+  labs(tag = "M") + theme(plot.tag.position = c(.12, .95),
+                          plot.tag = element_text(size = 10)) 
 @
@@
 
-Figure~\ref{fig:application_sizes} present tree size distributions for each of the partitioning approaches.
-Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern appears reversed for longRPart trees.
-For math and reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small ($N=250$).
-This pattern reverses with large sample size ($N=1,000$), then the best-performing LM(M) tree approach yields a larger number of splits than the best-performing longRPart approach.
-For the math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
-
-Given the very minor differences in performance for LM(M) tree employing clustered covariances with different specifications of the random effects (Figure~\ref{fig:application_MSEs}), Figure~\ref{fig:application_sizes} suggests that LM trees with clustered covariances may provide very good complexity-accuracy trade-off.
-
-Figure~\ref{fig:application_times} depicts the computation time distributions of the different partitioning algorithms.
-A clear computational advantage is observed for LM trees, with an average computation time of \Sexpr{round(mean_times["LM tree"], digits = 2)} seconds.
-Including estimation of random effects increases computation time, which was \Sexpr{round(mean_times["LMM tree"], digits = 2)} seconds on average for LMM trees.
-LongRPart trees required longest computation times, with an average of \Sexpr{round(mean_times["longRPart"], digits = 2)} seconds.
+Figure~\ref{fig:application_sizes} presents tree size distributions for each of the partitioning approaches.
+Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees.
+For Math and Reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small.
+This pattern reverses with large sample size, where the best-performing LM(M) tree approach yields a larger number of splits than the best-performing longRPart approach.
+For the Math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
+
+\subsubsection{Computation time}
+
+Figure~\ref{fig:application_times} depicts computation time distributions of the different partitioning algorithms.
+A clear computational advantage is observed for LM trees ($M =$ \Sexpr{round(mean_times["LM tree"], digits = 2)} seconds).
+LMM trees require longer computation times ($M =$ \Sexpr{round(mean_times["LMM tree"], digits = 2)} seconds).
+This increase is minor compared to LongRPart trees, which required longest computation times ($M =$ \Sexpr{round(mean_times["longRPart"], digits = 2)} seconds).
 
@@
 
-In Study I, we found the default estimation approach for GLMM trees proposed in \cite{FokkySmit18} yields high Type-I error rates when partitioning LGCMs.
-As hypothesized, performing parameter-stability tests using clustered covariances strongly improved performance.
-Also as hypothesized, we found initializing estimation with the random effects to be beneficial, but only when the random-effects specification comprised a random intercept only.
-Combining cluster-level covariances and random-effects initialization was not beneficial.
-The performance of of LM(M) trees using clustered covariances appeared largely unaffected by (mis-)specification of the random effects.
-
-In Study II, we found good performance of SEM trees in partitioning LGCMs, but a stronger sensitivity to mis-specification of the random effects, where underspecification yielded higher Type-I errors.
-In accordance with results of \cite{ArnoyVoel21}, we found score-based SEM trees to have somewhat lower power than LRT-based SEM trees.
-LongCART tended to be underpowered, which could be due to the trees not incorporating random slopes, or the parameter stability test for categorical covariates proposed by \cite{KundyHare19} may be underpowered.
-
-In Study III, we found higher predictive accuracy for LM(M) trees than for longRPart, but differences in performance were small.
-Similar to Studies I and II, we found LM(M) trees to be comparably insensitive to (mis-)specification of the random effects, which may provide an important practical advantage.
-
-All three studies showed substantial computational advantage of GLM(M) trees.
-This is likely due to the estimation approach employed by GLMM trees, where fixed-effects parameters are estimated locally within a node while random-effects parameters are estimated globally, using all observations.
+In line with expectations, the simulation study showed that the default estimation approach for GLMM trees yields too many splits when partitioning LGCMs.
+As hypothesized, performing parameter-stability tests using clustered covariances mitigated this overfitting.
+Also as expected, initializing estimation with the random effects mitigated overfitting, but too strongly when both random intercepts and slopes were estimated.
+The performance of of LM(M) trees using clustered covariances appeared largely unaffected by (mis-)specification of the random effects, and could therefore be the safest choice in practice.
+Combining cluster-level covariances and random-effects initialization worsened performance, os in practice users should use either one of the two approaches.
+
+The simulations showed comparable performance of LM(M) and SEM trees in partitioning LGCMs.
+SEM trees did however show stronger sensitivity to misspecification of the random effects, with underspecification yielding a higher number of splits.
+Furthermore, in line with results of \cite{ArnoyVoel21}, we found score-based SEM trees to have somewhat lower power than LRT-based SEM trees, but computationally much less burdensome.
+LongCART trees often selected the wrong partitioning variable for the first split, and were outperformed by LM(M) and SEM trees.
+The LongCART parameter stability tests \citep{KundyHare19} may be underpowered for detecting systematic instability of the fixed intercept, or for detecting instability with respect to categorical covariates.
+
+In the application, we found higher predictive accuracy for LM(M) trees than for longRPart, but differences in performance were small.
+In line with the simulation studies, we found performance of LM(M) trees did not depend on (mis-)specification of the random effects, which may provide an important practical advantage.
+
+Both simulation and application studies indicated a substantial computational advantage of GLM(M) trees.
+This is likely due to the local-global estimation approach employed by GLMM trees, where fixed-effects parameters are estimated locally within a node and random-effects parameters are estimated globally, using all observations.
 In contrast, SEM trees, LongCART and longRPart fit the full mixed-effects model in each node, which may substantially increase computational burden.
-
-\subsection{Possible points for further discussion}
-
-\begin{itemize}
-
-\item Lower performance of longRPart possibly due to lack of robust criterion for determining statistical significance of a split.
-Cross-validation procedures might mitigate this problem, but feasibility of such approaches currently impeded by lack of functionality for generating predictions and high computational burden.
-
-\item MELT \citep{EoyCho14} employs random-effects predictions as a partitioning criterion, which is effective, as our results on random-effects initialization illustrate that random effects can capture effects of substantive interest that we want to use for explanation.
-
-\item Future work:
-
-\begin{itemize}
-\item GLMM trees can be extended to allow for partitioning based on all mixed-effects model parameters simultaneously, by using the score-based tests for mixed-effects models proposed by \cite{WangyMerk18}.
-\item GLMM trees can be extended to allow for partitioning non-linear growth curve models / GAMs. 
-\item Performance of GLMM trees on non-gaussian responses (binomial or count responses) should be evaluated.
-\end{itemize}
-\end{itemize}
+The local-global estimation approach also reduces model complexity and the good performance of GLMM trees in the current study suggests this may be beneficial for stability and generalizability.
+
+Yet, a possible downside of the local-global estimation approach is that it does not allow for recovering subgroups with differences in random-effects parameters.
+When there is a specific interest in partitioning the random effects, score-based SEM trees should be preferred.
+Alternatively, researchers may want to use the parameter stability tests for mixed-effects models developed by \cite{WangyMerk18} \citep[see also ][]{WangyMerk21,WangyGrav22}.
+This will be useful, for example, when the number and/or distances between timepoints differs between respondents and SEM-based growth curve models can thus not be applied \cite{NeisyMatt18}.
+
+The focus of the current study was limited to Gaussian responses and LGCMs.
+Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses.
+We expect that the good performance of clustered covariances in the parameter stability tests generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures.
+Yet, whether our conclusions generalize to settings beyond LGCMs should be assessed in future studies.
+Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances.
+Future studies could assess whether computation and use of the full sandwich estimator further improves performance of the parameter stability tests.
 
@@
 
+\newpage
 \appendix
@@
 For the data-generating parameters, the effects on tree size were strongest for $\sigma_{b_0}$, followed by $N$, $p$, $\sigma_{b_1}$ and $\rho$.
-Increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ tend to yield higher numbers of splits, and the effect of $\rho$ is minimal.
-The main exception to these main effects is that increased values of $\sigma_{b_0}$ and $\sigma_{b_1}$ yield lower numbers of splits when random slopes are estimated, and estimation is initialized with the random effects.
-In these cases, most variance may be captured by the random effects already in the first iteration of the algorithm, and will no longer be picked up by splits in the tree.
-Also, increased values of $p$ and $\sigma_{b_1}$ yield a slightly lower number of splits when cluster-level covariances are employed.
-Cluster-level covariances appear to be beneficial in accurately recovering the true subgroups, but with higher $p$, the Bonferroni correction will further reduce power to detect.
-Increased variance of $\sigma_{b_1}$ may reduce the power to detect the second and third split in the tree, which are driven by subgroup differences in slopes.
+Higher values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ tend to yield higher numbers of splits, while the effect of $\rho$ is minimal.
 
 For SEM trees and LongCART (Figure~\ref{fig:tree_sizes_interact}), strongest effects were observed for $N$, followed by $\sigma_{b_0}$, $p_{noise}$, $\rho$ and $\sigma_{b_1}$.
-Results for $N$ were as expected for all methods: with increasing sample size, more splits are implemented.
+Results for $N$ were as expected for all methods: More splits are implemented with higher sample size.
 Both LRT- and score-based SEM trees seem only affected by levels of $\sigma_{b_0}$, $p_{noise}$ and $\rho$ under misspecification of the random effects.
@@
 This pattern seemed reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects were correctly specified.
-LongCART implemented more splits with higher levels of $p_{noise}$ and $\rho$, but was unaffected by levels of $\sigma_{b_0}$ and $\sigma_{b_1}$.
+LongCART implemented more splits with higher levels of $\rho$,  $\sigma_{b_1}$, $N$, $p_{noise}$ and $\sigma_{b_0}$.
 
@@
 cols <- rep(rainbow_hcl(10)[c(3,5)], times = 5)
-ggplot(tmp[LMM_ids, ]) +
-  geom_boxplot(aes(x=setting, y=tree_size, fill = sigma_int), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[0]))) + 
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_int), 
+               position=position_dodge(1), alpha = .5, width = .6, 
+               outlier.shape = NA, coef = NULL) +
+  geom_count(aes(group=sigma_int), position=position_dodge(1), colour = "black", 
+             fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) + 
   geom_hline(yintercept=3, col = "darkgrey") + 
   theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
-  labs(fill=expression(sigma[b[0]]^2))
+  labs(fill=expression(sigma[b[0]]^2)) +
+  stat_summary(aes(group=sigma_int), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
 @
@@
 <<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp[LMM_ids, ]) +
-  geom_boxplot(aes(x=setting, y=tree_size, fill = N), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(),
-                     breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits") +
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = N), 
+               position=position_dodge(1), alpha = .5, width = .6, 
+               outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=N), 
+             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits") +
   geom_hline(yintercept=3, col = "darkgrey") + 
   theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank())
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  stat_summary(aes(group=N), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
 @
@@
 <<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp[LMM_ids, ]) +
-  geom_boxplot(aes(x=setting, y=tree_size, fill = p_noise), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(),
-                     breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col = expression(p[noise])) +
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = p_noise), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=p_noise), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col = expression(p[noise])) +
   geom_hline(yintercept=3, col = "darkgrey") +
@@
         strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(p[noise]))
+  labs(fill=expression(p[noise])) +
+  stat_summary(aes(group=p_noise), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
 @
@@
 <<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp[LMM_ids, ]) +
-  geom_boxplot(aes(x=setting, y=tree_size, fill = sigma_slope), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(),
-                     breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[1]))) +
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_slope), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=sigma_slope), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
   geom_hline(yintercept=3, col = "darkgrey") + 
@@
         strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(sigma[b[1]]^2))
+  labs(fill=expression(sigma[b[1]]^2)) +
+  stat_summary(aes(group=sigma_slope), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+
 @
@@
 <<echo=FALSE, fig=TRUE, height=1.8>>=
-ggplot(tmp[LMM_ids, ]) +
-  geom_boxplot(aes(x=setting, y=tree_size, fill = rho), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[1]))) +
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = rho), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=rho), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
   geom_hline(yintercept=3, col = "darkgrey") + 
   theme(strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(rho))
+  labs(fill=expression(rho)) +
+  stat_summary(aes(group=rho), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
 theme_set(theme_gray(base_size = 11))
 @
-\end{subfigure}
-{\footnotesize \textit{Note.} Grey lines indicate true number of splits; distances on $y$-axis are on log scale.
+\end{subfigure}%
+\vspace{-.6cm}%
+{\singlespacing \small \textit{Note.} Black dots represent means, grey circles represent counts, grey lines represent true number of splits.
+Distances on $y$-axis are on log scale.
 $\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $N$ = sample size at level 2; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
@@
 <<echo=FALSE, fig=TRUE, height=1.7>>=
+lim <- c(0, 18)
 theme_set(theme_gray(base_size = 8))
-ggplot(tmp2) +
-  geom_boxplot(aes(x=random, y=tree_size, fill = N), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = N), position=position_dodge(1), alpha = .5, width = .6,
+               outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=N), 
+             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col=expression(N)) + 
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(fill=expression(N)) +
+  stat_summary(aes(group=N), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_int), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=sigma_int), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, limits = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(sigma[b[0]]^2)) +
+  stat_summary(aes(group=sigma_int), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = p_noise), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=p_noise), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col = expression(p[noise])) +
+  geom_hline(yintercept=3, col = "darkgrey") +
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(p[noise])) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  stat_summary(aes(group=p_noise), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = rho), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=rho), position=position_dodge(1), colour = "black", 
+             fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col=expression(rho)) +
+  geom_hline(yintercept=3, col = "darkgrey") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(rho)) + 
+  stat_summary(aes(group=rho), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=2.0>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_slope), 
+               position=position_dodge(1), alpha = .5, width = .6,
+               outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=sigma_slope), 
+             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
   scale_y_continuous(trans = pseudo_log_trans(), 
@@
   facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col=expression(N)) + 
-  geom_hline(yintercept=3, col = "darkgrey") + 
-  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
-  labs(fill=expression(N))
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp2) +
-  geom_boxplot(aes(x=random, y=tree_size, fill = sigma_int), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col=expression(sigma^2~(b[0]))) +
-  geom_hline(yintercept=3, col = "darkgrey") + 
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(sigma[b[0]]^2))
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp2) +
-  geom_boxplot(aes(x=random, y=tree_size, fill = p_noise), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col = expression(p[noise])) +
-  geom_hline(yintercept=3, col = "darkgrey") +
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(p[noise]))
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp2) +
-  geom_boxplot(aes(x=random, y=tree_size, fill = rho), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col=expression(rho)) +
-  geom_hline(yintercept=3, col = "darkgrey") + 
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(rho))
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.8>>=
-ggplot(tmp2) +
-  geom_boxplot(aes(x=random, y=tree_size, fill = sigma_slope), 
-               position=position_dodge(1), alpha = .5, width = .6) + 
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "# of splits", col=expression(sigma[b[1]]^2)) +
+  labs(x = "", y = "no. of splits", col=expression(sigma[b[1]]^2)) +
   geom_hline(yintercept=3, col = "darkgrey") +
   theme(strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(sigma[b[1]]^2))
+  labs(fill=expression(sigma[b[1]]^2)) + 
+  scale_x_discrete(labels = scales::parse_format()) +
+  stat_summary(aes(group=sigma_slope), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
 theme_set(theme_gray(base_size = 11))
 @
-\end{subfigure}
-{\footnotesize \textit{Note.} Grey lines indicate true number of splits; distances on $y$-axis are on log scale.
+\end{subfigure}%
+\vspace{-.6cm}%
+{ \singlespacing \small \textit{Note.} Black dots represent means, grey circles represent counts, grey lines represent true number of splits.
+Distances on $y$-axis are on log scale.
 $N$ = sample size at level 2; $\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
@@
 https://bbolker.github.io/mixedmodels-misc/notes/corr\_braindump.html
-
-<<eval=FALSE, echo=FALSE>>=
-## ECLSK are reading scores with minbucket = 500
-# load("predsECLSK")
-# load("tree_sizesECLSK")
-# MSEs <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
-# ses <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
-#          sqrt(sum(!is.na(x))))
-# Table <- data.frame(MSE = MSEs[-13], SE = ses[-13],
-#                     mean_tree_size = sapply(sizes, mean, na.rm = TRUE),
-#                     SD = sapply(sizes, sd, na.rm = TRUE))
-# names(Table) <- c("$MSE$", "$se_{MSE}$", "$M_{tree size}$", "$SD_{tree size}$")
-# rownames(Table) <- c("LMtree", "LMtree_c", "LMMtree", "LMMtree_c", "LMMtree_r",
-#                      "LMMtree_{cr}", "LMMtree_s", "LMMtree_{sc}", "LMMtree_{sr}",
-#                      "LMMtree_{scr}", "longRPart", "longRPart_s")
-# tmp_minbucket <- data.frame(ID = names(MSEs[-13]),
-#                             tree_size = colMeans(sizes, na.rm = TRUE),
-#                             MSE = MSEs[-13])
-@
-
-<<echo=FALSE>>=
-#' Print LaTeX tables with headings over two lines
-#' 
-#' @param .data A data frame with at least some of the column names having a \code{separator}.
-#' @param separator A regular expression that splits the top and bottom rows.
-If the separator is not found in the names of \code{.data}, the function returns an error (saying you should probably just use \code{print.xtable()})
-#' @param align Passed to \code{xtable}: Character vector of length equal to the number of columns of the resulting table, indicating the alignment of the corresponding columns.
-#' @param digits Passed to \code{xtable}
-#' @param label Passed to \code{xtable}
-#' @param caption Passed to \code{xtable}
-#' @param booktabs Should the tabular environment produced use booktabs? Set to TRUE for (my) convenience. This will cause an error if  \verb{\usepackage{booktabs}} has not been called in LaTeX.
-#' @param heading_command A (simple) LaTeX control sequence (properly escaped) to apply to each heading names.
-#' @param ... Arguments passed to \code{print.xtable}. You cannot pass \code{add.to.row}, \code{include.rownames}, or \code{include.colnames} as we make use of these options in this function.  
-#' @return Output intended for LaTeX. A table produced using xtable where groups of column names are put in the top row. 
-#' @author Hugh Parsonage
-#' @examples
-#' example_df <- 
-#' data.frame(yr = 2001:2005, 
-#'           Revenue__foo = 1:5, 
-#'           Revenue__bar = 11:15, 
-#'           Revenue__baz = 21:25, 
-#'           ordinary = 1:5,
-#'           Expense__foo = 1:5,
-#'           Expense__bar = 11:15, 
-#'           Expense__baz = 21:25, 
-#'           Last__foo = 1:5, 
-#'           Last__baz = 2:6,
-#'           last = 101:105)
-#' print_2heading_xtable(example_df, separator = "__")
-#' @export
-print_2heading_xtable <- function(.data, 
-                                  separator = "__", 
-                                  align = NULL, 
-                                  caption = NULL,
-                                  digits = NULL,
-                                  label = NULL,
-                                  booktabs = TRUE, 
-                                  heading_command = "\\textbf", ...){
-  orig_names <- names(.data)
-  if (!any(grepl(separator, orig_names))){
-    stop("No separator found in column names, so there is no point in using this function. Make sure you have specified the right separator; otherwise, just use print.xtable().")
-  }
-
-  if (any(c("add.to.row", "include.colnames", "include.rownames") %in% names(list(...)))){
-    stop("You should not pass add.to.row, include.colnames, or include.rownames to print.xtable() via this function.")
-  }
-
-  split_names <-  grep(separator, orig_names, value = TRUE)
-  split_positions <- grep(separator, orig_names, value = FALSE)
-
-  # get the names before the separator
-  top_headers <- gsub(paste0("^(.*)", separator, ".*$"), "\\1", split_names)
-  # Where in the original table is there a new top header?
-
-  orig_names_no_suffix <- 
-    gsub(paste0("^(.*)", separator, ".*$"), paste0("\\1", separator), orig_names)
-
-  # For cmidrule{}
-  position_of_header_instance <- 
-    # Need to test first column
-    which(orig_names_no_suffix == dplyr::lead(orig_names_no_suffix) & 
-            (orig_names_no_suffix != dplyr::lag(orig_names_no_suffix) | is.na(dplyr::lag(orig_names_no_suffix))))
-
-  position_of_header_final <- 
-    # Need to test final column
-    which((orig_names_no_suffix != dplyr::lead(orig_names_no_suffix) | is.na(dplyr::lead(orig_names_no_suffix))) &
-            orig_names_no_suffix == dplyr::lag(orig_names_no_suffix))
-
-  if (length(position_of_header_instance) != length(position_of_header_final)){
-    stop("This is a bug. Sorry. Please provide your data frame to the grattan package maintainer.")
-  }
-
-  double_row_column_names <- 
-    rbind(gsub("^(.*)__(.*)$", "\\1", orig_names), gsub("^(.*)__(.*)$", "\\2", orig_names))
-
-  # factor etc in table to preserve order
-  top_headers_widths <- 
-    as.data.frame(table(factor(double_row_column_names[1,], levels = unique(double_row_column_names[1,]))))
-
-  first_row <- 
-    unique(double_row_column_names[1,])
-
-  first_row_formatted <- 
-    paste0(heading_command, "{", first_row, "}")
-
-  top_row <- character(length(first_row))
-
-  # Could do paste0() directly but decided that it would 
-  # avoid the point which is to add \multicolumn only to the rows that call for it.
-  for (ii in seq_along(first_row)){
-    if (first_row[ii] %in% top_headers){
-      top_row[ii] <- paste0("\\multicolumn{", top_headers_widths$Freq[ii], "}{c}{", first_row_formatted[ii], "}")
-    }
-  }
-  rm(ii)
-
-  for_latex_top_row <- 
-    paste0(paste0(top_row, collapse = " & "), "\\\\")
-
-  if (booktabs){
-    # (lr) to avoid cmidrule touching adjacent groups
-    between_row <- paste0("\\cmidrule(lr){",  position_of_header_instance, "-", position_of_header_final, "}")
-  } else {
-    between_row <- paste0("\\cline{",  position_of_header_instance, "-", position_of_header_final, "}")
-  }
-  for_latex_between_row <- 
-    paste0(paste0(between_row, collapse = ""))
-
-  for_latex_second_row <- 
-    paste0(heading_command, "{", double_row_column_names[2,], "}")
-
-  for_latex_second_row <- 
-    paste0(paste0(for_latex_second_row, collapse = " & "), "\\\\")
-
-  addtorow <- list()
-  addtorow$pos <- list(0, 0, 0)
-  addtorow$command <- 
-    paste0(paste0(c(for_latex_top_row, for_latex_between_row, for_latex_second_row)), "\n")
-
-  xtable::print.xtable(xtable::xtable(.data, align = align, digits = digits, 
-                                      label = label, caption = caption), 
-                       type = "latex",
-                       add.to.row = addtorow, 
-                       include.colnames = FALSE, 
-                       include.rownames = FALSE,
-                       booktabs = booktabs,
-                       ...)
-}
-@
-
-<<echo=FALSE, eval=FALSE, results=tex>>=
-## ECLSK2 are reading scores with maxdepth = 4
-load("predsECLSK2")
-var_reading <- round(var(preds$observed), digits = 3)
-MSEs_read <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
-ses_read <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
-         sqrt(sum(!is.na(x))))
-load("preds_math")
-var_math <- round(var(preds$observed), digits = 3)
-MSEs_math <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
-ses_math <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
-         sqrt(sum(!is.na(x))))
-load("preds_science")
-var_science <- round(var(preds$observed), digits = 3)
-MSEs_scie <- sapply(preds, function(x) mean((x - preds$observed)^2, na.rm = TRUE))
-ses_scie <- sapply(preds, function(x) sd((x - preds$observed)^2, na.rm = TRUE) / 
-         sqrt(sum(!is.na(x))))
-Table2 <- data.frame(MSEs_read[-13], ses_read[-13],
-                     MSEs_math[-13], ses_math[-13],
-                     MSEs_scie[-13], ses_scie[-13])
-Table2$mean_rank <- 1/3*(rank(Table2$MSEs_read..13.) + rank(Table2$MSEs_math..13.) + 
-  rank(Table2$MSEs_scie..13.))
-names(Table2) <- c("Reading MSE", "SE", "Math MSE", "SE", "Science MSE", "SE", 
-                   "mean rank")
-Names <- cbind(sapply(strsplit(rownames(Table2), "_"), function(x) x[1]),
-                sapply(strsplit(sapply(strsplit(rownames(Table2), "_"), 
-                                       function(x) x[2]), ""), 
-                       function(x) paste(x, collapse = ", ")))
-Names[,2][Names[,2] == "NA"] <- ""
-Names[3:12, 2] <- paste("i,", Names[3:12, 2])
-
-Table3 <- cbind(Algorithm = paste(Names[,1], Names[,2]),
-                Table2, stringsAsFactors = FALSE)
-
-names(Table3) <- c("Algorithm",  "Reading__MSE", "Reading__SE", "Math__MSE", "Math__SE", 
-                   "Science__MSE", "Science__SE", "  __mean\nrank")
-
-print_2heading_xtable(Table3, separator = "__", 
-                      digits = c(0, 0, 4, 3, 4, 3, 4, 3, 1), 
-                      align = c("r", "l", rep("c", times = 7)), 
-                      caption = "Predictive accuracy for the ECLSK data.
-c indicates that parameter stabiliy tests were performed at the cluster level; r indicates that estimation was initialized with the random effects; i indicates that random intercepts were estimated; s indicates that random slopes were estimated.",
-                      label = "tab:ECLSK_accuracy",
-                      hline.after = c(-1, 0, nrow(Table3)))
-@
-
-<<echo=FALSE, eval=FALSE, results=tex>>=
-load("tree_sizesECLSK2")
-failed_longRpart_count <- sum(is.na(sizes$longRPart))
-failed_longRpart_s_count <- sum(is.na(sizes$longRPart_s))
-tree_sizes <- data.frame(
-  tree_size = colMeans(sizes, na.rm = TRUE),
-  SD = sapply(sizes, sd, na.rm = TRUE))
-load("sizes_math")
-failed_longRpart_count <- sum(failed_longRpart_count, sum(is.na(sizes$longRPart)))
-failed_longRpart_s_count <- sum(failed_longRpart_s_count, sum(is.na(sizes$longRPart_s)))
-tree_sizes <- cbind(tree_sizes,
-                    tree_size = colMeans(sizes, na.rm=TRUE),
-                    SD = sapply(sizes, sd, na.rm=TRUE))
-load("sizes_science")
-failed_longRpart_count <- sum(failed_longRpart_count, sum(is.na(sizes$longRPart)))
-failed_longRpart_s_count <- sum(failed_longRpart_s_count, sum(is.na(sizes$longRPart_s)))
-tree_sizes <- cbind(tree_sizes,
-                    tree_size = colMeans(sizes, na.rm=TRUE),
-                    SD = sapply(sizes, sd, na.rm=TRUE))
-tree_sizes$mean_rank <- 1/3*(rank(tree_sizes[,1]) + rank(tree_sizes[,3]) + rank(tree_sizes[,5]))
-
-tree_sizes <- cbind(Algorithm = paste(Names[,1], Names[,2]), tree_sizes, 
-                    stringsAsFactors = FALSE)
-names(tree_sizes) <- c("Algorithm",  "Reading__M", "Reading__SD", "Math__M", "Math__SD", 
-                   "Science__M", "Science__SD", "  __mean\nrank")
-
-print_2heading_xtable(tree_sizes, separator = "__", 
-                      digits = c(0, 0, 1, 1, 1, 1, 1, 1, 1), 
-                      align = c("r", "l", rep("c", times = 7)), 
-                      caption = "Tree sizes for the ECLSK data.
-c indicates that parameter stabiliy tests were performed at the cluster level; r indicates that estimation was initialized with the random effects; i indicates that random intercepts were estimated; s indicates that random slopes were estimated.",
-                      label = "tab:ECLSK_complexity",
-                      hline.after = c(-1, 0, nrow(tree_sizes)))
-@
-
-%Table~\ref{tab:ECLSK_accuracy} shows the cross-validated predictive accuracies for all RPMs.
-Prediction errors were largest for the science data, which also had the largest sample variance (...), while prediction errors were lowest for the reading data, which also had the smallest sample variance (...).
-The sample variance of the math score was ....
-
-%Several fitting attempts for longRPart trees resulted in estimation errors: For the trees with random intercepts estimated only, ...
-out of 30 tree-fitting attempts yielded an error, while for the trees with both random intercepts and slopes estimated, ...
-out of 30 tree-fitting attempts yielded an error.
-Results for longRPart are computed based on successfully fitted trees.
-
-%The MSEs differ significantly only between the LM(M) trees and longRPart trees.
-LM(M) trees provide somewhat lower prediction errors, but the tree sizes in Table~\ref{tab:ECLSK_complexity} indicate that this comes at the cost of larger tree size.
-
-%On average, LMM trees were more accurate but also larger than longRPart trees.
-In terms of average predictive accuracy, the tree which estimates random intercepts and initialized estimation with the random effects performed best for the LMM trees.
-For the longRPart trees, the model with both random intercepts and slopes performed best.
-
-%However, the settings that perform best for predicting reading and math scores, did not perform best for predicting science scores.
-Cluster sizes (i.e., number of observations per student) are lower for the science scores (3) than for the reading and math scores (5), which may partly explain the difference.
-For LM(M) trees, the trees which estimate both random intercepts and slopes while performing parameter stability tests on the cluster level, perform best in the reading and math data.
-However, for the science data, these settings perform among the worst, while highest accuracy was obtained by the LMM tree in which only a random intercept was estimated, and estimation was initialized with the random effects.
-Note that both these settings were found to perform best in the earlier simulation study.
-
-%For longPart, highest predictive accuracy was also obtained with trees in which both random intercepts and slopes were estimated for the reading and math data.
-For the science data, trees in which only random intercepts were estimated provided highest predictive accuracy.
-In all three datasets, the random-effects specification that yielded the largest longRPart trees, also yielded the highest predictive accuracy.
-
-\begin{figure}
-<<echo=FALSE, eval=FALSE, fig=TRUE>>=
-load("comp_timesECLSK2") ## Reading, maxdepth = 4
-times2 <- list(reading = times)
-load("times_math") 
-times2$math <- times
-load("times_science") 
-times2$science <- times
-times <- do.call(rbind, times2)
-comp_time <- reshape2::melt(times)
-p <- ggplot(comp_time, aes(variable, value))  
-p + geom_boxplot() + labs(x = "", y = "Computation time (in seconds)") +
-  scale_y_continuous(trans = "log", breaks = c(.1, 1, 10, 100, 1000, 10000)) +
-  theme(axis.text.x = element_text(angle = 90))
-@ 
-\caption{Computation time distributions for the different partitioning methods.}
-\label{fig:comp_timeECLSK}
-\end{figure}
-
-%In terms of computation time, Figure~\ref{fig:comp_timeECLSK} indicates a clear computational advantage for LM(M) trees.
-The outlying, lower computation times for longRPart are due to estimation errors early in the fitting process, after which computations are halted completely.

### 2023-01-20 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
improvements in abstract and intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3243 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 22883f9d5661d07da9d7f6258f4d71a06c2b8503

@@
 \authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
-\authorsaffiliations{{$^1$Leiden University}, {$^2$University of Innbruck}}
+\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}
 
 \abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time.
-Heterogeneity between subjects is likely encountered in such models and researchers are often interested in explaining or predicting this heterogeneity.
-We show how generalized linear mixed effects model (GLMM) trees can be used to identify heterogeneous subgroups in linear growth curve models.
-GLMM trees were originally proposed for clustered cross-sectional data; here we extend the method to allow for longitudinal data.
+Heterogeneity between subjects is often encountered in such models and researchers are typically interested in explaining or predicting this heterogeneity.
+We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups associated with different sets of parameters in linear growth curve models.
+Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data.
 The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.
 We evaluate GLMM tree's performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities.
-We compare the performance of GLMM trees with three alternative methods: SEM trees, LongCART and longRPart.
+We compare the performance of GLMM trees with three alternative methods: structural equation model (SEM) trees, LongCART, and longRPart.
 GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart.
@@
 
-Development over time is of prime interest in psychological researchers.
+Development over time is of prime interest in psychological research.
 For example, researchers may want to model and predict student's reading or math development in educational studies \citep{MorgyFark09}, or patient's symptom reduction over the course of clinical trails \citep{Fein09}.
 Mixed-effects or latent-variable models can be used to model development over time, and allow for explaining possible heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}.
-However, the covariates that are predictive of development may often not be known in advance, and identifying them may be of prime interest.
-
-Recursive partitioning methods (RPMs) allow for identifying relevant predictors from a potentially (very) large number of covariates.
+However, the covariates that are predictive of development may often not be known in advance, and methods for identifying them are needed.
+
+Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates.
 Several RPMs that allow for partitioning growth-curve models have been developed: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
@@
 
-Here we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for partitioning LGCMs.
+Here we extend generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} so that they can be applied to partitioning LGCMs.
 GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas.
-Yet, GLMM trees provide a more general method 2for subgroup detection in almost any type of mixed-effects model.
+Moreover, the general idea of GLMM trees is also appealing for subgroup detection in almost any type of mixed-effects model.
 Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates).
-In this paper, we propose and test two recent extensions of GLMM trees that account for these characteristics.
-We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wide range of settings where covariates are measured at higher levels and/or random-effects variance is substantial.
+In this paper, we propose and test two extensions of GLMM trees that account for these characteristics.
+We focus on the specific use case of partitioning LGCMs, but the extensions are also critical for a wide range of further settings where covariates are measured at higher levels and/or where random-effects variance is substantial.
 
@@
 Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations.
-This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes.
-With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \cite{FokkySmit18}.
-
-All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based the observations in that node only.
-This fully local estimation approach brings higher computational load and increased model complexity.
+This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes.
+With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \citep{FokkySmit18}.
+
+All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based on the observations in that node only.
+This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity.
 GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model.
@@
 
-The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees.
-These methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split.
-To choose the optimal split, a likelihood ratio test is computed for every possible cut point.
-Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}.
+The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and SEM trees based on likelihood ratio tests (LRTs).
+All of these methods employ an exhaustive split detection procedure: For every possible split in every available partitioning variable in the current node, the full parametric model is re-estimated in the two nodes that would result from the split.
+To choose the optimal split, some split criterion (such as a $p$-value from the corresponding LRT) is computed for every possible cutpoint.
+Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}.
 LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split.
@@
 
-In what follows, we first describe estimation of GLMM trees, and possible adjustments to better account for longitudinal data structures.
-Next, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees.
+In what follows, we first describe estimation of GLMM trees in general along with the novel adjustments geared towards longitudinal data structures.
+Subsequently, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees.
 In Simulation~Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
@@
 %
-Thus, in the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$.
+Throughout this paper we focus on the case with a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$.
+Therefore, the identity function can be used for the link $g$.
 Generalizations to other response variable types within the GLM are straightforward.
 Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$).
-We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$.
+We further assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$.
 Note that the number and spacing of observed timepoints may differ between subjects.

### 2023-01-20 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
improvements in estimation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3244 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7b8f56f862b0187b7e2e45826d7934fc2a724b76

@@
 
-The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the terminal nodes.
+The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the subgroups defined by the terminal nodes of the tree.
 Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations.
@@
 
-All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based on the observations in that node only.
+All other methods for partitioning LGCMs take a fully local estimation approach: Within every subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that node only.
 This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity.
@@
 Note that the number and spacing of observed timepoints may differ between subjects.
-$\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part.
+$\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on the subgroup/node~$j$ which observation~$i$ falls into.
 As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$.
@@
 
-The parameters of a traditional GLMM can be estimated by, for example, (restricted) maximum likelihood.
-The GLMM tree model is estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:
+The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML).
+
+Thus, when it is known which node~$j$ each observation~$i$ falls into, the GLMM specified by Equations~\ref{eq:expected_value}--\ref{eq:GLMMtree} can be fitted ``as usual'', yielding local subgroup-specific fixed-effect estimates $\hat \beta_j$ and global random-effect estimates $\hat b_i$.
+To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}.
+The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm where the initialization simply assumes zero random effects.
 
@@
 	\item Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
-	\item Set $r = r+1$. Estimate the partition using $Z_{i} \hat{b_i}_{(r-1)}$ as an offset.
-	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with terminal node $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
+	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b_i}_{(r-1)}$ as an offset.
+	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with the partition $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
 	\item Repeat Steps~1 and~2 until convergence.
@@
 
-The algorithm initializes by assuming zero random effects.
+%% The algorithm initializes by assuming zero random effects.
 In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2.
@@
 The fixed effects are estimated \textit{locally}, using the observations in the current node.
+
 Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3.
-Typically, this converges if the tree does not change from one iteration to the next.
+Typically, this converges when the partition~$j(r)$ from the GLM tree is the same as $j(r-1)$.
 

### 2023-01-20 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
make notation of b_i more consistent
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3245 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 304253a3c76033f9e2dc86fcbf6716e3ce23b7aa

@@
 Note that the number and spacing of observed timepoints may differ between subjects.
-$\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on the subgroup/node~$j$ which observation~$i$ falls into.
+$\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on the subgroup/node~$j$ which subject~$i$ falls into.
 As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$.
-Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.
+Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.
 
 The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML).
-
-Thus, when it is known which node~$j$ each observation~$i$ falls into, the GLMM specified by Equations~\ref{eq:expected_value}--\ref{eq:GLMMtree} can be fitted ``as usual'', yielding local subgroup-specific fixed-effect estimates $\hat \beta_j$ and global random-effect estimates $\hat b_i$.
+Thus, when it is known which node~$j$ each subject~$i$ falls into, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates $\hat \beta_j$ and \emph{global} random-effect estimates $\hat b_i$.
 To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}.
@@
 	\setcounter{enumi}{-1}
-	\item Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
-	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b_i}_{(r-1)}$ as an offset.
-	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with the partition $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
+	\item Initialize by setting $r$ and all values $\hat{b}_{i,(r)}$ to 0.
+	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset.
+	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with the partition $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{i,(r)}$ from the estimated model.
 	\item Repeat Steps~1 and~2 until convergence.
@@
 Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3.
-Typically, this converges when the partition~$j(r)$ from the GLM tree is the same as $j(r-1)$.
+Typically, this converges when the partition~$j(r)$ from the GLM tree is the same as $j(r-1)$ from the previous step.
 

### 2023-01-20 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
more improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3246 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 34da27aae162232c32964457ff24de8111deac64

@@
 
+In the following two subsections provide some more details on alternative initialization strategies for Step~0 and details of the GLM tree algorithm from Step~1, respecitvely.
+Each subsection first reviews the well-established methods and the proceeds to discuss modifications that may improve the performance when partitioning LGCMs using longitudinal data.
+
 \subsection{Initialization}
 
-Studies to date indicate that initializing estimation with zero random effects yields accurate final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
-\cite{SelaySimo12} assessed the impact of different initialization values and found minor impact that became smaller with increasing sample size.
+Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
+\cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that became decreased further with increasing sample size.
 In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data.
-With longitudinal data, however, random effects tend to be stronger: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data.
-If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that is difficult to overcome in subsequent iterations.
+With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more closely than observations nested within the same unit in cross-sectional data.
+If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations.
 Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy.
+More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
 
@@
 
-The subgroup structure in Step~1 is estimated using model-based recursive partitioning \citep[MOB; ][]{ZeilyHoth08}. The MOB algorithm cycles iteratively through the following steps:
+The subgroup structure in Step~1 is estimated by a GLM tree using the general model-based recursive partitioning (MOB) algorithm by \cite{ZeilyHoth08}.
+Here, we give a short general overview and then comment on the aspects of the algorithm that may be particularly relevant for LGCMs.
+In the case of GLMs (with random effects held constant in an offset), the MOB algorithm cycles iteratively through the following steps:
 
@@
 \renewcommand{\labelenumi}{\alph{enumi})}
-	\item Fit the parametric model to all observations in the dataset. 
-	\item Test for parameter instability with respect to each of the partitioning variables.
-	\item If there is some overall parameter instability, split the dataset with respect to the partitioning variable associated with the highest instability.
+	\item Fit the GLM to all observations in the current subgroup. 
+	\item Test for instability of the GLM parameters with respect to each of the partitioning variables.
+	\item If there is some overall parameter instability, split the subgroup with respect to the partitioning variable associated with the highest instability.
 	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
@@
 
-Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the model fitted in Step~(a).
-Under mild regularity conditions, the scores have an expected value of 0.
-The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or exhibit systematic deviations when ordered by the values of a covariate.
-For continuous covariates (or ordered covariates with a large enough number of unique values), this involves computing the cumulative score process $W_k(t)$ with respect to each potential partitioning variable $U_k$ \citep{ZeilyHoth08}:
-
+Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the GLM fitted in Step~(a).
+Under correct specification of the model and mild regularity conditions, the scores have an expected value of 0.
+The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or exhibit systematic deviations when ordered by the values of a covariate available for partitioning.
+For continuous covariates $U_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each potential partitioning variable \citep{ZeilyHoth08}:
+%
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in Step~(a), and $n_j$ gives the number of observations in the current node.
-Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes the estimated scores, with subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$.
-Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
-
-From the cumulative score process $W_k(t)$, a range of test statistics can be derived which capture non-random fluctuations.
-For numerical partitioning variables, a maximum Lagrange multiplier test statistic can be computed, which takes the maximum of the squared Euclidean norm of $W_k(t)$, weighted by it's variance \citep{ZeilyHorn07}.
+%
+where $\hat{J}$ is a suitable estimate of the covariance matrix of the parameter estimates, and $n_j$ gives the number of observations in the current subgroup.
+Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$.
+Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation associated with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
+
+From the cumulative score process $W_k(t)$, a range of test statistics can be derived which capture increased fluctuations (beyond the random fluctuation under parameter stability).
+For numerical partitioning variables, a maximum Lagrange multiplier test statistic can be computed, which takes the maximum of the squared Euclidean norm of $W_k(t)$, weighted by its variance \citep{ZeilyHorn07}.
 This statistic is referred to as the \textit{supLM} statistic, and is asymptotically equivalent to the maximum of likelihood-ratio statistics.

### 2023-01-29 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
revised estimation -> check again
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3249 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 21c48d74adfae687c7aea260924630296aa24514

@@
 %
-Throughout this paper we focus on the case with a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$.
-Therefore, the identity function can be used for the link $g$.
-Generalizations to other response variable types within the GLM are straightforward.
-Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$).
-We further assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$.
-Note that the number and spacing of observed timepoints may differ between subjects.
-$\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on the subgroup/node~$j$ which subject~$i$ falls into.
-As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$.
+Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon$.
+Therefore, the identity function can be used for the link $g$ but generalizations to other response variable types within the GLM are straightforward.
+In the general notation above, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject~$i$ ($i = {1, \dots, N}$), comprising $p$ regressors plus one column of 1s for the intercept.
+In the following, we assume that time is the predictor variable of interest (i.e., $p = 1$), where the number $n_i$ and spacing of observed timepoints may differ between subjects.
+The fixed-effects parameters $\beta_j$ (here, intercept and time slope) in GLMM trees are node-specific, i.e., their value depends on the subgroup/node~$j$ which subject~$i$ falls into.
+As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is the corresponding vector of random effects for subject $i$.
 Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.
@@
 The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML).
-Thus, when it is known which node~$j$ each subject~$i$ falls into, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates $\hat \beta_j$ and \emph{global} random-effect estimates $\hat b_i$.
+Thus, when it is known which node~$j$ each subject~$i$ falls into, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates~$\hat \beta_j$ and \emph{global} random-effect estimates~$\hat b_i$.
 To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}.
-The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm where the initialization simply assumes zero random effects.
-
+The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm.
+The initialization simply assumes zero random effects.
+%
 \begin{enumerate}
@@
 	\setcounter{enumi}{-1}
-	\item Initialize by setting $r$ and all values $\hat{b}_{i,(r)}$ to 0.
-	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset.
-	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with the partition $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{i,(r)}$ from the estimated model.
+	\item Initialize by setting step $r = 0$ and all random-effect estimates $\hat{b}_{i,(r)} = 0$.
+	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset, yielding the partition $j_{(r)}$.
+	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j, (r)} + Z_{i} b_{i, (r)}$ with the partition $j_{(r)}$ from Step~1.
+Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
 	\item Repeat Steps~1 and~2 until convergence.
 \end{enumerate}
-
-%% The algorithm initializes by assuming zero random effects.
-In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2.
-Thus, the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset.
-The fixed effects are estimated \textit{locally}, using the observations in the current node.
-
+%
 Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3.
-Typically, this converges when the partition~$j(r)$ from the GLM tree is the same as $j(r-1)$ from the previous step.
+Typically, this converges when the partition~$j_{(r)}$ from the GLM tree is the same as $j_{(r-1)}$ from the previous step.
 
 In the following two subsections provide some more details on alternative initialization strategies for Step~0 and details of the GLM tree algorithm from Step~1, respecitvely.
-Each subsection first reviews the well-established methods and the proceeds to discuss modifications that may improve the performance when partitioning LGCMs using longitudinal data.
+Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve the performance when partitioning LGCMs using longitudinal data.
 

### 2023-01-29 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
further improvements for initialization and partitioning subsections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3250 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ccc6feeffe593e508b23ae959a2680667248aac0

@@
 Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
-\cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that became decreased further with increasing sample size.
+\cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size.
 In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data.
@@
 More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
+The alternative initialization step is thus:
+%
+\begin{enumerate}
+\setlength\itemsep{0em}
+\renewcommand{\labelenumi}{\arabic{enumi}$'$.}
+	\setcounter{enumi}{-1}
+	\item Initialize by setting step $r = 0$ and fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta + Z_{i} b_{i, (r)}$ to the full sample.
+Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
+\end{enumerate}
 
@@
 
-The subgroup structure in Step~1 is estimated by a GLM tree using the general model-based recursive partitioning (MOB) algorithm by \cite{ZeilyHoth08}.
+The subgroup structure in Step~1 is estimated by a GLM tree using the general model-based recursive partitioning (MOB) algorithm of \cite{ZeilyHoth08}.
 Here, we give a short general overview and then comment on the aspects of the algorithm that may be particularly relevant for LGCMs.
 In the case of GLMs (with random effects held constant in an offset), the MOB algorithm cycles iteratively through the following steps:
-
+%
 \begin{enumerate}
 \setlength\itemsep{0em}
-\renewcommand{\labelenumi}{\alph{enumi})}
+\renewcommand{\labelenumi}{(\alph{enumi})}
 	\item Fit the GLM to all observations in the current subgroup. 
@@
 \end{enumerate}
-
+%
 Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the GLM fitted in Step~(a).
 Under correct specification of the model and mild regularity conditions, the scores have an expected value of 0.
-The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or exhibit systematic deviations when ordered by the values of a covariate available for partitioning.
+The parameter stability tests evaluate whether the scores fluctuate randomly around this mean of 0, or exhibit systematic deviations when ordered by the values of a covariate available for partitioning.
 For continuous covariates $U_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each potential partitioning variable \citep{ZeilyHoth08}:

### 2023-01-29 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
improved description of variations of methods
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3251 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9a84b505a41887e3d3afc1def06cd8c6f08fbe17

@@
 
-\begin{figure}[bt]
+\begin{figure}[t!]
 \caption{Design of the subgroup-specific fixed effects in the simulation.}
-<<fig=T, height=5, width=5, echo=F>>=
+<<fig=TRUE, height=5.5, width=5.5, echo=FALSE>>=
 plot(tree, tnex = 3.5, terminal_panel = combine_panels(
@@
 Every dataset comprised four non-overlapping subgroups, corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree}.
-The response was computed as:\\
-
+The response was computed as:
+%
 $$y_{i} =  X_{i} \beta_{j} + Z_{i} b_i + \epsilon_{i},$$
-
-\noindent where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part.
+%
+where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part.
 The bottom panels of Figure~\ref{fig:design_tree} present the subgroup-specific values of the fixed intercept and slope ($\beta_{j0}$ and $\beta_{j1}$, respectively).
 The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints.
-The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$\footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees in the next section.}.
-Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next.
+The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$.\footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees in the next section.} Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next.
 Values of $\epsilon_{i}$ were independently generated from a normal distribution with a mean of zero and a variance of 5.
@@
 We varied the following data-generating characteristics:
-
+%
 \begin{itemize}
@@
 \item Number of noise partitioning variables: small ($p = 5$) or large ($p = 25$).
-\item Intercorrelation between partitioning variables: absent ($\rho = 0$) or present ($\rho = .3$).
+\item Intercorrelation between partitioning variables: absent ($\rho = 0$) or present ($\rho = 0.3$).
 \end{itemize}
-
+%
 We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We performed all data generation and analysis in R \citep[version 4.1.2; ][]{R22}.
@@
 
-We applied ten partitioning approaches to every generated dataset, each employing one of three different random-effects specifications:
-
+We applied ten variations LM(M) trees to every generated dataset, where each variation combines one of the following three \emph{random-effects specifications} (none/intercepts/intercepts+slopes) with the \emph{random-effect initializations} (if any; all zero vs.\ full sample estimates) and \emph{covariance specifications in the parameter instability tests} (classic vs.\ clustered), as discussed in the previous section.
+%
 \begin{itemize}
-
-    \item $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$. That is, random effects and thus their variance were fixed to 0, and two fixed-effects linear model (LM) trees were fitted.
+  \item None: $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$.
+Random effects and thus their variance were fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications.
   \begin{itemize}
-        \item The default observation-level covariances were used to compute the parameter stability tests.
-        \item Clustered covariances were used to compute the parameter stability tests. 
+      \item Default: Classic observation-level covariances.
+      \item Alternative: Clustered covariances. 
   \end{itemize}
 
-    \item $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$.
-That is, four linear mixed-effects model (LMM) tree were fitted in which the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0:
+  \item Intercepts: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$.
+This yields linear mixed-effects model (LMM) trees in which only the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0.
+The four variations considered are the following.
   \begin{itemize}
-      \item Default settings were used, in which parameter stability tests employed observation-level covariances and estimation is initialized with the tree structure. 
-    \item Instead, clustered covariances were used to compute the parameter stability tests.
-    \item Instead, estimation was initialized with the random effects. 
-    \item Instead, clustered covariances were used to compute the parameter stability tests and estimation was initialized with the random effects.
+    \item Default: Classic observation-level covariances in the parameter stability tests and random-effect initialization with all zeros (0.). 
+    \item Alternative: Clustered covariances in the parameter stability tests.
+    \item Alternative: Random-effect initialization with the full-sample estimates (0$'$.). 
+    \item Alternative: Clustered covariances and random-effect initialization with the full-sample estimates.
   \end{itemize}
 
-    \item $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$.
-That is, four LMM trees were fitted in which the variance of both the random intercept and slope were freely estimated.
-The same four approaches were applied (default, clustered covariances, random-effects initialization, clustered covariances and random-effects initialization) as with the previous LMM tree specification.
+  \item Intercepts and slopes: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$.
+This yields LMM trees in which the variance of both the random intercept and slope were freely estimated.
+The four variations considered are the same as for the intercept-only LMM trees.
 \end{itemize}
-
+%
 To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}.
@@
 We evaluated tree accuracy by counting the number of splits in each tree.
-Trees with $> 3$ splits are indicative of Type-I error, while trees with $> 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables).
+Trees with $> 3$ splits are indicative of Type-I error, while trees with $< 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables).
 We also assessed the variable selected for the first split in every tree.
@@
 
-\begin{figure}[!ht]
+\begin{figure}[t!]
 \caption{Tree size distributions for LM trees (left panel) and LMM trees (middle and right panel).}
@@
 ## Plot
+theme_set(theme_bw())
 plot0 <- ggplot(tmp[LMM_ids, ]) +
@@
                alpha = 0.5, width = .6, outlier.shape = NA, coef = NULL) + 
-  geom_count(aes(x=setting, y=tree_size), col = "grey", alpha = .75) +
+  geom_count(aes(x=setting, y=tree_size), col = "gray", alpha = .75) +
   scale_y_continuous(trans = pseudo_log_trans(sigma = 1, base = exp(1)), 
@@
   geom_text(data = summ_stats, aes(x=setting, y=y, label = all),
-            size = 3, col = "grey44") +
+            size = 3, col = "gray44") +
   labs(x = "", y = "no. of splits") + theme(legend.position = "none") +
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   facet_grid(~ranef, scales = "free", space = "free", 
@@
 
-{\textit{Note.} Grey circles represent counts, grey lines represent true number of splits.
+{\textit{Note.} Gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
 Panel labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.
 Distances on $y$-axis are on the log scale.
-cl.cov.
-= cluster-level covariances employed in parameter stability tests; ran.eff.
-= estimation initialized with the random effects.
+cl.cov.~= cluster-level covariances employed in parameter stability tests; ran.eff.~= estimation initialized with the random effects on the full sample.
 MAD = Mean absolute deviation from true tree size.}
@@
 \end{figure}
+
+\emph{FIXME: Figure~2 and its caption are inconsistent because LM trees are not on the left but on the right.
+Also, the following paragraphs are inconsistent with respect to the discussion of MAD.
+The first occurrence of MAD is maybe M?}
 
@@
 In terms of MAD, however, the LMM tree estimating random intercepts only performed even slightly better than the latter two approaches.
-Thus, initializing estimation with the random effects can successfully mitigate the overfitting observed for the default LMM trees, but the effect depens strongly on the random-effects specification.
+Thus, initializing estimation with the random effects can successfully mitigate the overfitting observed for the default LMM trees, but the effect depends strongly on the random-effects specification.
 With estimation of both a random intercept and slope, random-effects initialization may lead to subgroup differences being captured by the random effects instead of the tree structure.
@@
   geom_boxplot(aes(x=random, y=tree_size), alpha = .5, width = .6, coef = NULL) +
-  geom_count(aes(x=random, y=tree_size), col = "grey", alpha = .75) +
-  geom_hline(yintercept=3, col = "darkgrey") +
+  geom_count(aes(x=random, y=tree_size), col = "gray", alpha = .75) +
+  geom_hline(yintercept=3, col = "darkgray") +
   geom_text(data = summ_stats, aes(x=random, y=y, label = all),
-            size = 3, col = "grey44") +
+            size = 3, col = "gray44") +
   scale_y_continuous(trans = pseudo_log_trans(), 
@@
 {
-\footnotesize \textit{Note.} Grey circles represent counts, grey lines indicate true number of splits.
+\footnotesize \textit{Note.} Gray circles represent counts, dark gray horizontal lines indicate true number of splits~3.
 Distances on the $y$-axis are on the log scale.
@@
 ggplot(times, aes(Var2, value)) + 
-  geom_violin(, alpha = .5, fill = "grey", width=.5) +
+  geom_violin(, alpha = .5, fill = "gray", width=.5) +
   labs(x = "", y = "Computation time (in seconds)") +
@@
   geom_text(data = summ_stats, aes(x=Var2, y=y, label = M),
-            size = 3, col = "darkgrey") 
+            size = 3, col = "darkgray") 
 @
@@
 
-theme_set(theme_gray(base_size = 10))
+theme_set(theme_bw(base_size = 10))
 ggplot(tmp3) +
@@
 ggplot(ECLSK_times) +
-  geom_violin(aes(x=ind, y=values), alpha = .5, fill = "grey", width=.7) +
+  geom_violin(aes(x=ind, y=values), alpha = .5, fill = "gray", width=.7) +
   scale_y_continuous(trans = "log", limits = c(0.1, 12000),
@@
   geom_text(data = summ_stats, aes(x=ind, y=y, label = M),
-            size = 3, col = "grey44") +
+            size = 3, col = "gray44") +
   labs(tag = "M") + theme(plot.tag.position = c(.12, .95),
@@
 <<echo=FALSE, fig=TRUE, height=1.7>>=
-theme_set(theme_gray(base_size = 8))
+theme_set(theme_bw(base_size = 8))
 cols <- rep(rainbow_hcl(10)[c(3,5)], times = 5)
@@
   labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) + 
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
@@
   labs(x = "", y = "no. of splits") +
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
@@
   labs(x = "", y = "no. of splits", col = expression(p[noise])) +
-  geom_hline(yintercept=3, col = "darkgrey") +
+  geom_hline(yintercept=3, col = "darkgray") +
   theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
@@
   labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
@@
   labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   theme(strip.background = element_blank(), strip.text.x = element_blank()) +
@@
                col = "black", shape=16, size = .4) 
-theme_set(theme_gray(base_size = 11))
+theme_set(theme_bw(base_size = 11))
 @
@@
 \vspace{-.6cm}%
-{\singlespacing \small \textit{Note.} Black dots represent means, grey circles represent counts, grey lines represent true number of splits.
+{\singlespacing \small \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
 Distances on $y$-axis are on log scale.
@@
 lim <- c(0, 18)
-theme_set(theme_gray(base_size = 8))
+theme_set(theme_bw(base_size = 8))
 ggplot(tmp2, aes(x=random, y=tree_size)) +
@@
   labs(x = "", y = "no. of splits", col=expression(N)) + 
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
@@
   labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) +
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
@@
   labs(x = "", y = "no. of splits", col=expression(rho)) +
-  geom_hline(yintercept=3, col = "darkgrey") + 
+  geom_hline(yintercept=3, col = "darkgray") + 
   theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
@@
   labs(x = "", y = "no. of splits", col=expression(sigma[b[1]]^2)) +
-  geom_hline(yintercept=3, col = "darkgrey") +
+  geom_hline(yintercept=3, col = "darkgray") +
   theme(strip.background = element_blank(), strip.text.x = element_blank()) +
@@
                col = "black", shape=16, size = .4)
-theme_set(theme_gray(base_size = 11))
+theme_set(theme_bw(base_size = 11))
 @
@@
 \vspace{-.6cm}%
-{ \singlespacing \small \textit{Note.} Black dots represent means, grey circles represent counts, grey lines represent true number of splits.
+{ \singlespacing \small \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
 Distances on $y$-axis are on log scale.

### 2023-02-10 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
small improvements and comments in Study II
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3253 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 280cf0ddd5185cdb155494d93405045938f33dde

@@
 
+\emph{FIXME: The `Note` below Table~1 (and~2) is too close to the lower horizontal rule, making it hard to read. The order of of the rows should match the order of the panels in
+Figure~2, possibly aided by horizontal rules (then each combination of sigmas only has to be spelled out once).}
+
 <<echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
@@
 probs <- sapply(probs, function(x) {
-  names(x)[is.na(names(x))] <- "no split"
+  names(x)[is.na(names(x))] <- "No split"
   return(x)
@@
 splits[is.na(splits)] <- 0
-for (u_set in list("x2","x3", paste0("x", 4:30), "no split")) {
+for (u_set in list("x2","x3", paste0("x", 4:30), "No split")) {
   splits <- cbind(splits, sapply(probs, function(x) sum(x[names(x) %in% u_set])))
@@
 colnames(splits) <- c("Algorithm", "Random effects", "Fitting approach", "$U_1$", 
-                      "$U_2$", "$U_3$", "$U_4$ - $U_{25}$", "no split")
+                      "$U_2$", "$U_3$", "$U_4$--$U_{25}$", "No split")
 
@@
 This allows us to evaluate the possible (dis)advantages of global versus local estimation of random-effects parameters, as well as the performance of the different splitting criteria employed by each method.
-The same data-generating design as in Simulation Study I was employed.
-To reduce the number of comparisons, we only include performance of LM(M) trees fitted using clustered covariances, because they showed good performance in Simulation Study I.
+The same data-generating design as in Simulation Study~I was employed.
+To reduce the number of comparisons, we only include performance of LM(M) trees fitted using clustered covariances, because they showed good performance in Simulation Study~I.
 
@@
 
-\item We used the default "naive" splitting approach which employs likelihood ratio tests (LRTs) as the splitting criterion \citep{BranyOert13}.
+\item We used the default ``naive'' splitting approach which employs likelihood ratio tests (LRTs) as the splitting criterion \citep{BranyOert13}.
 That is, the likelihood of the SEM fitted to the observations in the current node is compared against the likelihood of a range of two-group SEMs, in which the two groups are defined by each possible candidate split.
@@
 We fitted a single LongCART tree to each dataset.
-The LongCART function estimates node-specific models comprising a random intercept term; this default cannot be overridden.
+The LongCART function estimates node-specific models comprising a random intercept term; this default cannot be changed.
 A fixed-effects model was specified with the response regressed on time and a subject-specific random intercept.
@@
 Table~\ref{tab:first_splits2} presents variable selection frequencies for the first split in the fitted trees.
-For SEM trees, the LRT criterion yields almost perfect accuracy for the first split.
-The score-based approach selected the wrong variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$U_1$`[grep("s_sem", rownames(splits))])))}\% of datasets, but only when the random-effects structure was misspecified.
-When random effects were correctly specified, score-based SEM trees also provided perfect recovery of the first split.
+For SEM trees, the LRT criterion yields almost perfect accuracy for the first split as does the score-based criterion provided the random effects were correctly specified.
+Only in case of misspecification of the random-effects structure, the score-based SEM trees selected the wrong variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$U_1$`[grep("s_sem", rownames(splits))])))}\% of datasets.
 Closer inspection of stability tests seperately per model parameter suggested that the the score-based tests for SEMs are more sensitive to instability in the fixed slope than in the fixed intercept.
@@
 LongCART trees exhibit low accuracy for recovering the first split, selecting the wrong variable in all datasets where at least one split was implemented.
-LongCART showed an even stronger tendency than score-based SEM trees to select $U_2$ or $U_3$ for the first split.
-Of note, in the datasets where LongCART did not implement any splits, in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not reach the criterion for significance.
+LongCART showed a strong tendency to select $U_2$ or $U_3$ for the first split.
+It is noteworthy that in the datasets where LongCART did not implement any splits, in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not reach the criterion for significance.
 This suggests the parameter stability tests proposed by \cite{KundyHare19} are less sensitive to instabilities with respect to the fixed intercept, or less sensitive to instability with respect to a categorical covariate.
@@
 LMM trees require longer computation times because of the estimation of random effects.
-Yet, this increase is minor compared to the computation times required by LongCART and SEM trees: LRT-based SEM trees required strikingly computation times, while score-based SEM trees were computationally more efficient, as expected.
-Computation times for LongCART were in between LRT- and score-based SEM trees.
+Yet, this increase is minor compared to the computation times required by LongCART and SEM trees: LRT-based SEM trees required striking computation times, while score-based SEM trees were computationally more efficient, as expected.
 

### 2023-02-10 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
some improvements and comments in application
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3254 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 62318d166dfa1da296d90779a5c49a79c8f417c0

@@
 
-We analyzed longitudinal assessments of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998-1999 \citep[ECLS-K; ][]{NCES10}.
-Data were collected from 21,304 children from 1,018 schools across the USA were collected.
+We analyzed longitudinal assessments of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}.
+Data were collected from 21,304 children from 1,018 schools across the USA.
 Assessments took place from kindergarten in 1998 through 8th grade in 2007.
@@
 
-Response variables are reading, math and science abilities.
+Response variables are reading, math, and science abilities.
 These were assessed using multi-item cognitive tests, from which latent ability estimates were computed with a mean of zero and variance of one.
@@
 
-We used eleven time-invariant covariates as potential partitioning variables: Gender (51.1\% male); Age in months at baseline (range 53 to 96; M = 6.14 years at baseline); Race (8 categories); First time in kindergarten (yes/no); Socio-economic status (range -5 to 3); Fine motor skills (e.g., drawing figures; range 0 to 9); Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8); Interpersonal skills (range 1 to 4); Self-control (range 1 to 4); Internalizing problem behavior (range 1 to 4); Externalizing problem behavior (range 1 to 4).
+\emph{README: Are these power transformations for months by 1/2 or 2/3 standard in the literature? Should this be cited in some way or is this uncontroversial?}
+
+We used eleven time-invariant covariates as potential partitioning variables: Gender (51.1\% male); Age in months at baseline (range 53 to 96; M = 6.14 years at baseline); Race (8 categories); First time in kindergarten (yes/no); Socio-economic status (range $-$5 to 3); Fine motor skills (e.g., drawing figures; range 0 to 9); Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8); Interpersonal skills (range 1 to 4); Self-control (range 1 to 4); Internalizing problem behavior (range 1 to 4); Externalizing problem behavior (range 1 to 4).
 
@@
 We fitted four LM(M) trees using the approaches which showed most promise in the simulations: An LM tree ($\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$) using clustered covariances in the parameter-stability tests, two LMM trees using clustered covariances with different random-effects specifications ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$ and $\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$), and an LMM tree initializing estimation with the random effects and freely estimating a random intercept only ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$).
+
+\emph{FIXME: Maybe better clarify what `existing data' in the next sentence means.
+I think you are referring to data where we have to evaluate out-of-sample rather than with the `true' effects in-sample, right?}
 
@@
 
+\emph{README: The conclusion not to consider SEM trees which performed rather well in the simulation studies is a bit unsatisfying.
+Also, longRPart was not included in the simulation studies.
+Overall, the new insights gained compared to the simulation studies are not so huge.
+So maybe we should discuss again how to best present this?}
+
 Using package \textbf{longRPart} \citep[version 1.0; ][]{StewyAbdo12}, we fitted two trees with different random effects specifications: A tree comprising a random intercept only, and a tree comprising both a random intercept and slope with respect to student.
@@
 We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, likely more representative of real-world studies in psychology.
-For each combination of response variable (math, reading or science) and sample size, we performed 100 repetitions.
+For each combination of response variable (math, reading, or science) and sample size, we performed 100 repetitions.
 We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample of the current repetition.
@@
 ## for longRPart and longRPart_s are quite similar, and seem to differ
-## most for Science, which was most strongly affected by the mistake.
+## most for science, which was most strongly affected by the mistake.
 
@@
 
-LongRPart trees comprising both random intercepts and slopes yielded estimation errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the Math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the Science data.
+LongRPart trees comprising both random intercepts and slopes yielded convergence errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the science data.
 All other model-fitting attempts converged successfully.
@@
 Table~\ref{tab:application_MSEs} presents means and standard deviations of the MSEs.
-Only few differences in performance between LM(M) and longRPart trees are larger $> 1$ standard error: For Reading with $N=250$ and for Science with both sample sizes.
-Note that the standard errors in the pairwise $t$-tests are get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not.
+Only few differences in performance between LM(M) and longRPart trees are larger $> 1$ standard error: For reading with $N=250$ and for science with both sample sizes.
+Note that the standard errors in the pairwise $t$-tests get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not.
 The latter thus provide a more useful measure of effect size, while the former have higher power to detect differences between methods.
 
-Table~\ref{tab:application_MSEs} indicates that LMM trees estimating only a random intercept and employing cluster-level covariances performed best for predicting Math and Reading abilities.
+Table~\ref{tab:application_MSEs} indicates that LMM trees estimating only a random intercept and employing cluster-level covariances performed best for predicting math and reading abilities.
 For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best.
-Random-effects initialization was the worst-performing LMM tree approach for Math and Reading.
-
-For longRPart, trees comprising both a random intercept and slope performed best for predicting Math and Reading, trees with a random intercept only performed best for predicting Science.
-The differences in performance between Math and Reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of Science abilities.
+Random-effects initialization was the worst-performing LMM tree approach for math and reading.
+
+For longRPart, trees comprising both a random intercept and slope performed best for predicting math and reading, trees with a random intercept only performed best for predicting science.
+The differences in performance between math and reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of science abilities.
 
@@
 \begin{figure}[!bt]
-\caption{Tree sizes for trees fitted to Math (top), Reading (middle) and Science (bottom) ability trajectories.}
+\caption{Tree sizes for trees fitted to math (top), reading (middle), and science (bottom) ability trajectories.}
 \begin{subfigure}{1.2\textwidth}
@@
 Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees.
-For Math and Reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small.
+For math and reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small.
 This pattern reverses with large sample size, where the best-performing LM(M) tree approach yields a larger number of splits than the best-performing longRPart approach.
-For the Math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
+For the math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
 

### 2023-03-22 pkg/glmertree/R/glmertree.R
Version 1.2-2: Added experimental betamertree function
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3258 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6531527a0e69b00698bd34a78dca3b18408c10df

@@ -12,8 +12,8 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
   ## check if data is complete
   all_vars <- if (any(grepl(".", as.character(formula), fixed = TRUE))) 
     names(data) else all.vars(formula)
-  if (nrow(data) != sum(stats::complete.cases(data[ , all_vars]))) {
-    warning("'data' contains missing values, note that listwise deletion will be employed.", immediate. = TRUE) 
+  if (nrow(data[ , all_vars]) != sum(stats::complete.cases(data[ , all_vars]))) {
+    warning("some variables have missing values, note that listwise deletion will be employed.", immediate. = TRUE) 
     data_has_missings <- TRUE
   } else {
     data_has_missings <- FALSE
@@ -83,6 +83,7 @@ lmertree <- function(formula, data, weights = NULL, cluster = NULL,
     old_N <- nrow(data)
     data <- data[complete.cases(data[ , all_vars]), ]
     warning(paste0("New sample size is N = ", nrow(data), " (old sample size was N = ", old_N, ")."))
+    
   }
   
   ## initialization
@@ -238,7 +239,7 @@ glmertree <- function(formula, data, family = "binomial", weights = NULL,
   ## check if data is complete
   all_vars <- if (any(grepl(".", as.character(formula), fixed = TRUE))) 
     names(data) else all.vars(formula)
-  if (nrow(data) != sum(stats::complete.cases(data[ , all_vars]))) {
+  if (nrow(data[ , all_vars]) != sum(stats::complete.cases(data[ , all_vars]))) {
     warning("'data' contains missing values, note that listwise deletion will be employed.", immediate. = TRUE) 
     data_has_missings <- TRUE
   } else {
@@ -460,12 +461,13 @@ fixef.lmertree <- coef.lmertree <-
 
 
 
-VarCorr.lmertree <- VarCorr.glmertree <- function(object, ...) {
-  merMod_type <- ifelse(inherits(object, "lmertree"), "lmer", "glmer")
-  VarCorr(object[[merMod_type]], ...)
+VarCorr.lmertree <- VarCorr.glmertree <- function(x, ...) {
+  merMod_type <- ifelse(inherits(x, "lmertree"), "lmer", "glmer")
+  VarCorr(x[[merMod_type]], ...)
 }
 
 
+# ## Old version of plot lmertree function
 # plot.lmertree <- plot.glmertree <- function(x, which = "all", ask = TRUE, 
 #                                             type = "extended", ...) {    
 # 

### 2023-06-22 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Version 1.2-2: Added experimental betamertree function
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3258 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6531527a0e69b00698bd34a78dca3b18408c10df
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3262 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 8b30d46ca33a3409789ca7731f27d1ba4c5f31fc

@@
 
-\title{Recursive partitioning of linear growth curve models}
+\title{Subgroup detection in linear growth curve models with GLMM trees}
 \shorttitle{Recursive partitioning of linear growth curve models}
@@
 
-The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and SEM trees based on likelihood ratio tests (LRTs).
+The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees.
 All of these methods employ an exhaustive split detection procedure: For every possible split in every available partitioning variable in the current node, the full parametric model is re-estimated in the two nodes that would result from the split.
-To choose the optimal split, some split criterion (such as a $p$-value from the corresponding LRT) is computed for every possible cutpoint.
+To choose the optimal split, some split criterion (such as a $p$-value from a likelihood-ratio test) is computed for every possible cutpoint.
 Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}.
@@
 
-In summary, compared to other methods for partitioning LGCMs, GLMM trees may provide lower computational load, lower model complexity and instability, and unbiased variable selection.
-
-In what follows, we first describe estimation of GLMM trees in general along with the novel adjustments geared towards longitudinal data structures.
-Subsequently, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees.
-In Simulation~Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
-In the Application, we compare the performance of GLMM trees and longRPart in existing datasets on children's development of reading, math and science abilities.
-In the Discussion, we summarize our findings and implications.
+Compared to other methods for partitioning LGCMs, GLMM trees may provide lower computational load, lower model complexity and instability, and unbiased variable selection.
+As such, GLMM trees could prove particularly useful for subgroup detection in longitudinal trajectories.
+In the remainder of this section, we explain how GLMM trees are estimated and propose adjustments that may be particularly useful for partitioning longitudinal trajectories.
+We use a dataset on science ability trajectories te illustrate the relevance of the adjustments.
+Subsequently, we assess the performance of the proposed adjustments to GLMM trees in partitioning LGCMs: In Study~I, we assess performance of the proposed adaptations to GLMM trees in simulated datasets.
+In Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART, in simulated datasets.
+In the Empirical~Evaluation, we compare the performance of the proposed adjustments of GLMM trees in existing datasets on children's development of reading, math and science abilities.
+In the Discussion, we summarize our findings and discuss implications.
 
@@
 
-\section{Estimation}
+\section{Estimation of GLMM trees}
 \label{sec:estimation}
@@
 The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML).
-Thus, when it is known which node~$j$ each subject~$i$ falls into, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates~$\hat \beta_j$ and \emph{global} random-effect estimates~$\hat b_i$.
+Thus, when it is known into which node~$j$ each subject~$i$ falls, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates~$\hat \beta_j$ and \emph{global} random-effect estimates~$\hat b_i$.
 To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}.
@@
 
-In the following two subsections provide some more details on alternative initialization strategies for Step~0 and details of the GLM tree algorithm from Step~1, respecitvely.
+The following two subsections describe alternative initialization strategies for Step~0 and details of the GLM tree algorithm from Step~1, respectively.
 Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve the performance when partitioning LGCMs using longitudinal data.
@@
 Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
-\cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size.
+\cite{SelaySimo12} specifically assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size.
 In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data.
-With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more closely than observations nested within the same unit in cross-sectional data.
-If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations.
+With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data.
+If the random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations.
 Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy.
 More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
+
 The alternative initialization step is thus:
@@
 
-\section{Simulation Study I}
-
-We performed a simulation study to compare the performance GLM(M) trees estimated by initializing estimation with the subgroup structure versus the random effects, and employing observation- versus cluster-level covariances for the parameter stability tests.
+\subsection{Illustration on real data}
+
+To illustrate the proposed adjustments on real data, we use a subsample of 250 children from a much larger dataset of the Early Childhood Longitudinal Study \citep[ECLS-K; ][]{NCES10}.
+Here, we partition trajectories of science ability, which were assessed in 3rd, 5th and 8th grade.
+We used 11 potential partitioning variables, including sociodemographics, motor and interpersonal skills.
+The dataset is described and analyzed in more detail in Study~III.
+With a sample of 250 children, we expect to obtain a relatively small number of relevant subgroups that can more easily be depicted and interpreted.
+
+\begin{figure}
+\caption{Trees resulting from various GLMM tree fitting approaches.}
+\begin{subfigure}{1.1\textwidth}%
+<<fig=TRUE, height=5.5, width=10, echo=FALSE>>=
+library("glmertree")
+load("sizes, N = 250, Science.Rda")
+scie_sizes <- cbind(sizes, N = 250)
+load("bag_ids, N = 250, Science.Rda")
+load("Science ability data.Rdata")
+data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
+data$CHILDID <- factor(data$CHILDID)
+data$GENDER <- factor(data$GENDER)
+data$asmtmm <- factor(data$asmtmm)
+LMM_form <- score ~ months | CHILDID | GENDER + RACE + WKSESL + C1GMOTOR + 
+  C1FMOTOR + T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE
+
+lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
+plot(lmm_tree, which = "tree", type = "simple", gp = gpar(cex = .7))
+icc <- as.data.frame(VarCorr(lmm_tree))
+
+testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
+testdat$CHILDID <- factor(testdat$CHILDID)
+testdat$GENDER <- factor(testdat$GENDER)
+testdat$asmtmm <- factor(testdat$asmtmm)
+## Remove obs with levels of race in test that are not in train
+levs <- unique(testdat$RACE) %in% unique(data$RACE)
+if (!all(levs)) {
+  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
+}
+lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
+@
+\caption{GLMM tree resulting from application of default settings.}
+\end{subfigure}%
+
+\begin{subfigure}{0.8\textwidth}%
+<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
+lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
+plot(lmm_tree_r, which = "tree", type = "simple", gp = gpar(cex = .7))
+icc_r <- as.data.frame(VarCorr(lmm_tree_r))
+lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
+@
+\caption{GLMM tree resulting from use of random-effects initialization.}
+\end{subfigure}%
+
+\begin{subfigure}{0.8\textwidth}%
+<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
+lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
+plot(lmm_tree_c, which = "tree", type = "simple", gp = gpar(cex = .7))
+icc_c <- as.data.frame(VarCorr(lmm_tree_c))
+lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
+@
+\caption{GLMM tree resulting from use of clustered covariances.}
+\end{subfigure}%
+
+\label{fig:LMM_trees}
+\end{figure}
+
+Using the the default fitting approach, the tree in the top panel of Figure~\ref{fig:LMM_trees} was obtained, comprising \Sexpr{(length(lmm_tree$tree)+1)/2} terminal nodes.
+Each node has its own fixed estimate for the intercept (science ability at baseline) and slope of time in months.
+The most important partitioning variable is a measure for socio-economic backgroud (WKSESL), with lower values yielding lower values for the intercept, indicating that children with lower SES score tend to have lower science skills both at the baseline and later assessments.
+Further, we see effects of internalizing problems (T1INTERN), gender, race and fine motor skills (C1FMOTOR), which we do not inspect in detail here as we want to focus on cahnges in results with the proposed adjustments to the estimation approach.
+The proportion of variance in science skill scores explained in science ability scores by the tree is \Sexpr{round(1 - mean((lmm_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, when evaluated on observations not part of the training sample.
+%The intra-class correlation for the random intercept term is \Sexpr{icc$vcov[1]/sum(icc$vcov)}.
+
+The middle panel of Figure~\ref{fig:LMM_trees} depicts the GLMM tree fitted on the science trajectories, by initializing estimation with the random effects.
+A slightly different set of partitioning variables were selected and the tree has less terminal nodes.
+The percentage of variance in science skill scores explained by the tree is \Sexpr{round(1 - mean((lmm_r_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, slightly lower than for the GLMM tree fitted with default settings.
+%The intra-class correlation for the random intercept term is \Sexpr{round(icc_r$vcov[1]/sum(icc_r$vcov), digits = 2)}, slightly higher than for the previous GLMM tree.
+The bottom panel of Figure~\ref{fig:LMM_trees} depicts the GLMM tree fitted on the science trajectories, estimated using clustered covariances.
+The resulting tree is the smallest tree obtained, and selected SES and internalizing problems for partitioning.
+With a proportion of variance explained of \Sexpr{round(1 - mean((lmm_c_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, providing highest predictive accuracy of the three fitting approaches.
+%The intra-class correlation for the random intercept term is \Sexpr{round(icc_c$vcov[1]/sum(icc_c$vcov), digits = 3)}, higher than obtained with the previous two GLMM trees.
+
+\section{Study~1}
 
@@
   \item None: $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$.
-Random effects and thus their variance were fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications.
+Random effects and thus their variance were fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications:
   \begin{itemize}
@@
 This yields linear mixed-effects model (LMM) trees in which only the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0.
-The four variations considered are the following.
+The four variations considered are the following:
   \begin{itemize}
@@
 To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}.
-To compute clustered covariance, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}.
+To compute clustered covariances, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}.
 We employed outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.
@@
 
-We evaluated tree accuracy by counting the number of splits in each tree.
+We evaluated tree accuracy by counting the number of splits in each tree, and computing mean absolute deviation (MAD) from the true tree size (3 splits).
 Trees with $> 3$ splits are indicative of Type-I error, while trees with $< 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables).
@@
 
-\emph{FIXME: Figure~2 and its caption are inconsistent because LM trees are not on the left but on the right.
-Also, the following paragraphs are inconsistent with respect to the discussion of MAD.
-The first occurrence of MAD is maybe M?}
-
 Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach.
-Irrespective of the random-effect specification, the default fitting approaches (left-most box in every panel) tended to overfit, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average.
-This overfitting was successfully mitigated through use of cluster-level covariances in the parameter stability tests, irrespective of the random-effects specification.
-The average number of splits of LMM trees employing clustered covariances were closest to the true number of splits.
-In terms of mean absolute deviation (MAD) from the true tree size, LM trees with clustered covariances performed best, closely followed by the LMM trees employing clustered covariances.
-
-In terms of MAD, however, the LMM tree estimating random intercepts only performed even slightly better than the latter two approaches.
-Thus, initializing estimation with the random effects can successfully mitigate the overfitting observed for the default LMM trees, but the effect depends strongly on the random-effects specification.
+Irrespective of random-effects specification, the default fitting approaches (left-most box in every panel) tended to overfit, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average.
+This overfitting was successfully mitigated through use of cluster-level covariances in the parameter stability tests, again irrespective of the random-effects specification.
+The average number of splits in LMM trees employing clustered covariances was closest to the true number of splits.
+In terms of MAD, however, the LMM tree initializing estimation with the random effects, when only a random intercept was freely estimated, performed best.
+Thus, initializing estimation with the random effects can successfully mitigate overfitting observed for the default LMM trees, but this effect depends strongly on the random-effects specification.
 With estimation of both a random intercept and slope, random-effects initialization may lead to subgroup differences being captured by the random effects instead of the tree structure.
-Use of random-effects initialization and cluster-level covariances in tandem was less effective in reducing the number of splits than using either of the two approaches.
-
-Distributions of the number of splits, separated according to the levels of the data-generating design are presented and discussed in  Figure~\ref{fig:LMM_sizes_interact} and Appendix~\ref{sec:AppendixA}, because they show a pattern very similar to Figure~\ref{fig:LMM_sizes}.
-Use of cluster-level covariances provides the most accurate split recovery.
+
+Irrespective of the random-effects specification, combining the use random-effects initialization and cluster-level covariances was not very effective.
+Use of clustered covariances seems to provide the most reliably improvement in performance.
+
+Distributions of the number of splits, separated according to the levels of the data-generating design are omitted here because this revealed a pattern very similar to Figure~\ref{fig:LMM_sizes}.
+Thus, the optimal fitting approach did not seem dependent on data characteristics.
+
+These results are presented Figure~\ref{fig:LMM_sizes_interact} and discussed in Appendix~\ref{sec:AppendixA}.
 Of the data-generating parameters, $\sigma_{b_1}$ and $N$ showed the strongest effects, with higher values resulting in a higher number of splits.
@@
 Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split: Only very rarely is $U_1$ not selected for the first split.
-
-\emph{FIXME: The `Note` below Table~1 (and~2) is too close to the lower horizontal rule, making it hard to read. The order of of the rows should match the order of the panels in
-Figure~2, possibly aided by horizontal rules (then each combination of sigmas only has to be spelled out once).}
 
@@
 ## Print the table
-kable_styling(add_footnote(kable(splits[1:10, -1], format = "latex", booktabs = TRUE, 
+kable_styling(add_footnote(row_spec(kable(splits[1:10, -1], format = "latex", booktabs = TRUE, 
                                  label = "first_splits", align = "llccccc", 
         escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
-        caption = "Variables selected for the first split by each LM tree (top two rows) and LMM tree (bottom eight rows) estimation approach.", row.names = FALSE, digits = 3L),
+        caption = "Variables selected for the first split by each LM tree (top two rows) and LMM tree (bottom eight rows) estimation approach.", row.names = FALSE, digits = 3L), row = c(2, 6), hline_after =  TRUE),
   "\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3).
@@
 
-\section{Simulation Study II}
+\section{Study~II}
 
@@
 
-\section{Application}
+\section{Study~III}
 
@@
 Time was measured as the number of months since the baseline assessment.
-In order to obtain approximately linear trajectories, $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.
-
-\emph{README: Are these power transformations for months by 1/2 or 2/3 standard in the literature? Should this be cited in some way or is this uncontroversial?}
+In order to obtain approximately linear trajectories, we chose metrics of time based on visual inspection of the observed data: $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.
 
@@
 
-\subsubsection{Partitioning methods}
+\subsubsection{Fitting approaches}
 
@@
 
-\emph{FIXME: Maybe better clarify what `existing data' in the next sentence means.
-I think you are referring to data where we have to evaluate out-of-sample rather than with the `true' effects in-sample, right?}
-
-To compare the performance of LM(M) trees on existing data, we could not employ SEM tree or LongCART, so we opted for longRPart \citep{AbdoyLeBl02}.
+%To compare performance of LM(M) trees on real data, we could not employ SEM tree or LongCART, so we opted for longRPart \citep{AbdoyLeBl02}.
 Neither SEM trees, LongCART, longRPart or longRPart2 implementations provide functionality to obtain predictions for new observations.
@@
 
-\emph{README: The conclusion not to consider SEM trees which performed rather well in the simulation studies is a bit unsatisfying.
-Also, longRPart was not included in the simulation studies.
-Overall, the new insights gained compared to the simulation studies are not so huge.
-So maybe we should discuss again how to best present this?}
-
-Using package \textbf{longRPart} \citep[version 1.0; ][]{StewyAbdo12}, we fitted two trees with different random effects specifications: A tree comprising a random intercept only, and a tree comprising both a random intercept and slope with respect to student.
+%Using package \textbf{longRPart} \citep[version 1.0; ][]{StewyAbdo12}, we fitted two trees with different random effects specifications: A tree comprising a random intercept only, and a tree comprising both a random intercept and slope with respect to student.
 In order to implement a split, longRPart uses a criterion based on the minimum decrease in deviance, which is set to 0.01, by default.
@@
 The ECLS-K datasets have exceptionally large sample sizes.
-We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, likely more representative of real-world studies in psychology.
+We therefore employed random sampling to obtain training samples of $N=250$ children, likely more representative of real-world studies in psychology.
 For each combination of response variable (math, reading, or science) and sample size, we performed 100 repetitions.
@@
 Note that this measures the accuracy of the fixed-effects predictions only.
-Tree size was measured by the number of splits in each tree, and computation time was measured in seconds.
+Tree size was measured by the number of splits in each tree.
 
@@
 \caption{Mean squared errors for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
-\begin{subfigure}{1.2\textwidth}
+\begin{subfigure}{\textwidth}
 <<echo=FALSE, fig=TRUE, height=2, width=8>>=
@@
 math_MSEs <- cbind(MSEs, N = 250)
-load("MSEs, N = 1000, Math.Rda")
-math_MSEs <- rbind(math_MSEs, cbind(MSEs, N = 1000))
+#load("MSEs, N = 1000, Math.Rda")
+#math_MSEs <- rbind(math_MSEs, cbind(MSEs, N = 1000))
 #sapply(math_MSEs, function(x) table(is.na(x)))$longRPart
@@
 read_MSEs <- cbind(MSEs, N = 250)
-load("MSEs, N = 1000, Reading.Rda")
-read_MSEs <- rbind(read_MSEs, cbind(MSEs, N = 1000))
+#load("MSEs, N = 1000, Reading.Rda")
+#read_MSEs <- rbind(read_MSEs, cbind(MSEs, N = 1000))
 #sapply(read_MSEs, function(x) table(is.na(x)))
@@
 scie_MSEs <- cbind(MSEs, N = 250)
-load("MSEs, N = 1000, Science.Rda")
-scie_MSEs <- rbind(scie_MSEs, cbind(MSEs, N = 1000))
+#load("MSEs, N = 1000, Science.Rda")
+#scie_MSEs <- rbind(scie_MSEs, cbind(MSEs, N = 1000))
 #sapply(scie_MSEs, function(x) table(is.na(x)))$longRPart
@@
 ## Compute summary stats
-MSE_df <- data.frame(t(sapply(math_MSEs[ , -c(4,6,7,10)], 
-                              function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(math_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE)))))
-MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
-                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE)))))
-summ_stats <- MSE_df[ , c(1:2,5:6,9:10)]
-names(summ_stats) <- c("math, N = 250", "math, N = 1,000", 
-                       "read, N = 250", "read, N = 250", 
-                       "scie, N = 250", "scie, N = 1,000")
-
-tmp3 <- data.frame(stack(math_MSEs[ , -c(4,6,7,10)]),
+# MSE_df <- data.frame(t(sapply(math_MSEs[ , -c(4,6,7,10)], 
+#                               function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE)))))
+# MSE_df <- cbind(MSE_df, t(sapply(math_MSEs[ , -c(4,6,7,10)], 
+#                                function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE)))))
+# MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
+#                                function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE)))))
+# MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
+#                                function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE)))))
+# MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
+#                                function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE)))))
+# MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
+#                                function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE)))))
+# summ_stats <- MSE_df[ , c(1:2,5:6,9:10)]
+# names(summ_stats) <- c("math, N = 250", "math, N = 1,000", 
+#                        "read, N = 250", "read, N = 1,000", 
+#                        "scie, N = 250", "scie, N = 1,000")
+
+MSE_df <- data.frame(sapply(math_MSEs[ , -c(4,6,7,10)], 
+                              function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE))))
+MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE))))
+MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE))))
+MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE))))
+MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE))))
+MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE))))
+summ_stats <- MSE_df[ , c(1:2,5:6)]
+names(summ_stats) <- c("math, N = 250", "read, N = 250", "scie, N = 250")
+
+MSE_df <- MSE_df[-(5:6) , ] ## omit longRPart trees
+
+tmp3 <- data.frame(stack(math_MSEs[ , -c(4,6,7:10)]),
                     dataset_id = factor(rep(1:nrow(math_MSEs), times = 12)),
@@
 
-theme_set(theme_bw(base_size = 10))
+theme_set(theme_bw(base_size = 12))
 ggplot(tmp3) +
-  geom_violin(aes(x=random, y=MSE, fill = N), 
-               position=position_dodge(1), alpha = .5, width = .7) +
+  geom_violin(aes(x=random, y=MSE), 
+               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
   scale_y_continuous(trans = log_trans(),
@@
   facet_grid(~method2, scales = "free", space = "free") +
-  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
+        legend.position="none") +
   labs(x = "", y = "MSE")
@@
 \end{subfigure}
-\begin{subfigure}{1.2\textwidth}
+\begin{subfigure}{\textwidth}
 <<echo=FALSE, fig=TRUE, height=1.6, width=8>>=
-tmp4 <- data.frame(stack(read_MSEs[ , -c(4,6,7,10)]),
+tmp4 <- data.frame(stack(read_MSEs[ , -c(4,6,7:10)]),
                     dataset_id = factor(rep(1:nrow(read_MSEs), times = 12)),
@@
 ggplot(tmp4) +
-  geom_violin(aes(x=random, y=MSE, fill=N), 
-               position=position_dodge(1), alpha = .5, width = .7) +
+  geom_violin(aes(x=random, y=MSE), 
+               position=position_dodge(1), alpha = .5, width = .6, fill = "gray" ) +
   scale_y_continuous(trans = log_trans(),
@@
   theme(strip.background = element_blank(), strip.text.x = element_blank(),
-        axis.title.x=element_blank(), axis.text.x=element_blank()) +
+        axis.title.x=element_blank(), axis.text.x=element_blank(),
+        legend.position="none") +
   labs(x = "", y = "MSE")
@@
 \end{subfigure}
-\begin{subfigure}{1.2\textwidth}
+\begin{subfigure}{\textwidth}
 <<echo=FALSE, fig=TRUE, height=2, width=8>>=
-tmp5 <- data.frame(stack(scie_MSEs[ , -c(4,6,7,10)]),
+tmp5 <- data.frame(stack(scie_MSEs[ , -c(4,6,7:10)]),
                     dataset_id = factor(rep(1:nrow(scie_MSEs), times = 12)),
@@
 ggplot(tmp5, aes(x=random, y=MSE)) +
-  geom_violin(aes(fill=N), 
-               position=position_dodge(1), alpha = .5, width = .7) +
+  geom_violin(position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
   scale_y_continuous(trans = log_trans(),
@@
   facet_grid(~method2, scales = "free", space = "free") +
-  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  theme(strip.background = element_blank(), strip.text.x = element_blank(),
+        legend.position="none") +
   labs(x = "", y = "MSE") +
@@
 \vspace{-0.5cm}%
-{\footnotesize \textit{Note.
-}The second $y$-axis (far right) gives an estimate of the proportion of variance explained, computed as $1 - \frac{MSE}{var(y)}$.
+{\footnotesize \\ \textit{Note.
+}The second $y$-axis (far right) gives an estimate of the proportion of variance explained, \\computed as $1 - \frac{\text{MSE}}{\text{var}(y)}$.
 Distances on $y$-axis are on the log scale.}
@@
 library("kableExtra")
-MSE_df <- cbind(paste0(MSE_df[,1], " (", MSE_df[,3], ")"),
-                paste0(MSE_df[,2], " (", MSE_df[,4], ")"),
-                paste0(MSE_df[,5], " (", MSE_df[,7], ")"), 
-                paste0(MSE_df[,6], " (", MSE_df[,8], ")"), 
-                paste0(MSE_df[,9], " (", MSE_df[,11], ")"),
-                paste0(MSE_df[,10], " (", MSE_df[,12], ")"))
-rownames(MSE_df) <- c("LM tree$^c$", "LMM tree$^{i,c}$", "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$",  
-                      "longRPart$^i$", "longRPart$^{i,s}$")
-MSE_df <- MSE_df[c(1,2,4,3,5,6), ]
-
-kable_styling(add_header_above(add_header_above(add_footnote(
+# MSE_df <- cbind(paste0(MSE_df[,1], " (", MSE_df[,3], ")"),
+#                 paste0(MSE_df[,2], " (", MSE_df[,4], ")"),
+#                 paste0(MSE_df[,5], " (", MSE_df[,7], ")"), 
+#                 paste0(MSE_df[,6], " (", MSE_df[,8], ")"), 
+#                 paste0(MSE_df[,9], " (", MSE_df[,11], ")"),
+#                 paste0(MSE_df[,10], " (", MSE_df[,12], ")"))
+#MSE_df <- cbind(paste0(MSE_df[,1], " (", MSE_df[,3], ")"),
+#               paste0(MSE_df[,2], " (", MSE_df[,4], ")"),
+#               paste0(MSE_df[,5], " (", MSE_df[,7], ")"), 
+#               paste0(MSE_df[,6], " (", MSE_df[,8], ")"), 
+#               paste0(MSE_df[,9], " (", MSE_df[,11], ")"),
+#               paste0(MSE_df[,10], " (", MSE_df[,12], ")"))
+rownames(MSE_df) <- c("LM tree$^c$", "LMM tree$^{i,c}$", "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$")  
+                      #"longRPart$^i$", "longRPart$^{i,s}$")
+colnames(MSE_df) <- rep(c("M", "SD"), times = 3)
+MSE_df <- MSE_df[c(1,2,4,3), ]
+
+kable_styling(add_header_above(#add_header_above(
+  add_footnote(
   kable(MSE_df, format = "latex", booktabs = TRUE, label = "application_MSEs", align = "c", 
@@
   notation="none", threeparttable = TRUE, escape = FALSE),
-  c(" ", "N = 250"=1, "N = 1,000"=1, "N = 250"=1, "N = 1,000"=1,
-    "N = 250"=1, "N = 1,000"=1), align = "c"),
+  #c(" ", "N = 250"=1, "N = 1,000"=1, "N = 250"=1, "N = 1,000"=1,
+  #  "N = 250"=1, "N = 1,000"=1), align = "c"),
   c(" ", "Math" = 2, "Reading" = 2, "Science" = 2), align = "c"),
-  font_size = 9, full_width=FALSE)
+  font_size = 11, full_width=FALSE)
 @
@@
 
-\subsubsection{Predictive accuracy}
-
-LongRPart trees comprising both random intercepts and slopes yielded convergence errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the science data.
+%LongRPart trees comprising both random intercepts and slopes yielded convergence errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the science data.
 All other model-fitting attempts converged successfully.
@@
 
-Figure~\ref{fig:application_MSEs} presents MSE distributions for each of the partitioning approaches, separated by sample size.
-As expected, all methods show higher accuracy with larger sample sizes.
+Figure~\ref{fig:application_MSEs} presents MSE distributions for each of the estimation approaches and Table~\ref{tab:application_MSEs} presents means and standard deviations of the MSEs.
+%As expected, all methods show higher accuracy with larger sample sizes.
 LM(M) trees tend to show better predictive accuracy than longRPart trees.
@@
 However, these performance differences are small; the effects of sample size are much more obvious than the differences between methods in Figure~\ref{fig:application_MSEs}.
-
-Table~\ref{tab:application_MSEs} presents means and standard deviations of the MSEs.
-Only few differences in performance between LM(M) and longRPart trees are larger $> 1$ standard error: For reading with $N=250$ and for science with both sample sizes.
-Note that the standard errors in the pairwise $t$-tests get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not.
-The latter thus provide a more useful measure of effect size, while the former have higher power to detect differences between methods.
-
+Differences in performance are small, few differences are $> 1$ standard error.
+Note that standard errors in pairwise $t$-tests get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not.
+The latter thus provide a more useful measure of effect size, while the former has higher power to detect differences between methods.
 Table~\ref{tab:application_MSEs} indicates that LMM trees estimating only a random intercept and employing cluster-level covariances performed best for predicting math and reading abilities.
-For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best.
-Random-effects initialization was the worst-performing LMM tree approach for math and reading.
-
-For longRPart, trees comprising both a random intercept and slope performed best for predicting math and reading, trees with a random intercept only performed best for predicting science.
+For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best, while this was the worst-performing approach for math and reading.
+
+%For longRPart, trees comprising both a random intercept and slope performed best for predicting math and reading, trees with a random intercept only performed best for predicting science.
 The differences in performance between math and reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of science abilities.
 
-Again, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with random intercepts, slopes or neither.
+%Again, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with random intercepts, slopes or neither.
 Pairwise $t$-tests did not indicate significant differences between those three approaches for any of the response variables.
@@
 math_sizes <- cbind(sizes, N = 250)
-load("sizes, N = 1000, Math.Rda")
-math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
+#load("sizes, N = 1000, Math.Rda")
+#math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
 #sapply(math_sizes, function(x) table(is.na(x)))
@@
 read_sizes <- cbind(sizes, N = 250)
-load("sizes, N = 1000, Reading.Rda")
-read_sizes <- rbind(read_sizes, cbind(sizes, N = 1000))
+#load("sizes, N = 1000, Reading.Rda")
+#read_sizes <- rbind(read_sizes, cbind(sizes, N = 1000))
 #sapply(read_sizes, function(x) table(is.na(x)))
@@
 scie_sizes <- cbind(sizes, N = 250)
-load("sizes, N = 1000, Science.Rda")
-scie_sizes <- rbind(scie_sizes, cbind(sizes, N = 1000))
+#load("sizes, N = 1000, Science.Rda")
+#scie_sizes <- rbind(scie_sizes, cbind(sizes, N = 1000))
 #sapply(scie_sizes, function(x) table(is.na(x)))
@@
 
-tmp6 <- data.frame(stack(math_sizes[ , -c(4,6,7,10)]),
+tmp6 <- data.frame(stack(math_sizes[ , -c(4,6,7:10)]),
                     dataset_id = factor(rep(1:nrow(math_sizes), times = 12)),
@@
 ggplot(tmp6) +
-  geom_violin(aes(x=random, y=splits, fill = N), 
-               position=position_dodge(1), alpha = .5, width = .7) +
+  geom_violin(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
   facet_grid(~method2, scales = "free", space = "free") +
@@
 <<echo=FALSE, fig=TRUE, height=1.6, width=8>>=
-tmp7 <- data.frame(stack(read_sizes[ , -c(4,6,7,10)]),
+tmp7 <- data.frame(stack(read_sizes[ , -c(4,6,7:10)]),
                     dataset_id = factor(rep(1:nrow(read_sizes), times = 12)),
@@
 ggplot(tmp7) +
-  geom_violin(aes(x=random, y=splits, fill=N), 
-               position=position_dodge(1), alpha = .5, width = .7) +
+  geom_violin(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
   facet_grid(~method2, scales = "free", space = "free") +
@@
 <<echo=FALSE, fig=TRUE, height=2, width=8>>=
-tmp8 <- data.frame(stack(scie_sizes[ , -c(4,6,7,10)]),
+tmp8 <- data.frame(stack(scie_sizes[ , -c(4,6,7:10)]),
                     dataset_id = factor(rep(1:nrow(scie_sizes), times = 12)),
@@
 ggplot(tmp8) +
-  geom_violin(aes(x=random, y=splits, fill=N), 
-               position=position_dodge(1), alpha = .5, width = .7) +
+  geom_violin(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
   facet_grid(~method2, scales = "free", space = "free") +
@@
 
-\begin{figure}[h]
-\caption{Computation times for partitioning math, reading and science ability trajectories.}
-\begin{subfigure}{.5\textwidth}
-<<echo=FALSE, fig=TRUE, height=3.25, width=3>>=
-load("times, N = 250, Reading.Rda")
-read_times <- cbind(times, N = 250)
-load("times, N = 1000, Reading.Rda")
-read_times <- rbind(read_times, cbind(times, N = 1000))
-
-load("times, N = 250, Math.Rda")
-math_times <- cbind(times, N = 250)
-load("times, N = 1000, Math.Rda")
-math_times <- rbind(math_times, cbind(times, N = 1000))
-
-load("times, N = 250, Science.Rda")
-scie_times <- cbind(times, N = 250)
-load("times, N = 1000, Science.Rda")
-scie_times <- rbind(scie_times, cbind(times, N = 1000))
-
-## Combine all LMM trees
-ECLSK_times <- rbind(stack(math_times[ , -c(6, 10)]), 
-                     stack(read_times[ , -c(6, 10)]),
-                     stack(scie_times[ , -c(6, 10)]))
-ECLSK_times$ind <- as.character(ECLSK_times$ind)
-ECLSK_times$ind[grepl("LMMtree", ECLSK_times$ind)]<- "LMM tree"
-ECLSK_times$ind[grepl("longRPart", ECLSK_times$ind)]<- "longRPart"
-ECLSK_times$ind[ECLSK_times$ind == "LMtree_c"] <- "LM tree"
-ECLSK_times$ind <- factor(ECLSK_times$ind)
-
-## Compute summary stats
-mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
-summ_stats <- data.frame(M = sprintf(mean_times, fmt = '%#.2f'),
-                         ind = names(mean_times),
-                         y = 11000)
-## Plot
-ggplot(ECLSK_times) +
-  geom_violin(aes(x=ind, y=values), alpha = .5, fill = "gray", width=.7) +
-  scale_y_continuous(trans = "log", limits = c(0.1, 12000),
-                     breaks = c(.01, .1, 1, 10, 100, 1000)) +
-  labs(x = "", y = "Computation time (in seconds)") +
-  geom_text(data = summ_stats, aes(x=ind, y=y, label = M),
-            size = 3, col = "gray44") +
-  labs(tag = "M") + theme(plot.tag.position = c(.12, .95),
-                          plot.tag = element_text(size = 10)) 
-@
-\end{subfigure}
-\label{fig:application_times}
-\end{figure}
-
-Figure~\ref{fig:application_sizes} presents tree size distributions for each of the partitioning approaches.
-Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees.
+%\begin{figure}[h]
+%\caption{Computation times for partitioning math, reading and science ability trajectories.}
+%\begin{subfigure}{.5\textwidth}
+<<eval=FALSE, echo=FALSE, fig=TRUE, height=3.25, width=3>>=
+# load("times, N = 250, Reading.Rda")
+# read_times <- cbind(times, N = 250)
+# load("times, N = 1000, Reading.Rda")
+# read_times <- rbind(read_times, cbind(times, N = 1000))
+# 
+# load("times, N = 250, Math.Rda")
+# math_times <- cbind(times, N = 250)
+# load("times, N = 1000, Math.Rda")
+# math_times <- rbind(math_times, cbind(times, N = 1000))
+# 
+# load("times, N = 250, Science.Rda")
+# scie_times <- cbind(times, N = 250)
+# load("times, N = 1000, Science.Rda")
+# scie_times <- rbind(scie_times, cbind(times, N = 1000))
+# 
+# 
+# ## Combine all LMM trees
+# ECLSK_times <- rbind(stack(math_times[ , -c(6, 10)]), 
+#                      stack(read_times[ , -c(6, 10)]),
+#                      stack(scie_times[ , -c(6, 10)]))
+# ECLSK_times$ind <- as.character(ECLSK_times$ind)
+# ECLSK_times$ind[grepl("LMMtree", ECLSK_times$ind)]<- "LMM tree"
+# ECLSK_times$ind[grepl("longRPart", ECLSK_times$ind)]<- "longRPart"
+# ECLSK_times$ind[ECLSK_times$ind == "LMtree_c"] <- "LM tree"
+# ECLSK_times$ind <- factor(ECLSK_times$ind)
+# 
+# ## Compute summary stats
+# mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
+# summ_stats <- data.frame(M = sprintf(mean_times, fmt = '%#.2f'),
+#                          ind = names(mean_times),
+#                          y = 11000)
+# ## Plot
+# ggplot(ECLSK_times) +
+#   geom_violin(aes(x=ind, y=values), alpha = .5, fill = "gray", width=.7) +
+#   scale_y_continuous(trans = "log", limits = c(0.1, 12000),
+#                      breaks = c(.01, .1, 1, 10, 100, 1000)) +
+#   labs(x = "", y = "Computation time (in seconds)") +
+#   geom_text(data = summ_stats, aes(x=ind, y=y, label = M),
+#             size = 3, col = "gray44") +
+#   labs(tag = "M") + theme(plot.tag.position = c(.12, .95),
+#                           plot.tag = element_text(size = 10)) 
+@
+%\end{subfigure}
+%\label{fig:application_times}
+%\end{figure}
+
+Figure~\ref{fig:application_sizes} presents tree size distributions.
+%Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees.
 For math and reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small.
@@
 For the math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
-
-\subsubsection{Computation time}
-
-Figure~\ref{fig:application_times} depicts computation time distributions of the different partitioning algorithms.
-A clear computational advantage is observed for LM trees ($M =$ \Sexpr{round(mean_times["LM tree"], digits = 2)} seconds).
-LMM trees require longer computation times ($M =$ \Sexpr{round(mean_times["LMM tree"], digits = 2)} seconds).
-This increase is minor compared to LongRPart trees, which required longest computation times ($M =$ \Sexpr{round(mean_times["longRPart"], digits = 2)} seconds).
+Random-effects initialization tends to yield the largest trees, while LM trees with clustered covariances tend to yield the smallest trees.
+LMM trees with clustered covariances yields somewhat larger trees, but smaller than LMM trees with random-effects initialization.
+
+%\subsubsection{Computation time}
+
+%Figure~\ref{fig:application_times} depicts computation time distributions of the different partitioning algorithms.
+A clear computational advantage is observed for LM trees ($M =$ \Sexpr{#round(mean_times["LM tree"], digits = 2)} seconds).
+LMM trees require longer computation times ($M =$ \Sexpr{#round(mean_times["LMM tree"], digits = 2)} seconds).
+This increase is minor compared to LongRPart trees, which required longest computation times %($M =$ \Sexpr{#round(mean_times["longRPart"], digits = 2)} seconds).
 
@@
 
-In line with expectations, the simulation study showed that the default estimation approach for GLMM trees yields too many splits when partitioning LGCMs.
+Study~I showed that the default estimation approach for GLMM trees yields too many splits when partitioning LGCMs.
 As hypothesized, performing parameter-stability tests using clustered covariances mitigated this overfitting.
-Also as expected, initializing estimation with the random effects mitigated overfitting, but too strongly when both random intercepts and slopes were estimated.
-The performance of of LM(M) trees using clustered covariances appeared largely unaffected by (mis-)specification of the random effects, and could therefore be the safest choice in practice.
-Combining cluster-level covariances and random-effects initialization worsened performance, os in practice users should use either one of the two approaches.
-
-The simulations showed comparable performance of LM(M) and SEM trees in partitioning LGCMs.
+As hypothesized, initializing estimation with the random effects mitigated overfitting, but too strongly when both random intercepts and slopes were estimated.
+The performance of LM(M) trees using clustered covariances appeared largely unaffected by (mis-)specification of the random effects, and use of clustered covariances may therefore be the safest choice in practice.
+Combining cluster-level covariances and random-effects initialization worsened performance and is thus best avoided.
+The results also indicated that the optimal estimation approach for LMM trees does not depend on sample size, variance of the random effect(s), the number of potential partitioning variables or the intercorrelation between partioning variables.
+
+Study~II showed comparable performance of LM(M) and SEM trees in partitioning LGCMs.
 SEM trees did however show stronger sensitivity to misspecification of the random effects, with underspecification yielding a higher number of splits.
@@
 The LongCART parameter stability tests \citep{KundyHare19} may be underpowered for detecting systematic instability of the fixed intercept, or for detecting instability with respect to categorical covariates.
-
-In the application, we found higher predictive accuracy for LM(M) trees than for longRPart, but differences in performance were small.
-In line with the simulation studies, we found performance of LM(M) trees did not depend on (mis-)specification of the random effects, which may provide an important practical advantage.
-
-Both simulation and application studies indicated a substantial computational advantage of GLM(M) trees.
-This is likely due to the local-global estimation approach employed by GLMM trees, where fixed-effects parameters are estimated locally within a node and random-effects parameters are estimated globally, using all observations.
-In contrast, SEM trees, LongCART and longRPart fit the full mixed-effects model in each node, which may substantially increase computational burden.
-The local-global estimation approach also reduces model complexity and the good performance of GLMM trees in the current study suggests this may be beneficial for stability and generalizability.
+As in Study~I, these results were not strongly affected by sample size, variance of the random effect(s), the number of potential partitioning variables or the intercorrelation between partioning variables.
+
+Study~III, which evaluated LM(M) tree performance in partitioning math, reading and science trajectories, we found only small differences in terms of predictive accuracy.
+In line with the simulation studies, we found good performance of LMM trees employing clustered covariances.
+Although GLMM trees initializing estimation with the random effects performed best on partitioning science trajectories, this slight improvement came at the cost of substantial increases in tree size.
+
+Study~II indicated a clear computational advantage for LM(M) trees.
+This is in large part due to the local-global estimation approach employed by LMM trees, where fixed-effects parameters are estimated locally within a node and random-effects parameters are estimated globally, using all observations.
+In contrast, SEM trees, LongCART and longRPart fit the full mixed-effects model in each node, which substantially increases computational load.
+The local-global estimation approach also reduces model complexity and the good performance of LMM trees in the current study suggest it also benefits stability and generalizability.
 
@@
 Alternatively, researchers may want to use the parameter stability tests for mixed-effects models developed by \cite{WangyMerk18} \citep[see also ][]{WangyMerk21,WangyGrav22}.
-This will be useful, for example, when the number and/or distances between timepoints differs between respondents and SEM-based growth curve models can thus not be applied \cite{NeisyMatt18}.
-
-The focus of the current study was limited to Gaussian responses and LGCMs.
+This will be useful, for example, when the number and/or distances between timepoints differs between respondents and SEM-based growth curve models can thus not be applied \citep{NeisyMatt18}.
+
+The current evaluations were limited to Gaussian responses and LGCMs.
 Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses.
@@
 Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances.
-Future studies could assess whether computation and use of the full sandwich estimator further improves performance of the parameter stability tests.
+Future studies may assess whether computation and use of the full sandwich estimator further improves performance of the parameter stability tests.
 

### 2023-06-26 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
added new growth curve plot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3264 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fba33cd90a3336a2e4a788163a1fffcde8ef6392

@@
 
+<<growthplot, echo=FALSE, results=hide>>=
+growthplot <- function(object, cluster = NULL, ...) {
+  if(is.null(cluster)) cluster <- object$tree$fitted[["(cluster)"]]
+  if(is.null(cluster)) cluster <- Formula::model.part(Formula::Formula(object$formula),
+    data = model.frame(object$lmer), lhs = 0, rhs = 2L)[[1L]]
+  plot(object$tree, terminal_panel = node_growthplot, tp_args = list(cluster = cluster, ...))
+}
+
+node_growthplot <- function(mobobj, cluster = NULL,
+  id = TRUE, pop = TRUE, bg = "white",
+  curvecol = gray(0.2, alpha = 1/6), curvelwd = 2,
+  linecol = 2, linelwd = 3,
+  ylines = NULL, xlab = FALSE, ylab = FALSE,
+  mainlab = NULL, gp = gpar(...), ...)
+{
+  ## obtain dependent variable
+  mf <- model.frame(mobobj)
+  y <- Formula::model.part(mobobj$info$Formula, mf, lhs = 1L, rhs = 0L)
+  if(isTRUE(ylab)) ylab <- names(y)
+  if(identical(ylab, FALSE)) ylab <- ""
+  if(is.null(ylines)) ylines <- ifelse(identical(ylab, ""), 3, 5)
+  y <- y[[1L]]
+
+  ## obtain explanatory variables
+  X <- Formula::model.part(mobobj$info$Formula, mf, lhs = 0L, rhs = 1L)
+  X <- X[,NCOL(X),drop=FALSE]
+  xlab <- if(!identical(xlab, FALSE)) { if(isTRUE(xlab)) colnames(X) else xlab[1L] } else ""
+
+  ## axis scaling
+  xscale <- range(X) + c(-0.1, 0.1) * diff(range(X))
+  yscale <- range(y) + c(-0.1, 0.1) * diff(range(y))
+
+  ## other information
+  fitted <- mobobj$fitted[["(fitted)"]]
+  cf <- coef(mobobj)
+
+  rval <- function(node) {
+
+    ## node index
+    nid <- id_node(node)
+    ix <- fitted %in% nodeids(mobobj, from = nid, terminal = TRUE)
+
+    ## subsets
+    y <- y[ix]
+    x <- X[ix, 1L]
+    cf <- cf[as.character(nid), ]
+
+    ## set up top viewport
+    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3,
+		       widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")),
+                       heights = unit(c(1, 1), c("lines", "null"))),
+		       width = unit(1, "npc"),
+                       height = unit(1, "npc") - unit(2, "lines"),
+		       name = paste("node_growthplot", nid, sep = ""),
+                       gp = gp)
+    pushViewport(top_vp)
+    grid.rect(gp = gpar(fill = bg, col = 0))
+
+    ## main title
+    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
+    pushViewport(top)
+
+    if (is.null(mainlab)) { 
+      mainlab <- if(id) {
+    	function(id, nobs) sprintf("Node %s (n = %s)", id, nobs)
+      } else {
+    	function(id, nobs) sprintf("n = %s", nobs)
+      }
+    }
+    if (is.function(mainlab)) {
+      mainlab <- mainlab(nid, info_node(node)$nobs)
+    }
+    grid.text(mainlab)
+    popViewport()
+
+    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2,
+                     xscale = xscale, yscale = yscale,
+                     name = paste0("node_growthplot", nid, "plot"),
+                     clip = FALSE)
+    pushViewport(plot)
+
+    grid.xaxis(at = c(ceiling(xscale[1L] * 10), floor(xscale[2L] * 10))/10)
+    grid.yaxis(at = c(ceiling(yscale[1L] * 10), floor(yscale[2L] * 10))/10)
+    grid.rect(gp = gpar(fill = "transparent"))
+    grid.clip()
+
+    if(!is.null(cluster)) {
+      cluster <- factor(cluster[ix])
+      for(j in levels(cluster)) {
+        jx <- which(cluster == j)
+        grid.lines(unit(x[jx], "native"), unit(y[jx], "native"), gp = gpar(col = curvecol, lwd = curvelwd))
+      }
+    }
+    grid.abline(unit(cf[1], "native"), unit(cf[2], "native"), gp = gpar(col = linecol, lwd = linelwd))
+
+    if(ylab != "") grid.text(ylab, y = unit(0.5, "npc"), x = unit(-2.5, "lines"), rot = 90)
+    if(xlab != "") grid.text(xlab, x = unit(0.5, "npc"), y = unit(-2, "lines"))                
+    if(pop) popViewport(2L) else upViewport(2L)
+
+  }
+
+  return(rval)
+}
+class(node_growthplot) <- "grapcon_generator"
+@
+
 To illustrate the proposed adjustments on real data, we use a subsample of 250 children from a much larger dataset of the Early Childhood Longitudinal Study \citep[ECLS-K; ][]{NCES10}.
@@
 lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
-plot(lmm_tree, which = "tree", type = "simple", gp = gpar(cex = .7))
+growthplot(lmm_tree)
 icc <- as.data.frame(VarCorr(lmm_tree))
@@
 lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
-plot(lmm_tree_r, which = "tree", type = "simple", gp = gpar(cex = .7))
+growthplot(lmm_tree_r)
 icc_r <- as.data.frame(VarCorr(lmm_tree_r))
@@
 lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
-plot(lmm_tree_c, which = "tree", type = "simple", gp = gpar(cex = .7))
+growthplot(lmm_tree_c)
 icc_c <- as.data.frame(VarCorr(lmm_tree_c))
@@
 \label{fig:LMM_trees}
+\end{figure}
+
+\begin{figure}
+\caption{GLMM tree for linear growth curves of science abilities among 250~children in 3rd, 5th, and 8th grade.}
+
+<<fig=TRUE, height=9, width=15, echo=FALSE>>=
+growthplot(lmm_tree_c)
+@
+
+\label{fig:LMM_tree_c}
 \end{figure}

### 2023-08-14 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
LGCM partitioning paper update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3267 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6082b34566c1224f807facc12177ede44b1fce45

--- 2023-06-22 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
@@
 \usepackage{setspace}
+\usepackage{hyperref}
 
 \renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
+\renewcommand{\thefootnote}{\fnsymbol{footnote}} % uses symbols instead of numbers for footnotes
 
@@
 \abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time.
-Heterogeneity between subjects is often encountered in such models and researchers are typically interested in explaining or predicting this heterogeneity.
-We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups associated with different sets of parameters in linear growth curve models.
+Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity.
+We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with differently shaped trajectories in linear growth curve models.
 Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data.
 The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.
-We evaluate GLMM tree's performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities.
-We compare the performance of GLMM trees with three alternative methods: structural equation model (SEM) trees, LongCART, and longRPart.
-GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart.
-In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}
+We evaluate GLMM trees' performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities.
+We compare the performance of GLMM trees with alternative methods: structural equation model (SEM) trees and LongCART.
+GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART.
+In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.\\}
 
@@
 \section{Introduction}
+
 \label{sec:Introduction}
@@
 
+<<echo=FALSE>>=
+library("glmertree")
+load("bag_ids, N = 250, Science.Rda")
+load("Science ability data.Rdata")
+names(sciedata) <- c("CHILDID", "GENDER", "RACE", "SES", "GMOTOR", "FMOTOR", 
+                 "INTERN", "EXTERN", "INTERP", "CONTRO", "FIRKDG", 
+                 "AGEBASELINE", "science", "score", "asmtmm", "months")
+data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
+data$CHILDID <- factor(data$CHILDID)
+data$GENDER <- factor(data$GENDER)
+data$asmtmm <- factor(data$asmtmm)
+library("lmerTest")
+LMM <- lmer(score ~ months + (1|CHILDID), data = data)
+icc <- data.frame(VarCorr(LMM))
+@
+
 Development over time is of prime interest in psychological research.
-For example, researchers may want to model and predict student's reading or math development in educational studies \citep{MorgyFark09}, or patient's symptom reduction over the course of clinical trails \citep{Fein09}.
-Mixed-effects or latent-variable models can be used to model development over time, and allow for explaining possible heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}.
-However, the covariates that are predictive of development may often not be known in advance, and methods for identifying them are needed.
+For example, in educational studies researchers may want to model student's academic development over time; in clinical studies researchers may want to model patients' symptoms over time.
+Mixed-effects or latent-variable models can be used to model such trajectories and allow for explaining heterogeneity with covariates of a-priori known relevance \citep[e.g., ][]{NeisyMatt18}.
+However, when the covariates that can explain heterogeneity are not known in advance, methods for identifying them are needed.
+
+For example, Figure~\ref{fig:global_trajectories} shows trajectories of science knowledge and skills from 250 children who were assessed at three timepoints throughout grades 3 through 8\footnote{Further details on the source of these data and scaling of the response variable are provided in Study~III.}.
+The grey lines depict individual trajectories, while the red line depicts the estimated average trajectory.
+Although the red line describes the average trajectory well, the grey lines indicate substantial intra-individual variability in average levels and growth over time.
+A common research aim might be to identify which covariates can explain or predict this heterogeneity.
+
+\begin{figure}%
+\caption{Growth curves of science ability.}
+\begin{subfigure}{0.65\textwidth}%
+<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
+y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
+x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
+plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
+     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
+     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
+axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
+sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
+lines(x = c(min(data$months), max(data$months)), 
+     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
+     lwd = 3, col = 2)
+@
+\\\footnotesize{\textit{Note.} Grey lines depict observed individual trajectories.
+The red line depicts the average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.}
+\end{subfigure}%
+\label{fig:global_trajectories}
+\end{figure}%
+
+\begin{figure}%
+\caption{Partitioned growth curves of science abilities.}
+\begin{subfigure}{0.8\textwidth}%
+<<fig=TRUE, height=9, width=12, echo=FALSE>>=
+LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
+  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
+testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
+testdat$CHILDID <- factor(testdat$CHILDID)
+testdat$GENDER <- factor(testdat$GENDER)
+testdat$asmtmm <- factor(testdat$asmtmm)
+## Remove obs with levels of race in test that are not in train
+levs <- unique(testdat$RACE) %in% unique(data$RACE)
+if (!all(levs)) {
+  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
+}
+lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
+lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
+plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
+     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
+                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
+                                        xaxis.labs = c(0, 12, 36, 60)),
+     gp = gpar(cex = 1.5))
+icc_c <- as.data.frame(VarCorr(lmm_tree_c))
+lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
+lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)
+@
+\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
+Grey lines depict observed individual trajectories.
+Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercepts with respect to individuals.}
+\end{subfigure}%
+\label{fig:LMM_tree_c}
+\end{figure}%
+
+<<echo=FALSE, results=tex, fig.pos='b'>>=
+library("kableExtra")
+coefs <- coef(lmm_tree_c)
+coefs <- cbind(Node = as.numeric(rownames(coefs)), coefs)
+colnames(coefs) <- c("Node", "Intercept", "Slope")
+kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs", 
+                             align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
+        caption = "Fixed-effects coefficient estimates from the terminal nodes of Figure 2.", 
+        row.names = FALSE, digits = 3L),
+  font_size = 11, full_width=FALSE)
+@
 
 Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates.
-Several RPMs that allow for partitioning growth-curve models have been developed: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs).
-Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.
-
-Here we extend generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} so that they can be applied to partitioning LGCMs.
+Figure~\ref{fig:LMM_tree_c} shows an example tree, which identified socio-economic status (SES) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of these children, assessed at baseline.
+Four subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope (see Table~\ref{tab:local_coefs}).
+For children in the group with lower levels of SES, the intercept is also lowest, indicating lower science ability.
+For intermediate levels of SES (between -0.41 and 1.23), there is also an effect of internalizing problems, where children with higher internalizing problems have a lower intercept.
+Highest science ability intercepts are observed in the group with highest SES.
+It appears that nodes with higher intercepts also tend to have higher slopes, indicating that children with relatively higher ability also show larger ability gains over time.
+
+The tree in Figure~\ref{fig:LMM_tree_c} has been estimated with generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}.
 GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas.
-Moreover, the general idea of GLMM trees is also appealing for subgroup detection in almost any type of mixed-effects model.
+In the current paper we aim to extend GLMM trees, so that they can be applied to partitioning LGCMs.
+The general idea of GLMM trees is appealing for subgroup detection in almost any type of mixed-effects model.
 Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates).
 In this paper, we propose and test two extensions of GLMM trees that account for these characteristics.
-We focus on the specific use case of partitioning LGCMs, but the extensions are also critical for a wide range of further settings where covariates are measured at higher levels and/or where random-effects variance is substantial.
+We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wider range of settings where covariates are measured at higher levels and/or where the random effects have substantial variance.
+
+There are several alternative RPMs that can be used to partition linear growth-curve models (LGCMs): GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}.
+Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.
 
 The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the subgroups defined by the terminal nodes of the tree.
-Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations.
+Instead, fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations.
 This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes.
-With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \citep{FokkySmit18}.
-
-All other methods for partitioning LGCMs take a fully local estimation approach: Within every subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that node only.
+With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects in mixed-effects models \citep{FokkySmit18}.
+
+Other methods for partitioning LGCMs take a fully local estimation approach: Within every node or subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that subgroup only.
 This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity.
-GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model.
+GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results.
 Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.
 
-The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees.
-All of these methods employ an exhaustive split detection procedure: For every possible split in every available partitioning variable in the current node, the full parametric model is re-estimated in the two nodes that would result from the split.
-To choose the optimal split, some split criterion (such as a $p$-value from a likelihood-ratio test) is computed for every possible cutpoint.
+The computational advantage of GLMM trees is strongest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees.
+These methods employ an exhaustive split detection procedure, where for every possible split point in the current node, the full parametric model needs to be re-estimated in the two resulting nodes.
+To choose the optimal split, the splitting criterion (such as a $p$-value from a likelihood-ratio test) is derived from these two models.
 Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}.
-LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split.
-This not only reduces computational load, it also allows for separating variable and cutpoint selection, thus preventing selection bias.
-The GLMM tree algorithm also employs a two-step approach to split selection.
-
-Compared to other methods for partitioning LGCMs, GLMM trees may provide lower computational load, lower model complexity and instability, and unbiased variable selection.
-As such, GLMM trees could prove particularly useful for subgroup detection in longitudinal trajectories.
-In the remainder of this section, we explain how GLMM trees are estimated and propose adjustments that may be particularly useful for partitioning longitudinal trajectories.
-We use a dataset on science ability trajectories te illustrate the relevance of the adjustments.
-Subsequently, we assess the performance of the proposed adjustments to GLMM trees in partitioning LGCMs: In Study~I, we assess performance of the proposed adaptations to GLMM trees in simulated datasets.
-In Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART, in simulated datasets.
-In the Empirical~Evaluation, we compare the performance of the proposed adjustments of GLMM trees in existing datasets on children's development of reading, math and science abilities.
+LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not require model refitting for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split.
+This reduces computational load, while it also allows for separating variable and cutpoint selection, thus preventing selection bias.
+The GLMM tree algorithm also employs a two-step approach to split selection and thereby shares these advantages.
+
+Given this lower model complexity, lower computational burden and lack of variable selection bias, GLMM trees might be particularly useful for subgroup detection in LGCMs.
+In the remainder of this section, we explain how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories.
+Next, we assess the performance of the proposed adjustments to GLMM trees in partitioning LGCMs: In Study~I, we assess performance of the proposed adaptations to GLMM trees in simulated datasets.
+In Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
+In the Study~III, we compare the performance of the proposed adjustments of GLMM trees in existing datasets on children's development of reading, math and science abilities.
 In the Discussion, we summarize our findings and discuss implications.
@@
 In the following, we assume that time is the predictor variable of interest (i.e., $p = 1$), where the number $n_i$ and spacing of observed timepoints may differ between subjects.
-The fixed-effects parameters $\beta_j$ (here, intercept and time slope) in GLMM trees are node-specific, i.e., their value depends on the subgroup/node~$j$ which subject~$i$ falls into.
+The fixed-effects parameters $\beta_j$ (here, intercept and time slope) in GLMM trees are node-specific, i.e., their value depends on the subgroup/node~$j$ into which subject~$i$ falls.
 As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is the corresponding vector of random effects for subject $i$.
@@
 To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}.
-The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm.
-The initialization simply assumes zero random effects.
+The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:
 %
@@
 %
+This initialization simply assumes zero random effects.
 Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3.
@@
 
-The following two subsections describe alternative initialization strategies for Step~0 and details of the GLM tree algorithm from Step~1, respectively.
-Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve the performance when partitioning LGCMs using longitudinal data.
+The following two subsections describes an alternative initialization strategy for Step~0 and an alternative approach for fitting the GLM tree from Step~1.
+Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve performance when partitioning longitudinal data.
 
@@
 If the random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations.
-Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy.
-More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
+Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery.
+Specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
 
@@
 \end{enumerate}
+
+\begin{figure}%
+\caption{GLMM tree estimated by initializing with zero random effects.}
+<<fig=TRUE, height=7, width=12, echo=FALSE, out.width=12>>=
+lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
+plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
+     nodesize_level = "CHILDID", tp_args = 
+       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
+            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
+icc <- as.data.frame(VarCorr(lmm_tree))
+lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
+@
+\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability.
+Grey lines depict observed individual trajectories.
+Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
+\label{fig:LMM_tree}
+\end{figure}%
+
+\begin{figure}%
+\caption{GLMM tree initialized by estimating the random effects.}
+<<fig=TRUE, height=9, width=15, echo=FALSE>>=
+lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
+plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
+     nodesize_level = "CHILDID", tp_args = 
+       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
+            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
+icc_r <- as.data.frame(VarCorr(lmm_tree_r))
+lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
+@
+\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
+Grey lines depict observed individual trajectories.
+Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
+\label{fig:LMM_tree_r}
+\end{figure}%
+
+To illustrate how initialization may make a difference we applied both approaches to the dataset from Figures~\ref{fig:global_trajectories}~and~\ref{fig:LMM_tree_c}.
+Initializing estimation assuming zero random effects yielded the tree in Figure~\ref{fig:LMM_tree}, with \Sexpr{width(lmm_tree$tree)} subgroups and an estimated random-intercept variance of \Sexpr{formatC(data.frame(VarCorr(lmm_tree$lmer))$vcov[1], digits = 2)}.
+Additional splits were implemented based on gender, race, internalizing problems and fine motor skills.
+Considering the relatively large number of subgroups and relatively small sample sizes in some of the terminal nodes, this tree may not generalize well to other samples.
+Initializing with estimation of the random effects yielded the tree depicted in Figure~\ref{fig:LMM_tree_r}, with \Sexpr{width(lmm_tree_r$tree)} subgroups and an estimated random-intercept variance of \Sexpr{formatC(data.frame(VarCorr(lmm_tree_r$lmer))$vcov[1], digits = 2)}.
+Compared to the tree in Figure~\ref{fig:LMM_tree_c}, an additional split was implemented based on gross motor skills (GMOTOR): children with further developed motor skills at baseline have a higher intercept.
 
@@
 The parameter stability tests evaluate whether the scores fluctuate randomly around this mean of 0, or exhibit systematic deviations when ordered by the values of a covariate available for partitioning.
-For continuous covariates $U_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each potential partitioning variable \citep{ZeilyHoth08}:
+For continuous covariates $u_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each potential partitioning variable \citep{ZeilyHoth08}:
 %
@@
 \label{eq:efp}
-W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
+W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(u_{ik})}
 \end{eqnarray}
@@
 where $\hat{J}$ is a suitable estimate of the covariance matrix of the parameter estimates, and $n_j$ gives the number of observations in the current subgroup.
-Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$.
+Further, $\hat{\psi}_{\sigma(u_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(u_{ik})$} denoting their ordering by the values of partitioning variable $u_k$.
 Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation associated with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
@@
 
-When partitioning longitudinal data, covariates will often be measured at the subject level, which should be accounted for in computing the estimated covariance matrix $\hat{J}$.
+When partitioning longitudinal data, covariates will often be measured at the subject level (i.e., time-invariant covariates), which should be accounted for in computing the estimated covariance matrix $\hat{J}$.
 In general, this computation makes use of the scores.
@@
 This resembles a GEE-type approach with an independence correlation structure.
-Our expectation is that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve accuracy of the tree.
+Our expectation is that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve subgroup recovery.
+
+In fact, the tree in Figure~\ref{fig:LMM_tree_c} was obtained by application of GLMM trees with cluster-level parameter stability tests.
+Application of GLMM trees with both cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates.
+It provides the most parsimonious tree structure of the three estimation approaches employed thus far.
+The estimated random-intercept variance was \Sexpr{formatC(data.frame(VarCorr(lmm_tree_c$lmer))$vcov[1], digits = 2)}, the highest value obtained thus far, illustrating how some variation in the response variable can be captured by either the random effects or by the tree structure.
+In the next sections, we assess the performance of the original algorithm and proposed adaptations through empirical evaluations.
 
@@
 
-\subsection{Illustration on real data}
-
-To illustrate the proposed adjustments on real data, we use a subsample of 250 children from a much larger dataset of the Early Childhood Longitudinal Study \citep[ECLS-K; ][]{NCES10}.
-Here, we partition trajectories of science ability, which were assessed in 3rd, 5th and 8th grade.
-We used 11 potential partitioning variables, including sociodemographics, motor and interpersonal skills.
-The dataset is described and analyzed in more detail in Study~III.
-With a sample of 250 children, we expect to obtain a relatively small number of relevant subgroups that can more easily be depicted and interpreted.
-
-\begin{figure}
-\caption{Trees resulting from various GLMM tree fitting approaches.}
-\begin{subfigure}{1.1\textwidth}%
-<<fig=TRUE, height=5.5, width=10, echo=FALSE>>=
-library("glmertree")
-load("sizes, N = 250, Science.Rda")
-scie_sizes <- cbind(sizes, N = 250)
-load("bag_ids, N = 250, Science.Rda")
-load("Science ability data.Rdata")
-data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
-data$CHILDID <- factor(data$CHILDID)
-data$GENDER <- factor(data$GENDER)
-data$asmtmm <- factor(data$asmtmm)
-LMM_form <- score ~ months | CHILDID | GENDER + RACE + WKSESL + C1GMOTOR + 
-  C1FMOTOR + T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE
-
-lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
-plot(lmm_tree, which = "tree", type = "simple", gp = gpar(cex = .7))
-icc <- as.data.frame(VarCorr(lmm_tree))
-
-testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
-testdat$CHILDID <- factor(testdat$CHILDID)
-testdat$GENDER <- factor(testdat$GENDER)
-testdat$asmtmm <- factor(testdat$asmtmm)
-## Remove obs with levels of race in test that are not in train
-levs <- unique(testdat$RACE) %in% unique(data$RACE)
-if (!all(levs)) {
-  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
-}
-lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
-@
-\caption{GLMM tree resulting from application of default settings.}
-\end{subfigure}%
-
-\begin{subfigure}{0.8\textwidth}%
-<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
-lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
-plot(lmm_tree_r, which = "tree", type = "simple", gp = gpar(cex = .7))
-icc_r <- as.data.frame(VarCorr(lmm_tree_r))
-lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
-@
-\caption{GLMM tree resulting from use of random-effects initialization.}
-\end{subfigure}%
-
-\begin{subfigure}{0.8\textwidth}%
-<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
-lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
-plot(lmm_tree_c, which = "tree", type = "simple", gp = gpar(cex = .7))
-icc_c <- as.data.frame(VarCorr(lmm_tree_c))
-lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
-@
-\caption{GLMM tree resulting from use of clustered covariances.}
-\end{subfigure}%
-
-\label{fig:LMM_trees}
-\end{figure}
-
-Using the the default fitting approach, the tree in the top panel of Figure~\ref{fig:LMM_trees} was obtained, comprising \Sexpr{(length(lmm_tree$tree)+1)/2} terminal nodes.
-Each node has its own fixed estimate for the intercept (science ability at baseline) and slope of time in months.
-The most important partitioning variable is a measure for socio-economic backgroud (WKSESL), with lower values yielding lower values for the intercept, indicating that children with lower SES score tend to have lower science skills both at the baseline and later assessments.
-Further, we see effects of internalizing problems (T1INTERN), gender, race and fine motor skills (C1FMOTOR), which we do not inspect in detail here as we want to focus on cahnges in results with the proposed adjustments to the estimation approach.
-The proportion of variance in science skill scores explained in science ability scores by the tree is \Sexpr{round(1 - mean((lmm_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, when evaluated on observations not part of the training sample.
-%The intra-class correlation for the random intercept term is \Sexpr{icc$vcov[1]/sum(icc$vcov)}.
-
-The middle panel of Figure~\ref{fig:LMM_trees} depicts the GLMM tree fitted on the science trajectories, by initializing estimation with the random effects.
-A slightly different set of partitioning variables were selected and the tree has less terminal nodes.
-The percentage of variance in science skill scores explained by the tree is \Sexpr{round(1 - mean((lmm_r_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, slightly lower than for the GLMM tree fitted with default settings.
-%The intra-class correlation for the random intercept term is \Sexpr{round(icc_r$vcov[1]/sum(icc_r$vcov), digits = 2)}, slightly higher than for the previous GLMM tree.
-The bottom panel of Figure~\ref{fig:LMM_trees} depicts the GLMM tree fitted on the science trajectories, estimated using clustered covariances.
-The resulting tree is the smallest tree obtained, and selected SES and internalizing problems for partitioning.
-With a proportion of variance explained of \Sexpr{round(1 - mean((lmm_c_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, providing highest predictive accuracy of the three fitting approaches.
-%The intra-class correlation for the random intercept term is \Sexpr{round(icc_c$vcov[1]/sum(icc_c$vcov), digits = 3)}, higher than obtained with the previous two GLMM trees.
-
-\section{Study~1}
+\FloatBarrier
+\section{Study~I: Assessment of Subgroup Recovery}
 
@@
 set.seed(42)
-U1 <- sample(0:1, size = 1250, replace = TRUE) 
-U2 <- round(rnorm(1250))
-U3 <- round(rnorm(1250))
+u1 <- sample(0:1, size = 1250, replace = TRUE) 
+u2 <- round(rnorm(1250))
+u3 <- round(rnorm(1250))
 time <- rep(0:4, each = 250)
@@
 ## intercept = -1, slope = -1
-node3 <- as.numeric(U1 == 0 & U2 <= 0)
+node3 <- as.numeric(u1 == 0 & u2 <= 0)
 ## left-middle node: x1 == 0 & x2 > 0
 ## intercept = -1, slope = 0
-node4 <- as.numeric(U1 == 0 & U2 > 0)
+node4 <- as.numeric(u1 == 0 & u2 > 0)
 ## right-middle node: x1 == 1 & x3 =< 0
 ## intercept = 1, slope = 0
-node6 <- as.numeric(U1 == 1 & U3 <= 0)
+node6 <- as.numeric(u1 == 1 & u3 <= 0)
 ## right-most node: x1 == 1 & x3 > 0
 ## intercept = 1, slope = 1
-node7 <- as.numeric(U1 == 1 & U3 > 0)
+node7 <- as.numeric(u1 == 1 & u3 > 0)
 y <- -(node3 + node3*time) - node4 + node6 + (node7 + node7*time) 
-data <- data.frame(U1 = as.factor(U1), U2, U3, time, y)   
-tree <- lmtree(y ~ time | U1 + U2 + U3, data = data, maxdepth = 3)
+data <- data.frame(u1 = as.factor(u1), u2, u3, time, y)   
+tree <- lmtree(y ~ time | u1 + u2 + u3, data = data, maxdepth = 3)
+
 fig <- party(
@@
 
-\begin{figure}[t!]
+\begin{figure}[tb]
 \caption{Design of the subgroup-specific fixed effects in the simulation.}
@@
 plot(tree, tnex = 3.5, terminal_panel = combine_panels(
-    panel1 = node_bivplot(tree, xlab = "time", pointcol = "white",
-                        mainlab = function(id, nobs) paste("node", id)),
-    panel2 = node_terminal(fig, FUN = identity, height = 4, width = 12, id = FALSE),
+    panel1 = glmertree:::node_growthplot(tree, xaxis.at = c(0, 4), xscal = c(0, 4), ylines = 2),
+    panel2 = node_terminal(fig, FUN = identity, height = 4, width = 11, id = FALSE, fill = "white",
+                           align = "right", just = "top", top = 0.85, gp = gpar(lty = 0, fill = "white")),
     party2 = fig), 
@@
 Every dataset comprised four non-overlapping subgroups, corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree}.
+The subgroups are defined by the three true partitioning variables: $u_1$, $u_2$ and $u_3$.
+All partitioning variables were generated from a standard normal distribution with $\mu = 0$ and $\sigma^2 = 25$.
+To allow for assessing possible selection bias toward partitioning variables with a larger number of possible cutpoints, variable $u_1$ was transformed to a binary factor, with values below the mean set to 0 and values above the mean set to 1.
 The response was computed as:
@@
 where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part.
-The bottom panels of Figure~\ref{fig:design_tree} present the subgroup-specific values of the fixed intercept and slope ($\beta_{j0}$ and $\beta_{j1}$, respectively).
+These $\beta_j$ values are reported below the terminal node panels in Figure~\ref{fig:design_tree}.
 The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints.
-The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$.\footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees in the next section.} Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next.
-Values of $\epsilon_{i}$ were independently generated from a normal distribution with a mean of zero and a variance of 5.
-
-We varied the following data-generating characteristics:
+The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$\footnote{Although GLMM trees are not restricted to have the same set of timepoints for each subject, this design allowed for comparison with SEM trees in Study~II.}.
+Values of $b_i$ (random intercepts and slopes) were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described below.
+Values of $\epsilon_{i}$ were independently generated from a normal distribution with a mean of 0 and variance of 5.
+
+We varied the following five data-generating characteristics:
 %
@@
 \item Number of subjects: small ($N = 100$) or large ($N = 250$).
-\item Variance of the random intercept: small ($\sigma_{b_0}^2 = 1$) or large ($\sigma_{b_0}^2 = 2$).
+\item Variance of the random intercept: small ($\sigma_{b_0}^2 = 1$) or large ($\sigma_{b_0}^2 = 4$).
 \item Variance of the random slope: small ($\sigma_{b_1}^2 = 0.1$) or large ($\sigma_{b_1}^2 = 0.4$).
@@
 %
-We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We performed all data generation and analysis in R \citep[version 4.1.2; ][]{R22}.
+A full factorial design was employed, yielding $2^5 = 32$ cells of the design; 100 repetitions were performed per cell.
+All data generation and analysis was performed in R \citep[version 4.1.2; ][]{R22}.
 
@@
 
-We applied ten variations LM(M) trees to every generated dataset, where each variation combines one of the following three \emph{random-effects specifications} (none/intercepts/intercepts+slopes) with the \emph{random-effect initializations} (if any; all zero vs.\ full sample estimates) and \emph{covariance specifications in the parameter instability tests} (classic vs.\ clustered), as discussed in the previous section.
+We applied ten different fitting approaches to every generated dataset.
+Each variation combines one of three \emph{random-effects specifications} (none vs.
+intercepts vs.
+intercepts+slopes) with one of two \emph{random-effect initializations} (if any; all zero vs.\ full sample estimates) and one of two \emph{covariance specifications in the parameter instability tests} (classic vs.\ clustered):
 %
@@
   \item None: $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$.
-Random effects and thus their variance were fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications:
+Random effects are not estimated and their variances thus fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications:
   \begin{itemize}
@@
   \begin{itemize}
-    \item Default: Classic observation-level covariances in the parameter stability tests and random-effect initialization with all zeros (0.). 
+    \item Default: Classic observation-level covariances in the parameter stability tests and random-effect initialization with all zeros (original step 0.). 
     \item Alternative: Clustered covariances in the parameter stability tests.
-    \item Alternative: Random-effect initialization with the full-sample estimates (0$'$.). 
+    \item Alternative: Random-effect initialization with the full-sample estimates (alternative step 0$'$.). 
     \item Alternative: Clustered covariances and random-effect initialization with the full-sample estimates.
@@
 To compute clustered covariances, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}.
-We employed outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.
+We employed the outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.
 
@@
 
-We evaluated tree accuracy by counting the number of splits in each tree, and computing mean absolute deviation (MAD) from the true tree size (3 splits).
-Trees with $> 3$ splits are indicative of Type-I error, while trees with $< 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables).
-We also assessed the variable selected for the first split in every tree.
+We evaluated tree accuracy by counting the number of splits in each tree, and computing the standard (SD) and mean absolute deviation (MAD) from the true tree size (3 splits).
+Trees with $> 3$ splits are indicative of Type-I error, while trees with $< 3$ splits are indicative of Type-II errors (i.e., power too low to detect the true partitioning variables).
+We also assessed whether the variable selected for the first split in every tree was the true first splitting variable.
 
@@
                alternative = "two.sided", p.adjust = "bonferroni")
-##          lmer    lmer_c  lmer_r  lmer_cr lmer_s  lmer_s_c lmer_s_r lmer_s_cr lm
-## lmer_c    < 2e-16 -       -       -       -       -        -        -         -      
-## lmer_r    < 2e-16 1.0000  -       -       -       -        -        -         -      
-## lmer_cr   < 2e-16 < 2e-16 < 2e-16 -       -       -        -        -         -      
-## lmer_s    0.0773  < 2e-16 < 2e-16 < 2e-16 -       -        -        -         -      
-## lmer_s_c  < 2e-16 1.0000  0.0472  < 2e-16 < 2e-16 -        -        -         -      
-## lmer_s_r  < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16 < 2e-16  -        -         -      
-## lmer_s_cr < 2e-16 < 2e-16 < 2e-16 0.0027  < 2e-16 < 2e-16  < 2e-16  -         -      
-## lm        1.0000  < 2e-16 < 2e-16 < 2e-16 1.0000  < 2e-16  < 2e-16  < 2e-16   -      
-## lm_c      < 2e-16 2.4e-10 < 2e-16 < 2e-16 < 2e-16 4.6e-07  < 2e-16  < 2e-16   < 2e-16
 @
@@
                                       "lmer_s_cr", "lmer_s_r"))
-breaks <- c(0, 1, 3, 7, 20)
+breaks <- c(0, 1, 3, 7, 20, 38, 50, 64)
+labels <- c("0", "1", "3", "7", "20", "MAD", "SD", "M")
 lim <- c(0, 60)
@@
   scale_y_continuous(trans = pseudo_log_trans(sigma = 1, base = exp(1)), 
-                     breaks = breaks, lim = lim) +
+                     breaks = breaks, labels = labels, lim = lim) +
   geom_text(data = summ_stats, aes(x=setting, y=y, label = all),
@@
              labeller = label_parsed)  +
-  coord_cartesian(clip = 'off') +
-  labs(tag = "  M\n SD\nMAD") + theme(plot.tag.position = c(.04, .85),
-                                   plot.tag = element_text(size = 10))
+  coord_cartesian(clip = 'off') 
 plot0
@@
 
-{\textit{Note.} Gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
+{\textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= Mean absolute deviation from true tree size.
+Gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
 Panel labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.
@@
 cl.cov.~= cluster-level covariances employed in parameter stability tests; ran.eff.~= estimation initialized with the random effects on the full sample.
-MAD = Mean absolute deviation from true tree size.}
+}
 \label{fig:LMM_sizes}
@@
 Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach.
-Irrespective of random-effects specification, the default fitting approaches (left-most box in every panel) tended to overfit, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average.
-This overfitting was successfully mitigated through use of cluster-level covariances in the parameter stability tests, again irrespective of the random-effects specification.
-The average number of splits in LMM trees employing clustered covariances was closest to the true number of splits.
-In terms of MAD, however, the LMM tree initializing estimation with the random effects, when only a random intercept was freely estimated, performed best.
-Thus, initializing estimation with the random effects can successfully mitigate overfitting observed for the default LMM trees, but this effect depends strongly on the random-effects specification.
-With estimation of both a random intercept and slope, random-effects initialization may lead to subgroup differences being captured by the random effects instead of the tree structure.
-
-Irrespective of the random-effects specification, combining the use random-effects initialization and cluster-level covariances was not very effective.
-Use of clustered covariances seems to provide the most reliably improvement in performance.
-
-Distributions of the number of splits, separated according to the levels of the data-generating design are omitted here because this revealed a pattern very similar to Figure~\ref{fig:LMM_sizes}.
-Thus, the optimal fitting approach did not seem dependent on data characteristics.
-
-These results are presented Figure~\ref{fig:LMM_sizes_interact} and discussed in Appendix~\ref{sec:AppendixA}.
-Of the data-generating parameters, $\sigma_{b_1}$ and $N$ showed the strongest effects, with higher values resulting in a higher number of splits.
+The default fitting approach tends to overfit, irrespective of the random-effects specification, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average.
+The use of cluster-level covariances in the parameter stability tests successfully mitigated overfitting, irrespective of the random-effects specification.
+LMM trees with clustered covariances implemented an average number of splits that was closest to the true tree size, with little difference between random-effects specifications (i.e., with and without random slopes).
+
+In terms of MAD, however, the LMM tree initializing estimation with the random effects performed best, if only random intercepts (not random slopes) were estimated.
+Likely, initializing estimation with the random effects may be successful only when the random-effects specification is kept relatively simple.
+With a more complex random-effects specification (i.e., both random intercept and slope estimated; right panel in Figure~\ref{fig:LMM_sizes}, initializing estimation with the random effects resulted in underfitting.
+Subgroup differences are then more likely to be captured by the random effects than by the tree structure.
+
+The combined use of random-effects initialization and cluster-level covariances was not very effective, irrespective of the random-effects specification.
+All in all, use of clustered covariances appears to provide the most reliable improvement in performance.
+
+Distributions of the number of splits, separated according to the levels of the data-generating design are presented in Figure~\ref{fig:LMM_sizes_interact} and discussed in Appendix~\ref{sec:AppendixA}.
+The results show a pattern very similar to Figure~\ref{fig:LMM_sizes}; no substantial interactions between data-generating and model-specification parameters were observed.
+Main effects of the data-generating parameters were as expected: The strongest effects were for $\sigma_{b_1}$ and $N$, with higher values resulting in a higher number of splits.
 The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, and were involved in minor interactions: With the default fitting approach, they had a positive effect on the number of implemented splits; with clustered covariance or random-effects initialization, they had a negative effect on the number of splits.
@@
 
-Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split: Only very rarely is $U_1$ not selected for the first split.
+Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split: Only very rarely is $u_1$ not selected for the first split.
 
@@
 splits <- cbind(rows, splits)
-colnames(splits) <- c("Algorithm", "Random effects", "Fitting approach", "$U_1$", 
-                      "$U_2$", "$U_3$", "$U_4$--$U_{25}$", "No split")
+colnames(splits) <- c("Algorithm", "Random effects", "Fitting approach", "$u_1$", 
+                      "$u_2$", "$u_3$", "$u_4$--$u_{25}$", "No split")
 
@@
         caption = "Variables selected for the first split by each LM tree (top two rows) and LMM tree (bottom eight rows) estimation approach.", row.names = FALSE, digits = 3L), row = c(2, 6), hline_after =  TRUE),
-  "\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3).
+  "\\footnotesize \\\\ \\textit{Note.} $u_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $u_2$ and $u_3$ being true splitting variables (nodes 2 and 3).
 $\\hat{\\sigma}_{b_0}$ and $\\hat{\\sigma}_{b_1}$ are the estimated standard deviations of the random intercept and slope, respectively.",
@@
 
-\section{Study~II}
-
-Next, we compared performance of LM(M) trees with that of SEM trees and LongCART.
-This allows us to evaluate the possible (dis)advantages of global versus local estimation of random-effects parameters, as well as the performance of the different splitting criteria employed by each method.
+\section{Study~II: Comparison with Other Partitioning Methods}
+
+Next, we compared the performance of LM(M) trees with that of SEM trees and LongCART.
+This allowed for evaluating the possible (dis)advantages of global versus local estimation of random-effects parameters, as well as the performance of the different splitting criteria employed by each method.
 The same data-generating design as in Simulation Study~I was employed.
@@
 
-\item We used the default ``naive'' splitting approach which employs likelihood ratio tests (LRTs) as the splitting criterion \citep{BranyOert13}.
-That is, the likelihood of the SEM fitted to the observations in the current node is compared against the likelihood of a range of two-group SEMs, in which the two groups are defined by each possible candidate split.
-An LRT can thus be computed for each candidate split, which quantifies the improvement in fit that may be obtained by implementing the split.
-In each step, the candidate yielding the highest LRT is selected for splitting, and splitting is continued as long as a candidate split yields a $p$-value of the LRT above a pre-specified $\alpha$ level (0.05, by default).
-
-\item We used the score-based splitting approach of \cite{ArnoyVoel21}.
+\item The default ``naive'' splitting approach which employs likelihood ratio tests (LRTs) as the splitting criterion \citep{BranyOert13}.
+That is, for each candidate split, the log-likelihood of the SEM fitted to the observations in the current node is compared against the sum of the log-likelihoods of a two-group SEM, in which the two groups are defined by the candidate split.
+An LRT can thus be computed for each candidate split, which quantifies the improvement in fit that would result from implementing this split.
+In each step, the candidate split  yielding the highest LRT is selected for splitting, and splitting is continued as long as a candidate split yields a $p$-value of the LRT above a pre-specified $\alpha$ level (0.05, by default).
+
+\item The score-based splitting approach of \cite{ArnoyVoel21}.
 This approach uses the MOB algorithm described in the Introduction, where the parametric model fitted in step (a) is a SEM.
@@
 
-For each splitting approach, three different random-effects specifications were employed: Variances of latent intercepts and slopes both fixed to 0; variance of the latent intercept freely estimated and variance of the latent slope fixed to 0; variances of latent intercepts and slopes both freely estimated.
+For each splitting criterion, three different random-effects specifications were employed:
+%
+\begin{itemize}
+  \item None: $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$. Random effects were not estimated and their variances were thus fixed to 0.
+
+  \item Intercepts: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$. In every node, the variance of the random intercept was freely estimated; the variance of the random slope was fixed to 0.
+
+  \item Intercepts and slopes: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$. In every node, the variances of the random intercept and slope, as well as their correlation, were freely estimated. 
+\end{itemize}
+%
 To specify the node-specific models for SEM trees, we employed an LGCM specification with the response at each timepoint regressed on a latent intercept and slope.
@@
 
-\begin{figure}[!b]
+\begin{figure}[tb]
 \caption{Tree size distributions for LM(M) trees with clustered covariances, SEM trees and LongCART.}
@@
 
-breaks <- c(0, 1, 3, 7, 15)
+breaks <- c(0, 1, 3, 7, 15, 20, 25.5, 32)
+labels <- c("0", "1", "3", "7", "15", "MAD", "SD", "M")
 lim <- c(0, 30)
@@
   scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
+                     breaks = breaks, lim = lim, labels = labels) +
   facet_grid(~ method, scales = "free", space = "free") +
@@
   scale_x_discrete(labels = scales::parse_format()) +
-  coord_cartesian(clip = 'off') +
-  labs(tag = "  M\n SD\nMAD") + theme(plot.tag.position = c(.03, .87),
-                                   plot.tag = element_text(size = 10))
+  coord_cartesian(clip = 'off')
 plot1
@@
 {
-\footnotesize \textit{Note.} Gray circles represent counts, dark gray horizontal lines indicate true number of splits~3.
-Distances on the $y$-axis are on the log scale.
-$x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.
-MAD = Mean absolute deviation from true tree size.}
+\footnotesize \textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= Mean absolute deviation from true tree size.
+Grey circles represent counts, dark gray horizontal lines indicate true number of splits~3.
+Distances on the $y$-axis are on the log scale; $x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.}
 \label{fig:tree_sizes}
@@
 Figure~\ref{fig:tree_sizes} depicts tree size distributions for the different algorithms.
-Comparing the performance of LM(M) trees (leftmost panel) with the performance of SEM trees (middle two panels) indicates that performance of SEM trees is more strongly and systematically affected by the complexity of the random-effects structure than LM(M) trees, with more complex structure yielding lower tree size.
-When random intercepts and slopes are fixed to zero, both LRT- and score based SEM trees tend to implement too many splits.
-With the correct random-effects specification (both random intercepts and slopes freely estimated), LRT-based SEM trees recover the correct tree size on average, but with higher variance than LMM trees.
-Score-based SEM trees perform best when only random intercepts are estimated, implementing too few splits when random slopes are estimated, too.
-LongCART trees appear underpowered, implementing two splits on average.
-
-On average, LMM trees estimating a random intercept and/or slope implement an average number of splits very close to the true number of splits; the same goes for LRT-based SEM trees estimating both random intercept and slope, and score-based SEM trees estimating a random intercept only.
-According to the mean absolute deviation from the true number of splits (MAD), however, LM trees perform best, closely followed by LMM trees estimating both random slope and intercept, LRT-based SEM trees estimating a random intercept only, LMM trees estimating only a random intercept and score-based SEM trees estimating only a random intercept.
+SEM trees (middle two panels) performed well, exhibiting overfitting only when no random effects were specified.
+LRT-based SEM trees tended to implement more splits than score-based SEM trees.
+Both LRT- and score-based SEM trees yielded lower tree size with increasing complexity of the random-effect specification.
+With the correct random-effects specification (i.e., both random intercept and slope were estimated), LRT-based SEM trees showed average tree size very close to the true tree size, while under-specification (i.e., specification of only a random intercept) increased tree size by 0.2 splits on average.
+Score-based SEM trees perform best when only random intercepts are estimated, implementing too few splits when random slopes were (correctly) specified.
+LongCART trees seem to suffer from a lack of power, implementing only two splits on average.
+Overall, LM(M) trees with clustered covariances showed best performance, but are very closely followed by LRT-based SEM trees, which seem to be affected more strongly by mis-specification of the random effects.
 
@@
 
-Distributions of the number of splits, separated according to the levels of the data-generating parameters are depicted and discussed in Appendix~\ref{sec:AppendixA} and Figure~\ref{fig:tree_sizes_interact}, because they show a pattern very similar to Figure~\ref{fig:tree_sizes}.
-Of the data-generating parameters, $N$ and $\sigma_{b_0}^2$ showed the strongest effects and in the expected direction with higher values resulting in a higher number of splits.
-The number of splits implemented by SEM trees was most strongly affected by the data-generating parameters when the random effects were misspecified (i.e., random intercepts and/or slope fixed to 0).
+Distributions of the number of splits, separated according to the levels of the data-generating parameters are depicted and discussed in Appendix~\ref{sec:AppendixA} and Figure~\ref{fig:tree_sizes_interact}.
+They are omitted here, as they show a pattern very similar to Figure~\ref{fig:tree_sizes}.
+Of the four data-generating parameters, $N$ and $\sigma_{b_0}^2$ showed the strongest effects, with higher values resulting in a higher number of splits, as expected.
+The number of splits implemented by SEM trees was most strongly affected by the data-generating parameters when the random effects were mis-specified (i.e., random intercepts and/or slope fixed to 0).
 
@@
 @
+
+\FloatBarrier
 
@@
         caption = "Cross-validated mean squared errors for each of the response variables."),
-  "\\footnotesize \\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3).
+  "\\footnotesize \\\\ \\textit{Note.} $u_1$ is the true first splitting variable and is a binary factor; all other partitioning variables are continuous, with $u_2$ and $u_3$ being true splitting variables (nodes 2 and 3).
 The first column indicates wheter the random intercept and/or slope were estimated or not.",
@@
 Table~\ref{tab:first_splits2} presents variable selection frequencies for the first split in the fitted trees.
-For SEM trees, the LRT criterion yields almost perfect accuracy for the first split as does the score-based criterion provided the random effects were correctly specified.
-Only in case of misspecification of the random-effects structure, the score-based SEM trees selected the wrong variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$U_1$`[grep("s_sem", rownames(splits))])))}\% of datasets.
-Closer inspection of stability tests seperately per model parameter suggested that the the score-based tests for SEMs are more sensitive to instability in the fixed slope than in the fixed intercept.
+For SEM trees, the LRT criterion yields almost perfect accuracy for the first split.
+The score-based criterion for SEM trees provides near-perfect accuracy when random effects were correctly specified.
+When random effects were mis-specified, score-based SEM trees selected the wrong variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$u_1$`[grep("s_sem", rownames(splits))])))}\% of datasets.
+Closer inspection of stability tests for individual models and parameters suggested that the the score-based tests for SEMs are more sensitive to instability in the fixed slope than in the fixed intercept, explaining why $u_2$ or $u_3$ were often selected for the first split.
 
 LongCART trees exhibit low accuracy for recovering the first split, selecting the wrong variable in all datasets where at least one split was implemented.
-LongCART showed a strong tendency to select $U_2$ or $U_3$ for the first split.
-It is noteworthy that in the datasets where LongCART did not implement any splits, in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets $U_1$ was the strongest splitting candidate, but the parameter stability test simply did not reach the criterion for significance.
-This suggests the parameter stability tests proposed by \cite{KundyHare19} are less sensitive to instabilities with respect to the fixed intercept, or less sensitive to instability with respect to a categorical covariate.
+LongCART showed a strong tendency to select $u_2$ or $u_3$ for the first split.
+Closer inspection of the fitted LongCART trees revealed that in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets where no splits were implemented, $u_1$ was the strongest splitting candidate, but the parameter stability tests did not reach significance.
+This suggests the tests proposed by \cite{KundyHare19} are less sensitive to instability of the intercept (compared to instability of the slope), or less sensitive to instability with respect to categorical covariates (compared to instability with respect to continuous covariates).
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!ht]%
 \caption{Computation time distributions for the different partitioning methods.}
-\begin{subfigure}{.8\textwidth}
+\begin{subfigure}{.8\textwidth}%
 <<fig=TRUE, echo=FALSE, width=4, height=3.25>>=
@@
                      rep("SEM tree\n(score)", times = 3),
-                     "LongCART")
+                     "LongCRT")
 times <- reshape2::melt(times)
@@
                          y = 100000)
+
 ggplot(times, aes(Var2, value)) + 
-  geom_violin(, alpha = .5, fill = "gray", width=.5) +
+  geom_boxplot(, alpha = .5, fill = "gray", width=.5) +
   labs(x = "", y = "Computation time (in seconds)") +
   geom_text(data = summ_stats, aes(x=Var2, y=y, label = M),
-            size = 3, col = "black") +
+            size = 3, col = "gray44") +
   scale_y_continuous(trans = "log", limits = c(0.01, 100000), 
-                     breaks = c(.01, .1, 1, 10, 100, 1000, 10000)) +
-  labs(tag = "M") + theme(plot.tag.position = c(.12, .96),
-                          plot.tag = element_text(size = 10)) + 
-  geom_text(data = summ_stats, aes(x=Var2, y=y, label = M),
-            size = 3, col = "darkgray") 
-@
-\end{subfigure}
+                     breaks = c(.01, .1, 1, 10, 100, 1000, 10000, 100000),
+                     labels = c("1e-02", "1e-01", "1e+00", "1e+01", "1e+02", "1e+03", "1e+04", "M"))
+@
+\end{subfigure}%
+\\ \textit{Note.} $y$-axis is on the log scale; M = mean computation time in seconds. 
 \label{fig:comp_times}
-\end{figure}
+\end{figure}%
 
@@
 
-\section{Study~III}
+\section{Study~III: Cross-validated predictive accuracy on real data}
 
@@
 
-We analyzed longitudinal assessments of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}.
+We analyzed trajectories on children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}.
 Data were collected from 21,304 children from 1,018 schools across the USA.
@@
 Time was measured as the number of months since the baseline assessment.
-In order to obtain approximately linear trajectories, we chose metrics of time based on visual inspection of the observed data: $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.
+In order to obtain approximately linear trajectories, we chose the timing metric based on visual inspection of the observed data: $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.
 
@@
 
-We fitted four LM(M) trees using the approaches which showed most promise in the simulations: An LM tree ($\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$) using clustered covariances in the parameter-stability tests, two LMM trees using clustered covariances with different random-effects specifications ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$ and $\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$), and an LMM tree initializing estimation with the random effects and freely estimating a random intercept only ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$).
+We fitted four LM(M) trees using the approaches which showed most promise in the simulations: 
+% 
+\begin{itemize}
+  \item An LM tree ($\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$) using clustered covariances in the parameter-stability tests.
+  \item An LMM tree using clustered covariances, with a random intercept freely esimtated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$).
+  \item An LMM tree using clustered covariances, with both random intercept and slope freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$). 
+  \item An LMM tree using random-effect initialization with the full-sample estimate, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$). 
+\end{itemize}
+%
+As a benchmark, we also estimated two standard LMMs with a fixed effect of time: One LMM with a random intercept freely estimated, one LMM with both random intercept and slope freely estimated.
+The LMMs did not include additional predictors in the model, only timepoint and the child identifier.
 
@@
 For each combination of response variable (math, reading, or science) and sample size, we performed 100 repetitions.
-We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample of the current repetition.
-Note that this measures the accuracy of the fixed-effects predictions only.
+We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample in the current repetition.
+This separation of train and test observations does not allow for using the random effects in computing predictions, these were thus assumed 0; only accuracy of the fixed-effects parameters is thus qualified by the MSEs.
 Tree size was measured by the number of splits in each tree.
 
+\FloatBarrier
 \subsection{Results}
 
+Figure~\ref{fig:application_MSEs} presents MSE distributions for each of the estimation approaches and Table~\ref{tab:application_MSEs} presents the means and standard deviations.
+%As expected, all methods show higher accuracy with larger sample sizes.
+LM(M) trees tend to show better predictive accuracy than longRPart trees.
+Pairwise $t$-tests indicated significantly better performance for each of the LM(M) tree approaches employing clustered covariances, compared to both longRPart approaches (Bonferroni corrected $p$-values all $< .01$).
+However, these performance differences are small; the effects of sample size are much more obvious than the differences between methods in Figure~\ref{fig:application_MSEs}.
+Performance of the benchmark LMMs indicates that a substantial amount of variance can be explained by the single fixed linear effect of time.
+Further, all LM(M) trees outperform the benchmark LMMs.
+Differences in performance between LM(M) trees are small.
+Table~\ref{tab:application_MSEs} and Figure~\ref{fig:application_MSEs} indicate that LMM trees employing cluster-level covariances performed best for predicting math and reading abilities.
+Trees estimating only a random intercept performed best, but performance differed only about 1 SE compared to trees estimating both random intercept and slope.
+
+For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best, while this was the worst-performing approach for math and reading.
+LM trees with clustered covariances never performed best, illustrating the benefit of estimating random effects.
+
+%For longRPart, trees comprising both a random intercept and slope performed best for predicting math and reading, trees with a random intercept only performed best for predicting science.
+The differences in performance between math and reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of science abilities.
+
+%Again, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with random intercepts, slopes or neither.
+Pairwise $t$-tests did not indicate significant differences between those three approaches for any of the response variables.
+
 \begin{figure}[!bt]
-\caption{Mean squared errors for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
+\caption{Mean squared errors for trees fitted to math, reading and science ability trajectories.}
 \begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+<<echo=FALSE, fig=TRUE, height=2, width=5>>=
 ## Load datasets for computing variance of response variables
@@
 
-## Compute summary stats
-# MSE_df <- data.frame(t(sapply(math_MSEs[ , -c(4,6,7,10)], 
-#                               function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE)))))
-# MSE_df <- cbind(MSE_df, t(sapply(math_MSEs[ , -c(4,6,7,10)], 
-#                                function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE)))))
-# MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
-#                                function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE)))))
-# MSE_df <- cbind(MSE_df, t(sapply(read_MSEs[ , -c(4,6,7,10)], 
-#                                function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE)))))
-# MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
-#                                function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE)))))
-# MSE_df <- cbind(MSE_df, t(sapply(scie_MSEs[ , -c(4,6,7,10)], 
-#                                function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE)))))
-# summ_stats <- MSE_df[ , c(1:2,5:6,9:10)]
-# names(summ_stats) <- c("math, N = 250", "math, N = 1,000", 
-#                        "read, N = 250", "read, N = 1,000", 
-#                        "scie, N = 250", "scie, N = 1,000")
-
+## Math
 MSE_df <- data.frame(sapply(math_MSEs[ , -c(4,6,7,10)], 
@@
                                function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE))))
+MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, 
+                                                                  function(x) 1 - mean(x, na.rm = TRUE)/var_math))))
+## Reading
 MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , -c(4,6,7,10)], 
@@
                                function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE))))
+MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, 
+                                                                  function(x) 1 - mean(x, na.rm = TRUE)/var_read))))
+
+## Science 
 MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , -c(4,6,7,10)], 
@@
                                function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE))))
-summ_stats <- MSE_df[ , c(1:2,5:6)]
-names(summ_stats) <- c("math, N = 250", "read, N = 250", "scie, N = 250")
+MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , -c(4,6,7,10)], 
+                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, 
+                                                                  function(x) 1 - mean(x, na.rm = TRUE)/var_scie))))
 
@@
 
+## tmp3 is math
 tmp3 <- data.frame(stack(math_MSEs[ , -c(4,6,7:10)]),
@@
 tmp3$N <- factor(tmp3$N)
-
 theme_set(theme_bw(base_size = 12))
 ggplot(tmp3) +
-  geom_violin(aes(x=random, y=MSE), 
-               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
-  scale_y_continuous(trans = log_trans(),
+  geom_boxplot(aes(x=random, y=MSE), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
+  scale_y_continuous(#trans = log_trans(),
                      sec.axis = sec_axis(trans=~1-(./var_math))) +
@@
   theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
-        legend.position="none") +
+        legend.position="none", axis.text.y = element_text(size = 8)) +
   labs(x = "", y = "MSE")
@@
 \begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.6, width=8>>=
+<<echo=FALSE, fig=TRUE, height=1.6, width=5>>=
+## tmp4 is reading
 tmp4 <- data.frame(stack(read_MSEs[ , -c(4,6,7:10)]),
@@
 ggplot(tmp4) +
-  geom_violin(aes(x=random, y=MSE), 
-               position=position_dodge(1), alpha = .5, width = .6, fill = "gray" ) +
-  scale_y_continuous(trans = log_trans(),
+  geom_boxplot(aes(x=random, y=MSE), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
+  scale_y_continuous(#trans = log_trans(),
                        sec.axis = sec_axis(trans=~1-(./var_read))) +
@@
         axis.title.x=element_blank(), axis.text.x=element_blank(),
-        legend.position="none") +
+        legend.position="none", axis.text.y = element_text(size = 8)) +
   labs(x = "", y = "MSE")
@@
 \begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+<<echo=FALSE, fig=TRUE, height=2, width=5>>=
+## tmp5 is science
 tmp5 <- data.frame(stack(scie_MSEs[ , -c(4,6,7:10)]),
@@
 ggplot(tmp5, aes(x=random, y=MSE)) +
-  geom_violin(position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
-  scale_y_continuous(trans = log_trans(),
+  geom_boxplot(position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
+  scale_y_continuous(#trans = log_trans(),
                        sec.axis = sec_axis(trans=~1-(./var_scie))) +
@@
   theme(strip.background = element_blank(), strip.text.x = element_blank(),
-        legend.position="none") +
+        legend.position="none", axis.text.y = element_text(size = 8)) +
   labs(x = "", y = "MSE") +
@@
 {\footnotesize \\ \textit{Note.
-}The second $y$-axis (far right) gives an estimate of the proportion of variance explained, \\computed as $1 - \frac{\text{MSE}}{\text{var}(y)}$.
-Distances on $y$-axis are on the log scale.}
+} Top, middle and bottom panels depict math, reading and science ability trajectories, respectively.
+The secondary $y$-axis on the right quantifies the proportion of variance explained, computed as $1 - \frac{\text{MSE}}{\text{var}(y)}$.}
 \label{fig:application_MSEs}
@@
 pairwise.t.test(tmp3$MSE, tmp3$method, alternative = "two.sided", 
-                p.adjust = "bonferroni")
-##             LMtree_c LMMtree_c LMMtree_r LMMtree_sc longRPart
-## LMMtree_c   1.00000  -         -         -          -        
-## LMMtree_r   0.08664  0.00033   -         -          -        
-## LMMtree_sc  1.00000  1.00000   0.00801   -          -        
-## longRPart   1.7e-13  < 2e-16   8.6e-06   5.8e-16    -        
-## longRPart_s 0.00789  1.1e-05   1.00000   0.00046    0.00029
-
+                p.adjust = "bonferroni") ## math
 pairwise.t.test(tmp4$MSE, tmp4$method, alternative = "two.sided", 
-                p.adjust = "bonferroni")
-##            LMtree_c LMMtree_c LMMtree_r LMMtree_sc longRPart
-## LMMtree_c   1.00     -         -         -          -        
-## LMMtree_r   4.0e-10  1.2e-10   -         -          -        
-## LMMtree_sc  1.00     1.00      3.7e-09   -          -        
-## longRPart   < 2e-16  < 2e-16   1.9e-07   < 2e-16    -        
-## longRPart_s 1.9e-05  7.9e-06   0.98      1.0e-04    9.4e-13
-
+                p.adjust = "bonferroni") ## reading
 pairwise.t.test(tmp5$MSE, tmp5$method, alternative = "two.sided", 
-                p.adjust = "bonferroni")
-##            LMtree_c LMMtree_c LMMtree_r LMMtree_sc longRPart
-## LMMtree_c   1.000    -         -         -          -        
-## LMMtree_r   0.089    1.000     -         -          -        
-## LMMtree_sc  1.000    1.000     1.000     -          -        
-## longRPart   <2e-16   <2e-16    <2e-16    <2e-16     -        
-## longRPart_s <2e-16   <2e-16    <2e-16    <2e-16     <2e-16
-
+                p.adjust = "bonferroni") ## science
 @
@@
 library("kableExtra")
-# MSE_df <- cbind(paste0(MSE_df[,1], " (", MSE_df[,3], ")"),
-#                 paste0(MSE_df[,2], " (", MSE_df[,4], ")"),
-#                 paste0(MSE_df[,5], " (", MSE_df[,7], ")"), 
-#                 paste0(MSE_df[,6], " (", MSE_df[,8], ")"), 
-#                 paste0(MSE_df[,9], " (", MSE_df[,11], ")"),
-#                 paste0(MSE_df[,10], " (", MSE_df[,12], ")"))
-#MSE_df <- cbind(paste0(MSE_df[,1], " (", MSE_df[,3], ")"),
-#               paste0(MSE_df[,2], " (", MSE_df[,4], ")"),
-#               paste0(MSE_df[,5], " (", MSE_df[,7], ")"), 
-#               paste0(MSE_df[,6], " (", MSE_df[,8], ")"), 
-#               paste0(MSE_df[,9], " (", MSE_df[,11], ")"),
-#               paste0(MSE_df[,10], " (", MSE_df[,12], ")"))
 rownames(MSE_df) <- c("LM tree$^c$", "LMM tree$^{i,c}$", "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$")  
                       #"longRPart$^i$", "longRPart$^{i,s}$")
-colnames(MSE_df) <- rep(c("M", "SD"), times = 3)
+colnames(MSE_df) <- rep(c("M", "SD", "$R^2$"), times = 3)
 MSE_df <- MSE_df[c(1,2,4,3), ]
+
+## Add benchmark LMMs
+load("LMM MSEs, N = 250, Math.RDa")
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 1] <- format(colMeans(MSEs), digits = 4)
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 2] <- round(sapply(MSEs, sd), digits = 3)
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 3] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_math)
+
+load("LMM MSEs, N = 250, Reading.RDa")
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 4] <- format(colMeans(MSEs), digits = 4)
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 5] <- round(sapply(MSEs, sd), digits = 3)
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 6] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_read)
+
+load("LMM MSEs, N = 250, Science.RDa")
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 7] <- format(colMeans(MSEs), digits = 4)
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 8] <- round(sapply(MSEs, sd), digits = 3)
+MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 9] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_scie)
 
@@
   kable(MSE_df, format = "latex", booktabs = TRUE, label = "application_MSEs", align = "c", 
-        escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
+        escape=FALSE, linesep = "", # linesep command suppresses addlinesep every 5 rows
         caption = "Cross-validated mean squared errors for each of the response variables."),
-  "\\footnotesize \\textit{Note.} Values represent means over 100 repetitions, parenthesized values represent SDs and can be interpreted as standard errors.
-c = cluster-level covariances; r = estimation initialized with random effects; i =  random-intercept variance freely estimated; s = random-slope variance freely estimated.",
+  "\\footnotesize \\\\ \\textit{Note.} Means and standard deviations computed over 100 repetitions.
+Standard deviations of the MSE can be interpreted as standard errors.
+$R^2$ was computed as $\\frac{MSE}{var(y)}$.
+$^c$~cluster-level covariances; $^r$~estimation initialized with random effects; $^i$~random-intercept variance freely estimated; $^s$~random-slope variance freely estimated.",
   notation="none", threeparttable = TRUE, escape = FALSE),
@@
   #  "N = 250"=1, "N = 1,000"=1), align = "c"),
-  c(" ", "Math" = 2, "Reading" = 2, "Science" = 2), align = "c"),
+  c(" ", "Math" = 3, "Reading" = 3, "Science" = 3), align = "c"),
   font_size = 11, full_width=FALSE)
@@
 <<echo=FALSE, eval=FALSE>>=
-
 ###############################
@@
 ## most for science, which was most strongly affected by the mistake.
-
-@
-
-%LongRPart trees comprising both random intercepts and slopes yielded convergence errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the science data.
-All other model-fitting attempts converged successfully.
-Results are presented for successful model estimations.
-
-Figure~\ref{fig:application_MSEs} presents MSE distributions for each of the estimation approaches and Table~\ref{tab:application_MSEs} presents means and standard deviations of the MSEs.
-%As expected, all methods show higher accuracy with larger sample sizes.
-LM(M) trees tend to show better predictive accuracy than longRPart trees.
-Pairwise $t$-tests indicated significantly better performance for each of the LM(M) tree approaches employing clustered covariances, compared to both longRPart approaches (Bonferroni corrected $p$-values all $< .01$).
-However, these performance differences are small; the effects of sample size are much more obvious than the differences between methods in Figure~\ref{fig:application_MSEs}.
-Differences in performance are small, few differences are $> 1$ standard error.
-Note that standard errors in pairwise $t$-tests get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not.
-The latter thus provide a more useful measure of effect size, while the former has higher power to detect differences between methods.
-Table~\ref{tab:application_MSEs} indicates that LMM trees estimating only a random intercept and employing cluster-level covariances performed best for predicting math and reading abilities.
-For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best, while this was the worst-performing approach for math and reading.
-
-%For longRPart, trees comprising both a random intercept and slope performed best for predicting math and reading, trees with a random intercept only performed best for predicting science.
-The differences in performance between math and reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of science abilities.
-
-%Again, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with random intercepts, slopes or neither.
-Pairwise $t$-tests did not indicate significant differences between those three approaches for any of the response variables.
+@
 
@@
 
-\begin{figure}[!bt]
-\caption{Tree sizes for trees fitted to math (top), reading (middle), and science (bottom) ability trajectories.}
-\begin{subfigure}{1.2\textwidth}
-<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+\begin{figure}[!bt]%
+\caption{Tree sizes for trees fitted to math, reading, and science ability trajectories.}
+\begin{subfigure}{\textwidth}%
+<<echo=FALSE, fig=TRUE, height=2, width=5>>=
 ## Tree size
@@
 
+## tmp6 is math
 tmp6 <- data.frame(stack(math_sizes[ , -c(4,6,7:10)]),
@@
 ggplot(tmp6) +
-  geom_violin(aes(x=random, y=splits), 
-               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
+  geom_boxplot(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
   facet_grid(~method2, scales = "free", space = "free") +
-  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
+        axis.text.y = element_text(size = 8)) +
   labs(x = "", y = "no. of splits")
@@
 \end{subfigure}
-\begin{subfigure}{1.2\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.6, width=8>>=
+\begin{subfigure}{\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.6, width=5>>=
+## tmp7 is reading
 tmp7 <- data.frame(stack(read_sizes[ , -c(4,6,7:10)]),
@@
 ggplot(tmp7) +
-  geom_violin(aes(x=random, y=splits), 
-               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
+  geom_boxplot(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
   facet_grid(~method2, scales = "free", space = "free") +
@@
   theme(strip.background = element_blank(), strip.text.x = element_blank(),
-        axis.title.x=element_blank(), axis.text.x=element_blank())
+        axis.title.x=element_blank(), axis.text.x=element_blank(),
+        axis.text.y = element_text(size = 8))
 @
 \end{subfigure}
-\begin{subfigure}{1.2\textwidth}
-<<echo=FALSE, fig=TRUE, height=2, width=8>>=
+\begin{subfigure}{\textwidth}
+<<echo=FALSE, fig=TRUE, height=2, width=5>>=
+## tmp8 is science
 tmp8 <- data.frame(stack(scie_sizes[ , -c(4,6,7:10)]),
@@
 ggplot(tmp8) +
-  geom_violin(aes(x=random, y=splits), 
-               position=position_dodge(1), alpha = .5, width = .6, fill = "gray") +
+  geom_boxplot(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
   facet_grid(~method2, scales = "free", space = "free") +
-  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  theme(strip.background = element_blank(), strip.text.x = element_blank(),
+        axis.text.y = element_text(size = 8)) +
   labs(x = "", y = "no. of splits") +
@@
 @
-\end{subfigure}
+\end{subfigure}%
+{\\\footnotesize \textit{Note.} Top, middle and bottom panels depict math, reading and science ability trajectories, respectively.}
 \label{fig:application_sizes}
-\end{figure}
+\end{figure}%
 
@@
 Figure~\ref{fig:application_sizes} presents tree size distributions.
-%Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees.
+Random-effects initialization tends to yield the largest trees, while LM trees with clustered covariances tend to yield the smallest trees.
+LMM trees with clustered covariances yield tree sizes in between.%Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees.
 For math and reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small.
@@
 For the math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
-Random-effects initialization tends to yield the largest trees, while LM trees with clustered covariances tend to yield the smallest trees.
-LMM trees with clustered covariances yields somewhat larger trees, but smaller than LMM trees with random-effects initialization.
 
@@
 
+\FloatBarrier
+
 \section{Discussion}
 
-Study~I showed that the default estimation approach for GLMM trees yields too many splits when partitioning LGCMs.
-As hypothesized, performing parameter-stability tests using clustered covariances mitigated this overfitting.
-As hypothesized, initializing estimation with the random effects mitigated overfitting, but too strongly when both random intercepts and slopes were estimated.
-The performance of LM(M) trees using clustered covariances appeared largely unaffected by (mis-)specification of the random effects, and use of clustered covariances may therefore be the safest choice in practice.
-Combining cluster-level covariances and random-effects initialization worsened performance and is thus best avoided.
+The Simulations in Study~I showed that the proposed extensions of GLMM trees are effective for partitioning linear growth curve models.
+Use of clustered covariances in the parameter stability tests mitigated the strong overfitting that was observed using the standard covariances.
+Initializing estimation with the random effects, instead of assuming these to be zero was also successfull in mitigating overfitting, but only when the random-effects specification comprised only a random intercept.
+The performance of trees fitted using clustered covariances appeared largely unaffected by (mis-)specification of the random effects; use of clustered covariances may therefore be the safest choice in practice.
+Combining cluster-level covariances and random-effects initialization worsened performance and thus not recommended.
 The results also indicated that the optimal estimation approach for LMM trees does not depend on sample size, variance of the random effect(s), the number of potential partitioning variables or the intercorrelation between partioning variables.
@@
 Study~II showed comparable performance of LM(M) and SEM trees in partitioning LGCMs.
-SEM trees did however show stronger sensitivity to misspecification of the random effects, with underspecification yielding a higher number of splits.
-Furthermore, in line with results of \cite{ArnoyVoel21}, we found score-based SEM trees to have somewhat lower power than LRT-based SEM trees, but computationally much less burdensome.
+SEM trees appeared more sensitive to mis-specification of the random effects, with under-specification yielding too many splits being implemented.
+In line with results of \cite{ArnoyVoel21}, we found score-based SEM trees to have somewhat lower power than LRT-based SEM trees, but with much lower computational burgen.
 LongCART trees often selected the wrong partitioning variable for the first split, and were outperformed by LM(M) and SEM trees.
-The LongCART parameter stability tests \citep{KundyHare19} may be underpowered for detecting systematic instability of the fixed intercept, or for detecting instability with respect to categorical covariates.
+The LongCART parameter stability tests \citep{KundyHare19} may be underpowered for detecting instability of the fixed intercept, or for detecting instability with respect to categorical covariates.
 As in Study~I, these results were not strongly affected by sample size, variance of the random effect(s), the number of potential partitioning variables or the intercorrelation between partioning variables.
 
-Study~III, which evaluated LM(M) tree performance in partitioning math, reading and science trajectories, we found only small differences in terms of predictive accuracy.
-In line with the simulation studies, we found good performance of LMM trees employing clustered covariances.
-Although GLMM trees initializing estimation with the random effects performed best on partitioning science trajectories, this slight improvement came at the cost of substantial increases in tree size.
-
-Study~II indicated a clear computational advantage for LM(M) trees.
+Study~III assessed performance of LM(M) trees in partitioning children's trajectories of math, reading and science abilities.
+Similar to the results of the simulation studies, we found very similar performance between LM(M) trees which employed clustered covariances but with different random effects specifications, though best performance was observed for a random-effects specification comprising only a random intercept.
+However, when partitioning science trajectories, for which there were a smaller number of observations per child, initializing estimation with the random effects showed the best performance.
+A possible explanation is that the higher the number of observations per subject, the more the clustered covariances will differ from covariances assuming independent observations, creating greater potential for improving the parameter stability tests.
+With a smaller number of observations per subject, initializing estimation with the random effect may be more beneficial.
+However, initializing estimation with the random effects also resuled in a substantially higher number of splits in all three datasets.
+
+As expected, Study~II revealed a clear computational advantage for LM(M) trees.
 This is in large part due to the local-global estimation approach employed by LMM trees, where fixed-effects parameters are estimated locally within a node and random-effects parameters are estimated globally, using all observations.
-In contrast, SEM trees, LongCART and longRPart fit the full mixed-effects model in each node, which substantially increases computational load.
-The local-global estimation approach also reduces model complexity and the good performance of LMM trees in the current study suggest it also benefits stability and generalizability.
+In contrast, SEM trees and LongCART fit the full mixed-effects model in each node, which substantially increases computational load.
+The local-global estimation approach also reduces model complexity, because a lower number of random-effects parameters need to be estimated.
+The good performance of LMM trees in the current study suggest the lower complexity may also yield better stability and generalizability.
 
 Yet, a possible downside of the local-global estimation approach is that it does not allow for recovering subgroups with differences in random-effects parameters.
-When there is a specific interest in partitioning the random effects, score-based SEM trees should be preferred.
+When there is a specific interest in partitioning the random effects, score-based SEM trees may be preferred.
 Alternatively, researchers may want to use the parameter stability tests for mixed-effects models developed by \cite{WangyMerk18} \citep[see also ][]{WangyMerk21,WangyGrav22}.
@@
 We expect that the good performance of clustered covariances in the parameter stability tests generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures.
-Yet, whether our conclusions generalize to settings beyond LGCMs should be assessed in future studies.
+Yet, whether our conclusions generalize to settings beyond LGCMs remains to be evaluated.
 Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances.
-Future studies may assess whether computation and use of the full sandwich estimator further improves performance of the parameter stability tests.
+Though computationally more burdensome, future studies could assess whether use of the full sandwich estimator would further improve performance.
 
@@
 
-\section{Effects of data-generating parameters on tree size}
+\section{Appendix A: Effects of data-generating parameters on tree size}
 \label{sec:AppendixA}
@@
 \vspace{-.6cm}%
-{\singlespacing \small \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
+{\singlespacing \footnotesize \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
 Distances on $y$-axis are on log scale.
@@
 \vspace{-.6cm}%
-{ \singlespacing \small \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
+{ \singlespacing \footnotesize \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
 Distances on $y$-axis are on log scale.

### 2023-08-15 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
small touch-ups, esp grey -> gray
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3269 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c72b022b6fab2678f0fbbc00f4247c7da9e7535a

@@
 For example, Figure~\ref{fig:global_trajectories} shows trajectories of science knowledge and skills from 250 children who were assessed at three timepoints throughout grades 3 through 8\footnote{Further details on the source of these data and scaling of the response variable are provided in Study~III.}.
-The grey lines depict individual trajectories, while the red line depicts the estimated average trajectory.
-Although the red line describes the average trajectory well, the grey lines indicate substantial intra-individual variability in average levels and growth over time.
+The gray lines depict individual trajectories, while the red line depicts the estimated average trajectory.
+Although the red line describes the average trajectory well, the gray lines indicate substantial intra-individual variability in average levels and growth over time.
 A common research aim might be to identify which covariates can explain or predict this heterogeneity.
@@
 @
-\\\footnotesize{\textit{Note.} Grey lines depict observed individual trajectories.
+\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories.
 The red line depicts the average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.}
@@
 \\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
-Grey lines depict observed individual trajectories.
-Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercepts with respect to individuals.}
+Gray lines depict observed individual trajectories.
+Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.}
 \end{subfigure}%
@@
 For children in the group with lower levels of SES, the intercept is also lowest, indicating lower science ability.
-For intermediate levels of SES (between -0.41 and 1.23), there is also an effect of internalizing problems, where children with higher internalizing problems have a lower intercept.
+For intermediate levels of SES (between $-0.41$ and $1.23$), there is also an effect of internalizing problems, where children with higher internalizing problems have a lower intercept.
 Highest science ability intercepts are observed in the group with highest SES.
@@
 \\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability.
-Grey lines depict observed individual trajectories.
+Gray lines depict observed individual trajectories.
 Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
@@
 \\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
-Grey lines depict observed individual trajectories.
+Gray lines depict observed individual trajectories.
 Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
@@
 \footnotesize \textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= Mean absolute deviation from true tree size.
-Grey circles represent counts, dark gray horizontal lines indicate true number of splits~3.
+Gray circles represent counts, dark gray horizontal lines indicate true number of splits~3.
 Distances on the $y$-axis are on the log scale; $x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.}
@@
 ## Add benchmark LMMs
-load("LMM MSEs, N = 250, Math.RDa")
+load("LMM MSEs, N = 250, Math.Rda")
 MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 1] <- format(colMeans(MSEs), digits = 4)
@@
 
-load("LMM MSEs, N = 250, Reading.RDa")
+load("LMM MSEs, N = 250, Reading.Rda")
 MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 4] <- format(colMeans(MSEs), digits = 4)
@@
 
-load("LMM MSEs, N = 250, Science.RDa")
+load("LMM MSEs, N = 250, Science.Rda")
 MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 7] <- format(colMeans(MSEs), digits = 4)

### 2023-09-07 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Paper updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3275 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 36a1766808faf6884af890d2c27118e77548feff

@@
 \usepackage{hyperref}
-
 \renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
@@
 
-\title{Subgroup detection in linear growth curve models with GLMM trees}
+\title{Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees}
 \shorttitle{Recursive partitioning of linear growth curve models}
@@
 The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.
-We evaluate GLMM trees' performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities.
-We compare the performance of GLMM trees with alternative methods: structural equation model (SEM) trees and LongCART.
-GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART.
-In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.\\}
+In simulated and real-world data, we assess the performance of the extensions and compare against other partitioning methods for growth curve models.
+Extended GLMM trees perform more accurately than the original algorithm and LongCART, and similarly accurate as structural equation model (SEM) trees.
+In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and are much faster to compute.\\}
 
@@
 \maketitle
+\setlength{\tabcolsep}{3pt}
 
 \section{Introduction}
-
 \label{sec:Introduction}
@@
 Mixed-effects or latent-variable models can be used to model such trajectories and allow for explaining heterogeneity with covariates of a-priori known relevance \citep[e.g., ][]{NeisyMatt18}.
-However, when the covariates that can explain heterogeneity are not known in advance, methods for identifying them are needed.
-
-For example, Figure~\ref{fig:global_trajectories} shows trajectories of science knowledge and skills from 250 children who were assessed at three timepoints throughout grades 3 through 8\footnote{Further details on the source of these data and scaling of the response variable are provided in Study~III.}.
-The gray lines depict individual trajectories, while the red line depicts the estimated average trajectory.
-Although the red line describes the average trajectory well, the gray lines indicate substantial intra-individual variability in average levels and growth over time.
-A common research aim might be to identify which covariates can explain or predict this heterogeneity.
-
-\begin{figure}%
+However, when these covariates are not known in advance, methods for identifying them are needed.
+
+As an example, trajectories of science knowledge and skills from a sample of 250 children are depicted in Figure~\ref{fig:global_trajectories}.
+The children were assessed at three timepoints across grades 3 through 8\footnote{Further details on the source of these data are provided in Study~III.}.
+The red line depicts the estimated average trajectory, while the gray lines depict individual trajectories.
+The gray lines reveal substantial variability between the children, both in initial levels and growth over time.
+An obvious research aim would be to identify covariates that can explain or predict this heterogeneity.
+
+\begin{figure}[b]%
 \caption{Growth curves of science ability.}
@@
 @
-\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories.
-The red line depicts the average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.}
+\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.
+The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories.}
 \end{subfigure}%
@@
 
+%\FloatBarrier
+\subsection{Recursive Partitioning Methods for Growth Curve Models}
+
+Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates.
+Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline.
+Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope %(presented in Table~\ref{tab:local_coefs}).
+Groups of children with lower (higher) SES also have a lower (higher) intercept, indicating lower (higher) average science ability.
+For children with lower SES (node~2), there is also an effect of gross motor skills: The group with lower motor skills has a lower intercept.
+For children with intermediate levels of SES (node~6), there is also an effect of internalizing problems: The group with lower internalizing problems has a higher intercept.
+Also, the two groups (or nodes) with higher intercepts have higher slopes, indicating that children with relatively higher ability may also show larger gains in ability over time.
+
 \begin{figure}%
-\caption{Partitioned growth curves of science abilities.}
-\begin{subfigure}{0.8\textwidth}%
-<<fig=TRUE, height=9, width=12, echo=FALSE>>=
+\caption{Partitioned growth curves of science ability.}
+\begin{subfigure}{0.9\textwidth}%
+<<fig=TRUE, height=9, width=13.5, echo=FALSE>>=
 LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
@@
 }
+
+## Fit models
 lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
 lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
-plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
-     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
-                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
-                                        xaxis.labs = c(0, 12, 36, 60)),
-     gp = gpar(cex = 1.5))
+lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
+icc_r <- as.data.frame(VarCorr(lmm_tree_r))
+lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
 icc_c <- as.data.frame(VarCorr(lmm_tree_c))
@@
 lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)
-@
+
+## Plot tree
+plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
+     nodesize_level = "CHILDID", tp_args = 
+       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
+            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
+@
+\label{subfig:tree}
 \\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
@@
 \end{subfigure}%
-\label{fig:LMM_tree_c}
-\end{figure}%
-
+~
+\begin{subfigure}{0.25\textwidth}%
+\vspace{-2.95cm}
+\hspace{-2.5cm}
+\scriptsize
 <<echo=FALSE, results=tex, fig.pos='b'>>=
 library("kableExtra")
-coefs <- coef(lmm_tree_c)
-coefs <- cbind(Node = as.numeric(rownames(coefs)), coefs)
+coefs <- coef(lmm_tree_r)
+coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
 colnames(coefs) <- c("Node", "Intercept", "Slope")
-kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs", 
-                             align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
-        caption = "Fixed-effects coefficient estimates from the terminal nodes of Figure 2.", 
-        row.names = FALSE, digits = 3L),
-  font_size = 11, full_width=FALSE)
-@
-
-Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates.
-Figure~\ref{fig:LMM_tree_c} shows an example tree, which identified socio-economic status (SES) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of these children, assessed at baseline.
-Four subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope (see Table~\ref{tab:local_coefs}).
-For children in the group with lower levels of SES, the intercept is also lowest, indicating lower science ability.
-For intermediate levels of SES (between $-0.41$ and $1.23$), there is also an effect of internalizing problems, where children with higher internalizing problems have a lower intercept.
-Highest science ability intercepts are observed in the group with highest SES.
-It appears that nodes with higher intercepts also tend to have higher slopes, indicating that children with relatively higher ability also show larger ability gains over time.
-
-The tree in Figure~\ref{fig:LMM_tree_c} has been estimated with generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}.
-GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas.
-In the current paper we aim to extend GLMM trees, so that they can be applied to partitioning LGCMs.
+coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
+coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
+coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
+                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
+kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
+        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
+@
+\label{subfig:coefs}
+\end{subfigure}
+\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
+Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.
+The table presents numerical estimates of fixed intercepts and slopes.}
+\label{fig:LMM_tree_r}
+\end{figure}%
+
+<<echo=FALSE, results=tex, fig.pos='b'>>=
+# library("kableExtra")
+# coefs <- coef(lmm_tree_r)
+# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
+# colnames(coefs) <- c("Node", "Intercept", "Slope")
+# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
+# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
+# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
+#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
+# kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs",
+#                              align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
+#         caption = "Fixed-effects coefficient estimates from the terminal nodes of Figure 2.",
+#         row.names = FALSE, digits = 3L),
+#   font_size = 11, full_width=FALSE)
+@
+
+The tree in Figure~\ref{fig:LMM_tree_r} has been estimated with generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}.
+GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are nested in treatment centers, classrooms and/or geographical areas, for example.
+In the current paper we extend GLMM trees, so that they can be applied to partitioning LGCMs.
 The general idea of GLMM trees is appealing for subgroup detection in almost any type of mixed-effects model.
-Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates).
+Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and the predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates).
 In this paper, we propose and test two extensions of GLMM trees that account for these characteristics.
@@
 There are several alternative RPMs that can be used to partition linear growth-curve models (LGCMs): GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}.
-Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.
+Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods specifically target subgroups with different time-by-treatment interactions, so are not generally applicable for partitioning growth curve models.}.
 
@@
 Other methods for partitioning LGCMs take a fully local estimation approach: Within every node or subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that subgroup only.
-This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity.
-GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results.
-Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.
+This fully local estimation approach provides more flexibility, but yields higher computational burden and model complexity.
+In contrast, GLMM trees estimate a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results.
+Furthermore, the fully local estimation requires possible partitioning variables to be measured at the highest level of nesting, while GLMM trees' local-global estimation approach allows partitioning variables to be measured at any level.
 
@@
 This reduces computational load, while it also allows for separating variable and cutpoint selection, thus preventing selection bias.
-The GLMM tree algorithm also employs a two-step approach to split selection and thereby shares these advantages.
-
-Given this lower model complexity, lower computational burden and lack of variable selection bias, GLMM trees might be particularly useful for subgroup detection in LGCMs.
-In the remainder of this section, we explain how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories.
-Next, we assess the performance of the proposed adjustments to GLMM trees in partitioning LGCMs: In Study~I, we assess performance of the proposed adaptations to GLMM trees in simulated datasets.
-In Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
-In the Study~III, we compare the performance of the proposed adjustments of GLMM trees in existing datasets on children's development of reading, math and science abilities.
+The GLMM tree algorithm shares these advantages, because it also employs a two-step approach to split selection.
+
+Given their unbiased variable selection, lower model complexity and computational burden, GLMM trees might be particularly useful for subgroup detection in LGCMs.
+The next section explains how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories.
+Next, the performance of the proposed adjustments is evaluated: In Study~I, we assess performance in simulated datasets, in Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
+In Study~III, we assess performance of the proposed adjustments in existing datasets on children's development of reading, math and science abilities.
 In the Discussion, we summarize our findings and discuss implications.
@@
 
-\section{Estimation of GLMM trees}
+\FloatBarrier
+\section{Estimation of GLMM trees and Adaptations for Longitudinal Data}
 \label{sec:estimation}
@@
 
-The following two subsections describes an alternative initialization strategy for Step~0 and an alternative approach for fitting the GLM tree from Step~1.
+The following two subsections describe alternative approaches for the initialization in Step~0 and for fitting the GLM tree in Step~1.
 Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve performance when partitioning longitudinal data.
@@
 
-Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
-\cite{SelaySimo12} specifically assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size.
+Previous publications on mixed-effects recursive partitioning find that initializing the random-effect estimates with zero yields accurate estimates of subgroup memberships and the final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
+\cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased with increasing sample size.
 In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data.
 With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data.
-If the random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations.
+If random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations.
 Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery.
@@
 \caption{GLMM tree estimated by initializing with zero random effects.}
-<<fig=TRUE, height=7, width=12, echo=FALSE, out.width=12>>=
+<<fig=TRUE, height=8, width=14, echo=FALSE>>=
 lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
@@
 
+To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}.
+In fact, the tree presented in Figure~\ref{fig:LMM_tree_r} was estimated by initializing with the random effects.
+Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}.
+The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in  Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills.
+Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples.
+
+\FloatBarrier
+\subsection{Partitioning}
+
 \begin{figure}%
-\caption{GLMM tree initialized by estimating the random effects.}
-<<fig=TRUE, height=9, width=15, echo=FALSE>>=
-lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
-plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
-     nodesize_level = "CHILDID", tp_args = 
-       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
-            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
-icc_r <- as.data.frame(VarCorr(lmm_tree_r))
-lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
+\caption{GLMM tree estimated with cluster-level parameter stability tests.}
+<<fig=TRUE, height=8, width=12, echo=FALSE>>=
+plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
+     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
+                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
+                                        xaxis.labs = c(0, 12, 36, 60)),
+     gp = gpar(cex = 1.5))
 @
@@
 Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
-\label{fig:LMM_tree_r}
+\label{fig:LMM_tree_c}
 \end{figure}%
 
-To illustrate how initialization may make a difference we applied both approaches to the dataset from Figures~\ref{fig:global_trajectories}~and~\ref{fig:LMM_tree_c}.
-Initializing estimation assuming zero random effects yielded the tree in Figure~\ref{fig:LMM_tree}, with \Sexpr{width(lmm_tree$tree)} subgroups and an estimated random-intercept variance of \Sexpr{formatC(data.frame(VarCorr(lmm_tree$lmer))$vcov[1], digits = 2)}.
-Additional splits were implemented based on gender, race, internalizing problems and fine motor skills.
-Considering the relatively large number of subgroups and relatively small sample sizes in some of the terminal nodes, this tree may not generalize well to other samples.
-Initializing with estimation of the random effects yielded the tree depicted in Figure~\ref{fig:LMM_tree_r}, with \Sexpr{width(lmm_tree_r$tree)} subgroups and an estimated random-intercept variance of \Sexpr{formatC(data.frame(VarCorr(lmm_tree_r$lmer))$vcov[1], digits = 2)}.
-Compared to the tree in Figure~\ref{fig:LMM_tree_c}, an additional split was implemented based on gross motor skills (GMOTOR): children with further developed motor skills at baseline have a higher intercept.
-
-\subsection{Partitioning}
-
 The subgroup structure in Step~1 is estimated by a GLM tree using the general model-based recursive partitioning (MOB) algorithm of \cite{ZeilyHoth08}.
-Here, we give a short general overview and then comment on the aspects of the algorithm that may be particularly relevant for LGCMs.
+Here, we give a general overview and then comment on the aspects of the algorithm that may be particularly relevant for LGCMs.
 In the case of GLMs (with random effects held constant in an offset), the MOB algorithm cycles iteratively through the following steps:
@@
 where $\hat{J}$ is a suitable estimate of the covariance matrix of the parameter estimates, and $n_j$ gives the number of observations in the current subgroup.
-Further, $\hat{\psi}_{\sigma(u_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(u_{ik})$} denoting their ordering by the values of partitioning variable $u_k$.
+Further, $\hat{\psi}_{\sigma(u_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(u_{ik})$} denoting their ordering by the values of partitioning covariate $u_k$.
 Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation associated with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
@@
 Approximate asymptotic $p$-values for the \textit{supLM} statistic can be computed with the method of \cite{Hans97}.
-Categorical covariates do not provide an implicit ordering.
-Scores are therefore binned at each level of the covariate and a test statistic is computed that does not depend on the ordering of the levels \citep{MerkyFan14}.
+Categorical covariates do not provide an implicit ordering and scores are therefore binned at each level of the covariate.
+From these, a test statistic is computed that does not depend on the ordering of the levels \citep{MerkyFan14}.
 
@@
 
-In fact, the tree in Figure~\ref{fig:LMM_tree_c} was obtained by application of GLMM trees with cluster-level parameter stability tests.
+The tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests.
 Application of GLMM trees with both cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates.
-It provides the most parsimonious tree structure of the three estimation approaches employed thus far.
-The estimated random-intercept variance was \Sexpr{formatC(data.frame(VarCorr(lmm_tree_c$lmer))$vcov[1], digits = 2)}, the highest value obtained thus far, illustrating how some variation in the response variable can be captured by either the random effects or by the tree structure.
-In the next sections, we assess the performance of the original algorithm and proposed adaptations through empirical evaluations.
+Compared to Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, the cluster-level parameter stability tests provided the most parsimonious tree structure thus far.
+The estimated variances of the random intercept were \Sexpr{formatC(data.frame(VarCorr(lmm_tree_r$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree_r}, \Sexpr{formatC(data.frame(VarCorr(lmm_tree$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree}) and \Sexpr{formatC(data.frame(VarCorr(lmm_tree_c$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree_c}, indicating that the more parsimonious the tree structure, the more variance will be captured by the random effects.
+
+In the next sections, we will assess performance of the original algorithm and proposed adaptations through more extensive empirical evaluations.
 
@@
 \begin{figure}[tb]
-\caption{Design of the subgroup-specific fixed effects in the simulation.}
+\caption{Design of subgroups and fixed effects.}
 <<fig=TRUE, height=5.5, width=5.5, echo=FALSE>>=
@@
 where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part.
-These $\beta_j$ values are reported below the terminal node panels in Figure~\ref{fig:design_tree}.
+$\beta_j$ values are reported below the terminal node panels in Figure~\ref{fig:design_tree}.
 The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints.
@@
 Values of $b_i$ (random intercepts and slopes) were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described below.
-Values of $\epsilon_{i}$ were independently generated from a normal distribution with a mean of 0 and variance of 5.
+Values of $\epsilon_{i}$ were independently generated from a normal distribution with $\mu = 0$ and $\sigma^2 = 5$.
 
@@
 \item Variance of the random slope: small ($\sigma_{b_1}^2 = 0.1$) or large ($\sigma_{b_1}^2 = 0.4$).
-\item Number of noise partitioning variables: small ($p = 5$) or large ($p = 25$).
+\item Number of noise variables: small ($p = 5$) or large ($p = 25$).
 \item Intercorrelation between partitioning variables: absent ($\rho = 0$) or present ($\rho = 0.3$).
@@
 To compute clustered covariances, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}.
-We employed the outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.
+We employed the outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator for the covariances.
 
@@
 
+\FloatBarrier
 \subsection{Results}
@@
 
-\begin{figure}[t!]
+\begin{figure}[b]
 \caption{Tree size distributions for LM trees (left panel) and LMM trees (middle and right panel).}
@@
 
-{\textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= Mean absolute deviation from true tree size.
+{\footnotesize \textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= mean absolute deviation from true tree size.
 Gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
@@
 The default fitting approach tends to overfit, irrespective of the random-effects specification, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average.
-The use of cluster-level covariances in the parameter stability tests successfully mitigated overfitting, irrespective of the random-effects specification.
-LMM trees with clustered covariances implemented an average number of splits that was closest to the true tree size, with little difference between random-effects specifications (i.e., with and without random slopes).
-
-In terms of MAD, however, the LMM tree initializing estimation with the random effects performed best, if only random intercepts (not random slopes) were estimated.
-Likely, initializing estimation with the random effects may be successful only when the random-effects specification is kept relatively simple.
-With a more complex random-effects specification (i.e., both random intercept and slope estimated; right panel in Figure~\ref{fig:LMM_sizes}, initializing estimation with the random effects resulted in underfitting.
-Subgroup differences are then more likely to be captured by the random effects than by the tree structure.
-
-The combined use of random-effects initialization and cluster-level covariances was not very effective, irrespective of the random-effects specification.
-All in all, use of clustered covariances appears to provide the most reliable improvement in performance.
+The use of cluster-level covariances in the parameter stability tests successfully mitigated overfitting: LMM trees with clustered covariances with (middle panel) or without (right panel) random slopes showed an average number of splits closest to the true tree size.
+
+In terms of MAD, however, LMM trees initializing estimation with the random effects performed best, but only if the random slope was not estimated.
+Initializing estimation with the random effects may only be useful if the random-effects specification is kept relatively simple.
+With more complex random-effects specifications, initializing estimation seems to result in too few groups being detected; subgroup differences are then more likely to be captured by the random effects than by the tree structure.
+The second-lowest MAD was observed for LM trees with clustered covariances (left panel).
+Combined use of random-effects initialization and cluster-level covariances was not very effective, irrespective of the random-effects specification.
 
@@
 Main effects of the data-generating parameters were as expected: The strongest effects were for $\sigma_{b_1}$ and $N$, with higher values resulting in a higher number of splits.
-The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, and were involved in minor interactions: With the default fitting approach, they had a positive effect on the number of implemented splits; with clustered covariance or random-effects initialization, they had a negative effect on the number of splits.
-The correlation between partitioning variables hardly affected the number of implemented splits.
+The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, while the correlation between partitioning variables hardly affected the number of implemented splits.
 
@@
   facet_grid(~ method, scales = "free", space = "free") +
-  labs(x = "", y = "no. of splits") + theme(legend.position = "none") +
+  labs(x = "", y = "no. of splits") + theme(legend.position = "none",
+                       text=element_text(size=9)) +
   scale_x_discrete(labels = scales::parse_format()) +
@@
 \vspace{-0.5cm}%
-{
-\footnotesize \textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= Mean absolute deviation from true tree size.
-Gray circles represent counts, dark gray horizontal lines indicate true number of splits~3.
+{\footnotesize \textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= mean absolute deviation from true tree size.
+Gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
 Distances on the $y$-axis are on the log scale; $x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.}
@@
         escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
-        caption = "Cross-validated mean squared errors for each of the response variables."),
+        caption = "Variables selected for the first splits by each of the partitioning approaches."),
   "\\footnotesize \\\\ \\textit{Note.} $u_1$ is the true first splitting variable and is a binary factor; all other partitioning variables are continuous, with $u_2$ and $u_3$ being true splitting variables (nodes 2 and 3).
@@
 LongCART showed a strong tendency to select $u_2$ or $u_3$ for the first split.
-Closer inspection of the fitted LongCART trees revealed that in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets where no splits were implemented, $u_1$ was the strongest splitting candidate, but the parameter stability tests did not reach significance.
+Closer inspection of the fitted LongCART trees revealed that in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets in which no splits were implemented, $u_1$ was the strongest splitting candidate, but the parameter stability tests did not reach significance.
 This suggests the tests proposed by \cite{KundyHare19} are less sensitive to instability of the intercept (compared to instability of the slope), or less sensitive to instability with respect to categorical covariates (compared to instability with respect to continuous covariates).
@@
                      rep("SEM tree\n(score)", times = 3),
-                     "LongCRT")
+                     "LongCART")
 times <- reshape2::melt(times)
@@
             size = 3, col = "gray44") +
+  theme(text = element_text(size = 10)) +
   scale_y_continuous(trans = "log", limits = c(0.01, 100000), 
@@
 \end{subfigure}%
-\\ \textit{Note.} $y$-axis is on the log scale; M = mean computation time in seconds. 
+\\ {\footnotesize \textit{Note.} $y$-axis is on the log scale; M = mean computation time in seconds. }
 \label{fig:comp_times}
@@
 
-\section{Study~III: Cross-validated predictive accuracy on real data}
+\section{Study~III: Partitioning Academic Trajectories}
 
@@
 Data were collected from 21,304 children from 1,018 schools across the USA.
-Assessments took place from kindergarten in 1998 through 8th grade in 2007.
-In the current analyses, we focus on the assessments from kindergarten, 1st, 3rd, 5th and 8th grade.
-
-Response variables are reading, math, and science abilities.
-These were assessed using multi-item cognitive tests, from which latent ability estimates were computed with a mean of zero and variance of one.
-Reading and math abilities were assessed in all five rounds of data collection, while science knowledge was assessed in the 3rd, 5th and 8th grade only.
+Assessments took place from kindergarten in 1998 through 8th grade in 2007, here we focus on assessments from kindergarten, 1st, 3rd, 5th and 8th grade.
+
+Response variables are reading, math, and science abilities, which were assessed using multi-item cognitive tests.
+Latent ability estimates were computed with a mean of zero and variance of one.
+Reading and math abilities were assessed in all five rounds of data collection, science knowledge was assessed in 3rd, 5th and 8th grade.
 We analyzed data from children who completed all assessments yielding $N = 6,277$ for reading; $N = 6,512$ for math; $N = 6,625$ for science.
 
-Our analyses focus on predicting trajectories of reading, math and science abilities, based on baseline characteristics.
 Time was measured as the number of months since the baseline assessment.
@@
 
-We used eleven time-invariant covariates as potential partitioning variables: Gender (51.1\% male); Age in months at baseline (range 53 to 96; M = 6.14 years at baseline); Race (8 categories); First time in kindergarten (yes/no); Socio-economic status (range $-$5 to 3); Fine motor skills (e.g., drawing figures; range 0 to 9); Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8); Interpersonal skills (range 1 to 4); Self-control (range 1 to 4); Internalizing problem behavior (range 1 to 4); Externalizing problem behavior (range 1 to 4).
+We used 11 time-invariant covariates as potential partitioning variables, all assessed at baseline: Gender (51.1\% male); Age in months (range 53 to 96; M = 6.14 years); Race (8 categories); First time in kindergarten (yes/no); Socio-economic status (range $-$5 to 3); Fine motor skills (e.g., drawing figures; range 0 to 9); Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8); Interpersonal skills (range 1 to 4); Self-control (range 1 to 4); Internalizing problem behavior (range 1 to 4); Externalizing problem behavior (range 1 to 4).
 
@@
 
-We fitted four LM(M) trees using the approaches which showed most promise in the simulations: 
+We applied five LM(M) trees to the data, focusing on the original GLMM tree approach and the approaches that performed well in the simulations:
 % 
@@
   \item An LM tree ($\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$) using clustered covariances in the parameter-stability tests.
-  \item An LMM tree using clustered covariances, with a random intercept freely esimtated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$).
-  \item An LMM tree using clustered covariances, with both random intercept and slope freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$). 
-  \item An LMM tree using random-effect initialization with the full-sample estimate, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$). 
+  \item An LMM tree using observation-level covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and initialization assuming zero random effects.
+  \item An LMM tree using clustered covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and initialization assuming zero random effects.
+  \item An LMM tree using clustered covariances, with both random intercept and slope freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$) and initialization assuming zero random effects .
+  \item An LMM tree uwith a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and random-effect initialization with the full-sample estimate. 
 \end{itemize}
 %
-As a benchmark, we also estimated two standard LMMs with a fixed effect of time: One LMM with a random intercept freely estimated, one LMM with both random intercept and slope freely estimated.
-The LMMs did not include additional predictors in the model, only timepoint and the child identifier.
-
-%To compare performance of LM(M) trees on real data, we could not employ SEM tree or LongCART, so we opted for longRPart \citep{AbdoyLeBl02}.
-Neither SEM trees, LongCART, longRPart or longRPart2 implementations provide functionality to obtain predictions for new observations.
-Writing custom functionality to allow for prediction was most straightforward for longRPart.
-Furthermore, performance of longRPart will be representative of the other methods, in the sense that all four methods fit full mixed-effects models locally.
-Like LongCART and longRPart2, longRPart fits a mixed-effects model in each of the terminal nodes.
-Like LRT-based SEM trees, longRPart uses an exhaustive LRT-based search in every node to find the optimal next split.
-
-%Using package \textbf{longRPart} \citep[version 1.0; ][]{StewyAbdo12}, we fitted two trees with different random effects specifications: A tree comprising a random intercept only, and a tree comprising both a random intercept and slope with respect to student.
-In order to implement a split, longRPart uses a criterion based on the minimum decrease in deviance, which is set to 0.01, by default.
-This yielded a very large number of splits in the current analyses and poor predictive accuracy.
+Although LRT-based SEM trees performed very well in the simulations, they could not be used in this study because growth curve SEMs do not allow for incorporating continuous time.
+
+%We therefore opted for using longRPart \citep[version 1.0; ][]{AbdoyLeBl02}, which employs the exact same LRT-based splitting criterion as LRT-based SEM trees.
+It however fits a mixed-effects model in every node, instead of a SEM, allowing for the incorporation of time as a continuous variable.
+We fitted two longRPart trees with different random effects specifications:
+% 
+%\begin{itemize}
+%  \item A longRPart tree which freely estimates a random intercept ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$).
+%  \item A longRPart tree which freely estimates both a random intercept and slope ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$).
+%\end{itemize}
+%
+%The default longRPart settings yielded a very large number of splits in the current analyses and poor predictive accuracy.
 We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = \Sexpr{2^5} terminal nodes, or \Sexpr{(2^5)-1} splits.
-Significance of the LRT, or a cross-validation approach to determine whether to implement a split or not may yield better results, but this has not been implemented.
+We created a custom function for computing predictions from fitted longRPart trees, as this functionality was not included in the package itself.
+
+%As benchmarks, we included two LMMs with a fixed effect of time: 
+%
+%\begin{itemize}
+%  \item An LMM with a random intercept freely estimated.
+%  \item An LMM with both random intercept and slope freely estimated.
+%\end{itemize}
+%
+%Besides time and the child identifier, the LMMs did not include additional predictors in the model.
 
@@
 
-The ECLS-K datasets have exceptionally large sample sizes.
-We therefore employed random sampling to obtain training samples of $N=250$ children, likely more representative of real-world studies in psychology.
-For each combination of response variable (math, reading, or science) and sample size, we performed 100 repetitions.
-We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample in the current repetition.
-This separation of train and test observations does not allow for using the random effects in computing predictions, these were thus assumed 0; only accuracy of the fixed-effects parameters is thus qualified by the MSEs.
-Tree size was measured by the number of splits in each tree.
+The ECLS-K datasets have exceptionally large sample sizes, so we employed random sampling to obtain training samples of $N=250$ children, likely more representative of real-world studies in psychology.
+We performed 100 repetitions for each response variable (math, reading, or science).
+We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values (MSE) for all children not included in the training sample in the current repetition.
+This separation of train and test observations does not allow for using the random effects in computing predictions; the cross-validated MSEs only quantify accuracy of the fixed-effects parameters.
+Tree size was measured by counting the number of splits in each tree.
 
@@
 
-Figure~\ref{fig:application_MSEs} presents MSE distributions for each of the estimation approaches and Table~\ref{tab:application_MSEs} presents the means and standard deviations.
-%As expected, all methods show higher accuracy with larger sample sizes.
-LM(M) trees tend to show better predictive accuracy than longRPart trees.
-Pairwise $t$-tests indicated significantly better performance for each of the LM(M) tree approaches employing clustered covariances, compared to both longRPart approaches (Bonferroni corrected $p$-values all $< .01$).
-However, these performance differences are small; the effects of sample size are much more obvious than the differences between methods in Figure~\ref{fig:application_MSEs}.
-Performance of the benchmark LMMs indicates that a substantial amount of variance can be explained by the single fixed linear effect of time.
-Further, all LM(M) trees outperform the benchmark LMMs.
-Differences in performance between LM(M) trees are small.
-Table~\ref{tab:application_MSEs} and Figure~\ref{fig:application_MSEs} indicate that LMM trees employing cluster-level covariances performed best for predicting math and reading abilities.
-Trees estimating only a random intercept performed best, but performance differed only about 1 SE compared to trees estimating both random intercept and slope.
-
-For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best, while this was the worst-performing approach for math and reading.
-LM trees with clustered covariances never performed best, illustrating the benefit of estimating random effects.
-
-%For longRPart, trees comprising both a random intercept and slope performed best for predicting math and reading, trees with a random intercept only performed best for predicting science.
-The differences in performance between math and reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of science abilities.
-
-%Again, the performance of LM(M) trees employing cluster-level covariances seems hardly affected by the specification of the random-effects: Figure~\ref{fig:application_MSEs} shows very similar performance for LM(M) trees with random intercepts, slopes or neither.
-Pairwise $t$-tests did not indicate significant differences between those three approaches for any of the response variables.
+Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions.
+Differences in predictive performance are small, all $R^2$ differences are smaller than 0.01.
+Bonferroni-adjusted pairwise $t$-tests indicated no difference in performance between default LMM trees and LMM trees with random-effects initialization for any of the three outcomes.
+Incontract, the three LM(M) trees using clustered covariances performed significantly better for the math and reading outcomes.
+For the science outcomes, no significant differences were observed.
+Thus, cluster-level covariances seem to provide the most robust improvement of performance.
+
+Figure~\ref{fig:application_sizes} presents tree size distributions.
+Similar to the simulation study's results, default LMM trees implement the largest number of splits.
+Cluster-level covariances provide a robust reduction in the number of splits.
+Given their non-distinguishable predictive performance, LM trees with clustered covariances may be preferred if a obtaining a sparse result is critical.
+LMM trees with clustered covariances may provide less sparsity but better predictive accuracy, irrespective of whether random intercepts and/or slopes have been specified.
+Random-effects initialization yields tree sizes similar to the default fitting approach.
 
@@
 ## Math
-MSE_df <- data.frame(sapply(math_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- data.frame(sapply(math_MSEs[ , c(1:4, 6)], 
                               function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.3f", tapply(x, math_MSEs$N, 
@@
 ## Reading
-MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.3f", tapply(x, read_MSEs$N, 
@@
 ## Science 
-MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , -c(4,6,7,10)], 
+MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , c(1:4, 6)], 
                                function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, 
@@
 
-MSE_df <- MSE_df[-(5:6) , ] ## omit longRPart trees
-
 ## tmp3 is math
-tmp3 <- data.frame(stack(math_MSEs[ , -c(4,6,7:10)]),
-                    dataset_id = factor(rep(1:nrow(math_MSEs), times = 12)),
+tmp3 <- data.frame(stack(math_MSEs[ , c(1:4, 6)]),
+                    dataset_id = factor(rep(1:nrow(math_MSEs), times = 5)),
                    N = math_MSEs$N)
@@
 tmp3$method2 <- as.character(tmp3$method)
-tmp3$method2[grepl("LMtree", tmp3$method2)] <- "LM tree\n\n(cl.cov.)"
+
+tmp3$method2[tmp3$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
+tmp3$method2[tmp3$method2 == "LMMtree"] <- "LMM tree"
 tmp3$method2[tmp3$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cl.cov.)"
 tmp3$method2[tmp3$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
-tmp3$method2[tmp3$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cl.cov.+ran.eff.)"
-tmp3$method2[grepl("longR", tmp3$method2)] <- "longRPart"
 tmp3$random <- ifelse(tmp3$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
-                      ifelse(tmp3$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                      ifelse(tmp3$method %in% c("LMMtree", "LMMtree_c", "LMMtree_r"), 
                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp3$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+                             ifelse(tmp3$method %in% c("LMMtree_sc"), 
                                 "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
@@
   theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
-        legend.position="none", axis.text.y = element_text(size = 8)) +
+        legend.position="none", axis.text.y = element_text(size = 8),
+        axis.title.y = element_text(size = 10)) +
   labs(x = "", y = "MSE")
@@
 ## tmp4 is reading
-tmp4 <- data.frame(stack(read_MSEs[ , -c(4,6,7:10)]),
-                    dataset_id = factor(rep(1:nrow(read_MSEs), times = 12)),
+tmp4 <- data.frame(stack(read_MSEs[ , c(1:4,6)]),
+                    dataset_id = factor(rep(1:nrow(read_MSEs), times = 5)),
                    N = read_MSEs$N)
@@
 tmp4$method2[grepl("longR", tmp4$method2)] <- "longRPart"
+tmp4$method2[tmp4$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
+tmp4$method2[tmp4$method2 == "LMMtree"] <- "LMM tree"
+tmp4$method2[tmp4$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cl.cov.)"
+tmp4$method2[tmp4$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
 tmp4$random <- ifelse(tmp4$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
-                      ifelse(tmp4$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                      ifelse(tmp4$method %in% c("LMMtree", "LMMtree_c", "LMMtree_r"), 
                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp4$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+                             ifelse(tmp4$method %in% c("LMMtree_sc"), 
                                 "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
@@
         axis.title.x=element_blank(), axis.text.x=element_blank(),
-        legend.position="none", axis.text.y = element_text(size = 8)) +
+        legend.position="none", axis.text.y = element_text(size = 8),
+        axis.title.y = element_text(size = 10)) +
   labs(x = "", y = "MSE")
@@
 ## tmp5 is science
-tmp5 <- data.frame(stack(scie_MSEs[ , -c(4,6,7:10)]),
-                    dataset_id = factor(rep(1:nrow(scie_MSEs), times = 12)),
+tmp5 <- data.frame(stack(scie_MSEs[ , c(1:4,6)]), ## omit longRPart with 7:10
+                    dataset_id = factor(rep(1:nrow(scie_MSEs), times = 5)),
                    N = scie_MSEs$N)
@@
 tmp5$method2 <- as.character(tmp5$method)
-tmp5$method2[grepl("LMtree", tmp5$method2)] <- "LM tree\n\n(cluster)"
-tmp5$method2[tmp5$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
-tmp5$method2[tmp5$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
-tmp5$method2[tmp5$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
-tmp5$method2[grepl("longR", tmp5$method2)] <- "longRPart"
+tmp5$method2[tmp5$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
+tmp5$method2[tmp5$method2 == "LMMtree"] <- "LMM tree"
+tmp5$method2[tmp5$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cl.cov.)"
+tmp5$method2[tmp5$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
 tmp5$random <- ifelse(tmp5$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
-                      ifelse(tmp5$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
+                      ifelse(tmp5$method %in% c("LMMtree", "LMMtree_c", "LMMtree_r"), 
                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp5$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
+                             ifelse(tmp5$method %in% c("LMMtree_sc"), 
                                 "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
@@
   theme(strip.background = element_blank(), strip.text.x = element_blank(),
-        legend.position="none", axis.text.y = element_text(size = 8)) +
+        legend.position="none", axis.text.y = element_text(size = 9),
+        axis.text.x = element_text(size = 9), axis.title.y = element_text(size = 10)) +
   labs(x = "", y = "MSE") +
@@
 } Top, middle and bottom panels depict math, reading and science ability trajectories, respectively.
-The secondary $y$-axis on the right quantifies the proportion of variance explained, computed as $1 - \frac{\text{MSE}}{\text{var}(y)}$.}
+The secondary $y$-axis on the right quantifies the proportion of variance explained, computed as $1 - \frac{\text{mean(MSE)}}{\text{var}(y)}$.}
 \label{fig:application_MSEs}
@@
 <<echo=FALSE, eval=FALSE>>=
-## Pairwise difference tests
+## Pairwise difference tests of MSEs
 pairwise.t.test(tmp3$MSE, tmp3$method, alternative = "two.sided", 
                 p.adjust = "bonferroni") ## math
+#            LMtree_c LMMtree LMMtree_c LMMtree_r
+# LMMtree    3.0e-05  -       -         -        
+# LMMtree_c  1        3.5e-08 -         -        
+# LMMtree_r  2.9e-05  1       3.3e-08   -        
+# LMMtree_sc 1        1.1e-06 1         1.0e-06  
 pairwise.t.test(tmp4$MSE, tmp4$method, alternative = "two.sided", 
                 p.adjust = "bonferroni") ## reading
+#            LMtree_c LMMtree LMMtree_c LMMtree_r
+# LMMtree    2.0e-06  -       -         -        
+# LMMtree_c  1        7.2e-07 -         -        
+# LMMtree_r  2.1e-06  1       7.6e-07   -        
+# LMMtree_sc 1        7.8e-09 1         8.2e-09  
 pairwise.t.test(tmp5$MSE, tmp5$method, alternative = "two.sided", 
                 p.adjust = "bonferroni") ## science
-@
-
-\setlength{\tabcolsep}{3pt}
+#           LMtree_c LMMtree LMMtree_c LMMtree_r
+# LMMtree    0.49     -       -         -        
+# LMMtree_c  1.00     1.00    -         -        
+# LMMtree_r  0.31     1.00    1.00      -        
+# LMMtree_sc 0.61     1.00    1.00      1.00     
+@
 
@@
 library("kableExtra")
-rownames(MSE_df) <- c("LM tree$^c$", "LMM tree$^{i,c}$", "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$")  
-                      #"longRPart$^i$", "longRPart$^{i,s}$")
+rownames(MSE_df) <- c("LM tree$^c$", "LMM tree$^i$", "LMM tree$^{i,c}$", 
+                      "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$")
 colnames(MSE_df) <- rep(c("M", "SD", "$R^2$"), times = 3)
-MSE_df <- MSE_df[c(1,2,4,3), ]
 
 ## Add benchmark LMMs
-load("LMM MSEs, N = 250, Math.Rda")
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 1] <- format(colMeans(MSEs), digits = 4)
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 2] <- round(sapply(MSEs, sd), digits = 3)
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 3] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_math)
-
-load("LMM MSEs, N = 250, Reading.Rda")
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 4] <- format(colMeans(MSEs), digits = 4)
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 5] <- round(sapply(MSEs, sd), digits = 3)
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 6] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_read)
-
-load("LMM MSEs, N = 250, Science.Rda")
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 7] <- format(colMeans(MSEs), digits = 4)
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 8] <- round(sapply(MSEs, sd), digits = 3)
-MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 9] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_scie)
+#load("LMM MSEs, N = 250, Math.RDa")
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 1] <- format(colMeans(MSEs), digits = 4)
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 2] <- round(sapply(MSEs, sd), digits = 3)
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 3] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_math)
+
+#load("LMM MSEs, N = 250, Reading.RDa")
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 4] <- format(colMeans(MSEs), digits = 4)
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 5] <- round(sapply(MSEs, sd), digits = 3)
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 6] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_read)
+
+#load("LMM MSEs, N = 250, Science.RDa")
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 7] <- format(colMeans(MSEs), digits = 4)
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 8] <- round(sapply(MSEs, sd), digits = 3)
+#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 9] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_scie)
+
+## boldface highest R2 values
+best_ids <- apply(MSE_df[ , c(1,4,7)], 2, which.min)
+MSE_df[best_ids[1] , 1:3] <- paste0("\\textbf{", MSE_df[best_ids[1] , 1:3], "}")
+MSE_df[best_ids[2] , 4:6] <- paste0("\\textbf{", MSE_df[best_ids[2] , 4:6], "}")
+MSE_df[best_ids[3] , 7:9] <- paste0("\\textbf{", MSE_df[best_ids[3] , 7:9], "}")
 
@@
   add_footnote(
-  kable(MSE_df, format = "latex", booktabs = TRUE, label = "application_MSEs", align = "c", 
+  kable(MSE_df[c(1:3, 5, 4), ], format = "latex", booktabs = TRUE, label = "application_MSEs", align = "c", 
         escape=FALSE, linesep = "", # linesep command suppresses addlinesep every 5 rows
         caption = "Cross-validated mean squared errors for each of the response variables."),
-  "\\footnotesize \\\\ \\textit{Note.} Means and standard deviations computed over 100 repetitions.
-Standard deviations of the MSE can be interpreted as standard errors.
-$R^2$ was computed as $\\frac{MSE}{var(y)}$.
+  "\\footnotesize \\\\ \\textit{Note.} Means and standard deviations computed over 100 cross-validation repetitions.
+Boldfaced values indicate the best-performing method for each outcome.
+$R^2$ was computed as $\\frac{\\text{mean(MSE)}}{\\text{var}(y)}$.
 $^c$~cluster-level covariances; $^r$~estimation initialized with random effects; $^i$~random-intercept variance freely estimated; $^s$~random-slope variance freely estimated.",
@@
 @
+
+<<eval=FALSE, echo=FALSE>>=
+math_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = mathdata)
+(as.data.frame(VarCorr(math_LMM))$vcov / var_math)
+(as.data.frame(VarCorr(math_LMM))$vcov / as.data.frame(VarCorr(math_LMM))$vcov[4])
+
+read_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = readdata)
+(as.data.frame(VarCorr(read_LMM))$vcov / var_read)
+(as.data.frame(VarCorr(read_LMM))$vcov / as.data.frame(VarCorr(read_LMM))$vcov[4])
+
+scie_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = sciedata)
+(as.data.frame(VarCorr(scie_LMM))$vcov / var_scie)
+(as.data.frame(VarCorr(scie_LMM))$vcov / as.data.frame(VarCorr(scie_LMM))$vcov[4])
+@
+
+\begin{figure}[!bt]%
+\caption{Sizes of trees fitted to math, reading, and science ability trajectories.}
+\begin{subfigure}{\textwidth}%
+<<echo=FALSE, fig=TRUE, height=2, width=5>>=
+## Tree size
+load("sizes, N = 250, Math.Rda")
+math_sizes <- cbind(sizes, N = 250)
+#load("sizes, N = 1000, Math.Rda")
+#math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
+#sapply(math_sizes, function(x) table(is.na(x)))
+#sapply(math_sizes, function(x) tapply(x, math_sizes$N, mean, na.rm =TRUE))
+
+load("sizes, N = 250, Reading.Rda")
+read_sizes <- cbind(sizes, N = 250)
+#load("sizes, N = 1000, Reading.Rda")
+#read_sizes <- rbind(read_sizes, cbind(sizes, N = 1000))
+#sapply(read_sizes, function(x) table(is.na(x)))
+#sapply(read_sizes, function(x) tapply(x, read_sizes$N, mean, na.rm =TRUE))
+
+load("sizes, N = 250, Science.Rda")
+scie_sizes <- cbind(sizes, N = 250)
+#load("sizes, N = 1000, Science.Rda")
+#scie_sizes <- rbind(scie_sizes, cbind(sizes, N = 1000))
+#sapply(scie_sizes, function(x) table(is.na(x)))
+#sapply(scie_sizes, function(x) tapply(x, scie_sizes$N, mean, na.rm =TRUE))
+
+## Made a mistake in evaluating tree size in the Application scripts:
+## I took #splits = nrow(tree$frame)-1, assuming tree$frame contains a row for
+## each terminal node. It contains a row for each inner and terminal node.
+## Thus, #splits = (nrow(tree$frame)-1)/2; original number should be divided by 2:
+#math_sizes$longRPart <- math_sizes$longRPart/2 
+#math_sizes$longRPart_s <- math_sizes$longRPart_s/2 
+#read_sizes$longRPart <- read_sizes$longRPart/2 
+#read_sizes$longRPart_s <- read_sizes$longRPart_s/2 
+#scie_sizes$longRPart <- scie_sizes$longRPart/2 
+#scie_sizes$longRPart_s <- scie_sizes$longRPart_s/2
+
+## tmp6 is math
+tmp6 <- data.frame(stack(math_sizes[ , c(1:4,6)]),
+                    dataset_id = factor(rep(1:nrow(math_sizes), times = 5)),
+                   N = math_sizes$N)
+names(tmp6)[1:2] <- c("splits", "method") 
+tmp6$method2 <- as.character(tmp6$method)
+tmp6$method2[tmp6$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
+tmp6$method2[tmp6$method2 == "LMMtree"] <- "LMM tree"
+tmp6$method2[tmp6$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cl.cov.)"
+tmp6$method2[tmp6$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
+tmp6$random <- ifelse(tmp6$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp6$method %in% c("LMMtree", "LMMtree_c", "LMMtree_r"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
+                             ifelse(tmp6$method %in% c("LMMtree_sc"), 
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
+tmp6$N <- factor(tmp6$N)
+ggplot(tmp6) +
+  geom_boxplot(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
+  facet_grid(~method2, scales = "free", space = "free") +
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
+        axis.text.y = element_text(size = 8), axis.title.y = element_text(size = 10)) +
+  labs(x = "", y = "no. of splits")
+@
+\end{subfigure}
+\begin{subfigure}{\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.6, width=5>>=
+## tmp7 is reading
+tmp7 <- data.frame(stack(read_sizes[ , c(1:4,6)]),
+                    dataset_id = factor(rep(1:nrow(read_sizes), times = 5)),
+                   N = read_sizes$N)
+names(tmp7)[1:2] <- c("splits", "method") 
+tmp7$method2 <- as.character(tmp7$method)
+tmp7$method2[tmp7$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
+tmp7$method2[tmp7$method2 == "LMMtree"] <- "LMM tree"
+tmp7$method2[tmp7$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cl.cov.)"
+tmp7$method2[tmp7$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
+tmp7$random <- ifelse(tmp7$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp7$method %in% c("LMMtree", "LMMtree_c", "LMMtree_r"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
+                             ifelse(tmp7$method %in% c("LMMtree_sc"), 
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
+tmp7$N <- factor(tmp7$N)
+ggplot(tmp7) +
+  geom_boxplot(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
+  facet_grid(~method2, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits") +
+  theme(strip.background = element_blank(), strip.text.x = element_blank(),
+        axis.title.x=element_blank(), axis.text.x=element_blank(),
+        axis.text.y = element_text(size = 8), axis.title.y = element_text(size = 10))
+@
+\end{subfigure}
+\begin{subfigure}{\textwidth}
+<<echo=FALSE, fig=TRUE, height=2, width=5>>=
+## tmp8 is science
+tmp8 <- data.frame(stack(scie_sizes[ , c(1:4,6)]),
+                    dataset_id = factor(rep(1:nrow(scie_sizes), times = 5)),
+                   N = scie_sizes$N)
+names(tmp8)[1:2] <- c("splits", "method") 
+tmp8$method2 <- as.character(tmp8$method)
+tmp8$method2[tmp8$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
+tmp8$method2[tmp8$method2 == "LMMtree"] <- "LMM tree"
+tmp8$method2[tmp8$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cl.cov.)"
+tmp8$method2[tmp8$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
+tmp8$random <- ifelse(tmp8$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
+                      ifelse(tmp8$method %in% c("LMMtree", "LMMtree_c", "LMMtree_r"), 
+                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
+                             ifelse(tmp8$method %in% c("LMMtree_sc"), 
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
+tmp8$N <- factor(tmp8$N)
+ggplot(tmp8) +
+  geom_boxplot(aes(x=random, y=splits), 
+               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
+  facet_grid(~method2, scales = "free", space = "free") +
+  theme(strip.background = element_blank(), strip.text.x = element_blank(),
+        axis.text.y = element_text(size = 8), 
+        axis.text.x = element_text(size = 9), axis.title.y = element_text(size = 10)) +
+  labs(x = "", y = "no. of splits") +
+  scale_x_discrete(labels = scales::parse_format()) +
+  coord_cartesian(clip = 'off')
+@
+\end{subfigure}%
+{\\\footnotesize \textit{Note.} Top, middle and bottom panels depict math, reading and science ability trajectories, respectively.}
+\label{fig:application_sizes}
+\end{figure}%
+
+\FloatBarrier
+
+\section{Discussion}
+
+The simulations showed that the proposed extensions of GLMM trees are effective for partitioning LGCMs.
+Use of clustered covariances appears most effective and their good performance appears largely unaffected by (mis-)specification of the random effects; it may therefore be the safest choice in practice.
+Initializing estimation with the random effects was also effective, but only when the random-effects specification is kept simple (i.e., no estimation of random slopes).
+Combining cluster-level covariances and random-effects initialization worsened performance and is thus not recommended.
+
+Strong performance of clustered covariances was also observed in partitioning real-world academic trajectories.
+They provided substantially smaller trees for all outcomes and better or equal predictive accuracy.
+In comparison to cluster-level covariances, random-effects initialization resulted larger trees for all outcomes and worse performance for the reading and math outcomes.
+
+The simulations showed comparable performance of LM(M) and SEM trees in partitioning LGCMs.
+SEM trees may however be more sensitive to mis-specification of the random effects, with under-specification resulting in too many splits.
+In line with results of \cite{ArnoyVoel21}, we found score-based SEM trees to have somewhat lower power than LRT-based SEM trees, but at a much lower computational cost.
+LongCART trees often selected the wrong partitioning variable for the first split, and were outperformed by LM(M) and SEM trees.
+The LongCART parameter stability tests \citep{KundyHare19} may be underpowered for detecting instability of the fixed intercept, or for detecting instability with respect to categorical covariates.
+
+The simulations clearly illustrated the lower computational burden of GLMM trees.
+This is in large part due to their local-global estimation approach, where fixed-effects parameters are estimated locally within a node and random-effects parameters are estimated globally, using all observations.
+In contrast, SEM trees and LongCART fit the full mixed-effects model in each node, which substantially increases computational load.
+The local-global estimation approach also reduces model complexity, because a lower number of random-effects parameters need to be estimated.
+
+Yet, a possible downside of the local-global estimation approach is that it does not allow for recovering subgroups with differences in random-effects parameters.
+When there is a specific interest in partitioning the random effects, score-based SEM trees may be preferred.
+Alternatively, researchers may want to use the parameter stability tests for mixed-effects models developed by \cite{WangyMerk18} \citep[see also ][]{WangyMerk21,WangyGrav22}.
+This will be useful, for example, when the number of or distances between timepoints differ between respondents and SEM-based growth curve models cannot be applied \citep{NeisyMatt18}.
+
+The current evaluations were limited to Gaussian responses and LGCMs.
+Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses.
+Also, whether our conclusions generalize to settings beyond LGCMs remains to be evaluated.
+We expect that the strong performance of cluster-level covariances in the parameter stability tests generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures.
+Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances.
+Though computationally more burdensome, future work could assess potential benefits of using the full sandwich estimator.
+
+\bibliography{bib}
+
+\newpage
+\appendix
+
+\section{Appendix A: Effects of data-generating parameters on tree size}
+\label{sec:AppendixA}
+
+For LM(M) trees (Figure~\ref{fig:LMM_sizes_interact}), use of cluster-level covariances provided the most robust improvement in split recovery.
+For the data-generating parameters, the effects on tree size were strongest for $\sigma_{b_0}$, followed by $N$, $p$, $\sigma_{b_1}$ and $\rho$.
+Higher values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ tend to yield higher numbers of splits, while the effect of $\rho$ is minimal.
+
+For SEM trees and LongCART (Figure~\ref{fig:tree_sizes_interact}), strongest effects were observed for $N$, followed by $\sigma_{b_0}$, $p_{noise}$, $\rho$ and $\sigma_{b_1}$.
+Results for $N$ were as expected for all methods: More splits are implemented with higher sample size.
+Both LRT- and score-based SEM trees seem only affected by levels of $\sigma_{b_0}$, $p_{noise}$ and $\rho$ under misspecification of the random effects.
+Especially when both random intercept and slope variances are fixed to 0, higher levels of $\sigma_{b_0}$ and $p_{noise}$ yield more splits with both SEM tree approached.
+LRT-based SEM trees seem unaffected by levels of $\rho$, while score-based SEM trees seemed to implement a larger number of splits with increasing values of $\rho$.
+This pattern seemed reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects were correctly specified.
+LongCART implemented more splits with higher levels of $\rho$,  $\sigma_{b_1}$, $N$, $p_{noise}$ and $\sigma_{b_0}$.
+
+\begin{figure}[!ht]
+\caption{Effects of data-generating parameters on tree size for LM(M) trees.}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.7>>=
+theme_set(theme_bw(base_size = 8))
+cols <- rep(rainbow_hcl(10)[c(3,5)], times = 5)
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_int), 
+               position=position_dodge(1), alpha = .5, width = .6, 
+               outlier.shape = NA, coef = NULL) +
+  geom_count(aes(group=sigma_int), position=position_dodge(1), colour = "black", 
+             fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) + 
+  geom_hline(yintercept=3, col = "darkgray") + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(fill=expression(sigma[b[0]]^2)) +
+  stat_summary(aes(group=sigma_int), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = N), 
+               position=position_dodge(1), alpha = .5, width = .6, 
+               outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=N), 
+             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits") +
+  geom_hline(yintercept=3, col = "darkgray") + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  stat_summary(aes(group=N), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = p_noise), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=p_noise), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col = expression(p[noise])) +
+  geom_hline(yintercept=3, col = "darkgray") +
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(p[noise])) +
+  stat_summary(aes(group=p_noise), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_slope), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=sigma_slope), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
+  geom_hline(yintercept=3, col = "darkgray") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(sigma[b[1]]^2)) +
+  stat_summary(aes(group=sigma_slope), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.8>>=
+ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
+  geom_boxplot(aes(fill = rho), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=rho), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
+  geom_hline(yintercept=3, col = "darkgray") + 
+  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(rho)) +
+  stat_summary(aes(group=rho), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
+theme_set(theme_bw(base_size = 11))
+@
+\end{subfigure}%
+\vspace{-.6cm}%
+{\singlespacing \footnotesize \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
+Distances on $y$-axis are on log scale.
+$\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $N$ = sample size at level 2; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
+\label{fig:LMM_sizes_interact}
+\end{figure}
+
+\begin{figure}[!ht]
+\caption{Effect of data-generating parameters on tree size for LM(M), SEM and LongCART trees.}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.7>>=
+lim <- c(0, 18)
+theme_set(theme_bw(base_size = 8))
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = N), position=position_dodge(1), alpha = .5, width = .6,
+               outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=N), 
+             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col=expression(N)) + 
+  geom_hline(yintercept=3, col = "darkgray") + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
+  labs(fill=expression(N)) +
+  stat_summary(aes(group=N), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4) 
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_int), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=sigma_int), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, limits = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) +
+  geom_hline(yintercept=3, col = "darkgray") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(sigma[b[0]]^2)) +
+  stat_summary(aes(group=sigma_int), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = p_noise), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=p_noise), position=position_dodge(1), 
+             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col = expression(p[noise])) +
+  geom_hline(yintercept=3, col = "darkgray") +
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(p[noise])) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  stat_summary(aes(group=p_noise), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.5>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = rho), position=position_dodge(1), 
+               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=rho), position=position_dodge(1), colour = "black", 
+             fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col=expression(rho)) +
+  geom_hline(yintercept=3, col = "darkgray") + 
+  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
+        strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(rho)) + 
+  stat_summary(aes(group=rho), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+@
+\end{subfigure}
+\begin{subfigure}{1.25\textwidth}
+<<echo=FALSE, fig=TRUE, height=2.0>>=
+ggplot(tmp2, aes(x=random, y=tree_size)) +
+  geom_boxplot(aes(fill = sigma_slope), 
+               position=position_dodge(1), alpha = .5, width = .6,
+               outlier.shape = NA, coef=NULL) + 
+  geom_count(aes(group=sigma_slope), 
+             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
+  scale_y_continuous(trans = pseudo_log_trans(), 
+                     breaks = breaks, lim = lim) +
+  facet_grid(~method, scales = "free", space = "free") +
+  labs(x = "", y = "no. of splits", col=expression(sigma[b[1]]^2)) +
+  geom_hline(yintercept=3, col = "darkgray") +
+  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
+  labs(fill=expression(sigma[b[1]]^2)) + 
+  scale_x_discrete(labels = scales::parse_format()) +
+  stat_summary(aes(group=sigma_slope), position=position_dodge(1), fun="mean", 
+               col = "black", shape=16, size = .4)
+theme_set(theme_bw(base_size = 11))
+@
+\end{subfigure}%
+\vspace{-.6cm}%
+{ \singlespacing \footnotesize \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
+Distances on $y$-axis are on log scale.
+$N$ = sample size at level 2; $\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
+\label{fig:tree_sizes_interact}
+\end{figure}
+
+\FloatBarrier
+
+\end{document}
 
@@
 
-<<eval=FALSE, echo=FALSE>>=
-math_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = mathdata)
-(as.data.frame(VarCorr(math_LMM))$vcov / var_math)
-(as.data.frame(VarCorr(math_LMM))$vcov / as.data.frame(VarCorr(math_LMM))$vcov[4])
-
-read_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = readdata)
-(as.data.frame(VarCorr(read_LMM))$vcov / var_read)
-(as.data.frame(VarCorr(read_LMM))$vcov / as.data.frame(VarCorr(read_LMM))$vcov[4])
-
-scie_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = sciedata)
-(as.data.frame(VarCorr(scie_LMM))$vcov / var_scie)
-(as.data.frame(VarCorr(scie_LMM))$vcov / as.data.frame(VarCorr(scie_LMM))$vcov[4])
-@
-
-\begin{figure}[!bt]%
-\caption{Tree sizes for trees fitted to math, reading, and science ability trajectories.}
-\begin{subfigure}{\textwidth}%
-<<echo=FALSE, fig=TRUE, height=2, width=5>>=
-## Tree size
-load("sizes, N = 250, Math.Rda")
-math_sizes <- cbind(sizes, N = 250)
-#load("sizes, N = 1000, Math.Rda")
-#math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
-#sapply(math_sizes, function(x) table(is.na(x)))
-#sapply(math_sizes, function(x) tapply(x, math_sizes$N, mean, na.rm =TRUE))
-
-load("sizes, N = 250, Reading.Rda")
-read_sizes <- cbind(sizes, N = 250)
-#load("sizes, N = 1000, Reading.Rda")
-#read_sizes <- rbind(read_sizes, cbind(sizes, N = 1000))
-#sapply(read_sizes, function(x) table(is.na(x)))
-#sapply(read_sizes, function(x) tapply(x, read_sizes$N, mean, na.rm =TRUE))
-
-load("sizes, N = 250, Science.Rda")
-scie_sizes <- cbind(sizes, N = 250)
-#load("sizes, N = 1000, Science.Rda")
-#scie_sizes <- rbind(scie_sizes, cbind(sizes, N = 1000))
-#sapply(scie_sizes, function(x) table(is.na(x)))
-#sapply(scie_sizes, function(x) tapply(x, scie_sizes$N, mean, na.rm =TRUE))
-
-## Made a mistake in evaluating tree size in the Application scripts:
-## I took #splits = nrow(tree$frame)-1, assuming tree$frame contains a row for
-## each terminal node. It contains a row for each inner and terminal node.
-## Thus, #splits = (nrow(tree$frame)-1)/2; original number should be divided by 2:
-math_sizes$longRPart <- math_sizes$longRPart/2 
-math_sizes$longRPart_s <- math_sizes$longRPart_s/2 
-read_sizes$longRPart <- read_sizes$longRPart/2 
-read_sizes$longRPart_s <- read_sizes$longRPart_s/2 
-scie_sizes$longRPart <- scie_sizes$longRPart/2 
-scie_sizes$longRPart_s <- scie_sizes$longRPart_s/2
-
-## tmp6 is math
-tmp6 <- data.frame(stack(math_sizes[ , -c(4,6,7:10)]),
-                    dataset_id = factor(rep(1:nrow(math_sizes), times = 12)),
-                   N = math_sizes$N)
-names(tmp6)[1:2] <- c("splits", "method") 
-tmp6$method2 <- as.character(tmp6$method)
-tmp6$method2[grepl("LMtree", tmp6$method2)] <- "LM tree"
-tmp6$method2[tmp6$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
-tmp6$method2[tmp6$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
-tmp6$method2[tmp6$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
-tmp6$method2[grepl("longR", tmp6$method2)] <- "longRPart"
-tmp6$random <- ifelse(tmp6$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
-                      ifelse(tmp6$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
-                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp6$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp6$N <- factor(tmp6$N)
-ggplot(tmp6) +
-  geom_boxplot(aes(x=random, y=splits), 
-               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
-  facet_grid(~method2, scales = "free", space = "free") +
-  theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
-        axis.text.y = element_text(size = 8)) +
-  labs(x = "", y = "no. of splits")
-@
-\end{subfigure}
-\begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.6, width=5>>=
-## tmp7 is reading
-tmp7 <- data.frame(stack(read_sizes[ , -c(4,6,7:10)]),
-                    dataset_id = factor(rep(1:nrow(read_sizes), times = 12)),
-                   N = read_sizes$N)
-names(tmp7)[1:2] <- c("splits", "method") 
-tmp7$method2 <- as.character(tmp7$method)
-tmp7$method2[grepl("LMtree", tmp7$method2)] <- "LM tree"
-tmp7$method2[tmp7$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
-tmp7$method2[tmp7$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
-tmp7$method2[tmp7$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
-tmp7$method2[grepl("longR", tmp7$method2)] <- "longRPart"
-tmp7$random <- ifelse(tmp7$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
-                      ifelse(tmp7$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
-                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp7$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp7$N <- factor(tmp7$N)
-ggplot(tmp7) +
-  geom_boxplot(aes(x=random, y=splits), 
-               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
-  facet_grid(~method2, scales = "free", space = "free") +
-  labs(x = "", y = "no. of splits") +
-  theme(strip.background = element_blank(), strip.text.x = element_blank(),
-        axis.title.x=element_blank(), axis.text.x=element_blank(),
-        axis.text.y = element_text(size = 8))
-@
-\end{subfigure}
-\begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=2, width=5>>=
-## tmp8 is science
-tmp8 <- data.frame(stack(scie_sizes[ , -c(4,6,7:10)]),
-                    dataset_id = factor(rep(1:nrow(scie_sizes), times = 12)),
-                   N = scie_sizes$N)
-names(tmp8)[1:2] <- c("splits", "method") 
-tmp8$method2 <- as.character(tmp8$method)
-tmp8$method2[grepl("LMtree", tmp8$method2)] <- "LM tree"
-tmp8$method2[tmp8$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
-tmp8$method2[tmp8$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
-tmp8$method2[tmp8$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
-tmp8$method2[grepl("longR", tmp8$method2)] <- "longRPart"
-tmp8$random <- ifelse(tmp8$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
-                      ifelse(tmp8$method %in% c("LMMtree_c", "LMMtree_r", "LMMtree_cr", "longRPart"), 
-                             "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp8$method %in% c("LMMtree_sc", "LMMtree_scr", "longRPart_s"), 
-                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp8$N <- factor(tmp8$N)
-ggplot(tmp8) +
-  geom_boxplot(aes(x=random, y=splits), 
-               position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
-  facet_grid(~method2, scales = "free", space = "free") +
-  theme(strip.background = element_blank(), strip.text.x = element_blank(),
-        axis.text.y = element_text(size = 8)) +
-  labs(x = "", y = "no. of splits") +
-  scale_x_discrete(labels = scales::parse_format()) +
-  coord_cartesian(clip = 'off')
-@
-\end{subfigure}%
-{\\\footnotesize \textit{Note.} Top, middle and bottom panels depict math, reading and science ability trajectories, respectively.}
-\label{fig:application_sizes}
-\end{figure}%
-
 %\begin{figure}[h]
@@
 <<eval=FALSE, echo=FALSE, fig=TRUE, height=3.25, width=3>>=
-# load("times, N = 250, Reading.Rda")
-# read_times <- cbind(times, N = 250)
-# load("times, N = 1000, Reading.Rda")
-# read_times <- rbind(read_times, cbind(times, N = 1000))
-# 
-# load("times, N = 250, Math.Rda")
-# math_times <- cbind(times, N = 250)
-# load("times, N = 1000, Math.Rda")
-# math_times <- rbind(math_times, cbind(times, N = 1000))
-# 
-# load("times, N = 250, Science.Rda")
-# scie_times <- cbind(times, N = 250)
-# load("times, N = 1000, Science.Rda")
-# scie_times <- rbind(scie_times, cbind(times, N = 1000))
-# 
-# 
-# ## Combine all LMM trees
-# ECLSK_times <- rbind(stack(math_times[ , -c(6, 10)]), 
-#                      stack(read_times[ , -c(6, 10)]),
-#                      stack(scie_times[ , -c(6, 10)]))
-# ECLSK_times$ind <- as.character(ECLSK_times$ind)
-# ECLSK_times$ind[grepl("LMMtree", ECLSK_times$ind)]<- "LMM tree"
-# ECLSK_times$ind[grepl("longRPart", ECLSK_times$ind)]<- "longRPart"
-# ECLSK_times$ind[ECLSK_times$ind == "LMtree_c"] <- "LM tree"
-# ECLSK_times$ind <- factor(ECLSK_times$ind)
-# 
-# ## Compute summary stats
-# mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
-# summ_stats <- data.frame(M = sprintf(mean_times, fmt = '%#.2f'),
-#                          ind = names(mean_times),
-#                          y = 11000)
-# ## Plot
+load("times, N = 250, Reading.Rda")
+read_times <- cbind(times, N = 250)
+load("times, N = 1000, Reading.Rda")
+read_times <- rbind(read_times, cbind(times, N = 1000))
+
+load("times, N = 250, Math.Rda")
+math_times <- cbind(times, N = 250)
+load("times, N = 1000, Math.Rda")
+math_times <- rbind(math_times, cbind(times, N = 1000))
+
+load("times, N = 250, Science.Rda")
+scie_times <- cbind(times, N = 250)
+load("times, N = 1000, Science.Rda")
+scie_times <- rbind(scie_times, cbind(times, N = 1000))
+
+## Combine all LMM trees
+ECLSK_times <- rbind(stack(math_times[ , -c(6, 10)]),
+                     stack(read_times[ , -c(6, 10)]),
+                     stack(scie_times[ , -c(6, 10)]))
+ECLSK_times$ind <- as.character(ECLSK_times$ind)
+ECLSK_times$ind[grepl("LMMtree", ECLSK_times$ind)]<- "LMM tree"
+ECLSK_times$ind[grepl("longRPart", ECLSK_times$ind)]<- "longRPart"
+ECLSK_times$ind[ECLSK_times$ind == "LMtree_c"] <- "LM tree"
+ECLSK_times$ind <- factor(ECLSK_times$ind)
+
+## Compute summary stats
+mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
+#summ_stats <- data.frame(M = sprintf(mean_times, fmt = '%#.2f'),
+#                         ind = names(mean_times), y = 11000)
+## Plot
 # ggplot(ECLSK_times) +
@@
 %\end{figure}
-
-Figure~\ref{fig:application_sizes} presents tree size distributions.
-Random-effects initialization tends to yield the largest trees, while LM trees with clustered covariances tend to yield the smallest trees.
-LMM trees with clustered covariances yield tree sizes in between.%Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees.
-For math and reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small.
-This pattern reverses with large sample size, where the best-performing LM(M) tree approach yields a larger number of splits than the best-performing longRPart approach.
-For the math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
-
-%\subsubsection{Computation time}
-
-%Figure~\ref{fig:application_times} depicts computation time distributions of the different partitioning algorithms.
-A clear computational advantage is observed for LM trees ($M =$ \Sexpr{#round(mean_times["LM tree"], digits = 2)} seconds).
-LMM trees require longer computation times ($M =$ \Sexpr{#round(mean_times["LMM tree"], digits = 2)} seconds).
-This increase is minor compared to LongRPart trees, which required longest computation times %($M =$ \Sexpr{#round(mean_times["longRPart"], digits = 2)} seconds).
-
-\FloatBarrier
-
-\section{Discussion}
-
-The Simulations in Study~I showed that the proposed extensions of GLMM trees are effective for partitioning linear growth curve models.
-Use of clustered covariances in the parameter stability tests mitigated the strong overfitting that was observed using the standard covariances.
-Initializing estimation with the random effects, instead of assuming these to be zero was also successfull in mitigating overfitting, but only when the random-effects specification comprised only a random intercept.
-The performance of trees fitted using clustered covariances appeared largely unaffected by (mis-)specification of the random effects; use of clustered covariances may therefore be the safest choice in practice.
-Combining cluster-level covariances and random-effects initialization worsened performance and thus not recommended.
-The results also indicated that the optimal estimation approach for LMM trees does not depend on sample size, variance of the random effect(s), the number of potential partitioning variables or the intercorrelation between partioning variables.
-
-Study~II showed comparable performance of LM(M) and SEM trees in partitioning LGCMs.
-SEM trees appeared more sensitive to mis-specification of the random effects, with under-specification yielding too many splits being implemented.
-In line with results of \cite{ArnoyVoel21}, we found score-based SEM trees to have somewhat lower power than LRT-based SEM trees, but with much lower computational burgen.
-LongCART trees often selected the wrong partitioning variable for the first split, and were outperformed by LM(M) and SEM trees.
-The LongCART parameter stability tests \citep{KundyHare19} may be underpowered for detecting instability of the fixed intercept, or for detecting instability with respect to categorical covariates.
-As in Study~I, these results were not strongly affected by sample size, variance of the random effect(s), the number of potential partitioning variables or the intercorrelation between partioning variables.
-
-Study~III assessed performance of LM(M) trees in partitioning children's trajectories of math, reading and science abilities.
-Similar to the results of the simulation studies, we found very similar performance between LM(M) trees which employed clustered covariances but with different random effects specifications, though best performance was observed for a random-effects specification comprising only a random intercept.
-However, when partitioning science trajectories, for which there were a smaller number of observations per child, initializing estimation with the random effects showed the best performance.
-A possible explanation is that the higher the number of observations per subject, the more the clustered covariances will differ from covariances assuming independent observations, creating greater potential for improving the parameter stability tests.
-With a smaller number of observations per subject, initializing estimation with the random effect may be more beneficial.
-However, initializing estimation with the random effects also resuled in a substantially higher number of splits in all three datasets.
-
-As expected, Study~II revealed a clear computational advantage for LM(M) trees.
-This is in large part due to the local-global estimation approach employed by LMM trees, where fixed-effects parameters are estimated locally within a node and random-effects parameters are estimated globally, using all observations.
-In contrast, SEM trees and LongCART fit the full mixed-effects model in each node, which substantially increases computational load.
-The local-global estimation approach also reduces model complexity, because a lower number of random-effects parameters need to be estimated.
-The good performance of LMM trees in the current study suggest the lower complexity may also yield better stability and generalizability.
-
-Yet, a possible downside of the local-global estimation approach is that it does not allow for recovering subgroups with differences in random-effects parameters.
-When there is a specific interest in partitioning the random effects, score-based SEM trees may be preferred.
-Alternatively, researchers may want to use the parameter stability tests for mixed-effects models developed by \cite{WangyMerk18} \citep[see also ][]{WangyMerk21,WangyGrav22}.
-This will be useful, for example, when the number and/or distances between timepoints differs between respondents and SEM-based growth curve models can thus not be applied \citep{NeisyMatt18}.
-
-The current evaluations were limited to Gaussian responses and LGCMs.
-Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses.
-We expect that the good performance of clustered covariances in the parameter stability tests generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures.
-Yet, whether our conclusions generalize to settings beyond LGCMs remains to be evaluated.
-Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances.
-Though computationally more burdensome, future studies could assess whether use of the full sandwich estimator would further improve performance.
-
-\bibliography{bib}
-
-\newpage
-\appendix
-
-\section{Appendix A: Effects of data-generating parameters on tree size}
-\label{sec:AppendixA}
-
-For LM(M) trees (Figure~\ref{fig:LMM_sizes_interact}), use of cluster-level covariances provided the most robust improvement in split recovery.
-For the data-generating parameters, the effects on tree size were strongest for $\sigma_{b_0}$, followed by $N$, $p$, $\sigma_{b_1}$ and $\rho$.
-Higher values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ tend to yield higher numbers of splits, while the effect of $\rho$ is minimal.
-
-For SEM trees and LongCART (Figure~\ref{fig:tree_sizes_interact}), strongest effects were observed for $N$, followed by $\sigma_{b_0}$, $p_{noise}$, $\rho$ and $\sigma_{b_1}$.
-Results for $N$ were as expected for all methods: More splits are implemented with higher sample size.
-Both LRT- and score-based SEM trees seem only affected by levels of $\sigma_{b_0}$, $p_{noise}$ and $\rho$ under misspecification of the random effects.
-Especially when both random intercept and slope variances are fixed to 0, higher levels of $\sigma_{b_0}$ and $p_{noise}$ yield more splits with both SEM tree approached.
-LRT-based SEM trees seem unaffected by levels of $\rho$, while score-based SEM trees seemed to implement a larger number of splits with increasing values of $\rho$.
-This pattern seemed reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects were correctly specified.
-LongCART implemented more splits with higher levels of $\rho$,  $\sigma_{b_1}$, $N$, $p_{noise}$ and $\sigma_{b_0}$.
-
-\begin{figure}[!ht]
-\caption{Effects of data-generating parameters on tree size for LM(M) trees.}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.7>>=
-theme_set(theme_bw(base_size = 8))
-cols <- rep(rainbow_hcl(10)[c(3,5)], times = 5)
-ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
-  geom_boxplot(aes(fill = sigma_int), 
-               position=position_dodge(1), alpha = .5, width = .6, 
-               outlier.shape = NA, coef = NULL) +
-  geom_count(aes(group=sigma_int), position=position_dodge(1), colour = "black", 
-             fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
-  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) + 
-  geom_hline(yintercept=3, col = "darkgray") + 
-  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
-  labs(fill=expression(sigma[b[0]]^2)) +
-  stat_summary(aes(group=sigma_int), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4) 
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
-  geom_boxplot(aes(fill = N), 
-               position=position_dodge(1), alpha = .5, width = .6, 
-               outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=N), 
-             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
-  labs(x = "", y = "no. of splits") +
-  geom_hline(yintercept=3, col = "darkgray") + 
-  theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  stat_summary(aes(group=N), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4) 
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
-  geom_boxplot(aes(fill = p_noise), position=position_dodge(1), 
-               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=p_noise), position=position_dodge(1), 
-             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
-  labs(x = "", y = "no. of splits", col = expression(p[noise])) +
-  geom_hline(yintercept=3, col = "darkgray") +
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(p[noise])) +
-  stat_summary(aes(group=p_noise), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4) 
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
-  geom_boxplot(aes(fill = sigma_slope), position=position_dodge(1), 
-               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=sigma_slope), position=position_dodge(1), 
-             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
-  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
-  geom_hline(yintercept=3, col = "darkgray") + 
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(sigma[b[1]]^2)) +
-  stat_summary(aes(group=sigma_slope), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4)
-
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.8>>=
-ggplot(tmp[LMM_ids, ], aes(x=setting, y=tree_size)) +
-  geom_boxplot(aes(fill = rho), position=position_dodge(1), 
-               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=rho), position=position_dodge(1), 
-             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  facet_grid(~ranef, scales = "free", space = "free", labeller = label_parsed) +
-  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[1]))) +
-  geom_hline(yintercept=3, col = "darkgray") + 
-  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(rho)) +
-  stat_summary(aes(group=rho), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4) 
-theme_set(theme_bw(base_size = 11))
-@
-\end{subfigure}%
-\vspace{-.6cm}%
-{\singlespacing \footnotesize \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
-Distances on $y$-axis are on log scale.
-$\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $N$ = sample size at level 2; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
-\label{fig:LMM_sizes_interact}
-\end{figure}
-
-\begin{figure}[!ht]
-\caption{Effect of data-generating parameters on tree size for LM(M), SEM and LongCART trees.}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.7>>=
-lim <- c(0, 18)
-theme_set(theme_bw(base_size = 8))
-ggplot(tmp2, aes(x=random, y=tree_size)) +
-  geom_boxplot(aes(fill = N), position=position_dodge(1), alpha = .5, width = .6,
-               outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=N), 
-             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "no. of splits", col=expression(N)) + 
-  geom_hline(yintercept=3, col = "darkgray") + 
-  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
-  labs(fill=expression(N)) +
-  stat_summary(aes(group=N), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4) 
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp2, aes(x=random, y=tree_size)) +
-  geom_boxplot(aes(fill = sigma_int), position=position_dodge(1), 
-               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=sigma_int), position=position_dodge(1), 
-             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, limits = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "no. of splits", col=expression(sigma^2~(b[0]))) +
-  geom_hline(yintercept=3, col = "darkgray") + 
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(sigma[b[0]]^2)) +
-  stat_summary(aes(group=sigma_int), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4)
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp2, aes(x=random, y=tree_size)) +
-  geom_boxplot(aes(fill = p_noise), position=position_dodge(1), 
-               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=p_noise), position=position_dodge(1), 
-             colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "no. of splits", col = expression(p[noise])) +
-  geom_hline(yintercept=3, col = "darkgray") +
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(p[noise])) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  stat_summary(aes(group=p_noise), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4)
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.5>>=
-ggplot(tmp2, aes(x=random, y=tree_size)) +
-  geom_boxplot(aes(fill = rho), position=position_dodge(1), 
-               alpha = .5, width = .6, outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=rho), position=position_dodge(1), colour = "black", 
-             fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), breaks = breaks, lim = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "no. of splits", col=expression(rho)) +
-  geom_hline(yintercept=3, col = "darkgray") + 
-  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),
-        strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(rho)) + 
-  stat_summary(aes(group=rho), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4)
-@
-\end{subfigure}
-\begin{subfigure}{1.25\textwidth}
-<<echo=FALSE, fig=TRUE, height=2.0>>=
-ggplot(tmp2, aes(x=random, y=tree_size)) +
-  geom_boxplot(aes(fill = sigma_slope), 
-               position=position_dodge(1), alpha = .5, width = .6,
-               outlier.shape = NA, coef=NULL) + 
-  geom_count(aes(group=sigma_slope), 
-             position=position_dodge(1), colour = "black", fill=NA, show.legend = FALSE, alpha = .15) +
-  scale_y_continuous(trans = pseudo_log_trans(), 
-                     breaks = breaks, lim = lim) +
-  facet_grid(~method, scales = "free", space = "free") +
-  labs(x = "", y = "no. of splits", col=expression(sigma[b[1]]^2)) +
-  geom_hline(yintercept=3, col = "darkgray") +
-  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
-  labs(fill=expression(sigma[b[1]]^2)) + 
-  scale_x_discrete(labels = scales::parse_format()) +
-  stat_summary(aes(group=sigma_slope), position=position_dodge(1), fun="mean", 
-               col = "black", shape=16, size = .4)
-theme_set(theme_bw(base_size = 11))
-@
-\end{subfigure}%
-\vspace{-.6cm}%
-{ \singlespacing \footnotesize \textit{Note.} Black dots represent means, gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
-Distances on $y$-axis are on log scale.
-$N$ = sample size at level 2; $\sigma_{b_0}^2$ = variance of random intercept; $\sigma_{b_1}^2$ = variance of random slope; $p_{noise}$ = number of noise variables; $\rho$ = correlation between partitioning variables.}
-\label{fig:tree_sizes_interact}
-\end{figure}
-
-\FloatBarrier
-
-\end{document}
 

### 2023-09-08 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
LGCM paper updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3276 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bb13f5e03206caef93d0ccd9295ce5e69cdf2d4a

@@
 For children with intermediate levels of SES (node~6), there is also an effect of internalizing problems: The group with lower internalizing problems has a higher intercept.
-Also, the two groups (or nodes) with higher intercepts have higher slopes, indicating that children with relatively higher ability may also show larger gains in ability over time.
+The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher ability also gain more ability over time.
 
@@
 \label{subfig:tree}
-\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
-Gray lines depict observed individual trajectories.
-Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.}
 \end{subfigure}%
@@
 \caption{GLMM tree estimated by initializing with zero random effects.}
-<<fig=TRUE, height=8, width=14, echo=FALSE>>=
+\begin{subfigure}{1.2\textwidth}
+<<fig=TRUE, height=9, width=16, echo=FALSE>>=
 lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
@@
 @
+\end{subfigure}
 \\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability.
@@
 
-In the next sections, we will assess performance of the original algorithm and proposed adaptations through more extensive empirical evaluations.
+In the next sections, we assess performance of the original algorithm and proposed adaptations through more extensive empirical evaluations.
 

### 2023-09-12 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Paper updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3278 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e88d07fc867a04fa9afe5e1b7b944993f259565e

@@
 Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline.
-Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope %(presented in Table~\ref{tab:local_coefs}).
-Groups of children with lower (higher) SES also have a lower (higher) intercept, indicating lower (higher) average science ability.
-For children with lower SES (node~2), there is also an effect of gross motor skills: The group with lower motor skills has a lower intercept.
-For children with intermediate levels of SES (node~6), there is also an effect of internalizing problems: The group with lower internalizing problems has a higher intercept.
+Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope.
+Groups of children with higher SES also have higher intercepts, indicating higher average science ability.
+The group of children with lower SES (node~2) is further split based on gross motor skills, with higher motor skills resulting in a higher intercept.
+The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept.
 The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher ability also gain more ability over time.
@@
 Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.
+The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories.
 The table presents numerical estimates of fixed intercepts and slopes.}
@@
 In terms of MAD, however, LMM trees initializing estimation with the random effects performed best, but only if the random slope was not estimated.
-Initializing estimation with the random effects may only be useful if the random-effects specification is kept relatively simple.
-With more complex random-effects specifications, initializing estimation seems to result in too few groups being detected; subgroup differences are then more likely to be captured by the random effects than by the tree structure.
+Thus, initializing estimation with the random effects may only be useful if the random-effects specification is kept relatively simple.
+With more complex random-effects specifications, actual subgroup differences may be more likely captured by the random effects than by the tree structure.
 The second-lowest MAD was observed for LM trees with clustered covariances (left panel).
@@
 The results show a pattern very similar to Figure~\ref{fig:LMM_sizes}; no substantial interactions between data-generating and model-specification parameters were observed.
-Main effects of the data-generating parameters were as expected: The strongest effects were for $\sigma_{b_1}$ and $N$, with higher values resulting in a higher number of splits.
+Main effects of the data-generating parameters were as expected: Strongest effects were for $\sigma_{b_1}$ and $N$, with higher values resulting in a higher number of splits.
 The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, while the correlation between partitioning variables hardly affected the number of implemented splits.
 
-Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split: Only very rarely is $u_1$ not selected for the first split.
+Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split for all LM(M) tree fitting approaches.
+Only very rarely is $u_1$ not selected for the first split, if a first split was implemented.
 
@@
 
-We applied five LM(M) trees to the data, focusing on the original GLMM tree approach and the approaches that performed well in the simulations:
+We applied five LM(M) trees to the data, focusing on the original GLMM tree approach and approaches that performed well in the simulations:
 % 
@@
   \item An LMM tree using clustered covariances, with both random intercept and slope freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$) and initialization assuming zero random effects .
-  \item An LMM tree uwith a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and random-effect initialization with the full-sample estimate. 
+  \item An LMM tree using observation-level covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and random-effect initialization with the full-sample estimate.
 \end{itemize}
 %
-Although LRT-based SEM trees performed very well in the simulations, they could not be used in this study because growth curve SEMs do not allow for incorporating continuous time.
+Although LRT- and score-based SEM trees performed very well in the simulations, they could not be used in this study because growth-curve SEMs do not allow for incorporating continuous time.
 
@@
 Bonferroni-adjusted pairwise $t$-tests indicated no difference in performance between default LMM trees and LMM trees with random-effects initialization for any of the three outcomes.
-Incontract, the three LM(M) trees using clustered covariances performed significantly better for the math and reading outcomes.
+In contrast, the three LM(M) trees using clustered covariances performed significantly better for the math and reading outcomes.
 For the science outcomes, no significant differences were observed.
@@
 Given their non-distinguishable predictive performance, LM trees with clustered covariances may be preferred if a obtaining a sparse result is critical.
-LMM trees with clustered covariances may provide less sparsity but better predictive accuracy, irrespective of whether random intercepts and/or slopes have been specified.
-Random-effects initialization yields tree sizes similar to the default fitting approach.
+LMM trees with clustered covariances may provide less sparsity but better predictive accuracy, irrespective of whether random intercepts and/or slopes were estimated.
+In contrast to the simulation study results, random-effects initialization here yields tree sizes similar to the default fitting approach.
 
@@
 They provided substantially smaller trees for all outcomes and better or equal predictive accuracy.
-In comparison to cluster-level covariances, random-effects initialization resulted larger trees for all outcomes and worse performance for the reading and math outcomes.
+In comparison to cluster-level covariances, random-effects initialization resulted in larger trees for all outcomes and worse performance for the reading and math outcomes.
 
@@
 Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses.
-Also, whether our conclusions generalize to settings beyond LGCMs remains to be evaluated.
-We expect that the strong performance of cluster-level covariances in the parameter stability tests generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures.
+We expect that the strong performance of cluster-level covariances generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures, but whether our conclusions generalize to settings beyond LGCMs remains to be evaluated.
 Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances.

### 2024-02-03 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Paper revisions for BRM
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3279 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ea0ca6a39dd2e0dafa4274567df49d2023c9ee7d

@@
 \authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}
+\authornote{
+\textbf{Declarations}
+
+Funding: Not applicable.
+Conflicts of interest: Not applicable.
+Ethics approval: Not applicable, simulated and publicly available data from the ECLS-K study were analysed.
+Consent to participate: Not applicable.
+Consent for publication: Not applicable.
+Availability of data and materials: All methods used are implemented in R and associated packages, as described in the Method section.
+Code availability: Data and R code are available as supplementary materials.
+None of the experiments were preregistered.}
 
@@
 Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity.
-We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with differently shaped trajectories in linear growth curve models.
+We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with different trajectories in linear growth curve models.
 Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data.
@@
 LMM <- lmer(score ~ months + (1|CHILDID), data = data)
-icc <- data.frame(VarCorr(LMM))
 @
@@
 Mixed-effects or latent-variable models can be used to model such trajectories and allow for explaining heterogeneity with covariates of a-priori known relevance \citep[e.g., ][]{NeisyMatt18}.
-However, when these covariates are not known in advance, methods for identifying them are needed.
+However, when these covariates or their shape of association with the response are not known in advance, methods for identifying them are needed.
 
 As an example, trajectories of science knowledge and skills from a sample of 250 children are depicted in Figure~\ref{fig:global_trajectories}.
-The children were assessed at three timepoints across grades 3 through 8\footnote{Further details on the source of these data are provided in Study~III.}.
+The children were assessed at three timepoints across grades 3 through 8\footnote{Further details on the source of these data and the scaling of time are provided in Study~III.}.
 The red line depicts the estimated average trajectory, while the gray lines depict individual trajectories.
@@
 \\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.
-The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories.}
+The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here.
+To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
 \end{subfigure}%
@@
 Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline.
-Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope.
+Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope, which are presented in Table~\ref{tab:local_coefs}.
 Groups of children with higher SES also have higher intercepts, indicating higher average science ability.
@@
 The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept.
-The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher ability also gain more ability over time.
+The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher baseline ability also gain more ability over time.
 
@@
 \caption{Partitioned growth curves of science ability.}
-\begin{subfigure}{0.9\textwidth}%
+\begin{subfigure}{1\textwidth}%
 <<fig=TRUE, height=9, width=13.5, echo=FALSE>>=
@@
 lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
-icc_r <- as.data.frame(VarCorr(lmm_tree_r))
-lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
-icc_c <- as.data.frame(VarCorr(lmm_tree_c))
-lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
-lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)
 
@@
 \end{subfigure}%
-~
-\begin{subfigure}{0.25\textwidth}%
-\vspace{-2.95cm}
-\hspace{-2.5cm}
-\scriptsize
+%~
+%\begin{subfigure}{0.25\textwidth}%
+%\vspace{-2.95cm}
+%\hspace{-2.5cm}
+%\scriptsize
+<<echo=FALSE, results=tex, fig.pos='b'>>=
+# library("kableExtra")
+# coefs <- coef(lmm_tree_r)
+# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
+# colnames(coefs) <- c("Node", "Intercept", "Slope")
+# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
+# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
+# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
+#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
+# kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
+#         row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
+@
+%\label{subfig:coefs}
+%\end{subfigure}
+\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
+Gray lines depict observed individual trajectories.
+Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.
+The $x$-axis is not linear in the number of months, because time was scaled as $(\text{months after baseline})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here.
+To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
+\label{fig:LMM_tree_r}
+\end{figure}%
+
 <<echo=FALSE, results=tex, fig.pos='b'>>=
@@
                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
-kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
-        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
-@
-\label{subfig:coefs}
-\end{subfigure}
-\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
-Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.
-The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories.
-The table presents numerical estimates of fixed intercepts and slopes.}
-\label{fig:LMM_tree_r}
-\end{figure}%
-
-<<echo=FALSE, results=tex, fig.pos='b'>>=
-# library("kableExtra")
-# coefs <- coef(lmm_tree_r)
-# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
-# colnames(coefs) <- c("Node", "Intercept", "Slope")
-# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
-# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
-# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
-#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
-# kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs",
-#                              align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
-#         caption = "Fixed-effects coefficient estimates from the terminal nodes of Figure 2.",
-#         row.names = FALSE, digits = 3L),
-#   font_size = 11, full_width=FALSE)
+kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs",
+                    align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
+                    caption = "Estimated fixed-effects coefficients for the terminal nodes of Figure 2.",
+                    row.names = FALSE, digits = 3L),
+              font_size = 11, full_width=FALSE, position = "left")
 @
@@
 Other methods for partitioning LGCMs take a fully local estimation approach: Within every node or subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that subgroup only.
-This fully local estimation approach provides more flexibility, but yields higher computational burden and model complexity.
+This fully local estimation approach provides more flexibility, but also higher computational burden and model complexity.
 In contrast, GLMM trees estimate a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results.
@@
 %
-Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon$.
+
+Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon^2$.
 Therefore, the identity function can be used for the link $g$ but generalizations to other response variable types within the GLM are straightforward.
-In the general notation above, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject~$i$ ($i = {1, \dots, N}$), comprising $p$ regressors plus one column of 1s for the intercept.
-In the following, we assume that time is the predictor variable of interest (i.e., $p = 1$), where the number $n_i$ and spacing of observed timepoints may differ between subjects.
-The fixed-effects parameters $\beta_j$ (here, intercept and time slope) in GLMM trees are node-specific, i.e., their value depends on the subgroup/node~$j$ into which subject~$i$ falls.
-As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is the corresponding vector of random effects for subject $i$.
+In the general notation above, $X_i$ is the $n_i \times (p + 1)$ fixed-effects design matrix for subject i $(i = 1, \dots, N)$, comprising $p$ regressors plus one column of 1s for the intercept.
+In this paper, we assume that time is the only time-varying covariate of interest (i.e., $p = 1$), but other time-varying covariates can also be included in $X_i$.
+The number $n_i$ and spacing of observed timepoints may differ between subjects.
+Ttime-constant covariates are not contained in $X_i$, but enter the model through the subgroup indicator $j$.
+The value of the fixed-effects parameters $\beta$ (here, intercept and time slope) in GLMM trees are node-specific, their value depending on the subgroup/node $j$ into which subject $i$ falls.
+Subgroup membership is defined by the values of time-constant covariates\footnote{In principle, it is also possible to use time-varying covariates for partitioning with GLMM trees, but this might not be appropriate for partitioning growth curves, as participants changing subgroups over time yields discontinuous growth curves.}.
+
+As in the standard GLMM \citep[e.g., ][]{PinhyBate00}, GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of Xi, and bi is the corresponding vector of random effects for subject $i$.
 Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.
@@
             xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
-icc <- as.data.frame(VarCorr(lmm_tree))
-lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
+vcov_d <- as.data.frame(VarCorr(lmm_tree))
 @
@@
 The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in  Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills.
-Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples.
+Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples\footnote{Generalizability will be empirically evaluated using cross validation in Study~III.}.
 
@@
 The tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests.
-Application of GLMM trees with both cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates.
-Compared to Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, the cluster-level parameter stability tests provided the most parsimonious tree structure thus far.
-The estimated variances of the random intercept were \Sexpr{formatC(data.frame(VarCorr(lmm_tree_r$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree_r}, \Sexpr{formatC(data.frame(VarCorr(lmm_tree$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree}) and \Sexpr{formatC(data.frame(VarCorr(lmm_tree_c$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree_c}, indicating that the more parsimonious the tree structure, the more variance will be captured by the random effects.
-
-In the next sections, we assess performance of the original algorithm and proposed adaptations through more extensive empirical evaluations.
+Combining cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates.
+Compared to the trees in Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, the cluster-level parameter stability tests provided the most parsimonious tree structure thus far.
+Table~\ref{tab:ranef} presents the estimated random-effects parameters for the models fitted thus far.
+Not surprisingly, the estimated residual standard deviation ($\hat{\sigma}_\epsilon$) differs only very slightly between models.
+Yet, substantial differences can be seen for the standard deviation of the random intercept ($\hat{\sigma}_\epsilon$): With more subgroups or terminal nodes, less variance is captured by the random effects and more variance is captured by the tree structure.
+
+<<echo=FALSE>>=
+vcov_d <- as.data.frame(VarCorr(lmm_tree))
+vcov_r <- as.data.frame(VarCorr(lmm_tree_r))
+vcov_c <- as.data.frame(VarCorr(lmm_tree_c))
+vcv <- as.data.frame(VarCorr(LMM))
+tab <- data.frame(Figure = 1:4,
+                  Model = c("LMM", rep("LMM tree", times = 3)), 
+                  Estimation = c("default", "random effects initialization", 
+                                 "default", "clustered covariances"),
+                  n_subs = c(1L, (length(lmm_tree_r$tree)+1)/2, 
+                             (length(lmm_tree$tree)+1)/2, (length(lmm_tree_c$tree)+1)/2))
+tab$sig_b <- c(vcv[vcv$grp == "CHILDID", "sdcor"], vcov_r[vcov_r$grp == "CHILDID", "sdcor"], 
+               vcov_d[vcov_d$grp == "CHILDID", "sdcor"], vcov_c[vcov_c$grp == "CHILDID", "sdcor"])
+tab$sig_epsb <- c(vcv[vcv$grp == "Residual", "sdcor"], vcov_r[vcov_r$grp == "Residual", "sdcor"], 
+               vcov_d[vcov_d$grp == "Residual", "sdcor"], vcov_c[vcov_c$grp == "Residual", "sdcor"])
+tab$pool_sd <- c(sqrt(sum(vcv$vcov)), sqrt(sum(vcov_r$vcov)),
+                 sqrt(sum(vcov_d$vcov)), sqrt(sum(vcov_c$vcov)))
+ds <- dist(coef(lmm_tree_r)[ , "(Intercept)"], method = "manhattan") / tab$pool_sd[tab$Figure == 1L]
+@
+
+The variances in Table~\ref{tab:ranef} can also be used to quantify variability associated with the trajectories in the terminal nodes of the trees.
+The third column of Table~\ref{tab:ranef} presents pooled standard deviations that can be computed from variance estimates \citep{WestyKenn14}.
+These allow, for example, to gauge the magnitude of subgroup differences in the subgroup-specific intercepts in Table~\ref{tab:local_coefs}, which range from $d = \Sexpr{round(min(ds), digits = 2)}$ to $d = \Sexpr{round(max(ds), digits = 2)}$.
+Yet, these estimates are based on a modest sample size, and are based on training data only, so can be overly optimistic.
+In the next sections, we evaluate performance on simulated (Studies~I~and~II) and real data (Study~III).
+
+<<echo=FALSE, results=tex, fig.pos='b'>>=
+vcov_d <- as.data.frame(VarCorr(lmm_tree))
+vcov_r <- as.data.frame(VarCorr(lmm_tree_r))
+vcov_c <- as.data.frame(VarCorr(lmm_tree_c))
+vcv <- as.data.frame(VarCorr(LMM))
+tab <- data.frame(Figure = 1:4,
+                  Model = c("LMM", rep("LMM tree", times = 3)), 
+                  Estimation = c("default", "random effects initialization", 
+                                 "default", "clustered covariances"),
+                  n_subs = c(1L, (length(lmm_tree_r$tree)+1)/2, 
+                             (length(lmm_tree$tree)+1)/2, (length(lmm_tree_c$tree)+1)/2))
+tab$sig_b <- c(vcv[vcv$grp == "CHILDID", "sdcor"], vcov_r[vcov_r$grp == "CHILDID", "sdcor"], 
+               vcov_d[vcov_d$grp == "CHILDID", "sdcor"], vcov_c[vcov_c$grp == "CHILDID", "sdcor"])
+tab$sig_epsb <- c(vcv[vcv$grp == "Residual", "sdcor"], vcov_r[vcov_r$grp == "Residual", "sdcor"], 
+               vcov_d[vcov_d$grp == "Residual", "sdcor"], vcov_c[vcov_c$grp == "Residual", "sdcor"])
+tab$pool_sd <- c(sqrt(sum(vcv$vcov)), sqrt(sum(vcov_r$vcov)),
+                 sqrt(sum(vcov_d$vcov)), sqrt(sum(vcov_c$vcov)))
+colnames(tab)[4:7] <- c("# of nodes", "$\\hat{\\sigma}_b$", "$\\hat{\\sigma}_\\epsilon$", "$\\hat{\\sigma}_{\\text{pooled}}$")
+kable_styling(
+  kable(tab, format = "latex", linesep = "", row.names = FALSE, 
+        label = "ranef", digits = 3, align = c("cllccc"), booktabs = TRUE, escape =FALSE, 
+        caption = "Estimated random-effects parameters for fitted mixed-effects models in Figures 1 through 4."),
+  position = "left")
+@
+
+<<echo=FALSE, eval=FALSE>>=
+test_dat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
+test_dat$CHILDID <- factor(test_dat$CHILDID)
+test_dat$GENDER <- factor(test_dat$GENDER)
+test_dat <- test_dat[test_dat$RACE != 6, ]
+lmmt_preds <- predict(lmm_tree, re.form = NA, newdata = testdat)
+lmmt_c_preds <- predict(lmm_tree_c, re.form = NA, newdata = testdat)
+lmmt_r_preds <- predict(lmm_tree_r, re.form = NA, newdata = testdat)
+mean((test_dat$score - lmmt_preds)^2)
+mean((test_dat$score - lmmt_r_preds)^2)
+mean((test_dat$score - lmmt_c_preds)^2)
+var(test_dat$score)
+lmm_preds <- predict(LMM, newdata = test_dat, re.form = NA)
+mean((test_dat$score - lmm_preds)^2)
+@
 
@@
 \section{Study~I: Assessment of Subgroup Recovery}
+\label{sec:studyI}
 
@@
 %
+
+<<echo=FALSE, results=hide>>=
+#library("devtools")
+#install_github("brandmaier/semper")
+library("semper")
+
+## low to medium slope reliability for the model with the 
+## small random slope:
+
+## low sigma_b0 and low sigma_b1 
+model1 <- lgcm(timepoints=0:4, slope.variance = 0.1,
+              intercept.slope.covariance = 0, intercept.variance = 1,
+              residual.variance = 5)
+ecr(model1)
+##[1] 0.2857143
+
+## high sigma_b0 and low sigma_b
+model2 <- lgcm(timepoints=0:4, slope.variance = 0.1,
+              intercept.slope.covariance = 0, intercept.variance = 4,
+              residual.variance = 5)
+ecr(model2)
+##[1] 0.21875
+
+## medium to good slope reliability for the condition with large slope variance:
+model3 <- lgcm(timepoints=0:4, slope.variance = 0.4, intercept.slope.covariance = 0,
+              intercept.variance = 1, residual.variance = 5)
+ecr(model3)
+##[1] 0.6153846
+
+model4 <- lgcm(timepoints=0:4, slope.variance = 0.4, intercept.slope.covariance = 0,
+              intercept.variance = 4, residual.variance = 5)
+ecr(model4)
+##[1] 0.5283019
+@
+
+Chosen values of $\sigma_{b_0}$ are reflective of intra-class correlations of \Sexpr{round(1/(1+5), digits = 3)} and \Sexpr{round(4/(4+5), digits = 3)} in designs with zero random-slope variance.
+Values chosen for $\sigma_{b_1}$ are reflective of effective curve reliabilities of \Sexpr{round(ecr(model2), digits = 3)} and \Sexpr{round(ecr(model1), digits = 3)} for $\sigma_{b_0}^2 = 1$, and \Sexpr{round(ecr(model4), digits = 3)} and \Sexpr{round(ecr(model3), digits = 3)} for $\sigma_{b_1}^2 = 4$ \citep{BrandyVonO18}.
+
 A full factorial design was employed, yielding $2^5 = 32$ cells of the design; 100 repetitions were performed per cell.
@@
 rownames(splitvar) <- NULL
-
-## LongCART registers best splitting variable if no split was implemented, overwrite
-#table(splitvar[tree_size$Lcrt == 1, "Lcrt"])
-# x1 x14 x16  x2 x28  x5  x9 
-#607   1   1   1   2   1   1 
-#
-## Note: variable selection bias may not be a problem; power seems to be a problem
 Lcrt_splitvars <- splitvar[, "Lcrt"] ## save for later use
@@
 \section{Study~II: Comparison with Other Partitioning Methods}
+\label{sec:studyII}
 
@@
 
-<<echo=FALSE, eval=FALSE>>=
-## SEM trees
-SEM_ids <- which(tmp2$method_id %in% c("f_sem", "f_sem_i", "f_sem_s",
-                                      "s_sem", "s_sem_i", "s_sem_s"))
-mod_SEM <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
-  N + sigma_int + sigma_slope + p_noise + rho), 
-  data = tmp2[SEM_ids, ])
-summary(mod_SEM)
-coefs <- stdCoef(mod_SEM)[,1, drop = FALSE]^2
-r2_N <- sum(coefs[grep("N", rownames(coefs)), ])
-r2_si <- sum(coefs[grep("sigma_int", rownames(coefs)), ])
-r2_ss <- sum(coefs[grep("sigma_slope", rownames(coefs)), ])
-r2_p <- sum(coefs[grep("p_noise", rownames(coefs)), ])
-r2_rho <- sum(coefs[grep("rho", rownames(coefs)), ])
-
-## LongCART
-Lcrt_ids <- which(tmp2$method_id %in% "Lcrt")
-mod <- lm(tree_size ~ N + sigma_int + sigma_slope + p_noise + rho, 
-  data = tmp2[Lcrt_ids, ])
-summary(mod)
-coefs <- stdCoef(mod_LMM)[,1, drop = FALSE]^2
-r2_N <- sum(coefs[grep("N", rownames(coefs)), ])
-r2_si <- sum(coefs[grep("sigma_int", rownames(coefs)), ])
-r2_ss <- sum(coefs[grep("sigma_slope", rownames(coefs)), ])
-r2_p <- sum(coefs[grep("p_noise", rownames(coefs)), ])
-r2_rho <- sum(coefs[grep("rho", rownames(coefs)), ])
-@
-
-<<eval=FALSE, echo=FALSE>>=
-## LM(M) trees only
-mod_all <- lmer(tree_size ~ (1 | dataset_id) + method_id*(
-  N + sigma_int + sigma_slope + p_noise + rho), data = tmp2)
-#summary(mod_LMM)
-#fixef(mod_LMM)[order(abs(fixef(mod_LMM)))]
-coefs <- stdCoef(mod_all)[,1, drop = FALSE]^2
-
-## "Variance explained" by N, sigma_n0, sigma_b1, p_noise and rho:
-sum(coefs[grep("N", rownames(coefs)), ])
-sum(coefs[grep("sigma_int", rownames(coefs)), ])
-sum(coefs[grep("p_noise", rownames(coefs)), ])
-sum(coefs[grep("rho", rownames(coefs)), ])
-sum(coefs[grep("sigma_slope", rownames(coefs)), ])
-
-## Note, above is not really variance explained, but some measure 
-## of influence that does not add up to 1
-
-## Variance explained by main effects:
-coefs["N250", ] 
-coefs["sigma_int2", ]
-coefs["p_noise", ]
-coefs["sigma_slope0.632455532033676", ]
-coefs["rho0.3", ]
-
-round(coefs[order(coefs$stdcoef, decreasing = TRUE), , drop = FALSE], digits = 3)
-
-## Variance explained by random effects:
-varco <- as.data.frame(VarCorr(mod_all))$vcov /
-  var(tmp2[, "tree_size"])
-varco
+<<ech0=FALSE, eval=FALSE>>=
+## Identify which LongCART trees perform very badly. Latest version of LongCART
+## May not perform as bad and may have tgvars argument fixed.
+rng_lcrt <- range(tmp2$tree_size[tmp2$method_id == "Lcrt"])
+lcrt_min_size <- rng_lcrt[1]
+lcrt_max_size <- rng_lcrt[2]
+length(which(tmp2$tree_size == lcrt_min_size & tmp2$method_id == "Lcrt")) ## too many
+length(which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"))
+tmp2[which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"), ]
+## dataset_id has 3200 levels, so unique number for each repetition and row of design
+as.numeric(tmp2[which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"), ]$dataset_id) %% 32
+## So, this is the 30th dataset from the 
+ceiling(as.numeric(tmp2[which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"), ]$dataset_id) / 32)
+## 28th repetition, before the 28th repetition, there were 27:
+27*32+30
+
 @
@@
 \section{Study~III: Partitioning Academic Trajectories}
+\label{sec:studyIII}
 
@@
 
-We analyzed trajectories on children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}.
+We analyzed trajectories of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}.
 Data were collected from 21,304 children from 1,018 schools across the USA.
@@
 Time was measured as the number of months since the baseline assessment.
-In order to obtain approximately linear trajectories, we chose the timing metric based on visual inspection of the observed data: $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.
+In order to obtain approximately linear trajectories, we transformed the timing metric based on visual inspection of the observed data: $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.
+Figure~\ref{fig:time_scaling} in Appendix~\ref{sec:AppendixB} shows the trajectories before and after transformation.
+We chose a fractional exponential rather than a log transform, in order for the baseline assessment to retain a value of 0 and to avoid negative time values.
+Either transformation allows to increase distances between earlier timepoints, and decrease distances between later timepoints.
 
@@
 Boldfaced values indicate the best-performing method for each outcome.
-$R^2$ was computed as $\\frac{\\text{mean(MSE)}}{\\text{var}(y)}$.
+$R^2$ was computed as $1-\\frac{\\text{mean(MSE)}}{\\text{var}(y)}$.
 $^c$~cluster-level covariances; $^r$~estimation initialized with random effects; $^i$~random-intercept variance freely estimated; $^s$~random-slope variance freely estimated.",
@@
 Yet, a possible downside of the local-global estimation approach is that it does not allow for recovering subgroups with differences in random-effects parameters.
-When there is a specific interest in partitioning the random effects, score-based SEM trees may be preferred.
+When there is a specific interest in partitioning the random effects, score-based SEM trees should likely be preferred.
 Alternatively, researchers may want to use the parameter stability tests for mixed-effects models developed by \cite{WangyMerk18} \citep[see also ][]{WangyMerk21,WangyGrav22}.
-This will be useful, for example, when the number of or distances between timepoints differ between respondents and SEM-based growth curve models cannot be applied \citep{NeisyMatt18}.
+This will be useful, for example, when the number of or distances between timepoints differ between respondents so SEM-based growth curve models cannot be applied \citep{NeisyMatt18}.
 
@@
 Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses.
-We expect that the strong performance of cluster-level covariances generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures, but whether our conclusions generalize to settings beyond LGCMs remains to be evaluated.
+We expect that the strong performance of cluster-level covariances generalizes to other settings where partitioning covariates are measured at higher levels, in longitudinal as well as in otherwise nested or clustered data structures; yet, this remains to be empirically evaluated.
 Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances.
@@
 
-\section{Appendix A: Effects of data-generating parameters on tree size}
+\section{Effects of data-generating parameters on tree size}
 \label{sec:AppendixA}
@@
 \label{fig:tree_sizes_interact}
+\end{figure}
+
+\FloatBarrier
+\newpage
+\section{Time scaling for academic trajectories in the Appplication}
+\label{sec:AppendixB}
+
+\begin{figure}[!ht]
+\caption{Observed trajectories before and after transforming time for reading ($N = \Sexpr{nrow(readdata)/5}$), math ($N = \Sexpr{nrow(mathdata)/5}$) and science ($N = \Sexpr{nrow(sciedata)/3}$) abilities in the Application.}
+\begin{subfigure}{1.2\textwidth}%
+<<echo=FALSE, fig=TRUE, height=7>>=
+## Create plots illustrating scaling of time
+par(mfrow = c(3, 2))
+for (resp in c("Reading", "Math", "Science")) {
+
+  ## Load data and apply transformations if necessary
+  load(paste(resp, "ability data.Rdata"))
+  tmp_data <- get(ifelse(resp == "Reading", "readdata", 
+                         ifelse(resp == "Math", "mathdata", "sciedata")))
+  frac <- ifelse(resp == "Scie", 2/3, 1/2)
+  tmp_data$months <- tmp_data$months^(1/frac) 
+  xlab <- switch(resp, Reading = c("time", expression(time^{1/2})), 
+                 Math = c("time", expression(time^{2/3})),
+                 Science = c("time", expression(time^{2/3})))
+
+  ## Create plots
+  for (i in 1:2) {
+    y <- t(sapply(unique(tmp_data$CHILDID), function(x) 
+      tmp_data$score[tmp_data$CHILDID == x]))
+    x <- t(sapply(unique(tmp_data$CHILDID), function(x) 
+      tmp_data$months[tmp_data$CHILDID == x]))
+    plot(tmp_data$month, tmp_data$score, type = "n", main = resp, xlab = xlab[i], 
+         ylab = "Score", cex.axis = .7, cex.lab = .7, cex.main = .7)
+    sapply(1:nrow(x), function(row) 
+      lines(x[row, ], y[row, ], col = gray(0.5, alpha = 1/20), lwd = .05))
+    tmp_data$months <- tmp_data$months^if (i == 1L) frac else (1/frac)
+  }
+
+}
+@
+\end{subfigure}
+\textit{Note.} Before transformation, time is measured in months after baseline assessment.  
+\label{fig:time_scaling}
 \end{figure}

### 2024-02-08 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3280 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: eb01daae67197c936d81f48c4903114f9b805270

@@
 Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity.
-We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with different trajectories in linear growth curve models.
+We show how generalized linear mixed-effects model (GLMM) trees can be used to identify subgroups with different trajectories in linear growth curve models.
 Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data.
 The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.
-In simulated and real-world data, we assess the performance of the extensions and compare against other partitioning methods for growth curve models.
+In simulated and real-world data, we assess performance of the extensions and compare against other partitioning methods for growth curve models.
 Extended GLMM trees perform more accurately than the original algorithm and LongCART, and similarly accurate as structural equation model (SEM) trees.
@@
 library("glmertree")
-load("bag_ids, N = 250, Science.Rda")
 load("Science ability data.Rdata")
-names(sciedata) <- c("CHILDID", "GENDER", "RACE", "SES", "GMOTOR", "FMOTOR", 
-                 "INTERN", "EXTERN", "INTERP", "CONTRO", "FIRKDG", 
-                 "AGEBASELINE", "science", "score", "asmtmm", "months")
-data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
+load("sizesScience.Rda")
+Ns <- c(250, 1000)
+nreps <- 100L
+## Get sample previously used
+set.seed(42)
+for (resp in c("Math", "Reading", "Science")) {
+  for (N in Ns) {
+
+    ## Read in data and prepare objects for saving results
+    load(paste0(resp, " ability data.Rdata"))
+    resp_short <- ifelse(resp == "Math", "math", ifelse(resp == "Reading", "read", "scie"))
+    data <- get(paste0(resp_short, "data"))
     data$CHILDID <- factor(data$CHILDID)
-data$GENDER <- factor(data$GENDER)
-data$asmtmm <- factor(data$asmtmm)
+
+    ## Generate samples for each repetition
+    bag_ids <- sapply(1:nreps, function(x) sample(unique(data$CHILDID), size = N))
+    if (N == 250 && resp == "Science")
+      intro_sample <- bag_ids[ , 39]
+
+  }
+}
+sciedata$CHILDID <- factor(sciedata$CHILDID)
+sciedata$GENDER <- factor(sciedata$GENDER)
+sciedata$RACE <- factor(sciedata$RACE)
+## simplify variable names
+names(sciedata) <- gsub("P1", "", gsub("T1", "", gsub("C1", "", names(sciedata))))
+names(sciedata)[names(sciedata) == "WKSESL"] <- "SES"
+data <- sciedata[sciedata$CHILDID %in% intro_sample, ]
 library("lmerTest")
@@
 \\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.
-The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here.
-To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
+The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories.
+To facilitate interpretation, the labels on the $x$-axes represent values on the original time scale.}
 \end{subfigure}%
@@
 The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept.
-The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher baseline ability also gain more ability over time.
+The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher baseline ability may also gain more ability over time.
 
@@
   INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
-testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
-testdat$CHILDID <- factor(testdat$CHILDID)
-testdat$GENDER <- factor(testdat$GENDER)
-testdat$asmtmm <- factor(testdat$asmtmm)
-## Remove obs with levels of race in test that are not in train
-levs <- unique(testdat$RACE) %in% unique(data$RACE)
-if (!all(levs)) {
-  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
-}
 
 ## Fit models
-lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
-lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
-lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
+lmm_tree <- lmertree(LMM_form, data = data)
+lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID)
+lmm_tree_r <- lmertree(LMM_form, data = data, ranefstart = TRUE)
 
@@
 The $x$-axis is not linear in the number of months, because time was scaled as $(\text{months after baseline})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here.
-To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
+To facilitate interpretation, the labels on the $x$-axes represent values on the original time scale.}
 \label{fig:LMM_tree_r}
@@
 GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are nested in treatment centers, classrooms and/or geographical areas, for example.
-In the current paper we extend GLMM trees, so that they can be applied to partitioning LGCMs.
+In the current paper we extend GLMM trees, so that they can be applied to partitioning linear growth-curve models (LGCMs).
 The general idea of GLMM trees is appealing for subgroup detection in almost any type of mixed-effects model.
@@
 
-There are several alternative RPMs that can be used to partition linear growth-curve models (LGCMs): GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}.
+There are several alternative RPMs that can be used to partition LGCMs: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}.
 Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods specifically target subgroups with different time-by-treatment interactions, so are not generally applicable for partitioning growth curve models.}.
 
-The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the subgroups defined by the terminal nodes of the tree.
-Instead, fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations.
+The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is a local-global estimation approach: GLMM trees do not fit a full parametric model in each of the subgroups defined by the terminal nodes of the tree.
+Instead, fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, while the random-effects parameters are estimated \textit{globally}, using all observations.
 This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes.
@@
 The next section explains how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories.
-Next, the performance of the proposed adjustments is evaluated: In Study~I, we assess performance in simulated datasets, in Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART.
-In Study~III, we assess performance of the proposed adjustments in existing datasets on children's development of reading, math and science abilities.
+Next, the performance of the proposed adjustments is evaluated: In Study~I, we assess performance in simulated datasets, in Study~II, we compare performance of GLMM trees with that of two other partitioning methods: SEM trees and LongCART.
+In Study~III, we assess performance of the proposed adjustments in real datasets on children's development of reading, math and science abilities.
 In the Discussion, we summarize our findings and discuss implications.
@@
 Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon^2$.
-Therefore, the identity function can be used for the link $g$ but generalizations to other response variable types within the GLM are straightforward.
-In the general notation above, $X_i$ is the $n_i \times (p + 1)$ fixed-effects design matrix for subject i $(i = 1, \dots, N)$, comprising $p$ regressors plus one column of 1s for the intercept.
+Therefore, the identity function can be used for the link $g$, but generalizations to other response variable types within the GLM are straightforward.
+In the general notation above, $X_i$ is the $n_i \times (p + 1)$ fixed-effects design matrix for subject $i$ $(i = 1, \dots, N)$, comprising $p$ time-varying covariates plus one column of 1s for the intercept.
 In this paper, we assume that time is the only time-varying covariate of interest (i.e., $p = 1$), but other time-varying covariates can also be included in $X_i$.
 The number $n_i$ and spacing of observed timepoints may differ between subjects.
-Ttime-constant covariates are not contained in $X_i$, but enter the model through the subgroup indicator $j$.
+Time-constant covariates are not contained in $X_i$, but enter the model through the subgroup indicator $j$.
 The value of the fixed-effects parameters $\beta$ (here, intercept and time slope) in GLMM trees are node-specific, their value depending on the subgroup/node $j$ into which subject $i$ falls.
-Subgroup membership is defined by the values of time-constant covariates\footnote{In principle, it is also possible to use time-varying covariates for partitioning with GLMM trees, but this might not be appropriate for partitioning growth curves, as participants changing subgroups over time yields discontinuous growth curves.}.
-
-As in the standard GLMM \citep[e.g., ][]{PinhyBate00}, GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of Xi, and bi is the corresponding vector of random effects for subject $i$.
+In the current paper, subgroup membership is defined by the values of time-constant covariates\footnote{In principle, it is also possible to use time-varying covariates for partitioning with GLMM trees, but this might not be appropriate for partitioning growth curves, as participants changing subgroups over time would yield discontinuous growth curves.}.
+
+As in the standard GLMM \citep[e.g., ][]{PinhyBate00} $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is the corresponding vector of random effects for subject $i$.
 Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.
@@
 This initialization simply assumes zero random effects.
-Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3.
-Typically, this converges when the partition~$j_{(r)}$ from the GLM tree is the same as $j_{(r-1)}$ from the previous step.
+Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~2.
+Typically, this converges when the partition~$j_{(r)}$ from the GLM tree in Step~1 is the same as $j_{(r-1)}$ from the previous iteration.
 
@@
 
-Previous publications on mixed-effects recursive partitioning find that initializing the random-effect estimates with zero yields accurate estimates of subgroup memberships and the final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
+Previous studies on mixed-effects recursive partitioning find that initializing the random-effect estimates with zero yields accurate estimates of subgroup memberships and final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}.
 \cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased with increasing sample size.
@@
 If random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations.
-Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery.
-Specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
+We expect that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery.
+Specifically, that means the GLMM-tree algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
 
@@
 \end{figure}%
-
-To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}.
-In fact, the tree presented in Figure~\ref{fig:LMM_tree_r} was estimated by initializing with the random effects.
-Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}.
-The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in  Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills.
-Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples\footnote{Generalizability will be empirically evaluated using cross validation in Study~III.}.
-
-\FloatBarrier
-\subsection{Partitioning}
-
-\begin{figure}%
-\caption{GLMM tree estimated with cluster-level parameter stability tests.}
-<<fig=TRUE, height=8, width=12, echo=FALSE>>=
-plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
-     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
-                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
-                                        xaxis.labs = c(0, 12, 36, 60)),
-     gp = gpar(cex = 1.5))
-@
-\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
-Gray lines depict observed individual trajectories.
-Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
-\label{fig:LMM_tree_c}
-\end{figure}%
-
-The subgroup structure in Step~1 is estimated by a GLM tree using the general model-based recursive partitioning (MOB) algorithm of \cite{ZeilyHoth08}.
-Here, we give a general overview and then comment on the aspects of the algorithm that may be particularly relevant for LGCMs.
-In the case of GLMs (with random effects held constant in an offset), the MOB algorithm cycles iteratively through the following steps:
-%
-\begin{enumerate}
-\setlength\itemsep{0em}
-\renewcommand{\labelenumi}{(\alph{enumi})}
-	\item Fit the GLM to all observations in the current subgroup. 
-	\item Test for instability of the GLM parameters with respect to each of the partitioning variables.
-	\item If there is some overall parameter instability, split the subgroup with respect to the partitioning variable associated with the highest instability.
-	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
-\end{enumerate}
-%
-Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the GLM fitted in Step~(a).
-Under correct specification of the model and mild regularity conditions, the scores have an expected value of 0.
-The parameter stability tests evaluate whether the scores fluctuate randomly around this mean of 0, or exhibit systematic deviations when ordered by the values of a covariate available for partitioning.
-For continuous covariates $u_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each potential partitioning variable \citep{ZeilyHoth08}:
-%
-\begin{eqnarray}
-\label{eq:efp}
-W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(u_{ik})}
-\end{eqnarray}
-%
-where $\hat{J}$ is a suitable estimate of the covariance matrix of the parameter estimates, and $n_j$ gives the number of observations in the current subgroup.
-Further, $\hat{\psi}_{\sigma(u_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(u_{ik})$} denoting their ordering by the values of partitioning covariate $u_k$.
-Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation associated with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
-
-From the cumulative score process $W_k(t)$, a range of test statistics can be derived which capture increased fluctuations (beyond the random fluctuation under parameter stability).
-For numerical partitioning variables, a maximum Lagrange multiplier test statistic can be computed, which takes the maximum of the squared Euclidean norm of $W_k(t)$, weighted by its variance \citep{ZeilyHorn07}.
-This statistic is referred to as the \textit{supLM} statistic, and is asymptotically equivalent to the maximum of likelihood-ratio statistics.
-Approximate asymptotic $p$-values for the \textit{supLM} statistic can be computed with the method of \cite{Hans97}.
-Categorical covariates do not provide an implicit ordering and scores are therefore binned at each level of the covariate.
-From these, a test statistic is computed that does not depend on the ordering of the levels \citep{MerkyFan14}.
-
-When partitioning longitudinal data, covariates will often be measured at the subject level (i.e., time-invariant covariates), which should be accounted for in computing the estimated covariance matrix $\hat{J}$.
-In general, this computation makes use of the scores.
-By summing the scores within clusters prior to computation of the covariances, so-called \textit{clustered} covariances are obtained, which account for dependence between observations within the same cluster \citep{ZeilyKoll20}.
-This resembles a GEE-type approach with an independence correlation structure.
-Our expectation is that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve subgroup recovery.
-
-The tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests.
-Combining cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates.
-Compared to the trees in Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, the cluster-level parameter stability tests provided the most parsimonious tree structure thus far.
-Table~\ref{tab:ranef} presents the estimated random-effects parameters for the models fitted thus far.
-Not surprisingly, the estimated residual standard deviation ($\hat{\sigma}_\epsilon$) differs only very slightly between models.
-Yet, substantial differences can be seen for the standard deviation of the random intercept ($\hat{\sigma}_\epsilon$): With more subgroups or terminal nodes, less variance is captured by the random effects and more variance is captured by the tree structure.
 
@@
 
-The variances in Table~\ref{tab:ranef} can also be used to quantify variability associated with the trajectories in the terminal nodes of the trees.
-The third column of Table~\ref{tab:ranef} presents pooled standard deviations that can be computed from variance estimates \citep{WestyKenn14}.
-These allow, for example, to gauge the magnitude of subgroup differences in the subgroup-specific intercepts in Table~\ref{tab:local_coefs}, which range from $d = \Sexpr{round(min(ds), digits = 2)}$ to $d = \Sexpr{round(max(ds), digits = 2)}$.
-Yet, these estimates are based on a modest sample size, and are based on training data only, so can be overly optimistic.
-In the next sections, we evaluate performance on simulated (Studies~I~and~II) and real data (Study~III).
+To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}.
+The tree presented in Figure~\ref{fig:LMM_tree_r} was in fact estimated by initializing with the random effects.
+Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}.
+The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills.
+Table~\ref{tab:ranef} presents the estimated random-effects parameters for the models fitted thus far.
+Compared to the global LMM in Figure~\ref{fig:global_trajectories}, the trees in Figure~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree} have very similar residual variances, but clearly lower variance of the random intercept.
+Part of the inter-individual variation captured by the random effects in the global LMM is thus explained by subgroup-specific fixed effects in the trees.
+
+The pooled standard deviations in Table~\ref{tab:ranef} can also be used to compute effect sizes for the subgroup differences \citep{WestyKenn14}.
+The effect sizes for Figure~\ref{fig:LMM_tree_r} and Table~\ref{tab:local_coefs} range from $d = \Sexpr{round(min(ds), digits = 2)}$ for nodes~7~and~8, indicating a small to medium negative effect of internalizing symptoms; to $d = \Sexpr{round(max(ds), digits = 2)}$ for nodes~3~and~9, indicating a huge effect of socio-economic status and gross motor skills combined.
+Yet, these estimates are based on training data only, so can be overly optimistic.
+In light of the small number of respondents for some of the subgroups in Figure~\ref{fig:LMM_tree_r}, these effects may not generalize well to other samples.
+Therefore, generalizability of the tree structures obtained will be empirically evaluated using cross validation in Study~III.
+
+\FloatBarrier
+\subsection{Partitioning}
+
+\begin{figure}%
+\caption{GLMM tree estimated with cluster-level parameter stability tests.}
+<<fig=TRUE, height=8, width=12, echo=FALSE>>=
+plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
+     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
+                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
+                                        xaxis.labs = c(0, 12, 36, 60)),
+     gp = gpar(cex = 1.5))
+@
+\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability.
+Gray lines depict observed individual trajectories.
+Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
+\label{fig:LMM_tree_c}
+\end{figure}%
+
+The subgroup structure in Step~1 of the GLMM-tree algorithmis estimated by a GLM tree using the model-based recursive partitioning (MOB) algorithm of \cite{ZeilyHoth08}.
+Here, we give a general overview and then comment on aspects of the algorithm particularly relevant for partitioning LGCMs.
+In the case of GLMs (with random effects held constant in an offset for GLMMs), the MOB algorithm cycles iteratively through the following steps:
+%
+\begin{enumerate}
+\setlength\itemsep{0em}
+\renewcommand{\labelenumi}{(\alph{enumi})}
+	\item Fit the GLM to all observations in the current subgroup. 
+	\item Test for instability of the GLM parameters with respect to each of the partitioning variables.
+	\item If there is some overall parameter instability, split the subgroup with respect to the partitioning variable associated with the highest instability.
+	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
+\end{enumerate}
+%
+Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the GLM fitted in Step~(a).
+Under correct specification of the model and mild regularity conditions, the scores have an expected value of 0.
+The parameter stability tests evaluate whether the scores fluctuate randomly around this mean of 0, or exhibit systematic deviations when ordered by the values of a partitioning variable.
+For continuous covariates $u_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each of the potential partitioning variables \citep{ZeilyHoth08}:
+%
+\begin{eqnarray}
+\label{eq:efp}
+W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(u_{ik})}
+\end{eqnarray}
+%
+where $\hat{J}$ is a suitable estimate of the covariance matrix of the parameter estimates, and $n_j$ gives the number of observations in the current subgroup.
+Further, $\hat{\psi}_{\sigma(u_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(u_{ik})$} denoting their ordering by the values of partitioning covariate $u_k$.
+Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation associated with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
+
+From the cumulative score process $W_k(t)$, a range of test statistics can be derived which capture increased fluctuations (beyond the random fluctuation under parameter stability).
+For numerical partitioning variables, a maximum Lagrange multiplier test statistic can be computed, which takes the maximum of the squared Euclidean norm of $W_k(t)$, weighted by its variance \citep{ZeilyHorn07}.
+This statistic is referred to as the \textit{supLM} statistic, and is asymptotically equivalent to the maximum of likelihood-ratio statistics.
+Approximate asymptotic $p$-values for the \textit{supLM} statistic can be computed with the method of \cite{Hans97}.
+Categorical covariates do not provide an implicit ordering and scores are therefore binned at each level of the covariate.
+From these, a test statistic is computed that does not depend on the ordering of the levels \citep{MerkyFan14}.
+
+When partitioning longitudinal data, covariates will often be measured at the subject level (i.e., time-invariant covariates), which should be accounted for in computing the estimated covariance matrix $\hat{J}$.
+In general, this computation makes use of the scores.
+By summing the scores within clusters prior to computation of the covariances, so-called \textit{clustered} covariances are obtained, which account for dependence between observations within the same cluster \citep{ZeilyKoll20}.
+This resembles a GEE-type approach with an independence correlation structure.
+We expect that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve subgroup recovery.
+
+To illustrate, the tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests.
+Compared to the trees in Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, this yielded the most parsimonious tree structure thus far.
+The random-effects parameters in Table~\ref{tab:ranef} show that this smaller GLMM tree also has slightly higher variance of the random intercept.
+With less subgroups or terminal nodes, less variance can be captured by the tree structure, and more variance can be captured by the random effects.
+Fitting a GLMM tree with both cluster-level parameter stability tests and random-effects initialization resulted in the exact same tree structure and parameter estimates.
+In the next sections, we will evaluate whether cluster-level parameter stability tests and/or random-effects initialization provide more accurate and better generalizable GLMM trees.
 
 <<echo=FALSE, results=tex, fig.pos='b'>>=
-vcov_d <- as.data.frame(VarCorr(lmm_tree))
-vcov_r <- as.data.frame(VarCorr(lmm_tree_r))
-vcov_c <- as.data.frame(VarCorr(lmm_tree_c))
-vcv <- as.data.frame(VarCorr(LMM))
-tab <- data.frame(Figure = 1:4,
-                  Model = c("LMM", rep("LMM tree", times = 3)), 
-                  Estimation = c("default", "random effects initialization", 
-                                 "default", "clustered covariances"),
-                  n_subs = c(1L, (length(lmm_tree_r$tree)+1)/2, 
-                             (length(lmm_tree$tree)+1)/2, (length(lmm_tree_c$tree)+1)/2))
-tab$sig_b <- c(vcv[vcv$grp == "CHILDID", "sdcor"], vcov_r[vcov_r$grp == "CHILDID", "sdcor"], 
-               vcov_d[vcov_d$grp == "CHILDID", "sdcor"], vcov_c[vcov_c$grp == "CHILDID", "sdcor"])
-tab$sig_epsb <- c(vcv[vcv$grp == "Residual", "sdcor"], vcov_r[vcov_r$grp == "Residual", "sdcor"], 
-               vcov_d[vcov_d$grp == "Residual", "sdcor"], vcov_c[vcov_c$grp == "Residual", "sdcor"])
-tab$pool_sd <- c(sqrt(sum(vcv$vcov)), sqrt(sum(vcov_r$vcov)),
-                 sqrt(sum(vcov_d$vcov)), sqrt(sum(vcov_c$vcov)))
-colnames(tab)[4:7] <- c("# of nodes", "$\\hat{\\sigma}_b$", "$\\hat{\\sigma}_\\epsilon$", "$\\hat{\\sigma}_{\\text{pooled}}$")
-kable_styling(
+colnames(tab)[4:7] <- c("$J$", "$\\hat{\\sigma}_b$", "$\\hat{\\sigma}_\\epsilon$", "$\\hat{\\sigma}_{\\text{pooled}}$")
+kable_styling(add_footnote(
   kable(tab, format = "latex", linesep = "", row.names = FALSE, 
-        label = "ranef", digits = 3, align = c("cllccc"), booktabs = TRUE, escape =FALSE, 
-        caption = "Estimated random-effects parameters for fitted mixed-effects models in Figures 1 through 4."),
-  position = "left")
-@
-
-<<echo=FALSE, eval=FALSE>>=
-test_dat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
-test_dat$CHILDID <- factor(test_dat$CHILDID)
-test_dat$GENDER <- factor(test_dat$GENDER)
-test_dat <- test_dat[test_dat$RACE != 6, ]
-lmmt_preds <- predict(lmm_tree, re.form = NA, newdata = testdat)
-lmmt_c_preds <- predict(lmm_tree_c, re.form = NA, newdata = testdat)
-lmmt_r_preds <- predict(lmm_tree_r, re.form = NA, newdata = testdat)
-mean((test_dat$score - lmmt_preds)^2)
-mean((test_dat$score - lmmt_r_preds)^2)
-mean((test_dat$score - lmmt_c_preds)^2)
-var(test_dat$score)
-lmm_preds <- predict(LMM, newdata = test_dat, re.form = NA)
-mean((test_dat$score - lmm_preds)^2)
-@
-
-% $\textrm{Var} \hat{\beta}$ = (X^\top X)^{-1} X^\top \Omega X (X^\top X)^{-1}$
-
-% The $X^\top \Omega X$ is the meat, and the $(X^\top X)^{-1}$ is the bread.
+        label = "ranef", digits = 3, align = c("cllcccc"), booktabs = TRUE, escape = FALSE, 
+        caption = "Estimated random-effects parameters for the mixed-effects models in Figures 1 through 4."),
+  "\\footnotesize \\\\ \\textit{Note.} $J$ is the number of subgroups; $\\hat{\\sigma}_b$ is the estimated standard deviation of the random intercept; $\\hat{\\sigma}_\\epsilon$ is the estimated residual standard deviation; $\\hat{\\sigma}_{\\text{pooled}}$ is calculated as $\\sqrt{\\hat{\\sigma}^2_b + \\hat{\\sigma}^2_\\epsilon}$.", notation="none", threeparttable = TRUE, escape = FALSE),
+  font_size = 11, full_width=FALSE, position = "left")
+@
+
+%$\textrm{Var} \hat{\beta}$ = (X^\top X)^{-1} X^\top \Omega X (X^\top X)^{-1}$
+
+%The $X^\top \Omega X$ is the meat, and the $(X^\top X)^{-1}$ is the bread.
 For computing robust standard errors, we compute a new kind of meat.
@@
 We simulated data according to the subgroup structure depicted in Figure~\ref{fig:design_tree}.
-Every dataset comprised four non-overlapping subgroups, corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree}.
-The subgroups are defined by the three true partitioning variables: $u_1$, $u_2$ and $u_3$.
+Every dataset comprised four non-overlapping subgroups, corresponding to the terminal nodes of Figure~\ref{fig:design_tree}, defined by the three true partitioning variables: $u_1$, $u_2$ and $u_3$.
 All partitioning variables were generated from a standard normal distribution with $\mu = 0$ and $\sigma^2 = 25$.
@@
 %
-\begin{itemize}
+\begin{enumerate}
 \setlength\itemsep{0em}
@@
 \item Intercorrelation between partitioning variables: absent ($\rho = 0$) or present ($\rho = 0.3$).
-\end{itemize}
+\end{enumerate}
 %
@@
 
-Chosen values of $\sigma_{b_0}$ are reflective of intra-class correlations of \Sexpr{round(1/(1+5), digits = 3)} and \Sexpr{round(4/(4+5), digits = 3)} in designs with zero random-slope variance.
-Values chosen for $\sigma_{b_1}$ are reflective of effective curve reliabilities of \Sexpr{round(ecr(model2), digits = 3)} and \Sexpr{round(ecr(model1), digits = 3)} for $\sigma_{b_0}^2 = 1$, and \Sexpr{round(ecr(model4), digits = 3)} and \Sexpr{round(ecr(model3), digits = 3)} for $\sigma_{b_1}^2 = 4$ \citep{BrandyVonO18}.
-
+Expressed as effect sizes, the values chosen for $\sigma_{b_0}$ correspond to intra-class correlations of \Sexpr{round(1/(1+5), digits = 3)} and \Sexpr{round(4/(4+5), digits = 3)}, respectively.
+Values chosen for $\sigma_{b_1}$ correspond to effective curve reliabilities of \Sexpr{round(ecr(model2), digits = 3)} and \Sexpr{round(ecr(model1), digits = 3)} for $\sigma_{b_0}^2 = 1$, and \Sexpr{round(ecr(model4), digits = 3)} and \Sexpr{round(ecr(model3), digits = 3)} for $\sigma_{b_1}^2 = 4$, as computed per \citep{BrandyVonO18}.
 A full factorial design was employed, yielding $2^5 = 32$ cells of the design; 100 repetitions were performed per cell.
@@
   \item None: $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$.
-Random effects are not estimated and their variances thus fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications:
+Random effects are fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications:
   \begin{itemize}
@@
   \item Intercepts: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$.
-This yields linear mixed-effects model (LMM) trees in which only the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0.
+This yields linear mixed-effects model (LMM) trees in which the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0.
 The four variations considered are the following:
   \begin{itemize}
-    \item Default: Classic observation-level covariances in the parameter stability tests and random-effect initialization with all zeros (original step 0.). 
+    \item Default: Classic observation-level covariances in the parameter stability tests and random-effect initialization with all zeros (original Step~0). 
     \item Alternative: Clustered covariances in the parameter stability tests.
-    \item Alternative: Random-effect initialization with the full-sample estimates (alternative step 0$'$.). 
+    \item Alternative: Random-effect initialization with the full-sample estimates (alternative Step~0$'$). 
     \item Alternative: Clustered covariances and random-effect initialization with the full-sample estimates.
@@
   \item Intercepts and slopes: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$.
-This yields LMM trees in which the variance of both the random intercept and slope were freely estimated.
+This yields LMM trees in which the variance of both random intercept and slope were freely estimated.
 The four variations considered are the same as for the intercept-only LMM trees.
@@
 To compute clustered covariances, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}.
-We employed the outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator for the covariances.
+We employed the outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator.
 
@@
 
-We evaluated tree accuracy by counting the number of splits in each tree, and computing the standard (SD) and mean absolute deviation (MAD) from the true tree size (3 splits).
-Trees with $> 3$ splits are indicative of Type-I error, while trees with $< 3$ splits are indicative of Type-II errors (i.e., power too low to detect the true partitioning variables).
+We evaluated tree accuracy by counting the number of splits in every tree, and computing the standard (SD) and mean absolute deviation (MAD) from the true tree size of three splits.
+A tree with more than three splits was thus taken as Type-I error, while a tree with less than three splits was taken as a Type-II error (i.e., power too low to detect the true partitioning variables).
 We also assessed whether the variable selected for the first split in every tree was the true first splitting variable.
@@
 library("lmerTest")
+
 ## Compare means between settings
@@
 <<echo=FALSE, fig=TRUE, height=4>>=
-library("dplyr")
 
@@
 Distances on $y$-axis are on the log scale.
-cl.cov.~= cluster-level covariances employed in parameter stability tests; ran.eff.~= estimation initialized with the random effects on the full sample.
-}
+cl.cov.~= cluster-level covariances employed in parameter stability tests; ran.eff.~= estimation initialized with the random effects on the full sample.}
 \label{fig:LMM_sizes}
@@
 
-Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split for all LM(M) tree fitting approaches.
+Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for all LM(M) tree fitting approaches.
 Only very rarely is $u_1$ not selected for the first split, if a first split was implemented.
@@
 This allowed for evaluating the possible (dis)advantages of global versus local estimation of random-effects parameters, as well as the performance of the different splitting criteria employed by each method.
-The same data-generating design as in Simulation Study~I was employed.
-To reduce the number of comparisons, we only include performance of LM(M) trees fitted using clustered covariances, because they showed good performance in Simulation Study~I.
+The same data-generating design as in Simulation Study~I was employed (Figure~\ref{fig:design_tree}).
+To reduce the number of comparisons, we only included performance of LM(M) trees fitted using clustered covariances, because these showed good performance in Simulation Study~I.
 
@@
 
-We fitted a total of six SEM trees to every dataset. We used two different splitting criteria:
-
+We fitted a total of six SEM trees to every dataset, using two different splitting criteria: 
+%
 \begin{itemize} 
-
 \item The default ``naive'' splitting approach which employs likelihood ratio tests (LRTs) as the splitting criterion \citep{BranyOert13}.
@@
 An LRT can thus be computed for each candidate split, which quantifies the improvement in fit that would result from implementing this split.
-In each step, the candidate split  yielding the highest LRT is selected for splitting, and splitting is continued as long as a candidate split yields a $p$-value of the LRT above a pre-specified $\alpha$ level (0.05, by default).
+In each step, the candidate split  yielding the highest LRT is selected for splitting, and splitting is continued as long as a candidate split yields a $p$-value of the LRT above a pre-specified $\alpha$ level.
 
@@
 While for GLMM trees, parameter stability tests are computed for the fixed-effects parameters only, score-based SEM trees compute parameter stability tests based on both fixed- and random-effects parameters.
-
 \end{itemize}
-
+%
 For each splitting criterion, three different random-effects specifications were employed:
@@
 Errors were assumed uncorrelated between timepoints and an error variance was freely estimated for each timepoint.
-We used package \textbf{lavaan} \citep[version 0.6-11; ][]{Ross12} to fit the SEMs and we used package \textbf{semtree} \citep[version 0.9.17; ][]{BranyOert13} to fit the SEM trees.
+We used package \textbf{lavaan} \citep[version 0.6-11; ][]{Ross12} to fit the SEMs and package \textbf{semtree} \citep[version 0.9.17; ][]{BranyOert13} to fit the SEM trees.
 
 We fitted a single LongCART tree to each dataset.
-The LongCART function estimates node-specific models comprising a random intercept term; this default cannot be changed.
+The LongCART function estimates node-specific models comprising a random intercept term, a default that cannot be overruled.
 A fixed-effects model was specified with the response regressed on time and a subject-specific random intercept.
-We used package \textbf{LongCART} \citep[version 3.1][]{Kund21} to fit LongCART trees.
+We used package \textbf{LongCART} \citep[version 3.1][]{Kund21} to fit these trees.
+
+For all model-fitting parameters not discussed above, default settings were employed.
+The tree-fitting algorithms employ the same value of $\alpha=0.05$ for evaluating significance of potential splits and deciding whether splitting should continue.
+Yet, they differ in application of the Bonferroni correction: By default, \textbf{semtree} and \textbf{LongCART} do not employ a Bonferroni correction for the number of splitting candidates tried at every split, while \textbf{glmertree} does.
+Further, the algorithms have different defaults for the minimum number of observations in a terminal node (argument \verb|minbucket|).
+In the current experiments, the default setting for \textbf{semtree} implements at most 9 ($N=100$) and 24 ($N=250$) splits.
+Further, \textbf{LongCART} and \textbf{glmertree} both implement at most 24 ($N=100$) and 61 ($N=250$) splits.
+To evaluate the effects  of these differences, we will compare the number of implemented splits against these maxima.
 
@@
 tmp2$method_id <- factor(tmp2$method_id)
-@
-
-<<ech0=FALSE, eval=FALSE>>=
-## Identify which LongCART trees perform very badly. Latest version of LongCART
-## May not perform as bad and may have tgvars argument fixed.
-rng_lcrt <- range(tmp2$tree_size[tmp2$method_id == "Lcrt"])
-lcrt_min_size <- rng_lcrt[1]
-lcrt_max_size <- rng_lcrt[2]
-length(which(tmp2$tree_size == lcrt_min_size & tmp2$method_id == "Lcrt")) ## too many
-length(which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"))
-tmp2[which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"), ]
-## dataset_id has 3200 levels, so unique number for each repetition and row of design
-as.numeric(tmp2[which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"), ]$dataset_id) %% 32
-## So, this is the 30th dataset from the 
-ceiling(as.numeric(tmp2[which(tmp2$tree_size == lcrt_max_size & tmp2$method_id == "Lcrt"), ]$dataset_id) / 32)
-## 28th repetition, before the 28th repetition, there were 27:
-27*32+30
-
 @
@@
 Gray circles represent counts, dark gray horizontal lines represent true number of splits~3.
-Distances on the $y$-axis are on the log scale; $x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.}
+Distances on the $y$-axis are on the log scale; $x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0, or freely estimated.}
 \label{fig:tree_sizes}
@@
 LRT-based SEM trees tended to implement more splits than score-based SEM trees.
-Both LRT- and score-based SEM trees yielded lower tree size with increasing complexity of the random-effect specification.
-With the correct random-effects specification (i.e., both random intercept and slope were estimated), LRT-based SEM trees showed average tree size very close to the true tree size, while under-specification (i.e., specification of only a random intercept) increased tree size by 0.2 splits on average.
-Score-based SEM trees perform best when only random intercepts are estimated, implementing too few splits when random slopes were (correctly) specified.
+Both LRT- and score-based SEM trees yielded lower tree size with more complex random-effect specifications.
+With the correct random-effects specification (i.e., both random intercept and slope freely estimated), LRT-based SEM trees showed average tree size very close to the true tree size, while under-specification (i.e., only the random intercept freely estimated) increased tree size by 0.2 splits on average.
+Score-based SEM trees perform best when only random intercepts were estimated, implementing too few splits when random slopes were (correctly) specified.
 LongCART trees seem to suffer from a lack of power, implementing only two splits on average.
-Overall, LM(M) trees with clustered covariances showed best performance, but are very closely followed by LRT-based SEM trees, which seem to be affected more strongly by mis-specification of the random effects.
+Overall, LM(M) trees with clustered covariances showed best performance, but very closely followed by LRT-based SEM trees, which seem more strongly affected by mis-specification of the random effects.
 
@@
 The number of splits implemented by SEM trees was most strongly affected by the data-generating parameters when the random effects were mis-specified (i.e., random intercepts and/or slope fixed to 0).
+None of the algorithms reached the maximum number of splits retained as specified by the \verb|minbucket| argument in any of the simulated datasets, suggesting that this parameter does not affect tree size.
+The strong effect of sample size suggests the significance level $\alpha$ for continuation of splitting is much more important.
+With $N=100$, underfitting is more likely, while with $N=250$, overfitting is more likely, suggesting that the default $\alpha$ performs well in these simulations.
+Combined with the very small effect of the number of potential partitioning variables, the default (lack of) Bonferroni correction also performs well.
 
@@
 Table~\ref{tab:first_splits2} presents variable selection frequencies for the first split in the fitted trees.
-For SEM trees, the LRT criterion yields almost perfect accuracy for the first split.
-The score-based criterion for SEM trees provides near-perfect accuracy when random effects were correctly specified.
+LRT-based SEM trees provide almost perfect accuracy for the first split.
+Score-based SEM trees provide near-perfect accuracy when random effects were correctly specified.
 When random effects were mis-specified, score-based SEM trees selected the wrong variable for the first split in about \Sexpr{round(100*(1 - mean(splits$`$u_1$`[grep("s_sem", rownames(splits))])))}\% of datasets.
@@
 LongCART showed a strong tendency to select $u_2$ or $u_3$ for the first split.
-Closer inspection of the fitted LongCART trees revealed that in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets in which no splits were implemented, $u_1$ was the strongest splitting candidate, but the parameter stability tests did not reach significance.
-This suggests the tests proposed by \cite{KundyHare19} are less sensitive to instability of the intercept (compared to instability of the slope), or less sensitive to instability with respect to categorical covariates (compared to instability with respect to continuous covariates).
+Closer inspection of the fitted LongCART trees revealed that in \Sexpr{round(100*prop.table(table(Lcrt_splitvars[tree_size$Lcrt == 1L] == "x1"))["TRUE"])}\% percent of datasets in which no splits were implemented, $u_1$ was found to be the strongest splitting candidate, but the parameter stability tests did not reach significance.
+This suggests the tests proposed by \cite{KundyHare19} are less sensitive to instability of the intercept (compared to slope instability), or less sensitive to instability with respect to categorical covariates (compared to continuous covariates).
 
@@
 A clear computational advantage is observed for LM trees.
-LMM trees require longer computation times because of the estimation of random effects.
-Yet, this increase is minor compared to the computation times required by LongCART and SEM trees: LRT-based SEM trees required striking computation times, while score-based SEM trees were computationally more efficient, as expected.
+LMM trees require somewhat longer computation times because of the estimation of random effects.
+Yet, this increase seems minor compared to the computation times required by LongCART and score-based SEM trees.
+As expected, LRT-based SEM trees require strikingly long computation times.
 
@@
 
-We applied five LM(M) trees to the data, focusing on the original GLMM tree approach and approaches that performed well in the simulations:
+We applied five LM(M) trees to the data, focusing on the original GLMM tree approach and those that performed well in the simulations:
 % 
@@
   \item An LMM tree using clustered covariances, with both random intercept and slope freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$) and initialization assuming zero random effects .
-  \item An LMM tree using observation-level covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and random-effect initialization with the full-sample estimate.
+  \item An LMM tree using observation-level covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and random-effects initialization with the full-sample estimate.
 \end{itemize}
@@
 Although LRT- and score-based SEM trees performed very well in the simulations, they could not be used in this study because growth-curve SEMs do not allow for incorporating continuous time.
-
-%We therefore opted for using longRPart \citep[version 1.0; ][]{AbdoyLeBl02}, which employs the exact same LRT-based splitting criterion as LRT-based SEM trees.
-It however fits a mixed-effects model in every node, instead of a SEM, allowing for the incorporation of time as a continuous variable.
-We fitted two longRPart trees with different random effects specifications:
+For comparison, we therefore fitted two linear mixed models, both with random intercept and slope of time freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$):
 %
-%\begin{itemize}
-%  \item A longRPart tree which freely estimates a random intercept ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$).
-%  \item A longRPart tree which freely estimates both a random intercept and slope ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$).
-%\end{itemize}
+\begin{itemize}
+  \item A simple LMM, with a linear fixed effect of time, reflecting a situation where no subgroups are detected or used for prediction.
+  \item A complex LMM, with a linear fixed effect of time, as well as main effects and interactions with time of all possible partitioning variables.
+\end{itemize}
 %
-%The default longRPart settings yielded a very large number of splits in the current analyses and poor predictive accuracy.
-We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = \Sexpr{2^5} terminal nodes, or \Sexpr{(2^5)-1} splits.
-We created a custom function for computing predictions from fitted longRPart trees, as this functionality was not included in the package itself.
-
-%As benchmarks, we included two LMMs with a fixed effect of time: 
-%
-%\begin{itemize}
-%  \item An LMM with a random intercept freely estimated.
-%  \item An LMM with both random intercept and slope freely estimated.
-%\end{itemize}
-%
-%Besides time and the child identifier, the LMMs did not include additional predictors in the model.
 
@@
 We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values (MSE) for all children not included in the training sample in the current repetition.
-This separation of train and test observations does not allow for using the random effects in computing predictions; the cross-validated MSEs only quantify accuracy of the fixed-effects parameters.
-Tree size was measured by counting the number of splits in each tree.
+Such separation of train and test observations does not allow for using random effects in computing predictions; cross-validated MSEs thus only quantify accuracy of the fixed-effects parameters.
+Model complexity was measured by two indicators: Tree size as quantified by the number of splits in each tree, and the number of partitioning variables included in the final model.
+Pairwise $t$-tests with Bonferroni correction to account for multiple testing were applied to the MSEs of all methods to evaluate statistical significance of differences.
 
@@
 
-Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions.
-Differences in predictive performance are small, all $R^2$ differences are smaller than 0.01.
-Bonferroni-adjusted pairwise $t$-tests indicated no difference in performance between default LMM trees and LMM trees with random-effects initialization for any of the three outcomes.
-In contrast, the three LM(M) trees using clustered covariances performed significantly better for the math and reading outcomes.
-For the science outcomes, no significant differences were observed.
-Thus, cluster-level covariances seem to provide the most robust improvement of performance.
-
-Figure~\ref{fig:application_sizes} presents tree size distributions.
-Similar to the simulation study's results, default LMM trees implement the largest number of splits.
-Cluster-level covariances provide a robust reduction in the number of splits.
-Given their non-distinguishable predictive performance, LM trees with clustered covariances may be preferred if a obtaining a sparse result is critical.
-LMM trees with clustered covariances may provide less sparsity but better predictive accuracy, irrespective of whether random intercepts and/or slopes were estimated.
-In contrast to the simulation study results, random-effects initialization here yields tree sizes similar to the default fitting approach.
-
-\begin{figure}[!bt]
-\caption{Mean squared errors for trees fitted to math, reading and science ability trajectories.}
-\begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=2, width=5>>=
-## Load datasets for computing variance of response variables
+<<echo=FALSE>>=
+## Load MSEs
+load("MSEsMath.Rda")
+math_MSEs <- MSEs
+load("MSEsReading.Rda")
+read_MSEs <- MSEs
+load("MSEsScience.Rda")
+scie_MSEs <- MSEs
+
+## Load datasets
 load("Reading ability data.Rdata")
@@
 
-## MSE
-load("MSEs, N = 250, Math.Rda")
-math_MSEs <- cbind(MSEs, N = 250)
-#load("MSEs, N = 1000, Math.Rda")
-#math_MSEs <- rbind(math_MSEs, cbind(MSEs, N = 1000))
-#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart
-#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart_s
-#sapply(math_MSEs, function(x) tapply(x, math_MSEs$N, mean, na.rm =TRUE))
-
-load("MSEs, N = 250, Reading.Rda")
-read_MSEs <- cbind(MSEs, N = 250)
-#load("MSEs, N = 1000, Reading.Rda")
-#read_MSEs <- rbind(read_MSEs, cbind(MSEs, N = 1000))
-#sapply(read_MSEs, function(x) table(is.na(x)))
-#sapply(read_MSEs, function(x) tapply(x, read_MSEs$N, mean))
-
-load("MSEs, N = 250, Science.Rda")
-scie_MSEs <- cbind(MSEs, N = 250)
-#load("MSEs, N = 1000, Science.Rda")
-#scie_MSEs <- rbind(scie_MSEs, cbind(MSEs, N = 1000))
-#sapply(scie_MSEs, function(x) table(is.na(x)))$longRPart
-#sapply(scie_MSEs, function(x) table(is.na(x)))$longRPart_s
-#sapply(scie_MSEs, function(x) tapply(x, scie_MSEs$N, mean))
+R2 <- round(data.frame(scie_min = 1 - sapply(scie_MSEs, min)/var_scie, 
+                       read_min = 1 - sapply(read_MSEs, min)/var_read,
+                       math_min = 1 - sapply(math_MSEs, min)/var_math,
+                       scie_max = 1 - sapply(scie_MSEs, max)/var_scie,
+                       read_max = 1 - sapply(read_MSEs, max)/var_read,
+                       math_max = 1 - sapply(math_MSEs, max)/var_math), digits = 2)
+
+load("vars_selectedMath.Rda")
+vs_math <- vars_selected
+load("vars_selectedReading.Rda")
+vs_read <- vars_selected
+load("vars_selectedScience.Rda")
+vs_scie <- vars_selected
+@
+
+Table~\ref{tab:application_MSEs} and Figure~\ref{fig:application_MSEs} present MSE and $R^2$ distributions.
+Simple LMMs comprising only a fixed effect of time performed worst, but their $R^2$ values ranging from \Sexpr{min(R2["LMM_s", ])} to \Sexpr{max(R2["LMM_s", ])} illustrate that time is the major predictor of math, reading and science abilities.
+The complex LMMs, comprising fixed effects of time and all covariates as well as their interactions, showed best performance, with $R^2$ values ranging from \Sexpr{min(R2["LMM_s3", ])} to \Sexpr{max(R2["LMM_s3", ])}.
+Table~\ref{tab:application_vars_selected} shows the number of covariates used for prediction, and indicates this performance gain comes at a considerable cost in complexity.
+
+Predictive performance of LM(M) trees strikes a balance between the simple and complex LMMs.
+As shown in Table~\ref{tab:application_MSEs}, LM(M) trees with clustered covariances provide predictive accuracy closer to the complex than the simple LMMs, while using only a fraction of the covariates: \Sexpr{round(100*mean(apply(sapply(vs_math, \(x) apply(x, 1, sum)), 2, mean)[c(1,3,5)])/11)}\% of covariates for predicting math, \Sexpr{round(100*mean(apply(sapply(vs_read, \(x) apply(x, 1, sum)), 2, mean)[c(1,3,5)])/11)}\% of covariates for reading, and \Sexpr{round(100*mean(apply(sapply(vs_scie, \(x) apply(x, 1, sum)), 2, mean)[c(1,3,5)])/11)}\% of covariates for science trajectories, on average.
+As such, LMM trees substantially improve upon the simple, no-subgroup LMMs using only few partitioning variables.
+
+Between the different LMM-tree fitting approaches, differences in performance are small, all $R^2$ differences $< 0.01$.
+There were no significant differences in performance between default LMM trees and LMM trees with random-effects initialization for any of the three outcomes.
+In contrast, the three LM(M) trees using clustered covariances performed significantly better for the math and reading outcomes.
+For the science outcomes, no significant differences between LM(M) trees were observed.
+Thus, cluster-level covariances seem to provide the best predictive performance for LM(M) trees.
+
+Figure~\ref{fig:application_sizes} presents tree size distributions for the LM(M) trees.
+Similar to the simulation study's results, default LMM trees implement the largest number of splits.
+Cluster-level covariances provide a robust reduction in the number of splits.
+Given their non-distinguishable predictive performance, LM trees with clustered covariances may be preferred if obtaining a sparse result is critical.
+LMM trees with clustered covariances may provide less sparsity but better predictive accuracy, irrespective of whether random intercepts and/or slopes were estimated.
+In contrast to the simulation study results, random-effects initialization resulted in tree sizes similar to the default fitting approach in these experiments.
+
+\begin{figure}[!bt]
+\caption{Mean squared errors for trees fitted to math (top row), reading (middle row) and science (bottom row) ability trajectories.}
+\begin{subfigure}{1.15\textwidth}
+<<echo=FALSE, fig=TRUE, height=2, width=7>>=
+
+## Difference of LMM trees with simple and complex LMMs
+#round((sapply(math_MSEs, mean) - sapply(math_MSEs, mean)["LMM_s3"])/var_math, digits = 4)[c(1:5,7)]
+#round((sapply(read_MSEs, mean) - sapply(read_MSEs, mean)["LMM_s3"])/var_read, digits = 4)[c(1:5,7)]
+#round((sapply(scie_MSEs, mean) - sapply(scie_MSEs, mean)["LMM_s3"])/var_scie, digits = 4)[c(1:5,7)]
+## Distance between LMM trees and simple LMM is always bigger than LMM trees and complex LMM
+
+## Create table with M, SD, R2
 
 ## Math
-MSE_df <- data.frame(sapply(math_MSEs[ , c(1:4, 6)], 
-                              function(x) sprintf("%.4f", tapply(x, math_MSEs$N, mean, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, sd, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(math_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.3f", tapply(x, math_MSEs$N, 
-                                                                  function(x) 1 - mean(x, na.rm = TRUE)/var_math))))
-## Reading
-MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.4f", tapply(x, read_MSEs$N, mean, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, sd, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(read_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.3f", tapply(x, read_MSEs$N, 
-                                                                  function(x) 1 - mean(x, na.rm = TRUE)/var_read))))
+MSE_df <- sprintf("%.4f", sapply(math_MSEs[c(1:5, 7, 11)], mean, na.rm = TRUE))
+MSE_df <- cbind(MSE_df, sprintf("%.3f", sapply(math_MSEs[c(1:5, 7, 11)], sd, na.rm = TRUE)))
+MSE_df <- cbind(MSE_df, sprintf("%.3f", sapply(math_MSEs[c(1:5, 7, 11)], \(x) 1 - mean(x, na.rm = TRUE)/var_math)))
+
+## Read
+MSE_df <- cbind(MSE_df, sprintf("%.4f", sapply(read_MSEs[c(1:5, 7, 11)], mean, na.rm = TRUE)))
+MSE_df <- cbind(MSE_df, sprintf("%.3f", sapply(read_MSEs[c(1:5, 7, 11)], sd, na.rm = TRUE)))
+MSE_df <- cbind(MSE_df, sprintf("%.3f", sapply(read_MSEs[c(1:5, 7, 11)], \(x) 1 - mean(x, na.rm = TRUE)/var_read)))
 
 ## Science
-MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.4f", tapply(x, scie_MSEs$N, mean, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, sd, na.rm = TRUE))))
-MSE_df <- cbind(MSE_df, sapply(scie_MSEs[ , c(1:4, 6)], 
-                               function(x) sprintf("%.3f", tapply(x, scie_MSEs$N, 
-                                                                  function(x) 1 - mean(x, na.rm = TRUE)/var_scie))))
+MSE_df <- cbind(MSE_df, sprintf("%.4f", sapply(scie_MSEs[c(1:5, 7, 11)], mean, na.rm = TRUE)))
+MSE_df <- cbind(MSE_df, sprintf("%.3f", sapply(scie_MSEs[c(1:5, 7, 11)], sd, na.rm = TRUE)))
+MSE_df <- cbind(MSE_df, sprintf("%.3f", sapply(scie_MSEs[c(1:5, 7, 11)], \(x) 1 - mean(x, na.rm = TRUE)/var_scie)))
 
 ## tmp3 is math
-tmp3 <- data.frame(stack(math_MSEs[ , c(1:4, 6)]),
-                    dataset_id = factor(rep(1:nrow(math_MSEs), times = 5)),
-                   N = math_MSEs$N)
+tmp3 <- data.frame(stack(math_MSEs[ , c(1:5, 7, 11)]),
+                    dataset_id = factor(rep(1:nrow(math_MSEs), times = 7)))
 names(tmp3)[1:2] <- c("MSE", "method") 
 tmp3$method2 <- as.character(tmp3$method)
-
+tmp3$method2[grepl("LMtree", tmp3$method2)] <- "LM tree\n\n(cluster)"
+tmp3$method2[tmp3$method2 %in% c("LMMtree_c", "LMMtree_sc")] <- "LMM tree\n\n(cluster)"
+tmp3$method2[tmp3$method2 == "LMMtree_r"] <- "LMM tree\n\n(ranef)"
+tmp3$method2[tmp3$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
 tmp3$method2[tmp3$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
@@
 tmp3$method2[tmp3$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
+tmp3$method2[tmp3$method2 %in% c("LMM_s", "LMM_s3")] <- " LMM"
 tmp3$random <- ifelse(tmp3$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
@@
                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp3$method %in% c("LMMtree_sc"), 
-                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp3$N <- factor(tmp3$N)
+                             ifelse(tmp3$method == "LMMtree_sc", 
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)",
+                                ifelse(tmp3$method == "LMM_s", " time ", "all"))))
 theme_set(theme_bw(base_size = 12))
@@
 \end{subfigure}
-\begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=1.6, width=5>>=
+\begin{subfigure}{1.15\textwidth}
+<<echo=FALSE, fig=TRUE, height=1.6, width=7>>=
 ## tmp4 is reading
-tmp4 <- data.frame(stack(read_MSEs[ , c(1:4,6)]),
-                    dataset_id = factor(rep(1:nrow(read_MSEs), times = 5)),
-                   N = read_MSEs$N)
+tmp4 <- data.frame(stack(read_MSEs[ , c(1:5, 7, 11)]),
+                    dataset_id = factor(rep(1:nrow(read_MSEs), times = 7)))
 names(tmp4)[1:2] <- c("MSE", "method") 
@@
 tmp4$method2[tmp4$method2 %in% c("LMMtree_cr", "LMMtree_scr")] <- "LMM tree\n\n(cluster + ranef)"
-tmp4$method2[grepl("longR", tmp4$method2)] <- "longRPart"
 tmp4$method2[tmp4$method2 == "LMtree_c"] <- "LM tree\n\n(cl.cov.)"
@@
 tmp4$method2[tmp4$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
+tmp4$method2[tmp4$method2 %in% c("LMM_s", "LMM_s3")] <- " LMM"
 tmp4$random <- ifelse(tmp4$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
@@
                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp4$method %in% c("LMMtree_sc"), 
-                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp4$N <- factor(tmp4$N)
+                             ifelse(tmp4$method == "LMMtree_sc",
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", 
+                                ifelse(tmp4$method == "LMM_s", " time ", "all"))))
 ggplot(tmp4) +
@@
 \end{subfigure}
-\begin{subfigure}{\textwidth}
-<<echo=FALSE, fig=TRUE, height=2, width=5>>=
+\begin{subfigure}{1.15\textwidth}
+<<echo=FALSE, fig=TRUE, height=2, width=7>>=
 ## tmp5 is science
-tmp5 <- data.frame(stack(scie_MSEs[ , c(1:4,6)]), ## omit longRPart with 7:10
-                    dataset_id = factor(rep(1:nrow(scie_MSEs), times = 5)),
-                   N = scie_MSEs$N)
+tmp5 <- data.frame(stack(scie_MSEs[ , c(1:5, 7, 11)]),
+                    dataset_id = factor(rep(1:nrow(scie_MSEs), times = 7)))
 names(tmp5)[1:2] <- c("MSE", "method") 
@@
 tmp5$method2[tmp5$method2 == "LMMtree_r"] <- "LMM tree\n\n(ran.eff.)"
+tmp5$method2[tmp5$method2 %in% c("LMM_s", "LMM_s3")] <- " LMM"
 tmp5$random <- ifelse(tmp5$method == "LMtree_c", "atop(hat(sigma)[b[0]] == 0, hat(sigma)[b[1]] == 0)", 
@@
                              "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] == 0)", 
-                             ifelse(tmp5$method %in% c("LMMtree_sc"), 
-                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp5$N <- factor(tmp5$N)
-library("ggforce")
+                             ifelse(tmp5$method == "LMMtree_sc", 
+                                "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", 
+                                ifelse(tmp5$method == "LMM_s", " time ", "all"))))
 ggplot(tmp5, aes(x=random, y=MSE)) +
   geom_boxplot(position=position_dodge(1), alpha = .5, width = .45, fill = "gray") +
-  scale_y_continuous(#trans = log_trans(),
-                       sec.axis = sec_axis(trans=~1-(./var_scie))) +
+  scale_y_continuous(sec.axis = sec_axis(trans=~1-(./var_scie))) +
   facet_grid(~method2, scales = "free", space = "free") +
@@
 \vspace{-0.5cm}%
-{\footnotesize \\ \textit{Note.
-} Top, middle and bottom panels depict math, reading and science ability trajectories, respectively.
-The secondary $y$-axis on the right quantifies the proportion of variance explained, computed as $1 - \frac{\text{mean(MSE)}}{\text{var}(y)}$.}
+{\footnotesize \\ \textit{Note. } Secondary $y$-axis on the right quantifies the proportion of variance explained, computed as $1 - \frac{\text{mean(MSE)}}{\text{var}(y)}$.}
 \label{fig:application_MSEs}
@@
 rownames(MSE_df) <- c("LM tree$^c$", "LMM tree$^i$", "LMM tree$^{i,c}$", 
-                      "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$")
-colnames(MSE_df) <- rep(c("M", "SD", "$R^2$"), times = 3)
-
-## Add benchmark LMMs
-#load("LMM MSEs, N = 250, Math.RDa")
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 1] <- format(colMeans(MSEs), digits = 4)
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 2] <- round(sapply(MSEs, sd), digits = 3)
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 3] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_math)
-
-#load("LMM MSEs, N = 250, Reading.RDa")
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 4] <- format(colMeans(MSEs), digits = 4)
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 5] <- round(sapply(MSEs, sd), digits = 3)
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 6] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_read)
-
-#load("LMM MSEs, N = 250, Science.RDa")
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 7] <- format(colMeans(MSEs), digits = 4)
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 8] <- round(sapply(MSEs, sd), digits = 3)
-#MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 9] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_scie)
+                      "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$", "LMM$^t$", "LMM$^a$")
+colnames(MSE_df) <- rep(c("$M$", "$SD$", "$R^2$"), times = 3)
 
 ## boldface highest R2 values
-best_ids <- apply(MSE_df[ , c(1,4,7)], 2, which.min)
-MSE_df[best_ids[1] , 1:3] <- paste0("\\textbf{", MSE_df[best_ids[1] , 1:3], "}")
-MSE_df[best_ids[2] , 4:6] <- paste0("\\textbf{", MSE_df[best_ids[2] , 4:6], "}")
-MSE_df[best_ids[3] , 7:9] <- paste0("\\textbf{", MSE_df[best_ids[3] , 7:9], "}")
-
-kable_styling(add_header_above(#add_header_above(
+#best_ids <- apply(MSE_df[ , c(1,4,7)], 2, which.min)
+#MSE_df[best_ids[1] , 1:3] <- paste0("\\textbf{", MSE_df[best_ids[1] , 1:3], "}")
+#MSE_df[best_ids[2] , 4:6] <- paste0("\\textbf{", MSE_df[best_ids[2] , 4:6], "}")
+#MSE_df[best_ids[3] , 7:9] <- paste0("\\textbf{", MSE_df[best_ids[3] , 7:9], "}")
+
+kable_styling(
+  add_header_above(
   add_footnote(
-  kable(MSE_df[c(1:3, 5, 4), ], format = "latex", booktabs = TRUE, label = "application_MSEs", align = "c", 
+  row_spec(
+  kable(MSE_df[c(1:3, 5, 4, 6:7), ], format = "latex", booktabs = TRUE, label = "application_MSEs", align = "c", 
         escape=FALSE, linesep = "", # linesep command suppresses addlinesep every 5 rows
         caption = "Cross-validated mean squared errors for each of the response variables."),
+  5, hline_after = TRUE,), 
   "\\footnotesize \\\\ \\textit{Note.} Means and standard deviations computed over 100 cross-validation repetitions.
-Boldfaced values indicate the best-performing method for each outcome.
 $R^2$ was computed as $1-\\frac{\\text{mean(MSE)}}{\\text{var}(y)}$.
-$^c$~cluster-level covariances; $^r$~estimation initialized with random effects; $^i$~random-intercept variance freely estimated; $^s$~random-slope variance freely estimated.",
+$^c$~cluster-level covariances; $^r$~estimation initialized with random effects; $^i$~random-intercept variance freely estimated; $^s$~random-slope variance freely estimated; $^t$~LMM with fixed effect of time; $^a$~LMM with fixed effects of time, all covariates and all time-by-covariate interactions.",
   notation="none", threeparttable = TRUE, escape = FALSE), 
-  #c(" ", "N = 250"=1, "N = 1,000"=1, "N = 250"=1, "N = 1,000"=1,
-  #  "N = 250"=1, "N = 1,000"=1), align = "c"),
   c(" ", "Math" = 3, "Reading" = 3, "Science" = 3), align = "c"),
@@
 
-<<eval=FALSE, echo=FALSE>>=
-math_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = mathdata)
-(as.data.frame(VarCorr(math_LMM))$vcov / var_math)
-(as.data.frame(VarCorr(math_LMM))$vcov / as.data.frame(VarCorr(math_LMM))$vcov[4])
-
-read_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = readdata)
-(as.data.frame(VarCorr(read_LMM))$vcov / var_read)
-(as.data.frame(VarCorr(read_LMM))$vcov / as.data.frame(VarCorr(read_LMM))$vcov[4])
-
-scie_LMM <- lmer(score ~ months + (1 + months|CHILDID), data = sciedata)
-(as.data.frame(VarCorr(scie_LMM))$vcov / var_scie)
-(as.data.frame(VarCorr(scie_LMM))$vcov / as.data.frame(VarCorr(scie_LMM))$vcov[4])
+<<results=tex, echo=FALSE>>=
+vs <- cbind(sprintf("%.2f", c(apply(sapply(vs_math, \(x) apply(x, 1, sum)), 2, mean), 0, 11)),
+            sprintf("%.2f", c(apply(sapply(vs_math, \(x) apply(x, 1, sum)), 2, sd), 0, 0)),
+            sprintf("%.2f", c(apply(sapply(vs_read, \(x) apply(x, 1, sum)), 2, mean), 0, 11)),
+            sprintf("%.2f", c(apply(sapply(vs_read, \(x) apply(x, 1, sum)), 2, sd), 0, 0)),
+            sprintf("%.2f", c(apply(sapply(vs_scie, \(x) apply(x, 1, sum)), 2, mean), 0, 11)),
+            sprintf("%.2f", c(apply(sapply(vs_scie, \(x) apply(x, 1, sum)), 2, sd), 0, 0)))
+colnames(vs) <- rep(c("M", "SD"), times = 3)
+rownames(vs) <- c("LM tree$^c$", "LMM tree$^i$", "LMM tree$^{i,c}$", 
+                  "LMM tree$^{i,r}$", "LMM tree$^{i,s,c}$", "LMM$^{t}$", "LMM$^{a}$")
+kable_styling(
+  add_header_above(
+  add_footnote(
+  row_spec(
+  kable(vs, format = "latex", booktabs = TRUE, label = "application_vars_selected", align = "c", 
+        escape=FALSE, linesep = "", # linesep command suppresses addlinesep every 5 rows
+        caption = "Mean number of variables selected for partitioning."),
+  5, hline_after = TRUE),
+  "\\footnotesize \\\\ \\textit{Note.} Means and standard deviations computed over 100 cross-validation repetitions.
+$^c$~cluster-level covariances; $^r$~estimation initialized with random effects; $^i$~random-intercept variance freely estimated; $^s$~random-slope variance freely estimated; $^t$~LMM with fixed effect of time; $^a$~LMM with fixed effects of time, all covariates and all time-by-covariate interactions.",
+  notation="none", threeparttable = TRUE, escape = FALSE), 
+  c(" ", "Math" = 2, "Reading" = 2, "Science" = 2), align = "c"),
+  font_size = 11, full_width=FALSE)
 @
@@
 \begin{figure}[!bt]%
-\caption{Sizes of trees fitted to math, reading, and science ability trajectories.}
+\caption{Sizes of trees fitted to math (top row), reading (middle row), and science (bottom row) ability trajectories.}
 \begin{subfigure}{\textwidth}%
@@
 ## Tree size
-load("sizes, N = 250, Math.Rda")
-math_sizes <- cbind(sizes, N = 250)
-#load("sizes, N = 1000, Math.Rda")
-#math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
+load("sizesMath.Rda")
+math_sizes <- sizes
 #sapply(math_sizes, function(x) table(is.na(x)))
@@
 
-load("sizes, N = 250, Reading.Rda")
-read_sizes <- cbind(sizes, N = 250)
-#load("sizes, N = 1000, Reading.Rda")
-#read_sizes <- rbind(read_sizes, cbind(sizes, N = 1000))
+load("sizesReading.Rda")
+read_sizes <- sizes
 #sapply(read_sizes, function(x) table(is.na(x)))
@@
 
-load("sizes, N = 250, Science.Rda")
-scie_sizes <- cbind(sizes, N = 250)
-#load("sizes, N = 1000, Science.Rda")
-#scie_sizes <- rbind(scie_sizes, cbind(sizes, N = 1000))
+load("sizesScience.Rda")
+scie_sizes <- sizes
 #sapply(scie_sizes, function(x) table(is.na(x)))
@@
 
-## Made a mistake in evaluating tree size in the Application scripts:
-## I took #splits = nrow(tree$frame)-1, assuming tree$frame contains a row for
-## each terminal node. It contains a row for each inner and terminal node.
-## Thus, #splits = (nrow(tree$frame)-1)/2; original number should be divided by 2:
-#math_sizes$longRPart <- math_sizes$longRPart/2 
-#math_sizes$longRPart_s <- math_sizes$longRPart_s/2 
-#read_sizes$longRPart <- read_sizes$longRPart/2 
-#read_sizes$longRPart_s <- read_sizes$longRPart_s/2 
-#scie_sizes$longRPart <- scie_sizes$longRPart/2 
-#scie_sizes$longRPart_s <- scie_sizes$longRPart_s/2
-
 ## tmp6 is math
-tmp6 <- data.frame(stack(math_sizes[ , c(1:4,6)]),
-                    dataset_id = factor(rep(1:nrow(math_sizes), times = 5)),
-                   N = math_sizes$N)
+tmp6 <- data.frame(stack(math_sizes[ , 1:5]),
+                    dataset_id = factor(rep(1:nrow(math_sizes), times = 5)))
 names(tmp6)[1:2] <- c("splits", "method") 
@@
                                 "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp6$N <- factor(tmp6$N)
 ggplot(tmp6) +
@@
 ## tmp7 is reading
-tmp7 <- data.frame(stack(read_sizes[ , c(1:4,6)]),
-                    dataset_id = factor(rep(1:nrow(read_sizes), times = 5)),
-                   N = read_sizes$N)
+tmp7 <- data.frame(stack(read_sizes[ , 1:5]),
+                    dataset_id = factor(rep(1:nrow(read_sizes), times = 5)))
 names(tmp7)[1:2] <- c("splits", "method") 
@@
                                 "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp7$N <- factor(tmp7$N)
 ggplot(tmp7) +
@@
 ## tmp8 is science
-tmp8 <- data.frame(stack(scie_sizes[ , c(1:4,6)]),
-                    dataset_id = factor(rep(1:nrow(scie_sizes), times = 5)),
-                   N = scie_sizes$N)
+tmp8 <- data.frame(stack(scie_sizes[ , 1:5]),
+                    dataset_id = factor(rep(1:nrow(scie_sizes), times = 5)))
 names(tmp8)[1:2] <- c("splits", "method") 
@@
                                 "atop(hat(sigma)[b[0]] > 0, hat(sigma)[b[1]] > 0)", NA)))
-tmp8$N <- factor(tmp8$N)
 ggplot(tmp8) +
@@
 \end{subfigure}%
-{\\\footnotesize \textit{Note.} Top, middle and bottom panels depict math, reading and science ability trajectories, respectively.}
+%{\\\footnotesize \textit{Note.}}
 \label{fig:application_sizes}
@@
 The simulations showed that the proposed extensions of GLMM trees are effective for partitioning LGCMs.
-Use of clustered covariances appears most effective and their good performance appears largely unaffected by (mis-)specification of the random effects; it may therefore be the safest choice in practice.
+Use of clustered covariances seem most effective and their good performance was largely unaffected by (mis-)specification of the random effects.
+Use of clustered covariances may therefore be the optimal choice in most practical applications.
 Initializing estimation with the random effects was also effective, but only when the random-effects specification is kept simple (i.e., no estimation of random slopes).
@@
 In comparison to cluster-level covariances, random-effects initialization resulted in larger trees for all outcomes and worse performance for the reading and math outcomes.
+GLMM trees provided a substantial boost in predictive performance compared to global LGCMs, but were outperformed by LGCMs comprising main effects and interactions with time of all baseline covariates.
+Yet, GLMM trees employing clustered covariances selected at most $\frac{1}{3}$ of baseline covariates for distinguishing subgroups, thus providing simpler and easier to interpret results.
+This is in line with earlier research showing that trees may not provide state-of-the-art accuracy but do provide a very good trade-off between accuracy and complexity \cite{Hand06,FokkyIlie22}.
+Especially if there are a large number of possibly relevant covariates, or categorical covariates with many levels, GLMM trees may provide an interpretable alternative to LGCMs comprising all covariates.
 
@@
 Yet, a possible downside of the local-global estimation approach is that it does not allow for recovering subgroups with differences in random-effects parameters.
-When there is a specific interest in partitioning the random effects, score-based SEM trees should likely be preferred.
+When there is a specific interest in subgroups with different random-effects parameters, score-based SEM trees should likely be preferred.
 Alternatively, researchers may want to use the parameter stability tests for mixed-effects models developed by \cite{WangyMerk18} \citep[see also ][]{WangyMerk21,WangyGrav22}.
-This will be useful, for example, when the number of or distances between timepoints differ between respondents so SEM-based growth curve models cannot be applied \citep{NeisyMatt18}.
+These will be useful, for example, when the number of or distances between timepoints differ between respondents so SEM-based growth curve models cannot be applied \citep{NeisyMatt18}.
 
@@
 Especially when both random intercept and slope variances are fixed to 0, higher levels of $\sigma_{b_0}$ and $p_{noise}$ yield more splits with both SEM tree approached.
-LRT-based SEM trees seem unaffected by levels of $\rho$, while score-based SEM trees seemed to implement a larger number of splits with increasing values of $\rho$.
-This pattern seemed reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects were correctly specified.
+LRT-based SEM trees seem unaffected by levels of $\rho$, while score-based SEM trees seem to implement a larger number of splits with increasing values of $\rho$.
+This pattern seems reversed for increased magnitude of $\sigma_{b_1}$, which yields a lower number of splits for both SEM tree approaches, but only when the random effects are correctly specified.
 LongCART implemented more splits with higher levels of $\rho$,  $\sigma_{b_1}$, $N$, $p_{noise}$ and $\sigma_{b_0}$.
@@
 \begin{figure}[!ht]
-\caption{Observed trajectories before and after transforming time for reading ($N = \Sexpr{nrow(readdata)/5}$), math ($N = \Sexpr{nrow(mathdata)/5}$) and science ($N = \Sexpr{nrow(sciedata)/3}$) abilities in the Application.}
+\caption{Observed trajectories before (left column) and after (right column) transforming time for reading ($N = 6,277$), math ($N = 6,512$) and science ($N = 6,625$) abilities in the Application.}
+
 \begin{subfigure}{1.2\textwidth}%
@@
 \end{document}
-
-<<echo=FALSE, eval=FALSE>>=
-###############################
-##
-## Check estimation problems
-##  for longRPart
-##
-
-load("notes, N = 250, Math.Rda")
-math_notes <- cbind(notes, N = 250)
-load("notes, N = 1000, Math.Rda")
-math_notes <- rbind(math_notes, cbind(notes, N = 1000))
-sapply(math_notes, function(x) table(is.na(x)))
-## 8 Warnings for LM(M)tree and longRPart: level of race removed from test data
-## 76 Warnings for longRPart_s: 
-sapply(math_notes, function(x) table(x, useNA = "ifany"))
-## 4 warnings for race omitted from test data
-## 76-4=72 warnings for 1, 2 or 3 nodes estimation not converged
-
-load("notes, N = 250, Reading.Rda")
-read_notes <- cbind(notes, N = 250)
-load("notes, N = 1000, Reading.Rda")
-read_notes <- rbind(read_notes, cbind(notes, N = 1000))
-sapply(read_notes, function(x) table(is.na(x)))
-## 18 Warnings for LM(M)tree and longRPart: level of race removed from test data
-## 80 Warnings for longRPart_s: 
-sapply(read_notes, function(x) table(x, useNA = "ifany"))
-## 11 warnings for race omitted from test data
-## 80-11=69 warnings for 1 or 2 nodes estimation not converged
-
-load("notes, N = 250, Science.Rda")
-scie_notes <- cbind(notes, N = 250)
-load("notes, N = 1000, Science.Rda")
-scie_notes <- rbind(scie_notes, cbind(notes, N = 1000))
-sapply(scie_notes, function(x) table(is.na(x)))
-## 18 Warnings for LM(M)tree and longRPart: level of race removed from test data
-## 82 Warnings for longRPart_s: 
-sapply(scie_notes, function(x) table(x, useNA = "ifany"))
-## 12 warnings for race omitted from test data
-## 82-12=70 warnings for 1, 2 or 3 nodes estimation not converged
-
-## How does this correspond to MSEs?
-sapply(math_MSEs, function(x) table(is.na(x), math_MSEs$N))
-sapply(read_MSEs, function(x) table(is.na(x), read_MSEs$N))
-sapply(scie_MSEs, function(x) table(is.na(x), scie_MSEs$N))
-## Oops, we got MSE anyway
-
-## Identify a tree where longRPart did not converge in a node
-tapply(math_sizes$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
-       table, useNA = "ifany")
-tapply(math_sizes$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
-       mean, na.rm = TRUE)
-## Non-converged nodes are in larger trees, this makes sense (smaller nodes)
-tapply(math_MSEs$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
-       table, useNA = "ifany")
-tapply(math_MSEs$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
-       mean, na.rm = TRUE)
-tapply(math_MSEs$longRPart_s, factor(is.na(math_notes$longRPart_s)), 
-       sd, na.rm = TRUE)
-
-## [6] "NANode 0.221767719908376did not converge. "      
-math_sizes[6,]
-math_MSEs[6,]
-
-## Explanation: If lme model did not converge, no predictions were
-##   generated and written to preds. Thus, predictions from last lme
-##   were used for those nodes, which is always the lme with random 
-## intercepts only
-
-## Illustrate by refitting trees:
-load("bag_ids, N = 250, Math.Rda")
-library("longRPart")
-
-## Prepare formulas and data
-randomFormula <-  ~ 1 | CHILDID
-randomFormula_s <- ~ (1 + months) | CHILDID
-fixedFormula <- score ~ months
-rPartFormula <- ~ GENDER + RACE + WKSESL + C1GMOTOR + C1FMOTOR + 
-  T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE
-resp <- "Math"
-load(paste0(resp, " ability data.Rdata"))
-resp_short <- ifelse(resp == "Math", "math", ifelse(resp == "Reading", "read", "scie"))
-data <- get(paste0(resp_short, "data"))
-data$CHILDID <- factor(data$CHILDID)
-data$GENDER <- factor(data$GENDER)
-data$asmtmm <- factor(data$asmtmm)
-
-## Set up train and test data  
-train <- data[data$CHILDID %in% bag_ids[ , 6], ]
-test <- data[!data$CHILDID %in% bag_ids[ , 6], ]
-## Remove obs with levels of race in test are also in train
-levs <- unique(test$RACE) %in% unique(train$RACE)
-if (!all(levs)) {
-  test <- test[-which(!test$RACE %in% unique(test$RACE)[levs]), ]
-  print(paste0(notes[i, method], "Levels or race omitted from test data: ", 
-               unique(test$race)[!levs]))
-}
-
-## Fit longRPart
-lrpdata <- train[ , c("CHILDID", "months", "score", "GENDER", "RACE",
-                      "WKSESL", "C1GMOTOR", "C1FMOTOR", "T1INTERN",
-                      "T1EXTERN", "T1INTERP", "T1CONTRO", "P1FIRKDG",
-                      "AGEBASELINE")]
-lrpdata$CHILDID <- as.numeric(lrpdata$CHILDID)
-
-for (form in 1:2) {
-  rForm <- list(randomFormula, randomFormula_s)[[form]]
-  method <- c("longRPart", "longRPart_s")[form]
-  system.time(
-    tree <- try(longRPart(randomFormula = rForm, lmeFormula = fixedFormula,
-                    rPartFormula = rPartFormula, data = lrpdata, 
-                    control = rpart.control(maxdepth = 5))))["elapsed"]
-
-  ## Check size
-  (nrow(tree$frame)-1)/2
-  if (form == 1L) math_sizes$longRPart[6] else if (form == 2L) 
-    math_sizes$longRPart_s[6]
-
-  if (class(tree) != "try-error") {
-
-    ## Get predictions
-    train_nodes <- factor(predict(tree))
-    test_nodes <- factor(predict(tree, newdata = test),
-                         levels = levels(train_nodes))
-    preds <- rep(NA, times = nrow(test))
-    ## Fit lme to training observations in each node and predict test observations
-    for (j in unique(train_nodes)) {
-        lme_mod <- try(lme(random = tree$randomFormula,
-                           fixed = tree$lmeFormula,
-                           data = tree$data[train_nodes == j, ]))
-        if (class(lme_mod) != "try-error") {
-          preds[test_nodes == j] <- predict(lme_mod, 
-                                            newdata = test[test_nodes == j, ], 
-                                            level = 0)
-        } else {
-          print(paste0("Node ", j, "did not converge. "))
-        }
-    }
-    ## Check MSE:
-    mean((test$score - preds)^2)  
-    if (form == 1L) math_MSEs$longRPart[6] else if (form == 2L) math_MSEs$longRPart_s[6]
-  }
-}
-## MSEs and tree sizes match up
-
-## How many observations are affected?
-table(is.na(math_sizes$longRPart), math_sizes$N)
-table(is.na(math_sizes$longRPart_s), math_sizes$N)
-table(is.na(read_sizes$longRPart), read_sizes$N)
-table(is.na(read_sizes$longRPart_s), read_sizes$N)
-table(is.na(scie_sizes$longRPart), scie_sizes$N)
-table(is.na(scie_sizes$longRPart_s), scie_sizes$N)
-
-table(is.na(math_sizes$longRPart_s), is.na(math_MSEs$longRPart_s))
-table(is.na(read_sizes$longRPart_s), is.na(read_MSEs$longRPart_s))
-table(is.na(scie_sizes$longRPart_s), is.na(scie_MSEs$longRPart_s))
-
-## How strong is the influence?
-
-## Let's say on average, 2 nodes are affected (mean of 1:3)
-
-## Estimate total # of level-2 obs affected divide by total #:
-math_affected <- sum(2 * math_sizes$N[!is.na(math_notes$longRPart_s)] / 
-  (math_sizes$longRPart_s[!is.na(math_notes$longRPart_s)]+1))
-math_affected / sum(math_sizes$N)
-## [1] 0.06443202
-
-read_affected <- sum(2 * read_sizes$N[!is.na(read_notes$longRPart_s)] / 
-  (read_sizes$longRPart_s[!is.na(read_notes$longRPart_s)]+1))
-read_affected / sum(read_sizes$N)
-## [1] 0.111979
-
-scie_affected <- sum(2 * scie_sizes$N[!is.na(scie_notes$longRPart_s)] / 
-  (scie_sizes$longRPart_s[!is.na(scie_notes$longRPart_s)]+1), na.rm = TRUE)
-scie_affected / sum(scie_sizes$N)
-## [1] 0.2531524
-
-## Ideally, would refit all longRPart_s trees which had convergence errors
-## in one of the nodes. And then use NA for prediction, or refit a model
-## without random slopes. This might paint a slightly difference picture.
-## However, I do not expect dramatic differences, given that results
-## for longRPart and longRPart_s are quite similar, and seem to differ
-## most for science, which was most strongly affected by the mistake.
-@
-
-%\begin{figure}[h]
-%\caption{Computation times for partitioning math, reading and science ability trajectories.}
-%\begin{subfigure}{.5\textwidth}
-<<eval=FALSE, echo=FALSE, fig=TRUE, height=3.25, width=3>>=
-load("times, N = 250, Reading.Rda")
-read_times <- cbind(times, N = 250)
-load("times, N = 1000, Reading.Rda")
-read_times <- rbind(read_times, cbind(times, N = 1000))
-
-load("times, N = 250, Math.Rda")
-math_times <- cbind(times, N = 250)
-load("times, N = 1000, Math.Rda")
-math_times <- rbind(math_times, cbind(times, N = 1000))
-
-load("times, N = 250, Science.Rda")
-scie_times <- cbind(times, N = 250)
-load("times, N = 1000, Science.Rda")
-scie_times <- rbind(scie_times, cbind(times, N = 1000))
-
-## Combine all LMM trees
-ECLSK_times <- rbind(stack(math_times[ , -c(6, 10)]),
-                     stack(read_times[ , -c(6, 10)]),
-                     stack(scie_times[ , -c(6, 10)]))
-ECLSK_times$ind <- as.character(ECLSK_times$ind)
-ECLSK_times$ind[grepl("LMMtree", ECLSK_times$ind)]<- "LMM tree"
-ECLSK_times$ind[grepl("longRPart", ECLSK_times$ind)]<- "longRPart"
-ECLSK_times$ind[ECLSK_times$ind == "LMtree_c"] <- "LM tree"
-ECLSK_times$ind <- factor(ECLSK_times$ind)
-
-## Compute summary stats
-mean_times <- with(ECLSK_times, tapply(values, ind, mean, na.rm=TRUE))
-#summ_stats <- data.frame(M = sprintf(mean_times, fmt = '%#.2f'),
-#                         ind = names(mean_times), y = 11000)
-## Plot
-# ggplot(ECLSK_times) +
-#   geom_violin(aes(x=ind, y=values), alpha = .5, fill = "gray", width=.7) +
-#   scale_y_continuous(trans = "log", limits = c(0.1, 12000),
-#                      breaks = c(.01, .1, 1, 10, 100, 1000)) +
-#   labs(x = "", y = "Computation time (in seconds)") +
-#   geom_text(data = summ_stats, aes(x=ind, y=y, label = M),
-#             size = 3, col = "gray44") +
-#   labs(tag = "M") + theme(plot.tag.position = c(.12, .95),
-#                           plot.tag = element_text(size = 10))
-@
-%\end{subfigure}
-%\label{fig:application_times}
-%\end{figure}
 

### 2024-02-08 pkg/glmertree/paper/lgcm-tree/BRM/Revision 1/_5 Analysis of ECLSK data.R

@@ -1,0 +1,162 @@
+## Load libraries & save session info
+library("glmertree")
+library("partykit")
+sInfo <- sessionInfo()
+save(sInfo, file = "sessionInfo.Rda")
+list.rules <- pre:::list.rules
+
+
+
+#################################################
+##
+## Cross-validate performance of LM(M) trees
+##
+
+## Set up design
+methods <- c("LMtree_c", "LMMtree", "LMMtree_c", "LMMtree_r", "LMMtree_sc", 
+             "LMM", "LMM_s", "LMM2", "LMM_s2", "LMM3", "LMM_s3")
+nreps <- 100L
+part_vars <- c("months", "GENDER", "RACE", "WKSESL", "C1GMOTOR", "C1FMOTOR", 
+               "T1INTERN", "T1EXTERN", "T1INTERP", "T1CONTRO", "P1FIRKDG", "AGEBASELINE")
+
+## Prepare formulas
+LMt_form <- score ~ months | GENDER + RACE + WKSESL + C1GMOTOR + 
+  C1FMOTOR + T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE
+
+LMM_form <- score ~ (1|CHILDID) + months
+LMM_s_form <- score ~ (1 + months|CHILDID) + months
+
+LMM_form2 <- score ~ (1|CHILDID) + months + GENDER + RACE + WKSESL + C1GMOTOR + 
+  C1FMOTOR + T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE
+LMM_s_form2 <- score ~ (1 + months|CHILDID) + months + GENDER + RACE + WKSESL + 
+  C1GMOTOR + C1FMOTOR + T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE
+
+LMM_form3 <- score ~ (1|CHILDID) + months*(GENDER + RACE + WKSESL + C1GMOTOR + 
+  C1FMOTOR + T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE)
+LMM_s_form3 <- score ~ (1 + months|CHILDID) + months*(GENDER + RACE + WKSESL + 
[…]
